### 2007-03-18 doc/logparser.tex
A paper describing logparser.  Despite hours of work on it I haven't made all that much progress.

\documentclass[a4paper,12pt,draft]{article}

% Better typesetting of URLs.
\usepackage{url}
\usepackage[final]{graphicx}

\begin{document}

\title{Parsing Postfix log files}
\author{John Tobin \\ Computer Science Department \\ Trinity College \\
Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
\date{}
\maketitle

\section{Abstract}

Most mail server administrators will have performed some basic processing
of Postfix logs at one time or another, whether it was to figure out why
something isn't working, why mail is being rejected, or to check whether
their new anti-spam measures are working.  The more adventurous will have
generated some statistics to show how many hits each of their anti-spam
measures have gotten in the last week, and possibly even generated some
graphs to clearly illustrate the point to management or complaining
users\footnote{This was the author's first real foray into processing
Postfix logs}.  Very few will have performed in-depth parsing and analysis
of their logs, where the parsing needs to be per-connection or per-queueid,
rather than per-line.  One of the barriers to this type of processing is
the unstructured nature of Postfix's logs: each logging line was added on
an ad hoc basis as Wietse found a requirement or added new functionality.
This paper attempts to document the difficult process of parsing Postfix
logs, and presents a program which parses logs and places the resulting
data into a database for further processing.

\section{Background}

Postfix is a highly configurable, high performance, secure and scalable
Mail Transport Agent (MTA),  It features extensive anti-spam restrictions,
allowing a mail administrator to deploy those restrictions which they judge
suitable for their needs, rather than a fixed set chosen by Postfix's
author.  These restrictions can be applied and combined on a per-host,
per-recipient or per-sender basis, allowing clients/recipients to be
whitelisted with respect to specific restrictions.  Postfix leverages
simple lookup tables to support arbitrarily complicated user-defined
sequences of checks and exceptions, with the ultimate in flexibility being
the facility to consult an external process which can implement whatever
logic is required; if you wish you can restrict some users to sending mail
on the third Tuesday after pay day only, if that is of any use to you.
Administrators can also supply their own rejection messages, to make it
clear to senders why exactly their mail was rejected.  Unfortunately this
flexibility has a cost: complexity in the logs generated.  While it is easy
to use \texttt{grep(1)} to determine the fate of an individual email, and
even to extend that into generating some useful graphs, following the
progress of an email through Postfix can be quite difficult.  The logs tend
to follow a 90\%/10\% pattern: 90\% of the time the logging is simple, and
the other 10\% of the logging requires 90\% of the code.

This paper and the program it describes are part of a larger project to
optimise your Postfix restrictions, generate statistics and graphs, and
provide a platform on which new restrictions can be trialled and evaluated
to see if they are worth using in the fight against spam.  The program
parses Postfix logs and populates a database with the data gleamed from
those logs, providing a consistent and simple view of your logs which
future tools can utilise.

\section{Parsing details}

The complexity and variation in Postfix's logs requires similar flexibility
in the parser; decoupling the parsing rules from the associated actions
allows the rules to be loaded from an external source without modifications
to the parser source code (significantly lowering the barrier to entry for
new users), and simplifies both parser and rules.  The separation also
allows for a XXX REWRITE separation of functionality: rules handle low
level details of identifying a line, whereas the parser handles the higher
level details of following the path a mail takes through postfix,
assembling the required information, etc.

\subsection{Rules}

Rules are relatively simple, defining the following for each rule:

\begin{description}

    \item [Name] A short name for the rule.

    \item [Description] A longer, more detailed description of the rule.

    \item [Program] The program (smtp, smtpd, qmgr, etc) which the rule
        applies to.  This avoids needlessly trying rules which won't
        match, or worse, might match unintentionally.

    \item [Regex] The regex to compare the log line against.  The regex
        will first have several keywords expanded; this simplifies writing
        rules and avoids needless repetition of complex regexs.

    \item [Result\_cols and connection\_cols] This is how we extract the
        fields in the log line matched by the regex.  The format is:
        \newline 
        hostname = 1; helo = 2; sender = 4; \newline
        i.e. semi-colon separated assignment statements, with the column
        name on the left and the match from the regex (\$1, \$2 etc) on the
        right hand side (without \$).

    \item [Action] The action for the parser to take.

    \item [Queueid] The index of the field in regex giving the queueid.

    \item [Rule\_order] This is used to sort the rules, so that rules
        matching more commonly occurring log lines will be tried first, and
        is updated every time the program is run.

    \item [Priority] This is the user-configurable companion to
        rule\_order: rules with a higher priority will be tried first,
        overriding rule\_order, allowing more specific rules to take
        precedence over more general rules.

    \item [Result]  The result the log line represents: rejection,
        acceptance, info, etc.

    \item [Result\_data and connection\_data] Sometimes rules need to fake
        a piece of data which isn't present in the log line: the most
        common would be smtp\_code when mail is accepted.  The format is
        the same as result\_cols and connection\_cols.

\end{description}

\subsection{Algorithm}

Lets start with a flow chart describing the program:

XXX FLOW CHART GOES HERE

%\includegraphics{logparser-flow-chart.ps}

A high level view of the algorithm could be expressed as:

\begin{enumerate}

    \item Mail enters the system via SMTP or local submission.
    \item If the mail is rejected, log all data, and finish with the mail.
    \item Follow the progress of the mail until it's either delivered or
        bounced, log all data, and finish with the mail.

\end{enumerate}

Unfortunately it's not that easy.

\subsubsection{Complications}

\begin{enumerate}

    \item The mail lacks a queueid until the mail has been accepted, so log
        lines must first be correlated by the smtpd pid, then transition to
        being correlated by the queueid.  This is relatively minor, but
        does require two versions of several functions,
        \texttt{per\_pid} and \texttt{per\_queueid}.

    \item Multiple independent mails may be delivered during one
        connection; this requires cloning the current data, so that
        subsequent mails won't trample over each other.  This must be done
        every time a mail is accepted, as it's impossible to tell in
        advance which connections will accept multiple mails.  Happily once
        the mail has been accepted log entries won't be correlated by pid
        for that mail any more, so there isn't any ambiguity about which
        mail a given log line belongs to.  One possible (as yet unseen)
        difficulty is distinguishing between rejects which belong to the
        current mail, and rejects which belong to previous, unaccepted,
        mails.

    \item The most difficult complication is that mails are not always
        delivered directly to a mailbox: sometimes they are aliased and
        need to be delivered to an address (or addresses) on another
        server.  When this occurs a child mail will be injected into the
        postfix queue, but without the explicit logging smtpd or sendmail
        injected mails have, so the source is not immediately discernible;
        from a strictly linear reading of the logs it \textit{usually}
        appears as if the child mail has appeared from thin air.
        Subsequently the parent mail will log the creation of the child
        mail:

        \texttt{3FF7C4317: to=<username@cs.tcd.ie>, relay=local, \newline delay=0,
        status=sent (forwarded as 56F5B43FD)}

        Unfortunately while all log lines from an individual process appear
        in the logs in monotonic order, the order in which log lines from
        different processes are interleaved is subject to the vagaries of
        process scheduling.  In addition the first log line belonging to
        the child mail (the log line cited above properly belongs to the
        parent mail) is logged by qmgr, so depending on how busy qmgr (and
        the server generally) is, the parent's creation log line may appear
        before or after (generally after) the child's log line from qmgr.
        This has the effect that the parser cannot complain when it
        encounters a log line from qmgr for a previously unseen mail; it
        must mark the mail as potentially fake and subsequently clear the
        faked flag if and when the origin of the mail becomes clear.
        Obviously the parser could omit checking of where mails originate
        from, but I feel that it is better to require an explicit source,
        as bugs are more likely to be exposed.

        Process scheduling can have a still more confusing effect: quite
        often the child mail will be created, delivered and removed
        \textbf{before} the parent logs the creation line!  Thus mails
        marked as faked cannot be committed, instead they must be marked as
        commit ready and subsequently committed by the parent mail.  Quite
        apart from identifying mail injected in an unknown fashion, or bugs
        in the parser, faked mails need to be marked as such because they
        lack some data present in the parent mail, and must copy that data
        from their parent.

\end{enumerate}

XXX TODO what happens to the original connection?? XXX

XXX TODO is it possible to distinguish between different rejected mails on
one connection??  XXX

\appendix

\end{document}

### 2007-03-18 doc/logparser.tex
Wrote an outline/structure for the paper, and moved bits around to fit.  I now have an idea of how much progress I've made, yay!

@@
+% $Id$
 \documentclass[a4paper,12pt,draft]{article}
@@
 \usepackage{url}
+% Include images
 \usepackage[final]{graphicx}
@@
 Most mail server administrators will have performed some basic processing
-of Postfix logs at one time or another, whether it was to figure out why
-something isn't working, why mail is being rejected, or to check whether
-their new anti-spam measures are working.  The more adventurous will have
-generated some statistics to show how many hits each of their anti-spam
-measures have gotten in the last week, and possibly even generated some
-graphs to clearly illustrate the point to management or complaining
-users\footnote{This was the author's first real foray into processing
-Postfix logs}.  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing needs to be per-connection or per-queueid,
-rather than per-line.  One of the barriers to this type of processing is
-the unstructured nature of Postfix's logs: each logging line was added on
-an ad hoc basis as Wietse found a requirement or added new functionality.
-This paper attempts to document the difficult process of parsing Postfix
-logs, and presents a program which parses logs and places the resulting
-data into a database for further processing.
+of Postfix logs at one time or another, whether it was to debug out a
+problem, explain why mail is being rejected, or to check whether their new
+anti-spam measures are working.  The more adventurous will have generated
+some statistics to show how many hits each of their anti-spam measures have
+gotten in the last week, and possibly even generated some graphs to clearly
+illustrate the point to management or complaining users\footnote{This was
+the author's first real foray into processing Postfix logs}.  Very few will
+have performed in-depth parsing and analysis of their logs, where the
+parsing must correlate the log lines per-connection or per-queueid, rather
+than processing lines independently.  One of the barriers to this type of
+processing is the unstructured nature of Postfix's logs, where each logging
+line was added on an ad hoc basis as Wietse found a requirement or added
+new functionality, and further complication comes from user defined log
+messages.  This paper attempts to document the difficult process of parsing
+Postfix logs, and presents a program which parses logs and places the
+resulting data into a database for further processing.
 
@@
 
+\subsection{Postfix background}
+
 Postfix is a highly configurable, high performance, secure and scalable
-Mail Transport Agent (MTA),  It features extensive anti-spam restrictions,
+Mail Transport Agent.  It features extensive anti-spam restrictions,
 allowing a mail administrator to deploy those restrictions which they judge
@@
 simple lookup tables to support arbitrarily complicated user-defined
-sequences of checks and exceptions, with the ultimate in flexibility being
-the facility to consult an external process which can implement whatever
-logic is required; if you wish you can restrict some users to sending mail
-on the third Tuesday after pay day only, if that is of any use to you.
-Administrators can also supply their own rejection messages, to make it
-clear to senders why exactly their mail was rejected.  Unfortunately this
-flexibility has a cost: complexity in the logs generated.  While it is easy
-to use \texttt{grep(1)} to determine the fate of an individual email, and
-even to extend that into generating some useful graphs, following the
-progress of an email through Postfix can be quite difficult.  The logs tend
-to follow a 90\%/10\% pattern: 90\% of the time the logging is simple, and
-the other 10\% of the logging requires 90\% of the code.
+sequences of restrictions and exceptions, with the ultimate in flexibility
+being the facility to consult an external process which can implement
+whatever logic is required: if you wish you can restrict some users to
+sending mail on the third Tuesday after pay day only, if that is of any use
+to you.  Administrators can also supply their own rejection messages, to
+make it clear to senders why exactly their mail was rejected.
+Unfortunately this flexibility has a cost: complexity in the logs
+generated.  While it is easy to use \texttt{grep(1)} to determine the fate
+of an individual email, following the progress of an email through Postfix
+can be quite difficult.  The logs tend to follow a 90\%/10\% pattern: 90\%
+of the time following the log entries is simple, but the other 10\% of the
+logging requires 90\% of the code.
+
+\subsection{Paper background}
 
@@
 those logs, providing a consistent and simple view of your logs which
-future tools can utilise.
+future tools can utilise.  The program should be reasonably easily
+extensible, so if logging to a database doesn't meet your requirements, you
+can implement your own processing.
 
-\section{Parsing details}
+\section{Adaptable parsing: user defined rules}
 
@@
 to the parser source code (significantly lowering the barrier to entry for
-new users), and simplifies both parser and rules.  The separation also
-allows for a XXX REWRITE separation of functionality: rules handle low
-level details of identifying a line, whereas the parser handles the higher
-level details of following the path a mail takes through postfix,
-assembling the required information, etc.
-
-\subsection{Rules}
+new users), and greatly simplifies both parser and rules.  The separation
+also allows for a clear separation of functionality: rules handle low level
+details of identifying a line, whereas the parser handles the higher level
+details of following the path a mail takes through postfix, assembling the
+required information, etc.
 
@@
 
-    \item [Name] A short name for the rule.
+    \item [name] A short name for the rule.
 
-    \item [Description] A longer, more detailed description of the rule.
+    \item [description] A longer, more detailed description of the rule.
 
-    \item [Program] The program (smtp, smtpd, qmgr, etc) which the rule
+    \item [program] The program (smtp, smtpd, qmgr, etc) which the rule
         applies to.  This avoids needlessly trying rules which won't
@@
 
-    \item [Regex] The regex to compare the log line against.  The regex
+    \item [regex] The regex to compare the log line against.  The regex
         will first have several keywords expanded; this simplifies writing
-        rules and avoids needless repetition of complex regexs.
+        rules and avoids needless repetition of complex regex components.
 
-    \item [Result\_cols and connection\_cols] This is how we extract the
+    \item [result\_cols and connection\_cols] This is how we extract the
         fields in the log line matched by the regex.  The format is:
@@
 
-    \item [Action] The action for the parser to take.
+    \item [action] The action for the parser to take.
 
-    \item [Queueid] The index of the field in regex giving the queueid.
+    \item [queueid] The index of the field in regex giving the queueid.
 
-    \item [Rule\_order] This is used to sort the rules, so that rules
+    \item [rule\_order] This is used to sort the rules, so that rules
         matching more commonly occurring log lines will be tried first, and
@@
 
-    \item [Priority] This is the user-configurable companion to
+    \item [priority] This is the user-configurable companion to
         rule\_order: rules with a higher priority will be tried first,
@@
 
-    \item [Result]  The result the log line represents: rejection,
+    \item [result]  The result the log line represents: rejection,
         acceptance, info, etc.
 
-    \item [Result\_data and connection\_data] Sometimes rules need to fake
+    \item [result\_data and connection\_data] Sometimes rules need to fake
         a piece of data which isn't present in the log line: the most
@@
 
-\subsection{Algorithm}
+\section{Parsing algorithm}
 
-Lets start with a flow chart describing the program:
+The most substantive part of the parser is the parser al . . . . blah
 
-XXX FLOW CHART GOES HERE
-
-%\includegraphics{logparser-flow-chart.ps}
+\subsection{Naive approach}
 
@@
     \item Mail enters the system via SMTP or local submission.
+
     \item If the mail is rejected, log all data, and finish with the mail.
+
     \item Follow the progress of the mail until it's either delivered or
@@
 
-\subsubsection{Complications}
+\subsection{Complications encountered}
 
@@
 
-        \texttt{3FF7C4317: to=<username@cs.tcd.ie>, relay=local, \newline delay=0,
-        status=sent (forwarded as 56F5B43FD)}
+        \texttt{3FF7C4317: to=<username@cs.tcd.ie>, relay=local, \newline 
+        delay=0, status=sent (forwarded as 56F5B43FD)}
 
@@
 
+\subsection{Full algorithm}
+
+Lets start with a flow chart describing the program:
+
+\includegraphics{logparser-flow-chart.ps}
+
+\section{Limitations and improvements}
+
+\section{Summary}
+
 XXX TODO what happens to the original connection?? XXX

### 2007-03-21 doc/logparser.tex
Most of the structural changes come from today's meeting with Carl. Added an abstract and table of contents; previous abstract became introduction. Started writing a parser background section, suggested by Carl. Added limitations and possible improvements section (partially written), and empty conclusion section. Lots of tweaks and minor improvements, re-wording, etc.

@@
 \title{Parsing Postfix log files}
-\author{John Tobin \\ Computer Science Department \\ Trinity College \\
-Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
+\author{John Tobin \\ School of Computer Science and Statistics \\ 
+Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
 \date{}
@@
 
-\section{Abstract}
+\begin{abstract}
+
+    Parsing Postfix logs is much more difficult than it first appears, but
+    it is possible.  This paper describes the process required, documenting
+    the algorithm and difficulties encountered, with reference to an
+    implementation which stores data gathered from the logs in an SQL
+    database.
+
+\end{abstract}
+
+XXX WHAT IS THE CONCLUSION??? XXX
+
+\newpage
+\tableofcontents
+
+\section{Introduction}
 
 Most mail server administrators will have performed some basic processing
-of Postfix logs at one time or another, whether it was to debug out a
-problem, explain why mail is being rejected, or to check whether their new
+of Postfix logs at one time or another, whether it was to debug a problem,
+explain to a user why mail is being rejected, or to check whether their new
 anti-spam measures are working.  The more adventurous will have generated
@@
 gotten in the last week, and possibly even generated some graphs to clearly
-illustrate the point to management or complaining users\footnote{This was
-the author's first real foray into processing Postfix logs}.  Very few will
+illustrate the point to management or complaining users.\footnote{This was
+the author's first real foray into processing Postfix logs.}  Very few will
 have performed in-depth parsing and analysis of their logs, where the
@@
 processing is the unstructured nature of Postfix's logs, where each logging
-line was added on an ad hoc basis as Wietse found a requirement or added
-new functionality, and further complication comes from user defined log
-messages.  This paper attempts to document the difficult process of parsing
-Postfix logs, and presents a program which parses logs and places the
-resulting data into a database for further processing.
+line was added on an ad hoc basis as a requirement was discovered or new
+functionality was added.  Further complication arises the ability to have
+rejection messages defined by the administrator, and every Real time Black
+List returns a different message.  This paper documents the difficult
+process of parsing Postfix logs, and presents a program which parses logs
+and places the resulting data into a database for further processing by
+other applications.
 
@@
 of the time following the log entries is simple, but the other 10\% of the
-logging requires 90\% of the code.
+logging requires 90\% of the code.\footnote{These numbers don't have a
+solid scientific basis, they're based on gut feeling from writing the
+software.}
 
@@
 future tools can utilise.  The program should be reasonably easily
-extensible, so if logging to a database doesn't meet your requirements, you
-can implement your own processing.
+extensible, so if logging to a database doesn't meet requirements, one
+can implement one's own processing.
+
+\subsection{Parser background}
+
+Before getting into detail about the parser, a brief overview is in order.
+The parser is split into two parts: the parsing algorithm, and the rules
+which are applied to the lines.
+
+\begin{description}
+
+    \item [Rules] The first matching rule wins: no further rules are tried
+        against that line,\footnote{With one exception, detailed later.}
+        but there is a facility for the specifying the order of the rules,
+        so that more specific rules can be matched first.  Rules are
+        completely self-contained: there are no sub-rules, so the
+        complexity is linear.  Rules are similar to the tokeniser/Lex part
+        of a Lex/YACC parser.
+
+    \item [Algorithm]
+
+\end{description}
 
@@
 allows the rules to be loaded from an external source without modifications
-to the parser source code (significantly lowering the barrier to entry for
-new users), and greatly simplifies both parser and rules.  The separation
-also allows for a clear separation of functionality: rules handle low level
-details of identifying a line, whereas the parser handles the higher level
-details of following the path a mail takes through postfix, assembling the
-required information, etc.
+to the algorithm source code (significantly lowering the barrier to entry
+for new users), and greatly simplifies both algorithm and rules.  The
+separation also allows for a clear separation of functionality: rules
+handle low level details of identifying a line, whereas the algorithm
+handles the higher level details of following the path a mail takes through
+postfix, assembling the required information, etc.
 
@@
 
-The most substantive part of the parser is the parser al . . . . blah
+While the rules are more than double (treble with 2.3.x rules?) the size of
+the parser, the rules are quite simple, and each rule is very much
+independent of its fellows.  The parser is significantly more complicated,
+and highly interdependent.
 
@@
 
-    \item If the mail is rejected, log all data, and finish with the mail.
+    \item If the mail is rejected, log all data and finish.
 
     \item Follow the progress of the mail until it's either delivered or
-        bounced, log all data, and finish with the mail.
+        bounced, log all data, and finish.
 
@@
 
-    \item The mail lacks a queueid until the mail has been accepted, so log
-        lines must first be correlated by the smtpd pid, then transition to
-        being correlated by the queueid.  This is relatively minor, but
-        does require two versions of several functions,
-        \texttt{per\_pid} and \texttt{per\_queueid}.
+    \item The mail lacks a queueid until it has been accepted, so log lines
+        must first be correlated by the smtpd pid, then transition to being
+        correlated by the queueid.  This is relatively minor, but does
+        require two versions of several functions, \texttt{by\_pid} and
+        \texttt{by\_queueid}.
 
@@
         every time a mail is accepted, as it's impossible to tell in
-        advance which connections will accept multiple mails.  Happily once
-        the mail has been accepted log entries won't be correlated by pid
-        for that mail any more, so there isn't any ambiguity about which
-        mail a given log line belongs to.  One possible (as yet unseen)
-        difficulty is distinguishing between rejects which belong to the
-        current mail, and rejects which belong to previous, unaccepted,
-        mails.
+        advance which connections will accept multiple mails.  It is quite
+        easy to overlook this complication because only a small minority of
+        connections accept more than one mail. Happily once the mail has
+        been accepted log entries won't be correlated by pid for that mail
+        any more, so there isn't any ambiguity about which mail a given log
+        line belongs to.  The original connection will be discarded when
+        the client disconnects.  One possible (as yet unseen) difficulty is
+        distinguishing between rejects which belong to the current mail,
+        and rejects which belong to previous, unaccepted, mails.
 
     \item The most difficult complication is that mails are not always
-        delivered directly to a mailbox: sometimes they are aliased and
+        delivered directly to a mailbox (or program; there is very little
+        difference between the two): sometimes they are aliased and
         need to be delivered to an address (or addresses) on another
@@
 
-        \texttt{3FF7C4317: to=<username@cs.tcd.ie>, relay=local, \newline 
+        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline 
         delay=0, status=sent (forwarded as 56F5B43FD)}
@@
         the server generally) is, the parent's creation log line may appear
-        before or after (generally after) the child's log line from qmgr.
+        before or after (usually after) the child's log line from
+        qmgr.\footnote{Postfix is quite paranoid about mail delivery, an
+        excellent characteristic for an MTA to posess, so it won't log that
+        the child has been created until it is absolutely certain that the
+        mail has been written to disk.  Because the parent then has to be
+        informed that the mail has been committed, it's most likely that
+        the qmgr will log first.}
+
         This has the effect that the parser cannot complain when it
@@
 \includegraphics{logparser-flow-chart.ps}
+\label{flow-chart}
 
-\section{Limitations and improvements}
+\section{Limitations and possible improvements}
 
-\section{Summary}
+Every piece of software suffers from some limitations, and there is almost
+always room for improvement.
 
-XXX TODO what happens to the original connection?? XXX
+\subsection{Limitations}
 
-XXX TODO is it possible to distinguish between different rejected mails on
-one connection??  XXX
+\begin{itemize}
+
+    \item Each new Postfix release requires new rules to be written, to
+        cope with the new messages.
+
+    \item The program should save state at the end of each run, and reload
+        it at the start of the next run, to properly cope with mails which
+        span logs.  This might also require purging old mails once they've
+        been in the state tables for too long.
+
+    \item It appears that the hostname used in the HELO command is not
+        logged if the mail is accepted.\footnote{Tested with Postfix
+        2.2.10, this may possibly have changed in Postfix 2.3, or the
+        upcoming 2.4.}
+
+\end{itemize}
+
+\subsection{Possible improvements}
+\begin{itemize}
+
+    \item A progress bar would be useful when run interactively, as the
+        program takes roughly one minute per 10MB of
+        logs.\footnote{Approximately 80\% of the run time is consumed by
+        logging to the database.}  Obviously performance is entirely
+        dependant on the machine the program is running on.
+
+\end{itemize}
+
+\section{Conclusion}
 

### 2007-03-23 doc/logparser.tex
Improved parser overview. Move rule characteristics to rules section. Split rule attributes into their own subsection, and seriously improved them. Added subsections for the flow chart and a detailed description of actions; the latter is empty.

@@
 to see if they are worth using in the fight against spam.  The program
-parses Postfix logs and populates a database with the data gleamed from
+parses Postfix logs and populates a database with the data gleaned from
 those logs, providing a consistent and simple view of your logs which
 future tools can utilise.  The program should be reasonably easily
-extensible, so if logging to a database doesn't meet requirements, one
-can implement one's own processing.
+extensible, so if logging to a database doesn't meet requirements, one can
+implement one's own processing.
 
@@
 Before getting into detail about the parser, a brief overview is in order.
-The parser is split into two parts: the parsing algorithm, and the rules
-which are applied to the lines.
-
-\begin{description}
-
-    \item [Rules] The first matching rule wins: no further rules are tried
-        against that line,\footnote{With one exception, detailed later.}
-        but there is a facility for the specifying the order of the rules,
-        so that more specific rules can be matched first.  Rules are
-        completely self-contained: there are no sub-rules, so the
-        complexity is linear.  Rules are similar to the tokeniser/Lex part
-        of a Lex/YACC parser.
-
-    \item [Algorithm]
-
-\end{description}
+The parser is split into two parts: the parsing algorithm and the rules
+which are applied to the lines.  Rules can be thought of as the
+tokeniser/Lex part of a Lex/YACC parser; they identify the line and return
+some data, which the algorithm (the YACC part) receives, performing the
+correct action.  Rules are all about identifying a line and extracting data
+from it, whereas the algorithm is all about following the path the mail
+takes through Postfix, piecing the data together into a coherent whole, and
+performing housekeeping duties.
 
@@
 in the parser; decoupling the parsing rules from the associated actions
-allows the rules to be loaded from an external source without modifications
-to the algorithm source code (significantly lowering the barrier to entry
-for new users), and greatly simplifies both algorithm and rules.  The
-separation also allows for a clear separation of functionality: rules
-handle low level details of identifying a line, whereas the algorithm
-handles the higher level details of following the path a mail takes through
-postfix, assembling the required information, etc.
+allows new rules to be written and tested without modifications to the
+algorithm source code (significantly lowering the barrier to entry for new
+or casual users), and greatly simplifies both algorithm and rules.  It also
+allows for a clear separation of functionality: rules handle low level
+details of identifying a line, whereas the algorithm handles the higher
+level details of following the path a mail takes through postfix,
+assembling the required data, etc.
 
-Rules are relatively simple, defining the following for each rule:
+Rule characteristics:
+
+\begin{itemize}
+
+    \item The first matching rule wins: no further rules are tried against
+        that line,\footnote{With one exception, detailed later in Section
+        \ref{actions-in-detail}.} but there is a facility for the specifying
+        the order of the rules, so that more specific rules can be matched
+        first.
+
+    \item Rules are completely self-contained: there are no sub-rules, so
+        rules have linear complexity, and can be understood without
+        reference to any other rules.
+
+\end{itemize}
+
+\subsection{Rule attributes}
+
+Each rule defines the following:
 
@@
         applies to.  This avoids needlessly trying rules which won't
-        match, or worse, might match unintentionally.
+        match the line, or worse, might match unintentionally.
 
     \item [regex] The regex to compare the log line against.  The regex
-        will first have several keywords expanded; this simplifies writing
-        rules and avoids needless repetition of complex regex components.
+        will first have several keywords expanded; this simplifies reading
+        and writing rules, avoids needless repetition of complex regex
+        components, and allows the components to be corrected and/or
+        improved in one location.
 
-    \item [result\_cols and connection\_cols] This is how we extract the
-        fields in the log line matched by the regex.  The format is:
-        \newline 
-        hostname = 1; helo = 2; sender = 4; \newline
-        i.e. semi-colon separated assignment statements, with the column
-        name on the left and the match from the regex (\$1, \$2 etc) on the
+    \item [result\_cols, connection\_cols] This specifies how the fields in
+        the log line will be extracted.  The format is: \newline
+        client\_hostname = 1; helo = 2; sender = 4; \newline
+        i.e.  semi-colon separated assignment statements, with the variable
+        name on the left and the match from the regex (\$1, \$2 etc.) on the
         right hand side (without \$).
 
-    \item [action] The action for the parser to take.
+    \item [result\_data, connection\_data] Sometimes rules need to fake
+        a piece of data which isn't present in the log line: the most
+        common would be smtp\_code when mail is accepted.  The format of
+        the same as result\_cols and connection\_cols.
 
-    \item [queueid] The index of the field in regex giving the queueid.
+    \item [result]  XXX RESULT NEEDS TO BE RENAMED XXX This is the action
+        Postfix would have taken to generate this line.  Info is represents
+        an unspecified intermediate action that the parser is not
+        interested in per se, but which does log useful information.  The
+        extracted information will be used, but the other results will be
+        discarded.
 
-    \item [rule\_order] This is used to sort the rules, so that rules
-        matching more commonly occurring log lines will be tried first, and
-        is updated every time the program is run.
+    \item [action] The action the algorithm will take, e.g.
+        \begin{description}
+
+            \item [ignore] Ignore this line.  The simplest action, it is
+                required because every line from a program of interest to
+                the parser must be parsed properly; any unparsed line is
+                considered an error.
+
+            \item [save\_by\_pid] Use the pid from the log line to lookup a
+                connection, then save the data from the log line in the
+                connection data structure.
+
+            \item [commit] The mail is finished with, enter it in the
+                database.
+
+        \end{description}
+
+        A full list can be found in Section \ref{actions-in-detail}.
+
+    \item [queueid] Specifies the match from the regex which gives the
+        queueid, or zero if the log line doesn't contain a queueid.
+
+    \item [rule\_order] This is an efficiency measure.  A counter is
+        maintained for every rule and incremented each time the rule
+        successfully matches.  At the start of each run the program sorts
+        the rules in ascending rule\_order, and at the end of the run
+        updates every rule's rule\_order so that rules which matched more
+        frequently have a lower rule\_order.  Assuming that the 
+        distribution of log lines is reasonably consistent, rules matching
+        more commonly occurring log lines will be tried before rules
+        matching less commonly occurring log lines, lowering the program's
+        execution time.  XXX GIVE EXAMPLE TIMINGS XXX
 
@@
         precedence over more general rules.
-
-    \item [result]  The result the log line represents: rejection,
-        acceptance, info, etc.
-
-    \item [result\_data and connection\_data] Sometimes rules need to fake
-        a piece of data which isn't present in the log line: the most
-        common would be smtp\_code when mail is accepted.  The format is
-        the same as result\_cols and connection\_cols.
 
@@
 
-\subsection{Full algorithm}
-
-Lets start with a flow chart describing the program:
+\subsection{Algorithm flow chart}
 
@@
 \label{flow-chart}
+
+\subsection{Full algorithm}
+
+\label{full-algorithm}
+
+\subsection{Actions in detail}
+
+\label{actions-in-detail}
 

### 2007-03-25 doc/logparser.tex
Lots of clarifications, improvements, more detail. Added statistics about (non-)compiled regexs and sorted/reversed/shuffled rule_order.

@@
     it is possible.  This paper describes the process required, documenting
-    the algorithm and difficulties encountered, with reference to an
-    implementation which stores data gathered from the logs in an SQL
-    database.
+    the parsing algorithm and rules, and explaining the difficulties
+    encountered, with reference to an implementation which stores data
+    gathered from the logs in an SQL database for later processing.
 
@@
 of Postfix logs at one time or another, whether it was to debug a problem,
-explain to a user why mail is being rejected, or to check whether their new
-anti-spam measures are working.  The more adventurous will have generated
-some statistics to show how many hits each of their anti-spam measures have
-gotten in the last week, and possibly even generated some graphs to clearly
-illustrate the point to management or complaining users.\footnote{This was
-the author's first real foray into processing Postfix logs.}  Very few will
-have performed in-depth parsing and analysis of their logs, where the
-parsing must correlate the log lines per-connection or per-queueid, rather
-than processing lines independently.  One of the barriers to this type of
-processing is the unstructured nature of Postfix's logs, where each logging
-line was added on an ad hoc basis as a requirement was discovered or new
-functionality was added.  Further complication arises the ability to have
-rejection messages defined by the administrator, and every Real time Black
-List returns a different message.  This paper documents the difficult
-process of parsing Postfix logs, and presents a program which parses logs
-and places the resulting data into a database for further processing by
-other applications.
+explain to a user why their mail is being rejected, or to check whether
+their new anti-spam measures are working.  The more adventurous will have
+generated some statistics to show how many hits each of their anti-spam
+measures has gotten in the last week, and possibly even generated some
+graphs to clearly illustrate the point to management or complaining
+users.\footnote{This was the author's first real foray into processing
+Postfix logs.}  Very few will have performed in-depth parsing and analysis
+of their logs, where the parsing must correlate the log lines
+per-connection or per-queueid, rather than processing lines independently.
+One of the barriers to this type of processing is the unstructured nature
+of Postfix's logs, where each logging line was added on an ad hoc basis as
+a requirement was discovered or new functionality was added.  Further
+complication arises as a result of the facility to have rejection messages
+defined by the administrator, and every Real-time Black List returns a
+different explanatory message.  This paper documents the difficult process
+of parsing Postfix logs, and presents a program which parses logs and
+places the resulting data into a database for further processing by other
+applications.
 
@@
 suitable for their needs, rather than a fixed set chosen by Postfix's
-author.  These restrictions can be applied and combined on a per-host,
-per-recipient or per-sender basis, allowing clients/recipients to be
-whitelisted with respect to specific restrictions.  Postfix leverages
-simple lookup tables to support arbitrarily complicated user-defined
-sequences of restrictions and exceptions, with the ultimate in flexibility
-being the facility to consult an external process which can implement
-whatever logic is required: if you wish you can restrict some users to
-sending mail on the third Tuesday after pay day only, if that is of any use
-to you.  Administrators can also supply their own rejection messages, to
-make it clear to senders why exactly their mail was rejected.
+author.  These restrictions can be selectively applied, combined and
+bypassed on a per-client, per-recipient or per-sender basis, allowing
+varying levels of defense and or permissiveness.  Postfix leverages simple
+lookup tables to support arbitrarily complicated user-defined sequences of
+restrictions and exceptions, with the ultimate in flexibility being the
+facility to consult an external process which can implement whatever logic
+is required: some users can be restricted to sending mail on the third
+Tuesday after pay day only, if that is of any use.\footnote{This example
+may actually be useful in a payroll system.  More commonly encountered
+scenarios are restricting posting to large mailing lists (e.g. sending
+weekly offers to customers) to out of office hours, checking SPF records
+(see http://www.openspf.org/ or
+http://en.wikipedia.org/wiki/Sender\_Policy\_Framework for details), or
+rate limiting clients.}  Administrators can also supply their own rejection
+messages, to make it clear to senders why exactly their mail was rejected.
 Unfortunately this flexibility has a cost: complexity in the logs
 generated.  While it is easy to use \texttt{grep(1)} to determine the fate
-of an individual email, following the progress of an email through Postfix
-can be quite difficult.  The logs tend to follow a 90\%/10\% pattern: 90\%
-of the time following the log entries is simple, but the other 10\% of the
-logging requires 90\% of the code.\footnote{These numbers don't have a
-solid scientific basis, they're based on gut feeling from writing the
-software.}
+of an individual email, following the journey an email takes through
+Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
+pattern: 90\% of the time the journey is simple, but the other 10\% of the
+time requires 90\% of the code.\footnote{These numbers don't have a solid
+scientific basis, they're based on gut feeling from writing and debugging
+the software.}
 
@@
 This paper and the program it describes are part of a larger project to
-optimise your Postfix restrictions, generate statistics and graphs, and
-provide a platform on which new restrictions can be trialled and evaluated
-to see if they are worth using in the fight against spam.  The program
-parses Postfix logs and populates a database with the data gleaned from
-those logs, providing a consistent and simple view of your logs which
+optimise a server's Postfix restrictions, generate statistics and graphs,
+and provide a platform on which new restrictions can be trialled and
+evaluated to see if they are worth using in the fight against spam.  The
+program parses Postfix logs and populates a database with the data gleaned
+from those logs, providing a consistent and simple view of the logs which
 future tools can utilise.  The program should be reasonably easily
-extensible, so if logging to a database doesn't meet requirements, one can
-implement one's own processing.
+extensible, so if logging to a database doesn't meet requirements,
+appropriate functionality can be implemented.
 
@@
 
-Before getting into detail about the parser, a brief overview is in order.
+Before getting into detail about the parser a brief overview is in order.
 The parser is split into two parts: the parsing algorithm and the rules
-which are applied to the lines.  Rules can be thought of as the
-tokeniser/Lex part of a Lex/YACC parser; they identify the line and return
-some data, which the algorithm (the YACC part) receives, performing the
-correct action.  Rules are all about identifying a line and extracting data
-from it, whereas the algorithm is all about following the path the mail
-takes through Postfix, piecing the data together into a coherent whole, and
-performing housekeeping duties.
+which are applied to the lines.  Rules can be thought of as the Lex part of
+a Lex and YACC style parser; rules identify the line and return some data,
+which the algorithm (the YACC part) receives, performing the requested
+action.  Rules are solely concerned with identifying a line and extracting
+data from it, whereas the algorithm's task is to follow the journey each
+mail takes through Postfix, piecing the data together into a coherent
+whole, and performing housekeeping duties.
 
@@
 in the parser; decoupling the parsing rules from the associated actions
-allows new rules to be written and tested without modifications to the
-algorithm source code (significantly lowering the barrier to entry for new
-or casual users), and greatly simplifies both algorithm and rules.  It also
-allows for a clear separation of functionality: rules handle low level
-details of identifying a line, whereas the algorithm handles the higher
-level details of following the path a mail takes through postfix,
-assembling the required data, etc.
+allows new rules to be written and tested without requiring modifications
+to the algorithm source code (significantly lowering the barrier to entry
+for new or casual users), and greatly simplifies both algorithm and rules.
+It also allows for a clear separation of functionality: rules handle low
+level details of identifying and extracting data from a line, whereas the
+algorithm handles the higher level details of following the path a mail
+takes through postfix, assembling the required data, etc.
 
-Rule characteristics:
+Rule have certain characteristics which may help in understanding the
+parser:
 
@@
         that line,\footnote{With one exception, detailed later in Section
-        \ref{actions-in-detail}.} but there is a facility for the specifying
-        the order of the rules, so that more specific rules can be matched
-        first.
+        \ref{actions-in-detail}.} but there is a facility for the
+        specifying the order of the rules, so that more specific rules can
+        be tried first.
 
     \item Rules are completely self-contained: there are no sub-rules, so
-        rules have linear complexity, and can be understood without
-        reference to any other rules.
+        rules have linear complexity, and can be understood in isolation,
+        without reference to any other rules.
+
+    \item XXX EXPAND ON THIS SOME MORE XXX
 
@@
 
-    \item [program] The program (smtp, smtpd, qmgr, etc) which the rule
-        applies to.  This avoids needlessly trying rules which won't
-        match the line, or worse, might match unintentionally.
+    \item [program] The program (smtp, smtpd, qmgr, etc.) whose log lines
+        the rule applies to.  This avoids needlessly trying rules which
+        won't match the line, or worse, might match unintentionally.
 
-    \item [regex] The regex to compare the log line against.  The regex
-        will first have several keywords expanded; this simplifies reading
-        and writing rules, avoids needless repetition of complex regex
+    \item [regex] The regex to match the log line against.  The regex will
+        first have several keywords expanded: this simplifies reading and
+        writing rules, avoids needless repetition of complex regex
         components, and allows the components to be corrected and/or
-        improved in one location.
+        improved in one location.  For efficiency the keywords are expanded
+        and every rule's regex is compiled before attempting to parse the
+        log file - otherwise each regex would be recompiled each time it
+        was used, resulting in an 468.95\% slowdown.
 
-    \item [result\_cols, connection\_cols] This specifies how the fields in
+        Averaged over 10 test runs:
+
+        \begin{itemize}
+
+            \item Compiled run time was 55.47 seconds, standard deviation
+                was 0.247 seconds.
+
+            \item Non-compiled run time was 260.14 seconds (4 minutes,
+                20.14 seconds), standard deviation was 0.96 seconds.
+
+        \end{itemize}
+
+        Logging to the database was disabled for the test runs.
+
+    \item [result\_cols, connection\_cols] Specifies how the fields in
         the log line will be extracted.  The format is: \newline
-        client\_hostname = 1; helo = 2; sender = 4; \newline
+        client\_hostname = 1; recipient = 2; sender = 4; \newline
         i.e. semi-colon separated assignment statements, with the variable
-        name on the left and the match from the regex (\$1, \$2 etc.) on the
-        right hand side (without \$).
+        name on the left and the matching field from the regex on the
+        right hand side.
 
-    \item [result\_data, connection\_data] Sometimes rules need to fake
-        a piece of data which isn't present in the log line: the most
-        common would be smtp\_code when mail is accepted.  The format of
-        the same as result\_cols and connection\_cols.
+    \item [result\_data, connection\_data] Sometimes rules need to fake a
+        piece of data which isn't present in the log line: e.g. setting
+        smtp\_code when mail is accepted.  The format is the same as
+        result\_cols and connection\_cols, except that arbitrary data can
+        be specified rather than just a number.
 
     \item [result]  XXX RESULT NEEDS TO BE RENAMED XXX This is the action
-        Postfix would have taken to generate this line.  Info is represents
-        an unspecified intermediate action that the parser is not
-        interested in per se, but which does log useful information.  The
-        extracted information will be used, but the other results will be
-        discarded.
+        Postfix must have taken to generate this line, with two exceptions:
+
+        \begin{itemize}
+
+            \item [info] Represents an unspecified intermediate action that
+                the parser is not interested in per se, but which does log
+                useful information, supplementing other log lines.
+
+            \item [ignored] An action which is not only uninteresting in
+                itself, but which also provides no useful data.
+
+        \end{itemize}
+
+        Uninteresting lines are parsed rather than ignored completely so
+        that any lines the parser isn't capable of handling become
+        immediately obvious.
 
@@
                 connection, then save the data from the log line in the
-                connection data structure.
+                connection data structure thus found.
 
             \item [commit] The mail is finished with, enter it in the
-                database.
+                database, and remove the mail from memory.
 
@@
 
-    \item [rule\_order] This is an efficiency measure.  A counter is
+    \item [rule\_order] This is an efficiency measure.  This counter is
         maintained for every rule and incremented each time the rule
         successfully matches.  At the start of each run the program sorts
-        the rules in ascending rule\_order, and at the end of the run
-        updates every rule's rule\_order so that rules which matched more
-        frequently have a lower rule\_order.  Assuming that the 
+        the rules in descending rule\_order, and at the end of the run
+        updates every rule's rule\_order.  Assuming that the 
         distribution of log lines is reasonably consistent, rules matching
@@
         matching less commonly occurring log lines, lowering the program's
-        execution time.  XXX GIVE EXAMPLE TIMINGS XXX
+        execution time.
+
+        Averaged over 10 test runs:
+
+        \begin{itemize}
+
+            \item Ascending order run time was 56.684 seconds, standard
+                deviation was 0.29 seconds.
+
+            \item Descending order (reversed) run time was 64.037 seconds,
+                standard deviation was 0.16 seconds.
+
+            \item Shuffled order run time was 57.19 seconds, standard
+                deviation was 0.95 seconds.
+
+        \end{itemize}
+
+        Logging to the database was disabled for the test runs.
 
@@
 While the rules are more than double (treble with 2.3.x rules?) the size of
-the parser, the rules are quite simple, and each rule is very much
+the algorithm, the rules are quite simple, and each rule is completely
 independent of its fellows.  The parser is significantly more complicated,
-and highly interdependent.
+and highly internally interdependent.
 
@@
 
-    \item Follow the progress of the mail until it's either delivered or
+    \item Follow the progress of the accepted mail until it's either delivered or
         bounced, log all data, and finish.
@@
         correlated by the queueid.  This is relatively minor, but does
-        require two versions of several functions, \texttt{by\_pid} and
+        require:
+
+        \begin{itemize}
+
+            \item Two versions of several functions, \texttt{by\_pid} and
                 \texttt{by\_queueid}.
 
+            \item Two data structures to hold the data structures for each
+                connection.
+
+            \item Most importantly: every section of code must know whether
+                it is needs to lookup the data structures by pid or
+                queueid.
+
+        \end{itemize}
+
     \item Multiple independent mails may be delivered during one
-        connection; this requires cloning the current data, so that
+        connection: this requires cloning the current data, so that
         subsequent mails won't trample over each other.  This must be done
@@
         been accepted log entries won't be correlated by pid for that mail
-        any more, so there isn't any ambiguity about which mail a given log
-        line belongs to.  The original connection will be discarded when
-        the client disconnects.  One possible (as yet unseen) difficulty is
-        distinguishing between rejects which belong to the current mail,
-        and rejects which belong to previous, unaccepted, mails.
+        any more (queueid will be used instead), so there isn't any
+        ambiguity about which mail a given log line belongs to.  The
+        original connection will be discarded unsaved when the client
+        disconnects.  One unsolved difficulty is distinguishing between
+        different groups of rejections, e.g.
+
+        \begin{enumerate}
+
+            \item The client attempts to deliver a mail, but it is
+                rejected.
+
+            \item The client issues the RSET command to reset the session.
+
+            \item The client attempts to deliver another mail, likewise
+                rejected.
+
+        \end{enumerate}
+
+        There should probably be two different entires in the database
+        resulting from the above sequence, but currently there will only be
+        one.
 
@@
         Unfortunately while all log lines from an individual process appear
-        in the logs in monotonic order, the order in which log lines from
-        different processes are interleaved is subject to the vagaries of
-        process scheduling.  In addition the first log line belonging to
-        the child mail (the log line cited above properly belongs to the
-        parent mail) is logged by qmgr, so depending on how busy qmgr (and
-        the server generally) is, the parent's creation log line may appear
-        before or after (usually after) the child's log line from
-        qmgr.\footnote{Postfix is quite paranoid about mail delivery, an
-        excellent characteristic for an MTA to posess, so it won't log that
-        the child has been created until it is absolutely certain that the
-        mail has been written to disk.  Because the parent then has to be
-        informed that the mail has been committed, it's most likely that
-        the qmgr will log first.}
+        in chronological order, the order in which log lines from different
+        processes are interleaved is subject to the vagaries of process
+        scheduling.  In addition the first log line belonging to the child
+        mail (the log line cited above properly belongs to the parent mail)
+        is logged by qmgr,\footnote{Qmgr is the Postfix daemon which
+        manages the mail queue, determining which mails will have delivery
+        attempted next.} so the order also depends on how busy qmgr
+        is.\footnote{Postfix is quite paranoid about mail delivery, an
+        excellent characteristic for an MTA to possess, so it won't log
+        that the child has been created until it is absolutely certain that
+        the mail has been written to disk.}
 
@@
         \textbf{before} the parent logs the creation line!  Thus mails
-        marked as faked cannot be committed, instead they must be marked as
-        commit ready and subsequently committed by the parent mail.  Quite
-        apart from identifying mail injected in an unknown fashion, or bugs
-        in the parser, faked mails need to be marked as such because they
-        lack some data present in the parent mail, and must copy that data
-        from their parent.
+        marked as faked cannot be entered into the database; instead they
+        must be marked as database ready and subsequently entered by the
+        parent mail.  Quite apart from identifying mail injected in an
+        unknown fashion, or bugs in the parser, faked mails need to be
+        marked as such because they lack some data present in the parent
+        mail, and must copy that data from their parent.  XXX EITHER REMOVE
+        THE LAST LINE OR EXPAND ON IT SOME MORE, AND FIGURE OUT WHY MORE
+        DATA ISN'T REQUIRED XXX
 

### 2007-03-26 doc/logparser.tex
Added an example rule and documentation for the new restriction_name field in rules.

@@
 
+    \item [restriction\_name] The restriction which caused the mail to be
+        rejected.  Only applicable to rules which have a result of
+        \texttt{rejected}, other rules will have an empty string.
+
     \item [regex] The regex to match the log line against.  The regex will
@@
 \end{description}
+
+\subsection{Example rule}
+
+This example rule matches the message Postfix logs when it rejects mail
+from a sender address where the domain has neither an MX record nor an A
+record, i.e. mail could not be delivered to the sender's address.  
+For full details see
+http://www.postfix.org/postconf.5.html\#reject\_unknown\_sender\_domain
+
+% Don't reformat this!
+\begin{tabular}[]{ll}
+
+name                & Unknown sender domain                             \\
+description         & We do not accept mail from unknown domains        \\
+program             & postfix/smtpd                                     \\
+restriction\_name   & reject\_unknown\_sender\_domain                   \\
+regex               & \verb!^<(__SENDER__)>: Sender address rejected: ! \\
+                    & \verb!  Domain not found; from=<\1> !             \\
+                    & \verb!  to=<(__RECIPIENT__)> proto=E?SMTP !       \\
+                    & \verb!  helo=<(__HELO__)>$!                       \\
+result\_cols        & recipient = 2; sender = 1                         \\
+connection\_cols    & helo = 3                                          \\
+result\_data        &                                                   \\
+connection\_data    &                                                   \\
+result              & REJECTED                                          \\
+action              & SAVE\_BY\_PID                                     \\
+queueid             & 0                                                 \\
+rule\_order         & 0                                                 \\
+priority            & 0                                                 \\
+
+\end{tabular}
 

### 2007-04-03 doc/logparser.tex
Major work: two thirds of the full algorithm description is written; it definitely requires editing and proof reading by someone not intimately connected with the code. There are supporting changes to the flow chart; they are minor but pervasive. Added assumptions section to Background. Explanatory comment added to the flowchart to clarify what it describes. Added actions in detail section; the skeleton is there, but it needs to be filled out. Expanded the limitations and improvements sections. Various punctuation corrections based on reading Eats, Shoots and Leaves. Various minor improvements

@@
 
-    Parsing Postfix logs is much more difficult than it first appears, but
+    Parsing Postfix logs is much more difficult than it first appears but
     it is possible.  This paper describes the process required, documenting
-    the parsing algorithm and rules, and explaining the difficulties
+    the parsing algorithm and rules, explaining the difficulties
     encountered, with reference to an implementation which stores data
@@
 may actually be useful in a payroll system.  More commonly encountered
-scenarios are restricting posting to large mailing lists (e.g. sending
-weekly offers to customers) to out of office hours, checking SPF records
-(see http://www.openspf.org/ or
-http://en.wikipedia.org/wiki/Sender\_Policy\_Framework for details), or
-rate limiting clients.}  Administrators can also supply their own rejection
-messages, to make it clear to senders why exactly their mail was rejected.
-Unfortunately this flexibility has a cost: complexity in the logs
-generated.  While it is easy to use \texttt{grep(1)} to determine the fate
-of an individual email, following the journey an email takes through
-Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
-pattern: 90\% of the time the journey is simple, but the other 10\% of the
-time requires 90\% of the code.\footnote{These numbers don't have a solid
-scientific basis, they're based on gut feeling from writing and debugging
-the software.}
+scenarios are restricting posting to large mailing lists (e.g.\ sending
+weekly offers to customers) to out of office hours, rate limiting clients,
+or checking SPF records (see http://www.openspf.org/ or
+http://en.wikipedia.org/wiki/Sender\_Policy\_Framework for details)}
+Administrators can also supply their own rejection messages, to make it
+clear to senders why exactly their mail was rejected.  Unfortunately this
+flexibility has a cost: complexity in the logs generated.  While it is easy
+to use \texttt{grep(1)} to determine the fate of an individual email,
+following the journey an email takes through Postfix can be quite
+difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
+the journey is simple, but the other 10\% of the time requires 90\% of the
+code.\footnote{These numbers don't have a solid scientific basis, they're
+based on gut feeling from writing and debugging the software.}
 
@@
 
+\subsection{Assumptions}
+
+The algorithm described and the program implementing it make a small number
+of (hopefully safe and reasonable) assumptions:
+
+\begin{itemize}
+
+    \item The logs are whole and complete; nothing has been removed, either
+        deliberately or accidentally (e.g.\ log rotation gone awry, file
+        system filling up, logging unable to cope with the volume of logs).
+
+    \item Postfix logs sufficient information to make it possible to
+        accurately reconstruct the actions it has taken.
+
+    \item The Postfix queue has not been tampered with, causing unexplained
+        appearance or disappearance of mail.
+
+\end{itemize}
+
+In some ways this task is similar to reverse engineering a program, or
+replicating a black box program based solely on its inputs and outputs.
+Although the source code is available, reading and understanding it would
+require a significant investment of time:
+
+\begin{tabular}[]{lll}
+
+    Postfix 2.3.8   & Postfix 2.4.0 &                   \\
+    82224           & 83965         & lines of code     \\
+    67146           & 68675         & lines of comments \\
+    17647           & 18069         & lines are blank   \\
+    167017          & 170709        & lines in total    \\
+
+\end{tabular}
+
 \section{Adaptable parsing: user defined rules}
@@
     \item The first matching rule wins: no further rules are tried against
-        that line,\footnote{With one exception, detailed later in Section
-        \ref{actions-in-detail}.} but there is a facility for the
-        specifying the order of the rules, so that more specific rules can
-        be tried first.
+        that line,\footnote{With one exception, details are provided later
+        in Section~\ref{actions-in-detail}.} but there is a facility for
+        the specifying the order of the rules, so that more specific rules
+        can be tried first.
 
@@
         and every rule's regex is compiled before attempting to parse the
-        log file - otherwise each regex would be recompiled each time it
+        log file --- otherwise each regex would be recompiled each time it
         was used, resulting in an 468.95\% slowdown.
@@
 
-    \item [result\_cols, connection\_cols] Specifies how the fields in
-        the log line will be extracted.  The format is: \newline
+    \item [result\_cols, connection\_cols] Specifies how the fields in the
+        log line will be extracted.  The format is: \newline
         client\_hostname = 1; recipient = 2; sender = 4; \newline
         i.e. semi-colon separated assignment statements, with the variable
-        name on the left and the matching field from the regex on the
-        right hand side.
+        name on the left and the matching field from the regex on the right
+        hand side.
 
     \item [result\_data, connection\_data] Sometimes rules need to fake a
-        piece of data which isn't present in the log line: e.g. setting
+        piece of data which isn't present in the log line: e.g.\ setting
         smtp\_code when mail is accepted.  The format is the same as
@@
 
-        A full list can be found in Section \ref{actions-in-detail}.
+        A full list can be found in Section~\ref{actions-in-detail}.
 
@@
         the rules in descending rule\_order, and at the end of the run
-        updates every rule's rule\_order.  Assuming that the 
-        distribution of log lines is reasonably consistent, rules matching
-        more commonly occurring log lines will be tried before rules
-        matching less commonly occurring log lines, lowering the program's
-        execution time.
+        updates every rule's rule\_order.  Assuming that the distribution
+        of log lines is reasonably consistent, rules matching more commonly
+        occurring log lines will be tried before rules matching less
+        commonly occurring log lines, lowering the program's execution
+        time.
 
@@
 from a sender address where the domain has neither an MX record nor an A
-record, i.e. mail could not be delivered to the sender's address.  
-For full details see
+record, i.e.\ mail could not be delivered to the sender's address.  For full
+details see
 http://www.postfix.org/postconf.5.html\#reject\_unknown\_sender\_domain
+
+XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
+PERHAPS WHEN THE ALGORITHM IS EXPLAINED\@? XXX
 
@@
 While the rules are more than double (treble with 2.3.x rules?) the size of
-the algorithm, the rules are quite simple, and each rule is completely
+the algorithm, the rules are quite simple and each rule is completely
 independent of its fellows.  The parser is significantly more complicated,
@@
         injected mails have, so the source is not immediately discernible;
-        from a strictly linear reading of the logs it \textit{usually}
+        from a strictly linear reading of the logs it \textit{usually\/}
         appears as if the child mail has appeared from thin air.
@@
 
-\subsection{Algorithm flow chart}
+\subsection{Flow chart}
+
+\label{flow-chart}
+
+This flow chart shows the paths the data representing a mail/connection can
+take through the parser algorithm.
 
 \includegraphics{logparser-flow-chart.ps}
-\label{flow-chart}
 
@@
 
+The intermingling of log entries from different mails immediately rules out
+the possibility of handling each mail in isolation; the parser must be
+capable of handling multiple mails in parallel, each potentially at a
+different stage in its journey, without any interference between mails ---
+except in the minority of cases where intra-mail interference is required.
+I felt that the best way to implement this was to maintain state
+information for every unfinished mail and manipulate it appropriately for
+each log line encountered.  The parser thus requires both a method of
+mapping log lines to the correct mail and a method of specifying the action
+the log line represents.  The former is achieved by using the pid of the
+smtpd during the initial phase, then switching to the queueid once the mail
+has been accepted.  The latter uses the action field of the rule which
+matched the log line, executing the code in the function named by the
+action.  From a high level viewpoint this design shares significant
+similarity with how an editor maintains state for multiple files and
+performs the actions the user initiates on the correct file.
+
+XXX USE A DISPATCH TABLE TO MAKE THIS ACCURATE XXX
+
+Section~\ref{actions-in-detail} explains the actions in substantive detail;
+this section will omit such detail because it would clutter and confuse the
+algorithm description.  The flow chart in section~\ref{flow-chart} should
+be consulted also.
+
+\subsubsection{Mail enters the system}
+
+\label{mail-enters-the-system}
+
+Everything starts off with a mail entering the system, by local submission
+via postdrop/sendmail, by SMTP, or by re-injection due to forwarding.
+Local submission is the simplest of the three: a queueid is assigned
+immediately and the sender address is logged (action: pickup;
+flowchart:~2).  SMTP is more complicated:
+
+\begin{enumerate}
+
+    \item First there is a connection from the remote client
+        (action: connect; flowchart:~1).
+
+    \item This is followed either by rejections (action: save\_by\_pid;
+        flowchart:~4) or acceptance of one or more mails (action: clone,
+        then save\_by\_pid; flowchart:~5,~4), or a combination of both.
+
+    \item The client disconnects (action: disconnect; flowchart:~6).  As
+        soon as a mail is accepted it is assigned a queueid; if there is no
+        queueid Postfix didn't accept any mail over that connection, it
+        rejected all attempts due to the configured restrictions.  Absence
+        of a queueid is checked for during disconnection handling (action:
+        still disconnect; flowchart:~7) and if none is found the data is
+        cleaned up and entered in the database (action: still disconnect;
+        flowchart:~8; subroutines: fixup\_connection and
+        commit\_connection), then removed from memory (action: still
+        disconnect; flowchart:~9).
+
+    \item If one or more mails were accepted there will be more log entries
+        later.
+
+\end{enumerate}
+
+Re-injection due to forwarding sadly lacks explicit log lines of its own;
+there are two implicit indications:
+
+\begin{enumerate}
+
+    \item The more commonly occurring indicator is qmgr selecting a mail
+        with a previously unseen queueid for delivery (action:
+        qmgr\_chooses\_mail; flowchart:~3), in which case a new data
+        structure will be created.  The mail will be marked as faked; this
+        flag should be subsequently cleared.  The data is then saved
+        (action: still qmgr\_chooses\_mail; flowchart:~3).
+
+    \item This indicator will be seen for every mail that is re-injected
+        but is less common in identifying re-injected mail because it
+        typically occurs after the indicator explained above.  Local
+        delivery re-injects the mail and logs a successful delivery rather
+        than delivering directly to a mailbox or program (action: track;
+        flowchart:~11).  In this case the mail may already have been
+        created (by qmgr\_chooses\_mail, above); if not a new data
+        structure will be created.  In both cases the new mail is marked as
+        a child of the parent.
+
+\end{enumerate}
+
+Re-injection is somewhat awkward to explain because it overlaps both the
+mail entering and mail delivery sections.  See
+section~\ref{tracking-re-injected-mail} for a full discussion.
+
+\subsubsection{Mail delivery}
+
+The obvious counterpart to mail entering the system is mail leaving the
+system, whether it is by deletion, bouncing or delivery.  All three are
+handled in exactly the same way:
+
+\begin{enumerate}
+
+    \item The sender and recipient addresses will be logged separately
+        (action: save\_by\_queueid; flowchart:~10).
+
+    \item Sometimes mail is re-injected and the child mail needs to be
+        tracked by the parent mail (action: track; flowchart:~11) --- see
+        section~\ref{tracking-re-injected-mail}.
+
+    \item Eventually the mail will be delivered (even if re-injected),
+        bounced, or deleted by the administrator (action: commit;
+        flowchart:~12).  This is the last line in the logs for this
+        particular mail (it may be indirectly referred to if it was
+        re-injected).  If it is neither parent nor child of re-injection
+        the data is cleaned up and entered in the database (action: still
+        commit; flowchart:~14), then deleted from memory (action: still
+        commit; flowchart:~15).  For re-injected mails see 
+        section~\ref{tracking-re-injected-mail}.
+
+\end{enumerate}
+
+I'd like to reiterate that the actions above happen whether the mail is
+delivered to a mailbox, piped to a command, delivered to a remote server,
+bounced (due to a mail loop, delivery failure, or five day timeout), or
+deleted by the administrator.  The exception is when the mail is
+re-injected due to forwarding, and that deserves its own section, which is
+coming up next.
+
+\subsubsection{Tracking re-injected mail}
+
+\label{tracking-re-injected-mail}
+
+It's probably apparent by now that tracking re-injected mails is the single
+most complex part of the parser.
+
 \subsection{Actions in detail}
@@
 \label{actions-in-detail}
+
+\begin{description}
+
+    \item [ignore] Ignore this line, so that all log lines we're interested
+        in are parsed, but this line doesn't have any effect.
+
+    \item [connect] Handle a connection from a remote client.  Complains if
+        a connection already exists for this client
+
+    \item [disconnect]
+
+    \item [save\_by\_pid]
+
+    \item [save\_by\_queueid]
+
+    \item [commit]
+
+    \item [track]
+
+    \item [restriction\_start]
+
+    \item [qmgr\_chooses\_mail] 
+        however qmgr selects every mail for delivery, so most of the time
+        this action merely falls through to saving the information from the
+        log line (action: save\_by\_queueid; flowchart:~10).
+
+    \item [pickup]
+
+    \item [clone]
+
+\end{description}
 
@@
         span logs.  This might also require purging old mails once they've
-        been in the state tables for too long.
+        been in the state tables for too long, though the presence of very
+        old mails indicates a bug in either the implementation or the
+        specification.
 
@@
         logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, this may possibly have changed in Postfix 2.3, or the
-        upcoming 2.4.}
+        2.2.10; this may possibly have changed in Postfix 2.3, or the
+        upcoming 2.4.}  It should be reasonably simple to write a policy
+        server which causes Postfix to log a warning containing the HELO
+        hostname when the DATA command is accepted.
+
+    \item The algorithm does not distinguish between mails where one or
+        more mails are rejected and a subsequent mail is accepted; it will
+        appear in the database as one mail with lots of rejections followed
+        by acceptance.  I don't believe it's possible to make this
+        distinction given the data Postfix logs, though it might be
+        possible to write a policy server to provide additional logging.
 
@@
 
+    \item Write the policy server referred to in the third limitation
+        above.
+
 \end{itemize}

### 2007-04-08 doc/logparser.tex
Lots of corrections from proof-reading. Added a motivation section, needs to be expanded. Expanded the abstract and other sections with uses for the gathered data; this should make it blindingly obvious that the data is useful and the whole thing is more than an academic exercise. Added regex expansions. Added acceptable variable names,and enforced them in the parser. Added a sample for log line for the example rule. Ordered actions so they correspond with the order in which they're usually called. Some more limitations and improvements. Expanded on lots of things, clarified too.

@@
 \usepackage[final]{graphicx}
+\renewcommand{\labelenumii}{\roman{enumii}:}
 
@@
     encountered, with reference to an implementation which stores data
-    gathered from the logs in an SQL database for later processing.
+    gathered from the logs in an SQL database.  The gathered data can then
+    be used to optimise current anti-spam measures, provide a baseline to
+    test new anti-spam measures against, or to produce statistics showing
+    how effective those measures are.
 
@@
 
+XXX FIGURE OUT WHEN TO USE MAIL AND WHEN TO USE CONNECTION; POSSIBLY STATE
+THAT THEY'RE MORE OR LESS EQUIVALENT  XXX
+
 \newpage
@@
 
+\newpage
 \section{Introduction}
@@
 their new anti-spam measures are working.  The more adventurous will have
-generated some statistics to show how many hits each of their anti-spam
-measures has gotten in the last week, and possibly even generated some
-graphs to clearly illustrate the point to management or complaining
+generated statistics to show how many hits each of their anti-spam measures
+has gotten in the last week, and possibly even generated some graphs to
+clearly illustrate the point to management or complaining
 users.\footnote{This was the author's first real foray into processing
@@
 of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid, rather than processing lines independently.
+per-connection or per-queueid rather than processing lines independently.
 One of the barriers to this type of processing is the unstructured nature
-of Postfix's logs, where each logging line was added on an ad hoc basis as
-a requirement was discovered or new functionality was added.  Further
+of Postfix logs, where each logging line was added on an ad hoc basis as a
+requirement was discovered or new functionality was added.  Further
 complication arises as a result of the facility to have rejection messages
-defined by the administrator, and every Real-time Black List returns a
-different explanatory message.  This paper documents the difficult process
-of parsing Postfix logs, and presents a program which parses logs and
-places the resulting data into a database for further processing by other
-applications.
+defined by the administrator, every Real-time Black List returns a
+different explanatory message, policy servers may have different messages
+depending on the characteristics of the connection, \ldots.  This paper
+documents the difficult process of parsing Postfix logs, and presents a
+program which parses logs and places the resulting data into a database.
+The gathered data can then be used to optimise current anti-spam measures,
+provide a baseline to test new anti-spam measures against, or to produce
+statistics showing how effective those measures are.  There are numerous
+other uses for such data: improving server performance by identifying
+troublesome destinations and reconfiguring appropriately or dedicating
+transports to those destinations; identifying high volume uses (e.g.
+customer newsletters) and restricting those uses to off-peak times; as a
+base for billing customers on a shared server; etc.
+
+XXX MAYBE EXPLAIN WHAT WILL BE IN THE FOLLOWING SECTIONS??? XXX
 
 \section{Background}
+
+\subsection{Motivation}
+
+This paper and the program it describes are part of a larger project to
+optimise a server's Postfix restrictions, generate statistics and graphs,
+and provide a platform on which new restrictions can be trialled and
+evaluated to see if they are worth using in the fight against spam.  The
+program parses Postfix logs and populates a database with the data gleaned
+from those logs, providing a consistent and simple view of the logs which
+future tools can utilise.
+
+XXX EXPLAIN WHY THIS IS USEFUL AND IMPORTANT XXX
+
+XXX ADD A SECTION EXPLAINING THE IMMEDIATE BENEFITS \\
+\texttt{SELECT name, description, restriction\_name, rule\_order \\
+    FROM rules \\
+    ORDER BY rule\_order;}\\
+Tells you which of your restrictions are most effective.
+
+XXX COMPARE TO PREVIOUS WORK XXX
 
@@
 http://en.wikipedia.org/wiki/Sender\_Policy\_Framework for details)}
-Administrators can also supply their own rejection messages, to make it
+Administrators can also supply their own rejection messages to make it
 clear to senders why exactly their mail was rejected.  Unfortunately this
@@
 
-\subsection{Paper background}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are worth using in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The program should be reasonably easily
-extensible, so if logging to a database doesn't meet requirements,
-appropriate functionality can be implemented.
-
 \subsection{Parser background}
@@
 a Lex and YACC style parser; rules identify the line and return some data,
-which the algorithm (the YACC part) receives, performing the requested
+which the algorithm (the YACC part) receives and performs the requested
 action.  Rules are solely concerned with identifying a line and extracting
@@
 mail takes through Postfix, piecing the data together into a coherent
-whole, and performing housekeeping duties.
+whole, saving it in a useful and consistent form, and performing
+housekeeping duties.
 
@@
         deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging unable to cope with the volume of logs).
+        system filling up, logging system unable to cope with the volume of
+        logs).
 
@@
 
-In some ways this task is similar to reverse engineering a program, or
-replicating a black box program based solely on its inputs and outputs.
-Although the source code is available, reading and understanding it would
-require a significant investment of time:
+In some ways this task is similar to reverse engineering or replicating a
+black box program based solely on its inputs and outputs.  Although the
+source code is available, reading and understanding it would require a
+significant investment of time:
 
-\begin{tabular}[]{lll}
+\begin{tabular}[]{llll}
 
-    Postfix 2.3.8   & Postfix 2.4.0 &                   \\
-    82224           & 83965         & lines of code     \\
-    67146           & 68675         & lines of comments \\
-    17647           & 18069         & lines are blank   \\
-    167017          & 170709        & lines in total    \\
+    Postfix 2.2.11  & Postfix 2.3.8   & Postfix 2.4.0 &                   \\
+    71548           & 82224           & 83965         & lines of code     \\
+    60962           & 67146           & 68675         & lines of comments \\
+    16117           & 17647           & 18069         & lines are blank   \\
+    148627          & 167017          & 170709        & lines in total    \\
 
 \end{tabular}
+
+\subsection{Other parsers}
+
+XXX ADD INFORMATION ABOUT OTHER POSTFIX LOG PARSERS HERE XXX
 
@@
 for new or casual users), and greatly simplifies both algorithm and rules.
-It also allows for a clear separation of functionality: rules handle low
-level details of identifying and extracting data from a line, whereas the
+It also creates a clear separation of functionality: rules handle low level
+details of identifying and extracting data from a line, whereas the
 algorithm handles the higher level details of following the path a mail
-takes through postfix, assembling the required data, etc.
+takes through Postfix, assembling the required data, etc.
 
@@
         in Section~\ref{actions-in-detail}.} but there is a facility for
-        the specifying the order of the rules, so that more specific rules
+        the specifying the order of the rules so that more specific rules
         can be tried first.
 
-    \item Rules are completely self-contained: there are no sub-rules, so
-        rules have linear complexity, and can be understood in isolation,
-        without reference to any other rules.
+    \item Rules are completely self-contained and can be understood in
+        isolation, without reference to any other rules.
 
-    \item XXX EXPAND ON THIS SOME MORE XXX
+    \item There are no sub-rules, so rules have linear computational
+        complexity.
 
@@
 
-    \item [regex] The regex to match the log line against.  The regex will
-        first have several keywords expanded: this simplifies reading and
-        writing rules, avoids needless repetition of complex regex
-        components, and allows the components to be corrected and/or
-        improved in one location.  For efficiency the keywords are expanded
-        and every rule's regex is compiled before attempting to parse the
-        log file --- otherwise each regex would be recompiled each time it
-        was used, resulting in an 468.95\% slowdown.
+    \item [regex] The regular expression\footnote{See
+        http://en.wikipedia.org/wiki/Regular\_expression for more
+        information about regular expressions in general.} to match the log
+        line against.  The regex will first have several keywords expanded:
+        this simplifies reading and writing rules, avoids needless
+        repetition of complex regex components, and allows the components
+        to be corrected and/or improved in one location.
+
+        The following keywords are expanded:
+
+        \texttt{\_\_SENDER\_\_, \_\_RECIPIENT\_\_, \_\_HELO\_\_,
+        \_\_HOSTNAME\_\_, \_\_IP\_\_, \newline\_\_SMTP\_CODE\_\_,
+        \_\_QUEUEID\_\_ and \_\_COMMAND\_\_.}
+
+        \_\_COMMAND\_\_ is an SMTP command, e.g. MAIL FROM; hopefully the
+        others are reasonably self explanatory.
+
+        For efficiency the keywords are expanded and every rule's regex is
+        compiled before attempting to parse the log file --- otherwise each
+        regex would be recompiled each time it was used, resulting in an
+        468.95\% slowdown.
 
@@
 
-        \begin{itemize}
+        \begin{description}
 
-            \item Compiled run time was 55.47 seconds, standard deviation
-                was 0.247 seconds.
+            \item [Compiled once and saved:] run time was 55.47 seconds,
+                standard deviation was 0.247 seconds.
 
-            \item Non-compiled run time was 260.14 seconds (4 minutes,
-                20.14 seconds), standard deviation was 0.96 seconds.
+            \item [Compiled every time and discarded:] run time was 260.14
+                seconds (4 minutes, 20.14 seconds) (468.95\% slower),
+                standard deviation was 0.96 seconds.
 
-        \end{itemize}
+        \end{description}
 
@@
 
+        XXX DO I NEED TO SAY ANYTHING MORE ABOUT THIS??
+
     \item [result\_cols, connection\_cols] Specifies how the fields in the
         log line will be extracted.  The format is: \newline
-        client\_hostname = 1; recipient = 2; sender = 4; \newline
-        i.e. semi-colon separated assignment statements, with the variable
-        name on the left and the matching field from the regex on the right
-        hand side.
+        \texttt{client\_hostname = 1; recipient = 2, sender = 4;} \newline
+        i.e.\ semi-colon or comma separated assignment statements, with the
+        variable name on the left and the matching field from the regex on
+        the right hand side.  The list of acceptable variable names is:
 
-    \item [result\_data, connection\_data] Sometimes rules need to fake a
+        \texttt{client\_hostname, client\_ip, server\_ip, server\_hostname,
+        \newline helo, sender, recipient, smtp\_code, data and child.}
+
+    \item [result\_data, connection\_data] Sometimes rules need to supply a
         piece of data which isn't present in the log line: e.g.\ setting
-        smtp\_code when mail is accepted.  The format is the same as
-        result\_cols and connection\_cols, except that arbitrary data can
-        be specified rather than just a number.
+        \texttt{smtp\_code} when mail is accepted.  The format and allowed
+        variables are the same as for \texttt{result\_cols} and
+        \texttt{connection\_cols}, except that arbitrary
+        data\footnote{Commas and semi-colons cannot be escaped and thus
+        cannot be used.  This is intended for small amounts of data rather
+        than large, so dealing with escape sequences seemed unnecessary.}
+        is permitted on the right hand side of the assignment.
 
@@
 
-        Uninteresting lines are parsed rather than ignored completely so
-        that any lines the parser isn't capable of handling become
-        immediately obvious.
+        Uninteresting lines are parsed so that any lines the parser isn't
+        capable of handling become immediately obvious errors.
 
@@
 
-        \begin{itemize}
+        \begin{description}
 
-            \item Ascending order run time was 56.684 seconds, standard
-                deviation was 0.29 seconds.
+            \item [Ascending order (best):] run time was 56.684 seconds,
+                standard deviation was 0.29 seconds.
 
-            \item Descending order (reversed) run time was 64.037 seconds,
-                standard deviation was 0.16 seconds.
+            \item [Descending order (worst):] run time was 64.037
+                seconds (12.97\% slower), standard deviation was 0.16
+                seconds.
 
-            \item Shuffled order run time was 57.19 seconds, standard
-                deviation was 0.95 seconds.
+            \item [Shuffled order:] run time was 57.19 seconds
+                (0.89\% slower), standard deviation was 0.95 seconds.
 
-        \end{itemize}
+        \end{description}
+
+        XXX RERUN OVER MORE LOG FILES XXX
 
         Logging to the database was disabled for the test runs.
+
+        It is hoped that this will prove to have greater effectiveness as
+        the number of rules grows; currently the effect is negligible,
+        particularly in comparison to compiling each regex once.
 
@@
 
+This rule would match the following log line:
+
+\begin{verbatim}
+NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 
+  550 <foo@bar.baz>: Sender address rejected: Domain not found;
+  from=<foo@bar.baz> to=<info@example.com> proto=SMTP
+  helo=<smtp.bar.baz>
+\end{verbatim}
+
 XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
-PERHAPS WHEN THE ALGORITHM IS EXPLAINED\@? XXX
+PERHAPS WHEN THE ALGORITHM IS PROPERLY EXPLAINED IT'LL BE EASIER\@?  MAKE
+REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
+HOW THE RULE MATCHES, ETC.  XXX
 
@@
 
-While the rules are more than double (treble with 2.3.x rules?) the size of
+While the rules are more than double (treble with 2.3.x rules? XXX ADD
+2.3.X RULES AND CHECK XXX) the size of
 the algorithm, the rules are quite simple and each rule is completely
-independent of its fellows.  The parser is significantly more complicated,
-and highly internally interdependent.
+independent of its fellows.  The algorithm is significantly more
+complicated and highly internally interdependent.
 
-\subsection{Naive approach}
+\subsection{A naive approach}
 
@@
 
-    \item Follow the progress of the accepted mail until it's either delivered or
-        bounced, log all data, and finish.
+    \item Follow the progress of the accepted mail until it's either
+        delivered, bounced or deleted, log all data, and finish.
 
@@
 \subsection{Complications encountered}
+
+\label{complications}
 
@@
 
-            \item Two data structures to hold the data structures for each
-                connection.
+            \item Two containing data structures to hold the data
+                structure for each connection.
 
@@
     \item Multiple independent mails may be delivered during one
-        connection: this requires cloning the current data, so that
-        subsequent mails won't trample over each other.  This must be done
-        every time a mail is accepted, as it's impossible to tell in
-        advance which connections will accept multiple mails.  It is quite
-        easy to overlook this complication because only a small minority of
-        connections accept more than one mail. Happily once the mail has
-        been accepted log entries won't be correlated by pid for that mail
-        any more (queueid will be used instead), so there isn't any
-        ambiguity about which mail a given log line belongs to.  The
-        original connection will be discarded unsaved when the client
-        disconnects.  One unsolved difficulty is distinguishing between
-        different groups of rejections, e.g.
+        connection: this requires cloning the current data as soon as a
+        mail is accepted, so that subsequent mails won't trample over each
+        other's data.  This must be done every time a mail is accepted, as
+        it's impossible to tell in advance which connections will accept
+        multiple mails.  It is quite easy to overlook this complication
+        because only a small minority of connections accept more than one
+        mail. Happily once the mail has been accepted log entries won't be
+        correlated by pid for that mail any more (its queueid will be used
+        instead), so there isn't any ambiguity about which mail a given log
+        line belongs to.  The original connection will be discarded unsaved
+        when the client disconnects.  One unsolved difficulty is
+        distinguishing between different groups of rejections, e.g. when
+        dealing with the following sequence:
 
@@
         delivered directly to a mailbox (or program; there is very little
-        difference between the two): sometimes they are aliased and
-        need to be delivered to an address (or addresses) on another
-        server.  When this occurs a child mail will be injected into the
-        postfix queue, but without the explicit logging smtpd or sendmail
-        injected mails have, so the source is not immediately discernible;
-        from a strictly linear reading of the logs it \textit{usually\/}
-        appears as if the child mail has appeared from thin air.
-        Subsequently the parent mail will log the creation of the child
-        mail:
+        difference between the two): sometimes they are accepted for a
+        local address but need to be delivered to one or more remote
+        addresses due to aliases.  When this occurs a child mail will be
+        injected into the postfix queue, but without the explicit logging
+        smtpd or sendmail injected mails have, so the source is not
+        immediately discernible from the log line in which the mail first
+        appears; from a strictly linear reading of the logs it
+        \textit{usually\/} appears as if the child mail has appeared from
+        thin air.  Subsequently the parent mail will log the creation of
+        the child mail:
 
@@
 
-        This has the effect that the parser cannot complain when it
-        encounters a log line from qmgr for a previously unseen mail; it
-        must mark the mail as potentially fake and subsequently clear the
-        faked flag if and when the origin of the mail becomes clear.
-        Obviously the parser could omit checking of where mails originate
-        from, but I feel that it is better to require an explicit source,
-        as bugs are more likely to be exposed.
+        Because of this the parser cannot complain when it encounters a log
+        line from qmgr for a previously unseen mail; it must mark the mail
+        as potentially fake and subsequently clear the faked flag if and
+        when the origin of the mail becomes clear.  Obviously the parser
+        could omit checking of where mails originate from, but I feel that
+        it is better to require an explicit source, as bugs are more likely
+        to be exposed.
 
         Process scheduling can have a still more confusing effect: quite
-        often the child mail will be created, delivered and removed
-        \textbf{before} the parent logs the creation line!  Thus mails
-        marked as faked cannot be entered into the database; instead they
-        must be marked as database ready and subsequently entered by the
-        parent mail.  Quite apart from identifying mail injected in an
-        unknown fashion, or bugs in the parser, faked mails need to be
+        often the child mail will be created, delivered and entirely
+        finished with \textbf{before} the parent logs the creation line!
+        Thus mails marked as faked cannot be entered into the database when
+        their final log line is parsed; instead they must be marked as
+        database ready and subsequently entered by the parent mail once it
+        has been identified.  Quite apart from identifying mail injected in
+        an unknown fashion, or bugs in the parser, faked mails need to be
         marked as such because they lack some data present in the parent
-        mail, and must copy that data from their parent.  XXX EITHER REMOVE
-        THE LAST LINE OR EXPAND ON IT SOME MORE, AND FIGURE OUT WHY MORE
-        DATA ISN'T REQUIRED XXX
+        mail, and must copy that data from their parent before being
+        entered in the database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND
+        ON IT SOME MORE.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED.
+
+        The algorithm requirements are discussed in
+        section~\ref{tracking-re-injected-mail}.
 
@@
 
+\newpage
 \subsection{Flow chart}
@@
 I felt that the best way to implement this was to maintain state
-information for every unfinished mail and manipulate it appropriately for
-each log line encountered.  The parser thus requires both a method of
-mapping log lines to the correct mail and a method of specifying the action
-the log line represents.  The former is achieved by using the pid of the
-smtpd during the initial phase, then switching to the queueid once the mail
-has been accepted.  The latter uses the action field of the rule which
-matched the log line, executing the code in the function named by the
-action.  From a high level viewpoint this design shares significant
-similarity with how an editor maintains state for multiple files and
-performs the actions the user initiates on the correct file.
+information for every unfinished mail and manipulate the appropriate mail
+correctly for each log line encountered.  The parser thus requires both a
+method of mapping log lines to the correct mail and a method of specifying
+the action the log line represents.  The former is achieved by using the
+pid of the smtpd to identify the correct mail during the initial phase,
+then switching to the queueid once the mail has been accepted.  The latter
+uses the action field of the rule which matched the log line, executing the
+code in the function named by the action.  From a high level viewpoint this
+design shares significant similarity with how an editor maintains state for
+multiple files and performs the actions the user initiates on the correct
+file.
 
@@
 algorithm description.  The flow chart in section~\ref{flow-chart} should
-be consulted also.
+be consulted while reading this section also.
 
@@
 
-Everything starts off with a mail entering the system, by local submission
-via postdrop/sendmail, by SMTP, or by re-injection due to forwarding.
-Local submission is the simplest of the three: a queueid is assigned
-immediately and the sender address is logged (action: pickup;
-flowchart:~2).  SMTP is more complicated:
+Everything starts off with a mail entering the system, whether by local
+submission via postdrop/sendmail, by SMTP, or by re-injection due to
+forwarding.  Local submission is the simplest of the three: a queueid is
+assigned immediately and the sender address is logged (action: pickup;
+flowchart:~2).
+
+SMTP is more complicated:
 
@@
 
-    \item This is followed either by rejections (action: save\_by\_pid;
-        flowchart:~4) or acceptance of one or more mails (action: clone,
-        then save\_by\_pid; flowchart:~5,~4), or a combination of both.
+    \item This is followed by rejection of sender, recipients, client
+        address, etc. (action: save\_by\_pid; flowchart:~4); acceptance of
+        one or more mails (action: clone, then save\_by\_pid;
+        flowchart:~5,~4); or some interleaving of both.
 
@@
         queueid Postfix didn't accept any mail over that connection, it
-        rejected all attempts due to the configured restrictions.  Absence
-        of a queueid is checked for during disconnection handling (action:
-        still disconnect; flowchart:~7) and if none is found the data is
-        cleaned up and entered in the database (action: still disconnect;
-        flowchart:~8; subroutines: fixup\_connection and
-        commit\_connection), then removed from memory (action: still
-        disconnect; flowchart:~9).
+        rejected all attempted deliveries due to the configured
+        restrictions.  Absence of a queueid is checked for during
+        disconnection handling (action: still disconnect; flowchart:~7) and
+        if none is found the data is cleaned up and entered in the database
+        (action: still disconnect; flowchart:~8; subroutines:
+        fixup\_connection and commit\_connection), then removed from memory
+        (action: still disconnect; flowchart:~9).
 
     \item If one or more mails were accepted there will be more log entries
-        later.
+        later, see section~\ref{mail-delivery}.
 
@@
 
-Re-injection due to forwarding sadly lacks explicit log lines of its own;
+Re-injection due to forwarding sadly lacks explicit log lines of its
+own;\footnote{Previously discussed in section~\ref{complications},
+complication 3.}  Re-injection is somewhat awkward to explain because it
+overlaps both the mail acceptance and mail delivery sections.  See
+section~\ref{tracking-re-injected-mail} for a full discussion.
+
+\subsubsection{Mail delivery}
+
+\label{mail-delivery}
+
+The obvious counterpart to mail entering the system is mail leaving the
+system, whether it is by deletion, bouncing, local or remote delivery.  All
+three are handled in exactly the same way:
+
+\begin{enumerate}
+
+    \item The sender and recipient addresses will be logged separately
+        (action: save\_by\_queueid; flowchart:~10).
+
+    \item Sometimes mail is re-injected and the child mail needs to be
+        tracked by the parent mail (action: track; flowchart:~11) --- see
+        section~\ref{tracking-re-injected-mail}.
+
+    \item Eventually the mail will be delivered (even if re-injected),
+        bounced, or deleted by the administrator (action: commit;
+        flowchart:~12).  This is the last log line for this particular mail
+        (though it may be indirectly referred to if it was re-injected).
+        If it is neither parent nor child of re-injection the data is
+        cleaned up and entered in the database (action: still commit;
+        flowchart:~14; subroutines fixup\_connection and
+        commit\_connection), then deleted from memory (action: still
+        commit; flowchart:~15).  For re-injected mails see
+        section~\ref{tracking-re-injected-mail}.
+
+\end{enumerate}
+
+I'd like to reiterate that the actions above happen whether the mail is
+delivered to a mailbox, piped to a command, delivered to a remote server,
+bounced (due to a mail loop, delivery failure, or five day timeout), or
+deleted by the administrator.  The exception is when the mail is
+re-injected due to forwarding, and that deserves its own section, which is
+coming up next.
+
+\subsubsection{Tracking re-injected mail}
+
+\label{tracking-re-injected-mail}
+
+It's probably apparent by now that tracking re-injected mails is the single
+most complex part of the parser.  Section~\ref{complications} complication
+3 has most of the previous discussion, should you wish to review it.
+
 there are two implicit indications:
@@
 
-Re-injection is somewhat awkward to explain because it overlaps both the
-mail entering and mail delivery sections.  See
-section~\ref{tracking-re-injected-mail} for a full discussion.
-
-\subsubsection{Mail delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether it is by deletion, bouncing or delivery.  All three are
-handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item The sender and recipient addresses will be logged separately
-        (action: save\_by\_queueid; flowchart:~10).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: track; flowchart:~11) --- see
-        section~\ref{tracking-re-injected-mail}.
-
-    \item Eventually the mail will be delivered (even if re-injected),
-        bounced, or deleted by the administrator (action: commit;
-        flowchart:~12).  This is the last line in the logs for this
-        particular mail (it may be indirectly referred to if it was
-        re-injected).  If it is neither parent nor child of re-injection
-        the data is cleaned up and entered in the database (action: still
-        commit; flowchart:~14), then deleted from memory (action: still
-        commit; flowchart:~15).  For re-injected mails see 
-        section~\ref{tracking-re-injected-mail}.
-
-\end{enumerate}
-
-I'd like to reiterate that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator.  The exception is when the mail is
-re-injected due to forwarding, and that deserves its own section, which is
-coming up next.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking-re-injected-mail}
-
-It's probably apparent by now that tracking re-injected mails is the single
-most complex part of the parser.
-
 \subsection{Actions in detail}
@@
 
+XXX ADD A LIST IN LOAD\_RULES() SPECIFYING THE ALLOWED ACTIONS XXX
+
+This list contains all the actions available in the algorithm, in the order
+in which they tend to occur in practice.
+
 \begin{description}
 
-    \item [ignore] Ignore this line, so that all log lines we're interested
-        in are parsed, but this line doesn't have any effect.
+    \item [ignore] Do nothing with this line, merely parse it so that
+        unparsed lines can be reported.
 
@@
 
-    \item [disconnect]
+    \item [restriction\_start]
 
@@
 
+    \item [clone]
+
+    \item [disconnect]
+
     \item [save\_by\_queueid]
-
-    \item [commit]
-
-    \item [track]
-
-    \item [restriction\_start]
 
@@
 
+    \item [track]
+
+    \item [commit]
+
     \item [pickup]
-
-    \item [clone]
 
@@
 
-Every piece of software suffers from some limitations, and there is almost
+Every piece of software suffers from some limitations and there is almost
 always room for improvement.
@@
 
-    \item Each new Postfix release requires new rules to be written, to
-        cope with the new messages.
+    \item Each new Postfix release requires new rules to be written to cope
+        with the new messages.  Similarly new RBLs, new policy servers and
+        new administrator defined rejection messages require new rules.
 
-    \item The program should save state at the end of each run, and reload
-        it at the start of the next run, to properly cope with mails which
-        span logs.  This might also require purging old mails once they've
-        been in the state tables for too long, though the presence of very
-        old mails indicates a bug in either the implementation or the
-        specification.
+    \item The program doesn't cope with mails which span log files:
+        currently they are reported as unfinished mails at the end of the
+        first run and are complained about when encountered during the next
+        run because they don't have a source.
 
@@
         server which causes Postfix to log a warning containing the HELO
-        hostname when the DATA command is accepted.
+        hostname when the DATA command is accepted.\footnote{See
+        http://www.postfix.org/SMTPD\_POLICY\_README.html for full
+        documentation about policy servers.}
 
@@
         distinction given the data Postfix logs, though it might be
-        possible to write a policy server to provide additional logging.
+        possible to write a policy server to provide additional
+        logging.
+
+    \item The program will not detect parsing the same log file twice,
+        resulting in the database containing duplicate entries.
 
@@
 
-    \item Write the policy server referred to in the third limitation
-        above.
+    \item Write the policy servers referred to in the limitations above.
+
+    \item Save state at the end of each run; load it at the start of the
+        next run.  This should allow the program to cope with mails
+        spanning log files.  This might also require purging old mails once
+        they've been in the state tables for too long, though the presence
+        of very old mails indicates a bug in either the implementation or
+        the specification.
 
@@
 
+\section{References}
+
 \end{document}

### 2007-04-10 doc/logparser.tex
Added a run down of the contents of the paper in the introduction.

@@
 
-XXX MAYBE EXPLAIN WHAT WILL BE IN THE FOLLOWING SECTIONS??? XXX
+Section~\ref{background} provides background information about: the
+motivation for this project; the Postfix Mail Transport Agent; a short
+overview of the parser; the assumptions inherent in the design of the
+algorithm and program; and a list of other Postfix log parsers.
+
+Section~\ref{rules} discusses the parsing rules in detail, explaining their
+structure, giving an example rule and sample data it matches successfully
+against.
+
+Section~\ref{parsing-algorithm} contains the meat of the paper, describing
+a naive parsing algorithm and the complications encountered which shaped
+the full algorithm, followed by a comprehensive explanation of the
+different stages of the algorithm and the actions taken during the
+execution of the algorithm.
+
+Section~\ref{limitations-improvements} lists the limitations of the program
+and/or the algorithm, then suggests some ways of dealing with the
+limitations, with the goal of improving parsing and reproducing the journey
+a mail takes through the internals of Postfix.
+
+Section~\ref{conclusion} describes the results of the project to date.  XXX
+IMPROVE THIS.
+
+Section~\ref{references} catalogues resources used in developing the
+algorithm, writing the program and preparing this paper.  
+Any resources referred to in the paper are also given, plus some additional
+resources expected to be helpful in understanding Postfix, anti-spam
+techniques, or the paper.
 
 \section{Background}
+
+\label{background}
 
@@
 \section{Adaptable parsing: user defined rules}
+
+\label{rules}
 
@@
 \section{Parsing algorithm}
+
+\label{parsing-algorithm}
 
@@
 
+\label{limitations-improvements}
+
 Every piece of software suffers from some limitations and there is almost
@@
 
+\label{conclusion}
+
 \section{References}
 
+\label{references}
+
+XXX ADD RESOURCES ONCE THE PAPER IS REASONABLY COMPLETE AND THEN REFERENCE
+THEM FROM WITHIN THE PAPER XXX
+
 \end{document}

### 2007-04-12 doc/logparser.tex
Added "Database as API" section because it seemed warranted when trying to expand the motivation section. Added example SQL queries to show what can be done with the data.  Should probably be moved to an appendix at some stage. Fixed minor typographical issues picked up by chktex. Added more place-holder XXX items.

@@
 
-XXX WHAT IS THE CONCLUSION??? XXX
+XXX WHAT IS THE CONCLUSION???
 
-XXX FIGURE OUT WHEN TO USE MAIL AND WHEN TO USE CONNECTION; POSSIBLY STATE
-THAT THEY'RE MORE OR LESS EQUIVALENT  XXX
+XXX FIGURE OUT WHEN TO USE MAIL AND WHEN TO USE CONNECTION\@; POSSIBLY STATE
+THAT THEY'RE MORE OR LESS EQUIVALENT
 
@@
 \section{Introduction}
+
+\label{introduction}
 
@@
 Section~\ref{conclusion} describes the results of the project to date.  XXX
-IMPROVE THIS.
+IMPROVE THIS\@.
 
@@
 from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.
+future tools can utilise.  The gathered data can then be used to optimise
+current anti-spam measures, provide a baseline to test new anti-spam
+measures against, or to produce statistics showing how effective those
+measures are.\footnote{Section~\ref{introduction} lists some more example
+uses.}
 
-XXX EXPLAIN WHY THIS IS USEFUL AND IMPORTANT XXX
+A short example of the optimisation possible using data from the database
+is optimal ordering of Postfix restrictions:
 
-XXX ADD A SECTION EXPLAINING THE IMMEDIATE BENEFITS \\
-\texttt{SELECT name, description, restriction\_name, rule\_order \\
-    FROM rules \\
-    ORDER BY rule\_order;}\\
-Tells you which of your restrictions are most effective.
+\begin{verbatim}
+SELECT name, description, restriction_name, rule_order
+    FROM rules
+    WHERE postfix_action = 'REJECTED'
+    ORDER BY rule_order DESC;
+\end{verbatim}
 
-XXX COMPARE TO PREVIOUS WORK XXX
+If the database supports sub-selects percentages can be
+obtained:\footnote{This works for SQLite3, where || is the concatenation
+operator.  If the database containing the data does not support
+concatenation, simply remove ``|| '\%\''' from the query.}
+
+XXX STOP THIS FROM BEING SPLIT ACROSS PAGES IF POSSIBLE
+
+\begin{verbatim}
+SELECT name, description, restriction_name, rule_order,
+        (rule_order * 100.0 /
+            (SELECT SUM(rule_order)
+                FROM rules
+                WHERE postfix_action = 'REJECTED'
+            )
+        ) || '%' AS percentage
+    FROM rules
+    WHERE postfix_action = 'REJECTED'
+    ORDER BY rule_order DESC;
+\end{verbatim}
+
+Another example is determining which restrictions are not effective:
+
+\begin{verbatim}
+SELECT name, description, restriction_name, rule_order,
+        (rule_order * 100.0 /
+            (SELECT SUM(rule_order)
+                FROM rules
+                WHERE postfix_action = 'REJECTED'
+            )
+        ) || '%' AS percentage
+    FROM rules
+    WHERE postfix_action = 'REJECTED'
+        AND rule_order < 100
+    ORDER BY rule_order ASC;
+\end{verbatim}
+
+XXX COMPARE TO OTHER PARSERS
+
+\subsection{Database as Application Programming Interface}
+
+The database populated by this program provides a simple interface to
+Postfix logs.  Although the interface is a database schema, it is in effect
+quite similar to the API provided by a library: it insulates both user and
+provider of the API from changes in the implementation of the other.  The
+algorithm implemented by the program can be improved; support can be added
+for earlier or later releases of Postfix; bugs can be fixed or limitations
+removed from the program.  Statistics and/or graphs can be generated from
+the database; new restrictions can be tested and the results inspected;
+trends in the fight against spam can emerge from historical data saved in
+the database.  Using a database simplifies writing programs which need to
+interact with the data in several ways:
+
+\begin{enumerate}
+
+    \item Libraries exist for the majority of programming languages
+        allowing access to databases, whereas if this program was written
+        as a library a new interface layer would need to be written for
+        every programming language used with the API\@.
+
+    \item Databases provide complex querying and sorting functionality to
+        the user without requiring large amounts of programming.  All
+        databases provide one or more programs, of variable complexity and
+        sophistication, which can be used for ad hoc queries with minimal
+        investment of time.
+
+    \item Databases are easily extensible, e.g.:
+
+        \begin{itemize}
+
+            \item Other tables can be added to the database.
+
+            \item New columns can be added to the tables used by the
+                program, with sufficient DEFAULT clauses or a clever
+                TRIGGER or two.
+
+            \item A VIEW gives a custom arrangement of data with very
+                little effort.
+
+            \item If the database supports it, access can be granted on a
+                fine-grained basis so that the finance department can
+                produce invoices, the helpdesk can run limited queries as
+                part of dealing with support calls, and the administrators
+                have full access to the data.
+
+            \item Triggers can be written to perform actions when certain
+                events occur.  In pseudo-SQL\@:
+
+\begin{verbatim}
+CREATE TRIGGER ON INSERT INTO results
+    WHERE sender = 'boss@example.com'
+        AND postfix_action = 'REJECTED'
+    SEND PANIC EMAIL TO 'postmaster@example.com';
+\end{verbatim}
+
+        \end{itemize}
+
+\end{enumerate}
 
@@
 
-XXX ADD INFORMATION ABOUT OTHER POSTFIX LOG PARSERS HERE XXX
+XXX ADD INFORMATION ABOUT OTHER POSTFIX LOG PARSERS HERE
 
@@
 
-        \_\_COMMAND\_\_ is an SMTP command, e.g. MAIL FROM; hopefully the
+        \_\_COMMAND\_\_ is an SMTP command, e.g. MAIL FROM\@; hopefully the
         others are reasonably self explanatory.
@@
         Logging to the database was disabled for the test runs.
+
+        XXX RERUN OVER MORE LOG FILES
 
@@
 
-    \item [result]  XXX RESULT NEEDS TO BE RENAMED XXX This is the action
-        Postfix must have taken to generate this line, with two exceptions:
+    \item [postfix\_action] This is the action Postfix must have taken to
+        generate this line, with two exceptions:
 
@@
 
-        XXX RERUN OVER MORE LOG FILES XXX
+        XXX RERUN OVER MORE LOG FILES
 
@@
 REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
-HOW THE RULE MATCHES, ETC.  XXX
+HOW THE RULE MATCHES, ETC\@.
 
@@
 connection\_data    &                                                   \\
-result              & REJECTED                                          \\
+postfix\_action     & REJECTED                                          \\
 action              & SAVE\_BY\_PID                                     \\
@@
 
-While the rules are more than double (treble with 2.3.x rules? XXX ADD
-2.3.X RULES AND CHECK XXX) the size of
-the algorithm, the rules are quite simple and each rule is completely
-independent of its fellows.  The algorithm is significantly more
-complicated and highly internally interdependent.
+While the rules have more lines of code than the algorithm, the rules are
+quite simple and each rule is completely independent of its fellows.  The
+algorithm is significantly more complicated and highly internally
+interdependent.
 
@@
         when the client disconnects.  One unsolved difficulty is
-        distinguishing between different groups of rejections, e.g. when
+        distinguishing between different groups of rejections, e.g.\ when
         dealing with the following sequence:
@@
         entered in the database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND
-        ON IT SOME MORE.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED.
+        ON IT SOME MORE\@.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED\@.
 
@@
 file.
-
-XXX USE A DISPATCH TABLE TO MAKE THIS ACCURATE XXX
 
@@
 
+XXX EXPAND THIS
+
 there are two implicit indications:
@@
 
-XXX ADD A LIST IN LOAD\_RULES() SPECIFYING THE ALLOWED ACTIONS XXX
+XXX USE THE SAME DOCUMENTATION HERE AS IN Parser.pm
 
@@
 XXX ADD RESOURCES ONCE THE PAPER IS REASONABLY COMPLETE AND THEN REFERENCE
-THEM FROM WITHIN THE PAPER XXX
+THEM FROM WITHIN THE PAPER
 

### 2007-04-15 doc/logparser.tex
Added review of other parsers.

@@
     Parsing Postfix logs is much more difficult than it first appears but
-    it is possible.  This paper describes the process required, documenting
-    the parsing algorithm and rules, explaining the difficulties
-    encountered, with reference to an implementation which stores data
-    gathered from the logs in an SQL database.  The gathered data can then
-    be used to optimise current anti-spam measures, provide a baseline to
-    test new anti-spam measures against, or to produce statistics showing
-    how effective those measures are.
+    it is possible to achieve a high degree of accuracy in understanding
+    the logs and reconstructing the actions taken by Postfix to generate
+    the logs.  This paper describes the process required, documenting the
+    parsing algorithm and rules, explaining the difficulties encountered,
+    with reference to an implementation which stores data gathered from the
+    logs in an SQL database.  The gathered data can then be used to
+    optimise current anti-spam measures, provide a baseline to test new
+    anti-spam measures against, or to produce statistics showing how
+    effective those measures are.
 
@@
 complication arises as a result of the facility to have rejection messages
-defined by the administrator, every Real-time Black List returns a
-different explanatory message, policy servers may have different messages
-depending on the characteristics of the connection, \ldots.  This paper
-documents the difficult process of parsing Postfix logs, and presents a
-program which parses logs and places the resulting data into a database.
-The gathered data can then be used to optimise current anti-spam measures,
-provide a baseline to test new anti-spam measures against, or to produce
-statistics showing how effective those measures are.  There are numerous
-other uses for such data: improving server performance by identifying
-troublesome destinations and reconfiguring appropriately or dedicating
-transports to those destinations; identifying high volume uses (e.g.
-customer newsletters) and restricting those uses to off-peak times; as a
-base for billing customers on a shared server; etc.
+defined by the administrator; every Real-time Black List returns a
+different explanatory message; policy servers may have different messages
+depending on the characteristics of the connection; there are many ways in
+which the log lines may differ.  This paper documents the difficult process
+of parsing Postfix logs, presenting a program which parses logs and places
+the resulting data into a database.  The gathered data can then be used to
+optimise current anti-spam measures, provide a baseline to test new
+anti-spam measures against, or to produce statistics showing how effective
+those measures are.  There are numerous other uses for such data: improving
+server performance by identifying troublesome destinations and
+reconfiguring appropriately or dedicating transports to those destinations;
+identifying high volume uses (e.g.  customer newsletters) and restricting
+those uses to off-peak times; as a base for billing customers on a shared
+server; etc.
 
@@
 overview of the parser; the assumptions inherent in the design of the
-algorithm and program; and a list of other Postfix log parsers.
+algorithm and program; using a database schema as an API; and a comparison
+against other Postfix log parsers.
 
@@
 \end{verbatim}
-
-XXX COMPARE TO OTHER PARSERS
 
@@
 
-XXX ADD INFORMATION ABOUT OTHER POSTFIX LOG PARSERS HERE
+10 other parsers have been reviewed in Appendix~\ref{other-parsers} as part
+of the background research for this project.  None of the reviewed parsers
+perform the type of advanced parsing and log correlation described here;
+all are intended to perform a specific parsing and reporting task, rather
+than be a generic parser and leave generating reports from the data to
+other programs.  Some save the data extracted to a database but the
+majority discard all data once finished running, making historical analysis
+impossible.  The other parsers listed all produce a report of greater or
+lesser complexity and detail, whereas the program described here doesn't
+attempt to produce a report at all; that responsibility is deferred to a
+separate program, to be developed later.  The parsing algorithm and program
+described here are designed to enable much more detailed log analysis by
+providing a stable platform for subsequent programs to develop upon.
+
+XXX DO I NEED TO SAY MORE??
 
@@
 
-XXX ADD RESOURCES ONCE THE PAPER IS REASONABLY COMPLETE AND THEN REFERENCE
-THEM FROM WITHIN THE PAPER
+\subsection{Other Postfix log parsers}
+
+\label{other-parsers}
+
+\begin{description}
+
+    \item [pflogsumm.pl] \textit{pflogsumm.pl is designed to provide an
+        over-view of postfix activity, with just enough detail to give the
+        administrator a "heads up" for potential trouble spots.}
+        \url{http://jimsun.linxnet.com/postfix_contrib.html}
+
+    \item [Sawmill Universal Log File Analysis and Reporting] is a
+        general purpose commercial product which parses 687 log file
+        formats (correct as of 2007/04/15) and produces reports.  I have not
+        experimented with it due to its limited data extraction facilities;
+        it does have three different sets of data for Postfix (as of
+        2007/04/15, one is beta), but they do not appear to be interlinked,
+        nor does it save sufficient data for the purposes of this project.
+        The source code is available in an obfuscated form only (presumably
+        for a fee), and the product is quite expensive, as it requires a
+        license for each report which is to be generated. \newline
+        \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline
+        \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
+        \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
+
+    \item [Splunk] XXX INVESTIGATE FURTHER \url{http://www.splunk.com}
+
+    \item [Isoqlog] \textit{Isoqlog is an MTA log analysis program written
+        in C. It designed to scan qmail, postfix, sendmail and exim logfile
+        and produce usage statistics in HTML format for viewing through a
+        browser. It produces Top domains output according to Sender,
+        Receiver, Total mails and bytes; it keeps your main domain mail
+        statistics with regard to Days Top Domain, Top Users values for per
+        day, per month and years.}  \newline
+        \url{http://www.enderunix.org/isoqlog/}
+
+    \item [AWStats] \textit{AWStats is a free powerful and featureful tool
+        that generates advanced web, streaming, ftp or mail server
+        statistics, graphically.}
+
+        AWStats will produce simple graphs for many different services, but
+        that restricts it to supporting the Lowest Common Denominator: the
+        data it will extract from an MTA log file is:
+        \texttt{time2, email, email\_r, host, host\_r, method, url, code,
+        and bytesd.} \newline
+        \url{http://awstats.sourceforge.net/} \newline
+        \url{http://awstats.sourceforge.net/awstats.mail.html} \newline
+        \url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
+
+    \item [Log analyser - throughput monitor] This utility tracks the
+        number of events which occurred over a particular time and warns if
+        the frequency of events passes a certain threshold.  It's designed
+        to provide real time alerts when dictionary attacks, mail loops or
+        similar problems occur.  \newline
+        \url{http://home.uninet.ee/~ragnar/throughput_monitor/}
+
+    \item [Anteater] \textit{The Anteater project is a Mail Traffic
+        Analyser. Anteater supports currently the logformat produced by
+        Sendmail and by Postfix. The tool is written in 100\% C++ and is
+        very easy to customize. Input, output, and the analysis are modular
+        class objects with a clear interface. There are eight useful
+        analyse modules, writing the result in plain ASCII or HTML, to
+        stdout or to files.}
+
+        Anteater hasn't been updated since November 2003, and doesn't have
+        any English documentation.
+
+        \url{http://anteater.drzoom.ch/}
+
+    \item [Yet Another Advanced Logfile Analyser] uses a plugin based
+        system to analyse log files and produce output reports.  The core
+        code is merely 91 lines long, as all the parsing and report
+        generation is handled by modules.  Using YAALA as a base would be
+        only slightly less work as both input and output modules would need
+        to be written; it may even be more work to implement the parser
+        within the constraints of YAALA.
+
+        \url{http://yaala.org/}
+
+    \item [Logparser/Lire] Lire is another general purpose log parser, but
+        is developed under the GNU GPL.  Like other general purpose log
+        parsers it doesn't extract enough data from Postfix logs to be
+        worthwhile.  Development also seems to have stalled, with only a
+        minor update since October 2004.  \newline
+        \url{http://logreport.org/lire/}
+
+    \item [Logrep] \textit{Logrep is a secure multi-platform framework for
+        the collection, extraction, and presentation of information from
+        various log files.}
+
+        The parsing Logrep performs is extremely basic, counting the number
+        of different processes executed, from address, to addresses, number
+        of recipients, bytes transferred and delay.  \newline
+        \url{http://www.itefix.no/phpws/index.php}
+
+\end{description}
+
+\subsection{Postfix resources}
+
+\subsection{SMTP references}
+
+XXX ADD RESOURCES/REFERENCES HERE AND THEN MAKE REFERENCES TO THEM FROM WITHIN THE PAPER
 

### 2007-04-15 doc/logparser.tex
Tweaks suggested by chktex.

@@
 reconfiguring appropriately or dedicating transports to those destinations;
-identifying high volume uses (e.g.  customer newsletters) and restricting
+identifying high volume uses (e.g.\ customer newsletters) and restricting
 those uses to off-peak times; as a base for billing customers on a shared
@@
 overview of the parser; the assumptions inherent in the design of the
-algorithm and program; using a database schema as an API; and a comparison
-against other Postfix log parsers.
+algorithm and program; using a database schema as an API\@; and a
+comparison against other Postfix log parsers.
 
@@
         over-view of postfix activity, with just enough detail to give the
-        administrator a "heads up" for potential trouble spots.}
+        administrator a ``heads up'' for potential trouble spots.\/}
         \url{http://jimsun.linxnet.com/postfix_contrib.html}
@@
         statistics with regard to Days Top Domain, Top Users values for per
-        day, per month and years.}  \newline
+        day, per month and years.\/}  \newline
         \url{http://www.enderunix.org/isoqlog/}
@@
         that generates advanced web, streaming, ftp or mail server
-        statistics, graphically.}
+        statistics, graphically.\/}
 
@@
 
-    \item [Log analyser - throughput monitor] This utility tracks the
+    \item [Log analyser --- throughput monitor] This utility tracks the
         number of events which occurred over a particular time and warns if
@@
         analyse modules, writing the result in plain ASCII or HTML, to
-        stdout or to files.}
+        stdout or to files.\/}
 
@@
         to be written; it may even be more work to implement the parser
-        within the constraints of YAALA.
+        within the constraints of YAALA\@.
 
@@
     \item [Logparser/Lire] Lire is another general purpose log parser, but
-        is developed under the GNU GPL.  Like other general purpose log
+        is developed under the GNU GPL\@.  Like other general purpose log
         parsers it doesn't extract enough data from Postfix logs to be
@@
         the collection, extraction, and presentation of information from
-        various log files.}
+        various log files.\/}
 

### 2007-04-16 doc/logparser.tex
Improved the tracking section, and extended it with a description of the algorithm. Changed faked to "of unknown origin" to make it clearer. Removed unnecessary action references cluttering up the algorithm description. Rewrote some of the algorithm stuff so it's hopefully clearer.

@@
 operator.  If the database containing the data does not support
-concatenation, simply remove ``|| '\%\''' from the query.}
+this syntax, simply remove ``|| '\%\''' from the query; the results will be
+the same, just slightly less visually pleasing.}
 
@@
         Because of this the parser cannot complain when it encounters a log
-        line from qmgr for a previously unseen mail; it must mark the mail
-        as potentially fake and subsequently clear the faked flag if and
+        line from qmgr for a previously unseen mail; it must flag the mail
+        as coming from an unknown origin and subsequently clear the flag if and
         when the origin of the mail becomes clear.  Obviously the parser
@@
         finished with \textbf{before} the parent logs the creation line!
-        Thus mails marked as faked cannot be entered into the database when
-        their final log line is parsed; instead they must be marked as
-        database ready and subsequently entered by the parent mail once it
-        has been identified.  Quite apart from identifying mail injected in
-        an unknown fashion, or bugs in the parser, faked mails need to be
-        marked as such because they lack some data present in the parent
-        mail, and must copy that data from their parent before being
-        entered in the database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND
-        ON IT SOME MORE\@.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED\@.
+        Thus mails flagged as coming from an unknown origin cannot be
+        entered into the database when their final log line is parsed;
+        instead they must be marked as database ready and subsequently
+        entered by the parent mail once it has been identified.  Quite
+        apart from identifying mail injected in an unknown fashion, or bugs
+        in the parser, unknown origin mails need to be marked as such
+        because they lack some data present in the parent mail, and must
+        copy that data from their parent before being entered in the
+        database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND ON IT SOME
+        MORE\@.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED\@.
 
@@
 this section will omit such detail because it would clutter and confuse the
-algorithm description.  The flow chart in section~\ref{flow-chart} should
-be consulted while reading this section also.
+algorithm description.  The data flow chart in section~\ref{flow-chart}
+should be consulted while reading this section also.
 
@@
     \item The client disconnects (action: disconnect; flowchart:~6).  As
-        soon as a mail is accepted it is assigned a queueid; if there is no
-        queueid Postfix didn't accept any mail over that connection, it
-        rejected all attempted deliveries due to the configured
+        soon as a mail is accepted it is assigned a queueid; thus if there
+        is no queueid Postfix didn't accept any mail over that connection,
+        it rejected all attempted deliveries due to the configured
         restrictions.  Absence of a queueid is checked for during
-        disconnection handling (action: still disconnect; flowchart:~7) and
-        if none is found the data is cleaned up and entered in the database
-        (action: still disconnect; flowchart:~8; subroutines:
-        fixup\_connection and commit\_connection), then removed from memory
-        (action: still disconnect; flowchart:~9).
+        disconnection handling (flowchart:~7) and if none is found the data
+        is cleaned up and entered in the database (flowchart:~8;
+        subroutines: fixup\_connection and commit\_connection), then
+        removed from memory (flowchart:~9).
 
@@
         tracked by the parent mail (action: track; flowchart:~11) --- see
-        section~\ref{tracking-re-injected-mail}.
-
-    \item Eventually the mail will be delivered (even if re-injected),
-        bounced, or deleted by the administrator (action: commit;
-        flowchart:~12).  This is the last log line for this particular mail
-        (though it may be indirectly referred to if it was re-injected).
-        If it is neither parent nor child of re-injection the data is
-        cleaned up and entered in the database (action: still commit;
-        flowchart:~14; subroutines fixup\_connection and
-        commit\_connection), then deleted from memory (action: still
-        commit; flowchart:~15).  For re-injected mails see
-        section~\ref{tracking-re-injected-mail}.
+        section~\ref{tracking-re-injected-mail} for further details.
+
+    \item Eventually the mail will be delivered, bounced, or deleted by the
+        administrator (action: commit; flowchart:~12).  This is the last
+        log line for this particular mail (though it may be indirectly
+        referred to if it was re-injected).  If it is neither parent nor
+        child of re-injection the data is cleaned up and entered in the
+        database (flowchart:~14; subroutines fixup\_connection and
+        commit\_connection), then deleted from memory (flowchart:~15).  For
+        re-injected mails see section~\ref{tracking-re-injected-mail}.
 
@@
 bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator.  The exception is when the mail is
-re-injected due to forwarding, and that deserves its own section, which is
-coming up next.
+deleted by the administrator.  The exception is what happens after delivery
+to the parent or children of mail re-injected due to forwarding, and that
+is explained in section~\ref{tracking-re-injected-mail}.
 
@@
 It's probably apparent by now that tracking re-injected mails is the single
-most complex part of the parser.  Section~\ref{complications} complication
+most complex part of the parser.  Section~\ref{complications}, complication
 3 has most of the previous discussion, should you wish to review it.
 
-XXX EXPAND THIS
-
-there are two implicit indications:
+The crux of the problem is that re-injected mails appear in the logs
+without explicit logging indicating their source.  There are two implicit
+indications:
 
@@
 
-    \item The more commonly occurring indicator is qmgr selecting a mail
-        with a previously unseen queueid for delivery (action:
-        qmgr\_chooses\_mail; flowchart:~3), in which case a new data
-        structure will be created.  The mail will be marked as faked; this
-        flag should be subsequently cleared.  The data is then saved
-        (action: still qmgr\_chooses\_mail; flowchart:~3).
-
-    \item This indicator will be seen for every mail that is re-injected
-        but is less common in identifying re-injected mail because it
-        typically occurs after the indicator explained above.  Local
-        delivery re-injects the mail and logs a successful delivery rather
-        than delivering directly to a mailbox or program (action: track;
-        flowchart:~11).  In this case the mail may already have been
-        created (by qmgr\_chooses\_mail, above); if not a new data
-        structure will be created.  In both cases the new mail is marked as
-        a child of the parent.
+    \item The indicator which more commonly introduces re-injection is qmgr
+        selecting a mail with a previously unseen queueid for delivery
+        (action: qmgr\_chooses\_mail; flowchart:~3), in which case a new
+        data structure will be created.  The mail will be flagged as having
+        unknown origins; this flag should be subsequently cleared once the
+        origin has been established.  The data is then saved (flowchart:~3;
+        subroutine: save).
+
+    \item Local delivery re-injects the mail and logs a relayed delivery
+        rather than delivering directly to a mailbox or program as it
+        usually would (action: track; flowchart:~11).\footnote{Relayed
+        delivery is performed by the SMTP client; local delivery means
+        local to the server, i.e.\ an address the server accepts mail for.}
+        In this case the mail may already have been created (action:
+        qmgr\_chooses\_mail; flowchart:~3; described above) and the unknown
+        origin flag will be cleared; if not a new data structure will be
+        created.  In both cases the new mail is marked as a child of the
+        parent.  The log line in question is:
+
+        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline 
+        delay=0, status=sent (forwarded as 56F5B43FD)}
+
+        This second indicator always occurs for re-injected mail but
+        typically it occurs after the first indicator explained above, and
+        so rarely introduces re-injection.  This indicator is required to
+        tie the parent and child mails together and so is central to the
+        process of tracking re-injected mails.
 
 \end{enumerate}
+
+The algorithm for tracking and saving re-injected mail to the database will
+finally be described:
+
+\begin{itemize}
+
+    \item If the mail is of unknown origin it is a child mail whose parent
+        hasn't yet been identified (action: commit; flowchart:~16).  Mark
+        the mail as ready for entry in the database (flowchart:~17), and
+        wait for the parent to deal with it ( flowchart:~18).  The mail
+        should never have another directly associated log entry; only its
+        parent should refer to it.
+
+    \item If the mail is a child mail then it has already been tracked
+        (action: commit; flowchart:~19): the data is cleaned up, pulling
+        data from its parent if necessary, and entered in the database
+        (flowchart:~20; subroutines fixup\_connection and
+        commit\_connection), then deleted from memory (flowchart:~21).  The
+        child mail will inform the parent mail that it is finished with
+        (flowchart:~22); if this is the last child and the parent is also
+        finished the parent will delete itself.  The child mail is now
+        finished with.
+
+    \item The last alternative is that the mail is a parent mail (action:
+        commit; flowchart:~23).  Regardless of the state of its children
+        the data is cleaned up and entered in the database ( flowchart:~24;
+        subroutines fixup\_connection and commit\_connection).  The parent
+        may have children which are waiting to be entered in the database (
+        flowchart:~25); for each of those children their data is cleaned up
+        and entered in the database, then the child is finished with and
+        its data discarded.  The parent may also have outstanding children
+        which are not yet delivered, in which case (flowchart:~26) the
+        parent must wait for all children to finish.  As soon as the last
+        child is finished with the parent is also finished with
+        (flowchart:~27).
+
+\end{itemize}
+
+A parent mail can have multiple children, which may be delivered before or
+after the parent mail.
 

### 2007-04-18 doc/logparser.tex
Added Policy server section. Added more XXX items, including placeholder log coverage section.

@@
 \usepackage[final]{graphicx}
+% Change how nested enumerate environments are labelled.
 \renewcommand{\labelenumii}{\roman{enumii}:}
@@
 
-XXX WHAT IS THE CONCLUSION???
+XXX WHAT IS THE CONCLUSION???  FIRST THING IS TO COMPARE TO OTHER PARSERS
+REVIEWED
 
@@
 lookup tables to support arbitrarily complicated user-defined sequences of
-restrictions and exceptions, with the ultimate in flexibility being the
-facility to consult an external process which can implement whatever logic
-is required: some users can be restricted to sending mail on the third
-Tuesday after pay day only, if that is of any use.\footnote{This example
-may actually be useful in a payroll system.  More commonly encountered
-scenarios are restricting posting to large mailing lists (e.g.\ sending
-weekly offers to customers) to out of office hours, rate limiting clients,
-or checking SPF records (see http://www.openspf.org/ or
-http://en.wikipedia.org/wiki/Sender\_Policy\_Framework for details)}
-Administrators can also supply their own rejection messages to make it
-clear to senders why exactly their mail was rejected.  Unfortunately this
-flexibility has a cost: complexity in the logs generated.  While it is easy
-to use \texttt{grep(1)} to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
+restrictions and exceptions, with policy servers as the ultimate in
+flexibility.  Administrators can also supply their own rejection messages
+to make it clear to senders why exactly their mail was rejected.
+Unfortunately this flexibility has a cost: complexity in the logs
+generated.  While it is easy to use \texttt{grep(1)} to determine the fate
+of an individual email, following the journey an email takes through
+Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
+pattern: 90\% of the time the journey is simple, but the other 10\% of the
+time requires 90\% of the code.\footnote{These numbers don't have a solid
+scientific basis, they're based on gut feeling from writing and debugging
+the software.}
+
+A policy server is an external program that accepts state information from
+Postfix for each SMTP command not rejected by an earlier restriction; the
+policy server can utilise that state information to implement whatever
+logic is required.  E.g.\ some users can be restricted to sending mail on
+the third Tuesday after pay day only --- this example may actually be
+useful in a payroll system to prevent spam or (worse) phishing mails with
+faked sender addresses.  More commonly encountered scenarios are:
+
+\begin{itemize}
+
+    \item Checking Sender Policy Framework records.  SPF records specify
+        which mail servers are allowed to send mail claiming to be from a
+        particular domain.  The intention is to reduce faked sender
+        addresses, which will in turn reduce spam, backscatter and
+        joe-jobs; however there has been a lot of resistance to the
+        proposal because it breaks or vastly complicates some features of
+        SMTP.  For more details see \url{http://www.openspf.org/} or
+        \url{http://en.wikipedia.org/wiki/Sender_Policy_Framework} XXX ADD
+        REFERENCES FOR SPAM BACKSCATTER AND JOE-JOBS TO THE APPENDIX.
+
+    \item Greylisting, a technique that temporarily rejects mail where the
+        triple of (sender, recipient, remote IP address) is unknown; the
+        assumption is that it's uneconomical for a spammer to retry, but
+        that a legitimate mail server must retry.  Sadly spammers are using
+        increasingly complex and well written programs to distribute spam,
+        frequently using an ISP provided SMTP server from a compromised
+        machine on its network, so greylisting is slowly becoming less
+        useful, but it does block a large percentage of spam mail at the
+        moment.  See \url{http://www.greylisting.org/} for full details.
+
+    \item Using a scoring system such as Policyd-weight where 
+        tests accumulate points against the sending system --- if the
+        eventual score is too high the mail is rejected.
+        \url{http://www.policyd-weight.org/}
+
+    \item Policyd performs the following: \textit{Greylisting,
+        Sender-(envelope, SASL or host / ip)-based throttling (on messages
+        and/or volume per defined time unit), Recipient rate limiting,
+        Spamtrap monitoring / blacklisting, HELO auto blacklisting and HELO
+        randomization preventation.\/}  \url{http://policyd.sourceforge.net}
+
+\end{itemize}
+
+Example attributes taken from \newline
+\url{http://www.postfix.org/SMTPD_POLICY_README.html}:
+
+\begin{tabular}[]{ll}
+
+    request                 & smtpd\_access\_policy     \\
+    protocol\_state         & RCPT                      \\
+    protocol\_name          & SMTP                      \\
+    helo\_name              & some.domain.tld           \\
+    queue\_id               & 8045F2AB23                \\
+    sender                  & foo@bar.tld               \\
+    recipient               & bar@foo.tld               \\
+    recipient\_count        & 0                         \\
+    client\_address         & 1.2.3.4                   \\
+    client\_name            & another.domain.tld        \\
+    reverse\_client\_name   & another.domain.tld        \\
+    instance                & 123.456.7                 \\
+
+\end{tabular}
 
@@
 housekeeping duties.
+
+XXX REFER TO THE GRAMMAR COMPARISON SECTION ONCE ITS WRITTEN.
 
@@
 
+XXX COMPARE AND CONTRAST WITH CONTEXT FREE GRAMMARS, OTHER GRAMMARS.
+
+XXX DESCRIBE IT IN LEFT: RIGHT PRODUCTION TERMS, IF POSSIBLE.
+
 \subsection{Rule attributes}
@@
 
+        XXX RERUN AGAIN ONCE POSTFIX 2.3 RULES HAVE BEEN ADDED TO SEE WHAT
+        THE DIFFERENCE IS.  THERE ARE 110 POSTFIX 2.2 RULES AS OF
+        2007/04/18
+
         Logging to the database was disabled for the test runs.
@@
 REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
-HOW THE RULE MATCHES, ETC\@.
+HOW THE RULE MATCHES, AND SO ON\@.
 
@@
 
+\section{Parsing coverage}
+
+XXX THIS SECTION NEEDS TO BE WRITTEN
+
+\label{parsing-coverage}
+
+\subsection{Log lines covered}
+
+\label{log-lines-covered}
+
+\subsection{Mails covered}
+
+\label{mails-covered}
+
 \section{Limitations and possible improvements}
@@
     \item Each new Postfix release requires new rules to be written to cope
-        with the new messages.  Similarly new RBLs, new policy servers and
+        with the new log lines.  Similarly new RBLs, new policy servers and
         new administrator defined rejection messages require new rules.

### 2007-04-18 stats/time2secs.pl
Moved stats stuff into a directory of its own.

@@ -1,0 +1,28 @@
+#!/usr/bin/env perl
+
+# $Id$
+
+use strict;
+use warnings;
+
+use Regexp::Common qw(number);
+
+my $hms_re = qr/^($RE{num}{int}+):($RE{num}{int}+):($RE{num}{real}+)$/x;
+my  $ms_re = qr/^                 ($RE{num}{int}+):($RE{num}{real}+)$/x;
+
+sub time2secs {
+    my ($time) = @_;
+    my $secs;
+
+    if ($time =~ m/$hms_re/) {
+        my ($hour, $minute, $second) = ($1, $2, $3);
+        $secs = ($hour * 60 * 60) + ($minute * 60) + $second;
+    } elsif ($time =~ m/$ms_re/) {
+        my ($minute, $second) = ($1, $2);
+        $secs = ($minute * 60) + $second;
+    } else {
+        die qq{$0: time2secs: failed to parse $time\n};
+    }
+
+    return $secs;
+}

### 2007-05-14 doc/logparser.tex
Documented and discussed identifying bounce notifications.  It'll need a second draft, but this is the basis of it.

@@
 
+\subsection{XXX}
+
+\subsubsection{Identifying bounce notifications}
+
+\label{identifying-bounce-notifications}
+
+Postfix 2.2.x (and presumably previous versions) lacks explicit logging
+when bounce notifications are generated; suddenly there will be log entries
+for a mail which lacks an obvious source.  There are similarities to the
+problem of child mails discussed in section
+\ref{tracking-re-injected-mail}, but unlike the solution described therein
+bounce notifications do not eventually have a log line which identifies
+their source.  The solution is that heuristics must be used to identify
+them, and those heuristics are:
+
+\begin{enumerate}
+
+    \item The sender address will be <>.
+
+    \item Neither \texttt{postfix/smtpd} nor \texttt{postfix/pickup} will
+        have logged any messages associated with the mail, viz it was
+        generated internally by Postfix, not accepted via SMTP or submitted
+        locally by sendmail.
+
+    \item The mail will not have been tagged as a tracked mail (see section
+        \ref{tracking-re-injected-mail}).
+
+    \item The message-id has a specific format: \newline
+        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
+        e.g. \texttt{20070321125732.D168138A1@smtp.cs.tcd.ie}
+
+        The message-id is generated by Postfix for this bounce
+        notification, so the queueid in the message-id must be the same as
+        the queueid of the mail: this is what distinguishes bounce
+        notifications generated locally from bounce notifications which are
+        being forwarded inwards as a result of aliasing.  In the later case
+        the message-id will be unchanged, and so even if it happens to be
+        in the correct format (e.g. if it was generated by Postfix on
+        another server) the queueid in the message-id will not match the
+        queueid of the mail.
+
+\end{enumerate}
+
+Once a mail has been identified as a bounce notification the unknown origin
+flag is cleared and the mail can be entered in the database.
+
+There is a small chance that a mail will be incorrectly identified as a
+bounce notification, as the heuristics used may be too broad.  For this to
+occur the following conditions would have to be met:
+
+\begin{enumerate}
+
+    \item The mail must have been generated internally by Postfix.
+
+    \item The sender address must be <>.
+
+    \item The message-id must have the correct format and match the queueid
+        of the mail.  Whist it is possible that a mail sent from elsewhere
+        could easily have the correct message-id format, the chance that
+        the queueid in the message-id would match the queueid of the mail
+        is extremely small.
+
+\end{enumerate}
+
+The most likely cause of mis-identification is if a mail generated
+internally by Postfix is identified as a bounce notification when it is a
+different type of message; arguably this is a benefit rather than a
+drawback, as future mails generated internally by Postfix will be handled
+correctly.  XXX FIND OUT IF DSNS FIT THE PATTERN.
+
+\subsubsection{Aborted delivery attempts}
+
+\label{aborted-delivery-attempts}
+
+\subsubsection{Timeouts during DATA phase}
+
+\label{timeouts-during-data-phase}
+
+\subsubsection{Out of order log entries}
+
+\label{out-of-order-log-entries}
+
 \section{Parsing coverage}

### 2007-05-16 doc/logparser.tex
First attempt at describing timeout handling.  It'll need proofreading and editing for sure, but it's a start. Aborted delivery attempts section is written, but again will need work I'm sure. Minor cleanup on recent work.

@@
 
-\subsection{XXX}
+\subsection{Additional complications}
 
@@
 bounce notifications do not eventually have a log line which identifies
-their source.  The solution is that heuristics must be used to identify
-them, and those heuristics are:
+their source.  Heuristics must be used to identify bounce notifications,
+and those heuristics are:
 
@@
 
-    \item The sender address will be <>.
-
-    \item Neither \texttt{postfix/smtpd} nor \texttt{postfix/pickup} will
-        have logged any messages associated with the mail, viz it was
-        generated internally by Postfix, not accepted via SMTP or submitted
-        locally by sendmail.
+    \item The sender address will be $<>$.
+
+    \item Neither smtpd nor pickup will have logged any messages associated
+        with the mail, indicating it was generated internally by Postfix,
+        not accepted via SMTP or submitted locally by sendmail.
 
@@
 
-    \item The sender address must be <>.
+    \item The sender address must be $<>$.
 
@@
 
+Some mail clients appear to send the following sequence of commands during
+the SMTP session:
+
+\begin{verbatim}
+    EHLO client.hostname
+    MAIL FROM: <sender@address>
+    RCPT TO: <recipient@address>
+    RSET
+    MAIL FROM: <sender@address>
+    RCPT TO: <recipient@address>
+    DATA
+\end{verbatim}
+
+The client aborts the first delivery attempt after the first recipient is
+accepted, then makes a second delivery attempt which it continues with
+until the delivery is complete. XXX VERIFY THIS AND DETERMINE WHAT CLIENTS
+EXHIBIT THIS BEHAVIOUR XXX
+
+Once again Postfix does not log a message making the client's behaviour
+clear, so once again heuristics are required to identify the when this
+behaviour occurs.  In this case a list of all mails accepted during a
+connection are saved in the connection state, and the accepted mails
+examined when the disconnection line is parsed and the disconnection action
+executed.  Each mail is checked for the following characteristics:
+
+\begin{itemize}
+
+    \item Is the mail missing its cleanup log message?  Every mail which
+        passes through Postfix will have a cleanup line; lack of a cleanup
+        line is a sure sign the mail didn't make it too far.
+
+    \item Were there exactly two smtpd log lines for the mail?  There
+        should be a connection line and a mail accepted line.
+
+\end{itemize}
+
+If both checks are successful then the mail is assumed to be one of the
+offending bogus mails and is discarded.  There will be no further entries
+logged for such mails, so without identifying and discarding them they
+accumulate in the state table and will cause clashes if the queueid is
+reused.
+
 \subsubsection{Timeouts during DATA phase}
@@
 
-\subsubsection{Out of order log entries}
-
-\label{out-of-order-log-entries}
+The DATA phase of the SMTP conversation is where the headers and body of the
+mail are transferred.  Sometimes there is a timeout or the connection is
+lost\footnote{For brevity's sake timeout will be used throughout this
+section, but everything applies equally to lost connections.} during the
+DATA phase; when this occurs Postfix will discard the mail and the parser
+needs to discard the data associated with that mail.  It seems more
+intuitive to save the data to the database, but if a timeout occurs there
+won't yet have been any data gathered for the mail, so there is none
+available to save; the timeout is saved with the connection data instead.
+
+To deal properly with timeouts the parsing algorithm needs to do the
+following in the timeout action:
+
+\begin{enumerate}
+
+    \item Record the timeout and associated data in the connection's
+        results.
+
+    \item If no mails have been accepted yet nothing needs to be done; the
+        timeout action ends.  The timeout action is dependant on the clone
+        action keeping a list of all mails accepted on each connection.
+
+    \item Because of ESMTP pipelining the client can send MAIL FROM, RCPT
+        TO and DATA commands in one packet, instead of sending each command
+        individually and waiting for the reply before sending subsequent
+        commands.  If pipelining is used and Postfix rejects the sender
+        address or any of the recipient addresses there is no way for the
+        client to tell which command was rejected, nor does there appear to
+        be a way for the SMTP conversation to recover, so clients
+        disconnect uncleanly.
+
+        A timeout may thus apply either to an accepted mail, or a rejected
+        mail.  To distinguish between the two cases the algorithm compares
+        the timestamp of the last accepted mail against the timestamp of
+        the last line logged by smtpd for that connection.  If the smtpd
+        timestamp is later there was a rejection between the accepted mail
+        and the timeout, therefore the timeout applies to a rejected mail;
+        the timeout has already been recorded so the timeout action
+        finishes.  If the mail acceptance timestamp is greater then the timeout
+        applies to the just-accepted mail.
+
+    \item The mail needs to be discarded, but there is yet another
+        complication: sometimes the cleanup line is logged after the
+        timeout line.  If the cleanup line has already been seen the mail
+        can be discarded (both from global state and the connection's list
+        of accepted mails) and the timeout action finishes.
+
+        In the case where the cleanup line is still pending the algorithm
+        updates a global cache of queueids, adding the queueid and the
+        timestamp from the timeout line.  When the next cleanup line is
+        parsed for that queueid the cache will be checked, and will be
+        deemed part of the discarded mail and ignored if it meets the
+        following requirements:
+
+        \begin{itemize}
+
+            \item The queueid must not have been reused yet, i.e. there
+                isn't an entry in the state tables for the queueid.
+
+            \item The timestamp of the cleanup line must be within six
+                minutes of the mail acceptance timestamp.  Timeouts happen
+                after five minutes, so six gives a one minute leeway.
+
+        \end{itemize}
+
+        The next cleanup line must meet the criteria above for it to be
+        discarded because not every connection where a timeout occurs will
+        have a cleanup line logged for it; if the algorithm blindly
+        discarded the next cleanup line it would in some cases be mistaken.
+        Whether or not the line is discarded the queueid will be removed
+        from the cache of timeout queueids.
+
+\end{enumerate}
+
+\subsubsection{Pickup logging after cleanup}
+
+\label{pickup-logging-after-cleanup}
+
+Occasionally the pickup line logged when mail is submitted locally via
+sendmail appears later in the logfile than the cleanup line for that mail.
+This seems to occur during periods of particularly heavy load, so is most
+likely due to process scheduling vagaries.  Normally if the queueid given
+in the pickup line exists a warning is generated by the pickup action, but
+if the following conditions are met it is assumed that the lines are out of
+order:
+
+\begin{itemize}
+
+    \item The only program which logged anything for the mail is cleanup.
+
+    \item There is less than a five second difference in the timestamps of
+        the cleanup and pickup lines.
+
+\end{itemize}
+
+As always with heuristics there may be circumstances in which these
+heuristics incorrectly match, but none have been identified so far.
 

### 2007-05-24 doc/logparser.tex
Documented actions and identifying more aborted mails (arising as a result of restriction_start changes).

@@
 
-    \item [description] A longer, more detailed description of the rule.
-
-    \item [program] The program (smtp, smtpd, qmgr, etc.) whose log lines
-        the rule applies to.  This avoids needlessly trying rules which
-        won't match the line, or worse, might match unintentionally.
+    \item [description] Something must have occurred to cause Postfix to
+        log each line (e.g. a remote client connecting causes a connection
+        line to be logged).  The description field is generally used to
+        describe the action causing the log lines this rule matches.
 
@@
         \texttt{rejected}, other rules will have an empty string.
+
+    \item [program] The program (postfix/smtp, postfix/smtpd, postfix/qmgr,
+        etc.) whose log lines the rule applies to.  This avoids needlessly
+        trying rules which won't match the line, or worse, might match
+        unintentionally.
 
@@
 
-XXX USE THE SAME DOCUMENTATION HERE AS IN Parser.pm
-
-This list contains all the actions available in the algorithm, in the order
-in which they tend to occur in practice.
+This list contains all the actions available in the algorithm.  Each action
+is passed the same arguments:
 
@@
 
-    \item [ignore] Do nothing with this line, merely parse it so that
-        unparsed lines can be reported.
-
-    \item [connect] Handle a connection from a remote client.  Complains if
-        a connection already exists for this client
-
-    \item [restriction\_start]
-
-    \item [save\_by\_pid]
-
-    \item [clone]
-
-    \item [disconnect]
-
-    \item [save\_by\_queueid]
-
-    \item [qmgr\_chooses\_mail] 
-        however qmgr selects every mail for delivery, so most of the time
-        this action merely falls through to saving the information from the
-        log line (action: save\_by\_queueid; flowchart:~10).
-
-    \item [track]
-
-    \item [commit]
-
-    \item [pickup]
+    \item [\$line] The log line, with the standard syslog(3) fields parsed,
+        removed and made available.
+
+    \item [\$rule] The matching rule
+
+    \item [@matches] The fields in the line captured by \$rule's regex.
+
+\end{description}
+
+\begin{description}
+
+    \item [IGNORE] IGNORE just returns successfully; it is used when a line
+        needs to be parsed for completeness but doesn't either provide any
+        useful data or require anything to be done.
+
+    \item [CONNECT] Handle a remote client connecting: create a new state
+        table entry (indexed by smtpd pid) and save both the client
+        hostname and IP address.
+
+    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
+        connection in the database, perform any required cleanup, and
+        delete the connection from the state tables.
+
+    \item [SAVE\_BY\_PID] Use the pid from \$line to find the correct
+        connection and call \$self->save() with the appropriate arguments -
+        see save() in SUBROUTINES for more details. If the connection
+        doesn't exist a connection marked faked will be created and a
+        warning issued.
+
+    \item [SAVE\_BY\_QUEUEID] Use the queueid from \$rule and @matches to
+        find the correct connection and call \$self->save() with the
+        appropriate arguments - see save() in SUBROUTINES for more details.
+        If the connection doesn't exist a connection marked faked will be
+        created and a warning issued.
+
+    \item [COMMIT] Enter the data into the database. Entry may be postponed
+        if the mail is a child waiting to be tracked.
+
+    \item [TRACK] Track a mail when it is forwarded to another mail server;
+        this happens when a local address is aliased to a remote address.
+        TRACK will be called when dealing with the parent mail, and will
+        create the child mail if necessary. TRACK checks if the child has
+        already been tracked, either with this parent or with another
+        parent, and issues appropriate warnings in either case. Tracking
+        children is discussed extensively in the paper written about this
+        parser; details about obtaining the paper are given in the SEE ALSO
+        section.
+
+    \item [REJECTION] Deal with postfix rejecting an SMTP command from the
+        remote client: log the rejection with the accepted mail if there is
+        one, otherwise log it with the connection.
+
+    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
+        picking a mail from the queue to deliver. This action is used for
+        both pickup and cleanup due to out of order log lines.
+
+    \item [PICKUP] Pickup is the service which deals with mail submitted
+        locally via /usr/sbin/sendmail. This action creates a new state
+        table entry and saves data to it, unless out of order logging has
+        caused the cleanup line to be logged first. Lines are assumed to be
+        out of order if the only program seen thus far is cleanup and there
+        is less than five seconds difference between the timestamps of the
+        two lines.
+
+    \item [CLONE] Multiple mails may be accepted on a single connection, so
+        each time a mail is accepted the connection's state table entry
+        must be cloned; if the original data structure was used the second
+        and subsequent mails would corrupt the data structure.
+
+    \item [MAIL\_TOO\_LARGE] When the client tries to send a larger message
+        than the local server accepts the mail will be discarded.  See
+        TIMEOUT for further discussion; the two are handled in exactly the
+        same way.
+
+    \item [TIMEOUT] The connection timed out so the mail currently being
+        transferred must be discarded. The mail may have been accepted, in
+        which case there's a data structure to dispose of, or it may not in
+        which case there's none. The gory details can be found in the
+        internals documentation.
 
@@
 reused.
+
+\subsubsection{Further aborted delivery attempts}
+
+Some mail clients disconnect abruptly if a second or subsequent recipient
+is rejected; they may also disconnect after other errors, but such
+disconnections are already being handled.  Sadly Postfix doesn't log a
+message saying the mail has been discarded, as should be expected by now.
+The checks to identify this happening are:
+
+\begin{itemize}
+
+    \item Is the mail missing its cleanup log message?  Every mail which
+        passes through Postfix will have a cleanup line; lack of a cleanup
+        line is a sure sign the mail didn't make it too far.
+
+    \item Were there three or more smtpd log lines for the mail?  There
+        should be a connection line and a mail accepted line, followed by
+        one or more rejection lines.
+
+    \item Is the last smtpd log line a rejection line?
+
+\end{itemize}
+
+These checks are made during the DISCONNECT action: if all checks are
+successful then the mail is assumed to have been discarded when the client
+disconnected.  There will be no further entries logged for such mails, so
+without identifying and discarding them they accumulate in the state table
+and will cause clashes if the queueid is reused.
 

### 2007-05-27 doc/logparser.tex
Started my bibliography - unnecessarily hard in latex, I think.  It's nicely set up now though.
Worked through sections one and two adding bibliography references and doing some TODO items.

@@
 of Postfix logs at one time or another, whether it was to debug a problem,
-explain to a user why their mail is being rejected, or to check whether
-their new anti-spam measures are working.  The more adventurous will have
+explain to a user why their mail is being rejected, or check whether their
+new anti-spam measures are working.  The more adventurous will have
 generated statistics to show how many hits each of their anti-spam measures
@@
 requirement was discovered or new functionality was added.  Further
-complication arises as a result of the facility to have rejection messages
-defined by the administrator; every Real-time Black List returns a
-different explanatory message; policy servers may have different messages
-depending on the characteristics of the connection; there are many ways in
-which the log lines may differ.  This paper documents the difficult process
-of parsing Postfix logs, presenting a program which parses logs and places
-the resulting data into a database.  The gathered data can then be used to
-optimise current anti-spam measures, provide a baseline to test new
+complication because the list of rejection messages is not fixed: new
+messages can be added by the administrator; every Real-time Black List
+returns a different explanatory message; policy servers may have different
+messages depending on the characteristics of the connection; there are many
+ways in which the log lines may differ.  This paper documents the difficult
+process of parsing Postfix logs, presenting a program which parses logs and
+places the resulting data into a database.  The gathered data can then be
+used to optimise current anti-spam measures, provide a baseline to test new
 anti-spam measures against, or to produce statistics showing how effective
@@
 
-Section~\ref{background} provides background information about: the
-motivation for this project; the Postfix Mail Transport Agent; a short
-overview of the parser; the assumptions inherent in the design of the
-algorithm and program; using a database schema as an API\@; and a
-comparison against other Postfix log parsers.
+Section~\ref{background} provides background information about the
+following: the motivation for this project; the Postfix Mail Transport
+Agent; a short overview of the parser; the assumptions inherent in the
+design of the algorithm and program; using a database schema as an API\@;
+and a comparison against other Postfix log parsers.
 
@@
 
-Section~\ref{conclusion} describes the results of the project to date.  XXX
-IMPROVE THIS\@.
+Section~\ref{conclusion} describes the results of the project to date and
+contains the paper's conclusion.
 
@@
 If the database supports sub-selects percentages can be
-obtained:\footnote{This works for SQLite3, where || is the concatenation
-operator.  If the database containing the data does not support
-this syntax, simply remove ``|| '\%\''' from the query; the results will be
-the same, just slightly less visually pleasing.}
-
-XXX STOP THIS FROM BEING SPLIT ACROSS PAGES IF POSSIBLE
+obtained:
 
@@
 
-Another example is determining which restrictions are not effective:
+SQL note: $||$ is the concatenation operator in SQLite3; if the database
+containing the extracted data does not support this syntax, then simply
+remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the results will be
+the same, just slightly less visually pleasing.
+
+Another example is determining which restrictions are not effective: this
+example shows which rules had less than 100 hits on the last log file
+parsed.
 
@@
     \item Libraries exist for the majority of programming languages
-        allowing access to databases, whereas if this program was written
+        allowing access to databases, whereas if this parser was written
         as a library a new interface layer would need to be written for
-        every programming language used with the API\@.
+        every programming language using API\@.
 
@@
 Postfix is a highly configurable, high performance, secure and scalable
-Mail Transport Agent.  It features extensive anti-spam restrictions,
-allowing a mail administrator to deploy those restrictions which they judge
-suitable for their needs, rather than a fixed set chosen by Postfix's
-author.  These restrictions can be selectively applied, combined and
-bypassed on a per-client, per-recipient or per-sender basis, allowing
-varying levels of defense and or permissiveness.  Postfix leverages simple
-lookup tables to support arbitrarily complicated user-defined sequences of
-restrictions and exceptions, with policy servers as the ultimate in
-flexibility.  Administrators can also supply their own rejection messages
-to make it clear to senders why exactly their mail was rejected.
+Mail Transport Agent.  It features extensive optional anti-spam
+restrictions, allowing an administrator to deploy those restrictions which
+they judge suitable for their site's needs, rather than a fixed set chosen
+by Postfix's author.  These restrictions can be selectively applied,
+combined and bypassed on a per-client, per-recipient or per-sender basis,
+allowing varying levels of defense and/or permissiveness.  Postfix
+leverages simple lookup tables to support arbitrarily complicated
+user-defined sequences of restrictions and exceptions, with policy
+servers\footnote{Policy servers will be explained shortly.} as the ultimate
+in flexibility.  Administrators can also supply their own rejection
+messages to make it clear to senders why exactly their mail was rejected.
 Unfortunately this flexibility has a cost: complexity in the logs
@@
 
-A policy server is an external program that accepts state information from
-Postfix for each SMTP command not rejected by an earlier restriction; the
-policy server can utilise that state information to implement whatever
-logic is required.  E.g.\ some users can be restricted to sending mail on
-the third Tuesday after pay day only --- this example may actually be
-useful in a payroll system to prevent spam or (worse) phishing mails with
-faked sender addresses.  More commonly encountered scenarios are:
+\subsubsection{Policy servers}
+
+A policy server \cite{policy-servers} is an external program that accepts
+state information from Postfix for each SMTP command not rejected by an
+earlier restriction; the policy server can utilise that state information
+to implement whatever logic is required.  E.g.\ some users can be
+restricted to sending mail on the third Tuesday after pay day only --- this
+example may actually be useful in a payroll system to prevent spam or
+(worse) phishing mails with faked sender addresses.  More commonly
+encountered scenarios are:
 
@@
 
-    \item Checking Sender Policy Framework records.  SPF records specify
-        which mail servers are allowed to send mail claiming to be from a
-        particular domain.  The intention is to reduce faked sender
-        addresses, which will in turn reduce spam, backscatter and
-        joe-jobs; however there has been a lot of resistance to the
-        proposal because it breaks or vastly complicates some features of
-        SMTP.  For more details see \url{http://www.openspf.org/} or
-        \url{http://en.wikipedia.org/wiki/Sender_Policy_Framework} XXX ADD
-        REFERENCES FOR SPAM BACKSCATTER AND JOE-JOBS TO THE APPENDIX.
-
-    \item Greylisting, a technique that temporarily rejects mail where the
-        triple of (sender, recipient, remote IP address) is unknown; the
-        assumption is that it's uneconomical for a spammer to retry, but
-        that a legitimate mail server must retry.  Sadly spammers are using
-        increasingly complex and well written programs to distribute spam,
-        frequently using an ISP provided SMTP server from a compromised
-        machine on its network, so greylisting is slowly becoming less
-        useful, but it does block a large percentage of spam mail at the
-        moment.  See \url{http://www.greylisting.org/} for full details.
-
-    \item Using a scoring system such as Policyd-weight where 
-        tests accumulate points against the sending system --- if the
-        eventual score is too high the mail is rejected.
-        \url{http://www.policyd-weight.org/}
-
-    \item Policyd performs the following: \textit{Greylisting,
-        Sender-(envelope, SASL or host / ip)-based throttling (on messages
-        and/or volume per defined time unit), Recipient rate limiting,
-        Spamtrap monitoring / blacklisting, HELO auto blacklisting and HELO
-        randomization preventation.\/}  \url{http://policyd.sourceforge.net}
+    \item Checking Sender Policy Framework records \cite{openspf},
+        \cite{wikipedia-spf}.  SPF records specify which mail servers are
+        allowed to send mail claiming to be from a particular domain.  The
+        intention is to reduce spam from faked sender addresses,
+        backscatter \cite{postfix-backscatter} and joe-jobs
+        \cite{wikipedia-joe-job}; however there
+        has been a lot of resistance to the proposal because it breaks or
+        vastly complicates some features of SMTP.
+
+    \item Greylisting \cite{greylisting}, a technique that temporarily
+        rejects mail where the triple of (sender, recipient, remote IP
+        address) is unknown; the assumption is that it's uneconomical for a
+        spammer to retry, but that a legitimate mail server must retry.
+        Sadly spammers are using increasingly complex and well written
+        programs to distribute spam, frequently using an ISP provided SMTP
+        server from a compromised machine on its network.  Greylisting is
+        slowly becoming less useful, but it does block a large percentage
+        of spam mail at the moment.
+
+    \item Using a scoring system such as Policyd-weight
+        \cite{policyd-weight} where tests accumulate points against the
+        sending system --- if the eventual score is too high the mail is
+        rejected.
+
+    \item Rate limiting or throttling on a per-sender, per-client or
+        per-recipient basis as performed by Policyd \cite{policyd}.
 
@@
 
-Example attributes taken from \newline
-\url{http://www.postfix.org/SMTPD_POLICY_README.html}:
+Example attributes taken from \cite{policy-servers}:
 
@@
 whole, saving it in a useful and consistent form, and performing
-housekeeping duties.
-
-XXX REFER TO THE GRAMMAR COMPARISON SECTION ONCE ITS WRITTEN.
+housekeeping duties.  Readers familiar with context-free grammars may find
+Section \ref{comparison against context-free grammars} helpful.
 
@@
 
-XXX DO I NEED TO SAY MORE??
-
 \section{Adaptable parsing: user defined rules}
@@
 
-XXX COMPARE AND CONTRAST WITH CONTEXT FREE GRAMMARS, OTHER GRAMMARS.
+\subsection{Comparison against context-free grammars}
+
+\label{comparison against context-free grammars}
 
@@
 
-            \item [save\_by\_pid] Use the pid from the log line to lookup a
-                connection, then save the data from the log line in the
-                connection data structure thus found.
-
             \item [commit] The mail is finished with, enter it in the
@@
 REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
-HOW THE RULE MATCHES, AND SO ON\@.
+HOW THE RULE MATCHES, AND SO ON\@.  SAVE\_BY\_PID IS GONE NOW, USE A
+DIFFERENT EXAMPLE.
 
@@
 
+% Bibliography.
+\bibliographystyle{logparser-bibliography-style}
+\bibliography{logparser-bibliography}
+
 \end{document}

### 2007-05-29 doc/logparser.tex
Added cached vs discarded graphs for regex run time. Added Makefile rules to create the graphs. Improved related paper section.

@@
     \item The first matching rule wins: no further rules are tried against
-        that line,\footnote{With one exception, details are provided later
-        in Section~\ref{actions-in-detail}.} but there is a facility for
-        the specifying the order of the rules so that more specific rules
-        can be tried first.
+        that line, but there is a facility for specifying the order of
+        rules so that more specific rules can be tried first.
 
@@
         this simplifies reading and writing rules, avoids needless
-        repetition of complex regex components, and allows the components
-        to be corrected and/or improved in one location.
-
-        The following keywords are expanded:
-
-        \texttt{\_\_SENDER\_\_, \_\_RECIPIENT\_\_, \_\_HELO\_\_,
-        \_\_HOSTNAME\_\_, \_\_IP\_\_, \newline\_\_SMTP\_CODE\_\_,
-        \_\_QUEUEID\_\_ and \_\_COMMAND\_\_.}
-
-        \_\_COMMAND\_\_ is an SMTP command, e.g. MAIL FROM\@; hopefully the
-        others are reasonably self explanatory.
+        repetition of complex regex components, allows the components
+        to be corrected and/or improved in one location, and makes each
+        regex largely self-documenting.
+
+        The following keywords are expanded (full explanations can be found
+        in the source code):
+
+        \_\_SENDER\_\_, \_\_RECIPIENT\_\_, \_\_MESSAGE\_ID\_\_,
+        \_\_HELO\_\_, \newline \_\_EMAIL\_\_, \_\_HOSTNAME\_\_, \_\_IP\_\_,
+        \_\_IPv4\_\_, \_\_IPv6\_\_, \newline \_\_SMTP\_CODE\_\_,
+        \_\_RESTRICTION\_START\_\_, \_\_QUEUEID\_\_, \newline
+        \_\_COMMAND\_\_, \_\_SHORT\_CMD\_\_, \_\_DELAYS\_\_, \_\_DELAY\_\_,
+        \_\_DSN\_\_ and \_\_CONN\_USE\_\_.
 
@@
         compiled before attempting to parse the log file --- otherwise each
-        regex would be recompiled each time it was used, resulting in an
-        468.95\% slowdown.
-
-        Averaged over 10 test runs:
-
-        \begin{description}
-
-            \item [Compiled once and saved:] run time was 55.47 seconds,
-                standard deviation was 0.247 seconds.
-
-            \item [Compiled every time and discarded:] run time was 260.14
-                seconds (4 minutes, 20.14 seconds) (468.95\% slower),
-                standard deviation was 0.96 seconds.
-
-        \end{description}
-
-        Logging to the database was disabled for the test runs.
-
-        XXX RERUN OVER MORE LOG FILES
-
-        XXX DO I NEED TO SAY ANYTHING MORE ABOUT THIS??
+        regex would be recompiled each time it was used, resulting in a
+        large, data dependent slowdown.  Graph \ref{normal regex vs discard
+        regex} shows execution times gathered over 93 log files, with and
+        without caching the regex; also shown on the graph are the file
+        sizes and number of lines in each file.  Graph \ref{normal regex vs
+        disarded regex factor} shows the percentage execution time increase
+        when not caching each regex.  Logging to the database was disabled
+        for the test runs.
 
@@
 
+\label{conclusion}
+
 \appendix
-
-\label{conclusion}
 
@@
 
+\section{Graphs}
+
+\subsection{Regex: cached vs compiled every time}
+\label{normal regex vs discard regex}
+\includegraphics{plot-time.ps}
+
+\subsection{Regex: percentage execution time increase}
+\label{normal regex vs disarded regex factor}
+\includegraphics{plot-factor.ps}
+
 \end{document}

### 2007-05-29 doc/logparser.tex
Added normal vs shuffle vs reverse graph. Renamed plot-* and stats-* to more meaningful names. Updated Makefile appropriately.

@@
         log line will be extracted.  The format is: \newline
-        \texttt{client\_hostname = 1; recipient = 2, sender = 4;} \newline
+        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline
         i.e.\ semi-colon or comma separated assignment statements, with the
@@
         time.
+\ref{normal vs shuffled vs reversed ordering}
 
@@
 \label{normal regex vs discard regex}
-\includegraphics{plot-time.ps}
+\includegraphics{plot-cached-discarded.ps}
 
@@
 \label{normal regex vs disarded regex factor}
-\includegraphics{plot-factor.ps}
+\includegraphics{plot-cached-discarded-factor.ps}
+
+\subsection{Normal vs shuffled vs reversed ordering}
+\label{normal vs shuffled vs reversed ordering}
+\includegraphics{plot-normal-shuffle-reverse.ps}
 

### 2007-05-29 doc/logparser.tex
Added new graph - percentage increase of shuffled over normal and reversed over normal. Removed file sizes and lines from normal vs shuffled vs reversed. Appropriate Makefile changes, plus minor improvements. Documented averaging of run times and discarding of first run.

@@
         disarded regex factor} shows the percentage execution time increase
-        when not caching each regex.  Logging to the database was disabled
-        for the test runs.
+        when not caching each regex.  Each log file was parsed 10 times,
+        the first run discarded and the remaining 9 runs averaged.  Logging
+        to the database was disabled for the test runs.
 
@@
         updates every rule's rule\_order.  Assuming that the distribution
-        of log lines is reasonably consistent, rules matching more commonly
-        occurring log lines will be tried before rules matching less
-        commonly occurring log lines, lowering the program's execution
-        time.
-\ref{normal vs shuffled vs reversed ordering}
-
-        Averaged over 10 test runs:
-
-        \begin{description}
-
-            \item [Ascending order (best):] run time was 56.684 seconds,
-                standard deviation was 0.29 seconds.
-
-            \item [Descending order (worst):] run time was 64.037
-                seconds (12.97\% slower), standard deviation was 0.16
-                seconds.
-
-            \item [Shuffled order:] run time was 57.19 seconds
-                (0.89\% slower), standard deviation was 0.95 seconds.
-
-        \end{description}
-
-        XXX RERUN OVER MORE LOG FILES
-
-        XXX RERUN AGAIN ONCE POSTFIX 2.3 RULES HAVE BEEN ADDED TO SEE WHAT
-        THE DIFFERENCE IS.  THERE ARE 110 POSTFIX 2.2 RULES AS OF
-        2007/04/18
-
-        Logging to the database was disabled for the test runs.
+        of log lines is reasonably consistent between log files, rules
+        matching more commonly occurring log lines will be tried before
+        rules matching less commonly occurring log lines, lowering the
+        program's execution time.  Graph \ref{normal vs shuffled vs
+        reversed ordering} plots the execution time over 93 log files for
+        normal, shuffle and reverse ordering.  Graph \ref{normal vs
+        shuffled vs reversed ordering factor} shows the percentage
+        execution time increase for shuffled and reversed over normal.
+        Each log file was parsed 10 times, the first run discarded and the
+        remaining 9 runs averaged.  Logging to the database was disabled
+        for the test runs.
 
         It is hoped that this will prove to have greater effectiveness as
-        the number of rules grows; currently the effect is negligible,
-        particularly in comparison to compiling each regex once.
+        the number of rules grows; currently the effect is negligible
+        in comparison to caching each compiled regex, but it still provides
+        approximately a 10\% performance increase.
 
@@
 
+\subsection{Percentage increase of shuffled over normal and reversed over normal}
+\label{normal vs shuffled vs reversed ordering factor}
+\includegraphics{plot-normal-shuffle-reverse-factor.ps}
+
 \subsection{Normal vs shuffled vs reversed ordering}

### 2007-05-30 doc/logparser.tex
Corrected and improved flow chart. Added graphs, hopefully clarified graphs too. Improved paper with clearer wording. Described field usage in example rule. Expanded Makefile to cover new graphs. Stupidly deleted everything, had to redo the days work.

@@
         in comparison to caching each compiled regex, but it still provides
-        approximately a 10\% performance increase.
+        approximately a 7\% -- 10\% performance increase.
 
@@
         rule\_order: rules with a higher priority will be tried first,
-        overriding rule\_order, allowing more specific rules to take
+        overriding rule\_order.  This allows more specific rules to take
         precedence over more general rules.
@@
 
-XXX CARL SUGGESTS EXPANDING ON THIS SOME MORE, BUT I DON'T KNOW HOW\@.
-PERHAPS WHEN THE ALGORITHM IS PROPERLY EXPLAINED IT'LL BE EASIER\@?  MAKE
-REFERENCES TO THE ALGORITHM, EXPLAIN MORE CLEARLY HOW THE FIELDS ARE USED,
-HOW THE RULE MATCHES, AND SO ON\@.  SAVE\_BY\_PID IS GONE NOW, USE A
-DIFFERENT EXAMPLE.
-
 % Don't reformat this!
@@
 
+\textbf{Field}      & \textbf{Value}                                    \\
 name                & Unknown sender domain                             \\
 description         & We do not accept mail from unknown domains        \\
+restriction\_name   & reject\_unknown\_sender\_domain                   \\
+postfix\_action     & REJECTED                                          \\
 program             & postfix/smtpd                                     \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-regex               & \verb!^<(__SENDER__)>: Sender address rejected: ! \\
-                    & \verb!  Domain not found; from=<\1> !             \\
-                    & \verb!  to=<(__RECIPIENT__)> proto=E?SMTP !       \\
-                    & \verb!  helo=<(__HELO__)>$!                       \\
-result\_cols        & recipient = 2; sender = 1                         \\
-connection\_cols    & helo = 3                                          \\
+regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
+                    & \verb!Sender address rejected: Domain not found;! \\
+                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
+                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
+result\_cols        & recipient = 6; sender = 5                         \\
+connection\_cols    & helo = 7                                          \\
 result\_data        &                                                   \\
 connection\_data    &                                                   \\
-postfix\_action     & REJECTED                                          \\
-action              & SAVE\_BY\_PID                                     \\
-queueid             & 0                                                 \\
+action              & REJECTION                                         \\
+queueid             & 1                                                 \\
 rule\_order         & 0                                                 \\
@@
 \end{tabular}
+
+The various fields are used as follows;
+
+\begin{description}
+
+    \item [name, description, restriction\_name and postfix\_action:] are
+        not \newline used by the algorithm, they serve to document the rule.
+
+    \item [program and regex:] If the program equals the program which
+        logged the line the regex will be matched against the line; if
+        successful the action will be called, if not the next rule will be
+        tried.
+
+    \item [action:] will be called if the regex matches successfully (see
+        Section \ref{actions-in-detail} for full details).
+
+    \item [result\_cols, connection\_cols, result\_data and
+        connection\_data:] are \newline used by the action to extract and
+        save data matched by the regex.
+
+    \item [queueid:] is used the extract the queueid matched by the regex,
+        so that mails can be found by queueid and actions performed on the
+        mail.
+
+\end{description}
 
@@
         in the parser, unknown origin mails need to be marked as such
-        because they lack some data present in the parent mail, and must
+        because they lack some data present in the parent mail; they must
         copy that data from their parent before being entered in the
-        database.  XXX EITHER REMOVE THE LAST LINE OR EXPAND ON IT SOME
-        MORE\@.  FIGURE OUT WHY MORE DATA ISN'T REQUIRED\@.
+        database.
 
@@
 except in the minority of cases where intra-mail interference is required.
-I felt that the best way to implement this was to maintain state
-information for every unfinished mail and manipulate the appropriate mail
-correctly for each log line encountered.  The parser thus requires both a
-method of mapping log lines to the correct mail and a method of specifying
-the action the log line represents.  The former is achieved by using the
-pid of the smtpd to identify the correct mail during the initial phase,
-then switching to the queueid once the mail has been accepted.  The latter
-uses the action field of the rule which matched the log line, executing the
-code in the function named by the action.  From a high level viewpoint this
-design shares significant similarity with how an editor maintains state for
-multiple files and performs the actions the user initiates on the correct
-file.
+The best way to implement this is to maintain state information for every
+unfinished mail and manipulate the appropriate mail correctly for each log
+line encountered.  The parser thus requires both a method of mapping log
+lines to the correct mail and a method of specifying the action the log
+line represents.  The former is achieved by using the smtpd process id to
+identify the correct mail during the initial phase, then switching to the
+queueid once the mail has been accepted.  The latter uses the action field
+of the rule which matched the log line, executing the code in the function
+named by the action.  From a high level viewpoint this design shares
+significant similarity with how an editor maintains state for multiple
+files and performs the actions the user initiates on the correct file.
 
@@
 drawback, as future mails generated internally by Postfix will be handled
-correctly.  XXX FIND OUT IF DSNS FIT THE PATTERN.
+correctly.
 
@@
 
+XXX CHANGE REFERENCES TO BIBLIOGRAPHY??  MERGE PREVIOUS SECTIONS??  PUT
+BIBLIOGRAPHY IN ITS OWN SECTION??
+
 % Bibliography.
@@
 
+\subsection{execution time vs file size vs number lines}
+\label{execution time vs file size vs number lines}
+\includegraphics{plot-normal-filesize-numlines.ps}
+
+\subsection{execution time vs file size vs number lines factor}
+\label{execution time vs file size vs number lines factor}
+\includegraphics{plot-normal-filesize-numlines-factor.ps}
+
 \subsection{Regex: cached vs compiled every time}

### 2007-05-30 doc/logparser.tex
Fixed stupid bugs - I didn't fully reproduce the work I deleted earlier.

@@
 
-\subsection{execution time vs file size vs number lines}
+\subsection{Execution time vs file size vs number of lines.}
 \label{execution time vs file size vs number lines}
@@
 
-\subsection{execution time vs file size vs number lines factor}
+\subsection{Ratio of file size to execution time and number of lines to
+execution time}
 \label{execution time vs file size vs number lines factor}
@@
 
-\subsection{Regex: cached vs compiled every time}
+\subsection{Regex: cached vs discarded}
 \label{normal regex vs discard regex}
@@
 
-\subsection{Percentage increase of shuffled over normal and reversed over normal}
-\label{normal vs shuffled vs reversed ordering factor}
-\includegraphics{plot-normal-shuffle-reverse-factor.ps}
-
 \subsection{Normal vs shuffled vs reversed ordering}
@@
 
+\subsection{Percentage increase of shuffled over normal and reversed over
+normal}
+\label{normal vs shuffled vs reversed ordering factor}
+\includegraphics{plot-normal-shuffle-reverse-factor.ps}
+
 \end{document}

### 2007-06-04 doc/logparser.tex
logparser.tex: * Lots of small corrections and improvements throughout the text. * Lots of place-holder sections. * get the bibliography properly into the TOC.
logparser-flow-chart.dot: changes in wording.
logparser-bibliography.bib: added "Last checked" as a note to all entries.
Makefile: stop the make if latex or bibtex fail.

@@
 \renewcommand{\labelenumii}{\roman{enumii}:}
+\renewcommand{\refname}{Bibliography}
+\usepackage[section,numbib,nottoc]{tocbibind}
 
@@
 
-XXX FIGURE OUT WHEN TO USE MAIL AND WHEN TO USE CONNECTION\@; POSSIBLY STATE
-THAT THEY'RE MORE OR LESS EQUIVALENT
+XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS
 
@@
 requirement was discovered or new functionality was added.  Further
-complication because the list of rejection messages is not fixed: new
-messages can be added by the administrator; every Real-time Black List
-returns a different explanatory message; policy servers may have different
-messages depending on the characteristics of the connection; there are many
-ways in which the log lines may differ.  This paper documents the difficult
-process of parsing Postfix logs, presenting a program which parses logs and
-places the resulting data into a database.  The gathered data can then be
-used to optimise current anti-spam measures, provide a baseline to test new
-anti-spam measures against, or to produce statistics showing how effective
-those measures are.  There are numerous other uses for such data: improving
-server performance by identifying troublesome destinations and
-reconfiguring appropriately or dedicating transports to those destinations;
-identifying high volume uses (e.g.\ customer newsletters) and restricting
-those uses to off-peak times; as a base for billing customers on a shared
-server; etc.
-
-Section~\ref{background} provides background information about the
-following: the motivation for this project; the Postfix Mail Transport
-Agent; a short overview of the parser; the assumptions inherent in the
-design of the algorithm and program; using a database schema as an API\@;
-and a comparison against other Postfix log parsers.
+complication arises because the list of rejection messages is not fixed:
+new messages can be added by the administrator with custom checks; every
+Real-time Black List returns a different explanatory message; policy
+servers may have different messages depending on the characteristics of the
+connection; there are many ways in which the log lines may differ between
+servers.  This paper documents the difficult process of parsing Postfix
+logs, presenting a program which parses logs and places the resulting data
+into a database.  The gathered data can then be used to optimise current
+anti-spam measures, provide a baseline to test new anti-spam measures
+against, or to produce statistics showing how effective those measures are.
+There are numerous other uses for such data: improving server performance
+by identifying troublesome destinations and reconfiguring appropriately or
+dedicating transports to those destinations; identifying regular high
+volume uses (e.g.\ customer newsletters) and restricting those uses to
+off-peak times; detecting virus outbreaks which propagate via email; as a
+base for billing customers on a shared server.
+
+\vspace{1em}\noindent\textbf{Layout of the paper:}
+
+Section~\ref{background} provides background information useful in
+understanding the paper, parser and algorithm.
 
@@
 structure, giving an example rule and sample data it matches successfully
-against.
+against, followed by a discussion of rule efficiency concerns.
 
@@
 the full algorithm, followed by a comprehensive explanation of the
-different stages of the algorithm and the actions taken during the
-execution of the algorithm.
+different stages of the algorithm, the actions taken during the execution
+of the algorithm, and further complications whose solution completes the
+algorithm.
 
@@
 
-Section~\ref{references} catalogues resources used in developing the
-algorithm, writing the program and preparing this paper.  
-Any resources referred to in the paper are also given, plus some additional
-resources expected to be helpful in understanding Postfix, anti-spam
-techniques, or the paper.
+Section~\ref{other-parsers} discusses other parsers and why they were
+deemed unsuitable for the task.
+
+Section~\ref{bibliography} is the bibliography, containing the resources
+used in developing the algorithm, writing the program and preparing this
+paper.  Also listed are some additional resources expected to be helpful in
+understanding SMTP, Postfix, anti-spam techniques, or the paper.
+
+Section~\ref{graphs} contains graphs illustrating the topics discussed in
+section~\ref{rule efficiency}.
 
@@
 A short example of the optimisation possible using data from the database
-is optimal ordering of Postfix restrictions:
+is which Postfix restrictions reject the highest number of mails:
 
@@
         as a library a new interface layer would need to be written for
-        every programming language using API\@.
+        every programming language using the API\@.
 
@@
         the user without requiring large amounts of programming.  All
-        databases provide one or more programs, of variable complexity and
+        databases provide a program, of variable complexity and
         sophistication, which can be used for ad hoc queries with minimal
@@
 
+    \item SQL is reasonably standard and many people will already be
+        familiar with it; for those unfamiliar with it there are already
+        lots of readily available resources from which to learn.  Although
+        every vendor implements a different dialect of SQL, the basics are
+        the same everywhere.
+
 \end{enumerate}
+
+\subsection{Database schema}
+\label{database schema}
+
+XXX WRITE THIS
+
+\subsection{SMTP background}
+\label{smtp background}
+
+XXX WRITE THIS
 
@@
 combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of defense and/or permissiveness.  Postfix
+allowing varying levels of severity and/or permissiveness.  Postfix
 leverages simple lookup tables to support arbitrarily complicated
@@
 
+\subsubsection{Mixing and matching Postfix restrictions}
+
+XXX TO BE WRITTEN
+
 \subsubsection{Policy servers}
 
-A policy server \cite{policy-servers} is an external program that accepts
+A policy server~\cite{policy-servers} is an external program that accepts
 state information from Postfix for each SMTP command not rejected by an
@@
 restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent spam or
-(worse) phishing mails with faked sender addresses.  More commonly
+example may actually be useful in a payroll system to prevent problems from
+spam or (worse) phishing mails with faked sender addresses.  More commonly
 encountered scenarios are:
@@
 
-    \item Checking Sender Policy Framework records \cite{openspf},
-        \cite{wikipedia-spf}.  SPF records specify which mail servers are
-        allowed to send mail claiming to be from a particular domain.  The
-        intention is to reduce spam from faked sender addresses,
-        backscatter \cite{postfix-backscatter} and joe-jobs
-        \cite{wikipedia-joe-job}; however there
-        has been a lot of resistance to the proposal because it breaks or
-        vastly complicates some features of SMTP.
-
-    \item Greylisting \cite{greylisting}, a technique that temporarily
-        rejects mail where the triple of (sender, recipient, remote IP
-        address) is unknown; the assumption is that it's uneconomical for a
-        spammer to retry, but that a legitimate mail server must retry.
-        Sadly spammers are using increasingly complex and well written
-        programs to distribute spam, frequently using an ISP provided SMTP
-        server from a compromised machine on its network.  Greylisting is
-        slowly becoming less useful, but it does block a large percentage
-        of spam mail at the moment.
-
-    \item Using a scoring system such as Policyd-weight
-        \cite{policyd-weight} where tests accumulate points against the
-        sending system --- if the eventual score is too high the mail is
-        rejected.
+    \item Checking Sender Policy Framework
+        records~\cite{openspf},~\cite{wikipedia-spf}.  SPF records specify
+        which mail servers are allowed to send mail claiming to be from a
+        particular domain.  The intention is to reduce spam from faked
+        sender addresses, backscatter~\cite{postfix-backscatter} and
+        joe-jobs~\cite{wikipedia-joe-job}; however there has been a lot of
+        resistance to the proposal because it breaks or vastly complicates
+        some features of SMTP\@.
+
+    \item Greylisting~\cite{greylisting}, is a technique that temporarily
+        rejects mail when the triple of (sender, recipient, remote IP
+        address) is unknown; on the second and subsequent delivery attempts
+        from that triple the mail will be accepted.  The assumption is that
+        retrying after a temporary failure is uneconomical for a spammer,
+        but that a legitimate mail server must retry.  Sadly spammers are
+        using increasingly complex and well written programs to distribute
+        spam, frequently using an ISP provided SMTP server from a
+        compromised machine on the ISP's network.  Greylisting is slowly
+        becoming less useful, but it does block a large percentage of spam
+        mail at the moment.
+
+    \item Using a scoring system such as
+        Policyd-weight~\cite{policyd-weight} where tests accumulate points
+        against the sending system --- if the eventual score is too high
+        the mail is rejected.
 
     \item Rate limiting or throttling on a per-sender, per-client or
-        per-recipient basis as performed by Policyd \cite{policyd}.
+        per-recipient basis as performed by Policyd~\cite{policyd}.
 
@@
 
-Example attributes taken from \cite{policy-servers}:
+Example attributes taken from~\cite{policy-servers}:
 
@@
 which are applied to the lines.  Rules can be thought of as the Lex part of
-a Lex and YACC style parser; rules identify the line and return some data,
-which the algorithm (the YACC part) receives and performs the requested
-action.  Rules are solely concerned with identifying a line and extracting
-data from it, whereas the algorithm's task is to follow the journey each
-mail takes through Postfix, piecing the data together into a coherent
-whole, saving it in a useful and consistent form, and performing
+a Lex and YACC style parser: rules identify the line and return some data,
+which the algorithm (the YACC part) receives and uses when performing the
+requested action.  Rules are solely concerned with identifying a line and
+extracting data from it, whereas the algorithm's task is to follow the
+journey each mail takes through Postfix, piecing the data together into a
+coherent whole, saving it in a useful and consistent form, and performing
 housekeeping duties.  Readers familiar with context-free grammars may find
-Section \ref{comparison against context-free grammars} helpful.
+Section~\ref{comparison against context-free grammars} helpful.
 
@@
 
-    \item The logs are whole and complete; nothing has been removed, either
+    \item The logs are whole and complete: nothing has been removed, either
         deliberately or accidentally (e.g.\ log rotation gone awry, file
@@
 all are intended to perform a specific parsing and reporting task, rather
-than be a generic parser and leave generating reports from the data to
-other programs.  Some save the data extracted to a database but the
-majority discard all data once finished running, making historical analysis
-impossible.  The other parsers listed all produce a report of greater or
-lesser complexity and detail, whereas the program described here doesn't
-attempt to produce a report at all; that responsibility is deferred to a
-separate program, to be developed later.  The parsing algorithm and program
-described here are designed to enable much more detailed log analysis by
-providing a stable platform for subsequent programs to develop upon.
+than be a generic parser, extracting data and leaving generation of reports
+from the data to other programs.  Some parsers save the data extracted to a
+database but the majority discard all data once they have finished running,
+making historical analysis impossible.  The other parsers reviewed all
+produce a report of greater or lesser complexity and detail, whereas the
+program described here doesn't attempt to produce a report at all; that
+responsibility is deferred to a separate program, to be developed later.
+The parsing algorithm and program described here are designed to enable
+much more detailed log analysis by providing a stable platform for
+subsequent programs to develop upon.
+
+\subsection{Conventions used in the paper}
+
+The words \textit{connection\/} and \textit{mail\/} are often used
+interchangeably in this paper; in general the word used was chosen based on
+the context it appears in.
 
@@
 It also creates a clear separation of functionality: rules handle low level
-details of identifying and extracting data from a line, whereas the
-algorithm handles the higher level details of following the path a mail
+details of identifying log lines and extracting data from a line, whereas
+the algorithm handles the higher level details of following the path a mail
 takes through Postfix, assembling the required data, etc.
@@
 
-XXX DESCRIBE IT IN LEFT: RIGHT PRODUCTION TERMS, IF POSSIBLE.
+XXX DESCRIBE IT IN LEFT\@: RIGHT PRODUCTION TERMS, IF POSSIBLE\@.
 
@@
     \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g. a remote client connecting causes a connection
+        log each line (e.g.\ a remote client connecting causes a connection
         line to be logged).  The description field is generally used to
@@
         \texttt{rejected}, other rules will have an empty string.
+
+    \item [postfix\_action] This is the action Postfix must have taken to
+        generate this line, with two exceptions:
+
+        \begin{itemize}
+
+            \item [INFO] Represents an unspecified intermediate action that
+                the parser is not interested in per se, but which does log
+                useful information, supplementing other log lines.
+
+            \item [IGNORED] An action which is not only uninteresting in
+                itself, but which also provides no useful data.
+
+        \end{itemize}
+
+        Uninteresting lines are parsed so that any lines the parser isn't
+        capable of handling become immediately obvious errors.
 
@@
 
+        Additional fields are captured by \_\_RESTRICTION\_START\_\_, so
+        rules using it will start the fields in result\_cols,
+        connection\_cols, etc.\ at 5.
+
         For efficiency the keywords are expanded and every rule's regex is
@@
         regex would be recompiled each time it was used, resulting in a
-        large, data dependent slowdown.  Graph \ref{normal regex vs discard
-        regex} shows execution times gathered over 93 log files, with and
-        without caching the regex; also shown on the graph are the file
-        sizes and number of lines in each file.  Graph \ref{normal regex vs
-        disarded regex factor} shows the percentage execution time increase
-        when not caching each regex.  Each log file was parsed 10 times,
-        the first run discarded and the remaining 9 runs averaged.  Logging
-        to the database was disabled for the test runs.
+        large, data dependent slowdown.  Rule efficiency concerns are
+        discussed in section~\ref{rule efficiency}.
 
@@
 
-        \texttt{client\_hostname, client\_ip, server\_ip, server\_hostname,
-        \newline helo, sender, recipient, smtp\_code, data and child.}
+        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
+        \newline server\_hostname and helo.\newline} \texttt{result\_cols:
+        sender, recipient, smtp\_code, data \newline and child.}
 
@@
         data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for small amounts of data rather
-        than large, so dealing with escape sequences seemed unnecessary.}
-        is permitted on the right hand side of the assignment.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this line, with two exceptions:
-
-        \begin{itemize}
-
-            \item [info] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [ignored] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-        \end{itemize}
-
-        Uninteresting lines are parsed so that any lines the parser isn't
-        capable of handling become immediately obvious errors.
-
-    \item [action] The action the algorithm will take, e.g.
-        \begin{description}
-
-            \item [ignore] Ignore this line.  The simplest action, it is
-                required because every line from a program of interest to
-                the parser must be parsed properly; any unparsed line is
-                considered an error.
-
-            \item [commit] The mail is finished with, enter it in the
-                database, and remove the mail from memory.
-
-        \end{description}
-
-        A full list can be found in Section~\ref{actions-in-detail}.
-
-    \item [queueid] Specifies the match from the regex which gives the
-        queueid, or zero if the log line doesn't contain a queueid.
-
-    \item [rule\_order] This is an efficiency measure.  This counter is
+        cannot be used.  This is intended for use with small amounts of
+        data rather than large, so dealing with escape sequences seemed
+        unnecessary.} is permitted on the right hand side of the
+        assignment.
+
+    \item [action] The action the algorithm will take; a full list can be
+        found in Section~\ref{actions-in-detail}.
+
+    \item [queueid] Specifies the matching field from the regex which gives
+        the queueid, or zero if the log line doesn't contain a queueid.
+
+    \item [rule\_order] is an efficiency measure.  This counter is
         maintained for every rule and incremented each time the rule
@@
         rules matching less commonly occurring log lines, lowering the
-        program's execution time.  Graph \ref{normal vs shuffled vs
-        reversed ordering} plots the execution time over 93 log files for
-        normal, shuffle and reverse ordering.  Graph \ref{normal vs
-        shuffled vs reversed ordering factor} shows the percentage
-        execution time increase for shuffled and reversed over normal.
-        Each log file was parsed 10 times, the first run discarded and the
-        remaining 9 runs averaged.  Logging to the database was disabled
-        for the test runs.
-
-        It is hoped that this will prove to have greater effectiveness as
-        the number of rules grows; currently the effect is negligible
-        in comparison to caching each compiled regex, but it still provides
-        approximately a 7\% -- 10\% performance increase.
+        program's execution time.  Rule efficiency concerns are discussed
+        in section~\ref{rule efficiency}.
 
     \item [priority] This is the user-configurable companion to
-        rule\_order: rules with a higher priority will be tried first,
-        overriding rule\_order.  This allows more specific rules to take
-        precedence over more general rules.
+        rule\_order: rules will be tried in order of priority, overriding
+        rule\_order.  This allows more specific rules to take precedence
+        over more general rules.
 
@@
 
-This example rule matches the message Postfix logs when it rejects mail
-from a sender address where the domain has neither an MX record nor an A
-record, i.e.\ mail could not be delivered to the sender's address.  For full
-details see
-http://www.postfix.org/postconf.5.html\#reject\_unknown\_sender\_domain
+This example rule matches the message logged by Postfix when it rejects
+mail from a sender address where the domain has neither an MX record nor an
+A record, i.e.\ mail could not be delivered to the sender's address --- for
+full details see~\cite{reject-unknown-sender-domain}.
 
@@
     \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule.
-
-    \item [program and regex:] If the program equals the program which
-        logged the line the regex will be matched against the line; if
-        successful the action will be called, if not the next rule will be
-        tried.
-
-    \item [action:] will be called if the regex matches successfully (see
-        Section \ref{actions-in-detail} for full details).
+        not \newline used by the algorithm, they serve to document the rule
+        for the user's benefit.
+
+    \item [program and regex:] If the program in the rule equals the
+        program which logged the line the regex will be matched against the
+        line; if the match is successful the action will be called, if not
+        the next rule will be tried.
+
+    \item [action:] will be executed if the regex matches successfully (see
+        section~\ref{actions-in-detail} for full details).
 
@@
     \item [queueid:] is used the extract the queueid matched by the regex,
-        so that mails can be found by queueid and actions performed on the
-        mail.
+        so that the mail can be found by queueid and actions performed on
+        it.
 
 \end{description}
+
+Additional fields are captured by \_\_RESTRICTION\_START\_\_, hence the
+fields in result\_cols and connection\_cols start at 5.
+
+\subsection{Rule efficiency}
+
+\label{rule efficiency}
+
+XXX WRITE ABOUT\@: DYNAMICS OF THE SYSTEM
+
+XXX PRIORITY, OVERLAPPING RULES, MORE SPECIFIC VS \newline MORE GENERAL
+
+XXX EXPLAIN THE DIPS IN THE GRAPHS WITH LARGE LOG FILES
+
+XXX MENTION THAT THE PROGRAM HAS LINER COMPLEXITY AND SCALES LINEARLY WITH
+THE LOG FILE SIZE
+
+Graph~\ref{normal regex vs discard regex} shows execution times gathered
+over 93 log files, with and without caching the regex; also shown on the
+graph are the file sizes and number of lines in each file.
+Graph~\ref{normal regex vs disarded regex factor} shows the percentage
+execution time increase when not caching each regex.  Each log file was
+parsed 10 times, the first run discarded and the remaining 9 runs averaged.
+Logging to the database was disabled for the test runs.
+
+Graph~\ref{normal vs shuffled vs reversed ordering} plots the execution
+time over 93 log files for normal, shuffle and reverse ordering.
+Graph~\ref{normal vs shuffled vs reversed ordering factor} shows the
+percentage execution time increase for shuffled and reversed over normal.
+Each log file was parsed 10 times, the first run discarded and the
+remaining 9 runs averaged.  Logging to the database was disabled for the
+test runs.
+
+It is hoped that this will prove to have greater effectiveness as the
+number of rules grows; currently the effect is negligible in comparison to
+caching each compiled regex, but it still provides approximately a 7--10\%
+performance increase.
 
@@
 
-Unfortunately it's not that easy.
+Unfortunately it's not that straightforward.
 
@@
 
+\subsubsection{Queueid vs pid}
+
+The mail lacks a queueid until it has been accepted, so log lines must
+first be correlated by the smtpd pid, then transition to being correlated
+by the queueid.  This is relatively minor, but does require:
+
+\begin{itemize}
+
+    \item Two versions of several functions: \texttt{by\_pid} and
+        \texttt{by\_queueid}.
+
+    \item Two state tables to hold the data structure for each connection.
+
+    \item Most importantly: every section of code must know whether it
+        needs to lookup the data structures by pid or queueid.
+
+\end{itemize}
+
+\subsubsection{Connection reuse}
+
+Multiple independent mails may be delivered during one connection: this
+requires cloning the current data as soon as a mail is accepted, so that
+subsequent mails won't trample over each other's data.  This must be done
+every time a mail is accepted, as it's impossible to tell in advance which
+connections will accept multiple mails.  It is quite easy to overlook this
+complication because only a small minority of connections accept more than
+one mail. Happily once the mail has been accepted log entries won't be
+correlated by pid for that mail any more (its queueid will be used
+instead), so there isn't any ambiguity about which mail a given log line
+belongs to.\footnote{Unfortunately this statement is not completely
+accurate: see section~\ref{timeouts-during-data-phase} for details.
+However in general there isn't any ambiguity about which data structure
+should be used for a given log line.}  The original connection will be
+discarded unsaved when the client disconnects if there haven't been any
+rejections.  One unsolved difficulty is
+distinguishing between different groups of rejections, e.g.\ when dealing
+with the following sequence:
+
 \begin{enumerate}
 
-    \item The mail lacks a queueid until it has been accepted, so log lines
-        must first be correlated by the smtpd pid, then transition to being
-        correlated by the queueid.  This is relatively minor, but does
-        require:
-
-        \begin{itemize}
-
-            \item Two versions of several functions, \texttt{by\_pid} and
-                \texttt{by\_queueid}.
-
-            \item Two containing data structures to hold the data
-                structure for each connection.
-
-            \item Most importantly: every section of code must know whether
-                it is needs to lookup the data structures by pid or
-                queueid.
-
-        \end{itemize}
-
-    \item Multiple independent mails may be delivered during one
-        connection: this requires cloning the current data as soon as a
-        mail is accepted, so that subsequent mails won't trample over each
-        other's data.  This must be done every time a mail is accepted, as
-        it's impossible to tell in advance which connections will accept
-        multiple mails.  It is quite easy to overlook this complication
-        because only a small minority of connections accept more than one
-        mail. Happily once the mail has been accepted log entries won't be
-        correlated by pid for that mail any more (its queueid will be used
-        instead), so there isn't any ambiguity about which mail a given log
-        line belongs to.  The original connection will be discarded unsaved
-        when the client disconnects.  One unsolved difficulty is
-        distinguishing between different groups of rejections, e.g.\ when
-        dealing with the following sequence:
-
-        \begin{enumerate}
-
-            \item The client attempts to deliver a mail, but it is
-                rejected.
+    \item The client attempts to deliver a mail, but it is rejected.
 
@@
 
-            \item The client attempts to deliver another mail, likewise
-                rejected.
+    \item The client attempts to deliver another mail, likewise rejected.
 
@@
 
-        There should probably be two different entires in the database
-        resulting from the above sequence, but currently there will only be
-        one.
-
-    \item The most difficult complication is that mails are not always
-        delivered directly to a mailbox (or program; there is very little
-        difference between the two): sometimes they are accepted for a
-        local address but need to be delivered to one or more remote
-        addresses due to aliases.  When this occurs a child mail will be
-        injected into the postfix queue, but without the explicit logging
-        smtpd or sendmail injected mails have, so the source is not
-        immediately discernible from the log line in which the mail first
-        appears; from a strictly linear reading of the logs it
-        \textit{usually\/} appears as if the child mail has appeared from
-        thin air.  Subsequently the parent mail will log the creation of
-        the child mail:
+There should probably be two different entires in the database resulting
+from the above sequence, but currently there will only be one.
+
+\subsubsection{Re-injected mails}
+
+The most difficult complication is that locally addressed mails are not
+always delivered directly to a mailbox: sometimes they are accepted for a
+local address but need to be delivered to one or more remote addresses due
+to aliases.  When this occurs a child mail will be injected into the
+postfix queue, but without the explicit logging smtpd or sendmail injected
+mails have.  Thus the source is not immediately discernible from the log
+line in which the mail first appears; from a strictly chronological reading
+of the logs it \textit{usually\/} appears as if the child mail has appeared
+from thin air.  Subsequently the parent mail will log the creation of the
+child mail:
 
@@
 
-        Unfortunately while all log lines from an individual process appear
-        in chronological order, the order in which log lines from different
-        processes are interleaved is subject to the vagaries of process
-        scheduling.  In addition the first log line belonging to the child
-        mail (the log line cited above properly belongs to the parent mail)
-        is logged by qmgr,\footnote{Qmgr is the Postfix daemon which
-        manages the mail queue, determining which mails will have delivery
-        attempted next.} so the order also depends on how busy qmgr
-        is.\footnote{Postfix is quite paranoid about mail delivery, an
-        excellent characteristic for an MTA to possess, so it won't log
-        that the child has been created until it is absolutely certain that
-        the mail has been written to disk.}
-
-        Because of this the parser cannot complain when it encounters a log
-        line from qmgr for a previously unseen mail; it must flag the mail
-        as coming from an unknown origin and subsequently clear the flag if and
-        when the origin of the mail becomes clear.  Obviously the parser
-        could omit checking of where mails originate from, but I feel that
-        it is better to require an explicit source, as bugs are more likely
-        to be exposed.
-
-        Process scheduling can have a still more confusing effect: quite
-        often the child mail will be created, delivered and entirely
-        finished with \textbf{before} the parent logs the creation line!
-        Thus mails flagged as coming from an unknown origin cannot be
-        entered into the database when their final log line is parsed;
-        instead they must be marked as database ready and subsequently
-        entered by the parent mail once it has been identified.  Quite
-        apart from identifying mail injected in an unknown fashion, or bugs
-        in the parser, unknown origin mails need to be marked as such
-        because they lack some data present in the parent mail; they must
-        copy that data from their parent before being entered in the
-        database.
-
-        The algorithm requirements are discussed in
+Unfortunately while all log lines from an individual process appear in
+chronological order, the order in which log lines from different processes
+are interleaved is subject to the vagaries of process scheduling.  In
+addition the first log line belonging to the child mail (the log line cited
+above properly belongs to the parent mail) is logged by qmgr,\footnote{Qmgr
+is the Postfix daemon which manages the mail queue, determining which mails
+will have delivery attempted next.} so the order also depends on how busy
+qmgr is.\footnote{Postfix is quite paranoid about mail delivery, an
+excellent characteristic for an MTA to possess, so it won't log that the
+child has been created until it is absolutely certain that the mail has
+been written to disk.}
+
+Because of this the parser cannot complain when it encounters a log line
+from qmgr for a previously unseen mail; it must flag the mail as coming
+from an unknown origin and subsequently clear the flag if and when the
+origin of the mail becomes clear.  Obviously the parser could omit checking
+of where mails originate from, but the author believes that it is better to
+require an explicit source, as bugs are more likely to be exposed.
+
+Process scheduling can have a still more confusing effect: quite often the
+child mail will be created, delivered and entirely finished with
+\textbf{before} the parent logs the creation line!  Thus mails flagged as
+coming from an unknown origin cannot be entered into the database when
+their final log line is parsed; instead they must be marked as database
+ready and subsequently entered by the parent mail once it has been
+identified.  Quite apart from identifying mail injected in an unknown
+fashion, or bugs in the parser, unknown origin mails need to be marked as
+such because they lack some data present in the parent mail; they must copy
+that data from their parent before being entered in the database.
+The effect of this complication upon the algorithm is discussed in
 section~\ref{tracking-re-injected-mail}.
-
-\end{enumerate}
 
@@
 named by the action.  From a high level viewpoint this design shares
-significant similarity with how an editor maintains state for multiple
-files and performs the actions the user initiates on the correct file.
+significant similarity with how an event dispatch loop works in a GUI\@.
 
@@
 algorithm description.  The data flow chart in section~\ref{flow-chart}
-should be consulted while reading this section also.
+should also be consulted while reading this section.
 
@@
 Everything starts off with a mail entering the system, whether by local
-submission via postdrop/sendmail, by SMTP, or by re-injection due to
-forwarding.  Local submission is the simplest of the three: a queueid is
-assigned immediately and the sender address is logged (action: pickup;
-flowchart:~2).
+submission via postdrop/sendmail, by SMTP, by re-injection due to
+forwarding, or internally generated by Postfix.  Local submission is the
+simplest case: a queueid is assigned immediately and the sender address is
+logged (action: pickup; flowchart:~2).
 
@@
     \item This is followed by rejection of sender, recipients, client
-        address, etc. (action: save\_by\_pid; flowchart:~4); acceptance of
-        one or more mails (action: clone, then save\_by\_pid;
-        flowchart:~5,~4); or some interleaving of both.
-
-    \item The client disconnects (action: disconnect; flowchart:~6).  As
-        soon as a mail is accepted it is assigned a queueid; thus if there
-        is no queueid Postfix didn't accept any mail over that connection,
-        it rejected all attempted deliveries due to the configured
-        restrictions.  Absence of a queueid is checked for during
-        disconnection handling (flowchart:~7) and if none is found the data
-        is cleaned up and entered in the database (flowchart:~8;
-        subroutines: fixup\_connection and commit\_connection), then
-        removed from memory (flowchart:~9).
+        address, etc. (action: rejection; flowchart:~4); acceptance of one
+        or more mails (action: clone; flowchart:~5); or some interleaving
+        of both.
+
+    \item The client disconnects (action: disconnect; flowchart:~6).  If
+        Postfix has rejected any SMTP commands the data will be saved to
+        the database; if not there won't be any data to save (any mails
+        accepted will already have been cloned so their data is in another
+        data structure).
 
     \item If one or more mails were accepted there will be more log entries
-        later, see section~\ref{mail-delivery}.
+        for those mails later, see section~\ref{mail-delivery}.
 
@@
 own;\footnote{Previously discussed in section~\ref{complications},
-complication 3.}  Re-injection is somewhat awkward to explain because it
+complication 3.} re-injection is somewhat awkward to explain because it
 overlaps both the mail acceptance and mail delivery sections.  See
@@
 
+Internally generated mails lack any explicit origin and can only be
+detected using heuristics (see
+section~\ref{identifying-bounce-notifications} for details).  Bounce
+notifications are the primary example of internally generated mails, though
+there may be other types.
+
 \subsubsection{Mail delivery}
@@
 system, whether it is by deletion, bouncing, local or remote delivery.  All
-three are handled in exactly the same way:
+four are handled in exactly the same way:
 
@@
     \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: track; flowchart:~11) --- see
-        section~\ref{tracking-re-injected-mail} for further details.
+        tracked by the parent mail (action: track; flowchart:~11) ---
+        section~\ref{tracking-re-injected-mail} discusses this in detail.
 
@@
         child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14; subroutines fixup\_connection and
-        commit\_connection), then deleted from memory (flowchart:~15).  For
+        database (flowchart:~14), then deleted from memory (flowchart:~15).  For
         re-injected mails see section~\ref{tracking-re-injected-mail}.
@@
 
-I'd like to reiterate that the actions above happen whether the mail is
+It should be reiterated that the actions above happen whether the mail is
 delivered to a mailbox, piped to a command, delivered to a remote server,
@@
 It's probably apparent by now that tracking re-injected mails is the single
-most complex part of the parser.  Section~\ref{complications}, complication
-3 has most of the previous discussion, should you wish to review it.
+most complex part of the algorithm.  Section~\ref{complications} has most
+of the previous discussion, should you wish to review it.
 
@@
         selecting a mail with a previously unseen queueid for delivery
-        (action: qmgr\_chooses\_mail; flowchart:~3), in which case a new
-        data structure will be created.  The mail will be flagged as having
-        unknown origins; this flag should be subsequently cleared once the
-        origin has been established.  The data is then saved (flowchart:~3;
-        subroutine: save).
+        (action: mail\_picked\_for\_delivery; flowchart:~3), in which case
+        a new data structure will be created.  The mail will have been
+        flagged as having unknown origins; this flag should be subsequently
+        cleared once the origin has been established.  The data is then
+        saved.  This may also be an indicator that the mail is a bounce
+        notification, see section~\ref{identifying-bounce-notifications}
+        for details.
 
@@
 
-The algorithm for tracking and saving re-injected mail to the database will
+The algorithm for tracking and saving re-injected mail to the database can
 finally be described:
@@
 
-    \item If the mail is of unknown origin it is a child mail whose parent
-        hasn't yet been identified (action: commit; flowchart:~16).  Mark
-        the mail as ready for entry in the database (flowchart:~17), and
-        wait for the parent to deal with it ( flowchart:~18).  The mail
-        should never have another directly associated log entry; only its
-        parent should refer to it.
+    \item If the mail is of unknown origin it is assumed to be a child mail
+        whose parent hasn't yet been identified (action: commit;
+        flowchart:~16).  Mark the mail as ready for entry in the database
+        (flowchart:~17), and wait for the parent to deal with it
+        (flowchart:~18).  The mail should not have subsequent log entries;
+        only its parent should refer to it.
 
     \item If the mail is a child mail then it has already been tracked
-        (action: commit; flowchart:~19): the data is cleaned up, pulling
-        data from its parent if necessary, and entered in the database
-        (flowchart:~20; subroutines fixup\_connection and
-        commit\_connection), then deleted from memory (flowchart:~21).  The
-        child mail will inform the parent mail that it is finished with
-        (flowchart:~22); if this is the last child and the parent is also
-        finished the parent will delete itself.  The child mail is now
-        finished with.
+        (action: commit; flowchart:~19): the data is cleaned up, missing
+        data is copied from its parent if necessary, and the child is
+        entered in the database (flowchart:~20), then deleted from the
+        state tables (flowchart:~21).  The child mail will be removed from
+        the parent mail's list of children (flowchart:~22); if this is the
+        last child and the parent has also been entered in the database the
+        parent will be deleted from the state tables.
 
@@
         commit; flowchart:~23).  Regardless of the state of its children
-        the data is cleaned up and entered in the database ( flowchart:~24;
-        subroutines fixup\_connection and commit\_connection).  The parent
-        may have children which are waiting to be entered in the database (
-        flowchart:~25); for each of those children their data is cleaned up
-        and entered in the database, then the child is finished with and
-        its data discarded.  The parent may also have outstanding children
-        which are not yet delivered, in which case (flowchart:~26) the
-        parent must wait for all children to finish.  As soon as the last
-        child is finished with the parent is also finished with
-        (flowchart:~27).
+        the data is cleaned up and entered in the database (flowchart:~24).
+        The parent may have children which are waiting to be entered in the
+        database (flowchart:~25); for each of those children their data is
+        cleaned up and entered in the database, then deleted from the state
+        tables.  The parent may also have outstanding children which are
+        not yet delivered, in which case (flowchart:~26) the parent must
+        wait for those children to be finished with.  As soon as the last
+        child is deleted from the state tables the parent will also be
+        finished with (flowchart:~27).
 
@@
 
-This list contains all the actions available in the algorithm.  Each action
-is passed the same arguments:
+Each action is passed the same arguments:
 
@@
 
-    \item [\$rule] The matching rule
+    \item [\$rule] The matching rule.
 
@@
 
-    \item [IGNORE] IGNORE just returns successfully; it is used when a line
-        needs to be parsed for completeness but doesn't either provide any
-        useful data or require anything to be done.
+    \item [IGNORE] This rule just returns successfully; it is used when a
+        line needs to be parsed for completeness but doesn't either provide
+        any useful data or require anything to be done.
 
@@
 
-    \item [SAVE\_BY\_PID] Use the pid from \$line to find the correct
-        connection and call \$self->save() with the appropriate arguments -
-        see save() in SUBROUTINES for more details. If the connection
-        doesn't exist a connection marked faked will be created and a
-        warning issued.
-
-    \item [SAVE\_BY\_QUEUEID] Use the queueid from \$rule and @matches to
-        find the correct connection and call \$self->save() with the
-        appropriate arguments - see save() in SUBROUTINES for more details.
-        If the connection doesn't exist a connection marked faked will be
-        created and a warning issued.
-
-    \item [COMMIT] Enter the data into the database. Entry may be postponed
-        if the mail is a child waiting to be tracked.
-
-    \item [TRACK] Track a mail when it is forwarded to another mail server;
-        this happens when a local address is aliased to a remote address.
-        TRACK will be called when dealing with the parent mail, and will
-        create the child mail if necessary. TRACK checks if the child has
-        already been tracked, either with this parent or with another
-        parent, and issues appropriate warnings in either case. Tracking
-        children is discussed extensively in the paper written about this
-        parser; details about obtaining the paper are given in the SEE ALSO
-        section.
+    \item [SAVE\_BY\_QUEUEID] Save the data extracted by the regex to the
+        mail identified by the queueid from the log line.
+
+    \item [COMMIT] Enter the data from the mail into the database. Entry
+        will be postponed if the mail is a child waiting to be tracked.
+        Once entered, the mail will be deleted from the state tables.
+
+    \item [TRACK] Track a mail when it is re-injected for forwarding to
+        another mail server; this happens when a local address is aliased
+        to a remote address.  TRACK will be called when dealing with the
+        parent mail, and will create the child mail if necessary. TRACK
+        checks if the child has already been tracked, either with this
+        parent or with another parent, and issues appropriate warnings in
+        either case.
 
     \item [REJECTION] Deal with postfix rejecting an SMTP command from the
-        remote client: log the rejection with the accepted mail if there is
-        one, otherwise log it with the connection.
+        remote client: log the rejection with the last accepted mail or the
+        connection, as appropriate.
 
@@
         picking a mail from the queue to deliver. This action is used for
-        both pickup and cleanup due to out of order log lines.
+        both qmgr and cleanup due to out of order log lines; see
+        section~\ref{discarding-cleanup-lines} for details.
 
@@
         table entry and saves data to it, unless out of order logging has
-        caused the cleanup line to be logged first. Lines are assumed to be
-        out of order if the only program seen thus far is cleanup and there
-        is less than five seconds difference between the timestamps of the
-        two lines.
+        caused the cleanup line to be logged first, as discussed in
+        section~\ref{pickup-logging-after-cleanup}.
 
@@
         which case there's a data structure to dispose of, or it may not in
-        which case there's none. The gory details can be found in the
-        internals documentation.
+        which case there's none. See
+        section~\ref{timeouts-during-data-phase} for the gory details.
 
@@
 for a mail which lacks an obvious source.  There are similarities to the
-problem of child mails discussed in section
-\ref{tracking-re-injected-mail}, but unlike the solution described therein
-bounce notifications do not eventually have a log line which identifies
-their source.  Heuristics must be used to identify bounce notifications,
-and those heuristics are:
+problem of re-injected mails discussed in
+section~\ref{tracking-re-injected-mail}, but unlike the solution described
+therein bounce notifications do not eventually have a log line which
+identifies their source.  Heuristics must be used to identify bounce
+notifications, and those heuristics are:
 
@@
 
-    \item The mail will not have been tagged as a tracked mail (see section
-        \ref{tracking-re-injected-mail}).
+    \item The mail will not have been tagged as a tracked mail (see
+        section~\ref{tracking-re-injected-mail}).
 
@@
 
-        The message-id is generated by Postfix for this bounce
-        notification, so the queueid in the message-id must be the same as
-        the queueid of the mail: this is what distinguishes bounce
-        notifications generated locally from bounce notifications which are
-        being forwarded inwards as a result of aliasing.  In the later case
-        the message-id will be unchanged, and so even if it happens to be
-        in the correct format (e.g. if it was generated by Postfix on
-        another server) the queueid in the message-id will not match the
-        queueid of the mail.
+    \item The queueid in the message-id must be the same as the queueid of
+        the mail: this is what distinguishes bounce notifications generated
+        locally from bounce notifications which are being re-injected as a
+        result of aliasing.  In the later case the message-id will be
+        unchanged from the original bounce notification, and so even if it
+        happens to be in the correct format (e.g.\ if it was generated by
+        Postfix on another server) the queueid in the message-id will not
+        match the queueid of the mail.
 
@@
     \item The message-id must have the correct format and match the queueid
-        of the mail.  Whist it is possible that a mail sent from elsewhere
-        could easily have the correct message-id format, the chance that
-        the queueid in the message-id would match the queueid of the mail
-        is extremely small.
+        of the mail.  While a mail sent from elsewhere could easily have
+        the correct message-id format, the chance that the queueid in the
+        message-id would match the queueid of the mail is extremely small.
 
@@
 accepted, then makes a second delivery attempt which it continues with
-until the delivery is complete. XXX VERIFY THIS AND DETERMINE WHAT CLIENTS
-EXHIBIT THIS BEHAVIOUR XXX
+until the delivery is complete.
 
 Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify the when this
+clear, so once again heuristics are required to identify when this
 behaviour occurs.  In this case a list of all mails accepted during a
@@
 accumulate in the state table and will cause clashes if the queueid is
-reused.
+reused.  The mail cannot be entered in the database as the only data
+available is the client hostname and IP address, but the database schema
+requires many more fields be populated (see section~\ref{database schema}).
 
@@
 is rejected; they may also disconnect after other errors, but such
-disconnections are already being handled.  Sadly Postfix doesn't log a
-message saying the mail has been discarded, as should be expected by now.
-The checks to identify this happening are:
+disconnections are either unimportant or are handled elsewhere in the
+algorithm (section~\ref{timeouts-during-data-phase}).  Sadly Postfix
+doesn't log a message saying the mail has been discarded, as should be
+expected by now.  The checks to identify this happening are:
 
@@
 without identifying and discarding them they accumulate in the state table
-and will cause clashes if the queueid is reused.
+and will cause clashes if the queueid is reused.  The mail cannot be
+entered in the database as the only data available is the client hostname
+and IP address, but the database schema (see section~\ref{database schema})
+requires many more fields be populated.
 
@@
 To deal properly with timeouts the parsing algorithm needs to do the
-following in the timeout action:
+following in the TIMEOUT action:
 
@@
         address or any of the recipient addresses there is no way for the
-        client to tell which command was rejected, nor does there appear to
-        be a way for the SMTP conversation to recover, so clients
-        disconnect uncleanly.
-
-        A timeout may thus apply either to an accepted mail, or a rejected
+        client to tell which command was rejected.  Some clients make no
+        attempt to recover and disconnect uncleanly.
+
+        A timeout may thus apply either to an accepted mail or a rejected
         mail.  To distinguish between the two cases the algorithm compares
@@
         the timeout has already been recorded so the timeout action
-        finishes.  If the mail acceptance timestamp is greater then the timeout
-        applies to the just-accepted mail.
-
-    \item The mail needs to be discarded, but there is yet another
-        complication: sometimes the cleanup line is logged after the
-        timeout line.  If the cleanup line has already been seen the mail
-        can be discarded (both from global state and the connection's list
-        of accepted mails) and the timeout action finishes.
-
-        In the case where the cleanup line is still pending the algorithm
-        updates a global cache of queueids, adding the queueid and the
-        timestamp from the timeout line.  When the next cleanup line is
-        parsed for that queueid the cache will be checked, and will be
-        deemed part of the discarded mail and ignored if it meets the
-        following requirements:
+        finishes.  If the mail acceptance timestamp is greater then the
+        timeout applies to the just-accepted mail, which will be discarded.
+
+\end{enumerate}
+
+This complication is further complicated by out of order cleanup lines: see
+section~\ref{discarding-cleanup-lines} for details.
+
+\subsubsection{Discarding cleanup lines}
+
+\label{discarding-cleanup-lines}
+
+The author has only observed this complication occurring after a timeout,
+though there may be other circumstances which trigger it.  Sometimes the
+cleanup line is logged after the timeout line; parsing this line causes the
+creation of a new state table entry for the queueid in the log line.  This
+is incorrect because the line actually belongs to the mail which has just
+been discarded, and the next log line for that queueid will be seen when
+the queueid is reused, causing a queueid clash and the appropriate warning.
+
+In the case where the cleanup line is still pending the algorithm updates a
+global cache of queueids, adding the queueid and the timestamp from the
+timeout line.  When the next cleanup line is parsed for that queueid the
+cache will be checked, and the line will be deemed part of the discarded
+mail and discarded if it meets the following requirements:
 
@@
 
-            \item The queueid must not have been reused yet, i.e. there
-                isn't an entry in the state tables for the queueid.
-
-            \item The timestamp of the cleanup line must be within six
-                minutes of the mail acceptance timestamp.  Timeouts happen
-                after five minutes, so six gives a one minute leeway.
+    \item The queueid must not have been reused yet, i.e.\ there isn't an
+        entry in the state tables for the queueid.
+
+    \item The timestamp of the cleanup line must be within six minutes of
+        the mail acceptance timestamp.  Timeouts happen after five minutes,
+        so six minutes gives one minute of leeway.
 
@@
 
-        The next cleanup line must meet the criteria above for it to be
-        discarded because not every connection where a timeout occurs will
-        have a cleanup line logged for it; if the algorithm blindly
-        discarded the next cleanup line it would in some cases be mistaken.
-        Whether or not the line is discarded the queueid will be removed
-        from the cache of timeout queueids.
-
-\end{enumerate}
+The next cleanup line must meet the criteria above for it to be discarded
+because not every connection where a timeout occurs will have a cleanup
+line logged for it; if the algorithm blindly discarded the next cleanup
+line it would in some cases be mistaken.  Whether or not the next cleanup
+line is discarded the queueid will be removed from the cache of timeout
+queueids when the next pickup line is parsed.
 
@@
 
+XXX WRITE THIS
+
 \subsection{Mails covered}
@@
 \label{mails-covered}
+
+XXX WRITE THIS
 
@@
     \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10; this may possibly have changed in Postfix 2.3, or the
-        upcoming 2.4.}  It should be reasonably simple to write a policy
-        server which causes Postfix to log a warning containing the HELO
-        hostname when the DATA command is accepted.\footnote{See
-        http://www.postfix.org/SMTPD\_POLICY\_README.html for full
-        documentation about policy servers.}
+        logged if the mail is accepted.\footnote{Tested with Postfix 2.2.10
+        and 2.3.7; this may possibly have changed in Postfix 2.4.}  It
+        should be reasonably simple to write a policy server which causes
+        Postfix to log a warning containing the HELO hostname when the DATA
+        command is accepted.
 
@@
 \subsection{Possible improvements}
+
 \begin{itemize}
@@
 
+XXX WRITE THIS
+
 \appendix
 
-\section{References}
-
-\label{references}
-
-\subsection{Other Postfix log parsers}
+\section{Other Postfix log parsers reviewed}
+
+XXX THIS NEEDS TO BE EXPANDED
 
@@
 
-\subsection{Postfix resources}
-
-\subsection{SMTP references}
+\bibliographystyle{logparser-bibliography-style}
 
@@
 
-XXX CHANGE REFERENCES TO BIBLIOGRAPHY??  MERGE PREVIOUS SECTIONS??  PUT
-BIBLIOGRAPHY IN ITS OWN SECTION??
-
-% Bibliography.
-\bibliographystyle{logparser-bibliography-style}
+\label{bibliography}
 \bibliography{logparser-bibliography}
@@
 \section{Graphs}
+
+\label{graphs}
 

### 2007-06-11 doc/logparser.tex
logparser.tex: * Documented SQL schema. * The TOC is now linked for easy navigation, and there are bookmarks in Acrobat too. * Misc cleanups. Makefile: produce a PDF automatically.

@@
 
-% Better typesetting of URLs.
-\usepackage{url}
 % Include images
@@
 \renewcommand{\refname}{Bibliography}
+% Add the bibliography into the table of contents.
 \usepackage[section,numbib,nottoc]{tocbibind}
+% When creating a PDF make the table of contents into links to the pages
+% (without horrible red borders) and include bookmarks.
+\usepackage{hyperref}
+\hypersetup{%
+    pdftitle    = {Parsing Postfix log files},
+    pdfauthor   = {John Tobin},
+    final,
+    pdfborder   = {0 0 0},
+    breaklinks,
+}
 
@@
 XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS
+
+XXX LOTS MORE REFERENCES
 
@@
 
-XXX WRITE THIS
+The database schema can be conceptually divided in two: the rules which are
+used to parse log files, and the data saved from the parsing of log files.
+Rules have the fields required to parse the log lines, extract data to be
+saved, and cause the correct action to be executed; they also have several
+fields which aid the user in understanding what each rule parses.  The
+rules are described in detail in section~\ref{rules} and the fields in
+particular in section~\ref{rule attributes}, and that discussion won't be
+repeated here.  The data saved from parsing the logs is divided into two
+tables: connections and results.
+
+\subsubsection{Connections table}
+
+Every connection and/or mail will have a single entry in the connections
+table containing the following fields:
+
+\begin{description}
+
+    \item [id] This field uniquely identifies the row.
+
+    \item [server\_ip] The IP address (IPv4 or IPv6) of the server: the
+        local server when receiving mail, the remote server when sending
+        mail.
+
+    \item [server\_hostname] The hostname of the server, it will be
+        \textit{unknown} if the IP address could not be resolved to a
+        hostname via DNS.
+
+    \item [client\_ip] The client IP address (IPv4 or IPv6): the remote
+        server when receiving mail, the local server when sending mail.
+
+    \item [client\_hostname] The hostname of the client, it will be
+        \textit{unknown} if the IP address could not be resolved to a
+        hostname via DNS.
+
+    \item [helo] The hostname used in the HELO command.  The HELO
+        occasionally changes during a connection, presumably because spam
+        or virus senders think it's a good idea.  By default Postfix only
+        logs this when it rejects an SMTP command, but it is trivial to log
+        this for every connection:
+
+        \texttt{smtpd\_helo\_restrictions = \newline
+        \hspace*{2em} check\_helo\_access regexp:/etc/postfix/log\_helo.regex}
+
+        Where \texttt{/etc/postfix/log\_helo.regex} contains:
+
+        \texttt{/./ WARN Logging HELO for diagnostic purposes}
+
+    \item [queueid] The queueid of the mail, or \textit{NOQUEUE} if a mail
+        wasn't accepted.
+
+    \item [start] The timestamp of the first log line, in seconds since the
+        epoch.
+
+    \item [end] The timestamp of the last log line, in seconds since the
+        epoch.
+
+\end{description}
+
+\subsubsection{Results table}
+
+Every log line corresponding to Postfix performing an action other than
+INFO will have an entry in the results table, e.g. rejecting an SMTP
+command, delivering a mail, or bouncing a mail.  Any log line where the
+Postfix action is INFO is not interesting in and of itself, but provides
+additional information which will be saved with other results.
+
+\begin{description}
+
+    \item [connection\_id] A reference to the entry in the connections
+        table this results entry belongs to.
+
+    \item [rule\_id] A reference to the entry in the rules table which
+        matched the log line and created this result.
+
+    \item [warning] Postfix can log a warning instead of rejecting an SMTP
+        command - a facility that is quite useful for testing new
+        restrictions.  This field will be 1 if the log line parsed was a
+        warning rather than a real rejection, or 0 for a real rejection or
+        any other result.
+
+    \item [smtp\_code] The SMTP code extracted from the log line.  In
+        general an SMTP code is only present for a rejection or final
+        delivery; results missing an SMTP code will duplicate the SMTP code
+        of other results.  Some final delivery log lines don't contain an
+        SMTP code: in those cases the code is faked based on the success or
+        failure represented by the log line.
+
+    \item [sender] The sender's email address.  This can change during one
+        single connection, when the connection is reused to send multiple
+        mails.
+
+    \item [recipient] The recipient address.
+
+    \item [message\_id] The message-id of the accepted mail, or NULL if no
+        mail was accepted.
+
+    \item [data] A field available for anything not covered by other
+        fields, e.g. the rejection message from an RBL.
+
+    \item [timestamp] The time the line was logged, in seconds since the
+        epoch.
+
+\end{description}
 
@@
 
-    \item Checking Sender Policy Framework
-        records~\cite{openspf},~\cite{wikipedia-spf}.  SPF records specify
-        which mail servers are allowed to send mail claiming to be from a
-        particular domain.  The intention is to reduce spam from faked
-        sender addresses, backscatter~\cite{postfix-backscatter} and
+    \item Checking Sender Policy Framework records~\cite{openspf,
+        wikipedia-spf}.  SPF records specify which mail servers are allowed
+        to send mail claiming to be from a particular domain.  The
+        intention is to reduce spam from faked sender addresses,
+        backscatter~\cite{postfix-backscatter} and
         joe-jobs~\cite{wikipedia-joe-job}; however there has been a lot of
@@
 \subsection{Rule attributes}
+
+\label{rule attributes}
 
@@
         administrator a ``heads up'' for potential trouble spots.\/}
-        \url{http://jimsun.linxnet.com/postfix_contrib.html}
+        \newline \url{http://jimsun.linxnet.com/postfix_contrib.html}
 
@@
 
-\subsection{Ratio of file size to execution time and number of lines to
-execution time}
+\subsection{Ratio of file size and number of lines to execution time}
 \label{execution time vs file size vs number lines factor}
@@
 
-\subsection{Percentage increase of shuffled over normal and reversed over
-normal}
+\subsection{Percentage increase of shuffled and reversed over normal}
 \label{normal vs shuffled vs reversed ordering factor}

### 2007-06-12 doc/logparser.tex
Wrote the background section on Postfix restrictions.

@@
 
-XXX TO BE WRITTEN
+Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
+smtpd_per_user_control, policy-servers}, the following is a brief overview.
+
+Postfix uses several restriction lists (each containing zero or more
+restrictions) to decide the fate of an SMTP command, one list for each
+stage of the SMTP conversation: client connection, HELO command, MAIL FROM
+command, RCPT TO command (possibly multiple times), DATA command, and end
+of data.  By default restriction lists will not be evaluated until the
+first RCPT TO command is received, because some clients don't deal with
+earlier rejections properly; a side benefit of this delay is that Postfix
+has more information available when logging the rejection.\footnote{The
+\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
+uses simple lookup tables as the deciding factor for some restrictions,
+e.g. \texttt{check\_client\_access~\nolinebreak
+cidr:/etc/postfix/client\_access}\footnote{\texttt{cidr} is the type of the
+lookup table.} will check whether the IP address of the connected client
+matches the left hand side of each line in the file and will return the
+right hand side of the first matching line.  Other restrictions determine
+their result in different ways, e.g. \texttt{reject\_rbl\_client
+some.rbl.com} checks whether the IP address of the client is present in the
+listed RBL.
+
+Each restriction is evaluated to produce a result of \textit{reject\/},
+\textit{permit\/}, \textit{dunno\/} or the name of another restriction to
+be evaluated.\footnote{Other results are possible, for full details
+see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
+The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
+\textit{dunno\/} means to stop evaluating the current restriction and
+continue with the next restriction in the list, allowing more specific
+cases to be handled differently from more general cases.  When the result
+is another restriction Postfix will evaluate the new restriction, allowing
+restrictions to be chosen based on the client IP address, HELO hostname,
+sending address, recipient address, etc.  The administrator can define new
+restrictions as a list of existing restrictions, allowing arbitrarily long
+and complex sequences of lookups and restrictions.
+
+This description is necessarily brief, for further details
+see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
 

### 2007-06-18 doc/logparser.tex
Wrote the SMTP background section, plus supporting references.  Started on parsing efficiency, but didn't get far.  Added Log Mail Analyser to the list of reviewed parsers.  Improved one graph slightly.

@@
 
-XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS
+XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS.  REVIEW
+EACH PAPER, WRITING AN ABSTRACT FOR IT: WHAT'S THE SAME, WHAT'S DIFFERENT,
+PERFORMANCE AND \newline COMPLETENESS COMPARISON.  FROM THIS I'LL GET AN
+OVERALL PICTURE, AND I CAN WORK REFERENCES TO THE PAPERS INTO THIS PAPER,
+PLUS MAYBE A PRIOR ART SECTION.
+
+XXX TURN OFF RESTRICTIONS FOR ME AND SEE HOW WELL SPAMASSASSIN DEALS
+WITH THE RESULTING MAILS; FALSE POSITIVES, FALSE NEGATIVES, ETC.
 
@@
 
-XXX WRITE THIS
+The Simple Mail Transfer Protocol, originally defined in RFC
+821~\cite{RFC821} and later in RFC 2821~\cite{RFC2821}, is used for
+transferring mail between Mail Transport Agents (MTAs).  It is a simple,
+human readable, plain text protocol, making it quite easy to test and debug
+problems with it.  Despite the simplicity many virus and/or spam sending
+programs fail to implement it properly, so requiring strict adherence to
+the protocol specification is beneficial in protecting against spam and
+viruses.\footnote{This violates the principle of \textit{Be liberal in what
+you accept, and conservative in what you send\/} from RFC
+760~\cite{rfc760}, but unfortunately that principle was written in a
+friendlier time.}  A typical SMTP conversation resembles the following (the
+lines starting with a three digit number are sent by the server, all other
+lines are sent by the client):
+
+\begin{verbatim}
+220 smtp.example.com ESMTP
+HELO client.example.com
+250 smtp.example.com
+MAIL FROM: <alice@example.com>
+250 2.1.0 Ok
+RCPT TO: <bob@example.com>
+250 2.1.5 Ok
+DATA
+354 End data with <CR><LF>.<CR><LF>
+Message headers and body sent here.
+.
+250 2.0.0 Ok: queued as D7AFA38BA
+QUIT
+221 2.0.0 Bye
+\end{verbatim}
+
+A detailed description of SMTP is beyond the scope of this document:
+introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
 
@@
 
+Parsing efficiency is an obvious concern when the parser routinely needs to
+deal with 50 MB log files containing 300,000 log lines.  The graphs in
+section~\ref{graphs} illustrate the execution time for various
+configurations of the parser and should be referred to during the following
+discussion.
+
+An important property of a parser is how execution time scales with input
+size: does it scale linearly, polynomially, or exponentially?
+Graph~\ref{execution time vs file size vs number lines} shows the execution
+time in seconds, file size in MB and tens of thousands of lines per log
+file.  All three lines run roughly parallel, giving a visual impression
+that the algorithm scales linearly with input size.  This is borne(XXX) out
+by graph~\ref{execution time vs file size vs number lines factor}, which
+plots the ratio of file size in MB vs execution time in seconds and number
+of lines
+
 Graph~\ref{normal regex vs discard regex} shows execution times gathered
@@
 
+    \item [Log Mail Analyser] \cite{log-mail-analyser} attempts to
+        correlate Postfix log lines by queueid and produce both a CSV file
+        and either a Berkeley DB or MySQL database.  It aims to save
+        \textit{date and hour, DNS name and IP address host, mail server IP
+        address, sender, receiver and e-mail status (sent, rejected)\/}, so
+        its data extraction is reasonably complete.  The parsing is
+        hard-coded into the program; extending it requires familiarity with
+        a substantial portion of the code.  Producing a CSV file appears to
+        be quite limiting, but the intention is to provide a format
+        suitable for use with grep(1), awk(1) and the myriad of other Unix
+        utilities already available.  Supporting Berkeley DB is also
+        limiting, requiring an understanding of the (partially documented)
+        database structure and writing a program to extract the required
+        data - it appears to be the least useful of the three output
+        formats, though the authors of LMA did provide some queries:
+        IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT.  Unfortunately
+        when the author tested LMA it didn't work properly and after
+        exploring the code for some time the author reluctantly gave up.
+
 \end{description}

### 2007-06-18 doc/logparser.tex
Bunch of warnings from chktex fixed.

@@
 
-XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS.  REVIEW
-EACH PAPER, WRITING AN ABSTRACT FOR IT: WHAT'S THE SAME, WHAT'S DIFFERENT,
-PERFORMANCE AND \newline COMPLETENESS COMPARISON.  FROM THIS I'LL GET AN
-OVERALL PICTURE, AND I CAN WORK REFERENCES TO THE PAPERS INTO THIS PAPER,
-PLUS MAYBE A PRIOR ART SECTION.
-
-XXX TURN OFF RESTRICTIONS FOR ME AND SEE HOW WELL SPAMASSASSIN DEALS
-WITH THE RESULTING MAILS; FALSE POSITIVES, FALSE NEGATIVES, ETC.
+XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS\@.
+REVIEW EACH PAPER, WRITING AN ABSTRACT FOR IT\@: WHAT'S THE SAME, WHAT'S
+DIFFERENT, PERFORMANCE AND \newline COMPLETENESS COMPARISON\@.  FROM THIS
+I'LL GET AN OVERALL PICTURE, AND I CAN WORK REFERENCES TO THE PAPERS INTO
+THIS PAPER, PLUS MAYBE A PRIOR ART SECTION\@.
+
+XXX TURN OFF RESTRICTIONS FOR ME AND SEE HOW WELL SPAMASSASSIN DEALS WITH
+THE RESULTING MAILS\@; FALSE POSITIVES, FALSE NEGATIVES, ETC\@.
 
@@
     \item [server\_hostname] The hostname of the server, it will be
-        \textit{unknown} if the IP address could not be resolved to a
-        hostname via DNS.
+        \textit{unknown\/} if the IP address could not be resolved to a
+        hostname via DNS\@.
 
@@
     \item [client\_hostname] The hostname of the client, it will be
-        \textit{unknown} if the IP address could not be resolved to a
-        hostname via DNS.
+        \textit{unknown\/} if the IP address could not be resolved to a
+        hostname via DNS\@.
 
@@
 
-    \item [queueid] The queueid of the mail, or \textit{NOQUEUE} if a mail
-        wasn't accepted.
+    \item [queueid] The queueid of the mail, or \textit{NOQUEUE\/} if a
+        mail wasn't accepted.
 
@@
 Every log line corresponding to Postfix performing an action other than
-INFO will have an entry in the results table, e.g. rejecting an SMTP
+INFO will have an entry in the results table, e.g.\ rejecting an SMTP
 command, delivering a mail, or bouncing a mail.  Any log line where the
@@
     \item [warning] Postfix can log a warning instead of rejecting an SMTP
-        command - a facility that is quite useful for testing new
+        command --- a facility that is quite useful for testing new
         restrictions.  This field will be 1 if the log line parsed was a
@@
     \item [data] A field available for anything not covered by other
-        fields, e.g. the rejection message from an RBL.
+        fields, e.g.\ the rejection message from an RBL\@.
 
@@
 uses simple lookup tables as the deciding factor for some restrictions,
-e.g. \texttt{check\_client\_access~\nolinebreak
-cidr:/etc/postfix/client\_access}\footnote{\texttt{cidr} is the type of the
-lookup table.} will check whether the IP address of the connected client
-matches the left hand side of each line in the file and will return the
-right hand side of the first matching line.  Other restrictions determine
-their result in different ways, e.g. \texttt{reject\_rbl\_client
-some.rbl.com} checks whether the IP address of the client is present in the
-listed RBL.
-
-Each restriction is evaluated to produce a result of \textit{reject\/},
-\textit{permit\/}, \textit{dunno\/} or the name of another restriction to
-be evaluated.\footnote{Other results are possible, for full details
+e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
+\footnote{\texttt{cidr} is the type of the lookup table.} will check
+whether the IP address of the connected client matches the left hand side
+of each line in the file and will return the right hand side of the first
+matching line.  Other restrictions determine their result in different
+ways, e.g. \texttt{reject\_rbl\_client some.rbl.com} checks whether the IP
+address of the client is present in the listed RBL\@.
+
+Each restriction is evaluated to produce a result of \textit{reject},
+\textit{permit}, \textit{dunno\/} or the name of another restriction to be
+evaluated.\footnote{Other results are possible, for full details
 see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
@@
 file.  All three lines run roughly parallel, giving a visual impression
-that the algorithm scales linearly with input size.  This is borne(XXX) out
-by graph~\ref{execution time vs file size vs number lines factor}, which
-plots the ratio of file size in MB vs execution time in seconds and number
-of lines
+that the algorithm scales linearly with input size.  This is borne out by
+graph~\ref{execution time vs file size vs number lines factor}, which plots
+the ratio of file size in MB vs execution time in seconds and number of
+lines
 
@@
 
-    \item [Log Mail Analyser] \cite{log-mail-analyser} attempts to
+    \item [Log Mail Analyser]~\cite{log-mail-analyser} attempts to
         correlate Postfix log lines by queueid and produce both a CSV file
@@
         \textit{date and hour, DNS name and IP address host, mail server IP
-        address, sender, receiver and e-mail status (sent, rejected)\/}, so
+        address, sender, receiver and e-mail status (sent, rejected)}, so
         its data extraction is reasonably complete.  The parsing is
@@
         database structure and writing a program to extract the required
-        data - it appears to be the least useful of the three output
+        data --- it appears to be the least useful of the three output
         formats, though the authors of LMA did provide some queries:
-        IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT.  Unfortunately
-        when the author tested LMA it didn't work properly and after
-        exploring the code for some time the author reluctantly gave up.
+        IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.
+        Unfortunately when the author tested LMA it didn't work properly
+        and after exploring the code for some time the author reluctantly
+        gave up.
 

### 2007-06-22 doc/logparser.tex
Wrote overlapping rules section.

@@
 
+\subsection{Overlapping rules}
+
+The parser does not try to detect overlapping rules;\footnote{It may be
+possible to parse each rule's regex, generate an NFA, and compare all NFAs
+to determine if any overlap.  The author has not attempted to do this: such
+a project would probably qualify for a PhD, and may involve solving the
+Halting Problem and circumventing the Church-Turing Thesis.} that
+responsibility is left to the author of the rules.  Unintentionally
+overlapping rules lead to inconsistent parsing and data extraction because
+the order in which rules are tried against each line determines which rule
+matches.  Overlapping rules are frequently a requirement, allowing a more
+specific rule to match some lines and a more general rule to match the
+remainder, e.g. separating smtp delivery to a filtering proxy from smtp
+delivery to the rest of the world.  The algorithm provides a facility for
+handling overlapping rules: the priority field in each rule (defaults to
+zero).  Rules are tried in order of priority, highest first, overriding the
+rule\_order field maintained by the parser.  Negative priorities may also
+be useful for catchall rules.
+
+Detecting overlapping rules is difficult, but the following may be helpful:
+
+\begin{itemize}
+
+    \item Sort the regexs and visually inspect the list, e.g. with SQL
+        similar to: \textbf{select regex from rules order by regex;}
+
+    \item Compare the results of parsing using sorted, shuffled and
+        reversed rules.\footnote{See section~\ref{rule efficiency} for more
+        details of sorting the rules.}  Parse a number of log files using
+        normal sorting, dump a textual representation of the connections
+        and results tables, and delete everything from those tables.
+        Repeat with shuffled and reversed sorting.  If there are no
+        overlapping rules the tables from each run will be identical;
+        differences indicate overlapping rules, and the rule\_ids which
+        differ determine which rules overlap.  Unfortunately this method
+        cannot prove the absence of overlapping rules; it can detect
+        overlapping rules --- but only if there are log lines which match
+        more than one rule.
+
+\end{itemize}
+
 \subsection{Example rule}

### 2007-06-24 doc/logparser.tex
Wrote most of the rule efficiency section.  It needs a bit more work and proof reading, but it's mostly there now.  Changed the graphs slightly, added new graphs, and updated the Makefile.

@@
     pdfauthor   = {John Tobin},
-    final,
+    final       = true,
     pdfborder   = {0 0 0},
-    breaklinks,
+    breaklinks  = true,
 }
@@
 
-XXX LOOK AT citeseer.com AND scholar.google.com FOR RELATED PAPERS\@.
-REVIEW EACH PAPER, WRITING AN ABSTRACT FOR IT\@: WHAT'S THE SAME, WHAT'S
-DIFFERENT, PERFORMANCE AND \newline COMPLETENESS COMPARISON\@.  FROM THIS
-I'LL GET AN OVERALL PICTURE, AND I CAN WORK REFERENCES TO THE PAPERS INTO
-THIS PAPER, PLUS MAYBE A PRIOR ART SECTION\@.
+XXX MAYBE A PRIOR ART SECTION\@?
 
@@
 XXX LOTS MORE REFERENCES
+
+XXX MAYBE ADD A SECTION ABOUT OPTIMISATION OF THE DATABASE\@.
 
@@
         program's execution time.  Rule efficiency concerns are discussed
-        in section~\ref{rule efficiency}.
+        in section~\ref{rule ordering for efficiency}.
 
@@
 
+XXX WRITE ABOUT DATABASE OPTIMISATION??
+
 \label{rule efficiency}
 
-XXX WRITE ABOUT\@: DYNAMICS OF THE SYSTEM
-
-XXX PRIORITY, OVERLAPPING RULES, MORE SPECIFIC VS \newline MORE GENERAL
-
-XXX EXPLAIN THE DIPS IN THE GRAPHS WITH LARGE LOG FILES
-
-XXX MENTION THAT THE PROGRAM HAS LINER COMPLEXITY AND SCALES LINEARLY WITH
-THE LOG FILE SIZE
-
 Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 50 MB log files containing 300,000 log lines.  The graphs in
-section~\ref{graphs} illustrate the execution time for various
-configurations of the parser and should be referred to during the following
-discussion.
+deal with 50 MB log files containing 300,000 log lines (for a small mail
+server --- large scale mail servers would have much larger log files on a
+daily basis).  When generating the data for the graphs included in
+section~\ref{graphs}, 93 log files (totaling 10.08 GB, 60.72 million lines)
+were each parsed 10 times, the first run discarded and the remaining 9 runs
+averaged.  The first run is discarded for two reasons:
+
+\begin{enumerate}
+
+    \item The execution time will be higher because the log file must be
+        read from disk, whereas for subsequent runs the log file will be
+        cached in memory by the operating system.
+
+    \item The execution time will also be higher because the rule ordering
+        will be sub-optimal compared to subsequent runs.
+
+\end{enumerate}
+
+Saving results to the database was disabled for the test runs, as that
+dominates the run time of the program, and the tests are aimed at measuring
+the speed of the parser rather than the speed of the database.
+
+\subsubsection{Algorithmic complexity}
 
@@
 time in seconds, file size in MB and tens of thousands of lines per log
-file.  All three lines run roughly parallel, giving a visual impression
-that the algorithm scales linearly with input size.  This is borne out by
-graph~\ref{execution time vs file size vs number lines factor}, which plots
-the ratio of file size in MB vs execution time in seconds and number of
-lines
-
-Graph~\ref{normal regex vs discard regex} shows execution times gathered
-over 93 log files, with and without caching the regex; also shown on the
-graph are the file sizes and number of lines in each file.
-Graph~\ref{normal regex vs disarded regex factor} shows the percentage
-execution time increase when not caching each regex.  Each log file was
-parsed 10 times, the first run discarded and the remaining 9 runs averaged.
-Logging to the database was disabled for the test runs.
-
-Graph~\ref{normal vs shuffled vs reversed ordering} plots the execution
-time over 93 log files for normal, shuffle and reverse ordering.
-Graph~\ref{normal vs shuffled vs reversed ordering factor} shows the
-percentage execution time increase for shuffled and reversed over normal.
-Each log file was parsed 10 times, the first run discarded and the
-remaining 9 runs averaged.  Logging to the database was disabled for the
-test runs.
-
-It is hoped that this will prove to have greater effectiveness as the
-number of rules grows; currently the effect is negligible in comparison to
-caching each compiled regex, but it still provides approximately a 7--10\%
-performance increase.
+file per log file.  All three lines run roughly in parallel, giving a
+visual impression that the algorithm scales linearly with input size.  This
+impression is borne out by graph~\ref{execution time vs file size vs number
+lines factor}, which plots the ratio of file size vs execution time and
+number of lines vs execution time.  As the reader can see, the ratios are
+quite tightly banded, showing that the algorithm scales linearly: the much
+larger log files between points 60 and 70 on the X axis in
+graph~\ref{execution time vs file size vs number lines} don't cause any
+abnormality in the corresponding points in graph~\ref{execution time vs
+file size vs number lines factor}.
+
+\subsubsection{Rule ordering for efficiency}
+
+\label{rule ordering for efficiency}
+
+XXX DO A FULL RUN ONCE RULE\_ORDER ISN'T BEING TRASHED AND GRAPH RULE
+HITS\@.
+
+Rule ordering was briefly discussed in section~\ref{rule attributes} and
+it will be covered in greater detail in this section.  At the time of
+writing there are 131 different rules, some of which match much more
+frequently than others.  Assuming that the distribution of log lines is
+reasonably steady over time, program efficiency should benefit from trying
+more frequently matching rules before those which match less frequently.
+To test this hypothesis three full test runs were performed with different
+rule orderings:
+
+\begin{description}
+
+    \item [normal]  The most optimal order, according to the hypothesis:
+        rules which match most often will be tried first.
+
+    \item [shuffle] Random ordering --- the rules will be shuffled once
+        before use and will retain that ordering for the entirety of the
+        log file.  Note that the ordering will change every time the parser
+        is executed, so 10 different orderings will be generated for each
+        log file in the test run.  This is intended to represent an
+        unsorted rule set.
+
+    \item [reverse] Hypothetically the least optimal order: the most
+        frequently matching rules will be tried last.
+
+\end{description}
+
+The resulting execution times are plotted in graph~\ref{normal vs shuffled
+vs reversed ordering}, where the difference is not readily apparent, but
+graph~\ref{normal vs shuffled vs reversed ordering factor} shows the
+percentage increase of execution times:
+
+\begin{description}
+
+    \item [shuffled] 6--11\% increase; mean 8.34, standard deviation 2.44.
+
+    \item [reversed] 15--22\% increase; mean 17.97, standard deviation
+        2.94.
+
+\end{description}
+
+Overall this increases the parser's efficiency by 5.5--10\%; a modest but
+worthwhile increase.
+
+\subsubsection{Caching each regex}
+
+Graph~\ref{normal regex vs discard regex} shows execution times with and
+without caching the regex; also shown on the graph are the file sizes and
+number of lines in each file.  Caching the compiled regexs is obviously far
+more efficient; graph~\ref{normal regex vs disarded regex factor} shows the
+percentage execution time increase when not caching each regex is between
+350--570\% (mean 465.58\%, standard deviation 114.70\%).  Perl compiles the
+original regex into an internal representation, optimising the regex to
+improve the speed of matching, but this compilation and optimisation takes
+CPU time.  Perl automatically caches static regexs, but dynamic regexs need
+to be explicitly compiled and cached.
 
@@
 \bibliographystyle{logparser-bibliography-style}
-
-XXX ADD RESOURCES/REFERENCES HERE AND THEN MAKE REFERENCES TO THEM FROM WITHIN THE PAPER
-
 \label{bibliography}
@@
 
-\subsection{Execution time vs file size vs number of lines.}
+\subsection{Parser scalability}
+
+\subsubsection{Execution time vs file size vs number of lines.}
 \label{execution time vs file size vs number lines}
@@
 
-\subsection{Ratio of file size and number of lines to execution time}
+\subsubsection{Ratio of file size and number of lines to execution time}
 \label{execution time vs file size vs number lines factor}
@@
 
-\subsection{Regex: cached vs discarded}
+\subsection{Rule ordering}
+
+\subsubsection{Normal vs shuffled vs reversed ordering}
+\label{normal vs shuffled vs reversed ordering}
+\includegraphics{plot-normal-shuffle-reverse.ps}
+
+\subsubsection{Percentage increase of shuffled and reversed over normal}
+\label{normal vs shuffled vs reversed ordering factor}
+\includegraphics{plot-normal-shuffle-reverse-factor.ps}
+
+The worst-case performance of shuffled rules can be seen at log file 30,
+where it is equivalent to reversed rules.
+
+XXX THIS NEXT PARAGRAPH MIGHT NEED TO BE REWRITTEN\@.
+
+The dips at logfiles 22 and 62--68 correspond to peaks in log file size in
+graph~\ref{execution time vs file size vs number lines}.  The explanation
+took some time to arrive at, but turns out to be reasonably simple.
+Examining the database after parsing logfile 22 shows 64 rules matched
+during parsing, but the distribution of matching rules is extremely skewed,
+as shown in graph~\ref{log22 distribution by rule}.  Graph~\ref{rule
+distribution by program} shows rule distribution by program.  Comparing the
+number of log lines per program and rules per program shows that
+\textit{postfix/qmgr\/} has the highest number of log lines but has a low
+number of rules to try, so the average execution time for a
+\textit{postfix/qmgr\/} is quite low, reducing the overall average
+execution time per log line.
+
+XXX CREATE A GRAPH AVERAGE EXECUTION TIME PER LOG FILE\@.
+
+\subsubsection{Log 22 distribution by rule}
+\label{log22 distribution by rule}
+\includegraphics{plot-log22-distribution-by-rule.ps}
+
+The individual rules are not labelled on the x-axis because doing so makes
+the graph illegible.
+
+\subsubsection{Log 22 distribution by program}
+\label{log22 distribution by program}
+\includegraphics{plot-log22-distribution-by-program.ps}
+
+\subsubsection{Rule distribution by program}
+\label{rule distribution by program}
+\includegraphics{plot-log22-rule-count-by-program.ps}
+
+\subsection{Caching regexs}
+
+\subsubsection{Regex: cached vs discarded}
 \label{normal regex vs discard regex}
@@
 
-\subsection{Regex: percentage execution time increase}
+\subsubsection{Regex: percentage execution time increase}
 \label{normal regex vs disarded regex factor}
@@
 
-\subsection{Normal vs shuffled vs reversed ordering}
-\label{normal vs shuffled vs reversed ordering}
-\includegraphics{plot-normal-shuffle-reverse.ps}
-
-\subsection{Percentage increase of shuffled and reversed over normal}
-\label{normal vs shuffled vs reversed ordering factor}
-\includegraphics{plot-normal-shuffle-reverse-factor.ps}
+Two large dips can be seen in graph~\ref{normal regex vs disarded regex
+factor} at log files 22 and 62--68, corresponding to the spikes in log file
+size in graph~\ref{execution time vs file size vs number lines}.  The
+anomaly at files 62--68 is caused by a combination of a mail loop and a
+delivery failure which resulted in a huge number of log lines, all of which
+were matched by a single rule.  Sorting the rules by the number of
+successful matches has the effect that this rule will be tried first, so
+the average number of regexs compiled per line drops sharply for those
+seven log files, resulting in a much lower increase in relative execution
+time, as detailed below.
+
+\begin{tabular}[]{lrr}
+                                    & Mean      & Standard Deviation    \\
+    All logs                        & 465.58\%  & 114.70\%              \\
+    All logs except 22 and 62--68   & 497.43\%  & 50.58\%               \\
+    Logs 22 and 62--68              & 127.16\%  & 21.28\%               \\
+
+\end{tabular}
 

### 2007-06-26 doc/logparser.tex
Improved the graphs and explanation for the dips in normal/shuffle/reverse; added a better graph and removed the two it replaced. Added all the files necessary for rebuilding - generating some of them again would be painful, and the explanations are tied to the log files used now. Removed unnecessary files.

@@
 THE RESULTING MAILS\@; FALSE POSITIVES, FALSE NEGATIVES, ETC\@.
-
-XXX LOTS MORE REFERENCES
-
-XXX MAYBE ADD A SECTION ABOUT OPTIMISATION OF THE DATABASE\@.
 
@@
 specific rule to match some lines and a more general rule to match the
-remainder, e.g. separating smtp delivery to a filtering proxy from smtp
+remainder, e.g.\ separating smtp delivery to a filtering proxy from smtp
 delivery to the rest of the world.  The algorithm provides a facility for
@@
 
-    \item Sort the regexs and visually inspect the list, e.g. with SQL
+    \item Sort the regexs and visually inspect the list, e.g.\ with SQL
         similar to: \textbf{select regex from rules order by regex;}
@@
 \subsection{Rule efficiency}
-
-XXX WRITE ABOUT DATABASE OPTIMISATION??
 
@@
 
-XXX THIS NEXT PARAGRAPH MIGHT NEED TO BE REWRITTEN\@.
-
 The dips at logfiles 22 and 62--68 correspond to peaks in log file size in
 graph~\ref{execution time vs file size vs number lines}.  The explanation
-took some time to arrive at, but turns out to be reasonably simple.
-Examining the database after parsing logfile 22 shows 64 rules matched
-during parsing, but the distribution of matching rules is extremely skewed,
-as shown in graph~\ref{log22 distribution by rule}.  Graph~\ref{rule
-distribution by program} shows rule distribution by program.  Comparing the
-number of log lines per program and rules per program shows that
-\textit{postfix/qmgr\/} has the highest number of log lines but has a low
-number of rules to try, so the average execution time for a
-\textit{postfix/qmgr\/} is quite low, reducing the overall average
-execution time per log line.
-
-XXX CREATE A GRAPH AVERAGE EXECUTION TIME PER LOG FILE\@.
+for this took some time to arrive at, but it turns out to be reasonably
+simple.  Examining the database after parsing logfile~22 shows 64 rules
+matched during parsing, but the distribution of matching rules is extremely
+skewed, as shown in graph~\ref{log22 distribution by rule}.  Comparing the
+number of log lines per program versus rules per program in
+graph~\ref{log22 line and rule distribution by program} shows that
+\textit{postfix/qmgr\/} has the highest number of log lines; however it has
+a low number of rules, so the average parsing time for a
+\textit{postfix/qmgr\/} log line is quite low, reducing the overall average
+parsing time per log line.
+
+XXX FIGURE OUT THE DIP AT LOG 34, mail.log-20070228
 
@@
 
-\subsubsection{Log 22 distribution by program}
-\label{log22 distribution by program}
-\includegraphics{plot-log22-distribution-by-program.ps}
-
-\subsubsection{Rule distribution by program}
-\label{rule distribution by program}
-\includegraphics{plot-log22-rule-count-by-program.ps}
+\subsubsection{Log 22 line and rule distribution by program}
+\label{log22 line and rule distribution by program}
+\includegraphics{plot-log22-distribution-versus-rule-count.ps}
 

### 2007-06-29 doc/logparser.tex
Wrote a large chunk of the parsing coverage section; line coverage is pretty much done, mail coverage is started but needs lots of work (requires changes and improvements to the parser). Documented new actions. Documented situations when smtpd stops logging. Added context free grammar section. Attempted to make Latex accept the URLs in the bibliography without complaining. Misc cleanup.

@@
 
+% Useful stuff for math mode.
+\usepackage{amstext}
+% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
+\usepackage{url}
+\DeclareUrlCommand\urlwrap{}
+\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
 % Include images
@@
 % When creating a PDF make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.
+% (without horrible red borders) and include bookmarks.  The title and
+% author don't work - I think either gnuplot or graphviz clobbers it.
 \usepackage{hyperref}
@@
 matching line.  Other restrictions determine their result in different
-ways, e.g. \texttt{reject\_rbl\_client some.rbl.com} checks whether the IP
+ways, e.g.\ \texttt{reject\_rbl\_client some.rbl.com} checks whether the IP
 address of the client is present in the listed RBL\@.
@@
 housekeeping duties.  Readers familiar with context-free grammars may find
-Section~\ref{comparison against context-free grammars} helpful.
+section~\ref{comparison against context-free grammars} helpful.
 
@@
 
-\subsection{Comparison against context-free grammars}
-
 \label{comparison against context-free grammars}
 
-XXX DESCRIBE IT IN LEFT\@: RIGHT PRODUCTION TERMS, IF POSSIBLE\@.
+In context-free grammar terms the parser rules could be described as:
+
+$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
+\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
 
@@
         rules matching less commonly occurring log lines, lowering the
-        program's execution time.  Rule efficiency concerns are discussed
-        in section~\ref{rule ordering for efficiency}.
+        program's execution time.  Rule ordering for efficiency is
+        discussed in section~\ref{rule ordering for efficiency}.
 
@@
 
+    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
+        it kills all smtpds,\footnote{Possibly other programs are killed
+        also, but the parser is only affected by and interested in smtpd.}
+        requiring any active connections to be committed and deleted from
+        the state tables.
+
+    \item [SMTPD\_KILLED] Sometimes Postfix needs to forcefully kill an
+        smtpd; the active connection for that smtpd must be committed and
+        deleted from the state tables.
+
+    \item [SMTPD\_DIED] Sometimes an smtpd dies or exits unsuccessfully;
+        the active connection for that smtpd must be committed and deleted
+        from the state tables.
+
 \end{description}
@@
 \subsection{Additional complications}
+
+\label{additional complications}
 
@@
         \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g. \texttt{20070321125732.D168138A1@smtp.cs.tcd.ie}
+        e.g.\ \texttt{20070321125732.D168138A1@smtp.cs.tcd.ie}
 
@@
 
+\subsubsection{Smtpd stops logging}
+
+\label{smtpd stops logging}
+
+Occasionally an smtpd will just stop logging, without any explicit reason.
+After poring over logs for some time there are several reasons for this
+infrequent occurrence:
+
+\begin{enumerate}
+
+    \item Postfix is stopped or its configuration is reloaded.  When this
+        happens all smtpds exit, and all entries in the connections state
+        table must be committed and deleted from the state table.  The
+        queueid state table is untouched.
+
+    \item Postfix sometimes forcibly kills an smtpd, so the active
+        connection must be committed and deleted from the connections state
+        table.
+
+    \item Occasionally an smtpd will die or exit uncleanly, so the active
+        connection must be committed and deleted from the connections state
+        table.
+
+\end{enumerate}
+
+The above appear to cover all situations where an smtpd suddenly stops
+logging.
+
 \section{Parsing coverage}
 
-XXX THIS SECTION NEEDS TO BE WRITTEN
+The discussion of the parser's coverage of Postfix log files is separated
+into two parts: log lines covered and mails covered.  The first is
+important because the parser should handle all (relevant) log lines it's
+given; the second is equally important because the parser must properly
+deal with every mail if it is to be useful.  Improving the former is
+less intrusive, as it just requires new rules to be written; improving the
+latter is much more intrusive as it requires changes to the parser
+algorithm, and it can also be much harder to notice a deficiency.
 
@@
 
+Parsing a log line is a three stage process:
+
+\begin{enumerate}
+
+    \item Check if there are any rules for the program which produced the
+        log line; if not then skip the line.
+
+    \item Try each rule, ordered by priority and rule order (see
+        sections~\ref{rule attributes} and~\ref{rule ordering for
+        efficiency}), until a matching rule is found.
+
+    \item Execute the appropriate action.
+
+\end{enumerate}
+
+Full coverage of log lines requires the following:
+
+\begin{enumerate}
+
+    \item Each program of interest must have at least one rule or it's log
+        lines will be silently skipped; in the extreme case of zero rules
+        the parser would happily skip every log line.  There may be any
+        number of log lines from other programs intermingled in the log
+        file, and there are some Postfix programs which don't produce any
+        log lines of interest (currently \textit{postfix/anvil\/} and
+        \textit{postfix/scache\/}).
+
+    \item There must be a rule to match each different line produced by
+        each program; if a line is not successfully matched the parser will
+        issue a warning.  Rules should be as specific and tightly bound as
+        possible to ensure accurate parsing:\footnote{A rule which matches
+        zero or more of any character will successfully parse every log
+        line, but not in a meaningful way.} most log lines contain fixed
+        strings and have a rigid pattern, so this is not a problem.
+
+    \item The appropriate action to take --- this is discussed in
+        section~\ref{mails-covered}.
+
+\end{enumerate}
+
+Full coverage of log lines is both easy to achieve and hard to maintain.
+It is easy to achieve full coverage for a limited set of log files (at the
+time of writing the parser has 131 rules, fully parsing 93 contiguous log
+files from Postfix 2.2 and 2.3), and new rules are easy to add.
+Maintaining full coverage is hard because other servers have different
+restrictions with custom messages, RBLs change their messages over time,
+major releases of Postfix change warning messages (usually adding more
+information), etc.\ so over time the log lines drift and change.
+
+XXX GRAPH RULE HITS
+
+\subsection{Mails covered}
+
+\label{mails-covered}
+
 XXX WRITE THIS
 
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-XXX WRITE THIS
+Coverage of mails is much more difficult to determine accurately.  The
+parser can dump its state tables in a human readable form; examining these
+tables with reference to the log files is the best way to detect mails
+which were not handled properly (all the complications discussed in
+section~\ref{additional complications} were detected in this way).  The
+parser issues warnings when it detects any errors, some of which may alert
+the user to a problem, e.g.\ when a queueid is reused before the previous
+mail is fully dealt with, when a queueid/pid is not found, or when there
+are problems tracking a child mail (see
+section~\ref{tracking-re-injected-mail}).  There should be few or no
+warnings (warnings about mail where some of the log lines are not in the
+logfile can be ignored), and the state table should only contain entries
+for mails which had yet to be delivered when the logfiles ended.
+
+At the time of writing the parser is being tested with 93 log files.  The
+state table contains entries for mails not yet delivered at the end of the
 

### 2007-06-30 doc/logparser.tex
Documented the action that handles watchdog timeouts. Expanded the mails covered section, listing the warnings from test-some. Removed limitations and improvements which no longer apply.

@@
 
+    \item [SMTPD\_WATCHDOG] Smtpds have a watchdog timer to deal with
+        unusual situations --- after five hours the timer will expire and
+        the smtpd will exit.  This occurs very infrequently, as there are
+        many other timeouts which should occur in the intervening hours:
+        DNS timeouts, timeouts reading data, etc.  The active connection
+        for that smtpd must be committed and deleted from the state tables.
+
 \end{description}
@@
 
+Coverage of mails is much more difficult to determine accurately than
+coverage of lines.  The parser can dump its state tables in a human
+readable form; examining these tables with reference to the log files is
+the best way to detect mails which were not handled properly (all the
+complications discussed in section~\ref{additional complications} were
+diagnosed and sometimes detected in this way).  The parser issues warnings
+when it detects any errors, some of which may alert the user to a problem,
+e.g.\ when a queueid is reused before the previous mail is fully dealt
+with, when a queueid or pid is not found,\footnote{There will often be a
+few warnings about missing queueids or pids in the first few hundred or
+thousand log lines because the earlier log lines are in the previous log
+file; loading the saved state from the previous log file will solve this
+problem} or when there are problems tracking a child mail (see
+section~\ref{tracking-re-injected-mail}).  There should be few or no
+warnings, and the state table should only contain entries for mails which
+had yet to be delivered when the logfiles ended.
+
+At the time of writing the parser is being tested with 93 log files.  There
+are 25 warnings produced, but because the parser errs on the side of
+producing more warnings rather than fewer those 25 warnings represent only
+8 problems:
+
+\begin{itemize}
+
+    \item 3 connections (5 warnings) started before the first log file, so
+        their initial log entries are missing.
+
+    \item 5 mails (20 warnings) have out of order log entries, leading to
+        warnings about missing data and queueids.  The two log files where
+        the out of order log lines occur are more than an order of
+        magnitude larger than normal, due to the extremely large volume of
+        mail handled by the server on those days; process scheduling
+        vagaries and the undefined order of messages sent to syslog(8) may
+        explain why the lines were out of order.
+
+\end{itemize}
+
+The first of these problems cannot be dealt with; although out of order log
+lines (the second problem) are dealt with in other circumstances (see
+sections~\ref{tracking-re-injected-mail}, \ref{discarding-cleanup-lines}
+and \ref{pickup-logging-after-cleanup}), dealing with these lines would
+introduce a large amount of complication (deferred warnings, deferred
+committal, ambiguity) for a problem which occurs only four times in
+60721709 log lines.
+
 XXX WRITE THIS
 
-Coverage of mails is much more difficult to determine accurately.  The
-parser can dump its state tables in a human readable form; examining these
-tables with reference to the log files is the best way to detect mails
-which were not handled properly (all the complications discussed in
-section~\ref{additional complications} were detected in this way).  The
-parser issues warnings when it detects any errors, some of which may alert
-the user to a problem, e.g.\ when a queueid is reused before the previous
-mail is fully dealt with, when a queueid/pid is not found, or when there
-are problems tracking a child mail (see
-section~\ref{tracking-re-injected-mail}).  There should be few or no
-warnings (warnings about mail where some of the log lines are not in the
-logfile can be ignored), and the state table should only contain entries
-for mails which had yet to be delivered when the logfiles ended.
-
-At the time of writing the parser is being tested with 93 log files.  The
+The
 state table contains entries for mails not yet delivered at the end of the
@@
         new administrator defined rejection messages require new rules.
-
-    \item The program doesn't cope with mails which span log files:
-        currently they are reported as unfinished mails at the end of the
-        first run and are complained about when encountered during the next
-        run because they don't have a source.
 
@@
 
+        XXX EXPAND ON THIS
+
     \item The program will not detect parsing the same log file twice,
@@
     \item Write the policy servers referred to in the limitations above.
-
-    \item Save state at the end of each run; load it at the start of the
-        next run.  This should allow the program to cope with mails
-        spanning log files.  This might also require purging old mails once
-        they've been in the state tables for too long, though the presence
-        of very old mails indicates a bug in either the implementation or
-        the specification.
 

### 2007-07-01 doc/logparser.tex
Renamed rule_order to hits; added hits_total; updated paper appropriately. Now graphing rule hits.

@@
 
-    \item [rule\_order] is an efficiency measure.  This counter is
-        maintained for every rule and incremented each time the rule
-        successfully matches.  At the start of each run the program sorts
-        the rules in descending rule\_order, and at the end of the run
-        updates every rule's rule\_order.  Assuming that the distribution
-        of log lines is reasonably consistent between log files, rules
-        matching more commonly occurring log lines will be tried before
-        rules matching less commonly occurring log lines, lowering the
-        program's execution time.  Rule ordering for efficiency is
-        discussed in section~\ref{rule ordering for efficiency}.
-
-    \item [priority] This is the user-configurable companion to
-        rule\_order: rules will be tried in order of priority, overriding
-        rule\_order.  This allows more specific rules to take precedence
-        over more general rules.
+    \item [hits] is an efficiency measure.  This counter is maintained for
+        every rule and incremented each time the rule successfully matches.
+        At the start of each run the program sorts the rules in descending
+        hits, and at the end of the run updates every rule's hits.
+        Assuming that the distribution of log lines is reasonably
+        consistent between log files, rules matching more commonly
+        occurring log lines will be tried before rules matching less
+        commonly occurring log lines, lowering the program's execution
+        time.  Rule ordering for efficiency is discussed in
+        section~\ref{rule ordering for efficiency}.
+
+    \item [hits\_total] The total number of hits for this rule over all
+        runs of the parser.
+
+    \item [priority] This is the user-configurable companion to hits: rules
+        will be tried in order of priority, overriding hits.  This allows
+        more specific rules to take precedence over more general rules.
 
@@
 zero).  Rules are tried in order of priority, highest first, overriding the
-rule\_order field maintained by the parser.  Negative priorities may also
-be useful for catchall rules.
+hits field maintained by the parser.  Negative priorities may also be
+useful for catchall rules.
 
@@
 queueid             & 1                                                 \\
-rule\_order         & 0                                                 \\
+hits                & 0                                                 \\
+hits\_total         & 0                                                 \\
 priority            & 0                                                 \\
@@
         it.
+
+    \item [hits, hits\_total and priority:] hits and priority are used in
+        ordering the rules (see section~\ref{rule ordering for
+        efficiency}); hits\_total is updated at the end of each run but is
+        otherwise unused.
 
@@
 
-XXX DO A FULL RUN ONCE RULE\_ORDER ISN'T BEING TRASHED AND GRAPH RULE
-HITS\@.
-
-Rule ordering was briefly discussed in section~\ref{rule attributes} and
-it will be covered in greater detail in this section.  At the time of
-writing there are 131 different rules, some of which match much more
-frequently than others.  Assuming that the distribution of log lines is
-reasonably steady over time, program efficiency should benefit from trying
-more frequently matching rules before those which match less frequently.
-To test this hypothesis three full test runs were performed with different
-rule orderings:
+Rule ordering was briefly discussed in section~\ref{rule attributes} and it
+will be covered in greater detail in this section.  At the time of writing
+there are 131 different rules, the majority of which match several orders
+of magnitude less frequently than the top 10\%, as shown in graph~\ref{rule
+hits}.  Assuming that the distribution of log lines is reasonably steady
+over time, program efficiency should benefit from trying more frequently
+matching rules before those which match less frequently.  To test this
+hypothesis three full test runs were performed with different rule
+orderings:
 
@@
         both qmgr and cleanup due to out of order log lines; see
-        section~\ref{discarding-cleanup-lines} for details.
+        section~\ref{discarding cleanup lines} for details.
 
@@
         caused the cleanup line to be logged first, as discussed in
-        section~\ref{pickup-logging-after-cleanup}.
+        section~\ref{pickup logging after cleanup}.
 
@@
 This complication is further complicated by out of order cleanup lines: see
-section~\ref{discarding-cleanup-lines} for details.
+section~\ref{discarding cleanup lines} for details.
 
@@
 
-\label{discarding-cleanup-lines}
+\label{discarding cleanup lines}
 
@@
 
-\label{pickup-logging-after-cleanup}
+\label{pickup logging after cleanup}
 
@@
 information), etc.\ so over time the log lines drift and change.
-
-XXX GRAPH RULE HITS
+Graph~\ref{rule hits} shows the hits for each rule over all 93 log files;
+it's obvious that a small number of rules match the vast majority of the
+lines, and more than half the rules match fewer than 100 times.
 
@@
 lines (the second problem) are dealt with in other circumstances (see
-sections~\ref{tracking-re-injected-mail}, \ref{discarding-cleanup-lines}
-and \ref{pickup-logging-after-cleanup}), dealing with these lines would
-introduce a large amount of complication (deferred warnings, deferred
+sections~\ref{tracking-re-injected-mail},~\ref{discarding cleanup
+lines}~and~\ref{pickup logging after cleanup}), dealing with these lines
+would introduce a large amount of complication (deferred warnings, deferred
 committal, ambiguity) for a problem which occurs only four times in
@@
 
+\subsubsection{Rule hits}
+\label{rule hits}
+\includegraphics{plot-hits.ps}
+
 \subsubsection{Normal vs shuffled vs reversed ordering}

### 2007-07-02 doc/logparser.tex
Improved the explanation of dips in the normal/shuffle/reverse factor graph.

@@
 for this took some time to arrive at, but it turns out to be reasonably
-simple.  Examining the database after parsing logfile~22 shows 64 rules
-matched during parsing, but the distribution of matching rules is extremely
-skewed, as shown in graph~\ref{log22 distribution by rule}.  Comparing the
-number of log lines per program versus rules per program in
-graph~\ref{log22 line and rule distribution by program} shows that
-\textit{postfix/qmgr\/} has the highest number of log lines; however it has
-a low number of rules, so the average parsing time for a
-\textit{postfix/qmgr\/} log line is quite low, reducing the overall average
-parsing time per log line.
-
-XXX FIGURE OUT THE DIP AT LOG 34, mail.log-20070228
+simple.  Examining the database after parsing logfile~22 in isolation shows
+64 rules matched during parsing, but as shown in graph~\ref{log22
+distribution by rule} the distribution of matching rules is extremely
+skewed.  Comparing the number of log lines per program versus rules per
+program in graph~\ref{log22 line and rule distribution by program} shows
+that \textit{postfix/qmgr\/} has the highest number of log lines, however
+it has a low number of rules.  This combination results in a low average
+number of rules to try for \textit{postfix/qmgr\/} log lines, reducing the
+overall average number of rules to try per log line.  In turn this reduces
+the average parsing time per log line, and thus the parsing time for the
+log file.  While the parsing time is substantially greater than the time
+for the log files on either side, the increase from optimal ordering to
+reversed ordering is proportionally smaller, producing the dips in
+graph~\ref{normal vs shuffled vs reversed ordering factor}.
 
@@
 the average number of regexs compiled per line drops sharply for those
-seven log files, resulting in a much lower increase in relative execution
+eight log files, resulting in a much lower increase in relative execution
 time, as detailed below.

### 2007-07-05 doc/logparser.tex
Parse postfix/bounce messages.  Might need some bugs shaken out of it. Change the time period for discarding cleanup lines to seven minutes because evidence from the logs shows it needs to be that high. Log the number of entries in each data source when dumping state; change the Makefile to reflect that. Save the date in results, to make debugging easier. Refactored my_warn() and my_die() to use format_error(); added a backtrace to messages too. Updated Parser documentation appropriately. Updated the paper appropriately. Expanded the mail coverage section with a breakdown of the remaining stat

@@
 
+    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) log the
+        creation of bounce messages.  This action creates a new mail if
+        necessary, marking it as a bounce notification and removing the
+        unknown origin flag if it exists.
+
 \end{description}
@@
 correctly.
+
+Postfix 2.3 (and hopefully subsequent version) log the creation of a bounce
+message.
 
@@
 
-    \item The timestamp of the cleanup line must be within six minutes of
+    \item The timestamp of the cleanup line must be within seven minutes of
         the mail acceptance timestamp.  Timeouts happen after five minutes,
-        so six minutes gives one minute of leeway.
+        but some data may have been transferred slowly, and empirical
+        evidence suggests seven minutes is a good compromise between false
+        positives and false negatives.
 
@@
 
+XXX REDO THIS
+
 \begin{itemize}
@@
 
-The
-state table contains entries for mails not yet delivered at the end of the
+The state table contains entries for mails not yet delivered when the
+parser finishes execution.  Ideally all it should contain are mails which
+are awaiting delivery after the period covered by the logs, though it may
+also contain mails from the start of the logs whose initial entries are not
+contained in the logs.  Any other entries are evidence of a failure in
+parsing or an aberration in the logs.  After parsing the 93 test log files
+the state table contains XXX entries, breaking down into:
+
+\begin{itemize}
+
+    \item 1 connection which started only seconds before the logs ended and
+        had not completed.
+
+    \item 9 mails which had yet to be delivered due to repeated failures.
+
+    \item 4 mails resulting from the out of order log entries described
+        above.
+
+    \item 9 mails whose initial log entries are not present in the logs.
+
+\end{itemize}
 

### 2007-07-09 doc/logparser.tex
populate-rules.sql: Make the tracking rule provide defaults for {server,client}_{hostname,ip} to handle forwarded bounce notifications. logparser.tex: Updated to match changes.  May need a little more added, dunno yet. Parser.pm: * Timeout for discarding cleanup lines increased to 10 minutes. * Added cloned_by * Tidy up the smtpd and last cloned mail using tidy_after_timeout() whenever there's a timeout, smtpd dies, is killed, or watchdog triggers - if the pid still exists (sometimes it doesn't, dunno why, not too bothered).  Missing cloned mail won't be committed now, reducing warnings. Also

@@
 
-    \item The mail will not have been tagged as a tracked mail (see
-        section~\ref{tracking-re-injected-mail}).
-
     \item The message-id has a specific format: \newline
@@
 
-    \item The timestamp of the cleanup line must be within seven minutes of
+    \item The timestamp of the cleanup line must be within ten minutes of
         the mail acceptance timestamp.  Timeouts happen after five minutes,
         but some data may have been transferred slowly, and empirical
-        evidence suggests seven minutes is a good compromise between false
-        positives and false negatives.
+        evidence shows that ten minutes is not unreasonable; hopefully it
+        is a good compromise between false positives and false negatives.
 
@@
 
-Occasionally an smtpd will just stop logging, without any explicit reason.
-After poring over logs for some time there are several reasons for this
-infrequent occurrence:
+Occasionally an smtpd will just stop logging, without an immediately
+obvious reason.  After poring over logs for some time there are several
+reasons for this infrequent occurrence:
 
@@
 The above appear to cover all situations where an smtpd suddenly stops
-logging.
+logging.  In addition to removing an active connection the last accepted
+mail may need to be discarded, as detailed in
+section~\ref{timeouts-during-data-phase}.
 

### 2007-07-09 doc/logparser.tex
Updated the coverage section to reflect progress on reducing warnings and state table entries.

@@
 At the time of writing the parser is being tested with 93 log files.  There
-are 25 warnings produced, but because the parser errs on the side of
+are 28 warnings produced, but because the parser errs on the side of
 producing more warnings rather than fewer those 25 warnings represent only
-8 problems:
-
-XXX REDO THIS
+10 problems:
 
@@
 
-    \item 5 mails (20 warnings) have out of order log entries, leading to
+    \item 7 mails (23 warnings) have out of order log entries, leading to
         warnings about missing data and queueids.  The two log files where
@@
 would introduce a large amount of complication (deferred warnings, deferred
-committal, ambiguity) for a problem which occurs only four times in
+committal, ambiguity) for a problem which occurs only seven times in
 60721709 log lines.
 
-XXX WRITE THIS
-
-The state table contains entries for mails not yet delivered when the
-parser finishes execution.  Ideally all it should contain are mails which
-are awaiting delivery after the period covered by the logs, though it may
-also contain mails from the start of the logs whose initial entries are not
-contained in the logs.  Any other entries are evidence of a failure in
-parsing or an aberration in the logs.  After parsing the 93 test log files
-the state table contains XXX entries, breaking down into:
+The state tables contain entries for mails not yet delivered when the
+parser finishes execution.  Ideally all they should contain are mails which
+are awaiting delivery after the period covered by the logs, though they may
+also contain mails whose initial entries are not contained in the logs.
+Any other entries are evidence of a failure in parsing or an aberration in
+the logs.  After parsing the 93 test log files the state tables contain 27
+entries, breaking down into:
 
@@
 
-    \item 1 connection which started only seconds before the logs ended and
-        had not completed.
-
-    \item 9 mails which had yet to be delivered due to repeated failures.
-
-    \item 4 mails resulting from the out of order log entries described
+    \item 2 connections which started only seconds before the logs ended
+        and had not completed.
+
+    \item 7 mails which had yet to be delivered due to repeated failures.
+
+    \item 5 mails resulting from the out of order log entries described
         above.
@@
     \item 9 mails whose initial log entries are not present in the logs.
+
+    \item 4 mails present for unknown reasons.
 

### 2007-07-11 doc/logparser.tex
Added support for generic rules which will be tried after program specific rules. Documented generic rules. Changed all the LDAP rules to be generic rules.

@@
         trying rules which won't match the line, or worse, might match
-        unintentionally.
+        unintentionally.  Rules whose program is \texttt{*} will be tried
+        against any lines which aren't parsed by program specific rules.
 
@@
 
-    \item 7 mails (23 warnings) have out of order log entries, leading to
+    \item 5 mails (20 warnings) have out of order log entries, leading to
         warnings about missing data and queueids.  The two log files where
@@
 would introduce a large amount of complication (deferred warnings, deferred
-committal, ambiguity) for a problem which occurs only seven times in
+committal, ambiguity) for a problem which occurs only five times in
 60721709 log lines.
@@
 Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the 93 test log files the state tables contain 27
+the logs.  After parsing the 93 test log files the state tables contain 25
 entries, breaking down into:
@@
 
-    \item 2 connections which started only seconds before the logs ended
+    \item 1 connection which started only seconds before the logs ended
         and had not completed.
@@
 
-    \item 5 mails resulting from the out of order log entries described
+    \item 4 mails resulting from the out of order log entries described
         above.
@@
         logged if the mail is accepted.\footnote{Tested with Postfix 2.2.10
-        and 2.3.7; this may possibly have changed in Postfix 2.4.}  It
+        and 2.3.11; this may possibly have changed in Postfix 2.4.}  It
         should be reasonably simple to write a policy server which causes
@@
         logging.
-
-        XXX EXPAND ON THIS
 

### 2007-07-14 doc/logparser.tex
Wrote prior art section discussing Log Mail Analyser. Documented handling out of order log lines using valid combos. Updated mail coverage section now that valid combos are working and have reduced the number of warnings and mails in the state tables. Some cleanups. Another attempt to break URLs properly.

@@
     pdfborder   = {0 0 0},
-    breaklinks  = true,
+%    breaklinks  = true,
 }
+\usepackage{breakurl}
 
@@
 XXX MAYBE A PRIOR ART SECTION\@?
-
-XXX TURN OFF RESTRICTIONS FOR ME AND SEE HOW WELL SPAMASSASSIN DEALS WITH
-THE RESULTING MAILS\@; FALSE POSITIVES, FALSE NEGATIVES, ETC\@.
 
@@
 
+\subsection{Previous research in this area}
+
+There only appears to be one prior paper published about parsing Postfix
+log files: \textit{Log Mail Analyzer: Architecture and Practical
+Utilizations\/}~\cite{log-mail-analyser}.  The aim of Log Mail Analyzer is
+quite different from his parser: it attempts to present correlated log data
+in a form suitable for a systems administrator to search using grep(1),
+producing a CSV file\footnote{Comma-Separated Value, a very simple format
+where a record is stored in a single line, with fields separated by a comma
+or other punctuation symbol.  Problems with CSV files include the
+difficulty in escaping separators, providing multiple values for a field
+(e.g.\ multiple recipients), adding new fields, and lack of a bundled schema
+describing the fields.} which can subsequently be imported into a MySQL or
+Berkeley DB database.  The data stored is limited to time and date,
+hostname and IP address of client, sender and recipient addresses, and the
+SMTP code.  The decision to support both CSV and Berkeley DB appears to
+have been a limiting factor: the values which can be stored are limited,
+it's difficult to link different records (foreign keys in SQL enable this
+easily), and ad-hoc queries require extensive code to be written.  Berkeley
+DB is not an SQL database, it only supports (key, value) pairs; in the main
+table the value used is a CSV line, and the key is an integer referred to
+by other tables.  The secondary by-sender, by-recipient, by-date and by-IP
+tables use the sender/recipient/date/IP as the key, and a CSV list of
+integers referring to the main table is used as the value.  In effect this
+re-implements SQL foreign keys, but without the full functionality offered
+by even the most basic of SQL databases.  It also requires custom code to
+search on some combination of the above.
+
+Handling of multiple recipients, SMTP codes or remote servers\footnote{A
+single mail may be sent to multiple remote servers if it was addressed to
+recipients in different domains, or Postfix needs to try multiple servers
+for one or more recipients.} is not explained in the paper, but the options
+appear to be:
+
+\begin{enumerate}
+
+    \item Discard some of the data; this option is obviously undesirable,
+        and presumably was not chosen by the LMA authors.
+
+    \item Combine all data into a CSV line, which is then embedded into the
+        main CSV line for that mail, making searching even more difficult
+        and requiring a second separating character.
+
+\end{enumerate}
+
+The schema used with the MySQL database is undocumented, but at least it's
+possible to discover the schema with an SQL database, unlike with Berkeley
+DB\@.
+
+Parsing is hand-coded and requires major changes to the code to parse new
+log lines or extract additional data.  It does not appear to deal with any
+of the complications discussed in
+sections~\ref{complications}~and~\ref{additional complications}.  The
+program fails to parse Postfix 2.2.x or 2.3.x log files; indeed it produces
+a number of warnings and no usable output.
+
+In summary Log Mail Analyser appears to be a proof of concept, written to
+demonstrate the point of the paper (that having this information in an
+accessible fashion is useful to systems administrators), rather than a
+program designed to be extensible and useful in a production environment.
+
+% Literature review notes:
+%
+% Hard-coded parsing, requiring code changes to add more.  Attempts to
+% correlate log lines, saves data to database for data mining purposes.
+% Hard to extend/expand/understand.  Appears to only save: date and hour,
+% DNS name and IP address host, mail server IP address, sender, receiver
+% and e-mail status (sent, rejected).  Undocumented schema.  Design
+% decision to use CSV as an intermediate format between the log file and
+% the database seems to have been restrictive.  Appears to require a
+% queueid but majority of log entries (e.g. rejections) lack a queueid.
+% Supports whitelisting IP addresses when parsing logs, but whitelisting
+% when generating reports/data mining would be preferable.  Supporting
+% Berkeley DB is probably limiting the software - an example is the
+% difficulty in searching a pipe-delimited string, so they have
+% re-implemented foreign keys with tables keyed by ip address etc. pointing
+% at the main table - this also won't scale well.  There doesn't appear to
+% be any attempt to deal with the complications I've encountered: their
+% parsing isn't detailed enough to encounter them.  It doesn't run
+% properly; doesn't create any output; throws up errors.
+
 \section{Adaptable parsing: user defined rules}
@@
 The effect of this complication upon the algorithm is discussed in
-section~\ref{tracking-re-injected-mail}.
+section~\ref{tracking re-injected mail}.
 
@@
 overlaps both the mail acceptance and mail delivery sections.  See
-section~\ref{tracking-re-injected-mail} for a full discussion.
+section~\ref{tracking re-injected mail} for a full discussion.
 
@@
         tracked by the parent mail (action: track; flowchart:~11) ---
-        section~\ref{tracking-re-injected-mail} discusses this in detail.
+        section~\ref{tracking re-injected mail} discusses this in detail.
 
@@
         database (flowchart:~14), then deleted from memory (flowchart:~15).  For
-        re-injected mails see section~\ref{tracking-re-injected-mail}.
+        re-injected mails see section~\ref{tracking re-injected mail}.
 
@@
 to the parent or children of mail re-injected due to forwarding, and that
-is explained in section~\ref{tracking-re-injected-mail}.
+is explained in section~\ref{tracking re-injected mail}.
 
@@
 
-\label{tracking-re-injected-mail}
+\label{tracking re-injected mail}
 
@@
 problem of re-injected mails discussed in
-section~\ref{tracking-re-injected-mail}, but unlike the solution described
+section~\ref{tracking re-injected mail}, but unlike the solution described
 therein bounce notifications do not eventually have a log line which
@@
 section~\ref{timeouts-during-data-phase}.
+
+\subsubsection{Out of order log lines}
+
+\label{out of order log lines}
+
+Occasionally a log file will have out of order log lines which cannot be
+dealt with by the techniques described in sections~\ref{tracking
+re-injected mail},~\ref{discarding cleanup lines}~and~\ref{pickup logging
+after cleanup}.  In the 93 log files used for testing this occurs only five
+times in 60721709 log lines, but for completeness of the algorithm it
+should be dealt with.  The five occurrences in the test log files have the
+same characteristics: the postfix/local log line showing delivery to a
+local mailbox occurs after the postfix/qmgr log line showing removal of the
+mail from the queue because delivery is completed.  This causes problems:
+the mail is not complete, so entry into the database fails; a new mail is
+created when the postfix/local line is parsed and remains in the state
+tables; four warnings are issued per pair of out of order log lines.
+
+The solution to this problem is to examine the list of programs which have
+logged messages for each mail, comparing the list against a table of
+known-good combinations of programs.  If combination is found in the valid
+list the mail can be entered in the database; if the combination is not
+found entry must be postponed and the mail flagged.  The SAVE\_BY\_QUEUEID
+action checks for the flag and retries entry; if the additional log lines
+have caused the mail to reach a valid combination entry will succeed,
+otherwise it must be postponed once more.
+
+The list of valid combinations is given below.  Every mail will
+additionally have log entries from postfix/cleanup and postfix/qmgr; any
+mail may also have log entries from postfix/bounce, postfix/postsuper, or
+both.
+
+\begin{description}
+
+    \item [postfix/local:] Local delivery of bounce notification, or local
+        delivery of forwarded/tracked mail.
+
+    \item [postfix/local, postfix/pickup:] Locally submitted mail, locally
+        delivered.
+
+    \item [postfix/local, postfix/pickup, postfix/smtp:] Locally submitted
+        mail, \newline both local and remote delivery.
+
+    \item [postfix/local, postfix/smtp, postfix/smtpd:] Mail accepted from
+        a remote client, both local and remote delivery.
+
+    \item [postfix/local, postfix/smtpd:] Mail accepted from a remote
+        client, local delivery.
+
+    \item [postfix/pickup, postfix/smtp:] Locally submitted mail, remote
+        delivery.
+
+    \item [postfix/smtp:] Remote delivery of forwarded/tracked mail.
+
+    \item [postfix/smtp, postfix/smtpd:] Mail accepted from a remote
+        client, remotely delivered (relaying mail for clients on the local
+        network).
+
+\end{description}
+
+This applies to accepted mails only, not to rejected mails.
 
@@
 problem} or when there are problems tracking a child mail (see
-section~\ref{tracking-re-injected-mail}).  There should be few or no
+section~\ref{tracking re-injected mail}).  There should be few or no
 warnings, and the state table should only contain entries for mails which
@@
 At the time of writing the parser is being tested with 93 log files.  There
-are 28 warnings produced, but because the parser errs on the side of
-producing more warnings rather than fewer those 25 warnings represent only
-10 problems:
-
-\begin{itemize}
-
-    \item 3 connections (5 warnings) started before the first log file, so
-        their initial log entries are missing.
-
-    \item 5 mails (20 warnings) have out of order log entries, leading to
-        warnings about missing data and queueids.  The two log files where
-        the out of order log lines occur are more than an order of
-        magnitude larger than normal, due to the extremely large volume of
-        mail handled by the server on those days; process scheduling
-        vagaries and the undefined order of messages sent to syslog(8) may
-        explain why the lines were out of order.
-
-\end{itemize}
-
-The first of these problems cannot be dealt with; although out of order log
-lines (the second problem) are dealt with in other circumstances (see
-sections~\ref{tracking-re-injected-mail},~\ref{discarding cleanup
-lines}~and~\ref{pickup logging after cleanup}), dealing with these lines
-would introduce a large amount of complication (deferred warnings, deferred
-committal, ambiguity) for a problem which occurs only five times in
-60721709 log lines.
+are 5 warnings produced, but because the parser errs on the side of
+producing more warnings rather than fewer those 5 warnings represent only 3
+instances of 1 problem: 3 connections started before the first log file, so
+their initial log entries are missing, leading to warnings when their log
+lines are parsed.
 
@@
 Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the 93 test log files the state tables contain 25
+the logs.  After parsing the 93 test log files the state tables contain 22
 entries, breaking down into:
@@
 
-    \item 1 connection which started only seconds before the logs ended
+    \item 2 connections which started only seconds before the logs ended
         and had not completed.
 
+    \item 9 mails whose initial log entries were not present in the logs.
+
     \item 7 mails which had yet to be delivered due to repeated failures.
-
-    \item 4 mails resulting from the out of order log entries described
-        above.
-
-    \item 9 mails whose initial log entries are not present in the logs.
 

### 2007-07-22 doc/logparser.tex
Updated coverage section now that there are no unknown mails in the state table. Documented logging HELOs.

@@
 REVIEWED
-
-XXX MAYBE A PRIOR ART SECTION\@?
 
@@
 readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (all the
+the best way to detect mails which were not handled properly (many of the
 complications discussed in section~\ref{additional complications} were
-diagnosed and sometimes detected in this way).  The parser issues warnings
-when it detects any errors, some of which may alert the user to a problem,
-e.g.\ when a queueid is reused before the previous mail is fully dealt
-with, when a queueid or pid is not found,\footnote{There will often be a
-few warnings about missing queueids or pids in the first few hundred or
-thousand log lines because the earlier log lines are in the previous log
-file; loading the saved state from the previous log file will solve this
-problem} or when there are problems tracking a child mail (see
-section~\ref{tracking re-injected mail}).  There should be few or no
-warnings, and the state table should only contain entries for mails which
-had yet to be delivered when the logfiles ended.
+detected in this way).  The parser issues warnings when it detects any
+errors, some of which may alert the user to a problem, e.g.\ when a queueid
+is reused before the previous mail is fully dealt with, when a queueid or
+pid is not found,\footnote{There will often be warnings about missing
+queueids or pids in the first few hundred or thousand log lines because the
+earlier log lines are in the previous log file; loading the saved state
+from the previous log file will solve this problem.} or when there are
+problems tracking a child mail (see section~\ref{tracking re-injected
+mail}).  There should be few or no warnings, and the state table should
+only contain entries for mails which had yet to be delivered when the
+logfiles ended.
 
@@
 are 5 warnings produced, but because the parser errs on the side of
-producing more warnings rather than fewer those 5 warnings represent only 3
+producing more warnings rather than fewer those 5 warnings represent 3
 instances of 1 problem: 3 connections started before the first log file, so
@@
 Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the 93 test log files the state tables contain 22
+the logs.  After parsing the 93 test log files the state tables contain 18
 entries, breaking down into:
@@
 
-    \item 2 connections which started only seconds before the logs ended
-        and had not completed.
+    \item 1 connection which started only seconds before the logs ended and
+        had not yet completed.
+
+    \item 1 mail which had been accepted only seconds before the logs ended
+        and had not yet been delivered.
 
@@
 
-    \item 4 mails present for unknown reasons.
-
 \end{itemize}
+
+The XXX connections and mails handled by Postfix during this period are
+handled correctly by the parser.
 
@@
         logged if the mail is accepted.\footnote{Tested with Postfix 2.2.10
-        and 2.3.11; this may possibly have changed in Postfix 2.4.}  It
-        should be reasonably simple to write a policy server which causes
-        Postfix to log a warning containing the HELO hostname when the DATA
-        command is accepted.
+        and 2.3.11; this may possibly have changed in Postfix 2.4.}  It is
+        quite easy to rectify this:
+
+        \begin{enumerate}
+
+            \item Create \texttt{/etc/postfix/log\_helo.pcre}
+                containing:\newline
+                \hspace*{2em}\texttt{/./~~~~WARN~Logging~HELO}
+
+            \item Modify \texttt{smtpd\_sender\_restrictions} in
+                \texttt{/etc/postfix/main.cf} to contain\newline
+                \hspace*{2em}\texttt{check\_client\_access~/etc/postfix/log\_helo.pcre}
+
+        \end{enumerate}
+
+        The purpose of logging the HELO in
+        \texttt{smtpd\_sender\_restrictions} is to only log once per mail.
 

### 2007-07-22 doc/logparser.tex
Converted line endings.
Documented the last aborted delivery attempts.

@@
 
+\subsubsection{Yet more aborted delivery attempts}
+
+\label{yet-more-aborted-delivery-attempts}
+
+The aborted delivery attempts described in
+section~\ref{aborted-delivery-attempts} occur frequently, but the aborted
+delivery attempts described in this section only occur four times in the 93
+log files used for testing.  The symptoms are the same as in
+section~\ref{aborted-delivery-attempts}, except that there \textit{is\/} a
+\texttt{postfix/cleanup} log line; there does not appear to be anything in
+the log file to explain why there are no further log messages.  The only
+way to detect these mails is to periodically scan all mails in the state
+tables, deleting any mails displaying the following characteristics:
+
+\begin{itemize}
+
+    \item The timestamp of the last log line for the mail must be 12 hours
+        or more earlier than the last log line parsed.
+
+    \item There must be exactly two \texttt{postfix/smtpd} and one
+        \texttt{postfix/cleanup} log entries for the mail, with no
+        additional log entries.
+
+\end{itemize}
+
+12 hours is a somewhat arbitrary time period, but it is far longer than
+Postfix would delay delivery of a mail in the queue.  The state tables are
+scanned for mails matching the characteristics above each time the end of a
+log file is reached, and matching mails are deleted.
+
 \section{Parsing coverage}

### 2007-08-13 doc/logparser.tex
First round of corrections after proof reading; there are more to come.

@@
     it is possible to achieve a high degree of accuracy in understanding
-    the logs and reconstructing the actions taken by Postfix to generate
-    the logs.  This paper describes the process required, documenting the
-    parsing algorithm and rules, explaining the difficulties encountered,
-    with reference to an implementation which stores data gathered from the
-    logs in an SQL database.  The gathered data can then be used to
-    optimise current anti-spam measures, provide a baseline to test new
-    anti-spam measures against, or to produce statistics showing how
-    effective those measures are.
+    the logs, and thus reconstructing the actions taken by Postfix to
+    generate the logs.  This paper describes the creation of a parser,
+    documenting the parsing algorithm and rules, explaining the
+    difficulties encountered and the solutions developed, with reference to
+    an implementation which stores data gleaned from the logs in an SQL
+    database.  The gathered data can then be used to optimise current
+    anti-spam measures, provide a baseline to test new anti-spam measures
+    against, or to produce statistics showing how effective those measures
+    are.
 
@@
 has gotten in the last week, and possibly even generated some graphs to
-clearly illustrate the point to management or complaining
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing lines independently.
-One of the barriers to this type of processing is the unstructured nature
-of Postfix logs, where each logging line was added on an ad hoc basis as a
-requirement was discovered or new functionality was added.  Further
-complication arises because the list of rejection messages is not fixed:
-new messages can be added by the administrator with custom checks; every
-Real-time Black List returns a different explanatory message; policy
-servers may have different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers.  This paper documents the difficult process of parsing Postfix
-logs, presenting a program which parses logs and places the resulting data
-into a database.  The gathered data can then be used to optimise current
-anti-spam measures, provide a baseline to test new anti-spam measures
-against, or to produce statistics showing how effective those measures are.
-There are numerous other uses for such data: improving server performance
-by identifying troublesome destinations and reconfiguring appropriately or
-dedicating transports to those destinations; identifying regular high
-volume uses (e.g.\ customer newsletters) and restricting those uses to
-off-peak times; detecting virus outbreaks which propagate via email; as a
-base for billing customers on a shared server.
+clearly illustrate the point to management or users.\footnote{This was the
+author's first real foray into processing Postfix logs.}  Very few will
+have performed in-depth parsing and analysis of their logs, where the
+parsing must correlate the log lines per-connection or per-queueid rather
+than processing lines independently.  One of the barriers to this type of
+processing is the unstructured nature of Postfix logs, where each log line
+was added on an ad hoc basis as a requirement was discovered or new
+functionality was added.  Further complication arises because the set of
+rejection messages is not fixed: new messages can be added by the
+administrator with custom checks; every Real-time Black List returns a
+different explanatory message; policy servers may log different messages
+depending on the characteristics of the connection; there are many ways in
+which the log lines may differ between servers.  This paper documents the
+difficult process of parsing Postfix logs, presenting a program which
+parses logs and places the resulting data into a database.  The gathered
+data can then be used to optimise current anti-spam measures, provide a
+baseline to test new anti-spam measures against, or to produce statistics
+showing how effective those measures are.  There are numerous other uses
+for such data: improving server performance by identifying troublesome
+destinations and reconfiguring appropriately or dedicating transports to
+those destinations; identifying regular high volume uses (e.g.\ customer
+newsletters) and restricting those uses to off-peak times; detecting virus
+outbreaks which propagate via email; as a base for billing customers on a
+shared server.  Preserving the raw data enables users to develop a
+multitude of uses far beyond those conceived of by the author.
 
@@
 Section~\ref{rules} discusses the parsing rules in detail, explaining their
-structure, giving an example rule and sample data it matches successfully
+structure, using an example rule and sample data it matches successfully
 against, followed by a discussion of rule efficiency concerns.
@@
 and/or the algorithm, then suggests some ways of dealing with the
-limitations, with the goal of improving parsing and reproducing the journey
-a mail takes through the internals of Postfix.
+limitations, with the goal of improving parsing and reproduction of the
+journey a mail takes through Postfix.
 
@@
 
-Section~\ref{bibliography} is the bibliography, containing the resources
-used in developing the algorithm, writing the program and preparing this
-paper.  Also listed are some additional resources expected to be helpful in
-understanding SMTP, Postfix, anti-spam techniques, or the paper.
+The bibliography contains references to the resources used in developing
+the algorithm, writing the program and preparing this paper.  Also listed
+are some additional resources expected to be helpful in understanding SMTP,
+Postfix, anti-spam techniques, or the paper.
 
 Section~\ref{graphs} contains graphs illustrating the topics discussed in
-section~\ref{rule efficiency}.
+section~\ref{rule efficiency}, rule efficiency.
 
@@
 and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are worth using in the fight against spam.  The
+evaluated to see if they are beneficial in the fight against spam.  The
 program parses Postfix logs and populates a database with the data gleaned
@@
 \begin{verbatim}
-SELECT name, description, restriction_name, rule_order
+SELECT name, description, restriction_name, hits_total
     FROM rules
     WHERE postfix_action = 'REJECTED'
-    ORDER BY rule_order DESC;
+    ORDER BY hits_total DESC;
 \end{verbatim}
@@
 \begin{verbatim}
-SELECT name, description, restriction_name, rule_order,
-        (rule_order * 100.0 /
-            (SELECT SUM(rule_order)
+SELECT name, description, restriction_name, hits_total,
+        (hits_total * 100.0 /
+            (SELECT SUM(hits_total)
                 FROM rules
@@
     WHERE postfix_action = 'REJECTED'
-    ORDER BY rule_order DESC;
+    ORDER BY hits_total DESC;
 \end{verbatim}
@@
 \begin{verbatim}
-SELECT name, description, restriction_name, rule_order,
-        (rule_order * 100.0 /
-            (SELECT SUM(rule_order)
+SELECT name, description, restriction_name, hits,
+        (hits * 100.0 /
+            (SELECT SUM(hits)
                 FROM rules
@@
     WHERE postfix_action = 'REJECTED'
-        AND rule_order < 100
-    ORDER BY rule_order ASC;
+        AND hits < 100
+    ORDER BY hits ASC;
 \end{verbatim}
@@
 quite similar to the API provided by a library: it insulates both user and
-provider of the API from changes in the implementation of the other.  The
-algorithm implemented by the program can be improved; support can be added
+provider of the API from changes in the implementation of the API\@.  The
+algorithm implemented by the parser can be improved; support can be added
 for earlier or later releases of Postfix; bugs can be fixed or limitations
-removed from the program.  Statistics and/or graphs can be generated from
+removed from the parser.  Statistics and/or graphs can be generated from
 the database; new restrictions can be tested and the results inspected;
@@
     \item Libraries exist for the majority of programming languages
-        allowing access to databases, whereas if this parser was written
-        as a library a new interface layer would need to be written for
+        allowing access to databases, whereas if this parser was written as
+        a shared library a new interface layer would need to be written for
         every programming language using the API\@.
@@
         the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
+        databases provide a program, of varying complexity and
         sophistication, which can be used for ad hoc queries with minimal
@@
 
-            \item Other tables can be added to the database.
+            \item Other tables can be added to the database, e.g.\ to cache
+                historical data.
 
@@
 
+XXX SHOULD THIS SECTION BE MOVED ELSEWHERE\@?
+
 The database schema can be conceptually divided in two: the rules which are
@@
 Rules have the fields required to parse the log lines, extract data to be
-saved, and cause the correct action to be executed; they also have several
-fields which aid the user in understanding what each rule parses.  The
-rules are described in detail in section~\ref{rules} and the fields in
-particular in section~\ref{rule attributes}, and that discussion won't be
-repeated here.  The data saved from parsing the logs is divided into two
-tables: connections and results.
+saved, and the action to be executed; they also have several fields which
+aid the user in understanding what each rule parses.  The rules are
+described in detail in section~\ref{rules} and the fields in particular in
+section~\ref{rule attributes}, and that discussion won't be repeated here.
+The data saved from parsing the logs is divided into two tables described
+below: connections and results.
 
@@
         or virus senders think it's a good idea.  By default Postfix only
-        logs this when it rejects an SMTP command, but it is trivial to log
-        this for every connection:
-
-        \texttt{smtpd\_helo\_restrictions = \newline
-        \hspace*{2em} check\_helo\_access regexp:/etc/postfix/log\_helo.regex}
-
-        Where \texttt{/etc/postfix/log\_helo.regex} contains:
-
-        \texttt{/./ WARN Logging HELO for diagnostic purposes}
-
-    \item [queueid] The queueid of the mail, or \textit{NOQUEUE\/} if a
-        mail wasn't accepted.
+        logs the HELO when it rejects an SMTP command, but it is quite easy
+        to rectify this:
+
+\label{logging helo}
+
+        \begin{enumerate}
+
+            \item Create \texttt{/etc/postfix/log\_helo.pcre}
+                containing:\newline
+                \hspace*{2em}\texttt{/./~~~~WARN~Logging~HELO}
+
+            \item Modify \texttt{smtpd\_helo\_restrictions} in
+                \texttt{/etc/postfix/main.cf} to contain\newline
+                \hspace*{2em}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
+
+        \end{enumerate}
+
+        The purpose of logging the HELO in
+        \texttt{smtpd\_helo\_restrictions} is to log only once per mail; it
+        also seems like the natural place.
+
+    \item [queueid] The queueid of the mail if the conection represents an
+        accepted mail, or \textit{NOQUEUE\/} otherwise.
 
@@
 Every log line corresponding to Postfix performing an action other than
-INFO will have an entry in the results table, e.g.\ rejecting an SMTP
-command, delivering a mail, or bouncing a mail.  Any log line where the
-Postfix action is INFO is not interesting in and of itself, but provides
-additional information which will be saved with other results.
+INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
+SMTP command, delivering a mail, or bouncing a mail.  Any log line where
+the Postfix action is INFO is not interesting in and of itself, but
+provides additional information which will be saved with other results;
+IGNORED log lines don't even provide useful information.
 
@@
     \item [connection\_id] A reference to the entry in the connections
-        table this results entry belongs to.
+        table this results table entry belongs to.
 
@@
 
-    \item [warning] Postfix can log a warning instead of rejecting an SMTP
-        command --- a facility that is quite useful for testing new
-        restrictions.  This field will be 1 if the log line parsed was a
-        warning rather than a real rejection, or 0 for a real rejection or
-        any other result.
-
-    \item [smtp\_code] The SMTP code extracted from the log line.  In
+    \item [warning] Postfix can be configured to log a warning instead of
+        enforcing a restriction that would reject an SMTP command --- a
+        facility that is quite useful for testing new restrictions.  This
+        field will be 1 if the log line parsed was a warning rather than a
+        real rejection, or 0 for a real rejection or any other result.
+
+    \item [smtp\_code] The SMTP code associated with the log line.  In
         general an SMTP code is only present for a rejection or final
         delivery; results missing an SMTP code will duplicate the SMTP code
-        of other results.  Some final delivery log lines don't contain an
-        SMTP code: in those cases the code is faked based on the success or
-        failure represented by the log line.
+        of other results in the connection.  Some final delivery log lines
+        don't contain an SMTP code: in those cases the code is faked based
+        on the success or failure represented by the log line.
 
@@
 The Simple Mail Transfer Protocol, originally defined in RFC
-821~\cite{RFC821} and later in RFC 2821~\cite{RFC2821}, is used for
+821~\cite{RFC821} and updated in RFC 2821~\cite{RFC2821}, is used for
 transferring mail between Mail Transport Agents (MTAs).  It is a simple,
@@
 Postfix is a highly configurable, high performance, secure and scalable
-Mail Transport Agent.  It features extensive optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained shortly.} as the ultimate
-in flexibility.  Administrators can also supply their own rejection
-messages to make it clear to senders why exactly their mail was rejected.
-Unfortunately this flexibility has a cost: complexity in the logs
-generated.  While it is easy to use \texttt{grep(1)} to determine the fate
-of an individual email, following the journey an email takes through
-Postfix can be quite difficult.  The logs tend to follow a 90\%-10\%
-pattern: 90\% of the time the journey is simple, but the other 10\% of the
-time requires 90\% of the code.\footnote{These numbers don't have a solid
-scientific basis, they're based on gut feeling from writing and debugging
-the software.}
+Mail Transport Agent.  It features extensive and extensible optional
+anti-spam restrictions, allowing an administrator to deploy those
+restrictions which they judge suitable for their site's needs, rather than
+a fixed set chosen by Postfix's author.  These restrictions can be
+selectively applied, combined and bypassed on a per-client, per-recipient
+or per-sender basis, allowing varying levels of severity and/or
+permissiveness.  Postfix leverages simple lookup tables to support
+arbitrarily complicated user-defined sequences of restrictions and
+exceptions, with policy servers\footnote{Policy servers will be explained
+shortly.} as the ultimate in flexibility.  Administrators can also supply
+their own rejection messages to make it clear to senders why exactly their
+mail was rejected.  Unfortunately this flexibility has a cost: complexity
+in the logs generated.  While it is easy to use \texttt{grep(1)} to
+determine the fate of an individual email, following the journey an email
+takes through Postfix can be quite difficult.  The logs tend to follow a
+90\%-10\% pattern: 90\% of the time the journey is simple, but the other
+10\% of the time requires 90\% of the code.\footnote{These numbers don't
+have a solid scientific basis, they're based on gut feeling from writing
+and debugging the software.}
 
@@
 matching line.  Other restrictions determine their result in different
-ways, e.g.\ \texttt{reject\_rbl\_client some.rbl.com} checks whether the IP
-address of the client is present in the listed RBL\@.
+ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks whether the
+IP address of the client is present in the listed RBL\@.
 
@@
 restrictions to be chosen based on the client IP address, HELO hostname,
-sending address, recipient address, etc.  The administrator can define new
+sender address, recipient address, etc.  The administrator can define new
 restrictions as a list of existing restrictions, allowing arbitrarily long
@@
 
-    \item Greylisting~\cite{greylisting}, is a technique that temporarily
+    \item Greylisting~\cite{greylisting} is a technique that temporarily
         rejects mail when the triple of (sender, recipient, remote IP
@@
 
+XXX REWRITE THIS
+
 Before getting into detail about the parser a brief overview is in order.
@@
 coherent whole, saving it in a useful and consistent form, and performing
-housekeeping duties.  Readers familiar with context-free grammars may find
-section~\ref{comparison against context-free grammars} helpful.
+housekeeping duties.
 
@@
 Utilizations\/}~\cite{log-mail-analyser}.  The aim of Log Mail Analyzer is
-quite different from his parser: it attempts to present correlated log data
-in a form suitable for a systems administrator to search using grep(1),
-producing a CSV file\footnote{Comma-Separated Value, a very simple format
-where a record is stored in a single line, with fields separated by a comma
-or other punctuation symbol.  Problems with CSV files include the
+quite different from this parser: it attempts to present correlated log
+data in a form suitable for a systems administrator to search using
+grep(1), producing a CSV file\footnote{Comma-Separated Value, a very simple
+format where a record is stored in a single line, with fields separated by
+a comma or other punctuation symbol.  Problems with CSV files include the
 difficulty in escaping separators, providing multiple values for a field
-(e.g.\ multiple recipients), adding new fields, and lack of a bundled schema
-describing the fields.} which can subsequently be imported into a MySQL or
-Berkeley DB database.  The data stored is limited to time and date,
-hostname and IP address of client, sender and recipient addresses, and the
-SMTP code.  The decision to support both CSV and Berkeley DB appears to
-have been a limiting factor: the values which can be stored are limited,
-it's difficult to link different records (foreign keys in SQL enable this
-easily), and ad-hoc queries require extensive code to be written.  Berkeley
-DB is not an SQL database, it only supports (key, value) pairs; in the main
-table the value used is a CSV line, and the key is an integer referred to
-by other tables.  The secondary by-sender, by-recipient, by-date and by-IP
-tables use the sender/recipient/date/IP as the key, and a CSV list of
-integers referring to the main table is used as the value.  In effect this
-re-implements SQL foreign keys, but without the full functionality offered
-by even the most basic of SQL databases.  It also requires custom code to
-search on some combination of the above.
+(e.g.\ multiple recipients), adding new fields, and lack of a bundled
+schema describing the fields.} which can subsequently be imported into a
+MySQL or Berkeley DB database.  The data stored is limited to time and
+date, hostname and IP address of client, sender and recipient addresses,
+and the SMTP code.  The decision to support both CSV and Berkeley DB
+appears to have been a limiting factor: the values which can be stored are
+limited, it's difficult to link different records (foreign keys in SQL
+enable this easily), and ad-hoc queries require extensive code to be
+written.  Berkeley DB is not an SQL database, it only supports (key, value)
+pairs; in the main table the value used is a CSV line, and the key is an
+integer referred to by other tables.  The secondary by-sender,
+by-recipient, by-date and by-IP tables use the sender/recipient/date/IP as
+the key, and a CSV list of integers referring to the main table is used as
+the value.  In effect this re-implements SQL foreign keys, but without the
+full functionality offered by even the most basic of SQL databases.  It
+also requires custom code to search on some combination of the above.
 
@@
 program designed to be extensible and useful in a production environment.
+
+XXX EXPLAIN WHY ASO IS MORE COMPREHENSIVE, WHAT IT DOES BETTER AND WHY\@.
+MAKE THIS A COMPARISON INSTEAD OF A CRITIQUE\@.
 
@@
         log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  The description field is generally used to
-        describe the action causing the log lines this rule matches.
+        line to be logged).  This field describes the action causing the
+        log lines this rule matches.
 
@@
 
-    \item [program] The program (postfix/smtp, postfix/smtpd, postfix/qmgr,
-        etc.) whose log lines the rule applies to.  This avoids needlessly
-        trying rules which won't match the line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any lines which aren't parsed by program specific rules.
-
-    \item [regex] The regular expression\footnote{See
-        http://en.wikipedia.org/wiki/Regular\_expression for more
-        information about regular expressions in general.} to match the log
-        line against.  The regex will first have several keywords expanded:
-        this simplifies reading and writing rules, avoids needless
-        repetition of complex regex components, allows the components
-        to be corrected and/or improved in one location, and makes each
-        regex largely self-documenting.
+    \item [program] The program (\texttt{postfix/smtpd},
+        \texttt{postfix/qmgr}, etc.) whose log lines the rule applies to.
+        This avoids needlessly trying rules which won't match the line, or
+        worse, might match unintentionally.  Rules whose program is
+        \texttt{*} will be tried against any lines which aren't parsed by
+        program specific rules.
+
+    \item [regex] The regular expression\cite{wikipedia-regex, perlre} to
+        match the log line against.  The regex will first have several
+        keywords expanded: this simplifies reading and writing rules;
+        avoids needless repetition of complex regex components; allows the
+        components to be corrected and/or improved in one location; and
+        makes each regex largely self-documenting.
 
@@
         \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline server\_hostname and helo.\newline} \texttt{result\_cols:
-        sender, recipient, smtp\_code, data \newline and child.}
+        \newline \hspace*{2em} server\_hostname and helo.\newline}
+        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
+        \newline \hspace*{2em} pid\_regex, data and child.}
 
@@
         At the start of each run the program sorts the rules in descending
-        hits, and at the end of the run updates every rule's hits.
+        order of hits, and at the end of the run updates every rule's hits.
         Assuming that the distribution of log lines is reasonably
@@
         will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules.
+        more specific rules to take precedence over more general rules
+        (described in section~\ref{overlapping rules}).
 
@@
 \subsection{Overlapping rules}
+
+\label{overlapping rules}
 
@@
 specific rule to match some lines and a more general rule to match the
-remainder, e.g.\ separating smtp delivery to a filtering proxy from smtp
+majority, e.g.\ separating smtp delivery to a filtering proxy from smtp
 delivery to the rest of the world.  The algorithm provides a facility for
-handling overlapping rules: the priority field in each rule (defaults to
+ordering overlapping rules: the priority field in each rule (defaults to
 zero).  Rules are tried in order of priority, highest first, overriding the
@@
         cannot prove the absence of overlapping rules; it can detect
-        overlapping rules --- but only if there are log lines which match
-        more than one rule.
+        overlapping rules --- but only if there are log lines in the input
+        files which match more than one rule.
 
@@
 NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 
-  550 <foo@bar.baz>: Sender address rejected: Domain not found;
-  from=<foo@bar.baz> to=<info@example.com> proto=SMTP
-  helo=<smtp.bar.baz>
+  550 <foo@example.com>: Sender address rejected:
+  Domain not found; from=<foo@example.com>
+  to=<info@example.net> proto=SMTP
+  helo=<smtp.example.com>
 \end{verbatim}
@@
 
+\vspace{1em}
+
 The various fields are used as follows;
@@
         program which logged the line the regex will be matched against the
-        line; if the match is successful the action will be called, if not
-        the next rule will be tried.
+        line; if the match is successful the action will be executed, if not
+        the next rule will be tried.  If the program-specific rules don't
+        match the log line, the generic rules will be tried in the same
+        way.
 
@@
     \item [queueid:] is used the extract the queueid matched by the regex,
-        so that the mail can be found by queueid and actions performed on
-        it.
+        so that the correct mail can be found by queueid and actions
+        performed on it.
 
@@
 Additional fields are captured by \_\_RESTRICTION\_START\_\_, hence the
-fields in result\_cols and connection\_cols start at 5.
+fields in result\_cols and connection\_cols start at 5 in the example.
 
@@
 Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 50 MB log files containing 300,000 log lines (for a small mail
+deal with 75 MB log files containing 300,000 log lines (for a small mail
 server --- large scale mail servers would have much larger log files on a
@@
 
-Rule ordering was briefly discussed in section~\ref{rule attributes} and it
-will be covered in greater detail in this section.  At the time of writing
-there are 131 different rules, the majority of which match several orders
-of magnitude less frequently than the top 10\%, as shown in graph~\ref{rule
-hits}.  Assuming that the distribution of log lines is reasonably steady
-over time, program efficiency should benefit from trying more frequently
-matching rules before those which match less frequently.  To test this
-hypothesis three full test runs were performed with different rule
-orderings:
+Rule ordering was mentioned in section~\ref{rule attributes} and will be
+covered in greater detail in this section.  At the time of writing there
+are 139 different rules, with the top 10\% matching the vast majority of
+the log lines, and the remaining log lines split across the other 90\% of
+the rules (as shown in graph~\ref{rule hits}).  Assuming that the
+distribution of log lines is reasonably steady over time, program
+efficiency should benefit from trying more frequently matching rules before
+those which match less frequently.  To test this hypothesis three full test
+runs were performed with different rule orderings:
 
@@
 
+Perl compiles the original regex into an internal representation,
+optimising the regex to improve the speed of matching, but this compilation
+and optimisation takes CPU time.  Perl automatically caches static regexs,
+but dynamic regexs need to be explicitly compiled and cached.
 Graph~\ref{normal regex vs discard regex} shows execution times with and
@@
 percentage execution time increase when not caching each regex is between
-350--570\% (mean 465.58\%, standard deviation 114.70\%).  Perl compiles the
-original regex into an internal representation, optimising the regex to
-improve the speed of matching, but this compilation and optimisation takes
-CPU time.  Perl automatically caches static regexs, but dynamic regexs need
-to be explicitly compiled and cached.
+350--570\% (mean 465.58\%, standard deviation 114.70\%).
 
@@
 
-The most difficult complication is that locally addressed mails are not
-always delivered directly to a mailbox: sometimes they are accepted for a
-local address but need to be delivered to one or more remote addresses due
-to aliases.  When this occurs a child mail will be injected into the
-postfix queue, but without the explicit logging smtpd or sendmail injected
-mails have.  Thus the source is not immediately discernible from the log
-line in which the mail first appears; from a strictly chronological reading
-of the logs it \textit{usually\/} appears as if the child mail has appeared
-from thin air.  Subsequently the parent mail will log the creation of the
-child mail:
+The most difficult complication initially encountered is that locally
+addressed mails are not always delivered directly to a mailbox: sometimes
+they are accepted for a local address but need to be delivered to one or
+more remote addresses due to aliases.  When this occurs a child mail will
+be injected into the postfix queue, but without the explicit logging smtpd
+or sendmail injected mails have.  Thus the source is not immediately
+discernible from the log line in which the mail first appears; from a
+strictly chronological reading of the logs it \textit{usually\/} appears as
+if the child mail has appeared from thin air.  Subsequently the parent mail
+will log the creation of the child mail:
 
@@
 from qmgr for a previously unseen mail; it must flag the mail as coming
-from an unknown origin and subsequently clear the flag if and when the
+from an unknown origin, and subsequently clear the flag if and when the
 origin of the mail becomes clear.  Obviously the parser could omit checking
@@
 coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as database
-ready and subsequently entered by the parent mail once it has been
+their final log line is parsed; instead they must be marked as ready for
+entry and subsequently entered by the parent mail once it has been
 identified.  Quite apart from identifying mail injected in an unknown
@@
 such because they lack some data present in the parent mail; they must copy
-that data from their parent before being entered in the database.
-The effect of this complication upon the algorithm is discussed in
+that data from their parent before being entered in the database.  The
+effect of this complication upon the algorithm is discussed in
 section~\ref{tracking re-injected mail}.
@@
 of the rule which matched the log line, executing the code in the function
-named by the action.  From a high level viewpoint this design shares
-significant similarity with how an event dispatch loop works in a GUI\@.
+named by the action.
 
@@
 
-Internally generated mails lack any explicit origin and can only be
-detected using heuristics (see
+Internally generated mails lack any explicit origin (in Postfix 2.2.x) and
+can only be detected using heuristics (see
 section~\ref{identifying-bounce-notifications} for details).  Bounce
@@
         child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from memory (flowchart:~15).  For
-        re-injected mails see section~\ref{tracking re-injected mail}.
+        database (flowchart:~14), then deleted from memory (flowchart:~15).
+        Re-injected mails are described in section~\ref{tracking
+        re-injected mail}.
 
@@
 
-It's probably apparent by now that tracking re-injected mails is the single
-most complex part of the algorithm.  Section~\ref{complications} has most
-of the previous discussion, should you wish to review it.
-
 The crux of the problem is that re-injected mails appear in the logs
@@
         (action: mail\_picked\_for\_delivery; flowchart:~3), in which case
-        a new data structure will be created.  The mail will have been
-        flagged as having unknown origins; this flag should be subsequently
-        cleared once the origin has been established.  The data is then
-        saved.  This may also be an indicator that the mail is a bounce
-        notification, see section~\ref{identifying-bounce-notifications}
-        for details.
+        a new data structure will be created.  The mail will be flagged as
+        having unknown origins; this flag should be subsequently cleared
+        once the origin has been established.  This may also be an
+        indicator that the mail is a bounce notification, see
+        section~\ref{identifying-bounce-notifications} for details.
 
@@
         delivery is performed by the SMTP client; local delivery means
-        local to the server, i.e.\ an address the server accepts mail for.}
-        In this case the mail may already have been created (action:
-        qmgr\_chooses\_mail; flowchart:~3; described above) and the unknown
-        origin flag will be cleared; if not a new data structure will be
-        created.  In both cases the new mail is marked as a child of the
-        parent.  The log line in question is:
+        local to the server, i.e.\ an address the server is final
+        destination for.} In this case the mail may already have been
+        created (action: qmgr\_chooses\_mail; flowchart:~3; described
+        above) and the unknown origin flag will be cleared; if not a new
+        data structure will be created.  In both cases the new mail is
+        marked as a child of the parent.  The log line in question is:
 
@@
         child is deleted from the state tables the parent will also be
-        finished with (flowchart:~27).
+        finished with (flowchart:~27), and deleted from the state tables.
 
@@
 \end{description}
+
+The actions:
 
@@
 
-    \item [SAVE\_BY\_QUEUEID] Save the data extracted by the regex to the
-        mail identified by the queueid from the log line.
+    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
+        the log line, and save the data extracted by the regex to it.
 
@@
         parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either with this
-        parent or with another parent, and issues appropriate warnings in
-        either case.
+        checks if the child has already been tracked, either by this parent
+        or by another parent, and issues appropriate warnings in either
+        case.
 
     \item [REJECTION] Deal with postfix rejecting an SMTP command from the
-        remote client: log the rejection with the last accepted mail or the
-        connection, as appropriate.
+        remote client: log the rejection with a mail if there is a queueid
+        in the log line, or with the connection if not.
 
@@
         picking a mail from the queue to deliver. This action is used for
-        both qmgr and cleanup due to out of order log lines; see
-        section~\ref{discarding cleanup lines} for details.
+        both qmgr and cleanup as it needs to deal with out of order log
+        lines; see section~\ref{discarding cleanup lines} for details.
 
     \item [PICKUP] Pickup is the service which deals with mail submitted
-        locally via /usr/sbin/sendmail. This action creates a new state
-        table entry and saves data to it, unless out of order logging has
-        caused the cleanup line to be logged first, as discussed in
-        section~\ref{pickup logging after cleanup}.
+        locally via /usr/sbin/sendmail.
+
+        Out of order log entries may have caused the state table entry to
+        already exist (see section~\ref{pickup logging after cleanup});
+        otherwise it is created.  The data from the log line is then saved
+        to the state table entry.
 
@@
         must be cloned; if the original data structure was used the second
-        and subsequent mails would corrupt the data structure.
-
-    \item [MAIL\_TOO\_LARGE] When the client tries to send a larger message
+        and subsequent mails would overwrite one another's data.
+
+    \item [MAIL\_TOO\_LARGE] When the client tries to send a message larger
         than the local server accepts the mail will be discarded.  See
@@
         which case there's a data structure to dispose of, or it may not in
-        which case there's none. See
+        which case there is not.  See
         section~\ref{timeouts-during-data-phase} for the gory details.
@@
         it kills all smtpds,\footnote{Possibly other programs are killed
-        also, but the parser is only affected by and interested in smtpd.}
-        requiring any active connections to be committed and deleted from
-        the state tables.
+        also, but the parser is only affected by and interested in smtpds
+        exiting.} requiring any active connections to be committed and
+        deleted from the state tables.
 
@@
 
-    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) log the
+    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
         creation of bounce messages.  This action creates a new mail if
         necessary, marking it as a bounce notification and removing the
-        unknown origin flag if it exists.
+        unknown origin flag if it already exists.
 
@@
 
-    \item The sender address will be $<>$.
-
-    \item Neither smtpd nor pickup will have logged any messages associated
-        with the mail, indicating it was generated internally by Postfix,
-        not accepted via SMTP or submitted locally by sendmail.
+    \item The sender address is $<>$.
+
+    \item Neither smtpd nor pickup have logged any messages associated with
+        the mail, indicating it was generated internally by Postfix, not
+        accepted via SMTP or submitted locally by sendmail.
 
@@
         \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.cs.tcd.ie}
+        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
 
@@
 
-Postfix 2.3 (and hopefully subsequent version) log the creation of a bounce
-message.
+Postfix 2.3 (and hopefully subsequent versions) log the creation of a
+bounce message.
 
@@
 behaviour occurs.  In this case a list of all mails accepted during a
-connection are saved in the connection state, and the accepted mails
+connection are saved in the connection state, and the accepted mails are
 examined when the disconnection line is parsed and the disconnection action
@@
 line is discarded the queueid will be removed from the cache of timeout
-queueids when the next pickup line is parsed.
+queueids when the next pickup line containing that queueid is parsed.
 
@@
 after cleanup}.  In the 93 log files used for testing this occurs only five
-times in 60721709 log lines, but for completeness of the algorithm it
+times in 607,21,709 log lines, but for completeness of the algorithm it
 should be dealt with.  The five occurrences in the test log files have the
-same characteristics: the postfix/local log line showing delivery to a
-local mailbox occurs after the postfix/qmgr log line showing removal of the
-mail from the queue because delivery is completed.  This causes problems:
-the mail is not complete, so entry into the database fails; a new mail is
-created when the postfix/local line is parsed and remains in the state
-tables; four warnings are issued per pair of out of order log lines.
+same characteristics: the \texttt{postfix/local} log line showing delivery
+to a local mailbox occurs after the \texttt{postfix/qmgr} log line showing
+removal of the mail from the queue because delivery is completed.  This
+causes problems: the mail is not complete, so entry into the database
+fails; a new mail is created when the \texttt{postfix/local} line is parsed
+and remains in the state tables; four warnings are issued per pair of out
+of order log lines.
 
@@
 logged messages for each mail, comparing the list against a table of
-known-good combinations of programs.  If combination is found in the valid
-list the mail can be entered in the database; if the combination is not
-found entry must be postponed and the mail flagged.  The SAVE\_BY\_QUEUEID
-action checks for the flag and retries entry; if the additional log lines
-have caused the mail to reach a valid combination entry will succeed,
-otherwise it must be postponed once more.
+known-good combinations of programs.  If the mail's combination is found in
+the valid list the mail can be entered in the database; if the combination
+is not found entry must be postponed and the mail flagged.  The
+SAVE\_BY\_QUEUEID action checks for the flag and retries entry if it's
+found; if the additional log lines have caused the mail to reach a valid
+combination entry will succeed, otherwise it must be postponed once more.
 
 The list of valid combinations is given below.  Every mail will
-additionally have log entries from postfix/cleanup and postfix/qmgr; any
-mail may also have log entries from postfix/bounce, postfix/postsuper, or
-both.
+additionally have log entries from \texttt{postfix/cleanup} and
+\texttt{postfix/qmgr}; any mail may also have log entries from
+\texttt{postfix/bounce}, \texttt{postfix/postsuper}, or both.
 
@@
 12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.  The state tables are
-scanned for mails matching the characteristics above each time the end of a
-log file is reached, and matching mails are deleted.
-
-\section{Parsing coverage}
+Postfix would delay delivery of a mail in the queue.\footnote{This may be a
+problem if Postfix is not running for an extended period of time.}  The
+state tables are scanned for mails matching the characteristics above each
+time the end of a log file is reached, and matching mails are deleted.
+
+\section{Coverage}
 
@@
 
-    \item Try each rule, ordered by priority and rule order (see
-        sections~\ref{rule attributes} and~\ref{rule ordering for
-        efficiency}), until a matching rule is found.
-
-    \item Execute the appropriate action.
+    \item Try each rule until a matching rule is found.
+
+    \item Execute the action specified by the rule.
 
@@
 
-    \item Each program of interest must have at least one rule or it's log
+    \item Each program of interest must have at least one rule or its log
         lines will be silently skipped; in the extreme case of zero rules
@@
         file, and there are some Postfix programs which don't produce any
-        log lines of interest (currently \textit{postfix/anvil\/} and
-        \textit{postfix/scache\/}).
+        log lines of interest.
 
@@
 
-Full coverage of log lines is both easy to achieve and hard to maintain.
-It is easy to achieve full coverage for a limited set of log files (at the
-time of writing the parser has 131 rules, fully parsing 93 contiguous log
-files from Postfix 2.2 and 2.3), and new rules are easy to add.
-Maintaining full coverage is hard because other servers have different
-restrictions with custom messages, RBLs change their messages over time,
-major releases of Postfix change warning messages (usually adding more
-information), etc.\ so over time the log lines drift and change.
-Graph~\ref{rule hits} shows the hits for each rule over all 93 log files;
-it's obvious that a small number of rules match the vast majority of the
-lines, and more than half the rules match fewer than 100 times.
+Full coverage of log lines is easy to achieve yet hard to maintain.  It is
+easy to achieve full coverage for a limited set of log files (at the time
+of writing the parser has 139 rules, fully parsing 93 contiguous log files
+from Postfix 2.2 and 2.3), and new rules are easy to add.  Maintaining full
+coverage is hard because other servers have different restrictions with
+custom messages, RBLs change their messages over time, major releases of
+Postfix change warning messages (usually adding more information), etc.\ so
+over time the log lines drift and change.  Graph~\ref{rule hits} shows the
+number of hits for each rule over all 93 log files; it's obvious that a
+small number of rules match the vast majority of the lines, and more than
+half the rules match fewer than 100 times.
 
@@
 queueids or pids in the first few hundred or thousand log lines because the
-earlier log lines are in the previous log file; loading the saved state
-from the previous log file will solve this problem.} or when there are
-problems tracking a child mail (see section~\ref{tracking re-injected
-mail}).  There should be few or no warnings, and the state table should
-only contain entries for mails which had yet to be delivered when the
-logfiles ended.
+earlier log lines for those connections or mails are in the previous log
+file; loading the saved state from the previous log file will solve this
+problem.} or when there are problems tracking a child mail (see
+section~\ref{tracking re-injected mail}).  There should be few or no
+warnings, and the state table should only contain entries for mails which
+had yet to be delivered when the logfiles ended, or started before the
+logfiles began.
 
@@
 
-The XXX connections and mails handled by Postfix during this period are
-handled correctly by the parser.
+The 13,850,793 connections and mails accepted, rejected or delivered by
+Postfix during this period are handled correctly by the parser.
 
@@
         logged if the mail is accepted.\footnote{Tested with Postfix 2.2.10
-        and 2.3.11; this may possibly have changed in Postfix 2.4.}  It is
-        quite easy to rectify this:
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline
-                \hspace*{2em}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_sender\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \hspace*{2em}\texttt{check\_client\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        The purpose of logging the HELO in
-        \texttt{smtpd\_sender\_restrictions} is to only log once per mail.
+        and 2.3.11; this may possibly have changed in Postfix 2.4.}
+        Rectifying this has already been described in section~\ref{logging
+        helo}.
 
@@
 
-    \item A progress bar would be useful when run interactively, as the
-        program takes roughly one minute per 10MB of
-        logs.\footnote{Approximately 80\% of the run time is consumed by
-        logging to the database.}  Obviously performance is entirely
-        dependant on the machine the program is running on.
-
-    \item Write the policy servers referred to in the limitations above.
+    \item Write the policy server referred to in the limitations above.
 
@@
 
-XXX THIS NEEDS TO BE EXPANDED
+XXX THIS NEEDS TO BE EXPANDED\@; EXPLAIN WHY THEY'RE NOT GOOD ENOUGH
+
+XXX ADD DATES CHECKED FOR ALL PARSERS
 
@@
 \bibliographystyle{logparser-bibliography-style}
+\bibliography{logparser-bibliography}
 \label{bibliography}
-\bibliography{logparser-bibliography}
 
@@
 
-\subsubsection{Percentage increase of shuffled and reversed over normal}
+\subsubsection{Percentage increase of shuffled + reversed over normal}
 \label{normal vs shuffled vs reversed ordering factor}
@@
 program in graph~\ref{log22 line and rule distribution by program} shows
-that \textit{postfix/qmgr\/} has the highest number of log lines, however
-it has a low number of rules.  This combination results in a low average
-number of rules to try for \textit{postfix/qmgr\/} log lines, reducing the
+that \texttt{postfix/qmgr} has the highest number of log lines, however it
+has a low number of rules.  This combination results in a low average
+number of rules to try for \texttt{postfix/qmgr} log lines, reducing the
 overall average number of rules to try per log line.  In turn this reduces

### 2007-08-15 doc/logparser.tex
Second pass through the corrections; all the small stuff is done. Promoted database schema subsection to a full section. Merged LMA review into prior art section, and reworked the combination somewhat.

@@
 \usepackage{url}
-\DeclareUrlCommand\urlwrap{}
-\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
+%\DeclareUrlCommand\urlwrap{}
+%\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
 % Include images
@@
 % Add the bibliography into the table of contents.
-\usepackage[section,numbib,nottoc]{tocbibind}
+\usepackage[section,numbib]{tocbibind}
 % When creating a PDF make the table of contents into links to the pages
@@
 \usepackage{hyperref}
-\hypersetup{%
+\hypersetup{
     pdftitle    = {Parsing Postfix log files},
@@
 
-Section~\ref{rules} discusses the parsing rules in detail, explaining their
-structure, using an example rule and sample data it matches successfully
-against, followed by a discussion of rule efficiency concerns.
+Section~\ref{database schema} partially describes the database schema used by
+the parser, explaining in detail the tables used for storing the resulting
+data.  The remainder of the schema is laid out in section~\ref{rule
+attributes}.
+
+Section~\ref{rules} discusses the parsing rules in detail, describing the
+table used to hold the rules and explaining their structure, using an
+example rule and sample data it matches successfully against, followed by a
+discussion of rule efficiency concerns.
 
@@
 algorithm.
+
+Section~\ref{parsing coverage} analyses the coverage the parser achieves
+over a set of 93 log files taken from a mail server handling mail for over
+1000 users.  Coverage is described both in terms of the fraction of lines
+parsed and the fraction of mails and connections successfully handled by
+the parsing algorithm.
 
@@
     \item SQL is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are already
-        lots of readily available resources from which to learn.  Although
-        every vendor implements a different dialect of SQL, the basics are
-        the same everywhere.
+        familiar with it; for those unfamiliar with it there are lots of
+        readily available resources from which to learn.  Although every
+        vendor implements a different dialect of SQL, the basics are the
+        same everywhere.
 
 \end{enumerate}
-
-\subsection{Database schema}
-\label{database schema}
-
-XXX SHOULD THIS SECTION BE MOVED ELSEWHERE\@?
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in section~\ref{rules} and the fields in particular in
-section~\ref{rule attributes}, and that discussion won't be repeated here.
-The data saved from parsing the logs is divided into two tables described
-below: connections and results.
-
-\subsubsection{Connections table}
-
-Every connection and/or mail will have a single entry in the connections
-table containing the following fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The IP address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \textit{unknown\/} if the IP address could not be resolved to a
-        hostname via DNS\@.
-
-    \item [client\_ip] The client IP address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \textit{unknown\/} if the IP address could not be resolved to a
-        hostname via DNS\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO when it rejects an SMTP command, but it is quite easy
-        to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline
-                \hspace*{2em}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_helo\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \hspace*{2em}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        The purpose of logging the HELO in
-        \texttt{smtpd\_helo\_restrictions} is to log only once per mail; it
-        also seems like the natural place.
-
-    \item [queueid] The queueid of the mail if the conection represents an
-        accepted mail, or \textit{NOQUEUE\/} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch.
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsubsection{Results table}
-
-Every log line corresponding to Postfix performing an action other than
-INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
-SMTP command, delivering a mail, or bouncing a mail.  Any log line where
-the Postfix action is INFO is not interesting in and of itself, but
-provides additional information which will be saved with other results;
-IGNORED log lines don't even provide useful information.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the entry in the connections
-        table this results table entry belongs to.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an SMTP command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection or any other result.
-
-    \item [smtp\_code] The SMTP code associated with the log line.  In
-        general an SMTP code is only present for a rejection or final
-        delivery; results missing an SMTP code will duplicate the SMTP code
-        of other results in the connection.  Some final delivery log lines
-        don't contain an SMTP code: in those cases the code is faked based
-        on the success or failure represented by the log line.
-
-    \item [sender] The sender's email address.  This can change during one
-        single connection, when the connection is reused to send multiple
-        mails.
-
-    \item [recipient] The recipient address.
-
-    \item [message\_id] The message-id of the accepted mail, or NULL if no
-        mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an RBL\@.
-
-    \item [timestamp] The time the line was logged, in seconds since the
-        epoch.
-
-\end{description}
 
@@
 
-XXX REWRITE THIS
-
 Before getting into detail about the parser a brief overview is in order.
 The parser is split into two parts: the parsing algorithm and the rules
-which are applied to the lines.  Rules can be thought of as the Lex part of
-a Lex and YACC style parser: rules identify the line and return some data,
-which the algorithm (the YACC part) receives and uses when performing the
-requested action.  Rules are solely concerned with identifying a line and
-extracting data from it, whereas the algorithm's task is to follow the
-journey each mail takes through Postfix, piecing the data together into a
-coherent whole, saving it in a useful and consistent form, and performing
-housekeeping duties.
+which are applied to the lines.  Rules are solely concerned with
+identifying a line and extracting data from it, whereas the algorithm's
+task is to follow the journey each mail takes through Postfix, piecing the
+data extracted by rules into a coherent whole, saving it in a useful and
+consistent form, and performing housekeeping duties.
 
@@
 
+\label{prior art}
+
 There only appears to be one prior paper published about parsing Postfix
@@
 data in a form suitable for a systems administrator to search using
-grep(1), producing a CSV file\footnote{Comma-Separated Value, a very simple
-format where a record is stored in a single line, with fields separated by
-a comma or other punctuation symbol.  Problems with CSV files include the
+grep(1), awk(1) and the myriad of other Unix utilities already available,
+producing a CSV file which can subsequently be imported into a
+MySQL or Berkeley DB database.  The decision to support both CSV and
+Berkeley DB appears to have been a limiting factor:
+Comma-Separated Value is a very simple format
+where a record is stored in a single line, with fields separated by a comma
+or other punctuation symbol.  Problems with CSV files include the
 difficulty in escaping separators, providing multiple values for a field
 (e.g.\ multiple recipients), adding new fields, and lack of a bundled
-schema describing the fields.} which can subsequently be imported into a
-MySQL or Berkeley DB database.  The data stored is limited to time and
-date, hostname and IP address of client, sender and recipient addresses,
-and the SMTP code.  The decision to support both CSV and Berkeley DB
-appears to have been a limiting factor: the values which can be stored are
-limited, it's difficult to link different records (foreign keys in SQL
-enable this easily), and ad-hoc queries require extensive code to be
-written.  Berkeley DB is not an SQL database, it only supports (key, value)
-pairs; in the main table the value used is a CSV line, and the key is an
-integer referred to by other tables.  The secondary by-sender,
-by-recipient, by-date and by-IP tables use the sender/recipient/date/IP as
-the key, and a CSV list of integers referring to the main table is used as
-the value.  In effect this re-implements SQL foreign keys, but without the
-full functionality offered by even the most basic of SQL databases.  It
-also requires custom code to search on some combination of the above.
-
-Handling of multiple recipients, SMTP codes or remote servers\footnote{A
-single mail may be sent to multiple remote servers if it was addressed to
-recipients in different domains, or Postfix needs to try multiple servers
-for one or more recipients.} is not explained in the paper, but the options
-appear to be:
+schema describing the fields.
+
+Berkeley DB is not an SQL database, it only supports (key, value) pairs; in
+the main table the value used is a CSV line, and the key is an integer
+referred to by other tables.  The secondary by-sender, by-recipient,
+by-date and by-IP tables use the sender/recipient/date/IP as the key, and a
+CSV list of integers referring to the main table is used as the value.  In
+effect this re-implements SQL foreign keys, but without the full
+functionality offered by even the most basic of SQL databases.  It also
+requires custom code to search on some combination of the above, though the
+authors of LMA did provide some queries: IP-STORY, FROM-STORY, DAILY-EMAIL
+and DAILY-REJECT\@.  Berkeley DB appears to be the least useful of the
+three output formats
+
+The data stored is limited to time and date, hostname and IP address of
+client, sender and recipient addresses, and the SMTP code.  Handling of
+multiple recipients, SMTP codes or remote servers\footnote{A single mail
+may be sent to multiple remote servers if it was addressed to recipients in
+different domains, or Postfix needs to try multiple servers for one or more
+recipients.} is not explained in the paper, but the options appear to be:
 
@@
 
+\section{Database schema}
+\label{database schema}
+
+The database schema can be conceptually divided in two: the rules which are
+used to parse log files, and the data saved from the parsing of log files.
+Rules have the fields required to parse the log lines, extract data to be
+saved, and the action to be executed; they also have several fields which
+aid the user in understanding what each rule parses.  The rules are
+described in detail in section~\ref{rules} and the fields in particular in
+section~\ref{rule attributes}, and that discussion won't be repeated here.
+The data saved from parsing the logs is divided into two tables described
+below: connections and results.
+
+\subsection{Connections table}
+
+Every connection and/or mail will have a single entry in the connections
+table containing the following fields:
+
+\begin{description}
+
+    \item [id] This field uniquely identifies the row.
+
+    \item [server\_ip] The IP address (IPv4 or IPv6) of the server: the
+        local server when receiving mail, the remote server when sending
+        mail.
+
+    \item [server\_hostname] The hostname of the server, it will be
+        \textit{unknown\/} if the IP address could not be resolved to a
+        hostname via DNS\@.
+
+    \item [client\_ip] The client IP address (IPv4 or IPv6): the remote
+        server when receiving mail, the local server when sending mail.
+
+    \item [client\_hostname] The hostname of the client, it will be
+        \textit{unknown\/} if the IP address could not be resolved to a
+        hostname via DNS\@.
+
+    \item [helo] The hostname used in the HELO command.  The HELO
+        occasionally changes during a connection, presumably because spam
+        or virus senders think it's a good idea.  By default Postfix only
+        logs the HELO when it rejects an SMTP command, but it is quite easy
+        to rectify this:
+
+\label{logging helo}
+
+        \begin{enumerate}
+
+            \item Create \texttt{/etc/postfix/log\_helo.pcre}
+                containing:\newline
+                \hspace*{2em}\texttt{/./~~~~WARN~Logging~HELO}
+
+            \item Modify \texttt{smtpd\_helo\_restrictions} in
+                \texttt{/etc/postfix/main.cf} to contain\newline
+                \hspace*{2em}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
+
+        \end{enumerate}
+
+        The purpose of logging the HELO in
+        \texttt{smtpd\_helo\_restrictions} is to log only once per mail; it
+        also seems like the natural place.
+
+    \item [queueid] The queueid of the mail if the conection represents an
+        accepted mail, or \textit{NOQUEUE\/} otherwise.
+
+    \item [start] The timestamp of the first log line, in seconds since the
+        epoch.
+
+    \item [end] The timestamp of the last log line, in seconds since the
+        epoch.
+
+\end{description}
+
+\subsection{Results table}
+
+Every log line corresponding to Postfix performing an action other than
+INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
+SMTP command, delivering a mail, or bouncing a mail.  Any log line where
+the Postfix action is INFO is not interesting in and of itself, but
+provides additional information which will be saved with other results;
+IGNORED log lines don't even provide useful information.
+
+\begin{description}
+
+    \item [connection\_id] A reference to the entry in the connections
+        table this results table entry belongs to.
+
+    \item [rule\_id] A reference to the entry in the rules table which
+        matched the log line and created this result.
+
+    \item [warning] Postfix can be configured to log a warning instead of
+        enforcing a restriction that would reject an SMTP command --- a
+        facility that is quite useful for testing new restrictions.  This
+        field will be 1 if the log line parsed was a warning rather than a
+        real rejection, or 0 for a real rejection or any other result.
+
+    \item [smtp\_code] The SMTP code associated with the log line.  In
+        general an SMTP code is only present for a rejection or final
+        delivery; results missing an SMTP code will duplicate the SMTP code
+        of other results in the connection.  Some final delivery log lines
+        don't contain an SMTP code: in those cases the code is faked based
+        on the success or failure represented by the log line.
+
+    \item [sender] The sender's email address.  This can change during one
+        single connection, when the connection is reused to send multiple
+        mails.
+
+    \item [recipient] The recipient address.
+
+    \item [message\_id] The message-id of the accepted mail, or NULL if no
+        mail was accepted.
+
+    \item [data] A field available for anything not covered by other
+        fields, e.g.\ the rejection message from an RBL\@.
+
+    \item [timestamp] The time the line was logged, in seconds since the
+        epoch.
+
+\end{description}
+
 \section{Adaptable parsing: user defined rules}
@@
 
-While the rules have more lines of code than the algorithm, the rules are
-quite simple and each rule is completely independent of its fellows.  The
-algorithm is significantly more complicated and highly internally
-interdependent.
+Where the rules are quite simple and each rule is completely independent of
+its fellows, the algorithm is significantly more complicated and highly
+internally interdependent.  The algorithm deals with all the complications
+of parsing, the eccentricities and oddities of Postfix logs, and presents
+the resulting data in a normalised, easy to use representation.
 
@@
 should be used for a given log line.}  The original connection will be
-discarded unsaved when the client disconnects if there haven't been any
-rejections.  One unsolved difficulty is
+discarded unsaved when the client disconnects if it doesn't have any data
+worth saving, i.e.\ no rejections.  One unsolved difficulty is
 distinguishing between different groups of rejections, e.g.\ when dealing
@@
 
-    \item This is followed by rejection of sender, recipients, client
-        address, etc. (action: rejection; flowchart:~4); acceptance of one
-        or more mails (action: clone; flowchart:~5); or some interleaving
-        of both.
+    \item This is followed by rejection of sender address, recipient
+        addresses, client IP address or hostname, etc. (action: rejection;
+        flowchart:~4); acceptance of one or more mails (action: clone;
+        flowchart:~5); or some interleaving of both.
 
@@
 complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections.  See
-section~\ref{tracking re-injected mail} for a full discussion.
-
-Internally generated mails lack any explicit origin (in Postfix 2.2.x) and
-can only be detected using heuristics (see
+overlaps both the mail acceptance and mail delivery sections, so discussion
+is deferred to section~\ref{tracking re-injected mail}.
+
+Internally generated mails lack any explicit origin in Postfix 2.2.x and
+must be detected using heuristics (see
 section~\ref{identifying-bounce-notifications} for details).  Bounce
@@
 The obvious counterpart to mail entering the system is mail leaving the
-system, whether it is by deletion, bouncing, local or remote delivery.  All
-four are handled in exactly the same way:
+system, whether it is by deletion, bouncing, or local or remote delivery.
+All four are handled in exactly the same way:
 
@@
         child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from memory (flowchart:~15).
-        Re-injected mails are described in section~\ref{tracking
-        re-injected mail}.
+        database (flowchart:~14), then deleted from the state tables
+        (flowchart:~15).  Re-injected mails are described in
+        section~\ref{tracking re-injected mail}.
 
@@
 
-    \item The indicator which more commonly introduces re-injection is qmgr
-        selecting a mail with a previously unseen queueid for delivery
+    \item The indicator which more commonly introduces re-injection is when 
+        \texttt{postfix/qmgr}
+        selects a mail with a previously unseen queueid for delivery
         (action: mail\_picked\_for\_delivery; flowchart:~3), in which case
@@
         This second indicator always occurs for re-injected mail but
-        typically it occurs after the first indicator explained above, and
-        so rarely introduces re-injection.  This indicator is required to
-        tie the parent and child mails together and so is central to the
-        process of tracking re-injected mails.
+        typically after the first indicator explained above.  This
+        indicator is required to tie the parent and child mails together
+        and so is central to the process of tracking re-injected mails.
 
@@
         The parent may have children which are waiting to be entered in the
-        database (flowchart:~25); for each of those children their data is
-        cleaned up and entered in the database, then deleted from the state
-        tables.  The parent may also have outstanding children which are
-        not yet delivered, in which case (flowchart:~26) the parent must
-        wait for those children to be finished with.  As soon as the last
-        child is deleted from the state tables the parent will also be
-        finished with (flowchart:~27), and deleted from the state tables.
+        database (flowchart:~25); each of those children's data is cleaned
+        up and entered in the database, then deleted from the state tables.
+        The parent may also have outstanding children which are not yet
+        delivered, in which case (flowchart:~26) the parent must wait for
+        those children to be finished with.  As soon as the last child is
+        deleted from the state tables the parent will also be finished with
+        (flowchart:~27), and deleted from the state tables.
 
@@
 
+\label{parsing coverage}
+
 The discussion of the parser's coverage of Postfix log files is separated
@@
 algorithm, and it can also be much harder to notice a deficiency.
-
-\label{parsing-coverage}
 
@@
 
-XXX ADD DATES CHECKED FOR ALL PARSERS
-
 \label{other-parsers}
@@
         \newline \url{http://jimsun.linxnet.com/postfix_contrib.html}
-
-    \item [Sawmill Universal Log File Analysis and Reporting] is a
-        general purpose commercial product which parses 687 log file
-        formats (correct as of 2007/04/15) and produces reports.  I have not
+        \newline (Last checked 2007/08/13.)
+
+    \item [Sawmill Universal Log File Analysis and Reporting] is a general
+        purpose commercial product which parses 687 log file formats
+        (correct as of 2007/04/15) and produces reports.  I have not
         experimented with it due to its limited data extraction facilities;
@@
         \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
+        \newline (Last checked 2007/08/13.)
 
@@
         day, per month and years.\/}  \newline
-        \url{http://www.enderunix.org/isoqlog/}
+        \url{http://www.enderunix.org/isoqlog/} \newline (Last checked
+        2007/08/13.)
 
@@
         \url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
+        \newline (Last checked 2007/08/13.)
 
@@
         similar problems occur.  \newline
-        \url{http://home.uninet.ee/~ragnar/throughput_monitor/}
+        \url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline
+        (Last checked 2007/08/13.)
 
@@
 
-        \url{http://anteater.drzoom.ch/}
+        \url{http://anteater.drzoom.ch/} \newline (Last checked
+        2007/08/13.)
 
@@
 
-        \url{http://yaala.org/}
+        \url{http://yaala.org/} \newline (Last checked 2007/08/13.)
 
@@
         minor update since October 2004.  \newline
-        \url{http://logreport.org/lire/}
+        \url{http://logreport.org/lire.html} \newline (Last checked
+        2007/08/13.)
 
@@
         of recipients, bytes transferred and delay.  \newline
-        \url{http://www.itefix.no/phpws/index.php}
-
-    \item [Log Mail Analyser]~\cite{log-mail-analyser} attempts to
-        correlate Postfix log lines by queueid and produce both a CSV file
-        and either a Berkeley DB or MySQL database.  It aims to save
-        \textit{date and hour, DNS name and IP address host, mail server IP
-        address, sender, receiver and e-mail status (sent, rejected)}, so
-        its data extraction is reasonably complete.  The parsing is
-        hard-coded into the program; extending it requires familiarity with
-        a substantial portion of the code.  Producing a CSV file appears to
-        be quite limiting, but the intention is to provide a format
-        suitable for use with grep(1), awk(1) and the myriad of other Unix
-        utilities already available.  Supporting Berkeley DB is also
-        limiting, requiring an understanding of the (partially documented)
-        database structure and writing a program to extract the required
-        data --- it appears to be the least useful of the three output
-        formats, though the authors of LMA did provide some queries:
-        IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.
-        Unfortunately when the author tested LMA it didn't work properly
-        and after exploring the code for some time the author reluctantly
-        gave up.
+        \url{http://www.itefix.no/phpws/index.php} \newline (Last checked
+        2007/08/13.)
+
+    \item [Log Mail Analyser]~\cite{log-mail-analyser} has already been
+        discussed in section~\ref{prior art}.
 

### 2007-09-07 doc/logparser.tex
Improved stats: * Automated extracting data from directories of results; it's not quite right,   as making changes to the directories or contents doesn't cause a rebuild. * Split normal/shuffle/reverse factor graph into two, so they're legible. * Updated log 22 stats, and graphs. * Generate stats for logs 22 and 62-68 mean and standard deviation.
Added introductions and conclusions to several of the sections. Moved some subsections because it made more sense. Some minor improvements. Automated more stats generation. Made 'make veryclean' remove everything (or nearly everything) that's generate

@@
 
-    Parsing Postfix logs is much more difficult than it first appears but
+    Parsing Postfix logs is much more difficult than it first appears, but
     it is possible to achieve a high degree of accuracy in understanding
@@
 
+\subsection{Introduction}
+
+This section provides background information helpful in understanding the
+remainder of the paper.  It begins with a discussion of the motivation
+underlying the project, followed by some technical information: the use of
+a database as an API\@; a brief introduction to the SMTP protocol; and a
+longer introduction to Postfix, concentrating on the topics most relevant
+to this paper, namely Postfix restrictions and policy servers.  The
+assumptions made in designing and implementing the parser are explained, as
+are the conventions used in this paper.  Other projects which attempt to
+parse Postfix logs are summarised (full details are available in
+Section~\ref{other-parsers}), plus a review of previously published
+research in this area.
+
+XXX SHOULD THERE BE A REGEX SECTION\@?
+
 \subsection{Motivation}
@@
 e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\footnote{\texttt{cidr} is the type of the lookup table.} will check
+\footnote{\texttt{cidr} is the type of the lookup table.  XXX PROVIDE A
+REFERENCE} will check
 whether the IP address of the connected client matches the left hand side
@@
 
-\subsection{Parser background}
-
-Before getting into detail about the parser a brief overview is in order.
-The parser is split into two parts: the parsing algorithm and the rules
-which are applied to the lines.  Rules are solely concerned with
-identifying a line and extracting data from it, whereas the algorithm's
-task is to follow the journey each mail takes through Postfix, piecing the
-data extracted by rules into a coherent whole, saving it in a useful and
-consistent form, and performing housekeeping duties.
-
 \subsection{Assumptions}
@@
 \end{tabular}
+
+\subsection{Conventions used in the paper}
+
+The words \textit{connection\/} and \textit{mail\/} are often used
+interchangeably in this paper; in general the word used was chosen based on
+the context it appears in.
 
@@
 subsequent programs to develop upon.
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
 
@@
 
+\subsection{Conclusion}
+
+This section has provided background information on several topics relevant
+to the remainder of the paper.  It started with the motivation behind the
+project, continuing with explanations of:
+
+\begin{itemize}
+
+    \item Using a database as an application programming interface.
+
+    \item The SMTP protocol.
+
+    \item Postfix restrictions and policy servers.
+
+\end{itemize}
+
+The assumptions which must be satisfied for the parser to work correctly
+were discussed; the conventions used in the paper were listed; a brief
+comparison of this project against other Postfix parsers was provided,
+followed by a review of the previously published literature.
+
 \section{Database schema}
 \label{database schema}
+
+The database is an integral part of the parser: it stores the rules and the
+results of applying those rules to Postfix log files.  Understanding the
+database schema is important in understanding the actions of the parser,
+and essential to developing further applications which utilise the data
+gathered.
+
+\subsection{Introduction}
 
@@
 aid the user in understanding what each rule parses.  The rules are
-described in detail in section~\ref{rules} and the fields in particular in
-section~\ref{rule attributes}, and that discussion won't be repeated here.
-The data saved from parsing the logs is divided into two tables described
-below: connections and results.
-
-\subsection{Connections table}
-
-Every connection and/or mail will have a single entry in the connections
-table containing the following fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The IP address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \textit{unknown\/} if the IP address could not be resolved to a
-        hostname via DNS\@.
-
-    \item [client\_ip] The client IP address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \textit{unknown\/} if the IP address could not be resolved to a
-        hostname via DNS\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO when it rejects an SMTP command, but it is quite easy
-        to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline
-                \hspace*{2em}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_helo\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \hspace*{2em}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        The purpose of logging the HELO in
-        \texttt{smtpd\_helo\_restrictions} is to log only once per mail; it
-        also seems like the natural place.
-
-    \item [queueid] The queueid of the mail if the conection represents an
-        accepted mail, or \textit{NOQUEUE\/} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch.
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-Every log line corresponding to Postfix performing an action other than
-INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
-SMTP command, delivering a mail, or bouncing a mail.  Any log line where
-the Postfix action is INFO is not interesting in and of itself, but
-provides additional information which will be saved with other results;
-IGNORED log lines don't even provide useful information.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the entry in the connections
-        table this results table entry belongs to.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an SMTP command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection or any other result.
-
-    \item [smtp\_code] The SMTP code associated with the log line.  In
-        general an SMTP code is only present for a rejection or final
-        delivery; results missing an SMTP code will duplicate the SMTP code
-        of other results in the connection.  Some final delivery log lines
-        don't contain an SMTP code: in those cases the code is faked based
-        on the success or failure represented by the log line.
-
-    \item [sender] The sender's email address.  This can change during one
-        single connection, when the connection is reused to send multiple
-        mails.
-
-    \item [recipient] The recipient address.
-
-    \item [message\_id] The message-id of the accepted mail, or NULL if no
-        mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an RBL\@.
-
-    \item [timestamp] The time the line was logged, in seconds since the
-        epoch.
-
-\end{description}
-
-\section{Adaptable parsing: user defined rules}
-
-\label{rules}
-
-The complexity and variation in Postfix's logs requires similar flexibility
-in the parser; decoupling the parsing rules from the associated actions
-allows new rules to be written and tested without requiring modifications
-to the algorithm source code (significantly lowering the barrier to entry
-for new or casual users), and greatly simplifies both algorithm and rules.
-It also creates a clear separation of functionality: rules handle low level
-details of identifying log lines and extracting data from a line, whereas
-the algorithm handles the higher level details of following the path a mail
-takes through Postfix, assembling the required data, etc.
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item The first matching rule wins: no further rules are tried against
-        that line, but there is a facility for specifying the order of
-        rules so that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item There are no sub-rules, so rules have linear computational
-        complexity.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Rule attributes}
+described in detail in section~\ref{rules} but the fields are covered in
+section~\ref{rule attributes}.
+
+The data saved from parsing the logs is also divided into two tables as
+described below: connections and results.  The connections table contains a
+row for every mail accepted and every connection where there was a
+rejection; the individual fields will be described in the forthcoming
+section~\ref{connections table}.  The results table has one or more rows
+per row in the connections table, depending on the specifics of the
+mail/connection; the fields will be covered in detail in
+section~\ref{results table}.
+
+\subsection{Rules table}
 
 \label{rule attributes}
+
+Rules are discussed in detain in section~\ref{rules}, but the rules table
+is covered here.  Rules are created by the user, not the parser, and will
+not be modified by the parser (except for the hits and hits\_total fields).
+Rules parse the individual lines, telling the parser which fields to
+extract and what action to take for that line.
 
@@
 \end{description}
+
+\subsection{Connections table}
+
+\label{connections table}
+
+Every accepted mail and every connection where there was a rejection will
+have a single entry in the connections table containing the following
+fields:
+
+\begin{description}
+
+    \item [id] This field uniquely identifies the row.
+
+    \item [server\_ip] The IP address (IPv4 or IPv6) of the server: the
+        local server when receiving mail, the remote server when sending
+        mail.
+
+    \item [server\_hostname] The hostname of the server, it will be
+        \textit{unknown\/} if the IP address could not be resolved to a
+        hostname via DNS\@.
+
+    \item [client\_ip] The client IP address (IPv4 or IPv6): the remote
+        server when receiving mail, the local server when sending mail.
+
+    \item [client\_hostname] The hostname of the client, it will be
+        \textit{unknown\/} if the IP address could not be resolved to a
+        hostname via DNS\@.
+
+    \item [helo] The hostname used in the HELO command.  The HELO
+        occasionally changes during a connection, presumably because spam
+        or virus senders think it's a good idea.  By default Postfix only
+        logs the HELO when it rejects an SMTP command, but it is quite easy
+        to rectify this:
+
+\label{logging helo}
+
+        \begin{enumerate}
+
+            \item Create \texttt{/etc/postfix/log\_helo.pcre}
+                containing:\newline
+                \hspace*{2em}\texttt{/./~~~~WARN~Logging~HELO}
+
+            \item Modify \texttt{smtpd\_helo\_restrictions} in
+                \texttt{/etc/postfix/main.cf} to contain\newline
+                \hspace*{2em}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
+
+        \end{enumerate}
+
+        The purpose of logging the HELO in
+        \texttt{smtpd\_helo\_restrictions} is to log only once per mail; it
+        also seems like the natural place.
+
+    \item [queueid] The queueid of the mail if the conection represents an
+        accepted mail, or \textit{NOQUEUE\/} otherwise.
+
+    \item [start] The timestamp of the first log line, in seconds since the
+        epoch.
+
+    \item [end] The timestamp of the last log line, in seconds since the
+        epoch.
+
+\end{description}
+
+\subsection{Results table}
+
+\label{results table}
+
+Every log line corresponding to Postfix performing an action other than
+INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
+SMTP command, delivering a mail, or bouncing a mail.  Any log line where
+the Postfix action is INFO is not interesting in and of itself, but
+provides additional information which will be saved with other results;
+IGNORED log lines don't even provide useful information.  Each row is
+associated with a single connection, though there may be many results per
+connection.
+
+\begin{description}
+
+    \item [connection\_id] A reference to the entry in the connections
+        table this results table entry belongs to.
+
+    \item [rule\_id] A reference to the entry in the rules table which
+        matched the log line and created this result.
+
+    \item [warning] Postfix can be configured to log a warning instead of
+        enforcing a restriction that would reject an SMTP command --- a
+        facility that is quite useful for testing new restrictions.  This
+        field will be 1 if the log line parsed was a warning rather than a
+        real rejection, or 0 for a real rejection or any other result.
+
+    \item [smtp\_code] The SMTP code associated with the log line.  In
+        general an SMTP code is only present for a rejection or final
+        delivery; results missing an SMTP code will duplicate the SMTP code
+        of other results in the connection.  Some final delivery log lines
+        don't contain an SMTP code: in those cases the code is faked based
+        on the success or failure represented by the log line.
+
+    \item [sender] The sender's email address.  This can change during one
+        single connection, when the connection is reused to send multiple
+        mails.
+
+    \item [recipient] The recipient address.
+
+    \item [message\_id] The message-id of the accepted mail, or NULL if no
+        mail was accepted.
+
+    \item [data] A field available for anything not covered by other
+        fields, e.g.\ the rejection message from an RBL\@.
+
+    \item [timestamp] The time the line was logged, in seconds since the
+        epoch.
+
+\end{description}
+
+\subsection{Conclusion}
+
+The table containing the rules used by the parser, and the two tables
+containing the data extracted from the Postfix logs were described, with
+the purpose of each field discussed in detail.  A clear, comprehensible
+schema is essential in using the extracted data; it's more important for
+using the data than storing the data, because storing the data is a
+write-once operation, whereas utilising the data requires frequent
+searching, sorting and manipulation of the data to produce customised
+reports and/or statistics.
+
+\section{Adaptable parsing: user defined rules}
+
+\label{rules}
+
+\subsection{Introduction}
+
+This section discusses the rules used in parsing Postfix log files,
+starting with rule characteristics, continuing with detecting and dealing
+with overlapping rules, and the problems they can cause.  An example rule
+and a line it would match are provided, plus a description of how the
+fields in the rule are used when matching and performing the action (the
+details of the table containing the rules have already been described in
+section~\ref{rule attributes},).  This section finishes with a discussion
+of rule efficiency concerns, referring to the graphs in
+section~\ref{graphs}.
+
+\subsection{Rule characteristics}
+
+The parser is split into two parts: the parsing algorithm and the rules
+which are applied to the lines.  The complexity and variation in Postfix's
+logs requires similar flexibility in the parser; decoupling the parsing
+rules from the associated actions allows new rules to be written and tested
+without requiring modifications to the algorithm source code (significantly
+lowering the barrier to entry for new or casual users), and greatly
+simplifies both algorithm and rules.  It also creates a clear separation of
+functionality: rules handle low level details of identifying log lines and
+extracting data from a line, whereas the algorithm handles the higher level
+details of following the path a mail takes through Postfix, assembling the
+required data, etc.
+
+Rule have certain characteristics which may help in understanding the
+parser:
+
+\begin{itemize}
+
+    \item The first matching rule wins: no further rules are tried against
+        that line, but there is a facility for specifying the order of
+        rules so that more specific rules can be tried first.
+
+    \item Rules are completely self-contained and can be understood in
+        isolation, without reference to any other rules.
+
+    \item There are no sub-rules, so rules have linear computational
+        complexity.
+
+\end{itemize}
+
+\label{comparison against context-free grammars}
+
+In context-free grammar terms the parser rules could be described as:
+
+$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
+\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
 
@@
 
-\begin{description}
-
-    \item [shuffled] 6--11\% increase; mean 8.34, standard deviation 2.44.
-
-    \item [reversed] 15--22\% increase; mean 17.97, standard deviation
-        2.94.
-
-\end{description}
-
-Overall this increases the parser's efficiency by 5.5--10\%; a modest but
-worthwhile increase.
+\input{stats-normal-shuffle-reverse-percentage}
+
+Overall this provides a modest but worthwhile performance increase.
 
@@
 more efficient; graph~\ref{normal regex vs disarded regex factor} shows the
-percentage execution time increase when not caching each regex is between
-350--570\% (mean 465.58\%, standard deviation 114.70\%).
+percentage execution time increase when not caching each regex.  Over all
+93 log files the execution time increase when not caching each compiled
+regex works out at:
+
+XXX IMPROVE THIS\@.
+
+\input{stats-cached-discarded-include-for-discussion}
+
+\subsection{Conclusion}
+
+This section dealt with the rules used in parsing Postfix log files:
+
+\begin{itemize}
+
+    \item The characteristics of the rules were described.
+
+    \item Detecting overlapping rules and dealing with the problems they
+        can cause was covered, including a discussion of why overlapping
+        rules can be helpful as well as harmful.
+
+    \item An example log line and the rule matching it was given, followed
+        by a description of how the fields in the rule are used both in the
+        matching and the subsequent action that is executed.
+
+    \item The database table containing the rules is dealt with in
+        section~\ref{rule attributes}, and is not duplicated in this
+        section.
+
+    \item The final part of this section discusses rule efficiency,
+        covering the effects of caching compiled regexs and optimal
+        ordering ordering of rules, with reference to the graphs in
+        section~\ref{graphs}.
+
+\end{itemize}
 
@@
 of parsing, the eccentricities and oddities of Postfix logs, and presents
-the resulting data in a normalised, easy to use representation.
-
-\subsection{A naive approach}
+the resulting data in a normalised, easy to use representation.  The
+algorithm's task is to follow the journey each mail takes through Postfix,
+piecing the data extracted by rules into a coherent whole, saving it in a
+useful and consistent form, and performing housekeeping duties.
+
+\subsection{Introduction}
+
+This section covers the following topics:
+
+\begin{itemize}
+
+    \item A high level overview of the algorithm.
+
+    \item The first set of complications encountered: initially obvious
+        difficulties which had to be overcome.
+
+    \item A flow chart showing the paths a mail can take through the
+        algorithm.
+
+    \item A description of the full algorithm.
+
+    \item The actions which can be taken by the parser are described in
+        detail.
+
+    \item Additional complications which have arisen during the development
+        of this parser are described.
+
+\end{itemize}
+
+\subsection{A high level overview}
 
@@
 
-Unfortunately it's not that straightforward.
+Unfortunately that ignores the many complications encountered.
 
@@
 This flow chart shows the paths the data representing a mail/connection can
-take through the parser algorithm.
+take through the parser algorithm.  The flow chart covers the most common
+paths only; there are additional, uncommon paths which are excluded for the
+sake of clarity.
 
@@
 algorithm description.  The data flow chart in section~\ref{flow-chart}
-should also be consulted while reading this section.
+should also be consulted while reading this section.  For the sake of
+clarity this description covers the most common paths through the
+algorithm, as including every path would hinder understanding.  The more
+uncommon paths are caused by the complications described in
+section~\ref{additional complications}, and are covered therein.
 
@@
 \label{additional complications}
+
+XXX EXPLAIN EXACTLY WHERE IN THE ALGORITHM THE \newline CHECKS ETC\@. KICK
+IN\@.
 
@@
 
+\subsection{Conclusion}
+
 \section{Coverage}
@@
 \label{parsing coverage}
+
+\subsection{Introduction}
 
@@
 
+\subsection{Conclusion}
+
 \section{Limitations and possible improvements}
@@
 \label{limitations-improvements}
+
+\subsection{Introduction}
 
@@
 
+\subsection{Conclusion}
+
 \section{Conclusion}
@@
 \label{other-parsers}
+
+\subsection{Introduction}
+
+\subsection{Parsers reviewed}
 
@@
 
+\subsection{Conclusion}
+
 \bibliographystyle{logparser-bibliography-style}
@@
 
+\subsection{Introduction}
+
 \subsection{Parser scalability}
@@
 \label{normal vs shuffled vs reversed ordering factor}
-\includegraphics{plot-normal-shuffle-reverse-factor.ps}
+\includegraphics{plot-normal-shuffle-factor.ps}
+\newline
+\includegraphics{plot-normal-reverse-factor.ps}
 
@@
 
-\begin{tabular}[]{lrr}
-                                    & Mean      & Standard Deviation    \\
-    All logs                        & 465.58\%  & 114.70\%              \\
-    All logs except 22 and 62--68   & 497.43\%  & 50.58\%               \\
-    Logs 22 and 62--68              & 127.16\%  & 21.28\%               \\
-
-\end{tabular}
+\input{stats-cached-discarded-include-for-graph}
+
+\subsection{Conclusion}
 

### 2007-09-19 doc/logparser.tex
Moved all generated files to build subdirectory. Added introduction to graphs section. Added some more stats. Removed generated files from repository, they need to be rebuilt properly.

@@
 
-\input{stats-normal-shuffle-reverse-percentage}
+\input{build/stats-normal-shuffle-reverse-percentage}
 
@@
 
-\input{stats-cached-discarded-include-for-discussion}
+\input{build/stats-cached-discarded-include-for-discussion}
 
@@
 
-\includegraphics{logparser-flow-chart.ps}
+\includegraphics{build/logparser-flow-chart.ps}
 
@@
 
+Graphs are an excellent means of displaying data, transforming a
+meaningless stream of numbers into an easily comprehensible form, where
+anomalies and patterns are immediately obvious.  These graphs are used to
+illustrate the topics discussed in section~\ref{rule efficiency}.  The
+graphs in the first section cover parser scalability, demonstrating that
+its performance scales linearly with input size.  The impact of rule
+ordering is shown in section~\ref{rule ordering graphs}, and the lower
+impact of rule ordering on larger log files is explained.  The final group
+of graphs vividly shows the huge impact that caching compiled regexs has on
+parser performance.
+
 \subsection{Parser scalability}
@@
 \label{execution time vs file size vs number lines}
-\includegraphics{plot-normal-filesize-numlines.ps}
+\includegraphics{build/plot-normal-filesize-numlines.ps}
 
@@
 \label{execution time vs file size vs number lines factor}
-\includegraphics{plot-normal-filesize-numlines-factor.ps}
+\includegraphics{build/plot-normal-filesize-numlines-factor.ps}
+
+\input{build/stats-normal-filesize-line-count-include}
 
 \subsection{Rule ordering}
+
+\label{rule ordering graphs}
 
@@
 \label{rule hits}
-\includegraphics{plot-hits.ps}
+\includegraphics{build/plot-hits.ps}
 
@@
 \label{normal vs shuffled vs reversed ordering}
-\includegraphics{plot-normal-shuffle-reverse.ps}
+\includegraphics{build/plot-normal-shuffle-reverse.ps}
 
@@
 \label{normal vs shuffled vs reversed ordering factor}
-\includegraphics{plot-normal-shuffle-factor.ps}
+\includegraphics{build/plot-normal-shuffle-factor.ps}
 \newline
-\includegraphics{plot-normal-reverse-factor.ps}
-
-The worst-case performance of shuffled rules can be seen at log file 30,
-where it is equivalent to reversed rules.
+\includegraphics{build/plot-normal-reverse-factor.ps}
 
@@
 
+This non-intuitive effect will not be exhibited by log files in general: in
+a normally functioning mail system the impact of rule ordering will scale
+linearly with log file size.  The skewed distribution of log lines in the
+specific log files creates the anomaly described herein.
+
 \subsubsection{Log 22 distribution by rule}
 \label{log22 distribution by rule}
-\includegraphics{plot-log22-distribution-by-rule.ps}
+\includegraphics{build/plot-log22-distribution-by-rule.ps}
 
@@
 \label{log22 line and rule distribution by program}
-\includegraphics{plot-log22-distribution-versus-rule-count.ps}
+\includegraphics{build/plot-log22-distribution-versus-rule-count.ps}
 
@@
 \label{normal regex vs discard regex}
-\includegraphics{plot-cached-discarded.ps}
+\includegraphics{build/plot-cached-discarded.ps}
 
@@
 \label{normal regex vs disarded regex factor}
-\includegraphics{plot-cached-discarded-factor.ps}
+\includegraphics{build/plot-cached-discarded-factor.ps}
 
@@
 
-\input{stats-cached-discarded-include-for-graph}
+\input{build/stats-cached-discarded-include-for-graph}
 

### 2007-09-23 doc/logparser.tex
Improved stats, added new graphs, improved explanations and updated source data. Added conclusions to parsing algorithm, parser coverage, limitations and graphs sections.

@@
 e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\footnote{\texttt{cidr} is the type of the lookup table.  XXX PROVIDE A
-REFERENCE} will check
-whether the IP address of the connected client matches the left hand side
-of each line in the file and will return the right hand side of the first
-matching line.  Other restrictions determine their result in different
-ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks whether the
-IP address of the client is present in the listed RBL\@.
+\footnote{\texttt{cidr} is the type of the lookup table \cite{cidr_table}}
+will check whether the IP address of the connected client matches the left
+hand side of each line in the file and will return the right hand side of
+the first matching line.  Other restrictions determine their result in
+different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
+whether the IP address of the client is present in the listed RBL\@.
 
@@
 vs reversed ordering}, where the difference is not readily apparent, but
-graph~\ref{normal vs shuffled vs reversed ordering factor} shows the
-percentage increase of execution times:
-
-\input{build/stats-normal-shuffle-reverse-percentage}
-
-Overall this provides a modest but worthwhile performance increase.
+graphs~\ref{normal vs shuffled vs reversed ordering factor} show the
+percentage increase of execution times (mean and standard deviation are
+shown after the graph).  Overall this provides a modest but worthwhile
+performance increase.
 
@@
 Graph~\ref{normal regex vs discard regex} shows execution times with and
-without caching the regex; also shown on the graph are the file sizes and
-number of lines in each file.  Caching the compiled regexs is obviously far
+without caching the regex.  Caching the compiled regexs is obviously far
 more efficient; graph~\ref{normal regex vs disarded regex factor} shows the
-percentage execution time increase when not caching each regex.  Over all
-93 log files the execution time increase when not caching each compiled
-regex works out at:
-
-XXX IMPROVE THIS\@.
-
-\input{build/stats-cached-discarded-include-for-discussion}
+percentage execution time increase when not caching each regex.
+
+Caching the compiled regexs is quite simple, and is the single most
+effective optimisation tried in the parser.
 
@@
 
+This section has presented the parsing algorithm, divided into: a high
+level overview of the basic algorithm, followed by the first group of
+difficulties thwarted.  The revised algorithm was detailed, preceded by a
+flow chart depicting the paths a mail may take through the algorithm.  The
+actions which are available to rules were described in full, and finally
+the remaining complications discovered and overcome in the process of
+refining and completing the algorithm.  Detecting, diagnosing and defeating
+complications forms the largest single subsection of this document,
+mirroring the development of the parser,  The complications are described
+in the order they were overcome, with subsequent problems affecting fewer
+mails (often by an order or magnitude), though the time required to solve
+problems increased with each successive problem.
+
 \section{Coverage}
@@
 
+Parser coverage is divided into two topics in this section: log lines
+covered, and mails covered.  The former is initially more important, as the
+parser must successfully parse every line if it is to be complete, but
+subsequently the latter takes precedence because reproducing the path a
+mail takes through Postfix is the aim of the parser.  Increasing the
+percentage of log lines parsed is relatively simple and not intrusive:
+adding new rules or modifying existing rules is simplified by the parser
+implementation.  Improving the logical coverage is harder, as the actions
+taken by Postfix must be reconstructed, and the new sequence of actions
+integrated into the existing model without breaking the existing sequences.
+detecting a deficiency in the parsing algorithm is also significantly
+harder than detecting unparsed log lines, as the parser will warn about any
+unparsed line, whereas the author must spot a flaw in the parser by
+understanding the warnings generated.
+
 \section{Limitations and possible improvements}
@@
 \subsection{Conclusion}
+
+This section has covered the limitations of the parser and possible
+improvements which may be implemented in the future.
 
@@
 graphs in the first section cover parser scalability, demonstrating that
-its performance scales linearly with input size.  The impact of rule
-ordering is shown in section~\ref{rule ordering graphs}, and the lower
-impact of rule ordering on larger log files is explained.  The final group
-of graphs vividly shows the huge impact that caching compiled regexs has on
-parser performance.
+performance scales linearly with input size.  The impact of rule ordering
+is shown in section~\ref{rule ordering graphs}, and the lower impact of
+rule ordering on larger log files is explained.  The final group of graphs
+vividly shows the huge impact that caching compiled regexs has on parser
+performance.
 
@@
 \label{execution time vs file size vs number lines}
+
+The Y axis represents the following:
+
+\begin{enumerate}
+
+    \item The time required, in seconds, to parse the log file.
+
+    \item The size of the log file, in megabytes.
+
+    \item The number of lines in the log file, divided by 10000.
+
+\end{enumerate}
+
+\noindent
 \includegraphics{build/plot-normal-filesize-numlines.ps}
@@
 \label{execution time vs file size vs number lines factor}
+
+\noindent
 \includegraphics{build/plot-normal-filesize-numlines-factor.ps}
 
+\input
 \input{build/stats-normal-filesize-line-count-include}
@@
 \label{rule ordering graphs}
-
-\subsubsection{Rule hits}
-\label{rule hits}
-\includegraphics{build/plot-hits.ps}
 
@@
 
-\subsubsection{Log 22 distribution by rule}
+\noindent\input{build/stats-normal-shuffle-reverse-include.tex}
+
+\subsubsection{Log 22 distribution}
 \label{log22 distribution by rule}
@@
 \label{log22 line and rule distribution by program}
+
+\noindent
 \includegraphics{build/plot-log22-distribution-versus-rule-count.ps}
+
+\subsubsection{Log 22 lines per rule}
+\label{log22 lines per rule}
+
+As shown in this graph, the most common log lines (see graph~\ref{log22
+line and rule distribution by program}) are from programs which have a low
+number of rules, thus reducing the average parsing time per line, and the
+overall execution time relative to the input size.
+
+\noindent
+\includegraphics{build/plot-log22-distribution-versus-rule-count-factor.ps}
+
+\subsubsection{Rule hits}
+\label{rule hits}
+
+\noindent\includegraphics{build/plot-hits.ps}
 
@@
 size in graph~\ref{execution time vs file size vs number lines}.  The
-anomaly at files 62--68 is caused by a combination of a mail loop and a
-delivery failure which resulted in a huge number of log lines, all of which
-were matched by a single rule.  Sorting the rules by the number of
-successful matches has the effect that this rule will be tried first, so
-the average number of regexs compiled per line drops sharply for those
-eight log files, resulting in a much lower increase in relative execution
-time, as detailed below.
+anomalies are caused by a combination of a mail loop and a delivery failure
+which resulted in a huge number of log lines, all of which were matched by
+a single rule.  Sorting the rules by the number of successful matches has
+the effect that this rule will be tried first, so the average number of
+regexs compiled per line drops sharply for those eight log files, resulting
+in a much lower increase in relative execution time, as detailed below.
 
@@
 
+The graphs presented in this section illustrate the topics discussed in
+section~\ref{rule efficiency}.  The first collection of graphs are about
+parser scalability, showing the linearly relationship between execution
+time and input size.  Section~\ref{rule ordering graphs} relates to the
+effect of rule ordering on execution time, and the unexpected consequences
+of specific inputs.  The necessity of caching compiled regexs is attested
+to by the final group of graphs, where the difference between caching and
+discarding is staggering.
+
 \end{document}

### 2007-09-24 doc/logparser.tex
Expanded the prior art section, making it a comparison rather than a critique.

@@
 %\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
+\newcommand{\parsername}{ASO }
 % Include images
@@
 
-XXX SHOULD THERE BE A REGEX SECTION\@?
-
 \subsection{Motivation}
@@
 data in a form suitable for a systems administrator to search using
-grep(1), awk(1) and the myriad of other Unix utilities already available,
-producing a CSV file which can subsequently be imported into a
-MySQL or Berkeley DB database.  The decision to support both CSV and
-Berkeley DB appears to have been a limiting factor:
-Comma-Separated Value is a very simple format
-where a record is stored in a single line, with fields separated by a comma
-or other punctuation symbol.  Problems with CSV files include the
-difficulty in escaping separators, providing multiple values for a field
-(e.g.\ multiple recipients), adding new fields, and lack of a bundled
-schema describing the fields.
-
-Berkeley DB is not an SQL database, it only supports (key, value) pairs; in
-the main table the value used is a CSV line, and the key is an integer
-referred to by other tables.  The secondary by-sender, by-recipient,
-by-date and by-IP tables use the sender/recipient/date/IP as the key, and a
-CSV list of integers referring to the main table is used as the value.  In
-effect this re-implements SQL foreign keys, but without the full
-functionality offered by even the most basic of SQL databases.  It also
-requires custom code to search on some combination of the above, though the
-authors of LMA did provide some queries: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  Berkeley DB appears to be the least useful of the
-three output formats
+grep(1), awk(1) and the myriad of other standard Unix text processing
+utilities already available, producing a CSV file which can subsequently be
+imported into a MySQL or Berkeley DB database.  The decision to support
+both CSV and Berkeley DB appears to have been a limiting factor:
+Comma-Separated Value is a very simple format where a record is stored in a
+single line, with fields separated by a comma or other punctuation symbol.
+Problems with CSV files include the difficulty in escaping separators,
+providing multiple values for a field (e.g.\ multiple recipients), adding
+new fields, and lack of a bundled schema describing the fields.
+
+Berkeley DB is not an SQL database, supporting only (key, value) pairs; in
+the main table the key is an integer referred to by secondary tables and
+the value used is a CSV line containing all the data for that row.  The
+secondary by-sender, by-recipient, by-date and by-IP tables use the
+sender/recipient/date/IP as the key, and a CSV list of integers referring
+to the main table is the value.  In effect this re-implements SQL foreign
+keys, but without the functionality offered by even the most basic of SQL
+databases (joins, ordering, searches, etc.).  It also requires custom code
+to search on some combination of the above, though the authors of LMA did
+provide some queries: IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.
+Berkeley DB appears to be the least useful of the three output formats
 
@@
 
-Parsing is hand-coded and requires major changes to the code to parse new
-log lines or extract additional data.  It does not appear to deal with any
-of the complications discussed in
+\parsername is bundled with extensive documentation, including an in-depth
+explanation of the database schema.  It supports multiple values for every
+field for which multiple values are reasonable (e.g. the sender address
+should have one value only, whereas the recipient address can have
+multiple values).  The default database it uses is SQLite3, which supports
+most standard SQL features (its lacks full foreign key support), has
+libraries available for most common programming languages, and a simple
+text based client allowing ad-hoc queries with minimal effort.
+
+Parsing in LMA is hand-coded and requires major changes to the code to
+parse new log lines or extract additional data.  It does not appear to deal
+with any of the complications discussed in
 sections~\ref{complications}~and~\ref{additional complications}.  The
 program fails to parse Postfix 2.2.x or 2.3.x log files; indeed it produces
-a number of warnings and no usable output.
+a number of warnings and no usable output.  Parsing in \parsername is
+divided in two (see sections~\ref{rules} and ~\ref{parsing-algorithm} for a
+more detailed explanation): parsing a new log line is a trivial matter of
+adding a new rule, but code changes may be required if new complications
+(sections~\ref{complications} and~\ref{additional complications}) must be
+dealt with or a new action is required (section~\ref{actions-in-detail}).
+
+LMA does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL and
+DAILY-REJECT); the author was unable to parse any logs with LMA, so didn't
+have an opportunity to review the reports.  \parsername does not have any
+bundled reports, it delegates that task to a separate, dedicated program.
 
@@
 program designed to be extensible and useful in a production environment.
-
-XXX EXPLAIN WHY ASO IS MORE COMPREHENSIVE, WHAT IT DOES BETTER AND WHY\@.
-MAKE THIS A COMPARISON INSTEAD OF A CRITIQUE\@.
 

### 2007-09-24 doc/logparser.tex
Expanded the additional complications section, explaining where in the algorithm the checks are made, for those complications which didn't already have that information. Minor cleanups.

@@
 %\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
-\newcommand{\parsername}{ASO }
+\newcommand{\parsername}{ASO}
 % Include images
@@
 e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\footnote{\texttt{cidr} is the type of the lookup table \cite{cidr_table}}
+\footnote{\texttt{cidr} is the type of the lookup table~\cite{cidr_table}}
 will check whether the IP address of the connected client matches the left
@@
 
-\parsername is bundled with extensive documentation, including an in-depth
+\parsername{} is bundled with extensive documentation, including an in-depth
 explanation of the database schema.  It supports multiple values for every
-field for which multiple values are reasonable (e.g. the sender address
+field for which multiple values are reasonable (e.g.\ the sender address
 should have one value only, whereas the recipient address can have
@@
 program fails to parse Postfix 2.2.x or 2.3.x log files; indeed it produces
-a number of warnings and no usable output.  Parsing in \parsername is
-divided in two (see sections~\ref{rules} and ~\ref{parsing-algorithm} for a
+a number of warnings and no usable output.  Parsing in \parsername{} is
+divided in two (see sections~\ref{rules} and~\ref{parsing-algorithm} for a
 more detailed explanation): parsing a new log line is a trivial matter of
@@
 DAILY-REJECT); the author was unable to parse any logs with LMA, so didn't
-have an opportunity to review the reports.  \parsername does not have any
+have an opportunity to review the reports.  \parsername{} does not have any
 bundled reports, it delegates that task to a separate, dedicated program.
@@
 
-XXX EXPLAIN EXACTLY WHERE IN THE ALGORITHM THE \newline CHECKS ETC\@. KICK
-IN\@.
-
 \subsubsection{Identifying bounce notifications}
@@
 bounce message.
+
+This check is performed during the COMMIT action.
 
@@
 
+This check is performed in the DISCONNECT action; it requires support in
+the CLONE action where a list of cloned connections is maintained.
+
 \subsubsection{Further aborted delivery attempts}
@@
 disconnected.  There will be no further entries logged for such mails, so
-without identifying and discarding them they accumulate in the state table
-and will cause clashes if the queueid is reused.  The mail cannot be
-entered in the database as the only data available is the client hostname
-and IP address, but the database schema (see section~\ref{database schema})
-requires many more fields be populated.
+without identifying and entering them in the database immediately they
+accumulate in the state table and will cause clashes if the queueid is
+reused.
 
@@
 
+During the TIMEOUT action if the mail lacks a cleanup line, the queueid and
+other data about the mail, particularly the timestamp of the last log line,
+are added to a global cache of timeout queueids.  This cache is
+subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action, and
+if the queueid exists and the cleanup line passes the checks listed above
+it will be discarded.
+
 \subsubsection{Pickup logging after cleanup}
@@
 
+This complication is dealt with during the PICKUP action.
+
 \subsubsection{Smtpd stops logging}
@@
 
+These occurrences are handled by the three actions POSTFIX\_RELOAD,
+SMTPD\_DIED and SMTPD\_WATCHDOG\@.
+
 \subsubsection{Out of order log lines}
@@
 after cleanup}.  In the 93 log files used for testing this occurs only five
-times in 607,21,709 log lines, but for completeness of the algorithm it
+times in 60,721,709 log lines, but for completeness of the algorithm it
 should be dealt with.  The five occurrences in the test log files have the
@@
 logged messages for each mail, comparing the list against a table of
-known-good combinations of programs.  If the mail's combination is found in
-the valid list the mail can be entered in the database; if the combination
-is not found entry must be postponed and the mail flagged.  The
-SAVE\_BY\_QUEUEID action checks for the flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will succeed, otherwise it must be postponed once more.
+known-good combinations of programs (this check is performed during the
+COMMIT action).  If the mail's combination is found in the valid list the
+mail can be entered in the database; if the combination is not found entry
+must be postponed and the mail flagged.  The SAVE\_BY\_QUEUEID action
+checks for the flag and retries entry if it's found; if the additional log
+lines have caused the mail to reach a valid combination entry will succeed,
+otherwise it must be postponed once more.
 
@@
 
-\input
 \input{build/stats-normal-filesize-line-count-include}

### 2007-09-25 doc/logparser.tex
Critique of splunk.

@@
 
-    \item [Splunk] XXX INVESTIGATE FURTHER \url{http://www.splunk.com}
+    \item [Splunk] \textit{Splunk is an IT Search engine. It is software
+        that indexes any format of IT data from any source in real time,
+        including logs, configurations, scripts, code, messages, traps,
+        alerts, activity reports, stack traces and metrics from all of your
+        applications, servers and devices. Splunk lets you search,
+        navigate, alert and report on all your IT data in real time using
+        an AJAX web interface. You can also share knowledge and Splunk
+        solutions with other members of the Splunk community via
+        SplunkBase.\/}
+
+        Splunk aims to index all an organisation's logs, providing a
+        centralised view capable of searching and correlating diverse log
+        sources.  The web interface allows complicated searches, providing
+        statistics and graphs in real time.  Saved searches can be
+        configured to run periodically and the results emailed to a
+        recipient or sent to a shell script, which presumably can publish
+        the results as required (though possibly without the graphs and
+        detailed statistics).  In the specific case of parsing Postfix
+        logs, Splunk extracts some standard fields: to and from addresses,
+        HELO, date, host the logs were collected from, and protocol.  It
+        parses the standard syslog fields at the beginning of the line, and
+        extracts any \texttt{key=<value>} pairs; these are the fields
+        listed above.  Searches can be made based on the extracted fields,
+        and all text in the line is also available for searching.  Many
+        types of reports are available, though all are variations of a bar
+        or pie chart, with the exception of bubble and heatmap graphs.  It
+        is very easy to drill down through the graphs to extract a portion
+        of the data (e.g.\ select the hour with the largest number of
+        events, then select a particular host, and finally a specific
+        address).  It is not possible to search on partial words, though
+        the web interface does provide and refine a list of possible search
+        terms as you type.  The searches are quite fast, though as the free
+        version has a limit on the amount of data indexed per day the
+        volume of data to be searched was merely 45mb; the cheapest
+        licensed version costs \$6000, and it still limits the volume of
+        data indexed per day.
+
+        The interface is optimised for interactive rather than automated
+        queries and it does not appear to be possible to write independent
+        tools to utilise the Splunk database.  Splunk is a generic tool, so
+        it lacks any Postfix specific support over and above extracting
+        the \texttt{key=<value>} fields; most importantly it makes no
+        attempt to correlate log lines by queueid or pid, nor to handle any
+        of the myriad complications discussed in this document.
+        Correlation of log lines is left up to the user, and Splunk
+        certainly makes the job easier than using grep, awk and similar
+        Unix tools unless the user has a very high proficiency with those
+        tools.
+
+        \url{http://www.splunk.com}
 

### 2007-09-26 doc/logparser.tex
Improved the critiques of: pflogsumm, Sawmill, Isoqlog, AWStats and Anteater.

@@
 
-    \item [pflogsumm.pl] \textit{pflogsumm.pl is designed to provide an
-        over-view of postfix activity, with just enough detail to give the
+    \item [Pflogsumm] \textit{pflogsumm is designed to provide an over-view
+        of Postfix activity, with just enough detail to give the
         administrator a ``heads up'' for potential trouble spots.\/}
-        \newline \url{http://jimsun.linxnet.com/postfix_contrib.html}
+
+        Pflogsumm produces a report designed for troubleshooting, and does
+        not support saving any data; parsing is fixed and concentrates on
+        extracting only the data required to produce the report.  Both the
+        parsing and reporting are difficult to extend, as it is a
+        specialised tool.  It does not attempt to correlate log lines by
+        queueid or pid, nor does it need to deal with the complications
+        encountered during this project.  Pflogsumm produces a useful
+        report, and successfully dealt with all the Postfix logs tested
+        with.
+
+        \url{http://jimsun.linxnet.com/postfix_contrib.html}
         \newline (Last checked 2007/08/13.)
@@
         purpose commercial product which parses 687 log file formats
-        (correct as of 2007/04/15) and produces reports.  I have not
-        experimented with it due to its limited data extraction facilities;
-        it does have three different sets of data for Postfix (as of
-        2007/04/15, one is beta), but they do not appear to be interlinked,
-        nor does it save sufficient data for the purposes of this project.
+        (correct as of 2007/04/15) and produces reports.  Its data
+        extraction facilities are quite limited, though it does extract
+        three different sets of data for Postfix (as of 2007/04/15, one is
+        beta), but they do not appear to be interlinked, nor does it save
+        sufficient data for the purposes of this project.  Not attempt is
+        made to correlate log lines or deal with the difficulties
+        documented in sections~\ref{complications} and~\ref{additional
+        complications}.\footnote{If any attempt is made there is no
+        reference to it in the documentation available on the website.}
         The source code is available in an obfuscated form only (presumably
         for a fee), and the product is quite expensive, as it requires a
-        license for each report which is to be generated. \newline
+        license for each report which is to be generated.  The web
+        interface allows creation of dynamic reports based on any field,
+        but due to the associated cost the author has not experimented with
+        it.
+
         \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline
+        Fields extracted: from, to, server, UID, relay, status, number of
+        recipients, origin hostname, origin IP and virus.  The fields
+        \textit{server}, \textit{uid\/} and \textit{virus\/} are not
+        explained in their documentation: \textit{server\/} is probably the
+        server the mail is delivered to, and \textit{uid\/} might be the
+        uid of the user submitting mail locally.  Postfix does not perform
+        any form of virus checking (though it has many options for
+        cooperating with an external virus scanner), so the
+        \textit{virus\/} field is a mystery.
+
         \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
+        Fields extracted: from, to, RBL list, client hostname and client
+        IP\@.
+
         \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
-        \newline (Last checked 2007/08/13.)
+        \newline
+        Fields extracted: from, to, client hostname, client IP, relay
+        hostname, relay IP, status, response code, RBL list and message id.
+
+        (Last checked 2007/09/26.)
 
@@
 
-        \url{http://www.splunk.com}
+        \url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
 
@@
         statistics with regard to Days Top Domain, Top Users values for per
-        day, per month and years.\/}  \newline
+        day, per month and years.\/}
+
+        Due to the paucity of data gathered by this tool the author has not
+        experimented with it, but based on the documentation it is obvious
+        that it doesn't attempt to gather the breadth of data required for
+        in-depth analysis, and presumably doesn't attempt to correlate log
+        lines per-connection or per-mail.
+
         \url{http://www.enderunix.org/isoqlog/} \newline (Last checked
@@
         AWStats will produce simple graphs for many different services, but
-        that restricts it to supporting the Lowest Common Denominator: the
-        data it will extract from an MTA log file is:
-        \texttt{time2, email, email\_r, host, host\_r, method, url, code,
-        and bytesd.} \newline
+        that restricts it to supporting the Lowest Common Denominator.  The
+        data it will extract from an MTA log file is limited to: \newline
+        \hspace*{2em} time2, email, email\_r, host, host\_r, method, url,
+        code and bytesd.  \newline  There does not seem to be an
+        explanation of those fields in the documentation.  As AWStats is a
+        generic log processor it can safely be assumed that it has
+        no special purpose code to deal with the intricacies of Postfix
+        logs.
+
         \url{http://awstats.sourceforge.net/} \newline
@@
         to provide real time alerts when dictionary attacks, mail loops or
-        similar problems occur.  \newline
+        similar problems occur. It doesn't attempt to extract or save data,
+        correlate log lines, or any of the more advanced tasks described in
+        this document, because it is not designed to do so.
+
         \url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline
@@
         Anteater hasn't been updated since November 2003, and doesn't have
-        any English documentation.
+        any English documentation so it's difficult, nigh impossible, to
+        accurately comment on what analysis is performs.  It did not run
+        successfully when tested, and its parsing would certainly be out of
+        date as Postfix has evolved considerably since this tool was last
+        updated.  As it neither ran successfully nor has documentation the
+        author can read it is difficult to comment on the features it
+        provides.
 

### 2007-10-08 doc/logparser.tex
Minor graph changes

@@
 \usepackage{breakurl}
+\usepackage{lastpage}
 
@@
 integrated into the existing model without breaking the existing sequences.
-detecting a deficiency in the parsing algorithm is also significantly
+Detecting a deficiency in the parsing algorithm is also significantly
 harder than detecting unparsed log lines, as the parser will warn about any
@@
 skewed.  Comparing the number of log lines per program versus rules per
-program in graph~\ref{log22 line and rule distribution by program} shows
-that \texttt{postfix/qmgr} has the highest number of log lines, however it
-has a low number of rules.  This combination results in a low average
-number of rules to try for \texttt{postfix/qmgr} log lines, reducing the
-overall average number of rules to try per log line.  In turn this reduces
-the average parsing time per log line, and thus the parsing time for the
-log file.  While the parsing time is substantially greater than the time
-for the log files on either side, the increase from optimal ordering to
-reversed ordering is proportionally smaller, producing the dips in
-graph~\ref{normal vs shuffled vs reversed ordering factor}.
+program in graphs~\ref{log22 log line and rule distribution by program}
+shows that \texttt{postfix/qmgr} has the highest number of log lines,
+however it has a low number of rules.  This combination results in a low
+average number of rules to try for \texttt{postfix/qmgr} log lines,
+reducing the overall average number of rules to try per log line.  In turn
+this reduces the average parsing time per log line, and thus the parsing
+time for the log file.  While the parsing time is substantially greater
+than the time for the log files on either side, the increase from optimal
+ordering to reversed ordering is proportionally smaller, producing the dips
+in graph~\ref{normal vs shuffled vs reversed ordering factor}.
 
@@
 
-\subsubsection{Log 22 line and rule distribution by program}
-\label{log22 line and rule distribution by program}
+\subsubsection{Log 22 log line and rule distribution by program}
+\label{log22 log line and rule distribution by program}
 
 \noindent
-\includegraphics{build/plot-log22-distribution-versus-rule-count.ps}
-
-\subsubsection{Log 22 lines per rule}
-\label{log22 lines per rule}
-
-As shown in this graph, the most common log lines (see graph~\ref{log22
-line and rule distribution by program}) are from programs which have a low
-number of rules, thus reducing the average parsing time per line, and the
-overall execution time relative to the input size.
+\includegraphics{build/plot-log22-rules.ps}
 
 \noindent
-\includegraphics{build/plot-log22-distribution-versus-rule-count-factor.ps}
+\includegraphics{build/plot-log22-log-lines.ps}
+
+As shown in graphs above, the most common log lines are from programs which
+have a low number of rules, thus reducing the average parsing time per
+line, and the overall execution time relative to the input size.
 

### 2007-10-09 doc/logparser.tex
Added introduction and conclusion to other parsers section. Improved YAALA review.

@@
 
+It is important to compare and contrast newly developed programs,
+algorithms and parsers against those already available, to accurately judge
+what, if any, improvements are delivered by the newcomers.  There are not
+that many previously developed Postfix log parsers, indeed it was quite
+difficult to find ten parsers to review for this project, and the
+functionality offered ranges from quite basic to much more mature,
+depending on the needs of the creator.  The programs are not reviewed from
+an independent viewpoint; the objective is to compare and contrast each
+program with \parsername{}.
+
 \subsection{Parsers reviewed}
@@
 
-    \item [Yet Another Advanced Logfile Analyser] uses a plugin based
-        system to analyse log files and produce output reports.  The core
-        code is merely 91 lines long, as all the parsing and report
-        generation is handled by modules.  Using YAALA as a base would be
-        only slightly less work as both input and output modules would need
-        to be written; it may even be more work to implement the parser
-        within the constraints of YAALA\@.
-
-        \url{http://yaala.org/} \newline (Last checked 2007/08/13.)
+    \item [Yet Another Advanced Logfile Analyser] \textit{yaala is a very
+        flexible analyser for all kinds of logfiles. It uses parsers to
+        extract information from a logfile, an SQL-like query language to
+        relate the information to each other and an output-module to format
+        the information appropriately.\/}
+
+        YAALA uses a plugin based system to analyse log files and produce
+        HTML output reports.  The core code is merely 153 lines long, as
+        all the parsing and report generation is handled by modules.  YAALA
+        supports storing previously gathered data; it dumps its internal
+        data structures using Perl's Storable module~\cite{perl-storable},
+        so with some knowledge of the module responsible for parsing
+        Postfix log files it should be possible to extract data with an
+        external program.  Using YAALA as a base would be only slightly
+        less work as both input and output modules would need to be
+        written; it may even be more work to implement the parser within
+        the constraints of YAALA\@.  YAALA bundles a Postfix parser, which
+        uses hard-coded parsing to extract the following fields from
+        specific log lines:
+
+        Aggregations: count, bytes (the sum of bytes transferred; count is
+        not explained).\newline
+        Keyfields [sic]: date, hour, sender, recipient, defer\_count, delay,
+        incoming\_host, outgoing\_host.
+
+        YAALA provides an SQL like mini-language used when generating the
+        reports; sample reports can be seen at~\cite{yaala-samples}.
+        Example query: \newline \hspace*{2em} \texttt{requests BY file
+        WHERE host =\~{} Google}. \newline The mini-language is quite
+        limited, and cannot be used to extract data, merely to create
+        reports.
+
+        YAALA as distributed does not attempt the in-depth parsing
+        explained in this document, and implementing it within the confines
+        of YAALA would be at least as much work as implementing it
+        separately.
+
+        In summary YAALA provides a Postfix parser which is not extensible
+        unless the user is prepared to change the source
+        code,\footnote{\parsername{} requires source code changes for some
+        new features, but parsing new log lines is accomplished by adding
+        new rules.} and reasonably flexible report generation, but no
+        facilities to extract data for use in other tools.
+
+        \url{http://yaala.org/} \newline (Last checked 2007/10/09.)
 
@@
 \subsection{Conclusion}
+
+While there are other programs available which perform basic Postfix log
+parsing (some to a greater level of intricacy and detail than others), none
+attempt to correlate log lines by pid or queueid to produce an overall
+record of the journey each mail traverses through Postfix.  Most other
+parsers immediately generate a report and discard the data extracted from
+the logs; those which don't discard the data retain it in a format
+inaccessible to other tools.  All of the parsers reviewed can produce a
+report of greater or lesser detail and complexity, a facility not offered
+by \parsername{}; reporting is deferred to a subsequent application.
+
+The overriding difference between \parsername{} and the other parsers
+reviewed herein is that none of them aim to achieve the high level of
+understanding of Postfix logs achieved by \parsername{}.
 

### 2007-10-09 doc/logparser.tex
Improved Lire review

@@
 
-    \item [Logparser/Lire] Lire is another general purpose log parser, but
-        is developed under the GNU GPL\@.  Like other general purpose log
-        parsers it doesn't extract enough data from Postfix logs to be
-        worthwhile.  Development also seems to have stalled, with only a
-        minor update since October 2004.  \newline
+    \item [Logparser/Lire] \textit{As any good system administrator knows,
+        there's a lot more to keep track of in an active network than just
+        webservers. Lire is hands down the most versatile log analysis
+        software available today. Lire not only keeps you informed about
+        your HTTP, FTP, and mail traffic, it also reports on your
+        firewalls, your print servers, and your DNS activity. The ever
+        growing list of Lire-supported services clearly outstrips any other
+        software, in large part thanks to the numerous volunteers who have
+        pioneered many new services and features. Lire is a total solution
+        for your log analysis needs.\/}
+
+        Lire is a general purpose log parser supporting many different
+        types of log file.  Its Postfix parser extracts the following data
+        from Postfix logs: \textit{The email servers' reports will show you
+        the number of deliveries and the volume of email delivered by day,
+        the domains from which you receive or send the most emails, the
+        relays most used, etc.\/}; notably rejections are not mentioned.
+        It supports multiple output formats (text, HTML, PDF and Excel 95).
+        Logs can be imported and the extracted data saved, and reports
+        generated later.  Accessing this data from another application is
+        undocumented; given the source code it should be possible, with
+        time, to access the data.
+
+        Unlike other parsers reviewed, Lire does attempt to correlate log
+        lines by queueid, but not by pid, so e.g.\ the complete list of
+        recipients for a mail should be available.
+
+        Lire does attempt to correlate log lines by queueid, not pid, but
+        its hard coded parsing extracts only part of the available data and
+        makes no attempt to deal with the other complications described in
+        sections~\ref{complications} and~\ref{additional complications}.
+        It stores data, but not in an easily accessible format.  It
+        generates reports in multiple formats, but those reports do not
+        appear to be customisable.
+
         \url{http://logreport.org/lire.html} \newline (Last checked
-        2007/08/13.)
+        2007/10/09.)
 
@@
 
-    \item [Log Mail Analyser]~\cite{log-mail-analyser} has already been
-        discussed in section~\ref{prior art}.
+    \item [Log Mail Analyser] has already been discussed in
+        section~\ref{prior art}.
 

### 2007-10-09 doc/logparser.tex
Explained that rules are filtered according to the program which produced the log line. Mentioned that Postfix is separated into different programs which do one thing and do it well.

@@
 
+Postfix's design follows the Unix philosophy of \textit{Write programs that
+do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
+into various component programs to perform the tasks required of an MTA\@:
+receive mail, send mail, local delivery of mail, etc. --- full details can
+be found at~\cite{postfix-overview}.  Each log line contains the name of
+the Postfix program which produced the log line, and this information is
+used when determining which rules should be used to parse each log line
+(see section~\ref{rule characteristics} for details.)
+
 \subsubsection{Mixing and matching Postfix restrictions}
@@
 
+\label{rule characteristics}
+
 The parser is split into two parts: the parsing algorithm and the rules
@@
 \begin{itemize}
+
+    \item Rules are annotated with the name of a Postfix program, and will
+        only be used when parsing log lines produced by that
+        program.\footnote{There are also generic rules which are used when
+        parsing log lines produced by any Postfix program, but only if
+        there are also rules specific to that program, and those rules must
+        already have been tried and failed.}  Any given rule will only be
+        used to parse a subset of the log lines, and any given log line
+        will only be parsed by a subset of the rules.
 

### 2007-10-13 doc/logparser.tex
Improved AWStats review. Improved Logrep review. Improved conclusion of "review of other parsers" section.

@@
 
-XXX THIS NEEDS TO BE EXPANDED\@; EXPLAIN WHY THEY'RE NOT GOOD ENOUGH
-
 \label{other-parsers}
@@
         that generates advanced web, streaming, ftp or mail server
-        statistics, graphically.\/}
+        statistics, graphically. This log analyzer works as a CGI or from
+        command line and shows you all possible information your log
+        contains, in few graphical web pages. It uses a partial information
+        file to be able to process large log files, often and quickly. It
+        can analyze log files from all major server tools like Apache log
+        files (NCSA combined/XLF/ELF log format or common/CLF log format),
+        WebStar, IIS (W3C log format) and a lot of other web, proxy, wap,
+        streaming servers, mail servers and some ftp servers.\/}
 
@@
         code and bytesd.  \newline  There does not seem to be an
-        explanation of those fields in the documentation.  As AWStats is a
-        generic log processor it can safely be assumed that it has
-        no special purpose code to deal with the intricacies of Postfix
-        logs.
+        explanation of those fields in the documentation.  AWStats has no
+        special purpose code to deal with the intricacies of Postfix logs,
+        in fact it operates by coercing Postfix logs into
+        Apache\footnote{The Apache web server is the most popular HTTP
+        server in use for the past 10 years; more information is available
+        at \url{http://httpd.apache.org/}} format log files, for analysis
+        by its HTTP log file parser.  The converting parser is hard coded,
+        hard to extend, and only deals with a small portion of the log
+        lines generates by Postfix, but it does correlate log lines by
+        queueid.
 
@@
         \url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-        \newline (Last checked 2007/08/13.)
+        \newline (Last checked 2007/10/13.)
 
@@
         the collection, extraction, and presentation of information from
-        various log files.\/}
-
-        The parsing Logrep performs is extremely basic, counting the number
-        of different processes executed, from address, to addresses, number
-        of recipients, bytes transferred and delay.  \newline
+        various log files. It features HTML reports, multi dimensional
+        analysis, overview pages, SSH communication, and graphs, and
+        supports over 30 popular systems including Snort, Squid, Postfix,
+        Apache, Sendmail, syslog, ipchains, iptables, NT event logs,
+        Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
+
+        Logrep extracts the following fields from Postfix logs:
+
+        \begin{itemize}
+
+            \item For mail sent and received: from address, size, and
+                time and date.
+
+            \item For mail sent: to addresses, SMTP code, and delay.
+
+            \item For mail received: the hostname of the sender.
+
+        \end{itemize}
+
+        Fields are correlated based on the queueid (referred to as
+        sessionname within Logrep), but not by pid.  The parsing is hard
+        coded, hard to extend, and error prone: empty fields are saved when
+        the log line doesn't match the regex, though it appears that they
+        will not overwrite existing data.  It also counts the number of log
+        lines parsed and skipped.  Most notably rejections are completely
+        ignored, making it unsuitable for the purposes of this project.
+        Needless to say it doesn't attempt to address any of the
+        complications in sections~\ref{complications} and~\ref{additional
+        complications} except for correlating by queueid.
+
+        Logrep does not come with any documentation, though some scanty
+        documentation is available on its website.  It requires a web
+        browser to interact with it, ensuring that automated log processing
+        will be difficult.  Sadly all attempts to use Logrep failed, as it
+        was unable to access the log files selected; this appears to be a
+        bug rather than operator error.  As parsing failed it wasn't
+        possible to review the reports Logrep can generate (available in
+        HTML only), nor to examine the (undocumented) format in which it
+        can save extracted data for subsequent reuse.
+
+        Logrep extracts far less data from Postfix logs than \parsername{},
+        completely ignores rejections, is effectively undocumented, doesn't
+        deal with the more complicated aspects of Postfix logs, cannot be
+        easily extended by the user, and at the time of writing doesn't
+        work properly.
+
         \url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-        2007/08/13.)
-
-    \item [Log Mail Analyser] has already been discussed in
+        2007/10/13.)
+
+    \item [Log Mail Analyser] Please see the previous discussion in
         section~\ref{prior art}.
@@
 While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of intricacy and detail than others), none
-attempt to correlate log lines by pid or queueid to produce an overall
-record of the journey each mail traverses through Postfix.  Most other
-parsers immediately generate a report and discard the data extracted from
-the logs; those which don't discard the data retain it in a format
-inaccessible to other tools.  All of the parsers reviewed can produce a
-report of greater or lesser detail and complexity, a facility not offered
-by \parsername{}; reporting is deferred to a subsequent application.
+parsing (some to a greater level of detail than others), few
+attempt to correlate log lines queueid (none correlate by pid) to produce
+an overall record of the journey each mail traverses through Postfix.  No
+other parser collects the breadth of information gathered by \parsername{},
+nor makes it as easy to add new parsing rules.  Most other parsers
+immediately generate a report and discard the data extracted from the logs;
+those which don't discard the data retain it in a format inaccessible to
+other tools.  All of the parsers reviewed can produce a report of greater
+or lesser detail and complexity, a facility not offered by \parsername{};
+reporting is deferred to a subsequent application.
 

### 2007-10-14 stats/extract-stats.pl
Plot various stats from the output of the test runs.

@@ -1,0 +1,105 @@
+#!/usr/bin/env perl
+
+# $Id$
+
+use strict;
+use warnings;
+
+use IO::File;
+use IO::Dir;
+use File::Spec::Functions qw(catfile);
+use File::Slurp;
+use Regexp::Common;
+use List::Util qw(sum min max);
+
+sub usage {
+    warn qq{Usage: $0 <results directory> <output directory>};
+    warn qq{Process <results directory>/timing.*, creating graphs in <output directory>};
+    exit 1;
+}
+
+if (@ARGV != 2) {
+    usage();
+}
+
+my ($results_dir, $output_dir) = @ARGV;
+my $plot_file = catfile($output_dir, q{plot.gpi});
+my $plot_file_fh = IO::File->new(q{> } . $plot_file)
+    or die qq{$0: failed opening $plot_file: $!\n};
+
+print $plot_file_fh <<"HEADER";
+set xlabel "Log file"
+set grid
+# Don't continue the scales to the next round number; stop at the topmost point.
+set autoscale xfixmax
+# Put the legend at the top left, Left justified text.
+set key top left Left reverse
+
+
+HEADER
+
+my $dir_dh = IO::Dir->new($results_dir)
+    or die qq{$0: Failed opening $results_dir: $!\n};
+my @files = grep /^timing\./, $dir_dh->read();
+my %files_split = map { m/(.*)\.(\d+)$/;
+    ( $_ => { prefix => $1, number => $2 } ); }
+        @files;
+@files = sort { $files_split{$a}->{prefix} cmp $files_split{$b}->{prefix}
+        || $files_split{$a}->{number} <=> $files_split{$b}->{number} } @files;
+@files = map { catfile($results_dir, $_); } @files;
+use Data::Dumper;
+#print qq{files: }, Dumper(\@files);
+#print qq{files_split: }, Dumper(\%files_split);
+
+my %data;
+foreach my $file (@files) {
+    my @lines = read_file($file);
+    foreach my $line (@lines) {
+        chomp $line;
+        my ($field, $filename, $time) = split /:/, $line;
+        $data{$field} ||= [];
+        push @{$data{$field}}, $time;
+    }
+}
+#print qq{data: }, Dumper(\%data);
+
+my $plot_number = 1;
+foreach my $field (sort keys %data) {
+    my $values = $data{$field};
+    my $output_file = catfile($output_dir, $field);
+    my $output_file_fh = IO::File->new(q{> } . $output_file)
[…]

### 2007-10-16 doc/logparser.tex
Greatly improved the explanation of the dips in the normal/shuffle/reverse factor graphs.  I'm quite pleased with it, though the paragraph pulling it all together might need work.

@@
 
-The resulting execution times are plotted in graph~\ref{normal vs shuffled
-vs reversed ordering}, where the difference is not readily apparent, but
-graphs~\ref{normal vs shuffled vs reversed ordering factor} show the
+Graphs~\ref{normal vs shuffled vs reversed ordering factor} show the
 percentage increase of execution times (mean and standard deviation are
@@
 
-The Y axis represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
 \noindent
@@
 
+The Y axis represents the following:
+
+\begin{enumerate}
+
+    \item The time required, in seconds, to parse the log file.
+
+    \item The size of the log file, in megabytes.
+
+    \item The number of lines in the log file, divided by 10000.
+
+\end{enumerate}
+
 \subsubsection{Ratio of file size and number of lines to execution time}
@@
 \label{rule ordering graphs}
-
-\subsubsection{Normal vs shuffled vs reversed ordering}
-\label{normal vs shuffled vs reversed ordering}
-\includegraphics{build/plot-normal-shuffle-reverse.ps}
 
@@
 
+\subsubsection{Why are there dips in the graphs?}
+\label{Why are there dips in the graphs?}
+
 The dips at logfiles 22 and 62--68 correspond to peaks in log file size in
-graph~\ref{execution time vs file size vs number lines}.  The explanation
-for this took some time to arrive at, but it turns out to be reasonably
-simple.  Examining the database after parsing logfile~22 in isolation shows
-64 rules matched during parsing, but as shown in graph~\ref{log22
-distribution by rule} the distribution of matching rules is extremely
-skewed.  Comparing the number of log lines per program versus rules per
-program in graphs~\ref{log22 log line and rule distribution by program}
-shows that \texttt{postfix/qmgr} has the highest number of log lines,
-however it has a low number of rules.  This combination results in a low
-average number of rules to try for \texttt{postfix/qmgr} log lines,
-reducing the overall average number of rules to try per log line.  In turn
-this reduces the average parsing time per log line, and thus the parsing
-time for the log file.  While the parsing time is substantially greater
-than the time for the log files on either side, the increase from optimal
-ordering to reversed ordering is proportionally smaller, producing the dips
-in graph~\ref{normal vs shuffled vs reversed ordering factor}.
-
-This non-intuitive effect will not be exhibited by log files in general: in
-a normally functioning mail system the impact of rule ordering will scale
-linearly with log file size.  The skewed distribution of log lines in the
-specific log files creates the anomaly described herein.
+graph~\ref{execution time vs file size vs number lines}, and peaks in
+graph~\ref{execution time vs file size vs number lines factor} (where a
+peak means that more lines are processed per second, i.e.\ performance is
+better).  The explanation for this took some time to arrive at, but it
+turns out to be reasonably simple.  The large log files in question were
+caused by a mail forwarding loop, where the distribution of log lines is
+quite different to normal, resulting in different performance
+characteristics.
+
+The mail loop was set up by a user modifying his mail forwarding to:
+\newline \hspace*{2em}\texttt{$\backslash$username, username@domain}
+\newline This instructs Postfix to deliver the mail to the local user, and
+also forward to the remote address; this is generally not a problem except
+that the remote address in this case is the user's address, creating an
+infinite loop.  To prevent this happening Postfix examines the Delivered-To
+header in the mail and if the mail has already been delivered to the
+current address it is bounced back to the sender with the error message
+\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
+well, but unfortunately the user noticed they had not received any mail in
+a while, and opted to send a test mail to themselves, causing a loop not
+caught by Postfix as described below.
+
+\begin{enumerate}
+
+    \item Postfix accepts a mail from username@domain, for
+        username@domain.
+
+    \item Postfix delivers the mail to the local mailbox and
+        username@domain, as instructed by the user's forwarding
+        instructions. The forwarded mail has a
+        \texttt{Delivered-To:~username@domain} header added, and the
+        envelope sender address is username@domain.  Log lines are added by
+        \textit{postfix/local}, \textit{postfix/qmgr\/} (twice),
+        \textit{postfix/cleanup\/} and finally \textit{postfix/pickup}.
+
+    \item Postfix accepts the mail for username@domain, but while
+        delivering it notices that the \texttt{Delivered-To} header already
+        contains the address it's currently delivering to, and therefore
+        sends a bounce notification with sender address \textit{$<>$\/} to
+        the original sender: username@domain.  Log lines are added by
+        \textit{postfix/local}, \textit{postfix/qmgr\/} (twice) and
+        \textit{postfix/cleanup}.
+
+    \item Postfix delivers the bounce notification to both the local
+        mailbox and to username@domain, as instructed by the user's
+        forwarding instructions.  A \texttt{Delivered-To:~username@domain}
+        header is added to the forwarded bounce notification, and the
+        envelope sender address is now username@domain.  Log lines are
+        added by \textit{postfix/local},
+        \textit{postfix/qmgr\/} (twice), \textit{postfix/cleanup}, and finally
+        \textit{postfix/pickup}.
+
+    \item Postfix accepts the forwarded bounce notification but while
+        delivering it notices that the \texttt{Delivered-To} header already
+        contains the address currently being delivered to, and sends a
+        bounce notification to the sender: username@domain.  Log lines are
+        added by \textit{postfix/local}, \textit{postfix/qmgr\/} (twice)
+        and \textit{postfix/cleanup}.
+
+    \item GOTO 2.
+
+\end{enumerate}
+
+The sequence described above occurs extremely rapidly, as Postfix does not
+have to deliver the mail to an external system, so mails are delivered,
+bounced and generated as fast as the disks can keep up, resulting in a huge
+volume of logs.
+
+The vast majority of log lines when a mail loop occurs are from Postfix
+components which have a small number of rules associated with them, whereas
+in general the most common log lines are logged by \textit{postfix/smtpd},
+which also has the highest number of rules.  \textit{postfix/smtpd\/} log
+lines are distributed across rules much more evenly that
+\textit{postfix/qmgr}, \textit{postfix/local}, \textit{postfix/cleanup\/}
+or \textit{postfix/pickup}, so the average number of rules required to
+parse a \textit{postfix/smtpd\/} log line is much higher that the average
+number required to parse other log lines.  Thus the average number of rules
+required to parse a random line from logs containing a mail loop is much
+lower than the average number required to parse a random log line from logs
+without a mail loop, as can be seen in the peaks in graphs~\ref{execution
+time vs file size vs number lines factor} and~\ref{normal vs shuffled vs
+reversed ordering factor}.  The difference is more extreme in the reversed
+ordering case because the largest number of rules to be consulted for the
+commonly occurring log lines during a mail loop is far less than the
+largest number of rules that are consulted for commonly occurring lines
+outside of a mail loop.  The number of rules which needs to be consulted
+when the ordering is shuffled varies between the optimum and nadir, and
+the performance varies proportionally.
+
+The difference between logs with a mail loop and logs without can be seen
+in this table showing the increases for the different rule orderings and
+combinations of log files:
 
 \noindent\input{build/stats-normal-shuffle-reverse-include.tex}
-
-\subsubsection{Log 22 distribution}
-\label{log22 distribution by rule}
-\includegraphics{build/plot-log22-distribution-by-rule.ps}
-
-The individual rules are not labelled on the x-axis because doing so makes
-the graph illegible.
-
-\subsubsection{Log 22 log line and rule distribution by program}
-\label{log22 log line and rule distribution by program}
-
-\noindent
-\includegraphics{build/plot-log22-rules.ps}
-
-\noindent
-\includegraphics{build/plot-log22-log-lines.ps}
-
-As shown in graphs above, the most common log lines are from programs which
-have a low number of rules, thus reducing the average parsing time per
-line, and the overall execution time relative to the input size.
-
-\subsubsection{Rule hits}
-\label{rule hits}
-
-\noindent\includegraphics{build/plot-hits.ps}
 
@@
 
+\subsection{Rule hits}
+\label{rule hits}
+
+\noindent\includegraphics{build/plot-hits.ps}
+
 \subsection{Conclusion}

### 2007-10-16 doc/logparser.tex
Improved the cached/discarded section, referring to the normal/shuffle/reverse section.
Moved the rule hits graph to a subsection of its own, and added three more
graphs giving an improved breakdown of the rule hits, as in the overall graph
it's difficult to distinguish what's happening in the lower end.

Extended the introduction and conclusion with the new material.

@@
 performance scales linearly with input size.  The impact of rule ordering
-is shown in section~\ref{rule ordering graphs}, and the lower impact of
-rule ordering on larger log files is explained.  The final group of graphs
+is shown in section~\ref{rule ordering graphs}, and the anomalous dips and
+peaks apparent in some graphs are explained.  The next group of graphs
 vividly shows the huge impact that caching compiled regexs has on parser
-performance.
+performance.  This section concludes with a breakdown of the rule hits
+accumulated during a single test run of the parser.
 
@@
 
-The Y axis represents the following:
+The Y axis in the graph above represents the following:
 
@@
 
+The following graphs show the impact that not caching compiled regexs has
+on parser performance: on typical log files the execution time when not
+caching compiled regexs is 500--600\% of the execution time when caching;
+reversing the perspective shows that cached execution time is merely
+17--20\% of non-cached execution time.  Caching compiled regexs is probably
+the single most effective optimisation possible in the parser's
+implementation.
+
 \subsubsection{Regex: cached vs discarded}
@@
 size in graph~\ref{execution time vs file size vs number lines}.  The
-anomalies are caused by a combination of a mail loop and a delivery failure
-which resulted in a huge number of log lines, all of which were matched by
-a single rule.  Sorting the rules by the number of successful matches has
-the effect that this rule will be tried first, so the average number of
-regexs compiled per line drops sharply for those eight log files, resulting
-in a much lower increase in relative execution time, as detailed below.
+reason for the anomalous log files has already been explained in
+section~\ref{Why are there dips in the graphs?}.
+
+The distribution of log lines across rules is much different and the
+average number of rules consulted per log line is much lower when there is
+a mail loop; this results in far fewer regex compilations per line than
+when there isn't a mail loop, and a correspondingly decreased execution
+time.  The increases in execution time when not caching regexs for
+different combinations of log files are summarised in the table below:
 
@@
 
+The number of hits per rule is quite unevenly spread, resembling a Power
+Law distribution~\cite{powerlaw}.
+
 \noindent\includegraphics{build/plot-hits.ps}
+
+As the graph above is quite difficult to read it has been separated into
+three sections below: low, middle and high.
+
+\noindent\includegraphics{build/plot-hits-low.ps}
+
+It is apparent from the low graph that some rules have few or no hits;
+those with zero hits are rules which were written to parse log files used
+during development of the parser but not utilised in the test runs
+performed for this document.
+
+\noindent\includegraphics{build/plot-hits-middle.ps}
+
+\noindent\includegraphics{build/plot-hits-high.ps}
 
@@
 of specific inputs.  The necessity of caching compiled regexs is attested
-to by the final group of graphs, where the difference between caching and
-discarding is staggering.
+to by the third group of graphs, where the difference between caching and
+discarding is staggering.  This section concluded with a breakdown of the
+rule hits accumulated during a single test run of the parser.
 

### 2007-10-18 doc/logparser.tex
I'm now using the figure environment for graphs, so they flow better in the paper.  The paper needs proof reading now, as there may be old and broken references.

@@
 \usepackage{lastpage}
+
+% \showgraph{filename}{caption}{label}
+\newcommand{\showgraph}[3]{
+    \begin{figure}[hbt!]
+        \caption{#2}\label{#3}
+        \includegraphics{#1}
+    \end{figure}
+}
 
@@
 size: does it scale linearly, polynomially, or exponentially?
-Graph~\ref{execution time vs file size vs number lines} shows the execution
-time in seconds, file size in MB and tens of thousands of lines per log
-file per log file.  All three lines run roughly in parallel, giving a
-visual impression that the algorithm scales linearly with input size.  This
-impression is borne out by graph~\ref{execution time vs file size vs number
-lines factor}, which plots the ratio of file size vs execution time and
-number of lines vs execution time.  As the reader can see, the ratios are
-quite tightly banded, showing that the algorithm scales linearly: the much
-larger log files between points 60 and 70 on the X axis in
-graph~\ref{execution time vs file size vs number lines} don't cause any
-abnormality in the corresponding points in graph~\ref{execution time vs
-file size vs number lines factor}.
+Graph~\ref{execution time vs file size vs number of lines graph} shows the
+execution time in seconds, file size in MB and tens of thousands of lines
+per log file per log file.  All three lines run roughly in parallel, giving
+a visual impression that the algorithm scales linearly with input size.
+This impression is borne out by graph~\ref{execution time vs file size vs
+number lines factor}, which plots the ratio of file size vs execution time
+and number of lines vs execution time.  As the reader can see, the ratios
+are quite tightly banded, showing that the algorithm scales linearly: the
+much larger log files between points 60 and 70 on the X axis in
+graph~\ref{execution time vs file size vs number of lines graph} don't
+cause any abnormality in the corresponding points in graph~\ref{execution
+time vs file size vs number lines factor}.
 
@@
 
-Graphs~\ref{normal vs shuffled vs reversed ordering factor} show the
-percentage increase of execution times (mean and standard deviation are
-shown after the graph).  Overall this provides a modest but worthwhile
-performance increase.
+Graphs~\ref{percentage increase of shuffled over normal}
+and~\ref{percentage increase of reversed over normal} show the percentage
+increase of execution times (mean and standard deviation are shown after
+the graph).  Overall this provides a modest but worthwhile performance
+increase.
 
@@
 without caching the regex.  Caching the compiled regexs is obviously far
-more efficient; graph~\ref{normal regex vs disarded regex factor} shows the
-percentage execution time increase when not caching each regex.
+more efficient; graph~\ref{normal regex vs discarded regex factor} shows
+the percentage execution time increase when not caching each regex.
 
@@
 
+\renewcommand{\figurename}{Graph}
+
 \subsection{Introduction}
@@
 
-\subsubsection{Execution time vs file size vs number of lines.}
-\label{execution time vs file size vs number lines}
-
-\noindent
-\includegraphics{build/plot-normal-filesize-numlines.ps}
+\showgraph{build/plot-normal-filesize-numlines.ps}{Execution time vs file
+size vs number of lines}{execution time vs file size vs number of lines graph}
 
@@
 
-\subsubsection{Ratio of file size and number of lines to execution time}
-\label{execution time vs file size vs number lines factor}
-
-\noindent
-\includegraphics{build/plot-normal-filesize-numlines-factor.ps}
+\showgraph{build/plot-normal-filesize-numlines-factor.ps}{Ratio of file
+size and number of lines to execution time}{execution time vs file size vs
+number lines factor}
 
@@
 
+\clearpage
+
 \subsection{Rule ordering}
@@
 
-\subsubsection{Percentage increase of shuffled + reversed over normal}
-\label{normal vs shuffled vs reversed ordering factor}
-\includegraphics{build/plot-normal-shuffle-factor.ps}
-\newline
-\includegraphics{build/plot-normal-reverse-factor.ps}
+\showgraph{build/plot-normal-shuffle-factor.ps}{Percentage increase of
+shuffled over normal}{percentage increase of shuffled over normal}
+
+\showgraph{build/plot-normal-reverse-factor.ps}{Percentage increase of
+reversed over normal}{percentage increase of reversed over normal}
 
@@
 The dips at logfiles 22 and 62--68 correspond to peaks in log file size in
-graph~\ref{execution time vs file size vs number lines}, and peaks in
+graph~\ref{execution time vs file size vs number of lines graph}, and peaks in
 graph~\ref{execution time vs file size vs number lines factor} (where a
@@
 without a mail loop, as can be seen in the peaks in graphs~\ref{execution
-time vs file size vs number lines factor} and~\ref{normal vs shuffled vs
-reversed ordering factor}.  The difference is more extreme in the reversed
-ordering case because the largest number of rules to be consulted for the
-commonly occurring log lines during a mail loop is far less than the
-largest number of rules that are consulted for commonly occurring lines
-outside of a mail loop.  The number of rules which needs to be consulted
-when the ordering is shuffled varies between the optimum and nadir, and
-the performance varies proportionally.
+time vs file size vs number lines factor},~\ref{percentage increase of
+shuffled over normal} and~\ref{percentage increase of reversed over
+normal}.  The difference is more extreme in the reversed ordering case
+because the largest number of rules to be consulted for the commonly
+occurring log lines during a mail loop is far less than the largest number
+of rules that are consulted for commonly occurring lines outside of a mail
+loop.  The number of rules which needs to be consulted when the ordering is
+shuffled varies between the optimum and nadir, and the performance varies
+proportionally.
 
@@
 
-\subsubsection{Regex: cached vs discarded}
-\label{normal regex vs discard regex}
-\includegraphics{build/plot-cached-discarded.ps}
-
-\subsubsection{Regex: percentage execution time increase}
-\label{normal regex vs disarded regex factor}
-\includegraphics{build/plot-cached-discarded-factor.ps}
-
-Two large dips can be seen in graph~\ref{normal regex vs disarded regex
+\showgraph{build/plot-cached-discarded.ps}{Regex: cached vs
+discarded}{normal regex vs discard regex}
+
+\showgraph{build/plot-cached-discarded-factor.ps}{Regex: percentage
+execution time increase}{normal regex vs discarded regex factor}
+
+Two large dips can be seen in graph~\ref{normal regex vs discarded regex
 factor} at log files 22 and 62--68, corresponding to the spikes in log file
-size in graph~\ref{execution time vs file size vs number lines}.  The
-reason for the anomalous log files has already been explained in
-section~\ref{Why are there dips in the graphs?}.
+size in graphs~\ref{execution time vs file size vs number of lines graph}
+and~\ref{normal regex vs discard regex}.  The reason for the anomalous log
+files has already been explained in section~\ref{Why are there dips in the
+graphs?}.
 
@@
 
-\noindent\includegraphics{build/plot-hits.ps}
-
-As the graph above is quite difficult to read it has been separated into
-three sections below: low, middle and high.
-
-\noindent\includegraphics{build/plot-hits-low.ps}
-
-It is apparent from the low graph that some rules have few or no hits;
-those with zero hits are rules which were written to parse log files used
-during development of the parser but not utilised in the test runs
-performed for this document.
-
-\noindent\includegraphics{build/plot-hits-middle.ps}
-
-\noindent\includegraphics{build/plot-hits-high.ps}
+\showgraph{build/plot-hits.ps}{Hits per rule}{rule hits graph}
+
+As graph~\ref{rule hits graph} is quite difficult to read it has been
+separated into three sections: low, middle and high.
+
+\showgraph{build/plot-hits-low.ps}{Hits per rule (low)}{hits per rule low}
+
+It is apparent from the low graph (graph~\ref{hits per rule low}) that some
+rules have few or no hits; those with zero hits are rules which were
+written to parse log files used during development of the parser but not
+utilised in the test runs performed for this document.
+
+\showgraph{build/plot-hits-middle.ps}{Hits per rule (middle)}{hits per rule
+middle}
+
+\showgraph{build/plot-hits-high.ps}{Hits per rule (high)}{hits per rule
+high}
+
+\clearpage
 

### 2007-10-18 doc/logparser.tex
Changed the majority of \ref commands to my \refwithpage, which includes the page number automatically.  I like having the page number, but not how it interrupts the flow, so I think I'll have to work on the appearance.

@@
     \end{figure}
+}
+
+% Replacement for \ref{}, adds the page number too.
+\newcommand{\refwithpage}[1]{%
+    \empty{}\ref{#1} (page~\pageref{#1})%
 }
@@
 
-Section~\ref{database schema} partially describes the database schema used by
-the parser, explaining in detail the tables used for storing the resulting
-data.  The remainder of the schema is laid out in section~\ref{rule
-attributes}.
+Section~\ref{database schema} partially describes the database schema used
+by the parser, explaining in detail the tables used for storing the
+resulting data.  The remainder of the schema is laid out in
+section~\ref{rule attributes}.
 
@@
 parse Postfix logs are summarised (full details are available in
-Section~\ref{other-parsers}), plus a review of previously published
+section~\refwithpage{other-parsers}), plus a review of previously published
 research in this area.
@@
 measures against, or to produce statistics showing how effective those
-measures are.\footnote{Section~\ref{introduction} lists some more example
-uses.}
+measures are.\footnote{Section~\refwithpage{introduction} lists some more
+example uses.}
 
@@
 used when determining which rules should be used to parse each log line
-(see section~\ref{rule characteristics} for details.)
+(see section~\refwithpage{rule characteristics} for details.)
 
@@
 
-10 other parsers have been reviewed in Appendix~\ref{other-parsers} as part
-of the background research for this project.  None of the reviewed parsers
-perform the type of advanced parsing and log correlation described here;
-all are intended to perform a specific parsing and reporting task, rather
-than be a generic parser, extracting data and leaving generation of reports
-from the data to other programs.  Some parsers save the data extracted to a
-database but the majority discard all data once they have finished running,
-making historical analysis impossible.  The other parsers reviewed all
-produce a report of greater or lesser complexity and detail, whereas the
-program described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
+10 other parsers have been reviewed in Appendix~\refwithpage{other-parsers}
+as part of the background research for this project.  None of the reviewed
+parsers perform the type of advanced parsing and log correlation described
+here; all are intended to perform a specific parsing and reporting task,
+rather than be a generic parser, extracting data and leaving generation of
+reports from the data to other programs.  Some parsers save the data
+extracted to a database but the majority discard all data once they have
+finished running, making historical analysis impossible.  The other parsers
+reviewed all produce a report of greater or lesser complexity and detail,
+whereas the program described here doesn't attempt to produce a report at
+all; that responsibility is deferred to a separate program, to be developed
+later.  The parsing algorithm and program described here are designed to
+enable much more detailed log analysis by providing a stable platform for
 subsequent programs to develop upon.
@@
 
-\parsername{} is bundled with extensive documentation, including an in-depth
-explanation of the database schema.  It supports multiple values for every
-field for which multiple values are reasonable (e.g.\ the sender address
-should have one value only, whereas the recipient address can have
+\parsername{} is bundled with extensive documentation, including an
+in-depth explanation of the database schema.  It supports multiple values
+for every field for which multiple values are reasonable (e.g.\ the sender
+address should have one value only, whereas the recipient address can have
 multiple values).  The default database it uses is SQLite3, which supports
@@
 with any of the complications discussed in
-sections~\ref{complications}~and~\ref{additional complications}.  The
-program fails to parse Postfix 2.2.x or 2.3.x log files; indeed it produces
-a number of warnings and no usable output.  Parsing in \parsername{} is
-divided in two (see sections~\ref{rules} and~\ref{parsing-algorithm} for a
-more detailed explanation): parsing a new log line is a trivial matter of
-adding a new rule, but code changes may be required if new complications
-(sections~\ref{complications} and~\ref{additional complications}) must be
-dealt with or a new action is required (section~\ref{actions-in-detail}).
+sections~\refwithpage{complications} and~\refwithpage{additional
+complications}.  The program fails to parse Postfix 2.2.x or 2.3.x log
+files; indeed it produces a number of warnings and no usable output.
+Parsing in \parsername{} is divided in two (see
+sections~\refwithpage{rules} and~\refwithpage{parsing-algorithm} for a more
+detailed explanation): parsing a new log line is a trivial matter of adding
+a new rule, but code changes may be required if new complications
+(sections~\refwithpage{complications} and~\refwithpage{additional
+complications}) must be dealt with or a new action is required
+(section~\refwithpage{actions-in-detail}).
 
@@
 aid the user in understanding what each rule parses.  The rules are
-described in detail in section~\ref{rules} but the fields are covered in
-section~\ref{rule attributes}.
+described in detail in section~\refwithpage{rules} but the fields are
+covered in section~\refwithpage{rule attributes}.
 
@@
 rejection; the individual fields will be described in the forthcoming
-section~\ref{connections table}.  The results table has one or more rows
-per row in the connections table, depending on the specifics of the
+section~\refwithpage{connections table}.  The results table has one or more
+rows per row in the connections table, depending on the specifics of the
 mail/connection; the fields will be covered in detail in
-section~\ref{results table}.
+section~\refwithpage{results table}.
 
@@
 
-Rules are discussed in detain in section~\ref{rules}, but the rules table
-is covered here.  Rules are created by the user, not the parser, and will
-not be modified by the parser (except for the hits and hits\_total fields).
-Rules parse the individual lines, telling the parser which fields to
-extract and what action to take for that line.
+Rules are discussed in detain in section~\refwithpage{rules}, but the rules
+table is covered here.  Rules are created by the user, not the parser, and
+will not be modified by the parser (except for the hits and hits\_total
+fields).  Rules parse the individual lines, telling the parser which fields
+to extract and what action to take for that line.
 
@@
         large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in section~\ref{rule efficiency}.
+        discussed in section~\refwithpage{rule efficiency}.
 
@@
         log line will be extracted.  The format is: \newline
-        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline
-        i.e.\ semi-colon or comma separated assignment statements, with the
+        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline i.e.\
+        semi-colon or comma separated assignment statements, with the
         variable name on the left and the matching field from the regex on
@@
     \item [action] The action the algorithm will take; a full list can be
-        found in Section~\ref{actions-in-detail}.
+        found in Section~\refwithpage{actions-in-detail}.
 
@@
         time.  Rule ordering for efficiency is discussed in
-        section~\ref{rule ordering for efficiency}.
+        section~\refwithpage{rule ordering for efficiency}.
 
@@
         more specific rules to take precedence over more general rules
-        (described in section~\ref{overlapping rules}).
+        (described in section~\refwithpage{overlapping rules}).
 
@@
 details of the table containing the rules have already been described in
-section~\ref{rule attributes},).  This section finishes with a discussion
-of rule efficiency concerns, referring to the graphs in
-section~\ref{graphs}.
+section~\refwithpage{rule attributes},).  This section finishes with a
+discussion of rule efficiency concerns, referring to the graphs in
+section~\refwithpage{graphs}.
 
@@
     \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See section~\ref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, dump a textual representation of the connections
-        and results tables, and delete everything from those tables.
-        Repeat with shuffled and reversed sorting.  If there are no
-        overlapping rules the tables from each run will be identical;
+        reversed rules.\footnote{See section~\refwithpage{rule efficiency}
+        for more details of sorting the rules.}  Parse a number of log
+        files using normal sorting, dump a textual representation of the
+        connections and results tables, and delete everything from those
+        tables.  Repeat with shuffled and reversed sorting.  If there are
+        no overlapping rules the tables from each run will be identical;
         differences indicate overlapping rules, and the rule\_ids which
@@
     \item [action:] will be executed if the regex matches successfully (see
-        section~\ref{actions-in-detail} for full details).
+        section~\refwithpage{actions-in-detail} for full details).
 
@@
     \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see section~\ref{rule ordering for
+        ordering the rules (see section~\refwithpage{rule ordering for
         efficiency}); hits\_total is updated at the end of each run but is
@@
 daily basis).  When generating the data for the graphs included in
-section~\ref{graphs}, 93 log files (totaling 10.08 GB, 60.72 million lines)
-were each parsed 10 times, the first run discarded and the remaining 9 runs
-averaged.  The first run is discarded for two reasons:
+section~\refwithpage{graphs}, 93 log files (totaling 10.08 GB, 60.72
+million lines) were each parsed 10 times, the first run discarded and the
+remaining 9 runs averaged.  The first run is discarded for two reasons:
 
@@
 size: does it scale linearly, polynomially, or exponentially?
-Graph~\ref{execution time vs file size vs number of lines graph} shows the
-execution time in seconds, file size in MB and tens of thousands of lines
-per log file per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\ref{execution time vs file size vs
-number lines factor}, which plots the ratio of file size vs execution time
-and number of lines vs execution time.  As the reader can see, the ratios
-are quite tightly banded, showing that the algorithm scales linearly: the
-much larger log files between points 60 and 70 on the X axis in
-graph~\ref{execution time vs file size vs number of lines graph} don't
-cause any abnormality in the corresponding points in graph~\ref{execution
-time vs file size vs number lines factor}.
+Graph~\refwithpage{execution time vs file size vs number of lines graph}
+shows the execution time in seconds, file size in MB and tens of thousands
+of lines per log file per log file.  All three lines run roughly in
+parallel, giving a visual impression that the algorithm scales linearly
+with input size.  This impression is borne out by
+graph~\refwithpage{execution time vs file size vs number lines factor},
+which plots the ratio of file size vs execution time and number of lines vs
+execution time.  As the reader can see, the ratios are quite tightly
+banded, showing that the algorithm scales linearly: the much larger log
+files between points 60 and 70 on the X axis in
+graph~\refwithpage{execution time vs file size vs number of lines graph}
+don't cause any abnormality in the corresponding points in
+graph~\refwithpage{execution time vs file size vs number lines factor}.
 
@@
 
-Rule ordering was mentioned in section~\ref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are 139 different rules, with the top 10\% matching the vast majority of
-the log lines, and the remaining log lines split across the other 90\% of
-the rules (as shown in graph~\ref{rule hits}).  Assuming that the
-distribution of log lines is reasonably steady over time, program
+Rule ordering was mentioned in section~\refwithpage{rule attributes} and
+will be covered in greater detail in this section.  At the time of writing
+there are 139 different rules, with the top 10\% matching the vast majority
+of the log lines, and the remaining log lines split across the other 90\%
+of the rules (as shown in graph~\refwithpage{rule hits}).  Assuming that
+the distribution of log lines is reasonably steady over time, program
 efficiency should benefit from trying more frequently matching rules before
@@
 
-Graphs~\ref{percentage increase of shuffled over normal}
-and~\ref{percentage increase of reversed over normal} show the percentage
-increase of execution times (mean and standard deviation are shown after
-the graph).  Overall this provides a modest but worthwhile performance
-increase.
+Graphs~\refwithpage{percentage increase of shuffled over normal}
+and~\refwithpage{percentage increase of reversed over normal} show the
+percentage increase of execution times (mean and standard deviation are
+shown after the graph).  Overall this provides a modest but worthwhile
+performance increase.
 
@@
 but dynamic regexs need to be explicitly compiled and cached.
-Graph~\ref{normal regex vs discard regex} shows execution times with and
-without caching the regex.  Caching the compiled regexs is obviously far
-more efficient; graph~\ref{normal regex vs discarded regex factor} shows
-the percentage execution time increase when not caching each regex.
+Graph~\refwithpage{normal regex vs discard regex} shows execution times
+with and without caching the regex.  Caching the compiled regexs is
+obviously far more efficient; graph~\refwithpage{normal regex vs discarded
+regex factor} shows the percentage execution time increase when not caching
+each regex.
 
@@
     \item The database table containing the rules is dealt with in
-        section~\ref{rule attributes}, and is not duplicated in this
-        section.
+        section~\refwithpage{rule attributes}, and is not duplicated in
+        this section.
 
@@
         ordering ordering of rules, with reference to the graphs in
-        section~\ref{graphs}.
+        section~\refwithpage{graphs}.
 
@@
 belongs to.\footnote{Unfortunately this statement is not completely
-accurate: see section~\ref{timeouts-during-data-phase} for details.
+accurate: see section~\refwithpage{timeouts-during-data-phase} for details.
 However in general there isn't any ambiguity about which data structure
@@
 effect of this complication upon the algorithm is discussed in
-section~\ref{tracking re-injected mail}.
+section~\refwithpage{tracking re-injected mail}.
 
@@
 
-Section~\ref{actions-in-detail} explains the actions in substantive detail;
-this section will omit such detail because it would clutter and confuse the
-algorithm description.  The data flow chart in section~\ref{flow-chart}
-should also be consulted while reading this section.  For the sake of
-clarity this description covers the most common paths through the
-algorithm, as including every path would hinder understanding.  The more
-uncommon paths are caused by the complications described in
-section~\ref{additional complications}, and are covered therein.
+Section~\refwithpage{actions-in-detail} explains the actions in substantive
+detail; this section will omit such detail because it would clutter and
+confuse the algorithm description.  The data flow chart in
+section~\refwithpage{flow-chart} should also be consulted while reading
+this section.  For the sake of clarity this description covers the most
+common paths through the algorithm, as including every path would hinder
+understanding.  The more uncommon paths are caused by the complications
+described in section~\refwithpage{additional complications}, and are
+covered therein.
 
@@
     \item If one or more mails were accepted there will be more log entries
-        for those mails later, see section~\ref{mail-delivery}.
+        for those mails later, see section~\refwithpage{mail-delivery}.
 
@@
 Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in section~\ref{complications},
+own;\footnote{Previously discussed in section~\refwithpage{complications},
 complication 3.} re-injection is somewhat awkward to explain because it
 overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to section~\ref{tracking re-injected mail}.
+is deferred to section~\refwithpage{tracking re-injected mail}.
 
@@
 must be detected using heuristics (see
-section~\ref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.
+section~\refwithpage{identifying-bounce-notifications} for details).
+Bounce notifications are the primary example of internally generated mails,
+though there may be other types.
 
@@
         tracked by the parent mail (action: track; flowchart:~11) ---
-        section~\ref{tracking re-injected mail} discusses this in detail.
+        section~\refwithpage{tracking re-injected mail} discusses this in
+        detail.
 
@@
         (flowchart:~15).  Re-injected mails are described in
-        section~\ref{tracking re-injected mail}.
+        section~\refwithpage{tracking re-injected mail}.
 
@@
 to the parent or children of mail re-injected due to forwarding, and that
-is explained in section~\ref{tracking re-injected mail}.
+is explained in section~\refwithpage{tracking re-injected mail}.
 
@@
     \item The indicator which more commonly introduces re-injection is when
-        \texttt{postfix/qmgr}
-        selects a mail with a previously unseen queueid for delivery
-        (action: mail\_picked\_for\_delivery; flowchart:~3), in which case
-        a new data structure will be created.  The mail will be flagged as
-        having unknown origins; this flag should be subsequently cleared
-        once the origin has been established.  This may also be an
-        indicator that the mail is a bounce notification, see
-        section~\ref{identifying-bounce-notifications} for details.
+        \texttt{postfix/qmgr} selects a mail with a previously unseen
+        queueid for delivery (action: mail\_picked\_for\_delivery;
+        flowchart:~3), in which case a new data structure will be created.
+        The mail will be flagged as having unknown origins; this flag
+        should be subsequently cleared once the origin has been
+        established.  This may also be an indicator that the mail is a
+        bounce notification, see
+        section~\refwithpage{identifying-bounce-notifications} for details.
 
@@
         both qmgr and cleanup as it needs to deal with out of order log
-        lines; see section~\ref{discarding cleanup lines} for details.
+        lines; see section~\refwithpage{discarding cleanup lines} for
+        details.
 
@@
         Out of order log entries may have caused the state table entry to
-        already exist (see section~\ref{pickup logging after cleanup});
-        otherwise it is created.  The data from the log line is then saved
-        to the state table entry.
+        already exist (see section~\refwithpage{pickup logging after
+        cleanup}); otherwise it is created.  The data from the log line is
+        then saved to the state table entry.
 
@@
         which case there is not.  See
-        section~\ref{timeouts-during-data-phase} for the gory details.
+        section~\refwithpage{timeouts-during-data-phase} for the gory
+        details.
 
@@
 for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in
-section~\ref{tracking re-injected mail}, but unlike the solution described
-therein bounce notifications do not eventually have a log line which
-identifies their source.  Heuristics must be used to identify bounce
-notifications, and those heuristics are:
+problem of re-injected mails discussed in section~\refwithpage{tracking
+re-injected mail}, but unlike the solution described therein bounce
+notifications do not eventually have a log line which identifies their
+source.  Heuristics must be used to identify bounce notifications, and
+those heuristics are:
 
@@
 available is the client hostname and IP address, but the database schema
-requires many more fields be populated (see section~\ref{database schema}).
+requires many more fields be populated (see section~\refwithpage{database
+schema}).
 
@@
 disconnections are either unimportant or are handled elsewhere in the
-algorithm (section~\ref{timeouts-during-data-phase}).  Sadly Postfix
-doesn't log a message saying the mail has been discarded, as should be
-expected by now.  The checks to identify this happening are:
+algorithm (section~\refwithpage{timeouts-during-data-phase}).  Sadly
+Postfix doesn't log a message saying the mail has been discarded, as should
+be expected by now.  The checks to identify this happening are:
 
@@
 This complication is further complicated by out of order cleanup lines: see
-section~\ref{discarding cleanup lines} for details.
+section~\refwithpage{discarding cleanup lines} for details.
 
@@
 mail may need to be discarded, as detailed in
-section~\ref{timeouts-during-data-phase}.
+section~\refwithpage{timeouts-during-data-phase}.
 
@@
 Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in sections~\ref{tracking
-re-injected mail},~\ref{discarding cleanup lines}~and~\ref{pickup logging
-after cleanup}.  In the 93 log files used for testing this occurs only five
-times in 60,721,709 log lines, but for completeness of the algorithm it
-should be dealt with.  The five occurrences in the test log files have the
-same characteristics: the \texttt{postfix/local} log line showing delivery
-to a local mailbox occurs after the \texttt{postfix/qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \texttt{postfix/local} line is parsed
-and remains in the state tables; four warnings are issued per pair of out
-of order log lines.
+dealt with by the techniques described in sections~\refwithpage{tracking
+re-injected mail},~\refwithpage{discarding cleanup
+lines}~and~\refwithpage{pickup logging after cleanup}.  In the 93 log files
+used for testing this occurs only five times in 60,721,709 log lines, but
+for completeness of the algorithm it should be dealt with.  The five
+occurrences in the test log files have the same characteristics: the
+\texttt{postfix/local} log line showing delivery to a local mailbox occurs
+after the \texttt{postfix/qmgr} log line showing removal of the mail from
+the queue because delivery is completed.  This causes problems: the mail is
+not complete, so entry into the database fails; a new mail is created when
+the \texttt{postfix/local} line is parsed and remains in the state tables;
+four warnings are issued per pair of out of order log lines.
 
@@
 The aborted delivery attempts described in
-section~\ref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the 93
-log files used for testing.  The symptoms are the same as in
-section~\ref{aborted-delivery-attempts}, except that there \textit{is\/} a
-\texttt{postfix/cleanup} log line; there does not appear to be anything in
-the log file to explain why there are no further log messages.  The only
-way to detect these mails is to periodically scan all mails in the state
-tables, deleting any mails displaying the following characteristics:
+section~\refwithpage{aborted-delivery-attempts} occur frequently, but the
+aborted delivery attempts described in this section only occur four times
+in the 93 log files used for testing.  The symptoms are the same as in
+section~\refwithpage{aborted-delivery-attempts}, except that there
+\textit{is\/} a \texttt{postfix/cleanup} log line; there does not appear to
+be anything in the log file to explain why there are no further log
+messages.  The only way to detect these mails is to periodically scan all
+mails in the state tables, deleting any mails displaying the following
+characteristics:
 
@@
     \item The appropriate action to take --- this is discussed in
-        section~\ref{mails-covered}.
+        section~\refwithpage{mails-covered}.
 
@@
 Postfix change warning messages (usually adding more information), etc.\ so
-over time the log lines drift and change.  Graph~\ref{rule hits} shows the
-number of hits for each rule over all 93 log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
+over time the log lines drift and change.  Graph~\refwithpage{rule hits}
+shows the number of hits for each rule over all 93 log files; it's obvious
+that a small number of rules match the vast majority of the lines, and more
+than half the rules match fewer than 100 times.
 
@@
 the best way to detect mails which were not handled properly (many of the
-complications discussed in section~\ref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
+complications discussed in section~\refwithpage{additional complications}
+were detected in this way).  The parser issues warnings when it detects any
 errors, some of which may alert the user to a problem, e.g.\ when a queueid
@@
 problem.} or when there are problems tracking a child mail (see
-section~\ref{tracking re-injected mail}).  There should be few or no
-warnings, and the state table should only contain entries for mails which
-had yet to be delivered when the logfiles ended, or started before the
-logfiles began.
+section~\refwithpage{tracking re-injected mail}).  There should be few or
+no warnings, and the state table should only contain entries for mails
+which had yet to be delivered when the logfiles ended, or started before
+the logfiles began.
 
@@
         and 2.3.11; this may possibly have changed in Postfix 2.4.}
-        Rectifying this has already been described in section~\ref{logging
-        helo}.
+        Rectifying this has already been described in
+        section~\refwithpage{logging helo}.
 
@@
         distinction given the data Postfix logs, though it might be
-        possible to write a policy server to provide additional
-        logging.
+        possible to write a policy server to provide additional logging.
 
@@
 
-        \url{http://jimsun.linxnet.com/postfix_contrib.html}
-        \newline (Last checked 2007/08/13.)
+        \url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last
+        checked 2007/08/13.)
 
@@
         made to correlate log lines or deal with the difficulties
-        documented in sections~\ref{complications} and~\ref{additional
-        complications}.\footnote{If any attempt is made there is no
-        reference to it in the documentation available on the website.}
-        The source code is available in an obfuscated form only (presumably
-        for a fee), and the product is quite expensive, as it requires a
-        license for each report which is to be generated.  The web
-        interface allows creation of dynamic reports based on any field,
-        but due to the associated cost the author has not experimented with
-        it.
+        documented in sections~\refwithpage{complications}
+        and~\refwithpage{additional complications}.\footnote{If any attempt
+        is made there is no reference to it in the documentation available
+        on the website.} The source code is available in an obfuscated form
+        only (presumably for a fee), and the product is quite expensive, as
+        it requires a license for each report which is to be generated.
+        The web interface allows creation of dynamic reports based on any
+        field, but due to the associated cost the author has not
+        experimented with it.
 
@@
         \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
-        \newline
-        Fields extracted: from, to, client hostname, client IP, relay
-        hostname, relay IP, status, response code, RBL list and message id.
+        \newline Fields extracted: from, to, client hostname, client IP,
+        relay hostname, relay IP, status, response code, RBL list and
+        message id.
 
@@
         tools to utilise the Splunk database.  Splunk is a generic tool, so
-        it lacks any Postfix specific support over and above extracting
-        the \texttt{key=<value>} fields; most importantly it makes no
-        attempt to correlate log lines by queueid or pid, nor to handle any
-        of the myriad complications discussed in this document.
-        Correlation of log lines is left up to the user, and Splunk
-        certainly makes the job easier than using grep, awk and similar
-        Unix tools unless the user has a very high proficiency with those
-        tools.
+        it lacks any Postfix specific support over and above extracting the
+        \texttt{key=<value>} fields; most importantly it makes no attempt
+        to correlate log lines by queueid or pid, nor to handle any of the
+        myriad complications discussed in this document.  Correlation of
+        log lines is left up to the user, and Splunk certainly makes the
+        job easier than using grep, awk and similar Unix tools unless the
+        user has a very high proficiency with those tools.
 
@@
         Aggregations: count, bytes (the sum of bytes transferred; count is
-        not explained).\newline
-        Keyfields [sic]: date, hour, sender, recipient, defer\_count, delay,
-        incoming\_host, outgoing\_host.
+        not explained).\newline Keyfields [sic]: date, hour, sender,
+        recipient, defer\_count, delay, incoming\_host, outgoing\_host.
 
@@
         makes no attempt to deal with the other complications described in
-        sections~\ref{complications} and~\ref{additional complications}.
-        It stores data, but not in an easily accessible format.  It
-        generates reports in multiple formats, but those reports do not
-        appear to be customisable.
+        sections~\refwithpage{complications} and~\refwithpage{additional
+        complications}.  It stores data, but not in an easily accessible
+        format.  It generates reports in multiple formats, but those
+        reports do not appear to be customisable.
 
@@
 
-            \item For mail sent and received: from address, size, and
-                time and date.
+            \item For mail sent and received: from address, size, and time
+                and date.
 
@@
         Needless to say it doesn't attempt to address any of the
-        complications in sections~\ref{complications} and~\ref{additional
-        complications} except for correlating by queueid.
+        complications in sections~\refwithpage{complications}
+        and~\refwithpage{additional complications} except for correlating
+        by queueid.
 
@@
     \item [Log Mail Analyser] Please see the previous discussion in
-        section~\ref{prior art}.
+        section~\refwithpage{prior art}.
 
@@
 anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in section~\ref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in section~\ref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The next group of graphs
-vividly shows the huge impact that caching compiled regexs has on parser
-performance.  This section concludes with a breakdown of the rule hits
-accumulated during a single test run of the parser.
+illustrate the topics discussed in section~\refwithpage{rule efficiency}.
+The graphs in the first section cover parser scalability, demonstrating
+that performance scales linearly with input size.  The impact of rule
+ordering is shown in section~\refwithpage{rule ordering graphs}, and the
+anomalous dips and peaks apparent in some graphs are explained.  The next
+group of graphs vividly shows the huge impact that caching compiled regexs
+has on parser performance.  This section concludes with a breakdown of the
+rule hits accumulated during a single test run of the parser.
 
@@
 The dips at logfiles 22 and 62--68 correspond to peaks in log file size in
-graph~\ref{execution time vs file size vs number of lines graph}, and peaks in
-graph~\ref{execution time vs file size vs number lines factor} (where a
-peak means that more lines are processed per second, i.e.\ performance is
-better).  The explanation for this took some time to arrive at, but it
-turns out to be reasonably simple.  The large log files in question were
-caused by a mail forwarding loop, where the distribution of log lines is
-quite different to normal, resulting in different performance
+graph~\refwithpage{execution time vs file size vs number of lines graph},
+and peaks in graph~\refwithpage{execution time vs file size vs number lines
+factor} (where a peak means that more lines are processed per second, i.e.\
+performance is better).  The explanation for this took some time to arrive
+at, but it turns out to be reasonably simple.  The large log files in
+question were caused by a mail forwarding loop, where the distribution of
+log lines is quite different to normal, resulting in different performance
 characteristics.
@@
 
-    \item Postfix accepts a mail from username@domain, for
-        username@domain.
+    \item Postfix accepts a mail from username@domain, for username@domain.
 
@@
         envelope sender address is now username@domain.  Log lines are
-        added by \textit{postfix/local},
-        \textit{postfix/qmgr\/} (twice), \textit{postfix/cleanup}, and finally
-        \textit{postfix/pickup}.
+        added by \textit{postfix/local}, \textit{postfix/qmgr\/} (twice),
+        \textit{postfix/cleanup}, and finally \textit{postfix/pickup}.
 
@@
 lower than the average number required to parse a random log line from logs
-without a mail loop, as can be seen in the peaks in graphs~\ref{execution
-time vs file size vs number lines factor},~\ref{percentage increase of
-shuffled over normal} and~\ref{percentage increase of reversed over
-normal}.  The difference is more extreme in the reversed ordering case
-because the largest number of rules to be consulted for the commonly
-occurring log lines during a mail loop is far less than the largest number
-of rules that are consulted for commonly occurring lines outside of a mail
-loop.  The number of rules which needs to be consulted when the ordering is
-shuffled varies between the optimum and nadir, and the performance varies
+without a mail loop, as can be seen in the peaks in
+graphs~\refwithpage{execution time vs file size vs number lines
+factor},~\refwithpage{percentage increase of shuffled over normal}
+and~\refwithpage{percentage increase of reversed over normal}.  The
+difference is more extreme in the reversed ordering case because the
+largest number of rules to be consulted for the commonly occurring log
+lines during a mail loop is far less than the largest number of rules that
+are consulted for commonly occurring lines outside of a mail loop.  The
+number of rules which needs to be consulted when the ordering is shuffled
+varies between the optimum and nadir, and the performance varies
 proportionally.
@@
 
-Two large dips can be seen in graph~\ref{normal regex vs discarded regex
-factor} at log files 22 and 62--68, corresponding to the spikes in log file
-size in graphs~\ref{execution time vs file size vs number of lines graph}
-and~\ref{normal regex vs discard regex}.  The reason for the anomalous log
-files has already been explained in section~\ref{Why are there dips in the
-graphs?}.
+Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
+regex factor} at log files 22 and 62--68, corresponding to the spikes in
+log file size in graphs~\refwithpage{execution time vs file size vs number
+of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
+reason for the anomalous log files has already been explained in
+section~\refwithpage{Why are there dips in the graphs?}.
 
@@
 
-As graph~\ref{rule hits graph} is quite difficult to read it has been
-separated into three sections: low, middle and high.
+As graph~\refwithpage{rule hits graph} is quite difficult to read it has
+been separated into three sections: low, middle and high.
 
@@
 
-It is apparent from the low graph (graph~\ref{hits per rule low}) that some
-rules have few or no hits; those with zero hits are rules which were
-written to parse log files used during development of the parser but not
-utilised in the test runs performed for this document.
+It is apparent from the low graph (graph~\refwithpage{hits per rule low})
+that some rules have few or no hits; those with zero hits are rules which
+were written to parse log files used during development of the parser but
+not utilised in the test runs performed for this document.
 
@@
 The graphs presented in this section illustrate the topics discussed in
-section~\ref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linearly relationship between execution
-time and input size.  Section~\ref{rule ordering graphs} relates to the
-effect of rule ordering on execution time, and the unexpected consequences
-of specific inputs.  The necessity of caching compiled regexs is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  This section concluded with a breakdown of the
-rule hits accumulated during a single test run of the parser.
+section~\refwithpage{rule efficiency}.  The first collection of graphs are
+about parser scalability, showing the linearly relationship between
+execution time and input size.  Section~\refwithpage{rule ordering graphs}
+relates to the effect of rule ordering on execution time, and the
+unexpected consequences of specific inputs.  The necessity of caching
+compiled regexs is attested to by the third group of graphs, where the
+difference between caching and discarding is staggering.  This section
+concluded with a breakdown of the rule hits accumulated during a single
+test run of the parser.
 

### 2007-10-22 doc/logparser.tex
Made the flow chart into a figure, which resulted in minor tidying. Small cleanups for overfull hboxes etc.

@@
 \tableofcontents
+\listoffigures
 
@@
 
-This flow chart shows the paths the data representing a mail/connection can
-take through the parser algorithm.  The flow chart covers the most common
-paths only; there are additional, uncommon paths which are excluded for the
-sake of clarity.
-
-\includegraphics{build/logparser-flow-chart.ps}
+Figure~\ref{flow chart image} shows the paths the data representing a
+mail/connection can take through the parser algorithm.  The flow chart
+covers the most common paths only; there are additional, uncommon paths
+which are excluded for the sake of clarity, but are described in
+section~\refwithpage{additional complications}.
+
+% This is very likely to move around as the flow chart is as big as
+% possible, so if any preceding text changes the flow chart may end up on
+% the next page, which may not be a bad thing, maybe.  What woould be a bad
+% thing is the \clearpage seems to cause a blank page sometimes, if the
+% image is just slightly too big.
+\showgraph{build/logparser-flow-chart.ps}{Parser flow chart}{flow chart
+image}
+\clearpage
 
@@
 
-    \item The appropriate action to take --- this is discussed in
+    \item The appropriate action to take --- discussed in
         section~\refwithpage{mails-covered}.
@@
 
-    \item [Log Mail Analyser] Please see the previous discussion in
-        section~\refwithpage{prior art}.
+    \item [Log Mail Analyser] Please see the previous in-depth discussion
+        of Log Mail Analyser in section~\refwithpage{prior art}.
 

### 2007-10-23 doc/logparser.tex
Documented the EXPIRY action, including changes to the flow chart and references to the flow chart.

@@
     \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: commit; flowchart:~12).  This is the last
+        administrator (action: commit; flowchart:~13).  This is the last
         log line for this particular mail (though it may be indirectly
@@
         child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables
-        (flowchart:~15).  Re-injected mails are described in
+        database (flowchart:~15), then deleted from the state tables
+        (flowchart:~16).  Re-injected mails are described in
         section~\refwithpage{tracking re-injected mail}.
@@
         whose parent hasn't yet been identified (action: commit;
-        flowchart:~16).  Mark the mail as ready for entry in the database
-        (flowchart:~17), and wait for the parent to deal with it
-        (flowchart:~18).  The mail should not have subsequent log entries;
+        flowchart:~17).  Mark the mail as ready for entry in the database
+        (flowchart:~18), and wait for the parent to deal with it
+        (flowchart:~19).  The mail should not have subsequent log entries;
         only its parent should refer to it.
@@
     \item If the mail is a child mail then it has already been tracked
-        (action: commit; flowchart:~19): the data is cleaned up, missing
+        (action: commit; flowchart:~20): the data is cleaned up, missing
         data is copied from its parent if necessary, and the child is
-        entered in the database (flowchart:~20), then deleted from the
-        state tables (flowchart:~21).  The child mail will be removed from
-        the parent mail's list of children (flowchart:~22); if this is the
+        entered in the database (flowchart:~21), then deleted from the
+        state tables (flowchart:~22).  The child mail will be removed from
+        the parent mail's list of children (flowchart:~23); if this is the
         last child and the parent has also been entered in the database the
@@
     \item The last alternative is that the mail is a parent mail (action:
-        commit; flowchart:~23).  Regardless of the state of its children
-        the data is cleaned up and entered in the database (flowchart:~24).
+        commit; flowchart:~24).  Regardless of the state of its children
+        the data is cleaned up and entered in the database (flowchart:~25).
         The parent may have children which are waiting to be entered in the
-        database (flowchart:~25); each of those children's data is cleaned
+        database (flowchart:~26); each of those children's data is cleaned
         up and entered in the database, then deleted from the state tables.
         The parent may also have outstanding children which are not yet
-        delivered, in which case (flowchart:~26) the parent must wait for
+        delivered, in which case (flowchart:~27) the parent must wait for
         those children to be finished with.  As soon as the last child is
         deleted from the state tables the parent will also be finished with
-        (flowchart:~27), and deleted from the state tables.
+        (flowchart:~28), and deleted from the state tables.
 
@@
         in the log line, or with the connection if not.
+
+    \item [EXPIRY] If Postfix has not managed to deliver a mail after
+        trying for five days it will give up and return the mail to the
+        sender.  When this happens the mail will not have a combination of
+        Postfix programs which passes the valid combinations check (see
+        section~\refwithpage{out of order log lines}).  To ensure that the
+        mail can be committed the EXPIRY action sets a flag marking the
+        mail as expired; the flag later causes the valid combinations check
+        to be skipped, so the mail will be committed.
 

### 2007-11-06 doc/logparser-acronyms.tex
Added a glossary and acronym list. Started working on Carl's corrections.

% Warning: don't use acronyms within definitions, they don't work properly.

\newacronym{RBL}{Real-time Black List}{description={
    Real-time Black Lists are a simple collaborative anti-spam technique
    used to reject or penalise email sent from mail servers reported to
    have sent large volumes of spam.  To use an RBL the mail server makes a
    DNS request incorporating the IP address of the currently connected
    client; if the requested hostname exists the client is on the RBL, and
    the mail server can decide what course of action to take.
}}

\newacronym{API}{Application Programming Interface}{description={
    One of the fundamental concepts when writing programs is the reuse of
    existing code, so that each program does not reinvent the wheel.  When
    a body of code is intended to be reused repeatedly, the user of this
    code needs to be informed of the functionality provided by the code.
    An API defines the interface provided to the user, and acts a contract
    between the user and the provider: if the user adheres to the API the
    provider guarantees it will work, while the provider is free to change
    the implementation as long as the API is preserved.
}}

\newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
    SMTP is the protocol which transfers mail from the sender to the
    recipient across the Internet.  A brief introduction to SMTP is
    provided in section~\refwithpage{SMTP background}.
}}

\newacronym{MTA}{Mail Transfer Agent}{description={
    A Mail Transfer Agent sends and/or receives mail via SMTP\@.
}}

\newacronym{RFC}{Request For Comments}{description={
    The Request For Comments series is a series of proposals defining
    various protocols and file formats, e.g. SMTP\@.  The name is somewhat
    misleading: initially the authors were asking for peer review, but
    these documents are now the de facto standards the Internet runs on.
}}

### 2007-11-06 doc/logparser.tex
Added a glossary and acronym list. Started working on Carl's corrections.

@@
 \usepackage{url}
-%\DeclareUrlCommand\urlwrap{}
-%\newcommand{\urlwrapfill}[1]{\urlwrap{#1}\hfill}
 \newcommand{\parsername}{ASO}
@@
 \usepackage[section,numbib]{tocbibind}
+
+\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
+% Redefine the command used to produce the glossary title, because the
+% default command produces an unnumbered section whereas I want a numbered
+% section.
+\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
+\makeglossary{}
+\makeacronym{}
+
 % When creating a PDF make the table of contents into links to the pages
@@
 \newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} (page~\pageref{#1})%
+    \empty{}\ref{#1} [page~\pageref{#1}]%
 }
@@
 \begin{document}
+
+% Pull in the acronyms early, so they can be used throughout the text.
+\input{logparser-acronyms.tex}
 
@@
 Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\date{}
 \maketitle
@@
 was added on an ad hoc basis as a requirement was discovered or new
-functionality was added.  Further complication arises because the set of
-rejection messages is not fixed: new messages can be added by the
-administrator with custom checks; every Real-time Black List returns a
-different explanatory message; policy servers may log different messages
-depending on the characteristics of the connection; there are many ways in
-which the log lines may differ between servers.  This paper documents the
-difficult process of parsing Postfix logs, presenting a program which
-parses logs and places the resulting data into a database.  The gathered
-data can then be used to optimise current anti-spam measures, provide a
-baseline to test new anti-spam measures against, or to produce statistics
-showing how effective those measures are.  There are numerous other uses
-for such data: improving server performance by identifying troublesome
-destinations and reconfiguring appropriately or dedicating transports to
-those destinations; identifying regular high volume uses (e.g.\ customer
-newsletters) and restricting those uses to off-peak times; detecting virus
-outbreaks which propagate via email; as a base for billing customers on a
-shared server.  Preserving the raw data enables users to develop a
-multitude of uses far beyond those conceived of by the author.
+functionality was added.\footnote{A history of all changes made to Postfix
+is distributed with the source code, available from
+\url{http://www.postfix.org/}} Further complication arises because the set
+of rejection messages is not fixed: new messages can be added by the
+administrator with custom checks; every \RBL{} returns a different
+explanatory message; policy servers may log different messages depending on
+the characteristics of the connection; there are many ways in which the log
+lines may differ between servers, even within the same organisation when
+servers are configured differently, or are running different version of
+Postfix.  This paper documents the difficult process of parsing Postfix
+logs, presenting a program which parses logs and places the resulting data
+into a database.  The gathered data can then be used to optimise current
+anti-spam measures, provide a baseline to test new anti-spam measures
+against, or to produce statistics showing how effective those measures are.
+There are numerous other uses for such data: improving server performance
+by identifying troublesome destinations and reconfiguring appropriately or
+dedicating transports to those destinations; identifying regular high
+volume uses (e.g.\ customer newsletters) and restricting those uses to
+off-peak times; detecting virus outbreaks which propagate via email; as a
+base for billing customers on a shared server.  Preserving the raw data
+enables users to develop a multitude of uses far beyond those conceived of
+by the author.
 
@@
 Section~\ref{rules} discusses the parsing rules in detail, describing the
-table used to hold the rules and explaining their structure, using an
-example rule and sample data it matches successfully against, followed by a
-discussion of rule efficiency concerns.
+table in the database used to hold the rules and explaining their
+structure, referring to an example rule and sample data it matches
+successfully against, followed by a discussion of rule efficiency concerns.
 
@@
 
-Section~\ref{other-parsers} discusses other parsers and why they were
+Appendix~\ref{other-parsers} discusses other parsers and why they were
 deemed unsuitable for the task.
@@
 
-Section~\ref{graphs} contains graphs illustrating the topics discussed in
-section~\ref{rule efficiency}, rule efficiency.
+Appendix~\ref{graphs} contains graphs illustrating the topics previously
+discussed in section~\ref{rule efficiency}, rule efficiency, and explains
+the irregularities observed in the graphs.
+
+Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
 
@@
 underlying the project, followed by some technical information: the use of
-a database as an API\@; a brief introduction to the SMTP protocol; and a
-longer introduction to Postfix, concentrating on the topics most relevant
-to this paper, namely Postfix restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
+a database as an \API\@; a brief introduction to
+\ifacronymfirstuse{SMTP}{the}{} \SMTP\@; and a longer introduction to
+Postfix, concentrating on the topics most relevant to this paper, namely
+Postfix restrictions and policy servers.  The assumptions made in designing
+and implementing the parser are explained, as are the conventions used in
+this paper.  Other projects which attempt to parse Postfix logs are
+summarised (full details are available in
 section~\refwithpage{other-parsers}), plus a review of previously published
@@
 
+These database queries yield summary statistics about the efficiency of
+spam avoidance techniques that is far less feasible to assess directly from
+log files without prior pre-processing into a database along the lines
+proposed, implemented and tested herein.
+
 \subsection{Database as Application Programming Interface}
@@
 \subsection{SMTP background}
-\label{smtp background}
-
-The Simple Mail Transfer Protocol, originally defined in RFC
-821~\cite{RFC821} and updated in RFC 2821~\cite{RFC2821}, is used for
-transferring mail between Mail Transport Agents (MTAs).  It is a simple,
-human readable, plain text protocol, making it quite easy to test and debug
-problems with it.  Despite the simplicity many virus and/or spam sending
-programs fail to implement it properly, so requiring strict adherence to
-the protocol specification is beneficial in protecting against spam and
-viruses.\footnote{This violates the principle of \textit{Be liberal in what
-you accept, and conservative in what you send\/} from RFC
-760~\cite{rfc760}, but unfortunately that principle was written in a
-friendlier time.}  A typical SMTP conversation resembles the following (the
-lines starting with a three digit number are sent by the server, all other
-lines are sent by the client):
+\label{SMTP background}
+
+The \SMTPlong, originally defined in \RFC{} 821~\cite{RFC821} and updated
+in \RFC{} 2821~\cite{RFC2821}, is used for transferring mail between the
+sending and receiving \MTA\@.  It is a simple, human readable, plain text
+protocol, making it quite easy to test and debug problems with it.  Despite
+the simplicity many virus and/or spam sending programs fail to implement it
+properly, so requiring strict adherence to the protocol specification is
+beneficial in protecting against spam and viruses.\footnote{This violates
+the principle of \textit{Be liberal in what you accept, and conservative in
+what you send\/} from \RFC{} 760~\cite{rfc760}, but unfortunately that
+principle was written in a friendlier time.}  A typical SMTP conversation
+resembles the following (the lines starting with a three digit number are
+sent by the server, all other lines are sent by the client):
 
@@
 
-            \item Modify \texttt{smtpd\_helo\_restrictions} in
+            \item Modify \texttt{smtpd\_data\_restrictions} in
                 \texttt{/etc/postfix/main.cf} to contain\newline
@@
 
-        The purpose of logging the HELO in
-        \texttt{smtpd\_helo\_restrictions} is to log only once per mail; it
-        also seems like the natural place.
+        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
+        place to log the HELO hostname, there won't be a queueid associated
+        with the mail for the first recipient, so that log line cannot be
+        associated with the correct mail.  There is guaranteed to be
+        queueid when the DATA command has been reached, and thus it will be
+        logged by any restrictions taking effect in
+        \texttt{smtpd\_data\_restrictions}.
 
@@
 
+% The appropriate command has been redefined so this will produce a section
+% and label called glossary.
+\printglossary{}
+\printacronym{}
+
 \end{document}

### 2007-11-06 doc/logparser.tex
Replaced SMTP with \SMTP{} everywhere. Put the included tables into table environments and added a list of tables. Added the glossary and made it and the acronym list play together.

@@
 \usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
 \makeglossary{}
@@
 
+%\showtable{filename}{caption}{label}
+\newcommand{\showtable}[3]{
+    \begin{table}[ht]
+        \caption{#2}\label{#3}
+        \input{#1}
+    \end{table}
+}
+
 % Replacement for \ref{}, adds the page number too.
@@
 \listoffigures
+\listoftables
 
@@
 the algorithm, writing the program and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding SMTP,
-Postfix, anti-spam techniques, or the paper.
+are some additional resources expected to be helpful in understanding
+\SMTP{}, Postfix, anti-spam techniques, or the paper.
 
@@
 what you send\/} from \RFC{} 760~\cite{rfc760}, but unfortunately that
-principle was written in a friendlier time.}  A typical SMTP conversation
-resembles the following (the lines starting with a three digit number are
-sent by the server, all other lines are sent by the client):
+principle was written in a friendlier time.}  A typical \SMTP{}
+conversation resembles the following (the lines starting with a three digit
+number are sent by the server, all other lines are sent by the client):
 
@@
 
-A detailed description of SMTP is beyond the scope of this document:
+A detailed description of \SMTP{} is beyond the scope of this document:
 introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
@@
 Postfix uses several restriction lists (each containing zero or more
-restrictions) to decide the fate of an SMTP command, one list for each
-stage of the SMTP conversation: client connection, HELO command, MAIL FROM
-command, RCPT TO command (possibly multiple times), DATA command, and end
-of data.  By default restriction lists will not be evaluated until the
+restrictions) to decide the fate of an \SMTP{} command, one list for each
+stage of the \SMTP{} conversation: client connection, HELO command, MAIL
+FROM command, RCPT TO command (possibly multiple times), DATA command, and
+end of data.  By default restriction lists will not be evaluated until the
 first RCPT TO command is received, because some clients don't deal with
@@
 A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each SMTP command not rejected by an
+state information from Postfix for each \SMTP{} command not rejected by an
 earlier restriction; the policy server can utilise that state information
@@
         resistance to the proposal because it breaks or vastly complicates
-        some features of SMTP\@.
+        some features of \SMTP{}\@.
 
@@
         using increasingly complex and well written programs to distribute
-        spam, frequently using an ISP provided SMTP server from a
+        spam, frequently using an ISP provided \SMTP{} server from a
         compromised machine on the ISP's network.  Greylisting is slowly
@@
 The data stored is limited to time and date, hostname and IP address of
-client, sender and recipient addresses, and the SMTP code.  Handling of
-multiple recipients, SMTP codes or remote servers\footnote{A single mail
+client, sender and recipient addresses, and the \SMTP{} code.  Handling of
+multiple recipients, \SMTP{} codes or remote servers\footnote{A single mail
 may be sent to multiple remote servers if it was addressed to recipients in
@@
 
-    \item The SMTP protocol.
+    \item The \SMTP{} protocol.
 
@@
         or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO when it rejects an SMTP command, but it is quite easy
-        to rectify this:
+        logs the HELO when it rejects an \SMTP{} command, but it is quite
+        easy to rectify this:
 
@@
 INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
-SMTP command, delivering a mail, or bouncing a mail.  Any log line where
+\SMTP{} command, delivering a mail, or bouncing a mail.  Any log line where
 the Postfix action is INFO is not interesting in and of itself, but
@@
     \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an SMTP command --- a
+        enforcing a restriction that would reject an \SMTP{} command --- a
         facility that is quite useful for testing new restrictions.  This
@@
 
-    \item [smtp\_code] The SMTP code associated with the log line.  In
-        general an SMTP code is only present for a rejection or final
-        delivery; results missing an SMTP code will duplicate the SMTP code
-        of other results in the connection.  Some final delivery log lines
-        don't contain an SMTP code: in those cases the code is faked based
-        on the success or failure represented by the log line.
+    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
+        general an \SMTP{} code is only present for a rejection or final
+        delivery; results missing an \SMTP{} code will duplicate the
+        \SMTP{} code of other results in the connection.  Some final
+        delivery log lines don't contain an \SMTP{} code: in those cases
+        the code is faked based on the success or failure represented by
+        the log line.
 
@@
 specific rule to match some lines and a more general rule to match the
-majority, e.g.\ separating smtp delivery to a filtering proxy from smtp
-delivery to the rest of the world.  The algorithm provides a facility for
-ordering overlapping rules: the priority field in each rule (defaults to
-zero).  Rules are tried in order of priority, highest first, overriding the
-hits field maintained by the parser.  Negative priorities may also be
-useful for catchall rules.
+majority, e.g.\ separating \SMTP{} delivery to a filtering proxy from
+\SMTP{} delivery to the rest of the world.  The algorithm provides a
+facility for ordering overlapping rules: the priority field in each rule
+(defaults to zero).  Rules are tried in order of priority, highest first,
+overriding the hits field maintained by the parser.  Negative priorities
+may also be useful for catchall rules.
 
@@
 
-    \item Mail enters the system via SMTP or local submission.
+    \item Mail enters the system via \SMTP{} or local submission.
 
@@
 Everything starts off with a mail entering the system, whether by local
-submission via postdrop/sendmail, by SMTP, by re-injection due to
+submission via postdrop/sendmail, by \SMTP{}, by re-injection due to
 forwarding, or internally generated by Postfix.  Local submission is the
@@
 
-SMTP is more complicated:
+\SMTP{} is more complicated:
 
@@
     \item The client disconnects (action: disconnect; flowchart:~6).  If
-        Postfix has rejected any SMTP commands the data will be saved to
+        Postfix has rejected any \SMTP{} commands the data will be saved to
         the database; if not there won't be any data to save (any mails
@@
         usually would (action: track; flowchart:~11).\footnote{Relayed
-        delivery is performed by the SMTP client; local delivery means
+        delivery is performed by the \SMTP{} client; local delivery means
         local to the server, i.e.\ an address the server is final
@@
 
-    \item [REJECTION] Deal with postfix rejecting an SMTP command from the
-        remote client: log the rejection with a mail if there is a queueid
-        in the log line, or with the connection if not.
+    \item [REJECTION] Deal with postfix rejecting an \SMTP{} command from
+        the remote client: log the rejection with a mail if there is a
+        queueid in the log line, or with the connection if not.
 
@@
         the mail, indicating it was generated internally by Postfix, not
-        accepted via SMTP or submitted locally by sendmail.
+        accepted via \SMTP{} or submitted locally by sendmail.
 
@@
 Some mail clients appear to send the following sequence of commands during
-the SMTP session:
+the \SMTP{} session:
 
@@
 
-The DATA phase of the SMTP conversation is where the headers and body of the
-mail are transferred.  Sometimes there is a timeout or the connection is
-lost\footnote{For brevity's sake timeout will be used throughout this
+The DATA phase of the \SMTP{} conversation is where the headers and body of
+the mail are transferred.  Sometimes there is a timeout or the connection
+is lost\footnote{For brevity's sake timeout will be used throughout this
 section, but everything applies equally to lost connections.} during the
@@
 
-            \item For mail sent: to addresses, SMTP code, and delay.
+            \item For mail sent: to addresses, \SMTP{} code, and delay.
 
@@
 
-\input{build/stats-normal-filesize-line-count-include}
+\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
+size \& number of lines to execution time: statistics}{execution time vs
+file size vs number lines factor table}
 
@@
 The difference between logs with a mail loop and logs without can be seen
-in this table showing the increases for the different rule orderings and
-combinations of log files:
-
-\noindent\input{build/stats-normal-shuffle-reverse-include.tex}
+in table~\refwithpage{Rule ordering with and without a mail loop} showing
+the increases for the different rule orderings and combinations of log
+files:
+
+\showtable{build/stats-normal-shuffle-reverse-include}{Rule ordering with
+and without a mail loop}{Rule ordering with and without a mail loop}
 
@@
 
-\showgraph{build/plot-cached-discarded-factor.ps}{Regex: percentage
+\showgraph{build/plot-cached-discarded-factor.ps}{Regex caching: percentage
 execution time increase}{normal regex vs discarded regex factor}
@@
 
-\input{build/stats-cached-discarded-include-for-graph}
+\showtable{build/stats-cached-discarded-include-for-graph}
+{Regex caching/discarding with different groups of log files}
+{Regex caching/discarding with different groups of log files}
 
@@
 
-% The appropriate command has been redefined so this will produce a section
-% and label called glossary.
+% This fools the syntax highlighting so we get a new fold here, but it
+% doesn't create a new section.
+\ifacronymfirstuse{SMTP}{
+\section{Glossary}
+}{}
+% Redefine the command used to produce the glossary title, because the
+% default command produces an unnumbered section whereas I want a numbered
+% section.
+\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
 \printglossary{}
+% Redefine the command a second time to produce acronyms instead of a
+% glossary.
+\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
 \printacronym{}
 
+% The contents of the glossary.
+\glossary{name={ESMTP},description={
+    ESMTP is Extended \SMTP{}, defined in \RFC{}~1869~\cite{RFC1869}.
+}}
+
 \end{document}

### 2007-11-06 doc/logparser-acronyms.tex
Replaced SMTP with \SMTP{} everywhere. Put the included tables into table environments and added a list of tables. Added the glossary and made it and the acronym list play together.
Lots of corrections, including some of Carl's. Lots more acronyms; adding those causes large diffs without appreciable changes     in the processed document.

@@
+% % vim: set textwidth=75 spell :
 % Warning: don't use acronyms within definitions, they don't work properly.
@@
 }}
+
+\newacronym{SQL}{Structured Query Language}{description={
+    SQL is the standard language used for database querying, modification
+    and maintenance.  Some information about SQL, including its history,
+    can be found at \cite{wikipedia-sql}.
+}}
+
+\newacronym{SPF}{Sender Policy Framework}{description={
+    SPF is explained fully in~\cite{openspf, wikipedia-spf}.
+}}
+
+\newacronym{ISP}{Internet Service Provider}{description={
+    An ISP is a company which sells Internet access to consumers.
+}}
+
+\newacronym{LMA}{Log Mail Analyzer}{description={
+    One of the other Postfix log parsers reviewed and the only example of
+    published prior art~\cite{log-mail-analyser} found by the author.
+}}
+
+\newacronym{CSV}{Comma-Separated Value}{description={
+    The most basic form of database available, each record is a single line
+    in the file and the fields are separated by a special character,
+    typically a comma or colon.
+}}

### 2007-11-06 doc/logparser.tex
Lots of corrections, including some of Carl's. Lots more acronyms; adding those causes large diffs without appreciable changes     in the processed document.

@@
 the characteristics of the connection; there are many ways in which the log
-lines may differ between servers, even within the same organisation when
-servers are configured differently, or are running different version of
+lines may differ between servers, even within the same organisation:
+servers may be configured differently, or running different version of
 Postfix.  This paper documents the difficult process of parsing Postfix
 logs, presenting a program which parses logs and places the resulting data
-into a database.  The gathered data can then be used to optimise current
-anti-spam measures, provide a baseline to test new anti-spam measures
-against, or to produce statistics showing how effective those measures are.
-There are numerous other uses for such data: improving server performance
-by identifying troublesome destinations and reconfiguring appropriately or
-dedicating transports to those destinations; identifying regular high
-volume uses (e.g.\ customer newsletters) and restricting those uses to
-off-peak times; detecting virus outbreaks which propagate via email; as a
-base for billing customers on a shared server.  Preserving the raw data
-enables users to develop a multitude of uses far beyond those conceived of
-by the author.
+into a database for later use.  The gathered data can then be used to
+optimise current anti-spam measures, provide a baseline to test new
+anti-spam measures against, or to produce statistics showing how effective
+those measures are.  There are numerous other uses for such data: improving
+server performance by identifying troublesome destinations and
+reconfiguring appropriately or dedicating transports to those destinations;
+identifying regular high volume uses (e.g.\ customer newsletters) and
+restricting those uses to off-peak times; detecting virus outbreaks which
+propagate via email; as a base for billing customers on a shared server.
+Preserving the raw data enables users to develop a multitude of uses far
+beyond those conceived of by the author.
 
@@
 
-Section~\ref{database schema} partially describes the database schema used
-by the parser, explaining in detail the tables used for storing the
-resulting data.  The remainder of the schema is laid out in
-section~\ref{rule attributes}.
+Section~\ref{database schema} describes the database schema used by the
+parser, explaining in detail the tables used for storing the data gleaned
+from the logs and the table that stores the rules.
 
@@
 structure, referring to an example rule and sample data it matches
-successfully against, followed by a discussion of rule efficiency concerns.
+successfully against.  This section concludes with a discussion of rule
+efficiency concerns.
 
@@
 the full algorithm, followed by a comprehensive explanation of the
-different stages of the algorithm, the actions taken during the execution
-of the algorithm, and further complications whose solution completes the
+different stages of the algorithm, the actions taken during execution of
+the algorithm, and further complications whose solution completes the
 algorithm.
@@
 1000 users.  Coverage is described both in terms of the fraction of lines
-parsed and the fraction of mails and connections successfully handled by
-the parsing algorithm.
-
-Section~\ref{limitations-improvements} lists the limitations of the program
-and/or the algorithm, then suggests some ways of dealing with the
-limitations, with the goal of improving parsing and reproduction of the
-journey a mail takes through Postfix.
-
-Section~\ref{conclusion} describes the results of the project to date and
-contains the paper's conclusion.
+parsed and the fraction of mails and connections successfully reconstructed
+by the parsing algorithm.
+
+Section~\ref{limitations-improvements} lists the limitations of the
+algorithm, then suggests some ways of dealing with the limitations, with
+the goal of improving parsing and reproduction of the journey a mail takes
+through Postfix.
+
+Section~\ref{conclusion} is contains the paper's conclusion.
 
 Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task.
+deemed unsuitable for the task, including why they could not be improved or
+expanded upon.
 
@@
 the irregularities observed in the graphs.
+
+Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
 
@@
 Postfix, concentrating on the topics most relevant to this paper, namely
-Postfix restrictions and policy servers.  The assumptions made in designing
-and implementing the parser are explained, as are the conventions used in
-this paper.  Other projects which attempt to parse Postfix logs are
+Postfix anti-spam restrictions and policy servers.  The assumptions made in
+designing and implementing the parser are explained, as are the conventions
+used in this paper.  Other projects which attempt to parse Postfix logs are
 summarised (full details are available in
-section~\refwithpage{other-parsers}), plus a review of previously published
-research in this area.
+appendix~\refwithpage{other-parsers}), finishing with a review of
+previously published research in this area.
 
@@
 
-SQL note: $||$ is the concatenation operator in SQLite3; if the database
+\SQL{} note: $||$ is the concatenation operator in SQLite3; if the database
 containing the extracted data does not support this syntax, then simply
@@
 Another example is determining which restrictions are not effective: this
-example shows which rules had less than 100 hits on the last log file
-parsed.
+example shows which restrictions had fewer than 100 hits on the last log
+file parsed.
 
@@
 Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to the API provided by a library: it insulates both user and
-provider of the API from changes in the implementation of the API\@.  The
-algorithm implemented by the parser can be improved; support can be added
-for earlier or later releases of Postfix; bugs can be fixed or limitations
-removed from the parser.  Statistics and/or graphs can be generated from
-the database; new restrictions can be tested and the results inspected;
-trends in the fight against spam can emerge from historical data saved in
-the database.  Using a database simplifies writing programs which need to
-interact with the data in several ways:
+quite similar to any other \API{} provided by shared code: it insulates
+both user and provider of the \API{} from changes in the implementation of
+the \API{}\@.  The algorithm implemented by the parser can be improved;
+support can be added for earlier or later releases of Postfix; bugs can be
+fixed or limitations removed from the parser; these changes will not cause
+the user to be negatively impacted.  Statistics and/or graphs can be
+generated from the database; new restrictions can be tested and the results
+inspected; trends in the fight against spam can emerge from historical data
+saved in the database; the parser remains the same as the usage adapts.
+Using a database simplifies writing programs which need to interact with
+the data in several ways:
 
@@
 
-    \item Libraries exist for the majority of programming languages
-        allowing access to databases, whereas if this parser was written as
-        a shared library a new interface layer would need to be written for
-        every programming language using the API\@.
+    \item The majority of programming languages have existing code
+        available allowing access to databases, so the majority of
+        languages will be able to access the results obtained by running
+        the parser.  If this parser was written to be utilised as shared
+        code there would be little or no difficulty in using it when
+        writing in the programing language it is implemented in, but every
+        other programing language wishing to utilise it would require an
+        interface layer to be written.  This requirement would drastically
+        reduce the viability of any project wishing to build upon the work
+        of this project.
 
@@
                 program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.
+                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
+                for explanations of these terms, e.g.~\cite{wikipedia-sql}
+                XXX IMPROVE.}
 
@@
         vendor implements a different dialect of SQL, the basics are the
-        same everywhere.
+        same everywhere (analogous to the overall similarities and minor
+        differences between British English, American English and
+        Australian English).  XXX CITE SOME RESOURCES\@.
 
 \end{enumerate}
+
+Storing the results in a database will also increase the efficiency of
+using those results, as the logs need only be parsed once; indeed the
+results may be used by someone with no access to the original logs.
 
@@
 exceptions, with policy servers\footnote{Policy servers will be explained
-shortly.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use \texttt{grep(1)} to
-determine the fate of an individual email, following the journey an email
-takes through Postfix can be quite difficult.  The logs tend to follow a
-90\%-10\% pattern: 90\% of the time the journey is simple, but the other
-10\% of the time requires 90\% of the code.\footnote{These numbers don't
-have a solid scientific basis, they're based on gut feeling from writing
-and debugging the software.}
+in section~\refwithpage{policy servers}.} as the ultimate in flexibility.
+Administrators can also supply their own rejection messages to make it
+clear to senders why exactly their mail was rejected.  Unfortunately this
+flexibility has a cost: complexity in the logs generated.  While it is easy
+to use standard Unix text processing utilities to determine the fate of an
+individual email, following the journey an email takes through Postfix can
+be quite difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of
+the time the journey is simple, but the other 10\% of the time requires
+90\% of the code.\footnote{These numbers don't have a solid scientific
+basis, they're based on gut feeling from writing and debugging the
+software.}
 
@@
 be found at~\cite{postfix-overview}.  Each log line contains the name of
-the Postfix program which produced the log line, and this information is
-used when determining which rules should be used to parse each log line
-(see section~\refwithpage{rule characteristics} for details.)
+the Postfix program which produced it, and this information is used when
+determining which rules should be used to parse each log line (see
+section~\refwithpage{rule characteristics} for details.)
 
@@
 has more information available when logging the rejection.\footnote{The
-\texttt{smtpd\_delay\_reject} parameter controls this behaviour.}  Postfix
-uses simple lookup tables as the deciding factor for some restrictions,
-e.g.\ \texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\footnote{\texttt{cidr} is the type of the lookup table~\cite{cidr_table}}
+\texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
+this behaviour.}  Postfix uses simple lookup tables as the deciding factor
+for some restrictions, e.g.\
+\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
+\footnote{\texttt{cidr}~\cite{cidr_table} is the type of the lookup table.}
 will check whether the IP address of the connected client matches the left
@@
 continue with the next restriction in the list, allowing more specific
-cases to be handled differently from more general cases.  When the result
-is another restriction Postfix will evaluate the new restriction, allowing
-restrictions to be chosen based on the client IP address, HELO hostname,
-sender address, recipient address, etc.  The administrator can define new
-restrictions as a list of existing restrictions, allowing arbitrarily long
-and complex sequences of lookups and restrictions.
+cases to be used as exceptions to more general cases.  When the result is
+the name of another restriction Postfix will evaluate the new restriction,
+allowing restrictions to be chosen based on the client IP address, HELO
+hostname, sender address, recipient address, etc.  The administrator can
+define new restrictions as a list of existing restrictions, allowing
+arbitrarily long and complex sequences of lookups and restrictions.
 
@@
 \subsubsection{Policy servers}
+
+\label{policy servers}
 
@@
 
-    \item Checking Sender Policy Framework records~\cite{openspf,
-        wikipedia-spf}.  SPF records specify which mail servers are allowed
+    \item Checking \SPF{} records~\cite{openspf,
+        wikipedia-spf}.  \SPF{} records specify which mail servers are allowed
         to send mail claiming to be from a particular domain.  The
@@
         backscatter~\cite{postfix-backscatter} and
-        joe-jobs~\cite{wikipedia-joe-job}; however there has been a lot of
+        joe jobs~\cite{wikipedia-joe-job}; however there has been a lot of
         resistance to the proposal because it breaks or vastly complicates
@@
         rejects mail when the triple of (sender, recipient, remote IP
-        address) is unknown; on the second and subsequent delivery attempts
+        address) is unknown; on second and subsequent delivery attempts
         from that triple the mail will be accepted.  The assumption is that
-        retrying after a temporary failure is uneconomical for a spammer,
-        but that a legitimate mail server must retry.  Sadly spammers are
-        using increasingly complex and well written programs to distribute
-        spam, frequently using an ISP provided \SMTP{} server from a
-        compromised machine on the ISP's network.  Greylisting is slowly
+        maintaining a list of failed addresses and retrying after a
+        temporary failure is uneconomical for a spammer, but that a
+        legitimate mail server must retry.  Sadly spammers are using
+        increasingly complex and well written programs to distribute spam,
+        frequently using an \ISP{} provided \SMTP{} server from a
+        compromised machine on the \ISP{}'s network.  Greylisting is slowly
         becoming less useful, but it does block a large percentage of spam
@@
         system filling up, logging system unable to cope with the volume of
-        logs).
+        logs).  On a well run system it is extremely unlikely that any of
+        these problems will arise, though it is of course possible,
+        particularly when undergoing a deluge of spam.
 
@@
 
-10 other parsers have been reviewed in Appendix~\refwithpage{other-parsers}
-as part of the background research for this project.  None of the reviewed
-parsers perform the type of advanced parsing and log correlation described
-here; all are intended to perform a specific parsing and reporting task,
-rather than be a generic parser, extracting data and leaving generation of
-reports from the data to other programs.  Some parsers save the data
-extracted to a database but the majority discard all data once they have
-finished running, making historical analysis impossible.  The other parsers
-reviewed all produce a report of greater or lesser complexity and detail,
-whereas the program described here doesn't attempt to produce a report at
-all; that responsibility is deferred to a separate program, to be developed
-later.  The parsing algorithm and program described here are designed to
-enable much more detailed log analysis by providing a stable platform for
+Ten other parsers have been reviewed in
+appendix~\refwithpage{other-parsers} as part of the background research for
+this project.  None of the reviewed parsers perform the type of advanced
+parsing and log correlation described here; all are intended to perform a
+specific parsing and reporting task, rather than be a generic parser,
+extracting data and leaving generation of reports from the data to other
+programs.  Some parsers save the data extracted to a data store but the
+majority discard all data once they have finished running, making
+historical analysis impossible.  The other parsers reviewed all produce a
+report of greater or lesser complexity and detail, whereas the program
+described here doesn't attempt to produce a report at all; that
+responsibility is deferred to a separate program, to be developed later.
+The parsing algorithm and program described here are designed to enable
+much more detailed log analysis by providing a stable platform for
 subsequent programs to develop upon.
@@
 log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of Log Mail Analyzer is
-quite different from this parser: it attempts to present correlated log
-data in a form suitable for a systems administrator to search using
-grep(1), awk(1) and the myriad of other standard Unix text processing
-utilities already available, producing a CSV file which can subsequently be
-imported into a MySQL or Berkeley DB database.  The decision to support
-both CSV and Berkeley DB appears to have been a limiting factor:
-Comma-Separated Value is a very simple format where a record is stored in a
-single line, with fields separated by a comma or other punctuation symbol.
-Problems with CSV files include the difficulty in escaping separators,
-providing multiple values for a field (e.g.\ multiple recipients), adding
-new fields, and lack of a bundled schema describing the fields.
-
-Berkeley DB is not an SQL database, supporting only (key, value) pairs; in
-the main table the key is an integer referred to by secondary tables and
-the value used is a CSV line containing all the data for that row.  The
-secondary by-sender, by-recipient, by-date and by-IP tables use the
-sender/recipient/date/IP as the key, and a CSV list of integers referring
-to the main table is the value.  In effect this re-implements SQL foreign
-keys, but without the functionality offered by even the most basic of SQL
-databases (joins, ordering, searches, etc.).  It also requires custom code
-to search on some combination of the above, though the authors of LMA did
-provide some queries: IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.
-Berkeley DB appears to be the least useful of the three output formats
+Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
+different from this parser: it attempts to present correlated log data in a
+form suitable for a systems administrator to search using grep(1), awk(1)
+and the myriad of other standard Unix text processing utilities already
+available, producing a \CSV{} file which can subsequently be imported into
+a MySQL or Berkeley DB database.  The decision to support both \CSV{} and
+Berkeley DB appears to have been a limiting factor: \CSV{} is a very simple
+format where a record is stored in a single line, with fields separated by
+a comma or other punctuation symbol.  Problems with CSV files include the
+difficulty in escaping separators, providing multiple values for a field
+(e.g.\ multiple recipients), adding new fields, and lack of a bundled
+schema describing the fields.
+
+Berkeley DB is not an \SQL{} database, supporting only \textbf{(key,
+value)} pairs; in the main table the key is an integer referred to by
+secondary tables and the value used is a \CSV{} line containing all the
+data for that row.  The secondary by-sender, by-recipient, by-date and
+by-IP tables use the sender/recipient/date/IP as the key, and a \CSV{} list
+of integers referring to the main table is the value.  In effect this
+re-implements \SQL{} foreign keys, but without the functionality offered by
+even the most basic of \SQL{} databases (joins, ordering, searches, etc.).
+It also requires custom code to search on some combination of the above,
+though the authors of \LMA{} did provide some queries: IP-STORY,
+FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the
+least useful of the three output formats
 
@@
 possible to discover the schema with an SQL database, unlike with Berkeley
-DB\@.
+DB\@; all \SQL{} databases embed the database schema into the database and
+provide commands for displaying it.  Berkeley DB does not embed a schema,
+as there is neither requirement nor benefit; it provides \textbf{(key,
+value)} pairs, so any additional structuring of the data is imposed by the
+application, thus it behoves the application to document this structure.
 
@@
 
-Parsing in LMA is hand-coded and requires major changes to the code to
+Parsing in \LMA{} is hand-coded and requires major changes to the code to
 parse new log lines or extract additional data.  It does not appear to deal
@@
 
-LMA does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL and
-DAILY-REJECT); the author was unable to parse any logs with LMA, so didn't
-have an opportunity to review the reports.  \parsername{} does not have any
-bundled reports, it delegates that task to a separate, dedicated program.
-
-In summary Log Mail Analyser appears to be a proof of concept, written to
-demonstrate the point of the paper (that having this information in an
-accessible fashion is useful to systems administrators), rather than a
-program designed to be extensible and useful in a production environment.
+\LMA{} does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL
+and DAILY-REJECT); the author was unable to parse any logs with \LMA{}, so
+didn't have an opportunity to review the reports.  \parsername{} does not
+have any bundled reports, it delegates that task to a separate, dedicated
+program.
+
+In summary \LMA{} appears to be a proof of concept, written to demonstrate
+the point of the paper (that having this information in an accessible
+fashion is useful to systems administrators), rather than a program
+designed to be extensible and useful in a production environment.
 
@@
 
-    \item Using a database as an application programming interface.
-
-    \item The \SMTP{} protocol.
+    \item Using a database as an \API{}.
+
+    \item \SMTP{}.
 
@@
 
+\glossary{name={SQLite3},description={
+    \textit{SQLite is a small C library that implements a self-contained,
+    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
+    \SQL{} implementation focusing on correctness, simplicity and speed.
+    Unlike other \SQL{} implementations it does not require a separate
+    server process, greatly simplifying deployment of an application
+    utilising it.  More details can be found at~\cite{sqlite-features} or
+    \url{http://www.sqlite.org/}.
+}}
+
+\glossary{name={Phishing},description={
+    Phishing~\cite{wikipedia-phishing} is an attempt to acquire information
+    by masquerading as an entity trusted by the user, e.g.\ a bank.
+}}
+
+\glossary{name={Backscatter},description={
+    When a spammer or worm sends mail with forged sender addresses,
+    innocent sites are flooded with undeliverable mail notifications; this
+    is called backscatter mail.
+}}
+
+\glossary{name={Joe job},description={
+    A joe job is when spam mail is sent with a faked sender address with
+    the intention of sullying the good name of the owner of the address.
+    Joe jobs are a cause of backscatter, though by no means the only cause.
+}}
+
 \end{document}

### 2007-11-07 doc/logparser-acronyms.tex
More corrections and improvements.

@@
 }}
+
+\newacronym{FQDN}{Fully Qualified Domain Name}{description={
+    An FQDN is a hostname plus domain name, e.g.
+        example.com     is a domain name    \newline
+        www             is a hostname       \newline
+        www.example.com is a FQDN
+}}
+
+\newacronym{DFA}{Deterministic Finite Automata}{description={
+    XXX EXPLAIN DFA\@.  SHOULD IT BE NFA\@?
+}}
+
+\newacronym{DNS}{Domain Name System}{description={
+    The DNS converts between hostnames (www.example.com) and IP addresses
+    (10.1.2.3).
+}}

### 2007-11-07 doc/logparser.tex
More corrections and improvements.

@@
         frequently using an \ISP{} provided \SMTP{} server from a
-        compromised machine on the \ISP{}'s network.  Greylisting is slowly
-        becoming less useful, but it does block a large percentage of spam
-        mail at the moment.
+        compromised machine on the \ISP{}'s network.  Greylisting will
+        slowly become less useful, but it does block a large percentage of
+        spam mail at the moment; the most effective restrictions over the
+        93 log files used in testing the parser are shown in
+        table~\refwithpage{Summary of rejections}.  Greylisting is
+        obviously worth using, at least at the moment, particularly when
+        you factor in Greylisting's position as the final restriction which
+        a mail must overcome: Greylisting only takes effect for mails which
+        have passed every other restriction.
+
+        \begin{table}[ht]
+            \caption{Summary of rejections}\label{Summary of rejections}
+            \begin{tabular}[]{lr}
+                Restriction type    & Total rejections  \\
+                Total rejections    & 5283739           \\
+                Unknown recipient   & 2338000           \\
+                Unknown sender      & 864148            \\
+                \RBL{}s             & 807185            \\
+                Greylisting         & 497244            \\
+                Non-\FQDN{} HELO    & 403652            \\
+            \end{tabular}
+        \end{table}
 
@@
 The database is an integral part of the parser: it stores the rules and the
-results of applying those rules to Postfix log files.  Understanding the
-database schema is important in understanding the actions of the parser,
-and essential to developing further applications which utilise the data
-gathered.
+data gleaned by applying those rules to Postfix log files.  Understanding
+the database schema is important in understanding the actions of the
+parser, and essential to developing further applications which utilise the
+data gathered.
 
@@
 described in detail in section~\refwithpage{rules} but the fields are
-covered in section~\refwithpage{rule attributes}.
+covered here.
 
@@
 
-        \begin{itemize}
+        \begin{description}
 
@@
 
-        \end{itemize}
+        \end{description}
 
@@
 
-    \item [regex] The regular expression\cite{wikipedia-regex, perlre} to
+    \item [regex] The regular expression~\cite{wikipedia-regex, perlre} to
         match the log line against.  The regex will first have several
@@
         \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \hspace*{2em} server\_hostname and helo.\newline}
+        \newline \hspace*{2em} server\_hostname} and \texttt{helo.\newline}
         \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \hspace*{2em} pid\_regex, data and child.}
+        \newline \hspace*{2em} pid\_regex, data} and \texttt{child.}
 
@@
         cannot be used.  This is intended for use with small amounts of
-        data rather than large, so dealing with escape sequences seemed
-        unnecessary.} is permitted on the right hand side of the
+        data rather than large amounts, so dealing with escape sequences
+        seemed unnecessary.} is permitted on the right hand side of the
         assignment.
@@
     \item [server\_hostname] The hostname of the server, it will be
-        \textit{unknown\/} if the IP address could not be resolved to a
+        \texttt{unknown} if the IP address could not be resolved to a
         hostname via DNS\@.
@@
     \item [client\_hostname] The hostname of the client, it will be
-        \textit{unknown\/} if the IP address could not be resolved to a
+        \texttt{unknown} if the IP address could not be resolved to a
         hostname via DNS\@.
@@
         logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.
-
-    \item [queueid] The queueid of the mail if the conection represents an
-        accepted mail, or \textit{NOQUEUE\/} otherwise.
+        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
+        specifying a HELO-based restriction in
+        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
+        correctly.
+
+    \item [queueid] The queueid of the mail if the connection represents an
+        accepted mail, or \texttt{NOQUEUE} otherwise.
 
@@
 
-    \item [connection\_id] A reference to the entry in the connections
-        table this results table entry belongs to.
+    \item [connection\_id] A reference to the row in the connections table
+        this result's table row belongs to.
 
@@
 
-    \item [message\_id] The message-id of the accepted mail, or NULL if no
-        mail was accepted.
+    \item [message\_id] The message-id of the accepted mail, or
+        \texttt{NULL} if no mail was accepted.
 
@@
 
-The table containing the rules used by the parser, and the two tables
+The table containing the rules used by the parser, and both tables
 containing the data extracted from the Postfix logs were described, with
 the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential in using the extracted data; it's more important for
-using the data than storing the data, because storing the data is a
+schema is essential when using the extracted data; it's more important when
+using the data than when storing it, because storing the data is a
 write-once operation, whereas utilising the data requires frequent
@@
 
-\section{Adaptable parsing: user defined rules}
+\section{Parsing rules}
 
@@
 
-The parser is split into two parts: the parsing algorithm and the rules
-which are applied to the lines.  The complexity and variation in Postfix's
-logs requires similar flexibility in the parser; decoupling the parsing
-rules from the associated actions allows new rules to be written and tested
+The complexity and variation in Postfix's logs requires similar flexibility
+in the parser, which is split into two parts: the parsing algorithm and the
+parsing rules which are applied to the lines.  Decoupling the parsing rules
+from the associated actions allows new rules to be written and tested
 without requiring modifications to the algorithm source code (significantly
-lowering the barrier to entry for new or casual users), and greatly
-simplifies both algorithm and rules.  It also creates a clear separation of
-functionality: rules handle low level details of identifying log lines and
-extracting data from a line, whereas the algorithm handles the higher level
-details of following the path a mail takes through Postfix, assembling the
-required data, etc.
+lowering the barrier to entry for new or casual users who need to parse new
+log lines), and greatly simplifies both algorithm and rules.  Decoupling
+also creates a clear separation of functionality: rules handle low level
+details of identifying log lines and extracting data from a line, whereas
+the algorithm handles the higher level details of following the path a mail
+takes through Postfix, assembling the required data before storing it,
+dealing with complications arising, etc.
 
@@
         there are also rules specific to that program, and those rules must
-        already have been tried and failed.}  Any given rule will only be
-        used to parse a subset of the log lines, and any given log line
-        will only be parsed by a subset of the rules.
+        already have been tried and failed on the current line.}  Any given
+        rule will only be used to parse a subset of the log lines, and any
+        given log line will only be parsed by a subset of the rules.
 
@@
 The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's regex, generate an NFA, and compare all NFAs
-to determine if any overlap.  The author has not attempted to do this: such
-a project would probably qualify for a PhD, and may involve solving the
-Halting Problem and circumventing the Church-Turing Thesis.} that
+possible to parse each rule's regex, generate a \DFA{} from it, and examine
+every \DFA{} to determine if any overlap.  The author has not attempted to
+do this: such a project by itself would probably qualify for a PhD, and may
+involve solving the Halting Problem and circumventing the Church-Turing
+Thesis.  XXX REFERENCES FOR HALTING PROBLEM AND CHURCH-TURING\@.} that
 responsibility is left to the author of the rules.  Unintentionally
 overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line determines which rule
-matches.  Overlapping rules are frequently a requirement, allowing a more
-specific rule to match some lines and a more general rule to match the
-majority, e.g.\ separating \SMTP{} delivery to a filtering proxy from
-\SMTP{} delivery to the rest of the world.  The algorithm provides a
-facility for ordering overlapping rules: the priority field in each rule
-(defaults to zero).  Rules are tried in order of priority, highest first,
-overriding the hits field maintained by the parser.  Negative priorities
-may also be useful for catchall rules.
+the order in which rules are tried against each line is unspecified, and
+the first matching rule wins.  Overlapping rules are frequently a
+requirement, allowing a more specific rule to match some lines and a more
+general rule to match the majority, e.g.\ separating \SMTP{} delivery to a
+specific sites from \SMTP{} delivery to the rest of the world.  The
+algorithm provides a facility for ordering overlapping rules: the priority
+field in each rule (defaults to zero).  Rules are sorted by priority,
+highest first, and then rules with the same priority are sorted by the
+number of successful matches when parsing the previous log file.  Negative
+priorities may be useful for catchall rules.
 
@@
         no overlapping rules the tables from each run will be identical;
-        differences indicate overlapping rules, and the rule\_ids which
-        differ determine which rules overlap.  Unfortunately this method
-        cannot prove the absence of overlapping rules; it can detect
-        overlapping rules --- but only if there are log lines in the input
-        files which match more than one rule.
+        differences indicate overlapping rules.  Which rules overlap can be
+        determined by examining the differences in the tables: each result
+        contains the id of the rule which created it, so the rule
+        referenced in the normal table overlaps with the rule referenced in
+        the reversed table.  Unfortunately this method cannot prove the
+        absence of overlapping rules; it can detect overlapping rules, but
+        only if there are log lines in the input files which match more
+        than one rule.
 
@@
 This example rule matches the message logged by Postfix when it rejects
-mail from a sender address where the domain has neither an MX record nor an
-A record, i.e.\ mail could not be delivered to the sender's address --- for
-full details see~\cite{reject-unknown-sender-domain}.
+mail from a sender address because the appropriate \DNS{} entries are
+missing, i.e.\ mail could not be delivered to the sender's address (for
+full details see~\cite{reject-unknown-sender-domain}).
 
@@
     \item [program and regex:] If the program in the rule equals the
-        program which logged the line the regex will be matched against the
-        line; if the match is successful the action will be executed, if not
-        the next rule will be tried.  If the program-specific rules don't
-        match the log line, the generic rules will be tried in the same
-        way.
+        program which logged the line the regex will be tried against the
+        line; if the match is successful the action will be executed, if
+        not the next rule will be tried.  If the program-specific rules
+        don't match the log line, the generic rules will be tried in the
+        same way.
 
@@
 
-    \item [queueid:] is used the extract the queueid matched by the regex,
-        so that the correct mail can be found by queueid and actions
+    \item [queueid:] The index of the field in the regex which captured the
+        queueid, or zero if the line does not contain a queueid.  This
+        allows the correct mail can be found by queueid and actions
         performed on it.
@@
         ordering the rules (see section~\refwithpage{rule ordering for
-        efficiency}); hits\_total is updated at the end of each run but is
-        otherwise unused.
+        efficiency}); hits is set to the number of successful matches at
+        the end of the parsing run, and hits\_total has hits added to it,
+        but is otherwise unused by the algorithm.
 
@@
 daily basis).  When generating the data for the graphs included in
-section~\refwithpage{graphs}, 93 log files (totaling 10.08 GB, 60.72
-million lines) were each parsed 10 times, the first run discarded and the
+appendix~\refwithpage{graphs}, 93 log files (totaling 10.08 GB, 60.72
+million lines) were each parsed 10 times, the first run discarded, and the
 remaining 9 runs averaged.  The first run is discarded for two reasons:
@@
 of the log lines, and the remaining log lines split across the other 90\%
-of the rules (as shown in graph~\refwithpage{rule hits}).  Assuming that
-the distribution of log lines is reasonably steady over time, program
+of the rules (as shown in graph~\refwithpage{rule hits graph}).  Assuming
+that the distribution of log lines is reasonably steady over time, program
 efficiency should benefit from trying more frequently matching rules before
@@
 percentage increase of execution times (mean and standard deviation are
-shown after the graph).  Overall this provides a modest but worthwhile
+shown in table~\refwithpage{Regex caching/discarding with different groups
+of log files}).  Overall this provides a modest but worthwhile
 performance increase.
@@
 Caching the compiled regexs is quite simple, and is the single most
-effective optimisation tried in the parser.
+effective optimisation implemented in the parser.
 
@@
 
-    \item An example log line and the rule matching it was given, followed
-        by a description of how the fields in the rule are used both in the
-        matching and the subsequent action that is executed.
+    \item An example log line and the rule matching it illustrated a
+        description of how the fields in the rule are used both in the
+        matching phase and the subsequent action that is executed.
 
@@
         covering the effects of caching compiled regexs and optimal
-        ordering ordering of rules, with reference to the graphs in
-        section~\refwithpage{graphs}.
+        ordering of rules, with reference to the graphs in
+        appendix~\refwithpage{graphs}.
 
@@
 
-    \item A flow chart showing the paths a mail can take through the
+    \item A flow chart showing common paths a mail can take through the
         algorithm.
 
-    \item A description of the full algorithm.
-
-    \item The actions which can be taken by the parser are described in
+    \item An explanation of the paths shown in the flow chart.
+
+    \item The actions the parser makes available to rules are covered in
         detail.
@@
     \item Additional complications which have arisen during the development
-        of this parser are described.
+        of this parser are documented, including their solutions and where
+        those solutions are implemented in the algorithm
 
@@
 
+\glossary{name={Epoch},description={
+    Most operating systems store the current time and timestamps of files
+    etc.\ as seconds elapsed since the epoch, the beginning of time as far
+    as the operating system is concerned.  On Unix and Unix-derived systems
+    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
+    different.
+}}
+
+\glossary{name={NULL},description={
+    NULL is a special term used in \SQL{} databases indicating the absence
+    of data for the field.
+}}
+
 \end{document}

### 2007-11-11 doc/logparser.tex
More corrections. Capitalised Postfix in the correct places. Adopted a standard formatting for postfix/foo daemons.

@@
     \empty{}\ref{#1} [page~\pageref{#1}]%
+}
+
+% A command to format a Postfix daemon's name
+\newcommand{\daemon}[1]{%
+    \texttt{postfix/#1}%
 }
@@
 
-    \item [program] The program (\texttt{postfix/smtpd},
-        \texttt{postfix/qmgr}, etc.) whose log lines the rule applies to.
-        This avoids needlessly trying rules which won't match the line, or
-        worse, might match unintentionally.  Rules whose program is
-        \texttt{*} will be tried against any lines which aren't parsed by
-        program specific rules.
+    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
+        log lines the rule applies to.  This avoids needlessly trying rules
+        which won't match the line, or worse, might match unintentionally.
+        Rules whose program is \texttt{*} will be tried against any lines
+        which aren't parsed by program specific rules.
 
@@
 postfix\_action     & REJECTED                                          \\
-program             & postfix/smtpd                                     \\
+program             & \daemon{smtpd}                                    \\
 regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
@@
 
-    \item Mail enters the system via \SMTP{} or local submission.
+    \item Mail enters the system via \SMTP{} or local submission; and new
+        data structure is created for it.
 
@@
     \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, log all data, and finish.
+        delivered, bounced or deleted, then log all data, and finish.
 
@@
 addressed mails are not always delivered directly to a mailbox: sometimes
-they are accepted for a local address but need to be delivered to one or
-more remote addresses due to aliases.  When this occurs a child mail will
-be injected into the postfix queue, but without the explicit logging smtpd
-or sendmail injected mails have.  Thus the source is not immediately
-discernible from the log line in which the mail first appears; from a
-strictly chronological reading of the logs it \textit{usually\/} appears as
-if the child mail has appeared from thin air.  Subsequently the parent mail
-will log the creation of the child mail:
+they are addressed to and accepted for a local address but need to be
+delivered to one or more remote addresses due to aliases.  When this occurs
+a child mail will be injected into the Postfix queue, but without the
+explicit logging smtpd or postdrop injected mails have.  Thus the source is
+not immediately discernible from the log line in which the mail first
+appears; from a strictly chronological reading of the logs it
+\textit{usually\/} appears as if the child mail has appeared from thin air.
+Subsequently the parent mail will log the creation of the child mail:
 
@@
 qmgr is.\footnote{Postfix is quite paranoid about mail delivery, an
-excellent characteristic for an MTA to possess, so it won't log that the
+excellent characteristic for an \MTA{} to possess, so it won't log that the
 child has been created until it is absolutely certain that the mail has
@@
 of where mails originate from, but the author believes that it is better to
-require an explicit source, as bugs are more likely to be exposed.
+require an explicit source, as bugs in the parser are more likely to be
+exposed.
 
@@
 that data from their parent before being entered in the database.  The
-effect of this complication upon the algorithm is discussed in
+effect of this complication upon the algorithm is discussed more fully in
 section~\refwithpage{tracking re-injected mail}.
+
+XXX FIGURE OUT IF I NEED TO COPY DATA ANY MORE, AND DOCUMENT IT IF SO\@.
 
@@
 Section~\refwithpage{actions-in-detail} explains the actions in substantive
-detail; this section will omit such detail because it would clutter and
-confuse the algorithm description.  The data flow chart in
+detail; this section omits such detail because it would clutter and confuse
+the algorithm description.  The flow chart in
 section~\refwithpage{flow-chart} should also be consulted while reading
-this section.  For the sake of clarity this description covers the most
-common paths through the algorithm, as including every path would hinder
-understanding.  The more uncommon paths are caused by the complications
-described in section~\refwithpage{additional complications}, and are
-covered therein.
+this section.  For the sake of clarity this description covers only the
+most common paths through the algorithm, as including every path would
+hinder understanding; the more uncommon paths are caused by the
+complications described in section~\refwithpage{additional complications},
+and are covered therein.
 
@@
 Everything starts off with a mail entering the system, whether by local
-submission via postdrop/sendmail, by \SMTP{}, by re-injection due to
-forwarding, or internally generated by Postfix.  Local submission is the
-simplest case: a queueid is assigned immediately and the sender address is
-logged (action: pickup; flowchart:~2).
+submission via postdrop/sendmail,\footnote{Postfix provides a command that
+is compatible with the Sendmail (\url{http://www.sendmail.org/ XXX CHECK})
+mail submission program which all Unix commands which send mail depend on.}
+by \SMTP{}, by re-injection due to forwarding, or internally generated by
+Postfix.  Local submission is the simplest case: a queueid is assigned
+immediately and the sender address is logged (action: pickup;
+flowchart:~2).
 
@@
 The obvious counterpart to mail entering the system is mail leaving the
-system, whether it is by deletion, bouncing, or local or remote delivery.
+system, whether by deletion, bouncing, local delivery, or remote delivery.
 All four are handled in exactly the same way:
@@
 
-    \item The sender and recipient addresses will be logged separately
+    \item Postfix will log the sender and recipient addresses separately
         (action: save\_by\_queueid; flowchart:~10).
@@
 deleted by the administrator.  The exception is what happens after delivery
-to the parent or children of mail re-injected due to forwarding, and that
-is explained in section~\refwithpage{tracking re-injected mail}.
+to the parent or children of mail re-injected due to forwarding, as
+explained in section~\refwithpage{tracking re-injected mail}.
 
@@
     \item The indicator which more commonly introduces re-injection is when
-        \texttt{postfix/qmgr} selects a mail with a previously unseen
+        \daemon{qmgr} selects a mail with a previously unseen
         queueid for delivery (action: mail\_picked\_for\_delivery;
@@
         destination for.} In this case the mail may already have been
-        created (action: qmgr\_chooses\_mail; flowchart:~3; described
-        above) and the unknown origin flag will be cleared; if not a new
-        data structure will be created.  In both cases the new mail is
-        marked as a child of the parent.  The log line in question is:
+        created (described above) and the unknown origin flag will be
+        cleared; if not a new data structure will be created.  In both
+        cases the new mail is marked as a child of the parent.  The log
+        line in question is:
 
@@
         This second indicator always occurs for re-injected mail but
-        typically after the first indicator explained above.  This
+        typically occurs after the first indicator explained above.  This
         indicator is required to tie the parent and child mails together
@@
 
-    \item [REJECTION] Deal with postfix rejecting an \SMTP{} command from
+    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
         the remote client: log the rejection with a mail if there is a
@@
 occurrences in the test log files have the same characteristics: the
-\texttt{postfix/local} log line showing delivery to a local mailbox occurs
-after the \texttt{postfix/qmgr} log line showing removal of the mail from
-the queue because delivery is completed.  This causes problems: the mail is
-not complete, so entry into the database fails; a new mail is created when
-the \texttt{postfix/local} line is parsed and remains in the state tables;
-four warnings are issued per pair of out of order log lines.
+\daemon{local} log line showing delivery to a local mailbox occurs after
+the \daemon{qmgr} log line showing removal of the mail from the queue
+because delivery is completed.  This causes problems: the mail is not
+complete, so entry into the database fails; a new mail is created when the
+\daemon{local} line is parsed and remains in the state tables; four
+warnings are issued per pair of out of order log lines.
 
@@
 The list of valid combinations is given below.  Every mail will
-additionally have log entries from \texttt{postfix/cleanup} and
-\texttt{postfix/qmgr}; any mail may also have log entries from
-\texttt{postfix/bounce}, \texttt{postfix/postsuper}, or both.
+additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
+mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
+both.
 
@@
 
-    \item [postfix/local:] Local delivery of bounce notification, or local
+    \item [\daemon{local}:] Local delivery of bounce notification, or local
         delivery of forwarded/tracked mail.
 
-    \item [postfix/local, postfix/pickup:] Locally submitted mail, locally
-        delivered.
-
-    \item [postfix/local, postfix/pickup, postfix/smtp:] Locally submitted
-        mail, \newline both local and remote delivery.
-
-    \item [postfix/local, postfix/smtp, postfix/smtpd:] Mail accepted from
-        a remote client, both local and remote delivery.
-
-    \item [postfix/local, postfix/smtpd:] Mail accepted from a remote
+    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
+        server, delivered locally on the server.
+
+    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
+        submitted mail, \newline both local and remote delivery.
+
+    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
+        from a remote client, both local and remote delivery.
+
+    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
         client, local delivery.
 
-    \item [postfix/pickup, postfix/smtp:] Locally submitted mail, remote
+    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
         delivery.
 
-    \item [postfix/smtp:] Remote delivery of forwarded/tracked mail.
-
-    \item [postfix/smtp, postfix/smtpd:] Mail accepted from a remote
-        client, remotely delivered (relaying mail for clients on the local
-        network).
+    \item [\daemon{smtp}:] Remote delivery of forwarded/tracked mail.
+
+    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
+        client, then remotely delivered (typically relaying mail for
+        clients on the local network).
 
@@
 section~\refwithpage{aborted-delivery-attempts}, except that there
-\textit{is\/} a \texttt{postfix/cleanup} log line; there does not appear to
-be anything in the log file to explain why there are no further log
-messages.  The only way to detect these mails is to periodically scan all
-mails in the state tables, deleting any mails displaying the following
-characteristics:
+\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
+anything in the log file to explain why there are no further log messages.
+The only way to detect these mails is to periodically scan all mails in the
+state tables, deleting any mails displaying the following characteristics:
 
@@
 
-    \item There must be exactly two \texttt{postfix/smtpd} and one
-        \texttt{postfix/cleanup} log entries for the mail, with no
-        additional log entries.
+    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
+        log entries for the mail, with no additional log entries.
 
@@
         envelope sender address is username@domain.  Log lines are added by
-        \textit{postfix/local}, \textit{postfix/qmgr\/} (twice),
-        \textit{postfix/cleanup\/} and finally \textit{postfix/pickup}.
+        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
+        \daemon{pickup}.
 
@@
         the original sender: username@domain.  Log lines are added by
-        \textit{postfix/local}, \textit{postfix/qmgr\/} (twice) and
-        \textit{postfix/cleanup}.
+        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
 
@@
         forwarding instructions.  A \texttt{Delivered-To:~username@domain}
-        header is added to the forwarded bounce notification, and the
-        envelope sender address is now username@domain.  Log lines are
-        added by \textit{postfix/local}, \textit{postfix/qmgr\/} (twice),
-        \textit{postfix/cleanup}, and finally \textit{postfix/pickup}.
+        header is added to the forwarded bounce notification, which now has
+        an envelope sender address of username@domain.  Log lines are added
+        by \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup}, and
+        finally \daemon{pickup}.
 
     \item Postfix accepts the forwarded bounce notification but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address currently being delivered to, and sends a
-        bounce notification to the sender: username@domain.  Log lines are
-        added by \textit{postfix/local}, \textit{postfix/qmgr\/} (twice)
-        and \textit{postfix/cleanup}.
+        delivering the mail it notices that the \texttt{Delivered-To}
+        header already contains the address currently being delivered to,
+        and sends a bounce notification to the sender: username@domain.
+        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
+        \daemon{cleanup}.
 
@@
 components which have a small number of rules associated with them, whereas
-in general the most common log lines are logged by \textit{postfix/smtpd},
-which also has the highest number of rules.  \textit{postfix/smtpd\/} log
-lines are distributed across rules much more evenly that
-\textit{postfix/qmgr}, \textit{postfix/local}, \textit{postfix/cleanup\/}
-or \textit{postfix/pickup}, so the average number of rules required to
-parse a \textit{postfix/smtpd\/} log line is much higher that the average
-number required to parse other log lines.  Thus the average number of rules
-required to parse a random line from logs containing a mail loop is much
-lower than the average number required to parse a random log line from logs
-without a mail loop, as can be seen in the peaks in
+in general the most common log lines are logged by \daemon{smtpd}, which
+also has the highest number of rules.  \daemon{smtpd} log lines are
+distributed across rules much more evenly that \daemon{qmgr},
+\daemon{local}, \daemon{cleanup} or \daemon{pickup}, so the average number
+of rules required to parse a \daemon{smtpd} log line is much higher that
+the average number required to parse other log lines.  Thus the average
+number of rules required to parse a random line from logs containing a mail
+loop is much lower than the average number required to parse a random log
+line from logs without a mail loop, as can be seen in the peaks in
 graphs~\refwithpage{execution time vs file size vs number lines

### 2007-11-11 doc/logparser-acronyms.tex
More acronyms. Wrapped all Postfix programs except sendmail in \daemon{}. Added glossary entries for all Postfix programs, some need to be fleshed out. Some corrections. More references.

@@
 \newacronym{MTA}{Mail Transfer Agent}{description={
-    A Mail Transfer Agent sends and/or receives mail via SMTP\@.
+    A Mail Transfer Agent sends and/or receives mail via SMTP.
 }}
@@
     The Request For Comments series is a series of proposals defining
-    various protocols and file formats, e.g. SMTP\@.  The name is somewhat
+    various protocols and file formats, e.g. SMTP.  The name is somewhat
     misleading: initially the authors were asking for peer review, but
@@
 \newacronym{DFA}{Deterministic Finite Automata}{description={
-    XXX EXPLAIN DFA\@.  SHOULD IT BE NFA\@?
+    XXX EXPLAIN DFA.  SHOULD IT BE NFA?
 }}
@@
 }}
+
+\newacronym{pid}{Process IDentifier}{description={
+    There may be multiple copies of any program executing at any one time,
+    so the program's name is not suitable as a distinguishing identifier;
+    instead each process executing is given a pid which is guaranteed to be
+    unique for the lifetime of the process.  Once the process has
+    completed, the pid may be reused, as they are drawn from a finite pool.
+}}
+
+\newacronym{UCE}{Unsolicited Commercial Email}{description={
+    UCE is a more restrictive definition of spam than most people would
+    use: it only covers mail that is explicitly commercial, thus excluding
+    viruses, Bayesian poisoning mails, backscatter, and those annoying
+    chain letters you get from friends.
+}}
+
+\newacronym{regex}{Regular Expression}{description={
+    Regular expressions are a method of matching patterns in text that are
+    explained in detail in~\cite{wikipedia-regex, perlre} .
+}}
+
+\newacronym{regexs}{Regular Expressions}{description={
+    Regular expressions are a method of matching patterns in text that are
+    explained in detail in~\cite{wikipedia-regex, perlre} .
+}}

### 2007-11-11 doc/logparser.tex
More acronyms. Wrapped all Postfix programs except sendmail in \daemon{}. Added glossary entries for all Postfix programs, some need to be fleshed out. Some corrections. More references.

@@
     \texttt{postfix/#1}%
+}
+
+% This is ridiculous, but I can't put @ in glossary entries, so . . .
+\newcommand{\at}[0]{%
+    @%
 }
@@
 
-    \item Checking \SPF{} records~\cite{openspf,
-        wikipedia-spf}.  \SPF{} records specify which mail servers are allowed
-        to send mail claiming to be from a particular domain.  The
-        intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe jobs~\cite{wikipedia-joe-job}; however there has been a lot of
+    \item Checking \SPF{} records~\cite{openspf, wikipedia-spf}.  \SPF{}
+        records specify which mail servers are allowed to send mail
+        claiming to be from a particular domain.  The intention is to
+        reduce spam from faked sender addresses,
+        backscatter~\cite{postfix-backscatter} and joe
+        jobs~\cite{wikipedia-joe-job}; however there has been a lot of
         resistance to the proposal because it breaks or vastly complicates
@@
 
-    \item [regex] The regular expression~\cite{wikipedia-regex, perlre} to
-        match the log line against.  The regex will first have several
-        keywords expanded: this simplifies reading and writing rules;
-        avoids needless repetition of complex regex components; allows the
-        components to be corrected and/or improved in one location; and
-        makes each regex largely self-documenting.
+    \item [regex] The \regex{} to match the log line against.  The \regex{}
+        will first have several keywords expanded: this simplifies reading
+        and writing rules; avoids needless repetition of complex \regex{}
+        components; allows the components to be corrected and/or improved
+        in one location; and makes each \regex{} largely self-documenting.
 
@@
 
-        For efficiency the keywords are expanded and every rule's regex is
-        compiled before attempting to parse the log file --- otherwise each
-        regex would be recompiled each time it was used, resulting in a
-        large, data dependent slowdown.  Rule efficiency concerns are
+        For efficiency the keywords are expanded and every rule's \regex{}
+        is compiled before attempting to parse the log file --- otherwise
+        each \regex{} would be recompiled each time it was used, resulting
+        in a large, data dependent slowdown.  Rule efficiency concerns are
         discussed in section~\refwithpage{rule efficiency}.
@@
         semi-colon or comma separated assignment statements, with the
-        variable name on the left and the matching field from the regex on
-        the right hand side.  The list of acceptable variable names is:
+        variable name on the left and the matching field from the \regex{}
+        on the right hand side.  The list of acceptable variable names is:
 
@@
 
-    \item [queueid] Specifies the matching field from the regex which gives
-        the queueid, or zero if the log line doesn't contain a queueid.
+    \item [queueid] Specifies the matching field from the \regex{} which
+        gives the queueid, or zero if the log line doesn't contain a
+        queueid.
 
@@
 The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's regex, generate a \DFA{} from it, and examine
-every \DFA{} to determine if any overlap.  The author has not attempted to
-do this: such a project by itself would probably qualify for a PhD, and may
-involve solving the Halting Problem and circumventing the Church-Turing
-Thesis.  XXX REFERENCES FOR HALTING PROBLEM AND CHURCH-TURING\@.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line is unspecified, and
-the first matching rule wins.  Overlapping rules are frequently a
-requirement, allowing a more specific rule to match some lines and a more
-general rule to match the majority, e.g.\ separating \SMTP{} delivery to a
-specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
+possible to parse each rule's \regex{}, generate a \DFA{} from it, and
+examine every \DFA{} to determine if any overlap.  The author has not
+attempted to do this: such a project by itself would probably qualify for a
+PhD, and may involve solving the Halting Problem and circumventing the
+Church-Turing Thesis.  XXX REFERENCES FOR HALTING PROBLEM AND
+CHURCH-TURING\@.} that responsibility is left to the author of the rules.
+Unintentionally overlapping rules lead to inconsistent parsing and data
+extraction because the order in which rules are tried against each line is
+unspecified, and the first matching rule wins.  Overlapping rules are
+frequently a requirement, allowing a more specific rule to match some lines
+and a more general rule to match the majority, e.g.\ separating \SMTP{}
+delivery to a specific sites from \SMTP{} delivery to the rest of the
+world.  The algorithm provides a facility for ordering overlapping rules:
+the priority field in each rule (defaults to zero).  Rules are sorted by
+priority, highest first, and then rules with the same priority are sorted
+by the number of successful matches when parsing the previous log file.
+Negative priorities may be useful for catchall rules.
 
@@
 
-    \item Sort the regexs and visually inspect the list, e.g.\ with SQL
+    \item Sort by regex and visually inspect the list, e.g.\ with SQL
         similar to: \textbf{select regex from rules order by regex;}
@@
     \item [program and regex:] If the program in the rule equals the
-        program which logged the line the regex will be tried against the
-        line; if the match is successful the action will be executed, if
-        not the next rule will be tried.  If the program-specific rules
+        program which logged the line the \regex{} will be tried against
+        the line; if the match is successful the action will be executed,
+        if not the next rule will be tried.  If the program-specific rules
         don't match the log line, the generic rules will be tried in the
@@
 
-    \item [action:] will be executed if the regex matches successfully (see
-        section~\refwithpage{actions-in-detail} for full details).
+    \item [action:] will be executed if the \regex{} matches successfully
+        (see section~\refwithpage{actions-in-detail} for full details).
 
@@
 
-    \item [queueid:] The index of the field in the regex which captured the
-        queueid, or zero if the line does not contain a queueid.  This
+    \item [queueid:] The index of the field in the \regex{} which captured
+        the queueid, or zero if the line does not contain a queueid.  This
         allows the correct mail can be found by queueid and actions
@@
 
-Perl compiles the original regex into an internal representation,
-optimising the regex to improve the speed of matching, but this compilation
-and optimisation takes CPU time.  Perl automatically caches static regexs,
-but dynamic regexs need to be explicitly compiled and cached.
-Graph~\refwithpage{normal regex vs discard regex} shows execution times
-with and without caching the regex.  Caching the compiled regexs is
-obviously far more efficient; graph~\refwithpage{normal regex vs discarded
-regex factor} shows the percentage execution time increase when not caching
-each regex.
-
-Caching the compiled regexs is quite simple, and is the single most
+Perl compiles the original \regex{} into an internal representation,
+optimising the \regex{} to improve the speed of matching, but this
+compilation and optimisation takes CPU time.  Perl automatically caches
+static \regexs{}, but dynamic \regexs{} need to be explicitly compiled and
+cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
+times with and without caching the regex.  Caching the compiled \regexs{}
+is obviously far more efficient; graph~\refwithpage{normal regex vs
+discarded regex factor} shows the percentage execution time increase when
+not caching each regex.
+
+Caching the compiled \regexs{} is quite simple, and is the single most
 effective optimisation implemented in the parser.
@@
     \item The final part of this section discusses rule efficiency,
-        covering the effects of caching compiled regexs and optimal
+        covering the effects of caching compiled \regexs{} and optimal
         ordering of rules, with reference to the graphs in
@@
 The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the smtpd pid, then transition to being correlated
-by the queueid.  This is relatively minor, but does require:
+first be correlated by the \daemon{smtpd} \pid{}, then transition to being
+correlated by the queueid.  This is relatively minor, but does require:
 
@@
     \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by pid or queueid.
+        needs to lookup the data structures by \pid{} or queueid.
 
@@
 one mail. Happily once the mail has been accepted log entries won't be
-correlated by pid for that mail any more (its queueid will be used
+correlated by \pid{} for that mail any more (its queueid will be used
 instead), so there isn't any ambiguity about which mail a given log line
@@
 a child mail will be injected into the Postfix queue, but without the
-explicit logging smtpd or postdrop injected mails have.  Thus the source is
-not immediately discernible from the log line in which the mail first
-appears; from a strictly chronological reading of the logs it
-\textit{usually\/} appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
+explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
+Thus the source is not immediately discernible from the log line in which
+the mail first appears; from a strictly chronological reading of the logs
+it \textit{usually\/} appears as if the child mail has appeared from thin
+air.  Subsequently the parent mail will log the creation of the child mail:
 
@@
 addition the first log line belonging to the child mail (the log line cited
-above properly belongs to the parent mail) is logged by qmgr,\footnote{Qmgr
-is the Postfix daemon which manages the mail queue, determining which mails
-will have delivery attempted next.} so the order also depends on how busy
-qmgr is.\footnote{Postfix is quite paranoid about mail delivery, an
-excellent characteristic for an \MTA{} to possess, so it won't log that the
-child has been created until it is absolutely certain that the mail has
-been written to disk.}
+above properly belongs to the parent mail) is logged by \daemon{qmgr}, so
+the order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is
+quite paranoid about mail delivery, an excellent characteristic for an
+\MTA{} to possess, so it won't log that the child has been created until it
+is absolutely certain that the mail has been written to disk.}
 
 Because of this the parser cannot complain when it encounters a log line
-from qmgr for a previously unseen mail; it must flag the mail as coming
-from an unknown origin, and subsequently clear the flag if and when the
-origin of the mail becomes clear.  Obviously the parser could omit checking
-of where mails originate from, but the author believes that it is better to
-require an explicit source, as bugs in the parser are more likely to be
-exposed.
+from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
+coming from an unknown origin, and subsequently clear the flag if and when
+the origin of the mail becomes clear.  Obviously the parser could omit
+checking of where mails originate from, but the author believes that it is
+better to require an explicit source, as bugs in the parser are more likely
+to be exposed.
 
@@
 lines to the correct mail and a method of specifying the action the log
-line represents.  The former is achieved by using the smtpd process id to
-identify the correct mail during the initial phase, then switching to the
-queueid once the mail has been accepted.  The latter uses the action field
-of the rule which matched the log line, executing the code in the function
-named by the action.
+line represents.  The former is achieved by using the \daemon{smtpd}
+process id to identify the correct mail during the initial phase, then
+switching to the queueid once the mail has been accepted.  The latter uses
+the action field of the rule which matched the log line, executing the code
+in the function named by the action.
 
@@
 Everything starts off with a mail entering the system, whether by local
-submission via postdrop/sendmail,\footnote{Postfix provides a command that
-is compatible with the Sendmail (\url{http://www.sendmail.org/ XXX CHECK})
-mail submission program which all Unix commands which send mail depend on.}
-by \SMTP{}, by re-injection due to forwarding, or internally generated by
-Postfix.  Local submission is the simplest case: a queueid is assigned
-immediately and the sender address is logged (action: pickup;
-flowchart:~2).
+submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
+due to forwarding, or internally generated by Postfix.  Local submission is
+the simplest case: a queueid is assigned immediately and the sender address
+is logged (action: pickup; flowchart:~2).
 
@@
         commit; flowchart:~24).  Regardless of the state of its children
-        the data is cleaned up and entered in the database (flowchart:~25).
+        its data is cleaned up and entered in the database (flowchart:~25).
         The parent may have children which are waiting to be entered in the
@@
 
-    \item [\$line] The log line, with the standard syslog(3) fields parsed,
-        removed and made available.
-
-    \item [\$rule] The matching rule.
-
-    \item [@matches] The fields in the line captured by \$rule's regex.
+    \item [line] The log line, separated into fields:
+
+        \begin{description}
+
+            \item [timestamp] The time the line was logged at.
+
+            \item [host] The hostname of the server which logged the line.
+
+            \item [program] The name of the program which logged the line.
+
+            \item [pid] The \pid{} of the program which logged the line.
+
+            \item [text] The remainder of the line.
+
+        \end{description}
+
+    \item [rule] The matching rule.
+
+    \item [matches] The fields in the line captured by the rule's regex.
 
@@
     \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by smtpd pid) and save both the client
-        hostname and IP address.
+        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
+        client hostname and IP address.
 
@@
     \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the regex to it.
+        the log line, and save the data extracted by the \regex{} to it.
 
@@
         picking a mail from the queue to deliver. This action is used for
-        both qmgr and cleanup as it needs to deal with out of order log
-        lines; see section~\refwithpage{discarding cleanup lines} for
-        details.
-
-    \item [PICKUP] Pickup is the service which deals with mail submitted
-        locally via /usr/sbin/sendmail.
-
-        Out of order log entries may have caused the state table entry to
-        already exist (see section~\refwithpage{pickup logging after
-        cleanup}); otherwise it is created.  The data from the log line is
-        then saved to the state table entry.
+        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
+        out of order log lines; see section~\refwithpage{discarding cleanup
+        lines} for details.
+
+    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
+        service dealing with a locally submitted mail.  Out of order log
+        entries may have caused the state table entry to already exist (see
+        section~\refwithpage{pickup logging after cleanup}); otherwise it
+        is created.  The data from the log line is then saved to the state
+        table entry.
 
@@
     \item [MAIL\_TOO\_LARGE] When the client tries to send a message larger
-        than the local server accepts the mail will be discarded.  See
-        TIMEOUT for further discussion; the two are handled in exactly the
-        same way.
+        than the local server accepts, the mail will be discarded, and the
+        client informed.  See TIMEOUT for further discussion; the two are
+        handled in exactly the same way.
 
@@
     \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all smtpds,\footnote{Possibly other programs are killed
-        also, but the parser is only affected by and interested in smtpds
-        exiting.} requiring any active connections to be committed and
+        it kills all \daemon{smtpd} processes,\footnote{Possibly other
+        programs are killed also, but the parser is only affected by and
+        interested in \daemon{smtpd} processes exiting.} requiring any
+        active connections to be cleaned up, entered in the database, and
         deleted from the state tables.
 
-    \item [SMTPD\_KILLED] Sometimes Postfix needs to forcefully kill an
-        smtpd; the active connection for that smtpd must be committed and
+    \item [SMTPD\_KILLED] Sometimes an event occurs which requires Postfix
+        to forcefully kill a \daemon{smtpd}; that \daemon{smtpd} processes'
+        active connection must be cleaned up, entered in the database, and
         deleted from the state tables.
 
-    \item [SMTPD\_DIED] Sometimes an smtpd dies or exits unsuccessfully;
-        the active connection for that smtpd must be committed and deleted
-        from the state tables.
-
-    \item [SMTPD\_WATCHDOG] Smtpds have a watchdog timer to deal with
-        unusual situations --- after five hours the timer will expire and
-        the smtpd will exit.  This occurs very infrequently, as there are
-        many other timeouts which should occur in the intervening hours:
-        DNS timeouts, timeouts reading data, etc.  The active connection
-        for that smtpd must be committed and deleted from the state tables.
+    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
+        unsuccessfully; the active connection for that \daemon{smtpd} must
+        be cleaned up, entered in the database, and deleted from the state
+        tables.
+
+    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
+        to deal with unusual situations --- after five hours the timer will
+        expire and the \daemon{smtpd} will exit.  This occurs very
+        infrequently, as there are many other timeouts which should occur
+        in the intervening hours: \DNS{} timeouts, timeouts reading data
+        from the client, etc.  The active connection for that
+        \daemon{smtpd} must be cleaned up, entered in the database, and
+        deleted from the state tables.
 
@@
         necessary, marking it as a bounce notification and removing the
-        unknown origin flag if it already exists.
+        unknown origin flag if it exists.
 
@@
 
-    \item Neither smtpd nor pickup have logged any messages associated with
-        the mail, indicating it was generated internally by Postfix, not
-        accepted via \SMTP{} or submitted locally by sendmail.
+    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
+        messages associated with the mail, indicating it was generated
+        internally by Postfix, not accepted via \SMTP{} or submitted
+        locally by \daemon{postdrop} or sendmail.
 
@@
         locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the later case the message-id will be
+        result of aliasing.  In the latter case the message-id will be
         unchanged from the original bounce notification, and so even if it
@@
 Once a mail has been identified as a bounce notification the unknown origin
-flag is cleared and the mail can be entered in the database.
+flag is cleared and the mail can be cleaned up and entered in the database.
 
@@
 different type of message; arguably this is a benefit rather than a
-drawback, as future mails generated internally by Postfix will be handled
+drawback, as other mails generated internally by Postfix will be handled
 correctly.
@@
 
-    \item Is the mail missing its cleanup log message?  Every mail which
-        passes through Postfix will have a cleanup line; lack of a cleanup
-        line is a sure sign the mail didn't make it too far.
-
-    \item Were there exactly two smtpd log lines for the mail?  There
-        should be a connection line and a mail accepted line.
+    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
+        which passes through Postfix will have a \daemon{cleanup} line;
+        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
+        it too far.
+
+    \item Were there exactly two \daemon{smtpd} log lines for the mail?
+        There should be a connection line and a mail accepted line.
 
@@
 
-    \item Is the mail missing its cleanup log message?  Every mail which
-        passes through Postfix will have a cleanup line; lack of a cleanup
-        line is a sure sign the mail didn't make it too far.
-
-    \item Were there three or more smtpd log lines for the mail?  There
-        should be a connection line and a mail accepted line, followed by
-        one or more rejection lines.
-
-    \item Is the last smtpd log line a rejection line?
+    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
+        which passes through Postfix will have a \daemon{cleanup} line;
+        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
+        it too far.
+
+    \item Were there three or more \daemon{smtpd} log lines for the mail?
+        There should be a connection line and a mail accepted line,
+        followed by one or more rejection lines.
+
+    \item Is the last \daemon{smtpd} log line a rejection line?
 
@@
         the timestamp of the last accepted mail against the timestamp of
-        the last line logged by smtpd for that connection.  If the smtpd
-        timestamp is later there was a rejection between the accepted mail
-        and the timeout, therefore the timeout applies to a rejected mail;
-        the timeout has already been recorded so the timeout action
-        finishes.  If the mail acceptance timestamp is greater then the
-        timeout applies to the just-accepted mail, which will be discarded.
+        the last line logged by \daemon{smtpd} for that connection.  If the
+        \daemon{smtpd} timestamp is later there was a rejection between the
+        accepted mail and the timeout, therefore the timeout applies to a
+        rejected mail; the timeout has already been recorded so the timeout
+        action finishes.  If the mail acceptance timestamp is greater then
+        the timeout applies to the just-accepted mail, which will be
+        discarded.
 
@@
 
-This complication is further complicated by out of order cleanup lines: see
-section~\refwithpage{discarding cleanup lines} for details.
+This complication is further complicated by the presence of out of order \daemon{cleanup}
+lines: see section~\refwithpage{discarding cleanup lines} for details.
 
@@
 though there may be other circumstances which trigger it.  Sometimes the
-cleanup line is logged after the timeout line; parsing this line causes the
-creation of a new state table entry for the queueid in the log line.  This
-is incorrect because the line actually belongs to the mail which has just
-been discarded, and the next log line for that queueid will be seen when
-the queueid is reused, causing a queueid clash and the appropriate warning.
-
-In the case where the cleanup line is still pending the algorithm updates a
-global cache of queueids, adding the queueid and the timestamp from the
-timeout line.  When the next cleanup line is parsed for that queueid the
-cache will be checked, and the line will be deemed part of the discarded
-mail and discarded if it meets the following requirements:
+\daemon{cleanup} line is logged after the timeout line; parsing this line
+causes the creation of a new state table entry for the queueid in the log
+line.  This is incorrect because the line actually belongs to the mail
+which has just been discarded, and the next log line for that queueid will
+be seen when the queueid is reused, causing a queueid clash and the
+appropriate warning.
+
+In the case where the \daemon{cleanup} line is still pending the algorithm
+updates a global cache of queueids, adding the queueid and the timestamp
+from the timeout line.  When the next \daemon{cleanup} line is parsed for
+that queueid the cache will be checked, and the line will be deemed part of
+the discarded mail and discarded if it meets the following requirements:
 
@@
 
-    \item The timestamp of the cleanup line must be within ten minutes of
-        the mail acceptance timestamp.  Timeouts happen after five minutes,
-        but some data may have been transferred slowly, and empirical
-        evidence shows that ten minutes is not unreasonable; hopefully it
-        is a good compromise between false positives and false negatives.
+    \item The timestamp of the \daemon{cleanup} line must be within ten
+        minutes of the mail acceptance timestamp.  Timeouts happen after
+        five minutes, but some data may have been transferred slowly, and
+        empirical evidence shows that ten minutes is not unreasonable;
+        hopefully it is a good compromise between false positives and false
+        negatives.
 
@@
 
-The next cleanup line must meet the criteria above for it to be discarded
-because not every connection where a timeout occurs will have a cleanup
-line logged for it; if the algorithm blindly discarded the next cleanup
-line it would in some cases be mistaken.  Whether or not the next cleanup
-line is discarded the queueid will be removed from the cache of timeout
-queueids when the next pickup line containing that queueid is parsed.
-
-During the TIMEOUT action if the mail lacks a cleanup line, the queueid and
-other data about the mail, particularly the timestamp of the last log line,
-are added to a global cache of timeout queueids.  This cache is
-subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action, and
-if the queueid exists and the cleanup line passes the checks listed above
-it will be discarded.
+The next \daemon{cleanup} line must meet the criteria above for it to be
+discarded because not every connection where a timeout occurs will have a
+\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
+next \daemon{cleanup} line it would in some cases be mistaken.  Whether or
+not the next \daemon{cleanup} line is discarded the queueid will be removed
+from the cache of timeout queueids when the next \daemon{pickup} line
+containing that queueid is parsed.
+
+During the TIMEOUT action if the mail lacks a \daemon{cleanup} line, the
+queueid and other data about the mail, particularly the timestamp of the
+last log line, are added to a global cache of timeout queueids.  This cache
+is subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action,
+and if the queueid exists and the \daemon{cleanup} line passes the checks
+listed above it will be discarded.
 
@@
 
-Occasionally the pickup line logged when mail is submitted locally via
-sendmail appears later in the logfile than the cleanup line for that mail.
-This seems to occur during periods of particularly heavy load, so is most
-likely due to process scheduling vagaries.  Normally if the queueid given
-in the pickup line exists a warning is generated by the pickup action, but
-if the following conditions are met it is assumed that the lines are out of
-order:
+Occasionally the \daemon{pickup} line logged when mail is submitted locally
+via sendmail appears later in the logfile than the \daemon{cleanup} line
+for that mail.  This seems to occur during periods of particularly heavy
+load, so is most likely due to process scheduling vagaries.  Normally if
+the queueid given in the \daemon{pickup} line exists a warning is generated
+by the \daemon{pickup} action, but if the following conditions are met it
+is assumed that the lines are out of order:
 
@@
 
-    \item The only program which logged anything for the mail is cleanup.
+    \item The only program which has logged anything thus far for the mail
+        is \daemon{cleanup}.
 
     \item There is less than a five second difference in the timestamps of
-        the cleanup and pickup lines.
+        the \daemon{cleanup} and \daemon{pickup} lines.
 
@@
 
-Occasionally an smtpd will just stop logging, without an immediately
-obvious reason.  After poring over logs for some time there are several
-reasons for this infrequent occurrence:
+Occasionally a \daemon{smtpd} will just stop logging, without an
+immediately obvious reason.  After poring over logs for some time there are
+several reasons for this infrequent occurrence:
 
@@
     \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all smtpds exit, and all entries in the connections state
-        table must be committed and deleted from the state table.  The
-        queueid state table is untouched.
-
-    \item Postfix sometimes forcibly kills an smtpd, so the active
+        happens all \daemon{smtpd} processes exit, and all entries in the
+        connections state table must be committed and deleted from the
+        state table.  The queueid state table is untouched.
+
+    \item Postfix sometimes forcibly kills a \daemon{smtpd}, so the active
         connection must be committed and deleted from the connections state
@@
 
-    \item Occasionally an smtpd will die or exit uncleanly, so the active
-        connection must be committed and deleted from the connections state
-        table.
+    \item Occasionally a \daemon{smtpd} will die or exit uncleanly, so the
+        active connection must be committed and deleted from the
+        connections state table.
 
@@
 
-The above appear to cover all situations where an smtpd suddenly stops
-logging.  In addition to removing an active connection the last accepted
-mail may need to be discarded, as detailed in
+The above appear to cover all situations where a \daemon{smtpd} suddenly
+stops logging.  In addition to removing an active connection the last
+accepted mail may need to be discarded, as detailed in
 section~\refwithpage{timeouts-during-data-phase}.
@@
 is reused before the previous mail is fully dealt with, when a queueid or
-pid is not found,\footnote{There will often be warnings about missing
-queueids or pids in the first few hundred or thousand log lines because the
+\pid{} is not found,\footnote{There will often be warnings about a missing
+queueid or \pid{} in the first few hundred or thousand log lines because the
 earlier log lines for those connections or mails are in the previous log
@@
         specialised tool.  It does not attempt to correlate log lines by
-        queueid or pid, nor does it need to deal with the complications
+        queueid or \pid{}, nor does it need to deal with the complications
         encountered during this project.  Pflogsumm produces a useful
@@
         \texttt{key=<value>} fields; most importantly it makes no attempt
-        to correlate log lines by queueid or pid, nor to handle any of the
-        myriad complications discussed in this document.  Correlation of
-        log lines is left up to the user, and Splunk certainly makes the
+        to correlate log lines by queueid or \pid{}, nor to handle any of
+        the myriad complications discussed in this document.  Correlation
+        of log lines is left up to the user, and Splunk certainly makes the
         job easier than using grep, awk and similar Unix tools unless the
@@
         Unlike other parsers reviewed, Lire does attempt to correlate log
-        lines by queueid, but not by pid, so e.g.\ the complete list of
+        lines by queueid, but not by \pid{}, so e.g.\ the complete list of
         recipients for a mail should be available.
 
-        Lire does attempt to correlate log lines by queueid, not pid, but
-        its hard coded parsing extracts only part of the available data and
-        makes no attempt to deal with the other complications described in
-        sections~\refwithpage{complications} and~\refwithpage{additional
+        Lire does attempt to correlate log lines by queueid, not \pid{},
+        but its hard coded parsing extracts only part of the available data
+        and makes no attempt to deal with the other complications described
+        in sections~\refwithpage{complications} and~\refwithpage{additional
         complications}.  It stores data, but not in an easily accessible
@@
         Fields are correlated based on the queueid (referred to as
-        sessionname within Logrep), but not by pid.  The parsing is hard
-        coded, hard to extend, and error prone: empty fields are saved when
-        the log line doesn't match the regex, though it appears that they
-        will not overwrite existing data.  It also counts the number of log
-        lines parsed and skipped.  Most notably rejections are completely
-        ignored, making it unsuitable for the purposes of this project.
-        Needless to say it doesn't attempt to address any of the
-        complications in sections~\refwithpage{complications}
-        and~\refwithpage{additional complications} except for correlating
-        by queueid.
+        sessionname [sic] within Logrep), but not by \pid{}.  The parsing
+        is hard coded, hard to extend, and error prone: empty fields are
+        saved when the log line doesn't match the \regex{}, though it
+        appears that they will not overwrite existing data.  It also counts
+        the number of log lines parsed and skipped.  Most notably
+        rejections are completely ignored, making it unsuitable for the
+        purposes of this project.  Needless to say it doesn't attempt to
+        address any of the complications in
+        sections~\refwithpage{complications} and~\refwithpage{additional
+        complications} except for correlating by queueid.
 
@@
 While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few
-attempt to correlate log lines queueid (none correlate by pid) to produce
-an overall record of the journey each mail traverses through Postfix.  No
+parsing (some to a greater level of detail than others), few attempt to
+correlate log lines queueid (none correlate by \pid{}) to produce an
+overall record of the journey each mail traverses through Postfix.  No
 other parser collects the breadth of information gathered by \parsername{},
@@
 anomalous dips and peaks apparent in some graphs are explained.  The next
-group of graphs vividly shows the huge impact that caching compiled regexs
-has on parser performance.  This section concludes with a breakdown of the
-rule hits accumulated during a single test run of the parser.
+group of graphs vividly shows the huge impact that caching compiled
+\regexs{} has on parser performance.  This section concludes with a
+breakdown of the rule hits accumulated during a single test run of the
+parser.
 
@@
 \showgraph{build/plot-normal-filesize-numlines.ps}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines graph}
+size vs number of lines}{execution time vs file size vs number of lines
+graph}
 
@@
 
-The following graphs show the impact that not caching compiled regexs has
-on parser performance: on typical log files the execution time when not
-caching compiled regexs is 500--600\% of the execution time when caching;
-reversing the perspective shows that cached execution time is merely
-17--20\% of non-cached execution time.  Caching compiled regexs is probably
-the single most effective optimisation possible in the parser's
+The following graphs show the impact that not caching compiled \regexs{}
+has on parser performance: on typical log files the execution time when not
+caching compiled \regexs{} is 500--600\% of the execution time when
+caching; reversing the perspective shows that cached execution time is
+merely 17--20\% of non-cached execution time.  Caching compiled \regexs{}
+is probably the single most effective optimisation possible in the parser's
 implementation.
@@
 average number of rules consulted per log line is much lower when there is
-a mail loop; this results in far fewer regex compilations per line than
+a mail loop; this results in far fewer \regex{} compilations per line than
 when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching regexs for
+time.  The increases in execution time when not caching \regexs{} for
 different combinations of log files are summarised in the table below:
@@
 unexpected consequences of specific inputs.  The necessity of caching
-compiled regexs is attested to by the third group of graphs, where the
+compiled \regexs{} is attested to by the third group of graphs, where the
 difference between caching and discarding is staggering.  This section
@@
 
+% Postfix components
+\glossary{name={bounce},description={
+    The definitive documentation is \url{http://www.postfix.org/bounce.8.html}.
+}}
+
+\glossary{name={cleanup},description={
+    Cleanup processes all incoming mail, sitting between qmgr and the
+    programs which accept mail (smtpd and pickup).  It removes duplicate
+    recipient addresses, inserts missing headers, and optionally rewrites
+    addresses if configured to do so.  The definitive
+    documentation is \url{http://www.postfix.org/cleanup.8.html}.
+}}
+
+\glossary{name={local},description={
+    The definitive documentation is \url{http://www.postfix.org/local.8.html}.
+}}
+
+\glossary{name={pickup},description={
+    Pickup is the service which deals with mail submitted locally via
+    postdrop and sendmail; it passes all submitted mail to cleanup.  The
+    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
+}}
+
+\glossary{name={postdrop},description={
+    The definitive documentation is \url{http://www.postfix.org/postdrop.1.html}.
+}}
+
+\glossary{name={postsuper},description={
+    The definitive documentation is available at \url{http://www.postfix.org/postsuper.1.html}.
+}}
+
+\glossary{name={qmgr},description={
+    Qmgr is the Postfix daemon which manages the mail queue, determining
+    which mails will be delivered next.  Qmgr orders the mails based on the
+    recipient for local mails and the destination server for remote
+    addresses, ensuring that it balances the aims of achieving maximum
+    concurrency while avoiding overwhelming destinations or wasting time
+    and resources on non-responsive destinations.  The definitive
+    documentation is \url{http://www.postfix.org/qmgr.8.html}.
+}}
+
+\glossary{name={sendmail},description={
+    Postfix provides a command that is compatible with the Sendmail
+    (\url{http://www.sendmail.org/}) mail submission program
+    that all Unix commands which send mail depend on.  The definitive
+    documentation is \url{http://www.postfix.org/sendmail.1.html}.
+}}
+
+\glossary{name={smtp},description={
+    The definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
+}}
+
+\glossary{name={smtpd},description={
+    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
+    implements all the anti-spam restrictions Postfix provides.  The
+    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
+}}
+
+\glossary{name={Bayesian spam filtering},description={
+    Bayesian spam filtering is a method of classifying mail based on the
+    frequency that the words in the mail have previously appeared in a spam
+    corpus and a ham (non-spam) corpus.  A full description is beyond the
+    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
+    for a detailed explanation.
+}}
+
+\glossary{name={Bayesian poisoning},description={
+    Bayesian poisoning is the addition of innocuous or unrelated words to a
+    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
+    word Viagra would be firmly in the spam corpus for most people, but by
+    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
+    forward\/} and \textit{best business practices\/} to a mail received by
+    a manager, the Bayesian spam filter might tip the balance from bad to
+    good, because the non-spam words might outweigh the spam words.
+}}
+
+\glossary{name={$<>$},sort={<>},description={
+    $<>$ is the sender address used for mail which should not be replied
+    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
+    in $<>$, so \textit{username\at{}domain\/} becomes
+    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
+    address.  All mail servers must accept mail sent from $<>$, or they are
+    in violation of \SMTP{}.
+}}
+
 \end{document}

### 2007-11-12 doc/logparser.tex
More corrections.

@@
 
-Some mail clients appear to send the following sequence of commands during
+Some mail clients appear\footnote{Due to privacy concerns no attempt has
+been made to identify either the users or the software which exhibits this
+behaviour.} to send the following sequence of commands during
 the \SMTP{} session:
@@
 behaviour occurs.  In this case a list of all mails accepted during a
-connection are saved in the connection state, and the accepted mails are
-examined when the disconnection line is parsed and the disconnection action
-executed.  Each mail is checked for the following characteristics:
+connection is saved in the connection state, and the accepted mails are
+examined when the disconnection action is executed.  Each mail is checked
+for the following characteristics:
 
@@
     \item Were there exactly two \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line.
+        There should be a connection line and a mail accepted line, nothing
+        else.
 
@@
 available is the client hostname and IP address, but the database schema
-requires many more fields be populated (see section~\refwithpage{database
-schema}).
+requires many more fields be populated (see
+sections~\refwithpage{connections table} and~\refwithpage{results table}).
 
@@
 won't yet have been any data gathered for the mail, so there is none
-available to save; the timeout is saved with the connection data instead.
+available to save; the timeout is recorded with the connection data
+instead.
 
@@
 
-    \item Because of ESMTP pipelining the client can send MAIL FROM, RCPT
-        TO and DATA commands in one packet, instead of sending each command
+    \item \ESMTP{} pipelining allows the client to send MAIL FROM, RCPT TO
+        and DATA commands in one packet, instead of sending each command
         individually and waiting for the reply before sending subsequent
@@
         client to tell which command was rejected.  Some clients make no
-        attempt to recover and disconnect uncleanly.
+        attempt to recover and disconnect uncleanly; others presumably
+        stop pipelining and restart the mail sending operation.  The client
+        could attempt to parse the error message, but that would be
+        extremely error prone, whereas not pipelining makes the cause of
+        the error message obvious.
 
@@
 
-This complication is further complicated by the presence of out of order \daemon{cleanup}
-lines: see section~\refwithpage{discarding cleanup lines} for details.
+This complication is further complicated by the presence of out of order
+\daemon{cleanup} lines: see section~\refwithpage{discarding cleanup lines}
+for details.
 
@@
 
-In the case where the \daemon{cleanup} line is still pending the algorithm
-updates a global cache of queueids, adding the queueid and the timestamp
-from the timeout line.  When the next \daemon{cleanup} line is parsed for
-that queueid the cache will be checked, and the line will be deemed part of
-the discarded mail and discarded if it meets the following requirements:
+In the case where the \daemon{cleanup} line is still pending during the
+TIMEOUT action the algorithm updates a global cache of queueids, adding the
+queueid and the timestamp from the timeout line.  When the next
+\daemon{cleanup} line is parsed for that queueid the cache will be checked,
+and the line will be deemed part of the discarded mail and discarded if it
+meets the following requirements:
 
@@
 \daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line it would in some cases be mistaken.  Whether or
-not the next \daemon{cleanup} line is discarded the queueid will be removed
-from the cache of timeout queueids when the next \daemon{pickup} line
-containing that queueid is parsed.
+next \daemon{cleanup} line after a TIMEOUT it would in some cases be
+mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
+queueid will be removed from the cache of timeout queueids when the next
+\daemon{pickup} line containing that queueid is parsed.
 
@@
 is subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action,
-and if the queueid exists and the \daemon{cleanup} line passes the checks
-listed above it will be discarded.
+and if the queueid is found in the cache and the \daemon{cleanup} line
+passes the checks listed above the line will be discarded.
 
@@
 Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the logfile than the \daemon{cleanup} line
+via sendmail appears later in the log file than the \daemon{cleanup} line
 for that mail.  This seems to occur during periods of particularly heavy
@@
 
-    \item There is less than a five second difference in the timestamps of
-        the \daemon{cleanup} and \daemon{pickup} lines.
+    \item There is less than a five second difference between the
+        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
 
@@
 As always with heuristics there may be circumstances in which these
-heuristics incorrectly match, but none have been identified so far.
+heuristics match incorrectly,  but none have been identified so far.
 
@@
         happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be committed and deleted from the
-        state table.  The queueid state table is untouched.
-
-    \item Postfix sometimes forcibly kills a \daemon{smtpd}, so the active
-        connection must be committed and deleted from the connections state
-        table.
+        connections state table must be cleaned up, entered in the database
+        and deleted.  The queueid state table is untouched.
+
+    \item Postfix sometimes forcibly kills a \daemon{smtpd} XXX FIGURE OUT
+        WHY, so the active connection must be cleaned up, entered in the
+        database and deleted from the connections state table.
 
     \item Occasionally a \daemon{smtpd} will die or exit uncleanly, so the
-        active connection must be committed and deleted from the
-        connections state table.
+        active connection must be cleaned up, entered in the database and
+        deleted from the connections state table.
 
@@
 
-The above appear to cover all situations where a \daemon{smtpd} suddenly
-stops logging.  In addition to removing an active connection the last
-accepted mail may need to be discarded, as detailed in
-section~\refwithpage{timeouts-during-data-phase}.
+The above descriptions appear to cover all situations identified thus far
+where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
+active connection the last accepted mail may need to be discarded, as
+detailed in section~\refwithpage{timeouts-during-data-phase}.
 
@@
 re-injected mail},~\refwithpage{discarding cleanup
-lines}~and~\refwithpage{pickup logging after cleanup}.  In the 93 log files
+lines}~or~\refwithpage{pickup logging after cleanup}.  In the 93 log files
 used for testing this occurs only five times in 60,721,709 log lines, but
@@
 mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged.  The SAVE\_BY\_QUEUEID action
-checks for the flag and retries entry if it's found; if the additional log
-lines have caused the mail to reach a valid combination entry will succeed,
-otherwise it must be postponed once more.
-
-The list of valid combinations is given below.  Every mail will
+must be postponed and the mail flagged for later entry.  The
+SAVE\_BY\_QUEUEID action checks for the flag and retries entry if it's
+found; if the additional log lines have caused the mail to reach a valid
+combination entry will proceed, otherwise it must be postponed once more.
+
+The list of valid combinations is explained below.  Every mail will
 additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
@@
 
-    \item [\daemon{local}:] Local delivery of bounce notification, or local
-        delivery of forwarded/tracked mail.
+    \item [\daemon{local}:] Local delivery of a bounce notification, or
+        local delivery of a forwarded or tracked mail.
 
@@
     \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery.
+        client, local delivery only.
 
     \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded/tracked mail.
+        delivery only.
+
+    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
+        a bounce notification.
 
@@
 
-This applies to accepted mails only, not to rejected mails.
+This check applies to accepted mails only, not to rejected mails.
 
@@
 
-This section has presented the parsing algorithm, divided into: a high
-level overview of the basic algorithm, followed by the first group of
-difficulties thwarted.  The revised algorithm was detailed, preceded by a
-flow chart depicting the paths a mail may take through the algorithm.  The
-actions which are available to rules were described in full, and finally
-the remaining complications discovered and overcome in the process of
-refining and completing the algorithm.  Detecting, diagnosing and defeating
-complications forms the largest single subsection of this document,
-mirroring the development of the parser,  The complications are described
-in the order they were overcome, with subsequent problems affecting fewer
-mails (often by an order or magnitude), though the time required to solve
-problems increased with each successive problem.
+This section has presented the parsing algorithm, starting with a high
+level overview of the basic algorithm, and the first group of difficulties
+thwarted.  The revised algorithm was detailed, preceded by a flow chart
+depicting the paths a mail may take through the algorithm, and followed by
+the actions which are available to rules were described in full.  The
+section concludes with the remaining complications discovered and overcome
+in the process of refining and completing the algorithm.  Detecting,
+diagnosing and defeating complications forms the largest single portion of
+this document, mirroring the development of the parser,  The complications
+are described in the order they were overcome, with subsequent problems
+affecting fewer mails (often by an order or magnitude), though the time
+required to solve problems increased with each successive problem.
 
@@
 no warnings, and the state table should only contain entries for mails
-which had yet to be delivered when the logfiles ended, or started before
-the logfiles began.
+which had yet to be delivered when the log files ended, or started before
+the log files began.
 
@@
 
-The dips at logfiles 22 and 62--68 correspond to peaks in log file size in
+The dips at log files 22 and 62--68 correspond to peaks in log file size in
 graph~\refwithpage{execution time vs file size vs number of lines graph},
@@
 % The contents of the glossary.
-\glossary{name={ESMTP},description={
-    ESMTP is Extended \SMTP{}, defined in \RFC{}~1869~\cite{RFC1869}.
-}}
-
 \glossary{name={SQLite3},description={
@@
 
+\glossary{name={queueid},description={
+    Each mail in Postfix's queue is assigned a queueid to uniquely identify
+    it.  Queueids are assigned from a limited pool, so although they are
+    guaranteed to be unique for the lifetime of the mail, given sufficient
+    time they will be reused.
+}}
+
 \end{document}

### 2007-11-12 doc/logparser.tex
Corrections to Coverage, Limitations and Review of Other Parsers.

@@
 
+\label{connection reuse}
+
 Multiple independent mails may be delivered during one connection: this
@@
 coverage is hard because other servers have different restrictions with
-custom messages, RBLs change their messages over time, major releases of
-Postfix change warning messages (usually adding more information), etc.\ so
-over time the log lines drift and change.  Graph~\refwithpage{rule hits}
-shows the number of hits for each rule over all 93 log files; it's obvious
-that a small number of rules match the vast majority of the lines, and more
-than half the rules match fewer than 100 times.
+custom messages, \RBL{} messages change over time, major releases of
+Postfix change warning messages (usually adding more information), etc.,\
+so over time the log lines drift and change.  Graph~\refwithpage{rule hits
+graph} shows the number of hits for each rule over all 93 log files; it's
+obvious that a small number of rules match the vast majority of the lines,
+and more than half the rules match fewer than 100 times.
 
@@
 \pid{} is not found,\footnote{There will often be warnings about a missing
-queueid or \pid{} in the first few hundred or thousand log lines because the
-earlier log lines for those connections or mails are in the previous log
-file; loading the saved state from the previous log file will solve this
-problem.} or when there are problems tracking a child mail (see
+queueid or \pid{} in the first few hundred or thousand log lines because
+the earlier log lines for those connections or mails are in the previous
+log file; loading the saved state from the previous log file will solve
+this problem.} or when there are problems tracking a child mail (see
 section~\refwithpage{tracking re-injected mail}).  There should be few or
-no warnings, and the state table should only contain entries for mails
-which had yet to be delivered when the log files ended, or started before
-the log files began.
+no warnings when parsing, and when finished parsing the state table should
+only contain entries for mails which had yet to be delivered when the log
+files ended, or started before the log files began.
 
@@
 percentage of log lines parsed is relatively simple and not intrusive:
-adding new rules or modifying existing rules is simplified by the parser
-implementation.  Improving the logical coverage is harder, as the actions
-taken by Postfix must be reconstructed, and the new sequence of actions
-integrated into the existing model without breaking the existing sequences.
-Detecting a deficiency in the parsing algorithm is also significantly
-harder than detecting unparsed log lines, as the parser will warn about any
-unparsed line, whereas the author must spot a flaw in the parser by
-understanding the warnings generated.
+adding new rules or modifying existing rules is simplified by the
+separation of rules and algorithm.  Improving the logical coverage is
+harder, as the actions taken by Postfix must be reconstructed, and the new
+sequence of actions integrated into the existing model without breaking
+the existing sequences.  Detecting a deficiency in the parsing algorithm is
+also significantly harder than detecting unparsed log lines, as the parser
+will warn about any unparsed line, whereas discovering a flaw in the
+parser requires understanding of the warnings produced and the mails
+remaining in the state table.  Rectifying a flaw in the parser requires an
+understanding of both the parser and Postfix's logs, and investigative work
+to determine the cause of the deficiency, followed by further examination
+of the logs in developing a solution.
 
@@
 
-\begin{itemize}
+\begin{enumerate}
 
     \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly new RBLs, new policy servers and
-        new administrator defined rejection messages require new rules.
+        with the new log lines.  Similarly using a new \RBL{}, new policy
+        server or new administrator defined rejection messages require new
+        rules.
 
@@
         appear in the database as one mail with lots of rejections followed
-        by acceptance.  I don't believe it's possible to make this
-        distinction given the data Postfix logs, though it might be
-        possible to write a policy server to provide additional logging.
+        by acceptance (this has already been mentioned in
+        section~\refwithpage{connection reuse}).  I don't believe it's
+        possible to make this distinction given the data Postfix logs,
+        though it might be possible to write a policy server to provide
+        additional logging.
 
@@
 
-\end{itemize}
+\end{enumerate}
 
@@
 
-    \item Write the policy server referred to in the limitations above.
+    \item Write the policy server referred to in limitation 3 above.
 
@@
 
+XXX COMPARE AND CONTRAST RATHER THAN REVIEW\@.
+
+XXX HIGHLIGHT THE DIFFERENCE BETWEEN ADDING NEW RULES AND NEW PARSER
+FUNCTIONALITY\@.
+
+NEED A BETTER TERM THAN HARD-CODED, THOUGH THAT MIGHT ARISE FROM THE
+PREVIOUS TODO ITEM\@.
+
 \begin{description}
@@
 
-    \item [Sawmill Universal Log File Analysis and Reporting] is a general
-        purpose commercial product which parses 687 log file formats
-        (correct as of 2007/04/15) and produces reports.  Its data
-        extraction facilities are quite limited, though it does extract
-        three different sets of data for Postfix (as of 2007/04/15, one is
-        beta), but they do not appear to be interlinked, nor does it save
-        sufficient data for the purposes of this project.  Not attempt is
-        made to correlate log lines or deal with the difficulties
+    \item [Sawmill Universal Log File Analysis and Reporting] \textit{
+        Sawmill is a \newline Postfix log analyzer (it also support 686
+        other log formats).  It can process log files in Postfix format,
+        and generate dynamic statistics from them, analyzing and reporting
+        events.  Sawmill can parse Postfix logs, import them into a SQL
+        database (or its own built-in database), aggregate them, and
+        generate dynamically filtered reports, all through a web interface.
+        Sawmill can perform Postfix analysis on any platform, including
+        Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other UNIX, and
+        more.\/}
+
+        Sawmill is a general purpose commercial product which parses 687
+        log file formats (correct as of 2007/04/15) and produces reports.
+        Its data extraction facilities are quite limited, though it does
+        extract three different sets of data for Postfix (one is beta as of
+        2007/04/15), but they do not appear to be interlinked, nor does it
+        save sufficient data for the purposes of this project.  No attempt
+        is made to correlate log lines or deal with the difficulties
         documented in sections~\refwithpage{complications}
@@
         only (presumably for a fee), and the product is quite expensive, as
-        it requires a license for each report which is to be generated.
-        The web interface allows creation of dynamic reports based on any
-        field, but due to the associated cost the author has not
-        experimented with it.
+        it requires a license per report which is to be generated.  The web
+        interface allows creation of dynamic reports based on any field,
+        but due to the associated cost the author has not experimented with
+        it.
 
@@
         \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
-        Fields extracted: from, to, RBL list, client hostname and client
+        Fields extracted: from, to, \RBL{}, client hostname and client
         IP\@.
@@
         \newline Fields extracted: from, to, client hostname, client IP,
-        relay hostname, relay IP, status, response code, RBL list and
+        relay hostname, relay IP, status, response code, \RBL{} and
         message id.
@@
         logs, Splunk extracts some standard fields: to and from addresses,
-        HELO, date, host the logs were collected from, and protocol.  It
-        parses the standard syslog fields at the beginning of the line, and
-        extracts any \texttt{key=<value>} pairs; these are the fields
-        listed above.  Searches can be made based on the extracted fields,
-        and all text in the line is also available for searching.  Many
-        types of reports are available, though all are variations of a bar
-        or pie chart, with the exception of bubble and heatmap graphs.  It
-        is very easy to drill down through the graphs to extract a portion
-        of the data (e.g.\ select the hour with the largest number of
-        events, then select a particular host, and finally a specific
-        address).  It is not possible to search on partial words, though
-        the web interface does provide and refine a list of possible search
-        terms as you type.  The searches are quite fast, though as the free
-        version has a limit on the amount of data indexed per day the
-        volume of data to be searched was merely 45mb; the cheapest
-        licensed version costs \$6000, and it still limits the volume of
-        data indexed per day.
+        HELO hostname, date, host the logs were collected from, and
+        protocol.  It parses the standard syslog fields at the beginning of
+        the line, and extracts any \texttt{key=<value>} pairs; these are
+        the fields listed above.  Searches can be made based on the
+        extracted fields, and all text in the line is also available for
+        searching.  Many types of reports are available, though all are
+        variations of a bar or pie chart, with the exception of bubble and
+        heatmap graphs.  It is very easy to drill down through the graphs
+        to extract a portion of the data (e.g.\ select the hour with the
+        largest number of events, then select a particular host, and
+        finally a specific address).  It is not possible to search on
+        partial words, though the web interface does provide and refine a
+        list of possible search terms as you type.  The searches are quite
+        fast, though as the free version has a limit on the amount of data
+        indexed per day the volume of data to be searched was merely 45mb;
+        the cheapest licensed version costs \$6000, and it still limits the
+        volume of data indexed per day.
 
@@
         in C. It designed to scan qmail, postfix, sendmail and exim logfile
-        and produce usage statistics in HTML format for viewing through a
-        browser. It produces Top domains output according to Sender,
+        and produce usage statistics in \HTML{} format for viewing through
+        a browser. It produces Top domains output according to Sender,
         Receiver, Total mails and bytes; it keeps your main domain mail
@@
         at \url{http://httpd.apache.org/}} format log files, for analysis
-        by its HTTP log file parser.  The converting parser is hard coded,
-        hard to extend, and only deals with a small portion of the log
-        lines generates by Postfix, but it does correlate log lines by
+        by AWStats' HTTP log file parser.  The converting parser is hard
+        coded, hard to extend, and only deals with a small portion of the
+        log lines generated by Postfix, but it does correlate log lines by
         queueid.
@@
         YAALA uses a plugin based system to analyse log files and produce
-        HTML output reports.  The core code is merely 153 lines long, as
+        \HTML{} output reports.  The core code is merely 153 lines long, as
         all the parsing and report generation is handled by modules.  YAALA
@@
         data structures using Perl's Storable module~\cite{perl-storable},
-        so with some knowledge of the module responsible for parsing
-        Postfix log files it should be possible to extract data with an
-        external program.  Using YAALA as a base would be only slightly
-        less work as both input and output modules would need to be
-        written; it may even be more work to implement the parser within
-        the constraints of YAALA\@.  YAALA bundles a Postfix parser, which
-        uses hard-coded parsing to extract the following fields from
-        specific log lines:
-
-        Aggregations: count, bytes (the sum of bytes transferred; count is
-        not explained).\newline Keyfields [sic]: date, hour, sender,
+        so with some knowledge of the data structure created by the module
+        responsible for parsing Postfix log files it should be possible to
+        extract data with an external program.  Using YAALA as a base would
+        be only slightly less work than starting from scratch, as both
+        input and output modules would need to be written; it may even be
+        more work to implement the parser within the constraints of
+        YAALA\@.  YAALA bundles a Postfix parser, which uses hard-coded
+        parsing to extract the following fields from specific log lines:
+
+        Aggregations: count(count is not explained), bytes (the sum of
+        bytes transferred).\newline Keyfields [sic]: date, hour, sender,
         recipient, defer\_count, delay, incoming\_host, outgoing\_host.
@@
         WHERE host =\~{} Google}. \newline The mini-language is quite
-        limited, and cannot be used to extract data, merely to create
-        reports.
+        limited, and cannot be used to extract data for external use,
+        merely to create reports.
 
@@
         relays most used, etc.\/}; notably rejections are not mentioned.
-        It supports multiple output formats (text, HTML, PDF and Excel 95).
-        Logs can be imported and the extracted data saved, and reports
-        generated later.  Accessing this data from another application is
-        undocumented; given the source code it should be possible, with
-        time, to access the data.
-
-        Unlike other parsers reviewed, Lire does attempt to correlate log
-        lines by queueid, but not by \pid{}, so e.g.\ the complete list of
-        recipients for a mail should be available.
-
-        Lire does attempt to correlate log lines by queueid, not \pid{},
-        but its hard coded parsing extracts only part of the available data
-        and makes no attempt to deal with the other complications described
-        in sections~\refwithpage{complications} and~\refwithpage{additional
+        It supports multiple output formats (text, \HTML{}, PDF XXX and
+        Excel 95).  Logs can be imported and the extracted data saved, and
+        reports generated from it later.  Accessing this data from another
+        application is undocumented; given the source code it should be
+        possible, with time, to access the data.
+
+        Like AWStats and Logrep, Lire attempts to correlate log lines by
+        queueid, but not by \pid{}, so the complete list of recipients for
+        a mail should be available; however its hard coded parsing extracts
+        only part of the available data and makes no attempt to deal with
+        the other complications described in
+        sections~\refwithpage{complications} and~\refwithpage{additional
         complications}.  It stores data, but not in an easily accessible
-        format.  It generates reports in multiple formats, but those
+        format, and generates reports in multiple formats, but those
         reports do not appear to be customisable.
@@
         rejections are completely ignored, making it unsuitable for the
-        purposes of this project.  Needless to say it doesn't attempt to
-        address any of the complications in
-        sections~\refwithpage{complications} and~\refwithpage{additional
-        complications} except for correlating by queueid.
-
-        Logrep does not come with any documentation, though some scanty
+        purposes of this project.  It doesn't attempt to address any of the
+        complications in sections~\refwithpage{complications}
+        and~\refwithpage{additional complications} except for correlating
+        by queueid.
+
+        Logrep does not come with any documentation, though some scant
         documentation is available on its website.  It requires a web
@@
         possible to review the reports Logrep can generate (available in
-        HTML only), nor to examine the (undocumented) format in which it
+        \HTML{} only), nor to examine the (undocumented) format in which it
         can save extracted data for subsequent reuse.
@@
 parsing (some to a greater level of detail than others), few attempt to
-correlate log lines queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  No
-other parser collects the breadth of information gathered by \parsername{},
-nor makes it as easy to add new parsing rules.  Most other parsers
-immediately generate a report and discard the data extracted from the logs;
-those which don't discard the data retain it in a format inaccessible to
-other tools.  All of the parsers reviewed can produce a report of greater
-or lesser detail and complexity, a facility not offered by \parsername{};
-reporting is deferred to a subsequent application.
+correlate log lines by queueid (none correlate by \pid{}) to produce an
+overall record of the journey each mail traverses through Postfix.  None of
+the reviewed parsers collect the breadth of information gathered by
+\parsername{}, nor make it as easy to extend the parser.  Most other
+parsers immediately generate a report and discard the data extracted from
+the logs; those which don't discard the data retain it in a format
+inaccessible to other tools.  All of the parsers reviewed can produce a
+report of greater or lesser detail and complexity, a facility not offered
+by \parsername{}; reporting is deferred to a subsequent application.
 

### 2007-11-13 doc/logparser-acronyms.tex
More corrections.
Corrections to Coverage, Limitations and Review of Other Parsers.
Finished the corrections, or marked where corrections remain to be made. More acronyms, and updated acronyms I missed earlier.

@@
     code needs to be informed of the functionality provided by the code.
-    An API defines the interface provided to the user, and acts a contract
-    between the user and the provider: if the user adheres to the API the
-    provider guarantees it will work, while the provider is free to change
-    the implementation as long as the API is preserved.
+    An API defines the interface provided to the user, and acts as a
+    contract between the user and the provider: if the user adheres to the
+    API the provider guarantees it will work, while the provider is free to
+    change the implementation as long as the API is preserved.
 }}
@@
 \newacronym{FQDN}{Fully Qualified Domain Name}{description={
-    An FQDN is a hostname plus domain name, e.g.
+    An FQDN is a hostname plus domain name, e.g.    \newline
         example.com     is a domain name          \newline
@@
 
-\newacronym{regexs}{Regular Expressions}{description={
-    Regular expressions are a method of matching patterns in text that are
-    explained in detail in~\cite{wikipedia-regex, perlre} .
+% I think this is better than having both regex and regexs in the acronym
+% list.
+\newcommand{\regexs}{\regex{}s}
+
+\newacronym{ESMTP}{Extended SMTP}{description={
+    ESMTP is Extended SMTP, defined in RFC~1869~\cite{RFC1869}.
 }}
+
+\newacronym{HTML}{Hypertext Markup Language}{description={
+    HTML is the markup language designed for writing web
+    pages~\cite{wikipedia-html, w3schools-html}.
+}}
+
+\newacronym{IP}{Internet Protocol}{description={
+    XXX ADD A DEFINITION\@.
+}}

### 2007-11-13 doc/logparser.tex
Finished the corrections, or marked where corrections remain to be made. More acronyms, and updated acronyms I missed earlier.

@@
 underlying the project, followed by some technical information: the use of
-a database as an \API\@; a brief introduction to
-\ifacronymfirstuse{SMTP}{the}{} \SMTP\@; and a longer introduction to
+a database as an \API{}\@; a brief introduction to
+\ifacronymfirstuse{SMTP}{the}{} \SMTP{}\@; and a longer introduction to
 Postfix, concentrating on the topics most relevant to this paper, namely
@@
             \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-SQL\@:
+                events occur.  In pseudo-\SQL{}\@:
 
@@
 
-    \item SQL is reasonably standard and many people will already be
+    \item \SQL{} is reasonably standard and many people will already be
         familiar with it; for those unfamiliar with it there are lots of
         readily available resources from which to learn.  Although every
-        vendor implements a different dialect of SQL, the basics are the
+        vendor implements a different dialect of \SQL{}, the basics are the
         same everywhere (analogous to the overall similarities and minor
@@
 in \RFC{} 2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA\@.  It is a simple, human readable, plain text
+sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
 protocol, making it quite easy to test and debug problems with it.  Despite
@@
 do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
-into various component programs to perform the tasks required of an MTA\@:
-receive mail, send mail, local delivery of mail, etc. --- full details can
-be found at~\cite{postfix-overview}.  Each log line contains the name of
-the Postfix program which produced it, and this information is used when
-determining which rules should be used to parse each log line (see
-section~\refwithpage{rule characteristics} for details.)
+into various component programs to perform the tasks required of an
+\MTA{}\@: receive mail, send mail, local delivery of mail, etc. --- full
+details can be found at~\cite{postfix-overview}.  Each log line contains
+the name of the Postfix program which produced it, and this information is
+used when determining which rules should be used to parse each log line
+(see section~\refwithpage{rule characteristics} for details.)
 
@@
 \footnote{\texttt{cidr}~\cite{cidr_table} is the type of the lookup table.}
-will check whether the IP address of the connected client matches the left
-hand side of each line in the file and will return the right hand side of
-the first matching line.  Other restrictions determine their result in
+will check whether the \IP{} address of the connected client matches the
+left hand side of each line in the file and will return the right hand side
+of the first matching line.  Other restrictions determine their result in
 different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
-whether the IP address of the client is present in the listed RBL\@.
+whether the \IP{} address of the client is present in the listed \RBL{}\@.
 
@@
 the name of another restriction Postfix will evaluate the new restriction,
-allowing restrictions to be chosen based on the client IP address, HELO
+allowing restrictions to be chosen based on the client \IP{} address, HELO
 hostname, sender address, recipient address, etc.  The administrator can
@@
     \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender, recipient, remote IP
+        rejects mail when the triple of (sender, recipient, remote \IP{}
         address) is unknown; on second and subsequent delivery attempts
@@
 Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using grep(1), awk(1)
+different from this parser: it attempts to present correlated log data in
+a form suitable for a systems administrator to search using grep(1), awk(1)
 and the myriad of other standard Unix text processing utilities already
@@
 format where a record is stored in a single line, with fields separated by
-a comma or other punctuation symbol.  Problems with CSV files include the
-difficulty in escaping separators, providing multiple values for a field
-(e.g.\ multiple recipients), adding new fields, and lack of a bundled
+a comma or other punctuation symbol.  Problems with \CSV{} files include
+the difficulty in escaping separators, providing multiple values for a
+field (e.g.\ multiple recipients), adding new fields, and lack of a bundled
 schema describing the fields.
@@
 data for that row.  The secondary by-sender, by-recipient, by-date and
-by-IP tables use the sender/recipient/date/IP as the key, and a \CSV{} list
-of integers referring to the main table is the value.  In effect this
-re-implements \SQL{} foreign keys, but without the functionality offered by
-even the most basic of \SQL{} databases (joins, ordering, searches, etc.).
-It also requires custom code to search on some combination of the above,
-though the authors of \LMA{} did provide some queries: IP-STORY,
-FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the
-least useful of the three output formats
-
-The data stored is limited to time and date, hostname and IP address of
+by-\IP{} tables use the sender/recipient/date/\IP{} as the key, and a
+\CSV{} list of integers referring to the main table is the value.  In
+effect this re-implements \SQL{} foreign keys, but without the
+functionality offered by even the most basic of \SQL{} databases (joins,
+ordering, searches, etc.).  It also requires custom code to search on some
+combination of the above, though the authors of \LMA{} did provide some
+queries: IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB
+appears to be the least useful of the three output formats
+
+The data stored is limited to time and date, hostname and \IP{} address of
 client, sender and recipient addresses, and the \SMTP{} code.  Handling of
@@
     \item Discard some of the data; this option is obviously undesirable,
-        and presumably was not chosen by the LMA authors.
-
-    \item Combine all data into a CSV line, which is then embedded into the
-        main CSV line for that mail, making searching even more difficult
-        and requiring a second separating character.
+        and presumably was not chosen by the \LMA{} authors.
+
+    \item Combine all data into a \CSV{} line, which is then embedded into
+        the main \CSV{} line for that mail, making searching even more
+        difficult and requiring a second separating character.
 
@@
 The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an SQL database, unlike with Berkeley
-DB\@; all \SQL{} databases embed the database schema into the database and
-provide commands for displaying it.  Berkeley DB does not embed a schema,
-as there is neither requirement nor benefit; it provides \textbf{(key,
-value)} pairs, so any additional structuring of the data is imposed by the
-application, thus it behoves the application to document this structure.
+possible to discover the schema with an \SQL{} database, unlike with
+Berkeley DB\@; all \SQL{} databases embed the database schema into the
+database and provide commands for displaying it.  Berkeley DB does not
+embed a schema, as there is neither requirement nor benefit; it provides
+\textbf{(key, value)} pairs, so any additional structuring of the data is
+imposed by the application, thus it behoves the application to document
+this structure.
 
@@
 multiple values).  The default database it uses is SQLite3, which supports
-most standard SQL features (its lacks full foreign key support), has
+most standard \SQL{} features (its lacks full foreign key support), has
 libraries available for most common programming languages, and a simple
@@
 % Hard to extend/expand/understand.  Appears to only save: date and hour,
-% DNS name and IP address host, mail server IP address, sender, receiver
-% and e-mail status (sent, rejected).  Undocumented schema.  Design
-% decision to use CSV as an intermediate format between the log file and
-% the database seems to have been restrictive.  Appears to require a
-% queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting IP addresses when parsing logs, but whitelisting
+% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
+% receiver and e-mail status (sent, rejected).  Undocumented schema.
+% Design decision to use \CSV{} as an intermediate format between the log
+% file and the database seems to have been restrictive.  Appears to require
+% a queueid but majority of log entries (e.g. rejections) lack a queueid.
+% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
 % when generating reports/data mining would be preferable.  Supporting
@@
 
-    \item [server\_ip] The IP address (IPv4 or IPv6) of the server: the
+    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
         local server when receiving mail, the remote server when sending
@@
     \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the IP address could not be resolved to a
-        hostname via DNS\@.
-
-    \item [client\_ip] The client IP address (IPv4 or IPv6): the remote
+        \texttt{unknown} if the \IP{} address could not be resolved to a
+        hostname via \DNS{}\@.
+
+    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
         server when receiving mail, the local server when sending mail.
@@
     \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the IP address could not be resolved to a
-        hostname via DNS\@.
+        \texttt{unknown} if the \IP{} address could not be resolved to a
+        hostname via \DNS{}\@.
 
@@
     \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an RBL\@.
+        fields, e.g.\ the rejection message from an \RBL{}\@.
 
@@
 
-    \item Sort by regex and visually inspect the list, e.g.\ with SQL
+    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
         similar to: \textbf{select regex from rules order by regex;}
@@
         connection\_data:] are \newline used by the action to extract and
-        save data matched by the regex.
+        save data matched by the \regex{}.
 
@@
 cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the regex.  Caching the compiled \regexs{}
-is obviously far more efficient; graph~\refwithpage{normal regex vs
-discarded regex factor} shows the percentage execution time increase when
-not caching each regex.
+times with and without caching the \regex{}.  Caching the compiled
+\regexs{} is obviously far more efficient; graph~\refwithpage{normal regex
+vs discarded regex factor} shows the percentage execution time increase
+when not caching each \regex{}.
 
@@
     \item This is followed by rejection of sender address, recipient
-        addresses, client IP address or hostname, etc. (action: rejection;
-        flowchart:~4); acceptance of one or more mails (action: clone;
-        flowchart:~5); or some interleaving of both.
+        addresses, client \IP{} address or hostname, etc. (action:
+        rejection; flowchart:~4); acceptance of one or more mails (action:
+        clone; flowchart:~5); or some interleaving of both.
 
@@
         table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and IP address.
+        client hostname and \IP{} address.
 
@@
 reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and IP address, but the database schema
+available is the client hostname and \IP{} address, but the database schema
 requires many more fields be populated (see
@@
         Fields extracted: from, to, server, UID, relay, status, number of
-        recipients, origin hostname, origin IP and virus.  The fields
+        recipients, origin hostname, origin \IP{} and virus.  The fields
         \textit{server}, \textit{uid\/} and \textit{virus\/} are not
@@
         Fields extracted: from, to, \RBL{}, client hostname and client
-        IP\@.
+        \IP{}\@.
 
         \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
-        \newline Fields extracted: from, to, client hostname, client IP,
-        relay hostname, relay IP, status, response code, \RBL{} and
+        \newline Fields extracted: from, to, client hostname, client \IP{},
+        relay hostname, relay \IP{}, status, response code, \RBL{} and
         message id.
@@
         that restricts it to supporting the Lowest Common Denominator.  The
-        data it will extract from an MTA log file is limited to: \newline
-        \hspace*{2em} time2, email, email\_r, host, host\_r, method, url,
-        code and bytesd.  \newline  There does not seem to be an
-        explanation of those fields in the documentation.  AWStats has no
-        special purpose code to deal with the intricacies of Postfix logs,
-        in fact it operates by coercing Postfix logs into
+        data it will extract from an \MTA{} log file is limited to:
+        \newline \hspace*{2em} time2, email, email\_r, host, host\_r,
+        method, url, code and bytesd.  \newline  There does not seem to be
+        an explanation of those fields in the documentation.  AWStats has
+        no special purpose code to deal with the intricacies of Postfix
+        logs, in fact it operates by coercing Postfix logs into
         Apache\footnote{The Apache web server is the most popular HTTP
@@
         class objects with a clear interface. There are eight useful
-        analyse modules, writing the result in plain ASCII or HTML, to
+        analyse modules, writing the result in plain ASCII or \HTML{}, to
         stdout or to files.\/}
@@
 
-        YAALA provides an SQL like mini-language used when generating the
-        reports; sample reports can be seen at~\cite{yaala-samples}.
+        YAALA provides an \SQL{} like mini-language used when generating
+        the reports; sample reports can be seen at~\cite{yaala-samples}.
         Example query: \newline \hspace*{2em} \texttt{requests BY file
@@
 ordering is shown in section~\refwithpage{rule ordering graphs}, and the
-anomalous dips and peaks apparent in some graphs are explained.  The next
+anomalous dips and peaks apparent in some graphs are explained.  The third
 group of graphs vividly shows the huge impact that caching compiled
@@
 
-The Y axis in the graph above represents the following:
+The Y axis in graph~\refwithpage{execution time vs file size vs number of
+lines graph} represents the following:
 
@@
 \end{enumerate}
+
+Graph~\refwithpage{execution time vs file size vs number lines factor}
+shows the ratio of execution time vs file size and number of lines (higher
+is better, it means more bytes or lines processed per second).  The ratios
+are quite tightly banded with the exception of log files 22 and 62--68,
+where they are noticeably higher; graph~\refwithpage{execution time vs file
+size vs number lines factor} and table~\refwithpage{execution time vs file
+size vs number lines factor table} show that the parser's execution time
+scales linearly with input size.
 
@@
 \newline This instructs Postfix to deliver the mail to the local user, and
-also forward to the remote address; this is generally not a problem except
-that the remote address in this case is the user's address, creating an
-infinite loop.  To prevent this happening Postfix examines the Delivered-To
-header in the mail and if the mail has already been delivered to the
-current address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately the user noticed they had not received any mail in
-a while, and opted to send a test mail to themselves, causing a loop not
-caught by Postfix as described below.
+also forward it to the remote address; this is generally not a problem
+except that the remote address in this case is the user's address, creating
+an infinite loop.  To prevent this happening Postfix examines the
+Delivered-To header in the mail, and if the mail has already been delivered
+to the current address it is bounced back to the sender with the error
+message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
+works well, but unfortunately in this case the user noticed they had not
+received any mail in a while and opted to send a test mail to themselves,
+causing a loop not caught by Postfix as described below.
 
@@
 
-    \item Postfix delivers the bounce notification to both the local
-        mailbox and to username@domain, as instructed by the user's
-        forwarding instructions.  A \texttt{Delivered-To:~username@domain}
-        header is added to the forwarded bounce notification, which now has
-        an envelope sender address of username@domain.  Log lines are added
-        by \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup}, and
-        finally \daemon{pickup}.
+    \item Postfix accepts the bounce notification and delivers it to both
+        the local mailbox and to username@domain, as instructed by the
+        user's forwarding instructions.  A
+        \texttt{Delivered-To:~username@domain} header is added to the
+        forwarded bounce notification, which now has an envelope sender
+        address of username@domain.  Log lines are added by \daemon{local},
+        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
+        \daemon{pickup}.
 
@@
 
-The sequence described above occurs extremely rapidly, as Postfix does not
-have to deliver the mail to an external system, so mails are delivered,
+The sequence described above occurs extremely rapidly because Postfix does
+not have to deliver the mail to an external system, so mails are delivered,
 bounced and generated as fast as the disks can keep up, resulting in a huge
@@
 components which have a small number of rules associated with them, whereas
-in general the most common log lines are logged by \daemon{smtpd}, which
-also has the highest number of rules.  \daemon{smtpd} log lines are
-distributed across rules much more evenly that \daemon{qmgr},
-\daemon{local}, \daemon{cleanup} or \daemon{pickup}, so the average number
-of rules required to parse a \daemon{smtpd} log line is much higher that
-the average number required to parse other log lines.  Thus the average
-number of rules required to parse a random line from logs containing a mail
-loop is much lower than the average number required to parse a random log
-line from logs without a mail loop, as can be seen in the peaks in
-graphs~\refwithpage{execution time vs file size vs number lines
-factor},~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal}.  The
-difference is more extreme in the reversed ordering case because the
-largest number of rules to be consulted for the commonly occurring log
-lines during a mail loop is far less than the largest number of rules that
-are consulted for commonly occurring lines outside of a mail loop.  The
-number of rules which needs to be consulted when the ordering is shuffled
-varies between the optimum and nadir, and the performance varies
-proportionally.
+in general \daemon{smtpd} adds the majority of log lines, and also has the
+highest number of rules.  \daemon{smtpd} log lines are distributed across
+rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
+\daemon{cleanup} or \daemon{pickup}, so the average number of rules
+required to parse a \daemon{smtpd} log line is much higher that the average
+number required to parse other log lines.
+
+These two characteristics combine to reduce the average number or rules
+required to parse a log line when there is a mail loop, as shown by the
+peaks in graph~\refwithpage{execution time vs file size vs number lines
+factor}.  When the rule ordering is reversed the majority of log lines
+generated by a mail loop will be parsed with very few rules, whereas
+without a mail loop the majority of log lines require a large number of
+rules; this leads to a noticeable drop in the average time required to
+parse a log line, as shown in graph~\refwithpage{percentage increase of
+reversed over normal}.  The number of rules which need to be consulted when
+the ordering is shuffled varies between the optimum and nadir, and the
+performance varies proportionally.
 
@@
 
-The distribution of log lines across rules is much different and the
-average number of rules consulted per log line is much lower when there is
-a mail loop; this results in far fewer \regex{} compilations per line than
+The distribution of log lines across rules when there is a mail loop is
+much different and the average number of rules consulted per log line is
+much lower; this results in far fewer \regex{} compilations per line than
 when there isn't a mail loop, and a correspondingly decreased execution
@@
 section~\refwithpage{rule efficiency}.  The first collection of graphs are
-about parser scalability, showing the linearly relationship between
-execution time and input size.  Section~\refwithpage{rule ordering graphs}
-relates to the effect of rule ordering on execution time, and the
-unexpected consequences of specific inputs.  The necessity of caching
-compiled \regexs{} is attested to by the third group of graphs, where the
-difference between caching and discarding is staggering.  This section
-concluded with a breakdown of the rule hits accumulated during a single
-test run of the parser.
+about parser scalability, showing the linear relationship between execution
+time and input size.  Section~\refwithpage{rule ordering graphs} relates to
+the effect of rule ordering on execution time, and the unexpected
+consequences of specific inputs.  The necessity of caching compiled
+\regexs{} is attested to by the third group of graphs, where the difference
+between caching and discarding is staggering.  This section concluded with
+a breakdown of the rule hits accumulated during a single test run of the
+parser.
 
@@
 
+\glossary{name={IPv4},description={
+    XXX IPv4.
+}}
+
+\glossary{name={IPv6},description={
+    XXX IPv6.
+}}
+
 \end{document}

### 2007-11-15 doc/logparser-acronyms.tex
Dealt with a few TODO items, but added lots more. Added lots more references in the process.

@@
 
-\newacronym{IP}{Internet Protocol}{description={
-    XXX ADD A DEFINITION\@.
+\newacronym{IP}{Internet Protocol address}{description={
+    The Internet Protocol~\cite{wikipedia-ip} is the protocol used to
+    communicate between computers on the Internet.  An IP address is a
+    unique address assigned to a computer on the Internet, allowing it to
+    communicate with other computers on the Internet.  For more information
+    see~\cite{wikipedia-ip-address}.
 }}
+
+\newacronym{PDF}{Portable Document Format}{description={
+    PDF is the file format created by Adobe Systems in 1993 for document
+    exchange.  It aims to be device independent, so documents should look
+    the same whether viewed on screen or printed.
+}}

### 2007-11-15 doc/logparser.tex
Dealt with a few TODO items, but added lots more. Added lots more references in the process.

@@
 REVIEWED
+
+XXX MOVE plot-* TO A SUBDIRECTORY\@?
 
@@
                 TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms, e.g.~\cite{wikipedia-sql}
-                XXX IMPROVE.}
+                for explanations of these terms,
+                e.g.~\cite{sql-for-web-nerds}}
 
@@
         differences between British English, American English and
-        Australian English).  XXX CITE SOME RESOURCES\@.
+        Australian English).  A good introduction to \SQL{} can be found
+        at~\cite{sql-for-web-nerds}, others
+        are~\cite{w3schools-sql-tutorial, sqlcourse.com}.
 
@@
 details of the table containing the rules have already been described in
-section~\refwithpage{rule attributes},).  This section finishes with a
+section~\refwithpage{rule attributes},).  This section continues with a
 discussion of rule efficiency concerns, referring to the graphs in
-section~\refwithpage{graphs}.
+section~\refwithpage{graphs}, and finishes with an explanation of how to
+add or edit rules, including the algorithm used to generate a new \regex{}
+from unparsed lines.
 
@@
 attempted to do this: such a project by itself would probably qualify for a
-PhD, and may involve solving the Halting Problem and circumventing the
-Church-Turing Thesis.  XXX REFERENCES FOR HALTING PROBLEM AND
-CHURCH-TURING\@.} that responsibility is left to the author of the rules.
-Unintentionally overlapping rules lead to inconsistent parsing and data
-extraction because the order in which rules are tried against each line is
-unspecified, and the first matching rule wins.  Overlapping rules are
-frequently a requirement, allowing a more specific rule to match some lines
-and a more general rule to match the majority, e.g.\ separating \SMTP{}
-delivery to a specific sites from \SMTP{} delivery to the rest of the
-world.  The algorithm provides a facility for ordering overlapping rules:
-the priority field in each rule (defaults to zero).  Rules are sorted by
-priority, highest first, and then rules with the same priority are sorted
-by the number of successful matches when parsing the previous log file.
-Negative priorities may be useful for catchall rules.
+PhD, and may involve solving the Halting
+Problem~\cite{wikipedia-halting-problem} and circumventing the
+Church-Turing Thesis~\cite{wikipedia-church-turing-thesis}.} that
+responsibility is left to the author of the rules.  Unintentionally
+overlapping rules lead to inconsistent parsing and data extraction because
+the order in which rules are tried against each line is unspecified, and
+the first matching rule wins.  Overlapping rules are frequently a
+requirement, allowing a more specific rule to match some lines and a more
+general rule to match the majority, e.g.\ separating \SMTP{} delivery to a
+specific sites from \SMTP{} delivery to the rest of the world.  The
+algorithm provides a facility for ordering overlapping rules: the priority
+field in each rule (defaults to zero).  Rules are sorted by priority,
+highest first, and then rules with the same priority are sorted by the
+number of successful matches when parsing the previous log file.  Negative
+priorities may be useful for catchall rules.
 
@@
 \subsection{Example rule}
+
+\label{example rule}
 
@@
 
+\subsection{Creating new rules}
+
+The logs produced by Postfix differ from installation to installation, as
+administrators have the freedom to choose the subset of available
+restrictions which suits their needs, including using different \RBL{}
+services, policy servers, or custom rejection messages.  To facilitate
+this, the parser's design separates parsing rules from parsing actions:
+adding new actions is difficult, but adding new rules to parse new
+rejection messages is trivial (and occurs much more frequently).  The
+implementation provides three programs to ease the process of adding new
+rules: \texttt{add-new-rule}, \texttt{edit-rule} and \texttt{logs2regexs}
+XXX NEED BETTER NAMES\@.
+
+\texttt{add-new-rule} is the simplest of the three, it just starts the
+user's preferred editor with a template rule which the user modifies as
+required, and then inserts the new rule into the database.  The various
+fields have already been described in sections~\refwithpage{rule
+attributes},~\refwithpage{rule characteristics} and~\refwithpage{example
+rule}, and will not be repeated here.
+
+\texttt{edit-rule} extracts a rule from the database, saves it in a human
+readable form, and runs the user's preferred editor for the user to change
+the rule as required.  Once the user is satisfied with the changes, the
+database will be updated appropriately.
+
+\texttt{logs2regexs}
+
 \subsection{Conclusion}
@@
 
-    \item The final part of this section discusses rule efficiency,
-        covering the effects of caching compiled \regexs{} and optimal
-        ordering of rules, with reference to the graphs in
+    \item The topic of rule efficiency was discussed next, covering the
+        effects of caching compiled \regexs{} and optimal ordering of
+        rules, with reference to the graphs in
         appendix~\refwithpage{graphs}.
+
+    \item This section finished with an explanation of how to add or edit
+        rules, including the algorithm used to generate a new \regex{} from
+        unparsed lines.
 
@@
 
+    \item The parser does not distinguish between logs produced by
+        different sources when parsing; all results will be saved to the
+        same database.  This may be viewed as an advantage, as logs from
+        different sources can be combined in the same database, or it may
+        be viewed as a limitation as there is no facility to distinguish
+        between logs from different sources in the same database.  If the
+        results of parsing logs from different sources must remain
+        separate, the parser can easily be instructed to use a different
+        database to store the results in.
+
 \end{enumerate}
@@
 
-NEED A BETTER TERM THAN HARD-CODED, THOUGH THAT MIGHT ARISE FROM THE
+XXX NEED A BETTER TERM THAN HARD-CODED, THOUGH THAT MIGHT ARISE FROM THE
 PREVIOUS TODO ITEM\@.
@@
         relays most used, etc.\/}; notably rejections are not mentioned.
-        It supports multiple output formats (text, \HTML{}, PDF XXX and
+        It supports multiple output formats (text, \HTML{}, \PDF{} and
         Excel 95).  Logs can be imported and the extracted data saved, and
@@
 \glossary{name={bounce},description={
-    The definitive documentation is \url{http://www.postfix.org/bounce.8.html}.
+    XXX The definitive documentation is \url{http://www.postfix.org/bounce.8.html}.
 }}
@@
 \glossary{name={local},description={
-    The definitive documentation is \url{http://www.postfix.org/local.8.html}.
+    XXX The definitive documentation is \url{http://www.postfix.org/local.8.html}.
 }}
@@
 \glossary{name={postdrop},description={
-    The definitive documentation is \url{http://www.postfix.org/postdrop.1.html}.
+    XXX The definitive documentation is \url{http://www.postfix.org/postdrop.1.html}.
 }}
@@
 \glossary{name={postsuper},description={
-    The definitive documentation is available at \url{http://www.postfix.org/postsuper.1.html}.
+    XXX The definitive documentation is available at \url{http://www.postfix.org/postsuper.1.html}.
 }}
@@
 \glossary{name={smtp},description={
-    The definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
+    XXX The definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
 }}
@@
 \glossary{name={IPv4},description={
-    XXX IPv4.
+    Internet Protocol~\cite{wikipedia-ipv4} version 4 is the fourth version
+    of the Internet Protocol used to interconnect computers on the
+    Internet.  It is the first widely deployed version of IP, and has been
+    in use for over 25 years.
 }}
@@
 \glossary{name={IPv6},description={
-    XXX IPv6.
+    Internet Protocol~\cite{wikipedia-ipv6} version 6 is the latest version
+    of the Internet Protocol used to interconnect computers on the
+    Internet.  It is the successor to IPv4, bringing with it a greatly
+    expanded address space, allowing many more computers to use the
+    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
+    come as existing networks transition from the former to the latter.
 }}

### 2007-11-15 doc/logparser-acronyms.tex
Capitalised Wikipedia so it wouldn't be highlighted as a spelling error any more.  Silly?  Possibly.

@@
     and maintenance.  Some information about SQL, including its history,
-    can be found at \cite{wikipedia-sql}.
+    can be found at \cite{Wikipedia-sql}.
 }}
@@
 \newacronym{SPF}{Sender Policy Framework}{description={
-    SPF is explained fully in~\cite{openspf, wikipedia-spf}.
+    SPF is explained fully in~\cite{openspf, Wikipedia-spf}.
 }}
@@
     Regular expressions are a method of matching patterns in text that are
-    explained in detail in~\cite{wikipedia-regex, perlre} .
+    explained in detail in~\cite{Wikipedia-regex, perlre} .
 }}
@@
     HTML is the markup language designed for writing web
-    pages~\cite{wikipedia-html, w3schools-html}.
+    pages~\cite{Wikipedia-html, w3schools-html}.
 }}
@@
 \newacronym{IP}{Internet Protocol address}{description={
-    The Internet Protocol~\cite{wikipedia-ip} is the protocol used to
+    The Internet Protocol~\cite{Wikipedia-ip} is the protocol used to
     communicate between computers on the Internet.  An IP address is a
@@
     communicate with other computers on the Internet.  For more information
-    see~\cite{wikipedia-ip-address}.
+    see~\cite{Wikipedia-ip-address}.
 }}

### 2007-11-15 doc/logparser.tex
Capitalised Wikipedia so it wouldn't be highlighted as a spelling error any more.  Silly?  Possibly.

@@
 
-    \item Checking \SPF{} records~\cite{openspf, wikipedia-spf}.  \SPF{}
+    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.  \SPF{}
         records specify which mail servers are allowed to send mail
@@
         backscatter~\cite{postfix-backscatter} and joe
-        jobs~\cite{wikipedia-joe-job}; however there has been a lot of
+        jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
         resistance to the proposal because it breaks or vastly complicates
@@
 PhD, and may involve solving the Halting
-Problem~\cite{wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{wikipedia-church-turing-thesis}.} that
+Problem~\cite{Wikipedia-halting-problem} and circumventing the
+Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
 responsibility is left to the author of the rules.  Unintentionally
@@
 \glossary{name={Phishing},description={
-    Phishing~\cite{wikipedia-phishing} is an attempt to acquire information
+    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
     by masquerading as an entity trusted by the user, e.g.\ a bank.
@@
 \glossary{name={IPv4},description={
-    Internet Protocol~\cite{wikipedia-ipv4} version 4 is the fourth version
+    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
     of the Internet Protocol used to interconnect computers on the
@@
 \glossary{name={IPv6},description={
-    Internet Protocol~\cite{wikipedia-ipv6} version 6 is the latest version
+    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
     of the Internet Protocol used to interconnect computers on the

### 2007-11-18 doc/logparser.tex
Finished the glossary entries for Postfix components.

@@
 
-The \SMTPlong, originally defined in \RFC{} 821~\cite{RFC821} and updated
-in \RFC{} 2821~\cite{RFC2821}, is used for transferring mail between the
+The \SMTPlong, originally defined in \RFC{}~821~\cite{RFC821} and updated
+in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
 sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
@@
 the principle of \textit{Be liberal in what you accept, and conservative in
-what you send\/} from \RFC{} 760~\cite{rfc760}, but unfortunately that
+what you send\/} from \RFC{}~760~\cite{rfc760}, but unfortunately that
 principle was written in a friendlier time.}  A typical \SMTP{}
@@
 The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{}, generate a \DFA{} from it, and
-examine every \DFA{} to determine if any overlap.  The author has not
-attempted to do this: such a project by itself would probably qualify for a
-PhD, and may involve solving the Halting
+possible to parse each rule's \regex{}, and determine if any overlap.  The
+author has not attempted to do this: such a project by itself would
+probably qualify for a PhD, and may involve solving the Halting
 Problem~\cite{Wikipedia-halting-problem} and circumventing the
@@
 \glossary{name={bounce},description={
-    XXX The definitive documentation is \url{http://www.postfix.org/bounce.8.html}.
+    The bounce daemon is responsible for sending bounce notifications in
+    Postfix versions later than 2.2.  The definitive documentation is
+    \url{http://www.postfix.org/bounce.8.html}.
 }}
@@
 \glossary{name={local},description={
-    XXX The definitive documentation is \url{http://www.postfix.org/local.8.html}.
+    Local is the Postfix component responsible for local delivery of mail
+    (i.e.\ delivered on the server Postfix is running on), whether it be to
+    a user's mailbox or a program such as a mailing list manager or
+    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
+    processing of a user's \texttt{.forward} file.  The definitive
+    documentation is \url{http://www.postfix.org/local.8.html}.
 }}
@@
 \glossary{name={postdrop},description={
-    XXX The definitive documentation is \url{http://www.postfix.org/postdrop.1.html}.
+    Postdrop is used when submitting mail locally on the server: it creates
+    a new mail in the queue and copies its input into the mail.  Subsequent
+    delivery of the mail is the responsibility of other Postfix components.
+    The definitive documentation is
+    \url{http://www.postfix.org/postdrop.1.html}.
 }}
@@
 \glossary{name={postsuper},description={
-    XXX The definitive documentation is available at \url{http://www.postfix.org/postsuper.1.html}.
+    Maintenance task such as deleting mails from the queue, putting mail on
+    hold (no further delivery attempts will be made until it is released
+    from hold, also by postsuper), and consistency checking of the mail
+    queue.  The definitive documentation is available at
+    \url{http://www.postfix.org/postsuper.1.html}.
 }}
@@
     Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program
-    that all Unix commands which send mail depend on.  The definitive
-    documentation is \url{http://www.postfix.org/sendmail.1.html}.
+    (\url{http://www.sendmail.org/}) mail submission program that all Unix
+    commands which send mail depend on; Postfix sendmail executes postdrop
+    to place a new mail in the queue.  The definitive documentation is
+    \url{http://www.postfix.org/sendmail.1.html}.
 }}
@@
 \glossary{name={smtp},description={
-    XXX The definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
+    Delivery of mail over \SMTP{} is performed by the smtp component.  The
+    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
 }}
@@
     a manager, the Bayesian spam filter might tip the balance from bad to
-    good, because the non-spam words might outweigh the spam words.
+    good, if the non-spam words outweigh the spam words.
 }}
@@
     \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address.  All mail servers must accept mail sent from $<>$, or they are
-    in violation of \SMTP{}.
+    address, but is always written as $<>$ for clarity.  All mail servers
+    must accept mail sent from $<>$, or they are in violation of
+    \RFC{}~2821~\cite{RFC2821}.
 }}

### 2007-11-18 doc/logparser.tex
Dealt with some more todo items and improved the other parsers section some more.

@@
 REVIEWED
-
-XXX MOVE plot-* TO A SUBDIRECTORY\@?
 
@@
 
-    \item Postfix sometimes forcibly kills a \daemon{smtpd} XXX FIGURE OUT
-        WHY, so the active connection must be cleaned up, entered in the
-        database and deleted from the connections state table.
+    \item Sometimes an \daemon{smtpd} is killed by a signal, so the active
+        connection must be cleaned up, entered in the database and deleted
+        from the connections state table.
 
@@
 an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.
+program with \parsername{}.  One important difference between \parsername{}
+and the parsers reviewed is that only \parsername{} makes it possible to
+parse new log lines without modifying the core parsing algorithm.  Although
+this may seem like a trivial point, is it substantially more difficult to
+understand a program's entire parsing algorithm, identify the correct
+location to change, and make the appropriate changes without adversely
+affecting parsing, particularly as there may be edge cases which are not
+immediately obvious.\footnote{See
+section~\refwithpage{yet-more-aborted-delivery-attempts} for a complication
+which occurs only four times in 93 log files tested.}  Requiring changes to
+the parsing algorithm also complicates upgrades, as the changes must be
+preserved, and may clash with changes made by the developer.  \parsername{}
+allows the user to add new rules to the database without changing the
+parsing algorithm, unless the new log lines to be parsed require
+functionality not already provided by the algorithm.  If the new log lines
+do require new functionality, new actions can be added to the parser
+without modifying existing actions or other parts of the algorithm; only in
+the rare case that the new actions require support from other sections of
+the code will more extensive changes be required.
+
+XXX ADD THE PREVIOUS TWO SENTENCES TO THE PARSING RULES/ALGORITHM
+SECTION\@.
 
@@
 XXX COMPARE AND CONTRAST RATHER THAN REVIEW\@.
-
-XXX HIGHLIGHT THE DIFFERENCE BETWEEN ADDING NEW RULES AND NEW PARSER
-FUNCTIONALITY\@.
-
-XXX NEED A BETTER TERM THAN HARD-CODED, THOUGH THAT MIGHT ARISE FROM THE
-PREVIOUS TODO ITEM\@.
 
@@
         Pflogsumm produces a report designed for troubleshooting, and does
-        not support saving any data; parsing is fixed and concentrates on
-        extracting only the data required to produce the report.  Both the
-        parsing and reporting are difficult to extend, as it is a
-        specialised tool.  It does not attempt to correlate log lines by
-        queueid or \pid{}, nor does it need to deal with the complications
-        encountered during this project.  Pflogsumm produces a useful
-        report, and successfully dealt with all the Postfix logs tested
-        with.
+        not support saving any data; parsing concentrates on extracting
+        only the data required to produce the report.  Both the parsing and
+        reporting are difficult to extend, as it is a specialised tool.  It
+        does not attempt to correlate log lines by queueid or \pid{}, nor
+        does it need to deal with the complications encountered during this
+        project.  Pflogsumm produces a useful report, and successfully
+        dealt with all the Postfix logs tested with.
 
@@
         the line, and extracts any \texttt{key=<value>} pairs; these are
-        the fields listed above.  Searches can be made based on the
-        extracted fields, and all text in the line is also available for
-        searching.  Many types of reports are available, though all are
-        variations of a bar or pie chart, with the exception of bubble and
-        heatmap graphs.  It is very easy to drill down through the graphs
-        to extract a portion of the data (e.g.\ select the hour with the
-        largest number of events, then select a particular host, and
-        finally a specific address).  It is not possible to search on
-        partial words, though the web interface does provide and refine a
-        list of possible search terms as you type.  The searches are quite
-        fast, though as the free version has a limit on the amount of data
-        indexed per day the volume of data to be searched was merely 45mb;
-        the cheapest licensed version costs \$6000, and it still limits the
-        volume of data indexed per day.
+        the fields listed above.  Searches can be based on the extracted
+        fields, and all text in the line is also available for searching.
+        Many types of reports are available, though all are variations of a
+        bar or pie chart, with the exception of bubble and heatmap graphs.
+        It is very easy to drill down through the graphs to extract a
+        portion of the data (e.g.\ select the hour with the largest number
+        of events, then select a particular host, and finally a specific
+        address).  It is not possible to search on partial words, though
+        the web interface does provide and refine a list of possible search
+        terms as you type.  The searches are quite fast, though as the free
+        version has a limit on the amount of data indexed per day the
+        volume of data to be searched was merely 45mb; the cheapest
+        licensed version costs \$6000, and it still limits the volume of
+        data indexed per day.
 
@@
 
+        XXX TEST THIS\@: THE NEXT PARAGRAPH IS CRAP\@.
+
         Due to the paucity of data gathered by this tool the author has not
@@
         AWStats will produce simple graphs for many different services, but
-        that restricts it to supporting the Lowest Common Denominator.  The
-        data it will extract from an \MTA{} log file is limited to:
-        \newline \hspace*{2em} time2, email, email\_r, host, host\_r,
-        method, url, code and bytesd.  \newline  There does not seem to be
-        an explanation of those fields in the documentation.  AWStats has
-        no special purpose code to deal with the intricacies of Postfix
-        logs, in fact it operates by coercing Postfix logs into
-        Apache\footnote{The Apache web server is the most popular HTTP
-        server in use for the past 10 years; more information is available
-        at \url{http://httpd.apache.org/}} format log files, for analysis
-        by AWStats' HTTP log file parser.  The converting parser is hard
-        coded, hard to extend, and only deals with a small portion of the
-        log lines generated by Postfix, but it does correlate log lines by
-        queueid.
+        supporting many different services without special purpose code
+        restricts it to supporting the \LCD{}.  The data it will extract
+        from an \MTA{} log file is limited to: \newline \hspace*{2em}
+        time2, email, email\_r, host, host\_r, method, url, code and
+        bytesd.  \newline  There does not seem to be an explanation of
+        those fields in the documentation.  AWStats has no special purpose
+        code to deal with the intricacies of Postfix logs, in fact it
+        operates by coercing Postfix logs into Apache\footnote{The Apache
+        web server is the most popular HTTP server in use for the past 10
+        years; more information is available at
+        \url{http://httpd.apache.org/}} format log files, for analysis by
+        AWStats' HTTP log file parser.  The converting parser only deals
+        with a small portion of the log lines generated by Postfix, and
+        though it does correlate log lines by queueid, it does not deal
+        with any of the other complications described in this document.
 
@@
 
-        Anteater hasn't been updated since November 2003, and doesn't have
-        any English documentation so it's difficult, nigh impossible, to
-        accurately comment on what analysis is performs.  It did not run
-        successfully when tested, and its parsing would certainly be out of
-        date as Postfix has evolved considerably since this tool was last
-        updated.  As it neither ran successfully nor has documentation the
-        author can read it is difficult to comment on the features it
-        provides.
+        Anteater doesn't have any English documentation so it's difficult,
+        nigh impossible, to accurately comment on what analysis is
+        performs.  It did not run successfully when tested, and its parsing
+        would certainly be out of date as Postfix has evolved considerably
+        since this tool was last updated (November 2003).  As it neither
+        ran successfully nor has documentation the author can read a
+        detailed review cannot be provided.
 
@@
         be only slightly less work than starting from scratch, as both
-        input and output modules would need to be written; it may even be
-        more work to implement the parser within the constraints of
-        YAALA\@.  YAALA bundles a Postfix parser, which uses hard-coded
-        parsing to extract the following fields from specific log lines:
-
-        Aggregations: count(count is not explained), bytes (the sum of
-        bytes transferred).\newline Keyfields [sic]: date, hour, sender,
-        recipient, defer\_count, delay, incoming\_host, outgoing\_host.
+        input and output modules would need to be written specially; it may
+        even be more work to implement the parser within the constraints of
+        YAALA\@.  YAALA bundles a Postfix parser, which extracts the
+        following fields from specific log lines:
+
+        \hspace*{2em}Aggregations: count (not explained), bytes (sum of
+        bytes transferred).\newline \hspace*{2em}Keyfields [sic]: date,
+        hour, sender, recipient, defer\_count, delay, incoming\_host,
+        outgoing\_host.
 
@@
         WHERE host =\~{} Google}. \newline The mini-language is quite
-        limited, and cannot be used to extract data for external use,
-        merely to create reports.
+        limited and cannot be used to extract data for external use, merely
+        to create reports.
 
         YAALA as distributed does not attempt the in-depth parsing
-        explained in this document, and implementing it within the confines
-        of YAALA would be at least as much work as implementing it
-        separately.
-
-        In summary YAALA provides a Postfix parser which is not extensible
-        unless the user is prepared to change the source
-        code,\footnote{\parsername{} requires source code changes for some
-        new features, but parsing new log lines is accomplished by adding
-        new rules.} and reasonably flexible report generation, but no
-        facilities to extract data for use in other tools.
+        explained in this document, and implementing such parsing within
+        the confines of YAALA would be at least as much work as
+        implementing it separately.
+
+        In summary YAALA provides a Postfix parser which only handles the
+        most common Postfix log lines, provides reasonably flexible report
+        generation from the limited data extracted, but no facilities to
+        extract data for use in other tools.
 
@@
         the domains from which you receive or send the most emails, the
-        relays most used, etc.\/}; notably rejections are not mentioned.
-        It supports multiple output formats (text, \HTML{}, \PDF{} and
-        Excel 95).  Logs can be imported and the extracted data saved, and
-        reports generated from it later.  Accessing this data from another
-        application is undocumented; given the source code it should be
-        possible, with time, to access the data.
+        relays most used, etc.\/}; notably rejections are not mentioned or
+        handled.  It supports multiple output formats (text, \HTML{},
+        \PDF{} and Excel 95).  Logs can be parsed, the extracted data
+        saved, and reports generated from it later.  Accessing this data
+        from another application is undocumented; given the source code it
+        should be possible, with sufficient time and effort, to access the
+        data.
 
@@
         queueid, but not by \pid{}, so the complete list of recipients for
-        a mail should be available; however its hard coded parsing extracts
-        only part of the available data and makes no attempt to deal with
-        the other complications described in
-        sections~\refwithpage{complications} and~\refwithpage{additional
-        complications}.  It stores data, but not in an easily accessible
-        format, and generates reports in multiple formats, but those
-        reports do not appear to be customisable.
+        a mail should be available; however its parser extracts only part
+        of the available data and makes no attempt to deal with the other
+        complications described in sections~\refwithpage{complications}
+        and~\refwithpage{additional complications}.  It stores data, but
+        not in an easily accessible format, and generates reports in
+        multiple formats, but those reports do not appear to be
+        customisable.
 
@@
         sessionname [sic] within Logrep), but not by \pid{}.  The parsing
-        is hard coded, hard to extend, and error prone: empty fields are
-        saved when the log line doesn't match the \regex{}, though it
-        appears that they will not overwrite existing data.  It also counts
-        the number of log lines parsed and skipped.  Most notably
-        rejections are completely ignored, making it unsuitable for the
-        purposes of this project.  It doesn't attempt to address any of the
-        complications in sections~\refwithpage{complications}
-        and~\refwithpage{additional complications} except for correlating
-        by queueid.
+        is error prone: empty fields are saved when the log line doesn't
+        match the \regex{}, though it appears that they will not overwrite
+        existing data.  It also counts the number of log lines parsed and
+        skipped.  Most notably rejections are completely ignored, making it
+        unsuitable for the purposes of this project.  It doesn't attempt to
+        address any of the complications in
+        sections~\refwithpage{complications} and~\refwithpage{additional
+        complications} except for correlating by queueid.
 
@@
         was unable to access the log files selected; this appears to be a
-        bug rather than operator error.  As parsing failed it wasn't
-        possible to review the reports Logrep can generate (available in
-        \HTML{} only), nor to examine the (undocumented) format in which it
-        can save extracted data for subsequent reuse.
+        bug rather than operator error.\footnote{If it is caused by
+        operator error, the interface needs improvement as the (minimal)
+        instructions were followed as closely as possible.}  As parsing
+        failed it wasn't possible to review the reports Logrep can generate
+        (available in \HTML{} only), nor to examine the (undocumented)
+        format in which it can save extracted data for subsequent reuse.
 
@@
         completely ignores rejections, is effectively undocumented, doesn't
-        deal with the more complicated aspects of Postfix logs, cannot be
-        easily extended by the user, and at the time of writing doesn't
-        work properly.
+        deal with the more complicated aspects of Postfix logs, and at the
+        time of writing doesn't work properly.
 
         \url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-        2007/10/13.)
+        2007/11/18.)
 
@@
 the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser.  Most other
-parsers immediately generate a report and discard the data extracted from
-the logs; those which don't discard the data retain it in a format
-inaccessible to other tools.  All of the parsers reviewed can produce a
-report of greater or lesser detail and complexity, a facility not offered
-by \parsername{}; reporting is deferred to a subsequent application.
+\parsername{}, nor make it as easy to extend the parser to handle new log
+lines.  Most other parsers immediately generate a report and discard the
+data extracted from the logs; those which don't discard the data retain it
+in a format inaccessible to other tools.  All of the parsers reviewed can
+produce a report of greater or lesser detail and complexity, a facility not
+offered by \parsername{}; reporting is deferred to a subsequent
+application.
 

### 2007-11-20 doc/logparser.tex
More XXX items dealt with.

@@
 
+XXX WHERE SHOULD THE NEXT PARAGRAPH BE PLACED\@?
+
+\label{why separate rules and algorithm}
+
+Separating the rules and algorithm makes it possible to parse new log lines
+without modifying the core parsing algorithm.  Although this may seem like
+a trivial point, is it substantially more difficult to understand a
+program's entire parsing algorithm, identify the correct location to
+change, and make the appropriate changes without adversely affecting
+parsing, particularly as there may be edge cases which are not immediately
+obvious.\footnote{See
+section~\refwithpage{yet-more-aborted-delivery-attempts} for a complication
+which occurs only four times in 93 log files tested.}  Requiring changes to
+the parsing algorithm also complicates upgrades, as the changes must be
+preserved during the upgrade, and may clash with changes made by the
+developer.  \parsername{} allows the user to add new rules to the database
+without changing the parsing algorithm, unless the new log lines to be
+parsed require functionality not already provided by the algorithm.  If the
+new log lines do require new functionality, new actions can be added to the
+parser without modifying existing actions or other parts of the algorithm;
+only in the rare case that the new actions require support from other
+sections of the code will more extensive changes be required.
+
 \subsection{Introduction}
@@
 and the parsers reviewed is that only \parsername{} makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change, and make the appropriate changes without adversely
-affecting parsing, particularly as there may be edge cases which are not
-immediately obvious.\footnote{See
-section~\refwithpage{yet-more-aborted-delivery-attempts} for a complication
-which occurs only four times in 93 log files tested.}  Requiring changes to
-the parsing algorithm also complicates upgrades, as the changes must be
-preserved, and may clash with changes made by the developer.  \parsername{}
-allows the user to add new rules to the database without changing the
-parsing algorithm, unless the new log lines to be parsed require
-functionality not already provided by the algorithm.  If the new log lines
-do require new functionality, new actions can be added to the parser
-without modifying existing actions or other parts of the algorithm; only in
-the rare case that the new actions require support from other sections of
-the code will more extensive changes be required.
-
-XXX ADD THE PREVIOUS TWO SENTENCES TO THE PARSING RULES/ALGORITHM
-SECTION\@.
+parse new log lines without modifying the core parsing algorithm; see
+section~\refwithpage{why separate rules and algorithm} for a discussion of
+why the separation of rules from the algorithm is beneficial.
 
@@
 
-        XXX TEST THIS\@: THE NEXT PARAGRAPH IS CRAP\@.
-
-        Due to the paucity of data gathered by this tool the author has not
-        experimented with it, but based on the documentation it is obvious
-        that it doesn't attempt to gather the breadth of data required for
-        in-depth analysis, and presumably doesn't attempt to correlate log
-        lines per-connection or per-mail.
+        Isoqlog produces a report listing the number of mails sent by each
+        unique sender address, and separately the total number of bytes
+        transferred; both reports are produced for daily, monthly and
+        annual time spans, but only for the domains listed in its
+        configuration file.  It appears to ignore all log lines except for
+        those for the current day, though it does maintain a record of data
+        previously extracted, which the newly extracted data is merged into
+        (no information is provided on the format of the data store).  It
+        doesn't utilise rejection log lines in any way, so is unsuitable
+        for the purposes of this project.  It doesn't handle any of the
+        complications discussed in this document, doesn't gather the
+        breadth of data required for this project, and ignores the majority
+        of log lines produced by Postfix.
 

### 2007-11-22 doc/logparser-acronyms.tex
Finished the glossary entries for Postfix components.
Dealt with some more todo items and improved the other parsers section some more.
Further cleanup. Lots more XXX items. More work on logs2regexs documentation.

@@
 \newacronym{SPF}{Sender Policy Framework}{description={
-    SPF is explained fully in~\cite{openspf, Wikipedia-spf}.
+    SPF is introduced in section~\refwithpage{spf introduction} and
+    explained fully in~\cite{openspf, Wikipedia-spf}.
 }}
@@
         www.example.com is a FQDN
-}}
-
-\newacronym{DFA}{Deterministic Finite Automata}{description={
-    XXX EXPLAIN DFA.  SHOULD IT BE NFA?
 }}
@@
 }}
+
+\newacronym{LCD}{Lowest Common Denominator}{description={
+    LCD is a mathematical term which is used figuratively to mean the least
+    useful member of a set of alternatives.
+}}
+
+\newacronym{SLCT}{Simple Logfile Clustering Tool}{description={
+    SLCT~\cite{slct-paper} is a tool implementing an algorithm designed by
+    Risto Vaarandi~\cite{risto-vaarandi} for identifying and grouping
+    similar log files, then producing a regex for each group of lines.
+}}

### 2007-11-22 doc/logparser.tex
Ordered the actions alphabetically.
Further cleanup. Lots more XXX items. More work on logs2regexs documentation.

@@
 
-% When creating a PDF make the table of contents into links to the pages
+% When creating a \PDF{} make the table of contents into links to the pages
 % (without horrible red borders) and include bookmarks.  The title and
@@
     @%
+}
+
+\newcommand{\tab}[0]{%
+    \hspace*{2em}%
 }
@@
 discussed in section~\ref{rule efficiency}, rule efficiency, and explains
-the irregularities observed in the graphs.
+the irregularities observed in these graphs.
+
+Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
 
 Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
 
@@
 
-\subsection{SMTP background}
+\subsection{Simple Mail Transfer Protocol}
 \label{SMTP background}
 
-The \SMTPlong, originally defined in \RFC{}~821~\cite{RFC821} and updated
+The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
 in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
@@
 what you send\/} from \RFC{}~760~\cite{rfc760}, but unfortunately that
-principle was written in a friendlier time.}  A typical \SMTP{}
-conversation resembles the following (the lines starting with a three digit
-number are sent by the server, all other lines are sent by the client):
+principle was written in a friendlier time. XXX CARL WANTS ME TO SUGGEST A
+NEW PRINCIPAL.}  A typical \SMTP{} conversation resembles the following
+(the lines starting with a three digit number are sent by the server, all
+other lines are sent by the client):
 
@@
 
-\subsection{Postfix background}
+\subsection{Postfix}
 
@@
 this behaviour.}  Postfix uses simple lookup tables as the deciding factor
-for some restrictions, e.g.\
-\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
+for some restrictions, e.g.\ XXX IMPROVE THIS EXPLANATION OR REMOVE IT \\
+\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
 \footnote{\texttt{cidr}~\cite{cidr_table} is the type of the lookup table.}
-will check whether the \IP{} address of the connected client matches the
-left hand side of each line in the file and will return the right hand side
-of the first matching line.  Other restrictions determine their result in
-different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
-whether the \IP{} address of the client is present in the listed \RBL{}\@.
+\newline checks whether the \IP{} address of the connected client matches
+the left hand side of each line in the file and will return the right hand
+side of the first matching line.  Other restrictions determine their result
+in different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com}
+checks whether the \IP{} address of the client is present in the listed
+\RBL{}\@.
 
@@
 
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.  \SPF{}
+    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
+        \SPF{}\label{spf introduction}
         records specify which mail servers are allowed to send mail
@@
 
-\subsection{Other parsers}
+\subsection{Other Postfix log parsers}
 
@@
 
-Parsing in \LMA{} is hand-coded and requires major changes to the code to
-parse new log lines or extract additional data.  It does not appear to deal
-with any of the complications discussed in
-sections~\refwithpage{complications} and~\refwithpage{additional
-complications}.  The program fails to parse Postfix 2.2.x or 2.3.x log
-files; indeed it produces a number of warnings and no usable output.
-Parsing in \parsername{} is divided in two (see
+XXX IMPROVE THE NEXT PARAGRAPH BASED ON COMMENTS FROM CARL\@.
+
+Parsing in \LMA{} requires major changes to the code to parse new log lines
+or extract additional data.  It does not appear to deal with any of the
+complications discussed in sections~\refwithpage{complications}
+and~\refwithpage{additional complications}.  The program fails to parse
+Postfix 2.2 or 2.3 log files; indeed all it produces is a number of
+warning messages from the Perl interpreter, without any other output,
+useful or otherwise.  Parsing in \parsername{} is divided in two (see
 sections~\refwithpage{rules} and~\refwithpage{parsing-algorithm} for a more
@@
 didn't have an opportunity to review the reports.  \parsername{} does not
-have any bundled reports, it delegates that task to a separate, dedicated
+produce any reports, it delegates that task to a separate, dedicated
 program.
 
-In summary \LMA{} appears to be a proof of concept, written to demonstrate
-the point of the paper (that having this information in an accessible
+In summary, \LMA{} appears to be a proof of concept, written to demonstrate
+the point of their paper (that having this information in an accessible
 fashion is useful to systems administrators), rather than a program
@@
         \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \hspace*{2em} server\_hostname} and \texttt{helo.\newline}
+        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
         \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \hspace*{2em} pid\_regex, data} and \texttt{child.}
+        \newline \tab{} pid\_regex, data} and \texttt{child.}
 
@@
             \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline
-                \hspace*{2em}\texttt{/./~~~~WARN~Logging~HELO}
+                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
 
@@
                 \texttt{/etc/postfix/main.cf} to contain\newline
-                \hspace*{2em}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
+                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
 
@@
 rejection messages is trivial (and occurs much more frequently).  The
-implementation provides three programs to ease the process of adding new
-rules: \texttt{add-new-rule}, \texttt{edit-rule} and \texttt{logs2regexs}
-XXX NEED BETTER NAMES\@.
-
-\texttt{add-new-rule} is the simplest of the three, it just starts the
-user's preferred editor with a template rule which the user modifies as
-required, and then inserts the new rule into the database.  The various
-fields have already been described in sections~\refwithpage{rule
-attributes},~\refwithpage{rule characteristics} and~\refwithpage{example
-rule}, and will not be repeated here.
-
-\texttt{edit-rule} extracts a rule from the database, saves it in a human
-readable form, and runs the user's preferred editor for the user to change
-the rule as required.  Once the user is satisfied with the changes, the
-database will be updated appropriately.
-
-\texttt{logs2regexs}
+implementation provides a program to ease the process of creating new rules
+from unparsed lines, based on the algorithm developed by Risto
+Vaarandi~\cite{risto-vaarandi} for \SLCT{}~\cite{slct-paper}.  The
+differences between the two algorithms will be outlines as part of the
+general explanation below.
+
+The core of the algorithm was developed by Risto Vaarandi, and is quite
+simple: log lines are generally created by substituting variable words into
+a fixed pattern, and analysis of the frequency with which each word occurs
+determines whether the word is fixed or variable.  This classification can
+be used to group similar log lines and generate a \regex{} to match each
+group of lines.
+
+There are 4 steps in the algorithm:
+
+\begin{description}
+
+    \item [Pre-process the file]
+
+        In the original algorithm the preprocessing stage was
+
+    \item [Calculate the word frequency]
+
+    \item [XXX]
+
+    \item [Build regexs]
+
+\end{description}
 
@@
 
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
+    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
+        creation of bounce messages.  This action creates a new mail if
+        necessary; if the mail already exists the unknown origin flag will
+        be removed.  The action also marks the mail as a bounce
+        notification.
+
+    \item [CLONE] Multiple mails may be accepted on a single connection, so
+        each time a mail is accepted the connection's state table entry
+        must be cloned; if the original data structure was used the second
+        and subsequent mails would overwrite one another's data.
+
+    \item [COMMIT] Enter the data from the mail into the database. Entry
+        will be postponed if the mail is a child waiting to be tracked.
+        Once entered, the mail will be deleted from the state tables.
 
@@
         delete the connection from the state tables.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
 
@@
 
+    \item [IGNORE] This rule just returns successfully; it is used when a
+        line needs to be parsed for completeness but doesn't either provide
+        any useful data or require anything to be done.
+
     \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
@@
         lines} for details.
+
+    \item [MAIL\_TOO\_LARGE] When the client tries to send a message larger
+        than the local server accepts, the mail will be discarded, and the
+        client informed.  See TIMEOUT for further discussion; the two are
+        handled in exactly the same way.
 
@@
 
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned; if the original data structure was used the second
-        and subsequent mails would overwrite one another's data.
-
-    \item [MAIL\_TOO\_LARGE] When the client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there is not.  See
-        section~\refwithpage{timeouts-during-data-phase} for the gory
-        details.
-
     \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
@@
 
-    \item [SMTPD\_KILLED] Sometimes an event occurs which requires Postfix
-        to forcefully kill a \daemon{smtpd}; that \daemon{smtpd} processes'
-        active connection must be cleaned up, entered in the database, and
-        deleted from the state tables.
+    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
+        the remote client: log the rejection with a mail if there is a
+        queueid in the log line, or with the connection if not.
+
+    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
+        the log line, and save the data extracted by the \regex{} to it.
 
@@
         tables.
+
+    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
+        signal~\cite{Wikipedia-unix-signals}; the active connection for
+        that \daemon{smtpd} must be cleaned up, entered in the database,
+        and deleted from the state tables.
 
@@
 
-    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
-        creation of bounce messages.  This action creates a new mail if
-        necessary, marking it as a bounce notification and removing the
-        unknown origin flag if it exists.
+    \item [TIMEOUT] The connection timed out so the mail currently being
+        transferred must be discarded. The mail may have been accepted, in
+        which case there's a data structure to dispose of, or it may not in
+        which case there is not.  See
+        section~\refwithpage{timeouts-during-data-phase} for the gory
+        details.
+
+    \item [TRACK] Track a mail when it is re-injected for forwarding to
+        another mail server; this happens when a local address is aliased
+        to a remote address.  TRACK will be called when dealing with the
+        parent mail, and will create the child mail if necessary. TRACK
+        checks if the child has already been tracked, either by this parent
+        or by another parent, and issues appropriate warnings in either
+        case.
 
@@
         restricts it to supporting the \LCD{}.  The data it will extract
-        from an \MTA{} log file is limited to: \newline \hspace*{2em}
-        time2, email, email\_r, host, host\_r, method, url, code and
-        bytesd.  \newline  There does not seem to be an explanation of
-        those fields in the documentation.  AWStats has no special purpose
-        code to deal with the intricacies of Postfix logs, in fact it
-        operates by coercing Postfix logs into Apache\footnote{The Apache
-        web server is the most popular HTTP server in use for the past 10
-        years; more information is available at
-        \url{http://httpd.apache.org/}} format log files, for analysis by
-        AWStats' HTTP log file parser.  The converting parser only deals
-        with a small portion of the log lines generated by Postfix, and
-        though it does correlate log lines by queueid, it does not deal
-        with any of the other complications described in this document.
+        from an \MTA{} log file is limited to: \newline \tab{} time2,
+        email, email\_r, host, host\_r, method, url, code and bytesd.
+        \newline  There does not seem to be an explanation of those fields
+        in the documentation.  AWStats has no special purpose code to deal
+        with the intricacies of Postfix logs, in fact it operates by
+        coercing Postfix logs into Apache\footnote{The Apache web server is
+        the most popular HTTP server in use for the past 10 years; more
+        information is available at \url{http://httpd.apache.org/}} format
+        log files, for analysis by AWStats' HTTP log file parser.  The
+        converting parser only deals with a small portion of the log lines
+        generated by Postfix, and though it does correlate log lines by
+        queueid, it does not deal with any of the other complications
+        described in this document.
 
@@
 
-        \hspace*{2em}Aggregations: count (not explained), bytes (sum of
-        bytes transferred).\newline \hspace*{2em}Keyfields [sic]: date,
-        hour, sender, recipient, defer\_count, delay, incoming\_host,
-        outgoing\_host.
+        \tab{}Aggregations: count (not explained), bytes (sum of bytes
+        transferred).\newline \tab{}Keyfields [sic]: date, hour, sender,
+        recipient, defer\_count, delay, incoming\_host, outgoing\_host.
 
@@
         the reports; sample reports can be seen at~\cite{yaala-samples}.
-        Example query: \newline \hspace*{2em} \texttt{requests BY file
-        WHERE host =\~{} Google}. \newline The mini-language is quite
-        limited and cannot be used to extract data for external use, merely
-        to create reports.
+        Example query: \newline \tab{} \texttt{requests BY file WHERE host
+        =\~{} Google}. \newline The mini-language is quite limited and
+        cannot be used to extract data for external use, merely to create
+        reports.
 
@@
 
+XXX CAN I PUT \% ON THE Y AXIS\@?
+
 \showgraph{build/plot-normal-shuffle-factor.ps}{Percentage increase of
@@
 The mail loop was set up by a user modifying his mail forwarding to:
-\newline \hspace*{2em}\texttt{$\backslash$username, username@domain}
-\newline This instructs Postfix to deliver the mail to the local user, and
-also forward it to the remote address; this is generally not a problem
-except that the remote address in this case is the user's address, creating
-an infinite loop.  To prevent this happening Postfix examines the
-Delivered-To header in the mail, and if the mail has already been delivered
-to the current address it is bounced back to the sender with the error
-message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
-works well, but unfortunately in this case the user noticed they had not
-received any mail in a while and opted to send a test mail to themselves,
-causing a loop not caught by Postfix as described below.
+\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
+instructs Postfix to deliver the mail to the local user, and also forward
+it to the remote address; this is generally not a problem except that the
+remote address in this case is the user's address, creating an infinite
+loop.  To prevent this happening Postfix examines the Delivered-To header
+in the mail, and if the mail has already been delivered to the current
+address it is bounced back to the sender with the error message
+\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
+well, but unfortunately in this case the user noticed they had not received
+any mail in a while and opted to send a test mail to themselves, causing a
+loop not caught by Postfix as described below.
 
@@
 
-    \item GOTO 2.
+    \item GOTO 2.  XXX IMPROVE THIS, IT'S RUBBISH\@.
 

### 2007-11-25 doc/logparser.tex
Finished the explanation of the logs2regexs algorithm.

@@
 discussion of rule efficiency concerns, referring to the graphs in
-section~\refwithpage{graphs}, and finishes with an explanation of how to
-add or edit rules, including the algorithm used to generate a new \regex{}
-from unparsed lines.
+section~\refwithpage{graphs}, and finishes with an explanation of the
+algorithm used to generate a new \regex{} from unparsed lines.
 
@@
 optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time.  Perl automatically caches
-static \regexs{}, but dynamic \regexs{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexs{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
+compilation and optimisation takes CPU time; far more CPU time, in fact,
+than the actual matching takes.  Perl automatically caches static
+\regexs{}, but dynamic \regexs{} need to be explicitly compiled and cached.
+Graph~\refwithpage{normal regex vs discard regex} shows execution times
+with and without caching the \regex{}.  Caching the compiled \regexs{} is
+obviously far more efficient; graph~\refwithpage{normal regex vs discarded
+regex factor} shows the percentage execution time increase when not caching
+each \regex{}.
 
@@
 
-The logs produced by Postfix differ from installation to installation, as
-administrators have the freedom to choose the subset of available
+The logs produced by Postfix differ from installation to installation,
+because administrators have the freedom to choose the subset of available
 restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate
-this, the parser's design separates parsing rules from parsing actions:
-adding new actions is difficult, but adding new rules to parse new
-rejection messages is trivial (and occurs much more frequently).  The
-implementation provides a program to ease the process of creating new rules
-from unparsed lines, based on the algorithm developed by Risto
-Vaarandi~\cite{risto-vaarandi} for \SLCT{}~\cite{slct-paper}.  The
-differences between the two algorithms will be outlines as part of the
+services, policy servers, or custom rejection messages.  To facilitate easy
+parsing of new log lines, the parser's design separates parsing rules from
+parsing actions: adding new actions is difficult, but adding new rules to
+parse new rejection messages is trivial (and also occurs much more
+frequently).  The implementation provides a program to ease the process of
+creating new rules from unparsed lines, based on the algorithm developed by
+Risto Vaarandi~\cite{risto-vaarandi} for his \SLCT{}~\cite{slct-paper}.
+The differences between the two algorithms will be outlined as part of the
 general explanation below.
 
-The core of the algorithm was developed by Risto Vaarandi, and is quite
-simple: log lines are generally created by substituting variable words into
-a fixed pattern, and analysis of the frequency with which each word occurs
-determines whether the word is fixed or variable.  This classification can
-be used to group similar log lines and generate a \regex{} to match each
-group of lines.
+The core of the algorithm is quite simple: log lines are generally created
+by substituting variable words into a fixed pattern, and analysis of the
+frequency with which each word occurs can be used to determine whether the
+word is fixed or variable.  This classification can be used to group
+similar log lines and generate a \regex{} to match each group of lines.
 
@@
 
-    \item [Pre-process the file]
-
-        In the original algorithm the preprocessing stage was
-
-    \item [Calculate the word frequency]
-
-    \item [XXX]
-
-    \item [Build regexs]
+    \item [Pre-process the file]  The new algorithm leverages the knowledge
+        gained while writing rules and performs a large number of
+        substitutions on the input lines, replacing commonly occurring
+        variable terms (e.g.\ email addresses, \IP{} addresses, the
+        standard start of rejection messages, etc.) with \regex{} keywords
+        which the parser will expand when it loads the rule (see the
+        \regex{} entry in section~\refwithpage{rule attributes}).  The
+        purpose of this phase is to utilise the existing knowledge to
+        create more accurate \regexs{}.  The new log lines are written to a
+        temporary file, which all subsequent stages use instead of the
+        original input file.
+
+        In the original algorithm the purpose of the preprocessing stage
+        was to reduce the memory consumption of the program.  In the first
+        pass it generated a hash from a small range of values for each word
+        of each line, incrementing a counter for each hash.  The counters
+        will later be used to filter out words: if the word's hash does not
+        have a high frequency, the word itself cannot have a high
+        frequency, and there is no need to maintain a counter for it,
+        reducing the number of counters and thus the programme's memory
+        consumption.
+
+    \item [Calculate word frequencies]  The position of words within a line
+        is important: a common word does not indicate similarity between
+        lines unless it occupies the same position within both
+        lines.\footnote{If a variable term within a line contains spaces,
+        it will appear to the algorithm as two words rather than one.  This
+        will alter the position of subsequent words, so a word occurring in
+        different positions in two log lines may indicate similarity, but
+        the algorithm does not attempt to deal with this possibility.}  The
+        algorithm maintains a counter for each \textit{(word, word's
+        position within the line)\/} tuple, incrementing it each time that
+        word is occurs in that position.
+
+        The original algorithm only maintains counters for words whose hash
+        result has a high frequency from the previous phase; this reduces
+        the number of counters maintained by the algorithm, reducing the
+        memory requirements of the algorithm.  The modified algorithm omits
+        this check because the majority of unique or infrequently occurring
+        words will have been substituted with keywords during the first
+        phase, vastly reducing the number of tuples to maintain counters
+        for.
+
+    \item [Classify words based on their frequency]  The frequency of each
+        \textit{(word, word's position within the line)\/} tuple is
+        checked: if its frequency is greater than the threshold supplied by
+        the user (1\% of all lines is generally a good starting point) it
+        is classified as a fixed word, otherwise it is classified as a
+        variable term.  Variable terms are replaced by \texttt{.*}, which
+        means to match zero or more of any character.  The words are
+        reassembled to produce a \regex{} matching the line, and a counter
+        is maintained for each \regex{}.
+
+    \item [Build regexs]  Contiguous sequences of \texttt{.*} in the newly
+        reassembled \regexs{} are collapsed to a single \texttt{.*}; any
+        resulting duplicate \regexs{} are combined, and their counters
+        summed.  If the frequency of a \regex{} is lower than the threshold
+        supplied by the user the \regex{} is discarded.\footnote{This is a
+        second, independent threshold, but once again 1\% of input lines is
+        a good starting point; obviously the threshold depends on the
+        number and type of input lines.}  The new \regexs{} are printed for
+        the user to add to the database, either as new rules or merged into
+        the \regexs{} of existing rules; the counter for each \regex{} is
+        also printed, giving the user an indication of how many of the
+        input lines that \regex{} should match.  Discarding \regexs{} will
+        result in some of the input lines not being matched; this utility
+        should be run again once the unmatched lines have been isolated by
+        running the parser, including the new \regexs{}, with the same
+        input which produced the original set of unparsed lines.
 
 \end{description}
+
+A second utility is also provided which reads a list of new \regexs{} and
+the input given to the first utility.  It tries to match each input line
+against each \regex{}, counting the number of lines which match each
+\regex{}, warning the user if an input line is matched by more than one
+\regex{}, and additionally warning if an input line is not matched by any
+\regex{}.  It displays a summary of how many input lines each \regex{}
+matched, comparing it to the expected number of matches; this provides the
+user with an easy method of checking if the \regexs{} produced by the first
+utility are correctly matching the input lines they are based upon.  A
+future version of this utility will also group input lines by \regex{}, so
+the user can tweak the \regexs{} if required.
 
@@
 
-    \item This section finished with an explanation of how to add or edit
-        rules, including the algorithm used to generate a new \regex{} from
-        unparsed lines.
+    \item This section finished with an explanation of the algorithm used
+        to generate a new \regex{} from unparsed lines.
 
@@
 
+XXX SPLIT THIS IMAGE INTO TWO\@; IT'S UNREADABLE NOW, IT'S SO BIG\@.
+
 % This is very likely to move around as the flow chart is as big as
 % possible, so if any preceding text changes the flow chart may end up on
-% the next page, which may not be a bad thing, maybe.  What woould be a bad
+% the next page, which may not be a bad thing, maybe.  What would be a bad
 % thing is the \clearpage seems to cause a blank page sometimes, if the
@@
 
-    \item [matches] The fields in the line captured by the rule's regex.
+    \item [matches] The fields in the line captured by the rule's \regex{}.
 
@@
 
+\glossary{name={hash},description={
+    A hashing function transforms a string of characters to a number.
+    There are many possible uses for the resulting number: a common usage
+    is to maintain a data structure indexed by strings in an efficient
+    manner.  A full description is beyond the scope of this paper, further
+    information can be found at~\cite{hash-functions}.
+}}
+
 \end{document}

### 2007-11-25 doc/logparser.tex
Finished Carl's first set of corrections; some of them are added as XXX items, but at least I've gotten the vast majority done.

@@
 
+% This may keep hyperref happy about the Abstract entry in the table of
+% contents.
+\newcounter{dummy}
+
 \begin{document}
@@
 
+\refstepcounter{dummy}
+\addcontentsline{toc}{section}{Abstract}
 \begin{abstract}
@@
 
-Section~\ref{database schema} describes the database schema used by the
-parser, explaining in detail the tables used for storing the data gleaned
-from the logs and the table that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, describing the
-table in the database used to hold the rules and explaining their
-structure, referring to an example rule and sample data it matches
-successfully against.  This section concludes with a discussion of rule
-efficiency concerns.
+This algorithm requires a database for storing both the rules used when
+parsing and the results gleaned from parsing.  The database schema used is
+described in section~\ref{database schema}, explaining in detail the tables
+used for storing the data gleaned from the logs and the table that stores
+the rules.
+
+Section~\ref{rules} discusses the parsing rules in detail, explaining the
+purpose and usage of each field in a rule, referring to an example rule and
+sample data it matches successfully against.  This section concludes with a
+discussion of rule efficiency concerns.
 
@@
 parsed and the fraction of mails and connections successfully reconstructed
-by the parsing algorithm.
+by the parsing algorithm, including dealing with false negatives and a
+discussion of the difficulties in identifying false positives.
 
@@
         at~\cite{sql-for-web-nerds}, others
-        are~\cite{w3schools-sql-tutorial, sqlcourse.com}.
+        are~\cite{w3schools-sql-tutorial, sqlcourse.com}.  Depending on the
+        database in use there may be tools available which reduce or remove
+        the requirement to know \SQL{}; for SQLite (the default database
+        used by the implementation) there are several
+        available~\cite{sqlite-guis}.
 
@@
 principle was written in a friendlier time. XXX CARL WANTS ME TO SUGGEST A
-NEW PRINCIPAL.}  A typical \SMTP{} conversation resembles the following
-(the lines starting with a three digit number are sent by the server, all
-other lines are sent by the client):
+NEW PRINCIPAL AND POSSIBLY AN APPENDIX OF SPAM RELATED ISSUES\@.}  A
+typical \SMTP{} conversation resembles the following (the lines starting
+with a three digit number are sent by the server, all other lines are sent
+by the client):
 
@@
 
+An example deviation from the protocol:
+
+\begin{verbatim}
+220 smtp.example.com ESMTP
+HELO client.example.com
+250 smtp.example.com
+MAIL FROM: Alice N. Other alice@example.com
+501 5.1.7 Bad sender address syntax
+RCPT TO: Bob in Sales/Marketing bob@example.com
+503 5.5.1 Error: need MAIL command
+DATA
+503 5.5.1 Error: need RCPT command
+Message headers and body sent here.
+.
+502 5.5.2 Error: command not recognized
+QUIT
+221 2.0.0 Bye
+\end{verbatim}
+
+This client is so poorly written that not only does it present the sender
+and recipient addresses improperly, it ignores the error messages returned
+by the server and carries on regardless.
+
 A detailed description of \SMTP{} is beyond the scope of this document:
@@
 
-Postfix is a highly configurable, high performance, secure and scalable
-Mail Transport Agent.  It features extensive and extensible optional
-anti-spam restrictions, allowing an administrator to deploy those
-restrictions which they judge suitable for their site's needs, rather than
-a fixed set chosen by Postfix's author.  These restrictions can be
-selectively applied, combined and bypassed on a per-client, per-recipient
-or per-sender basis, allowing varying levels of severity and/or
-permissiveness.  Postfix leverages simple lookup tables to support
-arbitrarily complicated user-defined sequences of restrictions and
-exceptions, with policy servers\footnote{Policy servers will be explained
-in section~\refwithpage{policy servers}.} as the ultimate in flexibility.
+Postfix is a \MTA{} with the following design aims (in order of
+importance): security, flexibility of configuration, scalability, and high
+performance.  It features extensive, extensible, optional anti-spam
+restrictions, allowing an administrator to deploy those restrictions which
+they judge suitable for their site's needs, rather than a fixed set chosen
+by Postfix's author.  These restrictions can be selectively applied,
+combined and bypassed on a per-client, per-recipient or per-sender basis,
+allowing varying levels of severity and/or permissiveness.  Postfix
+leverages simple lookup tables to support arbitrarily complicated
+user-defined sequences of restrictions and exceptions, with policy
+servers\footnote{Policy servers will be explained in
+section~\refwithpage{policy servers}.} as the ultimate in flexibility.
 Administrators can also supply their own rejection messages to make it
@@
 black box program based solely on its inputs and outputs.  Although the
-source code is available, reading and understanding it would require a
-significant investment of time:
-
-\begin{tabular}[]{llll}
-
-    Postfix 2.2.11  & Postfix 2.3.8   & Postfix 2.4.0 &                   \\
-    71548           & 82224           & 83965         & lines of code     \\
-    60962           & 67146           & 68675         & lines of comments \\
-    16117           & 17647           & 18069         & lines are blank   \\
-    148627          & 167017          & 170709        & lines in total    \\
-
-\end{tabular}
+source code is available,\footnote{Reading and understanding the source
+code would require a significant investment of time; the source code for
+Postfix 2.4.6 is 16MB.} there are advantages to treating Postfix as a black
+box while developing the parser:
+
+\begin{itemize}
+
+    \item The parser is developed using real world logs rather than the
+        idealised logs someone would naturally envisage reading the source
+        code.
+
+    \item The source code cannot accurately communicate the variety of
+        orderings in which log lines are written to the log file, as
+        process scheduling interferes with it.
+
+    \item The parser acts as a second source of information, with the
+        information gathered from empirical evidence.  An interesting
+        project would be to compare the empirical knowledge inherent in the
+        parsing algorithm with the documentation and source code of
+        Postfix.
+
+\end{itemize}
 
@@
 XXX IMPROVE THE NEXT PARAGRAPH BASED ON COMMENTS FROM CARL\@.
+
+XXX REFER TO THE RULES/PARSER SPLIT ELSEWHERE RATHER THAN EXPLAINING IT
+AGAIN HERE\@.
+
+XXX``CATEGORISE THE COMPLICATIONS SECTIONS IN WORDS HERE'' --- CARL\@.
+
+XXX DISCUSS THE OUTPUT, OR LACK THEREOF, IN GREATER DETAIL, AFTER RERUNNING
+\LMA{}\@.
 
@@
 
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline 
-        delay=0, status=sent (forwarded as 56F5B43FD)}
+        \texttt{3FF7C4317: to=<username@example.com>, relay=local,
+        \newline delay=0, status=sent (forwarded as 56F5B43FD)}
 
@@
 \label{parsing coverage}
+
+XXX ADD SOMETHING ABOUT FALSE POSITIVES AND NEGATIVES\@.
 

### 2007-11-26 doc/logparser-acronyms.tex
Dealt with some XXX items.

@@
 
-\newacronym{IP}{Internet Protocol address}{description={
+\newacronym{IP}{Internet Protocol}{description={
     The Internet Protocol~\cite{Wikipedia-ip} is the protocol used to

### 2007-11-26 doc/logparser.tex
Dealt with some XXX items.

@@
 this behaviour.}  Postfix uses simple lookup tables as the deciding factor
-for some restrictions, e.g.\ XXX IMPROVE THIS EXPLANATION OR REMOVE IT \\
+for some restrictions, e.g.\newline
 \tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\footnote{\texttt{cidr}~\cite{cidr_table} is the type of the lookup table.}
-\newline checks whether the \IP{} address of the connected client matches
-the left hand side of each line in the file and will return the right hand
-side of the first matching line.  Other restrictions determine their result
-in different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com}
-checks whether the \IP{} address of the client is present in the listed
-\RBL{}\@.
+\newline The line above can be broken down as follows:
+
+\begin{description}
+
+    \item [check\_client\_access] The name of the restriction to evaluate.
+
+    \item [cidr] The type of the lookup table.
+
+    \item [/etc/postfix/client\_access] The file containing the lookup
+        table.
+
+\end{description}
+
+check\_client\_access checks whether the \IP{} address of the connected
+client matches the left hand side of each line in the file and returns the
+right hand side of the first matching line.  Other restrictions determine
+their result by consulting external sources, e.g.\
+\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
+address of the client is present in the listed \RBL{}, rejecting the
+command if the client is listed.
 
@@
 
-XXX IMPROVE THE NEXT PARAGRAPH BASED ON COMMENTS FROM CARL\@.
-
-XXX REFER TO THE RULES/PARSER SPLIT ELSEWHERE RATHER THAN EXPLAINING IT
-AGAIN HERE\@.
-
-XXX``CATEGORISE THE COMPLICATIONS SECTIONS IN WORDS HERE'' --- CARL\@.
-
 XXX DISCUSS THE OUTPUT, OR LACK THEREOF, IN GREATER DETAIL, AFTER RERUNNING
@@
 or extract additional data.  It does not appear to deal with any of the
-complications discussed in sections~\refwithpage{complications}
-and~\refwithpage{additional complications}.  The program fails to parse
-Postfix 2.2 or 2.3 log files; indeed all it produces is a number of
-warning messages from the Perl interpreter, without any other output,
-useful or otherwise.  Parsing in \parsername{} is divided in two (see
-sections~\refwithpage{rules} and~\refwithpage{parsing-algorithm} for a more
-detailed explanation): parsing a new log line is a trivial matter of adding
-a new rule, but code changes may be required if new complications
-(sections~\refwithpage{complications} and~\refwithpage{additional
-complications}) must be dealt with or a new action is required
-(section~\refwithpage{actions-in-detail}).
+complications discussed in sections~\refwithpage{complications} (initial
+complications) and~\refwithpage{additional complications} (additional
+complications).  The program fails to parse Postfix 2.2 or 2.3 log files;
+indeed all it produces is a number of warning messages from the Perl
+interpreter, without any other output, useful or otherwise.
 
@@
 entry and subsequently entered by the parent mail once it has been
-identified.  Quite apart from identifying mail injected in an unknown
-fashion, or bugs in the parser, unknown origin mails need to be marked as
-such because they lack some data present in the parent mail; they must copy
-that data from their parent before being entered in the database.  The
-effect of this complication upon the algorithm is discussed more fully in
-section~\refwithpage{tracking re-injected mail}.
-
-XXX FIGURE OUT IF I NEED TO COPY DATA ANY MORE, AND DOCUMENT IT IF SO\@.
+identified.
+
+XXX MAYBE REMOVE THE NEWPAGE BELOW WHEN I'VE SPLIT THE FLOW CHART\@.
 
@@
     \item If the mail is a child mail then it has already been tracked
-        (action: commit; flowchart:~20): the data is cleaned up, missing
-        data is copied from its parent if necessary, and the child is
-        entered in the database (flowchart:~21), then deleted from the
+        (action: commit; flowchart:~20): the data is cleaned up, the child
+        is entered in the database (flowchart:~21), then deleted from the
         state tables (flowchart:~22).  The child mail will be removed from

### 2007-11-28 doc/logparser.tex
Split the flow chart into two, and tweaked the parts to be clearer. Fixed up all the references to the flow chart in the algorithm section.

@@
 
-XXX MAYBE REMOVE THE NEWPAGE BELOW WHEN I'VE SPLIT THE FLOW CHART\@.
-
-\newpage
 \subsection{Flow chart}
@@
 
-Figure~\ref{flow chart image} shows the paths the data representing a
-mail/connection can take through the parser algorithm.  The flow chart
-covers the most common paths only; there are additional, uncommon paths
-which are excluded for the sake of clarity, but are described in
-section~\refwithpage{additional complications}.
-
-XXX SPLIT THIS IMAGE INTO TWO\@; IT'S UNREADABLE NOW, IT'S SO BIG\@.
+Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
+image part 2} shows the paths the data representing a mail/connection can
+take through the parser algorithm.  The flow chart covers the most common
+paths only; there are additional, uncommon paths which are excluded for the
+sake of clarity, but are described in section~\refwithpage{additional
+complications}.
 
@@
 % image is just slightly too big.
-\showgraph{build/logparser-flow-chart.ps}{Parser flow chart}{flow chart
-image}
+
+\showgraph{build/logparser-flow-chart-part-1.ps}{Parser flow chart part
+1}{flow chart image part 1}
+
+\showgraph{build/logparser-flow-chart-part-2.ps}{Parser flow chart part
+2}{flow chart image part 2}
+
 \clearpage
@@
     \item Postfix will log the sender and recipient addresses separately
-        (action: save\_by\_queueid; flowchart:~10).
+        (action: save\_by\_queueid; flowchart:~9).
 
     \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: track; flowchart:~11) ---
+        tracked by the parent mail (action: track; flowchart:~10) ---
         section~\refwithpage{tracking re-injected mail} discusses this in
@@
     \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: commit; flowchart:~13).  This is the last
+        administrator (action: commit; flowchart:~12).  This is the last
         log line for this particular mail (though it may be indirectly
@@
         child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~15), then deleted from the state tables
-        (flowchart:~16).  Re-injected mails are described in
-        section~\refwithpage{tracking re-injected mail}.
+        database (flowchart:~14), then deleted from the state tables.
+        Re-injected mails are described in section~\refwithpage{tracking
+        re-injected mail}.
 
@@
         rather than delivering directly to a mailbox or program as it
-        usually would (action: track; flowchart:~11).\footnote{Relayed
+        usually would (action: track; flowchart:~10).\footnote{Relayed
         delivery is performed by the \SMTP{} client; local delivery means
@@
         whose parent hasn't yet been identified (action: commit;
-        flowchart:~17).  Mark the mail as ready for entry in the database
-        (flowchart:~18), and wait for the parent to deal with it
-        (flowchart:~19).  The mail should not have subsequent log entries;
+        flowchart:~15).  Mark the mail as ready for entry in the database
+        (flowchart:~16), and wait for the parent to deal with it
+        (flowchart:~17).  The mail should not have subsequent log entries;
         only its parent should refer to it.
@@
     \item If the mail is a child mail then it has already been tracked
-        (action: commit; flowchart:~20): the data is cleaned up, the child
-        is entered in the database (flowchart:~21), then deleted from the
-        state tables (flowchart:~22).  The child mail will be removed from
-        the parent mail's list of children (flowchart:~23); if this is the
-        last child and the parent has also been entered in the database the
-        parent will be deleted from the state tables.
+        (action: commit; flowchart:~18): the data is cleaned up, the child
+        is entered in the database (flowchart:~19), then deleted from the
+        state tables.  The child mail will be removed from the parent
+        mail's list of children (flowchart:~20); if this is the last child
+        and the parent has also been entered in the database the parent
+        will be deleted from the state tables.
 
     \item The last alternative is that the mail is a parent mail (action:
-        commit; flowchart:~24).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~25).
+        commit; flowchart:~21).  Regardless of the state of its children
+        its data is cleaned up and entered in the database (flowchart:~22).
         The parent may have children which are waiting to be entered in the
-        database (flowchart:~26); each of those children's data is cleaned
+        database (flowchart:~23); each of those children's data is cleaned
         up and entered in the database, then deleted from the state tables.
         The parent may also have outstanding children which are not yet
-        delivered, in which case (flowchart:~27) the parent must wait for
+        delivered, in which case (flowchart:~24) the parent must wait for
         those children to be finished with.  As soon as the last child is
         deleted from the state tables the parent will also be finished with
-        (flowchart:~28), and deleted from the state tables.
+        (flowchart:~25), and deleted from the state tables.
 
@@
 
-XXX CAN I PUT \% ON THE Y AXIS\@?
-
 \showgraph{build/plot-normal-shuffle-factor.ps}{Percentage increase of

### 2007-12-02 doc/logparser.tex
Tested LMA again, got the code working this time. Improved the review and criticism, not that I have output to do something with.

@@
 REVIEWED
+
+XXX THINK OF A NAME\@.
 
@@
 Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in
-a form suitable for a systems administrator to search using grep(1), awk(1)
-and the myriad of other standard Unix text processing utilities already
-available, producing a \CSV{} file which can subsequently be imported into
-a MySQL or Berkeley DB database.  The decision to support both \CSV{} and
-Berkeley DB appears to have been a limiting factor: \CSV{} is a very simple
-format where a record is stored in a single line, with fields separated by
-a comma or other punctuation symbol.  Problems with \CSV{} files include
-the difficulty in escaping separators, providing multiple values for a
-field (e.g.\ multiple recipients), adding new fields, and lack of a bundled
-schema describing the fields.
+different from this parser: it attempts to present correlated log data in a
+form suitable for a systems administrator to search using the myriad of
+standard Unix text processing utilities already available, producing both a
+\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
+support both \CSV{} and Berkeley DB appears to have been a limiting factor:
+\CSV{} is a very simple format where a record is stored in a single line,
+with fields separated by a comma or other punctuation symbol.  Problems
+with \CSV{} files include the difficulty in escaping separators, providing
+multiple values for a field (e.g.\ multiple recipients), adding new fields,
+and lack of a bundled schema describing the fields.
 
@@
 The data stored is limited to time and date, hostname and \IP{} address of
-client, sender and recipient addresses, and the \SMTP{} code.  Handling of
-multiple recipients, \SMTP{} codes or remote servers\footnote{A single mail
-may be sent to multiple remote servers if it was addressed to recipients in
-different domains, or Postfix needs to try multiple servers for one or more
-recipients.} is not explained in the paper, but the options appear to be:
-
-\begin{enumerate}
-
-    \item Discard some of the data; this option is obviously undesirable,
-        and presumably was not chosen by the \LMA{} authors.
-
-    \item Combine all data into a \CSV{} line, which is then embedded into
-        the main \CSV{} line for that mail, making searching even more
-        difficult and requiring a second separating character.
-
-\end{enumerate}
+client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
+size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
+codes or remote servers\footnote{A single mail may be sent to multiple
+remote servers if it was addressed to recipients in different domains, or
+Postfix needs to try multiple servers for one or more recipients.} is not
+explained in the paper; experimental observation shows that multiple
+records are added when there are multiple recipients (sadly the records are
+not associated in any way), and presumably the same approach is taken when
+there are multiple destination servers.
 
@@
 database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it provides
-\textbf{(key, value)} pairs, so any additional structuring of the data is
-imposed by the application, thus it behoves the application to document
-this structure.
-
-\parsername{} is bundled with extensive documentation, including an
-in-depth explanation of the database schema.  It supports multiple values
-for every field for which multiple values are reasonable (e.g.\ the sender
-address should have one value only, whereas the recipient address can have
-multiple values).  The default database it uses is SQLite3, which supports
-most standard \SQL{} features (its lacks full foreign key support), has
-libraries available for most common programming languages, and a simple
-text based client allowing ad-hoc queries with minimal effort.
-
-XXX DISCUSS THE OUTPUT, OR LACK THEREOF, IN GREATER DETAIL, AFTER RERUNNING
-\LMA{}\@.
+embed a schema, as there is neither requirement nor benefit; it only
+provides \textbf{(key, value)} pairs, so any additional structuring of the
+data is imposed by the application, thus it behoves the application to
+document this structure.
 
@@
 complications) and~\refwithpage{additional complications} (additional
-complications).  The program fails to parse Postfix 2.2 or 2.3 log files;
-indeed all it produces is a number of warning messages from the Perl
-interpreter, without any other output, useful or otherwise.
-
-\LMA{} does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT); the author was unable to parse any logs with \LMA{}, so
-didn't have an opportunity to review the reports.  \parsername{} does not
-produce any reports, it delegates that task to a separate, dedicated
-program.
+complications).  It does not differentiate between different types of
+rejections, so it is not suitable for the purposes of this project; the
+data about which restriction caused the rejection is discarded, whereas an
+important goal of this project is to retain that data to aid optimisation
+and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
+files generated on Solaris hosts because the fields added automatically
+differ from those added on Linux hosts; logs from Solaris hosts (and
+possibly other operating systems) thus require preprocessing before parsing
+by \LMA{}.
+
+\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
+and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
+client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
+sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
+DAILY-REJECT initially failed with an error message from the Perl
+interpreter; after making some corrections to the code it worked, and
+produced the \CSV{} lines for the specified day where the \SMTP{} code
+signifies a rejection.  All of these reports are trivially simple to
+produce from the \CSV{} file using the standard Unix tool \texttt{awk},
+e.g.\ the most complicated, DAILY-REJECT, is merely:
+
+\begin{verbatim}
+awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/ 
+    { print $0; print " "; }' lma_output.txt
+\end{verbatim}
+
+Notes about the command above: there is an unnecessary line containing only
+a single space output after each matching line, for compatibility with the
+output of DAILY-REJECT\@.  Some records have an \SMTP{} code of 0, though
+this is not a valid \SMTP{} code, so presumably this is due to \LMA{} not
+parsing a line properly; \LMA{} considers this code to be a rejection, so
+the command above must too.  In a tiny proportion of records the \SMTP{}
+code is \textit{deferred\/} (again, not a valid code, and most likely due
+to mis-parsing), and is treated as a rejection by \LMA{}, so the equivalent
+\texttt{awk} command must do so too.
+
+There are a number of differences in the output from DAILY-REJECT and the
+\texttt{awk} command:
+
+\begin{itemize}
+
+    \item \LMA{} does not escape or replace the separators used in the
+        \CSV{} records when they are present in sender or recipient
+        addresses, leading to some mangled records; these records are not
+        present in the output from \texttt{awk}.
+
+    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
+        produces only one output line, whereas the \texttt{awk} command
+        produces all matching lines.
+
+\end{itemize}
 
@@
 
+\glossary{name={Solaris},description={
+    Solaris is a Unix-derived Operating System produced by Sun Microsystems
+    (\url{http://www.sun.com/software/solaris/}).
+}}
+
+\glossary{name={awk},description={AWK is a general purpose programming
+    language that is designed for processing text-based data, and is
+    available as a standard utility on all Unix systems.
+}}
+
 \end{document}

### 2007-12-03 doc/logparser.tex
Dealt with another couple of XXX items.

@@
 what you send\/} from \RFC{}~760~\cite{rfc760}, but unfortunately that
-principle was written in a friendlier time. XXX CARL WANTS ME TO SUGGEST A
-NEW PRINCIPAL AND POSSIBLY AN APPENDIX OF SPAM RELATED ISSUES\@.}  A
-typical \SMTP{} conversation resembles the following (the lines starting
-with a three digit number are sent by the server, all other lines are sent
-by the client):
+principle was written in a friendlier time.  Given the deluge of spam that
+mail servers are subjected to daily, a more appropriate maxim could be:
+\textit{Require strict adherence to \RFC{}~2821; implement the strongest
+restrictions you can; relax the restrictions and adherence only when
+legitimate mail is impeded.\/}  It's not as friendly, nor as catchy, but it
+more accurately reflects the current situation.} A typical \SMTP{}
+conversation resembles the following (the lines starting with a three digit
+number are sent by the server, all other lines are sent by the client):
 
@@
 
-    \item GOTO 2.  XXX IMPROVE THIS, IT'S RUBBISH\@.
+    \item Repeat from step two; this will continue indefinitely unless an
+        administrator intervenes and deletes the appropriate mails from the
+        queue.
 

### 2007-12-04 doc/logparser-acronyms.tex
Added false positive/negative stuff to the coverage section.

@@
 
-\newacronym{pid}{Process IDentifier}{description={
+\newacronym{pid}{Process Identifier}{description={
     There may be multiple copies of any program executing at any one time,
@@
 % I think this is better than having both regex and regexs in the acronym
-% list.
+% list.  Likewise for pids.
 \newcommand{\regexs}{\regex{}s}
+\newcommand{\pids}{\pid{}s}
 

### 2007-12-04 doc/logparser.tex
Added false positive/negative stuff to the coverage section.

@@
 
-XXX ADD SOMETHING ABOUT FALSE POSITIVES AND NEGATIVES\@.
-
 \subsection{Introduction}
@@
 and more than half the rules match fewer than 100 times.
+
+Warnings are issued for any lines which are not parsed; no warnings are
+issued for unparsed lines while testing with the 93 test log files, so it
+can be safely concluded that there are zero false negatives.  False
+positives are harder to quantify: short of examining each of the 60,721,709
+log lines and determining which \regex{} parsed it, there is no way to be
+sure that every line was parsed by the correct \regex{}, making it
+impossible to quantify the false positive rate.  The author is confident
+that the false positive rate is extremely low, approaching if not reaching
+zero, because the \regexs{} used are as specific as possible and care has
+been taken to ensure that there are no unintentionally overlapping
+\regexs{}.
 
@@
 
-The 13,850,793 connections and mails accepted, rejected or delivered by
-Postfix during this period are handled correctly by the parser.
+There are no mails in the state tables which should not be present, thus it
+can be concluded that there are zero false negatives.  Once again,
+determining the false positive rate is much harder, as manually checking
+the results of parsing 13,850,793 connections and mails accepted, rejected,
+bounced or delivered is infeasible.  There is considerable circumstantial
+evidence that the false positive rate is quite low:
+
+\begin{itemize}
+
+    \item The parser is quite verbose when complaining about known problems
+        (e.g.\ if a mail is tracked twice as described in
+        section~\refwithpage{tracking re-injected mail}), and no such
+        warnings are produced during the test runs.
+
+    \item Queueids and \pids{} naturally group together log lines belonging
+        to one mail or connection, respectively; it is extremely unlikely
+        that a log line would be associated with the wrong connection.
+
+    \item When dealing with the complications described in
+        sections~\refwithpage{complications} and~\refwithpage{additional
+        complications} the solutions are as specific and restrictive as
+        possible, with the goal of minimising the number of false
+        positives.  In addition the solution to the \textit{Out of order
+        log files\/} problem described in section~\refwithpage{out of order
+        log lines} imposes conditions which each reassembled mail must
+        comply with to be acceptable.
+
+    \item Every effort has been made while developing to make the parser as
+        precise, demanding and particular as possible.
+
+\end{itemize}
+
+Despite the lack of numerical evidence, the author is confident that the
+false positive rate when reconstructing a mail is exceedingly low, if not
+approaching zero.
 

### 2007-12-04 doc/logparser.tex
Improved the first half of the review of other parsers; it's more compare and contrast than review now (or getting there, at least).

@@
 \newcommand{\parsername}{ASO}
+\newcommand{\parsernames}{ASO's}
 % Include images
@@
 
-        Pflogsumm produces a report designed for troubleshooting, and does
-        not support saving any data; parsing concentrates on extracting
-        only the data required to produce the report.  Both the parsing and
-        reporting are difficult to extend, as it is a specialised tool.  It
-        does not attempt to correlate log lines by queueid or \pid{}, nor
-        does it need to deal with the complications encountered during this
-        project.  Pflogsumm produces a useful report, and successfully
-        dealt with all the Postfix logs tested with.
+        Pflogsumm produces a report designed for troubleshooting, rather
+        than for in-depth analysis.  It does not support saving any data,
+        nor does it extract more data than is required for producing the
+        report.  Both the parsing and reporting are difficult to extend as
+        it is a specialised tool, unlike the easily extensible design of
+        \parsername{}.  It does not attempt to correlate log lines by
+        queueid or \pid{}, nor does it need to deal with the complications
+        encountered during this project.  Pflogsumm produces a useful
+        report, and successfully dealt with all the Postfix logs tested
+        with, but is not a suitable base for this project.
 
@@
         Sawmill is a general purpose commercial product which parses 687
-        log file formats (correct as of 2007/04/15) and produces reports.
+        log file formats (correct as of 2007/12/04) and produces reports.
         Its data extraction facilities are quite limited, though it does
         extract three different sets of data for Postfix (one is beta as of
-        2007/04/15), but they do not appear to be interlinked, nor does it
+        2007/12/04), but they do not appear to be interlinked, nor does it
         save sufficient data for the purposes of this project.  No attempt
@@
         only (presumably for a fee), and the product is quite expensive, as
-        it requires a license per report which is to be generated.  The web
-        interface allows creation of dynamic reports based on any field,
-        but due to the associated cost the author has not experimented with
-        it.
+        it requires a license per report which is to be generated; in
+        contrast \parsername{} is free and the code is freely available.
+        The web interface allows creation of dynamic reports based on any
+        field, but due to the associated cost the author has not
+        experimented with it.  Presumably as the company charges per report
+        the user wishes to generate the data store (if there is one) is
+        deliberately inaccessible and undocumented to prevent the user
+        bypassing the program and generating their own reports.
 
@@
 
-        (Last checked 2007/09/26.)
+        Even if the three data sets were linked together Sawmill would
+        extract less data than \parsername{}, and it does not appear to
+        extract data about rejections except when the rejection is caused
+        by an \RBL{} check.
+
+        (Last checked 2007/12/04.)
 
@@
         centralised view capable of searching and correlating diverse log
-        sources.  The web interface allows complicated searches, providing
-        statistics and graphs in real time.  Saved searches can be
-        configured to run periodically and the results emailed to a
-        recipient or sent to a shell script, which presumably can publish
-        the results as required (though possibly without the graphs and
-        detailed statistics).  In the specific case of parsing Postfix
-        logs, Splunk extracts some standard fields: to and from addresses,
-        HELO hostname, date, host the logs were collected from, and
-        protocol.  It parses the standard syslog fields at the beginning of
-        the line, and extracts any \texttt{key=<value>} pairs; these are
-        the fields listed above.  Searches can be based on the extracted
-        fields, and all text in the line is also available for searching.
-        Many types of reports are available, though all are variations of a
-        bar or pie chart, with the exception of bubble and heatmap graphs.
-        It is very easy to drill down through the graphs to extract a
-        portion of the data (e.g.\ select the hour with the largest number
-        of events, then select a particular host, and finally a specific
-        address).  It is not possible to search on partial words, though
-        the web interface does provide and refine a list of possible search
-        terms as you type.  The searches are quite fast, though as the free
+        sources.  The web interface supports complicated searches,
+        providing statistics and graphs in real time, a facility not
+        provided by \parsername{} (report generation has been deferred to a
+        subsequent program).  Saved searches\footnote{The author was unable
+        to save searches, though that may have been due to limitations in
+        the free version.} can be run periodically and the results emailed
+        to a recipient or sent to a shell script, which presumably can
+        publish the results as required (though possibly without the graphs
+        and detailed statistics); \parsername{} provides the database and
+        leaves it to the user to utilise it in any way, whenever they want,
+        with no restrictions on usage of the data.  The interface is
+        optimised for interactive rather than automated queries and it does
+        not appear to be possible to write independent tools to utilise the
+        Splunk database.  Many types of reports are available, though most
+        are variations of a bar or pie chart, with the exception of bubble
+        and heatmap graphs.  It is very easy to drill down through the
+        graphs to extract a portion of the data (e.g.\ select the hour with
+        the largest number of events, then select a particular host, and
+        finally a specific address), though it is not possible to search on
+        partial words.  The searches are quite fast, though as the free
         version has a limit on the amount of data indexed per day the
-        volume of data to be searched was merely 45mb; the cheapest
-        licensed version costs \$6000, and it still limits the volume of
-        data indexed per day.
-
-        The interface is optimised for interactive rather than automated
-        queries and it does not appear to be possible to write independent
-        tools to utilise the Splunk database.  Splunk is a generic tool, so
-        it lacks any Postfix specific support over and above extracting the
-        \texttt{key=<value>} fields; most importantly it makes no attempt
-        to correlate log lines by queueid or \pid{}, nor to handle any of
-        the myriad complications discussed in this document.  Correlation
-        of log lines is left up to the user, and Splunk certainly makes the
-        job easier than using grep, awk and similar Unix tools unless the
-        user has a very high proficiency with those tools.
+        volume of data to be searched was merely 45MB\@; the cheapest
+        licensed version costs \$6000, and still limits the volume of data
+        indexed per day.
+
+        In the specific case of parsing Postfix logs, Splunk extracts some
+        standard fields: to and from addresses, HELO hostname, date, host
+        the logs were collected from, and protocol.  It parses the standard
+        syslog fields at the beginning of the line, and extracts any
+        \texttt{key=<value>} pairs occurring after the standard syslog
+        prologue; these pairs are the fields listed above.  Searches can be
+        based on the extracted fields, and all text in the line is also
+        available for searching.  \parsername{} extracts noticeably more
+        data, though it does not make the full text of the line available,
+        and the full power of \SQL{} is available when searching, allowing
+        the user to search on arbitrarily complicated conditions.
+
+        Splunk is a generic tool, so it lacks any Postfix specific support
+        over and above extracting the \texttt{key=<value>} fields; most
+        importantly it makes no attempt to correlate log lines by queueid
+        or \pid{}, nor to handle any of the myriad complications discussed
+        in this document.
 
@@
         annual time spans, but only for the domains listed in its
-        configuration file.  It appears to ignore all log lines except for
-        those for the current day, though it does maintain a record of data
-        previously extracted, which the newly extracted data is merged into
-        (no information is provided on the format of the data store).  It
-        doesn't utilise rejection log lines in any way, so is unsuitable
-        for the purposes of this project.  It doesn't handle any of the
+        configuration file (making it impossible to produce reports for
+        every sender domain).  It appears to ignore all log lines except
+        for those for the current day, though it does maintain a record of
+        data previously extracted, which the newly extracted data is merged
+        into (no information is provided on the format of the data store).
+        The data extracted appears to be limited to the number of mails
+        sent by each sender, unlike the copious amounts of data extracted
+        by \parsername{}.  It doesn't utilise rejection log lines in any
+        way, so is unsuitable for the purposes of this project.  Its
+        parsing is completely inextensible, indeed is almost
+        incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
+        low level string manipulation; it is the opposite end of the
+        spectrum to \parsernames{} parsing.  It doesn't handle any of the
         complications discussed in this document, doesn't gather the
@@
         restricts it to supporting the \LCD{}.  The data it will extract
-        from an \MTA{} log file is limited to: \newline \tab{} time2,
-        email, email\_r, host, host\_r, method, url, code and bytesd.
-        \newline  There does not seem to be an explanation of those fields
-        in the documentation.  AWStats has no special purpose code to deal
-        with the intricacies of Postfix logs, in fact it operates by
+        from an \MTA{} log file is limited in comparison to \parsername{}:
+        \newline \tab{} time2, email, email\_r, host, host\_r, method, url,
+        code and bytesd.\newline  There does not seem to be an explanation
+        of any of those fields in the documentation (\parsername{} provides
+        copious documentation).  AWStats has no special purpose code to
+        deal with the intricacies of Postfix logs, in fact it operates by
         coercing Postfix logs into Apache\footnote{The Apache web server is
         the most popular HTTP server in use for the past 10 years; more
-        information is available at \url{http://httpd.apache.org/}} format
+        information is available at \url{http://httpd.apache.org/}.} format
         log files, for analysis by AWStats' HTTP log file parser.  The
         converting parser only deals with a small portion of the log lines
-        generated by Postfix, and though it does correlate log lines by
-        queueid, it does not deal with any of the other complications
-        described in this document.
+        generated by Postfix, silently skipping those it cannot deal with,
+        and does not distinguish between different types of rejection.
+        Although though it does correlate log lines by queueid, it does not
+        deal with any of the other complications described in this
+        document.
 
@@
 
+\glossary{name={syslog},description={Syslog is the standard logging
+    mechanism on Unix systems: the program sends log messages to syslog,
+    then syslog filters and stores the messages according to the
+    configuration the administrator has chosen.
+}}
+
 \end{document}

### 2007-12-06 doc/logparser.tex
Improved other parsers review; it's more compare and contrast than it was. Picked Postfix Log Parser as the name. Added parser background section, introducing the rules/algorithm split. Hyperref doesn't work after upgrading to unstable, so it's disabled for now.

@@
 \usepackage{amstext}
-% Wrapping of URLs.  This is used in the bibliography instead of \url{}.
+% Wrapping of URLs.
 \usepackage{url}
-\newcommand{\parsername}{ASO}
-\newcommand{\parsernames}{ASO's}
+\newcommand{\parsername}{\PLP{}}
+\newcommand{\parsernames}{\PLP{}'s}
 % Include images
@@
 
+% XXX THIS DOESN'T WORK AFTER PACKAGE UPGRADES :(
 % When creating a \PDF{} make the table of contents into links to the pages
@@
 % author don't work - I think either gnuplot or graphviz clobbers it.
-\usepackage{hyperref}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-\usepackage{breakurl}
+%\usepackage{hyperref}
+%\hypersetup{
+%    pdftitle    = {Parsing Postfix log files},
+%    pdfauthor   = {John Tobin},
+%    final       = true,
+%    pdfborder   = {0 0 0},
+%%    breaklinks  = true,
+%}
+%\usepackage{breakurl}
 \usepackage{lastpage}
@@
     generate the logs.  This paper describes the creation of a parser,
-    documenting the parsing algorithm and rules, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
+    documenting the parsing algorithm, explaining the difficulties
+    encountered and the solutions developed, with reference to an
+    implementation which stores data gleaned from the logs in an SQL
     database.  The gathered data can then be used to optimise current
@@
 
-XXX THINK OF A NAME\@.
+XXX NOW SAVING THE SIZE OF MAILS\@.
 
@@
 
+\label{postfix background}
+
 Postfix is a \MTA{} with the following design aims (in order of
@@
 
+\subsection{Parser design}
+
+\label{parser design}
+
+It should be clear from the earlier Postfix background
+(section~\refwithpage{postfix background}) that logs produced by Postfix
+are not fixed; they vary widely from host to host, depending on the set of
+restrictions chosen by the administrator.  With this in mind, one of the
+parser's design aims was to make adding new rules as easy as possible, to
+enable administrators to properly parse their log lines.  The main design
+decision which enables this is the separation of rules from the parsing
+algorithm, and secondly the division of the algorithm into discrete
+actions.
+
+\label{why separate rules and algorithm}
+
+Separating the rules from the algorithm makes it possible to parse new log
+lines without modifying the core parsing algorithm.  Although this may seem
+like a trivial point, is it substantially more difficult to understand a
+program's entire parsing algorithm, identify the correct location to
+change, and make the appropriate changes without adversely affecting
+parsing, particularly as there may be edge cases which are not immediately
+obvious.\footnote{See
+section~\refwithpage{yet-more-aborted-delivery-attempts} for a complication
+which occurs only four times in 93 log files tested.}  Requiring changes to
+the parsing algorithm also complicates upgrades, as the changes must be
+preserved during the upgrade, and may clash with changes made by the
+developer.  \parsername{} allows the user to add new rules to the database
+without changing the parsing algorithm, unless the new log lines to be
+parsed require functionality not already provided by the algorithm.  If the
+new log lines do require new functionality, new actions can be added to the
+parser without modifying existing actions or other parts of the algorithm;
+only in the rare case that the new actions require support from other
+sections of the code will more extensive changes be required.
+
 \subsection{Conventions used in the paper}
@@
 
-XXX WHERE SHOULD THE NEXT PARAGRAPH BE PLACED\@?
-
-\label{why separate rules and algorithm}
-
-Separating the rules and algorithm makes it possible to parse new log lines
-without modifying the core parsing algorithm.  Although this may seem like
-a trivial point, is it substantially more difficult to understand a
-program's entire parsing algorithm, identify the correct location to
-change, and make the appropriate changes without adversely affecting
-parsing, particularly as there may be edge cases which are not immediately
-obvious.\footnote{See
-section~\refwithpage{yet-more-aborted-delivery-attempts} for a complication
-which occurs only four times in 93 log files tested.}  Requiring changes to
-the parsing algorithm also complicates upgrades, as the changes must be
-preserved during the upgrade, and may clash with changes made by the
-developer.  \parsername{} allows the user to add new rules to the database
-without changing the parsing algorithm, unless the new log lines to be
-parsed require functionality not already provided by the algorithm.  If the
-new log lines do require new functionality, new actions can be added to the
-parser without modifying existing actions or other parts of the algorithm;
-only in the rare case that the new actions require support from other
-sections of the code will more extensive changes be required.
-
 \subsection{Introduction}
@@
 \subsection{Parsers reviewed}
-
-XXX COMPARE AND CONTRAST RATHER THAN REVIEW\@.
 
@@
         generated by Postfix, silently skipping those it cannot deal with,
-        and does not distinguish between different types of rejection.
+        and does not distinguish between different types of rejection;
+        there is no easy way to extend it to handle new log lines.
         Although though it does correlate log lines by queueid, it does not
@@
         correlate log lines, or any of the more advanced tasks described in
-        this document, because it is not designed to do so.
+        this document, because it is not designed to do so.  The user must
+        construct the \regexs{} to match significant input lines before it
+        will parse any logs.  This program was not reviewed in further
+        detail because its aims are so far removed from the aims of this
+        project.
 
@@
         Anteater doesn't have any English documentation so it's difficult,
-        nigh impossible, to accurately comment on what analysis is
-        performs.  It did not run successfully when tested, and its parsing
-        would certainly be out of date as Postfix has evolved considerably
-        since this tool was last updated (November 2003).  As it neither
-        ran successfully nor has documentation the author can read a
-        detailed review cannot be provided.
+        nigh impossible, for this author to accurately comment on what
+        analysis it performs.  It did not run successfully when tested, and
+        its parsing would certainly be out of date as Postfix has evolved
+        considerably since this tool was last updated (November 2003).  As
+        it neither ran successfully nor has documentation the author can
+        read a detailed review cannot be provided.
 
@@
         YAALA uses a plugin based system to analyse log files and produce
-        \HTML{} output reports.  The core code is merely 153 lines long, as
-        all the parsing and report generation is handled by modules.  YAALA
-        supports storing previously gathered data; it dumps its internal
-        data structures using Perl's Storable module~\cite{perl-storable},
-        so with some knowledge of the data structure created by the module
-        responsible for parsing Postfix log files it should be possible to
-        extract data with an external program.  Using YAALA as a base would
-        be only slightly less work than starting from scratch, as both
-        input and output modules would need to be written specially; it may
-        even be more work to implement the parser within the constraints of
-        YAALA\@.  YAALA bundles a Postfix parser, which extracts the
-        following fields from specific log lines:
+        \HTML{} output reports, with all the parsing and report generation
+        handled by modules.  Using YAALA as a base would be only slightly
+        less work than starting from scratch, as both input and output
+        modules would need to be written specially; it may even be more
+        work to implement the parser within the constraints of YAALA\@.
+        YAALA supports storing previously gathered data using Perl's
+        Storable module~\cite{perl-storable}, so with enough knowledge of
+        the data structure YAALA stores it should be possible to extract
+        data with another Perl program also using the Storable module;
+        \parsername{} uses a well documented database which is accessible
+        from the majority of programming languages.
+
+        YAALA provides a Postfix parser which extracts the following fields
+        from specific log lines:
 
@@
 
-        YAALA provides an \SQL{} like mini-language used when generating
-        the reports; sample reports can be seen at~\cite{yaala-samples}.
-        Example query: \newline \tab{} \texttt{requests BY file WHERE host
-        =\~{} Google}. \newline The mini-language is quite limited and
-        cannot be used to extract data for external use, merely to create
-        reports.
-
-        YAALA as distributed does not attempt the in-depth parsing
-        explained in this document, and implementing such parsing within
-        the confines of YAALA would be at least as much work as
-        implementing it separately.
+        YAALA extracts most of the fields \parsername{} does, but it does
+        not maintain separate counters for each restriction like
+        \parsername{}, so its data cannot be used for optimisation, testing
+        or understanding of restrictions.  YAALA's Postfix parser does not
+        deal with the complications explained in this document, though it
+        does correlate log lines by queueid.
+
+        YAALA provides a mini-language based on \SQL{} that is used when
+        generating reports; sample reports can be seen
+        at~\cite{yaala-samples}.  Example query: \newline \tab{}
+        \texttt{requests BY file WHERE host =\~{} Google} \newline The
+        mini-language is quite limited and cannot be used to extract data
+        for external use, merely to create reports.
 
@@
         most common Postfix log lines, provides reasonably flexible report
-        generation from the limited data extracted, but no facilities to
-        extract data for use in other tools.
+        generation from the limited data extracted, but has no facilities
+        to extract data for use in other tools.
 
@@
         relays most used, etc.\/}; notably rejections are not mentioned or
-        handled.  It supports multiple output formats (text, \HTML{},
-        \PDF{} and Excel 95).  Logs can be parsed, the extracted data
-        saved, and reports generated from it later.  Accessing this data
-        from another application is undocumented; given the source code it
-        should be possible, with sufficient time and effort, to access the
-        data.
+        dealt with, and unlike \parsername{} there is no facility to extend
+        the parser.  It supports multiple output formats for generated
+        reports (text, \HTML{}, \PDF{} and Excel 95) but the reports do not
+        appear to be customisable; \parsername{} does not produce any
+        reports.  Lire supports saving extracted data for later report
+        generation, but accessing this data from another application is
+        undocumented; given the source code it should be possible, with
+        sufficient time and effort, to access the data from an external
+        program.
 
@@
         complications described in sections~\refwithpage{complications}
-        and~\refwithpage{additional complications}.  It stores data, but
-        not in an easily accessible format, and generates reports in
-        multiple formats, but those reports do not appear to be
-        customisable.
+        and~\refwithpage{additional complications}.  \parsername{} uses an
+        \SQL{} database to make accessing the extracted data as easy as
+        possible.
 
@@
 
-        Logrep extracts the following fields from Postfix logs:
+        Logrep extracts roughly half the fields \parsername{} does:
 
@@
 
-        Fields are correlated based on the queueid (referred to as
+        It also counts the number of log lines parsed and skipped.  Log
+        lines are correlated based on the queueid (referred to as
         sessionname [sic] within Logrep), but not by \pid{}.  The parsing
@@
         match the \regex{}, though it appears that they will not overwrite
-        existing data.  It also counts the number of log lines parsed and
-        skipped.  Most notably rejections are completely ignored, making it
-        unsuitable for the purposes of this project.  It doesn't attempt to
-        address any of the complications in
+        existing data.  Most notably rejections are completely ignored,
+        making it unsuitable for the purposes of this project.  It doesn't
+        attempt to address any of the complications in
         sections~\refwithpage{complications} and~\refwithpage{additional
@@
         Logrep does not come with any documentation, though some scant
-        documentation is available on its website.  It requires a web
-        browser to interact with it, ensuring that automated log processing
-        will be difficult.  Sadly all attempts to use Logrep failed, as it
-        was unable to access the log files selected; this appears to be a
-        bug rather than operator error.\footnote{If it is caused by
+        documentation is available on its website (\parsername{} provides
+        copious documentation).  It requires a web browser to interact with
+        it, ensuring that automated log processing will be difficult,
+        whereas automated processing is a key part of \parsernames{}
+        design.  Sadly all the author's attempts to use Logrep failed, as
+        it was unable to access the log files selected; this appears to be
+        a bug rather than operator error.\footnote{If it is caused by
         operator error, the interface needs improvement as the (minimal)
-        instructions were followed as closely as possible.}  As parsing
-        failed it wasn't possible to review the reports Logrep can generate
-        (available in \HTML{} only), nor to examine the (undocumented)
-        format in which it can save extracted data for subsequent reuse.
+        instructions were followed as closely as possible, and multiple
+        attempts were made.}  As parsing failed it wasn't possible to
+        review the reports Logrep can generate (available in \HTML{} only),
+        nor to examine the (undocumented) format in which it can save
+        extracted data for subsequent reuse.
 

### 2007-12-06 doc/logparser.tex
Now saving the size of mails delivered.
Replaced regexs with regexes.

@@
 REVIEWED
-
-XXX NOW SAVING THE SIZE OF MAILS\@.
 
@@
 
+    \item [size] The size of the mail; it will only be present for
+        delivered mails.
+
     \item [message\_id] The message-id of the accepted mail, or
@@
 than the actual matching takes.  Perl automatically caches static
-\regexs{}, but dynamic \regexs{} need to be explicitly compiled and cached.
-Graph~\refwithpage{normal regex vs discard regex} shows execution times
-with and without caching the \regex{}.  Caching the compiled \regexs{} is
-obviously far more efficient; graph~\refwithpage{normal regex vs discarded
-regex factor} shows the percentage execution time increase when not caching
-each \regex{}.
-
-Caching the compiled \regexs{} is quite simple, and is the single most
+\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
+cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
+times with and without caching the \regex{}.  Caching the compiled
+\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
+vs discarded regex factor} shows the percentage execution time increase
+when not caching each \regex{}.
+
+Caching the compiled \regexes{} is quite simple, and is the single most
 effective optimisation implemented in the parser.
@@
         purpose of this phase is to utilise the existing knowledge to
-        create more accurate \regexs{}.  The new log lines are written to a
-        temporary file, which all subsequent stages use instead of the
+        create more accurate \regexes{}.  The new log lines are written to
+        a temporary file, which all subsequent stages use instead of the
         original input file.
@@
 
-    \item [Build regexs]  Contiguous sequences of \texttt{.*} in the newly
-        reassembled \regexs{} are collapsed to a single \texttt{.*}; any
-        resulting duplicate \regexs{} are combined, and their counters
+    \item [Build regexes]  Contiguous sequences of \texttt{.*} in the newly
+        reassembled \regexes{} are collapsed to a single \texttt{.*}; any
+        resulting duplicate \regexes{} are combined, and their counters
         summed.  If the frequency of a \regex{} is lower than the threshold
@@
         a good starting point; obviously the threshold depends on the
-        number and type of input lines.}  The new \regexs{} are printed for
-        the user to add to the database, either as new rules or merged into
-        the \regexs{} of existing rules; the counter for each \regex{} is
-        also printed, giving the user an indication of how many of the
-        input lines that \regex{} should match.  Discarding \regexs{} will
-        result in some of the input lines not being matched; this utility
-        should be run again once the unmatched lines have been isolated by
-        running the parser, including the new \regexs{}, with the same
-        input which produced the original set of unparsed lines.
+        number and type of input lines.}  The new \regexes{} are printed
+        for the user to add to the database, either as new rules or merged
+        into the \regexes{} of existing rules; the counter for each
+        \regex{} is also printed, giving the user an indication of how many
+        of the input lines that \regex{} should match.  Discarding
+        \regexes{} will result in some of the input lines not being
+        matched; this utility should be run again once the unmatched lines
+        have been isolated by running the parser, including the new
+        \regexes{}, with the same input which produced the original set of
+        unparsed lines.
 
@@
 
-A second utility is also provided which reads a list of new \regexs{} and
+A second utility is also provided which reads a list of new \regexes{} and
 the input given to the first utility.  It tries to match each input line
@@
 matched, comparing it to the expected number of matches; this provides the
-user with an easy method of checking if the \regexs{} produced by the first
-utility are correctly matching the input lines they are based upon.  A
-future version of this utility will also group input lines by \regex{}, so
-the user can tweak the \regexs{} if required.
+user with an easy method of checking if the \regexes{} produced by the
+first utility are correctly matching the input lines they are based upon.
+A future version of this utility will also group input lines by \regex{},
+so the user can tweak the \regexes{} if required.
 
@@
     \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexs{} and optimal ordering of
+        effects of caching compiled \regexes{} and optimal ordering of
         rules, with reference to the graphs in
@@
 that the false positive rate is extremely low, approaching if not reaching
-zero, because the \regexs{} used are as specific as possible and care has
+zero, because the \regexes{} used are as specific as possible and care has
 been taken to ensure that there are no unintentionally overlapping
-\regexs{}.
+\regexes{}.
 
@@
         this document, because it is not designed to do so.  The user must
-        construct the \regexs{} to match significant input lines before it
+        construct the \regexes{} to match significant input lines before it
         will parse any logs.  This program was not reviewed in further
@@
 group of graphs vividly shows the huge impact that caching compiled
-\regexs{} has on parser performance.  This section concludes with a
+\regexes{} has on parser performance.  This section concludes with a
 breakdown of the rule hits accumulated during a single test run of the
@@
 
-\subsection{Caching regexs}
-
-The following graphs show the impact that not caching compiled \regexs{}
+\subsection{Caching regexes}
+
+The following graphs show the impact that not caching compiled \regexes{}
 has on parser performance: on typical log files the execution time when not
-caching compiled \regexs{} is 500--600\% of the execution time when
+caching compiled \regexes{} is 500--600\% of the execution time when
 caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexs{}
+merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
 is probably the single most effective optimisation possible in the parser's
@@
 when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexs{} for
+time.  The increases in execution time when not caching \regexes{} for
 different combinations of log files are summarised in the table below:
@@
 consequences of specific inputs.  The necessity of caching compiled
-\regexs{} is attested to by the third group of graphs, where the difference
-between caching and discarding is staggering.  This section concluded with
-a breakdown of the rule hits accumulated during a single test run of the
-parser.
+\regexes{} is attested to by the third group of graphs, where the
+difference between caching and discarding is staggering.  This section
+concluded with a breakdown of the rule hits accumulated during a single
+test run of the parser.
 

### 2007-12-07 doc/logparser.tex
Made hyperref work again.
I want to be able to use pdftex as well as latex, which requires major restructuring of the Makefile; as a side effect the build process will be greatly improved.  Having thoroughly broken the build process, it's now back in a reasonable and working state, so I'm committing even though the changes aren't finished.

@@
 \usepackage{amstext}
-% Wrapping of URLs.
-\usepackage{url}
+% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
+% package does.
+%\usepackage{url}
+% The name of the program, so I only have to change it in one place.
 \newcommand{\parsername}{\PLP{}}
@@
 
-% XXX THIS DOESN'T WORK AFTER PACKAGE UPGRADES :(
 % When creating a \PDF{} make the table of contents into links to the pages
@@
 % author don't work - I think either gnuplot or graphviz clobbers it.
-%\usepackage{hyperref}
-%\hypersetup{
-%    pdftitle    = {Parsing Postfix log files},
-%    pdfauthor   = {John Tobin},
-%    final       = true,
-%    pdfborder   = {0 0 0},
-%%    breaklinks  = true,
-%}
-%\usepackage{breakurl}
+% With later TeX versions the dvips argument is necessary; everywhere else
+% suggests using pdftex as the argument, but pdftex doesn't know how to
+% include postscript images, so I need to create a dvi, convert it to
+% postscript, and convert that to PDF.  I might lookup how to include
+% postscript with pdftex.
+\usepackage[dvips]{hyperref}
+\hypersetup{
+    pdftitle    = {Parsing Postfix log files},
+    pdfauthor   = {John Tobin},
+    final       = true,
+    pdfborder   = {0 0 0},
+%    breaklinks  = true,
+}
+% This is necessary for URLs in the bibliography; I dunno why the url
+% package doesn't work.
+\usepackage{breakurl}
 \usepackage{lastpage}
@@
 
-\showgraph{build/logparser-flow-chart-part-1.ps}{Parser flow chart part
+\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
 1}{flow chart image part 1}
 
-\showgraph{build/logparser-flow-chart-part-2.ps}{Parser flow chart part
+\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
 2}{flow chart image part 2}
@@
 
-\showgraph{build/plot-normal-filesize-numlines.ps}{Execution time vs file
+\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
 size vs number of lines}{execution time vs file size vs number of lines
@@
 
-\showgraph{build/plot-normal-filesize-numlines-factor.ps}{Ratio of file
+\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
 size and number of lines to execution time}{execution time vs file size vs
@@
 
-\showgraph{build/plot-normal-shuffle-factor.ps}{Percentage increase of
+\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
 shuffled over normal}{percentage increase of shuffled over normal}
 
-\showgraph{build/plot-normal-reverse-factor.ps}{Percentage increase of
+\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
 reversed over normal}{percentage increase of reversed over normal}
@@
 
-\showgraph{build/plot-cached-discarded.ps}{Regex: cached vs
+\showgraph{build/plot-cached-discarded}{Regex: cached vs
 discarded}{normal regex vs discard regex}
 
-\showgraph{build/plot-cached-discarded-factor.ps}{Regex caching: percentage
+\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
 execution time increase}{normal regex vs discarded regex factor}
@@
 
-\showgraph{build/plot-hits.ps}{Hits per rule}{rule hits graph}
+\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
 
@@
 
-\showgraph{build/plot-hits-low.ps}{Hits per rule (low)}{hits per rule low}
+\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
 
@@
 
-\showgraph{build/plot-hits-middle.ps}{Hits per rule (middle)}{hits per rule
+\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
 middle}
 
-\showgraph{build/plot-hits-high.ps}{Hits per rule (high)}{hits per rule
+\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
 high}

### 2007-12-14 doc/logparser.tex
Using pdflatex to generate a PDF directly works now. It needs some fine tuning: the graphs look a bit crap, and pdflatex complains     about the flow charts, but it's working. Also removed the range from the tables, because I now understand standard     deviation and I was using it incorrectly.
Improved the logs2regexs algorithm explanation.

@@
 \usepackage[section,numbib]{tocbibind}
+% Provides commands to distinguish between pdf and dvi output.
+\usepackage{ifpdf}
 
@@
 % author don't work - I think either gnuplot or graphviz clobbers it.
-% With later TeX versions the dvips argument is necessary; everywhere else
-% suggests using pdftex as the argument, but pdftex doesn't know how to
-% include postscript images, so I need to create a dvi, convert it to
-% postscript, and convert that to PDF.  I might lookup how to include
-% postscript with pdftex.
+\ifpdf{}
+    \usepackage[pdftex]{hyperref}
+\else{}
     \usepackage[dvips]{hyperref}
+\fi{}
 \hypersetup{
@@
 % This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work.
+% package doesn't work; don't include it when generating \PDF{} output.
+\ifpdf{}
+\else{}
     \usepackage{breakurl}
+\fi{}
 \usepackage{lastpage}
@@
         position within the line)\/} tuple, incrementing it each time that
-        word is occurs in that position.
+        word occurs in that position.
 
@@
         variable term.  Variable terms are replaced by \texttt{.*}, which
-        means to match zero or more of any character.  The words are
-        reassembled to produce a \regex{} matching the line, and a counter
-        is maintained for each \regex{}.
-
-    \item [Build regexes]  Contiguous sequences of \texttt{.*} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.*}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        summed.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second, independent threshold, but once again 1\% of input lines is
-        a good starting point; obviously the threshold depends on the
-        number and type of input lines.}  The new \regexes{} are printed
-        for the user to add to the database, either as new rules or merged
-        into the \regexes{} of existing rules; the counter for each
-        \regex{} is also printed, giving the user an indication of how many
-        of the input lines that \regex{} should match.  Discarding
-        \regexes{} will result in some of the input lines not being
-        matched; this utility should be run again once the unmatched lines
-        have been isolated by running the parser, including the new
-        \regexes{}, with the same input which produced the original set of
-        unparsed lines.
+        means to match zero or more of any character.
+
+    \item [Build regexes]  The words are reassembled to produce a \regex{}
+        matching the line, and a counter is maintained for each \regex{}.
+        Contiguous sequences of \texttt{.*} in the newly reassembled
+        \regexes{} are collapsed to a single \texttt{.*}; any resulting
+        duplicate \regexes{} are combined, and their counters summed.  If
+        the frequency of a \regex{} is lower than the threshold supplied by
+        the user the \regex{} is discarded.\footnote{This is a second,
+        independent threshold, but once again 1\% of input lines is a good
+        starting point; obviously the threshold depends on the number and
+        type of input lines.}  The new \regexes{} are printed for the user
+        to add to the database, either as new rules or merged into the
+        \regexes{} of existing rules; the counter for each \regex{} is also
+        printed, giving the user an indication of how many of the input
+        lines that \regex{} should match.  Discarding \regexes{} will
+        result in some of the input lines not being matched; this utility
+        should be run again once the unmatched lines have been isolated by
+        running the parser, including the new \regexes{}, with the same
+        input which produced the original set of unparsed lines.
 

### 2007-12-17 doc/logparser.tex
Use .+ instead of .* in the generated regexes.

@@
         is classified as a fixed word, otherwise it is classified as a
-        variable term.  Variable terms are replaced by \texttt{.*}, which
+        variable term.  Variable terms are replaced by \texttt{.+}, which
         means to match zero or more of any character.
@@
         matching the line, and a counter is maintained for each \regex{}.
-        Contiguous sequences of \texttt{.*} in the newly reassembled
-        \regexes{} are collapsed to a single \texttt{.*}; any resulting
+        Contiguous sequences of \texttt{.+} in the newly reassembled
+        \regexes{} are collapsed to a single \texttt{.+}; any resulting
         duplicate \regexes{} are combined, and their counters summed.  If

### 2007-12-18 doc/logparser.tex
Started on the overall conclusion.  I want to get Carl's opinion before taking it much further; I reckon I need to expand it to at least a page, if not two.

@@
 
+\newpage
 \section{Conclusion}
@@
 
-XXX WRITE THIS
+Parsing Postfix logs appears at first sight to be an almost trivial task,
+especially given previous experience in parsing logs, but it turns out to
+be a much more taxing project than initially expected.  The variety and
+breadth of log lines produced by Postfix is quite surprising, as a quick
+survey of sample log files gives the impression that the number of distinct
+log lines is quite small; this impression is due to the uneven distribution
+evidenced by logs produced in normal operation (see graph~\refwithpage{rule
+hits graph} for a vivid illustration of this).
+
+Enabling users to easily extend the parser to deal with new log lines is a
+design imperative, and despite the resulting increase in complexity the job
+is quite tractable, though it does raise interesting efficiency and
+optimisation questions (answered in section~\refwithpage{rule efficiency}).
+For the parser to be useful to the average user, parsing new log lines
+produced by Postfix must be as simple a task as configuring Postfix to
+produce those log lines was in the first place, otherwise users won't make
+to effort to utilise the parser.  Providing a tool to ease the generation
+of \regexes{} from unparsed log lines should help
+
+The real difficulties arise once the parser is successfully dealing with
+90\% of the log lines, as the irregularities and complications previously
+explained only begin to become apparent once the vast majority of log lines
+have been parsed successfully.  Dealing with numerous, infrequently
+occurring log lines is a simple, albeit tiresome task, whereas dealing with
+information which hasn't been logged, e.g.\ when delivery attempts are
+aborted.
 

### 2007-12-23 doc/logparser.tex
Improved the discussion of why rules and actions are separated, by incorporating the appropriate paragraph from the rules section; the rules and algorithm sections now refer to the parser design section. Extended the conclusion somewhat.

@@
 
+Decoupling the parsing rules from the associated actions allows new rules
+to be written and tested without requiring modifications to the algorithm
+source code (significantly lowering the barrier to entry for new or casual
+users who need to parse new log lines), and greatly simplifies both
+algorithm and rules.  Decoupling also creates a clear separation of
+functionality: rules handle low level details of identifying log lines and
+extracting data from a line, whereas the algorithm handles the higher level
+details of following the path a mail takes through Postfix, assembling the
+required data before storing it, dealing with complications arising, etc.
+
 Separating the rules from the algorithm makes it possible to parse new log
@@
 
+Please refer to section~\refwithpage{parser design} for a discussion of why
+the rules and actions have been separated in the parser's design.
+
 \subsection{Rule characteristics}
@@
 \label{rule characteristics}
-
-The complexity and variation in Postfix's logs requires similar flexibility
-in the parser, which is split into two parts: the parsing algorithm and the
-parsing rules which are applied to the lines.  Decoupling the parsing rules
-from the associated actions allows new rules to be written and tested
-without requiring modifications to the algorithm source code (significantly
-lowering the barrier to entry for new or casual users who need to parse new
-log lines), and greatly simplifies both algorithm and rules.  Decoupling
-also creates a clear separation of functionality: rules handle low level
-details of identifying log lines and extracting data from a line, whereas
-the algorithm handles the higher level details of following the path a mail
-takes through Postfix, assembling the required data before storing it,
-dealing with complications arising, etc.
 
@@
 
+Please refer to section~\refwithpage{parser design} for a discussion of why
+the rules and actions have been separated in the parser's design.
+
 \subsection{Introduction}
@@
 log lines is quite small; this impression is due to the uneven distribution
-evidenced by logs produced in normal operation (see graph~\refwithpage{rule
+exhibited by logs produced in normal operation (see graph~\refwithpage{rule
 hits graph} for a vivid illustration of this).
@@
 to effort to utilise the parser.  Providing a tool to ease the generation
-of \regexes{} from unparsed log lines should help
+of \regexes{} from unparsed log lines should greatly help users
+add rules to parse unparsed lines.
 
@@
 have been parsed successfully.  Dealing with numerous, infrequently
-occurring log lines is a simple, albeit tiresome task, whereas dealing with
-information which hasn't been logged, e.g.\ when delivery attempts are
-aborted.
+occurring log lines is a simple task, albeit tiresome, whereas dealing with
+mails where information appears to be missing is much more grueling.
+Trawling through the logs, looking for something out of the ordinary,
+possibly hundreds or even thousands of lines away from the last mention of
+the queueid in question, is extremely time consuming and error prone.  In
+many cases the task is not to spot the line which is unusual, but to spot
+that a line normally present is missing.  In all cases the evidence must be
+used to construct a hypothesis to explain the irregularities, and that
+hypothesis must then be tested in the parser; if successful the parser must
+be modified to deal with the irregularities, without adversely affecting
+the existing functionality.  The complications described in
+section~\refwithpage{additional complications} were solved in the order
+they are described in, and that order closely resembles the frequency with
+which they occur; the most frequently occurring complications dominate the
+warning messages produced, and so naturally they are the first
+complications to be dealt with.
 

### 2007-12-23 doc/logparser-acronyms.tex
Improved other parsers review; it's more compare and contrast than it was. Picked Postfix Log Parser as the name. Added parser background section, introducing the rules/algorithm split. Hyperref doesn't work after upgrading to unstable, so it's disabled for now.
Replaced regexs with regexes.
Added some discussion of the similarity between the parser's rules/actions/algorithm split and Wood's Augmented Transition Networks.

@@
 % Warning: don't use acronyms within definitions, they don't work properly.
+
+\newacronym{PLP}{Postfix Log Parser}{description={
+    Postfix Log Parser is the implementation of the algorithm described in
+    this paper.
+}}
 
@@
 
-% I think this is better than having both regex and regexs in the acronym
+% I think this is better than having both regex and regexes in the acronym
 % list.  Likewise for pids.
-\newcommand{\regexs}{\regex{}s}
+\newcommand{\regexes}{\regex{}es}
 \newcommand{\pids}{\pid{}s}
@@
 }}
+
+\newacronym{ATN}{Augmented Transition Networks}{description={
+    Augmented Transition Networks are a tool used in Computational
+    Linguistics; they are originally described in~\cite{atns} and further
+    in~\cite{nlpip}.
+}}

### 2007-12-23 doc/logparser.tex
Added some discussion of the similarity between the parser's rules/actions/algorithm split and Wood's Augmented Transition Networks.

@@
 
+There is some similarity between the parser's design and William Wood's
+\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
+creating grammars to parse or generate sentences.  The resemblance between
+\ATN{} and the parser is accidental, but it is interesting how two
+apparently different approaches share an underlying separation of concerns;
+this appears to be a natural division of responsibility and functionality.
+
+% Do Not Reformat!
+
+\begin{tabular}[]{lll}
+    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
+    Networks            & Algorithm         & Determines the sequence 
+                                              of transitions            \\
+                        &                   & or actions which 
+                                              constitutes a valid       \\
+                        &                   & input.                    \\
+    Transitions         & Actions           & Save data and impose
+                                              conditions the            \\
+                        &                   & input must meet to be
+                                              considered valid.         \\
+    Abbreviations       & Rules             & Responsible for 
+                                              classifying input.        \\
+\end{tabular}
+
 \subsection{Conventions used in the paper}

### 2008-01-16 doc/logparser.tex
Added sections on:     Emergent behaviour     Adding new actions     Benefits of storing rules in the database     Ordering of the additional complications Extended and improved the conclusion.

@@
 
-XXX WHAT IS THE CONCLUSION???  FIRST THING IS TO COMPARE TO OTHER PARSERS
-REVIEWED
+XXX USE \textsection{} INSTEAD OF SECTION WHEN REFERRING TO \\OTHER
+SECTIONS\@?
 
@@
 
+An important but easily overlooked benefit of storing the rules in the
+database is the connection between rules and results: if more information
+is required when examining a result, the rules which produced the database
+entries are available for inspection (and each result references the rule
+which created it).  There is no ambiguity about which rule resulted in a
+particular result, eliminating one potential source of confusion.
+
 \subsection{Rules table}
@@
 
+\label{creating new rules}
+
 The logs produced by Postfix differ from installation to installation,
@@
     \item A flow chart showing common paths a mail can take through the
-        algorithm.
+        algorithm, and discussing the emergent behaviour observed in the
+        parser.
 
@@
 
+The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
+which is far more complicated than the behaviour described by the
+algorithm.  The top-level parser could be written in pseudo-code like so
+(with indentation level indicating flow control):
+
+\begin{verbatim}
+for each line in the input files:
+    for each rule defined by the user:
+        if this rule matches the input line:
+            perform the action specified by the rule
+            skip the remaining rules
+            process the next input line
+    warn the user that the input line was not parsed
+\end{verbatim}
+
+Yet despite the brevity of the algorithm above the (simplified) flow chart
+below contains twenty one states, three entry points, five exit points,
+five loops, four explicit branches, and eight implicit branches.  Several
+more states and corresponding branches are omitted for the sake of clarity
+(these extra states are caused by solving the difficulties described in
+section~\refwithpage{additional complications}).
+
+The solution to the second last complication encountered
+(section~\refwithpage{out of order log lines}) requires that a list of the
+acceptable combinations of Postfix programs a mail can pass through;
+however this does not correspond to the \textit{actions} a mail must pass
+through.  Before that complication was overcome, i.e.\ for the majority of
+the development of this parser, there was nothing in the parser explicitly
+encoding the states or paths a mail could take; the vast majority of the
+parser has no need for this flow of data to be specified.  This eases the
+process of adding new actions (described in section~\refwithpage{adding new
+actions}), as there is no requirement to explicitly insert the action into
+an algorithmic version of the flow chart below.
+
 % This is very likely to move around as the flow chart is as big as
@@
 
-    \item First there is a connection from the remote client
-        (action: connect; flowchart:~1).
+    \item First there is a connection from the remote client (action:
+        connect; flowchart:~1).
 
@@
     \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen
-        queueid for delivery (action: mail\_picked\_for\_delivery;
-        flowchart:~3), in which case a new data structure will be created.
-        The mail will be flagged as having unknown origins; this flag
-        should be subsequently cleared once the origin has been
-        established.  This may also be an indicator that the mail is a
-        bounce notification, see
+        \daemon{qmgr} selects a mail with a previously unseen queueid for
+        delivery (action: mail\_picked\_for\_delivery; flowchart:~3), in
+        which case a new data structure will be created.  The mail will be
+        flagged as having unknown origins; this flag should be subsequently
+        cleared once the origin has been established.  This may also be an
+        indicator that the mail is a bounce notification, see
         section~\refwithpage{identifying-bounce-notifications} for details.
@@
 
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local,
-        \newline delay=0, status=sent (forwarded as 56F5B43FD)}
+        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
+        delay=0, status=sent (forwarded as 56F5B43FD)}
 
@@
 
+\subsection{Adding new actions}
+
+\label{adding new actions}
+
+Adding new actions is almost as simple as adding new rules, though actions
+are not stored in the database.  The implementor writes a subroutine which
+accepts the standard arguments given to actions, and registers it as an
+action by calling the parser method add\_actions() with the subroutine
+name as a parameter.  No other work is required on the implementor's part;
+all of their attention and effort can be focused on the correctness of
+their action.  The only negative aspect is that the process involves
+editing the parser source code, which makes upgrading to a later version of
+the parser more difficult, though by no means impossible.  If the author of
+the new action wishes they can take advantage of the parser's object
+oriented implementation~\cite{Wikipedia-object-orientation} by subclassing
+it and implementing their changes in the derived class, allowing future
+upgrading of the parser with greatly reduced chance of conflicts.  The
+action may need to extend the list of valid combinations described in
+section~\refwithpage{out of order log lines} if the addition creates a
+different set of acceptable programs, but this is unlikely to occur, as it
+would require parsing log lines from Postfix components the parser
+currently ignores.\footnote{The mail server used for development does not
+utilise the \daemon{lmtp} or \daemon{virtual} delivery agents, so this
+parser does not deal with those components.  Adding new rules to parse
+their log lines is a simple process, though if their behaviour differs
+significantly from the \daemon{smtp} or \daemon{local} delivery agents new
+actions may be required.}
+
 \subsection{Additional complications}
@@
 \label{additional complications}
+
+The complications described in this section are listed in the order in
+which they were encountered during development of the parser.  Each of
+these complications caused the parser to operate incorrectly, generating
+either warning messages or leaving mails in the state table.  The frequency
+of occurrence is much higher at the start of the list, with the first
+complication occurring several orders of magnitude more frequently than the
+last.  When deciding which problem to address next, the most common was
+always chosen, as resolving the most common problem would yield the biggest
+improvement in the parser, and also prune the greatest number of mails from
+the state tables and error messages.
 
@@
 the actions which are available to rules were described in full.  The
-section concludes with the remaining complications discovered and overcome
-in the process of refining and completing the algorithm.  Detecting,
-diagnosing and defeating complications forms the largest single portion of
-this document, mirroring the development of the parser,  The complications
-are described in the order they were overcome, with subsequent problems
-affecting fewer mails (often by an order or magnitude), though the time
-required to solve problems increased with each successive problem.
+parser exhibits emergent behaviour far more complicated than the behaviour
+encoded in the algorithm, and this behaviour is covered in the flow chart
+subsection.  The section concludes with the remaining complications
+discovered and overcome in the process of refining and completing the
+algorithm.  Detecting, diagnosing and defeating complications forms the
+largest single portion of this document, mirroring the development of the
+parser,  The complications are described in the order they were overcome,
+with subsequent problems affecting fewer mails (often by an order or
+magnitude), though the time required to solve problems increased with each
+successive problem.
 
@@
 and more than half the rules match fewer than 100 times.
+
+XXX PICK A RANDOM SAMPLE OF LOG LINES AND ENSURE THEY ARE PARSED
+PROPERLY\@.
 
@@
 
-Enabling users to easily extend the parser to deal with new log lines is a
-design imperative, and despite the resulting increase in complexity the job
-is quite tractable, though it does raise interesting efficiency and
-optimisation questions (answered in section~\refwithpage{rule efficiency}).
-For the parser to be useful to the average user, parsing new log lines
-produced by Postfix must be as simple a task as configuring Postfix to
-produce those log lines was in the first place, otherwise users won't make
-to effort to utilise the parser.  Providing a tool to ease the generation
-of \regexes{} from unparsed log lines should greatly help users
-add rules to parse unparsed lines.
+Given the diverse nature of the log lines and the ease with which
+administrators can cause new lines to be logged
+(section~\refwithpage{postfix background}), enabling users to easily extend
+the parser to deal with new log lines is a design imperative
+(section~\refwithpage{parser design}).  Despite the resulting initial
+increase in complexity the job is quite tractable, though it does raise
+interesting efficiency and optimisation questions (answered in
+section~\refwithpage{rule efficiency}).  Providing a tool
+(section~\refwithpage{creating new rules}) to ease the generation of
+\regexes{} from unparsed log lines should greatly help users add rules to
+parse formerly unparsed lines.
+
+The implementation not only substantially eases the parsing of new log
+lines, it makes adding new actions
+(section~\refwithpage{parsing-algorithm}) an almost trivial task.  The
+simplicity of adding a new action (section~\refwithpage{adding new
+actions}) frees the implementor from worrying about how their action might
+disrupt parsing of other lines or the behaviour of other actions, allowing
+their concentration to be focused on correctly implementing their action.
+
+The division of the parser into rules and actions is unusual because it
+divides the two so completely; although parsers are often divided (lex and
+yacc~\cite{lex-and-yacc} based parsers being an obvious example), the parts
+are usually quite internally interdependent, and will be combined into a
+complete parser by the compilation process; in contrast \parsername{} keeps
+the rules and actions separate until the parser runs.  XXX EXTEND THIS\@;
+SOMETHING ABOUT THE CONTROL FLOW BEING DIFFERENT, I.E. RULE DETERMINES
+ACTION, RATHER THAN ACTION EXPECTING INPUT\@.
+
+The emergent behaviour~\cite{Wikipedia-Emergence} exhibited by the rules
+and actions is also interesting, and is discussed beside the flow chart
+depicting said emergent behaviour (section~\refwithpage{flow-chart}).  This
+emergent behaviour greatly eases
 

### 2008-01-20 doc/logparser-acronyms.tex
Dealt with things which arose during my last meeting with Carl: * Improve the description of states in the emergent behaviour section. * Explained subclassing some more. * Added glossary and acronym entries for lmtp and virtual.

@@
     provided in section~\refwithpage{SMTP background}.
+}}
+
+\newacronym{LMTP}{Local Mail Transfer Protocol}{description={
+    LMTP is a protocol derived from SMTP that removes the need for the
+    server to maintain a mail delivery queue, instead relying on the client
+    to maintain it.  Typically the client would be an MTA, and the server
+    would be a delivery agent or a mail store.  Full details are available
+    in \cite{lmtp-rfc}.
 }}

### 2008-01-20 doc/logparser.tex
Dealt with things which arose during my last meeting with Carl: * Improve the description of states in the emergent behaviour section. * Explained subclassing some more. * Added glossary and acronym entries for lmtp and virtual.

@@
 Yet despite the brevity of the algorithm above the (simplified) flow chart
-below contains twenty one states, three entry points, five exit points,
-five loops, four explicit branches, and eight implicit branches.  Several
+below contains:
+
+\begin{itemize}
+
+    \item Twenty one states (not including explicit branches).
+
+    \item Three entry points (1, 2, 3).
+
+    \item Five exit points (8, 14, 17, 20, 25).
+
+    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
+
+    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
+        by an action, determining what will happen next to a mail.
+
+    \item Four implicit branches, where the transition is determined by the
+        next log line which is processed for that mail, e.g.\ state 1 can
+        transition to either state 4 or state 5.  (1--\{4,5\},
+        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
+
+\end{itemize}
+
+    Several
 more states and corresponding branches are omitted for the sake of clarity
@@
 acceptable combinations of Postfix programs a mail can pass through;
-however this does not correspond to the \textit{actions} a mail must pass
+however this does not correspond to the \textit{actions\/} a mail must pass
 through.  Before that complication was overcome, i.e.\ for the majority of
@@
 accepts the standard arguments given to actions, and registers it as an
-action by calling the parser method add\_actions() with the subroutine
-name as a parameter.  No other work is required on the implementor's part;
-all of their attention and effort can be focused on the correctness of
-their action.  The only negative aspect is that the process involves
-editing the parser source code, which makes upgrading to a later version of
-the parser more difficult, though by no means impossible.  If the author of
-the new action wishes they can take advantage of the parser's object
-oriented implementation~\cite{Wikipedia-object-orientation} by subclassing
-it and implementing their changes in the derived class, allowing future
-upgrading of the parser with greatly reduced chance of conflicts.  The
-action may need to extend the list of valid combinations described in
-section~\refwithpage{out of order log lines} if the addition creates a
-different set of acceptable programs, but this is unlikely to occur, as it
-would require parsing log lines from Postfix components the parser
-currently ignores.\footnote{The mail server used for development does not
-utilise the \daemon{lmtp} or \daemon{virtual} delivery agents, so this
-parser does not deal with those components.  Adding new rules to parse
-their log lines is a simple process, though if their behaviour differs
-significantly from the \daemon{smtp} or \daemon{local} delivery agents new
-actions may be required.}
+action by calling the parser subroutine add\_actions() with the name of the
+new action subroutine as a parameter.  No other work is required on the
+implementor's part; all of their attention and effort can be focused on the
+correctness of their action.  The only negative aspect is that the process
+involves editing the parser source code, which makes upgrading to a later
+version of the parser more difficult, though by no means impossible.  If
+the author of the new action wishes they can take advantage of the parser's
+object oriented implementation~\cite{Wikipedia-object-orientation} by
+subclassing it and implementing their changes in the derived class,
+allowing future upgrading of the parser with greatly reduced chance of
+conflicts.\footnote{The real difference between the two approaches is where
+the new code is placed.  The simpler option is to change the parser code
+directly, but those changes will then have to be made to subsequent
+versions of the parser, and as the scope of the changes increases so does
+the chance of conflict.  The more time consuming option is to write a
+subclass containing the new actions and change the program which invokes
+the parser so that it uses the subclass rather than the parser; the changes
+required to the program invoking the parser are minor and much less likely
+to lead to conflicts when upgrading to future versions of the parser.  An
+alternative is to submit new actions to the author of the parser for
+inclusion in future versions, resulting in two benefits: the new actions do
+not need to be maintained separately, and other users of the parser will
+avail of the new functionality.} The action may need to extend the list of
+valid combinations described in section~\refwithpage{out of order log
+lines} if the addition creates a different set of acceptable programs, but
+this is unlikely to occur, as it would require parsing log lines from
+Postfix components the parser currently ignores.\footnote{The mail server
+used for development does not utilise the \daemon{lmtp} or \daemon{virtual}
+delivery agents, so this parser does not have rules to handle log lines
+from those components.  Adding new rules to parse their log lines is a
+simple process, though if their behaviour differs significantly from the
+\daemon{smtp} or \daemon{local} delivery agents new actions may be
+required.  The mail server in question is a production mail server handling
+mail for a university department; the benefit is that the logs used exhibit
+the idiosyncrasies and foibles a mail server in the wild must deal with,
+but the downside is that significantly altering the configuration just to
+log messages from a different Postfix component is not an option.}
 
@@
 complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
+last.  When deciding which problem to address next the most common was
 always chosen, as resolving the most common problem would yield the biggest
@@
 
+\glossary{name={lmtp},description={
+    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
+    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
+}}
+
 \glossary{name={local},description={
@@
 
+\glossary{name={virtual},description={
+    Virtual is the Postfix component responsible for delivery of mails to
+    virtual domains.  With \daemon{local} delivery the destination is
+    determined only by the portion of the email address on the left side of
+    the \at{}, whereas with \daemon{virtual} delivery the destination is
+    determined by the entire email address, e.g.\ if the server considers
+    itself responsible for both \textbf{example.org} and
+    \textbf{example.net} domains: \daemon{local} considers
+    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
+    the same mailbox, whereas \daemon{virtual} considers them to have
+    different mailboxes.  Virtual delivery is used when a server hosts
+    multiple domains where a username may be present in more than one
+    domain but represent different users in each.  The definitive
+    documentation is
+    \url{http://www.postfix.org/virtual.8.html}.
+}}
+
 \glossary{name={Bayesian spam filtering},description={

### 2008-01-28 doc/logparser.tex
New command \sectionref{} which puts section squiggles in front of references, and all appropriate references changed to use it.

@@
 }
+% section references, automatically add \textsection
+\newcommand{\sectionref}[1]{%
+    \textsection{}\refwithpage{#1}%
+}
 
@@
 \end{abstract}
-
-XXX USE \textsection{} INSTEAD OF SECTION WHEN REFERRING TO \\OTHER
-SECTIONS\@?
 
@@
 measures against, or to produce statistics showing how effective those
-measures are.\footnote{Section~\refwithpage{introduction} lists some more
-example uses.}
+measures are.\footnote{\sectionref{introduction} lists some more example
+uses.}
 
@@
 user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in
-section~\refwithpage{policy servers}.} as the ultimate in flexibility.
-Administrators can also supply their own rejection messages to make it
-clear to senders why exactly their mail was rejected.  Unfortunately this
-flexibility has a cost: complexity in the logs generated.  While it is easy
-to use standard Unix text processing utilities to determine the fate of an
-individual email, following the journey an email takes through Postfix can
-be quite difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of
-the time the journey is simple, but the other 10\% of the time requires
-90\% of the code.\footnote{These numbers don't have a solid scientific
-basis, they're based on gut feeling from writing and debugging the
-software.}
+servers\footnote{Policy servers will be explained in \sectionref{policy
+servers}.} as the ultimate in flexibility.  Administrators can also supply
+their own rejection messages to make it clear to senders why exactly their
+mail was rejected.  Unfortunately this flexibility has a cost: complexity
+in the logs generated.  While it is easy to use standard Unix text
+processing utilities to determine the fate of an individual email,
+following the journey an email takes through Postfix can be quite
+difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
+the journey is simple, but the other 10\% of the time requires 90\% of the
+code.\footnote{These numbers don't have a solid scientific basis, they're
+based on gut feeling from writing and debugging the software.}
 
@@
 used when determining which rules should be used to parse each log line
-(see section~\refwithpage{rule characteristics} for details.)
+(see \sectionref{rule characteristics} for details.)
 
@@
 
-It should be clear from the earlier Postfix background
-(section~\refwithpage{postfix background}) that logs produced by Postfix
-are not fixed; they vary widely from host to host, depending on the set of
-restrictions chosen by the administrator.  With this in mind, one of the
-parser's design aims was to make adding new rules as easy as possible, to
-enable administrators to properly parse their log lines.  The main design
-decision which enables this is the separation of rules from the parsing
-algorithm, and secondly the division of the algorithm into discrete
-actions.
+It should be clear from the earlier Postfix background (\sectionref{postfix
+background}) that logs produced by Postfix are not fixed; they vary widely
+from host to host, depending on the set of restrictions chosen by the
+administrator.  With this in mind, one of the parser's design aims was to
+make adding new rules as easy as possible, to enable administrators to
+properly parse their log lines.  The main design decision which enables
+this is the separation of rules from the parsing algorithm, and secondly
+the division of the algorithm into discrete actions.
 
@@
 parsing, particularly as there may be edge cases which are not immediately
-obvious.\footnote{See
-section~\refwithpage{yet-more-aborted-delivery-attempts} for a complication
-which occurs only four times in 93 log files tested.}  Requiring changes to
-the parsing algorithm also complicates upgrades, as the changes must be
-preserved during the upgrade, and may clash with changes made by the
-developer.  \parsername{} allows the user to add new rules to the database
-without changing the parsing algorithm, unless the new log lines to be
-parsed require functionality not already provided by the algorithm.  If the
-new log lines do require new functionality, new actions can be added to the
-parser without modifying existing actions or other parts of the algorithm;
-only in the rare case that the new actions require support from other
-sections of the code will more extensive changes be required.
+obvious.\footnote{See \sectionref{yet-more-aborted-delivery-attempts} for a
+complication which occurs only four times in 93 log files tested.}
+Requiring changes to the parsing algorithm also complicates upgrades, as
+the changes must be preserved during the upgrade, and may clash with
+changes made by the developer.  \parsername{} allows the user to add new
+rules to the database without changing the parsing algorithm, unless the
+new log lines to be parsed require functionality not already provided by
+the algorithm.  If the new log lines do require new functionality, new
+actions can be added to the parser without modifying existing actions or
+other parts of the algorithm; only in the rare case that the new actions
+require support from other sections of the code will more extensive changes
+be required.
 
@@
 or extract additional data.  It does not appear to deal with any of the
-complications discussed in sections~\refwithpage{complications} (initial
-complications) and~\refwithpage{additional complications} (additional
+complications discussed in \sectionref{complications} (initial
+complications) and \sectionref{additional complications} (additional
 complications).  It does not differentiate between different types of
@@
 aid the user in understanding what each rule parses.  The rules are
-described in detail in section~\refwithpage{rules} but the fields are
-covered here.
+described in detail in \sectionref{rules} but the fields are covered here.
 
@@
 rejection; the individual fields will be described in the forthcoming
-section~\refwithpage{connections table}.  The results table has one or more
-rows per row in the connections table, depending on the specifics of the
+\sectionref{connections table}.  The results table has one or more rows per
+row in the connections table, depending on the specifics of the
 mail/connection; the fields will be covered in detail in
-section~\refwithpage{results table}.
+\sectionref{results table}.
 
@@
 
-Rules are discussed in detain in section~\refwithpage{rules}, but the rules
-table is covered here.  Rules are created by the user, not the parser, and
-will not be modified by the parser (except for the hits and hits\_total
-fields).  Rules parse the individual lines, telling the parser which fields
-to extract and what action to take for that line.
+Rules are discussed in detain in \sectionref{rules}, but the rules table is
+covered here.  Rules are created by the user, not the parser, and will not
+be modified by the parser (except for the hits and hits\_total fields).
+Rules parse the individual lines, telling the parser which fields to
+extract and what action to take for that line.
 
@@
         in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in section~\refwithpage{rule efficiency}.
+        discussed in \sectionref{rule efficiency}.
 
@@
     \item [action] The action the algorithm will take; a full list can be
-        found in Section~\refwithpage{actions-in-detail}.
+        found in \sectionref{actions-in-detail}.
 
@@
         time.  Rule ordering for efficiency is discussed in
-        section~\refwithpage{rule ordering for efficiency}.
+        \sectionref{rule ordering for efficiency}.
 
@@
         more specific rules to take precedence over more general rules
-        (described in section~\refwithpage{overlapping rules}).
+        (described in \sectionref{overlapping rules}).
 
@@
 details of the table containing the rules have already been described in
-section~\refwithpage{rule attributes},).  This section continues with a
-discussion of rule efficiency concerns, referring to the graphs in
-section~\refwithpage{graphs}, and finishes with an explanation of the
-algorithm used to generate a new \regex{} from unparsed lines.
-
-Please refer to section~\refwithpage{parser design} for a discussion of why
-the rules and actions have been separated in the parser's design.
+\sectionref{rule attributes},).  This section continues with a discussion
+of rule efficiency concerns, referring to the graphs in
+\sectionref{graphs}, and finishes with an explanation of the algorithm used
+to generate a new \regex{} from unparsed lines.
+
+Please refer to \sectionref{parser design} for a discussion of why the
+rules and actions have been separated in the parser's design.
 
@@
     \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See section~\refwithpage{rule efficiency}
-        for more details of sorting the rules.}  Parse a number of log
-        files using normal sorting, dump a textual representation of the
-        connections and results tables, and delete everything from those
-        tables.  Repeat with shuffled and reversed sorting.  If there are
-        no overlapping rules the tables from each run will be identical;
+        reversed rules.\footnote{See \sectionref{rule efficiency} for more
+        details of sorting the rules.}  Parse a number of log files using
+        normal sorting, dump a textual representation of the connections
+        and results tables, and delete everything from those tables.
+        Repeat with shuffled and reversed sorting.  If there are no
+        overlapping rules the tables from each run will be identical;
         differences indicate overlapping rules.  Which rules overlap can be
@@
     \item [action:] will be executed if the \regex{} matches successfully
-        (see section~\refwithpage{actions-in-detail} for full details).
+        (see \sectionref{actions-in-detail} for full details).
 
@@
     \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see section~\refwithpage{rule ordering for
-        efficiency}); hits is set to the number of successful matches at
-        the end of the parsing run, and hits\_total has hits added to it,
-        but is otherwise unused by the algorithm.
+        ordering the rules (see \sectionref{rule ordering for efficiency});
+        hits is set to the number of successful matches at the end of the
+        parsing run, and hits\_total has hits added to it, but is otherwise
+        unused by the algorithm.
 
@@
 
-Rule ordering was mentioned in section~\refwithpage{rule attributes} and
-will be covered in greater detail in this section.  At the time of writing
-there are 139 different rules, with the top 10\% matching the vast majority
-of the log lines, and the remaining log lines split across the other 90\%
-of the rules (as shown in graph~\refwithpage{rule hits graph}).  Assuming
-that the distribution of log lines is reasonably steady over time, program
+Rule ordering was mentioned in \sectionref{rule attributes} and will be
+covered in greater detail in this section.  At the time of writing there
+are 139 different rules, with the top 10\% matching the vast majority of
+the log lines, and the remaining log lines split across the other 90\% of
+the rules (as shown in graph~\refwithpage{rule hits graph}).  Assuming that
+the distribution of log lines is reasonably steady over time, program
 efficiency should benefit from trying more frequently matching rules before
@@
         which the parser will expand when it loads the rule (see the
-        \regex{} entry in section~\refwithpage{rule attributes}).  The
-        purpose of this phase is to utilise the existing knowledge to
-        create more accurate \regexes{}.  The new log lines are written to
-        a temporary file, which all subsequent stages use instead of the
-        original input file.
+        \regex{} entry in \sectionref{rule attributes}).  The purpose of
+        this phase is to utilise the existing knowledge to create more
+        accurate \regexes{}.  The new log lines are written to a temporary
+        file, which all subsequent stages use instead of the original input
+        file.
 
@@
     \item The database table containing the rules is dealt with in
-        section~\refwithpage{rule attributes}, and is not duplicated in
-        this section.
+        \sectionref{rule attributes}, and is not duplicated in this
+        section.
 
@@
 
-Please refer to section~\refwithpage{parser design} for a discussion of why
-the rules and actions have been separated in the parser's design.
+Please refer to \sectionref{parser design} for a discussion of why the
+rules and actions have been separated in the parser's design.
 
@@
 belongs to.\footnote{Unfortunately this statement is not completely
-accurate: see section~\refwithpage{timeouts-during-data-phase} for details.
-However in general there isn't any ambiguity about which data structure
-should be used for a given log line.}  The original connection will be
-discarded unsaved when the client disconnects if it doesn't have any data
-worth saving, i.e.\ no rejections.  One unsolved difficulty is
-distinguishing between different groups of rejections, e.g.\ when dealing
-with the following sequence:
+accurate: see \sectionref{timeouts-during-data-phase} for details.  However
+in general there isn't any ambiguity about which data structure should be
+used for a given log line.}  The original connection will be discarded
+unsaved when the client disconnects if it doesn't have any data worth
+saving, i.e.\ no rejections.  One unsolved difficulty is distinguishing
+between different groups of rejections, e.g.\ when dealing with the
+following sequence:
 
@@
 paths only; there are additional, uncommon paths which are excluded for the
-sake of clarity, but are described in section~\refwithpage{additional
+sake of clarity, but are described in \sectionref{additional
 complications}.
@@
 
-    Several
-more states and corresponding branches are omitted for the sake of clarity
-(these extra states are caused by solving the difficulties described in
-section~\refwithpage{additional complications}).
-
-The solution to the second last complication encountered
-(section~\refwithpage{out of order log lines}) requires that a list of the
-acceptable combinations of Postfix programs a mail can pass through;
-however this does not correspond to the \textit{actions\/} a mail must pass
-through.  Before that complication was overcome, i.e.\ for the majority of
-the development of this parser, there was nothing in the parser explicitly
-encoding the states or paths a mail could take; the vast majority of the
-parser has no need for this flow of data to be specified.  This eases the
-process of adding new actions (described in section~\refwithpage{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart below.
+Several more states and corresponding branches are omitted for the sake of
+clarity (these extra states are caused by solving the difficulties
+described in \sectionref{additional complications}).
+
+The solution to the \textit{out of order log log lines\/} complication
+encountered during development (\sectionref{out of order log lines})
+requires that a list of the acceptable combinations of Postfix programs a
+mail can pass through; however this does not correspond to the
+\textit{actions\/} a mail must pass through.  Before that complication was
+overcome, i.e.\ for the majority of the development of this parser, there
+was nothing in the parser explicitly encoding the states or paths a mail
+could take; the vast majority of the parser has no need for this flow of
+data to be specified.  This eases the process of adding new actions
+(described in \sectionref{adding new actions}), as there is no requirement
+to explicitly insert the action into an algorithmic version of the flow
+chart below.
 
@@
 
-Section~\refwithpage{actions-in-detail} explains the actions in substantive
-detail; this section omits such detail because it would clutter and confuse
-the algorithm description.  The flow chart in
-section~\refwithpage{flow-chart} should also be consulted while reading
-this section.  For the sake of clarity this description covers only the
-most common paths through the algorithm, as including every path would
-hinder understanding; the more uncommon paths are caused by the
-complications described in section~\refwithpage{additional complications},
-and are covered therein.
+\sectionref{actions-in-detail} explains the actions in substantive detail;
+this section omits such detail because it would clutter and confuse the
+algorithm description.  The flow chart in \sectionref{flow-chart} should
+also be consulted while reading this section.  For the sake of clarity this
+description covers only the most common paths through the algorithm, as
+including every path would hinder understanding; the more uncommon paths
+are caused by the complications described in \sectionref{additional
+complications}, and are covered therein.
 
@@
     \item If one or more mails were accepted there will be more log entries
-        for those mails later, see section~\refwithpage{mail-delivery}.
+        for those mails later, see \sectionref{mail-delivery}.
 
@@
 Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in section~\refwithpage{complications},
+own;\footnote{Previously discussed in \sectionref{complications},
 complication 3.} re-injection is somewhat awkward to explain because it
 overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to section~\refwithpage{tracking re-injected mail}.
+is deferred to \sectionref{tracking re-injected mail}.
 
@@
 must be detected using heuristics (see
-section~\refwithpage{identifying-bounce-notifications} for details).
-Bounce notifications are the primary example of internally generated mails,
-though there may be other types.
+\sectionref{identifying-bounce-notifications} for details).  Bounce
+notifications are the primary example of internally generated mails, though
+there may be other types.
 
@@
         tracked by the parent mail (action: track; flowchart:~10) ---
-        section~\refwithpage{tracking re-injected mail} discusses this in
+        \sectionref{tracking re-injected mail} discusses this in
         detail.
@@
         database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in section~\refwithpage{tracking
-        re-injected mail}.
+        Re-injected mails are described in \sectionref{tracking re-injected
+        mail}.
 
@@
 to the parent or children of mail re-injected due to forwarding, as
-explained in section~\refwithpage{tracking re-injected mail}.
+explained in \sectionref{tracking re-injected mail}.
 
@@
         indicator that the mail is a bounce notification, see
-        section~\refwithpage{identifying-bounce-notifications} for details.
+        \sectionref{identifying-bounce-notifications} for details.
 
@@
         Postfix programs which passes the valid combinations check (see
-        section~\refwithpage{out of order log lines}).  To ensure that the
-        mail can be committed the EXPIRY action sets a flag marking the
-        mail as expired; the flag later causes the valid combinations check
-        to be skipped, so the mail will be committed.
+        \sectionref{out of order log lines}).  To ensure that the mail can
+        be committed the EXPIRY action sets a flag marking the mail as
+        expired; the flag later causes the valid combinations check to be
+        skipped, so the mail will be committed.
 
@@
         both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see section~\refwithpage{discarding cleanup
-        lines} for details.
+        out of order log lines; see \sectionref{discarding cleanup lines}
+        for details.
 
@@
         entries may have caused the state table entry to already exist (see
-        section~\refwithpage{pickup logging after cleanup}); otherwise it
-        is created.  The data from the log line is then saved to the state
+        \sectionref{pickup logging after cleanup}); otherwise it is
+        created.  The data from the log line is then saved to the state
         table entry.
@@
         which case there is not.  See
-        section~\refwithpage{timeouts-during-data-phase} for the gory
-        details.
+        \sectionref{timeouts-during-data-phase} for the gory details.
 
@@
 avail of the new functionality.} The action may need to extend the list of
-valid combinations described in section~\refwithpage{out of order log
-lines} if the addition creates a different set of acceptable programs, but
-this is unlikely to occur, as it would require parsing log lines from
-Postfix components the parser currently ignores.\footnote{The mail server
-used for development does not utilise the \daemon{lmtp} or \daemon{virtual}
+valid combinations described in \sectionref{out of order log lines} if the
+addition creates a different set of acceptable programs, but this is
+unlikely to occur, as it would require parsing log lines from Postfix
+components the parser currently ignores.\footnote{The mail server used for
+development does not utilise either the \daemon{lmtp} or \daemon{virtual}
 delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse their log lines is a
-simple process, though if their behaviour differs significantly from the
-\daemon{smtp} or \daemon{local} delivery agents new actions may be
+from those components.  Adding new rules to parse those component's log
+lines is a simple process, though if their behaviour differs significantly
+from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
 required.  The mail server in question is a production mail server handling
@@
 for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in section~\refwithpage{tracking
-re-injected mail}, but unlike the solution described therein bounce
-notifications do not eventually have a log line which identifies their
-source.  Heuristics must be used to identify bounce notifications, and
-those heuristics are:
+problem of re-injected mails discussed in \sectionref{tracking re-injected
+mail}, but unlike the solution described therein bounce notifications do
+not eventually have a log line which identifies their source.  Heuristics
+must be used to identify bounce notifications, and those heuristics are:
 
@@
 available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see
-sections~\refwithpage{connections table} and~\refwithpage{results table}).
+requires many more fields be populated (see \sectionref{connections table}
+and \sectionref{results table}).
 
@@
 disconnections are either unimportant or are handled elsewhere in the
-algorithm (section~\refwithpage{timeouts-during-data-phase}).  Sadly
-Postfix doesn't log a message saying the mail has been discarded, as should
-be expected by now.  The checks to identify this happening are:
+algorithm (\sectionref{timeouts-during-data-phase}).  Sadly Postfix doesn't
+log a message saying the mail has been discarded, as should be expected by
+now.  The checks to identify this happening are:
 
@@
 This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see section~\refwithpage{discarding cleanup lines}
-for details.
+\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
+details.
 
@@
 active connection the last accepted mail may need to be discarded, as
-detailed in section~\refwithpage{timeouts-during-data-phase}.
+detailed in \sectionref{timeouts-during-data-phase}.
 
@@
 Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in sections~\refwithpage{tracking
-re-injected mail},~\refwithpage{discarding cleanup
-lines}~or~\refwithpage{pickup logging after cleanup}.  In the 93 log files
-used for testing this occurs only five times in 60,721,709 log lines, but
-for completeness of the algorithm it should be dealt with.  The five
-occurrences in the test log files have the same characteristics: the
-\daemon{local} log line showing delivery to a local mailbox occurs after
-the \daemon{qmgr} log line showing removal of the mail from the queue
-because delivery is completed.  This causes problems: the mail is not
-complete, so entry into the database fails; a new mail is created when the
-\daemon{local} line is parsed and remains in the state tables; four
-warnings are issued per pair of out of order log lines.
+dealt with by the techniques described in \sectionref{tracking re-injected
+mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
+after cleanup}.  In the 93 log files used for testing this occurs only five
+times in 60,721,709 log lines, but for completeness of the algorithm it
+should be dealt with.  The five occurrences in the test log files have the
+same characteristics: the \daemon{local} log line showing delivery to a
+local mailbox occurs after the \daemon{qmgr} log line showing removal of
+the mail from the queue because delivery is completed.  This causes
+problems: the mail is not complete, so entry into the database fails; a new
+mail is created when the \daemon{local} line is parsed and remains in the
+state tables; four warnings are issued per pair of out of order log lines.
 
@@
 The aborted delivery attempts described in
-section~\refwithpage{aborted-delivery-attempts} occur frequently, but the
-aborted delivery attempts described in this section only occur four times
-in the 93 log files used for testing.  The symptoms are the same as in
-section~\refwithpage{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
+\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
+delivery attempts described in this section only occur four times in the 93
+log files used for testing.  The symptoms are the same as in
+\sectionref{aborted-delivery-attempts}, except that there \textit{is\/} a
+\daemon{cleanup} log line; there does not appear to be anything in the log
+file to explain why there are no further log messages.  The only way to
+detect these mails is to periodically scan all mails in the state tables,
+deleting any mails displaying the following characteristics:
 
@@
     \item The appropriate action to take --- discussed in
-        section~\refwithpage{mails-covered}.
+        \sectionref{mails-covered}.
 
@@
 the best way to detect mails which were not handled properly (many of the
-complications discussed in section~\refwithpage{additional complications}
-were detected in this way).  The parser issues warnings when it detects any
+complications discussed in \sectionref{additional complications} were
+detected in this way).  The parser issues warnings when it detects any
 errors, some of which may alert the user to a problem, e.g.\ when a queueid
@@
 this problem.} or when there are problems tracking a child mail (see
-section~\refwithpage{tracking re-injected mail}).  There should be few or
-no warnings when parsing, and when finished parsing the state table should
+\sectionref{tracking re-injected mail}).  There should be few or no
+warnings when parsing, and when finished parsing the state table should
 only contain entries for mails which had yet to be delivered when the log
@@
         (e.g.\ if a mail is tracked twice as described in
-        section~\refwithpage{tracking re-injected mail}), and no such
-        warnings are produced during the test runs.
+        \sectionref{tracking re-injected mail}), and no such warnings are
+        produced during the test runs.
 
@@
     \item When dealing with the complications described in
-        sections~\refwithpage{complications} and~\refwithpage{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} problem described in section~\refwithpage{out of order
-        log lines} imposes conditions which each reassembled mail must
-        comply with to be acceptable.
+        \sectionref{complications} \sectionref{additional complications}
+        the solutions are as specific and restrictive as possible, with the
+        goal of minimising the number of false positives.  In addition the
+        solution to the \textit{Out of order log files\/} problem described
+        in \sectionref{out of order log lines} imposes conditions which
+        each reassembled mail must comply with to be acceptable.
 
@@
         and 2.3.11; this may possibly have changed in Postfix 2.4.}
-        Rectifying this has already been described in
-        section~\refwithpage{logging helo}.
+        Rectifying this has already been described in \sectionref{logging
+        helo}.
 
@@
         by acceptance (this has already been mentioned in
-        section~\refwithpage{connection reuse}).  I don't believe it's
-        possible to make this distinction given the data Postfix logs,
-        though it might be possible to write a policy server to provide
-        additional logging.
+        \sectionref{connection reuse}).  I don't believe it's possible to
+        make this distinction given the data Postfix logs, though it might
+        be possible to write a policy server to provide additional logging.
 
@@
 Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged
-(section~\refwithpage{postfix background}), enabling users to easily extend
-the parser to deal with new log lines is a design imperative
-(section~\refwithpage{parser design}).  Despite the resulting initial
-increase in complexity the job is quite tractable, though it does raise
-interesting efficiency and optimisation questions (answered in
-section~\refwithpage{rule efficiency}).  Providing a tool
-(section~\refwithpage{creating new rules}) to ease the generation of
-\regexes{} from unparsed log lines should greatly help users add rules to
-parse formerly unparsed lines.
+administrators can cause new lines to be logged (\sectionref{postfix
+background}), enabling users to easily extend the parser to deal with new
+log lines is a design imperative (\sectionref{parser design}).  Despite the
+resulting initial increase in complexity the job is quite tractable, though
+it does raise interesting efficiency and optimisation questions (answered
+in \sectionref{rule efficiency}).  Providing a tool (\sectionref{creating
+new rules}) to ease the generation of \regexes{} from unparsed log lines
+should greatly help users add rules to parse formerly unparsed lines.
 
 The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions
-(section~\refwithpage{parsing-algorithm}) an almost trivial task.  The
-simplicity of adding a new action (section~\refwithpage{adding new
-actions}) frees the implementor from worrying about how their action might
-disrupt parsing of other lines or the behaviour of other actions, allowing
-their concentration to be focused on correctly implementing their action.
+lines, it makes adding new actions (\sectionref{parsing-algorithm}) an
+almost trivial task.  The simplicity of adding a new action
+(\sectionref{adding new actions}) frees the implementor from worrying about
+how their action might disrupt parsing of other lines or the behaviour of
+other actions, allowing their concentration to be focused on correctly
+implementing their action.
 
@@
 and actions is also interesting, and is discussed beside the flow chart
-depicting said emergent behaviour (section~\refwithpage{flow-chart}).  This
-emergent behaviour greatly eases
+depicting said emergent behaviour (\sectionref{flow-chart}).  This emergent
+behaviour greatly eases
 
@@
 the existing functionality.  The complications described in
-section~\refwithpage{additional complications} were solved in the order
-they are described in, and that order closely resembles the frequency with
-which they occur; the most frequently occurring complications dominate the
+\sectionref{additional complications} were solved in the order they are
+described in, and that order closely resembles the frequency with which
+they occur; the most frequently occurring complications dominate the
 warning messages produced, and so naturally they are the first
@@
 parse new log lines without modifying the core parsing algorithm; see
-section~\refwithpage{why separate rules and algorithm} for a discussion of
-why the separation of rules from the algorithm is beneficial.
+\sectionref{why separate rules and algorithm} for a discussion of why the
+separation of rules from the algorithm is beneficial.
 
@@
         is made to correlate log lines or deal with the difficulties
-        documented in sections~\refwithpage{complications}
-        and~\refwithpage{additional complications}.\footnote{If any attempt
-        is made there is no reference to it in the documentation available
-        on the website.} The source code is available in an obfuscated form
-        only (presumably for a fee), and the product is quite expensive, as
-        it requires a license per report which is to be generated; in
-        contrast \parsername{} is free and the code is freely available.
-        The web interface allows creation of dynamic reports based on any
-        field, but due to the associated cost the author has not
-        experimented with it.  Presumably as the company charges per report
-        the user wishes to generate the data store (if there is one) is
-        deliberately inaccessible and undocumented to prevent the user
-        bypassing the program and generating their own reports.
+        documented in \sectionref{complications} \sectionref{additional
+        complications}.\footnote{If any attempt is made there is no
+        reference to it in the documentation available on the website.} The
+        source code is available in an obfuscated form only (presumably for
+        a fee), and the product is quite expensive, as it requires a
+        license per report which is to be generated; in contrast
+        \parsername{} is free and the code is freely available.  The web
+        interface allows creation of dynamic reports based on any field,
+        but due to the associated cost the author has not experimented with
+        it.  Presumably as the company charges per report the user wishes
+        to generate the data store (if there is one) is deliberately
+        inaccessible and undocumented to prevent the user bypassing the
+        program and generating their own reports.
 
@@
         of the available data and makes no attempt to deal with the other
-        complications described in sections~\refwithpage{complications}
-        and~\refwithpage{additional complications}.  \parsername{} uses an
+        complications described in \sectionref{complications}
+        \sectionref{additional complications}.  \parsername{} uses an
         \SQL{} database to make accessing the extracted data as easy as
@@
         attempt to address any of the complications in
-        sections~\refwithpage{complications} and~\refwithpage{additional
-        complications} except for correlating by queueid.
+        \sectionref{complications} \sectionref{additional complications}
+        except for correlating by queueid.
 
@@
     \item [Log Mail Analyser] Please see the previous in-depth discussion
-        of Log Mail Analyser in section~\refwithpage{prior art}.
+        of Log Mail Analyser in \sectionref{prior art}.
 
@@
 anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in section~\refwithpage{rule efficiency}.
-The graphs in the first section cover parser scalability, demonstrating
-that performance scales linearly with input size.  The impact of rule
-ordering is shown in section~\refwithpage{rule ordering graphs}, and the
-anomalous dips and peaks apparent in some graphs are explained.  The third
-group of graphs vividly shows the huge impact that caching compiled
-\regexes{} has on parser performance.  This section concludes with a
-breakdown of the rule hits accumulated during a single test run of the
-parser.
+illustrate the topics discussed in \sectionref{rule efficiency}.  The
+graphs in the first section cover parser scalability, demonstrating that
+performance scales linearly with input size.  The impact of rule ordering
+is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
+peaks apparent in some graphs are explained.  The third group of graphs
+vividly shows the huge impact that caching compiled \regexes{} has on
+parser performance.  This section concludes with a breakdown of the rule
+hits accumulated during a single test run of the parser.
 
@@
 reason for the anomalous log files has already been explained in
-section~\refwithpage{Why are there dips in the graphs?}.
+\sectionref{Why are there dips in the graphs?}.
 
@@
 The graphs presented in this section illustrate the topics discussed in
-section~\refwithpage{rule efficiency}.  The first collection of graphs are
-about parser scalability, showing the linear relationship between execution
-time and input size.  Section~\refwithpage{rule ordering graphs} relates to
-the effect of rule ordering on execution time, and the unexpected
-consequences of specific inputs.  The necessity of caching compiled
-\regexes{} is attested to by the third group of graphs, where the
-difference between caching and discarding is staggering.  This section
-concluded with a breakdown of the rule hits accumulated during a single
-test run of the parser.
+\sectionref{rule efficiency}.  The first collection of graphs are about
+parser scalability, showing the linear relationship between execution time
+and input size.  \sectionref{rule ordering graphs} relates to the effect of
+rule ordering on execution time, and the unexpected consequences of
+specific inputs.  The necessity of caching compiled \regexes{} is attested
+to by the third group of graphs, where the difference between caching and
+discarding is staggering.  This section concluded with a breakdown of the
+rule hits accumulated during a single test run of the parser.
 

### 2008-02-07 doc/logparser.tex
Added new commands giving the number of log files and number of rules, to make changing easier, and replaced all occurrences of then numbers. The table summarising the most effective restrictions is now automatically generated.

@@
     \hspace*{2em}%
+}
+
+\newcommand{\numberOFlogFILES}[0]{%
+    93%
+}
+
+\newcommand{\numberOFrules}[0]{%
+    142%
 }
@@
 Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of 93 log files taken from a mail server handling mail for over
-1000 users.  Coverage is described both in terms of the fraction of lines
-parsed and the fraction of mails and connections successfully reconstructed
-by the parsing algorithm, including dealing with false negatives and a
-discussion of the difficulties in identifying false positives.
+over a set of \numberOFlogFILES{} log files taken from a mail server
+handling mail for over 1000 users.  Coverage is described both in terms of
+the fraction of lines parsed and the fraction of mails and connections
+successfully reconstructed by the parsing algorithm, including dealing with
+false negatives and a discussion of the difficulties in identifying false
+positives.
 
@@
         spam mail at the moment; the most effective restrictions over the
-        93 log files used in testing the parser are shown in
-        table~\refwithpage{Summary of rejections}.  Greylisting is
+        \numberOFlogFILES{} log files used in testing the parser are shown
+        in table~\refwithpage{Summary of rejections}.  Greylisting is
         obviously worth using, at least at the moment, particularly when
@@
             \caption{Summary of rejections}\label{Summary of rejections}
-            \begin{tabular}[]{lr}
-                Restriction type    & Total rejections  \\
-                Total rejections    & 5283739           \\
-                Unknown recipient   & 2338000           \\
-                Unknown sender      & 864148            \\
-                \RBL{}s             & 807185            \\
-                Greylisting         & 497244            \\
-                Non-\FQDN{} HELO    & 403652            \\
-            \end{tabular}
+            \input{build/restriction-table-include.tex}
         \end{table}
@@
 obvious.\footnote{See \sectionref{yet-more-aborted-delivery-attempts} for a
-complication which occurs only four times in 93 log files tested.}
-Requiring changes to the parsing algorithm also complicates upgrades, as
-the changes must be preserved during the upgrade, and may clash with
-changes made by the developer.  \parsername{} allows the user to add new
-rules to the database without changing the parsing algorithm, unless the
-new log lines to be parsed require functionality not already provided by
-the algorithm.  If the new log lines do require new functionality, new
-actions can be added to the parser without modifying existing actions or
-other parts of the algorithm; only in the rare case that the new actions
-require support from other sections of the code will more extensive changes
-be required.
+complication which occurs only four times in \numberOFlogFILES{} log files
+tested.} Requiring changes to the parsing algorithm also complicates
+upgrades, as the changes must be preserved during the upgrade, and may
+clash with changes made by the developer.  \parsername{} allows the user to
+add new rules to the database without changing the parsing algorithm,
+unless the new log lines to be parsed require functionality not already
+provided by the algorithm.  If the new log lines do require new
+functionality, new actions can be added to the parser without modifying
+existing actions or other parts of the algorithm; only in the rare case
+that the new actions require support from other sections of the code will
+more extensive changes be required.
 
@@
 daily basis).  When generating the data for the graphs included in
-appendix~\refwithpage{graphs}, 93 log files (totaling 10.08 GB, 60.72
-million lines) were each parsed 10 times, the first run discarded, and the
-remaining 9 runs averaged.  The first run is discarded for two reasons:
+appendix~\refwithpage{graphs}, \numberOFlogFILES{} log files (totaling
+10.08 GB, 60.72 million lines) were each parsed 10 times, the first run
+discarded, and the remaining 9 runs averaged.  The first run is discarded
+for two reasons:
 
@@
 covered in greater detail in this section.  At the time of writing there
-are 139 different rules, with the top 10\% matching the vast majority of
-the log lines, and the remaining log lines split across the other 90\% of
-the rules (as shown in graph~\refwithpage{rule hits graph}).  Assuming that
-the distribution of log lines is reasonably steady over time, program
-efficiency should benefit from trying more frequently matching rules before
-those which match less frequently.  To test this hypothesis three full test
-runs were performed with different rule orderings:
+are \numberOFrules{} different rules, with the top 10\% matching the vast
+majority of the log lines, and the remaining log lines split across the
+other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
+Assuming that the distribution of log lines is reasonably steady over time,
+program efficiency should benefit from trying more frequently matching
+rules before those which match less frequently.  To test this hypothesis
+three full test runs were performed with different rule orderings:
 
@@
 mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the 93 log files used for testing this occurs only five
-times in 60,721,709 log lines, but for completeness of the algorithm it
-should be dealt with.  The five occurrences in the test log files have the
-same characteristics: the \daemon{local} log line showing delivery to a
-local mailbox occurs after the \daemon{qmgr} log line showing removal of
-the mail from the queue because delivery is completed.  This causes
-problems: the mail is not complete, so entry into the database fails; a new
-mail is created when the \daemon{local} line is parsed and remains in the
-state tables; four warnings are issued per pair of out of order log lines.
+after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
+occurs only five times in 60,721,709 log lines, but for completeness of the
+algorithm it should be dealt with.  The five occurrences in the test log
+files have the same characteristics: the \daemon{local} log line showing
+delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
+removal of the mail from the queue because delivery is completed.  This
+causes problems: the mail is not complete, so entry into the database
+fails; a new mail is created when the \daemon{local} line is parsed and
+remains in the state tables; four warnings are issued per pair of out of
+order log lines.
 
@@
 \sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the 93
-log files used for testing.  The symptoms are the same as in
-\sectionref{aborted-delivery-attempts}, except that there \textit{is\/} a
-\daemon{cleanup} log line; there does not appear to be anything in the log
-file to explain why there are no further log messages.  The only way to
-detect these mails is to periodically scan all mails in the state tables,
-deleting any mails displaying the following characteristics:
+delivery attempts described in this section only occur four times in the
+\numberOFlogFILES{} log files used for testing.  The symptoms are the same
+as in \sectionref{aborted-delivery-attempts}, except that there
+\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
+anything in the log file to explain why there are no further log messages.
+The only way to detect these mails is to periodically scan all mails in the
+state tables, deleting any mails displaying the following characteristics:
 
@@
 easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser has 139 rules, fully parsing 93 contiguous log files
-from Postfix 2.2 and 2.3), and new rules are easy to add.  Maintaining full
-coverage is hard because other servers have different restrictions with
-custom messages, \RBL{} messages change over time, major releases of
-Postfix change warning messages (usually adding more information), etc.,\
-so over time the log lines drift and change.  Graph~\refwithpage{rule hits
-graph} shows the number of hits for each rule over all 93 log files; it's
-obvious that a small number of rules match the vast majority of the lines,
-and more than half the rules match fewer than 100 times.
+of writing the parser has \numberOFrules{} rules, fully parsing
+\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
+rules are easy to add.  Maintaining full coverage is hard because other
+servers have different restrictions with custom messages, \RBL{} messages
+change over time, major releases of Postfix change warning messages
+(usually adding more information), etc.,\ so over time the log lines drift
+and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
+for each rule over all \numberOFlogFILES{} log files; it's obvious that a
+small number of rules match the vast majority of the lines, and more than
+half the rules match fewer than 100 times.
 
@@
 Warnings are issued for any lines which are not parsed; no warnings are
-issued for unparsed lines while testing with the 93 test log files, so it
-can be safely concluded that there are zero false negatives.  False
-positives are harder to quantify: short of examining each of the 60,721,709
-log lines and determining which \regex{} parsed it, there is no way to be
-sure that every line was parsed by the correct \regex{}, making it
-impossible to quantify the false positive rate.  The author is confident
-that the false positive rate is extremely low, approaching if not reaching
-zero, because the \regexes{} used are as specific as possible and care has
-been taken to ensure that there are no unintentionally overlapping
+issued for unparsed lines while testing with the \numberOFlogFILES{} test
+log files, so it can be safely concluded that there are zero false
+negatives.  False positives are harder to quantify: short of examining each
+of the 60,721,709 log lines and determining which \regex{} parsed it, there
+is no way to be sure that every line was parsed by the correct \regex{},
+making it impossible to quantify the false positive rate.  The author is
+confident that the false positive rate is extremely low, approaching if not
+reaching zero, because the \regexes{} used are as specific as possible and
+care has been taken to ensure that there are no unintentionally overlapping
 \regexes{}.
@@
 
-At the time of writing the parser is being tested with 93 log files.  There
-are 5 warnings produced, but because the parser errs on the side of
-producing more warnings rather than fewer those 5 warnings represent 3
-instances of 1 problem: 3 connections started before the first log file, so
-their initial log entries are missing, leading to warnings when their log
-lines are parsed.
+At the time of writing the parser is being tested with \numberOFlogFILES{}
+log files.  There are 5 warnings produced, but because the parser errs on
+the side of producing more warnings rather than fewer those 5 warnings
+represent 3 instances of 1 problem: 3 connections started before the first
+log file, so their initial log entries are missing, leading to warnings
+when their log lines are parsed.
 
@@
 Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the 93 test log files the state tables contain 18
-entries, breaking down into:
+the logs.  After parsing the \numberOFlogFILES{} test log files the state
+tables contain 18 entries, breaking down into:
 

### 2008-02-10 doc/logparser.tex
Documented the parsing and verification of a random sample of log lines.

@@
 
-XXX PICK A RANDOM SAMPLE OF LOG LINES AND ENSURE THEY ARE PARSED
-PROPERLY\@.
-
 Warnings are issued for any lines which are not parsed; no warnings are
@@
 is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate.  The author is
-confident that the false positive rate is extremely low, approaching if not
-reaching zero, because the \regexes{} used are as specific as possible and
-care has been taken to ensure that there are no unintentionally overlapping
-\regexes{}.
+making it impossible to quantify the false positive rate; however a random
+sample of 6039 log lines was parsed and the results checked manually to
+ensure that the correct \regex{} parsed each line.  The sample was
+generated by running the following command:
+
+\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
+
+\noindent{}to randomly extract roughly one line in every 10,000.  Although
+on initial appearances exercising only 36 rules (from a total of
+\numberOFrules{}) when parsing 6039 log lines seems quite low, after
+examining graph~\refwithpage{rule hits graph} it becomes apparent that such
+a low hit rate is to be expected; the reader should also bear in mind that
+even when parsing all \numberOFlogFILES{} log files not all the rules are
+exercised (some of the rules are for parsing log lines which only appear in
+other log files).
 
@@
 files ended, or started before the log files began.
+
+XXX PICK A RANDOM SAMPLE OF LOG LINES AND ENSURE THEY ARE PARSED
+PROPERLY\@.
 

### 2008-02-20 doc/logparser.tex
Documented DELETE action and associated complication.

@@
 \newcommand{\numberOFrules}[0]{%
-    142%
+    146%
 }
@@
 \end{abstract}
+
+XXX COVERAGE\@.
 
@@
         or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO when it rejects an \SMTP{} command, but it is quite
-        easy to rectify this:
+        logs the HELO hostname when it rejects an \SMTP{} command, but it
+        is quite easy to rectify this:
 
@@
         with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be
+        associated with the correct mail.  There is guaranteed to be a
         queueid when the DATA command has been reached, and thus it will be
@@
         correctly.
+
+        Logging the HELO hostname in this fashion also prevents the
+        complication described in \sectionref{Mail deleted before delivery
+        is attempted} from occurring, because when the HELO hostname is
+        logged one recipient address will be logged also.
 
@@
         case.
+
+    \item [DELETE] Deals with mail deleted by the administrator using the
+        administrative command \daemon{postsuper}.  This action adds a
+        dummy recipient address if required, then invokes the COMMIT action
+        to handle adding the mail to the database.  The complication this
+        action deals with is described fully in \sectionref{Mail deleted
+        before delivery is attempted}.
 
@@
 
+\subsubsection{Mail deleted before delivery is attempted}
+
+\label{Mail deleted before delivery is attempted}
+
+Postfix logs the recipient address when delivery of a mail is attempted, so
+if delivery has yet to be attempted the parser cannot determine the
+recipient address or addresses.  This is a problem when mail is arriving
+faster than Postfix can attempt delivery, and the administrator deletes
+some of the mail (because it's the result of a mail loop, mail bomb, or
+joe-job) before Postfix has had a chance to try to deliver it.  In this
+case the recipient address will not have been logged, so a dummy recipient
+address needs to be added, as every mail is required to have at least one
+recipient.  This complication is generally observed in few log files, but
+typically when it does occur there will be many instances of it, closely
+grouped.
+
 \subsection{Conclusion}

### 2008-02-20 doc/logparser.tex
Documented verification of correct parsing in the coverage section.

@@
 \end{abstract}
-
-XXX COVERAGE\@.
 
@@
 
-XXX PICK A RANDOM SAMPLE OF LOG LINES AND ENSURE THEY ARE PARSED
-PROPERLY\@.
-
 At the time of writing the parser is being tested with \numberOFlogFILES{}
@@
     \item When dealing with the complications described in
-        \sectionref{complications} \sectionref{additional complications}
-        the solutions are as specific and restrictive as possible, with the
-        goal of minimising the number of false positives.  In addition the
-        solution to the \textit{Out of order log files\/} problem described
-        in \sectionref{out of order log lines} imposes conditions which
-        each reassembled mail must comply with to be acceptable.
+        \sectionref{complications} and \sectionref{additional
+        complications} the solutions are as specific and restrictive as
+        possible, with the goal of minimising the number of false
+        positives.  In addition the solution to the \textit{Out of order
+        log files\/} problem described in \sectionref{out of order log
+        lines} imposes conditions which each reassembled mail must comply
+        with to be acceptable.
 
@@
 
-Despite the lack of numerical evidence, the author is confident that the
-false positive rate when reconstructing a mail is exceedingly low, if not
-approaching zero.
+Whereas verifying by inspection that the parser correctly deals with all
+60,721,709 lines in the test logs is infeasible, verifying a subset of
+those logs is a tractable, if extremely time consuming, task.  A sample of
+logs was obtained by randomly selecting a log file:
+
+\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
+\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
+
+The first 6000 lines of this log file (roughly 0.01\% of the total) was
+extracted:
+
+\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
+
+It is important that the log lines used are contiguous so that all log
+entries are present for as many of the connections/mails as possible.  This
+log segment was parsed with all debugging options enabled, resulting in
+167,448 lines of output (27.908 lines of output per line of input).  All
+167,448 lines were examined in conjunction with the input log and a dump of
+the resulting database, verifying that for each of the input lines the
+parser used the correct rule, executed the correct action, which in turn
+produced the correct result, and inserted the correct data in the database.
+The log segment produced 4 warnings, 10 mails remaining in the state
+tables, and 1625 connections entered in the database.
+
+Given the circumstantial and experimental evidence detailed above, the
+author is confident that the false positive rate when reconstructing a mail
+is exceedingly low, if not approaching zero.
 

### 2008-02-24 doc/logparser.tex
Corrected logging of recipients with HELO line.

@@
         complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring, because when the HELO hostname is
-        logged one recipient address will be logged also.
+        is attempted} from occurring when, but only in the case where there
+        is a single recipient; in that case the recipient address will be
+        logged also.
 

### 2008-02-25 doc/logparser.tex
Updated 'aborted delivery attempts' section: there's no longer a restriction on     the number of smtpd log lines. Documented out of order bounce line complication.

@@
 \newcommand{\numberOFrules}[0]{%
-    146%
+    157%
 }
@@
         be removed.  The action also marks the mail as a bounce
-        notification.
+        notification.  To deal with complication \sectionref{Further out of
+        order log lines} this action checks a cache of recent bounce mails
+        to avoid creating bogus bounce mails when lines are logged out of
+        order.
 
@@
 examined when the disconnection action is executed.  Each mail is checked
-for the following characteristics:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there exactly two \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line, nothing
-        else.
-
-\end{itemize}
-
-If both checks are successful then the mail is assumed to be one of the
+for the following: is the mail missing its \daemon{cleanup} log message?
+Every mail which passes through Postfix will have a \daemon{cleanup} line;
+lack of a \daemon{cleanup} line is a sure sign the mail didn't make it too
+far.
+
+If this check is successful then the mail is assumed to be one of the
 offending bogus mails and is discarded.  There will be no further entries
@@
 grouped.
+
+\subsubsection{Further out of order log lines}
+
+\label{Further out of order log lines}
+
+This is yet another complication which only occurs during periods of
+extremely high load, when process scheduling vagaries and even hard disk
+access times cause strange behaviour.  In this complication bounce
+notification mails are created, delivered and deleted from the queue,
+\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
+with this the COMMIT action maintains a cache of recently committed bounce
+notification mails, which the BOUNCE action subsequently checks if the
+bounce mail isn't already in the state tables. If a mail exists in the
+cache under the correct queueid, and its start time is less than ten
+seconds before the timestamp of the bounce log line, it is assumed that the
+bounce notification mail has already been processed and the BOUNCE action
+does not create one.  If the queueid exists it is removed from the cache,
+as it has either just been used or it is too old to use in future.  Whether
+the BOUNCE action creates a mail or finds one existing in the state tables,
+it flags the mail as having been seen by the BOUNCE action; if this flag is
+present the COMMIT action will not add the mail to the cache of recent
+bounce notification mails.\footnote{This is not required to correctly deal
+with the complication, but is an optimisation to reduce memory usage of the
+parser; on the occasions the author has observed this action occurring
+there have been a huge number of bounce notification mails generated; if
+every bounce notification mail was cached it would dramatically increase
+the memory requirements of the parser.}  The cache of bounce notification
+mails will be pruned whenever the parser's state is saved, though if the
+size of the cache ever becomes a problem it could be pruned periodically to
+keep it in check.
 

### 2008-02-26 doc/logparser.tex
Documented new valid combo. Improved introduction, referring to the new content, and pointing out novel     things.  Needs more work, I think.

@@
 Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.
+understanding the paper, parser and algorithm.  It introduces the idea of
+using a database schema as an \API{}, providing an interface to the data
+gathered that is language-neutral.  The unusual separation of rules,
+actions and algorithm is discussed, giving the reasons that approach was
+taken when designing the parser.
 
@@
 parsing and the results gleaned from parsing.  The database schema used is
-described in section~\ref{database schema}, explaining in detail the tables
+described in \sectionref{database schema}, explaining in detail the tables
 used for storing the data gleaned from the logs and the table that stores
@@
 purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  This section concludes with a
-discussion of rule efficiency concerns.
-
-Section~\ref{parsing-algorithm} contains the meat of the paper, describing
+sample data it matches successfully against.  The pros and cons of
+overlapping rules are discussed, including techniques for detecting
+unintentional overlaps.  Rule efficiency concerns are discussed, in
+particular the optimisations used by the algorithm, with reference to the
+graphs in appendix~\refwithpage{graphs}.  The final section is a
+description of using the tools provided with the parser to generate a
+\regex{} from related log lines for new rules.
+
+Section~\ref{parsing-algorithm} contains the core of the paper, describing
 a naive parsing algorithm and the complications encountered which shaped
-the full algorithm, followed by a comprehensive explanation of the
-different stages of the algorithm, the actions taken during execution of
-the algorithm, and further complications whose solution completes the
-algorithm.
+the full algorithm.  A flow chart and a discussion of the emergent
+behaviour exhibited by the algorithm accompanies a comprehensive
+explanation of the different stages of the algorithm.  The actions taken
+during execution of the algorithm are described, followed by the process of
+adding a new action.  The section concludes with an in-depth description of
+the further complications and their solutions which complete the algorithm.
 
@@
 false negatives and a discussion of the difficulties in identifying false
-positives.
+positives.  A random sampling of log lines are parsed, and the correctness
+of the results manually verified, as part of determining the coverage of
+the parser.
 
 Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with the limitations, with
-the goal of improving parsing and reproduction of the journey a mail takes
-through Postfix.
-
-Section~\ref{conclusion} is contains the paper's conclusion.
+algorithm, then suggests some ways of dealing with them, with the goal of
+improving parsing and reproduction of the journey a mail takes through
+Postfix.
+
+Section~\ref{conclusion} contains the paper's conclusion.
 
@@
 The bibliography contains references to the resources used in developing
-the algorithm, writing the program and preparing this paper.  Also listed
+the algorithm, writing the program, and preparing this paper.  Also listed
 are some additional resources expected to be helpful in understanding
@@
 Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in section~\ref{rule efficiency}, rule efficiency, and explains
+discussed in \sectionref{rule efficiency}, rule efficiency, and explains
 the irregularities observed in these graphs.
@@
 
+    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
+        client, then deleted by the administrator before any delivery
+        attempt (administrator deleting unwanted mail, typically due to a
+        mail loop or joe-job).  Notice that \daemon{postsuper} is required,
+        not optional, for this combination.
+
 \end{description}

### 2008-02-27 doc/logparser.tex
Minor changes noticed when reviewing changes for Carl.

@@
 \newcommand{\numberOFrules}[0]{%
-    157%
+    168%
 }
@@
         is a single recipient; in that case the recipient address will be
-        logged also.
+        logged also, but when there are multiple recipients none are
+        logged.
 
@@
 
+    \item [DELETE] Deals with mail deleted by the administrator using the
+        administrative command \daemon{postsuper}.  This action adds a
+        dummy recipient address if required, then invokes the COMMIT action
+        to handle adding the mail to the database.  The complication this
+        action deals with is described fully in \sectionref{Mail deleted
+        before delivery is attempted}.
+
     \item [DISCONNECT] Deal with the remote client disconnecting: enter the
@@
         case.
-
-    \item [DELETE] Deals with mail deleted by the administrator using the
-        administrative command \daemon{postsuper}.  This action adds a
-        dummy recipient address if required, then invokes the COMMIT action
-        to handle adding the mail to the database.  The complication this
-        action deals with is described fully in \sectionref{Mail deleted
-        before delivery is attempted}.
 
@@
         by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  I don't believe it's possible to
-        make this distinction given the data Postfix logs, though it might
-        be possible to write a policy server to provide additional logging.
+        \sectionref{connection reuse}).  It doesn't appear to be possible
+        to make this distinction given the data Postfix logs, though it
+        might be possible to write a policy server to provide additional
+        logging.
 

### 2008-03-03 doc/logparser.tex
Changes after proof-reading.

@@
     Parsing Postfix logs is much more difficult than it first appears, but
-    it is possible to achieve a high degree of accuracy in understanding
-    the logs, and thus reconstructing the actions taken by Postfix to
-    generate the logs.  This paper describes the creation of a parser,
-    documenting the parsing algorithm, explaining the difficulties
-    encountered and the solutions developed, with reference to an
-    implementation which stores data gleaned from the logs in an SQL
+    it \textit{is\/} possible to achieve a high degree of accuracy in
+    understanding the logs, and thus accuracy in reconstructing the actions
+    taken by Postfix to generate the logs.  This paper describes the
+    creation of a parser, documenting the parsing algorithm, explaining the
+    difficulties encountered and the solutions developed, with reference to
+    an implementation which stores data gleaned from the logs in an SQL
     database.  The gathered data can then be used to optimise current
@@
 Most mail server administrators will have performed some basic processing
-of Postfix logs at one time or another, whether it was to debug a problem,
-explain to a user why their mail is being rejected, or check whether their
-new anti-spam measures are working.  The more adventurous will have
-generated statistics to show how many hits each of their anti-spam measures
-has gotten in the last week, and possibly even generated some graphs to
-clearly illustrate the point to management or users.\footnote{This was the
-author's first real foray into processing Postfix logs.}  Very few will
-have performed in-depth parsing and analysis of their logs, where the
-parsing must correlate the log lines per-connection or per-queueid rather
-than processing lines independently.  One of the barriers to this type of
-processing is the unstructured nature of Postfix logs, where each log line
-was added on an ad hoc basis as a requirement was discovered or new
-functionality was added.\footnote{A history of all changes made to Postfix
-is distributed with the source code, available from
-\url{http://www.postfix.org/}} Further complication arises because the set
-of rejection messages is not fixed: new messages can be added by the
-administrator with custom checks; every \RBL{} returns a different
-explanatory message; policy servers may log different messages depending on
-the characteristics of the connection; there are many ways in which the log
-lines may differ between servers, even within the same organisation:
-servers may be configured differently, or running different version of
-Postfix.  This paper documents the difficult process of parsing Postfix
-logs, presenting a program which parses logs and places the resulting data
-into a database for later use.  The gathered data can then be used to
-optimise current anti-spam measures, provide a baseline to test new
-anti-spam measures against, or to produce statistics showing how effective
-those measures are.  There are numerous other uses for such data: improving
-server performance by identifying troublesome destinations and
-reconfiguring appropriately or dedicating transports to those destinations;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
+of the logs produced by their mail server at one time or another, whether
+it was to debug a problem, explain to a user why their mail is being
+rejected, or check whether new anti-spam measures are working.  The more
+adventurous will have generated statistics to show how many hits each of
+their anti-spam measures has gotten in the last week, and possibly even
+generated some graphs to clearly illustrate the point to management or
+users.\footnote{This was the author's first real foray into processing
+Postfix logs.}  Very few will have performed in-depth parsing and analysis
+of their logs, where the parsing must correlate the log lines
+per-connection or per-queueid rather than processing lines independently.
+One of the barriers to this type of processing is the unstructured nature
+of Postfix logs, where each log line was added on an ad hoc basis as a
+requirement was discovered or new functionality was added.\footnote{A
+history of all changes made to Postfix is distributed with the source code,
+available from \url{http://www.postfix.org/}} Further complication arises
+because the set of rejection messages is not fixed: new messages can be
+added by the administrator with custom checks; every \RBL{} returns a
+different explanatory message; policy servers may log different messages
+depending on the characteristics of the connection; there are many ways in
+which the log lines may differ between servers, even within the same
+organisation --- servers may be configured differently, or running
+different version of Postfix.  This paper documents the difficult process
+of parsing Postfix logs, presenting a program which parses logs and places
+the resulting data into a database for later use.  The gathered data can
+then be used to optimise current anti-spam measures, provide a baseline to
+test new anti-spam measures against, or to produce statistics showing how
+effective those measures are.  There are numerous other uses for such data:
+improving server performance by identifying troublesome destinations and
+reconfiguring appropriately; identifying regular high volume uses (e.g.\
+customer newsletters) and restricting those uses to off-peak times;
+detecting virus outbreaks which propagate via email; as a base for billing
+customers on a shared server.  Preserving the raw data enables users to
+develop a multitude of uses far beyond those conceived of by the author.
 
@@
 parsing and the results gleaned from parsing.  The database schema used is
-described in \sectionref{database schema}, explaining in detail the tables
+described in section~\refwithpage{database schema}, explaining in detail the tables
 used for storing the data gleaned from the logs and the table that stores
@@
 graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate a
-\regex{} from related log lines for new rules.
+description of using the tools provided with the parser to generate new
+rules (specifically the \regex{} in each rule) from unparsed log lines.
 
 Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications encountered which shaped
-the full algorithm.  A flow chart and a discussion of the emergent
+a naive parsing algorithm and the complications initially encountered which
+shaped the full algorithm.  A flow chart and a discussion of the emergent
 behaviour exhibited by the algorithm accompanies a comprehensive
@@
 during execution of the algorithm are described, followed by the process of
-adding a new action.  The section concludes with an in-depth description of
-the further complications and their solutions which complete the algorithm.
+adding a new action.  The section concludes with an in-depth description
+of the further complications discovered and their solutions which complete
+the algorithm.
 
@@
 false negatives and a discussion of the difficulties in identifying false
-positives.  A random sampling of log lines are parsed, and the correctness
+positives.  A random sampling of log lines was parsed, and the correctness
 of the results manually verified, as part of determining the coverage of
@@
 
-Section~\ref{conclusion} contains the paper's conclusion.
+Section~\ref{conclusion} contains the paper's conclusion, describing the
+results of the research, design and implementation of the parser.
 
@@
 Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency}, rule efficiency, and explains
+discussed in \sectionref{rule efficiency} (rule efficiency), and explains
 the irregularities observed in these graphs.
@@
 underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to
-\ifacronymfirstuse{SMTP}{the}{} \SMTP{}\@; and a longer introduction to
-Postfix, concentrating on the topics most relevant to this paper, namely
-Postfix anti-spam restrictions and policy servers.  The assumptions made in
-designing and implementing the parser are explained, as are the conventions
-used in this paper.  Other projects which attempt to parse Postfix logs are
-summarised (full details are available in
+a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
+introduction to Postfix, concentrating on the topics most relevant to this
+paper, namely Postfix anti-spam restrictions and policy servers.  The
+assumptions made in designing and implementing the parser are explained, as
+are the conventions used in this paper.  Other projects which attempt to
+parse Postfix logs are summarised (full details are available in
 appendix~\refwithpage{other-parsers}), finishing with a review of
@@
 A short example of the optimisation possible using data from the database
-is which Postfix restrictions reject the highest number of mails:
+is determining which Postfix restrictions reject the highest number of
+mails:
 
@@
 
-If the database supports sub-selects percentages can be
-obtained:
+If the database supports sub-selects percentages can be obtained:
+\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
+the database containing the extracted data does not support this syntax,
+then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
+results will be the same, just slightly less visually pleasing.}
 
@@
 
-\SQL{} note: $||$ is the concatenation operator in SQLite3; if the database
-containing the extracted data does not support this syntax, then simply
-remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the results will be
-the same, just slightly less visually pleasing.
-
 Another example is determining which restrictions are not effective: this
-example shows which restrictions had fewer than 100 hits on the last log
-file parsed.
+shows which restrictions had fewer than 100 hits on the last log file
+parsed.
 
@@
     \item The majority of programming languages have existing code
-        available allowing access to databases, so the majority of
-        languages will be able to access the results obtained by running
-        the parser.  If this parser was written to be utilised as shared
-        code there would be little or no difficulty in using it when
-        writing in the programing language it is implemented in, but every
-        other programing language wishing to utilise it would require an
+        available providing access to databases, so the majority of
+        languages will be able to access the results obtained by the
+        parser.  If this parser was written to be utilised as shared code
+        there would be little or no difficulty in using it when writing in
+        the programing language it is implemented in, but every other
+        programing language wishing to utilise it would require an
         interface layer to be written.  This requirement would drastically
@@
         the user without requiring large amounts of programming.  All
-        databases provide a program, of varying complexity and
+        databases provide a program, of variable complexity and
         sophistication, which can be used for ad hoc queries with minimal
@@
             \item Other tables can be added to the database, e.g.\ to cache
-                historical data.
+                historical, summary or computed data.
 
@@
                 part of dealing with support calls, and the administrators
-                have full access to the data.
+                can have full access to the data.
 
@@
         familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn.  Although every
-        vendor implements a different dialect of \SQL{}, the basics are the
-        same everywhere (analogous to the overall similarities and minor
-        differences between British English, American English and
-        Australian English).  A good introduction to \SQL{} can be found
-        at~\cite{sql-for-web-nerds}, others
-        are~\cite{w3schools-sql-tutorial, sqlcourse.com}.  Depending on the
-        database in use there may be tools available which reduce or remove
-        the requirement to know \SQL{}; for SQLite (the default database
-        used by the implementation) there are several
-        available~\cite{sqlite-guis}.
+        readily available resources from which to learn (a good
+        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
+        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
+        every vendor implements a different dialect of \SQL{}, the basics
+        are the same everywhere (analogous to the overall similarities and
+        minor differences between Irish English, British English, American
+        English and Australian English).  Depending on the database in use
+        there may be tools available which reduce or remove the requirement
+        to know \SQL{}; for SQLite (the default database used by the
+        implementation) there are several available~\cite{sqlite-guis}.
 
@@
 and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.
+by the server and carries on regardless.  There are many spam and virus
+sending programs which are this deficient --- unfortunately others
+(particularly newer programs) were written by competent programmers, or
+utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
+hosts, Microsoft Outlook on Windows hosts).
 
@@
 used when determining which rules should be used to parse each log line
-(see \sectionref{rule characteristics} for details.)
+(see \sectionref{rule characteristics} for details.)  This design has
+positive security implications also: those components which interact with
+other hosts are not privileged, and those components which are privileged
+do not interact with other hosts.
 
@@
 Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview.
+smtpd_per_user_control, policy-servers}, the following is a brief overview
+only.
 
@@
 first RCPT TO command is received, because some clients don't deal with
-earlier rejections properly; a side benefit of this delay is that Postfix
-has more information available when logging the rejection.\footnote{The
+earlier rejections properly; a benefit of this delay is that Postfix has
+more information available when logging the rejection.\footnote{The
 \texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
@@
 
-check\_client\_access checks whether the \IP{} address of the connected
-client matches the left hand side of each line in the file and returns the
-right hand side of the first matching line.  Other restrictions determine
-their result by consulting external sources, e.g.\
+The restriction \texttt{check\_client\_access} checks whether the \IP{}
+address of the connected client is found in the file and returned the
+associated action if found; the method of searching the file is dependant
+on the type of the file (\texttt{cidr} in the example) --- see
+\cite{postfix-lookup-tables} for more details.  Other restrictions
+determine their result by consulting external sources, e.g.\
 \texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the listed \RBL{}, rejecting the
-command if the client is listed.
+address of the client is present in the \RBL{} \texttt{rbl.example.com},
+rejecting the command if the client is listed.
 
@@
 \textit{dunno\/} means to stop evaluating the current restriction and
-continue with the next restriction in the list, allowing more specific
-cases to be used as exceptions to more general cases.  When the result is
-the name of another restriction Postfix will evaluate the new restriction,
-allowing restrictions to be chosen based on the client \IP{} address, HELO
-hostname, sender address, recipient address, etc.  The administrator can
-define new restrictions as a list of existing restrictions, allowing
-arbitrarily long and complex sequences of lookups and restrictions.
+continue processing the remainder of the restriction list, allowing more
+specific cases to be used as exceptions to more general cases.  When the
+result is the name of another restriction Postfix will evaluate the new
+restriction, allowing restrictions to be chosen based on the client \IP{}
+address, HELO hostname, sender address, recipient address, etc.  The
+administrator can define new restrictions as a list of existing
+restrictions, allowing arbitrarily long and complex sequences of lookups,
+restrictions and exceptions.
 
@@
 earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  E.g.\ some users can be
+to implement whatever logic is required.  For example some users can be
 restricted to sending mail on the third Tuesday after pay day only --- this
@@
     \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction}
-        records specify which mail servers are allowed to send mail
-        claiming to be from a particular domain.  The intention is to
-        reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and joe
-        jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
+        \SPF{}\label{spf introduction} records specify which mail servers
+        are allowed to send mail claiming to be from a particular domain.
+        The intention is to reduce spam from faked sender addresses,
+        backscatter~\cite{postfix-backscatter} and
+        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
         resistance to the proposal because it breaks or vastly complicates
@@
     \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender, recipient, remote \IP{}
-        address) is unknown; on second and subsequent delivery attempts
-        from that triple the mail will be accepted.  The assumption is that
-        maintaining a list of failed addresses and retrying after a
-        temporary failure is uneconomical for a spammer, but that a
-        legitimate mail server must retry.  Sadly spammers are using
-        increasingly complex and well written programs to distribute spam,
-        frequently using an \ISP{} provided \SMTP{} server from a
-        compromised machine on the \ISP{}'s network.  Greylisting will
-        slowly become less useful, but it does block a large percentage of
-        spam mail at the moment; the most effective restrictions over the
-        \numberOFlogFILES{} log files used in testing the parser are shown
-        in table~\refwithpage{Summary of rejections}.  Greylisting is
-        obviously worth using, at least at the moment, particularly when
-        you factor in Greylisting's position as the final restriction which
-        a mail must overcome: Greylisting only takes effect for mails which
-        have passed every other restriction.
+        rejects mail when the triple of (sender address, recipient address,
+        remote \IP{} address) is unknown; on second and subsequent delivery
+        attempts from that triple the mail will be accepted.  The
+        assumption is that maintaining a list of failed addresses and
+        retrying after a temporary failure is uneconomical for a spam
+        sender, but that a legitimate mail server must retry.  Sadly spam
+        senders are using increasingly complex and well written programs to
+        distribute spam, frequently using an \ISP{} provided \SMTP{} server
+        from a compromised machine on the \ISP{}'s network.  Greylisting
+        will slowly become less useful, but it does block a large
+        percentage of spam mail at the moment; the most effective
+        restrictions over the \numberOFlogFILES{} log files used in testing
+        the parser are shown in table~\refwithpage{Summary of rejections}.
+        Greylisting is obviously worth using, at least at the moment,
+        particularly when you factor in Greylisting's position as the final
+        restriction which a mail must overcome:\footnote{Greylisting is the
+        final restriction a mail must overcome in the configuration used on
+        the mail server the logs were obtained from; an administrator is
+        free to use Greylisting at whichever position in the restriction
+        list they feel is most appropriate for their mail system.}
+        Greylisting only takes effect for mails which have passed every
+        other restriction.
 
@@
         against the sending system --- if the eventual score is too high
-        the mail is rejected.
+        the mail is rejected.  Postfix's restrictions are binary, and the
+        administrator must manually whitelist clients if they are to bypass
+        a restriction; using a threshold which requires hitting several
+        restrictions frees the administrator from whitelisting clients
+        which fall foul of one restriction only.
 
@@
         these problems will arise, though it is of course possible,
-        particularly when undergoing a deluge of spam.
+        particularly when suffering a deluge of spam or a mail loop.
 
     \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.
+        accurately reconstruct the actions it has taken.  There are a
+        number of heuristics used when parsing; see
+        \sectionref{identifying-bounce-notifications},
+        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
+        logging after cleanup} for details.
 
@@
 code would require a significant investment of time; the source code for
-Postfix 2.4.6 is 16MB.} there are advantages to treating Postfix as a black
+Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
 box while developing the parser:
@@
         orderings in which log lines are written to the log file, as
-        process scheduling interferes with it.
+        process scheduling interferes with logging and other processing.
 
@@
 make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  The main design decision which enables
-this is the separation of rules from the parsing algorithm, and secondly
-the division of the algorithm into discrete actions.
-
-\label{why separate rules and algorithm}
+properly parse their log lines.  To enable this the parser is divided into
+three parts:
+
+\begin{description}
+
+    \item [rules] These match individual log lines and determine which
+        actions will be executed.  Rules provide an easily extensible
+        method of associating log lines and actions.  Rules are described
+        in detail in \sectionref{rules}.
+
+    \item [actions] Are invoked to deal with a log line once the line has
+        been identified by the rules: modifying data structures, handing
+        complications, and saving data to the database.  The actions
+        perform the work of reconstructing the journey a mail takes through
+        Postfix.  Full details of actions can be found in
+        \sectionref{parsing-algorithm}, particularly
+        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
+
+    \item [framework] Is responsible for loading rules, managing data
+        structures so actions can retrieve them, finding the rule which
+        matches each log line, invoking the correct action, etc.  The
+        framework provides the structure which actions and rules plug into.
+        The framework is not described in detail in this document because
+        it is not particularly interesting.
+
+\end{description}
+
+It may help to think of the rules and actions as components which plug into
+the framework.
+
+\label{why separate rules, actions and framework?}
 
 Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the algorithm
+to be written and tested without requiring modifications to the parser
 source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both
-algorithm and rules.  Decoupling also creates a clear separation of
-functionality: rules handle low level details of identifying log lines and
-extracting data from a line, whereas the algorithm handles the higher level
-details of following the path a mail takes through Postfix, assembling the
-required data before storing it, dealing with complications arising, etc.
-
-Separating the rules from the algorithm makes it possible to parse new log
-lines without modifying the core parsing algorithm.  Although this may seem
-like a trivial point, is it substantially more difficult to understand a
-program's entire parsing algorithm, identify the correct location to
-change, and make the appropriate changes without adversely affecting
-parsing, particularly as there may be edge cases which are not immediately
-obvious.\footnote{See \sectionref{yet-more-aborted-delivery-attempts} for a
-complication which occurs only four times in \numberOFlogFILES{} log files
-tested.} Requiring changes to the parsing algorithm also complicates
-upgrades, as the changes must be preserved during the upgrade, and may
-clash with changes made by the developer.  \parsername{} allows the user to
-add new rules to the database without changing the parsing algorithm,
-unless the new log lines to be parsed require functionality not already
-provided by the algorithm.  If the new log lines do require new
-functionality, new actions can be added to the parser without modifying
-existing actions or other parts of the algorithm; only in the rare case
-that the new actions require support from other sections of the code will
-more extensive changes be required.
+users who need to parse new log lines), and greatly simplifies both actions
+and rules.  Decoupling also creates a clear separation of functionality:
+rules handle low level details of identifying log lines and extracting data
+from a line, whereas the actions handle the higher level details of
+following the path a mail takes through Postfix, assembling the required
+data before storing it, dealing with complications arising, etc.
+
+Decoupling the actions from the framework simplifies both framework and
+actions: the framework provides services to the actions, and doesn't need
+to deal with the complications which arise, or the task of reconstructing a
+mail's journey through Postfix; actions benefit from having services
+provided by the framework, freeing them to concentrate on the task of
+accurately reconstructing each mail's journey through Postfix and dealing
+with the complications described in \sectionref{additional complications}.
+
+Separating the rules from the actions and framework makes it possible to
+parse new log lines without modifying the core parsing algorithm.  Although
+this may seem like a trivial point, is it substantially more difficult to
+understand a program's entire parsing algorithm, identify the correct
+location to change and make the appropriate changes, versus adding a new
+rule with the action to invoke, a \regex{} to match the lines, and the
+specification of the data to extract.  Bear in mind that the changes must
+be made without adversely affecting existing parsing, particularly as there
+may be edge cases which are not immediately obvious.\footnote{See
+\sectionref{yet-more-aborted-delivery-attempts} for a complication which
+occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
+changes to the parser's code also complicates upgrades, as the changes
+must be preserved during the upgrade, and may clash with changes made by
+the developer.\footnote{See \sectionref{adding new actions} for how to add
+new actions}.  \parsername{} allows the user to add new rules to the
+database without changing the parsing algorithm, unless the new log lines
+to be parsed require functionality not already provided by the algorithm.
+If the new log lines do require new functionality, new actions can be added
+to the parser without modifying existing actions or other parts of the
+algorithm; only in the rare case that the new actions require support from
+other sections of the code will more extensive changes be required.
 
@@
 
+    \item Assumptions made when designing and developing the parser.
+
+    \item A description of the parser's novel three section design.
+
+    \item Conventions used in this document.
+
+    \item A brief comparison of this parser versus other Postfix parsers.
+
+    \item A review of the literature previously published in this area.
+
 \end{itemize}
-
-The assumptions which must be satisfied for the parser to work correctly
-were discussed; the conventions used in the paper were listed; a brief
-comparison of this project against other Postfix parsers was provided,
-followed by a review of the previously published literature.
 
@@
 aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here.
+described in detail in \sectionref{rules} but the fields are covered here
+in \sectionref{rule attributes}.
 
@@
 row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in the forthcoming
+rejection; the individual fields will be described in
 \sectionref{connections table}.  The results table has one or more rows per
-row in the connections table, depending on the specifics of the
-mail/connection; the fields will be covered in detail in
+row in the connections table; the fields will be covered in detail in
 \sectionref{results table}.
@@
 
-Rules are discussed in detain in \sectionref{rules}, but the rules table is
-covered here.  Rules are created by the user, not the parser, and will not
-be modified by the parser (except for the hits and hits\_total fields).
-Rules parse the individual lines, telling the parser which fields to
-extract and what action to take for that line.
+Rules are discussed in detail in \sectionref{rules}, but the structure of
+the rules table is covered here.  Rules are created by the user, not the
+parser, and will not be modified by the parser (except for the hits and
+hits\_total fields).  Rules parse the individual lines, extracting data to
+be saved in the connections and results tables, and specifying action to
+take for that line.
 
@@
 \begin{description}
+
+    \item [id] A unique identifier for each rule which other tables can use
+        to refer to a specific rule.
 
@@
     \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have a result of
-        \texttt{rejected}, other rules will have an empty string.
+        rejected.  Only applicable to rules which have an action of
+        \texttt{REJECTION}, other rules should have an empty string.
 
     \item [postfix\_action] This is the action Postfix must have taken to
-        generate this line, with two exceptions:
+        generate this line.  This field is mostly ignored, but two values
+        (IGNORED and INFO) have special meaning, as described below in the
+        list of typical values.\label{postfix_action}
 
         \begin{description}
+
+            \item [ACCEPTED] Postfix has accepted a mail, and will
+                subsequently attempt to deliver it.
+
+            \item [BOUNCED] The mail has bounced, due to a mail loop,
+                delivery failure, or five day timeout.
+
+            \item [DELETED] The mail was deleted from the queue by an
+                administrator.
+
+            \item [DISCARDED] Postfix discarded the mail it was in the
+                process of accepting, because it was either larger than the
+                limit set by the administrator, or the client timed out or
+                disconnected.
+
+            \item [EXPIRED] The mail has been in the queue for five
+                days\footnote{As with much of Postfix's behaviour, this is
+                the default value but can be changed by the administrator
+                if they choose.} without successful delivery.  A bounce
+                mail will be generated and sent to the sender address.
 
@@
                 itself, but which also provides no useful data.
+
+            \item [POSTFIX\_RELOAD] The administrator has caused Postfix to
+                start or stop, and all existing \daemon{smtpd} processes
+                will be terminated.
+
+            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
+                see~\cite{postfix-cleanup} for details of the processing
+                performed by \daemon{cleanup}.
+
+            \item [REJECTED] Postfix rejected a command from the remote
+                client, causing at least one recipient to be rejected.
+
+            \item [SENT] Postfix has successfully sent a mail.
 
@@
 
-        The following keywords are expanded (full explanations can be found
-        in the source code):
-
-        \_\_SENDER\_\_, \_\_RECIPIENT\_\_, \_\_MESSAGE\_ID\_\_,
-        \_\_HELO\_\_, \newline \_\_EMAIL\_\_, \_\_HOSTNAME\_\_, \_\_IP\_\_,
-        \_\_IPv4\_\_, \_\_IPv6\_\_, \newline \_\_SMTP\_CODE\_\_,
-        \_\_RESTRICTION\_START\_\_, \_\_QUEUEID\_\_, \newline
-        \_\_COMMAND\_\_, \_\_SHORT\_CMD\_\_, \_\_DELAYS\_\_, \_\_DELAY\_\_,
-        \_\_DSN\_\_ and \_\_CONN\_USE\_\_.
-
-        Additional fields are captured by \_\_RESTRICTION\_START\_\_, so
-        rules using it will start the fields in result\_cols,
-        connection\_cols, etc.\ at 5.
-
         For efficiency the keywords are expanded and every rule's \regex{}
@@
         in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}.
+        discussed in \sectionref{rule efficiency}, with the impact of
+        compiling and caching \regexes{} covered in \sectionref{Caching
+        each regex}.
 
     \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is: \newline
-        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline i.e.\
-        semi-colon or comma separated assignment statements, with the
-        variable name on the left and the matching field from the \regex{}
-        on the right hand side.  The list of acceptable variable names is:
+        log line will be extracted from the log line.  The format is:
+        \newline \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
+        \newline i.e.\ semi-colon or comma separated assignment statements,
+        with the variable name on the left and the matching capture from
+        the \regex{} on the right hand side.  The list of acceptable
+        variable names is:
 
@@
         \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} pid\_regex, data} and \texttt{child.}
+        \newline \tab{} size,} and \texttt{data}
+
+        Additionally \texttt{pid\_regex} and \texttt{child} are used by the
+        \texttt{TRACK, BOUNCE} and \texttt{SMTPD\_DIED} actions.
 
@@
 
-    \item [action] The action the algorithm will take; a full list can be
-        found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching field from the \regex{} which
+    \item [action] The action that will be invoked when this rule matches a
+        line; a full list can be found in \sectionref{actions-in-detail}.
+
+    \item [queueid] Specifies the matching capture from the \regex{} which
         gives the queueid, or zero if the log line doesn't contain a
@@
 
-    \item [hits] is an efficiency measure.  This counter is maintained for
-        every rule and incremented each time the rule successfully matches.
-        At the start of each run the program sorts the rules in descending
-        order of hits, and at the end of the run updates every rule's hits.
-        Assuming that the distribution of log lines is reasonably
-        consistent between log files, rules matching more commonly
-        occurring log lines will be tried before rules matching less
-        commonly occurring log lines, lowering the program's execution
-        time.  Rule ordering for efficiency is discussed in
-        \sectionref{rule ordering for efficiency}.
+    \item [hits] This counter is maintained for every rule and incremented
+        each time the rule successfully matches.  At the start of each run
+        the program sorts the rules in descending order of hits, and at the
+        end of the run updates every rule's hits.  Assuming that the
+        distribution of log lines is reasonably consistent between log
+        files, rules matching more commonly occurring log lines will be
+        tried before rules matching less commonly occurring log lines,
+        lowering the program's execution time.  Rule ordering for
+        efficiency is discussed in \sectionref{rule ordering for
+        efficiency}.
 
@@
 
-    \item [helo] The hostname used in the HELO command.  The HELO
+    \item [helo] The hostname used in the HELO command.  The HELO hostname
         occasionally changes during a connection, presumably because spam
@@
         logged also, but when there are multiple recipients none are
-        logged.
+        logged.  It is also possible to warn for every recipient,
+        preventing the complication in \sectionref{Mail deleted before
+        delivery is attempted} entirely.
 
@@
 
-Every log line corresponding to Postfix performing an action other than
-INFO or IGNORED will have an entry in the results table, e.g.\ rejecting an
-\SMTP{} command, delivering a mail, or bouncing a mail.  Any log line where
-the Postfix action is INFO is not interesting in and of itself, but
-provides additional information which will be saved with other results;
-IGNORED log lines don't even provide useful information.  Each row is
-associated with a single connection, though there may be many results per
-connection.
+Every log line where the associated rule has a \texttt{postfix\_action}
+will have an entry in the results table, e.g.\ rejecting an \SMTP{}
+command, delivering a mail, or bouncing a mail (see
+\sectionref{postfix_action} for typical values of
+\texttt{postfix\_action}).  Each row is associated with a single
+connection, though there may be many results per connection.
 
@@
     \item [connection\_id] A reference to the row in the connections table
-        this result's table row belongs to.
+        this result is associated with.
 
@@
         matched the log line and created this result.
+
+    \item [id] A unique identifier for this result.
 
@@
         field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection or any other result.
+        real rejection, or 0 for a real rejection.  This field should be
+        ignored if the result is not a rejection.
 
@@
         general an \SMTP{} code is only present for a rejection or final
-        delivery; results missing an \SMTP{} code will duplicate the
-        \SMTP{} code of other results in the connection.  Some final
+        delivery; results initially missing an \SMTP{} code will duplicate
+        the \SMTP{} code of other results in the connection.  Some final
         delivery log lines don't contain an \SMTP{} code: in those cases
@@
 
-    \item [recipient] The recipient address.
+    \item [recipient] The recipient address; there may be multiple
+        recipient addresses per mail, but only one per result.
 
@@
 
-The table containing the rules used by the parser, and both tables
+The table containing the rules used by the parser and both tables
 containing the data extracted from the Postfix logs were described, with
@@
 \subsubsection{Caching each regex}
+
+\label{Caching each regex}
 
@@
         attempt (administrator deleting unwanted mail, typically due to a
-        mail loop or joe-job).  Notice that \daemon{postsuper} is required,
+        mail loop or joe~job).  Notice that \daemon{postsuper} is required,
         not optional, for this combination.
@@
 some of the mail (because it's the result of a mail loop, mail bomb, or
-joe-job) before Postfix has had a chance to try to deliver it.  In this
+joe~job) before Postfix has had a chance to try to deliver it.  In this
 case the recipient address will not have been logged, so a dummy recipient
@@
 parse new log lines without modifying the core parsing algorithm; see
-\sectionref{why separate rules and algorithm} for a discussion of why the
-separation of rules from the algorithm is beneficial.
+\sectionref{why separate rules, actions and framework?} for a discussion of
+why the separation of rules from the algorithm is beneficial.
 
@@
 
+\label{Caching regexes}
+
 The following graphs show the impact that not caching compiled \regexes{}
@@
 \glossary{name={Backscatter},description={
-    When a spammer or worm sends mail with forged sender addresses,
+    When a spam sender or worm sends mail with forged sender addresses,
     innocent sites are flooded with undeliverable mail notifications; this
@@
 
-\glossary{name={Joe job},description={
-    A joe job is when spam mail is sent with a faked sender address with
+\glossary{name={Joe~job},description={
+    A joe~job is when spam mail is sent with a faked sender address with
     the intention of sullying the good name of the owner of the address.
-    Joe jobs are a cause of backscatter, though by no means the only cause.
+    joe~jobs are a cause of backscatter, though by no means the only cause.
 }}

### 2008-03-04 doc/logparser.tex
Further corrections after proofreading.

@@
 \end{abstract}
+
+XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
 
@@
 over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 1000 users.  Coverage is described both in terms of
+handling mail for over 700 users.  Coverage is described both in terms of
 the fraction of lines parsed and the fraction of mails and connections
@@
 This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, continuing with detecting and dealing
-with overlapping rules, and the problems they can cause.  An example rule
-and a line it would match are provided, plus a description of how the
-fields in the rule are used when matching and performing the action (the
-details of the table containing the rules have already been described in
-\sectionref{rule attributes},).  This section continues with a discussion
-of rule efficiency concerns, referring to the graphs in
-\sectionref{graphs}, and finishes with an explanation of the algorithm used
-to generate a new \regex{} from unparsed lines.
+starting with rule characteristics, followed by the problems caused by
+overlapping rules, and techniques to detect and deal with such rules.  An
+example rule and a log line it would match are provided, plus a description
+of how the fields in the rule\footnote{The details of the table containing
+the rules have already been described in \sectionref{rule attributes}.} are
+used when matching a line and subsequently performing the requested action.
+The section continues with a discussion of rule efficiency concerns,
+referring to the graphs in \sectionref{graphs}, and finishes with an
+explanation of the algorithm used to generate new \regexes{} from unparsed
+lines.
 
@@
         parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules must
+        there are also rules specific to that program, and those rules have
         already have been tried and failed on the current line.}  Any given
@@
     \item The first matching rule wins: no further rules are tried against
-        that line, but there is a facility for specifying the order of
-        rules so that more specific rules can be tried first.
+        that line, but there is a facility for ordering the rules so that
+        more specific rules can be tried first.
 
@@
 The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{}, and determine if any overlap.  The
+possible to parse each rule's \regex{} and determine if any overlap.  The
 author has not attempted to do this: such a project by itself would
@@
 overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line is unspecified, and
-the first matching rule wins.  Overlapping rules are frequently a
-requirement, allowing a more specific rule to match some lines and a more
-general rule to match the majority, e.g.\ separating \SMTP{} delivery to a
+the order in which rules are tried against each line may change between log
+files, and the first matching rule wins.  Overlapping rules are frequently
+a requirement, allowing a more specific rule to match some lines and a more
+general rule to match the majority, e.g.\ separating \SMTP{} delivery to
 specific sites from \SMTP{} delivery to the rest of the world.  The
@@
 
-Detecting overlapping rules is difficult, but the following may be helpful:
+Detecting overlapping rules is difficult, but the following approaches may
+be helpful:
 
@@
         details of sorting the rules.}  Parse a number of log files using
-        normal sorting, dump a textual representation of the connections
-        and results tables, and delete everything from those tables.
-        Repeat with shuffled and reversed sorting.  If there are no
-        overlapping rules the tables from each run will be identical;
-        differences indicate overlapping rules.  Which rules overlap can be
-        determined by examining the differences in the tables: each result
-        contains the id of the rule which created it, so the rule
-        referenced in the normal table overlaps with the rule referenced in
-        the reversed table.  Unfortunately this method cannot prove the
-        absence of overlapping rules; it can detect overlapping rules, but
-        only if there are log lines in the input files which match more
-        than one rule.
+        normal sorting, then dump a textual representation of the rules,
+        connections and results tables.  Repeat with shuffled and reversed
+        rules, starting with a fresh database.  If there are no overlapping
+        rules the tables from each run will be identical; differences
+        indicate overlapping rules.  Which rules overlap can be determined
+        by examining the differences in the tables: each result contains a
+        reference to the rule which created it, if the references differ
+        between runs the two rules referenced in the differing records
+        overlap.  Unfortunately this method cannot prove the absence of
+        overlapping rules; it can detect overlapping rules, but only if
+        there are log lines in the input files which match more than one
+        rule.
 
@@
 \begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]:
-  550 <foo@example.com>: Sender address rejected:
-  Domain not found; from=<foo@example.com>
-  to=<info@example.net> proto=SMTP
-  helo=<smtp.example.com>
+NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
+  <foo@example.com>: Sender address rejected: Domain not found;
+  from=<foo@example.com> to=<info@example.net>
+  proto=SMTP helo=<smtp.example.com>
 \end{verbatim}
@@
 
-The various fields are used as follows;
+Additional data will be captured automatically when the \regex{} contains 
+\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
+connection\_cols start at 5.  The various fields are used as follows;
 
@@
     \item [program and regex:] If the program in the rule equals the
-        program which logged the line the \regex{} will be tried against
-        the line; if the match is successful the action will be executed,
-        if not the next rule will be tried.  If the program-specific rules
-        don't match the log line, the generic rules will be tried in the
-        same way.
+        program which logged the line a match using the \regex{} will be
+        attempted against the line; if the match is successful the action
+        will be executed, if not the next rule will be tried.  If the
+        program-specific rules don't match the log line, the parser will
+        fall back to generic rules; if those rules are unsuccessful a
+        warning will be issued.
 
@@
 
-    \item [queueid:] The index of the field in the \regex{} which captured
-        the queueid, or zero if the line does not contain a queueid.  This
-        allows the correct mail can be found by queueid and actions
-        performed on it.
+    \item [queueid:] The index of the capture in the \regex{} which
+        supplies the queueid, or zero if the line does not contain a
+        queueid.  This allows the correct mail can be found by queueid and
+        actions performed on it.
 
@@
         hits is set to the number of successful matches at the end of the
-        parsing run, and hits\_total has hits added to it, but is otherwise
-        unused by the algorithm.
+        parsing run; hits\_total is the sum of hits over every parsing run,
+        but is otherwise unused by the algorithm.
 
@@
 
-Additional fields are captured by \_\_RESTRICTION\_START\_\_, hence the
-fields in result\_cols and connection\_cols start at 5 in the example.
-
 \subsection{Rule efficiency}
@@
 Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (for a small mail
-server --- large scale mail servers would have much larger log files on a
-daily basis).  When generating the data for the graphs included in
-appendix~\refwithpage{graphs}, \numberOFlogFILES{} log files (totaling
-10.08 GB, 60.72 million lines) were each parsed 10 times, the first run
-discarded, and the remaining 9 runs averaged.  The first run is discarded
-for two reasons:
+deal with 75 MB log files containing 300,000 log lines (generated daily on
+a mail server handling mail for approximately 700 users --- large scale
+mail servers would have much larger log files on a daily basis).  When
+generating the data for the graphs included in \sectionref{graphs},
+\numberOFlogFILES{} log files (totaling 10.08 GB, 60.72 million lines) were
+each parsed 10 times, the first run discarded, and the execution time for
+the remaining 9 runs averaged.  The first run is discarded for two reasons:
 
@@
 dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database.
+the speed of the parser rather than the speed of the database and the disks
+the database is stored on.
 
@@
 
-An important property of a parser is how execution time scales with input
-size: does it scale linearly, polynomially, or exponentially?
+An important property of a parser is how execution time scales relative to
+input size: does it scale linearly, polynomially, or exponentially?
 Graph~\refwithpage{execution time vs file size vs number of lines graph}
 shows the execution time in seconds, file size in MB and tens of thousands
-of lines per log file per log file.  All three lines run roughly in
-parallel, giving a visual impression that the algorithm scales linearly
-with input size.  This impression is borne out by
-graph~\refwithpage{execution time vs file size vs number lines factor},
-which plots the ratio of file size vs execution time and number of lines vs
-execution time.  As the reader can see, the ratios are quite tightly
-banded, showing that the algorithm scales linearly: the much larger log
-files between points 60 and 70 on the X axis in
-graph~\refwithpage{execution time vs file size vs number of lines graph}
-don't cause any abnormality in the corresponding points in
-graph~\refwithpage{execution time vs file size vs number lines factor}.
+of lines per log file.  All three lines run roughly in parallel, giving a
+visual impression that the algorithm scales linearly with input size.  This
+impression is borne out by graph~\refwithpage{execution time vs file size
+vs number lines factor}, which plots the ratio of file size vs execution
+time and ratio of number of lines vs execution time (higher is
+better).\footnote{Table~\refwithpage{execution time vs file size vs number
+lines factor table} shows the ratios for different breakdowns of the log
+files.}  As the reader can see the ratios are quite tightly banded, showing
+that the algorithm scales linearly: the much larger log files between
+points 60 and 70 on the X axis in graph~\refwithpage{execution time vs file
+size vs number of lines graph} actually cause the ratio to increase, rather
+than decrease.  The strange behaviour where larger log files are parsed
+more efficiently is explained fully in \sectionref{Why are there dips in
+the graphs?}.
 
@@
 
-    \item [reverse] Hypothetically the least optimal order: the most
-        frequently matching rules will be tried last.
+    \item [reverse] Hypothetically the worst order: the most frequently
+        matching rules will be tried last.
 
@@
 and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times (mean and standard deviation are
-shown in table~\refwithpage{Regex caching/discarding with different groups
-of log files}).  Overall this provides a modest but worthwhile
-performance increase.
+percentage increase of execution times, with table~\refwithpage{Execution
+time increase for different rule orderings} showing the mean increases for
+different groupings of logs.  Overall this provides a modest but worthwhile
+performance increase of approximately 9\%, for a small investment in time
+and programming.
 
@@
 frequency with which each word occurs can be used to determine whether the
-word is fixed or variable.  This classification can be used to group
-similar log lines and generate a \regex{} to match each group of lines.
+word is variable or part of the fixed pattern.  This classification can be
+used to group similar log lines and generate a \regex{} to match each group
+of lines.
 
@@
         \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise the existing knowledge to create more
-        accurate \regexes{}.  The new log lines are written to a temporary
-        file, which all subsequent stages use instead of the original input
-        file.
+        this phase is to utilise existing knowledge to create more accurate
+        \regexes{}.  The new log lines are written to a temporary file,
+        which all subsequent stages use instead of the original input file.
 
@@
         of each line, incrementing a counter for each hash.  The counters
-        will later be used to filter out words: if the word's hash does not
-        have a high frequency, the word itself cannot have a high
-        frequency, and there is no need to maintain a counter for it,
-        reducing the number of counters and thus the programme's memory
+        will be used in the next pass to filter out words: if the word's
+        hash does not have a high frequency, the word itself cannot have a
+        high frequency, and there is no need to maintain a counter for it,
+        reducing the number of counters and thus the program's memory
         consumption.
@@
         will alter the position of subsequent words, so a word occurring in
-        different positions in two log lines may indicate similarity, but
-        the algorithm does not attempt to deal with this possibility.}  The
-        algorithm maintains a counter for each \textit{(word, word's
-        position within the line)\/} tuple, incrementing it each time that
-        word occurs in that position.
+        different positions in two log lines \textit{may\/} indicate
+        similarity, but the algorithm does not attempt to deal with this
+        possibility.}  The algorithm maintains a counter for each
+        \textit{(word, word's position within the line)\/} tuple,
+        incrementing it each time that word occurs in that position.
 
         The original algorithm only maintains counters for words whose hash
-        result has a high frequency from the previous phase; this reduces
+        result from the previous phase has a high frequency; this reduces
         the number of counters maintained by the algorithm, reducing the
@@
         is classified as a fixed word, otherwise it is classified as a
-        variable term.  Variable terms are replaced by \texttt{.+}, which
-        means to match zero or more of any character.
+        variable term.  If a variable term appears sufficiently often it
+        will be classified as a fixed term, but that should be noticed by
+        the user when reviewing the new \regexes{}.  Variable terms are
+        replaced by \texttt{.+}, which means to match zero or more of any
+        character.
 
@@
         \regexes{} are collapsed to a single \texttt{.+}; any resulting
-        duplicate \regexes{} are combined, and their counters summed.  If
+        duplicate \regexes{} are combined, and their counters added.  If
         the frequency of a \regex{} is lower than the threshold supplied by
-        the user the \regex{} is discarded.\footnote{This is a second,
-        independent threshold, but once again 1\% of input lines is a good
-        starting point; obviously the threshold depends on the number and
-        type of input lines.}  The new \regexes{} are printed for the user
-        to add to the database, either as new rules or merged into the
-        \regexes{} of existing rules; the counter for each \regex{} is also
-        printed, giving the user an indication of how many of the input
-        lines that \regex{} should match.  Discarding \regexes{} will
-        result in some of the input lines not being matched; this utility
-        should be run again once the unmatched lines have been isolated by
-        running the parser, including the new \regexes{}, with the same
-        input which produced the original set of unparsed lines.
+        the user the \regex{} is discarded.\footnote{This is a second
+        threshold, independent of the threshold used to differentiate
+        between fixed and variable words, but once again 1\% of input lines
+        is a good starting point; obviously the threshold depends on the
+        number and type of input lines.}  The new \regexes{} are printed
+        for the user to add to the database, either as new rules or merged
+        into the \regexes{} of existing rules; the counter for each
+        \regex{} is also printed, giving the user an indication of how many
+        of the input lines that \regex{} should match.  Discarding
+        \regexes{} will result in some of the input lines not being
+        matched; this utility should be run again once the unmatched lines
+        have been isolated by running the parser, including the new
+        \regexes{}, with the same input which produced the original set of
+        unparsed lines.
 
@@
         description of how the fields in the rule are used both in the
-        matching phase and the subsequent action that is executed.
-
-    \item The database table containing the rules is dealt with in
-        \sectionref{rule attributes}, and is not duplicated in this
+        matching phase and the action that is subsequently executed.
+
+    \item The structure of the database table containing the rules is dealt
+        with in \sectionref{rule attributes}, and is not duplicated in this
         section.
@@
 
+\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
+increase for different rule orderings}{Execution time increase for
+different rule orderings}
+
 \subsubsection{Why are there dips in the graphs?}
@@
 The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Rule ordering with and without a mail loop} showing
-the increases for the different rule orderings and combinations of log
-files:
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Rule ordering with
-and without a mail loop}{Rule ordering with and without a mail loop}
+in table~\refwithpage{Execution time increase for different rule orderings}
+showing the increases for the different rule orderings and combinations of
+log files:
 

### 2008-03-04 doc/logparser.tex
More corrections after proofreading.

@@
         The framework is not described in detail in this document because
-        it is not particularly interesting.
+        it is not particularly interesting, but there is some coverage in
+        \sectionref{framework}.
 
@@
 internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, and presents
-the resulting data in a normalised, easy to use representation.  The
+of parsing, the eccentricities and oddities of Postfix logs, presenting the
+resulting data in a normalised, easy to use representation.  The
 algorithm's task is to follow the journey each mail takes through Postfix,
-piecing the data extracted by rules into a coherent whole, saving it in a
+combining the data extracted by rules into a coherent whole, saving it in a
 useful and consistent form, and performing housekeeping duties.
@@
 Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
+rules, actions and framework have been separated in the parser's design.
+In this section algorithm can be taken to mean the combination of framework
+and actions.
 
@@
     \item A flow chart showing common paths a mail can take through the
-        algorithm, and discussing the emergent behaviour observed in the
-        parser.
-
-    \item An explanation of the paths shown in the flow chart.
+        algorithm, a discussion of the emergent behaviour observed in the
+        parser, and an explanation of the paths shown in the flow chart.
+
+    \item A brief description of the framework.
 
@@
     \item Additional complications which have arisen during the development
-        of this parser are documented, including their solutions and where
-        those solutions are implemented in the algorithm
+        of this parser are documented, the solutions to those
+        complications, and where those solutions are implemented in the
+        algorithm
 
@@
 
-A high level view of the algorithm could be expressed as:
+From the viewpoint of an individual mail passing through the parser the
+experience could be summarised as:
 
@@
 
-    \item Mail enters the system via \SMTP{} or local submission; and new
-        data structure is created for it.
+    \item Mail enters the system via \SMTP{} or local submission.
 
@@
 
-Unfortunately that ignores the many complications encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires cloning the current data as soon as a mail is accepted, so that
-subsequent mails won't trample over each other's data.  This must be done
-every time a mail is accepted, as it's impossible to tell in advance which
-connections will accept multiple mails.  It is quite easy to overlook this
-complication because only a small minority of connections accept more than
-one mail. Happily once the mail has been accepted log entries won't be
-correlated by \pid{} for that mail any more (its queueid will be used
-instead), so there isn't any ambiguity about which mail a given log line
-belongs to.\footnote{Unfortunately this statement is not completely
-accurate: see \sectionref{timeouts-during-data-phase} for details.  However
-in general there isn't any ambiguity about which data structure should be
-used for a given log line.}  The original connection will be discarded
-unsaved when the client disconnects if it doesn't have any data worth
-saving, i.e.\ no rejections.  One unsolved difficulty is distinguishing
-between different groups of rejections, e.g.\ when dealing with the
-following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it \textit{usually\/} appears as if the child mail has appeared from thin
-air.  Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-Unfortunately while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition the first log line belonging to the child mail (the log line cited
-above properly belongs to the parent mail) is logged by \daemon{qmgr}, so
-the order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is
-quite paranoid about mail delivery, an excellent characteristic for an
-\MTA{} to possess, so it won't log that the child has been created until it
-is absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation line!  Thus mails flagged as
-coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} shows the paths the data representing a mail/connection can
-take through the parser algorithm.  The flow chart covers the most common
-paths only; there are additional, uncommon paths which are excluded for the
-sake of clarity, but are described in \sectionref{additional
-complications}.
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour described by the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
+The framework's high level overview could be expressed as (indentation
+denoting flow of control):
 
@@
 
+Unfortunately both high level views ignore the many complications
+encountered.
+
+\subsection{Complications encountered}
+
+\label{complications}
+
+These complications were encountered early in the parser's implementation
+and guided its design and development.
+
+\subsubsection{Queueid vs pid}
+
+The mail lacks a queueid until it has been accepted, so log lines must
+first be correlated by the \daemon{smtpd} \pid{}, then transition to being
+correlated by the queueid.  This is relatively minor, but does require:
+
+\begin{itemize}
+
+    \item Two versions of several functions: \texttt{by\_pid} and
+        \texttt{by\_queueid}.
+
+    \item Two state tables to hold the data structure for each connection.
+
+    \item Most importantly: every section of code must know whether it
+        needs to lookup the data structures by \pid{} or queueid.
+
+\end{itemize}
+
+\subsubsection{Connection reuse}
+
+\label{connection reuse}
+
+Multiple independent mails may be delivered during one connection: this
+requires the algorithm to clone the current data as soon as a mail is
+accepted, so that subsequent mails won't trample over each other's data.
+This must be done every time a mail is accepted, as it's impossible to tell
+in advance which connections will accept multiple mails.  Happily once the
+mail has been accepted log entries won't be correlated by \pid{} for that
+mail any more (its queueid will be used instead), so there isn't any
+ambiguity about which mail a given log line belongs
+to.\footnote{Unfortunately this statement is not completely accurate: see
+\sectionref{timeouts-during-data-phase} for details.  However in general
+there isn't any ambiguity about which data structure should be used for a
+given log line.}  The original connection will be discarded unsaved when
+the client disconnects if it doesn't have any data worth saving, i.e.\ no
+rejections.  One unsolved difficulty is distinguishing between different
+groups of rejections, e.g.\ when dealing with the following sequence:
+
+\begin{enumerate}
+
+    \item The client attempts to deliver a mail, but it is rejected.
+
+    \item The client issues the RSET command to reset the session.
+
+    \item The client attempts to deliver another mail, likewise rejected.
+
+\end{enumerate}
+
+There should probably be two different entires in the database resulting
+from the above sequence, but currently there will only be one.
+
+\subsubsection{Re-injected mails}
+
+The most difficult complication initially encountered is that locally
+addressed mails are not always delivered directly to a mailbox: sometimes
+they are addressed to and accepted for a local address but need to be
+delivered to one or more remote addresses due to aliases.  When this occurs
+a child mail will be injected into the Postfix queue, but without the
+explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
+Thus the source is not immediately discernible from the log line in which
+the mail first appears; from a strictly chronological reading of the logs
+it usually appears as if the child mail has appeared from thin air.
+Subsequently the parent mail will log the creation of the child mail:
+
+\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
+delay=0, status=sent (forwarded as 56F5B43FD)}
+
+Unfortunately while all log lines from an individual process appear in
+chronological order, the order in which log lines from different processes
+are interleaved is subject to the vagaries of process scheduling.  In
+addition the first log line belonging to the child mail (the log line cited
+above belongs to the parent mail) is logged by \daemon{qmgr}, so the order
+also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
+paranoid about mail delivery, an excellent characteristic for an \MTA{} to
+possess, so it won't log that the child has been created until it is
+absolutely certain that the mail has been written to disk.}
+
+Because of this the parser cannot complain when it encounters a log line
+from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
+coming from an unknown origin, and subsequently clear the flag if and when
+the origin of the mail becomes clear.  Obviously the parser could omit
+checking of where mails originate from, but the author believes that it is
+better to require an explicit source, as bugs in the parser are more likely
+to be exposed.
+
+Process scheduling can have a still more confusing effect: quite often the
+child mail will be created, delivered and entirely finished with
+\textbf{before} the parent logs the creation line!  Thus mails flagged as
+coming from an unknown origin cannot be entered into the database when
+their final log line is parsed; instead they must be marked as ready for
+entry and subsequently entered by the parent mail once it has been
+identified.
+
+\subsection{Flow chart}
+
+\label{flow-chart}
+
+Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
+image part 2} show the paths the data representing a mail can take through
+the parser algorithm.  The flow chart covers the most common paths only;
+there are additional, uncommon paths which are excluded for the sake of
+clarity; details of the deviations can be found in \sectionref{additional
+complications}.
+
+\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
+1}{flow chart image part 1}
+
+\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
+2}{flow chart image part 2}
+
+\clearpage
+
+\subsubsection{Mail enters the system}
+
+\label{mail-enters-the-system}
+
+Everything starts off with a mail entering the system, whether by local
+submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
+due to forwarding, or internally generated by Postfix.  Local submission is
+the simplest case: a queueid is assigned immediately and the sender address
+is logged (action: PICKUP\@; flowchart:~2).
+
+\SMTP{} is more complicated:
+
+\begin{enumerate}
+
+    \item First there is a connection from the remote client (action:
+        CONNECT\@; flowchart:~1).
+
+    \item This is followed by rejection of sender address, recipient
+        addresses, client \IP{} address or hostname, HELO hostname, etc.\
+        (action: REJECTION\@; flowchart:~4); acceptance of one or more
+        mails (action: CLONE\@; flowchart:~5); or some interleaving of
+        both.
+
+    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
+        Postfix has rejected any \SMTP{} commands the data will be saved to
+        the database; if not there won't be any data to save (any mails
+        accepted will already have been cloned so their data is in another
+        data structure).
+
+    \item If one or more mails were accepted there will be more log entries
+        for those mails later, see \sectionref{mail-delivery}.
+
+\end{enumerate}
+
+Re-injection due to forwarding sadly lacks explicit log lines of its
+own;\footnote{Previously discussed in \sectionref{complications},
+complication 3.} re-injection is somewhat awkward to explain because it
+overlaps both the mail acceptance and mail delivery sections, so discussion
+is deferred to \sectionref{tracking re-injected mail}.
+
+Internally generated mails lack any explicit origin in Postfix 2.2.x and
+must be detected using heuristics (see
+\sectionref{identifying-bounce-notifications} for details).  Bounce
+notifications are the primary example of internally generated mails, though
+there may be other types.\footnote{Postfix may generate mails to the
+administrator when it encounters configuration errors, but such mails are
+presumably rare.}
+
+\subsubsection{Mail delivery}
+
+\label{mail-delivery}
+
+The obvious counterpart to mail entering the system is mail leaving the
+system, whether by deletion, bouncing, local delivery, or remote delivery.
+All four are handled in exactly the same way:
+
+\begin{enumerate}
+
+    \item Postfix will log the sender and recipient addresses separately
+        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
+
+    \item Sometimes mail is re-injected and the child mail needs to be
+        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
+        \sectionref{tracking re-injected mail} discusses this in
+        detail.
+
+    \item Eventually the mail will be delivered, bounced, or deleted by the
+        administrator (action: COMMIT\@; flowchart:~12).  This is the last
+        log line for this particular mail (though it may be indirectly
+        referred to if it was re-injected).  If it is neither parent nor
+        child of re-injection the data is cleaned up and entered in the
+        database (flowchart:~14), then deleted from the state tables.
+        Re-injected mails are described in \sectionref{tracking re-injected
+        mail}.
+
+\end{enumerate}
+
+It should be reiterated that the actions above happen whether the mail is
+delivered to a mailbox, piped to a command, delivered to a remote server,
+bounced (due to a mail loop, delivery failure, or five day timeout), or
+deleted by the administrator, \textit{unless\/} the mail is either parent
+or child of re-injection, as explained in \sectionref{tracking re-injected
+mail}.
+
+\subsubsection{Tracking re-injected mail}
+
+\label{tracking re-injected mail}
+
+The crux of the problem is that re-injected mails appear in the logs
+without explicit logging indicating their source.  There are two implicit
+indications:
+
+\begin{enumerate}
+
+    \item The indicator which more commonly introduces re-injection is when
+        \daemon{qmgr} selects a mail with a previously unseen queueid for
+        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
+        which case a new data structure will be created for that mail.  The
+        mail will be flagged as having unknown origins; this flag should be
+        subsequently cleared once the origin has been established.  This
+        may also be an indicator that the mail is a bounce notification,
+        see \sectionref{identifying-bounce-notifications} for details.
+
+    \item Local delivery re-injects the mail and logs a relayed delivery
+        rather than delivering directly to a mailbox or program as it
+        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
+        delivery is performed by the \SMTP{} client; local delivery means
+        local to the server, i.e.\ an address the server is final
+        destination for.} In this case the mail may already have been
+        created (described above) and the unknown origin flag will be
+        cleared; if not a new data structure will be created.  In both
+        cases the re-injected mail is marked as a child of the original
+        mail.  The log line in question is:
+
+        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
+        delay=0, status=sent (forwarded as 56F5B43FD)}
+
+        This second indicator always occurs for re-injected mail but
+        typically occurs after the first indicator explained above.  This
+        is required to tie the parent and child mails together and so is
+        central to the process of tracking re-injected mails.
+
+\end{enumerate}
+
+The algorithm for tracking and saving re-injected mail to the database can
+finally be described:
+
+\begin{itemize}
+
+    \item If the mail is of unknown origin it is assumed to be a child mail
+        whose parent hasn't yet been identified (action: COMMIT\@;
+        flowchart:~15).  Mark the mail as ready for entry in the database
+        (flowchart:~16), and wait for the parent to deal with it
+        (flowchart:~17).  The mail should not have subsequent log entries;
+        only its parent should refer to it.
+
+    \item If the mail is a child mail then it has already been tracked
+        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
+        is cleaned up, the child is entered in the database
+        (flowchart:~19), then deleted from the state tables.  The child
+        mail will be removed from the parent mail's list of children
+        (flowchart:~20); if this is the last child and the parent has also
+        been entered in the database the parent will be deleted from the
+        state tables.
+
+    \item The last alternative is that the mail is a parent mail (action:
+        COMMIT\@; flowchart:~21).  Regardless of the state of its children
+        its data is cleaned up and entered in the database (flowchart:~22).
+        The parent may have children which are waiting to be entered in the
+        database (flowchart:~23); each of those children's data is cleaned
+        up and entered in the database, then deleted from the state tables.
+        The parent may also have outstanding children which are not yet
+        delivered (flowchart:~24), in which case the parent must wait for
+        those children to be finished with.  As soon as the last child is
+        deleted from the state tables the parent will also be finished with
+        (flowchart:~25), and deleted from the state tables.
+
+\end{itemize}
+
+A parent mail can have multiple children, which may be delivered before or
+after the parent mail.
+
+\subsubsection{Emergent behaviour}
+
+The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
+which is far more complicated than the behaviour encoded in the algorithm.
+The top-level parser could be written in pseudo-code like so (with
+indentation level indicating flow control):
+
+\begin{verbatim}
+for each line in the input files:
+    for each rule defined by the user:
+        if this rule matches the input line:
+            perform the action specified by the rule
+            skip the remaining rules
+            process the next input line
+    warn the user that the input line was not parsed
+\end{verbatim}
+
 Yet despite the brevity of the algorithm above the (simplified) flow chart
-below contains:
+in \sectionref{flow-chart} contains:
 
@@
 Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the difficulties
+clarity (these extra states are caused by solving the complications
 described in \sectionref{additional complications}).
 
-The solution to the \textit{out of order log log lines\/} complication
-encountered during development (\sectionref{out of order log lines})
-requires that a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{actions\/} a mail must pass through.  Before that complication was
-overcome, i.e.\ for the majority of the development of this parser, there
-was nothing in the parser explicitly encoding the states or paths a mail
-could take; the vast majority of the parser has no need for this flow of
-data to be specified.  This eases the process of adding new actions
-(described in \sectionref{adding new actions}), as there is no requirement
-to explicitly insert the action into an algorithmic version of the flow
-chart below.
-
-% This is very likely to move around as the flow chart is as big as
-% possible, so if any preceding text changes the flow chart may end up on
-% the next page, which may not be a bad thing, maybe.  What would be a bad
-% thing is the \clearpage seems to cause a blank page sometimes, if the
-% image is just slightly too big.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsection{Full algorithm}
-
-\label{full-algorithm}
+The solution to the complication in \sectionref{out of order log lines}
+(out of order log lines) requires a list of the acceptable combinations of
+Postfix programs a mail can pass through; however this does not correspond
+to the \textit{actions\/} a mail must pass through.  Before that
+complication was overcome, i.e.\ for the majority of the development of
+this parser, there was nothing in the parser explicitly encoding the states
+or paths a mail could take; no other component in the parser has any need
+for this flow of data to be specified.  This eases the process of adding
+new actions (described in \sectionref{adding new actions}), as there is no
+requirement to explicitly insert the action into an algorithmic version of
+the flow chart.
+
+\subsection{Framework}
+
+\label{framework}
 
@@
 unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.  The parser thus requires both a method of mapping log
-lines to the correct mail and a method of specifying the action the log
-line represents.  The former is achieved by using the \daemon{smtpd}
-process id to identify the correct mail during the initial phase, then
-switching to the queueid once the mail has been accepted.  The latter uses
-the action field of the rule which matched the log line, executing the code
-in the function named by the action.
-
-\sectionref{actions-in-detail} explains the actions in substantive detail;
-this section omits such detail because it would clutter and confuse the
-algorithm description.  The flow chart in \sectionref{flow-chart} should
-also be consulted while reading this section.  For the sake of clarity this
-description covers only the most common paths through the algorithm, as
-including every path would hinder understanding; the more uncommon paths
-are caused by the complications described in \sectionref{additional
-complications}, and are covered therein.
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: pickup; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        connect; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, etc. (action:
-        rejection; flowchart:~4); acceptance of one or more mails (action:
-        clone; flowchart:~5); or some interleaving of both.
-
-    \item The client disconnects (action: disconnect; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: save\_by\_queueid; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: track; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: commit; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator.  The exception is what happens after delivery
-to the parent or children of mail re-injected due to forwarding, as
-explained in \sectionref{tracking re-injected mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: mail\_picked\_for\_delivery; flowchart:~3), in
-        which case a new data structure will be created.  The mail will be
-        flagged as having unknown origins; this flag should be subsequently
-        cleared once the origin has been established.  This may also be an
-        indicator that the mail is a bounce notification, see
-        \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: track; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the new mail is marked as a child of the parent.  The log
-        line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        indicator is required to tie the parent and child mails together
-        and so is central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: commit;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: commit; flowchart:~18): the data is cleaned up, the child
-        is entered in the database (flowchart:~19), then deleted from the
-        state tables.  The child mail will be removed from the parent
-        mail's list of children (flowchart:~20); if this is the last child
-        and the parent has also been entered in the database the parent
-        will be deleted from the state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        commit; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered, in which case (flowchart:~24) the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
+line encountered.
+
+This functionality is provided by the framework, which both drives parsing
+overall and provides services to the actions it invokes.  The framework is
+responsible for loading rules from the database and sanity checking them,
+reading log files, matching each rule against the current log line,
+invoking the correct action, maintaining state tables, loading and saving
+state, displaying a progress indicator, and miscellaneous other tasks.
+Actions use services provided by the framework, including storing and
+retrieving data in the state tables, extracting and saving data captured by
+rules, storage of global data, debugging functions, preparing a mail for
+entry in the database, and entering mails in the database.
+
+There is a similarity between this design and the event-driven programming
+paradigm commonly used in GUI programs, where one part of the program
+responds to events (mouse clicks in a GUI program, log lines being matched
+in the parser) and invokes the correct action.
 
@@
 
-    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
-        creation of bounce messages.  This action creates a new mail if
-        necessary; if the mail already exists the unknown origin flag will
-        be removed.  The action also marks the mail as a bounce
-        notification.  To deal with complication \sectionref{Further out of
-        order log lines} this action checks a cache of recent bounce mails
-        to avoid creating bogus bounce mails when lines are logged out of
-        order.
+    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
+        bounce messages.  This action creates a new mail if necessary; if
+        the mail already exists the unknown origin flag will be removed.
+        The action also marks the mail as a bounce notification.  To deal
+        with complication \sectionref{Further out of order log lines} this
+        action checks a cache of recent bounce mails to avoid creating
+        bogus bounce mails when lines are logged out of order.
 
@@
         each time a mail is accepted the connection's state table entry
-        must be cloned; if the original data structure was used the second
-        and subsequent mails would overwrite one another's data.
+        must be cloned and saved in the state tables under its queueid; if
+        the original data structure was used then second and subsequent
+        mails would overwrite one another's data.
 
@@
         Once entered, the mail will be deleted from the state tables.
+        Deletion will be postponed if the mail is the parent of re-injected
+        mail.
 
@@
 
-    \item [DELETE] Deals with mail deleted by the administrator using the
-        administrative command \daemon{postsuper}.  This action adds a
-        dummy recipient address if required, then invokes the COMMIT action
-        to handle adding the mail to the database.  The complication this
-        action deals with is described fully in \sectionref{Mail deleted
-        before delivery is attempted}.
+    \item [DELETE] Deals with mail deleted using Postfix's administrative
+        command \daemon{postsuper}.  This action adds a dummy recipient
+        address if required, then invokes the COMMIT action to handle
+        adding the mail to the database.  The complication this action
+        deals with is described fully in \sectionref{Mail deleted before
+        delivery is attempted}.
 
@@
         connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.
+        delete the connection from the state tables.  This action deals
+        with aborted delivery attempts
+        (\sectionref{aborted-delivery-attempts}).
 
@@
 
-    \item [MAIL\_TOO\_LARGE] When the client tries to send a message larger
+    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
         than the local server accepts, the mail will be discarded, and the
@@
         \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data from the log line is then saved to the state
-        table entry.
+        created.  The data extracted from the log line is then saved to the
+        state table entry.
 
@@
         which case there's a data structure to dispose of, or it may not in
-        which case there is not.  See
+        which case there isn't.  See
         \sectionref{timeouts-during-data-phase} for the gory details.
@@
 
-Adding new actions is almost as simple as adding new rules, though actions
-are not stored in the database.  The implementor writes a subroutine which
+Adding new actions is not as simple as adding new rules, though care has
+been taken in the parser's design and implementation to make adding new
+actions as painless as possible.  The implementor writes a subroutine which
 accepts the standard arguments given to actions, and registers it as an
-action by calling the parser subroutine add\_actions() with the name of the
-new action subroutine as a parameter.  No other work is required on the
-implementor's part; all of their attention and effort can be focused on the
-correctness of their action.  The only negative aspect is that the process
-involves editing the parser source code, which makes upgrading to a later
-version of the parser more difficult, though by no means impossible.  If
-the author of the new action wishes they can take advantage of the parser's
-object oriented implementation~\cite{Wikipedia-object-orientation} by
-subclassing it and implementing their changes in the derived class,
-allowing future upgrading of the parser with greatly reduced chance of
-conflicts.\footnote{The real difference between the two approaches is where
-the new code is placed.  The simpler option is to change the parser code
-directly, but those changes will then have to be made to subsequent
-versions of the parser, and as the scope of the changes increases so does
-the chance of conflict.  The more time consuming option is to write a
+action\footnote{The new action must be registered before the rules are
+loaded, as it is an error for a rule to reference an unregistered action;
+this helps catch mistakes made when adding new rules.} by calling the
+framework subroutine add\_actions() with the name of the new action
+subroutine as a parameter.  No other work is required on the implementor's
+part to integrate the action into the parser; all of their attention and
+effort can be focused on the correctness of their action.  The only
+negative aspect is that the process involves editing the parser source
+code, which makes upgrading to a later version of the parser more
+difficult, though by no means impossible.  If the author of the new action
+wishes they can take advantage of the parser's object oriented
+implementation~\cite{Wikipedia-object-orientation} by subclassing it and
+implementing their changes in the derived class, allowing future upgrading
+of the parser with greatly reduced chance of conflicts.\footnote{The real
+difference between the two approaches is where the new code is placed.  The
+simpler option is to change the parser code directly, but those changes
+will then have to be made to subsequent versions of the parser, and as the
+scope of the changes increases so does the chance of conflict, or mistakes
+when copying the action.  The more time consuming option is to write a
 subclass containing the new actions and change the program which invokes
@@
 mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and foibles a mail server in the wild must deal with,
-but the downside is that significantly altering the configuration just to
-log messages from a different Postfix component is not an option.}
+the idiosyncrasies and peculiarities a mail server in the wild must deal
+with, but the downside is that significantly altering the configuration
+just to log messages from a different Postfix component is not an option.}
 
@@
 always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, and also prune the greatest number of mails from
-the state tables and error messages.
+improvement in the parser, prune the greatest number of mails from
+the state tables and error messages, and make the remaining problems more
+apparent.
 
@@
         happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on another server) the queueid in the message-id will not
-        match the queueid of the mail.
+        Postfix on this or another server) the queueid in the message-id
+        will not match the queueid of the mail.
 
@@
 requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).
+and \sectionref{results table}).  This heuristic is weak, but thus far it
+has not caused any difficulties, and there does not seem to be another
+check available to extend the heuristics with.
 
@@
 disconnections are either unimportant or are handled elsewhere in the
-algorithm (\sectionref{timeouts-during-data-phase}).  Sadly Postfix doesn't
+parser (\sectionref{timeouts-during-data-phase}).  Sadly Postfix doesn't
 log a message saying the mail has been discarded, as should be expected by
@@
 the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake timeout will be used throughout this
-section, but everything applies equally to lost connections.} during the
-DATA phase; when this occurs Postfix will discard the mail and the parser
-needs to discard the data associated with that mail.  It seems more
-intuitive to save the data to the database, but if a timeout occurs there
-won't yet have been any data gathered for the mail, so there is none
-available to save; the timeout is recorded with the connection data
-instead.
+is lost\footnote{For brevity's sake \textit{timeout\/} will be used
+throughout this section, but everything applies equally to lost
+connections.} during the DATA phase; when this occurs Postfix will discard
+the mail and the parser needs to discard the data associated with that
+mail.  It seems more intuitive to save the data to the database, but if a
+timeout occurs there is no data available to save; the timeout is recorded
+with the connection data instead.
 
@@
     \item If no mails have been accepted yet nothing needs to be done; the
-        timeout action ends.  The timeout action is dependant on the clone
+        TIMEOUT action ends.  The TIMEOUT action is dependant on the CLONE
         action keeping a list of all mails accepted on each connection.
@@
         individually and waiting for the reply before sending subsequent
-        commands.  If pipelining is used and Postfix rejects the sender
-        address or any of the recipient addresses there is no way for the
-        client to tell which command was rejected.  Some clients make no
-        attempt to recover and disconnect uncleanly; others presumably
-        stop pipelining and restart the mail sending operation.  The client
-        could attempt to parse the error message, but that would be
-        extremely error prone, whereas not pipelining makes the cause of
-        the error message obvious.
+        commands.  If pipelining is used and Postfix rejects any command
+        there is no way for the client to tell which command was rejected.
+        Some clients make no attempt to recover and disconnect uncleanly;
+        others presumably stop pipelining and restart the mail sending
+        operation.  The client could attempt to parse the error message,
+        but that would be extremely error prone, whereas not pipelining
+        makes the cause of the error message obvious.
+
+        XXX CHECK IF THE DESCRIPTION ABOVE IS CORRECT\@.
 
@@
 line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded, and the next log line for that queueid will
-be seen when the queueid is reused, causing a queueid clash and the
-appropriate warning.
+which has just been discarded; the next log line for that queueid will be
+seen when the queueid is reused for a different mail, causing a queueid
+clash and the appropriate warning.
 
@@
 queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked,
+\daemon{cleanup} line is parsed for that queueid the cache will be checked
+(during the MAIL\_PICKED\_FOR\_DELIVERY action),
 and the line will be deemed part of the discarded mail and discarded if it
@@
 
-During the TIMEOUT action if the mail lacks a \daemon{cleanup} line, the
-queueid and other data about the mail, particularly the timestamp of the
-last log line, are added to a global cache of timeout queueids.  This cache
-is subsequently consulted during the MAIL\_PICKED\_FOR\_DELIVERY action,
-and if the queueid is found in the cache and the \daemon{cleanup} line
-passes the checks listed above the line will be discarded.
-
 \subsubsection{Pickup logging after cleanup}
@@
 load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the \daemon{pickup} line exists a warning is generated
-by the \daemon{pickup} action, but if the following conditions are met it
-is assumed that the lines are out of order:
+the queueid given in the PICKUP line exists in the state tables a warning
+is generated by the \daemon{pickup} action, but if the following conditions
+are met it is assumed that the lines were out of order:
 
@@
         connections state table must be cleaned up, entered in the database
-        and deleted.  The queueid state table is untouched.
-
-    \item Sometimes an \daemon{smtpd} is killed by a signal, so the active
-        connection must be cleaned up, entered in the database and deleted
-        from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will die or exit uncleanly, so the
-        active connection must be cleaned up, entered in the database and
-        deleted from the connections state table.
+        if there is sufficient data, and deleted.  The queueid state table
+        is untouched.
+
+    \item Sometimes a \daemon{smtpd} is killed by a signal, so its active
+        connection must be cleaned up, entered in the database if there is
+        sufficient data, and deleted from the connections state table.
+
+    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
+        connection must be cleaned up, entered in the database if there is
+        sufficient data, and deleted from the connections state table.
 
@@
 The solution to this problem is to examine the list of programs which have
-logged messages for each mail, comparing the list against a table of
+logged lines for each mail, comparing the list against a table of
 known-good combinations of programs (this check is performed during the
@@
 must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for the flag and retries entry if it's
+SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
 found; if the additional log lines have caused the mail to reach a valid
@@
         client, then deleted by the administrator before any delivery
-        attempt (administrator deleting unwanted mail, typically due to a
-        mail loop or joe~job).  Notice that \daemon{postsuper} is required,
-        not optional, for this combination.
+        attempt was made (the unwanted mail is typically due to a mail loop
+        or joe~job).  Notice that \daemon{postsuper} is required, not
+        optional, for this combination.
 
@@
 address needs to be added, as every mail is required to have at least one
-recipient.  This complication is generally observed in few log files, but
+recipient.  This complication has been observed in few log files, but
 typically when it does occur there will be many instances of it, closely
@@
 parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated; if
+there have been a huge number of bounce notification mails generated --- if
 every bounce notification mail was cached it would dramatically increase
@@
 size of the cache ever becomes a problem it could be pruned periodically to
-keep it in check.
+keep the size in check.
 
@@
 
-This section has presented the parsing algorithm, starting with a high
-level overview of the basic algorithm, and the first group of difficulties
-thwarted.  The revised algorithm was detailed, preceded by a flow chart
-depicting the paths a mail may take through the algorithm, and followed by
-the actions which are available to rules were described in full.  The
-parser exhibits emergent behaviour far more complicated than the behaviour
-encoded in the algorithm, and this behaviour is covered in the flow chart
-subsection.  The section concludes with the remaining complications
-discovered and overcome in the process of refining and completing the
-algorithm.  Detecting, diagnosing and defeating complications forms the
-largest single portion of this document, mirroring the development of the
-parser,  The complications are described in the order they were overcome,
-with subsequent problems affecting fewer mails (often by an order or
-magnitude), though the time required to solve problems increased with each
-successive problem.
+This section presented the core of the parser, starting with a very high
+level view and the initial complications which arose.  A flow chart showing
+the paths a mail may take through the nascent simplified algorithm was
+provided, followed by an explanation of those paths, and a discussion of
+the parser's emergent behaviour --- the data from the log files creates the
+paths in the flow chart, they aren't specified anywhere in the parser.  The
+framework which holds the parser together was covered next, after which
+came a description of the current actions provided by the parser, and the
+algorithm for analysing unparsed log lines to create \regexes{} for new
+rules.  Detecting, diagnosing and defeating complications forms the largest
+single portion of this document, mirroring the development of the parser,
+The complications are described in the order they were overcome, with
+subsequent problems affecting fewer mails (often by an order or magnitude),
+though the time required to solve problems increased with each successive
+problem.
 

### 2008-03-04 doc/logparser.tex
More corrections from proofreading.

@@
 of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing lines independently.
-One of the barriers to this type of processing is the unstructured nature
-of Postfix logs, where each log line was added on an ad hoc basis as a
-requirement was discovered or new functionality was added.\footnote{A
-history of all changes made to Postfix is distributed with the source code,
-available from \url{http://www.postfix.org/}} Further complication arises
-because the set of rejection messages is not fixed: new messages can be
-added by the administrator with custom checks; every \RBL{} returns a
-different explanatory message; policy servers may log different messages
-depending on the characteristics of the connection; there are many ways in
-which the log lines may differ between servers, even within the same
-organisation --- servers may be configured differently, or running
+per-connection or per-queueid rather than processing log lines
+independently.  One of the barriers to this type of processing is the
+unstructured nature of Postfix logs, where each log line was added on an ad
+hoc basis as a requirement was discovered or new functionality was
+added.\footnote{A history of all changes made to Postfix is distributed
+with the source code, available from \url{http://www.postfix.org/}} Further
+complication arises because the set of rejection messages is not fixed: new
+messages can be added by the administrator with custom checks; every \RBL{}
+returns a different explanatory message; policy servers may log different
+messages depending on the characteristics of the connection; there are many
+ways in which the log lines may differ between servers, even within the
+same organisation --- servers may be configured differently, or running
 different version of Postfix.  This paper documents the difficult process
@@
 parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail the tables
-used for storing the data gleaned from the logs and the table that stores
-the rules.
+described in section~\refwithpage{database schema}, explaining in detail
+the tables used for storing the data gleaned from the logs and the table
+that stores the rules.
 
@@
 during execution of the algorithm are described, followed by the process of
-adding a new action.  The section concludes with an in-depth description
-of the further complications discovered and their solutions which complete
-the algorithm.
+adding a new action.  The section concludes with an in-depth description of
+the further complications discovered and their solutions which complete the
+algorithm.
 
@@
 handling mail for over 700 users.  Coverage is described both in terms of
-the fraction of lines parsed and the fraction of mails and connections
+the fraction of log lines parsed and the fraction of mails and connections
 successfully reconstructed by the parsing algorithm, including dealing with
@@
 spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database along the lines
+log files without prior pre-processing into a database in the fashion
 proposed, implemented and tested herein.
@@
 
-    \item [actions] Are invoked to deal with a log line once the line has
-        been identified by the rules: modifying data structures, handing
+    \item [actions] Are invoked to deal with a log line once it has been
+        identified by the rules: modifying data structures, handing
         complications, and saving data to the database.  The actions
@@
 rules handle low level details of identifying log lines and extracting data
-from a line, whereas the actions handle the higher level details of
+from a log line, whereas the actions handle the higher level details of
 following the path a mail takes through Postfix, assembling the required
@@
 location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the lines, and the
+rule with the action to invoke, a \regex{} to match the log lines, and the
 specification of the data to extract.  Bear in mind that the changes must
@@
 parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual lines, extracting data to
-be saved in the connections and results tables, and specifying action to
-take for that line.
+hits\_total fields).  Rules parse the individual log lines, extracting data
+to be saved in the connections and results tables, and specifying action to
+take for that log line.
 
@@
     \item [postfix\_action] This is the action Postfix must have taken to
-        generate this line.  This field is mostly ignored, but two values
-        (IGNORED and INFO) have special meaning, as described below in the
-        list of typical values.\label{postfix_action}
+        generate this log line.  This field is mostly ignored, but two
+        values (IGNORED and INFO) have special meaning, as described below
+        in the list of typical values.\label{postfix_action}
 
@@
 
-        Uninteresting lines are parsed so that any lines the parser isn't
-        capable of handling become immediately obvious errors.
+        Uninteresting log lines are parsed so that any lines the parser
+        isn't capable of handling become immediately obvious errors.
 
@@
         log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the line, or worse, might match unintentionally.
-        Rules whose program is \texttt{*} will be tried against any lines
-        which aren't parsed by program specific rules.
+        which won't match the log line, or worse, might match
+        unintentionally.  Rules whose program is \texttt{*} will be tried
+        against any log lines which aren't parsed by program specific
+        rules.
 
@@
     \item [action] The action that will be invoked when this rule matches a
-        line; a full list can be found in \sectionref{actions-in-detail}.
+        log line; a full list can be found in
+        \sectionref{actions-in-detail}.
 
@@
 
-    \item [timestamp] The time the line was logged, in seconds since the
-        epoch.
+    \item [timestamp] The time the log line was logged, in seconds since
+        the epoch.
 
@@
 the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a line and subsequently performing the requested action.
-The section continues with a discussion of rule efficiency concerns,
-referring to the graphs in \sectionref{graphs}, and finishes with an
-explanation of the algorithm used to generate new \regexes{} from unparsed
-lines.
+used when matching a log line and subsequently performing the requested
+action.  The section continues with a discussion of rule efficiency
+concerns, referring to the graphs in \sectionref{graphs}, and finishes with
+an explanation of the algorithm used to generate new \regexes{} from
+unparsed lines.
 
@@
         there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current line.}  Any given
-        rule will only be used to parse a subset of the log lines, and any
-        given log line will only be parsed by a subset of the rules.
+        already have been tried and failed on the current log line.}  Any
+        given rule will only be used to parse a subset of the log lines,
+        and any given log line will only be parsed by a subset of the
+        rules.
 
     \item The first matching rule wins: no further rules are tried against
-        that line, but there is a facility for ordering the rules so that
-        more specific rules can be tried first.
+        that log line, but there is a facility for ordering the rules so
+        that more specific rules can be tried first.
 
@@
 files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some lines and a more
-general rule to match the majority, e.g.\ separating \SMTP{} delivery to
-specific sites from \SMTP{} delivery to the rest of the world.  The
+a requirement, allowing a more specific rule to match some log lines and a
+more general rule to match the majority, e.g.\ separating \SMTP{} delivery
+to specific sites from \SMTP{} delivery to the rest of the world.  The
 algorithm provides a facility for ordering overlapping rules: the priority
@@
     \item [program and regex:] If the program in the rule equals the
-        program which logged the line a match using the \regex{} will be
-        attempted against the line; if the match is successful the action
-        will be executed, if not the next rule will be tried.  If the
-        program-specific rules don't match the log line, the parser will
-        fall back to generic rules; if those rules are unsuccessful a
+        program which logged the log line a match using the \regex{} will
+        be attempted against the log line; if the match is successful the
+        action will be executed, if not the next rule will be tried.  If
+        the program-specific rules don't match the log line, the parser
+        will fall back to generic rules; if those rules are unsuccessful a
         warning will be issued.
@@
     \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the line does not contain a
+        supplies the queueid, or zero if the log line does not contain a
         queueid.  This allows the correct mail can be found by queueid and
@@
 generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, 60.72 million lines) were
-each parsed 10 times, the first run discarded, and the execution time for
-the remaining 9 runs averaged.  The first run is discarded for two reasons:
+\numberOFlogFILES{} log files (totaling 10.08 GB, 60.72 million log lines)
+were each parsed 10 times, the first run discarded, and the execution time
+for the remaining 9 runs averaged.  The first run is discarded for two
+reasons:
 
@@
 shows the execution time in seconds, file size in MB and tens of thousands
-of lines per log file.  All three lines run roughly in parallel, giving a
-visual impression that the algorithm scales linearly with input size.  This
-impression is borne out by graph~\refwithpage{execution time vs file size
-vs number lines factor}, which plots the ratio of file size vs execution
-time and ratio of number of lines vs execution time (higher is
-better).\footnote{Table~\refwithpage{execution time vs file size vs number
-lines factor table} shows the ratios for different breakdowns of the log
-files.}  As the reader can see the ratios are quite tightly banded, showing
-that the algorithm scales linearly: the much larger log files between
-points 60 and 70 on the X axis in graph~\refwithpage{execution time vs file
-size vs number of lines graph} actually cause the ratio to increase, rather
-than decrease.  The strange behaviour where larger log files are parsed
-more efficiently is explained fully in \sectionref{Why are there dips in
-the graphs?}.
+of log lines per log file.  All three lines run roughly in parallel, giving
+a visual impression that the algorithm scales linearly with input size.
+This impression is borne out by graph~\refwithpage{execution time vs file
+size vs number lines factor}, which plots the ratio of file size vs
+execution time and ratio of number of log lines vs execution time (higher
+is better).\footnote{Table~\refwithpage{execution time vs file size vs
+number lines factor table} shows the ratios for different breakdowns of the
+log files.}  As the reader can see the ratios are quite tightly banded,
+showing that the algorithm scales linearly: the much larger log files
+between points 60 and 70 on the X axis in graph~\refwithpage{execution time
+vs file size vs number of lines graph} actually cause the ratio to
+increase, rather than decrease.  The strange behaviour where larger log
+files are parsed more efficiently is explained fully in \sectionref{Why are
+there dips in the graphs?}.
 
@@
 frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed lines, based on the algorithm developed by
-Risto Vaarandi~\cite{risto-vaarandi} for his \SLCT{}~\cite{slct-paper}.
-The differences between the two algorithms will be outlined as part of the
-general explanation below.
+creating new rules from unparsed log lines, based on the algorithm
+developed by Risto Vaarandi~\cite{risto-vaarandi} for his
+\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
+be outlined as part of the general explanation below.
 
@@
 used to group similar log lines and generate a \regex{} to match each group
-of lines.
+of log lines.
 
@@
         gained while writing rules and performs a large number of
-        substitutions on the input lines, replacing commonly occurring
+        substitutions on the log input lines, replacing commonly occurring
         variable terms (e.g.\ email addresses, \IP{} addresses, the
@@
         pass it generated a hash from a small range of values for each word
-        of each line, incrementing a counter for each hash.  The counters
-        will be used in the next pass to filter out words: if the word's
-        hash does not have a high frequency, the word itself cannot have a
-        high frequency, and there is no need to maintain a counter for it,
-        reducing the number of counters and thus the program's memory
-        consumption.
-
-    \item [Calculate word frequencies]  The position of words within a line
-        is important: a common word does not indicate similarity between
-        lines unless it occupies the same position within both
-        lines.\footnote{If a variable term within a line contains spaces,
-        it will appear to the algorithm as two words rather than one.  This
-        will alter the position of subsequent words, so a word occurring in
-        different positions in two log lines \textit{may\/} indicate
-        similarity, but the algorithm does not attempt to deal with this
-        possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the line)\/} tuple,
+        of each log line, incrementing a counter for each hash.  The
+        counters will be used in the next pass to filter out words: if the
+        word's hash does not have a high frequency, the word itself cannot
+        have a high frequency, and there is no need to maintain a counter
+        for it, reducing the number of counters and thus the program's
+        memory consumption.
+
+    \item [Calculate word frequencies]  The position of words within a log
+        line is important: a common word does not indicate similarity
+        between log lines unless it occupies the same position within both
+        log lines.\footnote{If a variable term within a line contains
+        spaces, it will appear to the algorithm as two words rather than
+        one.  This will alter the position of subsequent words, so a word
+        occurring in different positions in two log lines \textit{may\/}
+        indicate similarity, but the algorithm does not attempt to deal
+        with this possibility.}  The algorithm maintains a counter for each
+        \textit{(word, word's position within the log line)\/} tuple,
         incrementing it each time that word occurs in that position.
@@
     \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the line)\/} tuple is
+        \textit{(word, word's position within the log line)\/} tuple is
         checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all lines is generally a good starting point) it
-        is classified as a fixed word, otherwise it is classified as a
+        the user (1\% of all log lines is generally a good starting point)
+        it is classified as a fixed word, otherwise it is classified as a
         variable term.  If a variable term appears sufficiently often it
@@
     \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the line, and a counter is maintained for each \regex{}.
-        Contiguous sequences of \texttt{.+} in the newly reassembled
-        \regexes{} are collapsed to a single \texttt{.+}; any resulting
-        duplicate \regexes{} are combined, and their counters added.  If
-        the frequency of a \regex{} is lower than the threshold supplied by
-        the user the \regex{} is discarded.\footnote{This is a second
-        threshold, independent of the threshold used to differentiate
-        between fixed and variable words, but once again 1\% of input lines
-        is a good starting point; obviously the threshold depends on the
-        number and type of input lines.}  The new \regexes{} are printed
-        for the user to add to the database, either as new rules or merged
-        into the \regexes{} of existing rules; the counter for each
-        \regex{} is also printed, giving the user an indication of how many
-        of the input lines that \regex{} should match.  Discarding
-        \regexes{} will result in some of the input lines not being
-        matched; this utility should be run again once the unmatched lines
-        have been isolated by running the parser, including the new
-        \regexes{}, with the same input which produced the original set of
-        unparsed lines.
+        matching the log line, and a counter is maintained for each
+        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
+        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
+        resulting duplicate \regexes{} are combined, and their counters
+        added.  If the frequency of a \regex{} is lower than the threshold
+        supplied by the user the \regex{} is discarded.\footnote{This is a
+        second threshold, independent of the threshold used to
+        differentiate between fixed and variable words, but once again 1\%
+        of input log lines is a good starting point; obviously the
+        threshold depends on the number and type of input log lines.}  The
+        new \regexes{} are printed for the user to add to the database,
+        either as new rules or merged into the \regexes{} of existing
+        rules; the counter for each \regex{} is also printed, giving the
+        user an indication of how many of the input log lines that \regex{}
+        should match.  Discarding \regexes{} will result in some of the
+        input log lines not being matched; this utility should be run again
+        once the unmatched log lines have been isolated by running the
+        parser, including the new \regexes{}, with the same input which
+        produced the original set of unparsed log lines.
 
@@
 A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input line
-against each \regex{}, counting the number of lines which match each
-\regex{}, warning the user if an input line is matched by more than one
-\regex{}, and additionally warning if an input line is not matched by any
-\regex{}.  It displays a summary of how many input lines each \regex{}
-matched, comparing it to the expected number of matches; this provides the
-user with an easy method of checking if the \regexes{} produced by the
-first utility are correctly matching the input lines they are based upon.
-A future version of this utility will also group input lines by \regex{},
-so the user can tweak the \regexes{} if required.
+the input given to the first utility.  It tries to match each input log
+line against each \regex{}, counting the number of log lines which match
+each \regex{}, warning the user if an input log line is matched by more
+than one \regex{}, and additionally warning if an input log line is not
+matched by any \regex{}.  It displays a summary of how many input log lines
+each \regex{} matched, comparing it to the expected number of matches; this
+provides the user with an easy method of checking if the \regexes{}
+produced by the first utility are correctly matching the input log lines
+they are based upon.  A future version of this utility will also group
+input log lines by \regex{}, so the user can tweak the \regexes{} if
+required.
 
@@
     \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed lines.
+        to generate a new \regex{} from unparsed log lines.
 
@@
 child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation line!  Thus mails flagged as
-coming from an unknown origin cannot be entered into the database when
+\textbf{before} the parent logs the creation log line!  Thus mails flagged
+as coming from an unknown origin cannot be entered into the database when
 their final log line is parsed; instead they must be marked as ready for
@@
 
-    \item Check if there are any rules for the program which produced the
-        log line; if not then skip the line.
-
-    \item Try each rule until a matching rule is found.
+    \item Check if there are any rules for the Postfix component which
+        produced the log line; if not then skip the log line.
+
+    \item Try each rule until a matching rule is found; if no match is
+        issue a warning and move to the next log line.
 
@@
 
-    \item Each program of interest must have at least one rule or its log
-        lines will be silently skipped; in the extreme case of zero rules
-        the parser would happily skip every log line.  There may be any
-        number of log lines from other programs intermingled in the log
-        file, and there are some Postfix programs which don't produce any
-        log lines of interest.
-
-    \item There must be a rule to match each different line produced by
-        each program; if a line is not successfully matched the parser will
-        issue a warning.  Rules should be as specific and tightly bound as
-        possible to ensure accurate parsing:\footnote{A rule which matches
-        zero or more of any character will successfully parse every log
-        line, but not in a meaningful way.} most log lines contain fixed
-        strings and have a rigid pattern, so this is not a problem.
+    \item Each Postfix component whose log lines are of interest must have
+        at least one rule or its log lines will be silently skipped; in the
+        extreme case of zero rules the parser would happily skip every log
+        line.  There may be any number of log lines from other programs
+        intermingled in the log file, and there are some Postfix programs
+        which don't produce any log lines of interest.
+
+    \item There must be a rule to match each different log line produced by
+        each program; if a log line is not successfully matched the parser
+        will issue a warning.  Rules should be as specific and tightly
+        bound as possible to ensure accurate parsing:\footnote{A rule which
+        matches zero or more of any character will successfully parse every
+        log line, but not in a meaningful way.} most log lines contain
+        fixed strings and have a rigid pattern, so this is not a problem.
 
@@
 easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser has \numberOFrules{} rules, fully parsing
+of writing the parser is tested with \numberOFrules{} rules, fully parsing
 \numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
@@
 
-Warnings are issued for any lines which are not parsed; no warnings are
-issued for unparsed lines while testing with the \numberOFlogFILES{} test
-log files, so it can be safely concluded that there are zero false
+Warnings are issued for any log lines which are not parsed; no warnings are
+issued for unparsed log lines while testing with the \numberOFlogFILES{}
+test log files, so it can be safely concluded that there are zero false
 negatives.  False positives are harder to quantify: short of examining each
@@
 sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each line.  The sample was
+ensure that the correct \regex{} parsed each log line.  The sample was
 generated by running the following command:
@@
 Coverage of mails is much more difficult to determine accurately than
-coverage of lines.  The parser can dump its state tables in a human
+coverage of log lines.  The parser can dump its state tables in a human
 readable form; examining these tables with reference to the log files is
@@
 is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found,\footnote{There will often be warnings about a missing
-queueid or \pid{} in the first few hundred or thousand log lines because
-the earlier log lines for those connections or mails are in the previous
-log file; loading the saved state from the previous log file will solve
-this problem.} or when there are problems tracking a child mail (see
-\sectionref{tracking re-injected mail}).  There should be few or no
-warnings when parsing, and when finished parsing the state table should
-only contain entries for mails which had yet to be delivered when the log
-files ended, or started before the log files began.
+\pid{} is not found in the state tables,\footnote{There will often be
+warnings about a missing queueid or \pid{} in the first few hundred or
+thousand log lines because the earlier log lines for those connections or
+mails are in the previous log file; loading the saved state from the
+previous log file will solve this problem.} or when there are problems
+tracking a child mail (see \sectionref{tracking re-injected mail}).  There
+should be few or no warnings when parsing, and when finished parsing the
+state table should only contain entries for mails which had yet to be
+delivered when the log files ended, or were accepted before the log files
+began.
 
@@
     \item 1 connection which started only seconds before the logs ended and
-        had not yet completed.
+        had not yet been fully transferred from client to server.
 
@@
     \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection, respectively; it is extremely unlikely
+        to one mail or connection respectively; it is extremely unlikely
         that a log line would be associated with the wrong connection.
@@
         positives.  In addition the solution to the \textit{Out of order
-        log files\/} problem described in \sectionref{out of order log
+        log files\/} complication described in \sectionref{out of order log
         lines} imposes conditions which each reassembled mail must comply
@@
 
-The first 6000 lines of this log file (roughly 0.01\% of the total) was
-extracted:
+The first 6000 lines of this log file (roughly 0.01\% of the total number
+of log lines used in testing) was extracted:
 
@@
 It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections/mails as possible.  This
-log segment was parsed with all debugging options enabled, resulting in
-167,448 lines of output (27.908 lines of output per line of input).  All
-167,448 lines were examined in conjunction with the input log and a dump of
-the resulting database, verifying that for each of the input lines the
-parser used the correct rule, executed the correct action, which in turn
-produced the correct result, and inserted the correct data in the database.
-The log segment produced 4 warnings, 10 mails remaining in the state
-tables, and 1625 connections entered in the database.
+entries are present for as many of the connections as possible.  This log
+segment was parsed with all debugging options enabled, resulting in 167,448
+lines of output.\footnote{A mean of 27.908 lines of output per line of
+input; each connection has 30 debugging lines, plus 21 debugging lines per
+result.  Connections which have been cloned will have the cloned connection
+in their debugging output, plus another 33 debugging lines.  Those numbers
+are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
+between the number of log lines and debugging lines: $33(connections) +
+30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
+and has not been rigorously verified.}  All 167,448 lines were examined in
+conjunction with the input log and a dump of the resulting database,
+verifying that for each of the input lines the parser used the correct rule
+and executed the correct action, which in turn produced the correct result
+and inserted the correct data in the database.  The log segment produced 4
+warnings, 10 mails remaining in the state tables, and 1625 connections
+correctly entered in the database.
 
@@
 Parser coverage is divided into two topics in this section: log lines
-covered, and mails covered.  The former is initially more important, as the
+covered and mails covered.  The former is initially more important, as the
 parser must successfully parse every line if it is to be complete, but
@@
 mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and not intrusive:
+percentage of log lines parsed is relatively simple and non-intrusive:
 adding new rules or modifying existing rules is simplified by the
-separation of rules and algorithm.  Improving the logical coverage is
-harder, as the actions taken by Postfix must be reconstructed, and the new
-sequence of actions integrated into the existing model without breaking
-the existing sequences.  Detecting a deficiency in the parsing algorithm is
-also significantly harder than detecting unparsed log lines, as the parser
-will warn about any unparsed line, whereas discovering a flaw in the
-parser requires understanding of the warnings produced and the mails
-remaining in the state table.  Rectifying a flaw in the parser requires an
-understanding of both the parser and Postfix's logs, and investigative work
-to determine the cause of the deficiency, followed by further examination
-of the logs in developing a solution.
+separation of rules, actions and framework.  Improving the logical coverage
+is harder, as the actions taken by Postfix must be reconstructed by the
+author, and the new sequence of actions integrated into the existing model
+without breaking the existing parsing.  Detecting a deficiency in the
+parsing algorithm is also significantly harder than detecting unparsed log
+lines, as the parser will warn about any unparsed line, whereas discovering
+a flaw in the parser requires understanding of the warnings produced and
+the mails remaining in the state table.  Rectifying a flaw in the parser
+requires an understanding of both the parser and Postfix's logs, and
+investigative work to determine the cause of the deficiency, followed by
+further examination of the logs in developing a solution.
 
@@
     \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix 2.2.10
-        and 2.3.11; this may possibly have changed in Postfix 2.4.}
-        Rectifying this has already been described in \sectionref{logging
-        helo}.
+        logged if the mail is accepted.\footnote{Tested with Postfix
+        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
+        2.5.}  Rectifying this has already been described in
+        \sectionref{logging helo}.
 
@@
 
-    \item Write the policy server referred to in limitation 3 above.
+    \item Investigate and write the policy server referred to in limitation 3 above.
 

### 2008-03-05 doc/logparser.tex
Corrections after proofreading section 8.

@@
 \subsubsection{Emergent behaviour}
+
+\label{Emergent behaviour}
 
@@
 Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially given previous experience in parsing logs, but it turns out to
-be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, as a quick
-survey of sample log files gives the impression that the number of distinct
-log lines is quite small; this impression is due to the uneven distribution
-exhibited by logs produced in normal operation (see graph~\refwithpage{rule
-hits graph} for a vivid illustration of this).
+especially if one has previous experience in parsing logs, but it turns out
+to be a much more taxing project than initially expected.  The variety and
+breadth of log lines produced by Postfix is quite surprising, because a
+quick survey of sample log files gives the impression that the number of
+distinct log lines is quite small; this impression is due to the uneven
+distribution exhibited by logs produced in normal operation (see
+graph~\refwithpage{rule hits graph} for a vivid illustration of this).
 
@@
 log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the job is quite tractable, though
-it does raise interesting efficiency and optimisation questions (answered
-in \sectionref{rule efficiency}).  Providing a tool (\sectionref{creating
-new rules}) to ease the generation of \regexes{} from unparsed log lines
-should greatly help users add rules to parse formerly unparsed lines.
+resulting initial increase in complexity the task is quite tractable,
+though it does raise interesting efficiency and optimisation questions
+(answered in \sectionref{rule efficiency}).  Providing a tool
+(\sectionref{creating new rules}) to ease the generation of \regexes{} from
+unparsed log lines should greatly help users add rules to parse formerly
+unparsed log lines.
 
 The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{parsing-algorithm}) an
-almost trivial task.  The simplicity of adding a new action
-(\sectionref{adding new actions}) frees the implementor from worrying about
-how their action might disrupt parsing of other lines or the behaviour of
-other actions, allowing their concentration to be focused on correctly
-implementing their action.
-
-The division of the parser into rules and actions is unusual because it
-divides the two so completely; although parsers are often divided (lex and
-yacc~\cite{lex-and-yacc} based parsers being an obvious example), the parts
-are usually quite internally interdependent, and will be combined into a
-complete parser by the compilation process; in contrast \parsername{} keeps
-the rules and actions separate until the parser runs.  XXX EXTEND THIS\@;
-SOMETHING ABOUT THE CONTROL FLOW BEING DIFFERENT, I.E. RULE DETERMINES
-ACTION, RATHER THAN ACTION EXPECTING INPUT\@.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} exhibited by the rules
-and actions is also interesting, and is discussed beside the flow chart
-depicting said emergent behaviour (\sectionref{flow-chart}).  This emergent
-behaviour greatly eases
+lines, it makes adding new actions (\sectionref{adding new actions}) a
+relatively easy task.  The simplicity of adding a new action frees the
+implementor from worrying about how their action might disrupt parsing of
+other log lines or the behaviour of other actions, allowing their
+concentration to focus on correctly implementing their new action.
+
+The division of the parser into rules, actions and framework is unusual
+because rules are separated so completely from the actions and framework;
+although parsers are often divided (lex and yacc~\cite{lex-and-yacc} based
+parsers being an obvious example), the parts are usually quite internally
+interdependent, and will be combined into a complete parser by the
+compilation process; in contrast \parsername{} keeps the rules and actions
+separate until the parser runs.  The actions and framework are not as
+completely separated, as the actions depend on services provided by the
+framework; however the actions and framework are not tightly integrated: it
+would be possible, with some work, to completely separate the two.
+
+XXX EXTEND THIS\@; SOMETHING ABOUT THE CONTROL FLOW BEING UNUSUAL, I.E.\
+RULE DETERMINES ACTION, RATHER THAN ACTION EXPECTING INPUT\@.
+
+The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
+behaviour}) exhibited by the rules and actions is also interesting, and is
+discussed after the flow chart (\sectionref{flow-chart}) and explanation of
+the paths through the parser.  This emergent behaviour greatly eases adding
+new actions (\sectionref{adding new actions}), as the actions do not have
+to be inserted into an explicit flow of control; new actions will naturally
+find their place in the paths through the parser available to a mail.
 
@@
 explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Dealing with numerous, infrequently
-occurring log lines is a simple task, albeit tiresome, whereas dealing with
-mails where information appears to be missing is much more grueling.
-Trawling through the logs, looking for something out of the ordinary,
-possibly hundreds or even thousands of lines away from the last mention of
-the queueid in question, is extremely time consuming and error prone.  In
-many cases the task is not to spot the line which is unusual, but to spot
-that a line normally present is missing.  In all cases the evidence must be
+have been parsed successfully.  Adding new rules to deal with numerous,
+infrequently occurring log lines is a simple task, albeit tiresome (though
+alleviated by the tool described in \sectionref{adding new actions}),
+whereas dealing with mails where information appears to be missing is much
+more grueling.  Trawling through the logs, looking for something out of the
+ordinary, possibly hundreds or even thousands of lines away from the last
+mention of the queueid in question, is extremely time consuming and error
+prone.  In some cases the task is not to spot the line which is unusual,
+but to spot that a line normally present is missing, i.e.\ to realise that
+one line amongst hundreds is absent.  In all cases the evidence must be
 used to construct a hypothesis to explain the irregularities, and that
@@
 complications to be dealt with.
+
+A parser's ability to correctly parse it's input is extremely important;
+the parser's coverage of its test log files is discussed in
+\sectionref{parsing coverage}.  Both its success at parsing individual log
+lines (\sectionref{log-lines-covered}) and its correctness in
+reconstructing each mail's journey through Postfix
+(\sectionref{mails-covered}) are described in detail, including the results
+of manually verifying the correct parsing of a subset of the test log
+files.
+
+\newpage
 

### 2008-03-08 doc/logparser.tex
Small changes

@@
 \end{abstract}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
 
@@
 behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the algorithm.  The actions taken
-during execution of the algorithm are described, followed by the process of
-adding a new action.  The section concludes with an in-depth description of
-the further complications discovered and their solutions which complete the
-algorithm.
+explanation of the different stages of the initial algorithm.  The
+framework which actions and rules fit into is documented, then the actions
+taken during execution of the algorithm are described, followed by the
+process of adding a new action.  The section concludes with an in-depth
+description of the further complications discovered and their solutions
+which complete the parser.
 
@@
 
+% XXX SHOULD I HAVE THIS \newpage?
 \newpage
@@
 
-A parser's ability to correctly parse it's input is extremely important;
-the parser's coverage of its test log files is discussed in
-\sectionref{parsing coverage}.  Both its success at parsing individual log
-lines (\sectionref{log-lines-covered}) and its correctness in
-reconstructing each mail's journey through Postfix
-(\sectionref{mails-covered}) are described in detail, including the results
-of manually verifying the correct parsing of a subset of the test log
-files.
+A parser's ability to correctly parse its input is extremely important; the
+parser's coverage of its test log files is discussed in \sectionref{parsing
+coverage}.  Both its success at parsing individual log lines
+(\sectionref{log-lines-covered}) and its correctness in reconstructing each
+mail's journey through Postfix (\sectionref{mails-covered}) are described
+in detail, including the results of manually verifying the correct parsing
+of a subset of the test log files.
 
@@
 
+XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
+
 \renewcommand{\figurename}{Graph}

### 2008-03-08 doc/logparser.tex
Diagnosed, fixed and documented another complication: mails deleted by postsuper during delivery.

@@
 typically when it does occur there will be many instances of it, closely
-grouped.
+grouped.  The DELETE action is responsible for handling this complication.
 
@@
 
+\subsubsection{Mails deleted during delivery}
+
+\label{Mails deleted during delivery}
+
+The administrator can delete mails using \daemon{postsuper}; occasionally
+mails in the process of being delivered will be deleted.  This results in
+the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
+\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
+removed from the state tables and saved in the database.  The DELETE action
+adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
+action, and the current log line discarded if the following conditions are
+met:
+
+\begin{enumerate}
+
+    \item The queueid is not found in the state tables.
+
+    \item The queueid is found in the cache of deleted mails.
+
+    \item The timestamp of the log line is within 5 minutes of the end time
+        of the mail.
+
+\end{enumerate}
+
+Sadly this solution involves discarding some data, but the complication
+only arises eight times (tightly grouped in one log file, which is not in
+the group used for testing); if this complication occurred more frequently
+it might be desirable to find the mail in the database and add the log
+line to it.
+
 \subsection{Conclusion}
@@
 
+    \item The solution to complication \sectionref{Mails deleted during
+        delivery} involves discarding data.
+
 \end{enumerate}
@@
 
-    \item Investigate and write the policy server referred to in limitation 3 above.
+    \item Investigate and write the policy server referred to in limitation
+        3 above.
+
+    \item Improve the solution to complication \sectionref{Mails deleted
+        during delivery} so that data is not discarded.
 

### 2008-03-10 doc/logparser.tex
Update the number of rules.
Corrected reference to pid_regex.

@@
 \newcommand{\numberOFrules}[0]{%
-    168%
+    161%
 }
@@
     \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted from the log line.  The format is:
-        \newline \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
+        log line will be extracted.  The format is:
+        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
         \newline i.e.\ semi-colon or comma separated assignment statements,
@@
 
-        Additionally \texttt{pid\_regex} and \texttt{child} are used by the
-        \texttt{TRACK, BOUNCE} and \texttt{SMTPD\_DIED} actions.
+        Additionally \texttt{child} and \texttt{pid} are used respectively
+        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
+        actions.
 

### 2008-04-01 doc/logparser.tex
Corrections from Carl for Sections 1 and 2.

@@
 % author don't work - I think either gnuplot or graphviz clobbers it.
+% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
+% still seem to work, somehow.
 \ifpdf{}
-    \usepackage[pdftex]{hyperref}
+    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
 \else{}
-    \usepackage[dvips]{hyperref}
+    \usepackage[dvips,hyperfootnotes=false]{hyperref}
 \fi{}
@@
 
+% Extra footnote functionality, including references to earlier footnotes.
+\usepackage{footmisc}
+
 % \showgraph{filename}{caption}{label}
@@
 \end{abstract}
+
+XXX PICK A NAME AND INTRODUCE IT\@.
 
@@
 complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every \RBL{}
-returns a different explanatory message; policy servers may log different
-messages depending on the characteristics of the connection; there are many
-ways in which the log lines may differ between servers, even within the
-same organisation --- servers may be configured differently, or running
-different version of Postfix.  This paper documents the difficult process
-of parsing Postfix logs, presenting a program which parses logs and places
-the resulting data into a database for later use.  The gathered data can
-then be used to optimise current anti-spam measures, provide a baseline to
-test new anti-spam measures against, or to produce statistics showing how
-effective those measures are.  There are numerous other uses for such data:
-improving server performance by identifying troublesome destinations and
-reconfiguring appropriately; identifying regular high volume uses (e.g.\
-customer newsletters) and restricting those uses to off-peak times;
-detecting virus outbreaks which propagate via email; as a base for billing
-customers on a shared server.  Preserving the raw data enables users to
-develop a multitude of uses far beyond those conceived of by the author.
+messages can be added by the administrator with custom checks; every
+\RBL{}\footnote{This document is supplied with a glossary, see
+appendix~\ref{Glossary}.} returns a different explanatory message; policy
+servers may log different messages depending on the characteristics of the
+connection; there are many ways in which the log lines may differ between
+servers, even within the same organisation --- servers may be configured
+differently, or running different version of Postfix.  This paper documents
+the difficult process of parsing Postfix logs, presenting a program which
+parses logs and places the resulting data into a database for later use.
+The gathered data can then be used to optimise current anti-spam measures,
+provide a baseline to test new anti-spam measures against, or to produce
+statistics showing how effective those measures are.  There are numerous
+other uses for such data: improving server performance by identifying
+troublesome destinations and reconfiguring appropriately; identifying
+regular high volume uses (e.g.\ customer newsletters) and restricting those
+uses to off-peak times; detecting virus outbreaks which propagate via
+email; as a base for billing customers on a shared server.  Preserving the
+raw data enables users to develop a multitude of uses far beyond those
+conceived of by the author.
 
@@
 which complete the parser.
+
+XXX HOW MANY MAILS PER DAY\@?
 
@@
 measures against, or to produce statistics showing how effective those
-measures are.\footnote{\sectionref{introduction} lists some more example
-uses.}
+measures are.
 
@@
 properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and viruses.\footnote{This violates
-the principle of \textit{Be liberal in what you accept, and conservative in
-what you send\/} from \RFC{}~760~\cite{rfc760}, but unfortunately that
-principle was written in a friendlier time.  Given the deluge of spam that
-mail servers are subjected to daily, a more appropriate maxim could be:
-\textit{Require strict adherence to \RFC{}~2821; implement the strongest
-restrictions you can; relax the restrictions and adherence only when
-legitimate mail is impeded.\/}  It's not as friendly, nor as catchy, but it
-more accurately reflects the current situation.} A typical \SMTP{}
-conversation resembles the following (the lines starting with a three digit
-number are sent by the server, all other lines are sent by the client):
+beneficial in protecting against spam and
+viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
+adhered to the principle of \textit{Be liberal in what you accept, and
+conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
+unfortunately that principle was written in a friendlier time.  Given the
+deluge of spam that mail servers are subjected to daily, a more appropriate
+maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
+the strongest restrictions you can; relax the restrictions and adherence
+only when legitimate mail is impeded.\/}  It's not as friendly, nor as
+catchy, but it more accurately reflects the current situation.} A typical
+\SMTP{} conversation resembles the following (the lines starting with a
+three digit number are sent by the server, all other lines are sent by the
+client):
 
@@
 utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).
+hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
+would have done its best to deal with deficient clients, with the intention
+of accepting as much mail destined for its users as
+possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
+command, or accepting sender or recipient addresses which were not enclosed
+in \texttt{<>}.
 
@@
 Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
-into various component programs to perform the tasks required of an
-\MTA{}\@: receive mail, send mail, local delivery of mail, etc. --- full
-details can be found at~\cite{postfix-overview}.  Each log line contains
-the name of the Postfix program which produced it, and this information is
-used when determining which rules should be used to parse each log line
-(see \sectionref{rule characteristics} for details.)  This design has
-positive security implications also: those components which interact with
-other hosts are not privileged, and those components which are privileged
-do not interact with other hosts.
+do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
+principle was also followed when designing the parser (\sectionref{parser
+design}), notably having separate actions rather than one huge action.} and
+is separated into various component programs to perform the tasks required
+of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
+full details can be found at~\cite{postfix-overview}.  Each log line
+contains the name of the Postfix program which produced it, and this
+information is used when determining which rules should be used to parse
+each log line (see \sectionref{rule characteristics} for details.)  This
+design has positive security implications also: those components which
+interact with other hosts are not privileged, and those components which
+are privileged do not interact with other hosts.
 
@@
 
-Postfix uses several restriction lists (each containing zero or more
-restrictions) to decide the fate of an \SMTP{} command, one list for each
-stage of the \SMTP{} conversation: client connection, HELO command, MAIL
-FROM command, RCPT TO command (possibly multiple times), DATA command, and
-end of data.  By default restriction lists will not be evaluated until the
-first RCPT TO command is received, because some clients don't deal with
-earlier rejections properly; a benefit of this delay is that Postfix has
-more information available when logging the rejection.\footnote{The
-\texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
-this behaviour.}  Postfix uses simple lookup tables as the deciding factor
-for some restrictions, e.g.\newline
+Postfix uses one restriction list (containing zero or more restrictions)
+for each stage of the \SMTP{} conversation: client connection, HELO
+command, MAIL FROM command, RCPT TO commands, DATA command, and end of
+data.  The appropriate restriction list is evaluated for each stage
+(evaluation will be explained shortly), though by default the restriction
+lists for client connection, HELO and MAIL FROM commands will not be
+evaluated until the first RCPT TO command is received, because some clients
+don't deal properly with rejections before RCPT TO\@; a benefit of this
+delay is that Postfix has more information available when logging the
+rejection.  Postfix uses simple lookup tables as the deciding factor when
+evaluating some restrictions, e.g.\newline
 \tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
@@
 restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.
+restrictions and exceptions.  Postfix tries to protect the administrator in
+as far as is reasonable, e.g.\ the restriction
+\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
+because the parameter it checks (the hostname given in the HELO command) is
+under the control of the remote client.  Despite this, it is possible for
+the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
+all users --- the administrator must remember that \textit{With great
+power comes great responsibility}.
 
@@
 
-    \item Rate limiting or throttling on a per-sender, per-client or
-        per-recipient basis as performed by Policyd~\cite{policyd}.
+    \item Rate limiting on a per-sender, per-client or per-recipient basis
+        as performed by Policyd~\cite{policyd}.
 
@@
     \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.
+        appearance or disappearance of mail.  This may happen if the
+        administrator deletes mail from the queue without using 
+        \daemon{postsuper}, or if there is filesystem corruption.
 
@@
         orderings in which log lines are written to the log file, as
-        process scheduling interferes with logging and other processing.
+        process scheduling independently interferes with logging and other
+        processing.
 
@@
 
-    \item [rules] These match individual log lines and determine which
+    \item [rules] Rules match individual log lines and determine which
         actions will be executed.  Rules provide an easily extensible
@@
 
-    \item [actions] Are invoked to deal with a log line once it has been
+    \item [actions] Actions invoked to deal with a log line once it has been
         identified by the rules: modifying data structures, handing
@@
 
-    \item [framework] Is responsible for loading rules, managing data
+    \item [framework] The framework is responsible for loading rules, managing data
         structures so actions can retrieve them, finding the rule which
@@
 the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}.  \parsername{} allows the user to add new rules to the
+new actions}  \parsername{} allows the user to add new rules to the
 database without changing the parsing algorithm, unless the new log lines
@@
 
-Berkeley DB is not an \SQL{} database, supporting only \textbf{(key,
-value)} pairs; in the main table the key is an integer referred to by
-secondary tables and the value used is a \CSV{} line containing all the
-data for that row.  The secondary by-sender, by-recipient, by-date and
-by-\IP{} tables use the sender/recipient/date/\IP{} as the key, and a
-\CSV{} list of integers referring to the main table is the value.  In
-effect this re-implements \SQL{} foreign keys, but without the
-functionality offered by even the most basic of \SQL{} databases (joins,
-ordering, searches, etc.).  It also requires custom code to search on some
-combination of the above, though the authors of \LMA{} did provide some
-queries: IP-STORY, FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB
-appears to be the least useful of the three output formats
+Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
+\SQL{} databases which store arbitrary tuples; in the main table the key is
+an integer referred to by secondary tables, and the value used is a \CSV{}
+line containing all of the data for that row.  The secondary by-sender,
+by-recipient, by-date and by-\IP{} tables use the
+sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
+referring to the main table is the value.  In effect this re-implements
+\SQL{} foreign keys, but without the functionality offered by even the most
+basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
+requires custom code to search on some combination of the above, though the
+authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
+DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
+of the three output formats
 
@@
 
-Notes about the command above: there is an unnecessary line containing only
-a single space output after each matching line, for compatibility with the
-output of DAILY-REJECT\@.  Some records have an \SMTP{} code of 0, though
-this is not a valid \SMTP{} code, so presumably this is due to \LMA{} not
-parsing a line properly; \LMA{} considers this code to be a rejection, so
-the command above must too.  In a tiny proportion of records the \SMTP{}
-code is \textit{deferred\/} (again, not a valid code, and most likely due
-to mis-parsing), and is treated as a rejection by \LMA{}, so the equivalent
-\texttt{awk} command must do so too.
+Notes about the command above:
+
+\begin{enumerate}
+
+    \item It outputs an unnecessary line containing only a single space
+        after each matching line, for compatibility with the output of
+        DAILY-REJECT\@.
+
+    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
+        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
+        line properly.  \LMA{} considers this code to be a rejection, so
+        the command above must, too.
+
+    \item In a tiny proportion of records the \SMTP{} code is
+        \textit{deferred\/} (again, not a valid code, and most likely due
+        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
+        so the equivalent \texttt{awk} command must do so, too.
+
+\end{enumerate}
 
@@
 
-    \item A description of the parser's novel three section design.
+    \item A description of a new parser's novel three section design.
 
@@
 
-The database is an integral part of the parser: it stores the rules and the
-data gleaned by applying those rules to Postfix log files.  Understanding
-the database schema is important in understanding the actions of the
-parser, and essential to developing further applications which utilise the
-data gathered.
+The database is an integral part of the parser presented here: it stores
+the rules and the data gleaned by applying those rules to Postfix log
+files.  Understanding the database schema is important in understanding the
+actions of the parser, and essential to developing further applications
+which utilise the data gathered.
 

### 2008-04-01 doc/logparser.tex
Corrections from Carl for sections 3 and 4.

@@
 
-            \item [POSTFIX\_RELOAD] The administrator has caused Postfix to
-                start or stop, and all existing \daemon{smtpd} processes
-                will be terminated.
+            \item [POSTFIX\_RELOAD] The administrator has instructed
+                Postfix to start or stop, and all existing \daemon{smtpd}
+                processes will be terminated.  This does not negatively
+                impact on the logs or mail queued by Postfix for delivery.
 
@@
         cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts, so dealing with escape sequences
-        seemed unnecessary.} is permitted on the right hand side of the
-        assignment.
+        data rather than large amounts in any one rule, so dealing with
+        escape sequences was deemed unnecessary.} is permitted on the right
+        hand side of the assignment.
 
     \item [action] The action that will be invoked when this rule matches a
-        log line; a full list can be found in
-        \sectionref{actions-in-detail}.
+        log line; a full list of actions and the parameters they are
+        invoked with can be found in \sectionref{actions-in-detail}.
 
@@
         gives the queueid, or zero if the log line doesn't contain a
-        queueid.
+        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
+        logged before a mail has been accepted (a queueid won't have been
+        allocated), or log lines which aren't tied to one particular mail.
 
@@
     \item [start] The timestamp of the first log line, in seconds since the
-        epoch.
+        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
 
@@
         single connection, when the connection is reused to send multiple
-        mails.
+        mails.  XXX BETTER EXPLANATION\@.
 
@@
 
-    \item There are no sub-rules, so rules have linear computational
-        complexity.
+    \item Rule processing time is a linear function of the number of rules.
 
@@
 
-This rule would match the following log line:
+The example rule below would match the following log line:
 
@@
 This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor}, which plots the ratio of file size vs
+size vs number lines factor} which plots the ratio of file size vs
 execution time and ratio of number of log lines vs execution time (higher

### 2008-04-01 doc/logparser.tex
Corrections from Carl for section 5.

@@
 This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily once the
+in advance which connections will accept multiple mails.  Happily, once the
 mail has been accepted log entries won't be correlated by \pid{} for that
@@
 
-Unfortunately while all log lines from an individual process appear in
+The child mail has been created with queueid \texttt{56F5B43FD}.  Different
+delivery methods result in different log lines:
+
+\begin{description}
+
+    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
+
+    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
+        BD07D3C49
+
+    \item [Local delivery:] delivered to command:
+        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
+
+\end{description}
+
+Unfortunately, while all log lines from an individual process appear in
 chronological order, the order in which log lines from different processes
 are interleaved is subject to the vagaries of process scheduling.  In
-addition the first log line belonging to the child mail (the log line cited
-above belongs to the parent mail) is logged by \daemon{qmgr}, so the order
-also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
+addition, the first log line belonging to the child mail (the log line
+cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
+order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
 paranoid about mail delivery, an excellent characteristic for an \MTA{} to
@@
 better to require an explicit source, as bugs in the parser are more likely
-to be exposed.
+to be exposed; such checks helped to identify the complications described
+in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
+after cleanup}.
 
@@
 child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus mails flagged
+\textbf{before} the parent logs the creation log line!  Thus, mails flagged
 as coming from an unknown origin cannot be entered into the database when
@@
 difficult, though by no means impossible.  If the author of the new action
-wishes they can take advantage of the parser's object oriented
+wishes, they can take advantage of the parser's object oriented
 implementation~\cite{Wikipedia-object-orientation} by subclassing it and
@@
 complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next the most common was
+last.  When deciding which problem to address next, the most common was
 always chosen, as resolving the most common problem would yield the biggest
@@
 
-Once a mail has been identified as a bounce notification the unknown origin
-flag is cleared and the mail can be cleaned up and entered in the database.
+Once a mail has been identified as a bounce notification, the unknown
+origin flag is cleared and the mail can be cleaned up and entered in the
+database.
 
@@
 The most likely cause of mis-identification is if a mail generated
-internally by Postfix is identified as a bounce notification when it is a
-different type of message; arguably this is a benefit rather than a
-drawback, as other mails generated internally by Postfix will be handled
-correctly.
+internally by Postfix is identified as a bounce notification but it is
+actually a different type of message; arguably this is a benefit rather
+than a drawback, as other mails generated internally by Postfix will be
+handled correctly.
 
@@
     EHLO client.hostname
-    MAIL FROM: <sender@address>
-    RCPT TO: <recipient@address>
+    MAIL FROM: <sender@example.com>
+    RCPT TO: <recipient@example.com>
     RSET
-    MAIL FROM: <sender@address>
-    RCPT TO: <recipient@address>
+    MAIL FROM: <sender@example.com>
+    RCPT TO: <recipient@example.com>
     DATA
@@
 disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly Postfix doesn't
+parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
 log a message saying the mail has been discarded, as should be expected by
@@
     \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.  The TIMEOUT action is dependant on the CLONE
-        action keeping a list of all mails accepted on each connection.
-
-    \item \ESMTP{} pipelining allows the client to send MAIL FROM, RCPT TO
-        and DATA commands in one packet, instead of sending each command
-        individually and waiting for the reply before sending subsequent
-        commands.  If pipelining is used and Postfix rejects any command
-        there is no way for the client to tell which command was rejected.
-        Some clients make no attempt to recover and disconnect uncleanly;
-        others presumably stop pipelining and restart the mail sending
-        operation.  The client could attempt to parse the error message,
-        but that would be extremely error prone, whereas not pipelining
-        makes the cause of the error message obvious.
-
-        XXX CHECK IF THE DESCRIPTION ABOVE IS CORRECT\@.
-
-        A timeout may thus apply either to an accepted mail or a rejected
-        mail.  To distinguish between the two cases the algorithm compares
-        the timestamp of the last accepted mail against the timestamp of
-        the last line logged by \daemon{smtpd} for that connection.  If the
-        \daemon{smtpd} timestamp is later there was a rejection between the
-        accepted mail and the timeout, therefore the timeout applies to a
-        rejected mail; the timeout has already been recorded so the timeout
-        action finishes.  If the mail acceptance timestamp is greater then
-        the timeout applies to the just-accepted mail, which will be
-        discarded.
+        TIMEOUT action ends.
+
+    \item If one or more recipients have been accepted Postfix will have
+        allocated a queueid for the incoming mail, and there will be a mail
+        in the state tables which needs to be dealt with.
 
 \end{enumerate}
+
+A timeout may thus apply either to an accepted mail or a rejected mail.  To
+distinguish between the two cases the algorithm compares the timestamp of
+the last accepted mail against the timestamp of the last line logged by
+\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
+CLONE action keeping a list of all mails accepted on each connection).  If
+the mail acceptance timestamp is later then the timeout applies to the
+just-accepted mail, which will be discarded.  If the \daemon{smtpd}
+timestamp is later there was a rejection between the accepted mail and the
+timeout: the action assumes that the timeout applies to a rejected delivery
+attempt and finishes.  This assumption is not necessarily correct, because
+Postfix may have accepted an earlier recipient and rejected a later one, in
+which case the timeout applies to the accepted mail, which should be
+discarded.  This has not been a problem in practice, though it may be in
+future.
 
@@
         minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly, and
-        empirical evidence shows that ten minutes is not unreasonable;
-        hopefully it is a good compromise between false positives and false
-        negatives.
+        five minutes, but some data may have been transferred slowly
+        (perhaps because either the client or server is suffering from
+        network congestion or rate limiting), and empirical evidence shows
+        that ten minutes is not unreasonable; hopefully it is a good
+        compromise between false positives and false negatives.
 
@@
 
-    \item Sometimes a \daemon{smtpd} is killed by a signal, so its active
+    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
+        for some reason, by the administrator, or by the OS), so its active
         connection must be cleaned up, entered in the database if there is
@@
 
+    \item The TIMEOUT action uses potentially incorrect heuristics to
+        decide whether the timeout applies to an accepted mail or not,
+        potentially leaving a mail in the state tables
+        (\sectionref{timeouts-during-data-phase}).
+
     \item The program will not detect parsing the same log file twice,
@@
         during delivery} so that data is not discarded.
+
+    \item Improve the heuristics used in
+        \sectionref{timeouts-during-data-phase}, or develop another
+        solution, to avoid incorrectly leaving a mail in the state tables.
 

### 2008-04-01 doc/logparser-acronyms.tex
The remainder of Carl's simple corrections.  There are some more I need to talk to him about.

@@
     Regular expressions are a method of matching patterns in text that are
-    explained in detail in~\cite{Wikipedia-regex, perlre} .
+    explained in detail in~\cite{Wikipedia-regex, perlre}.
 }}

### 2008-04-01 doc/logparser.tex
The remainder of Carl's simple corrections.  There are some more I need to talk to him about.

@@
 sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.  The sample was
+ensure that the correct \regex{} parsed each log line.\footnote{Each log
+line was examined and the correct \regex{} identified from the
+\numberOFrules{} rules in the database; the correct \regex{} was then
+compared to the \regex{} which was used by the parser.}  The sample was
 generated by running the following command:
@@
 log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer those 5 warnings
+the side of producing more warnings rather than fewer, those 5 warnings
 represent 3 instances of 1 problem: 3 connections started before the first
@@
 
-% XXX SHOULD I HAVE THIS \newpage?
-\newpage
 \section{Conclusion}
@@
 because rules are separated so completely from the actions and framework;
-although parsers are often divided (lex and yacc~\cite{lex-and-yacc} based
-parsers being an obvious example), the parts are usually quite internally
+although parsers are often divided (parsers based on the combination of
+\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
+being an obvious example), the parts are usually quite internally
 interdependent, and will be combined into a complete parser by the
@@
 used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful the parser must
-be modified to deal with the irregularities, without adversely affecting
-the existing functionality.  The complications described in
+hypothesis must then be tested in the parser; if successful, the parser
+must be modified to deal with the irregularities, without adversely
+affecting the existing functionality.  The complications described in
 \sectionref{additional complications} were solved in the order they are
@@
         but due to the associated cost the author has not experimented with
-        it.  Presumably as the company charges per report the user wishes
-        to generate the data store (if there is one) is deliberately
+        it.  Presumably, as the company charges per report the user wishes
+        to generate, the data store (if there is one) is deliberately
         inaccessible and undocumented to prevent the user bypassing the
@@
 is probably the single most effective optimisation possible in the parser's
-implementation.
+implementation; given that it only required sixteen extra lines of
+reasonably simple code,\footnote{The sixteen lines of code breaks down as
+follows: three lines to add caching, eight lines of error checking and
+reporting, three lines to optionally disable caching for debugging and
+performance measurement, and two lines of comments.} the investment in time
+was minimal.
 

### 2008-04-03 doc/logparser.tex
Further corrections from Carl; the majority are now done.  The remainder are more difficult.

@@
 
-XXX PICK A NAME AND INTRODUCE IT\@.
+XXX EITHER PICK A BETTER WORD OR GIVE A BETTER EXPLANATION OF EMERGENT
+BEHAVIOUR --- SECTIONS 5.4.4 AND 8\@.
+
+XXX FIGURE OUT THE CAUSE OF THE ABORTED DELIVERY ATTEMPTS IN SECTION
+5.8.2\@.
+
+XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@.
+
+XXX POSSIBLE CONFERENCES\@: SIGAI IN BRITAIN, AND AICS'S AI-2008 (CAMBRIDGE
+IN DECEMBER)\@.
 
@@
 differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting a program which
-parses logs and places the resulting data into a database for later use.
-The gathered data can then be used to optimise current anti-spam measures,
-provide a baseline to test new anti-spam measures against, or to produce
-statistics showing how effective those measures are.  There are numerous
-other uses for such data: improving server performance by identifying
-troublesome destinations and reconfiguring appropriately; identifying
-regular high volume uses (e.g.\ customer newsletters) and restricting those
-uses to off-peak times; detecting virus outbreaks which propagate via
-email; as a base for billing customers on a shared server.  Preserving the
-raw data enables users to develop a multitude of uses far beyond those
-conceived of by the author.
+the difficult process of parsing Postfix logs, presenting \PLP{}, a program
+which parses logs and places the resulting data into a database for later
+use.  The gathered data can then be used to optimise current anti-spam
+measures, provide a baseline to test new anti-spam measures against, or to
+produce statistics showing how effective those measures are.  There are
+numerous other uses for such data: improving server performance by
+identifying troublesome destinations and reconfiguring appropriately;
+identifying regular high volume uses (e.g.\ customer newsletters) and
+restricting those uses to off-peak times; detecting virus outbreaks which
+propagate via email; as a base for billing customers on a shared server.
+Preserving the raw data enables users to develop a multitude of uses far
+beyond those conceived of by the author.
 
@@
 \subsection{Motivation}
+
+\label{motivation}
 
@@
 information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details.)  This
+each log line (see \sectionref{rule characteristics} for details).  This
 design has positive security implications also: those components which
-interact with other hosts are not privileged, and those components which
-are privileged do not interact with other hosts.
+interact with other hosts are not privileged,\footnote{Privilege in this
+case means the power to perform actions that are limited to the
+administrator, and not available to ordinary users.} so bugs in those
+components will not give an attacker extra privileges; those components
+which are privileged do not interact with other hosts, making it much more
+difficult for an attacker to exploit any bugs which may exist in those
+components.
 
@@
 
-    \item [sender] The sender's email address.  This can change during one
-        single connection, when the connection is reused to send multiple
-        mails.  XXX BETTER EXPLANATION\@.
+    \item [sender] The sender's email address.  Because multiple mails may
+        be delivered during one connection, there may be different sender
+        addresses in the results for one connection; however there should
+        not be different sender addresses in the results for one email.
 
@@
 
-The most likely cause of mis-identification is if a mail generated
-internally by Postfix is identified as a bounce notification but it is
-actually a different type of message; arguably this is a benefit rather
-than a drawback, as other mails generated internally by Postfix will be
-handled correctly.
+If a mail is mis-classified as a bounce message it will almost certainly
+have been generated internally by Postfix; arguably mis-classification in
+this case is a benefit rather than a drawback, as other mails generated
+internally by Postfix will be handled correctly.
 
@@
 an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  One important difference between \parsername{}
-and the parsers reviewed is that only \parsername{} makes it possible to
-parse new log lines without modifying the core parsing algorithm; see
-\sectionref{why separate rules, actions and framework?} for a discussion of
-why the separation of rules from the algorithm is beneficial.
+program with \parsername{}.  There are some important differences between
+\parsername{} and the parsers reviewed here:
+
+\begin{enumerate}
+
+    \item Only \parsername{} makes it possible to parse new log lines
+        without modifying the core parsing algorithm; see \sectionref{why
+        separate rules, actions and framework?} for a discussion of why the
+        separation of rules from the algorithm is beneficial.
+
+    \item \parsername{} does not provide any reporting based on the data it
+        gathers.  This is by design: the parser is responsible for parsing
+        logs and extracting data --- further processing of that data is a
+        separate concern, deferred to another program.  Using an \SQL{}
+        database simplifies the process of generating such reports; some
+        simple examples are given in \sectionref{motivation}.
+
+\end{enumerate}
 
 \subsection{Parsers reviewed}
+
+In this section text in italics is a quote extracted from the appropriate
+program's documentation.
 
@@
         2007/12/04), but they do not appear to be interlinked, nor does it
-        save sufficient data for the purposes of this project.  No attempt
-        is made to correlate log lines or deal with the difficulties
-        documented in \sectionref{complications} \sectionref{additional
-        complications}.\footnote{If any attempt is made there is no
-        reference to it in the documentation available on the website.} The
-        source code is available in an obfuscated form only (presumably for
-        a fee), and the product is quite expensive, as it requires a
-        license per report which is to be generated; in contrast
-        \parsername{} is free and the code is freely available.  The web
-        interface allows creation of dynamic reports based on any field,
-        but due to the associated cost the author has not experimented with
-        it.  Presumably, as the company charges per report the user wishes
-        to generate, the data store (if there is one) is deliberately
-        inaccessible and undocumented to prevent the user bypassing the
-        program and generating their own reports.
+        save sufficient data for the purposes of this project.\footnote{The
+        data extracted by Sawmill is described later in this review, and
+        the data extracted by \PLP{} is described in
+        \sectionref{connections table} and \sectionref{results table}.}  No
+        attempt is made to correlate log lines or deal with the
+        difficulties documented in \sectionref{complications}
+        \sectionref{additional complications}.\footnote{If any attempt is
+        made there is no reference to it in the documentation available on
+        the website.} The source code is available in an obfuscated form
+        only (presumably for a fee), and the product is quite expensive, as
+        it requires a license per report which is to be generated; in
+        contrast \parsername{} is free and the code is freely available.
+        The web interface allows creation of dynamic reports based on any
+        field, but due to the associated cost the author has not
+        experimented with it.  Presumably, as the company charges per
+        report the user wishes to generate, the data store (if there is
+        one) is deliberately inaccessible and undocumented to prevent the
+        user bypassing the program and generating their own reports.
 
@@
         recipients, origin hostname, origin \IP{} and virus.  The fields
-        \textit{server}, \textit{uid\/} and \textit{virus\/} are not
-        explained in their documentation: \textit{server\/} is probably the
-        server the mail is delivered to, and \textit{uid\/} might be the
-        uid of the user submitting mail locally.  Postfix does not perform
-        any form of virus checking (though it has many options for
-        cooperating with an external virus scanner), so the
-        \textit{virus\/} field is a mystery.
+        \texttt{server}, \texttt{uid} and \texttt{virus} are not explained
+        in their documentation: \texttt{server} is probably the server the
+        mail is delivered to, and \texttt{uid} might be the uid of the user
+        submitting mail locally.  Postfix does not perform any form of
+        virus checking (though it has many options for cooperating with an
+        external virus scanner), so the \texttt{virus} field is a mystery.
 
@@
 \glossary{name={cleanup},description={
-    Cleanup processes all incoming mail, sitting between qmgr and the
-    programs which accept mail (smtpd and pickup).  It removes duplicate
-    recipient addresses, inserts missing headers, and optionally rewrites
-    addresses if configured to do so.  The definitive
-    documentation is \url{http://www.postfix.org/cleanup.8.html}.
+    Cleanup processes all incoming mail after it has been accepted and
+    before it is delivered.  It removes duplicate recipient addresses,
+    inserts missing headers, and optionally rewrites addresses if
+    configured to do so.  The definitive documentation is
+    \url{http://www.postfix.org/cleanup.8.html}.
 }}

### 2008-04-08 doc/logparser.tex
Added a graph showing the number of mails accepted per day, a table with standard deviation of all days, week days and weekend days, and references to them.

@@
 
-XXX HOW MANY MAILS PER DAY\@?
-
 Section~\ref{parsing coverage} analyses the coverage the parser achieves
 over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users.  Coverage is described both in terms of
-the fraction of log lines parsed and the fraction of mails and connections
-successfully reconstructed by the parsing algorithm, including dealing with
-false negatives and a discussion of the difficulties in identifying false
-positives.  A random sampling of log lines was parsed, and the correctness
-of the results manually verified, as part of determining the coverage of
-the parser.
+handling mail for over 700 users, averaging 8500 mails per day
+(graph~\refwithpage{Mails received per day}).  Coverage is described both
+in terms of the fraction of log lines parsed and the fraction of mails and
+connections successfully reconstructed by the parsing algorithm, including
+dealing with false negatives and a discussion of the difficulties in
+identifying false positives.  A random sampling of log lines was parsed,
+and the correctness of the results manually verified, as part of
+determining the coverage of the parser.
 
@@
 vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  This section concludes with a breakdown of the rule
-hits accumulated during a single test run of the parser.
+parser performance.  Miscellaneous graphs are presented in the final group
+of graphs; these graphs are referenced from various places in the test.
+This section concludes with a breakdown of the rule hits accumulated during
+a single test run of the parser.
 
@@
 
+\subsection{Miscellaneous graphs}
+
+\label{Miscellaneous graphs}
+
+\showgraph{build/plot-mails-received}{Mails received per day}{Mails
+received per day}
+
+As expected there are far more mails received on weekdays than at weekends.
+Note that this graph and the table below show the number of mails received
+by \SMTP{} only; in particular the mail loops noticeable in other graphs
+do not contribute to these figures.
+
+\showtable{build/mails-received-include-for-graph}{Number of mails received
+per day: statistics}{Number of mails received per day: statistics}
+
+\clearpage
+
 \subsection{Conclusion}
@@
 parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} relates to the effect of
-rule ordering on execution time, and the unexpected consequences of
+and input size.  \sectionref{rule ordering graphs} demonstrates the effect
+of rule ordering on execution time, and the unexpected consequences of
 specific inputs.  The necessity of caching compiled \regexes{} is attested
 to by the third group of graphs, where the difference between caching and
-discarding is staggering.  This section concluded with a breakdown of the
-rule hits accumulated during a single test run of the parser.
+discarding is staggering.  The penultimate section contains a breakdown of
+the rule hits accumulated during a single test run of the parser.
+Miscellaneous graphs expected to be useful are collected in the final
+section.
 

### 2008-04-08 doc/logparser.tex
Improved the emergent behaviour section (I hope) and added a piece to the conclusion about the unusual flow of control.

@@
 \end{abstract}
-
-XXX EITHER PICK A BETTER WORD OR GIVE A BETTER EXPLANATION OF EMERGENT
-BEHAVIOUR --- SECTIONS 5.4.4 AND 8\@.
 
@@
 The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour encoded in the algorithm.
-The top-level parser could be written in pseudo-code like so (with
-indentation level indicating flow control):
+which is far more complicated than the behaviour explicitly encoded in the
+algorithm.  The top-level parser could be written in pseudo-code like so
+(with indentation level indicating flow control):
 
@@
 
-The solution to the complication in \sectionref{out of order log lines}
-(out of order log lines) requires a list of the acceptable combinations of
-Postfix programs a mail can pass through; however this does not correspond
-to the \textit{actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the majority of the development of
+The actions introduce some of the complexity seen in the flow chart: all
+explicit branches and the transitions between some stages (19--20,
+21--22--23--24) in the second half of the flow chart
+(figure~\refwithpage{flow chart image part 2}) are encoded within the
+actions; the transitions between the other stages (16--17, 24--25) in the
+second half arise from the interaction between actions
+
+However the transitions between stages in the first half of the flow chart
+(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
+the parser; the transitions are determined by the ordering of log lines in
+the log file.  The complexity of the first 12 stages in the flow chart
+emerges from the interaction of simple rules and actions, easing the
+process of adding new actions (described in \sectionref{adding new
+actions}), as there is no requirement to explicitly insert the action into
+an algorithmic version of the flow chart.\footnote{The solution to the
+complication in \sectionref{out of order log lines} (out of order log
+lines) requires a list of the acceptable combinations of Postfix programs a
+mail can pass through; however this does not correspond to the
+\textit{parser actions\/} a mail must pass through.  Before that
+complication was overcome, i.e.\ for the first half of the development of
 this parser, there was nothing in the parser explicitly encoding the states
 or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.  This eases the process of adding
-new actions (described in \sectionref{adding new actions}), as there is no
-requirement to explicitly insert the action into an algorithmic version of
-the flow chart.
+for this flow of data to be specified.}
 
@@
 
-XXX EXTEND THIS\@; SOMETHING ABOUT THE CONTROL FLOW BEING UNUSUAL, I.E.\
-RULE DETERMINES ACTION, RATHER THAN ACTION EXPECTING INPUT\@.
-
 The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
@@
 discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases adding
-new actions (\sectionref{adding new actions}), as the actions do not have
-to be inserted into an explicit flow of control; new actions will naturally
-find their place in the paths through the parser available to a mail.
+the paths through the parser.  This emergent behaviour greatly eases the
+process of adding new actions (\sectionref{adding new actions}), as the
+actions do not have to be inserted into an explicit flow of control; new
+actions will naturally find their place in the paths through the parser.
+
+The flow of control in this parser is quite different from that of most
+other parsers.  In most traditional parsers the parser has a current state,
+and each state has a fixed set of acceptable next states, with unacceptable
+states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
+keyword \textbf{for} it expects to immediately see a left parenthesis, with
+any other input causing parsing to fail.\footnote{Comments will already
+have been removed by the preprocessor.}  This parser is different: the
+rule which matches the next log line for a mail dictates the action that
+will be invoked, which is equivalent to the next state in other parsers;
+thus the next log line for a mail dictates the next state for that mail.
 

### 2008-04-09 doc/logparser.tex
Reformatted other parsers section. Added TODO items.

@@
 
-XXX POSSIBLE CONFERENCES\@: SIGAI IN BRITAIN, AND AICS'S AI-2008 (CAMBRIDGE
-IN DECEMBER)\@.
+XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
+IN DECEMBER\@.
+
+XXX DOWNLOAD\@: PFLOGSUMM, SAWMILL AND ANTEATER\@.
+
+CHECK PARSING FOR THE FOLLOWING\@:
+
+\begin{enumerate}
+
+    \item Pflogsumm
+    \item Sawmill Universal Log File Analysis and Reporting
+    \item Splunk
+    \item Isoqlog
+    \item AWStats
+    \item Log analyser --- throughput monitor
+    \item Anteater
+    \item Yet Another Advanced Logfile Analyser
+    \item Logparser/Lire
+    \item Logrep
+    \item Log Mail Analyser
+
+\end{enumerate}
 
@@
 \subsection{Database as Application Programming Interface}
+
+\label{database as API}
 
@@
         separate rules, actions and framework?} for a discussion of why the
-        separation of rules from the algorithm is beneficial.
+        separation of the algorithm into rules, actions and framework is
+        beneficial.
 
@@
         separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports; some
+        database simplifies the process of generating such reports
+        (discussed in \sectionref{database as API}); some
         simple examples are given in \sectionref{motivation}.
@@
 
-In this section text in italics is a quote extracted from the appropriate
-program's documentation.
-
-\begin{description}
-
-    \item [Pflogsumm] \textit{pflogsumm is designed to provide an over-view
-        of Postfix activity, with just enough detail to give the
-        administrator a ``heads up'' for potential trouble spots.\/}
-
-        Pflogsumm produces a report designed for troubleshooting, rather
-        than for in-depth analysis.  It does not support saving any data,
-        nor does it extract more data than is required for producing the
-        report.  Both the parsing and reporting are difficult to extend as
-        it is a specialised tool, unlike the easily extensible design of
-        \parsername{}.  It does not attempt to correlate log lines by
-        queueid or \pid{}, nor does it need to deal with the complications
-        encountered during this project.  Pflogsumm produces a useful
-        report, and successfully dealt with all the Postfix logs tested
+In this section text in italics is a quote extracted from the documentation
+of the program being reviewed.
+
+\subsubsection{Pflogsumm}
+
+\textit{pflogsumm is designed to provide an over-view of Postfix activity,
+with just enough detail to give the administrator a ``heads up'' for
+potential trouble spots.\/}
+
+Pflogsumm produces a report designed for troubleshooting, rather than for
+in-depth analysis.  It does not support saving any data, nor does it
+extract more data than is required for producing the report.  Both the
+parsing and reporting are difficult to extend as it is a specialised tool,
+unlike the easily extensible design of \parsername{}.  It does not attempt
+to correlate log lines by queueid or \pid{}, nor does it need to deal with
+the complications encountered during this project.  Pflogsumm produces a
+useful report, and successfully dealt with all the Postfix logs tested
 with, but is not a suitable base for this project.
 
-        \url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last
-        checked 2007/08/13.)
-
-    \item [Sawmill Universal Log File Analysis and Reporting] \textit{
-        Sawmill is a \newline Postfix log analyzer (it also support 686
-        other log formats).  It can process log files in Postfix format,
-        and generate dynamic statistics from them, analyzing and reporting
-        events.  Sawmill can parse Postfix logs, import them into a SQL
-        database (or its own built-in database), aggregate them, and
-        generate dynamically filtered reports, all through a web interface.
-        Sawmill can perform Postfix analysis on any platform, including
-        Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other UNIX, and
-        more.\/}
-
-        Sawmill is a general purpose commercial product which parses 687
-        log file formats (correct as of 2007/12/04) and produces reports.
-        Its data extraction facilities are quite limited, though it does
-        extract three different sets of data for Postfix (one is beta as of
-        2007/12/04), but they do not appear to be interlinked, nor does it
-        save sufficient data for the purposes of this project.\footnote{The
-        data extracted by Sawmill is described later in this review, and
-        the data extracted by \PLP{} is described in
-        \sectionref{connections table} and \sectionref{results table}.}  No
-        attempt is made to correlate log lines or deal with the
+\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
+2007/08/13.)
+
+\subsubsection{Sawmill Universal Log File Analysis and Reporting}
+
+\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
+formats).  It can process log files in Postfix format, and generate dynamic
+statistics from them, analyzing and reporting events.  Sawmill can parse
+Postfix logs, import them into a SQL database (or its own built-in
+database), aggregate them, and generate dynamically filtered reports, all
+through a web interface.  Sawmill can perform Postfix analysis on any
+platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
+UNIX, and more.\/}
+
+Sawmill is a general purpose commercial product which parses 687 log file
+formats (correct as of 2007/12/04) and produces reports.  Its data
+extraction facilities are quite limited, though it does extract three
+different sets of data for Postfix (one is beta as of 2007/12/04), but they
+do not appear to be interlinked, nor does it save sufficient data for the
+purposes of this project.\footnote{The data extracted by Sawmill is
+described later in this review, and the data extracted by \PLP{} is
+described in \sectionref{connections table} and \sectionref{results
+table}.}  No attempt is made to correlate log lines or deal with the
 difficulties documented in \sectionref{complications}
-        \sectionref{additional complications}.\footnote{If any attempt is
-        made there is no reference to it in the documentation available on
-        the website.} The source code is available in an obfuscated form
-        only (presumably for a fee), and the product is quite expensive, as
-        it requires a license per report which is to be generated; in
-        contrast \parsername{} is free and the code is freely available.
-        The web interface allows creation of dynamic reports based on any
-        field, but due to the associated cost the author has not
-        experimented with it.  Presumably, as the company charges per
-        report the user wishes to generate, the data store (if there is
-        one) is deliberately inaccessible and undocumented to prevent the
-        user bypassing the program and generating their own reports.
-
-        \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline
-        Fields extracted: from, to, server, UID, relay, status, number of
-        recipients, origin hostname, origin \IP{} and virus.  The fields
-        \texttt{server}, \texttt{uid} and \texttt{virus} are not explained
-        in their documentation: \texttt{server} is probably the server the
-        mail is delivered to, and \texttt{uid} might be the uid of the user
-        submitting mail locally.  Postfix does not perform any form of
-        virus checking (though it has many options for cooperating with an
-        external virus scanner), so the \texttt{virus} field is a mystery.
-
-        \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline
-        Fields extracted: from, to, \RBL{}, client hostname and client
-        \IP{}\@.
-
-        \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
-        \newline Fields extracted: from, to, client hostname, client \IP{},
-        relay hostname, relay \IP{}, status, response code, \RBL{} and
-        message id.
-
-        Even if the three data sets were linked together Sawmill would
-        extract less data than \parsername{}, and it does not appear to
-        extract data about rejections except when the rejection is caused
-        by an \RBL{} check.
+\sectionref{additional complications}.\footnote{If any attempt is made
+there is no reference to it in the documentation available on the website.}
+The source code is available in an obfuscated form only (presumably for a
+fee), and the product is quite expensive, as it requires a license per
+report which is to be generated; in contrast \parsername{} is free and the
+code is freely available.  The web interface allows creation of dynamic
+reports based on any field, but due to the associated cost the author has
+not experimented with it.  Presumably, as the company charges per report
+the user wishes to generate, the data store (if there is one) is
+deliberately inaccessible and undocumented to prevent the user bypassing
+the program and generating their own reports.
+
+\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
+extracted: from, to, server, UID, relay, status, number of recipients,
+origin hostname, origin \IP{} and virus.  The fields \texttt{server},
+\texttt{uid} and \texttt{virus} are not explained in their documentation:
+\texttt{server} is probably the server the mail is delivered to, and
+\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
+does not perform any form of virus checking (though it has many options for
+cooperating with an external virus scanner), so the \texttt{virus} field is
+a mystery.
+
+\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
+extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
+
+\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
+extracted: from, to, client hostname, client \IP{}, relay hostname, relay
+\IP{}, status, response code, \RBL{} and message id.
+
+Even if the three data sets were linked together Sawmill would extract less
+data than \parsername{}, and it does not appear to extract data about
+rejections except when the rejection is caused by an \RBL{} check.
 
@@
 
-    \item [Splunk] \textit{Splunk is an IT Search engine. It is software
-        that indexes any format of IT data from any source in real time,
-        including logs, configurations, scripts, code, messages, traps,
-        alerts, activity reports, stack traces and metrics from all of your
-        applications, servers and devices. Splunk lets you search,
-        navigate, alert and report on all your IT data in real time using
-        an AJAX web interface. You can also share knowledge and Splunk
-        solutions with other members of the Splunk community via
+\subsubsection{Splunk}
+
+\textit{Splunk is an IT Search engine. It is software that indexes any
+format of IT data from any source in real time, including logs,
+configurations, scripts, code, messages, traps, alerts, activity reports,
+stack traces and metrics from all of your applications, servers and
+devices. Splunk lets you search, navigate, alert and report on all your IT
+data in real time using an AJAX web interface. You can also share knowledge
+and Splunk solutions with other members of the Splunk community via
 SplunkBase.\/}
 
-        Splunk aims to index all an organisation's logs, providing a
-        centralised view capable of searching and correlating diverse log
-        sources.  The web interface supports complicated searches,
-        providing statistics and graphs in real time, a facility not
-        provided by \parsername{} (report generation has been deferred to a
-        subsequent program).  Saved searches\footnote{The author was unable
-        to save searches, though that may have been due to limitations in
-        the free version.} can be run periodically and the results emailed
-        to a recipient or sent to a shell script, which presumably can
-        publish the results as required (though possibly without the graphs
-        and detailed statistics); \parsername{} provides the database and
-        leaves it to the user to utilise it in any way, whenever they want,
-        with no restrictions on usage of the data.  The interface is
-        optimised for interactive rather than automated queries and it does
-        not appear to be possible to write independent tools to utilise the
-        Splunk database.  Many types of reports are available, though most
-        are variations of a bar or pie chart, with the exception of bubble
-        and heatmap graphs.  It is very easy to drill down through the
-        graphs to extract a portion of the data (e.g.\ select the hour with
-        the largest number of events, then select a particular host, and
-        finally a specific address), though it is not possible to search on
-        partial words.  The searches are quite fast, though as the free
-        version has a limit on the amount of data indexed per day the
-        volume of data to be searched was merely 45MB\@; the cheapest
-        licensed version costs \$6000, and still limits the volume of data
-        indexed per day.
-
-        In the specific case of parsing Postfix logs, Splunk extracts some
-        standard fields: to and from addresses, HELO hostname, date, host
-        the logs were collected from, and protocol.  It parses the standard
-        syslog fields at the beginning of the line, and extracts any
-        \texttt{key=<value>} pairs occurring after the standard syslog
-        prologue; these pairs are the fields listed above.  Searches can be
-        based on the extracted fields, and all text in the line is also
-        available for searching.  \parsername{} extracts noticeably more
-        data, though it does not make the full text of the line available,
-        and the full power of \SQL{} is available when searching, allowing
-        the user to search on arbitrarily complicated conditions.
-
-        Splunk is a generic tool, so it lacks any Postfix specific support
-        over and above extracting the \texttt{key=<value>} fields; most
-        importantly it makes no attempt to correlate log lines by queueid
-        or \pid{}, nor to handle any of the myriad complications discussed
-        in this document.
+Splunk aims to index all an organisation's logs, providing a centralised
+view capable of searching and correlating diverse log sources.  The web
+interface supports complicated searches, providing statistics and graphs in
+real time, a facility not provided by \parsername{} (report generation has
+been deferred to a subsequent program).  Saved searches\footnote{The author
+was unable to save searches, though that may have been due to limitations
+in the free version.} can be run periodically and the results emailed to a
+recipient or sent to a shell script, which presumably can publish the
+results as required (though possibly without the graphs and detailed
+statistics); \parsername{} provides the database and leaves it to the user
+to utilise it in any way, whenever they want, with no restrictions on usage
+of the data.  The interface is optimised for interactive rather than
+automated queries and it does not appear to be possible to write
+independent tools to utilise the Splunk database.  Many types of reports
+are available, though most are variations of a bar or pie chart, with the
+exception of bubble and heatmap graphs.  It is very easy to drill down
+through the graphs to extract a portion of the data (e.g.\ select the hour
+with the largest number of events, then select a particular host, and
+finally a specific address), though it is not possible to search on partial
+words.  The searches are quite fast, though as the free version has a limit
+on the amount of data indexed per day the volume of data to be searched was
+merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
+the volume of data indexed per day.
+
+In the specific case of parsing Postfix logs, Splunk extracts some standard
+fields: to and from addresses, HELO hostname, date, host the logs were
+collected from, and protocol.  It parses the standard syslog fields at the
+beginning of the line, and extracts any \texttt{key=<value>} pairs
+occurring after the standard syslog prologue; these pairs are the fields
+listed above.  Searches can be based on the extracted fields, and all text
+in the line is also available for searching.  \parsername{} extracts
+noticeably more data, though it does not make the full text of the line
+available, and the full power of \SQL{} is available when searching,
+allowing the user to search on arbitrarily complicated conditions.
+
+Splunk is a generic tool, so it lacks any Postfix specific support over and
+above extracting the \texttt{key=<value>} fields; most importantly it makes
+no attempt to correlate log lines by queueid or \pid{}, nor to handle any
+of the myriad complications discussed in this document.
 
@@
 
-    \item [Isoqlog] \textit{Isoqlog is an MTA log analysis program written
-        in C. It designed to scan qmail, postfix, sendmail and exim logfile
-        and produce usage statistics in \HTML{} format for viewing through
-        a browser. It produces Top domains output according to Sender,
-        Receiver, Total mails and bytes; it keeps your main domain mail
-        statistics with regard to Days Top Domain, Top Users values for per
-        day, per month and years.\/}
-
-        Isoqlog produces a report listing the number of mails sent by each
-        unique sender address, and separately the total number of bytes
-        transferred; both reports are produced for daily, monthly and
-        annual time spans, but only for the domains listed in its
-        configuration file (making it impossible to produce reports for
-        every sender domain).  It appears to ignore all log lines except
-        for those for the current day, though it does maintain a record of
-        data previously extracted, which the newly extracted data is merged
-        into (no information is provided on the format of the data store).
-        The data extracted appears to be limited to the number of mails
-        sent by each sender, unlike the copious amounts of data extracted
-        by \parsername{}.  It doesn't utilise rejection log lines in any
-        way, so is unsuitable for the purposes of this project.  Its
-        parsing is completely inextensible, indeed is almost
-        incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-        low level string manipulation; it is the opposite end of the
-        spectrum to \parsernames{} parsing.  It doesn't handle any of the
-        complications discussed in this document, doesn't gather the
-        breadth of data required for this project, and ignores the majority
-        of log lines produced by Postfix.
-
-        \url{http://www.enderunix.org/isoqlog/} \newline (Last checked
-        2007/08/13.)
-
-    \item [AWStats] \textit{AWStats is a free powerful and featureful tool
-        that generates advanced web, streaming, ftp or mail server
-        statistics, graphically. This log analyzer works as a CGI or from
-        command line and shows you all possible information your log
-        contains, in few graphical web pages. It uses a partial information
-        file to be able to process large log files, often and quickly. It
-        can analyze log files from all major server tools like Apache log
-        files (NCSA combined/XLF/ELF log format or common/CLF log format),
-        WebStar, IIS (W3C log format) and a lot of other web, proxy, wap,
-        streaming servers, mail servers and some ftp servers.\/}
+\subsubsection{Isoqlog}
+
+\textit{Isoqlog is an MTA log analysis program written in C. It designed to
+scan qmail, postfix, sendmail and exim logfile and produce usage statistics
+in \HTML{} format for viewing through a browser. It produces Top domains
+output according to Sender, Receiver, Total mails and bytes; it keeps your
+main domain mail statistics with regard to Days Top Domain, Top Users
+values for per day, per month and years.\/}
+
+Isoqlog produces a report listing the number of mails sent by each unique
+sender address, and separately the total number of bytes transferred; both
+reports are produced for daily, monthly and annual time spans, but only for
+the domains listed in its configuration file (making it impossible to
+produce reports for every sender domain).  It appears to ignore all log
+lines except for those for the current day, though it does maintain a
+record of data previously extracted, which the newly extracted data is
+merged into (no information is provided on the format of the data store).
+The data extracted appears to be limited to the number of mails sent by
+each sender, unlike the copious amounts of data extracted by \parsername{}.
+It doesn't utilise rejection log lines in any way, so is unsuitable for the
+purposes of this project.  Its parsing is completely inextensible, indeed
+is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
+low level string manipulation; it is the opposite end of the spectrum to
+\parsernames{} parsing.  It doesn't handle any of the complications
+discussed in this document, doesn't gather the breadth of data required for
+this project, and ignores the majority of log lines produced by Postfix.
+
+\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
+
+\subsubsection{AWStats}
+
+\textit{AWStats is a free powerful and featureful tool that generates
+advanced web, streaming, ftp or mail server statistics, graphically. This
+log analyzer works as a CGI or from command line and shows you all possible
+information your log contains, in few graphical web pages. It uses a
+partial information file to be able to process large log files, often and
+quickly. It can analyze log files from all major server tools like Apache
+log files (NCSA combined/XLF/ELF log format or common/CLF log format),
+WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
+servers, mail servers and some ftp servers.\/}
 
 AWStats will produce simple graphs for many different services, but
-        supporting many different services without special purpose code
-        restricts it to supporting the \LCD{}.  The data it will extract
-        from an \MTA{} log file is limited in comparison to \parsername{}:
-        \newline \tab{} time2, email, email\_r, host, host\_r, method, url,
-        code and bytesd.\newline  There does not seem to be an explanation
-        of any of those fields in the documentation (\parsername{} provides
-        copious documentation).  AWStats has no special purpose code to
-        deal with the intricacies of Postfix logs, in fact it operates by
-        coercing Postfix logs into Apache\footnote{The Apache web server is
-        the most popular HTTP server in use for the past 10 years; more
-        information is available at \url{http://httpd.apache.org/}.} format
-        log files, for analysis by AWStats' HTTP log file parser.  The
-        converting parser only deals with a small portion of the log lines
-        generated by Postfix, silently skipping those it cannot deal with,
-        and does not distinguish between different types of rejection;
-        there is no easy way to extend it to handle new log lines.
-        Although though it does correlate log lines by queueid, it does not
-        deal with any of the other complications described in this
-        document.
-
-        \url{http://awstats.sourceforge.net/} \newline
+supporting many different services without special purpose code restricts
+it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
+file is limited in comparison to \parsername{}: \newline \tab{} time2,
+email, email\_r, host, host\_r, method, url, code and bytesd.\newline
+There does not seem to be an explanation of any of those fields in the
+documentation (\parsername{} provides copious documentation).  AWStats has
+no special purpose code to deal with the intricacies of Postfix logs, in
+fact it operates by coercing Postfix logs into Apache\footnote{The Apache
+web server is the most popular HTTP server in use for the past 10 years;
+more information is available at \url{http://httpd.apache.org/}.} format
+log files, for analysis by AWStats' HTTP log file parser.  The converting
+parser only deals with a small portion of the log lines generated by
+Postfix, silently skipping those it cannot deal with, and does not
+distinguish between different types of rejection; there is no easy way to
+extend it to handle new log lines.  Although though it does correlate log
+lines by queueid, it does not deal with any of the other complications
+described in this document.
+
+\noindent\url{http://awstats.sourceforge.net/} \newline
 \url{http://awstats.sourceforge.net/awstats.mail.html} \newline
@@
 
-    \item [Log analyser --- throughput monitor] This utility tracks the
-        number of events which occurred over a particular time and warns if
-        the frequency of events passes a certain threshold.  It's designed
-        to provide real time alerts when dictionary attacks, mail loops or
+\subsubsection{Log analyser --- throughput monitor}
+
+This utility tracks the number of events which occurred over a particular
+time and warns if the frequency of events passes a certain threshold.  It's
+designed to provide real time alerts when dictionary attacks, mail loops or
 similar problems occur. It doesn't attempt to extract or save data,
-        correlate log lines, or any of the more advanced tasks described in
-        this document, because it is not designed to do so.  The user must
-        construct the \regexes{} to match significant input lines before it
-        will parse any logs.  This program was not reviewed in further
-        detail because its aims are so far removed from the aims of this
-        project.
-
-        \url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline
-        (Last checked 2007/08/13.)
-
-    \item [Anteater] \textit{The Anteater project is a Mail Traffic
-        Analyser. Anteater supports currently the logformat produced by
-        Sendmail and by Postfix. The tool is written in 100\% C++ and is
-        very easy to customize. Input, output, and the analysis are modular
-        class objects with a clear interface. There are eight useful
-        analyse modules, writing the result in plain ASCII or \HTML{}, to
+correlate log lines, or any of the more advanced tasks described in this
+document, because it is not designed to do so.  The user must construct the
+\regexes{} to match significant input lines before it will parse any logs.
+This program was not reviewed in further detail because its aims are so far
+removed from the aims of this project.
+
+\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
+checked 2007/08/13.)
+
+\subsubsection{Anteater}
+
+\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
+currently the logformat produced by Sendmail and by Postfix. The tool is
+written in 100\% C++ and is very easy to customize. Input, output, and the
+analysis are modular class objects with a clear interface. There are eight
+useful analyse modules, writing the result in plain ASCII or \HTML{}, to
 stdout or to files.\/}
 
-        Anteater doesn't have any English documentation so it's difficult,
-        nigh impossible, for this author to accurately comment on what
-        analysis it performs.  It did not run successfully when tested, and
-        its parsing would certainly be out of date as Postfix has evolved
-        considerably since this tool was last updated (November 2003).  As
-        it neither ran successfully nor has documentation the author can
-        read a detailed review cannot be provided.
-
-        \url{http://anteater.drzoom.ch/} \newline (Last checked
-        2007/08/13.)
-
-    \item [Yet Another Advanced Logfile Analyser] \textit{yaala is a very
-        flexible analyser for all kinds of logfiles. It uses parsers to
-        extract information from a logfile, an SQL-like query language to
-        relate the information to each other and an output-module to format
-        the information appropriately.\/}
-
-        YAALA uses a plugin based system to analyse log files and produce
-        \HTML{} output reports, with all the parsing and report generation
-        handled by modules.  Using YAALA as a base would be only slightly
-        less work than starting from scratch, as both input and output
-        modules would need to be written specially; it may even be more
-        work to implement the parser within the constraints of YAALA\@.
-        YAALA supports storing previously gathered data using Perl's
-        Storable module~\cite{perl-storable}, so with enough knowledge of
-        the data structure YAALA stores it should be possible to extract
-        data with another Perl program also using the Storable module;
-        \parsername{} uses a well documented database which is accessible
-        from the majority of programming languages.
-
-        YAALA provides a Postfix parser which extracts the following fields
-        from specific log lines:
-
-        \tab{}Aggregations: count (not explained), bytes (sum of bytes
-        transferred).\newline \tab{}Keyfields [sic]: date, hour, sender,
-        recipient, defer\_count, delay, incoming\_host, outgoing\_host.
-
-        YAALA extracts most of the fields \parsername{} does, but it does
-        not maintain separate counters for each restriction like
-        \parsername{}, so its data cannot be used for optimisation, testing
-        or understanding of restrictions.  YAALA's Postfix parser does not
-        deal with the complications explained in this document, though it
-        does correlate log lines by queueid.
-
-        YAALA provides a mini-language based on \SQL{} that is used when
-        generating reports; sample reports can be seen
-        at~\cite{yaala-samples}.  Example query: \newline \tab{}
-        \texttt{requests BY file WHERE host =\~{} Google} \newline The
-        mini-language is quite limited and cannot be used to extract data
-        for external use, merely to create reports.
-
-        In summary YAALA provides a Postfix parser which only handles the
-        most common Postfix log lines, provides reasonably flexible report
-        generation from the limited data extracted, but has no facilities
-        to extract data for use in other tools.
+Anteater doesn't have any English documentation so it's difficult, nigh
+impossible, for this author to accurately comment on what analysis it
+performs.  It did not run successfully when tested, and its parsing would
+certainly be out of date as Postfix has evolved considerably since this
+tool was last updated (November 2003).  As it neither ran successfully nor
+has documentation the author can read a detailed review cannot be provided.
+
+\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
+
+\subsubsection{Yet Another Advanced Logfile Analyser}
+
+\textit{yaala is a very flexible analyser for all kinds of logfiles. It
+uses parsers to extract information from a logfile, an SQL-like query
+language to relate the information to each other and an output-module to
+format the information appropriately.\/}
+
+YAALA uses a plugin based system to analyse log files and produce \HTML{}
+output reports, with all the parsing and report generation handled by
+modules.  Using YAALA as a base would be only slightly less work than
+starting from scratch, as both input and output modules would need to be
+written specially; it may even be more work to implement the parser within
+the constraints of YAALA\@.  YAALA supports storing previously gathered
+data using Perl's Storable module~\cite{perl-storable}, so with enough
+knowledge of the data structure YAALA stores it should be possible to
+extract data with another Perl program also using the Storable module;
+\parsername{} uses a well documented database which is accessible from the
+majority of programming languages.
+
+YAALA provides a Postfix parser which extracts the following fields from
+specific log lines:
+
+\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
+transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
+outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
+
+YAALA extracts most of the fields \parsername{} does, but it does not
+maintain separate counters for each restriction like \parsername{}; this
+rules out the possibility of using the collected data for  optimisation,
+testing or understanding of restrictions.  YAALA's Postfix parser does not
+deal with the complications explained in this document, though it does
+correlate log lines by queueid.
+
+YAALA provides a mini-language based on \SQL{} that is used when generating
+reports; sample reports can be seen at~\cite{yaala-samples}.  Example
+query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
+\newline The mini-language is quite limited and cannot be used to extract
+data for external use, merely to create reports.
+
+In summary YAALA provides a Postfix parser which only handles the most
+common Postfix log lines, provides reasonably flexible report generation
+from the limited data extracted, but has no facilities to extract data for
+use in other tools.
 
@@
 
-    \item [Logparser/Lire] \textit{As any good system administrator knows,
-        there's a lot more to keep track of in an active network than just
-        webservers. Lire is hands down the most versatile log analysis
-        software available today. Lire not only keeps you informed about
-        your HTTP, FTP, and mail traffic, it also reports on your
-        firewalls, your print servers, and your DNS activity. The ever
-        growing list of Lire-supported services clearly outstrips any other
-        software, in large part thanks to the numerous volunteers who have
-        pioneered many new services and features. Lire is a total solution
-        for your log analysis needs.\/}
-
-        Lire is a general purpose log parser supporting many different
-        types of log file.  Its Postfix parser extracts the following data
-        from Postfix logs: \textit{The email servers' reports will show you
-        the number of deliveries and the volume of email delivered by day,
-        the domains from which you receive or send the most emails, the
-        relays most used, etc.\/}; notably rejections are not mentioned or
-        dealt with, and unlike \parsername{} there is no facility to extend
-        the parser.  It supports multiple output formats for generated
-        reports (text, \HTML{}, \PDF{} and Excel 95) but the reports do not
-        appear to be customisable; \parsername{} does not produce any
-        reports.  Lire supports saving extracted data for later report
-        generation, but accessing this data from another application is
-        undocumented; given the source code it should be possible, with
-        sufficient time and effort, to access the data from an external
-        program.
-
-        Like AWStats and Logrep, Lire attempts to correlate log lines by
-        queueid, but not by \pid{}, so the complete list of recipients for
-        a mail should be available; however its parser extracts only part
-        of the available data and makes no attempt to deal with the other
-        complications described in \sectionref{complications}
-        \sectionref{additional complications}.  \parsername{} uses an
-        \SQL{} database to make accessing the extracted data as easy as
-        possible.
-
-        \url{http://logreport.org/lire.html} \newline (Last checked
-        2007/10/09.)
-
-    \item [Logrep] \textit{Logrep is a secure multi-platform framework for
-        the collection, extraction, and presentation of information from
-        various log files. It features HTML reports, multi dimensional
-        analysis, overview pages, SSH communication, and graphs, and
-        supports over 30 popular systems including Snort, Squid, Postfix,
-        Apache, Sendmail, syslog, ipchains, iptables, NT event logs,
-        Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
+\subsubsection{Logparser/Lire}
+
+\textit{As any good system administrator knows, there's a lot more to keep
+track of in an active network than just webservers. Lire is hands down the
+most versatile log analysis software available today. Lire not only keeps
+you informed about your HTTP, FTP, and mail traffic, it also reports on
+your firewalls, your print servers, and your DNS activity. The ever growing
+list of Lire-supported services clearly outstrips any other software, in
+large part thanks to the numerous volunteers who have pioneered many new
+services and features. Lire is a total solution for your log analysis
+needs.\/}
+
+Lire is a general purpose log parser supporting many different types of log
+file.  Its Postfix parser extracts the following data from Postfix logs:
+\textit{The email servers' reports will show you the number of deliveries
+and the volume of email delivered by day, the domains from which you
+receive or send the most emails, the relays most used, etc.\/}; notably
+rejections are not mentioned or dealt with, and unlike \parsername{} there
+is no facility to extend the parser.  It supports multiple output formats
+for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
+do not appear to be customisable; \parsername{} does not produce any
+reports.  Lire supports saving extracted data for later report generation,
+but accessing this data from another application is undocumented; given the
+source code it should be possible, with sufficient time and effort, to
+access the data from an external program.
+
+Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
+but not by \pid{}, so the complete list of recipients for a mail should be
+available; however its parser extracts only part of the available data and
+makes no attempt to deal with the other complications described in
+\sectionref{complications} \sectionref{additional complications}.
+\parsername{} uses an \SQL{} database to make accessing the extracted data
+as easy as possible.
+
+\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
+
+\subsubsection{Logrep}
+
+\textit{Logrep is a secure multi-platform framework for the collection,
+extraction, and presentation of information from various log files. It
+features HTML reports, multi dimensional analysis, overview pages, SSH
+communication, and graphs, and supports over 30 popular systems including
+Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
+event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
 
@@
 
-            \item For mail sent and received: from address, size, and time
-                and date.
+    \item For mail sent and received: from address, size, and time and
+        date.
 
@@
 
-        It also counts the number of log lines parsed and skipped.  Log
-        lines are correlated based on the queueid (referred to as
-        sessionname [sic] within Logrep), but not by \pid{}.  The parsing
-        is error prone: empty fields are saved when the log line doesn't
-        match the \regex{}, though it appears that they will not overwrite
-        existing data.  Most notably rejections are completely ignored,
-        making it unsuitable for the purposes of this project.  It doesn't
-        attempt to address any of the complications in
-        \sectionref{complications} \sectionref{additional complications}
-        except for correlating by queueid.
+It also counts the number of log lines parsed and skipped.  Log lines are
+correlated based on the queueid (referred to as sessionname [sic] within
+Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
+saved when the log line doesn't match the \regex{}, though it appears that
+they will not overwrite existing data.  Most notably rejections are
+completely ignored, making it unsuitable for the purposes of this project.
+It doesn't attempt to address any of the complications in
+\sectionref{complications} \sectionref{additional complications} except for
+correlating by queueid.
 
 Logrep does not come with any documentation, though some scant
-        documentation is available on its website (\parsername{} provides
-        copious documentation).  It requires a web browser to interact with
-        it, ensuring that automated log processing will be difficult,
-        whereas automated processing is a key part of \parsernames{}
-        design.  Sadly all the author's attempts to use Logrep failed, as
-        it was unable to access the log files selected; this appears to be
-        a bug rather than operator error.\footnote{If it is caused by
-        operator error, the interface needs improvement as the (minimal)
-        instructions were followed as closely as possible, and multiple
-        attempts were made.}  As parsing failed it wasn't possible to
-        review the reports Logrep can generate (available in \HTML{} only),
-        nor to examine the (undocumented) format in which it can save
-        extracted data for subsequent reuse.
+documentation is available on its website (\parsername{} provides copious
+documentation).  It requires a web browser to interact with it, ensuring
+that automated log processing will be difficult, whereas automated
+processing is a key part of \parsernames{} design.  Sadly all the author's
+attempts to use Logrep failed, as it was unable to access the log files
+selected; this appears to be a bug rather than operator error.\footnote{If
+it is caused by operator error, the interface needs improvement as the
+(minimal) instructions were followed as closely as possible, and multiple
+attempts were made.}  As parsing failed it wasn't possible to review the
+reports Logrep can generate (available in \HTML{} only), nor to examine the
+(undocumented) format in which it can save extracted data for subsequent
+reuse.
 
 Logrep extracts far less data from Postfix logs than \parsername{},
-        completely ignores rejections, is effectively undocumented, doesn't
-        deal with the more complicated aspects of Postfix logs, and at the
-        time of writing doesn't work properly.
+completely ignores rejections, is effectively undocumented, doesn't deal
+with the more complicated aspects of Postfix logs, and at the time of
+writing doesn't work properly.
 
@@
 
-    \item [Log Mail Analyser] Please see the previous in-depth discussion
-        of Log Mail Analyser in \sectionref{prior art}.
-
-\end{description}
+\subsubsection{Log Mail Analyser}
+
+Please see the previous in-depth discussion of Log Mail Analyser in
+\sectionref{prior art}.
 

### 2008-04-10 doc/logparser.tex
Tested AWStats with my 93 log files, and wrote about it in the AWStats review.

@@
 \newcommand{\numberOFrules}[0]{%
-    161%
+    165%
+}
+
+\newcommand{\numberOFlogLINES}[0]{%
+    60,721,709%
+}
+
+\newcommand{\numberOFlogLINEShuman}[0]{%
+    60.72 million%
 }
@@
 
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@.
-
 XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
@@
 
-XXX DOWNLOAD\@: PFLOGSUMM, SAWMILL AND ANTEATER\@.
-
-CHECK PARSING FOR THE FOLLOWING\@:
+XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
 
@@
     \item Isoqlog
-    \item AWStats
     \item Log analyser --- throughput monitor
@@
 generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, 60.72 million log lines)
-were each parsed 10 times, the first run discarded, and the execution time
-for the remaining 9 runs averaged.  The first run is discarded for two
-reasons:
+\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
+log lines) were each parsed 10 times, the first run discarded, and the
+execution time for the remaining 9 runs averaged.  The first run is
+discarded for two reasons:
 
@@
 
+When tested with the \numberOFlogFILES{} test log files AWStats' report
+says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
+1,176,050 (11.29\%) corrupt lines; however there are actually
+\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
+actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
+The graphs it produces give an overview of mails received for the last
+calendar month, showing:
+
+\begin{itemize}
+
+    \item The number of mails accepted from each host.
+
+    \item How many mails were received by each recipient.
+
+    \item The average number of mails accepted by the server per day and
+        per hour.
+
+    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
+        entry in the list is: \texttt{Requested mail action not taken:
+        relaying not allowed, unknown recipient user, \ldots}
+
+\end{itemize}
+
+% Parsed lines in file: 10416129
+%  Found 4 dropped records,
+%  Found 1176050 corrupted records,
+%  Found 0 old records,
+%  Found 9240075 new qualified records.
+
 \noindent\url{http://awstats.sourceforge.net/} \newline

### 2008-04-11 doc/logparser.tex
Improved the "aborted delivery attempts" explanation. The heuristics used for detecting them still need work.

@@
 \end{abstract}
-
-XXX FIGURE OUT THE CAUSE OF THE ABORTED DELIVERY ATTEMPTS IN SECTION
-5.8.2\@.
 
@@
 
-Some mail clients appear\footnote{Due to privacy concerns no attempt has
-been made to identify either the users or the software which exhibits this
-behaviour.} to send the following sequence of commands during
-the \SMTP{} session:
+Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
+client aborts the first delivery attempt after the first recipient is
+accepted, then makes a second delivery attempt for the same recipient which
+it continues with until delivery is complete.  An example dialogue
+exhibiting this behaviour is shown presented below (lines starting with a
+three digit number are sent by the server, the other lines are sent by the
+client):
 
 \begin{verbatim}
-    EHLO client.hostname
+220 smtp.example.com ESMTP
+EHLO client.example.com
+250-smtp.example.com
+250-PIPELINING
+250-SIZE 15240000
+250-ENHANCEDSTATUSCODES
+250-8BITMIME
+250 DSN
 MAIL FROM: <sender@example.com>
-    RCPT TO: <recipient@example.com>
+250 2.1.0 Ok
+RCPT TO: <recipient@example.net>
+250 2.1.5 Ok
 RSET
+250 2.0.0 Ok
+RSET
+250 2.0.0 Ok
 MAIL FROM: <sender@example.com>
-    RCPT TO: <recipient@example.com>
+250 2.1.0 Ok
+RCPT TO: <recipient@example.net>
+250 2.1.5 Ok
 DATA
+354 End data with <CR><LF>.<CR><LF>
+The mail transfer is not shown.
+250 2.0.0 Ok: queued as 880223FA69
+QUIT
+221 2.0.0 Bye
 \end{verbatim}
-
-The client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt which it continues with
-until the delivery is complete.
 
@@
 examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message?
-Every mail which passes through Postfix will have a \daemon{cleanup} line;
-lack of a \daemon{cleanup} line is a sure sign the mail didn't make it too
-far.
+for the following: is the mail missing its \daemon{cleanup} log message,
+and are there exactly two \daemon{smtpd} log lines?  Every mail which
+passes through Postfix will have a \daemon{cleanup} line; lack of a
+\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
+two \daemon{smtpd} log lines will be a connection log line and a mail
+acceptance log line (Postfix logs acceptance as soon as the first recipient
+has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
 

### 2008-04-11 doc/logparser.tex
Improved pflogsumm section, having run it over the test log files.

@@
 
-    \item Pflogsumm
     \item Sawmill Universal Log File Analysis and Reporting
@@
 the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
+useful report, and successfully dealt with the \numberOFlogFILES{} logs
+files tested with,\footnote{The results it reported were not verified, but
+it did not report any errors, and has a very good reputation amongst
+Postfix users.} but is not a suitable base for this project.
+
+Pflogsumm's report includes the following:
+
+\begin{itemize}
+
+    \item Total number of mails accepted, delivered and rejected.  Total
+        bytes of mail accepted and delivered.
+
+    \item Total number of sender and recipient addresses and domains.
+
+    \item Per-hour averages and per-day summaries of the number of mails
+        received, delivered, deferred, bounced and rejected.
+
+    \item For received mail: per-domain totals for mails sent, deferred,
+        average delay, maximum delay and bytes delivered.  For received
+        mail: per-domain totals for mails received and bytes received.
+
+    \item Number of mails and bytes sent and received for each address.
+
+    \item Summary of why mail delivery was deferred or failed, why mails
+        were bounced, why mails were rejected, and warning messages.
+
+\end{itemize}
+
+There are many options available to include or exclude certain sections of
+the report.
 
 \url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
+2008/04/09.)
 
@@
 
-    \item The average number of mails accepted by the server per day and
-        per hour.
+    \item The average number of mails accepted by the server per-day and
+        per-hour.
 

### 2008-04-15 doc/logparser.tex
Better, tighter heuristics for detecting aborted delivery attempts.  There may be aborted delivery attempts which aren't detected by these heuristics; those should be obvious after another test run over all the log files.

@@
 examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
+for the following:
+
+\begin{itemize}
+
+    \item Are there two or more \daemon{smtpd} log lines?  Does the second
+        result have a postfix\_action attribute of ACCEPTED\@?  The first
+        two \daemon{smtpd} log lines will be a connection log line and a
+        mail acceptance log line (Postfix logs acceptance as soon as the
+        first recipient has been accepted).
+
+    \item Is \daemon{smtpd} the only Postfix component that produced a log
+        line for this mail?  Every mail which passes normally through
+        Postfix will have a \daemon{cleanup} line, and later a
+        \daemon{qmgr} log line; lack of a \daemon{cleanup} line is a sure
+        sign the mail didn't make it too far.
+
+    \item Does the queueid exist in the state tables?  If not it cannot be
+        an aborted delivery attempt.
+
+    \item If there are third and subsequent results, are all their
+        postfix\_action attributes equal to INFO\@?  If there are any log
+        lines after the first two they should be informational lines only.
+
+\end{itemize}
+
+If all the checks above are successful then the mail is assumed to be an
+aborted delivery attempt and is discarded.  There will be no further
+entries logged for such mails, so without identifying and discarding them
+they accumulate in the state table and will cause clashes if the queueid is
 reused.  The mail cannot be entered in the database as the only data
@@
 requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
+and \sectionref{results table}).  This heuristic is quite restrictive, and
+it appears there is little scope for false positives; if there are any
+false positives there will be warnings when the next log line for that mail
+is parsed.  False negatives are less likely to be detected: there may be
+queueid clashes (and warnings) if mails remain in the state tables after
+they should have been removed, otherwise the only way to detect false
+negatives is to examine the state tables after each parsing run.
 

### 2008-04-15 doc/logparser-background.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Pflogsumm
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
 \section{Background}
@@
 \end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per day and
-        per hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
-\section{Graphs}
-
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
-\end{document}

### 2008-04-15 doc/logparser-conclusion.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Pflogsumm
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
 \section{Conclusion}
@@
 \appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per day and
-        per hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
-\section{Graphs}
-
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
-\end{document}

### 2008-04-15 doc/logparser-coverage.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Pflogsumm
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
 \section{Coverage}
@@
 further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per day and
-        per hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
-\section{Graphs}
-
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
-\end{document}

### 2008-04-15 doc/logparser-database-schema.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Pflogsumm
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
 \section{Database schema}
@@
 reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per day and
-        per hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
-\section{Graphs}
-
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
-\end{document}

### 2008-04-15 doc/logparser-glossary.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Pflogsumm
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per day and
-        per hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
-\section{Graphs}
-
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
 % The contents of the glossary.
@@
 }}
-
-\end{document}

### 2008-04-15 doc/logparser-graphs.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Pflogsumm
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per day and
-        per hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
 \section{Graphs}
@@
 section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
-\end{document}

### 2008-04-15 doc/logparser-introduction.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Pflogsumm
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
 \section{Introduction}
@@
 Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per day and
-        per hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
-\section{Graphs}
-
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
-\end{document}

### 2008-04-15 doc/logparser-limitations.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Pflogsumm
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
 \section{Limitations and possible improvements}
@@
 improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per day and
-        per hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
-\section{Graphs}
-
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
-\end{document}

### 2008-04-15 doc/logparser-other-parsers-review.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
 \section{Other Postfix log parsers reviewed}
@@
 understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
-\section{Graphs}
-
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
-\end{document}

### 2008-04-15 doc/logparser-parsing-algorithm.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-15 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
 \section{Parsing algorithm}
@@
 problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with the \numberOFlogFILES{} logs
-files tested with,\footnote{The results it reported were not verified, but
-it did not report any errors, and has a very good reputation amongst
-Postfix users.} but is not a suitable base for this project.
-
-Pflogsumm's report includes the following:
-
-\begin{itemize}
-
-    \item Total number of mails accepted, delivered and rejected.  Total
-        bytes of mail accepted and delivered.
-
-    \item Total number of sender and recipient addresses and domains.
-
-    \item Per-hour averages and per-day summaries of the number of mails
-        received, delivered, deferred, bounced and rejected.
-
-    \item For received mail: per-domain totals for mails sent, deferred,
-        average delay, maximum delay and bytes delivered.  For received
-        mail: per-domain totals for mails received and bytes received.
-
-    \item Number of mails and bytes sent and received for each address.
-
-    \item Summary of why mail delivery was deferred or failed, why mails
-        were bounced, why mails were rejected, and warning messages.
-
-\end{itemize}
-
-There are many options available to include or exclude certain sections of
-the report.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2008/04/09.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per-day and
-        per-hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
-\section{Graphs}
-
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
-\end{document}

### 2008-04-15 doc/logparser-parsing-rules.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
-
-% Useful stuff for math mode.
-\usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
-% Include images
-\usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
-
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
-
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
-}
-
-%\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
-    \end{table}
-}
-
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
-% A command to format a Postfix daemon's name
-\newcommand{\daemon}[1]{%
-    \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-
-\newcommand{\tab}[0]{%
-    \hspace*{2em}%
-}
-
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
-\begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
-
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-\end{abstract}
-
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Pflogsumm
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
-
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
-
-\newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
 \section{Parsing rules}
@@
 \end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per day and
-        per hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
-
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
-\label{bibliography}
-
-\section{Graphs}
-
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
-\end{document}

### 2008-04-15 doc/logparser.tex
Split logparser.tex into one file per section - one file was too unwieldy, and latexsuite wasn't coping well with it any more.

--- 2008-04-11 doc/logparser.tex
@@
 \newpage
-\section{Introduction}
-
-\label{introduction}
-
-Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
-
-\vspace{1em}\noindent\textbf{Layout of the paper:}
-
-Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
-
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
-
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
-
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
-which complete the parser.
-
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
-
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
-
-Section~\ref{conclusion} contains the paper's conclusion, describing the
-results of the research, design and implementation of the parser.
-
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
-
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
-
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
-
-\section{Background}
-
-\label{background}
-
-\subsection{Introduction}
-
-This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
-
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
-
-\begin{enumerate}
-
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of variable complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
-                part of dealing with support calls, and the administrators
-                can have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
-
-\subsection{Simple Mail Transfer Protocol}
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{tabular}[]{ll}
-
-    request                 & smtpd\_access\_policy     \\
-    protocol\_state         & RCPT                      \\
-    protocol\_name          & SMTP                      \\
-    helo\_name              & some.domain.tld           \\
-    queue\_id               & 8045F2AB23                \\
-    sender                  & foo@bar.tld               \\
-    recipient               & bar@foo.tld               \\
-    recipient\_count        & 0                         \\
-    client\_address         & 1.2.3.4                   \\
-    client\_name            & another.domain.tld        \\
-    reverse\_client\_name   & another.domain.tld        \\
-    instance                & 123.456.7                 \\
-
-\end{tabular}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using 
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
-
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
-
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
-\subsection{Conventions used in the paper}
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
-
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
-
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it's
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
-
-\subsection{Conclusion}
-
-This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of a new parser's novel three section design.
-
-    \item Conventions used in this document.
-
-    \item A brief comparison of this parser versus other Postfix parsers.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
-
-\section{Database schema}
-\label{database schema}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered.
-
-\subsection{Introduction}
-
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
-
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-\subsection{Rules table}
-
-\label{rule attributes}
-
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
-
-Each rule defines the following:
-
-\begin{description}
-
-    \item [id] A unique identifier for each rule which other tables can use
-        to refer to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
-
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
-
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
-                delivery failure, or five day timeout.
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
-
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
-        rules.
-
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
-
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
-
-\end{description}
-
-\subsection{Connections table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{description}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
-        mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{description}
-
-\subsection{Results table}
-
-\label{results table}
-
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
-
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
-
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsection{Conclusion}
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
-
-\section{Parsing rules}
-
-\label{rules}
-
-\subsection{Introduction}
-
-This section discusses the rules used in parsing Postfix log files,
-starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% Don't reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-\end{description}
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The logs produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-\subsection{Conclusion}
-
-This section dealt with the rules used in parsing Postfix log files:
-
-\begin{itemize}
-
-    \item The characteristics of the rules were described.
-
-    \item Detecting overlapping rules and dealing with the problems they
-        can cause was covered, including a discussion of why overlapping
-        rules can be helpful as well as harmful.
-
-    \item An example log line and the rule matching it illustrated a
-        description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
-
-    \item The structure of the database table containing the rules is dealt
-        with in \sectionref{rule attributes}, and is not duplicated in this
-        section.
-
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
-
-    \item This section finished with an explanation of the algorithm used
-        to generate a new \regex{} from unparsed log lines.
-
-\end{itemize}
-
-\section{Parsing algorithm}
-
-\label{parsing-algorithm}
-
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsection{Introduction}
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications which have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsection{A high level overview}
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, log all data and finish.
-
-    \item Follow the progress of the accepted mail until it's either
-        delivered, bounced or deleted, then log all data, and finish.
-
-\end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsection{Complications encountered}
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should probably be two different entires in the database resulting
-from the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
-
-\subsection{Flow chart}
-
-\label{flow-chart}
-
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the logs
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\subsection{Actions in detail}
-
-\label{actions-in-detail}
-
-Each action is passed the same arguments:
-
-\begin{description}
-
-    \item [line] The log line, separated into fields:
-
-        \begin{description}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
-
-            \item [text] The remainder of the line.
-
-        \end{description}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The fields in the line captured by the rule's \regex{}.
-
-\end{description}
-
-The actions:
-
-\begin{description}
-
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database. Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
-        mail.
-
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
-
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
-
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
-
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded. The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary. TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
-
-\end{description}
-
-\subsection{Adding new actions}
-
-\label{adding new actions}
-
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
-
-\subsection{Additional complications}
-
-\label{additional complications}
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
-
-\subsubsection{Identifying bounce notifications}
-
-\label{identifying-bounce-notifications}
-
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is $<>$.
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
-
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be $<>$.
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsubsection{Aborted delivery attempts}
-
-\label{aborted-delivery-attempts}
-
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
-client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following: is the mail missing its \daemon{cleanup} log message,
-and are there exactly two \daemon{smtpd} log lines?  Every mail which
-passes through Postfix will have a \daemon{cleanup} line; lack of a
-\daemon{cleanup} line is a sure sign the mail didn't make it too far.  The
-two \daemon{smtpd} log lines will be a connection log line and a mail
-acceptance log line (Postfix logs acceptance as soon as the first recipient
-has been accepted); XXX IMPROVE THIS ONCE I HAVE BETTER HEURISTICS\@.
-
-If this check is successful then the mail is assumed to be one of the
-offending bogus mails and is discarded.  There will be no further entries
-logged for such mails, so without identifying and discarding them they
-accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is weak, but thus far it
-has not caused any difficulties, and there does not seem to be another
-check available to extend the heuristics with.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsubsection{Further aborted delivery attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
-
-\subsubsection{Timeouts during DATA phase}
-
-\label{timeouts-during-data-phase}
-
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
-
-\end{enumerate}
-
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
-
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
-
-\subsubsection{Pickup logging after cleanup}
-
-\label{pickup logging after cleanup}
-
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
-
-This complication is dealt with during the PICKUP action.
-
-\subsubsection{Smtpd stops logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-\end{enumerate}
-
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
-
-\subsubsection{Out of order log lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The solution to this problem is to examine the list of programs which have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.
-
-\subsubsection{Yet more aborted delivery attempts}
-
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
-
-\subsubsection{Mail deleted before delivery is attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
-typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
-
-This is yet another complication which only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsubsection{Mails deleted during delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
-
-\subsection{Conclusion}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
-\section{Coverage}
-
-\label{parsing coverage}
-
-\subsection{Introduction}
-
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-\subsection{Log lines covered}
-
-\label{log-lines-covered}
-
-Parsing a log line is a three stage process:
-
-\begin{enumerate}
-
-    \item Check if there are any rules for the Postfix component which
-        produced the log line; if not then skip the log line.
-
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
-
-    \item Execute the action specified by the rule.
-
-\end{enumerate}
-
-Full coverage of log lines requires the following:
-
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
-
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
-other log files).
-
-\subsection{Mails covered}
-
-\label{mails-covered}
-
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
-
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
-
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
-
-\begin{itemize}
-
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
-
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
-
-    \item 9 mails whose initial log entries were not present in the logs.
-
-    \item 7 mails which had yet to be delivered due to repeated failures.
-
-\end{itemize}
-
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
-
-\begin{itemize}
-
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
-
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
-
-    \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
-        with to be acceptable.
-
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
-
-\end{itemize}
-
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
-
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
-
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
-
-\subsection{Conclusion}
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
-
-\section{Limitations and possible improvements}
-
-\label{limitations-improvements}
-
-\subsection{Introduction}
-
-Every piece of software suffers from some limitations and there is almost
-always room for improvement.
-
-\subsection{Limitations}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The program will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
-
-    \item The parser does not distinguish between logs produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible improvements}
-
-\begin{itemize}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
-
-\end{itemize}
-
-\subsection{Conclusion}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
-\section{Conclusion}
-
-\label{conclusion}
-
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
-
-Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
-
-The implementation not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
-
-The division of the parser into rules, actions and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
-
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
-
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
-
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
-\newpage
-
-\appendix
-
-\section{Other Postfix log parsers reviewed}
-
-\label{other-parsers}
-
-\subsection{Introduction}
-
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
-
-\end{enumerate}
-
-\subsection{Parsers reviewed}
-
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
-
-\subsubsection{Pflogsumm}
-
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
-
-Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with all the Postfix logs tested
-with, but is not a suitable base for this project.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2007/08/13.)
-
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
-
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
-
-Sawmill is a general purpose commercial product which parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
-described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
-
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
-
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
-
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
-
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
-
-(Last checked 2007/12/04.)
-
-\subsubsection{Splunk}
-
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
-
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
-
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
-
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
-
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
-
-\subsubsection{Isoqlog}
-
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
-
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{AWStats}
-
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
-information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
-
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
-described in this document.
-
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
-The graphs it produces give an overview of mails received for the last
-calendar month, showing:
-
-\begin{itemize}
-
-    \item The number of mails accepted from each host.
-
-    \item How many mails were received by each recipient.
-
-    \item The average number of mails accepted by the server per day and
-        per hour.
-
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
-
-\end{itemize}
-
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
-
-\subsubsection{Log analyser --- throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
-
-\subsubsection{Anteater}
-
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
-
-Anteater doesn't have any English documentation so it's difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
-
-\subsubsection{Yet Another Advanced Logfile Analyser}
-
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
-
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
-
-YAALA provides a Postfix parser which extracts the following fields from
-specific log lines:
-
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
-
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
-
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
-
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logparser/Lire}
-
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
-
-Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
-
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
-
-\subsubsection{Logrep}
-
-\textit{Logrep is a secure multi-platform framework for the collection,
-extraction, and presentation of information from various log files. It
-features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
-
-Logrep extracts roughly half the fields \parsername{} does:
-
-\begin{itemize}
-
-    \item For mail sent and received: from address, size, and time and
-        date.
-
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
-
-    \item For mail received: the hostname of the sender.
-
-\end{itemize}
-
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
-they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
-
-Logrep does not come with any documentation, though some scant
-documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
-
-Logrep extracts far less data from Postfix logs than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
-writing doesn't work properly.
-
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
-\subsubsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
-\subsection{Conclusion}
-
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
+\input{logparser-introduction.tex}
+\input{logparser-background.tex}
+\input{logparser-database-schema.tex}
+\input{logparser-parsing-rules.tex}
+\input{logparser-parsing-algorithm.tex}
+\input{logparser-coverage.tex}
+\input{logparser-limitations.tex}
+\input{logparser-conclusion.tex}
+\input{logparser-other-parsers-review.tex}
 
@@
 
-\section{Graphs}
+\input{logparser-graphs.tex}
 
-\label{graphs}
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-\renewcommand{\figurename}{Graph}
-
-\subsection{Introduction}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-Graph~\refwithpage{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
-
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
-
-\subsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsection{Rule hits}
-\label{rule hits}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage
-
-\subsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\clearpage
-
-\subsection{Conclusion}
-
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
-
-% This fools the syntax highlighting so we get a new fold here, but it
-% doesn't create a new section.
-\ifacronymfirstuse{SMTP}{
-\section{Glossary}
-}{}
 % Redefine the command used to produce the glossary title, because the
@@
 
-% The contents of the glossary.
-\glossary{name={SQLite3},description={
-    \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
-}}
-
-\glossary{name={Phishing},description={
-    Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
-    by masquerading as an entity trusted by the user, e.g.\ a bank.
-}}
-
-\glossary{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
-}}
-
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
-}}
-
-\glossary{name={Epoch},description={
-    Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
-    different.
-}}
-
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
-}}
-
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
-\glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
-}}
-
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
-
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
-}}
-
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
-}}
-
+\input{logparser-glossary.tex}
+% logparser-acronyms.tex has been included in the preamble so that the
+% commands are available throughout the text.
 \end{document}

### 2008-04-29 doc/logparser-other-parsers-review.tex
Tested Splunk.

@@
 automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database.  Many types of reports
-are available, though most are variations of a bar or pie chart, with the
-exception of bubble and heatmap graphs.  It is very easy to drill down
+independent tools to utilise the Splunk database; there are additional
+reports available on \url{http://www.splunkbase.com/}, but the author was
+unable to find any --- every category was empty, even those which the
+interface claimed had reports available.  Many types of reports are bundled
+with the software, though most are variations of a bar or pie chart, with
+the exception of bubble and heatmap graphs.  It is very easy to drill down
 through the graphs to extract a portion of the data (e.g.\ select the hour
@@
 finally a specific address), though it is not possible to search on partial
-words.  The searches are quite fast, though as the free version has a limit
-on the amount of data indexed per day the volume of data to be searched was
-merely 45MB\@; the cheapest licensed version costs \$6000, and still limits
-the volume of data indexed per day.
+words.
+
+The web interface is quite attractive and easy to use when searching, but
+as an administrator it seems unnecessarily difficult to perform simple
+tasks.  When testing it took roughly 30 minutes to add a single log file to
+be indexed (log files must be indexed before searching), with the downside
+that the log file was copied into a spool directory, doubling the disk
+space usage.  The most suitable test would be to index all log files in a
+particular directory, but after three hours, numerous futile attempts, and
+reading all available documentation the author admitted defeat.  Using the
+\CLI{} was more successful: invoking the command \newline \tab{}
+\texttt{splunk find logs }\textit{log-directory\/}\newline added 40 of the
+\numberOFlogFILES{} log files to the queue for indexing.  Repeated attempts
+enqueued the same 40 log files, without explanation as to why the others
+were excluded.\footnote{The log files appear to have only been indexed
+once; presumably Splunk keeps track of the files it has indexed and
+discards requests to index files for a second time.  This may or may not be
+a useful feature for \parsername{}.} There did not appear to be an option
+to order the log files to ensure they would be processed in the order they
+were created.  Subsequently the author was successful in adding a single
+file at a time using the \CLI{}; a simple loop to add all desired log files
+was sufficient to index all files.  Splunk will periodically check all
+those files for updates unless they are manually removed from its list;
+this may or may not be useful behaviour.  Splunk did not appear to have any
+difficulty in parsing the log files, once instructed to do so.  All
+searches performed using the indexed data returned reasonable results.
+
+Splunk supports reading compressed log files without any configuration by
+the user.  The free version of Splunk limits the volume of data indexed per
+day to 500MB, though a trial Enterprise licence is available which allows
+indexing of up to 5GB of data per day.  The cheapest licensed version costs
+\$5000 (plus \$1000 support), and still limits the volume of data indexed
+per day to 500MB.
 
@@
 occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all text
-in the line is also available for searching.  \parsername{} extracts
+listed above.  Searches can be based on the extracted fields, and all the
+text in the line is also available for searching.  \parsername{} extracts
 noticeably more data, though it does not make the full text of the line
@@
 no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.
+of the myriad complications discussed in this document.  Some additional
+Postfix reports are supposedly available at
+\url{http://www.splunkbase.com/}, but the author was unable to find them
+(or any other reports).
 
-\url{http://www.splunk.com/} \newline (Last checked 2007/09/25.)
+\url{http://www.splunk.com/} \newline (Last checked 2008/04/29.)
 

### 2008-04-29 doc/logparser-other-parsers-review.tex
Tested Isoqlog and Anteater. Didn't test "Log analyser --- throughput monitor" - it would be pointless.

@@
 this project, and ignores the majority of log lines produced by Postfix.
+It ignores all log lines earlier than today, so testing with the test log
+files is pointless.
 
@@
 has documentation the author can read a detailed review cannot be provided.
+
+The Debian project (\url{http://www.debian.org/}) provides a manual page
+with the copy of anteater it distributes, so the author was at least able
+to run anteater with the correct arguments; sadly anteater produced zero
+for every statistic, presumably because it was unsuccessful in parsing the
+log lines.
 

### 2008-04-29 doc/logparser-other-parsers-review.tex
Tested lire. Logrep can't be tested, it doesn't work properly.

@@
 \parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
+as easy as possible.  When testing Lire on the \numberOFlogFILES{} test log
+files it performed reasonably well, producing summaries of:
 
-\url{http://logreport.org/lire.html} \newline (Last checked 2007/10/09.)
+\begin{itemize}
+
+    \item Delivery status and failed deliveries.
+
+    \item Sender and recipient domains and servers.
+
+    \item Number of deliveries and bytes per day and hour.
+
+    \item Recipients by domain.
+
+    \item Deliveries by relays, by size and by delay.
+
+    \item Delays by server and domain.
+
+    \item Most mails exchanged between a pair of correspondents.
+
+\end{itemize}
+
+The numbers it reports appear reasonable, and the subset verified by the
+author were correct.  Lire produces a report with less detail than
+Pflogsumm, and is considerable harder to configure.
+
+\url{http://logreport.org/lire.html} \newline (Last checked 2008/04/29.)
 

### 2008-04-30 doc/logparser-other-parsers-review.tex
Improved other parsers introduction. Sawmill tested. YAALA tested.

@@
         simple examples are given in \sectionref{motivation}.
+
+    \item Most of the parsers silently ignore log lines they cannot handle,
+        whereas \parsername{} complains loudly about every single line it
+        fails to parse.  The exception is AWStats, which outputs the
+        percentage of input lines it was unable to parse, but does not
+        output the lines themselves.
+
+    \item A minor difference is that most parsers don't handle compressed
+        files; both \parsername{} and Splunk handle them transparently,
+        without user intervention; Sawmill and Lire can be configured to
+        support compressed files, but Sawmill exhibits a dramatic slow down
+        when thus configured.  Although this is a minor difference, support
+        for reading compressed log files is quite helpful, as it
+        dramatically reduces the disk space required to store old logs.
 
@@
 rejections except when the rejection is caused by an \RBL{} check.
+
+When tested with the \numberOFlogFILES{} test log files it performed
+adequately, though the rate it processed log files at did slow down
+noticeably as it progressed.  Sawmill supports reading compressed log files
+but it exhibits a dramatic slow down when doing so: it took six hours to
+parse the first half of the log files; twelve hours to parse the next
+third; after twenty four hours parsing the remaining sixth it crashed due
+to lack of disk space.  On the second parsing attempt the log files were
+uncompressed beforehand and parsing took eight hours.
+
+Sawmill's web interface supports searching on any combination of the fields
+it extracts, but the interface is neither as simple to use nor as
+informative as the interface provided by Splunk.  The administrative
+interface is very easy to use --- it took only five minutes to start
+parsing a directory of log files.
 
@@
 \newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.
+data for external use, merely to create reports.  Only data selected by the
+query will be saved in the data store; other data will be discarded, and
+removed from the data store if already present.
 
-In summary YAALA provides a Postfix parser which only handles the most
-common Postfix log lines, provides reasonably flexible report generation
-from the limited data extracted, but has no facilities to extract data for
-use in other tools.
+Testing YAALA was unsuccessful because all the select clauses tried
+produced a similar error message:
+\newline\tab{}\texttt{lib/Yaala/Data/Core.pm: Unavailable aggregation
+requested:} \newline\tab{}\texttt{``bytes''. Returning 0.} \newline  The
+underlying reason for this is that YAALA only parsed 408 log lines of
+360632 log lines (0.11\%) in the first log file.
+
+In summary YAALA provides a Postfix parser which only attempts to parse the
+most common Postfix log lines (but fails miserably), provides reasonably
+flexible report generation from the limited data extracted, but has no
+facilities to extract data for use in other tools.
 

### 2008-05-01 doc/logparser-other-parsers-review.tex
Added summary table to other parsers' conclusion. Some more details about AWStats. Started testing LMA.

@@
 file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline
-There does not seem to be an explanation of any of those fields in the
+email, email\_r, host, host\_r, method, url, code and bytesd.\newline There
+does not seem to be an explanation of any of those fields in the
 documentation (\parsername{} provides copious documentation).  AWStats has
@@
 lines by queueid, it does not deal with any of the other complications
-described in this document.
+described in this document.  AWStats supports saving data but the format of
+the saved data is not documented (as far as the author could tell).  It
+also supports reading compressed log files, but that functionality was not
+tested.
 
@@
 1,176,050 (11.29\%) corrupt lines; however there are actually
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files.  AWStats
-actually parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
+\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files, so AWStats
+only parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
 The graphs it produces give an overview of mails received for the last
@@
 
+\begin{table}[ht]
+\begin{tabular}{l|l|l|l|l}
+    Parser              & Parsed     & Data        & Custom        & Compressed \\
+                        & test logs? & store?      & reports?      & logs       \\
+    \hline                                                                      \\
+    Pflogsumm           & Yes        & No          & No            & No         \\
+    Sawmill             & Yes        & Yes         & Searches      & Partially  \\
+    Splunk              & Yes        & Yes         & Searches      & Yes        \\
+                        &            &             & \& reports    &            \\
+    Isoqlog             & No         & Yes         & No            & No         \\
+    AWStats             & Partially  & Yes         & Partially\dag & Yes        \\
+    Throughput monitor  & N/A        & No          & No            & No         \\
+    Anteater            & No         & No          & No            & No         \\
+    YAALA               & No         & Yes\ddag    & Yes           & No         \\
+    Logparser/Lire      & Yes        & Yes         & Yes           & Yes        \\
+    Logrep              & No         & Yes         & No            & No         \\
+    Log Mail Analyser   & Yes        & Yes         & Partially     & No         \\ 
+    \parsername{}       & Yes        & Yes\dag\dag & No            & Yes        \\
+\end{tabular}
+
+\dag AWStats can be configured to omit a section from a report, but not to
+add extra sections.
+
+\ddag YAALA only stores the data required to produce the latest report;
+other data will be discarded.
+
+\dag\dag\parsername{} is the only parser with documentation for its data store.
+
+\caption{Summary of parsers' features}
+\label{Summary of parsers' features}
+\end{table}
+
 While there are other programs available which perform basic Postfix log
@@
 offered by \parsername{}; reporting is deferred to a subsequent
-application.
+application.  Table \refwithpage{Summary of parsers' features} provides a
+summary of the parsers' features.
 

### 2008-05-01 doc/logparser-background.tex
Added summary table to other parsers' conclusion. Some more details about AWStats. Started testing LMA.
Updates after checking results from LMA.

@@
 possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.
+by \LMA{}.  Once the preprocessing has been performed on the
+\numberOFlogFILES{} test log files \LMA{} parses the log files without
+complaint, though it does produce 32 entries in its output file for every
+rejection in the input log file; it also missed some 40\% of delivered
+mail.  Once these glaring deficiencies were discovered the author did not
+spend more time checking the results.
 

### 2008-05-01 doc/logparser-other-parsers-review.tex
Updates after checking results from LMA.

@@
     Logrep              & No         & Yes         & No            & No         \\
-    Log Mail Analyser   & Yes        & Yes         & Partially     & No         \\ 
+    Log Mail Analyser   & No         & Yes         & Partially     & No         \\ 
     \parsername{}       & Yes        & Yes\dag\dag & No            & Yes        \\

### 2008-05-01 doc/logparser-other-parsers-review.tex
Added cluster_groups to the rules table description and example rule. Improved Splunk review. Improved summary table in review of other parsers conclusion. Minor improvements in parser reviews. Improved review of other parsers conclusion.

@@
 reports available on \url{http://www.splunkbase.com/}, but the author was
-unable to find any --- every category was empty, even those which the
+unable to find any: every category was empty, even those which the
 interface claimed had reports available.  Many types of reports are bundled
@@
 searches performed using the indexed data returned reasonable results.
+
+Copious documentation is made available on \url{http://www.splunk.com/},
+but poor organisation and sheer abundance makes it extremely hard to find
+useful information.  Searching using the website's interface confusingly
+tends to return results from old documentation rather than new.  In
+general the documentation appears to have been written by someone
+intimately acquainted with the software and having difficulty understanding
+how a newcomer would approach tasks or the questions they would ask.
 
@@
 
-\subsubsection{Log analyser --- throughput monitor}
+\subsubsection{Throughput monitor}
 
@@
 \url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2007/08/13.)
+checked 2008/05/01.)
 
@@
 \parsername{} uses a well documented database which is accessible from the
-majority of programming languages.
+majority of programming languages.  This information was gleaned from the
+source code, as the documentation is sadly lacking.
 
@@
 
-\subsubsection{Logparser/Lire}
+\subsubsection{Lire}
 
@@
 \begin{table}[ht]
-\begin{tabular}{l|l|l|l|l}
-    Parser              & Parsed     & Data        & Custom        & Compressed \\
-                        & test logs? & store?      & reports?      & logs       \\
-    \hline                                                                      \\
-    Pflogsumm           & Yes        & No          & No            & No         \\
-    Sawmill             & Yes        & Yes         & Searches      & Partially  \\
-    Splunk              & Yes        & Yes         & Searches      & Yes        \\
-                        &            &             & \& reports    &            \\
-    Isoqlog             & No         & Yes         & No            & No         \\
-    AWStats             & Partially  & Yes         & Partially\dag & Yes        \\
-    Throughput monitor  & N/A        & No          & No            & No         \\
-    Anteater            & No         & No          & No            & No         \\
-    YAALA               & No         & Yes\ddag    & Yes           & No         \\
-    Logparser/Lire      & Yes        & Yes         & Yes           & Yes        \\
-    Logrep              & No         & Yes         & No            & No         \\
-    Log Mail Analyser   & No         & Yes         & Partially     & No         \\ 
-    \parsername{}       & Yes        & Yes\dag\dag & No            & Yes        \\
+\begin{tabular}{l|l|l|l|l|l}
+    Parser         & Parsed     & Data        & Custom        & Documentation? & Source \\
+                   & test logs? & store?      & reports?      &                & code?  \\
+    \hline         &            &             &               &                &        \\
+    Pflogsumm      & Yes        & No          & Partial\dag   & Good           & Yes    \\
+    Sawmill        & Yes        & Yes         & Searches      & Reasonable     & No     \\
+    Splunk         & Yes        & Yes         & Searches      & Abundant       & No     \\
+                   &            &             & \& reports    & but poor       &        \\
+    Isoqlog        & No         & Yes         & No            & No             & Yes    \\
+    AWStats        & Partially  & Yes         & Partial\dag   & Good           & Yes    \\
+    Throughput     & N/A        & No          & No            & Brief but      & Yes    \\
+    \tab{}monitor  &            &             &               & good           &        \\
+    Anteater       & No         & No          & No            & No             & Yes    \\
+    YAALA          & No         & Yes\ddag    & Searches      & Poor           & Yes    \\
+    Lire           & Yes        & Yes         & Yes           & Yes            & Yes    \\
+    Logrep         & No         & Yes         & No            & Barely         & Yes    \\
+    \LMA{}         & No         & Yes         & No            & No             & Yes    \\ 
+    \parsername{}  & Yes        & Yes\dag\dag & No            & Yes            & Yes    \\
 \end{tabular}
 
-\dag AWStats can be configured to omit a section from a report, but not to
-add extra sections.
+\dag Sections can be omitted from a report, but extra sections can not be
+added.
 
@@
 data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  All of the parsers reviewed can
-produce a report of greater or lesser detail and complexity, a facility not
-offered by \parsername{}; reporting is deferred to a subsequent
-application.  Table \refwithpage{Summary of parsers' features} provides a
-summary of the parsers' features.
+in a format inaccessible to other tools.  Nearly all of the parsers
+reviewed can produce a report of greater or lesser detail and complexity, a
+facility not offered by \parsername{}; reporting is deferred to a
+subsequent application.  The quality of the documentation offered by the
+subset of parsers which provide some varies from unusable to good; none of
+the parsers provide any documentation on the format of their data stores
+(if any).  Less than half of the parsers were capable of parsing the
+\numberOFlogFILES{} test log files, and improving or extending the parsers
+would have been quite a difficult task for any of the parsers.  Table
+\refwithpage{Summary of parsers' features} provides a summary of the
+parsers' features.
 

### 2008-05-01 doc/logparser-parsing-rules.tex
Added cluster_groups to the rules table description and example rule. Improved Splunk review. Improved summary table in review of other parsers conclusion. Minor improvements in parser reviews. Improved review of other parsers conclusion.

@@
 priority            & 0                                                 \\
+cluster\_group      & 400                                               \\
 
@@
         but is otherwise unused by the algorithm.
+
+    \item [cluster\_group] The cluster\_group attribute is used by the
+        Decision Tree algorithm described in a separate document; the
+        parser does not use it in any way.
 

### 2008-05-03 doc/logparser-parsing-algorithm.tex
Mention that Outlook makes aborted delivery attempts.

@@
 accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  An example dialogue
-exhibiting this behaviour is shown presented below (lines starting with a
-three digit number are sent by the server, the other lines are sent by the
+it continues with until delivery is complete.\footnote{Microsoft Outlook is
+one client which behaves in this fashion; no attempt has been made to
+identify other clients which act in a similar way.}  An example dialogue
+exhibiting this behaviour is presented below (lines starting with a three
+digit number are sent by the server, the other lines are sent by the
 client):

### 2008-05-13 papers/aics-2008/paper.tex
Brief outline for Carl to look at.

--- 2008-04-15 doc/logparser.tex
@@
-% $Id$
-\documentclass[a4paper,12pt,draft]{article}
+% $Id: logparser.tex 545 2008-05-01 20:00:22Z tobinjt $
+\documentclass[twocolumn,a4paper,11pt,draft]{article}
 
@@
 
-\newcommand{\numberOFlogFILES}[0]{%
-    93%
-}
-
-\newcommand{\numberOFrules}[0]{%
-    165%
-}
-
-\newcommand{\numberOFlogLINES}[0]{%
-    60,721,709%
-}
-
-\newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
-}
-
 % This may keep hyperref happy about the Abstract entry in the table of
@@
 
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-\title{Parsing Postfix log files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
+\title{Using Decision Trees for optimisation rather than classification
+XXX BETTER TITLE}
+\author{John Tobin \& Carl Vogel \\ School of Computer Science and Statistics \\
+Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie \&
+vogel@cs.tcd.ie }
 \maketitle
@@
 
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
+    Classification performed by evaluating user-defined sequences of tests
+    can be optimised by applying a Decision Tree algorithm to data gathered
+    from running the tests, providing the user with a more efficient or
+    more accurate sequence of tests.  This usage of Decision Trees is
+    illustrated by optimising Postfix~\cite{postfix} anti-spam
+    restrictions~\cite{smtpd_access_readme}, a task complicated by the
+    division of the tests into groups, restricting the Decision Tree
+    algorithm to examining subsets of attributes at a time.
 
@@
 
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
+\section{Introduction}
 
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
+Motivation, background on Postfix restrictions (appendix maybe?)
 
-\begin{enumerate}
+\section{Algorithm}
 
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
+Standard CART Decision Tree with modifications:
 
-\end{enumerate}
+\begin{itemize}
 
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
+    \item groups of attributes to examine
 
-\newpage
-\input{logparser-introduction.tex}
-\input{logparser-background.tex}
-\input{logparser-database-schema.tex}
-\input{logparser-parsing-rules.tex}
-\input{logparser-parsing-algorithm.tex}
-\input{logparser-coverage.tex}
-\input{logparser-limitations.tex}
-\input{logparser-conclusion.tex}
-\input{logparser-other-parsers-review.tex}
+    \item result already known --- non-rejections will be weeded out into
+        the false branches, eventually ending up in the bottom-left-most
+        leaf node.
 
-\bibliographystyle{logparser-bibliography-style}
-\bibliography{logparser-bibliography}
+    \item scoring functions
+
+\end{itemize}
+
+\section{Results}
+
+\begin{itemize}
+
+    \item duplicate rows collapsed; histograms of distribution
+
+    \item optimised ordering --- primary component analysis
+
+    \item image
+
+    \item different trees from different scoring functions
+
+\end{itemize}
+
+\section{Conclusion}
+
+Useful for optimisation and visualisation.
+
+\appendix
+
+\bibliographystyle{../common/bibliography-style}
+\bibliography{../common/bibliography}
 \label{bibliography}
 
-\input{logparser-graphs.tex}
-
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
-
-\input{logparser-glossary.tex}
-% logparser-acronyms.tex has been included in the preamble so that the
-% commands are available throughout the text.
 \end{document}

### 2008-05-14 papers/sgai-2008/paper.tex
Skeleton for Carl to read.

--- 2008-05-13 papers/aics-2008/paper.tex
@@
 
-\title{Using Decision Trees for optimisation rather than classification
-XXX BETTER TITLE}
+\title{A user-extensible and adaptable parser architecture}
 \author{John Tobin \& Carl Vogel \\ School of Computer Science and Statistics \\
@@
 
-    Classification performed by evaluating user-defined sequences of tests
-    can be optimised by applying a Decision Tree algorithm to data gathered
-    from running the tests, providing the user with a more efficient or
-    more accurate sequence of tests.  This usage of Decision Trees is
-    illustrated by optimising Postfix~\cite{postfix} anti-spam
-    restrictions~\cite{smtpd_access_readme}, a task complicated by the
-    division of the tests into groups, restricting the Decision Tree
-    algorithm to examining subsets of attributes at a time.
+    Some parsers need to be very precise and restrictive in what they
+    parse, yet allow users to easily adapt existing rules or add new rules
+    to parse new inputs, without requiring an in-depth knowledge of the
+    parser's internal workings.  This paper presents a novel parsing
+    architecture which aims to make the process of parsing new inputs as
+    simple as possible, enabling users to trivially add new rules (to parse
+    variants of existing inputs) and relatively easily add new actions (to
+    do something useful with a previously unknown class of input).
 
@@
 
-Motivation, background on Postfix restrictions (appendix maybe?)
+Background of the project --- emphasise the variety in the log lines.
+Motivation for creating the parser.  At some point I want to compare this
+architecture to recursive decent parsers, Lex/Yacc parsers, XXX parsers.
 
@@
 
-Standard CART Decision Tree with modifications:
-
-\begin{itemize}
-
-    \item groups of attributes to examine
-
-    \item result already known --- non-rejections will be weeded out into
-        the false branches, eventually ending up in the bottom-left-most
-        leaf node.
-
-    \item scoring functions
-
-\end{itemize}
+Split into framework, actions, \& rules.  Explain adding new rules and
+actions.  Cut down rules to the minimum possible.  Framework provides
+support functions, parsing loop, manages data --- boring stuff users don't
+need to concern themselves with.
 
@@
 
-\begin{itemize}
-
-    \item duplicate rows collapsed; histograms of distribution
-
-    \item optimised ordering --- primary component analysis
-
-    \item image
-
-    \item different trees from different scoring functions
-
-\end{itemize}
+Scalability, ordering, caching regexes.  Interesting to run with all rules
+versus running with the minimum number necessary to parse the test log
+files.  Then try that again with different orderings?
 
@@
 
-Useful for optimisation and visualisation.
+Makes adding new rules trivial, new actions tractable.  Knowledge of the
+framework is rarely necessary.
 
 \appendix
+
+\cite{postfix}
 

### 2008-05-20 doc/logparser-background.tex
:%s/logs/log files/gc

@@
 are the conventions used in this paper.  Other projects which attempt to
-parse Postfix logs are summarised (full details are available in
+parse Postfix log files are summarised (full details are available in
 appendix~\refwithpage{other-parsers}), finishing with a review of
@@
 evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix logs and populates a database with the data gleaned
-from those logs, providing a consistent and simple view of the logs which
-future tools can utilise.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.
+program parses Postfix log files and populates a database with the data
+gleaned from those log files, providing a consistent and simple view of the
+log files which future tools can utilise.  The gathered data can then be
+used to optimise current anti-spam measures, provide a baseline to test new
+anti-spam measures against, or to produce statistics showing how effective
+those measures are.
 
@@
 The database populated by this program provides a simple interface to
-Postfix logs.  Although the interface is a database schema, it is in effect
-quite similar to any other \API{} provided by shared code: it insulates
-both user and provider of the \API{} from changes in the implementation of
-the \API{}\@.  The algorithm implemented by the parser can be improved;
-support can be added for earlier or later releases of Postfix; bugs can be
-fixed or limitations removed from the parser; these changes will not cause
-the user to be negatively impacted.  Statistics and/or graphs can be
-generated from the database; new restrictions can be tested and the results
-inspected; trends in the fight against spam can emerge from historical data
-saved in the database; the parser remains the same as the usage adapts.
-Using a database simplifies writing programs which need to interact with
-the data in several ways:
+Postfix log files.  Although the interface is a database schema, it is in
+effect quite similar to any other \API{} provided by shared code: it
+insulates both user and provider of the \API{} from changes in the
+implementation of the \API{}\@.  The algorithm implemented by the parser
+can be improved; support can be added for earlier or later releases of
+Postfix; bugs can be fixed or limitations removed from the parser; these
+changes will not cause the user to be negatively impacted.  Statistics
+and/or graphs can be generated from the database; new restrictions can be
+tested and the results inspected; trends in the fight against spam can
+emerge from historical data saved in the database; the parser remains the
+same as the usage adapts.  Using a database simplifies writing programs
+which need to interact with the data in several ways:
 
@@
 Storing the results in a database will also increase the efficiency of
-using those results, as the logs need only be parsed once; indeed the
-results may be used by someone with no access to the original logs.
+using those results, as the log files need only be parsed once; indeed the
+results may be used by someone with no access to the original log files.
 
@@
 mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the logs generated.  While it is easy to use standard Unix text
+in the log files generated.  While it is easy to use standard Unix text
 processing utilities to determine the fate of an individual email,
 following the journey an email takes through Postfix can be quite
-difficult.  The logs tend to follow a 90\%-10\% pattern: 90\% of the time
-the journey is simple, but the other 10\% of the time requires 90\% of the
-code.\footnote{These numbers don't have a solid scientific basis, they're
-based on gut feeling from writing and debugging the software.}
+difficult.  The log files tend to follow a 90\%-10\% pattern: 90\% of the
+time the journey is simple, but the other 10\% of the time requires 90\% of
+the code.\footnote{These numbers don't have a solid scientific basis,
+they're based on gut feeling from writing and debugging the software.}
 
@@
         final restriction a mail must overcome in the configuration used on
-        the mail server the logs were obtained from; an administrator is
-        free to use Greylisting at whichever position in the restriction
+        the mail server the log files were obtained from; an administrator
+        is free to use Greylisting at whichever position in the restriction
         list they feel is most appropriate for their mail system.}
@@
 
-    \item The logs are whole and complete: nothing has been removed, either
-        deliberately or accidentally (e.g.\ log rotation gone awry, file
-        system filling up, logging system unable to cope with the volume of
-        logs).  On a well run system it is extremely unlikely that any of
-        these problems will arise, though it is of course possible,
-        particularly when suffering a deluge of spam or a mail loop.
+    \item The log files are whole and complete: nothing has been removed,
+        either deliberately or accidentally (e.g.\ log rotation gone awry,
+        file system filling up, logging system unable to cope with the
+        volume of log messages).  On a well run system it is extremely
+        unlikely that any of these problems will arise, though it is of
+        course possible, particularly when suffering a deluge of spam or a
+        mail loop.
 
@@
 
-    \item The parser is developed using real world logs rather than the
-        idealised logs someone would naturally envisage reading the source
-        code.
+    \item The parser is developed using real world log files rather than
+        the idealised log files someone would naturally envisage reading
+        the source code.
 
@@
 It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that logs produced by Postfix are not fixed; they vary widely
-from host to host, depending on the set of restrictions chosen by the
-administrator.  With this in mind, one of the parser's design aims was to
-make adding new rules as easy as possible, to enable administrators to
+background}) that log files produced by Postfix are not fixed; they vary
+widely from host to host, depending on the set of restrictions chosen by
+the administrator.  With this in mind, one of the parser's design aims was
+to make adding new rules as easy as possible, to enable administrators to
 properly parse their log lines.  To enable this the parser is divided into
@@
 
-    \item [actions] Actions invoked to deal with a log line once it has been
-        identified by the rules: modifying data structures, handing
+    \item [actions] Actions invoked to deal with a log line once it has
+        been identified by the rules: modifying data structures, handing
         complications, and saving data to the database.  The actions
@@
 
-    \item [framework] The framework is responsible for loading rules, managing data
-        structures so actions can retrieve them, finding the rule which
-        matches each log line, invoking the correct action, etc.  The
-        framework provides the structure which actions and rules plug into.
-        The framework is not described in detail in this document because
-        it is not particularly interesting, but there is some coverage in
-        \sectionref{framework}.
+    \item [framework] The framework is responsible for loading rules,
+        managing data structures so actions can retrieve them, finding the
+        rule which matches each log line, invoking the correct action, etc.
+        The framework provides the structure which actions and rules plug
+        into.  The framework is not described in detail in this document
+        because it is not particularly interesting, but there is some
+        coverage in \sectionref{framework}.
 
@@
 files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; logs from Solaris hosts (and
+differ from those added on Linux hosts; log files from Solaris hosts (and
 possibly other operating systems) thus require preprocessing before parsing
@@
 
-    \item Some records have an \SMTP{} code of \textit{0}, though this is not a
-        valid \SMTP{} code; presumably this is due to \LMA{} not parsing a
-        line properly.  \LMA{} considers this code to be a rejection, so
-        the command above must, too.
+    \item Some records have an \SMTP{} code of \textit{0}, though this is
+        not a valid \SMTP{} code; presumably this is due to \LMA{} not
+        parsing a line properly.  \LMA{} considers this code to be a
+        rejection, so the command above must, too.
 

### 2008-05-20 doc/logparser-conclusion.tex
Moved \appendix to a more appropriate place. Added postfix bibliography entry; I should refer to it from logparser.pdf.
:%s/logs/log files/gc

@@
 
-Parsing Postfix logs appears at first sight to be an almost trivial task,
-especially if one has previous experience in parsing logs, but it turns out
-to be a much more taxing project than initially expected.  The variety and
-breadth of log lines produced by Postfix is quite surprising, because a
-quick survey of sample log files gives the impression that the number of
-distinct log lines is quite small; this impression is due to the uneven
-distribution exhibited by logs produced in normal operation (see
-graph~\refwithpage{rule hits graph} for a vivid illustration of this).
+Parsing Postfix log files appears at first sight to be an almost trivial
+task, especially if one has previous experience in parsing log files, but
+it turns out to be a much more taxing project than initially expected.  The
+variety and breadth of log lines produced by Postfix is quite surprising,
+because a quick survey of sample log files gives the impression that the
+number of distinct log lines is quite small; this impression is due to the
+uneven distribution exhibited by log lines produced in normal operation
+(see graph~\refwithpage{rule hits graph} for a vivid illustration of this).
 
@@
 whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the logs, looking for something out of the
-ordinary, possibly hundreds or even thousands of lines away from the last
-mention of the queueid in question, is extremely time consuming and error
-prone.  In some cases the task is not to spot the line which is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
+more grueling.  Trawling through the log files, looking for something out
+of the ordinary, possibly hundreds or even thousands of lines away from the
+last mention of the queueid in question, is extremely time consuming and
+error prone.  In some cases the task is not to spot the line which is
+unusual, but to spot that a line normally present is missing, i.e.\ to
+realise that one line amongst hundreds is absent.  In all cases the
+evidence must be used to construct a hypothesis to explain the
+irregularities, and that hypothesis must then be tested in the parser; if
+successful, the parser must be modified to deal with the irregularities,
+without adversely affecting the existing functionality.  The complications
+described in \sectionref{additional complications} were solved in the order
+they are described in, and that order closely resembles the frequency with
+which they occur; the most frequently occurring complications dominate the
 warning messages produced, and so naturally they are the first
@@
 \newpage
-
-\appendix

### 2008-05-20 doc/logparser-coverage.tex
:%s/logs/log files/gc

@@
 parser finishes execution.  Ideally all they should contain are mails which
-are awaiting delivery after the period covered by the logs, though they may
-also contain mails whose initial entries are not contained in the logs.
-Any other entries are evidence of a failure in parsing or an aberration in
-the logs.  After parsing the \numberOFlogFILES{} test log files the state
-tables contain 18 entries, breaking down into:
+are awaiting delivery after the period covered by the log files, though
+they may also contain mails whose initial entries are not contained in the
+log files.  Any other entries are evidence of a failure in parsing or an
+aberration in the log files.  After parsing the \numberOFlogFILES{} test
+log files the state tables contain 18 entries, breaking down into:
 
@@
 
-    \item 1 connection which started only seconds before the logs ended and
-        had not yet been fully transferred from client to server.
+    \item 1 connection which started only seconds before the log files
+        ended and had not yet been fully transferred from client to server.
 
-    \item 1 mail which had been accepted only seconds before the logs ended
-        and had not yet been delivered.
+    \item 1 mail which had been accepted only seconds before the log files
+        ended and had not yet been delivered.
 
-    \item 9 mails whose initial log entries were not present in the logs.
+    \item 9 mails whose initial log entries were not present in the log
+        files.
 
@@
 Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test logs is infeasible, verifying a subset of
-those logs is a tractable, if extremely time consuming, task.  A sample of
-logs was obtained by randomly selecting a log file:
+60,721,709 lines in the test log files is infeasible, verifying a subset of
+those log files is a tractable, if extremely time consuming, task.  A
+sample of log lines was obtained by randomly selecting a log file:
 
@@
 the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's logs, and
+requires an understanding of both the parser and Postfix's log files, and
 investigative work to determine the cause of the deficiency, followed by
-further examination of the logs in developing a solution.
+further examination of the log files in developing a solution.

### 2008-05-20 doc/logparser-database-schema.tex
Added cluster_groups to the rules table description and example rule. Improved Splunk review. Improved summary table in review of other parsers conclusion. Minor improvements in parser reviews. Improved review of other parsers conclusion.
:%s/logs/log files/gc

@@
 
-The data saved from parsing the logs is also divided into two tables as
-described below: connections and results.  The connections table contains a
-row for every mail accepted and every connection where there was a
-rejection; the individual fields will be described in
+The data saved from parsing the log files is also divided into two tables
+as described below: connections and results.  The connections table
+contains a row for every mail accepted and every connection where there was
+a rejection; the individual fields will be described in
 \sectionref{connections table}.  The results table has one or more rows per
@@
                 processes will be terminated.  This does not negatively
-                impact on the logs or mail queued by Postfix for delivery.
+                impact on the log files or mail queued by Postfix for
+                delivery.
 
@@
         (described in \sectionref{overlapping rules}).
+
+    \item [cluster\_group] A reference to the \texttt{cluster\_group}
+        table.  That table is used by the Decision Tree algorithm described
+        in a separate document.
 
@@
 The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix logs were described, with
-the purpose of each field discussed in detail.  A clear, comprehensible
-schema is essential when using the extracted data; it's more important when
-using the data than when storing it, because storing the data is a
-write-once operation, whereas utilising the data requires frequent
-searching, sorting and manipulation of the data to produce customised
-reports and/or statistics.
+containing the data extracted from the Postfix log files were described,
+with the purpose of each field discussed in detail.  A clear,
+comprehensible schema is essential when using the extracted data; it's more
+important when using the data than when storing it, because storing the
+data is a write-once operation, whereas utilising the data requires
+frequent searching, sorting and manipulation of the data to produce
+customised reports and/or statistics.

### 2008-05-20 doc/logparser-graphs.tex
:%s/logs/log files/gc

@@
 bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of logs.
+volume of log lines.
 
@@
 
-The difference between logs with a mail loop and logs without can be seen
-in table~\refwithpage{Execution time increase for different rule orderings}
-showing the increases for the different rule orderings and combinations of
-log files:
+The difference between log files with a mail loop and log files without can
+be seen in table~\refwithpage{Execution time increase for different rule
+orderings} showing the increases for the different rule orderings and
+combinations of log files:
 

### 2008-05-20 doc/logparser-introduction.tex
:%s/logs/log files/gc

@@
 Most mail server administrators will have performed some basic processing
-of the logs produced by their mail server at one time or another, whether
-it was to debug a problem, explain to a user why their mail is being
-rejected, or check whether new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has gotten in the last week, and possibly even
+of the log files produced by their mail server at one time or another,
+whether it was to debug a problem, explain to a user why their mail is
+being rejected, or check whether new anti-spam measures are working.  The
+more adventurous will have generated statistics to show how many hits each
+of their anti-spam measures has gotten in the last week, and possibly even
 generated some graphs to clearly illustrate the point to management or
 users.\footnote{This was the author's first real foray into processing
-Postfix logs.}  Very few will have performed in-depth parsing and analysis
-of their logs, where the parsing must correlate the log lines
+Postfix log files.}  Very few will have performed in-depth parsing and
+analysis of their log files, where the parsing must correlate the log lines
 per-connection or per-queueid rather than processing log lines
 independently.  One of the barriers to this type of processing is the
-unstructured nature of Postfix logs, where each log line was added on an ad
-hoc basis as a requirement was discovered or new functionality was
+unstructured nature of Postfix log files, where each log line was added on
+an ad hoc basis as a requirement was discovered or new functionality was
 added.\footnote{A history of all changes made to Postfix is distributed
@@
 differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix logs, presenting \PLP{}, a program
-which parses logs and places the resulting data into a database for later
-use.  The gathered data can then be used to optimise current anti-spam
-measures, provide a baseline to test new anti-spam measures against, or to
-produce statistics showing how effective those measures are.  There are
-numerous other uses for such data: improving server performance by
-identifying troublesome destinations and reconfiguring appropriately;
-identifying regular high volume uses (e.g.\ customer newsletters) and
-restricting those uses to off-peak times; detecting virus outbreaks which
-propagate via email; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
+the difficult process of parsing Postfix log files, presenting \PLP{}, a
+program which parses log files and places the resulting data into a
+database for later use.  The gathered data can then be used to optimise
+current anti-spam measures, provide a baseline to test new anti-spam
+measures against, or to produce statistics showing how effective those
+measures are.  There are numerous other uses for such data: improving
+server performance by identifying troublesome destinations and
+reconfiguring appropriately; identifying regular high volume uses (e.g.\
+customer newsletters) and restricting those uses to off-peak times;
+detecting virus outbreaks which propagate via email; as a base for billing
+customers on a shared server.  Preserving the raw data enables users to
+develop a multitude of uses far beyond those conceived of by the author.
 
@@
 described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the logs and the table
-that stores the rules.
+the tables used for storing the data gleaned from the log files and the
+table that stores the rules.
 

### 2008-05-20 doc/logparser-limitations.tex
:%s/logs/log files/gc

@@
 
-    \item The parser does not distinguish between logs produced by
+    \item The parser does not distinguish between log files produced by
         different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as logs from
-        different sources can be combined in the same database, or it may
-        be viewed as a limitation as there is no facility to distinguish
-        between logs from different sources in the same database.  If the
-        results of parsing logs from different sources must remain
-        separate, the parser can easily be instructed to use a different
-        database to store the results in.
+        same database.  This may be viewed as an advantage, as log files
+        from different sources can be combined in the same database, or it
+        may be viewed as a limitation as there is no facility to
+        distinguish between log files from different sources in the same
+        database.  If the results of parsing log files from different
+        sources must remain separate, the parser can easily be instructed
+        to use a different database to store the results in.
 

### 2008-05-20 doc/logparser-other-parsers-review.tex
:%s/logs/log files/gc

@@
         gathers.  This is by design: the parser is responsible for parsing
-        logs and extracting data --- further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some
-        simple examples are given in \sectionref{motivation}.
+        log files and extracting data --- further processing of that data
+        is a separate concern, deferred to another program.  Using an
+        \SQL{} database simplifies the process of generating such reports
+        (discussed in \sectionref{database as API}); some simple examples
+        are given in \sectionref{motivation}.
 
@@
         for reading compressed log files is quite helpful, as it
-        dramatically reduces the disk space required to store old logs.
+        dramatically reduces the disk space required to store old log
+        files.
 
@@
 the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with the \numberOFlogFILES{} logs
-files tested with,\footnote{The results it reported were not verified, but
-it did not report any errors, and has a very good reputation amongst
-Postfix users.} but is not a suitable base for this project.
+useful report, and successfully dealt with the \numberOFlogFILES{} log
+files files tested with,\footnote{The results it reported were not
+verified, but it did not report any errors, and has a very good reputation
+amongst Postfix users.} but is not a suitable base for this project.
 
@@
 
-Splunk aims to index all an organisation's logs, providing a centralised
-view capable of searching and correlating diverse log sources.  The web
-interface supports complicated searches, providing statistics and graphs in
-real time, a facility not provided by \parsername{} (report generation has
-been deferred to a subsequent program).  Saved searches\footnote{The author
-was unable to save searches, though that may have been due to limitations
-in the free version.} can be run periodically and the results emailed to a
-recipient or sent to a shell script, which presumably can publish the
-results as required (though possibly without the graphs and detailed
-statistics); \parsername{} provides the database and leaves it to the user
-to utilise it in any way, whenever they want, with no restrictions on usage
-of the data.  The interface is optimised for interactive rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database; there are additional
-reports available on \url{http://www.splunkbase.com/}, but the author was
-unable to find any: every category was empty, even those which the
-interface claimed had reports available.  Many types of reports are bundled
-with the software, though most are variations of a bar or pie chart, with
-the exception of bubble and heatmap graphs.  It is very easy to drill down
-through the graphs to extract a portion of the data (e.g.\ select the hour
-with the largest number of events, then select a particular host, and
-finally a specific address), though it is not possible to search on partial
-words.
+Splunk aims to index all an organisation's log files, providing a
+centralised view capable of searching and correlating diverse log sources.
+The web interface supports complicated searches, providing statistics and
+graphs in real time, a facility not provided by \parsername{} (report
+generation has been deferred to a subsequent program).  Saved
+searches\footnote{The author was unable to save searches, though that may
+have been due to limitations in the free version.} can be run periodically
+and the results emailed to a recipient or sent to a shell script, which
+presumably can publish the results as required (though possibly without the
+graphs and detailed statistics); \parsername{} provides the database and
+leaves it to the user to utilise it in any way, whenever they want, with no
+restrictions on usage of the data.  The interface is optimised for
+interactive rather than automated queries and it does not appear to be
+possible to write independent tools to utilise the Splunk database; there
+are additional reports available on \url{http://www.splunkbase.com/}, but
+the author was unable to find any: every category was empty, even those
+which the interface claimed had reports available.  Many types of reports
+are bundled with the software, though most are variations of a bar or pie
+chart, with the exception of bubble and heatmap graphs.  It is very easy to
+drill down through the graphs to extract a portion of the data (e.g.\
+select the hour with the largest number of events, then select a particular
+host, and finally a specific address), though it is not possible to search
+on partial words.
 
@@
 
-In the specific case of parsing Postfix logs, Splunk extracts some standard
-fields: to and from addresses, HELO hostname, date, host the logs were
-collected from, and protocol.  It parses the standard syslog fields at the
-beginning of the line, and extracts any \texttt{key=<value>} pairs
-occurring after the standard syslog prologue; these pairs are the fields
-listed above.  Searches can be based on the extracted fields, and all the
-text in the line is also available for searching.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
+In the specific case of parsing Postfix log files, Splunk extracts some
+standard fields: to and from addresses, HELO hostname, date, host the log
+files were collected from, and protocol.  It parses the standard syslog
+fields at the beginning of the line, and extracts any \texttt{key=<value>}
+pairs occurring after the standard syslog prologue; these pairs are the
+fields listed above.  Searches can be based on the extracted fields, and
+all the text in the line is also available for searching.  \parsername{}
+extracts noticeably more data, though it does not make the full text of the
+line available, and the full power of \SQL{} is available when searching,
 allowing the user to search on arbitrarily complicated conditions.
@@
 documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix logs, in
-fact it operates by coercing Postfix logs into Apache\footnote{The Apache
-web server is the most popular HTTP server in use for the past 10 years;
-more information is available at \url{http://httpd.apache.org/}.} format
-log files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; there is no easy way to
-extend it to handle new log lines.  Although though it does correlate log
-lines by queueid, it does not deal with any of the other complications
+no special purpose code to deal with the intricacies of Postfix log files,
+in fact it operates by coercing Postfix log files into Apache\footnote{The
+Apache web server is the most popular HTTP server in use for the past 10
+years; more information is available at \url{http://httpd.apache.org/}.}
+format log files, for analysis by AWStats' HTTP log file parser.  The
+converting parser only deals with a small portion of the log lines
+generated by Postfix, silently skipping those it cannot deal with, and does
+not distinguish between different types of rejection; there is no easy way
+to extend it to handle new log lines.  Although though it does correlate
+log lines by queueid, it does not deal with any of the other complications
 described in this document.  AWStats supports saving data but the format of
@@
 document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any logs.
-This program was not reviewed in further detail because its aims are so far
-removed from the aims of this project.
+\regexes{} to match significant input lines before it will parse any log
+files.  This program was not reviewed in further detail because its aims
+are so far removed from the aims of this project.
 
@@
 Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix logs:
-\textit{The email servers' reports will show you the number of deliveries
-and the volume of email delivered by day, the domains from which you
-receive or send the most emails, the relays most used, etc.\/}; notably
+file.  Its Postfix parser extracts the following data from Postfix log
+files: \textit{The email servers' reports will show you the number of
+deliveries and the volume of email delivered by day, the domains from which
+you receive or send the most emails, the relays most used, etc.\/}; notably
 rejections are not mentioned or dealt with, and unlike \parsername{} there
@@
 
-Logrep extracts far less data from Postfix logs than \parsername{},
+Logrep extracts far less data from Postfix log files than \parsername{},
 completely ignores rejections, is effectively undocumented, doesn't deal
-with the more complicated aspects of Postfix logs, and at the time of
+with the more complicated aspects of Postfix log files, and at the time of
 writing doesn't work properly.
@@
 
-\dag\dag\parsername{} is the only parser with documentation for its data store.
+\dag\dag\parsername{} is the only parser with documentation for its data
+store.
 
@@
 lines.  Most other parsers immediately generate a report and discard the
-data extracted from the logs; those which don't discard the data retain it
-in a format inaccessible to other tools.  Nearly all of the parsers
-reviewed can produce a report of greater or lesser detail and complexity, a
-facility not offered by \parsername{}; reporting is deferred to a
-subsequent application.  The quality of the documentation offered by the
-subset of parsers which provide some varies from unusable to good; none of
-the parsers provide any documentation on the format of their data stores
+data extracted from the log files; those which don't discard the data
+retain it in a format inaccessible to other tools.  Nearly all of the
+parsers reviewed can produce a report of greater or lesser detail and
+complexity, a facility not offered by \parsername{}; reporting is deferred
+to a subsequent application.  The quality of the documentation offered by
+the subset of parsers which provide some varies from unusable to good; none
+of the parsers provide any documentation on the format of their data stores
 (if any).  Less than half of the parsers were capable of parsing the
@@
 reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix logs achieved by \parsername{}.
+understanding of Postfix log files achieved by \parsername{}.

### 2008-05-20 doc/logparser-parsing-algorithm.tex
:%s/logs/log files/gc

@@
 internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix logs, presenting the
-resulting data in a normalised, easy to use representation.  The
-algorithm's task is to follow the journey each mail takes through Postfix,
-combining the data extracted by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.
+of parsing, the eccentricities and oddities of Postfix log files,
+presenting the resulting data in a normalised, easy to use representation.
+The algorithm's task is to follow the journey each mail takes through
+Postfix, combining the data extracted by rules into a coherent whole,
+saving it in a useful and consistent form, and performing housekeeping
+duties.
 
@@
 Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the logs
-it usually appears as if the child mail has appeared from thin air.
+the mail first appears; from a strictly chronological reading of the log
+files it usually appears as if the child mail has appeared from thin air.
 Subsequently the parent mail will log the creation of the child mail:
@@
 
-The crux of the problem is that re-injected mails appear in the logs
+The crux of the problem is that re-injected mails appear in the log files
 without explicit logging indicating their source.  There are two implicit
@@
 required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the logs used exhibit
-the idiosyncrasies and peculiarities a mail server in the wild must deal
-with, but the downside is that significantly altering the configuration
-just to log messages from a different Postfix component is not an option.}
+mail for a university department; the benefit is that the log files used
+exhibit the idiosyncrasies and peculiarities a mail server in the wild must
+deal with, but the downside is that significantly altering the
+configuration just to log messages from a different Postfix component is
+not an option.}
 
@@
 Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over logs for some time there are
-several reasons for this infrequent occurrence:
+immediately obvious reason.  After poring over log files for some time
+there are several reasons for this infrequent occurrence:
 

### 2008-05-20 doc/logparser-parsing-rules.tex
:%s/logs/log files/gc

@@
 time increase for different rule orderings} showing the mean increases for
-different groupings of logs.  Overall this provides a modest but worthwhile
-performance increase of approximately 9\%, for a small investment in time
-and programming.
+different groupings of log files.  Overall this provides a modest but
+worthwhile performance increase of approximately 9\%, for a small
+investment in time and programming.
 
@@
 
-The logs produced by Postfix differ from installation to installation,
+The log files produced by Postfix differ from installation to installation,
 because administrators have the freedom to choose the subset of available

### 2008-05-20 doc/logparser.tex
Tested Splunk.
Tested Isoqlog and Anteater. Didn't test "Log analyser --- throughput monitor" - it would be pointless.
Tested lire. Logrep can't be tested, it doesn't work properly.
Improved other parsers introduction. Sawmill tested. YAALA tested.
Added summary table to other parsers' conclusion. Some more details about AWStats. Started testing LMA.
Added cluster_groups to the rules table description and example rule. Improved Splunk review. Improved summary table in review of other parsers conclusion. Minor improvements in parser reviews. Improved review of other parsers conclusion.
Moved \a

@@
 
-    Parsing Postfix logs is much more difficult than it first appears, but
-    it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the logs, and thus accuracy in reconstructing the actions
-    taken by Postfix to generate the logs.  This paper describes the
-    creation of a parser, documenting the parsing algorithm, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation which stores data gleaned from the logs in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
+    Parsing Postfix log files is much more difficult than it first appears,
+    but it \textit{is\/} possible to achieve a high degree of accuracy in
+    understanding the log files, and thus accuracy in reconstructing the
+    actions taken by Postfix to generate the log files.  This paper
+    describes the creation of a parser, documenting the parsing algorithm,
+    explaining the difficulties encountered and the solutions developed,
+    with reference to an implementation which stores data gleaned from the
+    log files in an SQL database.  The gathered data can then be used to
+    optimise current anti-spam measures, provide a baseline to test new
+    anti-spam measures against, or to produce statistics showing how
+    effective those measures are.
 
@@
 
-XXX TEST OTHER PARSERS ON THE 93 TEST LOG FILES\@:
-
-\begin{enumerate}
-
-    \item Sawmill Universal Log File Analysis and Reporting
-    \item Splunk
-    \item Isoqlog
-    \item Log analyser --- throughput monitor
-    \item Anteater
-    \item Yet Another Advanced Logfile Analyser
-    \item Logparser/Lire
-    \item Logrep
-    \item Log Mail Analyser
-
-\end{enumerate}
+XXX WEKA FOR DECISION TREE DISPLAY\@.
 
@@
 \input{logparser-conclusion.tex}
+
+\appendix
+
 \input{logparser-other-parsers-review.tex}

### 2008-05-20 doc/logparser-background.tex
Corrections to sections 1 and 2 from my proofreading.

@@
 This section provides background information helpful in understanding the
-remainder of the paper.  It begins with a discussion of the motivation
+remainder of the document.  It begins with a discussion of the motivation
 underlying the project, followed by some technical information: the use of
@@
 introduction to Postfix, concentrating on the topics most relevant to this
-paper, namely Postfix anti-spam restrictions and policy servers.  The
+document, namely Postfix anti-spam restrictions and policy servers.  The
 assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this paper.  Other projects which attempt to
+are the conventions used in this document.  Other projects which attempt to
 parse Postfix log files are summarised (full details are available in
@@
 
-This paper and the program it describes are part of a larger project to
-optimise a server's Postfix restrictions, generate statistics and graphs,
-and provide a platform on which new restrictions can be trialled and
-evaluated to see if they are beneficial in the fight against spam.  The
-program parses Postfix log files and populates a database with the data
-gleaned from those log files, providing a consistent and simple view of the
-log files which future tools can utilise.  The gathered data can then be
-used to optimise current anti-spam measures, provide a baseline to test new
-anti-spam measures against, or to produce statistics showing how effective
-those measures are.
+This document and the program it describes are part of a larger project to
+optimise a mail server's Postfix restrictions, generate statistics and
+graphs, and provide a platform on which new restrictions can be trialled
+and evaluated to determine if they are beneficial in the fight against
+spam.  The program parses Postfix log files and populates a database with
+the data gleaned from those log files, providing a consistent and simple
+view of the log files which future tools can utilise.  The gathered data
+can then be used to optimise current anti-spam measures, provide a baseline
+to test new anti-spam measures against, or to produce statistics showing
+how effective those measures are.
 
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
+A snippet of \SQL{} provides a short example of the optimisation possible
+using data from the database is determining which Postfix restrictions
+reject the highest number of mails:
 
@@
 
+XXX GIVE SAMPLE OUTPUT FOR THE SNIPPET ABOVE\@.
+
 Another example is determining which restrictions are not effective: this
 shows which restrictions had fewer than 100 hits on the last log file
-parsed.
+parsed, and the percentage of total rejections each restriction represents.
 
@@
 These database queries yield summary statistics about the efficiency of
-spam avoidance techniques that is far less feasible to assess directly from
-log files without prior pre-processing into a database in the fashion
-proposed, implemented and tested herein.
+spam avoidance techniques; statistics that are far less feasible to assess
+directly from log files without prior pre-processing into a database in the
+fashion proposed, implemented and tested herein.
 
@@
 
-    \item The majority of programming languages have existing code
-        available providing access to databases, so the majority of
-        languages will be able to access the results obtained by the
-        parser.  If this parser was written to be utilised as shared code
-        there would be little or no difficulty in using it when writing in
-        the programing language it is implemented in, but every other
-        programing language wishing to utilise it would require an
-        interface layer to be written.  This requirement would drastically
-        reduce the viability of any project wishing to build upon the work
-        of this project.
+    \item XXX REWRITE THIS PARAGRAPH\@.  The majority of programming
+        languages have existing code available providing access to
+        databases, so the majority of languages will be able to access the
+        results obtained by the parser.  If this parser was written to be
+        utilised as shared code there would be little or no difficulty in
+        using it when writing in the programing language it is implemented
+        in, but every other programing language wishing to utilise it would
+        require an interface layer to be written.  This requirement would
+        drastically reduce the viability of any project wishing to build
+        upon the work of this project.
 
@@
             \item If the database supports it, access can be granted on a
-                fine-grained basis so that the finance department can
-                produce invoices, the helpdesk can run limited queries as
+                fine-grained basis, e.g.\ allowing the finance department
+                to produce invoices, the helpdesk to run limited queries as
                 part of dealing with support calls, and the administrators
-                can have full access to the data.
+                to have full access to the data.
 
@@
 protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity many virus and/or spam sending programs fail to implement it
-properly, so requiring strict adherence to the protocol specification is
-beneficial in protecting against spam and
+the simplicity of the protocol many virus and/or spam sending programs fail
+to implement it properly, so requiring strict adherence to the protocol
+specification is beneficial in protecting against spam and
 viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
@@
 A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02}.
+introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02},
+the definitive references are~\cite{RFC821, RFC2821}.
 
@@
 combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of severity and/or permissiveness.  Postfix
+allowing varying levels of stricture and/or permissiveness.  Postfix
 leverages simple lookup tables to support arbitrarily complicated
@@
 difficult.  The log files tend to follow a 90\%-10\% pattern: 90\% of the
-time the journey is simple, but the other 10\% of the time requires 90\% of
-the code.\footnote{These numbers don't have a solid scientific basis,
-they're based on gut feeling from writing and debugging the software.}
+time the journey is simple, but the other 10\% of the time accounts for
+90\% of the code in the parser.\footnote{These numbers don't have a solid
+scientific basis, they're based on gut feeling from writing and debugging
+the software.}
 
@@
 principle was also followed when designing the parser (\sectionref{parser
-design}), notably having separate actions rather than one huge action.} and
-is separated into various component programs to perform the tasks required
-of an \MTA{}\@: receive mail, send mail, local delivery of mail, etc. ---
-full details can be found at~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix program which produced it, and this
+design}), notably in having separate actions rather than one huge action.}
+and is separated into various component programs to perform the tasks
+required of an \MTA{}\@: receive mail, send mail, local delivery of mail,
+etc. --- full details can be found at~\cite{postfix-overview}.  Each log
+line contains the name of the Postfix component which produced it, and this
 information is used when determining which rules should be used to parse
@@
 
+\label{Mixing and matching Postfix restrictions}
+
 Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}, the following is a brief overview
+smtpd_per_user_control, policy-servers}; the following is a brief overview
 only.
@@
 don't deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging the
-rejection.  Postfix uses simple lookup tables as the deciding factor when
-evaluating some restrictions, e.g.\newline
+delay is that Postfix has more information available when logging
+rejections.
+
+Each restriction is evaluated to produce a result of \textit{reject},
+\textit{permit}, \textit{dunno\/} or the name of another restriction to be
+evaluated.\footnote{Other results are possible as described
+in~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.} The
+meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
+\textit{dunno\/} means to stop evaluating the current restriction and
+continue processing the remainder of the restriction list, allowing more
+specific cases to be used as exceptions to more general cases.  When the
+result is the name of another restriction Postfix will evaluate the new
+restriction, allowing restrictions to be chosen based on the client \IP{}
+address, HELO hostname, sender address, recipient address,
+etc.\footnote{E.g.\ the administrator may require that all clients on the
+local network have valid \DNS{} entries, to prevent people sending mail
+from unknown machines.}  The administrator can define new restrictions as a
+list of existing restrictions, allowing arbitrarily long and complex
+sequences of lookups, restrictions and exceptions.  Postfix tries to
+protect the administrator in as far as is reasonable, e.g.\ the restriction
+\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
+because the parameter it checks (the hostname given in the HELO command) is
+under the control of the remote client.  Despite this, it is possible for
+the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
+all users --- the administrator must remember that \textit{With great power
+comes great responsibility}.
+
+Postfix uses simple lookup tables as the deciding factor when evaluating
+some restrictions, e.g.\ in the example restriction below\newline
 \tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
@@
 rejecting the command if the client is listed.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible, for full details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.}
-The meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address, etc.  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great
-power comes great responsibility}.
-
 This description is necessarily brief, for further details
@@
 
-A policy server~\cite{policy-servers} is an external program that accepts
-state information from Postfix for each \SMTP{} command not rejected by an
-earlier restriction; the policy server can utilise that state information
-to implement whatever logic is required.  For example some users can be
-restricted to sending mail on the third Tuesday after pay day only --- this
-example may actually be useful in a payroll system to prevent problems from
-spam or (worse) phishing mails with faked sender addresses.  More commonly
-encountered scenarios are:
+A \textit{policy server\/}~\cite{policy-servers} is an external program
+consulted by Postfix to determine the fate of an \SMTP{} command.  The
+policy server is given state information\footnote{Sample state information
+is shown in table~\refwithpage{Example attributes sent to policy servers}}
+and returns a verdict as described in \sectionref{Mixing and matching
+Postfix restrictions}.  The policy server can perform more complex checks
+than those provided by Postfix: a trivial example is restricting mail from
+addresses associated with the payroll system to sending mail on the third
+Tuesday after pay day only, to help prevent problems from spam or (worse)
+phishing mails with faked sender addresses.\footnote{A
+phishing mail might claim that the payroll system had a disastrous disk
+failure; until the server is replaced and restored all salary payments will
+have to be processed manually, so please reply to this mail with your name,
+address and back account details.}
+
+Some widely deployed policy servers:
 
@@
         resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}\@.
+        some features of \SMTP{}, e.g.\ forwarding mail from one company or
+        university to another when a user changes jobs.
 
@@
         Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is too high
-        the mail is rejected.  Postfix's restrictions are binary, and the
-        administrator must manually whitelist clients if they are to bypass
-        a restriction; using a threshold which requires hitting several
-        restrictions frees the administrator from whitelisting clients
-        which fall foul of one restriction only.
+        against the sending system --- if the eventual score is higher than
+        a threshold the mail is rejected.  Postfix's restrictions are
+        binary, and the administrator must manually whitelist clients if
+        they are to bypass a restriction; using a threshold which requires
+        hitting several restrictions frees the administrator from
+        whitelisting clients which fall foul of one restriction only.
 
@@
 Example attributes taken from~\cite{policy-servers}:
+
+\begin{table}[ht]
+
+    \caption{Example attributes sent to policy servers}\label{Example
+    attributes sent to policy servers}
 
@@
     \end{tabular}
+\end{table}
 
@@
 to make adding new rules as easy as possible, to enable administrators to
-properly parse their log lines.  To enable this the parser is divided into
+properly parse their log files.  To enable this the parser is divided into
 three parts:
@@
         actions will be executed.  Rules provide an easily extensible
-        method of associating log lines and actions.  Rules are described
-        in detail in \sectionref{rules}.
+        method of associating log lines with actions, and are described in
+        detail in \sectionref{rules}.
 
-    \item [actions] Actions invoked to deal with a log line once it has
-        been identified by the rules: modifying data structures, handing
-        complications, and saving data to the database.  The actions
-        perform the work of reconstructing the journey a mail takes through
-        Postfix.  Full details of actions can be found in
-        \sectionref{parsing-algorithm}, particularly
+    \item [actions] Actions are invoked to deal with a log line once it has
+        been identified by the rules: actions modify data structures,
+        handle complications, and cause data to be saved to the database.
+        The actions perform the work of reconstructing the journey a mail
+        takes through Postfix.  Full details of actions can be found in
         \sectionref{actions-in-detail} and \sectionref{adding new actions}.
@@
     \item [framework] The framework is responsible for loading rules,
-        managing data structures so actions can retrieve them, finding the
-        rule which matches each log line, invoking the correct action, etc.
-        The framework provides the structure which actions and rules plug
-        into.  The framework is not described in detail in this document
-        because it is not particularly interesting, but there is some
-        coverage in \sectionref{framework}.
+        managing data structures, finding the rule which matches each log
+        line, invoking the correct action, etc.  The framework provides the
+        structure which actions and rules plug into.  The framework is
+        described in detail in \sectionref{framework}.  XXX EXTEND
+        FRAMEWORK SECTION IF NECESSARY\@.
 
@@
 
-Decoupling the parsing rules from the associated actions allows new rules
-to be written and tested without requiring modifications to the parser
-source code (significantly lowering the barrier to entry for new or casual
-users who need to parse new log lines), and greatly simplifies both actions
-and rules.  Decoupling also creates a clear separation of functionality:
-rules handle low level details of identifying log lines and extracting data
-from a log line, whereas the actions handle the higher level details of
-following the path a mail takes through Postfix, assembling the required
-data before storing it, dealing with complications arising, etc.
+XXX MERGE THE NEXT TWO PARAGRAPHS\@.
+
+Decoupling the parsing rules from the associated actions and framework
+allows new rules to be written and tested without requiring modifications
+to the parser source code (significantly lowering the barrier to entry for
+new or casual users who need to parse new log lines), and greatly
+simplifies framework, actions and rules.  Decoupling also creates a clear
+separation of functionality: rules handle low level details of identifying
+log lines and extracting data from a log line; actions handle
+the higher level details of following the path a mail takes through
+Postfix, assembling the required data before storing it, dealing with
+complications arising, etc; the framework provides services to actions and
+stores data.
 
@@
 with the complications described in \sectionref{additional complications}.
+
+XXX CLARIFY THE SENTENCE STARTING WITH "Although this may seem" IN THE NEXT
+PARAGRAPH\@.
 
@@
 
-\subsection{Conventions used in the paper}
+\subsection{Conventions used in the document}
 
 The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this paper; in general the word used was chosen based on
-the context it appears in.
+interchangeably in this document; in general the word used was chosen based
+on the context it appears in.
 
@@
 extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store but the
-majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of greater or lesser complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
+programs.  Some parsers save the data extracted to a data store, but the
+many discard all data once they have finished running, making historical
+analysis impossible.  The other parsers reviewed all produce a report of
+varying complexity and detail, whereas the program described here doesn't
+attempt to produce a report at all; that responsibility is deferred to a
+separate program, to be developed later.  The parsing algorithm and program
+described here are designed to enable much more detailed log analysis by
+providing a stable platform for subsequent programs to develop upon.
+
+XXX EXTEND THIS\@.
 
@@
 with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients), adding new fields,
-and lack of a bundled schema describing the fields.
+multiple values for a field (e.g.\ multiple recipients) and adding new
+fields.  There is no facility to include documentation of the fields,
+separator, etc., unlike \SQL{} databases where every database includes a
+schema naming the fields and the type of data they store (integer, text,
+timestamp, etc.).
 
-Berkeley DB only supports storing \textbf{(key, value)} pairs, unlike
-\SQL{} databases which store arbitrary tuples; in the main table the key is
-an integer referred to by secondary tables, and the value used is a \CSV{}
-line containing all of the data for that row.  The secondary by-sender,
-by-recipient, by-date and by-\IP{} tables use the
+Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
+unlike \SQL{} databases which store arbitrary tuples; in \LMA{}'s main
+table the key is an integer referred to by secondary tables, and the value
+used is a \CSV{} line containing all of the data for that row.  The
+secondary by-sender, by-recipient, by-date and by-\IP{} tables use the
 sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
@@
 DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats
+of the three output formats: it doesn't provide the functionality of a
+basic \SQL{} database, not can it be used with standard Unix text
+processing tools.
 
@@
 records are added when there are multiple recipients (sadly the records are
-not associated in any way), and presumably the same approach is taken when
-there are multiple destination servers.
+not associated or linked in any way), and presumably the same approach is
+taken when there are multiple destination servers.
 
@@
 possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the database schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema, as there is neither requirement nor benefit; it only
-provides \textbf{(key, value)} pairs, so any additional structuring of the
-data is imposed by the application, thus it behoves the application to
-document this structure.
+Berkeley DB\@; all \SQL{} databases embed the schema into the database and
+provide commands for displaying it.  Berkeley DB does not embed a schema,
+as there is neither requirement nor benefit; it only provides \textbf{(key,
+value)} pairs, so any additional structuring of the data is imposed by the
+application, thus it behoves the application to document this structure.
 
@@
 This section has provided background information on several topics relevant
-to the remainder of the paper.  It started with the motivation behind the
-project, continuing with explanations of:
+to the remainder of the document.  It started with the motivation behind
+the project, continuing with explanations of:
 
@@
 
-    \item A description of a new parser's novel three section design.
+    \item A description of the parser's novel design.
 
@@
 
-    \item A brief comparison of this parser versus other Postfix parsers.
+    \item An in-depth comparison of this parser with \LMA{}, the parser
+        described in previously published research.
 

### 2008-05-20 doc/logparser-introduction.tex
Corrections to sections 1 and 2 from my proofreading.

@@
 of their anti-spam measures has gotten in the last week, and possibly even
-generated some graphs to clearly illustrate the point to management or
-users.\footnote{This was the author's first real foray into processing
+generated some graphs to clearly illustrate these statistics to management
+or users.\footnote{This was the author's first real foray into processing
 Postfix log files.}  Very few will have performed in-depth parsing and
@@
 servers, even within the same organisation --- servers may be configured
-differently, or running different version of Postfix.  This paper documents
-the difficult process of parsing Postfix log files, presenting \PLP{}, a
-program which parses log files and places the resulting data into a
-database for later use.  The gathered data can then be used to optimise
+differently, or running different versions of Postfix.  This paper
+documents the difficult process of parsing Postfix log files, presenting
+\PLP{}, a program which parses log files and places the resulting data into
+a database for later use.  The gathered data can then be used to optimise
 current anti-spam measures, provide a baseline to test new anti-spam
@@
 using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The unusual separation of rules,
-actions and algorithm is discussed, giving the reasons that approach was
-taken when designing the parser.
+gathered that is language-neutral.  The novel separation of rules, actions
+and framework is discussed, giving the reasons that approach was taken when
+designing the parser.
 
@@
 sample data it matches successfully against.  The pros and cons of
-overlapping rules are discussed, including techniques for detecting
+overlapping rules are considered, including techniques for detecting
 unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, with reference to the
-graphs in appendix~\refwithpage{graphs}.  The final section is a
-description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
+particular the optimisations used by the algorithm, utilising the graphs in
+appendix~\refwithpage{graphs} for illustrative purposes.  The section
+concludes with a description of using the tools provided with the parser to
+generate new rules (specifically the \regex{} in each rule) from unparsed
+log lines.
 
@@
 process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered and their solutions
+description of the further complications discovered, and their solutions
 which complete the parser.
@@
 in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm, including
-dealing with false negatives and a discussion of the difficulties in
-identifying false positives.  A random sampling of log lines was parsed,
-and the correctness of the results manually verified, as part of
-determining the coverage of the parser.
+connections successfully reconstructed by the parsing algorithm; dealing
+with false negatives and a discussion of the difficulties in identifying
+false positives is also included.  As part of determining the coverage of
+the parser a random sampling of log lines was parsed, and the correctness
+of the results manually verified.
 
@@
 
-Section~\ref{conclusion} contains the paper's conclusion, describing the
+Section~\ref{conclusion} contains the document's conclusion, describing the
 results of the research, design and implementation of the parser.
@@
 The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this paper.  Also listed
-are some additional resources expected to be helpful in understanding
-\SMTP{}, Postfix, anti-spam techniques, or the paper.
+the algorithm, writing the program, and preparing this document.  Also
+listed are some additional resources expected to be helpful in
+understanding \SMTP{}, Postfix, anti-spam techniques, or the document.
 
@@
 
-Appendix~\ref{Glossary} provides a glossary of terms used in the paper.
+Appendix~\ref{Glossary} provides a glossary of terms used in the document.
 
-Appendix~\ref{Acronyms} provides a list of acronyms used in the paper.
+Appendix~\ref{Acronyms} provides a list of acronyms used in the document.

### 2008-05-23 papers/sgai-2008/paper.tex
Changed to use Springer's formatting.

@@
 % $Id: logparser.tex 545 2008-05-01 20:00:22Z tobinjt $
-\documentclass[twocolumn,a4paper,11pt,draft]{article}
+%\documentclass[twocolumn,a4paper,11pt,draft]{article}
+\documentclass[]{svmult}
 
@@
 \usepackage{amstext}
-% Wrapping of URLs; this doesn't work in the bibliography, but the breakurl
-% package does.
-%\usepackage{url}
 % The name of the program, so I only have to change it in one place.
@@
 \usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
 \renewcommand{\refname}{Bibliography}
@@
 
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
-
-% When creating a \PDF{} make the table of contents into links to the pages
-% (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
-% hyperfootnotes need to be disabled to avoid breaking footmisc, but they
-% still seem to work, somehow.
-\ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
-\else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
-\fi{}
-\hypersetup{
-    pdftitle    = {Parsing Postfix log files},
-    pdfauthor   = {John Tobin},
-    final       = true,
-    pdfborder   = {0 0 0},
-%    breaklinks  = true,
-}
 % This is necessary for URLs in the bibliography; I dunno why the url
@@
 % Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
+\usepackage[bottom]{footmisc}
+
+% Extra packages recommended by Springer.
+\usepackage{mathptmx}
+\usepackage{helvet}
+\usepackage{courier}
+\usepackage{makeidx}
+\usepackage{multicol}
 
@@
 
-% This may keep hyperref happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
-
 \begin{document}
 
-\title{A user-extensible and adaptable parser architecture}
-\author{John Tobin \& Carl Vogel \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie \&
-vogel@cs.tcd.ie }
+\title*{A user-extensible and adaptable parser architecture}
+%\author{John Tobin \& Carl Vogel \\ School of Computer Science and Statistics \\
+%Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie \& vogel@cs.tcd.ie }
+\author{John Tobin and Carl Vogel}
+\institute{John Tobin \& Carl Vogel \\ School of Computer Science and Statistics \\
+Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie \& vogel@cs.tcd.ie }
 \maketitle
 
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
-\begin{abstract}
-
+\abstract{%
     Some parsers need to be very precise and restrictive in what they
@@
     do something useful with a previously unknown class of input).
+}
 
-\end{abstract}
+XXX DUE MONDAY JUNE 2$^{ND}$ 2008\@.
+
+PRESENTATION PAPERS\@: MAX 14 PAGES\@.
+
+POSTER PAPERS\@: MAX 6 PAGES\@.
+
+BLACK AND WHITE ONLY\@.
 

### 2008-05-23 papers/sgai-2008/paper.tex
Added some thoughts from Carl; it's now time to flesh everything out.

@@
 \usepackage{ifpdf}
+
+\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
+\makeglossary{}
+\makeacronym{}
 
@@
 \institute{John Tobin \& Carl Vogel \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie \& vogel@cs.tcd.ie }
+Trinity College \\ Dublin 2 \\ Ireland \\ \email{tobinjt@cs.tcd.ie}
+\email{vogel@cs.tcd.ie} }
 \maketitle
@@
 
+XXX FORMAT OF REFERENCES\@.
+
 \section{Introduction}
 
-Background of the project --- emphasise the variety in the log lines.
-Motivation for creating the parser.  At some point I want to compare this
+Background of the project --- emphasise the variety in the log lines, and
+the desire to achieve complete coverage, of both kinds.
+Motivation for creating the parser.  At some point I MIGHT compare this
 architecture to recursive decent parsers, Lex/Yacc parsers, XXX parsers.
@@
 Split into framework, actions, \& rules.  Explain adding new rules and
-actions.  Cut down rules to the minimum possible.  Framework provides
-support functions, parsing loop, manages data --- boring stuff users don't
-need to concern themselves with.
+actions.  Cut down rule attributes to the minimum possible.  Framework
+provides support functions, parsing loop, manages data --- boring stuff
+users don't need to concern themselves with.
+
+Thoughts from Carl:
+
+\begin{itemize}
+
+    \item context-free rules
+
+    \item partially context-aware actions, though not really
+
+    \item transduction: input log lines $\rightarrow$ database
+
+    \item Closer to NLP than fixed grammar.
+
+\end{itemize}
 

### 2008-05-23 doc/logparser-other-parsers-review.tex
Minor formatting changes to table.

@@
 \begin{tabular}{l|l|l|l|l|l}
-    Parser         & Parsed     & Data        & Custom        & Documentation? & Source \\
-                   & test logs? & store?      & reports?      &                & code?  \\
+    Parser         & Parsed test & Data        & Custom        & Documentation  & Source \\
+                   & log files?  & store?      & reports?      & quality?       & code?  \\
     \hline         &             &             &               &                &        \\
@@
     AWStats        & Partially   & Yes         & Partial\dag   & Good           & Yes    \\
-    Throughput     & N/A        & No          & No            & Brief but      & Yes    \\
-    \tab{}monitor  &            &             &               & good           &        \\
+    Throughput     & N/A         & No          & No            & Brief but good & Yes    \\
+    \tab{}monitor  &             &             &               &                &        \\
     Anteater       & No          & No          & No            & No             & Yes    \\

### 2008-05-23 papers/sgai-2008/paper.tex
Lots of content written in the Introduction and Architecture sections.

@@
-% $Id: logparser.tex 545 2008-05-01 20:00:22Z tobinjt $
-%\documentclass[twocolumn,a4paper,11pt,draft]{article}
+% $Id$
 \documentclass[]{svmult}
@@
 \usepackage{ifpdf}
+\usepackage{url}
 
@@
 \usepackage{multicol}
+\usepackage{cite}
 
@@
 \title*{A user-extensible and adaptable parser architecture}
-%\author{John Tobin \& Carl Vogel \\ School of Computer Science and Statistics \\
-%Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie \& vogel@cs.tcd.ie }
 \author{John Tobin and Carl Vogel}
-\institute{John Tobin \& Carl Vogel \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ \email{tobinjt@cs.tcd.ie}
-\email{vogel@cs.tcd.ie} }
+\institute{John Tobin \at{} School of Computer Science and Statistics,
+Trinity College, Dublin 2, Ireland. \newline{} \email{tobinjt@cs.tcd.ie} \and{}
+Carl Vogel \at{} School of Computer Science and Statistics,
+Trinity College, Dublin 2, Ireland. \newline{} \email{vogel@cs.tcd.ie}}
 \maketitle
@@
 \abstract{%
-    Some parsers need to be very precise and restrictive in what they
-    parse, yet allow users to easily adapt existing rules or add new rules
-    to parse new inputs, without requiring an in-depth knowledge of the
-    parser's internal workings.  This paper presents a novel parsing
-    architecture which aims to make the process of parsing new inputs as
-    simple as possible, enabling users to trivially add new rules (to parse
-    variants of existing inputs) and relatively easily add new actions (to
-    do something useful with a previously unknown class of input).
+    Some parsers need to be very precise and strict in what they parse, yet
+    must allow users to easily adapt existing rules or add new rules to
+    parse new inputs, without requiring the user to have an in-depth
+    knowledge and understanding of the parser's internal workings.  This
+    paper presents a novel parsing architecture which aims to make the
+    process of parsing new inputs as simple as possible, enabling users to
+    trivially add new rules or adapt existing rules (to parse variants of
+    existing inputs) and relatively easily add new actions (to do something
+    useful with a previously unknown class of input).  The architecture
+    provides a framework which actions and rules fit into: this framework
+    manages the parsing process, provides support functions for the
+    actions, and automatically performs some optimisations which improve
+    the speed of parsing.
 }
@@
 
+\input{../../doc/logparser-acronyms.tex}
+
 \section{Introduction}
 
-Background of the project --- emphasise the variety in the log lines, and
-the desire to achieve complete coverage, of both kinds.
-Motivation for creating the parser.  At some point I MIGHT compare this
-architecture to recursive decent parsers, Lex/Yacc parsers, XXX parsers.
+\label{Introduction}
 
-\section{Algorithm}
+The architecture described herein was developed as part of a larger project to
+improve anti-spam techniques by analysing the performance of the set of filters
+currently in use, optimising the order and membership of the set based on that
+analysis, and developing new measures to supplement the existing filters where
+there is a deficiency.  The approach chosen was to analyse the log files
+produced by the mail server, rather than modifying the mail server to generate
+performance statistics; this approach improves the chances of other sites
+testing and utilising the software.  The need arose for a parser capable of
+dealing with the huge variety of log lines produced by Postfix~\cite{postfix},
+the \MTA{}:
 
-Split into framework, actions, \& rules.  Explain adding new rules and
-actions.  Cut down rule attributes to the minimum possible.  Framework
-provides support functions, parsing loop, manages data --- boring stuff
-users don't need to concern themselves with.
+\begin{itemize}
+
+    \item Log lines differ between versions of Postfix.
+
+    \item The mail administrator can define custom rejection messages, and
+        thus custom log lines.
+
+    \item External resources utilised by Postfix (e.g.\ policy
+        servers~\cite{policy-servers} or \RBL{} lookups) can also change
+        their messages, without warning.
+
+\end{itemize}
+
+Existing parsers were considered, but their parsing coverage was too low,
+their parsing too inexact, and/or they did not extract sufficient data for
+the purposes of this project.\footnote{A review of the existing parsers
+considered for this project will be included in the thesis.} The effort
+required to adapt and improve an existing parser was judged to be greater
+than the effort to write one from scratch, as the techniques used by the
+existing parsers severely limited their potential: some ignored the
+majority of log lines, parsing specific log lines with great accuracy, but
+in an inextensible fashion; others sloppily parsed the majority of log
+lines, but were incapable of distinguishing between log lines of the same
+type, e.g.\ rejections, and the data they extracted was constrained to the
+\LCD{}.
+
+The solution developed is conceptually simple: provide a few generic functions,
+each capable of dealing with an entire \textit{category\/} of log lines (e.g.\
+rejections), accompanied by a multitude of precise patterns, each of which
+matches all log lines of a specific type, and only that type (e.g.\ rejection
+by a specific \RBL{}).
+
+XXX WHERE SHOULD I DISCUSS COMPLICATIONS\@?
+
+During development of the parser it became apparent that in addition to the
+obvious variety in log lines, there were a large number of complications to
+be overcome
+
+\section{Architecture}
+
+\label{Architecture}
+
+Explain adding new rules and actions.  Cut down rule attributes to the
+minimum possible.
 
@@
 
+XXX BEGIN\@:
+
+The architecture is split into three sections: framework, actions and
+rules.  Each will be discussed separately, but first an overview:
+
+\begin{description}
+
+    \item [Framework]  The framework is the structure which actions and
+        rules fit into.  The framework provides the parsing loop, loading
+        and validation of rules, shared data storage, storage of results,
+        and other support functions.
+
+    \item [Actions]  The actions perform the work required to deal with a
+        category of log lines, e.g.\ creating a new data structure when a
+        log line indicating a remote client connected is processed.
+
+    \item [Rules]  The rules are responsible for matching log lines,
+        specifying the action to invoke and the data to be extracted from
+        the log line.
+
+\end{description}
+
+Decoupling the parsing rules from the associated actions and framework allows
+new rules to be written and tested without requiring modifications to the
+parser source code (significantly lowering the barrier to entry for new or
+casual users who need to parse new log lines), and greatly simplifies
+framework, actions and rules. Decoupling the actions from the
+framework simplifies both framework and actions: the framework provides
+services to the actions, and doesn't need to deal with the complications which
+arise, or the task of reconstructing a mail's journey through Postfix; actions
+benefit from having services provided by the framework, freeing them to
+concentrate on the task of accurately reconstructing each mail's journey
+through Postfix and dealing with the complications arising (XXX REFERENCE).
+
+Decoupling also creates a clear separation of functionality: rules handle low
+level details of identifying log lines and extracting data from a log line;
+actions handle the higher level details of following the path a mail takes
+through Postfix, assembling the required data, dealing with complications
+arising, etc; the framework provides services to actions and stores data.
+
+There is some similarity between the parser's design and William Wood's
+\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
+creating grammars to parse or generate sentences.  The resemblance between
+\ATN{} and the parser is accidental, but it is interesting how two
+different approaches have a similar division of responsibilities, while
+having completely different implementations and semantics.
+
+% Do Not Reformat!
+
+\begin{table}[ht]
+    \caption{Similarities with ATN}\label{Similarities with ATN}
+    % The \smallskip{} below stop the text hitting the lines.
+    \begin{tabular}[]{lll}
+        \hline
+        \noalign{\smallskip}
+        \ATN{}        & Parser    & Similarity                          \\
+        \noalign{\smallskip}
+        \hline
+        \noalign{\smallskip}
+        Networks      & Framework & Determines the sequence of
+                                    transitions or actions which        \\
+                      &           & constitutes a valid input.          \\
+        Transitions   & Actions   & Save data and impose conditions the
+                                    input must meet to be               \\
+                      &           & considered valid.                   \\
+        Abbreviations & Rules     & Responsible for classifying input.  \\
+        \noalign{\smallskip}
+        \hline
+        \noalign{\smallskip}
+    \end{tabular}
+\end{table}
+
+\subsection{Framework}
+
+\label{Framework}
+
+The framework provides utility and support functions for the actions and
+rules.  It provides shared storage to pass data between actions, most
+notably the data structures used to represent each connection and mail,
+though there are some others.  It loads and validates rules, controls
+parsing of log files, invoking actions and tracking performance data, and
+stores results on request.
+
+The function which finds the rule matching the input line and invokes the
+requested action can be expressed in pseudo-code (with indentation denoting
+flow-of-control) as:
+
+\begin{verbatim}
+
+for each line in the input files:
+    for each rule defined by the user:
+        if this rule matches the input line:
+            perform the action specified by the rule
+            skip the remaining rules
+            process the next input line
+    warn the user that the input line was not parsed
+
+\end{verbatim}
+
+\subsection{Actions}
+
+\label{Actions}
+
+Actions contain the code required to deal with log lines from a particular
+category, e.g.\ rejections.
+
+\subsection{Rules}
+
+\label{Rules}
+
 \section{Results}
@@
 versus running with the minimum number necessary to parse the test log
-files.  Then try that again with different orderings?
+files.  Then try that again with different orderings?  Coverage.
 
@@
 framework is rarely necessary.
-
-\appendix
 
@@
 
+% Redefine the command used to produce the glossary title, because the
+% default command produces an unnumbered section whereas I want a numbered
+% section.
+\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
+\printglossary{}
+% Redefine the command a second time to produce acronyms instead of a
+% glossary.
+\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
+\printacronym{}
 \end{document}

### 2008-05-25 papers/sgai-2008/paper.tex
Fleshed out the rules section, mainly with content taken from logparser.pdf. Made it clear that rules cause cations to be invoked, based on feedback from     Ariane that it I gave the opposite impression at first. Added a little more to the framework section.

@@
 dealing with the huge variety of log lines produced by Postfix~\cite{postfix},
-the \MTA{}:
+the \MTA{} in use:
 
@@
 
-    \item The mail administrator can define custom rejection messages, and
-        thus custom log lines.
+    \item The mail administrator can define custom rejection messages,
+        resulting in custom log lines.
 
@@
 \end{description}
+
+The framework tries each rule in turn until it finds one which matches the
+input line, then invokes the action specified by that rule.
 
@@
 
+The function tracks the number of times each rule matches; this data is
+used during subsequent parsing to optimise the order the rules are tried
+in.  Rules specify the Postfix component which produces the log lines they
+parse; this avoids needlessly trying rules which won't match the log line,
+or worse, might match unintentionally.
+
 \subsection{Actions}
@@
 
+Rules are responsible for categorising input lines, specifying both the
+data to be extracted from each line and the action that should be invoked.
+This implementation stores the rules in the same \SQL{} database that the
+results are stored in; this removes any doubt as to which set of rules was
+used to produce a set of results.  Other implementations are free to store
+the rules in whatever fashion suits their needs.  Each rule defines several
+attributes:
+
+\begin{description}
+
+    \item [program] The Postfix program whose log lines the rule applies
+        to.  This avoids needlessly trying rules which won't match the log
+        line, or worse, might match unintentionally.  Rules whose program
+        is \texttt{*} will be tried against any log lines which aren't
+        parsed by program specific rules.
+
+    \item [regex] The \regex{} to match the log line against.  The \regex{}
+        will first have several keywords expanded: this simplifies reading
+        and writing rules; avoids needless repetition of complex \regex{}
+        components; allows the components to be corrected and/or improved
+        in one location; and makes each \regex{} largely self-documenting.
+
+        For efficiency the keywords are expanded and every rule's \regex{}
+        is compiled before attempting to parse the log file --- otherwise
+        each \regex{} would be recompiled each time it was used, resulting
+        in a large, data dependent slowdown.  Rule efficiency concerns are
+        discussed in \sectionref{rule efficiency}, with the impact of
+        compiling and caching \regexes{} covered in \sectionref{Caching
+        each regex}.
+
+    \item [result\_cols] Specifies how the fields in the
+        log line will be extracted.  The format is:
+        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
+        \newline i.e.\ semi-colon or comma separated assignment statements,
+        with the variable name on the left and the matching capture from
+        the \regex{} on the right hand side.
+
+    \item [result\_data] Sometimes rules need to supply a piece of data
+        which isn't present in the log line: e.g.\ setting
+        \texttt{smtp\_code} when mail is accepted.  The format and allowed
+        variables are the same as for \texttt{result\_cols}, except that
+        arbitrary data\footnote{Commas and semi-colons cannot be escaped
+        and thus cannot be used.  This is intended for use with small
+        amounts of data rather than large amounts in any one rule, so
+        dealing with escape sequences was deemed unnecessary.} is permitted
+        on the right hand side of the assignment.
+
+    \item [action] The action that will be invoked when this rule matches a
+        log line; actions are describe in \sectionref{Actions}.
+
+    \item [hits] This counter starts at zero and is incremented each time
+        the rule successfully matches.  At the start of each run the parser
+        sorts the rules in descending order of hits, and at the end of the
+        run updates every rule's hits.  Assuming that the distribution of
+        log lines is reasonably consistent between log files, rules
+        matching more commonly occurring log lines will be tried before
+        rules matching less commonly occurring log lines, lowering the
+        program's execution time.  Rule ordering for efficiency is
+        discussed in \sectionref{rule ordering for efficiency}.
+
+    \item [priority] This is the user-configurable companion to hits: rules
+        will be tried in order of priority, overriding hits.  This allows
+        more specific rules to take precedence over more general rules
+        (described in \sectionref{overlapping rules}).
+
+\end{description}
+
+The rules used by the Postfix log file parser define several additional
+attributes; each implementation is free to add any additional attributes it
+requires.
+
+\subsubsection{Overlapping rules}
+
+\label{overlapping rules}
+
+The parser does not try to detect overlapping rules; that responsibility is
+left to the author of the rules.\footnote{It may be possible to parse
+each rule's regex and determine if any overlap.  The author has not
+attempted to do this: given the tremendous power and flexibility of
+regex, particularly Perl's (XXX REFERENCE FOR PERL\@; GLOSSARY ENTRY?)
+regex, it would be an extremely difficult task.} Unintentionally
+overlapping rules lead to inconsistent parsing and data extraction because
+the order in which rules are tried against each line may change between log
+files, and the first matching rule wins.  Overlapping rules are frequently
+a requirement, allowing a more specific rule to match some log lines and a
+more general rule to match the majority, e.g.\ separating \SMTP{} delivery
+to specific sites from \SMTP{} delivery to the rest of the world.  The
+algorithm provides a facility for ordering overlapping rules: the priority
+field in each rule (defaults to zero).  Rules are sorted by priority,
+highest first, and then rules with the same priority are sorted by the
+number of successful matches when parsing the previous log file.  Negative
+priorities may be useful for catchall rules.
+
+Detecting overlapping rules is difficult, but the following approaches may
+be helpful:
+
+\begin{itemize}
+
+    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
+        similar to: \textbf{select regex from rules order by regex;}
+
+    \item Compare the results of parsing using sorted, shuffled and
+        reversed rules.\footnote{See \sectionref{rule efficiency} for more
+        details of sorting the rules.}  Parse a number of log files using
+        normal sorting, then dump a textual representation of the rules,
+        connections and results tables.  Repeat with shuffled and reversed
+        rules, starting with a fresh database.  If there are no overlapping
+        rules the tables from each run will be identical; differences
+        indicate overlapping rules.  Which rules overlap can be determined
+        by examining the differences in the tables: each result contains a
+        reference to the rule which created it, if the references differ
+        between runs the two rules referenced in the differing records
+        overlap.  Unfortunately this method cannot prove the absence of
+        overlapping rules; it can detect overlapping rules, but only if
+        there are log lines in the input files which match more than one
+        rule.
+
+\end{itemize}
+
 \section{Results}
@@
 framework is rarely necessary.
-
-\cite{postfix}
 

### 2008-05-26 papers/sgai-2008/paper.tex
Use Springer's bibliography style.
Added graph showing distribution of rules per action. Some other minor changes.

@@
 
+\newcommand{\numberOFlogFILES}[0]{%
+    93%
+}
+
+\newcommand{\numberOFrules}[0]{%
+    169%
+}
+
+\newcommand{\numberOFlogLINES}[0]{%
+    60,721,709%
+}
+
+\newcommand{\numberOFlogLINEShuman}[0]{%
+    60.72 million%
+}
+
+\newcommand{\numberOFactions}[0]{%
+    18%
+}
+
 \begin{document}
@@
 BLACK AND WHITE ONLY\@.
-
-XXX FORMAT OF REFERENCES\@.
 
@@
 lines, but were incapable of distinguishing between log lines of the same
-type, e.g.\ rejections, and the data they extracted was constrained to the
-\LCD{}.
+type, e.g.\ rejections, and they extracted only a portion of the data
+required for this project.
 
@@
 matches all log lines of a specific type, and only that type (e.g.\ rejection
-by a specific \RBL{}).
+by a specific \RBL{}).  XXX STOP \RBL{} BEING EXPANDED AGAIN\@.
 
@@
 
-Explain adding new rules and actions.  Cut down rule attributes to the
-minimum possible.
+Explain adding new rules and actions.
 
@@
 parsing of log files, invoking actions and tracking performance data, and
-stores results on request.
+stores results in the database on request.  The framework is independent of
+the parser being written: the same functionality will be required from the
+framework for most or all parsers.
 
@@
 requested action can be expressed in pseudo-code (with indentation denoting
-flow-of-control) as:
+flow of control) as:
 
@@
 Actions contain the code required to deal with log lines from a particular
-category, e.g.\ rejections.
+category, e.g.\ rejections.  The actions are parser-specific: each parser
+will need to write the actions it requires from scratch, unless it's
+extending an existing parser.  In the Postfix parser developed for this
+project there are \numberOFactions{} actions and \numberOFrules{} rules.
+
+\showgraph{build/plot-action-distribution}{Distribution of rules
+per action}{Distribution of rules per action}
 
@@
 
-\bibliographystyle{../common/bibliography-style}
+\bibliographystyle{spmpsci}
 \bibliography{../common/bibliography}

### 2008-05-27 doc/logparser-acronyms.tex
Tested Splunk.
Finished the architecture section, unless Carl has comments. Lots of minor changes.

@@
     SMTP is the protocol which transfers mail from the sender to the
-    recipient across the Internet.  A brief introduction to SMTP is
-    provided in section~\refwithpage{SMTP background}.
+    recipient across the Internet.  %XXX A brief introduction to SMTP is
+    %provided in \sectionref{SMTP background}.
 }}
@@
 \newacronym{SPF}{Sender Policy Framework}{description={
-    SPF is introduced in section~\refwithpage{spf introduction} and
-    explained fully in~\cite{openspf, Wikipedia-spf}.
+    SPF is introduced in \sectionref{spf introduction} and explained fully
+    in~\cite{openspf, Wikipedia-spf}.
 }}
@@
 }}
+
+\newacronym{CLI}{Command Line Interface}{description={
+    An computer interface based on typing commands, rather than using a
+    mouse.
+}}
+
+\newacronym{NLP}{Natural Language Processing}{description={
+    Natural Language Processing~\cite{Wikipedia-nlp} attempts to increase
+    our understanding of the languages normally used by humans (e.g.\
+    English, Spanish, Japanese), with the goal of writing programs which
+    can understand human languages.  Potential applications include:
+    programs which will automatically summarise a text; programs which
+    can be taught facts, will understand questions and formulate answers to
+    those questions.  It is in no way connected to Neuro-linguistic
+    programming, which is generally found when searching for NLP.
+}}

### 2008-05-27 doc/logparser-introduction.tex
Finished the architecture section, unless Carl has comments. Lots of minor changes.

@@
 parsing and the results gleaned from parsing.  The database schema used is
-described in section~\refwithpage{database schema}, explaining in detail
-the tables used for storing the data gleaned from the log files and the
-table that stores the rules.
+described in \sectionref{database schema}, explaining in detail the tables
+used for storing the data gleaned from the log files and the table that
+stores the rules.
 

### 2008-05-27 papers/sgai-2008/paper.tex
Finished the architecture section, unless Carl has comments. Lots of minor changes.

@@
 % $Id$
-\documentclass[]{svmult}
+\documentclass[draft]{svmult}
 
@@
 
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
+\usepackage[style=altlist,number=none,toc=true]{glossary}
 \makeglossary{}
-\makeacronym{}
 
@@
 \newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [page~\pageref{#1}]%
+    \empty{}\ref{#1} [p.~\pageref{#1}]%
 }
@@
 
-XXX DUE MONDAY JUNE 2$^{ND}$ 2008\@.
-
-PRESENTATION PAPERS\@: MAX 14 PAGES\@.
-
-POSTER PAPERS\@: MAX 6 PAGES\@.
-
-BLACK AND WHITE ONLY\@.
+XXX DUE MONDAY JUNE 2$^{ND}$ 2008\@.  PRESENTATION PAPERS\@: MAX 14
+PAGES\@.  POSTER PAPERS\@: MAX 6 PAGES\@.  BLACK AND WHITE ONLY\@.
 
@@
 matches all log lines of a specific type, and only that type (e.g.\ rejection
-by a specific \RBL{}).  XXX STOP \RBL{} BEING EXPANDED AGAIN\@.
-
-XXX WHERE SHOULD I DISCUSS COMPLICATIONS\@?
-
-During development of the parser it became apparent that in addition to the
-obvious variety in log lines, there were a large number of complications to
-be overcome
+by a specific \RBLshort{}).
 
@@
 \label{Architecture}
-
-Explain adding new rules and actions.
-
-Thoughts from Carl:
-
-\begin{itemize}
-
-    \item context-free rules
-
-    \item partially context-aware actions, though not really
-
-    \item transduction: input log lines $\rightarrow$ database
-
-    \item Closer to NLP than fixed grammar.
-
-\end{itemize}
-
-XXX BEGIN\@:
 
@@
 concentrate on the task of accurately reconstructing each mail's journey
-through Postfix and dealing with the complications arising (XXX REFERENCE).
+through Postfix and dealing with the complications arising (discussed in
+\sectionref{complications}).
 
@@
 project there are \numberOFactions{} actions and \numberOFrules{} rules.
+The distribution of rules to actions is very skewed, as can be seen in
+Fig.~\refwithpage{Distribution of rules per action}.
+\showgraph{build/plot-action-distribution}{Distribution of rules per
+action}{Distribution of rules per action} The distribution is skewed for
+the same reason the architecture is so successful in enabling users to
+parse new inputs: most new inputs can be parsed by a combination of a new
+rule and an existing action.  Unsurprisingly the action with the highest
+number of associated rules is \texttt{REJECTION}, the action which handles
+rejections; it's followed by \texttt{INFO}, the action responsible for
+handling informative inputs: these are log lines that supplement the data
+gathered from other log lines, but that don't themselves represent
+interesting events.  The third most common action is, perhaps surprisingly,
+\texttt{IGNORE}: there are many log lines which are not of interest to the
+parser, e.g.\ warning messages about \DNS{} mismatches, or log lines
+lacking a way to identify which data structure they should be associated
+with.\footnote{Postfix allocates a queueid (explained in the
+glossary,~\sectionref{Glossary}) to each mail; most log lines contain the
+queueid of the associated mail, but some log lines, particularly with older
+versions of Postfix, lack a queueid, preventing their association with a
+particular data structure.}  The parser warns about every unparsed input,
+to alert the user that they need to alter or extend their rules, hence the
+\texttt{IGNORE} action: the input has been parsed, but it is of no benefit
+so it is ignored.  The remaining actions have only one or two associated
+rules: in some cases this is because there will only ever be one type of
+input for that action, e.g.\ all log lines showing that a remote client has
+connected are matched by a single rule and handled by the \texttt{CONNECT}
+action; in other cases the action is required to address a deficiency in
+the log files, or a complication which arises during
+parsing.\footnote{\label{complications footnote}The complications which
+arose during development of the parser will be documented in detail in the
+author's thesis.}
 
-\showgraph{build/plot-action-distribution}{Distribution of rules
-per action}{Distribution of rules per action}
+The ability to add special purpose actions to deal with difficulties and
+new requirements which are discovered during parser development is one of
+the strengths of this architecture: instead of writing one monolithic
+function which must be modified to support new behaviour or resolve
+challenges, with all the attendant risks of adversely affecting the
+existing, working parser, a new, independent action can be written to
+satisfy the new requirements.  In some cases the new action will require
+the cooperation of other actions, e.g.\ to set or check a flag; this does
+introduce the possibility introducing failure by modifying existing
+actions, but modifications should be smaller and less frequent, and thus
+failures should be less likely and easier to test for and diagnose.  The
+parser can be written in an Object
+Oriented~\cite{Wikipedia-object-orientation} style, allowing sub-classes to
+extend or override actions in addition to adding new actions; because each
+action is an independent function, the sub-class need only modify the
+action it is interested in, rather than reproducing large chunks of
+functionality.
+
+\label{complications}
+
+During development of the parser it became apparent that in addition to the
+obvious variety in log lines, there were a large number of complications to
+be overcome.\footref{complications footnote}  Some were the result of
+deficiencies in Postfix's logging, which were improved or resolved by later
+versions of Postfix; others were due to the vagaries of process scheduling,
+client behaviour, and administrative actions.  All were successfully
+accommodated in the parser: adding new actions was enough for several of
+the complications; others required modifications to existing actions to
+work around the difficulties; the remainder were resolved by adapting
+actions to pass information to subsequent actions (via the framework).
 
@@
         in one location; and makes each \regex{} largely self-documenting.
-
         For efficiency the keywords are expanded and every rule's \regex{}
@@
     \item [action] The action that will be invoked when this rule matches a
-        log line; actions are describe in \sectionref{Actions}.
+        log line; actions are described in \sectionref{Actions}.
 
@@
 
+Adding new rules is simple: simply define all the required attributes and
+append the new rule to wherever the existing rules are stored.  The
+software developed for this project includes a utility based on
+\SLCT{}~\cite{slct-paper} to aid in producing \regexes{} to parse
+previously unparsed inputs.
+
 \subsubsection{Overlapping rules}
@@
 
+XXX DO I NEED A SECTION ABOUT OVERLAPPING RULES\@?
+
 The parser does not try to detect overlapping rules; that responsibility is
-left to the author of the rules.\footnote{It may be possible to parse
-each rule's regex and determine if any overlap.  The author has not
-attempted to do this: given the tremendous power and flexibility of
-regex, particularly Perl's (XXX REFERENCE FOR PERL\@; GLOSSARY ENTRY?)
-regex, it would be an extremely difficult task.} Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
+left to the author of the rules.\footnote{It may be possible to parse each
+rule's regex and determine if any overlap.  The author has not attempted to
+do this: given the tremendous power and flexibility of regexes,
+particularly Perl's regexes~\cite{perlre}, it would be an extremely
+difficult task.} Unintentionally overlapping rules lead to inconsistent
+parsing and data extraction because the order in which rules are tried
+against each line may change between log files, and the first matching rule
+wins.  Overlapping rules are frequently a requirement, allowing a more
+specific rule to match some log lines and a more general rule to match the
+majority, e.g.\ separating \SMTP{} delivery to specific sites from \SMTP{}
+delivery to the rest of the world.  The algorithm provides a facility for
+ordering overlapping rules: the priority field in each rule (defaults to
+zero).  Rules are sorted by priority, highest first, and then rules with
+the same priority are sorted by the number of successful matches when
+parsing the previous log file.  Negative priorities may be useful for
+catchall rules.
 
@@
 
+\subsection{Architecture characteristics}
+
+\label{Architecture characteristics}
+
+The architecture can be compared to existing approaches to aid in fully
+understanding it.
+
+\begin{description}
+
+    \item [Context-free rules:]  Rules are context free: they do not take
+        into account inputs which have previously been seen, nor do they
+        perform any kind of lookahead to future inputs.  In context-free
+        grammar terms the parser rules could be described as:
+
+        $\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
+        \text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
+
+    \item [Partially context-sensitive actions:]  XXX IS THIS THE CORRECT
+        TERMINOLOGY\@?  Actions can consult the results of previous actions
+        when executing, providing some context sensitivity.  In some cases
+        the lack of results from previous actions can be useful
+        information, e.g.\ in the Postfix parser when a client disconnects
+        if there have been no useful log lines for that client the data is
+        discarded, rather than trying to store empty records in the
+        database.
+
+    \item [Matching is simple:]  The first matching rule wins: there is no
+        backtracking to try alternate rules, no attempt made to pick a
+        \textit{best\/} rule; the first rule which matches the line is the
+        rule which parses it and determines the action that will be
+        invoked.\footnote{Rule authors can affect the order rules will be
+        tried in by setting the \texttt{priority} attribute of one or more
+        rules.}
+
+    \item [Transduction:]  The parser can be thought of as implementing
+        transduction: it takes data in one form (log files) and transforms
+        it to another form: a database in the case of the Postfix log
+        parser, though it could be normalised files or files which only
+        contain discrepancies or other interesting data.
+
+    \item [Closer to \NLP{} than fixed grammar:]  Unlike traditional
+        parsers such as those used when compiling a programming language,
+        this architecture does not have a fixed grammar specification which
+        inputs must adhere to.  The architecture is designed for easy user
+        alteration, in fact such alteration is expected.
+
+\end{description}
+
 \section{Results}
@@
 
+The architecture's greatest strength is the ease with which new inputs
+can be parsed and new actions written.
+
 \bibliographystyle{spmpsci}
@@
 \printglossary{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
+
+\glossary{name={queueid},description={
+    Each mail in Postfix's queue is assigned a queueid to uniquely identify
+    it.  Queueids are assigned from a limited pool, so although they are
+    guaranteed to be unique for the lifetime of the mail, given sufficient
+    time they will be reused.
+}}
+
 \end{document}

### 2008-05-27 papers/sgai-2008/paper.tex
Added lots of content from logparser.pdf to the results section, including lots of graphs.  I'll soon have to cut content out, but I'll leave it in for now and ask Carl's advice.

@@
         in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
+        discussed in \sectionref{rule efficiency}.
 
@@
 
+Parsing efficiency is an obvious concern when the parser routinely needs to
+deal with 75 MB log files containing 300,000 log lines (generated daily on
+a mail server handling mail for approximately 700 users --- large scale
+mail servers would have much larger log files on a daily basis).  When
+generating the data for the graphs included in this document,
+\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
+log lines) were each parsed 10 times, the first run discarded, and the
+execution time for the remaining 9 runs averaged.  The first run is
+discarded for two reasons:
+
+\begin{enumerate}
+
+    \item The execution time will be higher because the log file must be
+        read from disk, whereas for subsequent runs the log file will be
+        cached in memory by the operating system.
+
+    \item The execution time will also be higher because the rule ordering
+        will be sub-optimal compared to subsequent runs.
+
+\end{enumerate}
+
+Saving results to the database was disabled for the test runs, as that
+dominates the run time of the program, and the tests are aimed at measuring
+the speed of the parser rather than the speed of the database and the disks
+the database is stored on.
+
+\subsection{Architecture Scalability}
+
+MINIMUM VERSUS MAXIMUM NUMBER OF RULES\@.
+
+An important property of a parser is how execution time scales relative to
+input size: does it scale linearly, polynomially, or exponentially?
+Figure~\refwithpage{execution time vs file size vs number of lines graph}
+shows the execution time in seconds, file size in MB and tens of thousands
+of log lines per log file.  All three lines run roughly in parallel, giving
+a visual impression that the algorithm scales linearly with input size.
+This impression is borne out by fig.~\refwithpage{execution time vs file
+size vs number lines factor} which plots the ratio of file size vs
+execution time and ratio of number of log lines vs execution time (higher
+is better).  As the reader can see the ratios are quite tightly banded,
+showing that the algorithm scales linearly: the much larger log files
+between points 60 and 70 on the X axis in fig.~\refwithpage{execution time
+vs file size vs number of lines graph} actually cause the ratio to
+increase, rather than decrease.  The strange behaviour where larger log
+files are parsed more efficiently is explained fully in \sectionref{Why are
+there dips in the graphs?}.
+
+\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
+size vs number of lines}{execution time vs file size vs number of lines
+graph}
+
+\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
+size and number of lines to execution time}{execution time vs file size vs
+number lines factor}
+
+\subsection{Rule ordering}
+
+\label{Rule ordering}
+\label{rule ordering for efficiency}
+\label{rule efficiency}
+
+At the time of writing there are \numberOFrules{} different rules, with the
+top 10\% matching the vast majority of the log lines, and the remaining log
+lines split across the other 90\% of the rules (as shown in
+fig.~\refwithpage{rule hits graph}).  \showgraph{build/plot-hits}{Hits per
+rule}{rule hits graph} Assuming that the distribution of log lines is
+reasonably steady over time, program efficiency should benefit from trying
+more frequently matching rules before those which match less frequently.
+To test this hypothesis three full test runs were performed with different
+rule orderings:
+
+\begin{description}
+
+    \item [normal]  The most optimal order, according to the hypothesis:
+        rules which match most often will be tried first.
+
+    \item [shuffle] Random ordering --- the rules will be shuffled once
+        before use and will retain that ordering for the entirety of the
+        log file.  Note that the ordering will change every time the parser
+        is executed, so 10 different orderings will be generated for each
+        log file in the test run.  This is intended to represent an
+        unsorted rule set.
+
+    \item [reverse] Hypothetically the worst order: the most frequently
+        matching rules will be tried last.
+
+\end{description}
+
+Figure~\refwithpage{percentage increase of shuffled and reversed over
+normal} shows the percentage increase of execution times.  Overall this
+provides a modest but worthwhile performance increase of over 10\%, for a
+small investment in time and programming.
+\showgraph{build/plot-normal-shuffle-reverse-factor}{Percentage increase of
+shuffled and reversed over normal}{percentage increase of shuffled and
+reversed over normal}
+
+\subsubsection{Why are there dips in the graphs?}
+\label{Why are there dips in the graphs?}
+
+The dips at log files 22 and 62--68 correspond to peaks in log file size in
+graph~\refwithpage{execution time vs file size vs number of lines graph},
+and peaks in graph~\refwithpage{execution time vs file size vs number lines
+factor} (where a peak means that more lines are processed per second, i.e.\
+performance is better).  The explanation for this took some time to arrive
+at, but it turns out to be reasonably simple.  The large log files in
+question were caused by a mail forwarding loop, where the distribution of
+log lines is quite different to normal, resulting in different performance
+characteristics.  Fewer
+
+The vast majority of log lines when a mail loop occurs are from Postfix
+components which have a small number of rules associated with them, whereas
+in general \daemon{smtpd} adds the majority of log lines, and also has the
+highest number of rules.  \daemon{smtpd} log lines are distributed across
+rules much more evenly than the log lines which occur frequently during a
+mail loop, so the average number of rules
+required to parse a \daemon{smtpd} log line is much higher that the average
+number required to parse other log lines.
+
+These two characteristics combine to reduce the average number or rules
+required to parse a log line when there is a mail loop, as shown by the
+peaks in fig.~\refwithpage{execution time vs file size vs number lines
+factor}.  When the rule ordering is reversed the majority of log lines
+generated by a mail loop will be parsed with very few rules, whereas
+without a mail loop the majority of log lines require a large number of
+rules; this leads to a noticeable drop in the average time required to
+parse a log line, as shown in fig.~\refwithpage{percentage increase of
+shuffled and reversed over normal}.  The number of rules which need to be
+consulted when the ordering is shuffled varies between the optimum and
+nadir, and the performance varies proportionally.
+
+\subsection{Coverage}
+
 \section{Conclusion}

### 2008-05-27 papers/sgai-2008/paper.tex
Pruned some of the content.

@@
 
-Decoupling the parsing rules from the associated actions and framework allows
-new rules to be written and tested without requiring modifications to the
-parser source code (significantly lowering the barrier to entry for new or
-casual users who need to parse new log lines), and greatly simplifies
-framework, actions and rules. Decoupling the actions from the
-framework simplifies both framework and actions: the framework provides
-services to the actions, and doesn't need to deal with the complications which
-arise, or the task of reconstructing a mail's journey through Postfix; actions
-benefit from having services provided by the framework, freeing them to
-concentrate on the task of accurately reconstructing each mail's journey
-through Postfix and dealing with the complications arising (discussed in
-\sectionref{complications}).
+Decoupling the parsing rules from the associated actions and framework
+allows new rules to be written and tested without requiring modifications
+to the parser source code (significantly lowering the barrier to entry for
+new or casual users who need to parse new log lines), and greatly
+simplifies the framework, actions and rules. Decoupling the actions from
+the framework simplifies both framework and actions: the framework provides
+services to the actions, and doesn't need to deal with the complications
+which arise, or the task of reconstructing a mail's journey through
+Postfix; actions benefit from having services provided by the framework,
+freeing them to concentrate on the task of accurately reconstructing each
+mail's journey through Postfix and dealing with the complications arising
+(discussed in \sectionref{complications}).
 
@@
 
-The function tracks the number of times each rule matches; this data is
-used during subsequent parsing to optimise the order the rules are tried
-in.  Rules specify the Postfix component which produces the log lines they
-parse; this avoids needlessly trying rules which won't match the log line,
-or worse, might match unintentionally.
-
 \subsection{Actions}
@@
 to alert the user that they need to alter or extend their rules, hence the
-\texttt{IGNORE} action: the input has been parsed, but it is of no benefit
+\texttt{IGNORE} action: the input has been parsed, but it is of no benefit,
 so it is ignored.  The remaining actions have only one or two associated
@@
 work around the difficulties; the remainder were resolved by adapting
-actions to pass information to subsequent actions (via the framework).
+actions to pass information to subsequent actions (via the framework),
+which then check for the extra information and act appropriately.
 
@@
 
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}.
+    \item [regex] The \regex{} to match the log line against.  For
+        efficiency every rule's \regex{} is compiled before attempting to
+        parse the log file --- otherwise each \regex{} would be recompiled
+        each time it was used, resulting in a large, data dependent
+        slowdown.\footnote{This is discussed in more detail in the author's
+        thesis.}
 
@@
         log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
+        \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
         \newline i.e.\ semi-colon or comma separated assignment statements,
@@
         which isn't present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols}, except that
-        arbitrary data\footnote{Commas and semi-colons cannot be escaped
-        and thus cannot be used.  This is intended for use with small
-        amounts of data rather than large amounts in any one rule, so
-        dealing with escape sequences was deemed unnecessary.} is permitted
-        on the right hand side of the assignment.
+        \texttt{smtp\_code} when mail is accepted.  The format is the same
+        as for \texttt{result\_cols}, except that arbitrary
+        data\footnote{Commas and semi-colons cannot be escaped and thus
+        cannot be used.  This is intended for use with small amounts of
+        data rather than large amounts in any one rule, so dealing with
+        escape sequences was deemed unnecessary.} is permitted on the right
+        hand side of the assignment.
 
@@
 
-    \item [hits] This counter starts at zero and is incremented each time
-        the rule successfully matches.  At the start of each run the parser
-        sorts the rules in descending order of hits, and at the end of the
-        run updates every rule's hits.  Assuming that the distribution of
-        log lines is reasonably consistent between log files, rules
+    \item [hits] This counter is used to optimise the order in which rules
+        are tried against the input lines.  Assuming that the distribution
+        of log lines is reasonably consistent between log files, rules
         matching more commonly occurring log lines will be tried before
         rules matching less commonly occurring log lines, lowering the
-        program's execution time.  Rule ordering for efficiency is
-        discussed in \sectionref{rule ordering for efficiency}.
+        program's execution time.
 
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
-        more specific rules to take precedence over more general rules
+    \item [priority] This is the user-configurable companion to hits:
+        priority overrides hits when sorting rules.  This allows more
+        specific rules to take precedence over more general rules
         (described in \sectionref{overlapping rules}).
@@
 catchall rules.
+
+XXX MAYBE REMOVE THE SECTION ON DETECTING OVERLAPPING RULES\@?
 
@@
 
-Scalability, ordering, caching regexes.  Interesting to run with all rules
-versus running with the minimum number necessary to parse the test log
-files.  Then try that again with different orderings?  Coverage.
-
 Parsing efficiency is an obvious concern when the parser routinely needs to
@@
 
-MINIMUM VERSUS MAXIMUM NUMBER OF RULES\@.
+XXX MINIMUM VERSUS MAXIMUM NUMBER OF RULES\@.
 
@@
 \label{rule efficiency}
+
+XXX MINIMUM VERSUS MAXIMUM NUMBER OF RULES\@.
 
@@
 The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
+fig.~\refwithpage{execution time vs file size vs number of lines graph},
+and peaks in fig.~\refwithpage{execution time vs file size vs number lines
 factor} (where a peak means that more lines are processed per second, i.e.\
@@
 
+\label{coverage}
+
 \section{Conclusion}

### 2008-05-27 papers/sgai-2008/paper.tex
Added a chunk of logparser.pdf's coverage section.

@@
 
+The discussion of the parser's coverage of Postfix log files is separated
+into two parts: log lines covered and mails covered.  The first is
+important because the parser should handle all (relevant) log lines it's
+given; the second is equally important because the parser must properly
+deal with every mail if it is to be useful.  Improving the former is
+less intrusive, as it just requires new rules to be written; improving the
+latter is much more intrusive as it requires changes to the parser
+algorithm, and it can also be much harder to notice a deficiency.
+
+Warnings are issued for any log lines which are not parsed; no warnings are
+issued for unparsed log lines while testing with the \numberOFlogFILES{}
+test log files, so it can be safely concluded that there are zero false
+negatives.  False positives are harder to quantify: short of examining each
+of the 60,721,709 log lines and determining which \regex{} parsed it, there
+is no way to be sure that every line was parsed by the correct \regex{},
+making it impossible to quantify the false positive rate; however a random
+sample of 6039 log lines was parsed and the results checked manually to
+ensure that the correct \regex{} parsed each log line.
+
+Coverage of mails is much more difficult to determine accurately than
+coverage of log lines.  The parser can dump its state tables in a human
+readable form; examining these tables with reference to the log files is
+the best way to detect mails which were not handled properly.  The parser
+issues warnings when it detects any errors, some of which may alert the
+user to a problem.  There should be few or no warnings when parsing, and
+when finished parsing the state table should only contain entries for mails
+which had yet to be delivered when the log files ended, or were accepted
+before the log files began.
+
+Verifying by inspection that the parser correctly deals with all 60,721,709
+lines in the test log files is infeasible, but verifying a subset of those
+log files is a tractable, if extremely time consuming, task.  A sample of
+log lines was parsed with all debugging options enabled, resulting in
+167,448 lines of output.  All 167,448 lines were examined in conjunction
+with the input log and a dump of the resulting database, verifying that for
+each of the input lines the parser used the correct rule and executed the
+correct action, which in turn produced the correct result and inserted the
+correct data in the database.  The log segment produced 4 warnings, 10
+mails remaining in the state tables, and 1625 connections correctly entered
+in the database.
+
+Given the evidence detailed above, the author is confident that the false
+positive rate when reconstructing a mail is exceedingly low, if not
+approaching zero.
+
+Parser coverage is divided into two topics in this section: log lines
+covered and mails covered.\footnote{There is substantially more coverage
+discussion in the author's thesis.}  The former is initially more
+important, as the parser must successfully parse every line if it is to be
+complete, but subsequently the latter takes precedence because reproducing
+the path a mail takes through Postfix is the aim of the parser.  Increasing
+the percentage of log lines parsed is relatively simple and non-intrusive:
+adding new rules or modifying existing rules is simplified by the
+separation of rules, actions and framework.  Improving the logical coverage
+is harder, as the actions taken by Postfix must be reconstructed by the
+author, and the new sequence of actions integrated into the existing model
+without breaking the existing parsing.  Detecting a deficiency in the
+parsing algorithm is also significantly harder than detecting unparsed log
+lines, as the parser will warn about any unparsed line, whereas discovering
+a flaw in the parser requires understanding of the warnings produced and
+the mails remaining in the state table.  Rectifying a flaw in the parser
+requires an understanding of both the parser and Postfix's log files, and
+investigative work to determine the cause of the deficiency, followed by
+further examination of the log files in developing a solution.
+
 \section{Conclusion}

### 2008-05-28 doc/logparser-acronyms.tex
Corrections after proofreading.

@@
 \newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
-    SMTP is the protocol which transfers mail from the sender to the
-    recipient across the Internet.  %XXX A brief introduction to SMTP is
+    SMTP, originally defined in \RFC{}~821~\cite{RFC821} and updated in
+    \RFC{}~2821~\cite{RFC2821}, is the protocol which transfers mail from
+    the sender to the recipient across the Internet.  It is a simple, human
+    readable, plain text protocol, making it quite easy to test and debug
+    problems with it.  A detailed description of \SMTP{} is beyond the
+    scope of this document: introductory guides can be found
+    at~\cite{smtp-intro-01, smtp-intro-02}, the definitive references
+    are~\cite{RFC821, RFC2821}.
+    %XXX A brief introduction to SMTP is
     %provided in \sectionref{SMTP background}.
@@
     and maintenance.  Some information about SQL, including its history,
-    can be found at \cite{Wikipedia-sql}.
+    can be found at \cite{Wikipedia-sql}; a good introduction can be found
+    at~\cite{sql-for-web-nerds}, others are~\cite{w3schools-sql-tutorial,
+    sqlcourse.com}.
 }}
@@
 \newacronym{regex}{Regular Expression}{description={
-    Regular expressions are a method of matching patterns in text that are
-    explained in detail in~\cite{Wikipedia-regex, perlre}.
+    Regular expressions are a method of matching patterns against text that
+    are explained in detail in~\cite{Wikipedia-regex, perlre}.
 }}
@@
 \newacronym{ATN}{Augmented Transition Networks}{description={
-    Augmented Transition Networks are a tool used in Computational
-    Linguistics; they are originally described in~\cite{atns} and further
-    in~\cite{nlpip}.
+    Augmented Transition Networks, originally described in~\cite{atns} and
+    further in~\cite{nlpip}, are a tool used in Computational Linguistics
+    for creating grammars to parse or generate sentences,
 }}
@@
     English, Spanish, Japanese), with the goal of writing programs which
-    can understand human languages.  Potential applications include:
-    programs which will automatically summarise a text; programs which
-    can be taught facts, will understand questions and formulate answers to
-    those questions.  It is in no way connected to Neuro-linguistic
-    programming, which is generally found when searching for NLP.
+    can understand human languages.  It is in no way connected to
+    Neuro-linguistic programming, which is generally found when searching
+    for NLP.
 }}

### 2008-05-28 papers/sgai-2008/paper.tex
Corrections after proofreading.

@@
 % \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{
+\newcommand{\showgraph}[3]{%
     \begin{figure}[hbt!]
@@
     must allow users to easily adapt existing rules or add new rules to
-    parse new inputs, without requiring the user to have an in-depth
+    parse new inputs, without requiring that user have an in-depth
     knowledge and understanding of the parser's internal workings.  This
@@
     trivially add new rules or adapt existing rules (to parse variants of
-    existing inputs) and relatively easily add new actions (to do something
-    useful with a previously unknown class of input).  The architecture
-    provides a framework which actions and rules fit into: this framework
-    manages the parsing process, provides support functions for the
-    actions, and automatically performs some optimisations which improve
-    the speed of parsing.
+    existing inputs) and relatively easily add new actions (to process a
+    previously unknown class of input).  The architecture provides a
+    framework which actions and rules fit into: this framework manages the
+    parsing process, provides support functions for the actions, and
+    automatically performs some optimisations which improve the speed of
+    parsing.
 }
@@
 
-The architecture described herein was developed as part of a larger project to
-improve anti-spam techniques by analysing the performance of the set of filters
-currently in use, optimising the order and membership of the set based on that
-analysis, and developing new measures to supplement the existing filters where
-there is a deficiency.  The approach chosen was to analyse the log files
-produced by the mail server, rather than modifying the mail server to generate
-performance statistics; this approach improves the chances of other sites
-testing and utilising the software.  The need arose for a parser capable of
-dealing with the huge variety of log lines produced by Postfix~\cite{postfix},
-the \MTA{} in use:
+The architecture described herein was developed as part of a larger project
+to improve anti-spam techniques by analysing the performance of the set of
+filters currently in use, optimising the order and membership of the set
+based on that analysis, and developing new measures to supplement the
+existing filters where there is a deficiency.  The approach chosen to
+measure performance was to analyse the log files produced by the mail
+server, rather than modifying the mail server to generate performance
+statistics; this approach improves the chances of other sites testing and
+utilising the software.  The need arose for a parser capable of dealing
+with the huge variety of log lines produced by Postfix~\cite{postfix}, the
+mail server in use:
 
@@
         servers~\cite{policy-servers} or \RBL{} lookups) can also change
-        their messages, without warning.
+        their messages, sometimes without warning.
 
@@
 the purposes of this project.\footnote{A review of the existing parsers
-considered for this project will be included in the thesis.} The effort
-required to adapt and improve an existing parser was judged to be greater
-than the effort to write one from scratch, as the techniques used by the
-existing parsers severely limited their potential: some ignored the
+considered for this project will be included in the author's thesis.} The
+effort required to adapt and improve an existing parser was judged to be
+greater than the effort to write one from scratch, as the techniques used
+by the existing parsers severely limited their potential: some ignored the
 majority of log lines, parsing specific log lines with great accuracy, but
@@
 
-The solution developed is conceptually simple: provide a few generic functions,
-each capable of dealing with an entire \textit{category\/} of log lines (e.g.\
-rejections), accompanied by a multitude of precise patterns, each of which
-matches all log lines of a specific type, and only that type (e.g.\ rejection
-by a specific \RBLshort{}).
+The solution developed is conceptually simple: provide a few generic
+functions (\textit{actions\/}), each capable of dealing with an entire
+category of log lines (e.g.\ rejections), accompanied by a multitude of
+precise patterns (\textit{rules\/}), each of which matches all log lines of
+a specific type, and only that type (e.g.\ rejection by a specific
+\RBLshort{}).
 
@@
     \item [Framework]  The framework is the structure which actions and
-        rules fit into.  The framework provides the parsing loop, loading
-        and validation of rules, shared data storage, storage of results,
-        and other support functions.
+        rules fit into.  It provides the parsing loop, loading and
+        validation of rules, shared data storage, storage of results, and
+        other support functions.
 
@@
 
-Decoupling also creates a clear separation of functionality: rules handle low
-level details of identifying log lines and extracting data from a log line;
+Decoupling also creates a clear separation of functionality: rules handle
+low level details of identifying and extracting data from log lines;
 actions handle the higher level details of following the path a mail takes
@@
 rules.  It provides shared storage to pass data between actions, most
-notably the data structures used to represent each connection and mail,
-though there are some others.  It loads and validates rules, controls
-parsing of log files, invoking actions and tracking performance data, and
+notably the data structures used to represent each connection and mail.
+It loads and validates rules, controls
+parsing of log files, invokes actions, tracks how often each rule matches
+(to optimise rule ordering, see \sectionref{Rule ordering}), and
 stores results in the database on request.  The framework is independent of
@@
 
-The function which finds the rule matching the input line and invokes the
+The function which finds the rule matching the log line and invokes the
 requested action can be expressed in pseudo-code (with indentation denoting
@@
 
+The framework takes care of miscellaneous support functions and low level
+details of parsing, freeing the authors of actions to concentrate on
+writing productive code.  It links actions and rules, allowing either to be
+improved independently of the other.
+
 \subsection{Actions}
@@
 extending an existing parser.  In the Postfix parser developed for this
-project there are \numberOFactions{} actions and \numberOFrules{} rules.
-The distribution of rules to actions is very skewed, as can be seen in
+project there are \numberOFactions{} actions and \numberOFrules{} rules;
+the distribution of rules to actions is very skewed, as shown in
 Fig.~\refwithpage{Distribution of rules per action}.
@@
 the same reason the architecture is so successful in enabling users to
-parse new inputs: most new inputs can be parsed by a combination of a new
-rule and an existing action.  Unsurprisingly the action with the highest
-number of associated rules is \texttt{REJECTION}, the action which handles
-rejections; it's followed by \texttt{INFO}, the action responsible for
-handling informative inputs: these are log lines that supplement the data
-gathered from other log lines, but that don't themselves represent
-interesting events.  The third most common action is, perhaps surprisingly,
-\texttt{IGNORE}: there are many log lines which are not of interest to the
-parser, e.g.\ warning messages about \DNS{} mismatches, or log lines
-lacking a way to identify which data structure they should be associated
-with.\footnote{Postfix allocates a queueid (explained in the
+parse new log lines: most new log lines can be parsed by a combination of a
+new rule and an existing action.  Unsurprisingly the action with the
+highest number of associated rules is \texttt{REJECTION}, the action which
+handles rejections; it's followed by \texttt{INFO}, the action responsible
+for handling informative inputs: these are log lines that supplement the
+data gathered from other log lines (such as rejections), but that don't by
+themselves represent interesting events.  The third most common action is,
+perhaps surprisingly, \texttt{IGNORE}: there are many log lines which are
+not of interest to the parser, e.g.\ warning messages about DNS mismatches,
+or log lines lacking a way to identify which data structure they should be
+associated with.\footnote{Postfix allocates a queueid (explained in the
 glossary,~\sectionref{Glossary}) to each mail; most log lines contain the
-queueid of the associated mail, but some log lines, particularly with older
-versions of Postfix, lack a queueid, preventing their association with a
-particular data structure.}  The parser warns about every unparsed input,
-to alert the user that they need to alter or extend their rules, hence the
-\texttt{IGNORE} action: the input has been parsed, but it is of no benefit,
-so it is ignored.  The remaining actions have only one or two associated
-rules: in some cases this is because there will only ever be one type of
-input for that action, e.g.\ all log lines showing that a remote client has
-connected are matched by a single rule and handled by the \texttt{CONNECT}
-action; in other cases the action is required to address a deficiency in
-the log files, or a complication which arises during
-parsing.\footnote{\label{complications footnote}The complications which
-arose during development of the parser will be documented in detail in the
-author's thesis.}
+queueid of the associated mail, but some log lines (particularly with older
+versions of Postfix) lack a queueid, preventing their association with a
+particular data structure.}  The parser warns about every unparsed log
+line, to alert the user that they need to alter or extend their rules; the
+\texttt{IGNORE} action does nothing when executed, allowing uninteresting
+log lines to be parsed without any effects.  The remaining actions have
+only one or two associated rules: in some cases this is because there will
+only ever be one type of input for that action, e.g.\ all log lines showing
+that a remote client has connected are matched by a single rule and handled
+by the \texttt{CONNECT} action; in other cases the action is required to
+address a deficiency in the log files, or a complication which arises
+during parsing.\footnote{\label{complications footnote}The complications
+which arose during development of the parser will be documented in detail
+in the author's thesis.}
 
@@
 challenges, with all the attendant risks of adversely affecting the
-existing, working parser, a new, independent action can be written to
-satisfy the new requirements.  In some cases the new action will require
-the cooperation of other actions, e.g.\ to set or check a flag; this does
-introduce the possibility introducing failure by modifying existing
-actions, but modifications should be smaller and less frequent, and thus
-failures should be less likely and easier to test for and diagnose.  The
-parser can be written in an Object
+existing, working parser, when a new requirement arises a new, independent
+action can be written to satisfy it.  In some cases the new action will
+require the cooperation of other actions, e.g.\ to set or check a flag;
+there is a possibility of introducing failure by modifying existing
+actions, but modifications will be smaller and occur less frequently than
+with a monolithic architecture, and thus failures should be less likely and
+easier to test for and diagnose.  The parser can be written in an Object
 Oriented~\cite{Wikipedia-object-orientation} style, allowing sub-classes to
@@
         is \texttt{*} will be tried against any log lines which aren't
-        parsed by program specific rules.
+        parsed by program specific rules, allowing generic rules to be
+        written.
 
@@
         \texttt{smtp\_code} when mail is accepted.  The format is the same
-        as for \texttt{result\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
+        as for \texttt{result\_cols}, except that arbitrary data is
+        permitted on the right hand side of the assignment.
 
@@
 
-The architecture can be compared to existing approaches to aid in fully
-understanding it.
-
 \begin{description}
@@
         grammar terms the parser rules could be described as:
-
         $\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
@@
         the lack of results from previous actions can be useful
-        information, e.g.\ in the Postfix parser when a client disconnects
-        if there have been no useful log lines for that client the data is
-        discarded, rather than trying to store empty records in the
-        database.
+        information, XXX IS THE EXAMPLE NECESSARY\@?  E.g.\ in the Postfix
+        parser when a client disconnects if there have been no useful log
+        lines for that client the data is discarded, rather than trying to
+        store empty records in the database.
 
-    \item [Matching is simple:]  The first matching rule wins: there is no
-        backtracking to try alternate rules, no attempt made to pick a
-        \textit{best\/} rule; the first rule which matches the line is the
-        rule which parses it and determines the action that will be
-        invoked.\footnote{Rule authors can affect the order rules will be
-        tried in by setting the \texttt{priority} attribute of one or more
-        rules.}
+    \item [Matching rules to log lines is simple:]  The first matching rule
+        wins: there is no backtracking to try alternate rules, no attempt
+        is made to pick a \textit{best\/} rule; the first rule which
+        matches the line is the rule which parses it and determines the
+        action that will be invoked.
 
@@
         transduction: it takes data in one form (log files) and transforms
-        it to another form: a database in the case of the Postfix log
-        parser, though it could be normalised files or files which only
-        contain discrepancies or other interesting data.
+        it to another form (a database).  Other implementations are free to
+        choose the output format most suitable for their needs.
 
@@
         inputs must adhere to.  The architecture is designed for easy user
-        alteration, in fact such alteration is expected.
+        alteration, in fact such alteration is expected.  XXX NEED MORE
+        HERE\@.
+
+    \item [Line oriented:]  The architecture is line oriented at present:
+        there is no facility for rules to consume more input or push
+        unused input back onto the input stream.  This was not a deliberate
+        decision, rather a consequence of the line oriented nature of the
+        log files; other implementations could pursue a more flexible
+        approach.
 
@@
 a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
+mail servers would produce much larger log files on a daily basis).  When
 generating the data for the graphs included in this document,
@@
 log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
+execution time for the remaining 9 runs averaged.  XXX DO I NEED TO EXPLAIN
+WHY\@?  The first run is discarded for two reasons:
 
@@
 
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
+    \item [shuffle] Random ordering, intended to represent an unsorted rule
+        set.  The rules will be shuffled once before use and will retain
+        that ordering for the entirety of the log file.  Note that the
+        ordering will change every time the parser is executed, so 10
+        different orderings will be generated for each log file in the test
+        run.
 
@@
 Figure~\refwithpage{percentage increase of shuffled and reversed over
-normal} shows the percentage increase of execution times.  Overall this
-provides a modest but worthwhile performance increase of over 10\%, for a
-small investment in time and programming.
+normal} shows the percentage increase of execution times for shuffled and
+reversed rules.  Overall this optimisation provides a modest but worthwhile
+performance increase of over 10\%, for a small investment in time and
+programming.
 \showgraph{build/plot-normal-shuffle-reverse-factor}{Percentage increase of
-shuffled and reversed over normal}{percentage increase of shuffled and
-reversed over normal}
+execution time for different rule orderings}{percentage increase of
+shuffled and reversed over normal}
 

### 2008-05-28 doc/logparser-acronyms.tex
Lots of corrections after proofreading.

@@
     client; if the requested hostname exists the client is on the RBL, and
-    the mail server can decide what course of action to take.
+    the mail server can decide what course of action to take.  XXX
+    REFERENCES XXX.
 }}
@@
 \newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
-    SMTP, originally defined in \RFC{}~821~\cite{RFC821} and updated in
-    \RFC{}~2821~\cite{RFC2821}, is the protocol which transfers mail from
-    the sender to the recipient across the Internet.  It is a simple, human
-    readable, plain text protocol, making it quite easy to test and debug
-    problems with it.  A detailed description of \SMTP{} is beyond the
-    scope of this document: introductory guides can be found
-    at~\cite{smtp-intro-01, smtp-intro-02}, the definitive references
-    are~\cite{RFC821, RFC2821}.
-    %XXX A brief introduction to SMTP is
-    %provided in \sectionref{SMTP background}.
+    SMTP is the protocol which transfers mail from the sender to the
+    recipient across the Internet.  It is a simple, human readable, plain
+    text protocol, making it quite easy to test and debug problems with it.
+    A detailed description of \SMTP{} is beyond the scope of this document:
+    the original protocol definition is \RFC{}~821~\cite{RFC821}, updated
+    in \RFC{}~2821~\cite{RFC2821}; introductory guides can be found
+    at~\cite{smtp-intro-01,smtp-intro-02}.
 }}
@@
 \newacronym{regex}{Regular Expression}{description={
-    Regular expressions are a method of matching patterns against text that
-    are explained in detail in~\cite{Wikipedia-regex, perlre}.
+    Regular expressions are a powerful method of matching patterns against
+    text that are explained in detail in~\cite{Wikipedia-regex, perlre}.
 }}
@@
     SLCT~\cite{slct-paper} is a tool implementing an algorithm designed by
-    Risto Vaarandi~\cite{risto-vaarandi} for identifying and grouping
-    similar log files, then producing a regex for each group of lines.
+    Risto Vaarandi for identifying and grouping similar log files, then
+    producing a regex for each group of lines.
 }}
@@
     English, Spanish, Japanese), with the goal of writing programs which
-    can understand human languages.  It is in no way connected to
-    Neuro-linguistic programming, which is generally found when searching
-    for NLP.
+    can understand human languages.
 }}

### 2008-05-28 papers/sgai-2008/paper.tex
Lots of corrections after proofreading.

@@
 
-\usepackage[style=altlist,number=none,toc=true]{glossary}
+\usepackage[style=altlist,number=none,toc=true,global=true]{glossary}
 \makeglossary{}
@@
 \usepackage[bottom]{footmisc}
+
+% Tell Latex to use scalable fonts
+\usepackage{type1cm}
 
@@
 \institute{John Tobin \at{} School of Computer Science and Statistics,
-Trinity College, Dublin 2, Ireland. \newline{} \email{tobinjt@cs.tcd.ie} \and{}
-Carl Vogel \at{} School of Computer Science and Statistics,
+Trinity College, Dublin 2, Ireland. \newline{} \email{tobinjt@cs.tcd.ie}
+\and{} Carl Vogel \at{} School of Computer Science and Statistics,
 Trinity College, Dublin 2, Ireland. \newline{} \email{vogel@cs.tcd.ie}}
@@
     Some parsers need to be very precise and strict in what they parse, yet
-    must allow users to easily adapt existing rules or add new rules to
-    parse new inputs, without requiring that user have an in-depth
+    must allow users to easily adapt or extend the parser to parse new
+    inputs, without requiring that user have an in-depth
     knowledge and understanding of the parser's internal workings.  This
@@
     existing inputs) and relatively easily add new actions (to process a
-    previously unknown class of input).  The architecture provides a
-    framework which actions and rules fit into: this framework manages the
-    parsing process, provides support functions for the actions, and
-    automatically performs some optimisations which improve the speed of
-    parsing.
+    previously unknown class of input).  The architecture scales linearly
+    as the number of rules and/or size of input increases, making it
+    suitable for parsing large corpora or months of accumulated data.
 }
+
+XXX DOES THE ABSTRACT NEED TO BE LONGER\@?  IS IT A GOOD TITLE\@?
 
@@
 PAGES\@.  POSTER PAPERS\@: MAX 6 PAGES\@.  BLACK AND WHITE ONLY\@.
+
+XXX DO I NEED TO SAY THAT THINGS WILL BE EXPANDED UPON IN MY THESIS\@?
+
+XXX USE citeref AND A REFERENCE PACKAGE WHICH PUTS CLEVER PAGE NUMBERS ON
+REFERENCES\@.
+
+XXX CHECK EVERY URL IN THE BIBLIOGRAPHY\@.
 
@@
 utilising the software.  The need arose for a parser capable of dealing
-with the huge variety of log lines produced by Postfix~\cite{postfix}, the
-mail server in use:
+with the great number and variety of log lines produced by
+Postfix~\cite{postfix}, the \SMTP{} server in use.  The variety occurs for
+several reasons:
 
@@
     \item The mail administrator can define custom rejection messages,
-        resulting in custom log lines.
+        resulting in custom log lines~\cite{postfix-lookup-tables,
+        smtpd_access_readme, smtpd_per_user_control}.
 
@@
 \end{itemize}
+
+The parser must be designed so that adding support for parsing new log
+lines is a simple task, as the log lines will continue to evolve over time.
 
@@
 lines, but were incapable of distinguishing between log lines of the same
-type, e.g.\ rejections, and they extracted only a portion of the data
-required for this project.
+type, e.g.\ rejections, and extracted only a portion of the data required
+for this project.
 
@@
 The framework tries each rule in turn until it finds one which matches the
-input line, then invokes the action specified by that rule.
+input log line, then invokes the action specified by that rule.
 
@@
 the framework simplifies both framework and actions: the framework provides
-services to the actions, and doesn't need to deal with the complications
-which arise, or the task of reconstructing a mail's journey through
-Postfix; actions benefit from having services provided by the framework,
-freeing them to concentrate on the task of accurately reconstructing each
-mail's journey through Postfix and dealing with the complications arising
-(discussed in \sectionref{complications}).
+services to the actions, and doesn't need to perform any tasks specific to
+the input being parsed; actions benefit from having services provided by
+the framework, freeing them to concentrate on the task of accurately and
+correctly processing the information provided by rules.
 
@@
 low level details of identifying and extracting data from log lines;
-actions handle the higher level details of following the path a mail takes
-through Postfix, assembling the required data, dealing with complications
-arising, etc; the framework provides services to actions and stores data.
+actions handle the higher level details of assembling the required data,
+dealing with the intricacies of the input being parsed, complications
+arising, etc; the framework provides services to actions and manages the
+parsing process.
 
@@
 writing productive code.  It links actions and rules, allowing either to be
-improved independently of the other.
+improved independently of the other.  The framework is the core of the
+architecture and is deliberately quite simple: the rules deal with the
+varying log lines, and the actions deal with the intricacies and
+complications encountered when parsing.
 
@@
 
-Actions contain the code required to deal with log lines from a particular
-category, e.g.\ rejections.  The actions are parser-specific: each parser
-will need to write the actions it requires from scratch, unless it's
-extending an existing parser.  In the Postfix parser developed for this
-project there are \numberOFactions{} actions and \numberOFrules{} rules;
-the distribution of rules to actions is very skewed, as shown in
+Each action contains the code required to deal with log lines from a
+particular category, e.g.\ rejections.  The actions are parser-specific:
+each parser will need to write the actions it requires from scratch, unless
+it's extending an existing parser.  In the Postfix parser developed for
+this project there are \numberOFactions{} actions and \numberOFrules{}
+rules; the distribution of rules to actions is very skewed, as shown in
 Fig.~\refwithpage{Distribution of rules per action}.
@@
 highest number of associated rules is \texttt{REJECTION}, the action which
-handles rejections; it's followed by \texttt{INFO}, the action responsible
-for handling informative inputs: these are log lines that supplement the
-data gathered from other log lines (such as rejections), but that don't by
-themselves represent interesting events.  The third most common action is,
-perhaps surprisingly, \texttt{IGNORE}: there are many log lines which are
-not of interest to the parser, e.g.\ warning messages about DNS mismatches,
-or log lines lacking a way to identify which data structure they should be
-associated with.\footnote{Postfix allocates a queueid (explained in the
-glossary,~\sectionref{Glossary}) to each mail; most log lines contain the
-queueid of the associated mail, but some log lines (particularly with older
-versions of Postfix) lack a queueid, preventing their association with a
-particular data structure.}  The parser warns about every unparsed log
-line, to alert the user that they need to alter or extend their rules; the
-\texttt{IGNORE} action does nothing when executed, allowing uninteresting
-log lines to be parsed without any effects.  The remaining actions have
-only one or two associated rules: in some cases this is because there will
-only ever be one type of input for that action, e.g.\ all log lines showing
-that a remote client has connected are matched by a single rule and handled
-by the \texttt{CONNECT} action; in other cases the action is required to
-address a deficiency in the log files, or a complication which arises
-during parsing.\footnote{\label{complications footnote}The complications
-which arose during development of the parser will be documented in detail
-in the author's thesis.}
+handles rejections; it's followed by \texttt{SAVE\_BY\_QUEUEID}, the action
+responsible for handling informative inputs: these are log lines that
+supplement the data gathered from other log lines (such as rejections), but
+that don't by themselves represent interesting events.  The third most
+common action is, perhaps surprisingly, \texttt{IGNORE}: there are many log
+lines which are not of interest to the parser, e.g.\ warning messages about
+DNS mismatches, or log lines lacking a way to identify which data structure
+they should be associated with.\footnote{Postfix allocates a queueid
+(explained in the glossary,~\sectionref{Glossary}) to each mail; most log
+lines contain the queueid of the associated mail, but some log lines
+(particularly with older versions of Postfix) lack a queueid, preventing
+their association with a particular data structure.}  The parser warns
+about every unparsed log line, to alert the user that they need to alter or
+extend their rules; the \texttt{IGNORE} action does nothing when executed,
+allowing uninteresting log lines to be parsed without causing any effects.
+The remaining actions have only one or two associated rules: in some cases
+this is because there will only ever be one type of log line for that
+action, e.g.\ all log lines showing that a remote client has connected are
+matched by a single rule and handled by the \texttt{CONNECT} action; in
+other cases the action is required to address a deficiency in the log
+files, or a complication which arises during
+parsing.\footnote{\label{complications footnote}The complications which
+arose during development of the parser will be documented in detail in the
+author's thesis.}
 
@@
 accommodated in the parser: adding new actions was enough for several of
-the complications; others required modifications to existing actions to
-work around the difficulties; the remainder were resolved by adapting
-actions to pass information to subsequent actions (via the framework),
-which then check for the extra information and act appropriately.
+the complications; others required modifications to a single existing
+action to work around the difficulties; the remainder were resolved by
+adapting actions to pass information to subsequent actions (via the
+framework), which then check for the extra information and change their
+behaviour appropriately.
 
@@
 
-Rules are responsible for categorising input lines, specifying both the
-data to be extracted from each line and the action that should be invoked.
+Rules are responsible for categorising inputs, specifying both the data to
+be extracted from each log line and the action that should be invoked.
 This implementation stores the rules in the same \SQL{} database that the
@@
 
-    \item [program] The Postfix program whose log lines the rule applies
-        to.  This avoids needlessly trying rules which won't match the log
-        line, or worse, might match unintentionally.  Rules whose program
-        is \texttt{*} will be tried against any log lines which aren't
+    \item [program] The Postfix program whose log lines the rule should be
+        applied to.  This avoids needlessly trying rules that won't match
+        the log line, or worse, might match unintentionally.  Rules whose
+        program is \texttt{*} will be tried against any log lines not
         parsed by program specific rules, allowing generic rules to be
@@
         each time it was used, resulting in a large, data dependent
-        slowdown.\footnote{This is discussed in more detail in the author's
-        thesis.}
+        slowdown.\footnote{This will be discussed in more detail in the
+        author's thesis.}  XXX MAYBE SCRAP THE SECOND SENTENCE\@?
 
@@
         with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.
+        the \regex{} on the right.
 
@@
     \item [action] The action that will be invoked when this rule matches a
-        log line; actions are described in \sectionref{Actions}.
+        log line.
 
@@
         of log lines is reasonably consistent between log files, rules
-        matching more commonly occurring log lines will be tried before
-        rules matching less commonly occurring log lines, lowering the
-        program's execution time.
+        matching more frequently will be tried before rules matching less
+        frequently, lowering the parser's execution time.
 
@@
 
-\subsubsection{Overlapping rules}
-
 \label{overlapping rules}
 
-XXX DO I NEED A SECTION ABOUT OVERLAPPING RULES\@?
+XXX DO I NEED TO DISCUSS OVERLAPPING RULES\@?
 
@@
 difficult task.} Unintentionally overlapping rules lead to inconsistent
-parsing and data extraction because the order in which rules are tried
-against each line may change between log files, and the first matching rule
-wins.  Overlapping rules are frequently a requirement, allowing a more
+parsing and data extraction because the first matching rule wins and the
+order in which rules are tried against each line may change between log
+files.  Overlapping rules are frequently a requirement, allowing a more
 specific rule to match some log lines and a more general rule to match the
@@
 delivery to the rest of the world.  The algorithm provides a facility for
-ordering overlapping rules: the priority field in each rule (defaults to
-zero).  Rules are sorted by priority, highest first, and then rules with
-the same priority are sorted by the number of successful matches when
-parsing the previous log file.  Negative priorities may be useful for
+specifying the order of overlapping rules: the priority field in each rule
+(defaults to zero).  Rules are sorted by priority, highest first, and then
+rules with the same priority are sorted by the number of successful matches
+when parsing the previous log file.  Negative priorities may be useful for
 catchall rules.
@@
         grammar terms the parser rules could be described as:
-        $\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
+        $\text{\textless{}input\textgreater{}} \mapsto \text{rule-1} |
         \text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
@@
     \item [Partially context-sensitive actions:]  XXX IS THIS THE CORRECT
-        TERMINOLOGY\@?  Actions can consult the results of previous actions
-        when executing, providing some context sensitivity.  In some cases
-        the lack of results from previous actions can be useful
-        information, XXX IS THE EXAMPLE NECESSARY\@?  E.g.\ in the Postfix
-        parser when a client disconnects if there have been no useful log
-        lines for that client the data is discarded, rather than trying to
-        store empty records in the database.
+        TERMINOLOGY\@?  Actions can consult the results (or lack of
+        results) of previous actions when executing, providing some context
+        sensitivity.  Actions can not inspect past or upcoming inputs.
 
@@
         matches the line is the rule which parses it and determines the
-        action that will be invoked.
+        action that will be invoked.  The priority attribute may be used to
+        change the ordering of rules.
 
@@
         transduction: it takes data in one form (log files) and transforms
-        it to another form (a database).  Other implementations are free to
-        choose the output format most suitable for their needs.
+        it to another form (a database); other output formats may be more
+        suitable for other implementations.
 
@@
         this architecture does not have a fixed grammar specification which
-        inputs must adhere to.  The architecture is designed for easy user
-        alteration, in fact such alteration is expected.  XXX NEED MORE
-        HERE\@.
+        inputs must adhere to.  The architecture is capable of dealing with
+        overlapping inputs, out of order inputs, and ambiguous inputs where
+        heuristics must be applied --- all have arisen and been
+        successfully dealt with in the Postfix log parser.
 
@@
 \numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  XXX DO I NEED TO EXPLAIN
-WHY\@?  The first run is discarded for two reasons:
+log lines) were each parsed 10 times, the results from first run discarded,
+and the execution time for the remaining 9 runs averaged.  XXX DO I NEED TO
+EXPLAIN WHY\@?  The first run is discarded for two reasons:
 
@@
 
-XXX MINIMUM VERSUS MAXIMUM NUMBER OF RULES\@.
-
 An important property of a parser is how execution time scales relative to
@@
 Figure~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
+shows the execution time in seconds, file size in MB, and tens of thousands
 of log lines per log file.  All three lines run roughly in parallel, giving
@@
 size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better).  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in fig.~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
+execution time, and ratio of number of log lines vs execution time (higher
+is better in both cases).  As the reader can see the ratios are quite
+tightly banded across the graph, showing that the algorithm scales
+linearly: the much larger log files between points 60 and 70 on the X axis
+in fig.~\refwithpage{execution time vs file size vs number of lines graph}
+actually cause the ratio to increase (i.e.\ improve), rather than decrease.
+The strange behaviour where larger log files are parsed more efficiently is
+explained in \sectionref{Why are there dips in the graphs?}.
 
@@
 number lines factor}
+
+XXX MINIMUM VERSUS MAXIMUM NUMBER OF RULES\@.
 
@@
 top 10\% matching the vast majority of the log lines, and the remaining log
-lines split across the other 90\% of the rules (as shown in
-fig.~\refwithpage{rule hits graph}).  \showgraph{build/plot-hits}{Hits per
-rule}{rule hits graph} Assuming that the distribution of log lines is
-reasonably steady over time, program efficiency should benefit from trying
-more frequently matching rules before those which match less frequently.
-To test this hypothesis three full test runs were performed with different
-rule orderings:
+lines split across the other 90\% of the rules in a Power Law
+distribution~\cite{powerlaw} (as shown in fig.~\refwithpage{rule hits
+graph}).  \showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
+Assuming that the distribution of log lines is reasonably steady over time,
+parser efficiency should benefit from trying more frequently matching rules
+before those which match less frequently.  To test this hypothesis three
+full test runs were performed with different rule orderings:
 
@@
 
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
+    \item [normal]  Hypothetically the most optimal order: rules which
+        match most often will be tried first.
 
@@
 
-\subsubsection{Why are there dips in the graphs?}
+\noindent\textbf{Why are there dips in the graphs?}
+
 \label{Why are there dips in the graphs?}
 
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-fig.~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in fig.~\refwithpage{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.  Fewer
+The dips at log files 22 and 62--68 in fig.~\refwithpage{percentage
+increase of shuffled and reversed over normal} correspond to peaks in log
+file size in fig.~\refwithpage{execution time vs file size vs number of
+lines graph}, and peaks in fig.~\refwithpage{execution time vs file size vs
+number lines factor} (where a peak means that more lines are processed per
+second, i.e.\ performance is better).  The explanation for this took some
+time to arrive at, but it turns out to be reasonably simple.  The large log
+files in question were caused by a mail forwarding loop, where the
+distribution of log lines is quite different to normal, resulting in
+different performance characteristics.\footnote{The cause of the mail
+forwarding loop will be explained in the author's thesis.}
 
-The vast majority of log lines when a mail loop occurs are from Postfix
+XXX REWRITE THIS NEXT PARAGRAPH SO IT DOESN'T MENTION postfix/smtpd.  The
+vast majority of log lines when a mail loop occurs are from Postfix
 components which have a small number of rules associated with them, whereas
@@
 rules much more evenly than the log lines which occur frequently during a
-mail loop, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
+mail loop, so the average number of rules required to parse a
+\daemon{smtpd} log line is much higher that the average number required to
+parse other log lines.
 
@@
 peaks in fig.~\refwithpage{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in fig.~\refwithpage{percentage increase of
-shuffled and reversed over normal}.  The number of rules which need to be
-consulted when the ordering is shuffled varies between the optimum and
-nadir, and the performance varies proportionally.
+factor} and troughs in fig.~\refwithpage{percentage increase of shuffled
+and reversed over normal}.  When the rule ordering is reversed the majority
+of log lines generated by a mail loop will be parsed using very few rules,
+because there are very few rules for the Postfix components producing the
+log lines.  Without a mail loop the Postfix components producing the
+majority of log lines have a large number of rules associated with them,
+and reverse ordering deliberately starts with the least likely rules and
+moves towards the most likely.  This leads to a noticeable drop in the
+average time required to parse a log line during a mail loop, as shown in
+fig.~\refwithpage{percentage increase of shuffled and reversed over
+normal}.  The number of rules which need to be consulted when the ordering
+is shuffled varies between the optimum and the nadir, and the performance
+varies proportionally.
 
@@
 \label{coverage}
+
+XXX THIS NEEDS TO BE IMPROVED AND PROOFREAD\@.
 
@@
 Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.\footnote{There is substantially more coverage
-discussion in the author's thesis.}  The former is initially more
+covered and mails covered.\footnote{There will be substantially more
+coverage discussion in the author's thesis.}  The former is initially more
 important, as the parser must successfully parse every line if it is to be
@@
 \section{Conclusion}
+
+XXX WRITE THE CONCLUSION\@.
 

### 2008-05-29 papers/sgai-2008/paper.tex
Use varioref for references; still using the same format for distant references,     switched to the better format provided by varioref for close references. Removed tocbibind: no point in using it in this document, and it gives a smaller     bibliography. Minor other changes.

@@
 \usepackage[final]{graphicx}
-\renewcommand{\refname}{Bibliography}
+%\renewcommand{\refname}{Bibliography}
 % Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
+%\usepackage[section,numbib]{tocbibind}
 % Provides commands to distinguish between pdf and dvi output.
@@
 \usepackage{multicol}
+
+% Nicer citation lists - smaller spaces.
 \usepackage{cite}
+% Adds page numbers to bibliography entries, showing where they were cited.
+% Breaks when cite is used.
+% \usepackage{citeref}
+% Check for unused references.
+% \usepackage{refcheck}
+\usepackage{varioref}
+% Warnings rather than errors when references cross page boundaries.
+\vrefwarning{}
+% Errors when loops are encountered
+\vrefshowerrors{}
+% Show all the files/packages loaded.
+%\listfiles
 
@@
 \newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [p.~\pageref{#1}]%
+    %\empty{}\ref{#1} [p.~\pageref{#1}]%
+    \empty{}\vref{#1}%
+}
+% Shorten the text used for references with page numbers.
+\renewcommand{\reftextfaraway}[1]{%
+    [p.~\pageref{#1}]%
 }
@@
 \newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
+    %\textsection{}\refwithpage{#1}%
+    \textsection{}\vref*{#1}%
 }
@@
 XXX DO I NEED TO SAY THAT THINGS WILL BE EXPANDED UPON IN MY THESIS\@?
-
-XXX USE citeref AND A REFERENCE PACKAGE WHICH PUTS CLEVER PAGE NUMBERS ON
-REFERENCES\@.
 
@@
 tightly banded across the graph, showing that the algorithm scales
-linearly: the much larger log files between points 60 and 70 on the X axis
-in fig.~\refwithpage{execution time vs file size vs number of lines graph}
+linearly: the much larger log files at points 62--68 on the X axis in
+fig.~\refwithpage{execution time vs file size vs number of lines graph}
 actually cause the ratio to increase (i.e.\ improve), rather than decrease.

### 2008-05-29 papers/sgai-2008/paper.tex
Removed a lot of content to shorten the paper.

@@
 
+\today{}
+
 XXX DOES THE ABSTRACT NEED TO BE LONGER\@?  IS IT A GOOD TITLE\@?
@@
 PAGES\@.  POSTER PAPERS\@: MAX 6 PAGES\@.  BLACK AND WHITE ONLY\@.
-
-XXX DO I NEED TO SAY THAT THINGS WILL BE EXPANDED UPON IN MY THESIS\@?
 
@@
 
-    \item The mail administrator can define custom rejection messages,
-        resulting in custom log lines~\cite{postfix-lookup-tables,
-        smtpd_access_readme, smtpd_per_user_control}.
+    \item The mail administrator can define custom rejection
+        messages~\cite{postfix-lookup-tables, smtpd_access_readme,
+        smtpd_per_user_control}, resulting in custom log lines.
 
@@
 their parsing too inexact, and/or they did not extract sufficient data for
-the purposes of this project.\footnote{A review of the existing parsers
-considered for this project will be included in the author's thesis.} The
-effort required to adapt and improve an existing parser was judged to be
-greater than the effort to write one from scratch, as the techniques used
-by the existing parsers severely limited their potential: some ignored the
-majority of log lines, parsing specific log lines with great accuracy, but
-in an inextensible fashion; others sloppily parsed the majority of log
-lines, but were incapable of distinguishing between log lines of the same
-type, e.g.\ rejections, and extracted only a portion of the data required
-for this project.
+the purposes of this project.  The effort required to adapt and improve an
+existing parser was judged to be greater than the effort to write one from
+scratch, as the techniques used by the existing parsers severely limited
+their potential: some ignored the majority of log lines, parsing specific
+log lines with great accuracy, but in an inextensible fashion; others
+sloppily parsed the majority of log lines, but were incapable of
+distinguishing between log lines of the same type, e.g.\ rejections, and
+extracted only a portion of the data required for this project.
 
@@
 rules.  It provides shared storage to pass data between actions, most
-notably the data structures used to represent each connection and mail.
-It loads and validates rules, controls
-parsing of log files, invokes actions, tracks how often each rule matches
-(to optimise rule ordering, see \sectionref{Rule ordering}), and
-stores results in the database on request.  The framework is independent of
-the parser being written: the same functionality will be required from the
-framework for most or all parsers.
+notably the data structures used to represent each connection and mail.  It
+loads and validates rules, controls parsing of log files, invokes actions,
+tracks how often each rule matches (to optimise rule ordering, see
+\sectionref{Rule ordering}), and stores results in the database on request.
+The framework is independent of the parser being written: the same
+functionality will be required from the framework for most or all parsers.
 
@@
 requested action can be expressed in pseudo-code (with indentation denoting
-flow of control) as:
+flow of control):
+
+% DO NOT REFORMAT!
 
 \begin{verbatim}
-
 for each line in the input files: 
@@
     warn the user that the input line was not parsed
-
 \end{verbatim}
@@
 other cases the action is required to address a deficiency in the log
-files, or a complication which arises during
-parsing.\footnote{\label{complications footnote}The complications which
-arose during development of the parser will be documented in detail in the
-author's thesis.}
+files, or a complication which arises during parsing.
 
@@
 obvious variety in log lines, there were a large number of complications to
-be overcome.\footref{complications footnote}  Some were the result of
-deficiencies in Postfix's logging, which were improved or resolved by later
-versions of Postfix; others were due to the vagaries of process scheduling,
-client behaviour, and administrative actions.  All were successfully
-accommodated in the parser: adding new actions was enough for several of
-the complications; others required modifications to a single existing
-action to work around the difficulties; the remainder were resolved by
-adapting actions to pass information to subsequent actions (via the
-framework), which then check for the extra information and change their
-behaviour appropriately.
+be overcome.  Some were the result of deficiencies in Postfix's logging,
+which were improved or resolved by later versions of Postfix; others were
+due to the vagaries of process scheduling, client behaviour, and
+administrative actions.  All were successfully accommodated in the parser:
+adding new actions was enough for several of the complications; others
+required modifications to a single existing action to work around the
+difficulties; the remainder were resolved by adapting actions to pass
+information to subsequent actions (via the framework), which then check for
+the extra information and change their behaviour appropriately.
 
@@
 
-    \item [regex] The \regex{} to match the log line against.  For
-        efficiency every rule's \regex{} is compiled before attempting to
-        parse the log file --- otherwise each \regex{} would be recompiled
-        each time it was used, resulting in a large, data dependent
-        slowdown.\footnote{This will be discussed in more detail in the
-        author's thesis.}  XXX MAYBE SCRAP THE SECOND SENTENCE\@?
+    \item [regex] The \regex{} to match the log line against.
 
-    \item [result\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right.
+    \item [result\_cols] Specifies how the fields in the log line will be
+        extracted.  The format is:
+        \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline
+        i.e.\ semi-colon or comma separated assignment statements, with the
+        variable name on the left and the matching capture from the
+        \regex{} on the right.
 
@@
 
-XXX MAYBE REMOVE THE SECTION ON DETECTING OVERLAPPING RULES\@?
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
-
-\end{itemize}
-
 \subsection{Architecture characteristics}
@@
     \item [Line oriented:]  The architecture is line oriented at present:
-        there is no facility for rules to consume more input or push
-        unused input back onto the input stream.  This was not a deliberate
+        there is no facility for rules to consume more input or push unused
+        input back onto the input stream.  This was not a deliberate
         decision, rather a consequence of the line oriented nature of the
@@
 log lines) were each parsed 10 times, the results from first run discarded,
-and the execution time for the remaining 9 runs averaged.  XXX DO I NEED TO
-EXPLAIN WHY\@?  The first run is discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
+and the execution time for the remaining 9 runs averaged.
 
@@
 explained in \sectionref{Why are there dips in the graphs?}.
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
+\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file size
+vs number of lines}{execution time vs file size vs number of lines graph}
+\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file size
+and number of lines to execution time}{execution time vs file size vs
 number lines factor}
@@
 distribution of log lines is quite different to normal, resulting in
-different performance characteristics.\footnote{The cause of the mail
-forwarding loop will be explained in the author's thesis.}
+different performance characteristics.
 
@@
 given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
+deal with every mail if it is to be useful.  Improving the former is less
+intrusive, as it just requires new rules to be written; improving the
 latter is much more intrusive as it requires changes to the parser
@@
 Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.\footnote{There will be substantially more
-coverage discussion in the author's thesis.}  The former is initially more
-important, as the parser must successfully parse every line if it is to be
-complete, but subsequently the latter takes precedence because reproducing
-the path a mail takes through Postfix is the aim of the parser.  Increasing
-the percentage of log lines parsed is relatively simple and non-intrusive:
+covered and mails covered.  The former is initially more important, as the
+parser must successfully parse every line if it is to be complete, but
+subsequently the latter takes precedence because reproducing the path a
+mail takes through Postfix is the aim of the parser.  Increasing the
+percentage of log lines parsed is relatively simple and non-intrusive:
 adding new rules or modifying existing rules is simplified by the
@@
 
-The architecture's greatest strength is the ease with which new inputs
-can be parsed and new actions written.
+The architecture's greatest strength is the ease with which new inputs can
+be parsed and new actions written.
 

### 2008-05-29 doc/logparser-acronyms.tex
Changed RBL to DNSBL, and added a reference for it.

@@
 
-\newacronym{RBL}{Real-time Black List}{description={
-    Real-time Black Lists are a simple collaborative anti-spam technique
-    used to reject or penalise email sent from mail servers reported to
-    have sent large volumes of spam.  To use an RBL the mail server makes a
-    DNS request incorporating the IP address of the currently connected
-    client; if the requested hostname exists the client is on the RBL, and
-    the mail server can decide what course of action to take.  XXX
-    REFERENCES XXX.
+\newacronym{DNSBL}{DNS Blacklist}{description={
+    A DNS Blacklist~\cite{Wikipedia-DNSBL} is a simple collaborative
+    anti-spam technique used to reject or penalise email sent from mail
+    servers reported to have sent large volumes of spam.  To use a DNSBL
+    the mail server makes a DNS request incorporating the IP address of the
+    currently connected client; if the requested hostname exists the client
+    is on the DNSBL, and the mail server can decide what course of action
+    to take.
 }}

### 2008-05-29 doc/logparser-background.tex
Misc notes and minor changes.
Changed RBL to DNSBL, and added a reference for it.

@@
 determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client rbl.example.com} checks whether the \IP{}
-address of the client is present in the \RBL{} \texttt{rbl.example.com},
-rejecting the command if the client is listed.
+\texttt{reject\_rbl\_client dnsbl.example.com} checks whether the \IP{}
+address of the client is present in the \DNSBL{}
+\texttt{dnsbl.example.com}, rejecting the command if the client is listed.
 This description is necessarily brief, for further details
@@
 
+XXX CHANGE THIS; THE PARSER DESIGN IS NOVEL, DON'T STRESS THE SIMILARITY
+TOO MUCH\@.
+
 There is some similarity between the parser's design and William Wood's

### 2008-05-29 doc/logparser-coverage.tex
Changed RBL to DNSBL, and added a reference for it.

@@
 rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \RBL{} messages
+servers have different restrictions with custom messages, \DNSBL{} messages
 change over time, major releases of Postfix change warning messages

### 2008-05-29 doc/logparser-introduction.tex
Changed RBL to DNSBL, and added a reference for it.

@@
 messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
+\DNSBL{}\footnote{This document is supplied with a glossary, see
 appendix~\ref{Glossary}.} returns a different explanatory message; policy

### 2008-05-29 doc/logparser-other-parsers-review.tex
Changed RBL to DNSBL, and added a reference for it.

@@
 \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
-extracted: from, to, \RBL{}, client hostname and client \IP{}\@.
+extracted: from, to, RBL, client hostname and client \IP{}\@.
 
@@
 extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, \RBL{} and message id.
+\IP{}, status, response code, RBL and message id.
 
@@
 data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \RBL{} check.
+rejections except when the rejection is caused by an \DNSBL{} check.
 

### 2008-05-29 doc/logparser-parsing-rules.tex
Changed RBL to DNSBL, and added a reference for it.

@@
 because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \RBL{}
+restrictions which suits their needs, including using different \DNSBL{}
 services, policy servers, or custom rejection messages.  To facilitate easy

### 2008-05-29 papers/sgai-2008/paper.tex
Checked every URL in the bibliography.
Changed RBL to DNSBL, and added a reference for it.

@@
 
-XXX CHECK EVERY URL IN THE BIBLIOGRAPHY\@.
+XXX IN THE INTRODUCTION\@: WHY IS THIS RELEVANT TO AN AI CONFERENCE\@?
+
+XXX MAYBE MENTION THAT LOG LINES COULD BE MODIFIED AND RE-PARSED\@?  MAYBE
+WORK IT INTO THE ARCHITECTURE CHARACTERISTICS SECTION, AND ELSEWHERE\@.
+
+XXX MAYBE ADD AN EXAMPLE RULE AND LOG LINE, IF THERE'S SUFFICIENT SPACE\@.
 
@@
     \item External resources utilised by Postfix (e.g.\ policy
-        servers~\cite{policy-servers} or \RBL{} lookups) can also change
+        servers~\cite{policy-servers} or \DNSBL{} lookups) can also change
         their messages, sometimes without warning.
@@
 a specific type, and only that type (e.g.\ rejection by a specific
-\RBLshort{}).
+\DNSBL{}).
 
@@
 
+System could be used by mail administrators to monitor restrictions.
+
+Explain how someone could build upon it, e.g.\ to parse further Postfix
+stuff, or to parse other logs entirely.
+
+``This is a real world application, based on a real world need, rather than
+a toy application built as a proof of concept''  Put this in the
+introduction as well.
+
 \bibliographystyle{spmpsci}

### 2008-05-30 doc/logparser.tex
Better looking output from \refwithpage{} command.
Misc notes and minor changes.
Lots of corrections after proofreading.
Converted normal-vs-smaller graph to show the increase in execution time rather     than the decrease. Wrote about number of rules scalability. Wrapped labels on some graphs so they don't extend past the edge of the graph.

@@
 
-\usepackage[acronym=true,style=altlist,number=none,toc=true]{glossary}
+% global=true prevents acronyms being expanded twice
+\usepackage[global=true,acronym=true,style=altlist,number=none,toc=true]{glossary}
 \makeglossary{}
@@
 
+% Sort numbers where there are multiple citations.  Doesn't appear to have
+% any effect, though it does reduce the space between numbers.
+\usepackage{cite}
+
 % \showgraph{filename}{caption}{label}
@@
 \newcommand{\numberOFrules}[0]{%
-    165%
+    169%
 }
@@
 
+XXX IS THERE A SYMBOL FOR APPENDIX, LIKE \textsection{} FOR SECTIONS\@?
+
+XXX CHECK Y-AXIS LABELS ON GRAPHS TO ENSURE THEY CAN BE READ, AND WRAP IF
+NECESSARY\@.
+
 \newpage

### 2008-05-30 papers/sgai-2008/paper.tex
Converted normal-vs-smaller graph to show the increase in execution time rather     than the decrease. Wrote about number of rules scalability. Wrapped labels on some graphs so they don't extend past the edge of the graph.

@@
     169%
+}
+
+\newcommand{\numberOFrulesMINIMUM}[0]{%
+    119 XXX CHECK THIS NUMBER%
+}
+
+% \numberOFrulesMINIMUM as percentage of \numberOFrules
+\newcommand{\numberOFrulesMINIMUMpercentage}[0]{%
+    70.41\%%
+}
+
+% \numberOFrules as percentage increase of \numberOFrulesMINIMUM
+\newcommand{\numberOFrulesMAXIMUMpercentage}[0]{%
+    42.01\%%
 }
@@
 
-\subsection{Architecture Scalability}
+\subsection{Architecture Scalability: Input size}
 
@@
 
+\subsection{Architecture Scalability: Number of Rules}
+
+How any architecture scales as the number of rules increases is important,
+but it is particularly important in this architecture because it is
+expected that there will be a large number of rules.  There are
+\numberOFrules{} rules in the full ruleset, whereas the minimum number of
+rules required to parse the \numberOFlogFILES{} log files is
+\numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the total.  A
+second set of statistics was generated using the minimum ruleset and the
+results compared to the statistics generated using the full ruleset: the
+percentage execution increase when using the full ruleset instead of the
+minimal ruleset for normal, shuffled and reversed orderings is shown in
+fig.~\refwithpage{Percentage execution time increase of maximum ruleset
+over minimum ruleset}.  \showgraph{build/plot-normal-vs-smaller}{Percentage
+execution time increase of maximum ruleset over minimum ruleset}{Percentage
+execution time increase of maximum ruleset over minimum ruleset}
+
+It is clear from fig.~\refwithpage{Percentage execution time increase of
+maximum ruleset over minimum ruleset} that the increased number of rules
+makes a noticeable difference with the reverse ordering, and to a lesser
+extent with the shuffled ordering.  The normal ordering (where rules are
+sorted so that the most frequently matching rules are tried first) shows a
+\input{build/normal-vs-smaller-percentage-increase.tex} increase in
+execution time for a \numberOFrulesMAXIMUMpercentage{} increase in the
+number of rules.  These results show that the architecture scales very well
+as the number of rules increases, and that sorting the rules is an
+important optimisation enabling the parser to scale well.
+
 \subsection{Coverage}

### 2008-05-30 doc/logparser-acronyms.tex
Improved coverage section - it was better than I had thought. Rewrote parts of the dips in the graph explanation. Minor changes to shorten the paper.

@@
     the mail server makes a DNS request incorporating the IP address of the
-    currently connected client; if the requested hostname exists the client
-    is on the DNSBL, and the mail server can decide what course of action
-    to take.
+    client; if the requested hostname exists the client is on the DNSBL,
+    and the mail server can decide what course of action to take.
 }}
@@
     SLCT~\cite{slct-paper} is a tool implementing an algorithm designed by
-    Risto Vaarandi for identifying and grouping similar log files, then
-    producing a regex for each group of lines.
+    Risto Vaarandi for identifying, grouping and producing a regex to match
+    similar log lines.
 }}

### 2008-05-30 papers/sgai-2008/paper.tex
Improved coverage section - it was better than I had thought. Rewrote parts of the dips in the graph explanation. Minor changes to shorten the paper.

@@
 
-XXX MINIMUM VERSUS MAXIMUM NUMBER OF RULES\@.
-
 \subsection{Rule ordering}
@@
 \label{rule efficiency}
-
-XXX MINIMUM VERSUS MAXIMUM NUMBER OF RULES\@.
 
@@
 lines graph}, and peaks in fig.~\refwithpage{execution time vs file size vs
-number lines factor} (where a peak means that more lines are processed per
-second, i.e.\ performance is better).  The explanation for this took some
-time to arrive at, but it turns out to be reasonably simple.  The large log
-files in question were caused by a mail forwarding loop, where the
-distribution of log lines is quite different to normal, resulting in
-different performance characteristics.
-
-XXX REWRITE THIS NEXT PARAGRAPH SO IT DOESN'T MENTION postfix/smtpd.  The
-vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines which occur frequently during a
-mail loop, so the average number of rules required to parse a
-\daemon{smtpd} log line is much higher that the average number required to
-parse other log lines.
+number lines factor} (where a peak means that performance is better).  The
+explanation for this took some time to arrive at, but it turns out to be
+reasonably simple.  The large log files in question were caused by a mail
+forwarding loop, where the distribution of log lines is quite different to
+normal, resulting in different performance characteristics.  The vast
+majority of log lines when a mail loop occurs are from Postfix components
+which have a small number of rules associated with them, whereas in general
+the majority of log lines are from the Postfix component (\daemon{smtpd})
+with the highest number of rules.  Furthermore in normal log files the
+majority of log lines are distributed much more evenly across rules than
+the majority of log lines during a mail loop.
 
@@
 normal}.  The number of rules which need to be consulted when the ordering
-is shuffled varies between the optimum and the nadir, and the performance
-varies proportionally.
+is shuffled varies between the optimal and the pessimal, with the
+performance varying proportionally.
 
@@
 
-XXX THIS NEEDS TO BE IMPROVED AND PROOFREAD\@.
+The parser's coverage of Postfix log files is separated into two parts: log
+line coverage and mail coverage.  The first is important because the parser
+should handle all input log lines; the second is equally important because
+the parser must properly reconstruct every mail if it is to be useful.
+Improving the former is less intrusive, as it just requires new rules to be
+written; improving the latter is more intrusive as it requires changes to
+actions, and it can also be much harder to notice a deficiency.
 
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is less
-intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
-
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
-making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.
+The framework issues warnings for any log lines which are not parsed; no
+warnings are issued for unparsed log lines while parsing the
+\numberOFlogFILES{} test log files, so it can be safely concluded that
+there are zero false negatives.  False positives are harder to quantify:
+manually verifying that the correct \regex{} parsed each of the 60,721,709
+log lines is infeasible, making it impossible to quantify the false
+positive rate; however a random sample of 6039 log lines (0.00994\% of the
+total) was parsed and the results verified to ensure that the correct
+\regex{} parsed each log line.
 
@@
 which had yet to be delivered when the log files ended, or were accepted
-before the log files began.
+before the log files began.  Verifying by inspection that the parser
+correctly deals with all the mails in the test log files is infeasible, but
+verifying a subset of those mails is a tractable, if extremely time
+consuming, task.  A sample of 6000 log lines was parsed with all debugging
+options enabled, resulting in 167,448 lines of output.  All 167,448 lines
+were examined in conjunction with the input log file and a dump of the
+resulting database, verifying that for each of the input lines the parser
+used the correct rule and executed the correct action, which in turn
+produced the correct result and inserted the correct data in the database.
+The log segment produced 4 warnings (about mails which had started before
+the log file segment), 10 mails remaining in the state tables, and 1625
+connections correctly entered in the database.  Given the evidence detailed
+above, the author is confident that the false positive rate when
+reconstructing a mail is zero.
 
-Verifying by inspection that the parser correctly deals with all 60,721,709
-lines in the test log files is infeasible, but verifying a subset of those
-log files is a tractable, if extremely time consuming, task.  A sample of
-log lines was parsed with all debugging options enabled, resulting in
-167,448 lines of output.  All 167,448 lines were examined in conjunction
-with the input log and a dump of the resulting database, verifying that for
-each of the input lines the parser used the correct rule and executed the
-correct action, which in turn produced the correct result and inserted the
-correct data in the database.  The log segment produced 4 warnings, 10
-mails remaining in the state tables, and 1625 connections correctly entered
-in the database.
-
-Given the evidence detailed above, the author is confident that the false
-positive rate when reconstructing a mail is exceedingly low, if not
-approaching zero.
-
+XXX DO I NEED THIS PARAGRAPH\@?
 Parser coverage is divided into two topics in this section: log lines

### 2008-05-30 papers/sgai-2008/paper.tex
Mention that actions can return a modified input line to be parsed again.

@@
 XXX IN THE INTRODUCTION\@: WHY IS THIS RELEVANT TO AN AI CONFERENCE\@?
-
-XXX MAYBE MENTION THAT LOG LINES COULD BE MODIFIED AND RE-PARSED\@?  MAYBE
-WORK IT INTO THE ARCHITECTURE CHARACTERISTICS SECTION, AND ELSEWHERE\@.
 
@@
 
+Actions may also return a modified input line which will be parsed as if
+read from the input stream, allowing for cascaded parsing XXX REFERENCE\@.
+This is a powerful facility allowing a generic rule and action to parse
+part of the input, followed by more specific rules and actions parsing the
+remainder.
+
 \subsection{Rules}
@@
         results) of previous actions when executing, providing some context
-        sensitivity.  Actions can not inspect past or upcoming inputs.
+        sensitivity.  Actions can not inspect past or upcoming inputs,
+        however allowing actions can return a modified input line that will
+        be parsed as if read from the input stream, allowing for cascaded
+        parsing XXX REFERENCE\@.
 

### 2008-05-30 papers/sgai-2008/paper.tex
Wrote the conclusion. Added a little to the introduction.

@@
 \newcommand{\numberOFrulesMINIMUM}[0]{%
-    119 XXX CHECK THIS NUMBER%
+    115%
 }
 
+% XXX FIGURE OUT HOW TO DO THIS AUTOMATICALLY
 % \numberOFrulesMINIMUM as percentage of \numberOFrules
 \newcommand{\numberOFrulesMINIMUMpercentage}[0]{%
-    70.41\%%
+    68.04\%%
 }
@@
 \newcommand{\numberOFrulesMAXIMUMpercentage}[0]{%
-    42.01\%%
+    46.95\%%
 }
@@
 \DNSBL{}).
+
+This architecture is ideally suited to parsing inputs where the input is
+not fully understood or does not follow a fixed grammar: the architecture
+warns about unparsed inputs and other errors, but continues on parsing as
+best it can, allowing the developer of a new parser to decide which
+deficiencies are most important and require attention first, rather than
+being forced to fix the first error which arises.
 
@@
 
-XXX WRITE THE CONCLUSION\@.
+The architecture's greatest strength is the ease with which is can be
+adapted to deal with new requirements and inputs.  Parsing a variation of
+an existing input is a trivial task: simply modify an existing rule or add
+a new one with an appropriate \regex{} and the task is complete; knowledge
+of the framework is not required when writing rules.  Parsing a new
+category of input is achieved by writing a new action and appropriate
+rules; quite often the action will not need to interact directly with
+existing actions, but when interaction is required the framework provides
+the necessary facilities.  The decoupling of rules from actions allows
+different sets of rules to be used with the same actions: e.g.\ a parser
+might have actions to process versions one and two of a file format; the
+parser will then parse either version one or version two depending on the
+ruleset selected.  The architecture makes it possible to apply commonly
+used programming techniques such as object orientation, inheritance,
+composition, delegation and roles when designing and implementing a parser,
+simplifying the process of working within a team with shared responsibility
+or developing and testing additional functionality.
 
-Makes adding new rules trivial, new actions tractable.  Knowledge of the
-framework is rarely necessary.
+Developing a new parser based on this architecture follows a similar
+process to developing a new parser using any architecture: study the
+inputs, figure out what needs to happen for each category of input
+(actions), and write a specification identifying each type of input
+(rules); the existing framework should be suitable for use with the
+majority of parsers.  If the inputs to be parsed are suitable a large
+amount of work can be done with very little effort by using \SLCT{} to
+identify similar lines in the input and generate \regexes{} to match those
+lines; these \regexes{} will form the basis of the initial ruleset.  When
+the commonly occurring inputs are parsed by rules and actions, the process
+of identifying and rectifying parsing deficiencies can begin, with each
+iteration of the development process improving the parser, similar to
+Stepwise Refinement~\cite{stepwise-refinement}.
 
-The architecture's greatest strength is the ease with which new inputs can
-be parsed and new actions written.
-
-System could be used by mail administrators to monitor restrictions.
-
-Explain how someone could build upon it, e.g.\ to parse further Postfix
-stuff, or to parse other logs entirely.
-
-``This is a real world application, based on a real world need, rather than
-a toy application built as a proof of concept''  Put this in the
-introduction as well.
+XXX PUT THIS IN THE INTRODUCTION AS WELL\@.  The Postfix log file parser
+based on this architecture provides a basis for systems administrators to
+monitor the effectiveness of their anti-spam measures and adapt their
+defences to combat the new techniques used by those sending spam.  The
+parser is a fully usable application, built to address a genuine need,
+rather than a proof of concept whose sole purpose is to illustrate a new
+idea; it deals with the oddities and difficulties which occur in the real
+world, rather than a clean, idealised scenario developed to showcase the
+best features of a new approach.
 

### 2008-05-30 papers/sgai-2008/paper.tex
Resolved most XXX items.

@@
 
-XXX IN THE INTRODUCTION\@: WHY IS THIS RELEVANT TO AN AI CONFERENCE\@?
-
-XXX MAYBE ADD AN EXAMPLE RULE AND LOG LINE, IF THERE'S SUFFICIENT SPACE\@.
-
 \input{../../doc/logparser-acronyms.tex}
@@
 distinguishing between log lines of the same type, e.g.\ rejections, and
-extracted only a portion of the data required for this project.
+extracted only a portion of the data required for this project.  The parser
+written for this project is a fully usable application, built to address a
+genuine need, rather than a proof of concept whose sole purpose is to
+illustrate a new idea; it deals with the oddities and difficulties which
+occur in the real world, rather than a clean, idealised scenario developed
+to showcase the best features of a new approach.
 
@@
 a specific type, and only that type (e.g.\ rejection by a specific
-\DNSBL{}).
-
-This architecture is ideally suited to parsing inputs where the input is
-not fully understood or does not follow a fixed grammar: the architecture
-warns about unparsed inputs and other errors, but continues on parsing as
-best it can, allowing the developer of a new parser to decide which
-deficiencies are most important and require attention first, rather than
-being forced to fix the first error which arises.
+\DNSBL{}).  This architecture is ideally suited to parsing inputs where the
+input is not fully understood or does not follow a fixed grammar: the
+architecture warns about unparsed inputs and other errors, but continues on
+parsing as best it can, allowing the developer of a new parser to decide
+which deficiencies are most important and require attention first, rather
+than being forced to fix the first error which arises.
 
@@
 Actions may also return a modified input line which will be parsed as if
-read from the input stream, allowing for cascaded parsing XXX REFERENCE\@.
-This is a powerful facility allowing a generic rule and action to parse
-part of the input, followed by more specific rules and actions parsing the
-remainder.
+read from the input stream, allowing for cascaded
+parsing~\cite{cascaded-parsing}.  This is a powerful facility allowing a
+generic rule and action to parse part of the input, followed by more
+specific rules and actions parsing the remainder.
 
@@
 
-    \item [Partially context-sensitive actions:]  XXX IS THIS THE CORRECT
-        TERMINOLOGY\@?  Actions can consult the results (or lack of
-        results) of previous actions when executing, providing some context
-        sensitivity.  Actions can not inspect past or upcoming inputs,
-        however allowing actions can return a modified input line that will
-        be parsed as if read from the input stream, allowing for cascaded
-        parsing XXX REFERENCE\@.
+    \item [Partially context-sensitive actions and cascaded parsing:]
+        Actions can consult the results (or lack of results) of previous
+        actions when executing, providing some context sensitivity.
+        Actions can not inspect past or upcoming inputs, however allowing
+        actions can return a modified input line that will be parsed as if
+        read from the input stream, allowing for cascaded
+        parsing~\cite{cascaded-parsing}.
 
@@
 The parser's coverage of Postfix log files is separated into two parts: log
-line coverage and mail coverage.  The first is important because the parser
-should handle all input log lines; the second is equally important because
-the parser must properly reconstruct every mail if it is to be useful.
-Improving the former is less intrusive, as it just requires new rules to be
-written; improving the latter is more intrusive as it requires changes to
-actions, and it can also be much harder to notice a deficiency.
+line coverage and mail coverage.  The former is initially more important,
+as the parser must successfully parse every line if it is to be complete,
+but subsequently the latter takes precedence because reproducing the path a
+mail takes through Postfix is the aim of the parser.  Improving the former
+is less intrusive, as it just requires new rules to be written; improving
+the latter is more intrusive as it requires changes to actions, and it can
+also be much harder to notice a deficiency.
 
@@
 
-XXX DO I NEED THIS PARAGRAPH\@?
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's log files, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the log files in developing a solution.
-
 \section{Conclusion}
@@
 iteration of the development process improving the parser, similar to
-Stepwise Refinement~\cite{stepwise-refinement}.
+Stepwise Refinement~\cite{stepwise-refinement}.  This architecture is
+ideally suited to parsing inputs where the input is not fully understood or
+does not follow a fixed grammar: the architecture warns about unparsed
+inputs and other errors, but continues on parsing as best it can, allowing
+the developer of a new parser to decide which deficiencies are most
+important and require attention first, rather than being forced to fix the
+first error which arises.
 
-XXX PUT THIS IN THE INTRODUCTION AS WELL\@.  The Postfix log file parser
-based on this architecture provides a basis for systems administrators to
-monitor the effectiveness of their anti-spam measures and adapt their
-defences to combat the new techniques used by those sending spam.  The
-parser is a fully usable application, built to address a genuine need,
-rather than a proof of concept whose sole purpose is to illustrate a new
-idea; it deals with the oddities and difficulties which occur in the real
-world, rather than a clean, idealised scenario developed to showcase the
-best features of a new approach.
+The Postfix log file parser based on this architecture provides a basis for
+systems administrators to monitor the effectiveness of their anti-spam
+measures and adapt their defences to combat the new techniques used by
+those sending spam.  The parser is a fully usable application, built to
+address a genuine need, rather than a proof of concept whose sole purpose
+is to illustrate a new idea; it deals with the oddities and difficulties
+which occur in the real world, rather than a clean, idealised scenario
+developed to showcase the best features of a new approach.
 

### 2008-05-30 papers/sgai-2008/paper.tex
Small, simple, not too intrusive changes to shorten the paper. Still needs lots more shortening.

@@
     must allow users to easily adapt or extend the parser to parse new
-    inputs, without requiring that user have an in-depth
-    knowledge and understanding of the parser's internal workings.  This
-    paper presents a novel parsing architecture which aims to make the
-    process of parsing new inputs as simple as possible, enabling users to
-    trivially add new rules or adapt existing rules (to parse variants of
-    existing inputs) and relatively easily add new actions (to process a
-    previously unknown class of input).  The architecture scales linearly
-    as the number of rules and/or size of input increases, making it
-    suitable for parsing large corpora or months of accumulated data.
+    inputs, without requiring that the user have an in-depth knowledge and
+    understanding of the parser's internal workings.  This paper presents a
+    novel parsing architecture which aims to make the process of parsing
+    new inputs as simple as possible, enabling users to trivially add new
+    rules or adapt existing rules (to parse variants of existing inputs)
+    and relatively easily add new actions (to process a previously unknown
+    class of input).  The architecture scales linearly as the number of
+    rules and/or size of input increases, making it suitable for parsing
+    large corpora or months of accumulated data.
 }
 
-\today{}
-
-XXX DOES THE ABSTRACT NEED TO BE LONGER\@?  IS IT A GOOD TITLE\@?
-
-XXX DUE MONDAY JUNE 2$^{ND}$ 2008\@.  PRESENTATION PAPERS\@: MAX 14
-PAGES\@.  POSTER PAPERS\@: MAX 6 PAGES\@.  BLACK AND WHITE ONLY\@.
+XXX SHOULD THE ABSTRACT BE LONGER\@?  IS IT A GOOD TITLE\@?
 
@@
 with the great number and variety of log lines produced by
-Postfix~\cite{postfix}, the \SMTP{} server in use.  The variety occurs for
-several reasons:
+Postfix~\cite{postfix}, the \SMTP{} server in use.  The variety in log
+lines occurs for several reasons:
 
@@
     \item [Actions]  The actions perform the work required to deal with a
-        category of log lines, e.g.\ creating a new data structure when a
-        log line indicating a remote client connected is processed.
+        category of log lines, e.g.\ processing data from rejections.
 
-    \item [Rules]  The rules are responsible for matching log lines,
-        specifying the action to invoke and the data to be extracted from
-        the log line.
+    \item [Rules]  The rules are responsible for classifying or matching
+        log lines, specifying the action to invoke and the data to be
+        extracted from the log line.
 
@@
 tracks how often each rule matches (to optimise rule ordering, see
-\sectionref{Rule ordering}), and stores results in the database on request.
-The framework is independent of the parser being written: the same
-functionality will be required from the framework for most or all parsers.
+\sectionref{Rule ordering}), and stores results in the database.  The
+framework is independent of the parser being written: the same
+functionality will be required for most or all parsers.
 
@@
 details of parsing, freeing the authors of actions to concentrate on
-writing productive code.  It links actions and rules, allowing either to be
-improved independently of the other.  The framework is the core of the
-architecture and is deliberately quite simple: the rules deal with the
-varying log lines, and the actions deal with the intricacies and
-complications encountered when parsing.
+writing productive code.  It provides the link between actions and rules,
+allowing either to be improved independently of the other.  The framework
+is the core of the architecture and is deliberately quite simple: the rules
+deal with the varying log lines, and the actions deal with the intricacies
+and complications encountered when parsing.
 
@@
 DNS mismatches, or log lines lacking a way to identify which data structure
-they should be associated with.\footnote{Postfix allocates a queueid
-(explained in the glossary,~\sectionref{Glossary}) to each mail; most log
-lines contain the queueid of the associated mail, but some log lines
-(particularly with older versions of Postfix) lack a queueid, preventing
-their association with a particular data structure.}  The parser warns
-about every unparsed log line, to alert the user that they need to alter or
-extend their rules; the \texttt{IGNORE} action does nothing when executed,
-allowing uninteresting log lines to be parsed without causing any effects.
-The remaining actions have only one or two associated rules: in some cases
-this is because there will only ever be one type of log line for that
-action, e.g.\ all log lines showing that a remote client has connected are
-matched by a single rule and handled by the \texttt{CONNECT} action; in
-other cases the action is required to address a deficiency in the log
-files, or a complication which arises during parsing.
+they should be associated with.  The parser warns about every unparsed log
+line, to alert the user that they need to alter or extend their rules; the
+\texttt{IGNORE} action does nothing when executed, allowing uninteresting
+log lines to be parsed without causing any effects.  The remaining actions
+have only one or two associated rules: in some cases this is because there
+will only ever be one type of log line for that action, e.g.\ all log lines
+showing that a remote client has connected are matched by a single rule and
+handled by the \texttt{CONNECT} action; in other cases the action is
+required to address a deficiency in the log files, or a complication which
+arises during parsing.
 
@@
 challenges, with all the attendant risks of adversely affecting the
-existing, working parser, when a new requirement arises a new, independent
+existing, working parser, when a new requirement arises an independent
 action can be written to satisfy it.  In some cases the new action will
@@
 with a monolithic architecture, and thus failures should be less likely and
-easier to test for and diagnose.  The parser can be written in an Object
-Oriented~\cite{Wikipedia-object-orientation} style, allowing sub-classes to
-extend or override actions in addition to adding new actions; because each
-action is an independent function, the sub-class need only modify the
-action it is interested in, rather than reproducing large chunks of
-functionality.
+easier to test for and diagnose.  The parser can be written in an object
+oriented style, allowing sub-classes to extend or override actions in
+addition to adding new actions; because each action is an independent
+function, the sub-class need only modify the action it is interested in,
+rather than reproducing large chunks of functionality.
 
@@
 be overcome.  Some were the result of deficiencies in Postfix's logging,
-which were improved or resolved by later versions of Postfix; others were
-due to the vagaries of process scheduling, client behaviour, and
-administrative actions.  All were successfully accommodated in the parser:
-adding new actions was enough for several of the complications; others
-required modifications to a single existing action to work around the
-difficulties; the remainder were resolved by adapting actions to pass
-information to subsequent actions (via the framework), which then check for
-the extra information and change their behaviour appropriately.
+(some of those deficiencies were improved or resolved by later versions of
+Postfix); others were due to the vagaries of process scheduling, client
+behaviour, and administrative actions.  All were successfully accommodated
+in the parser: adding new actions was enough for several of the
+complications; others required modifications to a single existing action to
+work around the difficulties; the remainder were resolved by adapting
+actions to pass information to subsequent actions (via the framework),
+those subsequent actions then check for the extra information and change
+their behaviour appropriately.
 
@@
 Rules are responsible for categorising inputs, specifying both the data to
-be extracted from each log line and the action that should be invoked.
-This implementation stores the rules in the same \SQL{} database that the
+be extracted from each log line and the action that should be invoked.  The
+Postfix log parser stores the rules in the same \SQL{} database that the
 results are stored in; this removes any doubt as to which set of rules was
@@
     \item [priority] This is the user-configurable companion to hits:
-        priority overrides hits when sorting rules.  This allows more
-        specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
+        priority overrides hits when sorting rules, allowing more
+        specific rules to take precedence over more general rules.
 
@@
 software developed for this project includes a utility based on
-\SLCT{}~\cite{slct-paper} to aid in producing \regexes{} to parse
-previously unparsed inputs.
+SLCT~\cite{slct-paper} to aid in producing \regexes{} to parse previously
+unparsed inputs.
 
@@
 
-XXX DO I NEED TO DISCUSS OVERLAPPING RULES\@?
+%XXX DO I NEED TO DISCUSS OVERLAPPING RULES\@?
 
 The parser does not try to detect overlapping rules; that responsibility is
-left to the author of the rules.\footnote{It may be possible to parse each
-rule's regex and determine if any overlap.  The author has not attempted to
-do this: given the tremendous power and flexibility of regexes,
-particularly Perl's regexes~\cite{perlre}, it would be an extremely
-difficult task.} Unintentionally overlapping rules lead to inconsistent
-parsing and data extraction because the first matching rule wins and the
-order in which rules are tried against each line may change between log
-files.  Overlapping rules are frequently a requirement, allowing a more
-specific rule to match some log lines and a more general rule to match the
-majority, e.g.\ separating \SMTP{} delivery to specific sites from \SMTP{}
-delivery to the rest of the world.  The algorithm provides a facility for
-specifying the order of overlapping rules: the priority field in each rule
-(defaults to zero).  Rules are sorted by priority, highest first, and then
-rules with the same priority are sorted by the number of successful matches
-when parsing the previous log file.  Negative priorities may be useful for
-catchall rules.
+left to the author of the rules.  Unintentionally overlapping rules lead to
+inconsistent parsing and data extraction because the first matching rule
+wins and the order in which rules are tried against each line may change
+between log files.  Overlapping rules are frequently a requirement,
+allowing a more specific rule to match some log lines and a more general
+rule to match the majority, e.g.\ separating \SMTP{} delivery to specific
+sites from \SMTP{} delivery to the rest of the world.  The algorithm
+provides a facility for specifying the order of overlapping rules: the
+priority field in each rule (defaults to zero).  Rules are sorted by
+priority, highest first, and then rules with the same priority are sorted
+by the number of successful matches when parsing the previous log file.
+Negative priorities may be useful for catchall rules.
 
@@
 
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsection{Architecture Scalability: Input size}
+\subsection{Architecture Scalability: Input Size}
 
@@
 of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
+the impression that the algorithm scales linearly with input size.
 This impression is borne out by fig.~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time, and ratio of number of log lines vs execution time (higher
+size vs number lines factor} which plots the ratios of file size vs
+execution time, and number of log lines vs execution time (higher
 is better in both cases).  As the reader can see the ratios are quite

### 2008-05-31 papers/sgai-2008/paper.tex
Shortened by about 1.5 pages.  Reduced the glossary and references by half. Simple changes early in the paper ripple through and result in large differences later.

@@
 
-\input{../../doc/logparser-acronyms.tex}
+\newacronym{ATN}{Augmented Transition Networks}{description={
+    Augmented Transition Networks, originally described in~\cite{atns} and
+    further in~\cite{nlpip}, are a tool used in Computational Linguistics
+    for creating grammars to parse or generate sentences,
+}}
+
+\newacronym{DNSBL}{DNS Blacklist}{description={
+    A DNS Blacklist~\cite{Wikipedia-DNSBL} is a simple collaborative
+    anti-spam technique used to reject or penalise email sent from mail
+    servers reported to have sent large volumes of spam.
+}}
+
+\newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
+    SMTP is the protocol which transfers mail from the sender to the
+    recipient across the Internet.  It is a simple, human readable, plain
+    text protocol, making it quite easy to test and debug problems with it.
+    A detailed description of \SMTP{} is beyond the scope of this document:
+    the original protocol definition is RFC~821~\cite{RFC821}, updated in
+    RFC~2821~\cite{RFC2821}.
+}}
 
@@
 lines which are not of interest to the parser, e.g.\ warning messages about
-DNS mismatches, or log lines lacking a way to identify which data structure
-they should be associated with.  The parser warns about every unparsed log
-line, to alert the user that they need to alter or extend their rules; the
-\texttt{IGNORE} action does nothing when executed, allowing uninteresting
-log lines to be parsed without causing any effects.  The remaining actions
-have only one or two associated rules: in some cases this is because there
-will only ever be one type of log line for that action, e.g.\ all log lines
-showing that a remote client has connected are matched by a single rule and
-handled by the \texttt{CONNECT} action; in other cases the action is
-required to address a deficiency in the log files, or a complication which
-arises during parsing.
+DNS mismatches.  The parser warns about every unparsed log line, to alert
+the user that they need to alter or extend their rules; the \texttt{IGNORE}
+action does nothing when executed, allowing uninteresting log lines to be
+parsed without causing any effects.  The remaining actions have only one or
+two associated rules: in some cases this is because there will only ever be
+one type of log line for that action, e.g.\ all log lines showing that a
+remote client has connected are matched by a single rule and handled by the
+\texttt{CONNECT} action; in other cases the action is required to address a
+deficiency in the log files, or a complication which arises during parsing.
 
@@
 Actions may also return a modified input line which will be parsed as if
-read from the input stream, allowing for cascaded
+read from the input stream, allowing for a simplified version of cascaded
 parsing~\cite{cascaded-parsing}.  This is a powerful facility allowing a
@@
 be extracted from each log line and the action that should be invoked.  The
-Postfix log parser stores the rules in the same \SQL{} database that the
+Postfix log parser stores the rules in the same SQL database that the
 results are stored in; this removes any doubt as to which set of rules was
@@
 
-    \item [regex] The \regex{} to match the log line against.
+    \item [regex] The Regular Expression to match the log line against.
 
@@
         i.e.\ semi-colon or comma separated assignment statements, with the
-        variable name on the left and the matching capture from the
-        \regex{} on the right.
+        variable name on the left and the matching capture from the regex
+        on the right.
 
@@
 software developed for this project includes a utility based on
-SLCT~\cite{slct-paper} to aid in producing \regexes{} to parse previously
+SLCT~\cite{slct-paper} to aid in producing regexes to parse previously
 unparsed inputs.
@@
         actions can return a modified input line that will be parsed as if
-        read from the input stream, allowing for cascaded
-        parsing~\cite{cascaded-parsing}.
+        read from the input stream, allowing for a simplified version of
+        cascaded parsing~\cite{cascaded-parsing}.
 
@@
 
-    \item [Closer to \NLP{} than fixed grammar:]  Unlike traditional
-        parsers such as those used when compiling a programming language,
-        this architecture does not have a fixed grammar specification which
-        inputs must adhere to.  The architecture is capable of dealing with
-        overlapping inputs, out of order inputs, and ambiguous inputs where
-        heuristics must be applied --- all have arisen and been
-        successfully dealt with in the Postfix log parser.
+    \item [Closer to Natural Language Processing than using a fixed
+        grammar:] Unlike traditional parsers such as those used when
+        compiling a programming language, this architecture does not have a
+        fixed grammar specification which inputs must adhere to.  The
+        architecture is capable of dealing with overlapping inputs, out of
+        order inputs, and ambiguous inputs where heuristics must be applied
+        --- all have arisen and been successfully dealt with in the Postfix
+        log parser.
 
@@
 input size: does it scale linearly, polynomially, or exponentially?
-Figure~\refwithpage{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB, and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-the impression that the algorithm scales linearly with input size.
-This impression is borne out by fig.~\refwithpage{execution time vs file
-size vs number lines factor} which plots the ratios of file size vs
-execution time, and number of log lines vs execution time (higher
-is better in both cases).  As the reader can see the ratios are quite
-tightly banded across the graph, showing that the algorithm scales
-linearly: the much larger log files at points 62--68 on the X axis in
-fig.~\refwithpage{execution time vs file size vs number of lines graph}
-actually cause the ratio to increase (i.e.\ improve), rather than decrease.
-The strange behaviour where larger log files are parsed more efficiently is
-explained in \sectionref{Why are there dips in the graphs?}.
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file size
-vs number of lines}{execution time vs file size vs number of lines graph}
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file size
-and number of lines to execution time}{execution time vs file size vs
-number lines factor}
+Figure~\refwithpage{Execution time, file size, and number of lines} shows
+the execution time in seconds, file size in MB, and tens of thousands of
+log lines per log file.  All three lines run roughly in parallel, giving
+the impression that the algorithm scales linearly with input size.  This
+impression is borne out by fig.~\refwithpage{execution time vs file size vs
+number lines factor} which plots the ratios of file size vs execution time,
+and number of log lines vs execution time (higher is better in both cases).
+As the reader can see the ratios are quite tightly banded across the graph,
+showing that the algorithm scales linearly: the much larger log files at
+points 62--68 on the X axis in fig.~\refwithpage{Execution time, file size,
+and number of lines} actually cause the ratio to increase (i.e.\ improve),
+rather than decrease.  The strange behaviour where larger log files are
+parsed more efficiently is explained in \sectionref{Why are there dips in
+the graphs?}.  \showgraph{build/plot-normal-filesize-numlines}{Execution
+time, file size, and number of lines}{Execution time, file size, and number
+of lines} \showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of
+file size and number of lines to execution time}{execution time vs file
+size vs number lines factor}
 
@@
 top 10\% matching the vast majority of the log lines, and the remaining log
-lines split across the other 90\% of the rules in a Power Law
-distribution~\cite{powerlaw} (as shown in fig.~\refwithpage{rule hits
-graph}).  \showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-Assuming that the distribution of log lines is reasonably steady over time,
-parser efficiency should benefit from trying more frequently matching rules
-before those which match less frequently.  To test this hypothesis three
-full test runs were performed with different rule orderings:
+lines split across the other 90\% of the rules in a Power Law distribution
+(as shown in fig.~\refwithpage{rule hits graph}).
+\showgraph{build/plot-hits}{Hits per rule}{rule hits graph} Assuming that
+the distribution of log lines is reasonably steady over time, parser
+efficiency should benefit from trying more frequently matching rules before
+those which match less frequently.  To test this hypothesis three full test
+runs were performed with different rule orderings:
 
@@
 increase of shuffled and reversed over normal} correspond to peaks in log
-file size in fig.~\refwithpage{execution time vs file size vs number of
-lines graph}, and peaks in fig.~\refwithpage{execution time vs file size vs
+file size in fig.~\refwithpage{Execution time, file size, and number of
+lines}, and peaks in fig.~\refwithpage{execution time vs file size vs
 number lines factor} (where a peak means that performance is better).  The
@@
 maximum ruleset over minimum ruleset} that the increased number of rules
-makes a noticeable difference with the reverse ordering, and to a lesser
-extent with the shuffled ordering.  The normal ordering (where rules are
-sorted so that the most frequently matching rules are tried first) shows a
+has a noticeable performance impact with reverse ordering, and to a lesser
+extent with shuffled ordering.  The normal ordering (where are the most
+frequently matching rules are tried first) shows a
 \input{build/normal-vs-smaller-percentage-increase.tex} increase in
@@
 as the number of rules increases, and that sorting the rules is an
-important optimisation enabling the parser to scale well.
+important optimisation enabling this.
 
@@
 there are zero false negatives.  False positives are harder to quantify:
-manually verifying that the correct \regex{} parsed each of the 60,721,709
-log lines is infeasible, making it impossible to quantify the false
-positive rate; however a random sample of 6039 log lines (0.00994\% of the
-total) was parsed and the results verified to ensure that the correct
-\regex{} parsed each log line.
+manually verifying that the correct regex parsed each of the 60,721,709 log
+lines is infeasible, making it impossible to quantify the false positive
+rate; however a random sample of 6039 log lines (0.00994\% of the total)
+was parsed and the results verified to ensure that the correct regex parsed
+each log line.
 
@@
 which had yet to be delivered when the log files ended, or were accepted
-before the log files began.  Verifying by inspection that the parser
-correctly deals with all the mails in the test log files is infeasible, but
-verifying a subset of those mails is a tractable, if extremely time
-consuming, task.  A sample of 6000 log lines was parsed with all debugging
-options enabled, resulting in 167,448 lines of output.  All 167,448 lines
-were examined in conjunction with the input log file and a dump of the
-resulting database, verifying that for each of the input lines the parser
-used the correct rule and executed the correct action, which in turn
-produced the correct result and inserted the correct data in the database.
-The log segment produced 4 warnings (about mails which had started before
-the log file segment), 10 mails remaining in the state tables, and 1625
-connections correctly entered in the database.  Given the evidence detailed
-above, the author is confident that the false positive rate when
-reconstructing a mail is zero.
+before the log files began.  A sample of 6000 log lines was parsed with all
+debugging options enabled, resulting in 167,448 lines of output.  All
+167,448 lines were examined in conjunction with the input log file and a
+dump of the resulting database, verifying that for each of the input lines
+the parser used the correct rule and executed the correct action, which in
+turn produced the correct result and inserted the correct data in the
+database.  The log segment produced 4 warnings (about mails which had
+started before the log file segment), 10 mails remaining in the state
+tables, and 1625 connections correctly entered in the database.  Given the
+evidence detailed above, the author is confident that the false positive
+rate when reconstructing a mail is zero.
 
@@
 an existing input is a trivial task: simply modify an existing rule or add
-a new one with an appropriate \regex{} and the task is complete; knowledge
-of the framework is not required when writing rules.  Parsing a new
-category of input is achieved by writing a new action and appropriate
-rules; quite often the action will not need to interact directly with
-existing actions, but when interaction is required the framework provides
-the necessary facilities.  The decoupling of rules from actions allows
-different sets of rules to be used with the same actions: e.g.\ a parser
-might have actions to process versions one and two of a file format; the
-parser will then parse either version one or version two depending on the
-ruleset selected.  The architecture makes it possible to apply commonly
-used programming techniques such as object orientation, inheritance,
-composition, delegation and roles when designing and implementing a parser,
-simplifying the process of working within a team with shared responsibility
-or developing and testing additional functionality.
+a new one with an appropriate regex and the task is complete; knowledge of
+the framework is not required when writing rules.  Parsing a new category
+of input is achieved by writing a new action and appropriate rules; quite
+often the action will not need to interact directly with existing actions,
+but when interaction is required the framework provides the necessary
+facilities.  The decoupling of rules from actions allows different sets of
+rules to be used with the same actions: e.g.\ a parser might have actions
+to process versions one and two of a file format; the parser will then
+parse either version one or version two depending on the ruleset selected.
+The architecture makes it possible to apply commonly used programming
+techniques such as object orientation, inheritance, composition, delegation
+and roles when designing and implementing a parser, simplifying the process
+of working within a team with shared responsibility or developing and
+testing additional functionality.
 
@@
 majority of parsers.  If the inputs to be parsed are suitable a large
-amount of work can be done with very little effort by using \SLCT{} to
-identify similar lines in the input and generate \regexes{} to match those
-lines; these \regexes{} will form the basis of the initial ruleset.  When
-the commonly occurring inputs are parsed by rules and actions, the process
-of identifying and rectifying parsing deficiencies can begin, with each
+amount of work can be done with very little effort by using SLCT to
+identify similar lines in the input and generate regexes to match those
+lines; these regexes will form the basis of the initial ruleset.  When the
+commonly occurring inputs are parsed by rules and actions, the process of
+identifying and rectifying parsing deficiencies can begin, with each
 iteration of the development process improving the parser, similar to
@@
 
-\glossary{name={queueid},description={
-    Each mail in Postfix's queue is assigned a queueid to uniquely identify
-    it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
-}}
-
 \end{document}

### 2008-05-31 papers/sgai-2008/paper.tex
Addresses some of Carl's comments. Minor cleanup to packages used etc.

@@
 \usepackage{amstext}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
 % Include images
 \usepackage[final]{graphicx}
-%\renewcommand{\refname}{Bibliography}
-% Add the bibliography into the table of contents.
-%\usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
 \usepackage{url}
@@
 
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
 \usepackage{lastpage}
@@
 \usepackage{cite}
-% Adds page numbers to bibliography entries, showing where they were cited.
-% Breaks when cite is used.
-% \usepackage{citeref}
 % Check for unused references.
@@
 \vrefshowerrors{}
-% Show all the files/packages loaded.
-%\listfiles
 
@@
 \newcommand{\refwithpage}[1]{%
-    %\empty{}\ref{#1} [p.~\pageref{#1}]%
     \empty{}\vref{#1}%
@@
 \newcommand{\sectionref}[1]{%
-    %\textsection{}\refwithpage{#1}%
     \textsection{}\vref*{#1}%
@@
     \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
 }
@@
     new inputs as simple as possible, enabling users to trivially add new
-    rules or adapt existing rules (to parse variants of existing inputs)
-    and relatively easily add new actions (to process a previously unknown
-    class of input).  The architecture scales linearly as the number of
-    rules and/or size of input increases, making it suitable for parsing
-    large corpora or months of accumulated data.
+    rules (to parse variants of existing inputs) and relatively easily add
+    new actions (to process a previously unknown class of input).  The
+    architecture scales linearly as the number of rules and/or size of
+    input increases, making it suitable for parsing large corpora or months
+    of accumulated data.
 }
-
-XXX SHOULD THE ABSTRACT BE LONGER\@?  IS IT A GOOD TITLE\@?
 
@@
 \newacronym{DNSBL}{DNS Blacklist}{description={
-    A DNS Blacklist~\cite{Wikipedia-DNSBL} is a simple collaborative
-    anti-spam technique used to reject or penalise email sent from mail
-    servers reported to have sent large volumes of spam.
+    A DNS Blacklist is a simple collaborative anti-spam technique used to
+    reject or penalise email sent from mail servers reported to have sent
+    large volumes of spam.
 }}
@@
 based on that analysis, and developing new measures to supplement the
-existing filters where there is a deficiency.  The approach chosen to
+existing filters where there is a deficiency.  Most anti-spam techniques
+are content-based~\cite{a-plan-for-spam, word-stemming,
+relaxed-online-svms}, viz they examine the contents of each mail to
+determine whether it is spam or not.  Examining content requires accepting
+the mail in the first place, but rejecting rather than accepting mail is
+preferable in many cases: senders of non-spam mail that is mistakenly
+rejected will receive an immediate non-delivery notice; load is reduced on
+mail servers (allowing more intensive content-based measures to be used on
+mail that is accepted); users have less mail to wade through.  Improving
+the performance of anti-spam measures which are applied when mail is being
+transferred via \SMTP{} is the goal of this project, by providing a
+platform for reasoning about anti-spam filters.  The approach chosen to
 measure performance was to analyse the log files produced by the mail
@@
 with the great number and variety of log lines produced by
-Postfix~\cite{postfix}, the \SMTP{} server in use.  The variety in log
+Postfix~\cite{postfix}, the \SMTP{} server in use.  The parser must be
+designed so that adding support for parsing new log lines is a simple task,
+as the log lines will continue to evolve over time.  The variety in log
 lines occurs for several reasons:
@@
     \item The mail administrator can define custom rejection
-        messages~\cite{postfix-lookup-tables, smtpd_access_readme,
-        smtpd_per_user_control}, resulting in custom log lines.
+        messages, resulting in custom log lines.
 
@@
 \end{itemize}
-
-The parser must be designed so that adding support for parsing new log
-lines is a simple task, as the log lines will continue to evolve over time.
 
@@
 distinguishing between log lines of the same type, e.g.\ rejections, and
-extracted only a portion of the data required for this project.  The parser
-written for this project is a fully usable application, built to address a
-genuine need, rather than a proof of concept whose sole purpose is to
-illustrate a new idea; it deals with the oddities and difficulties which
-occur in the real world, rather than a clean, idealised scenario developed
-to showcase the best features of a new approach.
+extracted only a portion of the data required for this project.
 
@@
 
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
+% Redefine the command used to produce the glossary title so that it
+% doesn't produce a new section - this will save space.
+\renewcommand{\glossarytitle}{\noindent{}\textbf{\large{Glossary}}}
 \printglossary{}

### 2008-06-01 papers/sgai-2008/paper.tex
Addressed most of the remainder of Carl's comments. Still need to add examples, if possible.

@@
 allows new rules to be written and tested without requiring modifications
-to the parser source code (significantly lowering the barrier to entry for
-new or casual users who need to parse new log lines), and greatly
-simplifies the framework, actions and rules. Decoupling the actions from
-the framework simplifies both framework and actions: the framework provides
+to the parser source code, significantly lowering the barrier to entry for
+casual users who need to parse new log lines, e.g.\ part-time systems
+administrators attempting to reduce spam.  Decoupling the actions from the
+framework simplifies both framework and actions: the framework provides
 services to the actions, and doesn't need to perform any tasks specific to
@@
 of working within a team with shared responsibility or developing and
-testing additional functionality.
+testing additional functionality.  This architecture is ideally suited to
+parsing inputs where the input is not fully understood or does not follow a
+fixed grammar: the architecture warns about unparsed inputs and other
+errors, but continues on parsing as best it can, allowing the developer of
+a new parser to decide which deficiencies are most important and require
+attention first, rather than being forced to fix the first error which
+arises.
 
-Developing a new parser based on this architecture follows a similar
-process to developing a new parser using any architecture: study the
-inputs, figure out what needs to happen for each category of input
-(actions), and write a specification identifying each type of input
-(rules); the existing framework should be suitable for use with the
-majority of parsers.  If the inputs to be parsed are suitable a large
-amount of work can be done with very little effort by using SLCT to
-identify similar lines in the input and generate regexes to match those
-lines; these regexes will form the basis of the initial ruleset.  When the
-commonly occurring inputs are parsed by rules and actions, the process of
-identifying and rectifying parsing deficiencies can begin, with each
-iteration of the development process improving the parser, similar to
-Stepwise Refinement~\cite{stepwise-refinement}.  This architecture is
-ideally suited to parsing inputs where the input is not fully understood or
-does not follow a fixed grammar: the architecture warns about unparsed
-inputs and other errors, but continues on parsing as best it can, allowing
-the developer of a new parser to decide which deficiencies are most
-important and require attention first, rather than being forced to fix the
-first error which arises.
+The dataset gathered by the parser provides the foundation for the future
+of this project: applying machine-learning algorithms to the data to
+analyse and optimise the set of rules in use, followed by identifying
+patterns in the data which could be used to write new filters to recognise
+and reject spam rather than accepting it.  The current focus is on
+clustering and decision trees to optimise the order in which rules are
+applied; future efforts will involve using gathered data to train and test
+new filters.  This task is very similar to optimising a black box
+application based on its inputs and outputs, and this approach could be
+applied to optimising the behaviour of any system given sufficient logging
+to analyse.  An alternative approach to black box optimisation which uses
+application profiling in conjunction with the application's error messages
+to improve the error messages shown to users is described
+in~\cite{black-box-error-reporting}; profiling data may be useful in
+supplementing systems which fail to provide adequate log messages.
 

### 2008-06-01 papers/sgai-2008/paper.tex
Shaved another half a page off: * Removed the glossary entirely, moving the SMTP glossary entry to the introduction * Reduced the introduction by several lines, to fit in the SMTP footnote and   pull the start of the framework/actions/rules list onto page 2. This should give me a bit of space for an example action, rule and log line (or two).

@@
 \usepackage{url}
-
-\usepackage[style=altlist,number=none,toc=true,global=true]{glossary}
-\makeglossary{}
 
@@
 
-\newacronym{ATN}{Augmented Transition Networks}{description={
-    Augmented Transition Networks, originally described in~\cite{atns} and
-    further in~\cite{nlpip}, are a tool used in Computational Linguistics
-    for creating grammars to parse or generate sentences,
-}}
-
-\newacronym{DNSBL}{DNS Blacklist}{description={
-    A DNS Blacklist is a simple collaborative anti-spam technique used to
-    reject or penalise email sent from mail servers reported to have sent
-    large volumes of spam.
-}}
-
-\newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
-    SMTP is the protocol which transfers mail from the sender to the
-    recipient across the Internet.  It is a simple, human readable, plain
-    text protocol, making it quite easy to test and debug problems with it.
-    A detailed description of \SMTP{} is beyond the scope of this document:
-    the original protocol definition is RFC~821~\cite{RFC821}, updated in
-    RFC~2821~\cite{RFC2821}.
-}}
-
 \section{Introduction}
@@
 filters currently in use, optimising the order and membership of the set
-based on that analysis, and developing new measures to supplement the
-existing filters where there is a deficiency.  Most anti-spam techniques
-are content-based~\cite{a-plan-for-spam, word-stemming,
-relaxed-online-svms}, viz they examine the contents of each mail to
-determine whether it is spam or not.  Examining content requires accepting
-the mail in the first place, but rejecting rather than accepting mail is
+based on that analysis, and developing supplemental measures where there is
+a deficiency.  Most anti-spam techniques are
+content-based~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms},
+which requires accepting the mail in the first place, but rejecting is
 preferable in many cases: senders of non-spam mail that is mistakenly
 rejected will receive an immediate non-delivery notice; load is reduced on
-mail servers (allowing more intensive content-based measures to be used on
-mail that is accepted); users have less mail to wade through.  Improving
-the performance of anti-spam measures which are applied when mail is being
-transferred via \SMTP{} is the goal of this project, by providing a
+mail servers (allowing more intensive content-based measures to be used);
+users have less mail to wade through.  Improving the performance of
+anti-spam measures which are applied when mail is being transferred via
+SMTP\footref{smtp footnote} is the goal of this project, by providing a
 platform for reasoning about anti-spam filters.  The approach chosen to
@@
 with the great number and variety of log lines produced by
-Postfix~\cite{postfix}, the \SMTP{} server in use.  The parser must be
-designed so that adding support for parsing new log lines is a simple task,
-as the log lines will continue to evolve over time.  The variety in log
-lines occurs for several reasons:
+Postfix~\cite{postfix}, the SMTP\footnote{\label{smtp footnote}Simple Mail
+Transfer Protocol transfers mail across the Internet.  It is a simple,
+human readable, plain text protocol, making it quite easy to test and debug
+problems with it.  The original protocol definition is
+RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.} server in use.
+The parser must be designed so that adding support for parsing new log
+lines is a simple task, as the log lines will continue to evolve over time.
+The variety in log lines occurs for several reasons:
 
@@
     \item The mail administrator can define custom rejection
-        messages, resulting in custom log lines.
+        messages.
 
     \item External resources utilised by Postfix (e.g.\ policy
-        servers~\cite{policy-servers} or \DNSBL{} lookups) can also change
-        their messages, sometimes without warning.
+        servers~\cite{policy-servers} or DNSBLs) can also change their
+        messages, sometimes without warning.
 
@@
 Existing parsers were considered, but their parsing coverage was too low,
-their parsing too inexact, and/or they did not extract sufficient data for
-the purposes of this project.  The effort required to adapt and improve an
-existing parser was judged to be greater than the effort to write one from
-scratch, as the techniques used by the existing parsers severely limited
-their potential: some ignored the majority of log lines, parsing specific
-log lines with great accuracy, but in an inextensible fashion; others
-sloppily parsed the majority of log lines, but were incapable of
-distinguishing between log lines of the same type, e.g.\ rejections, and
-extracted only a portion of the data required for this project.
+their parsing too inexact, and/or they did not extract sufficient data.
+The effort required to adapt and improve an existing parser was judged to
+be greater than the effort to write one from scratch, as the techniques
+used by the existing parsers severely limited their potential: some ignored
+the majority of log lines, parsing specific log lines with great accuracy,
+but in an inextensible fashion; others sloppily parsed the majority of log
+lines, but were incapable of distinguishing between log lines of the same
+type, e.g.\ rejections, and extracted only a portion of the data required.
 
@@
 precise patterns (\textit{rules\/}), each of which matches all log lines of
-a specific type, and only that type (e.g.\ rejection by a specific
-\DNSBL{}).  This architecture is ideally suited to parsing inputs where the
-input is not fully understood or does not follow a fixed grammar: the
-architecture warns about unparsed inputs and other errors, but continues on
-parsing as best it can, allowing the developer of a new parser to decide
-which deficiencies are most important and require attention first, rather
-than being forced to fix the first error which arises.
+a specific type, and only that type (e.g.\ rejection by a specific DNSBL).
+This architecture is ideally suited to parsing inputs where the input is
+not fully understood or does not follow a fixed grammar: the architecture
+warns about unparsed inputs and other errors, but continues on parsing as
+best it can, allowing the developer of a new parser to decide which
+deficiencies are most important and require attention first, rather than
+being forced to fix the first error which arises.
 
@@
 There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-different approaches have a similar division of responsibilities, while
-having completely different implementations and semantics.
+Augmented Transition Networks (ATN)~\cite{atns, nlpip}, a tool used in
+Computational Linguistics for creating grammars to parse or generate
+sentences.  The resemblance between ATN and the parser is accidental, but
+it is interesting how two different approaches have a similar division of
+responsibilities, while having completely different implementations and
+semantics.
 
@@
         \noalign{\smallskip}
-        \ATN{}        & Parser    & Similarity                          \\
+        ATN           & Parser    & Similarity                          \\
         \noalign{\smallskip}
@@
 allowing a more specific rule to match some log lines and a more general
-rule to match the majority, e.g.\ separating \SMTP{} delivery to specific
-sites from \SMTP{} delivery to the rest of the world.  The algorithm
-provides a facility for specifying the order of overlapping rules: the
-priority field in each rule (defaults to zero).  Rules are sorted by
-priority, highest first, and then rules with the same priority are sorted
-by the number of successful matches when parsing the previous log file.
-Negative priorities may be useful for catchall rules.
+rule to match the majority, e.g.\ separating SMTP delivery to specific
+sites from SMTP delivery to the rest of the world.  The algorithm provides
+a facility for specifying the order of overlapping rules: the priority
+field in each rule (defaults to zero).  Rules are sorted by priority,
+highest first, and then rules with the same priority are sorted by the
+number of successful matches when parsing the previous log file.  Negative
+priorities may be useful for catchall rules.
 
@@
 \bibliography{../common/bibliography}
-\label{bibliography}
-
-% Redefine the command used to produce the glossary title so that it
-% doesn't produce a new section - this will save space.
-\renewcommand{\glossarytitle}{\noindent{}\textbf{\large{Glossary}}}
-\printglossary{}
 

### 2008-06-02 papers/sgai-2008/paper.tex
Improved the SMTP footnote so it isn't referenced twice.

@@
 anti-spam measures which are applied when mail is being transferred via
-SMTP\footref{smtp footnote} is the goal of this project, by providing a
-platform for reasoning about anti-spam filters.  The approach chosen to
-measure performance was to analyse the log files produced by the mail
-server, rather than modifying the mail server to generate performance
-statistics; this approach improves the chances of other sites testing and
-utilising the software.  The need arose for a parser capable of dealing
-with the great number and variety of log lines produced by
-Postfix~\cite{postfix}, the SMTP\footnote{\label{smtp footnote}Simple Mail
-Transfer Protocol transfers mail across the Internet.  It is a simple,
+SMTP\footnotemark{} is the goal of this project, by providing a platform
+for reasoning about anti-spam filters.  The approach chosen to measure
+performance was to analyse the log files produced by the mail server,
+rather than modifying the mail server to generate performance statistics;
+this approach improves the chances of other sites testing and utilising the
+software.  The need arose for a parser capable of dealing with the great
+number and variety of log lines produced by Postfix~\cite{postfix}, the
+SMTP server in use.  The parser must be designed so that adding support for
+parsing new log lines is a simple task, as the log lines will continue to
+evolve over time.  The variety in log lines occurs for several
+reasons:\footnotetext{Simple Mail Transfer Protocol transfers mail from
+sender to one or more recipients across the Internet.  It is a simple,
 human readable, plain text protocol, making it quite easy to test and debug
 problems with it.  The original protocol definition is
-RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.} server in use.
-The parser must be designed so that adding support for parsing new log
-lines is a simple task, as the log lines will continue to evolve over time.
-The variety in log lines occurs for several reasons:
+RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}
 

### 2008-06-02 papers/sgai-2008/paper.tex
Added an example action, rule and log line.

@@
 
+Using the \texttt{CONNECT} action as an example, it creates a new data
+structure for each client connecting, and saves the data extracted by the
+rule (shown in \sectionref{example rule}).  If the data structure already
+exists it is treated as a symptom of a bug, and the action issues a warning
+containing the full contents of the data structure, plus the log line which
+has just been parsed.
+
 The ability to add special purpose actions to deal with difficulties and
@@
 
-%XXX DO I NEED TO DISCUSS OVERLAPPING RULES\@?
-
 The parser does not try to detect overlapping rules; that responsibility is
@@
 priorities may be useful for catchall rules.
+
+\label{example rule}
+
+This example rule matches the message logged by Postfix when a remote
+client connects to deliver mail; it would match the following sample log
+line:
+
+\begin{verbatim}
+    connect from client.example.com[1.2.3.4]
+\end{verbatim}
+
+% Don't reformat this!
+% XXX CONSIDER USING A TABLE ENVIRONMENT; IT DIDN'T LOOK GREAT BEFORE.
+\begin{tabular}[]{ll}
+\textbf{Attribute}        & \textbf{Value}                                   \\
+program                   & \daemon{smtpd}                                   \\
+regex                     & \verb!^connect from (__HOSTNAME__)\[(__IP__)\]$! \\
+result\_cols              &                                                  \\
+connection\_cols~\empty{} & client\_hostname = 1; client\_ip = 2             \\
+action                    & CONNECT                                          \\
+priority                  & 0                                                \\
+\end{tabular}
+
+\noindent{}The \texttt{CONNECT} action is described in \sectionref{Actions}.
 

### 2008-06-02 papers/sgai-2008/paper.tex
Explained priority in a bit more detail.

@@
 
-    \item [priority] This is the user-configurable companion to hits:
-        priority overrides hits when sorting rules, allowing more
-        specific rules to take precedence over more general rules.
+    \item [priority] Priority overrides hits when sorting rules, allowing
+        more specific rules to take precedence over more general rules.  To
+        illustrate: most warnings are ignored by a rule whose regex matches
+        \verb!^warning:.*$!, but some warnings require a different action
+        to be taken, e.g.\ when a client tries to send a mail larger than
+        the maximum size the server accepts, the data structure for that
+        mail must be discarded because it won't have all the data required
+        to save it to the database.
 

### 2008-06-02 papers/sgai-2008/paper.tex
Corrections after final proofreading.

@@
 filters currently in use, optimising the order and membership of the set
-based on that analysis, and developing supplemental measures where there is
+based on that analysis, and developing supplemental filters where there is
 a deficiency.  Most anti-spam techniques are
@@
 rejected will receive an immediate non-delivery notice; load is reduced on
-mail servers (allowing more intensive content-based measures to be used);
+mail servers (allowing more intensive content-based techniques to be used);
 users have less mail to wade through.  Improving the performance of
-anti-spam measures which are applied when mail is being transferred via
+anti-spam techniques which are applied when mail is being transferred via
 SMTP\footnotemark{} is the goal of this project, by providing a platform
@@
 evolve over time.  The variety in log lines occurs for several
-reasons:\footnotetext{Simple Mail Transfer Protocol transfers mail from
+reasons:\footnotetext{Simple Mail Transfer Protocol transfers mail from the
 sender to one or more recipients across the Internet.  It is a simple,
@@
 
-Decoupling the parsing rules from the associated actions and framework
-allows new rules to be written and tested without requiring modifications
-to the parser source code, significantly lowering the barrier to entry for
-casual users who need to parse new log lines, e.g.\ part-time systems
-administrators attempting to reduce spam.  Decoupling the actions from the
-framework simplifies both framework and actions: the framework provides
-services to the actions, and doesn't need to perform any tasks specific to
-the input being parsed; actions benefit from having services provided by
-the framework, freeing them to concentrate on the task of accurately and
-correctly processing the information provided by rules.
+Decoupling the parsing rules from their associated actions and the
+framework allows new rules to be written and tested without requiring
+modifications to the parser source code, significantly lowering the barrier
+to entry for casual users who need to parse new log lines, e.g.\ part-time
+systems administrators attempting to reduce spam.  Decoupling the actions
+from the framework simplifies both framework and actions: the framework
+provides services to the actions, but doesn't need to perform any tasks
+specific to the input being parsed; actions benefit from having services
+provided by the framework, freeing them to concentrate on the task of
+accurately and correctly processing the information provided by rules.
 
@@
 dealing with the intricacies of the input being parsed, complications
-arising, etc; the framework provides services to actions and manages the
+arising, etc.;\ the framework provides services to actions and manages the
 parsing process.
@@
 loads and validates rules, controls parsing of log files, invokes actions,
-tracks how often each rule matches (to optimise rule ordering, see
+tracks how often each rule matches, to optimise rule ordering (see
 \sectionref{Rule ordering}), and stores results in the database.  The
@@
 DNS mismatches.  The parser warns about every unparsed log line, to alert
-the user that they need to alter or extend their rules; the \texttt{IGNORE}
-action does nothing when executed, allowing uninteresting log lines to be
-parsed without causing any effects.  The remaining actions have only one or
-two associated rules: in some cases this is because there will only ever be
-one type of log line for that action, e.g.\ all log lines showing that a
-remote client has connected are matched by a single rule and handled by the
-\texttt{CONNECT} action; in other cases the action is required to address a
-deficiency in the log files, or a complication which arises during parsing.
+the user that they need to alter or extend their ruleset; the
+\texttt{IGNORE} action does nothing when executed, allowing uninteresting
+log lines to be parsed without causing any effects.  The remaining actions
+have only one or two associated rules: in some cases this is because there
+will only ever be one type of log line for that action, e.g.\ all log lines
+showing that a remote client has connected are matched by a single rule and
+handled by the \texttt{CONNECT} action; in other cases the action is
+required to address a deficiency in the log files, or a complication which
+arises during parsing.
+
+\label{example action}
 
@@
 challenges, with all the attendant risks of adversely affecting the
-existing, working parser, when a new requirement arises an independent
-action can be written to satisfy it.  In some cases the new action will
-require the cooperation of other actions, e.g.\ to set or check a flag;
-there is a possibility of introducing failure by modifying existing
-actions, but modifications will be smaller and occur less frequently than
-with a monolithic architecture, and thus failures should be less likely and
-easier to test for and diagnose.  The parser can be written in an object
-oriented style, allowing sub-classes to extend or override actions in
-addition to adding new actions; because each action is an independent
-function, the sub-class need only modify the action it is interested in,
-rather than reproducing large chunks of functionality.
+existing parser, when a new requirement arises an independent action can be
+written to satisfy it.  In some cases the new action will require the
+cooperation of other actions, e.g.\ to set or check a flag; there is a
+possibility of introducing failure by modifying existing actions, but
+modifications will be smaller and occur less frequently than with a
+monolithic architecture, thus failures should be less likely, and easier to
+test for and diagnose.  The parser can be written in an object oriented
+style, allowing sub-classes to extend or override actions in addition to
+adding new actions; because each action is an independent function, the
+sub-class need only modify the action it is interested in, rather than
+reproducing large chunks of functionality.
 
@@
 obvious variety in log lines, there were a large number of complications to
-be overcome.  Some were the result of deficiencies in Postfix's logging,
+be overcome.  Some were the result of deficiencies in Postfix's logging
 (some of those deficiencies were improved or resolved by later versions of
@@
 Postfix log parser stores the rules in the same SQL database that the
-results are stored in; this removes any doubt as to which set of rules was
-used to produce a set of results.  Other implementations are free to store
-the rules in whatever fashion suits their needs.  Each rule defines several
+results are stored in, removing any doubt as to which set of rules was used
+to produce a set of results; other implementations are free to store the
+rules in whatever fashion suits their needs.  Each rule defines several
 attributes:
@@
         matching more frequently will be tried before rules matching less
-        frequently, lowering the parser's execution time.
+        frequently, reducing the parser's execution time.
 
@@
 inconsistent parsing and data extraction because the first matching rule
-wins and the order in which rules are tried against each line may change
+wins, and the order in which rules are tried against each line may change
 between log files.  Overlapping rules are frequently a requirement,
@@
 highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
+number of successful matches they had parsing the previous log file.  Negative
 priorities may be useful for catchall rules.
@@
 \begin{tabular}[]{ll}
-\textbf{Attribute}        & \textbf{Value}                                   \\
+    \hline
+    \noalign{\smallskip}
+    Attribute                 & Value                                            \\
+    \noalign{\smallskip}
+    \hline
+    \noalign{\smallskip}
     program                   & \daemon{smtpd}                                   \\
@@
     priority                  & 0                                                \\
+    \noalign{\smallskip}
+    \hline
+    \noalign{\smallskip}
 \end{tabular}
 
-\noindent{}The \texttt{CONNECT} action is described in \sectionref{Actions}.
+\noindent{}The \texttt{CONNECT} action is described in \sectionref{example
+action}.
 
-\subsection{Architecture characteristics}
+\subsection{Architecture Characteristics}
 
@@
         actions when executing, providing some context sensitivity.
-        Actions can not inspect past or upcoming inputs, however allowing
-        actions can return a modified input line that will be parsed as if
-        read from the input stream, allowing for a simplified version of
+        Actions can not inspect past or upcoming inputs, however actions
+        can return a modified input line that will be parsed as if read
+        from the input stream, allowing for a simplified version of
         cascaded parsing~\cite{cascaded-parsing}.
@@
         matches the line is the rule which parses it and determines the
-        action that will be invoked.  The priority attribute may be used to
-        change the ordering of rules.
+        action that will be invoked.
 
@@
         decision, rather a consequence of the line oriented nature of the
-        log files; other implementations could pursue a more flexible
-        approach.
+        log files; more flexible approaches could be pursued.
 
@@
 \numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the results from first run discarded,
-and the execution time for the remaining 9 runs averaged.
+log lines) were each parsed 10 times, the results from the first run
+discarded, and the execution time for the remaining 9 runs averaged.
 
@@
 
-\subsection{Rule ordering}
+\subsection{Rule Ordering}
 
@@
 lines}, and peaks in fig.~\refwithpage{execution time vs file size vs
-number lines factor} (where a peak means that performance is better).  The
+number lines factor}, where a peak means that performance is better.  The
 explanation for this took some time to arrive at, but it turns out to be
@@
 
-These two characteristics combine to reduce the average number or rules
+These two characteristics combine to reduce the average number of rules
 required to parse a log line when there is a mail loop, as shown by the
@@
 expected that there will be a large number of rules.  There are
-\numberOFrules{} rules in the full ruleset, whereas the minimum number of
-rules required to parse the \numberOFlogFILES{} log files is
+\numberOFrules{} rules in the full Postfix ruleset, whereas the minimum
+number of rules required to parse the \numberOFlogFILES{} log files is
 \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the total.  A
@@
 extent with shuffled ordering.  The normal ordering (where are the most
-frequently matching rules are tried first) shows a
-\input{build/normal-vs-smaller-percentage-increase.tex} increase in
+frequently matching rules are tried first) shows a mean increase of 
+\input{build/normal-vs-smaller-percentage-increase.tex} in
 execution time for a \numberOFrulesMAXIMUMpercentage{} increase in the
@@
 
-The architecture's greatest strength is the ease with which is can be
+The architecture's greatest strength is the ease with which it can be
 adapted to deal with new requirements and inputs.  Parsing a variation of
@@
 patterns in the data which could be used to write new filters to recognise
-and reject spam rather than accepting it.  The current focus is on
-clustering and decision trees to optimise the order in which rules are
-applied; future efforts will involve using gathered data to train and test
-new filters.  This task is very similar to optimising a black box
-application based on its inputs and outputs, and this approach could be
-applied to optimising the behaviour of any system given sufficient logging
-to analyse.  An alternative approach to black box optimisation which uses
-application profiling in conjunction with the application's error messages
-to improve the error messages shown to users is described
+and reject spam rather than accepting it.  The parser provides the data in
+a normalised form which is far easier to use as input to new or existing
+implementations of the machine-learning algorithms than trying to adapt the
+algorithms to extract data directly from the log files.  The current focus
+is on clustering and decision trees to optimise the order in which rules
+are applied; future efforts will involve using data gathered by the parser
+to train and test new filters.  This task is very similar to optimising a
+black box application based on its inputs and outputs, and this approach
+could be applied to optimising the behaviour of any system given sufficient
+logging to analyse.  An alternative approach to black box optimisation
+which uses application profiling in conjunction with the application's
+error messages to improve the error messages shown to users is described
 in~\cite{black-box-error-reporting}; profiling data may be useful in

### 2008-07-07 doc/logparser-background.tex
Changed appendix to sectionref in most places; it's more consistent, and will we correct when I reorganise things.

@@
 parse Postfix log files are summarised (full details are available in
-appendix~\refwithpage{other-parsers}), finishing with a review of
-previously published research in this area.
+\sectionref{other-parsers}), finishing with a review of previously
+published research in this area.
 
@@
 
-Ten other parsers have been reviewed in
-appendix~\refwithpage{other-parsers} as part of the background research for
-this project.  None of the reviewed parsers perform the type of advanced
-parsing and log correlation described here; all are intended to perform a
-specific parsing and reporting task, rather than be a generic parser,
-extracting data and leaving generation of reports from the data to other
-programs.  Some parsers save the data extracted to a data store, but the
-many discard all data once they have finished running, making historical
-analysis impossible.  The other parsers reviewed all produce a report of
-varying complexity and detail, whereas the program described here doesn't
-attempt to produce a report at all; that responsibility is deferred to a
-separate program, to be developed later.  The parsing algorithm and program
-described here are designed to enable much more detailed log analysis by
-providing a stable platform for subsequent programs to develop upon.
+Ten other parsers have been reviewed in \sectionref{other-parsers} as part
+of the background research for this project.  None of the reviewed parsers
+perform the type of advanced parsing and log correlation described here;
+all are intended to perform a specific parsing and reporting task, rather
+than be a generic parser, extracting data and leaving generation of reports
+from the data to other programs.  Some parsers save the data extracted to a
+data store, but the many discard all data once they have finished running,
+making historical analysis impossible.  The other parsers reviewed all
+produce a report of varying complexity and detail, whereas the program
+described here doesn't attempt to produce a report at all; that
+responsibility is deferred to a separate program, to be developed later.
+The parsing algorithm and program described here are designed to enable
+much more detailed log analysis by providing a stable platform for
+subsequent programs to develop upon.
 

### 2008-07-07 doc/logparser-database-schema.tex
Changed RBL to DNSBL, and added a reference for it.
Changed appendix to sectionref in most places; it's more consistent, and will we correct when I reorganise things.

@@
     \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, appendix~\refwithpage{Glossary}).
+        epoch (explained in the glossary, \sectionref{Glossary}).
 
@@
     \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \RBL{}\@.
+        fields, e.g.\ the rejection message from an \DNSBL{}\@.
 

### 2008-07-07 doc/logparser-introduction.tex
Changed appendix to sectionref in most places; it's more consistent, and will we correct when I reorganise things.

--- 2008-05-27 doc/logparser-introduction.tex
@@
 messages can be added by the administrator with custom checks; every
-\RBL{}\footnote{This document is supplied with a glossary, see
-appendix~\ref{Glossary}.} returns a different explanatory message; policy
+\DNSBL{}\footnote{This document is supplied with a glossary, see
+\textsection\ref{Glossary}.} returns a different explanatory message; policy
 servers may log different messages depending on the characteristics of the
@@
 particular the optimisations used by the algorithm, utilising the graphs in
-appendix~\refwithpage{graphs} for illustrative purposes.  The section
-concludes with a description of using the tools provided with the parser to
-generate new rules (specifically the \regex{} in each rule) from unparsed
-log lines.
+\sectionref{graphs} for illustrative purposes.  The section concludes with
+a description of using the tools provided with the parser to generate new
+rules (specifically the \regex{} in each rule) from unparsed log lines.
 

### 2008-07-07 doc/logparser-parsing-rules.tex
Changed appendix to sectionref in most places; it's more consistent, and will we correct when I reorganise things.

@@
         effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in
-        appendix~\refwithpage{graphs}.
+        rules, with reference to the graphs in \sectionref{graphs}.
 

### 2008-07-07 doc/logparser.tex
Copied some of the things I learned writing the SGAI paper. I haven't decided either way on using Times New Roman or not. Reordered some stuff to group similar things together. Switched to varioref for references. Extract some more warnings after running make.

@@
 \usepackage{amstext}
-% The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
 % Include images
 \usepackage[final]{graphicx}
-% Change how nested enumerate environments are labelled.
-\renewcommand{\labelenumii}{\roman{enumii}:}
-\renewcommand{\refname}{Bibliography}
 % Add the bibliography into the table of contents.
 \usepackage[section,numbib]{tocbibind}
-% Provides commands to distinguish between pdf and dvi output.
-\usepackage{ifpdf}
+% Create a label for the last page.  Might be useful for "page 23/79" or
+% something.
+\usepackage{lastpage}
 
+% Extra packages recommended by Springer: they change the fonts from
+% Computer Modern to something else and tells Latex to use scalable fonts.
+\usepackage{type1cm}
+%\usepackage{mathptmx}
+%\usepackage{helvet}
+%\usepackage{courier}
+% Extra footnote functionality, including references to earlier footnotes.
+\usepackage{footmisc}
+
+% Sort numbers where there are multiple citations.  Doesn't appear to have
+% any effect (probably clashes with varioref or hyperref), though it does
+% reduce the space between numbers.
+\usepackage{cite}
+
+% Acronyms and glossary entries.
 % global=true prevents acronyms being expanded twice
@@
 
+% Produce nicer references, e.g. "section 4.3 on the next page"
+% Must be loaded before hyperref.
+\usepackage{varioref}
+% Warnings rather than errors when references cross page boundaries.
+\vrefwarning{}
+% Errors when loops are encountered
+\vrefshowerrors{}
+% Shorten the text used for references with page numbers.
+\renewcommand{\reftextfaraway}[1]{%
+    [p.~\pageref{#1}]%
+}
+% Replacement for \ref{}, adds the page number too.
+\newcommand{\refwithpage}[1]{%
+    \empty{}\vref{#1}%
+    %\empty{}\ref{#1} [p.~\pageref{#1}]%
+}
+% section references, automatically add \textsection
+\newcommand{\sectionref}[1]{%
+    \textsection{}\vref*{#1}%
+    %\textsection{}\refwithpage{#1}%
+}
+
+% Provides commands to distinguish between pdf and dvi output.
+\usepackage{ifpdf}
 % When creating a \PDF{} make the table of contents into links to the pages
@@
     \usepackage[dvips,hyperfootnotes=false]{hyperref}
+    % This is necessary for wrapping URLs in the bibliography when
+    % producing a dvi, but causes problems when generating \PDF{} output.
+    \usepackage{breakurl}
 \fi{}
@@
     pdfborder   = {0 0 0},
-%    breaklinks  = true,
 }
-% This is necessary for URLs in the bibliography; I dunno why the url
-% package doesn't work; don't include it when generating \PDF{} output.
-\ifpdf{}
-\else{}
-    \usepackage{breakurl}
-\fi{}
-\usepackage{lastpage}
 
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
-
-% Sort numbers where there are multiple citations.  Doesn't appear to have
-% any effect, though it does reduce the space between numbers.
-\usepackage{cite}
+\renewcommand{\refname}{Bibliography}
+% New formatting commands.
 
@@
 
-% Replacement for \ref{}, adds the page number too.
-\newcommand{\refwithpage}[1]{%
-    \empty{}\ref{#1} [p.~\pageref{#1}]%
-}
-% section references, automatically add \textsection
-\newcommand{\sectionref}[1]{%
-    \textsection{}\refwithpage{#1}%
-}
-
 % A command to format a Postfix daemon's name
@@
 
+% Constant values.
 \newcommand{\numberOFlogFILES}[0]{%
@@
 
-% This may keep hyperref happy about the Abstract entry in the table of
+% The name of the program, so I only have to change it in one place.
+\newcommand{\parsername}{\PLP{}}
+\newcommand{\parsernames}{\PLP{}'s}
+
+% This keeps hyperref fairly happy about the Abstract entry in the table of
 % contents.
@@
 
-XXX POSSIBLE CONFERENCES\@: AICS'S AI-2008 IN CORK, AND SGAI IN CAMBRIDGE
-IN DECEMBER\@.
-
 XXX WEKA FOR DECISION TREE DISPLAY\@.
-
-XXX IS THERE A SYMBOL FOR APPENDIX, LIKE \textsection{} FOR SECTIONS\@?
 
@@
 NECESSARY\@.
+
+XXX EXPLAIN THE X-AXIS --- MADS THOUGHT IT WAS UNCLEAR\@.
 

### 2008-07-09 doc/logparser-acronyms.tex
Corrections from Mads.

@@
 % % vim: set textwidth=75 spell :
-% Warning: don't use acronyms within definitions, they don't work properly.
+% Warning: do not use acronyms within definitions, they do not work
+% properly.
 
@@
 \newacronym{regex}{Regular Expression}{description={
-    Regular expressions are a powerful method of matching patterns against
+    Regular expressions are a concise and powerful method of matching patterns against
     text that are explained in detail in~\cite{Wikipedia-regex, perlre}.

### 2008-07-09 doc/logparser-background.tex
Corrections from Mads.

@@
 the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It's not as friendly, nor as
+only when legitimate mail is impeded.\/}  it is not as friendly, nor as
 catchy, but it more accurately reflects the current situation.} A typical
@@
 following the journey an email takes through Postfix can be quite
-difficult.  The log files tend to follow a 90\%-10\% pattern: 90\% of the
-time the journey is simple, but the other 10\% of the time accounts for
-90\% of the code in the parser.\footnote{These numbers don't have a solid
-scientific basis, they're based on gut feeling from writing and debugging
-the software.}
+difficult.  For the majority of mails the journey is simple and brief, but
+the remaining minority can be quite complex (see \sectionref{additional
+complications} for details).
 
 Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy},\footnote{This
-principle was also followed when designing the parser (\sectionref{parser
-design}), notably in having separate actions rather than one huge action.}
-and is separated into various component programs to perform the tasks
-required of an \MTA{}\@: receive mail, send mail, local delivery of mail,
-etc. --- full details can be found at~\cite{postfix-overview}.  Each log
-line contains the name of the Postfix component which produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).  This
-design has positive security implications also: those components which
-interact with other hosts are not privileged,\footnote{Privilege in this
-case means the power to perform actions that are limited to the
-administrator, and not available to ordinary users.} so bugs in those
-components will not give an attacker extra privileges; those components
-which are privileged do not interact with other hosts, making it much more
-difficult for an attacker to exploit any bugs which may exist in those
-components.
+do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
+into various component programs to perform the tasks required of an
+\MTA{}\@: receive mail, send mail, local delivery of mail, etc. --- full
+details can be found in~\cite{postfix-overview}.  Each log line contains
+the name of the Postfix component which produced it, and this information
+is used when determining which rules should be used to parse each log line
+(see \sectionref{rule characteristics} for details).  This design has
+positive security implications also: those components which interact with
+other hosts are not privileged,\footnote{Privilege in this case means the
+power to perform actions that are limited to the administrator, and not
+available to ordinary users.} so bugs in those components will not give an
+attacker extra privileges; those components which are privileged do not
+interact with other hosts, making it much more difficult for an attacker to
+exploit any bugs which may exist in those components.
 
@@
 evaluated until the first RCPT TO command is received, because some clients
-don't deal properly with rejections before RCPT TO\@; a benefit of this
+do not deal properly with rejections before RCPT TO\@; a benefit of this
 delay is that Postfix has more information available when logging
@@
 the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must remember that \textit{With great power
-comes great responsibility}.
+all users --- the administrator must be cognisant of the ramifications of
+their configuration changes.  XXX UNIX DOESN'T STOP YOU DOING STUPID THINGS
+BECAUSE THAT WOULD ALSO STOP YOU DOING CLEVER THINGS\@.
 
@@
 policy server is given state information\footnote{Sample state information
-is shown in table~\refwithpage{Example attributes sent to policy servers}}
+is shown in \tableref{Example attributes sent to policy servers}}
 and returns a verdict as described in \sectionref{Mixing and matching
@@
 source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time; the source code for
-Postfix 2.5.1 is 17MB.} there are advantages to treating Postfix as a black
-box while developing the parser:
+code would require a significant investment of time: there are 375,750
+lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source code.}
+there are advantages to treating Postfix as a black box while developing
+the parser:
 
@@
         takes through Postfix.  Full details of actions can be found in
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}.
+        \sectionref{actions-in-detail} and \sectionref{adding new actions}
+        XXX IMPROVE THIS SENTENCE\@.
 
@@
 Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and doesn't need
+actions: the framework provides services to the actions, and does not need
 to deal with the complications which arise, or the task of reconstructing a
@@
 
-XXX CHANGE THIS; THE PARSER DESIGN IS NOVEL, DON'T STRESS THE SIMILARITY
+XXX CHANGE THIS; THE PARSER DESIGN IS NOVEL, DO NOT STRESS THE SIMILARITY
 TOO MUCH\@.
@@
 produce a report of varying complexity and detail, whereas the program
-described here doesn't attempt to produce a report at all; that
+described here does not attempt to produce a report at all; that
 responsibility is deferred to a separate program, to be developed later.
@@
 DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats: it doesn't provide the functionality of a
+of the three output formats: it does not provide the functionality of a
 basic \SQL{} database, not can it be used with standard Unix text
@@
 
-The schema used with the MySQL database is undocumented, but at least it's
+The schema used with the MySQL database is undocumented, but at least it is
 possible to discover the schema with an \SQL{} database, unlike with
@@
 % re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also won't scale well.  There doesn't appear to
-% be any attempt to deal with the complications I've encountered: their
-% parsing isn't detailed enough to encounter them.  It doesn't run
-% properly; doesn't create any output; throws up errors.
+% at the main table - this also will not scale well.  There does not appear
+% to be any attempt to deal with the complications I have encountered:
+% their parsing is not detailed enough to encounter them.  It does not run
+% properly; does not create any output; throws up errors.
 
-\subsection{Conclusion}
+\subsection{Summary}
 

### 2008-07-09 doc/logparser-coverage.tex
Corrections from Mads.

@@
 into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it's
+important because the parser should handle all (relevant) log lines it is
 given; the second is equally important because the parser must properly
@@
         intermingled in the log file, and there are some Postfix programs
-        which don't produce any log lines of interest.
+        which do not produce any log lines of interest.
 
@@
 and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it's obvious that a
+for each rule over all \numberOFlogFILES{} log files; it is obvious that a
 small number of rules match the vast majority of the lines, and more than
@@
 
-\subsection{Conclusion}
+\subsection{Summary}
 

### 2008-07-09 doc/logparser-database-schema.tex
Corrections from Mads.

@@
-\section{Database schema}
+\section{Database Schema}
 \label{database schema}
@@
         Uninteresting log lines are parsed so that any lines the parser
-        isn't capable of handling become immediately obvious errors.
+        is not capable of handling become immediately obvious errors.
 
@@
         log lines the rule applies to.  This avoids needlessly trying rules
-        which won't match the log line, or worse, might match
+        which will not match the log line, or worse, might match
         unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which aren't parsed by program specific
+        against any log lines which are not parsed by program specific
         rules.
@@
     \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which isn't present in the log line: e.g.\ setting
+        piece of data which is not present in the log line: e.g.\ setting
         \texttt{smtp\_code} when mail is accepted.  The format and allowed
@@
     \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line doesn't contain a
-        queueid.  Many log lines won't contain a queueid, e.g.\ rejections
-        logged before a mail has been accepted (a queueid won't have been
-        allocated), or log lines which aren't tied to one particular mail.
+        gives the queueid, or zero if the log line does not contain a
+        queueid.  Many log lines will not contain a queueid, e.g.\
+        rejections logged before a mail has been accepted (a queueid will
+        not have been allocated), or log lines which are not tied to one
+        particular mail.
 
@@
         occasionally changes during a connection, presumably because spam
-        or virus senders think it's a good idea.  By default Postfix only
+        or virus senders think it is a good idea.  By default Postfix only
         logs the HELO hostname when it rejects an \SMTP{} command, but it
@@
         Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there won't be a queueid associated
-        with the mail for the first recipient, so that log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus it will be
-        logged by any restrictions taking effect in
+        place to log the HELO hostname, there will not be a queueid
+        associated with the mail for the first recipient, so that log line
+        cannot be associated with the correct mail.  There is guaranteed to
+        be a queueid when the DATA command has been reached, and thus it
+        will be logged by any restrictions taking effect in
         \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
@@
         the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines don't contain an \SMTP{} code: in those cases
+        delivery log lines do not contain an \SMTP{} code: in those cases
         the code is faked based on the success or failure represented by
@@
 with the purpose of each field discussed in detail.  A clear,
-comprehensible schema is essential when using the extracted data; it's more
-important when using the data than when storing it, because storing the
-data is a write-once operation, whereas utilising the data requires
+comprehensible schema is essential when using the extracted data; it is
+more important when using the data than when storing it, because storing
+the data is a write-once operation, whereas utilising the data requires
 frequent searching, sorting and manipulation of the data to produce

### 2008-07-09 doc/logparser-graphs.tex
Added graph showing distribution of rules per action.
Added normal-vs-smaller graph; nothing written about it yet.
Corrections from Mads.

@@
 XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
+
+XXX DO SOMETHING WITH NORMAL VS SMALLER GRAPH\@.
+
+\showgraph{build/plot-normal-vs-smaller}{NORMAL VS SMALLER}{NORMAL VS
+SMALLER}
 
@@
         delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it's currently delivering to, and therefore
+        contains the address it is currently delivering to, and therefore
         sends a bounce notification with sender address \textit{$<>$\/} to
@@
 much lower; this results in far fewer \regex{} compilations per line than
-when there isn't a mail loop, and a correspondingly decreased execution
+when there is not a mail loop, and a correspondingly decreased execution
 time.  The increases in execution time when not caching \regexes{} for
@@
 
+\showgraph{build/plot-action-distribution}{Distribution of rules per
+action}{Distribution of rules per action}
+
 \clearpage
 
-\subsection{Conclusion}
+\subsection{Summary}
 

### 2008-07-09 doc/logparser-introduction.tex
Corrections from Mads.

@@
 \DNSBL{}\footnote{This document is supplied with a glossary, see
-\textsection\ref{Glossary}.} returns a different explanatory message; policy
-servers may log different messages depending on the characteristics of the
-connection; there are many ways in which the log lines may differ between
-servers, even within the same organisation --- servers may be configured
-differently, or running different versions of Postfix.  This paper
-documents the difficult process of parsing Postfix log files, presenting
-\PLP{}, a program which parses log files and places the resulting data into
-a database for later use.  The gathered data can then be used to optimise
-current anti-spam measures, provide a baseline to test new anti-spam
-measures against, or to produce statistics showing how effective those
-measures are.  There are numerous other uses for such data: improving
-server performance by identifying troublesome destinations and
+\textsection\ref{Glossary}.} returns a different explanatory message;
+policy servers may log different messages depending on the characteristics
+of the connection; there are many ways in which the log lines may differ
+between servers, even within the same organisation --- servers may be
+configured differently, or running different versions of Postfix.  This
+paper documents the difficult process of parsing Postfix log files,
+presenting \PLP{}, a program which parses log files and places the
+resulting data into a database for later use.  The gathered data can then
+be used to optimise current anti-spam measures, provide a baseline to test
+new anti-spam measures against, or to produce statistics showing how
+effective those measures are.  There are numerous other uses for such data:
+improving server performance by identifying troublesome destinations and
 reconfiguring appropriately; identifying regular high volume uses (e.g.\

### 2008-07-09 doc/logparser-limitations.tex
Changed RBL to DNSBL, and added a reference for it.
Corrections from Mads.

@@
     \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \RBL{}, new policy
+        with the new log lines.  Similarly using a new \DNSBL{}, new policy
         server or new administrator defined rejection messages require new
@@
         by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It doesn't appear to be possible
+        \sectionref{connection reuse}).  It does not appear to be possible
         to make this distinction given the data Postfix logs, though it

### 2008-07-09 doc/logparser-other-parsers-review.tex
Corrections from Mads.

@@
 
-    \item A minor difference is that most parsers don't handle compressed
+    \item A minor difference is that most parsers do not handle compressed
         files; both \parsername{} and Splunk handle them transparently,
@@
 each sender, unlike the copious amounts of data extracted by \parsername{}.
-It doesn't utilise rejection log lines in any way, so is unsuitable for the
-purposes of this project.  Its parsing is completely inextensible, indeed
-is almost incomprehensible, relying on \texttt{scanf(3)}, fixed offsets and
-low level string manipulation; it is the opposite end of the spectrum to
-\parsernames{} parsing.  It doesn't handle any of the complications
-discussed in this document, doesn't gather the breadth of data required for
-this project, and ignores the majority of log lines produced by Postfix.
-It ignores all log lines earlier than today, so testing with the test log
-files is pointless.
+It does not utilise rejection log lines in any way, so is unsuitable for
+the purposes of this project.  Its parsing is completely inextensible,
+indeed is almost incomprehensible, relying on \texttt{scanf(3)}, fixed
+offsets and low level string manipulation; it is the opposite end of the
+spectrum to \parsernames{} parsing.  It does not handle any of the
+complications discussed in this document, does not gather the breadth of
+data required for this project, and ignores the majority of log lines
+produced by Postfix.  It ignores all log lines earlier than today, so
+testing with the test log files is pointless.
 
@@
 This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It's
-designed to provide real time alerts when dictionary attacks, mail loops or
-similar problems occur. It doesn't attempt to extract or save data,
+time and warns if the frequency of events passes a certain threshold.  It
+is designed to provide real time alerts when dictionary attacks, mail loops
+or similar problems occur. It does not attempt to extract or save data,
 correlate log lines, or any of the more advanced tasks described in this
@@
 
-Anteater doesn't have any English documentation so it's difficult, nigh
+Anteater does not have any English documentation so it is difficult, nigh
 impossible, for this author to accurately comment on what analysis it
@@
 Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line doesn't match the \regex{}, though it appears that
+saved when the log line does not match the \regex{}, though it appears that
 they will not overwrite existing data.  Most notably rejections are
 completely ignored, making it unsuitable for the purposes of this project.
-It doesn't attempt to address any of the complications in
+It does not attempt to address any of the complications in
 \sectionref{complications} \sectionref{additional complications} except for
@@
 Logrep extracts far less data from Postfix log files than \parsername{},
-completely ignores rejections, is effectively undocumented, doesn't deal
+completely ignores rejections, is effectively undocumented, does not deal
 with the more complicated aspects of Postfix log files, and at the time of
-writing doesn't work properly.
+writing does not work properly.
 
@@
 lines.  Most other parsers immediately generate a report and discard the
-data extracted from the log files; those which don't discard the data
+data extracted from the log files; those which do not discard the data
 retain it in a format inaccessible to other tools.  Nearly all of the

### 2008-07-09 doc/logparser-parsing-algorithm.tex
Added graph showing distribution of rules per action.
Corrections from Mads.

@@
-\section{Parsing algorithm}
+\section{Parsing Algorithm}
 
@@
 
-Where the rules are quite simple and each rule is completely independent of
-its fellows, the algorithm is significantly more complicated and highly
-internally interdependent.  The algorithm deals with all the complications
-of parsing, the eccentricities and oddities of Postfix log files,
-presenting the resulting data in a normalised, easy to use representation.
-The algorithm's task is to follow the journey each mail takes through
-Postfix, combining the data extracted by rules into a coherent whole,
-saving it in a useful and consistent form, and performing housekeeping
-duties.
+Whereas the rules are quite simple and each rule is completely independent
+of the other rules, the algorithm is significantly more complicated and
+highly internally interdependent.  The algorithm deals with all the
+complications of parsing, the eccentricities and oddities of Postfix log
+files, presenting the resulting data in a normalised, easy to use
+representation.  The algorithm's task is to follow the journey each mail
+takes through Postfix, combining the data extracted by rules into a
+coherent whole, saving it in a useful and consistent form, and performing
+housekeeping duties.
 
@@
 
-    \item Follow the progress of the accepted mail until it's either
+    \item Follow the progress of the accepted mail until it is either
         delivered, bounced or deleted, then log all data, and finish.
@@
 requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails won't trample over each other's data.
-This must be done every time a mail is accepted, as it's impossible to tell
-in advance which connections will accept multiple mails.  Happily, once the
-mail has been accepted log entries won't be correlated by \pid{} for that
-mail any more (its queueid will be used instead), so there isn't any
-ambiguity about which mail a given log line belongs
+accepted, so that subsequent mails will not trample over each other's data.
+This must be done every time a mail is accepted, as it is impossible to
+tell in advance which connections will accept multiple mails.  Happily,
+once the mail has been accepted log entries will not be correlated by
+\pid{} for that mail any more (its queueid will be used instead), so there
+is not any ambiguity about which mail a given log line belongs
 to.\footnote{Unfortunately this statement is not completely accurate: see
 \sectionref{timeouts-during-data-phase} for details.  However in general
-there isn't any ambiguity about which data structure should be used for a
+there is not any ambiguity about which data structure should be used for a
 given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it doesn't have any data worth saving, i.e.\ no
+the client disconnects if it does not have any data worth saving, i.e.\ no
 rejections.  One unsolved difficulty is distinguishing between different
@@
 paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it won't log that the child has been created until it is
+possess, so it will not log that the child has been created until it is
 absolutely certain that the mail has been written to disk.}
@@
         Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there won't be any data to save (any mails
+        the database; if not there will not be any data to save (any mails
         accepted will already have been cloned so their data is in another
@@
     \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent hasn't yet been identified (action: COMMIT\@;
+        whose parent has not yet been identified (action: COMMIT\@;
         flowchart:~15).  Mark the mail as ready for entry in the database
@@
     \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but doesn't either provide
-        any useful data or require anything to be done.
+        line needs to be parsed for completeness but does not either
+        provide any useful data or require anything to be done.
 
@@
         which case there's a data structure to dispose of, or it may not in
-        which case there isn't.  See
+        which case there is not.  See
         \sectionref{timeouts-during-data-phase} for the gory details.
@@
 \end{description}
+
+The distribution of rules per action is shown in
+graph~\refwithpage{Distribution of rules per action}.
 
@@
         \daemon{qmgr} log line; lack of a \daemon{cleanup} line is a sure
-        sign the mail didn't make it too far.
+        sign the mail did not make it too far.
 
@@
 disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix doesn't
+parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix does not
 log a message saying the mail has been discarded, as should be expected by
@@
         which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail didn't make
-        it too far.
+        lack of a \daemon{cleanup} line is a sure sign the mail did not
+        make it too far.
 
@@
 
-    \item The queueid must not have been reused yet, i.e.\ there isn't an
+    \item The queueid must not have been reused yet, i.e.\ there is not an
         entry in the state tables for the queueid.
@@
 must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it's
+SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it is
 found; if the additional log lines have caused the mail to reach a valid
@@
 faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it's the result of a mail loop, mail bomb, or
+some of the mail (because it is the result of a mail loop, mail bomb, or
 joe~job) before Postfix has had a chance to try to deliver it.  In this
@@
 notification mails, which the BOUNCE action subsequently checks if the
-bounce mail isn't already in the state tables. If a mail exists in the
+bounce mail is not already in the state tables. If a mail exists in the
 cache under the correct queueid, and its start time is less than ten
@@
 the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they aren't specified anywhere in the parser.  The
-framework which holds the parser together was covered next, after which
+paths in the flow chart, they are not specified anywhere in the parser.
+The framework which holds the parser together was covered next, after which
 came a description of the current actions provided by the parser, and the

### 2008-07-09 doc/logparser-parsing-rules.tex
Corrections from Mads.

@@
-\section{Parsing rules}
+\section{Parsing Rules}
 
 \label{rules}
-
-\subsection{Introduction}
 
@@
 
-% Don't reformat this!
+% do not reformat this!
 \begin{tabular}[]{ll}
@@
         action will be executed, if not the next rule will be tried.  If
-        the program-specific rules don't match the log line, the parser
+        the program-specific rules do not match the log line, the parser
         will fall back to generic rules; if those rules are unsuccessful a

### 2008-07-09 doc/logparser-acronyms.tex
Corrections from Mads and chktex.

@@
     would be a delivery agent or a mail store.  Full details are available
-    in \cite{lmtp-rfc}.
+    in~\cite{lmtp-rfc}.
 }}
@@
 \newacronym{MTA}{Mail Transfer Agent}{description={
-    A Mail Transfer Agent sends and/or receives mail via SMTP.
+    A Mail Transfer Agent sends and/or receives mail via SMTP\backslashat{}.
 }}
@@
     The Request For Comments series is a series of proposals defining
-    various protocols and file formats, e.g. SMTP.  The name is somewhat
-    misleading: initially the authors were asking for peer review, but
-    these documents are now the de facto standards the Internet runs on.
+    various protocols and file formats, e.g.\ SMTP\backslashat{}.  The name
+    is somewhat misleading: initially the authors were asking for peer
+    review, but these documents are now the de facto standards the Internet
+    runs on.
 }}
@@
     and maintenance.  Some information about SQL, including its history,
-    can be found at \cite{Wikipedia-sql}; a good introduction can be found
+    can be found at~\cite{Wikipedia-sql}; a good introduction can be found
     at~\cite{sql-for-web-nerds}, others are~\cite{w3schools-sql-tutorial,

### 2008-07-09 doc/logparser-background.tex
Corrections from Mads and chktex.

@@
 
-XXX CLARIFY THE SENTENCE STARTING WITH "Although this may seem" IN THE NEXT
-PARAGRAPH\@.
+XXX CLARIFY THE SENTENCE STARTING WITH ``Although this may see'' IN THE
+NEXT PARAGRAPH\@.
 
@@
 
-XXX CHANGE THIS; THE PARSER DESIGN IS NOVEL, DO NOT STRESS THE SIMILARITY
+XXX CHANGE THIS\@; THE PARSER DESIGN IS NOVEL, DO NOT STRESS THE SIMILARITY
 TOO MUCH\@.
@@
 multiple values for a field (e.g.\ multiple recipients) and adding new
-fields.  There is no facility to include documentation of the fields,
-separator, etc., unlike \SQL{} databases where every database includes a
-schema naming the fields and the type of data they store (integer, text,
-timestamp, etc.).
+fields.  There is no facility to document the fields or the separator,
+unlike \SQL{} databases where every database includes a schema naming the
+fields and the type of data they store (integer, text, timestamp, etc.).
 

### 2008-07-09 doc/logparser-coverage.tex
Corrections from Mads and chktex.

@@
 \label{parsing coverage}
-
-\subsection{Introduction}
 
@@
 making it impossible to quantify the false positive rate; however a random
-sample of 6039 log lines was parsed and the results checked manually to
+sample of 6,039 log lines was parsed and the results checked manually to
 ensure that the correct \regex{} parsed each log line.\footnote{Each log

### 2008-07-09 doc/logparser-glossary.tex
Corrections from Mads and chktex.

@@
+% vim: set filetype=tex :
 % The contents of the glossary.
@@
 
-\glossary{name={awk},description={AWK is a general purpose programming
-    language that is designed for processing text-based data, and is
-    available as a standard utility on all Unix systems.
+\glossary{name={awk},description={
+    AWK is a general purpose programming language that is designed for
+    processing text-based data, and is available as a standard utility on
+    all Unix systems.
 }}
 
-\glossary{name={syslog},description={Syslog is the standard logging
-    mechanism on Unix systems: the program sends log messages to syslog,
-    then syslog filters and stores the messages according to the
-    configuration the administrator has chosen.
+\glossary{name={syslog},description={
+    Syslog is the standard logging mechanism on Unix systems: the program
+    sends log messages to syslog, then syslog filters and stores the
+    messages according to the configuration the administrator has chosen.
 }}
+
+\glossary{name={mail bomb},description={
+    A mail bomb occurs when an attacker inflicts a huge volume of mail on
+    the victim.  At best a mail bomb is irritating to the victim; at worst
+    the deluge of mail can be severe enough to: interrupt service for the
+    victim and/or other users; cause mail to be rejected because the victim
+    has reached a limit (e.g.\ too many mails, too much disk space
+    consumed); the victim may accidentally delete other mail while trying
+    to cope with the mail bomb.
+}}
+
+\glossary{name={mail loop},description={
+    Sometimes mail set to one address must be delivered to a different
+    address instead, e.g.\ because a person has changed jobs.  A mail loop
+    occurs when the recipient addresses constitute a cyclic directed graph;
+    the simplest example is when mail for \texttt{foo\at{}example.net} is
+    delivered to \texttt{bar\at{}example.com}, and mail for
+    \texttt{bar\at{}example.com} is delivered to
+    \texttt{bar\at{}example.com}.
+}}

### 2008-07-09 doc/logparser-other-parsers-review.tex
Corrections from Mads and chktex.

@@
 \label{other-parsers}
-
-\subsection{Introduction}
 
@@
 
-\subsection{Parsers reviewed}
+\subsection{Pflogsumm}
 
-In this section text in italics is a quote extracted from the documentation
-of the program being reviewed.
+\begin{quotation}
 
-\subsubsection{Pflogsumm}
+    pflogsumm is designed to provide an over-view of Postfix activity, with
+    just enough detail to give the administrator a ``heads up'' for
+    potential trouble spots.
 
-\textit{pflogsumm is designed to provide an over-view of Postfix activity,
-with just enough detail to give the administrator a ``heads up'' for
-potential trouble spots.\/}
+\end{quotation}
 
@@
 
-\subsubsection{Sawmill Universal Log File Analysis and Reporting}
+\subsection{Sawmill Universal Log File Analysis and Reporting}
 
-\textit{ Sawmill is a Postfix log analyzer (it also support 686 other log
-formats).  It can process log files in Postfix format, and generate dynamic
-statistics from them, analyzing and reporting events.  Sawmill can parse
-Postfix logs, import them into a SQL database (or its own built-in
-database), aggregate them, and generate dynamically filtered reports, all
-through a web interface.  Sawmill can perform Postfix analysis on any
-platform, including Window, Linux, FreeBSD, OpenBSD, Mac OS, Solaris, other
-UNIX, and more.\/}
+\begin{quotation}
+
+    Sawmill is a Postfix log analyzer (it also support [sic] 686 other log
+    formats).  It can process log files in Postfix format, and generate
+    dynamic statistics from them, analyzing and reporting events.  Sawmill
+    can parse Postfix logs, import them into a SQL database (or its own
+    built-in database), aggregate them, and generate dynamically filtered
+    reports, all through a web interface.  Sawmill can perform Postfix
+    analysis on any platform, including Window, Linux, FreeBSD, OpenBSD,
+    Mac OS, Solaris, other UNIX, and more.
+
+\end{quotation}
 
@@
 
-\subsubsection{Splunk}
+\subsection{Splunk}
 
-\textit{Splunk is an IT Search engine. It is software that indexes any
-format of IT data from any source in real time, including logs,
-configurations, scripts, code, messages, traps, alerts, activity reports,
-stack traces and metrics from all of your applications, servers and
-devices. Splunk lets you search, navigate, alert and report on all your IT
-data in real time using an AJAX web interface. You can also share knowledge
-and Splunk solutions with other members of the Splunk community via
-SplunkBase.\/}
+\begin{quotation}
+
+    Splunk is an IT Search engine. It is software that indexes any format
+    of IT data from any source in real time, including logs,
+    configurations, scripts, code, messages, traps, alerts, activity
+    reports, stack traces and metrics from all of your applications,
+    servers and devices. Splunk lets you search, navigate, alert and report
+    on all your IT data in real time using an AJAX web interface. You can
+    also share knowledge and Splunk solutions with other members of the
+    Splunk community via SplunkBase.
+
+\end{quotation}
 
@@
 \$5000 (plus \$1000 support), and still limits the volume of data indexed
-per day to 500MB.
+per day to 500MB\@.
 
@@
 
-\subsubsection{Isoqlog}
+\subsection{Isoqlog}
 
-\textit{Isoqlog is an MTA log analysis program written in C. It designed to
-scan qmail, postfix, sendmail and exim logfile and produce usage statistics
-in \HTML{} format for viewing through a browser. It produces Top domains
-output according to Sender, Receiver, Total mails and bytes; it keeps your
-main domain mail statistics with regard to Days Top Domain, Top Users
-values for per day, per month and years.\/}
+\begin{quotation}
+
+    Isoqlog is an MTA log analysis program written in C. It designed to
+    scan qmail, postfix, sendmail and exim logfile and produce usage
+    statistics in \HTML{} format for viewing through a browser. It produces
+    Top domains output according to Sender, Receiver, Total mails and
+    bytes; it keeps your main domain mail statistics with regard to Days
+    Top Domain, Top Users values for per day, per month and years.
+
+\end{quotation}
 
@@
 
-\subsubsection{AWStats}
+\subsection{AWStats}
 
-\textit{AWStats is a free powerful and featureful tool that generates
-advanced web, streaming, ftp or mail server statistics, graphically. This
-log analyzer works as a CGI or from command line and shows you all possible
+\begin{quotation}
+
+    AWStats is a free powerful and featureful tool that generates advanced
+    web, streaming, ftp or mail server statistics, graphically. This log
+    analyzer works as a CGI or from command line and shows you all possible
     information your log contains, in few graphical web pages. It uses a
-partial information file to be able to process large log files, often and
-quickly. It can analyze log files from all major server tools like Apache
-log files (NCSA combined/XLF/ELF log format or common/CLF log format),
-WebStar, IIS (W3C log format) and a lot of other web, proxy, wap, streaming
-servers, mail servers and some ftp servers.\/}
+    partial information file to be able to process large log files, often
+    and quickly. It can analyze log files from all major server tools like
+    Apache log files (NCSA combined/XLF/ELF log format or common/CLF log
+    format), WebStar, IIS (W3C log format) and a lot of other web, proxy,
+    wap, streaming servers, mail servers and some ftp servers.
+
+\end{quotation}
 
@@
 
-\subsubsection{Throughput monitor}
+\subsection{Throughput monitor}
 
@@
 
-\subsubsection{Anteater}
+\subsection{Anteater}
 
-\textit{The Anteater project is a Mail Traffic Analyser. Anteater supports
-currently the logformat produced by Sendmail and by Postfix. The tool is
-written in 100\% C++ and is very easy to customize. Input, output, and the
-analysis are modular class objects with a clear interface. There are eight
-useful analyse modules, writing the result in plain ASCII or \HTML{}, to
-stdout or to files.\/}
+\begin{quotation}
+
+    The Anteater project is a Mail Traffic Analyser. Anteater supports
+    currently the logformat produced by Sendmail and by Postfix. The tool
+    is written in 100\% C++ and is very easy to customize. Input, output,
+    and the analysis are modular class objects with a clear interface.
+    There are eight useful analyse modules, writing the result in plain
+    ASCII or \HTML{}, to stdout or to files.
+
+\end{quotation}
 
@@
 
-\subsubsection{Yet Another Advanced Logfile Analyser}
+\subsection{Yet Another Advanced Logfile Analyser}
 
-\textit{yaala is a very flexible analyser for all kinds of logfiles. It
-uses parsers to extract information from a logfile, an SQL-like query
-language to relate the information to each other and an output-module to
-format the information appropriately.\/}
+\begin{quotation}
+
+    yaala is a very flexible analyser for all kinds of logfiles. It uses
+    parsers to extract information from a logfile, an SQL-like query
+    language to relate the information to each other and an output-module
+    to format the information appropriately.
+
+\end{quotation}
 
@@
 
-\subsubsection{Lire}
+\subsection{Lire}
 
-\textit{As any good system administrator knows, there's a lot more to keep
-track of in an active network than just webservers. Lire is hands down the
-most versatile log analysis software available today. Lire not only keeps
-you informed about your HTTP, FTP, and mail traffic, it also reports on
-your firewalls, your print servers, and your DNS activity. The ever growing
-list of Lire-supported services clearly outstrips any other software, in
-large part thanks to the numerous volunteers who have pioneered many new
-services and features. Lire is a total solution for your log analysis
-needs.\/}
+\begin{quotation}
+
+    As any good system administrator knows, there's a lot more to keep
+    track of in an active network than just webservers. Lire is hands down
+    the most versatile log analysis software available today. Lire not only
+    keeps you informed about your HTTP, FTP, and mail traffic, it also
+    reports on your firewalls, your print servers, and your DNS activity.
+    The ever growing list of Lire-supported services clearly outstrips any
+    other software, in large part thanks to the numerous volunteers who
+    have pioneered many new services and features. Lire is a total solution
+    for your log analysis needs.
+
+\end{quotation}
 
@@
 
-\subsubsection{Logrep}
+\subsection{Logrep}
 
-\textit{Logrep is a secure multi-platform framework for the collection,
+\begin{quotation}
+
+    Logrep is a secure multi-platform framework for the collection,
     extraction, and presentation of information from various log files. It
     features HTML reports, multi dimensional analysis, overview pages, SSH
-communication, and graphs, and supports over 30 popular systems including
-Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains, iptables, NT
-event logs, Firewall-1, wtmp, xferlog, Oracle listener and Pix.\/}
+    communication, and graphs, and supports over 30 popular systems
+    including Snort, Squid, Postfix, Apache, Sendmail, syslog, ipchains,
+    iptables, NT event logs, Firewall-1, wtmp, xferlog, Oracle listener and
+    Pix.
+
+\end{quotation}
 
@@
 
-\subsubsection{Log Mail Analyser}
+\subsection{Log Mail Analyser}
 
@@
     \hline         &             &                 &               &                &        \\
-    Pflogsumm      & Yes         & No          & Partial\dag   & Good           & Yes    \\
+    Pflogsumm      & Yes         & No              & Partial\dag{} & Good           & Yes    \\
     Sawmill        & Yes         & Yes             & Searches      & Reasonable     & No     \\
@@
     Isoqlog        & No          & Yes             & No            & No             & Yes    \\
-    AWStats        & Partially   & Yes         & Partial\dag   & Good           & Yes    \\
+    AWStats        & Partially   & Yes             & Partial\dag{} & Good           & Yes    \\
     Throughput     & N/A         & No              & No            & Brief but good & Yes    \\
@@
     Anteater       & No          & No              & No            & No             & Yes    \\
-    YAALA          & No          & Yes\ddag    & Searches      & Poor           & Yes    \\
+    YAALA          & No          & Yes\ddag{}      & Searches      & Poor           & Yes    \\
     Lire           & Yes         & Yes             & Yes           & Yes            & Yes    \\
@@
     \LMA{}         & No          & Yes             & No            & No             & Yes    \\ 
-    \parsername{}  & Yes         & Yes\dag\dag & No            & Yes            & Yes    \\
+    \parsername{}  & Yes         & Yes\dag{}\dag{} & No            & Yes            & Yes    \\
 \end{tabular}
 
-\dag Sections can be omitted from a report, but extra sections can not be
+\dag{} Sections can be omitted from a report, but extra sections can not be
 added.
 
-\ddag YAALA only stores the data required to produce the latest report;
+\ddag{} YAALA only stores the data required to produce the latest report;
 other data will be discarded.
 
-\dag\dag\parsername{} is the only parser with documentation for its data
-store.
+\dag{}\dag{}\parsername{} is the only parser with documentation for its
+data store.
 

### 2008-07-09 doc/logparser-parsing-algorithm.tex
Corrections from Mads and chktex.

@@
 \subsection{Complications encountered}
+
+XXX DUNNO IF THIS SECTION SHOULD BE HERE\@.  QUEUEID VS PID ISN'T TOO
+IMPORTANT\@; CONNECTION REUSE COULD EASILY BE MOVED TO ADDITIONAL
+COMPLICATIONS\@; RE-INJECTED MAILS SHOULD PROBABLY BE MOVED TO ADDITIONAL
+COMPLICATIONS TOO\@.
 
@@
 the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but the author believes that it is
-better to require an explicit source, as bugs in the parser are more likely
-to be exposed; such checks helped to identify the complications described
-in \sectionref{discarding cleanup lines} and \sectionref{pickup logging
-after cleanup}.
+checking of where mails originate from, but requiring an explicit source
+helps to expose bugs in the parser; such checks helped to identify the
+complications described in \sectionref{discarding cleanup lines} and
+\sectionref{pickup logging after cleanup}.
 
@@
 
-Figures~\refwithpage{flow chart image part 1} and~\refwithpage{flow chart
-image part 2} show the paths the data representing a mail can take through
-the parser algorithm.  The flow chart covers the most common paths only;
-there are additional, uncommon paths which are excluded for the sake of
-clarity; details of the deviations can be found in \sectionref{additional
-complications}.
+XXX CONSIDER REMOVING PART TWO OF THE FLOW CHART\@: I DON'T THINK THAT
+RE-INJECTION IS IMPORTANT ENOUGH TO WARRANT IT\@.  THE EXPLANATION OF
+TRACKING (\textsection{}5.4.3) COULD BE MOVED TO ADDITIONAL
+COMPLICATIONS\@.
+
+Figure~\refwithpage{flow chart image part 1} and figure~\refwithpage{flow
+chart image part 2} show the paths the data representing a mail can take
+through the parser algorithm.  The flow chart covers the most common paths
+only; there are additional, uncommon paths which are excluded for the sake
+of clarity; details of the deviations can be found in
+\sectionref{additional complications}.
 

### 2008-07-10 doc/logparser-background.tex
Corrections from my last proofreading.

@@
 \label{background}
-
-\subsection{Introduction}
 
@@
 A snippet of \SQL{} provides a short example of the optimisation possible
-using data from the database is determining which Postfix restrictions
-reject the highest number of mails:
+using data from the database: determining which Postfix restrictions reject
+the highest number of mails:
 
@@
 
-    \item XXX REWRITE THIS PARAGRAPH\@.  The majority of programming
-        languages have existing code available providing access to
-        databases, so the majority of languages will be able to access the
-        results obtained by the parser.  If this parser was written to be
-        utilised as shared code there would be little or no difficulty in
-        using it when writing in the programing language it is implemented
-        in, but every other programing language wishing to utilise it would
-        require an interface layer to be written.  This requirement would
-        drastically reduce the viability of any project wishing to build
-        upon the work of this project.
+    \item It is already possible to access databases from the vast majority
+        of programming languages, allowing a developer to access the
+        data gathered using their preferred programming language, rather
+        than being restricted to the language the parser is written in.  It
+        is often possible to write an interface layer allowing code written
+        in one language to be used in another language, but this greatly
+        increases the effort required to use the parser.
 
@@
 Storing the results in a database will also increase the efficiency of
-using those results, as the log files need only be parsed once; indeed the
-results may be used by someone with no access to the original log files.
+using those results, as the log files need only be parsed once rather than
+each time the data is used; indeed the results may be used by someone with
+no access to the original log files.
 

### 2008-07-10 doc/logparser-database-schema.tex
Corrections from my last proofreading.

@@
 
-The database schema can be conceptually divided in two: the rules which are
-used to parse log files, and the data saved from the parsing of log files.
-Rules have the fields required to parse the log lines, extract data to be
-saved, and the action to be executed; they also have several fields which
-aid the user in understanding what each rule parses.  The rules are
-described in detail in \sectionref{rules} but the fields are covered here
-in \sectionref{rule attributes}.
+The database schema can be conceptually divided in two: the rules, which
+are used to parse log files, and the data saved from the parsing of log
+files.  Rules have the fields required to parse the log lines, extract data
+to be saved, and the action to be executed; they also have several fields
+which aid the user in understanding the meaning of the log lines parsed by
+each rule.  The rules are described in detail in \sectionref{rules} but the
+fields are covered in \sectionref{rule attributes}.
 
@@
 a rejection; the individual fields will be described in
-\sectionref{connections table}.  The results table has one or more rows per
-row in the connections table; the fields will be covered in detail in
-\sectionref{results table}.
+\sectionref{connections table}.  There will be at least one row in the
+results table for each row in the connections table; the fields will be
+covered in detail in \sectionref{results table}.
 
@@
 database is the connection between rules and results: if more information
-is required when examining a result, the rules which produced the database
-entries are available for inspection (and each result references the rule
-which created it).  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
+is required when examining a result, the rule which produced the result is
+available for inspection --- each result references the rule which created
+it.  There is no ambiguity about which rule resulted in a particular
+result, eliminating one potential source of confusion.
 
@@
 hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying action to
-take for that log line.
+to be saved in the connections and results tables, and specifying the
+action to take for that log line.
 
-Each rule defines the following:
+The following attributes are defined for each rule:
 
@@
         log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the action causing the
-        log lines this rule matches.
+        line to be logged).  This field describes the event causing the log
+        lines this rule matches.
 
@@
 
-            \item [PROCESSING] \daemon{cleanup} is processing a mail ---
-                see~\cite{postfix-cleanup} for details of the processing
-                performed by \daemon{cleanup}.
+            \item [PROCESSING] \daemon{cleanup} is processing a mail
+                (see~\cite{postfix-cleanup} for details).
 
@@
         discussed in \sectionref{rule efficiency}, with the impact of
-        compiling and caching \regexes{} covered in \sectionref{Caching
-        each regex}.
+        caching compiled \regexes{} covered in \sectionref{Caching each
+        regex}.
 
@@
         with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  The list of acceptable
-        variable names is:
+        the \regex{} on the right hand side.  Data specified by
+        \texttt{result\_cols} will be entered in the \texttt{results}
+        table, while \texttt{connection\_cols} populates the
+        \texttt{connection} table.  The list of acceptable variable names
+        is:
 
@@
 
-        Additionally \texttt{child} and \texttt{pid} are used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions.
+        Additionally \texttt{child} and \texttt{pid} are accepted and used
+        respectively by the \texttt{TRACK}, \texttt{BOUNCE} and
+        \texttt{SMTPD\_DIED} actions, but do not cause data to be entered
+        in the database.
 
@@
         data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This is intended for use with small amounts of
-        data rather than large amounts in any one rule, so dealing with
-        escape sequences was deemed unnecessary.} is permitted on the right
-        hand side of the assignment.
+        cannot be used.  This feature is intended for use with small
+        amounts of data rather than large amounts in any one rule, so
+        dealing with escape sequences was deemed unnecessary.} is permitted
+        on the right hand side of the assignment.
 
@@
 
-    \item [queueid] Specifies the matching capture from the \regex{} which
-        gives the queueid, or zero if the log line does not contain a
+    \item [queueid] Specifies the capturing field in the \regex{}
+        containing the queueid, or zero if the log line does not contain a
         queueid.  Many log lines will not contain a queueid, e.g.\
@@
         the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits.  Assuming that the
-        distribution of log lines is reasonably consistent between log
-        files, rules matching more commonly occurring log lines will be
-        tried before rules matching less commonly occurring log lines,
-        lowering the program's execution time.  Rule ordering for
-        efficiency is discussed in \sectionref{rule ordering for
-        efficiency}.
+        end of the run updates every rule's hits field in the database.
+        Assuming that the distribution of log lines is reasonably
+        consistent between log files, rules matching more commonly
+        occurring log lines will be tried before rules matching less
+        commonly occurring log lines, reducing the program's execution
+        time.  Rule ordering for efficiency is discussed in
+        \sectionref{rule ordering for efficiency}.
 
@@
 
-    \item [priority] This is the user-configurable companion to hits: rules
-        will be tried in order of priority, overriding hits.  This allows
+    \item [priority] This is the user-configurable companion to hits: when
+        the list of rules is sorted, priority overrides hits.  This allows
         more specific rules to take precedence over more general rules
@@
     \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local server when receiving mail, the remote server when sending
+        local address when receiving mail, the remote address when sending
         mail.
@@
     \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        server when receiving mail, the local server when sending mail.
+        address when receiving mail, the local address when sending mail.
 
@@
         logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this:
-
-\label{logging helo}
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there will not be a queueid
-        associated with the mail for the first recipient, so that log line
-        cannot be associated with the correct mail.  There is guaranteed to
-        be a queueid when the DATA command has been reached, and thus it
-        will be logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring when, but only in the case where there
-        is a single recipient; in that case the recipient address will be
-        logged also, but when there are multiple recipients none are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
+        is quite easy to rectify this, as described in \sectionref{logging
+        helo}.
 
@@
 
-Every log line where the associated rule has a \texttt{postfix\_action}
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
+Every interesting\footnote{All log lines are interesting unless the action
+of the rule which parses them is \texttt{INFO} or \texttt{IGNORE}, as
+described in \sectionref{actions-in-detail}.} log line will have an entry
+in the results table, e.g.\ rejecting an \SMTP{} command, delivering a
+mail, or bouncing a mail (see \sectionref{postfix_action} for typical
+values of \texttt{postfix\_action}).  Each row is associated with a single
 connection, though there may be many results per connection.
@@
 
-    \item [connection\_id] A reference to the row in the connections table
-        this result is associated with.
+    \item [connection\_id] The id of the row in the connections table this
+        result is associated with.
 
-    \item [rule\_id] A reference to the entry in the rules table which
-        matched the log line and created this result.
+    \item [rule\_id] The id of the entry in the rules table which matched
+        the log line and created this result.
 
@@
         real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection.
+        ignored if the result is not a rejection, i.e.\ the action field of
+        the associated rule is not \texttt{REJECTION}.
 
@@
         delivery log lines do not contain an \SMTP{} code: in those cases
-        the code is faked based on the success or failure represented by
-        the log line.
+        the code is specified by the rule, based on the success or failure
+        represented by the log line.
 
@@
         not be different sender addresses in the results for one email.
+        Mails sent over the same connection can be distinguished by their
+        queueid.
 

### 2008-07-10 doc/logparser-limitations.tex
Corrections from my last proofreading.

@@
 
+\label{logging helo}
+
 \begin{enumerate}
@@
         2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5.}  Rectifying this has already been described in
-        \sectionref{logging helo}.
+        2.5. XXX TEST THIS AGAIN.}  Rectifying this is relatively simple:
+        create a restriction which is guaranteed to warn for every accepted
+        mail, as follows:
+
+        \begin{enumerate}
+
+            \item Create \texttt{/etc/postfix/log\_helo.pcre}
+                containing:\newline \tab{}\texttt{/\^/~~~~WARN~Logging~HELO}
+
+            \item Modify \texttt{smtpd\_data\_restrictions} in
+                \texttt{/etc/postfix/main.cf} to contain\newline
+                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
+
+        \end{enumerate}
+
+        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
+        place to log the HELO hostname, there will not be a queueid
+        associated with the mail for the first recipient, so the log line
+        cannot be associated with the correct mail.  There is guaranteed
+        to be a queueid when the DATA command has been reached, and thus it
+        will be logged by any restrictions taking effect in
+        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
+        specifying a HELO-based restriction in
+        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
+        correctly.
+
+        Logging the HELO hostname in this fashion also prevents the
+        complication described in \sectionref{Mail deleted before delivery
+        is attempted} from occurring, but only in the case where there is a
+        single recipient; in that case the recipient address will be logged
+        also, but when there are multiple recipients no addresses are
+        logged.  It is also possible to warn for every recipient,
+        preventing the complication in \sectionref{Mail deleted before
+        delivery is attempted} entirely.
 

### 2008-07-10 doc/logparser-parsing-rules.tex
Corrections from my last proofreading.

@@
 starting with rule characteristics, followed by the problems caused by
-overlapping rules, and techniques to detect and deal with such rules.  An
-example rule and a log line it would match are provided, plus a description
-of how the fields in the rule\footnote{The details of the table containing
-the rules have already been described in \sectionref{rule attributes}.} are
-used when matching a log line and subsequently performing the requested
-action.  The section continues with a discussion of rule efficiency
-concerns, referring to the graphs in \sectionref{graphs}, and finishes with
-an explanation of the algorithm used to generate new \regexes{} from
-unparsed lines.
+overlapping rules, and techniques to detect and deal with such overlaps.
+An example rule and a log line it would match are provided, plus a
+description of how the fields in the rule\footnote{The table containing the
+rules has already been described in \sectionref{rule attributes}.} are used
+when matching a log line and subsequently performing the requested action.
+The section continues with a discussion of rule efficiency concerns, with
+reference to the graphs in \sectionref{graphs}, and finishes with an
+explanation of the algorithm used to generate new \regexes{} from unparsed
+log lines.
 
@@
         there are also rules specific to that program, and those rules have
-        already have been tried and failed on the current log line.}  Any
-        given rule will only be used to parse a subset of the log lines,
-        and any given log line will only be parsed by a subset of the
-        rules.
+        already have been tried and have failed on the current log line.}
+        Any given rule will only be used to parse a subset of the log
+        lines, and any given log line will only have a subset of the rules
+        tried against it.
 
     \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for ordering the rules so
-        that more specific rules can be tried first.
+        that log line, but there is a facility for prioritising the rules
+        so that more specific rules can be tried first.
 
@@
     \item Rule processing time is a linear function of the number of rules.
+        XXX IMPROVE THIS\@.
 
@@
 
-The parser does not try to detect overlapping rules;\footnote{It may be
-possible to parse each rule's \regex{} and determine if any overlap.  The
-author has not attempted to do this: such a project by itself would
-probably qualify for a PhD, and may involve solving the Halting
-Problem~\cite{Wikipedia-halting-problem} and circumventing the
-Church-Turing Thesis~\cite{Wikipedia-church-turing-thesis}.} that
+XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?
+
+The parser does not try to detect overlapping rules; that
 responsibility is left to the author of the rules.  Unintentionally
@@
 algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches when parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
+field in each rule (defaults to zero).  Negative priorities may be useful
+for catchall rules.
 
@@
     \item Compare the results of parsing using sorted, shuffled and
-        reversed rules.\footnote{See \sectionref{rule efficiency} for more
-        details of sorting the rules.}  Parse a number of log files using
-        normal sorting, then dump a textual representation of the rules,
-        connections and results tables.  Repeat with shuffled and reversed
-        rules, starting with a fresh database.  If there are no overlapping
-        rules the tables from each run will be identical; differences
-        indicate overlapping rules.  Which rules overlap can be determined
-        by examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing records
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
+        reversed rules (described in \sectionref{rule efficiency}).  Parse
+        a number of log files using normal sorting, then dump a textual
+        representation of the rules, connections and results tables.
+        Repeat with shuffled and reversed rules, starting with a fresh
+        database.  If there are no overlapping rules the tables from each
+        run will be identical; differences indicate overlapping rules.
+        Which rules overlap can be determined by examining the differences
+        in the tables: each result contains a reference to the rule which
+        created it, if the references differ between runs the two rules
+        referenced in the differing results overlap.  Unfortunately this
+        method cannot prove the absence of overlapping rules; it can detect
+        overlapping rules, but only if there are log lines in the input
+        files which match more than one rule.
 

### 2008-07-10 doc/logparser-background.tex
Corrections from proofreading. More reference commands for consistent formatting.

@@
     \item It is already possible to access databases from the vast majority
-        of programming languages, allowing a developer to access the
-        data gathered using their preferred programming language, rather
-        than being restricted to the language the parser is written in.  It
-        is often possible to write an interface layer allowing code written
-        in one language to be used in another language, but this greatly
+        of programming languages, allowing a developer to access the data
+        gathered using their preferred programming language, rather than
+        being restricted to the language the parser is written in.  It is
+        often possible to write an interface layer allowing code written in
+        one language to be used in another language, but this greatly
         increases the effort required to use the parser.
@@
 \subsection{Simple Mail Transfer Protocol}
+
 \label{SMTP background}
@@
 policy server is given state information\footnote{Sample state information
-is shown in \tableref{Example attributes sent to policy servers}}
-and returns a verdict as described in \sectionref{Mixing and matching
-Postfix restrictions}.  The policy server can perform more complex checks
-than those provided by Postfix: a trivial example is restricting mail from
+is shown in \tableref{Example attributes sent to policy servers}} and
+returns a verdict as described in \sectionref{Mixing and matching Postfix
+restrictions}.  The policy server can perform more complex checks than
+those provided by Postfix: a trivial example is restricting mail from
 addresses associated with the payroll system to sending mail on the third
 Tuesday after pay day only, to help prevent problems from spam or (worse)
-phishing mails with faked sender addresses.\footnote{A
-phishing mail might claim that the payroll system had a disastrous disk
-failure; until the server is replaced and restored all salary payments will
-have to be processed manually, so please reply to this mail with your name,
-address and back account details.}
+phishing mails with faked sender addresses.\footnote{A phishing mail might
+claim that the payroll system had a disastrous disk failure; until the
+server is replaced and restored all salary payments will have to be
+processed manually, so please reply to this mail with your name, address
+and back account details.}
 
@@
         restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in table~\refwithpage{Summary of rejections}.
+        the parser are shown in \tableref{Summary of rejections}.
         Greylisting is obviously worth using, at least at the moment,
@@
 code would require a significant investment of time: there are 375,750
-lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source code.}
-there are advantages to treating Postfix as a black box while developing
-the parser:
+lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source
+code.} there are advantages to treating Postfix as a black box while
+developing the parser:
 
@@
 separation of functionality: rules handle low level details of identifying
-log lines and extracting data from a log line; actions handle
-the higher level details of following the path a mail takes through
-Postfix, assembling the required data before storing it, dealing with
-complications arising, etc; the framework provides services to actions and
-stores data.
+log lines and extracting data from a log line; actions handle the higher
+level details of following the path a mail takes through Postfix,
+assembling the required data before storing it, dealing with complications
+arising, etc; the framework provides services to actions and stores data.
 
@@
 occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes
-must be preserved during the upgrade, and may clash with changes made by
-the developer.\footnote{See \sectionref{adding new actions} for how to add
-new actions}  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare case that the new actions require support from
-other sections of the code will more extensive changes be required.
+changes to the parser's code also complicates upgrades, as the changes must
+be preserved during the upgrade, and may clash with changes made by the
+developer.\footnote{See \sectionref{adding new actions} for how to add new
+actions}  \parsername{} allows the user to add new rules to the database
+without changing the parsing algorithm, unless the new log lines to be
+parsed require functionality not already provided by the algorithm.  If the
+new log lines do require new functionality, new actions can be added to the
+parser without modifying existing actions or other parts of the algorithm;
+only in the rare case that the new actions require support from other
+sections of the code will more extensive changes be required.
 

### 2008-07-10 doc/logparser-conclusion.tex
Corrections from proofreading. More reference commands for consistent formatting.

@@
 uneven distribution exhibited by log lines produced in normal operation
-(see graph~\refwithpage{rule hits graph} for a vivid illustration of this).
+(see \graphref{rule hits graph} for a vivid illustration of this).
 

### 2008-07-10 doc/logparser-coverage.tex
Corrections from proofreading. More reference commands for consistent formatting.

@@
 (usually adding more information), etc.,\ so over time the log lines drift
-and change.  Graph~\refwithpage{rule hits graph} shows the number of hits
+and change.  \Graphref{rule hits graph} shows the number of hits
 for each rule over all \numberOFlogFILES{} log files; it is obvious that a
@@
 \numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining graph~\refwithpage{rule hits graph} it becomes apparent that such
+examining \graphref{rule hits graph} it becomes apparent that such
 a low hit rate is to be expected; the reader should also bear in mind that

### 2008-07-10 doc/logparser-graphs.tex
Corrections from proofreading. More reference commands for consistent formatting.

@@
 
-The Y axis in graph~\refwithpage{execution time vs file size vs number of
+The Y axis in \graphref{execution time vs file size vs number of
 lines graph} represents the following:
@@
 
-Graph~\refwithpage{execution time vs file size vs number lines factor}
+\Graphref{execution time vs file size vs number lines factor}
 shows the ratio of execution time vs file size and number of lines (higher
@@
 are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; graph~\refwithpage{execution time vs file
-size vs number lines factor} and table~\refwithpage{execution time vs file
+where they are noticeably higher; \graphref{execution time vs file
+size vs number lines factor} and \tableref{execution time vs file
 size vs number lines factor table} show that the parser's execution time
@@
 The dips at log files 22 and 62--68 correspond to peaks in log file size in
-graph~\refwithpage{execution time vs file size vs number of lines graph},
-and peaks in graph~\refwithpage{execution time vs file size vs number lines
+\graphref{execution time vs file size vs number of lines graph},
+and peaks in \graphref{execution time vs file size vs number lines
 factor} (where a peak means that more lines are processed per second, i.e.\
@@
 required to parse a log line when there is a mail loop, as shown by the
-peaks in graph~\refwithpage{execution time vs file size vs number lines
+peaks in \graphref{execution time vs file size vs number lines
 factor}.  When the rule ordering is reversed the majority of log lines
@@
 rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in graph~\refwithpage{percentage increase of
+parse a log line, as shown in \graphref{percentage increase of
 reversed over normal}.  The number of rules which need to be consulted when
@@
 The difference between log files with a mail loop and log files without can
-be seen in table~\refwithpage{Execution time increase for different rule
+be seen in \tableref{Execution time increase for different rule
 orderings} showing the increases for the different rule orderings and
@@
 
-Two large dips can be seen in graph~\refwithpage{normal regex vs discarded
+Two large dips can be seen in \graphref{normal regex vs discarded
 regex factor} at log files 22 and 62--68, corresponding to the spikes in
@@
 
-As graph~\refwithpage{rule hits graph} is quite difficult to read it has
+As \graphref{rule hits graph} is quite difficult to read it has
 been separated into three sections: low, middle and high.
@@
 
-It is apparent from the low graph (graph~\refwithpage{hits per rule low})
+It is apparent from the low graph (\graphref{hits per rule low})
 that some rules have few or no hits; those with zero hits are rules which

### 2008-07-10 doc/logparser-introduction.tex
Corrections from proofreading. More reference commands for consistent formatting.

@@
 handling mail for over 700 users, averaging 8500 mails per day
-(graph~\refwithpage{Mails received per day}).  Coverage is described both
+(\graphref{Mails received per day}).  Coverage is described both
 in terms of the fraction of log lines parsed and the fraction of mails and

### 2008-07-10 doc/logparser-parsing-algorithm.tex
Corrections from proofreading. More reference commands for consistent formatting.

@@
 \label{parsing-algorithm}
+
+EMPHASISE THAT THE LOGS CAUSE THE INTERDEPENDENCIES, NOT THE
+ARCHITECTURE\@.
 
@@
 
-There should probably be two different entires in the database resulting
+There should probably be two different entries in the database resulting
 from the above sequence, but currently there will only be one.
@@
 
-Figure~\refwithpage{flow chart image part 1} and figure~\refwithpage{flow
+\Figureref{flow chart image part 1} and \figureref{flow
 chart image part 2} show the paths the data representing a mail can take
@@
 21--22--23--24) in the second half of the flow chart
-(figure~\refwithpage{flow chart image part 2}) are encoded within the
+(\figureref{flow chart image part 2}) are encoded within the
 actions; the transitions between the other stages (16--17, 24--25) in the
@@
 However the transitions between stages in the first half of the flow chart
-(figure~\refwithpage{flow chart image part 1}) are not encoded anywhere in
+(\figureref{flow chart image part 1}) are not encoded anywhere in
 the parser; the transitions are determined by the ordering of log lines in
@@
 The distribution of rules per action is shown in
-graph~\refwithpage{Distribution of rules per action}.
+\graphref{Distribution of rules per action}.
 

### 2008-07-10 doc/logparser-parsing-rules.tex
Corrections from proofreading. More reference commands for consistent formatting.

@@
     \item [program and regex:] If the program in the rule equals the
-        program which logged the log line a match using the \regex{} will
-        be attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules do not match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
+        program which logged the log line, a match using the rule's
+        \regex{} will be attempted against the log line; if the match is
+        successful the action will be executed, if not the next rule will
+        be tried.  If the program-specific rules do not match the log line,
+        the parser will fall back to generic rules; if those rules are
+        unsuccessful a warning will be issued.
 
-    \item [action:] will be executed if the \regex{} matches successfully
-        (see \sectionref{actions-in-detail} for full details).
+    \item [action:] The action to be executed if the \regex{} matches
+        successfully (see \sectionref{actions-in-detail} for full details).
 
@@
 \label{rule efficiency}
+
+XXX ADD TIMING INFORMATION\@.
 
@@
 input size: does it scale linearly, polynomially, or exponentially?
-Graph~\refwithpage{execution time vs file size vs number of lines graph}
+\Graphref{execution time vs file size vs number of lines graph}
 shows the execution time in seconds, file size in MB and tens of thousands
@@
 a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by graph~\refwithpage{execution time vs file
+This impression is borne out by \graphref{execution time vs file
 size vs number lines factor} which plots the ratio of file size vs
 execution time and ratio of number of log lines vs execution time (higher
-is better).\footnote{Table~\refwithpage{execution time vs file size vs
-number lines factor table} shows the ratios for different breakdowns of the
-log files.}  As the reader can see the ratios are quite tightly banded,
-showing that the algorithm scales linearly: the much larger log files
-between points 60 and 70 on the X axis in graph~\refwithpage{execution time
-vs file size vs number of lines graph} actually cause the ratio to
-increase, rather than decrease.  The strange behaviour where larger log
+is better); \tableref{execution time vs file size vs number lines
+factor table} shows the ratios for different breakdowns of the log files.
+As the reader can see the ratios are quite tightly banded, showing that the
+algorithm scales linearly: the much larger log files between points 60 and
+70 on the X axis in \graphref{execution time vs file size vs
+number of lines graph} actually cause the ratio to increase (i.e.\
+improve), rather than decrease.  The strange behaviour where larger log
 files are parsed more efficiently is explained fully in \sectionref{Why are
@@
 covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules, with the top 10\% matching the vast
-majority of the log lines, and the remaining log lines split across the
-other 90\% of the rules (as shown in graph~\refwithpage{rule hits graph}).
-Assuming that the distribution of log lines is reasonably steady over time,
-program efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
+are \numberOFrules{} different rules: the top 10\% match the vast majority
+of log lines, with the remaining log lines split across the other 90\% of
+the rules (as shown in \graphref{rule hits graph}).  Assuming that
+the distribution of log lines is reasonably steady over time, program
+efficiency should benefit from trying more frequently matching rules before
+those which match less frequently.  To test this hypothesis three full test
+runs were performed with different rule orderings:
 
@@
 
-    \item [shuffle] Random ordering --- the rules will be shuffled once
-        before use and will retain that ordering for the entirety of the
-        log file.  Note that the ordering will change every time the parser
-        is executed, so 10 different orderings will be generated for each
-        log file in the test run.  This is intended to represent an
-        unsorted rule set.
+    \item [shuffle] This is intended to represent a randomly ordered rule
+        set.  The rules will be shuffled once before use and will retain
+        that ordering for the entirety of the log file.  Note that the
+        ordering will change every time the parser is executed, so 10
+        different orderings will be generated for each log file in the test
+        run.
 
@@
 and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with table~\refwithpage{Execution
+percentage increase of execution times, with \tableref{Execution
 time increase for different rule orderings} showing the mean increases for
@@
 \regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  Graph~\refwithpage{normal regex vs discard regex} shows execution
+cached.  \Graphref{normal regex vs discard regex} shows execution
 times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; graph~\refwithpage{normal regex
+\regexes{} is obviously far more efficient; \graphref{normal regex
 vs discarded regex factor} shows the percentage execution time increase
@@
 effective optimisation implemented in the parser.
+
+XXX ADD A TABLE\@.
 
@@
 parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial (and also occurs much more
-frequently).  The implementation provides a program to ease the process of
+parse new rejection messages is trivial, and also occurs much more
+frequently.  The implementation provides a program to ease the process of
 creating new rules from unparsed log lines, based on the algorithm
@@
         the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.  The modified algorithm omits
-        this check because the majority of unique or infrequently occurring
-        words will have been substituted with keywords during the first
-        phase, vastly reducing the number of tuples to maintain counters
-        for.
+        memory requirements of the algorithm.
+
+        As time goes on the amount of memory typically available increases
+        and the requirement to reduce memory requirements decreases, hence
+        the modified algorithm omits the hashing step and maintains
+        counters for all tokens. In addition the majority of unique or
+        infrequently occurring words will have been substituted with
+        keywords during the first phase, vastly reducing the number of
+        tuples to maintain counters for.
 
@@
         variable term.  If a variable term appears sufficiently often it
-        will be classified as a fixed term, but that should be noticed by
-        the user when reviewing the new \regexes{}.  Variable terms are
+        will be misclassified as a fixed term, but that should be noticed
+        by the user when reviewing the new \regexes{}.  Variable terms are
         replaced by \texttt{.+}, which means to match zero or more of any
@@
     \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added.  If the frequency of a \regex{} is lower than the threshold
-        supplied by the user the \regex{} is discarded.\footnote{This is a
-        second threshold, independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of input log lines is a good starting point; obviously the
-        threshold depends on the number and type of input log lines.}  The
-        new \regexes{} are printed for the user to add to the database,
-        either as new rules or merged into the \regexes{} of existing
-        rules; the counter for each \regex{} is also printed, giving the
-        user an indication of how many of the input log lines that \regex{}
-        should match.  Discarding \regexes{} will result in some of the
-        input log lines not being matched; this utility should be run again
-        once the unmatched log lines have been isolated by running the
-        parser, including the new \regexes{}, with the same input which
-        produced the original set of unparsed log lines.
+        matching the log line, and a counter maintained for each \regex{}.
+        Contiguous sequences of \texttt{.+} in the newly reassembled
+        \regexes{} are collapsed to a single \texttt{.+}; any resulting
+        duplicate \regexes{} are combined, and their counters added
+        together.  If the \regex{}'s counter is lower than the threshold
+        supplied by the user the \regex{} is discarded; this second
+        threshold is independent of the threshold used to differentiate
+        between fixed and variable words, but once again 1\% of input log
+        lines is a good starting point; obviously the threshold depends on
+        the number and type of input log lines.  The new \regexes{} are
+        displayed for the user to add to the database, either as new rules
+        or merged into the \regexes{} of existing rules; the counter for
+        each \regex{} is also displayed, giving the user an indication of
+        how many of the input log lines that \regex{} should match.
+        Discarding \regexes{} will result in some of the input log lines
+        not being matched; the utility should be run again once the
+        unmatched log lines have been isolated by parsing the original log
+        files once the parser has been augmented with the new rules.
 
@@
 
+These utilities are not intended to create perfect \regexes{}, but they
+greatly reduce the effort required to parse new or different log lines.
+
 \subsection{Summary}
@@
         description of how the fields in the rule are used both in the
-        matching phase and the action that is subsequently executed.
+        matching phase and the action that is subsequently executed.  XXX
+        REWRITE ONCE EXAMPLE HAS CHANGED\@; MERGE WITH NEXT ITEM\@.
 
@@
 
-    \item The topic of rule efficiency was discussed next, covering the
-        effects of caching compiled \regexes{} and optimal ordering of
-        rules, with reference to the graphs in \sectionref{graphs}.
+    \item The topic of rule efficiency was discussed, covering the effects
+        of caching compiled \regexes{} and optimal ordering of rules, with
+        reference to the graphs in \sectionref{graphs}.
 

### 2008-07-11 doc/logparser-acronyms.tex
Corrections from proofreading. More reference commands for consistent formatting.
Corrections from proofreading. Worked out the new layout I want to use.

@@
 \newacronym{MTA}{Mail Transfer Agent}{description={
-    A Mail Transfer Agent sends and/or receives mail via SMTP\backslashat{}.
+    A Mail Transfer Agent sends and/or receives mail via
+    SMTP\backslashat{}.
 }}
@@
 \newacronym{FQDN}{Fully Qualified Domain Name}{description={
-    An FQDN is a hostname plus domain name, e.g.  \newline
+    An FQDN is a hostname plus domain name, e.g.\ \newline
         example.com     is a domain name          \newline
@@
 \newacronym{regex}{Regular Expression}{description={
-    Regular expressions are a concise and powerful method of matching patterns against
-    text that are explained in detail in~\cite{Wikipedia-regex, perlre}.
+    Regular expressions are a concise and powerful method of matching
+    patterns against text that are explained in detail
+    in~\cite{Wikipedia-regex, perlre}.
 }}

### 2008-07-11 doc/logparser-background.tex
Corrections from proofreading. Worked out the new layout I want to use.

@@
 into various component programs to perform the tasks required of an
-\MTA{}\@: receive mail, send mail, local delivery of mail, etc. --- full
+\MTA{}\@: receive mail, send mail, local delivery of mail, etc.\ --- full
 details can be found in~\cite{postfix-overview}.  Each log line contains
@@
         managing data structures, finding the rule which matches each log
-        line, invoking the correct action, etc.  The framework provides the
-        structure which actions and rules plug into.  The framework is
+        line, invoking the correct action, etc\@.  The framework provides
+        the structure which actions and rules plug into.  The framework is
         described in detail in \sectionref{framework}.  XXX EXTEND
@@
 % file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g. rejections) lack a queueid.
+% a queueid but majority of log entries (e.g.\ rejections) lack a queueid.
 % Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
@@
 % difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc. pointing
-% at the main table - this also will not scale well.  There does not appear
-% to be any attempt to deal with the complications I have encountered:
-% their parsing is not detailed enough to encounter them.  It does not run
-% properly; does not create any output; throws up errors.
+% re-implemented foreign keys with tables keyed by ip address etc.\
+% pointing at the main table - this also will not scale well.  There does
+% not appear to be any attempt to deal with the complications I have
+% encountered: their parsing is not detailed enough to encounter them.  It
+% does not run properly; does not create any output; throws up errors.
 

### 2008-07-11 doc/logparser-parsing-algorithm.tex
Corrections from proofreading. Worked out the new layout I want to use.

@@
 
-    \item If the mail is rejected, log all data and finish.
+    \item If the mail is rejected, save all data and finish.
 
     \item Follow the progress of the accepted mail until it is either
-        delivered, bounced or deleted, then log all data, and finish.
+        delivered, bounced, or deleted, then save all data, and finish.
 
@@
 This must be done every time a mail is accepted, as it is impossible to
-tell in advance which connections will accept multiple mails.  Happily,
-once the mail has been accepted log entries will not be correlated by
-\pid{} for that mail any more (its queueid will be used instead), so there
-is not any ambiguity about which mail a given log line belongs
-to.\footnote{Unfortunately this statement is not completely accurate: see
-\sectionref{timeouts-during-data-phase} for details.  However in general
-there is not any ambiguity about which data structure should be used for a
-given log line.}  The original connection will be discarded unsaved when
-the client disconnects if it does not have any data worth saving, i.e.\ no
-rejections.  One unsolved difficulty is distinguishing between different
-groups of rejections, e.g.\ when dealing with the following sequence:
+tell in advance which connections will accept multiple mails.  Once the
+mail has been accepted its log entries will not be correlated by \pid{} any
+more (its queueid will be used instead), so there is no longer any
+ambiguity about which mail a given log line belongs to (except when
+timeouts occur during the data phase
+\sectionref{timeouts-during-data-phase}).  If the original connection has
+any interesting data (e.g.\ rejections) it will be saved to the database
+when the client disconnects.  One unsolved difficulty is distinguishing
+between different groups of rejections, e.g.\ when dealing with the
+following sequence:
 
@@
 
-There should probably be two different entries in the database resulting
-from the above sequence, but currently there will only be one.
+There should ideally be two separate entries in the database resulting from
+the above sequence, but currently there will only be one.
 
@@
 cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how busy \daemon{qmgr} is.\footnote{Postfix is quite
-paranoid about mail delivery, an excellent characteristic for an \MTA{} to
-possess, so it will not log that the child has been created until it is
-absolutely certain that the mail has been written to disk.}
-
-Because of this the parser cannot complain when it encounters a log line
-from \daemon{qmgr} for a previously unseen mail; it must flag the mail as
-coming from an unknown origin, and subsequently clear the flag if and when
-the origin of the mail becomes clear.  Obviously the parser could omit
-checking of where mails originate from, but requiring an explicit source
-helps to expose bugs in the parser; such checks helped to identify the
-complications described in \sectionref{discarding cleanup lines} and
-\sectionref{pickup logging after cleanup}.
+order also depends on how soon \daemon{qmgr} processes the new mail.
+
+Because of the uncertain order the parser cannot complain when it
+encounters a log line from \daemon{qmgr} for a previously unseen mail;
+instead it must flag the mail as coming from an unknown origin, and
+subsequently clear the flag if and when the origin of the mail becomes
+clear.  Obviously the parser could omit checking where mails originate
+from, but requiring an explicit source helps to expose bugs in the parser;
+such checks helped to identify the complications described in
+\sectionref{discarding cleanup lines} and \sectionref{pickup logging after
+cleanup}.
 
@@
         in the intervening hours: \DNS{} timeouts, timeouts reading data
-        from the client, etc.  The active connection for that
+        from the client, etc\@.  The active connection for that
         \daemon{smtpd} must be cleaned up, entered in the database, and

### 2008-07-11 doc/logparser.tex
Corrections from Mads.
Corrections from Mads and chktex.
Corrections from proofreading. More reference commands for consistent formatting.
Corrections from proofreading. Worked out the new layout I want to use.

@@
 
-% Sort numbers where there are multiple citations.  Doesn't appear to have
+% Sort numbers where there are multiple citations.  does not appear to have
 % any effect (probably clashes with varioref or hyperref), though it does
@@
 
-% Produce nicer references, e.g. "section 4.3 on the next page"
+% Produce nicer references like "section 4.3 on the next page"
 % Must be loaded before hyperref.
@@
 }
+% A reference with a space between the label and the reference.
+\newcommand{\refwithlabel}[2]{%
+    #1~\vref{#2}%
+}
+% table references, for consistent formatting.
+\newcommand{\tableref}[1]{%
+    \refwithlabel{table}{#1}%
+}
+\newcommand{\Tableref}[1]{%
+    \refwithlabel{Table}{#1}%
+}
+% graph references
+\newcommand{\graphref}[1]{%
+    \refwithlabel{graph}{#1}%
+}
+\newcommand{\Graphref}[1]{%
+    \refwithlabel{Graph}{#1}%
+}
+% figure references
+\newcommand{\figureref}[1]{%
+    \refwithlabel{figure}{#1}%
+}
+\newcommand{\Figureref}[1]{%
+    \refwithlabel{Figure}{#1}%
+}
 
@@
 % (without horrible red borders) and include bookmarks.  The title and
-% author don't work - I think either gnuplot or graphviz clobbers it.
+% author do not work - I think either gnuplot or graphviz clobbers it.
 % hyperfootnotes need to be disabled to avoid breaking footmisc, but they
@@
 
-% This is ridiculous, but I can't put @ in glossary entries, so . . .
+% This is ridiculous, but I cannot put @ in glossary entries, so . . .
 \newcommand{\at}[0]{%
     @%
+}
+% This is ridiculous, but I cannot put @ in glossary entries, so . . .
+\newcommand{\backslashat}[0]{%
+    \@%
 }
@@
 
-XXX EXPLAIN THE X-AXIS --- MADS THOUGHT IT WAS UNCLEAR\@.
+XXX EXPLAIN THE X-AXIS ON GRAPHS --- MADS THOUGHT IT WAS UNCLEAR\@.
+
+XXX ``.'' AFTER ETC\@; SHOULD I HAVE A BACKSLASH AFTER THE ``.''?
+
+\newpage
+
+XXX NEW LAYOUT\@:
+
+\begin{verbatim}
+Introduction
+Background
+State of the Art Review
+Parser Architecture
+Postfix Parser Implementation
+    Database Schema
+    Rules
+    Actions
+    Complications
+Results - BETTER TITLE?
+    Coverage
+    Efficiency
+    Limitations and Improvements
+Conclusion
+References
+Glossary
+Acronyms
+\end{verbatim}
 

### 2008-07-11 doc/logparser-graphs.tex
More proofreading changes.

@@
 it to the remote address; this is generally not a problem except that the
-remote address in this case is the user's address, creating an infinite
-loop.  To prevent this happening Postfix examines the Delivered-To header
-in the mail, and if the mail has already been delivered to the current
-address it is bounced back to the sender with the error message
-\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
-well, but unfortunately in this case the user noticed they had not received
-any mail in a while and opted to send a test mail to themselves, causing a
-loop not caught by Postfix as described below.
+remote address in this case is the address the mail was originally sent to,
+creating an infinite loop.  To prevent this happening Postfix examines the
+Delivered-To header in the mail, and if the mail has already been delivered
+to the current address it is bounced back to the sender with the error
+message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
+works well, but unfortunately in this case the user noticed they had not
+received any mail in a while and opted to send a test mail to themselves,
+causing a loop not caught by Postfix:
 
@@
         contains the address it is currently delivering to, and therefore
-        sends a bounce notification with sender address \textit{$<>$\/} to
+        sends a bounce notification with sender address \texttt{<>} to
         the original sender: username@domain.  Log lines are added by
@@
 required to parse a log line when there is a mail loop, as shown by the
-peaks in \graphref{execution time vs file size vs number lines
-factor}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed with very few rules, whereas
-without a mail loop the majority of log lines require a large number of
-rules; this leads to a noticeable drop in the average time required to
-parse a log line, as shown in \graphref{percentage increase of
-reversed over normal}.  The number of rules which need to be consulted when
-the ordering is shuffled varies between the optimum and nadir, and the
-performance varies proportionally.
+peaks in \graphref{execution time vs file size vs number lines factor}.
+When the rule ordering is reversed the majority of log lines generated by a
+mail loop will be parsed with very few rules, whereas without a mail loop
+the majority of log lines require a large number of rules; this leads to a
+noticeable drop in the average time required to parse a log line, as shown
+in \graphref{percentage increase of reversed over normal}.  The number of
+rules which need to be consulted when the ordering is shuffled varies
+between the optimal and the pessimal, and the performance varies
+proportionally.
 
@@
 The distribution of log lines across rules when there is a mail loop is
-much different and the average number of rules consulted per log line is
-much lower; this results in far fewer \regex{} compilations per line than
-when there is not a mail loop, and a correspondingly decreased execution
-time.  The increases in execution time when not caching \regexes{} for
+much different to the norm and the average number of rules consulted per
+log line is much lower; this results in far fewer \regex{} compilations per
+line than when there is not a mail loop, and a correspondingly decreased
+execution time.  The unexpected outcome is that caching \regexes{} is
+proportionally less important when the log file contents were created by a
+mail loop.  The increases in execution time when not caching \regexes{} for
 different combinations of log files are summarised in the table below:
@@
 to by the third group of graphs, where the difference between caching and
-discarding is staggering.  The penultimate section contains a breakdown of
+not caching is staggering.  The penultimate section contains a breakdown of
 the rule hits accumulated during a single test run of the parser.

### 2008-07-21 doc/logparser-background.tex
Merged 'Other Postfix log parsers reviewed' and 'Previous research in this area' to create a 'State of the Art Review', which is now section 3.
Removed 'Throughput monitor' review.

Improved introduction and summary in 'State of the Art Review'; miscellaneous
related changes.

Moved 'Graphs' section closer to where it will eventually be, the 'Results'
section.

@@
 
-\subsection{Other Postfix log parsers}
-
-Ten other parsers have been reviewed in \sectionref{other-parsers} as part
-of the background research for this project.  None of the reviewed parsers
-perform the type of advanced parsing and log correlation described here;
-all are intended to perform a specific parsing and reporting task, rather
-than be a generic parser, extracting data and leaving generation of reports
-from the data to other programs.  Some parsers save the data extracted to a
-data store, but the many discard all data once they have finished running,
-making historical analysis impossible.  The other parsers reviewed all
-produce a report of varying complexity and detail, whereas the program
-described here does not attempt to produce a report at all; that
-responsibility is deferred to a separate program, to be developed later.
-The parsing algorithm and program described here are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
-
-XXX EXTEND THIS\@.
-
-\subsection{Previous research in this area}
-
-\label{prior art}
-
-There only appears to be one prior paper published about parsing Postfix
-log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients) and adding new
-fields.  There is no facility to document the fields or the separator,
-unlike \SQL{} databases where every database includes a schema naming the
-fields and the type of data they store (integer, text, timestamp, etc.).
-
-Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
-unlike \SQL{} databases which store arbitrary tuples; in \LMA{}'s main
-table the key is an integer referred to by secondary tables, and the value
-used is a \CSV{} line containing all of the data for that row.  The
-secondary by-sender, by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats: it does not provide the functionality of a
-basic \SQL{} database, not can it be used with standard Unix text
-processing tools.
-
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated or linked in any way), and presumably the same approach is
-taken when there are multiple destination servers.
-
-The schema used with the MySQL database is undocumented, but at least it is
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the schema into the database and
-provide commands for displaying it.  Berkeley DB does not embed a schema,
-as there is neither requirement nor benefit; it only provides \textbf{(key,
-value)} pairs, so any additional structuring of the data is imposed by the
-application, thus it behoves the application to document this structure.
-
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; log files from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.  Once the preprocessing has been performed on the
-\numberOFlogFILES{} test log files \LMA{} parses the log files without
-complaint, though it does produce 32 entries in its output file for every
-rejection in the input log file; it also missed some 40\% of delivered
-mail.  Once these glaring deficiencies were discovered the author did not
-spend more time checking the results.
-
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
-
-\begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
-\end{verbatim}
-
-Notes about the command above:
-
-\begin{enumerate}
-
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
-        DAILY-REJECT\@.
-
-    \item Some records have an \SMTP{} code of \textit{0}, though this is
-        not a valid \SMTP{} code; presumably this is due to \LMA{} not
-        parsing a line properly.  \LMA{} considers this code to be a
-        rejection, so the command above must, too.
-
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
-
-\end{enumerate}
-
-There are a number of differences in the output from DAILY-REJECT and the
-\texttt{awk} command:
-
-\begin{itemize}
-
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
-
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
-
-\end{itemize}
-
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
-
-% Literature review notes:
-%
-% Hard-coded parsing, requiring code changes to add more.  Attempts to
-% correlate log lines, saves data to database for data mining purposes.
-% Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g.\ rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc.\
-% pointing at the main table - this also will not scale well.  There does
-% not appear to be any attempt to deal with the complications I have
-% encountered: their parsing is not detailed enough to encounter them.  It
-% does not run properly; does not create any output; throws up errors.
-
 \subsection{Summary}

### 2008-07-21 doc/logparser-introduction.tex
Merged 'Other Postfix log parsers reviewed' and 'Previous research in this area' to create a 'State of the Art Review', which is now section 3.
Removed 'Throughput monitor' review.

Improved introduction and summary in 'State of the Art Review'; miscellaneous
related changes.

Moved 'Graphs' section closer to where it will eventually be, the 'Results'
section.

@@
 
+Section~\ref{state of the art review} reviews both the previously published
+research in this area and other available Postfix log parsers, discussing
+why they were deemed unsuitable for the task, including why they could not
+be improved or expanded upon.
+
 This algorithm requires a database for storing both the rules used when
@@
 
+Section~\ref{graphs} contains graphs illustrating the topics previously
+discussed in \sectionref{rule efficiency} (rule efficiency), and explains
+the irregularities observed in these graphs.
+
 Section~\ref{limitations-improvements} lists the limitations of the
@@
 
-Appendix~\ref{other-parsers} discusses other parsers and why they were
-deemed unsuitable for the task, including why they could not be improved or
-expanded upon.
-
 The bibliography contains references to the resources used in developing
@@
 
-Appendix~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
 Appendix~\ref{Glossary} provides a glossary of terms used in the document.

### 2008-07-21 doc/logparser-other-parsers-review.tex
Merged 'Other Postfix log parsers reviewed' and 'Previous research in this area' to create a 'State of the Art Review', which is now section 3.
Removed 'Throughput monitor' review.

Improved introduction and summary in 'State of the Art Review'; miscellaneous
related changes.

Moved 'Graphs' section closer to where it will eventually be, the 'Results'
section.

@@
-\section{Other Postfix log parsers reviewed}
+\section{State of the Art Review}
 
 \label{other-parsers}
+
+\label{state of the art review}
+
+Ten other parsers, starting with the only previously published research in
+this area, have been reviewed as part of the background research for this
+project.  None of the parsers reviewed perform the type of advanced parsing
+and log correlation required for this project; all bar one are intended to
+perform a specific parsing and reporting task, rather than to be a generic
+parser, extracting data and leaving generation of reports from the data to
+other programs.  Some parsers save the data extracted to a data store, but
+the majority discard all data once they have finished running, making
+historical analysis impossible.  The other parsers reviewed all produce a
+report of varying complexity and detail, whereas the program described here
+does not attempt to produce a report at all; that responsibility is
+deferred to a separate program, to be developed later.  The parsing
+algorithm and program developed for this project are designed to enable
+much more detailed log analysis by providing a stable platform for
+subsequent programs to develop upon.
+
+XXX EXTEND THIS\@.
+
+\subsection{Log Mail Analyser}
+
+\label{prior art}
+
+There only appears to be one prior paper published about parsing Postfix
+log files: \textit{Log Mail Analyzer: Architecture and Practical
+Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
+different from this parser: it attempts to present correlated log data in a
+form suitable for a systems administrator to search using the myriad of
+standard Unix text processing utilities already available, producing both a
+\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
+support both \CSV{} and Berkeley DB appears to have been a limiting factor:
+\CSV{} is a very simple format where a record is stored in a single line,
+with fields separated by a comma or other punctuation symbol.  Problems
+with \CSV{} files include the difficulty in escaping separators, providing
+multiple values for a field (e.g.\ multiple recipients) and adding new
+fields.  There is no facility to document the fields or the separator,
+unlike \SQL{} databases where every database includes a schema naming the
+fields and the type of data they store (integer, text, timestamp, etc.).
+
+Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
+unlike \SQL{} databases which store arbitrary tuples; in \LMA{}'s main
+table the key is an integer referred to by secondary tables, and the value
+used is a \CSV{} line containing all of the data for that row.  The
+secondary by-sender, by-recipient, by-date and by-\IP{} tables use the
+sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
+referring to the main table is the value.  In effect this re-implements
+\SQL{} foreign keys, but without the functionality offered by even the most
+basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
+requires custom code to search on some combination of the above, though the
+authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
+DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
+of the three output formats: it does not provide the functionality of a
+basic \SQL{} database, not can it be used with standard Unix text
+processing tools.
+
+The data stored is limited to time and date, hostname and \IP{} address of
+client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
+size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
+codes or remote servers\footnote{A single mail may be sent to multiple
+remote servers if it was addressed to recipients in different domains, or
+Postfix needs to try multiple servers for one or more recipients.} is not
+explained in the paper; experimental observation shows that multiple
+records are added when there are multiple recipients (sadly the records are
+not associated or linked in any way), and presumably the same approach is
+taken when there are multiple destination servers.
+
+The schema used with the MySQL database is undocumented, but at least it is
+possible to discover the schema with an \SQL{} database, unlike with
+Berkeley DB\@; all \SQL{} databases embed the schema into the database and
+provide commands for displaying it.  Berkeley DB does not embed a schema,
+as there is neither requirement nor benefit; it only provides \textbf{(key,
+value)} pairs, so any additional structuring of the data is imposed by the
+application, thus it behoves the application to document this structure.
+
+Parsing in \LMA{} requires major changes to the code to parse new log lines
+or extract additional data.  It does not appear to deal with any of the
+complications discussed in \sectionref{complications} (initial
+complications) and \sectionref{additional complications} (additional
+complications).  It does not differentiate between different types of
+rejections, so it is not suitable for the purposes of this project; the
+data about which restriction caused the rejection is discarded, whereas an
+important goal of this project is to retain that data to aid optimisation
+and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
+files generated on Solaris hosts because the fields added automatically
+differ from those added on Linux hosts; log files from Solaris hosts (and
+possibly other operating systems) thus require preprocessing before parsing
+by \LMA{}.  Once the preprocessing has been performed on the
+\numberOFlogFILES{} test log files \LMA{} parses the log files without
+complaint, though it does produce 32 entries in its output file for every
+rejection in the input log file; it also missed some 40\% of delivered
+mail.  Once these glaring deficiencies were discovered the author did not
+spend more time checking the results.
+
+\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
+and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
+client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
+sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
+DAILY-REJECT initially failed with an error message from the Perl
+interpreter; after making some corrections to the code it worked, and
+produced the \CSV{} lines for the specified day where the \SMTP{} code
+signifies a rejection.  All of these reports are trivially simple to
+produce from the \CSV{} file using the standard Unix tool \texttt{awk},
+e.g.\ the most complicated, DAILY-REJECT, is merely:
+
+\begin{verbatim}
+awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
+    { print $0; print " "; }' lma_output.txt
+\end{verbatim}
+
+Notes about the command above:
+
+\begin{enumerate}
+
+    \item It outputs an unnecessary line containing only a single space
+        after each matching line, for compatibility with the output of
+        DAILY-REJECT\@.
+
+    \item Some records have an \SMTP{} code of \textit{0}, though this is
+        not a valid \SMTP{} code; presumably this is due to \LMA{} not
+        parsing a line properly.  \LMA{} considers this code to be a
+        rejection, so the command above must, too.
+
+    \item In a tiny proportion of records the \SMTP{} code is
+        \textit{deferred\/} (again, not a valid code, and most likely due
+        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
+        so the equivalent \texttt{awk} command must do so, too.
+
+\end{enumerate}
+
+There are a number of differences in the output from DAILY-REJECT and the
+\texttt{awk} command:
+
+\begin{itemize}
+
+    \item \LMA{} does not escape or replace the separators used in the
+        \CSV{} records when they are present in sender or recipient
+        addresses, leading to some mangled records; these records are not
+        present in the output from \texttt{awk}.
+
+    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
+        produces only one output line, whereas the \texttt{awk} command
+        produces all matching lines.
+
+\end{itemize}
+
+In summary, \LMA{} appears to be a proof of concept, written to demonstrate
+the point of their paper (that having this information in an accessible
+fashion is useful to systems administrators), rather than a program
+designed to be extensible and useful in a production environment.
+
+% Literature review notes:
+%
+% Hard-coded parsing, requiring code changes to add more.  Attempts to
+% correlate log lines, saves data to database for data mining purposes.
+% Hard to extend/expand/understand.  Appears to only save: date and hour,
+% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
+% receiver and e-mail status (sent, rejected).  Undocumented schema.
+% Design decision to use \CSV{} as an intermediate format between the log
+% file and the database seems to have been restrictive.  Appears to require
+% a queueid but majority of log entries (e.g.\ rejections) lack a queueid.
+% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
+% when generating reports/data mining would be preferable.  Supporting
+% Berkeley DB is probably limiting the software - an example is the
+% difficulty in searching a pipe-delimited string, so they have
+% re-implemented foreign keys with tables keyed by ip address etc.\
+% pointing at the main table - this also will not scale well.  There does
+% not appear to be any attempt to deal with the complications I have
+% encountered: their parsing is not detailed enough to encounter them.  It
+% does not run properly; does not create any output; throws up errors.
 
@@
 what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers, indeed it was quite
-difficult to find ten parsers to review for this project, and the
+that many previously developed Postfix log parsers --- indeed it was quite
+difficult to find ten parsers to review for this project --- and the
 functionality offered ranges from quite basic to much more mature,
@@
 
-\subsection{Throughput monitor}
-
-This utility tracks the number of events which occurred over a particular
-time and warns if the frequency of events passes a certain threshold.  It
-is designed to provide real time alerts when dictionary attacks, mail loops
-or similar problems occur. It does not attempt to extract or save data,
-correlate log lines, or any of the more advanced tasks described in this
-document, because it is not designed to do so.  The user must construct the
-\regexes{} to match significant input lines before it will parse any log
-files.  This program was not reviewed in further detail because its aims
-are so far removed from the aims of this project.
-
-\url{http://home.uninet.ee/~ragnar/throughput_monitor/} \newline (Last
-checked 2008/05/01.)
-
 \subsection{Anteater}
@@
 
-\subsection{Log Mail Analyser}
-
-Please see the previous in-depth discussion of Log Mail Analyser in
-\sectionref{prior art}.
-
 \subsection{Summary}
@@
 \begin{table}[ht]
-\begin{tabular}{l|l|l|l|l|l}
+\begin{tabular}{llllll}
     Parser         & Parsed test & Data            & Custom        & Documentation  & Source \\
                    & log files?  & store?          & reports?      & quality?       & code?  \\
-    \hline         &             &                 &               &                &        \\
+    \tableline{}
+    \LMA{}         & No          & Yes             & No            & Poor           & Yes    \\ 
     Pflogsumm      & Yes         & No              & Partial\dag{} & Good           & Yes    \\
@@
                    &             &                 & \& reports    & but poor       &        \\
-    Isoqlog        & No          & Yes             & No            & No             & Yes    \\
+    Isoqlog        & No          & Yes             & No            & Poor           & Yes    \\
     AWStats        & Partially   & Yes             & Partial\dag{} & Good           & Yes    \\
-    Throughput     & N/A         & No              & No            & Brief but good & Yes    \\
-    \tab{}monitor  &             &                 &               &                &        \\
-    Anteater       & No          & No              & No            & No             & Yes    \\
+    Anteater       & No          & No              & No            & Poor           & Yes    \\
     YAALA          & No          & Yes\ddag{}      & Searches      & Poor           & Yes    \\
-    Lire           & Yes         & Yes             & Yes           & Yes            & Yes    \\
-    Logrep         & No          & Yes             & No            & Barely         & Yes    \\
-    \LMA{}         & No          & Yes             & No            & No             & Yes    \\ 
-    \parsername{}  & Yes         & Yes\dag{}\dag{} & No            & Yes            & Yes    \\
+    Lire           & Yes         & Yes             & Yes           & Reasonable     & Yes    \\
+    Logrep         & No          & Yes             & No            & Poor           & Yes    \\
+    \parsername{}  & Yes         & Yes\nialpha{}   & \nibeta{}     &                & Yes    \\
+    \tableline{}
 \end{tabular}
@@
 
-\dag{}\dag{}\parsername{} is the only parser with documentation for its
-data store.
+\nialpha{} \parsername{} is the only parser with documentation for its data
+store.
+
+\nibeta{} \parsername{} defers report generation to subsequent programs,
+but all the necessary data and documentation to produce reports is
+provided.
 

### 2008-07-22 doc/logparser-architecture.tex
Moved the architecture stuff into it's own section.

--- 2008-07-21 doc/logparser-background.tex
@@
-\section{Background}
-
-\label{background}
-
-This section provides background information helpful in understanding the
-remainder of the document.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-document, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this document.  Other projects which attempt to
-parse Postfix log files are summarised (full details are available in
-\sectionref{other-parsers}), finishing with a review of previously
-published research in this area.
-
-\subsection{Motivation}
-
-\label{motivation}
-
-This document and the program it describes are part of a larger project to
-optimise a mail server's Postfix restrictions, generate statistics and
-graphs, and provide a platform on which new restrictions can be trialled
-and evaluated to determine if they are beneficial in the fight against
-spam.  The program parses Postfix log files and populates a database with
-the data gleaned from those log files, providing a consistent and simple
-view of the log files which future tools can utilise.  The gathered data
-can then be used to optimise current anti-spam measures, provide a baseline
-to test new anti-spam measures against, or to produce statistics showing
-how effective those measures are.
-
-A snippet of \SQL{} provides a short example of the optimisation possible
-using data from the database: determining which Postfix restrictions reject
-the highest number of mails:
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
-\end{verbatim}
-
-XXX GIVE SAMPLE OUTPUT FOR THE SNIPPET ABOVE\@.
-
-Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed, and the percentage of total rejections each restriction represents.
-
-\begin{verbatim}
-SELECT name, description, restriction_name, hits,
-        (hits * 100.0 /
-            (SELECT SUM(hits)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-        AND hits < 100
-    ORDER BY hits ASC;
-\end{verbatim}
-
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques; statistics that are far less feasible to assess
-directly from log files without prior pre-processing into a database in the
-fashion proposed, implemented and tested herein.
-
-\subsection{Database as Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix log files.  Although the interface is a database schema, it is in
-effect quite similar to any other \API{} provided by shared code: it
-insulates both user and provider of the \API{} from changes in the
-implementation of the \API{}\@.  The algorithm implemented by the parser
-can be improved; support can be added for earlier or later releases of
-Postfix; bugs can be fixed or limitations removed from the parser; these
-changes will not cause the user to be negatively impacted.  Statistics
-and/or graphs can be generated from the database; new restrictions can be
-tested and the results inspected; trends in the fight against spam can
-emerge from historical data saved in the database; the parser remains the
-same as the usage adapts.  Using a database simplifies writing programs
-which need to interact with the data in several ways:
-
-\begin{enumerate}
-
-    \item It is already possible to access databases from the vast majority
-        of programming languages, allowing a developer to access the data
-        gathered using their preferred programming language, rather than
-        being restricted to the language the parser is written in.  It is
-        often possible to write an interface layer allowing code written in
-        one language to be used in another language, but this greatly
-        increases the effort required to use the parser.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of varying complexity and
-        sophistication, which can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-            \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
-
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis, e.g.\ allowing the finance department
-                to produce invoices, the helpdesk to run limited queries as
-                part of dealing with support calls, and the administrators
-                to have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-        \end{itemize}
-
-    \item \SQL{} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, as the log files need only be parsed once rather than
-each time the data is used; indeed the results may be used by someone with
-no access to the original log files.
-
-\subsection{Simple Mail Transfer Protocol}
-
-\label{SMTP background}
-
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity of the protocol many virus and/or spam sending programs fail
-to implement it properly, so requiring strict adherence to the protocol
-specification is beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  it is not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: <alice@example.com>
-250 2.1.0 Ok
-RCPT TO: <bob@example.com>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-Message headers and body sent here.
-.
-250 2.0.0 Ok: queued as D7AFA38BA
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-An example deviation from the protocol:
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-HELO client.example.com
-250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
-501 5.1.7 Bad sender address syntax
-RCPT TO: Bob in Sales/Marketing bob@example.com
-503 5.5.1 Error: need MAIL command
-DATA
-503 5.5.1 Error: need RCPT command
-Message headers and body sent here.
-.
-502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
-in \texttt{<>}.
-
-A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02},
-the definitive references are~\cite{RFC821, RFC2821}.
-
-\subsection{Postfix}
-
-\label{postfix background}
-
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of stricture and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the log files generated.  While it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  For the majority of mails the journey is simple and brief, but
-the remaining minority can be quite complex (see \sectionref{additional
-complications} for details).
-
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
-into various component programs to perform the tasks required of an
-\MTA{}\@: receive mail, send mail, local delivery of mail, etc.\ --- full
-details can be found in~\cite{postfix-overview}.  Each log line contains
-the name of the Postfix component which produced it, and this information
-is used when determining which rules should be used to parse each log line
-(see \sectionref{rule characteristics} for details).  This design has
-positive security implications also: those components which interact with
-other hosts are not privileged,\footnote{Privilege in this case means the
-power to perform actions that are limited to the administrator, and not
-available to ordinary users.} so bugs in those components will not give an
-attacker extra privileges; those components which are privileged do not
-interact with other hosts, making it much more difficult for an attacker to
-exploit any bugs which may exist in those components.
-
-\subsubsection{Mixing and matching Postfix restrictions}
-
-\label{Mixing and matching Postfix restrictions}
-
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}; the following is a brief overview
-only.
-
-Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
-command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-do not deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging
-rejections.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible as described
-in~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.} The
-meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address,
-etc.\footnote{E.g.\ the administrator may require that all clients on the
-local network have valid \DNS{} entries, to prevent people sending mail
-from unknown machines.}  The administrator can define new restrictions as a
-list of existing restrictions, allowing arbitrarily long and complex
-sequences of lookups, restrictions and exceptions.  Postfix tries to
-protect the administrator in as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must be cognisant of the ramifications of
-their configuration changes.  XXX UNIX DOESN'T STOP YOU DOING STUPID THINGS
-BECAUSE THAT WOULD ALSO STOP YOU DOING CLEVER THINGS\@.
-
-Postfix uses simple lookup tables as the deciding factor when evaluating
-some restrictions, e.g.\ in the example restriction below\newline
-\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
-
-\begin{description}
-
-    \item [check\_client\_access] The name of the restriction to evaluate.
-
-    \item [cidr] The type of the lookup table.
-
-    \item [/etc/postfix/client\_access] The file containing the lookup
-        table.
-
-\end{description}
-
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client dnsbl.example.com} checks whether the \IP{}
-address of the client is present in the \DNSBL{}
-\texttt{dnsbl.example.com}, rejecting the command if the client is listed.
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
-
-\subsubsection{Policy servers}
-
-\label{policy servers}
-
-A \textit{policy server\/}~\cite{policy-servers} is an external program
-consulted by Postfix to determine the fate of an \SMTP{} command.  The
-policy server is given state information\footnote{Sample state information
-is shown in \tableref{Example attributes sent to policy servers}} and
-returns a verdict as described in \sectionref{Mixing and matching Postfix
-restrictions}.  The policy server can perform more complex checks than
-those provided by Postfix: a trivial example is restricting mail from
-addresses associated with the payroll system to sending mail on the third
-Tuesday after pay day only, to help prevent problems from spam or (worse)
-phishing mails with faked sender addresses.\footnote{A phishing mail might
-claim that the payroll system had a disastrous disk failure; until the
-server is replaced and restored all salary payments will have to be
-processed manually, so please reply to this mail with your name, address
-and back account details.}
-
-Some widely deployed policy servers:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}, e.g.\ forwarding mail from one company or
-        university to another when a user changes jobs.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in \tableref{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the log files were obtained from; an administrator
-        is free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is higher than
-        a threshold the mail is rejected.  Postfix's restrictions are
-        binary, and the administrator must manually whitelist clients if
-        they are to bypass a restriction; using a threshold which requires
-        hitting several restrictions frees the administrator from
-        whitelisting clients which fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
-
-\begin{table}[ht]
-
-    \caption{Example attributes sent to policy servers}\label{Example
-    attributes sent to policy servers}
-
-    \begin{tabular}[]{ll}
-
-        request                 & smtpd\_access\_policy     \\
-        protocol\_state         & RCPT                      \\
-        protocol\_name          & SMTP                      \\
-        helo\_name              & some.domain.tld           \\
-        queue\_id               & 8045F2AB23                \\
-        sender                  & foo@bar.tld               \\
-        recipient               & bar@foo.tld               \\
-        recipient\_count        & 0                         \\
-        client\_address         & 1.2.3.4                   \\
-        client\_name            & another.domain.tld        \\
-        reverse\_client\_name   & another.domain.tld        \\
-        instance                & 123.456.7                 \\
-
-    \end{tabular}
-\end{table}
-
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The log files are whole and complete: nothing has been removed,
-        either deliberately or accidentally (e.g.\ log rotation gone awry,
-        file system filling up, logging system unable to cope with the
-        volume of log messages).  On a well run system it is extremely
-        unlikely that any of these problems will arise, though it is of
-        course possible, particularly when suffering a deluge of spam or a
-        mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time: there are 375,750
-lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source
-code.} there are advantages to treating Postfix as a black box while
-developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world log files rather than
-        the idealised log files someone would naturally envisage reading
-        the source code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
+\section{Parser Architecture}
 
@@
 
-\subsection{Conventions used in the document}
+\subsection{Rule characteristics}
 
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this document; in general the word used was chosen based
-on the context it appears in.
+\label{rule characteristics}
 
-\subsection{Summary}
-
-This section has provided background information on several topics relevant
-to the remainder of the document.  It started with the motivation behind
-the project, continuing with explanations of:
+Rule have certain characteristics which may help in understanding the
+parser:
 
@@
 
-    \item Using a database as an \API{}.
+    \item Rules are annotated with the name of a Postfix program, and will
+        only be used when parsing log lines produced by that
+        program.\footnote{There are also generic rules which are used when
+        parsing log lines produced by any Postfix program, but only if
+        there are also rules specific to that program, and those rules have
+        already have been tried and have failed on the current log line.}
+        Any given rule will only be used to parse a subset of the log
+        lines, and any given log line will only have a subset of the rules
+        tried against it.
 
-    \item \SMTP{}.
+    \item The first matching rule wins: no further rules are tried against
+        that log line, but there is a facility for prioritising the rules
+        so that more specific rules can be tried first.
 
-    \item Postfix restrictions and policy servers.
+    \item Rules are completely self-contained and can be understood in
+        isolation, without reference to any other rules.
 
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of the parser's novel design.
-
-    \item Conventions used in this document.
-
-    \item An in-depth comparison of this parser with \LMA{}, the parser
-        described in previously published research.
-
-    \item A review of the literature previously published in this area.
+    \item Rule processing time is a linear function of the number of rules.
+        XXX IMPROVE THIS\@.
 
 \end{itemize}
+
+\label{comparison against context-free grammars}
+
+In context-free grammar terms the parser rules could be described as:
+
+$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
+\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
+
+\subsection{Overlapping rules}
+
+\label{overlapping rules}
+
+XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?
+
+XXX DFA COMPARISON\@.
+
+The parser does not try to detect overlapping rules; that
+responsibility is left to the author of the rules.  Unintentionally
+overlapping rules lead to inconsistent parsing and data extraction because
+the order in which rules are tried against each line may change between log
+files, and the first matching rule wins.  Overlapping rules are frequently
+a requirement, allowing a more specific rule to match some log lines and a
+more general rule to match the majority, e.g.\ separating \SMTP{} delivery
+to specific sites from \SMTP{} delivery to the rest of the world.  The
+algorithm provides a facility for ordering overlapping rules: the priority
+field in each rule (defaults to zero).  Negative priorities may be useful
+for catchall rules.
+
+Detecting overlapping rules is difficult, but the following approaches may
+be helpful:
+
+\begin{itemize}
+
+    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
+        similar to: \textbf{select regex from rules order by regex;}
+
+    \item Compare the results of parsing using sorted, shuffled and
+        reversed rules (described in \sectionref{rule efficiency}).  Parse
+        a number of log files using normal sorting, then dump a textual
+        representation of the rules, connections and results tables.
+        Repeat with shuffled and reversed rules, starting with a fresh
+        database.  If there are no overlapping rules the tables from each
+        run will be identical; differences indicate overlapping rules.
+        Which rules overlap can be determined by examining the differences
+        in the tables: each result contains a reference to the rule which
+        created it, if the references differ between runs the two rules
+        referenced in the differing results overlap.  Unfortunately this
+        method cannot prove the absence of overlapping rules; it can detect
+        overlapping rules, but only if there are log lines in the input
+        files which match more than one rule.
+
+\end{itemize}
+
+\subsection{Framework}
+
+\label{framework}
+
+The intermingling of log entries from different mails immediately rules out
+the possibility of handling each mail in isolation; the parser must be
+capable of handling multiple mails in parallel, each potentially at a
+different stage in its journey, without any interference between mails ---
+except in the minority of cases where intra-mail interference is required.
+The best way to implement this is to maintain state information for every
+unfinished mail and manipulate the appropriate mail correctly for each log
+line encountered.
+
+This functionality is provided by the framework, which both drives parsing
+overall and provides services to the actions it invokes.  The framework is
+responsible for loading rules from the database and sanity checking them,
+reading log files, matching each rule against the current log line,
+invoking the correct action, maintaining state tables, loading and saving
+state, displaying a progress indicator, and miscellaneous other tasks.
+Actions use services provided by the framework, including storing and
+retrieving data in the state tables, extracting and saving data captured by
+rules, storage of global data, debugging functions, preparing a mail for
+entry in the database, and entering mails in the database.
+
+There is a similarity between this design and the event-driven programming
+paradigm commonly used in GUI programs, where one part of the program
+responds to events (mouse clicks in a GUI program, log lines being matched
+in the parser) and invokes the correct action.
+
+The framework's high level overview could be expressed as (indentation
+denoting flow of control):
+
+\begin{verbatim}
+for each line in the input files:
+    for each rule defined by the user:
+        if this rule matches the input line:
+            perform the action specified by the rule
+            skip the remaining rules
+            process the next input line
+    warn the user that the input line was not parsed
+\end{verbatim}

### 2008-07-22 doc/logparser-background.tex
Moved the architecture stuff into it's own section.

@@
 assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this document.  Other projects which attempt to
-parse Postfix log files are summarised (full details are available in
-\sectionref{other-parsers}), finishing with a review of previously
-published research in this area.
+are the conventions used in this document.
 
@@
 
-\subsection{Parser design}
-
-\label{parser design}
-
-It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that log files produced by Postfix are not fixed; they vary
-widely from host to host, depending on the set of restrictions chosen by
-the administrator.  With this in mind, one of the parser's design aims was
-to make adding new rules as easy as possible, to enable administrators to
-properly parse their log files.  To enable this the parser is divided into
-three parts:
-
-\begin{description}
-
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
-        method of associating log lines with actions, and are described in
-        detail in \sectionref{rules}.
-
-    \item [actions] Actions are invoked to deal with a log line once it has
-        been identified by the rules: actions modify data structures,
-        handle complications, and cause data to be saved to the database.
-        The actions perform the work of reconstructing the journey a mail
-        takes through Postfix.  Full details of actions can be found in
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}
-        XXX IMPROVE THIS SENTENCE\@.
-
-    \item [framework] The framework is responsible for loading rules,
-        managing data structures, finding the rule which matches each log
-        line, invoking the correct action, etc\@.  The framework provides
-        the structure which actions and rules plug into.  The framework is
-        described in detail in \sectionref{framework}.  XXX EXTEND
-        FRAMEWORK SECTION IF NECESSARY\@.
-
-\end{description}
-
-It may help to think of the rules and actions as components which plug into
-the framework.
-
-\label{why separate rules, actions and framework?}
-
-XXX MERGE THE NEXT TWO PARAGRAPHS\@.
-
-Decoupling the parsing rules from the associated actions and framework
-allows new rules to be written and tested without requiring modifications
-to the parser source code (significantly lowering the barrier to entry for
-new or casual users who need to parse new log lines), and greatly
-simplifies framework, actions and rules.  Decoupling also creates a clear
-separation of functionality: rules handle low level details of identifying
-log lines and extracting data from a log line; actions handle the higher
-level details of following the path a mail takes through Postfix,
-assembling the required data before storing it, dealing with complications
-arising, etc; the framework provides services to actions and stores data.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and does not need
-to deal with the complications which arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
-
-XXX CLARIFY THE SENTENCE STARTING WITH ``Although this may see'' IN THE
-NEXT PARAGRAPH\@.
-
-Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
-changes to the parser's code also complicates upgrades, as the changes must
-be preserved during the upgrade, and may clash with changes made by the
-developer.\footnote{See \sectionref{adding new actions} for how to add new
-actions}  \parsername{} allows the user to add new rules to the database
-without changing the parsing algorithm, unless the new log lines to be
-parsed require functionality not already provided by the algorithm.  If the
-new log lines do require new functionality, new actions can be added to the
-parser without modifying existing actions or other parts of the algorithm;
-only in the rare case that the new actions require support from other
-sections of the code will more extensive changes be required.
-
-XXX CHANGE THIS\@; THE PARSER DESIGN IS NOVEL, DO NOT STRESS THE SIMILARITY
-TOO MUCH\@.
-
-There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-
-% Do Not Reformat!
-
-\begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
-    Networks            & Algorithm         & Determines the sequence 
-                                              of transitions            \\
-                        &                   & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
-    Transitions         & Actions           & Save data and impose
-                                              conditions the            \\
-                        &                   & input must meet to be
-                                              considered valid.         \\
-    Abbreviations       & Rules             & Responsible for 
-                                              classifying input.        \\
-\end{tabular}
-
 \subsection{Conventions used in the document}

### 2008-07-22 doc/logparser-parsing-algorithm.tex
Moved the architecture stuff into it's own section.

@@
 \end{enumerate}
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
 
@@
 for this flow of data to be specified.}
-
-\subsection{Framework}
-
-\label{framework}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
 

### 2008-07-22 doc/logparser-parsing-rules.tex
Moved the architecture stuff into it's own section.

@@
 rules and actions have been separated in the parser's design.
-
-\subsection{Rule characteristics}
-
-\label{rule characteristics}
-
-Rule have certain characteristics which may help in understanding the
-parser:
-
-\begin{itemize}
-
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and have failed on the current log line.}
-        Any given rule will only be used to parse a subset of the log
-        lines, and any given log line will only have a subset of the rules
-        tried against it.
-
-    \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for prioritising the rules
-        so that more specific rules can be tried first.
-
-    \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
-
-    \item Rule processing time is a linear function of the number of rules.
-        XXX IMPROVE THIS\@.
-
-\end{itemize}
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\subsection{Overlapping rules}
-
-\label{overlapping rules}
-
-XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?
-
-The parser does not try to detect overlapping rules; that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Negative priorities may be useful
-for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
-        similar to: \textbf{select regex from rules order by regex;}
-
-    \item Compare the results of parsing using sorted, shuffled and
-        reversed rules (described in \sectionref{rule efficiency}).  Parse
-        a number of log files using normal sorting, then dump a textual
-        representation of the rules, connections and results tables.
-        Repeat with shuffled and reversed rules, starting with a fresh
-        database.  If there are no overlapping rules the tables from each
-        run will be identical; differences indicate overlapping rules.
-        Which rules overlap can be determined by examining the differences
-        in the tables: each result contains a reference to the rule which
-        created it, if the references differ between runs the two rules
-        referenced in the differing results overlap.  Unfortunately this
-        method cannot prove the absence of overlapping rules; it can detect
-        overlapping rules, but only if there are log lines in the input
-        files which match more than one rule.
-
-\end{itemize}
 

### 2008-07-22 doc/logparser-background.tex
Collected all the implementation bits into one section.

@@
 
-\subsection{Assumptions}
-
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
-
-\begin{itemize}
-
-    \item The log files are whole and complete: nothing has been removed,
-        either deliberately or accidentally (e.g.\ log rotation gone awry,
-        file system filling up, logging system unable to cope with the
-        volume of log messages).  On a well run system it is extremely
-        unlikely that any of these problems will arise, though it is of
-        course possible, particularly when suffering a deluge of spam or a
-        mail loop.
-
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
-        logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time: there are 375,750
-lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source
-code.} there are advantages to treating Postfix as a black box while
-developing the parser:
-
-\begin{itemize}
-
-    \item The parser is developed using real world log files rather than
-        the idealised log files someone would naturally envisage reading
-        the source code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
-
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
-
-\end{itemize}
-
 \subsection{Conventions used in the document}

### 2008-07-22 doc/logparser-implementation.tex
Collected all the implementation bits into one section.

--- 2008-07-22 doc/logparser-parsing-algorithm.tex
@@
-\section{Parsing Algorithm}
+\section{Parser Implementation}
+
+\subsection{Assumptions}
+
+The algorithm described and the program implementing it make a small number
+of (hopefully safe and reasonable) assumptions:
+
+\begin{itemize}
+
+    \item The log files are whole and complete: nothing has been removed,
+        either deliberately or accidentally (e.g.\ log rotation gone awry,
+        file system filling up, logging system unable to cope with the
+        volume of log messages).  On a well run system it is extremely
+        unlikely that any of these problems will arise, though it is of
+        course possible, particularly when suffering a deluge of spam or a
+        mail loop.
+
+    \item Postfix logs sufficient information to make it possible to
+        accurately reconstruct the actions it has taken.  There are a
+        number of heuristics used when parsing; see
+        \sectionref{identifying-bounce-notifications},
+        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
+        logging after cleanup} for details.
+
+    \item The Postfix queue has not been tampered with, causing unexplained
+        appearance or disappearance of mail.  This may happen if the
+        administrator deletes mail from the queue without using
+        \daemon{postsuper}, or if there is filesystem corruption.
+
+\end{itemize}
+
+In some ways this task is similar to reverse engineering or replicating a
+black box program based solely on its inputs and outputs.  Although the
+source code is available,\footnote{Reading and understanding the source
+code would require a significant investment of time: there are 375,750
+lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source
+code.} there are advantages to treating Postfix as a black box while
+developing the parser:
+
+\begin{itemize}
+
+    \item The parser is developed using real world log files rather than
+        the idealised log files someone would naturally envisage reading
+        the source code.
+
+    \item The source code cannot accurately communicate the variety of
+        orderings in which log lines are written to the log file, as
+        process scheduling independently interferes with logging and other
+        processing.
+
+    \item The parser acts as a second source of information, with the
+        information gathered from empirical evidence.  An interesting
+        project would be to compare the empirical knowledge inherent in the
+        parsing algorithm with the documentation and source code of
+        Postfix.
+
+\end{itemize}
+
+\subsection{Database Schema}
+\label{database schema}
+
+The database is an integral part of the parser presented here: it stores
+the rules and the data gleaned by applying those rules to Postfix log
+files.  Understanding the database schema is important in understanding the
+actions of the parser, and essential to developing further applications
+which utilise the data gathered.
+
+The database schema can be conceptually divided in two: the rules, which
+are used to parse log files, and the data saved from the parsing of log
+files.  Rules have the fields required to parse the log lines, extract data
+to be saved, and the action to be executed; they also have several fields
+which aid the user in understanding the meaning of the log lines parsed by
+each rule.  The rules are described in detail in \sectionref{rules} but the
+fields are covered in \sectionref{rule attributes}.
+
+The data saved from parsing the log files is also divided into two tables
+as described below: connections and results.  The connections table
+contains a row for every mail accepted and every connection where there was
+a rejection; the individual fields will be described in
+\sectionref{connections table}.  There will be at least one row in the
+results table for each row in the connections table; the fields will be
+covered in detail in \sectionref{results table}.
+
+An important but easily overlooked benefit of storing the rules in the
+database is the connection between rules and results: if more information
+is required when examining a result, the rule which produced the result is
+available for inspection --- each result references the rule which created
+it.  There is no ambiguity about which rule resulted in a particular
+result, eliminating one potential source of confusion.
+
+\subsubsection{Rules table}
+
+\label{rule attributes}
+
+Rules are discussed in detail in \sectionref{rules}, but the structure of
+the rules table is covered here.  Rules are created by the user, not the
+parser, and will not be modified by the parser (except for the hits and
+hits\_total fields).  Rules parse the individual log lines, extracting data
+to be saved in the connections and results tables, and specifying the
+action to take for that log line.
+
+The following attributes are defined for each rule:
+
+\begin{description}
+
+    \item [id] A unique identifier for each rule which other tables can use
+        to refer to a specific rule.
+
+    \item [name] A short name for the rule.
+
+    \item [description] Something must have occurred to cause Postfix to
+        log each line (e.g.\ a remote client connecting causes a connection
+        line to be logged).  This field describes the event causing the log
+        lines this rule matches.
+
+    \item [restriction\_name] The restriction which caused the mail to be
+        rejected.  Only applicable to rules which have an action of
+        \texttt{REJECTION}, other rules should have an empty string.
+
+    \item [postfix\_action] This is the action Postfix must have taken to
+        generate this log line.  This field is mostly ignored, but two
+        values (IGNORED and INFO) have special meaning, as described below
+        in the list of typical values.\label{postfix_action}
+
+        \begin{description}
+
+            \item [ACCEPTED] Postfix has accepted a mail, and will
+                subsequently attempt to deliver it.
+
+            \item [BOUNCED] The mail has bounced, due to a mail loop,
+                delivery failure, or five day timeout.
+
+            \item [DELETED] The mail was deleted from the queue by an
+                administrator.
+
+            \item [DISCARDED] Postfix discarded the mail it was in the
+                process of accepting, because it was either larger than the
+                limit set by the administrator, or the client timed out or
+                disconnected.
+
+            \item [EXPIRED] The mail has been in the queue for five
+                days\footnote{As with much of Postfix's behaviour, this is
+                the default value but can be changed by the administrator
+                if they choose.} without successful delivery.  A bounce
+                mail will be generated and sent to the sender address.
+
+            \item [INFO] Represents an unspecified intermediate action that
+                the parser is not interested in per se, but which does log
+                useful information, supplementing other log lines.
+
+            \item [IGNORED] An action which is not only uninteresting in
+                itself, but which also provides no useful data.
+
+            \item [POSTFIX\_RELOAD] The administrator has instructed
+                Postfix to start or stop, and all existing \daemon{smtpd}
+                processes will be terminated.  This does not negatively
+                impact on the log files or mail queued by Postfix for
+                delivery.
+
+            \item [PROCESSING] \daemon{cleanup} is processing a mail
+                (see~\cite{postfix-cleanup} for details).
+
+            \item [REJECTED] Postfix rejected a command from the remote
+                client, causing at least one recipient to be rejected.
+
+            \item [SENT] Postfix has successfully sent a mail.
+
+        \end{description}
+
+        Uninteresting log lines are parsed so that any lines the parser
+        is not capable of handling become immediately obvious errors.
+
+    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
+        log lines the rule applies to.  This avoids needlessly trying rules
+        which will not match the log line, or worse, might match
+        unintentionally.  Rules whose program is \texttt{*} will be tried
+        against any log lines which are not parsed by program specific
+        rules.
+
+    \item [regex] The \regex{} to match the log line against.  The \regex{}
+        will first have several keywords expanded: this simplifies reading
+        and writing rules; avoids needless repetition of complex \regex{}
+        components; allows the components to be corrected and/or improved
+        in one location; and makes each \regex{} largely self-documenting.
+
+        For efficiency the keywords are expanded and every rule's \regex{}
+        is compiled before attempting to parse the log file --- otherwise
+        each \regex{} would be recompiled each time it was used, resulting
+        in a large, data dependent slowdown.  Rule efficiency concerns are
+        discussed in \sectionref{rule efficiency}, with the impact of
+        caching compiled \regexes{} covered in \sectionref{Caching each
+        regex}.
+
+    \item [result\_cols, connection\_cols] Specifies how the fields in the
+        log line will be extracted.  The format is:
+        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
+        \newline i.e.\ semi-colon or comma separated assignment statements,
+        with the variable name on the left and the matching capture from
+        the \regex{} on the right hand side.  Data specified by
+        \texttt{result\_cols} will be entered in the \texttt{results}
+        table, while \texttt{connection\_cols} populates the
+        \texttt{connection} table.  The list of acceptable variable names
+        is:
+
+        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
+        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
+        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
+        \newline \tab{} size,} and \texttt{data}
+
+        Additionally \texttt{child} and \texttt{pid} are accepted and used
+        respectively by the \texttt{TRACK}, \texttt{BOUNCE} and
+        \texttt{SMTPD\_DIED} actions, but do not cause data to be entered
+        in the database.
+
+    \item [result\_data, connection\_data] Sometimes rules need to supply a
+        piece of data which is not present in the log line: e.g.\ setting
+        \texttt{smtp\_code} when mail is accepted.  The format and allowed
+        variables are the same as for \texttt{result\_cols} and
+        \texttt{connection\_cols}, except that arbitrary
+        data\footnote{Commas and semi-colons cannot be escaped and thus
+        cannot be used.  This feature is intended for use with small
+        amounts of data rather than large amounts in any one rule, so
+        dealing with escape sequences was deemed unnecessary.} is permitted
+        on the right hand side of the assignment.
+
+    \item [action] The action that will be invoked when this rule matches a
+        log line; a full list of actions and the parameters they are
+        invoked with can be found in \sectionref{actions-in-detail}.
+
+    \item [queueid] Specifies the capturing field in the \regex{}
+        containing the queueid, or zero if the log line does not contain a
+        queueid.  Many log lines will not contain a queueid, e.g.\
+        rejections logged before a mail has been accepted (a queueid will
+        not have been allocated), or log lines which are not tied to one
+        particular mail.
+
+    \item [hits] This counter is maintained for every rule and incremented
+        each time the rule successfully matches.  At the start of each run
+        the program sorts the rules in descending order of hits, and at the
+        end of the run updates every rule's hits field in the database.
+        Assuming that the distribution of log lines is reasonably
+        consistent between log files, rules matching more commonly
+        occurring log lines will be tried before rules matching less
+        commonly occurring log lines, reducing the program's execution
+        time.  Rule ordering for efficiency is discussed in
+        \sectionref{rule ordering for efficiency}.
+
+    \item [hits\_total] The total number of hits for this rule over all
+        runs of the parser.
+
+    \item [priority] This is the user-configurable companion to hits: when
+        the list of rules is sorted, priority overrides hits.  This allows
+        more specific rules to take precedence over more general rules
+        (described in \sectionref{overlapping rules}).
+
+    \item [cluster\_group] A reference to the \texttt{cluster\_group}
+        table.  That table is used by the Decision Tree algorithm described
+        in a separate document.
+
+\end{description}
+
+\subsubsection{Connections table}
+
+\label{connections table}
+
+Every accepted mail and every connection where there was a rejection will
+have a single entry in the connections table containing the following
+fields:
+
+\begin{description}
+
+    \item [id] This field uniquely identifies the row.
+
+    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
+        local address when receiving mail, the remote address when sending
+        mail.
+
+    \item [server\_hostname] The hostname of the server, it will be
+        \texttt{unknown} if the \IP{} address could not be resolved to a
+        hostname via \DNS{}\@.
+
+    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
+        address when receiving mail, the local address when sending mail.
+
+    \item [client\_hostname] The hostname of the client, it will be
+        \texttt{unknown} if the \IP{} address could not be resolved to a
+        hostname via \DNS{}\@.
+
+    \item [helo] The hostname used in the HELO command.  The HELO hostname
+        occasionally changes during a connection, presumably because spam
+        or virus senders think it is a good idea.  By default Postfix only
+        logs the HELO hostname when it rejects an \SMTP{} command, but it
+        is quite easy to rectify this, as described in \sectionref{logging
+        helo}.
+
+    \item [queueid] The queueid of the mail if the connection represents an
+        accepted mail, or \texttt{NOQUEUE} otherwise.
+
+    \item [start] The timestamp of the first log line, in seconds since the
+        epoch (explained in the glossary, \sectionref{Glossary}).
+
+    \item [end] The timestamp of the last log line, in seconds since the
+        epoch.
+
+\end{description}
+
+\subsubsection{Results table}
+
+\label{results table}
+
+Every interesting\footnote{All log lines are interesting unless the action
+of the rule which parses them is \texttt{INFO} or \texttt{IGNORE}, as
+described in \sectionref{actions-in-detail}.} log line will have an entry
+in the results table, e.g.\ rejecting an \SMTP{} command, delivering a
+mail, or bouncing a mail (see \sectionref{postfix_action} for typical
+values of \texttt{postfix\_action}).  Each row is associated with a single
+connection, though there may be many results per connection.
+
+\begin{description}
+
+    \item [connection\_id] The id of the row in the connections table this
+        result is associated with.
+
+    \item [rule\_id] The id of the entry in the rules table which matched
+        the log line and created this result.
+
+    \item [id] A unique identifier for this result.
+
+    \item [warning] Postfix can be configured to log a warning instead of
+        enforcing a restriction that would reject an \SMTP{} command --- a
+        facility that is quite useful for testing new restrictions.  This
+        field will be 1 if the log line parsed was a warning rather than a
+        real rejection, or 0 for a real rejection.  This field should be
+        ignored if the result is not a rejection, i.e.\ the action field of
+        the associated rule is not \texttt{REJECTION}.
+
+    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
+        general an \SMTP{} code is only present for a rejection or final
+        delivery; results initially missing an \SMTP{} code will duplicate
+        the \SMTP{} code of other results in the connection.  Some final
+        delivery log lines do not contain an \SMTP{} code: in those cases
+        the code is specified by the rule, based on the success or failure
+        represented by the log line.
+
+    \item [sender] The sender's email address.  Because multiple mails may
+        be delivered during one connection, there may be different sender
+        addresses in the results for one connection; however there should
+        not be different sender addresses in the results for one email.
+        Mails sent over the same connection can be distinguished by their
+        queueid.
+
+    \item [recipient] The recipient address; there may be multiple
+        recipient addresses per mail, but only one per result.
+
+    \item [size] The size of the mail; it will only be present for
+        delivered mails.
+
+    \item [message\_id] The message-id of the accepted mail, or
+        \texttt{NULL} if no mail was accepted.
+
+    \item [data] A field available for anything not covered by other
+        fields, e.g.\ the rejection message from an \DNSBL{}\@.
+
+    \item [timestamp] The time the log line was logged, in seconds since
+        the epoch.
+
+\end{description}
+
+\subsubsection{Summary}
+
+The table containing the rules used by the parser and both tables
+containing the data extracted from the Postfix log files were described,
+with the purpose of each field discussed in detail.  A clear,
+comprehensible schema is essential when using the extracted data; it is
+more important when using the data than when storing it, because storing
+the data is a write-once operation, whereas utilising the data requires
+frequent searching, sorting and manipulation of the data to produce
+customised reports and/or statistics.
+
+\subsection{Rules}
+
+\subsubsection{Example rule}
+
+\label{example rule}
+
+This example rule matches the message logged by Postfix when it rejects
+mail from a sender address because the appropriate \DNS{} entries are
+missing, i.e.\ mail could not be delivered to the sender's address (for
+full details see~\cite{reject-unknown-sender-domain}).
+
+The example rule below would match the following log line:
+
+\begin{verbatim}
+NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
+  <foo@example.com>: Sender address rejected: Domain not found;
+  from=<foo@example.com> to=<info@example.net>
+  proto=SMTP helo=<smtp.example.com>
+\end{verbatim}
+
+% do not reformat this!
+\begin{tabular}[]{ll}
+
+\textbf{Field}      & \textbf{Value}                                    \\
+name                & Unknown sender domain                             \\
+description         & We do not accept mail from unknown domains        \\
+restriction\_name   & reject\_unknown\_sender\_domain                   \\
+postfix\_action     & REJECTED                                          \\
+program             & \daemon{smtpd}                                    \\
+regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
+                    & \verb!Sender address rejected: Domain not found;! \\
+                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
+                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
+result\_cols        & recipient = 6; sender = 5                         \\
+connection\_cols    & helo = 7                                          \\
+result\_data        &                                                   \\
+connection\_data    &                                                   \\
+action              & REJECTION                                         \\
+queueid             & 1                                                 \\
+hits                & 0                                                 \\
+hits\_total         & 0                                                 \\
+priority            & 0                                                 \\
+cluster\_group      & 400                                               \\
+
+\end{tabular}
+
+\vspace{1em}
+
+Additional data will be captured automatically when the \regex{} contains 
+\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
+connection\_cols start at 5.  The various fields are used as follows;
+
+\begin{description}
+
+    \item [name, description, restriction\_name and postfix\_action:] are
+        not \newline used by the algorithm, they serve to document the rule
+        for the user's benefit.
+
+    \item [program and regex:] If the program in the rule equals the
+        program which logged the log line, a match using the rule's
+        \regex{} will be attempted against the log line; if the match is
+        successful the action will be executed, if not the next rule will
+        be tried.  If the program-specific rules do not match the log line,
+        the parser will fall back to generic rules; if those rules are
+        unsuccessful a warning will be issued.
+
+    \item [action:] The action to be executed if the \regex{} matches
+        successfully (see \sectionref{actions-in-detail} for full details).
+
+    \item [result\_cols, connection\_cols, result\_data and
+        connection\_data:] are \newline used by the action to extract and
+        save data matched by the \regex{}.
+
+    \item [queueid:] The index of the capture in the \regex{} which
+        supplies the queueid, or zero if the log line does not contain a
+        queueid.  This allows the correct mail can be found by queueid and
+        actions performed on it.
+
+    \item [hits, hits\_total and priority:] hits and priority are used in
+        ordering the rules (see \sectionref{rule ordering for efficiency});
+        hits is set to the number of successful matches at the end of the
+        parsing run; hits\_total is the sum of hits over every parsing run,
+        but is otherwise unused by the algorithm.
+
+    \item [cluster\_group] The cluster\_group attribute is used by the
+        Decision Tree algorithm described in a separate document; the
+        parser does not use it in any way.
+
+\end{description}
+
+\subsubsection{Creating new rules}
+
+\label{creating new rules}
+
+The log files produced by Postfix differ from installation to installation,
+because administrators have the freedom to choose the subset of available
+restrictions which suits their needs, including using different \DNSBL{}
+services, policy servers, or custom rejection messages.  To facilitate easy
+parsing of new log lines, the parser's design separates parsing rules from
+parsing actions: adding new actions is difficult, but adding new rules to
+parse new rejection messages is trivial, and also occurs much more
+frequently.  The implementation provides a program to ease the process of
+creating new rules from unparsed log lines, based on the algorithm
+developed by Risto Vaarandi~\cite{risto-vaarandi} for his
+\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
+be outlined as part of the general explanation below.
+
+The core of the algorithm is quite simple: log lines are generally created
+by substituting variable words into a fixed pattern, and analysis of the
+frequency with which each word occurs can be used to determine whether the
+word is variable or part of the fixed pattern.  This classification can be
+used to group similar log lines and generate a \regex{} to match each group
+of log lines.
+
+There are 4 steps in the algorithm:
+
+\begin{description}
+
+    \item [Pre-process the file]  The new algorithm leverages the knowledge
+        gained while writing rules and performs a large number of
+        substitutions on the log input lines, replacing commonly occurring
+        variable terms (e.g.\ email addresses, \IP{} addresses, the
+        standard start of rejection messages, etc.) with \regex{} keywords
+        which the parser will expand when it loads the rule (see the
+        \regex{} entry in \sectionref{rule attributes}).  The purpose of
+        this phase is to utilise existing knowledge to create more accurate
+        \regexes{}.  The new log lines are written to a temporary file,
+        which all subsequent stages use instead of the original input file.
+
+        In the original algorithm the purpose of the preprocessing stage
+        was to reduce the memory consumption of the program.  In the first
+        pass it generated a hash from a small range of values for each word
+        of each log line, incrementing a counter for each hash.  The
+        counters will be used in the next pass to filter out words: if the
+        word's hash does not have a high frequency, the word itself cannot
+        have a high frequency, and there is no need to maintain a counter
+        for it, reducing the number of counters and thus the program's
+        memory consumption.
+
+    \item [Calculate word frequencies]  The position of words within a log
+        line is important: a common word does not indicate similarity
+        between log lines unless it occupies the same position within both
+        log lines.\footnote{If a variable term within a line contains
+        spaces, it will appear to the algorithm as two words rather than
+        one.  This will alter the position of subsequent words, so a word
+        occurring in different positions in two log lines \textit{may\/}
+        indicate similarity, but the algorithm does not attempt to deal
+        with this possibility.}  The algorithm maintains a counter for each
+        \textit{(word, word's position within the log line)\/} tuple,
+        incrementing it each time that word occurs in that position.
+
+        The original algorithm only maintains counters for words whose hash
+        result from the previous phase has a high frequency; this reduces
+        the number of counters maintained by the algorithm, reducing the
+        memory requirements of the algorithm.
+
+        As time goes on the amount of memory typically available increases
+        and the requirement to reduce memory requirements decreases, hence
+        the modified algorithm omits the hashing step and maintains
+        counters for all tokens. In addition the majority of unique or
+        infrequently occurring words will have been substituted with
+        keywords during the first phase, vastly reducing the number of
+        tuples to maintain counters for.
+
+    \item [Classify words based on their frequency]  The frequency of each
+        \textit{(word, word's position within the log line)\/} tuple is
+        checked: if its frequency is greater than the threshold supplied by
+        the user (1\% of all log lines is generally a good starting point)
+        it is classified as a fixed word, otherwise it is classified as a
+        variable term.  If a variable term appears sufficiently often it
+        will be misclassified as a fixed term, but that should be noticed
+        by the user when reviewing the new \regexes{}.  Variable terms are
+        replaced by \texttt{.+}, which means to match zero or more of any
+        character.
+
+    \item [Build regexes]  The words are reassembled to produce a \regex{}
+        matching the log line, and a counter maintained for each \regex{}.
+        Contiguous sequences of \texttt{.+} in the newly reassembled
+        \regexes{} are collapsed to a single \texttt{.+}; any resulting
+        duplicate \regexes{} are combined, and their counters added
+        together.  If the \regex{}'s counter is lower than the threshold
+        supplied by the user the \regex{} is discarded; this second
+        threshold is independent of the threshold used to differentiate
+        between fixed and variable words, but once again 1\% of input log
+        lines is a good starting point; obviously the threshold depends on
+        the number and type of input log lines.  The new \regexes{} are
+        displayed for the user to add to the database, either as new rules
+        or merged into the \regexes{} of existing rules; the counter for
+        each \regex{} is also displayed, giving the user an indication of
+        how many of the input log lines that \regex{} should match.
+        Discarding \regexes{} will result in some of the input log lines
+        not being matched; the utility should be run again once the
+        unmatched log lines have been isolated by parsing the original log
+        files once the parser has been augmented with the new rules.
+
+\end{description}
+
+A second utility is also provided which reads a list of new \regexes{} and
+the input given to the first utility.  It tries to match each input log
+line against each \regex{}, counting the number of log lines which match
+each \regex{}, warning the user if an input log line is matched by more
+than one \regex{}, and additionally warning if an input log line is not
+matched by any \regex{}.  It displays a summary of how many input log lines
+each \regex{} matched, comparing it to the expected number of matches; this
+provides the user with an easy method of checking if the \regexes{}
+produced by the first utility are correctly matching the input log lines
+they are based upon.  A future version of this utility will also group
+input log lines by \regex{}, so the user can tweak the \regexes{} if
+required.
+
+These utilities are not intended to create perfect \regexes{}, but they
+greatly reduce the effort required to parse new or different log lines.
+
+\subsection{Parsing Algorithm}
 
@@
 
-\subsection{Introduction}
+\subsubsection{Introduction}
 
@@
 
-\subsection{A high level overview}
+\subsubsection{A high level overview}
 
@@
 
-\subsection{Complications encountered}
+\subsubsection{Complications encountered}
 
@@
 
-\subsection{Flow chart}
+\subsubsection{Flow chart}
 
@@
 
-\subsection{Actions in detail}
+\subsubsection{Actions in detail}
 
@@
 
-\subsection{Adding new actions}
+\subsubsection{Adding new actions}
 
@@
 
-\subsection{Additional complications}
+\subsubsection{Additional complications}
 
@@
 
-\subsection{Summary}
+\subsubsection{Summary}
 
@@
 problem.
+
+\subsection{Limitations and possible improvements}
+
+\label{limitations-improvements}
+
+Every piece of software suffers from some limitations and there is almost
+always room for improvement.
+
+\subsubsection{Limitations}
+
+\label{logging helo}
+
+\begin{enumerate}
+
+    \item Each new Postfix release requires new rules to be written to cope
+        with the new log lines.  Similarly using a new \DNSBL{}, new policy
+        server or new administrator defined rejection messages require new
+        rules.
+
+    \item It appears that the hostname used in the HELO command is not
+        logged if the mail is accepted.\footnote{Tested with Postfix
+        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
+        2.5. XXX TEST THIS AGAIN.}  Rectifying this is relatively simple:
+        create a restriction which is guaranteed to warn for every accepted
+        mail, as follows:
+
+        \begin{enumerate}
+
+            \item Create \texttt{/etc/postfix/log\_helo.pcre}
+                containing:\newline \tab{}\texttt{/\^/~~~~WARN~Logging~HELO}
+
+            \item Modify \texttt{smtpd\_data\_restrictions} in
+                \texttt{/etc/postfix/main.cf} to contain\newline
+                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
+
+        \end{enumerate}
+
+        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
+        place to log the HELO hostname, there will not be a queueid
+        associated with the mail for the first recipient, so the log line
+        cannot be associated with the correct mail.  There is guaranteed
+        to be a queueid when the DATA command has been reached, and thus it
+        will be logged by any restrictions taking effect in
+        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
+        specifying a HELO-based restriction in
+        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
+        correctly.
+
+        Logging the HELO hostname in this fashion also prevents the
+        complication described in \sectionref{Mail deleted before delivery
+        is attempted} from occurring, but only in the case where there is a
+        single recipient; in that case the recipient address will be logged
+        also, but when there are multiple recipients no addresses are
+        logged.  It is also possible to warn for every recipient,
+        preventing the complication in \sectionref{Mail deleted before
+        delivery is attempted} entirely.
+
+    \item The algorithm does not distinguish between mails where one or
+        more mails are rejected and a subsequent mail is accepted; it will
+        appear in the database as one mail with lots of rejections followed
+        by acceptance (this has already been mentioned in
+        \sectionref{connection reuse}).  It does not appear to be possible
+        to make this distinction given the data Postfix logs, though it
+        might be possible to write a policy server to provide additional
+        logging.
+
+    \item The TIMEOUT action uses potentially incorrect heuristics to
+        decide whether the timeout applies to an accepted mail or not,
+        potentially leaving a mail in the state tables
+        (\sectionref{timeouts-during-data-phase}).
+
+    \item The program will not detect parsing the same log file twice,
+        resulting in the database containing duplicate entries.
+
+    \item The parser does not distinguish between log files produced by
+        different sources when parsing; all results will be saved to the
+        same database.  This may be viewed as an advantage, as log files
+        from different sources can be combined in the same database, or it
+        may be viewed as a limitation as there is no facility to
+        distinguish between log files from different sources in the same
+        database.  If the results of parsing log files from different
+        sources must remain separate, the parser can easily be instructed
+        to use a different database to store the results in.
+
+    \item The solution to complication \sectionref{Mails deleted during
+        delivery} involves discarding data.
+
+\end{enumerate}
+
+\subsubsection{Possible improvements}
+
+\begin{itemize}
+
+    \item Investigate and write the policy server referred to in limitation
+        3 above.
+
+    \item Improve the solution to complication \sectionref{Mails deleted
+        during delivery} so that data is not discarded.
+
+    \item Improve the heuristics used in
+        \sectionref{timeouts-during-data-phase}, or develop another
+        solution, to avoid incorrectly leaving a mail in the state tables.
+
+\end{itemize}
+
+\subsubsection{Summary}
+
+This section has covered the limitations of the parser and possible
+improvements which may be implemented in the future.

### 2008-07-22 doc/logparser-parsing-rules.tex
Collected all the implementation bits into one section.

@@
 rules and actions have been separated in the parser's design.
-
-\subsection{Example rule}
-
-\label{example rule}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below would match the following log line:
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% do not reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-cluster\_group      & 400                                               \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
-
-\begin{description}
-
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program which logged the log line, a match using the rule's
-        \regex{} will be attempted against the log line; if the match is
-        successful the action will be executed, if not the next rule will
-        be tried.  If the program-specific rules do not match the log line,
-        the parser will fall back to generic rules; if those rules are
-        unsuccessful a warning will be issued.
-
-    \item [action:] The action to be executed if the \regex{} matches
-        successfully (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} which
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-    \item [cluster\_group] The cluster\_group attribute is used by the
-        Decision Tree algorithm described in a separate document; the
-        parser does not use it in any way.
-
-\end{description}
 
@@
 
-\subsection{Creating new rules}
-
-\label{creating new rules}
-
-The log files produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \DNSBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial, and also occurs much more
-frequently.  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the preprocessing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.
-
-        As time goes on the amount of memory typically available increases
-        and the requirement to reduce memory requirements decreases, hence
-        the modified algorithm omits the hashing step and maintains
-        counters for all tokens. In addition the majority of unique or
-        infrequently occurring words will have been substituted with
-        keywords during the first phase, vastly reducing the number of
-        tuples to maintain counters for.
-
-    \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be misclassified as a fixed term, but that should be noticed
-        by the user when reviewing the new \regexes{}.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter maintained for each \regex{}.
-        Contiguous sequences of \texttt{.+} in the newly reassembled
-        \regexes{} are collapsed to a single \texttt{.+}; any resulting
-        duplicate \regexes{} are combined, and their counters added
-        together.  If the \regex{}'s counter is lower than the threshold
-        supplied by the user the \regex{} is discarded; this second
-        threshold is independent of the threshold used to differentiate
-        between fixed and variable words, but once again 1\% of input log
-        lines is a good starting point; obviously the threshold depends on
-        the number and type of input log lines.  The new \regexes{} are
-        displayed for the user to add to the database, either as new rules
-        or merged into the \regexes{} of existing rules; the counter for
-        each \regex{} is also displayed, giving the user an indication of
-        how many of the input log lines that \regex{} should match.
-        Discarding \regexes{} will result in some of the input log lines
-        not being matched; the utility should be run again once the
-        unmatched log lines have been isolated by parsing the original log
-        files once the parser has been augmented with the new rules.
-
-\end{description}
-
-A second utility is also provided which reads a list of new \regexes{} and
-the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
-required.
-
-These utilities are not intended to create perfect \regexes{}, but they
-greatly reduce the effort required to parse new or different log lines.
-
 \subsection{Summary}

### 2008-07-22 doc/logparser-parsing-rules.tex
Gathered all the results bits into their own section.

@@
 rules and actions have been separated in the parser's design.
-
-\subsection{Rule efficiency}
-
-\label{rule efficiency}
-
-XXX ADD TIMING INFORMATION\@.
-
-Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\subsubsection{Algorithmic complexity}
-
-An important property of a parser is how execution time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-\Graphref{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by \graphref{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better); \tableref{execution time vs file size vs number lines
-factor table} shows the ratios for different breakdowns of the log files.
-As the reader can see the ratios are quite tightly banded, showing that the
-algorithm scales linearly: the much larger log files between points 60 and
-70 on the X axis in \graphref{execution time vs file size vs
-number of lines graph} actually cause the ratio to increase (i.e.\
-improve), rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
-
-\subsubsection{Rule ordering for efficiency}
-
-\label{rule ordering for efficiency}
-
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules: the top 10\% match the vast majority
-of log lines, with the remaining log lines split across the other 90\% of
-the rules (as shown in \graphref{rule hits graph}).  Assuming that
-the distribution of log lines is reasonably steady over time, program
-efficiency should benefit from trying more frequently matching rules before
-those which match less frequently.  To test this hypothesis three full test
-runs were performed with different rule orderings:
-
-\begin{description}
-
-    \item [normal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
-
-    \item [shuffle] This is intended to represent a randomly ordered rule
-        set.  The rules will be shuffled once before use and will retain
-        that ordering for the entirety of the log file.  Note that the
-        ordering will change every time the parser is executed, so 10
-        different orderings will be generated for each log file in the test
-        run.
-
-    \item [reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
-
-\end{description}
-
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with \tableref{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of log files.  Overall this provides a modest but
-worthwhile performance increase of approximately 9\%, for a small
-investment in time and programming.
-
-\subsubsection{Caching each regex}
-
-\label{Caching each regex}
-
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  \Graphref{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; \graphref{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
-
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
-
-XXX ADD A TABLE\@.
 

### 2008-07-22 doc/logparser-results.tex
Gathered all the results bits into their own section.

--- 2008-07-11 doc/logparser-graphs.tex
@@
-\section{Graphs}
+\section{Results}
+
+\subsection{Efficiency}
+
+\subsubsection{Rule efficiency}
+
+\label{rule efficiency}
+
+XXX ADD TIMING INFORMATION\@.
+
+Parsing efficiency is an obvious concern when the parser routinely needs to
+deal with 75 MB log files containing 300,000 log lines (generated daily on
+a mail server handling mail for approximately 700 users --- large scale
+mail servers would have much larger log files on a daily basis).  When
+generating the data for the graphs included in \sectionref{graphs},
+\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
+log lines) were each parsed 10 times, the first run discarded, and the
+execution time for the remaining 9 runs averaged.  The first run is
+discarded for two reasons:
+
+\begin{enumerate}
+
+    \item The execution time will be higher because the log file must be
+        read from disk, whereas for subsequent runs the log file will be
+        cached in memory by the operating system.
+
+    \item The execution time will also be higher because the rule ordering
+        will be sub-optimal compared to subsequent runs.
+
+\end{enumerate}
+
+Saving results to the database was disabled for the test runs, as that
+dominates the run time of the program, and the tests are aimed at measuring
+the speed of the parser rather than the speed of the database and the disks
+the database is stored on.
+
+\subsubsection{Algorithmic complexity}
+
+An important property of a parser is how execution time scales relative to
+input size: does it scale linearly, polynomially, or exponentially?
+\Graphref{execution time vs file size vs number of lines graph}
+shows the execution time in seconds, file size in MB and tens of thousands
+of log lines per log file.  All three lines run roughly in parallel, giving
+a visual impression that the algorithm scales linearly with input size.
+This impression is borne out by \graphref{execution time vs file
+size vs number lines factor} which plots the ratio of file size vs
+execution time and ratio of number of log lines vs execution time (higher
+is better); \tableref{execution time vs file size vs number lines
+factor table} shows the ratios for different breakdowns of the log files.
+As the reader can see the ratios are quite tightly banded, showing that the
+algorithm scales linearly: the much larger log files between points 60 and
+70 on the X axis in \graphref{execution time vs file size vs
+number of lines graph} actually cause the ratio to increase (i.e.\
+improve), rather than decrease.  The strange behaviour where larger log
+files are parsed more efficiently is explained fully in \sectionref{Why are
+there dips in the graphs?}.
+
+\subsubsection{Rule ordering for efficiency}
+
+\label{rule ordering for efficiency}
+
+Rule ordering was mentioned in \sectionref{rule attributes} and will be
+covered in greater detail in this section.  At the time of writing there
+are \numberOFrules{} different rules: the top 10\% match the vast majority
+of log lines, with the remaining log lines split across the other 90\% of
+the rules (as shown in \graphref{rule hits graph}).  Assuming that
+the distribution of log lines is reasonably steady over time, program
+efficiency should benefit from trying more frequently matching rules before
+those which match less frequently.  To test this hypothesis three full test
+runs were performed with different rule orderings:
+
+\begin{description}
+
+    \item [normal]  The most optimal order, according to the hypothesis:
+        rules which match most often will be tried first.
+
+    \item [shuffle] This is intended to represent a randomly ordered rule
+        set.  The rules will be shuffled once before use and will retain
+        that ordering for the entirety of the log file.  Note that the
+        ordering will change every time the parser is executed, so 10
+        different orderings will be generated for each log file in the test
+        run.
+
+    \item [reverse] Hypothetically the worst order: the most frequently
+        matching rules will be tried last.
+
+\end{description}
+
+Graphs~\refwithpage{percentage increase of shuffled over normal}
+and~\refwithpage{percentage increase of reversed over normal} show the
+percentage increase of execution times, with \tableref{Execution
+time increase for different rule orderings} showing the mean increases for
+different groupings of log files.  Overall this provides a modest but
+worthwhile performance increase of approximately 9\%, for a small
+investment in time and programming.
+
+\subsubsection{Caching each regex}
+
+\label{Caching each regex}
+
+Perl compiles the original \regex{} into an internal representation,
+optimising the \regex{} to improve the speed of matching, but this
+compilation and optimisation takes CPU time; far more CPU time, in fact,
+than the actual matching takes.  Perl automatically caches static
+\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
+cached.  \Graphref{normal regex vs discard regex} shows execution
+times with and without caching the \regex{}.  Caching the compiled
+\regexes{} is obviously far more efficient; \graphref{normal regex
+vs discarded regex factor} shows the percentage execution time increase
+when not caching each \regex{}.
+
+Caching the compiled \regexes{} is quite simple, and is the single most
+effective optimisation implemented in the parser.
+
+XXX ADD A TABLE\@.
+
+\subsection{Coverage}
+
+\label{parsing coverage}
+
+The discussion of the parser's coverage of Postfix log files is separated
+into two parts: log lines covered and mails covered.  The first is
+important because the parser should handle all (relevant) log lines it is
+given; the second is equally important because the parser must properly
+deal with every mail if it is to be useful.  Improving the former is
+less intrusive, as it just requires new rules to be written; improving the
+latter is much more intrusive as it requires changes to the parser
+algorithm, and it can also be much harder to notice a deficiency.
+
+\subsubsection{Log lines covered}
+
+\label{log-lines-covered}
+
+Parsing a log line is a three stage process:
+
+\begin{enumerate}
+
+    \item Check if there are any rules for the Postfix component which
+        produced the log line; if not then skip the log line.
+
+    \item Try each rule until a matching rule is found; if no match is
+        issue a warning and move to the next log line.
+
+    \item Execute the action specified by the rule.
+
+\end{enumerate}
+
+Full coverage of log lines requires the following:
+
+\begin{enumerate}
+
+    \item Each Postfix component whose log lines are of interest must have
+        at least one rule or its log lines will be silently skipped; in the
+        extreme case of zero rules the parser would happily skip every log
+        line.  There may be any number of log lines from other programs
+        intermingled in the log file, and there are some Postfix programs
+        which do not produce any log lines of interest.
+
+    \item There must be a rule to match each different log line produced by
+        each program; if a log line is not successfully matched the parser
+        will issue a warning.  Rules should be as specific and tightly
+        bound as possible to ensure accurate parsing:\footnote{A rule which
+        matches zero or more of any character will successfully parse every
+        log line, but not in a meaningful way.} most log lines contain
+        fixed strings and have a rigid pattern, so this is not a problem.
+
+    \item The appropriate action to take --- discussed in
+        \sectionref{mails-covered}.
+
+\end{enumerate}
+
+Full coverage of log lines is easy to achieve yet hard to maintain.  It is
+easy to achieve full coverage for a limited set of log files (at the time
+of writing the parser is tested with \numberOFrules{} rules, fully parsing
+\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
+rules are easy to add.  Maintaining full coverage is hard because other
+servers have different restrictions with custom messages, \DNSBL{} messages
+change over time, major releases of Postfix change warning messages
+(usually adding more information), etc.,\ so over time the log lines drift
+and change.  \Graphref{rule hits graph} shows the number of hits
+for each rule over all \numberOFlogFILES{} log files; it is obvious that a
+small number of rules match the vast majority of the lines, and more than
+half the rules match fewer than 100 times.
+
+Warnings are issued for any log lines which are not parsed; no warnings are
+issued for unparsed log lines while testing with the \numberOFlogFILES{}
+test log files, so it can be safely concluded that there are zero false
+negatives.  False positives are harder to quantify: short of examining each
+of the 60,721,709 log lines and determining which \regex{} parsed it, there
+is no way to be sure that every line was parsed by the correct \regex{},
+making it impossible to quantify the false positive rate; however a random
+sample of 6,039 log lines was parsed and the results checked manually to
+ensure that the correct \regex{} parsed each log line.\footnote{Each log
+line was examined and the correct \regex{} identified from the
+\numberOFrules{} rules in the database; the correct \regex{} was then
+compared to the \regex{} which was used by the parser.}  The sample was
+generated by running the following command:
+
+\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
+
+\noindent{}to randomly extract roughly one line in every 10,000.  Although
+on initial appearances exercising only 36 rules (from a total of
+\numberOFrules{}) when parsing 6039 log lines seems quite low, after
+examining \graphref{rule hits graph} it becomes apparent that such
+a low hit rate is to be expected; the reader should also bear in mind that
+even when parsing all \numberOFlogFILES{} log files not all the rules are
+exercised (some of the rules are for parsing log lines which only appear in
+other log files).
+
+\subsubsection{Mails covered}
+
+\label{mails-covered}
+
+Coverage of mails is much more difficult to determine accurately than
+coverage of log lines.  The parser can dump its state tables in a human
+readable form; examining these tables with reference to the log files is
+the best way to detect mails which were not handled properly (many of the
+complications discussed in \sectionref{additional complications} were
+detected in this way).  The parser issues warnings when it detects any
+errors, some of which may alert the user to a problem, e.g.\ when a queueid
+is reused before the previous mail is fully dealt with, when a queueid or
+\pid{} is not found in the state tables,\footnote{There will often be
+warnings about a missing queueid or \pid{} in the first few hundred or
+thousand log lines because the earlier log lines for those connections or
+mails are in the previous log file; loading the saved state from the
+previous log file will solve this problem.} or when there are problems
+tracking a child mail (see \sectionref{tracking re-injected mail}).  There
+should be few or no warnings when parsing, and when finished parsing the
+state table should only contain entries for mails which had yet to be
+delivered when the log files ended, or were accepted before the log files
+began.
+
+At the time of writing the parser is being tested with \numberOFlogFILES{}
+log files.  There are 5 warnings produced, but because the parser errs on
+the side of producing more warnings rather than fewer, those 5 warnings
+represent 3 instances of 1 problem: 3 connections started before the first
+log file, so their initial log entries are missing, leading to warnings
+when their log lines are parsed.
+
+The state tables contain entries for mails not yet delivered when the
+parser finishes execution.  Ideally all they should contain are mails which
+are awaiting delivery after the period covered by the log files, though
+they may also contain mails whose initial entries are not contained in the
+log files.  Any other entries are evidence of a failure in parsing or an
+aberration in the log files.  After parsing the \numberOFlogFILES{} test
+log files the state tables contain 18 entries, breaking down into:
+
+\begin{itemize}
+
+    \item 1 connection which started only seconds before the log files
+        ended and had not yet been fully transferred from client to server.
+
+    \item 1 mail which had been accepted only seconds before the log files
+        ended and had not yet been delivered.
+
+    \item 9 mails whose initial log entries were not present in the log
+        files.
+
+    \item 7 mails which had yet to be delivered due to repeated failures.
+
+\end{itemize}
+
+There are no mails in the state tables which should not be present, thus it
+can be concluded that there are zero false negatives.  Once again,
+determining the false positive rate is much harder, as manually checking
+the results of parsing 13,850,793 connections and mails accepted, rejected,
+bounced or delivered is infeasible.  There is considerable circumstantial
+evidence that the false positive rate is quite low:
+
+\begin{itemize}
+
+    \item The parser is quite verbose when complaining about known problems
+        (e.g.\ if a mail is tracked twice as described in
+        \sectionref{tracking re-injected mail}), and no such warnings are
+        produced during the test runs.
+
+    \item Queueids and \pids{} naturally group together log lines belonging
+        to one mail or connection respectively; it is extremely unlikely
+        that a log line would be associated with the wrong connection.
+
+    \item When dealing with the complications described in
+        \sectionref{complications} and \sectionref{additional
+        complications} the solutions are as specific and restrictive as
+        possible, with the goal of minimising the number of false
+        positives.  In addition the solution to the \textit{Out of order
+        log files\/} complication described in \sectionref{out of order log
+        lines} imposes conditions which each reassembled mail must comply
+        with to be acceptable.
+
+    \item Every effort has been made while developing to make the parser as
+        precise, demanding and particular as possible.
+
+\end{itemize}
+
+Whereas verifying by inspection that the parser correctly deals with all
+60,721,709 lines in the test log files is infeasible, verifying a subset of
+those log files is a tractable, if extremely time consuming, task.  A
+sample of log lines was obtained by randomly selecting a log file:
+
+\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
+\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
+
+The first 6000 lines of this log file (roughly 0.01\% of the total number
+of log lines used in testing) was extracted:
+
+\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
+
+It is important that the log lines used are contiguous so that all log
+entries are present for as many of the connections as possible.  This log
+segment was parsed with all debugging options enabled, resulting in 167,448
+lines of output.\footnote{A mean of 27.908 lines of output per line of
+input; each connection has 30 debugging lines, plus 21 debugging lines per
+result.  Connections which have been cloned will have the cloned connection
+in their debugging output, plus another 33 debugging lines.  Those numbers
+are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
+between the number of log lines and debugging lines: $33(connections) +
+30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
+and has not been rigorously verified.}  All 167,448 lines were examined in
+conjunction with the input log and a dump of the resulting database,
+verifying that for each of the input lines the parser used the correct rule
+and executed the correct action, which in turn produced the correct result
+and inserted the correct data in the database.  The log segment produced 4
+warnings, 10 mails remaining in the state tables, and 1625 connections
+correctly entered in the database.
+
+Given the circumstantial and experimental evidence detailed above, the
+author is confident that the false positive rate when reconstructing a mail
+is exceedingly low, if not approaching zero.
+
+\subsubsection{Summary}
+
+Parser coverage is divided into two topics in this section: log lines
+covered and mails covered.  The former is initially more important, as the
+parser must successfully parse every line if it is to be complete, but
+subsequently the latter takes precedence because reproducing the path a
+mail takes through Postfix is the aim of the parser.  Increasing the
+percentage of log lines parsed is relatively simple and non-intrusive:
+adding new rules or modifying existing rules is simplified by the
+separation of rules, actions and framework.  Improving the logical coverage
+is harder, as the actions taken by Postfix must be reconstructed by the
+author, and the new sequence of actions integrated into the existing model
+without breaking the existing parsing.  Detecting a deficiency in the
+parsing algorithm is also significantly harder than detecting unparsed log
+lines, as the parser will warn about any unparsed line, whereas discovering
+a flaw in the parser requires understanding of the warnings produced and
+the mails remaining in the state table.  Rectifying a flaw in the parser
+requires an understanding of both the parser and Postfix's log files, and
+investigative work to determine the cause of the deficiency, followed by
+further examination of the log files in developing a solution.
+
+\subsection{Graphs}
 
@@
 
+XXX EXPLAIN THE X-AXES ON THE GRAPHS\@.
+
 \showgraph{build/plot-normal-vs-smaller}{NORMAL VS SMALLER}{NORMAL VS
@@
 
-\subsection{Parser scalability}
+\newpage
+
+\subsubsection{Parser scalability}
 
@@
 
-\subsection{Rule ordering}
+\subsubsection{Rule ordering}
 
@@
 
-\subsection{Caching regexes}
+\subsubsection{Caching regexes}
 
@@
 
-\subsection{Rule hits}
+\subsubsection{Rule hits}
 \label{rule hits}
@@
 
-\subsection{Miscellaneous graphs}
+\subsubsection{Miscellaneous graphs}
 

### 2008-07-23 doc/logparser-conclusion.tex
Improved parts of the efficiency section and inlined graphs. Fixed up some references; added $Id$ to front page.

@@
 though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
+(answered in \sectionref{parser efficiency}).  Providing a tool
 (\sectionref{creating new rules}) to ease the generation of \regexes{} from

### 2008-07-23 doc/logparser-implementation.tex
Improved parts of the efficiency section and inlined graphs. Fixed up some references; added $Id$ to front page.

@@
         each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Rule efficiency concerns are
-        discussed in \sectionref{rule efficiency}, with the impact of
+        in a large, data dependent slowdown.  Efficiency concerns are
+        discussed in \sectionref{parser efficiency}, with the impact of
         caching compiled \regexes{} covered in \sectionref{Caching each
@@
 \subsection{Rules}
+
+\label{rules}
 

### 2008-07-23 doc/logparser-results.tex
Improved parts of the efficiency section and inlined graphs. Fixed up some references; added $Id$ to front page.

@@
 
-\subsection{Efficiency}
+XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
 
-\subsubsection{Rule efficiency}
+XXX DO SOMETHING WITH NORMAL VS SMALLER GRAPH\@.
 
-\label{rule efficiency}
+XXX EXPLAIN THE X-AXES ON THE GRAPHS\@.
+
+\showgraph{build/plot-normal-vs-smaller}{NORMAL VS SMALLER}{NORMAL VS
+SMALLER}
+
+\subsection{Parser Efficiency}
+
+\label{parser efficiency}
 
@@
 mail servers would have much larger log files on a daily basis).  When
-generating the data for the graphs included in \sectionref{graphs},
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the first run discarded, and the
+generating the timing data used in this section, \numberOFlogFILES{} log
+files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
+parsed 10 times, the results of the first first run discarded, and the
 execution time for the remaining 9 runs averaged.  The first run is
@@
 input size: does it scale linearly, polynomially, or exponentially?
-\Graphref{execution time vs file size vs number of lines graph}
-shows the execution time in seconds, file size in MB and tens of thousands
-of log lines per log file.  All three lines run roughly in parallel, giving
-a visual impression that the algorithm scales linearly with input size.
-This impression is borne out by \graphref{execution time vs file
-size vs number lines factor} which plots the ratio of file size vs
-execution time and ratio of number of log lines vs execution time (higher
-is better); \tableref{execution time vs file size vs number lines
-factor table} shows the ratios for different breakdowns of the log files.
-As the reader can see the ratios are quite tightly banded, showing that the
+\Graphref{execution time vs file size vs number of lines graph} shows the
+execution time in seconds, file size in MB, and tens of thousands of log
+lines per log file.  All three lines run roughly in parallel, giving the
+impression that the algorithm scales linearly with input size.  This
+impression is borne out by \graphref{execution time vs file size vs number
+lines factor} which plots both the ratio of file size vs execution time,
+and the ratio of number of log lines vs execution time (higher is better in
+both cases); \tableref{execution time vs file size vs number lines factor
+table} shows the ratios for different groups of log files.  As the
+reader can see the ratios are quite tightly banded, showing that the
 algorithm scales linearly: the much larger log files between points 60 and
-70 on the X axis in \graphref{execution time vs file size vs
-number of lines graph} actually cause the ratio to increase (i.e.\
-improve), rather than decrease.  The strange behaviour where larger log
-files are parsed more efficiently is explained fully in \sectionref{Why are
-there dips in the graphs?}.
+70 on the X axis in \graphref{execution time vs file size vs number of
+lines graph} actually cause the ratio to increase (i.e.\ improve), rather
+than decrease.  The strange behaviour where larger log files are parsed
+more efficiently is explained fully in \sectionref{Why are there dips in
+the graphs?}.
+
+\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
+size vs number of lines}{execution time vs file size vs number of lines
+graph}
+
+The Y axis in \graphref{execution time vs file size vs number of lines
+graph} represents the following:
+
+\begin{enumerate}
+
+    \item The time required, in seconds, to parse the log file.
+
+    \item The size of the log file, in megabytes.
+
+    \item The number of lines in the log file, divided by 10000.
+
+\end{enumerate}
+
+\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
+size and number of lines to execution time}{execution time vs file size vs
+number lines factor}
+
+\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
+size \& number of lines to execution time: statistics}{execution time vs
+file size vs number lines factor table}
 
@@
 
-Rule ordering was mentioned in \sectionref{rule attributes} and will be
-covered in greater detail in this section.  At the time of writing there
-are \numberOFrules{} different rules: the top 10\% match the vast majority
-of log lines, with the remaining log lines split across the other 90\% of
-the rules (as shown in \graphref{rule hits graph}).  Assuming that
-the distribution of log lines is reasonably steady over time, program
-efficiency should benefit from trying more frequently matching rules before
-those which match less frequently.  To test this hypothesis three full test
-runs were performed with different rule orderings:
+At the time of writing there are \numberOFrules{} different rules: the top
+10\% match the vast majority of log lines, with the remaining log lines
+split across the other 90\% of the rules (as shown in \graphref{rule hits
+graph}).  Assuming that the distribution of log lines is reasonably steady
+over time, parser efficiency should benefit from trying more frequently
+matching rules before those which match less frequently.  To test this
+hypothesis three full test runs were performed with different rule
+orderings:
 
@@
 
-Graphs~\refwithpage{percentage increase of shuffled over normal}
-and~\refwithpage{percentage increase of reversed over normal} show the
-percentage increase of execution times, with \tableref{Execution
-time increase for different rule orderings} showing the mean increases for
-different groupings of log files.  Overall this provides a modest but
-worthwhile performance increase of approximately 9\%, for a small
-investment in time and programming.
+\Graphref{percentage increase of shuffled over normal} shows the percentage
+increase of shuffled over normal, with the percentage increase of reverse
+over normal following in \graphref{percentage increase of reversed over
+normal}; the mean increases for different groupings of log files are given
+in \tableref{Execution time increase for different rule orderings}.
+Overall this optimisation provides a modest but worthwhile performance
+increase of approximately 10\%, for a small investment in time and
+programming.
+
+\label{rule ordering graphs}
+
+XXX SHOW NORMAL AS A REDUCTION FROM SHUFFLE, AND REVERSE AS AN INCREASE
+OVER SHUFFLE\@.
+
+\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
+shuffled over normal}{percentage increase of shuffled over normal}
+
+\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
+reversed over normal}{percentage increase of reversed over normal}
+
+\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
+increase for different rule orderings}{Execution time increase for
+different rule orderings}
 
@@
 
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-XXX DO SOMETHING WITH NORMAL VS SMALLER GRAPH\@.
-
-XXX EXPLAIN THE X-AXES ON THE GRAPHS\@.
-
-\showgraph{build/plot-normal-vs-smaller}{NORMAL VS SMALLER}{NORMAL VS
-SMALLER}
-
-\renewcommand{\figurename}{Graph}
-
-\subsubsection{Introduction}
-
 Graphs are an excellent means of displaying data, transforming a
 meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  These graphs are used to
-illustrate the topics discussed in \sectionref{rule efficiency}.  The
+anomalies and patterns are immediately obvious.  The
 graphs in the first section cover parser scalability, demonstrating that
@@
 a single test run of the parser.
-
-\newpage
-
-\subsubsection{Parser scalability}
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in \graphref{execution time vs file size vs number of
-lines graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-\Graphref{execution time vs file size vs number lines factor}
-shows the ratio of execution time vs file size and number of lines (higher
-is better, it means more bytes or lines processed per second).  The ratios
-are quite tightly banded with the exception of log files 22 and 62--68,
-where they are noticeably higher; \graphref{execution time vs file
-size vs number lines factor} and \tableref{execution time vs file
-size vs number lines factor table} show that the parser's execution time
-scales linearly with input size.
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
-
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
-
-\clearpage
-
-\subsubsection{Rule ordering}
-
-\label{rule ordering graphs}
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
-
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
 
@@
 
-The graphs presented in this section illustrate the topics discussed in
-\sectionref{rule efficiency}.  The first collection of graphs are about
+The first collection of graphs are about
 parser scalability, showing the linear relationship between execution time

### 2008-07-23 doc/logparser.tex
More proofreading changes.
Merged 'Other Postfix log parsers reviewed' and 'Previous research in this area' to create a 'State of the Art Review', which is now section 3.
Removed 'Throughput monitor' review.

Improved introduction and summary in 'State of the Art Review'; miscellaneous
related changes.

Moved 'Graphs' section closer to where it will eventually be, the 'Results'
section.
Moved the architecture stuff into it's own section.
Collected all the implementation bits into one section.
Gathered all the results bits into their own section.
Removed empty files.
Improved parts of the effic

@@
 \makeacronym{}
+
+% Embed SVN Id etc.
+\usepackage{svn}
 
@@
 % graph references
+% SHOULD THESE BE LABELLED WITH graph OR figure?
 \newcommand{\graphref}[1]{%
@@
 %\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
+\newcommand{\showtable}[3]{%
     \begin{table}[ht]
@@
     \end{table}
+}
+
+\newcommand{\tableline}[0]{%
+    \noalign{\smallskip}%
+    \hline%
+    \noalign{\smallskip}%
 }
@@
 
+% WSUIPA fonts.
+\input{ipamacs}
+
 \title{Parsing Postfix log files}
@@
 \maketitle
+\SVN$Id$
+\SVNId{}
 
@@
 
-XXX ``.'' AFTER ETC\@; SHOULD I HAVE A BACKSLASH AFTER THE ``.''?
-
 \newpage
@@
     Complications
+    Limitations and Improvements
 Results - BETTER TITLE?
@@
     Efficiency
-    Limitations and Improvements
 Conclusion
@@
 \input{logparser-background.tex}
-\input{logparser-database-schema.tex}
+\input{logparser-other-parsers-review.tex}
+\input{logparser-architecture.tex}
+\input{logparser-implementation.tex}
 \input{logparser-parsing-rules.tex}
-\input{logparser-parsing-algorithm.tex}
-\input{logparser-coverage.tex}
-\input{logparser-limitations.tex}
+\input{logparser-results.tex}
 \input{logparser-conclusion.tex}
@@
 
-\input{logparser-other-parsers-review.tex}
-
 \bibliographystyle{logparser-bibliography-style}
@@
 \label{bibliography}
-
-\input{logparser-graphs.tex}
 

### 2008-07-25 doc/logparser-implementation.tex
Lots of improvements to the efficiency section; lots more to come.

@@
-\section{Parser Implementation}
+\section{Postfix Parser Implementation}
 
@@
         discussed in \sectionref{parser efficiency}, with the impact of
-        caching compiled \regexes{} covered in \sectionref{Caching each
-        regex}.
+        caching compiled \regexes{} covered in \sectionref{Caching compiled
+        regexes}.
 

### 2008-07-25 doc/logparser-introduction.tex
Improved parts of the efficiency section and inlined graphs. Fixed up some references; added $Id$ to front page.
Lots of improvements to the efficiency section; lots more to come.

@@
 unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm, utilising the graphs in
-\sectionref{graphs} for illustrative purposes.  The section concludes with
-a description of using the tools provided with the parser to generate new
-rules (specifically the \regex{} in each rule) from unparsed log lines.
+particular the optimisations used by the algorithm.  The section concludes
+with a description of using the tools provided with the parser to generate
+new rules (specifically the \regex{} in each rule) from unparsed log lines.
 
@@
 
-Section~\ref{graphs} contains graphs illustrating the topics previously
-discussed in \sectionref{rule efficiency} (rule efficiency), and explains
-the irregularities observed in these graphs.
-
 Section~\ref{limitations-improvements} lists the limitations of the

### 2008-07-25 doc/logparser-results.tex
Lots of improvements to the efficiency section; lots more to come.

@@
 \section{Results}
+
+XXX NEED AN INTRODUCTION\@.
 
@@
 
-\showgraph{build/plot-normal-vs-smaller}{NORMAL VS SMALLER}{NORMAL VS
-SMALLER}
+XXX CHANGE normal TO optimal IN ORDERING STUFF, INCLUDING IN GRAPH CAPTIONS
+AND AXES LABELS\@.
 
@@
 
+XXX THIS SECTION NEEDS TO BE REWRITTEN\@.
+
 Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would have much larger log files on a daily basis).  When
-generating the timing data used in this section, \numberOFlogFILES{} log
-files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
+parse very large log files.  The server which generated the log files used
+in testing this parser accepts approximately 10,000 mails for 700 users a
+day, and on average it produces 75MB log files containing 300,000 log lines
+--- large scale mail servers would have much larger log files on a daily
+basis.
+
+\showgraph{build/plot-mails-received}{Mails received per day}{Mails
+received per day}
+
+As expected there are far more mails received on weekdays than at weekends.
+Note that this graph and the table below show the number of mails received
+by \SMTP{} only; in particular the mail loops noticeable in other graphs
+do not contribute to these figures.
+
+\showtable{build/mails-received-include-for-graph}{Number of mails received
+per day: statistics}{Number of mails received per day: statistics}
+
+\showgraph{build/plot-action-distribution}{Distribution of rules per
+action}{Distribution of rules per action}
+
+When generating the timing data used in this section, \numberOFlogFILES{}
+log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
 parsed 10 times, the results of the first first run discarded, and the
-execution time for the remaining 9 runs averaged.  The first run is
-discarded for two reasons:
+execution time for the remaining 9 runs averaged.  The results from the
+first run are discarded for two reasons:
 
@@
 the database is stored on.
+
+\newpage
 
@@
 than decrease.  The strange behaviour where larger log files are parsed
-more efficiently is explained fully in \sectionref{Why are there dips in
-the graphs?}.
+more efficiently is explained fully in \sectionref{Explaining the peaks in
+log file size}.
 
@@
 file size vs number lines factor table}
+
+\subsubsection{Explaining the peaks in log file size}
+
+\label{Explaining the peaks in log file size}
+
+\Graphref{execution time vs file size vs number of lines graph} shows log
+file size varying between 30--100MB, except for log files 22 (200MB) and
+62--68 (400--900MB).  These large log files were caused by a user
+unintentionally creating a mail forwarding loop, resulting in a huge number
+of mail deliveries and up to an order of magnitude more log lines than is
+usual.  The user specified his mail forwarding as: \newline
+\tab{}\texttt{$\backslash$username, username@domain} \newline This
+instructs Postfix to deliver the mail to the local user, and also forward
+it to the remote address; this is generally not a problem except that the
+remote address in this case is the address the mail was originally sent to,
+creating an infinite loop.  To prevent this happening Postfix examines the
+Delivered-To header in the mail, and if the mail has already been delivered
+to the current address it is bounced back to the sender with the error
+message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
+works well, but unfortunately in this case the user noticed they had not
+received any mail in a while and opted to send a test mail to themselves,
+causing a loop not caught by Postfix:
+
+\begin{enumerate}
+
+    \item Postfix accepts a mail from username@domain, for username@domain.
+
+    \item Postfix delivers the mail to the local mailbox and
+        username@domain, as instructed by the user's forwarding
+        instructions. The forwarded mail has a
+        \texttt{Delivered-To:~username@domain} header added, and the
+        envelope sender address is username@domain.
+
+    \item Postfix accepts the mail for username@domain, but while
+        delivering it notices that the \texttt{Delivered-To} header already
+        contains the address it is currently delivering to, and therefore
+        sends a bounce notification with sender address \texttt{<>} to
+        the original sender: username@domain.
+
+    \item Postfix accepts the bounce notification and delivers it to both
+        the local mailbox and to username@domain, as instructed by the
+        user's forwarding instructions.  A
+        \texttt{Delivered-To:~username@domain} header is added to the
+        forwarded bounce notification, which now has an envelope sender
+        address of username@domain.
+
+    \item Postfix accepts the forwarded bounce notification but while
+        delivering the mail it notices that the \texttt{Delivered-To}
+        header already contains the address currently being delivered to,
+        and sends a bounce notification to the sender: username@domain.
+
+    \item Repeat from step two; this will continue indefinitely unless an
+        administrator intervenes and deletes the appropriate mails from the
+        queue.
+
+\end{enumerate}
+
+The sequence described above occurs extremely rapidly because Postfix does
+not have to deliver the mail to an external system, so mails are delivered,
+bounced and generated as fast as the disks can keep up, resulting in a huge
+volume of log lines.
+
+The different distribution of log lines during a mail forwarding loop has
+interesting effects on execution time, described in detail in
+\sectionref{rule ordering for efficiency}, \sectionref{scalability as the
+number of rules rises}, and \sectionref{Caching compiled regexes}.
 
@@
 
-\Graphref{percentage increase of shuffled over normal} shows the percentage
-increase of shuffled over normal, with the percentage increase of reverse
-over normal following in \graphref{percentage increase of reversed over
-normal}; the mean increases for different groupings of log files are given
-in \tableref{Execution time increase for different rule orderings}.
-Overall this optimisation provides a modest but worthwhile performance
-increase of approximately 10\%, for a small investment in time and
-programming.
+\Graphref{Execution time relative to
+shuffled ordering graph} shows the execution time of normal and reverse orderings
+relative to shuffled ordering; the mean increases for different groupings
+of log files are given in \tableref{Execution time relative to shuffled
+ordering table}.  Overall this optimisation provides a modest but worthwhile
+performance increase of approximately 10\%, for a small investment in time
+and programming.
 
-\label{rule ordering graphs}
-
-XXX SHOW NORMAL AS A REDUCTION FROM SHUFFLE, AND REVERSE AS AN INCREASE
-OVER SHUFFLE\@.
-
-\showgraph{build/plot-normal-shuffle-factor}{Percentage increase of
-shuffled over normal}{percentage increase of shuffled over normal}
-
-\showgraph{build/plot-normal-reverse-factor}{Percentage increase of
-reversed over normal}{percentage increase of reversed over normal}
+\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Execution
+time of normal and reverse orderings relative to shuffled
+ordering}{Execution time relative to shuffled ordering graph}
 
 \showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-increase for different rule orderings}{Execution time increase for
-different rule orderings}
+relative to shuffled ordering}{Execution time relative to shuffled ordering
+table}
 
-\subsubsection{Caching each regex}
+XXX IMPROVE THE REMAINDER OF THIS SECTION\@; CLEARLY SEPARATE THE
+CHARACTERISTICS AND CLARIFY THEM\@.
 
-\label{Caching each regex}
+Different rule orderings have less effect when parsing log files 22 and
+62--68, due to the different distribution of log lines in those log files
+(see \sectionref{Explaining the peaks in log file size}).  The vast
+majority of log lines when a mail loop occurs are from Postfix components
+which have a small number of rules associated with them, whereas in general
+\daemon{smtpd} adds the majority of log lines, and also has the highest
+number of rules.  \daemon{smtpd} log lines are distributed across rules
+much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
+\daemon{cleanup} or \daemon{pickup}, so the average number of rules
+required to parse a \daemon{smtpd} log line is much higher than the average
+number required to parse other log lines.
 
-Perl compiles the original \regex{} into an internal representation,
-optimising the \regex{} to improve the speed of matching, but this
-compilation and optimisation takes CPU time; far more CPU time, in fact,
-than the actual matching takes.  Perl automatically caches static
-\regexes{}, but dynamic \regexes{} need to be explicitly compiled and
-cached.  \Graphref{normal regex vs discard regex} shows execution
-times with and without caching the \regex{}.  Caching the compiled
-\regexes{} is obviously far more efficient; \graphref{normal regex
-vs discarded regex factor} shows the percentage execution time increase
-when not caching each \regex{}.
+These two characteristics combine to reduce the average number of rules
+required to parse a log line when there is a mail loop.  The ratio of
+number of lines versus execution time in \graphref{execution time vs file
+size vs number lines factor} shows a noticeable improvement at log files 22
+and 62--68, where a higher ratio means more log lines parsed per unit of
+execution time.
 
-Caching the compiled \regexes{} is quite simple, and is the single most
-effective optimisation implemented in the parser.
+When the rule ordering is reversed the majority of log lines generated by a
+mail loop will be parsed with very few rules, whereas without a mail loop
+the majority of log lines require a large number of rules; this leads to a
+noticeable drop in the average time required to parse a log line, as shown
+in \graphref{Execution time relative to shuffled ordering graph}.  The
+number of rules which need to be consulted when the ordering is shuffled
+varies between the optimal and the pessimal, and the performance varies
+proportionally.
 
-XXX ADD A TABLE\@.
+\subsubsection{Scalability as the number of rules rises}
+
+\label{scalability as the number of rules rises}
+
+XXX WRITE THIS SECTION\@.
+
+\showgraph{build/plot-normal-vs-smaller}{NORMAL VS SMALLER}{NORMAL VS
+SMALLER}
+
+\subsubsection{Caching compiled regexs}
+
+\label{Caching compiled regexes}
+
+Before Perl attempts to match a \regex{} against a piece of text, the
+original \regex{} is compiled into an internal representation and optimised
+to improve the speed of matching, but this compilation and optimisation
+takes CPU time; far more CPU time, in fact, than the actual matching takes
+in many cases.  Perl automatically caches static \regexes{} at compile
+time, but \regexes{} loaded dynamically while the program need to be
+explicitly compiled and cached.
+
+The following graphs show the impact that not caching compiled \regexes{}
+has on parser performance: on typical log files the execution time when not
+caching compiled \regexes{} is 500--600\% of the execution time when
+caching; reversing the perspective shows that cached execution time is
+merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
+is probably the single most effective optimisation possible in the parser's
+implementation; given that it only required sixteen extra lines of
+reasonably simple code,\footnote{The sixteen lines of code breaks down as
+follows: three lines to add caching, eight lines of error checking and
+reporting, three lines to optionally disable caching for debugging and
+performance measurement, and two lines of comments.} the investment in time
+was minimal.
+
+\showgraph{build/plot-cached-discarded}{Regex: cached vs
+discarded}{normal regex vs discard regex}
+
+\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
+execution time increase}{normal regex vs discarded regex factor}
+
+Two large dips can be seen in \graphref{normal regex vs discarded
+regex factor} at log files 22 and 62--68, corresponding to the spikes in
+log file size in \graphref{execution time vs file size vs number
+of lines graph} and \graphref{normal regex vs discard regex}.  The
+reason for the anomalous log files has already been explained in
+\sectionref{Explaining the peaks in log file size}.
+
+The distribution of log lines across rules when there is a mail loop is
+much different to the norm and the average number of rules consulted per
+log line is much lower; this results in far fewer \regex{} compilations per
+line than when there is not a mail loop, and a correspondingly decreased
+execution time.  The unexpected outcome is that caching \regexes{} is
+proportionally less important when the log file contents were created by a
+mail loop.  The increases in execution time when not caching \regexes{} for
+different combinations of log files are summarised in the table below:
+
+XXX THIS IS AN AWFUL CAPTION\@.
+
+\showtable{build/stats-cached-discarded-include-for-graph}
+{Regex caching/discarding with different groups of log files}
+{Regex caching/discarding with different groups of log files}
+
+\subsubsection{Summary}
+
+This section began with parser scalability, showing the linear relationship
+between execution time and input size.  Demonstrates the effect of rule
+ordering on execution time, and the unexpected consequences of specific
+inputs.  The necessity of caching compiled \regexes{} is attested to by the
+third group of graphs, where the difference between caching and not caching
+is staggering.  The penultimate section contains a breakdown of the rule
+hits accumulated during a single test run of the parser.  Miscellaneous
+graphs expected to be useful are collected in the final section.
 
@@
 
-\label{graphs}
-
-Graphs are an excellent means of displaying data, transforming a
-meaningless stream of numbers into an easily comprehensible form, where
-anomalies and patterns are immediately obvious.  The
-graphs in the first section cover parser scalability, demonstrating that
-performance scales linearly with input size.  The impact of rule ordering
-is shown in \sectionref{rule ordering graphs}, and the anomalous dips and
-peaks apparent in some graphs are explained.  The third group of graphs
-vividly shows the huge impact that caching compiled \regexes{} has on
-parser performance.  Miscellaneous graphs are presented in the final group
-of graphs; these graphs are referenced from various places in the test.
-This section concludes with a breakdown of the rule hits accumulated during
-a single test run of the parser.
-
-\subsubsection{Why are there dips in the graphs?}
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 correspond to peaks in log file size in
-\graphref{execution time vs file size vs number of lines graph},
-and peaks in \graphref{execution time vs file size vs number lines
-factor} (where a peak means that more lines are processed per second, i.e.\
-performance is better).  The explanation for this took some time to arrive
-at, but it turns out to be reasonably simple.  The large log files in
-question were caused by a mail forwarding loop, where the distribution of
-log lines is quite different to normal, resulting in different performance
-characteristics.
-
-The mail loop was set up by a user modifying his mail forwarding to:
-\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
-remote address in this case is the address the mail was originally sent to,
-creating an infinite loop.  To prevent this happening Postfix examines the
-Delivered-To header in the mail, and if the mail has already been delivered
-to the current address it is bounced back to the sender with the error
-message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
-works well, but unfortunately in this case the user noticed they had not
-received any mail in a while and opted to send a test mail to themselves,
-causing a loop not caught by Postfix:
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions. The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice), \daemon{cleanup} and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it is currently delivering to, and therefore
-        sends a bounce notification with sender address \texttt{<>} to
-        the original sender: username@domain.  Log lines are added by
-        \daemon{local}, \daemon{qmgr} (twice) and \daemon{cleanup}.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.  Log lines are added by \daemon{local},
-        \daemon{qmgr} (twice), \daemon{cleanup}, and finally
-        \daemon{pickup}.
-
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
-        Log lines are added by \daemon{local}, \daemon{qmgr} (twice) and
-        \daemon{cleanup}.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
-volume of log lines.
-
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general \daemon{smtpd} adds the majority of log lines, and also has the
-highest number of rules.  \daemon{smtpd} log lines are distributed across
-rules much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher that the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number or rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in \graphref{execution time vs file size vs number lines factor}.
-When the rule ordering is reversed the majority of log lines generated by a
-mail loop will be parsed with very few rules, whereas without a mail loop
-the majority of log lines require a large number of rules; this leads to a
-noticeable drop in the average time required to parse a log line, as shown
-in \graphref{percentage increase of reversed over normal}.  The number of
-rules which need to be consulted when the ordering is shuffled varies
-between the optimal and the pessimal, and the performance varies
-proportionally.
-
-The difference between log files with a mail loop and log files without can
-be seen in \tableref{Execution time increase for different rule
-orderings} showing the increases for the different rule orderings and
-combinations of log files:
-
-\subsubsection{Caching regexes}
-
-\label{Caching regexes}
-
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
-
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
-
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
-
-Two large dips can be seen in \graphref{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in graphs~\refwithpage{execution time vs file size vs number
-of lines graph} and~\refwithpage{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Why are there dips in the graphs?}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different to the norm and the average number of rules consulted per
-log line is much lower; this results in far fewer \regex{} compilations per
-line than when there is not a mail loop, and a correspondingly decreased
-execution time.  The unexpected outcome is that caching \regexes{} is
-proportionally less important when the log file contents were created by a
-mail loop.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
-
-\subsubsection{Rule hits}
-\label{rule hits}
-
 The number of hits per rule is quite unevenly spread, resembling a Power
@@
 
-\clearpage
-
-\subsubsection{Miscellaneous graphs}
-
-\label{Miscellaneous graphs}
-
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
-
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\showgraph{build/plot-action-distribution}{Distribution of rules per
-action}{Distribution of rules per action}
-
-\clearpage
-
-\subsubsection{Summary}
-
-The first collection of graphs are about
-parser scalability, showing the linear relationship between execution time
-and input size.  \sectionref{rule ordering graphs} demonstrates the effect
-of rule ordering on execution time, and the unexpected consequences of
-specific inputs.  The necessity of caching compiled \regexes{} is attested
-to by the third group of graphs, where the difference between caching and
-not caching is staggering.  The penultimate section contains a breakdown of
-the rule hits accumulated during a single test run of the parser.
-Miscellaneous graphs expected to be useful are collected in the final
-section.
+\clearpage{}

### 2008-07-25 doc/logparser.tex
Lots of improvements to the efficiency section; lots more to come.

@@
 
-% Sort numbers where there are multiple citations.  does not appear to have
-% any effect (probably clashes with varioref or hyperref), though it does
-% reduce the space between numbers.
+% Sort numbers where there are multiple citations.  Does not appear to have
+% the expected effect (probably clashes with varioref or hyperref), though
+% it does reduce the space between numbers.
 \usepackage{cite}
@@
 \SVN$Id$
-\SVNId{}
+\begin{center}\SVNId{}\end{center}
 
@@
 
-\newpage
-
-XXX NEW LAYOUT\@:
-
-\begin{verbatim}
-Introduction
-Background
-State of the Art Review
-Parser Architecture
-Postfix Parser Implementation
-    Database Schema
-    Rules
-    Actions
-    Complications
-    Limitations and Improvements
-Results - BETTER TITLE?
-    Coverage
-    Efficiency
-Conclusion
-References
-Glossary
-Acronyms
-\end{verbatim}
+XXX CHECK THE CAPTIONS ON ALL GRAPHS AND TABLES --- SOME ARE AWFUL\@.
 

### 2008-07-26 doc/logparser-results.tex
Lots of improvements to the efficiency section - I'm happy with most of it now. I improved the graphs too, with better captions and labels.

@@
 XXX CHANGE normal TO optimal IN ORDERING STUFF, INCLUDING IN GRAPH CAPTIONS
-AND AXES LABELS\@.
+AND AXIS LABELS\@.
 
@@
 
-XXX THIS SECTION NEEDS TO BE REWRITTEN\@.
-
 Parsing efficiency is an obvious concern when the parser routinely needs to
@@
 in testing this parser accepts approximately 10,000 mails for 700 users a
-day, and on average it produces 75MB log files containing 300,000 log lines
---- large scale mail servers would have much larger log files on a daily
-basis.
+day; as expected there are far more mails received on weekdays than at
+weekends.   Median log file size is 50MB, containing 285,000 log lines ---
+large scale mail servers would have much larger log files.  Note that
+\graphref{Mails received per day} and \tableref{Number of mails received
+per day: statistics} show the number of mails received by \SMTP{} only; in
+particular the mail loops noticeable in other graphs do not contribute to
+these figures.
 
-\showgraph{build/plot-mails-received}{Mails received per day}{Mails
-received per day}
-
-As expected there are far more mails received on weekdays than at weekends.
-Note that this graph and the table below show the number of mails received
-by \SMTP{} only; in particular the mail loops noticeable in other graphs
-do not contribute to these figures.
+\showgraph{build/plot-mails-received}{Number of mails received per
+day}{Mails received per day}
 
 \showtable{build/mails-received-include-for-graph}{Number of mails received
-per day: statistics}{Number of mails received per day: statistics}
-
-\showgraph{build/plot-action-distribution}{Distribution of rules per
-action}{Distribution of rules per action}
+per day}{Number of mails received per day: statistics}
 
@@
 log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
-parsed 10 times, the results of the first first run discarded, and the
+parsed 10 times, the results from the first first run discarded, and the
 execution time for the remaining 9 runs averaged.  The results from the
-first run are discarded for two reasons:
-
-\begin{enumerate}
-
-    \item The execution time will be higher because the log file must be
-        read from disk, whereas for subsequent runs the log file will be
-        cached in memory by the operating system.
-
-    \item The execution time will also be higher because the rule ordering
-        will be sub-optimal compared to subsequent runs.
-
-\end{enumerate}
-
-Saving results to the database was disabled for the test runs, as that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of the parser rather than the speed of the database and the disks
-the database is stored on.
-
-\newpage
+first run were discarded because because the rule ordering will be
+sub-optimal compared to subsequent runs --- XXX MAYBE I SHOULDN'T DISCARD
+THE FIRST RESULT\@?  Saving results to the database was disabled for the
+test runs, as that dominates the run time of the program, and the tests are
+aimed at measuring the speed of the parser rather than the speed of the
+database and the disks the database is stored on.
 
@@
 input size: does it scale linearly, polynomially, or exponentially?
-\Graphref{execution time vs file size vs number of lines graph} shows the
-execution time in seconds, file size in MB, and tens of thousands of log
-lines per log file.  All three lines run roughly in parallel, giving the
-impression that the algorithm scales linearly with input size.  This
-impression is borne out by \graphref{execution time vs file size vs number
-lines factor} which plots both the ratio of file size vs execution time,
-and the ratio of number of log lines vs execution time (higher is better in
-both cases); \tableref{execution time vs file size vs number lines factor
-table} shows the ratios for different groups of log files.  As the
-reader can see the ratios are quite tightly banded, showing that the
-algorithm scales linearly: the much larger log files between points 60 and
-70 on the X axis in \graphref{execution time vs file size vs number of
-lines graph} actually cause the ratio to increase (i.e.\ improve), rather
-than decrease.  The strange behaviour where larger log files are parsed
-more efficiently is explained fully in \sectionref{Explaining the peaks in
-log file size}.
-
-\showgraph{build/plot-normal-filesize-numlines}{Execution time vs file
-size vs number of lines}{execution time vs file size vs number of lines
-graph}
-
-The Y axis in \graphref{execution time vs file size vs number of lines
-graph} represents the following:
-
-\begin{enumerate}
-
-    \item The time required, in seconds, to parse the log file.
-
-    \item The size of the log file, in megabytes.
-
-    \item The number of lines in the log file, divided by 10000.
-
-\end{enumerate}
-
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to execution time}{execution time vs file size vs
-number lines factor}
+\Graphref{parsing time vs file size vs number of lines graph} shows the
+parsing time in seconds, file size in MB, and tens of thousands of log
+lines for each of the \numberOFlogFILES{} log files.  All three lines run
+roughly in parallel, giving the impression that the algorithm scales
+linearly with input size.  This impression is borne out by
+\graphref{parsing time vs file size vs number lines factor} which plots
+both the ratio of file size vs parsing time, and the ratio of number of log
+lines vs parsing time (higher is better in both cases); \tableref{parsing
+time vs file size vs number lines factor table} shows the same ratios for
+different groups of log files.  As the reader can see the ratios are quite
+tightly banded, providing empirical evidence that the algorithm scales
+linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
+despite their larger than average size (shown in \graphref{parsing time vs
+file size vs number of lines graph}); this strange behaviour where larger
+log files are parsed more efficiently than smaller is explained fully in
+\sectionref{Explaining the peaks in log file size}.
 
 \showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size \& number of lines to execution time: statistics}{execution time vs
-file size vs number lines factor table}
+size and number of lines to parsing time}{parsing time vs file size vs
+number lines factor table}
+
+\showgraph{build/plot-normal-filesize-numlines}{Parsing time, file size,
+and number of lines}{parsing time vs file size vs number of lines graph}
+
+\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file size
+and number of lines to parsing time}{parsing time vs file size vs number
+lines factor}
 
@@
 
-\Graphref{execution time vs file size vs number of lines graph} shows log
+\Graphref{parsing time vs file size vs number of lines graph} shows log
 file size varying between 30--100MB, except for log files 22 (200MB) and
@@
 \tab{}\texttt{$\backslash$username, username@domain} \newline This
-instructs Postfix to deliver the mail to the local user, and also forward
-it to the remote address; this is generally not a problem except that the
+instructs Postfix to deliver the mail to the local user, and also forward a
+copy to the remote address; this is generally not a problem except that the
 remote address in this case is the address the mail was originally sent to,
@@
 to the current address it is bounced back to the sender with the error
-message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
+message \newline{} \tab{} \texttt{mail forwarding loop for
+username@domain}\newline{}  Ordinarily this
 works well, but unfortunately in this case the user noticed they had not
@@
 split across the other 90\% of the rules (as shown in \graphref{rule hits
-graph}).  Assuming that the distribution of log lines is reasonably steady
-over time, parser efficiency should benefit from trying more frequently
-matching rules before those which match less frequently.  To test this
-hypothesis three full test runs were performed with different rule
-orderings:
+graph}).  Assuming that the distribution of log lines is reasonably
+consistent over time, parser efficiency should benefit from trying more
+frequently matching rules before those which match less frequently.  To
+test this hypothesis three full test runs were performed with different
+rule orderings:
 
@@
         set.  The rules will be shuffled once before use and will retain
-        that ordering for the entirety of the log file.  Note that the
-        ordering will change every time the parser is executed, so 10
-        different orderings will be generated for each log file in the test
-        run.
+        that ordering until the parser exits.  Note that the ordering will
+        change every time the parser is executed, so 10 different rule
+        orderings will be generated for each log file in the test run.
 
@@
 
-\Graphref{Execution time relative to
-shuffled ordering graph} shows the execution time of normal and reverse orderings
-relative to shuffled ordering; the mean increases for different groupings
-of log files are given in \tableref{Execution time relative to shuffled
-ordering table}.  Overall this optimisation provides a modest but worthwhile
+\Graphref{Parsing time relative to shuffled ordering graph} shows the
+execution time of normal and reverse orderings relative to shuffled
+ordering; the mean relative execution times for different groupings of log
+files are given in \tableref{Parsing time relative to shuffled ordering
+table}.  Overall this optimisation provides a modest but worthwhile
 performance increase of approximately 10\%, for a small investment in time
@@
 
-\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Execution
-time of normal and reverse orderings relative to shuffled
-ordering}{Execution time relative to shuffled ordering graph}
+\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Parsing time
+relative to shuffled ordering}{Parsing time relative to shuffled ordering
+graph}
 
-\showtable{build/stats-normal-shuffle-reverse-include}{Execution time
-relative to shuffled ordering}{Execution time relative to shuffled ordering
+\showtable{build/stats-normal-shuffle-reverse-include}{Parsing time
+relative to shuffled ordering}{Parsing time relative to shuffled ordering
 table}
 
-XXX IMPROVE THE REMAINDER OF THIS SECTION\@; CLEARLY SEPARATE THE
-CHARACTERISTICS AND CLARIFY THEM\@.
-
-Different rule orderings have less effect when parsing log files 22 and
-62--68, due to the different distribution of log lines in those log files
-(see \sectionref{Explaining the peaks in log file size}).  The vast
-majority of log lines when a mail loop occurs are from Postfix components
-which have a small number of rules associated with them, whereas in general
-\daemon{smtpd} adds the majority of log lines, and also has the highest
-number of rules.  \daemon{smtpd} log lines are distributed across rules
-much more evenly than the log lines of \daemon{qmgr}, \daemon{local},
-\daemon{cleanup} or \daemon{pickup}, so the average number of rules
-required to parse a \daemon{smtpd} log line is much higher than the average
-number required to parse other log lines.
-
-These two characteristics combine to reduce the average number of rules
-required to parse a log line when there is a mail loop.  The ratio of
-number of lines versus execution time in \graphref{execution time vs file
-size vs number lines factor} shows a noticeable improvement at log files 22
-and 62--68, where a higher ratio means more log lines parsed per unit of
-execution time.
-
-When the rule ordering is reversed the majority of log lines generated by a
-mail loop will be parsed with very few rules, whereas without a mail loop
-the majority of log lines require a large number of rules; this leads to a
-noticeable drop in the average time required to parse a log line, as shown
-in \graphref{Execution time relative to shuffled ordering graph}.  The
-number of rules which need to be consulted when the ordering is shuffled
-varies between the optimal and the pessimal, and the performance varies
-proportionally.
+Differences in rule ordering have less effect on parsing time when parsing
+log files 22 and 62--68, due to the different distribution of log lines in
+those log files (see \sectionref{Explaining the peaks in log file size}).
+When a mail loop occurs the vast majority of log lines are from Postfix
+components which have a small number of rules associated with them, whereas
+in general the majority of log lines are from Postfix components which have
+a high number of associated rules; this results in the average number of
+rules required to parse a log line from log files 22 or 62--68 being
+considerably lower than usual.  The ratio of number of lines versus parsing
+time in \graphref{parsing time vs file size vs number lines factor} shows a
+noticeable improvement at log files 22 and 62--68, where a higher ratio
+means more log lines parsed per unit of time; similarly \graphref{Parsing
+time relative to shuffled ordering graph} shows much less difference
+between the orderings at log files 22 and 62--68.
 
@@
 
+\clearpage{}
+
 \subsubsection{Caching compiled regexs}
@@
 
-Before Perl attempts to match a \regex{} against a piece of text, the
-original \regex{} is compiled into an internal representation and optimised
-to improve the speed of matching, but this compilation and optimisation
-takes CPU time; far more CPU time, in fact, than the actual matching takes
-in many cases.  Perl automatically caches static \regexes{} at compile
-time, but \regexes{} loaded dynamically while the program need to be
-explicitly compiled and cached.
+Before the perl interpreter attempts to match a \regex{} against a piece of
+text, the \regex{} is compiled into an internal representation and
+optimised to improve the speed of matching.  This compilation and
+optimisation takes CPU time ---in many cases it takes far more CPU time
+than the actual matching.  If the interpreter is certain that a \regex{}
+will not change it will automatically cache the compilation results for
+later use.   The results of compiling a dynamically generated \regex{} can
+be cached and used in preference to the original \regex{}, but it is the
+responsibility of the programmer to do this.  The Postfix parser loads the
+rules it uses from a database, and thus each rule's \regex{} is compiled
+and the result cached to reduce parsing time.
 
-The following graphs show the impact that not caching compiled \regexes{}
-has on parser performance: on typical log files the execution time when not
-caching compiled \regexes{} is 500--600\% of the execution time when
-caching; reversing the perspective shows that cached execution time is
-merely 17--20\% of non-cached execution time.  Caching compiled \regexes{}
-is probably the single most effective optimisation possible in the parser's
-implementation; given that it only required sixteen extra lines of
-reasonably simple code,\footnote{The sixteen lines of code breaks down as
-follows: three lines to add caching, eight lines of error checking and
-reporting, three lines to optionally disable caching for debugging and
-performance measurement, and two lines of comments.} the investment in time
-was minimal.
+\Graphref{Increase in parsing time when not caching compiled regexes graph}
+shows the impact that not caching compiled \regexes{} has on parser
+performance: on typical log files the parsing time when not caching
+compiled \regexes{} is 400--600\% of the parsing time when caching.
+Caching compiled \regexes{} is probably the single most effective
+optimisation possible in the parser's implementation, and was quite simple
+to implement.
 
-\showgraph{build/plot-cached-discarded}{Regex: cached vs
-discarded}{normal regex vs discard regex}
+\showgraph{build/plot-cached-discarded-factor}{Increase in parsing time
+when not caching compiled regexes}{Increase in parsing time when not
+caching compiled regexes graph}
 
-\showgraph{build/plot-cached-discarded-factor}{Regex caching: percentage
-execution time increase}{normal regex vs discarded regex factor}
+Two large dips can be seen in \graphref{Increase in parsing time when not
+caching compiled regexes graph} at log files 22 and 62--68, corresponding
+to the spikes in log file size in \graphref{parsing time vs file size vs
+number of lines graph}.  The distribution of log lines across rules when
+there is a mail loop is much different to the norm, and the average number
+of rules consulted per log line is much lower: this results in far fewer
+\regex{} compilations per line when there is a mail loop, and a
+corresponding decrease in the average parsing time for a single log line.
+The outcome is that caching compiled \regexes{} is proportionally less
+important when the log file contents were created by a mail loop.  The
+increases in parsing time when not caching compiled \regexes{} for
+different groups of log files are summarised in \tableref{Increase in
+parsing time when not caching compiled regexes table}.
 
-Two large dips can be seen in \graphref{normal regex vs discarded
-regex factor} at log files 22 and 62--68, corresponding to the spikes in
-log file size in \graphref{execution time vs file size vs number
-of lines graph} and \graphref{normal regex vs discard regex}.  The
-reason for the anomalous log files has already been explained in
-\sectionref{Explaining the peaks in log file size}.
-
-The distribution of log lines across rules when there is a mail loop is
-much different to the norm and the average number of rules consulted per
-log line is much lower; this results in far fewer \regex{} compilations per
-line than when there is not a mail loop, and a correspondingly decreased
-execution time.  The unexpected outcome is that caching \regexes{} is
-proportionally less important when the log file contents were created by a
-mail loop.  The increases in execution time when not caching \regexes{} for
-different combinations of log files are summarised in the table below:
-
-XXX THIS IS AN AWFUL CAPTION\@.
-
-\showtable{build/stats-cached-discarded-include-for-graph}
-{Regex caching/discarding with different groups of log files}
-{Regex caching/discarding with different groups of log files}
+\showtable{build/stats-cached-discarded-include-for-graph} {Increase in
+parsing time when not caching compiled regexes} {Increase in parsing time
+when not caching compiled regexes table}
 
@@
 
+XXX FIGURE OUT WHERE TO PUT THESE GRAPHS\@; DO I NEED ALL THREE\@?
+
+\showgraph{build/plot-action-distribution}{Distribution of rules per
+action}{Distribution of rules per action}
+
 The number of hits per rule is quite unevenly spread, resembling a Power
@@
 
-As \graphref{rule hits graph} is quite difficult to read it has
-been separated into three sections: low, middle and high.
+As \graphref{rule hits graph} is quite difficult to read it has been
+separated into three sections: low, middle and high.
 

### 2008-07-28 doc/logparser-implementation.tex
Wrote minimum ruleset versus maximum ruleset section - largely taken from the SGAI paper.

@@
 
-XXX DUNNO IF THIS SECTION SHOULD BE HERE\@.  QUEUEID VS PID ISN'T TOO
+XXX DUNNO IF THIS SECTION SHOULD BE HERE\@.  QUEUEID VS PID IS NOT TOO
 IMPORTANT\@; CONNECTION REUSE COULD EASILY BE MOVED TO ADDITIONAL
@@
 
-XXX CONSIDER REMOVING PART TWO OF THE FLOW CHART\@: I DON'T THINK THAT
+XXX CONSIDER REMOVING PART TWO OF THE FLOW CHART\@: I DO NOT THINK THAT
 RE-INJECTION IS IMPORTANT ENOUGH TO WARRANT IT\@.  THE EXPLANATION OF

### 2008-07-28 doc/logparser-other-parsers-review.tex
Wrote minimum ruleset versus maximum ruleset section - largely taken from the SGAI paper.

@@
 (minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
+attempts were made.}  As parsing failed it was not possible to review the
 reports Logrep can generate (available in \HTML{} only), nor to examine the

### 2008-07-28 doc/logparser-results.tex
Renamed normal-vs-smaller graph to full-ruleset-vs-minimum-ruleset.
Wrote minimum ruleset versus maximum ruleset section - largely taken from the SGAI paper.

@@
 XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-XXX DO SOMETHING WITH NORMAL VS SMALLER GRAPH\@.
 
@@
 first run were discarded because because the rule ordering will be
-sub-optimal compared to subsequent runs --- XXX MAYBE I SHOULDN'T DISCARD
+sub-optimal compared to subsequent runs --- XXX MAYBE I SHOULD NOT DISCARD
 THE FIRST RESULT\@?  Saving results to the database was disabled for the
@@
 
-XXX WRITE THIS SECTION\@.
+How any architecture scales as the number of rules increases is important,
+but it is particularly important in this architecture because it is
+expected that the typical parser will have a large number of rules.  There
+are \numberOFrules{} rules in the full Postfix ruleset, whereas the minimum
+number of rules required to parse the \numberOFlogFILES{} log files used
+when generating the results in this document is \numberOFrulesMINIMUM{},
+\numberOFrulesMINIMUMpercentage{} of the full ruleset.  The full ruleset is
+larger because the Postfix parser is tested with \numberOFlogFILESall{} log
+files, to improve its parsing and ensure it correctly parses logs generated
+by later Postfix versions.  A second set of statistics was generated using
+the minimum ruleset and compared to the statistics generated using the full
+ruleset: the percentage execution increase when using the full ruleset
+instead of the minimal ruleset for normal, shuffled and reversed orderings
+is shown in \graphref{Percentage parsing time increase of maximum ruleset
+over minimum ruleset}.
 
-\showgraph{build/plot-normal-vs-smaller}{NORMAL VS SMALLER}{NORMAL VS
-SMALLER}
+\showgraph{build/plot-full-ruleset-vs-minimum-ruleset}{Percentage parsing
+time increase of maximum ruleset over minimum ruleset}{Percentage parsing
+time increase of maximum ruleset over minimum ruleset}
 
-\clearpage{}
+It is clear from \graphref{Percentage parsing time increase of maximum
+ruleset over minimum ruleset} that the increased number of rules noticeably
+increases parsing time with reverse ordering, and to a lesser extent with
+shuffled ordering.  The normal ordering (where are the most frequently
+matching rules are tried first) shows a mean increase of
+\input{build/full-ruleset-vs-minimum-ruleset-include.tex} in parsing time
+for a \numberOFrulesMAXIMUMpercentage{} increase in the number of rules.
+Log files 22 and 62--68 show much smaller increases for shuffled and
+reversed ordering than other log files do, because the majority of log
+lines in those log files are parsed by Postfix components with a small
+number of parsing rules, so removing unnecessary rules has little effect on
+the total number of rules used.  These results show that the architecture
+scales very well as the number of rules increases, and that optimally
+sorting the rules is an important optimisation contributing to this
+scalability.
 

### 2008-07-29 papers/sgai-2008/paper.tex
Removed contractions like don't. Updated graphs.

@@
 from the framework simplifies both framework and actions: the framework
-provides services to the actions, but doesn't need to perform any tasks
+provides services to the actions, but does not need to perform any tasks
 specific to the input being parsed; actions benefit from having services
@@
 each parser will need to write the actions it requires from scratch, unless
-it's extending an existing parser.  In the Postfix parser developed for
+it is extending an existing parser.  In the Postfix parser developed for
 this project there are \numberOFactions{} actions and \numberOFrules{}
@@
 highest number of associated rules is \texttt{REJECTION}, the action which
-handles rejections; it's followed by \texttt{SAVE\_BY\_QUEUEID}, the action
-responsible for handling informative inputs: these are log lines that
-supplement the data gathered from other log lines (such as rejections), but
-that don't by themselves represent interesting events.  The third most
-common action is, perhaps surprisingly, \texttt{IGNORE}: there are many log
-lines which are not of interest to the parser, e.g.\ warning messages about
-DNS mismatches.  The parser warns about every unparsed log line, to alert
-the user that they need to alter or extend their ruleset; the
-\texttt{IGNORE} action does nothing when executed, allowing uninteresting
-log lines to be parsed without causing any effects.  The remaining actions
-have only one or two associated rules: in some cases this is because there
-will only ever be one type of log line for that action, e.g.\ all log lines
-showing that a remote client has connected are matched by a single rule and
-handled by the \texttt{CONNECT} action; in other cases the action is
-required to address a deficiency in the log files, or a complication which
-arises during parsing.
+handles rejections; it is followed by \texttt{SAVE\_BY\_QUEUEID}, the
+action responsible for handling informative inputs: these are log lines
+that supplement the data gathered from other log lines (such as
+rejections), but that do not by themselves represent interesting events.
+The third most common action is, perhaps surprisingly, \texttt{IGNORE}:
+there are many log lines which are not of interest to the parser, e.g.\
+warning messages about DNS mismatches.  The parser warns about every
+unparsed log line, to alert the user that they need to alter or extend
+their ruleset; the \texttt{IGNORE} action does nothing when executed,
+allowing uninteresting log lines to be parsed without causing any effects.
+The remaining actions have only one or two associated rules: in some cases
+this is because there will only ever be one type of log line for that
+action, e.g.\ all log lines showing that a remote client has connected are
+matched by a single rule and handled by the \texttt{CONNECT} action; in
+other cases the action is required to address a deficiency in the log
+files, or a complication which arises during parsing.
 
@@
     \item [program] The Postfix program whose log lines the rule should be
-        applied to.  This avoids needlessly trying rules that won't match
-        the log line, or worse, might match unintentionally.  Rules whose
-        program is \texttt{*} will be tried against any log lines not
+        applied to.  This avoids needlessly trying rules that will not
+        match the log line, or worse, might match unintentionally.  Rules
+        whose program is \texttt{*} will be tried against any log lines not
         parsed by program specific rules, allowing generic rules to be
@@
     \item [result\_data] Sometimes rules need to supply a piece of data
-        which isn't present in the log line: e.g.\ setting
+        which is not present in the log line: e.g.\ setting
         \texttt{smtp\_code} when mail is accepted.  The format is the same
@@
         the maximum size the server accepts, the data structure for that
-        mail must be discarded because it won't have all the data required
-        to save it to the database.
+        mail must be discarded because it will not have all the data
+        required to save it to the database.
 
@@
 
-% Don't reformat this!
-% XXX CONSIDER USING A TABLE ENVIRONMENT; IT DIDN'T LOOK GREAT BEFORE.
+% Do not reformat this!
+% XXX CONSIDER USING A TABLE ENVIRONMENT; IT DID NOT LOOK GREAT BEFORE.
 \begin{tabular}[]{ll}
@@
 
-Figure~\refwithpage{percentage increase of shuffled and reversed over
-normal} shows the percentage increase of execution times for shuffled and
-reversed rules.  Overall this optimisation provides a modest but worthwhile
-performance increase of over 10\%, for a small investment in time and
-programming.
-\showgraph{build/plot-normal-shuffle-reverse-factor}{Percentage increase of
-execution time for different rule orderings}{percentage increase of
-shuffled and reversed over normal}
+Figure~\refwithpage{Parsing time relative to shuffled} shows the percentage
+increase of execution times for shuffled and reversed rules.  Overall this
+optimisation provides a modest but worthwhile performance increase of over
+10\%, for a small investment in time and programming.
+\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Parsing time
+relative to shuffled}{Parsing time relative to shuffled}
 
@@
 
-The dips at log files 22 and 62--68 in fig.~\refwithpage{percentage
-increase of shuffled and reversed over normal} correspond to peaks in log
-file size in fig.~\refwithpage{Execution time, file size, and number of
-lines}, and peaks in fig.~\refwithpage{execution time vs file size vs
-number lines factor}, where a peak means that performance is better.  The
-explanation for this took some time to arrive at, but it turns out to be
-reasonably simple.  The large log files in question were caused by a mail
-forwarding loop, where the distribution of log lines is quite different to
-normal, resulting in different performance characteristics.  The vast
-majority of log lines when a mail loop occurs are from Postfix components
-which have a small number of rules associated with them, whereas in general
-the majority of log lines are from the Postfix component (\daemon{smtpd})
-with the highest number of rules.  Furthermore in normal log files the
-majority of log lines are distributed much more evenly across rules than
-the majority of log lines during a mail loop.
+The dips at log files 22 and 62--68 in fig.~\refwithpage{Parsing time
+relative to shuffled} correspond to peaks in log file size in
+fig.~\refwithpage{Execution time, file size, and number of lines}, and
+peaks in fig.~\refwithpage{execution time vs file size vs number lines
+factor}, where a peak means that performance is better.  The explanation
+for this took some time to arrive at, but it turns out to be reasonably
+simple.  The large log files in question were caused by a mail forwarding
+loop, where the distribution of log lines is quite different to normal,
+resulting in different performance characteristics.  The vast majority of
+log lines when a mail loop occurs are from Postfix components which have a
+small number of rules associated with them, whereas in general the majority
+of log lines are from the Postfix component (\daemon{smtpd}) with the
+highest number of rules.  Furthermore in normal log files the majority of
+log lines are distributed much more evenly across rules than the majority
+of log lines during a mail loop.
 
@@
 peaks in fig.~\refwithpage{execution time vs file size vs number lines
-factor} and troughs in fig.~\refwithpage{percentage increase of shuffled
-and reversed over normal}.  When the rule ordering is reversed the majority
-of log lines generated by a mail loop will be parsed using very few rules,
-because there are very few rules for the Postfix components producing the
-log lines.  Without a mail loop the Postfix components producing the
-majority of log lines have a large number of rules associated with them,
-and reverse ordering deliberately starts with the least likely rules and
-moves towards the most likely.  This leads to a noticeable drop in the
-average time required to parse a log line during a mail loop, as shown in
-fig.~\refwithpage{percentage increase of shuffled and reversed over
-normal}.  The number of rules which need to be consulted when the ordering
-is shuffled varies between the optimal and the pessimal, with the
-performance varying proportionally.
+factor} and troughs in fig.~\refwithpage{Parsing time relative to
+shuffled}.  When the rule ordering is reversed the majority of log lines
+generated by a mail loop will be parsed using very few rules, because there
+are very few rules for the Postfix components producing the log lines.
+Without a mail loop the Postfix components producing the majority of log
+lines have a large number of rules associated with them, and reverse
+ordering deliberately starts with the least likely rules and moves towards
+the most likely.  This leads to a noticeable drop in the average time
+required to parse a log line during a mail loop, as shown in
+fig.~\refwithpage{Parsing time relative to shuffled}.  The number of rules
+which need to be consulted when the ordering is shuffled varies between the
+optimal and the pessimal, with the performance varying proportionally.
 

### 2008-07-29 doc/logparser-results.tex
Improvements copied from the thesis. Beginnings of a timing data section. Wrote a short piece about IBL. Some corrections from Carl. Tweaked graph sizes and spacing between floats to reduce page count.

@@
 aimed at measuring the speed of the parser rather than the speed of the
-database and the disks the database is stored on.
+database and the disks the database is stored on.  The computer used for
+test runs was a Dell Optiplex 745, equipped with:
 
-\subsubsection{Algorithmic complexity}
+\begin{tabular}[]{ll}
 
-An important property of a parser is how execution time scales relative to
+CPU         & One dual core 2.40GHz Intel(R) Core(TM)2 CPU, \\
+            & with 32KB L1 cache and 4MB L2 cache. \\
+RAM         & 2GB 667 MHz DDR RAM. \\
+Hard disk   & One XXX 232GB SATA disk. \\
+
+\end{tabular}
+
+In as far as possible the computer was dedicated to the task of gathering
+statistics from test runs, and was not used for any other purpose while
+test runs were ongoing.
+
+\subsubsection{Architecture Scalability: Input Size}
+
+An important property of a parser is how parsing time scales relative to
 input size: does it scale linearly, polynomially, or exponentially?

### 2008-07-29 papers/sgai-2008/paper.tex
Improvements copied from the thesis. Beginnings of a timing data section. Wrote a short piece about IBL. Some corrections from Carl. Tweaked graph sizes and spacing between floats to reduce page count.

@@
 
+\setlength{\floatsep}{0pt}
+\setlength{\intextsep}{0pt}
+\setlength{\textfloatsep}{0pt}
+
 % Nicer citation lists - smaller spaces.
@@
 \newcommand{\showgraph}[3]{%
-    \begin{figure}[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
-    \end{figure}
+    \begin{figure}[hbt!]%
+        \caption{#2}\label{#3}%
+        \includegraphics{#1}%
+    \end{figure}%
 }
@@
 
+\newcommand{\refwithlabel}[2]{%
+    #1~\vref{#2}%
+}
+% table references, for consistent formatting.
+\newcommand{\graphref}[1]{%
+    \refwithlabel{fig.}{#1}%
+}
+\newcommand{\Graphref}[1]{%
+    \refwithlabel{Figure}{#1}%
+}
+
 % A command to format a Postfix daemon's name
@@
     115%
+}
+
+\newcommand{\numberOFlogFILESall}[0]{%
+    522%
 }
@@
 a specific type, and only that type (e.g.\ rejection by a specific DNSBL).
-This architecture is ideally suited to parsing inputs where the input is
-not fully understood or does not follow a fixed grammar: the architecture
-warns about unparsed inputs and other errors, but continues on parsing as
-best it can, allowing the developer of a new parser to decide which
-deficiencies are most important and require attention first, rather than
-being forced to fix the first error which arises.
+It is an accepted standard to separate the parsing procedure from the
+declarative grammar it operates with.  Part of the novelty here is that the
+grammars (rules and actions) are themselves procedures.  This architecture
+is ideally suited to parsing inputs where the input is not fully understood
+or does not follow a traditional grammar: the architecture warns about
+unparsed inputs and other errors, but continues on parsing as best it can,
+allowing the developer of a new parser to decide which deficiencies are
+most important and require attention first, rather than being forced to fix
+the first error which arises.
 
@@
 particular category, e.g.\ rejections.  The actions are parser-specific:
-each parser will need to write the actions it requires from scratch, unless
-it is extending an existing parser.  In the Postfix parser developed for
+each parser author will need to write the actions required from scratch,
+unless extending an existing parser.  In the Postfix parser developed for
 this project there are \numberOFactions{} actions and \numberOFrules{}
 rules; the distribution of rules to actions is very skewed, as shown in
-Fig.~\refwithpage{Distribution of rules per action}.
+Fig.~\refwithpage{Distribution of rules per action}.  The distribution is
+skewed for the same reason the architecture is so successful in enabling
+users to parse new log lines: most new log lines can be parsed by a
+combination of a new rule and an existing action.
 \showgraph{build/plot-action-distribution}{Distribution of rules per
-action}{Distribution of rules per action} The distribution is skewed for
-the same reason the architecture is so successful in enabling users to
-parse new log lines: most new log lines can be parsed by a combination of a
-new rule and an existing action.  Unsurprisingly the action with the
-highest number of associated rules is \texttt{REJECTION}, the action which
-handles rejections; it is followed by \texttt{SAVE\_BY\_QUEUEID}, the
-action responsible for handling informative inputs: these are log lines
-that supplement the data gathered from other log lines (such as
-rejections), but that do not by themselves represent interesting events.
-The third most common action is, perhaps surprisingly, \texttt{IGNORE}:
-there are many log lines which are not of interest to the parser, e.g.\
-warning messages about DNS mismatches.  The parser warns about every
-unparsed log line, to alert the user that they need to alter or extend
-their ruleset; the \texttt{IGNORE} action does nothing when executed,
+action}{Distribution of rules per action} Unsurprisingly the action with
+the highest number of associated rules is \texttt{REJECTION}, the action
+which handles Postfix rejecting a mail; it is followed by
+\texttt{SAVE\_BY\_QUEUEID}, the action responsible for handling informative
+inputs: these are log lines that supplement the data gathered from other
+log lines (such as rejections), but that do not by themselves represent
+interesting events.  The third most common action is, perhaps surprisingly,
+\texttt{IGNORE}: the \texttt{IGNORE} action does nothing when executed,
 allowing uninteresting log lines to be parsed without causing any effects.
-The remaining actions have only one or two associated rules: in some cases
-this is because there will only ever be one type of log line for that
-action, e.g.\ all log lines showing that a remote client has connected are
-matched by a single rule and handled by the \texttt{CONNECT} action; in
-other cases the action is required to address a deficiency in the log
-files, or a complication which arises during parsing.
+The parser warns about every unparsed log line, to alert the user that they
+need to alter or extend their ruleset; the Postfix parser successfully
+parses every input line in the \numberOFlogFILESall{} log files it is
+currently being tested with.  The remaining actions have only one or two
+associated rules: in some cases this is because there will only ever be one
+type of log line for that action, e.g.\ all log lines showing that a remote
+client has connected are matched by a single rule and handled by the
+\texttt{CONNECT} action; in other cases the action is required to address a
+deficiency in the log files, or a complication which arises during parsing.
 
@@
 the strengths of this architecture: instead of writing one monolithic
-function which must be modified to support new behaviour or resolve
-challenges, with all the attendant risks of adversely affecting the
-existing parser, when a new requirement arises an independent action can be
-written to satisfy it.  In some cases the new action will require the
-cooperation of other actions, e.g.\ to set or check a flag; there is a
-possibility of introducing failure by modifying existing actions, but
-modifications will be smaller and occur less frequently than with a
-monolithic architecture, thus failures should be less likely, and easier to
-test for and diagnose.  The parser can be written in an object oriented
-style, allowing sub-classes to extend or override actions in addition to
-adding new actions; because each action is an independent function, the
-sub-class need only modify the action it is interested in, rather than
-reproducing large chunks of functionality.
+function which must be modified to support new behaviour, with all the
+attendant risks of adversely affecting the existing parser, when a new
+requirement arises an independent action can be written to satisfy it.  In
+some cases the new action will require the cooperation of other actions,
+e.g.\ to set or check a flag; there is a possibility of introducing failure
+by modifying existing actions, but modifications will be smaller and occur
+less frequently than with a monolithic architecture, thus failures should
+be less likely, and easier to test for and diagnose.  The parser can be
+written in an object oriented style, allowing sub-classes to extend or
+override actions in addition to adding new actions; because each action is
+an independent function, the sub-class need only modify the action it is
+interested in, rather than reproducing large chunks of functionality.
 
@@
 
-Rules are responsible for categorising inputs, specifying both the data to
-be extracted from each log line and the action that should be invoked.  The
-Postfix log parser stores the rules in the same SQL database that the
-results are stored in, removing any doubt as to which set of rules was used
-to produce a set of results; other implementations are free to store the
-rules in whatever fashion suits their needs.  Each rule defines several
-attributes:
+Rules categorise inputs, specifying both the data to be extracted from each
+log line and the action to invoke.  The Postfix log parser stores the rules
+in the same SQL database that the results are stored in, removing any doubt
+as to which set of rules was used to produce a set of results; other
+implementations are free to store the rules in whatever fashion suits their
+needs.
+
+XXX ADD SOMETHING ABOUT IMPLEMENTATIONS BEING FREE TO ADD OR REMOVE
+ATTRIBUTES\@.
 
@@
 
-    \item [program] The Postfix program whose log lines the rule should be
-        applied to.  This avoids needlessly trying rules that will not
+    \item [program] The Postfix program whose log lines the regex should be
+        applied to.  This avoids needlessly trying regexes that will not
         match the log line, or worse, might match unintentionally.  Rules
@@
 
-The rules used by the Postfix log file parser define several additional
-attributes; each implementation is free to add any additional attributes it
-requires.
-
 Adding new rules is simple: simply define all the required attributes and
@@
 SLCT~\cite{slct-paper} to aid in producing regexes to parse previously
-unparsed inputs.
+unparsed inputs.  Storing the rules separately allows other approaches to
+rule management to be used, e.g.\ instead of manually adding new rules when
+unparsed log lines are seen, machine learning techniques could be used to
+automatically generate new rules, based on the existing rules and input
+lines.  If this approach was taken the choice of machine learning technique
+would be constrained by the size of typical data sets: the
+\numberOFlogFILES{} log files used to generate statistics contain
+\numberOFlogLINEShuman{} log lines.  Techniques which require the full data
+set when training would be impractical; Instance Based Learning
+(IBL)~\cite{instance-based-learning} techniques which automatically
+determine which inputs from the training set are valuable and which can be
+discarded would probably reduce the data required to a manageable size.
+This avenue of research and development has not been pursued by the author,
+but could easily be independently undertaken.
 
@@
 Parsing efficiency is an obvious concern when the parser routinely needs to
-deal with 75 MB log files containing 300,000 log lines (generated daily on
-a mail server handling mail for approximately 700 users --- large scale
-mail servers would produce much larger log files on a daily basis).  When
-generating the data for the graphs included in this document,
-\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
-log lines) were each parsed 10 times, the results from the first run
-discarded, and the execution time for the remaining 9 runs averaged.
+parse very large log files.  The server which generated the log files used
+in testing this parser accepts approximately 10,000 mails for 700 users a
+day.  Median log file size is 50MB, containing 285,000 log lines --- large
+scale mail servers would have much larger log files.  When generating the
+timing data used in this section, \numberOFlogFILES{} log files (totaling
+10.08 GB, \numberOFlogLINEShuman{} log lines) were each parsed 10 times,
+the results from the first first run discarded, and the execution time for
+the remaining 9 runs averaged.  Saving results to the database was disabled
+for the test runs, as that dominates the run time of the program, and the
+tests are aimed at measuring the speed of the parser rather than the speed
+of the database and the disks the database is stored on.  The computer used
+for test runs was a Dell Optiplex 745, equipped with:
+
+\begin{tabular}[]{ll}
+
+CPU         & One dual core 2.40GHz Intel(R) Core(TM)2 CPU, \\
+            & with 32KB L1 cache and 4MB L2 cache.          \\
+RAM         & 2GB 667 MHz DDR RAM\@.                        \\
+Hard disk   & One XXX 232GB SATA disk.                      \\
+
+\end{tabular}
+
+In as far as possible the computer was dedicated to the task of gathering
+statistics from test runs, and was not used for any other purpose while
+test runs were ongoing.
+
+XXX TIMING INFORMATION\@.
+\newpage
 
@@
 
-An important property of a parser is how execution time scales relative to
+An important property of a parser is how parsing time scales relative to
 input size: does it scale linearly, polynomially, or exponentially?
-Figure~\refwithpage{Execution time, file size, and number of lines} shows
-the execution time in seconds, file size in MB, and tens of thousands of
-log lines per log file.  All three lines run roughly in parallel, giving
-the impression that the algorithm scales linearly with input size.  This
-impression is borne out by fig.~\refwithpage{execution time vs file size vs
-number lines factor} which plots the ratios of file size vs execution time,
-and number of log lines vs execution time (higher is better in both cases).
-As the reader can see the ratios are quite tightly banded across the graph,
-showing that the algorithm scales linearly: the much larger log files at
-points 62--68 on the X axis in fig.~\refwithpage{Execution time, file size,
-and number of lines} actually cause the ratio to increase (i.e.\ improve),
-rather than decrease.  The strange behaviour where larger log files are
-parsed more efficiently is explained in \sectionref{Why are there dips in
-the graphs?}.  \showgraph{build/plot-normal-filesize-numlines}{Execution
-time, file size, and number of lines}{Execution time, file size, and number
-of lines} \showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of
-file size and number of lines to execution time}{execution time vs file
-size vs number lines factor}
+\Graphref{Parsing time, file size, and number of lines} shows the parsing
+time in seconds, file size in MB, and tens of thousands of log lines for
+each of the \numberOFlogFILES{} log files.  All three lines run roughly in
+parallel, giving the impression that the algorithm scales linearly with
+input size.  This impression is borne out by \graphref{parsing time vs file
+size vs number lines factor} which plots both the ratio of file size vs
+parsing time, and the ratio of number of log lines vs parsing time (higher
+is better in both cases).  As the reader can see the ratios are quite
+tightly banded across the graph, showing that the algorithm scales
+linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
+despite their larger than average size (shown in \graphref{Parsing time,
+file size, and number of lines}); this strange behaviour where larger log
+files are parsed more efficiently is explained in \sectionref{Why are there
+dips in the graphs?}.
+\showgraph{build/plot-normal-filesize-numlines}{Parsing time, file size,
+and number of lines}{Parsing time, file size, and number of lines}
+\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file size
+and number of lines to parsing time}{parsing time vs file size vs number
+lines factor}
 
@@
 Figure~\refwithpage{Parsing time relative to shuffled} shows the percentage
-increase of execution times for shuffled and reversed rules.  Overall this
+increase of parsing times for shuffled and reversed rules.  Overall this
 optimisation provides a modest but worthwhile performance increase of over
@@
 relative to shuffled} correspond to peaks in log file size in
-fig.~\refwithpage{Execution time, file size, and number of lines}, and
-peaks in fig.~\refwithpage{execution time vs file size vs number lines
+fig.~\refwithpage{Parsing time, file size, and number of lines}, and
+peaks in fig.~\refwithpage{parsing time vs file size vs number lines
 factor}, where a peak means that performance is better.  The explanation
@@
 required to parse a log line when there is a mail loop, as shown by the
-peaks in fig.~\refwithpage{execution time vs file size vs number lines
+peaks in fig.~\refwithpage{parsing time vs file size vs number lines
 factor} and troughs in fig.~\refwithpage{Parsing time relative to

### 2008-07-30 doc/logparser-results.tex
Further corrections from Carl. Added paragraph about pathological rules. Some cleanups and XXX items dealt with.

@@
             & with 32KB L1 cache and 4MB L2 cache. \\
-RAM         & 2GB 667 MHz DDR RAM. \\
+RAM         & 2GB 667 MHz DDR RAM\@. \\
 Hard disk   & One XXX 232GB SATA disk. \\
@@
 time vs file size vs number lines factor table} shows the same ratios for
-different groups of log files.  As the reader can see the ratios are quite
-tightly banded, providing empirical evidence that the algorithm scales
-linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
-despite their larger than average size (shown in \graphref{parsing time vs
-file size vs number of lines graph}); this strange behaviour where larger
-log files are parsed more efficiently than smaller is explained fully in
-\sectionref{Explaining the peaks in log file size}.
+different groups of log files.  The ratios are quite tightly banded,
+providing empirical evidence that the algorithm scales linearly.  The ratio
+increases (i.e.\ improves) for log files 22 and 62--68 despite their larger
+than average size (shown in \graphref{parsing time vs file size vs number
+of lines graph}); this is explained fully in \sectionref{Explaining the
+peaks in log file size}.
 

### 2008-07-30 papers/sgai-2008/paper.tex
Further corrections from Carl. Added paragraph about pathological rules. Some cleanups and XXX items dealt with.

@@
 
-% XXX FIGURE OUT HOW TO DO THIS AUTOMATICALLY
 % \numberOFrulesMINIMUM as percentage of \numberOFrules
@@
 Rules categorise inputs, specifying both the data to be extracted from each
-log line and the action to invoke.  The Postfix log parser stores the rules
-in the same SQL database that the results are stored in, removing any doubt
-as to which set of rules was used to produce a set of results; other
+log line and the action to invoke.  The Postfix parser stores the rules in
+the same SQL database that the results are stored in, removing any doubt as
+to which set of rules was used to produce a set of results; other
 implementations are free to store the rules in whatever fashion suits their
-needs.
-
-XXX ADD SOMETHING ABOUT IMPLEMENTATIONS BEING FREE TO ADD OR REMOVE
-ATTRIBUTES\@.
+needs.  Similarly implementations are free to require additional attributes
+in their rules; the framework requires \texttt{regex} and \texttt{action}
+attributes, to determine whether a rule matches the current input line, and
+which action to invoke if it does.  Rules used by the Postfix parser
+define the following attributes: XXX DO I NEED TO LIST ALL THE
+ATTRIBUTES\@?
 
@@
 \numberOFlogLINEShuman{} log lines.  Techniques which require the full data
-set when training would be impractical; Instance Based Learning
-(IBL)~\cite{instance-based-learning} techniques which automatically
-determine which inputs from the training set are valuable and which can be
-discarded would probably reduce the data required to a manageable size.
-This avenue of research and development has not been pursued by the author,
-but could easily be independently undertaken.
+set when training would be impractical; Instance Based
+Learning~\cite{instance-based-learning} techniques which automatically
+determine which inputs from the training set are valuable and which inputs
+can be discarded would probably reduce the data required to a manageable
+size.  This avenue of research and development has not been pursued by the
+author, but could easily be independently undertaken.
 
@@
 highest first, and then rules with the same priority are sorted by the
-number of successful matches they had parsing the previous log file.  Negative
-priorities may be useful for catchall rules.
+number of successful matches they had parsing the previous log file.
+Negative priorities may be useful for catchall rules.  XXX DETECTING
+OVERLAP\@.
+
+It is possible to define pathological regexes, which fall into two main
+categories: regexes which match every input line, and regexes which consume
+large amounts of CPU time during matching.  Defining a regex which matches
+every input line is trivial: \verb!/^/! will match the start of the input
+line; every input line is guaranteed to start, so this regex would match
+every input line.  In general a regex will not consume a large amount of
+CPU time when the match is successful; usually CPU time is consumed by a
+regex with a lot of alteration and variable quantifiers failing to match.
+This topic is beyond the scope of this paper, see
+\cite{mastering-regular-expressions} for in-depth discussion.
 
@@
 XXX TIMING INFORMATION\@.
-\newpage
+
+XXX TIMING INFORMATION\@.
+
+XXX TIMING INFORMATION\@.
 
@@
 parsing time, and the ratio of number of log lines vs parsing time (higher
-is better in both cases).  As the reader can see the ratios are quite
-tightly banded across the graph, showing that the algorithm scales
-linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
-despite their larger than average size (shown in \graphref{Parsing time,
-file size, and number of lines}); this strange behaviour where larger log
-files are parsed more efficiently is explained in \sectionref{Why are there
-dips in the graphs?}.
-\showgraph{build/plot-normal-filesize-numlines}{Parsing time, file size,
-and number of lines}{Parsing time, file size, and number of lines}
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file size
-and number of lines to parsing time}{parsing time vs file size vs number
-lines factor}
+is better in both cases).  The ratios are quite tightly banded across the
+graph, showing that the algorithm scales linearly.  The ratio increases
+(i.e.\ improves) for log files 22 and 62--68 despite their larger than
+average size (shown in \graphref{Parsing time, file size, and number of
+lines}); this is explained in \sectionref{Why are there dips in the
+graphs?}.  \showgraph{build/plot-normal-filesize-numlines}{Parsing time,
+file size, and number of lines}{Parsing time, file size, and number of
+lines} \showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
+size and number of lines to parsing time}{parsing time vs file size vs
+number lines factor}
 

### 2008-07-30 doc/logparser-implementation.tex
Moved complications to their own subsection.

@@
 encountered.
-
-\subsubsection{Complications encountered}
-
-XXX DUNNO IF THIS SECTION SHOULD BE HERE\@.  QUEUEID VS PID IS NOT TOO
-IMPORTANT\@; CONNECTION REUSE COULD EASILY BE MOVED TO ADDITIONAL
-COMPLICATIONS\@; RE-INJECTED MAILS SHOULD PROBABLY BE MOVED TO ADDITIONAL
-COMPLICATIONS TOO\@.
-
-\label{complications}
-
-These complications were encountered early in the parser's implementation
-and guided its design and development.
-
-\subsubsection{Queueid vs pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
-
-\end{itemize}
-
-\subsubsection{Connection reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered during one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails will not trample over each other's data.
-This must be done every time a mail is accepted, as it is impossible to
-tell in advance which connections will accept multiple mails.  Once the
-mail has been accepted its log entries will not be correlated by \pid{} any
-more (its queueid will be used instead), so there is no longer any
-ambiguity about which mail a given log line belongs to (except when
-timeouts occur during the data phase
-\sectionref{timeouts-during-data-phase}).  If the original connection has
-any interesting data (e.g.\ rejections) it will be saved to the database
-when the client disconnects.  One unsolved difficulty is distinguishing
-between different groups of rejections, e.g.\ when dealing with the
-following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should ideally be two separate entries in the database resulting from
-the above sequence, but currently there will only be one.
-
-\subsubsection{Re-injected mails}
-
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the log
-files it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
-
-\end{description}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how soon \daemon{qmgr} processes the new mail.
-
-Because of the uncertain order the parser cannot complain when it
-encounters a log line from \daemon{qmgr} for a previously unseen mail;
-instead it must flag the mail as coming from an unknown origin, and
-subsequently clear the flag if and when the origin of the mail becomes
-clear.  Obviously the parser could omit checking where mails originate
-from, but requiring an explicit source helps to expose bugs in the parser;
-such checks helped to identify the complications described in
-\sectionref{discarding cleanup lines} and \sectionref{pickup logging after
-cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
 
@@
 
-\subsubsection{Additional complications}
+\subsubsection{Summary}
+
+This section presented the core of the parser, starting with a very high
+level view and the initial complications which arose.  A flow chart showing
+the paths a mail may take through the nascent simplified algorithm was
+provided, followed by an explanation of those paths, and a discussion of
+the parser's emergent behaviour --- the data from the log files creates the
+paths in the flow chart, they are not specified anywhere in the parser.
+The framework which holds the parser together was covered next, after which
+came a description of the current actions provided by the parser, and the
+algorithm for analysing unparsed log lines to create \regexes{} for new
+rules.  Detecting, diagnosing and defeating complications forms the largest
+single portion of this document, mirroring the development of the parser,
+The complications are described in the order they were overcome, with
+subsequent problems affecting fewer mails (often by an order or magnitude),
+though the time required to solve problems increased with each successive
+problem.
+
+\subsection{Complications encountered}
 
 \label{additional complications}
+\label{complications}
 
@@
 always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from
-the state tables and error messages, and make the remaining problems more
-apparent.
+improvement in the parser, prune the greatest number of mails from the
+state tables and error messages, and make the remaining problems more
+apparent.  The first three complications were encountered early in the
+parser's implementation and guided its design and development.
+
+\subsubsection{Queueid vs pid}
+
+The mail lacks a queueid until it has been accepted, so log lines must
+first be correlated by the \daemon{smtpd} \pid{}, then transition to being
+correlated by the queueid.  This is relatively minor, but does require:
+
+\begin{itemize}
+
+    \item Two versions of several functions: \texttt{by\_pid} and
+        \texttt{by\_queueid}.
+
+    \item Two state tables to hold the data structure for each connection.
+
+    \item Most importantly: every section of code must know whether it
+        needs to lookup the data structures by \pid{} or queueid.
+
+\end{itemize}
+
+\subsubsection{Connection reuse}
+
+\label{connection reuse}
+
+Multiple independent mails may be delivered during one connection: this
+requires the algorithm to clone the current data as soon as a mail is
+accepted, so that subsequent mails will not trample over each other's data.
+This must be done every time a mail is accepted, as it is impossible to
+tell in advance which connections will accept multiple mails.  Once the
+mail has been accepted its log entries will not be correlated by \pid{} any
+more (its queueid will be used instead), so there is no longer any
+ambiguity about which mail a given log line belongs to (except when
+timeouts occur during the data phase
+\sectionref{timeouts-during-data-phase}).  If the original connection has
+any interesting data (e.g.\ rejections) it will be saved to the database
+when the client disconnects.  One unsolved difficulty is distinguishing
+between different groups of rejections, e.g.\ when dealing with the
+following sequence:
+
+\begin{enumerate}
+
+    \item The client attempts to deliver a mail, but it is rejected.
+
+    \item The client issues the RSET command to reset the session.
+
+    \item The client attempts to deliver another mail, likewise rejected.
+
+\end{enumerate}
+
+There should ideally be two separate entries in the database resulting from
+the above sequence, but currently there will only be one.
+
+\subsubsection{Re-injected mails}
+
+The most difficult complication initially encountered is that locally
+addressed mails are not always delivered directly to a mailbox: sometimes
+they are addressed to and accepted for a local address but need to be
+delivered to one or more remote addresses due to aliases.  When this occurs
+a child mail will be injected into the Postfix queue, but without the
+explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
+Thus the source is not immediately discernible from the log line in which
+the mail first appears; from a strictly chronological reading of the log
+files it usually appears as if the child mail has appeared from thin air.
+Subsequently the parent mail will log the creation of the child mail:
+
+\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
+delay=0, status=sent (forwarded as 56F5B43FD)}
+
+The child mail has been created with queueid \texttt{56F5B43FD}.  Different
+delivery methods result in different log lines:
+
+\begin{description}
+
+    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
+
+    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
+        BD07D3C49
+
+    \item [Local delivery:] delivered to command:
+        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
+
+\end{description}
+
+Unfortunately, while all log lines from an individual process appear in
+chronological order, the order in which log lines from different processes
+are interleaved is subject to the vagaries of process scheduling.  In
+addition, the first log line belonging to the child mail (the log line
+cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
+order also depends on how soon \daemon{qmgr} processes the new mail.
+
+Because of the uncertain order the parser cannot complain when it
+encounters a log line from \daemon{qmgr} for a previously unseen mail;
+instead it must flag the mail as coming from an unknown origin, and
+subsequently clear the flag if and when the origin of the mail becomes
+clear.  Obviously the parser could omit checking where mails originate
+from, but requiring an explicit source helps to expose bugs in the parser;
+such checks helped to identify the complications described in
+\sectionref{discarding cleanup lines} and \sectionref{pickup logging after
+cleanup}.
+
+Process scheduling can have a still more confusing effect: quite often the
+child mail will be created, delivered and entirely finished with
+\textbf{before} the parent logs the creation log line!  Thus, mails flagged
+as coming from an unknown origin cannot be entered into the database when
+their final log line is parsed; instead they must be marked as ready for
+entry and subsequently entered by the parent mail once it has been
+identified.
 
@@
 
-\subsubsection{Summary}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they are not specified anywhere in the parser.
-The framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
-
 \subsection{Limitations and possible improvements}
@@
 improvements which may be implemented in the future.
+
+\subsection{Summary}
+
+XXX TO BE WRITTEN\@.

### 2008-07-30 doc/logparser-results.tex
Bunch of XXX items added.
Added timing data.

@@
 
+USE THE DATA COLLECTED ABOUT NUMBER OF RULES TRIED\@.  MAYBE INCLUDE IN
+ORDERINGS SECTION\@?
+
 \subsection{Parser Efficiency}
@@
 \label{parser efficiency}
-
-XXX ADD TIMING INFORMATION\@.
 
@@
 
-CPU         & One dual core 2.40GHz Intel(R) Core(TM)2 CPU, \\
+    CPU         & One dual core 2.40GHz Intel\textregistered{}
+                    Core\texttrademark{}2 CPU, \\
                 & with 32KB L1 cache and 4MB L2 cache. \\
     RAM         & 2GB 667 MHz DDR RAM\@. \\
-Hard disk   & One XXX 232GB SATA disk. \\
+    Hard disk   & One Seagate Barracuda 7200 RPM 250GB SATA disk. \\
 
@@
 
-In as far as possible the computer was dedicated to the task of gathering
-statistics from test runs, and was not used for any other purpose while
-test runs were ongoing.
+Parsing all \numberOFlogFILES{} log files in one run took
+\input{build/timing-run-include.tex}.
+
+The computer was dedicated to the task of gathering statistics from test
+runs, and was not used for any other purpose while test runs were ongoing;
+any services not necessary for running the tests were disabled.
 
@@
 \subsection{Coverage}
+
+I THINK THIS SECTION NEEDS SUBSTANTIAL WORK, IF NOT A RE-WRITE\@.
 

### 2008-07-30 papers/sgai-2008/paper.tex
Added timing data.

@@
 of the database and the disks the database is stored on.  The computer used
-for test runs was a Dell Optiplex 745, equipped with:
+for test runs was a Dell Optiplex 745:
 
@@
 
-CPU         & One dual core 2.40GHz Intel(R) Core(TM)2 CPU, \\
+    CPU         & One dual core 2.40GHz Intel\textregistered{}
+                    Core\texttrademark{}2 CPU, \\
                 & with 32KB L1 cache and 4MB L2 cache. \\
     RAM         & 2GB 667 MHz DDR RAM\@. \\
-Hard disk   & One XXX 232GB SATA disk.                      \\
+    Hard disk   & One Seagate Barracuda 7200 RPM 250GB SATA disk. \\
 
@@
 
-In as far as possible the computer was dedicated to the task of gathering
-statistics from test runs, and was not used for any other purpose while
-test runs were ongoing.
+Parsing all \numberOFlogFILES{} log files in one run took
+\input{build/timing-run-include.tex}.
 
-XXX TIMING INFORMATION\@.
-
-XXX TIMING INFORMATION\@.
-
-XXX TIMING INFORMATION\@.
+The computer was dedicated to the task of gathering statistics from test
+runs, and was not used for any other purpose while test runs were ongoing;
+any services not necessary for running the tests were disabled.
 
@@
 
+\newpage
 \subsection{Rule Ordering}
@@
 (as shown in fig.~\refwithpage{rule hits graph}).
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph} Assuming that
+Assuming that
 the distribution of log lines is reasonably steady over time, parser
@@
 10\%, for a small investment in time and programming.
+\showgraph{build/plot-hits}{Hits per rule}{rule hits graph} 
 \showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Parsing time

### 2008-07-30 doc/logparser-results.tex
Corrected the data used in normal-and-reverse-relative-to-shuffle graph, and added missing caption. Replaced \refwithpage with \graphref everywhere. Minor improvements.

@@
 \Graphref{Parsing time relative to shuffled ordering graph} shows the
-execution time of normal and reverse orderings relative to shuffled
-ordering; the mean relative execution times for different groupings of log
+parsing times of normal and reverse orderings relative to shuffled
+ordering; the mean relative parsing times for different groupings of log
 files are given in \tableref{Parsing time relative to shuffled ordering

### 2008-07-30 papers/sgai-2008/paper.tex
Corrected the data used in normal-and-reverse-relative-to-shuffle graph, and added missing caption. Replaced \refwithpage with \graphref everywhere. Minor improvements.

@@
 rules; the distribution of rules to actions is very skewed, as shown in
-Fig.~\refwithpage{Distribution of rules per action}.  The distribution is
-skewed for the same reason the architecture is so successful in enabling
-users to parse new log lines: most new log lines can be parsed by a
-combination of a new rule and an existing action.
+\graphref{Distribution of rules per action}.  The distribution is skewed
+for the same reason the architecture is so successful in enabling users to
+parse new log lines: most new log lines can be parsed by a combination of a
+new rule and an existing action.
 \showgraph{build/plot-action-distribution}{Distribution of rules per
@@
 lines split across the other 90\% of the rules in a Power Law distribution
-(as shown in fig.~\refwithpage{rule hits graph}).
-Assuming that
-the distribution of log lines is reasonably steady over time, parser
-efficiency should benefit from trying more frequently matching rules before
-those which match less frequently.  To test this hypothesis three full test
-runs were performed with different rule orderings:
+(as shown in \graphref{rule hits graph}).  Assuming that the distribution
+of log lines is reasonably steady over time, parser efficiency should
+benefit from trying more frequently matching rules before those which match
+less frequently.  To test this hypothesis three full test runs were
+performed with different rule orderings:
 
@@
 
-Figure~\refwithpage{Parsing time relative to shuffled} shows the percentage
-increase of parsing times for shuffled and reversed rules.  Overall this
-optimisation provides a modest but worthwhile performance increase of over
-10\%, for a small investment in time and programming.
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph} 
+\Graphref{Parsing time relative to shuffled} shows the parsing times of
+normal and reversed orderings relative to shuffled ordering.  Overall this
+optimisation provides a modest but worthwhile reduction in parsing time of
+approximately 10\%.  \showgraph{build/plot-hits}{Hits per rule}{rule hits
+graph}
 \showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Parsing time
-relative to shuffled}{Parsing time relative to shuffled}
+relative to shuffled ordering}{Parsing time relative to shuffled}
 
@@
 
-The dips at log files 22 and 62--68 in fig.~\refwithpage{Parsing time
-relative to shuffled} correspond to peaks in log file size in
-fig.~\refwithpage{Parsing time, file size, and number of lines}, and
-peaks in fig.~\refwithpage{parsing time vs file size vs number lines
-factor}, where a peak means that performance is better.  The explanation
-for this took some time to arrive at, but it turns out to be reasonably
-simple.  The large log files in question were caused by a mail forwarding
-loop, where the distribution of log lines is quite different to normal,
-resulting in different performance characteristics.  The vast majority of
-log lines when a mail loop occurs are from Postfix components which have a
-small number of rules associated with them, whereas in general the majority
-of log lines are from the Postfix component (\daemon{smtpd}) with the
-highest number of rules.  Furthermore in normal log files the majority of
-log lines are distributed much more evenly across rules than the majority
-of log lines during a mail loop.
+The dips at log files 22 and 62--68 in \graphref{Parsing time relative to
+shuffled} correspond to peaks in log file size in \graphref{Parsing time,
+file size, and number of lines}, and peaks in \graphref{parsing time vs
+file size vs number lines factor}, where a peak means that performance is
+better.  The explanation for this took some time to arrive at, but it turns
+out to be reasonably simple.  The large log files in question were caused
+by a mail forwarding loop, where the distribution of log lines is quite
+different to normal, resulting in different performance characteristics.
+The vast majority of log lines when a mail loop occurs are from Postfix
+components which have a small number of rules associated with them, whereas
+in general the majority of log lines are from the Postfix component
+(\daemon{smtpd}) with the highest number of rules.  Furthermore in normal
+log files the majority of log lines are distributed much more evenly across
+rules than the majority of log lines during a mail loop.
 
@@
 required to parse a log line when there is a mail loop, as shown by the
-peaks in fig.~\refwithpage{parsing time vs file size vs number lines
-factor} and troughs in fig.~\refwithpage{Parsing time relative to
-shuffled}.  When the rule ordering is reversed the majority of log lines
-generated by a mail loop will be parsed using very few rules, because there
-are very few rules for the Postfix components producing the log lines.
-Without a mail loop the Postfix components producing the majority of log
-lines have a large number of rules associated with them, and reverse
-ordering deliberately starts with the least likely rules and moves towards
-the most likely.  This leads to a noticeable drop in the average time
-required to parse a log line during a mail loop, as shown in
-fig.~\refwithpage{Parsing time relative to shuffled}.  The number of rules
-which need to be consulted when the ordering is shuffled varies between the
-optimal and the pessimal, with the performance varying proportionally.
+peaks in \graphref{parsing time vs file size vs number lines factor} and
+troughs in \graphref{Parsing time relative to shuffled}.  When the rule
+ordering is reversed the majority of log lines generated by a mail loop
+will be parsed using very few rules, because there are very few rules for
+the Postfix components producing the log lines.  Without a mail loop the
+Postfix components producing the majority of log lines have a large number
+of rules associated with them, and reverse ordering deliberately starts
+with the least likely rules and moves towards the most likely.  This leads
+to a noticeable drop in the average time required to parse a log line
+during a mail loop, as shown in \graphref{Parsing time relative to
+shuffled}.  The number of rules which need to be consulted when the
+ordering is shuffled varies between the optimal and the pessimal, with the
+performance varying proportionally.
 
@@
 minimal ruleset for normal, shuffled and reversed orderings is shown in
-fig.~\refwithpage{Percentage execution time increase of maximum ruleset
-over minimum ruleset}.  \showgraph{build/plot-normal-vs-smaller}{Percentage
+\graphref{Percentage execution time increase of maximum ruleset over
+minimum ruleset}.  \showgraph{build/plot-normal-vs-smaller}{Percentage
 execution time increase of maximum ruleset over minimum ruleset}{Percentage
@@
 
-It is clear from fig.~\refwithpage{Percentage execution time increase of
-maximum ruleset over minimum ruleset} that the increased number of rules
-has a noticeable performance impact with reverse ordering, and to a lesser
-extent with shuffled ordering.  The normal ordering (where are the most
-frequently matching rules are tried first) shows a mean increase of 
-\input{build/normal-vs-smaller-percentage-increase.tex} in
-execution time for a \numberOFrulesMAXIMUMpercentage{} increase in the
-number of rules.  These results show that the architecture scales very well
-as the number of rules increases, and that sorting the rules is an
-important optimisation enabling this.
+It is clear from \graphref{Percentage execution time increase of maximum
+ruleset over minimum ruleset} that the increased number of rules has a
+noticeable performance impact with reverse ordering, and to a lesser extent
+with shuffled ordering.  The normal ordering (where are the most frequently
+matching rules are tried first) shows a mean increase of
+\input{build/normal-vs-smaller-percentage-increase.tex} in execution time
+for a \numberOFrulesMAXIMUMpercentage{} increase in the number of rules.
+These results show that the architecture scales very well as the number of
+rules increases, and that sorting the rules is an important optimisation
+enabling this.
 

### 2008-07-30 doc/logparser-results.tex
Updated graph data. No longer discarding data from the first run. Minor corrections. Lots of XXX items added.

@@
 log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
-parsed 10 times, the results from the first first run discarded, and the
-execution time for the remaining 9 runs averaged.  The results from the
-first run were discarded because because the rule ordering will be
-sub-optimal compared to subsequent runs --- XXX MAYBE I SHOULD NOT DISCARD
-THE FIRST RESULT\@?  Saving results to the database was disabled for the
-test runs, as that dominates the run time of the program, and the tests are
-aimed at measuring the speed of the parser rather than the speed of the
-database and the disks the database is stored on.  The computer used for
-test runs was a Dell Optiplex 745, equipped with:
+parsed 10 times and the parsing times averaged.  Saving results to the
+database was disabled for the test runs, as that dominates the run time of
+the program, and the tests are aimed at measuring the speed of the parser
+rather than the speed of the database and the disks the database is stored
+on.  The computer used for test runs was a Dell Optiplex 745, equipped
+with:
 
@@
 The different distribution of log lines during a mail forwarding loop has
-interesting effects on execution time, described in detail in
+interesting effects on parsing time, described in detail in
 \sectionref{rule ordering for efficiency}, \sectionref{scalability as the
@@
 the minimum ruleset and compared to the statistics generated using the full
-ruleset: the percentage execution increase when using the full ruleset
+ruleset: the percentage parsing time increase when using the full ruleset
 instead of the minimal ruleset for normal, shuffled and reversed orderings
@@
 This section began with parser scalability, showing the linear relationship
-between execution time and input size.  Demonstrates the effect of rule
-ordering on execution time, and the unexpected consequences of specific
+between parsing time and input size.  Demonstrates the effect of rule
+ordering on parsing time, and the unexpected consequences of specific
 inputs.  The necessity of caching compiled \regexes{} is attested to by the

### 2008-07-30 papers/sgai-2008/paper.tex
Updated graph data. No longer discarding data from the first run. Minor corrections. Lots of XXX items added.

@@
 }
+
+XXX DISCUSS COMPLEXITY\@?  EVEN JUST MENTION THAT COMPLEXITY RISES WHEN
+INCREMENTAL PARSING IS USED, BUT THAT EXTERNAL TOOLS COULD BE USED TO DEAL
+WITH IT, AND THAT IT IS ENTIRELY DEPENDANT ON THE RULESET USED\@?
+
+XXX MAYBE MERGE THE AUTHOR SECTIONS IF DESPERATE FOR SPACE\@.
+
+XXX REVIEWER SUGGESTED COMPARING TO EXISTING WORK, BUT I DID NOT FIND ANY
+SIMILAR EXISTING WORK\@.  SHOULD SEARCH AGAIN\@.
+
+XXX OVERLAP DETECTION\@: L1 n L2 = E* --- ((E* --- L1) u (E* --- L2)).
+PERL REGEXES ARE AT LEAST AS POWERFUL AS A PUSHDOWN AUTOMATA --- SEE
+CHOMSKY'S HIERARCHY TO MAKE SURE\@.  CONTAINMENT CANNOT BE DETERMINED FOR
+CFGs, INTERSECTION IS POSSIBLE\@.
 
@@
 deal with the varying log lines, and the actions deal with the intricacies
-and complications encountered when parsing.
+and complications encountered when parsing.  XXX MAYBE REMOVE FIRST TWO
+SENTENCES\@.
 
@@
 \label{Actions}
+
+XXX CARL WONDERS IF THE NUMBERS HERE ARE TYPICAL\@?  ALSO SAY SOMETHING
+LIKE REJECTION AND IGNORE DO NOT IMPLY UNGRAMMATICAL\@.
 
@@
         match the log line, or worse, might match unintentionally.  Rules
-        whose program is \texttt{*} will be tried against any log lines not
-        parsed by program specific rules, allowing generic rules to be
-        written.
+        whose program is ``\texttt{*}'' will be tried against any log lines
+        not parsed by program specific rules, allowing generic rules to be
+        written.  This is a literal ``\texttt{*}'', not a wildcard or
+        regex.
 
@@
 size.  This avenue of research and development has not been pursued by the
-author, but could easily be independently undertaken.
+authors, but could easily be independently undertaken.
 
@@
 large amounts of CPU time during matching.  Defining a regex which matches
-every input line is trivial: \verb!/^/! will match the start of the input
-line; every input line is guaranteed to start, so this regex would match
-every input line.  In general a regex will not consume a large amount of
-CPU time when the match is successful; usually CPU time is consumed by a
-regex with a lot of alteration and variable quantifiers failing to match.
-This topic is beyond the scope of this paper, see
-\cite{mastering-regular-expressions} for in-depth discussion.
+every input line is trivial: \verb!/^/! will match the start of every input
+line.  In general a regex will not consume a large amount of CPU time when
+the match is successful; usually CPU time is consumed by a regex with a lot
+of alteration and variable quantifiers failing to match.  This topic is
+beyond the scope of this paper, see~\cite{mastering-regular-expressions}
+for in-depth discussion.
 
@@
 timing data used in this section, \numberOFlogFILES{} log files (totaling
-10.08 GB, \numberOFlogLINEShuman{} log lines) were each parsed 10 times,
-the results from the first first run discarded, and the execution time for
-the remaining 9 runs averaged.  Saving results to the database was disabled
+10.08 GB, \numberOFlogLINEShuman{} log lines) were each parsed 10 times and
+the parsing times averaged.  Saving results to the database was disabled
 for the test runs, as that dominates the run time of the program, and the
@@
 \begin{tabular}[]{ll}
-
+    \hline
+    \noalign{\smallskip}
+    XXX         & XXX NEED LABELS\@.                                \\
+    \noalign{\smallskip}
+    \hline
+    \noalign{\smallskip}
     CPU         & One dual core 2.40GHz Intel\textregistered{}
@@
     Hard disk   & One Seagate Barracuda 7200 RPM 250GB SATA disk.   \\
-
+    \noalign{\smallskip}
+    \hline
+    \noalign{\smallskip}
 \end{tabular}
@@
 Parsing all \numberOFlogFILES{} log files in one run took
-\input{build/timing-run-include.tex}.
-
-The computer was dedicated to the task of gathering statistics from test
-runs, and was not used for any other purpose while test runs were ongoing;
-any services not necessary for running the tests were disabled.
+\input{build/timing-run-include.tex}.  The computer was dedicated to the
+task of gathering statistics from test runs, and was not used for any other
+purpose while test runs were ongoing; any services not necessary for
+running the tests were disabled.
 
@@
 
+XXX MAYBE REDUCE THIS\@?
+
 \label{Why are there dips in the graphs?}
@@
 rules than the majority of log lines during a mail loop.
+
+\newpage
 
@@
 results compared to the statistics generated using the full ruleset: the
-percentage execution increase when using the full ruleset instead of the
+percentage parsing time increase when using the full ruleset instead of the
 minimal ruleset for normal, shuffled and reversed orderings is shown in
-\graphref{Percentage execution time increase of maximum ruleset over
-minimum ruleset}.  \showgraph{build/plot-normal-vs-smaller}{Percentage
-execution time increase of maximum ruleset over minimum ruleset}{Percentage
-execution time increase of maximum ruleset over minimum ruleset}
+\graphref{Percentage parsing time increase of maximum ruleset over minimum
+ruleset}.  \showgraph{build/plot-normal-vs-smaller}{Percentage parsing time
+increase of maximum ruleset over minimum ruleset}{Percentage parsing time
+increase of maximum ruleset over minimum ruleset}
 
-It is clear from \graphref{Percentage execution time increase of maximum
+It is clear from \graphref{Percentage parsing time increase of maximum
 ruleset over minimum ruleset} that the increased number of rules has a
@@
 matching rules are tried first) shows a mean increase of
-\input{build/normal-vs-smaller-percentage-increase.tex} in execution time
-for a \numberOFrulesMAXIMUMpercentage{} increase in the number of rules.
-These results show that the architecture scales very well as the number of
-rules increases, and that sorting the rules is an important optimisation
-enabling this.
+\input{build/normal-vs-smaller-percentage-increase.tex} in parsing time for
+a \numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
+results show that the architecture scales very well as the number of rules
+increases, and that sorting the rules is an important optimisation enabling
+this.
 
@@
 tables, and 1625 connections correctly entered in the database.  Given the
-evidence detailed above, the author is confident that the false positive
+evidence detailed above, the authors are confident that the false positive
 rate when reconstructing a mail is zero.

### 2008-07-31 doc/logparser-implementation.tex
Bunch of XXX items added.
Added timing data.
Removed the graphs showing breakdown of rule hits. Moved remaining graphs out of graphs section and removed graphs section. Minor corrections.

@@
+\newpage
 \section{Postfix Parser Implementation}
@@
 \subsubsection{Summary}
+
+XXX DO I NEED A SUMMARY HERE\@?
 
@@
 
+XXX MERGE WITH PREVIOUS PARAGRAPHS\@.
+
 This section covers the following topics:
@@
 TRACKING (\textsection{}5.4.3) COULD BE MOVED TO ADDITIONAL
-COMPLICATIONS\@.
+COMPLICATIONS\@.  POSSIBLY MERGE MAIL ENTRY AND DELIVERY IN HERE\@;
+TRACKING SHOULD BE IN COMPLICATIONS --- MERGE WITH RE-INJECTED MAILS\@.
 
@@
 
-The distribution of rules per action is shown in
-\graphref{Distribution of rules per action}.
+The distribution of rules per action is shown in \graphref{Distribution of
+rules per action}.
+
+\showgraph{build/plot-action-distribution}{Distribution of rules per
+action}{Distribution of rules per action}
 
@@
 
-\subsubsection{Summary}
-
-This section has covered the limitations of the parser and possible
-improvements which may be implemented in the future.
-
 \subsection{Summary}

### 2008-07-31 doc/logparser-results.tex
Removed the graphs showing breakdown of rule hits. Moved remaining graphs out of graphs section and removed graphs section. Minor corrections.

@@
 
-XXX EXPLAIN THE X-AXES ON THE GRAPHS\@.
-
 XXX CHANGE normal TO optimal IN ORDERING STUFF, INCLUDING IN GRAPH CAPTIONS
@@
 
-USE THE DATA COLLECTED ABOUT NUMBER OF RULES TRIED\@.  MAYBE INCLUDE IN
+XXX USE THE DATA COLLECTED ABOUT NUMBER OF RULES TRIED\@.  MAYBE INCLUDE IN
 ORDERINGS SECTION\@?
@@
 in testing this parser accepts approximately 10,000 mails for 700 users a
-day; as expected there are far more mails received on weekdays than at
-weekends.   Median log file size is 50MB, containing 285,000 log lines ---
-large scale mail servers would have much larger log files.  Note that
-\graphref{Mails received per day} and \tableref{Number of mails received
-per day: statistics} show the number of mails received by \SMTP{} only; in
-particular the mail loops noticeable in other graphs do not contribute to
-these figures.
+day; \graphref{Mails received per day} shows that as expected there are far
+more mails received on weekdays than at weekends.   Median log file size is
+50MB, containing 285,000 log lines --- large scale mail servers would have
+much larger log files.  Note that \graphref{Mails received per day} and
+\tableref{Number of mails received per day: statistics} show the number of
+mails received by \SMTP{} only; in particular the mail loops noticeable in
+other graphs do not contribute to these figures.
 
-\showgraph{build/plot-mails-received}{Number of mails received per
+\showgraph{build/plot-mails-received}{Number of mails received via SMTP per
 day}{Mails received per day}
@@
 \showtable{build/mails-received-include-for-graph}{Number of mails received
-per day}{Number of mails received per day: statistics}
+via SMTP per day}{Number of mails received per day: statistics}
 
@@
 of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILES{} contiguous log files from Postfix 2.2 and 2.3), and new
-rules are easy to add.  Maintaining full coverage is hard because other
-servers have different restrictions with custom messages, \DNSBL{} messages
-change over time, major releases of Postfix change warning messages
-(usually adding more information), etc.,\ so over time the log lines drift
-and change.  \Graphref{rule hits graph} shows the number of hits
-for each rule over all \numberOFlogFILES{} log files; it is obvious that a
-small number of rules match the vast majority of the lines, and more than
-half the rules match fewer than 100 times.
+\numberOFlogFILESall{} contiguous log files from Postfix 2.2 through to
+Postfix 2.5), and new rules are easy to add.  Maintaining full coverage is
+hard because other servers have different restrictions with custom
+messages, \DNSBL{} messages change over time, major releases of Postfix
+change warning messages (usually adding more information), etc.,\ so over
+time the log lines drift and change.  \Graphref{rule hits graph} shows the
+number of hits for each rule over all \numberOFlogFILES{} log files.  The
+number of hits per rule is quite unevenly spread, resembling a Power Law
+distribution~\cite{powerlaw} --- it is obvious that a small number of rules
+match the vast majority of the lines, and more than half the rules match
+fewer than 100 times.
+
+\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
 
@@
 
-\subsubsection{Summary}
+\subsection{Summary}
+
+XXX EXTEND TO SUMMARISE PARSER EFFICIENCY TOO\@.
+
+XXX REWRITE COVERAGE SUMARY IF NECESSARY\@.
 
@@
 further examination of the log files in developing a solution.
-
-\subsection{Graphs}
-
-XXX FIGURE OUT WHERE TO PUT THESE GRAPHS\@; DO I NEED ALL THREE\@?
-
-\showgraph{build/plot-action-distribution}{Distribution of rules per
-action}{Distribution of rules per action}
-
-The number of hits per rule is quite unevenly spread, resembling a Power
-Law distribution~\cite{powerlaw}.
-
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
-
-As \graphref{rule hits graph} is quite difficult to read it has been
-separated into three sections: low, middle and high.
-
-\showgraph{build/plot-hits-low}{Hits per rule (low)}{hits per rule low}
-
-It is apparent from the low graph (\graphref{hits per rule low})
-that some rules have few or no hits; those with zero hits are rules which
-were written to parse log files used during development of the parser but
-not utilised in the test runs performed for this document.
-
-\showgraph{build/plot-hits-middle}{Hits per rule (middle)}{hits per rule
-middle}
-
-\showgraph{build/plot-hits-high}{Hits per rule (high)}{hits per rule
-high}
-
-\clearpage{}

### 2008-08-01 doc/logparser-architecture.tex
Improved parts of the efficiency section and inlined graphs. Fixed up some references; added $Id$ to front page.
Bunch of XXX items added.
Dealt with some XXX items.

@@
 \section{Parser Architecture}
+
+XXX REWRITE AS NECESSARY TO MERGE AND SUPPLEMENT THE EXISTING CONTENT\@.
 
@@
         The actions perform the work of reconstructing the journey a mail
-        takes through Postfix.  Full details of actions can be found in
-        \sectionref{actions-in-detail} and \sectionref{adding new actions}
-        XXX IMPROVE THIS SENTENCE\@.
+        takes through Postfix.  Details of the actions available in the
+        Postfix parser can be found in \sectionref{actions-in-detail}, and
+        \sectionref{adding new actions} describes the process of adding new
+        actions.
 
@@
         managing data structures, finding the rule which matches each log
-        line, invoking the correct action, etc\@.  The framework provides
-        the structure which actions and rules plug into.  The framework is
+        line, invoking the correct action, etc\@.  The framework is
         described in detail in \sectionref{framework}.  XXX EXTEND
@@
 
-XXX CLARIFY THE SENTENCE STARTING WITH ``Although this may see'' IN THE
-NEXT PARAGRAPH\@.
-
 Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Although
-this may seem like a trivial point, is it substantially more difficult to
-understand a program's entire parsing algorithm, identify the correct
-location to change and make the appropriate changes, versus adding a new
-rule with the action to invoke, a \regex{} to match the log lines, and the
-specification of the data to extract.  Bear in mind that the changes must
-be made without adversely affecting existing parsing, particularly as there
-may be edge cases which are not immediately obvious.\footnote{See
-\sectionref{yet-more-aborted-delivery-attempts} for a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.}  Requiring
+parse new log lines without modifying the core parsing algorithm.  Adding a
+new rule with the action to invoke and a \regex{} to match the log lines is
+trivial in comparison to understanding a program's entire parsing
+algorithm, identifying the correct location to change and making the
+appropriate changes.  Bear in mind that the changes must be made without
+adversely affecting existing parsing, particularly as there may be edge
+cases which are not immediately obvious ---
+\sectionref{yet-more-aborted-delivery-attempts} describes a complication which
+occurs only four times in \numberOFlogFILES{} log files tested.  Requiring
 changes to the parser's code also complicates upgrades, as the changes must
 be preserved during the upgrade, and may clash with changes made by the
-developer.\footnote{See \sectionref{adding new actions} for how to add new
-actions}  \parsername{} allows the user to add new rules to the database
+developer (\sectionref{adding new actions} describes safely adding new
+actions).  \parsername{} allows the user to add new rules to the database
 without changing the parsing algorithm, unless the new log lines to be
@@
 
-XXX CHANGE THIS\@; THE PARSER DESIGN IS NOVEL, DO NOT STRESS THE SIMILARITY
-TOO MUCH\@.
-
 There is some similarity between the parser's design and William Wood's
@@
 this appears to be a natural division of responsibility and functionality.
+\Tableref{Similarities between ATN and this architecture} shows the
+similarities.
 
@@
 
+\begin{table}[ht]
+    \caption{Similarities between ATN and this architecture}
+    \empty{}\label{Similarities between ATN and this architecture}
     \begin{tabular}[]{lll}
-    \textit{\ATN{}\/}   & \textit{Parser\/} & \textit{Similarity\/}     \\
+        \tableline{}
+        \ATN{}          & Architecture  & Similarity                \\
+        \tableline{}
         Networks        & Algorithm     & Determines the sequence 
@@
                         &               & or actions which 
-                                              constitutes a valid       \\
-                        &                   & input.                    \\
+                                        constitutes a valid  input. \\
         Transitions     & Actions       & Save data and impose
@@
                                         classifying input.          \\
+        \tableline{}
     \end{tabular}
+\end{table}
 
@@
     \item Compare the results of parsing using sorted, shuffled and
-        reversed rules (described in \sectionref{rule efficiency}).  Parse
-        a number of log files using normal sorting, then dump a textual
-        representation of the rules, connections and results tables.
-        Repeat with shuffled and reversed rules, starting with a fresh
-        database.  If there are no overlapping rules the tables from each
-        run will be identical; differences indicate overlapping rules.
-        Which rules overlap can be determined by examining the differences
-        in the tables: each result contains a reference to the rule which
-        created it, if the references differ between runs the two rules
-        referenced in the differing results overlap.  Unfortunately this
-        method cannot prove the absence of overlapping rules; it can detect
-        overlapping rules, but only if there are log lines in the input
-        files which match more than one rule.
+        reversed rules (described in \sectionref{rule ordering for
+        efficiency}).  Parse a number of log files using normal sorting,
+        then dump a textual representation of the rules, connections and
+        results tables.  Repeat with shuffled and reversed rules, starting
+        with a fresh database.  If there are no overlapping rules the
+        tables from each run will be identical; differences indicate
+        overlapping rules.  Which rules overlap can be determined by
+        examining the differences in the tables: each result contains a
+        reference to the rule which created it, if the references differ
+        between runs the two rules referenced in the differing results
+        overlap.  Unfortunately this method cannot prove the absence of
+        overlapping rules; it can detect overlapping rules, but only if
+        there are log lines in the input files which match more than one
+        rule.
 
@@
 \end{verbatim}
+
+\subsection{Actions}
+
+XXX TO BE WRITTEN\@.
+
+\subsection{Rules}
+
+XXX TO BE WRITTEN\@.

### 2008-08-01 doc/logparser-background.tex
Wrote minimum ruleset versus maximum ruleset section - largely taken from the SGAI paper.
Dealt with some XXX items.

@@
 the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must be cognisant of the ramifications of
-their configuration changes.  XXX UNIX DOESN'T STOP YOU DOING STUPID THINGS
-BECAUSE THAT WOULD ALSO STOP YOU DOING CLEVER THINGS\@.
+all users --- the administrator must be cognisant of the effects of their
+configuration changes.  This is similar to one of UNIX's design
+philosophies: \textit{``UNIX was not designed to stop its users from doing
+stupid things, as that would also stop them from doing clever things''\/}
+--- Doug Gwyn~\cite{unix-philosophy}.
 

### 2008-08-01 doc/logparser.tex
Lots of improvements to the efficiency section - I'm happy with most of it now. I improved the graphs too, with better captions and labels.
Wrote minimum ruleset versus maximum ruleset section - largely taken from the SGAI paper.
Further corrections from Carl. Added paragraph about pathological rules. Some cleanups and XXX items dealt with.
Removed the graphs showing breakdown of rule hits. Moved remaining graphs out of graphs section and removed graphs section. Minor corrections.
Dealt with some XXX items.

@@
 
+\newcommand{\numberOFlogFILESall}[0]{%
+    522%
+}
+
 \newcommand{\numberOFrules}[0]{%
     169%
+}
+
+\newcommand{\numberOFrulesMINIMUM}[0]{%
+    115%
+}
+
+% \numberOFrulesMINIMUM as percentage of \numberOFrules
+\newcommand{\numberOFrulesMINIMUMpercentage}[0]{%
+    68.04\%%
+}
+
+% \numberOFrules as percentage increase of \numberOFrulesMINIMUM
+\newcommand{\numberOFrulesMAXIMUMpercentage}[0]{%
+    46.95\%%
 }
@@
 
-XXX CHECK Y-AXIS LABELS ON GRAPHS TO ENSURE THEY CAN BE READ, AND WRAP IF
-NECESSARY\@.
+XXX GENERATE RESULT\_COLS AND CONNECTION\_COLS AUTOMATICALLY\@.
 
-XXX EXPLAIN THE X-AXIS ON GRAPHS --- MADS THOUGHT IT WAS UNCLEAR\@.
-
-XXX CHECK THE CAPTIONS ON ALL GRAPHS AND TABLES --- SOME ARE AWFUL\@.
+XXX NEED A NAME FOR THE ARCHITECTURE, AND A DIFFERENT NAME FOR THE
+IMPLEMENTATION\@.
 

### 2008-08-01 doc/logparser-background.tex
Added sample output for one of the SQL examples.

@@
 
-If the database supports sub-selects percentages can be obtained:
-\footnote{\SQL{} note: $||$ is the concatenation operator in SQLite3; if
-the database containing the extracted data does not support this syntax,
-then simply remove `` $||$ '$\%$'\hspace{1ex}'' from the query --- the
-results will be the same, just slightly less visually pleasing.}
+If the database supports sub-selects percentages can be obtained for the
+top ten restrictions:
 
 \begin{verbatim}
-SELECT name, description, restriction_name, hits_total,
+SELECT restriction_name, hits_total,
         (hits_total * 100.0 /
@@
     WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC;
+    ORDER BY hits_total DESC
+    LIMIT 10;
 \end{verbatim}
 
-XXX GIVE SAMPLE OUTPUT FOR THE SNIPPET ABOVE\@.
+\begin{table}[ht]
+    \caption{Truncated sample output from SQL query showing percentages}
+    \empty{}\label{Truncated sample output from SQL query showing
+    percentages}
+    \begin{tabular}[]{lrr}
+        \tableline{}
+        Restriction & Number of hits & Percentage of hits \\
+        \tableline{}%
+        \input{build/sample-sql-output-include.tex}
+        \tableline{}
+    \end{tabular}
+\end{table}
 

### 2008-08-06 doc/logparser-background.tex
Part two of the great renaming complete. Graphs still need to be renamed.

@@
 If the database supports sub-selects percentages can be obtained for the
-top ten restrictions:
+top ten restrictions (sample output shown in \tableref{Truncated sample
+output from SQL query showing percentages}):
+
+XXX GENERATE THE FOLLOWING FROM THE SQL FILE FED TO SQLITE\@.
 
@@
         \tableline{}%
-        \input{build/sample-sql-output-include.tex}
+        \input{build/include-sample-sql-output}
         \tableline{}
@@
             \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/restriction-table-include.tex}
+            \input{build/include-restriction-summary-table}
         \end{table}

### 2008-08-06 doc/logparser-results.tex
Part two of the great renaming complete. Graphs still need to be renamed.

@@
 
-\showtable{build/mails-received-include-for-graph}{Number of mails received
-via SMTP per day}{Number of mails received per day: statistics}
+\showtable{build/include-mails-received-table}{Number of mails received via
+SMTP per day}{Number of mails received per day: statistics}
 
@@
 Parsing all \numberOFlogFILES{} log files in one run took
-\input{build/timing-run-include.tex}.  The computer was dedicated to the
-task of gathering statistics from test runs, and was not used for any other
-purpose while test runs were ongoing; any services not necessary for
+\input{build/include-timing-run-duration}.  The computer was dedicated to
+the task of gathering statistics from test runs, and was not used for any
+other purpose while test runs were ongoing; any services not necessary for
 running the tests were disabled.
@@
 
-\showtable{build/stats-normal-filesize-line-count-include}{Ratio of file
-size and number of lines to parsing time}{parsing time vs file size vs
-number lines factor table}
+\showtable{build/include-file-size-and-number-of-lines-vs-parsing-time}{Ratio
+of file size and number of lines to parsing time}{parsing time vs file size
+vs number lines factor table}
 
@@
 
-\showtable{build/stats-normal-shuffle-reverse-include}{Parsing time
+\showtable{build/include-normal-and-reverse-vs-shuffle}{Parsing time
 relative to shuffled ordering}{Parsing time relative to shuffled ordering
@@
 matching rules are tried first) shows a mean increase of
-\input{build/full-ruleset-vs-minimum-ruleset-include.tex} in parsing time
-for a \numberOFrulesMAXIMUMpercentage{} increase in the number of rules.
-Log files 22 and 62--68 show much smaller increases for shuffled and
-reversed ordering than other log files do, because the majority of log
-lines in those log files are parsed by Postfix components with a small
-number of parsing rules, so removing unnecessary rules has little effect on
-the total number of rules used.  These results show that the architecture
-scales very well as the number of rules increases, and that optimally
-sorting the rules is an important optimisation contributing to this
-scalability.
+\input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
+\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  Log
+files 22 and 62--68 show much smaller increases for shuffled and reversed
+ordering than other log files do, because the majority of log lines in
+those log files are parsed by Postfix components with a small number of
+parsing rules, so removing unnecessary rules has little effect on the total
+number of rules used.  These results show that the architecture scales very
+well as the number of rules increases, and that optimally sorting the rules
+is an important optimisation contributing to this scalability.
 
@@
 
-\showtable{build/stats-cached-discarded-include-for-graph} {Increase in
+\showtable{build/include-cached-regexes-vs-discarded-regexes} {Increase in
 parsing time when not caching compiled regexes} {Increase in parsing time

### 2008-08-07 doc/logparser-results.tex
Next step in the great renaming - renaming of graphs. In the process I removed some generated gnuplot input files from the repository.

@@
 
-\showgraph{build/plot-mails-received}{Number of mails received via SMTP per
+\showgraph{build/graph-mails-received}{Number of mails received via SMTP per
 day}{Mails received per day}
@@
 
-\showgraph{build/plot-normal-filesize-numlines}{Parsing time, file size,
+\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
 and number of lines}{parsing time vs file size vs number of lines graph}
 
-\showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file size
+\showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of file size
 and number of lines to parsing time}{parsing time vs file size vs number
@@
 
-\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Parsing time
-relative to shuffled ordering}{Parsing time relative to shuffled ordering
-graph}
+\showgraph{build/graph-normal-and-reverse-vs-shuffle}{Parsing time relative
+to shuffled ordering}{Parsing time relative to shuffled ordering graph}
 
@@
 
-\showgraph{build/plot-full-ruleset-vs-minimum-ruleset}{Percentage parsing
+\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
 time increase of maximum ruleset over minimum ruleset}{Percentage parsing
@@
 
-\showgraph{build/plot-cached-discarded-factor}{Increase in parsing time
-when not caching compiled regexes}{Increase in parsing time when not
-caching compiled regexes graph}
+\showgraph{build/graph-cached-regexes-vs-discarded-regexes}{Increase in
+parsing time when not caching compiled regexes}{Increase in parsing time
+when not caching compiled regexes graph}
 
@@
 
-\showgraph{build/plot-hits}{Hits per rule}{rule hits graph}
+\showgraph{build/graph-hits}{Hits per rule}{rule hits graph}
 

### 2008-08-07 doc/logparser-results.tex
Changed normal to optimal.

@@
 XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
-
-XXX CHANGE normal TO optimal IN ORDERING STUFF, INCLUDING IN GRAPH CAPTIONS
-AND AXIS LABELS\@.
 
@@
 
-    \item [normal]  The most optimal order, according to the hypothesis:
+    \item [optimal]  The most optimal order, according to the hypothesis:
         rules which match most often will be tried first.
@@
 \Graphref{Parsing time relative to shuffled ordering graph} shows the
-parsing times of normal and reverse orderings relative to shuffled
+parsing times of optimal and reverse orderings relative to shuffled
 ordering; the mean relative parsing times for different groupings of log
@@
 
-\showgraph{build/graph-normal-and-reverse-vs-shuffle}{Parsing time relative
+\showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time relative
 to shuffled ordering}{Parsing time relative to shuffled ordering graph}
 
-\showtable{build/include-normal-and-reverse-vs-shuffle}{Parsing time
+\showtable{build/include-optimal-and-reverse-vs-shuffle}{Parsing time
 relative to shuffled ordering}{Parsing time relative to shuffled ordering
@@
 ruleset: the percentage parsing time increase when using the full ruleset
-instead of the minimal ruleset for normal, shuffled and reversed orderings
+instead of the minimal ruleset for optimal, shuffled and reversed orderings
 is shown in \graphref{Percentage parsing time increase of maximum ruleset
@@
 increases parsing time with reverse ordering, and to a lesser extent with
-shuffled ordering.  The normal ordering (where are the most frequently
+shuffled ordering.  The optimal ordering (where the most frequently
 matching rules are tried first) shows a mean increase of

### 2008-08-07 doc/sgai-2008.tex
Moved SGAI 2008 paper to this directory, to avoid replicating lots of stuff in a different directory. Made the required modifications to have it all work.

--- 2008-07-30 papers/sgai-2008/paper.tex
@@
 XXX REVIEWER SUGGESTED COMPARING TO EXISTING WORK, BUT I DID NOT FIND ANY
-SIMILAR EXISTING WORK\@.  SHOULD SEARCH AGAIN\@.
+SIMILAR EXISTING WORK\@.  I SHOULD REALLY SEARCH AGAIN\@.
 
@@
 new rule and an existing action.
-\showgraph{build/plot-action-distribution}{Distribution of rules per
+\showgraph{build/graph-action-distribution}{Distribution of rules per
 action}{Distribution of rules per action} Unsurprisingly the action with
@@
 lines}); this is explained in \sectionref{Why are there dips in the
-graphs?}.  \showgraph{build/plot-normal-filesize-numlines}{Parsing time,
+graphs?}.  \showgraph{build/graph-input-size-vs-parsing-time}{Parsing time,
 file size, and number of lines}{Parsing time, file size, and number of
-lines} \showgraph{build/plot-normal-filesize-numlines-factor}{Ratio of file
-size and number of lines to parsing time}{parsing time vs file size vs
+lines} \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of
+file size and number of lines to parsing time}{parsing time vs file size vs
 number lines factor}
@@
 optimisation provides a modest but worthwhile reduction in parsing time of
-approximately 10\%.  \showgraph{build/plot-hits}{Hits per rule}{rule hits
-graph}
-\showgraph{build/plot-normal-and-reverse-relative-to-shuffle}{Parsing time
+approximately 10\%.  \showgraph{build/graph-hits}{Hits per rule}{rule hits
+graph} \showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
 relative to shuffled ordering}{Parsing time relative to shuffled}
+
+\newpage{}
 
@@
 \graphref{Percentage parsing time increase of maximum ruleset over minimum
-ruleset}.  \showgraph{build/plot-normal-vs-smaller}{Percentage parsing time
-increase of maximum ruleset over minimum ruleset}{Percentage parsing time
-increase of maximum ruleset over minimum ruleset}
+ruleset}.
+\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
+time increase of maximum ruleset over minimum ruleset}{Percentage parsing
+time increase of maximum ruleset over minimum ruleset}
 
@@
 matching rules are tried first) shows a mean increase of
-\input{build/normal-vs-smaller-percentage-increase.tex} in parsing time for
-a \numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
+\input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
+\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
 results show that the architecture scales very well as the number of rules
@@
 \bibliographystyle{spmpsci}
-\bibliography{../common/bibliography}
+\bibliography{logparser-bibliography.bib}
 

### 2008-08-11 doc/sgai-2008.tex
Dealt with some XXX items.

@@
 
+\newcommand{\tableline}[0]{%
+    \noalign{\smallskip}%
+    \hline%
+    \noalign{\smallskip}%
+}
+
 % Replacement for \ref{}, adds the page number too.
@@
     \refwithlabel{Figure}{#1}%
+}
+\newcommand{\tableref}[1]{%
+    \refwithlabel{tbl.}{#1}%
+}
+\newcommand{\Tableref}[1]{%
+    \refwithlabel{Table}{#1}%
 }
@@
 
-The framework provides utility and support functions for the actions and
-rules.  It provides shared storage to pass data between actions, most
-notably the data structures used to represent each connection and mail.  It
-loads and validates rules, controls parsing of log files, invokes actions,
-tracks how often each rule matches, to optimise rule ordering (see
-\sectionref{Rule ordering}), and stores results in the database.  The
-framework is independent of the parser being written: the same
-functionality will be required for most or all parsers.
+The framework takes care of miscellaneous support functions and low level
+details of parsing, freeing the authors of actions to concentrate on
+writing productive code.  It provides the link between actions and rules,
+allowing either to be improved independently of the other.  It provides
+shared storage to pass data between actions, loads and validates rules,
+controls parsing of log files, invokes actions, tracks how often each rule
+matches (to optimise rule ordering \sectionref{Rule ordering}), and
+stores results in the database.  The framework is largely independent of
+the parser being written: the same functionality will be required for most
+or all parsers.  The framework is the core of the architecture and is
+deliberately quite simple: the rules deal with the varying log lines, and
+the actions deal with the intricacies and complications encountered when
+parsing.
 
 The function which finds the rule matching the log line and invokes the
-requested action can be expressed in pseudo-code (with indentation denoting
-flow of control):
+requested action can be expressed in pseudo-code as:
 
@@
 
-The framework takes care of miscellaneous support functions and low level
-details of parsing, freeing the authors of actions to concentrate on
-writing productive code.  It provides the link between actions and rules,
-allowing either to be improved independently of the other.  The framework
-is the core of the architecture and is deliberately quite simple: the rules
-deal with the varying log lines, and the actions deal with the intricacies
-and complications encountered when parsing.  XXX MAYBE REMOVE FIRST TWO
-SENTENCES\@.
-
 \subsection{Actions}
@@
 \label{Actions}
-
-XXX CARL WONDERS IF THE NUMBERS HERE ARE TYPICAL\@?  ALSO SAY SOMETHING
-LIKE REJECTION AND IGNORE DO NOT IMPLY UNGRAMMATICAL\@.
 
@@
 parse new log lines: most new log lines can be parsed by a combination of a
-new rule and an existing action.
-\showgraph{build/graph-action-distribution}{Distribution of rules per
-action}{Distribution of rules per action} Unsurprisingly the action with
-the highest number of associated rules is \texttt{REJECTION}, the action
-which handles Postfix rejecting a mail; it is followed by
+new rule and an existing action.  It is expected that parsers utilising
+this architecture will have a high ratio of rules to actions, XXX FINISH
+THIS SENTENCE\@.  \showgraph{build/graph-action-distribution}{Distribution
+of rules per action}{Distribution of rules per action} Unsurprisingly the
+action with the highest number of associated rules is \texttt{REJECTION},
+the action which handles Postfix rejecting a mail; it is followed by
 \texttt{SAVE\_BY\_QUEUEID}, the action responsible for handling informative
@@
 \texttt{IGNORE}: the \texttt{IGNORE} action does nothing when executed,
-allowing uninteresting log lines to be parsed without causing any effects.
-The parser warns about every unparsed log line, to alert the user that they
-need to alter or extend their ruleset; the Postfix parser successfully
-parses every input line in the \numberOFlogFILESall{} log files it is
-currently being tested with.  The remaining actions have only one or two
-associated rules: in some cases this is because there will only ever be one
-type of log line for that action, e.g.\ all log lines showing that a remote
-client has connected are matched by a single rule and handled by the
-\texttt{CONNECT} action; in other cases the action is required to address a
-deficiency in the log files, or a complication which arises during parsing.
+allowing uninteresting log lines to be parsed without causing any effects
+(neither \texttt{REJECTION} nor \texttt{IGNORE} imply that the input is
+ungrammatical or unparsed).  The parser warns about every unparsed log
+line, to alert the user that they need to alter or extend their ruleset;
+the Postfix parser successfully parses every input line in the
+\numberOFlogFILESall{} log files it is currently being tested with.  The
+remaining actions have only one or two associated rules: in some cases this
+is because there will only ever be one type of log line for that action,
+e.g.\ all log lines showing that a remote client has connected are matched
+by a single rule and handled by the \texttt{CONNECT} action; in other cases
+the action is required to address a deficiency in the log files, or a
+complication which arises during parsing.
 
@@
 structure for each client connecting, and saves the data extracted by the
-rule (shown in \sectionref{example rule}).  If the data structure already
-exists it is treated as a symptom of a bug, and the action issues a warning
-containing the full contents of the data structure, plus the log line which
-has just been parsed.
+rule.  If the data structure already exists it is treated as a symptom of a
+bug, and the action issues a warning containing the full contents of the
+data structure, plus the log line which has just been parsed.
 
@@
 
-This example rule matches the message logged by Postfix when a remote
-client connects to deliver mail; it would match the following sample log
-line:
+The example rule in \tableref{Example rule} matches the message logged by
+Postfix when a remote client connects to deliver mail; it would match the
+following sample log line:
 
@@
 % Do not reformat this!
-% XXX CONSIDER USING A TABLE ENVIRONMENT; IT DID NOT LOOK GREAT BEFORE.
+\begin{table}[htbp]
+    \caption{Example rule}
+    \empty{}\label{Example rule}
     \begin{tabular}[]{ll}
-    \hline
-    \noalign{\smallskip}
+        \tableline{}%
         Attribute                 & Value                                            \\
-    \noalign{\smallskip}
-    \hline
-    \noalign{\smallskip}
+        \tableline{}%
         program                   & \daemon{smtpd}                                   \\
@@
         priority                  & 0                                                \\
-    \noalign{\smallskip}
-    \hline
-    \noalign{\smallskip}
+        \tableline{}%
     \end{tabular}
+\end{table}
 
@@
 of the database and the disks the database is stored on.  The computer used
-for test runs was a Dell Optiplex 745:
+for test runs was a Dell Optiplex 745 described in \tableref{Computer used
+to generate statistics}.
 
+\begin{table}[htbp]
+    \caption{Computer used to generate statistics}
+    \empty{}\label{Computer used to generate statistics}
     \begin{tabular}[]{ll}
-    \hline
-    \noalign{\smallskip}
-    XXX         & XXX NEED LABELS\@.                                \\
-    \noalign{\smallskip}
-    \hline
-    \noalign{\smallskip}
+        \tableline{}
+        Component type  & Component in use                                  \\
+        \tableline{}
         CPU             & One dual core 2.40GHz Intel\textregistered{}
@@
         Hard disk       & One Seagate Barracuda 7200 RPM 250GB SATA disk.   \\
-    \noalign{\smallskip}
-    \hline
-    \noalign{\smallskip}
+        \tableline{}
     \end{tabular}
+\end{table}
 
@@
 relative to shuffled ordering}{Parsing time relative to shuffled}
-
-\newpage{}
 

### 2008-08-11 doc/logparser-background.tex
Dealt with an XXX item - including a sample SQL query instead of duplicating it.

@@
 If the database supports sub-selects percentages can be obtained for the
-top ten restrictions (sample output shown in \tableref{Truncated sample
-output from SQL query showing percentages}):
+top ten restrictions (sample output shown in \tableref{Sample output from
+SQL query showing percentages}):
 
-XXX GENERATE THE FOLLOWING FROM THE SQL FILE FED TO SQLITE\@.
-
-\begin{verbatim}
-SELECT restriction_name, hits_total,
-        (hits_total * 100.0 /
-            (SELECT SUM(hits_total)
-                FROM rules
-                WHERE postfix_action = 'REJECTED'
-            )
-        ) || '%' AS percentage
-    FROM rules
-    WHERE postfix_action = 'REJECTED'
-    ORDER BY hits_total DESC
-    LIMIT 10;
-\end{verbatim}
+\input{build/include-sample-sql-query}
 
 \begin{table}[ht]
-    \caption{Truncated sample output from SQL query showing percentages}
-    \empty{}\label{Truncated sample output from SQL query showing
+    \caption{Sample output from SQL query showing percentages}
+    \empty{}\label{Sample output from SQL query showing
     percentages}

### 2008-08-12 doc/sgai-2008.tex
Dealt with some XXX items: * Mentioned prior art. * Added a little more about detecting overlapping regexes. * Removed "Why are there dips in the graphs?" paragraphs, they weren't   particularly important any more. * Changed normal to optimal.

@@
 \newcommand{\showgraph}[3]{%
-    \begin{figure}[hbt!]%
+    \begin{figure}[hbtp]%
         \caption{#2}\label{#3}%
@@
 \newcommand{\showtable}[3]{
-    \begin{table}[ht]
+    \begin{table}[hbtp]
         \caption{#2}\label{#3}
@@
 
-XXX DISCUSS COMPLEXITY\@?  EVEN JUST MENTION THAT COMPLEXITY RISES WHEN
-INCREMENTAL PARSING IS USED, BUT THAT EXTERNAL TOOLS COULD BE USED TO DEAL
-WITH IT, AND THAT IT IS ENTIRELY DEPENDANT ON THE RULESET USED\@?
+XXX DISCUSS COMPLEXITY\@?
 
-XXX MAYBE MERGE THE AUTHOR SECTIONS IF DESPERATE FOR SPACE\@.
-
-XXX REVIEWER SUGGESTED COMPARING TO EXISTING WORK, BUT I DID NOT FIND ANY
-SIMILAR EXISTING WORK\@.  I SHOULD REALLY SEARCH AGAIN\@.
-
-XXX OVERLAP DETECTION\@: L1 n L2 = E* --- ((E* --- L1) u (E* --- L2)).
-PERL REGEXES ARE AT LEAST AS POWERFUL AS A PUSHDOWN AUTOMATA --- SEE
-CHOMSKY'S HIERARCHY TO MAKE SURE\@.  CONTAINMENT CANNOT BE DETERMINED FOR
-CFGs, INTERSECTION IS POSSIBLE\@.
+XXX MAYBE MERGE THE AUTHOR SECTIONS IF DESPERATE\@.
 
@@
 type, e.g.\ rejections, and extracted only a portion of the data required.
+The only prior published work on the subject of parsing Postfix log files
+is~\cite{log-mail-analyser}, which aims to present correlated log data in a
+form suitable for a systems administrator to search using the myriad of
+standard Unix text processing utilities already available.  A full state of
+the art review is outside the scope of this paper but will be included in
+the thesis resulting from this work.
 
@@
 number of successful matches they had parsing the previous log file.
-Negative priorities may be useful for catchall rules.  XXX DETECTING
-OVERLAP\@.
+Negative priorities may be useful for catchall rules.
+
+Storing the rules separately to the parser allows external tools to be
+written to detect overlapping rules.  Traditional regexes are equivalent in
+computational power to Finite Automata (FA); it is possible to convert
+between regexes and FA, so regex overlap could be calculated by finding the
+intersection of two FA XXX REFERENCE\@.  Perl 5.10 regexes are more
+powerful than traditional regexes, e.g.\ it is possible to match correctly
+nested and balanced brackets to an arbitrary depth:
+\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
+Perl 5.10 regexes can maintain an arbitrary state stack and are thus
+equivalent in computational power to Pushdown Automata (PA), so overlap
+detection would require calculating the intersection of two PA.\@  XXX
+REFERENCE OR SOMETHING\@.
+
+XXX OVERLAP DETECTION\@: L1 n L2 = E* --- ((E* --- L1) u (E* --- L2)).
+CONTAINMENT CANNOT BE DETERMINED FOR CFGs, INTERSECTION IS POSSIBLE\@.
 
@@
         actions when executing, providing some context sensitivity.
-        Actions can not inspect past or upcoming inputs, however actions
+        Actions can not inspect past or future inputs, however actions
         can return a modified input line that will be parsed as if read
@@
 average size (shown in \graphref{Parsing time, file size, and number of
-lines}); this is explained in \sectionref{Why are there dips in the
-graphs?}.  \showgraph{build/graph-input-size-vs-parsing-time}{Parsing time,
+lines}).  \showgraph{build/graph-input-size-vs-parsing-time}{Parsing time,
 file size, and number of lines}{Parsing time, file size, and number of
@@
 
-\newpage
 \subsection{Rule Ordering}
@@
 
-    \item [normal]  Hypothetically the most optimal order: rules which
+    \item [optimal]  Hypothetically the most optimal order: rules which
         match most often will be tried first.
@@
 \Graphref{Parsing time relative to shuffled} shows the parsing times of
-normal and reversed orderings relative to shuffled ordering.  Overall this
+optimal and reversed orderings relative to shuffled ordering.  Overall this
 optimisation provides a modest but worthwhile reduction in parsing time of
@@
 relative to shuffled ordering}{Parsing time relative to shuffled}
-
-\noindent\textbf{Why are there dips in the graphs?}
-
-XXX MAYBE REDUCE THIS\@?
-
-\label{Why are there dips in the graphs?}
-
-The dips at log files 22 and 62--68 in \graphref{Parsing time relative to
-shuffled} correspond to peaks in log file size in \graphref{Parsing time,
-file size, and number of lines}, and peaks in \graphref{parsing time vs
-file size vs number lines factor}, where a peak means that performance is
-better.  The explanation for this took some time to arrive at, but it turns
-out to be reasonably simple.  The large log files in question were caused
-by a mail forwarding loop, where the distribution of log lines is quite
-different to normal, resulting in different performance characteristics.
-The vast majority of log lines when a mail loop occurs are from Postfix
-components which have a small number of rules associated with them, whereas
-in general the majority of log lines are from the Postfix component
-(\daemon{smtpd}) with the highest number of rules.  Furthermore in normal
-log files the majority of log lines are distributed much more evenly across
-rules than the majority of log lines during a mail loop.
-
-\newpage
-
-These two characteristics combine to reduce the average number of rules
-required to parse a log line when there is a mail loop, as shown by the
-peaks in \graphref{parsing time vs file size vs number lines factor} and
-troughs in \graphref{Parsing time relative to shuffled}.  When the rule
-ordering is reversed the majority of log lines generated by a mail loop
-will be parsed using very few rules, because there are very few rules for
-the Postfix components producing the log lines.  Without a mail loop the
-Postfix components producing the majority of log lines have a large number
-of rules associated with them, and reverse ordering deliberately starts
-with the least likely rules and moves towards the most likely.  This leads
-to a noticeable drop in the average time required to parse a log line
-during a mail loop, as shown in \graphref{Parsing time relative to
-shuffled}.  The number of rules which need to be consulted when the
-ordering is shuffled varies between the optimal and the pessimal, with the
-performance varying proportionally.
 
@@
 percentage parsing time increase when using the full ruleset instead of the
-minimal ruleset for normal, shuffled and reversed orderings is shown in
+minimal ruleset for optimal, shuffled and reversed orderings is shown in
 \graphref{Percentage parsing time increase of maximum ruleset over minimum
@@
 noticeable performance impact with reverse ordering, and to a lesser extent
-with shuffled ordering.  The normal ordering (where are the most frequently
+with shuffled ordering.  The optimal ordering (where the most frequently
 matching rules are tried first) shows a mean increase of

### 2008-08-12 doc/sgai-2008.tex
Some more improvements to regex overlap and the introduction.

@@
 
-XXX MAYBE MERGE THE AUTHOR SECTIONS IF DESPERATE\@.
-
 \section{Introduction}
@@
 performance was to analyse the log files produced by the mail server,
-rather than modifying the mail server to generate performance statistics;
+rather than modifying the mail server to generate performance statistics:
 this approach improves the chances of other sites testing and utilising the
@@
 SMTP server in use.  The parser must be designed so that adding support for
-parsing new log lines is a simple task, as the log lines will continue to
-evolve over time.  The variety in log lines occurs for several
+parsing new log lines is a simple task, as the log lines to be parsed will
+change over time.  The variety in log lines occurs for several
 reasons:\footnotetext{Simple Mail Transfer Protocol transfers mail from the
@@
 
-Existing parsers were considered, but their parsing coverage was too low,
-their parsing too inexact, and/or they did not extract sufficient data.
-The effort required to adapt and improve an existing parser was judged to
-be greater than the effort to write one from scratch, as the techniques
-used by the existing parsers severely limited their potential: some ignored
-the majority of log lines, parsing specific log lines with great accuracy,
-but in an inextensible fashion; others sloppily parsed the majority of log
-lines, but were incapable of distinguishing between log lines of the same
-type, e.g.\ rejections, and extracted only a portion of the data required.
-The only prior published work on the subject of parsing Postfix log files
+It was hoped to reuse an existing parser rather than starting from scratch:
+several parsers were considered and rejected because their parsing coverage
+was too low, their parsing too inexact, and/or they did not extract
+sufficient data.  The effort required to adapt and improve an existing
+parser was judged to be greater than the effort to write one, as the
+techniques used by the existing parsers severely limited their potential:
+some ignored the majority of log lines, parsing specific log lines with
+great accuracy, but without any provision for readily parsing new or
+similar log lines; others sloppily parsed the majority of log lines, but
+were incapable of distinguishing between log lines of the same type, e.g.\
+rejections, and extracted only a portion of the data required.  The only
+prior published work on the subject of parsing Postfix log files
 is~\cite{log-mail-analyser}, which aims to present correlated log data in a
@@
 new rule and an existing action.  It is expected that parsers utilising
-this architecture will have a high ratio of rules to actions, XXX FINISH
-THIS SENTENCE\@.  \showgraph{build/graph-action-distribution}{Distribution
+this architecture will have a high ratio of rules to actions, hopefully
+resulting in simpler rules and clearer distinctions between log lines.
+\showgraph{build/graph-action-distribution}{Distribution
 of rules per action}{Distribution of rules per action} Unsurprisingly the
@@
 computational power to Finite Automata (FA); it is possible to convert
-between regexes and FA, so regex overlap could be calculated by finding the
-intersection of two FA XXX REFERENCE\@.  Perl 5.10 regexes are more
-powerful than traditional regexes, e.g.\ it is possible to match correctly
-nested and balanced brackets to an arbitrary depth:
+between regexes and FA, so regex overlap could be calculated by finding a
+non-empty intersection of two FA\@.  Perl 5.10 regexes are more powerful
+than traditional regexes, e.g.\ it is possible to match correctly nested
+and balanced brackets to an arbitrary depth:
 \newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
@@
 equivalent in computational power to Pushdown Automata (PA), so overlap
-detection would require calculating the intersection of two PA.\@  XXX
-REFERENCE OR SOMETHING\@.
-
-XXX OVERLAP DETECTION\@: L1 n L2 = E* --- ((E* --- L1) u (E* --- L2)).
-CONTAINMENT CANNOT BE DETERMINED FOR CFGs, INTERSECTION IS POSSIBLE\@.
+detection may require calculating the intersection of two PA.\@  Whether
+intersection of FA or PA is required, detecting overlap between $n$ rules
+has $\mathcal{O}(n(n-1)/2)$ complexity.  XXX IMPROVE THE MATH DISPLAY\@.
 

### 2008-08-13 doc/sgai-2008.tex
Removed list of rule attributes. Improved lots of little pieces. Improved overlap discussion.

@@
 
+% Embed SVN Id etc.
+\usepackage{svn}
+
+% XXX REMOVE THESE AND CHECK RESULTING LENGTH
 \setlength{\floatsep}{0pt}
@@
 
+\newcommand{\numberOFruleINTERSECTIONS}[0]{%
+    14196%
+}
+
 \newcommand{\numberOFrulesMINIMUM}[0]{%
@@
 }
+
+\SVN$Id$
+\SVNId{}  \today{}
 
@@
 
-    \item [Actions]  The actions perform the work required to deal with a
+    \item [Actions]  Each action performs the work required to deal with a
         category of log lines, e.g.\ processing data from rejections.
@@
 controls parsing of log files, invokes actions, tracks how often each rule
-matches (to optimise rule ordering \sectionref{Rule ordering}), and
-stores results in the database.  The framework is largely independent of
-the parser being written: the same functionality will be required for most
-or all parsers.  The framework is the core of the architecture and is
-deliberately quite simple: the rules deal with the varying log lines, and
-the actions deal with the intricacies and complications encountered when
-parsing.
+matches to optimise rule ordering (\sectionref{Rule ordering}), and stores
+results in the database.  The framework is largely independent of the
+parser being written: the same functionality will be required for most or
+all parsers.  The framework is the core of the architecture and is
+deliberately quite simple: the rules deal with the variation in log lines,
+and the actions deal with the intricacies and complications encountered
+when parsing.
 
@@
 less frequently than with a monolithic architecture, thus failures should
-be less likely, and easier to test for and diagnose.  The parser can be
-written in an object oriented style, allowing sub-classes to extend or
-override actions in addition to adding new actions; because each action is
-an independent function, the sub-class need only modify the action it is
+be less likely, and will be easier to test for and diagnose.  The parser
+can be written in an object oriented style, allowing sub-classes to extend
+or override actions in addition to adding new actions; because each action
+is an independent function, the sub-class need only modify the action it is
 interested in, rather than reproducing large chunks of functionality.
@@
 be overcome.  Some were the result of deficiencies in Postfix's logging
-(some of those deficiencies were improved or resolved by later versions of
-Postfix); others were due to the vagaries of process scheduling, client
-behaviour, and administrative actions.  All were successfully accommodated
-in the parser: adding new actions was enough for several of the
-complications; others required modifications to a single existing action to
-work around the difficulties; the remainder were resolved by adapting
-actions to pass information to subsequent actions (via the framework),
-those subsequent actions then check for the extra information and change
-their behaviour appropriately.
+(some of which were improved on or resolved by later versions of Postfix);
+others were due to the vagaries of process scheduling, client behaviour,
+and administrative actions.  All were successfully accommodated in the
+parser: adding new actions was enough for several of the complications;
+others required modifications to a single existing action to work around
+the difficulties; the remainder were resolved by adapting actions to pass
+information to subsequent actions (via the framework), which check for the
+extra information and change their behaviour appropriately.
 
@@
 
-Rules categorise inputs, specifying both the data to be extracted from each
-log line and the action to invoke.  The Postfix parser stores the rules in
-the same SQL database that the results are stored in, removing any doubt as
-to which set of rules was used to produce a set of results; other
-implementations are free to store the rules in whatever fashion suits their
-needs.  Similarly implementations are free to require additional attributes
-in their rules; the framework requires \texttt{regex} and \texttt{action}
-attributes, to determine whether a rule matches the current input line, and
-which action to invoke if it does.  Rules used by the Postfix parser
-define the following attributes: XXX DO I NEED TO LIST ALL THE
-ATTRIBUTES\@?
+Rules categorise inputs, specifying both the regex to match against each
+log line and the action to invoke when the match is successful.  The
+Postfix parser stores its rules in the same SQL database the results are
+stored in, removing any doubt as to which set of rules was used to produce
+a set of results; other implementations are free to store the rules in
+whatever fashion suits their needs.
 
-\begin{description}
+The framework requires that each rule have \texttt{action} and
+\texttt{regex} attributes; each implementation is free to add any
+additional attributes it requires.  The Postfix parser adds several
+attributes to optimise rule ordering (\sectionref{Rule ordering}), to
+restrict which log lines rules can be matched against, and to describe each
+rule.  Data is automatically extracted and saved from the log line based on
+keywords in the rule's regex, but there is also a facility to specify extra
+data to be saved.
 
-    \item [program] The Postfix program whose log lines the regex should be
-        applied to.  This avoids needlessly trying regexes that will not
-        match the log line, or worse, might match unintentionally.  Rules
-        whose program is ``\texttt{*}'' will be tried against any log lines
-        not parsed by program specific rules, allowing generic rules to be
-        written.  This is a literal ``\texttt{*}'', not a wildcard or
-        regex.
-
-    \item [regex] The Regular Expression to match the log line against.
-
-    \item [result\_cols] Specifies how the fields in the log line will be
-        extracted.  The format is:
-        \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline
-        i.e.\ semi-colon or comma separated assignment statements, with the
-        variable name on the left and the matching capture from the regex
-        on the right.
-
-    \item [result\_data] Sometimes rules need to supply a piece of data
-        which is not present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format is the same
-        as for \texttt{result\_cols}, except that arbitrary data is
-        permitted on the right hand side of the assignment.
-
-    \item [action] The action that will be invoked when this rule matches a
-        log line.
-
-    \item [hits] This counter is used to optimise the order in which rules
-        are tried against the input lines.  Assuming that the distribution
-        of log lines is reasonably consistent between log files, rules
-        matching more frequently will be tried before rules matching less
-        frequently, reducing the parser's execution time.
-
-    \item [priority] Priority overrides hits when sorting rules, allowing
-        more specific rules to take precedence over more general rules.  To
-        illustrate: most warnings are ignored by a rule whose regex matches
-        \verb!^warning:.*$!, but some warnings require a different action
-        to be taken, e.g.\ when a client tries to send a mail larger than
-        the maximum size the server accepts, the data structure for that
-        mail must be discarded because it will not have all the data
-        required to save it to the database.
-
-\end{description}
-
-Adding new rules is simple: simply define all the required attributes and
-append the new rule to wherever the existing rules are stored.  The
-software developed for this project includes a utility based on
-SLCT~\cite{slct-paper} to aid in producing regexes to parse previously
-unparsed inputs.  Storing the rules separately allows other approaches to
-rule management to be used, e.g.\ instead of manually adding new rules when
-unparsed log lines are seen, machine learning techniques could be used to
-automatically generate new rules, based on the existing rules and input
-lines.  If this approach was taken the choice of machine learning technique
-would be constrained by the size of typical data sets: the
-\numberOFlogFILES{} log files used to generate statistics contain
-\numberOFlogLINEShuman{} log lines.  Techniques which require the full data
-set when training would be impractical; Instance Based
-Learning~\cite{instance-based-learning} techniques which automatically
-determine which inputs from the training set are valuable and which inputs
-can be discarded would probably reduce the data required to a manageable
-size.  This avenue of research and development has not been pursued by the
-authors, but could easily be independently undertaken.
+Adding new rules is simple: define all the required attributes and append
+the new rule to wherever the existing rules are stored.  The Postfix parser
+supplies a utility based on SLCT~\cite{slct-paper} to aid in producing
+regexes from previously unparsed inputs.  Storing the rules separately
+allows other approaches to rule management to be used, e.g.\ instead of
+manually adding new rules when unparsed log lines are seen, machine
+learning techniques could be used to automatically generate new rules,
+based on the existing rules and input lines.  If this approach was taken
+the choice of machine learning technique would be constrained by the size
+of typical data sets: the \numberOFlogFILES{} log files used to generate
+statistics contain \numberOFlogLINEShuman{} log lines.  Techniques which
+require the full data set when training would be impractical; Instance
+Based Learning~\cite{instance-based-learning} techniques which
+automatically determine which inputs from the training set are valuable and
+which inputs can be discarded would probably reduce the data required to a
+manageable size.  This avenue of research and development has not been
+pursued by the authors, but could easily be independently undertaken.
 
@@
 rule to match the majority, e.g.\ separating SMTP delivery to specific
-sites from SMTP delivery to the rest of the world.  The algorithm provides
-a facility for specifying the order of overlapping rules: the priority
-field in each rule (defaults to zero).  Rules are sorted by priority,
-highest first, and then rules with the same priority are sorted by the
-number of successful matches they had parsing the previous log file.
-Negative priorities may be useful for catchall rules.
+sites from SMTP delivery to the rest of the world.  Overlapping rules
+should have a priority attribute to specify the order in which they should
+be used; negative priorities may be useful for catchall rules.
 
@@
 computational power to Finite Automata (FA); it is possible to convert
-between regexes and FA, so regex overlap could be calculated by finding a
+between regexes and FA, so regex overlap could be detected by finding a
 non-empty intersection of two FA\@.  Perl 5.10 regexes are more powerful
@@
 equivalent in computational power to Pushdown Automata (PA), so overlap
-detection may require calculating the intersection of two PA.\@  Whether
-intersection of FA or PA is required, detecting overlap between $n$ rules
-has $\mathcal{O}(n(n-1)/2)$ complexity.  XXX IMPROVE THE MATH DISPLAY\@.
+detection may require calculating the intersection of two PA.\@  Detecting
+overlap between $n$ regexes requires $\frac{n(n-1)}{2}$ intersections to be
+calculated, giving $O(n^2)$ complexity.  This is certainly not a task to be
+performed on every parsing run: detecting overlap amongst the Postfix
+parser's \numberOFrules{} rules would require \numberOFruleINTERSECTIONS{}
+intersections to be calculated.
 

### 2008-08-14 doc/sgai-2008.tex
First batch of corrections from proofreading.

@@
 % XXX REMOVE THESE AND CHECK RESULTING LENGTH
-\setlength{\floatsep}{0pt}
-\setlength{\intextsep}{0pt}
-\setlength{\textfloatsep}{0pt}
+%\setlength{\floatsep}{0pt}
+%\setlength{\intextsep}{0pt}
+%\setlength{\textfloatsep}{0pt}
 
@@
 content-based~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms},
-which requires accepting the mail in the first place, but rejecting is
-preferable in many cases: senders of non-spam mail that is mistakenly
-rejected will receive an immediate non-delivery notice; load is reduced on
-mail servers (allowing more intensive content-based techniques to be used);
-users have less mail to wade through.  Improving the performance of
-anti-spam techniques which are applied when mail is being transferred via
-SMTP\footnotemark{} is the goal of this project, by providing a platform
-for reasoning about anti-spam filters.  The approach chosen to measure
-performance was to analyse the log files produced by the mail server,
-rather than modifying the mail server to generate performance statistics:
-this approach improves the chances of other sites testing and utilising the
-software.  The need arose for a parser capable of dealing with the great
-number and variety of log lines produced by Postfix~\cite{postfix}, the
-SMTP server in use.  The parser must be designed so that adding support for
-parsing new log lines is a simple task, as the log lines to be parsed will
-change over time.  The variety in log lines occurs for several
-reasons:\footnotetext{Simple Mail Transfer Protocol transfers mail from the
-sender to one or more recipients across the Internet.  It is a simple,
-human readable, plain text protocol, making it quite easy to test and debug
-problems with it.  The original protocol definition is
-RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}
+which requires accepting the mail in the first place, but rejecting mail
+during the delivery attempt is preferable: senders of non-spam mail that is
+mistakenly rejected will receive an immediate non-delivery notice; load is
+reduced on mail servers (allowing more intensive content-based techniques
+to be used); users have less mail to wade through.  Improving the
+performance of anti-spam techniques which are applied when mail is being
+transferred via SMTP\footnotemark{} is the goal of this project, by
+providing a platform for reasoning about anti-spam filters.  The approach
+chosen to measure performance was to analyse the log files produced by the
+mail server, rather than modifying the mail server to generate performance
+statistics: this approach improves the chances of other sites testing and
+utilising the software.  The need arose for a parser capable of dealing
+with the great number and variety of log lines produced by
+Postfix~\cite{postfix}, the SMTP server in use.  The parser must be
+designed so that adding support for parsing new log lines is a simple task,
+as the log lines to be parsed will change over time.  The variety in log
+lines occurs for several reasons:\footnotetext{Simple Mail Transfer
+Protocol transfers mail from the sender to one or more recipients across
+the Internet.  It is a simple, human readable, plain text protocol, making
+it quite easy to test and debug problems with it.  The original protocol
+definition is RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}
 
@@
 were incapable of distinguishing between log lines of the same type, e.g.\
-rejections, and extracted only a portion of the data required.  The only
+rejections, and extracted only a small portion of the data required.  The only
 prior published work on the subject of parsing Postfix log files
@@
 a specific type, and only that type (e.g.\ rejection by a specific DNSBL).
-It is an accepted standard to separate the parsing procedure from the
-declarative grammar it operates with.  Part of the novelty here is that the
-grammars (rules and actions) are themselves procedures.  This architecture
-is ideally suited to parsing inputs where the input is not fully understood
-or does not follow a traditional grammar: the architecture warns about
-unparsed inputs and other errors, but continues on parsing as best it can,
-allowing the developer of a new parser to decide which deficiencies are
-most important and require attention first, rather than being forced to fix
-the first error which arises.
+XXX IMPROVE:\@ It is an accepted standard to separate the parsing procedure
+from the declarative grammar it operates with.  Part of the novelty here is
+that the grammars (rules and actions) are themselves procedures.  This
+architecture is ideally suited to parsing inputs where the input is not
+fully understood or does not conform to a fixed grammar: the architecture
+warns about unparsed inputs and other errors, but continues on parsing as
+best it can, allowing the developer of a new parser to decide which
+deficiencies are most important and require attention first, rather than
+being forced to fix the first error which arises.
 
@@
     \item [Framework]  The framework is the structure which actions and
-        rules fit into.  It provides the parsing loop, loading and
+        rules plug into.  It provides the parsing loop, loading and
         validation of rules, shared data storage, storage of results, and
@@
 
-The framework tries each rule in turn until it finds one which matches the
-input log line, then invokes the action specified by that rule.
+For each input log line the framework tries each rule in turn until it
+finds one which matches the input log line, then invokes the action
+specified by that rule.
 
@@
 Decoupling also creates a clear separation of functionality: rules handle
-low level details of identifying and extracting data from log lines;
-actions handle the higher level details of assembling the required data,
-dealing with the intricacies of the input being parsed, complications
-arising, etc.;\ the framework provides services to actions and manages the
-parsing process.
+low level details of identifying log lines and extracting data; actions
+handle the higher level details of assembling the required data, dealing
+with the intricacies of the input being parsed, complications arising,
+etc.;\ the framework provides services to actions and manages the parsing
+process.
 
-There is some similarity between the parser's design and William Wood's
+There is some similarity between the architecture and William Wood's
 Augmented Transition Networks (ATN)~\cite{atns, nlpip}, a tool used in
@@
 it is interesting how two different approaches have a similar division of
-responsibilities, while having completely different implementations and
-semantics.
+responsibilities, while having different semantics.
 
@@
 parser being written: the same functionality will be required for most or
-all parsers.  The framework is the core of the architecture and is
-deliberately quite simple: the rules deal with the variation in log lines,
-and the actions deal with the intricacies and complications encountered
-when parsing.
+all parsers, though some specialised support functions may be provided.
+The framework is the core of the architecture and is deliberately quite
+simple: the rules deal with the variation in log lines, and the actions
+deal with the intricacies and complications encountered when parsing.
 
@@
 
-Each action contains the code required to deal with log lines from a
-particular category, e.g.\ rejections.  The actions are parser-specific:
-each parser author will need to write the actions required from scratch,
-unless extending an existing parser.  In the Postfix parser developed for
-this project there are \numberOFactions{} actions and \numberOFrules{}
-rules; the distribution of rules to actions is very skewed, as shown in
+Each action contains the code required to deal with a particular type of
+log lines, e.g.\ rejections.  The actions are parser-specific: each parser
+author will need to write the actions required from scratch, unless
+extending an existing parser.  In the Postfix parser developed for this
+project there are \numberOFactions{} actions and \numberOFrules{} rules,
+with a very skewed distribution of rules to actions as shown in
 \graphref{Distribution of rules per action}.  The distribution is skewed
@@
 resulting in simpler rules and clearer distinctions between log lines.
-\showgraph{build/graph-action-distribution}{Distribution
-of rules per action}{Distribution of rules per action} Unsurprisingly the
-action with the highest number of associated rules is \texttt{REJECTION},
-the action which handles Postfix rejecting a mail; it is followed by
+\showgraph{build/graph-action-distribution}{Distribution of rules per
+action}{Distribution of rules per action} Unsurprisingly the action with
+the highest number of associated rules is \texttt{REJECTION}, the action
+which handles Postfix rejecting a mail; it is followed by
 \texttt{SAVE\_BY\_QUEUEID}, the action responsible for handling informative
@@
 the action is required to address a deficiency in the log files, or a
-complication which arises during parsing.
-
-\label{example action}
-
-Using the \texttt{CONNECT} action as an example, it creates a new data
-structure for each client connecting, and saves the data extracted by the
-rule.  If the data structure already exists it is treated as a symptom of a
-bug, and the action issues a warning containing the full contents of the
-data structure, plus the log line which has just been parsed.
+complication which arises during parsing.  Using the \texttt{CONNECT}
+action as an example, it creates a new data structure for each client
+connecting, and saves the data extracted by the rule.  If the data
+structure already exists it is treated as a symptom of a bug, and the
+action issues a warning containing the full contents of the data structure,
+plus the log line which has just been parsed.
 
@@
 
-During development of the parser it became apparent that in addition to the
-obvious variety in log lines, there were a large number of complications to
-be overcome.  Some were the result of deficiencies in Postfix's logging
-(some of which were improved on or resolved by later versions of Postfix);
-others were due to the vagaries of process scheduling, client behaviour,
-and administrative actions.  All were successfully accommodated in the
-parser: adding new actions was enough for several of the complications;
-others required modifications to a single existing action to work around
-the difficulties; the remainder were resolved by adapting actions to pass
-information to subsequent actions (via the framework), which check for the
-extra information and change their behaviour appropriately.
+During development of the Postfix parser it became apparent that in
+addition to the obvious variety in log lines, there were a large number of
+complications to be overcome.  Some were the result of deficiencies in
+Postfix's logging (some of which were improved on or resolved by later
+versions of Postfix); others were due to the vagaries of process
+scheduling, client behaviour, and administrative actions.  All were
+successfully accommodated in the Postfix parser: adding new actions was
+enough to overcome several of the complications; others required
+modifications to a single existing action to work around the difficulties;
+the remainder were resolved by adapting actions to pass information to
+subsequent actions (via the framework), which check for the extra
+information and change their behaviour appropriately.
 
@@
 parsing~\cite{cascaded-parsing}.  This is a powerful facility allowing a
-generic rule and action to parse part of the input, followed by more
-specific rules and actions parsing the remainder.
+number of rules and actions to parse a single input line.
 
@@
 a set of results; other implementations are free to store the rules in
-whatever fashion suits their needs.
-
-The framework requires that each rule have \texttt{action} and
-\texttt{regex} attributes; each implementation is free to add any
-additional attributes it requires.  The Postfix parser adds several
-attributes to optimise rule ordering (\sectionref{Rule ordering}), to
-restrict which log lines rules can be matched against, and to describe each
-rule.  Data is automatically extracted and saved from the log line based on
-keywords in the rule's regex, but there is also a facility to specify extra
-data to be saved.
+whatever fashion suits their needs.  The framework requires each rule to
+have \texttt{action} and \texttt{regex} attributes; each implementation is
+free to add any additional attributes it requires.  The Postfix parser adds
+several attributes to optimise rule ordering (\sectionref{Rule ordering}),
+to restrict which log lines rules can be matched against, and to describe
+each rule.  Data is automatically extracted and saved from the log line
+based on keywords in the rule's regex, but there is also a facility to
+specify extra data to be saved.
 
@@
 
-The parser does not try to detect overlapping rules; that responsibility is
+The parser does not try to detect overlapping rules: that responsibility is
 left to the author of the rules.  Unintentionally overlapping rules lead to
@@
 overlap between $n$ regexes requires $\frac{n(n-1)}{2}$ intersections to be
-calculated, giving $O(n^2)$ complexity.  This is certainly not a task to be
-performed on every parsing run: detecting overlap amongst the Postfix
-parser's \numberOFrules{} rules would require \numberOFruleINTERSECTIONS{}
-intersections to be calculated.
+calculated, giving $O(n^2x)$ complexity, where $O(x)$ is the complexity of
+calculating intersection.  This is certainly not a task to be performed on
+every parsing run: detecting overlap amongst the Postfix parser's
+\numberOFrules{} rules would require calculating
+\numberOFruleINTERSECTIONS{} intersections.
 
@@
 large amounts of CPU time during matching.  Defining a regex which matches
-every input line is trivial: \verb!/^/! will match the start of every input
+every input line is trivial: \verb!/^/! matches the start of every input
 line.  In general a regex will not consume a large amount of CPU time when
@@
 
-The example rule in \tableref{Example rule} matches the message logged by
-Postfix when a remote client connects to deliver mail; it would match the
-following sample log line:
-
-\begin{verbatim}
-    connect from client.example.com[1.2.3.4]
-\end{verbatim}
+The example rule in \tableref{Example rule} matches the following sample
+log line logged by Postfix when a remote client connects to deliver mail:
+\newline{}\tab{}\verb!connect from client.example.com[1.2.3.4]!
 
@@
         \tableline{}%
-        program                   & \daemon{smtpd}                                   \\
+        %program                   & \daemon{smtpd}                                   \\
         regex                     & \verb!^connect from (__HOSTNAME__)\[(__IP__)\]$! \\
-        result\_cols              &                                                  \\
-        connection\_cols~\empty{} & client\_hostname = 1; client\_ip = 2             \\
         action                    & CONNECT                                          \\
-        priority                  & 0                                                \\
+        %priority                  & 0                                                \\
         \tableline{}%
@@
 \end{table}
-
-\noindent{}The \texttt{CONNECT} action is described in \sectionref{example
-action}.
 
@@
     \item [Context-free rules:]  Rules are context free: they do not take
-        into account inputs which have previously been seen, nor do they
-        perform any kind of lookahead to future inputs.  In context-free
+        into account past or future inputs.  In context-free
         grammar terms the parser rules could be described as:
@@
         fixed grammar specification which inputs must adhere to.  The
-        architecture is capable of dealing with overlapping inputs, out of
+        architecture is capable of dealing with interleaved inputs, out of
         order inputs, and ambiguous inputs where heuristics must be applied
@@
         there is no facility for rules to consume more input or push unused
-        input back onto the input stream.  This was not a deliberate
-        decision, rather a consequence of the line oriented nature of the
-        log files; more flexible approaches could be pursued.
+        input back onto the input stream (however actions may change input
+        lines and cause the modified line to be re-parsed).  This was not a
+        deliberate decision, rather a consequence of the line oriented
+        nature of the log files; more flexible approaches could be pursued.
 
@@
 Parsing efficiency is an obvious concern when the parser routinely needs to
-parse very large log files.  The server which generated the log files used
-in testing this parser accepts approximately 10,000 mails for 700 users a
-day.  Median log file size is 50MB, containing 285,000 log lines --- large
-scale mail servers would have much larger log files.  When generating the
-timing data used in this section, \numberOFlogFILES{} log files (totaling
-10.08 GB, \numberOFlogLINEShuman{} log lines) were each parsed 10 times and
-the parsing times averaged.  Saving results to the database was disabled
-for the test runs, as that dominates the run time of the program, and the
-tests are aimed at measuring the speed of the parser rather than the speed
-of the database and the disks the database is stored on.  The computer used
-for test runs was a Dell Optiplex 745 described in \tableref{Computer used
-to generate statistics}.
+parse very large log files.  The mail server which generated the log files
+used in testing this parser accepts approximately 10,000 mails for 700
+users a day.  Median log file size is 50MB, containing 285,000 log lines
+--- large scale mail servers would have much larger log files.  When
+generating the timing data used in this section, \numberOFlogFILES{} log
+files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
+parsed 10 times and the parsing times averaged.  Saving results to the
+database was disabled for the test runs, as that dominates the run time of
+the program, and the tests are aimed at measuring the speed of the parser
+rather than the speed of the database and the disks the database is stored
+on.  The computer used for test runs was a Dell Optiplex 745 described in
+\tableref{Computer used to generate statistics}.  Parsing all
+\numberOFlogFILES{} log files in one run took
+\input{build/include-timing-run-duration}.  The computer was dedicated to
+the task of gathering statistics from test runs, and was not used for any
+other purpose while test runs were ongoing; any services not necessary for
+running the tests were disabled.
 
@@
 
-Parsing all \numberOFlogFILES{} log files in one run took
-\input{build/include-timing-run-duration}.  The computer was dedicated to
-the task of gathering statistics from test runs, and was not used for any
-other purpose while test runs were ongoing; any services not necessary for
-running the tests were disabled.
-
 \subsection{Architecture Scalability: Input Size}
@@
 \Graphref{Parsing time, file size, and number of lines} shows the parsing
-time in seconds, file size in MB, and tens of thousands of log lines for
-each of the \numberOFlogFILES{} log files.  All three lines run roughly in
-parallel, giving the impression that the algorithm scales linearly with
-input size.  This impression is borne out by \graphref{parsing time vs file
-size vs number lines factor} which plots both the ratio of file size vs
-parsing time, and the ratio of number of log lines vs parsing time (higher
-is better in both cases).  The ratios are quite tightly banded across the
-graph, showing that the algorithm scales linearly.  The ratio increases
-(i.e.\ improves) for log files 22 and 62--68 despite their larger than
-average size (shown in \graphref{Parsing time, file size, and number of
-lines}).  \showgraph{build/graph-input-size-vs-parsing-time}{Parsing time,
-file size, and number of lines}{Parsing time, file size, and number of
-lines} \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of
-file size and number of lines to parsing time}{parsing time vs file size vs
-number lines factor}
+time in seconds, file size in MB, and number of log lines in tens of
+thousands, for each of the \numberOFlogFILES{} log files.  All three lines
+run roughly in parallel, giving the impression that the algorithm scales
+linearly with input size.  This impression is borne out by
+\graphref{parsing time vs file size vs number lines factor} which plots
+both the ratio of file size vs parsing time, and the ratio of number of log
+lines vs parsing time (higher is better in both cases).  The ratios are
+quite tightly banded across the graph, showing that the algorithm scales
+linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
+despite their larger than average size (shown in \graphref{Parsing time,
+file size, and number of lines}).
+
+\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
+and number of lines}{Parsing time, file size, and number of lines}
+\showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of number of
+lines and file size to parsing time}{parsing time vs file size vs number
+lines factor}
 
@@
 \Graphref{Parsing time relative to shuffled} shows the parsing times of
-optimal and reversed orderings relative to shuffled ordering.  Overall this
-optimisation provides a modest but worthwhile reduction in parsing time of
-approximately 10\%.  \showgraph{build/graph-hits}{Hits per rule}{rule hits
-graph} \showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
+optimal and reversed orderings as a percentage of the parsing time of
+shuffled ordering.  Overall this optimisation provides a modest but
+worthwhile reduction in parsing time of approximately 10\%.
+\showgraph{build/graph-hits}{Hits per rule}{rule hits graph}
+\showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
 relative to shuffled ordering}{Parsing time relative to shuffled}
@@
 expected that there will be a large number of rules.  There are
-\numberOFrules{} rules in the full Postfix ruleset, whereas the minimum
-number of rules required to parse the \numberOFlogFILES{} log files is
-\numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the total.  A
-second set of statistics was generated using the minimum ruleset and the
-results compared to the statistics generated using the full ruleset: the
-percentage parsing time increase when using the full ruleset instead of the
-minimal ruleset for optimal, shuffled and reversed orderings is shown in
-\graphref{Percentage parsing time increase of maximum ruleset over minimum
-ruleset}.
+\numberOFrules{} rules in the full Postfix parser ruleset, whereas the
+minimum number of rules required to parse the \numberOFlogFILES{} log files
+is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
+ruleset.  A second set of statistics was generated using the minimum
+ruleset and the parsing times compared to those generated using the full
+ruleset: the percentage parsing time increase when using the full ruleset
+instead of the minimal ruleset for optimal, shuffled and reversed orderings
+is shown in \graphref{Percentage parsing time increase of maximum ruleset
+over minimum ruleset}.
 \showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
@@
 ruleset over minimum ruleset} that the increased number of rules has a
-noticeable performance impact with reverse ordering, and to a lesser extent
-with shuffled ordering.  The optimal ordering (where the most frequently
-matching rules are tried first) shows a mean increase of
+noticeable performance impact with reverse ordering, and a lesser impact
+with shuffled ordering.  The optimal ordering shows a mean increase of
 \input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
@@
 
-The framework issues warnings for any log lines which are not parsed; no
+Warnings are issued for any log lines which are not parsed; no
 warnings are issued for unparsed log lines while parsing the
@@
 there are zero false negatives.  False positives are harder to quantify:
-manually verifying that the correct regex parsed each of the 60,721,709 log
-lines is infeasible, making it impossible to quantify the false positive
-rate; however a random sample of 6039 log lines (0.00994\% of the total)
-was parsed and the results verified to ensure that the correct regex parsed
-each log line.
+manually verifying that the correct regex parsed each of the
+\numberOFlogLINES{} log lines is infeasible, making it impossible to
+quantify the false positive rate; however a random sample of 6039 log lines
+(0.00994\% of the total) was parsed and the results manually verified by
+inspection to ensure that the correct regex parsed each log line.
 
@@
 started before the log file segment), 10 mails remaining in the state
-tables, and 1625 connections correctly entered in the database.  Given the
+tables, and 1625 connections correctly entered in the database.  No error
+messages were produced, therefore there were no false negatives.  Given the
 evidence detailed above, the authors are confident that the false positive
@@
 to process versions one and two of a file format; the parser will then
-parse either version one or version two depending on the ruleset selected.
-The architecture makes it possible to apply commonly used programming
-techniques such as object orientation, inheritance, composition, delegation
-and roles when designing and implementing a parser, simplifying the process
-of working within a team with shared responsibility or developing and
-testing additional functionality.  This architecture is ideally suited to
-parsing inputs where the input is not fully understood or does not follow a
-fixed grammar: the architecture warns about unparsed inputs and other
-errors, but continues on parsing as best it can, allowing the developer of
-a new parser to decide which deficiencies are most important and require
-attention first, rather than being forced to fix the first error which
-arises.
+parse either version one, version two, or both versions depending on the
+ruleset selected.  The architecture makes it possible to apply commonly
+used programming techniques such as object orientation, inheritance,
+composition, delegation and roles when designing and implementing a parser,
+simplifying the process of working within a team with shared
+responsibility, or when developing and testing additional functionality.
+This architecture is ideally suited to parsing inputs where the input is
+not fully understood or does not follow a fixed grammar: the architecture
+warns about unparsed inputs and other errors, but continues on parsing as
+best it can, allowing the developer of a new parser to decide which
+deficiencies are most important and require attention first, rather than
+being forced to fix the first error which arises.
 
@@
 a normalised form which is far easier to use as input to new or existing
-implementations of the machine-learning algorithms than trying to adapt the
-algorithms to extract data directly from the log files.  The current focus
+implementations of machine-learning algorithms than trying to adapt each
+algorithm to extract data directly from the log files.  The current focus
 is on clustering and decision trees to optimise the order in which rules

### 2008-08-15 doc/sgai-2008.tex
Finished my corrections.

@@
 were incapable of distinguishing between log lines of the same type, e.g.\
-rejections, and extracted only a small portion of the data required.  The only
-prior published work on the subject of parsing Postfix log files
+rejections, and extracted only a small portion of the data required.  The
+only prior published work on the subject of parsing Postfix log files
 is~\cite{log-mail-analyser}, which aims to present correlated log data in a
@@
 Computational Linguistics for creating grammars to parse or generate
-sentences.  The resemblance between ATN and the parser is accidental, but
-it is interesting how two different approaches have a similar division of
-responsibilities, while having different semantics.
+sentences.  The resemblance between ATN and the architecture is accidental,
+but it is interesting how two different approaches have a similar division
+of responsibilities, while having different semantics.
 
@@
         \noalign{\smallskip}
-        ATN           & Parser    & Similarity                          \\
+        ATN           & Architecture    & Similarity                          \\
         \noalign{\smallskip}
@@
 author will need to write the actions required from scratch, unless
-extending an existing parser.  In the Postfix parser developed for this
+extending an existing parser.  In the Postfix log parser developed for this
 project there are \numberOFactions{} actions and \numberOFrules{} rules,
@@
 (neither \texttt{REJECTION} nor \texttt{IGNORE} imply that the input is
-ungrammatical or unparsed).  The parser warns about every unparsed log
+ungrammatical or unparsed).  The framework warns about every unparsed log
 line, to alert the user that they need to alter or extend their ruleset;
-the Postfix parser successfully parses every input line in the
+the Postfix log parser successfully parses every input line in the
 \numberOFlogFILESall{} log files it is currently being tested with.  The
@@
 
-During development of the Postfix parser it became apparent that in
+During development of the Postfix log parser it became apparent that in
 addition to the obvious variety in log lines, there were a large number of
@@
 scheduling, client behaviour, and administrative actions.  All were
-successfully accommodated in the Postfix parser: adding new actions was
+successfully accommodated in the Postfix log parser: adding new actions was
 enough to overcome several of the complications; others required
@@
 log line and the action to invoke when the match is successful.  The
-Postfix parser stores its rules in the same SQL database the results are
-stored in, removing any doubt as to which set of rules was used to produce
-a set of results; other implementations are free to store the rules in
-whatever fashion suits their needs.  The framework requires each rule to
+Postfix log parser stores its rules in the same SQL database the results
+are stored in, removing any doubt as to which set of rules was used to
+produce a set of results; other implementations are free to store the rules
+in whatever fashion suits their needs.  The framework requires each rule to
 have \texttt{action} and \texttt{regex} attributes; each implementation is
-free to add any additional attributes it requires.  The Postfix parser adds
-several attributes to optimise rule ordering (\sectionref{Rule ordering}),
-to restrict which log lines rules can be matched against, and to describe
-each rule.  Data is automatically extracted and saved from the log line
-based on keywords in the rule's regex, but there is also a facility to
-specify extra data to be saved.
+free to add any additional attributes it requires.  The Postfix log parser
+adds several attributes to optimise rule ordering (\sectionref{Rule
+ordering}), to restrict which log lines rules can be matched against, and
+to describe each rule.  Data is automatically extracted and saved from the
+log line based on keywords in the rule's regex, but there is also a
+facility to specify extra data to be saved.
 
 Adding new rules is simple: define all the required attributes and append
-the new rule to wherever the existing rules are stored.  The Postfix parser
-supplies a utility based on SLCT~\cite{slct-paper} to aid in producing
-regexes from previously unparsed inputs.  Storing the rules separately
-allows other approaches to rule management to be used, e.g.\ instead of
-manually adding new rules when unparsed log lines are seen, machine
-learning techniques could be used to automatically generate new rules,
-based on the existing rules and input lines.  If this approach was taken
-the choice of machine learning technique would be constrained by the size
-of typical data sets: the \numberOFlogFILES{} log files used to generate
-statistics contain \numberOFlogLINEShuman{} log lines.  Techniques which
-require the full data set when training would be impractical; Instance
-Based Learning~\cite{instance-based-learning} techniques which
+the new rule to wherever the existing rules are stored.  The Postfix log
+parser supplies a utility based on SLCT~\cite{slct-paper} to aid in
+producing regexes from previously unparsed inputs.  Storing the rules
+separately allows other approaches to rule management to be used, e.g.\
+instead of manually adding new rules when unparsed log lines are seen,
+machine learning techniques could be used to automatically generate new
+rules, based on the existing rules and input lines.  If this approach was
+taken the choice of machine learning technique would be constrained by the
+size of typical data sets: the \numberOFlogFILES{} log files used to
+generate statistics contain \numberOFlogLINEShuman{} log lines.  Techniques
+which require the full data set when training would be impractical;
+Instance Based Learning~\cite{instance-based-learning} techniques which
 automatically determine which inputs from the training set are valuable and
@@
 
-The parser does not try to detect overlapping rules: that responsibility is
-left to the author of the rules.  Unintentionally overlapping rules lead to
-inconsistent parsing and data extraction because the first matching rule
-wins, and the order in which rules are tried against each line may change
-between log files.  Overlapping rules are frequently a requirement,
-allowing a more specific rule to match some log lines and a more general
-rule to match the majority, e.g.\ separating SMTP delivery to specific
-sites from SMTP delivery to the rest of the world.  Overlapping rules
-should have a priority attribute to specify the order in which they should
-be used; negative priorities may be useful for catchall rules.
+The architecture does not try to detect overlapping rules: that
+responsibility is left to the author of the rules.  Unintentionally
+overlapping rules lead to inconsistent parsing and data extraction because
+the first matching rule wins, and the order in which rules are tried
+against each line may change between log files.  Overlapping rules are
+frequently a requirement, allowing a more specific rule to match some log
+lines and a more general rule to match the majority, e.g.\ separating SMTP
+delivery to specific sites from SMTP delivery to the rest of the world.
+This can simplify both the general rule and the more specific rule;
+additionally rules from different sources can be combined with a minimum of
+prior cooperation or modification required.  Overlapping rules should have
+a priority attribute to specify the order in which they should be used;
+negative priorities may be useful for catchall rules.
 
@@
 calculating intersection.  This is certainly not a task to be performed on
-every parsing run: detecting overlap amongst the Postfix parser's
+every parsing run: detecting overlap amongst the Postfix log parser's
 \numberOFrules{} rules would require calculating
@@
 
-Parsing efficiency is an obvious concern when the parser routinely needs to
-parse very large log files.  The mail server which generated the log files
-used in testing this parser accepts approximately 10,000 mails for 700
-users a day.  Median log file size is 50MB, containing 285,000 log lines
---- large scale mail servers would have much larger log files.  When
-generating the timing data used in this section, \numberOFlogFILES{} log
-files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
-parsed 10 times and the parsing times averaged.  Saving results to the
-database was disabled for the test runs, as that dominates the run time of
-the program, and the tests are aimed at measuring the speed of the parser
-rather than the speed of the database and the disks the database is stored
-on.  The computer used for test runs was a Dell Optiplex 745 described in
+Parsing efficiency is an obvious concern when the Postfix log parser
+routinely needs to parse very large log files.  The mail server which
+generated the log files used in testing the Postfix log parser accepts
+approximately 10,000 mails for 700 users a day.  Median log file size is
+50MB, containing 285,000 log lines --- large scale mail servers would have
+much larger log files.  When generating the timing data used in this
+section, \numberOFlogFILES{} log files (totaling 10.08 GB,
+\numberOFlogLINEShuman{} log lines) were each parsed 10 times and the
+parsing times averaged.  Saving results to the database was disabled for
+the test runs, as that dominates the run time of the program, and the tests
+are aimed at measuring the speed of the Postfix log parser rather than the
+speed of the database and the disks the database is stored on.  The
+computer used for test runs was a Dell Optiplex 745 described in
 \tableref{Computer used to generate statistics}.  Parsing all
@@
 linearly with input size.  This impression is borne out by
-\graphref{parsing time vs file size vs number lines factor} which plots
-both the ratio of file size vs parsing time, and the ratio of number of log
-lines vs parsing time (higher is better in both cases).  The ratios are
+\graphref{parsing time vs file size vs number lines factor}: the ratios are
 quite tightly banded across the graph, showing that the algorithm scales
 linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
-despite their larger than average size (shown in \graphref{Parsing time,
-file size, and number of lines}).
+despite their larger than average size (due to a mail forwarding loop
+resulting in a greatly increased number of mails delivered and log lines
+generated).
 
@@
 lines factor}
+
+% XXX MAYBE ADD A TABLE SHOWING MEAN AND STDDEV OF THE RATIOS?
 
@@
 expected that there will be a large number of rules.  There are
-\numberOFrules{} rules in the full Postfix parser ruleset, whereas the
+\numberOFrules{} rules in the full Postfix log parser ruleset, whereas the
 minimum number of rules required to parse the \numberOFlogFILES{} log files
@@
 
-The parser's coverage of Postfix log files is separated into two parts: log
-line coverage and mail coverage.  The former is initially more important,
-as the parser must successfully parse every line if it is to be complete,
-but subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Improving the former
-is less intrusive, as it just requires new rules to be written; improving
-the latter is more intrusive as it requires changes to actions, and it can
-also be much harder to notice a deficiency.
+The Postfix log parser's coverage of log files is separated into two parts:
+log line coverage and mail coverage (the former is a requirement for the
+latter to be achieved).  Improving the former is less intrusive, as it just
+requires new rules to be written; improving the latter is more intrusive as
+it requires changes to actions, and it can also be much harder to notice a
+deficiency.
 
@@
 Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly.  The parser
-issues warnings when it detects any errors, some of which may alert the
-user to a problem.  There should be few or no warnings when parsing, and
-when finished parsing the state table should only contain entries for mails
-which had yet to be delivered when the log files ended, or were accepted
-before the log files began.  A sample of 6000 log lines was parsed with all
-debugging options enabled, resulting in 167,448 lines of output.  All
-167,448 lines were examined in conjunction with the input log file and a
-dump of the resulting database, verifying that for each of the input lines
-the parser used the correct rule and executed the correct action, which in
-turn produced the correct result and inserted the correct data in the
-database.  The log segment produced 4 warnings (about mails which had
-started before the log file segment), 10 mails remaining in the state
-tables, and 1625 connections correctly entered in the database.  No error
-messages were produced, therefore there were no false negatives.  Given the
-evidence detailed above, the authors are confident that the false positive
-rate when reconstructing a mail is zero.
+coverage of log lines.  The implementation can dump its state tables in a
+human readable form; examining these tables with reference to the log files
+is the best way to detect mails which were not handled properly.  The
+implementation issues warnings when it detects any errors, some of which
+may alert the user to a problem.  There should be few or no warnings when
+parsing, and when finished parsing the state table should only contain
+entries for mails which had yet to be delivered when the log files ended,
+or were accepted before the log files began.  A sample of 6000 log lines
+was parsed with all debugging options enabled, resulting in 167,448 lines
+of output.  All 167,448 lines were examined in conjunction with the input
+log file and a dump of the resulting database, verifying that for each of
+the input lines the implementation used the correct rule and executed the
+correct action, which in turn produced the correct result and inserted the
+correct data in the database.  The log segment produced 4 warnings (about
+mails which had started before the log file segment), 10 mails remaining in
+the state tables, and 1625 connections correctly entered in the database.
+No error messages were produced, therefore there were no false negatives.
+Given the evidence detailed above, the authors are confident that the false
+positive rate when reconstructing a mail is zero.
 
@@
 
-The dataset gathered by the parser provides the foundation for the future
-of this project: applying machine-learning algorithms to the data to
-analyse and optimise the set of rules in use, followed by identifying
-patterns in the data which could be used to write new filters to recognise
-and reject spam rather than accepting it.  The parser provides the data in
-a normalised form which is far easier to use as input to new or existing
-implementations of machine-learning algorithms than trying to adapt each
-algorithm to extract data directly from the log files.  The current focus
-is on clustering and decision trees to optimise the order in which rules
-are applied; future efforts will involve using data gathered by the parser
-to train and test new filters.  This task is very similar to optimising a
-black box application based on its inputs and outputs, and this approach
-could be applied to optimising the behaviour of any system given sufficient
-logging to analyse.  An alternative approach to black box optimisation
-which uses application profiling in conjunction with the application's
-error messages to improve the error messages shown to users is described
-in~\cite{black-box-error-reporting}; profiling data may be useful in
-supplementing systems which fail to provide adequate log messages.
+The dataset gathered by the Postfix log parser provides the foundation for
+the future of this project: applying machine-learning algorithms to the
+data to analyse and optimise the set of rules in use, followed by
+identifying patterns in the data which could be used to write new filters
+to recognise and reject spam rather than accepting it.  The parser provides
+the data in a normalised form which is far easier to use as input to new or
+existing implementations of machine-learning algorithms than trying to
+adapt each algorithm to extract data directly from the log files.  The
+current focus is on clustering and decision trees to optimise the order in
+which rules are applied; future efforts will involve using data gathered by
+the parser to train and test new filters.  This task is very similar to
+optimising a black box application based on its inputs and outputs, and
+this approach could be applied to optimising the behaviour of any system
+given sufficient logging to analyse.  An alternative approach to black box
+optimisation which uses application profiling in conjunction with the
+application's error messages to improve the error messages shown to users
+is described in~\cite{black-box-error-reporting}; profiling data may be
+useful in supplementing systems which fail to provide adequate log
+messages.
 

### 2008-08-15 doc/sgai-2008.tex
Some suggestions from Carl; more to follow.

@@
     understanding of the parser's internal workings.  This paper presents a
-    novel parsing architecture which aims to make the process of parsing
-    new inputs as simple as possible, enabling users to trivially add new
-    rules (to parse variants of existing inputs) and relatively easily add
-    new actions (to process a previously unknown class of input).  The
-    architecture scales linearly as the number of rules and/or size of
-    input increases, making it suitable for parsing large corpora or months
-    of accumulated data.
+    novel parsing architecture designed for Postfix log files which aims to
+    make the process of parsing new inputs as simple as possible, enabling
+    users to trivially add new rules (to parse variants of existing inputs)
+    and relatively easily add new actions (to process a previously unknown
+    class of input).  The architecture scales linearly as the number of
+    rules and/or size of input increases, making it suitable for parsing
+    large corpora or months of accumulated data.
 }
@@
 
-XXX DISCUSS COMPLEXITY\@?
+%XXX DISCUSS COMPLEXITY\@?
 
@@
 rejections, and extracted only a small portion of the data required.  The
-only prior published work on the subject of parsing Postfix log files
-is~\cite{log-mail-analyser}, which aims to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available.  A full state of
-the art review is outside the scope of this paper but will be included in
-the thesis resulting from this work.
+only prior published work on the subject of parsing Postfix log files that
+we are aware of is~\cite{log-mail-analyser}, which aims to present
+correlated log data in a form suitable for a systems administrator to
+search using the myriad of standard Unix text processing utilities already
+available.  A full state of the art review is outside the scope of this
+paper but will be included in the thesis resulting from this work.
 
@@
 from the declarative grammar it operates with.  Part of the novelty here is
-that the grammars (rules and actions) are themselves procedures.  This
-architecture is ideally suited to parsing inputs where the input is not
-fully understood or does not conform to a fixed grammar: the architecture
-warns about unparsed inputs and other errors, but continues on parsing as
-best it can, allowing the developer of a new parser to decide which
-deficiencies are most important and require attention first, rather than
-being forced to fix the first error which arises.
+in the way that the grammars (rules and actions) are themselves procedures.
+The relationship to another procedural framework (ATN) is discussed below.
+This architecture is ideally suited to parsing inputs where the input is
+not fully understood or does not conform to a fixed grammar: the
+architecture warns about unparsed inputs and other errors, but continues on
+parsing as best it can, allowing the developer of a new parser to decide
+which deficiencies are most important and require attention first, rather
+than being forced to fix the first error which arises.
 
@@
 
+XXX RENAME ``REJECTION'' TO ``MAIL\_REJECTED'' AND ``IGNORE'' TO
+``UNINTERESTING\_LOG\_LINE'' OR SOMETHING ELSE THAT MAKES IT BRUTALLY
+OBVIOUS THE PARSER IS SUCCESSFULLY DEALING WITH THE LOG LINES\@.  I SHOULD
+PROBABLY RENAME ``SAVE\_BY\_QUEUEID'' TOO\@.
+
 Each action contains the code required to deal with a particular type of
@@
 the action is required to address a deficiency in the log files, or a
-complication which arises during parsing.  Using the \texttt{CONNECT}
+complication which arises during parsing.  Taking the \texttt{CONNECT}
 action as an example, it creates a new data structure for each client
-connecting, and saves the data extracted by the rule.  If the data
-structure already exists it is treated as a symptom of a bug, and the
-action issues a warning containing the full contents of the data structure,
-plus the log line which has just been parsed.
+connecting, and saves the data extracted by the rule (in memory,
+subsequently to be entered into the database).  If the data structure
+already exists it is treated as a symptom of a bug, and the action issues a
+warning containing the full contents of the data structure, plus the log
+line which has just been parsed.
 
@@
 
-Warnings are issued for any log lines which are not parsed; no
-warnings are issued for unparsed log lines while parsing the
-\numberOFlogFILES{} test log files, so it can be safely concluded that
-there are zero false negatives.  False positives are harder to quantify:
-manually verifying that the correct regex parsed each of the
-\numberOFlogLINES{} log lines is infeasible, making it impossible to
-quantify the false positive rate; however a random sample of 6039 log lines
-(0.00994\% of the total) was parsed and the results manually verified by
-inspection to ensure that the correct regex parsed each log line.
+Warnings are issued for any log lines which are not parsed; no warnings are
+issued for unparsed log lines while parsing the \numberOFlogFILES{} test
+log files, so it can be safely concluded that there are zero false
+negatives.  False positives are harder to quantify: manually verifying that
+the correct regex parsed each of the \numberOFlogLINES{} log lines is
+infeasible, making it impossible to quantify the false positive rate.  A
+random sample of 6039 log lines (0.00994\% of the total) was parsed and the
+results manually verified by inspection to ensure that the correct regex
+parsed each log line.  The sample contained zero false positives, and this
+check has been automated to ensure continued accuracy.
 
@@
 entries for mails which had yet to be delivered when the log files ended,
-or were accepted before the log files began.  A sample of 6000 log lines
-was parsed with all debugging options enabled, resulting in 167,448 lines
-of output.  All 167,448 lines were examined in conjunction with the input
-log file and a dump of the resulting database, verifying that for each of
-the input lines the implementation used the correct rule and executed the
-correct action, which in turn produced the correct result and inserted the
-correct data in the database.  The log segment produced 4 warnings (about
-mails which had started before the log file segment), 10 mails remaining in
-the state tables, and 1625 connections correctly entered in the database.
-No error messages were produced, therefore there were no false negatives.
-Given the evidence detailed above, the authors are confident that the false
-positive rate when reconstructing a mail is zero.
+or were accepted before the log files began.  A second sample of 6000 log
+lines was parsed with all debugging options enabled, resulting in 167,448
+lines of output.  All 167,448 lines were examined in conjunction with the
+input log file and a dump of the resulting database, verifying that for
+each of the input lines the implementation used the correct rule and
+executed the correct action, which in turn produced the correct result and
+inserted the correct data in the database.  The log segment produced 4
+warnings (about mails which had started before the log file segment), 10
+mails remaining in the state tables, and 1625 connections correctly entered
+in the database.  No error messages were produced, therefore there were no
+false negatives.  Given the evidence detailed above, the authors are
+confident that the false positive rate when reconstructing a mail is zero.
 

### 2008-08-15 doc/logparser-implementation.tex
Next step in the great renaming - renaming of graphs. In the process I removed some generated gnuplot input files from the repository.
Renamed actions because Carl can't understand that the IGNORE action doesn't mean that the log line wasn't parsed, so other people might have the same problem:     IGNORE              => UNINTERESTING     REJECTION           => DELIVERY_REJECTED     SAVE_BY_QUEUEID     => SAVE_DATA

@@
 
-Every interesting\footnote{All log lines are interesting unless the action
-of the rule which parses them is \texttt{INFO} or \texttt{IGNORE}, as
-described in \sectionref{actions-in-detail}.} log line will have an entry
-in the results table, e.g.\ rejecting an \SMTP{} command, delivering a
-mail, or bouncing a mail (see \sectionref{postfix_action} for typical
-values of \texttt{postfix\_action}).  Each row is associated with a single
+Every interesting\footnote{All log lines are interesting unless the
+postfix\_action of the rule which parses them is \texttt{INFO} or
+\texttt{IGNORE}, as described in \sectionref{actions-in-detail}.} log line
+will have an entry in the results table, e.g.\ rejecting an \SMTP{}
+command, delivering a mail, or bouncing a mail (see
+\sectionref{postfix_action} for typical values of
+\texttt{postfix\_action}).  Each row is associated with a single
 connection, though there may be many results per connection.
@@
 
+    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an \SMTP{}
+        command from the remote client: log the rejection with a mail if
+        there is a queueid in the log line, or with the connection if not.
+
     \item [DISCONNECT] Deal with the remote client disconnecting: enter the
@@
 
-    \item [IGNORE] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but does not either
-        provide any useful data or require anything to be done.
-
     \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
@@
 
-    \item [REJECTION] Deal with Postfix rejecting an \SMTP{} command from
-        the remote client: log the rejection with a mail if there is a
-        queueid in the log line, or with the connection if not.
-
-    \item [SAVE\_BY\_QUEUEID] Find the correct mail based on the queueid in
+    \item [SAVE\_DATA] Find the correct mail based on the queueid in
         the log line, and save the data extracted by the \regex{} to it.
@@
 
+    \item [UNINTERESTING] This rule just returns successfully; it is used when a
+        line needs to be parsed for completeness but does not either
+        provide any useful data or require anything to be done.
+
 \end{description}
@@
 
-\showgraph{build/plot-action-distribution}{Distribution of rules per
+\showgraph{build/graph-action-distribution}{Distribution of rules per
 action}{Distribution of rules per action}

### 2008-08-15 doc/sgai-2008.tex
Renamed actions because Carl can't understand that the IGNORE action doesn't mean that the log line wasn't parsed, so other people might have the same problem:     IGNORE              => UNINTERESTING     REJECTION           => DELIVERY_REJECTED     SAVE_BY_QUEUEID     => SAVE_DATA

@@
 action}{Distribution of rules per action} Unsurprisingly the action with
-the highest number of associated rules is \texttt{REJECTION}, the action
+the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the action
 which handles Postfix rejecting a mail; it is followed by
-\texttt{SAVE\_BY\_QUEUEID}, the action responsible for handling informative
+\texttt{SAVE\_DATA}, the action responsible for handling informative
 inputs: these are log lines that supplement the data gathered from other
@@
 interesting events.  The third most common action is, perhaps surprisingly,
-\texttt{IGNORE}: the \texttt{IGNORE} action does nothing when executed,
+\texttt{UNINTERESTING}: this action does nothing when executed,
 allowing uninteresting log lines to be parsed without causing any effects
-(neither \texttt{REJECTION} nor \texttt{IGNORE} imply that the input is
+(it does not imply that the input is
 ungrammatical or unparsed).  The framework warns about every unparsed log

### 2008-08-15 doc/sgai-2008.tex
Removed some unused labels cluttering the paper. Minor cleanup.

@@
 % $Id$
+% XXX REMOVE draft OPTION
 \documentclass[draft]{svmult}
@@
 
-XXX RENAME ``REJECTION'' TO ``MAIL\_REJECTED'' AND ``IGNORE'' TO
-``UNINTERESTING\_LOG\_LINE'' OR SOMETHING ELSE THAT MAKES IT BRUTALLY
-OBVIOUS THE PARSER IS SUCCESSFULLY DEALING WITH THE LOG LINES\@.  I SHOULD
-PROBABLY RENAME ``SAVE\_BY\_QUEUEID'' TOO\@.
-
 Each action contains the code required to deal with a particular type of
@@
 action}{Distribution of rules per action} Unsurprisingly the action with
-the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the action
-which handles Postfix rejecting a mail; it is followed by
+the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the
+action which handles Postfix rejecting a mail; it is followed by
 \texttt{SAVE\_DATA}, the action responsible for handling informative
 inputs: these are log lines that supplement the data gathered from other
-log lines (such as rejections), but that do not by themselves represent
-interesting events.  The third most common action is, perhaps surprisingly,
-\texttt{UNINTERESTING}: this action does nothing when executed,
-allowing uninteresting log lines to be parsed without causing any effects
-(it does not imply that the input is
-ungrammatical or unparsed).  The framework warns about every unparsed log
-line, to alert the user that they need to alter or extend their ruleset;
-the Postfix log parser successfully parses every input line in the
-\numberOFlogFILESall{} log files it is currently being tested with.  The
-remaining actions have only one or two associated rules: in some cases this
-is because there will only ever be one type of log line for that action,
-e.g.\ all log lines showing that a remote client has connected are matched
-by a single rule and handled by the \texttt{CONNECT} action; in other cases
-the action is required to address a deficiency in the log files, or a
-complication which arises during parsing.  Taking the \texttt{CONNECT}
-action as an example, it creates a new data structure for each client
-connecting, and saves the data extracted by the rule (in memory,
-subsequently to be entered into the database).  If the data structure
-already exists it is treated as a symptom of a bug, and the action issues a
-warning containing the full contents of the data structure, plus the log
-line which has just been parsed.
+log lines, but that do not by themselves represent interesting events.  The
+third most common action is, perhaps surprisingly, \texttt{UNINTERESTING}:
+this action does nothing when executed, allowing uninteresting log lines to
+be parsed without causing any effects (it does not imply that the input is
+ungrammatical or unparsed).  Generally rules specifying the
+\texttt{UNINTERESTING} action parse log lines which are not associated with
+a specific mail, e.g.\ warnings about configuration files changing.  The
+framework warns about every unparsed log line, to alert the user that they
+need to alter or extend their ruleset; the Postfix log parser successfully
+parses every input line in the \numberOFlogFILESall{} log files it is
+currently being tested with.  The remaining actions have only one or two
+associated rules: in some cases this is because there will only ever be one
+type of log line for that action, e.g.\ all log lines showing that a remote
+client has connected are matched by a single rule and handled by the
+\texttt{CONNECT} action; in other cases the action is required to address a
+deficiency in the log files, or a complication which arises during parsing.
+Taking the \texttt{CONNECT} action as an example, it creates a new data
+structure for each client connecting, and saves the data extracted by the
+rule (in memory, subsequently to be entered into the database).  If the
+data structure already exists it is treated as a symptom of a bug, and the
+action issues a warning containing the full contents of the data structure,
+plus the log line which has just been parsed.
 
@@
 interested in, rather than reproducing large chunks of functionality.
-
-\label{complications}
 
@@
 
-\label{overlapping rules}
-
 The architecture does not try to detect overlapping rules: that
@@
 for in-depth discussion.
-
-\label{example rule}
 
@@
 \label{Rule ordering}
-\label{rule ordering for efficiency}
-\label{rule efficiency}
 

### 2008-08-16 doc/sgai-2008.tex
Added glossary and acronyms, to avoid manually managing them. Changed graph placement, saving a small amount of space. Improved the rule overlap section some more. Use address space reserved for examples for the example address.

@@
 
+% Acronyms and glossary entries.
+% global=true prevents acronyms being expanded twice
+\usepackage[global=true,acronym=true]{glossary}
+%\makeglossary{}
+%\makeacronym{}
+
 % Nicer citation lists - smaller spaces.
@@
 \newcommand{\showgraph}[3]{%
-    \begin{figure}[hbtp]%
+    \begin{figure}[btp]%
         \caption{#2}\label{#3}%
@@
 
+\newacronym{PDA}{Pushdown Automata}{}
+\newacronym{FA}{Finite Automata}{}
+\newacronym{CFL}{Context Free Languages}{}
+\newacronym{SLCT}{Simple Logfile Clustering Tool}{}
+\newacronym{SMTP}{Simple Mail Transfer Protocol}{}
+
 \section{Introduction}
@@
 performance of anti-spam techniques which are applied when mail is being
-transferred via SMTP\footnotemark{} is the goal of this project, by
+transferred via \SMTP{}\footnotemark{} is the goal of this project, by
 providing a platform for reasoning about anti-spam filters.  The approach
@@
 with the great number and variety of log lines produced by
-Postfix~\cite{postfix}, the SMTP server in use.  The parser must be
+Postfix~\cite{postfix}, the \SMTP{} server in use.  The parser must be
 designed so that adding support for parsing new log lines is a simple task,
@@
 the new rule to wherever the existing rules are stored.  The Postfix log
-parser supplies a utility based on SLCT~\cite{slct-paper} to aid in
+parser supplies a utility based on \SLCT{}~\cite{slct-paper} to aid in
 producing regexes from previously unparsed inputs.  Storing the rules
@@
 frequently a requirement, allowing a more specific rule to match some log
-lines and a more general rule to match the majority, e.g.\ separating SMTP
-delivery to specific sites from SMTP delivery to the rest of the world.
-This can simplify both the general rule and the more specific rule;
+lines and a more general rule to match the majority, e.g.\ separating
+\SMTP{} delivery to specific sites from \SMTP{} delivery to the rest of the
+world.  This can simplify both the general rule and the more specific rule;
 additionally rules from different sources can be combined with a minimum of
@@
 written to detect overlapping rules.  Traditional regexes are equivalent in
-computational power to Finite Automata (FA); it is possible to convert
-between regexes and FA, so regex overlap could be detected by finding a
-non-empty intersection of two FA\@.  Perl 5.10 regexes are more powerful
-than traditional regexes, e.g.\ it is possible to match correctly nested
-and balanced brackets to an arbitrary depth:
+computational power to \FA{}; it is possible to convert between regexes and
+\FA{}, so regex overlap could be detected by finding a non-empty
+intersection of two \FA{}\@.  One formula for \FA{} intersection is given
+in~\cite{intersection-of-NFA-using-Z}: $FA1 \cap{} FA2 =
+\overline{(\overline{FA1} \cup{} \overline{FA2})}$.  Perl 5.10 regexes are
+more powerful than traditional regexes, e.g.\ it is possible to match
+correctly nested and balanced brackets to an arbitrary depth:
 \newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
 Perl 5.10 regexes can maintain an arbitrary state stack and are thus
-equivalent in computational power to Pushdown Automata (PA), so overlap
-detection may require calculating the intersection of two PA.\@  Detecting
-overlap between $n$ regexes requires $\frac{n(n-1)}{2}$ intersections to be
-calculated, giving $O(n^2x)$ complexity, where $O(x)$ is the complexity of
-calculating intersection.  This is certainly not a task to be performed on
-every parsing run: detecting overlap amongst the Postfix log parser's
-\numberOFrules{} rules would require calculating
-\numberOFruleINTERSECTIONS{} intersections.
+equivalent in computational power to \PDA{}, so overlap detection may
+require calculating the intersection of two \PDA{}\@.  \CFL{} are
+equivalent in computational power to \PDA{}, and intersection of two
+\CFL{s} is not closed (i.e.\ intersection can produce a language which
+cannot be parsed by a \CFL{}), e.g.\ $a^{*}b^{n}c^{n} \cap a^{n}b^{n}c^{*}
+\rightarrow a^{n}b^{n}c^{n}$, so \PDA{} intersection may not be meaningful
+in some cases.  Detecting overlap between $n$ regexes requires
+$\frac{n(n-1)}{2}$ intersections to be calculated, giving $O(n^2x)$
+complexity, where $O(x)$ is the complexity of calculating intersection.
+This is certainly not a task to be performed on every parsing run:
+detecting overlap amongst the Postfix log parser's \numberOFrules{} rules
+would require calculating \numberOFruleINTERSECTIONS{} intersections.
 
@@
 log line logged by Postfix when a remote client connects to deliver mail:
-\newline{}\tab{}\verb!connect from client.example.com[1.2.3.4]!
+\newline{}\tab{}\verb!connect from client.example.com[192.0.2.3]!
 

### 2008-08-16 doc/sgai-2008.tex
Small improvements. Changes to reduce space consumption.

@@
 \newcommand{\showtable}[3]{
-    \begin{table}[hbtp]
+    \begin{table}[btp]
         \caption{#2}\label{#3}
@@
 
-\newcommand{\tableline}[0]{%
+\newcommand{\tabletopline}[0]{%
+    \hline%
+    \noalign{\smallskip}%
+}
+
+\newcommand{\tablebottomline}[0]{%
+    \noalign{\smallskip}%
+    \hline%
+}
+
+\newcommand{\tablemiddleline}[0]{%
     \noalign{\smallskip}%
@@
 
-\SVN$Id$
-\SVNId{}  \today{}
-
-%XXX DISCUSS COMPLEXITY\@?
+%\SVN$Id$
+%\SVNId{}  \today{}
 
@@
 paper but will be included in the thesis resulting from this work.
+The relationship to another procedural framework (ATN) is discussed below.
 
@@
 functions (\textit{actions\/}), each capable of dealing with an entire
-category of log lines (e.g.\ rejections), accompanied by a multitude of
-precise patterns (\textit{rules\/}), each of which matches all log lines of
-a specific type, and only that type (e.g.\ rejection by a specific DNSBL).
-XXX IMPROVE:\@ It is an accepted standard to separate the parsing procedure
-from the declarative grammar it operates with.  Part of the novelty here is
-in the way that the grammars (rules and actions) are themselves procedures.
-The relationship to another procedural framework (ATN) is discussed below.
-This architecture is ideally suited to parsing inputs where the input is
-not fully understood or does not conform to a fixed grammar: the
-architecture warns about unparsed inputs and other errors, but continues on
-parsing as best it can, allowing the developer of a new parser to decide
-which deficiencies are most important and require attention first, rather
-than being forced to fix the first error which arises.
+category of log lines (e.g.\ rejecting a mail), accompanied by a multitude
+of precise patterns (\textit{rules\/}), each of which matches all log lines
+of a specific type, and only that type (e.g.\ rejection by a specific
+DNSBL).  It is an accepted standard to separate the parsing procedure from
+the declarative grammar it operates with.  Part of the novelty here is in
+the way that the grammar is itself partially procedural (each action is a
+separate procedure).  This architecture is ideally suited to parsing inputs
+where the input is not fully understood or does not conform to a fixed
+grammar: the architecture warns about unparsed inputs and other errors, but
+continues on parsing as best it can, allowing the developer of a new parser
+to decide which deficiencies are most important and require attention
+first, rather than being forced to fix the first error which arises.
 
@@
 
-There is some similarity between the architecture and William Wood's
-Augmented Transition Networks (ATN)~\cite{atns, nlpip}, a tool used in
-Computational Linguistics for creating grammars to parse or generate
-sentences.  The resemblance between ATN and the architecture is accidental,
-but it is interesting how two different approaches have a similar division
-of responsibilities, while having different semantics.
+There is some similarity (shown in \tableref{Similarities with ATN})
+between the architecture and William Wood's Augmented Transition Networks
+(ATN)~\cite{atns, nlpip}, a tool used in Computational Linguistics for
+creating grammars to parse or generate sentences.  The resemblance between
+ATN and the architecture is accidental, but it is interesting how two
+different approaches have a similar division of responsibilities, while
+having different semantics.
 
@@
 
-\begin{table}[ht]
+\begin{table}[hbtp]
     \caption{Similarities with ATN}\label{Similarities with ATN}
@@
 ungrammatical or unparsed).  Generally rules specifying the
-\texttt{UNINTERESTING} action parse log lines which are not associated with
-a specific mail, e.g.\ warnings about configuration files changing.  The
-framework warns about every unparsed log line, to alert the user that they
-need to alter or extend their ruleset; the Postfix log parser successfully
-parses every input line in the \numberOFlogFILESall{} log files it is
-currently being tested with.  The remaining actions have only one or two
-associated rules: in some cases this is because there will only ever be one
-type of log line for that action, e.g.\ all log lines showing that a remote
-client has connected are matched by a single rule and handled by the
-\texttt{CONNECT} action; in other cases the action is required to address a
-deficiency in the log files, or a complication which arises during parsing.
-Taking the \texttt{CONNECT} action as an example, it creates a new data
-structure for each client connecting, and saves the data extracted by the
-rule (in memory, subsequently to be entered into the database).  If the
-data structure already exists it is treated as a symptom of a bug, and the
-action issues a warning containing the full contents of the data structure,
-plus the log line which has just been parsed.
+\texttt{UNINTERESTING} action are rules to parse log lines which are not
+associated with a specific mail, e.g.\ warnings about configuration files
+changing.  The framework warns about every unparsed log line, to alert the
+user that they need to alter or extend their ruleset; the Postfix log
+parser successfully parses every input line in the \numberOFlogFILESall{}
+log files it is currently being tested with.  The remaining actions have
+only one or two associated rules: in some cases this is because there will
+only ever be one type of log line for that action, e.g.\ all log lines
+showing that a remote client has connected are matched by a single rule and
+handled by the \texttt{CONNECT} action; in other cases the action is
+required to address a deficiency in the log files, or a complication which
+arises during parsing.  Taking the \texttt{CONNECT} action as an example,
+it creates a new data structure for each client connecting, and saves the
+data extracted by the rule (in memory, subsequently to be entered into the
+database).  If the data structure already exists it is treated as a symptom
+of a bug, and the action issues a warning containing the full contents of
+the data structure, plus the log line which has just been parsed.
 
@@
 more powerful than traditional regexes, e.g.\ it is possible to match
-correctly nested and balanced brackets to an arbitrary depth:
+correctly balanced brackets nested to an arbitrary depth:
 \newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
@@
     \begin{tabular}[]{ll}
-        \tableline{}%
+        \tabletopline{}%
         Attribute                 & Value                                            \\
-        \tableline{}%
+        \tablemiddleline{}%
         %program                   & \daemon{smtpd}                                   \\
@@
         %priority                  & 0                                                \\
-        \tableline{}%
+        \tablebottomline{}%
     \end{tabular}
@@
     \item [Matching rules to log lines is simple:]  The first matching rule
-        wins: there is no backtracking to try alternate rules, no attempt
-        is made to pick a \textit{best\/} rule; the first rule which
-        matches the line is the rule which parses it and determines the
-        action that will be invoked.
+        determines the action that will be invoked: there is no
+        backtracking to try alternate rules, no attempt is made to pick a
+        \textit{best\/} rule.
 
@@
 
+% XXX MAYBE CHANGE POSITIONING
 \begin{table}[htbp]
@@
     \begin{tabular}[]{ll}
-        \tableline{}
-        Component type  & Component in use                                  \\
-        \tableline{}
-        CPU             & One dual core 2.40GHz Intel\textregistered{}
-                            Core\texttrademark{}2 CPU,                      \\
-                        & with 32KB L1 cache and 4MB L2 cache.              \\
+        \tabletopline{}
+        Component  & Component used                                 \\
+        \tablemiddleline{}
+        CPU        & 1 dual core 2.40GHz Intel\textregistered{}
+                     Core\texttrademark{}2 CPU,                      
+                     with 32KB L1 and 4MB L2 cache.                 \\
         RAM        & 2GB 667 MHz DDR RAM\@.                         \\
-        Hard disk       & One Seagate Barracuda 7200 RPM 250GB SATA disk.   \\
-        \tableline{}
+        Hard disk  & 1 Seagate Barracuda 7200 RPM 250GB SATA disk.  \\
+        \tablebottomline{}
     \end{tabular}
@@
 lines factor}
-
-% XXX MAYBE ADD A TABLE SHOWING MEAN AND STDDEV OF THE RATIOS?
 
@@
 
+\newpage{} % XXX
 How any architecture scales as the number of rules increases is important,
@@
 is shown in \graphref{Percentage parsing time increase of maximum ruleset
-over minimum ruleset}.
+over minimum ruleset}.  It is clear from \graphref{Percentage parsing time
+increase of maximum ruleset over minimum ruleset} that the increased number
+of rules has a noticeable performance impact with reverse ordering, and a
+lesser impact with shuffled ordering.  The optimal ordering shows a mean
+increase of \input{build/include-full-ruleset-vs-minimum-ruleset} in
+parsing time for a \numberOFrulesMAXIMUMpercentage{} increase in the number
+of rules.  These results show that the architecture scales very well as the
+number of rules increases, and that sorting the rules is an important
+optimisation enabling this.
 \showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
@@
 time increase of maximum ruleset over minimum ruleset}
-
-It is clear from \graphref{Percentage parsing time increase of maximum
-ruleset over minimum ruleset} that the increased number of rules has a
-noticeable performance impact with reverse ordering, and a lesser impact
-with shuffled ordering.  The optimal ordering shows a mean increase of
-\input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
-\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
-results show that the architecture scales very well as the number of rules
-increases, and that sorting the rules is an important optimisation enabling
-this.
 

### 2008-08-16 doc/sgai-2008.tex
Small changes to reduce the paper to 14 pages.

@@
 Each action contains the code required to deal with a particular type of
-log lines, e.g.\ rejections.  The actions are parser-specific: each parser
+log line, e.g.\ rejections.  The actions are parser-specific: each parser
 author will need to write the actions required from scratch, unless
@@
 with a very skewed distribution of rules to actions as shown in
-\graphref{Distribution of rules per action}.  The distribution is skewed
-for the same reason the architecture is so successful in enabling users to
-parse new log lines: most new log lines can be parsed by a combination of a
-new rule and an existing action.  It is expected that parsers utilising
-this architecture will have a high ratio of rules to actions, hopefully
-resulting in simpler rules and clearer distinctions between log lines.
-\showgraph{build/graph-action-distribution}{Distribution of rules per
-action}{Distribution of rules per action} Unsurprisingly the action with
-the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the
-action which handles Postfix rejecting a mail; it is followed by
+\graphref{Distribution of rules per action}.  It is expected that parsers
+utilising this architecture will have a high ratio of rules to actions,
+hopefully resulting in simpler rules and clearer distinctions between
+inputs.  \showgraph{build/graph-action-distribution}{Distribution of rules
+per action}{Distribution of rules per action} Unsurprisingly the action
+with the highest number of associated rules is \texttt{DELIVERY\_REJECTED},
+the action which handles Postfix rejecting a mail; it is followed by
 \texttt{SAVE\_DATA}, the action responsible for handling informative
@@
         there is no facility for rules to consume more input or push unused
-        input back onto the input stream (however actions may change input
-        lines and cause the modified line to be re-parsed).  This was not a
-        deliberate decision, rather a consequence of the line oriented
-        nature of the log files; more flexible approaches could be pursued.
+        input back onto the input stream.  This was not a deliberate
+        decision, rather a consequence of the line oriented nature of the
+        log files; more flexible approaches could be pursued.
 
@@
 parsing times averaged.  Saving results to the database was disabled for
-the test runs, as that dominates the run time of the program, and the tests
-are aimed at measuring the speed of the Postfix log parser rather than the
-speed of the database and the disks the database is stored on.  The
-computer used for test runs was a Dell Optiplex 745 described in
-\tableref{Computer used to generate statistics}.  Parsing all
+the test runs, as the tests are aimed at measuring the speed of the Postfix
+log parser rather than the speed of the database and the disks the database
+is stored on.  The computer used for test runs was a Dell Optiplex 745
+described in \tableref{Computer used to generate statistics}.  Parsing all
 \numberOFlogFILES{} log files in one run took
@@
 
-\newpage{} % XXX
 How any architecture scales as the number of rules increases is important,

### 2008-08-19 doc/sgai-2008.tex
Corrections to the first half of the paper.

@@
     \begin{figure}[btp]%
+        \includegraphics{#1}%
         \caption{#2}\label{#3}%
-        \includegraphics{#1}%
     \end{figure}%
@@
 \newcommand{\tableref}[1]{%
-    \refwithlabel{tbl.}{#1}%
+    \refwithlabel{table}{#1}%
 }
@@
     understanding of the parser's internal workings.  This paper presents a
-    novel parsing architecture designed for Postfix log files which aims to
-    make the process of parsing new inputs as simple as possible, enabling
-    users to trivially add new rules (to parse variants of existing inputs)
-    and relatively easily add new actions (to process a previously unknown
-    class of input).  The architecture scales linearly as the number of
-    rules and/or size of input increases, making it suitable for parsing
-    large corpora or months of accumulated data.
+    novel parsing architecture, designed for parsing Postfix log files,
+    which aims to make the process of parsing new inputs as simple as
+    possible, enabling users to trivially add new rules (to parse variants
+    of existing inputs) and relatively easily add new actions (to process a
+    previously unknown class of input).  The architecture scales linearly
+    as the number of rules and/or size of input increases, making it
+    suitable for parsing large corpora or months of accumulated data.
 }
@@
 \newacronym{FA}{Finite Automata}{}
-\newacronym{CFL}{Context Free Languages}{}
+\newacronym{CFL}{Context-Free Languages}{}
 \newacronym{SLCT}{Simple Logfile Clustering Tool}{}
@@
 mistakenly rejected will receive an immediate non-delivery notice; load is
-reduced on mail servers (allowing more intensive content-based techniques
-to be used); users have less mail to wade through.  Improving the
-performance of anti-spam techniques which are applied when mail is being
-transferred via \SMTP{}\footnotemark{} is the goal of this project, by
-providing a platform for reasoning about anti-spam filters.  The approach
-chosen to measure performance was to analyse the log files produced by the
-mail server, rather than modifying the mail server to generate performance
-statistics: this approach improves the chances of other sites testing and
-utilising the software.  The need arose for a parser capable of dealing
-with the great number and variety of log lines produced by
-Postfix~\cite{postfix}, the \SMTP{} server in use.  The parser must be
-designed so that adding support for parsing new log lines is a simple task,
-as the log lines to be parsed will change over time.  The variety in log
-lines occurs for several reasons:\footnotetext{Simple Mail Transfer
-Protocol transfers mail from the sender to one or more recipients across
-the Internet.  It is a simple, human readable, plain text protocol, making
-it quite easy to test and debug problems with it.  The original protocol
-definition is RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}
+reduced on the accepting mail server (allowing more intensive content-based
+techniques to be used); users have less mail to wade through.  Improving
+the performance of anti-spam techniques which are applied when mail is
+being transferred via \SMTP{}\footnotemark{} is the goal of this project,
+by providing a platform for reasoning about anti-spam filters.  The
+approach chosen to measure performance is to analyse the log files produced
+by the mail server, rather than modifying Postfix~\cite{postfix}, the
+\SMTP{} server in use, to generate performance statistics: this approach
+improves the chances of other sites testing and utilising the software.
+The need arose for a parser capable of dealing with the great number and
+variety of log lines produced by Postfix; the parser must be designed so
+that adding support for parsing new inputs is a simple task, as the log
+lines to be parsed will change over time.  The variety in log lines occurs
+for several reasons:\footnotetext{Simple Mail Transfer Protocol transfers
+mail across the Internet from the sender to one or more recipients.  It is
+a simple, human readable, plain text protocol, making it quite easy to test
+and debug problems with it.  The original protocol definition is
+RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}
 
@@
 of precise patterns (\textit{rules\/}), each of which matches all log lines
-of a specific type, and only that type (e.g.\ rejection by a specific
+of a specific type and only that type (e.g.\ rejection by a specific
 DNSBL).  It is an accepted standard to separate the parsing procedure from
-the declarative grammar it operates with.  Part of the novelty here is in
+the declarative grammar it operates with; part of the novelty here is in
 the way that the grammar is itself partially procedural (each action is a
@@
     \item [Rules]  The rules are responsible for classifying or matching
-        log lines, specifying the action to invoke and the data to be
-        extracted from the log line.
+        log lines, specifying the action to invoke and the regex which
+        matches the log line and extracts data.
 
@@
 to entry for casual users who need to parse new log lines, e.g.\ part-time
-systems administrators attempting to reduce spam.  Decoupling the actions
-from the framework simplifies both framework and actions: the framework
-provides services to the actions, but does not need to perform any tasks
-specific to the input being parsed; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately and correctly processing the information provided by rules.
+systems administrators attempting to combat and reduce spam.  Decoupling
+the actions from the framework simplifies both framework and actions: the
+framework provides services to the actions, but does not need to perform
+any tasks specific to the input being parsed; actions benefit from having
+services provided by the framework, freeing them to concentrate on the task
+of accurately and correctly processing the information provided by rules.
 
@@
 ATN and the architecture is accidental, but it is interesting how two
-different approaches have a similar division of responsibilities, while
-having different semantics.
+different approaches share a similar division of responsibilities, while
+having different quite semantics.
 
@@
     \begin{tabular}[]{lll}
-        \hline
-        \noalign{\smallskip}
+        \tabletopline{}
         ATN           & Architecture    & Similarity                                \\
-        \noalign{\smallskip}
-        \hline
-        \noalign{\smallskip}
-        Networks      & Framework       & Determines the sequence of
-                                          transitions or actions which        \\
-                      &                 & constitutes a valid input.          \\
-        Transitions   & Actions         & Save data and impose conditions the
-                                          input must meet to be               \\
-                      &                 & considered valid.                   \\
+        \tablemiddleline{}
+        Networks      & Framework       & Determines the sequence of transitions or \\
+                      &                 & actions which constitutes a valid input.  \\
+        Transitions   & Actions         & Save data and impose conditions the       \\
+                      &                 & input must meet to be considered valid.   \\
         Abbreviations & Rules           & Responsible for classifying input.        \\
-        \noalign{\smallskip}
-        \hline
-        \noalign{\smallskip}
+        \tablebottomline{}
     \end{tabular}
@@
 log line, e.g.\ rejections.  The actions are parser-specific: each parser
-author will need to write the actions required from scratch, unless
+author will need to write the required actions from scratch, unless
 extending an existing parser.  In the Postfix log parser developed for this
 project there are \numberOFactions{} actions and \numberOFrules{} rules,
-with a very skewed distribution of rules to actions as shown in
+with a very uneven distribution of rules to actions as shown in
 \graphref{Distribution of rules per action}.  It is expected that parsers
 utilising this architecture will have a high ratio of rules to actions,
-hopefully resulting in simpler rules and clearer distinctions between
-inputs.  \showgraph{build/graph-action-distribution}{Distribution of rules
-per action}{Distribution of rules per action} Unsurprisingly the action
-with the highest number of associated rules is \texttt{DELIVERY\_REJECTED},
-the action which handles Postfix rejecting a mail; it is followed by
+hopefully resulting in simpler rules and clearer distinctions between the
+inputs parsed by different rules.
+\showgraph{build/graph-action-distribution}{Distribution of rules per
+action}{Distribution of rules per action} Unsurprisingly the action with
+the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the
+action which handles Postfix rejecting a mail; it is followed by
 \texttt{SAVE\_DATA}, the action responsible for handling informative
@@
 \texttt{UNINTERESTING} action are rules to parse log lines which are not
-associated with a specific mail, e.g.\ warnings about configuration files
+associated with a specific mail, e.g.\ notices about configuration files
 changing.  The framework warns about every unparsed log line, to alert the
@@
 parser successfully parses every input line in the \numberOFlogFILESall{}
-log files it is currently being tested with.  The remaining actions have
-only one or two associated rules: in some cases this is because there will
+log files it is currently being tested with.  The remaining actions only
+have one or two associated rules: in some cases this is because there will
 only ever be one type of log line for that action, e.g.\ all log lines
@@
 it creates a new data structure for each client connecting, and saves the
-data extracted by the rule (in memory, subsequently to be entered into the
-database).  If the data structure already exists it is treated as a symptom
-of a bug, and the action issues a warning containing the full contents of
-the data structure, plus the log line which has just been parsed.
+data extracted by the rule in memory, subsequently to be entered into the
+database.  If a data structure already exists for that connection it is
+treated as a symptom of a bug, and the action issues a warning containing
+the full contents of the existing data structure, plus the log line which
+has just been parsed.
 
@@
 less frequently than with a monolithic architecture, thus failures should
-be less likely, and will be easier to test for and diagnose.  The parser
-can be written in an object oriented style, allowing sub-classes to extend
-or override actions in addition to adding new actions; because each action
-is an independent function, the sub-class need only modify the action it is
-interested in, rather than reproducing large chunks of functionality.
+be less likely, and will be easier to test for and diagnose.  The parser is
+written in an object oriented style, allowing sub-classes to extend or
+override actions in addition to adding new actions; because each action is
+an independent procedure, the sub-class need only modify the action it is
+overriding, rather than reproducing large chunks of functionality.
 
@@
 modifications to a single existing action to work around the difficulties;
-the remainder were resolved by adapting actions to pass information to
-subsequent actions (via the framework), which check for the extra
-information and change their behaviour appropriately.
+the remainder were resolved by adapting existing actions to cooperate and
+exchange extra data, changing their behaviour as appropriate.
 
@@
 
-Adding new rules is simple: define all the required attributes and append
-the new rule to wherever the existing rules are stored.  The Postfix log
-parser supplies a utility based on \SLCT{}~\cite{slct-paper} to aid in
-producing regexes from previously unparsed inputs.  Storing the rules
-separately allows other approaches to rule management to be used, e.g.\
-instead of manually adding new rules when unparsed log lines are seen,
+Parsing new inputs is generally achieved by creating a new rule that pairs
+an existing action with a new regex.  Adding new rules is simple: define
+all the required attributes and append the new rule to wherever the
+existing rules are stored.  The Postfix log parser supplies a utility based
+on \SLCT{}~\cite{slct-paper} to aid in producing regexes from previously
+unparsed inputs.  Storing the rules separately allows other approaches to
+rule management to be used, e.g.\ instead of manually adding new rules,
 machine learning techniques could be used to automatically generate new
@@
 taken the choice of machine learning technique would be constrained by the
-size of typical data sets: the \numberOFlogFILES{} log files used to
-generate statistics contain \numberOFlogLINEShuman{} log lines.  Techniques
-which require the full data set when training would be impractical;
-Instance Based Learning~\cite{instance-based-learning} techniques which
+size of typical data sets (see \sectionref{Results}); techniques which
+require the full data set when training would be impractical; Instance
+Based Learning~\cite{instance-based-learning} techniques which
 automatically determine which inputs from the training set are valuable and
@@
 manageable size.  This avenue of research and development has not been
-pursued by the authors, but could easily be independently undertaken.
+pursued by the authors, but could easily be undertaken independently.
 
@@
 prior cooperation or modification required.  Overlapping rules should have
-a priority attribute to specify the order in which they should be used;
-negative priorities may be useful for catchall rules.
+a priority attribute to specify their relative ordering; negative
+priorities may be useful for catchall rules.
 
@@
 equivalent in computational power to \PDA{}, and intersection of two
-\CFL{s} is not closed (i.e.\ intersection can produce a language which
-cannot be parsed by a \CFL{}), e.g.\ $a^{*}b^{n}c^{n} \cap a^{n}b^{n}c^{*}
-\rightarrow a^{n}b^{n}c^{n}$, so \PDA{} intersection may not be meaningful
-in some cases.  Detecting overlap between $n$ regexes requires
-$\frac{n(n-1)}{2}$ intersections to be calculated, giving $O(n^2x)$
+\CFL{s} is not closed, i.e.\ intersection can produce a language which
+cannot be parsed by a \CFL{}, e.g.\ $a^{*}b^{n}c^{n} \cap a^{n}b^{n}c^{*}
+\rightarrow a^{n}b^{n}c^{n}$, so the results of \PDA{} intersection may not
+be meaningful in some cases.  Detecting overlap between $n$ regexes
+requires $n(n-1)/2$ intersections to be calculated, resulting in $O(n^2x)$
 complexity, where $O(x)$ is the complexity of calculating intersection.
@@
 categories: regexes which match every input line, and regexes which consume
-large amounts of CPU time during matching.  Defining a regex which matches
-every input line is trivial: \verb!/^/! matches the start of every input
-line.  In general a regex will not consume a large amount of CPU time when
-the match is successful; usually CPU time is consumed by a regex with a lot
-of alteration and variable quantifiers failing to match.  This topic is
-beyond the scope of this paper, see~\cite{mastering-regular-expressions}
-for in-depth discussion.
+excessive amounts of CPU time during matching.  Defining a regex which
+matches every input line is trivial: \verb!/^/! matches the start of every
+input line.  In general a regex will not consume much CPU time when the
+match is successful; usually CPU time is consumed by a regex with a lot of
+alteration and variable quantifiers failing to match.  This topic is beyond
+the scope of this paper, see~\cite{mastering-regular-expressions} for
+in-depth discussion.
 
@@
 
-    \item [Context-free rules:]  Rules are context free: they do not take
-        into account past or future inputs.  In context-free
-        grammar terms the parser rules could be described as:
+    \item [Context-free rules:]  Rules are context-free: they do not take
+        into account past or future inputs.  In context-free grammar terms
+        the parser rules could be described as:
         $\text{\textless{}input\textgreater{}} \mapsto \text{rule-1} |
@@
 
-    \item [Matching rules to log lines is simple:]  The first matching rule
-        determines the action that will be invoked: there is no
+    \item [Matching rules against log lines is simple:]  The first matching
+        rule determines the action that will be invoked: there is no
         backtracking to try alternate rules, no attempt is made to pick a
@@
         grammar:] Unlike traditional parsers such as those used when
-        compiling a programming language, this architecture does not have a
-        fixed grammar specification which inputs must adhere to.  The
-        architecture is capable of dealing with interleaved inputs, out of
-        order inputs, and ambiguous inputs where heuristics must be applied
-        --- all have arisen and been successfully dealt with in the Postfix
-        log parser.
+        compiling a programming language, this architecture does not
+        require a fixed grammar specification which inputs must adhere to.
+        The architecture is capable of dealing with interleaved inputs, out
+        of order inputs, and ambiguous inputs where heuristics must be
+        applied --- all have arisen and been successfully accommodated in
+        the Postfix log parser.
 
@@
 \section{Results}
+
+\label{Results}
 
@@
 linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
-despite their larger than average size (due to a mail forwarding loop
+despite their large size (due to a mail forwarding loop
 resulting in a greatly increased number of mails delivered and log lines

### 2008-08-20 doc/sgai-2008.tex
Corrections from the second half of the paper.

@@
 
-\title*{A user-extensible and adaptable parser architecture}
+\title*{A User-Extensible and Adaptable Parser Architecture}
 \author{John Tobin and Carl Vogel}
@@
     \begin{tabular}[]{lll}
-        \tabletopline{}
+        \tabletopline{}%
         ATN           & Architecture    & Similarity                                \\
-        \tablemiddleline{}
+        \tablemiddleline{}%
         Networks      & Framework       & Determines the sequence of transitions or \\
@@
         Abbreviations & Rules           & Responsible for classifying input.        \\
-        \tablebottomline{}
+        \tablebottomline{}%
     \end{tabular}
@@
 is stored on.  The computer used for test runs was a Dell Optiplex 745
-described in \tableref{Computer used to generate statistics}.  Parsing all
+described in \tableref{Computer used to generate statistics}; it was
+dedicated to the task of gathering statistics from test runs, and was not
+used for any other purpose while test runs were ongoing --- any services
+not necessary for running the tests were disabled.  Parsing all
 \numberOFlogFILES{} log files in one run took
-\input{build/include-timing-run-duration}.  The computer was dedicated to
-the task of gathering statistics from test runs, and was not used for any
-other purpose while test runs were ongoing; any services not necessary for
-running the tests were disabled.
+\input{build/include-timing-run-duration}.
 
@@
     \begin{tabular}[]{ll}
-        \tabletopline{}
+        \tabletopline{}%
         Component  & Component used                                 \\
-        \tablemiddleline{}
+        \tablemiddleline{}%
         CPU        & 1 dual core 2.40GHz Intel\textregistered{}
@@
         Hard disk  & 1 Seagate Barracuda 7200 RPM 250GB SATA disk.  \\
-        \tablebottomline{}
+        \tablebottomline{}%
     \end{tabular}
@@
 An important property of a parser is how parsing time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-\Graphref{Parsing time, file size, and number of lines} shows the parsing
-time in seconds, file size in MB, and number of log lines in tens of
-thousands, for each of the \numberOFlogFILES{} log files.  All three lines
-run roughly in parallel, giving the impression that the algorithm scales
-linearly with input size.  This impression is borne out by
-\graphref{parsing time vs file size vs number lines factor}: the ratios are
-quite tightly banded across the graph, showing that the algorithm scales
-linearly.  The ratio increases (i.e.\ improves) for log files 22 and 62--68
-despite their large size (due to a mail forwarding loop
-resulting in a greatly increased number of mails delivered and log lines
-generated).
+input size: linearly, polynomially, or exponentially?  \Graphref{Parsing
+time, file size, and number of lines} shows the parsing time in seconds,
+file size in MB, and number of log lines in tens of thousands, for each of
+the \numberOFlogFILES{} log files.  All three lines run roughly in
+parallel, giving the impression that the algorithm scales linearly with
+input size.  This impression is borne out by \graphref{parsing time vs file
+size vs number lines factor}: the ratios are tightly banded across the
+graph, showing that the algorithm scales linearly.  The ratios increase
+(i.e.\ improve) for log files 22 and 62--68 despite their large size due to
+a mail forwarding loop resulting in a greatly increased number of mails
+delivered and log lines generated.
 
@@
 
-At the time of writing there are \numberOFrules{} different rules, with the
-top 10\% matching the vast majority of the log lines, and the remaining log
+At the time of writing the Postfix log parser has \numberOFrules{} different rules, with the
+top 10\% matching the vast majority of log lines, and the remaining log
 lines split across the other 90\% of the rules in a Power Law distribution
 (as shown in \graphref{rule hits graph}).  Assuming that the distribution
-of log lines is reasonably steady over time, parser efficiency should
+of log lines is reasonably consistent over time, parser efficiency should
 benefit from trying more frequently matching rules before those which match
@@
 
-    \item [optimal]  Hypothetically the most optimal order: rules which
-        match most often will be tried first.
+    \item [optimal]  Hypothetically the best order: rules which match most
+        often will be tried first.
 
     \item [shuffle] Random ordering, intended to represent an unsorted rule
-        set.  The rules will be shuffled once before use and will retain
-        that ordering for the entirety of the log file.  Note that the
-        ordering will change every time the parser is executed, so 10
-        different orderings will be generated for each log file in the test
-        run.
+        set.  Note that the ordering will change every time the parser is
+        executed, so 10 different orderings will be generated for each log
+        file in the test run.
 
@@
 shuffled ordering.  Overall this optimisation provides a modest but
-worthwhile reduction in parsing time of approximately 10\%.
+worthwhile reduction in parsing time of approximately 10\%.  Optimal rule
+ordering is also beneficial in other ways, as shown in
+\sectionref{Architecture Scalability: Number of Rules}.
 \showgraph{build/graph-hits}{Hits per rule}{rule hits graph}
@@
 
+\label{Architecture Scalability: Number of Rules}
+
 How any architecture scales as the number of rules increases is important,
 but it is particularly important in this architecture because it is
-expected that there will be a large number of rules.  There are
+expected that typical parsers will have a large number of rules.  There are
 \numberOFrules{} rules in the full Postfix log parser ruleset, whereas the
@@
 is shown in \graphref{Percentage parsing time increase of maximum ruleset
-over minimum ruleset}.  It is clear from \graphref{Percentage parsing time
-increase of maximum ruleset over minimum ruleset} that the increased number
-of rules has a noticeable performance impact with reverse ordering, and a
-lesser impact with shuffled ordering.  The optimal ordering shows a mean
-increase of \input{build/include-full-ruleset-vs-minimum-ruleset} in
-parsing time for a \numberOFrulesMAXIMUMpercentage{} increase in the number
-of rules.  These results show that the architecture scales very well as the
-number of rules increases, and that sorting the rules is an important
-optimisation enabling this.
-\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
-time increase of maximum ruleset over minimum ruleset}{Percentage parsing
-time increase of maximum ruleset over minimum ruleset}
+over minimum ruleset}.  It is clear that the increased number of rules has
+a noticeable performance impact with reverse ordering, and a lesser impact
+with shuffled ordering.  The optimal ordering shows a mean increase of
+\input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
+\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
+results show that the architecture scales very well as the number of rules
+increases, and that sorting the rules is an important optimisation enabling
+this.  \showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage
+parsing time increase of maximum ruleset over minimum ruleset}{Percentage
+parsing time increase of maximum ruleset over minimum ruleset}
 
@@
 
-The Postfix log parser's coverage of log files is separated into two parts:
-log line coverage and mail coverage (the former is a requirement for the
-latter to be achieved).  Improving the former is less intrusive, as it just
-requires new rules to be written; improving the latter is more intrusive as
-it requires changes to actions, and it can also be much harder to notice a
-deficiency.
+The Postfix log parser requires two different types of coverage to be
+measured: log lines parsed and mail delivery attempts correctly understood
+(the former is a requirement for the latter to be achieved).  Improving the
+former is less intrusive, as it just requires new rules to be written;
+improving the latter is more intrusive as it requires changes to actions,
+and it can also be much harder to notice a deficiency.
 
-Warnings are issued for any log lines which are not parsed; no warnings are
-issued for unparsed log lines while parsing the \numberOFlogFILES{} test
-log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: manually verifying that
-the correct regex parsed each of the \numberOFlogLINES{} log lines is
-infeasible, making it impossible to quantify the false positive rate.  A
-random sample of 6039 log lines (0.00994\% of the total) was parsed and the
-results manually verified by inspection to ensure that the correct regex
-parsed each log line.  The sample contained zero false positives, and this
-check has been automated to ensure continued accuracy.
+Measuring Postfix log parser coverage begins with log line coverage.
+Warnings are issued for any log lines which are not parsed; no such
+warnings are issued while parsing the \numberOFlogFILES{} test log files,
+so it can be safely concluded that there are zero false negatives.  False
+positives are harder to quantify: manually verifying that the correct rule
+parsed each of the \numberOFlogLINES{} log lines is infeasible, making it
+impossible to quantify the false positive rate.  A random sample of 6039
+log lines (0.00994\% of the total) was parsed and the results manually
+verified by inspection to ensure that the correct rule parsed each log
+line.  The sample contained zero false positives, and this check has been
+automated to ensure continued accuracy.
 
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The implementation can dump its state tables in a
-human readable form; examining these tables with reference to the log files
-is the best way to detect mails which were not handled properly.  The
-implementation issues warnings when it detects any errors, some of which
-may alert the user to a problem.  There should be few or no warnings when
-parsing, and when finished parsing the state table should only contain
-entries for mails which had yet to be delivered when the log files ended,
-or were accepted before the log files began.  A second sample of 6000 log
-lines was parsed with all debugging options enabled, resulting in 167,448
-lines of output.  All 167,448 lines were examined in conjunction with the
-input log file and a dump of the resulting database, verifying that for
-each of the input lines the implementation used the correct rule and
-executed the correct action, which in turn produced the correct result and
-inserted the correct data in the database.  The log segment produced 4
-warnings (about mails which had started before the log file segment), 10
-mails remaining in the state tables, and 1625 connections correctly entered
-in the database.  No error messages were produced, therefore there were no
-false negatives.  Given the evidence detailed above, the authors are
-confident that the false positive rate when reconstructing a mail is zero.
+The proportion of mail delivery attempts correctly understood is much more
+difficult to determine accurately than coverage of log lines.  The
+implementation can dump its state tables in a human readable form;
+examining these tables with reference to the log files is the best way to
+detect mails which were not handled properly.  The implementation issues
+warnings when it detects any errors, which should alert the user to a
+problem.  There should be few or no warnings when parsing, and when
+finished parsing the state table should only contain entries for mails
+which had yet to be delivered when the log files ended, or were accepted
+before the log files began.  A second sample of 6000 log lines was parsed
+with all debugging options enabled, resulting in 167,448 lines of output.
+All 167,448 lines were examined in conjunction with the input log file and
+a dump of the resulting database, verifying that for each of the input
+lines the implementation used the correct rule and executed the correct
+action, which in turn produced the correct result and inserted the correct
+data in the database.  The log segment produced 4 warnings (about mails
+which had started before the log file segment), 10 mails remaining in the
+state tables, and 1625 connections correctly entered in the database.  No
+error messages were produced, therefore there were no false negatives.
+Given the evidence detailed above, the authors are confident that the false
+positive rate when reconstructing a mail is zero.
+
+Experience implementing the Postfix log parser shows that full input
+coverage is relatively easy to achieve with this architecture, and that
+with sufficient time and effort full understanding of the input is
+possible.  Postfix log files would require substantial time and effort to
+correctly parse regardless of the architecture used; this architecture
+enables an iterative approach to be used (similar to Stepwise
+Refinement~\cite{stepwise-refinement}), as is practiced in many other
+software engineering disciplines.
 
@@
 a new one with an appropriate regex and the task is complete; knowledge of
-the framework is not required when writing rules.  Parsing a new category
+the framework is not required.  Parsing a new category
 of input is achieved by writing a new action and appropriate rules; quite
@@
 facilities.  The decoupling of rules from actions allows different sets of
-rules to be used with the same actions: e.g.\ a parser might have actions
+rules to be used with the same actions, e.g.\ a parser might have actions
 to process versions one and two of a file format; the parser will then
-parse either version one, version two, or both versions depending on the
+parse either version one, version two, or both versions, depending on the
 ruleset selected.  The architecture makes it possible to apply commonly
 used programming techniques such as object orientation, inheritance,
-composition, delegation and roles when designing and implementing a parser,
+composition, delegation, and roles when designing and implementing a parser,
 simplifying the process of working within a team with shared
-responsibility, or when developing and testing additional functionality.
+responsibility for a parser, or when developing and testing additional functionality.
 This architecture is ideally suited to parsing inputs where the input is
@@
 this approach could be applied to optimising the behaviour of any system
-given sufficient logging to analyse.  An alternative approach to black box
+given sufficient log messages to analyse.  An alternative approach to black box
 optimisation which uses application profiling in conjunction with the
@@
 systems administrators to monitor the effectiveness of their anti-spam
-measures and adapt their defences to combat the new techniques used by
-those sending spam.  The parser is a fully usable application, built to
-address a genuine need, rather than a proof of concept whose sole purpose
-is to illustrate a new idea; it deals with the oddities and difficulties
-which occur in the real world, rather than a clean, idealised scenario
-developed to showcase the best features of a new approach.
+measures and adapt their defences to combat new techniques used by those
+sending spam.  This parser is a fully usable application, built to address
+a genuine need, rather than a proof of concept whose sole purpose is to
+illustrate a new idea; it deals with the oddities and difficulties which
+occur in the real world, rather than a clean, idealised scenario developed
+to showcase the best features of a new approach.
 

### 2008-08-20 doc/sgai-2008.tex
Final suggestions from Carl. Remaining corrections from my proofreading. Resolved all XXX items. Improved bibliography entries.

@@
 % $Id$
-% XXX REMOVE draft OPTION
-\documentclass[draft]{svmult}
+\documentclass{svmult}
+%\documentclass[draft]{svmult}
 
@@
 \usepackage{svn}
-
-% XXX REMOVE THESE AND CHECK RESULTING LENGTH
-%\setlength{\floatsep}{0pt}
-%\setlength{\intextsep}{0pt}
-%\setlength{\textfloatsep}{0pt}
 
@@
 \author{John Tobin and Carl Vogel}
-\institute{John Tobin \at{} School of Computer Science and Statistics,
-Trinity College, Dublin 2, Ireland. \newline{} \email{tobinjt@cs.tcd.ie}
-\and{} Carl Vogel \at{} School of Computer Science and Statistics,
-Trinity College, Dublin 2, Ireland. \newline{} \email{vogel@cs.tcd.ie}}
+\institute{John Tobin \email{tobinjt@cs.tcd.ie}
+            \and{} Carl Vogel \email{vogel@cs.tcd.ie}
+            \at{} School of Computer Science and Statistics,
+                    Trinity College, Dublin 2, Ireland.
+                    \newline{}
+                    Supported by Science Foundation Ireland RFP
+                    05/RF/CMS002.
+            }
+
 \maketitle
@@
 some cases the new action will require the cooperation of other actions,
-e.g.\ to set or check a flag; there is a possibility of introducing failure
-by modifying existing actions, but modifications will be smaller and occur
-less frequently than with a monolithic architecture, thus failures should
-be less likely, and will be easier to test for and diagnose.  The parser is
-written in an object oriented style, allowing sub-classes to extend or
-override actions in addition to adding new actions; because each action is
-an independent procedure, the sub-class need only modify the action it is
-overriding, rather than reproducing large chunks of functionality.
+e.g.\ to set or check a flag.  There is a possibility of introducing
+failure when modifying existing actions, but the modifications will be
+smaller and occur less frequently than with a monolithic architecture, thus
+failures will be less likely and will be easier to test for and diagnose.
+The parser is written in an object oriented style, allowing sub-classes to
+extend or override actions in addition to adding new actions; because each
+action is an independent procedure, the sub-class need only modify the
+action it is overriding, rather than reproducing large chunks of
+functionality.
 
@@
 intersection of two \FA{}\@.  One formula for \FA{} intersection is given
-in~\cite{intersection-of-NFA-using-Z}: $FA1 \cap{} FA2 =
+in for example~\cite{intersection-of-NFA-using-Z}: $FA1 \cap{} FA2 =
 \overline{(\overline{FA1} \cup{} \overline{FA2})}$.  Perl 5.10 regexes are
@@
 cannot be parsed by a \CFL{}, e.g.\ $a^{*}b^{n}c^{n} \cap a^{n}b^{n}c^{*}
-\rightarrow a^{n}b^{n}c^{n}$, so the results of \PDA{} intersection may not
-be meaningful in some cases.  Detecting overlap between $n$ regexes
-requires $n(n-1)/2$ intersections to be calculated, resulting in $O(n^2x)$
+\rightarrow a^{n}b^{n}c^{n}$, so the results of \PDA{} intersection may be
+intractable in some cases.  Detecting overlap between $n$ regexes requires
+$n(n-1)/2$ intersections to be calculated, resulting in $O(n^2x)$
 complexity, where $O(x)$ is the complexity of calculating intersection.
@@
 the test runs, as the tests are aimed at measuring the speed of the Postfix
-log parser rather than the speed of the database and the disks the database
-is stored on.  The computer used for test runs was a Dell Optiplex 745
-described in \tableref{Computer used to generate statistics}; it was
-dedicated to the task of gathering statistics from test runs, and was not
-used for any other purpose while test runs were ongoing --- any services
-not necessary for running the tests were disabled.  Parsing all
-\numberOFlogFILES{} log files in one run took
+log parser rather than the speed of the database.  The computer used for
+test runs was a Dell Optiplex 745 described in \tableref{Computer used to
+generate statistics}; it was dedicated to the task of gathering statistics
+from test runs, and was not used for any other purpose while test runs were
+ongoing --- any services not necessary for running the tests were disabled.
+Parsing all \numberOFlogFILES{} log files in one run took
 \input{build/include-timing-run-duration}.
 
-% XXX MAYBE CHANGE POSITIONING
 \begin{table}[htbp]
@@
 
-At the time of writing the Postfix log parser has \numberOFrules{} different rules, with the
-top 10\% matching the vast majority of log lines, and the remaining log
-lines split across the other 90\% of the rules in a Power Law distribution
-(as shown in \graphref{rule hits graph}).  Assuming that the distribution
-of log lines is reasonably consistent over time, parser efficiency should
-benefit from trying more frequently matching rules before those which match
-less frequently.  To test this hypothesis three full test runs were
-performed with different rule orderings:
+At the time of writing the Postfix log parser has \numberOFrules{}
+different rules: the top 10\% match the vast majority of log lines, with
+the remaining log lines split across the other 90\% of the rules.  The
+distribution resembles a Power Law distribution, as shown in \graphref{rule
+hits graph}.  Assuming that the distribution of log lines is reasonably
+consistent over time, parser efficiency should benefit from trying more
+frequently matching rules before those which match less frequently.  To
+test this hypothesis three full test runs were performed with different
+rule orderings:
 

### 2008-08-20 doc/sgai-2008.tex
Spelling checked. Checked URLs in bibliography. Run chktex on the bibliographies. Made ATN an acronym.

@@
 \newacronym{SMTP}{Simple Mail Transfer Protocol}{}
+\newacronym{ATN}{Augmented Transition Networks}{}
 
@@
 There is some similarity (shown in \tableref{Similarities with ATN})
-between the architecture and William Wood's Augmented Transition Networks
-(ATN)~\cite{atns, nlpip}, a tool used in Computational Linguistics for
+between the architecture and William Wood's 
+\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
 creating grammars to parse or generate sentences.  The resemblance between
-ATN and the architecture is accidental, but it is interesting how two
+\ATN{} and the architecture is accidental, but it is interesting how two
 different approaches share a similar division of responsibilities, while

### 2008-08-21 doc/sgai-2008.tex
Corrections after proofreading.

@@
 \newacronym{CFL}{Context-Free Languages}{}
-\newacronym{SLCT}{Simple Logfile Clustering Tool}{}
 \newacronym{SMTP}{Simple Mail Transfer Protocol}{}
@@
 filters currently in use, optimising the order and membership of the set
-based on that analysis, and developing supplemental filters where there is
-a deficiency.  Most anti-spam techniques are
-content-based~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms},
+based on that analysis, and developing supplemental filters where
+deficiencies are identified.  Most anti-spam techniques are
+content-based~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}
 which requires accepting the mail in the first place, but rejecting mail
@@
 approach chosen to measure performance is to analyse the log files produced
-by the mail server, rather than modifying Postfix~\cite{postfix}, the
-\SMTP{} server in use, to generate performance statistics: this approach
-improves the chances of other sites testing and utilising the software.
-The need arose for a parser capable of dealing with the great number and
-variety of log lines produced by Postfix; the parser must be designed so
-that adding support for parsing new inputs is a simple task, as the log
-lines to be parsed will change over time.  The variety in log lines occurs
-for several reasons:\footnotetext{Simple Mail Transfer Protocol transfers
-mail across the Internet from the sender to one or more recipients.  It is
-a simple, human readable, plain text protocol, making it quite easy to test
-and debug problems with it.  The original protocol definition is
+by the \SMTP{} server in use, Postfix~\cite{postfix}, rather than modifying
+it to generate statistics: this approach improves the chances of other
+sites testing and utilising the software.  The need arose for a parser
+capable of dealing with the great number and variety of log lines produced
+by Postfix: the parser must be designed so that adding support for parsing
+new inputs is a simple task, as the log lines to be parsed will change over
+time.  The variety in log lines occurs for several
+reasons:\footnotetext{Simple Mail Transfer Protocol transfers mail across
+the Internet from the sender to one or more recipients.  It is a simple,
+human readable, plain text protocol, making it quite easy to test and debug
+problems with it.  The original protocol definition is
 RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}
@@
 It was hoped to reuse an existing parser rather than starting from scratch:
-several parsers were considered and rejected because their parsing coverage
-was too low, their parsing too inexact, and/or they did not extract
-sufficient data.  The effort required to adapt and improve an existing
-parser was judged to be greater than the effort to write one, as the
-techniques used by the existing parsers severely limited their potential:
-some ignored the majority of log lines, parsing specific log lines with
-great accuracy, but without any provision for readily parsing new or
-similar log lines; others sloppily parsed the majority of log lines, but
-were incapable of distinguishing between log lines of the same type, e.g.\
-rejections, and extracted only a small portion of the data required.  The
-only prior published work on the subject of parsing Postfix log files that
-we are aware of is~\cite{log-mail-analyser}, which aims to present
+several parsers were considered and rejected because: they parsed too small
+a fraction of the log files; their parsing was too inexact; and/or they did
+not extract sufficient data.  The effort required to adapt and improve an
+existing parser was judged to be greater than the effort to write one, as
+the techniques used by the existing parsers severely limited their
+potential: some ignored the majority of log lines, parsing specific log
+lines with great accuracy, but without any provision for readily parsing
+new or similar log lines; others sloppily parsed the majority of log lines,
+but were incapable of distinguishing between log lines of the same type,
+e.g.\ rejections, and extracted only a small portion of the data required.
+The only prior published work on the subject of parsing Postfix log files
+that we are aware of is~\cite{log-mail-analyser}, which aims to present
 correlated log data in a form suitable for a systems administrator to
@@
 available.  A full state of the art review is outside the scope of this
-paper but will be included in the thesis resulting from this work.
-The relationship to another procedural framework (ATN) is discussed below.
+paper but will be included in the thesis resulting from this work; the
+relationship to another procedural framework (ATN) is discussed below.
 
@@
     \item [Actions]  Each action performs the work required to deal with a
-        category of log lines, e.g.\ processing data from rejections.
+        category of inputs, e.g.\ processing data from rejections.
 
     \item [Rules]  The rules are responsible for classifying or matching
-        log lines, specifying the action to invoke and the regex which
-        matches the log line and extracts data.
+        inputs, specifying the action to invoke and the regex which matches
+        the inputs and extracts data.
 
@@
 
-For each input log line the framework tries each rule in turn until it
-finds one which matches the input log line, then invokes the action
-specified by that rule.
+For each input the framework tries each rule in turn until it finds one
+which matches the input, then invokes the action specified by that rule.
 
@@
 modifications to the parser source code, significantly lowering the barrier
-to entry for casual users who need to parse new log lines, e.g.\ part-time
+to entry for casual users who need to parse new inputs, e.g.\ part-time
 systems administrators attempting to combat and reduce spam.  Decoupling
@@
 Decoupling also creates a clear separation of functionality: rules handle
-low level details of identifying log lines and extracting data; actions
-handle the higher level details of assembling the required data, dealing
-with the intricacies of the input being parsed, complications arising,
-etc.;\ the framework provides services to actions and manages the parsing
-process.
+low level details of identifying inputs and extracting data; actions handle
+the higher level details of assembling the required data, dealing with the
+intricacies of the input being parsed, complications arising, etc.;\ the
+framework provides services to actions and manages the parsing process.
 
 There is some similarity (shown in \tableref{Similarities with ATN})
-between the architecture and William Wood's 
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the architecture is accidental, but it is interesting how two
-different approaches share a similar division of responsibilities, while
-having different quite semantics.
+between the architecture and William Wood's \ATN{}~\cite{atns, nlpip}, a
+tool used in Computational Linguistics for creating grammars to parse or
+generate sentences.  The resemblance between \ATN{} and the architecture is
+accidental, but it is interesting how two different approaches share a
+similar division of responsibilities, while having different semantics.
 
@@
                       &                 & actions which constitutes a valid input.  \\
-        Transitions   & Actions         & Save data and impose conditions the       \\
+        Transitions   & Actions         & Saves data and imposes conditions the     \\
                       &                 & input must meet to be considered valid.   \\
@@
 shared storage to pass data between actions, loads and validates rules,
-controls parsing of log files, invokes actions, tracks how often each rule
-matches to optimise rule ordering (\sectionref{Rule ordering}), and stores
-results in the database.  The framework is largely independent of the
-parser being written: the same functionality will be required for most or
-all parsers, though some specialised support functions may be provided.
-The framework is the core of the architecture and is deliberately quite
-simple: the rules deal with the variation in log lines, and the actions
-deal with the intricacies and complications encountered when parsing.
+manages parsing, invokes actions, tracks how often each rule matches to
+optimise rule ordering (\sectionref{Rule ordering}), and stores results in
+the database.  The framework is largely independent of the parser being
+written: the same functionality will be required for most or all parsers,
+though some specialised support functions may be provided.  The framework
+is the core of the architecture and is deliberately quite simple: the rules
+deal with the variation in inputs, and the actions deal with the
+intricacies and complications encountered when parsing.
 
-The function which finds the rule matching the log line and invokes the
+The function which finds the rule matching the input and invokes the
 requested action can be expressed in pseudo-code as:
@@
 
-Each action contains the code required to deal with a particular type of
-log line, e.g.\ rejections.  The actions are parser-specific: each parser
+Each action contains the code required to deal with a particular category
+of input, e.g.\ rejections.  The actions are parser-specific: each parser
 author will need to write the required actions from scratch, unless
@@
 the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the
-action which handles Postfix rejecting a mail; it is followed by
-\texttt{SAVE\_DATA}, the action responsible for handling informative
+action which handles Postfix rejecting a delivery attempt; it is followed
+by \texttt{SAVE\_DATA}, the action responsible for handling informative
 inputs: these are log lines that supplement the data gathered from other
@@
 associated with a specific mail, e.g.\ notices about configuration files
-changing.  The framework warns about every unparsed log line, to alert the
+changing.  The framework warns about every unparsed input, to alert the
 user that they need to alter or extend their ruleset; the Postfix log
-parser successfully parses every input line in the \numberOFlogFILESall{}
-log files it is currently being tested with.  The remaining actions only
-have one or two associated rules: in some cases this is because there will
-only ever be one type of log line for that action, e.g.\ all log lines
-showing that a remote client has connected are matched by a single rule and
-handled by the \texttt{CONNECT} action; in other cases the action is
-required to address a deficiency in the log files, or a complication which
-arises during parsing.  Taking the \texttt{CONNECT} action as an example,
-it creates a new data structure for each client connecting, and saves the
-data extracted by the rule in memory, subsequently to be entered into the
-database.  If a data structure already exists for that connection it is
-treated as a symptom of a bug, and the action issues a warning containing
-the full contents of the existing data structure, plus the log line which
-has just been parsed.
+parser successfully parses every log line in the \numberOFlogFILESall{} log
+files it is currently being tested with.  The remaining actions only have
+one or two associated rules: in some cases this is because there will only
+ever be one log line variant for that action, e.g.\ all log lines showing
+that a remote client has connected are matched by a single rule and handled
+by the \texttt{CONNECT} action; in other cases the action is required to
+address a deficiency in the log files, or a complication which arises
+during parsing.  Taking the \texttt{CONNECT} action as an example, it
+creates a new data structure in memory for each client connecting, and
+saving the data extracted by the rule; this data will subsequently be
+entered into the database.  If a data structure already exists for that
+connection it is treated as a symptom of a bug, and the action issues a
+warning containing the full contents of the existing data structure, plus
+the log line which has just been parsed.
 
@@
 new requirements which are discovered during parser development is one of
-the strengths of this architecture: instead of writing one monolithic
+the strengths of this architecture.  Instead of writing one monolithic
 function which must be modified to support new behaviour, with all the
@@
 failures will be less likely and will be easier to test for and diagnose.
-The parser is written in an object oriented style, allowing sub-classes to
-extend or override actions in addition to adding new actions; because each
-action is an independent procedure, the sub-class need only modify the
-action it is overriding, rather than reproducing large chunks of
-functionality.
+The architecture can be implemented in an object oriented style, allowing
+sub-classes to extend or override actions in addition to adding new
+actions; because each action is an independent procedure, the sub-class
+need only modify the action it is overriding, rather than reproducing large
+chunks of functionality.
 
@@
 addition to the obvious variety in log lines, there were a large number of
-complications to be overcome.  Some were the result of deficiencies in
-Postfix's logging (some of which were improved on or resolved by later
-versions of Postfix); others were due to the vagaries of process
-scheduling, client behaviour, and administrative actions.  All were
-successfully accommodated in the Postfix log parser: adding new actions was
-enough to overcome several of the complications; others required
-modifications to a single existing action to work around the difficulties;
-the remainder were resolved by adapting existing actions to cooperate and
-exchange extra data, changing their behaviour as appropriate.
+complications to be overcome (covered in detail in the thesis).  Some were
+the result of deficiencies in Postfix's logging (some of which were
+improved on or resolved by later versions of Postfix); others were due to
+the vagaries of process scheduling, client behaviour, and administrative
+actions.  All were successfully accommodated in the Postfix log parser:
+adding new actions was enough to overcome several of the complications;
+others required modifications to a single existing action to work around
+the difficulties; the remainder were resolved by adapting existing actions
+to cooperate and exchange extra data, changing their behaviour as
+appropriate.
 
-Actions may also return a modified input line which will be parsed as if
-read from the input stream, allowing for a simplified version of cascaded
+Actions may return a modified input line which will be parsed as if read
+from the input stream, allowing for a simplified version of cascaded
 parsing~\cite{cascaded-parsing}.  This is a powerful facility allowing a
@@
 Rules categorise inputs, specifying both the regex to match against each
-log line and the action to invoke when the match is successful.  The
-Postfix log parser stores its rules in the same SQL database the results
-are stored in, removing any doubt as to which set of rules was used to
-produce a set of results; other implementations are free to store the rules
-in whatever fashion suits their needs.  The framework requires each rule to
-have \texttt{action} and \texttt{regex} attributes; each implementation is
-free to add any additional attributes it requires.  The Postfix log parser
-adds several attributes to optimise rule ordering (\sectionref{Rule
-ordering}), to restrict which log lines rules can be matched against, and
-to describe each rule.  Data is automatically extracted and saved from the
-log line based on keywords in the rule's regex, but there is also a
-facility to specify extra data to be saved.
+input and the action to invoke when the match is successful.  The Postfix
+log parser stores its rules in the same SQL database the results are stored
+in, removing any doubt as to which set of rules was used to produce a set
+of results; other implementations are free to store the rules in whatever
+fashion suits their needs.  The framework requires each rule to have
+\texttt{action} and \texttt{regex} attributes; each implementation is free
+to add any additional attributes it requires.  The Postfix log parser adds
+several attributes to optimise rule ordering (\sectionref{Rule ordering}),
+to restrict which log lines each rule can be matched against, and to
+describe each rule.  Data is automatically extracted and saved from the log
+line based on keywords in the rule's regex, but there is also a facility to
+specify extra data to be saved.
 
@@
 existing rules are stored.  The Postfix log parser supplies a utility based
-on \SLCT{}~\cite{slct-paper} to aid in producing regexes from previously
-unparsed inputs.  Storing the rules separately allows other approaches to
-rule management to be used, e.g.\ instead of manually adding new rules,
-machine learning techniques could be used to automatically generate new
-rules, based on the existing rules and input lines.  If this approach was
-taken the choice of machine learning technique would be constrained by the
-size of typical data sets (see \sectionref{Results}); techniques which
-require the full data set when training would be impractical; Instance
-Based Learning~\cite{instance-based-learning} techniques which
-automatically determine which inputs from the training set are valuable and
-which inputs can be discarded would probably reduce the data required to a
-manageable size.  This avenue of research and development has not been
-pursued by the authors, but could easily be undertaken independently.
+on Simple Logfile Clustering Tool~\cite{slct-paper} to aid in producing
+regexes from previously unparsed log lines.  Other approaches to rule
+management can be used, e.g.\ instead of manually adding new rules, machine
+learning techniques could be used to automatically generate new rules,
+based on the existing rules and input lines.  If this approach was taken
+the choice of machine learning technique would be constrained by the size
+of typical data sets (see \sectionref{Results}).  Techniques which require
+the full data set when training would be impractical: Instance Based
+Learning~\cite{instance-based-learning} techniques which automatically
+determine which inputs from the training set are valuable and which inputs
+can be discarded would probably reduce the data required to a manageable
+size.  This avenue of research and development has not been pursued by the
+authors, but could easily be undertaken independently.
 
@@
 the first matching rule wins, and the order in which rules are tried
-against each line may change between log files.  Overlapping rules are
-frequently a requirement, allowing a more specific rule to match some log
-lines and a more general rule to match the majority, e.g.\ separating
+against each input may change between parser invocations.  Overlapping
+rules are frequently a requirement, allowing a more specific rule to match
+some inputs and a more general rule to match the majority, e.g.\ separating
 \SMTP{} delivery to specific sites from \SMTP{} delivery to the rest of the
@@
 equivalent in computational power to \PDA{}, so overlap detection may
-require calculating the intersection of two \PDA{}\@.  \CFL{} are
-equivalent in computational power to \PDA{}, and intersection of two
-\CFL{s} is not closed, i.e.\ intersection can produce a language which
-cannot be parsed by a \CFL{}, e.g.\ $a^{*}b^{n}c^{n} \cap a^{n}b^{n}c^{*}
-\rightarrow a^{n}b^{n}c^{n}$, so the results of \PDA{} intersection may be
-intractable in some cases.  Detecting overlap between $n$ regexes requires
-$n(n-1)/2$ intersections to be calculated, resulting in $O(n^2x)$
-complexity, where $O(x)$ is the complexity of calculating intersection.
-This is certainly not a task to be performed on every parsing run:
-detecting overlap amongst the Postfix log parser's \numberOFrules{} rules
-would require calculating \numberOFruleINTERSECTIONS{} intersections.
+require calculating the intersection of two \PDA{}\@.  The results of
+\PDA{} intersection may be intractable in some cases: the intersection of
+two \CFL{} (equivalent in computational power to \PDA{}) is not closed,
+i.e.\ the resulting language cannot in all cases be parsed by a \CFL{}\@.
+A simple example is:
+$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
+
+Detecting overlap between $n$ regexes requires $n(n-1)/2$
+intersections to be calculated, resulting in $O(n^2x)$ complexity, where
+$O(x)$ is the complexity of calculating intersection.  This is certainly
+not a task to be performed on every parsing run: detecting overlap amongst
+the Postfix log parser's \numberOFrules{} rules would require calculating
+\numberOFruleINTERSECTIONS{} intersections.
 
@@
         %program                   & \daemon{smtpd}                                   \\
-        regex                     & \verb!^connect from (__HOSTNAME__)\[(__IP__)\]$! \\
+        regex                     & \verb!^connect from (__CLIENT_HOSTNAME__)\[(__CLIENT_IP__)\]$! \\
         action                    & CONNECT                                          \\
@@
 
-    \item [Matching rules against log lines is simple:]  The first matching
+    \item [Matching rules against inputs is simple:]  The first matching
         rule determines the action that will be invoked: there is no
@@
         decision, rather a consequence of the line oriented nature of the
-        log files; more flexible approaches could be pursued.
+        Postfix log files; more flexible approaches could be pursued.
 
@@
 Parsing efficiency is an obvious concern when the Postfix log parser
-routinely needs to parse very large log files.  The mail server which
-generated the log files used in testing the Postfix log parser accepts
-approximately 10,000 mails for 700 users a day.  Median log file size is
-50MB, containing 285,000 log lines --- large scale mail servers would have
-much larger log files.  When generating the timing data used in this
-section, \numberOFlogFILES{} log files (totaling 10.08 GB,
-\numberOFlogLINEShuman{} log lines) were each parsed 10 times and the
-parsing times averaged.  Saving results to the database was disabled for
-the test runs, as the tests are aimed at measuring the speed of the Postfix
-log parser rather than the speed of the database.  The computer used for
-test runs was a Dell Optiplex 745 described in \tableref{Computer used to
-generate statistics}; it was dedicated to the task of gathering statistics
-from test runs, and was not used for any other purpose while test runs were
-ongoing --- any services not necessary for running the tests were disabled.
-Parsing all \numberOFlogFILES{} log files in one run took
+routinely needs to parse large log files.  The mail server which generated
+the log files used in testing the Postfix log parser accepts approximately
+10,000 mails for 700 users per day.  Median log file size is 50MB,
+containing 285,000 log lines --- large scale mail servers would have much
+larger log files.  When generating the timing data used in this section,
+\numberOFlogFILES{} log files (totaling 10.08 GB, \numberOFlogLINEShuman{}
+log lines) were each parsed 10 times and the parsing times averaged.
+Saving results to the database was disabled for the test runs, as the tests
+are aimed at measuring the speed of the Postfix log parser rather than the
+speed of the database.  The computer used for test runs was a Dell Optiplex
+745 described in \tableref{Computer used to generate statistics}; it was
+dedicated to the task of gathering statistics from test runs, and was not
+used for any other purpose while test runs were ongoing --- any services
+not necessary for running the tests were disabled.  Parsing all
+\numberOFlogFILES{} log files in one run took
 \input{build/include-timing-run-duration}.
@@
 graph, showing that the algorithm scales linearly.  The ratios increase
-(i.e.\ improve) for log files 22 and 62--68 despite their large size due to
-a mail forwarding loop resulting in a greatly increased number of mails
-delivered and log lines generated.
+(i.e.\ improve) for log files 22 and 62--68 despite their large size; that
+unusually large size is due to a mail forwarding loop resulting in a
+greatly increased number of mails delivered and log lines generated.
 
@@
 \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of number of
-lines and file size to parsing time}{parsing time vs file size vs number
-lines factor}
+log lines and file size to parsing time}{parsing time vs file size vs
+number lines factor}
 
@@
 ruleset.  A second set of statistics was generated using the minimum
-ruleset and the parsing times compared to those generated using the full
+ruleset, and the parsing times compared to those generated using the full
 ruleset: the percentage parsing time increase when using the full ruleset
@@
 results show that the architecture scales very well as the number of rules
-increases, and that sorting the rules is an important optimisation enabling
-this.  \showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage
-parsing time increase of maximum ruleset over minimum ruleset}{Percentage
-parsing time increase of maximum ruleset over minimum ruleset}
+increases, and that optimally ordering the rules is an important
+optimisation enabling this.
+\showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
+time increase of maximum ruleset over minimum ruleset}{Percentage parsing
+time increase of maximum ruleset over minimum ruleset}
 
@@
 
-The Postfix log parser requires two different types of coverage to be
-measured: log lines parsed and mail delivery attempts correctly understood
+The Postfix log parser has two different types of coverage to be measured:
+log lines correctly parsed, and mail delivery attempts correctly understood
 (the former is a requirement for the latter to be achieved).  Improving the
-former is less intrusive, as it just requires new rules to be written;
-improving the latter is more intrusive as it requires changes to actions,
-and it can also be much harder to notice a deficiency.
+former is less intrusive, as usually it just requires new rules to be
+written; improving the latter is more intrusive as it requires changes to
+actions, and it can also be much harder to notice that a deficiency exists.
 
-Measuring Postfix log parser coverage begins with log line coverage.
-Warnings are issued for any log lines which are not parsed; no such
-warnings are issued while parsing the \numberOFlogFILES{} test log files,
-so it can be safely concluded that there are zero false negatives.  False
-positives are harder to quantify: manually verifying that the correct rule
-parsed each of the \numberOFlogLINES{} log lines is infeasible, making it
-impossible to quantify the false positive rate.  A random sample of 6039
-log lines (0.00994\% of the total) was parsed and the results manually
-verified by inspection to ensure that the correct rule parsed each log
-line.  The sample contained zero false positives, and this check has been
+Correct parsing of log lines must be measured first.  Warnings are issued
+for any log lines which are not parsed; no such warnings are issued while
+parsing the \numberOFlogFILES{} log files, so it can be safely concluded
+that there are zero false negatives.  False positives are harder to
+quantify: manually verifying that the correct rule parsed each of the
+\numberOFlogLINES{} log lines is infeasible, making it impossible to
+quantify the false positive rate.  A random sample of 6039 log lines
+(0.00994\% of the total) was parsed and the results manually verified by
+inspection to ensure that the correct rule parsed each log line.  The
+sample results contained zero false positives, and this check has been
 automated to ensure continued accuracy.
@@
 The proportion of mail delivery attempts correctly understood is much more
-difficult to determine accurately than coverage of log lines.  The
+difficult to determine accurately than log lines correctly parsed.  The
 implementation can dump its state tables in a human readable form;
-examining these tables with reference to the log files is the best way to
-detect mails which were not handled properly.  The implementation issues
-warnings when it detects any errors, which should alert the user to a
-problem.  There should be few or no warnings when parsing, and when
-finished parsing the state table should only contain entries for mails
-which had yet to be delivered when the log files ended, or were accepted
-before the log files began.  A second sample of 6000 log lines was parsed
-with all debugging options enabled, resulting in 167,448 lines of output.
-All 167,448 lines were examined in conjunction with the input log file and
-a dump of the resulting database, verifying that for each of the input
-lines the implementation used the correct rule and executed the correct
-action, which in turn produced the correct result and inserted the correct
-data in the database.  The log segment produced 4 warnings (about mails
-which had started before the log file segment), 10 mails remaining in the
-state tables, and 1625 connections correctly entered in the database.  No
-error messages were produced, therefore there were no false negatives.
-Given the evidence detailed above, the authors are confident that the false
-positive rate when reconstructing a mail is zero.
+examining these tables with reference to the log files and database is the
+best way to detect mails which were not handled properly.  The
+implementation issues warnings when it detects any errors, which should
+alert the user to a problem.  There should be few or no warnings during
+parsing, and when parsing is finished the state table should only contain
+entries for mails starting before or ending after the log file.  A second
+sample of 6000 log lines was parsed with all debugging options enabled,
+resulting in 167,448 lines of output.  All 167,448 lines were examined in
+conjunction with the input log file and a dump of the resulting database,
+verifying that for each of the log lines the implementation performed
+correctly.  The implementation produced 4 warnings about deficiencies in
+the log segment, 10 mails remaining in the state tables, and 1625 correct
+entries in the database.  No error messages were produced, therefore there
+were no false negatives.  Given the evidence detailed above, the authors
+are confident that the false positive rate when understanding a mail
+delivery attempt is zero.
 
@@
 
-The architecture's greatest strength is the ease with which it can be
+This architecture's greatest strength is the ease with which it can be
 adapted to deal with new requirements and inputs.  Parsing a variation of
 an existing input is a trivial task: simply modify an existing rule or add
-a new one with an appropriate regex and the task is complete; knowledge of
-the framework is not required.  Parsing a new category
-of input is achieved by writing a new action and appropriate rules; quite
-often the action will not need to interact directly with existing actions,
-but when interaction is required the framework provides the necessary
-facilities.  The decoupling of rules from actions allows different sets of
-rules to be used with the same actions, e.g.\ a parser might have actions
-to process versions one and two of a file format; the parser will then
-parse either version one, version two, or both versions, depending on the
-ruleset selected.  The architecture makes it possible to apply commonly
-used programming techniques such as object orientation, inheritance,
-composition, delegation, and roles when designing and implementing a parser,
-simplifying the process of working within a team with shared
-responsibility for a parser, or when developing and testing additional functionality.
-This architecture is ideally suited to parsing inputs where the input is
-not fully understood or does not follow a fixed grammar: the architecture
-warns about unparsed inputs and other errors, but continues on parsing as
-best it can, allowing the developer of a new parser to decide which
-deficiencies are most important and require attention first, rather than
-being forced to fix the first error which arises.
+a new rule with an appropriate regex and the task is complete; knowledge of
+the framework is not required.  Parsing a new category of input is achieved
+by writing a new action and appropriate rules; quite often the action will
+not need to interact with existing actions, but when interaction is
+required the framework provides the necessary facilities.  The decoupling
+of rules from actions allows different sets of rules to be used with the
+same actions, e.g.\ a parser might have actions to process versions one and
+two of a file format; the parser will then parse version one, version two,
+or both versions, depending on the ruleset selected.  The architecture
+makes it possible to apply commonly used programming techniques such as
+object orientation, inheritance, composition, delegation, and roles when
+designing and implementing a parser, simplifying the process of working
+within a team with shared responsibility for a parser, or when developing
+and testing additional functionality.  This architecture is ideally suited
+to parsing inputs where the input is not fully understood or does not
+follow a fixed grammar: the architecture warns about unparsed inputs and
+other errors, but continues on parsing as best it can, allowing the
+developer of a new parser to decide which deficiencies are most important
+and require attention first, rather than being forced to fix the first
+error which arises.
 

### 2008-08-22 doc/sgai-2008.tex
Corrections and improvements from proofreading.

@@
 \newacronym{ATN}{Augmented Transition Networks}{}
+\newacronym{DNSBL}{DNS Black List}{}
 
@@
 content-based~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}
-which requires accepting the mail in the first place, but rejecting mail
-during the delivery attempt is preferable: senders of non-spam mail that is
-mistakenly rejected will receive an immediate non-delivery notice; load is
-reduced on the accepting mail server (allowing more intensive content-based
-techniques to be used); users have less mail to wade through.  Improving
-the performance of anti-spam techniques which are applied when mail is
-being transferred via \SMTP{}\footnotemark{} is the goal of this project,
-by providing a platform for reasoning about anti-spam filters.  The
-approach chosen to measure performance is to analyse the log files produced
-by the \SMTP{} server in use, Postfix~\cite{postfix}, rather than modifying
-it to generate statistics: this approach improves the chances of other
-sites testing and utilising the software.  The need arose for a parser
-capable of dealing with the great number and variety of log lines produced
-by Postfix: the parser must be designed so that adding support for parsing
-new inputs is a simple task, as the log lines to be parsed will change over
-time.  The variety in log lines occurs for several
-reasons:\footnotetext{Simple Mail Transfer Protocol transfers mail across
-the Internet from the sender to one or more recipients.  It is a simple,
-human readable, plain text protocol, making it quite easy to test and debug
-problems with it.  The original protocol definition is
-RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}
+which requires accepting the mail in the first place to determine if it is
+spam, but rejecting mail during the delivery attempt is preferable: senders
+of non-spam mail that is mistakenly rejected will receive an immediate
+non-delivery notice; load is reduced on the accepting mail server (allowing
+more intensive content-based techniques to be used); users have less mail
+to wade through.  Improving the performance of anti-spam techniques which
+are applied when mail is being transferred via \SMTP{}\footnotemark{} is
+the goal of this project, by providing a platform for reasoning about
+anti-spam filters.  The approach chosen to measure performance is to
+analyse the log files produced by the \SMTP{} server in use,
+Postfix~\cite{postfix}, rather than modifying it to generate statistics:
+this approach improves the chances of other sites testing and utilising the
+software.  The need arose for a parser capable of dealing with the great
+number and variety of log lines produced by Postfix: the parser must be
+designed so that adding support for parsing new inputs is a simple task, as
+the log lines to be parsed will change over time.  The variety in log lines
+occurs for several reasons:\footnotetext{Simple Mail Transfer Protocol
+transfers mail across the Internet from the sender to one or more
+recipients.  It is a simple, human readable, plain text protocol, making it
+quite easy to test and debug problems with it.  The original protocol
+definition is RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}
 
@@
 
-    \item External resources utilised by Postfix (e.g.\ policy
-        servers~\cite{policy-servers} or DNSBLs) can also change their
-        messages, sometimes without warning.
+    \item External resources utilised by Postfix (e.g.\ \DNSBL{} or policy
+        servers~\cite{policy-servers}) can also change their messages,
+        sometimes without warning.
 
@@
 of a specific type and only that type (e.g.\ rejection by a specific
-DNSBL).  It is an accepted standard to separate the parsing procedure from
-the declarative grammar it operates with; part of the novelty here is in
-the way that the grammar is itself partially procedural (each action is a
-separate procedure).  This architecture is ideally suited to parsing inputs
-where the input is not fully understood or does not conform to a fixed
-grammar: the architecture warns about unparsed inputs and other errors, but
-continues on parsing as best it can, allowing the developer of a new parser
-to decide which deficiencies are most important and require attention
-first, rather than being forced to fix the first error which arises.
+\DNSBL{}).  It is an accepted standard to separate the parsing procedure
+from the declarative grammar it operates with; part of the novelty here is
+in the way that the grammar is itself partially procedural (each action is
+a separate procedure).  This architecture is ideally suited to parsing
+inputs where the input is not fully understood or does not conform to a
+fixed grammar: the architecture warns about unparsed inputs and other
+errors, but continues on parsing as best it can, allowing the developer of
+a new parser to decide which deficiencies are most important and require
+attention first, rather than being forced to fix the first error which
+arises.
 
@@
 
-For each input the framework tries each rule in turn until it finds one
+For each input the framework tries each rule in turn until it finds a rule
 which matches the input, then invokes the action specified by that rule.
@@
         \tabletopline{}%
-        ATN           & Architecture    & Similarity                                \\
+        ATN           & Parser Architecture & Similarity                                \\
         \tablemiddleline{}%
@@
 \begin{verbatim}
-for each line in the input files: 
+for each input:
     for each rule defined by the user: 
-        if this rule matches the input line:
+        if this rule matches the input:
             perform the action specified by the rule
             skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
+            process the next input
+    warn the user that the input was not parsed
 \end{verbatim}
@@
 author will need to write the required actions from scratch, unless
-extending an existing parser.  In the Postfix log parser developed for this
+extending an existing parser.  It is expected that parsers utilising this
+architecture will have a high ratio of rules to actions, hopefully
+resulting in simpler rules and clearer distinctions between the inputs
+parsed by different rules.  In the Postfix log parser developed for this
 project there are \numberOFactions{} actions and \numberOFrules{} rules,
 with a very uneven distribution of rules to actions as shown in
-\graphref{Distribution of rules per action}.  It is expected that parsers
-utilising this architecture will have a high ratio of rules to actions,
-hopefully resulting in simpler rules and clearer distinctions between the
-inputs parsed by different rules.
+\graphref{Distribution of rules per action}.
 \showgraph{build/graph-action-distribution}{Distribution of rules per
@@
 the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the
-action which handles Postfix rejecting a delivery attempt; it is followed
-by \texttt{SAVE\_DATA}, the action responsible for handling informative
-inputs: these are log lines that supplement the data gathered from other
-log lines, but that do not by themselves represent interesting events.  The
-third most common action is, perhaps surprisingly, \texttt{UNINTERESTING}:
-this action does nothing when executed, allowing uninteresting log lines to
-be parsed without causing any effects (it does not imply that the input is
+action which handles Postfix rejecting a mail delivery attempt; it is
+followed by \texttt{SAVE\_DATA}, the action responsible for handling
+informative log lines, which supplement the data gathered from other log
+lines but do not by themselves represent interesting events.  The third
+most common action is, perhaps surprisingly, \texttt{UNINTERESTING}: this
+action does nothing when executed, allowing uninteresting log lines to be
+parsed without causing any effects (it does not imply that the input is
 ungrammatical or unparsed).  Generally rules specifying the
@@
 associated with a specific mail, e.g.\ notices about configuration files
-changing.  The framework warns about every unparsed input, to alert the
-user that they need to alter or extend their ruleset; the Postfix log
-parser successfully parses every log line in the \numberOFlogFILESall{} log
-files it is currently being tested with.  The remaining actions only have
-one or two associated rules: in some cases this is because there will only
-ever be one log line variant for that action, e.g.\ all log lines showing
-that a remote client has connected are matched by a single rule and handled
-by the \texttt{CONNECT} action; in other cases the action is required to
-address a deficiency in the log files, or a complication which arises
-during parsing.  Taking the \texttt{CONNECT} action as an example, it
-creates a new data structure in memory for each client connecting, and
-saving the data extracted by the rule; this data will subsequently be
-entered into the database.  If a data structure already exists for that
-connection it is treated as a symptom of a bug, and the action issues a
-warning containing the full contents of the existing data structure, plus
-the log line which has just been parsed.
+changing.  The remaining actions only have one or two associated rules: in
+some cases this is because there will only ever be one log line variant for
+that action, e.g.\ all log lines showing that a remote client has connected
+are matched by a single rule and handled by the \texttt{CONNECT} action; in
+other cases the action is required to address a deficiency in the log
+files, or a complication which arises during parsing.  Taking the
+\texttt{CONNECT} action as an example, it creates a new data structure in
+memory for each client connecting, and saves the data extracted by the
+rule; this data will subsequently be entered into the database.  If a data
+structure already exists for that connection it is treated as a symptom of
+a bug, and the action issues a warning containing the full contents of the
+existing data structure, plus the log line which has just been parsed.
 
@@
 addition to the obvious variety in log lines, there were a large number of
-complications to be overcome (covered in detail in the thesis).  Some were
-the result of deficiencies in Postfix's logging (some of which were
-improved on or resolved by later versions of Postfix); others were due to
-the vagaries of process scheduling, client behaviour, and administrative
-actions.  All were successfully accommodated in the Postfix log parser:
-adding new actions was enough to overcome several of the complications;
-others required modifications to a single existing action to work around
-the difficulties; the remainder were resolved by adapting existing actions
-to cooperate and exchange extra data, changing their behaviour as
-appropriate.
+complications to be overcome.  Some were the result of deficiencies in
+Postfix's logging (some of which were improved on or resolved by later
+versions of Postfix); others were due to the vagaries of process
+scheduling, client behaviour, and administrative actions.  All were
+successfully accommodated in the Postfix log parser: adding new actions was
+enough to overcome several of the complications; others required
+modifications to a single existing action to work around the difficulties;
+the remainder were resolved by adapting existing actions to cooperate and
+exchange extra data, changing their behaviour as appropriate.
 
@@
 parsing~\cite{cascaded-parsing}.  This is a powerful facility allowing a
-number of rules and actions to parse a single input line.
+number of rules and actions to parse a single input.
 
@@
 of results; other implementations are free to store the rules in whatever
-fashion suits their needs.  The framework requires each rule to have
-\texttt{action} and \texttt{regex} attributes; each implementation is free
-to add any additional attributes it requires.  The Postfix log parser adds
-several attributes to optimise rule ordering (\sectionref{Rule ordering}),
-to restrict which log lines each rule can be matched against, and to
-describe each rule.  Data is automatically extracted and saved from the log
-line based on keywords in the rule's regex, but there is also a facility to
-specify extra data to be saved.
+fashion suits their needs.  The framework warns about every unparsed input,
+to alert the user that they need to alter or extend their ruleset; the
+Postfix log parser successfully parses every log line in the
+\numberOFlogFILESall{} log files it is currently being tested with.  The
+framework requires each rule to have \texttt{action} and \texttt{regex}
+attributes; each implementation is free to add any additional attributes it
+requires.  The Postfix log parser adds several attributes to optimise rule
+ordering (\sectionref{Rule ordering}), to restrict which log lines each
+rule can be matched against, and to describe each rule.  Data is
+automatically extracted and saved from the log line based on keywords in
+the rule's regex, but there is also a facility to specify extra data to be
+saved.
 
@@
 rules are frequently a requirement, allowing a more specific rule to match
-some inputs and a more general rule to match the majority, e.g.\ separating
-\SMTP{} delivery to specific sites from \SMTP{} delivery to the rest of the
-world.  This can simplify both the general rule and the more specific rule;
-additionally rules from different sources can be combined with a minimum of
-prior cooperation or modification required.  Overlapping rules should have
-a priority attribute to specify their relative ordering; negative
-priorities may be useful for catchall rules.
+some inputs and a more general rule to match the remainder, e.g.\
+separating \SMTP{} delivery to specific sites from \SMTP{} delivery to the
+rest of the world.  This can simplify both the general rule and the more
+specific rule; additionally rules from different sources can be combined
+with a minimum of prior cooperation or modification required.  Overlapping
+rules should have a priority attribute to specify their relative ordering;
+negative priorities may be useful for catchall rules.
 
@@
 written to detect overlapping rules.  Traditional regexes are equivalent in
-computational power to \FA{}; it is possible to convert between regexes and
-\FA{}, so regex overlap could be detected by finding a non-empty
-intersection of two \FA{}\@.  One formula for \FA{} intersection is given
-in for example~\cite{intersection-of-NFA-using-Z}: $FA1 \cap{} FA2 =
+computational power to \FA{}, and it is possible to convert between the
+two, so regex overlap could be detected by finding a non-empty intersection
+of two \FA{}\@.  The standard equation for \FA{} intersection (given for
+example in~\cite{intersection-of-NFA-using-Z}) is: $FA1 \cap{} FA2 =
 \overline{(\overline{FA1} \cup{} \overline{FA2})}$.  Perl 5.10 regexes are
@@
 Perl 5.10 regexes can maintain an arbitrary state stack and are thus
-equivalent in computational power to \PDA{}, so overlap detection may
-require calculating the intersection of two \PDA{}\@.  The results of
-\PDA{} intersection may be intractable in some cases: the intersection of
-two \CFL{} (equivalent in computational power to \PDA{}) is not closed,
-i.e.\ the resulting language cannot in all cases be parsed by a \CFL{}\@.
-A simple example is:
+equivalent in computational power to \PDA{} or \CFL{}, so detecting overlap
+may require calculating the intersection of two \PDA{} or \CFL{}\@.  The
+intersection of two \CFL{} is not closed, i.e.\ the resulting language
+cannot in all cases be parsed by a \CFL{}, so intersection may be
+intractable in some cases e.g.:
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
 
-Detecting overlap between $n$ regexes requires $n(n-1)/2$
-intersections to be calculated, resulting in $O(n^2x)$ complexity, where
-$O(x)$ is the complexity of calculating intersection.  This is certainly
-not a task to be performed on every parsing run: detecting overlap amongst
-the Postfix log parser's \numberOFrules{} rules would require calculating
+Detecting overlap between $n$ regexes requires $n(n-1)/2$ intersections to
+be calculated, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
+complexity of calculating intersection.  This is certainly not a task to be
+performed on every parsing run: detecting overlap amongst the Postfix log
+parser's \numberOFrules{} rules would require calculating
 \numberOFruleINTERSECTIONS{} intersections.
@@
 It is possible to define pathological regexes, which fall into two main
-categories: regexes which match every input line, and regexes which consume
+categories: regexes which match every input, and regexes which consume
 excessive amounts of CPU time during matching.  Defining a regex which
-matches every input line is trivial: \verb!/^/! matches the start of every
-input line.  In general a regex will not consume much CPU time when the
-match is successful; usually CPU time is consumed by a regex with a lot of
-alteration and variable quantifiers failing to match.  This topic is beyond
+matches every input is trivial: \verb!/^/! matches the start of every
+input.  In general a regex will not consume much CPU time when the match is
+successful; usually CPU time is consumed when a regex with a lot of
+alteration and variable quantifiers fails to match.  This topic is beyond
 the scope of this paper, see~\cite{mastering-regular-expressions} for
@@
         Actions can consult the results (or lack of results) of previous
-        actions when executing, providing some context sensitivity.
-        Actions can not inspect past or future inputs, however actions
-        can return a modified input line that will be parsed as if read
-        from the input stream, allowing for a simplified version of
-        cascaded parsing~\cite{cascaded-parsing}.
+        actions during execution, providing some context sensitivity.
+        Actions can not inspect past or future inputs, however actions can
+        return a modified input that will be parsed as if read from the
+        input stream, allowing for a simplified version of cascaded
+        parsing~\cite{cascaded-parsing}.
 
@@
 are aimed at measuring the speed of the Postfix log parser rather than the
-speed of the database.  The computer used for test runs was a Dell Optiplex
+speed of the database.  The computer used for test runs is a Dell Optiplex
 745 described in \tableref{Computer used to generate statistics}; it was
@@
 input size: linearly, polynomially, or exponentially?  \Graphref{Parsing
-time, file size, and number of lines} shows the parsing time in seconds,
-file size in MB, and number of log lines in tens of thousands, for each of
-the \numberOFlogFILES{} log files.  All three lines run roughly in
-parallel, giving the impression that the algorithm scales linearly with
-input size.  This impression is borne out by \graphref{parsing time vs file
-size vs number lines factor}: the ratios are tightly banded across the
-graph, showing that the algorithm scales linearly.  The ratios increase
-(i.e.\ improve) for log files 22 and 62--68 despite their large size; that
-unusually large size is due to a mail forwarding loop resulting in a
-greatly increased number of mails delivered and log lines generated.
+time, file size, and number of log lines} shows the parsing time in
+seconds, file size in MB, and number of log lines in tens of thousands, for
+each of the \numberOFlogFILES{} log files.  All three lines on the graph
+run roughly in parallel, giving the impression that the algorithm scales
+linearly with input size.  This impression is borne out by
+\graphref{parsing time vs file size vs number lines factor}: the ratios are
+tightly banded across the graph, showing that the algorithm scales
+linearly.  The ratios increase (i.e.\ improve) for log files 22 and 62--68
+despite their large size; that unusually large size is due to a mail
+forwarding loop resulting in a greatly increased number of mails delivered
+and log lines generated.
 
 \showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
-and number of lines}{Parsing time, file size, and number of lines}
+and number of log lines}{Parsing time, file size, and number of log lines}
 \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of number of
@@
 At the time of writing the Postfix log parser has \numberOFrules{}
-different rules: the top 10\% match the vast majority of log lines, with
-the remaining log lines split across the other 90\% of the rules.  The
-distribution resembles a Power Law distribution, as shown in \graphref{rule
-hits graph}.  Assuming that the distribution of log lines is reasonably
-consistent over time, parser efficiency should benefit from trying more
-frequently matching rules before those which match less frequently.  To
-test this hypothesis three full test runs were performed with different
-rule orderings:
+different rules: as shown in \graphref{rule hits graph} 10\% of the rules
+match the vast majority of log lines, with the remaining log lines split
+across the other 90\% of the rules, similar to a Power Law distribution.
+Assuming that the distribution of log lines is reasonably consistent over
+time, parser efficiency should benefit from trying more frequently matching
+rules before those which match less frequently.  To test this hypothesis
+three full test runs were performed with different rule orderings:
 
-\begin{description}
+\begin{description}[optimal]
 
@@
 worthwhile reduction in parsing time of approximately 10\%.  Optimal rule
-ordering is also beneficial in other ways, as shown in
+ordering is also beneficial in other ways, as described in
 \sectionref{Architecture Scalability: Number of Rules}.
-\showgraph{build/graph-hits}{Hits per rule}{rule hits graph}
-\showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
-relative to shuffled ordering}{Parsing time relative to shuffled}
+\showgraph{build/graph-hits}{Number of log lines matched by each rule}{rule
+hits graph} \showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing
+time relative to shuffled ordering}{Parsing time relative to shuffled}
 
@@
 How any architecture scales as the number of rules increases is important,
-but it is particularly important in this architecture because it is
+but it is particularly important for this architecture because it is
 expected that typical parsers will have a large number of rules.  There are
@@
 quantify: manually verifying that the correct rule parsed each of the
-\numberOFlogLINES{} log lines is infeasible, making it impossible to
-quantify the false positive rate.  A random sample of 6039 log lines
-(0.00994\% of the total) was parsed and the results manually verified by
-inspection to ensure that the correct rule parsed each log line.  The
-sample results contained zero false positives, and this check has been
-automated to ensure continued accuracy.
+\numberOFlogLINES{} log lines is infeasible.  A random sample of 6,039 log
+lines (0.00994\% of \numberOFlogLINES{}) was parsed and the results
+manually verified by inspection to ensure that the correct rule parsed each
+log line.  The sample results contained zero false positives, and this
+check has been automated to ensure continued accuracy.  The authors are
+confident that zero false positives occur when parsing the
+\numberOFlogFILES{} log files.
 
 The proportion of mail delivery attempts correctly understood is much more
-difficult to determine accurately than log lines correctly parsed.  The
-implementation can dump its state tables in a human readable form;
-examining these tables with reference to the log files and database is the
-best way to detect mails which were not handled properly.  The
-implementation issues warnings when it detects any errors, which should
-alert the user to a problem.  There should be few or no warnings during
-parsing, and when parsing is finished the state table should only contain
-entries for mails starting before or ending after the log file.  A second
-sample of 6000 log lines was parsed with all debugging options enabled,
-resulting in 167,448 lines of output.  All 167,448 lines were examined in
-conjunction with the input log file and a dump of the resulting database,
-verifying that for each of the log lines the implementation performed
-correctly.  The implementation produced 4 warnings about deficiencies in
-the log segment, 10 mails remaining in the state tables, and 1625 correct
-entries in the database.  No error messages were produced, therefore there
-were no false negatives.  Given the evidence detailed above, the authors
-are confident that the false positive rate when understanding a mail
-delivery attempt is zero.
+difficult to determine accurately than the proportion of log lines
+correctly parsed.  The implementation can dump its state tables in a human
+readable form; examining these tables with reference to the log files and
+database is the best way to detect mails which were not handled properly.
+The implementation issues warnings when it detects any errors or
+discrepancies, which should alert the user to a problem.  There should be
+few or no warnings during parsing, and when parsing is finished the state
+table should only contain entries for mails starting before or ending after
+the log file.  A second sample of 6000 log lines was parsed with all
+debugging options enabled, resulting in 167,448 lines of output.  All
+167,448 lines were examined in conjunction with the log segment and a dump
+of the resulting database, verifying that for each of the log lines the
+implementation performed correctly.  The implementation produced 4 warnings
+about deficiencies in the log segment, 10 mails correctly remaining in the
+state tables, and 1625 correct entries in the database.  No error messages
+were produced, therefore there were no false negatives.  Given the evidence
+detailed above, the authors are confident that the false positive rate when
+understanding a mail delivery attempt is zero.
 
@@
 an existing input is a trivial task: simply modify an existing rule or add
-a new rule with an appropriate regex and the task is complete; knowledge of
-the framework is not required.  Parsing a new category of input is achieved
-by writing a new action and appropriate rules; quite often the action will
-not need to interact with existing actions, but when interaction is
-required the framework provides the necessary facilities.  The decoupling
-of rules from actions allows different sets of rules to be used with the
-same actions, e.g.\ a parser might have actions to process versions one and
-two of a file format; the parser will then parse version one, version two,
-or both versions, depending on the ruleset selected.  The architecture
-makes it possible to apply commonly used programming techniques such as
-object orientation, inheritance, composition, delegation, and roles when
-designing and implementing a parser, simplifying the process of working
-within a team with shared responsibility for a parser, or when developing
-and testing additional functionality.  This architecture is ideally suited
-to parsing inputs where the input is not fully understood or does not
-follow a fixed grammar: the architecture warns about unparsed inputs and
-other errors, but continues on parsing as best it can, allowing the
-developer of a new parser to decide which deficiencies are most important
-and require attention first, rather than being forced to fix the first
-error which arises.
+a new rule with an appropriate regex and the task is complete.  Parsing a
+new category of input is achieved by writing a new action and appropriate
+rules; quite often the new action will not need to interact with existing
+actions, but when interaction is required the framework provides the
+necessary facilities.  The decoupling of rules from actions allows
+different sets of rules to be used with the same actions, e.g.\ a parser
+might have actions to process versions one and two of a file format; the
+parser will then parse version one, or version two, or both versions,
+depending on the ruleset selected.  The architecture makes it possible to
+apply commonly used programming techniques such as object orientation,
+inheritance, composition, delegation, and roles when designing and
+implementing a parser, simplifying the process of working within a team
+with shared responsibility for a parser, or when developing and testing
+additional functionality.  This architecture is ideally suited to parsing
+inputs where the input is not fully understood or does not follow a fixed
+grammar: the architecture warns about unparsed inputs and other errors, but
+continues on parsing as best it can, allowing the developer of a new parser
+to decide which deficiencies are most important and require attention
+first, rather than being forced to fix the first error which arises.
 
@@
 this approach could be applied to optimising the behaviour of any system
-given sufficient log messages to analyse.  An alternative approach to black box
-optimisation which uses application profiling in conjunction with the
+given sufficient log messages to analyse.  An alternative approach to black
+box optimisation which uses application profiling in conjunction with the
 application's error messages to improve the error messages shown to users

### 2008-08-22 doc/logparser-implementation.tex
Renamed MAIL_PICKED_FOR_DELIVERY to MAIL_QUEUED in thesis. Improved bibliography based on chktex suggestions.

@@
         \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_PICKED\_FOR\_DELIVERY\@; flowchart:~3), in
-        which case a new data structure will be created for that mail.  The
-        mail will be flagged as having unknown origins; this flag should be
-        subsequently cleared once the origin has been established.  This
-        may also be an indicator that the mail is a bounce notification,
-        see \sectionref{identifying-bounce-notifications} for details.
+        delivery (action: MAIL\_QUEUED\@; flowchart:~3), in which case a
+        new data structure will be created for that mail.  The mail will be
+        flagged as having unknown origins; this flag should be subsequently
+        cleared once the origin has been established.  This may also be an
+        indicator that the mail is a bounce notification, see
+        \sectionref{identifying-bounce-notifications} for details.
 
@@
 
-    \item [MAIL\_PICKED\_FOR\_DELIVERY] This action represents Postfix
+    \item [MAIL\_QUEUED] This action represents Postfix
         picking a mail from the queue to deliver. This action is used for
@@
 \daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_PICKED\_FOR\_DELIVERY action),
+(during the MAIL\_QUEUED action),
 and the line will be deemed part of the discarded mail and discarded if it

### 2008-08-22 doc/sgai-2008.tex
Changes suggested by the diction program.

@@
     novel parsing architecture, designed for parsing Postfix log files,
-    which aims to make the process of parsing new inputs as simple as
+    that aims to make the process of parsing new inputs as simple as
     possible, enabling users to trivially add new rules (to parse variants
@@
 more intensive content-based techniques to be used); users have less mail
-to wade through.  Improving the performance of anti-spam techniques which
+to wade through.  Improving the performance of anti-spam techniques that
 are applied when mail is being transferred via \SMTP{}\footnotemark{} is
@@
 a new parser to decide which deficiencies are most important and require
-attention first, rather than being forced to fix the first error which
+attention first, rather than being forced to fix the first error that
 arises.
@@
 
-    \item [Framework]  The framework is the structure which actions and
-        rules plug into.  It provides the parsing loop, loading and
-        validation of rules, shared data storage, storage of results, and
+    \item [Framework]  The framework is the structure that actions and
+        rules plug into.  It provides the parsing loop, shared data
+        storage, loading and validation of rules, storage of results, and
         other support functions.
@@
     \item [Rules]  The rules are responsible for classifying or matching
-        inputs, specifying the action to invoke and the regex which matches
+        inputs, specifying the action to invoke and the regex that matches
         the inputs and extracts data.
@@
 For each input the framework tries each rule in turn until it finds a rule
-which matches the input, then invokes the action specified by that rule.
+that matches the input, then invokes the action specified by that rule.
 
@@
 
-There is some similarity (shown in \tableref{Similarities with ATN})
-between the architecture and William Wood's \ATN{}~\cite{atns, nlpip}, a
-tool used in Computational Linguistics for creating grammars to parse or
-generate sentences.  The resemblance between \ATN{} and the architecture is
-accidental, but it is interesting how two different approaches share a
-similar division of responsibilities, while having different semantics.
+Some similarity exists between the architecture and William Wood's
+\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
+creating grammars to parse or generate sentences.  The resemblance between
+\ATN{} and the architecture (shown in \tableref{Similarities with ATN}) is
+accidental, but it is obvious that the two approaches share a similar
+division of responsibilities, although having different semantics.
 
@@
         Networks      & Framework           & Determines the sequence of transitions or \\
-                      &                     & actions which constitutes a valid input.  \\
+                      &                     & actions that constitutes a valid input.   \\
         Transitions   & Actions             & Saves data and imposes conditions the     \\
-                      &                     & input must meet to be considered valid.   \\
+                      &                     & input must meet to be accepted as valid.  \\
         Abbreviations & Rules               & Responsible for classifying input.        \\
@@
 The framework takes care of miscellaneous support functions and low level
-details of parsing, freeing the authors of actions to concentrate on
-writing productive code.  It provides the link between actions and rules,
-allowing either to be improved independently of the other.  It provides
-shared storage to pass data between actions, loads and validates rules,
-manages parsing, invokes actions, tracks how often each rule matches to
-optimise rule ordering (\sectionref{Rule ordering}), and stores results in
-the database.  The framework is largely independent of the parser being
+details of parsing, freeing the programmers writing actions to concentrate
+on writing productive code.  It provides the link between actions and
+rules, allowing either to be improved independently of the other.  It
+provides shared storage to pass data between actions, loads and validates
+rules, manages parsing, invokes actions, tracks how often each rule matches
+to optimise rule ordering (\sectionref{Rule ordering}), and stores results
+in the database.  The framework is largely independent of the parser being
 written: the same functionality will be required for most or all parsers,
@@
 
-The function which finds the rule matching the input and invokes the
+The function that finds the rule matching the input and invokes the
 requested action can be expressed in pseudo-code as:
@@
 author will need to write the required actions from scratch, unless
-extending an existing parser.  It is expected that parsers utilising this
-architecture will have a high ratio of rules to actions, hopefully
-resulting in simpler rules and clearer distinctions between the inputs
-parsed by different rules.  In the Postfix log parser developed for this
-project there are \numberOFactions{} actions and \numberOFrules{} rules,
-with a very uneven distribution of rules to actions as shown in
-\graphref{Distribution of rules per action}.
+extending an existing parser.  It is anticipated that parsers utilising
+this architecture will have a high ratio of rules to actions, with the aim
+of having simpler rules and clearer distinctions between the inputs parsed
+by different rules.  In the Postfix log parser developed for this project
+there are \numberOFactions{} actions and \numberOFrules{} rules, with an
+uneven distribution of rules to actions as shown in \graphref{Distribution
+of rules per action}.
 \showgraph{build/graph-action-distribution}{Distribution of rules per
 action}{Distribution of rules per action} Unsurprisingly the action with
-the highest number of associated rules is \texttt{DELIVERY\_REJECTED}, the
-action which handles Postfix rejecting a mail delivery attempt; it is
-followed by \texttt{SAVE\_DATA}, the action responsible for handling
-informative log lines, which supplement the data gathered from other log
-lines but do not by themselves represent interesting events.  The third
-most common action is, perhaps surprisingly, \texttt{UNINTERESTING}: this
-action does nothing when executed, allowing uninteresting log lines to be
-parsed without causing any effects (it does not imply that the input is
-ungrammatical or unparsed).  Generally rules specifying the
-\texttt{UNINTERESTING} action are rules to parse log lines which are not
-associated with a specific mail, e.g.\ notices about configuration files
-changing.  The remaining actions only have one or two associated rules: in
-some cases this is because there will only ever be one log line variant for
-that action, e.g.\ all log lines showing that a remote client has connected
-are matched by a single rule and handled by the \texttt{CONNECT} action; in
-other cases the action is required to address a deficiency in the log
-files, or a complication which arises during parsing.  Taking the
-\texttt{CONNECT} action as an example, it creates a new data structure in
-memory for each client connecting, and saves the data extracted by the
-rule; this data will subsequently be entered into the database.  If a data
-structure already exists for that connection it is treated as a symptom of
-a bug, and the action issues a warning containing the full contents of the
-existing data structure, plus the log line which has just been parsed.
+the most associated rules is \texttt{DELIVERY\_REJECTED}, the action that
+handles Postfix rejecting a mail delivery attempt; it is followed by
+\texttt{SAVE\_DATA}, the action responsible for handling informative log
+lines, that supplement the data gathered from other log lines but do not by
+themselves represent events.  The third most common action is, perhaps
+surprisingly, \texttt{UNINTERESTING}: this action does nothing when
+executed, allowing uninteresting log lines to be parsed without causing any
+effects (it does not imply that the input is ungrammatical or unparsed).
+Generally rules specifying the \texttt{UNINTERESTING} action are rules to
+parse log lines that are not associated with a specific mail, e.g.\
+notices about configuration files changing.  The remaining actions only
+have one or two associated rules: in some cases this is because there will
+only ever be one log line variant for that action, e.g.\ all log lines
+showing that a remote client has connected are matched by a single rule and
+handled by the \texttt{CONNECT} action; in other cases the action is
+required to address a deficiency in the log files, or a complication that
+arises during parsing.  Taking the \texttt{CONNECT} action as an example,
+it creates a new data structure in memory for each client connecting, and
+saves the data extracted by the rule; this data will subsequently be
+entered into the database.  If a data structure already exists for that
+connection it is treated as a symptom of a bug, and the action issues a
+warning containing the full contents of the existing data structure, plus
+the log line that has just been parsed.
 
 The ability to add special purpose actions to deal with difficulties and
-new requirements which are discovered during parser development is one of
-the strengths of this architecture.  Instead of writing one monolithic
-function which must be modified to support new behaviour, with all the
+new requirements that are discovered during parser development is one of
+the strengths of this architecture.  Instead of writing a single monolithic
+function that must be modified to support new behaviour, with all the
 attendant risks of adversely affecting the existing parser, when a new
-requirement arises an independent action can be written to satisfy it.  In
-some cases the new action will require the cooperation of other actions,
+requirement arises an independent action can be written to satisfy it.  
+Sometime the new action will require the cooperation of other actions,
 e.g.\ to set or check a flag.  There is a possibility of introducing
@@
 During development of the Postfix log parser it became apparent that in
-addition to the obvious variety in log lines, there were a large number of
+addition to the obvious variety in log lines, there were many 
 complications to be overcome.  Some were the result of deficiencies in
@@
 
-Actions may return a modified input line which will be parsed as if read
+Actions may return a modified input line that will be parsed as if read
 from the input stream, allowing for a simplified version of cascaded
-parsing~\cite{cascaded-parsing}.  This is a powerful facility allowing a
-number of rules and actions to parse a single input.
+parsing~\cite{cascaded-parsing}.  This is a powerful facility allowing 
+several rules and actions to parse a single input.
 
@@
 input and the action to invoke when the match is successful.  The Postfix
-log parser stores its rules in the same SQL database the results are stored
-in, removing any doubt as to which set of rules was used to produce a set
+log parser stores the rules it uses in the same SQL database the results are stored
+in, removing any doubt about which set of rules was used to produce a set
 of results; other implementations are free to store the rules in whatever
@@
 automatically extracted and saved from the log line based on keywords in
-the rule's regex, but there is also a facility to specify extra data to be
+the rule's regex, but there is also a mechanism to specify extra data to be
 saved.
@@
 the choice of machine learning technique would be constrained by the size
-of typical data sets (see \sectionref{Results}).  Techniques which require
+of typical data sets (see \sectionref{Results}).  Techniques that require
 the full data set when training would be impractical: Instance Based
-Learning~\cite{instance-based-learning} techniques which automatically
+Learning~\cite{instance-based-learning} techniques that automatically
 determine which inputs from the training set are valuable and which inputs
-can be discarded would probably reduce the data required to a manageable
+can be discarded might reduce the data required to a manageable
 size.  This avenue of research and development has not been pursued by the
@@
 the first matching rule wins, and the order in which rules are tried
-against each input may change between parser invocations.  Overlapping
+against each input might change between parser invocations.  Overlapping
 rules are frequently a requirement, allowing a more specific rule to match
@@
 computational power to \FA{}, and it is possible to convert between the
-two, so regex overlap could be detected by finding a non-empty intersection
+two, so regex overlap can be detected by finding a non-empty intersection
 of two \FA{}\@.  The standard equation for \FA{} intersection (given for
@@
 
-It is possible to define pathological regexes, which fall into two main
-categories: regexes which match every input, and regexes which consume
-excessive amounts of CPU time during matching.  Defining a regex which
+It is possible to define pathological regexes that fall into two main
+categories: regexes that match every input, and regexes that consume
+excessive amounts of CPU time during matching.  Defining a regex that
 matches every input is trivial: \verb!/^/! matches the start of every
@@
         actions during execution, providing some context sensitivity.
-        Actions can not inspect past or future inputs, however actions can
+        Actions can not inspect past or future inputs, but actions can
         return a modified input that will be parsed as if read from the
@@
         compiling a programming language, this architecture does not
-        require a fixed grammar specification which inputs must adhere to.
+        require a fixed grammar specification that inputs must adhere to.
         The architecture is capable of dealing with interleaved inputs, out
@@
 is shown in \graphref{Percentage parsing time increase of maximum ruleset
-over minimum ruleset}.  It is clear that the increased number of rules has
+over minimum ruleset}.  Clearly that the increased number of rules has
 a noticeable performance impact with reverse ordering, and a lesser impact
@@
 Correct parsing of log lines must be measured first.  Warnings are issued
-for any log lines which are not parsed; no such warnings are issued while
+for any log lines that are not parsed; no such warnings are issued while
 parsing the \numberOFlogFILES{} log files, so it can be safely concluded
@@
 readable form; examining these tables with reference to the log files and
-database is the best way to detect mails which were not handled properly.
+database is the best way to detect mails that were not handled properly.
 The implementation issues warnings when it detects any errors or
@@
 to decide which deficiencies are most important and require attention
-first, rather than being forced to fix the first error which arises.
+first, rather than being forced to fix the first error that arises.
 
@@
 data to analyse and optimise the set of rules in use, followed by
-identifying patterns in the data which could be used to write new filters
+identifying patterns in the data that could be used to write new filters
 to recognise and reject spam rather than accepting it.  The parser provides
-the data in a normalised form which is far easier to use as input to new or
+the data in a normalised form that is far easier to use as input to new or
 existing implementations of machine-learning algorithms than trying to
@@
 this approach could be applied to optimising the behaviour of any system
-given sufficient log messages to analyse.  An alternative approach to black
-box optimisation which uses application profiling in conjunction with the
+given sufficient log messages to analyse.  An alternate approach to black
+box optimisation that uses application profiling in conjunction with the
 application's error messages to improve the error messages shown to users
 is described in~\cite{black-box-error-reporting}; profiling data may be
-useful in supplementing systems which fail to provide adequate log
+useful in supplementing systems that fail to provide adequate log
 messages.
@@
 a genuine need, rather than a proof of concept whose sole purpose is to
-illustrate a new idea; it deals with the oddities and difficulties which
+illustrate a new idea; it deals with the oddities and difficulties that
 occur in the real world, rather than a clean, idealised scenario developed

### 2008-08-22 doc/sgai-2008.tex
Minor reformatting.

@@
 
-    \item The mail administrator can define custom rejection
-        messages.
+    \item The mail administrator can define custom rejection messages.
 
@@
 Generally rules specifying the \texttt{UNINTERESTING} action are rules to
-parse log lines that are not associated with a specific mail, e.g.\
-notices about configuration files changing.  The remaining actions only
-have one or two associated rules: in some cases this is because there will
-only ever be one log line variant for that action, e.g.\ all log lines
-showing that a remote client has connected are matched by a single rule and
-handled by the \texttt{CONNECT} action; in other cases the action is
-required to address a deficiency in the log files, or a complication that
-arises during parsing.  Taking the \texttt{CONNECT} action as an example,
-it creates a new data structure in memory for each client connecting, and
-saves the data extracted by the rule; this data will subsequently be
-entered into the database.  If a data structure already exists for that
-connection it is treated as a symptom of a bug, and the action issues a
-warning containing the full contents of the existing data structure, plus
-the log line that has just been parsed.
+parse log lines that are not associated with a specific mail, e.g.\ notices
+about configuration files changing.  The remaining actions only have one or
+two associated rules: in some cases this is because there will only ever be
+one log line variant for that action, e.g.\ all log lines showing that a
+remote client has connected are matched by a single rule and handled by the
+\texttt{CONNECT} action; in other cases the action is required to address a
+deficiency in the log files, or a complication that arises during parsing.
+Taking the \texttt{CONNECT} action as an example, it creates a new data
+structure in memory for each client connecting, and saves the data
+extracted by the rule; this data will subsequently be entered into the
+database.  If a data structure already exists for that connection it is
+treated as a symptom of a bug, and the action issues a warning containing
+the full contents of the existing data structure, plus the log line that
+has just been parsed.
 
@@
 During development of the Postfix log parser it became apparent that in
-addition to the obvious variety in log lines, there were many 
-complications to be overcome.  Some were the result of deficiencies in
-Postfix's logging (some of which were improved on or resolved by later
-versions of Postfix); others were due to the vagaries of process
-scheduling, client behaviour, and administrative actions.  All were
-successfully accommodated in the Postfix log parser: adding new actions was
-enough to overcome several of the complications; others required
-modifications to a single existing action to work around the difficulties;
-the remainder were resolved by adapting existing actions to cooperate and
-exchange extra data, changing their behaviour as appropriate.
+addition to the obvious variety in log lines, there were many complications
+to be overcome.  Some were the result of deficiencies in Postfix's logging
+(some of which were improved on or resolved by later versions of Postfix);
+others were due to the vagaries of process scheduling, client behaviour,
+and administrative actions.  All were successfully accommodated in the
+Postfix log parser: adding new actions was enough to overcome several of
+the complications; others required modifications to a single existing
+action to work around the difficulties; the remainder were resolved by
+adapting existing actions to cooperate and exchange extra data, changing
+their behaviour as appropriate.
 
@@
 input and the action to invoke when the match is successful.  The Postfix
-log parser stores the rules it uses in the same SQL database the results are stored
-in, removing any doubt about which set of rules was used to produce a set
-of results; other implementations are free to store the rules in whatever
-fashion suits their needs.  The framework warns about every unparsed input,
-to alert the user that they need to alter or extend their ruleset; the
-Postfix log parser successfully parses every log line in the
+log parser stores the rules it uses in the same SQL database the results
+are stored in, removing any doubt about which set of rules was used to
+produce a set of results; other implementations are free to store the rules
+in whatever fashion suits their needs.  The framework warns about every
+unparsed input, to alert the user that they need to alter or extend their
+ruleset; the Postfix log parser successfully parses every log line in the
 \numberOFlogFILESall{} log files it is currently being tested with.  The
@@
 determine which inputs from the training set are valuable and which inputs
-can be discarded might reduce the data required to a manageable
-size.  This avenue of research and development has not been pursued by the
-authors, but could easily be undertaken independently.
+can be discarded might reduce the data required to a manageable size.  This
+avenue of research and development has not been pursued by the authors, but
+could easily be undertaken independently.
 
@@
 is shown in \graphref{Percentage parsing time increase of maximum ruleset
-over minimum ruleset}.  Clearly that the increased number of rules has
-a noticeable performance impact with reverse ordering, and a lesser impact
+over minimum ruleset}.  Clearly that the increased number of rules has a
+noticeable performance impact with reverse ordering, and a lesser impact
 with shuffled ordering.  The optimal ordering shows a mean increase of
@@
 data to analyse and optimise the set of rules in use, followed by
-identifying patterns in the data that could be used to write new filters
-to recognise and reject spam rather than accepting it.  The parser provides
+identifying patterns in the data that could be used to write new filters to
+recognise and reject spam rather than accepting it.  The parser provides
 the data in a normalised form that is far easier to use as input to new or
@@
 is described in~\cite{black-box-error-reporting}; profiling data may be
-useful in supplementing systems that fail to provide adequate log
-messages.
+useful in supplementing systems that fail to provide adequate log messages.
 

### 2008-08-25 doc/logparser-acronyms.tex
Changes made following suggestions from diction.

@@
     contract between the user and the provider: if the user adheres to the
-    API the provider guarantees it will work, while the provider is free to
-    change the implementation as long as the API is preserved.
+    API the provider guarantees it will work, but is free to change the
+    implementation provided the API is preserved.
 }}
@@
 \newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
-    SMTP is the protocol which transfers mail from the sender to the
+    SMTP is the protocol that transfers mail from the sender to the
     recipient across the Internet.  It is a simple, human readable, plain
-    text protocol, making it quite easy to test and debug problems with it.
-    A detailed description of \SMTP{} is beyond the scope of this document:
-    the original protocol definition is \RFC{}~821~\cite{RFC821}, updated
-    in \RFC{}~2821~\cite{RFC2821}; introductory guides can be found
+    text protocol, making it quite simple to test and debug problems with
+    it.  A detailed description of \SMTP{} is beyond the scope of this
+    document: the original protocol definition is \RFC{}~821~\cite{RFC821},
+    updated in \RFC{}~2821~\cite{RFC2821}; introductory guides can be found
     at~\cite{smtp-intro-01,smtp-intro-02}.
@@
     server to maintain a mail delivery queue, instead relying on the client
-    to maintain it.  Typically the client would be an MTA, and the server
-    would be a delivery agent or a mail store.  Full details are available
+    to maintain it.  Typically the client is an MTA, and the server a
+    delivery agent or a mail store.  Full details are available
     in~\cite{lmtp-rfc}.
@@
 \newacronym{MTA}{Mail Transfer Agent}{description={
-    A Mail Transfer Agent sends and/or receives mail via
+    A Mail Transfer Agent sends and receives mail via
     SMTP\backslashat{}.
@@
 \newacronym{ISP}{Internet Service Provider}{description={
-    An ISP is a company which sells Internet access to consumers.
+    An ISP is a company that sells Internet access to consumers.
 }}
@@
     One of the other Postfix log parsers reviewed and the only example of
-    published prior art~\cite{log-mail-analyser} found by the author.
+    published prior art~\cite{log-mail-analyser} found that specifically
+    deals with parsing Postfix logs.
 }}
@@
     so the program's name is not suitable as a distinguishing identifier;
-    instead each process executing is given a pid which is guaranteed to be
+    instead each process executing is given a pid that is guaranteed to be
     unique for the lifetime of the process.  Once the process has
@@
 \newacronym{UCE}{Unsolicited Commercial Email}{description={
-    UCE is a more restrictive definition of spam than most people would
-    use: it only covers mail that is explicitly commercial, thus excluding
-    viruses, Bayesian poisoning mails, backscatter, and those annoying
-    chain letters you get from friends.
+    UCE is a more restrictive definition of spam than most people use: it
+    only covers mail that is explicitly commercial, thus excluding viruses,
+    Bayesian poisoning mails, backscatter, and those annoying chain letters
+    you get from friends.
 }}
@@
 \newacronym{LCD}{Lowest Common Denominator}{description={
-    LCD is a mathematical term which is used figuratively to mean the least
+    LCD is a mathematical term that is used figuratively to mean the least
     useful member of a set of alternatives.
@@
     our understanding of the languages normally used by humans (e.g.\
-    English, Spanish, Japanese), with the goal of writing programs which
+    English, Spanish, Japanese), with the goal of writing programs that
     can understand human languages.

### 2008-08-25 doc/logparser-architecture.tex
Changed normal to optimal.
Changes made following suggestions from diction.

@@
 
-XXX REWRITE AS NECESSARY TO MERGE AND SUPPLEMENT THE EXISTING CONTENT\@.
+XXX REWRITE TO MERGE AND SUPPLEMENT THE EXISTING CONTENT\@.
 
@@
         handle complications, and cause data to be saved to the database.
-        The actions perform the work of reconstructing the journey a mail
-        takes through Postfix.  Details of the actions available in the
-        Postfix parser can be found in \sectionref{actions-in-detail}, and
+        The actions do the work of reconstructing the journey a mail takes
+        through Postfix.  Details of the actions available in the Postfix
+        parser can be found in \sectionref{actions-in-detail}, and
         \sectionref{adding new actions} describes the process of adding new
@@
     \item [framework] The framework is responsible for loading rules,
-        managing data structures, finding the rule which matches each log
+        managing data structures, finding the rule that matches each log
         line, invoking the correct action, etc\@.  The framework is
@@
 
-It may help to think of the rules and actions as components which plug into
+It may help to think of the rules and actions as components that plug into
 the framework.
@@
 actions: the framework provides services to the actions, and does not need
-to deal with the complications which arise, or the task of reconstructing a
+to deal with the complications that arise, or the task of reconstructing a
 mail's journey through Postfix; actions benefit from having services
@@
 adversely affecting existing parsing, particularly as there may be edge
-cases which are not immediately obvious ---
-\sectionref{yet-more-aborted-delivery-attempts} describes a complication which
-occurs only four times in \numberOFlogFILES{} log files tested.  Requiring
-changes to the parser's code also complicates upgrades, as the changes must
-be preserved during the upgrade, and may clash with changes made by the
-developer (\sectionref{adding new actions} describes safely adding new
-actions).  \parsername{} allows the user to add new rules to the database
-without changing the parsing algorithm, unless the new log lines to be
-parsed require functionality not already provided by the algorithm.  If the
-new log lines do require new functionality, new actions can be added to the
-parser without modifying existing actions or other parts of the algorithm;
-only in the rare case that the new actions require support from other
-sections of the code will more extensive changes be required.
+cases that are not immediately obvious ---
+\sectionref{yet-more-aborted-delivery-attempts} describes a complication
+that occurs only four times in \numberOFlogFILES{} log files tested.
+Requiring changes to the parser's code also complicates upgrades, as the
+changes must be preserved during the upgrade, and may clash with changes
+made by the developer (\sectionref{adding new actions} describes safely
+adding new actions).  \parsername{} allows the user to add new rules to the
+database without changing the parsing algorithm, unless the new log lines
+to be parsed require functionality not already provided by the algorithm.
+If the new log lines do require new functionality, new actions can be added
+to the parser without modifying existing actions or other parts of the
+algorithm; only in the rare event that the new actions require support from
+other sections of the code will more extensive changes be required.
 
@@
 creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is interesting how two
-apparently different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
+\ATN{} and the parser is accidental, but it is obvious that the two
+different approaches share an underlying separation of concerns; this
+appears to be a natural division of responsibility and functionality.
 \Tableref{Similarities between ATN and this architecture} shows the
@@
                                         of transitions              \\
-                        &               & or actions which 
+                        &               & or actions that 
                                         constitutes a valid  input. \\
@@
                         &               & input must meet to be
-                                        considered valid.           \\
+                                        accepted as valid.          \\
         Abbreviations   & Rules         & Responsible for 
@@
 
-Rule have certain characteristics which may help in understanding the
+Rule have certain characteristics that may help in understanding the
 parser:
@@
     \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a facility for prioritising the rules
+        that log line, but there is a mechanism for prioritising the rules
         so that more specific rules can be tried first.
@@
 
-The parser does not try to detect overlapping rules; that
-responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the order in which rules are tried against each line may change between log
-files, and the first matching rule wins.  Overlapping rules are frequently
-a requirement, allowing a more specific rule to match some log lines and a
-more general rule to match the majority, e.g.\ separating \SMTP{} delivery
-to specific sites from \SMTP{} delivery to the rest of the world.  The
-algorithm provides a facility for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Negative priorities may be useful
-for catchall rules.
+The parser does not try to detect overlapping rules; that responsibility is
+left to the author of the rules.  Unintentionally overlapping rules lead to
+inconsistent parsing and data extraction because the order in which rules
+are tried against each line may change between log files, and the first
+matching rule wins.  Overlapping rules are frequently a requirement,
+allowing a more specific rule to match some log lines and a more general
+rule to match the majority, e.g.\ separating \SMTP{} delivery to specific
+sites from \SMTP{} delivery to the rest of the world.  The algorithm
+provides a mechanism for ordering overlapping rules: the priority field in
+each rule (defaults to zero).  Negative priorities may be useful for
+catchall rules.
 
@@
         reversed rules (described in \sectionref{rule ordering for
-        efficiency}).  Parse a number of log files using normal sorting,
-        then dump a textual representation of the rules, connections and
-        results tables.  Repeat with shuffled and reversed rules, starting
-        with a fresh database.  If there are no overlapping rules the
-        tables from each run will be identical; differences indicate
-        overlapping rules.  Which rules overlap can be determined by
-        examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing results
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files which match more than one
-        rule.
+        efficiency}).  Parse several log files using optimal ordering, then
+        dump a textual representation of the rules, connections and results
+        tables.  Repeat with shuffled and reversed ordering, starting with
+        a fresh database.  If there are no overlapping rules the tables
+        from each run will be identical; differences indicate overlapping
+        rules.  The rules that overlap can be determined by examining the
+        differences in the tables: each result contains a reference to the
+        rule which created it, if the references differ between runs the
+        two rules referenced in the differing results overlap.
+        Unfortunately this method cannot prove the absence of overlapping
+        rules; it can detect overlapping rules, but only if there are log
+        lines in the input files that match more than one rule.
 
@@
 The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be
-capable of handling multiple mails in parallel, each potentially at a
-different stage in its journey, without any interference between mails ---
-except in the minority of cases where intra-mail interference is required.
-The best way to implement this is to maintain state information for every
-unfinished mail and manipulate the appropriate mail correctly for each log
-line encountered.
+the possibility of handling each mail in isolation; the parser must be able
+to handle multiple mails in parallel, each potentially at a different stage
+in its journey, without any interference between mails --- except in the
+minority of cases where intra-mail interference is required.  The best way
+to implement this is to maintain state information for every unfinished
+mail and manipulate the appropriate mail correctly for each log line
+encountered.
 

### 2008-08-25 doc/logparser-background.tex
Changes made following suggestions from diction.

@@
 the data gleaned from those log files, providing a consistent and simple
-view of the log files which future tools can utilise.  The gathered data
+view of the log files that future tools can utilise.  The gathered data
 can then be used to optimise current anti-spam measures, provide a baseline
@@
 Postfix; bugs can be fixed or limitations removed from the parser; these
-changes will not cause the user to be negatively impacted.  Statistics
-and/or graphs can be generated from the database; new restrictions can be
+changes will not cause the user to be negatively affected.  Statistics
+and graphs can be generated from the database; new restrictions can be
 tested and the results inspected; trends in the fight against spam can
@@
 same as the usage adapts.  Using a database simplifies writing programs
-which need to interact with the data in several ways:
+that need to interact with the data in several ways:
 
@@
 
-    \item It is already possible to access databases from the vast majority
-        of programming languages, allowing a developer to access the data
-        gathered using their preferred programming language, rather than
-        being restricted to the language the parser is written in.  It is
-        often possible to write an interface layer allowing code written in
-        one language to be used in another language, but this greatly
-        increases the effort required to use the parser.
+    \item It is already possible to access databases from most programming
+        languages, allowing a developer to access the data gathered using
+        their preferred programming language, rather than being restricted
+        to the language the parser is written in.  It is often possible to
+        write an interface layer allowing code written in one language to
+        be used in another language, but this greatly increases the effort
+        required to use the parser.
 
@@
         databases provide a program, of varying complexity and
-        sophistication, which can be used for ad hoc queries with minimal
+        sophistication, that can be used for ad hoc queries with minimal
         investment of time.
@@
         English and Australian English).  Depending on the database in use
-        there may be tools available which reduce or remove the requirement
+        there may be tools available that reduce or remove the requirement
         to know \SQL{}; for SQLite (the default database used by the
-        implementation) there are several available~\cite{sqlite-guis}.
+        implementation) several are available~\cite{sqlite-guis}.
 
@@
 sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity of the protocol many virus and/or spam sending programs fail
+protocol, making it quite simple to test and debug problems with it.  Despite
+the simplicity of the protocol many virus and spam sending programs fail
 to implement it properly, so requiring strict adherence to the protocol
@@
 by the server and carries on regardless.  There are many spam and virus
-sending programs which are this deficient --- unfortunately others
+sending programs that are this deficient --- unfortunately others
 (particularly newer programs) were written by competent programmers, or
@@
 possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses which were not enclosed
+command, or accepting sender or recipient addresses that were not enclosed
 in \texttt{<>}.
@@
 combined and bypassed on a per-client, per-recipient or per-sender basis,
-allowing varying levels of stricture and/or permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions, with policy
+allowing varying levels of stricture and permissiveness.  Postfix leverages
+simple lookup tables to support arbitrarily complicated user-defined
+sequences of restrictions and exceptions, with policy
 servers\footnote{Policy servers will be explained in \sectionref{policy
@@
 mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the log files generated.  While it is easy to use standard Unix text
+in the log files generated.  Although it is easy to use standard Unix text
 processing utilities to determine the fate of an individual email,
 following the journey an email takes through Postfix can be quite
-difficult.  For the majority of mails the journey is simple and brief, but
-the remaining minority can be quite complex (see \sectionref{additional
+difficult.  For most mails the journey is simple and brief, but the
+remaining minority can be quite complex (see \sectionref{additional
 complications} for details).
@@
 details can be found in~\cite{postfix-overview}.  Each log line contains
-the name of the Postfix component which produced it, and this information
+the name of the Postfix component that produced it, and this information
 is used when determining which rules should be used to parse each log line
 (see \sectionref{rule characteristics} for details).  This design has
-positive security implications also: those components which interact with
+positive security implications also: those components that interact with
 other hosts are not privileged,\footnote{Privilege in this case means the
@@
 available to ordinary users.} so bugs in those components will not give an
-attacker extra privileges; those components which are privileged do not
+attacker extra privileges; those components that are privileged do not
 interact with other hosts, making it much more difficult for an attacker to
-exploit any bugs which may exist in those components.
+exploit any bugs that may exist in those components.
 
@@
 restriction, allowing restrictions to be chosen based on the client \IP{}
-address, HELO hostname, sender address, recipient address,
-etc.\footnote{E.g.\ the administrator may require that all clients on the
-local network have valid \DNS{} entries, to prevent people sending mail
+address, client hostname, HELO hostname, sender address, or recipient
+address.\footnote{E.g.\ the administrator may require that all clients on
+the local network have valid \DNS{} entries, to prevent people sending mail
 from unknown machines.}  The administrator can define new restrictions as a
@@
 
-Postfix uses simple lookup tables as the deciding factor when evaluating
-some restrictions, e.g.\ in the example restriction below\newline
+Postfix uses simple lookup tables to make decisions when evaluating some
+restrictions, e.g.\ in the example restriction below\newline
 \tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
@@
 
-The restriction \texttt{check\_client\_access} checks whether the \IP{}
-address of the connected client is found in the file and returned the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
+The restriction \texttt{check\_client\_access} checks if the \IP{} address
+of the connected client is found in the file and returns the associated
+action if found; the method of searching the file is dependant on the type
+of the file (\texttt{cidr} in the example) --- see
 \cite{postfix-lookup-tables} for more details.  Other restrictions
@@
         backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
+        joe~jobs~\cite{Wikipedia-joe-job}, but there has been considerable
         resistance to the proposal because it breaks or vastly complicates
@@
         Greylisting is obviously worth using, at least at the moment,
-        particularly when you factor in Greylisting's position as the final
-        restriction which a mail must overcome:\footnote{Greylisting is the
+        particularly when you consider Greylisting's position as the final
+        restriction that a mail must overcome:\footnote{Greylisting is the
         final restriction a mail must overcome in the configuration used on
@@
         list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails which have passed every
+        Greylisting only takes effect for mails that have passed every
         other restriction.
@@
         binary, and the administrator must manually whitelist clients if
-        they are to bypass a restriction; using a threshold which requires
-        hitting several restrictions frees the administrator from
-        whitelisting clients which fall foul of one restriction only.
+        they are to bypass a restriction; using a threshold that requires a
+        delivery attempt to hit several restrictions frees the
+        administrator from whitelisting clients that fall foul of one
+        restriction only.
 

### 2008-08-25 doc/logparser-conclusion.tex
Changes made following suggestions from diction.

--- 2008-07-10 doc/logparser-conclusion.tex
@@
 resulting initial increase in complexity the task is quite tractable,
-though it does raise interesting efficiency and optimisation questions
-(answered in \sectionref{rule efficiency}).  Providing a tool
-(\sectionref{creating new rules}) to ease the generation of \regexes{} from
-unparsed log lines should greatly help users add rules to parse formerly
-unparsed log lines.
+though it does raise efficiency and optimisation questions (answered in
+\sectionref{parser efficiency}).  Providing a tool (\sectionref{creating
+new rules}) to ease the generation of \regexes{} from unparsed log lines
+should greatly help users add rules to parse formerly unparsed log lines.
 
@@
 lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively easy task.  The simplicity of adding a new action frees the
+relatively simple task.  The simplicity of adding a new action frees the
 implementor from worrying about how their action might disrupt parsing of
@@
 last mention of the queueid in question, is extremely time consuming and
-error prone.  In some cases the task is not to spot the line which is
-unusual, but to spot that a line normally present is missing, i.e.\ to
-realise that one line amongst hundreds is absent.  In all cases the
-evidence must be used to construct a hypothesis to explain the
-irregularities, and that hypothesis must then be tested in the parser; if
-successful, the parser must be modified to deal with the irregularities,
-without adversely affecting the existing functionality.  The complications
-described in \sectionref{additional complications} were solved in the order
-they are described in, and that order closely resembles the frequency with
-which they occur; the most frequently occurring complications dominate the
+error prone.  Sometimes the task is not to spot the line that is unusual,
+but to spot that a line normally present is missing, i.e.\ to realise that
+one line amongst hundreds is absent.  In all cases the evidence must be
+used to construct a hypothesis to explain the irregularities, and that
+hypothesis must then be tested in the parser; if successful, the parser
+must be modified to deal with the irregularities, without adversely
+affecting the existing functionality.  The complications described in
+\sectionref{additional complications} were solved in the order they are
+described in, and that order closely resembles the frequency with which
+they occur; the most frequently occurring complications dominate the
 warning messages produced, and so naturally they are the first

### 2008-08-25 doc/logparser-glossary.tex
Changes made following suggestions from diction.

@@
 \glossary{name={pickup},description={
-    Pickup is the service which deals with mail submitted locally via
+    Pickup is the service that deals with mail submitted locally via
     postdrop and sendmail; it passes all submitted mail to cleanup.  The
@@
 \glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon which manages the mail queue, determining
+    Qmgr is the Postfix daemon that manages the mail queue, determining
     which mails will be delivered next.  Qmgr orders the mails based on the
@@
     (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands which send mail depend on; Postfix sendmail executes postdrop
+    commands that send mail depend on; Postfix sendmail executes postdrop
     to place a new mail in the queue.  The definitive documentation is
@@
 \glossary{name={smtpd},description={
-    Smtpd is the Postfix program which accepts mail via \SMTP{}, and
+    Smtpd is the Postfix program that accepts mail via \SMTP{}, and
     implements all the anti-spam restrictions Postfix provides.  The
@@
 \glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail which should not be replied
+    $<>$ is the sender address used for mail that should not be replied
     to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed

### 2008-08-25 doc/logparser-implementation.tex
Changes made following suggestions from diction.

@@
 
-    \item Postfix logs sufficient information to make it possible to
-        accurately reconstruct the actions it has taken.  There are a
-        number of heuristics used when parsing; see
+    \item Postfix logs enough information to make it possible to accurately
+        reconstruct the actions it has taken.  There are several heuristics
+        used when parsing; see
         \sectionref{identifying-bounce-notifications},
@@
 lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source
-code.} there are advantages to treating Postfix as a black box while
-developing the parser:
+code.} there are advantages to treating Postfix as a black box during
+parser development:
 
@@
     \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  An interesting
-        project would be to compare the empirical knowledge inherent in the
-        parsing algorithm with the documentation and source code of
-        Postfix.
+        information gathered from empirical evidence.  A separate project
+        could compare the empirical knowledge inherent in the parsing
+        algorithm with the documentation and source code of Postfix to see
+        how closely the two agree.
 
@@
 to be saved, and the action to be executed; they also have several fields
-which aid the user in understanding the meaning of the log lines parsed by
+that aid the user in understanding the meaning of the log lines parsed by
 each rule.  The rules are described in detail in \sectionref{rules} but the
@@
 database is the connection between rules and results: if more information
-is required when examining a result, the rule which produced the result is
-available for inspection --- each result references the rule which created
+is required when examining a result, the rule that produced the result is
+available for inspection --- each result references the rule that created
 it.  There is no ambiguity about which rule resulted in a particular
@@
 
-    \item [id] A unique identifier for each rule which other tables can use
+    \item [id] A unique identifier for each rule that other tables can use
         to refer to a specific rule.
@@
 
-    \item [restriction\_name] The restriction which caused the mail to be
-        rejected.  Only applicable to rules which have an action of
+    \item [restriction\_name] The restriction that caused the mail to be
+        rejected.  Only applicable to rules that have an action of
         \texttt{REJECTION}, other rules should have an empty string.
@@
 
-            \item [BOUNCED] The mail has bounced, due to a mail loop,
+            \item [BOUNCED] The mail has bounced, because of a mail loop,
                 delivery failure, or five day timeout.
@@
             \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but which does log
+                the parser is not interested in per se, but that does log
                 useful information, supplementing other log lines.
 
-            \item [IGNORED] An action which is not only uninteresting in
-                itself, but which also provides no useful data.
+            \item [IGNORED] An action that is not only uninteresting in
+                itself, but that also provides no useful data.
 
@@
                 processes will be terminated.  This does not negatively
-                impact on the log files or mail queued by Postfix for
+                affect the log files or mail queued by Postfix for
                 delivery.
@@
         log lines the rule applies to.  This avoids needlessly trying rules
-        which will not match the log line, or worse, might match
+        that will not match the log line, or worse, might match
         unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines which are not parsed by program specific
+        against any log lines that are not parsed by program specific
         rules.
@@
         in a large, data dependent slowdown.  Efficiency concerns are
-        discussed in \sectionref{parser efficiency}, with the impact of
-        caching compiled \regexes{} covered in \sectionref{Caching compiled
-        regexes}.
+        discussed in \sectionref{parser efficiency}, with caching compiled
+        \regexes{} covered in \sectionref{Caching compiled regexes}.
 
     \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is:
-        \newline \tab{} \texttt{smtp\_code = 1; recipient = 2, sender = 4;}
-        \newline i.e.\ semi-colon or comma separated assignment statements,
-        with the variable name on the left and the matching capture from
-        the \regex{} on the right hand side.  Data specified by
+        log line will be extracted.  The format is: \newline \tab{}
+        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline i.e.\
+        semi-colon or comma separated assignment statements, with the
+        variable name on the left and the matching capture from the
+        \regex{} on the right hand side.  Data specified by
         \texttt{result\_cols} will be entered in the \texttt{results}
-        table, while \texttt{connection\_cols} populates the
+        table, whereas \texttt{connection\_cols} populates the
         \texttt{connection} table.  The list of acceptable variable names
@@
     \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data which is not present in the log line: e.g.\ setting
+        piece of data that is not present in the log line: e.g.\ setting
         \texttt{smtp\_code} when mail is accepted.  The format and allowed
@@
         rejections logged before a mail has been accepted (a queueid will
-        not have been allocated), or log lines which are not tied to one
+        not have been allocated), or log lines that are not tied to one
         particular mail.
@@
         logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite easy to rectify this, as described in \sectionref{logging
-        helo}.
+        is quite simple to rectify this, as described in
+        \sectionref{logging helo}.
 
@@
 
+XXX REWRITE THE SENTENCE AND FOOT NOTE\@.
+
 Every interesting\footnote{All log lines are interesting unless the
-postfix\_action of the rule which parses them is \texttt{INFO} or
+postfix\_action of the rule that parses them is \texttt{INFO} or
 \texttt{IGNORE}, as described in \sectionref{actions-in-detail}.} log line
@@
 
-    \item [rule\_id] The id of the entry in the rules table which matched
+    \item [rule\_id] The id of the entry in the rules table that matched
         the log line and created this result.
@@
 frequent searching, sorting and manipulation of the data to produce
-customised reports and/or statistics.
+customised reports or statistics.
 
@@
 
-The example rule below would match the following log line:
+The example rule below matches the following log line:
 
@@
     \item [program and regex:] If the program in the rule equals the
-        program which logged the log line, a match using the rule's
+        program that logged the log line, a match using the rule's
         \regex{} will be attempted against the log line; if the match is
@@
 
-    \item [queueid:] The index of the capture in the \regex{} which
+    \item [queueid:] The index of the capture in the \regex{} that
         supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail can be found by queueid and
+        queueid.  This allows the correct mail to be found by queueid and
         actions performed on it.
@@
 restrictions which suits their needs, including using different \DNSBL{}
-services, policy servers, or custom rejection messages.  To facilitate easy
-parsing of new log lines, the parser's design separates parsing rules from
+services, policy servers, or custom rejection messages.  To facilitate
+parsing new log lines, the parser's design separates parsing rules from
 parsing actions: adding new actions is difficult, but adding new rules to
@@
     \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained while writing rules and performs a large number of
+        gained when writing rules and performs a large number of
         substitutions on the log input lines, replacing commonly occurring
@@
         standard start of rejection messages, etc.) with \regex{} keywords
-        which the parser will expand when it loads the rule (see the
+        that the parser will expand when it loads the rule (see the
         \regex{} entry in \sectionref{rule attributes}).  The purpose of
-        this phase is to utilise existing knowledge to create more accurate
+        this step is to utilise existing knowledge to create more accurate
         \regexes{}.  The new log lines are written to a temporary file,
@@
         The original algorithm only maintains counters for words whose hash
-        result from the previous phase has a high frequency; this reduces
+        result from the previous step has a high frequency; this reduces
         the number of counters maintained by the algorithm, reducing the
@@
         the modified algorithm omits the hashing step and maintains
-        counters for all tokens. In addition the majority of unique or
-        infrequently occurring words will have been substituted with
-        keywords during the first phase, vastly reducing the number of
-        tuples to maintain counters for.
+        counters for all tokens. In addition most uncommon or infrequently
+        occurring words will have been substituted with keywords during the
+        first step, vastly reducing the number of tuples to maintain
+        counters for.
 
@@
         lines is a good starting point; obviously the threshold depends on
-        the number and type of input log lines.  The new \regexes{} are
+        the number and content of input log lines.  The new \regexes{} are
         displayed for the user to add to the database, either as new rules
@@
 
-A second utility is also provided which reads a list of new \regexes{} and
+A second utility is also provided that reads a list of new \regexes{} and
 the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines which match
+line against each \regex{}, counting the number of log lines that match
 each \regex{}, warning the user if an input log line is matched by more
@@
 complications of parsing, the eccentricities and oddities of Postfix log
-files, presenting the resulting data in a normalised, easy to use
+files, presenting the resulting data in a normalised, simple to use
 representation.  The algorithm's task is to follow the journey each mail
@@
     \item The first set of complications encountered: initially obvious
-        difficulties which had to be overcome.
+        difficulties that had to be overcome.
 
@@
 
-    \item Additional complications which have arisen during the development
+    \item Additional complications that have arisen during the development
         of this parser are documented, the solutions to those
@@
 
-\Figureref{flow chart image part 1} and \figureref{flow
-chart image part 2} show the paths the data representing a mail can take
-through the parser algorithm.  The flow chart covers the most common paths
-only; there are additional, uncommon paths which are excluded for the sake
-of clarity; details of the deviations can be found in
-\sectionref{additional complications}.
+\Figureref{flow chart image part 1} and \figureref{flow chart image part 2}
+show the paths the data representing a mail can take through the parser
+algorithm.  The flow chart covers the most common paths only; there are
+additional, uncommon paths excluded for the sake of clarity; details of the
+deviations can be found in \sectionref{additional complications}.
 
@@
         its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children which are waiting to be entered in the
+        The parent may have children that are waiting to be entered in the
         database (flowchart:~23); each of those children's data is cleaned
@@
 The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-which is far more complicated than the behaviour explicitly encoded in the
+that is far more complicated than the behaviour explicitly encoded in the
 algorithm.  The top-level parser could be written in pseudo-code like so
@@
     \item Four implicit branches, where the transition is determined by the
-        next log line which is processed for that mail, e.g.\ state 1 can
+        next log line that is processed for that mail, e.g.\ state 1 can
         transition to either state 4 or state 5.  (1--\{4,5\},
@@
 
-            \item [host] The hostname of the server which logged the line.
-
-            \item [program] The name of the program which logged the line.
-
-            \item [pid] The \pid{} of the program which logged the line.
+            \item [host] The hostname of the server that logged the line.
+
+            \item [program] The name of the program that logged the line.
+
+            \item [pid] The \pid{} of the program that logged the line.
 
@@
         expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts which should occur
+        infrequently, as there are many other timeouts that should occur
         in the intervening hours: \DNS{} timeouts, timeouts reading data
@@
 been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine which
+actions as painless as possible.  The implementor writes a subroutine that
 accepts the standard arguments given to actions, and registers it as an
@@
 This section presented the core of the parser, starting with a very high
-level view and the initial complications which arose.  A flow chart showing
+level view and the initial complications that arose.  A flow chart showing
 the paths a mail may take through the nascent simplified algorithm was
@@
 when bounce notifications are generated; suddenly there will be log entries
-for a mail which lacks an obvious source.  There are similarities to the
+for a mail that lacks an obvious source.  There are similarities to the
 problem of re-injected mails discussed in \sectionref{tracking re-injected
 mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line which identifies their source.  Heuristics
+not eventually have a log line that identifies their source.  Heuristics
 must be used to identify bounce notifications, and those heuristics are:
@@
 it continues with until delivery is complete.\footnote{Microsoft Outlook is
-one client which behaves in this fashion; no attempt has been made to
-identify other clients which act in a similar way.}  An example dialogue
+one client that behaves in this fashion; no attempt has been made to
+identify other clients that act in a similar way.}  An example dialogue
 exhibiting this behaviour is presented below (lines starting with a three
@@
         allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables which needs to be dealt with.
+        in the state tables that needs to be dealt with.
 
@@
 The author has only observed this complication occurring after a timeout,
-though there may be other circumstances which trigger it.  Sometimes the
+though there may be other circumstances that trigger it.  Sometimes the
 \daemon{cleanup} line is logged after the timeout line; parsing this line
@@
 line.  This is incorrect because the line actually belongs to the mail
-which has just been discarded; the next log line for that queueid will be
+that has just been discarded; the next log line for that queueid will be
 seen when the queueid is reused for a different mail, causing a queueid
@@
 
-The solution to this problem is to examine the list of programs which have
+The solution to this problem is to examine the list of programs that have
 logged lines for each mail, comparing the list against a table of
@@
 
-This is yet another complication which only occurs during periods of
+This is yet another complication that only occurs during periods of
 extremely high load, when process scheduling vagaries and even hard disk
@@
         2.5. XXX TEST THIS AGAIN.}  Rectifying this is relatively simple:
-        create a restriction which is guaranteed to warn for every accepted
+        create a restriction that is guaranteed to warn for every accepted
         mail, as follows:

### 2008-08-25 doc/logparser-introduction.tex
Changes made following suggestions from diction.

@@
 whether it was to debug a problem, explain to a user why their mail is
-being rejected, or check whether new anti-spam measures are working.  The
-more adventurous will have generated statistics to show how many hits each
-of their anti-spam measures has gotten in the last week, and possibly even
+being rejected, or check if new anti-spam measures are working.  The more
+adventurous will have generated statistics to show how many hits each of
+their anti-spam measures has had in the last week, and possibly even
 generated some graphs to clearly illustrate these statistics to management
@@
 per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this type of processing is the
+independently.  One of the barriers to this kind of processing is the
 unstructured nature of Postfix log files, where each log line was added on
@@
 paper documents the difficult process of parsing Postfix log files,
-presenting \PLP{}, a program which parses log files and places the
-resulting data into a database for later use.  The gathered data can then
-be used to optimise current anti-spam measures, provide a baseline to test
-new anti-spam measures against, or to produce statistics showing how
-effective those measures are.  There are numerous other uses for such data:
+presenting \PLP{}, a program that parses log files and places the resulting
+data into a database for later use.  The gathered data can then be used to
+optimise current anti-spam measures, provide a baseline to test new
+anti-spam measures against, or to produce statistics showing how effective
+those measures are.  Numerous other uses are possible for such data:
 improving server performance by identifying troublesome destinations and
@@
 customer newsletters) and restricting those uses to off-peak times;
-detecting virus outbreaks which propagate via email; as a base for billing
+detecting virus outbreaks that propagate via email; as a base for billing
 customers on a shared server.  Preserving the raw data enables users to
@@
 Section~\ref{parsing-algorithm} contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered which
+a naive parsing algorithm and the complications initially encountered that
 shaped the full algorithm.  A flow chart and a discussion of the emergent
@@
 explanation of the different stages of the initial algorithm.  The
-framework which actions and rules fit into is documented, then the actions
+framework that actions and rules fit into is documented, then the actions
 taken during execution of the algorithm are described, followed by the
@@
 description of the further complications discovered, and their solutions
-which complete the parser.
+that complete the parser.
 

### 2008-08-25 doc/logparser-other-parsers-review.tex
Changes made following suggestions from diction.

--- 2008-07-21 doc/logparser-other-parsers-review.tex
@@
 this area, have been reviewed as part of the background research for this
-project.  None of the parsers reviewed perform the type of advanced parsing
+project.  None of the parsers reviewed perform the kind of advanced parsing
 and log correlation required for this project; all bar one are intended to
@@
 report of varying complexity and detail, whereas the program described here
-does not attempt to produce a report at all; that responsibility is
-deferred to a separate program, to be developed later.  The parsing
-algorithm and program developed for this project are designed to enable
-much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
+does not produce a report at all; that responsibility is deferred to a
+separate program, to be developed later.  The parsing algorithm and program
+developed for this project are designed to enable much more detailed log
+analysis by providing a stable platform for subsequent programs to develop
+upon.
 
@@
 \CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a limiting factor:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the difficulty in escaping separators, providing
-multiple values for a field (e.g.\ multiple recipients) and adding new
-fields.  There is no facility to document the fields or the separator,
-unlike \SQL{} databases where every database includes a schema naming the
-fields and the type of data they store (integer, text, timestamp, etc.).
+support both \CSV{} and Berkeley DB appears to have been a serious
+limitation: \CSV{} is a very simple format where a record is stored in a
+single line, with fields separated by a comma or other punctuation symbol.
+Problems with \CSV{} files include the difficulty in escaping separators,
+providing multiple values for a field (e.g.\ multiple recipients) and
+adding new fields.  There is no mechanism to document the fields or the
+separator, unlike \SQL{} databases where every database includes a schema
+naming the fields and the type of data they store (integer, text,
+timestamp, etc.).
 
 Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
-unlike \SQL{} databases which store arbitrary tuples; in \LMA{}'s main
+unlike \SQL{} databases that store arbitrary tuples; in \LMA{}'s main
 table the key is an integer referred to by secondary tables, and the value
@@
 sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  In effect this re-implements
+referring to the main table is the value.  This effectively re-implements
 \SQL{} foreign keys, but without the functionality offered by even the most
@@
     \item Some records have an \SMTP{} code of \textit{0}, though this is
-        not a valid \SMTP{} code; presumably this is due to \LMA{} not
+        not a valid \SMTP{} code; presumably this is because of \LMA{} not
         parsing a line properly.  \LMA{} considers this code to be a
@@
     \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely due
-        to mis-parsing by \LMA{}); it is treated as a rejection by \LMA{},
-        so the equivalent \texttt{awk} command must do so, too.
+        \textit{deferred\/} (again, not a valid code, and most likely
+        because of mis-parsing by \LMA{}); it is treated as a rejection by
+        \LMA{}, so the equivalent \texttt{awk} command must do so, too.
 
@@
 
-There are a number of differences in the output from DAILY-REJECT and the
+There are two differences in the output from DAILY-REJECT and the
 \texttt{awk} command:
@@
         produces only one output line, whereas the \texttt{awk} command
-        produces all matching lines.
+        above produces all matching lines.
 
@@
 Pflogsumm produces a report designed for troubleshooting, rather than for
-in-depth analysis.  It does not support saving any data, nor does it
-extract more data than is required for producing the report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not attempt
-to correlate log lines by queueid or \pid{}, nor does it need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully dealt with the \numberOFlogFILES{} log
-files files tested with,\footnote{The results it reported were not
-verified, but it did not report any errors, and has a very good reputation
-amongst Postfix users.} but is not a suitable base for this project.
+in-depth analysis.  It does not support saving any data or extracting more
+data than is required for producing the report.  Both the parsing and
+reporting are difficult to extend as it is a specialised tool, unlike the
+easily extensible design of \parsername{}.  It does not correlate log lines
+by queueid or \pid{}, and does not need to deal with the complications
+encountered during this project.  Pflogsumm produces a useful report, and
+successfully dealt with the \numberOFlogFILES{} log files tested
+with,\footnote{The results it reported were not verified, but it did not
+report any errors, and has a very good reputation amongst Postfix users.}
+but is not a suitable base for this project.
 
@@
 
-There are many options available to include or exclude certain sections of
-the report.
+Many options are available to include or exclude certain sections of the
+report.
 
@@
 
-Sawmill is a general purpose commercial product which parses 687 log file
+Sawmill is a general purpose commercial product that parses 687 log file
 formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are quite limited, though it does extract three
-different sets of data for Postfix (one is beta as of 2007/12/04), but they
-do not appear to be interlinked, nor does it save sufficient data for the
-purposes of this project.\footnote{The data extracted by Sawmill is
+extraction facilities are too limited to save sufficient data for the
+purposes of this project,\footnote{The data extracted by Sawmill is
 described later in this review, and the data extracted by \PLP{} is
 described in \sectionref{connections table} and \sectionref{results
-table}.}  No attempt is made to correlate log lines or deal with the
-difficulties documented in \sectionref{complications}
-\sectionref{additional complications}.\footnote{If any attempt is made
-there is no reference to it in the documentation available on the website.}
-The source code is available in an obfuscated form only (presumably for a
-fee), and the product is quite expensive, as it requires a license per
-report which is to be generated; in contrast \parsername{} is free and the
-code is freely available.  The web interface allows creation of dynamic
-reports based on any field, but due to the associated cost the author has
-not experimented with it.  Presumably, as the company charges per report
-the user wishes to generate, the data store (if there is one) is
-deliberately inaccessible and undocumented to prevent the user bypassing
-the program and generating their own reports.
+table}.} though it does extract three different sets of data for Postfix
+(one is beta as of 2007/12/04), but they do not appear to be interlinked.
+No attempt is made to correlate log lines or deal with the difficulties
+documented in \sectionref{complications} \sectionref{additional
+complications}.\footnote{If any attempt is made there is no reference to it
+in the documentation available on the website.} The source code is
+available in an obfuscated form only (presumably for a fee), and the
+product is quite expensive, as it requires a license per report to be
+generated; in contrast \parsername{} is free and the code is freely
+available.  The web interface allows creation of dynamic reports based on
+any field, but because of the associated cost the author has not
+experimented with it.  Presumably, as the company charges per report the
+user wishes to generate, the data store (if there is one) is deliberately
+inaccessible and undocumented to prevent the user bypassing the program and
+generating their own reports.
 
@@
 informative as the interface provided by Splunk.  The administrative
-interface is very easy to use --- it took only five minutes to start
-parsing a directory of log files.
+interface is easy to use --- it took only five minutes to start parsing a
+directory of log files.
 
@@
 have been due to limitations in the free version.} can be run periodically
-and the results emailed to a recipient or sent to a shell script, which
+and the results emailed to a recipient or sent to a shell script, that
 presumably can publish the results as required (though possibly without the
@@
 the author was unable to find any: every category was empty, even those
-which the interface claimed had reports available.  Many types of reports
+that the interface claimed had reports available.  Many types of reports
 are bundled with the software, though most are variations of a bar or pie
-chart, with the exception of bubble and heatmap graphs.  It is very easy to
-drill down through the graphs to extract a portion of the data (e.g.\
-select the hour with the largest number of events, then select a particular
-host, and finally a specific address), though it is not possible to search
-on partial words.
+chart, except bubble and heatmap graphs.  It is easy to drill down through
+the graphs to extract a portion of the data (e.g.\ select the hour with the
+largest number of events, then select a particular host, and finally a
+specific address), though it is not possible to search on partial words.
 
-The web interface is quite attractive and easy to use when searching, but
+The web interface is quite attractive and simple to use when searching, but
 as an administrator it seems unnecessarily difficult to perform simple
@@
 \numberOFlogFILES{} log files to the queue for indexing.  Repeated attempts
-enqueued the same 40 log files, without explanation as to why the others
+enqueued the same 40 log files, without explaining why the others
 were excluded.\footnote{The log files appear to have only been indexed
@@
 the user.  The free version of Splunk limits the volume of data indexed per
-day to 500MB, though a trial Enterprise licence is available which allows
+day to 500MB, though a trial Enterprise licence is available that allows
 indexing of up to 5GB of data per day.  The cheapest licensed version costs
@@
 
-In the specific case of parsing Postfix log files, Splunk extracts some
-standard fields: to and from addresses, HELO hostname, date, host the log
-files were collected from, and protocol.  It parses the standard syslog
-fields at the beginning of the line, and extracts any \texttt{key=<value>}
-pairs occurring after the standard syslog prologue; these pairs are the
-fields listed above.  Searches can be based on the extracted fields, and
-all the text in the line is also available for searching.  \parsername{}
-extracts noticeably more data, though it does not make the full text of the
-line available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
+When parsing Postfix log files Splunk extracts some standard fields: to and
+from addresses, HELO hostname, date, host the log files were collected
+from, and protocol.  It parses the standard syslog fields at the beginning
+of the line, and extracts any \texttt{key=<value>} pairs occurring after
+the standard syslog prologue; these pairs are the fields listed above.
+Searches can be based on the extracted fields, and all the text in the line
+is also available for searching.  \parsername{} extracts noticeably more
+data, though it does not make the full text of the line available, and the
+full power of \SQL{} is available when searching, allowing the user to
+search on arbitrarily complicated conditions.
 
@@
 above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, nor to handle any
-of the myriad complications discussed in this document.  Some additional
+no attempt to correlate log lines by queueid or \pid{}, or to handle any of
+the myriad complications discussed in this document.  Some additional
 Postfix reports are supposedly available at
@@
 generated by Postfix, silently skipping those it cannot deal with, and does
-not distinguish between different types of rejection; there is no easy way
-to extend it to handle new log lines.  Although though it does correlate
-log lines by queueid, it does not deal with any of the other complications
-described in this document.  AWStats supports saving data but the format of
-the saved data is not documented (as far as the author could tell).  It
-also supports reading compressed log files, but that functionality was not
-tested.
+not distinguish between different types of rejection; it would be a lot of
+work to extend it to handle new log lines.  Although though it does
+correlate log lines by queueid, it does not deal with any of the other
+complications described in this document.  AWStats supports saving data but
+the format of the saved data is not documented (as far as the author could
+tell).  It also supports reading compressed log files, but that
+functionality was not tested.
 
@@
 
-YAALA provides a Postfix parser which extracts the following fields from
+YAALA provides a Postfix parser that extracts the following fields from
 specific log lines:
@@
 
-In summary YAALA provides a Postfix parser which only attempts to parse the
+In summary YAALA provides a Postfix parser that only attempts to parse the
 most common Postfix log lines (but fails miserably), provides reasonably
@@
 but accessing this data from another application is undocumented; given the
-source code it should be possible, with sufficient time and effort, to
-access the data from an external program.
+source code it should be possible, with enough time and effort, to access
+the data from an external program.
 
@@
 completely ignored, making it unsuitable for the purposes of this project.
-It does not attempt to address any of the complications in
+It does not try to address any of the complications in
 \sectionref{complications} \sectionref{additional complications} except for
@@
 (minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it wasn't possible to review the
-reports Logrep can generate (available in \HTML{} only), nor to examine the
+attempts were made.}  As parsing failed it was not possible to review the
+reports Logrep can generate (available in \HTML{} only), or to examine the
 (undocumented) format in which it can save extracted data for subsequent
@@
 
-While there are other programs available which perform basic Postfix log
-parsing (some to a greater level of detail than others), few attempt to
+There are other programs available which perform basic Postfix log parsing
+(some to a greater level of detail than others), but few attempt to
 correlate log lines by queueid (none correlate by \pid{}) to produce an
@@
 the reviewed parsers collect the breadth of information gathered by
-\parsername{}, nor make it as easy to extend the parser to handle new log
+\parsername{}, or make it as easy to extend the parser to handle new log
 lines.  Most other parsers immediately generate a report and discard the
-data extracted from the log files; those which do not discard the data
+data extracted from the log files; those that do not discard the data
 retain it in a format inaccessible to other tools.  Nearly all of the
 parsers reviewed can produce a report of greater or lesser detail and
-complexity, a facility not offered by \parsername{}; reporting is deferred
-to a subsequent application.  The quality of the documentation offered by
-the subset of parsers which provide some varies from unusable to good; none
-of the parsers provide any documentation on the format of their data stores
-(if any).  Less than half of the parsers were capable of parsing the
-\numberOFlogFILES{} test log files, and improving or extending the parsers
-would have been quite a difficult task for any of the parsers.  Table
-\refwithpage{Summary of parsers' features} provides a summary of the
+complexity, unlike \parsername{}.  The quality of the documentation offered
+by the subset of parsers that provide some varies from unusable to good;
+none of the parsers provide any documentation on the format of their data
+stores (if any).  Fewer than half of the parsers were capable of parsing
+the \numberOFlogFILES{} test log files, and improving or extending the
+parsers would have been quite a difficult task for any of the parsers.
+Table \refwithpage{Summary of parsers' features} provides a summary of the
 parsers' features.

### 2008-08-25 doc/logparser-results.tex
Dealt with some XXX items: * Mentioned prior art. * Added a little more about detecting overlapping regexes. * Removed "Why are there dips in the graphs?" paragraphs, they weren't   particularly important any more. * Changed normal to optimal.
Changes made following suggestions from diction.

@@
 XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
+
+XXX COMPARE PARSING + ACTIONS TO JUST PARSING\@.
 
@@
 Parsing efficiency is an obvious concern when the parser routinely needs to
-parse very large log files.  The server which generated the log files used
+parse large log files.  The server that generated the log files used
 in testing this parser accepts approximately 10,000 mails for 700 users a
@@
 copy to the remote address; this is generally not a problem except that the
-remote address in this case is the address the mail was originally sent to,
-creating an infinite loop.  To prevent this happening Postfix examines the
-Delivered-To header in the mail, and if the mail has already been delivered
-to the current address it is bounced back to the sender with the error
-message \newline{} \tab{} \texttt{mail forwarding loop for
-username@domain}\newline{}  Ordinarily this
-works well, but unfortunately in this case the user noticed they had not
-received any mail in a while and opted to send a test mail to themselves,
-causing a loop not caught by Postfix:
+remote address in this instance is the address the mail was originally sent
+to, creating an infinite loop.  To prevent this happening Postfix examines
+the Delivered-To header in the mail, and if the mail has already been
+delivered to the current address it is bounced back to the sender with the
+error message \newline{} \tab{} \texttt{mail forwarding loop for
+username@domain}\newline{}  Ordinarily this works well, but unfortunately
+in this instance the user noticed they had not received any mail in a while
+and opted to send a test mail to themselves, causing a loop not caught by
+Postfix:
 
@@
 
-    \item Postfix accepts the mail for username@domain, but while
-        delivering it notices that the \texttt{Delivered-To} header already
-        contains the address it is currently delivering to, and therefore
-        sends a bounce notification with sender address \texttt{<>} to
-        the original sender: username@domain.
+    \item Postfix accepts the mail for username@domain, but during delivery
+        it notices that the \texttt{Delivered-To} header already contains
+        the address it is currently delivering to, and therefore sends a
+        bounce notification with sender address \texttt{<>} to the original
+        sender: username@domain.
 
@@
 
-    \item Postfix accepts the forwarded bounce notification but while
-        delivering the mail it notices that the \texttt{Delivered-To}
+    \item Postfix accepts the forwarded bounce notification but during
+        deliver the mail it notices that the \texttt{Delivered-To}
         header already contains the address currently being delivered to,
@@
 The different distribution of log lines during a mail forwarding loop has
-interesting effects on parsing time, described in detail in
+unexpected effects on parsing time, described in detail in
 \sectionref{rule ordering for efficiency}, \sectionref{scalability as the
@@
 consistent over time, parser efficiency should benefit from trying more
-frequently matching rules before those which match less frequently.  To
+frequently matching rules before those that match less frequently.  To
 test this hypothesis three full test runs were performed with different
@@
     \item [optimal]  The most optimal order, according to the hypothesis:
-        rules which match most often will be tried first.
+        rules that match most often will be tried first.
 
@@
 When a mail loop occurs the vast majority of log lines are from Postfix
-components which have a small number of rules associated with them, whereas
-in general the majority of log lines are from Postfix components which have
-a high number of associated rules; this results in the average number of
-rules required to parse a log line from log files 22 or 62--68 being
-considerably lower than usual.  The ratio of number of lines versus parsing
-time in \graphref{parsing time vs file size vs number lines factor} shows a
+components that have few rules associated with them, whereas in general the
+most log lines are from Postfix components that have a high number of
+associated rules; this results in the average number of rules required to
+parse a log line from log files 22 or 62--68 being considerably lower than
+usual.  The ratio of number of lines versus parsing time in
+\graphref{parsing time vs file size vs number lines factor} shows a
 noticeable improvement at log files 22 and 62--68, where a higher ratio
 means more log lines parsed per unit of time; similarly \graphref{Parsing
-time relative to shuffled ordering graph} shows much less difference
+time relative to shuffled ordering graph} shows a much smaller difference
 between the orderings at log files 22 and 62--68.
@@
 
-    \item Check if there are any rules for the Postfix component which
+    \item Check if there are any rules for the Postfix component that
         produced the log line; if not then skip the log line.
@@
         intermingled in the log file, and there are some Postfix programs
-        which do not produce any log lines of interest.
+        that do not produce any log lines of interest.
 
@@
         will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule which
+        bound as possible to ensure accurate parsing:\footnote{A rule that
         matches zero or more of any character will successfully parse every
@@
 
-Warnings are issued for any log lines which are not parsed; no warnings are
+Warnings are issued for any log lines that are not parsed; no warnings are
 issued for unparsed log lines while testing with the \numberOFlogFILES{}
@@
 \numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} which was used by the parser.}  The sample was
+compared to the \regex{} that was used by the parser.}  The sample was
 generated by running the following command:
@@
 even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines which only appear in
+exercised (some of the rules are for parsing log lines that only appear in
 other log files).
@@
 readable form; examining these tables with reference to the log files is
-the best way to detect mails which were not handled properly (many of the
+the best way to detect mails that were not handled properly (many of the
 complications discussed in \sectionref{additional complications} were
@@
 should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails which had yet to be
+state table should only contain entries for mails that had yet to be
 delivered when the log files ended, or were accepted before the log files
@@
 The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails which
+parser finishes execution.  Ideally all they should contain are mails that
 are awaiting delivery after the period covered by the log files, though
@@
 
-    \item 1 connection which started only seconds before the log files
+    \item 1 connection that started only seconds before the log files
         ended and had not yet been fully transferred from client to server.
 
-    \item 1 mail which had been accepted only seconds before the log files
+    \item 1 mail that had been accepted only seconds before the log files
         ended and had not yet been delivered.
@@
 
-    \item 7 mails which had yet to be delivered due to repeated failures.
+    \item 7 mails that had yet to be delivered due to repeated failures.
 
@@
         log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions which each reassembled mail must comply
+        lines} imposes conditions that each reassembled mail must comply
         with to be acceptable.
@@
 verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
+and executed the correct action, that in turn produced the correct result
 and inserted the correct data in the database.  The log segment produced 4

### 2008-08-25 doc/logparser.tex
Part two of the great renaming complete. Graphs still need to be renamed.
Renamed actions because Carl can't understand that the IGNORE action doesn't mean that the log line wasn't parsed, so other people might have the same problem:     IGNORE              => UNINTERESTING     REJECTION           => DELIVERY_REJECTED     SAVE_BY_QUEUEID     => SAVE_DATA
Removed some unused labels cluttering the paper. Minor cleanup.
Final suggestions from Carl. Remaining corrections from my proofreading. Resolved all XXX items. Improved bibliography entries.
Changes made following suggestions from diction.

@@
 }
+
+%\setlength{\floatsep}{0pt}
+%\setlength{\intextsep}{0pt}
+%\setlength{\textfloatsep}{0pt}
 
@@
     explaining the difficulties encountered and the solutions developed,
-    with reference to an implementation which stores data gleaned from the
+    with reference to an implementation that stores data gleaned from the
     log files in an SQL database.  The gathered data can then be used to
@@
 XXX NEED A NAME FOR THE ARCHITECTURE, AND A DIFFERENT NAME FOR THE
-IMPLEMENTATION\@.
+IMPLEMENTATION\@.  Parse::FAR\@???
+
+XXX CHECK ALL ``UNINTERESTING'' RULES TO SEE IF THEY COULD BE USED\@.
 
@@
 \input{logparser-implementation.tex}
-\input{logparser-parsing-rules.tex}
 \input{logparser-results.tex}

### 2008-08-27 doc/sgai-2008.tex
Corrections/improvements from proofreading.

@@
     previously unknown class of input).  The architecture scales linearly
-    as the number of rules and/or size of input increases, making it
-    suitable for parsing large corpora or months of accumulated data.
+    or better as the number of rules and/or size of input increases, making
+    it suitable for parsing large corpora or months of accumulated data.
 }
@@
 of non-spam mail that is mistakenly rejected will receive an immediate
-non-delivery notice; load is reduced on the accepting mail server (allowing
-more intensive content-based techniques to be used); users have less mail
-to wade through.  Improving the performance of anti-spam techniques that
-are applied when mail is being transferred via \SMTP{}\footnotemark{} is
-the goal of this project, by providing a platform for reasoning about
-anti-spam filters.  The approach chosen to measure performance is to
-analyse the log files produced by the \SMTP{} server in use,
-Postfix~\cite{postfix}, rather than modifying it to generate statistics:
-this approach improves the chances of other sites testing and utilising the
-software.  The need arose for a parser capable of dealing with the great
-number and variety of log lines produced by Postfix: the parser must be
-designed so that adding support for parsing new inputs is a simple task, as
-the log lines to be parsed will change over time.  The variety in log lines
-occurs for several reasons:\footnotetext{Simple Mail Transfer Protocol
-transfers mail across the Internet from the sender to one or more
-recipients.  It is a simple, human readable, plain text protocol, making it
-quite easy to test and debug problems with it.  The original protocol
-definition is RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821}.}
+non-delivery notice; resource usage is reduced on the accepting mail server
+(allowing more intensive content-based techniques to be used); users have
+less spam mail to wade through.  Improving the performance of anti-spam
+techniques that are applied when mail is being transferred via
+\SMTP{}\footnotemark{} is the goal of this project, by providing a platform
+for reasoning about anti-spam filters.  The approach chosen to measure
+performance is to analyse the log files produced by the \SMTP{} server in
+use, Postfix~\cite{postfix}, rather than modifying it to generate
+statistics: this approach improves the chances of other sites testing and
+utilising the software.  The need arose for a parser capable of dealing
+with the great number and variety of log lines produced by Postfix: the
+parser must be designed so that adding support for parsing new inputs is a
+simple task, as the log lines to be parsed will change over time.  The
+variety in log lines occurs for several reasons:\footnotetext{Simple Mail
+Transfer Protocol transfers mail across the Internet from the sender to one
+or more recipients.  It is a simple, human readable, plain text protocol,
+making it quite easy to test and debug problems with it.  The original
+protocol definition is RFC~821~\cite{RFC821}, updated in
+RFC~2821~\cite{RFC2821}.}
 
@@
     \item External resources utilised by Postfix (e.g.\ \DNSBL{} or policy
-        servers~\cite{policy-servers}) can also change their messages,
-        sometimes without warning.
+        servers~\cite{policy-servers}) can change their messages without
+        warning.
 
@@
 
-It was hoped to reuse an existing parser rather than starting from scratch:
-several parsers were considered and rejected because: they parsed too small
-a fraction of the log files; their parsing was too inexact; and/or they did
-not extract sufficient data.  The effort required to adapt and improve an
-existing parser was judged to be greater than the effort to write one, as
-the techniques used by the existing parsers severely limited their
-potential: some ignored the majority of log lines, parsing specific log
-lines with great accuracy, but without any provision for readily parsing
-new or similar log lines; others sloppily parsed the majority of log lines,
-but were incapable of distinguishing between log lines of the same type,
-e.g.\ rejections, and extracted only a small portion of the data required.
-The only prior published work on the subject of parsing Postfix log files
-that we are aware of is~\cite{log-mail-analyser}, which aims to present
-correlated log data in a form suitable for a systems administrator to
-search using the myriad of standard Unix text processing utilities already
-available.  A full state of the art review is outside the scope of this
-paper but will be included in the thesis resulting from this work; the
-relationship to another procedural framework (ATN) is discussed below.
+It was hoped to reuse an existing parser rather than writing one from
+scratch, but the existing parsers considered were rejected for one or more
+reasons: they parsed too small a fraction of the log files; their parsing
+was too inexact; they did not extract sufficient data.  The effort required
+to adapt and improve an existing parser was judged to be greater than the
+effort to write a new one, as the techniques used by the existing parsers
+severely limited their potential: some ignored the majority of log lines,
+parsing specific log lines with great accuracy, but without any provision
+for readily parsing new or similar log lines; others sloppily parsed the
+majority of log lines, but were incapable of distinguishing between log
+lines of the same category, e.g.\ rejections.  The only prior published
+work on the subject of parsing Postfix log files that the authors are aware
+of is~\cite{log-mail-analyser}, which aims to present correlated log data
+in a form suitable for a systems administrator to search using the myriad
+of standard Unix text processing utilities already available.  A full state
+of the art review is outside the scope of this paper but will be included
+in the thesis resulting from this work; the relationship to another
+procedural framework (ATN) is discussed below.
 
@@
 
-\begin{description}
+\begin{description}[Framework]
 
@@
 
-    \item [Rules]  The rules are responsible for classifying or matching
+    \item [Rules]  The rules are responsible for classifying
         inputs, specifying the action to invoke and the regex that matches
@@
 
-Some similarity exists between the architecture and William Wood's
+Some similarity exists between this architecture and William Wood's
 \ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
 creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the architecture (shown in \tableref{Similarities with ATN}) is
-accidental, but it is obvious that the two approaches share a similar
-division of responsibilities, although having different semantics.
+the two (shown in \tableref{Similarities with ATN}) is accidental, but it
+is obvious that the two approaches share a similar division of
+responsibilities, despite having different semantics.
 
@@
 details of parsing, freeing the programmers writing actions to concentrate
-on writing productive code.  It provides the link between actions and
-rules, allowing either to be improved independently of the other.  It
-provides shared storage to pass data between actions, loads and validates
-rules, manages parsing, invokes actions, tracks how often each rule matches
-to optimise rule ordering (\sectionref{Rule ordering}), and stores results
-in the database.  The framework is largely independent of the parser being
-written: the same functionality will be required for most or all parsers,
-though some specialised support functions may be provided.  The framework
-is the core of the architecture and is deliberately quite simple: the rules
-deal with the variation in inputs, and the actions deal with the
-intricacies and complications encountered when parsing.
+on writing productive code.  It links actions and rules, allowing either to
+be improved independently of the other.  It provides shared storage to pass
+data between actions, loads and validates rules, manages parsing, invokes
+actions, tracks how often each rule matches to optimise rule ordering
+(\sectionref{Rule ordering}), and stores results in the database.  Most
+parsers will require the same basic functionality from the framework, plus
+some specialised support functions.  The framework is the core of the
+architecture and is deliberately quite simple: the rules deal with the
+variation in inputs, and the actions deal with the intricacies and
+complications encountered when parsing.
 
@@
 
-Each action contains the code required to deal with a particular category
-of input, e.g.\ rejections.  The actions are parser-specific: each parser
-author will need to write the required actions from scratch, unless
+Each action is a subroutine written to deal with a particular category of
+input, e.g.\ rejections.  The actions are parser-specific: each parser
+author will need to write the required actions from scratch unless
 extending an existing parser.  It is anticipated that parsers utilising
@@
 \showgraph{build/graph-action-distribution}{Distribution of rules per
-action}{Distribution of rules per action} Unsurprisingly the action with
+action}{Distribution of rules per action} Unsurprisingly, the action with
 the most associated rules is \texttt{DELIVERY\_REJECTED}, the action that
@@
 \texttt{SAVE\_DATA}, the action responsible for handling informative log
-lines, that supplement the data gathered from other log lines but do not by
-themselves represent events.  The third most common action is, perhaps
-surprisingly, \texttt{UNINTERESTING}: this action does nothing when
-executed, allowing uninteresting log lines to be parsed without causing any
-effects (it does not imply that the input is ungrammatical or unparsed).
-Generally rules specifying the \texttt{UNINTERESTING} action are rules to
-parse log lines that are not associated with a specific mail, e.g.\ notices
-about configuration files changing.  The remaining actions only have one or
-two associated rules: in some cases this is because there will only ever be
-one log line variant for that action, e.g.\ all log lines showing that a
-remote client has connected are matched by a single rule and handled by the
-\texttt{CONNECT} action; in other cases the action is required to address a
-deficiency in the log files, or a complication that arises during parsing.
-Taking the \texttt{CONNECT} action as an example, it creates a new data
-structure in memory for each client connecting, and saves the data
-extracted by the rule; this data will subsequently be entered into the
-database.  If a data structure already exists for that connection it is
-treated as a symptom of a bug, and the action issues a warning containing
-the full contents of the existing data structure, plus the log line that
-has just been parsed.
+lines, supplementing the data gathered from other log lines.  The third
+most common action is, perhaps surprisingly, \texttt{UNINTERESTING}: this
+action does nothing when executed, allowing uninteresting log lines to be
+parsed without causing any effects (it does not imply that the input is
+ungrammatical or unparsed).  Generally rules specifying the
+\texttt{UNINTERESTING} action are rules to parse log lines that are not
+associated with a specific mail, e.g.\ notices about configuration files
+changing.  The remaining actions have only one or two associated rules:
+some actions are required to address a deficiency in the log files, or a
+complication that arises during parsing;  other actions will only ever have
+one log line variant, e.g.\ all log lines showing that a remote client has
+connected are matched by a single rule and handled by the \texttt{CONNECT}
+action.
+
+Taking the \texttt{CONNECT} action as an example: it creates a new data
+structure in memory for the client connecting, and saves the data extracted
+by the rule; this data will subsequently be entered into the database.  A
+data structure already existing for that connection is treated as a symptom
+of a bug, and the action issues a warning containing the full contents of
+the existing data structure, plus the log line that has just been parsed.
 
@@
 e.g.\ to set or check a flag.  There is a possibility of introducing
-failure when modifying existing actions, but the modifications will be
-smaller and occur less frequently than with a monolithic architecture, thus
-failures will be less likely and will be easier to test for and diagnose.
-The architecture can be implemented in an object oriented style, allowing
-sub-classes to extend or override actions in addition to adding new
-actions; because each action is an independent procedure, the sub-class
-need only modify the action it is overriding, rather than reproducing large
-chunks of functionality.
+failure when modifying existing actions in this way, but the modifications
+will be smaller and occur less frequently than with a monolithic
+architecture, thus failures will be less likely and will be easier to test
+for and diagnose.  The architecture can be implemented in an object
+oriented style, allowing sub-classes to extend or override actions in
+addition to adding new actions; because each action is an independent
+procedure, the sub-class need only modify the action it is overriding,
+rather than reproducing large chunks of functionality.
 
@@
 to be overcome.  Some were the result of deficiencies in Postfix's logging
-(some of which were improved on or resolved by later versions of Postfix);
-others were due to the vagaries of process scheduling, client behaviour,
-and administrative actions.  All were successfully accommodated in the
-Postfix log parser: adding new actions was enough to overcome several of
-the complications; others required modifications to a single existing
-action to work around the difficulties; the remainder were resolved by
-adapting existing actions to cooperate and exchange extra data, changing
-their behaviour as appropriate.
+(sometimes improved on or resolved by later versions of Postfix); others
+were due to the vagaries of process scheduling, client behaviour, and
+administrative actions.  All were successfully accommodated in the Postfix
+log parser: adding new actions was enough to overcome several of the
+complications; others required modifications to a single existing action to
+work around the difficulties; the remainder were resolved by adapting
+existing actions to cooperate and exchange extra data, changing their
+behaviour as appropriate.
 
@@
 attributes; each implementation is free to add any additional attributes it
-requires.  The Postfix log parser adds several attributes to optimise rule
-ordering (\sectionref{Rule ordering}), to restrict which log lines each
-rule can be matched against, and to describe each rule.  Data is
+requires.  The Postfix log parser adds several attributes: to optimise rule
+ordering (\sectionref{Rule ordering}); to restrict which log lines each
+rule can be matched against; and to describe each rule.  Data is
 automatically extracted and saved from the log line based on keywords in
@@
 on Simple Logfile Clustering Tool~\cite{slct-paper} to aid in producing
-regexes from previously unparsed log lines.  Other approaches to rule
-management can be used, e.g.\ instead of manually adding new rules, machine
-learning techniques could be used to automatically generate new rules,
-based on the existing rules and input lines.  If this approach was taken
-the choice of machine learning technique would be constrained by the size
-of typical data sets (see \sectionref{Results}).  Techniques that require
-the full data set when training would be impractical: Instance Based
+regexes from previously unparsed log lines.  Separating the rules from the
+actions and framework enables other approaches to rule management to be
+used, e.g.\ instead of manually adding new rules, machine learning
+techniques could be used to automatically generate new rules based on the
+existing rules and unparsed inputs.  If this approach was taken the choice
+of machine learning technique would be constrained by the size of typical
+data sets (see \sectionref{Results}).  Techniques requiring the full data
+set when training would be impractical; Instance Based
 Learning~\cite{instance-based-learning} techniques that automatically
@@
 written to detect overlapping rules.  Traditional regexes are equivalent in
-computational power to \FA{}, and it is possible to convert between the
-two, so regex overlap can be detected by finding a non-empty intersection
-of two \FA{}\@.  The standard equation for \FA{} intersection (given for
-example in~\cite{intersection-of-NFA-using-Z}) is: $FA1 \cap{} FA2 =
+computational power \FA{}, and can be converted to \FA{}, so regex overlap
+can be detected by finding a non-empty intersection of two \FA{}\@.  The
+standard equation for \FA{} intersection (given for example
+in~\cite{intersection-of-NFA-using-Z}) is: $FA1 \cap{} FA2 =
 \overline{(\overline{FA1} \cup{} \overline{FA2})}$.  Perl 5.10 regexes are
@@
 intersection of two \CFL{} is not closed, i.e.\ the resulting language
-cannot in all cases be parsed by a \CFL{}, so intersection may be
+cannot always be parsed by a \CFL{}, so intersection may be
 intractable in some cases e.g.:
@@
 
-Detecting overlap between $n$ regexes requires $n(n-1)/2$ intersections to
-be calculated, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
+Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
+intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
 complexity of calculating intersection.  This is certainly not a task to be
@@
 
-It is possible to define pathological regexes that fall into two main
+It is possible to define pathological regexes which fall into two main
 categories: regexes that match every input, and regexes that consume
@@
 matches every input is trivial: \verb!/^/! matches the start of every
-input.  In general a regex will not consume much CPU time when the match is
-successful; usually CPU time is consumed when a regex with a lot of
-alteration and variable quantifiers fails to match.  This topic is beyond
-the scope of this paper, see~\cite{mastering-regular-expressions} for
-in-depth discussion.
+input.  Usually excessive CPU time is consumed when a regex with a lot of
+alteration and variable quantifiers fails to match; successful matching is
+generally quite fast.  This topic is beyond the scope of this paper,
+see~\cite{mastering-regular-expressions} for in-depth discussion.
 
@@
 
-    \item [Context-free rules:]  Rules are context-free: they do not take
-        into account past or future inputs.  In context-free grammar terms
-        the parser rules could be described as:
-        $\text{\textless{}input\textgreater{}} \mapsto \text{rule-1} |
-        \text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
+    \item [Context-free rules:]  Rules do not take into account past or
+        future inputs.  In context-free grammar terms the parser rules
+        could be described as:
+        \newline{}$\text{\textless{}input\textgreater{}} \mapsto
+        \text{rule-1} | \text{rule-2} | \text{rule-3} | \dots |
+        \text{rule-n}$
 
@@
         Actions can not inspect past or future inputs, but actions can
-        return a modified input that will be parsed as if read from the
-        input stream, allowing for a simplified version of cascaded
+        return a modified input to be parsed as if read from the input
+        stream, allowing for a simplified version of cascaded
         parsing~\cite{cascaded-parsing}.
@@
         transduction: it takes data in one form (log files) and transforms
-        it to another form (a database); other output formats may be more
-        suitable for other implementations.
+        it to another form (a database); other formats may be more suitable
+        for other implementations.
 
@@
         there is no facility for rules to consume more input or push unused
-        input back onto the input stream.  This was not a deliberate
-        decision, rather a consequence of the line oriented nature of the
+        input back onto the input stream.  This was not a deliberate design
+        decision, rather a consequence of the line oriented nature of
         Postfix log files; more flexible approaches could be pursued.
@@
 the log files used in testing the Postfix log parser accepts approximately
-10,000 mails for 700 users per day.  Median log file size is 50MB,
+10,000 mails for 700 users per day; median log file size is 50MB,
 containing 285,000 log lines --- large scale mail servers would have much
@@
 \numberOFlogFILES{} log files in one run took
-\input{build/include-timing-run-duration}.
+\input{build/include-timing-run-duration}, averaging
+\input{build/include-timing-run-throughput.tex}.
 
@@
         \tabletopline{}%
-        Component  & Component used                                 \\
+        Component  & Component in use                               \\
         \tablemiddleline{}%
@@
 input size: linearly, polynomially, or exponentially?  \Graphref{Parsing
-time, file size, and number of log lines} shows the parsing time in
-seconds, file size in MB, and number of log lines in tens of thousands, for
-each of the \numberOFlogFILES{} log files.  All three lines on the graph
-run roughly in parallel, giving the impression that the algorithm scales
-linearly with input size.  This impression is borne out by
-\graphref{parsing time vs file size vs number lines factor}: the ratios are
-tightly banded across the graph, showing that the algorithm scales
+time, log file size, and number of log lines} shows the parsing time in
+seconds, log file size in MB, and number of log lines in tens of thousands,
+for each of the \numberOFlogFILES{} log files.  All three lines on the
+graph run roughly in parallel, giving the impression that the algorithm
+scales linearly with input size.  This impression is borne out by
+\graphref{parsing time vs log file size vs number lines factor}: the ratios
+are tightly banded across the graph, showing that the algorithm scales
 linearly.  The ratios increase (i.e.\ improve) for log files 22 and 62--68
-despite their large size; that unusually large size is due to a mail
-forwarding loop resulting in a greatly increased number of mails delivered
+despite their large size; that unusually large size is due to mail
+forwarding loops resulting in a greatly increased number of mails delivered
 and log lines generated.
 
-\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
-and number of log lines}{Parsing time, file size, and number of log lines}
+\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, log file size,
+and number of log lines}{Parsing time, log file size, and number of log lines}
 \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of number of
-log lines and file size to parsing time}{parsing time vs file size vs
+log lines and log file size to parsing time}{parsing time vs log file size vs
 number lines factor}
@@
 
-\begin{description}[optimal]
+\begin{description}[shuffled]
 
@@
 
-    \item [shuffle] Random ordering, intended to represent an unsorted rule
+    \item [shuffled] Random ordering, intended to represent an unsorted rule
         set.  Note that the ordering will change every time the parser is
@@
 \Graphref{Parsing time relative to shuffled} shows the parsing times of
-optimal and reversed orderings as a percentage of the parsing time of
+optimal and reverse orderings as a percentage of the parsing time of
 shuffled ordering.  Overall this optimisation provides a modest but
@@
 is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
-ruleset.  A second set of statistics was generated using the minimum
+ruleset.  A second set of test runs was performed using the minimum
 ruleset, and the parsing times compared to those generated using the full
 ruleset: the percentage parsing time increase when using the full ruleset
-instead of the minimal ruleset for optimal, shuffled and reversed orderings
+instead of the minimal ruleset for optimal, shuffled and reverse orderings
 is shown in \graphref{Percentage parsing time increase of maximum ruleset
-over minimum ruleset}.  Clearly that the increased number of rules has a
+over minimum ruleset}.  Clearly the increased number of rules has a
 noticeable performance impact with reverse ordering, and a lesser impact
@@
 former is less intrusive, as usually it just requires new rules to be
-written; improving the latter is more intrusive as it requires changes to
-actions, and it can also be much harder to notice that a deficiency exists.
+written; improving the latter is more intrusive as it requires adding or
+changing actions, and it can also be much harder to notice that a
+deficiency exists.
 
@@
 readable form; examining these tables with reference to the log files and
-database is the best way to detect mails that were not handled properly.
-The implementation issues warnings when it detects any errors or
-discrepancies, which should alert the user to a problem.  There should be
-few or no warnings during parsing, and when parsing is finished the state
-table should only contain entries for mails starting before or ending after
-the log file.  A second sample of 6000 log lines was parsed with all
-debugging options enabled, resulting in 167,448 lines of output.  All
-167,448 lines were examined in conjunction with the log segment and a dump
-of the resulting database, verifying that for each of the log lines the
-implementation performed correctly.  The implementation produced 4 warnings
-about deficiencies in the log segment, 10 mails correctly remaining in the
-state tables, and 1625 correct entries in the database.  No error messages
-were produced, therefore there were no false negatives.  Given the evidence
-detailed above, the authors are confident that the false positive rate when
-understanding a mail delivery attempt is zero.
+database is the best way to detect mail delivery attempts that were not
+understood properly.  The Postfix log parser issues warnings when it
+detects any errors or discrepancies, which should alert the user to a
+problem.  There should be few or no warnings during parsing, and when
+parsing is finished the state table should only contain entries for mails
+starting before or ending after the log file.  A second sample of 6000 log
+lines was parsed with all debugging options enabled, resulting in 167,448
+lines of output.  All 167,448 lines were examined in conjunction with the
+log segment and a dump of the resulting database, verifying that for each
+of the log lines the implementation performed correctly.  The
+implementation produced 4 warnings about deficiencies in the log segment,
+10 mails correctly remaining in the state tables, and 1625 correct entries
+in the database.  No error or warning messages were produced, therefore
+there were no false negatives.  Given the evidence detailed above, the
+authors are confident that the false positive rate when understanding a
+mail delivery attempt is zero.
 
@@
 coverage is relatively easy to achieve with this architecture, and that
-with sufficient time and effort full understanding of the input is
-possible.  Postfix log files would require substantial time and effort to
-correctly parse regardless of the architecture used; this architecture
-enables an iterative approach to be used (similar to Stepwise
+with enough time and effort full understanding of the input is possible.
+Postfix log files would require substantial time and effort to correctly
+parse regardless of the architecture used; this architecture enables an
+iterative approach to be used (similar to Stepwise
 Refinement~\cite{stepwise-refinement}), as is practiced in many other
@@
 parser will then parse version one, or version two, or both versions,
-depending on the ruleset selected.  The architecture makes it possible to
-apply commonly used programming techniques such as object orientation,
-inheritance, composition, delegation, and roles when designing and
-implementing a parser, simplifying the process of working within a team
-with shared responsibility for a parser, or when developing and testing
-additional functionality.  This architecture is ideally suited to parsing
-inputs where the input is not fully understood or does not follow a fixed
-grammar: the architecture warns about unparsed inputs and other errors, but
-continues on parsing as best it can, allowing the developer of a new parser
-to decide which deficiencies are most important and require attention
-first, rather than being forced to fix the first error that arises.
+depending on the ruleset selected.  Decoupling also allows other approaches
+to rule management, as discussed in \sectionref{Rules}.  The architecture
+makes it possible to apply commonly used programming techniques such as
+object orientation, inheritance, composition, delegation, and roles, when
+designing and implementing a parser, simplifying the process of working
+within a team or when developing and testing additional functionality.
+This architecture is ideally suited to parsing inputs where the input is
+not fully understood or does not follow a fixed grammar: the architecture
+warns about unparsed inputs and other errors, but continues on parsing as
+best it can, allowing the developer of a new parser to decide which
+deficiencies are most important and require attention first, rather than
+being forced to fix the first error that arises.
 
-The dataset gathered by the Postfix log parser provides the foundation for
-the future of this project: applying machine-learning algorithms to the
-data to analyse and optimise the set of rules in use, followed by
+The data gathered by the Postfix log parser provides the foundation for the
+future of this project: applying machine-learning algorithms to the data to
+analyse and optimise the set of anti-spam measures in use, followed by
 identifying patterns in the data that could be used to write new filters to

### 2008-08-28 doc/sgai-2008.tex
Corrections from proofreading.

@@
     of existing inputs) and relatively easily add new actions (to process a
-    previously unknown class of input).  The architecture scales linearly
-    or better as the number of rules and/or size of input increases, making
-    it suitable for parsing large corpora or months of accumulated data.
+    previously unknown category of input).  The architecture scales
+    linearly or better as the number of rules and size of input increases,
+    making it suitable for parsing large corpora or months of accumulated
+    data.
 }
@@
 based on that analysis, and developing supplemental filters where
-deficiencies are identified.  Most anti-spam techniques are
-content-based~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}
-which requires accepting the mail in the first place to determine if it is
-spam, but rejecting mail during the delivery attempt is preferable: senders
-of non-spam mail that is mistakenly rejected will receive an immediate
-non-delivery notice; resource usage is reduced on the accepting mail server
-(allowing more intensive content-based techniques to be used); users have
-less spam mail to wade through.  Improving the performance of anti-spam
-techniques that are applied when mail is being transferred via
-\SMTP{}\footnotemark{} is the goal of this project, by providing a platform
-for reasoning about anti-spam filters.  The approach chosen to measure
-performance is to analyse the log files produced by the \SMTP{} server in
-use, Postfix~\cite{postfix}, rather than modifying it to generate
-statistics: this approach improves the chances of other sites testing and
-utilising the software.  The need arose for a parser capable of dealing
-with the great number and variety of log lines produced by Postfix: the
-parser must be designed so that adding support for parsing new inputs is a
-simple task, as the log lines to be parsed will change over time.  The
+deficiencies are identified.  Most anti-spam techniques are content-based
+(e.g.~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}) requiring
+that the mail be accepted before determining if it is spam, but rejecting
+mail during the delivery attempt is preferable: senders of non-spam mail
+that is mistakenly rejected will receive an immediate non-delivery notice;
+resource usage is reduced on the accepting mail server (allowing more
+intensive content-based techniques to be used on the mail that is
+accepted); users have less spam mail to wade through.  Improving the
+performance of anti-spam techniques that are applied when mail is being
+transferred via \SMTP{}\footnotemark{} is the goal of this project, by
+providing a platform for reasoning about anti-spam filters.  The approach
+chosen to measure performance is to analyse the log files produced by the
+\SMTP{} server in use, Postfix~\cite{postfix}, rather than modifying it to
+generate statistics: this approach improves the chances of other sites
+testing and utilising the software.  The need arose for a parser capable of
+dealing with the great number and variety of log lines produced by Postfix:
+the parser must be designed so that adding support for parsing new inputs
+is a simple task, as the log lines to be parsed will change over time.  The
 variety in log lines occurs for several reasons:\footnotetext{Simple Mail
@@
 
-    \item Log lines differ between versions of Postfix.
+    \item Log lines differ amongst versions of Postfix.
 
@@
 scratch, but the existing parsers considered were rejected for one or more
-reasons: they parsed too small a fraction of the log files; their parsing
-was too inexact; they did not extract sufficient data.  The effort required
-to adapt and improve an existing parser was judged to be greater than the
-effort to write a new one, as the techniques used by the existing parsers
-severely limited their potential: some ignored the majority of log lines,
-parsing specific log lines with great accuracy, but without any provision
-for readily parsing new or similar log lines; others sloppily parsed the
-majority of log lines, but were incapable of distinguishing between log
-lines of the same category, e.g.\ rejections.  The only prior published
-work on the subject of parsing Postfix log files that the authors are aware
-of is~\cite{log-mail-analyser}, which aims to present correlated log data
-in a form suitable for a systems administrator to search using the myriad
-of standard Unix text processing utilities already available.  A full state
-of the art review is outside the scope of this paper but will be included
-in the thesis resulting from this work; the relationship to another
-procedural framework (ATN) is discussed below.
+of the following reasons: they parsed too small a fraction of the log
+files; their parsing was too inexact; they did not extract sufficient data.
+The effort required to adapt and improve an existing parser was judged to
+be greater than the effort to write a new one, because the techniques used
+by the existing parsers severely limited their potential: some ignored the
+majority of log lines, parsing specific log lines with great accuracy, but
+without any provision for readily parsing new or similar log lines; others
+sloppily parsed the majority of log lines, but were incapable of
+distinguishing between log lines of the same category, e.g.\ rejections.
+The only prior published work on the subject of parsing Postfix log files
+that the authors are aware of is~\cite{log-mail-analyser}, which aims to
+extract data from the log files, correlate it, and present it in a form
+suitable for a systems administrator to search using the myriad of standard
+Unix text processing utilities already available.  A full state of the art
+review is outside the scope of this paper but will be included in the
+thesis resulting from this work; the relationship to another procedural
+framework (ATN) is discussed below.
 
@@
 functions (\textit{actions\/}), each capable of dealing with an entire
-category of log lines (e.g.\ rejecting a mail), accompanied by a multitude
-of precise patterns (\textit{rules\/}), each of which matches all log lines
-of a specific type and only that type (e.g.\ rejection by a specific
+category of inputs (e.g.\ rejecting a mail), accompanied by a multitude of
+precise patterns (\textit{rules\/}), each of which matches all log lines of
+a specific type and only that type (e.g.\ rejection by a specific
 \DNSBL{}).  It is an accepted standard to separate the parsing procedure
@@
 
-Decoupling the parsing rules from their associated actions and the
-framework allows new rules to be written and tested without requiring
-modifications to the parser source code, significantly lowering the barrier
-to entry for casual users who need to parse new inputs, e.g.\ part-time
-systems administrators attempting to combat and reduce spam.  Decoupling
+Decoupling the parsing rules from their associated actions allows new rules
+to be written and tested without requiring modifications to the parser
+source code, significantly lowering the barrier to entry for casual users
+who need to parse new inputs, e.g.\ part-time systems administrators
+attempting to combat and reduce spam; it also allows companies to develop
+user-extensible parsers without divulging their source code.  Decoupling
 the actions from the framework simplifies both framework and actions: the
@@
 low level details of identifying inputs and extracting data; actions handle
-the higher level details of assembling the required data, dealing with the
+the higher level tasks of assembling the required data, dealing with the
 intricacies of the input being parsed, complications arising, etc.;\ the
@@
 Some similarity exists between this architecture and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-the two (shown in \tableref{Similarities with ATN}) is accidental, but it
-is obvious that the two approaches share a similar division of
+\ATN{}~\cite{atns, nlpip}, used in Computational Linguistics for creating
+grammars to parse or generate sentences.  The resemblance between the two
+(shown in \tableref{Similarities with ATN}) is accidental, but it is
+obvious that the two approaches share a similar division of
 responsibilities, despite having different semantics.
@@
                       &                     & actions that constitutes a valid input.   \\
-        Transitions   & Actions             & Saves data and imposes conditions the     \\
+        Transitions   & Actions             & Assembles data and imposes conditions the \\
                       &                     & input must meet to be accepted as valid.  \\
@@
 
-Each action is a subroutine written to deal with a particular category of
+Each action is a separate procedure written to deal with a particular category of
 input, e.g.\ rejections.  The actions are parser-specific: each parser
@@
 requirement arises an independent action can be written to satisfy it.
-Sometime the new action will require the cooperation of other actions,
+Sometimes the new action will require the cooperation of other actions,
 e.g.\ to set or check a flag.  There is a possibility of introducing
@@
 During development of the Postfix log parser it became apparent that in
-addition to the obvious variety in log lines, there were many complications
+addition to the obvious variety in log lines there were many complications
 to be overcome.  Some were the result of deficiencies in Postfix's logging
-(sometimes improved on or resolved by later versions of Postfix); others
+(some were improved on or resolved by later versions of Postfix); others
 were due to the vagaries of process scheduling, client behaviour, and
@@
 are stored in, removing any doubt about which set of rules was used to
-produce a set of results; other implementations are free to store the rules
-in whatever fashion suits their needs.  The framework warns about every
-unparsed input, to alert the user that they need to alter or extend their
-ruleset; the Postfix log parser successfully parses every log line in the
-\numberOFlogFILESall{} log files it is currently being tested with.  The
-framework requires each rule to have \texttt{action} and \texttt{regex}
+produce a set of results; other implementations are free to store their
+rules in whatever fashion suits their needs.  The framework warns about
+every unparsed input, to alert the user that they need to alter or extend
+their ruleset; the Postfix log parser successfully parses every log line in
+the \numberOFlogFILESall{} log files it is currently being tested with.
+The framework requires each rule to have \texttt{action} and \texttt{regex}
 attributes; each implementation is free to add any additional attributes it
@@
 rule can be matched against; and to describe each rule.  Data is
-automatically extracted and saved from the log line based on keywords in
-the rule's regex, but there is also a mechanism to specify extra data to be
-saved.
+automatically extracted and saved from the input based on keywords in the
+rule's regex, but the Postfix log parser provides a mechanism to specify
+extra data to be saved.
 
@@
 on Simple Logfile Clustering Tool~\cite{slct-paper} to aid in producing
-regexes from previously unparsed log lines.  Separating the rules from the
+regexes from unparsed log lines.  Separating the rules from the
 actions and framework enables other approaches to rule management to be
@@
 techniques could be used to automatically generate new rules based on the
-existing rules and unparsed inputs.  If this approach was taken the choice
+existing rules and the unparsed inputs.  If this approach was taken the choice
 of machine learning technique would be constrained by the size of typical
@@
 separating \SMTP{} delivery to specific sites from \SMTP{} delivery to the
-rest of the world.  This can simplify both the general rule and the more
-specific rule; additionally rules from different sources can be combined
-with a minimum of prior cooperation or modification required.  Overlapping
-rules should have a priority attribute to specify their relative ordering;
-negative priorities may be useful for catchall rules.
+rest of the world.  Allowing overlapping rules simplifies both the general
+rule and the more specific rule; additionally rules from different sources
+can be combined with a minimum of prior cooperation or modification
+required.  Overlapping rules should have a priority attribute to specify
+their relative ordering; negative priorities may be useful for catchall
+rules.
 
-Storing the rules separately to the parser allows external tools to be
-written to detect overlapping rules.  Traditional regexes are equivalent in
-computational power \FA{}, and can be converted to \FA{}, so regex overlap
-can be detected by finding a non-empty intersection of two \FA{}\@.  The
-standard equation for \FA{} intersection (given for example
+Decoupling the rules from the actions allows external tools to be written
+to detect overlapping rules.  Traditional regexes are equivalent in
+computational power to \FA{}, and can be converted to \FA{}, so regex
+overlap can be detected by finding a non-empty intersection of two \FA{}\@.
+The standard equation for \FA{} intersection (given for example
 in~\cite{intersection-of-NFA-using-Z}) is: $FA1 \cap{} FA2 =
@@
 intersection of two \CFL{} is not closed, i.e.\ the resulting language
-cannot always be parsed by a \CFL{}, so intersection may be
-intractable in some cases e.g.:
+cannot always be parsed by a \CFL{}, so intersection may be intractable in
+some cases e.g.:
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
@@
 complexity of calculating intersection.  This is certainly not a task to be
-performed on every parsing run: detecting overlap amongst the Postfix log
+performed every time the parser is used: detecting overlap amongst the Postfix log
 parser's \numberOFrules{} rules would require calculating
@@
 excessive amounts of CPU time during matching.  Defining a regex that
-matches every input is trivial: \verb!/^/! matches the start of every
-input.  Usually excessive CPU time is consumed when a regex with a lot of
+matches all inputs is trivial: \verb!/^/! matches the start of every input.
+Usually excessive CPU time is consumed when a regex with a lot of
 alteration and variable quantifiers fails to match; successful matching is
@@
 
-    \item [Context-free rules:]  Rules do not take into account past or
+    \item [Matching rules against inputs is simple:]  The first matching
+        rule determines the action that will be invoked: there is no
+        backtracking to try alternate rules, no attempt is made to pick a
+        \textit{best\/} rule.
+
+    \item [Line oriented:]  The architecture is line oriented at present:
+        there is no facility for rules to consume more input or push unused
+        input back onto the input stream.  This was not a deliberate design
+        decision, rather a consequence of the line oriented nature of
+        Postfix log files; more flexible approaches could be pursued.
+
+    \item [Context-free rules:]  Rules can not take into account past or
         future inputs.  In context-free grammar terms the parser rules
@@
 
-    \item [Partially context-sensitive actions and cascaded parsing:]
-        Actions can consult the results (or lack of results) of previous
-        actions during execution, providing some context sensitivity.
-        Actions can not inspect past or future inputs, but actions can
-        return a modified input to be parsed as if read from the input
-        stream, allowing for a simplified version of cascaded
-        parsing~\cite{cascaded-parsing}.
+    \item [Context-aware actions:] Actions can consult the results (or lack
+        of results) of previous actions during execution, providing some
+        context sensitivity.
 
-    \item [Matching rules against inputs is simple:]  The first matching
-        rule determines the action that will be invoked: there is no
-        backtracking to try alternate rules, no attempt is made to pick a
-        \textit{best\/} rule.
+    \item [Cascaded parsing:] Actions can not inspect past or future
+        inputs, but actions can return a modified input to be parsed as if
+        read from the input stream, allowing for a simplified version of
+        cascaded parsing~\cite{cascaded-parsing}.
 
-    \item [Transduction:]  The parser can be thought of as implementing
-        transduction: it takes data in one form (log files) and transforms
-        it to another form (a database); other formats may be more suitable
-        for other implementations.
+    \item [Transduction:]  The architecture can be thought of as
+        implementing transduction: it takes data in one form (log files)
+        and transforms it to another form (a database); other formats may
+        be more suitable for other implementations.
 
@@
 
-    \item [Line oriented:]  The architecture is line oriented at present:
-        there is no facility for rules to consume more input or push unused
-        input back onto the input stream.  This was not a deliberate design
-        decision, rather a consequence of the line oriented nature of
-        Postfix log files; more flexible approaches could be pursued.
-
 \end{description}
@@
 the log files used in testing the Postfix log parser accepts approximately
-10,000 mails for 700 users per day; median log file size is 50MB,
+10,000 mails for 700 users per day; median log file size is 50 MB,
 containing 285,000 log lines --- large scale mail servers would have much
@@
 log lines) were each parsed 10 times and the parsing times averaged.
-Saving results to the database was disabled for the test runs, as the tests
-are aimed at measuring the speed of the Postfix log parser rather than the
-speed of the database.  The computer used for test runs is a Dell Optiplex
-745 described in \tableref{Computer used to generate statistics}; it was
-dedicated to the task of gathering statistics from test runs, and was not
-used for any other purpose while test runs were ongoing --- any services
-not necessary for running the tests were disabled.  Parsing all
+Saving results to the database was disabled for the test runs, because the
+tests are aimed at measuring the speed of the Postfix log parser rather
+than the speed of the database.  The computer used for test runs is a Dell
+Optiplex 745 described in \tableref{Computer used to generate statistics};
+it was dedicated to the task of gathering statistics from test runs, and
+was not used for any other purpose while test runs were ongoing --- any
+services not necessary for running the tests were disabled.  Parsing all
 \numberOFlogFILES{} log files in one run took
@@
 seconds, log file size in MB, and number of log lines in tens of thousands,
-for each of the \numberOFlogFILES{} log files.  All three lines on the
-graph run roughly in parallel, giving the impression that the algorithm
-scales linearly with input size.  This impression is borne out by
+for each of the \numberOFlogFILES{} log files.  The lines on the graph run
+roughly in parallel, giving the impression that the algorithm scales
+linearly with input size.  This impression is borne out by
 \graphref{parsing time vs log file size vs number lines factor}: the ratios
@@
 different rules: as shown in \graphref{rule hits graph} 10\% of the rules
-match the vast majority of log lines, with the remaining log lines split
+XXX PERCENTAGE match the vast majority of log lines, with the remaining log lines split
 across the other 90\% of the rules, similar to a Power Law distribution.
@@
 expected that typical parsers will have a large number of rules.  There are
-\numberOFrules{} rules in the full Postfix log parser ruleset, whereas the
+\numberOFrules{} rules in the full Postfix log parser ruleset (it is tested
+with \numberOFlogFILESall{} log files), whereas the
 minimum number of rules required to parse the \numberOFlogFILES{} log files
@@
 readable form; examining these tables with reference to the log files and
-database is the best way to detect mail delivery attempts that were not
-understood properly.  The Postfix log parser issues warnings when it
-detects any errors or discrepancies, which should alert the user to a
-problem.  There should be few or no warnings during parsing, and when
-parsing is finished the state table should only contain entries for mails
-starting before or ending after the log file.  A second sample of 6000 log
-lines was parsed with all debugging options enabled, resulting in 167,448
-lines of output.  All 167,448 lines were examined in conjunction with the
-log segment and a dump of the resulting database, verifying that for each
-of the log lines the implementation performed correctly.  The
-implementation produced 4 warnings about deficiencies in the log segment,
-10 mails correctly remaining in the state tables, and 1625 correct entries
-in the database.  No error or warning messages were produced, therefore
-there were no false negatives.  Given the evidence detailed above, the
-authors are confident that the false positive rate when understanding a
-mail delivery attempt is zero.
+database is the best way to detect misunderstood mail delivery attempts.
+The Postfix log parser issues warnings when it detects any errors or
+discrepancies, alerting the user to the problem.  There should be few or no
+warnings during parsing, and when parsing is finished the state table
+should only contain entries for mails starting before or ending after the
+log file.  A second sample of 6000 log lines was parsed with all debugging
+options enabled, resulting in 167,448 lines of output.  All 167,448 lines
+were examined in conjunction with the log segment and a dump of the
+resulting database, verifying that for each of the log lines the Postfix
+log parser performed correctly.  The implementation produced 4 warnings
+about deficiencies in the log segment, 10 mails correctly remaining in the
+state tables, and 1625 correct entries in the database: it produced 0 false
+positives.  No error or warning messages were produced, therefore there
+were no false negatives.  Given the evidence detailed above, the authors
+are confident that zero false positives or negatives occur when parsing the
+\numberOFlogFILES{} log files.
 
@@
 different sets of rules to be used with the same actions, e.g.\ a parser
-might have actions to process versions one and two of a file format; the
-parser will then parse version one, or version two, or both versions,
-depending on the ruleset selected.  Decoupling also allows other approaches
-to rule management, as discussed in \sectionref{Rules}.  The architecture
-makes it possible to apply commonly used programming techniques such as
-object orientation, inheritance, composition, delegation, and roles, when
+might have actions to process versions one and two of a file format; by
+choosing the appropriate ruleset the parser will parse version one, or
+version two, or both versions.  Decoupling also allows other approaches to
+rule management, as discussed in \sectionref{Rules}.  The architecture
+makes it possible to apply commonly used programming techniques (such as
+object orientation, inheritance, composition, delegation, and roles) when
 designing and implementing a parser, simplifying the process of working
@@
 the data in a normalised form that is far easier to use as input to new or
-existing implementations of machine-learning algorithms than trying to
-adapt each algorithm to extract data directly from the log files.  The
-current focus is on clustering and decision trees to optimise the order in
-which rules are applied; future efforts will involve using data gathered by
-the parser to train and test new filters.  This task is very similar to
-optimising a black box application based on its inputs and outputs, and
-this approach could be applied to optimising the behaviour of any system
-given sufficient log messages to analyse.  An alternate approach to black
-box optimisation that uses application profiling in conjunction with the
-application's error messages to improve the error messages shown to users
-is described in~\cite{black-box-error-reporting}; profiling data may be
-useful in supplementing systems that fail to provide adequate log messages.
+existing algorithm implementations than trying to adapt each algorithm to
+extract data directly from the log files.  The current focus is on
+clustering and decision trees to optimise the order in which rules are
+applied; future efforts will involve using data gathered by the parser to
+train and test new filters.  This task is very similar to optimising a
+black box application based on its inputs and outputs, and this approach
+could be applied to optimising the behaviour of any system given sufficient
+log messages to analyse.  An alternate approach to black box optimisation
+that uses application profiling in conjunction with the application's error
+messages to improve the error messages shown to users is described
+in~\cite{black-box-error-reporting}; profiling data may be useful in
+supplementing systems that fail to provide adequate log messages.
 

### 2008-08-29 doc/sgai-2008.tex
Proofreading.

@@
 \abstract{%
-    Some parsers need to be very precise and strict in what they parse, yet
-    must allow users to easily adapt or extend the parser to parse new
-    inputs, without requiring that the user have an in-depth knowledge and
+    Some parsers need to be very precise and strict when parsing, yet must
+    allow users to easily adapt or extend the parser to parse new inputs,
+    without requiring that the user have an in-depth knowledge and
     understanding of the parser's internal workings.  This paper presents a
@@
 The architecture described herein was developed as part of a larger project
-to improve anti-spam techniques by analysing the performance of the set of
+to improve anti-spam defenses by analysing the performance of the set of
 filters currently in use, optimising the order and membership of the set
@@
 deficiencies are identified.  Most anti-spam techniques are content-based
-(e.g.~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}) requiring
-that the mail be accepted before determining if it is spam, but rejecting
-mail during the delivery attempt is preferable: senders of non-spam mail
-that is mistakenly rejected will receive an immediate non-delivery notice;
-resource usage is reduced on the accepting mail server (allowing more
-intensive content-based techniques to be used on the mail that is
-accepted); users have less spam mail to wade through.  Improving the
-performance of anti-spam techniques that are applied when mail is being
+(e.g.~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}) and
+require the mail to be accepted before determining if it is spam, but
+rejecting mail during the delivery attempt is preferable: senders of
+non-spam mail that is mistakenly rejected will receive an immediate
+non-delivery notice; resource usage is reduced on the accepting mail server
+(allowing more intensive content-based techniques to be used on the mail
+that is accepted); users have less spam mail to wade through.  Improving
+the performance of anti-spam techniques that are applied when mail is being
 transferred via \SMTP{}\footnotemark{} is the goal of this project, by
@@
 generate statistics: this approach improves the chances of other sites
-testing and utilising the software.  The need arose for a parser capable of
+testing and using the software.  The need arose for a parser capable of
 dealing with the great number and variety of log lines produced by Postfix:
 the parser must be designed so that adding support for parsing new inputs
-is a simple task, as the log lines to be parsed will change over time.  The
-variety in log lines occurs for several reasons:\footnotetext{Simple Mail
-Transfer Protocol transfers mail across the Internet from the sender to one
-or more recipients.  It is a simple, human readable, plain text protocol,
-making it quite easy to test and debug problems with it.  The original
-protocol definition is RFC~821~\cite{RFC821}, updated in
+is a simple task, because the log lines to be parsed will change over time.
+The variety in log lines occurs for several reasons:\footnotetext{Simple
+Mail Transfer Protocol transfers mail across the Internet from the sender
+to one or more recipients.  It is a simple, human readable, plain text
+protocol, making it quite easy to test and debug problems with it.  The
+original protocol definition is RFC~821~\cite{RFC821}, updated in
 RFC~2821~\cite{RFC2821}.}
@@
 by the existing parsers severely limited their potential: some ignored the
-majority of log lines, parsing specific log lines with great accuracy, but
-without any provision for readily parsing new or similar log lines; others
-sloppily parsed the majority of log lines, but were incapable of
-distinguishing between log lines of the same category, e.g.\ rejections.
-The only prior published work on the subject of parsing Postfix log files
-that the authors are aware of is~\cite{log-mail-analyser}, which aims to
-extract data from the log files, correlate it, and present it in a form
-suitable for a systems administrator to search using the myriad of standard
-Unix text processing utilities already available.  A full state of the art
-review is outside the scope of this paper but will be included in the
-thesis resulting from this work; the relationship to another procedural
-framework (ATN) is discussed below.
+majority of log lines, parsing specific log lines accurately, but without
+any provision for parsing new or similar log lines; others sloppily parsed
+the majority of log lines, but were incapable of distinguishing between log
+lines of the same category, e.g.\ rejections.  The only prior published
+work on the subject of parsing Postfix log files that the authors are aware
+of is \textit{Log Mail Analyser: Architecture and Practical
+Utilizations\/}~\cite{log-mail-analyser}, which aims to extract data from
+log files, correlate it, and present it in a form suitable for a systems
+administrator to search using the myriad of standard Unix text processing
+utilities already available.  A full state of the art review is outside the
+scope of this paper but will be included in the thesis resulting from this
+work.
 
@@
 functions (\textit{actions\/}), each capable of dealing with an entire
-category of inputs (e.g.\ rejecting a mail), accompanied by a multitude of
-precise patterns (\textit{rules\/}), each of which matches all log lines of
+category of inputs (e.g.\ rejecting a mail delivery attempt), accompanied by a multitude of
+precise patterns (\textit{rules\/}), each of which matches all inputs of
 a specific type and only that type (e.g.\ rejection by a specific
@@
 
-Each action is a separate procedure written to deal with a particular category of
-input, e.g.\ rejections.  The actions are parser-specific: each parser
-author will need to write the required actions from scratch unless
-extending an existing parser.  It is anticipated that parsers utilising
-this architecture will have a high ratio of rules to actions, with the aim
-of having simpler rules and clearer distinctions between the inputs parsed
-by different rules.  In the Postfix log parser developed for this project
+Each action is a separate procedure written to deal with a particular
+category of input, e.g.\ rejections.  The actions are parser-specific: each
+parser author will need to write the required actions from scratch unless
+extending an existing parser.  It is anticipated that parsers based on this
+architecture will have a high ratio of rules to actions, with the aim of
+having simpler rules and clearer distinctions between the inputs parsed by
+different rules.  In the Postfix log parser developed for this project
 there are \numberOFactions{} actions and \numberOFrules{} rules, with an
@@
 ungrammatical or unparsed).  Generally rules specifying the
-\texttt{UNINTERESTING} action are rules to parse log lines that are not
-associated with a specific mail, e.g.\ notices about configuration files
-changing.  The remaining actions have only one or two associated rules:
-some actions are required to address a deficiency in the log files, or a
-complication that arises during parsing;  other actions will only ever have
-one log line variant, e.g.\ all log lines showing that a remote client has
-connected are matched by a single rule and handled by the \texttt{CONNECT}
-action.
+\texttt{UNINTERESTING} action parse log lines that are not associated with
+a specific mail, e.g.\ notices about configuration files changing.  The
+remaining actions have only one or two associated rules: some actions are
+required to address a deficiency in the log files, or a complication that
+arises during parsing;  other actions will only ever have one log line
+variant, e.g.\ all log lines showing that a remote client has connected are
+matched by a single rule and handled by the \texttt{CONNECT} action.
 
-Taking the \texttt{CONNECT} action as an example: it creates a new data
-structure in memory for the client connecting, and saves the data extracted
-by the rule; this data will subsequently be entered into the database.  A
-data structure already existing for that connection is treated as a symptom
-of a bug, and the action issues a warning containing the full contents of
-the existing data structure, plus the log line that has just been parsed.
+Using the \texttt{CONNECT} action as an example: it creates a new data
+structure in memory for the new client connection, saving the data
+extracted by the rule into it; this data will be entered into the database
+when the mail delivery attempt is complete.  If a data structure already
+exists for the new connection it is treated as a symptom of a bug, and the
+action issues a warning containing the full contents of the existing data
+structure, plus the log line that has just been parsed.
 
@@
 to be overcome.  Some were the result of deficiencies in Postfix's logging
-(some were improved on or resolved by later versions of Postfix); others
-were due to the vagaries of process scheduling, client behaviour, and
+(some of which were rectified by later versions of Postfix); others were
+due to the vagaries of process scheduling, client behaviour, and
 administrative actions.  All were successfully accommodated in the Postfix
@@
 from the input stream, allowing for a simplified version of cascaded
-parsing~\cite{cascaded-parsing}.  This is a powerful facility allowing
-several rules and actions to parse a single input.
+parsing~\cite{cascaded-parsing}.  This powerful facility allows several
+rules and actions to parse a single input, potentially simplifying both
+rules and actions.
 
@@
 attributes; each implementation is free to add any additional attributes it
-requires.  The Postfix log parser adds several attributes: to optimise rule
-ordering (\sectionref{Rule ordering}); to restrict which log lines each
-rule can be matched against; and to describe each rule.  Data is
+requires.  The Postfix log parser adds several attributes: optimising rule
+ordering (\sectionref{Rule ordering}); restricting which log lines each
+rule can be matched against; and describing each rule.  Data is
 automatically extracted and saved from the input based on keywords in the
-rule's regex, but the Postfix log parser provides a mechanism to specify
-extra data to be saved.
+rule's regex, but the Postfix log parser also provides a mechanism to
+specify extra data to be saved.
 
 Parsing new inputs is generally achieved by creating a new rule that pairs
-an existing action with a new regex.  Adding new rules is simple: define
-all the required attributes and append the new rule to wherever the
-existing rules are stored.  The Postfix log parser supplies a utility based
-on Simple Logfile Clustering Tool~\cite{slct-paper} to aid in producing
-regexes from unparsed log lines.  Separating the rules from the
-actions and framework enables other approaches to rule management to be
-used, e.g.\ instead of manually adding new rules, machine learning
-techniques could be used to automatically generate new rules based on the
-existing rules and the unparsed inputs.  If this approach was taken the choice
-of machine learning technique would be constrained by the size of typical
-data sets (see \sectionref{Results}).  Techniques requiring the full data
-set when training would be impractical; Instance Based
+an existing action with a new regex.  The Postfix log parser supplies a
+utility based on Simple Logfile Clustering Tool~\cite{slct-paper} to aid in
+producing regexes from unparsed log lines.  Separating the rules from the
+actions and framework enables other rule management approaches to be used,
+e.g.\ instead of manually adding new rules, machine learning techniques
+could be used to automatically generate new rules.  If this approach was
+taken the choice of machine learning technique would be constrained by the
+size of typical data sets (see \sectionref{Results}).  Techniques requiring
+the full data set when training would be impractical; Instance Based
 Learning~\cite{instance-based-learning} techniques that automatically
 determine which inputs from the training set are valuable and which inputs
-can be discarded might reduce the data required to a manageable size.  This
-avenue of research and development has not been pursued by the authors, but
-could easily be undertaken independently.
+can be discarded might reduce the data required to a manageable size.  A
+parser might also dynamically create new rules in response to certain
+inputs, e.g.\ diagnostic messages indicating the source of the inputs has
+read a new configuration file.  These avenues of research and development
+has not been pursued by the authors, but could easily be undertaken
+independently.
 
@@
 to detect overlapping rules.  Traditional regexes are equivalent in
-computational power to \FA{}, and can be converted to \FA{}, so regex
+computational power to \FA{} and can be converted to \FA{}, so regex
 overlap can be detected by finding a non-empty intersection of two \FA{}\@.
@@
 in~\cite{intersection-of-NFA-using-Z}) is: $FA1 \cap{} FA2 =
-\overline{(\overline{FA1} \cup{} \overline{FA2})}$.  Perl 5.10 regexes are
-more powerful than traditional regexes, e.g.\ it is possible to match
-correctly balanced brackets nested to an arbitrary depth:
-\newline{}\tab{}\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!\newline{}
-Perl 5.10 regexes can maintain an arbitrary state stack and are thus
-equivalent in computational power to \PDA{} or \CFL{}, so detecting overlap
-may require calculating the intersection of two \PDA{} or \CFL{}\@.  The
-intersection of two \CFL{} is not closed, i.e.\ the resulting language
-cannot always be parsed by a \CFL{}, so intersection may be intractable in
-some cases e.g.:
+\overline{(\overline{FA1} \cup{} \overline{FA2})}$, which has considerable
+computation complexity.  Perl 5.10 regexes are more powerful than
+traditional regexes: it is possible to match correctly balanced brackets
+nested to an arbitrary depth, e.g.\ \verb!z<123<pq<>rs>j<r>ml>s! is matched
+by \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! Perl 5.10
+regexes can maintain an arbitrary state stack and are thus equivalent in
+computational power to \PDA{} or \CFL{}, so detecting overlap may require
+calculating the intersection of two \PDA{} or \CFL{}\@.  The intersection
+of two \CFL{} is not closed, i.e.\ the resulting language cannot always be
+parsed by a \CFL{}, so intersection may be intractable in some cases e.g.:
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
@@
 alteration and variable quantifiers fails to match; successful matching is
-generally quite fast.  This topic is beyond the scope of this paper,
-see~\cite{mastering-regular-expressions} for in-depth discussion.
+generally quite fast.  See~\cite{mastering-regular-expressions} for
+in-depth discussion.
 
@@
         \tablemiddleline{}%
-        %program                   & \daemon{smtpd}                                   \\
         regex                     & \verb!^connect from (__CLIENT_HOSTNAME__)\[(__CLIENT_IP__)\]$!  \\
-        action                    & CONNECT                                          \\
-        %priority                  & 0                                                \\
+        action                    & CONNECT \hspace{3em}(described in \sectionref{Actions})         \\
         \tablebottomline{}%
@@
 than the speed of the database.  The computer used for test runs is a Dell
-Optiplex 745 described in \tableref{Computer used to generate statistics};
-it was dedicated to the task of gathering statistics from test runs, and
-was not used for any other purpose while test runs were ongoing --- any
-services not necessary for running the tests were disabled.  Parsing all
+Optiplex 745 described in \tableref{Computer used to generate statistics},
+dedicated to the task of gathering statistics from test runs.  Parsing all
 \numberOFlogFILES{} log files in one run took
@@
 \input{build/include-timing-run-throughput.tex}.
+
+% DO NOT REFORMAT!
 
@@
         RAM        & 2GB 667 MHz DDR RAM\@.                             \\
-        Hard disk  & 1 Seagate Barracuda 7200 RPM 250GB SATA disk.  \\
+        Hard disk  & 1 Seagate Barracuda 7200 RPM 250GB SATA hard disk. \\
         \tablebottomline{}%
@@
 \Graphref{Parsing time relative to shuffled} shows the parsing times of
-optimal and reverse orderings as a percentage of the parsing time of
-shuffled ordering.  Overall this optimisation provides a modest but
-worthwhile reduction in parsing time of approximately 10\%.  Optimal rule
-ordering is also beneficial in other ways, as described in
-\sectionref{Architecture Scalability: Number of Rules}.
-\showgraph{build/graph-hits}{Number of log lines matched by each rule}{rule
-hits graph} \showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing
-time relative to shuffled ordering}{Parsing time relative to shuffled}
+optimal and reverse orderings as a percentage of shuffled ordering parsing
+time.  This optimisation provides a modest but worthwhile reduction in
+parsing time of approximately 10\% with normal log files, less when a mail
+loop occurs and the distribution of log lines is unusual.  Optimal rule
+ordering has other benefits, described in \sectionref{Architecture
+Scalability: Number of Rules}.  \showgraph{build/graph-hits}{Number of log
+lines matched by each rule}{rule hits graph}
+\showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
+relative to shuffled ordering}{Parsing time relative to shuffled}
 
@@
 (the former is a requirement for the latter to be achieved).  Improving the
-former is less intrusive, as usually it just requires new rules to be
-written; improving the latter is more intrusive as it requires adding or
-changing actions, and it can also be much harder to notice that a
-deficiency exists.
+former is less difficult, as usually it just requires new rules to be
+written; improving the latter is more difficult and intrusive as it
+requires adding or changing actions, and it can be much harder to notice
+that a deficiency exists.
 
@@
 for any log lines that are not parsed; no such warnings are issued while
-parsing the \numberOFlogFILES{} log files, so it can be safely concluded
-that there are zero false negatives.  False positives are harder to
+parsing the \numberOFlogFILES{} log files, therefore
+there are zero false negatives.  False positives are harder to
 quantify: manually verifying that the correct rule parsed each of the
@@
 warnings during parsing, and when parsing is finished the state table
-should only contain entries for mails starting before or ending after the
-log file.  A second sample of 6000 log lines was parsed with all debugging
-options enabled, resulting in 167,448 lines of output.  All 167,448 lines
-were examined in conjunction with the log segment and a dump of the
-resulting database, verifying that for each of the log lines the Postfix
-log parser performed correctly.  The implementation produced 4 warnings
-about deficiencies in the log segment, 10 mails correctly remaining in the
-state tables, and 1625 correct entries in the database: it produced 0 false
-positives.  No error or warning messages were produced, therefore there
-were no false negatives.  Given the evidence detailed above, the authors
-are confident that zero false positives or negatives occur when parsing the
-\numberOFlogFILES{} log files.
+should only contain entries for mail delivery attempts starting before or
+ending after the log file.  A second sample of 6000 log lines was parsed
+with all debugging options enabled, resulting in 167,448 lines of output.
+All 167,448 lines were examined in conjunction with the log segment and a
+dump of the resulting database, verifying that for each of the log lines
+the Postfix log parser performed correctly.  The implementation produced 4
+warnings about deficiencies in the log segment, 10 mails correctly
+remaining in the state tables, and 1625 correct entries in the database: it
+produced 0 false positives.  No error or warning messages were produced,
+therefore there were no false negatives.  Given the evidence detailed
+above, the authors are confident that zero false positives or negatives
+occur when parsing the \numberOFlogFILES{} log files.
 
@@
 makes it possible to apply commonly used programming techniques (such as
-object orientation, inheritance, composition, delegation, and roles) when
-designing and implementing a parser, simplifying the process of working
-within a team or when developing and testing additional functionality.
-This architecture is ideally suited to parsing inputs where the input is
-not fully understood or does not follow a fixed grammar: the architecture
-warns about unparsed inputs and other errors, but continues on parsing as
-best it can, allowing the developer of a new parser to decide which
-deficiencies are most important and require attention first, rather than
-being forced to fix the first error that arises.
+object orientation, inheritance, composition, delegation, roles,
+modularisation, or closures) when designing and implementing a parser,
+simplifying the process of working within a team or when developing and
+testing additional functionality.  This architecture is ideally suited to
+parsing inputs where the input is not fully understood or does not follow a
+fixed grammar: the architecture warns about unparsed inputs and other
+errors, but continues on parsing as best it can, allowing the developer of
+a new parser to decide which deficiencies are most important and require
+attention first, rather than being forced to fix the first error that
+arises.
 
@@
 applied; future efforts will involve using data gathered by the parser to
-train and test new filters.  This task is very similar to optimising a
-black box application based on its inputs and outputs, and this approach
-could be applied to optimising the behaviour of any system given sufficient
-log messages to analyse.  An alternate approach to black box optimisation
-that uses application profiling in conjunction with the application's error
+train and test new filters.  This task is similar to analysing a black
+box application based on its inputs and outputs, and this approach could be
+applied to analyse the behaviour of any system given sufficient log
+messages to parse.  An alternate approach to black box optimisation that
+uses application profiling in conjunction with the application's error
 messages to improve the error messages shown to users is described

### 2008-08-30 doc/sgai-2008.tex
Proofreading.

@@
 The architecture described herein was developed as part of a larger project
-to improve anti-spam defenses by analysing the performance of the set of
+to improve anti-spam defences by analysing the performance of the set of
 filters currently in use, optimising the order and membership of the set
@@
 functions (\textit{actions\/}), each capable of dealing with an entire
-category of inputs (e.g.\ rejecting a mail delivery attempt), accompanied by a multitude of
-precise patterns (\textit{rules\/}), each of which matches all inputs of
-a specific type and only that type (e.g.\ rejection by a specific
-\DNSBL{}).  It is an accepted standard to separate the parsing procedure
-from the declarative grammar it operates with; part of the novelty here is
-in the way that the grammar is itself partially procedural (each action is
-a separate procedure).  This architecture is ideally suited to parsing
-inputs where the input is not fully understood or does not conform to a
-fixed grammar: the architecture warns about unparsed inputs and other
-errors, but continues on parsing as best it can, allowing the developer of
-a new parser to decide which deficiencies are most important and require
-attention first, rather than being forced to fix the first error that
-arises.
+category of inputs (e.g.\ rejecting a mail delivery attempt), accompanied
+by a multitude of precise patterns (\textit{rules\/}), each of which
+matches all inputs of a specific type and only that type (e.g.\ rejection
+by a specific \DNSBL{}).  It is an accepted standard to separate the
+parsing procedure from the declarative grammar it operates with; part of
+the novelty here is in the way that the grammar is itself partially
+procedural (each action is a separate procedure).  This architecture is
+ideally suited to parsing inputs where the input is not fully understood or
+does not conform to a fixed grammar: the architecture warns about unparsed
+inputs and other errors, but continues parsing as best it can, allowing the
+developer of a new parser to decide which deficiencies are most important
+and require attention first, rather than being forced to fix the first
+error that arises.
 
@@
     \item [Actions]  Each action performs the work required to deal with a
-        category of inputs, e.g.\ processing data from rejections.
+        single category of inputs, e.g.\ processing data from rejections.
 
@@
 attributes; each implementation is free to add any additional attributes it
-requires.  The Postfix log parser adds several attributes: optimising rule
-ordering (\sectionref{Rule ordering}); restricting which log lines each
-rule can be matched against; and describing each rule.  Data is
-automatically extracted and saved from the input based on keywords in the
-rule's regex, but the Postfix log parser also provides a mechanism to
+requires.  The Postfix log parser adds attributes for several reasons:
+optimising rule ordering (\sectionref{Rule ordering}); restricting which
+log lines each rule can be matched against; and to describe each rule.
+Keywords in the rule's regex specify the data to be extracted from the
+input, but the Postfix log parser also provides a mechanism for rules to
 specify extra data to be saved.
@@
 utility based on Simple Logfile Clustering Tool~\cite{slct-paper} to aid in
-producing regexes from unparsed log lines.  Separating the rules from the
+producing regexes from unparsed log lines.  Decoupling the rules from the
 actions and framework enables other rule management approaches to be used,
@@
 traditional regexes: it is possible to match correctly balanced brackets
-nested to an arbitrary depth, e.g.\ \verb!z<123<pq<>rs>j<r>ml>s! is matched
-by \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! Perl 5.10
-regexes can maintain an arbitrary state stack and are thus equivalent in
-computational power to \PDA{} or \CFL{}, so detecting overlap may require
-calculating the intersection of two \PDA{} or \CFL{}\@.  The intersection
-of two \CFL{} is not closed, i.e.\ the resulting language cannot always be
-parsed by a \CFL{}, so intersection may be intractable in some cases e.g.:
+nested to an arbitrary depth, e.g.\
+\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
+\verb!z<123<pq<>rs>j<r>ml>s!. Perl 5.10 regexes can maintain an arbitrary
+state stack and are thus equivalent in computational power to \PDA{} or
+\CFL{}, so detecting overlap may require calculating the intersection of
+two \PDA{} or \CFL{}\@.  The intersection of two \CFL{} is not closed,
+i.e.\ the resulting language cannot always be parsed by a \CFL{}, so
+intersection may be intractable in some cases e.g.:
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
-
 Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
@@
 categories: regexes that match every input, and regexes that consume
-excessive amounts of CPU time during matching.  Defining a regex that
-matches all inputs is trivial: \verb!/^/! matches the start of every input.
+excessive amounts of CPU time during matching.  Defining a regex to match
+all inputs is trivial: \verb!/^/! matches the start of every input.
 Usually excessive CPU time is consumed when a regex with a lot of
-alteration and variable quantifiers fails to match; successful matching is
-generally quite fast.  See~\cite{mastering-regular-expressions} for
-in-depth discussion.
+alteration and variable quantifiers fails to match, but successful
+matching is generally quite fast (see~\cite{mastering-regular-expressions}
+for in-depth discussion).
 
@@
         \text{rule-1} | \text{rule-2} | \text{rule-3} | \dots |
-        \text{rule-n}$
+        \text{rule-n}$.
 
@@
 
-    \item [Cascaded parsing:] Actions can not inspect past or future
-        inputs, but actions can return a modified input to be parsed as if
-        read from the input stream, allowing for a simplified version of
-        cascaded parsing~\cite{cascaded-parsing}.
+    \item [Cascaded parsing:] Actions can return a modified input to be
+        parsed as if read from the input stream, allowing for a simplified
+        version of cascaded parsing~\cite{cascaded-parsing}.
 
@@
 \input{build/include-timing-run-duration}, averaging
-\input{build/include-timing-run-throughput.tex}.
+\input{build/include-timing-run-throughput}; median throughput when parsing
+files separately was \input{build/include-median-throughput-MB}
+(\input{build/include-median-throughput-log-lines}) parsed per minute.
 
@@
 optimal and reverse orderings as a percentage of shuffled ordering parsing
-time.  This optimisation provides a modest but worthwhile reduction in
-parsing time of approximately 10\% with normal log files, less when a mail
-loop occurs and the distribution of log lines is unusual.  Optimal rule
-ordering has other benefits, described in \sectionref{Architecture
-Scalability: Number of Rules}.  \showgraph{build/graph-hits}{Number of log
-lines matched by each rule}{rule hits graph}
-\showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
-relative to shuffled ordering}{Parsing time relative to shuffled}
+time.  This optimisation provides a mean reduction in parsing time of
+\input{build/include-optimal-ordering-parsing-time-reduction-other-logs}\%
+with normal log files,
+\input{build/include-optimal-ordering-parsing-time-reduction-logs-22-62-68}\%
+when a mail loop occurs and the distribution of log lines is unusual.
+Optimal rule ordering has other benefits, described in
+\sectionref{Architecture Scalability: Number of Rules}.
+\showgraph{build/graph-hits}{Number of log lines matched by each rule}{rule
+hits graph} \showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing
+time relative to shuffled ordering}{Parsing time relative to shuffled}
 
@@
 expected that typical parsers will have a large number of rules.  There are
-\numberOFrules{} rules in the full Postfix log parser ruleset (it is tested
-with \numberOFlogFILESall{} log files), whereas the
-minimum number of rules required to parse the \numberOFlogFILES{} log files
-is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
-ruleset.  A second set of test runs was performed using the minimum
-ruleset, and the parsing times compared to those generated using the full
-ruleset: the percentage parsing time increase when using the full ruleset
-instead of the minimal ruleset for optimal, shuffled and reverse orderings
-is shown in \graphref{Percentage parsing time increase of maximum ruleset
-over minimum ruleset}.  Clearly the increased number of rules has a
-noticeable performance impact with reverse ordering, and a lesser impact
-with shuffled ordering.  The optimal ordering shows a mean increase of
+\numberOFrules{} rules in the full Postfix log parser ruleset (parsing
+\numberOFlogFILESall{} log files), but the minimum number of rules required
+to parse the \numberOFlogFILES{} log files is \numberOFrulesMINIMUM{},
+\numberOFrulesMINIMUMpercentage{} of the full ruleset.  A second set of
+test runs was performed using the minimum ruleset, and the parsing times
+compared to those generated using the full ruleset: the percentage parsing
+time increase when using the full ruleset instead of the minimal ruleset
+for optimal, shuffled and reverse orderings is shown in
+\graphref{Percentage parsing time increase of maximum ruleset over minimum
+ruleset}.  Clearly the increased number of rules has a noticeable
+performance impact with reverse ordering, and a lesser impact with shuffled
+ordering.  The optimal ordering shows a mean increase of
 \input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
 \numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  These
-results show that the architecture scales very well as the number of rules
-increases, and that optimally ordering the rules is an important
+results show that the architecture scales extremely well as the number of
+rules increases, and that optimally ordering the rules is an important
 optimisation enabling this.
@@
 for any log lines that are not parsed; no such warnings are issued while
-parsing the \numberOFlogFILES{} log files, therefore
-there are zero false negatives.  False positives are harder to
-quantify: manually verifying that the correct rule parsed each of the
-\numberOFlogLINES{} log lines is infeasible.  A random sample of 6,039 log
-lines (0.00994\% of \numberOFlogLINES{}) was parsed and the results
-manually verified by inspection to ensure that the correct rule parsed each
-log line.  The sample results contained zero false positives, and this
-check has been automated to ensure continued accuracy.  The authors are
-confident that zero false positives occur when parsing the
-\numberOFlogFILES{} log files.
+parsing the \numberOFlogFILES{} log files, therefore there are zero false
+negatives.  False positives are harder to quantify: manually verifying that
+the correct rule parsed each of the \numberOFlogLINES{} log lines is
+infeasible.  A random sample of 6,039 log lines (0.00994\% of
+\numberOFlogLINES{}) was parsed and the results manually verified by
+inspection to ensure that the correct rule's regex matched each log line.
+The sample results contained zero false positives, and this check has been
+automated to ensure continued accuracy.  The authors are confident that
+zero false positives occur when parsing the \numberOFlogFILES{} log files.
 
@@
 fixed grammar: the architecture warns about unparsed inputs and other
-errors, but continues on parsing as best it can, allowing the developer of
+errors, but continues parsing as best it can, allowing the developer of
 a new parser to decide which deficiencies are most important and require
@@
 future of this project: applying machine-learning algorithms to the data to
-analyse and optimise the set of anti-spam measures in use, followed by
+analyse and optimise the set of anti-spam defences in use, followed by
 identifying patterns in the data that could be used to write new filters to

### 2008-09-01 doc/sgai-2008.tex
Proofreading.

@@
 the majority of log lines, but were incapable of distinguishing between log
-lines of the same category, e.g.\ rejections.  The only prior published
+lines of the same category, e.g.\ rejecting a mail delivery attempt.  The only prior published
 work on the subject of parsing Postfix log files that the authors are aware
@@
 
-    \item [Rules]  The rules are responsible for classifying
-        inputs, specifying the action to invoke and the regex that matches
-        the inputs and extracts data.
+    \item [Rules]  The rules are responsible for classifying inputs,
+        specifying the action to invoke and the regex that matches the
+        inputs and extracts data.
 
@@
 will be smaller and occur less frequently than with a monolithic
-architecture, thus failures will be less likely and will be easier to test
+architecture, thus failures will be less likely to occur and will be easier to test
 for and diagnose.  The architecture can be implemented in an object
@@
 existing actions to cooperate and exchange extra data, changing their
-behaviour as appropriate.
+behaviour as appropriate based on that extra data.
 
@@
 \numberOFlogFILES{} log files in one run took
-\input{build/include-timing-run-duration}, averaging
+\input{build/include-timing-run-duration}, mean throughput is
 \input{build/include-timing-run-throughput}; median throughput when parsing
@@
 
-At the time of writing the Postfix log parser has \numberOFrules{}
-different rules: as shown in \graphref{rule hits graph} 10\% of the rules
-XXX PERCENTAGE match the vast majority of log lines, with the remaining log lines split
-across the other 90\% of the rules, similar to a Power Law distribution.
-Assuming that the distribution of log lines is reasonably consistent over
-time, parser efficiency should benefit from trying more frequently matching
-rules before those which match less frequently.  To test this hypothesis
-three full test runs were performed with different rule orderings:
+\Graphref{rule hits graph} shows the number of log lines in the
+\numberOFlogFILES{} log files matched by each of the Postfix log parser's
+\numberOFrules{} rules.  The top ten rules match
+\input{build/include-top-ten-hits}\% of the log lines, with the remainder
+tailing off similar to a Power Law distribution.  Assuming that the
+distribution of log lines is reasonably consistent over time, parser
+efficiency should benefit from trying more frequently matching rules before
+those which match less frequently.  To test this hypothesis three full test
+runs were performed with different rule orderings:
 
@@
 results show that the architecture scales extremely well as the number of
-rules increases, and that optimally ordering the rules is an important
-optimisation enabling this.
+rules increases, and that optimally ordering the rules enables this.
 \showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing

### 2008-09-15 doc/sgai-2008-elsevier.tex
Converted SGAI paper from Springer format to Elsevier format. Added database required to generate some of the included statistics.

--- 2008-09-01 doc/sgai-2008.tex
@@
 % $Id$
-\documentclass{svmult}
-%\documentclass[draft]{svmult}
+\documentclass[preprint,draft,numbers]{elsarticle}
 
@@
 
+% Use eqlist instead of description lists to get lists with consistent
+% indentation.
+\usepackage{eqlist}
+
 % Nicer citation lists - smaller spaces.
-\usepackage{cite}
+% Breaks, probably due to conflict with natbib.
+%\usepackage{cite}
 % Check for unused references.
@@
 
-\title*{A User-Extensible and Adaptable Parser Architecture}
-\author{John Tobin and Carl Vogel}
-\institute{John Tobin \email{tobinjt@cs.tcd.ie}
-            \and{} Carl Vogel \email{vogel@cs.tcd.ie}
-            \at{} School of Computer Science and Statistics,
-                    Trinity College, Dublin 2, Ireland.
-                    \newline{}
-                    Supported by Science Foundation Ireland RFP
-                    05/RF/CMS002.
-            }
+\begin{frontmatter}
 
-\maketitle
+\title{A User-Extensible and Adaptable Parser Architecture}
+\author{John Tobin}
+\ead{tobinjt@cs.tcd.ie}
+\author{Carl Vogel}
+\ead{vogel@cs.tcd.ie}
+\address{School of Computer Science and Statistics, Trinity College, Dublin
+2, Ireland.}
+\address{Supported by Science Foundation Ireland RFP 05/RF/CMS002.}
 
-\abstract{%
+\begin{keyword}
+    Parsing
+\end{keyword}
+
+\begin{abstract}
     Some parsers need to be very precise and strict when parsing, yet must
@@
     data.
-}
+\end{abstract}
+
+\end{frontmatter}
 
@@
 
-\begin{description}[Framework]
+\begin{eqlist}
 
@@
 
-\end{description}
+\end{eqlist}
 
@@
 Some similarity exists between this architecture and William Wood's
-\ATN{}~\cite{atns, nlpip}, used in Computational Linguistics for creating
+\ATN{}~\cite{atns}, used in Computational Linguistics for creating
 grammars to parse or generate sentences.  The resemblance between the two
@@
     \caption{Similarities with ATN}\label{Similarities with ATN}
-    % The \smallskip{} below stop the text hitting the lines.
     \begin{tabular}[]{lll}
@@
 read a new configuration file.  These avenues of research and development
-has not been pursued by the authors, but could easily be undertaken
+have not been pursued by the authors, but could easily be undertaken
 independently.
@@
         \tablemiddleline{}%
-        regex                     & \verb!^connect from (__CLIENT_HOSTNAME__)\[(__CLIENT_IP__)\]$!  \\
+        regex                     & {\footnotesize \verb!^connect from (__CLIENT_HOSTNAME__)\[(__CLIENT_IP__)\]$!}  \\
         action                    & CONNECT \hspace{3em}(described in \sectionref{Actions})         \\
@@
     \item [Closer to Natural Language Processing than using a fixed
-        grammar:] Unlike traditional parsers such as those used when
-        compiling a programming language, this architecture does not
-        require a fixed grammar specification that inputs must adhere to.
-        The architecture is capable of dealing with interleaved inputs, out
-        of order inputs, and ambiguous inputs where heuristics must be
-        applied --- all have arisen and been successfully accommodated in
-        the Postfix log parser.
+        grammar:] \hfill{}In contrast \newline{} to traditional parsers
+        such as those used when compiling a programming language, this
+        architecture does not require a fixed grammar specification that
+        inputs must adhere to.  The architecture is capable of dealing with
+        interleaved inputs, out of order inputs, and ambiguous inputs where
+        heuristics must be applied --- all have arisen and been
+        successfully accommodated in the Postfix log parser.
 
@@
         CPU        & 1 dual core 2.40GHz Intel\textregistered{}
-                     Core\texttrademark{}2 CPU,                      
-                     with 32KB L1 and 4MB L2 cache.                     \\
+                     Core\texttrademark{}2 CPU,                         \\
+                   & with 32KB L1 and 4MB L2 cache.                     \\
         RAM        & 2GB 667 MHz DDR RAM\@.                             \\
@@
 
-\begin{description}[shuffled]
+\begin{eqlist}
 
@@
 
-\bibliographystyle{spmpsci}
-\bibliography{logparser-bibliography.bib}
+\bibliographystyle{sgai-2008-elsevier}
+\bibliography{logparser-bibliography}
 

### 2008-09-26 doc/sgai-2008-elsevier.tex
Changes needed to make the paper look right.

@@
 % $Id$
-\documentclass[preprint,draft,numbers]{elsarticle}
+\documentclass[preprint,draft,numbers,1p]{elsarticle}
 
@@
 % Extra footnote functionality, including references to earlier footnotes.
-\usepackage[bottom]{footmisc}
+%\usepackage[bottom]{footmisc}
 
@@
 \begin{keyword}
-    Parsing
+    Parsing \sep{} Postfix \sep{} Anti-Spam \sep{} log files
 \end{keyword}
@@
 deficiencies are identified.  Most anti-spam techniques are content-based
-(e.g.~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}) and
+(e.g.~\cite{a-plan-for-spam,word-stemming,relaxed-online-svms}) and
 require the mail to be accepted before determining if it is spam, but
@@
 
-    \item External resources utilised by Postfix (e.g.\ \DNSBL{} or policy
-        servers~\cite{policy-servers}) can change their messages without
-        warning.
+    \item External resources Postfix is configured to use (e.g.\ \DNSBL{}
+        or policy servers~\cite{policy-servers}) can change their messages
+        without warning.
 
@@
 the majority of log lines, but were incapable of distinguishing between log
-lines of the same category, e.g.\ rejecting a mail delivery attempt.  The only prior published
-work on the subject of parsing Postfix log files that the authors are aware
-of is \textit{Log Mail Analyser: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}, which aims to extract data from
-log files, correlate it, and present it in a form suitable for a systems
-administrator to search using the myriad of standard Unix text processing
-utilities already available.  A full state of the art review is outside the
-scope of this paper but will be included in the thesis resulting from this
-work.
+lines of the same category, e.g.\ rejecting a mail delivery attempt.  The
+only prior published work on the subject of parsing Postfix log files that
+the authors are aware of is \textit{Log Mail Analyser: Architecture and
+Practical Utilizations\/}~\cite{log-mail-analyser}, which aims to extract
+data from log files, correlate it, and present it in a form suitable for a
+systems administrator to search using the myriad of standard Unix text
+processing utilities already available.  A full state of the art review is
+outside the scope of this paper but will be included in the thesis
+resulting from this work.
 
@@
 will be smaller and occur less frequently than with a monolithic
-architecture, thus failures will be less likely to occur and will be easier to test
-for and diagnose.  The architecture can be implemented in an object
+architecture, thus failures will be less likely to occur and will be easier
+to test for and diagnose.  The architecture can be implemented in an object
 oriented style, allowing sub-classes to extend or override actions in
@@
 complexity of calculating intersection.  This is certainly not a task to be
-performed every time the parser is used: detecting overlap amongst the Postfix log
-parser's \numberOFrules{} rules would require calculating
+performed every time the parser is used: detecting overlap amongst the
+Postfix log parser's \numberOFrules{} rules would require calculating
 \numberOFruleINTERSECTIONS{} intersections.
@@
     \item [Closer to Natural Language Processing than using a fixed
-        grammar:] \hfill{}In contrast \newline{} to traditional parsers
+        grammar:] \empty{}~\newline{} In contrast to traditional parsers
         such as those used when compiling a programming language, this
@@
 \numberOFlogFILES{} log files in one run took
-\input{build/include-timing-run-duration}, mean throughput is
+\input{build/include-timing-run-duration}, mean throughput
 \input{build/include-timing-run-throughput}; median throughput when parsing
@@
 
-\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, log file size,
-and number of log lines}{Parsing time, log file size, and number of log lines}
-\showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of number of
-log lines and log file size to parsing time}{parsing time vs log file size vs
-number lines factor}
+\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, log file
+size, and number of log lines}{Parsing time, log file size, and number of
+log lines} \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio
+of number of log lines and log file size to parsing time}{parsing time vs
+log file size vs number lines factor}
 
@@
 
-    \item [shuffled] Random ordering, intended to represent an unsorted rule
-        set.  Note that the ordering will change every time the parser is
-        executed, so 10 different orderings will be generated for each log
-        file in the test run.
+    \item [shuffled] Random ordering, intended to represent an unsorted
+        rule set.  Note that the ordering will change every time the parser
+        is executed, so 10 different orderings will be generated for each
+        log file in the test run.
 

### 2008-10-28 doc/logparser-acronyms.tex
First pass through corrections for sections 1-3.

@@
     SPF is introduced in \sectionref{spf introduction} and explained fully
-    in~\cite{openspf, Wikipedia-spf}.
+    in~\cite{openspf,Wikipedia-spf}.
 }}
@@
 \newacronym{FQDN}{Fully Qualified Domain Name}{description={
-    An FQDN is a hostname plus domain name, e.g.\ \newline
-        example.com     is a domain name          \newline
-        www             is a hostname             \newline
+    An FQDN is a hostname plus domain name, e.g.\ \newline{}
+        example.com     is a domain name          \newline{}
+        www             is a hostname             \newline{}
         www.example.com is a FQDN
@@
     patterns against text that are explained in detail
-    in~\cite{Wikipedia-regex, perlre}.
+    in~\cite{Wikipedia-regex,perlre}.
 }}
@@
     HTML is the markup language designed for writing web
-    pages~\cite{Wikipedia-html, w3schools-html}.
+    pages~\cite{Wikipedia-html,w3schools-html}.
 }}

### 2008-10-28 doc/logparser-background.tex
First pass through corrections for sections 1-3.

@@
 a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on the topics most relevant to this
-document, namely Postfix anti-spam restrictions and policy servers.  The
-assumptions made in designing and implementing the parser are explained, as
-are the conventions used in this document.
+introduction to Postfix, concentrating on anti-spam restrictions and policy
+servers.  The assumptions made in designing and implementing the parser are
+explained, as are the conventions used in this document.
 
@@
 This document and the program it describes are part of a larger project to
-optimise a mail server's Postfix restrictions, generate statistics and
-graphs, and provide a platform on which new restrictions can be trialled
-and evaluated to determine if they are beneficial in the fight against
-spam.  The program parses Postfix log files and populates a database with
-the data gleaned from those log files, providing a consistent and simple
-view of the log files that future tools can utilise.  The gathered data
-can then be used to optimise current anti-spam measures, provide a baseline
-to test new anti-spam measures against, or to produce statistics showing
-how effective those measures are.
+optimise a mail server's Postfix-based anti-spam restrictions, generate
+statistics and graphs, and provide a platform on which new restrictions on
+trial can be evaluated to determine if they are beneficial in the fight
+against spam.  The program parses Postfix log files and populates a
+database with the data gleaned from those log files, providing a consistent
+and simple view of the log files that future tools can utilise.  The
+gathered data can be used to optimise current anti-spam measures, to
+provide a baseline to test new anti-spam measures against, or to produce
+statistics showing how effective those measures are.
 
-A snippet of \SQL{} provides a short example of the optimisation possible
-using data from the database: determining which Postfix restrictions reject
-the highest number of mails:
+Determining which Postfix restrictions reject the highest number of mails
+is a short example of the optimisation possible using data from the
+database:
 
@@
 Another example is determining which restrictions are not effective: this
-shows which restrictions had fewer than 100 hits on the last log file
-parsed, and the percentage of total rejections each restriction represents.
+sample query shows which restrictions had fewer than 100 rejections in the
+last log file parsed, and the percentage of total rejections each of those
+restrictions represents.
 
@@
 
-These database queries yield summary statistics about the efficiency of
-spam avoidance techniques; statistics that are far less feasible to assess
+The database queries above yield summary statistics about the efficiency of
+anti-spam techniques.  These statistics are far less feasible to assess
 directly from log files without prior pre-processing into a database in the
-fashion proposed, implemented and tested herein.
+fashion proposed, implemented, and tested herein.
 
@@
 Postfix log files.  Although the interface is a database schema, it is in
-effect quite similar to any other \API{} provided by shared code: it
-insulates both user and provider of the \API{} from changes in the
+effect quite similar to any other \APIlong{}~(\API{}) provided by shared
+code: it insulates both user and provider of the \API{} from changes in the
 implementation of the \API{}\@.  The algorithm implemented by the parser
@@
 Postfix; bugs can be fixed or limitations removed from the parser; these
-changes will not cause the user to be negatively affected.  Statistics
-and graphs can be generated from the database; new restrictions can be
-tested and the results inspected; trends in the fight against spam can
-emerge from historical data saved in the database; the parser remains the
-same as the usage adapts.  Using a database simplifies writing programs
-that need to interact with the data in several ways:
+changes will not cause the user to be negatively affected.  Statistics and
+graphs can be generated from the database; new restrictions can be tested
+and the results inspected; trends in the fight against spam can emerge from
+historical data saved in the database; the parser remains the same as the
+usage adapts.  Using a database simplifies writing programs that need to
+interact with the data in several ways:
 
@@
 
-    \item It is already possible to access databases from most programming
+    \item Facilities are provided to access databases from most programming
         languages, allowing a developer to access the data gathered using
@@
         \begin{itemize}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
 
@@
 
+            \item Other tables can be added to the database, e.g.\ to cache
+                historical, summary or computed data.
+
         \end{itemize}
@@
         introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial, sqlcourse.com}).  Although
+        others are~\cite{w3schools-sql-tutorial,sqlcourse.com}).  Although
         every vendor implements a different dialect of \SQL{}, the basics
@@
         minor differences between Irish English, British English, American
-        English and Australian English).  Depending on the database in use
+        English, and Australian English).  Depending on the database in use
         there may be tools available that reduce or remove the requirement
-        to know \SQL{}; for SQLite (the default database used by the
-        implementation) several are available~\cite{sqlite-guis}.
+        to know \SQL{}; several are available~\cite{sqlite-guis} for
+        SQLite (the default database used by the parser
+        implementation).
 
@@
 Storing the results in a database will also increase the efficiency of
-using those results, as the log files need only be parsed once rather than
-each time the data is used; indeed the results may be used by someone with
-no access to the original log files.
+using those results, because the log files need only be parsed once rather
+than each time the data is used; indeed the results may be used by someone
+with no access to the original log files.
 
@@
 sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite simple to test and debug problems with it.  Despite
+protocol, making it quite easy to test and debug problems with it.  Despite
 the simplicity of the protocol many virus and spam sending programs fail
@@
 
-This client is so poorly written that not only does it present the sender
-and recipient addresses improperly, it ignores the error messages returned
-by the server and carries on regardless.  There are many spam and virus
-sending programs that are this deficient --- unfortunately others
-(particularly newer programs) were written by competent programmers, or
-utilise competently written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as
+This example client is so poorly written that not only does it present the
+sender and recipient addresses improperly, it ignores the error messages
+returned by the server and carries on regardless.  Many spam and virus
+sending programs are this deficient --- unfortunately others (particularly
+newer programs) were written by competent programmers, or utilise
+competently written programs (e.g.\ Postfix or Sendmail on Unix hosts,
+Microsoft Outlook on Windows hosts).  Traditionally a mail server would
+have done its best to deal with deficient clients, with the intention of
+accepting as much mail destined for its users as
 possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or accepting sender or recipient addresses that were not enclosed
-in \texttt{<>}.
+command, or by accepting sender or recipient addresses that were not
+enclosed in \texttt{<>}.
 
 A detailed description of \SMTP{} is beyond the scope of this document:
-introductory guides can be found at~\cite{smtp-intro-01, smtp-intro-02},
-the definitive references are~\cite{RFC821, RFC2821}.
+introductory guides can be found at~\cite{smtp-intro-01,smtp-intro-02},
+and the definitive references are~\cite{RFC821,RFC2821}.
 
@@
 
-Postfix is a \MTA{} with the following design aims (in order of
-importance): security, flexibility of configuration, scalability, and high
-performance.  It features extensive, extensible, optional anti-spam
+Postfix is a \MTAlong{}~(\MTA{}) with the following design aims (in order
+of importance): security, flexibility of configuration, scalability, and
+high performance.  It features extensive, extensible, optional anti-spam
 restrictions, allowing an administrator to deploy those restrictions which
@@
 by Postfix's author.  These restrictions can be selectively applied,
-combined and bypassed on a per-client, per-recipient or per-sender basis,
+combined, and bypassed on a per-client, per-recipient, or per-sender basis,
 allowing varying levels of stricture and permissiveness.  Postfix leverages
@@
 difficult.  For most mails the journey is simple and brief, but the
-remaining minority can be quite complex (see \sectionref{additional
-complications} for details).
+remaining minority can have quite complex journeys (see
+\sectionref{additional complications} for details).
 
-Postfix's design follows the Unix philosophy of \textit{Write programs that
-do one thing and do it well\/}~\cite{unix-philosophy}, and is separated
-into various component programs to perform the tasks required of an
-\MTA{}\@: receive mail, send mail, local delivery of mail, etc.\ --- full
-details can be found in~\cite{postfix-overview}.  Each log line contains
-the name of the Postfix component that produced it, and this information
-is used when determining which rules should be used to parse each log line
-(see \sectionref{rule characteristics} for details).  This design has
-positive security implications also: those components that interact with
-other hosts are not privileged,\footnote{Privilege in this case means the
-power to perform actions that are limited to the administrator, and not
-available to ordinary users.} so bugs in those components will not give an
-attacker extra privileges; those components that are privileged do not
-interact with other hosts, making it much more difficult for an attacker to
-exploit any bugs that may exist in those components.
+Postfix's design follows the Unix philosophy of \textit{``Write programs
+that do one thing and do it well''\/}~\cite{unix-philosophy}, and is
+separated into various component programs to perform the tasks required of
+an \MTA{}\@: receive mail, send mail, local delivery of mail, etc.\ ---
+full details can be found in~\cite{postfix-overview}.  Each log line
+contains the name of the Postfix component that produced it, and this
+information is used when determining which rules should be used to parse
+each log line (see \sectionref{rule characteristics} for details).
+Postfix's design is strongly influenced by security concerns: those
+components that interact with other hosts are not
+privileged,\footnote{Privilege means the power to perform actions that are
+limited to the administrator, and not available to ordinary users.} so bugs
+in those components will not give an attacker extra privileges; those
+components that are privileged do not interact with other hosts, making it
+much more difficult for an attacker to exploit any bugs that may exist in
+those components.
 
@@
 evaluated until the first RCPT TO command is received, because some clients
-do not deal properly with rejections before RCPT TO\@; a benefit of this
-delay is that Postfix has more information available when logging
-rejections.
+do not deal properly with rejections before the first RCPT TO command; a
+benefit of this delay is that Postfix has more information available when
+logging rejections.
 
@@
 evaluated.\footnote{Other results are possible as described
-in~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.} The
+in~\cite{smtpd_access_readme,smtpd_per_user_control,policy-servers}.} The
 meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
 \textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing more
-specific cases to be used as exceptions to more general cases.  When the
-result is the name of another restriction Postfix will evaluate the new
-restriction, allowing restrictions to be chosen based on the client \IP{}
-address, client hostname, HELO hostname, sender address, or recipient
-address.\footnote{E.g.\ the administrator may require that all clients on
-the local network have valid \DNS{} entries, to prevent people sending mail
-from unknown machines.}  The administrator can define new restrictions as a
-list of existing restrictions, allowing arbitrarily long and complex
-sequences of lookups, restrictions and exceptions.  Postfix tries to
-protect the administrator in as far as is reasonable, e.g.\ the restriction
+continue processing the remainder of the restriction list, allowing
+exceptions to more general rules.  When the result is the name of another
+restriction Postfix will evaluate that restriction, allowing restrictions
+to be chosen based on the client \IP{} address, client hostname, HELO
+hostname, sender address, or recipient address.\footnote{E.g.\ the
+administrator may require that all clients on the local network have valid
+\DNS{} entries, to prevent people sending mail from unknown machines.}  The
+administrator can define new restrictions as a list of existing
+restrictions, allowing arbitrarily long and complex sequences of lookups,
+restrictions and exceptions.  Postfix tries to protect the administrator in
+as far as is reasonable, e.g.\ the restriction
 \texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
@@
 under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting mail for
-all users --- the administrator must be cognisant of the effects of their
+the administrator to make catastrophic mistakes, e.g.\ rejecting all mail
+--- the administrator must be cognisant of the effects of their
 configuration changes.  This is similar to one of UNIX's design
 philosophies: \textit{``UNIX was not designed to stop its users from doing
-stupid things, as that would also stop them from doing clever things''\/}
---- Doug Gwyn~\cite{unix-philosophy}.
+stupid things, as that would also stop them from doing clever
+things''\/}~\cite{unix-philosophy}.
 
 Postfix uses simple lookup tables to make decisions when evaluating some
-restrictions, e.g.\ in the example restriction below\newline
+restrictions, e.g.\newline{}
 \tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
-\newline The line above can be broken down as follows:
 
-\begin{description}
+\begin{eqlist}
 
@@
 
-\end{description}
+\end{eqlist}
 
 The restriction \texttt{check\_client\_access} checks if the \IP{} address
-of the connected client is found in the file and returns the associated
-action if found; the method of searching the file is dependant on the type
-of the file (\texttt{cidr} in the example) --- see
+of the connected client is found in the specified file and returns the
+associated action if found; the method of searching the file is dependant
+on the type of the file (\texttt{cidr} in the example) --- see
 \cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\
-\texttt{reject\_rbl\_client dnsbl.example.com} checks whether the \IP{}
-address of the client is present in the \DNSBL{}
-\texttt{dnsbl.example.com}, rejecting the command if the client is listed.
-This description is necessarily brief, for further details
-see~\cite{smtpd_access_readme, smtpd_per_user_control, policy-servers}.
+determine their result by consulting external sources, e.g.\newline{}
+\tab{}\texttt{reject\_rbl\_client dnsbl.example.com}\newline{} checks the
+\DNSBL{} \texttt{dnsbl.example.com}, rejecting the command if the client's
+\IP{} address is listed.  XXX DO I NEED THIS SENTENCE\@?  This description
+is necessarily brief, for further details see~\cite{smtpd_access_readme,
+smtpd_per_user_control, policy-servers}.
 
@@
 
-A \textit{policy server\/}~\cite{policy-servers} is an external program
-consulted by Postfix to determine the fate of an \SMTP{} command.  The
-policy server is given state information\footnote{Sample state information
-is shown in \tableref{Example attributes sent to policy servers}} and
-returns a verdict as described in \sectionref{Mixing and matching Postfix
+A policy server~\cite{policy-servers} is an external program consulted by
+Postfix to determine the fate of an \SMTP{} command.  The policy server is
+given state information by Postfix (sample state information is shown in
+\tableref{Example attributes sent to policy servers}) and returns a verdict
+from the set described in \sectionref{Mixing and matching Postfix
 restrictions}.  The policy server can perform more complex checks than
-those provided by Postfix: a trivial example is restricting mail from
-addresses associated with the payroll system to sending mail on the third
-Tuesday after pay day only, to help prevent problems from spam or (worse)
-phishing mails with faked sender addresses.\footnote{A phishing mail might
-claim that the payroll system had a disastrous disk failure; until the
-server is replaced and restored all salary payments will have to be
-processed manually, so please reply to this mail with your name, address
-and back account details.}
-
-Some widely deployed policy servers:
-
-\begin{itemize}
-
-    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter} and
-        joe~jobs~\cite{Wikipedia-joe-job}, but there has been considerable
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}, e.g.\ forwarding mail from one company or
-        university to another when a user changes jobs.
-
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects mail when the triple of (sender address, recipient address,
-        remote \IP{} address) is unknown; on second and subsequent delivery
-        attempts from that triple the mail will be accepted.  The
-        assumption is that maintaining a list of failed addresses and
-        retrying after a temporary failure is uneconomical for a spam
-        sender, but that a legitimate mail server must retry.  Sadly spam
-        senders are using increasingly complex and well written programs to
-        distribute spam, frequently using an \ISP{} provided \SMTP{} server
-        from a compromised machine on the \ISP{}'s network.  Greylisting
-        will slowly become less useful, but it does block a large
-        percentage of spam mail at the moment; the most effective
-        restrictions over the \numberOFlogFILES{} log files used in testing
-        the parser are shown in \tableref{Summary of rejections}.
-        Greylisting is obviously worth using, at least at the moment,
-        particularly when you consider Greylisting's position as the final
-        restriction that a mail must overcome:\footnote{Greylisting is the
-        final restriction a mail must overcome in the configuration used on
-        the mail server the log files were obtained from; an administrator
-        is free to use Greylisting at whichever position in the restriction
-        list they feel is most appropriate for their mail system.}
-        Greylisting only takes effect for mails that have passed every
-        other restriction.
-
-        \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
-            \input{build/include-restriction-summary-table}
-        \end{table}
-
-    \item Using a scoring system such as
-        Policyd-weight~\cite{policyd-weight} where tests accumulate points
-        against the sending system --- if the eventual score is higher than
-        a threshold the mail is rejected.  Postfix's restrictions are
-        binary, and the administrator must manually whitelist clients if
-        they are to bypass a restriction; using a threshold that requires a
-        delivery attempt to hit several restrictions frees the
-        administrator from whitelisting clients that fall foul of one
-        restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
-
-\end{itemize}
-
-Example attributes taken from~\cite{policy-servers}:
+those provided by Postfix: a trivial example is allowing addresses
+associated with the payroll system to send mail on the third Tuesday after
+pay day only, to help prevent problems from phishing mails using faked
+sender addresses.\footnote{A phishing mail might claim that the payroll
+system had a disastrous disk failure, and until the server is replaced all
+salary payments will have to be processed manually, so please reply to this
+mail with your name, address, and bank account details.}
 
@@
 
+    XXX HOW CAN I CENTRE THE TABLE\@?
     \begin{tabular}[]{ll}
@@
     \end{tabular}
+
 \end{table}
 
+Some widely deployed policy servers:
+
+\begin{itemize}
+
+    \item Checking \SPF{} records~\cite{openspf,Wikipedia-spf}.
+        \SPF{}\label{spf introduction} records specify which mail servers
+        are allowed to send mail claiming to be from a particular domain.
+        The intention is to reduce spam from faked sender addresses,
+        backscatter~\cite{postfix-backscatter}, and
+        joe~jobs~\cite{Wikipedia-joe-job}.  There has been considerable
+        resistance to the proposal because it breaks or vastly complicates
+        some features of \SMTP{}, e.g.\ forwarding mail from one company or
+        university to another when a user moves.
+
+    \item Greylisting~\cite{greylisting} is a technique that temporarily
+        rejects a delivery attempt when the triple of \newline{}
+        \tab{}\texttt{(sender address, recipient address, remote \IP{}
+        address)}\newline{} has not been seen before; on second and
+        subsequent delivery attempts from that triple the mail will be
+        accepted.  This blocks spam because maintaining a list of failed
+        addresses and retrying after a temporary failure is uneconomical
+        for a spam sender, but a legitimate mail server must retry
+        deliveries that temporarily failed.  Sadly spam senders are using
+        increasingly complex and well written programs to distribute spam,
+        frequently using an \ISP{} provided \SMTP{} server from a
+        compromised machine on the \ISP{}'s network.  Greylisting will
+        slowly become less useful, but it does block a large percentage of
+        spam mail at the moment; the most effective restrictions from the
+        \numberOFlogFILES{} log files used in testing the parser are shown
+        in \tableref{Summary of rejections}.  Greylisting is obviously
+        worth using, at least at the moment, particularly when you consider
+        Greylisting's position as the final restriction that a mail must
+        overcome:\footnote{Greylisting is the final restriction a mail must
+        overcome in the configuration used on the mail server the log files
+        were obtained from; an administrator is free to use Greylisting
+        wherever in the restriction list they feel is most appropriate for
+        their mail system.} Greylisting only takes effect for mails that
+        have passed every other restriction.  Some problems may be
+        encountered when using greylisting: some servers fail to retry;
+        legitimate mail may be delayed, particularly when coming from a
+        pool of servers.
+
+        \begin{table}[ht]
+            \caption{Summary of rejections}\label{Summary of rejections}
+            \input{build/include-restriction-summary-table}
+        \end{table}
+
+    \item Scoring systems such as Policyd-weight~\cite{policyd-weight}
+        perform tests on features of the delivery attempt (e.g.\ \IP{}
+        address, sender address), incrementing or decrementing a score
+        based on the results; if the eventual score is higher than a
+        threshold the mail is rejected.  The administrator must manually
+        whitelist clients if they are to bypass a Postfix restriction;
+        using a threshold that requires a delivery attempt to hit several
+        restrictions frees the administrator from whitelisting clients that
+        fall foul of one restriction only.
+
+    \item Rate limiting on a per-sender, per-client or per-recipient basis
+        as performed by Policyd~\cite{policyd}.
+
+\end{itemize}
+
 \subsection{Conventions used in the document}
+
+XXX MAYBE REWRITE SO THAT THIS IS NOT TRUE\@?  ARE THERE ANY OTHER
+CONVENTIONS\@?
 
@@
 
-This section has provided background information on several topics relevant
-to the remainder of the document.  It started with the motivation behind
-the project, continuing with explanations of:
-
-\begin{itemize}
-
-    \item Using a database as an \API{}.
-
-    \item \SMTP{}.
-
-    \item Postfix restrictions and policy servers.
-
-    \item Assumptions made when designing and developing the parser.
-
-    \item A description of the parser's novel design.
-
-    \item Conventions used in this document.
-
-    \item An in-depth comparison of this parser with \LMA{}, the parser
-        described in previously published research.
-
-    \item A review of the literature previously published in this area.
-
-\end{itemize}
+This section has provided background information on several topics,
+starting with the motivation behind the project, continuing with an
+explanation of the use of a database as an \API{} and an introduction to
+\SMTP{}.  The focus then switched to the \MTA{} in use, Postfix, and its
+restrictions and policy servers, finishing with the conventions used in
+this thesis.

### 2008-10-28 doc/logparser-implementation.tex
First pass through corrections for sections 1-3.

@@
     \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is: \newline \tab{}
-        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline i.e.\
+        log line will be extracted.  The format is: \newline{} \tab{}
+        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline{} i.e.\
         semi-colon or comma separated assignment statements, with the
@@
         \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
+        \newline{} \tab{} server\_hostname} and \texttt{helo}.\newline{}
         \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline \tab{} size,} and \texttt{data}
+        \newline{} \tab{} size,} and \texttt{data}
 
@@
     \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline used by the algorithm, they serve to document the rule
+        not \newline{} used by the algorithm, they serve to document the rule
         for the user's benefit.
@@
     \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline used by the action to extract and
+        connection\_data:] are \newline{} used by the action to extract and
         save data matched by the \regex{}.
@@
 
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
+        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline{}
         delay=0, status=sent (forwarded as 56F5B43FD)}
@@
 
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline
+\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline{}
 delay=0, status=sent (forwarded as 56F5B43FD)}
@@
 
-    \item The message-id has a specific format: \newline
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline
+    \item The message-id has a specific format: \newline{}
+        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline{}
         e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
@@
     \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline both local and remote delivery.
+        submitted mail, \newline{} both local and remote delivery.
 
@@
             \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline \tab{}\texttt{/\^/~~~~WARN~Logging~HELO}
+                containing:\newline{} \tab{}\texttt{/\^/~~~~WARN~Logging~HELO}
 
             \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline
+                \texttt{/etc/postfix/main.cf} to contain\newline{}
                 \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}

### 2008-10-28 doc/logparser-introduction.tex
First pass through corrections for sections 1-3.

@@
 being rejected, or check if new anti-spam measures are working.  The more
-adventurous will have generated statistics to show how many hits each of
-their anti-spam measures has had in the last week, and possibly even
+adventurous will have generated statistics to show how successful each of
+their anti-spam measures has been in the last week, and possibly even
 generated some graphs to clearly illustrate these statistics to management
@@
 paper documents the difficult process of parsing Postfix log files,
-presenting \PLP{}, a program that parses log files and places the resulting
+presenting \PLP{}, a program that parses Postfix log files and places the resulting
 data into a database for later use.  The gathered data can then be used to
@@
 
-\vspace{1em}\noindent\textbf{Layout of the paper:}
+\vspace{1em}\noindent\textbf{Layout:}
 
 Section~\ref{background} provides background information useful in
-understanding the paper, parser and algorithm.  It introduces the idea of
-using a database schema as an \API{}, providing an interface to the data
+understanding the thesis, parser, and architecture.  It introduces the idea
+of using a database schema as an \API{}, providing an interface to the data
 gathered that is language-neutral.  The novel separation of rules, actions

### 2008-10-28 doc/logparser-other-parsers-review.tex
Minor changes.
First pass through corrections for sections 1-3.

@@
 Ten other parsers, starting with the only previously published research in
-this area, have been reviewed as part of the background research for this
-project.  None of the parsers reviewed perform the kind of advanced parsing
-and log correlation required for this project; all bar one are intended to
-perform a specific parsing and reporting task, rather than to be a generic
+this area that the author is aware of, have been reviewed as part of the
+background research for this project.  None of the parsers reviewed perform
+the kind of advanced parsing and correlation required for this project; all
+except one (XXX IS THIS ACCURATE\@?) are intended to perform a specific
+parsing and reporting task, whereas \PLP{} is designed to be a generic
 parser, extracting data and leaving generation of reports from the data to
@@
 the majority discard all data once they have finished running, making
-historical analysis impossible.  The other parsers reviewed all produce a
-report of varying complexity and detail, whereas the program described here
-does not produce a report at all; that responsibility is deferred to a
-separate program, to be developed later.  The parsing algorithm and program
-developed for this project are designed to enable much more detailed log
-analysis by providing a stable platform for subsequent programs to develop
-upon.
+historical analysis impossible.  The parsers reviewed all produce a report
+of varying complexity and detail, whereas this project's parser does not
+produce a report at all; that responsibility is deferred to a separate
+program, to be developed later.  The parser developed for this project is
+designed to enable much more detailed log analysis by providing a stable
+platform for subsequent programs to develop upon.
+
+It is important to compare and contrast newly developed programs,
+algorithms, and parsers against those already available, to accurately
+judge what improvements, if any, are delivered by the newcomers.  There are
+not that many previously developed Postfix log parsers --- indeed it was
+quite difficult to find ten parsers to review for this project --- and the
+functionality offered ranges from quite basic to much more mature,
+depending on the needs of the creator.  The programs are not reviewed from
+an independent viewpoint; the objective is to compare and contrast each
+program with \parsername{}.  There are some important differences between
+\parsername{} and the parsers reviewed here:
+
+\begin{enumerate}
+
+    \item Only \parsername{} makes it possible to parse new log lines
+        without extensive and intrusive modifications to the parser;
+        \parsername{}'s architecture is described in \sectionref{why
+        separate rules, actions and framework?}.
+
+    \item \parsername{} does not provide any reports based on the data it
+        gathers.  This is by design: the parser is responsible for parsing
+        log files and extracting data; further processing of that data is a
+        separate concern, deferred to another program.  Using an \SQL{}
+        database simplifies the process of generating such reports
+        (discussed in \sectionref{database as API}); some simple examples
+        are given in \sectionref{motivation}.
+
+    \item Most of the parsers silently ignore log lines they cannot handle,
+        whereas \parsername{} complains loudly about every single log line
+        it fails to parse.  The exception is AWStats, which outputs the
+        percentage of input lines it was unable to parse, but does not
+        output the lines themselves.
+
+    \item A minor difference is that most parsers do not handle compressed
+        files; both \parsername{} and Splunk handle them transparently,
+        without user intervention; Sawmill and Lire can be configured to
+        support compressed files, but Sawmill exhibits a dramatic slow down
+        when thus configured.  Although this is a minor difference, support
+        for reading compressed log files is quite helpful, as it
+        dramatically reduces the disk space required to store old log
+        files.
+
+\end{enumerate}
 
@@
 
-There only appears to be one prior paper published about parsing Postfix
+There only appears to be one prior published paper about parsing Postfix
 log files: \textit{Log Mail Analyzer: Architecture and Practical
@@
 single line, with fields separated by a comma or other punctuation symbol.
-Problems with \CSV{} files include the difficulty in escaping separators,
-providing multiple values for a field (e.g.\ multiple recipients) and
-adding new fields.  There is no mechanism to document the fields or the
-separator, unlike \SQL{} databases where every database includes a schema
-naming the fields and the type of data they store (integer, text,
-timestamp, etc.).
+Problems with \CSV{} files include the need to escape separators in the
+data stored, providing multiple values for a field (e.g.\ multiple
+recipients), and adding new fields.  There is no mechanism to document the
+fields or the separator, unlike \SQL{} databases where every database
+includes a schema naming the fields and the type of data they store
+(integer, text, timestamp, etc.).
 
 Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
-unlike \SQL{} databases that store arbitrary tuples; in \LMA{}'s main
+unlike \SQL{} databases that store arbitrary tuples.  In \LMA{}'s main
 table the key is an integer referred to by secondary tables, and the value
-used is a \CSV{} line containing all of the data for that row.  The
-secondary by-sender, by-recipient, by-date and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and a \CSV{} list of integers
-referring to the main table is the value.  This effectively re-implements
+is a \CSV{} line containing all of the data for that row.  The
+secondary by-sender, by-recipient, by-date, and by-\IP{} tables use the
+sender/recipient/date/\IP{} as the key, and the value is a \CSV{} list of integers
+referring to the main table.  This effectively re-implements
 \SQL{} foreign keys, but without the functionality offered by even the most
@@
 
-The data stored is limited to time and date, hostname and \IP{} address of
-client, \IP{} of server, sender and recipient addresses, \SMTP{} code, and
-size (for accepted mails only).  Handling of multiple recipients, \SMTP{}
-codes or remote servers\footnote{A single mail may be sent to multiple
-remote servers if it was addressed to recipients in different domains, or
-Postfix needs to try multiple servers for one or more recipients.} is not
-explained in the paper; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
-not associated or linked in any way), and presumably the same approach is
-taken when there are multiple destination servers.
+\LMA{} stores the following date: time and date, hostname and \IP{} address
+of client, \IP{} of server, sender and recipient addresses, \SMTP{} code,
+and size (for accepted mails only).  It omits XXX XXX\@.  Handling of
+multiple recipients, \SMTP{} codes or remote servers\footnote{A single mail
+may be sent to multiple remote servers if it was addressed to recipients in
+different domains, or Postfix needs to try multiple servers for one or more
+recipients.} is not explained in the paper; experimental observation shows
+that multiple records are added when there are multiple recipients (sadly
+the records are not associated or linked in any way), and presumably the
+same approach is taken when there are multiple destination servers.
 
@@
 
-Parsing in \LMA{} requires major changes to the code to parse new log lines
-or extract additional data.  It does not appear to deal with any of the
-complications discussed in \sectionref{complications} (initial
-complications) and \sectionref{additional complications} (additional
-complications).  It does not differentiate between different types of
-rejections, so it is not suitable for the purposes of this project; the
-data about which restriction caused the rejection is discarded, whereas an
-important goal of this project is to retain that data to aid optimisation
-and evaluation of Postfix restrictions.  \LMA{} fails to parse Postfix log
-files generated on Solaris hosts because the fields added automatically
-differ from those added on Linux hosts; log files from Solaris hosts (and
-possibly other operating systems) thus require preprocessing before parsing
-by \LMA{}.  Once the preprocessing has been performed on the
-\numberOFlogFILES{} test log files \LMA{} parses the log files without
-complaint, though it does produce 32 entries in its output file for every
-rejection in the input log file; it also missed some 40\% of delivered
-mail.  Once these glaring deficiencies were discovered the author did not
-spend more time checking the results.
+\LMA{} requires major changes to the parser code to parse new log lines or
+extract additional data.  The code is structured as a long series of blocks
+that each handle all lines matching a single \regex{}, so parsing new log
+lines requires modifying an existing \regex{} or carefully inserting a new
+block in the correct place; extracting extra data will require modifying
+multiple blocks, \regexes{}, or both.
+
+\LMA{} does not appear to deal with any of the complications discussed in
+\sectionref{complications}.  It does not differentiate between different
+types of rejections, so it is not suitable for the purposes of this
+project; the data about which restriction caused the rejection is
+discarded, whereas the main goal of this project is to retain that data to
+aid optimisation and evaluation of Postfix restrictions.  \LMA{} fails to
+parse Postfix log files generated on Solaris hosts because the fields
+automatically prepended to each log line differ from those added on Linux
+hosts; log files from Solaris hosts (and possibly other operating systems)
+thus require preprocessing before parsing by \LMA{}.  Once the
+preprocessing has been performed on the \numberOFlogFILES{} test log files
+\LMA{} parses the log files without complaint, although it produced 32
+entries in its output file for every rejection in the input log file; it
+also missed some 40\% of delivered mail.  Once these glaring deficiencies
+were discovered the author did not waste any more time checking the
+results.
 
 \LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  IP-STORY shows the \CSV{} lines with the specified
-client \IP{} address; FROM-STORY shows the \CSV{} lines with the specified
-sender address; DAILY-EMAIL shows the \CSV{} lines for the specified day.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter; after making some corrections to the code it worked, and
-produced the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk},
-e.g.\ the most complicated, DAILY-REJECT, is merely:
+and DAILY-REJECT\@.  The first three extract \CSV{} lines for the specified
+client \IP{} address, sender address, or date respectively.  DAILY-REJECT
+initially failed with an error message from the Perl interpreter; after
+making some corrections to the code it worked, and extracted the \CSV{}
+lines for the specified day where the \SMTP{} code signifies a rejection.
+All of these reports are trivially simple to produce from the \CSV{} file
+using the standard Unix tool \texttt{awk}; the most complicated,
+DAILY-REJECT, is merely:
 
@@
     \item It outputs an unnecessary line containing only a single space
-        after each matching line, for compatibility with the output of
+        after each matching line, to accurately replicate the output of
         DAILY-REJECT\@.
@@
     \item Some records have an \SMTP{} code of \textit{0}, though this is
-        not a valid \SMTP{} code; presumably this is because of \LMA{} not
-        parsing a line properly.  \LMA{} considers this code to be a
-        rejection, so the command above must, too.
+        not a valid \SMTP{} code; presumably this is because \LMA{} did not
+        parse a log line properly.  DAILY-REJECT considers this code to be
+        a rejection, so the command above must too.
 
-    \item In a tiny proportion of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid code, and most likely
-        because of mis-parsing by \LMA{}); it is treated as a rejection by
-        \LMA{}, so the equivalent \texttt{awk} command must do so, too.
+    \item In a small fraction of records the \SMTP{} code is
+        \textit{deferred\/} (again, not a valid \SMTP{} code, and most
+        likely because of mis-parsing by \LMA{}); it is treated as a
+        rejection by DAILY-REJECT, so the equivalent \texttt{awk} command
+        must do so, too.
 
@@
 
-There are two differences in the output from DAILY-REJECT and the
+There are two differences between the output from DAILY-REJECT and the
 \texttt{awk} command:
@@
         addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.
+        present in the output from \texttt{awk}.  XXX WHY\@?  GIVE AN
+        EXAMPLE RECORD\@.
 
     \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        produces only one output line, whereas the \texttt{awk} command
-        above produces all matching lines.
+        extracts only one record, whereas the \texttt{awk} command above
+        extracts all matching records.
 
@@
 fashion is useful to systems administrators), rather than a program
-designed to be extensible and useful in a production environment.
+designed to be useful in a production environment; no consideration was
+given to extensibility when writing it.
 
@@
 
-It is important to compare and contrast newly developed programs,
-algorithms and parsers against those already available, to accurately judge
-what, if any, improvements are delivered by the newcomers.  There are not
-that many previously developed Postfix log parsers --- indeed it was quite
-difficult to find ten parsers to review for this project --- and the
-functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
-
-\begin{enumerate}
-
-    \item Only \parsername{} makes it possible to parse new log lines
-        without modifying the core parsing algorithm; see \sectionref{why
-        separate rules, actions and framework?} for a discussion of why the
-        separation of the algorithm into rules, actions and framework is
-        beneficial.
-
-    \item \parsername{} does not provide any reporting based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        log files and extracting data --- further processing of that data
-        is a separate concern, deferred to another program.  Using an
-        \SQL{} database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some simple examples
-        are given in \sectionref{motivation}.
-
-    \item Most of the parsers silently ignore log lines they cannot handle,
-        whereas \parsername{} complains loudly about every single line it
-        fails to parse.  The exception is AWStats, which outputs the
-        percentage of input lines it was unable to parse, but does not
-        output the lines themselves.
-
-    \item A minor difference is that most parsers do not handle compressed
-        files; both \parsername{} and Splunk handle them transparently,
-        without user intervention; Sawmill and Lire can be configured to
-        support compressed files, but Sawmill exhibits a dramatic slow down
-        when thus configured.  Although this is a minor difference, support
-        for reading compressed log files is quite helpful, as it
-        dramatically reduces the disk space required to store old log
-        files.
-
-\end{enumerate}
-
 \subsection{Pflogsumm}
@@
 
-Pflogsumm produces a report designed for troubleshooting, rather than for
+\noindent{}\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline{}
+(Last checked 2008/04/09.)
+
+Pflogsumm produces a report designed for troubleshooting rather than for
 in-depth analysis.  It does not support saving any data or extracting more
@@
 encountered during this project.  Pflogsumm produces a useful report, and
-successfully dealt with the \numberOFlogFILES{} log files tested
-with,\footnote{The results it reported were not verified, but it did not
-report any errors, and has a very good reputation amongst Postfix users.}
-but is not a suitable base for this project.
+successfully parsed the \numberOFlogFILES{} log files tested
+with,\footnote{The results it reported were not verified in detail, but it
+did not report any errors, and has a very good reputation amongst Postfix
+users.} but it is not a suitable base for this project.
 
@@
 
-    \item Total number of mails accepted, delivered and rejected.  Total
-        bytes of mail accepted and delivered.
+    \item Total number and bytes of mails accepted, delivered and rejected.
 
@@
     \item Per-hour averages and per-day summaries of the number of mails
-        received, delivered, deferred, bounced and rejected.
+        received, delivered, deferred, bounced, and rejected.
 
     \item For received mail: per-domain totals for mails sent, deferred,
-        average delay, maximum delay and bytes delivered.  For received
+        average delay, maximum delay, and bytes delivered.  For received
         mail: per-domain totals for mails received and bytes received.
@@
 report.
-
-\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline (Last checked
-2008/04/09.)
 
@@
 
+\noindent{}\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline{}
+\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline{}
+\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline{}
+(Last checked 2007/12/04.)
+
 Sawmill is a general purpose commercial product that parses 687 log file
@@
 (one is beta as of 2007/12/04), but they do not appear to be interlinked.
-No attempt is made to correlate log lines or deal with the difficulties
-documented in \sectionref{complications} \sectionref{additional
-complications}.\footnote{If any attempt is made there is no reference to it
-in the documentation available on the website.} The source code is
-available in an obfuscated form only (presumably for a fee), and the
-product is quite expensive, as it requires a license per report to be
-generated; in contrast \parsername{} is free and the code is freely
-available.  The web interface allows creation of dynamic reports based on
-any field, but because of the associated cost the author has not
-experimented with it.  Presumably, as the company charges per report the
-user wishes to generate, the data store (if there is one) is deliberately
-inaccessible and undocumented to prevent the user bypassing the program and
-generating their own reports.
+The documentation does not suggest that any attempt is made to correlate
+log lines or deal with the difficulties documented in
+\sectionref{complications}.  The source code is available in an obfuscated
+form only (presumably for a fee), and the product is quite expensive,
+requiring a license per report to be generated; in contrast \parsername{}
+is free and the code is freely available.  The web interface allows
+creation of dynamic reports based on any field, but because of the
+associated cost the author has not experimented with it.  Presumably, as
+the company charges per report the user wishes to generate, the data store
+(if there is one) is deliberately inaccessible and undocumented to prevent
+the user bypassing the program and generating their own reports.
 
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline Fields
+\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline{} Fields
 extracted: from, to, server, UID, relay, status, number of recipients,
 origin hostname, origin \IP{} and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in their documentation:
+\texttt{uid} and \texttt{virus} are not explained in the documentation:
 \texttt{server} is probably the server the mail is delivered to, and
@@
 
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline Fields
+\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline{} Fields
 extracted: from, to, RBL, client hostname and client \IP{}\@.
 
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline Fields
+\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline{} Fields
 extracted: from, to, client hostname, client \IP{}, relay hostname, relay
@@
 data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by an \DNSBL{} check.
+rejections except when the rejection is caused by a \DNSBL{} check.
 
@@
 but it exhibits a dramatic slow down when doing so: it took six hours to
-parse the first half of the log files; twelve hours to parse the next
+parse the first half of the log files, and twelve hours to parse the next
 third; after twenty four hours parsing the remaining sixth it crashed due
@@
 
-(Last checked 2007/12/04.)
+XXX ADD A SUMMARY\@.
 
@@
 
+\noindent{}\url{http://www.splunk.com/} \newline{}
+(Last checked 2008/04/29.)
+
 Splunk aims to index all an organisation's log files, providing a
@@
 graphs in real time, a facility not provided by \parsername{} (report
-generation has been deferred to a subsequent program).  Saved
-searches\footnote{The author was unable to save searches, though that may
-have been due to limitations in the free version.} can be run periodically
-and the results emailed to a recipient or sent to a shell script, that
-presumably can publish the results as required (though possibly without the
-graphs and detailed statistics); \parsername{} provides the database and
-leaves it to the user to utilise it in any way, whenever they want, with no
-restrictions on usage of the data.  The interface is optimised for
-interactive rather than automated queries and it does not appear to be
-possible to write independent tools to utilise the Splunk database; there
-are additional reports available on \url{http://www.splunkbase.com/}, but
-the author was unable to find any: every category was empty, even those
-that the interface claimed had reports available.  Many types of reports
-are bundled with the software, though most are variations of a bar or pie
-chart, except bubble and heatmap graphs.  It is easy to drill down through
-the graphs to extract a portion of the data (e.g.\ select the hour with the
-largest number of events, then select a particular host, and finally a
-specific address), though it is not possible to search on partial words.
+generation has been deferred to a subsequent program).  Saved searches can
+be run periodically and the results emailed to a recipient or sent to an
+external program for further processing (though maybe without the graphs
+and detailed statistics); the author was unable to save searches, though
+that may have been due to limitations in the free version.  The database is
+not available for use by external programs, whereas \parsername{} provides
+the database and leaves it to the user to utilise it without limit or
+restriction.  The interface is optimised for interactive rather than
+automated queries and it does not appear to be possible to write
+independent tools to utilise the Splunk database; there are additional
+reports available on \url{http://www.splunkbase.com/}, but the author was
+unable to find any: every category was empty, even those that the interface
+claimed had reports available.  Many types of reports are bundled with the
+software, though most are variations of a bar or pie chart, except bubble
+and heatmap graphs.  It is easy to drill down through the graphs to extract
+a portion of the data (e.g.\ select the hour with the largest number of
+events, then select a particular host, and finally a specific address),
+though it is not possible to search on partial words.
 
@@
 that the log file was copied into a spool directory, doubling the disk
-space usage.  The most suitable test would be to index all log files in a
-particular directory, but after three hours, numerous futile attempts, and
-reading all available documentation the author admitted defeat.  Using the
-\CLI{} was more successful: invoking the command \newline \tab{}
-\texttt{splunk find logs }\textit{log-directory\/}\newline added 40 of the
-\numberOFlogFILES{} log files to the queue for indexing.  Repeated attempts
-enqueued the same 40 log files, without explaining why the others
-were excluded.\footnote{The log files appear to have only been indexed
-once; presumably Splunk keeps track of the files it has indexed and
+space usage.  The most suitable test would be to index all the log files in
+a particular directory, but after three hours, numerous futile attempts,
+and reading all the available documentation the author admitted defeat.
+Using the \CLI{} was more successful: invoking the command \newline{}
+\tab{} \texttt{splunk find logs }\textit{log-directory\/}\newline{} added
+40 of the \numberOFlogFILES{} log files to the queue for indexing.
+Repeated attempts enqueued the same 40 log files, without explaining why
+the others were excluded.\footnote{The log files appear to have only been
+indexed once; presumably Splunk keeps track of the files it has indexed and
 discards requests to index files for a second time.  This may or may not be
@@
 \$5000 (plus \$1000 support), and still limits the volume of data indexed
-per day to 500MB\@.
+per day to 500MB\@.  Typical log file sizes for a small scale mail server
+are given in \sectionref{parser efficiency}.
 
 When parsing Postfix log files Splunk extracts some standard fields: to and
-from addresses, HELO hostname, date, host the log files were collected
-from, and protocol.  It parses the standard syslog fields at the beginning
-of the line, and extracts any \texttt{key=<value>} pairs occurring after
-the standard syslog prologue; these pairs are the fields listed above.
-Searches can be based on the extracted fields, and all the text in the line
-is also available for searching.  \parsername{} extracts noticeably more
-data, though it does not make the full text of the line available, and the
-full power of \SQL{} is available when searching, allowing the user to
-search on arbitrarily complicated conditions.
+from addresses, HELO hostname, time and date, the host the log files were
+generated on, and protocol (\SMTP{} or \ESMTP{}).  It parses the standard
+syslog fields at the beginning of the log line, and extracts any
+\texttt{key=<value>} pairs occurring after the standard syslog prologue;
+these pairs are the fields listed above.  Searches can be based on the
+extracted fields or the full text of the log line.  \parsername{} extracts
+noticeably more data, though it does not make the full text of the line
+available, and the full power of \SQL{} is available when searching,
+allowing the user to search on arbitrarily complicated conditions.
 
 Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields; most importantly it makes
-no attempt to correlate log lines by queueid or \pid{}, or to handle any of
-the myriad complications discussed in this document.  Some additional
-Postfix reports are supposedly available at
-\url{http://www.splunkbase.com/}, but the author was unable to find them
-(or any other reports).
-
-\url{http://www.splunk.com/} \newline (Last checked 2008/04/29.)
+above extracting the \texttt{key=<value>} fields from a log line; most
+importantly it makes no attempt to correlate log lines by queueid or
+\pid{}, or to handle any of the myriad complications discussed in
+\sectionref{complications}.  Some additional Postfix reports are supposedly
+available at \url{http://www.splunkbase.com/}, but the author was unable to
+find Postfix reports, or additional reports for any other log file types.
 
@@
 
-Isoqlog produces a report listing the number of mails sent by each unique
-sender address, and separately the total number of bytes transferred; both
-reports are produced for daily, monthly and annual time spans, but only for
-the domains listed in its configuration file (making it impossible to
-produce reports for every sender domain).  It appears to ignore all log
-lines except for those for the current day, though it does maintain a
-record of data previously extracted, which the newly extracted data is
-merged into (no information is provided on the format of the data store).
-The data extracted appears to be limited to the number of mails sent by
-each sender, unlike the copious amounts of data extracted by \parsername{}.
-It does not utilise rejection log lines in any way, so is unsuitable for
-the purposes of this project.  Its parsing is completely inextensible,
-indeed is almost incomprehensible, relying on \texttt{scanf(3)}, fixed
-offsets and low level string manipulation; it is the opposite end of the
-spectrum to \parsernames{} parsing.  It does not handle any of the
-complications discussed in this document, does not gather the breadth of
+\noindent{}\url{http://www.enderunix.org/isoqlog/} \newline{}
+(Last checked 2007/08/13.)
+
+Isoqlog's report lists daily, monthly, and annual totals of mails sent by
+each unique sender and bytes transferred (XXX EXPLAIN BYTES TRANSFERRED\@:
+TO/FROM, PER-SENDER\@?).  It only produces these reports for the domains
+listed in its configuration file, making it impossible to produce reports
+for every sender domain.  It ignores all log lines except those for today,
+so it is impossible to analyse historical logs, and testing with the
+\numberOFlogFILES{} test log files is pointless.  It does maintain a record
+of data previously extracted, which the newly extracted data is merged into
+(no information is provided on the format of the data store).  The data
+extracted is limited to the number of mails sent by each sender, unlike the
+copious amounts of data extracted by \parsername{}.  It does not utilise
+rejection log lines in any way, so is unsuitable for the purposes of this
+project.  Its parsing is completely inextensible, indeed is almost
+incomprehensible, relying on \texttt{scanf(3)}, unexplained fixed offsets,
+and low level string manipulation; it is the opposite end of the spectrum
+to \parsernames{} parsing.  It does not handle any of the complications
+discussed in \sectionref{complications}, does not gather the breadth of
 data required for this project, and ignores the majority of log lines
-produced by Postfix.  It ignores all log lines earlier than today, so
-testing with the test log files is pointless.
+produced by Postfix.
 
-\url{http://www.enderunix.org/isoqlog/} \newline (Last checked 2007/08/13.)
+XXX FINISH THIS\@.
 
@@
 
+\noindent{}\url{http://awstats.sourceforge.net/} \newline{}
+\url{http://awstats.sourceforge.net/awstats.mail.html} \newline{}
+\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL} \newline{}
+(Last checked 2007/10/13.)
+
 AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code restricts
-it to supporting the \LCD{}.  The data it will extract from an \MTA{} log
-file is limited in comparison to \parsername{}: \newline \tab{} time2,
-email, email\_r, host, host\_r, method, url, code and bytesd.\newline There
-does not seem to be an explanation of any of those fields in the
-documentation (\parsername{} provides copious documentation).  AWStats has
-no special purpose code to deal with the intricacies of Postfix log files,
-in fact it operates by coercing Postfix log files into Apache\footnote{The
-Apache web server is the most popular HTTP server in use for the past 10
-years; more information is available at \url{http://httpd.apache.org/}.}
-format log files, for analysis by AWStats' HTTP log file parser.  The
-converting parser only deals with a small portion of the log lines
-generated by Postfix, silently skipping those it cannot deal with, and does
-not distinguish between different types of rejection; it would be a lot of
-work to extend it to handle new log lines.  Although though it does
-correlate log lines by queueid, it does not deal with any of the other
-complications described in this document.  AWStats supports saving data but
-the format of the saved data is not documented (as far as the author could
-tell).  It also supports reading compressed log files, but that
-functionality was not tested.
+supporting many different services without special purpose code limits its
+functionality.  The data it will extract from an \MTA{} log file is limited
+in comparison to \parsername{}: \newline{} \tab{} time2, email, email\_r,
+host, host\_r, method, url, code, and bytesd.\newline{} XXX WHAT FIELDS ARE
+MISSING\@?  There does not seem to be an explanation of any of those fields
+in the documentation (\parsername{} provides copious documentation).
+AWStats coerces Postfix log files into Apache\footnote{The Apache web
+server is the most popular HTTP server in use over the past 10 years; more
+information is available at \url{http://httpd.apache.org/}.} format log
+files, for analysis by AWStats' HTTP log file parser.  The converting
+parser only deals with a small portion of the log lines generated by
+Postfix, silently skipping those it cannot deal with, and does not
+distinguish between different types of rejection; it would be a lot of work
+to extend it to handle new log lines.  Although it does correlate log lines
+by queueid (not by pid), it does not deal with any of the other
+complications described in \sectionref{complications}.  AWStats supports
+saving data but the format of the saved data is not documented, as far as
+the author could tell.  It also supports reading compressed log files, but
+that functionality was not tested.
 
-When tested with the \numberOFlogFILES{} test log files AWStats' report
-says it parsed 9,240,075 (88.70\%) log lines out of 10,416,129, with
-1,176,050 (11.29\%) corrupt lines; however there are actually
+When tested with the \numberOFlogFILES{} test log files AWStats' reported
+that it parsed 9,240,075 (88.70\%) of 10,416,129 log lines, skipping
+1,176,050 (11.29\%) corrupt log lines; however there are
 \numberOFlogLINES{} lines in the \numberOFlogFILES{} log files, so AWStats
-only parsed 17.15\% of the input lines, ignoring the remaining 82.85\%.
+parsed only 17.15\% of the input lines, ignoring the remaining 82.85\%.
+
 The graphs it produces give an overview of mails received for the last
@@
 
-    \item A top ten list of aggregated \SMTP{} error codes, e.g.\ the first
-        entry in the list is: \texttt{Requested mail action not taken:
-        relaying not allowed, unknown recipient user, \ldots}
+    \item The top ten \SMTP{} error codes (XXX SENDING, RECEIVING, OR
+        BOTH\@?), e.g.\ the first entry in the list is: \texttt{Requested
+        mail action not taken: relaying not allowed, unknown recipient
+        user, \ldots}
 
@@
 %  Found 9240075 new qualified records.
-
-\noindent\url{http://awstats.sourceforge.net/} \newline
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL}
-\newline (Last checked 2007/10/13.)
 
@@
 
-Anteater does not have any English documentation so it is difficult, nigh
-impossible, for this author to accurately comment on what analysis it
-performs.  It did not run successfully when tested, and its parsing would
-certainly be out of date as Postfix has evolved considerably since this
-tool was last updated (November 2003).  As it neither ran successfully nor
-has documentation the author can read a detailed review cannot be provided.
+\noindent{}\url{http://anteater.drzoom.ch/} \newline{}
+(Last checked 2007/08/13.)
+
+Anteater does not have any English documentation so it is impossible for
+this author to accurately comment on the analysis it performs.  It did not
+run successfully when tested, and its parsing would certainly be out of
+date as Postfix has evolved considerably since this tool was last updated
+(November 2003).  As it neither ran successfully nor has documentation the
+author can read a detailed review cannot be provided.
 
@@
 log lines.
-
-\url{http://anteater.drzoom.ch/} \newline (Last checked 2007/08/13.)
 
@@
 
+\noindent{}\url{http://yaala.org/} \newline{}
+(Last checked 2007/10/09.)
+
 YAALA uses a plugin based system to analyse log files and produce \HTML{}
@@
 the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so with enough
-knowledge of the data structure YAALA stores it should be possible to
-extract data with another Perl program also using the Storable module;
+data using Perl's Storable module~\cite{perl-storable}, so other Perl
+programs can use Storable to load, examine, and optionally modify the data;
 \parsername{} uses a well documented database which is accessible from the
@@
 
-YAALA provides a Postfix parser that extracts the following fields from
-specific log lines:
+YAALA provides a Postfix parser that extracts the following two types of
+fields from specific log lines:
 
-\noindent\tab{}Aggregations: count (not explained), bytes (sum of bytes
-transferred).\newline \tab{}Keyfields [sic]: incoming\_host,
-outgoing\_host, date, hour, sender, recipient, defer\_count, delay.
+\begin{eqlist}
 
-YAALA extracts most of the fields \parsername{} does, but it does not
-maintain separate counters for each restriction like \parsername{}; this
-rules out the possibility of using the collected data for  optimisation,
-testing or understanding of restrictions.  YAALA's Postfix parser does not
-deal with the complications explained in this document, though it does
-correlate log lines by queueid.
+    \item [Aggregations:] count (not explained), bytes (sum of bytes
+        transferred).
+
+    \item [Keyfields:] incoming\_host, outgoing\_host, date, hour, sender,
+        recipient, defer\_count, delay.
+
+\end{eqlist}
+
+YAALA extracts most of the fields \parsername{} does, but unlike
+\parsername{} it does not maintain separate counters for each restriction;
+this rules out the possibility of using the collected data for
+optimisation, testing or understanding of restrictions.  YAALA's Postfix
+parser does not deal with the complications explained in
+\sectionref{complications}, though it does correlate log lines by queueid.
 
 YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen at~\cite{yaala-samples}.  Example
-query: \newline \tab{} \texttt{requests BY file WHERE host =\~{} Google}
-\newline The mini-language is quite limited and cannot be used to extract
-data for external use, merely to create reports.  Only data selected by the
-query will be saved in the data store; other data will be discarded, and
-removed from the data store if already present.
+reports; sample reports can be seen
+at~\url{http://www.yaala.org/samples.html}.  Example query for HTTP proxy
+servers: \newline{} \tab{} \texttt{requests BY file WHERE host =\~{}
+Google} \newline{} The mini-language is quite limited and cannot be used to
+extract data for external use, merely to create reports.  Only data
+selected by the query will be saved in the data store; other data will be
+discarded, and removed from the data store if already present.
 
@@
 produced a similar error message:
-\newline\tab{}\texttt{lib/Yaala/Data/Core.pm: Unavailable aggregation
-requested:} \newline\tab{}\texttt{``bytes''. Returning 0.} \newline  The
-underlying reason for this is that YAALA only parsed 408 log lines of
-360632 log lines (0.11\%) in the first log file.
+\newline{}\tab{}\texttt{lib/Yaala/Data/Core.pm: Unavailable aggregation
+requested:} \newline{}\tab{}\tab{}\texttt{``bytes''. Returning 0.}
+\newline{}  The underlying reason for this is that YAALA only parsed 408
+(0.11\%) of 360632 log lines in the first log file; it was not tested with
+the remainder of the \numberOFlogFILES{} log files.
 
-In summary YAALA provides a Postfix parser that only attempts to parse the
-most common Postfix log lines (but fails miserably), provides reasonably
-flexible report generation from the limited data extracted, but has no
-facilities to extract data for use in other tools.
-
-\url{http://yaala.org/} \newline (Last checked 2007/10/09.)
+In summary YAALA provides a Postfix parser that tries to parse the most
+common Postfix log lines only, provides reasonably flexible report
+generation from the limited data extracted, but has no facilities to
+extract data for use in other tools.
 
@@
 
+    The email servers' reports will show you the number of deliveries and
+    the volume of email delivered by day, the domains from which you
+    receive or send the most emails, the relays most used, etc.
+
 \end{quotation}
 
+\noindent{}\url{http://logreport.org/lire.html} \newline{}
+(Last checked 2008/04/29.)
+
 Lire is a general purpose log parser supporting many different types of log
-file.  Its Postfix parser extracts the following data from Postfix log
-files: \textit{The email servers' reports will show you the number of
-deliveries and the volume of email delivered by day, the domains from which
-you receive or send the most emails, the relays most used, etc.\/}; notably
-rejections are not mentioned or dealt with, and unlike \parsername{} there
-is no facility to extend the parser.  It supports multiple output formats
-for generated reports (text, \HTML{}, \PDF{} and Excel 95) but the reports
-do not appear to be customisable; \parsername{} does not produce any
-reports.  Lire supports saving extracted data for later report generation,
-but accessing this data from another application is undocumented; given the
-source code it should be possible, with enough time and effort, to access
-the data from an external program.
+file.  The data extracted by its Postfix parser is shown in the quote
+above; notably rejections are not mentioned or dealt with, and unlike
+\parsername{} there is no facility to extend the parser.  It supports
+multiple output formats for generated reports (text, HTML, PDF, and Excel
+95) but the reports do not appear to be customisable; \parsername{} does
+not produce any reports.  Lire produces a report with less detail than
+Pflogsumm, and is considerable harder to configure.  Lire supports saving
+extracted data for later report generation, but accessing this data from
+another application is undocumented; given the source code it should be
+possible, with enough time and effort, to understand the format.
+\parsername{} uses an \SQL{} database to make accessing the extracted data
+as easy as possible.
 
@@
 makes no attempt to deal with the other complications described in
-\sectionref{complications} \sectionref{additional complications}.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.  When testing Lire on the \numberOFlogFILES{} test log
-files it performed reasonably well, producing summaries of:
+\sectionref{complications}.  When testing Lire on the \numberOFlogFILES{}
+test log files it performed reasonably well, producing summaries of:
 
@@
 
-    \item Number of deliveries and bytes per day and hour.
+    \item Number of deliveries and bytes per-day and per-hour.
 
@@
 
-    \item Deliveries by relays, by size and by delay.
+    \item Deliveries by relays, by size, and by delay.
 
-    \item Delays by server and domain.
+    \item Delays by server and by domain.
 
-    \item Most mails exchanged between a pair of correspondents.
+    \item Which pair of correspondents exchanged the highest number of
+        emails.
 
@@
 The numbers it reports appear reasonable, and the subset verified by the
-author were correct.  Lire produces a report with less detail than
-Pflogsumm, and is considerable harder to configure.
-
-\url{http://logreport.org/lire.html} \newline (Last checked 2008/04/29.)
+author were correct.
 
@@
 
+\noindent{}\url{http://www.itefix.no/phpws/index.php} \newline{}
+(Last checked 2007/11/18.)
+
 Logrep extracts roughly half the fields \parsername{} does:
+
+XXX WHAT FIELDS ARE MISSING\@?
 
@@
 It does not try to address any of the complications in
-\sectionref{complications} \sectionref{additional complications} except for
-correlating by queueid.
+\sectionref{complications} except for correlating by queueid.
 
@@
 documentation is available on its website (\parsername{} provides copious
-documentation).  It requires a web browser to interact with it, ensuring
-that automated log processing will be difficult, whereas automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.\footnote{If
-it is caused by operator error, the interface needs improvement as the
-(minimal) instructions were followed as closely as possible, and multiple
-attempts were made.}  As parsing failed it was not possible to review the
-reports Logrep can generate (available in \HTML{} only), or to examine the
+documentation).  It requires a web browser to interact with it, so
+automated log processing will be difficult, whereas automated processing is
+a key part of \parsernames{} design.  Sadly all the author's attempts to
+use Logrep failed, as it was unable to access the log files selected; this
+appears to be a bug rather than operator error.  If it was caused by
+operator error, the interface needs improvement as the (minimal)
+instructions were followed as closely as possible, and multiple attempts
+were made.  As parsing failed it was not possible to review the reports
+Logrep can generate (available in \HTML{} only), or to examine the
 (undocumented) format in which it can save extracted data for subsequent
@@
 
-\url{http://www.itefix.no/phpws/index.php} \newline (Last checked
-2007/11/18.)
-
 \subsection{Summary}
 
-\begin{table}[ht]
+There are other programs available which perform basic Postfix log parsing
+(some to a greater level of detail than others), but few attempt to
+correlate log lines by queueid (none correlate by \pid{}) to produce an
+overall record of the journey of each mail through Postfix.  None of the
+reviewed parsers collect the breadth of information gathered by
+\parsername{}, or make it as easy to extend the parser to handle new log
+lines.  Most of the parsers generate a report and immediately discard the
+data extracted from the log files; those that do not discard the data
+typically retain it in a format inaccessible to other tools.  Nearly all of
+the parsers reviewed can produce a report of greater or lesser detail and
+complexity, unlike \parsername{}.  The quality of the documentation offered
+by the subset of parsers that provide some varies from unusable to good;
+none of the parsers provide any documentation on the format of their data
+stores (if they have one).  Fewer than half of the parsers were capable of
+parsing the \numberOFlogFILES{} test log files, and improving or extending
+parsing would have been quite a difficult task for any of the parsers.
+Table \refwithpage{Summary of parsers' features} provides a summary of the
+parsers' features.
+
+The overriding difference between \parsername{} and the other parsers
+reviewed herein is that none of them aim for the high level of
+understanding of Postfix log files achieved by \parsername{}.
+
+\begin{table}[htb]
+    \caption{Summary of parsers' features}
+    \empty{}\label{Summary of parsers' features}
     \begin{tabular}{llllll}
@@
                         & log files?    & store?            & reports?          & quality?       & code?  \\
-    \tableline{}
+        \tableline{}%
         \LMA{}          & No            & Yes               & No                & Poor           & Yes    \\ 
-    Pflogsumm      & Yes         & No              & Partial\dag{} & Good           & Yes    \\
+        Pflogsumm       & Yes           & No                & Partial \dag{}    & Good           & Yes    \\
         Sawmill         & Yes           & Yes               & Searches          & Reasonable     & No     \\
@@
         Isoqlog         & No            & Yes               & No                & Poor           & Yes    \\
-    AWStats        & Partially   & Yes             & Partial\dag{} & Good           & Yes    \\
+        AWStats         & Partially     & Yes               & Partial \dag{}    & Good           & Yes    \\
         Anteater        & No            & No                & No                & Poor           & Yes    \\
-    YAALA          & No          & Yes\ddag{}      & Searches      & Poor           & Yes    \\
+        YAALA           & No            & Yes \ddag{}       & Searches          & Poor           & Yes    \\
         Lire            & Yes           & Yes               & Yes               & Reasonable     & Yes    \\
         Logrep          & No            & Yes               & No                & Poor           & Yes    \\
-    \parsername{}  & Yes         & Yes\nialpha{}   & \nibeta{}     &                & Yes    \\
+        \parsername{}   & Yes           & Yes \nialpha{}    & No \nibeta{}      & XXX            & Yes    \\
         \tableline{}
@@
 
-\dag{} Sections can be omitted from a report, but extra sections can not be
-added.
+    \begin{eqlist}
 
-\ddag{} YAALA only stores the data required to produce the latest report;
-other data will be discarded.
+        \item [\dag{}] Sections can be omitted from a report, but extra
+            sections can not be added.
 
-\nialpha{} \parsername{} is the only parser with documentation for its data
-store.
+        \item [\ddag{}] YAALA only stores the data required to produce the
+            latest report; other data will be discarded.
 
-\nibeta{} \parsername{} defers report generation to subsequent programs,
-but all the necessary data and documentation to produce reports is
-provided.
+        \item [\nialpha{}] \parsername{} is the only parser with
+            documentation for its data store.
 
-\caption{Summary of parsers' features}
-\label{Summary of parsers' features}
+        \item [\nibeta{}] \parsername{} defers report generation to
+            subsequent programs, but all the necessary data and
+            documentation to produce reports is provided.
+
+    \end{eqlist}
+
 \end{table}
-
-There are other programs available which perform basic Postfix log parsing
-(some to a greater level of detail than others), but few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
-overall record of the journey each mail traverses through Postfix.  None of
-the reviewed parsers collect the breadth of information gathered by
-\parsername{}, or make it as easy to extend the parser to handle new log
-lines.  Most other parsers immediately generate a report and discard the
-data extracted from the log files; those that do not discard the data
-retain it in a format inaccessible to other tools.  Nearly all of the
-parsers reviewed can produce a report of greater or lesser detail and
-complexity, unlike \parsername{}.  The quality of the documentation offered
-by the subset of parsers that provide some varies from unusable to good;
-none of the parsers provide any documentation on the format of their data
-stores (if any).  Fewer than half of the parsers were capable of parsing
-the \numberOFlogFILES{} test log files, and improving or extending the
-parsers would have been quite a difficult task for any of the parsers.
-Table \refwithpage{Summary of parsers' features} provides a summary of the
-parsers' features.
-
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim to achieve the high level of
-understanding of Postfix log files achieved by \parsername{}.

### 2008-10-28 doc/logparser-results.tex
First pass through corrections for sections 1-3.

@@
 of mail deliveries and up to an order of magnitude more log lines than is
-usual.  The user specified his mail forwarding as: \newline
-\tab{}\texttt{$\backslash$username, username@domain} \newline This
+usual.  The user specified his mail forwarding as: \newline{}
+\tab{}\texttt{$\backslash$username, username@domain} \newline{} This
 instructs Postfix to deliver the mail to the local user, and also forward a
@@
 
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline
+\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline{}
 \verb!            -e 'print [shuffle(@ARGV)]->[0];'!

### 2008-10-28 doc/logparser.tex
Minor changes.
First pass through corrections for sections 1-3.

@@
 \usepackage{footmisc}
+
+% Aligned list environments.
+\usepackage{eqlist}
 
@@
     actions taken by Postfix to generate the log files.  This paper
-    describes the creation of a parser, documenting the parsing algorithm,
-    explaining the difficulties encountered and the solutions developed,
-    with reference to an implementation that stores data gleaned from the
-    log files in an SQL database.  The gathered data can then be used to
+    describes the creation of a parser for Postfix log files, documenting
+    the architecture developed and the parser implementing it, explaining
+    the difficulties encountered and the solutions developed, with
+    reference to an implementation that stores data gleaned from the log
+    files in an SQL database.  The gathered data can then be used to
     optimise current anti-spam measures, provide a baseline to test new
@@
 
-\bibliographystyle{logparser-bibliography-style}
+%\bibliographystyle{logparser-bibliography-style}
+\bibliographystyle{sgai-2008-elsevier}
 \bibliography{logparser-bibliography}

### 2008-10-28 doc/logparser-architecture.tex
First pass through corrections for sections 1-3.
Second pass through corrections for sections 1-3.

@@
+\newpage{}
+
 \section{Parser Architecture}
@@
 accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{additional complications}.
+with the complications described in \sectionref{complications}.
 
@@
 There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns, nlpip}, a tool used in Computational Linguistics for
+\ATN{}~\cite{atns,nlpip}, a tool used in Computational Linguistics for
 creating grammars to parse or generate sentences.  The resemblance between

### 2008-10-28 doc/logparser-background.tex
Second pass through corrections for sections 1-3.

@@
 This section provides background information helpful in understanding the
-remainder of the document.  It begins with a discussion of the motivation
+remainder of the thesis.  It begins with a discussion of the motivation
 underlying the project, followed by some technical information: the use of
@@
 servers.  The assumptions made in designing and implementing the parser are
-explained, as are the conventions used in this document.
+explained, as are the conventions used in this thesis.
 
@@
 
-This document and the program it describes are part of a larger project to
+This thesis and the program it describes are part of a larger project to
 optimise a mail server's Postfix-based anti-spam restrictions, generate
@@
 
-If the database supports sub-selects percentages can be obtained for the
-top ten restrictions (sample output shown in \tableref{Sample output from
-SQL query showing percentages}):
+If the database supports sub-selects (where the results of one query are
+used in another), percentages can be obtained for the top ten restrictions
+(sample output shown in \tableref{Sample output from SQL query showing
+percentages}):
 
@@
 The database populated by this program provides a simple interface to
-Postfix log files.  Although the interface is a database schema, it is in
-effect quite similar to any other \APIlong{}~(\API{}) provided by shared
-code: it insulates both user and provider of the \API{} from changes in the
-implementation of the \API{}\@.  The algorithm implemented by the parser
-can be improved; support can be added for earlier or later releases of
-Postfix; bugs can be fixed or limitations removed from the parser; these
-changes will not cause the user to be negatively affected.  Statistics and
-graphs can be generated from the database; new restrictions can be tested
-and the results inspected; trends in the fight against spam can emerge from
-historical data saved in the database; the parser remains the same as the
-usage adapts.  Using a database simplifies writing programs that need to
-interact with the data in several ways:
+Postfix log files.  Although the interface is a database schema rather than
+a set of functions provided by shared code, it is in effect quite similar
+to any other \API{}: it provides a stable interface for users of the
+interface, allowing the implementation to be changed without adverse
+effects.  As long as the \API{} is maintained, the parser can be improved
+to handle additional complications; support can be added for earlier or
+later releases of Postfix; bugs can be fixed or limitations removed.
+
+Using a database simplifies writing programs that need to interact with the
+data in several ways:
 
@@
         there may be tools available that reduce or remove the requirement
-        to know \SQL{}; several are available~\cite{sqlite-guis} for
-        SQLite (the default database used by the parser
-        implementation).
+        to know \SQL{}; several are available for SQLite
+        (\url{http://www.sqlite.org/}), the default database used by the
+        parser implementation.
 
@@
 
-A detailed description of \SMTP{} is beyond the scope of this document:
+A detailed description of \SMTP{} is beyond the scope of this thesis:
 introductory guides can be found at~\cite{smtp-intro-01,smtp-intro-02},
@@
 of importance): security, flexibility of configuration, scalability, and
-high performance.  It features extensive, extensible, optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined, and bypassed on a per-client, per-recipient, or per-sender basis,
-allowing varying levels of stricture and permissiveness.  Postfix leverages
-simple lookup tables to support arbitrarily complicated user-defined
-sequences of restrictions and exceptions, with policy
-servers\footnote{Policy servers will be explained in \sectionref{policy
-servers}.} as the ultimate in flexibility.  Administrators can also supply
-their own rejection messages to make it clear to senders why exactly their
-mail was rejected.  Unfortunately this flexibility has a cost: complexity
-in the log files generated.  Although it is easy to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
+high performance.  It features extensive optional anti-spam restrictions,
+allowing an administrator to deploy those restrictions which they judge
+suitable for their site's needs, rather than a fixed set chosen by
+Postfix's author.  These restrictions can be selectively applied, combined,
+and bypassed on a per-client, per-recipient, or per-sender basis, allowing
+varying levels of stricture and permissiveness.  Postfix leverages simple
+lookup tables to support arbitrarily complicated user-defined sequences of
+restrictions and exceptions; policy servers (\sectionref{policy servers})
+provide a simple way to write new restrictions.  Administrators can also
+supply their own rejection messages to make it clear to senders exactly why
+their mail was rejected.  Unfortunately this flexibility has a cost:
+complexity in the log files generated.  Although it is easy to use standard
+Unix text processing utilities to determine the fate of an individual
+email, following the journey an email takes through Postfix can be quite
 difficult.  For most mails the journey is simple and brief, but the
 remaining minority can have quite complex journeys (see
-\sectionref{additional complications} for details).
+\sectionref{complications} for details).
 
@@
 
-\subsection{Conventions used in the document}
+\subsection{Conventions used}
 
@@
 The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably in this document; in general the word used was chosen based
+interchangeably; in general the word used was chosen based
 on the context it appears in.

### 2008-10-28 doc/logparser-conclusion.tex
Second pass through corrections for sections 1-3.

@@
 affecting the existing functionality.  The complications described in
-\sectionref{additional complications} were solved in the order they are
+\sectionref{complications} were solved in the order they are
 described in, and that order closely resembles the frequency with which

### 2008-10-28 doc/logparser-implementation.tex
Second pass through corrections for sections 1-3.

@@
 additional, uncommon paths excluded for the sake of clarity; details of the
-deviations can be found in \sectionref{additional complications}.
+deviations can be found in \sectionref{complications}.
 
@@
 clarity (these extra states are caused by solving the complications
-described in \sectionref{additional complications}).
+described in \sectionref{complications}).
 
@@
 rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this document, mirroring the development of the parser,
+single portion of this section, mirroring the development of the parser.
 The complications are described in the order they were overcome, with
@@
 
-\label{additional complications}
 \label{complications}

### 2008-10-28 doc/logparser-introduction.tex
Second pass through corrections for sections 1-3.

@@
 policy servers may log different messages depending on the characteristics
-of the connection; there are many ways in which the log lines may differ
-between servers, even within the same organisation --- servers may be
+of the connection.  There are many ways in which the log lines may differ
+between servers, even within the same organisation: servers may be
 configured differently, or running different versions of Postfix.  This
 paper documents the difficult process of parsing Postfix log files,
-presenting \PLP{}, a program that parses Postfix log files and places the resulting
-data into a database for later use.  The gathered data can then be used to
-optimise current anti-spam measures, provide a baseline to test new
-anti-spam measures against, or to produce statistics showing how effective
-those measures are.  Numerous other uses are possible for such data:
-improving server performance by identifying troublesome destinations and
-reconfiguring appropriately; identifying regular high volume uses (e.g.\
-customer newsletters) and restricting those uses to off-peak times;
+presenting \PLP{}, a program that parses Postfix log files and places the
+resulting data into a database for later use.  The gathered data can then
+be used to optimise current anti-spam measures, provide a baseline to test
+new anti-spam measures against, or to produce statistics showing how
+effective those measures are.  Numerous other uses are possible for such
+data: improving server performance by identifying troublesome destinations
+and reconfiguring appropriately; identifying regular high volume uses
+(e.g.\ customer newsletters) and restricting those uses to off-peak times;
 detecting virus outbreaks that propagate via email; as a base for billing
@@
 \vspace{1em}\noindent\textbf{Layout:}
+
+XXX THIS ALL NEEDS TO BE CHECKED AND UPDATED\@.
 
@@
 
-Section~\ref{conclusion} contains the document's conclusion, describing the
-results of the research, design and implementation of the parser.
+Section~\ref{conclusion} contains the conclusion of the thesis, describing
+the results of the research, design and implementation of the parser.
 
 The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this document.  Also
+the algorithm, writing the program, and preparing this thesis.  Also
 listed are some additional resources expected to be helpful in
-understanding \SMTP{}, Postfix, anti-spam techniques, or the document.
+understanding \SMTP{}, Postfix, anti-spam techniques, or the thesis.
 
-Appendix~\ref{Glossary} provides a glossary of terms used in the document.
+Appendix~\ref{Glossary} provides a glossary of terms used in the thesis.
 
-Appendix~\ref{Acronyms} provides a list of acronyms used in the document.
+Appendix~\ref{Acronyms} provides a list of acronyms used in the thesis.

### 2008-10-28 doc/logparser-other-parsers-review.tex
Second pass through corrections for sections 1-3.

@@
 \section{State of the Art Review}
-
-\label{other-parsers}
 
@@
 
+XXX UPDATE THE BLURBS\@.
+
 Ten other parsers, starting with the only previously published research in
 this area that the author is aware of, have been reviewed as part of the
-background research for this project.  None of the parsers reviewed perform
-the kind of advanced parsing and correlation required for this project; all
-except one (XXX IS THIS ACCURATE\@?) are intended to perform a specific
-parsing and reporting task, whereas \PLP{} is designed to be a generic
-parser, extracting data and leaving generation of reports from the data to
-other programs.  Some parsers save the data extracted to a data store, but
-the majority discard all data once they have finished running, making
-historical analysis impossible.  The parsers reviewed all produce a report
-of varying complexity and detail, whereas this project's parser does not
-produce a report at all; that responsibility is deferred to a separate
-program, to be developed later.  The parser developed for this project is
-designed to enable much more detailed log analysis by providing a stable
-platform for subsequent programs to develop upon.
+background research for this project.  Each of the parsers was tested with
+the \numberOFlogFILES{} test log files described in \sectionref{parser
+efficiency}.  None of the parsers reviewed perform the kind of advanced
+parsing and correlation required for this project; all except one (XXX IS
+THIS ACCURATE\@?) are intended to perform a specific parsing and reporting
+task, whereas \PLP{} is designed to be a generic parser, extracting data
+and leaving generation of reports from the data to other programs.  Some
+parsers save the data extracted to a data store, but the majority discard
+all data once they have finished running, making historical analysis
+impossible.  The parsers reviewed all produce a report of varying
+complexity and detail, whereas this project's parser does not produce a
+report at all; that responsibility is deferred to a separate program, to be
+developed later.  The parser developed for this project is designed to
+enable much more detailed log analysis by providing a stable platform for
+subsequent programs to develop upon.
 
@@
 
+XXX IS THE BDB STRUCTURE DOCUMENTED\@?
+
+XXX DESCRIBE THE \CSV{} FORMAT\@.
+
+XXX DESCRIBE THE MYSQL SCHEMA\@.
+
+IS THE SAME DATA STORED FOR ALL FORMATS\@?
+
 \label{prior art}
@@
 standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL or Berkeley DB database.  The decision to
-support both \CSV{} and Berkeley DB appears to have been a serious
-limitation: \CSV{} is a very simple format where a record is stored in a
-single line, with fields separated by a comma or other punctuation symbol.
-Problems with \CSV{} files include the need to escape separators in the
-data stored, providing multiple values for a field (e.g.\ multiple
-recipients), and adding new fields.  There is no mechanism to document the
-fields or the separator, unlike \SQL{} databases where every database
-includes a schema naming the fields and the type of data they store
-(integer, text, timestamp, etc.).
+\CSV{} file and either a MySQL (\url{http://www.mysql.com/}) or Berkeley DB
+(\url{http://www.oracle.com/database/berkeley-db/index.html}) database.
+XXX SHOULD THOSE URLS BE REFERENCES INSTEAD\@?  The decision to support
+both \CSV{} and Berkeley DB appears to have been a serious limitation:
+\CSV{} is a very simple format where a record is stored in a single line,
+with fields separated by a comma or other punctuation symbol.  Problems
+with \CSV{} files include the need to escape separators in the data stored,
+providing multiple values for a field (e.g.\ multiple recipients), and
+adding new fields.  There is no mechanism to document the fields or the
+separator, unlike \SQL{} databases where every database includes a schema
+naming the fields and the type of data they store (integer, text,
+timestamp, etc.).
 
@@
 
-\LMA{} stores the following date: time and date, hostname and \IP{} address
-of client, \IP{} of server, sender and recipient addresses, \SMTP{} code,
-and size (for accepted mails only).  It omits XXX XXX\@.  Handling of
+\LMA{} stores the following date: time and date, client hostname and \IP{}
+address, server \IP{} address, sender and recipient addresses, \SMTP{}
+code, and size (for accepted mails only).  It omits XXX XXX\@.  Handling of
 multiple recipients, \SMTP{} codes or remote servers\footnote{A single mail
@@
 client \IP{} address, sender address, or date respectively.  DAILY-REJECT
-initially failed with an error message from the Perl interpreter; after
-making some corrections to the code it worked, and extracted the \CSV{}
-lines for the specified day where the \SMTP{} code signifies a rejection.
-All of these reports are trivially simple to produce from the \CSV{} file
-using the standard Unix tool \texttt{awk}; the most complicated,
-DAILY-REJECT, is merely:
+initially failed with an error message (XXX ADD THE ERROR MESSAGE) from the
+Perl interpreter; after making some corrections to the code it worked, and
+extracted the \CSV{} lines for the specified day where the \SMTP{} code
+signifies a rejection.  All of these reports are trivially simple to
+produce from the \CSV{} file using the standard Unix tool \texttt{awk}; the
+most complicated, DAILY-REJECT, is merely:
 
@@
 
-    \item Total number and bytes of mails accepted, delivered and rejected.
-
-    \item Total number of sender and recipient addresses and domains.
+    \item Total number of mails accepted, delivered, and rejected.  Total
+        size of mails accepted and delivered.  Total number of sender and
+        recipient addresses and domains.
 
@@
         average delay, maximum delay, and bytes delivered.  For received
-        mail: per-domain totals for mails received and bytes received.
+        mail: per-domain totals for size and number of mails received.
 
-    \item Number of mails and bytes sent and received for each address.
+    \item Number and size of mails sent and received for each address.
 
@@
 
+XXX CHECK ALL THIS AGAIN\@; GIVE EXAMPLE PRICES\@.
+
 Sawmill is a general purpose commercial product that parses 687 log file
@@
 the user bypassing the program and generating their own reports.
+
+XXX THE BLURB MENTIONS \SQL{} DB\@; CAN THAT BE ACCESSED SEPARATELY, IS THE
+FORMAT DOCUMENTED, ETC\@?
 
@@
 
+XXX THE FIRST PARAGRAPH NEEDS TO BE REWRITTEN\@; PICK POINTS, THEN COMPARE
+AND CONTRAST\@.
+
 Splunk aims to index all an organisation's log files, providing a
@@
 events, then select a particular host, and finally a specific address),
-though it is not possible to search on partial words.
+though it is not possible to search on partial words.  All searches
+performed using the indexed data returned reasonable results.  The web
+interface is quite attractive and simple to use when searching, but as an
+administrator it seems unnecessarily difficult to perform simple tasks.
 
-The web interface is quite attractive and simple to use when searching, but
-as an administrator it seems unnecessarily difficult to perform simple
-tasks.  When testing it took roughly 30 minutes to add a single log file to
-be indexed (log files must be indexed before searching), with the downside
+When testing it took roughly 30 minutes to add a single log file to be
+indexed (log files must be indexed before searching), with the downside
 that the log file was copied into a spool directory, doubling the disk
@@
 Repeated attempts enqueued the same 40 log files, without explaining why
-the others were excluded.\footnote{The log files appear to have only been
-indexed once; presumably Splunk keeps track of the files it has indexed and
-discards requests to index files for a second time.  This may or may not be
-a useful feature for \parsername{}.} There did not appear to be an option
-to order the log files to ensure they would be processed in the order they
-were created.  Subsequently the author was successful in adding a single
+the others were excluded.\footnote{The log files appear to have been
+indexed once only; presumably Splunk keeps track of the files it has
+indexed and discards requests to index files for a second time.  This may
+or may not be a useful feature for \parsername{}.} There did not appear to
+be an option to order the log files to ensure they would be processed in
+the order they were created, though this may not be necessary or beneficial
+with Splunk.  Subsequently the author was successful in adding a single
 file at a time using the \CLI{}; a simple loop to add all desired log files
@@
 this may or may not be useful behaviour.  Splunk did not appear to have any
-difficulty in parsing the log files, once instructed to do so.  All
-searches performed using the indexed data returned reasonable results.
+difficulty in parsing the log files, once instructed to do so.
+\parsername{} parses the logs it is instructed to parse, in the order
+given; periodic parsing of logs is a task an administrator can easily
+achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.
 
@@
 day to 500MB, though a trial Enterprise licence is available that allows
-indexing of up to 5GB of data per day.  The cheapest licensed version costs
-\$5000 (plus \$1000 support), and still limits the volume of data indexed
-per day to 500MB\@.  Typical log file sizes for a small scale mail server
-are given in \sectionref{parser efficiency}.
+indexing of up to 5GB of data per day.  In 2007 the cheapest licensed
+version cost \$5000 plus \$1000 support, and limited the volume of data
+indexed per day to 500MB\@.  Prices were removed from the Splunk website
+during 2008; now Splunk's sales team must be contacted for a quote.
+Typical log file sizes for a small scale mail server are given in
+\sectionref{parser efficiency}.
 
@@
 find Postfix reports, or additional reports for any other log file types.
+
+\newpage{}
 

### 2008-10-28 doc/logparser-results.tex
Second pass through corrections for sections 1-3.

@@
 number of rules required to parse the \numberOFlogFILES{} log files used
-when generating the results in this document is \numberOFrulesMINIMUM{},
+when generating the results in this thesis is \numberOFrulesMINIMUM{},
 \numberOFrulesMINIMUMpercentage{} of the full ruleset.  The full ruleset is
@@
 the best way to detect mails that were not handled properly (many of the
-complications discussed in \sectionref{additional complications} were
+complications discussed in \sectionref{complications} were
 detected in this way).  The parser issues warnings when it detects any
@@
     \item When dealing with the complications described in
-        \sectionref{complications} and \sectionref{additional
-        complications} the solutions are as specific and restrictive as
-        possible, with the goal of minimising the number of false
-        positives.  In addition the solution to the \textit{Out of order
-        log files\/} complication described in \sectionref{out of order log
-        lines} imposes conditions that each reassembled mail must comply
-        with to be acceptable.
+        \sectionref{complications} the solutions are as specific and
+        restrictive as possible, with the goal of minimising the number of
+        false positives.  In addition the solution to the \textit{Out of
+        order log files\/} complication described in \sectionref{out of
+        order log lines} imposes conditions that each reassembled mail must
+        comply with to be acceptable.
 

### 2008-10-28 doc/logparser-architecture.tex
Corrections to section 4: mainly just a big bunch of TODO items.

@@
-\newpage{}
-
 \section{Parser Architecture}
@@
 
-\subsection{Parser design}
+XXX NEEDS AN INTRODUCTION ONCE CONTENT HAS SETTLED\@.
+
+XXX MERGE SOME OF THE PAPER'S CONTENT IN HERE\@.
+
+\subsection{Parser architecture and design}
 
@@
 It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that log files produced by Postfix are not fixed; they vary
-widely from host to host, depending on the set of restrictions chosen by
-the administrator.  With this in mind, one of the parser's design aims was
-to make adding new rules as easy as possible, to enable administrators to
-properly parse their log files.  To enable this the parser is divided into
-three parts:
+background}) that log files produced by Postfix vary widely from host to
+host, depending on the set of restrictions chosen by the administrator.
+With this in mind, one of the parser's design aims was to make adding new
+rules as easy as possible, to enable administrators to properly parse their
+own log files.  To enable this the architecture is divided into three
+parts:
+
+XXX UPDATE SECTION REFERENCES BELOW\@: THEY NEED TO REFER TO BOTH THE
+ARCHITECTURE AND THE IMPLEMENTATION\@.
 
 \begin{description}
+
+    \item [framework] The framework is responsible for loading rules,
+        managing data structures, finding the rule that matches each log
+        line, invoking the correct action, etc\@.  The framework is
+        described in detail in \sectionref{framework}.  XXX EXTEND
+        FRAMEWORK SECTION IF NECESSARY\@.
 
@@
 
-    \item [framework] The framework is responsible for loading rules,
-        managing data structures, finding the rule that matches each log
-        line, invoking the correct action, etc\@.  The framework is
-        described in detail in \sectionref{framework}.  XXX EXTEND
-        FRAMEWORK SECTION IF NECESSARY\@.
-
 \end{description}
@@
 It may help to think of the rules and actions as components that plug into
-the framework.
+the framework.  The framework tries each rule in turn until is finds one
+which matches the log line, then invokes the action specified by the rule.
 
@@
 
-XXX MERGE THE NEXT TWO PARAGRAPHS\@.
+XXX MERGE THE NEXT TWO PARAGRAPHS\@.  PULL CONTENT FROM THE PAPER\@.
 
@@
 
-\subsection{Rule characteristics}
+\subsection{Rules}
+
+XXX TO BE WRITTEN\@.  TAKE CONTENT FROM THE PAPER AND POSSIBLY FROM THE
+IMPLEMENTATION SECTION\@.
+
+\subsubsection{Rule characteristics}
 
@@
 
-\subsection{Overlapping rules}
+\subsubsection{Overlapping rules}
 
@@
 
+XXX PULL CONTENT FROM THE PAPER FOR HERE\@; IT'S BETTER WRITTEN\@.
+
 \label{framework}
@@
 
-XXX TO BE WRITTEN\@.
-
-\subsection{Rules}
-
-XXX TO BE WRITTEN\@.
+XXX TO BE WRITTEN\@.  START WITH THE CONTENT FROM THE PAPER, THEN CHECK THE
+IMPLEMENTATION SECTION\@.

### 2008-10-28 doc/logparser-implementation.tex
First half of the corrections to section 5.

@@
-\newpage
 \section{Postfix Parser Implementation}
 
+XXX WRITE AN INTRODUCTION\@.
+
+XXX UPDATE REFERENCES WHEN ARCHITECTURE AND IMPLEMENTATION ARE MORE FLESHED
+OUT\@.
+
+XXX FIGURE OUT WHEN TO USE ACTIONS AND WHEN TO USE ALGORITHM, ETC\@.
+
+XXX THIS MIGHT BE USEFUL SOMEWHERE\@:
+
+Whereas the rules are quite simple and each rule is completely independent
+of the other rules, the algorithm is significantly more complicated and
+highly internally interdependent.  The algorithm deals with all the
+complications of parsing, the eccentricities and oddities of Postfix log
+files, presenting the resulting data in a normalised, simple to use
+representation.  The algorithm's task is to follow the journey each mail
+takes through Postfix, combining the data extracted by rules into a
+coherent whole, saving it in a useful and consistent form, and performing
+housekeeping duties.
+
+Please refer to \sectionref{parser design} for a discussion of why the
+rules, actions and framework have been separated in the parser's design.
+In this section algorithm can be taken to mean the combination of framework
+and actions.
+
 \subsection{Assumptions}
 
-The algorithm described and the program implementing it make a small number
-of (hopefully safe and reasonable) assumptions:
+\parsername{} makes a small number of (hopefully safe and reasonable)
+assumptions:
 
@@
 In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Although the
-source code is available,\footnote{Reading and understanding the source
-code would require a significant investment of time: there are 375,750
-lines of code, documentation, etc.\ in Postfix 2.5.1's 17MB of source
-code.} there are advantages to treating Postfix as a black box during
-parser development:
+black box program based solely on its inputs and outputs.  There are advantages to
+treating Postfix as a black box during parser development:
 
 \begin{itemize}
+
+    \item Reading and understanding the source code would require a
+        significant investment of time: Postfix 2.5.5 has 17MB of source
+        code.
 
@@
 
-    \item The parser acts as a second source of information, with the
-        information gathered from empirical evidence.  A separate project
-        could compare the empirical knowledge inherent in the parsing
-        algorithm with the documentation and source code of Postfix to see
-        how closely the two agree.
+    \item The parser acts as a second source of information about Postfix's
+        operation, using information gathered from empirical evidence.  A
+        separate project could compare the empirical knowledge inherent in
+        the parsing algorithm with the documentation and source code of
+        Postfix to see how closely the two agree.
 
@@
 \label{database schema}
+
+XXX ADD A DIAGRAM OF THE DATABASE SCHEMA\@.
 
@@
 
-The database schema can be conceptually divided in two: the rules, which
-are used to parse log files, and the data saved from the parsing of log
-files.  Rules have the fields required to parse the log lines, extract data
-to be saved, and the action to be executed; they also have several fields
-that aid the user in understanding the meaning of the log lines parsed by
-each rule.  The rules are described in detail in \sectionref{rules} but the
+The database schema can be conceptually divided in two: the rules used to
+parse log files, and the data saved from the parsing of log files.  Rules
+have the fields required to parse the log lines, extract data to be saved,
+and specify the action to be executed; they also have several fields that
+aid the user in understanding the meaning of the log lines parsed by each
+rule.  The rules are described in detail in \sectionref{rules} but the
 fields are covered in \sectionref{rule attributes}.
@@
 An important but easily overlooked benefit of storing the rules in the
-database is the connection between rules and results: if more information
-is required when examining a result, the rule that produced the result is
-available for inspection --- each result references the rule that created
-it.  There is no ambiguity about which rule resulted in a particular
-result, eliminating one potential source of confusion.
+database is the link between rules and results: if more information is
+required when examining a result, the rule that produced the result is
+available for inspection: each result references the rule that created it.
+There is no ambiguity about which rule resulted in a particular result,
+eliminating one potential source of confusion.
+
+A clear, comprehensible schema is essential when using the extracted data;
+it is more important when using the data than when storing it, because
+storing the data is a write-once operation, whereas utilising the data
+requires frequent searching, sorting and manipulation of the data to
+produce customised reports or statistics.
 
@@
 \label{rule attributes}
+
+XXX SHOULD I USE TEXTTT OR SIMILAR ON FIELD NAMES\@?
 
@@
 
-\begin{description}
+\begin{eqlist}
 
     \item [id] A unique identifier for each rule that other tables can use
-        to refer to a specific rule.
+        when referring to a specific rule.
 
@@
 
+        XXX SHOULD THIS LIST BE HERE\@?  MAYBE MOVE TO AFTER THE FIELD
+        LIST\@?  DO I NEED THIS LIST AT ALL\@?
+
         \begin{description}
@@
             \item [BOUNCED] The mail has bounced, because of a mail loop,
-                delivery failure, or five day timeout.
+                delivery failure, or five day timeout.\footnote{As with
+                much of Postfix's behaviour, five days is the default value
+                but can be changed by the administrator if they choose.}
 
@@
                 process of accepting, because it was either larger than the
-                limit set by the administrator, or the client timed out or
-                disconnected.
+                size limit set by the administrator, or the client timed
+                out or disconnected.
 
             \item [EXPIRED] The mail has been in the queue for five
-                days\footnote{As with much of Postfix's behaviour, this is
-                the default value but can be changed by the administrator
-                if they choose.} without successful delivery.  A bounce
+                days\footnotemark[\value{footnote}]XXX CHECK IF THIS
+                FOOTNOTE MAKES SENSE without successful delivery.  A bounce
                 mail will be generated and sent to the sender address.
@@
 
-    \item [regex] The \regex{} to match the log line against.  The \regex{}
-        will first have several keywords expanded: this simplifies reading
-        and writing rules; avoids needless repetition of complex \regex{}
-        components; allows the components to be corrected and/or improved
-        in one location; and makes each \regex{} largely self-documenting.
-
-        For efficiency the keywords are expanded and every rule's \regex{}
-        is compiled before attempting to parse the log file --- otherwise
-        each \regex{} would be recompiled each time it was used, resulting
-        in a large, data dependent slowdown.  Efficiency concerns are
-        discussed in \sectionref{parser efficiency}, with caching compiled
-        \regexes{} covered in \sectionref{Caching compiled regexes}.
-
-    \item [result\_cols, connection\_cols] Specifies how the fields in the
-        log line will be extracted.  The format is: \newline{} \tab{}
-        \texttt{smtp\_code = 1; recipient = 2, sender = 4;} \newline{} i.e.\
-        semi-colon or comma separated assignment statements, with the
-        variable name on the left and the matching capture from the
-        \regex{} on the right hand side.  Data specified by
-        \texttt{result\_cols} will be entered in the \texttt{results}
-        table, whereas \texttt{connection\_cols} populates the
-        \texttt{connection} table.  The list of acceptable variable names
-        is:
-
-        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
-        \newline{} \tab{} server\_hostname} and \texttt{helo}.\newline{}
-        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
-        \newline{} \tab{} size,} and \texttt{data}
-
-        Additionally \texttt{child} and \texttt{pid} are accepted and used
-        respectively by the \texttt{TRACK}, \texttt{BOUNCE} and
-        \texttt{SMTPD\_DIED} actions, but do not cause data to be entered
-        in the database.
-
-    \item [result\_data, connection\_data] Sometimes rules need to supply a
-        piece of data that is not present in the log line: e.g.\ setting
-        \texttt{smtp\_code} when mail is accepted.  The format and allowed
-        variables are the same as for \texttt{result\_cols} and
-        \texttt{connection\_cols}, except that arbitrary
-        data\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be used.  This feature is intended for use with small
-        amounts of data rather than large amounts in any one rule, so
-        dealing with escape sequences was deemed unnecessary.} is permitted
-        on the right hand side of the assignment.
+    \item [regex] XXX THIS NEEDS TO BE UPDATED ONCE AUTOMATIC DATA
+        EXTRACTION IS FINISHED\@.  The \regex{} to match the log line
+        against.  The \regex{} will first have several keywords expanded:
+        this simplifies reading and writing rules; avoids needless
+        repetition of complex \regex{} components; allows the components to
+        be corrected and/or improved in one location; and makes each
+        \regex{} largely self-documenting.  For efficiency the keywords are
+        expanded and every rule's \regex{} is compiled before attempting to
+        parse the log file --- otherwise each \regex{} would be recompiled
+        each time it was used, resulting in a large, data dependent
+        slowdown.  Efficiency concerns are discussed in \sectionref{parser
+        efficiency}, with caching compiled \regexes{} covered in
+        \sectionref{Caching compiled regexes}.
+
+    \item [result\_data] Sometimes rules need to save a result that is not
+        present in the log line: e.g.\ setting \texttt{smtp\_code} when
+        mail is accepted.  The format is: \newline{} \tab{} \texttt{XXX
+        UPDATE EXAMPLE\@; smtp\_code = 550; recipient = 2, sender = 4;}
+        \newline{} i.e.\ semi-colon or comma separated assignment
+        statements, with the variable name on the left and the value on the
+        right hand side.\footnote{Commas and semi-colons cannot be escaped
+        and thus cannot be used.  This feature is intended for use with
+        small amounts of data rather than large amounts in any one rule, so
+        dealing with escape sequences was deemed unnecessary.}  Any field
+        in the results table can be set in this way; additionally
+        \texttt{child} and \texttt{pid} are accepted and used respectively
+        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
+        actions, but do not cause data to be entered in the database.
+
+    \item [connection\_data] The connections table equivalent of
+        \texttt{result\_data}.
 
@@
 
-    \item [queueid] Specifies the capturing field in the \regex{}
-        containing the queueid, or zero if the log line does not contain a
-        queueid.  Many log lines will not contain a queueid, e.g.\
-        rejections logged before a mail has been accepted (a queueid will
-        not have been allocated), or log lines that are not tied to one
-        particular mail.
-
     \item [hits] This counter is maintained for every rule and incremented
         each time the rule successfully matches.  At the start of each run
-        the program sorts the rules in descending order of hits, and at the
-        end of the run updates every rule's hits field in the database.
+        the parser sorts the rules in descending order of hits, and at the
+        end of the run it updates every rule's hits field in the database.
         Assuming that the distribution of log lines is reasonably
-        consistent between log files, rules matching more commonly
-        occurring log lines will be tried before rules matching less
-        commonly occurring log lines, reducing the program's execution
-        time.  Rule ordering for efficiency is discussed in
-        \sectionref{rule ordering for efficiency}.
+        consistent across log files, rules matching more commonly occurring
+        log lines will be tried before rules matching less commonly
+        occurring log lines, reducing the parser's execution time.  Rule
+        ordering for efficiency is discussed in \sectionref{rule ordering
+        for efficiency}.
 
@@
 
-    \item [cluster\_group] A reference to the \texttt{cluster\_group}
-        table.  That table is used by the Decision Tree algorithm described
-        in a separate document.
-
-\end{description}
+%    \item [cluster\_group] A reference to the \texttt{cluster\_group}
+%        table.  That table is used by the Decision Tree algorithm described
+%        in a separate document.
+
+\end{eqlist}
 
@@
 
-\begin{description}
+\begin{eqlist}
 
@@
 
-XXX REWRITE THE SENTENCE AND FOOT NOTE\@.
-
-Every interesting\footnote{All log lines are interesting unless the
-postfix\_action of the rule that parses them is \texttt{INFO} or
-\texttt{IGNORE}, as described in \sectionref{actions-in-detail}.} log line
-will have an entry in the results table, e.g.\ rejecting an \SMTP{}
-command, delivering a mail, or bouncing a mail (see
-\sectionref{postfix_action} for typical values of
-\texttt{postfix\_action}).  Each row is associated with a single
-connection, though there may be many results per connection.
-
-\begin{description}
+Every log line a row in the results table, unless the
+\texttt{postfix\_action} field of the rule which parsed it was INFO or
+IGNORE\@.  Each row in the results table is associated with a single
+connection, and there may be many results per connection.
+
+\begin{eqlist}
 
@@
         enforcing a restriction that would reject an \SMTP{} command --- a
-        facility that is quite useful for testing new restrictions.  This
+        mechanism that is quite useful for testing new restrictions.  This
         field will be 1 if the log line parsed was a warning rather than a
@@
     \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present for a rejection or final
-        delivery; results initially missing an \SMTP{} code will duplicate
-        the \SMTP{} code of other results in the connection.  Some final
-        delivery log lines do not contain an \SMTP{} code: in those cases
-        the code is specified by the rule, based on the success or failure
-        represented by the log line.
+        general an \SMTP{} code is only present in a log line for a
+        rejection or final delivery; results initially missing an \SMTP{}
+        code will duplicate the \SMTP{} code of other results in that
+        connection.  Some final delivery log lines do not contain an
+        \SMTP{} code: in those cases the code is specified by the rule,
+        based on the success or failure represented by the log line.
 
@@
     \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail, but only one per result.
+        recipient addresses per mail or connection.
 
@@
     \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from an \DNSBL{}\@.
-
-    \item [timestamp] The time the log line was logged, in seconds since
-        the epoch.
-
-\end{description}
-
-\subsubsection{Summary}
-
-XXX DO I NEED A SUMMARY HERE\@?
-
-The table containing the rules used by the parser and both tables
-containing the data extracted from the Postfix log files were described,
-with the purpose of each field discussed in detail.  A clear,
-comprehensible schema is essential when using the extracted data; it is
-more important when using the data than when storing it, because storing
-the data is a write-once operation, whereas utilising the data requires
-frequent searching, sorting and manipulation of the data to produce
-customised reports or statistics.
+        fields, e.g.\ the rejection message from a \DNSBL{}\@.
+
+    \item [timestamp] The time at which the log line was logged, in seconds
+        since the epoch.
+
+\end{eqlist}
 
@@
 
+XXX NEED AN INTRODUCTION\@.
+
 \subsubsection{Example rule}
@@
 This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the appropriate \DNS{} entries are
-missing, i.e.\ mail could not be delivered to the sender's address (for
-full details see~\cite{reject-unknown-sender-domain}).
-
-The example rule below matches the following log line:
+mail from a sender address because the domain in the sender address does
+not have an A, AAAA, or MX \DNS{} entry, i.e.\ mail could not be delivered
+to the sender's address (for full details
+see~\cite{reject-unknown-sender-domain}).
+
+Example log line matched by this rule:
+
+% RFC 3330 says that 192.0.2.0/24 is reserved for example use.
 
 \begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[10.1.1.1]: 550
+NOQUEUE: reject: RCPT from example.com[192.0.2.1]: 550
   <foo@example.com>: Sender address rejected: Domain not found;
@@
 program             & \daemon{smtpd}                                    \\
+XXX                 & UPDATE REGEX WHEN AUTOMATIC SAVING IS FINISHED    \\
 regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
@@
                     & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_cols        & recipient = 6; sender = 5                         \\
-connection\_cols    & helo = 7                                          \\
 result\_data        &                                                   \\
 connection\_data    &                                                   \\
-action              & REJECTION                                         \\
-queueid             & 1                                                 \\
+action              & DELIVERY\_REJECTED                                \\
 hits                & 0                                                 \\
@@
 priority            & 0                                                 \\
-cluster\_group      & 400                                               \\
+%cluster\_group      & 400                                               \\
 
@@
 
-Additional data will be captured automatically when the \regex{} contains 
-\_\_RESTRICTION\_START\_\_, hence the capture numbers in result\_cols and
-connection\_cols start at 5.  The various fields are used as follows;
+Fields matched by the regex will be automatically saved to the results and
+connections tables.
 
@@
 
-    \item [name, description, restriction\_name and postfix\_action:] are
-        not \newline{} used by the algorithm, they serve to document the rule
+    \item [name, description, restriction\_name, and postfix\_action:] are
+        not \newline{} used by the parser, they serve to document the rule
         for the user's benefit.
@@
     \item [program and regex:] If the program in the rule equals the
-        program that logged the log line, a match using the rule's
-        \regex{} will be attempted against the log line; if the match is
-        successful the action will be executed, if not the next rule will
-        be tried.  If the program-specific rules do not match the log line,
-        the parser will fall back to generic rules; if those rules are
-        unsuccessful a warning will be issued.
+        program in the log line, a match using the rule's \regex{} will be
+        attempted against the log line; if the match is successful the
+        action will be executed, if not the next rule will be tried.  If
+        the program-specific rules do not match the log line, the parser
+        will fall back to generic rules; if those rules are unsuccessful a
+        warning will be issued.
 
     \item [action:] The action to be executed if the \regex{} matches
-        successfully (see \sectionref{actions-in-detail} for full details).
-
-    \item [result\_cols, connection\_cols, result\_data and
-        connection\_data:] are \newline{} used by the action to extract and
-        save data matched by the \regex{}.
-
-    \item [queueid:] The index of the capture in the \regex{} that
-        supplies the queueid, or zero if the log line does not contain a
-        queueid.  This allows the correct mail to be found by queueid and
-        actions performed on it.
-
-    \item [hits, hits\_total and priority:] hits and priority are used in
+        successfully.  See \sectionref{actions-in-detail} for details of
+        the actions implemented by the parser, \sectionref{actions in
+        architecture} for the role of actions in the parser architecture.
+
+    \item [result\_data and connection\_data:] are used to provide XXX
+
+    \item [hits, hits\_total, and priority:] hits and priority are used in
         ordering the rules (see \sectionref{rule ordering for efficiency});
@@
         parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the algorithm.
-
-    \item [cluster\_group] The cluster\_group attribute is used by the
-        Decision Tree algorithm described in a separate document; the
-        parser does not use it in any way.
+        but is otherwise unused by the parser.
+
+%    \item [cluster\_group] The cluster\_group attribute is used by the
+%        Decision Tree algorithm described in a separate document; the
+%        parser does not use it in any way.
 
@@
 parsing new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions is difficult, but adding new rules to
-parse new rejection messages is trivial, and also occurs much more
-frequently.  The implementation provides a program to ease the process of
-creating new rules from unparsed log lines, based on the algorithm
-developed by Risto Vaarandi~\cite{risto-vaarandi} for his
-\SLCT{}~\cite{slct-paper}.  The differences between the two algorithms will
-be outlined as part of the general explanation below.
-
-The core of the algorithm is quite simple: log lines are generally created
-by substituting variable words into a fixed pattern, and analysis of the
-frequency with which each word occurs can be used to determine whether the
-word is variable or part of the fixed pattern.  This classification can be
-used to group similar log lines and generate a \regex{} to match each group
-of log lines.
+parsing actions: adding new actions can be difficult, but adding new rules
+to parse new rejection messages is trivial, and occurs much more
+frequently.  A program is provided to ease the process of creating new
+rules from unparsed log lines, based on the algorithm developed by Risto
+Vaarandi for his \SLCT{}~\cite{slct-paper}.  The differences between the
+two algorithms will be outlined as part of the general explanation below.
+
+The core of the \SLCT{} algorithm is quite simple: log lines are generally
+created by substituting variable words into a fixed pattern, and analysis
+of the frequency with which each word occurs can be used to determine
+whether the word is variable or part of the fixed pattern.  This
+classification can be used to group similar log lines and generate a
+\regex{} to match each group of log lines.
 
@@
         gained when writing rules and performs a large number of
-        substitutions on the log input lines, replacing commonly occurring
+        substitutions on the input log lines, replacing commonly occurring
         variable terms (e.g.\ email addresses, \IP{} addresses, the
@@
         this step is to utilise existing knowledge to create more accurate
-        \regexes{}.  The new log lines are written to a temporary file,
-        which all subsequent stages use instead of the original input file.
+        \regexes{}; it replaces a large number of variable words with fixed
+        words, improving the subsequent classification of words as fixed or
+        variable.  The new log lines are written to a temporary file, which
+        all subsequent stages use instead of the original input file.
 
@@
         word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, and there is no need to maintain a counter
-        for it, reducing the number of counters and thus the program's
-        memory consumption.
+        have a high frequency, so there is no need to maintain a counter
+        for it, reducing the number of counters required and thus the
+        program's memory consumption.
 
     \item [Calculate word frequencies]  The position of words within a log
-        line is important: a common word does not indicate similarity
-        between log lines unless it occupies the same position within both
-        log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two words rather than
+        line is important: a word occurring in two lines does not indicate
+        similarity unless it occupies the same position within both log
+        lines.\footnote{If a variable term within a line contains spaces,
+        it will appear to the algorithm as two or more words rather than
         one.  This will alter the position of subsequent words, so a word
@@
         with this possibility.}  The algorithm maintains a counter for each
-        \textit{(word, word's position within the log line)\/} tuple,
+        \texttt{(word, word's position within the log line)} tuple,
         incrementing it each time that word occurs in that position.
@@
         the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm.
+        memory requirements of the algorithm at a cost of increased CPU
+        usage.
 
@@
         the modified algorithm omits the hashing step and maintains
-        counters for all tokens. In addition most uncommon or infrequently
+        counters for all tokens. In addition most of the infrequently
         occurring words will have been substituted with keywords during the
@@
     \item [Classify words based on their frequency]  The frequency of each
-        \textit{(word, word's position within the log line)\/} tuple is
+        \texttt{(word, word's position within the log line)} tuple is
         checked: if its frequency is greater than the threshold supplied by
@@
         replaced by \texttt{.+}, which means to match zero or more of any
-        character.
+        character.  XXX SHOULD I USE \verb![\S]+! INSTEAD\@?
 
     \item [Build regexes]  The words are reassembled to produce a \regex{}
-        matching the log line, and a counter maintained for each \regex{}.
-        Contiguous sequences of \texttt{.+} in the newly reassembled
-        \regexes{} are collapsed to a single \texttt{.+}; any resulting
-        duplicate \regexes{} are combined, and their counters added
-        together.  If the \regex{}'s counter is lower than the threshold
-        supplied by the user the \regex{} is discarded; this second
-        threshold is independent of the threshold used to differentiate
-        between fixed and variable words, but once again 1\% of input log
-        lines is a good starting point; obviously the threshold depends on
-        the number and content of input log lines.  The new \regexes{} are
+        matching the log line, and a counter is maintained for each
+        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
+        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
+        resulting duplicate \regexes{} are combined, and their counters
+        added together.  If a \regex{}'s counter is lower than the
+        threshold supplied by the user the \regex{} is discarded; this
+        second threshold is independent of the threshold used to
+        differentiate between fixed and variable words, but once again 1\%
+        of log lines is a good starting point.  The new \regexes{} are
         displayed for the user to add to the database, either as new rules
@@
         each \regex{} is also displayed, giving the user an indication of
-        how many of the input log lines that \regex{} should match.
-        Discarding \regexes{} will result in some of the input log lines
-        not being matched; the utility should be run again once the
-        unmatched log lines have been isolated by parsing the original log
-        files once the parser has been augmented with the new rules.
+        how many of the log lines that \regex{} should match.  Discarding
+        \regexes{} will result in some of the log lines not being matched;
+        when the rules have been augmented with the new \regexes{}, the
+        original log files should be parsed again, and any remaining
+        unparsed log lines used as input to this utility.
 
 \end{description}
+
+XXX MERGE logs2regexes WITH check-regexes.
 
@@
 
-\subsection{Parsing Algorithm}
-
-\label{parsing-algorithm}
-
-EMPHASISE THAT THE LOGS CAUSE THE INTERDEPENDENCIES, NOT THE
-ARCHITECTURE\@.
-
-Whereas the rules are quite simple and each rule is completely independent
-of the other rules, the algorithm is significantly more complicated and
-highly internally interdependent.  The algorithm deals with all the
-complications of parsing, the eccentricities and oddities of Postfix log
-files, presenting the resulting data in a normalised, simple to use
-representation.  The algorithm's task is to follow the journey each mail
-takes through Postfix, combining the data extracted by rules into a
-coherent whole, saving it in a useful and consistent form, and performing
-housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-\subsubsection{Introduction}
-
-XXX MERGE WITH PREVIOUS PARAGRAPHS\@.
-
-This section covers the following topics:
-
-\begin{itemize}
-
-    \item A high level overview of the algorithm.
-
-    \item The first set of complications encountered: initially obvious
-        difficulties that had to be overcome.
-
-    \item A flow chart showing common paths a mail can take through the
-        algorithm, a discussion of the emergent behaviour observed in the
-        parser, and an explanation of the paths shown in the flow chart.
-
-    \item A brief description of the framework.
-
-    \item The actions the parser makes available to rules are covered in
-        detail.
-
-    \item Additional complications that have arisen during the development
-        of this parser are documented, the solutions to those
-        complications, and where those solutions are implemented in the
-        algorithm
-
-\end{itemize}
-
-\subsubsection{A high level overview}
+\subsection{Parser Flow Chart}
+
+\label{flow-chart}
+
+XXX MIGHT NEED A SMALL INTRO\@.
+
+XXX MOVE TO AFTER ACTIONS\@.
 
@@
 
-Unfortunately both high level views ignore the many complications
-encountered.
-
-\subsubsection{Flow chart}
-
-\label{flow-chart}
-
-XXX CONSIDER REMOVING PART TWO OF THE FLOW CHART\@: I DO NOT THINK THAT
-RE-INJECTION IS IMPORTANT ENOUGH TO WARRANT IT\@.  THE EXPLANATION OF
-TRACKING (\textsection{}5.4.3) COULD BE MOVED TO ADDITIONAL
-COMPLICATIONS\@.  POSSIBLY MERGE MAIL ENTRY AND DELIVERY IN HERE\@;
-TRACKING SHOULD BE IN COMPLICATIONS --- MERGE WITH RE-INJECTED MAILS\@.
-
-\Figureref{flow chart image part 1} and \figureref{flow chart image part 2}
-show the paths the data representing a mail can take through the parser
-algorithm.  The flow chart covers the most common paths only; there are
-additional, uncommon paths excluded for the sake of clarity; details of the
-deviations can be found in \sectionref{complications}.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart part
-1}{flow chart image part 1}
-
-\showgraph{build/logparser-flow-chart-part-2}{Parser flow chart part
-2}{flow chart image part 2}
+\Figureref{flow chart image} shows the paths the data representing a mail
+can take through the parser algorithm.  The flow chart covers the most
+common paths only, the complications described in
+\sectionref{complications} are excluded for the sake of clarity.  XXX WHERE
+ARE THE PATHS DESCRIBED\@?
+
+\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart}{flow chart
+image}
 
@@
 explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart
-(\figureref{flow chart image part 2}) are encoded within the
+21--22--23--24) in the second half of the flow chart are encoded within the
 actions; the transitions between the other stages (16--17, 24--25) in the
@@
 However the transitions between stages in the first half of the flow chart
-(\figureref{flow chart image part 1}) are not encoded anywhere in
+(\figureref{flow chart image}) are not encoded anywhere in
 the parser; the transitions are determined by the ordering of log lines in
@@
 
-    \item The program will not detect parsing the same log file twice,
+    \item The parser will not detect parsing the same log file twice,
         resulting in the database containing duplicate entries.

### 2008-10-28 doc/logparser-introduction.tex
First half of the corrections to section 5.

@@
 
-Section~\ref{parsing-algorithm} contains the core of the paper, describing
+Section~XXX contains the core of the paper, describing
 a naive parsing algorithm and the complications initially encountered that

### 2008-10-29 doc/logparser-implementation.tex
Second half of the corrections to section 5.

@@
 XXX THIS MIGHT BE USEFUL SOMEWHERE\@:
+
+XXX SHOULD I SAY ``cleaned up'' EVERYWHERE OR NOT\@?
 
@@
 
+XXX SHOULD THESE BE SEPARATE SUBSUBSECTIONS\@?
+
 Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or sendmail, by \SMTP{}, by re-injection
-due to forwarding, or internally generated by Postfix.  Local submission is
-the simplest case: a queueid is assigned immediately and the sender address
-is logged (action: PICKUP\@; flowchart:~2).
+submission via \daemon{postdrop}, by \SMTP{}, by re-injection due to
+forwarding, or internally generated by Postfix.  Local submission is the
+simplest case: a queueid is assigned immediately and the sender address is
+logged (action: PICKUP\@; flowchart:~2).
 
@@
 Re-injection due to forwarding sadly lacks explicit log lines of its
-own;\footnote{Previously discussed in \sectionref{complications},
-complication 3.} re-injection is somewhat awkward to explain because it
-overlaps both the mail acceptance and mail delivery sections, so discussion
-is deferred to \sectionref{tracking re-injected mail}.
+own; it is explained in \sectionref{Re-injected mails}.
 
 Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics (see
-\sectionref{identifying-bounce-notifications} for details).  Bounce
-notifications are the primary example of internally generated mails, though
-there may be other types.\footnote{Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.}
+must be detected using heuristics as described in
+\sectionref{identifying-bounce-notifications}.  Bounce notifications are
+the primary example of internally generated mails, though there may be
+other types.  Postfix may generate mails to the administrator when it
+encounters configuration errors, but such mails are presumably rare.
 
@@
     \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_BY\_QUEUEID\@; flowchart:~9).
+        (action: SAVE\_DATA\@; flowchart:~9).
 
@@
 
-\subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the log files
-without explicit logging indicating their source.  There are two implicit
-indications:
-
-\begin{enumerate}
-
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_QUEUED\@; flowchart:~3), in which case a
-        new data structure will be created for that mail.  The mail will be
-        flagged as having unknown origins; this flag should be subsequently
-        cleared once the origin has been established.  This may also be an
-        indicator that the mail is a bounce notification, see
-        \sectionref{identifying-bounce-notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline{}
-        delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
-
-\end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent has not yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children that are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
-
 \subsubsection{Emergent behaviour}
@@
 \label{Emergent behaviour}
+
+XXX DUNNO IF I SHOULD INCLUDE THIS SECTION\@; AT BEST IT NEEDS TO BE
+UPDATED, AT WORST IT'S RUBBISH\@.
 
@@
 
+\subsection{Actions}
+
+\label{actions in implementation}
+
+XXX INTRODUCTION\@.
+
 \subsubsection{Actions in detail}
@@
 
-\begin{description}
+%\begin{eqlist}[\def\makelabel#1{\bfseries#1:}]
+
+XXX IMPROVE HOW THIS LOOKS
+
+\begin{eqlist}
 
@@
 
-        \begin{description}
+        \begin{eqlist}
 
@@
 
-        \end{description}
+        \end{eqlist}
 
@@
 
-\end{description}
+\end{eqlist}
+
+All data extracted by the rule will be automatically saved, regardless of
+the action which is invoked.
+
+XXX MAKE ALL ACTION DESCRIPTIONS FLOW BETTER\@.
 
@@
         The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Further out of order log lines} this
-        action checks a cache of recent bounce mails to avoid creating
-        bogus bounce mails when lines are logged out of order.
+        with complication \sectionref{Bounce notification mails delivered
+        before their creation is logged} this action checks a cache of
+        recent bounce mails to avoid creating bogus bounce mails when lines
+        are logged out of order.
 
@@
         Deletion will be postponed if the mail is the parent of re-injected
-        mail.
+        mail (\sectionref{Re-injected mails}).
 
@@
 
-    \item [MAIL\_QUEUED] This action represents Postfix
-        picking a mail from the queue to deliver. This action is used for
-        both \daemon{qmgr} and \daemon{cleanup} as it needs to deal with
-        out of order log lines; see \sectionref{discarding cleanup lines}
-        for details.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a message larger
-        than the local server accepts, the mail will be discarded, and the
-        client informed.  See TIMEOUT for further discussion; the two are
-        handled in exactly the same way.
+    \item [MAIL\_QUEUED] This action represents Postfix picking a mail from
+        the queue to deliver. This action is used for both \daemon{qmgr}
+        and \daemon{cleanup} as it needs to deal with out of order log
+        lines; see \sectionref{discarding cleanup lines} for details.  XXX
+        EXPLAIN THIS BETTER\@.
+
+    \item [MAIL\_TOO\_LARGE] When a client tries to send a mail larger than
+        the local server accepts it will be discarded and the client
+        informed of the problem.  This is handled in exactly the same way
+        as the TIMEOUT action, further details are given there.
 
@@
 
-Adding new actions is not as simple as adding new rules, though care has
-been taken in the parser's design and implementation to make adding new
-actions as painless as possible.  The implementor writes a subroutine that
-accepts the standard arguments given to actions, and registers it as an
-action\footnote{The new action must be registered before the rules are
-loaded, as it is an error for a rule to reference an unregistered action;
-this helps catch mistakes made when adding new rules.} by calling the
-framework subroutine add\_actions() with the name of the new action
-subroutine as a parameter.  No other work is required on the implementor's
-part to integrate the action into the parser; all of their attention and
-effort can be focused on the correctness of their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation~\cite{Wikipedia-object-orientation} by subclassing it and
-implementing their changes in the derived class, allowing future upgrading
-of the parser with greatly reduced chance of conflicts.\footnote{The real
-difference between the two approaches is where the new code is placed.  The
-simpler option is to change the parser code directly, but those changes
-will then have to be made to subsequent versions of the parser, and as the
-scope of the changes increases so does the chance of conflict, or mistakes
-when copying the action.  The more time consuming option is to write a
-subclass containing the new actions and change the program which invokes
-the parser so that it uses the subclass rather than the parser; the changes
-required to the program invoking the parser are minor and much less likely
-to lead to conflicts when upgrading to future versions of the parser.  An
+XXX HALF OF THIS CONTENT IS IN FOOTNOTES, WHICH SHOULD BE INLINED\@.
+
+Adding new actions is not as easy as adding new rules, though care has been
+taken in the parser's design and implementation to make adding new actions
+as painless as possible.  The implementor writes a subroutine that accepts
+the standard arguments given to actions, and registers it as an action by
+calling the framework subroutine add\_actions() with the name of the new
+action as a parameter.  The new action must be registered before the rules
+are loaded, because it is an error for a rule to specify an unregistered
+action; this helps catch mistakes made when adding new rules.  No other
+work is required from the implementor to integrate the action into the
+parser; all of their attention and effort can be focused on correctly
+implementing their action.  The only negative aspect is that the process
+involves editing the parser source code, which makes upgrading to a later
+version of the parser more difficult, though by no means impossible.  If
+the author of the new action wishes, they can take advantage of the
+parser's object oriented implementation by subclassing it and implementing
+their changes in the derived class, allowing future upgrades of the parser
+with greatly reduced chance of conflicts.\footnote{The real difference
+between the two approaches is where the new code is placed.  The simpler
+option is to change the parser code directly, but those changes will then
+have to be made to subsequent versions of the parser, and as the scope of
+the changes increases so does the chance of conflict, or mistakes when
+copying the action.  The more time consuming option is to write a subclass
+containing the new actions and change the program that invokes the parser
+so that it uses the subclass rather than the parser; the changes required
+to the program invoking the parser are minor and much less likely to lead
+to conflicts when upgrading to future versions of the parser.  An
 alternative is to submit new actions to the author of the parser for
 inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser will
+not need to be maintained separately, and other users of the parser can
 avail of the new functionality.} The action may need to extend the list of
 valid combinations described in \sectionref{out of order log lines} if the
-addition creates a different set of acceptable programs, but this is
+new action creates a different set of acceptable programs, but this is
 unlikely to occur, as it would require parsing log lines from Postfix
@@
 required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit is that the log files used
-exhibit the idiosyncrasies and peculiarities a mail server in the wild must
-deal with, but the downside is that significantly altering the
-configuration just to log messages from a different Postfix component is
-not an option.}
-
-\subsubsection{Summary}
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications that arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they are not specified anywhere in the parser.
-The framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
-single portion of this section, mirroring the development of the parser.
-The complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
+mail for a university department; the benefit of using this server is that
+the log files used exhibit the idiosyncrasies and peculiarities a mail
+server in the wild must deal with, but the downside is that significantly
+altering the configuration just to log messages from a different Postfix
+component is not an option.}
 
@@
 \label{complications}
+
+XXX IMPROVE THE INTRO\@: ``needs more of a lead-in''.
+
+XXX SHOULD I HAVE A LIST OF HOW MANY COMPLICATIONS ARE HANDLED BY EACH
+ACTION\@?
 
@@
 
-    \item Two state tables to hold the data structure for each connection.
+    \item Two state tables to hold the data structure for each connection
+        and mail.
 
@@
 
-Multiple independent mails may be delivered during one connection: this
+Multiple independent mails may be delivered across one connection: this
 requires the algorithm to clone the current data as soon as a mail is
@@
 mail has been accepted its log entries will not be correlated by \pid{} any
-more (its queueid will be used instead), so there is no longer any
-ambiguity about which mail a given log line belongs to (except when
-timeouts occur during the data phase
-\sectionref{timeouts-during-data-phase}).  If the original connection has
-any interesting data (e.g.\ rejections) it will be saved to the database
-when the client disconnects.  One unsolved difficulty is distinguishing
-between different groups of rejections, e.g.\ when dealing with the
-following sequence:
+more, its queueid will be used instead (except when timeouts occur during
+the data phase \sectionref{timeouts-during-data-phase}).  If the original
+connection has any useful data (e.g.\ rejections) it will be saved to the
+database when the client disconnects.  One unsolved difficulty is
+distinguishing between different groups of rejections, e.g.\ when dealing
+with the following sequence:
 
@@
 
-    \item The client issues the RSET command to reset the session.
+    \item The client issues the RSET command to reset the \SMTP{} session.
 
@@
 \subsubsection{Re-injected mails}
+
+\label{Re-injected mails}
 
@@
 they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this occurs
-a child mail will be injected into the Postfix queue, but without the
-explicit logging \daemon{smtpd} or \daemon{postdrop} injected mails have.
-Thus the source is not immediately discernible from the log line in which
-the mail first appears; from a strictly chronological reading of the log
-files it usually appears as if the child mail has appeared from thin air.
-Subsequently the parent mail will log the creation of the child mail:
+delivered to one or more remote addresses due to aliases.  When this
+occurs a child mail will be injected into the Postfix queue, but without
+the explicit logging that \daemon{smtpd} or \daemon{postdrop} injected
+mails have.  Thus the source of the mail is not immediately discernible
+from the log line in which the mail first appears; from a strictly
+chronological reading of the log files it usually appears as if the child
+mail has appeared from thin air.  Subsequently the parent mail will log the
+creation of the child mail, e.g.\ parent mail 3FF7C4317 creates child mail
+56F5B43FD\@:
 
 \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline{}
-delay=0, status=sent (forwarded as 56F5B43FD)}
-
-The child mail has been created with queueid \texttt{56F5B43FD}.  Different
-delivery methods result in different log lines:
+\tab{} delay=0, status=sent (forwarded as 56F5B43FD)}
+
+XXX DO I NEED THIS LIST\@? IMPROVE HOW IT LOOKS IF I KEEP IT\@.
+
+Different delivery methods result in different log lines:
 
@@
 
-    \item [Local delivery:] delivered to command:
-        /mail/procmail/bin/procmail -p -t /mail/procmail/etc/procmailrc
+    \item [Local delivery:] delivered to command: /usr/bin/procmail
 
@@
 their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered by the parent mail once it has been
-identified.
+entry and subsequently entered once the parent mail has been identified.
+
+XXX MERGE THE REMAINDER OF THIS SECTION INTO THE PRIOR MATERIAL\@.
+
+subsubsection{Tracking re-injected mail}
+
+\label{tracking re-injected mail}
+
+The crux of the problem is that re-injected mails appear in the log files
+without explicit logging indicating their source.  There are two implicit
+indications:
+
+\begin{enumerate}
+
+    \item The indicator which more commonly introduces re-injection is when
+        \daemon{qmgr} selects a mail with a previously unseen queueid for
+        delivery (action: MAIL\_QUEUED\@; flowchart:~3), in which case a
+        new data structure will be created for that mail.  The mail will be
+        flagged as having unknown origins; this flag should be subsequently
+        cleared once the origin has been established.  This may also be an
+        indicator that the mail is a bounce notification, see
+        \sectionref{identifying-bounce-notifications} for details.
+
+    \item Local delivery re-injects the mail and logs a relayed delivery
+        rather than delivering directly to a mailbox or program as it
+        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
+        delivery is performed by the \SMTP{} client; local delivery means
+        local to the server, i.e.\ an address the server is final
+        destination for.} In this case the mail may already have been
+        created (described above) and the unknown origin flag will be
+        cleared; if not a new data structure will be created.  In both
+        cases the re-injected mail is marked as a child of the original
+        mail.  The log line in question is:
+
+        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline{}
+        delay=0, status=sent (forwarded as 56F5B43FD)}
+
+        This second indicator always occurs for re-injected mail but
+        typically occurs after the first indicator explained above.  This
+        is required to tie the parent and child mails together and so is
+        central to the process of tracking re-injected mails.
+
+\end{enumerate}
+
+The algorithm for tracking and saving re-injected mail to the database can
+finally be described:
+
+\begin{itemize}
+
+    \item If the mail is of unknown origin it is assumed to be a child mail
+        whose parent has not yet been identified (action: COMMIT\@;
+        flowchart:~15).  Mark the mail as ready for entry in the database
+        (flowchart:~16), and wait for the parent to deal with it
+        (flowchart:~17).  The mail should not have subsequent log entries;
+        only its parent should refer to it.
+
+    \item If the mail is a child mail then it has already been tracked
+        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
+        is cleaned up, the child is entered in the database
+        (flowchart:~19), then deleted from the state tables.  The child
+        mail will be removed from the parent mail's list of children
+        (flowchart:~20); if this is the last child and the parent has also
+        been entered in the database the parent will be deleted from the
+        state tables.
+
+    \item The last alternative is that the mail is a parent mail (action:
+        COMMIT\@; flowchart:~21).  Regardless of the state of its children
+        its data is cleaned up and entered in the database (flowchart:~22).
+        The parent may have children that are waiting to be entered in the
+        database (flowchart:~23); each of those children's data is cleaned
+        up and entered in the database, then deleted from the state tables.
+        The parent may also have outstanding children which are not yet
+        delivered (flowchart:~24), in which case the parent must wait for
+        those children to be finished with.  As soon as the last child is
+        deleted from the state tables the parent will also be finished with
+        (flowchart:~25), and deleted from the state tables.
+
+\end{itemize}
+
+A parent mail can have multiple children, which may be delivered before or
+after the parent mail.
 
@@
 
-Postfix 2.2.x (and presumably previous versions) lacks explicit logging
-when bounce notifications are generated; suddenly there will be log entries
-for a mail that lacks an obvious source.  There are similarities to the
-problem of re-injected mails discussed in \sectionref{tracking re-injected
-mail}, but unlike the solution described therein bounce notifications do
-not eventually have a log line that identifies their source.  Heuristics
-must be used to identify bounce notifications, and those heuristics are:
+Postfix 2.2.x (and presumably previous versions) does not generate a log
+line when it generates a bounce notification; suddenly there will be log
+entries for a mail that lacks an obvious source.  There are similarities to
+the problem of identifying re-injected mails discussed in
+\sectionref{tracking re-injected mail}, but unlike the solution described
+therein bounce notifications do not eventually have a log line that
+identifies their source.  Heuristics must be used to identify bounce
+notifications, and those heuristics are:
 
@@
 
-    \item The sender address is $<>$.
+    \item The sender address is \verb!<>!.
 
@@
         internally by Postfix, not accepted via \SMTP{} or submitted
-        locally by \daemon{postdrop} or sendmail.
+        locally by \daemon{postdrop}.
 
     \item The message-id has a specific format: \newline{}
-        \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline{}
+        \tab{} \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline{}
         e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
@@
         locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) the queueid in the message-id
-        will not match the queueid of the mail.
+        result of aliasing (described in \sectionref{Re-injected mails}).
+        In the latter case the message-id will be unchanged from the
+        original bounce notification, and so even if it happens to be in
+        the correct format (e.g.\ if it was generated by Postfix on this or
+        another server) it will not match the queueid of the mail.
 
@@
 Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be cleaned up and entered in the
-database.
+origin flag is cleared and the mail can be entered in the database.
 
@@
 
-    \item The sender address must be $<>$.
+    \item The sender address must be \verb!<>!.
 
@@
 this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.
-
-Postfix 2.3 (and hopefully subsequent versions) log the creation of a
-bounce message.
+internally by Postfix will be handled correctly.  Postfix 2.3 (and
+hopefully subsequent versions) log the creation of a bounce message.
 
@@
 accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.\footnote{Microsoft Outlook is
-one client that behaves in this fashion; no attempt has been made to
-identify other clients that act in a similar way.}  An example dialogue
-exhibiting this behaviour is presented below (lines starting with a three
-digit number are sent by the server, the other lines are sent by the
-client):
+it continues with until delivery is complete.  Microsoft Outlook is one
+client that behaves in this fashion; other clients may act in a similar
+way.  An example dialogue exhibiting this behaviour is presented below
+(lines starting with a three digit number are sent by the server, the other
+lines are sent by the client):
 
@@
 
-    \item Are there two or more \daemon{smtpd} log lines?  Does the second
-        result have a postfix\_action attribute of ACCEPTED\@?  The first
-        two \daemon{smtpd} log lines will be a connection log line and a
-        mail acceptance log line (Postfix logs acceptance as soon as the
-        first recipient has been accepted).
+    \item XXX IS THIS CORRECT NOW\@?  Are there two or more \daemon{smtpd}
+        log lines?  Does the second result have a postfix\_action attribute
+        of ACCEPTED\@?  The first two \daemon{smtpd} log lines will be a
+        connection log line and a mail acceptance log line (Postfix logs
+        acceptance as soon as the first recipient has been accepted).
 
@@
 
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix does not
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
+XXX MAYBE FIND OUT WHAT CLIENTS DISCONNECT\@?  Some mail clients disconnect
+abruptly if a second or subsequent recipient is rejected; they may also
+disconnect after other errors, but such disconnections are either
+unimportant or are handled elsewhere in the parser
+(\sectionref{timeouts-during-data-phase}).  Sadly, Postfix does not log a
+message saying the mail has been discarded, as should be expected by now.
+The checks to identify this happening are:
 
@@
 
-These checks are made during the DISCONNECT action: if all checks are
-successful then the mail is assumed to have been discarded when the client
-disconnected.  There will be no further entries logged for such mails, so
-without identifying and entering them in the database immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
+If all checks are successful then the mail is assumed to have been
+discarded when the client disconnected.  There will be no further entries
+logged for such mails, so without identifying and entering them in the
+database immediately they accumulate in the state table and will cause
+clashes if the queueid is reused.
+
+These checks are made during the DISCONNECT action.
 
@@
 the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the data to the database, but if a
-timeout occurs there is no data available to save; the timeout is recorded
-with the connection data instead.
+mail.  It seems more intuitive to save the mail's data to the database, but
+if a timeout occurs there is no data available to save; the timeout is
+recorded with the connection data instead, which is saved.
 
@@
 
-A timeout may thus apply either to an accepted mail or a rejected mail.  To
-distinguish between the two cases the algorithm compares the timestamp of
-the last accepted mail against the timestamp of the last line logged by
-\daemon{smtpd} for that connection (the TIMEOUT action is dependant on the
-CLONE action keeping a list of all mails accepted on each connection).  If
-the mail acceptance timestamp is later then the timeout applies to the
-just-accepted mail, which will be discarded.  If the \daemon{smtpd}
-timestamp is later there was a rejection between the accepted mail and the
-timeout: the action assumes that the timeout applies to a rejected delivery
-attempt and finishes.  This assumption is not necessarily correct, because
-Postfix may have accepted an earlier recipient and rejected a later one, in
-which case the timeout applies to the accepted mail, which should be
-discarded.  This has not been a problem in practice, though it may be in
-future.
-
-This complication is further complicated by the presence of out of order
-\daemon{cleanup} lines: see \sectionref{discarding cleanup lines} for
-details.
+XXX MAKE THIS PARAGRAPH CLEARER\@.  A timeout may thus apply either to an
+accepted mail or a rejected mail.  To distinguish between the two cases the
+algorithm compares the timestamp of the last accepted mail against the
+timestamp of the last line logged by \daemon{smtpd} for that connection
+(the TIMEOUT action is dependant on the CLONE action keeping a list of all
+mails accepted on each connection).  If the mail acceptance timestamp is
+later then the timeout applies to the just-accepted mail, which will be
+discarded.  If the \daemon{smtpd} timestamp is later there was a rejection
+between the accepted mail and the timeout: the action assumes that the
+timeout applies to a rejected delivery attempt and finishes.  This
+assumption is not necessarily correct, because Postfix may have accepted an
+earlier recipient and rejected a later one, in which case the timeout
+applies to the accepted mail, which should be discarded.  This has not been
+a problem in practice, though it may be in future.  This complication is
+further complicated by the presence of out of order \daemon{cleanup} lines:
+see \sectionref{discarding cleanup lines} for details.
+
+This complication is dealt with in the TIMEOUT action.
 
@@
 though there may be other circumstances that trigger it.  Sometimes the
-\daemon{cleanup} line is logged after the timeout line; parsing this line
-causes the creation of a new state table entry for the queueid in the log
-line.  This is incorrect because the line actually belongs to the mail
-that has just been discarded; the next log line for that queueid will be
-seen when the queueid is reused for a different mail, causing a queueid
-clash and the appropriate warning.
-
-In the case where the \daemon{cleanup} line is still pending during the
-TIMEOUT action the algorithm updates a global cache of queueids, adding the
-queueid and the timestamp from the timeout line.  When the next
-\daemon{cleanup} line is parsed for that queueid the cache will be checked
-(during the MAIL\_QUEUED action),
-and the line will be deemed part of the discarded mail and discarded if it
-meets the following requirements:
+\daemon{cleanup} line for a mail being accepted is logged after the timeout
+line; parsing this line causes the MAIL\_QUEUED action to create a new
+state table entry for the queueid in the log line.  This is incorrect
+because the line actually belongs to the mail that has just been discarded;
+the next log line for that queueid will be seen when the queueid is reused
+for a different mail, causing a queueid clash and the appropriate warning.
+
+When the \daemon{cleanup} line is still pending during the TIMEOUT action,
+the action updates a global list of queueids, adding the queueid and the
+timestamp from the log line.  When the next \daemon{cleanup} line is parsed
+for that queueid the list will be checked (during the MAIL\_QUEUED action),
+and the log line will be deemed part of the mail where the timeout occurred
+and discarded if it meets the following requirements:
 
@@
 
-    \item The queueid must not have been reused yet, i.e.\ there is not an
-        entry in the state tables for the queueid.
-
-    \item The timestamp of the \daemon{cleanup} line must be within ten
+    \item The queueid must not have been reused yet, i.e.\ it does not have
+        an entry in the state tables.
+
+    \item The timestamp of the \daemon{cleanup} log line must be within ten
         minutes of the mail acceptance timestamp.  Timeouts happen after
@@
         that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives and false negatives.
+        compromise between false positives (log lines incorrectly
+        discarded) and false negatives (new state table entries incorrectly
+        created).
 
@@
 The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because not every connection where a timeout occurs will have a
-\daemon{cleanup} line logged for it; if the algorithm blindly discarded the
-next \daemon{cleanup} line after a TIMEOUT it would in some cases be
-mistaken.  Whether or not the next \daemon{cleanup} line is discarded the
-queueid will be removed from the cache of timeout queueids when the next
-\daemon{pickup} line containing that queueid is parsed.
+discarded because some, but not all connections where a timeout occurs will
+have an associated \daemon{cleanup} line logged; if the algorithm blindly
+discarded the next \daemon{cleanup} line after a timeout it would sometimes
+be mistaken.  When the next \daemon{pickup} line containing that queueid is
+parsed the queueid will be removed from the cache of timeout queueids,
+regardless of whether it meets the criteria above.
+
+This complication is handled by the TIMEOUT and MAIL\_QUEUED actions.
 
@@
 
-Occasionally the \daemon{pickup} line logged when mail is submitted locally
-via sendmail appears later in the log file than the \daemon{cleanup} line
-for that mail.  This seems to occur during periods of particularly heavy
-load, so is most likely due to process scheduling vagaries.  Normally if
-the queueid given in the PICKUP line exists in the state tables a warning
-is generated by the \daemon{pickup} action, but if the following conditions
-are met it is assumed that the lines were out of order:
+When mail is submitted locally, \daemon{pickup} accepts the new mail and
+generates a log line showing the source.  Occasionally this log line will
+occur later in the log file than the \daemon{cleanup} log line, so the
+PICKUP action will find that a state table entry exists for that queueid.
+Normally if the queueid given in the PICKUP line exists in the state tables
+a warning is generated by the \daemon{pickup} action, but if the following
+conditions are met it is assumed that the lines were out of order:
 
@@
 As always with heuristics there may be circumstances in which these
-heuristics match incorrectly,  but none have been identified so far.
+heuristics match incorrectly, but none have been identified so far.  This
+complication seems to occur during periods of particularly heavy load, so
+is most likely caused by process scheduling vagaries.
 
@@
 immediately obvious reason.  After poring over log files for some time
-there are several reasons for this infrequent occurrence:
+several reasons have been found for this infrequent occurrence:
 
@@
     \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, and all entries in the
+        happens all \daemon{smtpd} processes exit, so all entries in the
         connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.  The queueid state table
-        is untouched.
+        if there is sufficient data, and deleted.
 
@@
 
+    \item Every Postfix process uses a watchdog which kills the process if
+        it is not reset for a considerable period of time (five hours by
+        default).  This safeguard prevents Postfix processes from running
+        indefinitely and consuming resources if a failure causes them to
+        enter a stuck state.  XXX IMPROVE THIS SOMEHOW\@.
+
 \end{enumerate}
 
-The above descriptions appear to cover all situations identified thus far
-where a \daemon{smtpd} suddenly stops logging.  In addition to removing an
-active connection the last accepted mail may need to be discarded, as
-detailed in \sectionref{timeouts-during-data-phase}.
+The circumstances above account for all occasions identified thus far where
+a \daemon{smtpd} suddenly stops logging.  In addition to removing an active
+connection the last accepted mail may need to be discarded, as detailed in
+\sectionref{timeouts-during-data-phase}; otherwise the queueid state table
+is untouched.
 
 These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED and SMTPD\_WATCHDOG\@.
+SMTPD\_DIED, and SMTPD\_WATCHDOG\@.
 
@@
 after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for completeness of the
-algorithm it should be dealt with.  The five occurrences in the test log
-files have the same characteristics: the \daemon{local} log line showing
-delivery to a local mailbox occurs after the \daemon{qmgr} log line showing
-removal of the mail from the queue because delivery is completed.  This
-causes problems: the mail is not complete, so entry into the database
+occurs only five times in 60,721,709 log lines, but for parser correctness
+it must be dealt with.  The five occurrences have the same characteristics:
+the \daemon{local} log line showing delivery to a local mailbox occurs
+after the \daemon{qmgr} log line showing removal of the mail from the queue
+because delivery is completed.  This causes problems: the data in the state
+tables for the mail is not complete, so entering it into the database
 fails; a new mail is created when the \daemon{local} line is parsed and
@@
 
-The solution to this problem is to examine the list of programs that have
-logged lines for each mail, comparing the list against a table of
-known-good combinations of programs (this check is performed during the
-COMMIT action).  If the mail's combination is found in the valid list the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
-SAVE\_BY\_QUEUEID action checks for that flag and retries entry if it is
-found; if the additional log lines have caused the mail to reach a valid
-combination entry will proceed, otherwise it must be postponed once more.
+The COMMIT action examines the list of programs that have logged lines for
+each mail, comparing the list against a table of known-good program
+combinations.  If the mail's combination is found in the table the mail can
+be entered in the database; if the combination is not found entry must be
+postponed and the mail flagged for later entry.  The SAVE\_DATA action
+checks for that flag; if the additional log lines have caused the mail to
+reach a valid combination then entry in the database will proceed,
+otherwise it must be postponed once more.
 
@@
 
+XXX WOULD IT BE BETTER TO BEGIN THE EXPLANATIONS ON THE FOLLOWING LINE\@?
+
 \begin{description}
@@
     \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a forwarded or tracked mail.
+        local delivery of a re-injected mail.
 
@@
 
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Locally
-        submitted mail, \newline{} both local and remote delivery.
+    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Mail submitted
+        locally on the server, \newline{} both local and remote delivery.
 
@@
 
-    \item [\daemon{pickup}, \daemon{smtp}:] Locally submitted mail, remote
-        delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of forwarded or tracked mail, or
+    \item [\daemon{pickup}, \daemon{smtp}:] Mail submitted locally on the
+        server, remote delivery only.
+
+    \item [\daemon{smtp}:] Remote delivery of either a re-injected mail or
         a bounce notification.
@@
         client, then remotely delivered (typically relaying mail for
-        clients on the local network).
+        clients on the local network to addresses outside the local
+        network).
 
@@
 
-This check applies to accepted mails only, not to rejected mails.
+This check applies to accepted mails only, not to rejected mails.  This
+check is performed during the COMMIT action.
 
@@
     \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log entries for the mail, with no additional log entries.
+        or more earlier than the last log line parsed from the current log
+        file.
+
+    \item IS THIS CORRECT NOW\@?  CAN THERE BE MORE smtpd LINES\@?  There
+        must be exactly two \daemon{smtpd} and one \daemon{cleanup} log
+        entries for the mail, with no additional log entries.
 
@@
 12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue.\footnote{This may be a
-problem if Postfix is not running for an extended period of time.}  The
-state tables are scanned for mails matching the characteristics above each
-time the end of a log file is reached, and matching mails are deleted.
+Postfix would delay delivery of a mail in the queue (unless Postfix is not
+running for an extended period of time).  The state tables are scanned for
+mails matching the characteristics above each time the end of a log file is
+reached, and matching mails are deleted.
 
@@
 case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required to have at least one
-recipient.  This complication has been observed in few log files, but
+address needs to be added, as every mail is required by the database schema
+(\sectionref{results table}) to have at least one recipient.  This
+complication has been observed in XXX/\numberOFlogFILES{} log files, but
 typically when it does occur there will be many instances of it, closely
-grouped.  The DELETE action is responsible for handling this complication.
-
-\subsubsection{Further out of order log lines}
-
-\label{Further out of order log lines}
+grouped.
+
+The DELETE action is responsible for handling this complication.
+
+\subsubsection{Bounce notification mails delivered before their creation is
+logged}
+
+\label{Bounce notification mails delivered before their creation is logged}
 
 This is yet another complication that only occurs during periods of
-extremely high load, when process scheduling vagaries and even hard disk
-access times cause strange behaviour.  In this complication bounce
-notification mails are created, delivered and deleted from the queue,
-\textit{before\/} the log line from \daemon{bounce} is logged.  To deal
-with this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail is not already in the state tables. If a mail exists in the
-cache under the correct queueid, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the BOUNCE action
-does not create one.  If the queueid exists it is removed from the cache,
-as it has either just been used or it is too old to use in future.  Whether
-the BOUNCE action creates a mail or finds one existing in the state tables,
+extremely high load, when process scheduling and even hard disk access
+times cause strange behaviour.  In this complication bounce notification
+mails are created, delivered, and deleted from the queue, \textit{before\/}
+the log line from \daemon{bounce} that explains their source is logged.  To
+deal with this the COMMIT action maintains a cache of recently committed
+bounce notification mails, which the BOUNCE action subsequently checks if
+the bounce mail is not already in the state tables. If the queueid exists
+in the cache, and its start time is less than ten seconds before the
+timestamp of the bounce log line, it is assumed that the bounce
+notification mail has already been processed and the BOUNCE action does not
+create one.  If the queueid exists it is removed from the cache, because it
+has either just been used or it is too old to use in future.  Whether the
+BOUNCE action creates a mail or finds an existing mail in the state tables,
 it flags the mail as having been seen by the BOUNCE action; if this flag is
@@
 bounce notification mails.\footnote{This is not required to correctly deal
-with the complication, but is an optimisation to reduce memory usage of the
-parser; on the occasions the author has observed this action occurring
+with the complication, but is an optimisation to reduce the parser's memory
+usage; on the occasions the author has observed this complication occurring
 there have been a huge number of bounce notification mails generated --- if
@@
 size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
+keep the size in check.  XXX EXPLAIN PRUNING\@.
 
@@
 The administrator can delete mails using \daemon{postsuper}; occasionally
-mails in the process of being delivered will be deleted.  This results in
-the log lines from the delivery agent (\daemon{local}, \daemon{virtual} or
-\daemon{smtp}) appearing in the log file \textit{after\/} the mail has been
-removed from the state tables and saved in the database.  The DELETE action
-adds deleted mails to a cache, which is checked by the SAVE\_BY\_QUEUEID
-action, and the current log line discarded if the following conditions are
-met:
+mails that are in the process of being delivered will be deleted.  This
+results in the log lines from the delivery agent (\daemon{local},
+\daemon{virtual} or \daemon{smtp}) appearing in the log file
+\textit{after\/} the mail has been removed from the state tables and saved
+in the database.  The DELETE action adds deleted mails to a cache, which is
+checked by the SAVE\_DATA action, and the current log line discarded if the
+following conditions are met:
 
@@
 
-    \item The timestamp of the log line is within 5 minutes of the end time
-        of the mail.
+    \item The timestamp of the log line is within 5 minutes of the final
+        timestamp of the mail.
 
@@
 Sadly this solution involves discarding some data, but the complication
-only arises eight times (tightly grouped in one log file, which is not in
-the group used for testing); if this complication occurred more frequently
-it might be desirable to find the mail in the database and add the log
-line to it.
+only arises eight times in quick succession in one log file, which is not
+in the \numberOFlogFILES{} log files used for testing; if this complication
+occurred more frequently it might be desirable to find the mail in the
+database and add the log line to it.
 
@@
 Every piece of software suffers from some limitations and there is almost
-always room for improvement.
+always room for improvement.  Below are the limitations and possible
+improvements that have been identified in the parser.
 
@@
 
-    \item Each new Postfix release requires new rules to be written to cope
-        with the new log lines.  Similarly using a new \DNSBL{}, new policy
-        server or new administrator defined rejection messages require new
-        rules.
-
-    \item It appears that the hostname used in the HELO command is not
-        logged if the mail is accepted.\footnote{Tested with Postfix
-        2.2.10, 2.3.11 and 2.4.7; this may possibly have changed in Postfix
-        2.5. XXX TEST THIS AGAIN.}  Rectifying this is relatively simple:
-        create a restriction that is guaranteed to warn for every accepted
-        mail, as follows:
+    \item Each new Postfix release requires new rules to be written or
+        existing rules modified to cope with the new or changed log lines.
+        Similarly using a new \DNSBL{}, a new policy server, or new
+        administrator-defined rejection messages require new rules.
+
+    \item The hostname used in the HELO command is not logged if the mail
+        is accepted.\footnote{Tested with Postfix 2.2.10, 2.3.11 and 2.4.7;
+        this may possibly have changed in Postfix 2.5. XXX TEST THIS
+        AGAIN.}  Rectifying this is relatively simple: create a restriction
+        that is guaranteed to warn for every accepted mail, as follows:
 
@@
             \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain\newline{}
-                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}
+                \texttt{/etc/postfix/main.cf} to contain:\newline{}
+                \tab{}\texttt{check\_helo\_access~pcre:/etc/postfix/log\_helo.pcre}
 
@@
         place to log the HELO hostname, there will not be a queueid
-        associated with the mail for the first recipient, so the log line
-        cannot be associated with the correct mail.  There is guaranteed
-        to be a queueid when the DATA command has been reached, and thus it
-        will be logged by any restrictions taking effect in
+        associated with the mail when \texttt{smtpd\_helo\_restrictions} is
+        evaluated for the first recipient, so the log line cannot be
+        associated with the correct mail.  There is guaranteed to be a
+        queueid when the DATA command has been reached, and thus the
+        queueid will be logged by any restrictions taking effect in
         \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
@@
 
-        Logging the HELO hostname in this fashion also prevents the
-        complication described in \sectionref{Mail deleted before delivery
-        is attempted} from occurring, but only in the case where there is a
+        Logging the HELO hostname in this fashion also partially prevents
+        the complication described in \sectionref{Mail deleted before
+        delivery is attempted} from occurring, but only when there is a
         single recipient; in that case the recipient address will be logged
@@
         preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.
-
-    \item The algorithm does not distinguish between mails where one or
-        more mails are rejected and a subsequent mail is accepted; it will
-        appear in the database as one mail with lots of rejections followed
-        by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It does not appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item The TIMEOUT action uses potentially incorrect heuristics to
-        decide whether the timeout applies to an accepted mail or not,
-        potentially leaving a mail in the state tables
-        (\sectionref{timeouts-during-data-phase}).
-
-    \item The parser will not detect parsing the same log file twice,
-        resulting in the database containing duplicate entries.
+        delivery is attempted} entirely.  XXX ADD A PARAGRAPH EXPLAINING
+        HOW\@.
+
+    \item The parser does not separate mails where one or more mails are
+        rejected and a subsequent mail is accepted; it will appear in the
+        database as one mail with lots of rejections followed by acceptance
+        (this has already been mentioned in \sectionref{connection reuse}).
+        It does not appear to be possible to make this distinction given
+        the data Postfix logs, though it might be possible to write a
+        policy server to provide additional logging.
+
+    \item The parser will not detect that it is parsing the same log file
+        twice, resulting in the database containing duplicate entries.
 
@@
 
-XXX TO BE WRITTEN\@.
+XXX TO BE WRITTEN\@.  START WITH THE OLD CONTENT BELOW\@.
+
+This section presented the core of the parser, starting with a very high
+level view and the initial complications that arose.  A flow chart showing
+the paths a mail may take through the nascent simplified algorithm was
+provided, followed by an explanation of those paths, and a discussion of
+the parser's emergent behaviour --- the data from the log files creates the
+paths in the flow chart, they are not specified anywhere in the parser.
+The framework which holds the parser together was covered next, after which
+came a description of the current actions provided by the parser, and the
+algorithm for analysing unparsed log lines to create \regexes{} for new
+rules.  Detecting, diagnosing and defeating complications forms the largest
+single portion of this section, mirroring the development of the parser.
+The complications are described in the order they were overcome, with
+subsequent problems affecting fewer mails (often by an order or magnitude),
+though the time required to solve problems increased with each successive
+problem.

### 2008-10-30 doc/logparser-implementation.tex
Fixed all overfull hbox warnings, and made some small improvements in the process.

@@
 
-        XXX SHOULD THIS LIST BE HERE\@?  MAYBE MOVE TO AFTER THE FIELD
-        LIST\@?  DO I NEED THIS LIST AT ALL\@?
+        XXX MAYBE MOVE THIS LIST TO AFTER THE FIELD LIST\@?  DO I NEED THIS
+        LIST AT ALL\@?
 
@@
             \item [EXPIRED] The mail has been in the queue for five
-                days\footnotemark[\value{footnote}]XXX CHECK IF THIS
-                FOOTNOTE MAKES SENSE without successful delivery.  A bounce
-                mail will be generated and sent to the sender address.
+                days\footnotemark[\value{footnote}] XXX CHECK IF THIS
+                FOOTNOTE WORKS without successful delivery.  A bounce mail
+                will be generated and sent to the sender address.
 
@@
             \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all existing \daemon{smtpd}
-                processes will be terminated.  This does not negatively
-                affect the log files or mail queued by Postfix for
-                delivery.
+                Postfix to start or stop, and all \daemon{smtpd} processes
+                will be terminated.  This does not negatively affect the
+                log files or mail queued by Postfix for delivery.
 
@@
         mail is accepted.  The format is: \newline{} \tab{} \texttt{XXX
-        UPDATE EXAMPLE\@; smtp\_code = 550; recipient = 2, sender = 4;}
-        \newline{} i.e.\ semi-colon or comma separated assignment
-        statements, with the variable name on the left and the value on the
-        right hand side.\footnote{Commas and semi-colons cannot be escaped
-        and thus cannot be used.  This feature is intended for use with
-        small amounts of data rather than large amounts in any one rule, so
-        dealing with escape sequences was deemed unnecessary.}  Any field
-        in the results table can be set in this way; additionally
-        \texttt{child} and \texttt{pid} are accepted and used respectively
-        by the \texttt{TRACK}, \texttt{BOUNCE} and \texttt{SMTPD\_DIED}
-        actions, but do not cause data to be entered in the database.
+        UPDATE EXAMPLE\@; smtp\_code = 550; \newline{} recipient = 2,
+        sender = 4;} \newline{} i.e.\ semi-colon or comma separated
+        assignment statements, with the variable name on the left and the
+        value on the right hand side.\footnote{Commas and semi-colons
+        cannot be escaped and thus cannot be used.  This feature is
+        intended for use with small amounts of data rather than large
+        amounts in any one rule, so dealing with escape sequences was
+        deemed unnecessary.}  Any field in the results table can be set in
+        this way; additionally \texttt{child} and \texttt{pid} are accepted
+        and used respectively by the \texttt{TRACK}, \texttt{BOUNCE} and
+        \texttt{SMTPD\_DIED} actions, but do not cause data to be entered
+        in the database.
 
@@
 program             & \daemon{smtpd}                                    \\
-XXX                 & UPDATE REGEX WHEN AUTOMATIC SAVING IS FINISHED    \\
+XXX                 & UPDATE REGEX FOR AUTOMATIC SAVING                 \\
 regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
@@
 
-    \item [Classify words based on their frequency]  The frequency of each
-        \texttt{(word, word's position within the log line)} tuple is
+    \item [Classify words based on their frequency]  The frequency of every
+        tuple \texttt{(word, word's position within the log line)} is
         checked: if its frequency is greater than the threshold supplied by
@@
 
-    \item XXX IS THIS CORRECT NOW\@?  Are there two or more \daemon{smtpd}
-        log lines?  Does the second result have a postfix\_action attribute
-        of ACCEPTED\@?  The first two \daemon{smtpd} log lines will be a
-        connection log line and a mail acceptance log line (Postfix logs
-        acceptance as soon as the first recipient has been accepted).
+    \item XXX IS THIS CORRECT NOW\@?\newline{}  Are there two or more
+        \daemon{smtpd} log lines?  Does the second result have a
+        postfix\_action attribute of ACCEPTED\@?  The first two
+        \daemon{smtpd} log lines will be a connection log line and a mail
+        acceptance log line (Postfix logs acceptance as soon as the first
+        recipient has been accepted).
 
@@
     \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Mail submitted
-        locally on the server, \newline{} both local and remote delivery.
+        locally \newline{} on the server, both local and remote delivery.
 

### 2008-10-30 doc/logparser-acronyms.tex
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

@@
     and maintenance.  Some information about SQL, including its history,
-    can be found at~\cite{Wikipedia-sql}; a good introduction can be found
-    at~\cite{sql-for-web-nerds}, others are~\cite{w3schools-sql-tutorial,
-    sqlcourse.com}.
+    can be found at \url{http://en.wikipedia.org/wiki/SQL}; a good
+    introduction can be found at \url{http://philip.greenspun.com/sql/}.
 }}
@@
 
-\newacronym{HTML}{Hypertext Markup Language}{description={
-    HTML is the markup language designed for writing web
-    pages~\cite{Wikipedia-html,w3schools-html}.
-}}
-
 \newacronym{IP}{Internet Protocol}{description={

### 2008-10-30 doc/logparser-background.tex
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

@@
                 TRIGGER or two.\footnote{Please refer to an \SQL{} guide
-                for explanations of these terms,
-                e.g.~\cite{sql-for-web-nerds}}
+                for explanations of these terms, e.g.\
+                \url{http://philip.greenspun.com/sql/}.}
 
@@
         readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at~\cite{sql-for-web-nerds},
-        others are~\cite{w3schools-sql-tutorial,sqlcourse.com}).  Although
+        introduction to \SQL{} can be found at
+        \url{http://philip.greenspun.com/sql/},
+        others are \url{http://www.w3schools.com/sql/default.asp},
+        \url{http://sqlcourse.com/}).  Although
         every vendor implements a different dialect of \SQL{}, the basics
@@
 
-    \caption{Example attributes sent to policy servers}\label{Example
-    attributes sent to policy servers}
+    \caption{Example attributes sent to policy servers}
+    \empty{}\label{Example attributes sent to policy servers}
 
-    XXX HOW CAN I CENTRE THE TABLE\@?
+    XXX HOW CAN I CENTRE THE TABLE\@?\hfill{}
+
     \begin{tabular}[]{ll}

### 2008-10-30 doc/logparser-conclusion.tex
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

@@
 \label{conclusion}
+
+XXX THIS NEEDS TO BE EXTENDED\@.
+
+XXX THERE IS NO MENTION OF RESULTS/EFFICIENCY\@.
 
@@
 though it does raise efficiency and optimisation questions (answered in
-\sectionref{parser efficiency}).  Providing a tool (\sectionref{creating
-new rules}) to ease the generation of \regexes{} from unparsed log lines
-should greatly help users add rules to parse formerly unparsed log lines.
+\sectionref{parser efficiency}).  Providing a tool to ease the generation
+of \regexes{} from unparsed log lines should greatly help users add rules
+to parse previously unparsed log lines (\sectionref{creating new rules}).
 
-The implementation not only substantially eases the parsing of new log
+The architecture not only substantially eases the parsing of new log
 lines, it makes adding new actions (\sectionref{adding new actions}) a
 relatively simple task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might disrupt parsing of
-other log lines or the behaviour of other actions, allowing their
-concentration to focus on correctly implementing their new action.
+implementor from worrying about how their action might interfere with
+existing parsing, so they can focus on correctly implementing their new
+action.
 
-The division of the parser into rules, actions and framework is unusual
+The division of the parser into rules, actions, and framework is unusual
 because rules are separated so completely from the actions and framework;
-although parsers are often divided (parsers based on the combination of
-\texttt{lex} and \texttt{yacc}~\cite{lex-and-yacc} 
-being an obvious example), the parts are usually quite internally
-interdependent, and will be combined into a complete parser by the
-compilation process; in contrast \parsername{} keeps the rules and actions
-separate until the parser runs.  The actions and framework are not as
-completely separated, as the actions depend on services provided by the
-framework; however the actions and framework are not tightly integrated: it
-would be possible, with some work, to completely separate the two.
+although parsers are often divided (the combination of \texttt{lex} and
+\texttt{yacc}~\cite{lex-and-yacc} being an obvious example), the parts are
+usually quite internally interdependent, and will be combined into a
+complete parser by the compilation process; in contrast \parsername{} keeps
+the rules and actions separate until the parser runs.  The actions and
+framework are not as completely separated, as the actions depend on
+services provided by the framework; however the actions and framework are
+not tightly integrated: it would be possible, with some work, to completely
+separate the two.  XXX I SHOULD DO THAT\@.
 
-The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
-behaviour}) exhibited by the rules and actions is also interesting, and is
-discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-the paths through the parser.  This emergent behaviour greatly eases the
-process of adding new actions (\sectionref{adding new actions}), as the
-actions do not have to be inserted into an explicit flow of control; new
-actions will naturally find their place in the paths through the parser.
+% XXX THIS SHOULD PROBABLY BE REMOVED\@.
+%The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
+%behaviour}) exhibited by the rules and actions is also interesting, and is
+%discussed after the flow chart (\sectionref{flow-chart}) and explanation of
+%the paths through the parser.  This emergent behaviour greatly eases the
+%process of adding new actions (\sectionref{adding new actions}), as the
+%actions do not have to be inserted into an explicit flow of control; new
+%actions will naturally find their place in the paths through the parser.
 
@@
 keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail.\footnote{Comments will already
-have been removed by the preprocessor.}  This parser is different: the
-rule which matches the next log line for a mail dictates the action that
-will be invoked, which is equivalent to the next state in other parsers;
-thus the next log line for a mail dictates the next state for that mail.
+any other input causing parsing to fail (comments will already have been
+removed by the preprocessor).  This parser is different: the rule which
+matches the next log line for a mail dictates the action that will be
+invoked, which is equivalent to the next state in other parsers; thus the
+next log line for a mail dictates the next state for that mail.  XXX THINK
+ABOUT HOW TO MAKE THE PARSER MORE STATEFUL IF REQUIRED\@; ADD IT TO THE
+ARCHITECTURE SECTION\@.
 
 The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications previously
-explained only begin to become apparent once the vast majority of log lines
-have been parsed successfully.  Adding new rules to deal with numerous,
-infrequently occurring log lines is a simple task, albeit tiresome (though
-alleviated by the tool described in \sectionref{adding new actions}),
-whereas dealing with mails where information appears to be missing is much
-more grueling.  Trawling through the log files, looking for something out
-of the ordinary, possibly hundreds or even thousands of lines away from the
-last mention of the queueid in question, is extremely time consuming and
-error prone.  Sometimes the task is not to spot the line that is unusual,
-but to spot that a line normally present is missing, i.e.\ to realise that
-one line amongst hundreds is absent.  In all cases the evidence must be
-used to construct a hypothesis to explain the irregularities, and that
-hypothesis must then be tested in the parser; if successful, the parser
-must be modified to deal with the irregularities, without adversely
-affecting the existing functionality.  The complications described in
-\sectionref{complications} were solved in the order they are
+90\% of the log lines, as the irregularities and complications explained in
+\sectionref{complications} only become apparent once the vast majority of
+log lines have been parsed successfully.  Adding new rules to deal with
+numerous, infrequently occurring log lines is a simple, albeit tiresome,
+task, (though alleviated by the tool described in \sectionref{adding new
+actions}), whereas dealing with mails where information appears to be
+missing is much more grueling.  Trawling through the log files, looking for
+something out of the ordinary, possibly hundreds or even thousands of lines
+away from the last mention of the queueid in question, is extremely time
+consuming and error prone.  Sometimes the task is not to spot the line that
+is unusual, but to spot that a line that is normally present is missing,
+i.e.\ to realise that one line amongst thousands is absent.  In all cases
+the evidence must be used to construct a hypothesis to explain the
+irregularities, and that hypothesis must then be tested in the parser; if
+successful, the parser must be modified to deal with the irregularities,
+without adversely affecting the existing functionality.  The complications
+described in \sectionref{complications} were solved in the order they are
 described in, and that order closely resembles the frequency with which

### 2008-10-30 doc/logparser-glossary.tex
First pass through corrections for sections 1-3.
Second pass through corrections for sections 1-3.
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

@@
     server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found at~\cite{sqlite-features} or
-    \url{http://www.sqlite.org/}.
+    utilising it.  More details can be found
+    at~\url{http://www.sqlite.org/}.
 }}
@@
     corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this document, see~\cite{bayesian-filtering, a-plan-for-spam}
+    scope of this thesis, see~\cite{bayesian-filtering,a-plan-for-spam}
     for a detailed explanation.
@@
 
-\glossary{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
-}}
+%\glossary{name={hash},description={
+%    A hashing function transforms a string of characters to a number.
+%    There are many possible uses for the resulting number: a common usage
+%    is to maintain a data structure indexed by strings in an efficient
+%    manner.  A full description is beyond the scope of this paper, further
+%    information can be found at~\cite{hash-functions}.
+%}}
 

### 2008-10-30 doc/logparser-other-parsers-review.tex
Fixed all overfull hbox warnings, and made some small improvements in the process.
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

@@
     scan qmail, postfix, sendmail and exim logfile and produce usage
-    statistics in \HTML{} format for viewing through a browser. It produces
+    statistics in HTML format for viewing through a browser. It produces
     Top domains output according to Sender, Receiver, Total mails and
@@
 
-    \item The top ten \SMTP{} error codes (XXX SENDING, RECEIVING, OR
-        BOTH\@?), e.g.\ the first entry in the list is: \texttt{Requested
-        mail action not taken: relaying not allowed, unknown recipient
-        user, \ldots}
+    \item A summary of the \SMTP{} codes used when rejecting delivery
+        attempts.
 
@@
 
-% Parsed lines in file: 10416129
-%  Found 4 dropped records,
-%  Found 1176050 corrupted records,
-%  Found 0 old records,
-%  Found 9240075 new qualified records.
+XXX NEED A CONCLUSION\@.
 
@@
     There are eight useful analyse modules, writing the result in plain
-    ASCII or \HTML{}, to stdout or to files.
+    ASCII or HTML, to stdout or to files.
 
@@
 
-YAALA uses a plugin based system to analyse log files and produce \HTML{}
+YAALA uses a plugin based system to analyse log files and produce HTML
 output reports, with all the parsing and report generation handled by
@@
 were made.  As parsing failed it was not possible to review the reports
-Logrep can generate (available in \HTML{} only), or to examine the
+Logrep can generate (available in HTML only), or to examine the
 (undocumented) format in which it can save extracted data for subsequent

### 2008-10-30 doc/logparser-results.tex
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).

@@
 ORDERINGS SECTION\@?
+
+XXX REUSE THE CONTENT FROM THE PAPER FOR THIS SECTION\@.
 
@@
 
+XXX REWRITE THIS SECTION SO THAT IT'S CORRECT\@: POINT 4 SAYS THAT THE
+ENVELOPE SENDER CHANGES, IS THAT CORRECT\@?
+
 \begin{enumerate}
@@
         \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, which now has an envelope sender
-        address of username@domain.
+        forwarded bounce notification, XXX which now has an envelope sender
+        address of username@domain.   XXX IS THIS CORRECT\@?
 
-    \item Postfix accepts the forwarded bounce notification but during
-        deliver the mail it notices that the \texttt{Delivered-To}
-        header already contains the address currently being delivered to,
-        and sends a bounce notification to the sender: username@domain.
+    \item Postfix accepts the forwarded bounce notification but during mail
+        delivery it notices that the \texttt{Delivered-To} header already
+        contains the address currently being delivered to, and sends a
+        bounce notification to the sender: username@domain.
 
@@
 not have to deliver the mail to an external system, so mails are delivered,
-bounced and generated as fast as the disks can keep up, resulting in a huge
+bounced, and generated as fast as the disks can keep up, resulting in a huge
 volume of log lines.
 
-The different distribution of log lines during a mail forwarding loop has
-unexpected effects on parsing time, described in detail in
+The different distribution of log lines caused by a mail forwarding loop
+has unexpected effects on parsing time, described in detail in
 \sectionref{rule ordering for efficiency}, \sectionref{scalability as the
@@
 
-I THINK THIS SECTION NEEDS SUBSTANTIAL WORK, IF NOT A RE-WRITE\@.
+XXX I THINK THIS SECTION NEEDS SUBSTANTIAL WORK, IF NOT A RE-WRITE\@; START
+WITH THE PAPER'S CONTENT\@.
 

### 2008-10-31 doc/logparser-acronyms.tex
Finished corrections; 112 XXX notes.

@@
 \newacronym{PLP}{Postfix Log Parser}{description={
-    Postfix Log Parser is the implementation of the algorithm described in
-    this paper.
+    Postfix Log Parser is the parser implementing the algorithm described
+    in this thesis.
 }}
@@
     A Mail Transfer Agent sends and receives mail via
-    SMTP\backslashat{}.
+    SMTP\backslashat{}.  XXX DO I NEED THIS\backslashat{}?
 }}
@@
 
-\newacronym{SPF}{Sender Policy Framework}{description={
-    SPF is introduced in \sectionref{spf introduction} and explained fully
-    in~\cite{openspf,Wikipedia-spf}.
-}}
-
-\newacronym{ISP}{Internet Service Provider}{description={
-    An ISP is a company that sells Internet access to consumers.
-}}
-
 \newacronym{LMA}{Log Mail Analyzer}{description={
     One of the other Postfix log parsers reviewed and the only example of
-    published prior art~\cite{log-mail-analyser} found that specifically
-    deals with parsing Postfix logs.
+    prior published paper~\cite{log-mail-analyser} found that specifically
+    deals with parsing Postfix log files.
 }}
@@
         www.example.com is a FQDN
-}}
-
-\newacronym{DNS}{Domain Name System}{description={
-    The DNS converts between hostnames (www.example.com) and IP addresses
-    (10.1.2.3).
 }}
@@
 
-\newacronym{regex}{Regular Expression}{description={
-    Regular expressions are a concise and powerful method of matching
-    patterns against text that are explained in detail
-    in~\cite{Wikipedia-regex,perlre}.
-}}
-
 % I think this is better than having both regex and regexes in the acronym
 % list.  Likewise for pids.
-\newcommand{\regexes}{\regex{}es}
 \newcommand{\pids}{\pid{}s}
@@
 \newacronym{IP}{Internet Protocol}{description={
-    The Internet Protocol~\cite{Wikipedia-ip} is the protocol used to
-    communicate between computers on the Internet.  An IP address is a
-    unique address assigned to a computer on the Internet, allowing it to
-    communicate with other computers on the Internet.  For more information
-    see~\cite{Wikipedia-ip-address}.
-}}
-
-\newacronym{PDF}{Portable Document Format}{description={
-    PDF is the file format created by Adobe Systems in 1993 for document
-    exchange.  It aims to be device independent, so documents should look
-    the same whether viewed on screen or printed.
-}}
-
-\newacronym{LCD}{Lowest Common Denominator}{description={
-    LCD is a mathematical term that is used figuratively to mean the least
-    useful member of a set of alternatives.
+    The Internet Protocol is the protocol used to communicate between
+    computers on the Internet.  An IP address is a unique address assigned
+    to a computer on the Internet, allowing it to communicate with other
+    computers on the Internet.
 }}

### 2008-10-31 doc/logparser-architecture.tex
First half of the corrections to section 5.
Fixed all overfull hbox warnings, and made some small improvements in the process.
Finished corrections; 112 XXX notes.

@@
 
-XXX REWRITE TO MERGE AND SUPPLEMENT THE EXISTING CONTENT\@.
+XXX REWRITE\@: MERGE CONTENT FROM ELSEWHERE IN THE THESIS AND PAPER\@.
 
-XXX NEEDS AN INTRODUCTION ONCE CONTENT HAS SETTLED\@.
+XXX NEEDS AN INTRODUCTION\@.
 
@@
 parse new log lines without modifying the core parsing algorithm.  Adding a
-new rule with the action to invoke and a \regex{} to match the log lines is
+new rule with the action to invoke and a regex to match the log lines is
 trivial in comparison to understanding a program's entire parsing
@@
 \subsection{Rules}
+
+\label{rule in architecture}
 
@@
 
-    \item Sort by \regex{} and visually inspect the list, e.g.\ with \SQL{}
+    \item Sort by regex and visually inspect the list, e.g.\ with \SQL{}
         similar to: \textbf{select regex from rules order by regex;}
@@
 
+\label{actions in architecture}
+
 XXX TO BE WRITTEN\@.  START WITH THE CONTENT FROM THE PAPER, THEN CHECK THE

### 2008-10-31 doc/logparser-background.tex
Finished corrections; 112 XXX notes.

@@
 administrator may require that all clients on the local network have valid
-\DNS{} entries, to prevent people sending mail from unknown machines.}  The
+DNS entries, to prevent people sending mail from unknown machines.}  The
 administrator can define new restrictions as a list of existing
@@
 
-    \item Checking \SPF{} records~\cite{openspf,Wikipedia-spf}.
-        \SPF{}\label{spf introduction} records specify which mail servers
-        are allowed to send mail claiming to be from a particular domain.
-        The intention is to reduce spam from faked sender addresses,
+    \item Sender Policy Framework (SPF)~\cite{openspf}.  SPF\label{spf
+        introduction} records specify which mail servers are allowed to
+        send mail claiming to be from a particular domain.  The intention
+        is to reduce spam from faked sender addresses,
         backscatter~\cite{postfix-backscatter}, and
@@
         increasingly complex and well written programs to distribute spam,
-        frequently using an \ISP{} provided \SMTP{} server from a
-        compromised machine on the \ISP{}'s network.  Greylisting will
-        slowly become less useful, but it does block a large percentage of
-        spam mail at the moment; the most effective restrictions from the
+        frequently using an ISP provided \SMTP{} server from a compromised
+        machine on the ISP's network.  Greylisting will slowly become less
+        useful, but it does block a large percentage of spam mail at the
+        moment; the most effective restrictions from the
         \numberOFlogFILES{} log files used in testing the parser are shown

### 2008-10-31 doc/logparser-conclusion.tex
Finished corrections; 112 XXX notes.

@@
 \sectionref{parser efficiency}).  Providing a tool to ease the generation
-of \regexes{} from unparsed log lines should greatly help users add rules
+of regexes from unparsed log lines should greatly help users add rules
 to parse previously unparsed log lines (\sectionref{creating new rules}).
@@
 although parsers are often divided (the combination of \texttt{lex} and
-\texttt{yacc}~\cite{lex-and-yacc} being an obvious example), the parts are
-usually quite internally interdependent, and will be combined into a
-complete parser by the compilation process; in contrast \parsername{} keeps
-the rules and actions separate until the parser runs.  The actions and
-framework are not as completely separated, as the actions depend on
-services provided by the framework; however the actions and framework are
-not tightly integrated: it would be possible, with some work, to completely
-separate the two.  XXX I SHOULD DO THAT\@.
+\texttt{yacc} XXX BOOK REF FOR LEX AND YACC being a common example), the
+parts are usually quite internally interdependent, and will be combined
+into a complete parser by the compilation process; in contrast
+\parsername{} keeps the rules and actions separate until the parser runs.
+The actions and framework are not as completely separated, as the actions
+depend on services provided by the framework; however the actions and
+framework are not tightly integrated: it would be possible, with some work,
+to completely separate the two.  XXX I SHOULD DO THAT\@.
 
 % XXX THIS SHOULD PROBABLY BE REMOVED\@.
-%The emergent behaviour~\cite{Wikipedia-Emergence} (\sectionref{Emergent
+%The emergent behaviour (\sectionref{Emergent
 %behaviour}) exhibited by the rules and actions is also interesting, and is

### 2008-10-31 doc/logparser-glossary.tex
Finished corrections; 112 XXX notes.

@@
 \glossary{name={postsuper},description={
-    Maintenance task such as deleting mails from the queue, putting mail on
-    hold (no further delivery attempts will be made until it is released
-    from hold, also by postsuper), and consistency checking of the mail
-    queue.  The definitive documentation is available at
+    Maintenance tasks such as deleting mails from the queue, putting mail
+    on hold and later releasing it (no further delivery attempts will be
+    made until it is released), and consistency checking of the mail queue.
+    The definitive documentation is available at
     \url{http://www.postfix.org/postsuper.1.html}.
@@
     (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands that send mail depend on; Postfix sendmail executes postdrop
-    to place a new mail in the queue.  The definitive documentation is
+    commands that send mail depend on; it executes \daemon{postdrop} to
+    place a new mail in the queue.  The definitive documentation is
     \url{http://www.postfix.org/sendmail.1.html}.
@@
     it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, given sufficient
-    time they will be reused.
+    guaranteed to be unique for the lifetime of the mail, they may be
+    reused later.
 }}
@@
 \glossary{name={IPv4},description={
-    Internet Protocol~\cite{Wikipedia-ipv4} version 4 is the fourth version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the first widely deployed version of IP, and has been
-    in use for over 25 years.
+    Internet Protocol version 4 is the fourth version of the Internet
+    Protocol used to interconnect computers on the Internet.  It is the
+    first widely deployed version of IP, and has been in use for over 25
+    years.
 }}
@@
 \glossary{name={IPv6},description={
-    Internet Protocol~\cite{Wikipedia-ipv6} version 6 is the latest version
-    of the Internet Protocol used to interconnect computers on the
-    Internet.  It is the successor to IPv4, bringing with it a greatly
-    expanded address space, allowing many more computers to use the
-    Internet simultaneously.  IPv4 and IPv6 will coexist for many years to
-    come as existing networks transition from the former to the latter.
+    Internet Protocol version 6 is the latest version of the Internet
+    Protocol used to interconnect computers on the Internet.  It is the
+    successor to IPv4, bringing with it a greatly expanded address space,
+    allowing many more computers to use the Internet simultaneously.  IPv4
+    and IPv6 will coexist for many years to come as existing networks
+    transition from the former to the latter.
 }}
@@
 %}}
-
-\glossary{name={Solaris},description={
-    Solaris is a Unix-derived Operating System produced by Sun Microsystems
-    (\url{http://www.sun.com/software/solaris/}).
-}}
 
@@
 \glossary{name={mail loop},description={
-    Sometimes mail set to one address must be delivered to a different
+    Sometimes mail sent to one address must be forwarded to a different
     address instead, e.g.\ because a person has changed jobs.  A mail loop
-    occurs when the recipient addresses constitute a cyclic directed graph;
-    the simplest example is when mail for \texttt{foo\at{}example.net} is
-    delivered to \texttt{bar\at{}example.com}, and mail for
-    \texttt{bar\at{}example.com} is delivered to
-    \texttt{bar\at{}example.com}.
+    occurs when the chain of recipient addresses constitute a cyclic
+    directed graph; the simplest example is when mail for
+    \texttt{alice\at{}example.net} is delivered to
+    \texttt{bob\at{}example.com}, and mail for \texttt{bob\at{}example.com}
+    is delivered to \texttt{alice\at{}example.net}.
 }}

### 2008-10-31 doc/logparser-implementation.tex
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).
Finished corrections; 112 XXX notes.

@@
 
-            \item [PROCESSING] \daemon{cleanup} is processing a mail
-                (see~\cite{postfix-cleanup} for details).
+            \item [PROCESSING] \daemon{cleanup} is processing a mail.
+                XXX ADD MORE DETAIL HERE
 
@@
     \item [regex] XXX THIS NEEDS TO BE UPDATED ONCE AUTOMATIC DATA
-        EXTRACTION IS FINISHED\@.  The \regex{} to match the log line
-        against.  The \regex{} will first have several keywords expanded:
+        EXTRACTION IS FINISHED\@.  The regex to match the log line
+        against.  The regex will first have several keywords expanded:
         this simplifies reading and writing rules; avoids needless
-        repetition of complex \regex{} components; allows the components to
+        repetition of complex regex components; allows the components to
         be corrected and/or improved in one location; and makes each
-        \regex{} largely self-documenting.  For efficiency the keywords are
-        expanded and every rule's \regex{} is compiled before attempting to
-        parse the log file --- otherwise each \regex{} would be recompiled
+        regex largely self-documenting.  For efficiency the keywords are
+        expanded and every rule's regex is compiled before attempting to
+        parse the log file --- otherwise each regex would be recompiled
         each time it was used, resulting in a large, data dependent
         slowdown.  Efficiency concerns are discussed in \sectionref{parser
-        efficiency}, with caching compiled \regexes{} covered in
+        efficiency}, with caching compiled regexes covered in
         \sectionref{Caching compiled regexes}.
@@
         \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via \DNS{}\@.
+        hostname via DNS\@.
 
@@
 mail from a sender address because the domain in the sender address does
-not have an A, AAAA, or MX \DNS{} entry, i.e.\ mail could not be delivered
+not have an A, AAAA, or MX DNS entry, i.e.\ mail could not be delivered
 to the sender's address (for full details
@@
     \item [program and regex:] If the program in the rule equals the
-        program in the log line, a match using the rule's \regex{} will be
+        program in the log line, a match using the rule's regex will be
         attempted against the log line; if the match is successful the
@@
 
-    \item [action:] The action to be executed if the \regex{} matches
+    \item [action:] The action to be executed if the regex matches
         successfully.  See \sectionref{actions-in-detail} for details of
@@
 classification can be used to group similar log lines and generate a
-\regex{} to match each group of log lines.
+regex to match each group of log lines.
 
@@
         variable terms (e.g.\ email addresses, \IP{} addresses, the
-        standard start of rejection messages, etc.) with \regex{} keywords
+        standard start of rejection messages, etc.) with regex keywords
         that the parser will expand when it loads the rule (see the
-        \regex{} entry in \sectionref{rule attributes}).  The purpose of
+        regex entry in \sectionref{rule attributes}).  The purpose of
         this step is to utilise existing knowledge to create more accurate
-        \regexes{}; it replaces a large number of variable words with fixed
+        regexes; it replaces a large number of variable words with fixed
         words, improving the subsequent classification of words as fixed or
@@
         was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash from a small range of values for each word
+        pass it generated a hash~\cite{hash-functions} from a small range of values for each word
         of each log line, incrementing a counter for each hash.  The
@@
         will be misclassified as a fixed term, but that should be noticed
-        by the user when reviewing the new \regexes{}.  Variable terms are
+        by the user when reviewing the new regexes.  Variable terms are
         replaced by \texttt{.+}, which means to match zero or more of any
@@
 
-    \item [Build regexes]  The words are reassembled to produce a \regex{}
+    \item [Build regexes]  The words are reassembled to produce a regex
         matching the log line, and a counter is maintained for each
-        \regex{}.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled \regexes{} are collapsed to a single \texttt{.+}; any
-        resulting duplicate \regexes{} are combined, and their counters
-        added together.  If a \regex{}'s counter is lower than the
-        threshold supplied by the user the \regex{} is discarded; this
+        regex.  Contiguous sequences of \texttt{.+} in the newly
+        reassembled regexes are collapsed to a single \texttt{.+}; any
+        resulting duplicate regexes are combined, and their counters
+        added together.  If a regex's counter is lower than the
+        threshold supplied by the user the regex is discarded; this
         second threshold is independent of the threshold used to
         differentiate between fixed and variable words, but once again 1\%
-        of log lines is a good starting point.  The new \regexes{} are
+        of log lines is a good starting point.  The new regexes are
         displayed for the user to add to the database, either as new rules
-        or merged into the \regexes{} of existing rules; the counter for
-        each \regex{} is also displayed, giving the user an indication of
-        how many of the log lines that \regex{} should match.  Discarding
-        \regexes{} will result in some of the log lines not being matched;
-        when the rules have been augmented with the new \regexes{}, the
+        or merged into the regexes of existing rules; the counter for
+        each regex is also displayed, giving the user an indication of
+        how many of the log lines that regex should match.  Discarding
+        regexes will result in some of the log lines not being matched;
+        when the rules have been augmented with the new regexes, the
         original log files should be parsed again, and any remaining
@@
 
-A second utility is also provided that reads a list of new \regexes{} and
+A second utility is also provided that reads a list of new regexes and
 the input given to the first utility.  It tries to match each input log
-line against each \regex{}, counting the number of log lines that match
-each \regex{}, warning the user if an input log line is matched by more
-than one \regex{}, and additionally warning if an input log line is not
-matched by any \regex{}.  It displays a summary of how many input log lines
-each \regex{} matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the \regexes{}
+line against each regex, counting the number of log lines that match
+each regex, warning the user if an input log line is matched by more
+than one regex, and additionally warning if an input log line is not
+matched by any regex.  It displays a summary of how many input log lines
+each regex matched, comparing it to the expected number of matches; this
+provides the user with an easy method of checking if the regexes
 produced by the first utility are correctly matching the input log lines
 they are based upon.  A future version of this utility will also group
-input log lines by \regex{}, so the user can tweak the \regexes{} if
+input log lines by regex, so the user can tweak the regexes if
 required.
 
-These utilities are not intended to create perfect \regexes{}, but they
+These utilities are not intended to create perfect regexes, but they
 greatly reduce the effort required to parse new or different log lines.
@@
 
-The rules and actions exhibit emergent behaviour~\cite{Wikipedia-Emergence}
-that is far more complicated than the behaviour explicitly encoded in the
-algorithm.  The top-level parser could be written in pseudo-code like so
-(with indentation level indicating flow control):
+The rules and actions exhibit emergent behaviour that is far more
+complicated than the behaviour explicitly encoded in the algorithm.  The
+top-level parser could be written in pseudo-code like so (with indentation
+level indicating flow control):
 
@@
 
-    \item [matches] The fields in the line captured by the rule's \regex{}.
+    \item [matches] The fields in the line captured by the rule's regex.
 
@@
     \item [SAVE\_DATA] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the \regex{} to it.
+        the log line, and save the data extracted by the regex to it.
 
@@
     \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal~\cite{Wikipedia-unix-signals}; the active connection for
+        signal; the active connection for
         that \daemon{smtpd} must be cleaned up, entered in the database,
@@
         infrequently, as there are many other timeouts that should occur
-        in the intervening hours: \DNS{} timeouts, timeouts reading data
+        in the intervening hours: DNS timeouts, timeouts reading data
         from the client, etc\@.  The active connection for that
@@
 came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create \regexes{} for new
+algorithm for analysing unparsed log lines to create regexes for new
 rules.  Detecting, diagnosing and defeating complications forms the largest

### 2008-10-31 doc/logparser-introduction.tex
Finished corrections; 112 XXX notes.

@@
 with a description of using the tools provided with the parser to generate
-new rules (specifically the \regex{} in each rule) from unparsed log lines.
+new rules (specifically the regex in each rule) from unparsed log lines.
 

### 2008-10-31 doc/logparser-other-parsers-review.tex
Finished corrections; 112 XXX notes.

@@
 extract additional data.  The code is structured as a long series of blocks
-that each handle all lines matching a single \regex{}, so parsing new log
-lines requires modifying an existing \regex{} or carefully inserting a new
+that each handle all lines matching a single regex, so parsing new log
+lines requires modifying an existing regex or carefully inserting a new
 block in the correct place; extracting extra data will require modifying
-multiple blocks, \regexes{}, or both.
+multiple blocks, regexes, or both.
 
@@
 % Hard to extend/expand/understand.  Appears to only save: date and hour,
-% \DNS{} name and \IP{} address host, mail server \IP{} address, sender,
+% DNS name and \IP{} address host, mail server \IP{} address, sender,
 % receiver and e-mail status (sent, rejected).  Undocumented schema.
@@
 Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line does not match the \regex{}, though it appears that
+saved when the log line does not match the regex, though it appears that
 they will not overwrite existing data.  Most notably rejections are

### 2008-10-31 doc/logparser-results.tex
Finished corrections; 112 XXX notes.

@@
 
-Before the perl interpreter attempts to match a \regex{} against a piece of
-text, the \regex{} is compiled into an internal representation and
+Before the perl interpreter attempts to match a regex against a piece of
+text, the regex is compiled into an internal representation and
 optimised to improve the speed of matching.  This compilation and
 optimisation takes CPU time ---in many cases it takes far more CPU time
-than the actual matching.  If the interpreter is certain that a \regex{}
+than the actual matching.  If the interpreter is certain that a regex
 will not change it will automatically cache the compilation results for
-later use.   The results of compiling a dynamically generated \regex{} can
-be cached and used in preference to the original \regex{}, but it is the
+later use.   The results of compiling a dynamically generated regex can
+be cached and used in preference to the original regex, but it is the
 responsibility of the programmer to do this.  The Postfix parser loads the
-rules it uses from a database, and thus each rule's \regex{} is compiled
+rules it uses from a database, and thus each rule's regex is compiled
 and the result cached to reduce parsing time.
@@
 \Graphref{Increase in parsing time when not caching compiled regexes graph}
-shows the impact that not caching compiled \regexes{} has on parser
+shows the impact that not caching compiled regexes has on parser
 performance: on typical log files the parsing time when not caching
-compiled \regexes{} is 400--600\% of the parsing time when caching.
-Caching compiled \regexes{} is probably the single most effective
+compiled regexes is 400--600\% of the parsing time when caching.
+Caching compiled regexes is probably the single most effective
 optimisation possible in the parser's implementation, and was quite simple
@@
 of rules consulted per log line is much lower: this results in far fewer
-\regex{} compilations per line when there is a mail loop, and a
+regex compilations per line when there is a mail loop, and a
 corresponding decrease in the average parsing time for a single log line.
-The outcome is that caching compiled \regexes{} is proportionally less
+The outcome is that caching compiled regexes is proportionally less
 important when the log file contents were created by a mail loop.  The
-increases in parsing time when not caching compiled \regexes{} for
+increases in parsing time when not caching compiled regexes for
 different groups of log files are summarised in \tableref{Increase in
@@
 ordering on parsing time, and the unexpected consequences of specific
-inputs.  The necessity of caching compiled \regexes{} is attested to by the
+inputs.  The necessity of caching compiled regexes is attested to by the
 third group of graphs, where the difference between caching and not caching
@@
 number of hits per rule is quite unevenly spread, resembling a Power Law
-distribution~\cite{powerlaw} --- it is obvious that a small number of rules
-match the vast majority of the lines, and more than half the rules match
-fewer than 100 times.
+distribution --- it is obvious that a small number of rules match the vast
+majority of the lines, and more than half the rules match fewer than 100
+times.
 
@@
 negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which \regex{} parsed it, there
-is no way to be sure that every line was parsed by the correct \regex{},
+of the 60,721,709 log lines and determining which regex parsed it, there
+is no way to be sure that every line was parsed by the correct regex,
 making it impossible to quantify the false positive rate; however a random
 sample of 6,039 log lines was parsed and the results checked manually to
-ensure that the correct \regex{} parsed each log line.\footnote{Each log
-line was examined and the correct \regex{} identified from the
-\numberOFrules{} rules in the database; the correct \regex{} was then
-compared to the \regex{} that was used by the parser.}  The sample was
+ensure that the correct regex parsed each log line.\footnote{Each log
+line was examined and the correct regex identified from the
+\numberOFrules{} rules in the database; the correct regex was then
+compared to the regex that was used by the parser.}  The sample was
 generated by running the following command:

### 2008-10-31 doc/logparser.tex
First half of the corrections to section 5.
Fixed underfull hbox warnings. Dealt with more corrections; nearly finished (if I ignore the 103 XXX notes).
Finished corrections; 112 XXX notes.

@@
 
+% Show the citation key (e.g. slct-paper) in the margin beside the
+% reference.
+\usepackage[nomsgs]{refcheck}
+% Show where references are used.
+% Neither work.
+%\usepackage{citeref}
+%\usepackage{backref}
 \renewcommand{\refname}{Bibliography}
+% Add space before urls to quell underfull hbox warnings.
+\newcommand{\urlprefix}[0]{\hfill{}}
 % New formatting commands.
@@
 
+XXX WOULD TABLES LOOK BETTER IF CAPTIONS WERE LEFT-ALIGNED\@?
+
+XXX SWITCH TO NEW GLOSSARY PACKAGE\@.
+
+XXX IS IT POSSIBLE TO HAVE ACRONYMS THAT ARE NOT INCLUDED IN THE ACRONYM
+LIST\@?
+
+XXX REDUCE THE LIST OF ACRONYMS\@.
+
+XXX CHECK THAT I SAY log line AND log file EVERYWHERE\@.
+
+XXX CHECK YEARS FOR BIBLIOGRAPHY ENTRIES\@.
+
+XXX IMPROVE BIBLIOGRAPHY APPEARANCE\@: NAMES LOOK UGLY (E.G.\ P.\ \empty{}.\
+Porters), URL SPACING IS MESSED UP --- URLS \\SHOULD PROBABLY BE LAST, ON THEIR
+OWN LINE\@.
+
 \newpage
@@
 \bibliography{logparser-bibliography}
-\label{bibliography}
 
@@
 % section.
+XXX MOVE POSTFIX DAEMONS INTO A SEPARATE SECTION\@; NEED TO SWITCH TO NEW
+GLOSSARY PACKAGE FIRST\@.
 \renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}

### 2008-11-06 doc/logparser-background.tex
Dealt with some XXX notes.

@@
 the administrator to make catastrophic mistakes, e.g.\ rejecting all mail
---- the administrator must be cognisant of the effects of their
-configuration changes.  This is similar to one of UNIX's design
-philosophies: \textit{``UNIX was not designed to stop its users from doing
-stupid things, as that would also stop them from doing clever
+--- the administrator must be cognisant of the effects their configuration
+changes will have.  This is similar to one of UNIX's design philosophies:
+\textit{``UNIX was not designed to stop its users from doing stupid things,
+as that would also stop them from doing clever
 things''\/}~\cite{unix-philosophy}.
@@
 The restriction \texttt{check\_client\_access} checks if the \IP{} address
-of the connected client is found in the specified file and returns the
+of the connected client is found in the specified table and returns the
 associated action if found; the method of searching the file is dependant
@@
 \tab{}\texttt{reject\_rbl\_client dnsbl.example.com}\newline{} checks the
-\DNSBL{} \texttt{dnsbl.example.com}, rejecting the command if the client's
-\IP{} address is listed.  XXX DO I NEED THIS SENTENCE\@?  This description
-is necessarily brief, for further details see~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}.
+\DNSBL{} \texttt{dnsbl.example.com} and rejects the command if the client's
+\IP{} address is listed.
+
+For further information about Postfix restrictions
+see~\cite{smtpd_access_readme,smtpd_per_user_control,policy-servers}.
 
@@
 
-    XXX HOW CAN I CENTRE THE TABLE\@?\hfill{}
+    \centering{}
 

### 2008-11-06 doc/logparser-conclusion.tex
Dealt with some XXX notes.

@@
 although parsers are often divided (the combination of \texttt{lex} and
-\texttt{yacc} XXX BOOK REF FOR LEX AND YACC being a common example), the
-parts are usually quite internally interdependent, and will be combined
-into a complete parser by the compilation process; in contrast
-\parsername{} keeps the rules and actions separate until the parser runs.
-The actions and framework are not as completely separated, as the actions
-depend on services provided by the framework; however the actions and
-framework are not tightly integrated: it would be possible, with some work,
-to completely separate the two.  XXX I SHOULD DO THAT\@.
+\texttt{yacc}~\cite{lex-and-yacc-book} being a common example), the parts
+are usually quite internally interdependent, and will be combined into a
+complete parser by the compilation process; in contrast \parsername{} keeps
+the rules and actions separate until the parser runs.  The actions and
+framework are not as completely separated, as the actions depend on
+services provided by the framework; however the actions and framework are
+not tightly integrated: it would be possible, with some work, to completely
+separate the two.  XXX I SHOULD DO THAT\@.
 

### 2008-11-06 doc/logparser-implementation.tex
Rearranged sections; no real content changes.
Dealt with some XXX notes.

@@
 \section{Postfix Parser Implementation}
+\label{Postfix Parser Implementation}
 
@@
 
-XXX SHOULD I USE TEXTTT OR SIMILAR ON FIELD NAMES\@?
-
 Rules are discussed in detail in \sectionref{rules}, but the structure of
@@
 
-    \item [postfix\_action] This is the action Postfix must have taken to
-        generate this log line.  This field is mostly ignored, but two
-        values (IGNORED and INFO) have special meaning, as described below
-        in the list of typical values.\label{postfix_action}
-
-        XXX MAYBE MOVE THIS LIST TO AFTER THE FIELD LIST\@?  DO I NEED THIS
-        LIST AT ALL\@?
-
-        \begin{description}
-
-            \item [ACCEPTED] Postfix has accepted a mail, and will
-                subsequently attempt to deliver it.
-
-            \item [BOUNCED] The mail has bounced, because of a mail loop,
-                delivery failure, or five day timeout.\footnote{As with
-                much of Postfix's behaviour, five days is the default value
-                but can be changed by the administrator if they choose.}
-
-            \item [DELETED] The mail was deleted from the queue by an
-                administrator.
-
-            \item [DISCARDED] Postfix discarded the mail it was in the
-                process of accepting, because it was either larger than the
-                size limit set by the administrator, or the client timed
-                out or disconnected.
-
-            \item [EXPIRED] The mail has been in the queue for five
-                days\footnotemark[\value{footnote}] XXX CHECK IF THIS
-                FOOTNOTE WORKS without successful delivery.  A bounce mail
-                will be generated and sent to the sender address.
-
-            \item [INFO] Represents an unspecified intermediate action that
-                the parser is not interested in per se, but that does log
-                useful information, supplementing other log lines.
-
-            \item [IGNORED] An action that is not only uninteresting in
-                itself, but that also provides no useful data.
-
-            \item [POSTFIX\_RELOAD] The administrator has instructed
-                Postfix to start or stop, and all \daemon{smtpd} processes
-                will be terminated.  This does not negatively affect the
-                log files or mail queued by Postfix for delivery.
-
-            \item [PROCESSING] \daemon{cleanup} is processing a mail.
-                XXX ADD MORE DETAIL HERE
-
-            \item [REJECTED] Postfix rejected a command from the remote
-                client, causing at least one recipient to be rejected.
-
-            \item [SENT] Postfix has successfully sent a mail.
-
-        \end{description}
-
-        Uninteresting log lines are parsed so that any lines the parser
-        is not capable of handling become immediately obvious errors.
+    \item [postfix\_action] The action Postfix must have taken to generate
+        this log line; the possible actions are described later in
+        \sectionref{list of postfix actions}.
 
@@
 \end{eqlist}
+
+\label{list of postfix actions}
+
+\begin{description}
+
+    \item [ACCEPTED] Postfix has accepted a mail, and will
+        subsequently attempt to deliver it.
+
+    \item [BOUNCED] The mail has bounced, because of a mail loop,
+        delivery failure, or five day timeout.\footnote{As with
+        much of Postfix's behaviour, five days is the default value
+        but can be changed by the administrator if they choose.}
+
+    \item [DELETED] The mail was deleted from the queue by an
+        administrator.
+
+    \item [DISCARDED] Postfix discarded the mail it was in the
+        process of accepting, because it was either larger than the
+        size limit set by the administrator, or the client timed
+        out or disconnected.
+
+    \item [EXPIRED] The mail has been in the queue for five
+        days\footnotemark[\value{footnote}] XXX CHECK IF THIS
+        FOOTNOTE WORKS without successful delivery.  A bounce mail
+        will be generated and sent to the sender address.
+
+    \item [INFO] Represents an unspecified intermediate action that
+        the parser is not interested in per se, but that does log
+        useful information, supplementing other log lines.
+
+    \item [IGNORED] An action that is not only uninteresting in
+        itself, but that also provides no useful data.
+
+    \item [POSTFIX\_RELOAD] The administrator has instructed
+        Postfix to start or stop, and all \daemon{smtpd} processes
+        will be terminated.  This does not negatively affect the
+        log files or mail queued by Postfix for delivery.
+
+    \item [PROCESSING] \daemon{cleanup} is processing a mail.
+        XXX ADD MORE DETAIL HERE
+
+    \item [REJECTED] Postfix rejected a command from the remote
+        client, causing at least one recipient to be rejected.
+
+    \item [SENT] Postfix has successfully sent a mail.
+
+\end{description}
 
@@
 
-\subsection{Parser Flow Chart}
-
-\label{flow-chart}
-
-XXX MIGHT NEED A SMALL INTRO\@.
-
-XXX MOVE TO AFTER ACTIONS\@.
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \SMTP{} or local submission.
-
-    \item If the mail is rejected, save all data and finish.
-
-    \item Follow the progress of the accepted mail until it is either
-        delivered, bounced, or deleted, then save all data, and finish.
-
-\end{enumerate}
-
-\Figureref{flow chart image} shows the paths the data representing a mail
-can take through the parser algorithm.  The flow chart covers the most
-common paths only, the complications described in
-\sectionref{complications} are excluded for the sake of clarity.  XXX WHERE
-ARE THE PATHS DESCRIBED\@?
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart}{flow chart
-image}
-
-\clearpage
-
-\subsubsection{Mail enters the system}
-
-\label{mail-enters-the-system}
-
-XXX SHOULD THESE BE SEPARATE SUBSUBSECTIONS\@?
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop}, by \SMTP{}, by re-injection due to
-forwarding, or internally generated by Postfix.  Local submission is the
-simplest case: a queueid is assigned immediately and the sender address is
-logged (action: PICKUP\@; flowchart:~2).
-
-\SMTP{} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there will not be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own; it is explained in \sectionref{Re-injected mails}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics as described in
-\sectionref{identifying-bounce-notifications}.  Bounce notifications are
-the primary example of internally generated mails, though there may be
-other types.  Postfix may generate mails to the administrator when it
-encounters configuration errors, but such mails are presumably rare.
-
-\subsubsection{Mail delivery}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_DATA\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
-
-\subsubsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-XXX DUNNO IF I SHOULD INCLUDE THIS SECTION\@; AT BEST IT NEEDS TO BE
-UPDATED, AT WORST IT'S RUBBISH\@.
-
-The rules and actions exhibit emergent behaviour that is far more
-complicated than the behaviour explicitly encoded in the algorithm.  The
-top-level parser could be written in pseudo-code like so (with indentation
-level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line that is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(\figureref{flow chart image}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
-
 \subsection{Actions}
@@
         and \daemon{cleanup} as it needs to deal with out of order log
-        lines; see \sectionref{discarding cleanup lines} for details.  XXX
-        EXPLAIN THIS BETTER\@.
+        lines; see \sectionref{discarding cleanup lines} for details.
+        XXXXXX EXPLAIN THIS BETTER\@.
 
@@
 
+\subsection{Framework}
+
+XXX TO BE WRITTEN\@.
+
+\subsection{Parser Flow Chart}
+
+\label{flow-chart}
+
+XXX MIGHT NEED A SMALL INTRO\@.
+
+From the viewpoint of an individual mail passing through the parser the
+experience could be summarised as:
+
+\begin{enumerate}
+
+    \item Mail enters the system via \SMTP{} or local submission.
+
+    \item If the mail is rejected, save all data and finish.
+
+    \item Follow the progress of the accepted mail until it is either
+        delivered, bounced, or deleted, then save all data, and finish.
+
+\end{enumerate}
+
+\Figureref{flow chart image} shows the paths the data representing a mail
+can take through the parser algorithm.  The flow chart covers the most
+common paths only, the complications described in
+\sectionref{complications} are excluded for the sake of clarity.  XXX WHERE
+ARE THE PATHS DESCRIBED\@?
+
+\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart}{flow chart
+image}
+
+\clearpage
+
+\subsubsection{Mail enters the system}
+
+\label{mail-enters-the-system}
+
+XXX SHOULD THESE BE SEPARATE SUBSUBSECTIONS\@?
+
+Everything starts off with a mail entering the system, whether by local
+submission via \daemon{postdrop}, by \SMTP{}, by re-injection due to
+forwarding, or internally generated by Postfix.  Local submission is the
+simplest case: a queueid is assigned immediately and the sender address is
+logged (action: PICKUP\@; flowchart:~2).
+
+\SMTP{} is more complicated:
+
+\begin{enumerate}
+
+    \item First there is a connection from the remote client (action:
+        CONNECT\@; flowchart:~1).
+
+    \item This is followed by rejection of sender address, recipient
+        addresses, client \IP{} address or hostname, HELO hostname, etc.\
+        (action: REJECTION\@; flowchart:~4); acceptance of one or more
+        mails (action: CLONE\@; flowchart:~5); or some interleaving of
+        both.
+
+    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
+        Postfix has rejected any \SMTP{} commands the data will be saved to
+        the database; if not there will not be any data to save (any mails
+        accepted will already have been cloned so their data is in another
+        data structure).
+
+    \item If one or more mails were accepted there will be more log entries
+        for those mails later, see \sectionref{mail-delivery}.
+
+\end{enumerate}
+
+Re-injection due to forwarding sadly lacks explicit log lines of its
+own; it is explained in \sectionref{Re-injected mails}.
+
+Internally generated mails lack any explicit origin in Postfix 2.2.x and
+must be detected using heuristics as described in
+\sectionref{identifying-bounce-notifications}.  Bounce notifications are
+the primary example of internally generated mails, though there may be
+other types.  Postfix may generate mails to the administrator when it
+encounters configuration errors, but such mails are presumably rare.
+
+\subsubsection{Mail delivery}
+
+\label{mail-delivery}
+
+The obvious counterpart to mail entering the system is mail leaving the
+system, whether by deletion, bouncing, local delivery, or remote delivery.
+All four are handled in exactly the same way:
+
+\begin{enumerate}
+
+    \item Postfix will log the sender and recipient addresses separately
+        (action: SAVE\_DATA\@; flowchart:~9).
+
+    \item Sometimes mail is re-injected and the child mail needs to be
+        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
+        \sectionref{tracking re-injected mail} discusses this in
+        detail.
+
+    \item Eventually the mail will be delivered, bounced, or deleted by the
+        administrator (action: COMMIT\@; flowchart:~12).  This is the last
+        log line for this particular mail (though it may be indirectly
+        referred to if it was re-injected).  If it is neither parent nor
+        child of re-injection the data is cleaned up and entered in the
+        database (flowchart:~14), then deleted from the state tables.
+        Re-injected mails are described in \sectionref{tracking re-injected
+        mail}.
+
+\end{enumerate}
+
+It should be reiterated that the actions above happen whether the mail is
+delivered to a mailbox, piped to a command, delivered to a remote server,
+bounced (due to a mail loop, delivery failure, or five day timeout), or
+deleted by the administrator, \textit{unless\/} the mail is either parent
+or child of re-injection, as explained in \sectionref{tracking re-injected
+mail}.
+
+\subsubsection{Emergent behaviour}
+
+\label{Emergent behaviour}
+
+XXX DUNNO IF I SHOULD INCLUDE THIS SECTION\@; AT BEST IT NEEDS TO BE
+UPDATED, AT WORST IT'S RUBBISH\@.
+
+The rules and actions exhibit emergent behaviour that is far more
+complicated than the behaviour explicitly encoded in the algorithm.  The
+top-level parser could be written in pseudo-code like so (with indentation
+level indicating flow control):
+
+\begin{verbatim}
+for each line in the input files:
+    for each rule defined by the user:
+        if this rule matches the input line:
+            perform the action specified by the rule
+            skip the remaining rules
+            process the next input line
+    warn the user that the input line was not parsed
+\end{verbatim}
+
+Yet despite the brevity of the algorithm above the (simplified) flow chart
+in \sectionref{flow-chart} contains:
+
+\begin{itemize}
+
+    \item Twenty one states (not including explicit branches).
+
+    \item Three entry points (1, 2, 3).
+
+    \item Five exit points (8, 14, 17, 20, 25).
+
+    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
+
+    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
+        by an action, determining what will happen next to a mail.
+
+    \item Four implicit branches, where the transition is determined by the
+        next log line that is processed for that mail, e.g.\ state 1 can
+        transition to either state 4 or state 5.  (1--\{4,5\},
+        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
+
+\end{itemize}
+
+Several more states and corresponding branches are omitted for the sake of
+clarity (these extra states are caused by solving the complications
+described in \sectionref{complications}).
+
+The actions introduce some of the complexity seen in the flow chart: all
+explicit branches and the transitions between some stages (19--20,
+21--22--23--24) in the second half of the flow chart are encoded within the
+actions; the transitions between the other stages (16--17, 24--25) in the
+second half arise from the interaction between actions
+
+However the transitions between stages in the first half of the flow chart
+(\figureref{flow chart image}) are not encoded anywhere in
+the parser; the transitions are determined by the ordering of log lines in
+the log file.  The complexity of the first 12 stages in the flow chart
+emerges from the interaction of simple rules and actions, easing the
+process of adding new actions (described in \sectionref{adding new
+actions}), as there is no requirement to explicitly insert the action into
+an algorithmic version of the flow chart.\footnote{The solution to the
+complication in \sectionref{out of order log lines} (out of order log
+lines) requires a list of the acceptable combinations of Postfix programs a
+mail can pass through; however this does not correspond to the
+\textit{parser actions\/} a mail must pass through.  Before that
+complication was overcome, i.e.\ for the first half of the development of
+this parser, there was nothing in the parser explicitly encoding the states
+or paths a mail could take; no other component in the parser has any need
+for this flow of data to be specified.}
+
 \subsection{Complications encountered}

### 2008-11-06 doc/logparser-introduction.tex
Dealt with some XXX notes.

@@
 
-Section~XXX contains the core of the paper, describing
-a naive parsing algorithm and the complications initially encountered that
-shaped the full algorithm.  A flow chart and a discussion of the emergent
-behaviour exhibited by the algorithm accompanies a comprehensive
-explanation of the different stages of the initial algorithm.  The
-framework that actions and rules fit into is documented, then the actions
-taken during execution of the algorithm are described, followed by the
-process of adding a new action.  The section concludes with an in-depth
+Section~\ref{Postfix Parser Implementation} contains the core of the paper,
+describing a naive parsing algorithm and the complications initially
+encountered that shaped the full algorithm.  A flow chart and a discussion
+of the emergent behaviour exhibited by the algorithm accompanies a
+comprehensive explanation of the different stages of the initial algorithm.
+The framework that actions and rules fit into is documented, then the
+actions taken during execution of the algorithm are described, followed by
+the process of adding a new action.  The section concludes with an in-depth
 description of the further complications discovered, and their solutions

### 2008-11-06 doc/logparser-other-parsers-review.tex
Dealt with some XXX notes.

@@
 (\url{http://www.oracle.com/database/berkeley-db/index.html}) database.
-XXX SHOULD THOSE URLS BE REFERENCES INSTEAD\@?  The decision to support
-both \CSV{} and Berkeley DB appears to have been a serious limitation:
-\CSV{} is a very simple format where a record is stored in a single line,
-with fields separated by a comma or other punctuation symbol.  Problems
-with \CSV{} files include the need to escape separators in the data stored,
-providing multiple values for a field (e.g.\ multiple recipients), and
-adding new fields.  There is no mechanism to document the fields or the
-separator, unlike \SQL{} databases where every database includes a schema
-naming the fields and the type of data they store (integer, text,
-timestamp, etc.).
+The decision to support both \CSV{} and Berkeley DB appears to have been a
+serious limitation: \CSV{} is a very simple format where a record is stored
+in a single line, with fields separated by a comma or other punctuation
+symbol.  Problems with \CSV{} files include the need to escape separators
+in the data stored, providing multiple values for a field (e.g.\ multiple
+recipients), and adding new fields.  There is no mechanism to document the
+fields or the separator, unlike \SQL{} databases where every database
+includes a schema naming the fields and the type of data they store
+(integer, text, timestamp, etc.).
 
@@
 
-Isoqlog's report lists daily, monthly, and annual totals of mails sent by
-each unique sender and bytes transferred (XXX EXPLAIN BYTES TRANSFERRED\@:
-TO/FROM, PER-SENDER\@?).  It only produces these reports for the domains
-listed in its configuration file, making it impossible to produce reports
-for every sender domain.  It ignores all log lines except those for today,
-so it is impossible to analyse historical logs, and testing with the
-\numberOFlogFILES{} test log files is pointless.  It does maintain a record
-of data previously extracted, which the newly extracted data is merged into
-(no information is provided on the format of the data store).  The data
-extracted is limited to the number of mails sent by each sender, unlike the
-copious amounts of data extracted by \parsername{}.  It does not utilise
-rejection log lines in any way, so is unsuitable for the purposes of this
-project.  Its parsing is completely inextensible, indeed is almost
-incomprehensible, relying on \texttt{scanf(3)}, unexplained fixed offsets,
-and low level string manipulation; it is the opposite end of the spectrum
-to \parsernames{} parsing.  It does not handle any of the complications
-discussed in \sectionref{complications}, does not gather the breadth of
-data required for this project, and ignores the majority of log lines
-produced by Postfix.
+Isoqlog's report misses most of the information gathered by \parsername{},
+and it only reports on the domains listed in its configuration file, making
+it impossible to produce reports for every sender domain.  It ignores all
+log lines except those for today, so it is impossible to analyse historical
+logs, and testing with the \numberOFlogFILES{} test log files is pointless.
+It does maintain a record of data previously extracted, which the newly
+extracted data is merged into (no information is provided on the format of
+the data store).  The data extracted is limited to the number of mails sent
+by each sender, unlike the copious amounts of data extracted by
+\parsername{}.  It does not utilise rejection log lines in any way, so is
+unsuitable for the purposes of this project.  Its parsing is completely
+inextensible, indeed is almost incomprehensible, relying on
+\texttt{scanf(3)}, unexplained fixed offsets, and low level string
+manipulation; it is the opposite end of the spectrum to \parsernames{}
+parsing.  It does not handle any of the complications discussed in
+\sectionref{complications}, does not gather the breadth of data required
+for this project, and ignores the majority of log lines produced by
+Postfix.
 

### 2008-11-22 doc/logparser-acronyms.tex
Dealt with some XXX notes.
Switched to new glossaries package. A couple of other XXX items have been done too.

@@
 % % vim: set textwidth=75 spell :
+% Note: every acronym is listed twice: once so it can be expanded, and once
+% so it will appear with a long description in the glossary.
 % Warning: do not use acronyms within definitions, they do not work
@@
 
-\newacronym{PLP}{Postfix Log Parser}{description={
-    Postfix Log Parser is the parser implementing the algorithm described
-    in this thesis.
-}}
+\newglossaryentry{PLP}{
+    type=\acronymtype,
+    name={PLP},
+    description={Postfix Log Parser\glsadd{PLP glossary}}
+}
+\newglossaryentry{PLP glossary}{
+    name={Postfix Log Parser},
+    description={
+        Postfix Log Parser is the parser implementing the algorithm
+        described in this thesis.
+    }
+}
 
-\newacronym{DNSBL}{DNS Blacklist}{description={
+\newglossaryentry{DNSBL}{
+    type=\acronymtype,
+    name={DNSBL},
+    description={DNS Blacklist\glsadd{DNSBL glossary}}
+}
+\newglossaryentry{DNSBL glossary}{
+    name={DNS Blacklist},
+    description={
         A DNS Blacklist~\cite{Wikipedia-DNSBL} is a simple collaborative
         anti-spam technique used to reject or penalise email sent from mail
-    servers reported to have sent large volumes of spam.  To use a DNSBL
-    the mail server makes a DNS request incorporating the IP address of the
-    client; if the requested hostname exists the client is on the DNSBL,
-    and the mail server can decide what course of action to take.
-}}
+        servers reported to have sent large volumes of spam.  To use a
+        DNSBL the mail server makes a DNS request incorporating the IP
+        address of the client; if the requested hostname exists the client
+        is on the DNSBL, and the mail server can decide what course of
+        action to take.
+    }
+}
 
-\newacronym{API}{Application Programming Interface}{description={
-    One of the fundamental concepts when writing programs is the reuse of
-    existing code, so that each program does not reinvent the wheel.  When
-    a body of code is intended to be reused repeatedly, the user of this
-    code needs to be informed of the functionality provided by the code.
-    An API defines the interface provided to the user, and acts as a
-    contract between the user and the provider: if the user adheres to the
-    API the provider guarantees it will work, but is free to change the
-    implementation provided the API is preserved.
-}}
+\newglossaryentry{API}{
+    type=\acronymtype,
+    name={API},
+    description={Application Programming Interface\glsadd{API glossary}}
+}
+\newglossaryentry{API glossary}{
+    name={Application Programming Interface},
+    description={
+        One of the fundamental concepts when writing programs is the reuse
+        of existing code, so that each program does not reinvent the wheel.
+        When a body of code is intended to be reused repeatedly, the user
+        of this code needs to be informed of the functionality provided by
+        the code.  An API defines the interface provided to the user, and
+        acts as a contract between the user and the provider: if the user
+        adheres to the API the provider guarantees it will work, but is
+        free to change the implementation provided the API is preserved.
+    }
+}
 
-\newacronym{SMTP}{Simple Mail Transfer Protocol}{description={
+\newglossaryentry{SMTP}{
+    type=\acronymtype,
+    name={SMTP},
+    description={Simple Mail Transfer Protocol\glsadd{SMTP glossary}}
+}
+\newglossaryentry{SMTP glossary}{
+    name={Simple Mail Transfer Protocol},
+    description={
         SMTP is the protocol that transfers mail from the sender to the
-    recipient across the Internet.  It is a simple, human readable, plain
-    text protocol, making it quite simple to test and debug problems with
-    it.  A detailed description of \SMTP{} is beyond the scope of this
-    document: the original protocol definition is \RFC{}~821~\cite{RFC821},
-    updated in \RFC{}~2821~\cite{RFC2821}; introductory guides can be found
+        recipient across the Internet.  It is a simple, human readable,
+        plain text protocol, making it quite simple to test and debug
+        problems with it.  A detailed description of SMTP is beyond the
+        scope of this document: the original protocol definition is
+        RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821};
+        introductory guides can be found
         at~\cite{smtp-intro-01,smtp-intro-02}.
-}}
+    }
+}
 
-\newacronym{LMTP}{Local Mail Transfer Protocol}{description={
+\newglossaryentry{LMTP}{
+    type=\acronymtype,
+    name={LMTP},
+    description={Local Mail Transfer Protocol\glsadd{LMTP glossary}}
+}
+\newglossaryentry{LMTP glossary}{
+    name={Local Mail Transfer Protocol},
+    description={
         LMTP is a protocol derived from SMTP that removes the need for the
-    server to maintain a mail delivery queue, instead relying on the client
-    to maintain it.  Typically the client is an MTA, and the server a
-    delivery agent or a mail store.  Full details are available
-    in~\cite{lmtp-rfc}.
-}}
+        server to maintain a mail delivery queue, instead relying on the
+        client to maintain it.  Typically the client is an MTA, and the
+        server a delivery agent or a mail store.  Full details are
+        available in~\cite{lmtp-rfc}.
+    }
+}
 
-\newacronym{MTA}{Mail Transfer Agent}{description={
+\newglossaryentry{MTA}{
+    type=\acronymtype,
+    name={MTA},
+    description={Mail Transfer Agent\glsadd{MTA glossary}}
+}
+\newglossaryentry{MTA glossary}{
+    name={Mail Transfer Agent},
+    description={
         A Mail Transfer Agent sends and receives mail via
-    SMTP\backslashat{}.  XXX DO I NEED THIS\backslashat{}?
-}}
+        SMTP\backslashat{}.  Users submit mail to an MTA via their mail
+        client (e.g.\ Microsoft Outlook, Thunderbird, webmail services);
+        the sending MTA transfers the mail to the receiving MTA (which may
+        in turn forward it to another MTA).
+    }
+}
 
-\newacronym{RFC}{Request For Comments}{description={
+\newglossaryentry{RFC}{
+    type=\acronymtype,
+    name={RFC},
+    description={Request For Comments\glsadd{RFC glossary}}
+}
+\newglossaryentry{RFC glossary}{
+    name={Request For Comments},
+    description={
         The Request For Comments series is a series of proposals defining
-    various protocols and file formats, e.g.\ SMTP\backslashat{}.  The name
-    is somewhat misleading: initially the authors were asking for peer
-    review, but these documents are now the de facto standards the Internet
-    runs on.
-}}
+        various protocols and file formats, e.g.\ SMTP\backslashat{}.  The
+        name is somewhat misleading: initially the authors were asking for
+        peer review, but these documents are now the de facto standards the
+        Internet runs on.
+    }
+}
 
-\newacronym{SQL}{Structured Query Language}{description={
-    SQL is the standard language used for database querying, modification
-    and maintenance.  Some information about SQL, including its history,
-    can be found at \url{http://en.wikipedia.org/wiki/SQL}; a good
-    introduction can be found at \url{http://philip.greenspun.com/sql/}.
-}}
+\newglossaryentry{SQL}{
+    type=\acronymtype,
+    name={SQL},
+    description={Structured Query Language\glsadd{SQL glossary}}
+}
+\newglossaryentry{SQL glossary}{
+    name={Structured Query Language},
+    description={
+        SQL is the standard language used for database querying,
+        modification and maintenance.  Some information about SQL,
+        including its history, can be found at
+        \url{http://en.wikipedia.org/wiki/SQL}; a good introduction can be
+        found at \url{http://philip.greenspun.com/sql/}.
+    }
+}
 
-\newacronym{LMA}{Log Mail Analyzer}{description={
-    One of the other Postfix log parsers reviewed and the only example of
-    prior published paper~\cite{log-mail-analyser} found that specifically
-    deals with parsing Postfix log files.
-}}
+\newglossaryentry{LMA}{
+    type=\acronymtype,
+    name={LMA},
+    description={Log Mail Analyzer\glsadd{LMA glossary}}
+}
+\newglossaryentry{LMA glossary}{
+    name={Log Mail Analyzer},
+    description={
+        One of the other Postfix log file parsers reviewed and the only
+        example of prior published paper~\cite{log-mail-analyser} found
+        that specifically deals with parsing Postfix log files.
+    }
+}
 
-\newacronym{CSV}{Comma-Separated Value}{description={
-    The most basic form of database available, each record is a single line
-    in the file and the fields are separated by a special character,
-    typically a comma or colon.
-}}
+\newglossaryentry{CSV}{
+    type=\acronymtype,
+    name={CSV},
+    description={Comma-Separated Value\glsadd{CSV glossary}}
+}
+\newglossaryentry{CSV glossary}{
+    name={Comma-Separated Value},
+    description={
+        The most basic form of database available, each record is a single
+        line in the file and the fields are separated by a special
+        character, typically a comma or colon.
+    }
+}
 
-\newacronym{FQDN}{Fully Qualified Domain Name}{description={
+\newglossaryentry{FQDN}{
+    type=\acronymtype,
+    name={FQDN},
+    description={Fully Qualified Domain Name\glsadd{FQDN glossary}}
+}
+\newglossaryentry{FQDN glossary}{
+    name={Fully Qualified Domain Name},
+    description={
         An FQDN is a hostname plus domain name, e.g.\ \newline{}
@@
             www.example.com is a FQDN
-}}
+    }
+}
 
-\newacronym{pid}{Process Identifier}{description={
-    There may be multiple copies of any program executing at any one time,
-    so the program's name is not suitable as a distinguishing identifier;
-    instead each process executing is given a pid that is guaranteed to be
-    unique for the lifetime of the process.  Once the process has
-    completed, the pid may be reused, as they are drawn from a finite pool.
-}}
+\newglossaryentry{pid}{
+    type=\acronymtype,
+    name={pid},
+    description={Process Identifier\glsadd{pid glossary}}
+}
+\newglossaryentry{pid glossary}{
+    name={Process Identifier},
+    description={
+        There may be multiple copies of any program executing at any one
+        time, so the program's name is not suitable as a distinguishing
+        identifier; instead each process executing is given a pid that is
+        guaranteed to be unique for the lifetime of the process.  Once the
+        process has completed, the pid may be reused, as they are drawn
+        from a finite pool.
+    }
+}
 
-\newacronym{UCE}{Unsolicited Commercial Email}{description={
-    UCE is a more restrictive definition of spam than most people use: it
-    only covers mail that is explicitly commercial, thus excluding viruses,
-    Bayesian poisoning mails, backscatter, and those annoying chain letters
-    you get from friends.
-}}
+\newglossaryentry{UCE}{
+    type=\acronymtype,
+    name={UCE},
+    description={Unsolicited Commercial Email\glsadd{UCE glossary}}
+}
+\newglossaryentry{UCE glossary}{
+    name={Unsolicited Commercial Email},
+    description={
+        UCE is a more restrictive definition of spam than most people use:
+        it only covers mail that is explicitly commercial, thus excluding
+        viruses, Bayesian poisoning mails, backscatter, and those annoying
+        chain letters you get from friends.
+    }
+}
 
-% I think this is better than having both regex and regexes in the acronym
-% list.  Likewise for pids.
-\newcommand{\pids}{\pid{}s}
+\newglossaryentry{ESMTP}{
+    type=\acronymtype,
+    name={ESMTP},
+    description={Extended SMTP\glsadd{ESMTP glossary}}
+}
+\newglossaryentry{ESMTP glossary}{
+    name={Extended SMTP},
+    description={
+        ESMTP is Extended SMTP, defined in RFC~1869~\cite{RFC1869}.
+    }
+}
 
-\newacronym{ESMTP}{Extended SMTP}{description={
-    ESMTP is Extended SMTP, defined in RFC~1869~\cite{RFC1869}.
-}}
+\newglossaryentry{IP}{
+    type=\acronymtype,
+    name={IP},
+    description={Internet Protocol\glsadd{IP glossary}}
+}
+\newglossaryentry{IP glossary}{
+    name={Internet Protocol},
+    description={
+        The Internet Protocol is the protocol used to communicate between
+        computers on the Internet.  An IP address is a unique address
+        assigned to a computer on the Internet, allowing it to communicate
+        with other computers on the Internet.
+    }
+}
 
-\newacronym{IP}{Internet Protocol}{description={
-    The Internet Protocol is the protocol used to communicate between
-    computers on the Internet.  An IP address is a unique address assigned
-    to a computer on the Internet, allowing it to communicate with other
-    computers on the Internet.
-}}
+\newglossaryentry{SLCT}{
+    type=\acronymtype,
+    name={SLCT},
+    description={Simple Logfile Clustering Tool\glsadd{SLCT glossary}}
+}
+\newglossaryentry{SLCT glossary}{
+    name={Simple Logfile Clustering Tool},
+    description={
+        SLCT~\cite{slct-paper} is a tool implementing an algorithm designed
+        by Risto Vaarandi for identifying, grouping and producing a regex
+        to match similar log lines.
+    }
+}
 
-\newacronym{SLCT}{Simple Logfile Clustering Tool}{description={
-    SLCT~\cite{slct-paper} is a tool implementing an algorithm designed by
-    Risto Vaarandi for identifying, grouping and producing a regex to match
-    similar log lines.
-}}
+\newglossaryentry{ATN}{
+    type=\acronymtype,
+    name={ATN},
+    description={Augmented Transition Networks\glsadd{ATN glossary}}
+}
+\newglossaryentry{ATN glossary}{
+    name={Augmented Transition Networks},
+    description={
+        Originally described in~\cite{atns} and further in~\cite{nlpip},
+        Augmented Transition Networks are a tool used in Computational
+        Linguistics for creating grammars to parse or generate sentences,
+    }
+}
 
-\newacronym{ATN}{Augmented Transition Networks}{description={
-    Augmented Transition Networks, originally described in~\cite{atns} and
-    further in~\cite{nlpip}, are a tool used in Computational Linguistics
-    for creating grammars to parse or generate sentences,
-}}
-
-\newacronym{CLI}{Command Line Interface}{description={
+\newglossaryentry{CLI}{
+    type=\acronymtype,
+    name={CLI},
+    description={Command Line Interface\glsadd{CLI glossary}}
+}
+\newglossaryentry{CLI glossary}{
+    name={Command Line Interface},
+    description={
         An computer interface based on typing commands, rather than using a
         mouse.
-}}
+    }
+}
 
-\newacronym{NLP}{Natural Language Processing}{description={
-    Natural Language Processing~\cite{Wikipedia-nlp} attempts to increase
-    our understanding of the languages normally used by humans (e.g.\
-    English, Spanish, Japanese), with the goal of writing programs that
-    can understand human languages.
-}}
+\newglossaryentry{NLP}{
+    type=\acronymtype,
+    name={NLP},
+    description={Natural Language Processing\glsadd{NLP glossary}}
+}
+\newglossaryentry{NLP glossary}{
+    name={Natural Language Processing},
+    description={
+        Natural Language Processing~\cite{Wikipedia-nlp} attempts to
+        increase our understanding of the languages normally used by humans
+        (e.g.\ English, Spanish, Japanese), with the goal of writing
+        programs that can understand human languages.
+    }
+}

### 2008-11-22 doc/logparser-architecture.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

@@
 There is some similarity between the parser's design and William Wood's
-\ATN{}~\cite{atns,nlpip}, a tool used in Computational Linguistics for
+\gls{ATN}~\cite{atns,nlpip}, a tool used in Computational Linguistics for
 creating grammars to parse or generate sentences.  The resemblance between
-\ATN{} and the parser is accidental, but it is obvious that the two
+\gls{ATN} and the parser is accidental, but it is obvious that the two
 different approaches share an underlying separation of concerns; this
@@
         \tableline{}
-        \ATN{}          & Architecture  & Similarity                \\
+        \gls{ATN}          & Architecture  & Similarity                \\
         \tableline{}
@@
 allowing a more specific rule to match some log lines and a more general
-rule to match the majority, e.g.\ separating \SMTP{} delivery to specific
-sites from \SMTP{} delivery to the rest of the world.  The algorithm
-provides a mechanism for ordering overlapping rules: the priority field in
-each rule (defaults to zero).  Negative priorities may be useful for
-catchall rules.
+rule to match the majority, e.g.\ separating \gls{SMTP} delivery to
+specific sites from \gls{SMTP} delivery to the rest of the world.  The
+algorithm provides a mechanism for ordering overlapping rules: the priority
+field in each rule (defaults to zero).  Negative priorities may be useful
+for catchall rules.
 
@@
 
-    \item Sort by regex and visually inspect the list, e.g.\ with \SQL{}
+    \item Sort by regex and visually inspect the list, e.g.\ with \gls{SQL}
         similar to: \textbf{select regex from rules order by regex;}

### 2008-11-22 doc/logparser-background.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

@@
 underlying the project, followed by some technical information: the use of
-a database as an \API{}\@; a brief introduction to \SMTP{}\@; a longer
-introduction to Postfix, concentrating on anti-spam restrictions and policy
-servers.  The assumptions made in designing and implementing the parser are
-explained, as are the conventions used in this thesis.
+a database as an \gls{API}\@; a brief introduction to \gls{SMTP}\@; a
+longer introduction to Postfix, concentrating on anti-spam restrictions and
+policy servers.  The assumptions made in designing and implementing the
+parser are explained, as are the conventions used in this thesis.
 
@@
 a set of functions provided by shared code, it is in effect quite similar
-to any other \API{}: it provides a stable interface for users of the
+to any other \gls{API}: it provides a stable interface for users of the
 interface, allowing the implementation to be changed without adverse
-effects.  As long as the \API{} is maintained, the parser can be improved
-to handle additional complications; support can be added for earlier or
-later releases of Postfix; bugs can be fixed or limitations removed.
+effects.  As long as the \gls{API} is maintained, the parser can be
+improved to handle additional complications; support can be added for
+earlier or later releases of Postfix; bugs can be fixed or limitations
+removed.
 
@@
                 program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \SQL{} guide
+                TRIGGER or two.\footnote{Please refer to an \gls{SQL} guide
                 for explanations of these terms, e.g.\
@@
             \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\SQL{}\@:
+                events occur.  In pseudo-\gls{SQL}\@:
 
@@
 
-    \item \SQL{} is reasonably standard and many people will already be
+    \item \gls{SQL} is reasonably standard and many people will already be
         familiar with it; for those unfamiliar with it there are lots of
         readily available resources from which to learn (a good
-        introduction to \SQL{} can be found at
-        \url{http://philip.greenspun.com/sql/},
-        others are \url{http://www.w3schools.com/sql/default.asp},
-        \url{http://sqlcourse.com/}).  Although
-        every vendor implements a different dialect of \SQL{}, the basics
-        are the same everywhere (analogous to the overall similarities and
-        minor differences between Irish English, British English, American
-        English, and Australian English).  Depending on the database in use
-        there may be tools available that reduce or remove the requirement
-        to know \SQL{}; several are available for SQLite
+        introduction to \gls{SQL} can be found at
+        \url{http://philip.greenspun.com/sql/}, others are
+        \url{http://www.w3schools.com/sql/default.asp},
+        \url{http://sqlcourse.com/}).  Although every vendor implements a
+        different dialect of \gls{SQL}, the basics are the same everywhere
+        (analogous to the overall similarities and minor differences
+        between Irish English, British English, American English, and
+        Australian English).  Depending on the database in use there may be
+        tools available that reduce or remove the requirement to know
+        \gls{SQL}; several are available for \gls{SQLite}
         (\url{http://www.sqlite.org/}), the default database used by the
@@
 
-The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
-in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
-sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
-protocol, making it quite easy to test and debug problems with it.  Despite
-the simplicity of the protocol many virus and spam sending programs fail
-to implement it properly, so requiring strict adherence to the protocol
-specification is beneficial in protecting against spam and
-viruses.\footnote{\label{footnote:rfc760}Originally all mail servers
-adhered to the principle of \textit{Be liberal in what you accept, and
-conservative in what you send\/} from \RFC{}~760~\cite{rfc760}, but
+The \acrlong{SMTP}, originally defined in \gls{RFC}~821~\cite{RFC821} and
+updated in \gls{RFC}~2821~\cite{RFC2821}, is used for transferring mail
+between the sending and receiving \gls{MTA}\@.  It is a simple, human
+readable, plain text protocol, making it quite easy to test and debug
+problems with it.  Despite the simplicity of the protocol many virus and
+spam sending programs fail to implement it properly, so requiring strict
+adherence to the protocol specification is beneficial in protecting against
+spam and viruses.\footnote{\label{footnote:rfc760}Originally all mail
+servers adhered to the principle of \textit{Be liberal in what you accept,
+and conservative in what you send\/} from \gls{RFC}~760~\cite{rfc760}, but
 unfortunately that principle was written in a friendlier time.  Given the
 deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \RFC{}~2821; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  it is not as friendly, nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\SMTP{} conversation resembles the following (the lines starting with a
-three digit number are sent by the server, all other lines are sent by the
-client):
+maxim could be: \textit{Require strict adherence to \gls{RFC}~2821;
+implement the strongest restrictions you can; relax the restrictions and
+adherence only when legitimate mail is impeded.\/}  it is not as friendly,
+nor as catchy, but it more accurately reflects the current situation.} A
+typical \gls{SMTP} conversation resembles the following (the lines starting
+with a three digit number are sent by the server, all other lines are sent
+by the client):
 
@@
 
-A detailed description of \SMTP{} is beyond the scope of this thesis:
+A detailed description of \gls{SMTP} is beyond the scope of this thesis:
 introductory guides can be found at~\cite{smtp-intro-01,smtp-intro-02},
@@
 
-Postfix is a \MTAlong{}~(\MTA{}) with the following design aims (in order
-of importance): security, flexibility of configuration, scalability, and
-high performance.  It features extensive optional anti-spam restrictions,
-allowing an administrator to deploy those restrictions which they judge
-suitable for their site's needs, rather than a fixed set chosen by
-Postfix's author.  These restrictions can be selectively applied, combined,
-and bypassed on a per-client, per-recipient, or per-sender basis, allowing
-varying levels of stricture and permissiveness.  Postfix leverages simple
-lookup tables to support arbitrarily complicated user-defined sequences of
-restrictions and exceptions; policy servers (\sectionref{policy servers})
-provide a simple way to write new restrictions.  Administrators can also
-supply their own rejection messages to make it clear to senders exactly why
-their mail was rejected.  Unfortunately this flexibility has a cost:
-complexity in the log files generated.  Although it is easy to use standard
-Unix text processing utilities to determine the fate of an individual
-email, following the journey an email takes through Postfix can be quite
-difficult.  For most mails the journey is simple and brief, but the
-remaining minority can have quite complex journeys (see
-\sectionref{complications} for details).
+Postfix is a \acrlong{MTA}~(\gls{MTA}) with the following design aims (in
+order of importance): security, flexibility of configuration, scalability,
+and high performance.  It features extensive optional anti-spam
+restrictions, allowing an administrator to deploy those restrictions which
+they judge suitable for their site's needs, rather than a fixed set chosen
+by Postfix's author.  These restrictions can be selectively applied,
+combined, and bypassed on a per-client, per-recipient, or per-sender basis,
+allowing varying levels of stricture and permissiveness.  Postfix leverages
+simple lookup tables to support arbitrarily complicated user-defined
+sequences of restrictions and exceptions; policy servers
+(\sectionref{policy servers}) provide a simple way to write new
+restrictions.  Administrators can also supply their own rejection messages
+to make it clear to senders exactly why their mail was rejected.
+Unfortunately this flexibility has a cost: complexity in the log files
+generated.  Although it is easy to use standard Unix text processing
+utilities to determine the fate of an individual email, following the
+journey an email takes through Postfix can be quite difficult.  For most
+mails the journey is simple and brief, but the remaining minority can have
+quite complex journeys (see \sectionref{complications} for details).
 
@@
 separated into various component programs to perform the tasks required of
-an \MTA{}\@: receive mail, send mail, local delivery of mail, etc.\ ---
+an \gls{MTA}\@: receive mail, send mail, local delivery of mail, etc.\ ---
 full details can be found in~\cite{postfix-overview}.  Each log line
@@
 Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \SMTP{} conversation: client connection, HELO
+for each stage of the \gls{SMTP} conversation: client connection, HELO
 command, MAIL FROM command, RCPT TO commands, DATA command, and end of
@@
 restriction Postfix will evaluate that restriction, allowing restrictions
-to be chosen based on the client \IP{} address, client hostname, HELO
+to be chosen based on the client \gls{IP} address, client hostname, HELO
 hostname, sender address, or recipient address.\footnote{E.g.\ the
@@
 
-The restriction \texttt{check\_client\_access} checks if the \IP{} address
-of the connected client is found in the specified table and returns the
-associated action if found; the method of searching the file is dependant
-on the type of the file (\texttt{cidr} in the example) --- see
+The restriction \texttt{check\_client\_access} checks if the \gls{IP}
+address of the connected client is found in the specified table and returns
+the associated action if found; the method of searching the file is
+dependant on the type of the file (\texttt{cidr} in the example) --- see
 \cite{postfix-lookup-tables} for more details.  Other restrictions
@@
 \tab{}\texttt{reject\_rbl\_client dnsbl.example.com}\newline{} checks the
-\DNSBL{} \texttt{dnsbl.example.com} and rejects the command if the client's
-\IP{} address is listed.
+\gls{DNSBL} \texttt{dnsbl.example.com} and rejects the command if the
+client's \gls{IP} address is listed.
 
@@
 A policy server~\cite{policy-servers} is an external program consulted by
-Postfix to determine the fate of an \SMTP{} command.  The policy server is
-given state information by Postfix (sample state information is shown in
+Postfix to determine the fate of an \gls{SMTP} command.  The policy server
+is given state information by Postfix (sample state information is shown in
 \tableref{Example attributes sent to policy servers}) and returns a verdict
@@
         backscatter~\cite{postfix-backscatter}, and
-        joe~jobs~\cite{Wikipedia-joe-job}.  There has been considerable
-        resistance to the proposal because it breaks or vastly complicates
-        some features of \SMTP{}, e.g.\ forwarding mail from one company or
-        university to another when a user moves.
+        joe~jobs~\cite{Wikipedia-joe-job}\glsadd{Joe job}.  There has been
+        considerable resistance to the proposal because it breaks or vastly
+        complicates some features of \gls{SMTP}, e.g.\ forwarding mail from
+        one company or university to another when a user moves.
 
@@
         rejects a delivery attempt when the triple of \newline{}
-        \tab{}\texttt{(sender address, recipient address, remote \IP{}
+        \tab{}\texttt{(sender address, recipient address, remote \gls{IP}
         address)}\newline{} has not been seen before; on second and
@@
         increasingly complex and well written programs to distribute spam,
-        frequently using an ISP provided \SMTP{} server from a compromised
-        machine on the ISP's network.  Greylisting will slowly become less
-        useful, but it does block a large percentage of spam mail at the
-        moment; the most effective restrictions from the
+        frequently using an ISP provided \gls{SMTP} server from a
+        compromised machine on the ISP's network.  Greylisting will slowly
+        become less useful, but it does block a large percentage of spam
+        mail at the moment; the most effective restrictions from the
         \numberOFlogFILES{} log files used in testing the parser are shown
@@
     \item Scoring systems such as Policyd-weight~\cite{policyd-weight}
-        perform tests on features of the delivery attempt (e.g.\ \IP{}
+        perform tests on features of the delivery attempt (e.g.\ \gls{IP}
         address, sender address), incrementing or decrementing a score
@@
 starting with the motivation behind the project, continuing with an
-explanation of the use of a database as an \API{} and an introduction to
-\SMTP{}.  The focus then switched to the \MTA{} in use, Postfix, and its
-restrictions and policy servers, finishing with the conventions used in
+explanation of the use of a database as an \gls{API} and an introduction to
+\gls{SMTP}.  The focus then switched to the \gls{MTA} in use, Postfix, and
+its restrictions and policy servers, finishing with the conventions used in
 this thesis.

### 2008-11-22 doc/logparser-glossary.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

@@
 % The contents of the glossary.
-\glossary{name={SQLite3},description={
+\newglossaryentry{SQLite}{name={SQLite},description={
     \textit{SQLite is a small C library that implements a self-contained,
-    embeddable, zero-configuration SQL database engine.\/} SQLite3 is an
-    \SQL{} implementation focusing on correctness, simplicity and speed.
-    Unlike other \SQL{} implementations it does not require a separate
+    embeddable, zero-configuration SQL database engine.\/} SQLite is an
+    \gls{SQL} implementation focusing on correctness, simplicity and speed.
+    Unlike other \gls{SQL} implementations it does not require a separate
     server process, greatly simplifying deployment of an application
@@
 
-\glossary{name={Phishing},description={
+\newglossaryentry{Phishing}{name={Phishing},description={
     Phishing~\cite{Wikipedia-phishing} is an attempt to acquire information
@@
 
-\glossary{name={Backscatter},description={
+\newglossaryentry{Backscatter}{name={Backscatter},description={
     When a spam sender or worm sends mail with forged sender addresses,
@@
 
-\glossary{name={Joe~job},description={
-    A joe~job is when spam mail is sent with a faked sender address with
-    the intention of sullying the good name of the owner of the address.
-    joe~jobs are a cause of backscatter, though by no means the only cause.
+\newglossaryentry{Joe job}{name={Joe job},description={
+    A joe~job describes spam mail sent using a faked sender address with
+    the intention of sullying the good name of the user of that address.
+    Joe~jobs are one cause of backscatter.
 }}
 
-\glossary{name={Epoch},description={
+\newglossaryentry{Epoch}{name={Epoch},description={
     Most operating systems store the current time and timestamps of files
@@
 
-\glossary{name={NULL},description={
-    NULL is a special term used in \SQL{} databases indicating the absence
-    of data for the field.
-}}
-
-% Postfix components
-\glossary{name={bounce},description={
-    The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
-}}
-
-\glossary{name={cleanup},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
-}}
-
-\glossary{name={lmtp},description={
-    Delivery of mail over \LMTP{} is performed by the lmtp component.  The
-    definitive documentation is \url{http://www.postfix.org/lmtp.8.html}.
-}}
-
-\glossary{name={local},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
-}}
-
-\glossary{name={pickup},description={
-    Pickup is the service that deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
-}}
-
-\glossary{name={postdrop},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
-}}
-
-\glossary{name={postsuper},description={
-    Maintenance tasks such as deleting mails from the queue, putting mail
-    on hold and later releasing it (no further delivery attempts will be
-    made until it is released), and consistency checking of the mail queue.
-    The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
-}}
-
-\glossary{name={qmgr},description={
-    Qmgr is the Postfix daemon that manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
-    recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
-}}
-
-\glossary{name={sendmail},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands that send mail depend on; it executes \daemon{postdrop} to
-    place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
-}}
-
-\glossary{name={smtp},description={
-    Delivery of mail over \SMTP{} is performed by the smtp component.  The
-    definitive documentation is \url{http://www.postfix.org/smtp.8.html}.
-}}
-
-\glossary{name={smtpd},description={
-    Smtpd is the Postfix program that accepts mail via \SMTP{}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
-}}
-
-\glossary{name={virtual},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
-    determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
-}}
-
-\glossary{name={Bayesian spam filtering},description={
-    Bayesian spam filtering is a method of classifying mail based on the
-    frequency that the words in the mail have previously appeared in a spam
-    corpus and a ham (non-spam) corpus.  A full description is beyond the
-    scope of this thesis, see~\cite{bayesian-filtering,a-plan-for-spam}
-    for a detailed explanation.
-}}
-
-\glossary{name={Bayesian poisoning},description={
-    Bayesian poisoning is the addition of innocuous or unrelated words to a
-    spam mail in the hope of defeating Bayesian spam filtering.  E.g.\ the
-    word Viagra would be firmly in the spam corpus for most people, but by
-    adding the words \textit{schedule}, \textit{meeting}, \textit{moving
-    forward\/} and \textit{best business practices\/} to a mail received by
-    a manager, the Bayesian spam filter might tip the balance from bad to
-    good, if the non-spam words outweigh the spam words.
-}}
-
-\glossary{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail that should not be replied
-    to, e.g.\ bounce notifications.  In \SMTP{} all addresses are enclosed
+\newglossaryentry{<>}{name={$<>$},sort={<>},description={
+    $<>$ is the sender address used for mail that should not be replied to,
+    e.g.\ bounce notifications.  In \gls{SMTP} all addresses are enclosed
     in $<>$, so \textit{username\at{}domain\/} becomes
@@
     must accept mail sent from $<>$, or they are in violation of
-    \RFC{}~2821~\cite{RFC2821}.
+    \gls{RFC}~2821~\cite{RFC2821}.
 }}
 
-\glossary{name={queueid},description={
+\newglossaryentry{queueid}{name={queueid},description={
     Each mail in Postfix's queue is assigned a queueid to uniquely identify
@@
 
-\glossary{name={IPv4},description={
-    Internet Protocol version 4 is the fourth version of the Internet
-    Protocol used to interconnect computers on the Internet.  It is the
-    first widely deployed version of IP, and has been in use for over 25
-    years.
-}}
-
-\glossary{name={IPv6},description={
-    Internet Protocol version 6 is the latest version of the Internet
-    Protocol used to interconnect computers on the Internet.  It is the
-    successor to IPv4, bringing with it a greatly expanded address space,
-    allowing many more computers to use the Internet simultaneously.  IPv4
-    and IPv6 will coexist for many years to come as existing networks
-    transition from the former to the latter.
-}}
-
-%\glossary{name={hash},description={
+%\newglossaryentry{hash}{name={hash},description={
 %    A hashing function transforms a string of characters to a number.
@@
 
-\glossary{name={awk},description={
+\newglossaryentry{awk}{name={awk},description={
     AWK is a general purpose programming language that is designed for
@@
 
-\glossary{name={syslog},description={
+\newglossaryentry{syslog}{name={syslog},description={
     Syslog is the standard logging mechanism on Unix systems: the program
@@
 
-\glossary{name={mail bomb},description={
+\newglossaryentry{mail bomb}{name={mail bomb},description={
     A mail bomb occurs when an attacker inflicts a huge volume of mail on
@@
 
-\glossary{name={mail loop},description={
+\newglossaryentry{mail loop}{name={mail loop},description={
     Sometimes mail sent to one address must be forwarded to a different

### 2008-11-22 doc/logparser-implementation.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

@@
     \item The log files are whole and complete: nothing has been removed,
-        either deliberately or accidentally (e.g.\ log rotation gone awry,
-        file system filling up, logging system unable to cope with the
-        volume of log messages).  On a well run system it is extremely
+        either deliberately or accidentally (e.g.\ log file rotation gone
+        awry, file system filling up, logging system unable to cope with
+        the volume of log messages).  On a well run system it is extremely
         unlikely that any of these problems will arise, though it is of
@@
 In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  There are advantages to
-treating Postfix as a black box during parser development:
+black box program based solely on its inputs and outputs.  There are
+advantages to treating Postfix as a black box during parser development:
 
@@
 %    \item [cluster\_group] A reference to the \texttt{cluster\_group}
-%        table.  That table is used by the Decision Tree algorithm described
-%        in a separate document.
+%        table.  That table is used by the Decision Tree algorithm
+%        described in a separate document.
 
@@
 
-    \item [server\_ip] The \IP{} address (IPv4 or IPv6) of the server: the
-        local address when receiving mail, the remote address when sending
+    \item [server\_ip] The \gls{IP} address (IPv4 or IPv6) of the server:
+        the local address when receiving mail, the remote address when
+        sending mail.
+
+    \item [server\_hostname] The hostname of the server, it will be
+        \texttt{unknown} if the \gls{IP} address could not be resolved to a
+        hostname via DNS\@.
+
+    \item [client\_ip] The client \gls{IP} address (IPv4 or IPv6): the
+        remote address when receiving mail, the local address when sending
         mail.
 
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
-        hostname via DNS\@.
-
-    \item [client\_ip] The client \IP{} address (IPv4 or IPv6): the remote
-        address when receiving mail, the local address when sending mail.
-
     \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \IP{} address could not be resolved to a
+        \texttt{unknown} if the \gls{IP} address could not be resolved to a
         hostname via DNS\@.
@@
         or virus senders think it is a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \SMTP{} command, but it
-        is quite simple to rectify this, as described in
+        logs the HELO hostname when it rejects an \gls{SMTP} command, but
+        it is quite simple to rectify this, as described in
         \sectionref{logging helo}.
@@
     \item [start] The timestamp of the first log line, in seconds since the
-        epoch (explained in the glossary, \sectionref{Glossary}).
+        epoch.\glsadd{Epoch}
 
@@
     \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \SMTP{} command --- a
-        mechanism that is quite useful for testing new restrictions.  This
-        field will be 1 if the log line parsed was a warning rather than a
-        real rejection, or 0 for a real rejection.  This field should be
-        ignored if the result is not a rejection, i.e.\ the action field of
-        the associated rule is not \texttt{REJECTION}.
-
-    \item [smtp\_code] The \SMTP{} code associated with the log line.  In
-        general an \SMTP{} code is only present in a log line for a
-        rejection or final delivery; results initially missing an \SMTP{}
-        code will duplicate the \SMTP{} code of other results in that
-        connection.  Some final delivery log lines do not contain an
-        \SMTP{} code: in those cases the code is specified by the rule,
-        based on the success or failure represented by the log line.
+        enforcing a restriction that would reject an \gls{SMTP} command ---
+        a mechanism that is quite useful for testing new restrictions.
+        This field will be 1 if the log line parsed was a warning rather
+        than a real rejection, or 0 for a real rejection.  This field
+        should be ignored if the result is not a rejection, i.e.\ the
+        action field of the associated rule is not \texttt{REJECTION}.
+
+    \item [smtp\_code] The \gls{SMTP} code associated with the log line.
+        In general an \gls{SMTP} code is only present in a log line for a
+        rejection or final delivery; results initially missing an
+        \gls{SMTP} code will duplicate the \gls{SMTP} code of other results
+        in that connection.  Some final delivery log lines do not contain
+        an \gls{SMTP} code: in those cases the code is specified by the
+        rule, based on the success or failure represented by the log line.
 
@@
     \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from a \DNSBL{}\@.
+        fields, e.g.\ the rejection message from a \gls{DNSBL}\@.
 
     \item [timestamp] The time at which the log line was logged, in seconds
-        since the epoch.
+        since the epoch.\glsadd{Epoch}
 
@@
 because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \DNSBL{}
+restrictions which suits their needs, including using different \gls{DNSBL}
 services, policy servers, or custom rejection messages.  To facilitate
@@
 rules from unparsed log lines, based on the algorithm developed by Risto
-Vaarandi for his \SLCT{}~\cite{slct-paper}.  The differences between the
+Vaarandi for his \gls{SLCT}~\cite{slct-paper}.  The differences between the
 two algorithms will be outlined as part of the general explanation below.
 
-The core of the \SLCT{} algorithm is quite simple: log lines are generally
-created by substituting variable words into a fixed pattern, and analysis
-of the frequency with which each word occurs can be used to determine
-whether the word is variable or part of the fixed pattern.  This
-classification can be used to group similar log lines and generate a
-regex to match each group of log lines.
+The core of the \gls{SLCT} algorithm is quite simple: log lines are
+generally created by substituting variable words into a fixed pattern, and
+analysis of the frequency with which each word occurs can be used to
+determine whether the word is variable or part of the fixed pattern.  This
+classification can be used to group similar log lines and generate a regex
+to match each group of log lines.
 
@@
         substitutions on the input log lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \IP{} addresses, the
+        variable terms (e.g.\ email addresses, \gls{IP} addresses, the
         standard start of rejection messages, etc.) with regex keywords
@@
         was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash~\cite{hash-functions} from a small range of values for each word
-        of each log line, incrementing a counter for each hash.  The
-        counters will be used in the next pass to filter out words: if the
-        word's hash does not have a high frequency, the word itself cannot
-        have a high frequency, so there is no need to maintain a counter
-        for it, reducing the number of counters required and thus the
-        program's memory consumption.
+        pass it generated a hash~\cite{hash-functions} from a small range
+        of values for each word of each log line, incrementing a counter
+        for each hash.  The counters will be used in the next pass to
+        filter out words: if the word's hash does not have a high
+        frequency, the word itself cannot have a high frequency, so there
+        is no need to maintain a counter for it, reducing the number of
+        counters required and thus the program's memory consumption.
 
@@
 
-            \item [pid] The \pid{} of the program that logged the line.
+            \item [pid] The \gls{pid} of the program that logged the line.
 
@@
     \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \pid{}) and save both the
-        client hostname and \IP{} address.
+        table entry (indexed by \daemon{smtpd} \gls{pid}) and save both the
+        client hostname and \gls{IP} address.
 
@@
 
-    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an \SMTP{}
+    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an \gls{SMTP}
         command from the remote client: log the rejection with a mail if
@@
 
-    \item [UNINTERESTING] This rule just returns successfully; it is used when a
-        line needs to be parsed for completeness but does not either
+    \item [UNINTERESTING] This rule just returns successfully; it is used
+        when a line needs to be parsed for completeness but does not either
         provide any useful data or require anything to be done.
@@
 
-    \item Mail enters the system via \SMTP{} or local submission.
+    \item Mail enters the system via \gls{SMTP} or local submission.
 
@@
 Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop}, by \SMTP{}, by re-injection due to
+submission via \daemon{postdrop}, by \gls{SMTP}, by re-injection due to
 forwarding, or internally generated by Postfix.  Local submission is the
@@
 
-\SMTP{} is more complicated:
+\gls{SMTP} is more complicated:
 
@@
     \item This is followed by rejection of sender address, recipient
-        addresses, client \IP{} address or hostname, HELO hostname, etc.\
-        (action: REJECTION\@; flowchart:~4); acceptance of one or more
-        mails (action: CLONE\@; flowchart:~5); or some interleaving of
+        addresses, client \gls{IP} address or hostname, HELO hostname,
+        etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
+        more mails (action: CLONE\@; flowchart:~5); or some interleaving of
         both.
@@
     \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \SMTP{} commands the data will be saved to
-        the database; if not there will not be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log entries
+        Postfix has rejected any \gls{SMTP} commands the data will be saved
+        to the database; if not there will not be any data to save (any
+        mails accepted will already have been cloned so their data is in
+        another data structure).
+
+    \item If one or more mails were accepted there will be more log lines
         for those mails later, see \sectionref{mail-delivery}.
@@
 The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \pid{}, then transition to being
-correlated by the queueid.  This is relatively minor, but does require:
+first be correlated by the \daemon{smtpd} \gls{pid}, then transition to
+being correlated by the queueid.  This is relatively minor, but does
+require:
 
@@
     \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \pid{} or queueid.
+        needs to lookup the data structures by \gls{pid} or queueid.
 
@@
 tell in advance which connections will accept multiple mails.  Once the
-mail has been accepted its log entries will not be correlated by \pid{} any
-more, its queueid will be used instead (except when timeouts occur during
-the data phase \sectionref{timeouts-during-data-phase}).  If the original
-connection has any useful data (e.g.\ rejections) it will be saved to the
-database when the client disconnects.  One unsolved difficulty is
+mail has been accepted its log lines will not be correlated by \gls{pid}
+any more, its queueid will be used instead (except when timeouts occur
+during the data phase \sectionref{timeouts-during-data-phase}).  If the
+original connection has any useful data (e.g.\ rejections) it will be saved
+to the database when the client disconnects.  One unsolved difficulty is
 distinguishing between different groups of rejections, e.g.\ when dealing
@@
 
-    \item The client issues the RSET command to reset the \SMTP{} session.
+    \item The client issues the RSET command to reset the \gls{SMTP}
+        session.
 
@@
 
-    \item [Delivered to a remote \SMTP{} server:] 250 Ok: queued as
+    \item [Delivered to a remote \gls{SMTP} server:] 250 Ok: queued as
         BD07D3C49
@@
         usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \SMTP{} client; local delivery means
-        local to the server, i.e.\ an address the server is final
+        delivery is performed by the \gls{SMTP} client; local delivery
+        means local to the server, i.e.\ an address the server is final
         destination for.} In this case the mail may already have been
@@
 
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline{}
-        delay=0, status=sent (forwarded as 56F5B43FD)}
+        \texttt{3FF7C4317: to=<username@example.com>, relay=local,
+        \newline{} delay=0, status=sent (forwarded as 56F5B43FD)}
 
@@
         (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log entries;
+        (flowchart:~17).  The mail should not have subsequent log lines;
         only its parent should refer to it.
@@
 
-    \item The sender address is \verb!<>!.
+    \item The sender address is \verb!<>!.\glsadd{<>}
 
@@
         messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \SMTP{} or submitted
+        internally by Postfix, not accepted via \gls{SMTP} or submitted
         locally by \daemon{postdrop}.
@@
 
-    \item The sender address must be \verb!<>!.
+    \item The sender address must be \verb!<>!.\glsadd{<>}
 
@@
 
-Some mail clients behave unexpectedly during the \SMTP{} dialogue: the
+Some mail clients behave unexpectedly during the \gls{SMTP} dialogue: the
 client aborts the first delivery attempt after the first recipient is
@@
 reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \IP{} address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  This heuristic is quite restrictive, and
-it appears there is little scope for false positives; if there are any
-false positives there will be warnings when the next log line for that mail
-is parsed.  False negatives are less likely to be detected: there may be
-queueid clashes (and warnings) if mails remain in the state tables after
-they should have been removed, otherwise the only way to detect false
-negatives is to examine the state tables after each parsing run.
+available is the client hostname and \gls{IP} address, but the database
+schema requires many more fields be populated (see \sectionref{connections
+table} and \sectionref{results table}).  This heuristic is quite
+restrictive, and it appears there is little scope for false positives; if
+there are any false positives there will be warnings when the next log line
+for that mail is parsed.  False negatives are less likely to be detected:
+there may be queueid clashes (and warnings) if mails remain in the state
+tables after they should have been removed, otherwise the only way to
+detect false negatives is to examine the state tables after each parsing
+run.
 
@@
 
-    \item Is the mail missing its \daemon{cleanup} log message?  Every mail
+    \item Is the mail missing its \daemon{cleanup} log line?  Every mail
         which passes through Postfix will have a \daemon{cleanup} line;
@@
 
-The DATA phase of the \SMTP{} conversation is where the headers and body of
-the mail are transferred.  Sometimes there is a timeout or the connection
-is lost\footnote{For brevity's sake \textit{timeout\/} will be used
-throughout this section, but everything applies equally to lost
+The DATA phase of the \gls{SMTP} conversation is where the headers and body
+of the mail are transferred.  Sometimes there is a timeout or the
+connection is lost\footnote{For brevity's sake \textit{timeout\/} will be
+used throughout this section, but everything applies equally to lost
 connections.} during the DATA phase; when this occurs Postfix will discard
@@
 The list of valid combinations is explained below.  Every mail will
-additionally have log entries from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log entries from \daemon{bounce}, \daemon{postsuper}, or
+additionally have log line from \daemon{cleanup} and \daemon{qmgr}; any
+mail may also have log line from \daemon{bounce}, \daemon{postsuper}, or
 both.
@@
 \textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log messages.
+anything in the log file to explain why there are no further log lines.
 The only way to detect these mails is to periodically scan all mails in the
@@
         must be exactly two \daemon{smtpd} and one \daemon{cleanup} log
-        entries for the mail, with no additional log entries.
+        lines for the mail, with no additional log lines.
 
@@
 faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it is the result of a mail loop, mail bomb, or
-joe~job) before Postfix has had a chance to try to deliver it.  In this
-case the recipient address will not have been logged, so a dummy recipient
-address needs to be added, as every mail is required by the database schema
+some of the mail (because it is the result of a mail loop\glsadd{mail
+loop}, mail bomb\glsadd{mail bomb}, or joe~job\glsadd{Joe job}) before
+Postfix has had a chance to try to deliver it.  In this case the recipient
+address will not have been logged, so a dummy recipient address needs to be
+added, as every mail is required by the database schema
 (\sectionref{results table}) to have at least one recipient.  This
@@
         existing rules modified to cope with the new or changed log lines.
-        Similarly using a new \DNSBL{}, a new policy server, or new
+        Similarly using a new \gls{DNSBL}, a new policy server, or new
         administrator-defined rejection messages require new rules.
@@
             \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline{} \tab{}\texttt{/\^/~~~~WARN~Logging~HELO}
+                containing:\newline{}
+                \tab{}\texttt{/\^/~~~~WARN~Logging~HELO}
 

### 2008-11-22 doc/logparser-introduction.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

@@
 messages can be added by the administrator with custom checks; every
-\DNSBL{}\footnote{This document is supplied with a glossary, see
+\gls{DNSBL}\footnote{This document is supplied with a glossary, see
 \textsection\ref{Glossary}.} returns a different explanatory message;
@@
 paper documents the difficult process of parsing Postfix log files,
-presenting \PLP{}, a program that parses Postfix log files and places the
-resulting data into a database for later use.  The gathered data can then
-be used to optimise current anti-spam measures, provide a baseline to test
-new anti-spam measures against, or to produce statistics showing how
+presenting \gls{PLP}, a program that parses Postfix log files and places
+the resulting data into a database for later use.  The gathered data can
+then be used to optimise current anti-spam measures, provide a baseline to
+test new anti-spam measures against, or to produce statistics showing how
 effective those measures are.  Numerous other uses are possible for such
@@
 understanding the thesis, parser, and architecture.  It introduces the idea
-of using a database schema as an \API{}, providing an interface to the data
-gathered that is language-neutral.  The novel separation of rules, actions
-and framework is discussed, giving the reasons that approach was taken when
-designing the parser.
+of using a database schema as an \gls{API}, providing an interface to the
+data gathered that is language-neutral.  The novel separation of rules,
+actions and framework is discussed, giving the reasons that approach was
+taken when designing the parser.
 
 Section~\ref{state of the art review} reviews both the previously published
-research in this area and other available Postfix log parsers, discussing
-why they were deemed unsuitable for the task, including why they could not
-be improved or expanded upon.
+research in this area and other available Postfix log file parsers,
+discussing why they were deemed unsuitable for the task, including why they
+could not be improved or expanded upon.
 
@@
 listed are some additional resources expected to be helpful in
-understanding \SMTP{}, Postfix, anti-spam techniques, or the thesis.
+understanding \gls{SMTP}, Postfix, anti-spam techniques, or the thesis.
 
-Appendix~\ref{Glossary} provides a glossary of terms used in the thesis.
+Appendix~XXX provides a glossary of terms used in the thesis.
 
-Appendix~\ref{Acronyms} provides a list of acronyms used in the thesis.
+Appendix~XXX provides a list of acronyms used in the thesis.

### 2008-11-22 doc/logparser-other-parsers-review.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

@@
 THIS ACCURATE\@?) are intended to perform a specific parsing and reporting
-task, whereas \PLP{} is designed to be a generic parser, extracting data
+task, whereas \gls{PLP} is designed to be a generic parser, extracting data
 and leaving generation of reports from the data to other programs.  Some
@@
 developed later.  The parser developed for this project is designed to
-enable much more detailed log analysis by providing a stable platform for
-subsequent programs to develop upon.
+enable much more detailed log file analysis by providing a stable platform
+for subsequent programs to develop upon.
 
@@
 judge what improvements, if any, are delivered by the newcomers.  There are
-not that many previously developed Postfix log parsers --- indeed it was
-quite difficult to find ten parsers to review for this project --- and the
-functionality offered ranges from quite basic to much more mature,
+not that many previously developed Postfix log file parsers --- indeed it
+was quite difficult to find ten parsers to review for this project --- and
+the functionality offered ranges from quite basic to much more mature,
 depending on the needs of the creator.  The programs are not reviewed from
@@
         log files and extracting data; further processing of that data is a
-        separate concern, deferred to another program.  Using an \SQL{}
+        separate concern, deferred to another program.  Using an \gls{SQL}
         database simplifies the process of generating such reports
@@
 
-XXX DESCRIBE THE \CSV{} FORMAT\@.
+XXX DESCRIBE THE \gls{CSV} FORMAT\@.
 
@@
 log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \LMA{} is quite
-different from this parser: it attempts to present correlated log data in a
-form suitable for a systems administrator to search using the myriad of
-standard Unix text processing utilities already available, producing both a
-\CSV{} file and either a MySQL (\url{http://www.mysql.com/}) or Berkeley DB
+Utilizations\/}~\cite{log-mail-analyser}.  The aim of \gls{LMA} is quite
+different from this parser: it attempts to present correlated data from log
+files in a form suitable for a systems administrator to search using the
+myriad of standard Unix text processing utilities already available,
+producing both a \gls{CSV} file and either a MySQL
+(\url{http://www.mysql.com/}) or Berkeley DB
 (\url{http://www.oracle.com/database/berkeley-db/index.html}) database.
-The decision to support both \CSV{} and Berkeley DB appears to have been a
-serious limitation: \CSV{} is a very simple format where a record is stored
-in a single line, with fields separated by a comma or other punctuation
-symbol.  Problems with \CSV{} files include the need to escape separators
-in the data stored, providing multiple values for a field (e.g.\ multiple
-recipients), and adding new fields.  There is no mechanism to document the
-fields or the separator, unlike \SQL{} databases where every database
-includes a schema naming the fields and the type of data they store
-(integer, text, timestamp, etc.).
+The decision to support both \gls{CSV} and Berkeley DB appears to have been
+a serious limitation: \gls{CSV} is a very simple format where a record is
+stored in a single line, with fields separated by a comma or other
+punctuation symbol.  Problems with \gls{CSV} files include the need to
+escape separators in the data stored, providing multiple values for a field
+(e.g.\ multiple recipients), and adding new fields.  There is no mechanism
+to document the fields or the separator, unlike \gls{SQL} databases where
+every database includes a schema naming the fields and the type of data
+they store (integer, text, timestamp, etc.).
 
 Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
-unlike \SQL{} databases that store arbitrary tuples.  In \LMA{}'s main
-table the key is an integer referred to by secondary tables, and the value
-is a \CSV{} line containing all of the data for that row.  The
-secondary by-sender, by-recipient, by-date, and by-\IP{} tables use the
-sender/recipient/date/\IP{} as the key, and the value is a \CSV{} list of integers
-referring to the main table.  This effectively re-implements
-\SQL{} foreign keys, but without the functionality offered by even the most
-basic of \SQL{} databases (joins, ordering, searches, etc.).  It also
-requires custom code to search on some combination of the above, though the
-authors of \LMA{} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the least useful
-of the three output formats: it does not provide the functionality of a
-basic \SQL{} database, not can it be used with standard Unix text
-processing tools.
+unlike \gls{SQL} databases that store arbitrary tuples.  In \gls{LMA}'s
+main table the key is an integer referred to by secondary tables, and the
+value is a \gls{CSV} line containing all of the data for that row.  The
+secondary by-sender, by-recipient, by-date, and by-\gls{IP} tables use the
+sender/recipient/date/\gls{IP} as the key, and the value is a \gls{CSV}
+list of integers referring to the main table.  This effectively
+re-implements \gls{SQL} foreign keys, but without the functionality offered
+by even the most basic of \gls{SQL} databases (joins, ordering, searches,
+etc.).  It also requires custom code to search on some combination of the
+above, though the authors of \gls{LMA} did provide some queries: IP-STORY,
+FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the
+least useful of the three output formats: it does not provide the
+functionality of a basic \gls{SQL} database, not can it be used with
+standard Unix text processing tools.
 
-\LMA{} stores the following date: time and date, client hostname and \IP{}
-address, server \IP{} address, sender and recipient addresses, \SMTP{}
-code, and size (for accepted mails only).  It omits XXX XXX\@.  Handling of
-multiple recipients, \SMTP{} codes or remote servers\footnote{A single mail
-may be sent to multiple remote servers if it was addressed to recipients in
-different domains, or Postfix needs to try multiple servers for one or more
-recipients.} is not explained in the paper; experimental observation shows
-that multiple records are added when there are multiple recipients (sadly
-the records are not associated or linked in any way), and presumably the
-same approach is taken when there are multiple destination servers.
+\gls{LMA} stores the following date: time and date, client hostname and
+\gls{IP} address, server \gls{IP} address, sender and recipient addresses,
+\gls{SMTP} code, and size (for accepted mails only).  It omits XXX XXX\@.
+Handling of multiple recipients, \gls{SMTP} codes or remote
+servers\footnote{A single mail may be sent to multiple remote servers if it
+was addressed to recipients in different domains, or Postfix needs to try
+multiple servers for one or more recipients.} is not explained in the
+paper; experimental observation shows that multiple records are added when
+there are multiple recipients (sadly the records are not associated or
+linked in any way), and presumably the same approach is taken when there
+are multiple destination servers.
 
 The schema used with the MySQL database is undocumented, but at least it is
-possible to discover the schema with an \SQL{} database, unlike with
-Berkeley DB\@; all \SQL{} databases embed the schema into the database and
-provide commands for displaying it.  Berkeley DB does not embed a schema,
-as there is neither requirement nor benefit; it only provides \textbf{(key,
-value)} pairs, so any additional structuring of the data is imposed by the
-application, thus it behoves the application to document this structure.
+possible to discover the schema with an \gls{SQL} database, unlike with
+Berkeley DB\@; all \gls{SQL} databases embed the schema into the database
+and provide commands for displaying it.  Berkeley DB does not embed a
+schema, as there is neither requirement nor benefit; it only provides
+\textbf{(key, value)} pairs, so any additional structuring of the data is
+imposed by the application, thus it behoves the application to document
+this structure.
 
-\LMA{} requires major changes to the parser code to parse new log lines or
-extract additional data.  The code is structured as a long series of blocks
-that each handle all lines matching a single regex, so parsing new log
-lines requires modifying an existing regex or carefully inserting a new
+\gls{LMA} requires major changes to the parser code to parse new log lines
+or extract additional data.  The code is structured as a long series of
+blocks that each handle all lines matching a single regex, so parsing new
+log lines requires modifying an existing regex or carefully inserting a new
 block in the correct place; extracting extra data will require modifying
@@
 
-\LMA{} does not appear to deal with any of the complications discussed in
-\sectionref{complications}.  It does not differentiate between different
+\gls{LMA} does not appear to deal with any of the complications discussed
+in \sectionref{complications}.  It does not differentiate between different
 types of rejections, so it is not suitable for the purposes of this
@@
 discarded, whereas the main goal of this project is to retain that data to
-aid optimisation and evaluation of Postfix restrictions.  \LMA{} fails to
-parse Postfix log files generated on Solaris hosts because the fields
+aid optimisation and evaluation of Postfix restrictions.  \gls{LMA} fails
+to parse Postfix log files generated on Solaris hosts because the fields
 automatically prepended to each log line differ from those added on Linux
 hosts; log files from Solaris hosts (and possibly other operating systems)
-thus require preprocessing before parsing by \LMA{}.  Once the
+thus require preprocessing before parsing by \gls{LMA}.  Once the
 preprocessing has been performed on the \numberOFlogFILES{} test log files
-\LMA{} parses the log files without complaint, although it produced 32
+\gls{LMA} parses the log files without complaint, although it produced 32
 entries in its output file for every rejection in the input log file; it
@@
 
-\LMA{} does provide some simple reports: IP-STORY, FROM-STORY, DAILY-EMAIL
-and DAILY-REJECT\@.  The first three extract \CSV{} lines for the specified
-client \IP{} address, sender address, or date respectively.  DAILY-REJECT
-initially failed with an error message (XXX ADD THE ERROR MESSAGE) from the
-Perl interpreter; after making some corrections to the code it worked, and
-extracted the \CSV{} lines for the specified day where the \SMTP{} code
-signifies a rejection.  All of these reports are trivially simple to
-produce from the \CSV{} file using the standard Unix tool \texttt{awk}; the
-most complicated, DAILY-REJECT, is merely:
+\gls{LMA} does provide some simple reports: IP-STORY, FROM-STORY,
+DAILY-EMAIL and DAILY-REJECT\@.  The first three extract \gls{CSV} lines
+for the specified client \gls{IP} address, sender address, or date
+respectively.  DAILY-REJECT initially failed with an error message (XXX ADD
+THE ERROR MESSAGE) from the Perl interpreter; after making some corrections
+to the code it worked, and extracted the \gls{CSV} lines for the specified
+day where the \gls{SMTP} code signifies a rejection.  All of these reports
+are trivially simple to produce from the \gls{CSV} file using the standard
+Unix tool \texttt{awk}\glsadd{awk}; the most complicated, DAILY-REJECT, is
+merely:
 
@@
 
-    \item Some records have an \SMTP{} code of \textit{0}, though this is
-        not a valid \SMTP{} code; presumably this is because \LMA{} did not
-        parse a log line properly.  DAILY-REJECT considers this code to be
-        a rejection, so the command above must too.
+    \item Some records have an \gls{SMTP} code of \textit{0}, though this
+        is not a valid \gls{SMTP} code; presumably this is because
+        \gls{LMA} did not parse a log line properly.  DAILY-REJECT
+        considers this code to be a rejection, so the command above must
+        too.
 
-    \item In a small fraction of records the \SMTP{} code is
-        \textit{deferred\/} (again, not a valid \SMTP{} code, and most
-        likely because of mis-parsing by \LMA{}); it is treated as a
+    \item In a small fraction of records the \gls{SMTP} code is
+        \textit{deferred\/} (again, not a valid \gls{SMTP} code, and most
+        likely because of mis-parsing by \gls{LMA}); it is treated as a
         rejection by DAILY-REJECT, so the equivalent \texttt{awk} command
@@
 
-    \item \LMA{} does not escape or replace the separators used in the
-        \CSV{} records when they are present in sender or recipient
+    \item \gls{LMA} does not escape or replace the separators used in the
+        \gls{CSV} records when they are present in sender or recipient
         addresses, leading to some mangled records; these records are not
@@
 
-    \item Where there are multiple records in the \CSV{} file, DAILY-REJECT
-        extracts only one record, whereas the \texttt{awk} command above
-        extracts all matching records.
+    \item Where there are multiple records in the \gls{CSV} file,
+        DAILY-REJECT extracts only one record, whereas the \texttt{awk}
+        command above extracts all matching records.
 
@@
 
-In summary, \LMA{} appears to be a proof of concept, written to demonstrate
-the point of their paper (that having this information in an accessible
-fashion is useful to systems administrators), rather than a program
-designed to be useful in a production environment; no consideration was
-given to extensibility when writing it.
+In summary, \gls{LMA} appears to be a proof of concept, written to
+demonstrate the point of their paper (that having this information in an
+accessible fashion is useful to systems administrators), rather than a
+program designed to be useful in a production environment; no consideration
+was given to extensibility when writing it.
 
@@
 % Hard to extend/expand/understand.  Appears to only save: date and hour,
-% DNS name and \IP{} address host, mail server \IP{} address, sender,
+% DNS name and \gls{IP} address host, mail server \gls{IP} address, sender,
 % receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \CSV{} as an intermediate format between the log
-% file and the database seems to have been restrictive.  Appears to require
-% a queueid but majority of log entries (e.g.\ rejections) lack a queueid.
-% Supports whitelisting \IP{} addresses when parsing logs, but whitelisting
-% when generating reports/data mining would be preferable.  Supporting
-% Berkeley DB is probably limiting the software - an example is the
-% difficulty in searching a pipe-delimited string, so they have
+% Design decision to use \gls{CSV} as an intermediate format between the
+% log file and the database seems to have been restrictive.  Appears to
+% require a queueid but majority of log lines (e.g.\ rejections) lack a
+% queueid.  Supports whitelisting \gls{IP} addresses when parsing logs, but
+% whitelisting when generating reports/data mining would be preferable.
+% Supporting Berkeley DB is probably limiting the software - an example is
+% the difficulty in searching a pipe-delimited string, so they have
 % re-implemented foreign keys with tables keyed by ip address etc.\
@@
 easily extensible design of \parsername{}.  It does not correlate log lines
-by queueid or \pid{}, and does not need to deal with the complications
+by queueid or \gls{pid}, and does not need to deal with the complications
 encountered during this project.  Pflogsumm produces a useful report, and
@@
 
-\noindent{}\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline{}
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline{}
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline{}
-(Last checked 2007/12/04.)
+\noindent{}\url{http://www.thesawmill.co.uk/formats/postfix.html}
+\newline{} \url{http://www.thesawmill.co.uk/formats/postfix_ii.html}
+\newline{} \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
+\newline{} (Last checked 2007/12/04.)
 
@@
 purposes of this project,\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \PLP{} is
+described later in this review, and the data extracted by \gls{PLP} is
 described in \sectionref{connections table} and \sectionref{results
@@
 
-XXX THE BLURB MENTIONS \SQL{} DB\@; CAN THAT BE ACCESSED SEPARATELY, IS THE
-FORMAT DOCUMENTED, ETC\@?
+XXX THE BLURB MENTIONS \gls{SQL} DB\@; CAN THAT BE ACCESSED SEPARATELY, IS
+THE FORMAT DOCUMENTED, ETC\@?
 
@@
 extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \IP{} and virus.  The fields \texttt{server},
+origin hostname, origin \gls{IP} and virus.  The fields \texttt{server},
 \texttt{uid} and \texttt{virus} are not explained in the documentation:
@@
 \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline{} Fields
-extracted: from, to, RBL, client hostname and client \IP{}\@.
+extracted: from, to, RBL, client hostname and client \gls{IP}\@.
 
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline{} Fields
-extracted: from, to, client hostname, client \IP{}, relay hostname, relay
-\IP{}, status, response code, RBL and message id.
+\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline{}
+Fields extracted: from, to, client hostname, client \gls{IP}, relay
+hostname, relay \gls{IP}, status, response code, RBL and message id.
 
@@
 data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by a \DNSBL{} check.
+rejections except when the rejection is caused by a \gls{DNSBL} check.
 
@@
 and reading all the available documentation the author admitted defeat.
-Using the \CLI{} was more successful: invoking the command \newline{}
+Using the \gls{CLI} was more successful: invoking the command \newline{}
 \tab{} \texttt{splunk find logs }\textit{log-directory\/}\newline{} added
@@
 with Splunk.  Subsequently the author was successful in adding a single
-file at a time using the \CLI{}; a simple loop to add all desired log files
-was sufficient to index all files.  Splunk will periodically check all
-those files for updates unless they are manually removed from its list;
+file at a time using the \gls{CLI}; a simple loop to add all desired log
+files was sufficient to index all files.  Splunk will periodically check
+all those files for updates unless they are manually removed from its list;
 this may or may not be useful behaviour.  Splunk did not appear to have any
@@
 from addresses, HELO hostname, time and date, the host the log files were
-generated on, and protocol (\SMTP{} or \ESMTP{}).  It parses the standard
-syslog fields at the beginning of the log line, and extracts any
-\texttt{key=<value>} pairs occurring after the standard syslog prologue;
-these pairs are the fields listed above.  Searches can be based on the
-extracted fields or the full text of the log line.  \parsername{} extracts
-noticeably more data, though it does not make the full text of the line
-available, and the full power of \SQL{} is available when searching,
-allowing the user to search on arbitrarily complicated conditions.
+generated on, and protocol (\gls{SMTP} or \gls{ESMTP}).  It parses the
+standard syslog\glsadd{syslog} fields at the beginning of the log line, and
+extracts any \texttt{key=<value>} pairs occurring after the standard syslog
+prologue; these pairs are the fields listed above.  Searches can be based
+on the extracted fields or the full text of the log line.  \parsername{}
+extracts noticeably more data, though it does not make the full text of the
+line available, and the full power of \gls{SQL} is available when
+searching, allowing the user to search on arbitrarily complicated
+conditions.
 
@@
 importantly it makes no attempt to correlate log lines by queueid or
-\pid{}, or to handle any of the myriad complications discussed in
+\gls{pid}, or to handle any of the myriad complications discussed in
 \sectionref{complications}.  Some additional Postfix reports are supposedly
@@
 supporting many different services without special purpose code limits its
-functionality.  The data it will extract from an \MTA{} log file is limited
-in comparison to \parsername{}: \newline{} \tab{} time2, email, email\_r,
-host, host\_r, method, url, code, and bytesd.\newline{} XXX WHAT FIELDS ARE
-MISSING\@?  There does not seem to be an explanation of any of those fields
-in the documentation (\parsername{} provides copious documentation).
-AWStats coerces Postfix log files into Apache\footnote{The Apache web
-server is the most popular HTTP server in use over the past 10 years; more
-information is available at \url{http://httpd.apache.org/}.} format log
-files, for analysis by AWStats' HTTP log file parser.  The converting
-parser only deals with a small portion of the log lines generated by
-Postfix, silently skipping those it cannot deal with, and does not
-distinguish between different types of rejection; it would be a lot of work
-to extend it to handle new log lines.  Although it does correlate log lines
-by queueid (not by pid), it does not deal with any of the other
+functionality.  The data it will extract from an \gls{MTA} log file is
+limited in comparison to \parsername{}: \newline{} \tab{} time2, email,
+email\_r, host, host\_r, method, url, code, and bytesd.\newline{} XXX WHAT
+FIELDS ARE MISSING\@?  There does not seem to be an explanation of any of
+those fields in the documentation (\parsername{} provides copious
+documentation).  AWStats coerces Postfix log files into Apache\footnote{The
+Apache web server is the most popular HTTP server in use over the past 10
+years; more information is available at \url{http://httpd.apache.org/}.}
+format log files, for analysis by AWStats' HTTP log file parser.  The
+converting parser only deals with a small portion of the log lines
+generated by Postfix, silently skipping those it cannot deal with, and does
+not distinguish between different types of rejection; it would be a lot of
+work to extend it to handle new log lines.  Although it does correlate log
+lines by queueid (not by pid), it does not deal with any of the other
 complications described in \sectionref{complications}.  AWStats supports
@@
 
-    \item A summary of the \SMTP{} codes used when rejecting delivery
+    \item A summary of the \gls{SMTP} codes used when rejecting delivery
         attempts.
@@
 
-YAALA provides a mini-language based on \SQL{} that is used when generating
-reports; sample reports can be seen
+YAALA provides a mini-language based on \gls{SQL} that is used when
+generating reports; sample reports can be seen
 at~\url{http://www.yaala.org/samples.html}.  Example query for HTTP proxy
@@
 
-Lire is a general purpose log parser supporting many different types of log
-file.  The data extracted by its Postfix parser is shown in the quote
-above; notably rejections are not mentioned or dealt with, and unlike
+Lire is a general purpose log file parser supporting many different types
+of log file.  The data extracted by its Postfix parser is shown in the
+quote above; notably rejections are not mentioned or dealt with, and unlike
 \parsername{} there is no facility to extend the parser.  It supports
@@
 possible, with enough time and effort, to understand the format.
-\parsername{} uses an \SQL{} database to make accessing the extracted data
-as easy as possible.
+\parsername{} uses an \gls{SQL} database to make accessing the extracted
+data as easy as possible.
 
 Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \pid{}, so the complete list of recipients for a mail should be
-available; however its parser extracts only part of the available data and
-makes no attempt to deal with the other complications described in
+but not by \gls{pid}, so the complete list of recipients for a mail should
+be available; however its parser extracts only part of the available data
+and makes no attempt to deal with the other complications described in
 \sectionref{complications}.  When testing Lire on the \numberOFlogFILES{}
@@
 
-    \item For mail sent: to addresses, \SMTP{} code, and delay.
+    \item For mail sent: to addresses, \gls{SMTP} code, and delay.
 
@@
 correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \pid{}.  The parsing is error prone: empty fields are
-saved when the log line does not match the regex, though it appears that
-they will not overwrite existing data.  Most notably rejections are
+Logrep), but not by \gls{pid}.  The parsing is error prone: empty fields
+are saved when the log line does not match the regex, though it appears
+that they will not overwrite existing data.  Most notably rejections are
 completely ignored, making it unsuitable for the purposes of this project.
@@
 documentation).  It requires a web browser to interact with it, so
-automated log processing will be difficult, whereas automated processing is
-a key part of \parsernames{} design.  Sadly all the author's attempts to
-use Logrep failed, as it was unable to access the log files selected; this
-appears to be a bug rather than operator error.  If it was caused by
-operator error, the interface needs improvement as the (minimal)
+automated log file processing will be difficult, whereas enabling automated
+processing is a key part of \parsernames{} design.  Sadly all the author's
+attempts to use Logrep failed, as it was unable to access the log files
+selected; this appears to be a bug rather than operator error.  If it was
+caused by operator error, the interface needs improvement as the (minimal)
 instructions were followed as closely as possible, and multiple attempts
@@
 
-There are other programs available which perform basic Postfix log parsing
-(some to a greater level of detail than others), but few attempt to
-correlate log lines by queueid (none correlate by \pid{}) to produce an
+There are other programs available which perform basic Postfix log file
+parsing (some to a greater level of detail than others), but few attempt to
+correlate log lines by queueid (none correlate by \gls{pid}) to produce an
 overall record of the journey of each mail through Postfix.  None of the
@@
         \tableline{}%
-        \LMA{}          & No            & Yes               & No                & Poor           & Yes    \\ 
+        \gls{LMA}          & No            & Yes               & No                & Poor           & Yes    \\ 
         Pflogsumm       & Yes           & No                & Partial \dag{}    & Good           & Yes    \\

### 2008-11-22 doc/logparser-postfix-daemons.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

% vim: set filetype=tex :
% Postfix components
\newglossaryentry{bounce}{name={bounce},type={postfix},description={
    The bounce daemon is responsible for sending bounce notifications in
    Postfix versions later than 2.2.  The definitive documentation is
    \url{http://www.postfix.org/bounce.8.html}.
}}

\newglossaryentry{cleanup}{name={cleanup},type={postfix},description={
    Cleanup processes all incoming mail after it has been accepted and
    before it is delivered.  It removes duplicate recipient addresses,
    inserts missing headers, and optionally rewrites addresses if
    configured to do so.  The definitive documentation is
    \url{http://www.postfix.org/cleanup.8.html}.
}}

\newglossaryentry{lmtp}{name={lmtp},type={postfix},description={
    Delivery of mail over \gls{LMTP} is performed by the lmtp component.
    The definitive documentation is
    \url{http://www.postfix.org/lmtp.8.html}.
}}

\newglossaryentry{local}{name={local},type={postfix},description={
    Local is the Postfix component responsible for local delivery of mail
    (i.e.\ delivered on the server Postfix is running on), whether it be to
    a user's mailbox or a program such as a mailing list manager or
    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
    processing of a user's \texttt{.forward} file.  The definitive
    documentation is \url{http://www.postfix.org/local.8.html}.
}}

\newglossaryentry{pickup}{name={pickup},type={postfix},description={
    Pickup is the service that deals with mail submitted locally via
    postdrop and sendmail; it passes all submitted mail to cleanup.  The
    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
}}

\newglossaryentry{postdrop}{name={postdrop},type={postfix},description={
    Postdrop is used when submitting mail locally on the server: it creates
    a new mail in the queue and copies its input into the mail.  Subsequent
    delivery of the mail is the responsibility of other Postfix components.
    The definitive documentation is
    \url{http://www.postfix.org/postdrop.1.html}.
}}

\newglossaryentry{postsuper}{name={postsuper},type={postfix},description={
    Maintenance tasks such as deleting mails from the queue, putting mail
    on hold and later releasing it (no further delivery attempts will be
    made until it is released), and consistency checking of the mail queue.
    The definitive documentation is available at
    \url{http://www.postfix.org/postsuper.1.html}.
}}

\newglossaryentry{qmgr}{name={qmgr},type={postfix},description={
    Qmgr is the Postfix daemon that manages the mail queue, determining
    which mails will be delivered next.  Qmgr orders the mails based on the
    recipient for local mails and the destination server for remote
    addresses, ensuring that it balances the aims of achieving maximum
    concurrency while avoiding overwhelming destinations or wasting time
    and resources on non-responsive destinations.  The definitive
    documentation is \url{http://www.postfix.org/qmgr.8.html}.
}}

\newglossaryentry{sendmail}{name={sendmail},type={postfix},description={
    Postfix provides a command that is compatible with the Sendmail
    (\url{http://www.sendmail.org/}) mail submission program that all Unix
    commands that send mail depend on; it executes \daemon{postdrop} to
    place a new mail in the queue.  The definitive documentation is
    \url{http://www.postfix.org/sendmail.1.html}.
}}

\newglossaryentry{smtp}{name={smtp},type={postfix},description={
    Delivery of mail over \gls{SMTP} is performed by the smtp component.
    The definitive documentation is
    \url{http://www.postfix.org/smtp.8.html}.
}}

\newglossaryentry{smtpd}{name={smtpd},type={postfix},description={
    Smtpd is the Postfix program that accepts mail via \gls{SMTP}, and
    implements all the anti-spam restrictions Postfix provides.  The
    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
}}

\newglossaryentry{virtual}{name={virtual},type={postfix},description={
    Virtual is the Postfix component responsible for delivery of mails to
    virtual domains.  With \daemon{local} delivery the destination is
    determined only by the portion of the email address on the left side of
    the \at{}, whereas with \daemon{virtual} delivery the destination is
    determined by the entire email address, e.g.\ if the server considers
    itself responsible for both \textbf{example.org} and
    \textbf{example.net} domains: \daemon{local} considers
    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
    the same mailbox, whereas \daemon{virtual} considers them to have
    different mailboxes.  Virtual delivery is used when a server hosts
    multiple domains where a username may be present in more than one
    domain but represent different users in each.  The definitive
    documentation is
    \url{http://www.postfix.org/virtual.8.html}.
}}

### 2008-11-22 doc/logparser-results.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

@@
 Parsing efficiency is an obvious concern when the parser routinely needs to
-parse large log files.  The server that generated the log files used
-in testing this parser accepts approximately 10,000 mails for 700 users a
-day; \graphref{Mails received per day} shows that as expected there are far
-more mails received on weekdays than at weekends.   Median log file size is
+parse large log files.  The server that generated the log files used in
+testing this parser accepts approximately 10,000 mails for 700 users a day;
+\graphref{Mails received per day} shows that as expected there are far more
+mails received on weekdays than at weekends.   Median log file size is
 50MB, containing 285,000 log lines --- large scale mail servers would have
@@
 \tableref{Number of mails received per day: statistics} show the number of
-mails received by \SMTP{} only; in particular the mail loops noticeable in
-other graphs do not contribute to these figures.
+mails received by \gls{SMTP} only; in particular the mail loops noticeable
+in other graphs do not contribute to these figures.
 
-\showgraph{build/graph-mails-received}{Number of mails received via SMTP per
-day}{Mails received per day}
+\showgraph{build/graph-mails-received}{Number of mails received via SMTP
+per day}{Mails received per day}
 
@@
         the address it is currently delivering to, and therefore sends a
-        bounce notification with sender address \texttt{<>} to the original
-        sender: username@domain.
+        bounce notification with sender address \texttt{<>}\glsadd{<>} to
+        the original sender: username@domain.
 
@@
 not have to deliver the mail to an external system, so mails are delivered,
-bounced, and generated as fast as the disks can keep up, resulting in a huge
-volume of log lines.
+bounced, and generated as fast as the disks can keep up, resulting in a
+huge volume of log lines.
 
@@
 
-\showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time relative
-to shuffled ordering}{Parsing time relative to shuffled ordering graph}
+\showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
+relative to shuffled ordering}{Parsing time relative to shuffled ordering
+graph}
 
@@
 hard because other servers have different restrictions with custom
-messages, \DNSBL{} messages change over time, major releases of Postfix
+messages, \gls{DNSBL} messages change over time, major releases of Postfix
 change warning messages (usually adding more information), etc.,\ so over
@@
 is reused before the previous mail is fully dealt with, when a queueid or
-\pid{} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \pid{} in the first few hundred or
+\gls{pid} is not found in the state tables,\footnote{There will often be
+warnings about a missing queueid or \gls{pid} in the first few hundred or
 thousand log lines because the earlier log lines for those connections or
@@
 represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log entries are missing, leading to warnings
-when their log lines are parsed.
+log file, so their initial log lines are missing, leading to warnings when
+their remaining log lines are parsed.
 
@@
 
-    \item 9 mails whose initial log entries were not present in the log
+    \item 9 mails whose initial log lines were not present in the log
         files.
@@
 
-    \item Queueids and \pids{} naturally group together log lines belonging
-        to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
+    \item Queueids and \glspl{pid} naturally group together log lines
+        belonging to one mail or connection respectively; it is extremely
+        unlikely that a log line would be associated with the wrong
+        connection.
 
@@
 and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log and a dump of the resulting database,
+conjunction with the input log file and a dump of the resulting database,
 verifying that for each of the input lines the parser used the correct rule
 and executed the correct action, that in turn produced the correct result
-and inserted the correct data in the database.  The log segment produced 4
-warnings, 10 mails remaining in the state tables, and 1625 connections
-correctly entered in the database.
+and inserted the correct data in the database.  The log file segment
+produced 4 warnings, 10 mails remaining in the state tables, and 1625
+connections correctly entered in the database.
 

### 2008-11-22 doc/logparser.tex
Switched to new glossaries package. A couple of other XXX items have been done too.

@@
 \usepackage{cite}
-
-% Acronyms and glossary entries.
-% global=true prevents acronyms being expanded twice
-\usepackage[global=true,acronym=true,style=altlist,number=none,toc=true]{glossary}
-\makeglossary{}
-\makeacronym{}
 
@@
 }
+
+% Acronyms and glossary entries.
+\usepackage[acronym,toc,numberedsection]{glossaries}
+\newglossary[plg]{postfix}{pin}{pout}{Postfix Daemons}
+\makeglossaries{}
 
@@
 % The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\PLP{}}
-\newcommand{\parsernames}{\PLP{}'s}
+\newcommand{\parsername}{\gls{PLP}}
+\newcommand{\parsernames}{\gls{PLP}'s}
 
@@
 \begin{document}
-
-% Pull in the acronyms early, so they can be used throughout the text.
-\input{logparser-acronyms.tex}
-
-% WSUIPA fonts.
-\input{ipamacs}
 
@@
 
-XXX SWITCH TO NEW GLOSSARY PACKAGE\@.
-
-XXX IS IT POSSIBLE TO HAVE ACRONYMS THAT ARE NOT INCLUDED IN THE ACRONYM
-LIST\@?
-
 XXX REDUCE THE LIST OF ACRONYMS\@.
-
-XXX CHECK THAT I SAY log line AND log file EVERYWHERE\@.
 
@@
 
-XXX IMPROVE BIBLIOGRAPHY APPEARANCE\@: NAMES LOOK UGLY (E.G.\ P.\ \empty{}.\
-Porters), URL SPACING IS MESSED UP --- URLS \\SHOULD PROBABLY BE LAST, ON THEIR
-OWN LINE\@.
+XXX IMPROVE BIBLIOGRAPHY APPEARANCE\@: NAMES LOOK UGLY (E.G.\ P.\
+\empty{}.\ Porters), URL SPACING IS MESSED UP --- URLS \\SHOULD PROBABLY BE
+LAST, ON THEIR OWN LINE\@.
 
@@
 \newpage
-\input{logparser-introduction.tex}
-\input{logparser-background.tex}
-\input{logparser-other-parsers-review.tex}
-\input{logparser-architecture.tex}
-\input{logparser-implementation.tex}
-\input{logparser-results.tex}
-\input{logparser-conclusion.tex}
+
+% Pull in the acronyms early, so they can be used throughout the text.
+\input{logparser-acronyms}
+\input{logparser-glossary}
+\input{logparser-postfix-daemons}
+\glsaddall[types={postfix}]
+
+% WSUIPA fonts.
+\input{ipamacs}
+
+\input{logparser-introduction}
+\input{logparser-background}
+\input{logparser-other-parsers-review}
+\input{logparser-architecture}
+\input{logparser-implementation}
+\input{logparser-results}
+\input{logparser-conclusion}
 
@@
 % section.
-XXX MOVE POSTFIX DAEMONS INTO A SEPARATE SECTION\@; NEED TO SWITCH TO NEW
-GLOSSARY PACKAGE FIRST\@.
-\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
-\printglossary{}
+%\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
+\label{Glossary}
+\printglossaries{}
 % Redefine the command a second time to produce acronyms instead of a
 % glossary.
-\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-\printacronym{}
+%\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
+%\printacronym{}
 
-\input{logparser-glossary.tex}
-% logparser-acronyms.tex has been included in the preamble so that the
-% commands are available throughout the text.
 \end{document}

### 2008-11-23 doc/logparser-acronyms.tex
Minor cleanups.

@@
-% % vim: set textwidth=75 spell :
+% vim: set filetype=tex :
 % Note: every acronym is listed twice: once so it can be expanded, and once
@@
     description={
-        SQL is the standard language used for database querying,
-        modification and maintenance.  Some information about SQL,
-        including its history, can be found at
+        The standard language used for database querying, modification, and
+        maintenance.  Some information and history can be found at
         \url{http://en.wikipedia.org/wiki/SQL}; a good introduction can be
@@
         SLCT~\cite{slct-paper} is a tool implementing an algorithm designed
-        by Risto Vaarandi for identifying, grouping and producing a regex
+        by Risto Vaarandi for identifying, grouping, and producing a regex
         to match similar log lines.

### 2008-11-23 doc/logparser-architecture.tex
Minor cleanups.

@@
 
-\label{why separate rules, actions and framework?}
+\label{why separate rules, actions, and framework?}
 
@@
 new or casual users who need to parse new log lines), and greatly
-simplifies framework, actions and rules.  Decoupling also creates a clear
+simplifies framework, actions, and rules.  Decoupling also creates a clear
 separation of functionality: rules handle low level details of identifying
@@
 
-    \item Compare the results of parsing using sorted, shuffled and
+    \item Compare the results of parsing using sorted, shuffled, and
         reversed rules (described in \sectionref{rule ordering for
         efficiency}).  Parse several log files using optimal ordering, then
-        dump a textual representation of the rules, connections and results
-        tables.  Repeat with shuffled and reversed ordering, starting with
-        a fresh database.  If there are no overlapping rules the tables
-        from each run will be identical; differences indicate overlapping
-        rules.  The rules that overlap can be determined by examining the
-        differences in the tables: each result contains a reference to the
-        rule which created it, if the references differ between runs the
-        two rules referenced in the differing results overlap.
-        Unfortunately this method cannot prove the absence of overlapping
-        rules; it can detect overlapping rules, but only if there are log
-        lines in the input files that match more than one rule.
+        dump a textual representation of the rules, connections, and
+        results tables.  Repeat with shuffled and reversed ordering,
+        starting with a fresh database.  If there are no overlapping rules
+        the tables from each run will be identical; differences indicate
+        overlapping rules.  The rules that overlap can be determined by
+        examining the differences in the tables: each result contains a
+        reference to the rule which created it, if the references differ
+        between runs the two rules referenced in the differing results
+        overlap.  Unfortunately this method cannot prove the absence of
+        overlapping rules; it can detect overlapping rules, but only if
+        there are log lines in the input files that match more than one
+        rule.
 

### 2008-11-23 doc/logparser-background.tex
Minor cleanups.

@@
 (evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO and MAIL FROM commands will not be
+lists for client connection, HELO, and MAIL FROM commands will not be
 evaluated until the first RCPT TO command is received, because some clients

### 2008-11-23 doc/logparser-implementation.tex
Minor cleanups.

@@
 Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions and framework have been separated in the parser's design.
+rules, actions, and framework have been separated in the parser's design.
 In this section algorithm can be taken to mean the combination of framework
@@
 storing the data is a write-once operation, whereas utilising the data
-requires frequent searching, sorting and manipulation of the data to
+requires frequent searching, sorting, and manipulation of the data to
 produce customised reports or statistics.
@@
         this way; additionally \texttt{child} and \texttt{pid} are accepted
-        and used respectively by the \texttt{TRACK}, \texttt{BOUNCE} and
+        and used respectively by the \texttt{TRACK}, \texttt{BOUNCE}, and
         \texttt{SMTPD\_DIED} actions, but do not cause data to be entered
@@
 Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered and entirely finished with
+child mail will be created, delivered, and entirely finished with
 \textbf{before} the parent logs the creation log line!  Thus, mails flagged
@@
     \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Mail submitted
-        locally \newline{} on the server, both local and remote delivery.
+        locally \newline{} on the server, for both local and remote
+        delivery.
 
     \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, both local and remote delivery.
+        from a remote client, for both local and remote delivery.
 
     \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, local delivery only.
+        client, for local delivery only.
 
     \item [\daemon{pickup}, \daemon{smtp}:] Mail submitted locally on the
-        server, remote delivery only.
+        server, for remote delivery only.
 
@@
     \item The hostname used in the HELO command is not logged if the mail
-        is accepted.\footnote{Tested with Postfix 2.2.10, 2.3.11 and 2.4.7;
+        is accepted.\footnote{Tested with Postfix 2.2.10, 2.3.11, and 2.4.7;
         this may possibly have changed in Postfix 2.5. XXX TEST THIS
@@
 algorithm for analysing unparsed log lines to create regexes for new
-rules.  Detecting, diagnosing and defeating complications forms the largest
+rules.  Detecting, diagnosing, and defeating complications forms the largest
 single portion of this section, mirroring the development of the parser.

### 2008-11-23 doc/logparser-introduction.tex
Minor cleanups.

@@
 understanding the thesis, parser, and architecture.  It introduces the idea
-of using a database schema as an \gls{API}, providing an interface to the
-data gathered that is language-neutral.  The novel separation of rules,
-actions and framework is discussed, giving the reasons that approach was
-taken when designing the parser.
+of using a database schema as an \gls{API}, to provide a language-neutral
+interface to the data gathered.  The novel separation of rules, actions,
+and framework is discussed, giving the reasons that approach was taken when
+designing the parser.
 
@@
 Section~\ref{conclusion} contains the conclusion of the thesis, describing
-the results of the research, design and implementation of the parser.
+the results of the research, design, and implementation of the parser.
 

### 2008-11-23 doc/logparser-other-parsers-review.tex
Minor cleanups.

@@
         \parsername{}'s architecture is described in \sectionref{why
-        separate rules, actions and framework?}.
+        separate rules, actions, and framework?}.
 
@@
 above, though the authors of \gls{LMA} did provide some queries: IP-STORY,
-FROM-STORY, DAILY-EMAIL and DAILY-REJECT\@.  Berkeley DB appears to be the
+FROM-STORY, DAILY-EMAIL, and DAILY-REJECT\@.  Berkeley DB appears to be the
 least useful of the three output formats: it does not provide the
@@
 extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \gls{IP} and virus.  The fields \texttt{server},
+origin hostname, origin \gls{IP}, and virus.  The fields \texttt{server},
 \texttt{uid} and \texttt{virus} are not explained in the documentation:
@@
 \url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline{} Fields
-extracted: from, to, RBL, client hostname and client \gls{IP}\@.
+extracted: from, to, RBL, client hostname, and client \gls{IP}\@.
 
@@
 Fields extracted: from, to, client hostname, client \gls{IP}, relay
-hostname, relay \gls{IP}, status, response code, RBL and message id.
+hostname, relay \gls{IP}, status, response code, RBL, and message id.
 
@@
 modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both input and output modules would need to be
-written specially; it may even be more work to implement the parser within
-the constraints of YAALA\@.  YAALA supports storing previously gathered
-data using Perl's Storable module~\cite{perl-storable}, so other Perl
-programs can use Storable to load, examine, and optionally modify the data;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.  This information was gleaned from the
-source code, as the documentation is sadly lacking.
+starting from scratch, as both the input and output modules would need to
+be written specially; it may even be more work to implement the parser
+within the constraints of YAALA\@.  YAALA supports storing previously
+gathered data using Perl's Storable module~\cite{perl-storable}, so other
+Perl programs can use Storable to load, examine, and optionally modify the
+data; \parsername{} uses a well documented database which is accessible
+from the majority of programming languages.  This information was gleaned
+from the source code, as the documentation is sadly lacking.
 

### 2008-11-23 doc/logparser-postfix-daemons.tex
Minor cleanups.

@@
     the \at{}, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address, e.g.\ if the server considers
-    itself responsible for both \textbf{example.org} and
-    \textbf{example.net} domains: \daemon{local} considers
-    \textbf{john\at{}example.org} and \textbf{john\at{}example.net} to have
-    the same mailbox, whereas \daemon{virtual} considers them to have
-    different mailboxes.  Virtual delivery is used when a server hosts
-    multiple domains where a username may be present in more than one
-    domain but represent different users in each.  The definitive
-    documentation is
+    determined by the entire email address.  E.g.\ if the server is
+    responsible for both the \textbf{example.org} and \textbf{example.net}
+    domains: \daemon{local} considers \textbf{john\at{}example.org} and
+    \textbf{john\at{}example.net} to have the same mailbox, whereas
+    \daemon{virtual} considers them to have different mailboxes.  Virtual
+    delivery is used when a server hosts multiple domains where a username
+    may be present in more than one domain but represent different users in
+    each.  The definitive documentation is
     \url{http://www.postfix.org/virtual.8.html}.

### 2008-11-23 doc/logparser-results.tex
Minor cleanups.

@@
 ruleset: the percentage parsing time increase when using the full ruleset
-instead of the minimal ruleset for optimal, shuffled and reversed orderings
-is shown in \graphref{Percentage parsing time increase of maximum ruleset
-over minimum ruleset}.
+instead of the minimal ruleset for optimal, shuffled, and reversed
+orderings is shown in \graphref{Percentage parsing time increase of maximum
+ruleset over minimum ruleset}.
 
@@
     \item Every effort has been made while developing to make the parser as
-        precise, demanding and particular as possible.
+        precise, demanding, and particular as possible.
 
@@
 adding new rules or modifying existing rules is simplified by the
-separation of rules, actions and framework.  Improving the logical coverage
+separation of rules, actions, and framework.  Improving the logical coverage
 is harder, as the actions taken by Postfix must be reconstructed by the

### 2008-11-23 doc/logparser-background.tex
Started updating blurbs etc. in the start of the art review. Dealt with a few XXX items.

@@
 underlying the project, followed by some technical information: the use of
-a database as an \gls{API}\@; a brief introduction to \gls{SMTP}\@; a
-longer introduction to Postfix, concentrating on anti-spam restrictions and
-policy servers.  The assumptions made in designing and implementing the
-parser are explained, as are the conventions used in this thesis.
+a database as an \gls{API}\@; a brief introduction to \gls{SMTP}\@.  It
+finishes with a longer introduction to Postfix, concentrating on anti-spam
+restrictions and policy servers.
 
@@
 
-\subsection{Conventions used}
-
-XXX MAYBE REWRITE SO THAT THIS IS NOT TRUE\@?  ARE THERE ANY OTHER
-CONVENTIONS\@?
-
-The words \textit{connection\/} and \textit{mail\/} are often used
-interchangeably; in general the word used was chosen based
-on the context it appears in.
-
 \subsection{Summary}
@@
 explanation of the use of a database as an \gls{API} and an introduction to
-\gls{SMTP}.  The focus then switched to the \gls{MTA} in use, Postfix, and
-its restrictions and policy servers, finishing with the conventions used in
-this thesis.
+\gls{SMTP}.  The focus then switched to Postfix, the \gls{MTA} in use, and
+its restrictions and policy servers.

### 2008-11-23 doc/logparser-other-parsers-review.tex
Started updating blurbs etc. in the start of the art review. Dealt with a few XXX items.

@@
 \noindent{}\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline{}
-(Last checked 2008/04/09.)
+(Last checked 2008/11/23.)
 
@@
 
-    Sawmill is a Postfix log analyzer (it also support [sic] 686 other log
+    Sawmill is a Postfix log analyzer (it also support 818 other log
     formats). It can process log files in Postfix format, and generate
@@
 \newline{} \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
-\newline{} (Last checked 2007/12/04.)
+\newline{} (Last checked 2008/11/23.)
 
-XXX CHECK ALL THIS AGAIN\@; GIVE EXAMPLE PRICES\@.
+Sawmill is a general purpose commercial product that parses 818 log file
+formats (as of 2008/11/23) and produces reports from the extracted data.
+Its data extraction facilities (described later) are too limited to save
+sufficient data for the purposes of this project; it extracts three
+different sets of data from Postfix log files, but they are not interlinked
+in any way.  The documentation does not suggest that any attempt is made to
+correlate log lines or deal with the difficulties documented in
+\sectionref{complications}.
 
-Sawmill is a general purpose commercial product that parses 687 log file
-formats (correct as of 2007/12/04) and produces reports.  Its data
-extraction facilities are too limited to save sufficient data for the
-purposes of this project,\footnote{The data extracted by Sawmill is
-described later in this review, and the data extracted by \gls{PLP} is
-described in \sectionref{connections table} and \sectionref{results
-table}.} though it does extract three different sets of data for Postfix
-(one is beta as of 2007/12/04), but they do not appear to be interlinked.
-The documentation does not suggest that any attempt is made to correlate
-log lines or deal with the difficulties documented in
-\sectionref{complications}.  The source code is available in an obfuscated
-form only (presumably for a fee), and the product is quite expensive,
-requiring a license per report to be generated; in contrast \parsername{}
-is free and the code is freely available.  The web interface allows
-creation of dynamic reports based on any field, but because of the
-associated cost the author has not experimented with it.  Presumably, as
-the company charges per report the user wishes to generate, the data store
-(if there is one) is deliberately inaccessible and undocumented to prevent
-the user bypassing the program and generating their own reports.
+Sawmill has three different Postfix log file parsers:
 
-XXX THE BLURB MENTIONS \gls{SQL} DB\@; CAN THAT BE ACCESSED SEPARATELY, IS
-THE FORMAT DOCUMENTED, ETC\@?
+\begin{enumerate}
 
-\url{http://www.thesawmill.co.uk/formats/postfix.html} \newline{} Fields
-extracted: from, to, server, UID, relay, status, number of recipients,
-origin hostname, origin \gls{IP}, and virus.  The fields \texttt{server},
-\texttt{uid} and \texttt{virus} are not explained in the documentation:
-\texttt{server} is probably the server the mail is delivered to, and
-\texttt{uid} might be the uid of the user submitting mail locally.  Postfix
-does not perform any form of virus checking (though it has many options for
-cooperating with an external virus scanner), so the \texttt{virus} field is
-a mystery.
+    \item \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline{}
+        Fields extracted: from, to, server, UID, relay, status, number of
+        recipients, origin hostname, origin \gls{IP}, and virus.  It also
+        counts the number of and total size of all mails delivered.  The
+        fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
+        \texttt{virus} are not explained in the documentation:
+        \texttt{server} is probably the server the mail is delivered to,
+        and \texttt{uid} might be the uid of the user submitting mail
+        locally.  Postfix does not perform any form of virus checking
+        (though it has many options for cooperating with an external virus
+        scanner), so the \texttt{virus} field is a mystery.
 
-\url{http://www.thesawmill.co.uk/formats/postfix_ii.html} \newline{} Fields
-extracted: from, to, RBL, client hostname, and client \gls{IP}\@.
+    \item \url{http://www.thesawmill.co.uk/formats/postfix_ii.html}
+        \newline{} Fields extracted: from, to, RBL list, client hostname,
+        and client \gls{IP}\@.  It also counts the number of and total size
+        of all mails delivered.
 
-\url{http://www.thesawmill.co.uk/formats/beta_postfix.html} \newline{}
-Fields extracted: from, to, client hostname, client \gls{IP}, relay
-hostname, relay \gls{IP}, status, response code, RBL, and message id.
+    \item \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
+        \newline{} Fields extracted: from, to, client hostname, client
+        \gls{IP}, relay hostname, relay \gls{IP}, status, response code,
+        RBL list, and message id.  It also counts the number and size of
+        all mails delivered, processed, blocked, expired, and bounced.
+
+\end{enumerate}
 
@@
 data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by a \gls{DNSBL} check.
+rejections except when the rejection is caused by a \gls{DNSBL} check
+(\textit{RBL list\/} in the list of fields).
 
@@
 uncompressed beforehand and parsing took eight hours.
+
+XXX THE BLURB MENTIONS \gls{SQL} DB\@; CAN THAT BE ACCESSED SEPARATELY, IS
+THE FORMAT DOCUMENTED, ETC\@?  XXX EXPERIMENT WITH THE REPORTS IN THE DEMO
+VERSION\@.  XXX EVERYTHING FROM HERE DOWN NEEDS TO BE CHECKED AND/OR
+REWRITTEN\@.
+
+The web interface allows creation of dynamic reports based on any field,
+but because of the associated cost the author has not experimented with it.
+As the company charges per report the user wishes to generate, the data
+store (if there is one) is deliberately inaccessible and undocumented to
+prevent the user bypassing the program and generating their own reports.
+
+The source code is available only in an obfuscated form (presumably for a
+fee), and the product is quite expensive, requiring a \euros{100} + VAT
+license per report to be generated; in contrast \parsername{} is free and
+the code is freely available.
 
@@
 
-    Splunk is an IT Search engine. It is software that indexes any format
-    of IT data from any source in real time, including logs,
-    configurations, scripts, code, messages, traps, alerts, activity
-    reports, stack traces and metrics from all of your applications,
-    servers and devices. Splunk lets you search, navigate, alert and report
-    on all your IT data in real time using an AJAX web interface. You can
-    also share knowledge and Splunk solutions with other members of the
-    Splunk community via SplunkBase.
+    Splunk is IT Search.
+
+    Search and navigate IT data from applications, servers and network
+    devices in real-time. Logs, configurations, messages, traps and alerts,
+    scripts, code, metrics and more. If a machine can generate it ---
+    Splunk can eat it. It's easy to download and use and it's very
+    powerful.
 
@@
 \noindent{}\url{http://www.splunk.com/} \newline{}
-(Last checked 2008/04/29.)
+(Last checked 2008/11/23.)
 
@@
 find Postfix reports, or additional reports for any other log file types.
-
-\newpage{}
 

### 2008-11-30 doc/logparser-other-parsers-review.tex
Improved the introduction and the LMA section.

@@
+\newpage{} %XXX
+
 \section{State of the Art Review}
@@
 
-Ten other parsers, starting with the only previously published research in
-this area that the author is aware of, have been reviewed as part of the
-background research for this project.  Each of the parsers was tested with
-the \numberOFlogFILES{} test log files described in \sectionref{parser
-efficiency}.  None of the parsers reviewed perform the kind of advanced
-parsing and correlation required for this project; all except one (XXX IS
-THIS ACCURATE\@?) are intended to perform a specific parsing and reporting
-task, whereas \gls{PLP} is designed to be a generic parser, extracting data
-and leaving generation of reports from the data to other programs.  Some
-parsers save the data extracted to a data store, but the majority discard
-all data once they have finished running, making historical analysis
-impossible.  The parsers reviewed all produce a report of varying
-complexity and detail, whereas this project's parser does not produce a
-report at all; that responsibility is deferred to a separate program, to be
-developed later.  The parser developed for this project is designed to
-enable much more detailed log file analysis by providing a stable platform
-for subsequent programs to develop upon.
-
 It is important to compare and contrast newly developed programs,
 algorithms, and parsers against those already available, to accurately
-judge what improvements, if any, are delivered by the newcomers.  There are
-not that many previously developed Postfix log file parsers --- indeed it
-was quite difficult to find ten parsers to review for this project --- and
-the functionality offered ranges from quite basic to much more mature,
-depending on the needs of the creator.  The programs are not reviewed from
-an independent viewpoint; the objective is to compare and contrast each
-program with \parsername{}.  There are some important differences between
-\parsername{} and the parsers reviewed here:
+judge what improvements, if any, are delivered by the newcomers.  Ten other
+parsers have been reviewed as part of the background research for this
+project, starting with the only previously published research in this area
+that the author is aware of.  There are not that many previously developed
+Postfix log file parsers --- indeed it was quite difficult to find ten
+parsers to review for this project --- and the functionality offered ranges
+from quite basic to much more mature, depending on the needs of the author
+of the parser.  The programs are not reviewed from an independent
+viewpoint; the objective is to compare and contrast each program with
+\parsername{}.  There are some important differences between \parsername{}
+and the parsers reviewed here:
 
 \begin{enumerate}
+
+    \item None of the parsers reviewed perform the kind of advanced parsing
+        and correlation required for this project (described in
+        \sectionref{complications})
 
@@
 
-    \item \parsername{} does not provide any reports based on the data it
-        gathers.  This is by design: the parser is responsible for parsing
-        log files and extracting data; further processing of that data is a
-        separate concern, deferred to another program.  Using an \gls{SQL}
-        database simplifies the process of generating such reports
-        (discussed in \sectionref{database as API}); some simple examples
-        are given in \sectionref{motivation}.
+    \item The parsers reviewed all produce a report of varying complexity
+        and detail, whereas \gls{PLP} is designed to be a generic parser,
+        extracting data and leaving generation of reports from the data to
+        other programs.  Using an \gls{SQL} database simplifies the process
+        of generating such reports (discussed in \sectionref{database as
+        API}); some simple examples are given in \sectionref{motivation}.
+
+        The parser developed for this project is designed to enable much
+        more detailed log file analysis by providing a stable platform for
+        subsequent programs to develop upon.
 
@@
         without user intervention; Sawmill and Lire can be configured to
-        support compressed files, but Sawmill exhibits a dramatic slow down
-        when thus configured.  Although this is a minor difference, support
-        for reading compressed log files is quite helpful, as it
-        dramatically reduces the disk space required to store old log
-        files.
+        support compressed files, but Sawmill exhibits a dramatic increase
+        in parsing time when doing so.  Although this is a minor
+        difference, support for reading compressed log files is quite
+        helpful, as it dramatically reduces the disk space required to
+        store old log files.
+
+    \item Some of the parsers reviewed save the data extracted to a data
+        store, but the majority discard all data once they have finished
+        generating the report, making historical analysis impossible.
 
@@
 
+Each of the parsers was tested with the \numberOFlogFILES{} test log files
+described in \sectionref{parser efficiency}.  The data extracted by
+\parsername{} is documented in \sectionref{connections table} and
+\sectionref{results table}.
+
+XXX LIST THE DATA NOT EXTRACTED BY EACH PARSER\@.
+
 \subsection{Log Mail Analyser}
-
-XXX IS THE BDB STRUCTURE DOCUMENTED\@?
-
-XXX DESCRIBE THE \gls{CSV} FORMAT\@.
-
-XXX DESCRIBE THE MYSQL SCHEMA\@.
-
-IS THE SAME DATA STORED FOR ALL FORMATS\@?
 
@@
 
+Very little documentation is provided with \gls{LMA}, though some
+documentation is available in~\cite{log-mail-analyser}.  Studying the
+source code is informative, though this author had difficulty as the
+authors of \gls{LMA} wrote in Italian.
+
 Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
@@
 
+The schema used with the MySQL database is undocumented, but at least it is
+possible to discover the schema with an \gls{SQL} database, unlike with
+Berkeley DB\@; all \gls{SQL} databases embed the schema into the database
+and provide commands for displaying it.  Berkeley DB does not embed a
+schema, as there is neither requirement nor benefit; it only provides
+\textbf{(key, value)} pairs, so any additional structuring of the data is
+imposed by the application, thus it behoves the application to document
+this structure.
+
+The \gls{CSV} record format is not documented, but the output file contains
+a comment giving the format:\newline{} \texttt{\# Timestamp|Nome Client|IP
+Client|IP Server|From|To|Status|Size} \newline{}\gls{LMA} treats lines
+starting with \texttt{\#} as comments, but not all \gls{CSV} parsers will,
+as there is no standard for comments in \gls{CSV} files.
+
 \gls{LMA} stores the following date: time and date, client hostname and
@@
 
-The schema used with the MySQL database is undocumented, but at least it is
-possible to discover the schema with an \gls{SQL} database, unlike with
-Berkeley DB\@; all \gls{SQL} databases embed the schema into the database
-and provide commands for displaying it.  Berkeley DB does not embed a
-schema, as there is neither requirement nor benefit; it only provides
-\textbf{(key, value)} pairs, so any additional structuring of the data is
-imposed by the application, thus it behoves the application to document
-this structure.
-
 \gls{LMA} requires major changes to the parser code to parse new log lines
@@
 
-\gls{LMA} does not appear to deal with any of the complications discussed
-in \sectionref{complications}.  It does not differentiate between different
+\gls{LMA} does not deal with any of the complications discussed in
+\sectionref{complications}.  It does not differentiate between different
 types of rejections, so it is not suitable for the purposes of this
@@
 for the specified client \gls{IP} address, sender address, or date
-respectively.  DAILY-REJECT initially failed with an error message (XXX ADD
-THE ERROR MESSAGE) from the Perl interpreter; after making some corrections
-to the code it worked, and extracted the \gls{CSV} lines for the specified
-day where the \gls{SMTP} code signifies a rejection.  All of these reports
-are trivially simple to produce from the \gls{CSV} file using the standard
-Unix tool \texttt{awk}\glsadd{awk}; the most complicated, DAILY-REJECT, is
-merely:
+respectively.  DAILY-REJECT initially failed with an error message from the
+Perl interpreter;\footnote{The error message was:
+\newline{}\texttt{Undefined subroutine \&main::LIST called at queryDB.pl
+line 372.}} after making some corrections to the code it worked, and
+extracted the \gls{CSV} lines for the specified day where the \gls{SMTP}
+code signifies a rejection.  All of these reports are trivially simple to
+produce from the \gls{CSV} file using the standard Unix tool
+\texttt{awk}\glsadd{awk}; the most complicated, DAILY-REJECT, is merely:
 
@@
 
+The source code is available in an obfuscated form, and the product is
+quite expensive, requiring a \euros{100} + VAT licence per report
+(discounts are available when buying multiple licences); in contrast
+\parsername{} is free and the code is freely available.  Sawmill is
+supplied with thorough and well written documentation; everything the
+author searched for was documented, except the MySQL database schema.  A
+commercial version of MySQL is required due to MySQL licensing
+restrictions, but Sawmill's documentation explains why and includes
+instructions on how to compile Sawmill so that it can use a non-commercial
+version of MySQL\@.
+
+Sawmill's web interface supports searching on any combination of the fields
+it extracts, but the interface is neither as simple to use nor as
+informative as the interface provided by Splunk.  The administrative
+interface is much easier to use than Splunk's: it took only five minutes to
+start parsing a directory of log files.  Searches performed using the web
+interface produced reasonable results, which were not verified in detail.
+
 When tested with the \numberOFlogFILES{} test log files it performed
@@
 uncompressed beforehand and parsing took eight hours.
-
-XXX THE BLURB MENTIONS \gls{SQL} DB\@; CAN THAT BE ACCESSED SEPARATELY, IS
-THE FORMAT DOCUMENTED, ETC\@?  XXX EXPERIMENT WITH THE REPORTS IN THE DEMO
-VERSION\@.  XXX EVERYTHING FROM HERE DOWN NEEDS TO BE CHECKED AND/OR
-REWRITTEN\@.
-
-The web interface allows creation of dynamic reports based on any field,
-but because of the associated cost the author has not experimented with it.
-As the company charges per report the user wishes to generate, the data
-store (if there is one) is deliberately inaccessible and undocumented to
-prevent the user bypassing the program and generating their own reports.
-
-The source code is available only in an obfuscated form (presumably for a
-fee), and the product is quite expensive, requiring a \euros{100} + VAT
-license per report to be generated; in contrast \parsername{} is free and
-the code is freely available.
-
-Sawmill's web interface supports searching on any combination of the fields
-it extracts, but the interface is neither as simple to use nor as
-informative as the interface provided by Splunk.  The administrative
-interface is easy to use --- it took only five minutes to start parsing a
-directory of log files.
 
@@
 day to 500MB, though a trial Enterprise licence is available that allows
-indexing of up to 5GB of data per day.  In 2007 the cheapest licensed
+indexing of up to 5GB of data per day.  In 2007 the cheapest licenced
 version cost \$5000 plus \$1000 support, and limited the volume of data
@@
         Pflogsumm       & Yes           & No                & Partial \dag{}    & Good           & Yes          \\
-        Sawmill         & Yes           & Yes               & Searches          & Reasonable     & No     \\
+        Sawmill         & Yes           & Yes               & Searches          & Very good      & \nialpha{}   \\
         Splunk          & Yes           & Yes               & Searches          & Abundant       & No           \\
@@
         Logrep          & No            & Yes               & No                & Poor           & Yes          \\
-        \parsername{}   & Yes           & Yes \nialpha{}    & No \nibeta{}      & XXX            & Yes    \\
+        \parsername{}   & Yes           & Yes \nibeta{}     & No \nichi{}       & XXX            & Yes          \\
         \tableline{}
@@
 
-        \item [\nialpha{}] \parsername{} is the only parser with
+        \item [\nialpha{}] Sawmill's source code is available in an
+            obfuscated form, so that customers can compile it on platforms
+            that pre-compiled binaries are not available for.
+
+        \item [\nibeta{}] \parsername{} is the only parser with
             documentation for its data store.
 
-        \item [\nibeta{}] \parsername{} defers report generation to
+        \item [\nichi{}] \parsername{} defers report generation to
             subsequent programs, but all the necessary data and
@@
 \end{table}
+
+\clearpage{}

### 2008-11-30 doc/sgai-2008-slides.tex
Slides for my SGAI paper.

\documentclass{beamer}
%\mode<presentation>
%{
%  \usetheme{Warsaw}
%  % or ...
%
%  \setbeamercovered{transparent}
%  % or whatever (possibly just delete it)
%}
\useoutertheme{infolines}

\usepackage[english]{babel}
% or whatever

\usepackage[latin1]{inputenc}
% or whatever

\usepackage{times}
\usepackage[T1]{fontenc}
% Or whatever. Note that the encoding and the font should match. If T1
% does not look nice, try deleting the line with the fontenc.

\title{A User-Extensible and Adaptable Parser Architecture}

\author{John Tobin \and Carl Vogel}
% - Give the names in the same order as the appear in the paper.
% - Use the \inst{?} command only if the authors have different
%   affiliation.

\institute[Trinity College]
{
  School of Computer Science and Statistics\\
  Trinity College, University of Dublin
}

\date[SGAI 2008]{Twenty-eighth SGAI International Conference on Artificial Intelligence}

\subject{Theoretical Computer Science}

% If you have a file called "university-logo-filename.xxx", where xxx
% is a graphic format that can be processed by latex or pdflatex,
% resp., then you can add a logo as follows:

% \pgfdeclareimage[height=0.5cm]{university-logo}{university-logo-filename}
% \logo{\pgfuseimage{university-logo}}

%% Delete this, if you do not want the table of contents to pop up at
%% the beginning of each subsection:
%\AtBeginSubsection[]
%{
%  \begin{frame}<beamer>{Outline}
%    \tableofcontents[currentsection,currentsubsection]
%  \end{frame}
%}

% If you wish to uncover everything in a step-wise fashion, uncomment
% the following command:

%\beamerdefaultoverlayspecification{<+->}

\begin{document}

\begin{frame}
    \titlepage{}
\end{frame}

% XXX MAYBE REMOVE THIS
\begin{frame}{Outline}
    \tableofcontents{}
    % You might wish to add the option [pausesections]
\end{frame}

% Structuring a talk is a difficult task and the following structure
% may not be suitable. Here are some rules that apply for this
% solution:

% - Exactly two or three sections (other than the summary).
% - At *most* three subsections per section.
% - Talk about 30s to 2min per frame. So there should be between about
%   15 and 30 frames, all told.

% - A conference audience is likely to know very little of what you
%   are going to talk about. So *simplify*!
% - In a 20min talk, getting the main ideas across is hard
%   enough. Leave out details, even if it means being less precise than
%   you think necessary.
% - If you omit details that are vital to the proof/implementation,
%   just say so once. Everybody will be happy with that.

\section{Background}

\begin{frame}{Background}

    \begin{description}

        \item [Aim] To improve SMTP-time anti-spam restriction / measures /
            techniques.

        \item [Method] Parse Postfix log files and analyse the data
            extracted.

    \end{description}

    This requires a parser for Postfix log files.

\end{frame}

\section{XXX}

\begin{frame}{XXX}

    \begin{itemize}

        \item Postfix log lines change over time and from site to site, so
            it must be easy for the end user to parse new lines.

        \item The processing required for a type of line (e.g.\ remote
            server connecting to send mail) rarely changes, but it can be
            quite complex.

        \item By separating parsing of lines from processing of lines, we
            can make parsing new lines as easy as possible, while still
            allowing for complicated processing when necessary.

    \end{itemize}

\end{frame}

\section{Architecture}

\begin{frame}{Architecture}

    \begin{description}

        \item [Framework] The framework handles loading rules and providing
            services to actions, e.g.\ data storage.  The framework tries
            each rule in turn until one matches, then it invokes the action
            specified by the rule.

        \item [Actions] Each action performs the processing required, e.g.\
            XXX GIVE SOME EXAMPLES\@.

        \item [Rules] Each rule matches one type of log line, e.g.\ the log
            line resulting from client connecting, a mail successfully
            delivered, or a mail rejected by an anti-spam measure.

    \end{description}

\end{frame}

\section{Features}

\begin{frame}{Features}

    \begin{itemize}

        \item Cascaded parsing.

        \item The parser does not need to be recompiled when rules are
            changed.

        \item Different rulesets can easily be used.

        \item Rules can be added by actions while the parser is running,
            allowing dynamically changing parsing, e.g.\ macros.

        \item XXX MORE FEATURES\@?

    \end{itemize}

\end{frame}

\section{Efficiency}

\begin{frame}{Efficiency}

    \begin{itemize}

        \item Rules are sorted by how frequently they match.

        \item Architecture scales linearly as the number of rules
            increases.

        \item Performance is \ldots{} XXX

        \item XXX MORE\@?

    \end{itemize}

\end{frame}

\section{Results}

\begin{frame}{Results.}

    \begin{itemize}

        \item 100\% coverage of log lines.

        \item 100\% coverage of all mails accepted, delivered, or rejected.

        \item XXX rules.

        \item XXX MORE\@?

    \end{itemize}

\end{frame}

\section*{Summary}

\begin{frame}{Summary}

    \begin{itemize}

        \item The \alert{first main message} of your talk in one or two lines.

        \item The \alert{second main message} of your talk in one or two lines.

        \item Perhaps a \alert{third message}, but not more than that.

    \end{itemize}

\end{frame}

\end{document}

### 2008-12-03 doc/sgai-2008-slides.tex
Lots of improvements to the slides.

@@
 \useoutertheme{infolines}
+\defbeamertemplate*{headline}{JT theme}{}
 
@@
 % does not look nice, try deleting the line with the fontenc.
+
+\newcommand{\timingnote}[1]{%
+    \textbf{#1}%
+}
 
@@
 
-%% Delete this, if you do not want the table of contents to pop up at
-%% the beginning of each subsection:
-%\AtBeginSubsection[]
-%{
-%  \begin{frame}<beamer>{Outline}
-%    \tableofcontents[currentsection,currentsubsection]
-%  \end{frame}
-%}
-
-% If you wish to uncover everything in a step-wise fashion, uncomment
-% the following command:
-
-%\beamerdefaultoverlayspecification{<+->}
-
 \begin{document}
@@
 \begin{frame}
+    \timingnote{30 seconds}
     \titlepage{}
@@
 
-% XXX MAYBE REMOVE THIS
-\begin{frame}{Outline}
+\begin{frame}{Outline of this talk}
+    \timingnote{30 seconds --- 1 minute}
     \tableofcontents{}
-    % You might wish to add the option [pausesections]
 \end{frame}
@@
 
+    \timingnote{1--2 minutes}
+
+    This parser architecture was developed as part of a larger project to
+    improve anti-spam techniques.
+
     \begin{description}
 
-        \item [Aim] To improve SMTP-time anti-spam restriction / measures /
-            techniques.
+        \item [Currently] Most anti-spam techniques are content-based, and
+            require each mail to be accepted before determining its spam
+            status.
+
+        \item [However] Rejecting mail at SMTP-time is more efficient, and
+            senders of non-spam mails that are incorrectly rejected will be
+            informed immediately.
+
+        \item [Project Aim] To improve SMTP-time anti-spam techniques, by
+            providing a platform for reasoning about anti-spam techniques.
 
@@
 
+        \item [XXX] This requires a parser for Postfix log files.
+
     \end{description}
-
-    This requires a parser for Postfix log files.
 
@@
 
-\section{XXX}
+\section{Input characteristics}
 
-\begin{frame}{XXX}
+\begin{frame}{Input characteristics}
+
+    \timingnote{1 minute}
 
@@
         \item Postfix log lines change over time and from site to site, so
-            it must be easy for the end user to parse new lines.
+            it must be easy for the end user to parse new log lines.
 
-        \item The processing required for a type of line (e.g.\ remote
-            server connecting to send mail) rarely changes, but it can be
-            quite complex.
+        \item The processing required for a type of log line (e.g.\ a
+            remote server connecting to send mail) rarely changes, but it
+            can be quite complex.
 
-        \item By separating parsing of lines from processing of lines, we
-            can make parsing new lines as easy as possible, while still
-            allowing for complicated processing when necessary.
+        \item By separating parsing of log lines from processing of log
+            lines, we can make parsing new or changed log lines as easy as
+            possible, while still allowing for complicated processing when
+            necessary.
 
@@
 
+    \timingnote{2 minutes}
+
+    The architecture is divided into three parts:
+
     \begin{description}
@@
         \item [Framework] The framework handles loading rules and providing
-            services to actions, e.g.\ data storage.  The framework tries
-            each rule in turn until one matches, then it invokes the action
-            specified by the rule.
+            services (XXX BETTER WORD\@?) to actions, e.g.\ data storage,
+            inter-action communication, XXX ANOTHER EXAMPLE\@.  For each
+            input log line, the framework tries each rule in turn until one
+            matches, then it invokes the action specified by the rule.
 
@@
 
-\section{Features}
+\section{Features of the Architecture}
 
-\begin{frame}{Features}
+\begin{frame}{Features of the Architecture}
+
+    \timingnote{2 minutes}
 
@@
 
-        \item Cascaded parsing.
+        \item Cascaded Parsing.  XXX SHOULD I HAVE A REFERENCE\@?
 
         \item The parser does not need to be recompiled when rules are
-            changed.
+            changed, allowing: quicker testing of modified rules; different
+            rulesets can easily be used; combining two or more rulesets.
 
-        \item Different rulesets can easily be used.
+        \item Rules can be changed or added by actions while the parser is
+            running, allowing dynamically changing parsing, e.g.\ macros.
+            This feature must be used with care.
 
-        \item Rules can be added by actions while the parser is running,
-            allowing dynamically changing parsing, e.g.\ macros.
+        \item The architecture is capable of parsing inputs that are not
+            completely understood, e.g.\ when developing a parser: the
+            framework will warn about unparsed inputs but will continue on
+            parsing as best it can.
+
+        \item Parsing can optionally be stateful, e.g.\ parsing a
+            programming language.
 
@@
 
+    \timingnote{2 minutes}
+
     \begin{itemize}
 
-        \item Rules are sorted by how frequently they match.
+        \item Median throughput when parsing 93 log files is 80.85~MB
+            (480,569 log lines) parsed per minute.
 
-        \item Architecture scales linearly as the number of rules
-            increases.
+        \item Rules are tagged with the name of a Postfix daemon, and will
+            only be used to parse log lines produced by that daemon.
 
-        \item Performance is \ldots{} XXX
+        \item Rules are sorted by how frequently they match, resulting in a
+            mean reduction of 14.78\% in parsing time compared to using
+            unsorted rules.
+
+        \item The architecture scales extremely well as the number of rules
+            increases: a 46.05\% increase in Postfix ruleset resulted in a
+            mean increase of 0.63\% in parsing time.
+
+        \item Parsing time scales linearly with input size, but is affected
+            by input content.
 
@@
 
-\begin{frame}{Results.}
+\begin{frame}{Results}
+
+    \timingnote{1--2 minutes}
 
@@
 
-        \item XXX rules.
+        \item 169 rules: the top 10\% match 85.03\% of log lines, the
+            remainder tailing off similar to a Power Law distribution.
 
         \item XXX MORE\@?
+
+    \end{itemize}
+
+\end{frame}
+
+\section{Future Directions}
+
+\begin{frame}{Future Directions}
+
+    \timingnote{1--2 minutes}
+
+    \begin{itemize}
+
+        \item Use Machine Learning instead of writing rules; Instance Based
+            Learning techniques a requirement due to the size of the data
+            set.
+
+        \item Detection of overlapping rules.
+
+        \item XXX REFERENCE IMPLEMENTATION\@?
 
@@
 
+    \timingnote{1 minute}
+
     \begin{itemize}
 
-        \item The \alert{first main message} of your talk in one or two lines.
+        \item East to develop a flexible yet strict parser.
 
-        \item The \alert{second main message} of your talk in one or two lines.
+        \item Easy for users to parse new or changed inputs.
 
-        \item Perhaps a \alert{third message}, but not more than that.
+        \item Parser can still be efficient and fast.
+
+    \end{itemize}
+
+    \timingnote{remainder}
+
+    \begin{itemize}
+
+        \item XXX MAYBE ON A SEPARATE SLIDE\@?
+
+        \item Questions?
+
+        \item Comments?
+
+        \item Suggestions?
 

### 2008-12-03 doc/sgai-2008-slides.tex
Improvements after meeting Carl.

@@
     \titlepage{}
+    XXX EMAIL ADDRESSES\@.
 \end{frame}
@@
 
-        \item [XXX] This requires a parser for Postfix log files.
+        \item [Conclusion] This requires a parser for Postfix log files.
 
@@
 
+    XXX SAMPLE LOG LINE\@; MAYBE HOW IT IS PARSED\@?
+
 \end{frame}
@@
         \item [Framework] The framework handles loading rules and providing
-            services (XXX BETTER WORD\@?) to actions, e.g.\ data storage,
-            inter-action communication, XXX ANOTHER EXAMPLE\@.  For each
-            input log line, the framework tries each rule in turn until one
-            matches, then it invokes the action specified by the rule.
+            support functions to actions, e.g.\ data storage, inter-action
+            communication, and utility functions.  For each input log line,
+            the framework tries each rule in turn until one matches, then
+            it invokes the action specified by the rule.
 
@@
 
-        \item Cascaded Parsing.  XXX SHOULD I HAVE A REFERENCE\@?
+        \item The architecture is capable of parsing inputs that are not
+            completely understood, e.g.\ when developing a parser: the
+            framework will warn about unparsed inputs but will continue on
+            parsing as best it can.  \timingnote{Emphasise this.}
+
+        \item Rules can be tagged with conditions which must hold before
+            the rule will be used to match an input.  XXX MAYBE EXTEND THIS
+            A LITTLE WITH AN EXAMPLE\@?  DO I NEED AN EXAMPLE, WHEN THERE'S
+            ONE IN A COUPLE OF SLIDE'S TIME\@?
 
         \item The parser does not need to be recompiled when rules are
-            changed, allowing: quicker testing of modified rules; different
-            rulesets can easily be used; combining two or more rulesets.
+            changed, allowing: quicker testing of modified rules; easing
+            interchange of rulesets; combining two or more rulesets.
+
+        \item Cascaded Parsing.
 
@@
             running, allowing dynamically changing parsing, e.g.\ macros.
-            This feature must be used with care.
+            This feature must be used with care.  \timingnote{Currying
+            functions in a calculator might be a good example.}
 
-        \item The architecture is capable of parsing inputs that are not
-            completely understood, e.g.\ when developing a parser: the
-            framework will warn about unparsed inputs but will continue on
-            parsing as best it can.
+    \end{itemize}
 
-        \item Parsing can optionally be stateful, e.g.\ parsing a
-            programming language.
+\end{frame}
 
-        \item XXX MORE FEATURES\@?
+\section{Results}
+
+\begin{frame}{Results}
+
+    \timingnote{1--2 minutes}
+
+    \begin{itemize}
+
+        \item 100\% coverage of log lines.
+
+        \item 100\% coverage of all mails accepted, delivered, or rejected.
+
+        \item 169 rules parsing 522 contiguous log files: the top 10\% of
+            the rules match 85.03\% of the log lines, the remainder tailing
+            off similar to a Power Law distribution.
+
+        \item Most actions are triggered by one or two rules, because they
+            deal with special cases.  Three actions are triggered by more
+            than 40 rules.  \timingnote{build/stats-action-distribution}
+            XXX EXPAND ON THIS\@?
+
+        \item XXX MORE\@?
 
@@
         \item The architecture scales extremely well as the number of rules
-            increases: a 46.05\% increase in Postfix ruleset resulted in a
-            mean increase of 0.63\% in parsing time.
+            increases: a 46.05\% increase in the Postfix ruleset ($115
+            \rightarrow{} 169$ rules) resulted in a mean increase of 0.63\%
+            in parsing time.
 
-        \item Parsing time scales linearly with input size, but is affected
-            by input content.
-
-        \item XXX MORE\@?
-
-    \end{itemize}
-
-\end{frame}
-
-\section{Results}
-
-\begin{frame}{Results}
-
-    \timingnote{1--2 minutes}
-
-    \begin{itemize}
-
-        \item 100\% coverage of log lines.
-
-        \item 100\% coverage of all mails accepted, delivered, or rejected.
-
-        \item 169 rules: the top 10\% match 85.03\% of log lines, the
-            remainder tailing off similar to a Power Law distribution.
+        \item Parsing time scales linearly with input size, but
+            \textit{is\/} affected by input content.
 
@@
 
+        \item Data mining to determine how well each anti-spam technique
+            works.
+
+        \item Data mining to search for combinations of features which can
+            be used to detect spam, and development of new anti-spam
+            techniques to defeat such spam.
+
         \item Use Machine Learning instead of writing rules; Instance Based
             Learning techniques a requirement due to the size of the data
-            set.
+            set.  \timingnote{Compare to SLCT-based tool.}
 
-        \item Detection of overlapping rules.
+        \item Detection of overlapping rules.  XXX MORE DETAILS\@.
 
@@
 
-\section*{Summary}
+\section{Summary}
 
@@
 
-        \item XXX MAYBE ON A SEPARATE SLIDE\@?
-
         \item Questions?
@@
         \item Comments?
-
-        \item Suggestions?
 

### 2008-12-07 doc/sgai-2008-slides.tex
Minor corrections and improvements; XXX items dealt with.

@@
 
-\author{John Tobin \and Carl Vogel}
+\author[John Tobin \and Carl Vogel]{John Tobin \and Carl Vogel\\
+    tobinjt@cs.tcd.ie vogel@cs.tcd.ie}
 % - Give the names in the same order as the appear in the paper.
@@
 {
-  School of Computer Science and Statistics\\
+    School of Computer Science and Statistics,\\
     Trinity College, University of Dublin
@@
     \titlepage{}
-    XXX EMAIL ADDRESSES\@.
 \end{frame}
 
-\begin{frame}{Outline of this talk}
+\begin{frame}{Outline Of This Talk}
     \timingnote{30 seconds --- 1 minute}
@@
 
-        \item [Currently] Most anti-spam techniques are content-based, and
-            require each mail to be accepted before determining its spam
-            status.
+        \item [Currently] Most commonly used anti-spam techniques are
+            content-based, and require each mail to be accepted before
+            determining its spam status.
 
         \item [However] Rejecting mail at SMTP-time is more efficient, and
-            senders of non-spam mails that are incorrectly rejected will be
-            informed immediately.
+            senders of non-spam mails that are incorrectly rejected will
+            receive a bounce notification immediately.
 
@@
 
-        \item [Method] Parse Postfix log files and analyse the data
-            extracted.
+        \item [Method] Parse Postfix log files and analyse the
+            extracted data.
 
-        \item [Conclusion] This requires a parser for Postfix log files.
+        \item [Conclusion] This method requires a parser for Postfix log
+            files.
 
@@
 
-\section{Input characteristics}
+\section{Input Characteristics}
 
-\begin{frame}{Input characteristics}
+\begin{frame}{Input Characteristics}
 
@@
 
-        \item The processing required for a type of log line (e.g.\ a
-            remote server connecting to send mail) rarely changes, but it
-            can be quite complex.
+        \item The processing required for a type of log line (e.g.\
+            rejecting a mail delivery attempt) rarely changes, but it can
+            be quite complex.
 
@@
 
-    XXX SAMPLE LOG LINE\@; MAYBE HOW IT IS PARSED\@?
+    Sample log line: \texttt{NOQUEUE\@: reject: RCPT from
+    client.example.net[192.0.2.1]: 554 5.7.1 <alice@example.net>: Relay
+    access denied; from=<eve@example.com> to=<alice@example.net> proto=SMTP
+    helo=<client.example.com>}
 
@@
 
-        \item [Framework] The framework handles loading rules and providing
-            support functions to actions, e.g.\ data storage, inter-action
+        \item [Framework] The framework manages rules and provides support
+            functions to actions, e.g.\ data storage, inter-action
             communication, and utility functions.  For each input log line,
@@
         \item [Actions] Each action performs the processing required, e.g.\
-            XXX GIVE SOME EXAMPLES\@.
+            create a new data structure when a client connects; record than
+            an anti-spam measure was successful; commit the data structure
+            when all processing is complete.
 
@@
         \item Rules can be tagged with conditions which must hold before
-            the rule will be used to match an input.  XXX MAYBE EXTEND THIS
-            A LITTLE WITH AN EXAMPLE\@?  DO I NEED AN EXAMPLE, WHEN THERE'S
-            ONE IN A COUPLE OF SLIDE'S TIME\@?
+            the rule will be used to match an input.
 
@@
 
-        \item Most actions are triggered by one or two rules, because they
-            deal with special cases.  Three actions are triggered by more
-            than 40 rules.  \timingnote{build/stats-action-distribution}
-            XXX EXPAND ON THIS\@?
+        \item Most actions are triggered by only one or two rules.
+
+        \item The top three actions are triggered by 41, 49, and 59 rules.
+            \timingnote{build/stats-action-distribution}
+            \timingnote{uninteresting, save\_data, delivery\_rejected}
 
@@
 
-        \item Median throughput when parsing 93 log files is 80.85~MB
-            (480,569 log lines) parsed per minute.
+        \item Median throughput is 80.85~MB (480,569 log lines) parsed per
+            minute.
 
-        \item Rules are tagged with the name of a Postfix daemon, and will
-            only be used to parse log lines produced by that daemon.
+        \item Rules are tagged with the name of a Postfix component, and
+            will only be used to parse log lines produced by that
+            component.
 
@@
             \textit{is\/} affected by input content.
-
-        \item XXX MORE\@?
 
@@
 
-        \item Use Machine Learning instead of writing rules; Instance Based
-            Learning techniques a requirement due to the size of the data
-            set.  \timingnote{Compare to SLCT-based tool.}
+        \item Use Machine Learning techniques instead of writing rules;
+            Instance Based Learning techniques must be used due to the size
+            of the data set.  \timingnote{Compare to SLCT-based tool.}
 
-        \item Detection of overlapping rules.  XXX MORE DETAILS\@.
+        \item Detection of overlapping rules: intersection of Finite
+            Automata or Push-Down Automata.
 
-        \item XXX REFERENCE IMPLEMENTATION\@?
+        \item Provide a reference implementation.
 
@@
 
-        \item East to develop a flexible yet strict parser.
+        \item Easy to develop a flexible yet strict parser.
 

### 2008-12-08 doc/sgai-2008-slides.tex
Some improvements.

@@
 
-        \item [Currently] Most commonly used anti-spam techniques are
-            content-based, and require each mail to be accepted before
+        \item [Currently] Most of the commonly used anti-spam techniques
+            are content-based, and require each mail to be accepted before
             determining its spam status.
@@
 
-        \item [Conclusion] This method requires a parser for Postfix log
+        \item [Therefore] This method requires a parser for Postfix log
             files.
@@
 
-\section{Input Characteristics}
+\section{Characteristics Of The Input: Postfix Log Files}
 
-\begin{frame}{Input Characteristics}
+\begin{frame}{Characteristics Of The Input: Postfix Log Files}
 
@@
 
-    Sample log line: \texttt{NOQUEUE\@: reject: RCPT from
-    client.example.net[192.0.2.1]: 554 5.7.1 <alice@example.net>: Relay
-    access denied; from=<eve@example.com> to=<alice@example.net> proto=SMTP
-    helo=<client.example.com>}
-
 \end{frame}
 
-\section{Architecture}
+\section{Parser Architecture}
 
-\begin{frame}{Architecture}
+\begin{frame}{Parser Architecture}
 
@@
 
-    The architecture is divided into three parts:
+    The parser architecture is divided into three parts: Framework,
+    Actions, and Rules.
 
@@
 
-        \item [Actions] Each action performs the processing required, e.g.\
-            create a new data structure when a client connects; record than
-            an anti-spam measure was successful; commit the data structure
-            when all processing is complete.
+        \item [Actions] Each action performs the processing required by one
+            type of log line.  Actions correspond to events: events within
+            Postfix create log lines, and actions perform the processing
+            required by those events.
 
-        \item [Rules] Each rule matches one type of log line, e.g.\ the log
+        \item [Rules] Each rule matches one log line variant, e.g.\ the log
             line resulting from client connecting, a mail successfully
@@
 
-\section{Features of the Architecture}
+\section{Features Of The Architecture}
 
-\begin{frame}{Features of the Architecture}
+\begin{frame}{Features Of The Architecture}
 
@@
 
-        \item The architecture is capable of parsing inputs that are not
-            completely understood, e.g.\ when developing a parser: the
-            framework will warn about unparsed inputs but will continue on
-            parsing as best it can.  \timingnote{Emphasise this.}
+        \item The architecture can be used when parsing input files that
+            are not completely understood or that do not have 100\% parsing
+            coverage.  \timingnote{Emphasise this.}
 
         \item Rules can be tagged with conditions which must hold before
-            the rule will be used to match an input.
+            the rule will be used when attempting to match an input.
 
@@
 
-        \item Rules can be changed or added by actions while the parser is
-            running, allowing dynamically changing parsing, e.g.\ macros.
-            This feature must be used with care.  \timingnote{Currying
-            functions in a calculator might be a good example.}
+        \item Rules can be added, changed, or removed by actions while the
+            parser is running, allowing dynamically changing parsing, e.g.\
+            macros.  This feature must be used with care.
+            \timingnote{Currying functions in a calculator might be a good
+            example.}
 
@@
 
-\section{Results}
+\section{Results Of Writing A Parser Using This Architecture}
 
-\begin{frame}{Results}
+\begin{frame}{Results Of Writing A Parser Using This Architecture}
 
@@
 
-        \item 169 rules parsing 522 contiguous log files: the top 10\% of
-            the rules match 85.03\% of the log lines, the remainder tailing
-            off similar to a Power Law distribution.
+        \item Zero false positives or false negatives.
 
-        \item Most actions are triggered by only one or two rules.
+        \item 169 rules parsing 522 contiguous log files: the top 10 rules
+            match 85.03\% of the log lines, the remainder tail off similar
+            to a Power Law distribution.
+
+        \item 18 actions; most are triggered by only one or two rules.
+            \timingnote{client connecting, postfix reloading}
 
         \item The top three actions are triggered by 41, 49, and 59 rules.
-            \timingnote{build/stats-action-distribution}
+            \timingnote{build/stats-action-distribution:}
             \timingnote{uninteresting, save\_data, delivery\_rejected}
-
-        \item XXX MORE\@?
 
@@
 
-\section{Efficiency}
+\section{Parser Efficiency}
 
-\begin{frame}{Efficiency}
+\begin{frame}{Parser Efficiency}
 
@@
         \item Median throughput is 80.85~MB (480,569 log lines) parsed per
-            minute.
+            minute; median log file size is 50 MB, containing 285,000 log
+            lines.
 
@@
             mean reduction of 14.78\% in parsing time compared to using
-            unsorted rules.
+            unsorted rules.  Priority attribute can be used to override
+            this.
 
@@
         \item Parsing time scales linearly with input size, but
-            \textit{is\/} affected by input content.
+            \textit{is\/} affected by input content: the content determines
+            the actions that will run, and some actions take longer to run
+            than others.
 
@@
 
-        \item Easy to develop a flexible yet strict parser.
+        \item Parser Architecture makes it easy to develop a flexible yet
+            strict parser.
 
-        \item Easy for users to parse new or changed inputs.
+        \item Parser Architecture makes it easy for users to parse new or
+            changed inputs.
 

### 2008-12-08 doc/sgai-2008-slides.tex
Improvements after dry runs with Carl.

@@
 
-\institute[Trinity College]
+\institute[TCD]
 {
@@
     \titlepage{}
+    \timingnote{Mention paper in journal}
 \end{frame}
@@
 \begin{frame}{Outline Of This Talk}
-    \timingnote{30 seconds --- 1 minute}
+    \timingnote{30 seconds --- 1 minute\newline{}}
+    \timingnote{Brief statement of motive during background}
     \tableofcontents{}
@@
 
-        \item [However] Rejecting mail at SMTP-time is more efficient, and
-            senders of non-spam mails that are incorrectly rejected will
-            receive a bounce notification immediately.
+        \item [However] Rejecting mail at SMTP-time (during the delivery
+            from sender to recipient) is more efficient, and senders of
+            non-spam mails that are incorrectly rejected will receive a
+            bounce notification immediately.
 
-        \item [Project Aim] To improve SMTP-time anti-spam techniques, by
-            providing a platform for reasoning about anti-spam techniques.
+        \item [Project Aim] To improve SMTP-time anti-spam techniques.
 
         \item [Method] Parse Postfix log files and analyse the
-            extracted data.
+            extracted data.  \timingnote{Explain where log lines come from}
 
@@
 
-        \item [Framework] The framework manages rules and provides support
-            functions to actions, e.g.\ data storage, inter-action
-            communication, and utility functions.  For each input log line,
-            the framework tries each rule in turn until one matches, then
-            it invokes the action specified by the rule.
+        \item [Framework] The framework manages rules, manages the parsing
+            process, and provides support functions to actions.  For each
+            input log line, the framework tries each rule in turn until one
+            matches, then it invokes the action specified by the matching
+            rule.
 
         \item [Actions] Each action performs the processing required by one
-            type of log line.  Actions correspond to events: events within
-            Postfix create log lines, and actions perform the processing
-            required by those events.
+            \alert{type} of log line.  Actions correspond to events: events
+            within Postfix create log lines, and actions perform the
+            processing required by those events.
 
-        \item [Rules] Each rule matches one log line variant, e.g.\ the log
-            line resulting from client connecting, a mail successfully
-            delivered, or a mail rejected by an anti-spam measure.
+        \item [Rules] Each rule matches one log line \alert{variant}; a
+            single log line \alert{type} may have many log line
+            \alert{variants}, e.g.\ rejection of a mail delivery attempt
+            has 59 variants.
 
@@
             the rule will be used when attempting to match an input.
+            \timingnote{Simple, but could be extended as necessary.}
 
-        \item The parser does not need to be recompiled when rules are
-            changed, allowing: quicker testing of modified rules; easing
-            interchange of rulesets; combining two or more rulesets.
+        \item Parsers and rulesets remain separate.
 
@@
             parser is running, allowing dynamically changing parsing, e.g.\
-            macros.  This feature must be used with care.
-            \timingnote{Currying functions in a calculator might be a good
-            example.}
+            macros.  \timingnote{Currying functions in a calculator might
+            be a good example.}
 
@@
 
-        \item 169 rules parsing 522 contiguous log files: the top 10 rules
-            match 85.03\% of the log lines, the remainder tail off similar
-            to a Power Law distribution.
+        \item 115 (now 169) rules parsing 93 (now 522) contiguous log
+            files: the 10 most frequently matching rules match 85.03\% of
+            the log lines, the remainder tail off similar to a Power Law
+            distribution.
 
-        \item 18 actions; most are triggered by only one or two rules.
+        \item 18 actions; 15 are triggered by only one or two rules.
             \timingnote{client connecting, postfix reloading}
 
-        \item The top three actions are triggered by 41, 49, and 59 rules.
-            \timingnote{build/stats-action-distribution:}
-            \timingnote{uninteresting, save\_data, delivery\_rejected}
+        \item 3 actions are triggered by an average of 50 rules:
+            uninteresting (41), save\_data (49), and delivery\_rejected
+            (59).
+            %\timingnote{build/stats-action-distribution:}
 
@@
             minute; median log file size is 50 MB, containing 285,000 log
-            lines.
+            lines; typically 6 or 7 log lines per mail.
 
@@
         \item Rules are sorted by how frequently they match, resulting in a
-            mean reduction of 14.78\% in parsing time compared to using
-            unsorted rules.  Priority attribute can be used to override
-            this.
+            mean reduction of 14.78\% in parsing time.  Priority attribute
+            can be used to override this.  \timingnote{change in ordering
+            is informative.  Described in paper}
 
@@
             \rightarrow{} 169$ rules) resulted in a mean increase of 0.63\%
-            in parsing time.
+            in parsing time.  \timingnote{less than 2/3 of 1\%; 2 orders of
+            magnitude}
 
         \item Parsing time scales linearly with input size, but
-            \textit{is\/} affected by input content: the content determines
-            the actions that will run, and some actions take longer to run
-            than others.
+            \textit{is\/} affected by input log lines: the log lines
+            determine the number of rules used and the actions that will
+            run, and some actions take longer to run than others.
 
@@
 
-\section{Future Directions}
+\section{Ongoing And Future Directions}
 
-\begin{frame}{Future Directions}
+\begin{frame}{Ongoing And Future Directions}
 
     \timingnote{1--2 minutes}
+
+    Ongoing:
 
@@
 
+    \end{itemize}
+
+    Future:
+
+    \begin{itemize}
+
         \item Use Machine Learning techniques instead of writing rules;
             Instance Based Learning techniques must be used due to the size
-            of the data set.  \timingnote{Compare to SLCT-based tool.}
+            of the data set.  This is one use of the ability to add or
+            change rules at run-time.  \timingnote{Compare to SLCT-based
+            tool.}
 
         \item Detection of overlapping rules: intersection of Finite
-            Automata or Push-Down Automata.
+            Automata or Push-Down Automata; how do rule conditions overlap?
 
@@
 
-        \item Parser Architecture makes it easy to develop a flexible yet
-            strict parser.
+        \item This Parser Architecture makes it easy to develop a flexible
+            yet strict parser.
 
-        \item Parser Architecture makes it easy for users to parse new or
-            changed inputs.
+        \item This Parser Architecture nevertheless makes it easy for users
+            to parse new or changed inputs.
 
-        \item Parser can still be efficient and fast.
+        \item Parsers written using this architecture can still be
+            efficient and fast.
+
+        \item The Postfix log file parser provides a basis for evaluating
+            SMTP-time anti-spam techniques, something that has not been
+            available before.
 

### 2008-12-12 doc/sgai-2008-slides.tex
Final changes before presentation.

@@
 \newcommand{\timingnote}[1]{%
-    \textbf{#1}%
+    %\textbf{#1}%
 }
@@
         \item Postfix log lines change over time and from site to site, so
-            it must be easy for the end user to parse new log lines.
+            it must be easy for the end user to parse new or changed log
+            lines.
 
@@
             lines, we can make parsing new or changed log lines as easy as
-            possible, while still allowing for complicated processing when
+            possible, while still allowing for complex processing when
             necessary.
@@
 
-    The parser architecture is divided into three parts: Framework,
-    Actions, and Rules.
+    The parser architecture is divided into three parts: \newline{}
+    Framework, Actions, and Rules.
 
@@
         \item [Framework] The framework manages rules, manages the parsing
-            process, and provides support functions to actions.  For each
-            input log line, the framework tries each rule in turn until one
-            matches, then it invokes the action specified by the matching
-            rule.
+            process, and provides support functions to actions.  \newline{}
+            For each input log line, the framework tries each rule in turn
+            until one matches, then it invokes the action specified by the
+            matching rule.
 
@@
         \item The architecture can be used when parsing input files that
-            are not completely understood or that do not have 100\% parsing
+            are not completely understood or do not have 100\% parsing
             coverage.  \timingnote{Emphasise this.}
@@
         \item Rules can be added, changed, or removed by actions while the
-            parser is running, allowing dynamically changing parsing, e.g.\
-            macros.  \timingnote{Currying functions in a calculator might
-            be a good example.}
+            parser is running, allowing dynamically changing parsing.
+            \timingnote{Currying functions in a calculator might be a good
+            example.}
 
@@
             files: the 10 most frequently matching rules match 85.03\% of
-            the log lines, the remainder tail off similar to a Power Law
-            distribution.
+            the log lines, the remaining rules tail off similar to a Power
+            Law distribution.
 
-        \item 18 actions; 15 are triggered by only one or two rules.
-            \timingnote{client connecting, postfix reloading}
+        \item 18 actions, of which 15 are triggered by only one or two
+            rules.  \timingnote{client connecting, postfix reloading}
 
-        \item 3 actions are triggered by an average of 50 rules:
-            uninteresting (41), save\_data (49), and delivery\_rejected
-            (59).
+        \item 3 actions are triggered by a high number of rules:
+            uninteresting~(41), save\_data~(49), and
+            delivery\_rejected~(59).
             %\timingnote{build/stats-action-distribution:}
@@
         \item Rules are sorted by how frequently they match, resulting in a
-            mean reduction of 14.78\% in parsing time.  Priority attribute
-            can be used to override this.  \timingnote{change in ordering
-            is informative.  Described in paper}
+            mean reduction of 14.78\% in parsing time.  A priority
+            attribute can be used to override this.  \timingnote{change in
+            ordering is informative.  Described in paper}
 
@@
             \textit{is\/} affected by input log lines: the log lines
-            determine the number of rules used and the actions that will
-            run, and some actions take longer to run than others.
+            determine the number of rules used and the actions that run,
+            and some actions take longer to run than others.
 
@@
 
-        \item Data mining to determine how well each anti-spam technique
-            works.
+        \item Data mining to determine how effective each of the anti-spam
+            techniques we currently use is.
 
@@
             be used to detect spam, and development of new anti-spam
-            techniques to defeat such spam.
+            techniques to reject such spam.
 
@@
 
-        \item Use Machine Learning techniques instead of writing rules;
-            Instance Based Learning techniques must be used due to the size
-            of the data set.  This is one use of the ability to add or
-            change rules at run-time.  \timingnote{Compare to SLCT-based
-            tool.}
+        \item Use Machine Learning techniques instead of writing rules by
+            hand; Instance Based Learning techniques must be used due to
+            the size of the data set.  This is one potential use of the
+            ability to add or change rules at run-time.
+            \timingnote{Compare to SLCT-based tool.}
 

### 2008-12-15 doc/logparser-introduction.tex
Improvements to the glossaries appearance. Minor fixes.

@@
 
-Appendix~XXX provides a glossary of terms used in the thesis.
+Appendix~\ref{\glsautoprefix{main}} provides a glossary of terms used in
+the thesis.
 
-Appendix~XXX provides a list of acronyms used in the thesis.
+Appendix~\ref{\glsautoprefix{\acronymtype}} provides a list of acronyms
+used in the thesis.
+
+Appendix~\ref{\glsautoprefix{postfix}} gives a brief description of Postfix
+daemons.

### 2008-12-15 doc/logparser-other-parsers-review.tex
Some minor changes based on the SGAI paper.
Improvements to the glossaries appearance. Minor fixes.

@@
-\newpage{} %XXX
-
 \section{State of the Art Review}
@@
 \label{state of the art review}
-
-XXX UPDATE THE BLURBS\@.
 
@@
         without extensive and intrusive modifications to the parser;
-        \parsername{}'s architecture is described in \sectionref{why
+        \parsernames{} architecture is described in \sectionref{why
         separate rules, actions, and framework?}.
@@
 
-XXX EXTEND THIS\@.
-
-Each of the parsers was tested with the \numberOFlogFILES{} test log files
+Each of the reviewed parsers was tested with the \numberOFlogFILES{} test log files
 described in \sectionref{parser efficiency}.  The data extracted by
 \parsername{} is documented in \sectionref{connections table} and
-\sectionref{results table}.
-
-XXX LIST THE DATA NOT EXTRACTED BY EACH PARSER\@.
+\sectionref{results table}; that list is repeated here, without the
+explanations:  server\_ip, server\_hostname, client\_ip, client\_hostname,
+helo, queueid, start, end, warning, smtp\_code, sender, recipient, size,
+message\_id, data, timestamp.
 
@@
 \label{prior art}
+
+XXX LIST THE DATA NOT EXTRACTED BY EACH PARSER FROM HERE ON\@.
 
@@
 
+XXX UPDATE THE BLURBS FROM HERE ON\@.
+
 \begin{quotation}
@@
     \begin{tabular}{llllll}
+        \tabletopline{}%
         Parser          & Parsed test   & Data              & Custom            & Documentation  & Source       \\
                         & log files?    & store?            & reports?          & quality?       & code?        \\
-        \tableline{}%
+        \tablemiddleline{}%
         \gls{LMA}       & No            & Yes               & No                & Poor           & Yes          \\ 
@@
         \parsername{}   & Yes           & Yes \nibeta{}     & No \nichi{}       & XXX            & Yes          \\
-        \tableline{}
+        \tablebottomline{}%
     \end{tabular}

### 2009-01-09 doc/logparser-acronyms.tex
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.

@@
     description={
-        A Mail Transfer Agent sends and receives mail via
-        SMTP\backslashat{}.  Users submit mail to an MTA via their mail
-        client (e.g.\ Microsoft Outlook, Thunderbird, webmail services);
-        the sending MTA transfers the mail to the receiving MTA (which may
-        in turn forward it to another MTA).
+        A Mail Transfer Agent sends and receives mail via SMTP\@.  Users
+        submit mail to an MTA via their mail client (e.g.\ Microsoft
+        Outlook, Thunderbird, webmail services); the sending MTA transfers
+        the mail to the receiving MTA (which may in turn forward it to
+        another MTA).
     }
@@
         The Request For Comments series is a series of proposals defining
-        various protocols and file formats, e.g.\ SMTP\backslashat{}.  The
-        name is somewhat misleading: initially the authors were asking for
-        peer review, but these documents are now the de facto standards the
+        various protocols and file formats, e.g.\ SMTP\@.  The name is
+        somewhat misleading: initially the authors were asking for peer
+        review, but these documents are now the de facto standards the
         Internet runs on.

### 2009-01-09 doc/logparser-glossary.tex
Minor cleanups.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.

@@
     embeddable, zero-configuration SQL database engine.\/} SQLite is an
-    \gls{SQL} implementation focusing on correctness, simplicity and speed.
-    Unlike other \gls{SQL} implementations it does not require a separate
-    server process, greatly simplifying deployment of an application
-    utilising it.  More details can be found
+    \gls{SQL} implementation focusing on correctness, simplicity, and
+    speed.  Unlike other \gls{SQL} implementations it does not require a
+    separate server process, greatly simplifying deployment of an
+    application utilising it.  More details can be found
     at~\url{http://www.sqlite.org/}.
@@
     e.g.\ bounce notifications.  In \gls{SMTP} all addresses are enclosed
-    in $<>$, so \textit{username\at{}domain\/} becomes
-    \textit{$<$username\at{}domain$>$\/}; thus $<>$ is actually an empty
+    in $<>$, so \textit{username@domain\/} becomes
+    \textit{$<$username@domain$>$\/}; thus $<>$ is actually an empty
     address, but is always written as $<>$ for clarity.  All mail servers
@@
     directed graph; the simplest example is when mail for
-    \texttt{alice\at{}example.net} is delivered to
-    \texttt{bob\at{}example.com}, and mail for \texttt{bob\at{}example.com}
-    is delivered to \texttt{alice\at{}example.net}.
+    \texttt{alice@example.net} is delivered to \texttt{bob@example.com},
+    and mail for \texttt{bob@example.com} is delivered to
+    \texttt{alice@example.net}.
 }}

### 2009-01-09 doc/logparser-introduction.tex
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.

--- 2008-11-23 doc/logparser-introduction.tex
@@
-\section{Introduction}
+\chapter{Introduction}
 
@@
 
-Appendix~XXX provides a glossary of terms used in the thesis.
+Appendix~\ref{Glossary} provides a glossary of terms used in the thesis.
 
-Appendix~XXX provides a list of acronyms used in the thesis.
+Appendix~\ref{Acronyms} provides a list of acronyms used in the thesis.
+
+Appendix~\ref{Postfix Daemons} gives a brief description of Postfix
+daemons.

### 2009-01-09 doc/logparser-postfix-daemons.tex
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.

@@
     determined only by the portion of the email address on the left side of
-    the \at{}, whereas with \daemon{virtual} delivery the destination is
+    the @, whereas with \daemon{virtual} delivery the destination is
     determined by the entire email address.  E.g.\ if the server is
     responsible for both the \textbf{example.org} and \textbf{example.net}
-    domains: \daemon{local} considers \textbf{john\at{}example.org} and
-    \textbf{john\at{}example.net} to have the same mailbox, whereas
+    domains: \daemon{local} considers \textbf{john@example.org} and
+    \textbf{john@example.net} to have the same mailbox, whereas
     \daemon{virtual} considers them to have different mailboxes.  Virtual

### 2009-01-09 doc/logparser.tex
Started updating blurbs etc. in the start of the art review. Dealt with a few XXX items.
Improved the appearance of URLs and notes in the bibliography.
Improvements after dry runs with Carl.
Some minor changes based on the SGAI paper.
Improvements to the glossaries appearance. Minor fixes.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.

@@
 % $Id$
-\documentclass[a4paper,12pt,draft]{article}
+\documentclass[a4paper,12pt,draft]{report}
 
@@
 \usepackage[final]{graphicx}
+\usepackage{url}
 % Add the bibliography into the table of contents.
-\usepackage[section,numbib]{tocbibind}
+\usepackage[chapter,numbib]{tocbibind}
 % Create a label for the last page.  Might be useful for "page 23/79" or
@@
 
-% Extra packages recommended by Springer: they change the fonts from
-% Computer Modern to something else and tells Latex to use scalable fonts.
+% Extra footnote functionality, including references to earlier footnotes.
+\usepackage[bottom]{footmisc}
+% The Euro symbol
+\usepackage{eurofont}
+
+% Aligned list environments.
+\usepackage{eqlist}
+
+% Tell Latex to use scalable fonts
 \usepackage{type1cm}
+
+% Extra packages recommended by Springer.
 %\usepackage{mathptmx}
@@
 %\usepackage{courier}
-% Extra footnote functionality, including references to earlier footnotes.
-\usepackage{footmisc}
+%\usepackage{makeidx}
+%\usepackage{multicol}
 
-% Aligned list environments.
-\usepackage{eqlist}
+% Embed SVN Id etc.
+\usepackage{svn}
 
@@
 \usepackage{cite}
-
-% Embed SVN Id etc.
-\usepackage{svn}
+% Check for unused references.
+% \usepackage{refcheck}
 
@@
 
+% Acronyms and glossary entries.  Useful options:
+% * nonumberlist disables the list of page numbers after each entry.
+\usepackage[acronym,toc,numberedsection=autolabel,style=list,nonumberlist]{glossaries}
+\renewcommand{\glspostdescription}[0]{}
+\renewcommand{\glsautoprefix}[1]{glossary:#1}
+\newglossary[plg]{postfix}{pin}{pout}{Postfix Daemons}
+\makeglossaries{}
+
 % Provides commands to distinguish between pdf and dvi output.
@@
 \ifpdf{}
-    \usepackage[pdftex,hyperfootnotes=false]{hyperref}
+    \usepackage[pdftex,hyperfootnotes=false,plainpages=false,pdfpagelabels]{hyperref}
 \else{}
-    \usepackage[dvips,hyperfootnotes=false]{hyperref}
+    \usepackage[dvips,hyperfootnotes=false,plainpages=false,pdfpagelabels]{hyperref}
     % This is necessary for wrapping URLs in the bibliography when
@@
 
-% Acronyms and glossary entries.
-\usepackage[acronym,toc,numberedsection]{glossaries}
-\newglossary[plg]{postfix}{pin}{pout}{Postfix Daemons}
-\makeglossaries{}
-
 %\setlength{\floatsep}{0pt}
@@
 % reference.
-\usepackage[nomsgs]{refcheck}
+%\usepackage[nomsgs]{refcheck}
 % Show where references are used.
@@
 %\usepackage{backref}
-\renewcommand{\refname}{Bibliography}
-% Add space before urls to quell underfull hbox warnings.
-\newcommand{\urlprefix}[0]{\hfill{}}
+%\renewcommand{\refname}{Bibliography}
+% Put each URL in the bibliography on a new line.
+\newcommand{\urlprefix}[0]{\newline{}}
 % New formatting commands.
@@
 %\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{%
+\newcommand{\showtable}[3]{
     \begin{table}[ht]
@@
     \texttt{postfix/#1}%
-}
-
-% This is ridiculous, but I cannot put @ in glossary entries, so . . .
-\newcommand{\at}[0]{%
-    @%
-}
-% This is ridiculous, but I cannot put @ in glossary entries, so . . .
-\newcommand{\backslashat}[0]{%
-    \@%
 }
@@
 \newcommand{\parsername}{\gls{PLP}}
-\newcommand{\parsernames}{\gls{PLP}'s}
-
-% This keeps hyperref fairly happy about the Abstract entry in the table of
-% contents.
-\newcounter{dummy}
+% XXX SHOULD THIS BE \glspl{PLP} or \gls{PLP}'s ??
+\newcommand{\parsernames}{\glspl{PLP}}
 
@@
 \maketitle
-\SVN$Id$
-\begin{center}\SVNId{}\end{center}
 
-\refstepcounter{dummy}
-\addcontentsline{toc}{section}{Abstract}
 \begin{abstract}
+
+    \phantomsection{}
+    \addcontentsline{toc}{chapter}{Abstract}
 
@@
 
+\SVN$Id$
+\begin{center}
+    \SVNId{}
+\end{center}
+
+XXX WHY DOES \\ref{} NOT WORK PROPERLY FOR CHAPTERS IN THE APPENDIX\@?
+
 XXX WEKA FOR DECISION TREE DISPLAY\@.
@@
 
-XXX WOULD TABLES LOOK BETTER IF CAPTIONS WERE LEFT-ALIGNED\@?
-
-XXX REDUCE THE LIST OF ACRONYMS\@.
-
 XXX CHECK YEARS FOR BIBLIOGRAPHY ENTRIES\@.
@@
 XXX IMPROVE BIBLIOGRAPHY APPEARANCE\@: NAMES LOOK UGLY (E.G.\ P.\
-\empty{}.\ Porters), URL SPACING IS MESSED UP --- URLS \\SHOULD PROBABLY BE
-LAST, ON THEIR OWN LINE\@.
+\empty{}.\ Porters)\@.
+
+XXX INVESTIGATE Doug Lenat.
 
@@
 
-%\bibliographystyle{logparser-bibliography-style}
-\bibliographystyle{sgai-2008-elsevier}
+\bibliographystyle{logparser-bibliography-style}
 \bibliography{logparser-bibliography}
 
-% Redefine the command used to produce the glossary title, because the
-% default command produces an unnumbered section whereas I want a numbered
-% section.
-%\renewcommand{\glossarytitle}{\section{Glossary}\label{Glossary}}
 \label{Glossary}
-\printglossaries{}
-% Redefine the command a second time to produce acronyms instead of a
-% glossary.
-%\renewcommand{\glossarytitle}{\section{Acronyms}\label{Acronyms}}
-%\printacronym{}
+\printglossary{}
+\label{Acronyms}
+\printglossary[type=\acronymtype]{}
+\label{Postfix Daemons}
+\printglossary[type=postfix]{}
 

### 2009-01-11 doc/logparser-other-parsers-review.tex
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Cleanups and improvements to the review of other parsers.

@@
-\section{State of the Art Review}
+\chapter{State of the Art Review}
 
@@
 algorithms, and parsers against those already available, to accurately
-judge what improvements, if any, are delivered by the newcomers.  Ten other
+judge what improvements, if any, are delivered by the newcomers.  Ten
 parsers have been reviewed as part of the background research for this
-project, starting with the only previously published research in this area
-that the author is aware of.  There are not that many previously developed
-Postfix log file parsers --- indeed it was quite difficult to find ten
-parsers to review for this project --- and the functionality offered ranges
-from quite basic to much more mature, depending on the needs of the author
-of the parser.  The programs are not reviewed from an independent
-viewpoint; the objective is to compare and contrast each program with
-\parsername{}.  There are some important differences between \parsername{}
-and the parsers reviewed here:
+project, comparing and contrasting each parser with \parsername{}; the
+first parser reviewed is the only previously published research in this
+area that the author is aware of.    There are not many previously
+developed Postfix log file parsers --- indeed it was quite difficult to
+find ten parsers to review for this project --- and the functionality
+offered ranges from quite basic to much more mature, depending on the needs
+of the author of the parser.  There are some important differences between
+\parsername{} and the parsers reviewed here:
+
+XXX SHOULD THIS BE IN THE CONCLUSION INSTEAD\@?
 
@@
 
-Each of the reviewed parsers was tested with the \numberOFlogFILES{} test log files
-described in \sectionref{parser efficiency}.  The data extracted by
-\parsername{} is documented in \sectionref{connections table} and
-\sectionref{results table}; that list is repeated here, without the
-explanations:  server\_ip, server\_hostname, client\_ip, client\_hostname,
-helo, queueid, start, end, warning, smtp\_code, sender, recipient, size,
-message\_id, data, timestamp.
+Each of the reviewed parsers was tested with the \numberOFlogFILES{} test
+log files described in \sectionref{parser efficiency}.  The data extracted
+by \parsername{} is documented in \sectionref{connections table} and
+\sectionref{results table}; for convenience that list is repeated here:
+server\_ip, server\_hostname, client\_ip, client\_hostname, helo hostname,
+queueid, start time, end time, smtp\_code, sender, recipient, size,
+message\_id.
 
-\subsection{Log Mail Analyser}
+\section{Log Mail Analyser}
 
 \label{prior art}
-
-XXX LIST THE DATA NOT EXTRACTED BY EACH PARSER FROM HERE ON\@.
 
@@
 least useful of the three output formats: it does not provide the
-functionality of a basic \gls{SQL} database, not can it be used with
+functionality of a basic \gls{SQL} database, nor can it be used with
 standard Unix text processing tools.
@@
 \textbf{(key, value)} pairs, so any additional structuring of the data is
-imposed by the application, thus it behoves the application to document
-this structure.
+imposed by the application, thus the application must document this
+structure.
 
@@
 \gls{IP} address, server \gls{IP} address, sender and recipient addresses,
-\gls{SMTP} code, and size (for accepted mails only).  It omits XXX XXX\@.
-Handling of multiple recipients, \gls{SMTP} codes or remote
-servers\footnote{A single mail may be sent to multiple remote servers if it
-was addressed to recipients in different domains, or Postfix needs to try
-multiple servers for one or more recipients.} is not explained in the
-paper; experimental observation shows that multiple records are added when
-there are multiple recipients (sadly the records are not associated or
-linked in any way), and presumably the same approach is taken when there
-are multiple destination servers.
+\gls{SMTP} code, and size (for accepted mails only); unlike \parsername{}
+it does not store the server hostname, helo hostname, queueid, start and
+end times, or message id (for accepted mails only).  Handling of multiple
+recipients, \gls{SMTP} codes or remote servers\footnote{A single mail may
+be sent to multiple remote servers if it was addressed to recipients in
+different domains, or Postfix needs to try multiple servers for one or more
+recipients.} is not explained in the paper; experimental observation shows
+that multiple records are added when there are multiple recipients (sadly
+the records are not associated or linked in any way), and presumably the
+same approach is taken when there are multiple destination servers.
 
@@
 
-\begin{enumerate}
+\begin{itemize}
 
@@
 
-\end{enumerate}
+\end{itemize}
 
@@
 
-\begin{itemize}
+\begin{enumerate}
 
@@
 
-\subsection{Pflogsumm}
+\section{Pflogsumm}
 
@@
 
-Pflogsumm produces a report designed for troubleshooting rather than for
-in-depth analysis.  It does not support saving any data or extracting more
-data than is required for producing the report.  Both the parsing and
-reporting are difficult to extend as it is a specialised tool, unlike the
-easily extensible design of \parsername{}.  It does not correlate log lines
-by queueid or \gls{pid}, and does not need to deal with the complications
-encountered during this project.  Pflogsumm produces a useful report, and
-successfully parsed the \numberOFlogFILES{} log files tested
-with,\footnote{The results it reported were not verified in detail, but it
-did not report any errors, and has a very good reputation amongst Postfix
-users.} but it is not a suitable base for this project.
+Pflogsumm produces a report designed for troubleshooting rather than
+in-depth analysis.  It does not support saving any data, and it does not
+extract any data than is not required for producing its report.  Both the
+parsing and reporting are difficult to extend as it is a specialised tool,
+unlike the easily extensible design of \parsername{}.  It does not
+correlate log lines by queueid or \gls{pid}, and does not need to deal with
+the complications encountered during this project.  Pflogsumm produces a
+useful report, and successfully parsed the \numberOFlogFILES{} log files
+tested with,\footnote{The results it reported were not verified in detail,
+but it did not report any errors, and has a very good reputation amongst
+Postfix users.} but it is not a suitable base for this project.  It does
+not extract the helo hostname, queueid, start and end times, or message id.
 
-Pflogsumm's report includes the following:
+Pflogsumm has many options are available to include or exclude certain
+sections of the report; by default it includes the following:
 
@@
 
-Many options are available to include or exclude certain sections of the
-report.
-
-\subsection{Sawmill Universal Log File Analysis and Reporting}
+\section{Sawmill Universal Log File Analysis and Reporting}
 
@@
 Its data extraction facilities (described later) are too limited to save
-sufficient data for the purposes of this project; it extracts three
-different sets of data from Postfix log files, but they are not interlinked
-in any way.  The documentation does not suggest that any attempt is made to
-correlate log lines or deal with the difficulties documented in
-\sectionref{complications}.
+sufficient data for the purposes of this project: although it can extract
+three different sets of data from Postfix log files, they are not
+interlinked in any way.  The documentation does not suggest that any
+attempt is made to correlate log lines or deal with the difficulties
+documented in \sectionref{complications}.
 
-Sawmill has three different Postfix log file parsers:
+Sawmill has three different Postfix log file parsers, extracting three
+different sets to data:
 
@@
 
-Even if the three data sets were linked together Sawmill would extract less
-data than \parsername{}, and it does not appear to extract data about
-rejections except when the rejection is caused by a \gls{DNSBL} check
-(\textit{RBL list\/} in the list of fields).
+Even if the three data sets were combined Sawmill would extract less data
+than \parsername{}: it omits the helo hostname, queueid, and start and end
+times.  Sawmill does not extract any data about rejections except when the
+rejection is caused by a \gls{DNSBL} check (\texttt{RBL list} in the list
+of fields).
 
@@
 (discounts are available when buying multiple licences); in contrast
-\parsername{} is free and the code is freely available.  Sawmill is
+\parsername{} is free to use and the code is freely available.  Sawmill is
 supplied with thorough and well written documentation; everything the
@@
 instructions on how to compile Sawmill so that it can use a non-commercial
-version of MySQL\@.
+version of MySQL (this was not attempted during the review process).
 
 Sawmill's web interface supports searching on any combination of the fields
-it extracts, but the interface is neither as simple to use nor as
-informative as the interface provided by Splunk.  The administrative
-interface is much easier to use than Splunk's: it took only five minutes to
-start parsing a directory of log files.  Searches performed using the web
-interface produced reasonable results, which were not verified in detail.
+it extracts, and the results obtained from searches were accurate.  The
+interface for searching is neither as simple to use nor as informative as
+the interface provided by Splunk.  The administrative interface is much
+easier to use than Splunk's: it took only five minutes to start parsing a
+whole directory of log files.  Searches performed using the web interface
+produced reasonable results, which were not verified in detail.
 
@@
 
-XXX ADD A SUMMARY\@.
+In summary Sawmill suffers from being a general purpose product; it is
+probably much more useful when parsing log files where each log line is
+self-contained, rather than linked to other log lines (e.g.\ web server log
+files).  It is not suitable as a base for this parser, as the source code
+made available is obfuscated and not intended for modification; in addition
+the architecture would probably need to be overhauled to deal with
+correlating log lines.
 
-\subsection{Splunk}
+\section{Splunk}
 
@@
 
-Splunk aims to index all an organisation's log files, providing a
+Splunk aims to index all of an organisation's log files, providing a
 centralised view capable of searching and correlating diverse log sources.
@@
 the database and leaves it to the user to utilise it without limit or
-restriction.  The interface is optimised for interactive rather than
+restriction.  The interface is optimised for interactive use rather than
 automated queries and it does not appear to be possible to write
@@
 
-When testing it took roughly 30 minutes to add a single log file to be
-indexed (log files must be indexed before searching), with the downside
-that the log file was copied into a spool directory, doubling the disk
-space usage.  The most suitable test would be to index all the log files in
-a particular directory, but after three hours, numerous futile attempts,
-and reading all the available documentation the author admitted defeat.
-Using the \gls{CLI} was more successful: invoking the command \newline{}
-\tab{} \texttt{splunk find logs }\textit{log-directory\/}\newline{} added
-40 of the \numberOFlogFILES{} log files to the queue for indexing.
-Repeated attempts enqueued the same 40 log files, without explaining why
-the others were excluded.\footnote{The log files appear to have been
-indexed once only; presumably Splunk keeps track of the files it has
-indexed and discards requests to index files for a second time.  This may
-or may not be a useful feature for \parsername{}.} There did not appear to
-be an option to order the log files to ensure they would be processed in
-the order they were created, though this may not be necessary or beneficial
-with Splunk.  Subsequently the author was successful in adding a single
-file at a time using the \gls{CLI}; a simple loop to add all desired log
-files was sufficient to index all files.  Splunk will periodically check
-all those files for updates unless they are manually removed from its list;
-this may or may not be useful behaviour.  Splunk did not appear to have any
-difficulty in parsing the log files, once instructed to do so.
+When testing Splunk it took roughly 30 minutes to figure out how to add a
+single log file to be indexed so that it could be searched, with the
+downside that the log file was copied into a spool directory before
+indexing, doubling the disk space usage.  The next test was to index all
+the log files in a particular directory, but after three hours, numerous
+futile attempts, and reading all the available documentation, the author
+admitted defeat.  Using the \gls{CLI} rather than the web interface was
+more successful: invoking the command \newline{} \tab{} \texttt{splunk find
+logs }\textit{log-directory\/}\newline{} added 40 of the
+\numberOFlogFILES{} log files to the queue for indexing.  Further attempts
+enqueued the same 40 log files, without explaining why the others were
+excluded.\footnote{The log files appear to have been indexed once only;
+presumably Splunk keeps track of the files it has indexed and discards
+requests to index files for a second time.  This may or may not be a useful
+feature for \parsername{}.} There did not appear to be an option to ensure
+the log files would be processed in the order they were created, though
+this may be neither necessary nor beneficial with Splunk.  Subsequently the
+author was successful in adding a single file at a time using the
+\gls{CLI}; a simple loop to add all desired log files was sufficient to
+index all files.  Splunk will periodically check all those files for
+updates unless they are manually removed from its list; this may or may not
+be useful behaviour.  Splunk did not appear to have any difficulty in
+indexing the log files, once they had been successfully added to its queue.
 \parsername{} parses the logs it is instructed to parse, in the order
@@
 useful information.  Searching using the website's interface confusingly
-tends to return results from old documentation rather than new.  In
-general the documentation appears to have been written by someone
-intimately acquainted with the software and having difficulty understanding
-how a newcomer would approach tasks or the questions they would ask.
+tends to return results from old documentation rather than new.  In general
+the documentation appears to have been written by someone intimately
+acquainted with the software who has difficulty understanding how a
+newcomer would approach tasks or the questions they would ask.
 
@@
 on the extracted fields or the full text of the log line.  \parsername{}
-extracts noticeably more data, though it does not make the full text of the
-line available, and the full power of \gls{SQL} is available when
-searching, allowing the user to search on arbitrarily complicated
-conditions.
+extracts noticeably more data (client and server \gls{IP} and hostname,
+queueid, start and end times, \gls{SMTP} code, and message ID), though it
+does not make the full text of the line available (this could be trivially
+added if desired, but would greatly increase the size of the resulting
+database).  The full power of \gls{SQL} is available when searching the
+data extracted by \parsername{}, allowing the user to search on arbitrarily
+complicated conditions.
 
@@
 
-\subsection{Isoqlog}
-
-XXX UPDATE THE BLURBS FROM HERE ON\@.
+\section{Isoqlog}
 
@@
 \noindent{}\url{http://www.enderunix.org/isoqlog/} \newline{}
-(Last checked 2007/08/13.)
+(Last checked 2009/01/11.)
 
@@
 
-XXX FINISH THIS\@.
-
-\subsection{AWStats}
+\section{AWStats}
 
@@
 \url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL} \newline{}
-(Last checked 2007/10/13.)
+(Last checked 2009/01/11.)
 
@@
 functionality.  The data it will extract from an \gls{MTA} log file is
-limited in comparison to \parsername{}: \newline{} \tab{} time2, email,
-email\_r, host, host\_r, method, url, code, and bytesd.\newline{} XXX WHAT
-FIELDS ARE MISSING\@?  There does not seem to be an explanation of any of
+limited in comparison to \parsername{}: time2, email, email\_r, host,
+host\_r, method, url, code, and bytesd.  There is no explanation for any of
 those fields in the documentation (\parsername{} provides copious
-documentation).  AWStats coerces Postfix log files into Apache\footnote{The
-Apache web server is the most popular HTTP server in use over the past 10
-years; more information is available at \url{http://httpd.apache.org/}.}
-format log files, for analysis by AWStats' HTTP log file parser.  The
-converting parser only deals with a small portion of the log lines
-generated by Postfix, silently skipping those it cannot deal with, and does
-not distinguish between different types of rejection; it would be a lot of
-work to extend it to handle new log lines.  Although it does correlate log
-lines by queueid (not by pid), it does not deal with any of the other
-complications described in \sectionref{complications}.  AWStats supports
-saving data but the format of the saved data is not documented, as far as
-the author could tell.  It also supports reading compressed log files, but
-that functionality was not tested.
+documentation), so the author could not understand the extracted data, nor
+determine what data is missing in comparison to \parsername{}.  AWStats
+coerces Postfix log files into Apache\footnote{The Apache web server is the
+most popular HTTP server in use over the past 10 years; more information is
+available at \url{http://httpd.apache.org/}.} format log files, for
+analysis by AWStats' HTTP log file parser.  The converting parser only
+deals with a small portion of the log lines generated by Postfix, silently
+skipping those it cannot deal with, and does not distinguish between
+different types of rejection; it would be a lot of work to extend it to
+handle new log lines.  Although it does correlate log lines by queueid (not
+by pid), it does not deal with any of the other complications described in
+\sectionref{complications}.  AWStats supports saving data but the format of
+the saved data is not documented, as far as the author could tell.  It also
+supports reading compressed log files, but that functionality was not
+tested.
 
@@
 
-XXX NEED A CONCLUSION\@.
+AWStats was not a suitable base for this project, because it assumes that
+all log files can be rewritten to be compatible with web server log files,
+and will contain similar data; coercing Postfix log files into web server
+log files, without substantial data loss, would require fully parsing the
+Postfix log files, so AWStats would not be required.  It may be possible to
+use AWStats graphing capabilities to generate reports, by generating log
+files to use as input to AWStats from the data extracted by \parsername{}.
 
-\subsection{Anteater}
+\section{Anteater}
+
+XXX LIST THE DATA NOT EXTRACTED BY EACH PARSER FROM HERE ON\@.
+
+XXX UPDATE THE BLURBS FROM HERE ON\@.
 
@@
 
-\subsection{Yet Another Advanced Logfile Analyser}
+\section{Yet Another Advanced Logfile Analyser}
 
@@
 
-\subsection{Lire}
+\section{Lire}
 
@@
 
-\subsection{Logrep}
+\section{Logrep}
 
@@
 
-\subsection{Summary}
+\section{Summary}
 

### 2009-01-11 doc/logparser-other-parsers-review.tex
Finished updating blurbs and noting which data each parser doesn't extract. Minor other cleanups and improvements.

@@
 
-XXX LIST THE DATA NOT EXTRACTED BY EACH PARSER FROM HERE ON\@.
-
-XXX UPDATE THE BLURBS FROM HERE ON\@.
-
 \begin{quotation}
@@
 \noindent{}\url{http://anteater.drzoom.ch/} \newline{}
-(Last checked 2007/08/13.)
+(Last checked 2009/01/11.)
 
@@
 date as Postfix has evolved considerably since this tool was last updated
-(November 2003).  As it neither ran successfully nor has documentation the
+(2003/11/06).  As it neither ran successfully nor has documentation the
 author can read a detailed review cannot be provided.
@@
 \noindent{}\url{http://yaala.org/} \newline{}
-(Last checked 2007/10/09.)
+(Last checked 2009/01/11.)
 
-YAALA uses a plugin based system to analyse log files and produce HTML
+YAALA uses a plugin-based system to analyse log files and produce HTML
 output reports, with all the parsing and report generation handled by
@@
 
-YAALA extracts most of the fields \parsername{} does, but unlike
-\parsername{} it does not maintain separate counters for each restriction;
-this rules out the possibility of using the collected data for
-optimisation, testing or understanding of restrictions.  YAALA's Postfix
-parser does not deal with the complications explained in
-\sectionref{complications}, though it does correlate log lines by queueid.
+\noindent{}YAALA's Postfix parser extracts some of the fields \parsername{}
+does: it stores either the \gls{IP} or the hostname for client and server,
+not both; it omits the helo hostname, queueid, \gls{SMTP} code, size of
+each accepted mail, and message ID\@.  It extracts some data that
+\parsername{} does not: the delay in delivering each mail, and how many
+times delivery was deferred for each mail; these could easily be extracted
+by \parsername{} if desired.  XXX EXTRACT DELAY AND DEFER ONCE AUTOMATIC
+EXTRACTION HAS BEEN FINISHED\@.  Unlike \parsername{}, YAALA does not
+maintain separate counters for each restriction; this rules out the
+possibility of using the collected data for optimisation, testing or
+understanding of restrictions.  YAALA's Postfix parser does not deal with
+the complications explained in \sectionref{complications}, though it does
+correlate log lines by queueid.
 
@@
 
+\end{quotation}
+
+\noindent{}\url{http://logreport.org/lire.html} \newline{}
+(Last checked 2009/01/11.)
+
+Lire is a general purpose log file parser supporting many different types
+of log file.  It takes a similar approach to YAALA, using plugins to parse
+different log file types.  The data extracted by its Postfix parser is not
+clearly documented; the manual says only:
+
+\begin{quotation}
+
     The email servers' reports will show you the number of deliveries and
@@
 
-\noindent{}\url{http://logreport.org/lire.html} \newline{}
-(Last checked 2008/04/29.)
+\noindent{}Examining the source code reveals that the parser looks for
+\texttt{<key>=<value>} pairs in each log line, extracts them, and
+correlates the data by queueid.  This approach will find the following
+fields: helo hostname, queueid, \gls{SMTP} code, sender and recipient
+addresses, and size of accepted mails.  It is unclear if the parser will
+extract any further data.  Lire misses the following fields extracted by
+\parsername{}: client and server \gls{IP} and hostname, start and end
+times, and message ID\@.
 
-Lire is a general purpose log file parser supporting many different types
-of log file.  The data extracted by its Postfix parser is shown in the
-quote above; notably rejections are not mentioned or dealt with, and unlike
-\parsername{} there is no facility to extend the parser.  It supports
-multiple output formats for generated reports (text, HTML, PDF, and Excel
-95) but the reports do not appear to be customisable; \parsername{} does
-not produce any reports.  Lire produces a report with less detail than
-Pflogsumm, and is considerable harder to configure.  Lire supports saving
-extracted data for later report generation, but accessing this data from
-another application is undocumented; given the source code it should be
-possible, with enough time and effort, to understand the format.
-\parsername{} uses an \gls{SQL} database to make accessing the extracted
-data as easy as possible.
+Lire supports multiple output formats for generated reports (text, HTML,
+PDF, and Excel 95) but the reports do not appear to be customisable;
+\parsername{} does not produce any reports.  Lire's report contains less
+detail than Pflogsumm, and is considerable harder to configure.  Lire
+supports saving extracted data for later report generation, but accessing
+this data from another application is undocumented; given the source code
+it should be possible, with enough time and effort, to understand the
+format.  \parsername{} uses an \gls{SQL} database to make accessing the
+extracted data as easy as possible.
 
@@
 \sectionref{complications}.  When testing Lire on the \numberOFlogFILES{}
-test log files it performed reasonably well, producing summaries of:
+test log files it performed reasonably well: the numbers it reports appear
+reasonable, and the subset verified by the author were correct.  Its report
+provided summaries of:
 
@@
 
-The numbers it reports appear reasonable, and the subset verified by the
-author were correct.
+Lire would not be a suitable base for this project: it does not extract
+enough data; does not deal with rejections in any way; does not make the
+extracted data easily available to other programs.  Its parser is
+in-extensible but could easily be replaced, however that would require
+writing a parser from scratch, so would not be any easier.
 
@@
 
-\noindent{}\url{http://www.itefix.no/phpws/index.php} \newline{}
-(Last checked 2007/11/18.)
+\noindent{}\url{http://www.itefix.no/i2/index.php} \newline{}
+(Last checked 2009/01/11.)
 
-Logrep extracts roughly half the fields \parsername{} does:
-
-XXX WHAT FIELDS ARE MISSING\@?
+Logrep extracts less than half the fields \parsername{} does:
 
@@
 
-It also counts the number of log lines parsed and skipped.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \gls{pid}.  The parsing is error prone: empty fields
-are saved when the log line does not match the regex, though it appears
-that they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It does not try to address any of the complications in
-\sectionref{complications} except for correlating by queueid.
+It also counts the number of log lines parsed and skipped.  It omits client
+\gls{IP} and hostname, server \gls{IP}, helo hostname, queueid, and message
+ID\@.  It extracts the delay for delivered mails, which \parsername{} does
+not.  Log lines are correlated based on the queueid (referred to as
+sessionname [sic] within Logrep), but not by \gls{pid}.  The parsing is
+error prone: empty fields are saved when the log line does not match the
+regex, though it appears that they will not overwrite existing data.  Most
+notably rejections are completely ignored, making it unsuitable for the
+purposes of this project.  It does not try to address any of the
+complications in \sectionref{complications} except for correlating by
+queueid.
 

### 2009-01-12 doc/logparser-other-parsers-review.tex
Proofread and corrected the introduction and LMA section of the state of the art review.

@@
 
-It is important to compare and contrast newly developed programs,
-algorithms, and parsers against those already available, to accurately
-judge what improvements, if any, are delivered by the newcomers.  Ten
-parsers have been reviewed as part of the background research for this
-project, comparing and contrasting each parser with \parsername{}; the
-first parser reviewed is the only previously published research in this
-area that the author is aware of.    There are not many previously
-developed Postfix log file parsers --- indeed it was quite difficult to
-find ten parsers to review for this project --- and the functionality
-offered ranges from quite basic to much more mature, depending on the needs
-of the author of the parser.  There are some important differences between
-\parsername{} and the parsers reviewed here:
+At the start of this project ten Postfix log file parsers were tested, with
+the hope of finding a suitable parser to build upon, rather than starting
+from scratch.  There are not many Postfix log file parsers available ---
+indeed it was quite difficult to find ten parsers to review for this
+project --- and the functionality offered ranges from quite basic to much
+more mature, depending on the needs of the author of the parser.  None of
+those parsers were suitable for this project, so the decision was taken to
+write a new parser.  The first parser reviewed is the only previously
+published research in this area that the author is aware of; it parses
+Postfix log files, but aim of the research is to show that presenting
+extracted data in a more easily accessible format is useful to systems
+administrators, rather than to improve anti-spam techniques..
+
+The same ten parsers have been reviewed and compared to this project's
+finished parser, to show how much effort would have been required to fulfil
+the aims and requirements of this project.  It is important to compare and
+contrast newly developed algorithms and parsers against those already
+available, to accurately judge what improvements, if any, are delivered by
+the newcomers.
 
 XXX SHOULD THIS BE IN THE CONCLUSION INSTEAD\@?
+
+There are some important differences between \parsername{} and the parsers
+reviewed here:
 
@@
     \item None of the parsers reviewed perform the kind of advanced parsing
-        and correlation required for this project (described in
-        \sectionref{complications})
+        required for this project or deal with the complications described
+        in \sectionref{complications}.
 
-    \item Only \parsername{} makes it possible to parse new log lines
-        without extensive and intrusive modifications to the parser;
-        \parsernames{} architecture is described in \sectionref{why
-        separate rules, actions, and framework?}.
+    \item Only \parsername{} enables parsing of new log lines without
+        extensive and intrusive modifications to the parser; \parsernames{}
+        architecture is described in \sectionref{why separate rules,
+        actions, and framework?}.
 
     \item The parsers reviewed all produce a report of varying complexity
-        and detail, whereas \gls{PLP} is designed to be a generic parser,
-        extracting data and leaving generation of reports from the data to
-        other programs.  Using an \gls{SQL} database simplifies the process
-        of generating such reports (discussed in \sectionref{database as
-        API}); some simple examples are given in \sectionref{motivation}.
+        and detail, whereas \gls{PLP} does not; it extracts data and leaves
+        generation of reports from the data to other programs.  Using an
+        \gls{SQL} database simplifies the process of generating such
+        reports (discussed in \sectionref{database as API}); some sample
+        queries are given in \sectionref{motivation}.  The parser developed
+        for this project is designed to enable much more detailed log file
+        analysis by providing a stable platform for subsequent programs to
+        develop upon.
 
-        The parser developed for this project is designed to enable much
-        more detailed log file analysis by providing a stable platform for
-        subsequent programs to develop upon.
-
-    \item Most of the parsers silently ignore log lines they cannot handle,
-        whereas \parsername{} complains loudly about every single log line
-        it fails to parse.  The exception is AWStats, which outputs the
-        percentage of input lines it was unable to parse, but does not
-        output the lines themselves.
+    \item Most of the reviewed parsers silently ignore log lines they
+        cannot handle, whereas \parsername{} complains loudly about every
+        single log line it fails to parse.  The exception is AWStats, which
+        outputs the percentage of input lines it was unable to parse, but
+        does not output the lines themselves.
 
@@
         in parsing time when doing so.  Although this is a minor
-        difference, support for reading compressed log files is quite
+        disparity, support for reading compressed log files is quite
         helpful, as it dramatically reduces the disk space required to
-        store old log files.
+        store historical log files.
 
-    \item Some of the parsers reviewed save the data extracted to a data
+    \item Some of the parsers reviewed save the extracted data to a data
         store, but the majority discard all data once they have finished
-        generating the report, making historical analysis impossible.
+        generating their report, making historical analysis impossible
+        without parsing all log files every time.
 
@@
 \sectionref{results table}; for convenience that list is repeated here:
-server\_ip, server\_hostname, client\_ip, client\_hostname, helo hostname,
-queueid, start time, end time, smtp\_code, sender, recipient, size,
-message\_id.
+server \gls{IP}, server hostname, client \gls{IP}, client hostname, helo
+hostname, queueid, start time, end time, \gls{SMTP} code, sender,
+recipient, size, message ID\@.
 
@@
 Utilizations\/}~\cite{log-mail-analyser}.  The aim of \gls{LMA} is quite
-different from this parser: it attempts to present correlated data from log
-files in a form suitable for a systems administrator to search using the
-myriad of standard Unix text processing utilities already available,
-producing both a \gls{CSV} file and either a MySQL
+different from \parsername{}: it attempts to present correlated data from
+log files in a form suitable for a systems administrator to search using
+the myriad of standard Unix text processing utilities already available.
+It produces a \gls{CSV} file and either a MySQL
 (\url{http://www.mysql.com/}) or Berkeley DB
@@
 The decision to support both \gls{CSV} and Berkeley DB appears to have been
-a serious limitation: \gls{CSV} is a very simple format where a record is
-stored in a single line, with fields separated by a comma or other
-punctuation symbol.  Problems with \gls{CSV} files include the need to
-escape separators in the data stored, providing multiple values for a field
-(e.g.\ multiple recipients), and adding new fields.  There is no mechanism
-to document the fields or the separator, unlike \gls{SQL} databases where
-every database includes a schema naming the fields and the type of data
-they store (integer, text, timestamp, etc.).
-
-Very little documentation is provided with \gls{LMA}, though some
-documentation is available in~\cite{log-mail-analyser}.  Studying the
+a serious limitation: XXX EXTEND\@: LIMITATIONS WILL BE EXPLAINED LATER OR
+SOMETHING\@.  Very little documentation is provided with \gls{LMA}, though
+some documentation is available in~\cite{log-mail-analyser}.  Studying the
 source code is informative, though this author had difficulty as the
 authors of \gls{LMA} wrote in Italian.
+
+\gls{CSV} is a very simple format where each record is stored in a single
+line, with fields separated by a comma or other punctuation symbol.
+Problems with \gls{CSV} files include the need to escape separators in the
+data stored, providing multiple values for a field (e.g.\ multiple
+recipients), and adding new fields.  There is no standard mechanism to
+document the fields or the separator, unlike \gls{SQL} databases where
+every database includes a schema naming the fields and the type of data
+they store (integer, text, timestamp, etc.).  The \gls{CSV} record format
+is not documented, but the output file contains a comment giving the
+format:\newline{} \texttt{\# Timestamp|Nome Client|IP Client|IP
+Server|From|To|Status|Size} \newline{}\gls{LMA} treats lines starting with
+\texttt{\#} as comments, but not all \gls{CSV} parsers will.
 
@@
 least useful of the three output formats: it does not provide the
-functionality of a basic \gls{SQL} database, nor can it be used with
-standard Unix text processing tools.
+functionality of a basic \gls{SQL} database, and unlike \gls{CSV} files it
+can not be used with standard Unix text processing tools.
 
 The schema used with the MySQL database is undocumented, but at least it is
-possible to discover the schema with an \gls{SQL} database, unlike with
+possible to discover the schema with an existing \gls{SQL} database, unlike with
 Berkeley DB\@; all \gls{SQL} databases embed the schema into the database
@@
 imposed by the application, thus the application must document this
-structure.
+structure.  MySQL support was not tested because there is no documentation
+on the schema required.
 
-The \gls{CSV} record format is not documented, but the output file contains
-a comment giving the format:\newline{} \texttt{\# Timestamp|Nome Client|IP
-Client|IP Server|From|To|Status|Size} \newline{}\gls{LMA} treats lines
-starting with \texttt{\#} as comments, but not all \gls{CSV} parsers will,
-as there is no standard for comments in \gls{CSV} files.
-
-\gls{LMA} stores the following date: time and date, client hostname and
-\gls{IP} address, server \gls{IP} address, sender and recipient addresses,
-\gls{SMTP} code, and size (for accepted mails only); unlike \parsername{}
-it does not store the server hostname, helo hostname, queueid, start and
-end times, or message id (for accepted mails only).  Handling of multiple
-recipients, \gls{SMTP} codes or remote servers\footnote{A single mail may
-be sent to multiple remote servers if it was addressed to recipients in
-different domains, or Postfix needs to try multiple servers for one or more
-recipients.} is not explained in the paper; experimental observation shows
-that multiple records are added when there are multiple recipients (sadly
-the records are not associated or linked in any way), and presumably the
-same approach is taken when there are multiple destination servers.
+Whether a MySQL database or Berkeley DB table is chosen in addition to the
+\gls{CSV} output, \gls{LMA} stores the following data: time and date,
+client hostname and \gls{IP} address, server \gls{IP} address, sender and
+recipient addresses, \gls{SMTP} code, and size (for accepted mails only).
+It is unclear which time and date is stored: start time, end time, or
+delivery time?  Unlike \parsername{} it does not store the server hostname,
+helo hostname, queueid, start and end times, timestamps for each log line,
+or message id (for accepted mails only).  Handling of multiple recipients,
+\gls{SMTP} codes, or remote servers\footnote{A single mail may be sent to
+multiple remote servers if it was addressed to recipients in different
+domains, or Postfix needs to try multiple servers for one or more
+recipients.} is not explained; experimental observation shows that multiple
+records are added when there are multiple recipients (sadly the records are
+not associated or linked in any way), and presumably the same approach is
+taken when there are multiple destination servers.
 
 \gls{LMA} requires major changes to the parser code to parse new log lines
-or extract additional data.  The code is structured as a long series of
-blocks that each handle all lines matching a single regex, so parsing new
-log lines requires modifying an existing regex or carefully inserting a new
-block in the correct place; extracting extra data will require modifying
-multiple blocks, regexes, or both.
+or to extract additional data.  The code is structured as a long series of
+blocks that each handle all log lines matching a single regex, so parsing
+new log lines requires modifying an existing regex or carefully inserting a
+new block in the correct place; extracting extra data will require
+modifying multiple blocks, regexes, or both.
 
 \gls{LMA} does not deal with any of the complications discussed in
-\sectionref{complications}.  It does not differentiate between different
-types of rejections, so it is not suitable for the purposes of this
-project; the data about which restriction caused the rejection is
-discarded, whereas the main goal of this project is to retain that data to
-aid optimisation and evaluation of Postfix restrictions.  \gls{LMA} fails
-to parse Postfix log files generated on Solaris hosts because the fields
-automatically prepended to each log line differ from those added on Linux
-hosts; log files from Solaris hosts (and possibly other operating systems)
-thus require preprocessing before parsing by \gls{LMA}.  Once the
-preprocessing has been performed on the \numberOFlogFILES{} test log files
-\gls{LMA} parses the log files without complaint, although it produced 32
-entries in its output file for every rejection in the input log file; it
-also missed some 40\% of delivered mail.  Once these glaring deficiencies
-were discovered the author did not waste any more time checking the
-results.
+\sectionref{complications}, except for correlating log lines by queueid;
+not correlating log lines by pid means it cannot correlate most rejections.
+It does not differentiate between different types of rejections, so it is
+not suitable for the purposes of this project; the data about which
+restriction caused the rejection is discarded, whereas the main goal of
+this project is to retain that data to aid optimisation and evaluation of
+anti-spam techniques.  \gls{LMA} fails to parse Postfix log files generated
+on Solaris hosts because the fields automatically prepended to each log
+line differ from those added on Linux hosts; log files from Solaris hosts
+(and possibly other operating systems) thus require preprocessing before
+parsing by \gls{LMA}.  Once the preprocessing has been performed on the
+\numberOFlogFILES{} test log files \gls{LMA} parses the log files without
+complaint, although it produced 32 entries in its output file for every
+rejection in the input log file; it also missed some 40\% of delivered
+mail.  Once these glaring deficiencies were discovered the author did not
+waste any more time checking the results.
 
 \gls{LMA} does provide some simple reports: IP-STORY, FROM-STORY,
-DAILY-EMAIL and DAILY-REJECT\@.  The first three extract \gls{CSV} lines
-for the specified client \gls{IP} address, sender address, or date
-respectively.  DAILY-REJECT initially failed with an error message from the
-Perl interpreter;\footnote{The error message was:
-\newline{}\texttt{Undefined subroutine \&main::LIST called at queryDB.pl
-line 372.}} after making some corrections to the code it worked, and
-extracted the \gls{CSV} lines for the specified day where the \gls{SMTP}
-code signifies a rejection.  All of these reports are trivially simple to
-produce from the \gls{CSV} file using the standard Unix tool
+DAILY-EMAIL and DAILY-REJECT\@.  These reports search the Berkeley DB files
+for matching records: the first three extract \gls{CSV} lines for the
+specified client \gls{IP} address, sender address, or date respectively.
+DAILY-REJECT initially failed with an error message from the Perl
+interpreter;\footnote{The error messages were: \newline{}\texttt{Undefined
+subroutine \&main::LIST called at queryDB.pl line
+372.}\newline{}\texttt{Undefined subroutine \&main::EXTRACT\_FROM\_DB
+called at queryDB.pl line 379.}} after making corrections to the code it
+worked, extracting the \gls{CSV} lines for the specified day where the
+\gls{SMTP} code signifies a rejection.  All of these reports are trivially
+simple to produce from the \gls{CSV} file using the standard Unix tool
 \texttt{awk}\glsadd{awk}; the most complicated, DAILY-REJECT, is merely:
 
+% perl queryDB.pl -dayreject 2007-01-26 > lma-query
+
 \begin{verbatim}
-awk -F\| '$1 ~ /2007-01-26/ && $7 ~ /^[450]|deferred/
-    { print $0; print " "; }' lma_output.txt
+awk -F\| 'BEGIN { previous = "" };
+    $1 ~ /2007-01-26/ && $7 != "250" && $0 != previous 
+    { print $0; print " "; previous = $0; }' lma_output.txt
 \end{verbatim}
@@
 
-    \item It outputs an unnecessary line containing only a single space
-        after each matching line, to accurately replicate the output of
+    \item It outputs a line containing only a single space after each
+        matching record, to accurately replicate the output of
         DAILY-REJECT\@.
 
-    \item Some records have an \gls{SMTP} code of \textit{0}, though this
-        is not a valid \gls{SMTP} code; presumably this is because
-        \gls{LMA} did not parse a log line properly.  DAILY-REJECT
-        considers this code to be a rejection, so the command above must
-        too.
+    \item DAILY-REJECT considers all \gls{SMTP} codes except \texttt{250}
+        to be rejections; this includes invalid \gls{SMTP} codes such as
+        \texttt{0} and \texttt{deferred}, so the awk command does too.
+        These invalid \gls{SMTP} codes are most likely present because of
+        incorrect parsing by \gls{LMA}.
 
-    \item In a small fraction of records the \gls{SMTP} code is
-        \textit{deferred\/} (again, not a valid \gls{SMTP} code, and most
-        likely because of mis-parsing by \gls{LMA}); it is treated as a
-        rejection by DAILY-REJECT, so the equivalent \texttt{awk} command
-        must do so, too.
+    \item \gls{LMA} produces 32 output lines in its \gls{CSV} file for
+        every single line it should have produced; the command above
+        suppresses duplicate sequential lines.
 
@@
 
-There are two differences between the output from DAILY-REJECT and the
-\texttt{awk} command:
+There are some differences between the output from DAILY-REJECT and the
+\texttt{awk} command; the author did not spend substantial time attempting
+to explain these differences.
 
@@
 
-    \item \gls{LMA} does not escape or replace the separators used in the
-        \gls{CSV} records when they are present in sender or recipient
-        addresses, leading to some mangled records; these records are not
-        present in the output from \texttt{awk}.  XXX WHY\@?  GIVE AN
-        EXAMPLE RECORD\@.
+    \item The output from DAILY-REJECT is missing some records which are
+        present in the \gls{CSV} file; this may be because it uses the
+        Berkeley DB files instead, and there may be differences between the
+        contents.
 
-    \item Where there are multiple records in the \gls{CSV} file,
-        DAILY-REJECT extracts only one record, whereas the \texttt{awk}
-        command above extracts all matching records.
+    \item Some records output by DAILY-REJECT are truncated: they are
+        missing the last | separating fields and the newline following it,
+        so the line containing only a single space is concatenated with the
+        record.
 
@@
 accessible fashion is useful to systems administrators), rather than a
-program designed to be useful in a production environment; no consideration
-was given to extensibility when writing it.
+program designed to be useful in a production environment.
 
@@
 Postfix users.} but it is not a suitable base for this project.  It does
-not extract the helo hostname, queueid, start and end times, or message id.
+not extract the helo hostname, queueid, start and end times, timestamps for
+each log line, or message id.
 
@@
 extracts noticeably more data (client and server \gls{IP} and hostname,
-queueid, start and end times, \gls{SMTP} code, and message ID), though it
-does not make the full text of the line available (this could be trivially
-added if desired, but would greatly increase the size of the resulting
-database).  The full power of \gls{SQL} is available when searching the
-data extracted by \parsername{}, allowing the user to search on arbitrarily
-complicated conditions.
+queueid, start and end times, timestamps for each log line, \gls{SMTP}
+code, and message ID), though it does not make the full text of the line
+available (this could be trivially added if desired, but would greatly
+increase the size of the resulting database).  The full power of \gls{SQL}
+is available when searching the data extracted by \parsername{}, allowing
+the user to search on arbitrarily complicated conditions.
 
@@
     \item [Keyfields:] incoming\_host, outgoing\_host, date, hour, sender,
-        recipient, defer\_count, delay.
+        recipient, defer\_count, delay.  Which date and hour are stored is
+        not documented: start time, end time, delivery time, or another
+        time?
 
@@
 not both; it omits the helo hostname, queueid, \gls{SMTP} code, size of
-each accepted mail, and message ID\@.  It extracts some data that
-\parsername{} does not: the delay in delivering each mail, and how many
-times delivery was deferred for each mail; these could easily be extracted
-by \parsername{} if desired.  XXX EXTRACT DELAY AND DEFER ONCE AUTOMATIC
-EXTRACTION HAS BEEN FINISHED\@.  Unlike \parsername{}, YAALA does not
-maintain separate counters for each restriction; this rules out the
-possibility of using the collected data for optimisation, testing or
-understanding of restrictions.  YAALA's Postfix parser does not deal with
-the complications explained in \sectionref{complications}, though it does
-correlate log lines by queueid.
+each accepted mail, start and end times, timestamps for each log line, and
+message ID\@.  It extracts some data that \parsername{} does not: the delay
+in delivering each mail, and how many times delivery was deferred for each
+mail; these could easily be extracted by \parsername{} if desired.  XXX
+EXTRACT DELAY AND DEFER ONCE AUTOMATIC EXTRACTION HAS BEEN FINISHED\@.
+Unlike \parsername{}, YAALA does not maintain separate counters for each
+restriction; this rules out the possibility of using the collected data for
+optimisation, testing or understanding of restrictions.  YAALA's Postfix
+parser does not deal with the complications explained in
+\sectionref{complications}, though it does correlate log lines by queueid.
 
@@
 \parsername{}: client and server \gls{IP} and hostname, start and end
-times, and message ID\@.
+times, timestamps of each log line, and message ID\@.
 
@@
     \item For mail sent and received: from address, size, and time and
-        date.
+        date.  Which date and hour are stored is not documented: start
+        time, end time, delivery time, or another time?
 

### 2009-01-13 doc/logparser-other-parsers-review.tex
Further corrections and improvements after proofreading.

@@
 \sectionref{results table}; for convenience that list is repeated here:
-server \gls{IP}, server hostname, client \gls{IP}, client hostname, helo
+server \gls{IP}, server hostname, client \gls{IP}, client hostname, HELO
 hostname, queueid, start time, end time, \gls{SMTP} code, sender,
@@
 delivery time?  Unlike \parsername{} it does not store the server hostname,
-helo hostname, queueid, start and end times, timestamps for each log line,
+HELO hostname, queueid, start and end times, timestamps for each log line,
 or message id (for accepted mails only).  Handling of multiple recipients,
@@
 in-depth analysis.  It does not support saving any data, and it does not
-extract any data than is not required for producing its report.  Both the
-parsing and reporting are difficult to extend as it is a specialised tool,
-unlike the easily extensible design of \parsername{}.  It does not
-correlate log lines by queueid or \gls{pid}, and does not need to deal with
-the complications encountered during this project.  Pflogsumm produces a
-useful report, and successfully parsed the \numberOFlogFILES{} log files
-tested with,\footnote{The results it reported were not verified in detail,
-but it did not report any errors, and has a very good reputation amongst
-Postfix users.} but it is not a suitable base for this project.  It does
-not extract the helo hostname, queueid, start and end times, timestamps for
-each log line, or message id.
-
-Pflogsumm has many options are available to include or exclude certain
+extract any data that it does not require to produce its report, e.g.\ it
+does not extract the HELO hostname, queueid, start and end times,
+timestamps for each log line, or message id.  Both the parsing and
+reporting are difficult to extend because it is a specialised tool, unlike
+the easily extensible design of \parsername{}.  It does not correlate log
+lines by queueid or \gls{pid}, and does not need to deal with the
+complications encountered during this project (\sectionref{complications}).
+Pflogsumm produces a useful report, and successfully parsed the
+\numberOFlogFILES{} log files it was tested with.\footnote{The results it reported
+were not verified in detail, but it did not report any errors, and has a
+very good reputation amongst Postfix users.}
+Pflogsumm has many options to include or exclude certain
 sections of the report; by default it includes the following:
@@
 interlinked in any way.  The documentation does not suggest that any
-attempt is made to correlate log lines or deal with the difficulties
-documented in \sectionref{complications}.
+attempt is made to correlate log lines by either queueid or pid or to deal
+with the difficulties documented in \sectionref{complications}.
 
@@
         \texttt{virus} are not explained in the documentation:
-        \texttt{server} is probably the server the mail is delivered to,
-        and \texttt{uid} might be the uid of the user submitting mail
-        locally.  Postfix does not perform any form of virus checking
-        (though it has many options for cooperating with an external virus
-        scanner), so the \texttt{virus} field is a mystery.
+        \texttt{server} is probably the hostname or \gls{IP} of the server
+        the mail is delivered to; \texttt{relay} might be the delivery
+        method: \gls{SMTP}, local delivery, or \gls{LMTP}; \texttt{uid}
+        might be the uid of the user submitting mail locally.  Postfix does
+        not perform any form of virus checking (though it has many options
+        for cooperating with an external virus scanner), so the
+        \texttt{virus} field is a mystery.
 
@@
         \newline{} Fields extracted: from, to, RBL list, client hostname,
-        and client \gls{IP}\@.  It also counts the number of and total size
-        of all mails delivered.
+        and client \gls{IP}\@.  It also counts the number and total size of
+        all mails delivered.
 
@@
 Even if the three data sets were combined Sawmill would extract less data
-than \parsername{}: it omits the helo hostname, queueid, and start and end
+than \parsername{}: it omits the HELO hostname, queueid, and start and end
 times.  Sawmill does not extract any data about rejections except when the
@@
 Sawmill's web interface supports searching on any combination of the fields
-it extracts, and the results obtained from searches were accurate.  The
-interface for searching is neither as simple to use nor as informative as
-the interface provided by Splunk.  The administrative interface is much
-easier to use than Splunk's: it took only five minutes to start parsing a
-whole directory of log files.  Searches performed using the web interface
-produced reasonable results, which were not verified in detail.
+it extracts, and searches performed using the web interface produced
+accurate results.  The interface for searching is neither as simple to use
+nor as informative as the interface provided by Splunk.  The administrative
+interface is much easier to use than Splunk's: it took only five minutes to
+start parsing a whole directory of log files.
 
@@
 probably much more useful when parsing log files where each log line is
-self-contained, rather than linked to other log lines (e.g.\ web server log
-files).  It is not suitable as a base for this parser, as the source code
-made available is obfuscated and not intended for modification; in addition
-the architecture would probably need to be overhauled to deal with
-correlating log lines.
+self-contained (e.g.\ web server log files), rather than log files
+containing interlinked log lines.  It is not suitable as a base for this
+parser, as the source code made available is obfuscated and not intended
+for modification; in addition the architecture would probably need to be
+overhauled or replaced to deal with correlating log lines.
 
@@
 graphs in real time, a facility not provided by \parsername{} (report
-generation has been deferred to a subsequent program).  Saved searches can
-be run periodically and the results emailed to a recipient or sent to an
-external program for further processing (though maybe without the graphs
-and detailed statistics); the author was unable to save searches, though
-that may have been due to limitations in the free version.  The database is
-not available for use by external programs, whereas \parsername{} provides
-the database and leaves it to the user to utilise it without limit or
-restriction.  The interface is optimised for interactive use rather than
-automated queries and it does not appear to be possible to write
-independent tools to utilise the Splunk database; there are additional
-reports available on \url{http://www.splunkbase.com/}, but the author was
-unable to find any: every category was empty, even those that the interface
-claimed had reports available.  Many types of reports are bundled with the
-software, though most are variations of a bar or pie chart, except bubble
-and heatmap graphs.  It is easy to drill down through the graphs to extract
-a portion of the data (e.g.\ select the hour with the largest number of
-events, then select a particular host, and finally a specific address),
-though it is not possible to search on partial words.  All searches
-performed using the indexed data returned reasonable results.  The web
-interface is quite attractive and simple to use when searching, but as an
-administrator it seems unnecessarily difficult to perform simple tasks.
+generation has been deferred to a subsequent program).  Searches can be
+based on the fields extracted by Splunk or the full text of the log line.
+Splunk allows quite complicated searches, but does not make the raw data
+available in an accessible form.  Saved searches can be run periodically
+and the results emailed to a recipient or sent to an external program for
+further processing (though maybe without the graphs and detailed
+statistics); the author was unable to save searches, though that may have
+been due to limitations in the free version.  The database is not available
+for use by external programs, whereas \parsername{} provides the database
+and leaves it to the user to utilise it without limit or restriction.  The
+interface is optimised for interactive use rather than automated queries
+and it does not appear to be possible to write independent tools to utilise
+the Splunk database;.  Some additional Postfix reports are supposedly
+available at \url{http://www.splunkbase.com/}, but the author was unable to
+find any Postfix reports, or indeed for any other log file types: every
+category was empty, even those that the web site claimed had many reports
+available.  Many types of graphs can be generated, though most are
+variations of a bar or pie chart, except bubble and heatmap graphs.  It is
+easy to drill down through the graphs to extract a portion of the data
+(e.g.\ select the hour with the largest number of events, then select a
+particular host, and finally a specific address), though it is not possible
+to search on partial words.  All searches performed using the indexed data
+returned reasonable results.
 
-When testing Splunk it took roughly 30 minutes to figure out how to add a
-single log file to be indexed so that it could be searched, with the
+The web interface is quite attractive and simple to use when searching, but
+as an administrator it seems unnecessarily difficult to perform simple
+tasks.  When testing Splunk it took roughly 30 minutes to figure out how to
+add a single log file to be indexed so that it could be searched, with the
 downside that the log file was copied into a spool directory before
@@
 admitted defeat.  Using the \gls{CLI} rather than the web interface was
-more successful: invoking the command \newline{} \tab{} \texttt{splunk find
+partially successful: the command \newline{} \tab{} \texttt{splunk find
 logs }\textit{log-directory\/}\newline{} added 40 of the
@@
 author was successful in adding a single file at a time using the
-\gls{CLI}; a simple loop to add all desired log files was sufficient to
-index all files.  Splunk will periodically check all those files for
+\gls{CLI}:\newline{} \tab{} \texttt{splunk add tail
+}\textit{filename\/}\newline{} A simple loop was then enough to add all the
+desired log files.  Splunk will periodically check all indexed files for
 updates unless they are manually removed from its list; this may or may not
@@
 but poor organisation and sheer abundance makes it extremely hard to find
-useful information.  Searching using the website's interface confusingly
-tends to return results from old documentation rather than new.  In general
-the documentation appears to have been written by someone intimately
-acquainted with the software who has difficulty understanding how a
-newcomer would approach tasks or the questions they would ask.
+useful information.  Searches confusingly tended to return results from old
+documentation rather than new.  In general the documentation appears to
+have been written by someone intimately acquainted with the software, who
+has difficulty understanding how a newcomer would approach tasks or the
+questions they would ask.
 
@@
 
-When parsing Postfix log files Splunk extracts some standard fields: to and
-from addresses, HELO hostname, time and date, the host the log files were
-generated on, and protocol (\gls{SMTP} or \gls{ESMTP}).  It parses the
-standard syslog\glsadd{syslog} fields at the beginning of the log line, and
-extracts any \texttt{key=<value>} pairs occurring after the standard syslog
-prologue; these pairs are the fields listed above.  Searches can be based
-on the extracted fields or the full text of the log line.  \parsername{}
-extracts noticeably more data (client and server \gls{IP} and hostname,
-queueid, start and end times, timestamps for each log line, \gls{SMTP}
-code, and message ID), though it does not make the full text of the line
-available (this could be trivially added if desired, but would greatly
-increase the size of the resulting database).  The full power of \gls{SQL}
-is available when searching the data extracted by \parsername{}, allowing
-the user to search on arbitrarily complicated conditions.
+When parsing Postfix log files Splunk parses the standard
+syslog\glsadd{syslog} fields at the beginning of the log line, and extracts
+any \texttt{key=value} pairs occurring after the standard syslog prologue:
+to and from addresses, HELO hostname, and protocol (\gls{SMTP} or
+\gls{ESMTP}).  \parsername{} extracts noticeably more data (client and
+server \gls{IP} and hostname, queueid, start and end times, timestamps for
+each log line, \gls{SMTP} code, and message ID), though it does not make
+the full text of the line available (this could be trivially added if
+desired, but would greatly increase the size of the resulting database).
+The full power of \gls{SQL} is available when searching the data extracted
+by \parsername{}, allowing the user to search on arbitrarily complicated
+conditions.
 
 Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=<value>} fields from a log line; most
+above extracting the \texttt{key=value} fields from a log line; most
 importantly it makes no attempt to correlate log lines by queueid or
-\gls{pid}, or to handle any of the myriad complications discussed in
-\sectionref{complications}.  Some additional Postfix reports are supposedly
-available at \url{http://www.splunkbase.com/}, but the author was unable to
-find Postfix reports, or additional reports for any other log file types.
+\gls{pid}, or to handle any of the other myriad complications discussed in
+\sectionref{complications}.  Its source code is unavailable, so it could
+not be used as a base for this project, even if it fulfilled all other
+requirements.
 
@@
 
-Isoqlog's report misses most of the information gathered by \parsername{},
-and it only reports on the domains listed in its configuration file, making
-it impossible to produce reports for every sender domain.  It ignores all
-log lines except those for today, so it is impossible to analyse historical
-logs, and testing with the \numberOFlogFILES{} test log files is pointless.
-It does maintain a record of data previously extracted, which the newly
-extracted data is merged into (no information is provided on the format of
-the data store).  The data extracted is limited to the number of mails sent
-by each sender, unlike the copious amounts of data extracted by
-\parsername{}.  It does not utilise rejection log lines in any way, so is
-unsuitable for the purposes of this project.  Its parsing is completely
-inextensible, indeed is almost incomprehensible, relying on
+Isoqlog's report misses most of the information gathered by \parsername{}:
+the data extracted is limited to the number of mails sent by each sender,
+and it only reports on senders from the domains listed in its configuration
+file, making it impossible to produce complete reports.  It ignores all log
+lines except those with today's date, so it is impossible to analyse
+historical log files, and testing with the \numberOFlogFILES{} test log
+files was pointless.  It does maintain a record of data previously
+extracted, which the newly extracted data is merged into; the format of the
+data store is undocumented.  It does not utilise rejection log lines in any
+way, so is unsuitable for the purposes of this project.  Its parsing is
+completely inextensible, indeed is almost incomprehensible, relying on
 \texttt{scanf(3)}, unexplained fixed offsets, and low level string
@@
 does: it stores either the \gls{IP} or the hostname for client and server,
-not both; it omits the helo hostname, queueid, \gls{SMTP} code, size of
+not both; it omits the HELO hostname, queueid, \gls{SMTP} code, size of
 each accepted mail, start and end times, timestamps for each log line, and
@@
 correlates the data by queueid.  This approach will find the following
-fields: helo hostname, queueid, \gls{SMTP} code, sender and recipient
+fields: HELO hostname, queueid, \gls{SMTP} code, sender and recipient
 addresses, and size of accepted mails.  It is unclear if the parser will
@@
 It also counts the number of log lines parsed and skipped.  It omits client
-\gls{IP} and hostname, server \gls{IP}, helo hostname, queueid, and message
+\gls{IP} and hostname, server \gls{IP}, HELO hostname, queueid, and message
 ID\@.  It extracts the delay for delivered mails, which \parsername{} does

### 2009-01-13 doc/logparser-other-parsers-review.tex
Further corrections after proofreading.

@@
 \label{state of the art review}
+
+XXX DO ANY OF THE PARSERS DISTINGUISH BETWEEN DIFFERENT TYPES OF
+REJECTIONS\@?
 
@@
 secondary by-sender, by-recipient, by-date, and by-\gls{IP} tables use the
-sender/recipient/date/\gls{IP} as the key, and the value is a \gls{CSV}
-list of integers referring to the main table.  This effectively
+sender/recipient/date/\gls{IP} address as the key, and the value is a
+\gls{CSV} list of integers referring to the main table.  This effectively
 re-implements \gls{SQL} foreign keys, but without the functionality offered
@@
         \texttt{virus} are not explained in the documentation:
-        \texttt{server} is probably the hostname or \gls{IP} of the server
-        the mail is delivered to; \texttt{relay} might be the delivery
-        method: \gls{SMTP}, local delivery, or \gls{LMTP}; \texttt{uid}
-        might be the uid of the user submitting mail locally.  Postfix does
-        not perform any form of virus checking (though it has many options
-        for cooperating with an external virus scanner), so the
-        \texttt{virus} field is a mystery.
+        \texttt{server} is probably the hostname or \gls{IP} address of the
+        server the mail is delivered to; \texttt{relay} might be the
+        delivery method: \gls{SMTP}, local delivery, or \gls{LMTP};
+        \texttt{uid} might be the uid of the user submitting mail locally.
+        Postfix does not perform any form of virus checking (though it has
+        many options for cooperating with an external virus scanner), so
+        the \texttt{virus} field is a mystery.
 
@@
 \gls{ESMTP}).  \parsername{} extracts noticeably more data (client and
-server \gls{IP} and hostname, queueid, start and end times, timestamps for
-each log line, \gls{SMTP} code, and message ID), though it does not make
-the full text of the line available (this could be trivially added if
-desired, but would greatly increase the size of the resulting database).
-The full power of \gls{SQL} is available when searching the data extracted
-by \parsername{}, allowing the user to search on arbitrarily complicated
-conditions.
+server \gls{IP} address and hostname, queueid, start and end times,
+timestamps for each log line, \gls{SMTP} code, and message ID), though it
+does not make the full text of the line available (this could be trivially
+added if desired, but would greatly increase the size of the resulting
+database).  The full power of \gls{SQL} is available when searching the
+data extracted by \parsername{}, allowing the user to search on arbitrarily
+complicated conditions.
 
@@
 
+XXX DOCUMENTATION\@?
+
 \section{AWStats}
@@
 
-AWStats will produce simple graphs for many different services, but
-supporting many different services without special purpose code limits its
-functionality.  The data it will extract from an \gls{MTA} log file is
-limited in comparison to \parsername{}: time2, email, email\_r, host,
-host\_r, method, url, code, and bytesd.  There is no explanation for any of
-those fields in the documentation (\parsername{} provides copious
-documentation), so the author could not understand the extracted data, nor
-determine what data is missing in comparison to \parsername{}.  AWStats
-coerces Postfix log files into Apache\footnote{The Apache web server is the
-most popular HTTP server in use over the past 10 years; more information is
-available at \url{http://httpd.apache.org/}.} format log files, for
-analysis by AWStats' HTTP log file parser.  The converting parser only
-deals with a small portion of the log lines generated by Postfix, silently
-skipping those it cannot deal with, and does not distinguish between
-different types of rejection; it would be a lot of work to extend it to
-handle new log lines.  Although it does correlate log lines by queueid (not
-by pid), it does not deal with any of the other complications described in
-\sectionref{complications}.  AWStats supports saving data but the format of
-the saved data is not documented, as far as the author could tell.  It also
-supports reading compressed log files, but that functionality was not
-tested.
+AWStats can produce simple graphs from many different services' log files,
+but supporting numerous log files formats without special purpose code
+limits its functionality.  The data it can extract from an \gls{MTA} log
+file is limited in comparison to \parsername{}: time2, email, email\_r,
+host, host\_r, method, url, code, and bytesd.  There is no explanation for
+any of those fields in the documentation (in contrast \parsername{}
+documents everything it extracts), so the author could not understand what
+data is extracted, nor determine what data is missing in comparison to
+\parsername{}.  AWStats coerces Postfix log files into Apache\footnote{The
+Apache web server is the most popular HTTP server in use over the past 10
+years; more information is available at \url{http://httpd.apache.org/}.}
+format log files, for analysis by AWStats' HTTP log file parser.  The
+converting parser only deals with a small portion of the log lines
+generated by Postfix, silently skipping those it cannot deal with, and does
+not distinguish between different types of rejection; extending it to parse
+all log lines would be at least as much work as writing a new parser.  It
+does correlate log lines by queueid (not by pid), but it does not deal with
+any of the other complications described in \sectionref{complications}.
+AWStats supports saving data, but the format of the data store is not
+documented.  It also supports reading compressed log files, but that
+functionality was not tested.
 
@@
 \numberOFlogLINES{} lines in the \numberOFlogFILES{} log files, so AWStats
-parsed only 17.15\% of the input lines, ignoring the remaining 82.85\%.
+parsed only 15.21\% of the input lines, declared 1.93\% were corrupt, and
+ignored the remaining 82.85\%.
 
@@
 log files, without substantial data loss, would require fully parsing the
-Postfix log files, so AWStats would not be required.  It may be possible to
-use AWStats graphing capabilities to generate reports, by generating log
-files to use as input to AWStats from the data extracted by \parsername{}.
+Postfix log files without using AWStats, i.e.\ would require writing a
+separate parser.  It may be possible to use AWStats' graphing capabilities
+to generate reports, by generating input for AWStats from the data
+extracted by \parsername{}.
 
@@
 run successfully when tested, and its parsing would certainly be out of
-date as Postfix has evolved considerably since this tool was last updated
-(2003/11/06).  As it neither ran successfully nor has documentation the
-author can read a detailed review cannot be provided.
+date because Postfix has evolved considerably since this tool was last
+updated (2003/11/06).  As it neither ran successfully nor has documentation
+the author can read, a detailed review cannot be provided.
 
@@
 
-YAALA uses a plugin-based system to analyse log files and produce HTML
-output reports, with all the parsing and report generation handled by
-modules.  Using YAALA as a base would be only slightly less work than
-starting from scratch, as both the input and output modules would need to
-be written specially; it may even be more work to implement the parser
-within the constraints of YAALA\@.  YAALA supports storing previously
-gathered data using Perl's Storable module~\cite{perl-storable}, so other
-Perl programs can use Storable to load, examine, and optionally modify the
-data; \parsername{} uses a well documented database which is accessible
-from the majority of programming languages.  This information was gleaned
-from the source code, as the documentation is sadly lacking.
+YAALA uses a plugin-based system to analyse log files and produce reports
+in HTML format, with all the parsing and report generation handled by
+plugins.  Using YAALA as a base for this project would be only slightly
+less work than starting from scratch, as both the input and output modules
+would need to be written specially; it may be more work to implement a
+parser within the constraints of YAALA\@.  YAALA supports storing
+previously gathered data using Perl's Storable module, so other Perl
+programs can use Storable to load, examine, and optionally modify the data;
+\parsername{} uses a well documented database which is accessible from the
+majority of programming languages.  This information was gleaned from the
+source code, as the documentation is sadly lacking.
 
-YAALA provides a Postfix parser that extracts the following two types of
-fields from specific log lines:
+YAALA provides a Postfix parser that extracts the following types of fields
+from specific log lines:
 
@@
 \noindent{}YAALA's Postfix parser extracts some of the fields \parsername{}
-does: it stores either the \gls{IP} or the hostname for client and server,
-not both; it omits the HELO hostname, queueid, \gls{SMTP} code, size of
-each accepted mail, start and end times, timestamps for each log line, and
-message ID\@.  It extracts some data that \parsername{} does not: the delay
-in delivering each mail, and how many times delivery was deferred for each
-mail; these could easily be extracted by \parsername{} if desired.  XXX
-EXTRACT DELAY AND DEFER ONCE AUTOMATIC EXTRACTION HAS BEEN FINISHED\@.
-Unlike \parsername{}, YAALA does not maintain separate counters for each
-restriction; this rules out the possibility of using the collected data for
-optimisation, testing or understanding of restrictions.  YAALA's Postfix
-parser does not deal with the complications explained in
-\sectionref{complications}, though it does correlate log lines by queueid.
+does: for client and server it stores either the \gls{IP} address or the
+hostname, not both; it omits the HELO hostname, queueid, \gls{SMTP} code,
+size of each accepted mail, start and end times, timestamps for each log
+line, and message ID\@.  It extracts some data that \parsername{} does not:
+the delay in delivering each mail, and how many times delivery was deferred
+for each mail; these could easily be extracted by \parsername{} if desired.
+XXX DELAY AND DEFERRAL\@.  Unlike \parsername{}, YAALA does not maintain
+separate counters for each restriction; this rules out the possibility of
+using the collected data for optimisation, testing, or understanding of
+restrictions.  YAALA's Postfix parser does not deal with the complications
+explained in \sectionref{complications}, though it does correlate log lines
+by queueid.
 
@@
 \texttt{<key>=<value>} pairs in each log line, extracts them, and
-correlates the data by queueid.  This approach will find the following
-fields: HELO hostname, queueid, \gls{SMTP} code, sender and recipient
+correlates the data by queueid.  This approach will extract the following
+data: HELO hostname, queueid, \gls{SMTP} code, sender and recipient
 addresses, and size of accepted mails.  It is unclear if the parser will
 extract any further data.  Lire misses the following fields extracted by
-\parsername{}: client and server \gls{IP} and hostname, start and end
-times, timestamps of each log line, and message ID\@.
+\parsername{}: client and server \gls{IP} address and hostname, start and
+end times, timestamps of each log line, and message ID\@.
 
@@
 \parsername{} does not produce any reports.  Lire's report contains less
-detail than Pflogsumm, and is considerable harder to configure.  Lire
-supports saving extracted data for later report generation, but accessing
-this data from another application is undocumented; given the source code
-it should be possible, with enough time and effort, to understand the
-format.  \parsername{} uses an \gls{SQL} database to make accessing the
-extracted data as easy as possible.
+detail than Pflogsumm, and it is considerable harder to configure.  Lire
+supports saving extracted data for later report generation, but the format
+of this data store is undocumented; given the source code it should be
+possible, with enough time and effort, to understand the format.
+\parsername{} uses an \gls{SQL} database to make accessing the extracted
+data as easy as possible.
 
@@
 It also counts the number of log lines parsed and skipped.  It omits client
-\gls{IP} and hostname, server \gls{IP}, HELO hostname, queueid, and message
-ID\@.  It extracts the delay for delivered mails, which \parsername{} does
-not.  Log lines are correlated based on the queueid (referred to as
-sessionname [sic] within Logrep), but not by \gls{pid}.  The parsing is
-error prone: empty fields are saved when the log line does not match the
-regex, though it appears that they will not overwrite existing data.  Most
-notably rejections are completely ignored, making it unsuitable for the
-purposes of this project.  It does not try to address any of the
-complications in \sectionref{complications} except for correlating by
-queueid.
+\gls{IP} address and hostname, server \gls{IP}, HELO hostname, queueid,
+timestamps of each log line, and message ID\@.  It extracts the delay for
+delivered mails, which \parsername{} does not.  Log lines are correlated
+based on the queueid (referred to as sessionname [sic] within Logrep), but
+not by \gls{pid}.  The parsing is error prone: empty fields are saved when
+the log line does not match the regex, though it appears that they will not
+overwrite existing data.  Most notably rejections are completely ignored,
+making it unsuitable for the purposes of this project.  It does not try to
+address any of the complications in \sectionref{complications} except for
+correlating by queueid.
 
@@
 instructions were followed as closely as possible, and multiple attempts
-were made.  As parsing failed it was not possible to review the reports
-Logrep can generate (available in HTML only), or to examine the
+were made.  Because parsing failed it was not possible to review the
+reports Logrep can generate (available in HTML only), or to examine the
 (undocumented) format in which it can save extracted data for subsequent
@@
 There are other programs available which perform basic Postfix log file
-parsing (some to a greater level of detail than others), but few attempt to
-correlate log lines by queueid (none correlate by \gls{pid}) to produce an
-overall record of the journey of each mail through Postfix.  None of the
+parsing (some to a greater level of detail than others).   None of the
 reviewed parsers collect the breadth of information gathered by
 \parsername{}, or make it as easy to extend the parser to handle new log
-lines.  Most of the parsers generate a report and immediately discard the
-data extracted from the log files; those that do not discard the data
-typically retain it in a format inaccessible to other tools.  Nearly all of
-the parsers reviewed can produce a report of greater or lesser detail and
-complexity, unlike \parsername{}.  The quality of the documentation offered
-by the subset of parsers that provide some varies from unusable to good;
-none of the parsers provide any documentation on the format of their data
-stores (if they have one).  Fewer than half of the parsers were capable of
+lines.  Some correlate log lines by queueid (none correlate by pid); none
+deal with any of the other complications described in XXX REFERENCE
+complications.  All of the reviewed parsers generate a report, and some
+provide a greater or lesser degree of customisation.  Most have a data
+store, but only \gls{LMA} provides any documentation on its format; some
+deliberately make the data store inaccessible to other tools.  Most but not
+all of the parsers provide documentation, with the quality varying from
+unusable to very good.  Fewer than half of the parsers were capable of
 parsing the \numberOFlogFILES{} test log files, and improving or extending
@@
 
-The overriding difference between \parsername{} and the other parsers
-reviewed herein is that none of them aim for the high level of
+XXX REPHRASE The overriding difference between \parsername{} and the other
+parsers reviewed herein is that none of them aim for the high level of
 understanding of Postfix log files achieved by \parsername{}.
@@
                         &               &                   & \& reports        & but poor       &              \\
-        Isoqlog         & No            & Yes               & No                & Poor           & Yes          \\
+        Isoqlog         & No            & Yes               & No                & No English     & Yes          \\
+                        &               &                   &                   & documentation  &              \\
         AWStats         & Partially     & Yes               & Partial \dag{}    & Good           & Yes          \\
@@
         Lire            & Yes           & Yes               & Yes               & Reasonable     & Yes          \\
-        Logrep          & No            & Yes               & No                & Poor           & Yes          \\
+        Logrep          & No            & Yes               & No                & None           & Yes          \\
         \parsername{}   & Yes           & Yes \nibeta{}     & No \nichi{}       & XXX            & Yes          \\

### 2009-01-14 doc/logparser-other-parsers-review.tex
Further corrections and improvements from proofreading.

@@
 \label{state of the art review}
-
-XXX DO ANY OF THE PARSERS DISTINGUISH BETWEEN DIFFERENT TYPES OF
-REJECTIONS\@?
 
@@
 
-XXX SHOULD THIS BE IN THE CONCLUSION INSTEAD\@?
+XXX SHOULD THIS LIST BE IN THE CONCLUSION INSTEAD\@?
 
@@
         support compressed files, but Sawmill exhibits a dramatic increase
-        in parsing time when doing so.  Although this is a minor
-        disparity, support for reading compressed log files is quite
-        helpful, as it dramatically reduces the disk space required to
-        store historical log files.
+        in parsing time when doing so.  Support for reading compressed log
+        files is quite helpful, as it dramatically reduces the disk space
+        required to store historical log files.
 
-    \item Some of the parsers reviewed save the extracted data to a data
-        store, but the majority discard all data once they have finished
-        generating their report, making historical analysis impossible
-        without parsing all log files every time.
+    \item Most of the reviewed parsers do not distinguish between different
+        delivery attempt rejections, so they can not be used to determine
+        the success rate of different anti-spam techniques.  The exception
+        is Pflogsumm, which provides a summary of why delivery attempts
+        were rejected.
 
@@
 \sectionref{results table}; for convenience that list is repeated here:
-server \gls{IP}, server hostname, client \gls{IP}, client hostname, HELO
-hostname, queueid, start time, end time, \gls{SMTP} code, sender,
-recipient, size, message ID\@.
+client and server \gls{IP} address and hostname, HELO hostname, queueid,
+start time, end time, \gls{SMTP} code, sender, recipient, size, message
+ID\@.
 
@@
 The decision to support both \gls{CSV} and Berkeley DB appears to have been
-a serious limitation: XXX EXTEND\@: LIMITATIONS WILL BE EXPLAINED LATER OR
-SOMETHING\@.  Very little documentation is provided with \gls{LMA}, though
-some documentation is available in~\cite{log-mail-analyser}.  Studying the
+a serious limitation: both formats have limitations that will be explained
+later.  Very little documentation is provided with \gls{LMA}, though some
+documentation is available in~\cite{log-mail-analyser}.  Studying the
 source code is informative, though this author had difficulty as the
@@
 Pflogsumm produces a useful report, and successfully parsed the
-\numberOFlogFILES{} log files it was tested with.\footnote{The results it reported
-were not verified in detail, but it did not report any errors, and has a
-very good reputation amongst Postfix users.}
-Pflogsumm has many options to include or exclude certain
-sections of the report; by default it includes the following:
+\numberOFlogFILES{} log files it was tested with.\footnote{The results it
+reported were not verified in detail, but it did not report any errors, and
+has a very good reputation amongst Postfix users.}  Pflogsumm has many
+options to include or exclude certain sections of the report, all clearly
+documented; by default it includes the following:
 
@@
 \end{itemize}
+
+Pflogsumm is the only reviewed parser which distinguishes between different
+rejections.
 
@@
         Fields extracted: from, to, server, UID, relay, status, number of
-        recipients, origin hostname, origin \gls{IP}, and virus.  It also
-        counts the number of and total size of all mails delivered.  The
-        fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
+        recipients, origin hostname, origin \gls{IP} address, and virus.
+        It also counts the number of and total size of all mails delivered.
+        The fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
         \texttt{virus} are not explained in the documentation:
@@
         \newline{} Fields extracted: from, to, RBL list, client hostname,
-        and client \gls{IP}\@.  It also counts the number and total size of
-        all mails delivered.
+        and client \gls{IP} address.  It also counts the number and total
+        size of all mails delivered.
 
@@
         \newline{} Fields extracted: from, to, client hostname, client
-        \gls{IP}, relay hostname, relay \gls{IP}, status, response code,
-        RBL list, and message id.  It also counts the number and size of
-        all mails delivered, processed, blocked, expired, and bounced.
+        \gls{IP} address, relay hostname, relay \gls{IP} address, status,
+        response code, RBL list, and message id.  It also counts the number
+        and size of all mails delivered, processed, blocked, expired, and
+        bounced.
 
@@
 supplied with thorough and well written documentation; everything the
-author searched for was documented, except the MySQL database schema.  A
-commercial version of MySQL is required due to MySQL licensing
-restrictions, but Sawmill's documentation explains why and includes
-instructions on how to compile Sawmill so that it can use a non-commercial
-version of MySQL (this was not attempted during the review process).
+author searched for was documented, except the MySQL database schema and
+some details of the data extracted by the parser.  A commercial version of
+MySQL is required due to MySQL licensing restrictions, but Sawmill's
+documentation explains why and includes instructions on how to compile
+Sawmill so that it can use a non-commercial version of MySQL (this was not
+attempted during the review process).
 
@@
 extracted, which the newly extracted data is merged into; the format of the
-data store is undocumented.  It does not utilise rejection log lines in any
-way, so is unsuitable for the purposes of this project.  Its parsing is
-completely inextensible, indeed is almost incomprehensible, relying on
-\texttt{scanf(3)}, unexplained fixed offsets, and low level string
-manipulation; it is the opposite end of the spectrum to \parsernames{}
-parsing.  It does not handle any of the complications discussed in
-\sectionref{complications}, does not gather the breadth of data required
-for this project, and ignores the majority of log lines produced by
-Postfix.
-
-XXX DOCUMENTATION\@?
+data store is undocumented.  Very little documentation is provided with
+Isoqlog, little more than installation instructions.  It does not utilise
+rejection log lines in any way, so is unsuitable for the purposes of this
+project.  Its parsing is completely inextensible, indeed is almost
+incomprehensible, relying on \texttt{scanf(3)}, unexplained fixed offsets,
+and low level string manipulation; it is the opposite end of the spectrum
+to \parsernames{} parsing.  It does not handle any of the complications
+discussed in \sectionref{complications}, does not gather the breadth of
+data required for this project, and ignores the majority of log lines
+produced by Postfix.
 
@@
 for each mail; these could easily be extracted by \parsername{} if desired.
-XXX DELAY AND DEFERRAL\@.  Unlike \parsername{}, YAALA does not maintain
-separate counters for each restriction; this rules out the possibility of
-using the collected data for optimisation, testing, or understanding of
-restrictions.  YAALA's Postfix parser does not deal with the complications
-explained in \sectionref{complications}, though it does correlate log lines
-by queueid.
+Unlike \parsername{}, YAALA does not maintain separate counters for each
+restriction; this rules out the possibility of using the collected data for
+optimisation, testing, or understanding of restrictions.  YAALA's Postfix
+parser does not deal with the complications explained in
+\sectionref{complications}, though it does correlate log lines by queueid.
 
@@
 \parsername{} uses an \gls{SQL} database to make accessing the extracted
-data as easy as possible.
+data as easy as possible.  In general Lire has poor documentation.
 
@@
 It also counts the number of log lines parsed and skipped.  It omits client
-\gls{IP} address and hostname, server \gls{IP}, HELO hostname, queueid,
-timestamps of each log line, and message ID\@.  It extracts the delay for
-delivered mails, which \parsername{} does not.  Log lines are correlated
-based on the queueid (referred to as sessionname [sic] within Logrep), but
-not by \gls{pid}.  The parsing is error prone: empty fields are saved when
-the log line does not match the regex, though it appears that they will not
-overwrite existing data.  Most notably rejections are completely ignored,
-making it unsuitable for the purposes of this project.  It does not try to
-address any of the complications in \sectionref{complications} except for
-correlating by queueid.
+\gls{IP} address and hostname, server \gls{IP} address, HELO hostname,
+queueid, timestamps of each log line, and message ID\@.  It extracts the
+delay for delivered mails, which \parsername{} does not.  Log lines are
+correlated based on the queueid (referred to as sessionname [sic] within
+Logrep), but not by \gls{pid}.  The parsing is error prone: empty fields
+are saved when the log line does not match the regex, though it appears
+that they will not overwrite existing data.  Most notably rejections are
+completely ignored, making it unsuitable for the purposes of this project.
+It does not try to address any of the complications in
+\sectionref{complications} except for correlating by queueid.
 
@@
 lines.  Some correlate log lines by queueid (none correlate by pid); none
-deal with any of the other complications described in XXX REFERENCE
-complications.  All of the reviewed parsers generate a report, and some
-provide a greater or lesser degree of customisation.  Most have a data
-store, but only \gls{LMA} provides any documentation on its format; some
-deliberately make the data store inaccessible to other tools.  Most but not
-all of the parsers provide documentation, with the quality varying from
-unusable to very good.  Fewer than half of the parsers were capable of
+deal with any of the other complications described in
+\sectionref{complications}.  All of the reviewed parsers generate a report,
+and some provide a greater or lesser degree of customisation.  Most have a
+data store, but only \gls{LMA} provides any documentation on its format;
+some deliberately make the data store inaccessible to other tools.  Most
+but not all of the parsers provide documentation, with the quality varying
+from unusable to very good.  Fewer than half of the parsers were capable of
 parsing the \numberOFlogFILES{} test log files, and improving or extending
@@
 Table \refwithpage{Summary of parsers' features} provides a summary of the
-parsers' features.
-
-XXX REPHRASE The overriding difference between \parsername{} and the other
-parsers reviewed herein is that none of them aim for the high level of
-understanding of Postfix log files achieved by \parsername{}.
+parsers' features.  The overriding difference between \parsername{} and the
+other parsers reviewed herein is that none of them aim for the high level
+of understanding of Postfix log files achieved by \parsername{}.
 
@@
         AWStats         & Partially     & Yes               & Partial \dag{}    & Good           & Yes          \\
-        Anteater        & No            & No                & No                & Poor           & Yes          \\
+        Anteater        & No            & No                & No                & None           & Yes          \\
         YAALA           & No            & Yes \ddag{}       & Searches          & Poor           & Yes          \\
+                        &               &                   & \& reports        &                &              \\
         Lire            & Yes           & Yes               & Yes               & Reasonable     & Yes          \\
         Logrep          & No            & Yes               & No                & None           & Yes          \\
-        \parsername{}   & Yes           & Yes \nibeta{}     & No \nichi{}       & XXX            & Yes          \\
+        \parsername{}   & Yes           & Yes \nibeta{}     & No \nichi{}       & \niepsilon{}   & Yes          \\
         \tablebottomline{}%
@@
 
+        \item [\niepsilon{}] \parsername{} aims to have thorough and
+            complete documentation, but the author cannot provide an
+            unbiased review.
+
     \end{eqlist}

### 2009-01-14 doc/logparser-other-parsers-review.tex
Replaced quotation environments around blurbs, they're quite obviously blurbs now.  Other improvements.

@@
 available, to accurately judge what improvements, if any, are delivered by
-the newcomers.
-
-XXX SHOULD THIS LIST BE IN THE CONCLUSION INSTEAD\@?
+the newcomers.  Each of the reviewed parsers was tested with the
+\numberOFlogFILES{} test log files described in \sectionref{parser
+efficiency}.  \Tableref{Summary of parsers' features} summarises the
+results.
 
@@
 
-Each of the reviewed parsers was tested with the \numberOFlogFILES{} test
-log files described in \sectionref{parser efficiency}.  The data extracted
-by \parsername{} is documented in \sectionref{connections table} and
-\sectionref{results table}; for convenience that list is repeated here:
-client and server \gls{IP} address and hostname, HELO hostname, queueid,
-start time, end time, \gls{SMTP} code, sender, recipient, size, message
-ID\@.
+The data extracted by \parsername{} is documented in
+\sectionref{connections table} and \sectionref{results table}; for
+convenience that list is repeated here: client and server \gls{IP} address
+and hostname, HELO hostname, queueid, start time, end time, \gls{SMTP}
+code, sender, recipient, size, message ID\@.
 
@@
         every single line it should have produced; the command above
-        suppresses duplicate sequential lines.
+        suppresses duplicate sequential lines.  DAILY-REJECT produces the
+        correct number of output lines, probably because it uses the
+        Berkeley DB files as input rather than the \gls{CSV} file.
 
@@
 
-\begin{quotation}
-
+\parserblurb{%
     pflogsumm is designed to provide an over-view of Postfix activity, with
@@
     potential trouble spots.
-
-\end{quotation}
-
-\noindent{}\url{http://jimsun.linxnet.com/postfix_contrib.html} \newline{}
-(Last checked 2008/11/23.)
+}
+{http://jimsun.linxnet.com/postfix_contrib.html}
+{2008/11/23}
 
@@
 
-\begin{quotation}
-
+\parserblurb{%
     Sawmill is a Postfix log analyzer (it also support 818 other log
@@
     Mac OS, Solaris, other UNIX, and more.
-
-\end{quotation}
-
-\noindent{}\url{http://www.thesawmill.co.uk/formats/postfix.html}
-\newline{} \url{http://www.thesawmill.co.uk/formats/postfix_ii.html}
-\newline{} \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
-\newline{} (Last checked 2008/11/23.)
+}
+{http://www.thesawmill.co.uk/formats/postfix.html}
+{2008/11/23}
 
@@
 attempt is made to correlate log lines by either queueid or pid or to deal
-with the difficulties documented in \sectionref{complications}.
+with the other difficulties documented in \sectionref{complications}.
 
@@
 
-\begin{quotation}
-
-    Splunk is IT Search.
-
-    Search and navigate IT data from applications, servers and network
-    devices in real-time. Logs, configurations, messages, traps and alerts,
-    scripts, code, metrics and more. If a machine can generate it ---
-    Splunk can eat it. It's easy to download and use and it's very
-    powerful.
-
-\end{quotation}
-
-\noindent{}\url{http://www.splunk.com/} \newline{}
-(Last checked 2008/11/23.)
+\parserblurb{%
+    Splunk is IT Search.  Search and navigate IT data from applications,
+    servers and network devices in real-time. Logs, configurations,
+    messages, traps and alerts, scripts, code, metrics and more. If a
+    machine can generate it --- Splunk can eat it. It's easy to download
+    and use and it's very powerful.
+}
+{http://www.splunk.com/}
+{2008/11/23}
 
@@
 
-\begin{quotation}
-
+\parserblurb{%
     Isoqlog is an MTA log analysis program written in C. It designed to
@@
     Top Domain, Top Users values for per day, per month and years.
-
-\end{quotation}
-
-\noindent{}\url{http://www.enderunix.org/isoqlog/} \newline{}
-(Last checked 2009/01/11.)
+}
+{http://www.enderunix.org/isoqlog/}
+{2009/01/11}
 
@@
 
-\begin{quotation}
-
+\parserblurb{%
     AWStats is a free powerful and featureful tool that generates advanced
@@
     wap, streaming servers, mail servers and some ftp servers.
-
-\end{quotation}
-
-\noindent{}\url{http://awstats.sourceforge.net/} \newline{}
+}
+{http://awstats.sourceforge.net/}
 \url{http://awstats.sourceforge.net/awstats.mail.html} \newline{}
@@
 
-\begin{quotation}
-
+\parserblurb{%
     The Anteater project is a Mail Traffic Analyser. Anteater supports
@@
     ASCII or HTML, to stdout or to files.
-
-\end{quotation}
-
-\noindent{}\url{http://anteater.drzoom.ch/} \newline{}
-(Last checked 2009/01/11.)
+}
+{http://anteater.drzoom.ch/}
+{2009/01/11}
 
@@
 
-\begin{quotation}
-
+\parserblurb{%
     yaala is a very flexible analyser for all kinds of logfiles. It uses
@@
     to format the information appropriately.
-
-\end{quotation}
-
-\noindent{}\url{http://yaala.org/} \newline{}
-(Last checked 2009/01/11.)
+}
+{http://yaala.org/}
+{2009/01/11}
 
@@
 less work than starting from scratch, as both the input and output modules
-would need to be written specially; it may be more work to implement a
+would need to be written specially; it might be more work to implement a
 parser within the constraints of YAALA\@.  YAALA supports storing
@@
 parser does not deal with the complications explained in
-\sectionref{complications}, though it does correlate log lines by queueid.
+\sectionref{complications}, except it does correlate log lines by queueid.
 
@@
 
+It might be possible to use \parsername{} as a plugin with YAALA, perhaps
+with an intermediate plugin interfacing between the two, but YAALA's data
+store is insufficient for \parsernames{} needs: \parsername{} uses two
+separate tables, whereas YAALA assumes all data will reside in one
+structure; YAALA's querying mini-language might not be able to deal with
+data in separate structures either.  This approach has not been attempted
+by the author.
+
 In summary YAALA provides a Postfix parser that tries to parse the most
@@
 
-\begin{quotation}
-
+\parserblurb{%
     As any good system administrator knows, there's a lot more to keep
@@
     for your log analysis needs.
-
-\end{quotation}
-
-\noindent{}\url{http://logreport.org/lire.html} \newline{}
-(Last checked 2009/01/11.)
+}
+{http://logreport.org/lire.html}
+{2009/01/11}
 
@@
 
-\begin{quotation}
-
+\parserblurb{%
     The email servers' reports will show you the number of deliveries and
@@
     receive or send the most emails, the relays most used, etc.
-
-\end{quotation}
-
+}
 \noindent{}Examining the source code reveals that the parser looks for
@@
 in-extensible but could easily be replaced, however that would require
-writing a parser from scratch, so would not be any easier.
+writing a parser from scratch, so would not be any easier.  \parsername{}
+could possibly be used to parse Postfix log files for Lire, but the
+difficulty may outweigh the benefits.  Lire's data store may not be
+suitable for storing the data extracted by \parsername{}, but the lack of
+documentation hinders any evaluation.  As with YAALA this approach has not
+been attempted by the author.
 
@@
 
-\begin{quotation}
-
+\parserblurb{%
     Logrep is a secure multi-platform framework for the collection,
@@
     Pix.
-
-\end{quotation}
-
-\noindent{}\url{http://www.itefix.no/i2/index.php} \newline{}
-(Last checked 2009/01/11.)
+}
+{http://www.itefix.no/i2/index.php}
+{2009/01/11}
 

### 2009-01-16 doc/logparser-architecture.tex
Some minor changes based on the SGAI paper.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of included files too, to avoid false positives.

@@
-\section{Parser Architecture}
+\chapter{Parser Architecture}
 
@@
 
-\subsection{Parser architecture and design}
+\section{Parser architecture and design}
 
@@
     \begin{tabular}[]{lll}
-        \tableline{}
+        \tabletopline{}%
         \gls{ATN}          & Architecture  & Similarity                \\
-        \tableline{}
+        \tablemiddleline{}%
         Networks        & Algorithm     & Determines the sequence 
@@
                                         classifying input.          \\
-        \tableline{}
+        \tablebottomline{}%
     \end{tabular}
@@
 
-\subsection{Rules}
+\section{Rules}
 
@@
 
-\subsubsection{Rule characteristics}
+\subsection{Rule characteristics}
 
@@
 
-\subsubsection{Overlapping rules}
+\subsection{Overlapping rules}
 
@@
         overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the input files that match more than one
-        rule.
+        there are log lines in the log files that are matched by more than
+        one rule.
 
@@
 
-\subsection{Framework}
+\section{Framework}
 
-XXX PULL CONTENT FROM THE PAPER FOR HERE\@; IT'S BETTER WRITTEN\@.
+XXX PULL CONTENT FROM THE PAPER FOR HERE\@; IT IS BETTER WRITTEN\@.
 
@@
 
-\subsection{Actions}
+\section{Actions}
 

### 2009-01-16 doc/logparser-background.tex
Some minor changes based on the SGAI paper.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of included files too, to avoid false positives.

@@
-\section{Background}
+\chapter{Background}
 
@@
 
-\subsection{Motivation}
+\section{Motivation}
 
@@
     \begin{tabular}[]{lrr}
-        \tableline{}
+        \tabletopline{}%
         Restriction & Number of hits & Percentage of hits \\
-        \tableline{}%
+        \tablemiddleline{}%
         \input{build/include-sample-sql-output}
-        \tableline{}
+        \tablebottomline{}%
     \end{tabular}
@@
 
-\subsection{Database as Application Programming Interface}
+\section{Using a Database as an Application Programming Interface}
 
@@
 Postfix log files.  Although the interface is a database schema rather than
-a set of functions provided by shared code, it is in effect quite similar
-to any other \gls{API}: it provides a stable interface for users of the
+a set of functions provided by shared code, it performs the same function
+as any other \gls{API}: it provides a stable interface for users of the
 interface, allowing the implementation to be changed without adverse
-effects.  As long as the \gls{API} is maintained, the parser can be
-improved to handle additional complications; support can be added for
-earlier or later releases of Postfix; bugs can be fixed or limitations
-removed.
+effects.  If as the \gls{API} is maintained, the parser can be improved to
+handle additional complications; support can be added for earlier or later
+releases of Postfix; bugs can be fixed or limitations removed.
 
@@
             \item New columns can be added to the tables used by the
-                program, with sufficient DEFAULT clauses or a clever
-                TRIGGER or two.\footnote{Please refer to an \gls{SQL} guide
-                for explanations of these terms, e.g.\
+                program, using DEFAULT clauses or TRIGGERS to populate
+                them.\footnote{Please refer to an \gls{SQL} guide for
+                explanations of these terms, e.g.\
                 \url{http://philip.greenspun.com/sql/}.}
 
-            \item A VIEW gives a custom arrangement of data with very
-                little effort.
+            \item A VIEW gives a custom arrangement of data with minimal
+                effort.
 
@@
 
-\subsection{Simple Mail Transfer Protocol}
+\section{Simple Mail Transfer Protocol}
 
@@
 between the sending and receiving \gls{MTA}\@.  It is a simple, human
-readable, plain text protocol, making it quite easy to test and debug
-problems with it.  Despite the simplicity of the protocol many virus and
-spam sending programs fail to implement it properly, so requiring strict
+readable, plain text protocol, simplifying testing and debugging problems
+with it.  Despite the simplicity of the protocol many virus and spam
+sending programs fail to implement it properly, so requiring strict
 adherence to the protocol specification is beneficial in protecting against
@@
 
-\subsection{Postfix}
+\section{Postfix}
 
@@
 combined, and bypassed on a per-client, per-recipient, or per-sender basis,
-allowing varying levels of stricture and permissiveness.  Postfix leverages
-simple lookup tables to support arbitrarily complicated user-defined
-sequences of restrictions and exceptions; policy servers
+allowing different levels of stricture and permissiveness.  Postfix
+leverages simple lookup tables to support arbitrarily complicated
+user-defined sequences of restrictions and exceptions; policy servers
 (\sectionref{policy servers}) provide a simple way to write new
@@
 Unfortunately this flexibility has a cost: complexity in the log files
-generated.  Although it is easy to use standard Unix text processing
+generated.  Although it is simple to use standard Unix text processing
 utilities to determine the fate of an individual email, following the
@@
 
-\subsubsection{Mixing and matching Postfix restrictions}
+\subsection{Mixing and matching Postfix restrictions}
 
@@
 
-\subsubsection{Policy servers}
+\subsection{Policy servers}
 
@@
         in \tableref{Summary of rejections}.  Greylisting is obviously
-        worth using, at least at the moment, particularly when you consider
-        Greylisting's position as the final restriction that a mail must
-        overcome:\footnote{Greylisting is the final restriction a mail must
-        overcome in the configuration used on the mail server the log files
-        were obtained from; an administrator is free to use Greylisting
-        wherever in the restriction list they feel is most appropriate for
-        their mail system.} Greylisting only takes effect for mails that
-        have passed every other restriction.  Some problems may be
-        encountered when using greylisting: some servers fail to retry;
-        legitimate mail may be delayed, particularly when coming from a
-        pool of servers.
+        worth using, at least at the moment, particularly when you take
+        into account Greylisting's position as the final restriction that a
+        mail must overcome:\footnote{Greylisting is the final restriction a
+        mail must overcome in the configuration used on the mail server the
+        log files were obtained from; an administrator is free to use
+        Greylisting wherever in the restriction list they feel is most
+        appropriate for their mail system.} Greylisting only takes effect
+        for mails that have passed every other restriction.  Some problems
+        may be encountered when using greylisting: some servers fail to
+        retry; legitimate mail may be delayed, particularly when coming
+        from a pool of servers.
 

### 2009-01-16 doc/logparser-conclusion.tex
Minor cleanups.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of included files too, to avoid false positives. Improved the check for cont

@@
-\section{Conclusion}
+\chapter{Conclusion}
 
@@
 Given the diverse nature of the log lines and the ease with which
-administrators can cause new lines to be logged (\sectionref{postfix
+administrators can cause new log lines to be logged (\sectionref{postfix
 background}), enabling users to easily extend the parser to deal with new
@@
 % XXX THIS SHOULD PROBABLY BE REMOVED\@.
-%The emergent behaviour (\sectionref{Emergent
-%behaviour}) exhibited by the rules and actions is also interesting, and is
-%discussed after the flow chart (\sectionref{flow-chart}) and explanation of
-%the paths through the parser.  This emergent behaviour greatly eases the
-%process of adding new actions (\sectionref{adding new actions}), as the
-%actions do not have to be inserted into an explicit flow of control; new
-%actions will naturally find their place in the paths through the parser.
+%The emergent behaviour (\sectionref{Emergent behaviour}) exhibited by the
+%rules and actions is also interesting, and is discussed after the flow
+%chart (\sectionref{flow-chart}) and explanation of the paths through the
+%parser.  This emergent behaviour greatly eases the process of adding new
+%actions (\sectionref{adding new actions}), as the actions do not have to
+%be inserted into an explicit flow of control; new actions will naturally
+%find their place in the paths through the parser.
 
@@
 missing is much more grueling.  Trawling through the log files, looking for
-something out of the ordinary, possibly hundreds or even thousands of lines
-away from the last mention of the queueid in question, is extremely time
-consuming and error prone.  Sometimes the task is not to spot the line that
-is unusual, but to spot that a line that is normally present is missing,
-i.e.\ to realise that one line amongst thousands is absent.  In all cases
-the evidence must be used to construct a hypothesis to explain the
-irregularities, and that hypothesis must then be tested in the parser; if
-successful, the parser must be modified to deal with the irregularities,
+something out of the ordinary, possibly hundreds or even thousands of log
+lines away from the last mention of the queueid in question, is extremely
+time consuming and error prone.  Sometimes the task is not to spot the line
+that is unusual, but to spot that a line that is normally present is
+missing, i.e.\ to realise that one line amongst thousands is absent.  In
+all cases the evidence must be used to construct a hypothesis to explain
+the irregularities, and that hypothesis must then be tested in the parser;
+if successful, the parser must be modified to deal with the irregularities,
 without adversely affecting the existing functionality.  The complications

### 2009-01-16 doc/logparser-implementation.tex
Improvements to the glossaries appearance. Minor fixes.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of included files too, to avoid fals

@@
-\section{Postfix Parser Implementation}
+\chapter{Postfix Parser Implementation}
 \label{Postfix Parser Implementation}
@@
 
-\subsection{Assumptions}
+\section{Assumptions}
 
@@
 
-\subsection{Database Schema}
+\section{Database Schema}
 \label{database schema}
@@
 
-\subsubsection{Rules table}
+\subsection{Rules table}
 
@@
 
-\subsubsection{Connections table}
+\subsection{Connections table}
 
@@
 
-\subsubsection{Results table}
+\subsection{Results table}
 
@@
 
-\subsubsection{Example rule}
+\subsection{Example rule}
 
@@
 
-\subsubsection{Creating new rules}
+\subsection{Creating new rules}
 
@@
     \item [Calculate word frequencies]  The position of words within a log
-        line is important: a word occurring in two lines does not indicate
-        similarity unless it occupies the same position within both log
-        lines.\footnote{If a variable term within a line contains spaces,
-        it will appear to the algorithm as two or more words rather than
-        one.  This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.}  The algorithm maintains a counter for each
-        \texttt{(word, word's position within the log line)} tuple,
-        incrementing it each time that word occurs in that position.
+        line is important: a word occurring in two log lines does not
+        indicate similarity unless it occupies the same position within
+        both log lines.\footnote{If a variable term within a line contains
+        spaces, it will appear to the algorithm as two or more words rather
+        than one.  This will alter the position of subsequent words, so a
+        word occurring in different positions in two log lines
+        \textit{may\/} indicate similarity, but the algorithm does not
+        attempt to deal with this possibility.}  The algorithm maintains a
+        counter for each \texttt{(word, word's position within the log
+        line)} tuple, incrementing it each time that word occurs in that
+        position.
 
@@
 
-\subsubsection{Actions in detail}
+\subsection{Actions in detail}
 
@@
         before their creation is logged} this action checks a cache of
-        recent bounce mails to avoid creating bogus bounce mails when lines
-        are logged out of order.
+        recent bounce mails to avoid creating bogus bounce mails when log
+        lines are out of order.
 
@@
         and \daemon{cleanup} as it needs to deal with out of order log
-        lines; see \sectionref{discarding cleanup lines} for details.
-        XXXXXX EXPLAIN THIS BETTER\@.
+        lines; see \sectionref{discarding cleanup log lines} for details.
+        XXX EXPLAIN THIS BETTER\@.
 
@@
 
-\subsubsection{Adding new actions}
+\subsection{Adding new actions}
 
@@
 
-\subsection{Framework}
+\section{Framework}
 
@@
 
-\subsection{Parser Flow Chart}
+\section{Parser Flow Chart}
 
@@
 
-\subsubsection{Mail enters the system}
+\subsection{Mail enters the system}
 
@@
 
-\subsubsection{Mail delivery}
+\subsection{Mail delivery}
 
@@
 
-\subsubsection{Emergent behaviour}
+\subsection{Emergent behaviour}
 
@@
 XXX DUNNO IF I SHOULD INCLUDE THIS SECTION\@; AT BEST IT NEEDS TO BE
-UPDATED, AT WORST IT'S RUBBISH\@.
+UPDATED, AT WORST IT IS RUBBISH\@.
 
@@
 
-\subsection{Complications encountered}
+\section{Complications encountered}
 
@@
 
-\subsubsection{Queueid vs pid}
+\subsection{Queueid vs pid}
 
@@
 
-\subsubsection{Connection reuse}
+\subsection{Connection reuse}
 
@@
 
-\subsubsection{Re-injected mails}
+\subsection{Re-injected mails}
 
@@
 such checks helped to identify the complications described in
-\sectionref{discarding cleanup lines} and \sectionref{pickup logging after
-cleanup}.
+\sectionref{discarding cleanup log lines} and \sectionref{pickup logging
+after cleanup}.
 
@@
 
-\subsubsection{Identifying bounce notifications}
+\subsection{Identifying bounce notifications}
 
@@
 
-\subsubsection{Aborted delivery attempts}
+\subsection{Aborted delivery attempts}
 
@@
         postfix\_action attributes equal to INFO\@?  If there are any log
-        lines after the first two they should be informational lines only.
+        lines after the first two they should be informational only.
 
@@
 
-\subsubsection{Further aborted delivery attempts}
+\subsection{Further aborted delivery attempts}
 
@@
     \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection line and a mail accepted line,
-        followed by one or more rejection lines.
+        There should be a connection log line and a mail accepted log line,
+        followed by one or more rejection log lines.
 
@@
 
-\subsubsection{Timeouts during DATA phase}
+\subsection{Timeouts during DATA phase}
 
@@
 a problem in practice, though it may be in future.  This complication is
-further complicated by the presence of out of order \daemon{cleanup} lines:
-see \sectionref{discarding cleanup lines} for details.
+further complicated by the presence of out of order \daemon{cleanup} log
+lines: see \sectionref{discarding cleanup log lines} for details.
 
@@
 
-\subsubsection{Discarding cleanup lines}
-
-\label{discarding cleanup lines}
+\subsection{Discarding cleanup log lines}
+
+\label{discarding cleanup log lines}
 
@@
 
-\subsubsection{Pickup logging after cleanup}
+\subsection{Pickup logging after cleanup}
 
@@
 a warning is generated by the \daemon{pickup} action, but if the following
-conditions are met it is assumed that the lines were out of order:
+conditions are met it is assumed that the log lines were out of order:
 
@@
     \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} lines.
+        timestamps of the \daemon{cleanup} and \daemon{pickup} log lines.
 
@@
 
-\subsubsection{Smtpd stops logging}
+\subsection{Smtpd stops logging}
 
@@
 
-\subsubsection{Out of order log lines}
+\subsection{Out of order log lines}
 
@@
 dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup lines} or \sectionref{pickup logging
-after cleanup}.  In the \numberOFlogFILES{} log files used for testing this
-occurs only five times in 60,721,709 log lines, but for parser correctness
-it must be dealt with.  The five occurrences have the same characteristics:
-the \daemon{local} log line showing delivery to a local mailbox occurs
-after the \daemon{qmgr} log line showing removal of the mail from the queue
-because delivery is completed.  This causes problems: the data in the state
-tables for the mail is not complete, so entering it into the database
-fails; a new mail is created when the \daemon{local} line is parsed and
-remains in the state tables; four warnings are issued per pair of out of
-order log lines.
-
-The COMMIT action examines the list of programs that have logged lines for
-each mail, comparing the list against a table of known-good program
-combinations.  If the mail's combination is found in the table the mail can
-be entered in the database; if the combination is not found entry must be
-postponed and the mail flagged for later entry.  The SAVE\_DATA action
-checks for that flag; if the additional log lines have caused the mail to
-reach a valid combination then entry in the database will proceed,
+mail}, \sectionref{discarding cleanup log lines} or \sectionref{pickup
+logging after cleanup}.  In the \numberOFlogFILES{} log files used for
+testing this occurs only five times in 60,721,709 log lines, but for parser
+correctness it must be dealt with.  The five occurrences have the same
+characteristics: the \daemon{local} log line showing delivery to a local
+mailbox occurs after the \daemon{qmgr} log line showing removal of the mail
+from the queue because delivery is completed.  This causes problems: the
+data in the state tables for the mail is not complete, so entering it into
+the database fails; a new mail is created when the \daemon{local} line is
+parsed and remains in the state tables; four warnings are issued per pair
+of out of order log lines.
+
+The COMMIT action examines the list of programs that have produced a log
+lines for each mail, comparing the list against a table of known-good
+program combinations.  If the mail's combination is found in the table the
+mail can be entered in the database; if the combination is not found entry
+must be postponed and the mail flagged for later entry.  The SAVE\_DATA
+action checks for that flag; if the additional log lines have caused the
+mail to reach a valid combination then entry in the database will proceed,
 otherwise it must be postponed once more.
@@
 
-\subsubsection{Yet more aborted delivery attempts}
+\subsection{Yet more aborted delivery attempts}
 
@@
 
-    \item IS THIS CORRECT NOW\@?  CAN THERE BE MORE smtpd LINES\@?  There
-        must be exactly two \daemon{smtpd} and one \daemon{cleanup} log
-        lines for the mail, with no additional log lines.
+    \item XXX IS THIS CORRECT NOW\@?  CAN THERE BE MORE smtpd LOG LINES\@?
+        There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
+        log lines for the mail, with no additional log lines.
 
@@
 
-\subsubsection{Mail deleted before delivery is attempted}
+\subsection{Mail deleted before delivery is attempted}
 
@@
 
-\subsubsection{Bounce notification mails delivered before their creation is
+\subsection{Bounce notification mails delivered before their creation is
 logged}
@@
 
-\subsubsection{Mails deleted during delivery}
+\subsection{Mails deleted during delivery}
 
@@
 
-\subsection{Limitations and possible improvements}
+\section{Limitations and possible improvements}
 
@@
 
-\subsubsection{Limitations}
+\subsection{Limitations}
 
@@
 
-\subsubsection{Possible improvements}
+\subsection{Possible improvements}
 

### 2009-01-16 doc/logparser-other-parsers-review.tex
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of included files too, to avoid false positives. Improved the check for contractions and fixed any it found. Made running make somewhat quieter, to avoid clutter that conceals more useful     output. Improved the underfull/overfull check, a

@@
 the hope of finding a suitable parser to build upon, rather than starting
-from scratch.  There are not many Postfix log file parsers available ---
-indeed it was quite difficult to find ten parsers to review for this
-project --- and the functionality offered ranges from quite basic to much
-more mature, depending on the needs of the author of the parser.  None of
-those parsers were suitable for this project, so the decision was taken to
-write a new parser.  The first parser reviewed is the only previously
-published research in this area that the author is aware of; it parses
-Postfix log files, but aim of the research is to show that presenting
-extracted data in a more easily accessible format is useful to systems
-administrators, rather than to improve anti-spam techniques..
+from scratch.  It was quite difficult to find ten parsers to review for
+this project, and the functionality offered by those parsers ranges from
+quite basic to much more mature, depending on the needs of the author of
+the parser.  None of those parsers were suitable for this project, so the
+decision was taken to write a new parser.  The first parser reviewed is the
+only previously published research in this area that the author is aware
+of; it parses Postfix log files, but aim of the research is to show that
+presenting extracted data in a more easily accessible format is useful to
+systems administrators, rather than to improve anti-spam techniques.
 
-The same ten parsers have been reviewed and compared to this project's
-finished parser, to show how much effort would have been required to fulfil
-the aims and requirements of this project.  It is important to compare and
-contrast newly developed algorithms and parsers against those already
-available, to accurately judge what improvements, if any, are delivered by
-the newcomers.  Each of the reviewed parsers was tested with the
-\numberOFlogFILES{} test log files described in \sectionref{parser
-efficiency}.  \Tableref{Summary of parsers' features} summarises the
-results.
+The same ten parsers have been reviewed and compared with \parsername{},
+this project's finished parser, to show how much effort would have been
+required to fulfil the aims and requirements of this project.  It is
+important to compare and contrast newly developed algorithms and parsers
+against those already available, to accurately judge what improvements, if
+any, are delivered by the newcomers.  Each of the reviewed parsers was
+tested with the \numberOFlogFILES{} test log files described in
+\sectionref{parser efficiency}.  \Tableref{Summary of parsers' features}
+summarises the results.
 
-There are some important differences between \parsername{} and the parsers
-reviewed here:
+Some important differences exist between \parsername{} and most or all of
+the parsers reviewed here:
 
@@
         single log line it fails to parse.  The exception is AWStats, which
-        outputs the percentage of input lines it was unable to parse, but
-        does not output the lines themselves.
+        outputs the percentage of log lines it was unable to parse, but
+        does not output the log lines themselves.
 
     \item A minor difference is that most parsers do not handle compressed
-        files; both \parsername{} and Splunk handle them transparently,
+        log files; both \parsername{} and Splunk handle them transparently,
         without user intervention; Sawmill and Lire can be configured to
-        support compressed files, but Sawmill exhibits a dramatic increase
-        in parsing time when doing so.  Support for reading compressed log
-        files is quite helpful, as it dramatically reduces the disk space
-        required to store historical log files.
+        support compressed log files, but Sawmill exhibits a dramatic
+        increase in parsing time when doing so.  Support for reading
+        compressed log files is quite helpful, as it dramatically reduces
+        the disk space required to store historical log files.
 
     \item Most of the reviewed parsers do not distinguish between different
-        delivery attempt rejections, so they can not be used to determine
+        delivery attempt rejections, so they cannot be used to determine
         the success rate of different anti-spam techniques.  The exception
@@
 a serious limitation: both formats have limitations that will be explained
-later.  Very little documentation is provided with \gls{LMA}, though some
+later.  Hardly any documentation is provided with \gls{LMA}, though some
 documentation is available in~\cite{log-mail-analyser}.  Studying the
@@
 data stored, providing multiple values for a field (e.g.\ multiple
-recipients), and adding new fields.  There is no standard mechanism to
-document the fields or the separator, unlike \gls{SQL} databases where
-every database includes a schema naming the fields and the type of data
-they store (integer, text, timestamp, etc.).  The \gls{CSV} record format
-is not documented, but the output file contains a comment giving the
-format:\newline{} \texttt{\# Timestamp|Nome Client|IP Client|IP
-Server|From|To|Status|Size} \newline{}\gls{LMA} treats lines starting with
-\texttt{\#} as comments, but not all \gls{CSV} parsers will.
+recipients), and adding new fields.  \gls{CSV} files do not have a standard
+mechanism to document the fields or the separator, unlike \gls{SQL}
+databases where every database includes a schema naming the fields and the
+type of data they store (e.g.\ integer, text, timestamp).  The \gls{CSV}
+record format is not documented, but the output file contains a comment
+giving the format:\newline{} \texttt{\# Timestamp|Nome Client|IP Client|IP
+Server|From|To|Status|Size} \newline{}\gls{LMA} treats \gls{CSV} lines
+starting with \texttt{\#} as comments, but not all \gls{CSV} parsers will.
 
@@
 re-implements \gls{SQL} foreign keys, but without the functionality offered
-by even the most basic of \gls{SQL} databases (joins, ordering, searches,
-etc.).  It also requires custom code to search on some combination of the
-above, though the authors of \gls{LMA} did provide some queries: IP-STORY,
-FROM-STORY, DAILY-EMAIL, and DAILY-REJECT\@.  Berkeley DB appears to be the
-least useful of the three output formats: it does not provide the
+by even the most basic of \gls{SQL} databases (e.g.\ joins, ordering,
+searches).  It also requires custom code to search on some combination of
+the above, though the authors of \gls{LMA} did provide some queries:
+IP-STORY, FROM-STORY, DAILY-EMAIL, and DAILY-REJECT\@.  Berkeley DB appears
+to be the least useful of the three output formats: it does not provide the
 functionality of a basic \gls{SQL} database, and unlike \gls{CSV} files it
-can not be used with standard Unix text processing tools.
+cannot be used with standard Unix text processing tools.
 
 The schema used with the MySQL database is undocumented, but at least it is
-possible to discover the schema with an existing \gls{SQL} database, unlike with
-Berkeley DB\@; all \gls{SQL} databases embed the schema into the database
-and provide commands for displaying it.  Berkeley DB does not embed a
-schema, as there is neither requirement nor benefit; it only provides
-\textbf{(key, value)} pairs, so any additional structuring of the data is
-imposed by the application, thus the application must document this
-structure.  MySQL support was not tested because there is no documentation
-on the schema required.
+possible to discover the schema with an existing \gls{SQL} database, unlike
+with Berkeley DB\@; all \gls{SQL} databases embed the schema into the
+database and provide commands for displaying it.  Berkeley DB does not
+embed a schema in its files, because there is neither requirement nor
+benefit; it only provides \textbf{(key, value)} pairs, so any additional
+structuring of the data is imposed by the application, thus the application
+must document this structure.  MySQL support was not tested because the
+schema required is not documented.
 
@@
 recipient addresses, \gls{SMTP} code, and size (for accepted mails only).
-It is unclear which time and date is stored: start time, end time, or
+It is unclear what time and date is stored: start time, end time, or
 delivery time?  Unlike \parsername{} it does not store the server hostname,
@@
 recipients.} is not explained; experimental observation shows that multiple
-records are added when there are multiple recipients (sadly the records are
+records are added when a mail had multiple recipients (sadly the records are
 not associated or linked in any way), and presumably the same approach is
-taken when there are multiple destination servers.
+taken when there were multiple destination servers.
 
@@
 (and possibly other operating systems) thus require preprocessing before
-parsing by \gls{LMA}.  Once the preprocessing has been performed on the
-\numberOFlogFILES{} test log files \gls{LMA} parses the log files without
-complaint, although it produced 32 entries in its output file for every
-rejection in the input log file; it also missed some 40\% of delivered
-mail.  Once these glaring deficiencies were discovered the author did not
-waste any more time checking the results.
+parsing by \gls{LMA}.  The \numberOFlogFILES{} preprocessed test log files
+were parsed without complaint by \gls{LMA}, although it produced 32 entries
+in the output \gls{CSV} file for every rejection in the input log file; it
+also missed some 40\% of delivered mail.  Once these glaring deficiencies
+were discovered the author did not waste any more time checking the
+results.
 
@@
 
-    \item \gls{LMA} produces 32 output lines in its \gls{CSV} file for
-        every single line it should have produced; the command above
-        suppresses duplicate sequential lines.  DAILY-REJECT produces the
-        correct number of output lines, probably because it uses the
-        Berkeley DB files as input rather than the \gls{CSV} file.
+    \item \gls{LMA} produces 32 lines in its \gls{CSV} file for every
+        single line it should have produced; the command above suppresses
+        duplicate sequential lines.  DAILY-REJECT produces the correct
+        number of output lines, probably because it uses the Berkeley DB
+        files as input rather than the \gls{CSV} file.
 
@@
 
-There are some differences between the output from DAILY-REJECT and the
-\texttt{awk} command; the author did not spend substantial time attempting
-to explain these differences.
+The output from DAILY-REJECT and the \texttt{awk} command is not exactly
+the same; the author did not spend substantial time attempting to explain
+these differences.
 
@@
 
-    \item The output from DAILY-REJECT is missing some records which are
+    \item The output from DAILY-REJECT is missing some records that are
         present in the \gls{CSV} file; this may be because it uses the
@@
 reported were not verified in detail, but it did not report any errors, and
-has a very good reputation amongst Postfix users.}  Pflogsumm has many
+has an excellent reputation amongst Postfix users.}  Pflogsumm has many
 options to include or exclude certain sections of the report, all clearly
@@
 
-Pflogsumm is the only reviewed parser which distinguishes between different
+Pflogsumm is the only reviewed parser that distinguishes between different
 rejections.
@@
 Its data extraction facilities (described later) are too limited to save
-sufficient data for the purposes of this project: although it can extract
-three different sets of data from Postfix log files, they are not
-interlinked in any way.  The documentation does not suggest that any
-attempt is made to correlate log lines by either queueid or pid or to deal
-with the other difficulties documented in \sectionref{complications}.
+enough data for the purposes of this project: although it can extract three
+different sets of data from Postfix log files, they are not interlinked in
+any way.  The documentation does not suggest that Sawmill correlates log
+lines by either queueid or pid or deals with the other difficulties
+documented in \sectionref{complications}.
 
@@
         \texttt{uid} might be the uid of the user submitting mail locally.
-        Postfix does not perform any form of virus checking (though it has
-        many options for cooperating with an external virus scanner), so
-        the \texttt{virus} field is a mystery.
+        Postfix does not do any form of virus checking (though it has many
+        options for cooperating with an external virus scanner), so the
+        \texttt{virus} field is a mystery.
 
@@
 some details of the data extracted by the parser.  A commercial version of
-MySQL is required due to MySQL licensing restrictions, but Sawmill's
+MySQL is required because of MySQL licensing restrictions, but Sawmill's
 documentation explains why and includes instructions on how to compile
@@
 Sawmill's web interface supports searching on any combination of the fields
-it extracts, and searches performed using the web interface produced
-accurate results.  The interface for searching is neither as simple to use
-nor as informative as the interface provided by Splunk.  The administrative
-interface is much easier to use than Splunk's: it took only five minutes to
-start parsing a whole directory of log files.
+it extracts, and all searches produced accurate results.  The interface for
+searching is neither as simple to use nor as informative as the interface
+provided by Splunk.  The administrative interface is much easier to use
+than Splunk's: it took only five minutes to start parsing a whole directory
+of log files.
 
@@
     messages, traps and alerts, scripts, code, metrics and more. If a
-    machine can generate it --- Splunk can eat it. It's easy to download
-    and use and it's very powerful.
+    machine can generate it --- Splunk can eat it. It\empty{}'s easy to
+    download and use and it\empty{}'s very powerful.
 }
@@
 
-XXX THE FIRST PARAGRAPH NEEDS TO BE REWRITTEN\@; PICK POINTS, THEN COMPARE
-AND CONTRAST\@.
-
 Splunk aims to index all of an organisation's log files, providing a
 centralised view capable of searching and correlating diverse log sources.
-The web interface supports complicated searches, providing statistics and
-graphs in real time, a facility not provided by \parsername{} (report
-generation has been deferred to a subsequent program).  Searches can be
-based on the fields extracted by Splunk or the full text of the log line.
-Splunk allows quite complicated searches, but does not make the raw data
-available in an accessible form.  Saved searches can be run periodically
-and the results emailed to a recipient or sent to an external program for
-further processing (though maybe without the graphs and detailed
-statistics); the author was unable to save searches, though that may have
-been due to limitations in the free version.  The database is not available
-for use by external programs, whereas \parsername{} provides the database
-and leaves it to the user to utilise it without limit or restriction.  The
-interface is optimised for interactive use rather than automated queries
-and it does not appear to be possible to write independent tools to utilise
-the Splunk database;.  Some additional Postfix reports are supposedly
-available at \url{http://www.splunkbase.com/}, but the author was unable to
-find any Postfix reports, or indeed for any other log file types: every
-category was empty, even those that the web site claimed had many reports
-available.  Many types of graphs can be generated, though most are
-variations of a bar or pie chart, except bubble and heatmap graphs.  It is
-easy to drill down through the graphs to extract a portion of the data
+The web interface provides search functionality, generating statistics and
+graphs in real time, a facility not provided by \parsername{}.  Splunk
+allows quite complicated searches, based on the fields extracted by Splunk
+(described later) or the full text of the log line, though it is not
+possible to search on partial words.  Search results are not available for
+use with other tools.  Searches can be saved for reuse; saved searches can
+be run periodically and the results emailed to a recipient or sent to an
+external program for further processing.  The author was unable to save
+searches, though that may have been because of limitations in the free
+version.  The data store is not available for use by external programs,
+whereas \parsername{} provides the database and leaves it to the user to
+utilise it without limit or restriction.  The web interface is optimised
+for interactive use rather than automated queries, and it does not appear
+to be possible to write independent tools to utilise the Splunk database.
+Some additional Postfix reports are supposedly available on
+\url{http://www.splunkbase.com/}, but the author was unable to find any
+Postfix reports, or indeed reports for any other log file types: every
+category was empty, even those that the web site claimed had numerous
+reports available.  Many types of graphs can be generated, though most are
+variations of a bar or pie chart, except the bubble and heatmap graphs.  It
+is easy to drill down through the graphs to select a portion of the data
 (e.g.\ select the hour with the largest number of events, then select a
-particular host, and finally a specific address), though it is not possible
-to search on partial words.  All searches performed using the indexed data
-returned reasonable results.
+particular host, and finally a specific sender address).  All searches
+performed using the indexed data returned reasonable results.
 
@@
 excluded.\footnote{The log files appear to have been indexed once only;
-presumably Splunk keeps track of the files it has indexed and discards
-requests to index files for a second time.  This may or may not be a useful
-feature for \parsername{}.} There did not appear to be an option to ensure
-the log files would be processed in the order they were created, though
-this may be neither necessary nor beneficial with Splunk.  Subsequently the
-author was successful in adding a single file at a time using the
-\gls{CLI}:\newline{} \tab{} \texttt{splunk add tail
+presumably Splunk keeps track of the log files it has indexed and discards
+requests to index log files for a second time.  This may or may not be a
+useful feature for \parsername{}.} There did not appear to be an option to
+ensure the log files would be processed in the order they were created,
+though this may be neither necessary nor beneficial with Splunk.
+Subsequently the author was successful in adding a single file at a time
+using the \gls{CLI}:\newline{} \tab{} \texttt{splunk add tail
 }\textit{filename\/}\newline{} A simple loop was then enough to add all the
-desired log files.  Splunk will periodically check all indexed files for
-updates unless they are manually removed from its list; this may or may not
-be useful behaviour.  Splunk did not appear to have any difficulty in
+desired log files.  Splunk will periodically check all indexed log files
+for updates unless they are manually removed from its list; this may or may
+not be useful behaviour.  Splunk did not appear to have any difficulty in
 indexing the log files, once they had been successfully added to its queue.
@@
 indexed per day to 500MB\@.  Prices were removed from the Splunk website
-during 2008; now Splunk's sales team must be contacted for a quote.
-Typical log file sizes for a small scale mail server are given in
+during 2008; now Splunk's sales team must be asked for a quote.  Typical
+log file sizes for a small scale mail server are given in
 \sectionref{parser efficiency}.
@@
 Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=value} fields from a log line; most
-importantly it makes no attempt to correlate log lines by queueid or
-\gls{pid}, or to handle any of the other myriad complications discussed in
-\sectionref{complications}.  Its source code is unavailable, so it could
-not be used as a base for this project, even if it fulfilled all other
-requirements.
+above extracting the \texttt{key=value} fields from a log line; it makes no
+attempt to correlate log lines by queueid or \gls{pid}, or to handle any of
+the other myriad complications discussed in \sectionref{complications}.
+Its source code is unavailable, so it could not be used as a base for this
+project, even if it fulfilled all other requirements.
 
@@
 files was pointless.  It does maintain a record of data previously
-extracted, which the newly extracted data is merged into; the format of the
-data store is undocumented.  Very little documentation is provided with
+extracted and the newly extracted data is merged into it; the format of the
+data store is undocumented.  Almost no documentation is provided with
 Isoqlog, little more than installation instructions.  It does not utilise
@@
 discussed in \sectionref{complications}, does not gather the breadth of
-data required for this project, and ignores the majority of log lines
-produced by Postfix.
+data required for this project, and ignores most of the log lines produced
+by Postfix.
 
@@
 }
-{http://awstats.sourceforge.net/}
-\url{http://awstats.sourceforge.net/awstats.mail.html} \newline{}
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL} \newline{}
-(Last checked 2009/01/11.)
+{http://awstats.sourceforge.net/awstats.mail.html}
+{2009/01/11}
 
@@
 file is limited in comparison to \parsername{}: time2, email, email\_r,
-host, host\_r, method, url, code, and bytesd.  There is no explanation for
-any of those fields in the documentation (in contrast \parsername{}
-documents everything it extracts), so the author could not understand what
-data is extracted, nor determine what data is missing in comparison to
-\parsername{}.  AWStats coerces Postfix log files into Apache\footnote{The
-Apache web server is the most popular HTTP server in use over the past 10
-years; more information is available at \url{http://httpd.apache.org/}.}
-format log files, for analysis by AWStats' HTTP log file parser.  The
-converting parser only deals with a small portion of the log lines
-generated by Postfix, silently skipping those it cannot deal with, and does
-not distinguish between different types of rejection; extending it to parse
-all log lines would be at least as much work as writing a new parser.  It
-does correlate log lines by queueid (not by pid), but it does not deal with
-any of the other complications described in \sectionref{complications}.
-AWStats supports saving data, but the format of the data store is not
-documented.  It also supports reading compressed log files, but that
-functionality was not tested.
+host, host\_r, method, url, code, and bytesd.  No explanation for any of
+those fields is provided in the documentation at
+\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL} (in
+contrast \parsername{} fully documents everything it extracts), so the
+author could not understand what data is extracted, or determine what data
+is missing in comparison to \parsername{}.  AWStats coerces Postfix log
+files into Apache\footnote{The Apache web server is the most popular HTTP
+server in use over the past 10 years; more information is available at
+\url{http://httpd.apache.org/}.} format log files, for analysis by AWStats'
+HTTP log file parser.  The converting parser only deals with a small
+portion of the log lines generated by Postfix, silently skipping those it
+cannot deal with, and does not distinguish between different types of
+rejection; extending it to parse all log lines would be at least as much
+work as writing a new parser.  It does correlate log lines by queueid (not
+by pid), but it does not deal with any of the other complications described
+in \sectionref{complications}.  AWStats supports saving data, but the
+format of the data store is not documented.  It also supports reading
+compressed log files, but that functionality was not tested.
 
@@
 that it parsed 9,240,075 (88.70\%) of 10,416,129 log lines, skipping
-1,176,050 (11.29\%) corrupt log lines; however there are
-\numberOFlogLINES{} lines in the \numberOFlogFILES{} log files, so AWStats
-parsed only 15.21\% of the input lines, declared 1.93\% were corrupt, and
-ignored the remaining 82.85\%.
+1,176,050 (11.29\%) corrupt log lines; the \numberOFlogFILES{} log files
+contain \numberOFlogLINES{} log lines, so AWStats parsed only 15.21\% of
+the log lines, declared 1.93\% were corrupt, and ignored the remaining
+82.85\%.
 
@@
 programs can use Storable to load, examine, and optionally modify the data;
-\parsername{} uses a well documented database which is accessible from the
-majority of programming languages.  This information was gleaned from the
-source code, as the documentation is sadly lacking.
+\parsername{} uses a well documented database that is accessible from most
+common programming languages.  This information was gleaned from the source
+code, as the documentation is sadly lacking.
 
@@
 separate tables, whereas YAALA assumes all data will reside in one
-structure; YAALA's querying mini-language might not be able to deal with
-data in separate structures either.  This approach has not been attempted
-by the author.
+structure; YAALA's querying mini-language might not deal successfully with
+data in separate structures.  This approach has not been attempted by the
+author.
 
@@
 different log file types.  The data extracted by its Postfix parser is not
-clearly documented; the manual says only:
+clearly documented:
 
-\parserblurb{%
-    The email servers' reports will show you the number of deliveries and
+\begin{quotation}
+
+    \noindent{} The email servers' reports will show you the number of deliveries and
     the volume of email delivered by day, the domains from which you
     receive or send the most emails, the relays most used, etc.
-}
+
+\end{quotation}
+
 \noindent{}Examining the source code reveals that the parser looks for
@@
 PDF, and Excel 95) but the reports do not appear to be customisable;
-\parsername{} does not produce any reports.  Lire's report contains less
-detail than Pflogsumm, and it is considerable harder to configure.  Lire
+\parsername{} does not produce any reports.  Lire's report is not as
+detailed as Pflogsumm's, and it is considerable harder to configure.  Lire
 supports saving extracted data for later report generation, but the format
@@
 
-Like AWStats and Logrep, Lire attempts to correlate log lines by queueid,
-but not by \gls{pid}, so the complete list of recipients for a mail should
-be available; however its parser extracts only part of the available data
-and makes no attempt to deal with the other complications described in
+Similar to AWStats and Logrep, Lire attempts to correlate log lines by
+queueid, but not by \gls{pid}, so the complete list of recipients for a
+mail should be available; its parser extracts only part of the available
+data and makes no attempt to deal with the other complications described in
 \sectionref{complications}.  When testing Lire on the \numberOFlogFILES{}
@@
 
-    \item Which pair of correspondents exchanged the highest number of
+    \item The pair of correspondents exchanged the highest number of
         emails.
@@
 extracted data easily available to other programs.  Its parser is
-in-extensible but could easily be replaced, however that would require
-writing a parser from scratch, so would not be any easier.  \parsername{}
-could possibly be used to parse Postfix log files for Lire, but the
-difficulty may outweigh the benefits.  Lire's data store may not be
-suitable for storing the data extracted by \parsername{}, but the lack of
-documentation hinders any evaluation.  As with YAALA this approach has not
-been attempted by the author.
+in-extensible but could easily be replaced, but that would require writing
+a parser from scratch, so would not be any less work.  \parsername{} could
+possibly be used to parse Postfix log files for Lire, but the difficulty
+may outweigh the benefits.  Lire's data store may not be suitable for
+storing the data extracted by \parsername{}, but the lack of documentation
+hinders any evaluation.  As with YAALA this approach has not been attempted
+by the author.
 
@@
 
-Logrep extracts less than half the fields \parsername{} does:
+Logrep extracts fewer than half the fields \parsername{} does:
 
@@
 delay for delivered mails, which \parsername{} does not.  Log lines are
-correlated based on the queueid (referred to as sessionname [sic] within
-Logrep), but not by \gls{pid}.  The parsing is error prone: empty fields
-are saved when the log line does not match the regex, though it appears
-that they will not overwrite existing data.  Most notably rejections are
-completely ignored, making it unsuitable for the purposes of this project.
-It does not try to address any of the complications in
-\sectionref{complications} except for correlating by queueid.
+correlated based on the queueid (called sessionname [sic] within Logrep),
+but not by \gls{pid}.  The parsing is error prone: empty fields are saved
+when the log line does not match the regex, though it appears that they
+will not overwrite existing data.  Most notably rejections are completely
+ignored, making it unsuitable for the purposes of this project.  It does
+not try to address any of the complications in \sectionref{complications}
+except for correlating by queueid.
 
@@
 reports Logrep can generate (available in HTML only), or to examine the
-(undocumented) format in which it can save extracted data for subsequent
-reuse.
+(undocumented) format it uses to save extracted data for subsequent reuse.
 
@@
 
-There are other programs available which perform basic Postfix log file
+This chapter has reviewed ten programs that perform basic Postfix log file
 parsing (some to a greater level of detail than others).   None of the
@@
 some deliberately make the data store inaccessible to other tools.  Most
-but not all of the parsers provide documentation, with the quality varying
+but not all of the parsers provide documentation, with the quality ranging
 from unusable to very good.  Fewer than half of the parsers were capable of
-parsing the \numberOFlogFILES{} test log files, and improving or extending
-parsing would have been quite a difficult task for any of the parsers.
-Table \refwithpage{Summary of parsers' features} provides a summary of the
-parsers' features.  The overriding difference between \parsername{} and the
-other parsers reviewed herein is that none of them aim for the high level
-of understanding of Postfix log files achieved by \parsername{}.
+parsing the \numberOFlogFILES{} test log files; improving or extending
+parsing would have been quite a difficult task for any of the parsers, and
+one that the author did not have the time to attempt.  \Tableref{Summary of
+parsers' features} provides a summary of the parsers' features.  The
+overriding difference between \parsername{} and the other parsers reviewed
+herein is that none of them aim for the high level of understanding of
+Postfix log files achieved by \parsername{}.
 
@@
         \item [\dag{}] Sections can be omitted from a report, but extra
-            sections can not be added.
+            sections cannot be added.
 

### 2009-01-16 doc/logparser-results.tex
Improvements after meeting Carl.
Converted from article to report, and sections to chapters. Some cleanups. @ doesn't need escaping in glossary entries with new glossaries package.
Further corrections and improvements from proofreading.
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files' or 'lines', and added a make target to help with it.  This is     supported by paragraphs-to-lines, which will hopefully find other uses.  I     renamed a couple of

@@
-\section{Results}
+\chapter{Results}
 
@@
 
+XXX COMPARE BEST AND WORST POSSIBLE RESULTS WITH REAL RESULTS\@.
+
 XXX REUSE THE CONTENT FROM THE PAPER FOR THIS SECTION\@.
 
-\subsection{Parser Efficiency}
+\section{Parser Efficiency}
 
@@
 
-\subsubsection{Architecture Scalability: Input Size}
+\clearpage{}
+
+\subsection{Architecture Scalability: Input Size}
 
@@
 input size: does it scale linearly, polynomially, or exponentially?
-\Graphref{parsing time vs file size vs number of lines graph} shows the
+\Graphref{parsing time vs file size vs number of log lines graph} shows the
 parsing time in seconds, file size in MB, and tens of thousands of log
@@
 linearly with input size.  This impression is borne out by
-\graphref{parsing time vs file size vs number lines factor} which plots
-both the ratio of file size vs parsing time, and the ratio of number of log
-lines vs parsing time (higher is better in both cases); \tableref{parsing
-time vs file size vs number lines factor table} shows the same ratios for
-different groups of log files.  The ratios are quite tightly banded,
-providing empirical evidence that the algorithm scales linearly.  The ratio
-increases (i.e.\ improves) for log files 22 and 62--68 despite their larger
-than average size (shown in \graphref{parsing time vs file size vs number
-of lines graph}); this is explained fully in \sectionref{Explaining the
-peaks in log file size}.
+\graphref{parsing time vs file size vs number of log lines factor} which
+plots both the ratio of file size vs parsing time, and the ratio of number
+of log lines vs parsing time (higher is better in both cases);
+\tableref{parsing time vs file size vs number of log lines factor table}
+shows the same ratios for different groups of log files.  The ratios are
+quite tightly banded, providing empirical evidence that the algorithm
+scales linearly.  The ratio increases (i.e.\ improves) for log files 22 and
+62--68 despite their larger than average size (shown in \graphref{parsing
+time vs file size vs number of log lines graph}); this is explained fully
+in \sectionref{Explaining the peaks in log file size}.
 
-\showtable{build/include-file-size-and-number-of-lines-vs-parsing-time}{Ratio
-of file size and number of lines to parsing time}{parsing time vs file size
-vs number lines factor table}
+\showtable{build/include-file-size-and-number-of-log-lines-vs-parsing-time}{Ratio
+of file size and number of log lines to parsing time}{parsing time vs file
+size vs number of log lines factor table}
 
 \showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
-and number of lines}{parsing time vs file size vs number of lines graph}
+and number of log lines}{parsing time vs file size vs number of log lines
+graph}
 
 \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of file size
-and number of lines to parsing time}{parsing time vs file size vs number
-lines factor}
+and number of log lines to parsing time}{parsing time vs file size vs
+number of log lines factor}
 
-\subsubsection{Explaining the peaks in log file size}
+\subsection{Explaining the peaks in log file size}
 
@@
 
-\Graphref{parsing time vs file size vs number of lines graph} shows log
+\Graphref{parsing time vs file size vs number of log lines graph} shows log
 file size varying between 30--100MB, except for log files 22 (200MB) and
@@
 
-XXX REWRITE THIS SECTION SO THAT IT'S CORRECT\@: POINT 4 SAYS THAT THE
+XXX REWRITE THIS SECTION SO THAT IT IS CORRECT\@: POINT 4 SAYS THAT THE
 ENVELOPE SENDER CHANGES, IS THAT CORRECT\@?
@@
 
-\subsubsection{Rule ordering for efficiency}
+\subsection{Rule ordering for efficiency}
 
@@
 parse a log line from log files 22 or 62--68 being considerably lower than
-usual.  The ratio of number of lines versus parsing time in
-\graphref{parsing time vs file size vs number lines factor} shows a
+usual.  The ratio of number of log lines versus parsing time in
+\graphref{parsing time vs file size vs number of log lines factor} shows a
 noticeable improvement at log files 22 and 62--68, where a higher ratio
@@
 
-\subsubsection{Scalability as the number of rules rises}
+\subsection{Scalability as the number of rules rises}
 
@@
 
-\subsubsection{Caching compiled regexs}
+\newpage{} % XXX
+\subsection{Caching compiled regexs}
 
@@
 to the spikes in log file size in \graphref{parsing time vs file size vs
-number of lines graph}.  The distribution of log lines across rules when
-there is a mail loop is much different to the norm, and the average number
-of rules consulted per log line is much lower: this results in far fewer
-regex compilations per line when there is a mail loop, and a
+number of log lines graph}.  The distribution of log lines across rules
+when there is a mail loop is much different to the norm, and the average
+number of rules consulted per log line is much lower: this results in far
+fewer regex compilations per line when there is a mail loop, and a
 corresponding decrease in the average parsing time for a single log line.
@@
 important when the log file contents were created by a mail loop.  The
-increases in parsing time when not caching compiled regexes for
-different groups of log files are summarised in \tableref{Increase in
-parsing time when not caching compiled regexes table}.
+increases in parsing time when not caching compiled regexes for different
+groups of log files are summarised in \tableref{Increase in parsing time
+when not caching compiled regexes table}.
 
@@
 
-\subsubsection{Summary}
+\subsection{Summary}
 
@@
 
-\subsection{Coverage}
+\section{Coverage}
 
@@
 
-\subsubsection{Log lines covered}
+\subsection{Log lines covered}
 
@@
 distribution --- it is obvious that a small number of rules match the vast
-majority of the lines, and more than half the rules match fewer than 100
-times.
+majority of the log lines, and more than half the rules match fewer than
+100 times.
 
@@
 
-\subsubsection{Mails covered}
+\subsection{Mails covered}
 
@@
 Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 lines in the test log files is infeasible, verifying a subset of
-those log files is a tractable, if extremely time consuming, task.  A
-sample of log lines was obtained by randomly selecting a log file:
+60,721,709 log lines in the test log files is infeasible, verifying a
+subset of those log files is a tractable, if extremely time consuming,
+task.  A sample of log lines was obtained by randomly selecting a log file:
 
@@
 
-The first 6000 lines of this log file (roughly 0.01\% of the total number
-of log lines used in testing) was extracted:
+The first 6000 log lines of this log file (roughly 0.01\% of the total
+number of log lines used in testing) was extracted:
 
@@
 conjunction with the input log file and a dump of the resulting database,
-verifying that for each of the input lines the parser used the correct rule
-and executed the correct action, that in turn produced the correct result
+verifying that for each of the log lines the parser used the correct rule
+and executed the correct action, which in turn produced the correct result
 and inserted the correct data in the database.  The log file segment

### 2009-01-16 doc/logparser.tex
Cleanups and improvements to the review of other parsers.
Proofread and corrected the introduction and LMA section of the state of the art review.
Further corrections and improvements after proofreading.
Further corrections after proofreading.
Further corrections and improvements from proofreading.
Replaced quotation environments around blurbs, they're quite obviously blurbs now.  Other improvements.
Lots of minor improvements and cleanups to the State of the Art Review,     including running diction on it. Checked that I've written 'log files' and 'log lines' everywhere, instead of     'files

@@
 % $Id$
+% XXX WOULD A SMALLER FONT BE BETTER?
 \documentclass[a4paper,12pt,draft]{report}
+\pagestyle{headings}
 
@@
 % \usepackage{refcheck}
+
+\newcommand{\parserblurb}[3]{%
+    \begin{quotation}
+        \noindent{}\textit{#1\/}
+
+        \noindent{}\url{#2}\newline
+        Last checked #3.
+    \end{quotation}
+}
 
@@
 \newcommand{\parsername}{\gls{PLP}}
-% XXX SHOULD THIS BE \glspl{PLP} or \gls{PLP}'s ??
-\newcommand{\parsernames}{\glspl{PLP}}
+\newcommand{\parsernames}{\gls{PLP}'s}
 
@@
 
-XXX WEKA FOR DECISION TREE DISPLAY\@.
-
 XXX GENERATE RESULT\_COLS AND CONNECTION\_COLS AUTOMATICALLY\@.
@@
 
-XXX CHECK YEARS FOR BIBLIOGRAPHY ENTRIES\@.
+XXX CHECK YEARS FOR BIBLIOGRAPHY ENTRIES\@.  WHAT DO I MEAN BY THIS\@?
 
@@
 
+XXX MAYBE DO SOMETHING ABOUT underfull hboxes IN BIBLIOGRAPHY WHEN URLS
+WRAP\@?
+
 XXX INVESTIGATE Doug Lenat.
+
+XXX EXTRACT DELAY AND DEFER WHEN AUTOMATIC EXTRACTION HAS BEEN FINISHED,
+THEN UPDATE FIELDS IN BOTH DATABASE AND REVIEW OF OTHER PARSERS\@.
+
+XXX POSSIBLY PUBLISH THE STATE OF THE ART REVIEW IN A JOURNAL\@?
 

### 2009-02-09 logparser/retest.pl
Test script used when debugging regular expressions.

@@ -1,0 +1,29 @@
+#!/usr/bin/env perl
+# vim: set textwidth=10000 :
+
+# $Id$
+
+use strict;
+use warnings;
+
+use lib q{..};
+use Data::Dumper;
+use ASO::Parser;
+
+my $parser = ASO::Parser->new({
+    data_source => q{dbi:SQLite:dbname=../sql/db.sq3}
+});
+
+my $line = q{NOQUEUE: reject_warning: RCPT from scan14.cs.tcd.ie[134.226.54.40]: 550 5.7.1 <134.226.54.40>: Helo command rejected: You are not in my network.  Go away.; from=<Niaz> to=<notredamean@gmail.com> proto=SMTP helo=<134.226.54.40>};
+my $regex = q{^__RESTRICTION_START__ <__HELO__>: Helo command rejected: You are not in my network. +Go away.; from=<(__SENDER__)> to=<(__RECIPIENT__)> proto=E?SMTP helo=<(__HELO__)>$};
+my $regex_filtered = ASO::Parser::filter_regex($parser, $regex);
+warn $regex_filtered, qq{\n};
+if ($line =~ m/$regex_filtered/) {
+    warn qq{success\n};
+    warn qq{\$`: $`\n};
+    warn qq{\$&: $&\n};
+    warn qq{\$': $'\n};
+    warn q{%+: }, Dumper(\%+);
+} else {
+    warn qq{failure\n};
+}

### 2009-02-10 logparser/retest.pl
New or improved rules to parse the remaining unparsed log lines. Small improvements to retest.pl.

@@ -14,15 +14,16 @@ my $parser = ASO::Parser->new({
     data_source => q{dbi:SQLite:dbname=../sql/db.sq3}
 });
 
-my $line = q{NOQUEUE: reject_warning: RCPT from scan14.cs.tcd.ie[134.226.54.40]: 550 5.7.1 <134.226.54.40>: Helo command rejected: You are not in my network.  Go away.; from=<Niaz> to=<notredamean@gmail.com> proto=SMTP helo=<134.226.54.40>};
-my $regex = q{^__RESTRICTION_START__ <__HELO__>: Helo command rejected: You are not in my network. +Go away.; from=<(__SENDER__)> to=<(__RECIPIENT__)> proto=E?SMTP helo=<(__HELO__)>$};
+my $line = q{143C0F3E27: to=<arash@dcs.qmul.ac.uk>, relay=mail.dcs.qmul.ac.uk[138.37.95.139]:25, delay=4.4, delays=0.05/0.02/3.3/1.1, dsn=4.0.0, status=deferred (host mail.dcs.qmul.ac.uk[138.37.95.139] said: 451-response to "RCPT TO:<fg07participants-bounces+arash=dcs.qmul.ac.uk@cs.tcd.ie>" from smtp.cs.tcd.ie [134.226.32.56] was: 450 4.2.0 <fg07participants-bounces+arash=dcs.qmul.ac.uk@cs.tcd.ie>: Recipient address rejected: Greylisted, see http://postgrey.schweikert.ch/help/cs.tcd.ie.html 451-Could not complete sender verify callout for 451-<fg07participants-bounces+arash=dcs.qmul.ac.uk@cs.tcd.ie>. 451-The mail server(s) for the domain may be temporarily unreachable, or 451-they may be permanently unreachable from this server. In the latter case, 451-you need to change the address or create an MX record for its domain 451-if it is supposed to be generally accessible from the Internet. 451 Talk to your mail administrator for details. (in reply to RCP};
[…]

### 2009-02-10 logparser/retest.pl
Further improvements to retest.pl.

@@ -8,23 +8,36 @@ use warnings;
 
 use lib q{..};
 use Data::Dumper;
+use feature qw(say);
 use ASO::Parser;
 
 my $parser = ASO::Parser->new({
     data_source => q{dbi:SQLite:dbname=../sql/db.sq3}
 });
 
-my $line = q{143C0F3E27: to=<arash@dcs.qmul.ac.uk>, relay=mail.dcs.qmul.ac.uk[138.37.95.139]:25, delay=4.4, delays=0.05/0.02/3.3/1.1, dsn=4.0.0, status=deferred (host mail.dcs.qmul.ac.uk[138.37.95.139] said: 451-response to "RCPT TO:<fg07participants-bounces+arash=dcs.qmul.ac.uk@cs.tcd.ie>" from smtp.cs.tcd.ie [134.226.32.56] was: 450 4.2.0 <fg07participants-bounces+arash=dcs.qmul.ac.uk@cs.tcd.ie>: Recipient address rejected: Greylisted, see http://postgrey.schweikert.ch/help/cs.tcd.ie.html 451-Could not complete sender verify callout for 451-<fg07participants-bounces+arash=dcs.qmul.ac.uk@cs.tcd.ie>. 451-The mail server(s) for the domain may be temporarily unreachable, or 451-they may be permanently unreachable from this server. In the latter case, 451-you need to change the address or create an MX record for its domain 451-if it is supposed to be generally accessible from the Internet. 451 Talk to your mail administrator for details. (in reply to RCP};
-my $regex = q{^(__QUEUEID__): to=<(__RECIPIENT__)>,(?: orig_to=<(__RECIPIENT__)>,)? relay=(__CLIENT_HOSTNAME__)\[(__CLIENT_IP__)\](?::\d+)?, (?:__CONN_USE__)?__DELAY__(?:__DELAYS__)?(?:dsn=__DSN__,\s)?status=deferred \(host (__CLIENT_HOSTNAME__)\[(__CLIENT_IP__)\] said: (__SMTP_CODE__)(__DATA__.*451.Could not complete sender verify callout for .*)$};
-my $regex_filtered = ASO::Parser::filter_regex($parser, $regex);
-warn $line, qq{\n};
-warn $regex_filtered, qq{\n};
+my ($left_arrow, $right_arrow) = qw(-->> <<--);
+
[…]

### 2009-02-10 doc/logparser-conclusion.tex
Dealt with some XXX items.

@@
 the rules and actions separate until the parser runs.  The actions and
-framework are not as completely separated, as the actions depend on
+framework are not as completely separated, because the actions depend on
 services provided by the framework; however the actions and framework are
 not tightly integrated: it would be possible, with some work, to completely
-separate the two.  XXX I SHOULD DO THAT\@.
-
-% XXX THIS SHOULD PROBABLY BE REMOVED\@.
-%The emergent behaviour (\sectionref{Emergent behaviour}) exhibited by the
-%rules and actions is also interesting, and is discussed after the flow
-%chart (\sectionref{flow-chart}) and explanation of the paths through the
-%parser.  This emergent behaviour greatly eases the process of adding new
-%actions (\sectionref{adding new actions}), as the actions do not have to
-%be inserted into an explicit flow of control; new actions will naturally
-%find their place in the paths through the parser.
+separate the two.
 

### 2009-02-10 doc/logparser-implementation.tex
Dealt with some XXX items.

@@
 
-    \item [regex] XXX THIS NEEDS TO BE UPDATED ONCE AUTOMATIC DATA
-        EXTRACTION IS FINISHED\@.  The regex to match the log line
-        against.  The regex will first have several keywords expanded:
-        this simplifies reading and writing rules; avoids needless
-        repetition of complex regex components; allows the components to
-        be corrected and/or improved in one location; and makes each
-        regex largely self-documenting.  For efficiency the keywords are
-        expanded and every rule's regex is compiled before attempting to
-        parse the log file --- otherwise each regex would be recompiled
-        each time it was used, resulting in a large, data dependent
-        slowdown.  Efficiency concerns are discussed in \sectionref{parser
-        efficiency}, with caching compiled regexes covered in
-        \sectionref{Caching compiled regexes}.
-
-    \item [result\_data] Sometimes rules need to save a result that is not
-        present in the log line: e.g.\ setting \texttt{smtp\_code} when
-        mail is accepted.  The format is: \newline{} \tab{} \texttt{XXX
-        UPDATE EXAMPLE\@; smtp\_code = 550; \newline{} recipient = 2,
-        sender = 4;} \newline{} i.e.\ semi-colon or comma separated
+    \item [regex] The regex to match the log line against.  The regex will
+        have several keywords expanded when the rules are loaded: this
+        simplifies reading and writing rules; avoids needless repetition of
+        complex regex components; allows the components to be corrected
+        and/or improved in one location; enables automatic extraction and
+        saving of data; and makes each regex largely self-documenting.  For
+        efficiency the keywords are expanded and every rule's regex is
+        compiled before attempting to parse the log file --- otherwise each
+        regex would be recompiled each time it was used, resulting in a
+        large, data dependent slowdown.  Efficiency concerns are discussed
+        in \sectionref{parser efficiency}, with caching compiled regexes
+        covered in \sectionref{Caching compiled regexes}.
+
+    \item [connection\_data] Sometimes rules need to save data that is not
+        present in the log line: e.g.\ setting \texttt{client\_ip} when a
+        mail is being delivered to another server.  The format is:
+        \newline{} \tab{} \texttt{ client\_hostname = localhost, client\_ip
+        = 127.0.0.1} \newline{} i.e.\ semi-colon or comma separated
         assignment statements, with the variable name on the left and the
@@
         amounts in any one rule, so dealing with escape sequences was
-        deemed unnecessary.}  Any field in the results table can be set in
-        this way; additionally \texttt{child} and \texttt{pid} are accepted
-        and used respectively by the \texttt{TRACK}, \texttt{BOUNCE}, and
-        \texttt{SMTPD\_DIED} actions, but do not cause data to be entered
-        in the database.
-
-    \item [connection\_data] The connections table equivalent of
-        \texttt{result\_data}.
+        deemed unnecessary.}  Any field in the connections table can be set
+        in this way.
+
+    \item [result\_data] The result table equivalent of
+        \texttt{connection\_data}.
 
@@
 
-\label{list of postfix actions}
-
-\begin{description}
-
-    \item [ACCEPTED] Postfix has accepted a mail, and will
-        subsequently attempt to deliver it.
-
-    \item [BOUNCED] The mail has bounced, because of a mail loop,
-        delivery failure, or five day timeout.\footnote{As with
-        much of Postfix's behaviour, five days is the default value
-        but can be changed by the administrator if they choose.}
-
-    \item [DELETED] The mail was deleted from the queue by an
-        administrator.
-
-    \item [DISCARDED] Postfix discarded the mail it was in the
-        process of accepting, because it was either larger than the
-        size limit set by the administrator, or the client timed
-        out or disconnected.
-
-    \item [EXPIRED] The mail has been in the queue for five
-        days\footnotemark[\value{footnote}] XXX CHECK IF THIS
-        FOOTNOTE WORKS without successful delivery.  A bounce mail
-        will be generated and sent to the sender address.
-
-    \item [INFO] Represents an unspecified intermediate action that
-        the parser is not interested in per se, but that does log
-        useful information, supplementing other log lines.
-
-    \item [IGNORED] An action that is not only uninteresting in
-        itself, but that also provides no useful data.
-
-    \item [POSTFIX\_RELOAD] The administrator has instructed
-        Postfix to start or stop, and all \daemon{smtpd} processes
-        will be terminated.  This does not negatively affect the
-        log files or mail queued by Postfix for delivery.
-
-    \item [PROCESSING] \daemon{cleanup} is processing a mail.
-        XXX ADD MORE DETAIL HERE
-
-    \item [REJECTED] Postfix rejected a command from the remote
-        client, causing at least one recipient to be rejected.
-
-    \item [SENT] Postfix has successfully sent a mail.
-
-\end{description}
-
 \subsection{Connections table}
@@
 program             & \daemon{smtpd}                                    \\
-XXX                 & UPDATE REGEX FOR AUTOMATIC SAVING                 \\
 regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
                     & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\5> to=<(__RECIPIENT__)> !            \\
+                    & \verb!from=<\k<sender>> to=<(__RECIPIENT__)> !    \\
                     & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
@@
 
-    \item [matches] The fields in the line captured by the rule's regex.
+    \item [matches] The data extracted from the line by the rule's regex.
 
@@
 action}{Distribution of rules per action}
+
+\subsection{Postfix actions}
+
+\label{list of postfix actions}
+
+XXX WHY DO I NEED HAVE postfix\_actions AT ALL\@?
+
+THERE IS NEARLY A 1--1 MAPPING BETWEEN action AND postfix\_action.
+
+\begin{tabular}{lll}
+
+1 & BOUNCE & BOUNCED \\
+1 & CLONE & ACCEPTED \\
+3 & COMMIT & DISCARDED \\
+1 & COMMIT & INFO \\
+1 & CONNECT & INFO \\
+1 & DELETE & DELETED \\
+64 & DELIVERY\_REJECTED & REJECTED \\
+1 & DISCONNECT & INFO \\
+1 & EXPIRY & EXPIRED \\
+1 & MAIL\_QUEUED & INFO \\
+1 & MAIL\_QUEUED & PROCESSING \\
+1 & MAIL\_TOO\_LARGE & DISCARDED \\
+1 & PICKUP & INFO \\
+2 & POSTFIX\_RELOAD & POSTFIX\_RELOAD \\
+13 & SAVE\_DATA & BOUNCED \\
+38 & SAVE\_DATA & INFO \\
+1 & SAVE\_DATA & REJECTED \\
+5 & SAVE\_DATA & SENT \\
+2 & SMTPD\_DIED & IGNORED \\
+1 & SMTPD\_WATCHDOG & IGNORED \\
+2 & TIMEOUT & DISCARDED \\
+1 & TRACK & SENT \\
+42 & UNINTERESTING & IGNORED \\
+
+\end{tabular}
+
+postfix\_action IS USED IN A FEW ACTIONS, BUT IT COULD PROBABLY BE EASILY
+REMOVED, OR MAYBE I SHOULD KEEP TRACK OF actions INSTEAD\@?
+
+XXX IF I KEEP THESE I NEED TO EXPLAIN WHY I HAVE THEM\@.
+
+\begin{description}
+
+    \item [ACCEPTED] Postfix has accepted a mail, and will
+        subsequently attempt to deliver it.
+
+    \item [BOUNCED] The mail has bounced, because of a mail loop,
+        delivery failure, or five day timeout.\footnote{As with
+        much of Postfix's behaviour, five days is the default value
+        but can be changed by the administrator if they choose.}
+
+    \item [DELETED] The mail was deleted from the queue by an
+        administrator.
+
+    \item [DISCARDED] Postfix discarded the mail it was in the
+        process of accepting, because it was either larger than the
+        size limit set by the administrator, or the client timed
+        out or disconnected.
+
+    \item [EXPIRED] The mail has been in the queue for five days without
+        successful delivery.  A bounce mail will be generated and sent to
+        the sender address.
+
+    \item [INFO] Represents an unspecified intermediate action that
+        the parser is not interested in per se, but that does log
+        useful information, supplementing other log lines.
+
+    \item [IGNORED] An action that is not only uninteresting in
+        itself, but that also provides no useful data.
+
+    \item [POSTFIX\_RELOAD] The administrator has instructed
+        Postfix to start or stop, and all \daemon{smtpd} processes
+        will be terminated.  This does not negatively affect the
+        log files or mail queued by Postfix for delivery.
+
+    \item [PROCESSING] \daemon{cleanup} is processing a mail.
+        XXX ADD MORE DETAIL HERE
+
+    \item [REJECTED] Postfix rejected a command from the remote
+        client, causing at least one recipient to be rejected.
+
+    \item [SENT] Postfix has successfully sent a mail.
+
+\end{description}
 
@@
 
-\Figureref{flow chart image} shows the paths the data representing a mail
-can take through the parser algorithm.  The flow chart covers the most
-common paths only, the complications described in
-\sectionref{complications} are excluded for the sake of clarity.  XXX WHERE
-ARE THE PATHS DESCRIBED\@?
+\Figureref{flow chart image} shows the most common paths the data
+representing a mail can take through the parser; the complications
+described in \sectionref{complications} are excluded for the sake of
+clarity.
 
@@
 
-\clearpage
-
-\subsection{Mail enters the system}
-
 \label{mail-enters-the-system}
-
-XXX SHOULD THESE BE SEPARATE SUBSUBSECTIONS\@?
 
@@
 simplest case: a queueid is assigned immediately and the sender address is
-logged (action: PICKUP\@; flowchart:~2).
-
-\gls{SMTP} is more complicated:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \gls{IP} address or hostname, HELO hostname,
-        etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
-        more mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \gls{SMTP} commands the data will be saved
-        to the database; if not there will not be any data to save (any
-        mails accepted will already have been cloned so their data is in
-        another data structure).
-
-    \item If one or more mails were accepted there will be more log lines
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-Re-injection due to forwarding sadly lacks explicit log lines of its
-own; it is explained in \sectionref{Re-injected mails}.
+logged (action: PICKUP\@; flowchart:~2).  Re-injection due to forwarding
+sadly lacks explicit log lines of its own; it is explained in
+\sectionref{Re-injected mails}.
 
@@
 
-\subsection{Mail delivery}
+\gls{SMTP} is more complicated than local submission:
+
+\begin{enumerate}
+
+    \item First there is a connection from the remote client (action:
+        CONNECT\@; flowchart:~1).
+
+    \item This is followed by rejection of sender address, recipient
+        addresses, client \gls{IP} address or hostname, HELO hostname,
+        etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
+        more mails (action: CLONE\@; flowchart:~5); or some interleaving of
+        both.
+
+    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
+        Postfix has rejected any \gls{SMTP} commands the data will be saved
+        to the database; if not there will not be any data to save (any
+        mails accepted will already have been cloned so their data is in
+        another data structure).
+
+    \item If one or more mails were accepted there will be more log lines
+        for those mails later, see \sectionref{mail-delivery}.
+
+\end{enumerate}
 
@@
 mail}.
-
-\subsection{Emergent behaviour}
-
-\label{Emergent behaviour}
-
-XXX DUNNO IF I SHOULD INCLUDE THIS SECTION\@; AT BEST IT NEEDS TO BE
-UPDATED, AT WORST IT IS RUBBISH\@.
-
-The rules and actions exhibit emergent behaviour that is far more
-complicated than the behaviour explicitly encoded in the algorithm.  The
-top-level parser could be written in pseudo-code like so (with indentation
-level indicating flow control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-Yet despite the brevity of the algorithm above the (simplified) flow chart
-in \sectionref{flow-chart} contains:
-
-\begin{itemize}
-
-    \item Twenty one states (not including explicit branches).
-
-    \item Three entry points (1, 2, 3).
-
-    \item Five exit points (8, 14, 17, 20, 25).
-
-    \item Five loops (4--4, 4--5--4, 5--5, 9--9, 9--10--9).
-
-    \item Four explicit branches (7, 13, 15, 18); these are decisions taken
-        by an action, determining what will happen next to a mail.
-
-    \item Four implicit branches, where the transition is determined by the
-        next log line that is processed for that mail, e.g.\ state 1 can
-        transition to either state 4 or state 5.  (1--\{4,5\},
-        4--\{4,5,6\}, 5--\{4,5,6\}, 9--\{9,10,11,12\})
-
-\end{itemize}
-
-Several more states and corresponding branches are omitted for the sake of
-clarity (these extra states are caused by solving the complications
-described in \sectionref{complications}).
-
-The actions introduce some of the complexity seen in the flow chart: all
-explicit branches and the transitions between some stages (19--20,
-21--22--23--24) in the second half of the flow chart are encoded within the
-actions; the transitions between the other stages (16--17, 24--25) in the
-second half arise from the interaction between actions
-
-However the transitions between stages in the first half of the flow chart
-(\figureref{flow chart image}) are not encoded anywhere in
-the parser; the transitions are determined by the ordering of log lines in
-the log file.  The complexity of the first 12 stages in the flow chart
-emerges from the interaction of simple rules and actions, easing the
-process of adding new actions (described in \sectionref{adding new
-actions}), as there is no requirement to explicitly insert the action into
-an algorithmic version of the flow chart.\footnote{The solution to the
-complication in \sectionref{out of order log lines} (out of order log
-lines) requires a list of the acceptable combinations of Postfix programs a
-mail can pass through; however this does not correspond to the
-\textit{parser actions\/} a mail must pass through.  Before that
-complication was overcome, i.e.\ for the first half of the development of
-this parser, there was nothing in the parser explicitly encoding the states
-or paths a mail could take; no other component in the parser has any need
-for this flow of data to be specified.}
 
@@
 
+\label{tracking re-injected mail}
+
 The most difficult complication initially encountered is that locally
@@
 
-XXX DO I NEED THIS LIST\@? IMPROVE HOW IT LOOKS IF I KEEP IT\@.
-
-Different delivery methods result in different log lines:
-
-\begin{description}
-
-    \item [Re-injected for forwarding:] forwarded as 56F5B43FD
-
-    \item [Delivered to a remote \gls{SMTP} server:] 250 Ok: queued as
-        BD07D3C49
-
-    \item [Local delivery:] delivered to command: /usr/bin/procmail
-
-\end{description}
-
 Unfortunately, while all log lines from an individual process appear in
@@
 
-subsubsection{Tracking re-injected mail}
-
-\label{tracking re-injected mail}
-
-The crux of the problem is that re-injected mails appear in the log files
-without explicit logging indicating their source.  There are two implicit
-indications:
+The crux of this complication is that re-injected mails appear in the log
+files without explicit logging indicating their source.  There are two
+implicit indications:
 
@@
         \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery (action: MAIL\_QUEUED\@; flowchart:~3), in which case a
-        new data structure will be created for that mail.  The mail will be
-        flagged as having unknown origins; this flag should be subsequently
-        cleared once the origin has been established.  This may also be an
-        indicator that the mail is a bounce notification, see
+        delivery, in which case a new data structure will be created for
+        that mail.  The mail will be flagged as having unknown origins;
+        this flag will subsequently be cleared once the origin has been
+        established.  This may also be an indicator that the mail is a
+        bounce notification, see
         \sectionref{identifying-bounce-notifications} for details.
@@
         rather than delivering directly to a mailbox or program as it
-        usually would (action: TRACK\@; flowchart:~10).\footnote{Relayed
-        delivery is performed by the \gls{SMTP} client; local delivery
-        means local to the server, i.e.\ an address the server is final
-        destination for.} In this case the mail may already have been
-        created (described above) and the unknown origin flag will be
-        cleared; if not a new data structure will be created.  In both
-        cases the re-injected mail is marked as a child of the original
-        mail.  The log line in question is:
+        usually would.\footnote{Relayed delivery is performed by the
+        \gls{SMTP} client; local delivery means local to the server, i.e.\
+        an address the server is final destination for.}  In this case the
+        mail may already have been created (described above) and the
+        unknown origin flag will be cleared; if not a new data structure
+        will be created.  In both cases the re-injected mail is marked as a
+        child of the original mail.  The log line in question is:
 
         \texttt{3FF7C4317: to=<username@example.com>, relay=local,
-        \newline{} delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This second indicator always occurs for re-injected mail but
-        typically occurs after the first indicator explained above.  This
-        is required to tie the parent and child mails together and so is
-        central to the process of tracking re-injected mails.
+        \newline{} \tab{} delay=0, status=sent (forwarded as 56F5B43FD)}
+
+        This always occurs for re-injected mail but typically occurs after
+        the first indicator.  This log line is required to tie the parent
+        and child mails together and so is central to the process of
+        tracking re-injected mails.
 
@@
     \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent has not yet been identified (action: COMMIT\@;
-        flowchart:~15).  Mark the mail as ready for entry in the database
-        (flowchart:~16), and wait for the parent to deal with it
-        (flowchart:~17).  The mail should not have subsequent log lines;
-        only its parent should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked
-        (action: COMMIT\@; flowchart:~18): as with all other mail, the data
-        is cleaned up, the child is entered in the database
-        (flowchart:~19), then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children
-        (flowchart:~20); if this is the last child and the parent has also
-        been entered in the database the parent will be deleted from the
-        state tables.
-
-    \item The last alternative is that the mail is a parent mail (action:
-        COMMIT\@; flowchart:~21).  Regardless of the state of its children
-        its data is cleaned up and entered in the database (flowchart:~22).
-        The parent may have children that are waiting to be entered in the
-        database (flowchart:~23); each of those children's data is cleaned
-        up and entered in the database, then deleted from the state tables.
-        The parent may also have outstanding children which are not yet
-        delivered (flowchart:~24), in which case the parent must wait for
-        those children to be finished with.  As soon as the last child is
-        deleted from the state tables the parent will also be finished with
-        (flowchart:~25), and deleted from the state tables.
+        whose parent has not yet been identified.  Mark the mail as ready
+        for entry in the database and wait for the parent to deal with it.
+        The mail should not have any subsequent log lines; only its parent
+        should refer to it.
+
+    \item If the mail is a child mail then it has already been tracked: as
+        with all other mail, the data is cleaned up, the child is entered
+        in the database, and then deleted from the state tables.  The child
+        mail will be removed from the parent mail's list of children; if
+        this is the last child and the parent has already been entered in
+        the database, the parent will also be deleted from the state
+        tables.
+
+    \item The last alternative is that the mail is a parent mail.
+        Regardless of the state of its children its data is cleaned up and
+        entered in the database.  The parent may have children that are
+        waiting to be entered in the database; the data for each of those
+        children is cleaned up and entered in the database, then deleted
+        from the state tables.  The parent may also have outstanding
+        children which are not yet delivered, in which case the parent must
+        be retained in the state tables until those children are finished
+        with.  As soon as the last child is deleted from the state tables
+        the parent will also be deleted from the state tables.
 

### 2009-02-12 doc/logparser-acronyms.tex
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

@@
     name={PLP},
-    description={Postfix Log Parser\glsadd{PLP glossary}}
+    description={Postfix Log Parser},
+    first={Postfix Log Parser (PLP)}
 }
@@
     name={DNSBL},
-    description={DNS Blacklist\glsadd{DNSBL glossary}}
+    description={DNS Blacklist},
+    first={DNS Blacklist (DNSBL)}
 }
@@
     name={API},
-    description={Application Programming Interface\glsadd{API glossary}}
+    description={Application Programming Interface},
+    first={Application Programming Interface (API)}
 }
@@
     name={SMTP},
-    description={Simple Mail Transfer Protocol\glsadd{SMTP glossary}}
+    description={Simple Mail Transfer Protocol},
+    first={Simple Mail Transfer Protocol (SMTP)}
 }
@@
     name={LMTP},
-    description={Local Mail Transfer Protocol\glsadd{LMTP glossary}}
+    description={Local Mail Transfer Protocol},
+    first={Local Mail Transfer Protocol (LMTP)}
 }
@@
     name={MTA},
-    description={Mail Transfer Agent\glsadd{MTA glossary}}
+    description={Mail Transfer Agent},
+    first={Mail Transfer Agent (MTA)}
 }
@@
     name={RFC},
-    description={Request For Comments\glsadd{RFC glossary}}
+    description={Request For Comments},
+    first={Request For Comments (RFC)}
 }
@@
     name={SQL},
-    description={Structured Query Language\glsadd{SQL glossary}}
+    description={Structured Query Language},
+    first={Structured Query Language (SQL)}
 }
@@
     name={LMA},
-    description={Log Mail Analyzer\glsadd{LMA glossary}}
+    description={Log Mail Analyzer},
+    first={Log Mail Analyzer (LMA)}
 }
@@
     name={CSV},
-    description={Comma-Separated Value\glsadd{CSV glossary}}
+    description={Comma-Separated Value},
+    first={Comma-Separated Value (CSV)}
 }
@@
     name={FQDN},
-    description={Fully Qualified Domain Name\glsadd{FQDN glossary}}
+    description={Fully Qualified Domain Name},
+    first={Fully Qualified Domain Name (FQDN)}
 }
@@
     name={pid},
-    description={Process Identifier\glsadd{pid glossary}}
+    description={Process Identifier},
+    first={Process Identifier (pid)}
 }
@@
     name={UCE},
-    description={Unsolicited Commercial Email\glsadd{UCE glossary}}
+    description={Unsolicited Commercial Email},
+    first={Unsolicited Commercial Email (UCE)}
 }
@@
     name={ESMTP},
-    description={Extended SMTP\glsadd{ESMTP glossary}}
+    description={Extended SMTP},
+    first={Extended SMTP (ESMTP)}
 }
@@
     name={IP},
-    description={Internet Protocol\glsadd{IP glossary}}
+    description={Internet Protocol},
+    first={Internet Protocol (IP)}
 }
@@
     name={SLCT},
-    description={Simple Logfile Clustering Tool\glsadd{SLCT glossary}}
+    description={Simple Logfile Clustering Tool},
+    first={Simple Logfile Clustering Tool (SLCT)}
 }
@@
     name={ATN},
-    description={Augmented Transition Networks\glsadd{ATN glossary}}
+    description={Augmented Transition Networks},
+    first={Augmented Transition Networks (ATN)}
 }
@@
     name={CLI},
-    description={Command Line Interface\glsadd{CLI glossary}}
+    description={Command Line Interface},
+    first={Command Line Interface (CLI)}
 }
@@
     name={NLP},
-    description={Natural Language Processing\glsadd{NLP glossary}}
+    description={Natural Language Processing},
+    first={Natural Language Processing (NLP)}
 }

### 2009-02-12 doc/logparser-architecture.tex
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

@@
 
-\section{Parser architecture and design}
+\section{Parser Architecture and Design}
 
@@
 There is some similarity between the parser's design and William Wood's
-\gls{ATN}~\cite{atns,nlpip}, a tool used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
-\gls{ATN} and the parser is accidental, but it is obvious that the two
-different approaches share an underlying separation of concerns; this
-appears to be a natural division of responsibility and functionality.
+\acronym{ATN}~\cite{atns,nlpip}, a tool used in Computational Linguistics
+for creating grammars to parse or generate sentences.  The resemblance
+between \acronym{ATN} and the parser is accidental, but it is obvious that
+the two different approaches share an underlying separation of concerns;
+this appears to be a natural division of responsibility and functionality.
 \Tableref{Similarities between ATN and this architecture} shows the
@@
         \tabletopline{}%
-        \gls{ATN}          & Architecture  & Similarity                \\
+        \acronym{ATN}   & Architecture  & Similarity                \\
         \tablemiddleline{}%
@@
 
-\subsection{Rule characteristics}
+\subsection{Rule Characteristics}
 
@@
 
-\subsection{Overlapping rules}
+\subsection{Overlapping Rules}
 
@@
 allowing a more specific rule to match some log lines and a more general
-rule to match the majority, e.g.\ separating \gls{SMTP} delivery to
-specific sites from \gls{SMTP} delivery to the rest of the world.  The
+rule to match the majority, e.g.\ separating \acronym{SMTP} delivery to
+specific sites from \acronym{SMTP} delivery to the rest of the world.  The
 algorithm provides a mechanism for ordering overlapping rules: the priority
@@
 
-    \item Sort by regex and visually inspect the list, e.g.\ with \gls{SQL}
-        similar to: \textbf{select regex from rules order by regex;}
+    \item Sort by regex and visually inspect the list, e.g.\ with
+        \acronym{SQL} similar to: \textbf{select regex from rules order by
+        regex;}
 

### 2009-02-12 doc/logparser-background.tex
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

@@
 This section provides background information helpful in understanding the
-remainder of the thesis.  It begins with a discussion of the motivation
-underlying the project, followed by some technical information: the use of
-a database as an \gls{API}\@; a brief introduction to \gls{SMTP}\@.  It
-finishes with a longer introduction to Postfix, concentrating on anti-spam
-restrictions and policy servers.
+remainder of this thesis.  It begins with a discussion of the motivation
+underlying the project, followed by a brief introduction to
+\acronym{SMTP}\@.  It finishes with a longer introduction to Postfix,
+concentrating on anti-spam restrictions and policy servers.
 
@@
 optimise a mail server's Postfix-based anti-spam restrictions, generate
-statistics and graphs, and provide a platform on which new restrictions on
-trial can be evaluated to determine if they are beneficial in the fight
-against spam.  The program parses Postfix log files and populates a
-database with the data gleaned from those log files, providing a consistent
-and simple view of the log files that future tools can utilise.  The
-gathered data can be used to optimise current anti-spam measures, to
-provide a baseline to test new anti-spam measures against, or to produce
-statistics showing how effective those measures are.
+statistics and graphs, and provide a platform on which new restrictions can
+be evaluated to determine if they are beneficial in the fight against spam.
+The program parses Postfix log files and populates a database with the data
+gleaned from those log files, providing a consistent and simple view of the
+log files that future tools can utilise.  The gathered data can be used to
+optimise current anti-spam measures, to provide a baseline to test new
+anti-spam measures against, or to produce statistics showing how effective
+those measures are.
 
@@
 
-\section{Using a Database as an Application Programming Interface}
-
-\label{database as API}
-
-The database populated by this program provides a simple interface to
-Postfix log files.  Although the interface is a database schema rather than
-a set of functions provided by shared code, it performs the same function
-as any other \gls{API}: it provides a stable interface for users of the
-interface, allowing the implementation to be changed without adverse
-effects.  If as the \gls{API} is maintained, the parser can be improved to
-handle additional complications; support can be added for earlier or later
-releases of Postfix; bugs can be fixed or limitations removed.
-
-Using a database simplifies writing programs that need to interact with the
-data in several ways:
-
-\begin{enumerate}
-
-    \item Facilities are provided to access databases from most programming
-        languages, allowing a developer to access the data gathered using
-        their preferred programming language, rather than being restricted
-        to the language the parser is written in.  It is often possible to
-        write an interface layer allowing code written in one language to
-        be used in another language, but this greatly increases the effort
-        required to use the parser.
-
-    \item Databases provide complex querying and sorting functionality to
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of varying complexity and
-        sophistication, that can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item New columns can be added to the tables used by the
-                program, using DEFAULT clauses or TRIGGERS to populate
-                them.\footnote{Please refer to an \gls{SQL} guide for
-                explanations of these terms, e.g.\
-                \url{http://philip.greenspun.com/sql/}.}
-
-            \item A VIEW gives a custom arrangement of data with minimal
-                effort.
-
-            \item If the database supports it, access can be granted on a
-                fine-grained basis, e.g.\ allowing the finance department
-                to produce invoices, the helpdesk to run limited queries as
-                part of dealing with support calls, and the administrators
-                to have full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\gls{SQL}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-            \item Other tables can be added to the database, e.g.\ to cache
-                historical, summary or computed data.
-
-        \end{itemize}
-
-    \item \gls{SQL} is reasonably standard and many people will already be
-        familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
-        introduction to \gls{SQL} can be found at
-        \url{http://philip.greenspun.com/sql/}, others are
-        \url{http://www.w3schools.com/sql/default.asp},
-        \url{http://sqlcourse.com/}).  Although every vendor implements a
-        different dialect of \gls{SQL}, the basics are the same everywhere
-        (analogous to the overall similarities and minor differences
-        between Irish English, British English, American English, and
-        Australian English).  Depending on the database in use there may be
-        tools available that reduce or remove the requirement to know
-        \gls{SQL}; several are available for \gls{SQLite}
-        (\url{http://www.sqlite.org/}), the default database used by the
-        parser implementation.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, because the log files need only be parsed once rather
-than each time the data is used; indeed the results may be used by someone
-with no access to the original log files.
-
 \section{Simple Mail Transfer Protocol}
@@
 
-The \acrlong{SMTP}, originally defined in \gls{RFC}~821~\cite{RFC821} and
-updated in \gls{RFC}~2821~\cite{RFC2821}, is used for transferring mail
-between the sending and receiving \gls{MTA}\@.  It is a simple, human
-readable, plain text protocol, simplifying testing and debugging problems
-with it.  Despite the simplicity of the protocol many virus and spam
-sending programs fail to implement it properly, so requiring strict
+The \glsfirst{SMTP}, originally defined in \acronym{RFC}~821~\cite{RFC821}
+and updated in \acronym{RFC}~2821~\cite{RFC2821}, is used for transferring
+mail between the sending and receiving \acronym{MTA}\@.  It is a simple,
+human readable, plain text protocol, simplifying testing and debugging
+problems with it.  Despite the simplicity of the protocol many virus and
+spam sending programs fail to implement it properly, so requiring strict
 adherence to the protocol specification is beneficial in protecting against
@@
 servers adhered to the principle of \textit{Be liberal in what you accept,
-and conservative in what you send\/} from \gls{RFC}~760~\cite{rfc760}, but
-unfortunately that principle was written in a friendlier time.  Given the
-deluge of spam that mail servers are subjected to daily, a more appropriate
-maxim could be: \textit{Require strict adherence to \gls{RFC}~2821;
-implement the strongest restrictions you can; relax the restrictions and
-adherence only when legitimate mail is impeded.\/}  it is not as friendly,
-nor as catchy, but it more accurately reflects the current situation.} A
-typical \gls{SMTP} conversation resembles the following (the lines starting
-with a three digit number are sent by the server, all other lines are sent
-by the client):
+and conservative in what you send\/} from \acronym{RFC}~760~\cite{rfc760},
+but unfortunately that principle was written in a friendlier time.  Given
+the deluge of spam that mail servers are subjected to daily, a more
+appropriate maxim could be: \textit{Require strict adherence to
+\acronym{RFC}~2821; implement the strongest restrictions you can; relax the
+restrictions and adherence only when legitimate mail is impeded.\/}  It is
+not as friendly, nor as catchy, but it more accurately reflects the current
+situation.} A typical \acronym{SMTP} conversation resembles the following
+(the lines starting with a three digit number are sent by the server, all
+other lines are sent by the client):
 
@@
 returned by the server and carries on regardless.  Many spam and virus
-sending programs are this deficient --- unfortunately others (particularly
-newer programs) were written by competent programmers, or utilise
-competently written programs (e.g.\ Postfix or Sendmail on Unix hosts,
-Microsoft Outlook on Windows hosts).  Traditionally a mail server would
-have done its best to deal with deficient clients, with the intention of
-accepting as much mail destined for its users as
-possible,\footref{footnote:rfc760} e.g.\ by ignoring the absence of a HELO
-command, or by accepting sender or recipient addresses that were not
-enclosed in \texttt{<>}.
+sending programs have serious deficiencies; unfortunately others
+(particularly newer programs) were written by competent programmers, or
+send mail using well written programs (e.g.\ Postfix or Sendmail on Unix
+hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
+would have done its best to deal with deficient clients, with the intention
+of accepting as much mail destined for its users as possible, e.g.\ by
+ignoring the absence of a HELO command, or by accepting sender or recipient
+addresses that were not enclosed in \texttt{<>}.
 
-A detailed description of \gls{SMTP} is beyond the scope of this thesis:
-introductory guides can be found at~\cite{smtp-intro-01,smtp-intro-02},
-and the definitive references are~\cite{RFC821,RFC2821}.
+A detailed description of \acronym{SMTP} is beyond the scope of this
+thesis: introductory guides can be found
+at~\cite{smtp-intro-01,smtp-intro-02}, and the definitive references
+are~\cite{RFC821,RFC2821}.
 
@@
 
-Postfix is a \acrlong{MTA}~(\gls{MTA}) with the following design aims (in
-order of importance): security, flexibility of configuration, scalability,
-and high performance.  It features extensive optional anti-spam
-restrictions, allowing an administrator to deploy those restrictions which
-they judge suitable for their site's needs, rather than a fixed set chosen
-by Postfix's author.  These restrictions can be selectively applied,
-combined, and bypassed on a per-client, per-recipient, or per-sender basis,
-allowing different levels of stricture and permissiveness.  Postfix
-leverages simple lookup tables to support arbitrarily complicated
-user-defined sequences of restrictions and exceptions; policy servers
+Postfix is a \glsfirst{MTA} with the following design aims
+(in order of importance): security, flexibility of configuration,
+scalability, and high performance.  It features extensive optional
+anti-spam restrictions, allowing an administrator to employ those
+restrictions which they judge suitable for their site's needs, rather than
+a fixed set chosen by Postfix's author.  These restrictions can be
+selectively applied, combined, and bypassed on a per-client, per-recipient,
+or per-sender basis, allowing different levels of stricture and
+permissiveness.  Postfix leverages simple lookup tables to support
+arbitrarily complicated user-defined sequences of restrictions and
+exceptions.  Administrators can supply their own rejection messages to make
+it clear to senders exactly why their mail was rejected.  Policy servers
 (\sectionref{policy servers}) provide a simple way to write new
-restrictions.  Administrators can also supply their own rejection messages
-to make it clear to senders exactly why their mail was rejected.
-Unfortunately this flexibility has a cost: complexity in the log files
-generated.  Although it is simple to use standard Unix text processing
-utilities to determine the fate of an individual email, following the
-journey an email takes through Postfix can be quite difficult.  For most
-mails the journey is simple and brief, but the remaining minority can have
-quite complex journeys (see \sectionref{complications} for details).
+restrictions without having to modify Postfix's source code.  Unfortunately
+this flexibility has a cost: complexity in the log files generated.
+Although it is relatively simple in most cases to use standard Unix text
+processing utilities to determine the fate of an individual email,
+following the journey an email takes through Postfix can be quite
+difficult.  For most mails the journey is simple and brief, but the
+remaining minority can have quite complex journeys (see
+\sectionref{complications} for details).
 
 Postfix's design follows the Unix philosophy of \textit{``Write programs
-that do one thing and do it well''\/}~\cite{unix-philosophy}, and is
-separated into various component programs to perform the tasks required of
-an \gls{MTA}\@: receive mail, send mail, local delivery of mail, etc.\ ---
-full details can be found in~\cite{postfix-overview}.  Each log line
-contains the name of the Postfix component that produced it, and this
-information is used when determining which rules should be used to parse
-each log line (see \sectionref{rule characteristics} for details).
+that do one thing and do it well''\/}~\cite{unix-philosophy}, and it is
+separated into multiple programs that each perform one of the tasks
+required of an \acronym{MTA}\@: receive mail, send mail, local delivery of
+mail, etc.\ (full details can be found in~\cite{postfix-overview}).  Each
+log line contains the name of the Postfix component that produced it, and
+this information is used when determining which rules should be used to
+parse each log line (see \sectionref{rule characteristics} for details).
 Postfix's design is strongly influenced by security concerns: those
@@
 
-\subsection{Mixing and matching Postfix restrictions}
+\subsection{Mixing and Matching Postfix Restrictions}
 
@@
 Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}; the following is a brief overview
-only.
+smtpd_per_user_control, policy-servers}; the following is only a brief
+introduction.
 
 Postfix uses one restriction list (containing zero or more restrictions)
-for each stage of the \gls{SMTP} conversation: client connection, HELO
+for each stage of the \acronym{SMTP} conversation: client connection, HELO
 command, MAIL FROM command, RCPT TO commands, DATA command, and end of
-data.  The appropriate restriction list is evaluated for each stage
-(evaluation will be explained shortly), though by default the restriction
-lists for client connection, HELO, and MAIL FROM commands will not be
-evaluated until the first RCPT TO command is received, because some clients
-do not deal properly with rejections before the first RCPT TO command; a
-benefit of this delay is that Postfix has more information available when
-logging rejections.
-
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno\/} or the name of another restriction to be
-evaluated.\footnote{Other results are possible as described
-in~\cite{smtpd_access_readme,smtpd_per_user_control,policy-servers}.} The
-meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing
-exceptions to more general rules.  When the result is the name of another
-restriction Postfix will evaluate that restriction, allowing restrictions
-to be chosen based on the client \gls{IP} address, client hostname, HELO
-hostname, sender address, or recipient address.\footnote{E.g.\ the
-administrator may require that all clients on the local network have valid
-DNS entries, to prevent people sending mail from unknown machines.}  The
-administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator in
-as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting all mail
---- the administrator must be cognisant of the effects their configuration
-changes will have.  This is similar to one of UNIX's design philosophies:
-\textit{``UNIX was not designed to stop its users from doing stupid things,
-as that would also stop them from doing clever
-things''\/}~\cite{unix-philosophy}.
+data.  The appropriate restriction list is evaluated for each stage, though
+by default the restriction lists for client connection, HELO, and MAIL FROM
+commands will not be evaluated until the first RCPT TO command is received,
+because some clients do not deal properly with rejections before this
+stage; a benefit of this delay is that Postfix has more information
+available when logging rejections.
 
@@
 
-The restriction \texttt{check\_client\_access} checks if the \gls{IP}
+The restriction \texttt{check\_client\_access} checks if the \acronym{IP}
 address of the connected client is found in the specified table and returns
@@
 \tab{}\texttt{reject\_rbl\_client dnsbl.example.com}\newline{} checks the
-\gls{DNSBL} \texttt{dnsbl.example.com} and rejects the command if the
-client's \gls{IP} address is listed.
+\acronym{DNSBL} \texttt{dnsbl.example.com} and rejects the command if the
+client's \acronym{IP} address is listed.
+
+Each restriction is evaluated to produce a result of \textit{reject},
+\textit{permit}, \textit{dunno}, or the name of another restriction to be
+evaluated; other less commonly used results are possible as described
+in~\cite{smtpd_access_readme,smtpd_per_user_control,policy-servers}. The
+meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
+\textit{dunno\/} means to stop evaluating the current restriction and
+continue processing the remainder of the restriction list, allowing
+exceptions to more general rules.  XXX GIVE AN EXAMPLE\@.  When the result
+is the name of another restriction Postfix will evaluate that restriction,
+allowing restrictions to be chosen based on the client \acronym{IP}
+address, client hostname, HELO hostname, sender address, or recipient
+address.  E.g.\ the administrator may require that all clients on the local
+network have valid DNS entries, to prevent people sending mail from unknown
+machines XXX EXPAND THIS WITH DETAILS\@.
+
+The administrator can define new restrictions as a list of existing
+restrictions, allowing arbitrarily long and complex sequences of lookups,
+restrictions and exceptions.  Postfix tries to protect the administrator
+from misconfiguration in as far as is reasonable, e.g.\ the restriction
+\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
+because the parameter it checks (the hostname given in the HELO command) is
+under the control of the remote client.  Despite this, it is possible for
+the administrator to make catastrophic mistakes, e.g.\ rejecting all mail
+--- the administrator must be cognisant of the effects their configuration
+changes will have.  This is similar to one of UNIX's design philosophies:
+\textit{``UNIX was not designed to stop its users from doing stupid things,
+as that would also stop them from doing clever
+things''\/}~\cite{unix-philosophy}.
 
@@
 
-\subsection{Policy servers}
+\subsection{Policy Servers}
 
@@
 A policy server~\cite{policy-servers} is an external program consulted by
-Postfix to determine the fate of an \gls{SMTP} command.  The policy server
-is given state information by Postfix (sample state information is shown in
-\tableref{Example attributes sent to policy servers}) and returns a verdict
-from the set described in \sectionref{Mixing and matching Postfix
-restrictions}.  The policy server can perform more complex checks than
-those provided by Postfix: a trivial example is allowing addresses
-associated with the payroll system to send mail on the third Tuesday after
-pay day only, to help prevent problems from phishing mails using faked
-sender addresses.\footnote{A phishing mail might claim that the payroll
-system had a disastrous disk failure, and until the server is replaced all
-salary payments will have to be processed manually, so please reply to this
-mail with your name, address, and bank account details.}
+Postfix to determine the fate of an \acronym{SMTP} command.  The policy
+server is given state information by Postfix (sample state information is
+shown in \tableref{Example attributes sent to policy servers}) and returns
+a result (reject, permit, dunno, a rejection name) as described in
+\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
+perform more complex checks than those provided by Postfix: a trivial
+example is allowing addresses associated with the payroll system to send
+mail on the third Tuesday after pay day only, to help prevent problems from
+phishing mails using faked sender addresses.\footnote{A phishing mail might
+claim that the payroll system had a disastrous disk failure, and until the
+server is replaced all salary payments will have to be processed manually,
+so please reply to this mail with your name, address, and bank account
+details.}
 
@@
         considerable resistance to the proposal because it breaks or vastly
-        complicates some features of \gls{SMTP}, e.g.\ forwarding mail from
-        one company or university to another when a user moves.
+        complicates some features of \acronym{SMTP}, e.g.\ forwarding mail
+        from one company or university to another when a user moves.
 
@@
         rejects a delivery attempt when the triple of \newline{}
-        \tab{}\texttt{(sender address, recipient address, remote \gls{IP}
-        address)}\newline{} has not been seen before; on second and
-        subsequent delivery attempts from that triple the mail will be
-        accepted.  This blocks spam because maintaining a list of failed
-        addresses and retrying after a temporary failure is uneconomical
-        for a spam sender, but a legitimate mail server must retry
-        deliveries that temporarily failed.  Sadly spam senders are using
-        increasingly complex and well written programs to distribute spam,
-        frequently using an ISP provided \gls{SMTP} server from a
+        \tab{}\texttt{(sender address, recipient address, remote
+        \acronym{IP} address)}\newline{} has not been seen before; on
+        second and subsequent delivery attempts from that triple the mail
+        will be accepted.  This blocks spam because maintaining a list of
+        failed addresses and retrying after a temporary failure is
+        uneconomical for a spam sender, but a legitimate mail server must
+        retry deliveries that temporarily failed.  Sadly spam senders are
+        using increasingly complex and well written programs to distribute
+        spam, frequently using an ISP provided \acronym{SMTP} server from a
         compromised machine on the ISP's network.  Greylisting will slowly
@@
     \item Scoring systems such as Policyd-weight~\cite{policyd-weight}
-        perform tests on features of the delivery attempt (e.g.\ \gls{IP}
-        address, sender address), incrementing or decrementing a score
-        based on the results; if the eventual score is higher than a
-        threshold the mail is rejected.  The administrator must manually
+        perform tests on features of the delivery attempt (e.g.\
+        \acronym{IP} address, sender address), incrementing or decrementing
+        a score based on the results; if the eventual score is higher than
+        a threshold the mail is rejected.  The administrator must manually
         whitelist clients if they are to bypass a Postfix restriction;
@@
 starting with the motivation behind the project, continuing with an
-explanation of the use of a database as an \gls{API} and an introduction to
-\gls{SMTP}.  The focus then switched to Postfix, the \gls{MTA} in use, and
-its restrictions and policy servers.
+explanation of the use of a database as an \acronym{API} and an
+introduction to \acronym{SMTP}.  The focus then switched to Postfix, the
+\acronym{MTA} in use, and its restrictions and policy servers.

### 2009-02-12 doc/logparser-glossary.tex
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

@@
     embeddable, zero-configuration SQL database engine.\/} SQLite is an
-    \gls{SQL} implementation focusing on correctness, simplicity, and
-    speed.  Unlike other \gls{SQL} implementations it does not require a
-    separate server process, greatly simplifying deployment of an
+    \acronym{SQL} implementation focusing on correctness, simplicity, and
+    speed.  Unlike other \acronym{SQL} implementations it does not require
+    a separate server process, greatly simplifying deployment of an
     application utilising it.  More details can be found
@@
     $<>$ is the sender address used for mail that should not be replied to,
-    e.g.\ bounce notifications.  In \gls{SMTP} all addresses are enclosed
-    in $<>$, so \textit{username@domain\/} becomes
+    e.g.\ bounce notifications.  In \acronym{SMTP} all addresses are
+    enclosed in $<>$, so \textit{username@domain\/} becomes
     \textit{$<$username@domain$>$\/}; thus $<>$ is actually an empty
@@
     must accept mail sent from $<>$, or they are in violation of
-    \gls{RFC}~2821~\cite{RFC2821}.
+    \acronym{RFC}~2821~\cite{RFC2821}.
 }}

### 2009-02-12 doc/logparser-implementation.tex
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

@@
 
-\subsection{Rules table}
+\subsection{Using a Database as an Application Programming Interface}
+
+\label{database as API}
+
+XXX LAST CHECKED DATES FOR ALL URLS\@.
+
+The database populated by this program provides a simple interface to
+Postfix log files.  Although the interface is a database schema rather than
+a set of functions provided by a library, it performs the same function as
+any other \acronym{API}: it provides a stable interface, allowing code on
+either side of the interface to be changed without adverse effects, as long
+as the interface is maintained.  The parser can be improved to handle
+additional complications, support earlier or later releases of Postfix,
+bugs can be fixed or limitations removed.  Programs that use the database
+can range from the simple examples in \sectionref{motivation} to far more
+complex data mining tools.
+
+Using a database simplifies writing programs that need to interact with the
+data in several ways:
+
+\begin{enumerate}
+
+    \item Facilities are provided to access databases from most programming
+        languages, allowing a developer to access the data gathered using
+        their preferred programming language, rather than being restricted
+        to the language the parser is written in.  It is often possible to
+        write an interface layer allowing code written in one language to
+        be used in another language, but this greatly increases the effort
+        required to use the parser.
+
+    \item Databases provide complex querying and sorting functionality to
+        the user without requiring large amounts of programming.  All
+        databases provide a program, of varying complexity and
+        sophistication, that can be used for ad hoc queries with minimal
+        investment of time.
+
+    \item Databases are easily extensible, e.g.:
+
+        \begin{itemize}
+
+            \item New columns can be added to the tables used by the
+                program, using DEFAULT clauses or TRIGGERS to populate
+                them.
+
+            \item A VIEW gives a custom arrangement of data with minimal
+                effort.
+
+            \item If the database supports it, access can be granted on a
+                fine-grained basis, e.g.\ allowing the finance department
+                to produce invoices, the helpdesk to run limited queries as
+                part of dealing with support calls, and the administrators
+                to have full access to the data.
+
+            \item Triggers can be written to perform actions when certain
+                events occur.  In pseudo-\acronym{SQL}\@:
+
+\begin{verbatim}
+CREATE TRIGGER ON INSERT INTO results
+    WHERE sender = 'boss@example.com'
+        AND postfix_action = 'REJECTED'
+    SEND PANIC EMAIL TO 'postmaster@example.com';
+\end{verbatim}
+
+            \item Other tables can be added to the database, e.g.\ to cache
+                historical, summary, or computed data.
+
+        \end{itemize}
+
+    \item \acronym{SQL} is reasonably standard and many people will already
+        be familiar with it; for those unfamiliar with it there are lots of
+        readily available resources from which to learn (a good
+        introduction to \acronym{SQL} can be found at
+        \url{http://philip.greenspun.com/sql/}, others are
+        \url{http://www.w3schools.com/sql/default.asp},
+        \url{http://sqlcourse.com/}).  Although every vendor implements a
+        different dialect of \acronym{SQL}, the basics are the same
+        everywhere (analogous to the overall similarities and minor
+        differences amongst Irish English, British English, American
+        English, and Australian English).  Depending on the database in use
+        there may be tools available that reduce or remove the requirement
+        to know \acronym{SQL}; several are available for \gls{SQLite}, the
+        default database used by the parser implementation.
+
+\end{enumerate}
+
+Storing the results in a database will also increase the efficiency of
+using those results, because the log files need only be parsed once rather
+than each time the data is used; indeed the results may be used by someone
+with no access to the original log files.
+
+\subsection{Rules Table}
 
@@
 
-\subsection{Connections table}
+\subsection{Connections Table}
 
@@
 
-    \item [server\_ip] The \gls{IP} address (IPv4 or IPv6) of the server:
-        the local address when receiving mail, the remote address when
-        sending mail.
+    \item [server\_ip] The \acronym{IP} address (IPv4 or IPv6) of the
+        server: the local address when receiving mail, the remote address
+        when sending mail.
 
     \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \gls{IP} address could not be resolved to a
-        hostname via DNS\@.
-
-    \item [client\_ip] The client \gls{IP} address (IPv4 or IPv6): the
+        \texttt{unknown} if the \acronym{IP} address could not be resolved
+        to a hostname via DNS\@.
+
+    \item [client\_ip] The client \acronym{IP} address (IPv4 or IPv6): the
         remote address when receiving mail, the local address when sending
@@
     \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \gls{IP} address could not be resolved to a
-        hostname via DNS\@.
+        \texttt{unknown} if the \acronym{IP} address could not be resolved
+        to a hostname via DNS\@.
 
@@
         or virus senders think it is a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \gls{SMTP} command, but
-        it is quite simple to rectify this, as described in
+        logs the HELO hostname when it rejects an \acronym{SMTP} command,
+        but it is quite simple to rectify this, as described in
         \sectionref{logging helo}.
@@
 
-\subsection{Results table}
+\subsection{Results Table}
 
@@
     \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \gls{SMTP} command ---
-        a mechanism that is quite useful for testing new restrictions.
+        enforcing a restriction that would reject an \acronym{SMTP} command
+        --- a mechanism that is quite useful for testing new restrictions.
         This field will be 1 if the log line parsed was a warning rather
@@
 
-    \item [smtp\_code] The \gls{SMTP} code associated with the log line.
-        In general an \gls{SMTP} code is only present in a log line for a
-        rejection or final delivery; results initially missing an
-        \gls{SMTP} code will duplicate the \gls{SMTP} code of other results
-        in that connection.  Some final delivery log lines do not contain
-        an \gls{SMTP} code: in those cases the code is specified by the
-        rule, based on the success or failure represented by the log line.
+    \item [smtp\_code] The \acronym{SMTP} code associated with the log
+        line.  In general an \acronym{SMTP} code is only present in a log
+        line for a rejection or final delivery; results initially missing
+        an \acronym{SMTP} code will duplicate the \acronym{SMTP} code of
+        other results in that connection.  Some final delivery log lines do
+        not contain an \acronym{SMTP} code: in those cases the code is
+        specified by the rule, based on the success or failure represented
+        by the log line.
 
@@
     \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from a \gls{DNSBL}\@.
+        fields, e.g.\ the rejection message from a \acronym{DNSBL}\@.
 
@@
 
-\subsection{Example rule}
+\subsection{Example Rule}
 
@@
 
-\subsection{Creating new rules}
+\subsection{Creating New Rules}
 
@@
 because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different \gls{DNSBL}
-services, policy servers, or custom rejection messages.  To facilitate
-parsing new log lines, the parser's design separates parsing rules from
-parsing actions: adding new actions can be difficult, but adding new rules
-to parse new rejection messages is trivial, and occurs much more
+restrictions which suits their needs, including using different
+\acronym{DNSBL} services, policy servers, or custom rejection messages.  To
+facilitate parsing new log lines, the parser's design separates parsing
+rules from parsing actions: adding new actions can be difficult, but adding
+new rules to parse new rejection messages is trivial, and occurs much more
 frequently.  A program is provided to ease the process of creating new
 rules from unparsed log lines, based on the algorithm developed by Risto
-Vaarandi for his \gls{SLCT}~\cite{slct-paper}.  The differences between the
-two algorithms will be outlined as part of the general explanation below.
-
-The core of the \gls{SLCT} algorithm is quite simple: log lines are
+Vaarandi for his \acronym{SLCT}~\cite{slct-paper}.  The differences between
+the two algorithms will be outlined as part of the general explanation
+below.
+
+The core of the \acronym{SLCT} algorithm is quite simple: log lines are
 generally created by substituting variable words into a fixed pattern, and
@@
         substitutions on the input log lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \gls{IP} addresses, the
+        variable terms (e.g.\ email addresses, \acronym{IP} addresses, the
         standard start of rejection messages, etc.) with regex keywords
@@
 
-\subsection{Actions in detail}
+\subsection{Actions in Detail}
 
@@
 
-            \item [pid] The \gls{pid} of the program that logged the line.
+            \item [pid] The \acronym{pid} of the program that logged the
+                line.
 
@@
     \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \gls{pid}) and save both the
-        client hostname and \gls{IP} address.
+        table entry (indexed by \daemon{smtpd} \acronym{pid}) and save both
+        the client hostname and \acronym{IP} address.
 
@@
 
-    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an \gls{SMTP}
-        command from the remote client: log the rejection with a mail if
-        there is a queueid in the log line, or with the connection if not.
+    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
+        \acronym{SMTP} command from the remote client: log the rejection
+        with a mail if there is a queueid in the log line, or with the
+        connection if not.
 
@@
 
-\subsection{Postfix actions}
+\subsection{Postfix Actions}
 
@@
 
-\subsection{Adding new actions}
+\subsection{Adding New Actions}
 
@@
 
-    \item Mail enters the system via \gls{SMTP} or local submission.
+    \item Mail enters the system via \acronym{SMTP} or local submission.
 
@@
 Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop}, by \gls{SMTP}, by re-injection due to
+submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
 forwarding, or internally generated by Postfix.  Local submission is the
@@
 
-\gls{SMTP} is more complicated than local submission:
+\acronym{SMTP} is more complicated than local submission:
 
@@
     \item This is followed by rejection of sender address, recipient
-        addresses, client \gls{IP} address or hostname, HELO hostname,
+        addresses, client \acronym{IP} address or hostname, HELO hostname,
         etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
@@
     \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \gls{SMTP} commands the data will be saved
-        to the database; if not there will not be any data to save (any
-        mails accepted will already have been cloned so their data is in
-        another data structure).
+        Postfix has rejected any \acronym{SMTP} commands the data will be
+        saved to the database; if not there will not be any data to save
+        (any mails accepted will already have been cloned so their data is
+        in another data structure).
 
@@
 
-\section{Complications encountered}
+\section{Complications Encountered}
 
@@
 
-\subsection{Queueid vs pid}
+\subsection{Queueid Vs Pid}
 
 The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \gls{pid}, then transition to
+first be correlated by the \daemon{smtpd} \acronym{pid}, then transition to
 being correlated by the queueid.  This is relatively minor, but does
@@
     \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \gls{pid} or queueid.
+        needs to lookup the data structures by \acronym{pid} or queueid.
 
@@
 
-\subsection{Connection reuse}
+\subsection{Connection Reuse}
 
@@
 tell in advance which connections will accept multiple mails.  Once the
-mail has been accepted its log lines will not be correlated by \gls{pid}
-any more, its queueid will be used instead (except when timeouts occur
-during the data phase \sectionref{timeouts-during-data-phase}).  If the
-original connection has any useful data (e.g.\ rejections) it will be saved
-to the database when the client disconnects.  One unsolved difficulty is
-distinguishing between different groups of rejections, e.g.\ when dealing
-with the following sequence:
+mail has been accepted its log lines will not be correlated by
+\acronym{pid} any more, its queueid will be used instead (except when
+timeouts occur during the data phase
+\sectionref{timeouts-during-data-phase}).  If the original connection has
+any useful data (e.g.\ rejections) it will be saved to the database when
+the client disconnects.  One unsolved difficulty is distinguishing between
+different groups of rejections, e.g.\ when dealing with the following
+sequence:
 
@@
 
-    \item The client issues the RSET command to reset the \gls{SMTP}
+    \item The client issues the RSET command to reset the \acronym{SMTP}
         session.
@@
 
-\subsection{Re-injected mails}
+\subsection{Re-injected Mails}
 
@@
         usually would.\footnote{Relayed delivery is performed by the
-        \gls{SMTP} client; local delivery means local to the server, i.e.\
-        an address the server is final destination for.}  In this case the
-        mail may already have been created (described above) and the
-        unknown origin flag will be cleared; if not a new data structure
-        will be created.  In both cases the re-injected mail is marked as a
-        child of the original mail.  The log line in question is:
+        \acronym{SMTP} client; local delivery means local to the server,
+        i.e.\ an address the server is final destination for.}  In this
+        case the mail may already have been created (described above) and
+        the unknown origin flag will be cleared; if not a new data
+        structure will be created.  In both cases the re-injected mail is
+        marked as a child of the original mail.  The log line in question
+        is:
 
@@
 
-\subsection{Identifying bounce notifications}
+\subsection{Identifying Bounce Notifications}
 
@@
         messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \gls{SMTP} or submitted
+        internally by Postfix, not accepted via \acronym{SMTP} or submitted
         locally by \daemon{postdrop}.
@@
 
-\subsection{Aborted delivery attempts}
+\subsection{Aborted Delivery Attempts}
 
@@
 
-Some mail clients behave unexpectedly during the \gls{SMTP} dialogue: the
-client aborts the first delivery attempt after the first recipient is
+Some mail clients behave unexpectedly during the \acronym{SMTP} dialogue:
+the client aborts the first delivery attempt after the first recipient is
 accepted, then makes a second delivery attempt for the same recipient which
@@
 reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \gls{IP} address, but the database
+available is the client hostname and \acronym{IP} address, but the database
 schema requires many more fields be populated (see \sectionref{connections
@@
 
-\subsection{Further aborted delivery attempts}
+\newpage{} % XXX
+
+\subsection{Further Aborted Delivery Attempts}
 
@@
 
-\subsection{Timeouts during DATA phase}
+\subsection{Timeouts During DATA Phase}
 
@@
 
-The DATA phase of the \gls{SMTP} conversation is where the headers and body
-of the mail are transferred.  Sometimes there is a timeout or the
+The DATA phase of the \acronym{SMTP} conversation is where the headers and
+body of the mail are transferred.  Sometimes there is a timeout or the
 connection is lost\footnote{For brevity's sake \textit{timeout\/} will be
@@
 
-\subsection{Discarding cleanup log lines}
+\subsection{Discarding Cleanup Log Lines}
 
@@
 
-\subsection{Pickup logging after cleanup}
+\subsection{Pickup Logging After Cleanup}
 
@@
 
-\subsection{Smtpd stops logging}
+\subsection{Smtpd Stops Logging}
 
@@
 
-\subsection{Out of order log lines}
+\subsection{Out of Order Log Lines}
 
@@
 
-\subsection{Yet more aborted delivery attempts}
+\subsection{Yet More Aborted Delivery Attempts}
 
@@
 
-\subsection{Mail deleted before delivery is attempted}
+\subsection{Mail Deleted Before Delivery is Attempted}
 
@@
 
-\subsection{Bounce notification mails delivered before their creation is
-logged}
+\subsection{Bounce Notification Mails Delivered Before Their Creation Is
+Logged}
 
@@
 
-\subsection{Mails deleted during delivery}
+\subsection{Mails Deleted During Delivery}
 
@@
 
-\section{Limitations and possible improvements}
+\section{Limitations and Possible Improvements}
 
@@
         existing rules modified to cope with the new or changed log lines.
-        Similarly using a new \gls{DNSBL}, a new policy server, or new
+        Similarly using a new \acronym{DNSBL}, a new policy server, or new
         administrator-defined rejection messages require new rules.
@@
     \item The hostname used in the HELO command is not logged if the mail
-        is accepted.\footnote{Tested with Postfix 2.2.10, 2.3.11, and 2.4.7;
-        this may possibly have changed in Postfix 2.5. XXX TEST THIS
+        is accepted.\footnote{Tested with Postfix 2.2.10, 2.3.11, and
+        2.4.7; this may possibly have changed in Postfix 2.5. XXX TEST THIS
         AGAIN.}  Rectifying this is relatively simple: create a restriction
@@
 
-\subsection{Possible improvements}
+\subsection{Possible Improvements}
 
@@
 came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create regexes for new
-rules.  Detecting, diagnosing, and defeating complications forms the largest
-single portion of this section, mirroring the development of the parser.
-The complications are described in the order they were overcome, with
+algorithm for analysing unparsed log lines to create regexes for new rules.
+Detecting, diagnosing, and defeating complications forms the largest single
+portion of this section, mirroring the development of the parser.  The
+complications are described in the order they were overcome, with
 subsequent problems affecting fewer mails (often by an order or magnitude),

### 2009-02-12 doc/logparser-introduction.tex
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

@@
 messages can be added by the administrator with custom checks; every
-\gls{DNSBL}\footnote{This document is supplied with a glossary, see
+\acronym{DNSBL}\footnote{This document is supplied with a glossary, see
 \textsection\ref{Glossary}.} returns a different explanatory message;
@@
 paper documents the difficult process of parsing Postfix log files,
-presenting \gls{PLP}, a program that parses Postfix log files and places
-the resulting data into a database for later use.  The gathered data can
-then be used to optimise current anti-spam measures, provide a baseline to
-test new anti-spam measures against, or to produce statistics showing how
-effective those measures are.  Numerous other uses are possible for such
-data: improving server performance by identifying troublesome destinations
-and reconfiguring appropriately; identifying regular high volume uses
-(e.g.\ customer newsletters) and restricting those uses to off-peak times;
-detecting virus outbreaks that propagate via email; as a base for billing
-customers on a shared server.  Preserving the raw data enables users to
-develop a multitude of uses far beyond those conceived of by the author.
+presenting \parsername{}, a program that parses Postfix log files and
+places the resulting data into a database for later use.  The gathered data
+can then be used to optimise current anti-spam measures, provide a baseline
+to test new anti-spam measures against, or to produce statistics showing
+how effective those measures are.  Numerous other uses are possible for
+such data: improving server performance by identifying troublesome
+destinations and reconfiguring appropriately; identifying regular high
+volume uses (e.g.\ customer newsletters) and restricting those uses to
+off-peak times; detecting virus outbreaks that propagate via email; as a
+base for billing customers on a shared server.  Preserving the raw data
+enables users to develop a multitude of uses far beyond those conceived of
+by the author.
 
@@
 understanding the thesis, parser, and architecture.  It introduces the idea
-of using a database schema as an \gls{API}, to provide a language-neutral
-interface to the data gathered.  The novel separation of rules, actions,
-and framework is discussed, giving the reasons that approach was taken when
-designing the parser.
+of using a database schema as an \acronym{API}, to provide a
+language-neutral interface to the data gathered.  The novel separation of
+rules, actions, and framework is discussed, giving the reasons that
+approach was taken when designing the parser.
 
@@
 listed are some additional resources expected to be helpful in
-understanding \gls{SMTP}, Postfix, anti-spam techniques, or the thesis.
+understanding \acronym{SMTP}, Postfix, anti-spam techniques, or the thesis.
 

### 2009-02-12 doc/logparser-other-parsers-review.tex
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

@@
     \item The parsers reviewed all produce a report of varying complexity
-        and detail, whereas \gls{PLP} does not; it extracts data and leaves
-        generation of reports from the data to other programs.  Using an
-        \gls{SQL} database simplifies the process of generating such
-        reports (discussed in \sectionref{database as API}); some sample
-        queries are given in \sectionref{motivation}.  The parser developed
-        for this project is designed to enable much more detailed log file
-        analysis by providing a stable platform for subsequent programs to
-        develop upon.
+        and detail, whereas \acronym{PLP} does not; it extracts data and
+        leaves generation of reports from the data to other programs.
+        Using an \acronym{SQL} database simplifies the process of
+        generating such reports (discussed in \sectionref{database as
+        API}); some sample queries are given in \sectionref{motivation}.
+        The parser developed for this project is designed to enable much
+        more detailed log file analysis by providing a stable platform for
+        subsequent programs to develop upon.
 
@@
 \sectionref{connections table} and \sectionref{results table}; for
-convenience that list is repeated here: client and server \gls{IP} address
-and hostname, HELO hostname, queueid, start time, end time, \gls{SMTP}
-code, sender, recipient, size, message ID\@.
+convenience that list is repeated here: client and server \acronym{IP}
+address and hostname, HELO hostname, queueid, start time, end time,
+\acronym{SMTP} code, sender, recipient, size, message ID\@.
 
@@
 log files: \textit{Log Mail Analyzer: Architecture and Practical
-Utilizations\/}~\cite{log-mail-analyser}.  The aim of \gls{LMA} is quite
-different from \parsername{}: it attempts to present correlated data from
-log files in a form suitable for a systems administrator to search using
-the myriad of standard Unix text processing utilities already available.
-It produces a \gls{CSV} file and either a MySQL
+Utilizations\/}~\cite{log-mail-analyser}.  The aim of \acronym{LMA} is
+quite different from \parsername{}: it attempts to present correlated data
+from log files in a form suitable for a systems administrator to search
+using the myriad of standard Unix text processing utilities already
+available.  It produces a \acronym{CSV} file and either a MySQL
 (\url{http://www.mysql.com/}) or Berkeley DB
 (\url{http://www.oracle.com/database/berkeley-db/index.html}) database.
-The decision to support both \gls{CSV} and Berkeley DB appears to have been
-a serious limitation: both formats have limitations that will be explained
-later.  Hardly any documentation is provided with \gls{LMA}, though some
-documentation is available in~\cite{log-mail-analyser}.  Studying the
-source code is informative, though this author had difficulty as the
-authors of \gls{LMA} wrote in Italian.
+The decision to support both \acronym{CSV} and Berkeley DB appears to have
+been a serious limitation: both formats have limitations that will be
+explained later.  Hardly any documentation is provided with \acronym{LMA},
+though some documentation is available in~\cite{log-mail-analyser}.
+Studying the source code is informative, though this author had difficulty
+as the authors of \acronym{LMA} wrote in Italian.
 
-\gls{CSV} is a very simple format where each record is stored in a single
-line, with fields separated by a comma or other punctuation symbol.
-Problems with \gls{CSV} files include the need to escape separators in the
-data stored, providing multiple values for a field (e.g.\ multiple
-recipients), and adding new fields.  \gls{CSV} files do not have a standard
-mechanism to document the fields or the separator, unlike \gls{SQL}
-databases where every database includes a schema naming the fields and the
-type of data they store (e.g.\ integer, text, timestamp).  The \gls{CSV}
-record format is not documented, but the output file contains a comment
-giving the format:\newline{} \texttt{\# Timestamp|Nome Client|IP Client|IP
-Server|From|To|Status|Size} \newline{}\gls{LMA} treats \gls{CSV} lines
-starting with \texttt{\#} as comments, but not all \gls{CSV} parsers will.
+\acronym{CSV} is a very simple format where each record is stored in a
+single line, with fields separated by a comma or other punctuation symbol.
+Problems with \acronym{CSV} files include the need to escape separators in
+the data stored, providing multiple values for a field (e.g.\ multiple
+recipients), and adding new fields.  \acronym{CSV} files do not have a
+standard mechanism to document the fields or the separator, unlike
+\acronym{SQL} databases where every database includes a schema naming the
+fields and the type of data they store (e.g.\ integer, text, timestamp).
+The \acronym{CSV} record format is not documented, but the output file
+contains a comment giving the format:\newline{} \texttt{\# Timestamp|Nome
+Client|IP Client|IP Server|From|To|Status|Size} \newline{}\acronym{LMA}
+treats \acronym{CSV} lines starting with \texttt{\#} as comments, but not
+all \acronym{CSV} parsers will.
 
 Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
-unlike \gls{SQL} databases that store arbitrary tuples.  In \gls{LMA}'s
-main table the key is an integer referred to by secondary tables, and the
-value is a \gls{CSV} line containing all of the data for that row.  The
-secondary by-sender, by-recipient, by-date, and by-\gls{IP} tables use the
-sender/recipient/date/\gls{IP} address as the key, and the value is a
-\gls{CSV} list of integers referring to the main table.  This effectively
-re-implements \gls{SQL} foreign keys, but without the functionality offered
-by even the most basic of \gls{SQL} databases (e.g.\ joins, ordering,
-searches).  It also requires custom code to search on some combination of
-the above, though the authors of \gls{LMA} did provide some queries:
-IP-STORY, FROM-STORY, DAILY-EMAIL, and DAILY-REJECT\@.  Berkeley DB appears
-to be the least useful of the three output formats: it does not provide the
-functionality of a basic \gls{SQL} database, and unlike \gls{CSV} files it
-cannot be used with standard Unix text processing tools.
+unlike \acronym{SQL} databases that store arbitrary tuples.  In
+\acronym{LMA}'s main table the key is an integer referred to by secondary
+tables, and the value is a \acronym{CSV} line containing all of the data
+for that row.  The secondary by-sender, by-recipient, by-date, and
+by-\acronym{IP} tables use the sender/recipient/date/\acronym{IP} address
+as the key, and the value is a \acronym{CSV} list of integers referring to
+the main table.  This effectively re-implements \acronym{SQL} foreign keys,
+but without the functionality offered by even the most basic of
+\acronym{SQL} databases (e.g.\ joins, ordering, searches).  It also
+requires custom code to search on some combination of the above, though the
+authors of \acronym{LMA} did provide some queries: IP-STORY, FROM-STORY,
+DAILY-EMAIL, and DAILY-REJECT\@.  Berkeley DB appears to be the least
+useful of the three output formats: it does not provide the functionality
+of a basic \acronym{SQL} database, and unlike \acronym{CSV} files it cannot
+be used with standard Unix text processing tools.
 
 The schema used with the MySQL database is undocumented, but at least it is
-possible to discover the schema with an existing \gls{SQL} database, unlike
-with Berkeley DB\@; all \gls{SQL} databases embed the schema into the
-database and provide commands for displaying it.  Berkeley DB does not
-embed a schema in its files, because there is neither requirement nor
+possible to discover the schema with an existing \acronym{SQL} database,
+unlike with Berkeley DB\@; all \acronym{SQL} databases embed the schema
+into the database and provide commands for displaying it.  Berkeley DB does
+not embed a schema in its files, because there is neither requirement nor
 benefit; it only provides \textbf{(key, value)} pairs, so any additional
@@
 Whether a MySQL database or Berkeley DB table is chosen in addition to the
-\gls{CSV} output, \gls{LMA} stores the following data: time and date,
-client hostname and \gls{IP} address, server \gls{IP} address, sender and
-recipient addresses, \gls{SMTP} code, and size (for accepted mails only).
-It is unclear what time and date is stored: start time, end time, or
-delivery time?  Unlike \parsername{} it does not store the server hostname,
-HELO hostname, queueid, start and end times, timestamps for each log line,
-or message id (for accepted mails only).  Handling of multiple recipients,
-\gls{SMTP} codes, or remote servers\footnote{A single mail may be sent to
-multiple remote servers if it was addressed to recipients in different
-domains, or Postfix needs to try multiple servers for one or more
-recipients.} is not explained; experimental observation shows that multiple
-records are added when a mail had multiple recipients (sadly the records are
-not associated or linked in any way), and presumably the same approach is
-taken when there were multiple destination servers.
+\acronym{CSV} output, \acronym{LMA} stores the following data: time and
+date, client hostname and \acronym{IP} address, server \acronym{IP}
+address, sender and recipient addresses, \acronym{SMTP} code, and size (for
+accepted mails only).  It is unclear what time and date is stored: start
+time, end time, or delivery time?  Unlike \parsername{} it does not store
+the server hostname, HELO hostname, queueid, start and end times,
+timestamps for each log line, or message id (for accepted mails only).
+Handling of multiple recipients, \acronym{SMTP} codes, or remote
+servers\footnote{A single mail may be sent to multiple remote servers if it
+was addressed to recipients in different domains, or Postfix needs to try
+multiple servers for one or more recipients.} is not explained;
+experimental observation shows that multiple records are added when a mail
+had multiple recipients (sadly the records are not associated or linked in
+any way), and presumably the same approach is taken when there were
+multiple destination servers.
 
-\gls{LMA} requires major changes to the parser code to parse new log lines
+\acronym{LMA} requires major changes to the parser code to parse new log lines
 or to extract additional data.  The code is structured as a long series of
@@
 
-\gls{LMA} does not deal with any of the complications discussed in
+\acronym{LMA} does not deal with any of the complications discussed in
 \sectionref{complications}, except for correlating log lines by queueid;
@@
 this project is to retain that data to aid optimisation and evaluation of
-anti-spam techniques.  \gls{LMA} fails to parse Postfix log files generated
-on Solaris hosts because the fields automatically prepended to each log
-line differ from those added on Linux hosts; log files from Solaris hosts
-(and possibly other operating systems) thus require preprocessing before
-parsing by \gls{LMA}.  The \numberOFlogFILES{} preprocessed test log files
-were parsed without complaint by \gls{LMA}, although it produced 32 entries
-in the output \gls{CSV} file for every rejection in the input log file; it
-also missed some 40\% of delivered mail.  Once these glaring deficiencies
-were discovered the author did not waste any more time checking the
-results.
+anti-spam techniques.  \acronym{LMA} fails to parse Postfix log files
+generated on Solaris hosts because the fields automatically prepended to
+each log line differ from those added on Linux hosts; log files from
+Solaris hosts (and possibly other operating systems) thus require
+preprocessing before parsing by \acronym{LMA}.  The \numberOFlogFILES{}
+preprocessed test log files were parsed without complaint by \acronym{LMA},
+although it produced 32 entries in the output \acronym{CSV} file for every
+rejection in the input log file; it also missed some 40\% of delivered
+mail.  Once these glaring deficiencies were discovered the author did not
+waste any more time checking the results.
 
-\gls{LMA} does provide some simple reports: IP-STORY, FROM-STORY,
+\acronym{LMA} does provide some simple reports: IP-STORY, FROM-STORY,
 DAILY-EMAIL and DAILY-REJECT\@.  These reports search the Berkeley DB files
-for matching records: the first three extract \gls{CSV} lines for the
-specified client \gls{IP} address, sender address, or date respectively.
-DAILY-REJECT initially failed with an error message from the Perl
-interpreter;\footnote{The error messages were: \newline{}\texttt{Undefined
-subroutine \&main::LIST called at queryDB.pl line
-372.}\newline{}\texttt{Undefined subroutine \&main::EXTRACT\_FROM\_DB
+for matching records: the first three extract \acronym{CSV} lines for the
+specified client \acronym{IP} address, sender address, or date
+respectively.  DAILY-REJECT initially failed with an error message from the
+Perl interpreter;\footnote{The error messages were:
+\newline{}\texttt{Undefined subroutine \&main::LIST called at queryDB.pl
+line 372.}\newline{}\texttt{Undefined subroutine \&main::EXTRACT\_FROM\_DB
 called at queryDB.pl line 379.}} after making corrections to the code it
-worked, extracting the \gls{CSV} lines for the specified day where the
-\gls{SMTP} code signifies a rejection.  All of these reports are trivially
-simple to produce from the \gls{CSV} file using the standard Unix tool
-\texttt{awk}\glsadd{awk}; the most complicated, DAILY-REJECT, is merely:
+worked, extracting the \acronym{CSV} lines for the specified day where the
+\acronym{SMTP} code signifies a rejection.  All of these reports are
+trivially simple to produce from the \acronym{CSV} file using the standard
+Unix tool \texttt{awk}\glsadd{awk}; the most complicated, DAILY-REJECT, is
+merely:
 
@@
 
-    \item DAILY-REJECT considers all \gls{SMTP} codes except \texttt{250}
-        to be rejections; this includes invalid \gls{SMTP} codes such as
-        \texttt{0} and \texttt{deferred}, so the awk command does too.
-        These invalid \gls{SMTP} codes are most likely present because of
-        incorrect parsing by \gls{LMA}.
+    \item DAILY-REJECT considers all \acronym{SMTP} codes except
+        \texttt{250} to be rejections; this includes invalid \acronym{SMTP}
+        codes such as \texttt{0} and \texttt{deferred}, so the awk command
+        does too.  These invalid \acronym{SMTP} codes are most likely
+        present because of incorrect parsing by \acronym{LMA}.
 
-    \item \gls{LMA} produces 32 lines in its \gls{CSV} file for every
-        single line it should have produced; the command above suppresses
-        duplicate sequential lines.  DAILY-REJECT produces the correct
-        number of output lines, probably because it uses the Berkeley DB
-        files as input rather than the \gls{CSV} file.
+    \item \acronym{LMA} produces 32 lines in its \acronym{CSV} file for
+        every single line it should have produced; the command above
+        suppresses duplicate sequential lines.  DAILY-REJECT produces the
+        correct number of output lines, probably because it uses the
+        Berkeley DB files as input rather than the \acronym{CSV} file.
 
@@
     \item The output from DAILY-REJECT is missing some records that are
-        present in the \gls{CSV} file; this may be because it uses the
+        present in the \acronym{CSV} file; this may be because it uses the
         Berkeley DB files instead, and there may be differences between the
@@
 
-In summary, \gls{LMA} appears to be a proof of concept, written to
+In summary, \acronym{LMA} appears to be a proof of concept, written to
 demonstrate the point of their paper (that having this information in an
@@
 % Hard to extend/expand/understand.  Appears to only save: date and hour,
-% DNS name and \gls{IP} address host, mail server \gls{IP} address, sender,
-% receiver and e-mail status (sent, rejected).  Undocumented schema.
-% Design decision to use \gls{CSV} as an intermediate format between the
-% log file and the database seems to have been restrictive.  Appears to
-% require a queueid but majority of log lines (e.g.\ rejections) lack a
-% queueid.  Supports whitelisting \gls{IP} addresses when parsing logs, but
-% whitelisting when generating reports/data mining would be preferable.
-% Supporting Berkeley DB is probably limiting the software - an example is
-% the difficulty in searching a pipe-delimited string, so they have
-% re-implemented foreign keys with tables keyed by ip address etc.\
-% pointing at the main table - this also will not scale well.  There does
-% not appear to be any attempt to deal with the complications I have
+% DNS name and \acronym{IP} address host, mail server \acronym{IP} address,
+% sender, receiver and e-mail status (sent, rejected).  Undocumented
+% schema.  Design decision to use \acronym{CSV} as an intermediate format
+% between the log file and the database seems to have been restrictive.
+% Appears to require a queueid but majority of log lines (e.g.\ rejections)
+% lack a queueid.  Supports whitelisting \acronym{IP} addresses when
+% parsing logs, but whitelisting when generating reports/data mining would
+% be preferable.  Supporting Berkeley DB is probably limiting the software;
+% an example is the difficulty in searching a pipe-delimited string, so
+% they have re-implemented foreign keys with tables keyed by ip address
+% etc.\ pointing at the main table - this also will not scale well.  There
+% does not appear to be any attempt to deal with the complications I have
 % encountered: their parsing is not detailed enough to encounter them.  It
@@
 the easily extensible design of \parsername{}.  It does not correlate log
-lines by queueid or \gls{pid}, and does not need to deal with the
+lines by queueid or \acronym{pid}, and does not need to deal with the
 complications encountered during this project (\sectionref{complications}).
@@
         Fields extracted: from, to, server, UID, relay, status, number of
-        recipients, origin hostname, origin \gls{IP} address, and virus.
-        It also counts the number of and total size of all mails delivered.
-        The fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
-        \texttt{virus} are not explained in the documentation:
-        \texttt{server} is probably the hostname or \gls{IP} address of the
-        server the mail is delivered to; \texttt{relay} might be the
-        delivery method: \gls{SMTP}, local delivery, or \gls{LMTP};
-        \texttt{uid} might be the uid of the user submitting mail locally.
-        Postfix does not do any form of virus checking (though it has many
-        options for cooperating with an external virus scanner), so the
-        \texttt{virus} field is a mystery.
+        recipients, origin hostname, origin \acronym{IP} address, and
+        virus.  It also counts the number of and total size of all mails
+        delivered.  The fields \texttt{server}, \texttt{uid},
+        \texttt{relay}, and \texttt{virus} are not explained in the
+        documentation: \texttt{server} is probably the hostname or
+        \acronym{IP} address of the server the mail is delivered to;
+        \texttt{relay} might be the delivery method: \acronym{SMTP}, local
+        delivery, or \acronym{LMTP}; \texttt{uid} might be the uid of the
+        user submitting mail locally.  Postfix does not do any form of
+        virus checking (though it has many options for cooperating with an
+        external virus scanner), so the \texttt{virus} field is a mystery.
 
@@
         \newline{} Fields extracted: from, to, RBL list, client hostname,
-        and client \gls{IP} address.  It also counts the number and total
-        size of all mails delivered.
+        and client \acronym{IP} address.  It also counts the number and
+        total size of all mails delivered.
 
@@
         \newline{} Fields extracted: from, to, client hostname, client
-        \gls{IP} address, relay hostname, relay \gls{IP} address, status,
-        response code, RBL list, and message id.  It also counts the number
-        and size of all mails delivered, processed, blocked, expired, and
-        bounced.
+        \acronym{IP} address, relay hostname, relay \acronym{IP} address,
+        status, response code, RBL list, and message id.  It also counts
+        the number and size of all mails delivered, processed, blocked,
+        expired, and bounced.
 
@@
 times.  Sawmill does not extract any data about rejections except when the
-rejection is caused by a \gls{DNSBL} check (\texttt{RBL list} in the list
-of fields).
+rejection is caused by a \acronym{DNSBL} check (\texttt{RBL list} in the
+list of fields).
 
@@
 futile attempts, and reading all the available documentation, the author
-admitted defeat.  Using the \gls{CLI} rather than the web interface was
+admitted defeat.  Using the \acronym{CLI} rather than the web interface was
 partially successful: the command \newline{} \tab{} \texttt{splunk find
@@
 Subsequently the author was successful in adding a single file at a time
-using the \gls{CLI}:\newline{} \tab{} \texttt{splunk add tail
+using the \acronym{CLI}:\newline{} \tab{} \texttt{splunk add tail
 }\textit{filename\/}\newline{} A simple loop was then enough to add all the
@@
 any \texttt{key=value} pairs occurring after the standard syslog prologue:
-to and from addresses, HELO hostname, and protocol (\gls{SMTP} or
-\gls{ESMTP}).  \parsername{} extracts noticeably more data (client and
-server \gls{IP} address and hostname, queueid, start and end times,
-timestamps for each log line, \gls{SMTP} code, and message ID), though it
-does not make the full text of the line available (this could be trivially
-added if desired, but would greatly increase the size of the resulting
-database).  The full power of \gls{SQL} is available when searching the
-data extracted by \parsername{}, allowing the user to search on arbitrarily
-complicated conditions.
+to and from addresses, HELO hostname, and protocol (\acronym{SMTP} or
+\acronym{ESMTP}).  \parsername{} extracts noticeably more data (client and
+server \acronym{IP} address and hostname, queueid, start and end times,
+timestamps for each log line, \acronym{SMTP} code, and message ID), though
+it does not make the full text of the line available (this could be
+trivially added if desired, but would greatly increase the size of the
+resulting database).  The full power of \acronym{SQL} is available when
+searching the data extracted by \parsername{}, allowing the user to search
+on arbitrarily complicated conditions.
 
@@
 above extracting the \texttt{key=value} fields from a log line; it makes no
-attempt to correlate log lines by queueid or \gls{pid}, or to handle any of
-the other myriad complications discussed in \sectionref{complications}.
-Its source code is unavailable, so it could not be used as a base for this
-project, even if it fulfilled all other requirements.
+attempt to correlate log lines by queueid or \acronym{pid}, or to handle
+any of the other myriad complications discussed in
+\sectionref{complications}.  Its source code is unavailable, so it could
+not be used as a base for this project, even if it fulfilled all other
+requirements.
 
@@
 but supporting numerous log files formats without special purpose code
-limits its functionality.  The data it can extract from an \gls{MTA} log
-file is limited in comparison to \parsername{}: time2, email, email\_r,
+limits its functionality.  The data it can extract from an \acronym{MTA}
+log file is limited in comparison to \parsername{}: time2, email, email\_r,
 host, host\_r, method, url, code, and bytesd.  No explanation for any of
@@
 
-    \item A summary of the \gls{SMTP} codes used when rejecting delivery
-        attempts.
+    \item A summary of the \acronym{SMTP} codes used when rejecting
+        delivery attempts.
 
@@
 \noindent{}YAALA's Postfix parser extracts some of the fields \parsername{}
-does: for client and server it stores either the \gls{IP} address or the
-hostname, not both; it omits the HELO hostname, queueid, \gls{SMTP} code,
-size of each accepted mail, start and end times, timestamps for each log
-line, and message ID\@.  It extracts some data that \parsername{} does not:
-the delay in delivering each mail, and how many times delivery was deferred
-for each mail; these could easily be extracted by \parsername{} if desired.
-Unlike \parsername{}, YAALA does not maintain separate counters for each
-restriction; this rules out the possibility of using the collected data for
-optimisation, testing, or understanding of restrictions.  YAALA's Postfix
-parser does not deal with the complications explained in
+does: for client and server it stores either the \acronym{IP} address or
+the hostname, not both; it omits the HELO hostname, queueid, \acronym{SMTP}
+code, size of each accepted mail, start and end times, timestamps for each
+log line, and message ID\@.  It extracts some data that \parsername{} does
+not: the delay in delivering each mail, and how many times delivery was
+deferred for each mail; these could easily be extracted by \parsername{} if
+desired.  Unlike \parsername{}, YAALA does not maintain separate counters
+for each restriction; this rules out the possibility of using the collected
+data for optimisation, testing, or understanding of restrictions.  YAALA's
+Postfix parser does not deal with the complications explained in
 \sectionref{complications}, except it does correlate log lines by queueid.
 
-YAALA provides a mini-language based on \gls{SQL} that is used when
+YAALA provides a mini-language based on \acronym{SQL} that is used when
 generating reports; sample reports can be seen
@@
 
-    \noindent{} The email servers' reports will show you the number of deliveries and
-    the volume of email delivered by day, the domains from which you
-    receive or send the most emails, the relays most used, etc.
+    \noindent{} The email servers' reports will show you the number of
+    deliveries and the volume of email delivered by day, the domains from
+    which you receive or send the most emails, the relays most used, etc.
 
@@
 correlates the data by queueid.  This approach will extract the following
-data: HELO hostname, queueid, \gls{SMTP} code, sender and recipient
+data: HELO hostname, queueid, \acronym{SMTP} code, sender and recipient
 addresses, and size of accepted mails.  It is unclear if the parser will
 extract any further data.  Lire misses the following fields extracted by
-\parsername{}: client and server \gls{IP} address and hostname, start and
-end times, timestamps of each log line, and message ID\@.
+\parsername{}: client and server \acronym{IP} address and hostname, start
+and end times, timestamps of each log line, and message ID\@.
 
@@
 possible, with enough time and effort, to understand the format.
-\parsername{} uses an \gls{SQL} database to make accessing the extracted
-data as easy as possible.  In general Lire has poor documentation.
+\parsername{} uses an \acronym{SQL} database to make accessing the
+extracted data as easy as possible.  In general Lire has poor
+documentation.
 
 Similar to AWStats and Logrep, Lire attempts to correlate log lines by
-queueid, but not by \gls{pid}, so the complete list of recipients for a
+queueid, but not by \acronym{pid}, so the complete list of recipients for a
 mail should be available; its parser extracts only part of the available
@@
 
-    \item For mail sent: to addresses, \gls{SMTP} code, and delay.
+    \item For mail sent: to addresses, \acronym{SMTP} code, and delay.
 
@@
 It also counts the number of log lines parsed and skipped.  It omits client
-\gls{IP} address and hostname, server \gls{IP} address, HELO hostname,
-queueid, timestamps of each log line, and message ID\@.  It extracts the
-delay for delivered mails, which \parsername{} does not.  Log lines are
-correlated based on the queueid (called sessionname [sic] within Logrep),
-but not by \gls{pid}.  The parsing is error prone: empty fields are saved
-when the log line does not match the regex, though it appears that they
-will not overwrite existing data.  Most notably rejections are completely
-ignored, making it unsuitable for the purposes of this project.  It does
-not try to address any of the complications in \sectionref{complications}
-except for correlating by queueid.
+\acronym{IP} address and hostname, server \acronym{IP} address, HELO
+hostname, queueid, timestamps of each log line, and message ID\@.  It
+extracts the delay for delivered mails, which \parsername{} does not.  Log
+lines are correlated based on the queueid (called sessionname [sic] within
+Logrep), but not by \acronym{pid}.  The parsing is error prone: empty
+fields are saved when the log line does not match the regex, though it
+appears that they will not overwrite existing data.  Most notably
+rejections are completely ignored, making it unsuitable for the purposes of
+this project.  It does not try to address any of the complications in
+\sectionref{complications} except for correlating by queueid.
 
@@
 and some provide a greater or lesser degree of customisation.  Most have a
-data store, but only \gls{LMA} provides any documentation on its format;
-some deliberately make the data store inaccessible to other tools.  Most
-but not all of the parsers provide documentation, with the quality ranging
-from unusable to very good.  Fewer than half of the parsers were capable of
-parsing the \numberOFlogFILES{} test log files; improving or extending
-parsing would have been quite a difficult task for any of the parsers, and
-one that the author did not have the time to attempt.  \Tableref{Summary of
-parsers' features} provides a summary of the parsers' features.  The
-overriding difference between \parsername{} and the other parsers reviewed
-herein is that none of them aim for the high level of understanding of
-Postfix log files achieved by \parsername{}.
+data store, but only \acronym{LMA} provides any documentation on its
+format; some deliberately make the data store inaccessible to other tools.
+Most but not all of the parsers provide documentation, with the quality
+ranging from unusable to very good.  Fewer than half of the parsers were
+capable of parsing the \numberOFlogFILES{} test log files; improving or
+extending parsing would have been quite a difficult task for any of the
+parsers, and one that the author did not have the time to attempt.
+\Tableref{Summary of parsers' features} provides a summary of the parsers'
+features.  The overriding difference between \parsername{} and the other
+parsers reviewed herein is that none of them aim for the high level of
+understanding of Postfix log files achieved by \parsername{}.
 
@@
         \tablemiddleline{}%
-        \gls{LMA}       & No            & Yes               & No                & Poor           & Yes          \\ 
+        \acronym{LMA}       & No            & Yes               & No                & Poor           & Yes          \\ 
         Pflogsumm       & Yes           & No                & Partial \dag{}    & Good           & Yes          \\

### 2009-02-12 doc/logparser-postfix-daemons.tex
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

@@
 \newglossaryentry{lmtp}{name={lmtp},type={postfix},description={
-    Delivery of mail over \gls{LMTP} is performed by the lmtp component.
-    The definitive documentation is
+    Delivery of mail over \acronym{LMTP} is performed by the lmtp
+    component.  The definitive documentation is
     \url{http://www.postfix.org/lmtp.8.html}.
@@
 \newglossaryentry{smtp}{name={smtp},type={postfix},description={
-    Delivery of mail over \gls{SMTP} is performed by the smtp component.
-    The definitive documentation is
+    Delivery of mail over \acronym{SMTP} is performed by the smtp
+    component.  The definitive documentation is
     \url{http://www.postfix.org/smtp.8.html}.
@@
 \newglossaryentry{smtpd}{name={smtpd},type={postfix},description={
-    Smtpd is the Postfix program that accepts mail via \gls{SMTP}, and
+    Smtpd is the Postfix program that accepts mail via \acronym{SMTP}, and
     implements all the anti-spam restrictions Postfix provides.  The

### 2009-02-12 doc/logparser-results.tex
Dealt with some XXX items.
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.

@@
 \tableref{Number of mails received per day: statistics} show the number of
-mails received by \gls{SMTP} only; in particular the mail loops noticeable
-in other graphs do not contribute to these figures.
+mails received by \acronym{SMTP} only; in particular the mail loops
+noticeable in other graphs do not contribute to these figures.
 
@@
 
-\subsection{Explaining the peaks in log file size}
+\subsection{Explaining the Peaks in Log File Size}
 
@@
 
-\subsection{Rule ordering for efficiency}
+\subsection{Rule Ordering For Efficiency}
 
@@
 
-\subsection{Scalability as the number of rules rises}
+\subsection{Scalability as the Number of Rules Rises}
 
@@
 
-\newpage{} % XXX
-\subsection{Caching compiled regexs}
+\subsection{Caching Compiled Regexs}
 
@@
 
-\subsection{Log lines covered}
+\subsection{Log Lines Covered}
 
@@
 hard because other servers have different restrictions with custom
-messages, \gls{DNSBL} messages change over time, major releases of Postfix
-change warning messages (usually adding more information), etc.,\ so over
-time the log lines drift and change.  \Graphref{rule hits graph} shows the
-number of hits for each rule over all \numberOFlogFILES{} log files.  The
-number of hits per rule is quite unevenly spread, resembling a Power Law
-distribution --- it is obvious that a small number of rules match the vast
-majority of the log lines, and more than half the rules match fewer than
-100 times.
+messages, \acronym{DNSBL} messages change over time, major releases of
+Postfix change warning messages (usually adding more information), etc.,\
+so over time the log lines drift and change.  \Graphref{rule hits graph}
+shows the number of hits for each rule over all \numberOFlogFILES{} log
+files.  The number of hits per rule is quite unevenly spread, resembling a
+Power Law distribution --- it is obvious that a small number of rules match
+the vast majority of the log lines, and more than half the rules match
+fewer than 100 times.
 
@@
 
-\subsection{Mails covered}
+\subsection{Mails Covered}
 
@@
 the best way to detect mails that were not handled properly (many of the
-complications discussed in \sectionref{complications} were
-detected in this way).  The parser issues warnings when it detects any
-errors, some of which may alert the user to a problem, e.g.\ when a queueid
-is reused before the previous mail is fully dealt with, when a queueid or
-\gls{pid} is not found in the state tables,\footnote{There will often be
-warnings about a missing queueid or \gls{pid} in the first few hundred or
-thousand log lines because the earlier log lines for those connections or
-mails are in the previous log file; loading the saved state from the
-previous log file will solve this problem.} or when there are problems
-tracking a child mail (see \sectionref{tracking re-injected mail}).  There
-should be few or no warnings when parsing, and when finished parsing the
-state table should only contain entries for mails that had yet to be
-delivered when the log files ended, or were accepted before the log files
-began.
+complications discussed in \sectionref{complications} were detected in this
+way).  The parser issues warnings when it detects any errors, some of which
+may alert the user to a problem, e.g.\ when a queueid is reused before the
+previous mail is fully dealt with, when a queueid or \acronym{pid} is not
+found in the state tables,\footnote{There will often be warnings about a
+missing queueid or \acronym{pid} in the first few hundred or thousand log
+lines because the earlier log lines for those connections or mails are in
+the previous log file; loading the saved state from the previous log file
+will solve this problem.} or when there are problems tracking a child mail
+(see \sectionref{tracking re-injected mail}).  There should be few or no
+warnings when parsing, and when finished parsing the state table should
+only contain entries for mails that had yet to be delivered when the log
+files ended, or were accepted before the log files began.
 
@@
 adding new rules or modifying existing rules is simplified by the
-separation of rules, actions, and framework.  Improving the logical coverage
-is harder, as the actions taken by Postfix must be reconstructed by the
-author, and the new sequence of actions integrated into the existing model
-without breaking the existing parsing.  Detecting a deficiency in the
+separation of rules, actions, and framework.  Improving the logical
+coverage is harder, as the actions taken by Postfix must be reconstructed
+by the author, and the new sequence of actions integrated into the existing
+model without breaking the existing parsing.  Detecting a deficiency in the
 parsing algorithm is also significantly harder than detecting unparsed log

### 2009-02-13 doc/logparser-implementation.tex
Fixed some 'overfull hbox' problems.

@@
         mail is being delivered to another server.  The format is:
-        \newline{} \tab{} \texttt{ client\_hostname = localhost, client\_ip
+        \newline{} \tab{} \texttt{ client\_hostname = localhost,}
+        \newline{} \tab{} \tab{} \texttt{client\_ip
         = 127.0.0.1} \newline{} i.e.\ semi-colon or comma separated
@@
 
-\subsection{Bounce Notification Mails Delivered Before Their Creation Is
-Logged}
+\subsection{Bounce Notification Mails Are Delivered Before Their Creation
+Is Logged}
 

### 2009-02-13 doc/logparser-results.tex
Fixed some 'overfull hbox' problems.

@@
 
-    \item [optimal]  The most optimal order, according to the hypothesis:
+    \item [Optimal]  The most optimal order, according to the hypothesis:
         rules that match most often will be tried first.
 
-    \item [shuffle] This is intended to represent a randomly ordered rule
+    \item [Shuffle] This is intended to represent a randomly ordered rule
         set.  The rules will be shuffled once before use and will retain
@@
 
-    \item [reverse] Hypothetically the worst order: the most frequently
+    \item [Reverse] Hypothetically the worst order: the most frequently
         matching rules will be tried last.

### 2009-02-16 doc/logparser-background.tex
Remainder of the corrections to the background chapter.

@@
         complicates some features of \acronym{SMTP}, e.g.\ forwarding mail
-        from one company or university to another when a user moves.
+        from one domain to another when a user moves.
 
@@
         second and subsequent delivery attempts from that triple the mail
-        will be accepted.  This blocks spam because maintaining a list of
-        failed addresses and retrying after a temporary failure is
-        uneconomical for a spam sender, but a legitimate mail server must
-        retry deliveries that temporarily failed.  Sadly spam senders are
-        using increasingly complex and well written programs to distribute
-        spam, frequently using an ISP provided \acronym{SMTP} server from a
-        compromised machine on the ISP's network.  Greylisting will slowly
-        become less useful, but it does block a large percentage of spam
-        mail at the moment; the most effective restrictions from the
-        \numberOFlogFILES{} log files used in testing the parser are shown
-        in \tableref{Summary of rejections}.  Greylisting is obviously
-        worth using, at least at the moment, particularly when you take
-        into account Greylisting's position as the final restriction that a
-        mail must overcome:\footnote{Greylisting is the final restriction a
-        mail must overcome in the configuration used on the mail server the
-        log files were obtained from; an administrator is free to use
-        Greylisting wherever in the restriction list they feel is most
-        appropriate for their mail system.} Greylisting only takes effect
-        for mails that have passed every other restriction.  Some problems
-        may be encountered when using greylisting: some servers fail to
-        retry; legitimate mail may be delayed, particularly when coming
-        from a pool of servers.
+        will be accepted.  This blocks spam from some senders because
+        maintaining a list of failed addresses and retrying after a
+        temporary failure is uneconomical for a spam sender, but a
+        legitimate mail server must retry deliveries that failed
+        temporarily.  Sadly spam senders are using increasingly complex and
+        well written programs to distribute spam, frequently using an ISP
+        provided \acronym{SMTP} server from a compromised machine on the
+        ISP's network.  Greylisting will slowly become less useful, but it
+        does block a large percentage of spam mail at the moment; the most
+        effective restrictions from the \numberOFlogFILES{} log files used
+        when generating the results in \sectionref{Results} are shown in
+        \tableref{Summary of rejections}.  The table shows that greylisting
+        is worth using at the moment, particularly when you take into
+        account its position as the final restriction that a mail must
+        overcome in the configuration used on the mail server that
+        generated the log files: on that server greylisting only takes
+        effect for mails that have passed all other restrictions.  Some
+        problems may be encountered when using greylisting: some servers
+        fail to retry after a temporary failure, or legitimate mail may be
+        delayed, particularly when coming from a pool of servers.
 
@@
         a threshold the mail is rejected.  The administrator must manually
-        whitelist clients if they are to bypass a Postfix restriction;
-        using a threshold that requires a delivery attempt to hit several
-        restrictions frees the administrator from whitelisting clients that
-        fall foul of one restriction only.
-
-    \item Rate limiting on a per-sender, per-client or per-recipient basis
-        as performed by Policyd~\cite{policyd}.
+        whitelist clients if they are to bypass a Postfix restriction,
+        whereas using a threshold that requires a delivery attempt to hit
+        several scored restrictions frees the administrator from
+        whitelisting clients that fall foul of one binary restriction only.
 
@@
 
-This section has provided background information on several topics,
-starting with the motivation behind the project, continuing with an
-explanation of the use of a database as an \acronym{API} and an
-introduction to \acronym{SMTP}.  The focus then switched to Postfix, the
-\acronym{MTA} in use, and its restrictions and policy servers.
+This section has provided background information useful in understanding
+this thesis, starting with the motivation behind the project, continuing
+with an introduction to \acronym{SMTP}, and finishing with Postfix, its
+anti-spam restrictions, and its support for policy servers.

### 2009-02-16 doc/logparser.tex
Dealt with some XXX items.
Fixed acronyms so they work properly, get expanded when appropriate, and add the     correct entries to the glossary and acronym appendices. Correct capitalisation of most section and subsection headings. Corrections made to background chapter; more remain to be made.
Fixed some 'overfull hbox' problems.
Remainder of the corrections to the background chapter.

@@
 \makeglossaries{}
+\newcommand{\acronym}[1]{%
+    \gls{#1}\glsadd{#1 glossary}%
+}
 
@@
 \newcommand{\tab}[0]{%
-    \hspace*{2em}%
+    \hspace*{1em}%
 }
@@
 % The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\gls{PLP}}
-\newcommand{\parsernames}{\gls{PLP}'s}
+\newcommand{\parsername}{\acronym{PLP}}
+\newcommand{\parsernames}{\acronym{PLP}'s}
 
@@
 
-XXX GENERATE RESULT\_COLS AND CONNECTION\_COLS AUTOMATICALLY\@.
-
 XXX NEED A NAME FOR THE ARCHITECTURE, AND A DIFFERENT NAME FOR THE
@@
 
-XXX IMPROVE BIBLIOGRAPHY APPEARANCE\@: NAMES LOOK UGLY (E.G.\ P.\
+XXX CHECK URLS FOR BIBLIOGRAPHY ENTRIES, AND UPDATE LAST CHECKED DATES\@.
+
+XXX SHOULD I SCRAP postfix\_action?
+
+XXX ANNOTATE ALL URLS WITH LAST CHECKED DATE\@.
+
+XXX STANDARDISE ON ``email'' OR ``mail'' EVERYWHERE\@?
+
+xXX IMPROVE BIBLIOGRAPHY APPEARANCE\@: NAMES LOOK UGLY (E.G.\ P.\
 \empty{}.\ Porters)\@.
+
+XXX CHECK WITH CARL ABOUT CAPITALISATION IN SECTION AND SUBSECTION
+HEADINGS\@.
 

### 2009-02-16 doc/logparser-architecture.tex
Renamed labels in architecture and implementation chapters and fixed references. Improvements to Makefile, and added new helper targets. Some small improvements to architecture chapter.

@@
 XXX MERGE SOME OF THE PAPER'S CONTENT IN HERE\@.
+
+XXX WHEN FINISHED, CHECK ALL REFERENCES TO ENSURE THEY POINT AT THE CORRECT
+CHAPTER\@.
 
@@
 own log files.  To enable this the architecture is divided into three
-parts:
+parts: framework, actions and rules.  Each will be discussed separately,
+but first an overview:
 
-XXX UPDATE SECTION REFERENCES BELOW\@: THEY NEED TO REFER TO BOTH THE
-ARCHITECTURE AND THE IMPLEMENTATION\@.
+XXX UPDATE SECTION REFERENCES BELOW\@: THEY SHOULD ONLY REFER TO THE
+ARCHITECTURE, THE IMPLEMENTATION CAN REFER BACK TO HERE\@.
 
-\begin{description}
+\begin{eqlist}
 
-    \item [framework] The framework is responsible for loading rules,
-        managing data structures, finding the rule that matches each log
-        line, invoking the correct action, etc\@.  The framework is
-        described in detail in \sectionref{framework}.  XXX EXTEND
-        FRAMEWORK SECTION IF NECESSARY\@.
+    \item [Framework]  The framework is the structure that actions and
+        rules plug into.  It provides the parsing loop, shared data
+        storage, loading and validation of rules, storage of results, and
+        other support functions.  The framework is described in detail in
+        \sectionref{framework in architecture}.
 
-    \item [rules] Rules match individual log lines and determine which
-        actions will be executed.  Rules provide an easily extensible
+    \item [Actions] Each action performs the work required to deal with a
+        single category of inputs, e.g.\ processing data from rejections.
+        Actions are invoked to deal with a log line once it has been
+        identified by the rules: actions modify data structures, handle
+        complications, and cause data to be saved to the database.
+
+    \item [Rules]  The rules are responsible for classifying inputs,
+        specifying the action to invoke and the regex that matches the
+        inputs and extracts data.  Rules provide an easily extensible
         method of associating log lines with actions, and are described in
-        detail in \sectionref{rules}.
+        detail in \sectionref{rules in architecture}.
 
-    \item [actions] Actions are invoked to deal with a log line once it has
-        been identified by the rules: actions modify data structures,
-        handle complications, and cause data to be saved to the database.
-        The actions do the work of reconstructing the journey a mail takes
-        through Postfix.  Details of the actions available in the Postfix
-        parser can be found in \sectionref{actions-in-detail}, and
-        \sectionref{adding new actions} describes the process of adding new
-        actions.
-
-\end{description}
+\end{eqlist}
 
@@
 changes must be preserved during the upgrade, and may clash with changes
-made by the developer (\sectionref{adding new actions} describes safely
-adding new actions).  \parsername{} allows the user to add new rules to the
-database without changing the parsing algorithm, unless the new log lines
-to be parsed require functionality not already provided by the algorithm.
-If the new log lines do require new functionality, new actions can be added
-to the parser without modifying existing actions or other parts of the
-algorithm; only in the rare event that the new actions require support from
-other sections of the code will more extensive changes be required.
+made by the developer (\sectionref{adding new actions in implementation}
+describes safely adding new actions).  \parsername{} allows the user to add
+new rules to the database without changing the parsing algorithm, unless
+the new log lines to be parsed require functionality not already provided
+by the algorithm.  If the new log lines do require new functionality, new
+actions can be added to the parser without modifying existing actions or
+other parts of the algorithm; only in the rare event that the new actions
+require support from other sections of the code will more extensive changes
+be required.
 
@@
 
-\label{rule in architecture}
+\label{rules in architecture}
 
@@
 
-\label{rule characteristics}
+\label{rule characteristics in architecture}
 
@@
 
-\label{overlapping rules}
+\label{overlapping rules in architecture}
 
@@
 
-\label{framework}
+XXX EXTEND FRAMEWORK SECTION IF NECESSARY\@.
+
+\label{framework in architecture}
 

### 2009-02-16 doc/logparser-background.tex
Renamed labels in architecture and implementation chapters and fixed references. Improvements to Makefile, and added new helper targets. Some small improvements to architecture chapter.

@@
 this information is used when determining which rules should be used to
-parse each log line (see \sectionref{rule characteristics} for details).
-Postfix's design is strongly influenced by security concerns: those
-components that interact with other hosts are not
+parse each log line (see \sectionref{rule characteristics in architecture}
+for details).  Postfix's design is strongly influenced by security
+concerns: those components that interact with other hosts are not
 privileged,\footnote{Privilege means the power to perform actions that are

### 2009-02-16 doc/logparser-conclusion.tex
Renamed labels in architecture and implementation chapters and fixed references. Improvements to Makefile, and added new helper targets. Some small improvements to architecture chapter.

@@
 of regexes from unparsed log lines should greatly help users add rules
-to parse previously unparsed log lines (\sectionref{creating new rules}).
+to parse previously unparsed log lines (\sectionref{creating new rules in
+implementation}).
 
-The architecture not only substantially eases the parsing of new log
-lines, it makes adding new actions (\sectionref{adding new actions}) a
-relatively simple task.  The simplicity of adding a new action frees the
-implementor from worrying about how their action might interfere with
-existing parsing, so they can focus on correctly implementing their new
-action.
+The architecture not only substantially eases the parsing of new log lines,
+it makes adding new actions (\sectionref{adding new actions in
+implementation}) a relatively simple task.  The simplicity of adding a new
+action frees the implementor from worrying about how their action might
+interfere with existing parsing, so they can focus on correctly
+implementing their new action.
 
@@
 task, (though alleviated by the tool described in \sectionref{adding new
-actions}), whereas dealing with mails where information appears to be
-missing is much more grueling.  Trawling through the log files, looking for
-something out of the ordinary, possibly hundreds or even thousands of log
-lines away from the last mention of the queueid in question, is extremely
-time consuming and error prone.  Sometimes the task is not to spot the line
-that is unusual, but to spot that a line that is normally present is
-missing, i.e.\ to realise that one line amongst thousands is absent.  In
-all cases the evidence must be used to construct a hypothesis to explain
-the irregularities, and that hypothesis must then be tested in the parser;
-if successful, the parser must be modified to deal with the irregularities,
-without adversely affecting the existing functionality.  The complications
-described in \sectionref{complications} were solved in the order they are
-described in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they are the first
-complications to be dealt with.
+actions in implementation}), whereas dealing with mails where information
+appears to be missing is much more grueling.  Trawling through the log
+files, looking for something out of the ordinary, possibly hundreds or even
+thousands of log lines away from the last mention of the queueid in
+question, is extremely time consuming and error prone.  Sometimes the task
+is not to spot the line that is unusual, but to spot that a line that is
+normally present is missing, i.e.\ to realise that one line amongst
+thousands is absent.  In all cases the evidence must be used to construct a
+hypothesis to explain the irregularities, and that hypothesis must then be
+tested in the parser; if successful, the parser must be modified to deal
+with the irregularities, without adversely affecting the existing
+functionality.  The complications described in \sectionref{complications}
+were solved in the order they are described in, and that order closely
+resembles the frequency with which they occur; the most frequently
+occurring complications dominate the warning messages produced, and so
+naturally they are the first complications to be dealt with.
 

### 2009-02-16 doc/logparser-implementation.tex
Renamed labels in architecture and implementation chapters and fixed references. Improvements to Makefile, and added new helper targets. Some small improvements to architecture chapter.

@@
 XXX SHOULD I SAY ``cleaned up'' EVERYWHERE OR NOT\@?
+
+XXX WHEN FINISHED, CHECK ALL REFERENCES TO ENSURE THEY POINT AT THE CORRECT
+CHAPTER\@.
 
@@
 aid the user in understanding the meaning of the log lines parsed by each
-rule.  The rules are described in detail in \sectionref{rules} but the
-fields are covered in \sectionref{rule attributes}.
+rule.  The rules are described in detail in \sectionref{rules in
+implementation} but the fields are covered in \sectionref{rule attributes}.
 
@@
 
-Rules are discussed in detail in \sectionref{rules}, but the structure of
-the rules table is covered here.  Rules are created by the user, not the
-parser, and will not be modified by the parser (except for the hits and
-hits\_total fields).  Rules parse the individual log lines, extracting data
-to be saved in the connections and results tables, and specifying the
-action to take for that log line.
+Rules are discussed in detail in \sectionref{rules in implementation}, but
+the structure of the rules table is covered here.  Rules are created by the
+user, not the parser, and will not be modified by the parser (except for
+the hits and hits\_total fields).  Rules parse the individual log lines,
+extracting data to be saved in the connections and results tables, and
+specifying the action to take for that log line.
 
@@
         this log line; the possible actions are described later in
-        \sectionref{list of postfix actions}.
+        \sectionref{list of postfix actions in implementation}.
 
@@
         log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions-in-detail}.
+        invoked with can be found in \sectionref{actions in detail in
+        implementation}.
 
@@
         more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules}).
+        (described in \sectionref{overlapping rules in architecture}).
 
@@
 
-\label{rules}
+\label{rules in implementation}
 
@@
 
-\label{example rule}
+\label{example rule in implementation}
 
@@
     \item [action:] The action to be executed if the regex matches
-        successfully.  See \sectionref{actions-in-detail} for details of
-        the actions implemented by the parser, \sectionref{actions in
-        architecture} for the role of actions in the parser architecture.
+        successfully.  See \sectionref{actions in detail in implementation}
+        for details of the actions implemented by the parser,
+        \sectionref{actions in architecture} for the role of actions in the
+        parser architecture.
 
@@
 
-\label{creating new rules}
+\label{creating new rules in implementation}
 
@@
 
+The actions do the work of reconstructing the journey a mail takes through
+Postfix.  Details of the actions available in the Postfix parser can be
+found in \sectionref{actions in detail in implementation}, and
+\sectionref{adding new actions in implementation} describes the process of
+adding new actions.
+
 \subsection{Actions in Detail}
 
-\label{actions-in-detail}
+\label{actions in detail in implementation}
 
@@
 
-\label{list of postfix actions}
+\label{list of postfix actions in implementation}
 
@@
 
-\label{adding new actions}
+\label{adding new actions in implementation}
 
@@
 
-\label{limitations-improvements}
+\label{limitations and improvements in implementation}
 

### 2009-02-16 doc/logparser-introduction.tex
Fixed some 'overfull hbox' problems.
Renamed labels in architecture and implementation chapters and fixed references. Improvements to Makefile, and added new helper targets. Some small improvements to architecture chapter.

@@
 Section~\ref{background} provides background information useful in
-understanding the thesis, parser, and architecture.  It introduces the idea
-of using a database schema as an \acronym{API}, to provide a
-language-neutral interface to the data gathered.  The novel separation of
-rules, actions, and framework is discussed, giving the reasons that
-approach was taken when designing the parser.
+understanding the thesis, parser, and architecture.
 
@@
 
-Section~\ref{rules} discusses the parsing rules in detail, explaining the
-purpose and usage of each field in a rule, referring to an example rule and
-sample data it matches successfully against.  The pros and cons of
-overlapping rules are considered, including techniques for detecting
-unintentional overlaps.  Rule efficiency concerns are discussed, in
-particular the optimisations used by the algorithm.  The section concludes
-with a description of using the tools provided with the parser to generate
-new rules (specifically the regex in each rule) from unparsed log lines.
+Section~\ref{rules in architecture} discusses the parsing rules in detail,
+explaining the purpose and usage of each field in a rule, referring to an
+example rule and sample data it matches successfully against.  The pros and
+cons of overlapping rules are considered, including techniques for
+detecting unintentional overlaps.  Rule efficiency concerns are discussed,
+in particular the optimisations used by the algorithm.  The section
+concludes with a description of using the tools provided with the parser to
+generate new rules (specifically the regex in each rule) from unparsed log
+lines.
 
@@
 
-Section~\ref{limitations-improvements} lists the limitations of the
-algorithm, then suggests some ways of dealing with them, with the goal of
-improving parsing and reproduction of the journey a mail takes through
-Postfix.
+Section~\ref{limitations and improvements in implementation} lists the
+limitations of the algorithm, then suggests some ways of dealing with them,
+with the goal of improving parsing and reproduction of the journey a mail
+takes through Postfix.
 

### 2009-02-16 doc/logparser-architecture.tex
Improved the first section. Rearranged remaining sections. Started introduction.

@@
 \chapter{Parser Architecture}
+
+\label{parser architecture}
 
@@
 
-XXX NEEDS AN INTRODUCTION\@.
-
-XXX MERGE SOME OF THE PAPER'S CONTENT IN HERE\@.
-
 XXX WHEN FINISHED, CHECK ALL REFERENCES TO ENSURE THEY POINT AT THE CORRECT
 CHAPTER\@.
+
+To avoid cluttering the explanation of the parser architecture with details
+of the difficulties encountered while parsing Postfix log files, the
+description of the parser architecture has been separated from the
+description of \parsername{}, the implementation developed for this
+project.  This chapter presents the architecture that has been designed and
+developed for this project, beginning with the overall architecture and
+design, followed by the three components of the architecture: Framework,
+Actions, and finally Rules.  The discussion in this chapter centers on the
+theoretical, XXX EXTEND WHEN THE CHAPTER IS FINISHED\@.
+
+The practical difficulties of writing a parser for Postfix log files are
+covered in detail in \sectionref{Postfix Parser Implementation}.
 
@@
 
-XXX UPDATE SECTION REFERENCES BELOW\@: THEY SHOULD ONLY REFER TO THE
-ARCHITECTURE, THE IMPLEMENTATION CAN REFER BACK TO HERE\@.
-
 \begin{eqlist}
@@
         storage, loading and validation of rules, storage of results, and
-        other support functions.  The framework is described in detail in
-        \sectionref{framework in architecture}.
+        other support functions.
 
@@
 
-    \item [Rules]  The rules are responsible for classifying inputs,
-        specifying the action to invoke and the regex that matches the
-        inputs and extracts data.  Rules provide an easily extensible
-        method of associating log lines with actions, and are described in
-        detail in \sectionref{rules in architecture}.
+    \item [Rules]  The rules are responsible for classifying inputs; they
+        specify the action to invoke and the regex that matches the inputs
+        and extracts data.  Rules provide an easily extensible method of
+        associating log lines with actions.
 
@@
 
-It may help to think of the rules and actions as components that plug into
-the framework.  The framework tries each rule in turn until is finds one
-which matches the log line, then invokes the action specified by the rule.
+For each input the framework tries each rule in turn until it finds a rule
+that matches the input, then invokes the action specified by that rule.
 
@@
 
-XXX MERGE THE NEXT TWO PARAGRAPHS\@.  PULL CONTENT FROM THE PAPER\@.
-
-Decoupling the parsing rules from the associated actions and framework
-allows new rules to be written and tested without requiring modifications
-to the parser source code (significantly lowering the barrier to entry for
-new or casual users who need to parse new log lines), and greatly
-simplifies framework, actions, and rules.  Decoupling also creates a clear
-separation of functionality: rules handle low level details of identifying
-log lines and extracting data from a log line; actions handle the higher
-level details of following the path a mail takes through Postfix,
-assembling the required data before storing it, dealing with complications
-arising, etc; the framework provides services to actions and stores data.
-
-Decoupling the actions from the framework simplifies both framework and
-actions: the framework provides services to the actions, and does not need
-to deal with the complications that arise, or the task of reconstructing a
-mail's journey through Postfix; actions benefit from having services
-provided by the framework, freeing them to concentrate on the task of
-accurately reconstructing each mail's journey through Postfix and dealing
-with the complications described in \sectionref{complications}.
+Decoupling the parsing rules from their associated actions allows new rules
+to be written and tested without requiring modifications to the parser
+source code, significantly lowering the barrier to entry for casual users
+who need to parse new inputs, e.g.\ part-time systems administrators
+attempting to combat and reduce spam; it also allows companies to develop
+user-extensible parsers without divulging their source code.  Decoupling
+the framework, actions, and rules simplifies all three and creates a clear
+separation of functionality: the framework provides services to the
+actions, but does not need to perform any tasks specific to the input being
+parsed; actions benefit from having services provided by the framework,
+freeing them to concentrate on the task of accurately and correctly
+processing the information provided by rules; rules handle the low level
+details of classifying inputs and extracting data from those inputs.
 
@@
 new rule with the action to invoke and a regex to match the log lines is
-trivial in comparison to understanding a program's entire parsing
-algorithm, identifying the correct location to change and making the
-appropriate changes.  Bear in mind that the changes must be made without
+trivial in comparison to understanding a program's entire parser,
+identifying the correct location to change, and making the appropriate
+changes.  Bear in mind that changes to the parser must be made without
 adversely affecting existing parsing, particularly as there may be edge
@@
 \sectionref{yet-more-aborted-delivery-attempts} describes a complication
-that occurs only four times in \numberOFlogFILES{} log files tested.
-Requiring changes to the parser's code also complicates upgrades, as the
+that occurs only four times in \numberOFlogFILES{} log files.  Requiring
+changes to the parser's source code also complicates upgrades, as the
 changes must be preserved during the upgrade, and may clash with changes
-made by the developer (\sectionref{adding new actions in implementation}
-describes safely adding new actions).  \parsername{} allows the user to add
-new rules to the database without changing the parsing algorithm, unless
-the new log lines to be parsed require functionality not already provided
-by the algorithm.  If the new log lines do require new functionality, new
-actions can be added to the parser without modifying existing actions or
-other parts of the algorithm; only in the rare event that the new actions
-require support from other sections of the code will more extensive changes
-be required.
+made by the developer.  This architecture allows the user to add new rules
+without changing the parser, unless the new log lines require functionality
+not already provided by the existing actions.  If the new log lines do
+require new functionality, new actions can be added to the parser without
+modifying existing actions (\sectionref{adding new actions in
+implementation} describes how to safely add new actions); only in the rare
+event that the new actions require support from other sections of the code
+will more extensive changes be required.
 
 There is some similarity between the parser's design and William Wood's
-\acronym{ATN}~\cite{atns,nlpip}, a tool used in Computational Linguistics
-for creating grammars to parse or generate sentences.  The resemblance
-between \acronym{ATN} and the parser is accidental, but it is obvious that
-the two different approaches share an underlying separation of concerns;
-this appears to be a natural division of responsibility and functionality.
-\Tableref{Similarities between ATN and this architecture} shows the
-similarities.
+\acronym{ATN}~\cite{atns}, used in Computational Linguistics for creating
+grammars to parse or generate sentences.  The resemblance between the two
+(shown in \tableref{Similarities between ATN and this architecture}) is
+accidental, but it is obvious that the two different approaches share a
+similar division of responsibilities, despite having different semantics.
 
@@
 \end{table}
+
+\section{Framework}
+
+XXX PULL CONTENT FROM THE PAPER FOR HERE\@; IT IS BETTER WRITTEN\@.
+
+XXX EXTEND FRAMEWORK SECTION IF NECESSARY\@.
+
+\label{framework in architecture}
+
+The intermingling of log entries from different mails immediately rules out
+the possibility of handling each mail in isolation; the parser must be able
+to handle multiple mails in parallel, each potentially at a different stage
+in its journey, without any interference between mails --- except in the
+minority of cases where intra-mail interference is required.  The best way
+to implement this is to maintain state information for every unfinished
+mail and manipulate the appropriate mail correctly for each log line
+encountered.
+
+This functionality is provided by the framework, which both drives parsing
+overall and provides services to the actions it invokes.  The framework is
+responsible for loading rules from the database and sanity checking them,
+reading log files, matching each rule against the current log line,
+invoking the correct action, maintaining state tables, loading and saving
+state, displaying a progress indicator, and miscellaneous other tasks.
+Actions use services provided by the framework, including storing and
+retrieving data in the state tables, extracting and saving data captured by
+rules, storage of global data, debugging functions, preparing a mail for
+entry in the database, and entering mails in the database.
+
+There is a similarity between this design and the event-driven programming
+paradigm commonly used in GUI programs, where one part of the program
+responds to events (mouse clicks in a GUI program, log lines being matched
+in the parser) and invokes the correct action.
+
+The framework's high level overview could be expressed as (indentation
+denoting flow of control):
+
+\begin{verbatim}
+for each line in the input files:
+    for each rule defined by the user:
+        if this rule matches the input line:
+            perform the action specified by the rule
+            skip the remaining rules
+            process the next input line
+    warn the user that the input line was not parsed
+\end{verbatim}
+
+\section{Actions}
+
+\label{actions in architecture}
+
+XXX TO BE WRITTEN\@.  START WITH THE CONTENT FROM THE PAPER, THEN CHECK THE
+IMPLEMENTATION SECTION\@.
 
@@
 
-\section{Framework}
+\section{Conclusion}
 
-XXX PULL CONTENT FROM THE PAPER FOR HERE\@; IT IS BETTER WRITTEN\@.
-
-XXX EXTEND FRAMEWORK SECTION IF NECESSARY\@.
-
-\label{framework in architecture}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be able
-to handle multiple mails in parallel, each potentially at a different stage
-in its journey, without any interference between mails --- except in the
-minority of cases where intra-mail interference is required.  The best way
-to implement this is to maintain state information for every unfinished
-mail and manipulate the appropriate mail correctly for each log line
-encountered.
-
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
-
-\begin{verbatim}
-for each line in the input files:
-    for each rule defined by the user:
-        if this rule matches the input line:
-            perform the action specified by the rule
-            skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
-\end{verbatim}
-
-\section{Actions}
-
-\label{actions in architecture}
-
-XXX TO BE WRITTEN\@.  START WITH THE CONTENT FROM THE PAPER, THEN CHECK THE
-IMPLEMENTATION SECTION\@.
+XXX TO BE WRITTEN\@.

### 2009-02-16 doc/logparser-architecture.tex
Improved the framework section in the architecture chapter; it still needs work, but I'm moving on to actions.

@@
 theoretical, XXX EXTEND WHEN THE CHAPTER IS FINISHED\@.
-
 The practical difficulties of writing a parser for Postfix log files are
@@
         \tablemiddleline{}%
-        Networks        & Algorithm     & Determines the sequence 
+        Networks        & Overall       & Determines the sequence 
                                           of transitions              \\
-                        &               & or actions that 
+                        & Algorithm     & or actions that 
                                           constitutes a valid input.  \\
-        Transitions     & Actions       & Save data and impose
-                                        conditions the              \\
-                        &               & input must meet to be
-                                        accepted as valid.          \\
+        Transitions     & Actions       & Assembles data and
+                                          imposes conditions          \\
+                        &               & the input must meet to be
+                                          accepted as                 \\
+                        &               & valid.                      \\
         Abbreviations   & Rules         & Responsible for 
@@
 
-XXX PULL CONTENT FROM THE PAPER FOR HERE\@; IT IS BETTER WRITTEN\@.
-
-XXX EXTEND FRAMEWORK SECTION IF NECESSARY\@.
+XXX EXTEND FRAMEWORK SECTION\@: EXPAND ON ALL OF THE POINTS IN THE LIST IN
+THE NEXT PARAGRAPH\@.
 
@@
 
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be able
-to handle multiple mails in parallel, each potentially at a different stage
-in its journey, without any interference between mails --- except in the
-minority of cases where intra-mail interference is required.  The best way
-to implement this is to maintain state information for every unfinished
-mail and manipulate the appropriate mail correctly for each log line
-encountered.
+The framework takes care of miscellaneous support functions and low level
+details of parsing, freeing the programmers writing actions to concentrate
+on writing productive code.  It links actions and rules, allowing either to
+be improved independently of the other.  It provides shared storage to pass
+data between actions, loads and saves state, loads and validates rules,
+manages parsing, invokes actions, tracks how often each rule matches to
+optimise rule ordering (\sectionref{rule ordering for efficiency}), stores
+results of parsing, and miscellaneous other tasks.
+Most parsers will require the same basic functionality from the framework,
+plus some specialised support functions.  The framework is the core of the
+architecture and is deliberately quite simple: the rules deal with the
+variation in inputs, and the actions deal with the intricacies and
+complications encountered when parsing.
 
-This functionality is provided by the framework, which both drives parsing
-overall and provides services to the actions it invokes.  The framework is
-responsible for loading rules from the database and sanity checking them,
-reading log files, matching each rule against the current log line,
-invoking the correct action, maintaining state tables, loading and saving
-state, displaying a progress indicator, and miscellaneous other tasks.
-Actions use services provided by the framework, including storing and
-retrieving data in the state tables, extracting and saving data captured by
-rules, storage of global data, debugging functions, preparing a mail for
-entry in the database, and entering mails in the database.
+The function that finds the rule matching the input and invokes the
+requested action can be expressed in pseudo-code (indentation denoting flow
+of control) as:
 
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-The framework's high level overview could be expressed as (indentation
-denoting flow of control):
+% DO NOT REFORMAT!
 
 \begin{verbatim}
-for each line in the input files:
+for each input:
     for each rule defined by the user: 
-        if this rule matches the input line:
+        if this rule matches the input:
             perform the action specified by the rule
             skip the remaining rules
-            process the next input line
-    warn the user that the input line was not parsed
+            process the next input
+    warn the user that the input was not parsed
 \end{verbatim}

### 2009-02-16 doc/logparser-implementation.tex
Improved the framework section in the architecture chapter; it still needs work, but I'm moving on to actions.

@@
 \chapter{Postfix Parser Implementation}
+
 \label{Postfix Parser Implementation}
@@
 and actions.
+
+The intermingling of log entries from different mails immediately rules out
+the possibility of handling each mail in isolation; the parser must be able
+to handle multiple mails in parallel, each potentially at a different stage
+in its journey, without any interference between mails --- except in the
+minority of cases where intra-mail interference is required.  The best way
+to implement this is to maintain state information for every unfinished
+mail and manipulate the appropriate mail correctly for each log line
+encountered.
+
+There is a similarity between this design and the event-driven programming
+paradigm commonly used in GUI programs, where one part of the program
+responds to events (mouse clicks in a GUI program, log lines being matched
+in the parser) and invokes the correct action.
 
@@
 
-The actions do the work of reconstructing the journey a mail takes through
-Postfix.  Details of the actions available in the Postfix parser can be
-found in \sectionref{actions in detail in implementation}, and
-\sectionref{adding new actions in implementation} describes the process of
-adding new actions.
+The actions reconstruct the journey a mail takes through Postfix.  Details
+of the actions available in the Postfix parser can be found in
+\sectionref{actions in detail in implementation}, and \sectionref{adding
+new actions in implementation} describes the process of adding new actions.
 

### 2009-02-16 doc/logparser-architecture.tex
Started actions section in architecture section, merging content from paper, and moving content to implementation chapter.

@@
 
-XXX TO BE WRITTEN\@.  START WITH THE CONTENT FROM THE PAPER, THEN CHECK THE
-IMPLEMENTATION SECTION\@.
+XXX TO BE WRITTEN --- CHECK THE IMPLEMENTATION SECTION\@.
+
+Each action is a separate procedure written to deal with a particular
+category of input, e.g.\ rejections.  The actions are parser-specific: each
+parser author will need to write the required actions from scratch unless
+extending an existing parser.  It is anticipated that parsers based on this
+architecture will have a high ratio of rules to actions, with the aim of
+having simpler rules and clearer distinctions between the inputs parsed by
+different rules.
+
+The ability to add special purpose actions to deal with difficulties and
+new requirements that are discovered during parser development is one of
+the strengths of this architecture.  Instead of writing a single monolithic
+function that must be modified to support new behaviour, with all the
+attendant risks of adversely affecting the existing parser, when a new
+requirement arises an independent action can be written to satisfy it.
+Sometimes the new action will require the cooperation of other actions,
+e.g.\ to set or check a flag.  There is a possibility of introducing
+failure when modifying existing actions in this way, but the modifications
+will be smaller and occur less frequently than with a monolithic
+architecture, thus failures will be less likely to occur and will be easier
+to test for and diagnose.  The architecture can be implemented in an object
+oriented style, allowing sub-classes to extend or override actions in
+addition to adding new actions; because each action is an independent
+procedure, the sub-class need only modify the action it is overriding,
+rather than reproducing large chunks of functionality.
+
+During development of the Postfix log parser it became apparent that in
+addition to the obvious variety in log lines there were many complications
+to be overcome.  Some were the result of deficiencies in Postfix's logging
+(some of which were rectified by later versions of Postfix); others were
+due to the vagaries of process scheduling, client behaviour, and
+administrative actions.  All were successfully accommodated in the Postfix
+log parser: adding new actions was enough to overcome several of the
+complications; others required modifications to a single existing action to
+work around the difficulties; the remainder were resolved by adapting
+existing actions to cooperate and exchange extra data (via the framework),
+changing their behaviour as appropriate based on that extra data.
+
+Actions may return a modified input line that will be parsed as if read
+from the input stream, allowing for a simplified version of cascaded
+parsing~\cite{cascaded-parsing}.  This powerful facility allows several
+rules and actions to parse a single input, potentially simplifying both
+rules and actions.
 

### 2009-02-16 doc/logparser-implementation.tex
Started actions section in architecture section, merging content from paper, and moving content to implementation chapter.

@@
 
-\subsection{Actions in Detail}
-
-\label{actions in detail in implementation}
+In the Postfix log parser developed for this project there are
+\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
+distribution of rules to actions as shown in \graphref{Distribution of
+rules per action}.  Unsurprisingly, the action with the most associated
+rules is \texttt{DELIVERY\_REJECTED}, the action that handles Postfix
+rejecting a mail delivery attempt; it is followed by \texttt{SAVE\_DATA},
+the action responsible for handling informative log lines, supplementing
+the data gathered from other log lines.  The third most common action is,
+perhaps surprisingly, \texttt{UNINTERESTING}: this action does nothing when
+executed, allowing uninteresting log lines to be parsed without causing any
+effects (it does not imply that the input is ungrammatical or unparsed).
+Generally rules specifying the \texttt{UNINTERESTING} action parse log
+lines that are not associated with a specific mail, e.g.\ notices about
+configuration files changing.  The remaining actions have only one or two
+associated rules: some actions are required to address a deficiency in the
+log files, or a complication that arises during parsing;  other actions
+will only ever have one log line variant, e.g.\ all log lines showing that
+a remote client has connected are matched by a single rule and handled by
+the \texttt{CONNECT} action.
+
+Using the \texttt{CONNECT} action as an example: it creates a new data
+structure in memory for the new client connection, saving the data
+extracted by the rule into it; this data will be entered into the database
+when the mail delivery attempt is complete.  If a data structure already
+exists for the new connection it is treated as a symptom of a bug, and the
+action issues a warning containing the full contents of the existing data
+structure, plus the log line that has just been parsed.
 
@@
 
-All data extracted by the rule will be automatically saved, regardless of
-the action which is invoked.
+\subsection{Actions in Detail}
+
+\label{actions in detail in implementation}
+
+Each action saves all data extracted by the rule's regex, if there is
+enough information in the log line to identify the correct connection to
+save the data to.
 
 XXX MAKE ALL ACTION DESCRIPTIONS FLOW BETTER\@.
-
-The actions:
 
@@
 
-THERE IS NEARLY A 1--1 MAPPING BETWEEN action AND postfix\_action.
+SEE TABLE (PROBABLY ON NEXT PAGE): THERE IS NEARLY A 1--1 MAPPING BETWEEN
+action AND postfix\_action.  NEW ACTIONS COULD BE ADDED TO FORCE A 1--1
+MAPPING, E.G. COMMIT + DISCARDED COULD BECOME A DISCARD ACTION\@.
+MAIL\_QUEUED HAVING INFO AND PROCESSING CAN PROBABLY BE DEALT WITH, THE
+BIGGEST DIFFICULTY WILL BE WITH SAVE\_DATA\@.
 

### 2009-02-22 doc/logparser-acronyms.tex
The beginning of the architecture rules section.

@@
     description={
-        Natural Language Processing~\cite{Wikipedia-nlp} attempts to
-        increase our understanding of the languages normally used by humans
-        (e.g.\ English, Spanish, Japanese), with the goal of writing
-        programs that can understand human languages.
+        Natural Language Processing~\cite{Wikipedia-nlp} is an area of
+        research that attempts to increase our understanding of the
+        languages normally used by humans (e.g.\ English, Spanish,
+        Japanese), with the goal of writing programs that can understand
+        human languages.
     }
 }
+
+\newglossaryentry{FA}{
+    type=\acronymtype,
+    name={FA},
+    description={Finite Automata},
+    first={Finite Automata (FA)}
+}
+\newglossaryentry{FA glossary}{
+    name={Finite Automata},
+    description={
+        Finite automata are computing devices that accept or recognize
+        regular languages.  They lack any form of storage, so while they
+        can recognise languages such as ${(ab)}^{*}$ or
+        $a^{*}{(b^{*}|c)}b$, they cannot count and so cannot recognise
+        languages such as $a^{n}b^{n}$.
+    }
+}
+
+\newglossaryentry{CFL}{
+    type=\acronymtype,
+    name={CFL},
+    description={Context Free Language},
+    first={Context Free Language (CFL)}
+}
+\newglossaryentry{CFL glossary}{
+    name={Context Free Language},
+    description={
+        XXX WRITE THIS DESCRIPTION
+    }
+}
+
+\newglossaryentry{PDA}{
+    type=\acronymtype,
+    name={PDA},
+    description={Push-Down Automata},
+    first={Push-Down Automata (PDA)}
+}
+\newglossaryentry{PDA glossary}{
+    name={Push-Down Automata},
+    description={
+        XXX WRITE THIS DESCRIPTION
+    }
+}
+
+%\newglossaryentry{}{
+%    type=\acronymtype,
+%    name={},
+%    description={},
+%    first={}
+%}
+%\newglossaryentry{}{
+%    name={},
+%    description={
+%    }
+%}

### 2009-02-22 doc/logparser-architecture.tex
The beginning of the architecture rules section.

@@
 
+Rules categorise inputs, specifying both the regex to match against each
+input and the action to invoke when the match is successful.  Parsing new
+inputs is generally achieved by creating a new rule that pairs an existing
+action with a new regex.  Decoupling the rules from the actions and
+framework enables other rule management approaches to be used, e.g.\
+instead of manually adding new rules, machine learning techniques could be
+used to automatically generate new rules.  If this approach was taken the
+choice of machine learning technique would be constrained by the size of
+typical data sets (see \sectionref{Results} XXX CHECK THIS REFERENCE WHEN
+THE RESULTS SECTION IS FINISHED).  Techniques requiring the full data set
+when training would be impractical; Instance Based
+Learning~\cite{instance-based-learning} techniques that automatically
+determine which inputs from the training set are valuable and which inputs
+can be discarded might reduce the data required to a manageable size.  A
+parser might also dynamically create new rules in response to certain
+inputs, e.g.\ diagnostic messages indicating the program which produced the
+input being parsed had read a new configuration file.  These avenues of
+research and development has not been pursued by the author, but could
+easily be undertaken independently.
+
+The architecture does not try to detect overlapping rules: that
+responsibility is left to the author of the rules.  Unintentionally
+overlapping rules lead to inconsistent parsing and data extraction because
+the first matching rule wins, and the order in which rules are tried
+against each input might change between parser invocations.  Overlapping
+rules are frequently a requirement, allowing a more specific rule to match
+some inputs and a more general rule to match the remainder, e.g.\
+separating \acronym{SMTP} delivery to specific sites from \acronym{SMTP}
+delivery to the rest of the world.  Allowing overlapping rules simplifies
+both the general rule and the more specific rule; additionally rules from
+different sources can be combined with a minimum of prior cooperation or
+modification required.  Overlapping rules should have a priority attribute
+to specify their relative ordering; negative priorities may be useful for
+catchall rules.
+
+Decoupling the rules from the actions allows external tools to be written
+to detect overlapping rules.  Traditional regexes are equivalent in
+computational power to \acronym{FA} and can be converted to \acronym{FA},
+so regex overlap can be detected by finding a non-empty intersection of two
+\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
+for example in~\cite{intersection-of-NFA-using-Z}) is: $FA1 \cap{} FA2 =
+\overline{(\overline{FA1} \cup{} \overline{FA2})}$, which has considerable
+computation complexity.  Perl 5.10 regexes are more powerful than
+traditional regexes: it is possible to match correctly balanced brackets
+nested to an arbitrary depth, e.g.\
+\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
+\verb!z<123<pq<>rs>j<r>ml>s!.  Perl 5.10 regexes can maintain an arbitrary
+state stack and are thus equivalent in computational power to \acronym{PDA}
+or \acronym{CFL}, so detecting overlap may require calculating the
+intersection of two \acronym{PDA} or \acronym{CFL}s.  The intersection of
+two \acronym{CFL}s is not closed, i.e.\ the resulting language cannot
+always be parsed by a \acronym{CFL}, so intersection may be intractable in
+some cases e.g.:
+$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
+Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
+intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
+complexity of calculating intersection.  This is certainly not a task to be
+performed every time the parser is used: detecting overlap amongst the
+Postfix log parser's \numberOFrules{} rules would require calculating
+\numberOFruleINTERSECTIONS{} intersections.
+
+The framework requires each rule to have \texttt{action} and \texttt{regex}
+attributes; each implementation is free to add any additional attributes it
+requires.
+
+It is possible to define pathological regexes which fall into two main
+categories: regexes that match every input, and regexes that consume
+excessive amounts of CPU time during matching.  Defining a regex to match
+all inputs is trivial: \verb!/^/! matches the start of every input.
+Usually excessive CPU time is consumed when a regex with a lot of
+alteration and variable quantifiers fails to match, but successful
+matching is generally quite fast (see~\cite{mastering-regular-expressions}
+for in-depth discussion).
+
+XXX RULES CAN HAVE CONDITIONS ATTACHED
+
+\label{comparison against context-free grammars}
+
+In context-free grammar terms the parser rules could be described as:
+
+$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
+\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
+
 \subsection{Rule Characteristics}
@@
 Rule have certain characteristics that may help in understanding the
-parser:
+parser architecture:
+
+XXX IMPROVE THIS SECTION\@; SHOULD IT BE A SUBSECTION\@?
 
@@
         only be used when parsing log lines produced by that
-        program.\footnote{There are also generic rules which are used when
-        parsing log lines produced by any Postfix program, but only if
-        there are also rules specific to that program, and those rules have
-        already have been tried and have failed on the current log line.}
-        Any given rule will only be used to parse a subset of the log
-        lines, and any given log line will only have a subset of the rules
-        tried against it.
+        program.
 
@@
 
-\label{comparison against context-free grammars}
+\subsection{Architecture Characteristics}
 
-In context-free grammar terms the parser rules could be described as:
+\label{Architecture characteristics}
 
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
+\begin{description}
 
-\subsection{Overlapping Rules}
+    \item [Matching rules against inputs is simple:]  The first matching
+        rule determines the action that will be invoked: there is no
+        backtracking to try alternate rules, no attempt is made to pick a
+        \textit{best\/} rule.
 
-\label{overlapping rules in architecture}
+    \item [Line oriented:]  The architecture is line oriented at present:
+        there is no facility for rules to consume more input or push unused
+        input back onto the input stream.  This was not a deliberate design
+        decision, rather a consequence of the line oriented nature of
+        Postfix log files; more flexible approaches could be pursued.
 
-XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?
+    \item [Context-free rules:]  Rules can not take into account past or
+        future inputs.  In context-free grammar terms the parser rules
+        could be described as:
+        \newline{}$\text{\textless{}input\textgreater{}} \mapsto
+        \text{rule-1} | \text{rule-2} | \text{rule-3} | \dots |
+        \text{rule-n}$.
 
-XXX DFA COMPARISON\@.
+    \item [Context-aware actions:] Actions can consult the results (or lack
+        of results) of previous actions during execution, providing some
+        context sensitivity.
 
-The parser does not try to detect overlapping rules; that responsibility is
-left to the author of the rules.  Unintentionally overlapping rules lead to
-inconsistent parsing and data extraction because the order in which rules
-are tried against each line may change between log files, and the first
-matching rule wins.  Overlapping rules are frequently a requirement,
-allowing a more specific rule to match some log lines and a more general
-rule to match the majority, e.g.\ separating \acronym{SMTP} delivery to
-specific sites from \acronym{SMTP} delivery to the rest of the world.  The
-algorithm provides a mechanism for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Negative priorities may be useful
-for catchall rules.
+    \item [Cascaded parsing:] Actions can return a modified input to be
+        parsed as if read from the input stream, allowing for a simplified
+        version of cascaded parsing~\cite{cascaded-parsing}.
 
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
+    \item [Transduction:]  The architecture can be thought of as
+        implementing transduction: it takes data in one form (log files)
+        and transforms it to another form (a database); other formats may
+        be more suitable for other implementations.
 
-\begin{itemize}
+    \item [Similarity to \acronym{NLP}:] \hfill{} \newline{}
+        Unlike traditional parsers such as those used when compiling a
+        programming language, this architecture does not require a fixed
+        grammar specification that inputs must adhere to.  The architecture
+        is capable of dealing with interleaved inputs, out of order inputs,
+        and ambiguous inputs where heuristics must be applied --- all have
+        arisen and been successfully accommodated in the Postfix log
+        parser.
 
-    \item Sort by regex and visually inspect the list, e.g.\ with
-        \acronym{SQL} similar to: \textbf{select regex from rules order by
-        regex;}
-
-    \item Compare the results of parsing using sorted, shuffled, and
-        reversed rules (described in \sectionref{rule ordering for
-        efficiency}).  Parse several log files using optimal ordering, then
-        dump a textual representation of the rules, connections, and
-        results tables.  Repeat with shuffled and reversed ordering,
-        starting with a fresh database.  If there are no overlapping rules
-        the tables from each run will be identical; differences indicate
-        overlapping rules.  The rules that overlap can be determined by
-        examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing results
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the log files that are matched by more than
-        one rule.
-
-\end{itemize}
+\end{description}
 

### 2009-02-22 doc/logparser-implementation.tex
The beginning of the architecture rules section.

@@
 
-\subsection{Using a Database as an Application Programming Interface}
+\subsection{Using A Database To Provide An Application Programming Interface}
 
@@
 XXX NEED AN INTRODUCTION\@.
+
+\subsection{Overlapping Rules}
+
+\label{overlapping rules in architecture}
+
+XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?
+
+XXX DFA COMPARISON\@.
+
+The parser does not try to detect overlapping rules; that responsibility is
+left to the author of the rules.  Unintentionally overlapping rules lead to
+inconsistent parsing and data extraction because the order in which rules
+are tried against each line may change between log files, and the first
+matching rule wins.  Overlapping rules are frequently a requirement,
+allowing a more specific rule to match some log lines and a more general
+rule to match the majority, e.g.\ separating \acronym{SMTP} delivery to
+specific sites from \acronym{SMTP} delivery to the rest of the world.  The
+algorithm provides a mechanism for ordering overlapping rules: the priority
+field in each rule (defaults to zero).  Negative priorities may be useful
+for catchall rules.
+
+Detecting overlapping rules is difficult, but the following approaches may
+be helpful:
+
+\begin{itemize}
+
+    \item Sort by regex and visually inspect the list, e.g.\ with
+        \acronym{SQL} similar to: \textbf{select regex from rules order by
+        regex;}
+
+    \item Compare the results of parsing using sorted, shuffled, and
+        reversed rules (described in \sectionref{rule ordering for
+        efficiency}).  Parse several log files using optimal ordering, then
+        dump a textual representation of the rules, connections, and
+        results tables.  Repeat with shuffled and reversed ordering,
+        starting with a fresh database.  If there are no overlapping rules
+        the tables from each run will be identical; differences indicate
+        overlapping rules.  The rules that overlap can be determined by
+        examining the differences in the tables: each result contains a
+        reference to the rule which created it, if the references differ
+        between runs the two rules referenced in the differing results
+        overlap.  Unfortunately this method cannot prove the absence of
+        overlapping rules; it can detect overlapping rules, but only if
+        there are log lines in the log files that are matched by more than
+        one rule.
+
+\end{itemize}
 

### 2009-02-23 doc/logparser-acronyms.tex
Added a graph that utilises the data about the number of rules tried, lines     parsed, etc. Added examples in background section to deal with remaining XXX items. Dealt with some other XXX items, and added some more. Improved the Makefile and fixed some dependency mistakes. Started adding "last checked" dates to all URLs. Improved the introduction in the architecture chapter.

@@
         A DNS Blacklist~\cite{Wikipedia-DNSBL} is a simple collaborative
-        anti-spam technique used to reject or penalise email sent from mail
-        servers reported to have sent large volumes of spam.  To use a
-        DNSBL the mail server makes a DNS request incorporating the IP
-        address of the client; if the requested hostname exists the client
-        is on the DNSBL, and the mail server can decide what course of
-        action to take.
+        anti-spam technique used to reject or penalise mail sent from mail
+        servers believed to be the source of large volumes of spam.  The
+        criteria used when deciding whether or not to include an IP address
+        on a DNSBL vary widely between DNSBLs, so before using one it is
+        very important to check their listing policies.  To use a DNSBL the
+        mail server makes a DNS request incorporating the IP address of the
+        client; if the requested hostname exists the client is on the
+        DNSBL, and the mail server can decide what course of action to
+        take.
     }
@@
         maintenance.  Some information and history can be found at
-        \url{http://en.wikipedia.org/wiki/SQL}; a good introduction can be
-        found at \url{http://philip.greenspun.com/sql/}.
+        \urlLastChecked{http://en.wikipedia.org/wiki/SQL}{XXX}; a good
+        introduction can be found at
+        \urlLastChecked{http://philip.greenspun.com/sql/}{XXX}.
     }

### 2009-02-23 doc/logparser-architecture.tex
Added a graph that utilises the data about the number of rules tried, lines     parsed, etc. Added examples in background section to deal with remaining XXX items. Dealt with some other XXX items, and added some more. Improved the Makefile and fixed some dependency mistakes. Started adding "last checked" dates to all URLs. Improved the introduction in the architecture chapter.

@@
 
-To avoid cluttering the explanation of the parser architecture with details
-of the difficulties encountered while parsing Postfix log files, the
-description of the parser architecture has been separated from the
-description of \parsername{}, the implementation developed for this
-project.  This chapter presents the architecture that has been designed and
-developed for this project, beginning with the overall architecture and
-design, followed by the three components of the architecture: Framework,
-Actions, and finally Rules.  The discussion in this chapter centers on the
-theoretical, XXX EXTEND WHEN THE CHAPTER IS FINISHED\@.
-The practical difficulties of writing a parser for Postfix log files are
-covered in detail in \sectionref{Postfix Parser Implementation}.
+To avoid cluttering the explanation of the parser architecture with the
+details involved in parsing Postfix log files and the description of the
+parser implemented for this project, the two have been separated.  This
+chapter presents the architecture designed and developed for this project,
+beginning with the overall architecture and design, followed by the three
+components of the architecture: Framework, Actions, and Rules, and
+finishes by describing the characteristics of this architecture.
+
+This chapter centers on the theoretical, implementation-independent aspects
+of the architecture; the practical difficulties of writing a parser for
+Postfix log files are covered in detail in \sectionref{Postfix Parser
+Implementation}.
 
@@
 
-\subsection{Rule Characteristics}
-
-\label{rule characteristics in architecture}
-
 Rule have certain characteristics that may help in understanding the
@@
 
-\subsection{Architecture Characteristics}
+\section{Architecture Characteristics}
 

### 2009-02-23 doc/logparser-background.tex
Added a graph that utilises the data about the number of rules tried, lines     parsed, etc. Added examples in background section to deal with remaining XXX items. Dealt with some other XXX items, and added some more. Improved the Makefile and fixed some dependency mistakes. Started adding "last checked" dates to all URLs. Improved the introduction in the architecture chapter.

--- 2009-02-16 doc/logparser-background.tex
@@
 
-Postfix is a \glsfirst{MTA} with the following design aims
-(in order of importance): security, flexibility of configuration,
-scalability, and high performance.  It features extensive optional
-anti-spam restrictions, allowing an administrator to employ those
-restrictions which they judge suitable for their site's needs, rather than
-a fixed set chosen by Postfix's author.  These restrictions can be
-selectively applied, combined, and bypassed on a per-client, per-recipient,
-or per-sender basis, allowing different levels of stricture and
-permissiveness.  Postfix leverages simple lookup tables to support
-arbitrarily complicated user-defined sequences of restrictions and
-exceptions.  Administrators can supply their own rejection messages to make
-it clear to senders exactly why their mail was rejected.  Policy servers
-(\sectionref{policy servers}) provide a simple way to write new
-restrictions without having to modify Postfix's source code.  Unfortunately
-this flexibility has a cost: complexity in the log files generated.
-Although it is relatively simple in most cases to use standard Unix text
-processing utilities to determine the fate of an individual email,
-following the journey an email takes through Postfix can be quite
-difficult.  For most mails the journey is simple and brief, but the
-remaining minority can have quite complex journeys (see
-\sectionref{complications} for details).
+Postfix is a \glsfirst{MTA} with the following design aims (in order of
+importance): security, flexibility of configuration, scalability, and high
+performance.  It features extensive optional anti-spam restrictions,
+allowing an administrator to employ those restrictions which they judge
+suitable for their site's needs, rather than a fixed set chosen by
+Postfix's author.  These restrictions can be selectively applied, combined,
+and bypassed on a per-client, per-recipient, or per-sender basis, allowing
+different levels of stricture and permissiveness.  Postfix leverages simple
+lookup tables to support arbitrarily complicated user-defined sequences of
+restrictions and exceptions.  Administrators can supply their own rejection
+messages to make it clear to senders exactly why their mail was rejected.
+Policy servers (\sectionref{policy servers}) provide a simple way to write
+new restrictions without having to modify Postfix's source code.
+Unfortunately this flexibility has a cost: complexity in the log files
+generated.  Although it is relatively simple in most cases to use standard
+Unix text processing utilities to determine the fate of an individual mail,
+following the journey a mail takes through Postfix can in some cases be
+quite difficult.  For most mails the journey is simple and brief, but the
+remainder can have quite complex journeys (see \sectionref{complications}
+for details).
 
@@
 this information is used when determining which rules should be used to
-parse each log line (see \sectionref{rule characteristics} for details).
+parse each log line (see \sectionref{rules in architecture} for details).
 Postfix's design is strongly influenced by security concerns: those
@@
 continue processing the remainder of the restriction list, allowing
-exceptions to more general rules.  XXX GIVE AN EXAMPLE\@.  When the result
-is the name of another restriction Postfix will evaluate that restriction,
-allowing restrictions to be chosen based on the client \acronym{IP}
-address, client hostname, HELO hostname, sender address, or recipient
-address.  E.g.\ the administrator may require that all clients on the local
-network have valid DNS entries, to prevent people sending mail from unknown
-machines XXX EXPAND THIS WITH DETAILS\@.
+exceptions to more general rules.  E.g.\ the general rule is that machines
+on the local network must authenticate before sending mail, except for the
+web server, because the legacy applications running on it lack
+authentication support; this configuration example shows how:
+
+\begin{verbatim}
+  main.cf:
+  smtpd_client_restrictions =
+    ...
+    permit_sasl_authenticated,
+    check_client_access /etc/postfix/allow_webserver.cidr,
+    ...
+
+  /etc/postfix/allow_webserver.cidr:
+    192.0.2.80/32   dunno
+    192.0.2.0/24    reject "Please authenticate to send mail"
+\end{verbatim}
+
+That example also shows how easy it is to supply a custom rejection
+message.  When the result is the name of another restriction Postfix will
+evaluate that restriction, allowing restrictions to be chosen based on the
+client \acronym{IP} address, client hostname, HELO hostname, sender
+address, or recipient address.  E.g.\ the administrator may require that
+all clients on the local network have valid DNS entries, to prevent people
+sending mail from unknown machines; one example of how to achieve this is:
+
+\begin{verbatim}
+  main.cf:
+  smtpd_client_restrictions =
+    ...
+    check_client_access /etc/postfix/require_dns_entries.cidr,
+    ...
+
+  /etc/postfix/require_dns_entries.cidr:
+    192.0.2.0/24    reject_unknown_client_hostname
+\end{verbatim}
 
@@
 restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions and exceptions.  Postfix tries to protect the administrator
+restrictions, and exceptions.  Postfix tries to protect the administrator
 from misconfiguration in as far as is reasonable, e.g.\ the restriction

### 2009-02-23 doc/logparser-implementation.tex
Added a graph that utilises the data about the number of rules tried, lines     parsed, etc. Added examples in background section to deal with remaining XXX items. Dealt with some other XXX items, and added some more. Improved the Makefile and fixed some dependency mistakes. Started adding "last checked" dates to all URLs. Improved the introduction in the architecture chapter.

@@
         addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one email.
+        not be different sender addresses in the results for one mail.
         Mails sent over the same connection can be distinguished by their

### 2009-02-23 doc/logparser-introduction.tex
Added a graph that utilises the data about the number of rules tried, lines     parsed, etc. Added examples in background section to deal with remaining XXX items. Dealt with some other XXX items, and added some more. Improved the Makefile and fixed some dependency mistakes. Started adding "last checked" dates to all URLs. Improved the introduction in the architecture chapter.

@@
 volume uses (e.g.\ customer newsletters) and restricting those uses to
-off-peak times; detecting virus outbreaks that propagate via email; as a
+off-peak times; detecting virus outbreaks that propagate via mail; as a
 base for billing customers on a shared server.  Preserving the raw data

### 2009-02-23 doc/logparser-other-parsers-review.tex
Fixed some 'overfull hbox' problems.
Added a graph that utilises the data about the number of rules tried, lines     parsed, etc. Added examples in background section to deal with remaining XXX items. Dealt with some other XXX items, and added some more. Improved the Makefile and fixed some dependency mistakes. Started adding "last checked" dates to all URLs. Improved the introduction in the architecture chapter.

@@
 use with other tools.  Searches can be saved for reuse; saved searches can
-be run periodically and the results emailed to a recipient or sent to an
+be run periodically and the results mailed to a recipient or sent to an
 external program for further processing.  The author was unable to save
@@
 
+\sloppy{}%
 \parserblurb{%
@@
 {2009/01/11}
+\fussy{}
 
@@
 
-    \item The pair of correspondents exchanged the highest number of
-        emails.
+    \item The pair of correspondents that exchanged the highest number of
+        mails.
 

### 2009-02-23 doc/logparser-results.tex
Remainder of the corrections to the background chapter.
The beginning of the architecture rules section.
Added a graph that utilises the data about the number of rules tried, lines     parsed, etc. Added examples in background section to deal with remaining XXX items. Dealt with some other XXX items, and added some more. Improved the Makefile and fixed some dependency mistakes. Started adding "last checked" dates to all URLs. Improved the introduction in the architecture chapter.

@@
 \chapter{Results}
+
+\label{Results}
+
+XXX SHOULD THIS CHAPTER BE CALLED ``Results'' OR ``Evaluation''\@?
 
@@
 
+XXX EXPAND THIS PARAGRAPH\@; THERE MAY BE CONTENT IN THE PAPER\@.  REFER TO
+THE NEW GRAPH BELOW TOO\@.
+
 The different distribution of log lines caused by a mail forwarding loop
@@
 number of rules rises}, and \sectionref{Caching compiled regexes}.
+
+\showgraph{build/graph-rules-tried-per-log-line}{Mean number of rules tried
+per log line}{Mean number of rules tried per log line}
 
@@
 
-\subsection{Scalability as the Number of Rules Rises}
+\subsection{Scalability When The Number Of Rules Rises}
 
@@
 when not caching compiled regexes graph}
+
+\newpage{} % XXX
 

### 2009-02-23 doc/logparser.tex
Improved the framework section in the architecture chapter; it still needs work, but I'm moving on to actions.
Updated rule numbers.
Enable pdflatex's kerning features for a better looking document.
The beginning of the architecture rules section.
Added a graph that utilises the data about the number of rules tried, lines     parsed, etc. Added examples in background section to deal with remaining XXX items. Dealt with some other XXX items, and added some more. Improved the Makefile and fixed some dependency mistakes. Started adding "last checked" dates to all URLs. Improved the introduction i

@@
 % $Id$
-% XXX WOULD A SMALLER FONT BE BETTER?
 \documentclass[a4paper,12pt,draft]{report}
@@
 \usepackage[final]{graphicx}
-\usepackage{url}
 % Add the bibliography into the table of contents.
@@
 % Extra footnote functionality, including references to earlier footnotes.
-\usepackage[bottom]{footmisc}
+% Removed for now, because it generates a warning:
+% "LaTeX Warning: Command \@makecol  has changed."
+%\usepackage[bottom]{footmisc}
 % The Euro symbol
@@
 \usepackage{type1cm}
+% Enable nice kerning features.
+\usepackage[final]{microtype}
 
@@
 
+% Add last checked dates to all URLs
+\newcommand{\urlLastChecked}[2]{%
+    \url{#1} (last checked #2)%
+}
+
 \newcommand{\tab}[0]{%
@@
 \newcommand{\numberOFrules}[0]{%
-    169%
+    185%
 }
@@
 \newcommand{\numberOFrulesMINIMUM}[0]{%
-    115%
+    116%
 }
@@
 \newcommand{\numberOFrulesMINIMUMpercentage}[0]{%
-    68.04\%%
+    62.70\%%
 }
@@
 \newcommand{\numberOFrulesMAXIMUMpercentage}[0]{%
-    46.95\%%
+    59.48\%%
 }
@@
     60.72 million%
+}
+
+\newcommand{\numberOFactions}[0]{%
+    18%
+}
+
+\newcommand{\numberOFruleINTERSECTIONS}[0]{%
+    14196%
 }
@@
 
-XXX WHY DOES \\ref{} NOT WORK PROPERLY FOR CHAPTERS IN THE APPENDIX\@?
+XXX WHY DOES $\backslash{}$ref\{\} NOT WORK PROPERLY FOR CHAPTERS IN THE APPENDIX\@?
 
@@
 XXX STANDARDISE ON ``email'' OR ``mail'' EVERYWHERE\@?
-
-xXX IMPROVE BIBLIOGRAPHY APPEARANCE\@: NAMES LOOK UGLY (E.G.\ P.\
-\empty{}.\ Porters)\@.
-
-XXX CHECK WITH CARL ABOUT CAPITALISATION IN SECTION AND SUBSECTION
-HEADINGS\@.
-
-XXX MAYBE DO SOMETHING ABOUT underfull hboxes IN BIBLIOGRAPHY WHEN URLS
-WRAP\@?
 

### 2009-02-23 doc/logparser-introduction.tex
Annotated all URLs with last checked date.

@@
 added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from \url{http://www.postfix.org/}} Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
+with the source code, available from
+\urlLastChecked{http://www.postfix.org/}{2009/02/23}} Further complication
+arises because the set of rejection messages is not fixed: new messages can
+be added by the administrator with custom checks; every
 \acronym{DNSBL}\footnote{This document is supplied with a glossary, see

### 2009-02-23 doc/logparser-other-parsers-review.tex
Annotated all URLs with last checked date.

@@
 available.  It produces a \acronym{CSV} file and either a MySQL
-(\url{http://www.mysql.com/}) or Berkeley DB
-(\url{http://www.oracle.com/database/berkeley-db/index.html}) database.
-The decision to support both \acronym{CSV} and Berkeley DB appears to have
-been a serious limitation: both formats have limitations that will be
-explained later.  Hardly any documentation is provided with \acronym{LMA},
-though some documentation is available in~\cite{log-mail-analyser}.
-Studying the source code is informative, though this author had difficulty
-as the authors of \acronym{LMA} wrote in Italian.
+(\urlLastChecked{http://www.mysql.com/}{2008/11/23}) or Berkeley DB
+(\urlLastChecked{http://www.oracle.com/database/berkeley-db/index.html}{2008/11/23})
+database.  The decision to support both \acronym{CSV} and Berkeley DB
+appears to have been a serious limitation: both formats have limitations
+that will be explained later.  Hardly any documentation is provided with
+\acronym{LMA}, though some documentation is available
+in~\cite{log-mail-analyser}.  Studying the source code is informative,
+though this author had difficulty as the authors of \acronym{LMA} wrote in
+Italian.
 
@@
 
-    \item \url{http://www.thesawmill.co.uk/formats/postfix.html} \newline{}
-        Fields extracted: from, to, server, UID, relay, status, number of
-        recipients, origin hostname, origin \acronym{IP} address, and
-        virus.  It also counts the number of and total size of all mails
-        delivered.  The fields \texttt{server}, \texttt{uid},
+    \item
+        \urlLastChecked{http://www.thesawmill.co.uk/formats/postfix.html}{2008/11/23}
+        \newline{} Fields extracted: from, to, server, UID, relay, status,
+        number of recipients, origin hostname, origin \acronym{IP} address,
+        and virus.  It also counts the number of and total size of all
+        mails delivered.  The fields \texttt{server}, \texttt{uid},
         \texttt{relay}, and \texttt{virus} are not explained in the
@@
 
-    \item \url{http://www.thesawmill.co.uk/formats/postfix_ii.html}
+    \item
+        \urlLastChecked{http://www.thesawmill.co.uk/formats/postfix_ii.html}{2008/11/23}
         \newline{} Fields extracted: from, to, RBL list, client hostname,
@@
 
-    \item \url{http://www.thesawmill.co.uk/formats/beta_postfix.html}
+    \item
+        \urlLastChecked{http://www.thesawmill.co.uk/formats/beta_postfix.html}{2008/11/23}
         \newline{} Fields extracted: from, to, client hostname, client
@@
 Some additional Postfix reports are supposedly available on
-\url{http://www.splunkbase.com/}, but the author was unable to find any
-Postfix reports, or indeed reports for any other log file types: every
-category was empty, even those that the web site claimed had numerous
-reports available.  Many types of graphs can be generated, though most are
-variations of a bar or pie chart, except the bubble and heatmap graphs.  It
-is easy to drill down through the graphs to select a portion of the data
-(e.g.\ select the hour with the largest number of events, then select a
-particular host, and finally a specific sender address).  All searches
-performed using the indexed data returned reasonable results.
+\urlLastChecked{http://www.splunkbase.com/}{2008/11/23}, but the author was
+unable to find any Postfix reports, or indeed reports for any other log
+file types: every category was empty, even those that the web site claimed
+had numerous reports available.  Many types of graphs can be generated,
+though most are variations of a bar or pie chart, except the bubble and
+heatmap graphs.  It is easy to drill down through the graphs to select a
+portion of the data (e.g.\ select the hour with the largest number of
+events, then select a particular host, and finally a specific sender
+address).  All searches performed using the indexed data returned
+reasonable results.
 
@@
 
-Copious documentation is made available on \url{http://www.splunk.com/},
-but poor organisation and sheer abundance makes it extremely hard to find
-useful information.  Searches confusingly tended to return results from old
-documentation rather than new.  In general the documentation appears to
-have been written by someone intimately acquainted with the software, who
-has difficulty understanding how a newcomer would approach tasks or the
-questions they would ask.
+Copious documentation is made available on
+\urlLastChecked{http://www.splunk.com/}{2008/11/23}, but poor organisation
+and sheer abundance makes it extremely hard to find useful information.
+Searches confusingly tended to return results from old documentation rather
+than new.  In general the documentation appears to have been written by
+someone intimately acquainted with the software, who has difficulty
+understanding how a newcomer would approach tasks or the questions they
+would ask.
 
@@
 those fields is provided in the documentation at
-\url{http://awstats.sourceforge.net/docs/awstats_faq.html#MAIL} (in
-contrast \parsername{} fully documents everything it extracts), so the
+\urlLastChecked{http://awstats.sourceforge.net/docs/awstats_faq.html\#MAIL}{2008/11/23}
+(in contrast \parsername{} fully documents everything it extracts), so the
 author could not understand what data is extracted, or determine what data
@@
 server in use over the past 10 years; more information is available at
-\url{http://httpd.apache.org/}.} format log files, for analysis by AWStats'
-HTTP log file parser.  The converting parser only deals with a small
-portion of the log lines generated by Postfix, silently skipping those it
-cannot deal with, and does not distinguish between different types of
-rejection; extending it to parse all log lines would be at least as much
-work as writing a new parser.  It does correlate log lines by queueid (not
-by pid), but it does not deal with any of the other complications described
-in \sectionref{complications}.  AWStats supports saving data, but the
-format of the data store is not documented.  It also supports reading
-compressed log files, but that functionality was not tested.
+\urlLastChecked{http://httpd.apache.org/}{2008/11/23}.} format log files,
+for analysis by AWStats' HTTP log file parser.  The converting parser only
+deals with a small portion of the log lines generated by Postfix, silently
+skipping those it cannot deal with, and does not distinguish between
+different types of rejection; extending it to parse all log lines would be
+at least as much work as writing a new parser.  It does correlate log lines
+by queueid (not by pid), but it does not deal with any of the other
+complications described in \sectionref{complications}.  AWStats supports
+saving data, but the format of the data store is not documented.  It also
+supports reading compressed log files, but that functionality was not
+tested.
 
@@
 
-The Debian project (\url{http://www.debian.org/}) provides a manual page
-with the copy of anteater it distributes, so the author was at least able
-to run anteater with the correct arguments; sadly anteater produced zero
-for every statistic, presumably because it was unsuccessful in parsing the
-log lines.
+The Debian project (\urlLastChecked{http://www.debian.org/}{2008/11/23})
+provides a manual page with the copy of anteater it distributes, so the
+author was at least able to run anteater with the correct arguments; sadly
+anteater produced zero for every statistic, presumably because it was
+unsuccessful in parsing the log lines.
 
@@
 generating reports; sample reports can be seen
-at~\url{http://www.yaala.org/samples.html}.  Example query for HTTP proxy
-servers: \newline{} \tab{} \texttt{requests BY file WHERE host =\~{}
-Google} \newline{} The mini-language is quite limited and cannot be used to
-extract data for external use, merely to create reports.  Only data
-selected by the query will be saved in the data store; other data will be
-discarded, and removed from the data store if already present.
+at~\urlLastChecked{http://www.yaala.org/samples.html}{2008/11/23}.  Example
+query for HTTP proxy servers: \newline{} \tab{} \texttt{requests BY file
+WHERE host =\~{} Google} \newline{} The mini-language is quite limited and
+cannot be used to extract data for external use, merely to create reports.
+Only data selected by the query will be saved in the data store; other data
+will be discarded, and removed from the data store if already present.
 

### 2009-02-23 doc/logparser-postfix-daemons.tex
Annotated all URLs with last checked date.

@@
     Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
+    \urlLastChecked{http://www.postfix.org/bounce.8.html}{2009/02/23}.
 }}
@@
     configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
+    \urlLastChecked{http://www.postfix.org/cleanup.8.html}{2009/02/23}.
 }}
@@
     component.  The definitive documentation is
-    \url{http://www.postfix.org/lmtp.8.html}.
+    \urlLastChecked{http://www.postfix.org/lmtp.8.html}{2009/02/23}.
 }}
@@
     a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
+    procmail (\urlLastChecked{http://www.procmail.org/}{2009/02/23}).  It
+    also handles aliases and processing of a user's \texttt{.forward} file.
+    The definitive documentation is
+    \urlLastChecked{http://www.postfix.org/local.8.html}{2009/02/23}.
 }}
@@
     postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
+    definitive documentation is
+    \urlLastChecked{http://www.postfix.org/pickup.8.html}{2009/02/23}.
 }}
@@
     The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
+    \urlLastChecked{http://www.postfix.org/postdrop.1.html}{2009/02/23}.
 }}
@@
     The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
+    \urlLastChecked{http://www.postfix.org/postsuper.1.html}{2009/02/23}.
 }}
@@
     and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
+    documentation is
+    \urlLastChecked{http://www.postfix.org/qmgr.8.html}{2009/02/23}.
 }}
@@
     Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands that send mail depend on; it executes \daemon{postdrop} to
-    place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
+    (\urlLastChecked{http://www.sendmail.org/}{2009/02/23}) mail submission
+    program that all Unix commands that send mail depend on; it executes
+    \daemon{postdrop} to place a new mail in the queue.  The definitive
+    documentation is
+    \urlLastChecked{http://www.postfix.org/sendmail.1.html}{2009/02/23}.
 }}
@@
     component.  The definitive documentation is
-    \url{http://www.postfix.org/smtp.8.html}.
+    \urlLastChecked{http://www.postfix.org/smtp.8.html}{2009/02/23}.
 }}
@@
     implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
+    definitive documentation is
+    \urlLastChecked{http://www.postfix.org/smtpd.8.html}{2009/02/23}.
 }}
@@
     each.  The definitive documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
+    \urlLastChecked{http://www.postfix.org/virtual.8.html}{2009/02/23}.
 }}

### 2009-02-23 doc/logparser-other-parsers-review.tex
Fixes for overfull hboxes.

@@
 Sawmill has three different Postfix log file parsers, extracting three
-different sets to data:
+different sets of data:
 
@@
 
-    \item
-        \urlLastChecked{http://www.thesawmill.co.uk/formats/postfix.html}{2008/11/23}
-        \newline{} Fields extracted: from, to, server, UID, relay, status,
-        number of recipients, origin hostname, origin \acronym{IP} address,
-        and virus.  It also counts the number of and total size of all
-        mails delivered.  The fields \texttt{server}, \texttt{uid},
+    \item \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/postfix.html}{2008/11/23}.
+        Fields extracted: from, to, server, UID, relay, status, number of
+        recipients, origin hostname, origin \acronym{IP} address, and
+        virus.  It also counts the number of and total size of all mails
+        delivered.  The fields \texttt{server}, \texttt{uid},
         \texttt{relay}, and \texttt{virus} are not explained in the
@@
 
-    \item
-        \urlLastChecked{http://www.thesawmill.co.uk/formats/postfix_ii.html}{2008/11/23}
-        \newline{} Fields extracted: from, to, RBL list, client hostname,
-        and client \acronym{IP} address.  It also counts the number and
-        total size of all mails delivered.
+    \item \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/postfix_ii.html}{2008/11/23}.
+        Fields extracted: from, to, RBL list, client hostname, and client
+        \acronym{IP} address.  It also counts the number and total size of
+        all mails delivered.
 
-    \item
-        \urlLastChecked{http://www.thesawmill.co.uk/formats/beta_postfix.html}{2008/11/23}
-        \newline{} Fields extracted: from, to, client hostname, client
-        \acronym{IP} address, relay hostname, relay \acronym{IP} address,
-        status, response code, RBL list, and message id.  It also counts
-        the number and size of all mails delivered, processed, blocked,
-        expired, and bounced.
+    \item \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/beta_postfix.html}{2008/11/23}.
+        Fields extracted: from, to, client hostname, client \acronym{IP}
+        address, relay hostname, relay \acronym{IP} address, status,
+        response code, RBL list, and message id.  It also counts the number
+        and size of all mails delivered, processed, blocked, expired, and
+        bounced.
 
@@
 
-The Debian project (\urlLastChecked{http://www.debian.org/}{2008/11/23})
-provides a manual page with the copy of anteater it distributes, so the
-author was at least able to run anteater with the correct arguments; sadly
-anteater produced zero for every statistic, presumably because it was
-unsuccessful in parsing the log lines.
+The Debian Linux distribution
+(\urlLastChecked{http://www.debian.org/}{2008/11/23}) provides a manual
+page with the copy of anteater it distributes, so the author was at least
+able to run anteater with the correct arguments; sadly anteater produced
+zero for every statistic, presumably because it was unsuccessful in parsing
+the log lines.
 

### 2009-02-23 doc/logparser-acronyms.tex
Annotated all URLs with last checked date.
Dealt with a couple of XXX items, and removed another couple because they're unimportant or I don't understand them any more.

@@
         maintenance.  Some information and history can be found at
-        \urlLastChecked{http://en.wikipedia.org/wiki/SQL}{XXX}; a good
-        introduction can be found at
-        \urlLastChecked{http://philip.greenspun.com/sql/}{XXX}.
+        \urlLastChecked{http://en.wikipedia.org/wiki/SQL}{2009/02/23}; a
+        good introduction can be found at
+        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
     }
@@
 
+\newglossaryentry{CFG}{
+    type=\acronymtype,
+    name={CFG},
+    description={Context Free Grammar},
+    first={Context Free Grammar (CFG)}
+}
+\newglossaryentry{CFG glossary}{
+    name={Context Free Grammar},
+    description={
+        See Context Free Language.
+    }
+}
+
 \newglossaryentry{CFL}{
@@
     description={
-        XXX WRITE THIS DESCRIPTION
+        A Context Free Language is a language generated or parsed by a
+        \acronym{CFG}.  Every transition in a CFG takes the form
+        $L\rightarrow{}R$, where $L$ is a single non-terminal symbol, and
+        $R$ is a (possibly empty) string of terminal and/or non-terminal
+        symbols.  Every CFG can be recognised by a \acronym{PDA}.
     }
@@
     description={
-        XXX WRITE THIS DESCRIPTION
+        A Push-Down Automation is a computational device similar to a
+        \acronym{FA}, but it can additionally use a stack to store data.
+        A PDA can manipulate the stack during state transitions (adding or
+        removing a single piece of data), and can use 
+        the piece of data on the top of the stack in addition to the input
+        when determining which state transition to take.  PDA can recognise
+        languages of the form $a^{n}b^{n}$, i.e.\ any number of one symbol
+        followed by the same number of another symbol.
     }

### 2009-02-23 doc/logparser-introduction.tex
Fixed the problem where \label{} and \ref{} didn't work properly with glossary chapters.

@@
 
-The bibliography contains references to the resources used in developing
-the algorithm, writing the program, and preparing this thesis.  Also
-listed are some additional resources expected to be helpful in
-understanding \acronym{SMTP}, Postfix, anti-spam techniques, or the thesis.
+The bibliography (appendix~\ref{bibliography}) contains references to the
+resources used in developing the algorithm, writing the program, and
+preparing this thesis.  Also listed are some additional resources expected
+to be helpful in understanding \acronym{SMTP}, Postfix, anti-spam
+techniques, or the thesis.
 

### 2009-02-23 doc/logparser.tex
Annotated all URLs with last checked date.
Fixes for overfull hboxes.
Dealt with a couple of XXX items, and removed another couple because they're unimportant or I don't understand them any more.
Fixed the problem where \label{} and \ref{} didn't work properly with glossary chapters.

@@
 % * nonumberlist disables the list of page numbers after each entry.
-\usepackage[acronym,toc,numberedsection=autolabel,style=list,nonumberlist]{glossaries}
+\usepackage[acronym,toc,numberedsection=nolabel,style=list,nonumberlist]{glossaries}
 \renewcommand{\glspostdescription}[0]{}
@@
 % Add last checked dates to all URLs
-\newcommand{\urlLastChecked}[2]{%
-    \url{#1} (last checked #2)%
+\newcommand{\urlLastChecked}[3][ ]{%
+    \url{#2}#1(last checked #3)%
 }
@@
 
-XXX WHY DOES $\backslash{}$ref\{\} NOT WORK PROPERLY FOR CHAPTERS IN THE APPENDIX\@?
-
-XXX NEED A NAME FOR THE ARCHITECTURE, AND A DIFFERENT NAME FOR THE
-IMPLEMENTATION\@.  Parse::FAR\@???
-
 XXX CHECK ALL ``UNINTERESTING'' RULES TO SEE IF THEY COULD BE USED\@.
 
-XXX CHECK YEARS FOR BIBLIOGRAPHY ENTRIES\@.  WHAT DO I MEAN BY THIS\@?
-
-XXX CHECK URLS FOR BIBLIOGRAPHY ENTRIES, AND UPDATE LAST CHECKED DATES\@.
-
 XXX SHOULD I SCRAP postfix\_action?
-
-XXX ANNOTATE ALL URLS WITH LAST CHECKED DATE\@.
-
-XXX STANDARDISE ON ``email'' OR ``mail'' EVERYWHERE\@?
-
-XXX INVESTIGATE Doug Lenat.
 
@@
 
-\label{Glossary}
+\renewcommand{\glossarypostamble}{\label{Glossary}}
 \printglossary{}
-\label{Acronyms}
+\renewcommand{\glossarypostamble}{\label{Acronyms}}
 \printglossary[type=\acronymtype]{}
-\label{Postfix Daemons}
+\renewcommand{\glossarypostamble}{\label{Postfix Daemons}}
 \printglossary[type=postfix]{}

### 2009-02-24 doc/logparser-implementation.tex
Annotated all URLs with last checked date.
Added delay, delays, and dsn to the results table; they're automatically     extracted and saved because filter_regex() fiddles the existing regex     snippets. Updated the thesis with new result columns, including the comparisons in the     state of the art review. Changed how the acronyms glossary is formatted by adding a new style that uses     eqlist. A couple of minor fixes.

@@
 XXX UPDATE REFERENCES WHEN ARCHITECTURE AND IMPLEMENTATION ARE MORE FLESHED
-OUT\@.
+OUT\@.  XXX WHEN FINISHED, CHECK ALL REFERENCES TO ENSURE THEY POINT AT THE
+CORRECT CHAPTER\@.
 
@@
 XXX SHOULD I SAY ``cleaned up'' EVERYWHERE OR NOT\@?
-
-XXX WHEN FINISHED, CHECK ALL REFERENCES TO ENSURE THEY POINT AT THE CORRECT
-CHAPTER\@.
 
@@
 \label{database as API}
-
-XXX LAST CHECKED DATES FOR ALL URLS\@.
 
@@
         introduction to \acronym{SQL} can be found at
-        \url{http://philip.greenspun.com/sql/}, others are
-        \url{http://www.w3schools.com/sql/default.asp},
-        \url{http://sqlcourse.com/}).  Although every vendor implements a
-        different dialect of \acronym{SQL}, the basics are the same
-        everywhere (analogous to the overall similarities and minor
-        differences amongst Irish English, British English, American
-        English, and Australian English).  Depending on the database in use
-        there may be tools available that reduce or remove the requirement
-        to know \acronym{SQL}; several are available for \gls{SQLite}, the
-        default database used by the parser implementation.
+        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23},
+        others are
+        \urlLastChecked{http://www.w3schools.com/sql/default.asp}{2009/02/23},
+        \urlLastChecked{http://sqlcourse.com/}{2009/02/23}).  Although
+        every vendor implements a different dialect of \acronym{SQL}, the
+        basics are the same everywhere (analogous to the overall
+        similarities and minor differences amongst Irish English, British
+        English, American English, and Australian English).  Depending on
+        the database in use there may be tools available that reduce or
+        remove the requirement to know \acronym{SQL}; several are available
+        for \gls{SQLite}, the default database used by the parser
+        implementation.
 
@@
 
+    \item [dsn] The \acronym{DSN} code, similar to the \acronym{SMTP} code;
+        it will only be present for log lines indicating delivery or
+        rejection.  \acronym{DSN} support was added to Postfix in version
+        2.3; logs generated by previous versions will not contain any DSN
+        information.
+
     \item [sender] The sender's email address.  Because multiple mails may
@@
         delivered mails.
+
+    \item [delay] How long the mail was delayed for while it was being
+        delivered.  This will only be present for delivered mails.
+
+    \item [delays] More details information about how long the mail was
+        delayed for while it was being delivered  This will only be present
+        for delivered mails..
 
@@
 \begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[192.0.2.1]: 550
+NOQUEUE: reject: RCPT from example.com[192.0.2.1]: 550 5.1.8
   <foo@example.com>: Sender address rejected: Domain not found;
@@
 
-SEE TABLE (PROBABLY ON NEXT PAGE): THERE IS NEARLY A 1--1 MAPPING BETWEEN
-action AND postfix\_action.  NEW ACTIONS COULD BE ADDED TO FORCE A 1--1
-MAPPING, E.G. COMMIT + DISCARDED COULD BECOME A DISCARD ACTION\@.
-MAIL\_QUEUED HAVING INFO AND PROCESSING CAN PROBABLY BE DEALT WITH, THE
-BIGGEST DIFFICULTY WILL BE WITH SAVE\_DATA\@.
+XXX SEE TABLE BELOW\@: THERE IS NEARLY A 1--1 MAPPING BETWEEN action AND
+postfix\_action.  NEW ACTIONS COULD BE ADDED TO FORCE A 1--1 MAPPING, E.G.
+COMMIT + DISCARDED COULD BECOME A DISCARD ACTION\@.  MAIL\_QUEUED HAVING
+INFO AND PROCESSING CAN PROBABLY BE DEALT WITH, THE BIGGEST DIFFICULTY WILL
+BE WITH SAVE\_DATA\@.
 

### 2009-02-24 doc/logparser-other-parsers-review.tex
Added delay, delays, and dsn to the results table; they're automatically     extracted and saved because filter_regex() fiddles the existing regex     snippets. Updated the thesis with new result columns, including the comparisons in the     state of the art review. Changed how the acronyms glossary is formatted by adding a new style that uses     eqlist. A couple of minor fixes.

@@
 \sectionref{parser efficiency}.  \Tableref{Summary of parsers' features}
-summarises the results.
+summarises the results of this review.
 
@@
         extensive and intrusive modifications to the parser; \parsernames{}
-        architecture is described in \sectionref{why separate rules,
-        actions, and framework?}.
+        architecture is described in \sectionref{parser architecture}.
 
@@
 address and hostname, HELO hostname, queueid, start time, end time,
-\acronym{SMTP} code, sender, recipient, size, message ID\@.
+\acronym{SMTP} code, \acronym{DSN} code, sender, recipient, size, message
+ID, delay and delays.
 
@@
 the server hostname, HELO hostname, queueid, start and end times,
-timestamps for each log line, or message id (for accepted mails only).
-Handling of multiple recipients, \acronym{SMTP} codes, or remote
-servers\footnote{A single mail may be sent to multiple remote servers if it
-was addressed to recipients in different domains, or Postfix needs to try
-multiple servers for one or more recipients.} is not explained;
-experimental observation shows that multiple records are added when a mail
-had multiple recipients (sadly the records are not associated or linked in
-any way), and presumably the same approach is taken when there were
-multiple destination servers.
+timestamps for each log line, \acronym{DSN} code, delivery delays, or
+message id (for accepted mails only).  Handling of multiple recipients,
+\acronym{SMTP} codes, or remote servers\footnote{A single mail may be sent
+to multiple remote servers if it was addressed to recipients in different
+domains, or Postfix needs to try multiple servers for one or more
+recipients.} is not explained; experimental observation shows that multiple
+records are added when a mail had multiple recipients (sadly the records
+are not associated or linked in any way), and presumably the same approach
+is taken when there were multiple destination servers.
 
@@
     \item Some records output by DAILY-REJECT are truncated: they are
-        missing the last | separating fields and the newline following it,
-        so the line containing only a single space is concatenated with the
-        record.
+        missing the last $|$ that separates the fields and the newline
+        following it, so the line containing only a single space is
+        concatenated with the record.
 
@@
 Even if the three data sets were combined Sawmill would extract less data
-than \parsername{}: it omits the HELO hostname, queueid, and start and end
-times.  Sawmill does not extract any data about rejections except when the
-rejection is caused by a \acronym{DNSBL} check (\texttt{RBL list} in the
-list of fields).
+than \parsername{}: it omits the HELO hostname, queueid, \acronym{DSN}
+code, delivery delays, and start and end times.  Sawmill does not extract
+any data about rejections except when the rejection is caused by a
+\acronym{DNSBL} check (\texttt{RBL list} in the list of fields).
 
@@
 server \acronym{IP} address and hostname, queueid, start and end times,
-timestamps for each log line, \acronym{SMTP} code, and message ID), though
-it does not make the full text of the line available (this could be
-trivially added if desired, but would greatly increase the size of the
-resulting database).  The full power of \acronym{SQL} is available when
-searching the data extracted by \parsername{}, allowing the user to search
-on arbitrarily complicated conditions.
+timestamps for each log line, \acronym{SMTP} and \acronym{DSN} codes,
+delivery delays, and message ID), though it does not make the full text of
+the line available (this could be trivially added if desired, but would
+greatly increase the size of the resulting database).  The full power of
+\acronym{SQL} is available when searching the data extracted by
+\parsername{}, allowing the user to search on arbitrarily complicated
+conditions.
 
@@
 the hostname, not both; it omits the HELO hostname, queueid, \acronym{SMTP}
-code, size of each accepted mail, start and end times, timestamps for each
-log line, and message ID\@.  It extracts some data that \parsername{} does
-not: the delay in delivering each mail, and how many times delivery was
-deferred for each mail; these could easily be extracted by \parsername{} if
-desired.  Unlike \parsername{}, YAALA does not maintain separate counters
-for each restriction; this rules out the possibility of using the collected
-data for optimisation, testing, or understanding of restrictions.  YAALA's
-Postfix parser does not deal with the complications explained in
-\sectionref{complications}, except it does correlate log lines by queueid.
+and \acronym{DSN} codes, size of each accepted mail, start and end times,
+timestamps for each log line, and message ID\@.  It extracts some data that
+\parsername{} does not: how many times delivery was deferred for each mail;
+this information can be determined from the database populated by
+\parsername{} if desired.  Unlike \parsername{}, YAALA does not maintain
+separate counters for each restriction; this rules out the possibility of
+using the collected data for optimisation, testing, or understanding of
+restrictions.  YAALA's Postfix parser does not deal with the complications
+explained in \sectionref{complications}, except it does correlate log lines
+by queueid.
 
@@
 \parsername{}: client and server \acronym{IP} address and hostname, start
-and end times, timestamps of each log line, and message ID\@.
+and end times, \acronym{DSN} code, delivery delays, timestamps of each log
+line, and message ID\@.
 
@@
 \acronym{IP} address and hostname, server \acronym{IP} address, HELO
-hostname, queueid, timestamps of each log line, and message ID\@.  It
-extracts the delay for delivered mails, which \parsername{} does not.  Log
-lines are correlated based on the queueid (called sessionname [sic] within
-Logrep), but not by \acronym{pid}.  The parsing is error prone: empty
-fields are saved when the log line does not match the regex, though it
-appears that they will not overwrite existing data.  Most notably
-rejections are completely ignored, making it unsuitable for the purposes of
-this project.  It does not try to address any of the complications in
-\sectionref{complications} except for correlating by queueid.
+hostname, queueid, timestamps of each log line, \acronym{DSN} code, and
+message ID\@.  Log lines are correlated based on the queueid (called
+sessionname [sic] within Logrep), but not by \acronym{pid}.  The parsing is
+error prone: empty fields are saved when the log line does not match the
+regex, though it appears that they will not overwrite existing data.  Most
+notably rejections are completely ignored, making it unsuitable for the
+purposes of this project.  It does not try to address any of the
+complications in \sectionref{complications} except for correlating by
+queueid.
 

### 2009-02-24 doc/logparser-implementation.tex
What I described as DSN is actually Enhanced States Codes.

@@
 
-    \item [dsn] The \acronym{DSN} code, similar to the \acronym{SMTP} code;
-        it will only be present for log lines indicating delivery or
-        rejection.  \acronym{DSN} support was added to Postfix in version
-        2.3; logs generated by previous versions will not contain any DSN
-        information.
+    \item [enhanced\_status\_code] The enhanced status code~\cite{RFC3463}
+        is similar to the \acronym{SMTP} code, but is intended to be parsed
+        by mail clients so that error messages can be clearly conveyed to
+        the user.  Enhanced status code support was added to Postfix in
+        version 2.3; logs generated by previous versions will not contain
+        any enhanced status codes.
 

### 2009-02-24 doc/logparser-other-parsers-review.tex
What I described as DSN is actually Enhanced States Codes.

@@
 address and hostname, HELO hostname, queueid, start time, end time,
-\acronym{SMTP} code, \acronym{DSN} code, sender, recipient, size, message
+\acronym{SMTP} code, enhanced status code, sender, recipient, size, message
 ID, delay and delays.
@@
 the server hostname, HELO hostname, queueid, start and end times,
-timestamps for each log line, \acronym{DSN} code, delivery delays, or
+timestamps for each log line, enhanced status code, delivery delays, or
 message id (for accepted mails only).  Handling of multiple recipients,
@@
 Even if the three data sets were combined Sawmill would extract less data
-than \parsername{}: it omits the HELO hostname, queueid, \acronym{DSN}
+than \parsername{}: it omits the HELO hostname, queueid, enhanced status
 code, delivery delays, and start and end times.  Sawmill does not extract
@@
 server \acronym{IP} address and hostname, queueid, start and end times,
-timestamps for each log line, \acronym{SMTP} and \acronym{DSN} codes,
+timestamps for each log line, \acronym{SMTP} and enhanced status codes,
 delivery delays, and message ID), though it does not make the full text of
@@
 the hostname, not both; it omits the HELO hostname, queueid, \acronym{SMTP}
-and \acronym{DSN} codes, size of each accepted mail, start and end times,
+and enhanced status codes, size of each accepted mail, start and end times,
 timestamps for each log line, and message ID\@.  It extracts some data that
@@
 \parsername{}: client and server \acronym{IP} address and hostname, start
-and end times, \acronym{DSN} code, delivery delays, timestamps of each log
-line, and message ID\@.
+and end times, enhanced status code, delivery delays, timestamps of each
+log line, and message ID\@.
 
@@
 \acronym{IP} address and hostname, server \acronym{IP} address, HELO
-hostname, queueid, timestamps of each log line, \acronym{DSN} code, and
+hostname, queueid, timestamps of each log line, enhanced status code, and
 message ID\@.  Log lines are correlated based on the queueid (called

### 2009-02-24 doc/logparser-postfix-daemons.tex
Proofread postfix daemons appendix and made corrections/improvements.

--- 2009-01-09 doc/logparser-postfix-daemons.tex
@@
     The bounce daemon is responsible for sending bounce notifications in
-    Postfix versions later than 2.2.  The definitive documentation is
-    \url{http://www.postfix.org/bounce.8.html}.
+    Postfix version 2.3 and later.
+    \daemonDocURL{http://www.postfix.org/bounce.8.html}{2009/02/23}.
 }}
@@
 \newglossaryentry{cleanup}{name={cleanup},type={postfix},description={
-    Cleanup processes all incoming mail after it has been accepted and
-    before it is delivered.  It removes duplicate recipient addresses,
-    inserts missing headers, and optionally rewrites addresses if
-    configured to do so.  The definitive documentation is
-    \url{http://www.postfix.org/cleanup.8.html}.
+    The cleanup daemon processes all incoming mail after it has been
+    accepted and before it is delivered.  It removes duplicate recipient
+    addresses, inserts missing headers, and rewrites addresses if
+    configured to do so.
+    \daemonDocURL{http://www.postfix.org/cleanup.8.html}{2009/02/23}.
 }}
@@
 \newglossaryentry{lmtp}{name={lmtp},type={postfix},description={
-    Delivery of mail over \gls{LMTP} is performed by the lmtp component.
-    The definitive documentation is
-    \url{http://www.postfix.org/lmtp.8.html}.
+    Delivers mail using the \acronym{LMTP} protocol.
+    \daemonDocURL{http://www.postfix.org/lmtp.8.html}{2009/02/23}.
 }}
@@
     Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ delivered on the server Postfix is running on), whether it be to
-    a user's mailbox or a program such as a mailing list manager or
-    procmail (\url{http://www.procmail.org/}).  It also handles aliases and
-    processing of a user's \texttt{.forward} file.  The definitive
-    documentation is \url{http://www.postfix.org/local.8.html}.
+    (i.e.\ mail delivered on the server Postfix is running on); this
+    includes alias expansion, processing of a user's \texttt{.forward}
+    file, and delivery of the mail, whether it be to a user's mailbox or a
+    program such as a mailing list manager.
+    \daemonDocURL{http://www.postfix.org/local.8.html}{2009/02/23}.
 }}
@@
 \newglossaryentry{pickup}{name={pickup},type={postfix},description={
-    Pickup is the service that deals with mail submitted locally via
-    postdrop and sendmail; it passes all submitted mail to cleanup.  The
-    definitive documentation is \url{http://www.postfix.org/pickup.8.html}.
+    Pickup is the daemon that deals with mail submitted locally via
+    \daemon{postdrop} and \daemon{sendmail,} passing the mail on to
+    \daemon{cleanup} for further processing.
+    \daemonDocURL{http://www.postfix.org/pickup.8.html}{2009/02/23}.
 }}
@@
 \newglossaryentry{postdrop}{name={postdrop},type={postfix},description={
-    Postdrop is used when submitting mail locally on the server: it creates
-    a new mail in the queue and copies its input into the mail.  Subsequent
-    delivery of the mail is the responsibility of other Postfix components.
-    The definitive documentation is
-    \url{http://www.postfix.org/postdrop.1.html}.
+    Postdrop is used when submitting mail locally on the server: it copies
+    its input into a newly created mail in the queue, for processing by
+    \daemon{pickup} and subsequent delivery by one or more of
+    \daemon{lmtp}, \daemon{local}, \daemon{smtp}, or \daemon{virtual}.
+    \daemonDocURL[\newline{}]{http://www.postfix.org/postdrop.1.html}{2009/02/23}.
 }}
@@
 \newglossaryentry{postsuper}{name={postsuper},type={postfix},description={
-    Maintenance tasks such as deleting mails from the queue, putting mail
-    on hold and later releasing it (no further delivery attempts will be
-    made until it is released), and consistency checking of the mail queue.
-    The definitive documentation is available at
-    \url{http://www.postfix.org/postsuper.1.html}.
+    Used by the administrator for maintenance tasks such as deleting mails
+    from the queue, putting mail on hold and later releasing it, and
+    consistency checking of the mail queue.
+    \daemonDocURL[\newline{}]{http://www.postfix.org/postsuper.1.html}{2009/02/23}.
 }}
@@
     Qmgr is the Postfix daemon that manages the mail queue, determining
-    which mails will be delivered next.  Qmgr orders the mails based on the
+    which mails will be delivered next.  Qmgr groups mail based on the
     recipient for local mails and the destination server for remote
-    addresses, ensuring that it balances the aims of achieving maximum
-    concurrency while avoiding overwhelming destinations or wasting time
-    and resources on non-responsive destinations.  The definitive
-    documentation is \url{http://www.postfix.org/qmgr.8.html}.
+    addresses, ensuring that it achieves maximum concurrency without
+    overwhelming destinations or wasting resources on non-responsive
+    destinations.
+    \daemonDocURL{http://www.postfix.org/qmgr.8.html}{2009/02/23}.
 }}
@@
 \newglossaryentry{sendmail}{name={sendmail},type={postfix},description={
-    Postfix provides a command that is compatible with the Sendmail
-    (\url{http://www.sendmail.org/}) mail submission program that all Unix
-    commands that send mail depend on; it executes \daemon{postdrop} to
-    place a new mail in the queue.  The definitive documentation is
-    \url{http://www.postfix.org/sendmail.1.html}.
+    A Postfix component that is compatible with the Sendmail
+    (\urlLastCheckedNoParens{http://www.sendmail.org/}{2009/02/23}) mail
+    submission program which all Unix commands that need to send mail use;
+    it executes \daemon{postdrop} to place a new mail in the queue.
+    \daemonDocURL[\newline{}]{http://www.postfix.org/sendmail.1.html}{2009/02/23}.
 }}
@@
 \newglossaryentry{smtp}{name={smtp},type={postfix},description={
-    Delivery of mail over \gls{SMTP} is performed by the smtp component.
-    The definitive documentation is
-    \url{http://www.postfix.org/smtp.8.html}.
+    Delivers mail using the \acronym{SMTP} protocol.
+    \daemonDocURL{http://www.postfix.org/smtp.8.html}{2009/02/23}.
 }}
@@
 \newglossaryentry{smtpd}{name={smtpd},type={postfix},description={
-    Smtpd is the Postfix program that accepts mail via \gls{SMTP}, and
-    implements all the anti-spam restrictions Postfix provides.  The
-    definitive documentation is \url{http://www.postfix.org/smtpd.8.html}.
+    The Postfix component that accepts mail via \acronym{SMTP}, and
+    implements most of the anti-spam restrictions Postfix provides.
+    \daemonDocURL{http://www.postfix.org/smtpd.8.html}{2009/02/23}.
 }}
@@
 \newglossaryentry{virtual}{name={virtual},type={postfix},description={
-    Virtual is the Postfix component responsible for delivery of mails to
-    virtual domains.  With \daemon{local} delivery the destination is
+    The Postfix component responsible for delivery of mails to virtual
+    domains.  When \daemon{local} delivers mail the destination is
     determined only by the portion of the email address on the left side of
-    the @, whereas with \daemon{virtual} delivery the destination is
-    determined by the entire email address.  E.g.\ if the server is
-    responsible for both the \textbf{example.org} and \textbf{example.net}
-    domains: \daemon{local} considers \textbf{john@example.org} and
-    \textbf{john@example.net} to have the same mailbox, whereas
-    \daemon{virtual} considers them to have different mailboxes.  Virtual
-    delivery is used when a server hosts multiple domains where a username
-    may be present in more than one domain but represent different users in
-    each.  The definitive documentation is
-    \url{http://www.postfix.org/virtual.8.html}.
+    the \texttt{@}, whereas with \daemon{virtual} delivery the destination
+    is determined by the entire email address.  E.g.\ if the server is
+    responsible for both the \texttt{example.org} and \texttt{example.net}
+    domains: \daemon{local} would deliver mail for
+    \texttt{john@example.org} and \texttt{john@example.net} to the same
+    mailbox, whereas \daemon{virtual} would deliver mail for those
+    addresses to different mailboxes.  Virtual delivery is used where the
+    local part of an address may be present in multiple domains, and each
+    must be delivered to different users.
+    \daemonDocURL{http://www.postfix.org/virtual.8.html}{2009/02/23}.
 }}

### 2009-02-25 doc/logparser-acronyms.tex
Added delay, delays, and dsn to the results table; they're automatically     extracted and saved because filter_regex() fiddles the existing regex     snippets. Updated the thesis with new result columns, including the comparisons in the     state of the art review. Changed how the acronyms glossary is formatted by adding a new style that uses     eqlist. A couple of minor fixes.
Proofread the glossary and made corrections. Removed some unnecessary glossary entries for acronyms.

@@
     first={Postfix Log Parser (PLP)}
-}
-\newglossaryentry{PLP glossary}{
-    name={Postfix Log Parser},
-    description={
-        Postfix Log Parser is the parser implementing the algorithm
-        described in this thesis.
-    }
 }
@@
         One of the fundamental concepts when writing programs is the reuse
-        of existing code, so that each program does not reinvent the wheel.
-        When a body of code is intended to be reused repeatedly, the user
-        of this code needs to be informed of the functionality provided by
-        the code.  An API defines the interface provided to the user, and
-        acts as a contract between the user and the provider: if the user
-        adheres to the API the provider guarantees it will work, but is
-        free to change the implementation provided the API is preserved.
+        of existing code, so that each new program does not reinvent
+        existing wheels.  An API defines the interface provided to the user
+        of the existing code, and acts as a contract between the user and
+        the provider: if the user adheres to the API the provider
+        guarantees it will work, but is free to change the implementation
+        as long as the API is preserved.
     }
@@
         problems with it.  A detailed description of SMTP is beyond the
-        scope of this document: the original protocol definition is
+        scope of this thesis: the original protocol definition is
         RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821};
@@
         various protocols and file formats, e.g.\ SMTP\@.  The name is
-        somewhat misleading: initially the authors were asking for peer
-        review, but these documents are now the de facto standards the
+        somewhat misleading nowadays: initially the authors were asking for
+        peer review, but these documents are now the de facto standards the
         Internet runs on.
@@
 }
-\newglossaryentry{LMA glossary}{
-    name={Log Mail Analyzer},
-    description={
-        One of the other Postfix log file parsers reviewed and the only
-        example of prior published paper~\cite{log-mail-analyser} found
-        that specifically deals with parsing Postfix log files.
-    }
-}
 
@@
     first={Comma-Separated Value (CSV)}
-}
-\newglossaryentry{CSV glossary}{
-    name={Comma-Separated Value},
-    description={
-        The most basic form of database available, each record is a single
-        line in the file and the fields are separated by a special
-        character, typically a comma or colon.
-    }
 }
@@
         it only covers mail that is explicitly commercial, thus excluding
-        viruses, Bayesian poisoning mails, backscatter, and those annoying
-        chain letters you get from friends.
+        viruses, Bayesian poisoning mails, backscatter\glsadd{Backscatter},
+        and those annoying chain letters you get from friends.
     }
@@
     description={
-        ESMTP is Extended SMTP, defined in RFC~1869~\cite{RFC1869}.
+        \acronym{ESMTP}~\cite{RFC1869}, provides a flexible mechanism for
+        \acronym{SMTP} to be extended with new functionality, allowing new
+        features to be tested without having to be included in the standard
+        protocol.  \acronym{ESMTP} is backwards compatible with
+        \acronym{SMTP}: \acronym{ESMTP} clients and servers can interact
+        with \acronym{SMTP} clients and servers without difficulty.
     }
@@
 }
-\newglossaryentry{IP glossary}{
-    name={Internet Protocol},
-    description={
-        The Internet Protocol is the protocol used to communicate between
-        computers on the Internet.  An IP address is a unique address
-        assigned to a computer on the Internet, allowing it to communicate
-        with other computers on the Internet.
-    }
-}
 
@@
     first={Simple Logfile Clustering Tool (SLCT)}
-}
-\newglossaryentry{SLCT glossary}{
-    name={Simple Logfile Clustering Tool},
-    description={
-        SLCT~\cite{slct-paper} is a tool implementing an algorithm designed
-        by Risto Vaarandi for identifying, grouping, and producing a regex
-        to match similar log lines.
-    }
 }
@@
 }
-\newglossaryentry{ATN glossary}{
-    name={Augmented Transition Networks},
-    description={
-        Originally described in~\cite{atns} and further in~\cite{nlpip},
-        Augmented Transition Networks are a tool used in Computational
-        Linguistics for creating grammars to parse or generate sentences,
-    }
-}
 
@@
     first={Command Line Interface (CLI)}
-}
-\newglossaryentry{CLI glossary}{
-    name={Command Line Interface},
-    description={
-        An computer interface based on typing commands, rather than using a
-        mouse.
-    }
 }
@@
         $R$ is a (possibly empty) string of terminal and/or non-terminal
-        symbols.  Every CFG can be recognised by a \acronym{PDA}.
+        symbols.  Every CFG can be recognised by a \acronym{PDA}, and
+        \acronyms{PDA} cannot recognise more complicated grammars, so
+        \acronyms{CFL}, \acronyms{CFG}, and \acronyms{PDA} are equivalent.
     }
@@
         A Push-Down Automation is a computational device similar to a
-        \acronym{FA}, but it can additionally use a stack to store data.
-        A PDA can manipulate the stack during state transitions (adding or
-        removing a single piece of data), and can use 
-        the piece of data on the top of the stack in addition to the input
-        when determining which state transition to take.  PDA can recognise
-        languages of the form $a^{n}b^{n}$, i.e.\ any number of one symbol
-        followed by the same number of another symbol.
+        \acronym{FA}, but it can additionally use a stack to store data.  A
+        \acronym{PDA} can manipulate the stack during state transitions
+        (adding or removing a single piece of data), and when determining
+        which state transition to take it can use the piece of data on the
+        top of the stack in addition to the input.  \acronyms{PDA} can
+        recognise languages of the form $a^{n}b^{n}$, i.e.\ any number of
+        one symbol followed by the same number of another symbol.
     }

### 2009-02-25 doc/logparser-glossary.tex
Annotated all URLs with last checked date.
Proofread the glossary and made corrections. Removed some unnecessary glossary entries for acronyms.

@@
 % The contents of the glossary.
+% XXX CHECK FOR ENTRIES THAT SHOULD BE INCLUDED.
 \newglossaryentry{SQLite}{name={SQLite},description={
@@
     application utilising it.  More details can be found
-    at~\url{http://www.sqlite.org/}.
+    at~\urlLastChecked{http://www.sqlite.org/}{2009/02/23}.
 }}
@@
 
-\newglossaryentry{Joe job}{name={Joe job},description={
+\newglossaryentry{Joe job}{name={Joe Job},description={
     A joe~job describes spam mail sent using a faked sender address with
     the intention of sullying the good name of the user of that address.
-    Joe~jobs are one cause of backscatter.
+    Joe~jobs are one cause of backscatter\glsadd{Backscatter}.
 }}
@@
     $<>$ is the sender address used for mail that should not be replied to,
-    e.g.\ bounce notifications.  In \acronym{SMTP} all addresses are
-    enclosed in $<>$, so \textit{username@domain\/} becomes
-    \textit{$<$username@domain$>$\/}; thus $<>$ is actually an empty
-    address, but is always written as $<>$ for clarity.  All mail servers
-    must accept mail sent from $<>$, or they are in violation of
+    e.g.\ bounce notifications.  In the \acronym{SMTP} dialog all addresses
+    are enclosed in $<>$, so \texttt{username@domain} is sent as
+    \texttt{$<$username@domain$>$}; thus $<>$ is actually an empty address,
+    but is always written as $<>$ for clarity.  All mail servers must
+    accept mail sent from $<>$, or they are in violation of
     \acronym{RFC}~2821~\cite{RFC2821}.
@@
 
-\newglossaryentry{awk}{name={awk},description={
+\newglossaryentry{awk}{name={AWK},description={
     AWK is a general purpose programming language that is designed for
-    processing text-based data, and is available as a standard utility on
-    all Unix systems.
+    processing text, and is available as a standard utility on all Unix
+    systems.
 }}
 
-\newglossaryentry{syslog}{name={syslog},description={
-    Syslog is the standard logging mechanism on Unix systems: the program
+\newglossaryentry{syslog}{name={Syslog},description={
+    Syslog is the standard logging mechanism on Unix systems: a program
     sends log messages to syslog, then syslog filters and stores the
-    messages according to the configuration the administrator has chosen.
+    messages as configured by the administrator.
 }}
 
-\newglossaryentry{mail bomb}{name={mail bomb},description={
-    A mail bomb occurs when an attacker inflicts a huge volume of mail on
-    the victim.  At best a mail bomb is irritating to the victim; at worst
-    the deluge of mail can be severe enough to: interrupt service for the
-    victim and/or other users; cause mail to be rejected because the victim
-    has reached a limit (e.g.\ too many mails, too much disk space
-    consumed); the victim may accidentally delete other mail while trying
-    to cope with the mail bomb.
+\newglossaryentry{mail bomb}{name={Mail Bomb},description={
+    A mail bomb occurs when an attacker sends a huge volume of mail to a
+    victim.  At best a mail bomb is irritating to the victim; at worst the
+    deluge of mail can be severe enough to interrupt service for the victim
+    and/or other users, by causing mail to be delayed or rejected because
+    the victim or the server has reached a limit, e.g.\ too many mails in
+    their mailbox, too much disk space consumed.
 }}
 
-\newglossaryentry{mail loop}{name={mail loop},description={
+\newglossaryentry{mail loop}{name={Mail Loop},description={
     Sometimes mail sent to one address must be forwarded to a different
@@
     and mail for \texttt{bob@example.com} is delivered to
-    \texttt{alice@example.net}.
+    \texttt{alice@example.net}.  XXX EXPLAIN WHY THIS IS NOT AUTOMATICLLY
+    DETECTED SOMETIMES\@.
 }}

### 2009-02-25 doc/logparser-acronyms.tex
Proofread bibliography and made corrections.

@@
     description={
-        A DNS Blacklist~\cite{Wikipedia-DNSBL} is a simple collaborative
-        anti-spam technique used to reject or penalise mail sent from mail
-        servers believed to be the source of large volumes of spam.  The
-        criteria used when deciding whether or not to include an IP address
-        on a DNSBL vary widely between DNSBLs, so before using one it is
-        very important to check their listing policies.  To use a DNSBL the
-        mail server makes a DNS request incorporating the IP address of the
-        client; if the requested hostname exists the client is on the
-        DNSBL, and the mail server can decide what course of action to
-        take.
+        A DNS Blacklist is a simple collaborative anti-spam technique used
+        to reject or penalise mail sent from mail servers believed to be
+        the source of large volumes of spam.  The criteria used when
+        deciding whether or not to include an IP address on a DNSBL vary
+        widely between DNSBLs, so before using one it is very important to
+        check their listing policies.  To use a DNSBL the mail server makes
+        a DNS request incorporating the IP address of the client; if the
+        requested hostname exists the client is on the DNSBL, and the mail
+        server can decide what course of action to take.
     }
@@
     description={
-        Natural Language Processing~\cite{Wikipedia-nlp} is an area of
-        research that attempts to increase our understanding of the
-        languages normally used by humans (e.g.\ English, Spanish,
-        Japanese), with the goal of writing programs that can understand
-        human languages.
+        Natural Language Processing is an area of research that attempts to
+        increase our understanding of the languages normally used by humans
+        (e.g.\ English, Spanish, Japanese), with the goal of writing
+        programs that can understand human languages.
     }

### 2009-02-25 doc/logparser-glossary.tex
Proofread bibliography and made corrections.

@@
 
-\newglossaryentry{Joe job}{name={Joe Job},description={
+\newglossaryentry{Joe-job}{name={Joe~Job},description={
     A joe~job describes spam mail sent using a faked sender address with

### 2009-02-25 doc/logparser-implementation.tex
Proofread bibliography and made corrections.

@@
         attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job).  Notice that \daemon{postsuper} is required, not
-        optional, for this combination.
+        or joe~job\glsadd{Joe-job}).  Notice that \daemon{postsuper} is
+        required, not optional, for this combination.
 
@@
 some of the mail (because it is the result of a mail loop\glsadd{mail
-loop}, mail bomb\glsadd{mail bomb}, or joe~job\glsadd{Joe job}) before
+loop}, mail bomb\glsadd{mail bomb}, or joe~job\glsadd{Joe-job}) before
 Postfix has had a chance to try to deliver it.  In this case the recipient

### 2009-02-25 doc/logparser-acronyms.tex
Corrections to the background chapter.

@@
 }
+\newcommand{\SMTPglossaryDescription}[0]{%
+    \acronym{SMTP} is the protocol used for transferring mail between the
+    sending and receiving \acronym{MTA}\@.  It is a simple, human readable,
+    plain text protocol, making it quite simple to test and debug problems
+    with it.  A detailed description of SMTP is beyond the scope of this
+    thesis: the original protocol definition is in RFC~821~\cite{RFC821},
+    updated in RFC~2821~\cite{RFC2821}; introductory guides can be found
+    at~\cite{smtp-intro-01,smtp-intro-02}.%
+}
 \newglossaryentry{SMTP glossary}{
@@
     description={
-        SMTP is the protocol that transfers mail from the sender to the
-        recipient across the Internet.  It is a simple, human readable,
-        plain text protocol, making it quite simple to test and debug
-        problems with it.  A detailed description of SMTP is beyond the
-        scope of this thesis: the original protocol definition is
-        RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821};
-        introductory guides can be found
-        at~\cite{smtp-intro-01,smtp-intro-02}.
+        \SMTPglossaryDescription{}
     }
@@
 
+\newglossaryentry{SPF}{
+    type=\acronymtype,
+    name={SPF},
+    description={Sender Policy Framework},
+    first={Sender Policy Framework (SPF)}
+}
+
 %\newglossaryentry{}{

### 2009-02-25 doc/logparser-background.tex
Proofread bibliography and made corrections.
Corrections to the background chapter.

@@
 remainder of this thesis.  It begins with a discussion of the motivation
-underlying the project, followed by a brief introduction to
-\acronym{SMTP}\@.  It finishes with a longer introduction to Postfix,
-concentrating on anti-spam restrictions and policy servers.
+underlying the project, followed by a brief introduction to \acronym{SMTP},
+and finishes with a longer introduction to Postfix, concentrating on
+anti-spam restrictions and policy servers.
 
@@
 
-This thesis and the program it describes are part of a larger project to
-optimise a mail server's Postfix-based anti-spam restrictions, generate
-statistics and graphs, and provide a platform on which new restrictions can
-be evaluated to determine if they are beneficial in the fight against spam.
-The program parses Postfix log files and populates a database with the data
-gleaned from those log files, providing a consistent and simple view of the
-log files that future tools can utilise.  The gathered data can be used to
-optimise current anti-spam measures, to provide a baseline to test new
-anti-spam measures against, or to produce statistics showing how effective
-those measures are.
+This work is part of a larger project to optimise a mail server's
+Postfix-based anti-spam restrictions, generate statistics and graphs, and
+provide a platform on which new restrictions can be evaluated to determine
+if they are beneficial in the fight against spam.  The program written for
+this project, \parsername{}, parses Postfix log files and populates a
+database with the data gleaned from those log files, providing a consistent
+and simple view of the log files that future tools can utilise.  The
+gathered data can be used to optimise current anti-spam measures, to
+provide a baseline to test new anti-spam measures against, or to produce
+statistics showing how effective those measures are.
 
-Determining which Postfix restrictions reject the highest number of mails
-is a short example of the optimisation possible using data from the
-database:
+A short example of the optimisation possible using data from the database
+is determining which Postfix restrictions reject the highest number of
+mails:
 
@@
     FROM rules
-    WHERE postfix_action = 'REJECTED'
+    WHERE action = 'DELIVERY_REJECTED'
     ORDER BY hits_total DESC;
@@
 If the database supports sub-selects (where the results of one query are
-used in another), percentages can be obtained for the top ten restrictions
-(sample output shown in \tableref{Sample output from SQL query showing
-percentages}):
+used as a parameter in another), percentages can be obtained for the top
+ten restrictions (sample output shown in \tableref{Sample output from SQL
+query}):
+
+XXX db.full-run.sq3 NEEDS TO BE UPDATED AND EVERYTHING REBUILT\@.
 
@@
 \begin{table}[ht]
-    \caption{Sample output from SQL query showing percentages}
-    \empty{}\label{Sample output from SQL query showing
-    percentages}
+    \caption{Sample output from SQL query}
+    \empty{}\label{Sample output from SQL query}
     \begin{tabular}[]{lrr}
         \tabletopline{}%
-        Restriction & Number of hits & Percentage of hits \\
+        Restriction & Number of  & Percentage of    \\
+                    & rejections & total rejections \\
         \tablemiddleline{}%
@@
                 FROM rules
-                WHERE postfix_action = 'REJECTED'
+                WHERE action = 'DELIVERY_REJECTED'
             )
@@
     FROM rules
-    WHERE postfix_action = 'REJECTED'
+    WHERE action = 'DELIVERY_REJECTED'
         AND hits < 100
@@
 
-\section{Simple Mail Transfer Protocol}
+\section{Simple Mail Transfer Protocol (SMTP)}
 
@@
 
-The \glsfirst{SMTP}, originally defined in \acronym{RFC}~821~\cite{RFC821}
-and updated in \acronym{RFC}~2821~\cite{RFC2821}, is used for transferring
-mail between the sending and receiving \acronym{MTA}\@.  It is a simple,
-human readable, plain text protocol, simplifying testing and debugging
-problems with it.  Despite the simplicity of the protocol many virus and
-spam sending programs fail to implement it properly, so requiring strict
-adherence to the protocol specification is beneficial in protecting against
-spam and viruses.\footnote{\label{footnote:rfc760}Originally all mail
-servers adhered to the principle of \textit{Be liberal in what you accept,
-and conservative in what you send\/} from \acronym{RFC}~760~\cite{rfc760},
-but unfortunately that principle was written in a friendlier time.  Given
-the deluge of spam that mail servers are subjected to daily, a more
-appropriate maxim could be: \textit{Require strict adherence to
-\acronym{RFC}~2821; implement the strongest restrictions you can; relax the
-restrictions and adherence only when legitimate mail is impeded.\/}  It is
-not as friendly, nor as catchy, but it more accurately reflects the current
-situation.} A typical \acronym{SMTP} conversation resembles the following
-(the lines starting with a three digit number are sent by the server, all
-other lines are sent by the client):
+\SMTPglossaryDescription{} Despite the simplicity of the protocol many
+virus and spam sending programs fail to implement it properly, so requiring
+strict adherence to the protocol specification is beneficial in protecting
+against spam and viruses.\footnote{\label{footnote:rfc760}Originally all
+mail servers adhered to the principle of \textit{Be liberal in what you
+accept, and conservative in what you send\/} from
+\acronym{RFC}~760~\cite{rfc760}, but unfortunately that principle was
+written in a friendlier time.  Given the deluge of spam that mail servers
+are subjected to daily, a more appropriate maxim could be: \textit{Require
+strict adherence to relevant \acronyms{RFC}; implement the strongest
+restrictions you can; relax the restrictions and adherence only when
+legitimate mail is impeded.\/}  It is neither as friendly nor as catchy,
+but it more accurately reflects the current situation.} A typical
+\acronym{SMTP} conversation resembles the following (the lines starting
+with a three digit number are sent by the server, all other lines are sent
+by the client):
 
@@
 returned by the server and carries on regardless.  Many spam and virus
-sending programs have serious deficiencies; unfortunately others
-(particularly newer programs) were written by competent programmers, or
-send mail using well written programs (e.g.\ Postfix or Sendmail on Unix
-hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
-would have done its best to deal with deficient clients, with the intention
-of accepting as much mail destined for its users as possible, e.g.\ by
-ignoring the absence of a HELO command, or by accepting sender or recipient
-addresses that were not enclosed in \texttt{<>}.
-
-A detailed description of \acronym{SMTP} is beyond the scope of this
-thesis: introductory guides can be found
-at~\cite{smtp-intro-01,smtp-intro-02}, and the definitive references
-are~\cite{RFC821,RFC2821}.
+sending programs have serious deficiencies; unfortunately other programs,
+particularly newer programs, were written by competent programmers or send
+mail using well written programs (e.g.\ Postfix or Sendmail on Unix hosts,
+Microsoft Outlook on Windows hosts).  Traditionally a mail server would
+have done its best to deal with deficient clients, with the intention of
+accepting as much mail sent to its users as possible, e.g.\ by ignoring the
+absence of a HELO command, or by accepting sender or recipient addresses
+that were not enclosed in \texttt{<>}.
 
@@
 and bypassed on a per-client, per-recipient, or per-sender basis, allowing
-different levels of stricture and permissiveness.  Postfix leverages simple
-lookup tables to support arbitrarily complicated user-defined sequences of
-restrictions and exceptions.  Administrators can supply their own rejection
-messages to make it clear to senders exactly why their mail was rejected.
-Policy servers (\sectionref{policy servers}) provide a simple way to write
-new restrictions without having to modify Postfix's source code.
-Unfortunately this flexibility has a cost: complexity in the log files
-generated.  Although it is relatively simple in most cases to use standard
-Unix text processing utilities to determine the fate of an individual mail,
-following the journey a mail takes through Postfix can in some cases be
-quite difficult.  For most mails the journey is simple and brief, but the
-remainder can have quite complex journeys (see \sectionref{complications}
-for details).
+different levels of stricture and permissiveness.  Administrators can
+supply their own rejection messages to make it clear to senders exactly why
+their mail was rejected.  Policy servers (\sectionref{policy servers})
+provide a simple way to write new restrictions without having to modify
+Postfix's source code.  Unfortunately this flexibility has a cost:
+complexity in the log files generated.  Although it is relatively simple in
+most cases to use standard Unix text processing utilities to determine the
+fate of an individual mail, in some cases it can be quite difficult.  For
+most mails the journey is simple and brief, but the remainder can have
+quite complex journeys (see \sectionref{complications} for details).
 
@@
 that do one thing and do it well''\/}~\cite{unix-philosophy}, and it is
-separated into multiple programs that each perform one of the tasks
+separated into multiple components that each perform one of the tasks
 required of an \acronym{MTA}\@: receive mail, send mail, local delivery of
-mail, etc.\ (full details can be found in~\cite{postfix-overview}).  Each
-log line contains the name of the Postfix component that produced it, and
-this information is used when determining which rules should be used to
-parse each log line (see \sectionref{rules in architecture} for details).
+mail, etc.;\ full details can be found in~\cite{postfix-overview}.
 Postfix's design is strongly influenced by security concerns: those
@@
 The administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex sequences of lookups,
-restrictions, and exceptions.  Postfix tries to protect the administrator
-from misconfiguration in as far as is reasonable, e.g.\ the restriction
-\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
-because the parameter it checks (the hostname given in the HELO command) is
-under the control of the remote client.  Despite this, it is possible for
-the administrator to make catastrophic mistakes, e.g.\ rejecting all mail
---- the administrator must be cognisant of the effects their configuration
-changes will have.  This is similar to one of UNIX's design philosophies:
-\textit{``UNIX was not designed to stop its users from doing stupid things,
-as that would also stop them from doing clever
+restrictions, allowing arbitrarily long and complex user-defined sequences
+of lookups, restrictions, and exceptions.  Postfix tries to protect the
+administrator from misconfiguration in as far as is reasonable, e.g.\ the
+restriction \texttt{check\_helo\_mx\_access} cannot cause a mail to be
+accepted, because the parameter it checks (the hostname given in the HELO
+command) is under the control of the remote client.  Despite this, it is
+possible for the administrator to make catastrophic mistakes, e.g.\
+rejecting all mail --- the administrator must be cognisant of the effects
+their configuration changes will have.  This is similar to one of UNIX's
+design philosophies: \textit{``UNIX was not designed to stop its users from
+doing stupid things, as that would also stop them from doing clever
 things''\/}~\cite{unix-philosophy}.
-
-For further information about Postfix restrictions
-see~\cite{smtpd_access_readme,smtpd_per_user_control,policy-servers}.
 
@@
 mail on the third Tuesday after pay day only, to help prevent problems from
-phishing mails using faked sender addresses.\footnote{A phishing mail might
+phishing mails using faked sender addresses. E.g.\ a phishing mail might
 claim that the payroll system had a disastrous disk failure, and until the
@@
 so please reply to this mail with your name, address, and bank account
-details.}
+details; the criminal can then use any details sent to him to help with
+identity theft.
 
@@
 
-    \caption{Example attributes sent to policy servers}
+    \caption{Example attributes sent to policy servers, taken
+    from~\cite{policy-servers}}
     \empty{}\label{Example attributes sent to policy servers}
@@
 
-    \item Sender Policy Framework (SPF)~\cite{openspf}.  SPF\label{spf
+    \item \acronym{SPF}~\cite{openspf}.  \acronym{SPF}\label{spf
         introduction} records specify which mail servers are allowed to
-        send mail claiming to be from a particular domain.  The intention
-        is to reduce spam from faked sender addresses,
+        send mail using sender addresses from a particular domain.  The
+        intention is to reduce spam from faked sender addresses,
         backscatter~\cite{postfix-backscatter}, and
-        joe~jobs~\cite{Wikipedia-joe-job}\glsadd{Joe job}.  There has been
-        considerable resistance to the proposal because it breaks or vastly
-        complicates some features of \acronym{SMTP}, e.g.\ forwarding mail
+        Joe~jobs~\glsadd{Joe-job}.  There has been considerable resistance
+        to \acronym{SPF} because it breaks or vastly complicates some
+        commonly-used features of \acronym{SMTP}, e.g.\ forwarding mail
         from one domain to another when a user moves.
@@
         provided \acronym{SMTP} server from a compromised machine on the
-        ISP's network.  Greylisting will slowly become less useful, but it
-        does block a large percentage of spam mail at the moment; the most
-        effective restrictions from the \numberOFlogFILES{} log files used
-        when generating the results in \sectionref{Results} are shown in
-        \tableref{Summary of rejections}.  The table shows that greylisting
-        is worth using at the moment, particularly when you take into
-        account its position as the final restriction that a mail must
-        overcome in the configuration used on the mail server that
-        generated the log files: on that server greylisting only takes
-        effect for mails that have passed all other restrictions.  Some
-        problems may be encountered when using greylisting: some servers
-        fail to retry after a temporary failure, or legitimate mail may be
-        delayed, particularly when coming from a pool of servers.
+        ISP's network.  Greylisting will slowly become less effective as
+        spam senders adapt, but it does block a large percentage of spam
+        mail at the moment; the most effective restrictions from the
+        \numberOFlogFILES{} log files used when generating the results in
+        \sectionref{Results} are shown in \tableref{Summary of rejections}.
+        The table shows that greylisting is worth using at the moment,
+        particularly when you take into account its position as the final
+        restriction that a mail must overcome in the configuration used on
+        the mail server that generated the log files: on that server
+        greylisting only takes effect for mails that have passed all other
+        restrictions.  Some problems may be encountered when using
+        greylisting: some servers fail to retry after a temporary failure,
+        or legitimate mail may be delayed, particularly when coming from a
+        pool of servers.
 
@@
         whereas using a threshold that requires a delivery attempt to hit
-        several scored restrictions frees the administrator from
-        whitelisting clients that fall foul of one binary restriction only.
+        several scored restrictions will allow mail from clients that would
+        fall foul of one binary restriction only.
+
+        XXX SHOULD I URLS OR BIBLIOGRAPHY ENTRIES\@?  SHOULD I SCRAP ALL
+        THE URLS AND MOVE EVERYTHING INTO THE BIBLIOGRAPHY\@?  ASK CARL\@.
 

### 2009-02-26 doc/logparser-architecture.tex
Proofread the glossary and made corrections. Removed some unnecessary glossary entries for acronyms.
Checks for lines containing only spaces and "log line" in the architecture chapter.
Corrections from proofreading for first couple of sections.

@@
 details involved in parsing Postfix log files and the description of the
-parser implemented for this project, the two have been separated.  This
-chapter presents the architecture designed and developed for this project,
-beginning with the overall architecture and design, followed by the three
-components of the architecture: Framework, Actions, and Rules, and
-finishes by describing the characteristics of this architecture.
+parser implemented for this project, the two topics have been separated.
+This chapter presents the architecture designed and developed for this
+project, beginning with the overall architecture and design, followed by
+the three components of the architecture: Framework, Actions, and Rules,
+and finishes by describing the characteristics of this architecture.
 
@@
     \item [Framework]  The framework is the structure that actions and
-        rules plug into.  It provides the parsing loop, shared data
-        storage, loading and validation of rules, storage of results, and
-        other support functions.
+        rules plug into.  It manages the parsing process, providing shared
+        data storage, loading and validation of rules, storage of results,
+        and other support functions.
 
     \item [Actions] Each action performs the work required to deal with a
-        single category of inputs, e.g.\ processing data from rejections.
-        Actions are invoked to deal with a log line once it has been
-        identified by the rules: actions modify data structures, handle
-        complications, and cause data to be saved to the database.
+        single category of inputs, e.g.\ rejecting a delivery attempt.
+        Actions are invoked to process an input once it has been classified
+        by a rule.
 
@@
         and extracts data.  Rules provide an easily extensible method of
-        associating log lines with actions.
+        associating inputs with actions.
 
@@
 separation of functionality: the framework provides services to the
-actions, but does not need to perform any tasks specific to the input being
+actions, but does not need to perform anything specific to the input being
 parsed; actions benefit from having services provided by the framework,
 freeing them to concentrate on the task of accurately and correctly
-processing the information provided by rules; rules handle the low level
-details of classifying inputs and extracting data from those inputs.
+processing the inputs and the information provided by rules; rules handle
+the low level details of classifying inputs and extracting data from those
+inputs, for processing by actions.
 
 Separating the rules from the actions and framework makes it possible to
-parse new log lines without modifying the core parsing algorithm.  Adding a
-new rule with the action to invoke and a regex to match the log lines is
-trivial in comparison to understanding a program's entire parser,
-identifying the correct location to change, and making the appropriate
-changes.  Bear in mind that changes to the parser must be made without
-adversely affecting existing parsing, particularly as there may be edge
-cases that are not immediately obvious ---
-\sectionref{yet-more-aborted-delivery-attempts} describes a complication
-that occurs only four times in \numberOFlogFILES{} log files.  Requiring
-changes to the parser's source code also complicates upgrades, as the
-changes must be preserved during the upgrade, and may clash with changes
-made by the developer.  This architecture allows the user to add new rules
-without changing the parser, unless the new log lines require functionality
-not already provided by the existing actions.  If the new log lines do
+parse new inputs without modifying the core parsing algorithm.  Adding a
+new rule with the action to invoke and a regex to match the inputs is
+trivial in comparison to understanding an entire parser, identifying the
+correct location to change, and making the appropriate changes.  Bear in
+mind that changes to a parser must be made without adversely affecting
+existing parsing, particularly as there may be edge cases that are not
+immediately obvious --- \sectionref{yet-more-aborted-delivery-attempts}
+describes a complication that occurs only four times in \numberOFlogFILES{}
+log files.  Requiring changes to a parser's source code also complicates
+upgrades, as the changes must be preserved during the upgrade, and may
+clash with changes made by the developer.  This architecture allows the
+user to add new rules without editing a parser, unless the new inputs
+require cannot be processed by the existing actions.  If the new inputs do
 require new functionality, new actions can be added to the parser without
-modifying existing actions (\sectionref{adding new actions in
-implementation} describes how to safely add new actions); only in the rare
-event that the new actions require support from other sections of the code
-will more extensive changes be required.
+having to modify existing actions; only in the rare event that the new
+actions require support from other actions will more extensive changes be
+required.
 
-There is some similarity between the parser's design and William Wood's
-\acronym{ATN}~\cite{atns}, used in Computational Linguistics for creating
-grammars to parse or generate sentences.  The resemblance between the two
-(shown in \tableref{Similarities between ATN and this architecture}) is
-accidental, but it is obvious that the two different approaches share a
-similar division of responsibilities, despite having different semantics.
+There is some similarity between this architecture and William Wood's
+\acronym{ATN}~\cite{atns,nlpip}, used in Computational Linguistics for
+creating grammars to parse or generate sentences.  The resemblance between
+the two (shown in \tableref{Similarities between ATN and this
+architecture}) is accidental, but it is clear that the two different
+approaches share a similar division of responsibilities, despite having
+different semantics.
 
@@
         \tablemiddleline{}%
-        Networks        & Overall       & Determines the sequence 
+        Networks        & Parser        & Determines the sequence 
                                           of transitions              \\
-                        & Algorithm     & or actions that 
+                        &               & or actions that 
                                           constitutes a valid input.  \\
-        Transitions     & Actions       & Assembles data and
-                                          imposes conditions          \\
+        Transitions     & Actions       & Assemble data and
+                                          impose conditions           \\
                         &               & the input must meet to be
@@
 
-The framework takes care of miscellaneous support functions and low level
-details of parsing, freeing the programmers writing actions to concentrate
+The framework takes care of miscellaneous support functions and managing
+the parsing process, freeing the programmers writing actions to concentrate
 on writing productive code.  It links actions and rules, allowing either to
 be improved independently of the other.  It provides shared storage to pass
-data between actions, loads and saves state, loads and validates rules,
+data between actions, saves and loads state, loads and validates rules,
 manages parsing, invokes actions, tracks how often each rule matches to
 optimise rule ordering (\sectionref{rule ordering for efficiency}), stores
-results of parsing, and miscellaneous other tasks.
-Most parsers will require the same basic functionality from the framework,
-plus some specialised support functions.  The framework is the core of the
-architecture and is deliberately quite simple: the rules deal with the
+results of parsing, and miscellaneous other tasks.  Most parsers will
+require the same basic functionality from the framework, plus some
+specialised support functions XXX EXPAND\@.  The framework is the core of
+the architecture and is deliberately quite simple: the rules deal with the
 variation in inputs, and the actions deal with the intricacies and
@@
 The function that finds the rule matching the input and invokes the
-requested action can be expressed in pseudo-code (indentation denoting flow
-of control) as:
-
-% DO NOT REFORMAT!
+requested action can be expressed in pseudo-code (with indentation denoting
+flow of control) as:
 

### 2009-02-26 doc/logparser-introduction.tex
Cite my paper - it needs to be added to the bibliography. Change section to chapter.

@@
 
-\vspace{1em}\noindent\textbf{Layout:}
+\section{Thesis Layout}
 
@@
 
-Section~\ref{background} provides background information useful in
+Chapter~\ref{background} provides background information useful in
 understanding the thesis, parser, and architecture.
 
-Section~\ref{state of the art review} reviews both the previously published
+Chapter~\ref{state of the art review} reviews both the previously published
 research in this area and other available Postfix log file parsers,
@@
 
-Section~\ref{rules in architecture} discusses the parsing rules in detail,
+Chapter~\ref{rules in architecture} discusses the parsing rules in detail,
 explaining the purpose and usage of each field in a rule, referring to an
@@
 
-Section~\ref{Postfix Parser Implementation} contains the core of the paper,
+Chapter~\ref{Postfix Parser Implementation} contains the core of the paper,
 describing a naive parsing algorithm and the complications initially
@@
 
-Section~\ref{parsing coverage} analyses the coverage the parser achieves
+Chapter~\ref{parsing coverage} analyses the coverage the parser achieves
 over a set of \numberOFlogFILES{} log files taken from a mail server
@@
 
-Section~\ref{limitations and improvements in implementation} lists the
+Chapter~\ref{limitations and improvements in implementation} lists the
 limitations of the algorithm, then suggests some ways of dealing with them,
@@
 
-Section~\ref{conclusion} contains the conclusion of the thesis, describing
+Chapter~\ref{conclusion} contains the conclusion of the thesis, describing
 the results of the research, design, and implementation of the parser.
@@
 daemons.
+
+\section{XXX NOTES}
+
+Portions of \sectionref{parser architecture}, \sectionref{Postfix Parser
+Implementation}, and \sectionref{Results} have previously been published at
+a conference~\cite{sgai-2008}, and later reprinted in a
+journal~\cite{elsevier-2009}.  Publication of the conference paper and
+journal article above was supported by Science Foundation Ireland
+RFP~05/RF/CMS002.

### 2009-02-26 doc/logparser-architecture.tex
Corrections, improvements, and new content in architecture chapter. Added SGAI and Elsevier entries to bibliography. Some minor cleanup elsewhere.

@@
 
-Each action is a separate procedure written to deal with a particular
-category of input, e.g.\ rejections.  The actions are parser-specific: each
-parser author will need to write the required actions from scratch unless
-extending an existing parser.  It is anticipated that parsers based on this
-architecture will have a high ratio of rules to actions, with the aim of
-having simpler rules and clearer distinctions between the inputs parsed by
-different rules.
+Each action is a separate procedure written to process a particular
+category of input, e.g.\ rejections.  There may be many input variants
+within one input category; in general each action will handle one category
+of inputs, with each rule rule recognising one input variant.  It is
+anticipated that parsers based on this architecture will have a high ratio
+of rules to actions, with the aim of having simple rules and clear
+distinctions between the inputs parsed by different rules.  An action may
+process different input variants in slightly different ways, but large
+changes in processing indicate the need for a new action and a new category
+of input; if an action becomes unnecessarily complicated is starts to turn
+into a monolithic parser, with too much logic contained in a single
+procedure.
 
-The ability to add special purpose actions to deal with difficulties and
-new requirements that are discovered during parser development is one of
-the strengths of this architecture.  Instead of writing a single monolithic
-function that must be modified to support new behaviour, with all the
-attendant risks of adversely affecting the existing parser, when a new
-requirement arises an independent action can be written to satisfy it.
-Sometimes the new action will require the cooperation of other actions,
-e.g.\ to set or check a flag.  There is a possibility of introducing
+The ability to easily add special purpose actions to deal with difficulties
+and new requirements that are discovered during parser development is one
+of the strengths of this architecture.  XXX REWRITE, EXPAND, AND CLARIFY UP
+TO THE NEXT XXX\@.  Instead of writing a single monolithic function that
+must be modified to support new behaviour, with all the attendant risks of
+adversely affecting the existing parser, when a new requirement arises an
+independent action can be written to satisfy it.  Sometimes the new action
+will require the cooperation of other actions, e.g.\ to set or check a
+flag, or save some extra data.  There is a possibility of introducing
 failure when modifying existing actions in this way, but the modifications
@@
 architecture, thus failures will be less likely to occur and will be easier
-to test for and diagnose.  The architecture can be implemented in an object
-oriented style, allowing sub-classes to extend or override actions in
-addition to adding new actions; because each action is an independent
-procedure, the sub-class need only modify the action it is overriding,
-rather than reproducing large chunks of functionality.
+to test for and diagnose.  XXX FINISH REWRITING\@.  The architecture can be
+implemented in an object oriented style, allowing sub-classes to extend or
+override actions in addition to adding new actions; because each action is
+an independent procedure, the sub-class need only modify the action it is
+overriding, rather than reproducing large chunks of functionality.
 
 During development of the Postfix log parser it became apparent that in
-addition to the obvious variety in log lines there were many complications
-to be overcome.  Some were the result of deficiencies in Postfix's logging
-(some of which were rectified by later versions of Postfix); others were
+addition to the obvious variety in log \empty{fool check}lines there were
+many complications to be overcome.  Some were the result of deficiencies in
+Postfix's logging (some of which were rectified by later versions of
+Postfix, e.g.\ identifying bounce
+notifications~\sectionref{identifying-bounce-notifications}); others were
 due to the vagaries of process scheduling, client behaviour, and
@@
 existing actions to cooperate and exchange extra data (via the framework),
-changing their behaviour as appropriate based on that extra data.
+changing their behaviour as appropriate based on that extra data.  Every
+architecture should aim to make the hard things possible as well as making
+the easy things easy (XXX MAYBE QUOTE LARRY\@?); the successful
+implementation of the Postfix log parser demonstrates that this
+architecture achieves that aim.
 
@@
 rules and actions to parse a single input, potentially simplifying both
-rules and actions.
+rules and actions.  XXX EXPAND WITH CHEMICAL FORMULAE EXAMPLE\@.
 
@@
 
-Rules categorise inputs, specifying both the regex to match against each
-input and the action to invoke when the match is successful.  Parsing new
-inputs is generally achieved by creating a new rule that pairs an existing
-action with a new regex.  Decoupling the rules from the actions and
-framework enables other rule management approaches to be used, e.g.\
-instead of manually adding new rules, machine learning techniques could be
-used to automatically generate new rules.  If this approach was taken the
-choice of machine learning technique would be constrained by the size of
-typical data sets (see \sectionref{Results} XXX CHECK THIS REFERENCE WHEN
-THE RESULTS SECTION IS FINISHED).  Techniques requiring the full data set
+XXX DO I NEED TO SUBDIVIDE THIS INTO SUBSECTIONS\@?
+
+Rules are responsible for categorising inputs: each rule should recognise
+one and only input variant; an input category with multiple input variants
+should have multiple rules, one for each variant.  Rules will typically use
+a regex when classifying inputs, but other approaches may prove useful for
+some applications, e.g.\ comparing fixed strings to the input, or checking
+the length of the input; for the remainder of this thesis it will be
+assumed that a regex is used.  A rule must specify both the regex to match
+against each input and the action to invoke when the match is successful,
+but implementations are free to add any other attributes they require;
+\sectionref{rule attributes} describes the attributes used in
+\parsername{}, and some potential attributes will be discussed later in
+this section.
+
+The framework issues a warning for each unparsed input, so it is evident
+when the ruleset needs to be augmented.  Parsing new inputs is achieved in
+one of three ways:
+
+\begin{enumerate}
+
+    \item Modify an existing rule's regex, because the new input is part of
+        an existing variant.
+
+    \item Write a new rule that pairs an existing action with a new regex;
+        this adds a new variant to an existing category.
+
+    \item Create a new category of inputs and a new action to process
+        inputs from the new category; write a new rule pairing the new
+        action with a new regex.
+
+\end{enumerate}
+
+Decoupling the rules from the actions and framework enables other rule
+management approaches to be used, e.g.\ instead of manually adding new
+rules, machine learning techniques could be used to automatically generate
+new rules.  If this approach was taken the choice of machine learning
+technique would be constrained by the size of typical data sets (see
+\sectionref{parser efficiency}).  Techniques requiring the full data set
 when training would be impractical; Instance Based
@@
 parser might also dynamically create new rules in response to certain
-inputs, e.g.\ diagnostic messages indicating the program which produced the
-input being parsed had read a new configuration file.  These avenues of
-research and development has not been pursued by the author, but could
-easily be undertaken independently.
+inputs, e.g.\ diagnostic messages indicating that the program which
+produced the input being parsed had read a new configuration file, and had
+changed its behaviour appropriately.  These avenues of research and
+development have not been pursued by the author, but but the architecture
+allows them to easily be undertaken independently.
 
-The architecture does not try to detect overlapping rules: that
-responsibility is left to the author of the rules.  Unintentionally
+XXX ATTACHING CONDITIONS TO RULES\@; GIVE PLP EXAMPLE AND REFER TO
+IMPLEMENTATION AND RESULTS SECTIONS\@?
+
+XXX STATEFUL PARSING ENABLED BY CONDITIONS, E.G\@.  PARSING C COMMENTS\@.
+
+XXX PRIORITY ATTRIBUTE AND RULE ORDERING\@.
+
+When adding new rules, the rule author must be aware that the new rule may
+overlap with an existing rule, i.e.\ some inputs could be parsed by more
+than one rule.  The architecture does not try to detect overlapping rules:
+that responsibility is left to the author of the rules.  Unintentionally
 overlapping rules lead to inconsistent parsing and data extraction because
@@
 
+XXX CHECK WITH CARL WHETHER I SHOULD SAY CFL OR CFG IN THE NEXT
+PARAGRAPH\@.
+
 Decoupling the rules from the actions allows external tools to be written
@@
 \overline{(\overline{FA1} \cup{} \overline{FA2})}$, which has considerable
-computation complexity.  Perl 5.10 regexes are more powerful than
+computation complexity XXX MENTION DEMORGAN''S LAW\@?  COMMENT ON THE
+COMPUTATION COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE
+SPINNING WHEEL BOOK\@.  Perl 5.10 regexes are more powerful than
 traditional regexes: it is possible to match correctly balanced brackets
@@
 \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
-\verb!z<123<pq<>rs>j<r>ml>s!.  Perl 5.10 regexes can maintain an arbitrary
-state stack and are thus equivalent in computational power to \acronym{PDA}
-or \acronym{CFL}, so detecting overlap may require calculating the
-intersection of two \acronym{PDA} or \acronym{CFL}s.  The intersection of
-two \acronym{CFL}s is not closed, i.e.\ the resulting language cannot
-always be parsed by a \acronym{CFL}, so intersection may be intractable in
-some cases e.g.:
+\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
+regex engine to maintain state on a stack, so Perl 5.10 regexes are thus
+equivalent in computational power to \acronym{PDA} or a \acronym{CFL};
+detecting overlap may require calculating the intersection of two
+\acronym{PDA} or \acronyms{CFL} instead of two \acronym{FA}.  The
+intersection of two \acronyms{CFL} is not closed, i.e.\ the resulting
+language cannot always be parsed by a \acronym{CFL}, so intersection may be
+intractable in some cases, e.g.:
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
@@
 intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
-complexity of calculating intersection.  This is certainly not a task to be
-performed every time the parser is used: detecting overlap amongst the
-Postfix log parser's \numberOFrules{} rules would require calculating
-\numberOFruleINTERSECTIONS{} intersections.
+complexity of calculating \acronym{FA} or \acronym{PDA} intersection.
+This is certainly not a task to be performed every time the parser runs:
+naive detection of overlap amongst the Postfix log parser's
+\numberOFrules{} rules would require calculating
+\numberOFruleINTERSECTIONS{} intersections.  When detecting overlap any
+conditions attached to rules must be taken into account, because two rules
+whose regexes overlap may have conditions attached which prevent the
+overlap in practice.  A less naive approach to overlap detection would
+first check for overlapping conditions amongst rules, and then check for
+overlapping regexes amongst the rules in each group.  Depending on the
+conditions employed a rule may fall into more than one group, e.g.\ given
+three conditions:
 
-The framework requires each rule to have \texttt{action} and \texttt{regex}
-attributes; each implementation is free to add any additional attributes it
-requires.
+\begin{enumerate}
+
+    \item $total < 10$
+
+    \item $total > 20$
+
+    \item $total < 30$
+
+\end{enumerate}
+
+\noindent{}The first and second conditions do not overlap, but the third
+condition does overlap with both the first and second conditions.  If the
+rules are divided into sets based on how their conditions overlap, the
+complexity of detecting overlap amongst $n$ rules is:
+
+$$O(n^{2}y)~+~\sum{O(|s|^{2}x)~\forall{}~s~\in{}~overlaps}$$
+
+where:
+
+\begin{tabular}[]{rcl}
+
+            y & = & cost of checking for overlap between two conditions \\
+            x & = & cost of checking for overlap between two regexes    \\
+     overlaps & = & set of sets of rules with overlapping conditions    \\
+
+\end{tabular}
 
@@
 excessive amounts of CPU time during matching.  Defining a regex to match
-all inputs is trivial: \verb!/^/! matches the start of every input.
-Usually excessive CPU time is consumed when a regex with a lot of
-alteration and variable quantifiers fails to match, but successful
-matching is generally quite fast (see~\cite{mastering-regular-expressions}
-for in-depth discussion).
-
-XXX RULES CAN HAVE CONDITIONS ATTACHED
-
-\label{comparison against context-free grammars}
-
-In context-free grammar terms the parser rules could be described as:
-
-$\text{\textless{}log-line\textgreater{}} \mapsto \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
+all inputs is trivial: \verb!/^/! matches the start of every input.  This
+regex would be highlighted by a tool that detects overlapping rules, and
+would easily be noticed by visual inspection, but more complex regexes
+would be harder to spot.  XXX MAYBE GIVE AN EXAMPLE REGEX\@?  Usually
+excessive CPU time is consumed when a regex with a lot of alteration and
+variable quantifiers fails to match, but successful matching is generally
+quite fast (see~\cite{mastering-regular-expressions} for in-depth
+discussion).  XXX THIS NEEDS TO BE EXPANDED, OR MOVED TO FIT IN
+ELSEWHERE\@.
 
@@
 
-XXX IMPROVE THIS SECTION\@; SHOULD IT BE A SUBSECTION\@?
+XXX IMPROVE THIS\@; SHOULD IT BE A SUBSECTION\@?  EXPAND ON ALL THESE, OR
+MOVE THEM TO FIT IN ELSEWHERE\@.  ADD MORE IF POSSIBLE\@.
 
@@
 
-    \item Rules are annotated with the name of a Postfix program, and will
-        only be used when parsing log lines produced by that
-        program.
-
     \item The first matching rule wins: no further rules are tried against
-        that log line, but there is a mechanism for prioritising the rules
-        so that more specific rules can be tried first.
+        that input, but there is a mechanism for prioritising the rules so
+        that more specific rules can be tried first.  XXX MOVE THIS TO THE
+        PRIORITY SECTION\@.
 
     \item Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.
+        isolation, without reference to any other rules.  This is similar
+        to a \acronym{CFG}; in \acronym{CFG} terms the parser rules could
+        be described as:
 
-    \item Rule processing time is a linear function of the number of rules.
-        XXX IMPROVE THIS\@.
+        $\text{\textless{}input\textgreater{}} \mapsto \text{rule-1} |
+        \text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
 
@@
 
+XXX EXPAND ON ALL THESE\@.
+
 \begin{description}
@@
         rule determines the action that will be invoked: there is no
-        backtracking to try alternate rules, no attempt is made to pick a
+        backtracking to try alternate rules; no attempt is made to pick a
         \textit{best\/} rule.
@@
         there is no facility for rules to consume more input or push unused
-        input back onto the input stream.  This was not a deliberate design
-        decision, rather a consequence of the line oriented nature of
-        Postfix log files; more flexible approaches could be pursued.
+        input back onto the input stream.  The current input can be
+        manipulated by actions, using the cascaded parsing feature.  This
+        was not a deliberate design decision, rather a consequence of the
+        line oriented nature of Postfix log files; more flexible approaches
+        could be pursued.
 
     \item [Context-free rules:]  Rules can not take into account past or
-        future inputs.  In context-free grammar terms the parser rules
-        could be described as:
-        \newline{}$\text{\textless{}input\textgreater{}} \mapsto
-        \text{rule-1} | \text{rule-2} | \text{rule-3} | \dots |
-        \text{rule-n}$.
+        future inputs when classifying the current input, though conditions
+        can affect which rules will be used to classify each input.
 
@@
     \item [Transduction:]  The architecture can be thought of as
-        implementing transduction: it takes data in one form (log files)
-        and transforms it to another form (a database); other formats may
-        be more suitable for other implementations.
+        implementing transduction: it takes data in one form
+        and transforms it to another form; \parsername{} transforms log
+        files to a database.
 
-    \item [Similarity to \acronym{NLP}:] \hfill{} \newline{}
-        Unlike traditional parsers such as those used when compiling a
-        programming language, this architecture does not require a fixed
-        grammar specification that inputs must adhere to.  The architecture
-        is capable of dealing with interleaved inputs, out of order inputs,
-        and ambiguous inputs where heuristics must be applied --- all have
-        arisen and been successfully accommodated in the Postfix log
-        parser.
+    \item [Similarity to Natural Language Processing:] Unlike traditional
+        parsers such as those used when compiling a programming language,
+        this architecture does not require a fixed grammar specification
+        that inputs must adhere to.  The architecture is capable of dealing
+        with interleaved inputs, out of order inputs, and ambiguous inputs
+        where heuristics must be applied --- all have arisen and been
+        successfully accommodated in the Postfix log file parser.
 

### 2009-02-27 doc/logparser-architecture.tex
Divided rules section in architecture into subsections.

@@
 \label{parser architecture}
-
-XXX REWRITE\@: MERGE CONTENT FROM ELSEWHERE IN THE THESIS AND PAPER\@.
 
@@
 
-XXX TO BE WRITTEN --- CHECK THE IMPLEMENTATION SECTION\@.
-
 Each action is a separate procedure written to process a particular
@@
 within one input category; in general each action will handle one category
-of inputs, with each rule rule recognising one input variant.  It is
-anticipated that parsers based on this architecture will have a high ratio
-of rules to actions, with the aim of having simple rules and clear
-distinctions between the inputs parsed by different rules.  An action may
-process different input variants in slightly different ways, but large
-changes in processing indicate the need for a new action and a new category
-of input; if an action becomes unnecessarily complicated is starts to turn
-into a monolithic parser, with too much logic contained in a single
-procedure.
+of inputs, with each rule recognising one input variant.  It is anticipated
+that parsers based on this architecture will have a high ratio of rules to
+actions, with the aim of having simple rules and clear distinctions between
+the inputs parsed by different rules.  An action may process different
+input variants in slightly different ways, but large changes in processing
+indicate the need for a new action and a new category of input; if an
+action becomes unnecessarily complicated is starts to turn into a
+monolithic parser, with too much logic contained in a single procedure.
 
@@
 architecture should aim to make the hard things possible as well as making
-the easy things easy (XXX MAYBE QUOTE LARRY\@?); the successful
-implementation of the Postfix log parser demonstrates that this
-architecture achieves that aim.
+the easy things easy; the successful implementation of the Postfix log
+parser demonstrates that this architecture achieves that aim.
 
@@
 
+Glucose: $C_{6}H_{12}O_{6}$
+Hydrocarbon: $CH_{3}{(CH_{2})}_{50}CH_{3}$
+
 \section{Rules}
@@
 \label{rules in architecture}
-
-XXX TO BE WRITTEN\@.  TAKE CONTENT FROM THE PAPER AND POSSIBLY FROM THE
-IMPLEMENTATION SECTION\@.
-
-XXX DO I NEED TO SUBDIVIDE THIS INTO SUBSECTIONS\@?
 
@@
 this section.
+
+\subsection{Adding New Rules}
 
@@
 changed its behaviour appropriately.  These avenues of research and
-development have not been pursued by the author, but but the architecture
+development have not been pursued by the author, but the architecture
 allows them to easily be undertaken independently.
+
+\subsection{Attaching Conditions To Rules}
 
@@
 
+\subsection{Prioritisation and Rule Ordering}
+
 XXX PRIORITY ATTRIBUTE AND RULE ORDERING\@.
+
+The first matching rule wins: no further rules are tried against that
+input, but there is a mechanism for prioritising the rules so that more
+specific rules can be tried first.
+
+\subsection{Overlapping Rules}
 
@@
 
+\subsection{Pathological Rules}
+
 It is possible to define pathological regexes which fall into two main
@@
 
-Rule have certain characteristics that may help in understanding the
-parser architecture:
-
-XXX IMPROVE THIS\@; SHOULD IT BE A SUBSECTION\@?  EXPAND ON ALL THESE, OR
-MOVE THEM TO FIT IN ELSEWHERE\@.  ADD MORE IF POSSIBLE\@.
-
-\begin{itemize}
-
-    \item The first matching rule wins: no further rules are tried against
-        that input, but there is a mechanism for prioritising the rules so
-        that more specific rules can be tried first.  XXX MOVE THIS TO THE
-        PRIORITY SECTION\@.
-
-    \item Rules are completely self-contained and can be understood in
+    Rules are completely self-contained and can be understood in
         isolation, without reference to any other rules.  This is similar
@@
         \text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-\end{itemize}
 

### 2009-02-27 doc/logparser.tex
Added delay, delays, and dsn to the results table; they're automatically     extracted and saved because filter_regex() fiddles the existing regex     snippets. Updated the thesis with new result columns, including the comparisons in the     state of the art review. Changed how the acronyms glossary is formatted by adding a new style that uses     eqlist. A couple of minor fixes.
Proofread postfix daemons appendix and made corrections/improvements.
Remove the space between glossary groups, so the three glossaries look better.
Proofread the glossary and made corrections. Removed some unnecessar

@@
 \makeglossaries{}
-\newcommand{\acronym}[1]{%
-    \gls{#1}\glsadd{#1 glossary}%
+\newcommand{\acronyms}[1]{%
+    \acronym[s]{#1}%
+}
+\newcommand{\acronym}[2][]{%
+    \gls{#2}#1%
+    % If there's a glossary entry for this acronym add it, otherwise
+    % do nothing.
+    \ifglsentryexists{#2 glossary}%
+        {\glsadd{#2 glossary}}%
+        {}%
+}
+% Define a new glossary style that uses eqlist, so that acronyms line up
+% nicely.
+\newglossarystyle{eqlist}{%
+    \glossarystyle{list}%
+    \renewenvironment{theglossary}{\begin{eqlist}}{\end{eqlist}}%
+    \renewcommand*{\glossaryentryfield}[5]{%
+        \item [\textbf{##2}] ##3%
+    }%
+    \renewcommand{\glsgroupskip}[0]{}%
+}
+% Define a new glossary style that removes the space between glossary
+% groups.
+\newglossarystyle{nospacelist}{%
+    \glossarystyle{list}%
+%    \renewenvironment{theglossary}{\begin{eqlist}}{\end{eqlist}}%
+%    \renewcommand*{\glossaryentryfield}[5]{%
+%        \item [\textbf{##2}] ##3%
+%    }%
+    \renewcommand{\glsgroupskip}[0]{}%
 }
@@
     final       = true,
-    pdfborder   = {0 0 0},
+    pdfborder   = {0, 0, 0},
 }
@@
 }
+\newcommand{\daemonDocURL}[3][]{%
+    \hfill{} \newline{} \url{#2}%
+    \hfill{} #1 Last checked #3%
+}
+\newcommand{\urlLastCheckedNoParens}[3][ ]{%
+    \url{#2}#1last checked #3%
+}
 
@@
 \newcommand{\numberOFruleINTERSECTIONS}[0]{%
-    14196%
+    17020%
 }
@@
 
-XXX EXTRACT DELAY AND DEFER WHEN AUTOMATIC EXTRACTION HAS BEEN FINISHED,
-THEN UPDATE FIELDS IN BOTH DATABASE AND REVIEW OF OTHER PARSERS\@.
-
 XXX POSSIBLY PUBLISH THE STATE OF THE ART REVIEW IN A JOURNAL\@?
@@
 \renewcommand{\glossarypostamble}{\label{Glossary}}
-\printglossary{}
+\printglossary[style=nospacelist]{}
 \renewcommand{\glossarypostamble}{\label{Acronyms}}
-\printglossary[type=\acronymtype]{}
+\printglossary[type=\acronymtype,style=eqlist]{}
 \renewcommand{\glossarypostamble}{\label{Postfix Daemons}}
-\printglossary[type=postfix]{}
+\printglossary[type=postfix,style=nospacelist]{}
 

### 2009-02-27 doc/logparser-architecture.tex
Corrections from proofreading of architecture chapter.

@@
 To avoid cluttering the explanation of the parser architecture with the
-details involved in parsing Postfix log files and the description of the
-parser implemented for this project, the two topics have been separated.
-This chapter presents the architecture designed and developed for this
-project, beginning with the overall architecture and design, followed by
-the three components of the architecture: Framework, Actions, and Rules,
-and finishes by describing the characteristics of this architecture.
-
-This chapter centers on the theoretical, implementation-independent aspects
-of the architecture; the practical difficulties of writing a parser for
-Postfix log files are covered in detail in \sectionref{Postfix Parser
-Implementation}.
+details involved in implementing a parser for Postfix log files, the two
+topics have been separated.  This chapter presents the architecture
+developed for this project, beginning with the overall architecture and
+design, followed by the three components of the architecture: Framework,
+Actions, and Rules.  This chapter centers on the theoretical,
+implementation-independent aspects of the architecture; the practical
+difficulties of writing a parser for Postfix log files are covered in
+detail in \sectionref{Postfix Parser Implementation}.
 
@@
 It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that log files produced by Postfix vary widely from host to
-host, depending on the set of restrictions chosen by the administrator.
-With this in mind, one of the parser's design aims was to make adding new
-rules as easy as possible, to enable administrators to properly parse their
-own log files.  To enable this the architecture is divided into three
-parts: framework, actions and rules.  Each will be discussed separately,
-but first an overview:
+background}) that log files produced by Postfix may vary widely from host
+to host, depending on the set of restrictions chosen by the administrator.
+With this in mind, one of the architecture's design aims was to make adding
+new rules to parse new inputs as easy as possible, to enable administrators
+to properly parse their own log files.  To enable this the architecture is
+divided into three parts: Framework, Actions, and Rules.  Each will be
+discussed separately, but first an overview:
 
@@
 
-    \item [Rules]  The rules are responsible for classifying inputs; they
-        specify the action to invoke and the regex that matches the inputs
-        and extracts data.  Rules provide an easily extensible method of
-        associating inputs with actions.
+    \item [Rules]  The rules are responsible for classifying inputs; each
+        rule recognises one input variant, and specifies the action that
+        will be invoked to process the input. Rules provide an easily
+        extensible method of associating inputs with actions.
 
@@
 For each input the framework tries each rule in turn until it finds a rule
-that matches the input, then invokes the action specified by that rule.
+that recognises the input, then invokes the action specified by that rule.
+If the input is not recognised by any of the rules, the framework issues a
+warning; the framework can continue parsing, although for some inputs
+stopping immediately may be preferable.
 
@@
 the framework, actions, and rules simplifies all three and creates a clear
-separation of functionality: the framework provides services to the
-actions, but does not need to perform anything specific to the input being
-parsed; actions benefit from having services provided by the framework,
-freeing them to concentrate on the task of accurately and correctly
-processing the inputs and the information provided by rules; rules handle
-the low level details of classifying inputs and extracting data from those
-inputs, for processing by actions.
+separation of functionality: the framework manages the parsing process and
+provides services to the actions, but does not need to perform anything
+specific to the input being parsed; actions benefit from having services
+provided by the framework, freeing them to concentrate on the task of
+accurately and correctly processing the inputs and the information provided
+by rules; rules are responsible for classifying inputs, and extracting data
+from those inputs for processing by actions.
 
@@
 parse new inputs without modifying the core parsing algorithm.  Adding a
-new rule with the action to invoke and a regex to match the inputs is
+new rule with the action to invoke and a regex to recognise the inputs is
 trivial in comparison to understanding an entire parser, identifying the
@@
 log files.  Requiring changes to a parser's source code also complicates
-upgrades, as the changes must be preserved during the upgrade, and may
-clash with changes made by the developer.  This architecture allows the
-user to add new rules without editing a parser, unless the new inputs
-require cannot be processed by the existing actions.  If the new inputs do
-require new functionality, new actions can be added to the parser without
-having to modify existing actions; only in the rare event that the new
-actions require support from other actions will more extensive changes be
-required.
+upgrades of the parser, because the changes must be preserved during the
+upgrade, and they may clash with changes made by the developer.  This
+architecture allows the user to add new rules without having to edit a
+parser, unless the new inputs cannot be processed by the existing actions.
+If the new inputs do require new functionality, new actions can be added to
+the parser without having to modify existing actions; only in the rare
+event that the new actions need to cooperate with existing actions will
+more extensive changes be required.
 
@@
         Transitions     & Actions       & Assemble data and
-                                          impose conditions           \\
-                        &               & the input must meet to be
-                                          accepted as                 \\
-                        &               & valid.                      \\
+                                          impose conditions the       \\
+                        &               & input must meet to be
+                                          accepted as valid.          \\
         Abbreviations   & Rules         & Responsible for
@@
 \end{table}
+
+The architecture can be thought of as implementing transduction: it takes
+data in one form and transforms it to another form; \parsername{}
+transforms log files to a database.  XXX NEED A REFERENCE FOR
+TRANSDUCTION\@.
+
+Unlike traditional parsers such as those used when compiling a programming
+language, this architecture does not require a fixed grammar specification
+that inputs must adhere to.  The architecture is capable of dealing with
+interleaved inputs, out of order inputs, and ambiguous inputs where
+heuristics must be applied --- all have arisen and been successfully
+accommodated in the Postfix log file parser.
 
@@
 
-The framework takes care of miscellaneous support functions and managing
-the parsing process, freeing the programmers writing actions to concentrate
-on writing productive code.  It links actions and rules, allowing either to
-be improved independently of the other.  It provides shared storage to pass
-data between actions, saves and loads state, loads and validates rules,
-manages parsing, invokes actions, tracks how often each rule matches to
-optimise rule ordering (\sectionref{rule ordering for efficiency}), stores
-results of parsing, and miscellaneous other tasks.  Most parsers will
-require the same basic functionality from the framework, plus some
-specialised support functions XXX EXPAND\@.  The framework is the core of
-the architecture and is deliberately quite simple: the rules deal with the
-variation in inputs, and the actions deal with the intricacies and
-complications encountered when parsing.
+The framework manages the parsing process and provides support functions,
+freeing the programmers writing actions to concentrate on writing
+productive code.  It links actions and rules, allowing either to be
+improved independently of the other, and allows new rules to be written
+without needing changes to the source code of a parser.  The framework is
+the core of the architecture and is deliberately quite simple: the rules
+deal with the variation in inputs, and the actions deal with the
+intricacies and complications encountered during parsing.
 
 The function that finds the rule matching the input and invokes the
-requested action can be expressed in pseudo-code (with indentation denoting
-flow of control) as:
+requested action can be expressed in pseudo-code as:
 
 \begin{verbatim}
-for each input:
-    for each rule defined by the user:
-        if this rule matches the input:
+INPUT:
+for each input {
+    for each rule defined by the user {
+        if this rule matches the input {
             perform the action specified by the rule
-            skip the remaining rules
-            process the next input
+            next INPUT
+        }
+    }
     warn the user that the input was not parsed
+}
 \end{verbatim}
+
+Most parsers will require the same basic functionality from the framework,
+plus some specialised support functions.  XXX EXPAND THIS\@.
+
+XXX TURN THIS INTO A LIST AND EXPAND THE ITEMS\@.  It provides shared
+storage to pass data between actions, saves and loads state, loads and
+validates rules, manages parsing, invokes actions, stores results of
+parsing, and miscellaneous other tasks.
 
@@
 Each action is a separate procedure written to process a particular
-category of input, e.g.\ rejections.  There may be many input variants
-within one input category; in general each action will handle one category
-of inputs, with each rule recognising one input variant.  It is anticipated
-that parsers based on this architecture will have a high ratio of rules to
-actions, with the aim of having simple rules and clear distinctions between
-the inputs parsed by different rules.  An action may process different
-input variants in slightly different ways, but large changes in processing
-indicate the need for a new action and a new category of input; if an
-action becomes unnecessarily complicated is starts to turn into a
-monolithic parser, with too much logic contained in a single procedure.
+category of input, e.g.\ rejecting a delivery attempt.  There may be many
+input variants within one input category; in general each action will
+handle one category of inputs, with each rule recognising one input
+variant.  XXX EXPAND AND CLARIFY\@; THE MESSAGE IS HERE, IT JUST NEEDS TO
+BE CLEARER\@.  It is anticipated that parsers based on this architecture
+will have a high ratio of rules to actions, with the aim of having simple
+rules and clear distinctions between the inputs parsed by different rules.
+An action may process different input variants in slightly different ways,
+but large changes in processing indicate the need for a new action and a
+new category of input; if an action becomes unnecessarily complicated it
+starts to turn into a monolithic parser, with too much logic contained in a
+single procedure.
 
@@
 an independent procedure, the sub-class need only modify the action it is
-overriding, rather than reproducing large chunks of functionality.
+overriding, and can inherit the others rather than reproducing large chunks
+of functionality.
 
 During development of the Postfix log parser it became apparent that in
-addition to the obvious variety in log \empty{fool check}lines there were
+addition to the obvious variety in log~lines there were
 many complications to be overcome.  Some were the result of deficiencies in
@@
 
+Actions are not context-sensitive, but they can consult the results (or
+lack of results) of previous actions during execution, providing some
+context sensitivity.  XXX EXPAND\@.
+
 Actions may return a modified input line that will be parsed as if read
@@
 Rules are responsible for categorising inputs: each rule should recognise
-one and only input variant; an input category with multiple input variants
+one and only input variant.  An input category with multiple input variants
 should have multiple rules, one for each variant.  Rules will typically use
@@
 the length of the input; for the remainder of this thesis it will be
-assumed that a regex is used.  A rule must specify both the regex to match
-against each input and the action to invoke when the match is successful,
-but implementations are free to add any other attributes they require;
-\sectionref{rule attributes} describes the attributes used in
+assumed that a regex is used.  Each rule must specify at least the regex to
+recognise inputs and the action to invoke when classification is
+successful, but implementations are free to add any other attributes they
+require; \sectionref{rule attributes} describes the attributes used in
 \parsername{}, and some potential attributes will be discussed later in
-this section.
+this section (\sectionref{attaching conditions to rules} and
+\sectionref{prioritisation and rule ordering}).  Classifying inputs using
+the rules is simple: the first rule to recognise the input determines the
+action that will be invoked: there is no backtracking to try alternate
+rules; no attempt is made to pick a \textit{best\/} rule.
 
@@
 Decoupling the rules from the actions and framework enables other rule
-management approaches to be used, e.g.\ instead of manually adding new
-rules, machine learning techniques could be used to automatically generate
-new rules.  If this approach was taken the choice of machine learning
-technique would be constrained by the size of typical data sets (see
-\sectionref{parser efficiency}).  Techniques requiring the full data set
-when training would be impractical; Instance Based
+management approaches to be used, e.g.\ instead of manually editing
+existing rules or adding new rules, machine learning techniques could be
+used to automate the process.  If this approach was taken the choice of
+machine learning technique would be constrained by the size of typical data
+sets (see \sectionref{parser efficiency}).  Techniques requiring the full
+data set when training would be impractical; Instance Based
 Learning~\cite{instance-based-learning} techniques that automatically
@@
 can be discarded might reduce the data required to a manageable size.  A
-parser might also dynamically create new rules in response to certain
-inputs, e.g.\ diagnostic messages indicating that the program which
-produced the input being parsed had read a new configuration file, and had
-changed its behaviour appropriately.  These avenues of research and
+parser could also dynamically create new rules in response to certain
+inputs, e.g.\ XXX GIVE A GOOD EXAMPLE\@.  These avenues of research and
 development have not been pursued by the author, but the architecture
@@
 \subsection{Attaching Conditions To Rules}
+
+\label{attaching conditions to rules}
 
@@
 \subsection{Prioritisation and Rule Ordering}
+
+\label{prioritisation and rule ordering}
 
@@
 
+\label{overlapping rules in architecture}
+
+XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?
+
 When adding new rules, the rule author must be aware that the new rule may
-overlap with an existing rule, i.e.\ some inputs could be parsed by more
-than one rule.  The architecture does not try to detect overlapping rules:
-that responsibility is left to the author of the rules.  Unintentionally
-overlapping rules lead to inconsistent parsing and data extraction because
-the first matching rule wins, and the order in which rules are tried
-against each input might change between parser invocations.  Overlapping
-rules are frequently a requirement, allowing a more specific rule to match
-some inputs and a more general rule to match the remainder, e.g.\
-separating \acronym{SMTP} delivery to specific sites from \acronym{SMTP}
-delivery to the rest of the world.  Allowing overlapping rules simplifies
-both the general rule and the more specific rule; additionally rules from
-different sources can be combined with a minimum of prior cooperation or
-modification required.  Overlapping rules should have a priority attribute
-to specify their relative ordering; negative priorities may be useful for
-catchall rules.
+overlap with one or more existing rules, i.e.\ some inputs could be parsed
+by more than one rule.  The architecture does not try to detect overlapping
+rules: that responsibility is left to the author of the rules.
+Unintentionally overlapping rules lead to inconsistent parsing and data
+extraction because the first matching rule wins, and the order in which
+rules are tried against each input might change between parser invocations.
+Overlapping rules are frequently a requirement, allowing a more specific
+rule to match some inputs and a more general rule to match the remainder,
+e.g.\ separating \acronym{SMTP} delivery to specific sites from
+\acronym{SMTP} delivery to the rest of the world.  Allowing overlapping
+rules simplifies both the general rule and the more specific rule;
+additionally rules from different sources can be combined with a minimum of
+prior cooperation or modification required.  Overlapping rules should have
+a priority attribute to specify their relative ordering; negative
+priorities may be useful for catchall rules.
 
 XXX CHECK WITH CARL WHETHER I SHOULD SAY CFL OR CFG IN THE NEXT
-PARAGRAPH\@.
+PARAGRAPH\@.  XXX DO I NEED TO MENTION CFG/CFL AT ALL\@?
 
@@
 \acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
-for example in~\cite{intersection-of-NFA-using-Z}) is: $FA1 \cap{} FA2 =
-\overline{(\overline{FA1} \cup{} \overline{FA2})}$, which has considerable
-computation complexity XXX MENTION DEMORGAN''S LAW\@?  COMMENT ON THE
-COMPUTATION COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE
-SPINNING WHEEL BOOK\@.  Perl 5.10 regexes are more powerful than
-traditional regexes: it is possible to match correctly balanced brackets
-nested to an arbitrary depth, e.g.\
-\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
+for example in~\cite{intersection-of-NFA-using-Z}) uses DeMorgan's law:
+$FA1~\cap{}~FA2~=~\overline{(\overline{FA1}~\cup{}~\overline{FA2})}$, which
+has considerable computation complexity.  XXX COMMENT ON THE COMPUTATIONAL
+COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE SPINNING WHEEL
+BOOK\@.  Perl 5.10 regexes are more powerful than traditional regexes: it
+is possible to match correctly balanced brackets nested to an arbitrary
+depth, e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
 \verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
-regex engine to maintain state on a stack, so Perl 5.10 regexes are thus
+regex engine to maintain state on a stack, so Perl 5.10 regexes are
 equivalent in computational power to \acronym{PDA} or a \acronym{CFL};
@@
 intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
-complexity of calculating \acronym{FA} or \acronym{PDA} intersection.
-This is certainly not a task to be performed every time the parser runs:
-naive detection of overlap amongst the Postfix log parser's
-\numberOFrules{} rules would require calculating
-\numberOFruleINTERSECTIONS{} intersections.  When detecting overlap any
-conditions attached to rules must be taken into account, because two rules
-whose regexes overlap may have conditions attached which prevent the
-overlap in practice.  A less naive approach to overlap detection would
-first check for overlapping conditions amongst rules, and then check for
-overlapping regexes amongst the rules in each group.  Depending on the
-conditions employed a rule may fall into more than one group, e.g.\ given
-three conditions:
+complexity of calculating \acronym{FA} or \acronym{PDA} intersection.  This
+is certainly not a task to be performed every time the parser runs: naive
+detection of overlap amongst the Postfix log parser's \numberOFrules{}
+rules would require calculating \numberOFruleINTERSECTIONS{} intersections.
+When detecting overlap any conditions attached to rules must be taken into
+account, because two rules whose regexes overlap may have conditions
+attached which prevent the overlap in practice.  A less naive approach to
+overlap detection would first check for overlapping conditions amongst
+rules, and then check for overlapping regexes amongst the rules in each set
+of rules with overlapping conditions.  Depending on the conditions employed
+a rule may fall into more than one group, e.g.\ given these three
+conditions:
 
 \begin{enumerate}
+
+    % Reduce inter-item spacing because each item is a single line.
+    \setlength{\itemsep}{0pt}
 
@@
 
-            y & = & cost of checking for overlap between two conditions \\
-            x & = & cost of checking for overlap between two regexes    \\
-     overlaps & = & set of sets of rules with overlapping conditions    \\
+            $y$ & = & cost of checking for overlap between two conditions \\
+            $x$ & = & cost of checking for overlap between two regexes    \\
+     $overlaps$ & = & set of sets of rules with overlapping conditions    \\
 
@@
 It is possible to define pathological regexes which fall into two main
-categories: regexes that match every input, and regexes that consume
-excessive amounts of CPU time during matching.  Defining a regex to match
-all inputs is trivial: \verb!/^/! matches the start of every input.  This
-regex would be highlighted by a tool that detects overlapping rules, and
-would easily be noticed by visual inspection, but more complex regexes
-would be harder to spot.  XXX MAYBE GIVE AN EXAMPLE REGEX\@?  Usually
-excessive CPU time is consumed when a regex with a lot of alteration and
-variable quantifiers fails to match, but successful matching is generally
-quite fast (see~\cite{mastering-regular-expressions} for in-depth
-discussion).  XXX THIS NEEDS TO BE EXPANDED, OR MOVED TO FIT IN
-ELSEWHERE\@.
+categories: regexes that match inputs they should not, and regexes that
+consume excessive amounts of CPU time during matching.  Defining a regex
+that matches inputs it should not is trivial: \verb!/^/! matches the start
+of every input.  This regex would be highlighted by a tool that detects
+overlapping rules, and would easily be noticed by visual inspection, but
+more complex regexes would be harder to spot.  Usually excessive CPU time
+is consumed when a regex with a lot of alteration and variable quantifiers
+fails to match, but successful matching is generally quite fast
+(see~\cite{mastering-regular-expressions} for in-depth discussion).  XXX
+THIS NEEDS TO BE EXPANDED, AND GIVE AN EXAMPLE\@.
 
-    Rules are completely self-contained and can be understood in
-        isolation, without reference to any other rules.  This is similar
-        to a \acronym{CFG}; in \acronym{CFG} terms the parser rules could
-        be described as:
+XXX PATHOLOGICAL CONDITIONS\@.
+
+\subsection{XXX TO BE DEALT WITH}
+
+XXX DUNNO WHERE TO PUT THIS\@; IT SHOULD PROBABLY GO IN THE RULES SECTION,
+BUT WHERE\@?
+
+Rules are self-contained and can be understood in isolation, without
+reference to other rules.  This is similar to a \acronym{CFG}; in
+\acronym{CFG} terms the parser rules could be described as:
 
@@
 
-\section{Architecture Characteristics}
-
-\label{Architecture characteristics}
-
-XXX EXPAND ON ALL THESE\@.
-
-\begin{description}
-
-    \item [Matching rules against inputs is simple:]  The first matching
-        rule determines the action that will be invoked: there is no
-        backtracking to try alternate rules; no attempt is made to pick a
-        \textit{best\/} rule.
-
-    \item [Line oriented:]  The architecture is line oriented at present:
-        there is no facility for rules to consume more input or push unused
-        input back onto the input stream.  The current input can be
-        manipulated by actions, using the cascaded parsing feature.  This
-        was not a deliberate design decision, rather a consequence of the
-        line oriented nature of Postfix log files; more flexible approaches
-        could be pursued.
-
-    \item [Context-free rules:]  Rules can not take into account past or
-        future inputs when classifying the current input, though conditions
-        can affect which rules will be used to classify each input.
-
-    \item [Context-aware actions:] Actions can consult the results (or lack
-        of results) of previous actions during execution, providing some
-        context sensitivity.
-
-    \item [Cascaded parsing:] Actions can return a modified input to be
-        parsed as if read from the input stream, allowing for a simplified
-        version of cascaded parsing~\cite{cascaded-parsing}.
-
-    \item [Transduction:]  The architecture can be thought of as
-        implementing transduction: it takes data in one form
-        and transforms it to another form; \parsername{} transforms log
-        files to a database.
-
-    \item [Similarity to Natural Language Processing:] Unlike traditional
-        parsers such as those used when compiling a programming language,
-        this architecture does not require a fixed grammar specification
-        that inputs must adhere to.  The architecture is capable of dealing
-        with interleaved inputs, out of order inputs, and ambiguous inputs
-        where heuristics must be applied --- all have arisen and been
-        successfully accommodated in the Postfix log file parser.
-
-\end{description}
+Rules can not take into account past or future inputs when classifying the
+current input, though conditions can affect which rules will be used to
+classify each input.  The architecture is line oriented at present: there
+is no facility for rules to consume more input or push unused input back
+onto the input stream.  The current input can be manipulated by actions,
+using the cascaded parsing feature.  This was not a deliberate design
+decision, rather a consequence of the line oriented nature of Postfix log
+files; more flexible approaches could be pursued.
 

### 2009-02-28 doc/logparser-acronyms.tex
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

@@
     thesis: the original protocol definition is in RFC~821~\cite{RFC821},
-    updated in RFC~2821~\cite{RFC2821}; introductory guides can be found
-    at~\cite{smtp-intro-01,smtp-intro-02}.%
+    updated in RFC~2821~\cite{RFC2821}; introductory guides can be found at
+    \urlLastChecked{http://db.glug-bom.org/lug-authors/philip/docs/mail-stuff/smtp-intro.html}{2009/02/28}
+    if more information is needed.%
 }

### 2009-02-28 doc/logparser-architecture.tex
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

@@
 CHAPTER\@.
+
+XXX INTEGRATE THE NEXT PARAGRAPH INTO THIS SECTION\@.
+
+The solution developed is conceptually simple: provide a few generic
+functions (\textit{actions\/}), each capable of dealing with an entire
+category of inputs (e.g.\ rejecting a mail delivery attempt), accompanied
+by a multitude of precise patterns (\textit{rules\/}), each of which
+matches all inputs of a specific type and only that type (e.g.\ rejection
+by a specific \acronym{DNSBL}).  It is an accepted standard to separate the
+parsing procedure from the declarative grammar it operates with; part of
+the novelty here is in the way that the grammar is itself partially
+procedural (each action is a separate procedure).  This architecture is
+ideally suited to parsing inputs where the input is not fully understood or
+does not conform to a fixed grammar: the architecture warns about unparsed
+inputs and other errors, but continues parsing as best it can, allowing the
+developer of a new parser to decide which deficiencies are most important
+and require attention first, rather than being forced to fix the first
+error that arises.
 
@@
 COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE SPINNING WHEEL
-BOOK\@.  Perl 5.10 regexes are more powerful than traditional regexes: it
-is possible to match correctly balanced brackets nested to an arbitrary
-depth, e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/! matches
-\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
-regex engine to maintain state on a stack, so Perl 5.10 regexes are
+BOOK\@.  Perl 5.10 regexes~\cite{perlre} are more powerful than traditional
+regexes: it is possible to match correctly balanced brackets nested to an
+arbitrary depth, e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!
+matches \verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires
+the regex engine to maintain state on a stack, so Perl 5.10 regexes are
 equivalent in computational power to \acronym{PDA} or a \acronym{CFL};

### 2009-02-28 doc/logparser-glossary.tex
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

@@
 % The contents of the glossary.
-% XXX CHECK FOR ENTRIES THAT SHOULD BE INCLUDED.
 \newglossaryentry{SQLite}{name={SQLite},description={
@@
 
-%\newglossaryentry{hash}{name={hash},description={
-%    A hashing function transforms a string of characters to a number.
-%    There are many possible uses for the resulting number: a common usage
-%    is to maintain a data structure indexed by strings in an efficient
-%    manner.  A full description is beyond the scope of this paper, further
-%    information can be found at~\cite{hash-functions}.
-%}}
+\newglossaryentry{hash}{name={hash},description={
+    A hashing function transforms a string of characters to a number.
+    There are many possible uses for the resulting number: a common usage
+    is to maintain a data structure indexed by strings in an efficient
+    manner.  A full description is beyond the scope of this paper, further
+    information can be found at~\cite{hash-functions}.
+}}
 
@@
     the victim or the server has reached a limit, e.g.\ too many mails in
-    their mailbox, too much disk space consumed.
+    their mailbox, or too much disk space consumed by the mail queue.
 }}
@@
     and mail for \texttt{bob@example.com} is delivered to
-    \texttt{alice@example.net}.  XXX EXPLAIN WHY THIS IS NOT AUTOMATICLLY
-    DETECTED SOMETIMES\@.
+    \texttt{alice@example.net}.  Postfix adds a \texttt{Delivered-To:}
+    header to all mails it delivers locally to detect mail loops, but in
+    some cases this header is removed by the program that forwards the
+    mail.
 }}

### 2009-02-28 doc/logparser-implementation.tex
Corrections to the background chapter.
Checks for lines containing only spaces and "log line" in the architecture chapter.
Corrections, improvements, and new content in architecture chapter. Added SGAI and Elsevier entries to bibliography. Some minor cleanup elsewhere.
Corrections from proofreading of architecture chapter.
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

@@
 
+Each log line contains the name of the Postfix component that produced it,
+and this information is used when determining which rules should be used to
+parse each log line (see \sectionref{rules in implementation} for details).
+
 \subsection{Overlapping Rules}
 
-\label{overlapping rules in architecture}
-
-XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?
-
-XXX DFA COMPARISON\@.
+\label{overlapping rules in implementation}
 
@@
         was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash~\cite{hash-functions} from a small range
-        of values for each word of each log line, incrementing a counter
-        for each hash.  The counters will be used in the next pass to
-        filter out words: if the word's hash does not have a high
+        pass it generated a hash~\cite{hash-functions}\glsadd{hash} from a
+        small range of values for each word of each log line, incrementing
+        a counter for each hash.  The counters will be used in the next
+        pass to filter out words: if the word's hash does not have a high
         frequency, the word itself cannot have a high frequency, so there

### 2009-02-28 doc/logparser-introduction.tex
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

@@
 \label{introduction}
+
+The architecture described herein was developed as part of a larger project
+to improve anti-spam defences by analysing the performance of the set of
+filters currently in use, optimising the order and membership of the set
+based on that analysis, and developing supplemental filters where
+deficiencies are identified.  Most anti-spam techniques are content-based
+(e.g.~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}) and
+require the mail to be accepted before determining if it is spam, but
+rejecting mail during the delivery attempt is preferable: senders of
+non-spam mail that is mistakenly rejected will receive an immediate
+non-delivery notice; resource usage is reduced on the accepting mail server
+(allowing more intensive content-based techniques to be used on the mail
+that is accepted); users have less spam mail to wade through.  Improving
+the performance of anti-spam techniques that are applied when mail is being
+transferred via \acronym{SMTP} is the goal of this project, by providing a
+platform for reasoning about anti-spam filters.  The approach chosen to
+measure performance is to analyse the log files produced by the
+\acronym{SMTP} server in use, Postfix
+(\urlLastCheckedNoParens{http://www.postfix.org/}{2009/02/28}), rather than
+modifying its source code to generate statistics: this approach improves
+the chances of other Postfix users testing and using the software developed
+for this project.  The need arose for a parser capable of dealing with the
+great number and variety of log lines produced by Postfix: the parser must
+be designed so that adding support for parsing new inputs is a simple task,
+because the log lines to be parsed will change over time.
 
@@
 with the source code, available from
-\urlLastChecked{http://www.postfix.org/}{2009/02/23}} Further complication
-arises because the set of rejection messages is not fixed: new messages can
-be added by the administrator with custom checks; every
-\acronym{DNSBL}\footnote{This document is supplied with a glossary, see
-\textsection\ref{Glossary}.} returns a different explanatory message;
+\urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
+complication arises because the set of rejection messages is not fixed: new
+messages can be added by the administrator with custom checks; every
+\acronym{DNSBL}\footnote{This thesis is supplied with a glossary
+(\textsection\ref{Glossary}) and a list of acronyms
+(\textsection\ref{Acronyms}).} returns a different explanatory message;
 policy servers may log different messages depending on the characteristics
@@
 between servers, even within the same organisation: servers may be
-configured differently, or running different versions of Postfix.  This
-paper documents the difficult process of parsing Postfix log files,
-presenting \parsername{}, a program that parses Postfix log files and
-places the resulting data into a database for later use.  The gathered data
-can then be used to optimise current anti-spam measures, provide a baseline
-to test new anti-spam measures against, or to produce statistics showing
-how effective those measures are.  Numerous other uses are possible for
-such data: improving server performance by identifying troublesome
-destinations and reconfiguring appropriately; identifying regular high
-volume uses (e.g.\ customer newsletters) and restricting those uses to
-off-peak times; detecting virus outbreaks that propagate via mail; as a
-base for billing customers on a shared server.  Preserving the raw data
-enables users to develop a multitude of uses far beyond those conceived of
-by the author.
+configured differently, or running different versions of Postfix.
+
+This thesis documents the difficult process of parsing Postfix log files,
+presenting an architecture designed to enable the users of a parser to
+easily extend it to parse their particular inputs, without requiring a lot
+of work or a high level of understanding of the parsing process and the
+parser's internal workings.  This architecture is the basis of
+\parsername{}, a program that parses Postfix log files and places the
+resulting data into a database for later analysis.  The gathered data can
+be used to optimise current anti-spam measures, provide a baseline to test
+new anti-spam measures against, or to produce statistics showing how
+effective those measures are.  Numerous other uses are possible for such
+data: improving server performance by identifying troublesome destinations
+and reconfiguring appropriately; identifying regular high volume uses
+(e.g.\ customer newsletters) and restricting those uses to off-peak times
+or providing a dedicated service for them; detecting virus outbreaks that
+propagate via mail; as a base for billing customers on a shared server.
+Preserving the raw data enables users to develop a multitude of uses far
+beyond those conceived of by the author.
 
@@
 
-XXX THIS ALL NEEDS TO BE CHECKED AND UPDATED\@.
-
 Chapter~\ref{background} provides background information useful in
-understanding the thesis, parser, and architecture.
+understanding the motivation behind the project, \acronym{SMTP}, and
+Postfix.
 
 Chapter~\ref{state of the art review} reviews both the previously published
-research in this area and other available Postfix log file parsers,
-discussing why they were deemed unsuitable for the task, including why they
-could not be improved or expanded upon.
+research in this area and other Postfix log file parsers, discussing why
+they were deemed unsuitable for the task, including why they could not be
+improved or expanded upon.
+
+XXX THIS ALL NEEDS TO BE CHECKED AND UPDATED\@.
 
@@
 
-\section{XXX NOTES}
+\section{Previously Published Work}
 
-Portions of \sectionref{parser architecture}, \sectionref{Postfix Parser
-Implementation}, and \sectionref{Results} have previously been published at
-a conference~\cite{sgai-2008}, and later reprinted in a
+Portions of \textsection\ref{introduction}, \textsection\ref{parser
+architecture}, \textsection\ref{Postfix Parser Implementation}, and
+\textsection\ref{Results} have previously been published at a
+conference~\cite{sgai-2008}, and later reprinted in a
 journal~\cite{elsevier-2009}.  Publication of the conference paper and
-journal article above was supported by Science Foundation Ireland
+journal article was supported by Science Foundation Ireland
 RFP~05/RF/CMS002.

### 2009-02-28 doc/logparser-other-parsers-review.tex
Corrections, improvements, and new content in architecture chapter. Added SGAI and Elsevier entries to bibliography. Some minor cleanup elsewhere.
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.

@@
 quite basic to much more mature, depending on the needs of the author of
-the parser.  None of those parsers were suitable for this project, so the
-decision was taken to write a new parser.  The first parser reviewed is the
-only previously published research in this area that the author is aware
-of; it parses Postfix log files, but aim of the research is to show that
-presenting extracted data in a more easily accessible format is useful to
-systems administrators, rather than to improve anti-spam techniques.
+the parser.
 
-The same ten parsers have been reviewed and compared with \parsername{},
-this project's finished parser, to show how much effort would have been
-required to fulfil the aims and requirements of this project.  It is
+It was hoped to reuse an existing parser rather than writing one from
+scratch, but the existing parsers considered were rejected for one or more
+reasons.  The effort required to adapt and improve an existing parser was
+judged to be greater than the effort to write a new one, because the
+techniques used by the existing parsers severely limited their potential:
+some ignored the majority of log lines, parsing specific log lines
+accurately, but without any provision for parsing new or similar log lines;
+others sloppily parsed the majority of log lines, but were incapable of
+distinguishing between log lines of the same category, e.g.\ rejecting a
+mail delivery attempt.  The first parser reviewed~\cite{log-mail-analyser}
+is the only previously published research in this area that the author is
+aware of; their research aims to show that providing the data from log
+files in a more accessible form is helpful to systems administrators.
+
+The ten parsers have been reviewed and compared with \parsername{}, this
+project's finished parser, to show how much effort would have been required
+to use those to fulfil the aims and requirements of this project.  It is
 important to compare and contrast newly developed algorithms and parsers
 against those already available, to accurately judge what improvements, if
-any, are delivered by the newcomers.  Each of the reviewed parsers was
-tested with the \numberOFlogFILES{} test log files described in
-\sectionref{parser efficiency}.  \Tableref{Summary of parsers' features}
-summarises the results of this review.
+any, are delivered by the newcomers.
 
@@
 \end{enumerate}
+
+Each of the reviewed parsers was tested with the \numberOFlogFILES{} test
+log files described in \sectionref{parser efficiency}; \tableref{Summary of
+parsers' features} summarises the results of this review.
 

### 2009-03-02 doc/logparser-background.tex
Updated full run database.
Corrections, improvements, and new content in architecture chapter. Added SGAI and Elsevier entries to bibliography. Some minor cleanup elsewhere.
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.
Some minor formatting fixes. Documented some of 'make check'. Added check for sentences with only one space separating them.

@@
 query}):
-
-XXX db.full-run.sq3 NEEDS TO BE UPDATED AND EVERYTHING REBUILT\@.
 
@@
 250 smtp.example.com
-MAIL FROM: Alice N. Other alice@example.com
+MAIL FROM: Alice in Distribution alice@example.com
 501 5.1.7 Bad sender address syntax
@@
 
-        XXX SHOULD I URLS OR BIBLIOGRAPHY ENTRIES\@?  SHOULD I SCRAP ALL
-        THE URLS AND MOVE EVERYTHING INTO THE BIBLIOGRAPHY\@?  ASK CARL\@.
-
 \end{itemize}

### 2009-03-02 doc/logparser-architecture.tex
Improved the Framework section of the Rules chapter, and made some minor supporting changes elsewhere.

--- 2009-02-27 doc/logparser-architecture.tex
@@
 new rules to parse new inputs as easy as possible, to enable administrators
-to properly parse their own log files.  To enable this the architecture is
-divided into three parts: Framework, Actions, and Rules.  Each will be
-discussed separately, but first an overview:
+to properly parse their own log files.  The solution developed is to divide
+the architecture into three parts: Framework, Actions, and Rules.  Each
+will be discussed separately, but first an overview:
 
-\begin{eqlist}
+\begin{eqlist}[\def\makelabel#1{\bfseries#1}]
 
@@
     \item [Actions] Each action performs the work required to deal with a
-        single category of inputs, e.g.\ rejecting a delivery attempt.
-        Actions are invoked to process an input once it has been classified
-        by a rule.
+        single \textit{category\/} of inputs, e.g.\ rejecting a delivery
+        attempt.  Actions are invoked to process an input once it has been
+        classified by a rule.
 
     \item [Rules]  The rules are responsible for classifying inputs; each
-        rule recognises one input variant, and specifies the action that
-        will be invoked to process the input. Rules provide an easily
+        rule recognises one input \textit{variant}, and specifies the
+        action that will be invoked to process the input.  Each input
+        category may have many input variants.  Rules provide an easily
         extensible method of associating inputs with actions.
@@
 heuristics must be applied --- all have arisen and been successfully
-accommodated in the Postfix log file parser.
+accommodated in the Postfix log file parser.  This architecture is ideally
+suited to parsing inputs where the input is not fully understood or does
+not conform to a fixed grammar: the architecture warns about unparsed
+inputs and other errors, but continues parsing as best it can, allowing the
+developer of a new parser to decide which deficiencies are most important
+and the order to address them in, rather than being forced to fix the first
+error that arises.
 
 \section{Framework}
-
-XXX EXTEND FRAMEWORK SECTION\@: EXPAND ON ALL OF THE POINTS IN THE LIST IN
-THE NEXT PARAGRAPH\@.
 
@@
 deal with the variation in inputs, and the actions deal with the
-intricacies and complications encountered during parsing.
-
-The function that finds the rule matching the input and invokes the
-requested action can be expressed in pseudo-code as:
+intricacies and complications encountered during parsing.  The function
+that finds the rule matching the input and invokes the requested action can
+be expressed in pseudo-code as:
 
@@
     for each rule defined by the user {
-        if this rule matches the input {
+        if this rule recognises the input {
             perform the action specified by the rule
@@
 
-Most parsers will require the same basic functionality from the framework,
-plus some specialised support functions.  XXX EXPAND THIS\@.
+Most parsers will require the same basic functionality from the framework;
+it is responsible for managing the parsing process from start to finish,
+which generally will involve the following steps:
 
-XXX TURN THIS INTO A LIST AND EXPAND THE ITEMS\@.  It provides shared
-storage to pass data between actions, saves and loads state, loads and
-validates rules, manages parsing, invokes actions, stores results of
-parsing, and miscellaneous other tasks.
+\begin{description}
+
+    \item [Register actions]  Each action needs to be registered with the
+        framework so that the framework knows about it; the list of
+        registered actions will alsobe used when validating rules.
+
+    \item [Load and validate rules]  The framework loads the rules from
+        wherever they are stored: a simple file, a database, or possibly
+        even a web server or other network service --- though that would
+        have serious security implications.  It validates each rule to
+        hopefully catch problems as early in the parsing process as
+        possible; the checks will be implementation-specific to some
+        extent, but will generally include some of the following:
+
+        \begin{itemize}
+
+            \squeezeitems{}
+
+            \item Ensuring the specified action has been registered.
+
+            \item Checking for conflict in the data to be extracted, e.g.\
+                setting the same variable twice.
+
+            \item Checking that the regex used in the rule is valid.
+
+            \item Some optimisation steps may also be performed during
+                loading of rules, as discussed in \sectionref{parser
+                efficiency}.
+
+        \end{itemize}
+
+    \item [Classify the input]  The pseudo-code above shows how rules are
+        successively tried until one is found that recognises the current
+        input.  That pseudo-code is very simple: there may be efficiency
+        concerns, (\sectionref{parser efficiency}), rule conditions
+        (\sectionref{attaching conditions to rules}), or prioritisations
+        (\sectionref{prioritisation and rule ordering}) that complicate the
+        process.
+
+    \item [Invoke actions]  Once a rule has been found that recognises the
+        current input, the specified action must be invoked.  The framework
+        marshals the data extracted by the rule, invokes the action, and
+        repeats the parsing process if cascaded parsing
+        (\sectionref{cascaded parsing}) is used.
+
+    \item [Shared storage]  Parsers generally need to save some state
+        information about the input being parsed, e.g.\ a compiler tracking
+        which variables are in lexical scope as it moves from one lexical
+        block to another.  The framework provides shared storage to deal
+        with this and any other storage needs the actions have.  Actions
+        may need to exchange data to correctly parse the input, e.g.\
+        setting or clearing flags, maintaining a list of previously used
+        identifiers, or ensuring at a higher level that the input being
+        parsed meets a list of requirements.
+
+    \item [Save and load state]  The architecture can save current state,
+        viz the shared storage it provides for actions, and reload it
+        later, so that information is not lost between parsing runs.
+
+    \item [Specialised support functions]  Actions may need support or
+        utility functions, and in general those functions should be
+        available to all action.  Depending on the implementation the
+        framework may be a good location for support functions; if there is
+        another way to make those functions available to all actions it may
+        be preferable to use it, maintaining a clear separation of
+        concerns.
+
+\end{description}
 
@@
 
+\subsection{Cascaded Parsing}
+
+\label{cascaded parsing}
+
 Actions may return a modified input line that will be parsed as if read
@@
 Glucose: $C_{6}H_{12}O_{6}$
-Hydrocarbon: $CH_{3}{(CH_{2})}_{50}CH_{3}$
+Hydrocarbon (polymer): $CH_{3}{(CH_{2})}_{50}CH_{3}$
 
@@
     % Reduce inter-item spacing because each item is a single line.
-    \setlength{\itemsep}{0pt}
+    \squeezeitems{}
 

### 2009-03-02 doc/logparser-architecture.tex
Rewrote some of the actions section to deal with XXX items. Wrote the cascaded parsing section. Added script to demonstrate cascaded parsing.

@@
 handle one category of inputs, with each rule recognising one input
-variant.  XXX EXPAND AND CLARIFY\@; THE MESSAGE IS HERE, IT JUST NEEDS TO
-BE CLEARER\@.  It is anticipated that parsers based on this architecture
-will have a high ratio of rules to actions, with the aim of having simple
-rules and clear distinctions between the inputs parsed by different rules.
-An action may process different input variants in slightly different ways,
-but large changes in processing indicate the need for a new action and a
-new category of input; if an action becomes unnecessarily complicated it
-starts to turn into a monolithic parser, with too much logic contained in a
-single procedure.
+variant.  It is anticipated that parsers based on this architecture will
+have a high ratio of rules to actions, with the aim of having simple rules
+and tightly focused actions.  An action may need to process different input
+variants in slightly different ways, but large changes in processing
+indicate the need for a new action and a new category of input; if an
+action becomes overly complicated it starts to turn into a monolithic
+parser, with too much logic contained in a single procedure.
 
@@
 and new requirements that are discovered during parser development is one
-of the strengths of this architecture.  XXX REWRITE, EXPAND, AND CLARIFY UP
-TO THE NEXT XXX\@.  Instead of writing a single monolithic function that
-must be modified to support new behaviour, with all the attendant risks of
-adversely affecting the existing parser, when a new requirement arises an
-independent action can be written to satisfy it.  Sometimes the new action
-will require the cooperation of other actions, e.g.\ to set or check a
-flag, or save some extra data.  There is a possibility of introducing
-failure when modifying existing actions in this way, but the modifications
-will be smaller and occur less frequently than with a monolithic
-architecture, thus failures will be less likely to occur and will be easier
-to test for and diagnose.  XXX FINISH REWRITING\@.  The architecture can be
-implemented in an object oriented style, allowing sub-classes to extend or
-override actions in addition to adding new actions; because each action is
-an independent procedure, the sub-class need only modify the action it is
-overriding, and can inherit the others rather than reproducing large chunks
-of functionality.
+of the strengths of this architecture.  Instead of writing a single
+monolithic function that processes every input and must be modified to
+support new behaviour, with all the attendant risks of adversely affecting
+the existing parsing, when a new requirement arises an independent action
+can be written to satisfy it.  Sometimes the new action will require the
+cooperation of other actions, e.g.\ to set or check a flag, so actions are
+not always self-contained, but there will still be a far lower degree of
+coupling and interdependency than in a monolithic parser.
 
 During development of the Postfix log parser it became apparent that in
-addition to the obvious variety in log~lines there were
-many complications to be overcome.  Some were the result of deficiencies in
-Postfix's logging (some of which were rectified by later versions of
-Postfix, e.g.\ identifying bounce
-notifications~\sectionref{identifying-bounce-notifications}); others were
-due to the vagaries of process scheduling, client behaviour, and
-administrative actions.  All were successfully accommodated in the Postfix
-log parser: adding new actions was enough to overcome several of the
-complications; others required modifications to a single existing action to
-work around the difficulties; the remainder were resolved by adapting
-existing actions to cooperate and exchange extra data (via the framework),
-changing their behaviour as appropriate based on that extra data.  Every
-architecture should aim to make the hard things possible as well as making
-the easy things easy; the successful implementation of the Postfix log
-parser demonstrates that this architecture achieves that aim.
-
-Actions are not context-sensitive, but they can consult the results (or
-lack of results) of previous actions during execution, providing some
-context sensitivity.  XXX EXPAND\@.
+addition to the obvious variety in log~lines there were many complications
+to be overcome.  Some were the result of deficiencies in Postfix's logging,
+and some of those were rectified by later versions of Postfix, e.g.\
+identifying bounce notifications
+(\sectionref{identifying-bounce-notifications}); others were due to the
+vagaries of process scheduling, client behaviour, and administrative
+actions.  All were successfully accommodated in the Postfix log parser:
+adding new actions was enough to overcome several of the complications;
+others required modifications to a single existing action to work around a
+difficulty; the remainder were resolved by adapting existing actions to
+cooperate and exchange extra data (via the framework), changing their
+behaviour as appropriate based on that extra data.  Every architecture
+should aim to make the hard things possible as well as making the easy
+things easy; the successful implementation of the Postfix log parser
+demonstrates that this architecture achieves that aim.
 
@@
 
-Actions may return a modified input line that will be parsed as if read
-from the input stream, allowing for a simplified version of cascaded
-parsing~\cite{cascaded-parsing}.  This powerful facility allows several
-rules and actions to parse a single input, potentially simplifying both
-rules and actions.  XXX EXPAND WITH CHEMICAL FORMULAE EXAMPLE\@.
+Actions may return a modified input line to the framework that will be
+parsed as if read from the input stream, allowing for a simplified version
+of cascaded parsing~\cite{cascaded-parsing}.  This powerful facility allows
+several rules and actions to parse a single input, potentially simplifying
+both rules and actions.  XXX EXPAND WITH CHEMICAL FORMULAE EXAMPLE\@.
 
-Glucose: $C_{6}H_{12}O_{6}$
-Hydrocarbon (polymer): $CH_{3}{(CH_{2})}_{50}CH_{3}$
+A simple example is to have one rule and action removing comments from
+input lines, so that other rules and actions do not have to handle comments
+at all; obviously if comment characters can be escaped or embedded in
+quoted strings the implementation needs to be more complex, but for some
+inputs this can greatly simplify subsequent parsing.
+
+XXX IS WEIGHT THE CORRECT WORD HERE\@?  SHOULD IT BE xxx WEIGHT\@?  GET
+GREG TO PROOFREAD THIS SECTION\@.
+
+A more complicated example is determining the weight of a compound given
+its chemical formula.  A chemical formula is a sequence of one or two
+character chemical element abbreviations, each optionally followed by a
+number indicating how many of that element are present in the formula (if
+omitted the number is assumed to be one).  Some parts of a formula may be
+repeated: they are enclosed in parentheses, and the closing parenthesis is
+followed by a number indicating how many are present in the formula.  A
+monomer is a simple compound, i.e.\ it does not have any repeated portions;
+compounds with repeated portions are called polymers.  E.g.\ glucose (a
+monomer) has the chemical formula $C_{6}H_{12}O_{6}$ and hydrocarbon (a
+polymer) is $CH_{3}{(CH_{2})}_{50}CH_{3}$.  The atomic weight of an
+individual element is given in periodic tables, but one element in
+particular will be important in the later explanation: Hydrogen,
+abbreviated H, has an atomic weight of 1.  The weight of a given compound
+is the sum of the atomic weights of each of the elements multiplied by the
+number of times they occur; this is simple to calculate for a monomer, but
+it is harder for a polymer because the weight of the repeated monomer must
+be multiplied by the number of times the monomer is present in the compound
+(e.g.\ for hydrocarbon the weight of $CH_{2}$ must be multiplied by $50$).
+Calculating the weight of a monomer requires a single rule and action: the
+action separates each element in the monomer, looks up its weight,
+multiplies it by the number of times it is present (one if unspecified),
+and adds up the results.  The rule's regex is \verb!^([A-Z]([a-z])?\d*)+$!,
+which matches:
+
+\begin{enumerate}
+
+    \squeezeitems{}
+
+    \item The start of the input.
+
+    \item One or more of: an uppercase character, optionally followed by a
+        lowercase character, followed by zero or more digits.
+
+    \item The end of the input.
+
+\end{enumerate}
+
+The regex does not guarantee that the chemical element abbreviations in the
+formula are correct; the regex could be modified to check that by replacing
+\verb![A-Z]([a-z])?! with all the valid abbreviations, simplifying the
+action at the cost of complicating the regex (which could be automatically
+generated).
+
+Parsing a polymer is more difficult because of the repeated monomers: each
+monomer's weight needs to be calculated, multiplied by the number of times
+it is present in the compound, and that figure added to the sum.  XXX CAN
+POLYMERS BE NESTED\@?  Without cascaded parsing the action would need to
+maintain state when recursively parsing the monomers embedded in the
+polymer.  With cascaded parsing available three rules and actions could be
+used as shown in \tableref{Regexes and Actions to calculate the weight of a
+chemical compound}:
+
+\newlength{\regexwidth}
+% This figure from experimentation; set the length to 0pt, see how big a
+% hbox latex complains about, and then set the length.  Stupid latex cannot
+% do \settowidth{\verb}.  Poo.
+\setlength{\regexwidth}{130pt}
+\newlength{\actionwidth}
+\setlength{\actionwidth}{\textwidth}
+\addtolength{\actionwidth}{-\regexwidth}
+% This is the width of the number column, roughly.
+\addtolength{\actionwidth}{-30pt}
+% This figure also from experimentation.
+\addtolength{\actionwidth}{-24pt}
+\begin{table}[ht]
+    \caption{Regexes and Actions to calculate the weight of a chemical
+    compound}
+    \empty{}\label{Regexes and Actions to calculate the weight of a
+    chemical compound}
+    \begin{tabular}{lp{\regexwidth}p{\actionwidth}}
+
+        \tabletopline{}%
+        No.   & Regex                       & Action \\
+        \tablemiddleline{}%
+        1     & \verb'((?!H\d+([()]|$))'    & Parse a monomer as described above,  \\
+              & \verb' ([A-Z]([a-z])?\d*)+)'& returning \verb!Hnnn!, where \verb!nnn! is
+                                              the weight of the monomer.  The first part
+                                              of the regex prevents it from parsing what
+                                              it returns (\verb!Hnnn!) and the parsing
+                                              process getting stuck in a loop.  \\
+        2     & \verb!\(H(\d+)\)(\d+)!      & Parse a repeated monomer that has been
+                                              rewritten by the first rule and action,
+                                              returning \verb!Hnnn!, where \verb!nnn! is
+                                              the product of the first and second numbers
+                                              captured by the regex.  \\
+        3     & \verb!^H(\d+)+$!            & Repeated applications of rules and
+                                              actions one and two will have rewritten the
+                                              input to a sequence of one or more
+                                              \verb!Hnnn!, where \verb!nnn! is the weight
+                                              of the compound.  This action should do
+                                              whatever is required with the calculated
+                                              weight.  \\
+        \tablebottomline{}%
+
+    \end{tabular}
+\end{table}
+
+Applying the rules and actions above to the hydrocarbon formula results in
+this sequence:
+
+\begin{tabular}{lrl}
+
+    1 & Start with              & $CH_{3}{(CH_{2})}_{50}CH_{3}$ \\
+    2 & Apply rule and action 1 & $H_{15}{(CH_{2})}_{50}CH_{3}$ \\
+    3 & Apply rule and action 1 & $H_{15}{(H_{14})}_{50}CH_{3}$ \\
+    3 & Apply rule and action 1 & $H_{15}{(H_{14})}_{50}H_{15}$ \\
+    5 & Apply rule and action 2 & $H_{15}H_{700}H_{15}$ \\
+    6 & Apply rule and action 1 & $H_{730}$ \\
+    7 & Apply rule and action 3 & 730 \\
+
+\end{tabular}
+
+XXX ARE THERE ANY PROBLEMS TO BE ADDRESSED HERE\@?  I DO NOT THINK THAT I
+NEED TO WORRY ABOUT ORDERING OR PRIORITY, BUT I SHOULD REVIEW TO BE SURE\@.
 

### 2009-03-03 doc/logparser-architecture.tex
Wrote the first half of the rule conditions section; it needs to be augmented with an example, and maybe some more explanation.

@@
 several rules and actions to parse a single input, potentially simplifying
-both rules and actions.  XXX EXPAND WITH CHEMICAL FORMULAE EXAMPLE\@.
+both rules and actions.
 
@@
 (e.g.\ for hydrocarbon the weight of $CH_{2}$ must be multiplied by $50$).
+
 Calculating the weight of a monomer requires a single rule and action: the
@@
 formula are correct; the regex could be modified to check that by replacing
-\verb![A-Z]([a-z])?! with all the valid abbreviations, simplifying the
-action at the cost of complicating the regex (which could be automatically
-generated).
+\verb![A-Z]([a-z])?! with all the valid abbreviations, slightly simplifying
+the action at the cost of complicating the regex (which could be
+automatically generated).
 
@@
 POLYMERS BE NESTED\@?  Without cascaded parsing the action would need to
-maintain state when recursively parsing the monomers embedded in the
+maintain state while recursively parsing the monomers embedded in the
 polymer.  With cascaded parsing available three rules and actions could be
@@
 \newlength{\regexwidth}
-% This figure from experimentation; set the length to 0pt, see how big a
-% hbox latex complains about, and then set the length.  Stupid latex cannot
-% do \settowidth{\verb}.  Poo.
+% These figures come from experimentation; set the length to 0pt, see how
+% big a hbox latex complains about, and then set the length.  Stupid latex
+% cannot do \settowidth{\verb}.  Poo.
 \setlength{\regexwidth}{130pt}
@@
 \addtolength{\actionwidth}{-30pt}
-% This figure also from experimentation.
+% This is the overhead of space around columns, on either side, etc.
 \addtolength{\actionwidth}{-24pt}
+
 \begin{table}[ht]
@@
                                               the weight of the monomer.  The first part
-                                              of the regex prevents it from parsing what
-                                              it returns (\verb!Hnnn!) and the parsing
-                                              process getting stuck in a loop.  \\
+                                              of the regex prevents it from parsing the
+                                              return value of this action (\verb!Hnnn!),
+                                              so the parsing process does not get stuck in
+                                              a loop.  \\
         2     & \verb!\(H(\d+)\)(\d+)!      & Parse a repeated monomer that has been
@@
                                               actions one and two will have rewritten the
-                                              input to a sequence of one or more
-                                              \verb!Hnnn!, where \verb!nnn! is the weight
-                                              of the compound.  This action should do
-                                              whatever is required with the calculated
-                                              weight.  \\
+                                              input to \verb!Hnnn!, where \verb!nnn! is
+                                              the weight of the compound.  This action
+                                              should do whatever is required with the
+                                              calculated weight.  \\
         \tablebottomline{}%
@@
 \end{tabular}
-
-XXX ARE THERE ANY PROBLEMS TO BE ADDRESSED HERE\@?  I DO NOT THINK THAT I
-NEED TO WORRY ABOUT ORDERING OR PRIORITY, BUT I SHOULD REVIEW TO BE SURE\@.
 
@@
 parser could also dynamically create new rules in response to certain
-inputs, e.g.\ XXX GIVE A GOOD EXAMPLE\@.  These avenues of research and
-development have not been pursued by the author, but the architecture
-allows them to easily be undertaken independently.
+inputs, e.g.\ parsing a subroutine declaration could cause a rule to be
+created that parses calls to that subroutine, checking that the arguments
+used agree with the subroutine's signature.
+
+These avenues of research and development have not been pursued by the
+author, but the architecture allows them to easily be undertaken
+independently.
 
@@
 
-XXX ATTACHING CONDITIONS TO RULES\@; GIVE PLP EXAMPLE AND REFER TO
-IMPLEMENTATION AND RESULTS SECTIONS\@?
+Rules can specify conditions that will be evaluated by the framework before
+attempting to use a rule to recognise an input: if the condition evaluates
+to true the rule will be used, if not the rule will be skipped.  Conditions
+can be as simple or complex as the parser requires, though naturally as the
+complexity rises so does the difficulty in understanding how everything
+interacts.  The framework has to evaluate each condition, so as the
+complexity of conditions increases so does the complexity of the code
+required to evaluate them; beyond a certain level of complexity conditions
+should probably be written in a proper programming language, e.g.\ taking
+advantage of dynamic languages' support for evaluating code at run-time, or
+embedding a language like Lua
+(\urlLastCheckedNoParens{http://ww.lua.org/}{2009/03/03}).  If an
+implementation is going to use conditions so complex that they will require
+a full programming language to implement them, the design may need to be
+re-thought, including the choice to use this architecture --- there may be
+other architectures more suitable.
+
+Conditions that examine the input will be the easiest to understand,
+because they can be understood in isolation --- they do not depend on
+variables set by actions or other rules.  Conditions that examine the input
+may be complex, but simple conditions can be quite useful too, e.g.\ every
+Postfix log \empty{}line contains the name of the Postfix component that
+produced it, so every rule used in \parsername{} specifies the component
+whose log \empty{}lines it recognises as a condition, reducing the number
+of rules that will be used when classifying an input (see \sectionref{rules
+in implementation} for details).
+
+Conditions can also check the value of variables that have been set by
+either actions or rules; it will probably be easier to understand how a
+variable's value will be used and changed if it is set by rules only rather
+than by actions, because the chain of checking and setting can be followed
+from rule to rule.  The downside to actions setting variables used in rule
+conditions is action at a distance: to understand when a rule's condition
+will be true requires understanding not just every rule but also every
+action.  The framework probably does not need to support the same level of
+complexity and flexibility when setting variables from rules (actions can
+obviously use complex code if they are setting variables); however, if the
+framework allows complex conditions, that code can probably be easily
+extended to support complex variable assignments too.
 

### 2009-03-03 doc/logparser-architecture.tex
Wrote C-style comment parsing example for rule conditions. Tidied up in a couple of other places.

@@
 
-\newlength{\regexwidth}
 % These figures come from experimentation; set the length to 0pt, see how
@@
 % cannot do \settowidth{\verb}.  Poo.
-\setlength{\regexwidth}{130pt}
-\newlength{\actionwidth}
-\setlength{\actionwidth}{\textwidth}
-\addtolength{\actionwidth}{-\regexwidth}
+\newlength{\chemicalactionwidth}
+\setlength{\chemicalactionwidth}{\textwidth}
 % This is the width of the number column, roughly.
-\addtolength{\actionwidth}{-30pt}
+\addtolength{\chemicalactionwidth}{-30pt}
+% The width of the regex column.
+\addtolength{\chemicalactionwidth}{-130pt}
 % This is the overhead of space around columns, on either side, etc.
-\addtolength{\actionwidth}{-24pt}
+\addtolength{\chemicalactionwidth}{-24pt}
 
@@
     chemical compound}
-    \begin{tabular}{lp{\regexwidth}p{\actionwidth}}
+    \begin{tabular}{llp{\chemicalactionwidth}}
 
@@
 
+An example of how rule conditions can be useful is parsing C-style
+comments, which start with \texttt{/*} and end with \texttt{*/}; the start
+and end tokens can be on one line, or may have many lines between them.
+\Tableref{Rules to parse C-style comments} shows the regexes, conditions,
+and state changes of the four rules required to parse C-style comments.
+These rules are somewhat simplified, e.g.\ rules one and two will recognise
+the comment start token even if it is within a quoted string, which is
+incorrect.  Rules one and two will only be used when the parser is parsing
+code, not comments: rule one recognises a comment that is contained within
+one line and leaves the parser's state unchanged; rule two recognises the
+start of a command and changes the parser's state to parsing comments.
+Rules three and four will be used when the parser is parsing a comment:
+rule three recognises the end of a comment and switches parsing state back
+to parsing code; rule four recognises a comment line without an end token
+and keeps the parser's unchanged.  It is important that the rules are
+applied in the order listed in \tableref{Rules to parse C-style comments};
+\sectionref{prioritisation and rule ordering} explains how this is
+achieved.
+
+\begin{table}[ht]
+    \caption{Rules to parse C-style comments}
+    \empty{}\label{Rules to parse C-style comments}
+    \begin{tabular}{llll}
+        \tabletopline{}%
+        No.   & Regex             & Condition         & New State         \\
+        \tablemiddleline{}%
+        1     & \verb!/\*.*?\*/!  & state == code     & state = code      \\
+        2     & \verb!/\*.*!      & state == code     & state = comment   \\
+        3     & \verb!.*\*/!      & state == comment  & state = code      \\
+        4     & \verb!.*!         & state == comment  & state = comment   \\
+        \tablebottomline{}%
+    \end{tabular}
+\end{table}
+
 Conditions can also check the value of variables that have been set by
@@
 extended to support complex variable assignments too.
-
-XXX STATEFUL PARSING ENABLED BY CONDITIONS, E.G\@.  PARSING C COMMENTS\@.
 

### 2009-03-03 doc/logparser-architecture.tex
Wrote priority/rule ordering paragraph; it turned out that a whole section was overkill.

@@
         (\sectionref{attaching conditions to rules}), or prioritisations
-        (\sectionref{prioritisation and rule ordering}) that complicate the
-        process.
+        (\sectionref{rules in architecture}) that complicate the process.
 
@@
 \parsername{}, and some potential attributes will be discussed later in
-this section (\sectionref{attaching conditions to rules} and
-\sectionref{prioritisation and rule ordering}).  Classifying inputs using
-the rules is simple: the first rule to recognise the input determines the
-action that will be invoked: there is no backtracking to try alternate
-rules; no attempt is made to pick a \textit{best\/} rule.
+\sectionref{attaching conditions to rules}.
+
+Classifying inputs using the rules is simple: the first rule to recognise
+the input determines the action that will be invoked: there is no
+backtracking to try alternate rules; no attempt is made to pick a
+\textit{best\/} rule.  \sectionref{attaching conditions to rules} contains
+an example which requires that the rules are used in a specific order to
+correctly parse the input, so a mechanism is needed to allow the author of
+the rules to specify that ordering.  The framework should be written so
+that it sorts the rules according to a priority attribute specified by the
+rule author, allowing fine-grained control over the order that rules are
+used in.  The priority attribute may be implemented as a number, or as a
+range of values, e.g.\ low, medium, and high, or in a different fashion
+entirely if it suits the implementation.  Rule ordering for efficiency is a
+separate topic that is covered in \sectionref{rule ordering for
+efficiency}; overlapping rules are discussed in \sectionref{overlapping
+rules in architecture}.
 
@@
 applied in the order listed in \tableref{Rules to parse C-style comments};
-\sectionref{prioritisation and rule ordering} explains how this is
-achieved.
+\sectionref{rules in architecture} has explained how this is achieved, and
+\sectionref{overlapping rules in architecture} discusses overlapping rules.
 
@@
 extended to support complex variable assignments too.
-
-\subsection{Prioritisation and Rule Ordering}
-
-\label{prioritisation and rule ordering}
-
-XXX PRIORITY ATTRIBUTE AND RULE ORDERING\@.
-
-The first matching rule wins: no further rules are tried against that
-input, but there is a mechanism for prioritising the rules so that more
-specific rules can be tried first.
 

### 2009-03-03 doc/logparser-architecture.tex
Dealt with some XXX items, and improved the pathological rules section.

@@
 
-XXX IS WEIGHT THE CORRECT WORD HERE\@?  SHOULD IT BE xxx WEIGHT\@?  GET
-GREG TO PROOFREAD THIS SECTION\@.
+XXX GET GREG TO PROOFREAD THIS SECTION\@.
 
-A more complicated example is determining the weight of a compound given
-its chemical formula.  A chemical formula is a sequence of one or two
+A more complicated example is determining the atomic weight of a compound
+given its chemical formula.  A chemical formula is a sequence of one or two
 character chemical element abbreviations, each optionally followed by a
@@
 particular will be important in the later explanation: Hydrogen,
-abbreviated H, has an atomic weight of 1.  The weight of a given compound
-is the sum of the atomic weights of each of the elements multiplied by the
-number of times they occur; this is simple to calculate for a monomer, but
-it is harder for a polymer because the weight of the repeated monomer must
-be multiplied by the number of times the monomer is present in the compound
-(e.g.\ for hydrocarbon the weight of $CH_{2}$ must be multiplied by $50$).
+abbreviated H, has an atomic weight of 1.  The atomic weight of a given
+compound is the sum of the atomic weights of each of the elements
+multiplied by the number of times they occur; this is simple to calculate
+for a monomer, but it is harder for a polymer because the atomic weight of
+the repeated monomer must be multiplied by the number of times the monomer
+is present in the compound (e.g.\ for hydrocarbon the atomic weight of
+$CH_{2}$ must be multiplied by $50$).
 
-Calculating the weight of a monomer requires a single rule and action: the
-action separates each element in the monomer, looks up its weight,
-multiplies it by the number of times it is present (one if unspecified),
-and adds up the results.  The rule's regex is \verb!^([A-Z]([a-z])?\d*)+$!,
-which matches:
+Calculating the atomic weight of a monomer requires a single rule and
+action: the action separates each element in the monomer, looks up its
+atomic weight, multiplies it by the number of times it is present (one if
+unspecified), and adds up the results.  The rule's regex is
+\verb!^([A-Z]([a-z])?\d*)+$!, which matches:
 
@@
 Parsing a polymer is more difficult because of the repeated monomers: each
-monomer's weight needs to be calculated, multiplied by the number of times
-it is present in the compound, and that figure added to the sum.  XXX CAN
-POLYMERS BE NESTED\@?  Without cascaded parsing the action would need to
-maintain state while recursively parsing the monomers embedded in the
-polymer.  With cascaded parsing available three rules and actions could be
-used as shown in \tableref{Regexes and Actions to calculate the weight of a
-chemical compound}:
+monomer's atomic weight needs to be calculated, multiplied by the number of
+times it is present in the compound, and that figure added to the sum.
+Without cascaded parsing the action would need to maintain state while
+recursively parsing the monomers embedded in the polymer.  With cascaded
+parsing available three rules and actions could be used as shown in
+\tableref{Regexes and Actions to calculate the atomic weight of a chemical
+compound}:
 
@@
 \begin{table}[ht]
-    \caption{Regexes and Actions to calculate the weight of a chemical
-    compound}
-    \empty{}\label{Regexes and Actions to calculate the weight of a
+    \caption{Regexes and Actions to calculate the atomic weight of a
+    chemical compound}
+    \empty{}\label{Regexes and Actions to calculate the atomic weight of a
     chemical compound}
@@
               & \verb' ([A-Z]([a-z])?\d*)+)'& returning \verb!Hnnn!, where \verb!nnn! is
-                                              the weight of the monomer.  The first part
-                                              of the regex prevents it from parsing the
-                                              return value of this action (\verb!Hnnn!),
-                                              so the parsing process does not get stuck in
-                                              a loop.  \\
+                                              the atomic weight of the monomer.  The first
+                                              part of the regex prevents it from parsing
+                                              the return value of this action
+                                              (\verb!Hnnn!), ensuring the parsing process
+                                              does not get stuck in a loop.  \\
         2     & \verb!\(H(\d+)\)(\d+)!      & Parse a repeated monomer that has been
@@
                                               input to \verb!Hnnn!, where \verb!nnn! is
-                                              the weight of the compound.  This action
-                                              should do whatever is required with the
-                                              calculated weight.  \\
+                                              the atomic weight of the compound.  This
+                                              action should do whatever is required with
+                                              the calculated atomic weight.  \\
         \tablebottomline{}%
@@
 
-XXX CHECK WITH CARL WHETHER I SHOULD SAY CFL OR CFG IN THE NEXT
-PARAGRAPH\@.  XXX DO I NEED TO MENTION CFG/CFL AT ALL\@?
-
 Decoupling the rules from the actions allows external tools to be written
@@
 the regex engine to maintain state on a stack, so Perl 5.10 regexes are
-equivalent in computational power to \acronym{PDA} or a \acronym{CFL};
-detecting overlap may require calculating the intersection of two
-\acronym{PDA} or \acronyms{CFL} instead of two \acronym{FA}.  The
-intersection of two \acronyms{CFL} is not closed, i.e.\ the resulting
-language cannot always be parsed by a \acronym{CFL}, so intersection may be
-intractable in some cases, e.g.:
+equivalent in computational power to \acronym{PDA}; detecting overlap may
+require calculating the intersection of two \acronym{PDA} instead of two
+\acronym{FA}.  The intersection of two \acronym{PDA} is not closed, i.e.\
+the result cannot always be implemented using a \acronym{PDA}, so
+intersection may be intractable in some cases, e.g.:
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
@@
 is certainly not a task to be performed every time the parser runs: naive
-detection of overlap amongst the Postfix log parser's \numberOFrules{}
-rules would require calculating \numberOFruleINTERSECTIONS{} intersections.
-When detecting overlap any conditions attached to rules must be taken into
+detection of overlap amongst \parsernames{} \numberOFrules{} rules would
+require calculating \numberOFruleINTERSECTIONS{} intersections.  When
+detecting overlap any conditions attached to rules must be taken into
 account, because two rules whose regexes overlap may have conditions
@@
 that matches inputs it should not is trivial: \verb!/^/! matches the start
-of every input.  This regex would be highlighted by a tool that detects
+of every input.  This regex would be found by a tool that detects
 overlapping rules, and would easily be noticed by visual inspection, but
-more complex regexes would be harder to spot.  Usually excessive CPU time
-is consumed when a regex with a lot of alteration and variable quantifiers
-fails to match, but successful matching is generally quite fast
-(see~\cite{mastering-regular-expressions} for in-depth discussion).  XXX
-THIS NEEDS TO BE EXPANDED, AND GIVE AN EXAMPLE\@.
+more complex regexes would be harder to spot.  Regexes that match more than
+they should are a problem not because of excessive resource usage, but
+because they may prevent the correct rule from recognising the input.  If
+an adaptive ordering system is used (see \sectionref{rule ordering for
+efficiency}) then a rule with a regex that matches too much may be promoted
+up through the list, displacing an increasing number of correct rules as it
+moves.
+
+Usually excessive CPU time is consumed when a regex with a lot of
+backtracking, caused by alteration and/or nested quantifiers, fails to
+match; successful matching is generally quite fast with such regexes, so
+problematic regexes may go unnoticed for some time.  E.g.\ with most regex
+engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"! is very
+fast when there is a match, but when the match fails its computational
+complexity is $O(2^{n})$; see~\cite{mastering-regular-expressions} for
+in-depth discussion of nested quantifiers, backtracking, alteration, and
+capturing groups.  Pathological regexes which consume excessive CPU time
+can be difficult to detect, both by visual inspection and by machine
+inspection, but if a regex is converted to a \acronym{FA} or the internal
+representation used by the regex engine, it may be possible to determine if
+there are nested quantifiers or other troublesome constructs present.
+Modern regex engines have addressed many of these problems, e.g.\ the regex
+to match double quoted strings given above fails immediately with Perl
+5.10, regardless of the input length, because the regex engine looks for
+the required double quotes first.  Similarly Perl 5.10's regex engine
+optimises alterations starting with literal text into a trie, which has
+constant matching time, whereas the matching time for alteration is
+proportional to the number of alternatives.  Perl 5.10 regexes can use
+\verb!(?>pattern)!, which matches \verb!pattern! the first time the regex
+engine passes over it, and does not change what it originally matched if
+the regex engine backtracks over it, so troublesome regexes can be modified
+to use this to reduce the impact of backtracking; Prolog
+(\urlLastCheckedNoParens{http://en.wikipedia.org/wiki/Prolog}{2009/03/03})
+users will notice a similarity to the \verb'!' (cut) operator.  A
+presentation showing some of Perl 5.10's regex features is available at
+\urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.
 
@@
 
-$\text{\textless{}input\textgreater{}} \mapsto \text{rule-1} |
+$\text{\textless{}input\textgreater{}} \mapsto{} \text{rule-1} |
 \text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$

### 2009-03-04 doc/logparser-architecture.tex
Wrote about pathological conditions. Dealt with XXX items.

@@
 \label{parser architecture}
-
-XXX WHEN FINISHED, CHECK ALL REFERENCES TO ENSURE THEY POINT AT THE CORRECT
-CHAPTER\@.
 
@@
 \parsername{}, and some potential attributes will be discussed later in
-\sectionref{attaching conditions to rules}.
+this section.
+
+The architecture is line oriented at present: there is no facility for
+rules to consume more input or push unused input back onto the input
+stream.  This was not a deliberate design decision, rather a consequence of
+the line oriented nature of Postfix log files; more flexible approaches
+could be pursued.
 
 Classifying inputs using the rules is simple: the first rule to recognise
-the input determines the action that will be invoked: there is no
+the input determines the action that will be invoked; there is no
 backtracking to try alternate rules; no attempt is made to pick a
@@
 rules in architecture}.
+
+In \acronym{CFG} terms the rules could be described as:
+
+$\text{\textless{}input\textgreater{}}~\mapsto{}~\text{rule-1}~|~\text{rule-2}~|~\text{rule-3}~|~\dots~|~\text{rule-n}$
+
+This is not entirely correct because the rules are not truly context free:
+rule conditions (described in \sectionref{attaching conditions to rules})
+restrict which rules will be used to recognise each input, imposing a
+context of sorts.
 
@@
 to parsing code; rule four recognises a comment line without an end token
-and keeps the parser's unchanged.  It is important that the rules are
+and keeps the parser's state unchanged.  It is important that the rules are
 applied in the order listed in \tableref{Rules to parse C-style comments};
 \sectionref{rules in architecture} has explained how this is achieved, and
-\sectionref{overlapping rules in architecture} discusses overlapping rules.
+\sectionref{overlapping rules in architecture} will discuss overlapping
+rules.
 
@@
 Conditions can also check the value of variables that have been set by
-either actions or rules; it will probably be easier to understand how a
-variable's value will be used and changed if it is set by rules only rather
-than by actions, because the chain of checking and setting can be followed
-from rule to rule.  The downside to actions setting variables used in rule
+either actions or rules; it will be easier to understand how a variable's
+value will be used and changed if it is set by rules only rather than by
+actions, because the chain of checking and setting can be followed from
+rule to rule.  The downside to actions setting variables used in rule
 conditions is action at a distance: to understand when a rule's condition
@@
 
+Once conditions pass a certain level of complexity determining whether two
+conditions overlap becomes intractable, because it requires so much
+knowledge of other rules and possibly even actions.  E.g.\ given two rules
+with conditions \verb!verbose == true! and \verb!silent == true!, it
+appears that these rules cannot overlap, yet there is nothing to stop both
+variables being set to true by one or more other rules or actions.  If
+variables used in conditions can be set by actions, determining whether two
+conditions overlap is impossible, because understanding how an action will
+change the variables is impossible; the Church-Turing Thesis XXX NEED A BIB
+ENTRY FOR CHURCH-TURING explains why.
+
 \subsection{Pathological Rules}
@@
 
-XXX PATHOLOGICAL CONDITIONS\@.
+Conditions can vary in complexity from simple equality through to a
+Turing-equivalent language, so enumerating pathological conditions is
+difficult if not pointless.  Conditions that examine the input using
+regexes can suffer from the same problems as rule regexes; conditions that
+check variables in uncomplicated ways may exhibit unexpected or incorrect
+behaviour, but are unlikely to exhibit pathological behaviour.  Deciding
+whether a more complex condition's behaviour is pathological or simply a
+bug is difficult and to some extent is a matter of opinion.  The author has
+not experimented with more complex conditions, so this thesis will not
+address this point any further.
 
-\subsection{XXX TO BE DEALT WITH}
-
-XXX DUNNO WHERE TO PUT THIS\@; IT SHOULD PROBABLY GO IN THE RULES SECTION,
-BUT WHERE\@?
-
-Rules are self-contained and can be understood in isolation, without
-reference to other rules.  This is similar to a \acronym{CFG}; in
-\acronym{CFG} terms the parser rules could be described as:
-
-$\text{\textless{}input\textgreater{}} \mapsto{} \text{rule-1} |
-\text{rule-2} | \text{rule-3} | \dots | \text{rule-n}$
-
-Rules can not take into account past or future inputs when classifying the
-current input, though conditions can affect which rules will be used to
-classify each input.  The architecture is line oriented at present: there
-is no facility for rules to consume more input or push unused input back
-onto the input stream.  The current input can be manipulated by actions,
-using the cascaded parsing feature.  This was not a deliberate design
-decision, rather a consequence of the line oriented nature of Postfix log
-files; more flexible approaches could be pursued.
-
-\section{Conclusion}
+\section{Summary}
 

### 2009-03-04 doc/logparser.tex
Dealt with some XXX items. Integrated some more content from the paper in the introduction section.
Improved the Framework section of the Rules chapter, and made some minor supporting changes elsewhere.
Dealt with some XXX items, and improved the pathological rules section.
Wrote about pathological conditions. Dealt with XXX items.

@@
 
-% Show the citation key (e.g. slct-paper) in the margin beside the
+% Show the citation key (e.g.\ slct-paper) in the margin beside the
 % reference.
@@
 %\renewcommand{\refname}{Bibliography}
+
+% Reduce the space between items in a list; this is useful when each item
+% is a single line, because then the defaul spacing makes the list look
+% very sparse.  This command needs to be the first thing in a list to take
+% effect.
+\newcommand{\squeezeitems}[0]{%
+    \setlength{\itemsep}{0pt}%
+}
+
 % Put each URL in the bibliography on a new line.
@@
 }
+% XXX CAN I MAKE THIS LOOK BETTER?  MAYBE USE [] LIKE \pageref{} DOES?
 \newcommand{\urlLastCheckedNoParens}[3][ ]{%
@@
 
-XXX POSSIBLY PUBLISH THE STATE OF THE ART REVIEW IN A JOURNAL\@?
+XXX ASK CARL ABOUT\@:
+
+\begin{itemize}
+
+    \item XXX WHERE SHOULD I USE URLS AND WHERE SHOULD THEY BE BIBLIOGRAPHY
+        ENTRIES\@?
+
+    \item XXX REFERENCE FOR TRANSDUCTION\@.
+
+    \item XXX PREVIOUS RESEARCH IN OVERLAPPING RULES\@?
+
+    \item XXX BORROW SPINNING WHEEL BOOK\@; NEED TO COMMENT ON THE
+        COMPUTATIONAL COMPLEXITY OF FA UNION AND INVERSION\@; MAYBE COMMENT
+        ON COMPUTATIONAL COMPLEXITY OF PDA UNION, INVERSION, AND
+        INTERSECTION\@?
+
+    \item XXX BIB ENTRY FOR CHURCH-TURING\@.
+
+\end{itemize}
 
@@
 
+% Add some glossary entries that should be present, but lack an appropriate
+% place in the text to mark them.
+\glsadd{queueid}
 \renewcommand{\glossarypostamble}{\label{Glossary}}

### 2009-03-05 doc/logparser-acronyms.tex
Removed URLs that are the first Google hit for a search term.

@@
     first={Structured Query Language (SQL)}
-}
-\newglossaryentry{SQL glossary}{
-    name={Structured Query Language},
-    description={
-        The standard language used for database querying, modification, and
-        maintenance.  Some information and history can be found at
-        \urlLastChecked{http://en.wikipedia.org/wiki/SQL}{2009/02/23}; a
-        good introduction can be found at
-        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
-    }
 }

### 2009-03-05 doc/logparser-architecture.tex
Removed URLs that are the first Google hit for a search term.

@@
 advantage of dynamic languages' support for evaluating code at run-time, or
-embedding a language like Lua
-(\urlLastCheckedNoParens{http://ww.lua.org/}{2009/03/03}).  If an
-implementation is going to use conditions so complex that they will require
-a full programming language to implement them, the design may need to be
-re-thought, including the choice to use this architecture --- there may be
-other architectures more suitable.
+embedding a language like Lua.  If an implementation is going to use
+conditions so complex that they will require a full programming language to
+implement them, the design may need to be re-thought, including the choice
+to use this architecture --- there may be other architectures more
+suitable.
 
@@
 the regex engine backtracks over it, so troublesome regexes can be modified
-to use this to reduce the impact of backtracking; Prolog
-(\urlLastCheckedNoParens{http://en.wikipedia.org/wiki/Prolog}{2009/03/03})
-users will notice a similarity to the \verb'!' (cut) operator.  A
-presentation showing some of Perl 5.10's regex features is available at
+to use this to reduce the impact of backtracking; Prolog users will notice
+a similarity to the \verb'!' (cut) operator.  A presentation showing some
+of Perl 5.10's regex features is available at
 \urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.

### 2009-03-05 doc/logparser-implementation.tex
Some minor formatting fixes. Documented some of 'make check'. Added check for sentences with only one space separating them.
Wrote the first half of the rule conditions section; it needs to be augmented with an example, and maybe some more explanation.
Removed URLs that are the first Google hit for a search term.

@@
         introduction to \acronym{SQL} can be found at
-        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23},
-        others are
-        \urlLastChecked{http://www.w3schools.com/sql/default.asp}{2009/02/23},
-        \urlLastChecked{http://sqlcourse.com/}{2009/02/23}).  Although
-        every vendor implements a different dialect of \acronym{SQL}, the
-        basics are the same everywhere (analogous to the overall
-        similarities and minor differences amongst Irish English, British
-        English, American English, and Australian English).  Depending on
-        the database in use there may be tools available that reduce or
-        remove the requirement to know \acronym{SQL}; several are available
-        for \gls{SQLite}, the default database used by the parser
+        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
+        Although every vendor implements a different dialect of
+        \acronym{SQL}, the basics are the same everywhere (analogous to the
+        overall similarities and minor differences amongst Irish English,
+        British English, American English, and Australian English).
+        Depending on the database in use there may be tools available that
+        reduce or remove the requirement to know \acronym{SQL}; several are
+        available for \gls{SQLite}, the default database used by the parser
         implementation.
@@
 
+XXX ONCE THIS PARAGRAPH HAS BEEN INTEGRATED AND EXPANDED UPDATE THE
+REFERENCE IN ARCHITECTURE->RULES->ATTACHING CONDITIONS TO RULES\@.
 Each log line contains the name of the Postfix component that produced it,
@@
         is accepted.\footnote{Tested with Postfix 2.2.10, 2.3.11, and
-        2.4.7; this may possibly have changed in Postfix 2.5. XXX TEST THIS
-        AGAIN.}  Rectifying this is relatively simple: create a restriction
-        that is guaranteed to warn for every accepted mail, as follows:
+        2.4.7; this may possibly have changed in Postfix 2.5.  XXX TEST
+        THIS AGAIN.}  Rectifying this is relatively simple: create a
+        restriction that is guaranteed to warn for every accepted mail, as
+        follows:
 

### 2009-03-05 doc/logparser-introduction.tex
Removed URLs that are the first Google hit for a search term.

@@
 measure performance is to analyse the log files produced by the
-\acronym{SMTP} server in use, Postfix
-(\urlLastCheckedNoParens{http://www.postfix.org/}{2009/02/28}), rather than
-modifying its source code to generate statistics: this approach improves
-the chances of other Postfix users testing and using the software developed
-for this project.  The need arose for a parser capable of dealing with the
-great number and variety of log lines produced by Postfix: the parser must
-be designed so that adding support for parsing new inputs is a simple task,
-because the log lines to be parsed will change over time.
+\acronym{SMTP} server in use, Postfix, rather than modifying its source
+code to generate statistics: this approach improves the chances of other
+Postfix users testing and using the software developed for this project.
+The need arose for a parser capable of dealing with the great number and
+variety of log lines produced by Postfix: the parser must be designed so
+that adding support for parsing new inputs is a simple task, because the
+log lines to be parsed will change over time.
 

### 2009-03-05 doc/logparser-postfix-daemons.tex
Removed URLs that are the first Google hit for a search term.

@@
 \newglossaryentry{sendmail}{name={sendmail},type={postfix},description={
-    A Postfix component that is compatible with the Sendmail
-    (\urlLastCheckedNoParens{http://www.sendmail.org/}{2009/02/23}) mail
+    A Postfix component that is compatible with the Sendmail mail
     submission program which all Unix commands that need to send mail use;

### 2009-03-05 doc/logparser-architecture.tex
Corrections to the first half of the architecture chapter.

@@
         attempt.  Actions are invoked to process an input once it has been
-        classified by a rule.
+        recognised by a rule.
 
-    \item [Rules]  The rules are responsible for classifying inputs; each
-        rule recognises one input \textit{variant}, and specifies the
-        action that will be invoked to process the input.  Each input
-        category may have many input variants.  Rules provide an easily
-        extensible method of associating inputs with actions.
+    \item [Rules]  Rules are responsible for classifying inputs: each rule
+        recognises one input \textit{variant\/} (a single input category
+        may have many input variants).  Each rule also specifies the action
+        to be invoked when an input has been recognised; rules thus provide
+        an extensible method of associating inputs with actions.
 
@@
 If the input is not recognised by any of the rules, the framework issues a
-warning; the framework can continue parsing, although for some inputs
-stopping immediately may be preferable.
+warning; the framework can continue parsing after this, although for some
+inputs stopping immediately may be preferable.
 
@@
 separation of functionality: the framework manages the parsing process and
-provides services to the actions, but does not need to perform anything
-specific to the input being parsed; actions benefit from having services
+provides services to the actions; actions benefit from having services
 provided by the framework, freeing them to concentrate on the task of
@@
 mind that changes to a parser must be made without adversely affecting
-existing parsing, particularly as there may be edge cases that are not
-immediately obvious --- \sectionref{yet-more-aborted-delivery-attempts}
-describes a complication that occurs only four times in \numberOFlogFILES{}
-log files.  Requiring changes to a parser's source code also complicates
-upgrades of the parser, because the changes must be preserved during the
-upgrade, and they may clash with changes made by the developer.  This
-architecture allows the user to add new rules without having to edit a
-parser, unless the new inputs cannot be processed by the existing actions.
-If the new inputs do require new functionality, new actions can be added to
-the parser without having to modify existing actions; only in the rare
-event that the new actions need to cooperate with existing actions will
-more extensive changes be required.
+existing parsing, including any edge cases that are not immediately obvious
+--- \sectionref{yet-more-aborted-delivery-attempts} describes a
+complication that occurs only four times in \numberOFlogFILES{} log files.
+Requiring changes to a parser's source code also complicates upgrades of
+the parser, because the changes must be preserved during the upgrade, and
+they may clash with changes made by the developer.  This architecture
+allows the user to add new rules without having to edit a parser, unless
+the new inputs cannot be processed by the existing actions.  If the new
+inputs do require new functionality, new actions can be added to the parser
+without having to modify existing actions; only in the rare event that the
+new actions need to cooperate with existing actions will more extensive
+changes be required.
 
@@
 heuristics must be applied --- all have arisen and been successfully
-accommodated in the Postfix log file parser.  This architecture is ideally
-suited to parsing inputs where the input is not fully understood or does
-not conform to a fixed grammar: the architecture warns about unparsed
-inputs and other errors, but continues parsing as best it can, allowing the
-developer of a new parser to decide which deficiencies are most important
-and the order to address them in, rather than being forced to fix the first
-error that arises.
+accommodated in \parsername{}.  This architecture is ideally suited to
+parsing inputs where the input is not fully understood or does not conform
+to a fixed grammar: the architecture warns about unparsed inputs and other
+errors, but continues parsing as best it can, allowing the developer of a
+new parser to decide which deficiencies are most important and the order to
+address them in, rather than being forced to fix the first error that
+arises.
 
@@
 it is responsible for managing the parsing process from start to finish,
-which generally will involve the following steps:
+which will generally involve the following:
 
@@
         framework so that the framework knows about it; the list of
-        registered actions will alsobe used when validating rules.
+        registered actions will also be used when validating rules.
 
@@
 
-            \item Ensuring the specified action has been registered.
+            \item Ensuring the action specified by the rule has been
+                registered with the framework.
 
-            \item Checking for conflict in the data to be extracted, e.g.\
+            \item Checking for conflicts in the data to be extracted, e.g.\
                 setting the same variable twice.
 
-            \item Checking that the regex used in the rule is valid.
+            \item Checking that the regex in the rule is valid.
 
@@
         input.  That pseudo-code is very simple: there may be efficiency
-        concerns, (\sectionref{parser efficiency}), rule conditions
+        concerns (\sectionref{parser efficiency}), rule conditions
         (\sectionref{attaching conditions to rules}), or prioritisations
@@
 
-    \item [Shared storage]  Parsers generally need to save some state
+    \item [Shared storage]  Parsers commonly need to save some state
         information about the input being parsed, e.g.\ a compiler tracking
@@
         identifiers, or ensuring at a higher level that the input being
-        parsed meets a list of requirements.
+        parsed meets requirements.
 
-    \item [Save and load state]  The architecture can save current state,
-        viz the shared storage it provides for actions, and reload it
-        later, so that information is not lost between parsing runs.
+    \item [Save and load state]  The architecture can save its current
+        state, i.e.\ the shared storage it provides for actions, and reload
+        it later, so that information is not lost between parsing runs.
 
     \item [Specialised support functions]  Actions may need support or
-        utility functions, and in general those functions should be
-        available to all action.  Depending on the implementation the
-        framework may be a good location for support functions; if there is
-        another way to make those functions available to all actions it may
-        be preferable to use it, maintaining a clear separation of
-        concerns.
+        utility functions; the framework may be a good location for support
+        functions, but if there is another way to make those functions
+        available to all actions it may be preferable to use it instead,
+        maintaining a clear separation of concerns.
 
@@
 input variants within one input category; in general each action will
-handle one category of inputs, with each rule recognising one input
-variant.  It is anticipated that parsers based on this architecture will
-have a high ratio of rules to actions, with the aim of having simple rules
-and tightly focused actions.  An action may need to process different input
-variants in slightly different ways, but large changes in processing
-indicate the need for a new action and a new category of input; if an
-action becomes overly complicated it starts to turn into a monolithic
-parser, with too much logic contained in a single procedure.
+handle one input category, with each rule recognising one input variant.
+It is anticipated that parsers based on this architecture will have a high
+ratio of rules to actions, with the aim of having simple rules and tightly
+focused actions.  An action may need to process different input variants in
+slightly different ways, but large changes in processing indicate the need
+for a new action and a new category of input; if an action becomes overly
+complicated it starts to turn into a monolithic parser, with too much logic
+contained in a single procedure.
 
@@
 monolithic function that processes every input and must be modified to
-support new behaviour, with all the attendant risks of adversely affecting
-the existing parsing, when a new requirement arises an independent action
-can be written to satisfy it.  Sometimes the new action will require the
-cooperation of other actions, e.g.\ to set or check a flag, so actions are
-not always self-contained, but there will still be a far lower degree of
-coupling and interdependency than in a monolithic parser.
+support any new behaviour, with all the attendant risks of adversely
+affecting the existing parsing, when a new requirement arises an
+independent action can be written to satisfy it.  Sometimes the new action
+will require the cooperation of other actions, e.g.\ to set or check a
+flag, so actions are not always self-contained, but there will still be a
+far lower degree of coupling and interdependency than in a monolithic
+parser.
 
-During development of the Postfix log parser it became apparent that in
-addition to the obvious variety in log~lines there were many complications
-to be overcome.  Some were the result of deficiencies in Postfix's logging,
-and some of those were rectified by later versions of Postfix, e.g.\
-identifying bounce notifications
+During development of \parsername{} it became apparent that in addition to
+the obvious variety in log~lines there were many complications to be
+overcome.  Some were the result of deficiencies in Postfix's logging, and
+some of those deficiencies were rectified by later versions of Postfix,
+e.g.\ identifying bounce notifications
 (\sectionref{identifying-bounce-notifications}); others were due to the
 vagaries of process scheduling, client behaviour, and administrative
-actions.  All were successfully accommodated in the Postfix log parser:
-adding new actions was enough to overcome several of the complications;
-others required modifications to a single existing action to work around a
+actions.  All were successfully accommodated in \parsername{}: adding new
+actions was enough to overcome several of the complications; others
+required modifications to a single existing action to work around a
 difficulty; the remainder were resolved by adapting existing actions to
@@
 behaviour as appropriate based on that extra data.  Every architecture
-should aim to make the hard things possible as well as making the easy
-things easy; the successful implementation of the Postfix log parser
-demonstrates that this architecture achieves that aim.
+should aim to make the easy things easy and the hard things possible; the
+successful implementation of \parsername{} demonstrates that this
+architecture achieves that aim.
 
@@
 several rules and actions to parse a single input, potentially simplifying
-both rules and actions.
-
-A simple example is to have one rule and action removing comments from
-input lines, so that other rules and actions do not have to handle comments
-at all; obviously if comment characters can be escaped or embedded in
-quoted strings the implementation needs to be more complex, but for some
-inputs this can greatly simplify subsequent parsing.
+both rules and actions.  A simple example is to have one rule and action
+removing comments from input lines, so that other rules and actions do not
+have to handle comments at all; obviously if comment characters can be
+escaped or embedded in quoted strings the implementation needs to be more
+complex, but for some inputs this can greatly simplify subsequent parsing.
 
@@
 A more complicated example is determining the atomic weight of a compound
-given its chemical formula.  A chemical formula is a sequence of one or two
-character chemical element abbreviations, each optionally followed by a
-number indicating how many of that element are present in the formula (if
-omitted the number is assumed to be one).  Some parts of a formula may be
-repeated: they are enclosed in parentheses, and the closing parenthesis is
-followed by a number indicating how many are present in the formula.  A
-monomer is a simple compound, i.e.\ it does not have any repeated portions;
-compounds with repeated portions are called polymers.  E.g.\ glucose (a
-monomer) has the chemical formula $C_{6}H_{12}O_{6}$ and hydrocarbon (a
-polymer) is $CH_{3}{(CH_{2})}_{50}CH_{3}$.  The atomic weight of an
-individual element is given in periodic tables, but one element in
-particular will be important in the later explanation: Hydrogen,
-abbreviated H, has an atomic weight of 1.  The atomic weight of a given
-compound is the sum of the atomic weights of each of the elements
-multiplied by the number of times they occur; this is simple to calculate
-for a monomer, but it is harder for a polymer because the atomic weight of
-the repeated monomer must be multiplied by the number of times the monomer
-is present in the compound (e.g.\ for hydrocarbon the atomic weight of
-$CH_{2}$ must be multiplied by $50$).
+given its chemical formula.  A chemical formula is a sequence of chemical
+element abbreviations, each optionally followed by a number indicating how
+many of that element are present in the formula (if omitted the number is
+assumed to be one).  Some parts of a formula may be repeated: they are
+enclosed in parentheses, and the closing parenthesis is followed by a
+number indicating how many are present in the formula.  A monomer is a
+simple compound, i.e.\ it does not have any repeated portions; compounds
+with repeated portions are called polymers.  E.g.\ glucose, a monomer, has
+the chemical formula $C_{6}H_{12}O_{6}$, and hydrocarbon, a polymer, is
+$CH_{3}{(CH_{2})}_{50}CH_{3}$.  The atomic weight of an individual element
+is given in periodic tables, but one element in particular will be
+important in the later explanation: Hydrogen, abbreviated H, has an atomic
+weight of 1.  The atomic weight of a given compound is the sum of the
+atomic weights of each of its elements multiplied by the number of times
+they occur; this is simple to calculate for a monomer, but it is harder for
+a polymer because the atomic weight of the repeated monomer must be
+multiplied by the number of times the monomer is present in the compound
+(e.g.\ for hydrocarbon the atomic weight of $CH_{2}$ must be multiplied by
+$50$).
 
 Calculating the atomic weight of a monomer requires a single rule and
-action: the action separates each element in the monomer, looks up its
+action.  The action separates each element in the monomer, looks up its
 atomic weight, multiplies it by the number of times it is present (one if
 unspecified), and adds up the results.  The rule's regex is
-\verb!^([A-Z]([a-z])?\d*)+$!, which matches:
-
-\begin{enumerate}
-
-    \squeezeitems{}
-
-    \item The start of the input.
-
-    \item One or more of: an uppercase character, optionally followed by a
-        lowercase character, followed by zero or more digits.
-
-    \item The end of the input.
-
-\end{enumerate}
-
-The regex does not guarantee that the chemical element abbreviations in the
-formula are correct; the regex could be modified to check that by replacing
-\verb![A-Z]([a-z])?! with all the valid abbreviations, slightly simplifying
-the action at the cost of complicating the regex (which could be
-automatically generated).
+\verb!^([A-Z]([a-z])?\d*)+$!, which does not guarantee that the chemical
+element abbreviations in the formula are correct; it could be modified to
+check that by replacing \verb![A-Z]([a-z])?! with all the valid
+abbreviations, slightly simplifying the action at the cost of complicating
+the regex, which could be automatically generated.
 
 Parsing a polymer is more difficult because of the repeated monomers: each
-monomer's atomic weight needs to be calculated, multiplied by the number of
-times it is present in the compound, and that figure added to the sum.
-Without cascaded parsing the action would need to maintain state while
-recursively parsing the monomers embedded in the polymer.  With cascaded
-parsing available three rules and actions could be used as shown in
+monomer's atomic weight needs to be calculated separately, multiplied by
+the number of times it is present in the compound, and that figure added to
+the sum.  Without cascaded parsing the action would need to maintain state
+while recursively parsing the monomers embedded in the polymer.  Using
+cascaded parsing, three rules and actions can be used instead as shown in
 \tableref{Regexes and Actions to calculate the atomic weight of a chemical
-compound}:
+compound}.
 
@@
 Rules are responsible for categorising inputs: each rule should recognise
-one and only input variant.  An input category with multiple input variants
-should have multiple rules, one for each variant.  Rules will typically use
-a regex when classifying inputs, but other approaches may prove useful for
-some applications, e.g.\ comparing fixed strings to the input, or checking
-the length of the input; for the remainder of this thesis it will be
-assumed that a regex is used.  Each rule must specify at least the regex to
-recognise inputs and the action to invoke when classification is
-successful, but implementations are free to add any other attributes they
-require; \sectionref{rule attributes} describes the attributes used in
-\parsername{}, and some potential attributes will be discussed later in
-this section.
+one and only one input variant; an input category with multiple input
+variants should have multiple rules, one for each variant.  Rules will
+typically use a regex when recognising inputs, but other approaches may
+prove useful for some applications, e.g.\ comparing fixed strings to the
+input, or checking the length of the input; for the remainder of this
+thesis it will be assumed that a regex is used.  Each rule must specify at
+least the regex to recognise inputs and the action to invoke when
+recognition is successful, but implementations are free to add any other
+attributes they require; \sectionref{rule attributes} describes the
+attributes used in \parsername{}, and some generally useful attributes will
+be discussed later in this section.
 
@@
 the line oriented nature of Postfix log files; more flexible approaches
-could be pursued.
-
-Classifying inputs using the rules is simple: the first rule to recognise
-the input determines the action that will be invoked; there is no
-backtracking to try alternate rules; no attempt is made to pick a
-\textit{best\/} rule.  \sectionref{attaching conditions to rules} contains
-an example which requires that the rules are used in a specific order to
-correctly parse the input, so a mechanism is needed to allow the author of
-the rules to specify that ordering.  The framework should be written so
-that it sorts the rules according to a priority attribute specified by the
-rule author, allowing fine-grained control over the order that rules are
-used in.  The priority attribute may be implemented as a number, or as a
-range of values, e.g.\ low, medium, and high, or in a different fashion
-entirely if it suits the implementation.  Rule ordering for efficiency is a
-separate topic that is covered in \sectionref{rule ordering for
-efficiency}; overlapping rules are discussed in \sectionref{overlapping
-rules in architecture}.
+could be pursued.  Classifying inputs using the rules is simple: the first
+rule to recognise the input determines the action that will be invoked;
+there is no backtracking to try alternate rules; no attempt is made to pick
+a \textit{best\/} rule.  \sectionref{attaching conditions to rules}
+contains an example which requires that the rules are used in a specific
+order to correctly parse the input, so a mechanism is needed to allow the
+author of the rules to specify that ordering.  The framework should sort
+the rules according to a priority attribute specified by the rule author,
+allowing fine-grained control over the order that rules are used in.  The
+priority attribute may be implemented as a number, or as a range of values,
+e.g.\ low, medium, and high, or in a different fashion entirely if it suits
+the implementation.  Rule ordering for efficiency is a separate topic that
+is covered in \sectionref{rule ordering for efficiency}; overlapping rules
+are discussed in \sectionref{overlapping rules in architecture}.
 
@@
 
-The framework issues a warning for each unparsed input, so it is evident
-when the ruleset needs to be augmented.  Parsing new inputs is achieved in
-one of three ways:
+The framework issues a warning for each unparsed input, so it is clearly
+evident when the ruleset needs to be augmented.  Parsing new inputs is
+achieved in one of three ways:
 
@@
 
-    \item Write a new rule that pairs an existing action with a new regex;
-        this adds a new variant to an existing category.
+    \item Write a new rule that pairs an existing action with a new regex,
+        adding a new variant to an existing category.
 
     \item Create a new category of inputs and a new action to process
-        inputs from the new category; write a new rule pairing the new
+        inputs from the new category, and write a new rule pairing the new
         action with a new regex.
@@
 created that parses calls to that subroutine, checking that the arguments
-used agree with the subroutine's signature.
-
-These avenues of research and development have not been pursued by the
-author, but the architecture allows them to easily be undertaken
-independently.
+used agree with the subroutine's signature.  These avenues of research and
+development have not been pursued by the author, but the architecture
+allows them to easily be undertaken independently.
 

### 2009-03-05 doc/logparser-implementation.tex
Corrections to the first half of the architecture chapter.

@@
         be familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn (a good
+        readily available resources from which to learn: a good
         introduction to \acronym{SQL} can be found at

### 2009-03-05 doc/logparser-architecture.tex
Corrections to second half of architecture chapter.

@@
 
-Rules can specify conditions that will be evaluated by the framework before
-attempting to use a rule to recognise an input: if the condition evaluates
-to true the rule will be used, if not the rule will be skipped.  Conditions
+Rules can have conditions attached that will be evaluated by the framework
+before attempting to use a rule to recognise an input: if the condition is
+true the rule will be used, if not the rule will be skipped.  Conditions
 can be as simple or complex as the parser requires, though naturally as the
-complexity rises so does the difficulty in understanding how everything
+complexity rises so too does the difficulty in understanding how everything
 interacts.  The framework has to evaluate each condition, so as the
-complexity of conditions increases so does the complexity of the code
-required to evaluate them; beyond a certain level of complexity conditions
-should probably be written in a proper programming language, e.g.\ taking
-advantage of dynamic languages' support for evaluating code at run-time, or
-embedding a language like Lua.  If an implementation is going to use
-conditions so complex that they will require a full programming language to
-implement them, the design may need to be re-thought, including the choice
-to use this architecture --- there may be other architectures more
-suitable.
+complexity of conditions increases so will the complexity of the code
+required to evaluate them.  Beyond a certain level of complexity,
+conditions should probably be written in a proper programming language,
+e.g.\ taking advantage of dynamic languages' support for evaluating code at
+run-time, or embedding a language like Lua.  If an implementation is going
+to use conditions so complex that they will require a Turing-equivilent
+programming language to implement them, the design may need to be
+re-thought, including the decision to use this architecture --- there may
+be other architectures more suitable.
 
 Conditions that examine the input will be the easiest to understand,
-because they can be understood in isolation --- they do not depend on
+because they can be understood in isolation; they do not depend on
 variables set by actions or other rules.  Conditions that examine the input
-may be complex, but simple conditions can be quite useful too, e.g.\ every
-Postfix log \empty{}line contains the name of the Postfix component that
-produced it, so every rule used in \parsername{} specifies the component
-whose log \empty{}lines it recognises as a condition, reducing the number
-of rules that will be used when classifying an input (see \sectionref{rules
-in implementation} for details).
+can be complex if required, but simple conditions can be quite useful too,
+e.g.\ every Postfix log \empty{}line contains the name of the Postfix
+component that produced it, so every rule used in \parsername{} has a
+condition specifying the component whose log \empty{}lines it recognises,
+reducing the number of rules that will be used when classifying an input
+(see \sectionref{rules in implementation} for details) and increasing the
+chance that the input will be correctly recognised..
+
+Conditions can also check the value of variables that have been set by
+either actions or rules; it is easier to understand how a variable's value
+will be used and changed if it is set by rules only rather than by actions,
+because the chain of checking and setting can be followed from rule to
+rule.  The downside to actions setting variables used in rule conditions is
+action at a distance: understanding when a rule's condition will be true or
+false requires understanding not just every other rule but also every
+action.  The framework probably does not need to support a high level of
+complexity and flexibility when rules are setting variables; however, if
+the framework supports complex conditions, that code can probably be easily
+extended to support complex variable assignments too.  The level of
+complexity the framework supports when evaluating conditions and setting
+variables has two costs that must be taken into account when designing a
+parser: the difficulty of implementation, and the difficulty of
+understanding rules and writing correct rules.  XXX NEEDS TO BE IMPROVED\@.
 
@@
 the comment start token even if it is within a quoted string, which is
-incorrect.  Rules one and two will only be used when the parser is parsing
-code, not comments: rule one recognises a comment that is contained within
-one line and leaves the parser's state unchanged; rule two recognises the
-start of a command and changes the parser's state to parsing comments.
-Rules three and four will be used when the parser is parsing a comment:
-rule three recognises the end of a comment and switches parsing state back
+incorrect.  Rules one and two will be used when the parser is parsing code,
+not comments: rule one recognises a comment that is contained within one
+line and leaves the parser's state unchanged; rule two recognises the start
+of a comment and changes the parser's state to parsing comments.  Rules
+three and four will be used when the parser is parsing a comment: rule
+three recognises the end of a comment and switches the parser's state back
 to parsing code; rule four recognises a comment line without an end token
@@
         \tabletopline{}%
-        No.   & Regex             & Condition         & New State         \\
+        No.   & Regex             & Condition           & Variable Changes    \\
         \tablemiddleline{}%
-        1     & \verb!/\*.*?\*/!  & state == code     & state = code      \\
-        2     & \verb!/\*.*!      & state == code     & state = comment   \\
-        3     & \verb!.*\*/!      & state == comment  & state = code      \\
-        4     & \verb!.*!         & state == comment  & state = comment   \\
+        % XXX WHAT SHOULD THE VARIABLE BE CALLED?
+        1     & \verb!/\*.*?\*/!  & parsing == code     & parsing = code      \\
+        2     & \verb!/\*.*!      & parsing == code     & parsing = comment   \\
+        3     & \verb!.*\*/!      & parsing == comment  & parsing = code      \\
+        4     & \verb!.*!         & parsing == comment  & parsing = comment   \\
         \tablebottomline{}%
@@
 \end{table}
-
-Conditions can also check the value of variables that have been set by
-either actions or rules; it will be easier to understand how a variable's
-value will be used and changed if it is set by rules only rather than by
-actions, because the chain of checking and setting can be followed from
-rule to rule.  The downside to actions setting variables used in rule
-conditions is action at a distance: to understand when a rule's condition
-will be true requires understanding not just every rule but also every
-action.  The framework probably does not need to support the same level of
-complexity and flexibility when setting variables from rules (actions can
-obviously use complex code if they are setting variables); however, if the
-framework allows complex conditions, that code can probably be easily
-extended to support complex variable assignments too.
 
@@
 overlap with one or more existing rules, i.e.\ some inputs could be parsed
-by more than one rule.  The architecture does not try to detect overlapping
-rules: that responsibility is left to the author of the rules.
-Unintentionally overlapping rules lead to inconsistent parsing and data
-extraction because the first matching rule wins, and the order in which
-rules are tried against each input might change between parser invocations.
-Overlapping rules are frequently a requirement, allowing a more specific
-rule to match some inputs and a more general rule to match the remainder,
-e.g.\ separating \acronym{SMTP} delivery to specific sites from
-\acronym{SMTP} delivery to the rest of the world.  Allowing overlapping
-rules simplifies both the general rule and the more specific rule;
-additionally rules from different sources can be combined with a minimum of
-prior cooperation or modification required.  Overlapping rules should have
-a priority attribute to specify their relative ordering; negative
-priorities may be useful for catchall rules.
+by more than one rule.  Unintentionally overlapping rules lead to
+inconsistent parsing and data extraction because the first matching rule
+wins, and the order in which rules are tried against each input might
+change between parser invocations.  Overlapping rules are frequently a
+requirement, allowing a more specific rule to match some inputs and a more
+general rule to match the remainder, e.g.\ separating \acronym{SMTP}
+delivery to specific sites from \acronym{SMTP} delivery to the rest of the
+world.  Allowing overlapping rules simplifies both the general rule and the
+more specific rule; additionally rules from different sources can be
+combined with a minimum of prior cooperation or modification required.
+Overlapping rules should have a priority attribute to specify their
+relative ordering; negative priorities may be useful for catchall rules.
+The architecture does not try to detect overlapping rules: that
+responsibility is left to the author of the rules.
 
-Decoupling the rules from the actions allows external tools to be written
-to detect overlapping rules.  Traditional regexes are equivalent in
-computational power to \acronym{FA} and can be converted to \acronym{FA},
-so regex overlap can be detected by finding a non-empty intersection of two
-\acronym{FA}\@.  The standard equation for \acronym{FA} intersection (given
-for example in~\cite{intersection-of-NFA-using-Z}) uses DeMorgan's law:
-$FA1~\cap{}~FA2~=~\overline{(\overline{FA1}~\cup{}~\overline{FA2})}$, which
-has considerable computation complexity.  XXX COMMENT ON THE COMPUTATIONAL
-COMPLEXITY OF INVERSION AND UNION\@; ADD A REFERENCE TO THE SPINNING WHEEL
-BOOK\@.  Perl 5.10 regexes~\cite{perlre} are more powerful than traditional
-regexes: it is possible to match correctly balanced brackets nested to an
-arbitrary depth, e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!
-matches \verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires
-the regex engine to maintain state on a stack, so Perl 5.10 regexes are
+Overlapping rules can be detected by visual inspection, or a program may be
+written to analyse the rule's regexes.  Traditional regexes are equivalent
+in computational power to \acronym{FA} and can be converted to
+\acronym{FA}, so regex overlap can be detected by finding a non-empty
+intersection of two \acronym{FA}\@.  Perl 5.10 regexes~\cite{perlre} are
+more powerful than traditional regexes: it is possible to match correctly
+balanced brackets nested to an arbitrary depth, e.g.\
+\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
+\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
+regex engine to maintain state on a stack, so Perl 5.10 regexes are
 equivalent in computational power to \acronym{PDA}; detecting overlap may
 require calculating the intersection of two \acronym{PDA} instead of two
-\acronym{FA}.  The intersection of two \acronym{PDA} is not closed, i.e.\
-the result cannot always be implemented using a \acronym{PDA}, so
-intersection may be intractable in some cases, e.g.:
+\acronym{FA}.  \acronym{PDA} intersection is not closed, i.e.\ the result
+cannot always be implemented using a \acronym{PDA}, so intersection may be
+intractable in some cases, e.g.:
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
@@
 complexity of calculating \acronym{FA} or \acronym{PDA} intersection.  This
-is certainly not a task to be performed every time the parser runs: naive
+is certainly not a task to be performed every time a parser runs: naive
 detection of overlap amongst \parsernames{} \numberOFrules{} rules would
-require calculating \numberOFruleINTERSECTIONS{} intersections.  When
-detecting overlap any conditions attached to rules must be taken into
+require calculating \numberOFruleINTERSECTIONS{} intersections.
+
+When detecting overlap any conditions attached to rules must be taken into
 account, because two rules whose regexes overlap may have conditions
-attached which prevent the overlap in practice.  A less naive approach to
+attached which prevent the rules overlapping.  A less naive approach to
 overlap detection would first check for overlapping conditions amongst
-rules, and then check for overlapping regexes amongst the rules in each set
-of rules with overlapping conditions.  Depending on the conditions employed
-a rule may fall into more than one group, e.g.\ given these three
-conditions:
+rules, and then check for overlap between the regexes of each pair of rules
+with overlapping conditions.  Depending on the conditions employed a rule
+may overlap with multiple rules, and rule overlap is not transitive, e.g.\
+given these three conditions:
 
@@
 
+XXX CARL SUGGESTS COMPARING THIS TO THE POWERSET\@; IS THAT REASONABLE\@?
+
 \noindent{}The first and second conditions do not overlap, but the third
-condition does overlap with both the first and second conditions.  If the
-rules are divided into sets based on how their conditions overlap, the
-complexity of detecting overlap amongst $n$ rules is:
+condition overlaps with both the first and second conditions.  When rules
+are divided into sets based on how their conditions overlap, the complexity
+of detecting overlap amongst $n$ rules is:
 
-$$O(n^{2}y)~+~\sum{O(|s|^{2}x)~\forall{}~s~\in{}~overlaps}$$
+XXX IN THE FORMULA BELOW $s \in{} overlaps$ SHOULD BE UNDER $\sum$
+
+$$O(n^{2}y) + O\left(\sum{|s|^{2}x \forall{} s \in{} overlaps}\right)$$
 
@@
 
+            $x$ & = & cost of checking for overlap between two regexes    \\
             $y$ & = & cost of checking for overlap between two conditions \\
-            $x$ & = & cost of checking for overlap between two regexes    \\
      $overlaps$ & = & set of sets of rules with overlapping conditions    \\
@@
 
+For this approach to be more efficient than the naive approach $y$ must be
+significantly lower than $x$.  If $y$ is higher than $x$ then the opposite
+approach should be taken: check for regex overlap first, then check
+for condition overlap only between rules that have overlapping regexes.
+
 Once conditions pass a certain level of complexity determining whether two
@@
 knowledge of other rules and possibly even actions.  E.g.\ given two rules
-with conditions \verb!verbose == true! and \verb!silent == true!, it
-appears that these rules cannot overlap, yet there is nothing to stop both
-variables being set to true by one or more other rules or actions.  If
-variables used in conditions can be set by actions, determining whether two
-conditions overlap is impossible, because understanding how an action will
-change the variables is impossible; the Church-Turing Thesis XXX NEED A BIB
-ENTRY FOR CHURCH-TURING explains why.
+with conditions \verb!verbose == true! and \verb!silent == true!, logically
+these rules should not overlap, yet there is nothing to stop both variables
+being set to true by one or more rules or actions.  If variables used in
+conditions can be set by actions, determining whether two conditions
+overlap is impossible, because understanding how an action will change the
+variables is impossible.  XXX DO I NEED TO MENTION CHURCH-TURING OR
+ANYTHING\@?
 
@@
 because they may prevent the correct rule from recognising the input.  If
-an adaptive ordering system is used (see \sectionref{rule ordering for
-efficiency}) then a rule with a regex that matches too much may be promoted
-up through the list, displacing an increasing number of correct rules as it
+an adaptive ordering system is used that prioritises rules that frequently
+recognise inputs (see \sectionref{rule ordering for efficiency}), then a
+rule with a regex that matches inputs it should not may be promoted up
+through the list, displacing an increasing number of correct rules as it
 moves.
@@
 5.10, regardless of the input length, because the regex engine looks for
-the required double quotes first.  Similarly Perl 5.10's regex engine
-optimises alterations starting with literal text into a trie, which has
-constant matching time, whereas the matching time for alteration is
-proportional to the number of alternatives.  Perl 5.10 regexes can use
-\verb!(?>pattern)!, which matches \verb!pattern! the first time the regex
-engine passes over it, and does not change what it originally matched if
-the regex engine backtracks over it, so troublesome regexes can be modified
-to use this to reduce the impact of backtracking; Prolog users will notice
-a similarity to the \verb'!' (cut) operator.  A presentation showing some
-of Perl 5.10's regex features is available at
+both of the required double quotes first.  Similarly Perl 5.10's regex
+engine optimises alterations starting with literal text into a trie, which
+has matching time proportional to the length of the alternatives, rather
+than the number of alternatives.  Perl regexes can use \verb!(?>pattern)!,
+which matches \verb!pattern! the first time the regex engine passes over
+it, and does not change what it originally matched if the regex engine
+backtracks over it, so troublesome regexes can use this to reduce the
+impact of backtracking; Prolog users will notice a similarity to the
+\verb'!' (cut) operator.  A presentation showing some of Perl 5.10's new
+regex features is available at
 \urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.
@@
 whether a more complex condition's behaviour is pathological or simply a
-bug is difficult and to some extent is a matter of opinion.  The author has
-not experimented with more complex conditions, so this thesis will not
-address this point any further.
+bug is difficult and to some extent is a matter of opinion.  When this
+architecture has received more widespread usage consensus should be reached
+on the topic of pathological conditions.
 

### 2009-03-05 doc/logparser-architecture.tex
Corrected the formulae in the overlapping rules section, and corrected the formatting.

@@
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
-Detecting overlap amongst $n$ regexes requires calculating $n(n-1)/2$
-intersections, resulting in $O(n^2x)$ complexity, where $O(x)$ is the
-complexity of calculating \acronym{FA} or \acronym{PDA} intersection.  This
-is certainly not a task to be performed every time a parser runs: naive
-detection of overlap amongst \parsernames{} \numberOFrules{} rules would
-require calculating \numberOFruleINTERSECTIONS{} intersections.
+Detecting overlap amongst $n$ regexes requires calculating
+$\frac{n\left(n-1\right)}{2}$ intersections, resulting in
+$O\left(n^{2}x\right)$ complexity, where $x$ is the cost of calculating
+\acronym{FA} or \acronym{PDA} intersection.  This is certainly not a task
+to be performed every time a parser runs: naive detection of overlap
+amongst \parsernames{} \numberOFrules{} rules would require calculating
+\numberOFruleINTERSECTIONS{} intersections.
 
@@
 
-XXX CARL SUGGESTS COMPARING THIS TO THE POWERSET\@; IS THAT REASONABLE\@?
-
 \noindent{}The first and second conditions do not overlap, but the third
 condition overlaps with both the first and second conditions.  When rules
-are divided into sets based on how their conditions overlap, the complexity
-of detecting overlap amongst $n$ rules is:
-
-XXX IN THE FORMULA BELOW $s \in{} overlaps$ SHOULD BE UNDER $\sum$
-
-$$O(n^{2}y) + O\left(\sum{|s|^{2}x \forall{} s \in{} overlaps}\right)$$
-
-where:
+are paired based on how their conditions overlap, the complexity of
+detecting overlap amongst $n$ rules is $O\left(n^{2}y+|o|x\right)$, where:
 
@@
     $y$ & = & cost of checking for overlap between two conditions \\
-     $overlaps$ & = & set of sets of rules with overlapping conditions    \\
+    $o$ & = & set of pairs of rules with overlapping conditions   \\
 
@@
 approach should be taken: check for regex overlap first, then check
-for condition overlap only between rules that have overlapping regexes.
+for condition overlap only between pairs of rules with overlapping regexes.
 

### 2009-03-06 doc/logparser-architecture.tex
Wrote architecture conclusion.

@@
 complex, but for some inputs this can greatly simplify subsequent parsing.
+Actions do not need to be specially registered with the framework or be
+declared in a particular way to use cascaded parsing: actions that do not
+use cascaded parsing will return nothing, those that do will simply return
+a string to be parsed.
 
@@
 
-XXX TO BE WRITTEN\@.
+This chapter has presented the parser architecture developed for this
+project.  It started with a high level view of the architecture, describing
+how it achieves its design aim of being easily extensible for users, and
+the advantages that being easily extensible brings to parser authors too.
+The three main components of the architecture were documented in detail,
+explaining each component's responsibilities and the functionality it
+provides, plus any difficulties associated with the components.  The
+framework, unsurprisingly, was quite quick to document, because it is the
+simplest of the three components.  The actions are also quite easy to
+understand, because the architecture does not impose any structure or
+requirements upon them: parser authors are free to do anything they want
+within an action.  The architecture's support for cascaded parsing was
+described in the actions section, with an example to illustrate how it can
+be useful.  The rules section was the longest, because although the rules
+appear conceptually simple --- recognise an input and specify the action
+that handles that input --- they have subtle behaviour that needs to be
+clearly explained.  When extending a ruleset a decision needs to be taken
+about whether the input should be recognised by extending an existing rule,
+by adding a new rule to an existing rule category, or by adding a new rule
+category, rule, and action.  Rules can have conditions attached to them,
+restricting the set of rules used to recognise an input; the complexity of
+the conditions used greatly influences the difficulty of writing a correct
+ruleset or understanding and extending an existing ruleset.  Overlapping
+rules are frequently a requirement in a parser, and their use can greatly
+simplify some rules, but they can be a source of bugs because they can
+recognise inputs unexpectedly.  The framework does not try to detect
+overlapping rules, because overlap amongst rules may be valid and is quite
+often intentional; that responsibility falls to the author of the rules.
+The difficulty involved in detecting overlap is proportional to the
+complexity of a ruleset's regexes and conditions, and may be possible for a
+human yet intractable or impossible for a program.  The rules section
+finishes with a discussion of pathological rules, concentrating on
+pathological regexes.

### 2009-03-06 doc/logparser-acronyms.tex
Dealt with some XXX items.

@@
     thesis: the original protocol definition is in RFC~821~\cite{RFC821},
-    updated in RFC~2821~\cite{RFC2821}; introductory guides can be found at
+    updated in RFC~2821~\cite{RFC2821}; an introductory guide can be found
+    at
     \urlLastChecked{http://db.glug-bom.org/lug-authors/philip/docs/mail-stuff/smtp-intro.html}{2009/02/28}

### 2009-03-06 doc/logparser-architecture.tex
Dealt with some XXX items.

@@
 data in one form and transforms it to another form; \parsername{}
-transforms log files to a database.  XXX NEED A REFERENCE FOR
-TRANSDUCTION\@.
+transforms log files to a database.
 
@@
 parser: the difficulty of implementation, and the difficulty of
-understanding rules and writing correct rules.  XXX NEEDS TO BE IMPROVED\@.
+understanding writing correct rules; the flexibility provided by complex
+conditions may be outweighed by the difficulty in understanding the
+interactions between rules that use complex conditions.
 
@@
         \tablemiddleline{}%
-        % XXX WHAT SHOULD THE VARIABLE BE CALLED?
-        1     & \verb!/\*.*?\*/!  & parsing == code     & parsing = code      \\
-        2     & \verb!/\*.*!      & parsing == code     & parsing = comment   \\
-        3     & \verb!.*\*/!      & parsing == comment  & parsing = code      \\
-        4     & \verb!.*!         & parsing == comment  & parsing = comment   \\
+        1     & \verb!/\*.*?\*/!  & expecting == code     & expecting = code      \\
+        2     & \verb!/\*.*!      & expecting == code     & expecting = comment   \\
+        3     & \verb!.*\*/!      & expecting == comment  & expecting = code      \\
+        4     & \verb!.*!         & expecting == comment  & expecting = comment   \\
         \tablebottomline{}%
@@
 \label{overlapping rules in architecture}
-
-XXX IS THERE ANY PREVIOUS RESEARCH IN THIS AREA\@?
 
@@
 being set to true by one or more rules or actions.  If variables used in
-conditions can be set by actions, determining whether two conditions
-overlap is impossible, because understanding how an action will change the
-variables is impossible.  XXX DO I NEED TO MENTION CHURCH-TURING OR
-ANYTHING\@?
+conditions can be set by actions, determining whether or not two conditions
+overlap is impossible: the Halting Problem show that it is impossible for
+one program to answer questions about another program's behaviour if the
+inquiring program is implemented on a compuational machine of the equal or
+lesser power to the program being inquired into.
 

### 2009-03-06 doc/logparser.tex
Corrections to second half of architecture chapter.
Dealt with some XXX items.

@@
 }
-% XXX CAN I MAKE THIS LOOK BETTER?  MAYBE USE [] LIKE \pageref{} DOES?
-\newcommand{\urlLastCheckedNoParens}[3][ ]{%
-    \url{#2}#1last checked #3%
-}
 
@@
 \newcommand{\numberOFruleINTERSECTIONS}[0]{%
-    17020%
+    17,020%
 }
@@
 
-XXX ASK CARL ABOUT\@:
-
-\begin{itemize}
-
-    \item XXX WHERE SHOULD I USE URLS AND WHERE SHOULD THEY BE BIBLIOGRAPHY
-        ENTRIES\@?
-
-    \item XXX REFERENCE FOR TRANSDUCTION\@.
-
-    \item XXX PREVIOUS RESEARCH IN OVERLAPPING RULES\@?
-
-    \item XXX BORROW SPINNING WHEEL BOOK\@; NEED TO COMMENT ON THE
-        COMPUTATIONAL COMPLEXITY OF FA UNION AND INVERSION\@; MAYBE COMMENT
-        ON COMPUTATIONAL COMPLEXITY OF PDA UNION, INVERSION, AND
-        INTERSECTION\@?
-
-    \item XXX BIB ENTRY FOR CHURCH-TURING\@.
-
-\end{itemize}
-
 \newpage

### 2009-03-09 doc/logparser-acronyms.tex
Corrections based on comments from Carl, plus a few I noticed in the process.

@@
     description={
-        See Context Free Language.
-    }
-}
-
-\newglossaryentry{CFL}{
-    type=\acronymtype,
-    name={CFL},
-    description={Context Free Language},
-    first={Context Free Language (CFL)}
-}
-\newglossaryentry{CFL glossary}{
-    name={Context Free Language},
-    description={
-        A Context Free Language is a language generated or parsed by a
-        \acronym{CFG}.  Every transition in a CFG takes the form
-        $L\rightarrow{}R$, where $L$ is a single non-terminal symbol, and
-        $R$ is a (possibly empty) string of terminal and/or non-terminal
-        symbols.  Every CFG can be recognised by a \acronym{PDA}, and
-        \acronyms{PDA} cannot recognise more complicated grammars, so
-        \acronyms{CFL}, \acronyms{CFG}, and \acronyms{PDA} are equivalent.
+        A CFG is a grammar in which every transition or production takes
+        the form $L\rightarrow{}R$, where $L$ is a single non-terminal
+        symbol, and $R$ is a (possibly empty) string of terminal and/or
+        non-terminal symbols.  Every CFG can be recognised by a
+        \acronym{PDA}, and \acronyms{PDA} cannot recognise more complicated
+        grammars, so \acronyms{CFG} and \acronyms{PDA} are equivalent.
     }

### 2009-03-09 doc/logparser-background.tex
Corrections based on comments from Carl, plus a few I noticed in the process.

@@
 \begin{table}[ht]
-    \caption{Sample output from SQL query}
+    \caption{Sample output from SQL query showing percentage of delivery
+    attempts rejected by each restriction}
     \empty{}\label{Sample output from SQL query}
@@
     \end{tabular}
+
+    All percentages above are exclusive, i.e.\ the first restriction's
+    number of rejections does not include the second restriction's number
+    of rejections.
+
 \end{table}
@@
 returned by the server and carries on regardless.  Many spam and virus
-sending programs have serious deficiencies; unfortunately other programs,
+sending programs have serious deficiencies; unfortunately, other programs,
 particularly newer programs, were written by competent programmers or send

### 2009-03-09 doc/logparser-introduction.tex
Corrections based on comments from Carl, plus a few I noticed in the process.

@@
 generated some graphs to clearly illustrate these statistics to management
-or users.\footnote{This was the author's first real foray into processing
-Postfix log files.}  Very few will have performed in-depth parsing and
-analysis of their log files, where the parsing must correlate the log lines
-per-connection or per-queueid rather than processing log lines
-independently.  One of the barriers to this kind of processing is the
-unstructured nature of Postfix log files, where each log line was added on
-an ad hoc basis as a requirement was discovered or new functionality was
-added.\footnote{A history of all changes made to Postfix is distributed
-with the source code, available from
+or users.\footnote{This was the first real foray the author, a Systems
+Administrator for a network of over 2,000 computers and over 1,800 users,
+took into processing Postfix log files.}  Very few will have performed
+in-depth parsing and analysis of their log files, where the parsing must
+correlate the log lines per-connection or per-queueid rather than
+processing log lines independently.  One of the barriers to this kind of
+processing is the unstructured nature of Postfix log files, where each log
+line was added on an ad hoc basis as a requirement was discovered or new
+functionality was added.\footnote{A history of all changes made to Postfix
+is distributed with the source code, available from
 \urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
@@
 
+Chapter~\ref{parser architecture} describes the parser architecture
+developed for this project, beginning with an overview of the architecture,
+then describing the components of the architecture in detail.
+
 XXX THIS ALL NEEDS TO BE CHECKED AND UPDATED\@.
 
+Chapter~\ref{Postfix Parser Implementation} discusses the parsing rules in
+detail, explaining the purpose and usage of each field in a rule, referring
+to an example rule and sample data it matches successfully against.  The
+pros and cons of overlapping rules are considered, including techniques for
+detecting unintentional overlaps.  Rule efficiency concerns are discussed,
+in particular the optimisations used by the algorithm.  The section
+concludes with a description of using the tools provided with the parser to
+generate new rules (specifically the regex in each rule) from unparsed log
+lines.  It contains the core of the paper, describing a naive parsing
+algorithm and the complications initially encountered that shaped the full
+algorithm.  A flow chart and a discussion of the emergent behaviour
+exhibited by the algorithm accompanies a comprehensive explanation of the
+different stages of the initial algorithm.  The framework that actions and
+rules fit into is documented, then the actions taken during execution of
+the algorithm are described, followed by the process of adding a new
+action.  The section concludes with an in-depth description of the further
+complications discovered, and their solutions that complete the parser.
 This algorithm requires a database for storing both the rules used when
@@
 stores the rules.
-
-Chapter~\ref{rules in architecture} discusses the parsing rules in detail,
-explaining the purpose and usage of each field in a rule, referring to an
-example rule and sample data it matches successfully against.  The pros and
-cons of overlapping rules are considered, including techniques for
-detecting unintentional overlaps.  Rule efficiency concerns are discussed,
-in particular the optimisations used by the algorithm.  The section
-concludes with a description of using the tools provided with the parser to
-generate new rules (specifically the regex in each rule) from unparsed log
-lines.
-
-Chapter~\ref{Postfix Parser Implementation} contains the core of the paper,
-describing a naive parsing algorithm and the complications initially
-encountered that shaped the full algorithm.  A flow chart and a discussion
-of the emergent behaviour exhibited by the algorithm accompanies a
-comprehensive explanation of the different stages of the initial algorithm.
-The framework that actions and rules fit into is documented, then the
-actions taken during execution of the algorithm are described, followed by
-the process of adding a new action.  The section concludes with an in-depth
-description of the further complications discovered, and their solutions
-that complete the parser.
 
@@
 
-Chapter~\ref{limitations and improvements in implementation} lists the
-limitations of the algorithm, then suggests some ways of dealing with them,
-with the goal of improving parsing and reproduction of the journey a mail
-takes through Postfix.
-
 Chapter~\ref{conclusion} contains the conclusion of the thesis, describing
@@
 resources used in developing the algorithm, writing the program, and
-preparing this thesis.  Also listed are some additional resources expected
-to be helpful in understanding \acronym{SMTP}, Postfix, anti-spam
-techniques, or the thesis.
+preparing this thesis.
 

### 2009-03-09 doc/logparser-other-parsers-review.tex
Some minor formatting fixes. Documented some of 'make check'. Added check for sentences with only one space separating them.
Removed URLs that are the first Google hit for a search term.
Corrections based on comments from Carl, plus a few I noticed in the process.

@@
 using the myriad of standard Unix text processing utilities already
-available.  It produces a \acronym{CSV} file and either a MySQL
-(\urlLastChecked{http://www.mysql.com/}{2008/11/23}) or Berkeley DB
-(\urlLastChecked{http://www.oracle.com/database/berkeley-db/index.html}{2008/11/23})
-database.  The decision to support both \acronym{CSV} and Berkeley DB
+available.  It produces a \acronym{CSV} file and either a MySQL or Berkeley
+DB database.  The decision to support both \acronym{CSV} and Berkeley DB
 appears to have been a serious limitation: both formats have limitations
@@
 \acronym{CSV} output, \acronym{LMA} stores the following data: time and
-date, client hostname and \acronym{IP} address, server \acronym{IP}
-address, sender and recipient addresses, \acronym{SMTP} code, and size (for
-accepted mails only).  It is unclear what time and date is stored: start
-time, end time, or delivery time?  Unlike \parsername{} it does not store
+date of the log line, client hostname and \acronym{IP} address, server
+\acronym{IP} address, sender and recipient addresses, \acronym{SMTP} code,
+and size (for accepted mails only).  Unlike \parsername{} it does not store
 the server hostname, HELO hostname, queueid, start and end times,
@@
 rejection in the input log file; it also missed some 40\% of delivered
-mail.  Once these glaring deficiencies were discovered the author did not
-waste any more time checking the results.
+mail.  Once these deficiencies were discovered the author did not spend any
+more time checking the results.
 
@@
 line 372.}\newline{}\texttt{Undefined subroutine \&main::EXTRACT\_FROM\_DB
-called at queryDB.pl line 379.}} after making corrections to the code it
+called at queryDB.pl line 379.}} after correcting the errors in the code it
 worked, extracting the \acronym{CSV} lines for the specified day where the
@@
 
-    \item DAILY-REJECT considers all \acronym{SMTP} codes except
-        \texttt{250} to be rejections; this includes invalid \acronym{SMTP}
-        codes such as \texttt{0} and \texttt{deferred}, so the awk command
-        does too.  These invalid \acronym{SMTP} codes are most likely
-        present because of incorrect parsing by \acronym{LMA}.
+    \item DAILY-REJECT considers all \acronym{SMTP} codes except ``250'' to
+        be rejections; this includes invalid \acronym{SMTP} codes such as
+        \texttt{0} and \texttt{deferred}, so the awk command does too.
+        These invalid \acronym{SMTP} codes are most likely present because
+        of incorrect parsing by \acronym{LMA}.
 
@@
 in-depth analysis.  It does not support saving any data, and it does not
-extract any data that it does not require to produce its report, e.g.\ it
+extract any data that it does not require to produce its report; e.g.\ it
 does not extract the HELO hostname, queueid, start and end times,
@@
     reports, all through a web interface.  Sawmill can perform Postfix
-    analysis on any platform, including Window, Linux, FreeBSD, OpenBSD,
-    Mac OS, Solaris, other UNIX, and more.
+    analysis on any platform, including Window~[sic], Linux, FreeBSD,
+    OpenBSD, Mac OS, Solaris, other UNIX, and more.
 }
@@
 
-The source code is available in an obfuscated form, and the product is
-quite expensive, requiring a \euros{100} + VAT licence per report
-(discounts are available when buying multiple licences); in contrast
+The source code is only available in an encrypted form, to support people
+who wish to use Sawmill on operating systems or machine architectures the
+company do not provide executables for.  Sawmill is quite expensive,
+requiring a \euros{100} + VAT licence per report, with discounts available
+when buying multiple licences (correct as of 2008/11/23); in contrast,
 \parsername{} is free to use and the code is freely available.  Sawmill is
@@
 searching is neither as simple to use nor as informative as the interface
-provided by Splunk.  The administrative interface is much easier to use
-than Splunk's: it took only five minutes to start parsing a whole directory
-of log files.
+provided by Splunk (see \sectionref{Splunk review}).  The administrative
+interface is much easier to use than Splunk's: it took only five minutes to
+start parsing a whole directory of log files.
 
@@
 
-In summary Sawmill suffers from being a general purpose product; it is
-probably much more useful when parsing log files where each log line is
+In summary, Sawmill suffers from supporting so many types of log files: it
+parses many types of log files, but none of them very well; it is probably
+much more useful when parsing log files where each log line is
 self-contained (e.g.\ web server log files), rather than log files
@@
 \section{Splunk}
+
+\label{Splunk review}
 
@@
 futile attempts, and reading all the available documentation, the author
-admitted defeat.  Using the \acronym{CLI} rather than the web interface was
-partially successful: the command \newline{} \tab{} \texttt{splunk find
-logs }\textit{log-directory\/}\newline{} added 40 of the
-\numberOFlogFILES{} log files to the queue for indexing.  Further attempts
-enqueued the same 40 log files, without explaining why the others were
-excluded.\footnote{The log files appear to have been indexed once only;
-presumably Splunk keeps track of the log files it has indexed and discards
-requests to index log files for a second time.  This may or may not be a
-useful feature for \parsername{}.} There did not appear to be an option to
-ensure the log files would be processed in the order they were created,
-though this may be neither necessary nor beneficial with Splunk.
-Subsequently the author was successful in adding a single file at a time
-using the \acronym{CLI}:\newline{} \tab{} \texttt{splunk add tail
-}\textit{filename\/}\newline{} A simple loop was then enough to add all the
-desired log files.  Splunk will periodically check all indexed log files
-for updates unless they are manually removed from its list; this may or may
-not be useful behaviour.  Splunk did not appear to have any difficulty in
-indexing the log files, once they had been successfully added to its queue.
+was still unable to index all the log files in a directory using the web
+interface.  Using the \acronym{CLI} rather than the web interface was
+partially successful: the command ``\texttt{splunk find logs
+}\textit{log-directory\/}'' added 40 of the \numberOFlogFILES{} log files
+to the queue for indexing.  Further attempts enqueued the same 40 log
+files, without explaining why the others were excluded.\footnote{The log
+files appear to have been indexed once only; presumably Splunk keeps track
+of the log files it has indexed and discards requests to index log files
+for a second time.  This may or may not be a useful feature for
+\parsername{}.} There did not appear to be an option to ensure the log
+files would be processed in the order they were created, though this may be
+neither necessary nor beneficial with Splunk.  Subsequently the author was
+successful in adding a single file at a time using the \acronym{CLI} and
+the command ``\texttt{splunk add tail }\textit{filename\/}''.  A simple
+loop using that command was enough to add all the desired log files.
+Splunk will periodically check all indexed log files for updates unless
+they are manually removed from its list; this may or may not be useful
+behaviour.  Splunk did not appear to have any difficulty in indexing the
+log files, once they had been successfully added to its queue.
 \parsername{} parses the logs it is instructed to parse, in the order
@@
 Copious documentation is made available on
-\urlLastChecked{http://www.splunk.com/}{2008/11/23}, but poor organisation
-and sheer abundance makes it extremely hard to find useful information.
-Searches confusingly tended to return results from old documentation rather
-than new.  In general the documentation appears to have been written by
-someone intimately acquainted with the software, who has difficulty
-understanding how a newcomer would approach tasks or the questions they
-would ask.
+\urlLastChecked{http://www.splunk.com/}{2008/11/23}, but the organisation
+and abundance of material does not support easy identification of useful
+information.  Searches confusingly tended to return results from old
+documentation rather than new.  In general the documentation appears to
+have been written by someone intimately acquainted with the software, who
+has difficulty understanding how a newcomer would approach tasks or the
+questions they would ask.
 
@@
 day to 500MB, though a trial Enterprise licence is available that allows
-indexing of up to 5GB of data per day.  In 2007 the cheapest licenced
+indexing of up to 5GB of data per day.  In 2007, the cheapest licenced
 version cost \$5000 plus \$1000 support, and limited the volume of data
@@
 those fields is provided in the documentation at
-\urlLastChecked{http://awstats.sourceforge.net/docs/awstats_faq.html\#MAIL}{2008/11/23}
-(in contrast \parsername{} fully documents everything it extracts), so the
-author could not understand what data is extracted, or determine what data
-is missing in comparison to \parsername{}.  AWStats coerces Postfix log
-files into Apache\footnote{The Apache web server is the most popular HTTP
-server in use over the past 10 years; more information is available at
-\urlLastChecked{http://httpd.apache.org/}{2008/11/23}.} format log files,
-for analysis by AWStats' HTTP log file parser.  The converting parser only
-deals with a small portion of the log lines generated by Postfix, silently
-skipping those it cannot deal with, and does not distinguish between
-different types of rejection; extending it to parse all log lines would be
-at least as much work as writing a new parser.  It does correlate log lines
-by queueid (not by pid), but it does not deal with any of the other
-complications described in \sectionref{complications}.  AWStats supports
-saving data, but the format of the data store is not documented.  It also
-supports reading compressed log files, but that functionality was not
-tested.
+\urlLastChecked{http://awstats.sourceforge.net/docs/awstats_faq.html\#MAIL}{2008/11/23},
+so the author could neither understand what data is extracted, nor
+determine what data is missing in comparison to \parsername{} (which fully
+documents everything it extracts).  AWStats coerces Postfix log files into
+the log file format used by the Apache web server, for analysis by AWStats'
+HTTP log file parser.  The converting parser only deals with a small
+portion of the log lines generated by Postfix, silently skipping those it
+cannot deal with, and does not distinguish between different types of
+rejection; extending it to parse all log lines would be at least as much
+work as writing a new parser.  It does correlate log lines by queueid (not
+by pid), but it does not deal with any of the other complications described
+in \sectionref{complications}.  AWStats supports saving data, but the
+format of the data store is not documented.  It also supports reading
+compressed log files, but that functionality was not tested.
 
@@
 the log lines, declared 1.93\% were corrupt, and ignored the remaining
-82.85\%.
+82.85\%.  The author did not verify the correctness of the parsing of those
+log lines AWStats did parse.
 
@@
 
-Anteater does not have any English documentation so it is impossible for
-this author to accurately comment on the analysis it performs.  It did not
-run successfully when tested, and its parsing would certainly be out of
-date because Postfix has evolved considerably since this tool was last
-updated (2003/11/06).  As it neither ran successfully nor has documentation
-the author can read, a detailed review cannot be provided.
+Anteater does not have any English documentation except for the quote above
+so it is impossible for this author to accurately comment on the analysis
+it performs.  It did not run successfully when tested, and its parsing
+would certainly be out of date because Postfix has evolved considerably
+since this tool was last updated (2003/11/06).  As it neither ran
+successfully nor has documentation the author can read, a detailed review
+cannot be provided.
 
-The Debian Linux distribution
-(\urlLastChecked{http://www.debian.org/}{2008/11/23}) provides a manual
-page with the copy of anteater it distributes, so the author was at least
-able to run anteater with the correct arguments; sadly anteater produced
-zero for every statistic, presumably because it was unsuccessful in parsing
-the log lines.
+The Debian Linux distribution provides a translated manual page with the
+copy of anteater it distributes, so the author was at least able to run
+anteater with the correct arguments; sadly anteater produced zero for every
+statistic, presumably because it was unsuccessful in parsing the log lines.
 
@@
 in HTML format, with all the parsing and report generation handled by
-plugins.  Using YAALA as a base for this project would be only slightly
-less work than starting from scratch, as both the input and output modules
-would need to be written specially; it might be more work to implement a
-parser within the constraints of YAALA\@.  YAALA supports storing
-previously gathered data using Perl's Storable module, so other Perl
-programs can use Storable to load, examine, and optionally modify the data;
-\parsername{} uses a well documented database that is accessible from most
-common programming languages.  This information was gleaned from the source
-code, as the documentation is sadly lacking.
+plugins.  Using YAALA as a base for this project would have been as much
+work as starting from scratch, as both the input and output modules would
+need to be written specially; it might be more work to implement a parser
+within the constraints of YAALA\@.  YAALA supports storing previously
+gathered data using Perl's Storable module, so other Perl programs can use
+Storable to load, examine, and optionally modify the data; \parsername{}
+uses a well documented database that is accessible from most common
+programming languages.  Information about how YAALA stores data was gleaned
+from the source code, as the documentation is sadly lacking.
 
-YAALA provides a Postfix parser that extracts the following types of fields
-from specific log lines:
+YAALA provides a Postfix parser that extracts the following of fields from
+specific log lines:
 
@@
 
-In summary YAALA provides a Postfix parser that tries to parse the most
+In summary, YAALA provides a Postfix parser that tries to parse the most
 common Postfix log lines only, provides reasonably flexible report
@@
 \parsername{} uses an \acronym{SQL} database to make accessing the
-extracted data as easy as possible.  In general Lire has poor
+extracted data as easy as possible.  In general, Lire has poor
 documentation.

### 2009-03-09 doc/logparser-architecture.tex
Corrections and changes after comments from Carl.

@@
 
-For each input the framework tries each rule in turn until it finds a rule
+For each input, the framework tries each rule in turn until it finds a rule
 that recognises the input, then invokes the action specified by that rule.
@@
 trivial in comparison to understanding an entire parser, identifying the
-correct location to change, and making the appropriate changes.  Bear in
-mind that changes to a parser must be made without adversely affecting
-existing parsing, including any edge cases that are not immediately obvious
---- \sectionref{yet-more-aborted-delivery-attempts} describes a
-complication that occurs only four times in \numberOFlogFILES{} log files.
-Requiring changes to a parser's source code also complicates upgrades of
-the parser, because the changes must be preserved during the upgrade, and
-they may clash with changes made by the developer.  This architecture
-allows the user to add new rules without having to edit a parser, unless
-the new inputs cannot be processed by the existing actions.  If the new
-inputs do require new functionality, new actions can be added to the parser
-without having to modify existing actions; only in the rare event that the
-new actions need to cooperate with existing actions will more extensive
-changes be required.
+correct location to change, and making the appropriate changes.  Additions
+to a parser must be made without adversely affecting existing parsing,
+including any edge cases that are not immediately obvious; an edge case
+that occurs only four times in \numberOFlogFILES{} log files is described
+in Yet More Aborted Delivery Attempts
+(\sectionref{yet-more-aborted-delivery-attempts}).  More intrusive changes
+are more likely to introduce a bug, so reducing the extent of the changes
+is important.  Requiring changes to a parser's source code also complicates
+upgrades of the parser, because the changes must be preserved during the
+upgrade, and they may clash with changes made by the developer.  This
+architecture allows the user to add new rules without having to edit a
+parser, unless the new inputs cannot be processed by the existing actions.
+If the new inputs do require new functionality, new actions can be added to
+the parser without having to modify existing actions; only in the rare
+event that the new actions need to cooperate with existing actions will
+more extensive changes be required.
 
@@
 
+    \item [Convert physical inputs to logical inputs] Each rule recognises
+        a single input at a time: there is no facility for rules to consume
+        more input or push unused input back onto the input stream,
+        although actions may use cascaded parsing (\sectionref{cascaded
+        parsing}) to effectively push input back onto the input stream.  A
+        physical input (i.e.\ a single line read from the input stream) may
+        contain multiple or partial logical inputs, and the framework must
+        pre-process these to provide a logical input for the rules to
+        recognise.  This simplifies the rules and actions considerably, at
+        the cost of added complexity in the framework; during the design
+        phase it was decided that it was easier to deal with the problem of
+        parsing multiple or partial inputs once, rather than dealing with
+        it in every rule and action.  This is easy in the case of Postfix
+        log files because there is a one-to-one mapping between physical
+        and logical inputs; mapping between physical and logical inputs may
+        be more difficult for other types of input.  Some input types may
+        require pre-processing equivalent to parsing the physical inputs;
+        in those cases the framework should behave similarly to other
+        architectures: present the entire input stream at once, and discard
+        the portion recognised by the successful rule.
+
     \item [Classify the input]  The pseudo-code above shows how rules are
@@
         it later, so that information is not lost between parsing runs.
+        \parsername{} uses this functionality because mails may take some
+        time to deliver and thus have their log lines split between log
+        files; a compiler might store data structures it builds as it
+        parses different files.
 
@@
 handle one input category, with each rule recognising one input variant.
-It is anticipated that parsers based on this architecture will have a high
-ratio of rules to actions, with the aim of having simple rules and tightly
-focused actions.  An action may need to process different input variants in
-slightly different ways, but large changes in processing indicate the need
-for a new action and a new category of input; if an action becomes overly
-complicated it starts to turn into a monolithic parser, with too much logic
-contained in a single procedure.
+It is expected that parsers based on this architecture will have many
+actions, and each action may be specified by many rules, with the aim of
+having simple rules and tightly focused actions.  An action may need to
+process different input variants in slightly different ways, but large
+changes in processing indicate the need for a new action and a new category
+of input; if an action becomes overly complicated it starts to turn into a
+monolithic parser, with too much logic contained in a single procedure.
 
@@
 complex, but for some inputs this can greatly simplify subsequent parsing.
-Actions do not need to be specially registered with the framework or be
-declared in a particular way to use cascaded parsing: actions that do not
-use cascaded parsing will return nothing, those that do will simply return
-a string to be parsed.
+A more complex example is evaluating simple arithmetic expressions, where
+sub-expressions enclosed in parentheses must be evaluated first; cascaded
+parsing can be used to parse and evaluate sub-expressions, substituting the
+result into the original expression.  Actions do not need to be specially
+registered with the framework or be declared in a particular way to use
+cascaded parsing: actions that do not use cascaded parsing will return
+nothing, those that do will simply return a string to be parsed.
+
+XXX MAYBE MOVE THIS EXAMPLE TO AN APPENDIX\@?  DO I EVEN NEED IT\@?  ASK
+CARL ABOUT IT\@.
 
@@
 
-The architecture is line oriented at present: there is no facility for
-rules to consume more input or push unused input back onto the input
-stream.  This was not a deliberate design decision, rather a consequence of
-the line oriented nature of Postfix log files; more flexible approaches
-could be pursued.  Classifying inputs using the rules is simple: the first
+Classifying inputs using the rules is simple: the first
 rule to recognise the input determines the action that will be invoked;
@@
 can be complex if required, but simple conditions can be quite useful too,
-e.g.\ every Postfix log \empty{}line contains the name of the Postfix
-component that produced it, so every rule used in \parsername{} has a
-condition specifying the component whose log \empty{}lines it recognises,
-reducing the number of rules that will be used when classifying an input
-(see \sectionref{rules in implementation} for details) and increasing the
-chance that the input will be correctly recognised..
+e.g.\ every Postfix log line contains the name of the Postfix component
+that produced it, so every rule used in \parsername{} has a condition
+specifying the component whose log lines it recognises, reducing the number
+of rules that will be used when classifying an input (see \sectionref{rules
+in implementation} for details) and increasing the chance that the input
+will be correctly recognised.
 

### 2009-03-09 doc/logparser-background.tex
Improved checks, added a check, dropped a check, and made corrections based on the new check.

@@
   smtpd_client_restrictions =
-    ...
+    <other restrictions>
     permit_sasl_authenticated,
     check_client_access /etc/postfix/allow_webserver.cidr,
-    ...
+    <other restrictions>
 
@@
   smtpd_client_restrictions =
-    ...
+    <other restrictions>
     check_client_access /etc/postfix/require_dns_entries.cidr,
-    ...
+    <other restrictions>
 

### 2009-03-09 doc/logparser-implementation.tex
Improved checks, added a check, dropped a check, and made corrections based on the new check.

@@
         delayed for while it was being delivered  This will only be present
-        for delivered mails..
+        for delivered mails.
 

### 2009-03-09 doc/logparser-architecture.tex
Changes following more comments from Carl

@@
 
+\newpage{} % XXX
 \subsection{Attaching Conditions To Rules}
@@
 run-time, or embedding a language like Lua.  If an implementation is going
-to use conditions so complex that they will require a Turing-equivilent
+to use conditions so complex that they will require a Turing-equivalent
 programming language to implement them, the design may need to be
@@
 the comment start token even if it is within a quoted string, which is
-incorrect.  Rules one and two will be used when the parser is parsing code,
-not comments: rule one recognises a comment that is contained within one
-line and leaves the parser's state unchanged; rule two recognises the start
-of a comment and changes the parser's state to parsing comments.  Rules
-three and four will be used when the parser is parsing a comment: rule
-three recognises the end of a comment and switches the parser's state back
-to parsing code; rule four recognises a comment line without an end token
-and keeps the parser's state unchanged.  It is important that the rules are
-applied in the order listed in \tableref{Rules to parse C-style comments};
-\sectionref{rules in architecture} has explained how this is achieved, and
-\sectionref{overlapping rules in architecture} will discuss overlapping
-rules.
+incorrect; the rules could be extended with new rules to parse quoted
+strings, excluding comments from within strings.
+
+Rules one and two will be used when the parser is parsing code, not
+comments: rule one recognises a comment that is contained within one line
+and leaves the parser's state unchanged; rule two recognises the start of a
+comment and changes the parser's state to parsing comments.  Rules three
+and four will be used when the parser is parsing a comment: rule three
+recognises the end of a comment and switches the parser's state back to
+parsing code; rule four recognises a comment line without an end token and
+keeps the parser's state unchanged.  It is important that the rules are
+applied in the order listed in \tableref{Rules to parse C-style comments}
+because rule 2 overlaps with rule 1, so rule 1 must be used first to
+correctly parse comments that are contained within one line;
+\sectionref{rules in architecture} has explained how this is achieved.
+\sectionref{overlapping rules in architecture} will discuss the benefits
+and difficulties of using overlapping rules.
 
@@
 may overlap with multiple rules, and rule overlap is not transitive, e.g.\
-given these three conditions:
+given these three artificial conditions:
 
@@
 
-    \item $total < 10$
+    \item total $<$ 10
 
-    \item $total > 20$
+    \item total $>$ 20
 
-    \item $total < 30$
+    \item total $<$ 30
 
@@
 
-For this approach to be more efficient than the naive approach $y$ must be
-significantly lower than $x$.  If $y$ is higher than $x$ then the opposite
-approach should be taken: check for regex overlap first, then check
-for condition overlap only between pairs of rules with overlapping regexes.
+In the worst case no overlap will be found between conditions, so all pairs
+or regexes must be checked for overlap and $|o|$ above will be equal to
+$n^{2}$.  For this approach to be more efficient than the naive approach
+$y$ must be significantly lower than $x$.  If $y$ is higher than $x$ then
+the checks for overlap should be performed in the opposite order: check for
+regex overlap first, then check for condition overlap only between pairs of
+rules with overlapping regexes.
 
@@
 fast when there is a match, but when the match fails its computational
-complexity is $O(2^{n})$; see~\cite{mastering-regular-expressions} for
-in-depth discussion of nested quantifiers, backtracking, alteration, and
-capturing groups.  Pathological regexes which consume excessive CPU time
-can be difficult to detect, both by visual inspection and by machine
-inspection, but if a regex is converted to a \acronym{FA} or the internal
-representation used by the regex engine, it may be possible to determine if
-there are nested quantifiers or other troublesome constructs present.
-Modern regex engines have addressed many of these problems, e.g.\ the regex
-to match double quoted strings given above fails immediately with Perl
-5.10, regardless of the input length, because the regex engine looks for
-both of the required double quotes first.  Similarly Perl 5.10's regex
-engine optimises alterations starting with literal text into a trie, which
-has matching time proportional to the length of the alternatives, rather
-than the number of alternatives.  Perl regexes can use \verb!(?>pattern)!,
-which matches \verb!pattern! the first time the regex engine passes over
-it, and does not change what it originally matched if the regex engine
-backtracks over it, so troublesome regexes can use this to reduce the
-impact of backtracking; Prolog users will notice a similarity to the
-\verb'!' (cut) operator.  A presentation showing some of Perl 5.10's new
-regex features is available at
+complexity for a string of length $n$ is $O(2^{n})$;
+see~\cite{mastering-regular-expressions} for in-depth discussion of nested
+quantifiers, backtracking, alteration, and capturing groups.  Pathological
+regexes which consume excessive CPU time can be difficult to detect, both
+by visual inspection and by machine inspection, but if a regex is converted
+to a \acronym{FA} or the internal representation used by the regex engine,
+it may be possible to determine if there are nested quantifiers or other
+troublesome constructs present.  Modern regex engines have addressed many
+of these problems, e.g.\ the regex to match double quoted strings given
+above fails immediately with Perl 5.10, regardless of the input length,
+because the regex engine looks for both of the required double quotes
+first.  Similarly Perl 5.10's regex engine optimises alterations starting
+with literal text into a trie, which has matching time proportional to the
+length of the alternatives, rather than the number of alternatives.  Perl
+regexes can use \verb!(?>pattern)!, which matches \verb!pattern! the first
+time the regex engine passes over it, and does not change what it
+originally matched if the regex engine backtracks over it, so troublesome
+regexes can use this to reduce the impact of backtracking; Prolog users
+will notice a similarity to the \verb'!' (cut) operator.  A presentation
+showing some of Perl 5.10's new regex features is available at
 \urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.
@@
 provides, plus any difficulties associated with the components.  The
-framework, unsurprisingly, was quite quick to document, because it is the
-simplest of the three components.  The actions are also quite easy to
-understand, because the architecture does not impose any structure or
-requirements upon them: parser authors are free to do anything they want
-within an action.  The architecture's support for cascaded parsing was
-described in the actions section, with an example to illustrate how it can
-be useful.  The rules section was the longest, because although the rules
-appear conceptually simple --- recognise an input and specify the action
-that handles that input --- they have subtle behaviour that needs to be
-clearly explained.  When extending a ruleset a decision needs to be taken
-about whether the input should be recognised by extending an existing rule,
-by adding a new rule to an existing rule category, or by adding a new rule
-category, rule, and action.  Rules can have conditions attached to them,
-restricting the set of rules used to recognise an input; the complexity of
-the conditions used greatly influences the difficulty of writing a correct
-ruleset or understanding and extending an existing ruleset.  Overlapping
-rules are frequently a requirement in a parser, and their use can greatly
-simplify some rules, but they can be a source of bugs because they can
-recognise inputs unexpectedly.  The framework does not try to detect
-overlapping rules, because overlap amongst rules may be valid and is quite
-often intentional; that responsibility falls to the author of the rules.
-The difficulty involved in detecting overlap is proportional to the
-complexity of a ruleset's regexes and conditions, and may be possible for a
-human yet intractable or impossible for a program.  The rules section
-finishes with a discussion of pathological rules, concentrating on
-pathological regexes.
+framework provides a number of support functions and manages the parsing
+process, enabling simple rules and actions to be written.  The actions are
+quite easy to understand, because the architecture does not impose any
+structure or requirements upon them: parser authors are free to do anything
+they want within an action.  The architecture's support for cascaded
+parsing was described in the actions section, with an example to illustrate
+how it can be useful for general parsing.  The rules section was the
+longest, because although the rules appear conceptually simple ---
+recognise an input and specify the action that handles that input --- they
+have subtle behaviour that needs to be clearly explained.  When extending a
+ruleset a decision needs to be taken about whether the input should be
+recognised by extending an existing rule, by adding a new rule to an
+existing rule category, or by adding a new rule category, rule, and action.
+Rules can have conditions attached to them, restricting the set of rules
+used to recognise an input; the complexity of the conditions used greatly
+influences the difficulty of writing a correct ruleset or understanding and
+extending an existing ruleset.  Overlapping rules are frequently a
+requirement in a parser, and their use can greatly simplify some rules, but
+they can be a source of bugs because they can recognise inputs
+unexpectedly.  The framework does not try to detect overlapping rules,
+because overlap amongst rules may be valid and is quite often intentional;
+that responsibility falls to the author of the rules.  The difficulty
+involved in detecting overlap is proportional to the complexity of a
+ruleset's regexes and conditions, and may be possible for a human yet
+intractable or impossible for a program.  The rules section finishes with a
+discussion of pathological rules, concentrating on pathological regexes.

### 2009-03-09 doc/logparser-introduction.tex
Changes following more comments from Carl

@@
 developed for this project, beginning with an overview of the architecture,
-then describing the components of the architecture in detail.
+then describing each of the components of the architecture in detail.
 

### 2009-03-09 doc/logparser-architecture.tex
Removed the overly long cascaded parsing example.

@@
         more input or push unused input back onto the input stream,
-        although actions may use cascaded parsing (\sectionref{cascaded
-        parsing}) to effectively push input back onto the input stream.  A
-        physical input (i.e.\ a single line read from the input stream) may
-        contain multiple or partial logical inputs, and the framework must
-        pre-process these to provide a logical input for the rules to
-        recognise.  This simplifies the rules and actions considerably, at
-        the cost of added complexity in the framework; during the design
-        phase it was decided that it was easier to deal with the problem of
-        parsing multiple or partial inputs once, rather than dealing with
-        it in every rule and action.  This is easy in the case of Postfix
-        log files because there is a one-to-one mapping between physical
-        and logical inputs; mapping between physical and logical inputs may
-        be more difficult for other types of input.  Some input types may
-        require pre-processing equivalent to parsing the physical inputs;
-        in those cases the framework should behave similarly to other
-        architectures: present the entire input stream at once, and discard
-        the portion recognised by the successful rule.
+        although actions may use cascaded parsing (explained in
+        \sectionref{actions in architecture}) to effectively push input
+        back onto the input stream.  A physical input (i.e.\ a single line
+        read from the input stream) may contain multiple or partial logical
+        inputs, and the framework must pre-process these to provide a
+        logical input for the rules to recognise.  This simplifies the
+        rules and actions considerably, at the cost of added complexity in
+        the framework; during the design phase it was decided that it was
+        easier to deal with the problem of parsing multiple or partial
+        inputs once, rather than dealing with it in every rule and action.
+        This is easy in the case of Postfix log files because there is a
+        one-to-one mapping between physical and logical inputs; mapping
+        between physical and logical inputs may be more difficult for other
+        types of input.  Some input types may require pre-processing
+        equivalent to parsing the physical inputs; in those cases the
+        framework should behave similarly to other architectures: present
+        the entire input stream at once, and discard the portion recognised
+        by the successful rule.
 
@@
         marshals the data extracted by the rule, invokes the action, and
-        repeats the parsing process if cascaded parsing
-        (\sectionref{cascaded parsing}) is used.
+        repeats the parsing process if cascaded parsing (see
+        \sectionref{actions in architecture}) is used.
 
@@
 
-\subsection{Cascaded Parsing}
-
-\label{cascaded parsing}
-
 Actions may return a modified input line to the framework that will be
@@
 have to handle comments at all; obviously if comment characters can be
-escaped or embedded in quoted strings the implementation needs to be more
-complex, but for some inputs this can greatly simplify subsequent parsing.
-A more complex example is evaluating simple arithmetic expressions, where
-sub-expressions enclosed in parentheses must be evaluated first; cascaded
-parsing can be used to parse and evaluate sub-expressions, substituting the
-result into the original expression.  Actions do not need to be specially
-registered with the framework or be declared in a particular way to use
-cascaded parsing: actions that do not use cascaded parsing will return
-nothing, those that do will simply return a string to be parsed.
-
-XXX MAYBE MOVE THIS EXAMPLE TO AN APPENDIX\@?  DO I EVEN NEED IT\@?  ASK
-CARL ABOUT IT\@.
-
-XXX GET GREG TO PROOFREAD THIS SECTION\@.
-
-A more complicated example is determining the atomic weight of a compound
-given its chemical formula.  A chemical formula is a sequence of chemical
-element abbreviations, each optionally followed by a number indicating how
-many of that element are present in the formula (if omitted the number is
-assumed to be one).  Some parts of a formula may be repeated: they are
-enclosed in parentheses, and the closing parenthesis is followed by a
-number indicating how many are present in the formula.  A monomer is a
-simple compound, i.e.\ it does not have any repeated portions; compounds
-with repeated portions are called polymers.  E.g.\ glucose, a monomer, has
-the chemical formula $C_{6}H_{12}O_{6}$, and hydrocarbon, a polymer, is
-$CH_{3}{(CH_{2})}_{50}CH_{3}$.  The atomic weight of an individual element
-is given in periodic tables, but one element in particular will be
-important in the later explanation: Hydrogen, abbreviated H, has an atomic
-weight of 1.  The atomic weight of a given compound is the sum of the
-atomic weights of each of its elements multiplied by the number of times
-they occur; this is simple to calculate for a monomer, but it is harder for
-a polymer because the atomic weight of the repeated monomer must be
-multiplied by the number of times the monomer is present in the compound
-(e.g.\ for hydrocarbon the atomic weight of $CH_{2}$ must be multiplied by
-$50$).
-
-Calculating the atomic weight of a monomer requires a single rule and
-action.  The action separates each element in the monomer, looks up its
-atomic weight, multiplies it by the number of times it is present (one if
-unspecified), and adds up the results.  The rule's regex is
-\verb!^([A-Z]([a-z])?\d*)+$!, which does not guarantee that the chemical
-element abbreviations in the formula are correct; it could be modified to
-check that by replacing \verb![A-Z]([a-z])?! with all the valid
-abbreviations, slightly simplifying the action at the cost of complicating
-the regex, which could be automatically generated.
-
-Parsing a polymer is more difficult because of the repeated monomers: each
-monomer's atomic weight needs to be calculated separately, multiplied by
-the number of times it is present in the compound, and that figure added to
-the sum.  Without cascaded parsing the action would need to maintain state
-while recursively parsing the monomers embedded in the polymer.  Using
-cascaded parsing, three rules and actions can be used instead as shown in
-\tableref{Regexes and Actions to calculate the atomic weight of a chemical
-compound}.
-
-% These figures come from experimentation; set the length to 0pt, see how
-% big a hbox latex complains about, and then set the length.  Stupid latex
-% cannot do \settowidth{\verb}.  Poo.
-\newlength{\chemicalactionwidth}
-\setlength{\chemicalactionwidth}{\textwidth}
-% This is the width of the number column, roughly.
-\addtolength{\chemicalactionwidth}{-30pt}
-% The width of the regex column.
-\addtolength{\chemicalactionwidth}{-130pt}
-% This is the overhead of space around columns, on either side, etc.
-\addtolength{\chemicalactionwidth}{-24pt}
-
-\begin{table}[ht]
-    \caption{Regexes and Actions to calculate the atomic weight of a
-    chemical compound}
-    \empty{}\label{Regexes and Actions to calculate the atomic weight of a
-    chemical compound}
-    \begin{tabular}{llp{\chemicalactionwidth}}
-
-        \tabletopline{}%
-        No.   & Regex                       & Action \\
-        \tablemiddleline{}%
-        1     & \verb'((?!H\d+([()]|$))'    & Parse a monomer as described above,  \\
-              & \verb' ([A-Z]([a-z])?\d*)+)'& returning \verb!Hnnn!, where \verb!nnn! is
-                                              the atomic weight of the monomer.  The first
-                                              part of the regex prevents it from parsing
-                                              the return value of this action
-                                              (\verb!Hnnn!), ensuring the parsing process
-                                              does not get stuck in a loop.  \\
-        2     & \verb!\(H(\d+)\)(\d+)!      & Parse a repeated monomer that has been
-                                              rewritten by the first rule and action,
-                                              returning \verb!Hnnn!, where \verb!nnn! is
-                                              the product of the first and second numbers
-                                              captured by the regex.  \\
-        3     & \verb!^H(\d+)+$!            & Repeated applications of rules and
-                                              actions one and two will have rewritten the
-                                              input to \verb!Hnnn!, where \verb!nnn! is
-                                              the atomic weight of the compound.  This
-                                              action should do whatever is required with
-                                              the calculated atomic weight.  \\
-        \tablebottomline{}%
-
-    \end{tabular}
-\end{table}
-
-Applying the rules and actions above to the hydrocarbon formula results in
-this sequence:
-
-\begin{tabular}{lrl}
-
-    1 & Start with              & $CH_{3}{(CH_{2})}_{50}CH_{3}$ \\
-    2 & Apply rule and action 1 & $H_{15}{(CH_{2})}_{50}CH_{3}$ \\
-    3 & Apply rule and action 1 & $H_{15}{(H_{14})}_{50}CH_{3}$ \\
-    3 & Apply rule and action 1 & $H_{15}{(H_{14})}_{50}H_{15}$ \\
-    5 & Apply rule and action 2 & $H_{15}H_{700}H_{15}$ \\
-    6 & Apply rule and action 1 & $H_{730}$ \\
-    7 & Apply rule and action 3 & 730 \\
-
-\end{tabular}
+escaped or embedded in quoted strings the implementation must be careful
+not to remove those.  For some inputs this type of pre-processing can
+greatly simplify parsing, echoing the simplification provided by the
+framework presenting rules and actions with logical inputs rather than
+physical inputs.  A more complex example use of cascaded parsing is
+evaluating simple arithmetic expressions, where sub-expressions enclosed in
+parentheses must be evaluated first; cascaded parsing can be used to parse
+and evaluate the sub-expressions, substituting the result into the original
+expression for subsequent re-evaluation.  Actions do not need to be
+specially registered with the framework or be declared in a particular way
+to use cascaded parsing: actions that do not use cascaded parsing will
+return nothing, those that do will simply return a string to be parsed.
 
@@
 
-\newpage{} % XXX
 \subsection{Attaching Conditions To Rules}

### 2009-03-09 doc/logparser-architecture.tex
Spell check.

@@
         concerns (\sectionref{parser efficiency}), rule conditions
-        (\sectionref{attaching conditions to rules}), or prioritisations
+        (\sectionref{attaching conditions to rules}), or rule priorities
         (\sectionref{rules in architecture}) that complicate the process.
@@
 one program to answer questions about another program's behaviour if the
-inquiring program is implemented on a compuational machine of the equal or
+inquiring program is implemented on a computational machine of the equal or
 lesser power to the program being inquired into.

### 2009-03-09 doc/logparser-conclusion.tex
Divided rules section in architecture into subsections.
Spell check.

@@
 XXX THERE IS NO MENTION OF RESULTS/EFFICIENCY\@.
+
+XXX PULL CONTENT FROM THE PAPER\@.
 
@@
 implementation}) a relatively simple task.  The simplicity of adding a new
-action frees the implementor from worrying about how their action might
+action frees the implementer from worrying about how their action might
 interfere with existing parsing, so they can focus on correctly

### 2009-03-09 doc/logparser-implementation.tex
Spell check.

@@
 
-        In the original algorithm the purpose of the preprocessing stage
+        In the original algorithm the purpose of the pre-processing stage
         was to reduce the memory consumption of the program.  In the first
@@
 taken in the parser's design and implementation to make adding new actions
-as painless as possible.  The implementor writes a subroutine that accepts
+as painless as possible.  The implementer writes a subroutine that accepts
 the standard arguments given to actions, and registers it as an action by
@@
 action; this helps catch mistakes made when adding new rules.  No other
-work is required from the implementor to integrate the action into the
+work is required from the implementer to integrate the action into the
 parser; all of their attention and effort can be focused on correctly
@@
 the author of the new action wishes, they can take advantage of the
-parser's object oriented implementation by subclassing it and implementing
+parser's object oriented implementation by sub-classing it and implementing
 their changes in the derived class, allowing future upgrades of the parser
@@
 
-If a mail is mis-classified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably mis-classification in
+If a mail is misclassified as a bounce message it will almost certainly
+have been generated internally by Postfix; arguably misclassification in
 this case is a benefit rather than a drawback, as other mails generated

### 2009-03-09 doc/logparser-other-parsers-review.tex
Spell check.

@@
 Solaris hosts (and possibly other operating systems) thus require
-preprocessing before parsing by \acronym{LMA}.  The \numberOFlogFILES{}
-preprocessed test log files were parsed without complaint by \acronym{LMA},
-although it produced 32 entries in the output \acronym{CSV} file for every
-rejection in the input log file; it also missed some 40\% of delivered
-mail.  Once these deficiencies were discovered the author did not spend any
-more time checking the results.
+pre-processing before parsing by \acronym{LMA}.  The \numberOFlogFILES{}
+pre-processed test log files were parsed without complaint by
+\acronym{LMA}, although it produced 32 entries in the output \acronym{CSV}
+file for every rejection in the input log file; it also missed some 40\% of
+delivered mail.  Once these deficiencies were discovered the author did not
+spend any more time checking the results.
 
@@
 containing interlinked log lines.  It is not suitable as a base for this
-parser, as the source code made available is obfuscated and not intended
+parser, as the source code made available is encrypted and not intended
 for modification; in addition the architecture would probably need to be
@@
         \item [\nialpha{}] Sawmill's source code is available in an
-            obfuscated form, so that customers can compile it on platforms
+            encrypted form, so that customers can compile it on platforms
             that pre-compiled binaries are not available for.

### 2009-03-09 doc/logparser-results.tex
Fixed the problem where \label{} and \ref{} didn't work properly with glossary chapters.
Corrections, improvements, and new content in architecture chapter. Added SGAI and Elsevier entries to bibliography. Some minor cleanup elsewhere.
Some minor formatting fixes. Documented some of 'make check'. Added check for sentences with only one space separating them.
Spell check.

@@
 
-\subsection{Caching Compiled Regexs}
+\subsection{Caching Compiled Regexes}
 
@@
 
-Before the perl interpreter attempts to match a regex against a piece of
+Before the Perl interpreter attempts to match a regex against a piece of
 text, the regex is compiled into an internal representation and
@@
 when not caching compiled regexes graph}
-
-\newpage{} % XXX
 
@@
 
-XXX REWRITE COVERAGE SUMARY IF NECESSARY\@.
+XXX REWRITE COVERAGE SUMMARY IF NECESSARY\@.
 

### 2009-03-10 doc/logparser-implementation.tex
Some minor changes.

@@
         replaced by \texttt{.+}, which means to match zero or more of any
-        character.  XXX SHOULD I USE \verb![\S]+! INSTEAD\@?
+        character.
 
@@
 
-\begin{eqlist}
+\begin{eqlist}[\def\makelabel#1{\bfseries#1:}]
+
+    \squeezeitems{}
+
+    \item [rule] The matching rule.
+
+    \item [matches] The data extracted from the line by the rule's regex.
 
@@
 
-        \begin{eqlist}
+        \begin{eqlist}[\def\makelabel#1{\bfseries#1:}]
+
+            \squeezeitems{}
 
@@
         \end{eqlist}
-
-    \item [rule] The matching rule.
-
-    \item [matches] The data extracted from the line by the rule's regex.
 
@@
 
-\newpage{} % XXX
-
 \subsection{Further Aborted Delivery Attempts}
 
-XXX MAYBE FIND OUT WHAT CLIENTS DISCONNECT\@?  Some mail clients disconnect
-abruptly if a second or subsequent recipient is rejected; they may also
-disconnect after other errors, but such disconnections are either
-unimportant or are handled elsewhere in the parser
-(\sectionref{timeouts-during-data-phase}).  Sadly, Postfix does not log a
-message saying the mail has been discarded, as should be expected by now.
-The checks to identify this happening are:
+Some mail clients disconnect abruptly if a second or subsequent recipient
+is rejected; they may also disconnect after other errors, but such
+disconnections are either unimportant or are handled elsewhere in the
+parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix does not
+log a message saying the mail has been discarded, as should be expected by
+now.  The checks to identify this happening are:
 
@@
     \item The hostname used in the HELO command is not logged if the mail
-        is accepted.\footnote{Tested with Postfix 2.2.10, 2.3.11, and
-        2.4.7; this may possibly have changed in Postfix 2.5.  XXX TEST
-        THIS AGAIN.}  Rectifying this is relatively simple: create a
+        is accepted.  Rectifying this is relatively simple: create a
         restriction that is guaranteed to warn for every accepted mail, as

### 2009-03-11 doc/logparser-implementation.tex
Rearranged sections; the content hasn't changed, just the ordering.
Corrections for the assumptions and database sections of the implementation chapter.

@@
 
-XXX FIGURE OUT WHEN TO USE ACTIONS AND WHEN TO USE ALGORITHM, ETC\@.
-
-XXX THIS MIGHT BE USEFUL SOMEWHERE\@:
-
-XXX SHOULD I SAY ``cleaned up'' EVERYWHERE OR NOT\@?
-
-Whereas the rules are quite simple and each rule is completely independent
-of the other rules, the algorithm is significantly more complicated and
-highly internally interdependent.  The algorithm deals with all the
-complications of parsing, the eccentricities and oddities of Postfix log
-files, presenting the resulting data in a normalised, simple to use
-representation.  The algorithm's task is to follow the journey each mail
-takes through Postfix, combining the data extracted by rules into a
-coherent whole, saving it in a useful and consistent form, and performing
-housekeeping duties.
+XXX FIGURE OUT WHEN TO USE ACTIONS AND WHEN TO USE ALGORITHM, PARSER,
+ETC\@.
+
+The parser deals with all the complications of parsing, the eccentricities
+and oddities of Postfix log files, presenting the resulting data in a
+normalised, simple to use representation.  The parser's task is to follow
+the journey each mail takes through Postfix, combining the data extracted
+by rules into a coherent whole, saving it in a useful and consistent form,
+and performing housekeeping duties.
 
@@
         unlikely that any of these problems will arise, though it is of
-        course possible, particularly when suffering a deluge of spam or a
-        mail loop.
+        course possible, particularly when suffering from a deluge of spam
+        or a mail loop.
 
@@
         \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts} and \sectionref{pickup
+        \sectionref{aborted-delivery-attempts}, and \sectionref{pickup
         logging after cleanup} for details.
@@
 In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  There are
-advantages to treating Postfix as a black box during parser development:
+black box program based solely on its inputs and outputs.  XXX REFER TO
+BLACK BOX PAPER\@?  There are some advantages to treating Postfix as a
+black box during parser development:
 
@@
         the idealised log files someone would naturally envisage reading
-        the source code.
-
-    \item The source code cannot accurately communicate the variety of
-        orderings in which log lines are written to the log file, as
-        process scheduling independently interferes with logging and other
-        processing.
+        the source code.  The source code cannot accurately communicate the
+        variety of orderings in which log lines are written to the log
+        file, as process scheduling independently interferes with logging
+        and other processing.
 
@@
         separate project could compare the empirical knowledge inherent in
-        the parsing algorithm with the documentation and source code of
-        Postfix to see how closely the two agree.
+        the parsing algorithm with Postfix's documentation and source code
+        to see how closely the two agree.
 
@@
 
-\section{Database Schema}
-\label{database schema}
-
-XXX ADD A DIAGRAM OF THE DATABASE SCHEMA\@.
+\section{Parser Flow Chart}
+
+\label{flow-chart}
+
+XXX MIGHT NEED A SMALL INTRO\@.
+
+From the viewpoint of an individual mail passing through the parser the
+experience could be summarised as:
+
+\begin{enumerate}
+
+    \item Mail enters the system via \acronym{SMTP} or local submission.
+
+    \item If the mail is rejected, save all data and finish.
+
+    \item Follow the progress of the accepted mail until it is either
+        delivered, bounced, or deleted, then save all data, and finish.
+
+\end{enumerate}
+
+\Figureref{flow chart image} shows the most common paths the data
+representing a mail can take through the parser; the complications
+described in \sectionref{complications} are excluded for the sake of
+clarity.
+
+\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart}{flow chart
+image}
+
+\label{mail-enters-the-system}
+
+Everything starts off with a mail entering the system, whether by local
+submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
+forwarding, or internally generated by Postfix.  Local submission is the
+simplest case: a queueid is assigned immediately and the sender address is
+logged (action: PICKUP\@; flowchart:~2).  Re-injection due to forwarding
+sadly lacks explicit log lines of its own; it is explained in
+\sectionref{Re-injected mails}.
+
+Internally generated mails lack any explicit origin in Postfix 2.2.x and
+must be detected using heuristics as described in
+\sectionref{identifying-bounce-notifications}.  Bounce notifications are
+the primary example of internally generated mails, though there may be
+other types.  Postfix may generate mails to the administrator when it
+encounters configuration errors, but such mails are presumably rare.
+
+\acronym{SMTP} is more complicated than local submission:
+
+\begin{enumerate}
+
+    \item First there is a connection from the remote client (action:
+        CONNECT\@; flowchart:~1).
+
+    \item This is followed by rejection of sender address, recipient
+        addresses, client \acronym{IP} address or hostname, HELO hostname,
+        etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
+        more mails (action: CLONE\@; flowchart:~5); or some interleaving of
+        both.
+
+    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
+        Postfix has rejected any \acronym{SMTP} commands the data will be
+        saved to the database; if not there will not be any data to save
+        (any mails accepted will already have been cloned so their data is
+        in another data structure).
+
+    \item If one or more mails were accepted there will be more log lines
+        for those mails later, see \sectionref{mail-delivery}.
+
+\end{enumerate}
+
+\label{mail-delivery}
+
+The obvious counterpart to mail entering the system is mail leaving the
+system, whether by deletion, bouncing, local delivery, or remote delivery.
+All four are handled in exactly the same way:
+
+\begin{enumerate}
+
+    \item Postfix will log the sender and recipient addresses separately
+        (action: SAVE\_DATA\@; flowchart:~9).
+
+    \item Sometimes mail is re-injected and the child mail needs to be
+        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
+        \sectionref{tracking re-injected mail} discusses this in
+        detail.
+
+    \item Eventually the mail will be delivered, bounced, or deleted by the
+        administrator (action: COMMIT\@; flowchart:~12).  This is the last
+        log line for this particular mail (though it may be indirectly
+        referred to if it was re-injected).  If it is neither parent nor
+        child of re-injection the data is cleaned up and entered in the
+        database (flowchart:~14), then deleted from the state tables.
+        Re-injected mails are described in \sectionref{tracking re-injected
+        mail}.
+
+\end{enumerate}
+
+It should be reiterated that the actions above happen whether the mail is
+delivered to a mailbox, piped to a command, delivered to a remote server,
+bounced (due to a mail loop, delivery failure, or five day timeout), or
+deleted by the administrator, \textit{unless\/} the mail is either parent
+or child of re-injection, as explained in \sectionref{tracking re-injected
+mail}.
+
+\section{Database}
+
+\label{database}
 
 The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by applying those rules to Postfix log
+the rules and the data gleaned by using those rules to parse Postfix log
 files.  Understanding the database schema is important in understanding the
 actions of the parser, and essential to developing further applications
-which utilise the data gathered.
+which utilise the data gathered; \sectionref{database as API} describes
+how the database schema functions as an \acronym{API}.
 
 The database schema can be conceptually divided in two: the rules used to
-parse log files, and the data saved from the parsing of log files.  Rules
-have the fields required to parse the log lines, extract data to be saved,
-and specify the action to be executed; they also have several fields that
-aid the user in understanding the meaning of the log lines parsed by each
-rule.  The rules are described in detail in \sectionref{rules in
-implementation} but the fields are covered in \sectionref{rule attributes}.
-
-The data saved from parsing the log files is also divided into two tables
-as described below: connections and results.  The connections table
-contains a row for every mail accepted and every connection where there was
-a rejection; the individual fields will be described in
-\sectionref{connections table}.  There will be at least one row in the
-results table for each row in the connections table; the fields will be
-covered in detail in \sectionref{results table}.
+recognise log lines, and the data saved from the parsing of log files.
+Rules have the fields required to recognise log lines, extract data to be
+saved, and specify the action to be invoked; they also have several fields
+that aid the user in understanding the meaning of the log lines recognised
+by each rule.  The rules are described in detail in \sectionref{rules in
+implementation}, but the fields are covered with the rest of the database
+schema in this section.
+
+The data saved from parsing the log files is also divided into two tables:
+connections and results.  The connections table contains an entry for every
+mail accepted and every connection where there was a rejection; the
+individual fields will be described in \sectionref{connections table}.
+There will be at least one entry in the results table for each entry in the
+connections table; the result table's fields will be covered in detail in
+\sectionref{results table}.  A diagram of the database schema is provided
+in \sectionref{Diagram of the Database Schema} before the in-depth
+descriptions of each table.
 
@@
 required when examining a result, the rule that produced the result is
-available for inspection: each result references the rule that created it.
-There is no ambiguity about which rule resulted in a particular result,
-eliminating one potential source of confusion.
+available for inspection because each result references the rule that
+created it.  There is no ambiguity about which rule resulted in a
+particular result, eliminating one potential source of confusion.
 
@@
 
-The database populated by this program provides a simple interface to
+The database populated by \parsername{} provides a simple interface to
 Postfix log files.  Although the interface is a database schema rather than
-a set of functions provided by a library, it performs the same function as
-any other \acronym{API}: it provides a stable interface, allowing code on
-either side of the interface to be changed without adverse effects, as long
-as the interface is maintained.  The parser can be improved to handle
-additional complications, support earlier or later releases of Postfix,
-bugs can be fixed or limitations removed.  Programs that use the database
-can range from the simple examples in \sectionref{motivation} to far more
-complex data mining tools.
+a set of functions in a library, it has the same benefits as any other
+\acronym{API}: it provides a stable interface, allowing code on either side
+of the interface to be changed without adverse effects, as long as the
+interface is adhered to.  Programs that use the database can range from the
+simple examples in \sectionref{motivation} to far more complex data mining
+tools.
 
@@
 
-    \item Facilities are provided to access databases from most programming
-        languages, allowing a developer to access the data gathered using
-        their preferred programming language, rather than being restricted
-        to the language the parser is written in.  It is often possible to
-        write an interface layer allowing code written in one language to
-        be used in another language, but this greatly increases the effort
-        required to use the parser.
-
-    \item Databases provide complex querying and sorting functionality to
+    \item Most programming languages have facilities for database access,
+        allowing a developer to write programs that use the gathered data
+        in their preferred programming language, rather than being
+        restricted to the language the parser is written in.
+
+    \item Databases provide complex querying and sorting functionality for
         the user without requiring large amounts of programming.  All
@@
 
-            \item If the database supports it, access can be granted on a
-                fine-grained basis, e.g.\ allowing the finance department
-                to produce invoices, the helpdesk to run limited queries as
-                part of dealing with support calls, and the administrators
-                to have full access to the data.
+            \item Some databases support granting access on a fine-grained
+                basis, e.g.\ allowing the finance department to produce
+                invoices, the helpdesk to run limited queries as part of
+                dealing with support calls, and the administrators to have
+                full access to the data.
 
@@
     WHERE sender = 'boss@example.com'
-        AND postfix_action = 'REJECTED'
+        AND action = 'DELIVERY_REJECTED'
     SEND PANIC EMAIL TO 'postmaster@example.com';
@@
 
-            \item Other tables can be added to the database, e.g.\ to cache
+            \item Other tables can be added to the database, to cache
                 historical, summary, or computed data.
@@
         be familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn: a good
-        introduction to \acronym{SQL} can be found at
+        readily available resources from which to learn, e.g.\
         \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
         Although every vendor implements a different dialect of
-        \acronym{SQL}, the basics are the same everywhere (analogous to the
+        \acronym{SQL}, the basics are the same everywhere, analogous to the
         overall similarities and minor differences amongst Irish English,
-        British English, American English, and Australian English).
+        British English, American English, and Australian English.
         Depending on the database in use there may be tools available that
         reduce or remove the requirement to know \acronym{SQL}; several are
-        available for \gls{SQLite}, the default database used by the parser
-        implementation.
+        available for \gls{SQLite}, the default database used by
+        \parsername{}.
 
@@
 
+\subsection{Diagram of the Database Schema}
+
+\label{Diagram of the Database Schema}
+
+XXX ADD A DIAGRAM OF THE DATABASE SCHEMA\@.
+
 \subsection{Rules Table}
@@
 Rules are discussed in detail in \sectionref{rules in implementation}, but
-the structure of the rules table is covered here.  Rules are created by the
-user, not the parser, and will not be modified by the parser (except for
-the hits and hits\_total fields).  Rules parse the individual log lines,
-extracting data to be saved in the connections and results tables, and
-specifying the action to take for that log line.
+the structure of the rules table is covered here with the other tables in
+the database.  Rules are created by the user, not the parser, and will not
+be modified by the parser (except for the hits and hits\_total fields).
+Rules classify the individual log lines, extracting data to be saved in the
+connections and results tables, and specifying the action to take for that
+log line.
 
@@
 
-\begin{eqlist}
+\begin{boldeqlist}
 
@@
     \item [restriction\_name] The restriction that caused the mail to be
-        rejected.  Only applicable to rules that have an action of
-        \texttt{REJECTION}, other rules should have an empty string.
+        rejected.  Only applicable to rules that recognise rejection log
+        lines, i.e.\ rules that have an action of
+        \texttt{DELIVERY\_REJECTED}, other rules should have an empty
+        string.
 
@@
 
-    \item [program] The program (\daemon{smtpd}, \daemon{qmgr}, etc.) whose
-        log lines the rule applies to.  This avoids needlessly trying rules
+    \item [program] The Postfix component (e.g.\ \daemon{smtpd}) whose log
+        lines the rule applies to.  This avoids needlessly trying rules
         that will not match the log line, or worse, might match
@@
         against any log lines that are not parsed by program specific
-        rules.
-
-    \item [regex] The regex to match the log line against.  The regex will
+        rules.  This is a simple example of the rule conditions described
+        in \sectionref{attaching conditions to rules}.
+
+    \item [regex] The regex to recognise log lines with.  The regex will
         have several keywords expanded when the rules are loaded: this
@@
         efficiency the keywords are expanded and every rule's regex is
-        compiled before attempting to parse the log file --- otherwise each
+        compiled before attempting to parse the log file; otherwise every
         regex would be recompiled each time it was used, resulting in a
-        large, data dependent slowdown.  Efficiency concerns are discussed
-        in \sectionref{parser efficiency}, with caching compiled regexes
-        covered in \sectionref{Caching compiled regexes}.
+        large, data dependent slowdown, as shown in \sectionref{Caching
+        compiled regexes}.
 
@@
         \newline{} \tab{} \texttt{ client\_hostname = localhost,}
-        \newline{} \tab{} \tab{} \texttt{client\_ip
-        = 127.0.0.1} \newline{} i.e.\ semi-colon or comma separated
-        assignment statements, with the variable name on the left and the
-        value on the right hand side.\footnote{Commas and semi-colons
-        cannot be escaped and thus cannot be used.  This feature is
-        intended for use with small amounts of data rather than large
-        amounts in any one rule, so dealing with escape sequences was
-        deemed unnecessary.}  Any field in the connections table can be set
-        in this way.
+        \newline{} \tab{} \tab{} \texttt{client\_ip = 127.0.0.1} \newline{}
+        i.e.\ semi-colon or comma separated assignment statements, with the
+        variable name on the left and the value on the right hand
+        side.\footnote{Commas and semi-colons cannot be escaped and thus
+        cannot be included in data.  This feature is intended for use with
+        small amounts of data, so dealing with escape sequences was deemed
+        unnecessary.}  Any field in the connections table can be set in
+        this way.
 
@@
 
-    \item [action] The action that will be invoked when this rule matches a
-        log line; a full list of actions and the parameters they are
-        invoked with can be found in \sectionref{actions in detail in
-        implementation}.
+    \item [action] The action that will be invoked when this rule
+        recognises a log line; a full list of actions and the parameters
+        they are invoked with can be found in \sectionref{actions in detail
+        in implementation}.
 
     \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully matches.  At the start of each run
-        the parser sorts the rules in descending order of hits, and at the
-        end of the run it updates every rule's hits field in the database.
-        Assuming that the distribution of log lines is reasonably
-        consistent across log files, rules matching more commonly occurring
-        log lines will be tried before rules matching less commonly
-        occurring log lines, reducing the parser's execution time.  Rule
-        ordering for efficiency is discussed in \sectionref{rule ordering
-        for efficiency}.
+        each time the rule successfully recognises a log line.  At the
+        start of each run the parser sorts the rules by how frequently each
+        rule recognised log lines, and at the end of the run it updates
+        every rule's hits field in the database.  Assuming that the
+        distribution of log lines is reasonably consistent across log
+        files, ordering rules by their recognition frequency will reduce
+        the parser's execution time.  Rule ordering for efficiency is
+        discussed in \sectionref{rule ordering for efficiency}.
 
@@
         more specific rules to take precedence over more general rules
-        (described in \sectionref{overlapping rules in architecture}).
-
-%    \item [cluster\_group] A reference to the \texttt{cluster\_group}
-%        table.  That table is used by the Decision Tree algorithm
-%        described in a separate document.
-
-\end{eqlist}
+        (described in \sectionref{rules in architecture}).
+
+\end{boldeqlist}
 
@@
 
-\end{eqlist}
+\end{boldeqlist}
 
@@
 
-Every log line a row in the results table, unless the
+XXX REWRITE THIS PARAGRAPH\@; FIGURE OUT WHAT THE HELL IS HAPPENING WITH
+POSTFIX\_ACTION\@.
+
+Every log line will have a row in the results table, unless the \newline{}
 \texttt{postfix\_action} field of the rule which parsed it was INFO or
 IGNORE\@.  Each row in the results table is associated with a single
-connection, and there may be many results per connection.
-
-\begin{eqlist}
+connection; a single connection will have at least one result associated
+with it, but will usually have several more, and may have hundreds.
+
+\begin{boldeqlist}
 
@@
 
-    \item [rule\_id] The id of the entry in the rules table that matched
+    \item [rule\_id] The id of the rule in the rules table that recognised
         the log line and created this result.
@@
         --- a mechanism that is quite useful for testing new restrictions.
-        This field will be 1 if the log line parsed was a warning rather
-        than a real rejection, or 0 for a real rejection.  This field
+        This field will be set to 1 if the log line parsed was a warning rather
+        than a real rejection, or to 0 for a real rejection.  This field
         should be ignored if the result is not a rejection, i.e.\ the
-        action field of the associated rule is not \texttt{REJECTION}.
+        action field of the associated rule is not
+        \texttt{DELIVERY\_REJECTED}.
 
@@
         line.  In general an \acronym{SMTP} code is only present in a log
-        line for a rejection or final delivery; results initially missing
-        an \acronym{SMTP} code will duplicate the \acronym{SMTP} code of
-        other results in that connection.  Some final delivery log lines do
-        not contain an \acronym{SMTP} code: in those cases the code is
-        specified by the rule, based on the success or failure represented
-        by the log line.
+        line for a rejection or final delivery; results whose log line did
+        not contain an \acronym{SMTP} code will duplicate the
+        \acronym{SMTP} code of other results in that connection.  Some
+        final delivery log lines do not contain an \acronym{SMTP} code: in
+        those cases the \acronym{SMTP} code is specified by the rule's
+        \texttt{result\_data} field, based on the success or failure
+        represented by the log line.
 
@@
         Mails sent over the same connection can be distinguished by their
-        queueid.
+        queueids.
 
@@
 
-    \item [delay] How long the mail was delayed for while it was being
+    \item [delay] How long the mail was delayed while it was being
         delivered.  This will only be present for delivered mails.
 
-    \item [delays] More details information about how long the mail was
-        delayed for while it was being delivered  This will only be present
-        for delivered mails.
+    \item [delays] More detailed information about how long the mail was
+        delayed while it was being delivered, again only present for
+        delivered mails.
 
@@
 
-\end{eqlist}
-
-\section{Rules}
-
-\label{rules in implementation}
-
-XXX NEED AN INTRODUCTION\@.
-
-XXX ONCE THIS PARAGRAPH HAS BEEN INTEGRATED AND EXPANDED UPDATE THE
-REFERENCE IN ARCHITECTURE->RULES->ATTACHING CONDITIONS TO RULES\@.
-Each log line contains the name of the Postfix component that produced it,
-and this information is used when determining which rules should be used to
-parse each log line (see \sectionref{rules in implementation} for details).
-
-\subsection{Overlapping Rules}
-
-\label{overlapping rules in implementation}
-
-The parser does not try to detect overlapping rules; that responsibility is
-left to the author of the rules.  Unintentionally overlapping rules lead to
-inconsistent parsing and data extraction because the order in which rules
-are tried against each line may change between log files, and the first
-matching rule wins.  Overlapping rules are frequently a requirement,
-allowing a more specific rule to match some log lines and a more general
-rule to match the majority, e.g.\ separating \acronym{SMTP} delivery to
-specific sites from \acronym{SMTP} delivery to the rest of the world.  The
-algorithm provides a mechanism for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Negative priorities may be useful
-for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by regex and visually inspect the list, e.g.\ with
-        \acronym{SQL} similar to: \textbf{select regex from rules order by
-        regex;}
-
-    \item Compare the results of parsing using sorted, shuffled, and
-        reversed rules (described in \sectionref{rule ordering for
-        efficiency}).  Parse several log files using optimal ordering, then
-        dump a textual representation of the rules, connections, and
-        results tables.  Repeat with shuffled and reversed ordering,
-        starting with a fresh database.  If there are no overlapping rules
-        the tables from each run will be identical; differences indicate
-        overlapping rules.  The rules that overlap can be determined by
-        examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing results
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the log files that are matched by more than
-        one rule.
-
-\end{itemize}
-
-\subsection{Example Rule}
-
-\label{example rule in implementation}
-
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the domain in the sender address does
-not have an A, AAAA, or MX DNS entry, i.e.\ mail could not be delivered
-to the sender's address (for full details
-see~\cite{reject-unknown-sender-domain}).
-
-Example log line matched by this rule:
-
-% RFC 3330 says that 192.0.2.0/24 is reserved for example use.
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[192.0.2.1]: 550 5.1.8
-  <foo@example.com>: Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% do not reformat this!
-\begin{tabular}[]{ll}
-
-\textbf{Field}      & \textbf{Value}                                    \\
-name                & Unknown sender domain                             \\
-description         & We do not accept mail from unknown domains        \\
-restriction\_name   & reject\_unknown\_sender\_domain                   \\
-postfix\_action     & REJECTED                                          \\
-program             & \daemon{smtpd}                                    \\
-regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                    & \verb!Sender address rejected: Domain not found;! \\
-                    & \verb!from=<\k<sender>> to=<(__RECIPIENT__)> !    \\
-                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-result\_data        &                                                   \\
-connection\_data    &                                                   \\
-action              & DELIVERY\_REJECTED                                \\
-hits                & 0                                                 \\
-hits\_total         & 0                                                 \\
-priority            & 0                                                 \\
-%cluster\_group      & 400                                               \\
-
-\end{tabular}
-
-\vspace{1em}
-
-Fields matched by the regex will be automatically saved to the results and
-connections tables.
-
-\begin{description}
-
-    \item [name, description, restriction\_name, and postfix\_action:] are
-        not \newline{} used by the parser, they serve to document the rule
-        for the user's benefit.
-
-    \item [program and regex:] If the program in the rule equals the
-        program in the log line, a match using the rule's regex will be
-        attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules do not match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] The action to be executed if the regex matches
-        successfully.  See \sectionref{actions in detail in implementation}
-        for details of the actions implemented by the parser,
-        \sectionref{actions in architecture} for the role of actions in the
-        parser architecture.
-
-    \item [result\_data and connection\_data:] are used to provide XXX
-
-    \item [hits, hits\_total, and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the parser.
-
-%    \item [cluster\_group] The cluster\_group attribute is used by the
-%        Decision Tree algorithm described in a separate document; the
-%        parser does not use it in any way.
-
-\end{description}
-
-\subsection{Creating New Rules}
-
-\label{creating new rules in implementation}
-
-The log files produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different
-\acronym{DNSBL} services, policy servers, or custom rejection messages.  To
-facilitate parsing new log lines, the parser's design separates parsing
-rules from parsing actions: adding new actions can be difficult, but adding
-new rules to parse new rejection messages is trivial, and occurs much more
-frequently.  A program is provided to ease the process of creating new
-rules from unparsed log lines, based on the algorithm developed by Risto
-Vaarandi for his \acronym{SLCT}~\cite{slct-paper}.  The differences between
-the two algorithms will be outlined as part of the general explanation
-below.
-
-The core of the \acronym{SLCT} algorithm is quite simple: log lines are
-generally created by substituting variable words into a fixed pattern, and
-analysis of the frequency with which each word occurs can be used to
-determine whether the word is variable or part of the fixed pattern.  This
-classification can be used to group similar log lines and generate a regex
-to match each group of log lines.
-
-There are 4 steps in the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained when writing rules and performs a large number of
-        substitutions on the input log lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \acronym{IP} addresses, the
-        standard start of rejection messages, etc.) with regex keywords
-        that the parser will expand when it loads the rule (see the
-        regex entry in \sectionref{rule attributes}).  The purpose of
-        this step is to utilise existing knowledge to create more accurate
-        regexes; it replaces a large number of variable words with fixed
-        words, improving the subsequent classification of words as fixed or
-        variable.  The new log lines are written to a temporary file, which
-        all subsequent stages use instead of the original input file.
-
-        In the original algorithm the purpose of the pre-processing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash~\cite{hash-functions}\glsadd{hash} from a
-        small range of values for each word of each log line, incrementing
-        a counter for each hash.  The counters will be used in the next
-        pass to filter out words: if the word's hash does not have a high
-        frequency, the word itself cannot have a high frequency, so there
-        is no need to maintain a counter for it, reducing the number of
-        counters required and thus the program's memory consumption.
-
-    \item [Calculate word frequencies]  The position of words within a log
-        line is important: a word occurring in two log lines does not
-        indicate similarity unless it occupies the same position within
-        both log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two or more words rather
-        than one.  This will alter the position of subsequent words, so a
-        word occurring in different positions in two log lines
-        \textit{may\/} indicate similarity, but the algorithm does not
-        attempt to deal with this possibility.}  The algorithm maintains a
-        counter for each \texttt{(word, word's position within the log
-        line)} tuple, incrementing it each time that word occurs in that
-        position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous step has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm at a cost of increased CPU
-        usage.
-
-        As time goes on the amount of memory typically available increases
-        and the requirement to reduce memory requirements decreases, hence
-        the modified algorithm omits the hashing step and maintains
-        counters for all tokens.  In addition most of the infrequently
-        occurring words will have been substituted with keywords during the
-        first step, vastly reducing the number of tuples to maintain
-        counters for.
-
-    \item [Classify words based on their frequency]  The frequency of every
-        tuple \texttt{(word, word's position within the log line)} is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be misclassified as a fixed term, but that should be noticed
-        by the user when reviewing the new regexes.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a regex
-        matching the log line, and a counter is maintained for each
-        regex.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled regexes are collapsed to a single \texttt{.+}; any
-        resulting duplicate regexes are combined, and their counters
-        added together.  If a regex's counter is lower than the
-        threshold supplied by the user the regex is discarded; this
-        second threshold is independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of log lines is a good starting point.  The new regexes are
-        displayed for the user to add to the database, either as new rules
-        or merged into the regexes of existing rules; the counter for
-        each regex is also displayed, giving the user an indication of
-        how many of the log lines that regex should match.  Discarding
-        regexes will result in some of the log lines not being matched;
-        when the rules have been augmented with the new regexes, the
-        original log files should be parsed again, and any remaining
-        unparsed log lines used as input to this utility.
-
-\end{description}
-
-XXX MERGE logs2regexes WITH check-regexes.
-
-A second utility is also provided that reads a list of new regexes and
-the input given to the first utility.  It tries to match each input log
-line against each regex, counting the number of log lines that match
-each regex, warning the user if an input log line is matched by more
-than one regex, and additionally warning if an input log line is not
-matched by any regex.  It displays a summary of how many input log lines
-each regex matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the regexes
-produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by regex, so the user can tweak the regexes if
-required.
-
-These utilities are not intended to create perfect regexes, but they
-greatly reduce the effort required to parse new or different log lines.
+\end{boldeqlist}
+
+\section{Framework}
+
+XXX TO BE WRITTEN\@.
 
@@
 
-\section{Framework}
-
-XXX TO BE WRITTEN\@.
-
-\section{Parser Flow Chart}
-
-\label{flow-chart}
-
-XXX MIGHT NEED A SMALL INTRO\@.
-
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \acronym{SMTP} or local submission.
-
-    \item If the mail is rejected, save all data and finish.
-
-    \item Follow the progress of the accepted mail until it is either
-        delivered, bounced, or deleted, then save all data, and finish.
-
-\end{enumerate}
-
-\Figureref{flow chart image} shows the most common paths the data
-representing a mail can take through the parser; the complications
-described in \sectionref{complications} are excluded for the sake of
-clarity.
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart}{flow chart
-image}
-
-\label{mail-enters-the-system}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
-forwarding, or internally generated by Postfix.  Local submission is the
-simplest case: a queueid is assigned immediately and the sender address is
-logged (action: PICKUP\@; flowchart:~2).  Re-injection due to forwarding
-sadly lacks explicit log lines of its own; it is explained in
-\sectionref{Re-injected mails}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics as described in
-\sectionref{identifying-bounce-notifications}.  Bounce notifications are
-the primary example of internally generated mails, though there may be
-other types.  Postfix may generate mails to the administrator when it
-encounters configuration errors, but such mails are presumably rare.
-
-\acronym{SMTP} is more complicated than local submission:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \acronym{IP} address or hostname, HELO hostname,
-        etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
-        more mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \acronym{SMTP} commands the data will be
-        saved to the database; if not there will not be any data to save
-        (any mails accepted will already have been cloned so their data is
-        in another data structure).
-
-    \item If one or more mails were accepted there will be more log lines
-        for those mails later, see \sectionref{mail-delivery}.
-
-\end{enumerate}
-
-\label{mail-delivery}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_DATA\@; flowchart:~9).
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
-
-\end{enumerate}
-
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
+\section{Rules}
+
+\label{rules in implementation}
+
+XXX NEED AN INTRODUCTION\@.
+
+XXX ONCE THIS PARAGRAPH HAS BEEN INTEGRATED AND EXPANDED UPDATE THE
+REFERENCE IN ARCHITECTURE->RULES->ATTACHING CONDITIONS TO RULES\@.
+Each log line contains the name of the Postfix component that produced it,
+and this information is used when determining which rules should be used to
+parse each log line (see \sectionref{rules in implementation} for details).
+
+\subsection{Overlapping Rules}
+
+\label{overlapping rules in implementation}
+
+The parser does not try to detect overlapping rules; that responsibility is
+left to the author of the rules.  Unintentionally overlapping rules lead to
+inconsistent parsing and data extraction because the order in which rules
+are tried against each line may change between log files, and the first
+matching rule wins.  Overlapping rules are frequently a requirement,
+allowing a more specific rule to match some log lines and a more general
+rule to match the majority, e.g.\ separating \acronym{SMTP} delivery to
+specific sites from \acronym{SMTP} delivery to the rest of the world.  The
+algorithm provides a mechanism for ordering overlapping rules: the priority
+field in each rule (defaults to zero).  Negative priorities may be useful
+for catchall rules.
+
+Detecting overlapping rules is difficult, but the following approaches may
+be helpful:
+
+\begin{itemize}
+
+    \item Sort by regex and visually inspect the list, e.g.\ with
+        \acronym{SQL} similar to: \textbf{select regex from rules order by
+        regex;}
+
+    \item Compare the results of parsing using sorted, shuffled, and
+        reversed rules (described in \sectionref{rule ordering for
+        efficiency}).  Parse several log files using optimal ordering, then
+        dump a textual representation of the rules, connections, and
+        results tables.  Repeat with shuffled and reversed ordering,
+        starting with a fresh database.  If there are no overlapping rules
+        the tables from each run will be identical; differences indicate
+        overlapping rules.  The rules that overlap can be determined by
+        examining the differences in the tables: each result contains a
+        reference to the rule which created it, if the references differ
+        between runs the two rules referenced in the differing results
+        overlap.  Unfortunately this method cannot prove the absence of
+        overlapping rules; it can detect overlapping rules, but only if
+        there are log lines in the log files that are matched by more than
+        one rule.
+
+\end{itemize}
+
+\subsection{Example Rule}
+
+\label{example rule in implementation}
+
+This example rule matches the message logged by Postfix when it rejects
+mail from a sender address because the domain in the sender address does
+not have an A, AAAA, or MX DNS entry, i.e.\ mail could not be delivered
+to the sender's address (for full details
+see~\cite{reject-unknown-sender-domain}).
+
+Example log line matched by this rule:
+
+% RFC 3330 says that 192.0.2.0/24 is reserved for example use.
+
+\begin{verbatim}
+NOQUEUE: reject: RCPT from example.com[192.0.2.1]: 550 5.1.8
+  <foo@example.com>: Sender address rejected: Domain not found;
+  from=<foo@example.com> to=<info@example.net>
+  proto=SMTP helo=<smtp.example.com>
+\end{verbatim}
+
+% do not reformat this!
+\begin{tabular}[]{ll}
+
+\textbf{Field}      & \textbf{Value}                                    \\
+name                & Unknown sender domain                             \\
+description         & We do not accept mail from unknown domains        \\
+restriction\_name   & reject\_unknown\_sender\_domain                   \\
+postfix\_action     & REJECTED                                          \\
+program             & \daemon{smtpd}                                    \\
+regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
+                    & \verb!Sender address rejected: Domain not found;! \\
+                    & \verb!from=<\k<sender>> to=<(__RECIPIENT__)> !    \\
+                    & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
+result\_data        &                                                   \\
+connection\_data    &                                                   \\
+action              & DELIVERY\_REJECTED                                \\
+hits                & 0                                                 \\
+hits\_total         & 0                                                 \\
+priority            & 0                                                 \\
+%cluster\_group      & 400                                               \\
+
+\end{tabular}
+
+\vspace{1em}
+
+Fields matched by the regex will be automatically saved to the results and
+connections tables.
+
+\begin{description}
+
+    \item [name, description, restriction\_name, and postfix\_action:] are
+        not \newline{} used by the parser, they serve to document the rule
+        for the user's benefit.
+
+    \item [program and regex:] If the program in the rule equals the
+        program in the log line, a match using the rule's regex will be
+        attempted against the log line; if the match is successful the
+        action will be executed, if not the next rule will be tried.  If
+        the program-specific rules do not match the log line, the parser
+        will fall back to generic rules; if those rules are unsuccessful a
+        warning will be issued.
+
+    \item [action:] The action to be executed if the regex matches
+        successfully.  See \sectionref{actions in detail in implementation}
+        for details of the actions implemented by the parser,
+        \sectionref{actions in architecture} for the role of actions in the
+        parser architecture.
+
+    \item [result\_data and connection\_data:] are used to provide XXX
+
+    \item [hits, hits\_total, and priority:] hits and priority are used in
+        ordering the rules (see \sectionref{rule ordering for efficiency});
+        hits is set to the number of successful matches at the end of the
+        parsing run; hits\_total is the sum of hits over every parsing run,
+        but is otherwise unused by the parser.
+
+%    \item [cluster\_group] The cluster\_group attribute is used by the
+%        Decision Tree algorithm described in a separate document; the
+%        parser does not use it in any way.
+
+\end{description}
+
+\subsection{Creating New Rules}
+
+\label{creating new rules in implementation}
+
+The log files produced by Postfix differ from installation to installation,
+because administrators have the freedom to choose the subset of available
+restrictions which suits their needs, including using different
+\acronym{DNSBL} services, policy servers, or custom rejection messages.  To
+facilitate parsing new log lines, the parser's design separates parsing
+rules from parsing actions: adding new actions can be difficult, but adding
+new rules to parse new rejection messages is trivial, and occurs much more
+frequently.  A program is provided to ease the process of creating new
+rules from unparsed log lines, based on the algorithm developed by Risto
+Vaarandi for his \acronym{SLCT}~\cite{slct-paper}.  The differences between
+the two algorithms will be outlined as part of the general explanation
+below.
+
+The core of the \acronym{SLCT} algorithm is quite simple: log lines are
+generally created by substituting variable words into a fixed pattern, and
+analysis of the frequency with which each word occurs can be used to
+determine whether the word is variable or part of the fixed pattern.  This
+classification can be used to group similar log lines and generate a regex
+to match each group of log lines.
+
+There are 4 steps in the algorithm:
+
+\begin{description}
+
+    \item [Pre-process the file]  The new algorithm leverages the knowledge
+        gained when writing rules and performs a large number of
+        substitutions on the input log lines, replacing commonly occurring
+        variable terms (e.g.\ email addresses, \acronym{IP} addresses, the
+        standard start of rejection messages, etc.) with regex keywords
+        that the parser will expand when it loads the rule (see the
+        regex entry in \sectionref{rule attributes}).  The purpose of
+        this step is to utilise existing knowledge to create more accurate
+        regexes; it replaces a large number of variable words with fixed
+        words, improving the subsequent classification of words as fixed or
+        variable.  The new log lines are written to a temporary file, which
+        all subsequent stages use instead of the original input file.
+
+        In the original algorithm the purpose of the pre-processing stage
+        was to reduce the memory consumption of the program.  In the first
+        pass it generated a hash~\cite{hash-functions}\glsadd{hash} from a
+        small range of values for each word of each log line, incrementing
+        a counter for each hash.  The counters will be used in the next
+        pass to filter out words: if the word's hash does not have a high
+        frequency, the word itself cannot have a high frequency, so there
+        is no need to maintain a counter for it, reducing the number of
+        counters required and thus the program's memory consumption.
+
+    \item [Calculate word frequencies]  The position of words within a log
+        line is important: a word occurring in two log lines does not
+        indicate similarity unless it occupies the same position within
+        both log lines.\footnote{If a variable term within a line contains
+        spaces, it will appear to the algorithm as two or more words rather
+        than one.  This will alter the position of subsequent words, so a
+        word occurring in different positions in two log lines
+        \textit{may\/} indicate similarity, but the algorithm does not
+        attempt to deal with this possibility.}  The algorithm maintains a
+        counter for each \texttt{(word, word's position within the log
+        line)} tuple, incrementing it each time that word occurs in that
+        position.
+
+        The original algorithm only maintains counters for words whose hash
+        result from the previous step has a high frequency; this reduces
+        the number of counters maintained by the algorithm, reducing the
+        memory requirements of the algorithm at a cost of increased CPU
+        usage.
+
+        As time goes on the amount of memory typically available increases
+        and the requirement to reduce memory requirements decreases, hence
+        the modified algorithm omits the hashing step and maintains
+        counters for all tokens.  In addition most of the infrequently
+        occurring words will have been substituted with keywords during the
+        first step, vastly reducing the number of tuples to maintain
+        counters for.
+
+    \item [Classify words based on their frequency]  The frequency of every
+        tuple \texttt{(word, word's position within the log line)} is
+        checked: if its frequency is greater than the threshold supplied by
+        the user (1\% of all log lines is generally a good starting point)
+        it is classified as a fixed word, otherwise it is classified as a
+        variable term.  If a variable term appears sufficiently often it
+        will be misclassified as a fixed term, but that should be noticed
+        by the user when reviewing the new regexes.  Variable terms are
+        replaced by \texttt{.+}, which means to match zero or more of any
+        character.
+
+    \item [Build regexes]  The words are reassembled to produce a regex
+        matching the log line, and a counter is maintained for each
+        regex.  Contiguous sequences of \texttt{.+} in the newly
+        reassembled regexes are collapsed to a single \texttt{.+}; any
+        resulting duplicate regexes are combined, and their counters
+        added together.  If a regex's counter is lower than the
+        threshold supplied by the user the regex is discarded; this
+        second threshold is independent of the threshold used to
+        differentiate between fixed and variable words, but once again 1\%
+        of log lines is a good starting point.  The new regexes are
+        displayed for the user to add to the database, either as new rules
+        or merged into the regexes of existing rules; the counter for
+        each regex is also displayed, giving the user an indication of
+        how many of the log lines that regex should match.  Discarding
+        regexes will result in some of the log lines not being matched;
+        when the rules have been augmented with the new regexes, the
+        original log files should be parsed again, and any remaining
+        unparsed log lines used as input to this utility.
+
+\end{description}
+
+XXX MERGE logs2regexes WITH check-regexes.
+
+A second utility is also provided that reads a list of new regexes and
+the input given to the first utility.  It tries to match each input log
+line against each regex, counting the number of log lines that match
+each regex, warning the user if an input log line is matched by more
+than one regex, and additionally warning if an input log line is not
+matched by any regex.  It displays a summary of how many input log lines
+each regex matched, comparing it to the expected number of matches; this
+provides the user with an easy method of checking if the regexes
+produced by the first utility are correctly matching the input log lines
+they are based upon.  A future version of this utility will also group
+input log lines by regex, so the user can tweak the regexes if
+required.
+
+These utilities are not intended to create perfect regexes, but they
+greatly reduce the effort required to parse new or different log lines.
 

### 2009-03-11 doc/logparser-introduction.tex
Corrections for the assumptions and database sections of the implementation chapter.

@@
 parsing and the results gleaned from parsing.  The database schema used is
-described in \sectionref{database schema}, explaining in detail the tables
-used for storing the data gleaned from the log files and the table that
-stores the rules.
+described in \sectionref{database}, explaining in detail the tables used
+for storing the data gleaned from the log files and the table that stores
+the rules.
 

### 2009-03-11 doc/logparser-architecture.tex
Improved Makefile to rerun latex if the list of figures or tables changes. Rearranged some parts of the rules section of the implementation chapter, and made corrections to it from proofreading.

@@
 recognition is successful, but implementations are free to add any other
-attributes they require; \sectionref{rule attributes} describes the
-attributes used in \parsername{}, and some generally useful attributes will
-be discussed later in this section.
+attributes they require; \sectionref{rules table} describes the attributes
+used in \parsername{}, and some generally useful attributes will be
+discussed later in this section.
 

### 2009-03-11 doc/logparser-implementation.tex
Improved Makefile to rerun latex if the list of figures or tables changes. Rearranged some parts of the rules section of the implementation chapter, and made corrections to it from proofreading.

@@
 
-\label{rule attributes}
+\label{rules table}
 
@@
 
-%\begin{eqlist}[\def\makelabel#1{\bfseries#1:}]
-
-XXX IMPROVE HOW THIS LOOKS
-
-\begin{eqlist}[\def\makelabel#1{\bfseries#1:}]
+\begin{boldeqlist}
 
@@
 
-\end{eqlist}
-
-\subsection{Actions in Detail}
+\end{boldeqlist}
+
+\subsection{Actions in the \parsernamelong{}}
 
@@
 
-XXX ONCE THIS PARAGRAPH HAS BEEN INTEGRATED AND EXPANDED UPDATE THE
-REFERENCE IN ARCHITECTURE->RULES->ATTACHING CONDITIONS TO RULES\@.
-Each log line contains the name of the Postfix component that produced it,
-and this information is used when determining which rules should be used to
-parse each log line (see \sectionref{rules in implementation} for details).
-
-\subsection{Overlapping Rules}
-
-\label{overlapping rules in implementation}
-
-The parser does not try to detect overlapping rules; that responsibility is
-left to the author of the rules.  Unintentionally overlapping rules lead to
-inconsistent parsing and data extraction because the order in which rules
-are tried against each line may change between log files, and the first
-matching rule wins.  Overlapping rules are frequently a requirement,
-allowing a more specific rule to match some log lines and a more general
-rule to match the majority, e.g.\ separating \acronym{SMTP} delivery to
-specific sites from \acronym{SMTP} delivery to the rest of the world.  The
-algorithm provides a mechanism for ordering overlapping rules: the priority
-field in each rule (defaults to zero).  Negative priorities may be useful
-for catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by regex and visually inspect the list, e.g.\ with
-        \acronym{SQL} similar to: \textbf{select regex from rules order by
-        regex;}
-
-    \item Compare the results of parsing using sorted, shuffled, and
-        reversed rules (described in \sectionref{rule ordering for
-        efficiency}).  Parse several log files using optimal ordering, then
-        dump a textual representation of the rules, connections, and
-        results tables.  Repeat with shuffled and reversed ordering,
-        starting with a fresh database.  If there are no overlapping rules
-        the tables from each run will be identical; differences indicate
-        overlapping rules.  The rules that overlap can be determined by
-        examining the differences in the tables: each result contains a
-        reference to the rule which created it, if the references differ
-        between runs the two rules referenced in the differing results
-        overlap.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the log files that are matched by more than
-        one rule.
-
-\end{itemize}
+The rules are responsible for classifying each log line and specifying the
+correct action to be invoked.  For any parser implemented using this
+architecture, the rules are the most visible component and most likely to
+be modified by users.  Rules need to be as simple as possible so that users
+can easily modify or add to them, but the implementation must balance that
+simplicity with the need to provide enough flexibility and power to
+successfully parse the inputs.
+
+The theoretical aspects of this component are covered in detail in
+\sectionref{rules in architecture}; this section is concerned with the
+practical aspects of how rules are implemented and used in \parsername{}.
+The structure of the rules table has already been documented in
+\sectionref{rules table}; that abstract description will not be repeated
+here, but should be fleshed out by the concrete example rule in
+\sectionref{example rule in implementation}.  The process of creating new
+rules from unparsed log lines is dealt with in \sectionref{creating new
+rules in implementation}; the implementation of the utility supplied with
+\parsername{} to create regexes from unparsed log lines is also described
+in that section.  The final portion of this section provides some
+suggestions for how to detect overlapping rules.
 
@@
 
-This example rule matches the message logged by Postfix when it rejects
-mail from a sender address because the domain in the sender address does
-not have an A, AAAA, or MX DNS entry, i.e.\ mail could not be delivered
-to the sender's address (for full details
-see~\cite{reject-unknown-sender-domain}).
-
-Example log line matched by this rule:
+The example rule in \tableref{Example rule in implementation table}
+recognises the log line resulting from Postfix rejecting a delivery attempt
+because the domain in the sender address does not have an A, AAAA,
+or MX DNS entry, i.e.\ mail could not be delivered to the sender's address
+(for full details see~\cite{reject-unknown-sender-domain}).
+
+Example log line recognised by this rule:
 
@@
 \begin{verbatim}
-NOQUEUE: reject: RCPT from example.com[192.0.2.1]: 550 5.1.8
-  <foo@example.com>: Sender address rejected: Domain not found;
+NOQUEUE: reject: RCPT from smtp.example.com[192.0.2.1]:
+  550 5.1.8 <foo@example.com>:
+  Sender address rejected: Domain not found;
   from=<foo@example.com> to=<info@example.net>
@@
 % do not reformat this!
-\begin{tabular}[]{ll}
+\begin{table}[ht]
+
+    \caption{Example rule}
+    \empty{}\label{Example rule in implementation table}
+    \begin{tabular}{ll}
 
@@
 
-\vspace{1em}
-
-Fields matched by the regex will be automatically saved to the results and
-connections tables.
+\end{table}
+
+The fields in \tableref{Example rule in implementation table} are used as
+follows:
 
@@
         will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.
-
-    \item [action:] The action to be executed if the regex matches
-        successfully.  See \sectionref{actions in detail in implementation}
-        for details of the actions implemented by the parser,
+        warning will be issued.  Data captured by the regex (sender,
+        recipient, etc.) will be automatically saved to the results and
+        connections tables.
+
+    \item [action:] The action to be invoked when the rule recognises a log
+        line.  See \sectionref{actions in detail in implementation} for
+        details of the actions implemented by \parsername{}, and
         \sectionref{actions in architecture} for the role of actions in the
@@
 
-    \item [result\_data and connection\_data:] are used to provide XXX
-
-    \item [hits, hits\_total, and priority:] hits and priority are used in
-        ordering the rules (see \sectionref{rule ordering for efficiency});
-        hits is set to the number of successful matches at the end of the
-        parsing run; hits\_total is the sum of hits over every parsing run,
+    \item [result\_data and connection\_data:] are used to provide data not
+        present in the log line, and are unused in this rule.
+
+    \item [hits, hits\_total, and priority:] hits and priority are used
+        when ordering the rules for more efficient parsing (see
+        \sectionref{rule ordering for efficiency}).  At the end of each
+        parsing run hits is set to the number of log lines recognised by
+        the rule.  Hits\_total is the sum of hits over every parsing run,
         but is otherwise unused by the parser.
-
-%    \item [cluster\_group] The cluster\_group attribute is used by the
-%        Decision Tree algorithm described in a separate document; the
-%        parser does not use it in any way.
 
@@
 \acronym{DNSBL} services, policy servers, or custom rejection messages.  To
-facilitate parsing new log lines, the parser's design separates parsing
-rules from parsing actions: adding new actions can be difficult, but adding
-new rules to parse new rejection messages is trivial, and occurs much more
-frequently.  A program is provided to ease the process of creating new
-rules from unparsed log lines, based on the algorithm developed by Risto
-Vaarandi for his \acronym{SLCT}~\cite{slct-paper}.  The differences between
-the two algorithms will be outlined as part of the general explanation
-below.
-
-The core of the \acronym{SLCT} algorithm is quite simple: log lines are
-generally created by substituting variable words into a fixed pattern, and
-analysis of the frequency with which each word occurs can be used to
-determine whether the word is variable or part of the fixed pattern.  This
-classification can be used to group similar log lines and generate a regex
-to match each group of log lines.
-
-There are 4 steps in the algorithm:
+facilitate parsing new log lines, the architecture separates rules from
+actions: adding new actions can be difficult, but adding new rules to
+recognise new log lines is trivial, and occurs much more frequently.
+
+XXX DESCRIBE HOW TO GO ABOUT ADDING A NEW RULE\@.
+
+A program is provided with \parsername{} to ease the process of creating
+new rules from unrecognised log lines, based on the algorithm developed by
+Risto Vaarandi for his \acronym{SLCT}~\cite{slct-paper}.  The differences
+between the two algorithms will be outlined as part of the general
+explanation below.  The core of the \acronym{SLCT} algorithm is quite
+simple: log lines are generally created by substituting variable words into
+a fixed pattern, and analysis of the frequency with which each word occurs
+can be used to determine whether the word is variable or part of the fixed
+pattern.  This classification can be used to group similar log lines and
+generate a regex to match each group of log lines.  There are 4 steps in
+the algorithm:
 
@@
 
-    \item [Pre-process the file]  The new algorithm leverages the knowledge
-        gained when writing rules and performs a large number of
-        substitutions on the input log lines, replacing commonly occurring
-        variable terms (e.g.\ email addresses, \acronym{IP} addresses, the
-        standard start of rejection messages, etc.) with regex keywords
-        that the parser will expand when it loads the rule (see the
-        regex entry in \sectionref{rule attributes}).  The purpose of
-        this step is to utilise existing knowledge to create more accurate
-        regexes; it replaces a large number of variable words with fixed
-        words, improving the subsequent classification of words as fixed or
-        variable.  The new log lines are written to a temporary file, which
-        all subsequent stages use instead of the original input file.
+    \item [Pre-process the file.]  The new algorithm leverages the
+        knowledge gained when writing rules and developing \parsername{};
+        it performs a large number of substitutions on the input log lines,
+        replacing commonly occurring variable terms (e.g.\ email addresses,
+        \acronym{IP} addresses, the standard start of rejection messages)
+        with keywords that \parsername{} will expand when it loads the new
+        rule.  The purpose of this step is to utilise existing knowledge to
+        create more accurate regexes; it replaces a large number of
+        variable words with fixed words, improving the subsequent
+        classification of words as fixed or variable.  The altered log
+        lines are written to a temporary file, which subsequent stages will
+        use instead of the original input file.
 
@@
         was to reduce the memory consumption of the program.  In the first
-        pass it generated a hash~\cite{hash-functions}\glsadd{hash} from a
-        small range of values for each word of each log line, incrementing
+        pass it generates a hash~\cite{hash-functions}\glsadd{hash} (from a
+        small range of values) for each word of each log line, incrementing
         a counter for each hash.  The counters will be used in the next
@@
 
-    \item [Calculate word frequencies]  The position of words within a log
+    \item [Calculate word frequencies.]  The position of words within a log
         line is important: a word occurring in two log lines does not
-        indicate similarity unless it occupies the same position within
-        both log lines.\footnote{If a variable term within a line contains
-        spaces, it will appear to the algorithm as two or more words rather
-        than one.  This will alter the position of subsequent words, so a
-        word occurring in different positions in two log lines
-        \textit{may\/} indicate similarity, but the algorithm does not
-        attempt to deal with this possibility.}  The algorithm maintains a
-        counter for each \texttt{(word, word's position within the log
-        line)} tuple, incrementing it each time that word occurs in that
-        position.
-
-        The original algorithm only maintains counters for words whose hash
-        result from the previous step has a high frequency; this reduces
-        the number of counters maintained by the algorithm, reducing the
-        memory requirements of the algorithm at a cost of increased CPU
-        usage.
+        indicate similarity unless it occupies the same position in both
+        log lines.  If a variable term within a line contains
+        spaces, it will appear to the algorithm as more than one word.
+        This will alter the position of subsequent words, so a word
+        occurring in different positions in two log lines \textit{may\/}
+        indicate similarity, but the algorithm does not attempt to deal
+        with this possibility.  The algorithm maintains a counter for each
+        \texttt{(word, word's position within the log line)} tuple,
+        incrementing it each time that word occurs in that position.
+
+        The original algorithm hashes each word and checks that result's
+        counter from the previous pass; if that counter has a high
+        frequency a separate counter will be maintained for this word in
+        this position.  This reduces the number of counters maintained by
+        the algorithm in this step, reducing the memory requirements of the
+        algorithm at a cost of increased CPU usage.
 
         As time goes on the amount of memory typically available increases
-        and the requirement to reduce memory requirements decreases, hence
-        the modified algorithm omits the hashing step and maintains
-        counters for all tokens.  In addition most of the infrequently
-        occurring words will have been substituted with keywords during the
-        first step, vastly reducing the number of tuples to maintain
-        counters for.
-
-    \item [Classify words based on their frequency]  The frequency of every
-        tuple \texttt{(word, word's position within the log line)} is
+        and the need to reduce memory requirements decreases, so the
+        modified algorithm omits the hashing step and maintains counters
+        for all tokens.  In addition most of the infrequently occurring
+        words will have been substituted with keywords during the first
+        step, vastly reducing the number of tuples to maintain counters
+        for.
+
+    \item [Classify words based on their frequency.]  The frequency of
+        every tuple \texttt{(word, word's position within the log line)} is
         checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point)
+        the user (1\% of all log lines is generally a good starting point),
         it is classified as a fixed word, otherwise it is classified as a
@@
         by the user when reviewing the new regexes.  Variable terms are
-        replaced by \texttt{.+}, which means to match zero or more of any
-        character.
-
-    \item [Build regexes]  The words are reassembled to produce a regex
-        matching the log line, and a counter is maintained for each
-        regex.  Contiguous sequences of \texttt{.+} in the newly
-        reassembled regexes are collapsed to a single \texttt{.+}; any
-        resulting duplicate regexes are combined, and their counters
-        added together.  If a regex's counter is lower than the
-        threshold supplied by the user the regex is discarded; this
-        second threshold is independent of the threshold used to
-        differentiate between fixed and variable words, but once again 1\%
-        of log lines is a good starting point.  The new regexes are
-        displayed for the user to add to the database, either as new rules
-        or merged into the regexes of existing rules; the counter for
-        each regex is also displayed, giving the user an indication of
-        how many of the log lines that regex should match.  Discarding
-        regexes will result in some of the log lines not being matched;
-        when the rules have been augmented with the new regexes, the
-        original log files should be parsed again, and any remaining
+        replaced by \texttt{.+} to match one or more of any character.
+
+    \item [Build regexes.]  The words are reassembled to produce a regex
+        matching the log line, and a counter is maintained for each regex.
+        Contiguous sequences of \texttt{.+} in the newly reassembled
+        regexes are collapsed to a single \texttt{.+}; any resulting
+        duplicate regexes are combined, and their counters added together.
+        If a regex's counter is lower than the threshold supplied by the
+        user the regex is discarded; this second threshold is independent
+        of the threshold used to differentiate between fixed and variable
+        words, but once again 1\% of log lines is a good starting point.
+        The new regexes are displayed for the user to add to the database,
+        either as new rules or merged into the regexes of existing rules;
+        the counter for each regex is also displayed, giving the user an
+        indication of how many of the log lines that regex should match.
+        Discarding regexes will result in some of the log lines not being
+        matched; when the rules have been augmented with the new regexes,
+        the original log files should be parsed again, and any remaining
         unparsed log lines used as input to this utility.
@@
 produced by the first utility are correctly matching the input log lines
-they are based upon.  A future version of this utility will also group
-input log lines by regex, so the user can tweak the regexes if
-required.
+they are based upon.
 
@@
 greatly reduce the effort required to parse new or different log lines.
+
+\subsection{Overlapping Rules}
+
+\label{overlapping rules in implementation}
+
+\parsername{} does not try to detect overlapping rules; that responsibility
+is left to the author of the rules.  The advantages and difficulties of
+overlapping rules have already been addressed in \sectionref{overlapping
+rules in architecture} and will not be repeated here.  \parsername{}
+provides a mechanism for ordering overlapping rules: the priority field in
+each rule (defaulting to zero); negative priorities may be useful for
+catchall rules.
+
+Detecting overlapping rules is difficult, but the following approaches may
+be helpful:
+
+\begin{itemize}
+
+    \item Sort by program and regex, then visually inspect the list, e.g.\
+        with \acronym{SQL} similar to:      \newline{}
+        \verb!SELECT program, regex!        \newline{}
+        \verb!    FROM rules!               \newline{}
+        \verb!    ORDER BY program, regex;! \newline{}
+        The rules are sorted first by program, then by regex, because rules
+        with overlapping regexes will not overlap if their programs are
+        different.  Note that this does not properly deal with rules whose
+        program is \texttt{*}; those rules will be used on all log lines
+        that have not been recognised by program-specific rules.
+
+    \item Compare the results of parsing using different rule orderings
+        (described in \sectionref{rule ordering for efficiency}).  Parse
+        several log files using optimal ordering, then dump a textual
+        representation of the rules, connections, and results tables.
+        Repeat with shuffled and reversed ordering, starting with a fresh
+        database.  If there are no overlapping rules the tables from each
+        run will be identical; differences indicate overlapping rules.  The
+        rules that overlap can be determined by examining the differences
+        in the tables: each result contains a reference to the rule which
+        created it, which will change if that rule overlaps with another.
+        Unfortunately this method cannot prove the absence of overlapping
+        rules; it can detect overlapping rules, but only if there are log
+        lines in the log files that are recognised by more than one rule.
+
+\end{itemize}
 

### 2009-03-11 doc/logparser-other-parsers-review.tex
Improved Makefile to rerun latex if the list of figures or tables changes. Rearranged some parts of the rules section of the implementation chapter, and made corrections to it from proofreading.

@@
     \item The parsers reviewed all produce a report of varying complexity
-        and detail, whereas \acronym{PLP} does not; it extracts data and
+        and detail, whereas \parsername{} does not; it extracts data and
         leaves generation of reports from the data to other programs.

### 2009-03-12 doc/logparser-architecture.tex
Corrections to the actions section of the implementation chapter. Removed a large chunk explaining postfix_action; remaining references will need to be excised over time. Renamed some labels, resulting in some churn without any real effect; in particular some long paragraphs were reformatted.

@@
 in Yet More Aborted Delivery Attempts
-(\sectionref{yet-more-aborted-delivery-attempts}).  More intrusive changes
+(\sectionref{yet more aborted delivery attempts}).  More intrusive changes
 are more likely to introduce a bug, so reducing the extent of the changes
@@
 some of those deficiencies were rectified by later versions of Postfix,
-e.g.\ identifying bounce notifications
-(\sectionref{identifying-bounce-notifications}); others were due to the
-vagaries of process scheduling, client behaviour, and administrative
-actions.  All were successfully accommodated in \parsername{}: adding new
-actions was enough to overcome several of the complications; others
-required modifications to a single existing action to work around a
-difficulty; the remainder were resolved by adapting existing actions to
-cooperate and exchange extra data (via the framework), changing their
-behaviour as appropriate based on that extra data.  Every architecture
-should aim to make the easy things easy and the hard things possible; the
-successful implementation of \parsername{} demonstrates that this
-architecture achieves that aim.
+e.g.\ identifying bounce notifications (\sectionref{identifying bounce
+notifications}); others were due to the vagaries of process scheduling,
+client behaviour, and administrative actions.  All were successfully
+accommodated in \parsername{}: adding new actions was enough to overcome
+several of the complications; others required modifications to a single
+existing action to work around a difficulty; the remainder were resolved by
+adapting existing actions to cooperate and exchange extra data (via the
+framework), changing their behaviour as appropriate based on that extra
+data.  Every architecture should aim to make the easy things easy and the
+hard things possible; the successful implementation of \parsername{}
+demonstrates that this architecture achieves that aim.
 

### 2009-03-12 doc/logparser-implementation.tex
Corrections to the actions section of the implementation chapter. Removed a large chunk explaining postfix_action; remaining references will need to be excised over time. Renamed some labels, resulting in some churn without any real effect; in particular some long paragraphs were reformatted.

@@
         reconstruct the actions it has taken.  There are several heuristics
-        used when parsing; see
-        \sectionref{identifying-bounce-notifications},
-        \sectionref{aborted-delivery-attempts}, and \sectionref{pickup
-        logging after cleanup} for details.
+        used when parsing; see \sectionref{identifying bounce
+        notifications}, \sectionref{aborted delivery attempts}, and
+        \sectionref{pickup logging after cleanup} for details.
 
@@
 
-\label{flow-chart}
+\label{flow chart}
 
@@
 
-\label{mail-enters-the-system}
+\label{mail enters the system}
 
@@
 Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics as described in
-\sectionref{identifying-bounce-notifications}.  Bounce notifications are
-the primary example of internally generated mails, though there may be
-other types.  Postfix may generate mails to the administrator when it
-encounters configuration errors, but such mails are presumably rare.
+must be detected using heuristics as described in \sectionref{identifying
+bounce notifications}.  Bounce notifications are the primary example of
+internally generated mails, though there may be other types.  Postfix may
+generate mails to the administrator when it encounters configuration
+errors, but such mails are presumably rare.
 
@@
     \item If one or more mails were accepted there will be more log lines
-        for those mails later, see \sectionref{mail-delivery}.
+        for those mails later, see \sectionref{mail delivery}.
 
@@
 
-\label{mail-delivery}
+\label{mail delivery}
 
@@
 
-    \item [postfix\_action] The action Postfix must have taken to generate
-        this log line; the possible actions are described later in
-        \sectionref{list of postfix actions in implementation}.
-
     \item [program] The Postfix component (e.g.\ \daemon{smtpd}) whose log
@@
 
-The actions reconstruct the journey a mail takes through Postfix.  Details
-of the actions available in the Postfix parser can be found in
-\sectionref{actions in detail in implementation}, and \sectionref{adding
-new actions in implementation} describes the process of adding new actions.
-
-In the Postfix log parser developed for this project there are
+Actions are the component of this architecture responsible for processing
+all of the inputs recognised by rules; in \parsername{} they reconstruct
+the journey each mail takes through Postfix, dealing with all the
+complications and difficulties that arise.  \parsername{} has
 \numberOFactions{} actions and \numberOFrules{} rules, with an uneven
@@
 rules per action}.  Unsurprisingly, the action with the most associated
-rules is \texttt{DELIVERY\_REJECTED}, the action that handles Postfix
-rejecting a mail delivery attempt; it is followed by \texttt{SAVE\_DATA},
-the action responsible for handling informative log lines, supplementing
-the data gathered from other log lines.  The third most common action is,
-perhaps surprisingly, \texttt{UNINTERESTING}: this action does nothing when
-executed, allowing uninteresting log lines to be parsed without causing any
-effects (it does not imply that the input is ungrammatical or unparsed).
-Generally rules specifying the \texttt{UNINTERESTING} action parse log
-lines that are not associated with a specific mail, e.g.\ notices about
-configuration files changing.  The remaining actions have only one or two
-associated rules: some actions are required to address a deficiency in the
-log files, or a complication that arises during parsing;  other actions
-will only ever have one log line variant, e.g.\ all log lines showing that
-a remote client has connected are matched by a single rule and handled by
-the \texttt{CONNECT} action.
-
-Using the \texttt{CONNECT} action as an example: it creates a new data
-structure in memory for the new client connection, saving the data
-extracted by the rule into it; this data will be entered into the database
-when the mail delivery attempt is complete.  If a data structure already
-exists for the new connection it is treated as a symptom of a bug, and the
-action issues a warning containing the full contents of the existing data
-structure, plus the log line that has just been parsed.
-
-Each action is passed the same arguments:
+rules is \texttt{DELIVERY\_REJECTED}, the action that processes Postfix
+rejecting a mail delivery attempt; next is \texttt{SAVE\_DATA}, the action
+responsible for processing informative log lines, supplementing the data
+gathered from other log lines.  The third most common action is, perhaps
+surprisingly, \texttt{UNINTERESTING}: this action does nothing when
+invoked, allowing uninteresting log lines to be parsed without any effects
+(it does not imply that the input is ungrammatical or unparsed).  Generally
+rules specifying the \texttt{UNINTERESTING} action recognise log lines that
+are not associated with a specific mail, e.g.\ notices about configuration
+files changing; these log lines are recognised and processed so that the
+framework can warn about unrecognised log lines, informing the user that
+they need to add to or extend their ruleset.  The remaining actions have
+only one or two associated rules: some are specialised actions required to
+address a deficiency in the log files, or a complication that arises during
+parsing; other actions will only ever have one log line variant, e.g.\ all
+log lines showing that a remote client has connected are recognised by a
+single rule and handled by the \texttt{CONNECT} action.  \parsernames{}
+actions are described in detail in \sectionref{actions in detail in
+implementation}, and \sectionref{adding new actions in implementation}
+covers the process of adding a new action.
+
+\showgraph{build/graph-action-distribution}{Distribution of rules per
+action}{Distribution of rules per action}
+
+\subsection{Actions in the \parsernamelong{}}
+
+\label{actions in detail in implementation}
+
+This section explains the actions found in \parsername{}; it may help to
+revisit the flow chart in \sectionref{flow chart} to see how a mail passes
+from one action to another as its log lines are recognised.  The words
+\textit{mail\/} and \textit{connection\/} are used in the actions
+descriptions below because they are less unwieldy and more specific than
+\textit{state table entry\/}; a connection becomes a mail during the
+\texttt{CLONE} action, which processes Postfix accepting a delivery
+attempt, and the data structure moves from one state table to another
+within the parser.
+
+The complications and difficulties that arose when parsing log files
+generated on a production mail server are documented in
+\sectionref{complications}; some action descriptions refer to specific
+difficulties they address.  The complications are documented in a separate
+section to avoid overwhelming the action descriptions.
+
+If there is enough information in the log line to identify the correct
+connection or mail, each action will save all the data captured by the
+recognising rule's regex.  Each action is passed the same arguments:
 
@@
 
-    \item [rule] The matching rule.
-
-    \item [matches] The data extracted from the line by the rule's regex.
+    \item [rule] The recognising rule.
+
+    \item [data] The data captured from the log line by the rule's regex.
 
@@
 
-        \begin{eqlist}[\def\makelabel#1{\bfseries#1:}]
+        \begin{boldeqlist}
 
@@
             \item [timestamp] The time the line was logged at.
-
-            \item [host] The hostname of the server that logged the line.
 
@@
 
-            \item [text] The remainder of the line.
-
-        \end{eqlist}
+            \item [host] The hostname of the server the line was logged on.
+
+            \item [text] The remainder of the log line, i.e.\ the message
+                logged by the program and recognised by the rule.
 
@@
 
-\subsection{Actions in the \parsernamelong{}}
-
-\label{actions in detail in implementation}
-
-Each action saves all data extracted by the rule's regex, if there is
-enough information in the log line to identify the correct connection to
-save the data to.
+\end{boldeqlist}
 
@@
 
-    \item [BOUNCE] Postfix 2.3 and subsequent versions log the creation of
-        bounce messages.  This action creates a new mail if necessary; if
-        the mail already exists the unknown origin flag will be removed.
-        The action also marks the mail as a bounce notification.  To deal
-        with complication \sectionref{Bounce notification mails delivered
-        before their creation is logged} this action checks a cache of
-        recent bounce mails to avoid creating bogus bounce mails when log
-        lines are out of order.
+    \item [BOUNCE\_CREATED] Postfix 2.3 and subsequent versions log the
+        creation of bounce messages.  This action creates a new mail if
+        necessary; if the mail already exists the unknown origin flag will
+        be removed.  The action also marks the mail as a bounce
+        notification.  To deal with complication \sectionref{Bounce
+        notification mails delivered before their creation is logged} this
+        action checks a cache of recent bounce mails, to avoid creating
+        bogus bounce mails when log lines are out of order.
 
@@
     \item [COMMIT] Enter the data from the mail into the database.  Entry
-        will be postponed if the mail is a child waiting to be tracked.
-        Once entered, the mail will be deleted from the state tables.
-        Deletion will be postponed if the mail is the parent of re-injected
+        will be postponed if the mail is a child waiting to be tracked
+        (\sectionref{Re-injected mails}).  Once entered in the database,
+        the mail will be usually be deleted from the state tables, but
+        deletion will be postponed if the mail is the parent of re-injected
         mail (\sectionref{Re-injected mails}).
 
-    \item [CONNECT] Handle a remote client connecting: create a new state
-        table entry (indexed by \daemon{smtpd} \acronym{pid}) and save both
-        the client hostname and \acronym{IP} address.
+    \item [CONNECT] Handle a remote client connecting: create a new
+        connection indexed by \daemon{smtpd} \acronym{pid}, and save the
+        data captured by the rule (\acronym{IP} address and hostname of the
+        client).  If a connection already exists it is treated as a symptom
+        of a bug, and the action issues a warning containing the full
+        contents of the existing connection plus the log line that has
+        just been parsed.
 
@@
         command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to handle
-        adding the mail to the database.  The complication this action
-        deals with is described fully in \sectionref{Mail deleted before
-        delivery is attempted}.
+        address if required, then invokes the COMMIT action to save the
+        mail to the database.  The complication this action deals with is
+        described fully in \sectionref{Mail deleted before delivery is
+        attempted}.
 
@@
         with a mail if there is a queueid in the log line, or with the
-        connection if not.
-
-    \item [DISCONNECT] Deal with the remote client disconnecting: enter the
-        connection in the database, perform any required cleanup, and
-        delete the connection from the state tables.  This action deals
-        with aborted delivery attempts
-        (\sectionref{aborted-delivery-attempts}).
+        connection if not.  XXX VERY IMPORTANT ACTION\@; EXPAND AND
+        REWRITE\@.
+
+    \item [DISCONNECT] Invoked when the remote client disconnects, it
+        enters the connection in the database, performs any required
+        cleanup, and deletes the connection from the state tables.  This
+        action deals with aborted delivery attempts (\sectionref{aborted
+        delivery attempts}).
 
@@
         be committed the EXPIRY action sets a flag marking the mail as
-        expired; the flag later causes the valid combinations check to be
-        skipped, so the mail will be committed.
+        expired; this flag later causes the valid combinations check to be
+        skipped.  The mail will be committed when a log line triggers the
+        \texttt{COMMIT} action; if the \texttt{COMMIT} action has already
+        run and refused to enter the mail in the database because it failed
+        the valid program combinations check, the \texttt{SAVE\_DATA}
+        action will re-invoke the \texttt{COMMIT} action the next time it
+        is invoked.
+
+    \item [MAIL\_BOUNCED] XXX WRITE THIS\@; INVOKES SAVE\_DATA ACTION\@.
+
+    \item [MAIL\_DISCARDED] XXX WRITE THIS\@; INVOKES COMMIT ACTION\@.
 
@@
         the queue to deliver.  This action is used for both \daemon{qmgr}
-        and \daemon{cleanup} as it needs to deal with out of order log
+        and \daemon{cleanup} because it needs to deal with out of order log
         lines; see \sectionref{discarding cleanup log lines} for details.
@@
 
+    \item [MAIL\_SENT] XXX WRITE THIS\@; INVOKES SAVE\_DATA ACTION\@.
+
     \item [MAIL\_TOO\_LARGE] When a client tries to send a mail larger than
         the local server accepts it will be discarded and the client
-        informed of the problem.  This is handled in exactly the same way
-        as the TIMEOUT action, further details are given there.
+        informed of the problem.  The mail may have been accepted and
+        partially transferred, in which case there's a data structure to
+        dispose of, otherwise there will not be.  See \sectionref{timeouts
+        during data phase} for the gory details; although that describes
+        timeouts the processing is the same for mails that are too large.
 
@@
         entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); otherwise it is
-        created.  The data extracted from the log line is then saved to the
-        state table entry.
+        \sectionref{pickup logging after cleanup}); if it does not exist it
+        will be created.
 
     \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes,\footnote{Possibly other
-        programs are killed also, but the parser is only affected by and
-        interested in \daemon{smtpd} processes exiting.} requiring any
-        active connections to be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [SAVE\_DATA] Find the correct mail based on the queueid in
-        the log line, and save the data extracted by the regex to it.
+        it kills all \daemon{smtpd} processes, requiring any active
+        connections to be cleaned up, entered in the database, and deleted
+        from the state tables.  Postfix probably kills all the other
+        components too, but the parser is only affected by \daemon{smtpd}
+        processes exiting.
+
+    \item [SAVE\_DATA] Every action that can locate the correct entry in
+        the state tables saves any data captured by the recognising rule's
+        regex to it.  The \texttt{SAVE\_DATA} saves data in this way but
+        does not do anything else; it is invoked for log lines that contain
+        useful data but do not require any further processing.
 
@@
         unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database, and deleted from the state
-        tables.
+        be cleaned up, entered in the database if it has sufficient data,
+        and deleted from the state tables.  In some cases the connection
+        will not have enough data to satisfy the database schema, so it
+        cannot be entered into the database; unfortunately this means that
+        some data is lost.
 
     \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal; the active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
+        signal; this is processed in the same way as the
+        \texttt{SMTPD\_DIED} action.
 
     \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations --- after five hours the timer will
+        to deal with unusual situations; after five hours the timer will
         expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts that should occur
-        in the intervening hours: DNS timeouts, timeouts reading data
-        from the client, etc\@.  The active connection for that
+        infrequently, as there are many other timeouts that should occur in
+        the intervening hours, e.g.\ timeouts for DNS requests, timeouts
+        reading data from the client.  The active connection for that
         \daemon{smtpd} must be cleaned up, entered in the database, and
@@
         which case there's a data structure to dispose of, or it may not in
-        which case there is not.  See
-        \sectionref{timeouts-during-data-phase} for the gory details.
+        which case there is not.  See \sectionref{timeouts during data
+        phase} for the gory details.
 
@@
         another mail server; this happens when a local address is aliased
-        to a remote address.  TRACK will be called when dealing with the
-        parent mail, and will create the child mail if necessary.  TRACK
-        checks if the child has already been tracked, either by this parent
-        or by another parent, and issues appropriate warnings in either
-        case.
+        to a remote address (see \sectionref{tracking re-injected mail} for
+        a full explanation).  \texttt{TRACK} will be called when dealing
+        with the parent mail, and will create the child mail if necessary.
+        \texttt{TRACK} checks if the child has already been tracked, either
+        by this parent or by another parent, and issues appropriate
+        warnings in either case.
 
@@
 
-The distribution of rules per action is shown in \graphref{Distribution of
-rules per action}.
-
-\showgraph{build/graph-action-distribution}{Distribution of rules per
-action}{Distribution of rules per action}
-
-\subsection{Postfix Actions}
-
-\label{list of postfix actions in implementation}
-
-XXX WHY DO I NEED HAVE postfix\_actions AT ALL\@?
-
-XXX SEE TABLE BELOW\@: THERE IS NEARLY A 1--1 MAPPING BETWEEN action AND
-postfix\_action.  NEW ACTIONS COULD BE ADDED TO FORCE A 1--1 MAPPING, E.G.
-COMMIT + DISCARDED COULD BECOME A DISCARD ACTION\@.  MAIL\_QUEUED HAVING
-INFO AND PROCESSING CAN PROBABLY BE DEALT WITH, THE BIGGEST DIFFICULTY WILL
-BE WITH SAVE\_DATA\@.
-
-\begin{tabular}{lll}
-
-1 & BOUNCE & BOUNCED \\
-1 & CLONE & ACCEPTED \\
-3 & COMMIT & DISCARDED \\
-1 & COMMIT & INFO \\
-1 & CONNECT & INFO \\
-1 & DELETE & DELETED \\
-64 & DELIVERY\_REJECTED & REJECTED \\
-1 & DISCONNECT & INFO \\
-1 & EXPIRY & EXPIRED \\
-1 & MAIL\_QUEUED & INFO \\
-1 & MAIL\_QUEUED & PROCESSING \\
-1 & MAIL\_TOO\_LARGE & DISCARDED \\
-1 & PICKUP & INFO \\
-2 & POSTFIX\_RELOAD & POSTFIX\_RELOAD \\
-13 & SAVE\_DATA & BOUNCED \\
-38 & SAVE\_DATA & INFO \\
-1 & SAVE\_DATA & REJECTED \\
-5 & SAVE\_DATA & SENT \\
-2 & SMTPD\_DIED & IGNORED \\
-1 & SMTPD\_WATCHDOG & IGNORED \\
-2 & TIMEOUT & DISCARDED \\
-1 & TRACK & SENT \\
-42 & UNINTERESTING & IGNORED \\
-
-\end{tabular}
-
-postfix\_action IS USED IN A FEW ACTIONS, BUT IT COULD PROBABLY BE EASILY
-REMOVED, OR MAYBE I SHOULD KEEP TRACK OF actions INSTEAD\@?
-
-XXX IF I KEEP THESE I NEED TO EXPLAIN WHY I HAVE THEM\@.
-
-\begin{description}
-
-    \item [ACCEPTED] Postfix has accepted a mail, and will
-        subsequently attempt to deliver it.
-
-    \item [BOUNCED] The mail has bounced, because of a mail loop,
-        delivery failure, or five day timeout.\footnote{As with
-        much of Postfix's behaviour, five days is the default value
-        but can be changed by the administrator if they choose.}
-
-    \item [DELETED] The mail was deleted from the queue by an
-        administrator.
-
-    \item [DISCARDED] Postfix discarded the mail it was in the
-        process of accepting, because it was either larger than the
-        size limit set by the administrator, or the client timed
-        out or disconnected.
-
-    \item [EXPIRED] The mail has been in the queue for five days without
-        successful delivery.  A bounce mail will be generated and sent to
-        the sender address.
-
-    \item [INFO] Represents an unspecified intermediate action that
-        the parser is not interested in per se, but that does log
-        useful information, supplementing other log lines.
-
-    \item [IGNORED] An action that is not only uninteresting in
-        itself, but that also provides no useful data.
-
-    \item [POSTFIX\_RELOAD] The administrator has instructed
-        Postfix to start or stop, and all \daemon{smtpd} processes
-        will be terminated.  This does not negatively affect the
-        log files or mail queued by Postfix for delivery.
-
-    \item [PROCESSING] \daemon{cleanup} is processing a mail.
-        XXX ADD MORE DETAIL HERE
-
-    \item [REJECTED] Postfix rejected a command from the remote
-        client, causing at least one recipient to be rejected.
-
-    \item [SENT] Postfix has successfully sent a mail.
-
-\end{description}
-
 \subsection{Adding New Actions}
@@
 Adding new actions is not as easy as adding new rules, though care has been
-taken in the parser's design and implementation to make adding new actions
-as painless as possible.  The implementer writes a subroutine that accepts
-the standard arguments given to actions, and registers it as an action by
-calling the framework subroutine add\_actions() with the name of the new
-action as a parameter.  The new action must be registered before the rules
-are loaded, because it is an error for a rule to specify an unregistered
-action; this helps catch mistakes made when adding new rules.  No other
-work is required from the implementer to integrate the action into the
-parser; all of their attention and effort can be focused on correctly
-implementing their action.  The only negative aspect is that the process
-involves editing the parser source code, which makes upgrading to a later
-version of the parser more difficult, though by no means impossible.  If
-the author of the new action wishes, they can take advantage of the
-parser's object oriented implementation by sub-classing it and implementing
-their changes in the derived class, allowing future upgrades of the parser
-with greatly reduced chance of conflicts.\footnote{The real difference
-between the two approaches is where the new code is placed.  The simpler
-option is to change the parser code directly, but those changes will then
-have to be made to subsequent versions of the parser, and as the scope of
-the changes increases so does the chance of conflict, or mistakes when
-copying the action.  The more time consuming option is to write a subclass
-containing the new actions and change the program that invokes the parser
-so that it uses the subclass rather than the parser; the changes required
-to the program invoking the parser are minor and much less likely to lead
-to conflicts when upgrading to future versions of the parser.  An
-alternative is to submit new actions to the author of the parser for
-inclusion in future versions, resulting in two benefits: the new actions do
-not need to be maintained separately, and other users of the parser can
-avail of the new functionality.} The action may need to extend the list of
-valid combinations described in \sectionref{out of order log lines} if the
-new action creates a different set of acceptable programs, but this is
-unlikely to occur, as it would require parsing log lines from Postfix
-components the parser currently ignores.\footnote{The mail server used for
-development does not utilise either the \daemon{lmtp} or \daemon{virtual}
-delivery agents, so this parser does not have rules to handle log lines
-from those components.  Adding new rules to parse those component's log
-lines is a simple process, though if their behaviour differs significantly
-from the \daemon{smtp} or \daemon{local} delivery agents new actions may be
-required.  The mail server in question is a production mail server handling
-mail for a university department; the benefit of using this server is that
-the log files used exhibit the idiosyncrasies and peculiarities a mail
-server in the wild must deal with, but the downside is that significantly
-altering the configuration just to log messages from a different Postfix
-component is not an option.}
+taken in the architecture and implementation to make adding new actions as
+painless as possible.  The implementer writes a subroutine that accepts the
+standard arguments given to actions, and registers it with the framework.
+The new action must be registered before the rules are loaded, because it
+is an error for a rule to specify an unregistered action; this helps catch
+mistakes made when adding new rules.  No other work is required from the
+implementer to integrate the action into the parser; all of their attention
+and effort can be focused on correctly implementing their action.  The only
+negative aspect is that the process involves editing the parser source
+code, which makes upgrading to a later version of the parser more
+difficult, though by no means impossible.  If the author of the new action
+wishes, they can take advantage of the parser's object oriented
+implementation by sub-classing it and implementing their changes in the
+derived class, allowing future upgrades of the parser with greatly reduced
+chance of conflicts.\footnote{The real difference between the two
+approaches is where the new code is placed.  The simpler option is to
+change the parser code directly, but those changes will then have to be
+made to subsequent versions of the parser, and as the scope of the changes
+increases so does the chance of conflict or mistakes.  The second option
+requires more effort initially but less effort in the long run: write a
+subclass containing the new actions and change the program that invokes the
+parser so that it uses the subclass rather than the parent class; the
+changes required are minor and much less likely to lead to conflicts when
+upgrading to future versions of the parser.  An alternative is to submit
+new actions to the author \parsername{} for inclusion in future versions so
+that the new actions do not need to be maintained separately, and other
+users of the parser can benefit from the new functionality.} The action may
+need to extend the list of valid combinations described in \sectionref{out
+of order log lines} if the new action creates a different set of acceptable
+programs, but this is unlikely to occur, as it would require parsing log
+lines from Postfix components the parser currently ignores.\footnote{The
+mail server used for development does not utilise either the \daemon{lmtp}
+or \daemon{virtual} delivery agents, so this parser does not have rules to
+handle log lines from those components.  Adding new rules to parse those
+component's log lines is a simple process, though if their behaviour
+differs significantly from the \daemon{smtp} or \daemon{local} delivery
+agents new actions may be required.  The mail server in question is a
+production mail server handling mail for a university department; the
+benefit of using this server is that the log files used exhibit the
+idiosyncrasies and peculiarities a mail server in the wild must deal with,
+but the downside is that significantly altering the configuration just to
+log messages from a different Postfix component is not an option.}
 
@@
 \acronym{pid} any more, its queueid will be used instead (except when
-timeouts occur during the data phase
-\sectionref{timeouts-during-data-phase}).  If the original connection has
-any useful data (e.g.\ rejections) it will be saved to the database when
-the client disconnects.  One unsolved difficulty is distinguishing between
-different groups of rejections, e.g.\ when dealing with the following
-sequence:
+timeouts occur during the data phase \sectionref{timeouts during data
+phase}).  If the original connection has any useful data (e.g.\ rejections)
+it will be saved to the database when the client disconnects.  One unsolved
+difficulty is distinguishing between different groups of rejections, e.g.\
+when dealing with the following sequence:
 
@@
         bounce notification, see
-        \sectionref{identifying-bounce-notifications} for details.
+        \sectionref{identifying bounce notifications} for details.
 
@@
 
-\label{identifying-bounce-notifications}
+\label{identifying bounce notifications}
 
@@
 
-\label{aborted-delivery-attempts}
+\label{aborted delivery attempts}
 
@@
 disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts-during-data-phase}).  Sadly, Postfix does not
+parser (\sectionref{timeouts during data phase}).  Sadly, Postfix does not
 log a message saying the mail has been discarded, as should be expected by
@@
 
-\label{timeouts-during-data-phase}
+\label{timeouts during data phase}
 
@@
 connection the last accepted mail may need to be discarded, as detailed in
-\sectionref{timeouts-during-data-phase}; otherwise the queueid state table
+\sectionref{timeouts during data phase}; otherwise the queueid state table
 is untouched.
@@
 
-\label{yet-more-aborted-delivery-attempts}
-
-The aborted delivery attempts described in
-\sectionref{aborted-delivery-attempts} occur frequently, but the aborted
-delivery attempts described in this section only occur four times in the
-\numberOFlogFILES{} log files used for testing.  The symptoms are the same
-as in \sectionref{aborted-delivery-attempts}, except that there
-\textit{is\/} a \daemon{cleanup} log line; there does not appear to be
-anything in the log file to explain why there are no further log lines.
-The only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails displaying the following characteristics:
+\label{yet more aborted delivery attempts}
+
+The aborted delivery attempts described in \sectionref{aborted delivery
+attempts} occur frequently, but the aborted delivery attempts described in
+this section only occur four times in the \numberOFlogFILES{} log files
+used for testing.  The symptoms are the same as in \sectionref{aborted
+delivery attempts}, except that there \textit{is\/} a \daemon{cleanup} log
+line; there does not appear to be anything in the log file to explain why
+there are no further log lines.  The only way to detect these mails is to
+periodically scan all mails in the state tables, deleting any mails
+displaying the following characteristics:
 
@@
 
-    \item Improve the heuristics used in
-        \sectionref{timeouts-during-data-phase}, or develop another
-        solution, to avoid incorrectly leaving a mail in the state tables.
+    \item Improve the heuristics used in \sectionref{timeouts during data
+        phase}, or develop another solution, to avoid incorrectly leaving a
+        mail in the state tables.
 

### 2009-03-13 doc/logparser-implementation.tex
Corrections to flowchart section of implementation chapter. Improvements to graph size so they fill the width of the page. More consistent formatting of references to the flowchart. Alphabetical ordering for the actions with the same number of rules in the action distribution graph. Added more XXX items. Dealt with a few XXX items, particularly in the actions in PLP section of the implementation chapter. Improved the appearance of the flowchart a little. Updated the number of actions so that it is correct.

@@
 
-XXX WRITE AN INTRODUCTION\@.
+XXX WRITE AN INTRODUCTION\@.  MOST OF THIS IS CRAP, BUT I MIGHT BE ABLE TO
+RESCUE SOME OF IT\@.
 
@@
 In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  XXX REFER TO
-BLACK BOX PAPER\@?  There are some advantages to treating Postfix as a
-black box during parser development:
+black box program based solely on its inputs and outputs.  Thus far
+analysis of the log files has been sufficient to reconstruct Postfix's
+behaviour, but for more difficult programs the techniques described in
+\cite{black-box-error-reporting} may be useful.  There are some advantages
+to treating Postfix as a black box during parser development:
 
@@
 
-XXX MIGHT NEED A SMALL INTRO\@.
+XXX NEEDS AN INTRO\@.
 
@@
 
-    \item If the mail is rejected, save all data and finish.
+    \item If the mail is rejected, save all data to the database and
+        delete the connection entry from the state table.
 
     \item Follow the progress of the accepted mail until it is either
-        delivered, bounced, or deleted, then save all data, and finish.
+        delivered, bounced, or deleted, then save all data to the database,
+        and delete the mail's entry from the state table.
 
@@
 simplest case: a queueid is assigned immediately and the sender address is
-logged (action: PICKUP\@; flowchart:~2).  Re-injection due to forwarding
-sadly lacks explicit log lines of its own; it is explained in
-\sectionref{Re-injected mails}.
-
-Internally generated mails lack any explicit origin in Postfix 2.2.x and
-must be detected using heuristics as described in \sectionref{identifying
-bounce notifications}.  Bounce notifications are the primary example of
-internally generated mails, though there may be other types.  Postfix may
-generate mails to the administrator when it encounters configuration
-errors, but such mails are presumably rare.
-
-\acronym{SMTP} is more complicated than local submission:
+logged \flowchart{PICKUP}{2}.  Re-injection due to forwarding sadly lacks
+explicit log lines of its own; it is explained in \sectionref{Re-injected
+mails}.  Internally generated mails lack any explicit origin in Postfix
+2.2.x and must be detected using heuristics as described in
+\sectionref{identifying bounce notifications}.  Bounce notifications are
+the primary example of internally generated mails, though there may be
+other types.  Postfix may generate mails to the administrator when it
+encounters configuration errors, but such mails are presumably rare.
+
+\acronym{SMTP} is more complicated than the others:
 
@@
 
-    \item First there is a connection from the remote client (action:
-        CONNECT\@; flowchart:~1).
+    \item First there is a connection from the remote client
+        \flowchart{CONNECT}{1}.
 
@@
         addresses, client \acronym{IP} address or hostname, HELO hostname,
-        etc.\ (action: REJECTION\@; flowchart:~4); acceptance of one or
-        more mails (action: CLONE\@; flowchart:~5); or some interleaving of
-        both.
-
-    \item The client disconnects (action: DISCONNECT\@; flowchart:~6).  If
-        Postfix has rejected any \acronym{SMTP} commands the data will be
-        saved to the database; if not there will not be any data to save
-        (any mails accepted will already have been cloned so their data is
-        in another data structure).
+        etc.\ \flowchart{REJECTION}{4}; acceptance of one or more mails
+        \flowchart{CLONE}{5}; or some interleaving of both.
+
+    \item The client disconnects \flowchart{DISCONNECT}{6}.  If Postfix has
+        rejected any \acronym{SMTP} commands the data will be saved to the
+        database; if not there will not be any data to save (any mails
+        accepted will already have been cloned so their data is in another
+        data structure).
 
@@
     \item Postfix will log the sender and recipient addresses separately
-        (action: SAVE\_DATA\@; flowchart:~9).
+        \flowchart{SAVE\_DATA}{9}.
 
     \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail (action: TRACK\@; flowchart:~10) ---
-        \sectionref{tracking re-injected mail} discusses this in
-        detail.
+        tracked by the parent mail \flowchart{TRACK}{10};
+        \sectionref{tracking re-injected mail} discusses this in detail.
 
     \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator (action: COMMIT\@; flowchart:~12).  This is the last
-        log line for this particular mail (though it may be indirectly
-        referred to if it was re-injected).  If it is neither parent nor
-        child of re-injection the data is cleaned up and entered in the
-        database (flowchart:~14), then deleted from the state tables.
-        Re-injected mails are described in \sectionref{tracking re-injected
-        mail}.
+        administrator \flowchart{COMMIT}{12}.  This is the last log line
+        for this particular mail (though it may be indirectly referred to
+        if it was re-injected).  If it is neither parent nor child of
+        re-injection the data is cleaned up and entered in the database,
+        then deleted from the state tables.  The handling of re-injected
+        mails is described in \sectionref{tracking re-injected mail}.
 
@@
 
-It should be reiterated that the actions above happen whether the mail is
-delivered to a mailbox, piped to a command, delivered to a remote server,
-bounced (due to a mail loop, delivery failure, or five day timeout), or
-deleted by the administrator, \textit{unless\/} the mail is either parent
-or child of re-injection, as explained in \sectionref{tracking re-injected
-mail}.
+It should be emphasised that the mail delivery process above happens
+whether the mail is delivered to a mailbox, piped to a command, delivered
+to a remote server, bounced (due to a mail loop, delivery failure, or five
+day timeout), or deleted by the administrator, \textit{unless\/} the mail
+is either parent or child of re-injection, as explained in
+\sectionref{tracking re-injected mail}.
 
@@
 
+XXX GENERIC RULES\@.
+
+XXX COMPRESSED FILES\@.
+
+XXX STATE TABLES\@.
+
+XXX LOAD AND VERIFY RULES\@.
+
 \section{Actions}
@@
 \label{actions in implementation}
-
-XXX INTRODUCTION\@.
 
@@
 rules is \texttt{DELIVERY\_REJECTED}, the action that processes Postfix
-rejecting a mail delivery attempt; next is \texttt{SAVE\_DATA}, the action
-responsible for processing informative log lines, supplementing the data
-gathered from other log lines.  The third most common action is, perhaps
+rejecting a mail delivery attempt.  The next most common action is, perhaps
 surprisingly, \texttt{UNINTERESTING}: this action does nothing when
@@
 framework can warn about unrecognised log lines, informing the user that
-they need to add to or extend their ruleset.  The remaining actions have
-only one or two associated rules: some are specialised actions required to
-address a deficiency in the log files, or a complication that arises during
-parsing; other actions will only ever have one log line variant, e.g.\ all
-log lines showing that a remote client has connected are recognised by a
-single rule and handled by the \texttt{CONNECT} action.  \parsernames{}
-actions are described in detail in \sectionref{actions in detail in
-implementation}, and \sectionref{adding new actions in implementation}
-covers the process of adding a new action.
+they need to add to or extend their ruleset.  Most of the remaining actions
+have only one or two associated rules: some are specialised actions
+required to address a deficiency in the log files, or a complication that
+arises during parsing; other actions will only ever have one log line
+variant, e.g.\ all log lines showing that a remote client has connected are
+recognised by a single rule and handled by the \texttt{CONNECT} action.
+\parsernames{} actions are described in detail in \sectionref{actions in
+detail in implementation}, and \sectionref{adding new actions in
+implementation} covers the process of adding a new action.
 
@@
 
-    \item [MAIL\_BOUNCED] XXX WRITE THIS\@; INVOKES SAVE\_DATA ACTION\@.
-
-    \item [MAIL\_DISCARDED] XXX WRITE THIS\@; INVOKES COMMIT ACTION\@.
+    \item [MAIL\_BOUNCED] This action behaves exactly like the
+        \texttt{SAVE\_DATA} action; it saves all data captured by the
+        recognising rule's regex, and does nothing more.  It is a separate
+        action to help distinguish delivery attempts that bounce from other
+        delivery attempts.
+
+    \item [MAIL\_DISCARDED] Sometimes mail is discarded, e.g.\ mail
+        submitted locally that is larger than the limit configured by the
+        administrator.  This action is used for those mails; it invokes the
+        \texttt{COMMIT} action, but is a separate action to simplify
+        further analysis.
 
@@
 
-    \item [MAIL\_SENT] XXX WRITE THIS\@; INVOKES SAVE\_DATA ACTION\@.
+    \item [MAIL\_SENT] This action behaves exactly like the
+        \texttt{SAVE\_DATA} action; it saves all data captured by the
+        recognising rule's regex, and does nothing more.  It is a separate
+        action to help distinguish successful delivery attempts from other
+        delivery attempts.
 

### 2009-03-13 doc/logparser-background.tex
Removed some unused references.

@@
 
-\label{SMTP background}
-
 \SMTPglossaryDescription{} Despite the simplicity of the protocol many
@@
 strict adherence to the protocol specification is beneficial in protecting
-against spam and viruses.\footnote{\label{footnote:rfc760}Originally all
-mail servers adhered to the principle of \textit{Be liberal in what you
-accept, and conservative in what you send\/} from
-\acronym{RFC}~760~\cite{rfc760}, but unfortunately that principle was
-written in a friendlier time.  Given the deluge of spam that mail servers
-are subjected to daily, a more appropriate maxim could be: \textit{Require
-strict adherence to relevant \acronyms{RFC}; implement the strongest
-restrictions you can; relax the restrictions and adherence only when
-legitimate mail is impeded.\/}  It is neither as friendly nor as catchy,
-but it more accurately reflects the current situation.} A typical
+against spam and viruses.\footnote{Originally all mail servers adhered to
+the principle of \textit{Be liberal in what you accept, and conservative in
+what you send\/} from \acronym{RFC}~760~\cite{rfc760}, but unfortunately
+that principle was written in a friendlier time.  Given the deluge of spam
+that mail servers are subjected to daily, a more appropriate maxim could
+be: \textit{Require strict adherence to relevant \acronyms{RFC}; implement
+the strongest restrictions you can; relax the restrictions and adherence
+only when legitimate mail is impeded.\/}  It is neither as friendly nor as
+catchy, but it more accurately reflects the current situation.} A typical
 \acronym{SMTP} conversation resembles the following (the lines starting
@@
 
-    \item \acronym{SPF}~\cite{openspf}.  \acronym{SPF}\label{spf
-        introduction} records specify which mail servers are allowed to
-        send mail using sender addresses from a particular domain.  The
-        intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter}, and
+    \item \acronym{SPF}~\cite{openspf}.  \acronym{SPF} records specify
+        which mail servers are allowed to send mail using sender addresses
+        from a particular domain.  The intention is to reduce spam from
+        faked sender addresses, backscatter~\cite{postfix-backscatter}, and
         Joe~jobs~\glsadd{Joe-job}.  There has been considerable resistance

### 2009-03-13 doc/logparser-implementation.tex
Removed some unused references.
Improved introduction and flow chart diagram in the flow chart section of the implementation chapter.

@@
 CORRECT CHAPTER\@.
-
-XXX FIGURE OUT WHEN TO USE ACTIONS AND WHEN TO USE ALGORITHM, PARSER,
-ETC\@.
 
@@
 
-From the viewpoint of an individual mail passing through the parser the
-experience could be summarised as:
-
-\begin{enumerate}
-
-    \item Mail enters the system via \acronym{SMTP} or local submission.
-
-    \item If the mail is rejected, save all data to the database and
-        delete the connection entry from the state table.
-
-    \item Follow the progress of the accepted mail until it is either
-        delivered, bounced, or deleted, then save all data to the database,
-        and delete the mail's entry from the state table.
-
-\end{enumerate}
-
-\Figureref{flow chart image} shows the most common paths the data
-representing a mail can take through the parser; the complications
-described in \sectionref{complications} are excluded for the sake of
-clarity.
+A picture is said to be worth a thousand words: the flow chart in
+\figureref{flow chart image} shows the most common paths a connection or
+mail can take through \parsername{}; the parsing complications described in
+\sectionref{complications} are excluded for the sake of clarity.
 
@@
 image}
-
-\label{mail enters the system}
 
@@
 logged \flowchart{PICKUP}{2}.  Re-injection due to forwarding sadly lacks
-explicit log lines of its own; it is explained in \sectionref{Re-injected
-mails}.  Internally generated mails lack any explicit origin in Postfix
-2.2.x and must be detected using heuristics as described in
-\sectionref{identifying bounce notifications}.  Bounce notifications are
-the primary example of internally generated mails, though there may be
-other types.  Postfix may generate mails to the administrator when it
-encounters configuration errors, but such mails are presumably rare.
+explicit log lines of its own; it is explained fully in
+\sectionref{Re-injected mails}.  Internally generated mails lack any
+explicit origin in Postfix 2.2.x and must be detected using heuristics as
+described in \sectionref{identifying bounce notifications}; later versions
+of Postfix do provide log lines for internally generated mails.    Bounce
+notifications are the primary example of internally generated mails, though
+there are other types, e.g.\ Postfix may generate mails to the
+administrator when it encounters configuration errors, but such mails are
+presumably rare.
 
@@
         addresses, client \acronym{IP} address or hostname, HELO hostname,
-        etc.\ \flowchart{REJECTION}{4}; acceptance of one or more mails
-        \flowchart{CLONE}{5}; or some interleaving of both.
+        etc.\ \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more
+        mails \flowchart{CLONE}{5}; or some interleaving of both.
 

### 2009-03-14 doc/logparser-implementation.tex
Removed all mention of postfix_action, rewriting as necessary.

@@
 
-XXX REWRITE THIS PARAGRAPH\@; FIGURE OUT WHAT THE HELL IS HAPPENING WITH
-POSTFIX\_ACTION\@.
-
-Every log line will have a row in the results table, unless the \newline{}
-\texttt{postfix\_action} field of the rule which parsed it was INFO or
-IGNORE\@.  Each row in the results table is associated with a single
-connection; a single connection will have at least one result associated
-with it, but will usually have several more, and may have hundreds.
+Every recognised log line will have a row in the results table, associated
+with a single connection; a single connection will have at least one result
+associated with it, but will usually have several more, and may have
+hundreds.
 
@@
     restriction\_name   & reject\_unknown\_sender\_domain                   \\
-    postfix\_action     & REJECTED                                          \\
     program             & \daemon{smtpd}                                    \\
@@
 
-    \item [name, description, restriction\_name, and postfix\_action:] are
-        not \newline{} used by the parser, they serve to document the rule
-        for the user's benefit.
+    \item [name, description, and restriction\_name:] are not used by the
+        parser, they serve to document the rule for the user's benefit.
 
@@
 
-    \item XXX IS THIS CORRECT NOW\@?\newline{}  Are there two or more
-        \daemon{smtpd} log lines?  Does the second result have a
-        postfix\_action attribute of ACCEPTED\@?  The first two
-        \daemon{smtpd} log lines will be a connection log line and a mail
-        acceptance log line (Postfix logs acceptance as soon as the first
-        recipient has been accepted).
+    \item Was the second result processed by the \texttt{CLONE} action?
+        The first two \daemon{smtpd} log lines will be a connection log
+        line and a mail acceptance log line; Postfix logs acceptance as
+        soon as the first recipient has been accepted.
 
@@
 
-    \item If there are third and subsequent results, are all their
-        postfix\_action attributes equal to INFO\@?  If there are any log
+    \item If there are third and subsequent results, were those log lines
+        processed by the \texttt{SAVE\_DATA} action?  If there are any log
         lines after the first two they should be informational only.
@@
         indefinitely and consuming resources if a failure causes them to
-        enter a stuck state.  XXX IMPROVE THIS SOMEHOW\@.
+        enter a stuck state.
 

### 2009-03-14 doc/logparser-implementation.tex
Dealt with some XXX items.

@@
         connection indexed by \daemon{smtpd} \acronym{pid}, and save the
-        data captured by the rule (\acronym{IP} address and hostname of the
-        client).  If a connection already exists it is treated as a symptom
-        of a bug, and the action issues a warning containing the full
-        contents of the existing connection plus the log line that has
-        just been parsed.
+        data captured by the rule's regex (\acronym{IP} address and
+        hostname of the client).  If a connection already exists it is
+        treated as a symptom of a bug in \parsername{}, and the action will
+        issue a warning containing the full contents of the existing
+        connection plus the log line that has just been parsed.
 
@@
 
-XXX DESCRIBE HOW TO GO ABOUT ADDING A NEW RULE\@.
+To add a new rule a new row must be added to the rules table in the
+database, containing all the required attributes: action, name,
+description, program, and regex.  The name and description fields should be
+set based on the meaning of the log line, to help others understand which
+log lines this rule will recognise; the program will be obvious from the
+unrecognised log lines.  The action depends on the what the log line
+represents, e.g.\ a delivery rejection, a mail being delivered, some useful
+information, or something else; examine the list of actions in
+\sectionref{actions in detail in implementation} to determine the correct
+one.  The regex will be constructed based on the log line, but see below
+for a tool to ease the process.
+
+There are other attributes that may be required in a rule:
+connection\_data, result\_data, priority, or restriction\_name.  In general
+it will only become clear that connection\_data or result\_data are
+required when \parsername{} warns about an entry in the connections or
+results tables that is missing some required fields, because values for the
+missing fields are not present in any of the log lines.  E.g.\ the rule
+recognising the \daemon{pickup} component processing a mail sets the
+client\_hostname to \texttt{localhost} and the client\_ip to
+\texttt{127.0.0.1}, because the mail originates on the local server.  If
+the new rule deliberately uses the architecture's overlapping rules feature
+it will be obvious that the priority field needs to be set, on this rule and
+possibly others; the priority field may be needed on unintentionally
+overlapping rules too, but that is more difficult to determine.  Finally
+the restriction\_name field should be set when the rule's action is
+\texttt{DELIVERY\_REJECTED}; hopefully the restriction in question will be
+clear from the content of the log line.
 
@@
 XXX IMPROVE THE INTRO\@: ``needs more of a lead-in''.
-
-XXX SHOULD I HAVE A LIST OF HOW MANY COMPLICATIONS ARE HANDLED BY EACH
-ACTION\@?
 
@@
 
+\newpage{} % XXX ADDED BECAUSE OF FLIP-FLOPPING LAYOUT.
+
 \subsection{Timeouts During DATA Phase}
@@
 
-XXX WOULD IT BE BETTER TO BEGIN THE EXPLANATIONS ON THE FOLLOWING LINE\@?
-
+% This will put the text on the line following the item name, if the
+% enumitem package is loaded.
+%\begin{description}[style=nextline]
 \begin{description}
@@
 
-    \item XXX IS THIS CORRECT NOW\@?  CAN THERE BE MORE smtpd LOG LINES\@?
-        There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
+    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
         log lines for the mail, with no additional log lines.
@@
 complication has been observed in XXX/\numberOFlogFILES{} log files, but
-typically when it does occur there will be many instances of it, closely
+typically when it does occur there were many instances of it, closely
 grouped.
+
+This lack of information cannot easily be overcome: it is trivial to log a
+warning for every recipient accepted, but when the warning for the first
+recipient is logged Postfix will not yet have allocated a queue file and
+queueid for the mail, so the warning will be associated with the connection
+rather than the accepted mail.  A queue file and queueid will be allocated
+after Postfix accepts the MAIL FROM command if
+\texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is set to ``no'', but that
+setting will cause disk IO for almost every delivery attempt, instead of
+just for delivery attempts where recipients are accepted, and consequently
+a drastic reduction in the performance of the mail server.
 
@@
 present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.\footnote{This is not required to correctly deal
+bounce notification mails.  This is not required to correctly deal
 with the complication, but is an optimisation to reduce the parser's memory
@@
 every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.}  The cache of bounce notification
+the memory requirements of the parser.  The cache of bounce notification
 mails will be pruned whenever the parser's state is saved, though if the
 size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.  XXX EXPLAIN PRUNING\@.
+keep the size in check.
 
@@
         Similarly using a new \acronym{DNSBL}, a new policy server, or new
-        administrator-defined rejection messages require new rules.
+        administrator-defined rejection messages requires new rules.
 
@@
         queueid will be logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}.  There is no difficulty in
+        \texttt{smtpd\_data\_restrictions}, and the log line can be
+        associated with the correct mail.  There is no difficulty in
         specifying a HELO-based restriction in
@@
         also, but when there are multiple recipients no addresses are
-        logged.  It is also possible to warn for every recipient,
-        preventing the complication in \sectionref{Mail deleted before
-        delivery is attempted} entirely.  XXX ADD A PARAGRAPH EXPLAINING
-        HOW\@.
+        logged.
 
@@
         sources must remain separate, the parser can easily be instructed
-        to use a different database to store the results in.
+        to store the results in a different database.
 

### 2009-03-17 doc/logparser-implementation.tex
Added a diagram of the database schema.  It took a long time to find a tool to draw it with, longer to draw it, and even longer still to make it fit on the page properly.
Checked all references in the implementation chapter.

Finished the flow chart introduction.

Some minor changes and corrections.

@@
 RESCUE SOME OF IT\@.
-
-XXX UPDATE REFERENCES WHEN ARCHITECTURE AND IMPLEMENTATION ARE MORE FLESHED
-OUT\@.  XXX WHEN FINISHED, CHECK ALL REFERENCES TO ENSURE THEY POINT AT THE
-CORRECT CHAPTER\@.
 
@@
 normalised, simple to use representation.  The parser's task is to follow
-the journey each mail takes through Postfix, combining the data extracted
-by rules into a coherent whole, saving it in a useful and consistent form,
-and performing housekeeping duties.
+the journey each mail takes through Postfix, combining the data captured by
+rules into a coherent whole, saving it in a useful and consistent form, and
+performing housekeeping duties.
 
@@
 
-XXX NEEDS AN INTRO\@.
-
 A picture is said to be worth a thousand words: the flow chart in
@@
 mail can take through \parsername{}; the parsing complications described in
-\sectionref{complications} are excluded for the sake of clarity.
+\sectionref{complications} are excluded for the sake of clarity.  The flow
+chart is intended to be a graphical overview of how a mail progresses
+through both Postfix and \parsername{}, providing an overall context into
+which the detailed descriptions in the remainder of this chapter will fit,
+in particular the actions (\sectionref{actions in implementation}) and
+complications (\sectionref{complications}).
 
@@
     \item If one or more mails were accepted there will be more log lines
-        for those mails later, see \sectionref{mail delivery}.
+        for those mails.
 
 \end{enumerate}
-
-\label{mail delivery}
 
@@
     \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail \flowchart{TRACK}{10};
-        \sectionref{tracking re-injected mail} discusses this in detail.
+        tracked by the parent mail \flowchart{TRACK}{10}; the handling of
+        re-injected mails is described in \sectionref{tracking re-injected
+        mail}.
 
@@
         re-injection the data is cleaned up and entered in the database,
-        then deleted from the state tables.  The handling of re-injected
-        mails is described in \sectionref{tracking re-injected mail}.
+        then deleted from the state tables.
 
@@
 recognise log lines, and the data saved from the parsing of log files.
-Rules have the fields required to recognise log lines, extract data to be
-saved, and specify the action to be invoked; they also have several fields
-that aid the user in understanding the meaning of the log lines recognised
-by each rule.  The rules are described in detail in \sectionref{rules in
-implementation}, but the fields are covered with the rest of the database
-schema in this section.
+Each rule has a regex to recognise log lines and capture data from them,
+and specifies the action to be invoked when a log line is recognised; they
+also have several fields that aid the user in understanding the meaning of
+the log lines recognised by each rule.  The rules are described in detail
+in \sectionref{rules in implementation}, but the rules table is documented
+in \sectionref{rules table} with the rest of the database schema.
 
@@
 \sectionref{results table}.  A diagram of the database schema is provided
-in \sectionref{Diagram of the Database Schema} before the in-depth
-descriptions of each table.
+in \figureref{Diagram of the Database Schema picture}, alongside the
+in-depth descriptions of each table.
 
@@
 
-A clear, comprehensible schema is essential when using the extracted data;
-it is more important when using the data than when storing it, because
-storing the data is a write-once operation, whereas utilising the data
-requires frequent searching, sorting, and manipulation of the data to
-produce customised reports or statistics.
+A clear, comprehensible schema is essential when using the data extracted
+from log lines; it is more important when using the data than when storing
+it, because storing the data is a write-once operation, whereas utilising
+the data requires frequent searching, sorting, and manipulation of the data
+to produce customised reports or statistics.
 
@@
 
-\subsection{Diagram of the Database Schema}
-
-\label{Diagram of the Database Schema}
-
-XXX ADD A DIAGRAM OF THE DATABASE SCHEMA\@.
-
 \subsection{Rules Table}
@@
 
+\newlength{\belowcaptionskipORIG}
+\setlength{\belowcaptionskipORIG}{\belowcaptionskip}
+\setlength{\belowcaptionskip}{10pt}
+\showgraph{database-schema}{Diagram of the Database Schema}{Diagram
+of the Database Schema picture}
+\setlength{\belowcaptionskip}{\belowcaptionskipORIG}
+
 Rules are discussed in detail in \sectionref{rules in implementation}, but
-the structure of the rules table is covered here with the other tables in
-the database.  Rules are created by the user, not the parser, and will not
-be modified by the parser (except for the hits and hits\_total fields).
-Rules classify the individual log lines, extracting data to be saved in the
-connections and results tables, and specifying the action to take for that
-log line.
+the structure of the rules table is documented here along with the other
+tables in the database.  Rules are created by the user, not the parser, and
+will not be modified by the parser (except for the hits and hits\_total
+fields).  Rules classify the individual log lines, capturing data to be
+saved in the connections and results tables, and specifying the action to
+take for that log line.
 
@@
         and/or improved in one location; enables automatic extraction and
-        saving of data; and makes each regex largely self-documenting.  For
-        efficiency the keywords are expanded and every rule's regex is
-        compiled before attempting to parse the log file; otherwise every
-        regex would be recompiled each time it was used, resulting in a
-        large, data dependent slowdown, as shown in \sectionref{Caching
-        compiled regexes}.
+        saving of captured data; and makes each regex largely
+        self-documenting.  For efficiency the keywords are expanded and
+        every rule's regex is compiled before attempting to parse the log
+        file; otherwise every regex would be recompiled each time it was
+        used, resulting in a large, data dependent slowdown, as shown in
+        \sectionref{Caching compiled regexes}.
 
@@
         i.e.\ semi-colon or comma separated assignment statements, with the
-        variable name on the left and the value on the right hand
-        side.\footnote{Commas and semi-colons cannot be escaped and thus
-        cannot be included in data.  This feature is intended for use with
-        small amounts of data, so dealing with escape sequences was deemed
-        unnecessary.}  Any field in the connections table can be set in
-        this way.
+        variable name on the left and the value on the right hand side.
+        Any field in the connections table can be set in this way.  Commas
+        and semi-colons cannot be escaped and thus cannot be included in
+        data.  This feature is intended for use with small amounts of data,
+        so dealing with escape sequences was deemed unnecessary.
 
@@
         established.  This may also be an indicator that the mail is a
-        bounce notification, see
-        \sectionref{identifying bounce notifications} for details.
+        bounce notification, see \sectionref{identifying bounce
+        notifications} for details.
 
@@
         locally from bounce notifications which are being re-injected as a
-        result of aliasing (described in \sectionref{Re-injected mails}).
-        In the latter case the message-id will be unchanged from the
-        original bounce notification, and so even if it happens to be in
-        the correct format (e.g.\ if it was generated by Postfix on this or
-        another server) it will not match the queueid of the mail.
+        result of aliasing.  In the latter case the message-id will be
+        unchanged from the original bounce notification, and so even if it
+        happens to be in the correct format (e.g.\ if it was generated by
+        Postfix on this or another server) it will not match the queueid of
+        the mail.
 
@@
 delivery attempts}, except that there \textit{is\/} a \daemon{cleanup} log
-line; there does not appear to be anything in the log file to explain why
-there are no further log lines.  The only way to detect these mails is to
-periodically scan all mails in the state tables, deleting any mails
-displaying the following characteristics:
+line; there is nothing in the log file to explain why there are no further
+log lines.  The only way to detect these mails is to periodically scan all
+mails in the state tables, deleting any mails displaying the following
+characteristics:
 
@@
 added, as every mail is required by the database schema
-(\sectionref{results table}) to have at least one recipient.  This
-complication has been observed in XXX/\numberOFlogFILES{} log files, but
-typically when it does occur there were many instances of it, closely
+(\sectionref{results table}) to have at least one recipient.  Typically
+when this complication occurs there are many instances of it, closely
 grouped.

### 2009-03-17 doc/logparser-implementation.tex
Wrote the framework section of the implementation chapter.

@@
 
-XXX TO BE WRITTEN\@.
-
-XXX GENERIC RULES\@.
-
-XXX COMPRESSED FILES\@.
-
-XXX STATE TABLES\@.
-
-XXX LOAD AND VERIFY RULES\@.
+The role of the framework in this architecture was described in detail in
+\sectionref{framework in architecture}; this section is concerned with the
+implementation of the framework within \parsername{}.  The framework
+manages the parsing process, taking care of the drudgery and boring tasks,
+provides services to the actions, and implements some efficiency measures.
+
+Each time the parser is run the framework performs some initialisation
+tasks, setting up data structures that will later be used by actions,
+including several state tables.  Data about the connections and mails being
+processed is held in \texttt{connections} and \texttt{queueids}
+respectively.  There are also state tables used when working around
+complications in parsing: \texttt{timeout\_queueids} is used when dealing
+with connections that time out during the DATA phase (\sectionref{timeouts
+during data phase}); \texttt{bounce\_queueids} is part of the solution to
+bounce notification mails being delivered before their creation is logged
+(\sectionref{Bounce notification mails delivered before their creation is
+logged}); and \texttt{postsuper\_deleted\_queueids} caches information
+about mails recently deleted by the administrator so that subsequent log
+lines processed by the SAVE\_DATA action can be discarded (\sectionref{Mail
+deleted before delivery is attempted}).
+
+When the framework loads the ruleset, it verifies each of the rules,
+checking:
+
+\begin{itemize}
+
+    \squeezeitems{}
+
+    \item That the specified action is registered with the framework.
+
+    \item Ensuring the regex is valid and compiling it for efficiency
+        (\sectionref{Caching compiled regexes}).
+
+    \item For overlap between the data captured by the regex and additional
+        data specified in either \texttt{connection\_data} or
+        \texttt{result\_data}.
+
+    \item That \texttt{connection\_data}, \texttt{result\_data}, and the
+        regex captures specify valid fields to save data to.
+
+\end{itemize}
+
+If there are state tables from a previous \parsername{} run to be loaded,
+they will be loaded now; obviously the framework supports saving its state
+tables too.  The need to track re-injected mail (\sectionref{Re-injected
+mails}) complicates the process of saving state, as the relationships
+between mails must be maintained; loading state is complicated by dealing
+with aborted delivery attempts (\sectionref{aborted delivery attempts}),
+because a different set of relationships between connections and mails must
+be re-created when the state tables are restored.
+
+The framework will now move on to parsing each log file.  For each log line
+it will use those rules whose \texttt{program} field matches the program in
+the log line (see \sectionref{rule conditions in implementation}), falling
+back to generic rules if necessary, and warning if the log line is
+unrecognised.  The rules are sorted for increased efficiency; see
+\sectionref{rule ordering for efficiency} for details.  The repetitive
+nature of log files gives them high compression ratios; the framework
+supports reading compressed files, simplifying the process of parsing them
+because they do not have to be uncompressed to a temporary file before
+parsing begins.  The framework also displays a progress bar to show how far
+parsing has progressed through the log file, and to give the user an
+indication of how long the remainder of the parsing process is likely to
+take.  The progress bar is not as reliable when parsing compressed files
+because a compressed block may uncompress to a variable number of log
+lines; there is also variation between the recognition and processing time
+of individual log lines, but overall the progress bar is a useful addition.
+
+The framework collects data used when evaluating \parsernames{} efficiency
+for the results chapter (\sectionref{Results}); in addition some of the
+techniques used to improve parsing speed can be turned off or altered to
+show the effect they have.  Three sets of data are collected:
+
+\begin{enumerate}
+
+    \item How long it takes to parse each log file.
+
+    \item The number of rules tried when recognising log lines; the
+        framework may need to try multiple rules for a given log line
+        before it finds one that recognises it.
+
+    \item How many log lines did the framework attempt to recognise, how
+        many were skipped because there were no rules whose condition
+        matched that log line's program, how many were successfully
+        recognised, and how many were not recognised.
+
+\end{enumerate}
+
+There are five ways the framework can adapt its behaviour to demonstrate
+how effective the optimisations it uses are, and how efficient
+\parsername{} is:
+
+\begin{enumerate}
+
+    \item The rule ordering used can be changed from optimal (the default,
+        most efficient) to shuffled (intended to represent an unordered
+        ruleset) or reverse (reverse of optimal, least efficient).  The
+        results of this are shown in \sectionref{rule ordering for
+        efficiency}.
+
+    \item The framework can record which rule recognised each log line, and
+        then on a subsequent run use that information so that it uses the
+        correct rule for each log line.  This gives the best possible
+        running time, as only one rule is used for each log line.  A
+        variation of this is to use the correct rule last, to get the worst
+        possible running time.  \sectionref{perfect rule ordering}
+        discusses this in detail.
+
+    \item Each regex is compiled and the result cached when the rules are
+        loaded; this can be disabled and the regex compiled every time it
+        is used when recognising log lines, to demonstrate the effect this
+        optimisation has on parser efficiency, as shown in
+        \sectionref{Caching compiled regexes}.
+
+    \item Normally once a log line has been recognised, the framework
+        invokes the action specified by the rule.  Invoking the action can
+        be skipped if desired, so the timing information shows how long is
+        spent recognising log lines; this time can be subtracted from the
+        time taken by a normal run to show how long is spent processing log
+        lines.  This data is analysed in \sectionref{recognising vs
+        processing}.
+
+    \item The framework can skip inserting data into the database, because
+        that dominates the execution time of the parser, and the evaluation
+        in \sectionref{Results} is measuring the speed of \parsername{},
+        not the speed of the database or the disks it resides on.
+
+\end{enumerate}
+
+The framework also provides several debugging options, to aid in writing or
+correcting rules, or figuring out why the parser is not behaving as
+expected.  In increasing order of severity they are:
+
+\begin{enumerate}
+
+    \item The regex from the recognising rule and the log line it
+        recognised can be printed, so that the user can verify that the
+        correct rule recognised each log line.  A possible variant of this
+        would be to mark specific rules so only their regexes and the log
+        lines they recognise would be printed.
+
+    \item Each result can be extended with extra debugging information, so
+        that when a mail's data is dumped for inspection more information
+        is available.  The information currently added is: the log line's
+        timestamp in human readable form; the entire log line; the name of
+        the log file and the number of the log line within it.
+
+    \item Every connection and result added to the database can be dumped
+        in a human readable form.  This will result in a huge amount of
+        debugging information, so it is only useful for small log file
+        snippets, because otherwise the amount of information is
+        overwhelming.
+
+\end{enumerate}
 
@@
 
-XXX HALF OF THIS CONTENT IS IN FOOTNOTES, WHICH SHOULD BE INLINED\@.
+XXX HALF OF THIS CONTENT IS IN FOOTNOTES, WHICH SHOULD BE INLINED\@; IT
+ALSO NEEDS TO BE PROOFREAD\@.
 
@@
 
+\label{rule conditions in implementation}
+
 XXX NEED AN INTRODUCTION\@.
+
+XXX MENTION RULE CONDITIONS HERE\@; STRIP THE EXISTING STUFF FROM THE RULES
+TABLE AND EXAMPLE RULE, THEN HAVE THEM REFER TO HERE\@.
 
@@
 
-\newpage{} % XXX ADDED BECAUSE OF FLIP-FLOPPING LAYOUT.
-
 \subsection{Timeouts During DATA Phase}

### 2009-03-17 doc/logparser-results.tex
Added a diagram of the database schema.  It took a long time to find a tool to draw it with, longer to draw it, and even longer still to make it fit on the page properly.
Checked all references in the implementation chapter.

Finished the flow chart introduction.

Some minor changes and corrections.
Wrote the framework section of the implementation chapter.

@@
 XXX NEED AN INTRODUCTION\@.
-
-XXX COMPARE SAVING RESULTS TO RAM DISK VERSUS SAVING TO HARD DISK\@.
 
@@
 XXX REUSE THE CONTENT FROM THE PAPER FOR THIS SECTION\@.
+
+XXX DO A TEST-SOME RUN AND SAVE THE RESULTS TO THE DATABASE, THEN COMPARE
+THE TIME WITH A RUN THAT DOES NOT SAVE RESULTS\@; PUT THE COMPARISON IN THE
+INTRODUCTION, NEAR THE MACHINE SPECS\@.
 
@@
 between the orderings at log files 22 and 62--68.
+
+\subsection{Perfect Rule Ordering}
+
+\label{perfect rule ordering}
+
+XXX WRITE ABOUT PERFECT RULE ORDERING\@; DUNNO WHETHER TO PUT IT IN WITH
+RULE ORDERING FOR EFFICIENCY, OR IN ITS OWN SECTION\@.
+
+\subsection{Where Is The Time Spent: Recognising Or Processing Log Lines?}
+
+\label{recognising vs processing}
+
+XXX EXAMINE THE TIME SPENT RECOGNISING LOG LINES VS THE TIME SPENT
+PROCESSING THEM\@.
 

### 2009-03-18 doc/logparser-architecture.tex
Removed some unused references.
Dealt with some XXX items, and added more.

@@
 inputs stopping immediately may be preferable.
-
-\label{why separate rules, actions, and framework?}
 
@@
 
+\label{rule conditions in architecture}
+
 Rules can have conditions attached that will be evaluated by the framework
@@
 run-time, or embedding a language like Lua.  If an implementation is going
-to use conditions so complex that they will require a Turing-equivalent
-programming language to implement them, the design may need to be
+to use conditions so complex that they will require a programming language
+capable of implementing a Turing machine, the design may need to be
 re-thought, including the decision to use this architecture --- there may

### 2009-03-18 doc/logparser-implementation.tex
Dealt with some XXX items, and added more.

@@
     \item [program] The Postfix component (e.g.\ \daemon{smtpd}) whose log
-        lines the rule applies to.  This avoids needlessly trying rules
-        that will not match the log line, or worse, might match
-        unintentionally.  Rules whose program is \texttt{*} will be tried
-        against any log lines that are not parsed by program specific
-        rules.  This is a simple example of the rule conditions described
-        in \sectionref{attaching conditions to rules}.
+        lines the rule applies to; see \sectionref{rule conditions in
+        implementation} for full details of how this attribute is used.
 
@@
 
+    \item [CLEANUP\_PROCESSING] \daemon{cleanup} processes every mail that
+        passes through Postfix; details of what it does are available in
+        \sectionref{Postfix Daemons}.  This action saves all data captured
+        by the rule's regex if the log line has not come after a timeout
+        (see \sectionref{discarding cleanup log lines}); it also creates
+        the mail if necessary (see \sectionref{pickup logging after
+        cleanup}).
+
     \item [CLONE] Multiple mails may be accepted on a single connection, so
@@
     \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
-        \acronym{SMTP} command from the remote client: log the rejection
-        with a mail if there is a queueid in the log line, or with the
-        connection if not.  XXX VERY IMPORTANT ACTION\@; EXPAND AND
-        REWRITE\@.
+        \acronym{SMTP} command from the remote client.  This is the action
+        most frequently specified by rules, because so many different
+        restrictions are used to reject delivery attempts.  This action is
+        quite simple: if the log line contains a queueid, save the data
+        captured by the rule's regex to the mail identified by that
+        queueid; otherwise save it to the connection identified by the
+        \acronym{pid} in the log line.  No further processing of the log
+        line is required.
 
@@
     \item [MAIL\_QUEUED] This action represents Postfix picking a mail from
-        the queue to deliver.  This action is used for both \daemon{qmgr}
-        and \daemon{cleanup} because it needs to deal with out of order log
-        lines; see \sectionref{discarding cleanup log lines} for details.
-        XXX EXPLAIN THIS BETTER\@.
+        the queue to deliver.  This action needs to deal with out of order
+        log lines when mail is re-injected for forwarding; see
+        \sectionref{Re-injected mails} for details.
 
@@
 
-XXX HALF OF THIS CONTENT IS IN FOOTNOTES, WHICH SHOULD BE INLINED\@; IT
-ALSO NEEDS TO BE PROOFREAD\@.
-
 Adding new actions is not as easy as adding new rules, though care has been
@@
 implementer to integrate the action into the parser; all of their attention
-and effort can be focused on correctly implementing their action.  The only
-negative aspect is that the process involves editing the parser source
-code, which makes upgrading to a later version of the parser more
-difficult, though by no means impossible.  If the author of the new action
-wishes, they can take advantage of the parser's object oriented
-implementation by sub-classing it and implementing their changes in the
-derived class, allowing future upgrades of the parser with greatly reduced
-chance of conflicts.\footnote{The real difference between the two
-approaches is where the new code is placed.  The simpler option is to
-change the parser code directly, but those changes will then have to be
-made to subsequent versions of the parser, and as the scope of the changes
-increases so does the chance of conflict or mistakes.  The second option
-requires more effort initially but less effort in the long run: write a
-subclass containing the new actions and change the program that invokes the
-parser so that it uses the subclass rather than the parent class; the
-changes required are minor and much less likely to lead to conflicts when
-upgrading to future versions of the parser.  An alternative is to submit
-new actions to the author \parsername{} for inclusion in future versions so
-that the new actions do not need to be maintained separately, and other
-users of the parser can benefit from the new functionality.} The action may
-need to extend the list of valid combinations described in \sectionref{out
-of order log lines} if the new action creates a different set of acceptable
-programs, but this is unlikely to occur, as it would require parsing log
-lines from Postfix components the parser currently ignores.\footnote{The
-mail server used for development does not utilise either the \daemon{lmtp}
-or \daemon{virtual} delivery agents, so this parser does not have rules to
-handle log lines from those components.  Adding new rules to parse those
-component's log lines is a simple process, though if their behaviour
-differs significantly from the \daemon{smtp} or \daemon{local} delivery
-agents new actions may be required.  The mail server in question is a
-production mail server handling mail for a university department; the
-benefit of using this server is that the log files used exhibit the
-idiosyncrasies and peculiarities a mail server in the wild must deal with,
-but the downside is that significantly altering the configuration just to
-log messages from a different Postfix component is not an option.}
+and effort can be focused on correctly implementing their new action.  The
+action may need to extend the list of valid combinations described in
+\sectionref{out of order log lines} if the new action creates a different
+set of acceptable programs, but this would only be necessary if the new
+action processes log lines from Postfix components \parsername{} currently
+does not have rules for.
 
@@
 
-\label{rule conditions in implementation}
-
 XXX NEED AN INTRODUCTION\@.
-
-XXX MENTION RULE CONDITIONS HERE\@; STRIP THE EXISTING STUFF FROM THE RULES
-TABLE AND EXAMPLE RULE, THEN HAVE THEM REFER TO HERE\@.
 
@@
 \parsername{} to create regexes from unparsed log lines is also described
-in that section.  The final portion of this section provides some
+in that section.  The rule conditions used in \parsername{} are explained
+next, and the final portion of this section provides some
 suggestions for how to detect overlapping rules.
@@
 
-    \item [program and regex:] If the program in the rule equals the
-        program in the log line, a match using the rule's regex will be
-        attempted against the log line; if the match is successful the
-        action will be executed, if not the next rule will be tried.  If
-        the program-specific rules do not match the log line, the parser
-        will fall back to generic rules; if those rules are unsuccessful a
-        warning will be issued.  Data captured by the regex (sender,
-        recipient, etc.) will be automatically saved to the results and
-        connections tables.
+    \item [program and regex:] The program is used to restrict the log
+        lines this rule will be used to recognise; see \sectionref{rule
+        conditions in implementation} for details.  The regex does the
+        actual recognition of the log lines, and data captured by the regex
+        (sender, recipient, etc.) will be automatically saved to the
+        results and connections tables.
 
@@
 greatly reduce the effort required to parse new or different log lines.
+
+\subsection{Rule Conditions}
+
+\label{rule conditions in implementation}
+
+Rule conditions as part of the architecture have already been documented in
+\sectionref{rule conditions in architecture}, and they can be very complex
+and difficult to evaluate.  In contrast, \parsername{} uses very simple
+rule conditions: each rule has a \texttt{program} attribute that specifies
+the Postfix component whose log lines it recognises, and each log line
+contains the name of the Postfix component that produced it; the framework
+will only use rules where the two are equal when trying to recognise log
+lines.  This avoids needlessly trying rules that will not match the log
+line, or worse, might match unintentionally.  In addition the framework
+supports generic rules, whose program attribute is ``*''; these will be
+used if none of the program-specific rules recognise the log line.  If none
+of the rules are successful the framework will warn the user.
+
+\subsection{Regex Components}
+
+XXX WRITE THIS BIT, AND REDUCE THE ENTRY IN THE RULES TABLE\@.
+
+    The regex to recognise log lines with.  The regex will
+        have several keywords expanded when the rules are loaded: this
+        simplifies reading and writing rules; avoids needless repetition of
+        complex regex components; allows the components to be corrected
+        and/or improved in one location; enables automatic extraction and
+        saving of captured data; and makes each regex largely
+        self-documenting.  For efficiency the keywords are expanded and
+        every rule's regex is compiled before attempting to parse the log
+        file; otherwise every regex would be recompiled each time it was
+        used, resulting in a large, data dependent slowdown, as shown in
+        \sectionref{Caching compiled regexes}.
 
@@
 
-Every piece of software suffers from some limitations and there is almost
+Every piece of software suffers from some limitations, and there is almost
 always room for improvement.  Below are the limitations and possible
-improvements that have been identified in the parser.
+improvements that have been identified in \parsername{}.
 
@@
 
-    \item Each new Postfix release requires new rules to be written or
-        existing rules modified to cope with the new or changed log lines.
+    \item Each new Postfix release requires writing new rules or modifying
+        existing rules to cope with the new or changed log lines.
         Similarly using a new \acronym{DNSBL}, a new policy server, or new
@@
 
-    \item The hostname used in the HELO command is not logged if the mail
-        is accepted.  Rectifying this is relatively simple: create a
-        restriction that is guaranteed to warn for every accepted mail, as
-        follows:
+    \item The hostname used in the HELO command is not logged if the
+        incoming delivery attempt is successful.  Configuring Postfix to do
+        this is relatively simple: create a restriction that is guaranteed
+        to warn for every accepted mail, as follows:
 
@@
         Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there will not be a queueid
+        place to log the HELO hostname, there will not yet be a queueid
         associated with the mail when \texttt{smtpd\_helo\_restrictions} is
@@
 
-    \item The parser does not separate mails where one or more mails are
-        rejected and a subsequent mail is accepted; it will appear in the
-        database as one mail with lots of rejections followed by acceptance
-        (this has already been mentioned in \sectionref{connection reuse}).
-        It does not appear to be possible to make this distinction given
-        the data Postfix logs, though it might be possible to write a
-        policy server to provide additional logging.
-
-    \item The parser will not detect that it is parsing the same log file
+    \item \parsername{} does not create separate mails where one or more
+        delivery attempts are rejected and subsequently a mail is accepted;
+        it will appear in the database as one mail with lots of rejections
+        followed by acceptance (this has already been mentioned in
+        \sectionref{connection reuse}).  It does not appear to be possible
+        to make this distinction given the data Postfix logs, though it
+        might be possible to write a policy server to provide additional
+        logging.
+
+    \item \parsername{} will not detect that it is parsing the same log file
         twice, resulting in the database containing duplicate entries.
 
-    \item The parser does not distinguish between log files produced by
+    \item \parsername{} does not distinguish between log files produced by
         different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, as log files
-        from different sources can be combined in the same database, or it
-        may be viewed as a limitation as there is no facility to
+        same database.  This may be viewed as an advantage, because log
+        files from different sources can be combined in the same database,
+        or it may be viewed as a limitation because there is no facility to
         distinguish between log files from different sources in the same
@@
         sources must remain separate, the parser can easily be instructed
-        to store the results in a different database.
+        to use a different database.
 
@@
 
-    \item Improve the heuristics used in \sectionref{timeouts during data
-        phase}, or develop another solution, to avoid incorrectly leaving a
-        mail in the state tables.
-
-\end{itemize}
+\end{enumerate}
 

### 2009-03-18 doc/logparser-results.tex
Dealt with some XXX items, and added more.

@@
-\chapter{Results}
+\chapter{Results and Evaluation}
 
 \label{Results}
-
-XXX SHOULD THIS CHAPTER BE CALLED ``Results'' OR ``Evaluation''\@?
 
@@
 XXX USE THE DATA COLLECTED ABOUT NUMBER OF RULES TRIED\@.  MAYBE INCLUDE IN
-ORDERINGS SECTION\@?
+ORDERINGS SECTION\@?  MIX WITH RECOGNISING VS RECOGNISING + ACTION, AND
+KEEP COUNTERS FOR EACH PROGRAM TOO --- MAYBE SOMETHING USEFUL WILL COME
+FROM IT\@.
 
@@
 \graphref{Mails received per day} shows that as expected there are far more
-mails received on weekdays than at weekends.   Median log file size is
-50MB, containing 285,000 log lines --- large scale mail servers would have
-much larger log files.  Note that \graphref{Mails received per day} and
-\tableref{Number of mails received per day: statistics} show the number of
-mails received by \acronym{SMTP} only; in particular the mail loops
-noticeable in other graphs do not contribute to these figures.
+mails received on weekdays than at weekends.  The mail server in question
+is a production mail server handling mail for a university department; the
+benefit of using this server is that the log files from it exhibit the
+idiosyncrasies and peculiarities a mail server in the wild must deal with,
+but the downside is that significantly altering the configuration just to
+log messages from a different Postfix component is not an option.  Median
+log file size is 50MB, containing 285,000 log lines --- large scale mail
+servers would have much larger log files.  Note that \graphref{Mails
+received per day} and \tableref{Number of mails received per day:
+statistics} show the number of mails received by \acronym{SMTP} only; in
+particular the mail loops noticeable in other graphs do not contribute to
+these figures.
 

### 2009-03-19 doc/logparser-implementation.tex
Dealt with more XXX items.

@@
 
-    \item [regex] The regex to recognise log lines with.  The regex will
-        have several keywords expanded when the rules are loaded: this
-        simplifies reading and writing rules; avoids needless repetition of
-        complex regex components; allows the components to be corrected
-        and/or improved in one location; enables automatic extraction and
-        saving of captured data; and makes each regex largely
-        self-documenting.  For efficiency the keywords are expanded and
-        every rule's regex is compiled before attempting to parse the log
-        file; otherwise every regex would be recompiled each time it was
-        used, resulting in a large, data dependent slowdown, as shown in
-        \sectionref{Caching compiled regexes}.
+    \item [regex] The regex to recognise log lines with, documented in
+        \sectionref{regex components}.
 
@@
 
-XXX MAKE ALL ACTION DESCRIPTIONS FLOW BETTER\@.
-
 \begin{description}
@@
         (see \sectionref{discarding cleanup log lines}); it also creates
-        the mail if necessary (see \sectionref{pickup logging after
-        cleanup}).
+        the mail if necessary, setting its unknown origin flag (see
+        \sectionref{pickup logging after cleanup}).
 
@@
     \item [CONNECT] Handle a remote client connecting: create a new
-        connection indexed by \daemon{smtpd} \acronym{pid}, and save the
-        data captured by the rule's regex (\acronym{IP} address and
-        hostname of the client).  If a connection already exists it is
-        treated as a symptom of a bug in \parsername{}, and the action will
-        issue a warning containing the full contents of the existing
-        connection plus the log line that has just been parsed.
+        connection, indexed by \daemon{smtpd} \acronym{pid}.  If a
+        connection already exists it is treated as a symptom of a bug in
+        \parsername{}, and the action will issue a warning containing the
+        full contents of the existing connection plus the log line that has
+        just been parsed.
 
@@
         sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check (see
-        \sectionref{out of order log lines}).  To ensure that the mail can
-        be committed the EXPIRY action sets a flag marking the mail as
-        expired; this flag later causes the valid combinations check to be
-        skipped.  The mail will be committed when a log line triggers the
-        \texttt{COMMIT} action; if the \texttt{COMMIT} action has already
-        run and refused to enter the mail in the database because it failed
-        the valid program combinations check, the \texttt{SAVE\_DATA}
-        action will re-invoke the \texttt{COMMIT} action the next time it
-        is invoked.
+        Postfix programs which passes the valid combinations check as
+        described in \sectionref{out of order log lines}; this action, with
+        cooperation from the COMMIT and SAVE\_DATA actions, deals with that
+        complication.
 
@@
     \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  Out of order log
-        entries may have caused the state table entry to already exist (see
-        \sectionref{pickup logging after cleanup}); if it does not exist it
-        will be created.
+        service dealing with a locally submitted mail.  A new mail will be
+        created, although out of order log entries may have caused the
+        state table entry to already exist, as documented in
+        \sectionref{pickup logging after cleanup}.
 
@@
         from the state tables.  Postfix probably kills all the other
-        components too, but the parser is only affected by \daemon{smtpd}
+        components too, but \parsername{} is only affected by \daemon{smtpd}
         processes exiting.
@@
 
-    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
-        signal; this is processed in the same way as the
-        \texttt{SMTPD\_DIED} action.
+    \item [SMTPD\_KILLED] Sometimes a \daemon{smtpd} is killed by a signal;
+        this is processed in the same way as a \daemon{smtpd} dying, see
+        the \texttt{SMTPD\_DIED} action for details.
 
@@
         infrequently, as there are many other timeouts that should occur in
-        the intervening hours, e.g.\ timeouts for DNS requests, timeouts
+        the intervening hours, e.g.\ timeouts for DNS requests or timeouts
         reading data from the client.  The active connection for that
@@
 
-    \item [TIMEOUT] The connection timed out so the mail currently being
-        transferred must be discarded.  The mail may have been accepted, in
-        which case there's a data structure to dispose of, or it may not in
-        which case there is not.  See \sectionref{timeouts during data
-        phase} for the gory details.
+    \item [TIMEOUT] The connection with the remote client timed out so the
+        mail currently being transferred must be discarded.  The mail may
+        have been accepted, in which case there's a data structure to
+        dispose of, or it may not in which case there is not.  See
+        \sectionref{timeouts during data phase} for full details.
 
@@
 
-    \item [UNINTERESTING] This rule just returns successfully; it is used
-        when a line needs to be parsed for completeness but does not either
-        provide any useful data or require anything to be done.
+    \item [UNINTERESTING] This action just returns successfully: it is used
+        when a log line needs to be parsed for completeness, but does not
+        either provide any useful data to be saved or require anything to
+        be done.
 
@@
 
-XXX NEED AN INTRODUCTION\@.
-
 The rules are responsible for classifying each log line and specifying the
@@
 
-The theoretical aspects of this component are covered in detail in
+The role of the rules in the architecture is covered in detail in
 \sectionref{rules in architecture}; this section is concerned with the
@@
 
-XXX WRITE THIS BIT, AND REDUCE THE ENTRY IN THE RULES TABLE\@.
-
-    The regex to recognise log lines with.  The regex will
-        have several keywords expanded when the rules are loaded: this
-        simplifies reading and writing rules; avoids needless repetition of
-        complex regex components; allows the components to be corrected
-        and/or improved in one location; enables automatic extraction and
-        saving of captured data; and makes each regex largely
-        self-documenting.  For efficiency the keywords are expanded and
-        every rule's regex is compiled before attempting to parse the log
-        file; otherwise every regex would be recompiled each time it was
-        used, resulting in a large, data dependent slowdown, as shown in
-        \sectionref{Caching compiled regexes}.
+\label{regex components}
+
+Each rule's regex will have keywords expanded when the rules are loaded,
+for several reasons:
+
+\begin{itemize}
+
+    \item It simplifies both reading and writing of regexes and makes each
+        regex largely self-documenting.  E.g.\ the meaning of
+        \_\_CLIENT\_HOSTNAME\_\_ is immediately clear, whereas
+        \verb!(?:unknown|(?:[-.\w]+))! needs to be deciphered.
+
+    \item It avoids needless repetition of complex regex components, and
+        allows the components to be corrected or improved in one
+        location.  E.g.\ \_\_SENDER\_\_ is used in 68 rules; if a mistake
+        is discovered it can be corrected in one place only.
+
+    \item It enables automatic extraction and saving of captured data.  The
+        regex snippets use Perl 5.10's named capture buffers~\cite{perlre},
+        so the mapping between captures and fields does not need to be
+        explicitly specified by the rule.
+
+\end{itemize}
+
+For efficiency the keywords are expanded and every rule's regex is compiled
+before attempting to parse the log file, otherwise every regex would be
+recompiled each time it was used, resulting in a large, data dependent
+slowdown, as shown in \sectionref{Caching compiled regexes}.  Most of the
+keywords are named after the fields in the connections or results tables
+they populate: \_\_CLIENT\_HOSTNAME\_\_, \_\_CLIENT\_IP\_\_, \_\_DELAY\_\_,
+\_\_DELAYS\_\_, \_\_ENHANCED\_STATUS\_CODE\_\_, \_\_HELO\_\_,
+\_\_MESSAGE\_ID\_\_, \_\_QUEUEID\_\_, \_\_RECIPIENT\_\_, \_\_SENDER\_\_,
+\_\_SERVER\_HOSTNAME\_\_, \_\_SERVER\_IP\_\_, \_\_SIZE\_\_, and
+\_\_SMTP\_CODE\_\_.
+
+Some of the keywords need more explanation:
+
+\begin{eqlist}
+
+    \squeezeitems{}
+
+    \item [\_\_CHILD\_\_]  The queueid of a child mail; see
+        \sectionref{Re-injected mails}.
+
+    \item [\_\_COMMAND\_\_]  All \acronym{SMTP} commands, except
+        \texttt{DATA}, which typically has a more specific rule.
+        Priorities could have been used instead of excluding \texttt{DATA}.
+
+    \item [\_\_CONN\_USE\_\_]  How many times the connection was reused;
+        Postfix tries to reuse connections whenever possible to reduce the
+        load on both the sending and receiving servers.
+
+    \item [\_\_DATA\_\_]  This snippet is special: it matches a zero-length
+        string, so it needs to be followed by a pattern that matches
+        appropriately, but it captures whatever is matched so that it can
+        be saved.  E.g.\ \verb!/(__DATA__connection (?:refused|reset!
+        \newline{} \tab{}\verb!by peer))/! matches either ``connection
+        refused'' or ``connection reset by peer'' and causes it to be saved
+        to the data field of that log line's result.
+
+    \item [\_\_PID\_\_]  The \acronym{pid} of a \daemon{smtpd} process that
+        dies or is killed; see the SMTPD\_DIED action in
+        \sectionref{actions in detail in implementation}.
+
+    \item [\_\_RESTRICTION\_START\_\_]  Matches the start of a delivery
+        rejected log line: XXX SHOULD I INCLUDE THIS\@? \newline{}
+        \verb!/(__QUEUEID__): reject(?:_warning)?: ! \newline{}
+        \verb!(?:RCPT|DATA) from ! \newline{}
+        \verb!(?>(__CLIENT_HOSTNAME__)\\[)! \newline{}
+        \verb!(?>(__CLIENT_IP__)\\]): (__SMTP_CODE__)! \newline{}
+        \verb!(?: __ENHANCED_STATUS_CODE__)?/!
+
+    \item [\_\_SHORT\_CMD\_\_]  Postfix sometimes logs \acronym{SMTP}
+        commands in a short, single word form.
+
+\end{eqlist}
 

### 2009-03-20 doc/logparser-implementation.tex
Dealt with more XXX items.

@@
 action processes log lines from Postfix components \parsername{} currently
-does not have rules for.
+does not have rules for, e.g.\ \daemon{virtual} or \daemon{lmtp}.
 
@@
 
-    \item [\_\_RESTRICTION\_START\_\_]  Matches the start of a delivery
-        rejected log line: XXX SHOULD I INCLUDE THIS\@? \newline{}
-        \verb!/(__QUEUEID__): reject(?:_warning)?: ! \newline{}
-        \verb!(?:RCPT|DATA) from ! \newline{}
-        \verb!(?>(__CLIENT_HOSTNAME__)\\[)! \newline{}
-        \verb!(?>(__CLIENT_IP__)\\]): (__SMTP_CODE__)! \newline{}
-        \verb!(?: __ENHANCED_STATUS_CODE__)?/!
+    \item [\_\_RESTRICTION\_START\_\_]  Matches the standard information
+        Postfix includes at the start of almost all log lines resulting
+        from rejecting a delivery attempt.
 
@@
 
-XXX IMPROVE THE INTRO\@: ``needs more of a lead-in''.
-
-The complications described in this section are listed in the order in
-which they were encountered during development of the parser.  Each of
-these complications caused the parser to operate incorrectly, generating
-either warning messages or leaving mails in the state table.  The frequency
-of occurrence is much higher at the start of the list, with the first
-complication occurring several orders of magnitude more frequently than the
-last.  When deciding which problem to address next, the most common was
-always chosen, as resolving the most common problem would yield the biggest
-improvement in the parser, prune the greatest number of mails from the
-state tables and error messages, and make the remaining problems more
-apparent.  The first three complications were encountered early in the
-parser's implementation and guided its design and development.
+It was initially expected that parsing Postfix log files would be a
+relatively simple task, requiring a couple of months of work.  The author
+had found Postfix log files useful when investigating problems reported by
+users, and an examination of the log files gave the impression that they
+would be straightforward to parse, process, and understand.  The large
+variation in log lines was not apparent, because the majority of log lines
+are recognised by a small set of rules, as shown in \figureref{rule hits
+graph}.  Most of the myriad complications and difficulties documented in
+this section were discovered during \parsernames{} development, but the
+first three complications were identified during the planning and design
+phase of this project, influencing the architecture's design.
+
+Each of these complications caused \parsername{} to operate incorrectly,
+generate warning messages, or leave mails in the state table.  The
+complications are listed in the order in which they were overcome during
+development of \parsername{}, with the first complication occurring several
+orders of magnitude more frequently than the last.  When deciding which
+problem to tackle next, the problem causing the highest number of warning
+messages or mails incorrectly remaining in the state tables was always
+chosen, because that approach yielded the biggest improvement in the
+parser, and made the remaining problems more apparent.
 
@@
 
-The most difficult complication initially encountered is that locally
-addressed mails are not always delivered directly to a mailbox: sometimes
-they are addressed to and accepted for a local address but need to be
-delivered to one or more remote addresses due to aliases.  When this
-occurs a child mail will be injected into the Postfix queue, but without
-the explicit logging that \daemon{smtpd} or \daemon{postdrop} injected
-mails have.  Thus the source of the mail is not immediately discernible
-from the log line in which the mail first appears; from a strictly
-chronological reading of the log files it usually appears as if the child
-mail has appeared from thin air.  Subsequently the parent mail will log the
-creation of the child mail, e.g.\ parent mail 3FF7C4317 creates child mail
-56F5B43FD\@:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline{}
-\tab{} delay=0, status=sent (forwarded as 56F5B43FD)}
+Mails sent to local addresses are not always delivered directly to a
+mailbox: sometimes they are sent to and accepted for a local address, but
+need to be delivered to one or more remote addresses due to aliasing.  When
+this occurs, a child mail will be injected into the Postfix queue, but
+without the explicit logging that mails injected by \daemon{smtpd} or
+\daemon{postdrop} have.  Thus the source of the mail is not immediately
+discernible from the log line in which the mail's queueid first appears:
+from a strictly chronological reading of the log lines it usually appears
+as if the child mail has been created without a source.  Subsequently the
+parent mail will log the creation of the child mail, e.g.\ parent mail
+3FF7C4317 creates child mail 56F5B43FD\@:
+
+\texttt{3FF7C4317: to=<username@example.com>, relay=local, \hfill{}
+\newline{} \tab{} \tab{} delay=0, status=sent (forwarded as 56F5B43FD)}
 
@@
 are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the log line
-cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
-order also depends on how soon \daemon{qmgr} processes the new mail.
-
-Because of the uncertain order the parser cannot complain when it
-encounters a log line from \daemon{qmgr} for a previously unseen mail;
-instead it must flag the mail as coming from an unknown origin, and
-subsequently clear the flag if and when the origin of the mail becomes
-clear.  Obviously the parser could omit checking where mails originate
-from, but requiring an explicit source helps to expose bugs in the parser;
-such checks helped to identify the complications described in
-\sectionref{discarding cleanup log lines} and \sectionref{pickup logging
-after cleanup}.
+addition, the first log line belonging to the child mail (the example log
+line above belongs to the parent mail) is logged by \daemon{qmgr} or
+\daemon{cleanup}, so the order also depends on how soon they process the
+new mail.
+
+Because of the uncertain order that log lines can appear in, \parsername{}
+cannot complain when it encounters a log line from either \daemon{qmgr} or
+\daemon{cleanup} for a previously unseen mail; instead it must flag the
+mail as coming from an unknown origin, and subsequently clear that flag if
+and when the origin of the mail becomes clear.  Obviously the parser could
+omit checking where mails originate from, but requiring an explicit source
+helps to expose bugs in the parser; such checks helped to identify the
+complications described in \sectionref{discarding cleanup log lines} and
+\sectionref{pickup logging after cleanup}.
 
 Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered, and entirely finished with
-\textbf{before} the parent logs the creation log line!  Thus, mails flagged
-as coming from an unknown origin cannot be entered into the database when
-their final log line is parsed; instead they must be marked as ready for
-entry and subsequently entered once the parent mail has been identified.
-
-XXX MERGE THE REMAINDER OF THIS SECTION INTO THE PRIOR MATERIAL\@.
-
-The crux of this complication is that re-injected mails appear in the log
-files without explicit logging indicating their source.  There are two
-implicit indications:
+child mail will be created, delivered, and entirely finished with,
+\textit{before\/} the parent mail logs the creation log line!  Thus, mails
+flagged as coming from an unknown origin cannot be entered into the
+database when their final log line is processed; instead they must be
+marked as ready for entry and subsequently entered once their origin has
+been identified.  The crux of this complication is that re-injected mails
+appear in the log files without explicit logging indicating their source.
+
+Tracking re-injected mail requires \parsername{} to do the following in the
+COMMIT action:
 
@@
 
-    \item The indicator which more commonly introduces re-injection is when
-        \daemon{qmgr} selects a mail with a previously unseen queueid for
-        delivery, in which case a new data structure will be created for
-        that mail.  The mail will be flagged as having unknown origins;
-        this flag will subsequently be cleared once the origin has been
-        established.  This may also be an indicator that the mail is a
-        bounce notification, see \sectionref{identifying bounce
-        notifications} for details.
-
-    \item Local delivery re-injects the mail and logs a relayed delivery
-        rather than delivering directly to a mailbox or program as it
-        usually would.\footnote{Relayed delivery is performed by the
-        \acronym{SMTP} client; local delivery means local to the server,
-        i.e.\ an address the server is final destination for.}  In this
-        case the mail may already have been created (described above) and
-        the unknown origin flag will be cleared; if not a new data
-        structure will be created.  In both cases the re-injected mail is
-        marked as a child of the original mail.  The log line in question
-        is:
-
-        \texttt{3FF7C4317: to=<username@example.com>, relay=local,
-        \newline{} \tab{} delay=0, status=sent (forwarded as 56F5B43FD)}
-
-        This always occurs for re-injected mail but typically occurs after
-        the first indicator.  This log line is required to tie the parent
-        and child mails together and so is central to the process of
-        tracking re-injected mails.
+    \item If a mail is tagged with the unknown origin flag it is assumed to
+        be a child mail whose parent has not yet been identified.  The mail
+        is tagged as ready to be entered in the database, but entry is
+        postponed until the parent is identified.  The child mail should
+        not have any subsequent log lines: only its parent should refer to
+        it.
+
+    \item If the mail is a child mail with an identified parent it is
+        entered in the database as usual, then removed from its parent's
+        list of children.  If this child is the last mail on that list, and
+        the parent has already been entered in the database, the parent
+        will be removed from the state tables.
+
+    \item If the mail is a parent it is entered in the database as usual,
+        because there will be no further log lines for it.  There may be
+        child mails waiting to be entered in the database; these are
+        entered as normal, and removed from the state tables.  If there are
+        more child mails, the parent's removal from the state tables will
+        be postponed until the last child has been entered.
 
 \end{enumerate}
-
-The algorithm for tracking and saving re-injected mail to the database can
-finally be described:
-
-\begin{itemize}
-
-    \item If the mail is of unknown origin it is assumed to be a child mail
-        whose parent has not yet been identified.  Mark the mail as ready
-        for entry in the database and wait for the parent to deal with it.
-        The mail should not have any subsequent log lines; only its parent
-        should refer to it.
-
-    \item If the mail is a child mail then it has already been tracked: as
-        with all other mail, the data is cleaned up, the child is entered
-        in the database, and then deleted from the state tables.  The child
-        mail will be removed from the parent mail's list of children; if
-        this is the last child and the parent has already been entered in
-        the database, the parent will also be deleted from the state
-        tables.
-
-    \item The last alternative is that the mail is a parent mail.
-        Regardless of the state of its children its data is cleaned up and
-        entered in the database.  The parent may have children that are
-        waiting to be entered in the database; the data for each of those
-        children is cleaned up and entered in the database, then deleted
-        from the state tables.  The parent may also have outstanding
-        children which are not yet delivered, in which case the parent must
-        be retained in the state tables until those children are finished
-        with.  As soon as the last child is deleted from the state tables
-        the parent will also be deleted from the state tables.
-
-\end{itemize}
-
-A parent mail can have multiple children, which may be delivered before or
-after the parent mail.
 
@@
 body of the mail are transferred.  Sometimes there is a timeout or the
-connection is lost\footnote{For brevity's sake \textit{timeout\/} will be
-used throughout this section, but everything applies equally to lost
-connections.} during the DATA phase; when this occurs Postfix will discard
+connection is lost during the DATA phase;\footnote{For brevity's sake
+\textit{timeout\/} will be used throughout this section, but everything
+applies equally to lost connections.} when this occurs Postfix will discard
 the mail and the parser needs to discard the data associated with that
@@
 if a timeout occurs there is no data available to save; the timeout is
-recorded with the connection data instead, which is saved.
-
-To deal properly with timeouts the parsing algorithm needs to do the
-following in the TIMEOUT action:
+recorded and saved with the connection instead.  To deal properly with
+timeouts the TIMEOUT action does the following:
 
@@
 
-    \item If one or more recipients have been accepted Postfix will have
-        allocated a queueid for the incoming mail, and there will be a mail
-        in the state tables that needs to be dealt with.
+    \item If one or more recipients have been accepted, Postfix will have
+        allocated a queueid for the incoming mail, and there may be a mail
+        in the state tables that needs to be discarded.  The timeout may
+        have interrupted transfer of an accepted delivery attempt, or it
+        may have occurred after a mail delivery attempt was rejected.  To
+        distinguish between the two cases the action compares the timestamp
+        of the last accepted mail against the timestamp of the last log
+        line from \daemon{smtpd} for that connection; this depends on the
+        CLONE action keeping a list of all mails accepted for each
+        connection.  If the mail acceptance timestamp is later than the
+        timeout timestamp, the timeout applies to the just-accepted mail,
+        which will be removed from the state tables, without entering it in
+        the database.  If the \daemon{smtpd} timestamp is later there was a
+        rejection between the accepted mail and the timeout: the action
+        assumes that the timeout applies to the rejected delivery attempt
+        and finishes.  This assumption is not necessarily correct, because
+        Postfix may have accepted an earlier recipient and rejected a later
+        one, in which case the timeout applies to the accepted mail, which
+        should be discarded; this has not been a problem in practice.
+        Processing timeouts is further complicated by the presence of out
+        of order \daemon{cleanup} log lines: see \sectionref{discarding
+        cleanup log lines} for details.
 
@@
 
-XXX MAKE THIS PARAGRAPH CLEARER\@.  A timeout may thus apply either to an
-accepted mail or a rejected mail.  To distinguish between the two cases the
-algorithm compares the timestamp of the last accepted mail against the
-timestamp of the last line logged by \daemon{smtpd} for that connection
-(the TIMEOUT action is dependant on the CLONE action keeping a list of all
-mails accepted on each connection).  If the mail acceptance timestamp is
-later then the timeout applies to the just-accepted mail, which will be
-discarded.  If the \daemon{smtpd} timestamp is later there was a rejection
-between the accepted mail and the timeout: the action assumes that the
-timeout applies to a rejected delivery attempt and finishes.  This
-assumption is not necessarily correct, because Postfix may have accepted an
-earlier recipient and rejected a later one, in which case the timeout
-applies to the accepted mail, which should be discarded.  This has not been
-a problem in practice, though it may be in future.  This complication is
-further complicated by the presence of out of order \daemon{cleanup} log
-lines: see \sectionref{discarding cleanup log lines} for details.
-
-This complication is dealt with in the TIMEOUT action.
+This complication is dealt with by the TIMEOUT action, with help from the
+CLONE action.
 

### 2009-03-25 doc/logparser-complications.tex
Moved complications to a separate file, because of their size.

--- 2009-03-20 doc/logparser-implementation.tex
@@
-\chapter{Postfix Parser Implementation}
-
-\label{Postfix Parser Implementation}
-
-XXX WRITE AN INTRODUCTION\@.  MOST OF THIS IS CRAP, BUT I MIGHT BE ABLE TO
-RESCUE SOME OF IT\@.
-
-The parser deals with all the complications of parsing, the eccentricities
-and oddities of Postfix log files, presenting the resulting data in a
-normalised, simple to use representation.  The parser's task is to follow
-the journey each mail takes through Postfix, combining the data captured by
-rules into a coherent whole, saving it in a useful and consistent form, and
-performing housekeeping duties.
-
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions, and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be able
-to handle multiple mails in parallel, each potentially at a different stage
-in its journey, without any interference between mails --- except in the
-minority of cases where intra-mail interference is required.  The best way
-to implement this is to maintain state information for every unfinished
-mail and manipulate the appropriate mail correctly for each log line
-encountered.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
-
-\section{Assumptions}
-
-\parsername{} makes a small number of (hopefully safe and reasonable)
-assumptions:
-
-\begin{itemize}
-
-    \item The log files are whole and complete: nothing has been removed,
-        either deliberately or accidentally (e.g.\ log file rotation gone
-        awry, file system filling up, logging system unable to cope with
-        the volume of log messages).  On a well run system it is extremely
-        unlikely that any of these problems will arise, though it is of
-        course possible, particularly when suffering from a deluge of spam
-        or a mail loop.
-
-    \item Postfix logs enough information to make it possible to accurately
-        reconstruct the actions it has taken.  There are several heuristics
-        used when parsing; see \sectionref{identifying bounce
-        notifications}, \sectionref{aborted delivery attempts}, and
-        \sectionref{pickup logging after cleanup} for details.
-
-    \item The Postfix queue has not been tampered with, causing unexplained
-        appearance or disappearance of mail.  This may happen if the
-        administrator deletes mail from the queue without using
-        \daemon{postsuper}, or if there is filesystem corruption.
-
-\end{itemize}
-
-In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Thus far
-analysis of the log files has been sufficient to reconstruct Postfix's
-behaviour, but for more difficult programs the techniques described in
-\cite{black-box-error-reporting} may be useful.  There are some advantages
-to treating Postfix as a black box during parser development:
-
-\begin{itemize}
-
-    \item Reading and understanding the source code would require a
-        significant investment of time: Postfix 2.5.5 has 17MB of source
-        code.
-
-    \item The parser is developed using real world log files rather than
-        the idealised log files someone would naturally envisage reading
-        the source code.  The source code cannot accurately communicate the
-        variety of orderings in which log lines are written to the log
-        file, as process scheduling independently interferes with logging
-        and other processing.
-
-    \item The parser acts as a second source of information about Postfix's
-        operation, using information gathered from empirical evidence.  A
-        separate project could compare the empirical knowledge inherent in
-        the parsing algorithm with Postfix's documentation and source code
-        to see how closely the two agree.
-
-\end{itemize}
-
-\section{Parser Flow Chart}
-
-\label{flow chart}
-
-A picture is said to be worth a thousand words: the flow chart in
-\figureref{flow chart image} shows the most common paths a connection or
-mail can take through \parsername{}; the parsing complications described in
-\sectionref{complications} are excluded for the sake of clarity.  The flow
-chart is intended to be a graphical overview of how a mail progresses
-through both Postfix and \parsername{}, providing an overall context into
-which the detailed descriptions in the remainder of this chapter will fit,
-in particular the actions (\sectionref{actions in implementation}) and
-complications (\sectionref{complications}).
-
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart}{flow chart
-image}
-
-Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
-forwarding, or internally generated by Postfix.  Local submission is the
-simplest case: a queueid is assigned immediately and the sender address is
-logged \flowchart{PICKUP}{2}.  Re-injection due to forwarding sadly lacks
-explicit log lines of its own; it is explained fully in
-\sectionref{Re-injected mails}.  Internally generated mails lack any
-explicit origin in Postfix 2.2.x and must be detected using heuristics as
-described in \sectionref{identifying bounce notifications}; later versions
-of Postfix do provide log lines for internally generated mails.  Bounce
-notifications are the primary example of internally generated mails, though
-there are other types, e.g.\ Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.
-
-\acronym{SMTP} is more complicated than the others:
-
-\begin{enumerate}
-
-    \item First there is a connection from the remote client
-        \flowchart{CONNECT}{1}.
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \acronym{IP} address or hostname, HELO hostname,
-        etc.\ \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more
-        mails \flowchart{CLONE}{5}; or some interleaving of both.
-
-    \item The client disconnects \flowchart{DISCONNECT}{6}.  If Postfix has
-        rejected any \acronym{SMTP} commands the data will be saved to the
-        database; if not there will not be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log lines
-        for those mails.
-
-\end{enumerate}
-
-The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
-
-\begin{enumerate}
-
-    \item Postfix will log the sender and recipient addresses separately
-        \flowchart{SAVE\_DATA}{9}.
-
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail \flowchart{TRACK}{10}; the handling of
-        re-injected mails is described in \sectionref{tracking re-injected
-        mail}.
-
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator \flowchart{COMMIT}{12}.  This is the last log line
-        for this particular mail (though it may be indirectly referred to
-        if it was re-injected).  If it is neither parent nor child of
-        re-injection the data is cleaned up and entered in the database,
-        then deleted from the state tables.
-
-\end{enumerate}
-
-It should be emphasised that the mail delivery process above happens
-whether the mail is delivered to a mailbox, piped to a command, delivered
-to a remote server, bounced (due to a mail loop, delivery failure, or five
-day timeout), or deleted by the administrator, \textit{unless\/} the mail
-is either parent or child of re-injection, as explained in
-\sectionref{tracking re-injected mail}.
-
-\section{Database}
-
-\label{database}
-
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by using those rules to parse Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered; \sectionref{database as API} describes
-how the database schema functions as an \acronym{API}.
-
-The database schema can be conceptually divided in two: the rules used to
-recognise log lines, and the data saved from the parsing of log files.
-Each rule has a regex to recognise log lines and capture data from them,
-and specifies the action to be invoked when a log line is recognised; they
-also have several fields that aid the user in understanding the meaning of
-the log lines recognised by each rule.  The rules are described in detail
-in \sectionref{rules in implementation}, but the rules table is documented
-in \sectionref{rules table} with the rest of the database schema.
-
-The data saved from parsing the log files is also divided into two tables:
-connections and results.  The connections table contains an entry for every
-mail accepted and every connection where there was a rejection; the
-individual fields will be described in \sectionref{connections table}.
-There will be at least one entry in the results table for each entry in the
-connections table; the result table's fields will be covered in detail in
-\sectionref{results table}.  A diagram of the database schema is provided
-in \figureref{Diagram of the Database Schema picture}, alongside the
-in-depth descriptions of each table.
-
-An important but easily overlooked benefit of storing the rules in the
-database is the link between rules and results: if more information is
-required when examining a result, the rule that produced the result is
-available for inspection because each result references the rule that
-created it.  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
-
-A clear, comprehensible schema is essential when using the data extracted
-from log lines; it is more important when using the data than when storing
-it, because storing the data is a write-once operation, whereas utilising
-the data requires frequent searching, sorting, and manipulation of the data
-to produce customised reports or statistics.
-
-\subsection{Using A Database To Provide An Application Programming Interface}
-
-\label{database as API}
-
-The database populated by \parsername{} provides a simple interface to
-Postfix log files.  Although the interface is a database schema rather than
-a set of functions in a library, it has the same benefits as any other
-\acronym{API}: it provides a stable interface, allowing code on either side
-of the interface to be changed without adverse effects, as long as the
-interface is adhered to.  Programs that use the database can range from the
-simple examples in \sectionref{motivation} to far more complex data mining
-tools.
-
-Using a database simplifies writing programs that need to interact with the
-data in several ways:
-
-\begin{enumerate}
-
-    \item Most programming languages have facilities for database access,
-        allowing a developer to write programs that use the gathered data
-        in their preferred programming language, rather than being
-        restricted to the language the parser is written in.
-
-    \item Databases provide complex querying and sorting functionality for
-        the user without requiring large amounts of programming.  All
-        databases provide a program, of varying complexity and
-        sophistication, that can be used for ad hoc queries with minimal
-        investment of time.
-
-    \item Databases are easily extensible, e.g.:
-
-        \begin{itemize}
-
-            \item New columns can be added to the tables used by the
-                program, using DEFAULT clauses or TRIGGERS to populate
-                them.
-
-            \item A VIEW gives a custom arrangement of data with minimal
-                effort.
-
-            \item Some databases support granting access on a fine-grained
-                basis, e.g.\ allowing the finance department to produce
-                invoices, the helpdesk to run limited queries as part of
-                dealing with support calls, and the administrators to have
-                full access to the data.
-
-            \item Triggers can be written to perform actions when certain
-                events occur.  In pseudo-\acronym{SQL}\@:
-
-\begin{verbatim}
-CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND action = 'DELIVERY_REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
-\end{verbatim}
-
-            \item Other tables can be added to the database, to cache
-                historical, summary, or computed data.
-
-        \end{itemize}
-
-    \item \acronym{SQL} is reasonably standard and many people will already
-        be familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn, e.g.\
-        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
-        Although every vendor implements a different dialect of
-        \acronym{SQL}, the basics are the same everywhere, analogous to the
-        overall similarities and minor differences amongst Irish English,
-        British English, American English, and Australian English.
-        Depending on the database in use there may be tools available that
-        reduce or remove the requirement to know \acronym{SQL}; several are
-        available for \gls{SQLite}, the default database used by
-        \parsername{}.
-
-\end{enumerate}
-
-Storing the results in a database will also increase the efficiency of
-using those results, because the log files need only be parsed once rather
-than each time the data is used; indeed the results may be used by someone
-with no access to the original log files.
-
-\subsection{Rules Table}
-
-\label{rules table}
-
-\newlength{\belowcaptionskipORIG}
-\setlength{\belowcaptionskipORIG}{\belowcaptionskip}
-\setlength{\belowcaptionskip}{10pt}
-\showgraph{database-schema}{Diagram of the Database Schema}{Diagram
-of the Database Schema picture}
-\setlength{\belowcaptionskip}{\belowcaptionskipORIG}
-
-Rules are discussed in detail in \sectionref{rules in implementation}, but
-the structure of the rules table is documented here along with the other
-tables in the database.  Rules are created by the user, not the parser, and
-will not be modified by the parser (except for the hits and hits\_total
-fields).  Rules classify the individual log lines, capturing data to be
-saved in the connections and results tables, and specifying the action to
-take for that log line.
-
-The following attributes are defined for each rule:
-
-\begin{boldeqlist}
-
-    \item [id] A unique identifier for each rule that other tables can use
-        when referring to a specific rule.
-
-    \item [name] A short name for the rule.
-
-    \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the event causing the log
-        lines this rule matches.
-
-    \item [restriction\_name] The restriction that caused the mail to be
-        rejected.  Only applicable to rules that recognise rejection log
-        lines, i.e.\ rules that have an action of
-        \texttt{DELIVERY\_REJECTED}, other rules should have an empty
-        string.
-
-    \item [program] The Postfix component (e.g.\ \daemon{smtpd}) whose log
-        lines the rule applies to; see \sectionref{rule conditions in
-        implementation} for full details of how this attribute is used.
-
-    \item [regex] The regex to recognise log lines with, documented in
-        \sectionref{regex components}.
-
-    \item [connection\_data] Sometimes rules need to save data that is not
-        present in the log line: e.g.\ setting \texttt{client\_ip} when a
-        mail is being delivered to another server.  The format is:
-        \newline{} \tab{} \texttt{ client\_hostname = localhost,}
-        \newline{} \tab{} \tab{} \texttt{client\_ip = 127.0.0.1} \newline{}
-        i.e.\ semi-colon or comma separated assignment statements, with the
-        variable name on the left and the value on the right hand side.
-        Any field in the connections table can be set in this way.  Commas
-        and semi-colons cannot be escaped and thus cannot be included in
-        data.  This feature is intended for use with small amounts of data,
-        so dealing with escape sequences was deemed unnecessary.
-
-    \item [result\_data] The result table equivalent of
-        \texttt{connection\_data}.
-
-    \item [action] The action that will be invoked when this rule
-        recognises a log line; a full list of actions and the parameters
-        they are invoked with can be found in \sectionref{actions in detail
-        in implementation}.
-
-    \item [hits] This counter is maintained for every rule and incremented
-        each time the rule successfully recognises a log line.  At the
-        start of each run the parser sorts the rules by how frequently each
-        rule recognised log lines, and at the end of the run it updates
-        every rule's hits field in the database.  Assuming that the
-        distribution of log lines is reasonably consistent across log
-        files, ordering rules by their recognition frequency will reduce
-        the parser's execution time.  Rule ordering for efficiency is
-        discussed in \sectionref{rule ordering for efficiency}.
-
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
-
-    \item [priority] This is the user-configurable companion to hits: when
-        the list of rules is sorted, priority overrides hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{rules in architecture}).
-
-\end{boldeqlist}
-
-\subsection{Connections Table}
-
-\label{connections table}
-
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
-
-\begin{boldeqlist}
-
-    \item [id] This field uniquely identifies the row.
-
-    \item [server\_ip] The \acronym{IP} address (IPv4 or IPv6) of the
-        server: the local address when receiving mail, the remote address
-        when sending mail.
-
-    \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \acronym{IP} address could not be resolved
-        to a hostname via DNS\@.
-
-    \item [client\_ip] The client \acronym{IP} address (IPv4 or IPv6): the
-        remote address when receiving mail, the local address when sending
-        mail.
-
-    \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \acronym{IP} address could not be resolved
-        to a hostname via DNS\@.
-
-    \item [helo] The hostname used in the HELO command.  The HELO hostname
-        occasionally changes during a connection, presumably because spam
-        or virus senders think it is a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \acronym{SMTP} command,
-        but it is quite simple to rectify this, as described in
-        \sectionref{logging helo}.
-
-    \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch.\glsadd{Epoch}
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
-
-\end{boldeqlist}
-
-\subsection{Results Table}
-
-\label{results table}
-
-Every recognised log line will have a row in the results table, associated
-with a single connection; a single connection will have at least one result
-associated with it, but will usually have several more, and may have
-hundreds.
-
-\begin{boldeqlist}
-
-    \item [connection\_id] The id of the row in the connections table this
-        result is associated with.
-
-    \item [rule\_id] The id of the rule in the rules table that recognised
-        the log line and created this result.
-
-    \item [id] A unique identifier for this result.
-
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \acronym{SMTP} command
-        --- a mechanism that is quite useful for testing new restrictions.
-        This field will be set to 1 if the log line parsed was a warning rather
-        than a real rejection, or to 0 for a real rejection.  This field
-        should be ignored if the result is not a rejection, i.e.\ the
-        action field of the associated rule is not
-        \texttt{DELIVERY\_REJECTED}.
-
-    \item [smtp\_code] The \acronym{SMTP} code associated with the log
-        line.  In general an \acronym{SMTP} code is only present in a log
-        line for a rejection or final delivery; results whose log line did
-        not contain an \acronym{SMTP} code will duplicate the
-        \acronym{SMTP} code of other results in that connection.  Some
-        final delivery log lines do not contain an \acronym{SMTP} code: in
-        those cases the \acronym{SMTP} code is specified by the rule's
-        \texttt{result\_data} field, based on the success or failure
-        represented by the log line.
-
-    \item [enhanced\_status\_code] The enhanced status code~\cite{RFC3463}
-        is similar to the \acronym{SMTP} code, but is intended to be parsed
-        by mail clients so that error messages can be clearly conveyed to
-        the user.  Enhanced status code support was added to Postfix in
-        version 2.3; logs generated by previous versions will not contain
-        any enhanced status codes.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one mail.
-        Mails sent over the same connection can be distinguished by their
-        queueids.
-
-    \item [recipient] The recipient address; there may be multiple
-        recipient addresses per mail or connection.
-
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
-
-    \item [delay] How long the mail was delayed while it was being
-        delivered.  This will only be present for delivered mails.
-
-    \item [delays] More detailed information about how long the mail was
-        delayed while it was being delivered, again only present for
-        delivered mails.
-
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from a \acronym{DNSBL}\@.
-
-    \item [timestamp] The time at which the log line was logged, in seconds
-        since the epoch.\glsadd{Epoch}
-
-\end{boldeqlist}
-
-\section{Framework}
-
-The role of the framework in this architecture was described in detail in
-\sectionref{framework in architecture}; this section is concerned with the
-implementation of the framework within \parsername{}.  The framework
-manages the parsing process, taking care of the drudgery and boring tasks,
-provides services to the actions, and implements some efficiency measures.
-
-Each time the parser is run the framework performs some initialisation
-tasks, setting up data structures that will later be used by actions,
-including several state tables.  Data about the connections and mails being
-processed is held in \texttt{connections} and \texttt{queueids}
-respectively.  There are also state tables used when working around
-complications in parsing: \texttt{timeout\_queueids} is used when dealing
-with connections that time out during the DATA phase (\sectionref{timeouts
-during data phase}); \texttt{bounce\_queueids} is part of the solution to
-bounce notification mails being delivered before their creation is logged
-(\sectionref{Bounce notification mails delivered before their creation is
-logged}); and \texttt{postsuper\_deleted\_queueids} caches information
-about mails recently deleted by the administrator so that subsequent log
-lines processed by the SAVE\_DATA action can be discarded (\sectionref{Mail
-deleted before delivery is attempted}).
-
-When the framework loads the ruleset, it verifies each of the rules,
-checking:
-
-\begin{itemize}
-
-    \squeezeitems{}
-
-    \item That the specified action is registered with the framework.
-
-    \item Ensuring the regex is valid and compiling it for efficiency
-        (\sectionref{Caching compiled regexes}).
-
-    \item For overlap between the data captured by the regex and additional
-        data specified in either \texttt{connection\_data} or
-        \texttt{result\_data}.
-
-    \item That \texttt{connection\_data}, \texttt{result\_data}, and the
-        regex captures specify valid fields to save data to.
-
-\end{itemize}
-
-If there are state tables from a previous \parsername{} run to be loaded,
-they will be loaded now; obviously the framework supports saving its state
-tables too.  The need to track re-injected mail (\sectionref{Re-injected
-mails}) complicates the process of saving state, as the relationships
-between mails must be maintained; loading state is complicated by dealing
-with aborted delivery attempts (\sectionref{aborted delivery attempts}),
-because a different set of relationships between connections and mails must
-be re-created when the state tables are restored.
-
-The framework will now move on to parsing each log file.  For each log line
-it will use those rules whose \texttt{program} field matches the program in
-the log line (see \sectionref{rule conditions in implementation}), falling
-back to generic rules if necessary, and warning if the log line is
-unrecognised.  The rules are sorted for increased efficiency; see
-\sectionref{rule ordering for efficiency} for details.  The repetitive
-nature of log files gives them high compression ratios; the framework
-supports reading compressed files, simplifying the process of parsing them
-because they do not have to be uncompressed to a temporary file before
-parsing begins.  The framework also displays a progress bar to show how far
-parsing has progressed through the log file, and to give the user an
-indication of how long the remainder of the parsing process is likely to
-take.  The progress bar is not as reliable when parsing compressed files
-because a compressed block may uncompress to a variable number of log
-lines; there is also variation between the recognition and processing time
-of individual log lines, but overall the progress bar is a useful addition.
-
-The framework collects data used when evaluating \parsernames{} efficiency
-for the results chapter (\sectionref{Results}); in addition some of the
-techniques used to improve parsing speed can be turned off or altered to
-show the effect they have.  Three sets of data are collected:
-
-\begin{enumerate}
-
-    \item How long it takes to parse each log file.
-
-    \item The number of rules tried when recognising log lines; the
-        framework may need to try multiple rules for a given log line
-        before it finds one that recognises it.
-
-    \item How many log lines did the framework attempt to recognise, how
-        many were skipped because there were no rules whose condition
-        matched that log line's program, how many were successfully
-        recognised, and how many were not recognised.
-
-\end{enumerate}
-
-There are five ways the framework can adapt its behaviour to demonstrate
-how effective the optimisations it uses are, and how efficient
-\parsername{} is:
-
-\begin{enumerate}
-
-    \item The rule ordering used can be changed from optimal (the default,
-        most efficient) to shuffled (intended to represent an unordered
-        ruleset) or reverse (reverse of optimal, least efficient).  The
-        results of this are shown in \sectionref{rule ordering for
-        efficiency}.
-
-    \item The framework can record which rule recognised each log line, and
-        then on a subsequent run use that information so that it uses the
-        correct rule for each log line.  This gives the best possible
-        running time, as only one rule is used for each log line.  A
-        variation of this is to use the correct rule last, to get the worst
-        possible running time.  \sectionref{perfect rule ordering}
-        discusses this in detail.
-
-    \item Each regex is compiled and the result cached when the rules are
-        loaded; this can be disabled and the regex compiled every time it
-        is used when recognising log lines, to demonstrate the effect this
-        optimisation has on parser efficiency, as shown in
-        \sectionref{Caching compiled regexes}.
-
-    \item Normally once a log line has been recognised, the framework
-        invokes the action specified by the rule.  Invoking the action can
-        be skipped if desired, so the timing information shows how long is
-        spent recognising log lines; this time can be subtracted from the
-        time taken by a normal run to show how long is spent processing log
-        lines.  This data is analysed in \sectionref{recognising vs
-        processing}.
-
-    \item The framework can skip inserting data into the database, because
-        that dominates the execution time of the parser, and the evaluation
-        in \sectionref{Results} is measuring the speed of \parsername{},
-        not the speed of the database or the disks it resides on.
-
-\end{enumerate}
-
-The framework also provides several debugging options, to aid in writing or
-correcting rules, or figuring out why the parser is not behaving as
-expected.  In increasing order of severity they are:
-
-\begin{enumerate}
-
-    \item The regex from the recognising rule and the log line it
-        recognised can be printed, so that the user can verify that the
-        correct rule recognised each log line.  A possible variant of this
-        would be to mark specific rules so only their regexes and the log
-        lines they recognise would be printed.
-
-    \item Each result can be extended with extra debugging information, so
-        that when a mail's data is dumped for inspection more information
-        is available.  The information currently added is: the log line's
-        timestamp in human readable form; the entire log line; the name of
-        the log file and the number of the log line within it.
-
-    \item Every connection and result added to the database can be dumped
-        in a human readable form.  This will result in a huge amount of
-        debugging information, so it is only useful for small log file
-        snippets, because otherwise the amount of information is
-        overwhelming.
-
-\end{enumerate}
-
-\section{Actions}
-
-\label{actions in implementation}
-
-Actions are the component of this architecture responsible for processing
-all of the inputs recognised by rules; in \parsername{} they reconstruct
-the journey each mail takes through Postfix, dealing with all the
-complications and difficulties that arise.  \parsername{} has
-\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
-distribution of rules to actions as shown in \graphref{Distribution of
-rules per action}.  Unsurprisingly, the action with the most associated
-rules is \texttt{DELIVERY\_REJECTED}, the action that processes Postfix
-rejecting a mail delivery attempt.  The next most common action is, perhaps
-surprisingly, \texttt{UNINTERESTING}: this action does nothing when
-invoked, allowing uninteresting log lines to be parsed without any effects
-(it does not imply that the input is ungrammatical or unparsed).  Generally
-rules specifying the \texttt{UNINTERESTING} action recognise log lines that
-are not associated with a specific mail, e.g.\ notices about configuration
-files changing; these log lines are recognised and processed so that the
-framework can warn about unrecognised log lines, informing the user that
-they need to add to or extend their ruleset.  Most of the remaining actions
-have only one or two associated rules: some are specialised actions
-required to address a deficiency in the log files, or a complication that
-arises during parsing; other actions will only ever have one log line
-variant, e.g.\ all log lines showing that a remote client has connected are
-recognised by a single rule and handled by the \texttt{CONNECT} action.
-\parsernames{} actions are described in detail in \sectionref{actions in
-detail in implementation}, and \sectionref{adding new actions in
-implementation} covers the process of adding a new action.
-
-\showgraph{build/graph-action-distribution}{Distribution of rules per
-action}{Distribution of rules per action}
-
-\subsection{Actions in the \parsernamelong{}}
-
-\label{actions in detail in implementation}
-
-This section explains the actions found in \parsername{}; it may help to
-revisit the flow chart in \sectionref{flow chart} to see how a mail passes
-from one action to another as its log lines are recognised.  The words
-\textit{mail\/} and \textit{connection\/} are used in the actions
-descriptions below because they are less unwieldy and more specific than
-\textit{state table entry\/}; a connection becomes a mail during the
-\texttt{CLONE} action, which processes Postfix accepting a delivery
-attempt, and the data structure moves from one state table to another
-within the parser.
-
-The complications and difficulties that arose when parsing log files
-generated on a production mail server are documented in
-\sectionref{complications}; some action descriptions refer to specific
-difficulties they address.  The complications are documented in a separate
-section to avoid overwhelming the action descriptions.
-
-If there is enough information in the log line to identify the correct
-connection or mail, each action will save all the data captured by the
-recognising rule's regex.  Each action is passed the same arguments:
-
-\begin{boldeqlist}
-
-    \squeezeitems{}
-
-    \item [rule] The recognising rule.
-
-    \item [data] The data captured from the log line by the rule's regex.
-
-    \item [line] The log line, separated into fields:
-
-        \begin{boldeqlist}
-
-            \squeezeitems{}
-
-            \item [timestamp] The time the line was logged at.
-
-            \item [program] The name of the program that logged the line.
-
-            \item [pid] The \acronym{pid} of the program that logged the
-                line.
-
-            \item [host] The hostname of the server the line was logged on.
-
-            \item [text] The remainder of the log line, i.e.\ the message
-                logged by the program and recognised by the rule.
-
-        \end{boldeqlist}
-
-\end{boldeqlist}
-
-\begin{description}
-
-    \item [BOUNCE\_CREATED] Postfix 2.3 and subsequent versions log the
-        creation of bounce messages.  This action creates a new mail if
-        necessary; if the mail already exists the unknown origin flag will
-        be removed.  The action also marks the mail as a bounce
-        notification.  To deal with complication \sectionref{Bounce
-        notification mails delivered before their creation is logged} this
-        action checks a cache of recent bounce mails, to avoid creating
-        bogus bounce mails when log lines are out of order.
-
-    \item [CLEANUP\_PROCESSING] \daemon{cleanup} processes every mail that
-        passes through Postfix; details of what it does are available in
-        \sectionref{Postfix Daemons}.  This action saves all data captured
-        by the rule's regex if the log line has not come after a timeout
-        (see \sectionref{discarding cleanup log lines}); it also creates
-        the mail if necessary, setting its unknown origin flag (see
-        \sectionref{pickup logging after cleanup}).
-
-    \item [CLONE] Multiple mails may be accepted on a single connection, so
-        each time a mail is accepted the connection's state table entry
-        must be cloned and saved in the state tables under its queueid; if
-        the original data structure was used then second and subsequent
-        mails would overwrite one another's data.
-
-    \item [COMMIT] Enter the data from the mail into the database.  Entry
-        will be postponed if the mail is a child waiting to be tracked
-        (\sectionref{Re-injected mails}).  Once entered in the database,
-        the mail will be usually be deleted from the state tables, but
-        deletion will be postponed if the mail is the parent of re-injected
-        mail (\sectionref{Re-injected mails}).
-
-    \item [CONNECT] Handle a remote client connecting: create a new
-        connection, indexed by \daemon{smtpd} \acronym{pid}.  If a
-        connection already exists it is treated as a symptom of a bug in
-        \parsername{}, and the action will issue a warning containing the
-        full contents of the existing connection plus the log line that has
-        just been parsed.
-
-    \item [DELETE] Deals with mail deleted using Postfix's administrative
-        command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to save the
-        mail to the database.  The complication this action deals with is
-        described fully in \sectionref{Mail deleted before delivery is
-        attempted}.
-
-    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
-        \acronym{SMTP} command from the remote client.  This is the action
-        most frequently specified by rules, because so many different
-        restrictions are used to reject delivery attempts.  This action is
-        quite simple: if the log line contains a queueid, save the data
-        captured by the rule's regex to the mail identified by that
-        queueid; otherwise save it to the connection identified by the
-        \acronym{pid} in the log line.  No further processing of the log
-        line is required.
-
-    \item [DISCONNECT] Invoked when the remote client disconnects, it
-        enters the connection in the database, performs any required
-        cleanup, and deletes the connection from the state tables.  This
-        action deals with aborted delivery attempts (\sectionref{aborted
-        delivery attempts}).
-
-    \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
-        sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check as
-        described in \sectionref{out of order log lines}; this action, with
-        cooperation from the COMMIT and SAVE\_DATA actions, deals with that
-        complication.
-
-    \item [MAIL\_BOUNCED] This action behaves exactly like the
-        \texttt{SAVE\_DATA} action; it saves all data captured by the
-        recognising rule's regex, and does nothing more.  It is a separate
-        action to help distinguish delivery attempts that bounce from other
-        delivery attempts.
-
-    \item [MAIL\_DISCARDED] Sometimes mail is discarded, e.g.\ mail
-        submitted locally that is larger than the limit configured by the
-        administrator.  This action is used for those mails; it invokes the
-        \texttt{COMMIT} action, but is a separate action to simplify
-        further analysis.
-
-    \item [MAIL\_QUEUED] This action represents Postfix picking a mail from
-        the queue to deliver.  This action needs to deal with out of order
-        log lines when mail is re-injected for forwarding; see
-        \sectionref{Re-injected mails} for details.
-
-    \item [MAIL\_SENT] This action behaves exactly like the
-        \texttt{SAVE\_DATA} action; it saves all data captured by the
-        recognising rule's regex, and does nothing more.  It is a separate
-        action to help distinguish successful delivery attempts from other
-        delivery attempts.
-
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a mail larger than
-        the local server accepts it will be discarded and the client
-        informed of the problem.  The mail may have been accepted and
-        partially transferred, in which case there's a data structure to
-        dispose of, otherwise there will not be.  See \sectionref{timeouts
-        during data phase} for the gory details; although that describes
-        timeouts the processing is the same for mails that are too large.
-
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  A new mail will be
-        created, although out of order log entries may have caused the
-        state table entry to already exist, as documented in
-        \sectionref{pickup logging after cleanup}.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes, requiring any active
-        connections to be cleaned up, entered in the database, and deleted
-        from the state tables.  Postfix probably kills all the other
-        components too, but \parsername{} is only affected by \daemon{smtpd}
-        processes exiting.
-
-    \item [SAVE\_DATA] Every action that can locate the correct entry in
-        the state tables saves any data captured by the recognising rule's
-        regex to it.  The \texttt{SAVE\_DATA} saves data in this way but
-        does not do anything else; it is invoked for log lines that contain
-        useful data but do not require any further processing.
-
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database if it has sufficient data,
-        and deleted from the state tables.  In some cases the connection
-        will not have enough data to satisfy the database schema, so it
-        cannot be entered into the database; unfortunately this means that
-        some data is lost.
-
-    \item [SMTPD\_KILLED] Sometimes a \daemon{smtpd} is killed by a signal;
-        this is processed in the same way as a \daemon{smtpd} dying, see
-        the \texttt{SMTPD\_DIED} action for details.
-
-    \item [SMTPD\_WATCHDOG] \daemon{smtpd} processes have a watchdog timer
-        to deal with unusual situations; after five hours the timer will
-        expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts that should occur in
-        the intervening hours, e.g.\ timeouts for DNS requests or timeouts
-        reading data from the client.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
-
-    \item [TIMEOUT] The connection with the remote client timed out so the
-        mail currently being transferred must be discarded.  The mail may
-        have been accepted, in which case there's a data structure to
-        dispose of, or it may not in which case there is not.  See
-        \sectionref{timeouts during data phase} for full details.
-
-    \item [TRACK] Track a mail when it is re-injected for forwarding to
-        another mail server; this happens when a local address is aliased
-        to a remote address (see \sectionref{tracking re-injected mail} for
-        a full explanation).  \texttt{TRACK} will be called when dealing
-        with the parent mail, and will create the child mail if necessary.
-        \texttt{TRACK} checks if the child has already been tracked, either
-        by this parent or by another parent, and issues appropriate
-        warnings in either case.
-
-    \item [UNINTERESTING] This action just returns successfully: it is used
-        when a log line needs to be parsed for completeness, but does not
-        either provide any useful data to be saved or require anything to
-        be done.
-
-\end{description}
-
-\subsection{Adding New Actions}
-
-\label{adding new actions in implementation}
-
-Adding new actions is not as easy as adding new rules, though care has been
-taken in the architecture and implementation to make adding new actions as
-painless as possible.  The implementer writes a subroutine that accepts the
-standard arguments given to actions, and registers it with the framework.
-The new action must be registered before the rules are loaded, because it
-is an error for a rule to specify an unregistered action; this helps catch
-mistakes made when adding new rules.  No other work is required from the
-implementer to integrate the action into the parser; all of their attention
-and effort can be focused on correctly implementing their new action.  The
-action may need to extend the list of valid combinations described in
-\sectionref{out of order log lines} if the new action creates a different
-set of acceptable programs, but this would only be necessary if the new
-action processes log lines from Postfix components \parsername{} currently
-does not have rules for, e.g.\ \daemon{virtual} or \daemon{lmtp}.
-
-\section{Rules}
-
-\label{rules in implementation}
-
-The rules are responsible for classifying each log line and specifying the
-correct action to be invoked.  For any parser implemented using this
-architecture, the rules are the most visible component and most likely to
-be modified by users.  Rules need to be as simple as possible so that users
-can easily modify or add to them, but the implementation must balance that
-simplicity with the need to provide enough flexibility and power to
-successfully parse the inputs.
-
-The role of the rules in the architecture is covered in detail in
-\sectionref{rules in architecture}; this section is concerned with the
-practical aspects of how rules are implemented and used in \parsername{}.
-The structure of the rules table has already been documented in
-\sectionref{rules table}; that abstract description will not be repeated
-here, but should be fleshed out by the concrete example rule in
-\sectionref{example rule in implementation}.  The process of creating new
-rules from unparsed log lines is dealt with in \sectionref{creating new
-rules in implementation}; the implementation of the utility supplied with
-\parsername{} to create regexes from unparsed log lines is also described
-in that section.  The rule conditions used in \parsername{} are explained
-next, and the final portion of this section provides some
-suggestions for how to detect overlapping rules.
-
-\subsection{Example Rule}
-
-\label{example rule in implementation}
-
-The example rule in \tableref{Example rule in implementation table}
-recognises the log line resulting from Postfix rejecting a delivery attempt
-because the domain in the sender address does not have an A, AAAA,
-or MX DNS entry, i.e.\ mail could not be delivered to the sender's address
-(for full details see~\cite{reject-unknown-sender-domain}).
-
-Example log line recognised by this rule:
-
-% RFC 3330 says that 192.0.2.0/24 is reserved for example use.
-
-\begin{verbatim}
-NOQUEUE: reject: RCPT from smtp.example.com[192.0.2.1]:
-  550 5.1.8 <foo@example.com>:
-  Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
-  proto=SMTP helo=<smtp.example.com>
-\end{verbatim}
-
-% do not reformat this!
-\begin{table}[ht]
-
-    \caption{Example rule}
-    \empty{}\label{Example rule in implementation table}
-    \begin{tabular}{ll}
-
-    \textbf{Field}      & \textbf{Value}                                    \\
-    name                & Unknown sender domain                             \\
-    description         & We do not accept mail from unknown domains        \\
-    restriction\_name   & reject\_unknown\_sender\_domain                   \\
-    program             & \daemon{smtpd}                                    \\
-    regex               & \verb!^__RESTRICTION_START__ <(__SENDER__)>: !    \\
-                        & \verb!Sender address rejected: Domain not found;! \\
-                        & \verb!from=<\k<sender>> to=<(__RECIPIENT__)> !    \\
-                        & \verb!proto=E?SMTP helo=<(__HELO__)>$!            \\
-    result\_data        &                                                   \\
-    connection\_data    &                                                   \\
-    action              & DELIVERY\_REJECTED                                \\
-    hits                & 0                                                 \\
-    hits\_total         & 0                                                 \\
-    priority            & 0                                                 \\
-    %cluster\_group      & 400                                               \\
-
-    \end{tabular}
-
-\end{table}
-
-The fields in \tableref{Example rule in implementation table} are used as
-follows:
-
-\begin{description}
-
-    \item [name, description, and restriction\_name:] are not used by the
-        parser, they serve to document the rule for the user's benefit.
-
-    \item [program and regex:] The program is used to restrict the log
-        lines this rule will be used to recognise; see \sectionref{rule
-        conditions in implementation} for details.  The regex does the
-        actual recognition of the log lines, and data captured by the regex
-        (sender, recipient, etc.) will be automatically saved to the
-        results and connections tables.
-
-    \item [action:] The action to be invoked when the rule recognises a log
-        line.  See \sectionref{actions in detail in implementation} for
-        details of the actions implemented by \parsername{}, and
-        \sectionref{actions in architecture} for the role of actions in the
-        parser architecture.
-
-    \item [result\_data and connection\_data:] are used to provide data not
-        present in the log line, and are unused in this rule.
-
-    \item [hits, hits\_total, and priority:] hits and priority are used
-        when ordering the rules for more efficient parsing (see
-        \sectionref{rule ordering for efficiency}).  At the end of each
-        parsing run hits is set to the number of log lines recognised by
-        the rule.  Hits\_total is the sum of hits over every parsing run,
-        but is otherwise unused by the parser.
-
-\end{description}
-
-\subsection{Creating New Rules}
-
-\label{creating new rules in implementation}
-
-The log files produced by Postfix differ from installation to installation,
-because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different
-\acronym{DNSBL} services, policy servers, or custom rejection messages.  To
-facilitate parsing new log lines, the architecture separates rules from
-actions: adding new actions can be difficult, but adding new rules to
-recognise new log lines is trivial, and occurs much more frequently.
-
-To add a new rule a new row must be added to the rules table in the
-database, containing all the required attributes: action, name,
-description, program, and regex.  The name and description fields should be
-set based on the meaning of the log line, to help others understand which
-log lines this rule will recognise; the program will be obvious from the
-unrecognised log lines.  The action depends on the what the log line
-represents, e.g.\ a delivery rejection, a mail being delivered, some useful
-information, or something else; examine the list of actions in
-\sectionref{actions in detail in implementation} to determine the correct
-one.  The regex will be constructed based on the log line, but see below
-for a tool to ease the process.
-
-There are other attributes that may be required in a rule:
-connection\_data, result\_data, priority, or restriction\_name.  In general
-it will only become clear that connection\_data or result\_data are
-required when \parsername{} warns about an entry in the connections or
-results tables that is missing some required fields, because values for the
-missing fields are not present in any of the log lines.  E.g.\ the rule
-recognising the \daemon{pickup} component processing a mail sets the
-client\_hostname to \texttt{localhost} and the client\_ip to
-\texttt{127.0.0.1}, because the mail originates on the local server.  If
-the new rule deliberately uses the architecture's overlapping rules feature
-it will be obvious that the priority field needs to be set, on this rule and
-possibly others; the priority field may be needed on unintentionally
-overlapping rules too, but that is more difficult to determine.  Finally
-the restriction\_name field should be set when the rule's action is
-\texttt{DELIVERY\_REJECTED}; hopefully the restriction in question will be
-clear from the content of the log line.
-
-A program is provided with \parsername{} to ease the process of creating
-new rules from unrecognised log lines, based on the algorithm developed by
-Risto Vaarandi for his \acronym{SLCT}~\cite{slct-paper}.  The differences
-between the two algorithms will be outlined as part of the general
-explanation below.  The core of the \acronym{SLCT} algorithm is quite
-simple: log lines are generally created by substituting variable words into
-a fixed pattern, and analysis of the frequency with which each word occurs
-can be used to determine whether the word is variable or part of the fixed
-pattern.  This classification can be used to group similar log lines and
-generate a regex to match each group of log lines.  There are 4 steps in
-the algorithm:
-
-\begin{description}
-
-    \item [Pre-process the file.]  The new algorithm leverages the
-        knowledge gained when writing rules and developing \parsername{};
-        it performs a large number of substitutions on the input log lines,
-        replacing commonly occurring variable terms (e.g.\ email addresses,
-        \acronym{IP} addresses, the standard start of rejection messages)
-        with keywords that \parsername{} will expand when it loads the new
-        rule.  The purpose of this step is to utilise existing knowledge to
-        create more accurate regexes; it replaces a large number of
-        variable words with fixed words, improving the subsequent
-        classification of words as fixed or variable.  The altered log
-        lines are written to a temporary file, which subsequent stages will
-        use instead of the original input file.
-
-        In the original algorithm the purpose of the pre-processing stage
-        was to reduce the memory consumption of the program.  In the first
-        pass it generates a hash~\cite{hash-functions}\glsadd{hash} (from a
-        small range of values) for each word of each log line, incrementing
-        a counter for each hash.  The counters will be used in the next
-        pass to filter out words: if the word's hash does not have a high
-        frequency, the word itself cannot have a high frequency, so there
-        is no need to maintain a counter for it, reducing the number of
-        counters required and thus the program's memory consumption.
-
-    \item [Calculate word frequencies.]  The position of words within a log
-        line is important: a word occurring in two log lines does not
-        indicate similarity unless it occupies the same position in both
-        log lines.  If a variable term within a line contains
-        spaces, it will appear to the algorithm as more than one word.
-        This will alter the position of subsequent words, so a word
-        occurring in different positions in two log lines \textit{may\/}
-        indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.  The algorithm maintains a counter for each
-        \texttt{(word, word's position within the log line)} tuple,
-        incrementing it each time that word occurs in that position.
-
-        The original algorithm hashes each word and checks that result's
-        counter from the previous pass; if that counter has a high
-        frequency a separate counter will be maintained for this word in
-        this position.  This reduces the number of counters maintained by
-        the algorithm in this step, reducing the memory requirements of the
-        algorithm at a cost of increased CPU usage.
-
-        As time goes on the amount of memory typically available increases
-        and the need to reduce memory requirements decreases, so the
-        modified algorithm omits the hashing step and maintains counters
-        for all tokens.  In addition most of the infrequently occurring
-        words will have been substituted with keywords during the first
-        step, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency.]  The frequency of
-        every tuple \texttt{(word, word's position within the log line)} is
-        checked: if its frequency is greater than the threshold supplied by
-        the user (1\% of all log lines is generally a good starting point),
-        it is classified as a fixed word, otherwise it is classified as a
-        variable term.  If a variable term appears sufficiently often it
-        will be misclassified as a fixed term, but that should be noticed
-        by the user when reviewing the new regexes.  Variable terms are
-        replaced by \texttt{.+} to match one or more of any character.
-
-    \item [Build regexes.]  The words are reassembled to produce a regex
-        matching the log line, and a counter is maintained for each regex.
-        Contiguous sequences of \texttt{.+} in the newly reassembled
-        regexes are collapsed to a single \texttt{.+}; any resulting
-        duplicate regexes are combined, and their counters added together.
-        If a regex's counter is lower than the threshold supplied by the
-        user the regex is discarded; this second threshold is independent
-        of the threshold used to differentiate between fixed and variable
-        words, but once again 1\% of log lines is a good starting point.
-        The new regexes are displayed for the user to add to the database,
-        either as new rules or merged into the regexes of existing rules;
-        the counter for each regex is also displayed, giving the user an
-        indication of how many of the log lines that regex should match.
-        Discarding regexes will result in some of the log lines not being
-        matched; when the rules have been augmented with the new regexes,
-        the original log files should be parsed again, and any remaining
-        unparsed log lines used as input to this utility.
-
-\end{description}
-
-XXX MERGE logs2regexes WITH check-regexes.
-
-A second utility is also provided that reads a list of new regexes and
-the input given to the first utility.  It tries to match each input log
-line against each regex, counting the number of log lines that match
-each regex, warning the user if an input log line is matched by more
-than one regex, and additionally warning if an input log line is not
-matched by any regex.  It displays a summary of how many input log lines
-each regex matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the regexes
-produced by the first utility are correctly matching the input log lines
-they are based upon.
-
-These utilities are not intended to create perfect regexes, but they
-greatly reduce the effort required to parse new or different log lines.
-
-\subsection{Rule Conditions}
-
-\label{rule conditions in implementation}
-
-Rule conditions as part of the architecture have already been documented in
-\sectionref{rule conditions in architecture}, and they can be very complex
-and difficult to evaluate.  In contrast, \parsername{} uses very simple
-rule conditions: each rule has a \texttt{program} attribute that specifies
-the Postfix component whose log lines it recognises, and each log line
-contains the name of the Postfix component that produced it; the framework
-will only use rules where the two are equal when trying to recognise log
-lines.  This avoids needlessly trying rules that will not match the log
-line, or worse, might match unintentionally.  In addition the framework
-supports generic rules, whose program attribute is ``*''; these will be
-used if none of the program-specific rules recognise the log line.  If none
-of the rules are successful the framework will warn the user.
-
-\subsection{Regex Components}
-
-\label{regex components}
-
-Each rule's regex will have keywords expanded when the rules are loaded,
-for several reasons:
-
-\begin{itemize}
-
-    \item It simplifies both reading and writing of regexes and makes each
-        regex largely self-documenting.  E.g.\ the meaning of
-        \_\_CLIENT\_HOSTNAME\_\_ is immediately clear, whereas
-        \verb!(?:unknown|(?:[-.\w]+))! needs to be deciphered.
-
-    \item It avoids needless repetition of complex regex components, and
-        allows the components to be corrected or improved in one
-        location.  E.g.\ \_\_SENDER\_\_ is used in 68 rules; if a mistake
-        is discovered it can be corrected in one place only.
-
-    \item It enables automatic extraction and saving of captured data.  The
-        regex snippets use Perl 5.10's named capture buffers~\cite{perlre},
-        so the mapping between captures and fields does not need to be
-        explicitly specified by the rule.
-
-\end{itemize}
-
-For efficiency the keywords are expanded and every rule's regex is compiled
-before attempting to parse the log file, otherwise every regex would be
-recompiled each time it was used, resulting in a large, data dependent
-slowdown, as shown in \sectionref{Caching compiled regexes}.  Most of the
-keywords are named after the fields in the connections or results tables
-they populate: \_\_CLIENT\_HOSTNAME\_\_, \_\_CLIENT\_IP\_\_, \_\_DELAY\_\_,
-\_\_DELAYS\_\_, \_\_ENHANCED\_STATUS\_CODE\_\_, \_\_HELO\_\_,
-\_\_MESSAGE\_ID\_\_, \_\_QUEUEID\_\_, \_\_RECIPIENT\_\_, \_\_SENDER\_\_,
-\_\_SERVER\_HOSTNAME\_\_, \_\_SERVER\_IP\_\_, \_\_SIZE\_\_, and
-\_\_SMTP\_CODE\_\_.
-
-Some of the keywords need more explanation:
-
-\begin{eqlist}
-
-    \squeezeitems{}
-
-    \item [\_\_CHILD\_\_]  The queueid of a child mail; see
-        \sectionref{Re-injected mails}.
-
-    \item [\_\_COMMAND\_\_]  All \acronym{SMTP} commands, except
-        \texttt{DATA}, which typically has a more specific rule.
-        Priorities could have been used instead of excluding \texttt{DATA}.
-
-    \item [\_\_CONN\_USE\_\_]  How many times the connection was reused;
-        Postfix tries to reuse connections whenever possible to reduce the
-        load on both the sending and receiving servers.
-
-    \item [\_\_DATA\_\_]  This snippet is special: it matches a zero-length
-        string, so it needs to be followed by a pattern that matches
-        appropriately, but it captures whatever is matched so that it can
-        be saved.  E.g.\ \verb!/(__DATA__connection (?:refused|reset!
-        \newline{} \tab{}\verb!by peer))/! matches either ``connection
-        refused'' or ``connection reset by peer'' and causes it to be saved
-        to the data field of that log line's result.
-
-    \item [\_\_PID\_\_]  The \acronym{pid} of a \daemon{smtpd} process that
-        dies or is killed; see the SMTPD\_DIED action in
-        \sectionref{actions in detail in implementation}.
-
-    \item [\_\_RESTRICTION\_START\_\_]  Matches the standard information
-        Postfix includes at the start of almost all log lines resulting
-        from rejecting a delivery attempt.
-
-    \item [\_\_SHORT\_CMD\_\_]  Postfix sometimes logs \acronym{SMTP}
-        commands in a short, single word form.
-
-\end{eqlist}
-
-\subsection{Overlapping Rules}
-
-\label{overlapping rules in implementation}
-
-\parsername{} does not try to detect overlapping rules; that responsibility
-is left to the author of the rules.  The advantages and difficulties of
-overlapping rules have already been addressed in \sectionref{overlapping
-rules in architecture} and will not be repeated here.  \parsername{}
-provides a mechanism for ordering overlapping rules: the priority field in
-each rule (defaulting to zero); negative priorities may be useful for
-catchall rules.
-
-Detecting overlapping rules is difficult, but the following approaches may
-be helpful:
-
-\begin{itemize}
-
-    \item Sort by program and regex, then visually inspect the list, e.g.\
-        with \acronym{SQL} similar to:      \newline{}
-        \verb!SELECT program, regex!        \newline{}
-        \verb!    FROM rules!               \newline{}
-        \verb!    ORDER BY program, regex;! \newline{}
-        The rules are sorted first by program, then by regex, because rules
-        with overlapping regexes will not overlap if their programs are
-        different.  Note that this does not properly deal with rules whose
-        program is \texttt{*}; those rules will be used on all log lines
-        that have not been recognised by program-specific rules.
-
-    \item Compare the results of parsing using different rule orderings
-        (described in \sectionref{rule ordering for efficiency}).  Parse
-        several log files using optimal ordering, then dump a textual
-        representation of the rules, connections, and results tables.
-        Repeat with shuffled and reversed ordering, starting with a fresh
-        database.  If there are no overlapping rules the tables from each
-        run will be identical; differences indicate overlapping rules.  The
-        rules that overlap can be determined by examining the differences
-        in the tables: each result contains a reference to the rule which
-        created it, which will change if that rule overlaps with another.
-        Unfortunately this method cannot prove the absence of overlapping
-        rules; it can detect overlapping rules, but only if there are log
-        lines in the log files that are recognised by more than one rule.
-
-\end{itemize}
-
 \section{Complications Encountered}
@@
 database and add the log line to it.
-
-\section{Limitations and Possible Improvements}
-
-\label{limitations and improvements in implementation}
-
-Every piece of software suffers from some limitations, and there is almost
-always room for improvement.  Below are the limitations and possible
-improvements that have been identified in \parsername{}.
-
-\subsection{Limitations}
-
-\label{logging helo}
-
-\begin{enumerate}
-
-    \item Each new Postfix release requires writing new rules or modifying
-        existing rules to cope with the new or changed log lines.
-        Similarly using a new \acronym{DNSBL}, a new policy server, or new
-        administrator-defined rejection messages requires new rules.
-
-    \item The hostname used in the HELO command is not logged if the
-        incoming delivery attempt is successful.  Configuring Postfix to do
-        this is relatively simple: create a restriction that is guaranteed
-        to warn for every accepted mail, as follows:
-
-        \begin{enumerate}
-
-            \item Create \texttt{/etc/postfix/log\_helo.pcre}
-                containing:\newline{}
-                \tab{}\texttt{/\^/~~~~WARN~Logging~HELO}
-
-            \item Modify \texttt{smtpd\_data\_restrictions} in
-                \texttt{/etc/postfix/main.cf} to contain:\newline{}
-                \tab{}\texttt{check\_helo\_access~pcre:/etc/postfix/log\_helo.pcre}
-
-        \end{enumerate}
-
-        Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there will not yet be a queueid
-        associated with the mail when \texttt{smtpd\_helo\_restrictions} is
-        evaluated for the first recipient, so the log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus the
-        queueid will be logged by any restrictions taking effect in
-        \texttt{smtpd\_data\_restrictions}, and the log line can be
-        associated with the correct mail.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
-
-        Logging the HELO hostname in this fashion also partially prevents
-        the complication described in \sectionref{Mail deleted before
-        delivery is attempted} from occurring, but only when there is a
-        single recipient; in that case the recipient address will be logged
-        also, but when there are multiple recipients no addresses are
-        logged.
-
-    \item \parsername{} does not create separate mails where one or more
-        delivery attempts are rejected and subsequently a mail is accepted;
-        it will appear in the database as one mail with lots of rejections
-        followed by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It does not appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
-    \item \parsername{} will not detect that it is parsing the same log file
-        twice, resulting in the database containing duplicate entries.
-
-    \item \parsername{} does not distinguish between log files produced by
-        different sources when parsing; all results will be saved to the
-        same database.  This may be viewed as an advantage, because log
-        files from different sources can be combined in the same database,
-        or it may be viewed as a limitation because there is no facility to
-        distinguish between log files from different sources in the same
-        database.  If the results of parsing log files from different
-        sources must remain separate, the parser can easily be instructed
-        to use a different database.
-
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
-\end{enumerate}
-
-\subsection{Possible Improvements}
-
-\begin{enumerate}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
-
-\end{enumerate}
-
-\section{Summary}
-
-XXX TO BE WRITTEN\@.  START WITH THE OLD CONTENT BELOW\@.
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications that arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they are not specified anywhere in the parser.
-The framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create regexes for new rules.
-Detecting, diagnosing, and defeating complications forms the largest single
-portion of this section, mirroring the development of the parser.  The
-complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.

### 2009-03-25 doc/logparser-implementation.tex
Moved complications to a separate file, because of their size.

@@
 
-\section{Complications Encountered}
-
-\label{complications}
-
-It was initially expected that parsing Postfix log files would be a
-relatively simple task, requiring a couple of months of work.  The author
-had found Postfix log files useful when investigating problems reported by
-users, and an examination of the log files gave the impression that they
-would be straightforward to parse, process, and understand.  The large
-variation in log lines was not apparent, because the majority of log lines
-are recognised by a small set of rules, as shown in \figureref{rule hits
-graph}.  Most of the myriad complications and difficulties documented in
-this section were discovered during \parsernames{} development, but the
-first three complications were identified during the planning and design
-phase of this project, influencing the architecture's design.
-
-Each of these complications caused \parsername{} to operate incorrectly,
-generate warning messages, or leave mails in the state table.  The
-complications are listed in the order in which they were overcome during
-development of \parsername{}, with the first complication occurring several
-orders of magnitude more frequently than the last.  When deciding which
-problem to tackle next, the problem causing the highest number of warning
-messages or mails incorrectly remaining in the state tables was always
-chosen, because that approach yielded the biggest improvement in the
-parser, and made the remaining problems more apparent.
-
-\subsection{Queueid Vs Pid}
-
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \acronym{pid}, then transition to
-being correlated by the queueid.  This is relatively minor, but does
-require:
-
-\begin{itemize}
-
-    \item Two versions of several functions: \texttt{by\_pid} and
-        \texttt{by\_queueid}.
-
-    \item Two state tables to hold the data structure for each connection
-        and mail.
-
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \acronym{pid} or queueid.
-
-\end{itemize}
-
-\subsection{Connection Reuse}
-
-\label{connection reuse}
-
-Multiple independent mails may be delivered across one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails will not trample over each other's data.
-This must be done every time a mail is accepted, as it is impossible to
-tell in advance which connections will accept multiple mails.  Once the
-mail has been accepted its log lines will not be correlated by
-\acronym{pid} any more, its queueid will be used instead (except when
-timeouts occur during the data phase \sectionref{timeouts during data
-phase}).  If the original connection has any useful data (e.g.\ rejections)
-it will be saved to the database when the client disconnects.  One unsolved
-difficulty is distinguishing between different groups of rejections, e.g.\
-when dealing with the following sequence:
-
-\begin{enumerate}
-
-    \item The client attempts to deliver a mail, but it is rejected.
-
-    \item The client issues the RSET command to reset the \acronym{SMTP}
-        session.
-
-    \item The client attempts to deliver another mail, likewise rejected.
-
-\end{enumerate}
-
-There should ideally be two separate entries in the database resulting from
-the above sequence, but currently there will only be one.
-
-\subsection{Re-injected Mails}
-
-\label{Re-injected mails}
-
-\label{tracking re-injected mail}
-
-Mails sent to local addresses are not always delivered directly to a
-mailbox: sometimes they are sent to and accepted for a local address, but
-need to be delivered to one or more remote addresses due to aliasing.  When
-this occurs, a child mail will be injected into the Postfix queue, but
-without the explicit logging that mails injected by \daemon{smtpd} or
-\daemon{postdrop} have.  Thus the source of the mail is not immediately
-discernible from the log line in which the mail's queueid first appears:
-from a strictly chronological reading of the log lines it usually appears
-as if the child mail has been created without a source.  Subsequently the
-parent mail will log the creation of the child mail, e.g.\ parent mail
-3FF7C4317 creates child mail 56F5B43FD\@:
-
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \hfill{}
-\newline{} \tab{} \tab{} delay=0, status=sent (forwarded as 56F5B43FD)}
-
-Unfortunately, while all log lines from an individual process appear in
-chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the example log
-line above belongs to the parent mail) is logged by \daemon{qmgr} or
-\daemon{cleanup}, so the order also depends on how soon they process the
-new mail.
-
-Because of the uncertain order that log lines can appear in, \parsername{}
-cannot complain when it encounters a log line from either \daemon{qmgr} or
-\daemon{cleanup} for a previously unseen mail; instead it must flag the
-mail as coming from an unknown origin, and subsequently clear that flag if
-and when the origin of the mail becomes clear.  Obviously the parser could
-omit checking where mails originate from, but requiring an explicit source
-helps to expose bugs in the parser; such checks helped to identify the
-complications described in \sectionref{discarding cleanup log lines} and
-\sectionref{pickup logging after cleanup}.
-
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered, and entirely finished with,
-\textit{before\/} the parent mail logs the creation log line!  Thus, mails
-flagged as coming from an unknown origin cannot be entered into the
-database when their final log line is processed; instead they must be
-marked as ready for entry and subsequently entered once their origin has
-been identified.  The crux of this complication is that re-injected mails
-appear in the log files without explicit logging indicating their source.
-
-Tracking re-injected mail requires \parsername{} to do the following in the
-COMMIT action:
-
-\begin{enumerate}
-
-    \item If a mail is tagged with the unknown origin flag it is assumed to
-        be a child mail whose parent has not yet been identified.  The mail
-        is tagged as ready to be entered in the database, but entry is
-        postponed until the parent is identified.  The child mail should
-        not have any subsequent log lines: only its parent should refer to
-        it.
-
-    \item If the mail is a child mail with an identified parent it is
-        entered in the database as usual, then removed from its parent's
-        list of children.  If this child is the last mail on that list, and
-        the parent has already been entered in the database, the parent
-        will be removed from the state tables.
-
-    \item If the mail is a parent it is entered in the database as usual,
-        because there will be no further log lines for it.  There may be
-        child mails waiting to be entered in the database; these are
-        entered as normal, and removed from the state tables.  If there are
-        more child mails, the parent's removal from the state tables will
-        be postponed until the last child has been entered.
-
-\end{enumerate}
-
-\subsection{Identifying Bounce Notifications}
-
-\label{identifying bounce notifications}
-
-Postfix 2.2.x (and presumably previous versions) does not generate a log
-line when it generates a bounce notification; suddenly there will be log
-entries for a mail that lacks an obvious source.  There are similarities to
-the problem of identifying re-injected mails discussed in
-\sectionref{tracking re-injected mail}, but unlike the solution described
-therein bounce notifications do not eventually have a log line that
-identifies their source.  Heuristics must be used to identify bounce
-notifications, and those heuristics are:
-
-\begin{enumerate}
-
-    \item The sender address is \verb!<>!.\glsadd{<>}
-
-    \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \acronym{SMTP} or submitted
-        locally by \daemon{postdrop}.
-
-    \item The message-id has a specific format: \newline{}
-        \tab{} \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline{}
-        e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
-
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) it will not match the queueid of
-        the mail.
-
-\end{enumerate}
-
-Once a mail has been identified as a bounce notification, the unknown
-origin flag is cleared and the mail can be entered in the database.
-
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
-occur the following conditions would have to be met:
-
-\begin{enumerate}
-
-    \item The mail must have been generated internally by Postfix.
-
-    \item The sender address must be \verb!<>!.\glsadd{<>}
-
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
-
-\end{enumerate}
-
-If a mail is misclassified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably misclassification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.  Postfix 2.3 (and
-hopefully subsequent versions) log the creation of a bounce message.
-
-This check is performed during the COMMIT action.
-
-\subsection{Aborted Delivery Attempts}
-
-\label{aborted delivery attempts}
-
-Some mail clients behave unexpectedly during the \acronym{SMTP} dialogue:
-the client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
-it continues with until delivery is complete.  Microsoft Outlook is one
-client that behaves in this fashion; other clients may act in a similar
-way.  An example dialogue exhibiting this behaviour is presented below
-(lines starting with a three digit number are sent by the server, the other
-lines are sent by the client):
-
-\begin{verbatim}
-220 smtp.example.com ESMTP
-EHLO client.example.com
-250-smtp.example.com
-250-PIPELINING
-250-SIZE 15240000
-250-ENHANCEDSTATUSCODES
-250-8BITMIME
-250 DSN
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-RSET
-250 2.0.0 Ok
-RSET
-250 2.0.0 Ok
-MAIL FROM: <sender@example.com>
-250 2.1.0 Ok
-RCPT TO: <recipient@example.net>
-250 2.1.5 Ok
-DATA
-354 End data with <CR><LF>.<CR><LF>
-The mail transfer is not shown.
-250 2.0.0 Ok: queued as 880223FA69
-QUIT
-221 2.0.0 Bye
-\end{verbatim}
-
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
-for the following:
-
-\begin{itemize}
-
-    \item Was the second result processed by the \texttt{CLONE} action?
-        The first two \daemon{smtpd} log lines will be a connection log
-        line and a mail acceptance log line; Postfix logs acceptance as
-        soon as the first recipient has been accepted.
-
-    \item Is \daemon{smtpd} the only Postfix component that produced a log
-        line for this mail?  Every mail which passes normally through
-        Postfix will have a \daemon{cleanup} line, and later a
-        \daemon{qmgr} log line; lack of a \daemon{cleanup} line is a sure
-        sign the mail did not make it too far.
-
-    \item Does the queueid exist in the state tables?  If not it cannot be
-        an aborted delivery attempt.
-
-    \item If there are third and subsequent results, were those log lines
-        processed by the \texttt{SAVE\_DATA} action?  If there are any log
-        lines after the first two they should be informational only.
-
-\end{itemize}
-
-If all the checks above are successful then the mail is assumed to be an
-aborted delivery attempt and is discarded.  There will be no further
-entries logged for such mails, so without identifying and discarding them
-they accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \acronym{IP} address, but the database
-schema requires many more fields be populated (see \sectionref{connections
-table} and \sectionref{results table}).  This heuristic is quite
-restrictive, and it appears there is little scope for false positives; if
-there are any false positives there will be warnings when the next log line
-for that mail is parsed.  False negatives are less likely to be detected:
-there may be queueid clashes (and warnings) if mails remain in the state
-tables after they should have been removed, otherwise the only way to
-detect false negatives is to examine the state tables after each parsing
-run.
-
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
-
-\subsection{Further Aborted Delivery Attempts}
-
-Some mail clients disconnect abruptly if a second or subsequent recipient
-is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts during data phase}).  Sadly, Postfix does not
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
-
-\begin{itemize}
-
-    \item Is the mail missing its \daemon{cleanup} log line?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail did not
-        make it too far.
-
-    \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection log line and a mail accepted log line,
-        followed by one or more rejection log lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
-
-\end{itemize}
-
-If all checks are successful then the mail is assumed to have been
-discarded when the client disconnected.  There will be no further entries
-logged for such mails, so without identifying and entering them in the
-database immediately they accumulate in the state table and will cause
-clashes if the queueid is reused.
-
-These checks are made during the DISCONNECT action.
-
-\subsection{Timeouts During DATA Phase}
-
-\label{timeouts during data phase}
-
-The DATA phase of the \acronym{SMTP} conversation is where the headers and
-body of the mail are transferred.  Sometimes there is a timeout or the
-connection is lost during the DATA phase;\footnote{For brevity's sake
-\textit{timeout\/} will be used throughout this section, but everything
-applies equally to lost connections.} when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
-mail.  It seems more intuitive to save the mail's data to the database, but
-if a timeout occurs there is no data available to save; the timeout is
-recorded and saved with the connection instead.  To deal properly with
-timeouts the TIMEOUT action does the following:
-
-\begin{enumerate}
-
-    \item Record the timeout and associated data in the connection's
-        results.
-
-    \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
-
-    \item If one or more recipients have been accepted, Postfix will have
-        allocated a queueid for the incoming mail, and there may be a mail
-        in the state tables that needs to be discarded.  The timeout may
-        have interrupted transfer of an accepted delivery attempt, or it
-        may have occurred after a mail delivery attempt was rejected.  To
-        distinguish between the two cases the action compares the timestamp
-        of the last accepted mail against the timestamp of the last log
-        line from \daemon{smtpd} for that connection; this depends on the
-        CLONE action keeping a list of all mails accepted for each
-        connection.  If the mail acceptance timestamp is later than the
-        timeout timestamp, the timeout applies to the just-accepted mail,
-        which will be removed from the state tables, without entering it in
-        the database.  If the \daemon{smtpd} timestamp is later there was a
-        rejection between the accepted mail and the timeout: the action
-        assumes that the timeout applies to the rejected delivery attempt
-        and finishes.  This assumption is not necessarily correct, because
-        Postfix may have accepted an earlier recipient and rejected a later
-        one, in which case the timeout applies to the accepted mail, which
-        should be discarded; this has not been a problem in practice.
-        Processing timeouts is further complicated by the presence of out
-        of order \daemon{cleanup} log lines: see \sectionref{discarding
-        cleanup log lines} for details.
-
-\end{enumerate}
-
-This complication is dealt with by the TIMEOUT action, with help from the
-CLONE action.
-
-\subsection{Discarding Cleanup Log Lines}
-
-\label{discarding cleanup log lines}
-
-The author has only observed this complication occurring after a timeout,
-though there may be other circumstances that trigger it.  Sometimes the
-\daemon{cleanup} line for a mail being accepted is logged after the timeout
-line; parsing this line causes the MAIL\_QUEUED action to create a new
-state table entry for the queueid in the log line.  This is incorrect
-because the line actually belongs to the mail that has just been discarded;
-the next log line for that queueid will be seen when the queueid is reused
-for a different mail, causing a queueid clash and the appropriate warning.
-
-When the \daemon{cleanup} line is still pending during the TIMEOUT action,
-the action updates a global list of queueids, adding the queueid and the
-timestamp from the log line.  When the next \daemon{cleanup} line is parsed
-for that queueid the list will be checked (during the MAIL\_QUEUED action),
-and the log line will be deemed part of the mail where the timeout occurred
-and discarded if it meets the following requirements:
-
-\begin{itemize}
-
-    \item The queueid must not have been reused yet, i.e.\ it does not have
-        an entry in the state tables.
-
-    \item The timestamp of the \daemon{cleanup} log line must be within ten
-        minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives (log lines incorrectly
-        discarded) and false negatives (new state table entries incorrectly
-        created).
-
-\end{itemize}
-
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because some, but not all connections where a timeout occurs will
-have an associated \daemon{cleanup} line logged; if the algorithm blindly
-discarded the next \daemon{cleanup} line after a timeout it would sometimes
-be mistaken.  When the next \daemon{pickup} line containing that queueid is
-parsed the queueid will be removed from the cache of timeout queueids,
-regardless of whether it meets the criteria above.
-
-This complication is handled by the TIMEOUT and MAIL\_QUEUED actions.
-
-\subsection{Pickup Logging After Cleanup}
-
-\label{pickup logging after cleanup}
-
-When mail is submitted locally, \daemon{pickup} accepts the new mail and
-generates a log line showing the source.  Occasionally this log line will
-occur later in the log file than the \daemon{cleanup} log line, so the
-PICKUP action will find that a state table entry exists for that queueid.
-Normally if the queueid given in the PICKUP line exists in the state tables
-a warning is generated by the \daemon{pickup} action, but if the following
-conditions are met it is assumed that the log lines were out of order:
-
-\begin{itemize}
-
-    \item The only program which has logged anything thus far for the mail
-        is \daemon{cleanup}.
-
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} log lines.
-
-\end{itemize}
-
-As always with heuristics there may be circumstances in which these
-heuristics match incorrectly, but none have been identified so far.  This
-complication seems to occur during periods of particularly heavy load, so
-is most likely caused by process scheduling vagaries.
-
-This complication is dealt with during the PICKUP action.
-
-\subsection{Smtpd Stops Logging}
-
-\label{smtpd stops logging}
-
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over log files for some time
-several reasons have been found for this infrequent occurrence:
-
-\begin{enumerate}
-
-    \item Postfix is stopped or its configuration is reloaded.  When this
-        happens all \daemon{smtpd} processes exit, so all entries in the
-        connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.
-
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
-
-    \item Every Postfix process uses a watchdog which kills the process if
-        it is not reset for a considerable period of time (five hours by
-        default).  This safeguard prevents Postfix processes from running
-        indefinitely and consuming resources if a failure causes them to
-        enter a stuck state.
-
-\end{enumerate}
-
-The circumstances above account for all occasions identified thus far where
-a \daemon{smtpd} suddenly stops logging.  In addition to removing an active
-connection the last accepted mail may need to be discarded, as detailed in
-\sectionref{timeouts during data phase}; otherwise the queueid state table
-is untouched.
-
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED, and SMTPD\_WATCHDOG\@.
-
-\subsection{Out of Order Log Lines}
-
-\label{out of order log lines}
-
-Occasionally a log file will have out of order log lines which cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup log lines} or \sectionref{pickup
-logging after cleanup}.  In the \numberOFlogFILES{} log files used for
-testing this occurs only five times in 60,721,709 log lines, but for parser
-correctness it must be dealt with.  The five occurrences have the same
-characteristics: the \daemon{local} log line showing delivery to a local
-mailbox occurs after the \daemon{qmgr} log line showing removal of the mail
-from the queue because delivery is completed.  This causes problems: the
-data in the state tables for the mail is not complete, so entering it into
-the database fails; a new mail is created when the \daemon{local} line is
-parsed and remains in the state tables; four warnings are issued per pair
-of out of order log lines.
-
-The COMMIT action examines the list of programs that have produced a log
-lines for each mail, comparing the list against a table of known-good
-program combinations.  If the mail's combination is found in the table the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The SAVE\_DATA
-action checks for that flag; if the additional log lines have caused the
-mail to reach a valid combination then entry in the database will proceed,
-otherwise it must be postponed once more.
-
-The list of valid combinations is explained below.  Every mail will
-additionally have log line from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log line from \daemon{bounce}, \daemon{postsuper}, or
-both.
-
-% This will put the text on the line following the item name, if the
-% enumitem package is loaded.
-%\begin{description}[style=nextline]
-\begin{description}
-
-    \item [\daemon{local}:] Local delivery of a bounce notification, or
-        local delivery of a re-injected mail.
-
-    \item [\daemon{local}, \daemon{pickup}:] Mail submitted locally on the
-        server, delivered locally on the server.
-
-    \item [\daemon{local}, \daemon{pickup}, \daemon{smtp}:] Mail submitted
-        locally \newline{} on the server, for both local and remote
-        delivery.
-
-    \item [\daemon{local}, \daemon{smtp}, \daemon{smtpd}:] Mail accepted
-        from a remote client, for both local and remote delivery.
-
-    \item [\daemon{local}, \daemon{smtpd}:] Mail accepted from a remote
-        client, for local delivery only.
-
-    \item [\daemon{pickup}, \daemon{smtp}:] Mail submitted locally on the
-        server, for remote delivery only.
-
-    \item [\daemon{smtp}:] Remote delivery of either a re-injected mail or
-        a bounce notification.
-
-    \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network to addresses outside the local
-        network).
-
-    \item [\daemon{smtpd}, \daemon{postsuper}:] Mail accepted from a remote
-        client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job\glsadd{Joe-job}).  Notice that \daemon{postsuper} is
-        required, not optional, for this combination.
-
-\end{description}
-
-This check applies to accepted mails only, not to rejected mails.  This
-check is performed during the COMMIT action.
-
-\subsection{Yet More Aborted Delivery Attempts}
-
-\label{yet more aborted delivery attempts}
-
-The aborted delivery attempts described in \sectionref{aborted delivery
-attempts} occur frequently, but the aborted delivery attempts described in
-this section only occur four times in the \numberOFlogFILES{} log files
-used for testing.  The symptoms are the same as in \sectionref{aborted
-delivery attempts}, except that there \textit{is\/} a \daemon{cleanup} log
-line; there is nothing in the log file to explain why there are no further
-log lines.  The only way to detect these mails is to periodically scan all
-mails in the state tables, deleting any mails displaying the following
-characteristics:
-
-\begin{itemize}
-
-    \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed from the current log
-        file.
-
-    \item There must be exactly two \daemon{smtpd} and one \daemon{cleanup}
-        log lines for the mail, with no additional log lines.
-
-\end{itemize}
-
-12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue (unless Postfix is not
-running for an extended period of time).  The state tables are scanned for
-mails matching the characteristics above each time the end of a log file is
-reached, and matching mails are deleted.
-
-\subsection{Mail Deleted Before Delivery is Attempted}
-
-\label{Mail deleted before delivery is attempted}
-
-Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
-faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it is the result of a mail loop\glsadd{mail
-loop}, mail bomb\glsadd{mail bomb}, or joe~job\glsadd{Joe-job}) before
-Postfix has had a chance to try to deliver it.  In this case the recipient
-address will not have been logged, so a dummy recipient address needs to be
-added, as every mail is required by the database schema
-(\sectionref{results table}) to have at least one recipient.  Typically
-when this complication occurs there are many instances of it, closely
-grouped.
-
-This lack of information cannot easily be overcome: it is trivial to log a
-warning for every recipient accepted, but when the warning for the first
-recipient is logged Postfix will not yet have allocated a queue file and
-queueid for the mail, so the warning will be associated with the connection
-rather than the accepted mail.  A queue file and queueid will be allocated
-after Postfix accepts the MAIL FROM command if
-\texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is set to ``no'', but that
-setting will cause disk IO for almost every delivery attempt, instead of
-just for delivery attempts where recipients are accepted, and consequently
-a drastic reduction in the performance of the mail server.
-
-The DELETE action is responsible for handling this complication.
-
-\subsection{Bounce Notification Mails Are Delivered Before Their Creation
-Is Logged}
-
-\label{Bounce notification mails delivered before their creation is logged}
-
-This is yet another complication that only occurs during periods of
-extremely high load, when process scheduling and even hard disk access
-times cause strange behaviour.  In this complication bounce notification
-mails are created, delivered, and deleted from the queue, \textit{before\/}
-the log line from \daemon{bounce} that explains their source is logged.  To
-deal with this the COMMIT action maintains a cache of recently committed
-bounce notification mails, which the BOUNCE action subsequently checks if
-the bounce mail is not already in the state tables.  If the queueid exists
-in the cache, and its start time is less than ten seconds before the
-timestamp of the bounce log line, it is assumed that the bounce
-notification mail has already been processed and the BOUNCE action does not
-create one.  If the queueid exists it is removed from the cache, because it
-has either just been used or it is too old to use in future.  Whether the
-BOUNCE action creates a mail or finds an existing mail in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.  This is not required to correctly deal
-with the complication, but is an optimisation to reduce the parser's memory
-usage; on the occasions the author has observed this complication occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
-
-\subsection{Mails Deleted During Delivery}
-
-\label{Mails deleted during delivery}
-
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails that are in the process of being delivered will be deleted.  This
-results in the log lines from the delivery agent (\daemon{local},
-\daemon{virtual} or \daemon{smtp}) appearing in the log file
-\textit{after\/} the mail has been removed from the state tables and saved
-in the database.  The DELETE action adds deleted mails to a cache, which is
-checked by the SAVE\_DATA action, and the current log line discarded if the
-following conditions are met:
-
-\begin{enumerate}
-
-    \item The queueid is not found in the state tables.
-
-    \item The queueid is found in the cache of deleted mails.
-
-    \item The timestamp of the log line is within 5 minutes of the final
-        timestamp of the mail.
-
-\end{enumerate}
-
-Sadly this solution involves discarding some data, but the complication
-only arises eight times in quick succession in one log file, which is not
-in the \numberOFlogFILES{} log files used for testing; if this complication
-occurred more frequently it might be desirable to find the mail in the
-database and add the log line to it.
+\input{logparser-complications}
 

### 2009-03-25 doc/logparser-implementation.tex
Corrections from proofreading, with lots more to come.

@@
 RESCUE SOME OF IT\@.
+
+XXX make log-files-lines
+
+XXX grep matches logparser*.tex
 
@@
 
-Please refer to \sectionref{parser design} for a discussion of why the
-rules, actions, and framework have been separated in the parser's design.
-In this section algorithm can be taken to mean the combination of framework
-and actions.
+XXX DID I SAY THE STUFF IN THE NEXT TWO PARAGRAPHS ALREADY\@?
 
@@
     \item Postfix logs enough information to make it possible to accurately
-        reconstruct the actions it has taken.  There are several heuristics
-        used when parsing; see \sectionref{identifying bounce
+        reconstruct the actions it has taken.  Heuristics are used in
+        several places when parsing; see \sectionref{identifying bounce
         notifications}, \sectionref{aborted delivery attempts}, and
-        \sectionref{pickup logging after cleanup} for details.
+        \sectionref{pickup logging after cleanup} for details.  At least
+        one difficulty encountered while writing \parsername{}
+        (\sectionref{yet more aborted delivery attempts}) could not be
+        solved using the data in the log files, and requires a brute force
+        approach.
 
@@
 In some ways this task is similar to reverse engineering or replicating a
-black box program based solely on its inputs and outputs.  Thus far
+black box system based solely on its inputs and outputs.  Thus far
 analysis of the log files has been sufficient to reconstruct Postfix's
@@
 
-    \item The parser is developed using real world log files rather than
-        the idealised log files someone would naturally envisage reading
-        the source code.  The source code cannot accurately communicate the
-        variety of orderings in which log lines are written to the log
-        file, as process scheduling independently interferes with logging
-        and other processing.
+    \item \parsername{} is developed using real world log files rather than
+        the idealised log files someone would naturally envisage when
+        reading the source code, which cannot accurately communicate the
+        variety of orderings in which log lines are found in log files.
 
     \item The parser acts as a second source of information about Postfix's
-        operation, using information gathered from empirical evidence.  A
-        separate project could compare the empirical knowledge inherent in
-        the parsing algorithm with Postfix's documentation and source code
-        to see how closely the two agree.
+        operation, based on empirical evidence.  A separate project could
+        compare the empirical knowledge inherent in \parsername{} with
+        Postfix's documentation and source code to see how closely the two
+        agree.
 
@@
 \figureref{flow chart image} shows the most common paths a connection or
-mail can take through \parsername{}; the parsing complications described in
+mail can take through \parsername{}; the difficulties described in
 \sectionref{complications} are excluded for the sake of clarity.  The flow
@@
 forwarding, or internally generated by Postfix.  Local submission is the
-simplest case: a queueid is assigned immediately and the sender address is
-logged \flowchart{PICKUP}{2}.  Re-injection due to forwarding sadly lacks
-explicit log lines of its own; it is explained fully in
+simplest case: a queueid is assigned immediately \flowchart{PICKUP}{2}, and
+the mail moves on to the delivery stage.  Re-injection due to forwarding
+sadly lacks explicit log lines of its own; it is explained fully in
 \sectionref{Re-injected mails}.  Internally generated mails lack any
@@
 administrator when it encounters configuration errors, but such mails are
-presumably rare.
+presumably rare.  XXX WOULD THIS PARAGRAPH BE BETTER AS A LIST\@?
 
@@
     \item The client disconnects \flowchart{DISCONNECT}{6}.  If Postfix has
-        rejected any \acronym{SMTP} commands the data will be saved to the
-        database; if not there will not be any data to save (any mails
-        accepted will already have been cloned so their data is in another
-        data structure).
-
-    \item If one or more mails were accepted there will be more log lines
-        for those mails.
+        rejected any \acronym{SMTP} commands the resulting data will be
+        saved to the database; if there were no rejections there will not
+        be any data to save.  Any mails accepted will already have been
+        cloned so their data is in another data structure, nd will be
+        delivered in the same way as mails that entered the system by any
+        other route.
 
@@
     \item Postfix will log the sender and recipient addresses separately
-        \flowchart{SAVE\_DATA}{9}.
+        \flowchart{SAVE\_DATA}{8}.
 
     \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail \flowchart{TRACK}{10}; the handling of
+        tracked by the parent mail \flowchart{TRACK}{9}; the handling of
         re-injected mails is described in \sectionref{tracking re-injected
@@
     \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator \flowchart{COMMIT}{12}.  This is the last log line
-        for this particular mail (though it may be indirectly referred to
-        if it was re-injected).  If it is neither parent nor child of
-        re-injection the data is cleaned up and entered in the database,
-        then deleted from the state tables.
+        administrator \flowchart{COMMIT}{11}.  This is the last log line
+        for this particular mail, though it may be indirectly referred to
+        if it was re-injected.  The mail is cleaned up and entered in the
+        database, then deleted from the state tables.
 
@@
 
-The database is an integral part of the parser presented here: it stores
-the rules and the data gleaned by using those rules to parse Postfix log
-files.  Understanding the database schema is important in understanding the
-actions of the parser, and essential to developing further applications
-which utilise the data gathered; \sectionref{database as API} describes
-how the database schema functions as an \acronym{API}.
+A database is used to store both the rules and the data gleaned by parsing
+Postfix log files.  Understanding the database schema is important in
+understanding the actions of the parser, and essential to developing
+further applications which utilise the data gathered; \sectionref{database
+as API} describes how the database schema functions as an \acronym{API}.
 
@@
 mail accepted and every connection where there was a rejection; the
-individual fields will be described in \sectionref{connections table}.
-There will be at least one entry in the results table for each entry in the
-connections table; the result table's fields will be covered in detail in
+individual fields will be described in \sectionref{connections table}.  The
+results table will have at least one entry for each entry in the
+connections table; its fields will be covered in detail in
 \sectionref{results table}.  A diagram of the database schema is provided
-in \figureref{Diagram of the Database Schema picture}, alongside the
+in \figureref{Diagram of the Database Schema picture}, to complement the
 in-depth descriptions of each table.
@@
 Postfix log files.  Although the interface is a database schema rather than
-a set of functions in a library, it has the same benefits as any other
-\acronym{API}: it provides a stable interface, allowing code on either side
-of the interface to be changed without adverse effects, as long as the
-interface is adhered to.  Programs that use the database can range from the
-simple examples in \sectionref{motivation} to far more complex data mining
-tools.
+a set of functions in a library, it gives the same benefits as any other
+\acronym{API}: it provides a stable interface between the user and the
+creator of the data, allowing code on either side of the interface to be
+changed without adverse effects, as long as the interface is adhered to.
+Programs that use the database can range from the simple examples in
+\sectionref{motivation} to far more complex data mining tools.
 
@@
 CREATE TRIGGER ON INSERT INTO results
-    WHERE sender = 'boss@example.com'
-        AND action = 'DELIVERY_REJECTED'
-    SEND PANIC EMAIL TO 'postmaster@example.com';
+    WHERE sender = "boss@example.com"
+        AND action = "DELIVERY_REJECTED"
+    SEND PANIC EMAIL TO "postmaster@example.com";
 \end{verbatim}
 
-            \item Other tables can be added to the database, to cache
-                historical, summary, or computed data.
+            \item Other tables can be added to the database, e.g.\ to cache
+                historical or computed data, or to link in data from other
+                sources.
 
@@
         \acronym{SQL}, the basics are the same everywhere, analogous to the
-        overall similarities and minor differences amongst Irish English,
+        overall similarity and minor differences amongst Irish English,
         British English, American English, and Australian English.
         Depending on the database in use there may be tools available that
-        reduce or remove the requirement to know \acronym{SQL}; several are
-        available for \gls{SQLite}, the default database used by
-        \parsername{}.
+        reduce or remove the requirement to know \acronym{SQL}.
 
@@
 tables in the database.  Rules are created by the user, not the parser, and
-will not be modified by the parser (except for the hits and hits\_total
+will not be modified by \parsername{} (except for the hits and hits\_total
 fields).  Rules classify the individual log lines, capturing data to be
 saved in the connections and results tables, and specifying the action to
-take for that log line.
-
-The following attributes are defined for each rule:
+take for each recognised log line.
+
+Each rule some or all of these fields; all fields are required unless
+otherwise stated.
 
@@
     \item [description] Something must have occurred to cause Postfix to
-        log each line (e.g.\ a remote client connecting causes a connection
-        line to be logged).  This field describes the event causing the log
-        lines this rule matches.
-
-    \item [restriction\_name] The restriction that caused the mail to be
-        rejected.  Only applicable to rules that recognise rejection log
-        lines, i.e.\ rules that have an action of
-        \texttt{DELIVERY\_REJECTED}, other rules should have an empty
-        string.
+        log each line, this field should describe the event causing the log
+        lines this rule recognises.
+
+    \item [restriction\_name] The name of the Postfix restriction that
+        caused the mail to be rejected.  Only applicable to rules that
+        recognise rejection log lines, i.e.\ rules that have an action of
+        \texttt{DELIVERY\_REJECTED}; other rules should not set this field.
 
@@
 
-    \item [regex] The regex to recognise log lines with, documented in
+    \item [regex] The regex to recognise log lines with, as documented in
         \sectionref{regex components}.
 
-    \item [connection\_data] Sometimes rules need to save data that is not
-        present in the log line: e.g.\ setting \texttt{client\_ip} when a
-        mail is being delivered to another server.  The format is:
+    \item [connection\_data] Sometimes rules need to provide data that is
+        not present in the log line: e.g.\ setting \texttt{client\_ip} when
+        a mail is being delivered to another server.  Any field in the
+        connections table can be set in this way.  The format is:
         \newline{} \tab{} \texttt{ client\_hostname = localhost,}
         \newline{} \tab{} \tab{} \texttt{client\_ip = 127.0.0.1} \newline{}
-        i.e.\ semi-colon or comma separated assignment statements, with the
-        variable name on the left and the value on the right hand side.
-        Any field in the connections table can be set in this way.  Commas
+        i.e.\ semi-colon or comma separated assignment statements.  Commas
         and semi-colons cannot be escaped and thus cannot be included in
-        data.  This feature is intended for use with small amounts of data,
-        so dealing with escape sequences was deemed unnecessary.
+        data, this feature is intended for use with small amounts of data,
+        so dealing with escape sequences was deemed unnecessary.  This
+        field is optional.
 
     \item [result\_data] The result table equivalent of
-        \texttt{connection\_data}.
+        \texttt{connection\_data}, also optional.
 
@@
         each time the rule successfully recognises a log line.  At the
-        start of each run the parser sorts the rules by how frequently each
-        rule recognised log lines, and at the end of the run it updates
+        start of each run \parsername{} sorts the rules by hits, and at the end of the run it updates
         every rule's hits field in the database.  Assuming that the
@@
         the parser's execution time.  Rule ordering for efficiency is
-        discussed in \sectionref{rule ordering for efficiency}.
+        discussed in \sectionref{rule ordering for efficiency}.  This field
+        will be set by the parser rather than the rule author.
 
     \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser.
+        runs of the parser; hits is reset to zero each time the parser is
+        run, but hits\_total is not.  This field will be set by the parser
+        rather than the rule author.
 
@@
         the list of rules is sorted, priority overrides hits.  This allows
-        more specific rules to take precedence over more general rules
-        (described in \sectionref{rules in architecture}).
+        more specific rules to take precedence over more general rules, as
+        described in \sectionref{rules in architecture}.  This field is
+        optional.
 
@@
 Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing the following
-fields:
+have a single entry in the connections table containing all of the
+following fields:
 
@@
 
-    \item [server\_ip] The \acronym{IP} address (IPv4 or IPv6) of the
-        server: the local address when receiving mail, the remote address
-        when sending mail.
+    \item [server\_ip] The \acronym{IP} address of the server: the local
+        address when receiving mail, the remote address when sending mail.
 
@@
 
-    \item [client\_ip] The client \acronym{IP} address (IPv4 or IPv6): the
-        remote address when receiving mail, the local address when sending
-        mail.
+    \item [client\_ip] The client \acronym{IP} address: the remote address
+        when receiving mail, the local address when sending mail.
 
@@
         logs the HELO hostname when it rejects an \acronym{SMTP} command,
-        but it is quite simple to rectify this, as described in
+        but it is quite simple to rectify this as described in
         \sectionref{logging helo}.
@@
     \item [queueid] The queueid of the mail if the connection represents an
-        accepted mail, or \texttt{NOQUEUE} otherwise.
-
-    \item [start] The timestamp of the first log line, in seconds since the
-        epoch.\glsadd{Epoch}
-
-    \item [end] The timestamp of the last log line, in seconds since the
-        epoch.
+        accepted mail, or \texttt{NOQUEUE} if not.
+
+    \item [start] The timestamp of the first log line.
+
+    \item [end] The timestamp of the last log line.
 
@@
         --- a mechanism that is quite useful for testing new restrictions.
-        This field will be set to 1 if the log line parsed was a warning rather
-        than a real rejection, or to 0 for a real rejection.  This field
-        should be ignored if the result is not a rejection, i.e.\ the
-        action field of the associated rule is not
-        \texttt{DELIVERY\_REJECTED}.
+        This field will be false for a real rejection, or true if the log
+        line was a warning.  This field should be ignored if the result is
+        not a rejection, i.e.\ the action field of the associated rule is
+        not \texttt{DELIVERY\_REJECTED}.
 
@@
         line.  In general an \acronym{SMTP} code is only present in a log
-        line for a rejection or final delivery; results whose log line did
-        not contain an \acronym{SMTP} code will duplicate the
+        line representing a rejection or final delivery; results whose log
+        line did not contain an \acronym{SMTP} code will duplicate the
         \acronym{SMTP} code of other results in that connection.  Some
@@
         \texttt{result\_data} field, based on the success or failure
-        represented by the log line.
+        represented by the log line.  XXX GIVE AN EXAMPLE LOG LINE\@.
 
     \item [enhanced\_status\_code] The enhanced status code~\cite{RFC3463}
-        is similar to the \acronym{SMTP} code, but is intended to be parsed
-        by mail clients so that error messages can be clearly conveyed to
-        the user.  Enhanced status code support was added to Postfix in
-        version 2.3; logs generated by previous versions will not contain
-        any enhanced status codes.
-
-    \item [sender] The sender's email address.  Because multiple mails may
-        be delivered during one connection, there may be different sender
-        addresses in the results for one connection; however there should
-        not be different sender addresses in the results for one mail.
+        is similar to the \acronym{SMTP} code, but is intended to be
+        interpreted by mail clients so that error messages can be clearly
+        conveyed to the user.  Enhanced status code support was added to
+        Postfix in version 2.3; log files from previous versions will not
+        contain any enhanced status codes.
+
+    \item [sender] The sender's email address.  This may change during a
+        connection if the client uses different sender addresses for
+        multiple rejected delivery attempts; however there should not be
+        different sender addresses in the results for one accepted mail.
         Mails sent over the same connection can be distinguished by their
@@
 
-    \item [size] The size of the mail; it will only be present for
-        delivered mails.
+    \item [size] The size of the mail, present for delivered mails only.
 
@@
 
-    \item [message\_id] The message-id of the accepted mail, or
-        \texttt{NULL} if no mail was accepted.
-
-    \item [data] A field available for anything not covered by other
-        fields, e.g.\ the rejection message from a \acronym{DNSBL}\@.
-
-    \item [timestamp] The time at which the log line was logged, in seconds
-        since the epoch.\glsadd{Epoch}
+    \item [message\_id] The message-id of the accepted mail, again present
+        for delivered mails only.
+
+    \item [data] A field available to store a piece of captured data that
+        does not have a specific field, e.g.\ the rejection message from a
+        \acronym{DNSBL}\@.
+
+    \item [timestamp] The time at which the log line was logged.
 
@@
 processed is held in \texttt{connections} and \texttt{queueids}
-respectively.  There are also state tables used when working around
+respectively.  The other state tables used when working around
 complications in parsing: \texttt{timeout\_queueids} is used when dealing
@@
 
-    \item Ensuring the regex is valid and compiling it for efficiency
+    \item The regex is valid and compiling it for efficiency
         (\sectionref{Caching compiled regexes}).
@@
 If there are state tables from a previous \parsername{} run to be loaded,
-they will be loaded now; obviously the framework supports saving its state
-tables too.  The need to track re-injected mail (\sectionref{Re-injected
-mails}) complicates the process of saving state, as the relationships
-between mails must be maintained; loading state is complicated by dealing
-with aborted delivery attempts (\sectionref{aborted delivery attempts}),
-because a different set of relationships between connections and mails must
-be re-created when the state tables are restored.
-
-The framework will now move on to parsing each log file.  For each log line
-it will use those rules whose \texttt{program} field matches the program in
-the log line (see \sectionref{rule conditions in implementation}), falling
-back to generic rules if necessary, and warning if the log line is
-unrecognised.  The rules are sorted for increased efficiency; see
-\sectionref{rule ordering for efficiency} for details.  The repetitive
-nature of log files gives them high compression ratios; the framework
-supports reading compressed files, simplifying the process of parsing them
-because they do not have to be uncompressed to a temporary file before
-parsing begins.  The framework also displays a progress bar to show how far
-parsing has progressed through the log file, and to give the user an
-indication of how long the remainder of the parsing process is likely to
-take.  The progress bar is not as reliable when parsing compressed files
-because a compressed block may uncompress to a variable number of log
-lines; there is also variation between the recognition and processing time
-of individual log lines, but overall the progress bar is a useful addition.
+they will be loaded now; the framework supports saving state at any time,
+without adversely affecting the parsing process.  The need to track
+re-injected mail (\sectionref{Re-injected mails}) complicates the process
+of saving state, as the relationships between mails must be maintained;
+loading state is also complicated by dealing with aborted delivery attempts
+(\sectionref{aborted delivery attempts}), because a separate set of
+relationships between connections and mails must be re-created when the
+state tables are restored.
+
+The framework is now ready to begin parsing.  For each log line it will use
+those rules whose \texttt{program} field matches the program in the log
+line (see \sectionref{rule conditions in implementation}), falling back to
+generic rules if necessary, and warning if the log line is unrecognised.
+The rules are sorted for increased efficiency; see \sectionref{rule
+ordering for efficiency} for details.  The repetitive nature of log files
+gives them high compression ratios; the framework supports reading
+compressed files, simplifying the process of parsing them because they do
+not have to be uncompressed to a temporary file before parsing begins.  The
+framework also displays a progress bar to show how far parsing has
+progressed through the log file, and to give the user an indication of how
+long the remainder of the parsing process is likely to take.  The progress
+bar is not as reliable when parsing compressed files, because a compressed
+block may uncompress to a variable number of log lines; there is also
+variation between the recognition and processing time of individual log
+lines, but overall the progress bar is a useful addition.
 
 The framework collects data used when evaluating \parsernames{} efficiency
-for the results chapter (\sectionref{Results}); in addition some of the
-techniques used to improve parsing speed can be turned off or altered to
-show the effect they have.  Three sets of data are collected:
+for the results chapter (\sectionref{Results}); some of the techniques used
+to improve parsing speed can be turned off or altered to measure the effect
+they have.  Three sets of data are collected:
 
@@
     \item The framework can record which rule recognised each log line, and
-        then on a subsequent run use that information so that it uses the
-        correct rule for each log line.  This gives the best possible
-        running time, as only one rule is used for each log line.  A
+        then on a subsequent run consult that information so that it uses
+        the correct rule for each log line.  This gives the best possible
+        running time, because only one rule is used for each log line.  A
         variation of this is to use the correct rule last, to get the worst
@@
         loaded; this can be disabled and the regex compiled every time it
-        is used when recognising log lines, to demonstrate the effect this
-        optimisation has on parser efficiency, as shown in
-        \sectionref{Caching compiled regexes}.
-
-    \item Normally once a log line has been recognised, the framework
+        is used when recognising log lines.  The effect of this
+        optimisation is shown in \sectionref{Caching compiled regexes}.
+
+    \item Normally when a log line has been recognised the framework
         invokes the action specified by the rule.  Invoking the action can
         be skipped if desired, so the timing information shows how long is
-        spent recognising log lines; this time can be subtracted from the
-        time taken by a normal run to show how long is spent processing log
-        lines.  This data is analysed in \sectionref{recognising vs
+        spent recognising log lines only; this time can be subtracted from
+        the time taken by a normal run to show how long is spent processing
+        log lines.  This data is analysed in \sectionref{recognising vs
         processing}.
@@
     \item The framework can skip inserting data into the database, because
-        that dominates the execution time of the parser, and the evaluation
-        in \sectionref{Results} is measuring the speed of \parsername{},
-        not the speed of the database or the disks it resides on.
+        that dominates the execution time of the parser; the evaluation
+        chapter measures the speed of \parsername{}, not the speed of the
+        database or the disks it resides on.
 
@@
 
-    \item Each result can be extended with extra debugging information, so
-        that when a mail's data is dumped for inspection more information
-        is available.  The information currently added is: the log line's
-        timestamp in human readable form; the entire log line; the name of
-        the log file and the number of the log line within it.
+    \item Each result can be extended with extra debugging information,
+        which is useful when a warning dumps a mail's data for inspection.
+        The information currently added is: the log line's timestamp in
+        human readable form; the entire log line; the name of the log file
+        and the number of the log line within it.  This information is not
+        stored in the database.
 
@@
 \texttt{CLONE} action, which processes Postfix accepting a delivery
-attempt, and the data structure moves from one state table to another
-within the parser.
-
-The complications and difficulties that arose when parsing log files
-generated on a production mail server are documented in
-\sectionref{complications}; some action descriptions refer to specific
-difficulties they address.  The complications are documented in a separate
-section to avoid overwhelming the action descriptions.
+attempt, and the data structure moves from the connections state table to
+the queueids state table.
+
+The complications and difficulties that arose when parsing real-world log
+files are documented in \sectionref{complications}; some action
+descriptions refer to specific difficulties they address.  The
+complications are documented in a separate section to avoid overwhelming
+the action descriptions.
 
@@
 
-            \item [program] The name of the program that logged the line.
-
-            \item [pid] The \acronym{pid} of the program that logged the
+            \item [program] The name of the program that generated the log
+                line.
+
+            \item [pid] The \acronym{pid} of the process that logged the
                 line.
@@
         necessary; if the mail already exists the unknown origin flag will
-        be removed.  The action also marks the mail as a bounce
+        be removed from it.  The action also marks the mail as a bounce
         notification.  To deal with complication \sectionref{Bounce
@@
         command \daemon{postsuper}.  This action adds a dummy recipient
-        address if required, then invokes the COMMIT action to save the
-        mail to the database.  The complication this action deals with is
-        described fully in \sectionref{Mail deleted before delivery is
-        attempted}.
+        address if required (see \sectionref{Mail deleted before delivery
+        is attempted}), then invokes the COMMIT action to save the mail to
+        the database.
 

### 2009-03-25 doc/logparser.tex
Some minor changes.
Corrections for the assumptions and database sections of the implementation chapter.
Improved Makefile to rerun latex if the list of figures or tables changes. Rearranged some parts of the rules section of the implementation chapter, and made corrections to it from proofreading.
Corrections to flowchart section of implementation chapter. Improvements to graph size so they fill the width of the page. More consistent formatting of references to the flowchart. Alphabetical ordering for the actions with the same number of rules in the action distribution graph. Added more XXX i

@@
 \usepackage{eqlist}
+% This voodoo below gives us an eqlist environment with bold items.
+\newcommand{\bolditem}[1]{%
+    \bfseries#1%
+}
+\newenvironment{boldeqlist}
+    {\begin{eqlist}[\renewcommand{\makelabel}{\bolditem}]}
+    {\end{eqlist}}
 
+% Enable customised lists.
+\usepackage{enumitem}
 % Tell Latex to use scalable fonts
@@
 \usepackage{svn}
-
-% Sort numbers where there are multiple citations.  Does not appear to have
-% the expected effect (probably clashes with varioref or hyperref), though
-% it does reduce the space between numbers.
-\usepackage{cite}
-% Check for unused references.
-% \usepackage{refcheck}
 
@@
 
-%\setlength{\floatsep}{0pt}
-%\setlength{\intextsep}{0pt}
-%\setlength{\textfloatsep}{0pt}
-
-% Show the citation key (e.g.\ slct-paper) in the margin beside the
-% reference.
+% Sort numbers where there are multiple citations.  Does not appear to have
+% the expected effect (probably clashes with varioref or hyperref), though
+% it does reduce the space between numbers.
+\usepackage{cite}
+% Check for unused references, and how the citation key (e.g.\ slct-paper)
+% in the margin beside the reference.
 %\usepackage[nomsgs]{refcheck}
-% Show where references are used.
-% Neither work.
+% Show where references are used; neither work.
 %\usepackage{citeref}
@@
     \setlength{\itemsep}{0pt}%
+    \setlength{\topsep}{0pt}%
+    \setlength{\partopsep}{0pt}%
 }
@@
 \newcommand{\showgraph}[3]{%
-    \begin{figure}[hbt!]
+    % XXX REVISIT THE PLACEMENT OF FIGURES, THOUGH I THINK LATEX'S DEFAULTS
+    % AR EPROBABLY PRETTY GOOD.
+    \begin{figure}%[hbt!]
         \caption{#2}\label{#3}
@@
 
+\newcommand{\flowchart}[2]{%
+    (action: #1\@; flowchart:~#2)%
+}
+
 % Add last checked dates to all URLs
@@
 \newcommand{\numberOFrules}[0]{%
-    185%
+    184%
 }
@@
 \newcommand{\numberOFrulesMINIMUM}[0]{%
-    116%
+    115%
 }
@@
 \newcommand{\numberOFrulesMINIMUMpercentage}[0]{%
-    62.70\%%
+    62.50\%%
 }
@@
 \newcommand{\numberOFrulesMAXIMUMpercentage}[0]{%
-    59.48\%%
+    60.00\%%
 }
@@
 \newcommand{\numberOFactions}[0]{%
-    18%
+    22%
 }
@@
 \newcommand{\numberOFruleINTERSECTIONS}[0]{%
-    17,020%
+    16,836%
 }
@@
 \newcommand{\parsername}{\acronym{PLP}}
-\newcommand{\parsernames}{\acronym{PLP}'s}
+\newcommand{\parsernames}{\acronyms{PLP}}
+\newcommand{\parsernamelong}{Postfix Log Parser}
 
@@
 
+    \setcounter{page}{2}
     \phantomsection{}
@@
 
-\end{abstract}
-
     \SVN$Id$
@@
 
-XXX SHOULD I SCRAP postfix\_action?
+    XXX ADD A $\backslash{}$action\{\} COMMAND TO FORMAT ACTION NAMES
+    CONSISTENTLY\@.
 
-\newpage
-\tableofcontents
-\listoffigures
-\listoftables
+    XXX FIX THE PROBLEMS THAT AROSE FROM THE LAST all-logs RUN\@.
 
-\newpage
+    XXX PLPs SHOULD BE PLP's IN A LOT OF PLACES\@; ADD A NEW COMMAND AND
+    CHECK THE CURRENT USAGE\@.
+
+    XXX SHOULD I HAVE A ``FUTURE WORK'' SECTION SOMEWHERE IN THE
+    CONCLUSION\@?
+
+\end{abstract}
 
@@
 \glsaddall[types={postfix}]
+
+\setcounter{page}{3}
+\tableofcontents
+\listoffigures
+\listoftables
+
+\newpage
 

### 2009-03-26 doc/logparser-complications.tex
Corrections from proofreading, with lots more to come.
More corrections from proofreading.

@@
-\section{Complications Encountered}
+\section{Complications Encountered While Writing \parsernamelong{}}
 
@@
 
-The mail lacks a queueid until it has been accepted, so log lines must
-first be correlated by the \daemon{smtpd} \acronym{pid}, then transition to
-being correlated by the queueid.  This is relatively minor, but does
-require:
+\label{queueid vs pid}
+
+A delivery attempt lacks a queueid until one recipient has been accepted,
+so log lines must first be correlated by \daemon{smtpd} \acronym{pid}, then
+transition to being correlated by their queueid.  This is relatively minor,
+but does require:
 
@@
 
-    \item Two state tables to hold the data structure for each connection
-        and mail.
+    \item Two state tables to hold data structures.
 
-    \item Most importantly: every section of code must know whether it
-        needs to lookup the data structures by \acronym{pid} or queueid.
+    \item Most importantly: every section of code must know which state
+        table it should query.
 
@@
 Multiple independent mails may be delivered across one connection: this
-requires the algorithm to clone the current data as soon as a mail is
-accepted, so that subsequent mails will not trample over each other's data.
+requires \parsername{} to clone the connection's data as soon as a mail is
+accepted, so that subsequent mails will not overwrite each other's data.
 This must be done every time a mail is accepted, as it is impossible to
-tell in advance which connections will accept multiple mails.  Once the
-mail has been accepted its log lines will not be correlated by
-\acronym{pid} any more, its queueid will be used instead (except when
-timeouts occur during the data phase \sectionref{timeouts during data
-phase}).  If the original connection has any useful data (e.g.\ rejections)
+tell in advance which connections will accept multiple mails.  Once a mail
+has been accepted its log lines will not be correlated by \acronym{pid} any
+more, its queueid will be used instead, as described in \sectionref{queueid
+vs pid}.  If the original connection has any useful data (e.g.\ rejections)
 it will be saved to the database when the client disconnects.  One unsolved
@@
 parent mail will log the creation of the child mail, e.g.\ parent mail
-3FF7C4317 creates child mail 56F5B43FD\@:
+\texttt{3FF7C4317} creates child mail \texttt{56F5B43FD}\@:
 
@@
 addition, the first log line belonging to the child mail (the example log
-line above belongs to the parent mail) is logged by \daemon{qmgr} or
+line above belongs to the parent mail) is logged by either \daemon{qmgr} or
 \daemon{cleanup}, so the order also depends on how soon they process the
@@
 child mail will be created, delivered, and entirely finished with,
-\textit{before\/} the parent mail logs the creation log line!  Thus, mails
-flagged as coming from an unknown origin cannot be entered into the
-database when their final log line is processed; instead they must be
-marked as ready for entry and subsequently entered once their origin has
-been identified.  The crux of this complication is that re-injected mails
-appear in the log files without explicit logging indicating their source.
+\textit{before\/} the parent mail logs its creation!  Thus, mails flagged
+as coming from an unknown origin cannot be entered into the database when
+their final log line is processed, nor can \parsername{} warn the user;
+instead they must be marked as ready for entry and subsequently entered
+once their origin has been identified.  The crux of this complication is
+that re-injected mails appear in the log files without explicit logging
+indicating their source.
 
@@
 
-    \item If a mail is tagged with the unknown origin flag it is assumed to
-        be a child mail whose parent has not yet been identified.  The mail
-        is tagged as ready to be entered in the database, but entry is
+    \item If a mail is tagged with the unknown origin flag, it is assumed
+        to be a child mail whose parent has not yet been identified.  The
+        mail is tagged as ready to be entered in the database, but entry is
         postponed until the parent is identified.  The child mail should
-        not have any subsequent log lines: only its parent should refer to
+        not have any subsequent log lines: only its parent will refer to
         it.
 
-    \item If the mail is a child mail with an identified parent it is
-        entered in the database as usual, then removed from its parent's
+    \item If the mail is a child mail whose parent has been identified, it
+        is entered in the database as usual, then removed from its parent's
         list of children.  If this child is the last mail on that list, and
@@
 
-    \item If the mail is a parent it is entered in the database as usual,
+    \item If the mail is a parent, it is entered in the database as usual
         because there will be no further log lines for it.  There may be
@@
         entered as normal, and removed from the state tables.  If there are
-        more child mails, the parent's removal from the state tables will
-        be postponed until the last child has been entered.
+        incomplete child mails, the parent's removal from the state tables
+        will be postponed until the last child has been entered.
 
@@
 identifies their source.  Heuristics must be used to identify bounce
-notifications, and those heuristics are:
+notifications:
 
@@
     \item The message-id has a specific format: \newline{}
-        \tab{} \texttt{YYYYMMDDhhmmss.queueid@server.hostname} \newline{}
+        \tab{} \texttt{YYYYMMDDhhmmss.queueid@server\_hostname} \newline{}
         e.g.\ \texttt{20070321125732.D168138A1@smtp.example.com}
@@
     \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes bounce notifications generated
-        locally from bounce notifications which are being re-injected as a
-        result of aliasing.  In the latter case the message-id will be
-        unchanged from the original bounce notification, and so even if it
-        happens to be in the correct format (e.g.\ if it was generated by
-        Postfix on this or another server) it will not match the queueid of
-        the mail.
+        the mail: this is what distinguishes a new bounce notification from
+        a bounce notification that is being re-injected as a result of
+        aliasing.  In the latter case the message-id will be unchanged from
+        the original bounce notification, and so even if it happens to be
+        in the correct format, e.g.\ if it was generated by Postfix on this
+        or another server, it will not match the queueid of the mail.
 
@@
 
-Some mail clients behave unexpectedly during the \acronym{SMTP} dialogue:
-the client aborts the first delivery attempt after the first recipient is
-accepted, then makes a second delivery attempt for the same recipient which
+Some mail clients behave strangely during the \acronym{SMTP} dialogue: the
+client aborts the first delivery attempt after the first recipient is
+accepted, then makes a second delivery attempt for the same recipient that
 it continues with until delivery is complete.  Microsoft Outlook is one
@@
 clear, so once again heuristics are required to identify when this
-behaviour occurs.  In this case a list of all mails accepted during a
-connection is saved in the connection state, and the accepted mails are
-examined when the disconnection action is executed.  Each mail is checked
+behaviour occurs.  A list of all mails accepted during a connection is
+saved in the connection's state table entry, and the accepted mails are
+examined when the disconnection action is invoked.  Each mail is checked
 for the following:
@@
         The first two \daemon{smtpd} log lines will be a connection log
-        line and a mail acceptance log line; Postfix logs acceptance as
-        soon as the first recipient has been accepted.
+        line and a mail acceptance log line.
 
@@
 
-    \item Does the queueid exist in the state tables?  If not it cannot be
+    \item Can the mail be found in the state tables?  If not it cannot be
         an aborted delivery attempt.
@@
 schema requires many more fields be populated (see \sectionref{connections
-table} and \sectionref{results table}).  This heuristic is quite
+table} and \sectionref{results table}).  These heuristics are quite
 restrictive, and it appears there is little scope for false positives; if
-there are any false positives there will be warnings when the next log line
-for that mail is parsed.  False negatives are less likely to be detected:
-there may be queueid clashes (and warnings) if mails remain in the state
-tables after they should have been removed, otherwise the only way to
-detect false negatives is to examine the state tables after each parsing
-run.
+there are any false positives there will also be warnings when the next log
+line for that mail is parsed.  False negatives are less likely to be
+detected: there may be queueid clashes (and warnings) if mails remain in
+the state tables after they should have been removed, otherwise the only
+way to detect false negatives is to examine the state tables after each
+parsing run.
 
 This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where a list of cloned connections is maintained.
+the CLONE action where the list of accepted mails is maintained.
 
@@
 disconnections are either unimportant or are handled elsewhere in the
-parser (\sectionref{timeouts during data phase}).  Sadly, Postfix does not
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
+parser, e.g.\ \sectionref{timeouts during data phase}.  Sadly, Postfix does
+not log a message saying the mail has been discarded, as should be expected
+by now.  The checks to identify this happening are:
 
@@
     \item Is the mail missing its \daemon{cleanup} log line?  Every mail
-        which passes through Postfix will have a \daemon{cleanup} line;
-        lack of a \daemon{cleanup} line is a sure sign the mail did not
-        make it too far.
+        which passes normally through Postfix will have a \daemon{cleanup}
+        line; lack of one is a sure sign the mail did not make it too far.
 
     \item Were there three or more \daemon{smtpd} log lines for the mail?
-        There should be a connection log line and a mail accepted log line,
-        followed by one or more rejection log lines.
-
-    \item Is the last \daemon{smtpd} log line a rejection line?
+        There should be a connection log line and a mail acceptance log
+        line, followed by one or more delivery rejected log lines.
 
@@
 
-If all checks are successful then the mail is assumed to have been
-discarded when the client disconnected.  There will be no further entries
-logged for such mails, so without identifying and entering them in the
-database immediately they accumulate in the state table and will cause
-clashes if the queueid is reused.
+If both checks are successful then the mail is assumed to have been
+discarded when the client disconnected.  There will be no further log lines
+for such mails, so without identifying and entering them in the database
+immediately they accumulate in the state table and will cause clashes if
+the queueid is reused.
 
@@
 body of the mail are transferred.  Sometimes there is a timeout or the
-connection is lost during the DATA phase;\footnote{For brevity's sake
+connection is lost during the DATA phase;\footnote{For the sake of brevity
 \textit{timeout\/} will be used throughout this section, but everything
 applies equally to lost connections.} when this occurs Postfix will discard
-the mail and the parser needs to discard the data associated with that
+the mail and \parsername{} needs to discard the data associated with that
 mail.  It seems more intuitive to save the mail's data to the database, but
@@
 
-    \item Record the timeout and associated data in the connection's
-        results.
+    \item Record the timeout and captured data in the connection's results.
 
@@
         have interrupted transfer of an accepted delivery attempt, or it
-        may have occurred after a mail delivery attempt was rejected.  To
-        distinguish between the two cases the action compares the timestamp
-        of the last accepted mail against the timestamp of the last log
-        line from \daemon{smtpd} for that connection; this depends on the
-        CLONE action keeping a list of all mails accepted for each
+        may have occurred after a mail delivery attempt was rejected.
+
+        XXX THE REMAINDER OF THIS DOES NOT MAKE SENSE\@; CHECK AND CORRECT
+        IT\@.  To distinguish between the two cases the action compares the
+        timestamp of the last accepted mail against the timestamp of the
+        last log line from \daemon{smtpd} for that connection; this depends
+        on the CLONE action keeping a list of all mails accepted for each
         connection.  If the mail acceptance timestamp is later than the
@@
 though there may be other circumstances that trigger it.  Sometimes the
-\daemon{cleanup} line for a mail being accepted is logged after the timeout
-line; parsing this line causes the MAIL\_QUEUED action to create a new
-state table entry for the queueid in the log line.  This is incorrect
-because the line actually belongs to the mail that has just been discarded;
-the next log line for that queueid will be seen when the queueid is reused
-for a different mail, causing a queueid clash and the appropriate warning.
+\daemon{cleanup} log line for a mail being accepted is logged after the
+timeout log line, by which time \parsername{} has discarded the mail;
+parsing the \daemon{cleanup} log line causes the CLEANUP\_PROCESSING action
+to create a new state table entry, to help deal with re-injected mails
+(\sectionref{Re-injected mails}).  This is incorrect because the log line
+actually belongs to the mail that has just been discarded; if the queueid
+is reused there will be a queueid clash, otherwise the new mail will just
+remain in the state tables.
 
-When the \daemon{cleanup} line is still pending during the TIMEOUT action,
-the action updates a global list of queueids, adding the queueid and the
-timestamp from the log line.  When the next \daemon{cleanup} line is parsed
-for that queueid the list will be checked (during the MAIL\_QUEUED action),
-and the log line will be deemed part of the mail where the timeout occurred
-and discarded if it meets the following requirements:
+During the TIMEOUT action, if the \daemon{cleanup} line is still pending,
+the action updates the \texttt{timeout\_queueids} state table, adding the
+queueid and the timestamp from the log line.  When the next
+\daemon{cleanup} line is parsed for that queueid, the list will be checked
+and if the log line meets the following requirements it will be deemed part
+of the mail where the timeout occurred and discarded.
 
@@
     \item The queueid must not have been reused yet, i.e.\ it does not have
-        an entry in the state tables.
+        a mail in the state tables.
 
@@
         minutes of the mail acceptance timestamp.  Timeouts happen after
-        five minutes, but some data may have been transferred slowly
-        (perhaps because either the client or server is suffering from
-        network congestion or rate limiting), and empirical evidence shows
-        that ten minutes is not unreasonable; hopefully it is a good
-        compromise between false positives (log lines incorrectly
-        discarded) and false negatives (new state table entries incorrectly
-        created).
+        five minutes, but some data may have been transferred slowly, and
+        empirical evidence shows that ten minutes is not unreasonable;
+        hopefully it is a good compromise between false positives (log
+        lines incorrectly discarded) and false negatives (new state table
+        entries incorrectly created).
 
@@
 
-The next \daemon{cleanup} line must meet the criteria above for it to be
-discarded because some, but not all connections where a timeout occurs will
-have an associated \daemon{cleanup} line logged; if the algorithm blindly
-discarded the next \daemon{cleanup} line after a timeout it would sometimes
-be mistaken.  When the next \daemon{pickup} line containing that queueid is
-parsed the queueid will be removed from the cache of timeout queueids,
-regardless of whether it meets the criteria above.
+The \daemon{cleanup} log line must meet the criteria above for it to be
+discarded because some, but not all, connections where a timeout occurs
+will have an associated \daemon{cleanup} log line; if the
+CLEANUP\_PROCESSING action blindly discarded the next \daemon{cleanup} log
+line after a timeout it would sometimes be mistaken.  When the next
+\daemon{pickup} log line containing that queueid is processed the queueid
+will be removed from the cache of timeout queueids, regardless of whether
+it meets the criteria above.
 
-This complication is handled by the TIMEOUT and MAIL\_QUEUED actions.
+This complication is handled by the CLEANUP\_PROCESSING and TIMEOUT
+actions.
 
@@
 SMTPD\_DIED, and SMTPD\_WATCHDOG\@.
+
+\newpage{} %XXX VARIOREF PROBLEM
 

### 2009-03-26 doc/logparser-implementation.tex
More corrections from proofreading.

@@
 XXX WRITE AN INTRODUCTION\@.  MOST OF THIS IS CRAP, BUT I MIGHT BE ABLE TO
-RESCUE SOME OF IT\@.
+RESCUE SOME OF IT\@.  IT SHOULD HAVE A BLURB, FOLLOWED BY A PREVIEW OF THE
+CHAPTER\@.
+
+XXX MAIL MADS ABOUT EXAMINERS\@.
 
@@
     \item [DISCONNECT] Invoked when the remote client disconnects, it
-        enters the connection in the database, performs any required
-        cleanup, and deletes the connection from the state tables.  This
-        action deals with aborted delivery attempts (\sectionref{aborted
-        delivery attempts}).
+        enters the connection in the database if it has any useful data,
+        performs any required cleanup, and deletes the connection from the
+        state tables.  This action deals with aborted delivery attempts
+        (\sectionref{aborted delivery attempts}).
 
     \item [EXPIRY] If Postfix has not managed to deliver a mail after
-        trying for five days it will give up and return the mail to the
+        trying for five days, it will give up and return the mail to the
         sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check as
-        described in \sectionref{out of order log lines}; this action, with
-        cooperation from the COMMIT and SAVE\_DATA actions, deals with that
-        complication.
+        Postfix programs which passes the valid combinations check,
+        implemented to deal with the complication described in
+        \sectionref{out of order log lines}; this action tags the mail as
+        having expired, so the COMMIT action will skip the valid
+        combinations check.
 
@@
         recognising rule's regex, and does nothing more.  It is a separate
-        action to help distinguish delivery attempts that bounce from other
+        action to distinguish delivery attempts that bounce from other
         delivery attempts.
@@
         recognising rule's regex, and does nothing more.  It is a separate
-        action to help distinguish successful delivery attempts from other
+        action to distinguish successful delivery attempts from other
         delivery attempts.
@@
     \item [MAIL\_TOO\_LARGE] When a client tries to send a mail larger than
-        the local server accepts it will be discarded and the client
+        the local server accepts, the mail will be discarded and the client
         informed of the problem.  The mail may have been accepted and
         partially transferred, in which case there's a data structure to
-        dispose of, otherwise there will not be.  See \sectionref{timeouts
-        during data phase} for the gory details; although that describes
-        timeouts the processing is the same for mails that are too large.
+        dispose of.  See \sectionref{timeouts during data phase} for full
+        details; although that describes timeouts the processing is the
+        same for mails that are too large.
 
@@
         service dealing with a locally submitted mail.  A new mail will be
-        created, although out of order log entries may have caused the
-        state table entry to already exist, as documented in
-        \sectionref{pickup logging after cleanup}.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its configuration
-        it kills all \daemon{smtpd} processes, requiring any active
+        created, although out of order log entries may have caused it to
+        already exist, as documented in \sectionref{pickup logging after
+        cleanup}.
+
+    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its
+        configuration, it kills all \daemon{smtpd} processes, requiring all
         connections to be cleaned up, entered in the database, and deleted
         from the state tables.  Postfix probably kills all the other
-        components too, but \parsername{} is only affected by \daemon{smtpd}
-        processes exiting.
+        components too, but \parsername{} is only affected by
+        \daemon{smtpd} processes exiting.
 
@@
 
-    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies or exits
-        unsuccessfully; the active connection for that \daemon{smtpd} must
-        be cleaned up, entered in the database if it has sufficient data,
-        and deleted from the state tables.  In some cases the connection
-        will not have enough data to satisfy the database schema, so it
-        cannot be entered into the database; unfortunately this means that
-        some data is lost.
-
-    \item [SMTPD\_KILLED] Sometimes a \daemon{smtpd} is killed by a signal;
-        this is processed in the same way as a \daemon{smtpd} dying, see
-        the \texttt{SMTPD\_DIED} action for details.
+    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies, is killed by a
+        signal, or exits unsuccessfully; the associated connection must be
+        cleaned up, entered in the database if it has sufficient data, and
+        deleted from the state tables.  In some cases the connection will
+        not have enough data to satisfy the database schema, so it cannot
+        be entered into the database; unfortunately this means that some
+        data is lost.
 
@@
 
-    \item [TIMEOUT] The connection with the remote client timed out so the
+    \item [TIMEOUT] The connection with the remote client timed out, so the
         mail currently being transferred must be discarded.  The mail may
         have been accepted, in which case there's a data structure to
-        dispose of, or it may not in which case there is not.  See
-        \sectionref{timeouts during data phase} for full details.
+        dispose of.  See \sectionref{timeouts during data phase} for full
+        details.
 
@@
 The rules are responsible for classifying each log line and specifying the
-correct action to be invoked.  For any parser implemented using this
-architecture, the rules are the most visible component and most likely to
-be modified by users.  Rules need to be as simple as possible so that users
-can easily modify or add to them, but the implementation must balance that
-simplicity with the need to provide enough flexibility and power to
-successfully parse the inputs.
+correct action to be invoked.  In any parser implemented using this
+architecture, the rules will be the most visible component and most likely
+to be modified by users.  Rules need to be as simple as possible so that
+users can easily modify or add to them, but the implementation must balance
+that simplicity with the need to provide enough flexibility and power to
+successfully parse inputs.
 
@@
 The structure of the rules table has already been documented in
-\sectionref{rules table}; that abstract description will not be repeated
-here, but should be fleshed out by the concrete example rule in
-\sectionref{example rule in implementation}.  The process of creating new
-rules from unparsed log lines is dealt with in \sectionref{creating new
-rules in implementation}; the implementation of the utility supplied with
-\parsername{} to create regexes from unparsed log lines is also described
-in that section.  The rule conditions used in \parsername{} are explained
-next, and the final portion of this section provides some
-suggestions for how to detect overlapping rules.
+\sectionref{rules table}; that description will not be repeated here, but
+should be fleshed out by the example rule in \sectionref{example rule in
+implementation}.  The process of creating new rules from unparsed log lines
+is dealt with in \sectionref{creating new rules in implementation}; the
+implementation of the utility supplied with \parsername{} to create regexes
+from unparsed log lines is also described in that section.  The rule
+conditions used in \parsername{} are explained next, and the final portion
+of this section provides some suggestions for how to detect overlapping
+rules.
 
@@
 recognises the log line resulting from Postfix rejecting a delivery attempt
-because the domain in the sender address does not have an A, AAAA,
-or MX DNS entry, i.e.\ mail could not be delivered to the sender's address
-(for full details see~\cite{reject-unknown-sender-domain}).
+because the domain in the sender address does not have an A, AAAA, or MX
+DNS entry, i.e.\ mail could not be delivered to any address in the sender's
+domain (for full details see~\cite{reject-unknown-sender-domain}).
 
@@
 \sectionref{actions in detail in implementation} to determine the correct
-one.  The regex will be constructed based on the log line, but see below
-for a tool to ease the process.
-
-There are other attributes that may be required in a rule:
-connection\_data, result\_data, priority, or restriction\_name.  In general
-it will only become clear that connection\_data or result\_data are
-required when \parsername{} warns about an entry in the connections or
+one.  The regex needs to be constructed based on the log line, but see
+below for a tool to ease the process.
+
+There are other attributes that may be required to make a rule work
+correctly: connection\_data, result\_data, priority, or restriction\_name.
+In general it will only become clear that connection\_data or result\_data
+are required when \parsername{} warns about an entry in the connections or
 results tables that is missing some required fields, because values for the
@@
 the new rule deliberately uses the architecture's overlapping rules feature
-it will be obvious that the priority field needs to be set, on this rule and
-possibly others; the priority field may be needed on unintentionally
+it will be obvious that the priority field needs to be set, on this rule
+and possibly others; the priority field may be needed on unintentionally
 overlapping rules too, but that is more difficult to determine.  Finally
 the restriction\_name field should be set when the rule's action is
-\texttt{DELIVERY\_REJECTED}; hopefully the restriction in question will be
+\texttt{DELIVERY\_REJECTED}; hopefully the name of the restriction will be
 clear from the content of the log line.
@@
 explanation below.  The core of the \acronym{SLCT} algorithm is quite
-simple: log lines are generally created by substituting variable words into
-a fixed pattern, and analysis of the frequency with which each word occurs
-can be used to determine whether the word is variable or part of the fixed
-pattern.  This classification can be used to group similar log lines and
-generate a regex to match each group of log lines.  There are 4 steps in
-the algorithm:
+simple: programs generally create log lines by substituting variable words
+into a fixed pattern, and analysis of the frequency with which each word
+occurs can be used to determine whether the word is variable or part of the
+fixed pattern.  This classification can be used to group similar log lines
+and generate a regex to match each group of log lines.  There are 4 steps
+in the algorithm:
 
@@
         \acronym{IP} addresses, the standard start of rejection messages)
-        with keywords that \parsername{} will expand when it loads the new
-        rule.  The purpose of this step is to utilise existing knowledge to
-        create more accurate regexes; it replaces a large number of
-        variable words with fixed words, improving the subsequent
-        classification of words as fixed or variable.  The altered log
-        lines are written to a temporary file, which subsequent stages will
-        use instead of the original input file.
+        with keywords described in \sectionref{regex components}.  The
+        purpose of this step is to utilise existing knowledge to create
+        more accurate regexes; it replaces a large number of variable words
+        with fixed words, improving the subsequent classification of words
+        as fixed or variable.  The altered log lines are written to a
+        temporary file, which subsequent stages will use instead of the
+        original input file.
 
@@
         indicate similarity unless it occupies the same position in both
-        log lines.  If a variable term within a line contains
+        log lines.  If a variable term substituted into a log line contains
         spaces, it will appear to the algorithm as more than one word.
@@
         counter from the previous pass; if that counter has a high
-        frequency a separate counter will be maintained for this word in
-        this position.  This reduces the number of counters maintained by
-        the algorithm in this step, reducing the memory requirements of the
-        algorithm at a cost of increased CPU usage.
-
-        As time goes on the amount of memory typically available increases
-        and the need to reduce memory requirements decreases, so the
-        modified algorithm omits the hashing step and maintains counters
-        for all tokens.  In addition most of the infrequently occurring
-        words will have been substituted with keywords during the first
-        step, vastly reducing the number of tuples to maintain counters
-        for.
-
-    \item [Classify words based on their frequency.]  The frequency of
-        every tuple \texttt{(word, word's position within the log line)} is
+        frequency a separate counter will be maintained for this
+        \texttt{(word, word's position within the log line)} tuple.  This
+        reduces the number of counters maintained by the algorithm in this
+        step, reducing the memory requirements of the algorithm at the cost
+        of increased CPU usage.
+
+        As time goes by the amount of memory typically available to a
+        program or algorithm increases, and the need to reduce memory
+        requirements correspondingly decreases, so the modified algorithm
+        omits the hashing step and maintains counters for all tokens.  Most
+        of the infrequently occurring words will have been substituted with
+        keywords during the first step, vastly reducing the number of
+        tuples to maintain counters for; the original algorithm does not
+        have the detailed knowledge leveraged by the new algorithm, because
+        it is a generic tool.
+
+    \item [Classify words based on their frequency.]  The frequency of each
+        \texttt{(word, word's position within the log line)} tuple is
         checked: if its frequency is greater than the threshold supplied by
@@
         variable term.  If a variable term appears sufficiently often it
-        will be misclassified as a fixed term, but that should be noticed
-        by the user when reviewing the new regexes.  Variable terms are
-        replaced by \texttt{.+} to match one or more of any character.
+        will be misclassified as a fixed term, but that will hopefully be
+        noticed by the user when reviewing the new regexes, and will be
+        obvious when the new rules do not recognise some log lines they are
+        expected to.  Variable terms are replaced by \texttt{.+} to match
+        one or more of any character.
 
@@
         matching the log line, and a counter is maintained for each regex.
-        Contiguous sequences of \texttt{.+} in the newly reassembled
-        regexes are collapsed to a single \texttt{.+}; any resulting
-        duplicate regexes are combined, and their counters added together.
-        If a regex's counter is lower than the threshold supplied by the
-        user the regex is discarded; this second threshold is independent
-        of the threshold used to differentiate between fixed and variable
-        words, but once again 1\% of log lines is a good starting point.
-        The new regexes are displayed for the user to add to the database,
-        either as new rules or merged into the regexes of existing rules;
-        the counter for each regex is also displayed, giving the user an
+        Contiguous sequences of \texttt{.+} in the newly assembled regexes
+        are collapsed to a single \texttt{.+}; any resulting duplicate
+        regexes are combined, and their counters added together.  If a
+        regex's counter is lower than the threshold supplied by the user
+        the regex is discarded; this second threshold is independent of the
+        threshold used to differentiate between fixed and variable words,
+        but once again 1\% of log lines is a good starting point.  The new
+        regexes are displayed for the user to add to the database, either
+        as new rules or merged into the regexes of existing rules; the
+        counter for each regex is also displayed, giving the user an
         indication of how many of the log lines that regex should match.
@@
 
-XXX MERGE logs2regexes WITH check-regexes.
+XXX MERGE logs2regexes WITH check-regexes.  CHECK IF SLCT DOES THIS STEP\@.
 
@@
 the Postfix component whose log lines it recognises, and each log line
-contains the name of the Postfix component that produced it; the framework
-will only use rules where the two are equal when trying to recognise log
-lines.  This avoids needlessly trying rules that will not match the log
-line, or worse, might match unintentionally.  In addition the framework
-supports generic rules, whose program attribute is ``*''; these will be
-used if none of the program-specific rules recognise the log line.  If none
-of the rules are successful the framework will warn the user.
+contains the name of the Postfix component that produced it; when trying to
+recognise log lines, the framework will only use rules where the two are
+equal.  This avoids needlessly trying rules that will not recognise the log
+line, or worse, might recognise it unintentionally.  In addition the
+framework supports generic rules, whose program attribute is ``*''; these
+will be used if none of the program-specific rules recognise the log line.
+If none of the rules are successful the framework will warn the user.
 
@@
         \_\_CLIENT\_HOSTNAME\_\_ is immediately clear, whereas
-        \verb!(?:unknown|(?:[-.\w]+))! needs to be deciphered.
+        \verb!(?:unknown|(?:[-.\w]+))! needs to be deciphered each time it
+        is encountered.
 
     \item It avoids needless repetition of complex regex components, and
-        allows the components to be corrected or improved in one
-        location.  E.g.\ \_\_SENDER\_\_ is used in 68 rules; if a mistake
-        is discovered it can be corrected in one place only.
+        allows the components to be corrected or improved in one location.
+        E.g.\ \_\_SENDER\_\_ is used in 68 rules; if a mistake is
+        discovered it only needs to be corrected in one place.
 
     \item It enables automatic extraction and saving of captured data.  The
-        regex snippets use Perl 5.10's named capture buffers~\cite{perlre},
-        so the mapping between captures and fields does not need to be
-        explicitly specified by the rule.
+        regex snippets use Perl 5.10's named capture buffers~\cite{perlre}
+        to capture data, so the mapping between captures and fields does
+        not need to be explicitly specified by the rule.
 
@@
 
-    \item [\_\_COMMAND\_\_]  All \acronym{SMTP} commands, except
-        \texttt{DATA}, which typically has a more specific rule.
-        Priorities could have been used instead of excluding \texttt{DATA}.
+    \item [\_\_COMMAND\_\_]  All \acronym{SMTP} commands.
 
@@
     \item [\_\_SHORT\_CMD\_\_]  Postfix sometimes logs \acronym{SMTP}
-        commands in a short, single word form.
+        commands in a short, single word form; this snippet matches all of
+        those, except \texttt{DATA}, which typically has a more specific
+        rule.  Priorities could have been used instead of excluding
+        \texttt{DATA}.
 
@@
 
-\parsername{} does not try to detect overlapping rules; that responsibility
-is left to the author of the rules.  The advantages and difficulties of
-overlapping rules have already been addressed in \sectionref{overlapping
-rules in architecture} and will not be repeated here.  \parsername{}
-provides a mechanism for ordering overlapping rules: the priority field in
-each rule (defaulting to zero); negative priorities may be useful for
-catchall rules.
+The advantages and difficulties of overlapping rules have already been
+addressed in \sectionref{overlapping rules in architecture} and will not be
+repeated here.  \parsername{} does not try to detect overlapping rules;
+that responsibility is left to the author of the rules.  A mechanism is
+provided for ordering overlapping rules: the priority field in each rule;
+negative priorities may be useful for catchall rules.
 
@@
 
-    \item Sort by program and regex, then visually inspect the list, e.g.\
-        with \acronym{SQL} similar to:      \newline{}
+    \item Sort rules by program and regex, then visually inspect the list,
+        e.g.\ with \acronym{SQL} similar to:    \newline{}
         \verb!SELECT program, regex!            \newline{}
@@
         with overlapping regexes will not overlap if their programs are
-        different.  Note that this does not properly deal with rules whose
-        program is \texttt{*}; those rules will be used on all log lines
-        that have not been recognised by program-specific rules.
-
-    \item Compare the results of parsing using different rule orderings
-        (described in \sectionref{rule ordering for efficiency}).  Parse
+        different.  Note that this does not properly deal with generic
+        rules whose program is \texttt{*}; those rules will be used on all
+        log lines that have not been recognised by program-specific rules.
+
+    \item Compare the results of parsing using different rule orderings, as
+        described in \sectionref{rule ordering for efficiency}.  Parse
         several log files using optimal ordering, then dump a textual
@@
         database.  If there are no overlapping rules the tables from each
-        run will be identical; differences indicate overlapping rules.  The
-        rules that overlap can be determined by examining the differences
-        in the tables: each result contains a reference to the rule which
-        created it, which will change if that rule overlaps with another.
-        Unfortunately this method cannot prove the absence of overlapping
-        rules; it can detect overlapping rules, but only if there are log
-        lines in the log files that are recognised by more than one rule.
+        run will be identical; differences indicate overlapping rules.
+        The rules that overlap can be determined by examining the
+        differences in the tables: each result contains a reference to the
+        rule which created it, which will change if that rule overlaps with
+        another.  Unfortunately this method cannot prove the absence of
+        overlapping rules; it can detect overlapping rules, but only if
+        there are log lines in the log files that are recognised by more
+        than one rule.
 

### 2009-03-26 doc/logparser-architecture.tex
Changes from proofreading, and a couple of other checks.

@@
 intricacies and complications encountered during parsing.  The function
-that finds the rule matching the input and invokes the requested action can
-be expressed in pseudo-code as:
+that finds the rule recognising the input and invokes the requested action
+can be expressed in pseudo-code as:
 
@@
 During development of \parsername{} it became apparent that in addition to
-the obvious variety in log~lines there were many complications to be
+the obvious variety in log lines there were many complications to be
 overcome.  Some were the result of deficiencies in Postfix's logging, and
@@
 both rules and actions.  A simple example is to have one rule and action
-removing comments from input lines, so that other rules and actions do not
+removing comments from inputs, so that other rules and actions do not
 have to handle comments at all; obviously if comment characters can be
@@
 by more than one rule.  Unintentionally overlapping rules lead to
-inconsistent parsing and data extraction because the first matching rule
-wins, and the order in which rules are tried against each input might
-change between parser invocations.  Overlapping rules are frequently a
-requirement, allowing a more specific rule to match some inputs and a more
-general rule to match the remainder, e.g.\ separating \acronym{SMTP}
-delivery to specific sites from \acronym{SMTP} delivery to the rest of the
-world.  Allowing overlapping rules simplifies both the general rule and the
-more specific rule; additionally rules from different sources can be
-combined with a minimum of prior cooperation or modification required.
-Overlapping rules should have a priority attribute to specify their
-relative ordering; negative priorities may be useful for catchall rules.
-The architecture does not try to detect overlapping rules: that
+inconsistent parsing and data extraction because the first rule to
+recognise the input wins, and the order in which rules are tried against
+each input might change between parser invocations.  Overlapping rules are
+frequently a requirement, allowing a more specific rule to recognise some
+inputs and a more general rule to recognise the remainder, e.g.\ separating
+\acronym{SMTP} delivery to specific sites from \acronym{SMTP} delivery to
+the rest of the world.  Allowing overlapping rules simplifies both the
+general rule and the more specific rule; additionally rules from different
+sources can be combined with a minimum of prior cooperation or modification
+required.  Overlapping rules should have a priority attribute to specify
+their relative ordering; negative priorities may be useful for catchall
+rules.  The architecture does not try to detect overlapping rules: that
 responsibility is left to the author of the rules.

### 2009-03-26 doc/logparser-complications.tex
Changes from proofreading, and a couple of other checks.

@@
         in the correct format, e.g.\ if it was generated by Postfix on this
-        or another server, it will not match the queueid of the mail.
+        or another server, it will not correspond with the queueid of the
+        mail.
 
@@
 
-    \item The message-id must have the correct format and match the queueid
-        of the mail.  While a mail sent from elsewhere could easily have
-        the correct message-id format, the chance that the queueid in the
-        message-id would match the queueid of the mail is extremely small.
+    \item The message-id must have the correct format and correspond with
+        the queueid of the mail.  While a mail sent from elsewhere could
+        easily have the correct message-id format, the chance that the
+        queueid in the message-id would correspond with the queueid of the
+        mail is extremely small.
 
@@
 When mail is submitted locally, \daemon{pickup} accepts the new mail and
-generates a log line showing the source.  Occasionally this log line will
-occur later in the log file than the \daemon{cleanup} log line, so the
-PICKUP action will find that a state table entry exists for that queueid.
-Normally if the queueid given in the PICKUP line exists in the state tables
-a warning is generated by the \daemon{pickup} action, but if the following
-conditions are met it is assumed that the log lines were out of order:
+generates a log line.  Occasionally this log line will occur later in the
+log file than the \daemon{cleanup} log line, so the PICKUP action will find
+that a state table entry already exists for that queueid.  Normally when
+this happens a warning is generated by the PICKUP action, but if the
+following conditions are met it is assumed that the log lines were out of
+order:
 
@@
 
-This complication is dealt with during the PICKUP action.
+This complication is dealt with by the PICKUP action.
 
@@
 
-The circumstances above account for all occasions identified thus far where
-a \daemon{smtpd} suddenly stops logging.  In addition to removing an active
-connection the last accepted mail may need to be discarded, as detailed in
+The circumstances above account for all occasions where a \daemon{smtpd}
+suddenly stops logging.  In addition to removing an active connection the
+last accepted mail may need to be discarded, as detailed in
 \sectionref{timeouts during data phase}; otherwise the queueid state table
@@
 
-These occurrences are handled by the three actions POSTFIX\_RELOAD,
-SMTPD\_DIED, and SMTPD\_WATCHDOG\@.
+These occurrences are handled by the actions POSTFIX\_RELOAD (1),
+SMTPD\_DIED (2 \& 3), and SMTPD\_WATCHDOG (4).
 
@@
 dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup log lines} or \sectionref{pickup
+mail}, \sectionref{discarding cleanup log lines}, or \sectionref{pickup
 logging after cleanup}.  In the \numberOFlogFILES{} log files used for
-testing this occurs only five times in 60,721,709 log lines, but for parser
-correctness it must be dealt with.  The five occurrences have the same
-characteristics: the \daemon{local} log line showing delivery to a local
-mailbox occurs after the \daemon{qmgr} log line showing removal of the mail
-from the queue because delivery is completed.  This causes problems: the
-data in the state tables for the mail is not complete, so entering it into
-the database fails; a new mail is created when the \daemon{local} line is
-parsed and remains in the state tables; four warnings are issued per pair
-of out of order log lines.
+testing this problem occurs only five times in 60,721,709 log lines, but
+for parser correctness it must be dealt with.  All five occurrences have
+the same characteristics: the \daemon{local} log line showing delivery to a
+local mailbox comes after the \daemon{qmgr} log line showing removal of the
+mail from the queue because delivery is complete.  This causes problems:
+the data in the state tables for the mail is not complete, so it cannot be
+entered into the database; a new mail is created when the \daemon{local}
+log line is processed and remains in the state tables; four warnings are
+issued per pair of out of order log lines.
 
-The COMMIT action examines the list of programs that have produced a log
-lines for each mail, comparing the list against a table of known-good
-program combinations.  If the mail's combination is found in the table the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The SAVE\_DATA
-action checks for that flag; if the additional log lines have caused the
-mail to reach a valid combination then entry in the database will proceed,
-otherwise it must be postponed once more.
+The COMMIT action examines the list of programs that have produced log lines
+for each mail, comparing the list against a table of known-good program
+combinations.  If the mail's combination is found in the table the mail can be
+entered in the database; if the combination is not found entry must be
+postponed and the mail flagged for later entry.  The MAIL\_DELIVERED action
+checks for that flag; if the additional log lines have caused the mail to reach
+a valid combination then entry in the database will proceed, otherwise it must
+be postponed once more.
 
 The list of valid combinations is explained below.  Every mail will
-additionally have log line from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have log line from \daemon{bounce}, \daemon{postsuper}, or
+additionally have log lines from \daemon{cleanup} and \daemon{qmgr}; any
+mail may also have a log line from \daemon{bounce}, \daemon{postsuper}, or
 both.
@@
         client, then deleted by the administrator before any delivery
-        attempt was made (the unwanted mail is typically due to a mail loop
-        or joe~job\glsadd{Joe-job}).  Notice that \daemon{postsuper} is
-        required, not optional, for this combination.
+        attempt was made.  Notice that \daemon{postsuper} is required, not
+        optional, for this combination.
 
@@
 This check applies to accepted mails only, not to rejected mails.  This
-check is performed during the COMMIT action.
+check is performed during the COMMIT action, with support from the
+MAIL\_DELIVERED action.
 
@@
 The aborted delivery attempts described in \sectionref{aborted delivery
-attempts} occur frequently, but the aborted delivery attempts described in
-this section only occur four times in the \numberOFlogFILES{} log files
-used for testing.  The symptoms are the same as in \sectionref{aborted
-delivery attempts}, except that there \textit{is\/} a \daemon{cleanup} log
-line; there is nothing in the log file to explain why there are no further
-log lines.  The only way to detect these mails is to periodically scan all
-mails in the state tables, deleting any mails displaying the following
-characteristics:
+attempts} occur frequently, but those described in this section only occur
+four times in the \numberOFlogFILES{} log files used for testing.  The
+symptoms are the same as in \sectionref{aborted delivery attempts}, except
+that there \textit{is\/} a \daemon{cleanup} log line; there is nothing in
+the log file to explain why there are no further log lines.  The only way
+to detect these mails is to periodically scan all mails in the state
+tables, deleting any mails with the following characteristics:
 
@@
     \item The timestamp of the last log line for the mail must be 12 hours
-        or more earlier than the last log line parsed from the current log
-        file.
+        or more earlier than the timestamp of the last log line parsed from
+        the current log file.
 
@@
 12 hours is a somewhat arbitrary time period, but it is far longer than
-Postfix would delay delivery of a mail in the queue (unless Postfix is not
-running for an extended period of time).  The state tables are scanned for
-mails matching the characteristics above each time the end of a log file is
-reached, and matching mails are deleted.
+Postfix would delay delivery of a mail in the queue, unless it was not
+running for an extended period of time.  Each time the end of a log file is
+reached, the state tables are scanned for mails matching the
+characteristics above, and any mails found are deleted.
 
@@
 Postfix logs the recipient address when delivery of a mail is attempted, so
-if delivery has yet to be attempted the parser cannot determine the
+if delivery has yet to be attempted \parsername{} cannot determine the
 recipient address or addresses.  This is a problem when mail is arriving
 faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail (because it is the result of a mail loop\glsadd{mail
-loop}, mail bomb\glsadd{mail bomb}, or joe~job\glsadd{Joe-job}) before
-Postfix has had a chance to try to deliver it.  In this case the recipient
-address will not have been logged, so a dummy recipient address needs to be
-added, as every mail is required by the database schema
-(\sectionref{results table}) to have at least one recipient.  Typically
-when this complication occurs there are many instances of it, closely
-grouped.
+some of the mail before Postfix has had a chance to try to deliver it.  In
+this case the recipient address will not have been logged, so a dummy
+recipient address needs to be added, as every mail is required by the
+database schema (\sectionref{results table}) to have at least one
+recipient.  Typically when this complication occurs there are many
+instances of it, closely grouped.
 
 This lack of information cannot easily be overcome: it is trivial to log a
-warning for every recipient accepted, but when the warning for the first
-recipient is logged Postfix will not yet have allocated a queue file and
-queueid for the mail, so the warning will be associated with the connection
-rather than the accepted mail.  A queue file and queueid will be allocated
-after Postfix accepts the MAIL FROM command if
+warning for every accepted recipient, but Postfix will not yet have
+allocated a queueid for the mail when the warning for the first recipient
+is logged, so the warning will be associated with the connection rather
+than the accepted mail.  A queue file and queueid will be allocated after
+Postfix accepts the MAIL FROM command if
 \texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is set to ``no'', but that
-setting will cause disk IO for almost every delivery attempt, instead of
-just for delivery attempts where recipients are accepted, and consequently
-a drastic reduction in the performance of the mail server.
+setting will cause disk IO for every delivery attempt, instead of just for
+delivery attempts where recipients are accepted, and consequently a drastic
+reduction in the performance of the mail server.
 
@@
 extremely high load, when process scheduling and even hard disk access
-times cause strange behaviour.  In this complication bounce notification
+times cause strange behaviour.  In this complication, bounce notification
 mails are created, delivered, and deleted from the queue, \textit{before\/}
-the log line from \daemon{bounce} that explains their source is logged.  To
-deal with this the COMMIT action maintains a cache of recently committed
-bounce notification mails, which the BOUNCE action subsequently checks if
-the bounce mail is not already in the state tables.  If the queueid exists
-in the cache, and its start time is less than ten seconds before the
-timestamp of the bounce log line, it is assumed that the bounce
-notification mail has already been processed and the BOUNCE action does not
-create one.  If the queueid exists it is removed from the cache, because it
-has either just been used or it is too old to use in future.  Whether the
-BOUNCE action creates a mail or finds an existing mail in the state tables,
-it flags the mail as having been seen by the BOUNCE action; if this flag is
-present the COMMIT action will not add the mail to the cache of recent
-bounce notification mails.  This is not required to correctly deal
-with the complication, but is an optimisation to reduce the parser's memory
-usage; on the occasions the author has observed this complication occurring
-there have been a huge number of bounce notification mails generated --- if
-every bounce notification mail was cached it would dramatically increase
-the memory requirements of the parser.  The cache of bounce notification
-mails will be pruned whenever the parser's state is saved, though if the
-size of the cache ever becomes a problem it could be pruned periodically to
-keep the size in check.
+the log line from \daemon{bounce} that explains their source.  To deal with
+this the COMMIT action maintains a cache of recently committed bounce
+notification mails, which the BOUNCE action subsequently checks if the
+bounce mail is not found in the state tables.  If the queueid exists in the
+cache, and its start time is less than ten seconds before the timestamp of
+the bounce log line, it is assumed that the bounce notification mail has
+already been processed and the BOUNCE action does not create one.  If the
+queueid exists in the cache it is removed, because it has either just been
+used or the problem did not occur for that mail.  Whether the BOUNCE action
+creates a mail or finds an existing mail in the state tables, it flags the
+mail as having been seen by the BOUNCE action; if this flag is present the
+COMMIT action will not add the mail to the cache of recent bounce
+notification mails.  This is not required to correctly deal with the
+complication, but is an optimisation to reduce the parser's memory usage.
+On the occasions the author has observed this complication occurring there
+have been a huge number of bounce notification mails generated: if every
+bounce notification mail was cached it would dramatically increase the
+memory requirements of the parser.  The cache of bounce notification mails
+will be pruned whenever the parser's state is saved, though if the size of
+the cache ever becomes a problem it could be pruned periodically to keep it
+in check.
 

### 2009-03-26 doc/logparser-conclusion.tex
Some minor changes.
Changes from proofreading, and a couple of other checks.

@@
 removed by the preprocessor).  This parser is different: the rule which
-matches the next log line for a mail dictates the action that will be
+recognises the next log line for a mail dictates the action that will be
 invoked, which is equivalent to the next state in other parsers; thus the
-next log line for a mail dictates the next state for that mail.  XXX THINK
-ABOUT HOW TO MAKE THE PARSER MORE STATEFUL IF REQUIRED\@; ADD IT TO THE
-ARCHITECTURE SECTION\@.
+next log line for a mail dictates the next state for that mail.
 

### 2009-03-26 doc/logparser-implementation.tex
Changes from proofreading, and a couple of other checks.

@@
 CHAPTER\@.
-
-XXX MAIL MADS ABOUT EXAMINERS\@.
-
-XXX make log-files-lines
-
-XXX grep matches logparser*.tex
 
@@
 performing housekeeping duties.
-
-XXX DID I SAY THE STUFF IN THE NEXT TWO PARAGRAPHS ALREADY\@?
 
@@
 paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being matched
-in the parser) and invokes the correct action.
+responds to events (mouse clicks in a GUI program, log lines being
+recognised in \parsername{}) and invokes the correct action.
 
@@
 The framework is now ready to begin parsing.  For each log line it will use
-those rules whose \texttt{program} field matches the program in the log
-line (see \sectionref{rule conditions in implementation}), falling back to
+those rules whose \texttt{program} field equals the program in the log line
+(see \sectionref{rule conditions in implementation}), falling back to
 generic rules if necessary, and warning if the log line is unrecognised.
@@
 gives them high compression ratios; the framework supports reading
-compressed files, simplifying the process of parsing them because they do
-not have to be uncompressed to a temporary file before parsing begins.  The
-framework also displays a progress bar to show how far parsing has
+compressed log files, simplifying the process of parsing them because they
+do not have to be uncompressed to a temporary file before parsing begins.
+The framework also displays a progress bar to show how far parsing has
 progressed through the log file, and to give the user an indication of how
 long the remainder of the parsing process is likely to take.  The progress
-bar is not as reliable when parsing compressed files, because a compressed
-block may uncompress to a variable number of log lines; there is also
-variation between the recognition and processing time of individual log
-lines, but overall the progress bar is a useful addition.
+bar is not as reliable when parsing compressed log files, because a
+compressed block may uncompress to a variable number of log lines; there is
+also variation between the recognition and processing time of individual
+log lines, but overall the progress bar is a useful addition.
 
@@
     \item How many log lines did the framework attempt to recognise, how
-        many were skipped because there were no rules whose condition
-        matched that log line's program, how many were successfully
+        many were skipped because there were no rules whose condition was
+        equal to that log line's program, how many were successfully
         recognised, and how many were not recognised.
@@
 
-A second utility is also provided that reads a list of new regexes and
-the input given to the first utility.  It tries to match each input log
-line against each regex, counting the number of log lines that match
-each regex, warning the user if an input log line is matched by more
-than one regex, and additionally warning if an input log line is not
-matched by any regex.  It displays a summary of how many input log lines
-each regex matched, comparing it to the expected number of matches; this
-provides the user with an easy method of checking if the regexes
-produced by the first utility are correctly matching the input log lines
-they are based upon.
+A second utility is also provided that reads a list of new regexes and the
+input given to the first utility.  It tries to match each input log line
+with each regex, counting the number of log lines that match each regex,
+warning the user if an input log line is matched by more than one regex,
+and additionally warning if an input log line is not matched by any regex.
+It displays a summary of how many input log lines each regex matched,
+comparing it to the expected number of matches; this provides the user with
+an easy method of checking if the regexes produced by the first utility are
+correctly matching the input log lines they are based upon.
 
@@
         Similarly using a new \acronym{DNSBL}, a new policy server, or new
-        administrator-defined rejection messages requires new rules.
+        administrator-defined rejection messages also requires new rules.
 
@@
         incoming delivery attempt is successful.  Configuring Postfix to do
-        this is relatively simple: create a restriction that is guaranteed
+        this is relatively simple; create a restriction that is guaranteed
         to warn for every accepted mail, as follows:
@@
         Although \texttt{smtpd\_helo\_restrictions} seems like the natural
-        place to log the HELO hostname, there will not yet be a queueid
-        associated with the mail when \texttt{smtpd\_helo\_restrictions} is
-        evaluated for the first recipient, so the log line cannot be
-        associated with the correct mail.  There is guaranteed to be a
-        queueid when the DATA command has been reached, and thus the
-        queueid will be logged by any restrictions taking effect in
+        place to log the HELO hostname, when it is evaluated for the first
+        recipient there will not yet be a queueid allocated for the
+        delivery attempt, so the log line cannot be associated with the
+        correct mail.  There is guaranteed to be a queueid when the DATA
+        command has been reached, and thus the queueid will be logged by
+        any restrictions taking effect in
         \texttt{smtpd\_data\_restrictions}, and the log line can be
@@
         database.  If the results of parsing log files from different
-        sources must remain separate, the parser can easily be instructed
-        to use a different database.
+        sources must remain separate, \parsername{} can easily be
+        instructed to use a different database.
 
@@
 
-XXX TO BE WRITTEN\@.  START WITH THE OLD CONTENT BELOW\@.
-
-This section presented the core of the parser, starting with a very high
-level view and the initial complications that arose.  A flow chart showing
-the paths a mail may take through the nascent simplified algorithm was
-provided, followed by an explanation of those paths, and a discussion of
-the parser's emergent behaviour --- the data from the log files creates the
-paths in the flow chart, they are not specified anywhere in the parser.
-The framework which holds the parser together was covered next, after which
-came a description of the current actions provided by the parser, and the
-algorithm for analysing unparsed log lines to create regexes for new rules.
-Detecting, diagnosing, and defeating complications forms the largest single
-portion of this section, mirroring the development of the parser.  The
-complications are described in the order they were overcome, with
-subsequent problems affecting fewer mails (often by an order or magnitude),
-though the time required to solve problems increased with each successive
-problem.
+This section has presented \parsername{}, the parser implemented for this
+project, beginning with the assumptions made during its development.  A
+simplified flowchart shows the most common paths taken through Postfix and
+\parsername{}, accompanied by a description of the stages and transitions.
+A database provides storage for rules and for data gathered from log files;
+any further use of that data is dependent on a clear understanding of the
+database schema, so the resemblance between a database schema and an
+\acronym{API} is described, followed by a detailed description and a
+diagram of the database schema.  The framework is documented next,
+including the steps it takes during initialisation, the parsing process,
+and the conveniences it offers to users.  The data collected by the
+framework is described, as are the ways in which various optimisations can
+be disabled to demonstrate their effect, and the debugging options the
+framework provides.  The implementation of actions in \parsername{} is
+documented, including how frequently each action is specified by rules, and
+brief descriptions of why some actions are more popular than others.  All
+the actions that are part of \parsername{} are described in detail,
+followed by the process of adding a new action to the parser.  The final
+component of the architecture, the rules, is also the most visible
+component, and its implementation is examined in detail with: an example
+rule and an explanation of every field; a description of how to add new
+rules and determine the values that should be used for each field; an
+explanation of the algorithm used by the utility that creates new regexes
+based on unparsed log lines, and how it differs from the original algorithm
+it is based on.  How \parsername{} uses rule conditions and overlapping
+rules is discussed, accompanied by a description of the regex snippets the
+framework provides to simplify the regexes used in rules.  The many
+complications and difficulties encountered while writing \parsername{}, and
+the solutions developed to overcome them, are documented in depth, also
+documented are how some solutions interact and which action or actions the
+solution is implemented in.  This chapter concludes with a list of the
+limitations identified in \parsername{}, followed by some possible
+improvements suggested by the author.

### 2009-03-26 doc/logparser-introduction.tex
Changes from proofreading, and a couple of other checks.

@@
 detail, explaining the purpose and usage of each field in a rule, referring
-to an example rule and sample data it matches successfully against.  The
+to an example rule and sample log line it successfully recognises.  The
 pros and cons of overlapping rules are considered, including techniques for

### 2009-03-26 doc/logparser-results.tex
Track the number of lines per program too.
Changes from proofreading, and a couple of other checks.

@@
 KEEP COUNTERS FOR EACH PROGRAM TOO --- MAYBE SOMETHING USEFUL WILL COME
-FROM IT\@.
+FROM IT\@.  COMPARE NUMBER OF RULES TRIED FOR DIFFERENT RULE ORDERINGS\@.
 
@@
 At the time of writing there are \numberOFrules{} different rules: the top
-10\% match the vast majority of log lines, with the remaining log lines
+10\% recognise the vast majority of log lines, with the remaining log lines
 split across the other 90\% of the rules (as shown in \graphref{rule hits
@@
 consistent over time, parser efficiency should benefit from trying more
-frequently matching rules before those that match less frequently.  To
-test this hypothesis three full test runs were performed with different
+frequently recognising rules before those that recognise less frequently.
+To test this hypothesis three full test runs were performed with different
 rule orderings:
@@
     \item [Optimal]  The most optimal order, according to the hypothesis:
-        rules that match most often will be tried first.
+        rules that recognise log lines most often will be tried first.
 
@@
 
-    \item [Reverse] Hypothetically the worst order: the most frequently
-        matching rules will be tried last.
+    \item [Reverse] Hypothetically the worst order: the rules that
+        recognise log lines most frequently will be tried last.
 
@@
 increases parsing time with reverse ordering, and to a lesser extent with
-shuffled ordering.  The optimal ordering (where the most frequently
-matching rules are tried first) shows a mean increase of
+shuffled ordering.  The optimal ordering (where the rules that recognise
+log lines most frequently are tried first) shows a mean increase of
 \input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
@@
 
-    \item Try each rule until a matching rule is found; if no match is
-        issue a warning and move to the next log line.
+    \item Try each rule until a recognising rule is found; if the log line
+        is not recognised, issue a warning and move to the next log line.
 
@@
 
-    \item There must be a rule to match each different log line produced by
-        each program; if a log line is not successfully matched the parser
-        will issue a warning.  Rules should be as specific and tightly
-        bound as possible to ensure accurate parsing:\footnote{A rule that
-        matches zero or more of any character will successfully parse every
-        log line, but not in a meaningful way.} most log lines contain
-        fixed strings and have a rigid pattern, so this is not a problem.
+    \item There must be a rule to recognise each different log line
+        produced by each program; if a log line is not successfully
+        recognised the parser will issue a warning.  Rules should be as
+        specific and tightly bound as possible to ensure accurate
+        parsing:\footnote{A rule with a regex that matches zero or more of
+        any character will successfully recognise every log line, but not
+        in a meaningful way.} most log lines contain fixed strings and have
+        a rigid pattern, so this is not a problem.
 
@@
 files.  The number of hits per rule is quite unevenly spread, resembling a
-Power Law distribution --- it is obvious that a small number of rules match
-the vast majority of the log lines, and more than half the rules match
-fewer than 100 times.
+Power Law distribution --- it is obvious that a small number of rules
+recognise the vast majority of the log lines, and more than half the rules
+recognise fewer than 100 log lines.
 

### 2009-03-26 doc/logparser-complications.tex
Added a \action{} command to have consistent formatting of action names. Changes \parsernames{} to have an ' before the s.

@@
 Tracking re-injected mail requires \parsername{} to do the following in the
-COMMIT action:
+\action{COMMIT} action:
 
@@
 
-This check is performed during the COMMIT action.
+This check is performed during the \action{COMMIT} action.
 
@@
 
-    \item Was the second result processed by the \texttt{CLONE} action?
+    \item Was the second result processed by the \action{CLONE} action?
         The first two \daemon{smtpd} log lines will be a connection log
@@
     \item If there are third and subsequent results, were those log lines
-        processed by the \texttt{SAVE\_DATA} action?  If there are any log
+        processed by the \action{SAVE\_DATA} action?  If there are any log
         lines after the first two they should be informational only.
@@
 
-This check is performed in the DISCONNECT action; it requires support in
-the CLONE action where the list of accepted mails is maintained.
+This check is performed in the \action{DISCONNECT} action; it requires
+support in the \action{CLONE} action where the list of accepted mails is
+maintained.
 
@@
 
-These checks are made during the DISCONNECT action.
+These checks are made during the \action{DISCONNECT} action.
 
@@
 recorded and saved with the connection instead.  To deal properly with
-timeouts the TIMEOUT action does the following:
+timeouts the \action{TIMEOUT} action does the following:
 
@@
     \item If no mails have been accepted yet nothing needs to be done; the
-        TIMEOUT action ends.
+        \action{TIMEOUT} action ends.
 
@@
         last log line from \daemon{smtpd} for that connection; this depends
-        on the CLONE action keeping a list of all mails accepted for each
-        connection.  If the mail acceptance timestamp is later than the
-        timeout timestamp, the timeout applies to the just-accepted mail,
-        which will be removed from the state tables, without entering it in
-        the database.  If the \daemon{smtpd} timestamp is later there was a
-        rejection between the accepted mail and the timeout: the action
-        assumes that the timeout applies to the rejected delivery attempt
-        and finishes.  This assumption is not necessarily correct, because
-        Postfix may have accepted an earlier recipient and rejected a later
-        one, in which case the timeout applies to the accepted mail, which
-        should be discarded; this has not been a problem in practice.
-        Processing timeouts is further complicated by the presence of out
-        of order \daemon{cleanup} log lines: see \sectionref{discarding
-        cleanup log lines} for details.
+        on the \action{CLONE} action keeping a list of all mails accepted
+        for each connection.  If the mail acceptance timestamp is later
+        than the timeout timestamp, the timeout applies to the
+        just-accepted mail, which will be removed from the state tables,
+        without entering it in the database.  If the \daemon{smtpd}
+        timestamp is later there was a rejection between the accepted mail
+        and the timeout: the action assumes that the timeout applies to the
+        rejected delivery attempt and finishes.  This assumption is not
+        necessarily correct, because Postfix may have accepted an earlier
+        recipient and rejected a later one, in which case the timeout
+        applies to the accepted mail, which should be discarded; this has
+        not been a problem in practice.  Processing timeouts is further
+        complicated by the presence of out of order \daemon{cleanup} log
+        lines: see \sectionref{discarding cleanup log lines} for details.
 
@@
 
-This complication is dealt with by the TIMEOUT action, with help from the
-CLONE action.
+This complication is dealt with by the \action{TIMEOUT} action, with help
+from the \action{CLONE} action.
 
@@
 timeout log line, by which time \parsername{} has discarded the mail;
-parsing the \daemon{cleanup} log line causes the CLEANUP\_PROCESSING action
-to create a new state table entry, to help deal with re-injected mails
-(\sectionref{Re-injected mails}).  This is incorrect because the log line
-actually belongs to the mail that has just been discarded; if the queueid
-is reused there will be a queueid clash, otherwise the new mail will just
-remain in the state tables.
+parsing the \daemon{cleanup} log line causes the
+\action{CLEANUP\_PROCESSING} action to create a new state table entry, to
+help deal with re-injected mails (\sectionref{Re-injected mails}).  This is
+incorrect because the log line actually belongs to the mail that has just
+been discarded; if the queueid is reused there will be a queueid clash,
+otherwise the new mail will just remain in the state tables.
 
-During the TIMEOUT action, if the \daemon{cleanup} line is still pending,
-the action updates the \texttt{timeout\_queueids} state table, adding the
-queueid and the timestamp from the log line.  When the next
+During the \action{TIMEOUT} action, if the \daemon{cleanup} line is still
+pending, the action updates the \texttt{timeout\_queueids} state table,
+adding the queueid and the timestamp from the log line.  When the next
 \daemon{cleanup} line is parsed for that queueid, the list will be checked
@@
 will have an associated \daemon{cleanup} log line; if the
-CLEANUP\_PROCESSING action blindly discarded the next \daemon{cleanup} log
-line after a timeout it would sometimes be mistaken.  When the next
-\daemon{pickup} log line containing that queueid is processed the queueid
-will be removed from the cache of timeout queueids, regardless of whether
-it meets the criteria above.
+\action{CLEANUP\_PROCESSING} action blindly discarded the next
+\daemon{cleanup} log line after a timeout it would sometimes be mistaken.
+When the next \daemon{pickup} log line containing that queueid is processed
+the queueid will be removed from the cache of timeout queueids, regardless
+of whether it meets the criteria above.
 
-This complication is handled by the CLEANUP\_PROCESSING and TIMEOUT
-actions.
+This complication is handled by the \action{CLEANUP\_PROCESSING} and
+\action{TIMEOUT} actions.
 
@@
 generates a log line.  Occasionally this log line will occur later in the
-log file than the \daemon{cleanup} log line, so the PICKUP action will find
-that a state table entry already exists for that queueid.  Normally when
-this happens a warning is generated by the PICKUP action, but if the
-following conditions are met it is assumed that the log lines were out of
-order:
+log file than the \daemon{cleanup} log line, so the \action{PICKUP} action
+will find that a state table entry already exists for that queueid.
+Normally when this happens a warning is generated by the \action{PICKUP}
+action, but if the following conditions are met it is assumed that the log
+lines were out of order:
 
@@
 
-This complication is dealt with by the PICKUP action.
+This complication is dealt with by the \action{PICKUP} action.
 
@@
 
-These occurrences are handled by the actions POSTFIX\_RELOAD (1),
-SMTPD\_DIED (2 \& 3), and SMTPD\_WATCHDOG (4).
+This complication is handled by several actions: \action{POSTFIX\_RELOAD} (1),
+\action{SMTPD\_DIED} (2 \& 3), and \action{SMTPD\_WATCHDOG} (4).
 
@@
 
-The COMMIT action examines the list of programs that have produced log lines
-for each mail, comparing the list against a table of known-good program
-combinations.  If the mail's combination is found in the table the mail can be
-entered in the database; if the combination is not found entry must be
-postponed and the mail flagged for later entry.  The MAIL\_DELIVERED action
-checks for that flag; if the additional log lines have caused the mail to reach
-a valid combination then entry in the database will proceed, otherwise it must
-be postponed once more.
+The \action{COMMIT} action examines the list of programs that have produced
+log lines for each mail, comparing the list against a table of known-good
+program combinations.  If the mail's combination is found in the table the
+mail can be entered in the database; if the combination is not found entry
+must be postponed and the mail flagged for later entry.  The
+\action{MAIL\_DELIVERED} action checks for that flag; if the additional log
+lines have caused the mail to reach a valid combination then entry in the
+database will proceed, otherwise it must be postponed once more.
 
@@
 This check applies to accepted mails only, not to rejected mails.  This
-check is performed during the COMMIT action, with support from the
-MAIL\_DELIVERED action.
+check is performed during the \action{COMMIT} action, with support from the
+\action{MAIL\_DELIVERED} action.
 
@@
 
-The DELETE action is responsible for handling this complication.
+The \action{DELETE} action is responsible for handling this complication.
 
@@
 the log line from \daemon{bounce} that explains their source.  To deal with
-this the COMMIT action maintains a cache of recently committed bounce
-notification mails, which the BOUNCE action subsequently checks if the
-bounce mail is not found in the state tables.  If the queueid exists in the
-cache, and its start time is less than ten seconds before the timestamp of
-the bounce log line, it is assumed that the bounce notification mail has
-already been processed and the BOUNCE action does not create one.  If the
-queueid exists in the cache it is removed, because it has either just been
-used or the problem did not occur for that mail.  Whether the BOUNCE action
-creates a mail or finds an existing mail in the state tables, it flags the
-mail as having been seen by the BOUNCE action; if this flag is present the
-COMMIT action will not add the mail to the cache of recent bounce
-notification mails.  This is not required to correctly deal with the
-complication, but is an optimisation to reduce the parser's memory usage.
-On the occasions the author has observed this complication occurring there
-have been a huge number of bounce notification mails generated: if every
-bounce notification mail was cached it would dramatically increase the
-memory requirements of the parser.  The cache of bounce notification mails
-will be pruned whenever the parser's state is saved, though if the size of
-the cache ever becomes a problem it could be pruned periodically to keep it
-in check.
+this the \action{COMMIT} action maintains a cache of recently committed
+bounce notification mails, which the \action{BOUNCE\_CREATED} action
+subsequently checks if the bounce mail is not found in the state tables.
+If the queueid exists in the cache, and its start time is less than ten
+seconds before the timestamp of the bounce log line, it is assumed that the
+bounce notification mail has already been processed and the
+\action{BOUNCE\_CREATED} action does not create one.  If the queueid exists
+in the cache it is removed, because it has either just been used or the
+problem did not occur for that mail.  Whether the \action{BOUNCE\_CREATED}
+action creates a mail or finds an existing mail in the state tables, it
+flags the mail as having been seen by the \action{BOUNCE\_CREATED} action;
+if this flag is present the \action{COMMIT} action will not add the mail to
+the cache of recent bounce notification mails.  This is not required to
+correctly deal with the complication, but is an optimisation to reduce the
+parser's memory usage.  On the occasions the author has observed this
+complication occurring there have been a huge number of bounce notification
+mails generated: if every bounce notification mail was cached it would
+dramatically increase the memory requirements of the parser.  The cache of
+bounce notification mails will be pruned whenever the parser's state is
+saved, though if the size of the cache ever becomes a problem it could be
+pruned periodically to keep it in check.
 
@@
 \textit{after\/} the mail has been removed from the state tables and saved
-in the database.  The DELETE action adds deleted mails to a cache, which is
-checked by the SAVE\_DATA action, and the current log line discarded if the
-following conditions are met:
+in the database.  The \action{DELETE} action adds deleted mails to a cache,
+which is checked by the \action{SAVE\_DATA} action, and the current log
+line discarded if the following conditions are met:
 

### 2009-03-26 doc/logparser-implementation.tex
Added a \action{} command to have consistent formatting of action names. Changes \parsernames{} to have an ' before the s.

@@
         recognise rejection log lines, i.e.\ rules that have an action of
-        \texttt{DELIVERY\_REJECTED}; other rules should not set this field.
+        \action{DELIVERY\_REJECTED}; other rules should not set this field.
 
@@
         not a rejection, i.e.\ the action field of the associated rule is
-        not \texttt{DELIVERY\_REJECTED}.
+        not \action{DELIVERY\_REJECTED}.
 
@@
 about mails recently deleted by the administrator so that subsequent log
-lines processed by the SAVE\_DATA action can be discarded (\sectionref{Mail
-deleted before delivery is attempted}).
+lines processed by the \action{SAVE\_DATA} action can be discarded
+(\sectionref{Mail deleted before delivery is attempted}).
 
@@
 rules per action}.  Unsurprisingly, the action with the most associated
-rules is \texttt{DELIVERY\_REJECTED}, the action that processes Postfix
+rules is \action{DELIVERY\_REJECTED}, the action that processes Postfix
 rejecting a mail delivery attempt.  The next most common action is, perhaps
-surprisingly, \texttt{UNINTERESTING}: this action does nothing when
+surprisingly, \action{UNINTERESTING}: this action does nothing when
 invoked, allowing uninteresting log lines to be parsed without any effects
 (it does not imply that the input is ungrammatical or unparsed).  Generally
-rules specifying the \texttt{UNINTERESTING} action recognise log lines that
+rules specifying the \action{UNINTERESTING} action recognise log lines that
 are not associated with a specific mail, e.g.\ notices about configuration
@@
 variant, e.g.\ all log lines showing that a remote client has connected are
-recognised by a single rule and handled by the \texttt{CONNECT} action.
+recognised by a single rule and handled by the \action{CONNECT} action.
 \parsernames{} actions are described in detail in \sectionref{actions in
@@
 \textit{state table entry\/}; a connection becomes a mail during the
-\texttt{CLONE} action, which processes Postfix accepting a delivery
+\action{CLONE} action, which processes Postfix accepting a delivery
 attempt, and the data structure moves from the connections state table to
@@
         address if required (see \sectionref{Mail deleted before delivery
-        is attempted}), then invokes the COMMIT action to save the mail to
-        the database.
+        is attempted}), then invokes the \action{COMMIT} action to save the
+        mail to the database.
 
@@
         \sectionref{out of order log lines}; this action tags the mail as
-        having expired, so the COMMIT action will skip the valid
+        having expired, so the \action{COMMIT} action will skip the valid
         combinations check.
@@
     \item [MAIL\_BOUNCED] This action behaves exactly like the
-        \texttt{SAVE\_DATA} action; it saves all data captured by the
+        \action{SAVE\_DATA} action; it saves all data captured by the
         recognising rule's regex, and does nothing more.  It is a separate
@@
         administrator.  This action is used for those mails; it invokes the
-        \texttt{COMMIT} action, but is a separate action to simplify
+        \action{COMMIT} action, but is a separate action to simplify
         further analysis.
@@
     \item [MAIL\_SENT] This action behaves exactly like the
-        \texttt{SAVE\_DATA} action; it saves all data captured by the
+        \action{SAVE\_DATA} action; it saves all data captured by the
         recognising rule's regex, and does nothing more.  It is a separate
@@
 
-    \item [PICKUP] The PICKUP action corresponds to the \daemon{pickup}
-        service dealing with a locally submitted mail.  A new mail will be
-        created, although out of order log entries may have caused it to
-        already exist, as documented in \sectionref{pickup logging after
-        cleanup}.
+    \item [PICKUP] The \action{PICKUP} action corresponds to the
+        \daemon{pickup} service dealing with a locally submitted mail.  A
+        new mail will be created, although out of order log entries may
+        have caused it to already exist, as documented in
+        \sectionref{pickup logging after cleanup}.
 
@@
         the state tables saves any data captured by the recognising rule's
-        regex to it.  The \texttt{SAVE\_DATA} saves data in this way but
+        regex to it.  The \action{SAVE\_DATA} saves data in this way but
         does not do anything else; it is invoked for log lines that contain
@@
         to a remote address (see \sectionref{tracking re-injected mail} for
-        a full explanation).  \texttt{TRACK} will be called when dealing
+        a full explanation).  \action{TRACK} will be called when dealing
         with the parent mail, and will create the child mail if necessary.
-        \texttt{TRACK} checks if the child has already been tracked, either
+        \action{TRACK} checks if the child has already been tracked, either
         by this parent or by another parent, and issues appropriate
@@
     connection\_data    &                                                   \\
-    action              & DELIVERY\_REJECTED                                \\
+    action              & \action{DELIVERY\_REJECTED}                       \\
     hits                & 0                                                 \\
@@
 the restriction\_name field should be set when the rule's action is
-\texttt{DELIVERY\_REJECTED}; hopefully the name of the restriction will be
+\action{DELIVERY\_REJECTED}; hopefully the name of the restriction will be
 clear from the content of the log line.
@@
     \item [\_\_PID\_\_]  The \acronym{pid} of a \daemon{smtpd} process that
-        dies or is killed; see the SMTPD\_DIED action in
+        dies or is killed; see the \action{SMTPD\_DIED} action in
         \sectionref{actions in detail in implementation}.

### 2009-03-26 doc/logparser-complications.tex
Dealt with some XXX items.

@@
         have interrupted transfer of an accepted delivery attempt, or it
-        may have occurred after a mail delivery attempt was rejected.
+        may have occurred after a mail delivery attempt was rejected.  If a
+        mail needs to be discarded, the following checks will all pass:
 
-        XXX THE REMAINDER OF THIS DOES NOT MAKE SENSE\@; CHECK AND CORRECT
-        IT\@.  To distinguish between the two cases the action compares the
-        timestamp of the last accepted mail against the timestamp of the
-        last log line from \daemon{smtpd} for that connection; this depends
-        on the \action{CLONE} action keeping a list of all mails accepted
-        for each connection.  If the mail acceptance timestamp is later
-        than the timeout timestamp, the timeout applies to the
-        just-accepted mail, which will be removed from the state tables,
-        without entering it in the database.  If the \daemon{smtpd}
-        timestamp is later there was a rejection between the accepted mail
-        and the timeout: the action assumes that the timeout applies to the
-        rejected delivery attempt and finishes.  This assumption is not
+        \begin{itemize}
+
+            \item The timestamp of the log line preceding the timeout log
+                line is earlier than the timestamp of the last accepted
+                delivery attempt, i.e.\ there have not been any rejections
+                since then the delivery attempt was accepted.
+
+            \item The mail must exist in the state tables.
+
+            \item The mail does not have a \daemon{qmgr} log line.
+
+        \end{itemize}
+
+        If all the checks are passed the mail will be discarded from the
+        state table, and will not be entered in the database.  If one or
+        more checks are not passed it is assumed that the timeout happened
+        after a rejected delivery attempt.  This assumption is not
         necessarily correct, because Postfix may have accepted an earlier
         recipient and rejected a later one, in which case the timeout
-        applies to the accepted mail, which should be discarded; this has
-        not been a problem in practice.  Processing timeouts is further
-        complicated by the presence of out of order \daemon{cleanup} log
-        lines: see \sectionref{discarding cleanup log lines} for details.
+        applies to the partially accepted mail, which should be discarded;
+        this has not been a problem in practice.  Processing timeouts is
+        further complicated by the presence of out of order
+        \daemon{cleanup} log lines: see \sectionref{discarding cleanup log
+        lines} for details.
 
@@
 \action{SMTPD\_DIED} (2 \& 3), and \action{SMTPD\_WATCHDOG} (4).
-
-\newpage{} %XXX VARIOREF PROBLEM
 

### 2009-03-26 doc/logparser-implementation.tex
Dealt with some XXX items.

@@
 submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
-forwarding, or internally generated by Postfix.  Local submission is the
+forwarding, or generated internally by Postfix.  Local submission is the
 simplest case: a queueid is assigned immediately \flowchart{PICKUP}{2}, and
@@
 administrator when it encounters configuration errors, but such mails are
-presumably rare.  XXX WOULD THIS PARAGRAPH BE BETTER AS A LIST\@?
+presumably rare.
 
@@
     \item This is followed by rejection of sender address, recipient
-        addresses, client \acronym{IP} address or hostname, HELO hostname,
-        etc.\ \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more
-        mails \flowchart{CLONE}{5}; or some interleaving of both.
+        addresses, client \acronym{IP} address or hostname, etc.\
+        \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
+        \flowchart{CLONE}{5}; or some interleaving of both.
 
@@
         \acronym{SMTP} code of other results in that connection.  Some
-        final delivery log lines do not contain an \acronym{SMTP} code: in
-        those cases the \acronym{SMTP} code is specified by the rule's
+        final delivery log lines do not contain an \acronym{SMTP} code,
+        e.g.\ when Postfix delivers to a command; in those cases the
+        \acronym{SMTP} code is specified by the rule's
         \texttt{result\_data} field, based on the success or failure
-        represented by the log line.  XXX GIVE AN EXAMPLE LOG LINE\@.
+        represented by the log line.
 

### 2009-03-27 doc/logparser-implementation.tex
Wrote introduction for implementation chapter.

@@
 
-XXX WRITE AN INTRODUCTION\@.  MOST OF THIS IS CRAP, BUT I MIGHT BE ABLE TO
-RESCUE SOME OF IT\@.  IT SHOULD HAVE A BLURB, FOLLOWED BY A PREVIEW OF THE
-CHAPTER\@.
-
-The parser deals with all the complications of parsing, the eccentricities
-and oddities of Postfix log files, presenting the resulting data in a
-normalised, simple to use representation.  The parser's task is to follow
-the journey each mail takes through Postfix, combining the data captured by
-rules into a coherent whole, saving it in a useful and consistent form, and
-performing housekeeping duties.
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be able
-to handle multiple mails in parallel, each potentially at a different stage
-in its journey, without any interference between mails --- except in the
-minority of cases where intra-mail interference is required.  The best way
-to implement this is to maintain state information for every unfinished
-mail and manipulate the appropriate mail correctly for each log line
-encountered.
-
-There is a similarity between this design and the event-driven programming
-paradigm commonly used in GUI programs, where one part of the program
-responds to events (mouse clicks in a GUI program, log lines being
-recognised in \parsername{}) and invokes the correct action.
+This chapter documents the implementation of \parsername{}, based on the
+architecture described in \sectionref{parser architecture}.  Any design may
+look good when examined in the abstract, but the real test of the design
+comes with the first concrete implementation; only then do any practical
+difficulties come to light.  Implementing this architecture was relatively
+straightforward, with only minor changes required: rule conditions and
+cascaded parsing were added, although later the implementation stopped
+using cascaded parsing.  The difficulties came not from the architecture
+and design, but from anomalies in the log files and Postfix's behaviour, as
+described in \sectionref{complications}.
+
+\parsername{} deals with all the complications of parsing, the
+eccentricities and oddities of Postfix log files, presenting the resulting
+data in a normalised, simple to use representation.  The parser's task is
+to follow the journey each mail takes through Postfix, combining the data
+captured by rules into a coherent whole, saving it in a useful and
+consistent form, and performing housekeeping duties.
+
+This chapter begins with the assumptions under which \parsername{} was
+designed and written, followed by a flowchart showing the most common paths
+taken through Postfix and \parsername{}, with a description of the stages
+and transitions.  The topic of the next section is the database that
+provides storage for the parser: any future work that analyses gathered
+data will do so using the database, so the database schema acts as an
+\acronym{API}.  A diagram of the database schema is provided, plus
+documentation for every table and field.
+
+The next three sections document the implementation of the three components
+of the architecture.  First is the framework, including its initialisation
+phase, the parsing process, and the conveniences it offers to users; that
+section finishes with the features that are important for this thesis but
+not strictly necessary for parsing: the performance data collected by the
+framework, the optimisations that can be disabled to show their effect, and
+the debugging options it provides.  The second component of the
+architecture is the actions: their implementation is documented, with a
+graph of how often each action is specified by rules, and a description of
+why some actions are more popular than others.  Every action that is part
+of \parsername{} is documented, and the actions section concludes with the
+process of adding a new action, and how to determine the values for each of
+the required fields.  Rules are the final component of the architecture,
+and also the most visible to users, because it is likely that users will
+need to add their own rules to recognise their log lines.  An example rule
+used by \parsername{} is shown, with every field clearly documented; this
+is followed by a description of adding new rules, and how to determine the
+value of each of the fields.  \parsername{} provides a utility to create
+regexes from unrecognised log lines; the algorithm it uses is documented,
+including the differences between it and the original algorithm.  The rules
+section finishes with a discussion of how \parsername{} uses rule
+conditions and overlapping rules, plus a description of the regex snippets
+provided to ease the process of writing regexes.
+
+On first inspection, Postfix log files look like they will be simple to
+parse, but this is not the case.  The many complications and difficulties
+encountered while writing \parsername{}, and the solutions developed to
+overcome them, are documented in \sectionref{complications}.  This chapter
+finishes with a list of \parsernames{} limitations, and some possible
+improvements that could be implemented.
 
@@
 image}
+
+The intermingling of log entries from different mails immediately rules out
+the possibility of handling each mail in isolation; the parser must be able
+to handle multiple mails in parallel, each potentially at a different stage
+in its journey, without any interference between mails --- except in the
+minority of cases where intra-mail interference is required.  The best way
+to implement this is to maintain state information for every unfinished
+mail and manipulate the appropriate mail correctly for each log line
+encountered.
 
@@
 
-This section has presented \parsername{}, the parser implemented for this
+This chapter has presented \parsername{}, the parser implemented for this
 project, beginning with the assumptions made during its development.  A
@@
 including the steps it takes during initialisation, the parsing process,
-and the conveniences it offers to users.  The data collected by the
-framework is described, as are the ways in which various optimisations can
-be disabled to demonstrate their effect, and the debugging options the
+and the conveniences it offers to users.  The performance data collected by
+the framework is described, as are the ways in which various optimisations
+can be disabled to demonstrate their effect, and the debugging options the
 framework provides.  The implementation of actions in \parsername{} is
@@
 complications and difficulties encountered while writing \parsername{}, and
-the solutions developed to overcome them, are documented in depth, also
-documented are how some solutions interact and which action or actions the
+the solutions developed to overcome them, are documented in depth; also
+documented are how some solutions interact, and which action or actions the
 solution is implemented in.  This chapter concludes with a list of the

### 2009-03-27 doc/logparser-implementation.tex
Merged check-regexs into logs2regexs, and updated the documentation.

@@
 fixed pattern.  This classification can be used to group similar log lines
-and generate a regex to match each group of log lines.  There are 4 steps
+and generate a regex to match each group of log lines.  There are five steps
 in the algorithm:
@@
         threshold used to differentiate between fixed and variable words,
-        but once again 1\% of log lines is a good starting point.  The new
-        regexes are displayed for the user to add to the database, either
-        as new rules or merged into the regexes of existing rules; the
-        counter for each regex is also displayed, giving the user an
-        indication of how many of the log lines that regex should match.
-        Discarding regexes will result in some of the log lines not being
-        matched; when the rules have been augmented with the new regexes,
-        the original log files should be parsed again, and any remaining
-        unparsed log lines used as input to this utility.
+        but once again 1\% of log lines is a good starting point.
+
+    \item [Test the new regexes.]  The final step replaces keywords in the
+        new regexes in the same way as \parsername{} does, and compiles
+        each regex to check they are valid.  A match will be attempted
+        between all of the new regexes and each of the unparsed log lines
+        the regexes were built from; the user will be warned if a log line
+        is not matched by any regex, or if a log line is matched by more
+        than one regex.  The number of log lines each regex matches is
+        counted, as is the number of log lines matched by all regexes,
+        though a log line is counted once only, even if matched by more
+        than one regex.
+
+        This step is not performed by the original algorithm.
 
@@
 
-XXX MERGE logs2regexes WITH check-regexes.  CHECK IF SLCT DOES THIS STEP\@.
-
-A second utility is also provided that reads a list of new regexes and the
-input given to the first utility.  It tries to match each input log line
-with each regex, counting the number of log lines that match each regex,
-warning the user if an input log line is matched by more than one regex,
-and additionally warning if an input log line is not matched by any regex.
-It displays a summary of how many input log lines each regex matched,
-comparing it to the expected number of matches; this provides the user with
-an easy method of checking if the regexes produced by the first utility are
-correctly matching the input log lines they are based upon.
-
-These utilities are not intended to create perfect regexes, but they
-greatly reduce the effort required to parse new or different log lines.
+The new regexes are displayed for the user to add to the database, either
+as new rules or merged into the regexes of existing rules; also displayed
+are the number of log lines each regex was expected to match, and the
+number it actually matched, to help the user notice problems in the new
+regexes.  Discarding regexes will result in some of the log lines not being
+matched; when the rules have been augmented with the new regexes, the
+original log files should be parsed again, and any remaining unparsed log
+lines used as input to this utility.
+
+This utility is not intended to create perfect regexes, but it greatly
+reduces the effort required to deal with unrecognised log lines.
 

### 2009-03-28 doc/logparser-results.tex
Changes made after reading my paper.

@@
 XXX COMPARE BEST AND WORST POSSIBLE RESULTS WITH REAL RESULTS\@.
-
-XXX REUSE THE CONTENT FROM THE PAPER FOR THIS SECTION\@.
 
@@
 running the tests were disabled.
-
-\clearpage{}
 
@@
 At the time of writing there are \numberOFrules{} different rules: the top
-10\% recognise the vast majority of log lines, with the remaining log lines
-split across the other 90\% of the rules (as shown in \graphref{rule hits
-graph}).  Assuming that the distribution of log lines is reasonably
-consistent over time, parser efficiency should benefit from trying more
-frequently recognising rules before those that recognise less frequently.
-To test this hypothesis three full test runs were performed with different
-rule orderings:
+10\% recognise \input{build/include-top-ten-hits}\% of the log lines, with
+the remaining log lines split across the other 90\% of the rules (as shown
+in \graphref{rule hits graph}).  Assuming that the distribution of log
+lines is reasonably consistent over time, parser efficiency should benefit
+from trying more frequently recognising rules before those that recognise
+less frequently.  To test this hypothesis three full test runs were
+performed with different rule orderings:
 
@@
 files are given in \tableref{Parsing time relative to shuffled ordering
-table}.  Overall this optimisation provides a modest but worthwhile
-performance increase of approximately 10\%, for a small investment in time
-and programming.
+table}.  This optimisation provides a mean reduction in parsing time of
+\input{build/include-optimal-ordering-parsing-time-reduction-other-logs}\%
+with normal log files,
+\input{build/include-optimal-ordering-parsing-time-reduction-logs-22-62-68}\%
+when a mail loop occurs and the distribution of log lines is unusual.
+Optimal rule ordering has other benefits, described in
+\sectionref{scalability as the number of rules rises}.
 
@@
 files, to improve its parsing and ensure it correctly parses logs generated
-by later Postfix versions.  A second set of statistics was generated using
-the minimum ruleset and compared to the statistics generated using the full
-ruleset: the percentage parsing time increase when using the full ruleset
-instead of the minimal ruleset for optimal, shuffled, and reversed
-orderings is shown in \graphref{Percentage parsing time increase of maximum
-ruleset over minimum ruleset}.
+by later Postfix versions.  A second set of test runs was performed using
+the minimum ruleset, and the parsing times compared to those generated
+using the full ruleset: the percentage parsing time increase when using the
+full ruleset instead of the minimal ruleset for optimal, shuffled and
+reverse orderings is shown in \graphref{Percentage parsing time increase of
+maximum ruleset over minimum ruleset}.
 
@@
 
-It is clear from \graphref{Percentage parsing time increase of maximum
-ruleset over minimum ruleset} that the increased number of rules noticeably
-increases parsing time with reverse ordering, and to a lesser extent with
-shuffled ordering.  The optimal ordering (where the rules that recognise
-log lines most frequently are tried first) shows a mean increase of
+Clearly the increased number of rules has a noticeable performance impact
+with reverse ordering, and a lesser impact with shuffled ordering.  The
+optimal ordering (where the rules that recognise log lines most frequently
+are tried first) shows a mean increase of
 \input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a

### 2009-03-29 doc/logparser-results.tex
Corrections after proofreading.

@@
 
+% XXX IS CHANGING FIGURE TO GRAPH A GOOD IDEA?
+\renewcommand{\figurename}{Graph}
+
 XXX NEED AN INTRODUCTION\@.
-
-XXX COMPARE PARSING + ACTIONS TO JUST PARSING\@.
 
@@
 
-XXX COMPARE BEST AND WORST POSSIBLE RESULTS WITH REAL RESULTS\@.
-
 XXX DO A TEST-SOME RUN AND SAVE THE RESULTS TO THE DATABASE, THEN COMPARE
@@
 INTRODUCTION, NEAR THE MACHINE SPECS\@.
+
+XXX THINK ABOUT THE ORDERING OF TABLES AND GRAPHS\@; MAYBE REARRANGE THEM
+SO THAT THEY APPEAR IN THE ORDER THEY ARE REFERENCED\@?
+
+XXX SHOULD I USE $\backslash{}$clearpage{} OR SOMETHING TO AFFECT
+PLACEMENT\@?  GRAPHS AND TABLES TEND TO MOVE QUITE FAR IN SOME OF THE
+SUBSECTIONS, THOUGH MAKING THE TEXT LONGER MIGHT HELP TOO\@.  IDEALLY I
+WANT TO HAVE TEXT, THEN ALL GRAPHS AND TABLES, THEN THE NEXT SECTION\@; I
+DO NOT LIKE THAT GRAPHS/TABLES SOMETIMES APPEAR BEFORE THEY ARE
+REFERENCED\@.  I MIGHT NEED TO REDUCE THE HEIGHT OF THE GRAPHS TOO\@.
 
@@
 testing this parser accepts approximately 10,000 mails for 700 users a day;
-\graphref{Mails received per day} shows that as expected there are far more
-mails received on weekdays than at weekends.  The mail server in question
-is a production mail server handling mail for a university department; the
-benefit of using this server is that the log files from it exhibit the
+\graphref{Mails received per day} shows that, as expected, far more mails
+are received on weekdays than at weekends.  The mail server in question is
+a production mail server handling mail for a university department; the
+benefit of using this server is that its log files exhibit the
 idiosyncrasies and peculiarities a mail server in the wild must deal with,
-but the downside is that significantly altering the configuration just to
-log messages from a different Postfix component is not an option.  Median
-log file size is 50MB, containing 285,000 log lines --- large scale mail
-servers would have much larger log files.  Note that \graphref{Mails
-received per day} and \tableref{Number of mails received per day:
-statistics} show the number of mails received by \acronym{SMTP} only; in
-particular the mail loops noticeable in other graphs do not contribute to
-these figures.
+but the downside is that significantly altering the configuration to
+accommodate this project is not an option.  Median log file size is 50MB,
+containing 285,000 log lines; large scale mail servers would have much
+larger log files.  Note that \graphref{Mails received per day} and
+\tableref{Number of mails received per day: statistics} show the number of
+mails received by \acronym{SMTP} only; in particular the mail loops
+noticeable in later graphs do not contribute to these figures.
 
@@
 log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
-parsed 10 times and the parsing times averaged.  Saving results to the
-database was disabled for the test runs, as that dominates the run time of
-the program, and the tests are aimed at measuring the speed of the parser
-rather than the speed of the database and the disks the database is stored
-on.  The computer used for test runs was a Dell Optiplex 745, equipped
-with:
+parsed 10 times, and the parsing times from the 10 runs averaged.  Saving
+results to the database was disabled for the test runs, because that
+dominates the run time of the program, and the tests are aimed at measuring
+the speed of \parsername{} rather than the speed of the database and the
+disks the database is stored on.  The computer used for test runs was a
+Dell Optiplex 745, shown in \tableref{Computer used to generate
+statistics}.
 
+\begin{table}[htbp]
+    \caption{Computer used to generate statistics}
+    \empty{}\label{Computer used to generate statistics}
     \begin{tabular}[]{ll}
-
+        \tabletopline{}%
+        Component  & Component in use                                   \\
+        \tablemiddleline{}%
         CPU         & One dual core 2.40GHz Intel\textregistered{}
@@
         Hard disk   & One Seagate Barracuda 7200 RPM 250GB SATA disk.   \\
-
+        \tablebottomline{}%
     \end{tabular}
+\end{table}
 
@@
 \Graphref{parsing time vs file size vs number of log lines graph} shows the
-parsing time in seconds, file size in MB, and tens of thousands of log
-lines for each of the \numberOFlogFILES{} log files.  All three lines run
-roughly in parallel, giving the impression that the algorithm scales
-linearly with input size.  This impression is borne out by
-\graphref{parsing time vs file size vs number of log lines factor} which
+parsing time in seconds, file size in MB, and number of log lines in tens
+of thousands, for each of the \numberOFlogFILES{} log files.  The three
+lines run roughly in parallel, giving the impression that the algorithm
+scales linearly with input size.  This impression is borne out by
+\graphref{parsing time vs file size vs number of log lines factor}, which
 plots both the ratio of file size vs parsing time, and the ratio of number
@@
 shows the same ratios for different groups of log files.  The ratios are
-quite tightly banded, providing empirical evidence that the algorithm
-scales linearly.  The ratio increases (i.e.\ improves) for log files 22 and
-62--68 despite their larger than average size (shown in \graphref{parsing
-time vs file size vs number of log lines graph}); this is explained fully
-in \sectionref{Explaining the peaks in log file size}.
+quite tightly banded, showing that the algorithm scales linearly.  The
+ratios increase (i.e.\ improve) for log files 22 and 62--68 despite their
+larger than average size (shown in \graphref{parsing time vs file size vs
+number of log lines graph}); this is explained fully in
+\sectionref{Explaining the peaks in log file size}.
 
@@
 
+XXX SHOULD THIS BE A SEPARATE SECTION OR INCLUDED IN THE PREVIOUS ONE\@?
+MAYBE INCLUDE GRAPHS OF RULES PER FILE OR SOMETHING TO EXTEND THIS\@?
+
+XXX REWRITE WHEN STATS RUNS HAVE FINISHED\@; REWRITE ALL THE REFERENCES
+TOO\@.
+
 \Graphref{parsing time vs file size vs number of log lines graph} shows log
@@
 error message \newline{} \tab{} \texttt{mail forwarding loop for
-username@domain}\newline{}  Ordinarily this works well, but unfortunately
-in this instance the user noticed they had not received any mail in a while
-and opted to send a test mail to themselves, causing a loop not caught by
-Postfix:
+username@domain}\newline{}  Normally this works well, but the forwarding
+method in use at that time stripped the Delivered-To header from the mail,
+breaking loop detection.
 
-XXX REWRITE THIS SECTION SO THAT IT IS CORRECT\@: POINT 4 SAYS THAT THE
-ENVELOPE SENDER CHANGES, IS THAT CORRECT\@?
-
-\begin{enumerate}
-
-    \item Postfix accepts a mail from username@domain, for username@domain.
-
-    \item Postfix delivers the mail to the local mailbox and
-        username@domain, as instructed by the user's forwarding
-        instructions.  The forwarded mail has a
-        \texttt{Delivered-To:~username@domain} header added, and the
-        envelope sender address is username@domain.
-
-    \item Postfix accepts the mail for username@domain, but during delivery
-        it notices that the \texttt{Delivered-To} header already contains
-        the address it is currently delivering to, and therefore sends a
-        bounce notification with sender address \texttt{<>}\glsadd{<>} to
-        the original sender: username@domain.
-
-    \item Postfix accepts the bounce notification and delivers it to both
-        the local mailbox and to username@domain, as instructed by the
-        user's forwarding instructions.  A
-        \texttt{Delivered-To:~username@domain} header is added to the
-        forwarded bounce notification, XXX which now has an envelope sender
-        address of username@domain.   XXX IS THIS CORRECT\@?
-
-    \item Postfix accepts the forwarded bounce notification but during mail
-        delivery it notices that the \texttt{Delivered-To} header already
-        contains the address currently being delivered to, and sends a
-        bounce notification to the sender: username@domain.
-
-    \item Repeat from step two; this will continue indefinitely unless an
-        administrator intervenes and deletes the appropriate mails from the
-        queue.
-
-\end{enumerate}
-
-The sequence described above occurs extremely rapidly because Postfix does
-not have to deliver the mail to an external system, so mails are delivered,
-bounced, and generated as fast as the disks can keep up, resulting in a
-huge volume of log lines.
+When the recipient in a mail loop is local, Postfix does not have to
+deliver the mail to an external system, so mails are delivered as fast as
+the disks can keep up, resulting in a huge volume of log lines.
 
@@
 in \graphref{rule hits graph}).  Assuming that the distribution of log
-lines is reasonably consistent over time, parser efficiency should benefit
-from trying more frequently recognising rules before those that recognise
-less frequently.  To test this hypothesis three full test runs were
-performed with different rule orderings:
+lines is reasonably consistent over time, \parsernames{} efficiency should
+benefit from trying more frequently recognising rules before those that
+recognise less frequently.  To test this hypothesis three full test runs
+were performed with different rule orderings:
 
@@
 
-    \item [Shuffle] This is intended to represent a randomly ordered rule
-        set.  The rules will be shuffled once before use and will retain
-        that ordering until the parser exits.  Note that the ordering will
-        change every time the parser is executed, so 10 different rule
-        orderings will be generated for each log file in the test run.
+    \item [Shuffle] This ordering is intended to represent a randomly
+        ordered rule set.  The rules will be shuffled once before use and
+        will retain that ordering until the parser exits.  Note that the
+        ordering will change every time the parser is executed, so 10
+        different rule orderings will be generated for each log file in the
+        test run.
 
@@
 
+\showgraph{build/graph-hits}{Hits per rule}{rule hits graph}
+
 \showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
@@
 
+\newpage{} % XXX VARIOREF PROBLEM
+
 Differences in rule ordering have less effect on parsing time when parsing
@@
 When a mail loop occurs the vast majority of log lines are from Postfix
-components that have few rules associated with them, whereas in general the
+components that have few rules associated with them, whereas in general
 most log lines are from Postfix components that have a high number of
@@
 parse a log line from log files 22 or 62--68 being considerably lower than
-usual.  The ratio of number of log lines versus parsing time in
+usual.  The ratio of number of log lines vs.\ parsing time in
 \graphref{parsing time vs file size vs number of log lines factor} shows a
@@
 expected that the typical parser will have a large number of rules.  There
-are \numberOFrules{} rules in the full Postfix ruleset, whereas the minimum
+are \numberOFrules{} rules in the full \parsername{} ruleset, whereas the minimum
 number of rules required to parse the \numberOFlogFILES{} log files used
@@
 \numberOFrulesMINIMUMpercentage{} of the full ruleset.  The full ruleset is
-larger because the Postfix parser is tested with \numberOFlogFILESall{} log
+larger because \parsername{} is tested with \numberOFlogFILESall{} log
 files, to improve its parsing and ensure it correctly parses logs generated
@@
 with reverse ordering, and a lesser impact with shuffled ordering.  The
-optimal ordering (where the rules that recognise log lines most frequently
-are tried first) shows a mean increase of
+optimal ordering shows a mean increase of
 \input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
-\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.  Log
-files 22 and 62--68 show much smaller increases for shuffled and reversed
-ordering than other log files do, because the majority of log lines in
-those log files are parsed by Postfix components with a small number of
-parsing rules, so removing unnecessary rules has little effect on the total
-number of rules used.  These results show that the architecture scales very
-well as the number of rules increases, and that optimally sorting the rules
-is an important optimisation contributing to this scalability.
+\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.
+XXXXXXXXXXXX IMPROVE THIS EXPLANATION\@.  Log files 22 and 62--68 show much
+smaller increases for shuffled and reversed ordering than other log files
+do, because the majority of log lines in those log files are parsed by
+Postfix components with a small number of parsing rules, so removing
+unnecessary rules has little effect on the total number of rules used.
+These results show that the architecture scales very well as the number of
+rules increases, and that optimally sorting the rules is an important
+optimisation contributing to this scalability.
 
@@
 Before the Perl interpreter attempts to match a regex against a piece of
-text, the regex is compiled into an internal representation and
-optimised to improve the speed of matching.  This compilation and
-optimisation takes CPU time ---in many cases it takes far more CPU time
-than the actual matching.  If the interpreter is certain that a regex
-will not change it will automatically cache the compilation results for
-later use.   The results of compiling a dynamically generated regex can
-be cached and used in preference to the original regex, but it is the
-responsibility of the programmer to do this.  The Postfix parser loads the
-rules it uses from a database, and thus each rule's regex is compiled
-and the result cached to reduce parsing time.
+text, the regex is compiled into an internal representation and optimised
+to improve the speed of matching.  This compilation and optimisation takes
+CPU time: in many cases it takes far more CPU time than the actual
+matching.  If the interpreter is certain that a regex will not change, it
+will automatically cache the compilation results for later use.   The
+results of compiling a dynamically generated regex can be cached and used
+in preference to the original regex, but it is the responsibility of the
+programmer to do this.  \parsername{} loads the rules it uses from a
+database, so it compiles each rule's regex and caches the result to reduce
+parsing time.
 
@@
 shows the impact that not caching compiled regexes has on parser
-performance: on typical log files the parsing time when not caching
-compiled regexes is 400--600\% of the parsing time when caching.
-Caching compiled regexes is probably the single most effective
-optimisation possible in the parser's implementation, and was quite simple
-to implement.
+performance: on typical log files, the parsing time when not caching
+compiled regexes is 400--600\% of the parsing time when caching.  Caching
+compiled regexes is probably the single most effective optimisation
+possible in \parsername{}, and was quite simple to implement.
 
@@
 
-Two large dips can be seen in \graphref{Increase in parsing time when not
-caching compiled regexes graph} at log files 22 and 62--68, corresponding
-to the spikes in log file size in \graphref{parsing time vs file size vs
-number of log lines graph}.  The distribution of log lines across rules
-when there is a mail loop is much different to the norm, and the average
-number of rules consulted per log line is much lower: this results in far
-fewer regex compilations per line when there is a mail loop, and a
-corresponding decrease in the average parsing time for a single log line.
-The outcome is that caching compiled regexes is proportionally less
+XXX UPDATE THIS\@.  Two large dips can be seen in \graphref{Increase in
+parsing time when not caching compiled regexes graph} at log files 22 and
+62--68, corresponding to the spikes in log file size in \graphref{parsing
+time vs file size vs number of log lines graph}.  The distribution of log
+lines across rules when there is a mail loop is much different to the norm,
+and the average number of rules consulted per log line is much lower: this
+results in far fewer regex compilations per line when there is a mail loop,
+and a corresponding decrease in the average parsing time for a single log
+line.  The outcome is that caching compiled regexes is proportionally less
 important when the log file contents were created by a mail loop.  The
@@
 
-\showtable{build/include-cached-regexes-vs-discarded-regexes} {Increase in
-parsing time when not caching compiled regexes} {Increase in parsing time
+\showtable{build/include-cached-regexes-vs-discarded-regexes}{Increase in
+parsing time when not caching compiled regexes}{Increase in parsing time
 when not caching compiled regexes table}
@@
 \subsection{Summary}
+
+XXX REWRITE WHEN THIS CHAPTER IS FINISHED\@.
 
@@
 recognise fewer than 100 log lines.
-
-\showgraph{build/graph-hits}{Hits per rule}{rule hits graph}
 

### 2009-03-30 doc/logparser-introduction.tex
Wrote introduction for implementation chapter.
Major work on the introduction chapter: the blurb at the start has been extended with content from the SGAI paper, and the layout section is more or less finished.

@@
 
-The architecture described herein was developed as part of a larger project
-to improve anti-spam defences by analysing the performance of the set of
-filters currently in use, optimising the order and membership of the set
-based on that analysis, and developing supplemental filters where
-deficiencies are identified.  Most anti-spam techniques are content-based
-(e.g.~\cite{a-plan-for-spam, word-stemming, relaxed-online-svms}) and
-require the mail to be accepted before determining if it is spam, but
-rejecting mail during the delivery attempt is preferable: senders of
-non-spam mail that is mistakenly rejected will receive an immediate
-non-delivery notice; resource usage is reduced on the accepting mail server
-(allowing more intensive content-based techniques to be used on the mail
-that is accepted); users have less spam mail to wade through.  Improving
-the performance of anti-spam techniques that are applied when mail is being
-transferred via \acronym{SMTP} is the goal of this project, by providing a
-platform for reasoning about anti-spam filters.  The approach chosen to
-measure performance is to analyse the log files produced by the
-\acronym{SMTP} server in use, Postfix, rather than modifying its source
-code to generate statistics: this approach improves the chances of other
-Postfix users testing and using the software developed for this project.
-The need arose for a parser capable of dealing with the great number and
-variety of log lines produced by Postfix: the parser must be designed so
-that adding support for parsing new inputs is a simple task, because the
-log lines to be parsed will change over time.
+The architecture and implementation described in this thesis was developed
+as the foundation of a larger project to improve anti-spam defences, by
+analysing the performance of the set of filters currently in use,
+optimising the order and membership of the set based on that analysis, and
+developing supplemental filters where deficiencies are identified.  Most
+anti-spam techniques are content-based (e.g.~\cite{a-plan-for-spam,
+relaxed-online-svms, word-stemming}) and require the mail to be accepted
+before determining if it is spam, but rejecting mail during the delivery
+attempt is preferable: senders of non-spam mail that is mistakenly rejected
+will receive an immediate non-delivery notice; resource usage is reduced on
+the accepting mail server, allowing more intensive content-based techniques
+to be used on the mail that is accepted; users have less spam mail to wade
+through.  Improving the performance of anti-spam techniques that are
+applied when mail is being transferred via \acronym{SMTP} is the goal of
+this project, by providing a platform for reasoning about anti-spam
+filters.
+
+The approach chosen to measure performance is to analyse the log files
+produced by the \acronym{SMTP} server in use, Postfix, rather than
+modifying its source code to generate statistics: this approach improves
+the chances of other Postfix users testing and using the software developed
+for this project.  The need arose for a parser capable of dealing with the
+great number and variety of log lines produced by Postfix: the parser must
+be designed so that adding support for parsing new inputs is a simple task,
+because the log lines to be parsed will change over time.  This variety in
+log lines occurs for several reasons:
+
+\begin{itemize}
+
+    \squeezeitems{}
+
+    \item Log lines differ amongst versions of Postfix.
+
+    \item The mail administrator can define custom rejection messages.
+
+    \item Policy servers~\cite{policy-servers} may log different messages
+        depending on the characteristics of the connection.
+
+    \item Every \acronym{DNSBL}\footnote{This thesis is supplied with a
+        glossary (\textsection\ref{Glossary}) and a list of acronyms
+        (\textsection\ref{Acronyms}).} returns a different explanatory
+        message.
+
+\end{itemize}
 
@@
 \urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
-complication arises because the set of rejection messages is not fixed: new
-messages can be added by the administrator with custom checks; every
-\acronym{DNSBL}\footnote{This thesis is supplied with a glossary
-(\textsection\ref{Glossary}) and a list of acronyms
-(\textsection\ref{Acronyms}).} returns a different explanatory message;
-policy servers may log different messages depending on the characteristics
-of the connection.  There are many ways in which the log lines may differ
-between servers, even within the same organisation: servers may be
-configured differently, or running different versions of Postfix.
+complication arises because the set of rejection messages is not fixed, and
+there are many ways in which the log lines may differ between servers, even
+within the same organisation, where servers may be configured differently,
+or even running different versions of Postfix.
+
+It was hoped to reuse an existing parser rather than writing one from
+scratch, but the effort required to adapt and improve an existing parser
+was judged to be greater than the effort to write a new one, because the
+techniques used by the existing parsers severely limited their potential:
+
+\begin{itemize}
+
+    \item Some ignored the majority of log lines, parsing specific log
+        lines accurately, but without any provision for parsing new or
+        similar log lines.
+
+    \item Others sloppily parsed the majority of log lines, but were
+        incapable of distinguishing between log lines of the same category,
+        e.g.\ rejecting a mail delivery attempt.
+
+    \item None of the reviewed parsers extracted enough data for the
+        purposes of this project.
+
+\end{itemize}
+
+The only prior published work on the subject of parsing Postfix log files
+that the author is aware of is \textit{Log Mail Analyser: Architecture and
+Practical Utilizations\/}~\cite{log-mail-analyser}, which aims to extract
+data from log files, correlate it, and present it in a form suitable for a
+systems administrator to search using the myriad of standard Unix text
+processing utilities already available.  A full state of the art review is
+provided in \sectionref{state of the art review}, beginning with a review
+of \textit{Log Mail Analyser}.
+
+The architecture developed is conceptually simple: provide a few generic
+functions (\textit{actions\/}), each capable of dealing with an entire
+category of inputs (e.g.\ rejecting a mail delivery attempt), accompanied
+by a multitude of precise patterns (\textit{rules\/}), each of which
+recognises one input variant and only that variant (e.g.\ rejection
+by a specific \acronym{DNSBL}).  This architecture is ideally suited to
+parsing inputs where the input is not fully understood or does not conform
+to a fixed grammar: the architecture warns about unparsed inputs and other
+errors, but continues parsing as best it can, allowing the developer of a
+new parser to decide which deficiencies are most important and require
+immediate attention, rather than being forced to fix the first error that
+arises.
 
@@
 and reconfiguring appropriately; identifying regular high volume uses
-(e.g.\ customer newsletters) and restricting those uses to off-peak times
+(e.g.\ customer newsletters) and restricting those uses to off-peak times,
 or providing a dedicated service for them; detecting virus outbreaks that
@@
 Chapter~\ref{parser architecture} describes the parser architecture
-developed for this project, beginning with an overview of the architecture,
-then describing each of the components of the architecture in detail.
+developed for this project, beginning with an overview, then describing
+each of the components of the architecture in detail.  This chapter
+concentrates on the abstract, theoretical, implementation-independent
+aspects of the architecture; the practical aspects are deferred until
+chapter~\ref{Postfix Parser Implementation}.
 
-XXX THIS ALL NEEDS TO BE CHECKED AND UPDATED\@.
+Chapter~\ref{Postfix Parser Implementation} documents \parsername{}, the
+parser based on the architecture described in chapter~\ref{parser
+architecture}.  The practical difficulties of implementing each of the
+components of the architecture are described, accompanied by the many
+complications encountered when parsing Postfix log files, and other
+important details of the implementation.
 
-Chapter~\ref{Postfix Parser Implementation} discusses the parsing rules in
-detail, explaining the purpose and usage of each field in a rule, referring
-to an example rule and sample log line it successfully recognises.  The
-pros and cons of overlapping rules are considered, including techniques for
-detecting unintentional overlaps.  Rule efficiency concerns are discussed,
-in particular the optimisations used by the algorithm.  The section
-concludes with a description of using the tools provided with the parser to
-generate new rules (specifically the regex in each rule) from unparsed log
-lines.  It contains the core of the paper, describing a naive parsing
-algorithm and the complications initially encountered that shaped the full
-algorithm.  A flow chart and a discussion of the emergent behaviour
-exhibited by the algorithm accompanies a comprehensive explanation of the
-different stages of the initial algorithm.  The framework that actions and
-rules fit into is documented, then the actions taken during execution of
-the algorithm are described, followed by the process of adding a new
-action.  The section concludes with an in-depth description of the further
-complications discovered, and their solutions that complete the parser.
-This algorithm requires a database for storing both the rules used when
-parsing and the results gleaned from parsing.  The database schema used is
-described in \sectionref{database}, explaining in detail the tables used
-for storing the data gleaned from the log files and the table that stores
-the rules.
+Chapter~\ref{Results} evaluates \parsernames{} efficiency, discussing the
+various optimisations implemented in the framework, and the effect they
+have.  It also discusses the coverage achieved by \parsername{} over
+\numberOFlogFILES{} log files, with separate discussions of the number of
+log lines correctly recognised, and the number of connections and mails
+whose journey through Postfix was correctly reconstructed.
 
-Chapter~\ref{parsing coverage} analyses the coverage the parser achieves
-over a set of \numberOFlogFILES{} log files taken from a mail server
-handling mail for over 700 users, averaging 8500 mails per day
-(\graphref{Mails received per day}).  Coverage is described both
-in terms of the fraction of log lines parsed and the fraction of mails and
-connections successfully reconstructed by the parsing algorithm; dealing
-with false negatives and a discussion of the difficulties in identifying
-false positives is also included.  As part of determining the coverage of
-the parser a random sampling of log lines was parsed, and the correctness
-of the results manually verified.
-
-Chapter~\ref{conclusion} contains the conclusion of the thesis, describing
-the results of the research, design, and implementation of the parser.
+Chapter~\ref{conclusion} contains the conclusion of the thesis.
 
 The bibliography (appendix~\ref{bibliography}) contains references to the
-resources used in developing the algorithm, writing the program, and
-preparing this thesis.
+resources used in designing the architecture and writing \parsername{}.
 
@@
 
-Appendix~\ref{Acronyms} provides a list of acronyms used in the thesis.
+Appendix~\ref{Acronyms} contains a list of acronyms used in the thesis;
+uncommon acronyms will have an entry in the glossary too.
 
-Appendix~\ref{Postfix Daemons} gives a brief description of Postfix
-daemons.
+Appendix~\ref{Postfix Daemons} concludes this thesis with a brief
+description of Postfix daemons.
 
@@
 
-Portions of \textsection\ref{introduction}, \textsection\ref{parser
-architecture}, \textsection\ref{Postfix Parser Implementation}, and
-\textsection\ref{Results} have previously been published at a
+Portions of chapters~\ref{introduction} and~\ref{parser
+architecture}--\ref{Results} have previously been published at a
 conference~\cite{sgai-2008}, and later reprinted in a
-journal~\cite{elsevier-2009}.  Publication of the conference paper and
-journal article was supported by Science Foundation Ireland
-RFP~05/RF/CMS002.
+journal~\cite{elsevier-2009}.  Publication of the conference paper was
+supported by Science Foundation Ireland RFP~05/RF/CMS002.

### 2009-03-30 doc/logparser.tex
Changes from proofreading, and a couple of other checks.
Added a \action{} command to have consistent formatting of action names. Changes \parsernames{} to have an ' before the s.
Dealt with some XXX items.
Changes made after reading my paper.
Corrections after proofreading.
Major work on the introduction chapter: the blurb at the start has been extended with content from the SGAI paper, and the layout section is more or less finished.

@@
 % Sort numbers where there are multiple citations.  Does not appear to have
-% the expected effect (probably clashes with varioref or hyperref), though
-% it does reduce the space between numbers.
+% the expected effect (probably clashes with hyperref), though it does
+% reduce the space between numbers.
 \usepackage{cite}
@@
 \newcommand{\showgraph}[3]{%
-    % XXX REVISIT THE PLACEMENT OF FIGURES, THOUGH I THINK LATEX'S DEFAULTS
-    % AR EPROBABLY PRETTY GOOD.
+    % XXX REVISIT THE PLACEMENT OF FIGURES AND TABLES, THOUGH I THINK
+    % LATEX'S DEFAULTS ARE PROBABLY PRETTY GOOD.
     \begin{figure}%[hbt!]
@@
 \newcommand{\showtable}[3]{
-    \begin{table}[ht]
+    \begin{table}%[ht]
         \caption{#2}\label{#3}
@@
 \newcommand{\flowchart}[2]{%
-    (action: #1\@; flowchart:~#2)%
+    (action: \action{#1}\@; flowchart:~#2)%
+}
+
+\newcommand{\action}[1]{%
+    \texttt{#1}%
 }
@@
 
+% XXX UPDATE THIS WHEN all-logs PASSES
 \newcommand{\numberOFlogFILESall}[0]{%
@@
 \newcommand{\parsername}{\acronym{PLP}}
-\newcommand{\parsernames}{\acronyms{PLP}}
+\newcommand{\parsernames}{\acronym['s]{PLP}}
 \newcommand{\parsernamelong}{Postfix Log Parser}
@@
 
+    XXX REWRITE WHEN THESIS IS FINISHED\@.
+
     Parsing Postfix log files is much more difficult than it first appears,
@@
 
+    XXX ABSTRACT FROM SGAI PAPER\@.
+
+    Some parsers need to be very precise and strict when parsing, yet must
+    allow users to easily adapt or extend the parser to parse new inputs,
+    without requiring that the user have an in-depth knowledge and
+    understanding of the parser's internal workings.  This paper presents a
+    novel parsing architecture, designed for parsing Postfix log files,
+    that aims to make the process of parsing new inputs as simple as
+    possible, enabling users to trivially add new rules (to parse variants
+    of existing inputs) and relatively easily add new actions (to process a
+    previously unknown category of input).  The architecture scales
+    linearly or better as the number of rules and size of input increases,
+    making it suitable for parsing large corpora or months of accumulated
+    data.
+
     \SVN$Id$
@@
 
-    XXX ADD A $\backslash{}$action\{\} COMMAND TO FORMAT ACTION NAMES
-    CONSISTENTLY\@.
-
     XXX FIX THE PROBLEMS THAT AROSE FROM THE LAST all-logs RUN\@.
-
-    XXX PLPs SHOULD BE PLP's IN A LOT OF PLACES\@; ADD A NEW COMMAND AND
-    CHECK THE CURRENT USAGE\@.
 
@@
     CONCLUSION\@?
+
+    XXX MAIL MADS ABOUT EXAMINERS\@.
 

### 2009-03-31 doc/logparser-background.tex
Changes from proofreading of introduction chapter.

@@
 \begin{table}[ht]
-    \caption{Sample output from SQL query showing percentage of delivery
-    attempts rejected by each restriction}
+    \caption{Sample output from SQL query, showing the percentage of
+    delivery attempts rejected by top ten restrictions}
     \empty{}\label{Sample output from SQL query}
@@
 
-\subsection{Mixing and Matching Postfix Restrictions}
+\subsection{Mixing And Matching Postfix Restrictions}
 
@@
         \numberOFlogFILES{} log files used when generating the results in
-        \sectionref{Results} are shown in \tableref{Summary of rejections}.
-        The table shows that greylisting is worth using at the moment,
-        particularly when you take into account its position as the final
-        restriction that a mail must overcome in the configuration used on
-        the mail server that generated the log files: on that server
-        greylisting only takes effect for mails that have passed all other
-        restrictions.  Some problems may be encountered when using
-        greylisting: some servers fail to retry after a temporary failure,
-        or legitimate mail may be delayed, particularly when coming from a
-        pool of servers.
+        \sectionref{Results} are shown in \tableref{Most effective Postfix
+        restrictions}.  The table shows that greylisting is worth using at
+        the moment, particularly when you take into account its position as
+        the final restriction that a mail must overcome in the
+        configuration used on the mail server that generated the log files:
+        on that server greylisting only takes effect for mails that have
+        passed all other restrictions.  Some problems may be encountered
+        when using greylisting: some servers fail to retry after a
+        temporary failure, or legitimate mail may be delayed, particularly
+        when coming from a pool of servers.
 
         \begin{table}[ht]
-            \caption{Summary of rejections}\label{Summary of rejections}
+            \caption{Most effective Postfix restrictions}
+            \empty{}\label{Most effective Postfix restrictions}
             \input{build/include-restriction-summary-table}

### 2009-03-31 doc/logparser-introduction.tex
Changes from proofreading of introduction chapter.

@@
 
-The architecture and implementation described in this thesis was developed
+The architecture and implementation described in this thesis were developed
 as the foundation of a larger project to improve anti-spam defences, by
-analysing the performance of the set of filters currently in use,
-optimising the order and membership of the set based on that analysis, and
-developing supplemental filters where deficiencies are identified.  Most
-anti-spam techniques are content-based (e.g.~\cite{a-plan-for-spam,
-relaxed-online-svms, word-stemming}) and require the mail to be accepted
-before determining if it is spam, but rejecting mail during the delivery
-attempt is preferable: senders of non-spam mail that is mistakenly rejected
-will receive an immediate non-delivery notice; resource usage is reduced on
-the accepting mail server, allowing more intensive content-based techniques
-to be used on the mail that is accepted; users have less spam mail to wade
-through.  Improving the performance of anti-spam techniques that are
-applied when mail is being transferred via \acronym{SMTP} is the goal of
-this project, by providing a platform for reasoning about anti-spam
-filters.
+analysing the performance of the set of anti-spam techniques currently in
+use, optimising the order and membership of the set based on that analysis,
+and developing supplemental anti-spam techniques where deficiencies are
+identified.  Most anti-spam techniques are content-based
+(e.g.~\cite{a-plan-for-spam, relaxed-online-svms, word-stemming}) and
+require a mail to be accepted before determining if it is spam, but
+rejecting mail during the delivery attempt is preferable: senders of
+non-spam mail that is mistakenly rejected will receive an immediate
+non-delivery notice; resource usage is reduced on the accepting mail
+server, allowing more intensive content-based techniques to be used on the
+remaining mail that is accepted; users have less spam mail to wade through.
+Improving the performance of anti-spam techniques that are applied when
+mail is being transferred via \acronym{SMTP} is the goal of this project,
+by providing a platform for reasoning about the performance of anti-spam
+techniques.
 
@@
 great number and variety of log lines produced by Postfix: the parser must
-be designed so that adding support for parsing new inputs is a simple task,
-because the log lines to be parsed will change over time.  This variety in
-log lines occurs for several reasons:
+be designed and implemented so that adding support for parsing new inputs
+is a simple task, because the log lines to be parsed will change over time.
+This variety in log lines occurs for several reasons:
 
@@
 
-    \item The mail administrator can define custom rejection messages.
+    \item The mail server administrator can define custom rejection
+        messages.
 
@@
 whether it was to debug a problem, explain to a user why their mail is
-being rejected, or check if new anti-spam measures are working.  The more
+being rejected, or check if new anti-spam techniques are working.  The more
 adventurous will have generated statistics to show how successful each of
@@
 or users.\footnote{This was the first real foray the author, a Systems
-Administrator for a network of over 2,000 computers and over 1,800 users,
+Administrator for a network of over 2000 computers and over 1800 users,
 took into processing Postfix log files.}  Very few will have performed
@@
 \urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
-complication arises because the set of rejection messages is not fixed, and
-there are many ways in which the log lines may differ between servers, even
-within the same organisation, where servers may be configured differently,
-or even running different versions of Postfix.
+complication arises because the set of log lines is not fixed, and there
+are many ways in which log lines may differ between servers, even within
+the same organisation, where servers may be configured differently or even
+running different versions of Postfix.
 
@@
 
-    \item Some ignored the majority of log lines, parsing specific log
-        lines accurately, but without any provision for parsing new or
+    \item Some parsers ignored the majority of log lines, parsing specific
+        log lines accurately, but without any provision for parsing new or
         similar log lines.
 
-    \item Others sloppily parsed the majority of log lines, but were
-        incapable of distinguishing between log lines of the same category,
-        e.g.\ rejecting a mail delivery attempt.
+    \item Other parsers sloppily parsed the majority of log lines, but were
+        incapable of distinguishing between log lines in one same category,
+        e.g.\ not distinguishing between different anti-spam techniques
+        causing the rejection of delivery attempts.
 
-    \item None of the reviewed parsers extracted enough data for the
-        purposes of this project.
+    \item None of the reviewed parsers extracted enough data for future
+        in-depth analysis.
 
@@
 systems administrator to search using the myriad of standard Unix text
-processing utilities already available.  A full state of the art review is
-provided in \sectionref{state of the art review}, beginning with a review
-of \textit{Log Mail Analyser}.
+processing utilities already available; it is reviewed alongside the other
+parsers in the State of the Art Review in chapter~\ref{state of the art
+review}.
 
-The architecture developed is conceptually simple: provide a few generic
-functions (\textit{actions\/}), each capable of dealing with an entire
-category of inputs (e.g.\ rejecting a mail delivery attempt), accompanied
-by a multitude of precise patterns (\textit{rules\/}), each of which
-recognises one input variant and only that variant (e.g.\ rejection
-by a specific \acronym{DNSBL}).  This architecture is ideally suited to
-parsing inputs where the input is not fully understood or does not conform
-to a fixed grammar: the architecture warns about unparsed inputs and other
-errors, but continues parsing as best it can, allowing the developer of a
-new parser to decide which deficiencies are most important and require
-immediate attention, rather than being forced to fix the first error that
-arises.
+Once it was decided that a new parser would be written, an architecture was
+required to base the implementation on.  Existing architectures are
+tailored towards parsing inputs with a fixed grammar or a tightly
+constrained format, whereas Postfix log files lack any form of constraint,
+as outlined earlier.  A new architecture was designed and developed for
+this parser, with the hope that it will be useful to others.  The resulting
+architecture is conceptually simple: provide a few generic functions
+(\textit{actions\/}), each capable of dealing with an entire category of
+inputs (e.g.\ rejecting a mail delivery attempt), accompanied by a
+multitude of precise patterns (\textit{rules\/}), each of which recognises
+one input variant and only that variant (e.g.\ rejection by a specific
+\acronym{DNSBL}), and specifies which action will process the inputs it
+recognises.  This architecture is ideally suited to parsing inputs
+where the input is not fully understood or does not conform to a fixed
+grammar: the architecture warns about unparsed inputs and other errors, but
+continues parsing as best it can, allowing the developer of a new parser to
+decide which deficiencies are most important and require immediate
+attention, rather than being forced to fix the first error that arises.
 
@@
 resulting data into a database for later analysis.  The gathered data can
-be used to optimise current anti-spam measures, provide a baseline to test
-new anti-spam measures against, or to produce statistics showing how
-effective those measures are.  Numerous other uses are possible for such
-data: improving server performance by identifying troublesome destinations
-and reconfiguring appropriately; identifying regular high volume uses
-(e.g.\ customer newsletters) and restricting those uses to off-peak times,
-or providing a dedicated service for them; detecting virus outbreaks that
-propagate via mail; as a base for billing customers on a shared server.
-Preserving the raw data enables users to develop a multitude of uses far
-beyond those conceived of by the author.
+be used to optimise current anti-spam defences, to produce statistics
+showing how effective each technique in use is, or to provide a baseline to
+test new anti-spam measures against.  Numerous other uses are possible for
+such data: improving server performance by identifying troublesome
+destinations and reconfiguring appropriately; identifying regular high
+volume uses (e.g.\ customer newsletters) and restricting those uses to
+off-peak times, or providing a dedicated service for them; detecting virus
+outbreaks that propagate via mail; as a base for billing customers on a
+shared server.  Preserving the raw data enables users to develop a
+multitude of uses far beyond those conceived of by the author.
 
@@
 Chapter~\ref{background} provides background information useful in
-understanding the motivation behind the project, \acronym{SMTP}, and
-Postfix.
+understanding \acronym{SMTP}, Postfix, and the motivation behind the
+project.
 
-Chapter~\ref{state of the art review} reviews both the previously published
+Chapter~\ref{state of the art review} reviews the previously published
 research in this area and other Postfix log file parsers, discussing why
@@
 concentrates on the abstract, theoretical, implementation-independent
-aspects of the architecture; the practical aspects are deferred until
-chapter~\ref{Postfix Parser Implementation}.
+aspects of the architecture; discussion of the practical aspects is
+deferred until chapter~\ref{Postfix Parser Implementation}.
 
@@
 components of the architecture are described, accompanied by the many
-complications encountered when parsing Postfix log files, and other
-important details of the implementation.
+complications encountered when parsing Postfix log files, and other details
+of the implementation.
 
@@
 have.  It also discusses the coverage achieved by \parsername{} over
-\numberOFlogFILES{} log files, with separate discussions of the number of
-log lines correctly recognised, and the number of connections and mails
-whose journey through Postfix was correctly reconstructed.
+\numberOFlogFILES{} log files, with separate sections for the number of log
+lines correctly recognised, and the number of connections and mails whose
+journey through Postfix was correctly reconstructed.
 

### 2009-03-31 doc/logparser-results.tex
Major work on the introduction chapter: the blurb at the start has been extended with content from the SGAI paper, and the layout section is more or less finished.
Changes from proofreading of introduction chapter.

@@
-\chapter{Results and Evaluation}
+\chapter{Results And Evaluation}
 
@@
 disks the database is stored on.  The computer used for test runs was a
-Dell Optiplex 745, shown in \tableref{Computer used to generate
-statistics}.
+Dell Optiplex 745, shown in \tableref{Details of the computer used to
+generate statistics}.
 
 \begin{table}[htbp]
-    \caption{Computer used to generate statistics}
-    \empty{}\label{Computer used to generate statistics}
+    \caption{Details of the computer used to generate statistics}
+    \empty{}\label{Details of the computer used to generate statistics}
     \begin{tabular}[]{ll}
@@
 \showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
-and number of log lines}{parsing time vs file size vs number of log lines
-graph}
+and number of log lines for \numberOFlogFILES{} log files}{parsing time vs
+file size vs number of log lines graph}
 
@@
 
-\subsection{Explaining the Peaks in Log File Size}
+\subsection{Explaining The Peaks In Log File Size}
 
@@
 
+\clearpage{} % XXX GRAPHS MOVE AROUND A LOT
 \subsection{Rule Ordering For Efficiency}
@@
 
-\showgraph{build/graph-hits}{Hits per rule}{rule hits graph}
+\clearpage{} % XXX GRAPHS MOVE AROUND DURING REBUILDS
+
+\showgraph{build/graph-hits}{Log lines recognised per rule}{rule hits graph}
 

### 2009-03-31 doc/logparser.tex
Changes from proofreading of introduction chapter.

@@
 \newcommand{\parsernamelong}{Postfix Log Parser}
+\newcommand{\parsernameshort}{PLP}
 
@@
     but it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the log files, and thus accuracy in reconstructing the
-    actions taken by Postfix to generate the log files.  This paper
-    describes the creation of a parser for Postfix log files, documenting
-    the architecture developed and the parser implementing it, explaining
-    the difficulties encountered and the solutions developed, with
-    reference to an implementation that stores data gleaned from the log
-    files in an SQL database.  The gathered data can then be used to
-    optimise current anti-spam measures, provide a baseline to test new
-    anti-spam measures against, or to produce statistics showing how
-    effective those measures are.
+    understanding the log files and reconstructing the actions taken by
+    Postfix to generate the log files.  This paper describes the creation
+    of a parser for Postfix log files, documenting the architecture
+    developed and the parser implementing it, explaining the difficulties
+    encountered and the solutions developed, with reference to an
+    implementation that stores data gleaned from the log files in an SQL
+    database.  The gathered data can then be used to optimise current
+    anti-spam measures, provide a baseline to test new anti-spam measures
+    against, or to produce statistics showing how effective those measures
+    are.
 

### 2009-03-31 doc/logparser-implementation.tex
Changes from proofreading of introduction chapter.
Added a new action, DELIVERY_ERROR, to save data from delivery errors that     happen before a disconnection. Changed some UNINTERESTING rules to use DELIVERY_ERROR and SAVE_DATA. Removed connection_data fields from UNINTERESTING rules. Added a debug field to rules, so that individual rules can have debugging     information dumped when they recognise a log line. Improved the abstract, though it still needs work. Documented everything in the thesis.

@@
 
-\showgraph{build/logparser-flow-chart-part-1}{Parser flow chart}{flow chart
-image}
+\showgraph{build/logparser-flow-chart-part-1}{\parsernamelong{} flow
+chart}{flow chart image}
 
@@
         optional.
+
+    \item [debug] If this field is true, a warning will be issued with
+        information about the rule and the log line every time this rule
+        recognises a log line.  This field is optional.
 
@@
         recognised can be printed, so that the user can verify that the
-        correct rule recognised each log line.  A possible variant of this
-        would be to mark specific rules so only their regexes and the log
-        lines they recognise would be printed.
+        correct rule recognised each log line.  Individual rules can also
+        set their debug field to true, and debugging information will be
+        printed each time they recognise a log line.
 
@@
 
-\subsection{Actions in the \parsernamelong{}}
+\subsection{Actions In The \parsernamelong{}}
 
@@
         mail to the database.
+
+    \item [DELIVERY\_ERROR] Process log lines indicating that an error
+        occurred and the remote client disconnected; the log lines
+        processed by this action will be followed by log lines processed by
+        the \action{DISCONNECT} action, so all this action does is save
+        data from the log line.
 
@@
 
-    \caption{Example rule}
+    \caption{Example of a rule used by \parsernameshort{}}
     \empty{}\label{Example rule in implementation table}
@@
     priority            & 0                                                 \\
+    debug               & 0                                                 \\
     %cluster\_group      & 400                                               \\
@@
         but is otherwise unused by the parser.
+
+    \item [debug] controls the debugging information that can be generated
+        when this rule recognises a log line.
 
@@
 
-\section{Limitations and Possible Improvements}
+\section{Limitations And Possible Improvements}
 

### 2009-03-31 doc/logparser.tex
Added a new action, DELIVERY_ERROR, to save data from delivery errors that     happen before a disconnection. Changed some UNINTERESTING rules to use DELIVERY_ERROR and SAVE_DATA. Removed connection_data fields from UNINTERESTING rules. Added a debug field to rules, so that individual rules can have debugging     information dumped when they recognise a log line. Improved the abstract, though it still needs work. Documented everything in the thesis.

@@
 \newcommand{\numberOFactions}[0]{%
-    22%
+    23%
 }
@@
     but it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding the log files and reconstructing the actions taken by
-    Postfix to generate the log files.  This paper describes the creation
-    of a parser for Postfix log files, documenting the architecture
-    developed and the parser implementing it, explaining the difficulties
-    encountered and the solutions developed, with reference to an
-    implementation that stores data gleaned from the log files in an SQL
-    database.  The gathered data can then be used to optimise current
-    anti-spam measures, provide a baseline to test new anti-spam measures
-    against, or to produce statistics showing how effective those measures
-    are.
-
-    XXX ABSTRACT FROM SGAI PAPER\@.
-
-    Some parsers need to be very precise and strict when parsing, yet must
-    allow users to easily adapt or extend the parser to parse new inputs,
-    without requiring that the user have an in-depth knowledge and
-    understanding of the parser's internal workings.  This paper presents a
-    novel parsing architecture, designed for parsing Postfix log files,
-    that aims to make the process of parsing new inputs as simple as
-    possible, enabling users to trivially add new rules (to parse variants
-    of existing inputs) and relatively easily add new actions (to process a
-    previously unknown category of input).  The architecture scales
-    linearly or better as the number of rules and size of input increases,
-    making it suitable for parsing large corpora or months of accumulated
-    data.
+    understanding Postfix log files and reconstructing the actions taken by
+    Postfix that generated the log files.  This thesis describes the
+    creation of a parser for Postfix log files, documenting the
+    architecture developed and the parser implementing it, explaining the
+    difficulties encountered and the solutions developed, with reference to
+    an implementation that stores data gleaned from the log files in an SQL
+    database.  The gathered data can be used to optimise current anti-spam
+    measures, to produce statistics showing how effective those measures
+    are, or to provide a baseline to test new anti-spam measures against.
+    The Postfix log file parser needs to be very precise and strict when
+    parsing, yet must allow users to easily adapt or extend it to parse new
+    inputs, without requiring that the user have an in-depth knowledge and
+    understanding of the parser's internal workings.  The newly developed
+    architecture aims to make the process of parsing new inputs as simple
+    as possible, enabling users to trivially add new rules (to parse
+    variants of existing inputs) and relatively easily add new actions (to
+    process a previously unknown category of inputs).  The architecture
+    scales linearly or better as the number of rules and size of input
+    increases, making it suitable for parsing large corpora or months of
+    accumulated data.
 
@@
 
-    XXX CHECK ALL ``UNINTERESTING'' RULES TO SEE IF THEY COULD BE USED\@.
+    XXX FIX THE PROBLEMS THAT AROSE FROM THE LAST all-logs RUN\@.
 
-    XXX FIX THE PROBLEMS THAT AROSE FROM THE LAST all-logs RUN\@.
+    XXX RUN ``test-some'' AGAIN TO UPDATE THE DB\@.
 

### 2009-04-02 doc/logparser-implementation.tex
Updated results after stats-max run finished. Added perfect best and perfect worst graph, and a table showing the std dev etc. Worked on graph and table placement, without much success.  I think manual     tweaking will be required, though adding 'h' back into the placement list     has helped. Added a graph for the number of times each action is invoked.

@@
 \setlength{\belowcaptionskip}{10pt}
-\showgraph{database-schema}{Diagram of the Database Schema}{Diagram
+\showgraph{build/database-schema}{Diagram of the Database Schema}{Diagram
 of the Database Schema picture}
@@
 rules is \action{DELIVERY\_REJECTED}, the action that processes Postfix
-rejecting a mail delivery attempt.  The next most common action is, perhaps
-surprisingly, \action{UNINTERESTING}: this action does nothing when
-invoked, allowing uninteresting log lines to be parsed without any effects
-(it does not imply that the input is ungrammatical or unparsed).  Generally
-rules specifying the \action{UNINTERESTING} action recognise log lines that
-are not associated with a specific mail, e.g.\ notices about configuration
-files changing; these log lines are recognised and processed so that the
-framework can warn about unrecognised log lines, informing the user that
-they need to add to or extend their ruleset.  Most of the remaining actions
-have only one or two associated rules: some are specialised actions
-required to address a deficiency in the log files, or a complication that
-arises during parsing; other actions will only ever have one log line
-variant, e.g.\ all log lines showing that a remote client has connected are
-recognised by a single rule and handled by the \action{CONNECT} action.
+rejecting a mail delivery attempt; next is \action{SAVE\_DATA}, the action
+that saves useful information without doing any other processing.  The
+third most common action is, perhaps surprisingly, \action{UNINTERESTING}:
+this action does nothing when invoked, allowing uninteresting log lines to
+be parsed without any effects (it does not imply that the input is
+ungrammatical or unparsed).  Generally rules specifying the
+\action{UNINTERESTING} action recognise log lines that are not associated
+with a specific mail, e.g.\ notices about configuration files changing;
+these log lines are recognised and processed so that the framework can warn
+about unrecognised log lines, informing the user that they need to add to
+or extend their ruleset.  Most of the remaining actions have only one or
+two associated rules: some are specialised actions required to address a
+deficiency in the log files, or a complication that arises during parsing;
+other actions will only ever have one log line variant, e.g.\ all log lines
+showing that a remote client has connected are recognised by a single rule
+and handled by the \action{CONNECT} action.
+
+XXX RUN test-some, STITCH THE COUNTS OF NUMBER OF RULES INVOKED INTO THE
+REAL TIMING DATA, UPDATE THE NUMBER OF INVOCATIONS GRAPH, AND WRITE ABOUT
+IT\@.
+
 \parsernames{} actions are described in detail in \sectionref{actions in
@@
 action}{Distribution of rules per action}
+
+\showgraph{build/graph-number-of-action-invocations}{Number of times each
+action was invoked when parsing \numberOFlogFILES{} log files}{Number of times
+each action was invoked}
 

### 2009-04-02 doc/logparser-results.tex
Updated results after stats-max run finished. Added perfect best and perfect worst graph, and a table showing the std dev etc. Worked on graph and table placement, without much success.  I think manual     tweaking will be required, though adding 'h' back into the placement list     has helped. Added a graph for the number of times each action is invoked.

@@
 
-\clearpage{} % XXX GRAPHS MOVE AROUND A LOT
 \subsection{Rule Ordering For Efficiency}
@@
 
-\clearpage{} % XXX GRAPHS MOVE AROUND DURING REBUILDS
-
 \showgraph{build/graph-hits}{Log lines recognised per rule}{rule hits graph}
@@
 RULE ORDERING FOR EFFICIENCY, OR IN ITS OWN SECTION\@.
+
+XXX ADD A TABLE SHOWING PERFECT WORST RELATIVE TO PERFECT BEST\@.
+
+\showgraph{build/graph-perfect-best-and-perfect-worst-and-optimal-and-reverse-vs-shuffled}{Parsing
+time including perfect best and worst relative to shuffled
+ordering}{Parsing time relative to shuffled ordering including perfect best
+and worst graph}
+
+\showtable{build/include-perfect-best-and-optimal-and-reverse-and-perfect-worst-vs-shuffle}{Parsing
+time including perfect best and worst relative to shuffled
+ordering}{Parsing time including perfect best and worst relative to
+shuffled ordering table}
+
+% XXX \FloatBarrier{} MIGHT BE USEFUL TO CONTROL WHERE FLOATS ARE PLACED.
 

### 2009-04-02 doc/logparser-implementation.tex
Added graph showing per-file action invocations. Wrote about action invocation numbers.

@@
 
-XXX RUN test-some, STITCH THE COUNTS OF NUMBER OF RULES INVOKED INTO THE
-REAL TIMING DATA, UPDATE THE NUMBER OF INVOCATIONS GRAPH, AND WRITE ABOUT
-IT\@.
-
-\parsernames{} actions are described in detail in \sectionref{actions in
-detail in implementation}, and \sectionref{adding new actions in
-implementation} covers the process of adding a new action.
+There is no correlation between how often an action is specified by rules
+and how often an action is invoked because a rule recognises a log line.
+\Graphref{Number of times each action was invoked} shows the number of
+times each action was invoked when parsing the \numberOFlogFILES{} log
+files used to generate the statistics in \sectionref{parser efficiency}.
+The action most commonly specified by rules, \action{DELIVERY\_REJECTED},
+is the sixth most commonly invoked, but the second and third most commonly
+specified actions are halfway down the list of most commonly invoked
+actions, with \action{SAVE\_DATA} being invoked for a mere 0.0903\% of
+recognised log lines.
+
+The action invoked most often is \action{MAIL\_QUEUED}; there is only a
+0.0051\% difference between it and \action{CLEANUP\_PROCESSING}, and a
+further 0.5203\% drop to \action{COMMIT}.  The near equality of these
+numbers is to be expected, because every mail delivered by Postfix will be
+managed in the queue by \daemon{qmgr} (\action{MAIL\_QUEUED}), processed by
+\daemon{cleanup} (\action{CLEANUP\_PROCESSING}), and removed from the queue
+by \daemon{qmgr} (\action{COMMIT}); the reason for the small disparity
+between \action{MAIL\_QUEUED} and \action{COMMIT} is mainly because mails
+are sometimes manually deleted from the queue.  Similarly \action{CONNECT}
+and \action{DISCONNECT} have a mere 0.0004\% difference between their
+number of invocations, because every connection will have a matching
+disconnection unless the \daemon{smtpd} process is killed for some reason.
+Unexpectedly there were far fewer connections (\action{CONNECT}) than there
+were mails in the queue (\action{MAIL\_QUEUED}); this is because of two
+local mail loops (\sectionref{Explaining the peaks in log file size}),
+which did not involve delivery via \acronym{SMTP}, and skew the total
+numbers.
 

### 2009-04-03 doc/logparser-background.tex
Added a table showing mean and stddev for the number of invocations of each action, for different groups of log files.

@@
     \begin{tabular}[]{ll}
-
+        \tabletopline{}%
+        Attribute name          & Attribute value           \\
+        \tablemiddleline{}%
         request                 & smtpd\_access\_policy     \\
@@
         instance                & 123.456.7                 \\
-
+        \tablebottomline{}%
     \end{tabular}

### 2009-04-03 doc/logparser-implementation.tex
Added a table showing mean and stddev for the number of invocations of each action, for different groups of log files.

@@
 
+XXX WRITE ABOUT THE TABLE\@.
+
 \showgraph{build/graph-action-distribution}{Distribution of rules per
@@
 each action was invoked}
+
+\newlength{\tabcolsepORIG}
+\setlength{\tabcolsepORIG}{\tabcolsep}
+\addtolength{\tabcolsep}{-1pt}
+
+\showtable{build/include-per-file-action-invocation-numbers-table}{Mean and
+standard deviation of the number of invocations of each action, for
+different groups of log files}{Mean and standard deviation of the number of
+invocations of each action, for different groups of log files}
+
+\setlength{\tabcolsep}{\tabcolsepORIG}
 
@@
     \begin{tabular}{ll}
-
+        \tabletopline{}%
         \textbf{Field}      & \textbf{Value}                                    \\
+        \tablemiddleline{}%
         name                & Unknown sender domain                             \\
@@
         %cluster\_group      & 400                                               \\
-
+        \tablebottomline{}%
     \end{tabular}

### 2009-04-04 doc/logparser-implementation.tex
Added graph showing number of action invocations, excluding log files 22 and 62-68. Removed table and old action invocations graph. Removed old discussion of number of action invocations, and wrote a new one referring to the new graph.

--- 2009-04-02 doc/logparser-implementation.tex
@@
 
-XXX RUN test-some, STITCH THE COUNTS OF NUMBER OF RULES INVOKED INTO THE
-REAL TIMING DATA, UPDATE THE NUMBER OF INVOCATIONS GRAPH, AND WRITE ABOUT
-IT\@.
-
-\parsernames{} actions are described in detail in \sectionref{actions in
-detail in implementation}, and \sectionref{adding new actions in
-implementation} covers the process of adding a new action.
-
-\showgraph{build/graph-action-distribution}{Distribution of rules per
+There is no correlation between how often an action is specified by rules
+and how often an action is invoked because a rule recognises a log line.
+\Graphref{Number of times each action was invoked excluding 22 and 62--68}
+shows the number of times each action was invoked when parsing the
+\numberOFlogFILES{} log files used to generate the statistics in
+\sectionref{parser efficiency}, \textit{excluding\/} log files 22 and
+62--68, because their content is extremely skewed by two mail loops (see
+\sectionref{Explaining the peaks in log file size}).  The action most
+commonly specified by rules, \action{DELIVERY\_REJECTED}, is the third most
+commonly invoked action; the most commonly invoked actions,
+\action{CONNECT} and \action{DISCONNECT}, are each specified by only one
+rule.  The \action{UNINTERESTING} action, the third most commonly specified
+action, is halfway down the graph.
+
+As expected, some actions have been invoked almost exactly the same number
+of times: almost every \action{CONNECT} will have a \action{DISCONNECT},
+and there is only a 0.00042\% difference between the number of times the
+two were invoked; every mail that enters Postfix's queue will be managed by
+\daemon{qmgr} (\action{MAIL\_QUEUED}) and processed by \daemon{cleanup}
+(\action{CLEANUP\_PROCESSING}), and again there is only a 0.01648\%
+difference between their invocations.  Confusingly,
+\action{CLEANUP\_PROCESSING} is invoked more often than
+\action{MAIL\_QUEUED}: occasionally, accepted mail in the process of being
+transferred is interrupted by a timeout, and there is a log line from
+\daemon{cleanup} but not from \daemon{qmgr}, as described in
+\sectionref{discarding cleanup log lines}.
+
+\showgraph{build/graph-action-distribution}{Number of rules specifying each
 action}{Distribution of rules per action}
 
-\showgraph{build/graph-number-of-action-invocations}{Number of times each
-action was invoked when parsing \numberOFlogFILES{} log files}{Number of times
-each action was invoked}
+\showgraph{build/graph-number-of-action-invocations-excluding-22-and-62--68}{Number
+of times each action was invoked when parsing \numberOFlogFILES{} log
+files, excluding log files 22 and 62--68}{Number of times each action was
+invoked excluding 22 and 62--68}
 

### 2009-04-04 doc/logparser-implementation.tex
Automatically generate the list of graphs, tables, and other included files.
Added graph showing average number of rules tried per program. Wrote a brief comparison between inserting and not inserting results. Removed the explanation of why mail loops occur, it's not important. Improved the blurb about the machine used for generating statistics.

@@
 \sectionref{parser efficiency}, \textit{excluding\/} log files 22 and
-62--68, because their content is extremely skewed by two mail loops (see
-\sectionref{Explaining the peaks in log file size}).  The action most
-commonly specified by rules, \action{DELIVERY\_REJECTED}, is the third most
-commonly invoked action; the most commonly invoked actions,
+62--68, because their content is extremely skewed by two mail loops.  The
+action most commonly specified by rules, \action{DELIVERY\_REJECTED}, is
+the third most commonly invoked action; the most commonly invoked actions,
 \action{CONNECT} and \action{DISCONNECT}, are each specified by only one
@@
 \begin{table}[ht]
-
     \caption{Example of a rule used by \parsernameshort{}}
@@
     \end{tabular}
-
 \end{table}

### 2009-04-04 doc/logparser-results.tex
Added graph showing average number of rules tried per program. Wrote a brief comparison between inserting and not inserting results. Removed the explanation of why mail loops occur, it's not important. Improved the blurb about the machine used for generating statistics.

@@
 
-XXX DO A TEST-SOME RUN AND SAVE THE RESULTS TO THE DATABASE, THEN COMPARE
-THE TIME WITH A RUN THAT DOES NOT SAVE RESULTS\@; PUT THE COMPARISON IN THE
-INTRODUCTION, NEAR THE MACHINE SPECS\@.
-
-XXX THINK ABOUT THE ORDERING OF TABLES AND GRAPHS\@; MAYBE REARRANGE THEM
-SO THAT THEY APPEAR IN THE ORDER THEY ARE REFERENCED\@?
-
-XXX SHOULD I USE $\backslash{}$clearpage{} OR SOMETHING TO AFFECT
+XXX SHOULD I USE $\backslash{}$clearpage{} OR SOMETHING TO AFFECT FLOAT
 PLACEMENT\@?  GRAPHS AND TABLES TEND TO MOVE QUITE FAR IN SOME OF THE
 SUBSECTIONS, THOUGH MAKING THE TEXT LONGER MIGHT HELP TOO\@.  IDEALLY I
-WANT TO HAVE TEXT, THEN ALL GRAPHS AND TABLES, THEN THE NEXT SECTION\@; I
-DO NOT LIKE THAT GRAPHS/TABLES SOMETIMES APPEAR BEFORE THEY ARE
-REFERENCED\@.  I MIGHT NEED TO REDUCE THE HEIGHT OF THE GRAPHS TOO\@.
+WANT TO HAVE TEXT, THEN ALL GRAPHS AND TABLES, THEN THE NEXT SECTION\@.
+
+\newpage{} % XXX VARIOREF PROBLEMS
 
@@
 testing this parser accepts approximately 10,000 mails for 700 users a day;
-\graphref{Mails received per day} shows that, as expected, far more mails
-are received on weekdays than at weekends.  The mail server in question is
-a production mail server handling mail for a university department; the
-benefit of using this server is that its log files exhibit the
-idiosyncrasies and peculiarities a mail server in the wild must deal with,
-but the downside is that significantly altering the configuration to
-accommodate this project is not an option.  Median log file size is 50MB,
-containing 285,000 log lines; large scale mail servers would have much
-larger log files.  Note that \graphref{Mails received per day} and
-\tableref{Number of mails received per day: statistics} show the number of
-mails received by \acronym{SMTP} only; in particular the mail loops
-noticeable in later graphs do not contribute to these figures.
+\graphref{Mails received per day} and \tableref{Number of mails received
+per day: statistics} show that, as expected, far more mails are received on
+weekdays than at weekends.  The mail server in question is a production
+mail server handling mail for a university department; the benefit of using
+this server is that its log files exhibit the idiosyncrasies and
+peculiarities a mail server in the wild must deal with, but the downside is
+that significantly altering the configuration to accommodate this project
+is not an option.  Median log file size is 50MB, containing 285,000 log
+lines; large scale mail servers would have much larger log files.  Note
+that \graphref{Mails received per day} and \tableref{Number of mails
+received per day: statistics} show the number of mails received by
+\acronym{SMTP} only; in particular the mail loops noticeable in later
+graphs do not contribute to these figures.
 
@@
 SMTP per day}{Number of mails received per day: statistics}
-
-When generating the timing data used in this section, \numberOFlogFILES{}
-log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
-parsed 10 times, and the parsing times from the 10 runs averaged.  Saving
-results to the database was disabled for the test runs, because that
-dominates the run time of the program, and the tests are aimed at measuring
-the speed of \parsername{} rather than the speed of the database and the
-disks the database is stored on.  The computer used for test runs was a
-Dell Optiplex 745, shown in \tableref{Details of the computer used to
-generate statistics}.
 
@@
 
-Parsing all \numberOFlogFILES{} log files in one run took
-\input{build/include-timing-run-duration}.  The computer was dedicated to
-the task of gathering statistics from test runs, and was not used for any
-other purpose while test runs were ongoing; any services not necessary for
-running the tests were disabled.
+When generating the timing data used in this section, \numberOFlogFILES{}
+log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
+parsed 10 times, and the times from the 10 runs averaged.  The computer
+used for test runs was a Dell Optiplex 745, shown in \tableref{Details of
+the computer used to generate statistics}; it was dedicated to the task of
+gathering statistics from test runs, and was not used for any other purpose
+while test runs were ongoing.  Saving results to the database was disabled
+for the test runs, because that dominates the run time of the parser, and
+the tests are aimed at measuring the speed of \parsername{} rather than the
+speed of the database and the disks the database is stored on.  Parsing all
+\numberOFlogFILES{} log files in one run without saving results to the
+database took \input{build/include-full-run-duration}, mean throughput is
+\input{build/include-full-run-throughput}; median throughput when parsing
+files separately was \input{build/include-median-throughput-MB}
+(\input{build/include-median-throughput-log-lines}) parsed per minute.  In
+contrast, when saving results to the database, parsing all
+\numberOFlogFILES{} log files took
+\input{build/include-insert-results-duration}, with mean throughput of
+\input{build/include-insert-results-throughput}; parsing makes up
+\input{build/include-skip-inserting-as-percentage-of-inserting} of the
+execution time when saving results to the database, so a 50\% improvement
+in the speed of parsing would result in only a
+\input{build/include-skip-inserting-with-50-percent-improvement--as-percentage-of-inserting}
+improvement in speed overall.
 
@@
 shows the same ratios for different groups of log files.  The ratios are
-quite tightly banded, showing that the algorithm scales linearly.  The
-ratios increase (i.e.\ improve) for log files 22 and 62--68 despite their
-larger than average size (shown in \graphref{parsing time vs file size vs
-number of log lines graph}); this is explained fully in
-\sectionref{Explaining the peaks in log file size}.
+quite tightly banded, showing that the algorithm scales linearly; the
+ratios increase (i.e.\ improve) for log files 22 and 62--68, despite their
+larger than average size.  In both groups of log files there was a mail
+loop because a user set up mail forwarding incorrectly, resulting in a very
+different distribution of log lines: normally most log lines are generated
+by mail delivery attempts from other hosts, but when the mail loops
+occurred most of the log lines resulted from failed delivery of mail
+generated on the server itself.  The Postfix components that generated most
+of the log lines during the mail loop have fewer associated rules than the
+Postfix components whose log lines normally make up the bulk of each log
+file; \graphref{Average number of rules tried per log line} shows the drop
+in the average number of rules tried per log line for log files containing
+a mail loop.
 
@@
 
-\subsection{Explaining The Peaks In Log File Size}
+\showgraph{build/graph-average-number-of-rules-tried-per-log-line}{Average
+number of rules tried per log line}{Average number of rules tried per log
+line}
 
-\label{Explaining the peaks in log file size}
+\showgraph{build/graph-average-number-of-rules-tried-per-program}{Average
+number of rules tried per Postfix component per log line}{Average number of
+rules tried per Postfix component per log line}
 
-XXX SHOULD THIS BE A SEPARATE SECTION OR INCLUDED IN THE PREVIOUS ONE\@?
-MAYBE INCLUDE GRAPHS OF RULES PER FILE OR SOMETHING TO EXTEND THIS\@?
-
-XXX REWRITE WHEN STATS RUNS HAVE FINISHED\@; REWRITE ALL THE REFERENCES
-TOO\@.
-
-\Graphref{parsing time vs file size vs number of log lines graph} shows log
-file size varying between 30--100MB, except for log files 22 (200MB) and
-62--68 (400--900MB).  These large log files were caused by a user
-unintentionally creating a mail forwarding loop, resulting in a huge number
-of mail deliveries and up to an order of magnitude more log lines than is
-usual.  The user specified his mail forwarding as: \newline{}
-\tab{}\texttt{$\backslash$username, username@domain} \newline{} This
-instructs Postfix to deliver the mail to the local user, and also forward a
-copy to the remote address; this is generally not a problem except that the
-remote address in this instance is the address the mail was originally sent
-to, creating an infinite loop.  To prevent this happening Postfix examines
-the Delivered-To header in the mail, and if the mail has already been
-delivered to the current address it is bounced back to the sender with the
-error message \newline{} \tab{} \texttt{mail forwarding loop for
-username@domain}\newline{}  Normally this works well, but the forwarding
-method in use at that time stripped the Delivered-To header from the mail,
-breaking loop detection.
-
-When the recipient in a mail loop is local, Postfix does not have to
-deliver the mail to an external system, so mails are delivered as fast as
-the disks can keep up, resulting in a huge volume of log lines.
-
-XXX EXPAND THIS PARAGRAPH\@; THERE MAY BE CONTENT IN THE PAPER\@.  REFER TO
-THE NEW GRAPH BELOW TOO\@.
-
-The different distribution of log lines caused by a mail forwarding loop
-has unexpected effects on parsing time, described in detail in
-\sectionref{rule ordering for efficiency}, \sectionref{scalability as the
-number of rules rises}, and \sectionref{Caching compiled regexes}.
-
-\showgraph{build/graph-rules-tried-per-log-line}{Mean number of rules tried
-per log line}{Mean number of rules tried per log line}
+\FloatBarrier{}
 
@@
 
+\newpage{} % XXX GRAPH PLACEMENT FLIP-FLOPPING
+
 \Graphref{Parsing time relative to shuffled ordering graph} shows the
@@
 log files 22 and 62--68, due to the different distribution of log lines in
-those log files (see \sectionref{Explaining the peaks in log file size}).
+those log files.
 When a mail loop occurs the vast majority of log lines are from Postfix

### 2009-04-04 doc/logparser-results.tex
Further improvement to the explanation of better parsing performance for mail loops.

@@
 
-% XXX IS CHANGING FIGURE TO GRAPH A GOOD IDEA?
 \renewcommand{\figurename}{Graph}
@@
 ratios increase (i.e.\ improve) for log files 22 and 62--68, despite their
-larger than average size.  In both groups of log files there was a mail
-loop because a user set up mail forwarding incorrectly, resulting in a very
-different distribution of log lines: normally most log lines are generated
-by mail delivery attempts from other hosts, but when the mail loops
-occurred most of the log lines resulted from failed delivery of mail
-generated on the server itself.  The Postfix components that generated most
-of the log lines during the mail loop have fewer associated rules than the
-Postfix components whose log lines normally make up the bulk of each log
-file; \graphref{Average number of rules tried per log line} shows the drop
-in the average number of rules tried per log line for log files containing
-a mail loop.
+larger than average size.  Both groups of log files are much larger than
+usual due to a mail loop caused by a user who set up mail forwarding
+incorrectly, resulting in a very different distribution of log lines:
+normally most log lines are generated by mail delivery attempts from other
+hosts, but when the mail loops occurred most of the log lines resulted from
+failed delivery of mail generated on the server itself.  The Postfix
+components that generated most of the log lines during the mail loop have
+fewer associated rules than the Postfix components whose log lines normally
+make up the bulk of each log file, so the average number of rules tried per
+log line reduces and so does the average parsing time per log line.
+\Graphref{Average number of rules tried per log line} shows the drop in the
+average number of rules tried per log line for log files containing a mail
+loop, and \graphref{Average number of rules tried per log line for each
+Postfix component} shows the average number of rules tried per log line for
+each Postfix component.
 
@@
 \showgraph{build/graph-average-number-of-rules-tried-per-program}{Average
-number of rules tried per Postfix component per log line}{Average number of
-rules tried per Postfix component per log line}
+number of rules tried per log line for each Postfix component}{Average
+number of rules tried per log line for each Postfix component}
 

### 2009-04-04 doc/logparser-implementation.tex
Corrections/improvements based on Carl's comments. Improved the flow chart, rewriting labels and making it easier to read.

@@
 difficulties come to light.  Implementing this architecture was relatively
-straightforward, with only minor changes required: rule conditions and
-cascaded parsing were added, although later the implementation stopped
-using cascaded parsing.  The difficulties came not from the architecture
-and design, but from anomalies in the log files and Postfix's behaviour, as
-described in \sectionref{complications}.
+straightforward: the difficulties came not from the architecture and
+design, but from anomalies in the log files and Postfix's behaviour, as
+described in \sectionref{complications}.  \parsername{} successfully deals
+with all the difficulties that were discovered during its development, but
+further difficulties may arise during future usage; unfortunately solving
+all potential difficulties that may arise is an intractable task, but the
+descriptions of the solutions developed thus far (see
+\sectionref{complications}) should help if someone needs to develop a
+solution of their own.
 
@@
 captured by rules into a coherent whole, saving it in a useful and
-consistent form, and performing housekeeping duties.
+consistent form, and performing housekeeping duties.  The intermingling of
+log entries from different mails immediately rules out the possibility of
+handling each mail in isolation; the parser must be able to handle multiple
+mails in parallel, each potentially at a different stage in its journey,
+without any interference between mails --- except in the minority of cases
+where intra-mail interference is required, e.g.\ mail re-injected for
+forwarding (\sectionref{Re-injected mails}).  The best way to deal with
+intermingling is to maintain state information for every unfinished mail
+and manipulate the appropriate mail correctly for each log line
+encountered.
 
@@
 the required fields.  Rules are the final component of the architecture,
-and also the most visible to users, because it is likely that users will
-need to add their own rules to recognise their log lines.  An example rule
-used by \parsername{} is shown, with every field clearly documented; this
-is followed by a description of adding new rules, and how to determine the
-value of each of the fields.  \parsername{} provides a utility to create
-regexes from unrecognised log lines; the algorithm it uses is documented,
-including the differences between it and the original algorithm.  The rules
-section finishes with a discussion of how \parsername{} uses rule
-conditions and overlapping rules, plus a description of the regex snippets
-provided to ease the process of writing regexes.
+and also the most visible to advanced users, e.g.\ systems administrators,
+because it is likely that they will need to add their own rules to
+recognise their log lines.  An example rule used by \parsername{} is shown,
+with every field clearly documented; this is followed by a description of
+adding new rules, and how to determine the value of each of the fields.
+\parsername{} provides a utility to create regexes from unrecognised log
+lines; the algorithm it uses is documented, including the differences
+between it and the original algorithm.  The rules section finishes with a
+discussion of how \parsername{} uses rule conditions and overlapping rules,
+plus a description of the regex snippets provided to ease the process of
+writing regexes.
 
@@
         course possible, particularly when suffering from a deluge of spam
-        or a mail loop.
+        or a mail loop.  When parsing individual log files in isolation,
+        it is highly likely that some mails will have log lines in previous
+        log files, and others will have log lines in subsequent log files;
+        to alleviate this problem \parsername{} supports saving and loading
+        its state tables, so they can be preserved between log files.
 
@@
 
-A picture is said to be worth a thousand words: the flow chart in
-\figureref{flow chart image} shows the most common paths a connection or
-mail can take through \parsername{}; the difficulties described in
-\sectionref{complications} are excluded for the sake of clarity.  The flow
-chart is intended to be a graphical overview of how a mail progresses
-through both Postfix and \parsername{}, providing an overall context into
-which the detailed descriptions in the remainder of this chapter will fit,
-in particular the actions (\sectionref{actions in implementation}) and
-complications (\sectionref{complications}).
+The flow chart in \figureref{flow chart image} shows the most common paths
+a connection or mail can take through \parsername{}; the difficulties
+described in \sectionref{complications} are excluded for the sake of
+clarity.  The flow chart is intended to be a graphical overview of how a
+mail progresses through both Postfix and \parsername{}, providing an
+overall context into which the detailed descriptions in the remainder of
+this chapter will fit, in particular the actions (\sectionref{actions in
+implementation}) and complications (\sectionref{complications}).  The
+states and transitions shown in \figureref{flow chart image} will be
+explained later in this section.
 
@@
 chart}{flow chart image}
-
-The intermingling of log entries from different mails immediately rules out
-the possibility of handling each mail in isolation; the parser must be able
-to handle multiple mails in parallel, each potentially at a different stage
-in its journey, without any interference between mails --- except in the
-minority of cases where intra-mail interference is required.  The best way
-to implement this is to maintain state information for every unfinished
-mail and manipulate the appropriate mail correctly for each log line
-encountered.
 
@@
 the mail moves on to the delivery stage.  Re-injection due to forwarding
-sadly lacks explicit log lines of its own; it is explained fully in
+lacks explicit log lines of its own; it is explained fully in
 \sectionref{Re-injected mails}.  Internally generated mails lack any
@@
 described in \sectionref{identifying bounce notifications}; later versions
-of Postfix do provide log lines for internally generated mails.  Bounce
-notifications are the primary example of internally generated mails, though
-there are other types, e.g.\ Postfix may generate mails to the
-administrator when it encounters configuration errors, but such mails are
-presumably rare.
+of Postfix do provide log lines for internally generated mails
+\flowchart{BOUNCE\_CREATED}{3}.  Bounce notifications are the primary
+example of internally generated mails, though there are other types, e.g.\
+Postfix may generate mails to the administrator when it encounters
+configuration errors.
 
@@
         \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
-        \flowchart{CLONE}{5}; or some interleaving of both.
+        \flowchart{CLONE}{5}; or some random interleaving of both.
 
@@
         be any data to save.  Any mails accepted will already have been
-        cloned so their data is in another data structure, nd will be
+        cloned so their data is in another data structure, and will be
         delivered in the same way as mails that entered the system by any
@@
     \item Sometimes mail is re-injected and the child mail needs to be
-        tracked by the parent mail \flowchart{TRACK}{9}; the handling of
+        tracked with the parent mail \flowchart{TRACK}{9}; the handling of
         re-injected mails is described in \sectionref{tracking re-injected
@@
 
-            \item Some databases support granting access on a fine-grained
-                basis, e.g.\ allowing the finance department to produce
-                invoices, the helpdesk to run limited queries as part of
-                dealing with support calls, and the administrators to have
-                full access to the data.
-
             \item Triggers can be written to perform actions when certain
@@
 
+    \item Some databases support granting access on a fine-grained basis,
+        e.g.\ allowing the finance department to produce invoices, the
+        helpdesk to run limited queries as part of dealing with support
+        calls, and the administrators to have full access to the data.
+
     \item \acronym{SQL} is reasonably standard and many people will already
@@
         Although every vendor implements a different dialect of
-        \acronym{SQL}, the basics are the same everywhere, analogous to the
-        overall similarity and minor differences amongst Irish English,
-        British English, American English, and Australian English.
-        Depending on the database in use there may be tools available that
-        reduce or remove the requirement to know \acronym{SQL}.
+        \acronym{SQL}, the basics are the same everywhere.  Depending on
+        the database in use there may be tools available that reduce or
+        remove the requirement to know \acronym{SQL}.
 
@@
 
-    \item [warning] Postfix can be configured to log a warning instead of
-        enforcing a restriction that would reject an \acronym{SMTP} command
-        --- a mechanism that is quite useful for testing new restrictions.
-        This field will be false for a real rejection, or true if the log
-        line was a warning.  This field should be ignored if the result is
-        not a rejection, i.e.\ the action field of the associated rule is
-        not \action{DELIVERY\_REJECTED}.
+    \item [warning] Administrators can configure Postfix to log a warning
+        instead of enforcing a restriction that would reject an
+        \acronym{SMTP} command --- a mechanism that is quite useful for
+        testing new restrictions.  This field will be false for a real
+        rejection, or true if the log line was a warning.  This field
+        should be ignored if the result is not a rejection, i.e.\ the
+        action field of the associated rule is not
+        \action{DELIVERY\_REJECTED}.
 
@@
 ordering for efficiency} for details.  The repetitive nature of log files
-gives them high compression ratios; the framework supports reading
-compressed log files, simplifying the process of parsing them because they
-do not have to be uncompressed to a temporary file before parsing begins.
-The framework also displays a progress bar to show how far parsing has
-progressed through the log file, and to give the user an indication of how
-long the remainder of the parsing process is likely to take.  The progress
-bar is not as reliable when parsing compressed log files, because a
-compressed block may uncompress to a variable number of log lines; there is
-also variation between the recognition and processing time of individual
-log lines, but overall the progress bar is a useful addition.
+gives them high compression ratios; the framework can read compressed log
+files, simplifying the process of parsing them because they do not have to
+be uncompressed to a temporary file before parsing begins.  When used
+interactively, the framework also displays a progress bar to show how far
+parsing has progressed through the log file and how long the remainder of
+the parsing process is likely to take.  The progress bar is not as reliable
+when parsing compressed log files, because a compressed block may
+uncompress to a variable number of log lines; there is also variation
+between the recognition and processing time of individual log lines, but
+overall the progress bar is a useful addition.
 
@@
 or extend their ruleset.  Most of the remaining actions have only one or
-two associated rules: some are specialised actions required to address a
-deficiency in the log files, or a complication that arises during parsing;
-other actions will only ever have one log line variant, e.g.\ all log lines
-showing that a remote client has connected are recognised by a single rule
-and handled by the \action{CONNECT} action.
+two associated rules, because they will only ever have one or two log line
+variants, e.g.\ all log lines showing that a remote client has connected
+are recognised by a single rule and handled by the \action{CONNECT} action.
 
@@
         queueid; otherwise save it to the connection identified by the
-        \acronym{pid} in the log line.  No further processing of the log
-        line is required.
+        \acronym{pid} in the log line.
 
@@
 
-    \item [MAIL\_DISCARDED] Sometimes mail is discarded, e.g.\ mail
-        submitted locally that is larger than the limit configured by the
-        administrator.  This action is used for those mails; it invokes the
-        \action{COMMIT} action, but is a separate action to simplify
+    \item [MAIL\_DISCARDED] Sometimes mail is discarded by Postfix, e.g.\
+        mail submitted locally that is larger than the limit configured by
+        the administrator.  This action is used for those mails; it invokes
+        the \action{COMMIT} action, but is a separate action to simplify
         further analysis.
@@
         not have enough data to satisfy the database schema, so it cannot
-        be entered into the database; unfortunately this means that some
-        data is lost.
+        be entered into the database for future analysis; unfortunately
+        this means that the small amount of data that has been gathered
+        will be lost.
 
@@
 taken in the architecture and implementation to make adding new actions as
-painless as possible.  The implementer writes a subroutine that accepts the
-standard arguments given to actions, and registers it with the framework.
-The new action must be registered before the rules are loaded, because it
-is an error for a rule to specify an unregistered action; this helps catch
-mistakes made when adding new rules.  No other work is required from the
-implementer to integrate the action into the parser; all of their attention
-and effort can be focused on correctly implementing their new action.  The
-action may need to extend the list of valid combinations described in
-\sectionref{out of order log lines} if the new action creates a different
-set of acceptable programs, but this would only be necessary if the new
-action processes log lines from Postfix components \parsername{} currently
-does not have rules for, e.g.\ \daemon{virtual} or \daemon{lmtp}.
+painless as possible; one of the few limitations is that \parsername{} is
+written in Perl, so new actions must also be written in Perl.  The
+implementer writes a subroutine that accepts the standard arguments given
+to actions, and registers it with the framework by calling the framework's
+\texttt{add\_actions()} subroutine.  The new action must be registered
+before the rules are loaded, because it is an error for a rule to specify
+an unregistered action; this helps catch mistakes made when adding new
+rules.  No other work is required from the implementer to integrate the
+action into the parser; all of their attention and effort can be focused on
+correctly implementing their new action.  The action may need to extend the
+list of valid combinations described in \sectionref{out of order log lines}
+if the new action creates a different set of acceptable programs, but this
+would only be necessary if the new action processes log lines from Postfix
+components \parsername{} currently does not have rules for, e.g.\
+\daemon{virtual} or \daemon{lmtp}.
 
@@
 results tables that is missing some required fields, because values for the
-missing fields are not present in any of the log lines.  E.g.\ the rule
-recognising the \daemon{pickup} component processing a mail sets the
+missing fields are not present in any of the log lines.  For example, the
+rule recognising the \daemon{pickup} component processing a mail sets the
 client\_hostname to \texttt{localhost} and the client\_ip to
@@
 will be used if none of the program-specific rules recognise the log line.
-If none of the rules are successful the framework will warn the user.
+If none of the rules are successful the framework will warn the user,
+informing them that they need to augment their ruleset, and alerting them
+that because \parsername{} failed to recognise some of their log lines, the
+results stored in the database may be incomplete.
 
@@
         delivery} involves discarding data.
+
+    \item XXX FURTHER COMPLICATIONS MAY ARISE\@.
+
+    \item XXX DATABASE GROWS WITHOUT BOUNDS\@.
 

### 2009-04-06 doc/logparser-complications.tex
Changes from proofreading of introduction chapter.
Added a new action, DELIVERY_ERROR, to save data from delivery errors that     happen before a disconnection. Changed some UNINTERESTING rules to use DELIVERY_ERROR and SAVE_DATA. Removed connection_data fields from UNINTERESTING rules. Added a debug field to rules, so that individual rules can have debugging     information dumped when they recognise a log line. Improved the abstract, though it still needs work. Documented everything in the thesis.
Added a table showing mean and stddev for the number of invocations of each action, for differe

@@
-\section{Complications Encountered While Writing \parsernamelong{}}
+\section{Complications Encountered While Writing The \parsernamelong{}}
 
@@
 
+The fifteen complications documented in this section can be divided into
+three broad classes:
+
+\begin{description}
+
+    \item [Recreating Postfix behaviour] Three complications are caused by
+        \parsername{} accurately recreating Postfix's behaviour.
+
+    \item [Lack of information in the log files] Seven complications are
+        caused by lack of information in the log files.  In general some
+        extra logging by Postfix could remove these complications, and in
+        one case extra logging in later versions of Postfix does remove the
+        complication (\sectionref{identifying bounce notifications}); three
+        of the complications would be removed if Postfix generated a log
+        line whenever it discarded a mail.
+
+    \item [Order log lines are found in log files] Five complications are
+        caused by out of order log lines.  Postfix is not to blame for
+        these, they are caused by process scheduling, inter-process
+        communication delays, and are more likely to occur when the mail
+        server is heavily loaded.
+
+\end{description}
+
 \subsection{Queueid Vs Pid}
@@
 
-    \item Most importantly: every section of code must know which state
-        table it should query.
+    \item Most importantly: each action needs to know whether it deals with
+        accepted mails, and should find the mail in the by\_queueid state
+        tables, or deals with delivery attempts, and thus will find the
+        connection in the by\_pid state tables.  Some actions, e.g.\
+        \action{DELIVERY\_REJECTED}, need to query both tables.  Support
+        functions fall into two groups: those that deal with state tables,
+        and so need to know which one to operate on; the other group are
+        passed a specific mail or connection to work with, and do not need
+        to know about state tables at all.
 
@@
 
-Once again Postfix does not log a message making the client's behaviour
-clear, so once again heuristics are required to identify when this
-behaviour occurs.  A list of all mails accepted during a connection is
+Postfix does not log a message making the client's behaviour clear, so
+again heuristics are required to identify when a delivery attempt is
+aborted in this way.  A list of all mails accepted during a connection is
 saved in the connection's state table entry, and the accepted mails are
@@
 disconnections are either unimportant or are handled elsewhere in the
-parser, e.g.\ \sectionref{timeouts during data phase}.  Sadly, Postfix does
-not log a message saying the mail has been discarded, as should be expected
-by now.  The checks to identify this happening are:
+parser, e.g.\ \sectionref{timeouts during data phase}.  Postfix does not
+log a message saying the mail has been discarded, as should be expected by
+now.  The checks to identify this happening are:
 
@@
 \daemon{cleanup} log line for a mail being accepted is logged after the
-timeout log line, by which time \parsername{} has discarded the mail;
-parsing the \daemon{cleanup} log line causes the
-\action{CLEANUP\_PROCESSING} action to create a new state table entry, to
-help deal with re-injected mails (\sectionref{Re-injected mails}).  This is
-incorrect because the log line actually belongs to the mail that has just
-been discarded; if the queueid is reused there will be a queueid clash,
-otherwise the new mail will just remain in the state tables.
+timeout log line, by which time \parsername{} has discarded the mail
+because it will not have enough data to satisfy the database schema (see
+\sectionref{timeouts during data phase}); parsing the \daemon{cleanup} log
+line causes the \action{CLEANUP\_PROCESSING} action to create a new state
+table entry, to help deal with re-injected mails (\sectionref{Re-injected
+mails}).  This is incorrect because the log line actually belongs to the
+mail that has just been discarded; if the queueid is reused there will be a
+queueid clash, otherwise the new mail will just remain in the state tables.
 
@@
 
-\subsection{Out of Order Log Lines}
+% \newpage{} % There's always a fucking varioref problem here
+
+\subsection{Out Of Order Log Lines}
 
@@
 
-\subsection{Mail Deleted Before Delivery is Attempted}
+\subsection{Mail Deleted Before Delivery Is Attempted}
 
@@
 database and add the log line to it.
+
+\subsection{Summary}
+
+This section has described the complications encountered while implementing
+\parsername{}.  Five of the fifteen complications are caused by log lines
+appearing in an unexpected or abnormal order in log files, often because of
+heavy system load: these complications typically caused warnings and a new
+state table entry.  Three of the fifteen complications are required to
+correctly recreate Postfix's behaviour.  The remaining seven complications
+are caused by deficiencies in Postfix's logging, and some could easily be
+resolved by adding additional logging to Postfix; these seven are
+dealt with by applying heuristics to specific log lines and/or mails, in
+some cases with the aid of information cached when mails are removed from
+the state tables.

### 2009-04-06 doc/logparser-implementation.tex
Corrections/improvements based on Carl's comments.

@@
         significant investment of time: Postfix 2.5.5 has 17MB of source
-        code.
+        code.  Each subsequent version would require further work to
+        investigate the changes; many of those changes, although they
+        improve Postfix's internals, would not have any effect on its
+        externally observable behaviour.
 
@@
         \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
-        \flowchart{CLONE}{5}; or some random interleaving of both.
-
-    \item The client disconnects \flowchart{DISCONNECT}{6}.  If Postfix has
-        rejected any \acronym{SMTP} commands the resulting data will be
-        saved to the database; if there were no rejections there will not
-        be any data to save.  Any mails accepted will already have been
-        cloned so their data is in another data structure, and will be
-        delivered in the same way as mails that entered the system by any
-        other route.
+        \flowchart{CLONE}{5}; failure of the remote client's connection
+        \flowchart{DELIVERY\_ERROR}{6}; or some random interleaving of two
+        or more of the above.
+
+    \item The client disconnects \flowchart{DISCONNECT, TIMEOUT}{6}, either
+        normally or with an error.  If Postfix has rejected any
+        \acronym{SMTP} commands the resulting data will be saved to the
+        database; if there were no rejections there will not be any data to
+        save.  Any mails accepted will already have been cloned so their
+        data is in another data structure, and will be delivered in the
+        same way as mails that entered the system by any other route.
 
@@
 
-    \item Postfix will log the sender and recipient addresses separately
-        \flowchart{SAVE\_DATA}{8}.
+    \item The mail will have one or more delivery attempts
+        \flowchart{MAIL\_SENT, SAVE\_DATA}{8}.
 
@@
 
-    \item Eventually the mail will be delivered, bounced, or deleted by the
-        administrator \flowchart{COMMIT}{11}.  This is the last log line
-        for this particular mail, though it may be indirectly referred to
-        if it was re-injected.  The mail is cleaned up and entered in the
-        database, then deleted from the state tables.
+    \item After one or more delivery attempts the mail will be delivered
+        \flowchart{MAIL\_SENT}{8}, bounced \flowchart{MAIL\_BOUNCED}{8},
+        expired \flowchart{EXPIRY}{8}, or deleted by the administrator
+        \flowchart{DELETE}{8}.
+
+    \item This is the last log line for this particular mail, though it may
+        be indirectly referred to if it was re-injected.  The mail is
+        cleaned up and entered in the database, then deleted from the state
+        tables \flowchart{COMMIT}{11}.
 
@@
     \item [restriction\_name] The name of the Postfix restriction that
-        caused the mail to be rejected.  Only applicable to rules that
-        recognise rejection log lines, i.e.\ rules that have an action of
-        \action{DELIVERY\_REJECTED}; other rules should not set this field.
+        caused the mail to be rejected.  This field is meaningful only for
+        rules that recognise rejection log lines, i.e.\ rules that have an
+        action of \action{DELIVERY\_REJECTED}.
 
@@
         conveyed to the user.  Enhanced status code support was added to
-        Postfix in version 2.3; log files from previous versions will not
+        Postfix in version 2.3; log lines from previous versions will not
         contain any enhanced status codes.
@@
 those rules whose \texttt{program} field equals the program in the log line
-(see \sectionref{rule conditions in implementation}), falling back to
-generic rules if necessary, and warning if the log line is unrecognised.
-The rules are sorted for increased efficiency; see \sectionref{rule
-ordering for efficiency} for details.  The repetitive nature of log files
-gives them high compression ratios; the framework can read compressed log
-files, simplifying the process of parsing them because they do not have to
-be uncompressed to a temporary file before parsing begins.  When used
-interactively, the framework also displays a progress bar to show how far
-parsing has progressed through the log file and how long the remainder of
-the parsing process is likely to take.  The progress bar is not as reliable
-when parsing compressed log files, because a compressed block may
-uncompress to a variable number of log lines; there is also variation
-between the recognition and processing time of individual log lines, but
-overall the progress bar is a useful addition.
+(previously described in \sectionref{rule conditions in implementation}),
+falling back to generic rules if necessary, and warning if the log line is
+unrecognised.  The rules are sorted for increased efficiency; see
+\sectionref{rule ordering for efficiency} for details.  The repetitive
+nature of log files gives them high compression ratios; the framework uses
+a Perl module named \texttt{IO::Uncompress::AnyUncompress} to read
+compressed log files; this saves users the trouble of uncompressing them to
+a temporary file before parsing begins.  When used interactively, the
+framework also displays a progress bar to show how far parsing has
+progressed through the log file and how long the remainder of the parsing
+process is likely to take.  The progress bar is not as reliable when
+parsing compressed log files, because a compressed block may uncompress to
+a variable number of log lines; there is also variation between the
+recognition and processing time of individual log lines, but overall the
+progress bar is a useful addition.
 
@@
 they have.  Three sets of data are collected:
+
+XXX CARL SUGGESTS ADDING GRAPHS AND STATS TABLES FOR THESE\@.
 
@@
 
-    \item The regex from the recognising rule and the log line it
-        recognised can be printed, so that the user can verify that the
-        correct rule recognised each log line.  Individual rules can also
-        set their debug field to true, and debugging information will be
-        printed each time they recognise a log line.
+    \item Individual rules can set their debug field to true, and debugging
+        information will be printed each time they recognise a log line.
 
@@
 
+    \item Every time a log line is recognised, the recognising rule's regex
+        and the log line can be printed, so that the user can verify that
+        the correct rule recognised each log line.
+
     \item Every connection and result added to the database can be dumped
@@
 \end{enumerate}
+
+A user, typically a mail administrator, would use these options when having
+difficulty adding or modifying rules to recognise log lines not recognised
+by \parsernames{} \numberOFrules{} rules.  Option 1 is useful for debugging
+a single rule, to check if it is recognising log lines it should not;
+figuring out that a rule is not recognising log lines that it should
+recognise is a more difficult task.  Option 2 is useful when \parsername{}
+is warning about parsing problems, because all the information about a mail
+will be present in the warning message: the full text of every log line
+that is part of the mail; the rule that recognised each log line; the
+filename of the log file and the line number within it for each log line.
+Option 3 is less useful because of the volume of data it produces, though
+for small log file snippets, e.g.\ 1000 log lines, it is possible to
+manually verify that the correct rule recognised each log line.  Option 4
+is not useful for most users: it is for debugging problems with the data
+that is saved to the database, and there are sufficient safeguards in place
+that there should be plenty of warning messages explaining what is wrong
+without using this option.
 
@@
 
-    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
-        \acronym{SMTP} command from the remote client.  This is the action
-        most frequently specified by rules, because so many different
-        restrictions are used to reject delivery attempts.  This action is
-        quite simple: if the log line contains a queueid, save the data
-        captured by the rule's regex to the mail identified by that
-        queueid; otherwise save it to the connection identified by the
-        \acronym{pid} in the log line.
+    \item [DELIVERY\_REJECTED] Postfix rejected an \acronym{SMTP} command
+        from the remote client.  This is the action most frequently
+        specified by rules, because so many different restrictions are used
+        to reject delivery attempts.  This action is quite simple: if the
+        log line contains a queueid, save the data captured by the rule's
+        regex to the mail identified by that queueid; otherwise save it to
+        the connection identified by the \acronym{pid} in the log line.
 
@@
 limitations identified in \parsername{}, followed by some possible
-improvements suggested by the author.
+improvements suggested by the author.  The next chapter will evaluate this
+implementation, examining both \parsernames{} efficiency and the coverage
+it achieves when parsing log files.

### 2009-04-06 doc/logparser-implementation.tex
Dealt with some XXX items. Improvements in results section. Added table showing number of rules per program.

@@
 
-    \item The regex is valid and compiling it for efficiency
-        (\sectionref{Caching compiled regexes}).
+    \item That the regex is valid.  The regex will also be compiled for
+        efficiency (\sectionref{Caching compiled regexes}).
 
@@
 
-XXX CARL SUGGESTS ADDING GRAPHS AND STATS TABLES FOR THESE\@.
+XXX ASK CARL IF I SHOULD HAVE MEAN, STANDARD DEVIATION, MORE GRAPHS\@?
 
@@
 
-    \item How long it takes to parse each log file.
-
-    \item The number of rules tried when recognising log lines; the
+    \item How long it takes to parse each log file (\graphref{parsing time
+        vs file size vs number of log lines graph}).
+
+    \item The number of rules tried when recognising log lines
+        (\graphref{Average number of rules tried per log line}); the
         framework may need to try multiple rules for a given log line
@@
 
-    \item How many log lines did the framework attempt to recognise, how
-        many were skipped because there were no rules whose condition was
-        equal to that log line's program, how many were successfully
-        recognised, and how many were not recognised.
+    \item The number of log lines and rules tried for each Postfix
+        component (\graphref{Average number of rules tried per log line for
+        each Postfix component}).
 
@@
 
-    \item XXX FURTHER COMPLICATIONS MAY ARISE\@.
-
-    \item XXX DATABASE GROWS WITHOUT BOUNDS\@.
+    \item Further complications may arise when parsing log files, and
+        \parsername{} will need to be modified to deal with them.
+
+    \item \parsername{} does not limit the size of the database, which will
+        grow without bounds unless the user deletes connections and results
+        from it.  This is both a limitation and a benefit: the benefit is
+        that data will never be unexpectedly deleted, but the limitation is
+        that the user must manage the size of the database.
 

### 2009-04-06 doc/logparser-results.tex
Dealt with some XXX items. Improvements in results section. Added table showing number of rules per program.

@@
 parse large log files.  The server that generated the log files used in
-testing this parser accepts approximately 10,000 mails for 700 users a day;
-\graphref{Mails received per day} and \tableref{Number of mails received
-per day: statistics} show that, as expected, far more mails are received on
-weekdays than at weekends.  The mail server in question is a production
-mail server handling mail for a university department; the benefit of using
-this server is that its log files exhibit the idiosyncrasies and
-peculiarities a mail server in the wild must deal with, but the downside is
-that significantly altering the configuration to accommodate this project
-is not an option.  Median log file size is 50MB, containing 285,000 log
-lines; large scale mail servers would have much larger log files.  Note
-that \graphref{Mails received per day} and \tableref{Number of mails
-received per day: statistics} show the number of mails received by
-\acronym{SMTP} only; in particular the mail loops noticeable in later
-graphs do not contribute to these figures.
+testing this parser accepts approximately 10,000 mails for 700 users per
+day; \graphref{Mails received per day} and \tableref{Number of mails
+received per day: statistics} show that, as expected, far more mails are
+received on weekdays than at weekends.  The mail server in question is a
+production mail server handling mail for a university department; the
+benefit of using this server is that its log files exhibit the
+idiosyncrasies and peculiarities a mail server in the wild must deal with,
+but the downside is that significantly altering the configuration to
+accommodate this project is not an option.  Median log file size is 50MB,
+containing 285,000 log lines; large scale mail servers would have much
+larger log files.  Note that \graphref{Mails received per day} and
+\tableref{Number of mails received per day: statistics} show the number of
+mails received by \acronym{SMTP} only, and the mail loops noticeable in
+later graphs do not contribute to these figures.
 
@@
 \numberOFlogFILES{} log files in one run without saving results to the
-database took \input{build/include-full-run-duration}, mean throughput is
+database took \input{build/include-full-run-duration}, mean throughput was
 \input{build/include-full-run-throughput}; median throughput when parsing
@@
 \input{build/include-insert-results-duration}, with mean throughput of
-\input{build/include-insert-results-throughput}; parsing makes up
+\input{build/include-insert-results-throughput}.  Parsing makes up only
 \input{build/include-skip-inserting-as-percentage-of-inserting} of the
@@
 log line reduces and so does the average parsing time per log line.
-\Graphref{Average number of rules tried per log line} shows the drop in the
-average number of rules tried per log line for log files containing a mail
-loop, and \graphref{Average number of rules tried per log line for each
-Postfix component} shows the average number of rules tried per log line for
-each Postfix component.
+\Graphref{Number of rules per Postfix component} shows the number of rules
+per Postfix component, \graphref{Average number of rules tried per log
+line} shows the drop in the average number of rules tried per log line for
+log files containing a mail loop, and \graphref{Average number of rules
+tried per log line for each Postfix component} shows the average number of
+rules tried per log line for each Postfix component.
 
@@
 
+\begin{table}[htbp]
+    \caption{Number of rules per Postfix component}
+    \empty{}\label{Number of rules per Postfix component}
+    \centering{}
+    \begin{tabular}{lr}
+        \tabletopline{}%
+        Postfix component & Number of rules \\
+        \tablemiddleline{}%
+        \input{build/include-number-of-rules-per-program}
+        \tablebottomline{}%
+    \end{tabular}
+\end{table}
+
 \FloatBarrier{}
@@
 
-At the time of writing there are \numberOFrules{} different rules: the top
-10\% recognise \input{build/include-top-ten-hits}\% of the log lines, with
-the remaining log lines split across the other 90\% of the rules (as shown
-in \graphref{rule hits graph}).  Assuming that the distribution of log
-lines is reasonably consistent over time, \parsernames{} efficiency should
-benefit from trying more frequently recognising rules before those that
-recognise less frequently.  To test this hypothesis three full test runs
-were performed with different rule orderings:
+\parsername{} has \numberOFrules{} rules: the top 10\% recognise
+\input{build/include-top-ten-hits}\% of the log lines, with the remaining
+log lines split across the other 90\% of the rules (as shown in
+\graphref{rule hits graph}).  Assuming that the distribution of log lines
+is reasonably consistent over time, \parsernames{} efficiency should
+benefit from trying rules that recognise log lines more frequently before
+those rules that recognise log lines less frequently.  To test this
+hypothesis, three full test runs were performed with different rule
+orderings:
 
@@
 
-\newpage{} % XXX GRAPH PLACEMENT FLIP-FLOPPING
-
 \Graphref{Parsing time relative to shuffled ordering graph} shows the
@@
 Optimal rule ordering has other benefits, described in
-\sectionref{scalability as the number of rules rises}.
+\sectionref{scalability as the number of rules rises}.  Differences in rule
+ordering have less effect on parsing time when parsing log files 22 and
+62--68, due to the different distribution of log lines in those log files.
 
@@
 table}
-
-\newpage{} % XXX VARIOREF PROBLEM
-
-Differences in rule ordering have less effect on parsing time when parsing
-log files 22 and 62--68, due to the different distribution of log lines in
-those log files.
-When a mail loop occurs the vast majority of log lines are from Postfix
-components that have few rules associated with them, whereas in general
-most log lines are from Postfix components that have a high number of
-associated rules; this results in the average number of rules required to
-parse a log line from log files 22 or 62--68 being considerably lower than
-usual.  The ratio of number of log lines vs.\ parsing time in
-\graphref{parsing time vs file size vs number of log lines factor} shows a
-noticeable improvement at log files 22 and 62--68, where a higher ratio
-means more log lines parsed per unit of time; similarly \graphref{Parsing
-time relative to shuffled ordering graph} shows a much smaller difference
-between the orderings at log files 22 and 62--68.
 

### 2009-04-06 doc/logparser-results.tex
Consistent float placement.
Added graphs showing parsing time of perfect best and worst vs optimal and reverse orderings. Started writing about perfect best and worst, but didn't get it finished.

@@
 
-\begin{table}[htbp]
+\begin{table}[thbp]
     \caption{Details of the computer used to generate statistics}
@@
 
-\begin{table}[htbp]
+\begin{table}[thbp]
     \caption{Number of rules per Postfix component}
@@
 Optimal rule ordering has other benefits, described in
-\sectionref{scalability as the number of rules rises}.  Differences in rule
+\sectionref{scalability as the number of rules rises}.  \Tableref{Parsing
+time relative to shuffled ordering table} shows that differences in rule
 ordering have less effect on parsing time when parsing log files 22 and
@@
 
-XXX WRITE ABOUT PERFECT RULE ORDERING\@; DUNNO WHETHER TO PUT IT IN WITH
-RULE ORDERING FOR EFFICIENCY, OR IN ITS OWN SECTION\@.
+XXX WRITE ABOUT PERFECT RULE ORDERING\@.
 
-XXX ADD A TABLE SHOWING PERFECT WORST RELATIVE TO PERFECT BEST\@.
+Optimal rule ordering, as described in \sectionref{rule ordering for
+efficiency}, is the best rule ordering it is possible to achieve without
+having an oracle that magically divines which rule should be used to
+recognise each log line.  Such an oracle would give perfect performance,
+because only one rule needs to be used to recognise each log line.
+\parsername{} can save a list showing which rule recognised each log line,
+and use that list to simulate having an oracle and improve parsing speed
+the second time a log file is parsed.   This does not provide a practical
+benefit, but it does provide a means to evaluate the performance of optimal
+rule ordering in comparison to an oracle.  The knowledge can also be used
+to generate a perfect worst oracle, where every rule (subject to rule
+conditions) is used before the correct rule.
+
+\Graphref{Parsing time relative to shuffled ordering including perfect best
+and worst graph} shows the parsing times of the perfect best oracle and
+perfect worst oracle relative to shuffled ordering, in addition to the
+parsing times of optimal and reverse orderings.  As expected, the perfect
+best oracle is more efficient than optimal ordering, and likewise the
+perfect worst oracle is less efficient than reverse ordering.
+
+XXX REFER TO SECOND AND THIRD GRAPHS, AND ADD TABLES SHOWING MEAN AND
+STDDEV FOR THE THREE GROUPS OF LOGS FILES\@.
+
+XXX WRITE SOMETHING TO MAKE IT EASIER TO GENERATE MEAN AND STDDEV FOR THREE
+GROUPS OF LOG FILES\@.
+
+\Graphref{Parsing time of perfect best oracle relative to optimal ordering}
 
 \showgraph{build/graph-perfect-best-and-perfect-worst-and-optimal-and-reverse-vs-shuffled}{Parsing
-time including perfect best and worst relative to shuffled
-ordering}{Parsing time relative to shuffled ordering including perfect best
-and worst graph}
+times relative to shuffled ordering}{Parsing time relative to shuffled
+ordering including perfect best and worst graph}
+
+\showgraph{build/graph-perfect-best-vs-optimal}{Parsing time of perfect
+best oracle relative to optimal ordering}{Parsing time of perfect best
+oracle relative to optimal ordering}
+
+\showgraph{build/graph-perfect-worst-vs-reverse}{Parsing time of perfect
+worst oracle relative to reverse ordering}{Parsing time of perfect worst
+oracle relative to reverse ordering}
 
 \showtable{build/include-perfect-best-and-optimal-and-reverse-and-perfect-worst-vs-shuffle}{Parsing
-time including perfect best and worst relative to shuffled
-ordering}{Parsing time including perfect best and worst relative to
+time including perfect best and worst oracles relative to shuffled
+ordering}{Parsing time including perfect best and worst oracles relative to
 shuffled ordering table}
 
-% XXX \FloatBarrier{} MIGHT BE USEFUL TO CONTROL WHERE FLOATS ARE PLACED.
+\FloatBarrier{}
 

### 2009-04-07 doc/logparser-implementation.tex
Consistent float placement.
Perfect oracle section is more or less finished, at last. Am now wondering if I should remove the perfect worst stuff.

@@
 consistent form, and performing housekeeping duties.  The intermingling of
-log entries from different mails immediately rules out the possibility of
+log lines from different mails immediately rules out the possibility of
 handling each mail in isolation; the parser must be able to handle multiple
@@
         \daemon{pickup} service dealing with a locally submitted mail.  A
-        new mail will be created, although out of order log entries may
-        have caused it to already exist, as documented in
-        \sectionref{pickup logging after cleanup}.
+        new mail will be created, although out of order log lines may have
+        caused it to already exist, as documented in \sectionref{pickup
+        logging after cleanup}.
 
@@
 % do not reformat this!
-\begin{table}[ht]
+\begin{table}[thbp]
     \caption{Example of a rule used by \parsernameshort{}}

### 2009-04-07 doc/logparser-results.tex
Perfect oracle section is more or less finished, at last. Am now wondering if I should remove the perfect worst stuff.

@@
 
+% Reduce the gap between columns in tables.
+\addtolength{\tabcolsep}{-1pt}
+
 XXX NEED AN INTRODUCTION\@.
-
-XXX USE THE DATA COLLECTED ABOUT NUMBER OF RULES TRIED\@.  MAYBE INCLUDE IN
-ORDERINGS SECTION\@?  MIX WITH RECOGNISING VS RECOGNISING + ACTION, AND
-KEEP COUNTERS FOR EACH PROGRAM TOO --- MAYBE SOMETHING USEFUL WILL COME
-FROM IT\@.  COMPARE NUMBER OF RULES TRIED FOR DIFFERENT RULE ORDERINGS\@.
-
-XXX SHOULD I USE $\backslash{}$clearpage{} OR SOMETHING TO AFFECT FLOAT
-PLACEMENT\@?  GRAPHS AND TABLES TEND TO MOVE QUITE FAR IN SOME OF THE
-SUBSECTIONS, THOUGH MAKING THE TEXT LONGER MIGHT HELP TOO\@.  IDEALLY I
-WANT TO HAVE TEXT, THEN ALL GRAPHS AND TABLES, THEN THE NEXT SECTION\@.
-
-\newpage{} % XXX VARIOREF PROBLEMS
 
@@
     \empty{}\label{Details of the computer used to generate statistics}
+    \centering{}
     \begin{tabular}[]{ll}
@@
 quite tightly banded, showing that the algorithm scales linearly; the
-ratios increase (i.e.\ improve) for log files 22 and 62--68, despite their
+ratios increase (i.e.\ improve) for log files 22 \& 62--68, despite their
 larger than average size.  Both groups of log files are much larger than
@@
 time relative to shuffled ordering table} shows that differences in rule
-ordering have less effect on parsing time when parsing log files 22 and
+ordering have less effect on parsing time when parsing log files 22 \&
 62--68, due to the different distribution of log lines in those log files.
@@
 
-\subsection{Perfect Rule Ordering}
+\FloatBarrier{}
+
+\subsection{Comparing Optimal Ordering Against A Perfect Oracle}
 
@@
 
-XXX WRITE ABOUT PERFECT RULE ORDERING\@.
+XXX SHOULD I BOTHER WITH PERFECT WORST\@?
 
@@
 \Graphref{Parsing time relative to shuffled ordering including perfect best
-and worst graph} shows the parsing times of the perfect best oracle and
-perfect worst oracle relative to shuffled ordering, in addition to the
-parsing times of optimal and reverse orderings.  As expected, the perfect
+and worst graph} shows the parsing times of the perfect best and perfect
+worst oracles, and the optimal and reverse orderings relative to shuffled
+ordering; \tableref{Parsing time relative to shuffled ordering, including
+oracles table} shows mean and standard deviation.  As expected, the perfect
 best oracle is more efficient than optimal ordering, and likewise the
 perfect worst oracle is less efficient than reverse ordering.
+\Graphref{Percentage increase in parsing time when using optimal ordering
+instead of perfect best oracle graph} and \graphref{Percentage decrease in
+parsing time when using reverse ordering instead of perfect worst oracle
+graph} show the percentage increase and decrease in parsing time when
+using, respectively, optimal ordering instead of the perfect best oracle,
+and reverse ordering instead of the perfect worst oracle; mean and standard
+deviation are given for different groups of log files in
+\tableref{Percentage increase in parsing time when using optimal ordering
+instead of perfect best oracle table} and \tableref{Percentage decrease in
+parsing time when using reverse ordering instead of perfect worst oracle
+table}.
 
-XXX REFER TO SECOND AND THIRD GRAPHS, AND ADD TABLES SHOWING MEAN AND
-STDDEV FOR THE THREE GROUPS OF LOGS FILES\@.
+Once again, the difference between the oracles and optimal or reversed
+orderings at its lowest when parsing log files resulting from a mail loop
+(log files 22 \& 62--68), because the average number of rules tried per log
+line is lower (see \graphref{Average number of rules tried per log line}).
+The performance when parsing the first log file is different to the
+remainder of the non-mail loop log files, because the hits field of every
+rule is zero; optimal ordering is less efficient, and reverse ordering is
+less inefficient.
 
-XXX WRITE SOMETHING TO MAKE IT EASIER TO GENERATE MEAN AND STDDEV FOR THREE
-GROUPS OF LOG FILES\@.
-
-\Graphref{Parsing time of perfect best oracle relative to optimal ordering}
+Optimal ordering proves to be quite efficient: \tableref{Percentage
+increase in parsing time when using optimal ordering instead of perfect
+best oracle table} shows that optimal ordering is less than
+\input{build/include-perfect-best-vs-optimal-mean} slower than parsing
+using a magical oracle that divines the correct rule to use for each log
+line.
 
 \showgraph{build/graph-perfect-best-and-perfect-worst-and-optimal-and-reverse-vs-shuffled}{Parsing
-times relative to shuffled ordering}{Parsing time relative to shuffled
-ordering including perfect best and worst graph}
-
-\showgraph{build/graph-perfect-best-vs-optimal}{Parsing time of perfect
-best oracle relative to optimal ordering}{Parsing time of perfect best
-oracle relative to optimal ordering}
-
-\showgraph{build/graph-perfect-worst-vs-reverse}{Parsing time of perfect
-worst oracle relative to reverse ordering}{Parsing time of perfect worst
-oracle relative to reverse ordering}
+time relative to shuffled ordering, including oracles}{Parsing time
+relative to shuffled ordering including perfect best and worst graph}
 
 \showtable{build/include-perfect-best-and-optimal-and-reverse-and-perfect-worst-vs-shuffle}{Parsing
-time including perfect best and worst oracles relative to shuffled
-ordering}{Parsing time including perfect best and worst oracles relative to
-shuffled ordering table}
+time relative to shuffled ordering, including oracles}{Parsing time
+relative to shuffled ordering, including oracles table}
+
+\showgraph{build/graph-perfect-best-vs-optimal}{Percentage increase in
+parsing time when using optimal ordering instead of perfect best
+oracle}{Percentage increase in parsing time when using optimal ordering
+instead of perfect best oracle graph}
+
+\showtable{build/include-perfect-best-vs-optimal-stddev}{Percentage
+increase in parsing time when using optimal ordering instead of perfect
+best oracle}{Percentage increase in parsing time when using optimal
+ordering instead of perfect best oracle table}
+
+\showgraph{build/graph-perfect-worst-vs-reverse}{Percentage decrease in
+parsing time when using reverse ordering instead of perfect worst
+oracle}{Percentage decrease in parsing time when using reverse ordering
+instead of perfect worst oracle graph}
+
+\showtable{build/include-perfect-worst-vs-reverse-stddev}{Percentage
+decrease in parsing time when using reverse ordering instead of perfect
+worst oracle}{Percentage decrease in parsing time when using reverse
+ordering instead of perfect worst oracle table}
 
@@
 expected that the typical parser will have a large number of rules.  There
-are \numberOFrules{} rules in the full \parsername{} ruleset, whereas the minimum
-number of rules required to parse the \numberOFlogFILES{} log files used
-when generating the results in this thesis is \numberOFrulesMINIMUM{},
+are \numberOFrules{} rules in the full \parsername{} ruleset, whereas the
+minimum number of rules required to parse the \numberOFlogFILES{} log files
+used when generating the results in this thesis is \numberOFrulesMINIMUM{},
 \numberOFrulesMINIMUMpercentage{} of the full ruleset.  The full ruleset is
 larger because \parsername{} is tested with \numberOFlogFILESall{} log
-files, to improve its parsing and ensure it correctly parses logs generated
-by later Postfix versions.  A second set of test runs was performed using
-the minimum ruleset, and the parsing times compared to those generated
-using the full ruleset: the percentage parsing time increase when using the
-full ruleset instead of the minimal ruleset for optimal, shuffled and
-reverse orderings is shown in \graphref{Percentage parsing time increase of
-maximum ruleset over minimum ruleset}.
+files, to improve its parsing and ensure it correctly parses log files
+generated by later Postfix versions.  A second set of test runs was
+performed using the minimum ruleset, and the parsing times compared to
+those generated using the full ruleset: the percentage parsing time
+increase when using the full ruleset instead of the minimal ruleset for
+optimal, shuffled and reverse orderings is shown in \graphref{Percentage
+parsing time increase of maximum ruleset over minimum ruleset}.
 
@@
 
+\newpage{} % XXX
+
 \subsection{Log Lines Covered}
@@
 other log files).
+
+\newpage{} % XXX
 

### 2009-04-07 doc/logparser-results.tex
Removed perfect worst and reverse ordering from the oracle section.

@@
 
-XXX SHOULD I BOTHER WITH PERFECT WORST\@?
-
 Optimal rule ordering, as described in \sectionref{rule ordering for
@@
 \parsername{} can save a list showing which rule recognised each log line,
-and use that list to simulate having an oracle and improve parsing speed
-the second time a log file is parsed.   This does not provide a practical
+and use that list to simulate an oracle and improve parsing speed the
+second time a log file is parsed.  This does not provide a practical
 benefit, but it does provide a means to evaluate the performance of optimal
-rule ordering in comparison to an oracle.  The knowledge can also be used
-to generate a perfect worst oracle, where every rule (subject to rule
-conditions) is used before the correct rule.
+rule ordering in comparison to an oracle.
 
-\Graphref{Parsing time relative to shuffled ordering including perfect best
-and worst graph} shows the parsing times of the perfect best and perfect
-worst oracles, and the optimal and reverse orderings relative to shuffled
-ordering; \tableref{Parsing time relative to shuffled ordering, including
-oracles table} shows mean and standard deviation.  As expected, the perfect
-best oracle is more efficient than optimal ordering, and likewise the
-perfect worst oracle is less efficient than reverse ordering.
-\Graphref{Percentage increase in parsing time when using optimal ordering
-instead of perfect best oracle graph} and \graphref{Percentage decrease in
-parsing time when using reverse ordering instead of perfect worst oracle
-graph} show the percentage increase and decrease in parsing time when
-using, respectively, optimal ordering instead of the perfect best oracle,
-and reverse ordering instead of the perfect worst oracle; mean and standard
-deviation are given for different groups of log files in
-\tableref{Percentage increase in parsing time when using optimal ordering
-instead of perfect best oracle table} and \tableref{Percentage decrease in
-parsing time when using reverse ordering instead of perfect worst oracle
-table}.
+\Graphref{Parsing time of perfect best oracle and optimal ordering relative
+to shuffled ordering graph} shows the parsing times of the perfect best
+oracle and the optimal ordering, relative to shuffled ordering;
+\tableref{Parsing time of perfect best oracle and optimal ordering relative
+to shuffled ordering table} shows mean and standard deviation.  As
+expected, the perfect best oracle is more efficient than optimal ordering.
+\Graphref{Percentage increase in parsing time when using optimal
+ordering instead of perfect best oracle graph} shows the percentage
+increase in parsing time when using optimal ordering instead of the perfect
+best oracle; mean and standard deviation are given for different groups of
+log files in \tableref{Percentage increase in parsing time when using
+optimal ordering instead of perfect best oracle table}.
 
-Once again, the difference between the oracles and optimal or reversed
-orderings at its lowest when parsing log files resulting from a mail loop
+Once again, the difference between the perfect best oracle and optimal
+ordering is at its lowest when parsing log files resulting from a mail loop
 (log files 22 \& 62--68), because the average number of rules tried per log
 line is lower (see \graphref{Average number of rules tried per log line}).
-The performance when parsing the first log file is different to the
-remainder of the non-mail loop log files, because the hits field of every
-rule is zero; optimal ordering is less efficient, and reverse ordering is
-less inefficient.
+The performance of optimal ordering when parsing the first log file is much
+worse than the remainder of the log files, because the hits field of every
+rule is zero, so optimal ordering performs similarly to shuffled ordering
+for that log file.
 
@@
 
-\showgraph{build/graph-perfect-best-and-perfect-worst-and-optimal-and-reverse-vs-shuffled}{Parsing
-time relative to shuffled ordering, including oracles}{Parsing time
-relative to shuffled ordering including perfect best and worst graph}
+\showgraph{build/graph-perfect-best-and-optimal-vs-shuffled}{Parsing time
+of perfect best oracle and optimal ordering relative to shuffled
+ordering}{Parsing time of perfect best oracle and optimal ordering relative
+to shuffled ordering graph}
 
-\showtable{build/include-perfect-best-and-optimal-and-reverse-and-perfect-worst-vs-shuffle}{Parsing
-time relative to shuffled ordering, including oracles}{Parsing time
-relative to shuffled ordering, including oracles table}
+\showtable{build/include-perfect-best-and-optimal-vs-shuffle}{Parsing time
+of perfect best oracle and optimal ordering relative to shuffled
+ordering}{Parsing time of perfect best oracle and optimal ordering relative
+to shuffled ordering table}
 
@@
 
-\showgraph{build/graph-perfect-worst-vs-reverse}{Percentage decrease in
-parsing time when using reverse ordering instead of perfect worst
-oracle}{Percentage decrease in parsing time when using reverse ordering
-instead of perfect worst oracle graph}
-
-\showtable{build/include-perfect-worst-vs-reverse-stddev}{Percentage
-decrease in parsing time when using reverse ordering instead of perfect
-worst oracle}{Percentage decrease in parsing time when using reverse
-ordering instead of perfect worst oracle table}
-
 \FloatBarrier{}
@@
 PROCESSING THEM\@.
+
+\showgraph{build/graph-percentage-time-spent-recognising-log-lines}{XXX}{XXX}
 

### 2009-04-08 doc/logparser-results.tex
Finished (I hope) the minimum vs maximum ruleset section.

@@
 
-\subsection{Where Is The Time Spent: Recognising Or Processing Log Lines?}
-
-\label{recognising vs processing}
-
-XXX EXAMINE THE TIME SPENT RECOGNISING LOG LINES VS THE TIME SPENT
-PROCESSING THEM\@.
-
-\showgraph{build/graph-percentage-time-spent-recognising-log-lines}{XXX}{XXX}
-
 \subsection{Scalability When The Number Of Rules Rises}
@@
 minimum number of rules required to parse the \numberOFlogFILES{} log files
-used when generating the results in this thesis is \numberOFrulesMINIMUM{},
-\numberOFrulesMINIMUMpercentage{} of the full ruleset.  The full ruleset is
-larger because \parsername{} is tested with \numberOFlogFILESall{} log
-files, to improve its parsing and ensure it correctly parses log files
-generated by later Postfix versions.  A second set of test runs was
-performed using the minimum ruleset, and the parsing times compared to
-those generated using the full ruleset: the percentage parsing time
-increase when using the full ruleset instead of the minimal ruleset for
-optimal, shuffled and reverse orderings is shown in \graphref{Percentage
-parsing time increase of maximum ruleset over minimum ruleset}.
+used when generating the data in this chapter is \numberOFrulesMINIMUM{}
+rules, \numberOFrulesMINIMUMpercentage{} of the full ruleset.  The full
+ruleset is larger because \parsername{} is tested with
+\numberOFlogFILESall{} log files; testing with more log files increases the
+chance of finding bugs in the parser or new complications to be overcome.
+The \numberOFlogFILES{} log files were each parsed 10 times using the
+minimum ruleset, and the parsing times compared to those generated using
+the full ruleset: the percentage parsing time increase when using the full
+ruleset instead of the minimal ruleset for optimal, shuffled and reverse
+orderings is shown in \graphref{Percentage parsing time increase of maximum
+ruleset over minimum ruleset}; mean and standard deviation are shown in
+\tableref{Percentage parsing time increase of maximum ruleset over minimum
+ruleset table}.
+
+Clearly the increased number of rules has a noticeable performance impact
+with reverse ordering, and a lesser impact with shuffled ordering, whereas
+optimal ordering shows very little change.  Log files 22 \& 62--68 show
+much smaller increases than other log files do, because the majority of log
+lines in those log files are produced by Postfix components with a small
+number of rules, so removing unnecessary rules has little effect on the
+total number of rules used; \tableref{Number of rules per Postfix component
+in the maximum and minimum rulesets} shows the number of rules per Postfix
+component for each ruleset.  The optimal ordering shows a mean increase of
+\input{build/include-full-ruleset-vs-minimum-ruleset-mean} in parsing time
+for a \numberOFrulesMAXIMUMpercentage{} increase in the number of rules.
+These results show that the architecture scales very well as the number of
+rules increases, and that optimally sorting the rules is an important
+optimisation contributing to this scalability.
 
@@
 
-Clearly the increased number of rules has a noticeable performance impact
-with reverse ordering, and a lesser impact with shuffled ordering.  The
-optimal ordering shows a mean increase of
-\input{build/include-full-ruleset-vs-minimum-ruleset} in parsing time for a
-\numberOFrulesMAXIMUMpercentage{} increase in the number of rules.
-XXXXXXXXXXXX IMPROVE THIS EXPLANATION\@.  Log files 22 and 62--68 show much
-smaller increases for shuffled and reversed ordering than other log files
-do, because the majority of log lines in those log files are parsed by
-Postfix components with a small number of parsing rules, so removing
-unnecessary rules has little effect on the total number of rules used.
-These results show that the architecture scales very well as the number of
-rules increases, and that optimally sorting the rules is an important
-optimisation contributing to this scalability.
+\showtable{build/include-full-ruleset-vs-minimum-ruleset}{Percentage parsing
+time increase of maximum ruleset over minimum ruleset}{Percentage parsing
+time increase of maximum ruleset over minimum ruleset table}
+
+\begin{table}[thbp]
+    \caption{Number of rules per Postfix component in the maximum and
+    minimum rulesets}
+    \empty{}\label{Number of rules per Postfix component in the maximum and
+    minimum rulesets}
+    \centering{}
+    \begin{tabular}{lrr}
+        \tabletopline{}%
+        Postfix component & Maximum ruleset & Minimum ruleset \\
+        \tablemiddleline{}%
+        \input{build/include-number-of-rules-per-program-minimum-ruleset}
+        \tablebottomline{}%
+    \end{tabular}
+\end{table}
+
+\FloatBarrier{}
 
@@
 XXX UPDATE THIS\@.  Two large dips can be seen in \graphref{Increase in
-parsing time when not caching compiled regexes graph} at log files 22 and
+parsing time when not caching compiled regexes graph} at log files 22 \&
 62--68, corresponding to the spikes in log file size in \graphref{parsing
@@
 when not caching compiled regexes table}
+
+\subsection{Where Is The Time Spent: Recognising Or Processing Log Lines?}
+
+\label{recognising vs processing}
+
+XXX COMMENT THAT THE OPTIMISATION TIME HAS BEEN SPENT ON RULES, AND THAT IS
+GOOD BECAUSE THAT IS WHERE THE MAJORITY OF EXECUTION TIME IS SPENT\@.
+ALSO, RULE OPTIMISATION IS GENERIC, WHEREAS ACTION OPTIMISATION IS
+PARSER-SPECIFIC\@.
+
+XXX EXAMINE THE TIME SPENT RECOGNISING LOG LINES VS THE TIME SPENT
+PROCESSING THEM\@.
+
+\showgraph{build/graph-percentage-time-spent-recognising-log-lines}{XXX}{XXX}
 

### 2009-04-08 doc/logparser.tex
Updated results after stats-max run finished. Added perfect best and perfect worst graph, and a table showing the std dev etc. Worked on graph and table placement, without much success.  I think manual     tweaking will be required, though adding 'h' back into the placement list     has helped. Added a graph for the number of times each action is invoked.
Added a table showing mean and stddev for the number of invocations of each action, for different groups of log files.
Use the booktabs packages to provide better horizontal rules for tables.
Removed some make todo false positives. Removed so

@@
 \usepackage[final]{graphicx}
+% Provides \FloatBarrier that stops floats going past it.
+\usepackage{placeins}
+% Relax the parameters Latex uses when placing floats.
+\usepackage{flexiblefloats}
+% Prevent floats from appearing above the line that includes them
+\usepackage{flafter}
 % Add the bibliography into the table of contents.
@@
 \usepackage{lastpage}
+
+% Better looking horizontal rules for tables.
+\usepackage{booktabs}
 
@@
 \usepackage{varioref}
-% Warnings rather than errors when references cross page boundaries.
-\vrefwarning{}
-% Errors when loops are encountered
-\vrefshowerrors{}
 % Shorten the text used for references with page numbers.
@@
 % Reduce the space between items in a list; this is useful when each item
-% is a single line, because then the defaul spacing makes the list look
+% is a single line, because then the default spacing makes the list look
 % very sparse.  This command needs to be the first thing in a list to take
@@
 % \showgraph{filename}{caption}{label}
-\newcommand{\showgraph}[3]{%
-    % XXX REVISIT THE PLACEMENT OF FIGURES AND TABLES, THOUGH I THINK
-    % LATEX'S DEFAULTS ARE PROBABLY PRETTY GOOD.
-    \begin{figure}%[hbt!]
-        \caption{#2}\label{#3}
-        \includegraphics{#1}
+\newcommand{\showgraph}[4][thbp]{%
+    \begin{figure}[#1]
+        \caption{#3}\label{#4}
+        \includegraphics{#2}
     \end{figure}
@@
 %\showtable{filename}{caption}{label}
-\newcommand{\showtable}[3]{
-    \begin{table}%[ht]
-        \caption{#2}\label{#3}
-        \input{#1}
+\newcommand{\showtable}[4][thbp]{%
+    \begin{table}[#1]
+        \caption{#3}\label{#4}
+        \input{#2}
     \end{table}
@@
 \newcommand{\tabletopline}[0]{%
-    \hline%
-    \noalign{\smallskip}%
+    \toprule{}%
 }
@@
 \newcommand{\tablebottomline}[0]{%
-    \noalign{\smallskip}%
-    \hline%
+    \bottomrule{}%
 }
@@
 \newcommand{\tablemiddleline}[0]{%
-    \noalign{\smallskip}%
-    \hline%
-    \noalign{\smallskip}%
+    \midrule{}%
 }
@@
 
-% XXX UPDATE THIS WHEN all-logs PASSES
 \newcommand{\numberOFlogFILESall}[0]{%
-    522%
+    774%
 }
@@
 
-    XXX FIX THE PROBLEMS THAT AROSE FROM THE LAST all-logs RUN\@.
-
-    XXX RUN ``test-some'' AGAIN TO UPDATE THE DB\@.
-
-    XXX SHOULD I HAVE A ``FUTURE WORK'' SECTION SOMEWHERE IN THE
-    CONCLUSION\@?
+    XXX ASK CARL\@: SHOULD I HAVE A ``FUTURE WORK'' SECTION SOMEWHERE IN
+    THE CONCLUSION\@?
 
     XXX MAIL MADS ABOUT EXAMINERS\@.
+
+    XXX SPELL CHECK AND RUN DICTION\@.
+
+    XXX MAKE IT EASY TO REDUCE THE HEIGHT OF GRAPHS\@.
 

### 2009-04-08 doc/logparser-results.tex
Worked on caching compiled regexes, and I'm pretty happy with it. Referred to the strange performance parsing the first log file in other sections.

@@
 when a mail loop occurs and the distribution of log lines is unusual.
-Optimal rule ordering has other benefits, described in
-\sectionref{scalability as the number of rules rises}.  \Tableref{Parsing
-time relative to shuffled ordering table} shows that differences in rule
-ordering have less effect on parsing time when parsing log files 22 \&
-62--68, due to the different distribution of log lines in those log files.
+\Tableref{Parsing time relative to shuffled ordering table} shows that
+differences in rule ordering have less effect on parsing time when parsing
+log files 22 \& 62--68, due to the different distribution of log lines in
+those log files.  A careful examination of \graphref{Parsing time relative
+to shuffled ordering graph} shows that, for the first log file, optimal and
+reverse orderings perform identically: this is because the hits field of
+each rule is zero for the first log file, so optimal and reverse orderings
+are identical.  For the first log file, shuffled ordering is most
+efficient, but that is accidental and cannot be relied upon.
 
@@
 worse than the remainder of the log files, because the hits field of every
-rule is zero, so optimal ordering performs similarly to shuffled ordering
-for that log file.
+rule is zero, so optimal ordering does not provide any benefit for that log
+file.
 
@@
 in the maximum and minimum rulesets} shows the number of rules per Postfix
-component for each ruleset.  The optimal ordering shows a mean increase of
+component for each ruleset.  Once again, for the first log file, optimal
+and reverse orderings have identical performance, because the hits field of
+every rule is zero.
+
+The optimal ordering shows a mean increase of just
 \input{build/include-full-ruleset-vs-minimum-ruleset-mean} in parsing time
@@
 CPU time: in many cases it takes far more CPU time than the actual
-matching.  If the interpreter is certain that a regex will not change, it
-will automatically cache the compilation results for later use.   The
-results of compiling a dynamically generated regex can be cached and used
-in preference to the original regex, but it is the responsibility of the
-programmer to do this.  \parsername{} loads the rules it uses from a
-database, so it compiles each rule's regex and caches the result to reduce
-parsing time.
+matching.  If the interpreter is certain that a regex will not change while
+the program is running, it will automatically cache the results of
+compiling and optimising the regex for later use.  The results of compiling
+a dynamically generated regex can be cached and used in preference to the
+original regex, but it is the responsibility of the programmer to do this;
+\parsername{} does this with every rule's regex when the rules are loaded
+from the database.
 
@@
 shows the impact that not caching compiled regexes has on parser
-performance: on typical log files, the parsing time when not caching
-compiled regexes is 400--600\% of the parsing time when caching.  Caching
+performance, with mean and standard deviation in \tableref{Increase in
+parsing time when not caching compiled regexes table}.  For typical log
+files, the mean increase in parsing time when not caching compiled regexes
+is \input{build/include-cached-regexes-vs-discarded-regexes-mean}.  Caching
 compiled regexes is probably the single most effective optimisation
-possible in \parsername{}, and was quite simple to implement.
+possible in \parsername{}, and was quite simple to implement.  As with
+previous optimisations, log files 22 \& 62--68 do not suffer such a large
+increase in parsing time when not caching compiled regexes; this is
+because, on average, fewer regexes are compiled per log line for those log
+files.  The increase in parsing time when parsing the first log file is
+much greater than for the other log files; again, this is because every
+rule's hits field is zero, so optimal ordering is less efficient than
+usual, and the average number of rules tried for each log line will be
+higher than usual.
 
@@
 
-XXX UPDATE THIS\@.  Two large dips can be seen in \graphref{Increase in
-parsing time when not caching compiled regexes graph} at log files 22 \&
-62--68, corresponding to the spikes in log file size in \graphref{parsing
-time vs file size vs number of log lines graph}.  The distribution of log
-lines across rules when there is a mail loop is much different to the norm,
-and the average number of rules consulted per log line is much lower: this
-results in far fewer regex compilations per line when there is a mail loop,
-and a corresponding decrease in the average parsing time for a single log
-line.  The outcome is that caching compiled regexes is proportionally less
-important when the log file contents were created by a mail loop.  The
-increases in parsing time when not caching compiled regexes for different
-groups of log files are summarised in \tableref{Increase in parsing time
-when not caching compiled regexes table}.
-
 \showtable{build/include-cached-regexes-vs-discarded-regexes}{Increase in
@@
 when not caching compiled regexes table}
+
+\FloatBarrier{}
 
@@
 \showgraph{build/graph-percentage-time-spent-recognising-log-lines}{XXX}{XXX}
+
+\FloatBarrier{}
 

### 2009-04-08 doc/logparser-complications.tex
Final corrections/improvements based on Carl's comments.

@@
 chosen, because that approach yielded the biggest improvement in the
-parser, and made the remaining problems more apparent.
+parser, and made the remaining problems more apparent.  Several of the
+complications were discovered because \parsername{} is very careful to
+check the origin of each mail; if it does not know the origin of a mail,
+that mail will be tagged as fake, and a warning issued if the mail's origin
+has not been determined before trying to save it to the database.  Some of
+the solutions to these complications require discarding log lines because
+they are out of order, and, less frequently, discarding a data structure
+containing data gathered about a connection or a mail, because of the
+paucity of information contained in the data structure.
 

### 2009-04-08 doc/logparser-implementation.tex
Final corrections/improvements based on Carl's comments.

@@
 eccentricities and oddities of Postfix log files, presenting the resulting
-data in a normalised, simple to use representation.  The parser's task is
-to follow the journey each mail takes through Postfix, combining the data
-captured by rules into a coherent whole, saving it in a useful and
-consistent form, and performing housekeeping duties.  The intermingling of
-log lines from different mails immediately rules out the possibility of
-handling each mail in isolation; the parser must be able to handle multiple
-mails in parallel, each potentially at a different stage in its journey,
-without any interference between mails --- except in the minority of cases
-where intra-mail interference is required, e.g.\ mail re-injected for
-forwarding (\sectionref{Re-injected mails}).  The best way to deal with
-intermingling is to maintain state information for every unfinished mail
-and manipulate the appropriate mail correctly for each log line
+data in a normalised, simple to use representation.  Unfortunately dealing
+with the complications that arise sometimes requires the parser to discard
+log lines (e.g.\ \sectionref{aborted delivery attempts}), and, less
+frequently, to discard a data structure (e.g.\ \sectionref{timeouts during
+data phase}).  \parsername{} is capable of parsing log files from Postfix
+version 2.2 through to version 2.5, and should be capable of parsing log
+files from later versions with only minor modifications or an updated
+ruleset.  The parser's task is to follow the journey each mail takes
+through Postfix, combining the data captured by rules into a coherent
+whole, saving it in a useful and consistent form, and performing
+housekeeping duties.  The intermingling of log lines from different mails
+immediately rules out the possibility of handling each mail in isolation;
+the parser must be able to handle multiple mails in parallel, each
+potentially at a different stage in its journey, without any interference
+between mails --- except in the minority of cases where intra-mail
+interference is required, e.g.\ mail re-injected for forwarding
+(\sectionref{Re-injected mails}).  The best way to deal with intermingling
+of log lines is to maintain state information for every unfinished mail,
+and to manipulate the appropriate mail correctly for each log line
 encountered.
@@
 The flow chart in \figureref{flow chart image} shows the most common paths
-a connection or mail can take through \parsername{}; the difficulties
-described in \sectionref{complications} are excluded for the sake of
-clarity.  The flow chart is intended to be a graphical overview of how a
-mail progresses through both Postfix and \parsername{}, providing an
+a connection or mail can take through \parsername{}; decision boxes and the
+difficulties described in \sectionref{complications} are excluded for the
+sake of clarity.  The flow chart is intended to be a graphical overview of
+how a mail progresses through both Postfix and \parsername{}, providing an
 overall context into which the detailed descriptions in the remainder of
@@
 processed is held in \texttt{connections} and \texttt{queueids}
-respectively.  The other state tables used when working around
+respectively.  The other state tables are used when working around
 complications in parsing: \texttt{timeout\_queueids} is used when dealing
@@
 recognition and processing time of individual log lines, but overall the
-progress bar is a useful addition.
+progress bar is a useful addition.  \parsername{} is primarily intended for
+parsing complete log files, but with minor modifications it could be used
+to parse a live log file, periodically checking it for new log lines; this
+could be very useful for programs that work best with up to date data,
+e.g.\ a program for monitoring the health of the mail system.
 
@@
 to improve parsing speed can be turned off or altered to measure the effect
-they have.  Three sets of data are collected:
-
-XXX ASK CARL IF I SHOULD HAVE MEAN, STANDARD DEVIATION, MORE GRAPHS\@?
+they have.  In each case the data gathered with the optimisation disabled
+will be compared to the data gathered with it enabled, to quantify the
+benefit each optimisation provides.  Three sets of data are collected:
 
@@
 There are five ways the framework can adapt its behaviour to demonstrate
-how effective the optimisations it uses are, and how efficient
-\parsername{} is:
+how effective each optimisation is:
 
@@
         chapter measures the speed of \parsername{}, not the speed of the
-        database or the disks it resides on.
+        database or the disks it resides on.  The impact of storing results
+        in the database is described at the beginning of
+        \sectionref{Results}.
 
@@
 This utility is not intended to create perfect regexes, but it greatly
-reduces the effort required to deal with unrecognised log lines.
+reduces the effort required to deal with unrecognised log lines.  The
+regexes it generates will be self-contained: a parser that relies on using
+cascaded parsing would require a modified algorithm, perhaps by replacing
+the pre-processing stage with one that applies existing cascading rules to
+each log line, and uses the resulting modified log lines for the remainder
+of the algorithm.
 
@@
 \end{eqlist}
+
+There is some similarity between regex components and cascaded parsing:
+each regex component resembles a rule that recognises part of a log line
+and consumes it, leaving the remainder to be recognised by other components
+or cascaded rules.  The major difference between the two is that regex
+components are explicitly used by the author of the ruleset, whereas
+cascaded parsing would be dynamically applied by the framework.
 

### 2009-04-08 doc/logparser-results.tex
Wrote the section about recognition vs recognition plus processing.

@@
 % Reduce the gap between columns in tables.
-\addtolength{\tabcolsep}{-1pt}
+\addtolength{\tabcolsep}{-2pt}
 
@@
 
-XXX COMMENT THAT THE OPTIMISATION TIME HAS BEEN SPENT ON RULES, AND THAT IS
-GOOD BECAUSE THAT IS WHERE THE MAJORITY OF EXECUTION TIME IS SPENT\@.
-ALSO, RULE OPTIMISATION IS GENERIC, WHEREAS ACTION OPTIMISATION IS
-PARSER-SPECIFIC\@.
+The optimisations described in this chapter have optimised the process of
+recognising log lines, but have not affected the actions at all.  There are
+two reasons for this:
 
-XXX EXAMINE THE TIME SPENT RECOGNISING LOG LINES VS THE TIME SPENT
-PROCESSING THEM\@.
+\begin{enumerate}
 
-\showgraph{build/graph-percentage-time-spent-recognising-log-lines}{XXX}{XXX}
+    \item With the optimisations described in this chapter enabled,
+        recognising log lines still dominates the execution time of the
+        parser.  \Graphref{Percentage of parsing time spent recognising log
+        lines graph} shows the percentage of parsing time spent recognising
+        log lines for each of the \numberOFlogFILES{} log files, with mean
+        and standard deviation shown in \tableref{Percentage of parsing
+        time spent recognising log lines table}; for normal log files
+        \input{build/include-percentage-time-spent-recognising-log-lines-mean}
+        of parsing time is spent recognising log lines.  Optimal ordering
+        reduces parsing time by
+        \input{build/include-optimal-ordering-parsing-time-reduction-other-logs}\%
+        relative to shuffled ordering; processing recognised log lines
+        occupies
+        \input{build/include-percentage-time-spent-processing-log-lines-mean}
+        of parsing time, less if using shuffled ordering, so optimising
+        actions could not possibly provide as big a performance increase as
+        optimally ordering rules.
+
+    \item The process of recognising log lines is not parser-specific
+        (excluding evaluation of rule conditions), so the optimisations
+        described in this chapter are applicable to all parsers based on
+        this architecture.  Actions are parser-specific, so it is unlikely
+        that any optimisations made to actions would be portable to other
+        parsers.
+
+\end{enumerate}
+
+Individual actions or the framework could and have been optimised, but
+there is plenty of existing literature on optimising programs, so the
+subject will not be dealt with here.
+
+\showgraph{build/graph-percentage-time-spent-recognising-log-lines}{Percentage
+of parsing time spent recognising log lines}{Percentage of parsing time
+spent recognising log lines graph}
+
+\showtable{build/include-percentage-time-spent-recognising-log-lines-table}{Percentage
+of parsing time spent recognising log lines}{Percentage of parsing time
+spent recognising log lines table}
 

### 2009-04-10 doc/logparser-results.tex
Started introduction.

@@
 
-XXX NEED AN INTRODUCTION\@.
+This chapter evaluates \parsername{} on two criteria: efficiency and
+coverage.  \parsername{} is intended to be used in a production
+environment, and must be capable of parsing log files generated by a high
+volume mail server in a reasonable time period.  Accurate but slow parsing
+is preferable to sloppy but quick parsing, so the goal of better parser
+efficiency must be balanced against the requirement for precise and correct
+processing of log files.
+
+The performance evaluation begins by describing the characteristics of the
+mail server that produced the log files used to test and evaluate the
+parser's performance, and also the computer that the tests were run on.
+How the parser scales as the size of log files increases is described next,
+with an explanation of why \parsername{} has better performance with the
+larger log files in the group.  The effect of using different rule
+orderings is explored, and the use of optimal rule ordering is compared to
+use of an oracle that allows the parser to use only one rule when
+recognising each log line.  When \parsername{} is used by other mail
+administrators they will need to extend the ruleset to parse their own log
+lines, so the next section addresses the question of how parser performance
+is affected as the number of rules in the ruleset rises.  The simple
+optimisation of caching the results of compiling regexes, and the huge
+impact it has on parser efficiency, is the penultimate topic to be covered.
+The performance evaluation concludes by examining where the parsing time is
+spent: recognition of log lines or their subsequent processing?
+
+XXX INTRODUCE COVERAGE\@.
 
@@
 received per day: statistics} show that, as expected, far more mails are
-received on weekdays than at weekends.  The mail server in question is a
-production mail server handling mail for a university department; the
-benefit of using this server is that its log files exhibit the
-idiosyncrasies and peculiarities a mail server in the wild must deal with,
-but the downside is that significantly altering the configuration to
-accommodate this project is not an option.  Median log file size is 50MB,
-containing 285,000 log lines; large scale mail servers would have much
-larger log files.  Note that \graphref{Mails received per day} and
-\tableref{Number of mails received per day: statistics} show the number of
-mails received by \acronym{SMTP} only, and the mail loops noticeable in
-later graphs do not contribute to these figures.
+received on weekdays than at weekends.  Note that these figures count mails
+received by \acronym{SMTP} only, and the mail loops noticeable in later
+graphs are not included.  The mail server in question is a production mail
+server handling mail for a university department; the benefit of using this
+server is that its log files exhibit the idiosyncrasies and peculiarities a
+mail server in the wild must deal with, but the downside is that
+significantly altering the configuration to accommodate this project is not
+an option.  Median log file size is 50MB, containing 285,000 log lines;
+large scale mail servers would have much larger log files.
 
@@
         CPU         & One dual core 2.40GHz Intel\textregistered{}
-                        Core\texttrademark{}2 CPU,                      \\
+                        Core\texttrademark{}~2 CPU,                     \\
                     & with 32KB L1 cache and 4MB L2 cache.              \\
@@
 \input{build/include-insert-results-duration}, with mean throughput of
-\input{build/include-insert-results-throughput}.  Parsing makes up only
+\input{build/include-insert-results-throughput} --- parsing makes up only
 \input{build/include-skip-inserting-as-percentage-of-inserting} of the
-execution time when saving results to the database, so a 50\% improvement
-in the speed of parsing would result in only a
-\input{build/include-skip-inserting-with-50-percent-improvement--as-percentage-of-inserting}
-improvement in speed overall.
+execution time when saving results to the database.
 
@@
 
-\subsection{Summary}
-
-XXX REWRITE WHEN THIS CHAPTER IS FINISHED\@.
-
-This section began with parser scalability, showing the linear relationship
-between parsing time and input size.  Demonstrates the effect of rule
-ordering on parsing time, and the unexpected consequences of specific
-inputs.  The necessity of caching compiled regexes is attested to by the
-third group of graphs, where the difference between caching and not caching
-is staggering.  The penultimate section contains a breakdown of the rule
-hits accumulated during a single test run of the parser.  Miscellaneous
-graphs expected to be useful are collected in the final section.
-
 \section{Coverage}
@@
 algorithm, and it can also be much harder to notice a deficiency.
-
-\newpage{} % XXX
 
@@
 other log files).
-
-\newpage{} % XXX
 
@@
 
+XXX REWRITE WHEN THIS CHAPTER IS FINISHED\@.
+
+This section began with parser scalability, showing the linear relationship
+between parsing time and input size.  Demonstrates the effect of rule
+ordering on parsing time, and the unexpected consequences of specific
+inputs.  The necessity of caching compiled regexes is attested to by the
+third group of graphs, where the difference between caching and not caching
+is staggering.  The penultimate section contains a breakdown of the rule
+hits accumulated during a single test run of the parser.  Miscellaneous
+graphs expected to be useful are collected in the final section.
+
 XXX REWRITE COVERAGE SUMMARY IF NECESSARY\@.

### 2009-04-10 doc/logparser-implementation.tex
Corrections from proofreading efficiency sections.

@@
     \item The number of rules tried when recognising log lines
-        (\graphref{Average number of rules tried per log line}); the
-        framework may need to try multiple rules for a given log line
-        before it finds one that recognises it.
+        (\graphref{Mean number of rules tried per log line}); the framework
+        may need to try multiple rules for a given log line before it finds
+        one that recognises it.
 
     \item The number of log lines and rules tried for each Postfix
-        component (\graphref{Average number of rules tried per log line for
+        component (\graphref{Mean number of rules tried per log line for
         each Postfix component}).

### 2009-04-10 doc/logparser-results.tex
Corrections from proofreading efficiency sections.

@@
 parse large log files.  The server that generated the log files used in
-testing this parser accepts approximately 10,000 mails for 700 users per
-day; \graphref{Mails received per day} and \tableref{Number of mails
+testing this parser accepts approximately 10,000 mails for 700 users each
+weekday; \graphref{Mails received per day} and \tableref{Number of mails
 received per day: statistics} show that, as expected, far more mails are
@@
 received by \acronym{SMTP} only, and the mail loops noticeable in later
-graphs are not included.  The mail server in question is a production mail
-server handling mail for a university department; the benefit of using this
-server is that its log files exhibit the idiosyncrasies and peculiarities a
-mail server in the wild must deal with, but the downside is that
-significantly altering the configuration to accommodate this project is not
-an option.  Median log file size is 50MB, containing 285,000 log lines;
-large scale mail servers would have much larger log files.
+graphs are not included.  Median log file size is 50MB, containing 285,000
+log lines; large scale mail servers would have much larger log files.  The
+mail server in question is a production mail server handling mail for a
+university department; the benefit of using this server is that its log
+files exhibit the idiosyncrasies and peculiarities a mail server in the
+wild must deal with, but the downside is that significantly altering the
+configuration to accommodate this project is not an option.
 
@@
 log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
-parsed 10 times, and the times from the 10 runs averaged.  The computer
-used for test runs was a Dell Optiplex 745, shown in \tableref{Details of
-the computer used to generate statistics}; it was dedicated to the task of
-gathering statistics from test runs, and was not used for any other purpose
+parsed 10 times, and the mean parsing time used.  The computer used for
+test runs was a Dell Optiplex 745, shown in \tableref{Details of the
+computer used to generate statistics}; it was dedicated to the task of
+gathering statistics from test runs, and did not run any other programs
 while test runs were ongoing.  Saving results to the database was disabled
@@
 \numberOFlogFILES{} log files in one run without saving results to the
-database took \input{build/include-full-run-duration}, mean throughput was
-\input{build/include-full-run-throughput}; median throughput when parsing
-files separately was \input{build/include-median-throughput-MB}
-(\input{build/include-median-throughput-log-lines}) parsed per minute.  In
-contrast, when saving results to the database, parsing all
-\numberOFlogFILES{} log files took
+database took \input{build/include-full-run-duration}, with mean throughput
+of \input{build/include-full-run-throughput}.  In contrast, when saving
+results to the database, parsing all \numberOFlogFILES{} log files took
 \input{build/include-insert-results-duration}, with mean throughput of
@@
 \subsection{Architecture Scalability: Input Size}
+
+XXX THIS SECTION DOES NOT SHOW ANY VARIETY IN INPUT SIZE, SO DOES NOT SHOW
+SCALABILITY\@.  MAYBE I SHOULD RENAME IT TO ``Characteristics Of The Input
+Log Files'' OR SOMETHING\@?  THE GRAPHS SHOW THAT LOGS HAVE CONSISTENT
+CONTENT EXCEPT WHEN THERE IS A MAIL LOOP\@.
 
@@
 ratios increase (i.e.\ improve) for log files 22 \& 62--68, despite their
-larger than average size.  Both groups of log files are much larger than
+larger than usual size.  Both groups of log files are much larger than
 usual due to a mail loop caused by a user who set up mail forwarding
@@
 fewer associated rules than the Postfix components whose log lines normally
-make up the bulk of each log file, so the average number of rules tried per
-log line reduces and so does the average parsing time per log line.
+make up the bulk of each log file, so the mean number of rules tried per
+log line reduces and so does the mean parsing time per log line.
 \Graphref{Number of rules per Postfix component} shows the number of rules
-per Postfix component, \graphref{Average number of rules tried per log
-line} shows the drop in the average number of rules tried per log line for
-log files containing a mail loop, and \graphref{Average number of rules
-tried per log line for each Postfix component} shows the average number of
-rules tried per log line for each Postfix component.
+per Postfix component, \graphref{Mean number of rules tried per log line}
+shows the drop in the mean number of rules tried per log line for log files
+containing a mail loop, and \graphref{Mean number of rules tried per log
+line for each Postfix component} shows the mean number of rules tried per
+log line for each Postfix component.
 
@@
 \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of file size
-and number of log lines to parsing time}{parsing time vs file size vs
-number of log lines factor}
+and number of log lines to parsing time (higher is better)}{parsing time vs
+file size vs number of log lines factor}
 
-\showgraph{build/graph-average-number-of-rules-tried-per-log-line}{Average
-number of rules tried per log line}{Average number of rules tried per log
+\showgraph{build/graph-average-number-of-rules-tried-per-log-line}{Mean
+number of rules tried per log line}{Mean number of rules tried per log
 line}
 
-\showgraph{build/graph-average-number-of-rules-tried-per-program}{Average
-number of rules tried per log line for each Postfix component}{Average
-number of rules tried per log line for each Postfix component}
+\showgraph{build/graph-average-number-of-rules-tried-per-program}{Mean
+number of rules tried per log line for each Postfix component}{Mean number
+of rules tried per log line for each Postfix component}
 
@@
 \parsername{} has \numberOFrules{} rules: the top 10\% recognise
-\input{build/include-top-ten-hits}\% of the log lines, with the remaining
-log lines split across the other 90\% of the rules (as shown in
-\graphref{rule hits graph}).  Assuming that the distribution of log lines
-is reasonably consistent over time, \parsernames{} efficiency should
-benefit from trying rules that recognise log lines more frequently before
-those rules that recognise log lines less frequently.  To test this
-hypothesis, three full test runs were performed with different rule
-orderings:
+\input{build/include-top-ten-hits}\% of the log lines in the
+\numberOFlogFILES{} log files, with the remaining log lines split across
+the other 90\% of the rules, as shown in \graphref{rule hits graph}.
+Assuming that the distribution of log lines is reasonably consistent over
+time, \parsernames{} efficiency should benefit from trying rules that
+recognise log lines more frequently before those rules that recognise log
+lines less frequently.  To test this hypothesis, three full test runs were
+performed with different rule orderings:
 
@@
 
-    \item [Reverse] Hypothetically the worst order: the rules that
-        recognise log lines most frequently will be tried last.
+    \item [Reverse] Hypothetically the worst order: rules that recognise
+        log lines most frequently will be tried last.
 
@@
 each rule is zero for the first log file, so optimal and reverse orderings
-are identical.  For the first log file, shuffled ordering is most
-efficient, but that is accidental and cannot be relied upon.
+produce identical rule orderings.  For the first log file, shuffled
+ordering is the most efficient of the three, but that is accidental and
+cannot be relied upon.
 
@@
 recognise each log line.  Such an oracle would give perfect performance,
-because only one rule needs to be used to recognise each log line.
+because only one rule would need to be used to recognise each log line.
 \parsername{} can save a list showing which rule recognised each log line,
 and use that list to simulate an oracle and improve parsing speed the
-second time a log file is parsed.  This does not provide a practical
-benefit, but it does provide a means to evaluate the performance of optimal
-rule ordering in comparison to an oracle.
+\textit{second\/} time a log file is parsed.  This does not provide a
+practical benefit, but it does provide a means to evaluate the performance
+of optimal rule ordering in comparison to an oracle.
 
-\Graphref{Parsing time of perfect best oracle and optimal ordering relative
-to shuffled ordering graph} shows the parsing times of the perfect best
-oracle and the optimal ordering, relative to shuffled ordering;
-\tableref{Parsing time of perfect best oracle and optimal ordering relative
-to shuffled ordering table} shows mean and standard deviation.  As
-expected, the perfect best oracle is more efficient than optimal ordering.
-\Graphref{Percentage increase in parsing time when using optimal
-ordering instead of perfect best oracle graph} shows the percentage
-increase in parsing time when using optimal ordering instead of the perfect
-best oracle; mean and standard deviation are given for different groups of
-log files in \tableref{Percentage increase in parsing time when using
-optimal ordering instead of perfect best oracle table}.
+\Graphref{Parsing time of oracle and optimal ordering relative to shuffled
+ordering graph} shows the parsing times of the oracle and the optimal
+ordering, relative to shuffled ordering; \tableref{Parsing time of oracle
+and optimal ordering relative to shuffled ordering table} shows mean and
+standard deviation.  As expected, the oracle is more efficient than optimal
+ordering, but not by much.  \Graphref{Percentage increase in parsing time
+when using optimal ordering instead of oracle graph} shows the percentage
+increase in parsing time when using optimal ordering instead of the oracle,
+with mean and standard deviation in \tableref{Percentage increase in
+parsing time when using optimal ordering instead of oracle table}.
 
-Once again, the difference between the perfect best oracle and optimal
-ordering is at its lowest when parsing log files resulting from a mail loop
-(log files 22 \& 62--68), because the average number of rules tried per log
-line is lower (see \graphref{Average number of rules tried per log line}).
-The performance of optimal ordering when parsing the first log file is much
-worse than the remainder of the log files, because the hits field of every
-rule is zero, so optimal ordering does not provide any benefit for that log
-file.
+Once again, the difference between the oracle and optimal ordering is at
+its lowest when parsing log files resulting from a mail loop (log files 22
+\& 62--68), because the mean number of rules tried per log line is lower
+(see \graphref{Mean number of rules tried per log line}).  When parsing the
+first log file, the performance of optimal ordering relative to the oracle
+is much worse than for the remainder of the log files, because for the
+first log file the hits field of every rule is zero, so optimal ordering
+does not provide any benefit for that log file; the oracle, in contrast, is
+flawless for every log file.
 
 Optimal ordering proves to be quite efficient: \tableref{Percentage
-increase in parsing time when using optimal ordering instead of perfect
-best oracle table} shows that optimal ordering is less than
+increase in parsing time when using optimal ordering instead of oracle
+table} shows that optimal ordering is less than
 \input{build/include-perfect-best-vs-optimal-mean} slower than parsing
@@
 \showgraph{build/graph-perfect-best-and-optimal-vs-shuffled}{Parsing time
-of perfect best oracle and optimal ordering relative to shuffled
-ordering}{Parsing time of perfect best oracle and optimal ordering relative
-to shuffled ordering graph}
+of oracle and optimal ordering relative to shuffled ordering}{Parsing time
+of oracle and optimal ordering relative to shuffled ordering graph}
 
 \showtable{build/include-perfect-best-and-optimal-vs-shuffle}{Parsing time
-of perfect best oracle and optimal ordering relative to shuffled
-ordering}{Parsing time of perfect best oracle and optimal ordering relative
-to shuffled ordering table}
+of oracle and optimal ordering relative to shuffled ordering}{Parsing time
+of oracle and optimal ordering relative to shuffled ordering table}
 
 \showgraph{build/graph-perfect-best-vs-optimal}{Percentage increase in
-parsing time when using optimal ordering instead of perfect best
-oracle}{Percentage increase in parsing time when using optimal ordering
-instead of perfect best oracle graph}
+parsing time when using optimal ordering instead of oracle}{Percentage
+increase in parsing time when using optimal ordering instead of oracle
+graph}
 
 \showtable{build/include-perfect-best-vs-optimal-stddev}{Percentage
-increase in parsing time when using optimal ordering instead of perfect
-best oracle}{Percentage increase in parsing time when using optimal
-ordering instead of perfect best oracle table}
+increase in parsing time when using optimal ordering instead of
+oracle}{Percentage increase in parsing time when using optimal ordering
+instead of oracle table}
 
@@
 
-\subsection{Scalability When The Number Of Rules Rises}
+\subsection{Scalability As The Number Of Rules Rises}
 
@@
 minimum number of rules required to parse the \numberOFlogFILES{} log files
-used when generating the data in this chapter is \numberOFrulesMINIMUM{}
-rules, \numberOFrulesMINIMUMpercentage{} of the full ruleset.  The full
-ruleset is larger because \parsername{} is tested with
+is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
+ruleset.  The full ruleset is larger because \parsername{} is tested with
 \numberOFlogFILESall{} log files; testing with more log files increases the
@@
 the full ruleset: the percentage parsing time increase when using the full
-ruleset instead of the minimal ruleset for optimal, shuffled and reverse
-orderings is shown in \graphref{Percentage parsing time increase of maximum
-ruleset over minimum ruleset}; mean and standard deviation are shown in
-\tableref{Percentage parsing time increase of maximum ruleset over minimum
-ruleset table}.
+ruleset instead of the minimal ruleset for optimal, shuffled, and reverse
+orderings is shown in \graphref{Percentage parsing time increase when using
+the maximum ruleset instead of the minimum ruleset}, with mean and standard
+deviation in \tableref{Percentage parsing time increase when using the
+maximum ruleset instead of the minimum ruleset table}.
 
@@
 optimal ordering shows very little change.  Log files 22 \& 62--68 show
-much smaller increases than other log files do, because the majority of log
-lines in those log files are produced by Postfix components with a small
-number of rules, so removing unnecessary rules has little effect on the
-total number of rules used; \tableref{Number of rules per Postfix component
-in the maximum and minimum rulesets} shows the number of rules per Postfix
-component for each ruleset.  Once again, for the first log file, optimal
-and reverse orderings have identical performance, because the hits field of
-every rule is zero.
+much smaller increases in parsing time than other log files do, because the
+majority of log lines in those log files are produced by Postfix components
+with a small number of rules, so removing unnecessary rules has little
+effect on the total number of rules used; \tableref{Number of rules per
+Postfix component in the maximum and minimum rulesets} shows the number of
+rules per Postfix component for each ruleset.  Once again, for the first
+log file, optimal and reverse orderings have identical performance, because
+the hits field of every rule is zero.
 
@@
 for a \numberOFrulesMAXIMUMpercentage{} increase in the number of rules.
-These results show that the architecture scales very well as the number of
-rules increases, and that optimally sorting the rules is an important
-optimisation contributing to this scalability.
+These results show that both the architecture and \parsername{} scale very
+well as the number of rules increases, and that optimally sorting the rules
+is an important optimisation contributing to this scalability.
 
 \showgraph{build/graph-full-ruleset-vs-minimum-ruleset}{Percentage parsing
-time increase of maximum ruleset over minimum ruleset}{Percentage parsing
-time increase of maximum ruleset over minimum ruleset}
+time increase when using the maximum ruleset instead of the minimum
+ruleset}{Percentage parsing time increase when using the maximum ruleset
+instead of the minimum ruleset}
 
-\showtable{build/include-full-ruleset-vs-minimum-ruleset}{Percentage parsing
-time increase of maximum ruleset over minimum ruleset}{Percentage parsing
-time increase of maximum ruleset over minimum ruleset table}
+\showtable{build/include-full-ruleset-vs-minimum-ruleset}{Percentage
+parsing time increase when using the maximum ruleset instead of the minimum
+ruleset}{Percentage parsing time increase when using the maximum ruleset
+instead of the minimum ruleset table}
 
@@
 rule's hits field is zero, so optimal ordering is less efficient than
-usual, and the average number of rules tried for each log line will be
-higher than usual.
+usual, and the mean number of rules tried for each log line will be higher
+than usual.
 
@@
 The optimisations described in this chapter have optimised the process of
-recognising log lines, but have not affected the actions at all.  There are
+recognising log lines, but have not optimised actions at all.  There are
 two reasons for this:
@@
         and standard deviation shown in \tableref{Percentage of parsing
-        time spent recognising log lines table}; for normal log files
+        time spent recognising log lines table}; for normal log files,
         \input{build/include-percentage-time-spent-recognising-log-lines-mean}
@@
         actions could not possibly provide as big a performance increase as
-        optimally ordering rules.
+        optimally ordering rules.  Similarly, caching compiled regexes
+        provides an
+        \input{build/include-cached-regexes-vs-discarded-regexes-mean-reduction}
+        reduction in parsing time; not invoking actions at all reduces the
+        most optimised parsing time by less than half that.  If processing
+        of recognised log lines was optimised to 1\% of its original
+        parsing time, it would be slightly better than optimal ordering,
+        but it would be vastly harder to implement; optimising actions
+        would not provide a sufficient reduction in parsing time to justify
+        the amount of effort required.
 

### 2009-04-11 doc/logparser-results.tex
Corrections to the first coverage section.

@@
 
+XXX IS THERE A BETTER WORD THAN ``coverage''?
+
 XXX I THINK THIS SECTION NEEDS SUBSTANTIAL WORK, IF NOT A RE-WRITE\@; START
@@
 
-The discussion of the parser's coverage of Postfix log files is separated
-into two parts: log lines covered and mails covered.  The first is
-important because the parser should handle all (relevant) log lines it is
-given; the second is equally important because the parser must properly
-deal with every mail if it is to be useful.  Improving the former is
-less intrusive, as it just requires new rules to be written; improving the
-latter is much more intrusive as it requires changes to the parser
-algorithm, and it can also be much harder to notice a deficiency.
+The discussion of \parsernames{} coverage of Postfix log files is separated
+into two parts: log lines correctly recognised, and mail delivery attempts
+correctly understood --- the former is a requirement for the latter to be
+achieved.  Correctly understanding and reconstructing every mail delivery
+attempt, whether it was successful or not, is important so that the
+information in the database is correct and complete.  Improving the
+proportion of log lines correctly recognised is the less difficult of the
+two, because usually it just requires new rules to be written or existing
+rules to be changed.  Improving the proportion of correctly understood and
+reconstructed mail delivery attempts is more difficult and intrusive,
+because it requires adding or changing actions, and it can be much harder
+to realise that a deficiency exists and needs to be addressed.
 
-\subsection{Log Lines Covered}
+\subsection{Log Lines Correctly Recognised}
 
@@
 
-    \item Check if there are any rules for the Postfix component that
-        produced the log line; if not then skip the log line.
+    \squeezeitems{}
+
+    \item Skip the log line if the ruleset does not contain any rules for
+        the Postfix component that produced the log line.
 
     \item Try each rule until a recognising rule is found; if the log line
-        is not recognised, issue a warning and move to the next log line.
+        is not recognised, issue a warning and move on to the next log
+        line.
 
-    \item Execute the action specified by the rule.
+    \item Invoke the action specified by the recognising rule.
 
@@
 
-Full coverage of log lines requires the following:
+Correctly recognising all log lines requires that each Postfix component
+whose log lines are of interest must have at least one rule, or its log
+lines will be silently skipped; in the extreme case of an empty ruleset the
+parser would skip every log line.  \parsername{} skips those log lines
+because there may be any number of log lines from other programs
+intermingled in the log file, and there are some Postfix components that do
+not produce any log lines of interest.  There must be a rule to recognise
+each log line variant produced by each Postfix component; if a log line is
+not recognised the parser will issue a warning, to inform the user that
+they need to extend their ruleset.  Currently, \parsername{} does not parse
+log lines from non-Postfix programs, e.g.\ Amavisd-new or SpamAssassin; it
+could easily be extended to do so, if a method could be developed to
+correctly associate such log lines with existing state table entries.  Each
+rule's regex should be as specific and precise as possible, to ensure
+accurate parsing: a rule with a regex that matches zero or more of any
+character will recognise every log line, but not in a meaningful way.  Most
+log lines contain fixed strings, so this is not a problem in practice.
 
-\begin{enumerate}
-
-    \item Each Postfix component whose log lines are of interest must have
-        at least one rule or its log lines will be silently skipped; in the
-        extreme case of zero rules the parser would happily skip every log
-        line.  There may be any number of log lines from other programs
-        intermingled in the log file, and there are some Postfix programs
-        that do not produce any log lines of interest.
-
-    \item There must be a rule to recognise each different log line
-        produced by each program; if a log line is not successfully
-        recognised the parser will issue a warning.  Rules should be as
-        specific and tightly bound as possible to ensure accurate
-        parsing:\footnote{A rule with a regex that matches zero or more of
-        any character will successfully recognise every log line, but not
-        in a meaningful way.} most log lines contain fixed strings and have
-        a rigid pattern, so this is not a problem.
-
-    \item The appropriate action to take --- discussed in
-        \sectionref{mails-covered}.
-
-\end{enumerate}
-
-Full coverage of log lines is easy to achieve yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files (at the time
-of writing the parser is tested with \numberOFrules{} rules, fully parsing
-\numberOFlogFILESall{} contiguous log files from Postfix 2.2 through to
-Postfix 2.5), and new rules are easy to add.  Maintaining full coverage is
-hard because other servers have different restrictions with custom
-messages, \acronym{DNSBL} messages change over time, major releases of
-Postfix change warning messages (usually adding more information), etc.,\
-so over time the log lines drift and change.  \Graphref{rule hits graph}
-shows the number of hits for each rule over all \numberOFlogFILES{} log
-files.  The number of hits per rule is quite unevenly spread, resembling a
-Power Law distribution --- it is obvious that a small number of rules
-recognise the vast majority of the log lines, and more than half the rules
-recognise fewer than 100 log lines.
-
-Warnings are issued for any log lines that are not parsed; no warnings are
-issued for unparsed log lines while testing with the \numberOFlogFILES{}
-test log files, so it can be safely concluded that there are zero false
-negatives.  False positives are harder to quantify: short of examining each
-of the 60,721,709 log lines and determining which regex parsed it, there
-is no way to be sure that every line was parsed by the correct regex,
-making it impossible to quantify the false positive rate; however a random
-sample of 6,039 log lines was parsed and the results checked manually to
-ensure that the correct regex parsed each log line.\footnote{Each log
-line was examined and the correct regex identified from the
-\numberOFrules{} rules in the database; the correct regex was then
-compared to the regex that was used by the parser.}  The sample was
-generated by running the following command:
-
-\verb!    perl -e 'print if (rand 1 < 0.0001)' -n LOG_FILES!
-
-\noindent{}to randomly extract roughly one line in every 10,000.  Although
-on initial appearances exercising only 36 rules (from a total of
-\numberOFrules{}) when parsing 6039 log lines seems quite low, after
-examining \graphref{rule hits graph} it becomes apparent that such
-a low hit rate is to be expected; the reader should also bear in mind that
-even when parsing all \numberOFlogFILES{} log files not all the rules are
-exercised (some of the rules are for parsing log lines that only appear in
-other log files).
+Full coverage of log lines is easy to achieve, yet hard to maintain.  It is
+easy to achieve full coverage for a limited set of log files; \parsername{}
+has \numberOFrules{} rules, fully parsing \numberOFlogFILESall{} log files
+from Postfix 2.2 through to Postfix 2.5.  Maintaining full coverage is hard
+because log lines change over time, e.g.\ administrators add restrictions
+with custom messages, \acronym{DNSBL} messages change, or major releases of
+Postfix change log lines (usually adding more information).  Warnings are
+issued for any log lines that are not recognised; no warnings are issued
+for unrecognised log lines while parsing the \numberOFlogFILES{} log files,
+so it can be safely concluded that there are zero false negatives.  False
+positives are harder to quantify: short of examining each of the 60,721,709
+log lines and determining which rule recognised it, there is no way to be
+sure that every log line was recognised by the correct rule, making it
+impossible to quantify the false positive rate.  However, a random sample
+of 6,039 log lines was parsed, and the results manually verified by
+inspection to ensure that the correct rule recognised each log line.  The
+sample was generated by running the command: \texttt{perl -n -e
+\singlequote{}print if (rand 1 < 0.0001)\singlequote{} LOG\_FILES} to
+randomly extract roughly one log line in every 10,000 (it actually
+extracted 0.00994\% instead of 0.001\%).  Each log line was examined and
+the correct rule identified from the \numberOFrules{} rules in the
+database; the correct rule was then compared to the rule that recognised
+the log line when parsing.  The sample results contained zero false
+positives, and this check has been automated to ensure continued accuracy.
+Based on this, the author is confident that zero false positives occur when
+parsing the \numberOFlogFILES{} log files.  On initial appearances,
+exercising only 36 rules from a total of \numberOFrules{} when parsing
+6,039 log lines seems low, but after examining \graphref{rule hits graph}
+it becomes apparent that parsing using such a low number of rules is to be
+expected.  The reader should also bear in mind that even when parsing all
+\numberOFlogFILES{} log files, only \numberOFrulesMINIMUM{} are used,
+because some of the rules are for recognising log lines that only appear in
+other log files.
 

### 2009-04-11 doc/logparser-results.tex
Corrections/improvements after proofreading second coverage section.

@@
 impossible to quantify the false positive rate.  However, a random sample
-of 6,039 log lines was parsed, and the results manually verified by
+of 6039 log lines was parsed, and the results manually verified by
 inspection to ensure that the correct rule recognised each log line.  The
@@
 randomly extract roughly one log line in every 10,000 (it actually
-extracted 0.00994\% instead of 0.001\%).  Each log line was examined and
-the correct rule identified from the \numberOFrules{} rules in the
-database; the correct rule was then compared to the rule that recognised
-the log line when parsing.  The sample results contained zero false
-positives, and this check has been automated to ensure continued accuracy.
-Based on this, the author is confident that zero false positives occur when
-parsing the \numberOFlogFILES{} log files.  On initial appearances,
-exercising only 36 rules from a total of \numberOFrules{} when parsing
-6,039 log lines seems low, but after examining \graphref{rule hits graph}
-it becomes apparent that parsing using such a low number of rules is to be
-expected.  The reader should also bear in mind that even when parsing all
+extracted 0.00994\% instead of 0.01\%).  Each log line was examined and the
+correct rule identified from the \numberOFrules{} rules in the database;
+the correct rule was then compared to the rule that recognised the log line
+when parsing.  The sample results contained zero false positives, and this
+check has been automated to ensure continued accuracy.  Based on this, the
+author is confident that zero false positives occur when parsing the
+\numberOFlogFILES{} log files.  On initial appearances, exercising only 36
+rules from a total of \numberOFrules{} when parsing 6039 log lines seems
+low, but after examining \graphref{rule hits graph} it becomes apparent
+that parsing using such a low number of rules is to be expected.  The
+reader should also bear in mind that even when parsing all
 \numberOFlogFILES{} log files, only \numberOFrulesMINIMUM{} are used,
@@
 
-\subsection{Mails Covered}
+\subsection{Mail Delivery Attempts Correctly Understood And Reconstructed}
 
@@
 
-Coverage of mails is much more difficult to determine accurately than
-coverage of log lines.  The parser can dump its state tables in a human
-readable form; examining these tables with reference to the log files is
-the best way to detect mails that were not handled properly (many of the
-complications discussed in \sectionref{complications} were detected in this
-way).  The parser issues warnings when it detects any errors, some of which
-may alert the user to a problem, e.g.\ when a queueid is reused before the
-previous mail is fully dealt with, when a queueid or \acronym{pid} is not
-found in the state tables,\footnote{There will often be warnings about a
-missing queueid or \acronym{pid} in the first few hundred or thousand log
-lines because the earlier log lines for those connections or mails are in
-the previous log file; loading the saved state from the previous log file
-will solve this problem.} or when there are problems tracking a child mail
-(see \sectionref{tracking re-injected mail}).  There should be few or no
-warnings when parsing, and when finished parsing the state table should
-only contain entries for mails that had yet to be delivered when the log
-files ended, or were accepted before the log files began.
+The proportion of mail delivery attempts that are correctly understood and
+reconstructed is much more difficult to determine accurately than the
+proportion of log lines that are correctly recognised.  The parser can dump
+its state tables in a human readable form; examining those tables with
+reference to the log files is the best way to detect mails that were not
+handled properly (many of the complications discussed in
+\sectionref{complications} were detected in this way).  \parsername{}
+issues warnings when it detects any errors or discrepancies, alerting the
+user to the problem, e.g.\ when a queueid is reused but the previous mail
+remains in the state tables, when a queueid or \acronym{pid} is not found
+in the state tables, or when a mail does not include sufficient data to
+satisfy the database schema.  There should be few or no warnings during
+parsing, and when finished parsing the state tables should only contain
+entries for mails that have log lines in subsequent log files.  There will
+often be warnings about a missing queueid or \acronym{pid} in the first few
+thousand log lines, because the earlier log lines for those connections or
+mails are in a previous log file; loading saved state tables from the
+previous log file will solve this problem.
 
-At the time of writing the parser is being tested with \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because the parser errs on
-the side of producing more warnings rather than fewer, those 5 warnings
-represent 3 instances of 1 problem: 3 connections started before the first
-log file, so their initial log lines are missing, leading to warnings when
-their remaining log lines are parsed.
+The data used in this chapter is generated by parsing \numberOFlogFILES{}
+log files.  There are 5 warnings produced, but because \parsername{} errs
+on the side of producing more warnings rather than fewer, those 5 warnings
+represent 3 instances of 1 problem: 3 connections that started before the
+first log file, so their initial log lines are missing, leading to warnings
+when their remaining log lines are parsed.  There are no false positives.
 
-The state tables contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally all they should contain are mails that
-are awaiting delivery after the period covered by the log files, though
-they may also contain mails whose initial entries are not contained in the
-log files.  Any other entries are evidence of a failure in parsing or an
-aberration in the log files.  After parsing the \numberOFlogFILES{} test
-log files the state tables contain 18 entries, breaking down into:
+The state tables will contain entries for mails not yet delivered when the
+parser finishes execution.  Ideally, that is all they will contain, though
+they may also contain mails whose initial log lines are not contained in
+the log files.  Any other entries in the state tables are evidence of
+either a failure in parsing, or an aberration in the log files.  After
+parsing the \numberOFlogFILES{} log files, the state tables contain 18
+entries, breaking down into:
 
 \begin{itemize}
+
+    \squeezeitems{}
 
@@
     \item 9 mails whose initial log lines were not present in the log
-        files.
+        files.  XXX WHY WERE THERE NO WARNINGS ABOUT THESE\@?
 
@@
 the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced or delivered is infeasible.  There is considerable circumstantial
-evidence that the false positive rate is quite low:
+bounced, or delivered is infeasible.  There is considerable evidence that
+the false positive rate is extremely low:
 
@@
 
-    \item The parser is quite verbose when complaining about known problems
-        (e.g.\ if a mail is tracked twice as described in
-        \sectionref{tracking re-injected mail}), and no such warnings are
-        produced during the test runs.
+    \item The parser is quite verbose when complaining about known
+        problems, e.g.\ if a mail is missing required data, and no such
+        warnings are produced during the test runs.
 
-    \item Queueids and \glspl{pid} naturally group together log lines
-        belonging to one mail or connection respectively; it is extremely
-        unlikely that a log line would be associated with the wrong
-        connection.
+    \item Queueids and \glspl{pid} naturally identify log lines belonging
+        to one mail or connection respectively; it is extremely unlikely
+        that a log line would be associated with the wrong connection.
 
     \item When dealing with the complications described in
-        \sectionref{complications} the solutions are as specific and
+        \sectionref{complications}, the solutions are as specific and
         restrictive as possible, with the goal of minimising the number of
-        false positives.  In addition the solution to the \textit{Out of
-        order log files\/} complication described in \sectionref{out of
-        order log lines} imposes conditions that each reassembled mail must
-        comply with to be acceptable.
+        false positives.  In addition, the solution to the complication
+        described in \sectionref{out of order log lines} imposes conditions
+        that each reconstructed mail must comply with to be acceptable.
 
-    \item Every effort has been made while developing to make the parser as
-        precise, demanding, and particular as possible.
+    \item Every effort has been made to make \parsername{} as precise,
+        demanding, and particular as possible.
 
@@
 
-Whereas verifying by inspection that the parser correctly deals with all
-60,721,709 log lines in the test log files is infeasible, verifying a
-subset of those log files is a tractable, if extremely time consuming,
-task.  A sample of log lines was obtained by randomly selecting a log file:
+%\verb!perl -Mstrict -Mwarnings -MList::Util=shuffle -e! \newline{}
+%\verb!     '@ARGV = [shuffle(@ARGV)]->[0];!             \newline{}
+%\verb!      while (<>) {!                               \newline{}
+%\verb!          if (rand 1 < 0.0001) {!                 \newline{}
+%\verb!              foreach my $i (1 .!.  6000) {!        \newline{}
+%\verb!                  print scalar <>;!               \newline{}
+%\verb!              }!                                  \newline{}
+%\verb!              exit;!                              \newline{}
+%\verb!          }!                                      \newline{}
+%\verb!      }' LOG_FILES!
 
-\verb!    perl -Mstrict -Mwarnings -MList::Util=shuffle \!\newline{}
-\verb!            -e 'print [shuffle(@ARGV)]->[0];'!
+Verifying by inspection that the parser correctly deals with all 60,721,709
+log lines in the \numberOFlogFILES{} log files is infeasible, but verifying
+the parsing of a sample from those log lines is a tractable, if extremely
+time consuming, task.  A sample of log lines was obtained by randomly
+selecting a log file, and then randomly selecting a block of 6000
+contiguous log lines from it (0.00988\% of the total number of log lines).
+It is important that the log lines are contiguous, so that all log lines
+are present for as many of the mail delivery attempts contained in the
+block as possible.  This log segment was parsed with all debugging options
+enabled, resulting in 167,448 lines of output.\footnote{A mean of 27.908
+lines of output per line of input; each connection has 30 debugging lines,
+plus 21 debugging lines per result.  Connections which have been cloned
+will have the cloned connection in their debugging output, plus another 33
+debugging lines.  Those numbers are approximate, and may vary $\pm{}$ 2.
+There is a linear relationship between the number of log lines and
+debugging lines: $33(connections) + 30(accepted~~mails) + 21(results)$.
+This formula is an approximation only, and has not been rigorously
+verified.}  All 167,448 lines were examined in conjunction with the log
+file segment and a dump of the resulting database, verifying that for each
+of the log lines \parsername{} used the correct rule and invoked the
+correct action, which in turn produced the correct result, and the correct
+data was inserted in the database.  The log file segment produced 4
+warnings, 10 mails remaining in the state tables, 1625 mail delivery
+attempts correctly entered in the database, zero false positives, and zero
+false negatives.
 
-The first 6000 log lines of this log file (roughly 0.01\% of the total
-number of log lines used in testing) was extracted:
-
-\verb!    sed -n -e '1,6000p' logfile > test-log-segment!
-
-It is important that the log lines used are contiguous so that all log
-entries are present for as many of the connections as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per line of
-input; each connection has 30 debugging lines, plus 21 debugging lines per
-result.  Connections which have been cloned will have the cloned connection
-in their debugging output, plus another 33 debugging lines.  Those numbers
-are approximate, and may vary $\pm{}$ 2.  There is a linear relationship
-between the number of log lines and debugging lines: $33(connections) +
-30(accepted~~mails) + 21(results)$.  This formula is an approximation only,
-and has not been rigorously verified.}  All 167,448 lines were examined in
-conjunction with the input log file and a dump of the resulting database,
-verifying that for each of the log lines the parser used the correct rule
-and executed the correct action, which in turn produced the correct result
-and inserted the correct data in the database.  The log file segment
-produced 4 warnings, 10 mails remaining in the state tables, and 1625
-connections correctly entered in the database.
-
-Given the circumstantial and experimental evidence detailed above, the
-author is confident that the false positive rate when reconstructing a mail
-is exceedingly low, if not approaching zero.
+Given the evidence detailed above, the author is confident that the false
+positive rate when reconstructing a mail delivery attempt from the
+\numberOFlogFILES{} log files is exceedingly low, if not zero.
 
@@
 further examination of the log files in developing a solution.
+
+Experience implementing the Postfix log parser shows that full input
+coverage is relatively easy to achieve with this architecture, and that
+with enough time and effort full understanding of the input is possible.
+Postfix log files would require substantial time and effort to correctly
+parse regardless of the architecture used; this architecture enables an
+iterative approach to be used (similar to Stepwise
+Refinement~\cite{stepwise-refinement}), as is practiced in many other
+software engineering disciplines.

### 2009-04-11 doc/logparser-conclusion.tex
Minor changes.
Corrections to the first coverage section.
Renamed Results chapter to Evaluation. Finished evaluation chapter introduction and summary.

@@
 XXX PULL CONTENT FROM THE PAPER\@.
+
+XXX INCLUDE A TIGHTLY FOCUSED FUTURE WORK SECTION\@.  PARSE LOG LINES FROM
+OTHER PROGRAMS IF A WAY CAN BE FOUND TO ASSOCIATE THEM WITH EXISTING DATA
+STRUCTURES\@.
 
@@
 
-\newpage
+Experience implementing \parsername{} shows that full input coverage is
+relatively easy to achieve with this architecture, and that with enough
+time and effort a full understanding of the input is possible.  Postfix log
+files would require substantial time and effort to correctly parse
+regardless of the architecture used; this architecture enables an iterative
+approach to be used (similar to Stepwise
+Refinement~\cite{stepwise-refinement}), as is practiced in many other
+software engineering disciplines.

### 2009-04-11 doc/logparser-implementation.tex
Renamed Results chapter to Evaluation. Finished evaluation chapter introduction and summary.

@@
 The framework collects data used when evaluating \parsernames{} efficiency
-for the results chapter (\sectionref{Results}); some of the techniques used
-to improve parsing speed can be turned off or altered to measure the effect
-they have.  In each case the data gathered with the optimisation disabled
-will be compared to the data gathered with it enabled, to quantify the
-benefit each optimisation provides.  Three sets of data are collected:
+for the evaluation chapter (\sectionref{Evaluation}); some of the
+techniques used to improve parsing speed can be turned off or altered to
+measure the effect they have.  In each case the data gathered with the
+optimisation disabled will be compared to the data gathered with it
+enabled, to quantify the benefit each optimisation provides.  Three sets of
+data are collected:
 
@@
         in the database is described at the beginning of
-        \sectionref{Results}.
+        \sectionref{Evaluation}.
 

### 2009-04-11 doc/logparser-introduction.tex
Renamed Results chapter to Evaluation. Finished evaluation chapter introduction and summary.

@@
 
-Chapter~\ref{Results} evaluates \parsernames{} efficiency, discussing the
-various optimisations implemented in the framework, and the effect they
+Chapter~\ref{Evaluation} evaluates \parsernames{} efficiency, discussing
+the various optimisations implemented in the framework, and the effect they
 have.  It also discusses the coverage achieved by \parsername{} over
@@
 Portions of chapters~\ref{introduction} and~\ref{parser
-architecture}--\ref{Results} have previously been published at a
+architecture}--\ref{Evaluation} have previously been published at a
 conference~\cite{sgai-2008}, and later reprinted in a

### 2009-04-11 doc/logparser-results.tex
Renamed Results chapter to Evaluation. Finished evaluation chapter introduction and summary.

@@
-\chapter{Results And Evaluation}
+\chapter{Evaluation}
 
-\label{Results}
+\label{Evaluation}
 
@@
 
-XXX INTRODUCE COVERAGE\@.
+The second criterion the parser is evaluated on is its coverage of Postfix
+log files.  This topic consists of two sections: what proportion of log
+lines are correctly recognised by the ruleset, and what proportion of mail
+delivery attempts are correctly understood and reconstructed by the
+actions?  The former is a requirement before the latter can be achieved,
+and the latter is important because the data provided by \parsername{} must
+be both complete and correct for it to be of use to others.
 
@@
 
-XXX IS THERE A BETTER WORD THAN ``coverage''?
-
-XXX I THINK THIS SECTION NEEDS SUBSTANTIAL WORK, IF NOT A RE-WRITE\@; START
-WITH THE PAPER'S CONTENT\@.
-
 \label{parsing coverage}
@@
     \item 9 mails whose initial log lines were not present in the log
-        files.  XXX WHY WERE THERE NO WARNINGS ABOUT THESE\@?
+        files.  There were no warnings for these mails because they were
+        resemble child mails waiting to be tracked with a parent, see
+        \sectionref{Re-injected mails} for details.
 
@@
 
-XXX EXTEND TO SUMMARISE PARSER EFFICIENCY TOO\@.
+This chapter evaluated \parsername{} on two criteria: efficiency, and
+coverage of Postfix log files.  The former began by describing the mail
+server the log files were taken from, the computer used to generate the
+statistics in this chapter, and how the parser scales as the size of log
+files increases, including why performance is better on the larger log
+files.  The framework optimises the order in which rules are used when
+trying to recognise each log line, and the effect that optimisation has is
+explored; this is followed by the impact that adding more rules to the
+ruleset has on parser performance.  The simplest and most effective of the
+optimisations, caching compiled regexes, is described next, and the
+efficiency evaluation concludes with an examination of where parsing time
+is spent: recognising log lines or processing them?
 
-XXX REWRITE WHEN THIS CHAPTER IS FINISHED\@.
+Coverage of Postfix log files is divided into two topics in this chapter:
+log lines correctly recognised, and mail delivery attempts correctly
+understood and reconstructed.  The former is initially more important,
+because the parser must correctly recognise every log line if it is to be
+complete, but subsequently the latter takes precedence because correctly
+reconstructing the journey a mail delivery attempt takes through Postfix is
+the aim of the parser.  Increasing the proportion of log lines correctly
+recognised is relatively simple and non-intrusive: adding new rules or
+modifying existing rules is very easy because of the separation of rules,
+actions, and framework.  Improving the understanding and reconstruction of
+mail delivery attempts is harder, because Postfix's behaviour must be
+analysed and figured out, and the new behaviour integrated into the actions
+without breaking the existing parsing.  Detecting a deficiency in the
+parser is also significantly harder, because the parser will warn about
+unrecognised log lines, whereas discovering a flaw in the parser requires
+careful study of any warnings produced and the entries remaining in the
+state table.  Rectifying a flaw in the parser requires a deep understanding
+of both the parser and Postfix's log files, investigative work to determine
+the cause of the deficiency, further examination of the log files to aid in
+developing a solution, and finally implementation, integration, and testing
+of the solution.
 
-This section began with parser scalability, showing the linear relationship
-between parsing time and input size.  Demonstrates the effect of rule
-ordering on parsing time, and the unexpected consequences of specific
-inputs.  The necessity of caching compiled regexes is attested to by the
-third group of graphs, where the difference between caching and not caching
-is staggering.  The penultimate section contains a breakdown of the rule
-hits accumulated during a single test run of the parser.  Miscellaneous
-graphs expected to be useful are collected in the final section.
-
-XXX REWRITE COVERAGE SUMMARY IF NECESSARY\@.
-
-Parser coverage is divided into two topics in this section: log lines
-covered and mails covered.  The former is initially more important, as the
-parser must successfully parse every line if it is to be complete, but
-subsequently the latter takes precedence because reproducing the path a
-mail takes through Postfix is the aim of the parser.  Increasing the
-percentage of log lines parsed is relatively simple and non-intrusive:
-adding new rules or modifying existing rules is simplified by the
-separation of rules, actions, and framework.  Improving the logical
-coverage is harder, as the actions taken by Postfix must be reconstructed
-by the author, and the new sequence of actions integrated into the existing
-model without breaking the existing parsing.  Detecting a deficiency in the
-parsing algorithm is also significantly harder than detecting unparsed log
-lines, as the parser will warn about any unparsed line, whereas discovering
-a flaw in the parser requires understanding of the warnings produced and
-the mails remaining in the state table.  Rectifying a flaw in the parser
-requires an understanding of both the parser and Postfix's log files, and
-investigative work to determine the cause of the deficiency, followed by
-further examination of the log files in developing a solution.
-
-Experience implementing the Postfix log parser shows that full input
-coverage is relatively easy to achieve with this architecture, and that
-with enough time and effort full understanding of the input is possible.
-Postfix log files would require substantial time and effort to correctly
-parse regardless of the architecture used; this architecture enables an
-iterative approach to be used (similar to Stepwise
-Refinement~\cite{stepwise-refinement}), as is practiced in many other
-software engineering disciplines.
+This chapter shows that it is possible to balance the conflicting goals of
+efficient and accurate parsing, and that one does not have to be sacrificed
+to achieve the other.

### 2009-04-12 doc/logparser-acronyms.tex
Ran diction and made changes based on its suggestions.

@@
         the source of large volumes of spam.  The criteria used when
-        deciding whether or not to include an IP address on a DNSBL vary
-        widely between DNSBLs, so before using one it is very important to
-        check their listing policies.  To use a DNSBL the mail server makes
-        a DNS request incorporating the IP address of the client; if the
+        deciding if an IP address should be included on a DNSBL vary widely
+        between DNSBLs, so before using one it is very important to check
+        their listing policies.  To use a DNSBL the mail server makes a DNS
+        request incorporating the IP address of the client; if the
         requested hostname exists the client is on the DNSBL, and the mail
@@
         guarantees it will work, but is free to change the implementation
-        as long as the API is preserved.
+        if the API is preserved.
     }
@@
         Finite automata are computing devices that accept or recognize
-        regular languages.  They lack any form of storage, so while they
+        regular languages.  They lack any form of storage, so although they
         can recognise languages such as ${(ab)}^{*}$ or
@@
         the form $L\rightarrow{}R$, where $L$ is a single non-terminal
-        symbol, and $R$ is a (possibly empty) string of terminal and/or
+        symbol, and $R$ is a (possibly empty) string of terminal or
         non-terminal symbols.  Every CFG can be recognised by a

### 2009-04-12 doc/logparser-architecture.tex
Changes from proofreading of introduction chapter.
Consistent float placement.
Ran diction and made changes based on its suggestions.

@@
 
-\section{Parser Architecture and Design}
+\section{Architecture Overview}
 
@@
 With this in mind, one of the architecture's design aims was to make adding
-new rules to parse new inputs as easy as possible, to enable administrators
-to properly parse their own log files.  The solution developed is to divide
-the architecture into three parts: Framework, Actions, and Rules.  Each
-will be discussed separately, but first an overview:
+new rules to parse new inputs as effortless as possible, to enable
+administrators to properly parse their own log files.  The solution
+developed is to divide the architecture into three parts: Framework,
+Actions, and Rules.  Each will be discussed separately, but first an
+overview:
 
@@
 
-There is some similarity between this architecture and William Wood's
+Some similarity exists between this architecture and William Wood's
 \acronym{ATN}~\cite{atns,nlpip}, used in Computational Linguistics for
@@
 the two (shown in \tableref{Similarities between ATN and this
-architecture}) is accidental, but it is clear that the two different
-approaches share a similar division of responsibilities, despite having
-different semantics.
+architecture}) is accidental, but clearly the two different approaches
+share a similar division of responsibilities, despite having different
+semantics.
 
@@
 
-\begin{table}[ht]
+\begin{table}[thbp]
     \caption{Similarities between ATN and this architecture}
@@
         have serious security implications.  It validates each rule to
-        hopefully catch problems as early in the parsing process as
-        possible; the checks will be implementation-specific to some
-        extent, but will generally include some of the following:
+        catch problems as early in the parsing process as possible; the
+        checks will be implementation-specific to some extent, but will
+        generally include some of the following:
 
@@
         inputs once, rather than dealing with it in every rule and action.
-        This is easy in the case of Postfix log files because there is a
+        This is trivial for Postfix log files because they have a
         one-to-one mapping between physical and logical inputs; mapping
@@
         utility functions; the framework may be a good location for support
-        functions, but if there is another way to make those functions
+        functions, but if another way exists to make those functions
         available to all actions it may be preferable to use it instead,
@@
 handle one input category, with each rule recognising one input variant.
-It is expected that parsers based on this architecture will have many
+It is anticipated that parsers based on this architecture will have many
 actions, and each action may be specified by many rules, with the aim of
@@
 escaped or embedded in quoted strings the implementation must be careful
-not to remove those.  For some inputs this type of pre-processing can
+not to remove those.  For some inputs this kind of pre-processing can
 greatly simplify parsing, echoing the simplification provided by the
@@
 
-\begin{table}[ht]
+\begin{table}[thbp]
     \caption{Rules to parse C-style comments}
@@
 cannot always be implemented using a \acronym{PDA}, so intersection may be
-intractable in some cases, e.g.:
+intractable for some combinations, e.g.:
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
@@
 account, because two rules whose regexes overlap may have conditions
-attached which prevent the rules overlapping.  A less naive approach to
+attached that prevent the rules overlapping.  A less naive approach to
 overlap detection would first check for overlapping conditions amongst
@@
 
-Once conditions pass a certain level of complexity determining whether two
+Once conditions pass a certain level of complexity determining if two
 conditions overlap becomes intractable, because it requires so much
@@
 being set to true by one or more rules or actions.  If variables used in
-conditions can be set by actions, determining whether or not two conditions
-overlap is impossible: the Halting Problem show that it is impossible for
-one program to answer questions about another program's behaviour if the
-inquiring program is implemented on a computational machine of the equal or
-lesser power to the program being inquired into.
+conditions can be set by actions, determining if two conditions overlap is
+impossible: the Halting Problem show that it is impossible for one program
+to answer questions about another program's behaviour if the inquiring
+program is implemented on a computational machine of the equal or lesser
+power to the program being inquired into.
 
@@
 
-It is possible to define pathological regexes which fall into two main
+It is possible to define pathological regexes, which fall into two main
 categories: regexes that match inputs they should not, and regexes that
@@
 
-Usually excessive CPU time is consumed when a regex with a lot of
-backtracking, caused by alteration and/or nested quantifiers, fails to
-match; successful matching is generally quite fast with such regexes, so
-problematic regexes may go unnoticed for some time.  E.g.\ with most regex
-engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"! is very
-fast when there is a match, but when the match fails its computational
-complexity for a string of length $n$ is $O(2^{n})$;
+Usually excessive CPU time is consumed when a regex fails to match, and the
+regex engine backtracks many times because of alteration or nested
+quantifiers; successful matching is generally quite fast with such regexes,
+so problematic regexes may go unnoticed for some time.  E.g.\ with most
+regex engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"!
+is very fast when there is a match, but when the match fails its
+computational complexity for a string of length $n$ is $O(2^{n})$;
 see~\cite{mastering-regular-expressions} for in-depth discussion of nested
 quantifiers, backtracking, alteration, and capturing groups.  Pathological
-regexes which consume excessive CPU time can be difficult to detect, both
-by visual inspection and by machine inspection, but if a regex is converted
-to a \acronym{FA} or the internal representation used by the regex engine,
-it may be possible to determine if there are nested quantifiers or other
-troublesome constructs present.  Modern regex engines have addressed many
-of these problems, e.g.\ the regex to match double quoted strings given
-above fails immediately with Perl 5.10, regardless of the input length,
-because the regex engine looks for both of the required double quotes
-first.  Similarly Perl 5.10's regex engine optimises alterations starting
-with literal text into a trie, which has matching time proportional to the
-length of the alternatives, rather than the number of alternatives.  Perl
-regexes can use \verb!(?>pattern)!, which matches \verb!pattern! the first
-time the regex engine passes over it, and does not change what it
-originally matched if the regex engine backtracks over it, so troublesome
-regexes can use this to reduce the impact of backtracking; Prolog users
-will notice a similarity to the \verb'!' (cut) operator.  A presentation
-showing some of Perl 5.10's new regex features is available at
+regexes that consume excessive CPU time can be difficult to detect, both by
+visual inspection and by machine inspection, but if a regex is converted to
+a \acronym{FA} or the internal representation used by the regex engine, it
+may be possible to determine if nested quantifiers or other troublesome
+constructs are present.  Modern regex engines have addressed many of these
+problems, e.g.\ the regex to match double quoted strings given above fails
+immediately with Perl 5.10, regardless of the input length, because the
+regex engine looks for both of the required double quotes first.  Similarly
+Perl 5.10's regex engine optimises alterations starting with literal text
+into a trie, which has matching time proportional to the length of the
+alternatives, rather than the number of alternatives.  Perl regexes can use
+\verb!(?>pattern)!, which matches \verb!pattern! the first time the regex
+engine passes over it, and does not change what it originally matched if
+the regex engine backtracks over it, so troublesome regexes can use this to
+reduce the impact of backtracking; Prolog users will notice a similarity to
+the \verb'!' (cut) operator.  A presentation showing some of Perl 5.10's
+new regex features is available at
 \urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.
@@
 check variables in uncomplicated ways may exhibit unexpected or incorrect
-behaviour, but are unlikely to exhibit pathological behaviour.  Deciding
-whether a more complex condition's behaviour is pathological or simply a
-bug is difficult and to some extent is a matter of opinion.  When this
+behaviour, but are unlikely to exhibit pathological behaviour.  Deciding if
+a more complex condition's behaviour is pathological or simply a bug is
+difficult and to some extent is a matter of opinion.  When this
 architecture has received more widespread usage consensus should be reached
@@
 provides, plus any difficulties associated with the components.  The
-framework provides a number of support functions and manages the parsing
+framework provides several support functions and manages the parsing
 process, enabling simple rules and actions to be written.  The actions are
-quite easy to understand, because the architecture does not impose any
+simple to understand, because the architecture does not impose any
 structure or requirements upon them: parser authors are free to do anything

### 2009-04-12 doc/logparser-background.tex
Automatically generate the list of graphs, tables, and other included files.
Dealt with some XXX items. Improvements in results section. Added table showing number of rules per program.
Consistent float placement.
Renamed Results chapter to Evaluation. Finished evaluation chapter introduction and summary.
Ran diction and made changes based on its suggestions.

@@
 
-\begin{table}[ht]
+\begin{table}[thbp]
     \caption{Sample output from SQL query, showing the percentage of
@@
 allowing an administrator to employ those restrictions which they judge
-suitable for their site's needs, rather than a fixed set chosen by
+suitable for their server's needs, rather than a fixed set chosen by
 Postfix's author.  These restrictions can be selectively applied, combined,
@@
 most cases to use standard Unix text processing utilities to determine the
-fate of an individual mail, in some cases it can be quite difficult.  For
+fate of an individual mail, sometimes it can be quite difficult.  For
 most mails the journey is simple and brief, but the remainder can have
@@
 separated into multiple components that each perform one of the tasks
-required of an \acronym{MTA}\@: receive mail, send mail, local delivery of
-mail, etc.;\ full details can be found in~\cite{postfix-overview}.
-Postfix's design is strongly influenced by security concerns: those
-components that interact with other hosts are not
-privileged,\footnote{Privilege means the power to perform actions that are
-limited to the administrator, and not available to ordinary users.} so bugs
-in those components will not give an attacker extra privileges; those
-components that are privileged do not interact with other hosts, making it
-much more difficult for an attacker to exploit any bugs that may exist in
-those components.
+required of an \acronym{MTA}, e.g.\ receive mail, send mail, local delivery
+of mail;\ full details can be found in~\cite{postfix-overview}.  Postfix's
+design is strongly influenced by security concerns: those components that
+interact with other hosts are not privileged,\footnote{Privilege means the
+power to perform actions that are limited to the administrator, and not
+available to ordinary users.} so bugs in those components will not give an
+attacker extra privileges; those components that are privileged do not
+interact with other hosts, making it much more difficult for an attacker to
+exploit any bugs that may exist in those components.
 
@@
 
-That example also shows how easy it is to supply a custom rejection
+That example also shows that it is trivial to supply a custom rejection
 message.  When the result is the name of another restriction Postfix will
@@
 
-\begin{table}[ht]
-
+\begin{table}[thbp]
     \caption{Example attributes sent to policy servers, taken
@@
     \empty{}\label{Example attributes sent to policy servers}
-
     \centering{}
-
     \begin{tabular}[]{ll}
@@
         \numberOFlogFILES{} log files used when generating the results in
-        \sectionref{Results} are shown in \tableref{Most effective Postfix
-        restrictions}.  The table shows that greylisting is worth using at
-        the moment, particularly when you take into account its position as
-        the final restriction that a mail must overcome in the
+        \sectionref{Evaluation} are shown in \tableref{Most effective
+        Postfix restrictions}.  The table shows that greylisting is worth
+        using at the moment, particularly when you take into account its
+        position as the final restriction that a mail must overcome in the
         configuration used on the mail server that generated the log files:
@@
 
-        \begin{table}[ht]
-            \caption{Most effective Postfix restrictions}
-            \empty{}\label{Most effective Postfix restrictions}
-            \input{build/include-restriction-summary-table}
-        \end{table}
+        \showtable{build/include-restriction-summary-table}{Most effective
+        Postfix restrictions}{Most effective Postfix restrictions}
 

### 2009-04-12 doc/logparser-complications.tex
Ran diction and made changes based on its suggestions.

@@
 would be straightforward to parse, process, and understand.  The large
-variation in log lines was not apparent, because the majority of log lines
-are recognised by a small set of rules, as shown in \figureref{rule hits
+variation in log lines was not apparent, because most log lines are
+recognised by a small set of rules, as shown in \figureref{rule hits
 graph}.  Most of the myriad complications and difficulties documented in
@@
         caused by lack of information in the log files.  In general some
-        extra logging by Postfix could remove these complications, and in
-        one case extra logging in later versions of Postfix does remove the
-        complication (\sectionref{identifying bounce notifications}); three
-        of the complications would be removed if Postfix generated a log
-        line whenever it discarded a mail.
+        extra logging by Postfix could remove these complications, and for
+        one complication, extra logging in later versions of Postfix does
+        remove it (\sectionref{identifying bounce notifications}); three of
+        the complications would be removed if Postfix generated a log line
+        whenever it discarded a mail.
 
@@
 
-    \item Most importantly: each action needs to know whether it deals with
-        accepted mails, and should find the mail in the by\_queueid state
-        tables, or deals with delivery attempts, and thus will find the
-        connection in the by\_pid state tables.  Some actions, e.g.\
+    \item Each action needs to know whether it deals with accepted mails,
+        and should find the mail in the by\_queueid state tables, or deals
+        with delivery attempts, and thus will find the connection in the
+        by\_pid state tables.  Some actions, e.g.\
         \action{DELIVERY\_REJECTED}, need to query both tables.  Support
@@
 There should ideally be two separate entries in the database resulting from
-the above sequence, but currently there will only be one.
+the above sequence, but there will only be one.
 
@@
 
-Unfortunately, while all log lines from an individual process appear in
+Unfortunately, although all log lines from an individual process appear in
 chronological order, the order in which log lines from different processes
@@
 mail as coming from an unknown origin, and subsequently clear that flag if
-and when the origin of the mail becomes clear.  Obviously the parser could
-omit checking where mails originate from, but requiring an explicit source
-helps to expose bugs in the parser; such checks helped to identify the
+the origin of the mail becomes clear.  Obviously the parser could omit
+checking where mails originate from, but requiring an explicit source helps
+to expose bugs in the parser; such checks helped to identify the
 complications described in \sectionref{discarding cleanup log lines} and
@@
 as coming from an unknown origin cannot be entered into the database when
-their final log line is processed, nor can \parsername{} warn the user;
+their final log line is processed, and \parsername{} cannot warn the user;
 instead they must be marked as ready for entry and subsequently entered
@@
         child mails waiting to be entered in the database; these are
-        entered as normal, and removed from the state tables.  If there are
-        incomplete child mails, the parent's removal from the state tables
-        will be postponed until the last child has been entered.
+        entered as usual, and removed from the state tables.  If the state
+        tables contain incomplete child mails, the parent's removal from
+        the state tables will be postponed until the last child has been
+        entered.
 
@@
 line when it generates a bounce notification; suddenly there will be log
-entries for a mail that lacks an obvious source.  There are similarities to
-the problem of identifying re-injected mails discussed in
-\sectionref{tracking re-injected mail}, but unlike the solution described
-therein bounce notifications do not eventually have a log line that
-identifies their source.  Heuristics must be used to identify bounce
-notifications:
+entries for a mail that lacks an obvious source.  Similarities exist to the
+problem of identifying re-injected mails discussed in \sectionref{tracking
+re-injected mail}, but unlike the solution described therein bounce
+notifications do not eventually have a log line that identifies their
+source.  Heuristics must be used to identify bounce notifications:
 
@@
         a bounce notification that is being re-injected as a result of
-        aliasing.  In the latter case the message-id will be unchanged from
+        aliasing.  For the latter, the message-id will be unchanged from
         the original bounce notification, and so even if it happens to be
@@
 
-There is a small chance that a mail will be incorrectly identified as a
-bounce notification, as the heuristics used may be too broad.  For this to
+A small chance exists that a mail will be incorrectly identified as a
+bounce notification, because the heuristics used may be too broad.  For this to
 occur the following conditions would have to be met:
@@
     \item The message-id must have the correct format and correspond with
-        the queueid of the mail.  While a mail sent from elsewhere could
+        the queueid of the mail.  Although a mail sent from elsewhere could
         easily have the correct message-id format, the chance that the
@@
 If a mail is misclassified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably misclassification in
-this case is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.  Postfix 2.3 (and
-hopefully subsequent versions) log the creation of a bounce message.
+have been generated internally by Postfix; arguably misclassification of
+this kind is a benefit rather than a drawback, as other mails generated
+internally by Postfix will be handled correctly.  Postfix 2.3 and
+subsequent versions log the creation of a bounce message.
 
@@
     \item Is \daemon{smtpd} the only Postfix component that produced a log
-        line for this mail?  Every mail which passes normally through
+        line for this mail?  Every mail that passes normally through
         Postfix will have a \daemon{cleanup} line, and later a
@@
 
-    \item If there are third and subsequent results, were those log lines
-        processed by the \action{SAVE\_DATA} action?  If there are any log
-        lines after the first two they should be informational only.
+    \item If third and subsequent results exist, were those log lines
+        processed by the \action{SAVE\_DATA} action?  Any log lines after
+        the first two they should be informational only.
 
@@
 table} and \sectionref{results table}).  These heuristics are quite
-restrictive, and it appears there is little scope for false positives; if
-there are any false positives there will also be warnings when the next log
-line for that mail is parsed.  False negatives are less likely to be
-detected: there may be queueid clashes (and warnings) if mails remain in
-the state tables after they should have been removed, otherwise the only
-way to detect false negatives is to examine the state tables after each
-parsing run.
+restrictive, and it appears to have little scope for false positives; any
+false positives will cause a warning when the next log line for that mail
+is parsed.  False negatives are less likely to be detected: there may be
+queueid clashes (and warnings) if mails remain in the state tables after
+they should have been removed, otherwise the only way to detect false
+negatives is to examine the state tables after each parsing run.
 
@@
     \item Is the mail missing its \daemon{cleanup} log line?  Every mail
-        which passes normally through Postfix will have a \daemon{cleanup}
+        that passes normally through Postfix will have a \daemon{cleanup}
         line; lack of one is a sure sign the mail did not make it too far.
@@
 The DATA phase of the \acronym{SMTP} conversation is where the headers and
-body of the mail are transferred.  Sometimes there is a timeout or the
+body of the mail are transferred.  Sometimes a timeout occurs or the
 connection is lost during the DATA phase;\footnote{For the sake of brevity
@@
 mail.  It seems more intuitive to save the mail's data to the database, but
-if a timeout occurs there is no data available to save; the timeout is
+if a timeout occurs no data will be available to save; the timeout is
 recorded and saved with the connection instead.  To deal properly with
@@
         necessarily correct, because Postfix may have accepted an earlier
-        recipient and rejected a later one, in which case the timeout
-        applies to the partially accepted mail, which should be discarded;
-        this has not been a problem in practice.  Processing timeouts is
-        further complicated by the presence of out of order
-        \daemon{cleanup} log lines: see \sectionref{discarding cleanup log
-        lines} for details.
+        recipient and rejected a later one, so the timeout applies to the
+        partially accepted mail, which should be discarded; this has not
+        been a problem in practice.  Processing timeouts is further
+        complicated by the presence of out of order \daemon{cleanup} log
+        lines: see \sectionref{discarding cleanup log lines} for details.
 
@@
 
-    \item The only program which has logged anything thus far for the mail
+    \item The only program that has logged anything thus far for the mail
         is \daemon{cleanup}.
 
-    \item There is less than a five second difference between the
-        timestamps of the \daemon{cleanup} and \daemon{pickup} log lines.
+    \item The difference between the timestamps of the \daemon{cleanup} and
+        \daemon{pickup} log lines is less than five seconds.
 
@@
 immediately obvious reason.  After poring over log files for some time
-several reasons have been found for this infrequent occurrence:
+several reasons have been found for this rare event:
 
@@
         connections state table must be cleaned up, entered in the database
-        if there is sufficient data, and deleted.
+        if they have enough data, and deleted.
 
@@
         for some reason, by the administrator, or by the OS), so its active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
+        connection must be cleaned up, entered in the database if it has
+        enough data, and deleted from the connections state table.
 
     \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if there is
-        sufficient data, and deleted from the connections state table.
+        connection must be cleaned up, entered in the database if it has
+        enough data, and deleted from the connections state table.
 
@@
 
-Occasionally a log file will have out of order log lines which cannot be
+Occasionally a log file will have out of order log lines that cannot be
 dealt with by the techniques described in \sectionref{tracking re-injected
@@
 The \action{COMMIT} action examines the list of programs that have produced
-log lines for each mail, comparing the list against a table of known-good
-program combinations.  If the mail's combination is found in the table the
-mail can be entered in the database; if the combination is not found entry
-must be postponed and the mail flagged for later entry.  The
+log lines for each mail, checking the list against a table of known-good
+Postfix component combinations.  If the mail's combination is found in the
+table the mail can be entered in the database; if the combination is not
+found entry must be postponed and the mail flagged for later entry.  The
 \action{MAIL\_DELIVERED} action checks for that flag; if the additional log
@@
 symptoms are the same as in \sectionref{aborted delivery attempts}, except
-that there \textit{is\/} a \daemon{cleanup} log line; there is nothing in
-the log file to explain why there are no further log lines.  The only way
-to detect these mails is to periodically scan all mails in the state
-tables, deleting any mails with the following characteristics:
+that the mail has a \daemon{cleanup} log line; nothing can be found in the
+log file to explain why this mail does not have any further log lines.  The
+only way to detect these mails is to periodically scan all mails in the
+state tables, deleting any mails with the following characteristics:
 
@@
 faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail before Postfix has had a chance to try to deliver it.  In
-this case the recipient address will not have been logged, so a dummy
-recipient address needs to be added, as every mail is required by the
-database schema (\sectionref{results table}) to have at least one
-recipient.  Typically when this complication occurs there are many
-instances of it, closely grouped.
+some of the mail before Postfix has had a chance to try to deliver it.  The
+recipient address will not have been logged, so a dummy recipient address
+needs to be added, because every mail is required by the database schema
+(\sectionref{results table}) to have at least one recipient.  Typically
+when this complication occurs the log file will show many instances of it,
+closely grouped.
 
@@
 only arises eight times in quick succession in one log file, which is not
-in the \numberOFlogFILES{} log files used for testing; if this complication
-occurred more frequently it might be desirable to find the mail in the
-database and add the log line to it.
+in the \numberOFlogFILES{} log files used for evaluating the parser; if
+this complication occurred more frequently it might be desirable to find
+the mail in the database and add the log line to it.
 
@@
 are caused by deficiencies in Postfix's logging, and some could easily be
-resolved by adding additional logging to Postfix; these seven are
-dealt with by applying heuristics to specific log lines and/or mails, in
-some cases with the aid of information cached when mails are removed from
-the state tables.
+resolved by adding additional logging to Postfix; these seven are dealt
+with by applying heuristics to specific log lines or mails, sometimes in
+conjunction with information cached when mails are removed from the state
+tables.

### 2009-04-12 doc/logparser-conclusion.tex
Ran diction and made changes based on its suggestions.

@@
 because a quick survey of sample log files gives the impression that the
-number of distinct log lines is quite small; this impression is due to the
-uneven distribution exhibited by log lines produced in normal operation
+number of distinct log lines is quite small; this impression is because of
+the uneven distribution exhibited by log lines produced in normal operation
 (see \graphref{rule hits graph} for a vivid illustration of this).

### 2009-04-12 doc/logparser-glossary.tex
Dealt with some XXX items.
Ran diction and made changes based on its suggestions.

@@
     a separate server process, greatly simplifying deployment of an
-    application utilising it.  More details can be found
-    at~\urlLastChecked{http://www.sqlite.org/}{2009/02/23}.
+    application utilising it.
 }}
@@
     Most operating systems store the current time and timestamps of files
-    etc.\ as seconds elapsed since the epoch, the beginning of time as far
-    as the operating system is concerned.  On Unix and Unix-derived systems
-    the epoch is 1970/01/01 00:00:00; on other operating systems it may be
+    as seconds elapsed since the epoch, the beginning of time as far as the
+    operating system is concerned.  On Unix and Unix-derived systems the
+    epoch is 1970/01/01 00:00:00; on other operating systems it may be
     different.
@@
 \newglossaryentry{hash}{name={hash},description={
-    A hashing function transforms a string of characters to a number.
-    There are many possible uses for the resulting number: a common usage
-    is to maintain a data structure indexed by strings in an efficient
-    manner.  A full description is beyond the scope of this paper, further
-    information can be found at~\cite{hash-functions}.
+    A hashing function transforms a string of characters to a number.  A
+    common usage is to maintain a data structure indexed by strings in an
+    efficient manner.  A full description is beyond the scope of this
+    paper, further information can be found at~\cite{hash-functions}.
 }}
@@
     deluge of mail can be severe enough to interrupt service for the victim
-    and/or other users, by causing mail to be delayed or rejected because
-    the victim or the server has reached a limit, e.g.\ too many mails in
-    their mailbox, or too much disk space consumed by the mail queue.
+    or other users, by causing mail to be delayed or rejected because the
+    victim or the server has reached a limit, e.g.\ too many mails in their
+    mailbox, or too much disk space consumed by the mail queue.
 }}
@@
     occurs when the chain of recipient addresses constitute a cyclic
-    directed graph; the simplest example is when mail for
-    \texttt{alice@example.net} is delivered to \texttt{bob@example.com},
-    and mail for \texttt{bob@example.com} is delivered to
-    \texttt{alice@example.net}.  Postfix adds a \texttt{Delivered-To:}
-    header to all mails it delivers locally to detect mail loops, but in
-    some cases this header is removed by the program that forwards the
-    mail.
+    directed graph; e.g.\ mail for \texttt{alice@example.net} is delivered
+    to \texttt{bob@example.com}, and mail for \texttt{bob@example.com} is
+    delivered to \texttt{alice@example.net}.  Postfix adds a
+    \texttt{Delivered-To:} header to all mails it delivers locally to
+    detect mail loops, but in some cases this header is removed by the
+    program that forwards the mail.
 }}

### 2009-04-12 doc/logparser-implementation.tex
Ran diction and made changes based on its suggestions.

@@
 frequently, to discard a data structure (e.g.\ \sectionref{timeouts during
-data phase}).  \parsername{} is capable of parsing log files from Postfix
-version 2.2 through to version 2.5, and should be capable of parsing log
-files from later versions with only minor modifications or an updated
-ruleset.  The parser's task is to follow the journey each mail takes
-through Postfix, combining the data captured by rules into a coherent
-whole, saving it in a useful and consistent form, and performing
-housekeeping duties.  The intermingling of log lines from different mails
-immediately rules out the possibility of handling each mail in isolation;
-the parser must be able to handle multiple mails in parallel, each
-potentially at a different stage in its journey, without any interference
-between mails --- except in the minority of cases where intra-mail
-interference is required, e.g.\ mail re-injected for forwarding
-(\sectionref{Re-injected mails}).  The best way to deal with intermingling
-of log lines is to maintain state information for every unfinished mail,
-and to manipulate the appropriate mail correctly for each log line
-encountered.
+data phase}).  \parsername{} can parse log files from Postfix version 2.2
+through to version 2.5, and should be capable of parsing log files from
+later versions with only minor modifications or an updated ruleset.  The
+parser's task is to follow the journey each mail takes through Postfix,
+combining the data captured by rules into a coherent whole, saving it in a
+useful and consistent form, and performing housekeeping duties.  The
+intermingling of log lines from different mails immediately rules out the
+possibility of handling each mail in isolation; the parser must handle
+multiple mails in parallel, each potentially at a different stage in its
+journey, without any interference between mails --- except in the minority
+of cases where intra-mail interference is required, e.g.\ mail re-injected
+for forwarding (\sectionref{Re-injected mails}).  The best way to deal with
+intermingling of log lines is to maintain state information for every
+unfinished mail, and to manipulate the appropriate mail correctly for each
+log line encountered.
 
@@
 On first inspection, Postfix log files look like they will be simple to
-parse, but this is not the case.  The many complications and difficulties
-encountered while writing \parsername{}, and the solutions developed to
-overcome them, are documented in \sectionref{complications}.  This chapter
-finishes with a list of \parsernames{} limitations, and some possible
-improvements that could be implemented.
+parse, but this impression turns out to be incorrect.  The many
+complications and difficulties encountered while writing \parsername{}, and
+the solutions developed to overcome them, are documented in
+\sectionref{complications}.  This chapter finishes with a list of
+\parsernames{} limitations, and some possible improvements that could be
+implemented.
 
@@
         administrator deletes mail from the queue without using
-        \daemon{postsuper}, or if there is filesystem corruption.
+        \daemon{postsuper}, or if the server suffers from filesystem
+        corruption.
 
@@
 In some ways this task is similar to reverse engineering or replicating a
-black box system based solely on its inputs and outputs.  Thus far
-analysis of the log files has been sufficient to reconstruct Postfix's
+black box system based solely on its inputs and outputs.  Thus far,
+analysis of the log files has been enough to reconstruct Postfix's
 behaviour, but for more difficult programs the techniques described in
-\cite{black-box-error-reporting} may be useful.  There are some advantages
-to treating Postfix as a black box during parser development:
+\cite{black-box-error-reporting} may be useful.  Some advantages come from
+treating Postfix as a black box during parser development:
 
@@
 Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
-forwarding, or generated internally by Postfix.  Local submission is the
-simplest case: a queueid is assigned immediately \flowchart{PICKUP}{2}, and
-the mail moves on to the delivery stage.  Re-injection due to forwarding
-lacks explicit log lines of its own; it is explained fully in
-\sectionref{Re-injected mails}.  Internally generated mails lack any
-explicit origin in Postfix 2.2.x and must be detected using heuristics as
-described in \sectionref{identifying bounce notifications}; later versions
-of Postfix do provide log lines for internally generated mails
-\flowchart{BOUNCE\_CREATED}{3}.  Bounce notifications are the primary
-example of internally generated mails, though there are other types, e.g.\
+submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection
+because of forwarding, or generated internally by Postfix.  Local
+submission is the simplest of the four: a queueid is assigned immediately
+\flowchart{PICKUP}{2}, and the mail moves on to the delivery stage.
+Re-injection because of forwarding lacks explicit log lines of its own; it
+is explained fully in \sectionref{Re-injected mails}.  Internally generated
+mails lack any explicit origin in Postfix 2.2.x and must be detected using
+heuristics as described in \sectionref{identifying bounce notifications};
+later versions of Postfix do provide log lines for internally generated
+mails \flowchart{BOUNCE\_CREATED}{3}.  Bounce notifications are the primary
+example of internally generated mails, though other types exist, e.g.\
 Postfix may generate mails to the administrator when it encounters
@@
 
-    \item First there is a connection from the remote client
-        \flowchart{CONNECT}{1}.
-
-    \item This is followed by rejection of sender address, recipient
-        addresses, client \acronym{IP} address or hostname, etc.\
+    \item The remote client connects \flowchart{CONNECT}{1}.
+
+    \item This is followed by rejection of \acronym{SMTP} commands by the
+        restrictions configured by the administrator
         \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
@@
 whether the mail is delivered to a mailbox, piped to a command, delivered
-to a remote server, bounced (due to a mail loop, delivery failure, or five
-day timeout), or deleted by the administrator, \textit{unless\/} the mail
-is either parent or child of re-injection, as explained in
+to a remote server, bounced (because of a mail loop, delivery failure, or
+five day timeout), or deleted by the administrator, \textit{unless\/} the
+mail is either parent or child of re-injection, as explained in
 \sectionref{tracking re-injected mail}.
@@
 understanding the actions of the parser, and essential to developing
-further applications which utilise the data gathered; \sectionref{database
+further applications that utilise the data gathered; \sectionref{database
 as API} describes how the database schema functions as an \acronym{API}.
@@
 connections and results.  The connections table contains an entry for every
-mail accepted and every connection where there was a rejection; the
-individual fields will be described in \sectionref{connections table}.  The
-results table will have at least one entry for each entry in the
+mail accepted and every connection that rejected an \acronym{SMTP} command;
+the individual fields will be described in \sectionref{connections table}.
+The results table will have at least one entry for each entry in the
 connections table; its fields will be covered in detail in
@@
 available for inspection because each result references the rule that
-created it.  There is no ambiguity about which rule resulted in a
-particular result, eliminating one potential source of confusion.
+created it.  No ambiguity or confusion is possible about which rule
+resulted in a particular result, eliminating one potential source of
+confusion.
 
@@
     \item \acronym{SQL} is reasonably standard and many people will already
-        be familiar with it; for those unfamiliar with it there are lots of
-        readily available resources from which to learn, e.g.\
+        be familiar with it; for those unfamiliar with \acronym{SQL}, lots of
+        resources are available from which to learn, e.g.\
         \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
@@
     \item [restriction\_name] The name of the Postfix restriction that
-        caused the mail to be rejected.  This field is meaningful only for
-        rules that recognise rejection log lines, i.e.\ rules that have an
-        action of \action{DELIVERY\_REJECTED}.
+        caused the mail to be rejected.  This field is valid only for rules
+        that recognise rejection log lines, i.e.\ rules that have an action
+        of \action{DELIVERY\_REJECTED}.
 
@@
 
-Every accepted mail and every connection where there was a rejection will
-have a single entry in the connections table containing all of the
-following fields:
+Every accepted mail and every connection that rejected an \acronym{SMTP}
+command will have a single entry in the connections table containing all of
+the following fields:
 
@@
 
-If there are state tables from a previous \parsername{} run to be loaded,
-they will be loaded now; the framework supports saving state at any time,
-without adversely affecting the parsing process.  The need to track
-re-injected mail (\sectionref{Re-injected mails}) complicates the process
-of saving state, as the relationships between mails must be maintained;
-loading state is also complicated by dealing with aborted delivery attempts
+State tables from a previous \parsername{} run, if any, will be loaded now;
+the framework supports saving state at any time, without adversely
+affecting the parsing process.  The need to track re-injected mail
+(\sectionref{Re-injected mails}) complicates the process of saving state,
+as the relationships between mails must be maintained; loading state is
+also complicated by dealing with aborted delivery attempts
 (\sectionref{aborted delivery attempts}), because a separate set of
@@
 parsing compressed log files, because a compressed block may uncompress to
-a variable number of log lines; there is also variation between the
-recognition and processing time of individual log lines, but overall the
-progress bar is a useful addition.  \parsername{} is primarily intended for
-parsing complete log files, but with minor modifications it could be used
-to parse a live log file, periodically checking it for new log lines; this
-could be very useful for programs that work best with up to date data,
-e.g.\ a program for monitoring the health of the mail system.
+a variable number of log lines; variation between the recognition and
+processing time of individual log lines also affects its accuracy, but
+overall the progress bar is a useful addition.  \parsername{} is primarily
+intended for parsing complete log files, but with minor modifications it
+could be used to parse a live log file, periodically checking it for new
+log lines; this could be very useful for programs that work best with up to
+date data, e.g.\ a program for monitoring the health of the mail system.
 
@@
 techniques used to improve parsing speed can be turned off or altered to
-measure the effect they have.  In each case the data gathered with the
-optimisation disabled will be compared to the data gathered with it
-enabled, to quantify the benefit each optimisation provides.  Three sets of
-data are collected:
+measure the effect they have.  The data gathered with each optimisation
+disabled will be compared to the data gathered with it enabled, to quantify
+the benefit each optimisation provides.  Three sets of data are collected:
 
@@
 
-There are five ways the framework can adapt its behaviour to demonstrate
+The framework has five ways that it can adapt its behaviour to demonstrate
 how effective each optimisation is:
@@
         chapter measures the speed of \parsername{}, not the speed of the
-        database or the disks it resides on.  The impact of storing results
-        in the database is described at the beginning of
+        database or the disks it resides on.  The effect on parsing time of
+        storing results in the database is described at the beginning of
         \sectionref{Evaluation}.
@@
         which is useful when a warning dumps a mail's data for inspection.
-        The information currently added is: the log line's timestamp in
-        human readable form; the entire log line; the name of the log file
-        and the number of the log line within it.  This information is not
+        The information that is added is: the log line's timestamp in human
+        readable form; the entire log line; the name of the log file and
+        the number of the log line within it.  This information is not
         stored in the database.
@@
 is not useful for most users: it is for debugging problems with the data
-that is saved to the database, and there are sufficient safeguards in place
-that there should be plenty of warning messages explaining what is wrong
-without using this option.
+that is saved to the database, and sufficient safeguards are in place that
+there should be plenty of warning messages explaining what is wrong without
+using this option.
 
@@
 
-There is no correlation between how often an action is specified by rules
-and how often an action is invoked because a rule recognises a log line.
+No correlation exists between how often an action is specified by rules and
+how often an action is invoked because a rule recognises a log line.
 \Graphref{Number of times each action was invoked excluding 22 and 62--68}
@@
 of times: almost every \action{CONNECT} will have a \action{DISCONNECT},
-and there is only a 0.00042\% difference between the number of times the
-two were invoked; every mail that enters Postfix's queue will be managed by
+with only a 0.00042\% difference between the number of times the two were
+invoked; every mail that enters Postfix's queue will be managed by
 \daemon{qmgr} (\action{MAIL\_QUEUED}) and processed by \daemon{cleanup}
-(\action{CLEANUP\_PROCESSING}), and again there is only a 0.01648\%
+(\action{CLEANUP\_PROCESSING}), and again the two have only a 0.01648\%
 difference between their invocations.  Confusingly,
@@
 
-If there is enough information in the log line to identify the correct
-connection or mail, each action will save all the data captured by the
-recognising rule's regex.  Each action is passed the same arguments:
+If the log line has enough information to identify the correct connection
+or mail, each action will save all the data captured by the recognising
+rule's regex.  Each action is passed the same arguments:
 
@@
         sender.  When this happens the mail will not have a combination of
-        Postfix programs which passes the valid combinations check,
+        Postfix programs that passes the valid combinations check,
         implemented to deal with the complication described in
@@
 
-    \item [MAIL\_BOUNCED] This action behaves exactly like the
-        \action{SAVE\_DATA} action; it saves all data captured by the
+    \item [MAIL\_BOUNCED] This action behaves in exactly the same way as
+        the \action{SAVE\_DATA} action; it saves all data captured by the
         recognising rule's regex, and does nothing more.  It is a separate
@@
 
-    \item [MAIL\_SENT] This action behaves exactly like the
+    \item [MAIL\_SENT] This action behaves in exactly the same way as the
         \action{SAVE\_DATA} action; it saves all data captured by the
@@
         informed of the problem.  The mail may have been accepted and
-        partially transferred, in which case there's a data structure to
-        dispose of.  See \sectionref{timeouts during data phase} for full
-        details; although that describes timeouts the processing is the
-        same for mails that are too large.
+        partially transferred; if so a data structure must be disposed of.
+        See \sectionref{timeouts during data phase} for full details;
+        although that describes timeouts the processing is the same for
+        mails that are too large.
 
@@
         signal, or exits unsuccessfully; the associated connection must be
-        cleaned up, entered in the database if it has sufficient data, and
-        deleted from the state tables.  In some cases the connection will
-        not have enough data to satisfy the database schema, so it cannot
-        be entered into the database for future analysis; unfortunately
-        this means that the small amount of data that has been gathered
-        will be lost.
+        cleaned up, entered in the database if it has enough data, and
+        deleted from the state tables.  Sometimes the connection will not
+        have enough data to satisfy the database schema, so it cannot be
+        entered into the database for future analysis; unfortunately this
+        means that the small amount of data that has been gathered will be
+        lost.
 
@@
     \item [TIMEOUT] The connection with the remote client timed out, so the
-        mail currently being transferred must be discarded.  The mail may
-        have been accepted, in which case there's a data structure to
-        dispose of.  See \sectionref{timeouts during data phase} for full
-        details.
+        mail being transferred must be discarded.  The mail may have been
+        accepted: if so there is a data structure to dispose of.  See
+        \sectionref{timeouts during data phase} for full details.
 
@@
         \action{TRACK} checks if the child has already been tracked, either
-        by this parent or by another parent, and issues appropriate
-        warnings in either case.
+        with this parent or with another parent, and issues appropriate
+        warnings if so.
 
@@
 would only be necessary if the new action processes log lines from Postfix
-components \parsername{} currently does not have rules for, e.g.\
+components that \parsername{} does not have rules for, e.g.\
 \daemon{virtual} or \daemon{lmtp}.
@@
         actual recognition of the log lines, and data captured by the regex
-        (sender, recipient, etc.) will be automatically saved to the
+        (e.g.\ sender, recipient) will be automatically saved to the
         results and connections tables.
@@
 because administrators have the freedom to choose the subset of available
-restrictions which suits their needs, including using different
+restrictions that suits their needs, including using different
 \acronym{DNSBL} services, policy servers, or custom rejection messages.  To
-facilitate parsing new log lines, the architecture separates rules from
-actions: adding new actions can be difficult, but adding new rules to
+ease the process of parsing new log lines, the architecture separates rules
+from actions: adding new actions can be difficult, but adding new rules to
 recognise new log lines is trivial, and occurs much more frequently.
@@
 
-There are other attributes that may be required to make a rule work
-correctly: connection\_data, result\_data, priority, or restriction\_name.
-In general it will only become clear that connection\_data or result\_data
-are required when \parsername{} warns about an entry in the connections or
+Other attributes may be required to make a rule work correctly:
+connection\_data, result\_data, priority, or restriction\_name.  In general
+it will only become clear that connection\_data or result\_data are
+required when \parsername{} warns about an entry in the connections or
 results tables that is missing some required fields, because values for the
@@
 the restriction\_name field should be set when the rule's action is
-\action{DELIVERY\_REJECTED}; hopefully the name of the restriction will be
-clear from the content of the log line.
+\action{DELIVERY\_REJECTED}; the name of the restriction should be clear
+from the content of the log line.
 
@@
 fixed pattern.  This classification can be used to group similar log lines
-and generate a regex to match each group of log lines.  There are five steps
-in the algorithm:
+and generate a regex to match each group of log lines.  The algorithm has
+five steps:
 
@@
         knowledge gained when writing rules and developing \parsername{};
-        it performs a large number of substitutions on the input log lines,
-        replacing commonly occurring variable terms (e.g.\ email addresses,
+        it performs many substitutions on the input log lines, replacing
+        commonly occurring variable terms (e.g.\ email addresses,
         \acronym{IP} addresses, the standard start of rejection messages)
@@
         purpose of this step is to utilise existing knowledge to create
-        more accurate regexes; it replaces a large number of variable words
-        with fixed words, improving the subsequent classification of words
-        as fixed or variable.  The altered log lines are written to a
-        temporary file, which subsequent stages will use instead of the
-        original input file.
+        more accurate regexes; it replaces many variable words with fixed
+        words, improving the subsequent classification of words as fixed or
+        variable.  The altered log lines are written to a temporary file,
+        which subsequent stages will use instead of the original input
+        file.
 
@@
         variable term.  If a variable term appears sufficiently often it
-        will be misclassified as a fixed term, but that will hopefully be
-        noticed by the user when reviewing the new regexes, and will be
-        obvious when the new rules do not recognise some log lines they are
+        will be misclassified as a fixed term, but that should be noticed
+        by the user when reviewing the new regexes, and will be obvious
+        when the new rules do not recognise some log lines they are
         expected to.  Variable terms are replaced by \texttt{.+} to match
@@
 \sectionref{rule conditions in architecture}, and they can be very complex
-and difficult to evaluate.  In contrast, \parsername{} uses very simple
+and difficult to evaluate.  In contrast, \parsername{} uses quite simple
 rule conditions: each rule has a \texttt{program} attribute that specifies
@@
 recompiled each time it was used, resulting in a large, data dependent
-slowdown, as shown in \sectionref{Caching compiled regexes}.  Most of the
-keywords are named after the fields in the connections or results tables
-they populate: \_\_CLIENT\_HOSTNAME\_\_, \_\_CLIENT\_IP\_\_, \_\_DELAY\_\_,
-\_\_DELAYS\_\_, \_\_ENHANCED\_STATUS\_CODE\_\_, \_\_HELO\_\_,
-\_\_MESSAGE\_ID\_\_, \_\_QUEUEID\_\_, \_\_RECIPIENT\_\_, \_\_SENDER\_\_,
-\_\_SERVER\_HOSTNAME\_\_, \_\_SERVER\_IP\_\_, \_\_SIZE\_\_, and
-\_\_SMTP\_CODE\_\_.
+slowdown, as described in \sectionref{Caching compiled regexes}.  Most of
+the keywords are named after the fields in the connections or results
+tables they populate: \_\_CLIENT\_HOSTNAME\_\_, \_\_CLIENT\_IP\_\_,
+\_\_DELAY\_\_, \_\_DELAYS\_\_, \_\_ENHANCED\_STATUS\_CODE\_\_,
+\_\_HELO\_\_, \_\_MESSAGE\_ID\_\_, \_\_QUEUEID\_\_, \_\_RECIPIENT\_\_,
+\_\_SENDER\_\_, \_\_SERVER\_HOSTNAME\_\_, \_\_SERVER\_IP\_\_, \_\_SIZE\_\_,
+and \_\_SMTP\_CODE\_\_.
 
@@
 
-There is some similarity between regex components and cascaded parsing:
-each regex component resembles a rule that recognises part of a log line
-and consumes it, leaving the remainder to be recognised by other components
-or cascaded rules.  The major difference between the two is that regex
+Some similarity exists between regex components and cascaded parsing: each
+regex component resembles a rule that recognises part of a log line and
+consumes it, leaving the remainder to be recognised by other components or
+cascaded rules.  The major difference between the two is that regex
 components are explicitly used by the author of the ruleset, whereas
@@
         Repeat with shuffled and reversed ordering, starting with a fresh
-        database.  If there are no overlapping rules the tables from each
-        run will be identical; differences indicate overlapping rules.
-        The rules that overlap can be determined by examining the
-        differences in the tables: each result contains a reference to the
-        rule which created it, which will change if that rule overlaps with
-        another.  Unfortunately this method cannot prove the absence of
-        overlapping rules; it can detect overlapping rules, but only if
-        there are log lines in the log files that are recognised by more
-        than one rule.
+        database.  If the ruleset does not have overlapping rules the
+        tables from each run will be identical; differences indicate
+        overlapping rules.  The rules that overlap can be determined by
+        examining the differences in the tables: each result contains a
+        reference to the rule which created it, which will change if that
+        rule overlaps with another.  Unfortunately this method cannot prove
+        the absence of overlapping rules; it can detect overlapping rules,
+        but only if the log files have log lines that are recognised by
+        more than one rule.
 
@@
 
-Every piece of software suffers from some limitations, and there is almost
-always room for improvement.  Below are the limitations and possible
+Every piece of software suffers from some limitations, and almost always
+has room for improvement.  Below are the limitations and possible
 improvements that have been identified in \parsername{}.
@@
         delivery attempt, so the log line cannot be associated with the
-        correct mail.  There is guaranteed to be a queueid when the DATA
-        command has been reached, and thus the queueid will be logged by
-        any restrictions taking effect in
+        correct mail.  A queueid is guaranteed to have been allocated when
+        the DATA command has been reached, and thus the queueid will be
+        logged by any restrictions taking effect in
         \texttt{smtpd\_data\_restrictions}, and the log line can be
-        associated with the correct mail.  There is no difficulty in
-        specifying a HELO-based restriction in
-        \texttt{smtpd\_data\_restrictions}, Postfix will perform the check
-        correctly.
+        associated with the correct mail.  Specifying a HELO-based
+        restriction in \texttt{smtpd\_data\_restrictions} does not cause
+        any problems; Postfix will perform the check correctly.
 
@@
         the complication described in \sectionref{Mail deleted before
-        delivery is attempted} from occurring, but only when there is a
-        single recipient; in that case the recipient address will be logged
-        also, but when there are multiple recipients no addresses are
-        logged.
+        delivery is attempted} from occurring, but only when the mail has a
+        single recipient; when a mail has a single recipient address it
+        will be logged, but when a mail has multiple recipients no
+        addresses are logged.
 
@@
         files from different sources can be combined in the same database,
-        or it may be viewed as a limitation because there is no facility to
+        or it may be viewed as a limitation because it is impossible to
         distinguish between log files from different sources in the same

### 2009-04-12 doc/logparser-introduction.tex
Ran diction and made changes based on its suggestions.

@@
 \urlLastChecked{http://www.postfix.org/}{2009/02/23}.}  Further
-complication arises because the set of log lines is not fixed, and there
-are many ways in which log lines may differ between servers, even within
-the same organisation, where servers may be configured differently or even
-running different versions of Postfix.
+complication arises because the set of log lines is not fixed, and log
+lines can differ in many ways between servers, even within the same
+organisation, where servers may be configured differently or running
+different versions of Postfix.
 
@@
 presenting an architecture designed to enable the users of a parser to
-easily extend it to parse their particular inputs, without requiring a lot
-of work or a high level of understanding of the parsing process and the
+easily extend it to parse their particular inputs, without requiring much
+work or a high level of understanding of the parsing process and the
 parser's internal workings.  This architecture is the basis of

### 2009-04-12 doc/logparser-other-parsers-review.tex
Removed some unused references.
Changes from proofreading of introduction chapter.
Consistent float placement.
Ran diction and made changes based on its suggestions.

@@
-\chapter{State of the Art Review}
+\chapter{State Of The Art Review}
 
@@
 
-\label{prior art}
+\label{log mail analyser}
 
@@
 into the database and provide commands for displaying it.  Berkeley DB does
-not embed a schema in its files, because there is neither requirement nor
-benefit; it only provides \textbf{(key, value)} pairs, so any additional
-structuring of the data is imposed by the application, thus the application
-must document this structure.  MySQL support was not tested because the
-schema required is not documented.
+not embed a schema in its files, because it does not need one, and
+embedding one would not provide any benefit; it only provides \textbf{(key,
+value)} pairs, so any additional structuring of the data is imposed by the
+application, thus the application must document this structure.  MySQL
+support was not tested because the schema required is not documented.
 
@@
 % an example is the difficulty in searching a pipe-delimited string, so
-% they have re-implemented foreign keys with tables keyed by ip address
-% etc.\ pointing at the main table - this also will not scale well.  There
+% they have re-implemented foreign keys with tables keyed by ip address and
+% so on pointing at the main table - this also will not scale well.  There
 % does not appear to be any attempt to deal with the complications I have
@@
 
-\section{Sawmill Universal Log File Analysis and Reporting}
+\section{Sawmill Universal Log File Analysis And Reporting}
 
@@
 though most are variations of a bar or pie chart, except the bubble and
-heatmap graphs.  It is easy to drill down through the graphs to select a
+heatmap graphs.  It is trivial to drill down through the graphs to select a
 portion of the data (e.g.\ select the hour with the largest number of
@@
 Copious documentation is made available on
-\urlLastChecked{http://www.splunk.com/}{2008/11/23}, but the organisation
-and abundance of material does not support easy identification of useful
-information.  Searches confusingly tended to return results from old
+\urlLastChecked{http://www.splunk.com/}{2008/11/23}, but the abundance of
+material and lack of organisation makes it hard to find the topic being
+sought.  Searches confusingly tended to return results from old
 documentation rather than new.  In general the documentation appears to
@@
 so the author could neither understand what data is extracted, nor
-determine what data is missing in comparison to \parsername{} (which fully
-documents everything it extracts).  AWStats coerces Postfix log files into
-the log file format used by the Apache web server, for analysis by AWStats'
-HTTP log file parser.  The converting parser only deals with a small
-portion of the log lines generated by Postfix, silently skipping those it
-cannot deal with, and does not distinguish between different types of
-rejection; extending it to parse all log lines would be at least as much
+determine what data is missing in comparison to \parsername{}, which fully
+documents all data that it extracts.  AWStats coerces Postfix log files
+into the log file format used by the Apache web server, for analysis by
+AWStats' HTTP log file parser.  The converting parser only deals with a
+small portion of the log lines generated by Postfix, silently skipping
+those it cannot deal with, and does not distinguish between different types
+of rejection; extending it to parse all log lines would be at least as much
 work as writing a new parser.  It does correlate log lines by queueid (not
@@
 would certainly be out of date because Postfix has evolved considerably
-since this tool was last updated (2003/11/06).  As it neither ran
+since this tool was last updated (2003/11/06).  Because it neither ran
 successfully nor has documentation the author can read, a detailed review
@@
     deliveries and the volume of email delivered by day, the domains from
-    which you receive or send the most emails, the relays most used, etc.
+    which you receive or send the most emails, the relays most used,
+    etc\empty{}.
 
@@
 \parsername{} uses an \acronym{SQL} database to make accessing the
-extracted data as easy as possible.  In general, Lire has poor
+extracted data as effortless as possible.  In general, Lire has poor
 documentation.
@@
 reviewed parsers collect the breadth of information gathered by
-\parsername{}, or make it as easy to extend the parser to handle new log
+\parsername{}, and are not designed to be extensible to handle new log
 lines.  Some correlate log lines by queueid (none correlate by pid); none
@@
 Most but not all of the parsers provide documentation, with the quality
-ranging from unusable to very good.  Fewer than half of the parsers were
+ranging from unusable to excellent.  Fewer than half of the parsers were
 capable of parsing the \numberOFlogFILES{} test log files; improving or
@@
 
-\begin{table}[htb]
-    \caption{Summary of parsers' features}
+\begin{table}[thbp]
+    \caption{Summary of reviewed parsers' features}
     \empty{}\label{Summary of parsers' features}
@@
         Pflogsumm       & Yes           & No                & Partial \dag{}    & Good           & Yes          \\
-        Sawmill         & Yes           & Yes               & Searches          & Very good      & \nialpha{}   \\
+        Sawmill         & Yes           & Yes               & Searches          & Excellent      & \nialpha{}   \\
         Splunk          & Yes           & Yes               & Searches          & Abundant       & No           \\

### 2009-04-12 doc/logparser-postfix-daemons.tex
Ran diction and made changes based on its suggestions.

@@
     includes alias expansion, processing of a user's \texttt{.forward}
-    file, and delivery of the mail, whether it be to a user's mailbox or a
+    file, and delivery of the mail, whether to a user's mailbox or a
     program such as a mailing list manager.

### 2009-04-12 doc/logparser-results.tex
Ran diction and made changes based on its suggestions.

@@
 optimisation of caching the results of compiling regexes, and the huge
-impact it has on parser efficiency, is the penultimate topic to be covered.
+effect it has on parser efficiency, is the penultimate topic to be covered.
 The performance evaluation concludes by examining where the parsing time is
@@
 gathering statistics from test runs, and did not run any other programs
-while test runs were ongoing.  Saving results to the database was disabled
-for the test runs, because that dominates the run time of the parser, and
-the tests are aimed at measuring the speed of \parsername{} rather than the
-speed of the database and the disks the database is stored on.  Parsing all
-\numberOFlogFILES{} log files in one run without saving results to the
-database took \input{build/include-full-run-duration}, with mean throughput
-of \input{build/include-full-run-throughput}.  In contrast, when saving
-results to the database, parsing all \numberOFlogFILES{} log files took
-\input{build/include-insert-results-duration}, with mean throughput of
-\input{build/include-insert-results-throughput} --- parsing makes up only
+while test runs were in progress.  Saving results to the database was
+disabled for the test runs, because that dominates the run time of the
+parser, and the tests are aimed at measuring the speed of \parsername{}
+rather than the speed of the database and the disks the database is stored
+on.  Parsing all \numberOFlogFILES{} log files in one run without saving
+results to the database took \input{build/include-full-run-duration}, with
+mean throughput of \input{build/include-full-run-throughput}.  In contrast,
+when saving results to the database, parsing all \numberOFlogFILES{} log
+files took \input{build/include-insert-results-duration}, with mean
+throughput of \input{build/include-insert-results-throughput} --- parsing
+makes up only
 \input{build/include-skip-inserting-as-percentage-of-inserting} of the
@@
 larger than usual size.  Both groups of log files are much larger than
-usual due to a mail loop caused by a user who set up mail forwarding
+usual because of a mail loop caused by a user who set up mail forwarding
 incorrectly, resulting in a very different distribution of log lines:
@@
 differences in rule ordering have less effect on parsing time when parsing
-log files 22 \& 62--68, due to the different distribution of log lines in
-those log files.  A careful examination of \graphref{Parsing time relative
-to shuffled ordering graph} shows that, for the first log file, optimal and
-reverse orderings perform identically: this is because the hits field of
-each rule is zero for the first log file, so optimal and reverse orderings
-produce identical rule orderings.  For the first log file, shuffled
-ordering is the most efficient of the three, but that is accidental and
-cannot be relied upon.
+log files 22 \& 62--68, because of the different distribution of log lines
+in those log files.  A careful examination of \graphref{Parsing time
+relative to shuffled ordering graph} shows that, for the first log file,
+optimal and reverse orderings perform identically: this is because the hits
+field of each rule is zero for the first log file, so optimal and reverse
+orderings produce identical rule orderings.  For the first log file,
+shuffled ordering is the most efficient of the three, but that is
+accidental and cannot be relied upon.
 
@@
 but it is particularly important in this architecture because it is
-expected that the typical parser will have a large number of rules.  There
-are \numberOFrules{} rules in the full \parsername{} ruleset, whereas the
-minimum number of rules required to parse the \numberOFlogFILES{} log files
-is \numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
+expected that the typical parser will have a large ruleset.  The full
+\parsername{} ruleset has \numberOFrules{} rules, whereas the minimum
+number of rules required to parse the \numberOFlogFILES{} log files is
+\numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
 ruleset.  The full ruleset is larger because \parsername{} is tested with
@@
 
-Clearly the increased number of rules has a noticeable performance impact
-with reverse ordering, and a lesser impact with shuffled ordering, whereas
-optimal ordering shows very little change.  Log files 22 \& 62--68 show
-much smaller increases in parsing time than other log files do, because the
-majority of log lines in those log files are produced by Postfix components
-with a small number of rules, so removing unnecessary rules has little
-effect on the total number of rules used; \tableref{Number of rules per
-Postfix component in the maximum and minimum rulesets} shows the number of
-rules per Postfix component for each ruleset.  Once again, for the first
-log file, optimal and reverse orderings have identical performance, because
-the hits field of every rule is zero.
+Clearly the increased number of rules causes a noticeable performance
+decrease with reverse ordering, and a lesser decrease with shuffled
+ordering, whereas optimal ordering shows scant change.  Log files 22 \&
+62--68 show much smaller increases in parsing time than other log files do,
+because most of the log lines in those log files are produced by Postfix
+components with few rules, so removing unnecessary rules has little effect
+on the total number of rules used; \tableref{Number of rules per Postfix
+component in the maximum and minimum rulesets} shows the number of rules
+per Postfix component for each ruleset.  Once again, for the first log
+file, optimal and reverse orderings have identical performance, because the
+hits field of every rule is zero.
 
-The optimal ordering shows a mean increase of just
+The optimal ordering has a mean increase of just
 \input{build/include-full-ruleset-vs-minimum-ruleset-mean} in parsing time
 for a \numberOFrulesMAXIMUMpercentage{} increase in the number of rules.
-These results show that both the architecture and \parsername{} scale very
-well as the number of rules increases, and that optimally sorting the rules
-is an important optimisation contributing to this scalability.
+These results show that both the architecture and \parsername{} scale
+extremely well as the number of rules increases, and that optimally sorting
+the rules is an important optimisation contributing to this scalability.
 
@@
 \Graphref{Increase in parsing time when not caching compiled regexes graph}
-shows the impact that not caching compiled regexes has on parser
+shows the effect that not caching compiled regexes has on parser
 performance, with mean and standard deviation in \tableref{Increase in
@@
 The optimisations described in this chapter have optimised the process of
-recognising log lines, but have not optimised actions at all.  There are
-two reasons for this:
+recognising log lines, but have not optimised actions at all.  Optimisation
+efforts have concentrated on recognition of log lines for two reasons:
 
@@
         but it would be vastly harder to implement; optimising actions
-        would not provide a sufficient reduction in parsing time to justify
-        the amount of effort required.
+        would not provide enough reduction in parsing time to justify the
+        amount of effort required.
 
@@
 Individual actions or the framework could and have been optimised, but
-there is plenty of existing literature on optimising programs, so the
-subject will not be dealt with here.
+plenty of existing literature is available on the topic of optimising
+programs, so the subject will not be dealt with here.
 
@@
 because there may be any number of log lines from other programs
-intermingled in the log file, and there are some Postfix components that do
-not produce any log lines of interest.  There must be a rule to recognise
-each log line variant produced by each Postfix component; if a log line is
-not recognised the parser will issue a warning, to inform the user that
-they need to extend their ruleset.  Currently, \parsername{} does not parse
-log lines from non-Postfix programs, e.g.\ Amavisd-new or SpamAssassin; it
-could easily be extended to do so, if a method could be developed to
-correctly associate such log lines with existing state table entries.  Each
-rule's regex should be as specific and precise as possible, to ensure
-accurate parsing: a rule with a regex that matches zero or more of any
-character will recognise every log line, but not in a meaningful way.  Most
-log lines contain fixed strings, so this is not a problem in practice.
+intermingled in the log file, and some Postfix components that do not
+produce any log lines of interest.  There must be a rule to recognise each
+log line variant produced by each Postfix component; if a log line is not
+recognised the parser will issue a warning, to inform the user that they
+need to extend their ruleset.  \parsername{} does not parse log lines from
+non-Postfix programs, e.g.\ Amavisd-new or SpamAssassin; it could easily be
+extended to do so, if a method could be developed to correctly associate
+such log lines with existing state table entries.  Each rule's regex should
+be as specific and precise as possible, to ensure accurate parsing: a rule
+with a regex that matches zero or more of any character will recognise
+every log line, but not in a meaningful way.  Most log lines contain fixed
+strings, so this is not a problem in practice.
 
-Full coverage of log lines is easy to achieve, yet hard to maintain.  It is
-easy to achieve full coverage for a limited set of log files; \parsername{}
-has \numberOFrules{} rules, fully parsing \numberOFlogFILESall{} log files
-from Postfix 2.2 through to Postfix 2.5.  Maintaining full coverage is hard
-because log lines change over time, e.g.\ administrators add restrictions
-with custom messages, \acronym{DNSBL} messages change, or major releases of
-Postfix change log lines (usually adding more information).  Warnings are
-issued for any log lines that are not recognised; no warnings are issued
-for unrecognised log lines while parsing the \numberOFlogFILES{} log files,
-so it can be safely concluded that there are zero false negatives.  False
-positives are harder to quantify: short of examining each of the 60,721,709
-log lines and determining which rule recognised it, there is no way to be
-sure that every log line was recognised by the correct rule, making it
-impossible to quantify the false positive rate.  However, a random sample
-of 6039 log lines was parsed, and the results manually verified by
-inspection to ensure that the correct rule recognised each log line.  The
-sample was generated by running the command: \texttt{perl -n -e
+Full coverage of log lines can be achieved without undue effort, yet it is
+hard to maintain.  Maintaining full coverage is hard because log lines
+change over time, e.g.\ administrators add restrictions with custom
+messages, \acronym{DNSBL} messages change, or major releases of Postfix
+change log lines (usually adding more information).  Warnings are issued
+for any log lines that are not recognised; no warnings are issued for
+unrecognised log lines while parsing the \numberOFlogFILES{} log files, so
+it can be safely concluded that zero false negatives arise.  False
+positives are harder to quantify, short of examining each of the 60,721,709
+log lines and ensuring that the correct rule recognised it.  However, a
+random sample of 6039 log lines was parsed, and the results manually
+verified by inspection to ensure that the correct rule recognised each log
+line.  The sample was generated by running the command: \texttt{perl -n -e
 \singlequote{}print if (rand 1 < 0.0001)\singlequote{} LOG\_FILES} to
@@
 in the state tables, or when a mail does not include sufficient data to
-satisfy the database schema.  There should be few or no warnings during
-parsing, and when finished parsing the state tables should only contain
-entries for mails that have log lines in subsequent log files.  There will
-often be warnings about a missing queueid or \acronym{pid} in the first few
-thousand log lines, because the earlier log lines for those connections or
-mails are in a previous log file; loading saved state tables from the
-previous log file will solve this problem.
+satisfy the database schema.  The parser should produce few or no warnings
+during parsing, and when finished parsing the state tables should only
+contain entries for mails that have log lines in subsequent log files.
+There will often be warnings about a missing queueid or \acronym{pid} in
+the first few thousand log lines, because the earlier log lines for those
+connections or mails are in a previous log file; loading saved state tables
+from the previous log file will solve this problem.
 
 The data used in this chapter is generated by parsing \numberOFlogFILES{}
-log files.  There are 5 warnings produced, but because \parsername{} errs
-on the side of producing more warnings rather than fewer, those 5 warnings
+log files.  5 are warnings produced, but because \parsername{} errs on the
+side of producing more warnings rather than fewer, those 5 warnings
 represent 3 instances of 1 problem: 3 connections that started before the
 first log file, so their initial log lines are missing, leading to warnings
-when their remaining log lines are parsed.  There are no false positives.
+when their remaining log lines are parsed.  None of the warnings are false
+positives.
 
@@
     \item 9 mails whose initial log lines were not present in the log
-        files.  There were no warnings for these mails because they were
-        resemble child mails waiting to be tracked with a parent, see
+        files.  These mails did not produce warnings because they resemble
+        child mails waiting to be tracked with a parent; see
         \sectionref{Re-injected mails} for details.
 
-    \item 7 mails that had yet to be delivered due to repeated failures.
+    \item 7 mails that had yet to be delivered because of repeated
+        failures.
 
@@
 
-There are no mails in the state tables which should not be present, thus it
-can be concluded that there are zero false negatives.  Once again,
-determining the false positive rate is much harder, as manually checking
-the results of parsing 13,850,793 connections and mails accepted, rejected,
-bounced, or delivered is infeasible.  There is considerable evidence that
-the false positive rate is extremely low:
+None of the mails in the state tables should not be present, thus it can be
+concluded that zero false negatives occur when parsing the
+\numberOFlogFILES{} log files.  Once again, determining the false positive
+rate is much harder, as manually checking the results of parsing 13,850,793
+connections and mails accepted, rejected, bounced, or delivered is
+infeasible.  Considerable evidence exists that the false positive rate is
+extremely low:
 
@@
 debugging lines.  Those numbers are approximate, and may vary $\pm{}$ 2.
-There is a linear relationship between the number of log lines and
-debugging lines: $33(connections) + 30(accepted~~mails) + 21(results)$.
-This formula is an approximation only, and has not been rigorously
-verified.}  All 167,448 lines were examined in conjunction with the log
-file segment and a dump of the resulting database, verifying that for each
-of the log lines \parsername{} used the correct rule and invoked the
-correct action, which in turn produced the correct result, and the correct
-data was inserted in the database.  The log file segment produced 4
-warnings, 10 mails remaining in the state tables, 1625 mail delivery
-attempts correctly entered in the database, zero false positives, and zero
-false negatives.
+An approximate linear relationship between the number of log lines and
+debugging lines is: $33(connections) + 30(accepted~~mails) + 21(results)$.}
+All 167,448 lines were examined in conjunction with the log file segment
+and a dump of the resulting database, verifying that for each of the log
+lines \parsername{} used the correct rule and invoked the correct action,
+which in turn produced the correct result, and the correct data was
+inserted in the database.  The log file segment produced 4 warnings, 10
+mails remaining in the state tables, 1625 mail delivery attempts correctly
+entered in the database, zero false positives, and zero false negatives.
 
@@
 files.  The framework optimises the order in which rules are used when
-trying to recognise each log line, and the effect that optimisation has is
-explored; this is followed by the impact that adding more rules to the
-ruleset has on parser performance.  The simplest and most effective of the
-optimisations, caching compiled regexes, is described next, and the
-efficiency evaluation concludes with an examination of where parsing time
-is spent: recognising log lines or processing them?
+trying to recognise each log line, and the effect that optimisation has on
+parsing time is explored; this is followed by the effect that adding more
+rules to the ruleset has on parser performance.  The simplest and most
+effective of the optimisations, caching compiled regexes, is described
+next, and the efficiency evaluation concludes with an examination of where
+parsing time is spent: recognising log lines or processing them?
 

### 2009-04-12 doc/logparser-acronyms.tex
Made corrections based on Ariane's comments for chapters 1 and 2. Acronyms are now expanded the first time they're used in each chapter, rather than just the first time they're used in the thesis.

@@
     with it.  A detailed description of SMTP is beyond the scope of this
-    thesis: the original protocol definition is in RFC~821~\cite{RFC821},
-    updated in RFC~2821~\cite{RFC2821}; an introductory guide can be found
-    at
-    \urlLastChecked{http://db.glug-bom.org/lug-authors/philip/docs/mail-stuff/smtp-intro.html}{2009/02/28}
-    if more information is needed.%
+    thesis: the original protocol definition is in
+    \acronym{RFC}~821~\cite{RFC821}, which was updated in
+    \acronym{RFC}~2821~\cite{RFC2821}.%
 }

### 2009-04-12 doc/logparser-background.tex
Made corrections based on Ariane's comments for chapters 1 and 2. Acronyms are now expanded the first time they're used in each chapter, rather than just the first time they're used in the thesis.

@@
 
-This section provides background information helpful in understanding the
+This chapter provides background information helpful in understanding the
 remainder of this thesis.  It begins with a discussion of the motivation
@@
 
-\SMTPglossaryDescription{} Despite the simplicity of the protocol many
+\SMTPglossaryDescription{} Despite the simplicity of the protocol, many
 virus and spam sending programs fail to implement it properly, so requiring
@@
 have done its best to deal with deficient clients, with the intention of
-accepting as much mail sent to its users as possible, e.g.\ by ignoring the
-absence of a HELO command, or by accepting sender or recipient addresses
-that were not enclosed in \texttt{<>}.
+accepting as much mail sent to its users as possible, e.g.\ by accepting
+sender or recipient addresses that were not enclosed in \texttt{<>}, or by
+ignoring the absence of a HELO command.  The HELO command is the first
+command sent by the client in the \acronym{SMTP} conversation; it is
+required by the protocol, but its original purpose was to detect a server
+trying to deliver mail to itself, so some implementations omitted it.
 

### 2009-04-12 doc/logparser-introduction.tex
Made corrections based on Ariane's comments for chapters 1 and 2. Acronyms are now expanded the first time they're used in each chapter, rather than just the first time they're used in the thesis.

@@
 Portions of chapters~\ref{introduction} and~\ref{parser
-architecture}--\ref{Evaluation} have previously been published at a
-conference~\cite{sgai-2008}, and later reprinted in a
+architecture}--\ref{Evaluation} have previously been published at an
+international conference~\cite{sgai-2008}, and later reprinted in a
 journal~\cite{elsevier-2009}.  Publication of the conference paper was

### 2009-04-12 doc/logparser-results.tex
Made corrections based on Ariane's comments for chapters 1 and 2. Acronyms are now expanded the first time they're used in each chapter, rather than just the first time they're used in the thesis.

@@
 enabled, resulting in 167,448 lines of output.\footnote{A mean of 27.908
-lines of output per line of input; each connection has 30 debugging lines,
-plus 21 debugging lines per result.  Connections which have been cloned
-will have the cloned connection in their debugging output, plus another 33
+lines of output per log line; each connection has 30 debugging lines, plus
+21 debugging lines per result.  Connections which have been cloned will
+have the cloned connection in their debugging output, plus another 33
 debugging lines.  Those numbers are approximate, and may vary $\pm{}$ 2.

### 2009-04-13 doc/logparser-complications.tex
Some minor changes.

@@
         line for this mail?  Every mail that passes normally through
-        Postfix will have a \daemon{cleanup} line, and later a
-        \daemon{qmgr} log line; lack of a \daemon{cleanup} line is a sure
-        sign the mail did not make it too far.
+        Postfix will have a \daemon{cleanup} log line, and later a
+        \daemon{qmgr} log line; lack of a \daemon{cleanup} log line is a
+        sure sign the mail did not make it too far.
 
@@
 
-During the \action{TIMEOUT} action, if the \daemon{cleanup} line is still
-pending, the action updates the \texttt{timeout\_queueids} state table,
-adding the queueid and the timestamp from the log line.  When the next
-\daemon{cleanup} line is parsed for that queueid, the list will be checked
-and if the log line meets the following requirements it will be deemed part
-of the mail where the timeout occurred and discarded.
+During the \action{TIMEOUT} action, if the \daemon{cleanup} log line is
+still pending, the action updates the \texttt{timeout\_queueids} state
+table, adding the queueid and the timestamp from the log line.  When the
+next \daemon{cleanup} log line is parsed for that queueid, the list will be
+checked and if the log line meets the following requirements it will be
+deemed part of the mail where the timeout occurred and discarded.
 

### 2009-04-13 doc/logparser-other-parsers-review.tex
Some minor changes.

@@
 neither necessary nor beneficial with Splunk.  Subsequently the author was
-successful in adding a single file at a time using the \acronym{CLI} and
-the command ``\texttt{splunk add tail }\textit{filename\/}''.  A simple
+successful in adding a single log file at a time using the \acronym{CLI}
+and the command ``\texttt{splunk add tail }\textit{filename\/}''.  A simple
 loop using that command was enough to add all the desired log files.
@@
 delivery delays, and message ID), though it does not make the full text of
-the line available (this could be trivially added if desired, but would
+the log line available (this could be trivially added if desired, but would
 greatly increase the size of the resulting database).  The full power of

### 2009-04-13 doc/logparser-results.tex
Some minor changes.

@@
 input size: does it scale linearly, polynomially, or exponentially?
-\Graphref{parsing time vs file size vs number of log lines graph} shows the
-parsing time in seconds, file size in MB, and number of log lines in tens
-of thousands, for each of the \numberOFlogFILES{} log files.  The three
-lines run roughly in parallel, giving the impression that the algorithm
-scales linearly with input size.  This impression is borne out by
-\graphref{parsing time vs file size vs number of log lines factor}, which
-plots both the ratio of file size vs parsing time, and the ratio of number
-of log lines vs parsing time (higher is better in both cases);
-\tableref{parsing time vs file size vs number of log lines factor table}
-shows the same ratios for different groups of log files.  The ratios are
-quite tightly banded, showing that the algorithm scales linearly; the
+\Graphref{parsing time vs log file size vs number of log lines graph} shows
+the parsing time in seconds, log file size in MB, and number of log lines
+in tens of thousands, for each of the \numberOFlogFILES{} log files.  The
+three lines run roughly in parallel, giving the impression that the
+algorithm scales linearly with input size.  This impression is borne out by
+\graphref{parsing time vs log file size vs number of log lines factor},
+which plots both the ratio of log file size vs parsing time, and the ratio
+of number of log lines vs parsing time (higher is better in both cases);
+\tableref{parsing time vs log file size vs number of log lines factor
+table} shows the same ratios for different groups of log files.  The ratios
+are quite tightly banded, showing that the algorithm scales linearly; the
 ratios increase (i.e.\ improve) for log files 22 \& 62--68, despite their
@@
 \showtable{build/include-file-size-and-number-of-log-lines-vs-parsing-time}{Ratio
-of file size and number of log lines to parsing time}{parsing time vs file
-size vs number of log lines factor table}
+of log file size and number of log lines to parsing time}{parsing time vs
+log file size vs number of log lines factor table}
 
-\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, file size,
-and number of log lines for \numberOFlogFILES{} log files}{parsing time vs
-file size vs number of log lines graph}
+\showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, log file
+size, and number of log lines for \numberOFlogFILES{} log files}{parsing
+time vs log file size vs number of log lines graph}
 
-\showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of file size
-and number of log lines to parsing time (higher is better)}{parsing time vs
-file size vs number of log lines factor}
+\showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of log file
+size and number of log lines to parsing time (higher is better)}{parsing
+time vs log file size vs number of log lines factor}
 

### 2009-04-13 doc/logparser.tex
Minor changes.
Corrections to the first coverage section.
Renamed Results chapter to Evaluation. Finished evaluation chapter introduction and summary.
Ran diction and made changes based on its suggestions.
Made corrections based on Ariane's comments for chapters 1 and 2. Acronyms are now expanded the first time they're used in each chapter, rather than just the first time they're used in the thesis.
Some minor changes.

@@
 
-% Embed SVN Id etc.
+% Embed SVN Id
 \usepackage{svn}
@@
 \hypersetup{
-    pdftitle    = {Parsing Postfix log files},
+    pdftitle    = {Parsing Postfix Log Files},
     pdfauthor   = {John Tobin},
@@
 
+% This is silly, but it keeps chktex happy.
+\newcommand{\singlequote}[0]{'}%
+
 % Constant values.
@@
 \newcommand{\numberOFrulesMINIMUMpercentage}[0]{%
-    62.50\%%
+    62.500\%%
 }
@@
 \newcommand{\numberOFrulesMAXIMUMpercentage}[0]{%
-    60.00\%%
+    60.000\%%
 }
@@
 
-\title{Parsing Postfix log files}
+\title{Parsing Postfix Log Files}
 \author{John Tobin \\ School of Computer Science and Statistics \\
@@
 
-    XXX ASK CARL\@: SHOULD I HAVE A ``FUTURE WORK'' SECTION SOMEWHERE IN
-    THE CONCLUSION\@?
-
     XXX MAIL MADS ABOUT EXAMINERS\@.
-
-    XXX SPELL CHECK AND RUN DICTION\@.
-
-    XXX MAKE IT EASY TO REDUCE THE HEIGHT OF GRAPHS\@.
 
@@
 \input{logparser-introduction}
+% Reset the list of acronyms that have been used, so they will be expanded
+% again the first time they are used.  The idea is to have them expanded
+% once per chapter; I do not reset them before the appendices, but some may
+% not have been used since they were reset, and thus will be expanded
+% again.
+\glsresetall{}
 \input{logparser-background}
+\glsresetall{}
 \input{logparser-other-parsers-review}
+\glsresetall{}
 \input{logparser-architecture}
+\glsresetall{}
 \input{logparser-implementation}
+\glsresetall{}
 \input{logparser-results}
+\glsresetall{}
 \input{logparser-conclusion}

### 2009-04-13 doc/logparser-conclusion.tex
Pulled the paper's conclusion into the thesis.

@@
 question, is extremely time consuming and error prone.  Sometimes the task
-is not to spot the line that is unusual, but to spot that a line that is
-normally present is missing, i.e.\ to realise that one line amongst
-thousands is absent.  In all cases the evidence must be used to construct a
-hypothesis to explain the irregularities, and that hypothesis must then be
-tested in the parser; if successful, the parser must be modified to deal
-with the irregularities, without adversely affecting the existing
-functionality.  The complications described in \sectionref{complications}
-were solved in the order they are described in, and that order closely
-resembles the frequency with which they occur; the most frequently
-occurring complications dominate the warning messages produced, and so
-naturally they are the first complications to be dealt with.
+is not to spot the log line that is unusual, but to spot that a log line
+that is normally present is missing, i.e.\ to realise that one log line
+amongst thousands is absent.  In all cases the evidence must be used to
+construct a hypothesis to explain the irregularities, and that hypothesis
+must then be tested in the parser; if successful, the parser must be
+modified to deal with the irregularities, without adversely affecting the
+existing functionality.  The complications described in
+\sectionref{complications} were solved in the order they are described in,
+and that order closely resembles the frequency with which they occur; the
+most frequently occurring complications dominate the warning messages
+produced, and so naturally they are the first complications to be dealt
+with.
 
@@
 software engineering disciplines.
+
+This architecture's greatest strength is the ease with which it can be
+adapted to deal with new requirements and inputs.  Parsing a variation of
+an existing input is a trivial task: simply modify an existing rule or add
+a new rule with an appropriate regex and the task is complete.  Parsing a
+new category of input is achieved by writing a new action and appropriate
+rules; quite often the new action will not need to interact with existing
+actions, but when interaction is required the framework provides the
+necessary facilities.  The decoupling of rules from actions allows
+different sets of rules to be used with the same actions, e.g.\ a parser
+might have actions to process versions one and two of a file format; by
+choosing the appropriate ruleset the parser will parse version one, or
+version two, or both versions.  Decoupling also allows other approaches to
+rule management, as discussed in \sectionref{rules in architecture}.  The
+architecture makes it possible to apply commonly used programming
+techniques (such as object orientation, inheritance, composition,
+delegation, roles, modularisation, or closures) when designing and
+implementing a parser, simplifying the process of working within a team or
+when developing and testing additional functionality.  This architecture is
+ideally suited to parsing inputs where the input is not fully understood or
+does not follow a fixed grammar: the architecture warns about unparsed
+inputs and other errors, but continues parsing as best it can, allowing the
+developer of a new parser to decide which deficiencies are most important
+and require attention first, rather than being forced to fix the first
+error that arises.
+
+The data gathered by the Postfix log parser provides the foundation for the
+future of this project: applying machine-learning algorithms to the data to
+analyse and optimise the set of anti-spam defences in use, followed by
+identifying patterns in the data that could be used to write new filters to
+recognise and reject spam rather than accepting it.  The parser provides
+the data in a normalised form that is far easier to use as input to new or
+existing algorithm implementations than trying to adapt each algorithm to
+extract data directly from the log files.  The current focus is on
+clustering and decision trees to optimise the order in which rules are
+applied; future efforts will involve using data gathered by the parser to
+train and test new filters.  This task is similar to analysing a black
+box application based on its inputs and outputs, and this approach could be
+applied to analyse the behaviour of any system given sufficient log
+messages to parse.  An alternate approach to black box optimisation that
+uses application profiling in conjunction with the application's error
+messages to improve the error messages shown to users is described
+in~\cite{black-box-error-reporting}; profiling data may be useful in
+supplementing systems that fail to provide adequate log messages.
+
+The Postfix log file parser based on this architecture provides a basis for
+systems administrators to monitor the effectiveness of their anti-spam
+measures and adapt their defences to combat new techniques used by those
+sending spam.  This parser is a fully usable application, built to address
+a genuine need, rather than a proof of concept whose sole purpose is to
+illustrate a new idea; it deals with the oddities and difficulties that
+occur in the real world, rather than a clean, idealised scenario developed
+to showcase the best features of a new approach.

### 2009-04-15 doc/logparser-results.tex
Corrections from proofreading.

@@
 
-This chapter evaluates \parsername{} on two criteria: efficiency and
-coverage.  \parsername{} is intended to be used in a production
-environment, and must be capable of parsing log files generated by a high
-volume mail server in a reasonable time period.  Accurate but slow parsing
-is preferable to sloppy but quick parsing, so the goal of better parser
-efficiency must be balanced against the requirement for precise and correct
-processing of log files.
+This chapter evaluates \parsername{} on two criteria: efficiency, and
+coverage of  Postfix log files.  \parsername{} is intended to be used in a
+production environment, and must be capable of parsing log files generated
+by a high volume mail server in a reasonable time period.  Accurate but
+slow parsing is preferable to sloppy but quick parsing, so the goal of
+better parser efficiency must be balanced against the requirement for
+precise and correct processing of log files.
 
 The performance evaluation begins by describing the characteristics of the
-mail server that produced the log files used to test and evaluate the
-parser's performance, and also the computer that the tests were run on.
-How the parser scales as the size of log files increases is described next,
-with an explanation of why \parsername{} has better performance with the
-larger log files in the group.  The effect of using different rule
-orderings is explored, and the use of optimal rule ordering is compared to
-use of an oracle that allows the parser to use only one rule when
-recognising each log line.  When \parsername{} is used by other mail
-administrators they will need to extend the ruleset to parse their own log
+mail server that produced the log files used evaluate the parser's
+performance and coverage; the computer that the tests were run on is also
+described.  The characteristics of the \numberOFlogFILES{} log files are
+described next, with an explanation of why \parsername{} has better
+performance with the larger log files in the group.  The effect of using
+different rule orderings is explored, and the use of optimal rule ordering
+is compared to use of an oracle that allows the parser to use only one rule
+when recognising each log line.  When \parsername{} is used by other mail
+administrators, they will need to extend the ruleset to parse their own log
 lines, so the next section addresses the question of how parser performance
-is affected as the number of rules in the ruleset rises.  The simple
-optimisation of caching the results of compiling regexes, and the huge
-effect it has on parser efficiency, is the penultimate topic to be covered.
-The performance evaluation concludes by examining where the parsing time is
-spent: recognition of log lines or their subsequent processing?
+is affected by an increase in the number of rules in the ruleset.  The
+penultimate topic to be covered is the simple optimisation of caching the
+results of compiling each regex, and the huge effect it has on parser
+efficiency.  The performance evaluation concludes by examining where
+parsing time is spent: recognition of log lines or their subsequent
+processing?
 
@@
 delivery attempts are correctly understood and reconstructed by the
-actions?  The former is a requirement before the latter can be achieved,
-and the latter is important because the data provided by \parsername{} must
-be both complete and correct for it to be of use to others.
+actions?  The former is a requirement for the latter to be achieved, and
+the latter is important because the data provided by \parsername{} must be
+both complete and correct for it to be of use to others.
 
@@
 
-Parsing efficiency is an obvious concern when the parser routinely needs to
-parse large log files.  The server that generated the log files used in
-testing this parser accepts approximately 10,000 mails for 700 users each
-weekday; \graphref{Mails received per day} and \tableref{Number of mails
-received per day: statistics} show that, as expected, far more mails are
-received on weekdays than at weekends.  Note that these figures count mails
-received by \acronym{SMTP} only, and the mail loops noticeable in later
-graphs are not included.  Median log file size is 50MB, containing 285,000
-log lines; large scale mail servers would have much larger log files.  The
-mail server in question is a production mail server handling mail for a
-university department; the benefit of using this server is that its log
-files exhibit the idiosyncrasies and peculiarities a mail server in the
-wild must deal with, but the downside is that significantly altering the
-configuration to accommodate this project is not an option.
+Efficiency is an obvious concern when the parser routinely needs to parse
+large log files.  The server that generated the log files used in testing
+this parser accepts approximately 10,000 mails for 700 users each weekday;
+\graphref{Mails received per day} and \tableref{Number of mails received
+per day: statistics} show that, as expected, far more mails are received on
+weekdays than at weekends.  Note that these figures only count mails
+received by \acronym{SMTP}, and the mail loops noticeable in later graphs
+do not affect the figures.  Median log file size is
+\input{build/include-median-file-size}, containing
+\input{build/include-median-number-of-log-lines} log lines; large scale
+mail servers would have much larger log files.  The mail server in question
+is a production mail server handling mail for a university department; the
+benefit of using this mail server is that its log files exhibit the
+idiosyncrasies and peculiarities a mail server in the wild must deal with,
+but the downside is that significantly altering the configuration to
+accommodate this project is not an option.
 
@@
 When generating the timing data used in this section, \numberOFlogFILES{}
-log files (totaling 10.08 GB, \numberOFlogLINEShuman{} log lines) were each
-parsed 10 times, and the mean parsing time used.  The computer used for
-test runs was a Dell Optiplex 745, shown in \tableref{Details of the
-computer used to generate statistics}; it was dedicated to the task of
-gathering statistics from test runs, and did not run any other programs
-while test runs were in progress.  Saving results to the database was
-disabled for the test runs, because that dominates the run time of the
-parser, and the tests are aimed at measuring the speed of \parsername{}
-rather than the speed of the database and the disks the database is stored
-on.  Parsing all \numberOFlogFILES{} log files in one run without saving
-results to the database took \input{build/include-full-run-duration}, with
-mean throughput of \input{build/include-full-run-throughput}.  In contrast,
-when saving results to the database, parsing all \numberOFlogFILES{} log
-files took \input{build/include-insert-results-duration}, with mean
-throughput of \input{build/include-insert-results-throughput} --- parsing
-makes up only
+log files (totaling \input{build/include-total-size-of-93-log-files},
+\numberOFlogLINEShuman{} log lines) were each parsed 10 times, and the mean
+parsing time.  The computer used for test runs was a Dell Optiplex 745,
+shown in \tableref{Details of the computer used to generate statistics}; it
+was dedicated to the task of gathering statistics from test runs, and did
+not run any other programs while test runs were in progress.  Saving
+results to the database was disabled for the test runs, because that
+dominates the run time of the parser, and the tests are aimed at measuring
+the speed of \parsername{} rather than the speed of the database and the
+disks it is stored on.  Parsing all \numberOFlogFILES{} log files in one
+run without saving results to the database took
+\input{build/include-full-run-duration}, with mean throughput of
+\input{build/include-full-run-throughput}.  In contrast, when saving
+results to the database, parsing all \numberOFlogFILES{} log files took
+\input{build/include-insert-results-duration}, with mean throughput of
+\input{build/include-insert-results-throughput} --- parsing makes up only
 \input{build/include-skip-inserting-as-percentage-of-inserting} of the
@@
 
-\subsection{Architecture Scalability: Input Size}
+\subsection{Characteristics Of The Input Log Files}
 
-XXX THIS SECTION DOES NOT SHOW ANY VARIETY IN INPUT SIZE, SO DOES NOT SHOW
-SCALABILITY\@.  MAYBE I SHOULD RENAME IT TO ``Characteristics Of The Input
-Log Files'' OR SOMETHING\@?  THE GRAPHS SHOW THAT LOGS HAVE CONSISTENT
-CONTENT EXCEPT WHEN THERE IS A MAIL LOOP\@.
-
-An important property of a parser is how parsing time scales relative to
-input size: does it scale linearly, polynomially, or exponentially?
-\Graphref{parsing time vs log file size vs number of log lines graph} shows
-the parsing time in seconds, log file size in MB, and number of log lines
-in tens of thousands, for each of the \numberOFlogFILES{} log files.  The
-three lines run roughly in parallel, giving the impression that the
-algorithm scales linearly with input size.  This impression is borne out by
-\graphref{parsing time vs log file size vs number of log lines factor},
-which plots both the ratio of log file size vs parsing time, and the ratio
-of number of log lines vs parsing time (higher is better in both cases);
-\tableref{parsing time vs log file size vs number of log lines factor
-table} shows the same ratios for different groups of log files.  The ratios
-are quite tightly banded, showing that the algorithm scales linearly; the
-ratios increase (i.e.\ improve) for log files 22 \& 62--68, despite their
-larger than usual size.  Both groups of log files are much larger than
-usual because of a mail loop caused by a user who set up mail forwarding
+The \numberOFlogFILES{} log files used when generating statistics have
+fairly consistent sizes and contents, except for two groups of log files:
+22 \& 62--68.  \Graphref{parsing time vs log file size vs number of log
+lines graph} shows the parsing time in seconds, log file size in MB, and
+number of log lines in tens of thousands, for each of the
+\numberOFlogFILES{} log files.  \Graphref{parsing time vs log file size vs
+number of log lines factor} plots both the ratio of log file size vs
+parsing time, and the ratio of number of log lines vs parsing time (higher
+is more efficient in both cases); \tableref{parsing time vs log file size
+vs number of log lines factor table} shows the same ratios for different
+groups of log files.  The ratios are quite tightly banded, except they
+increase (i.e.\ improve) for log files 22 \& 62--68, despite their larger
+than usual size.  Both groups of log files are much larger than usual
+because of mail loops caused by users who set up mail forwarding
 incorrectly, resulting in a very different distribution of log lines:
@@
 failed delivery of mail generated on the server itself.  The Postfix
-components that generated most of the log lines during the mail loop have
+components that generated most of the log lines during the mail loops have
 fewer associated rules than the Postfix components whose log lines normally
 make up the bulk of each log file, so the mean number of rules tried per
-log line reduces and so does the mean parsing time per log line.
-\Graphref{Number of rules per Postfix component} shows the number of rules
-per Postfix component, \graphref{Mean number of rules tried per log line}
-shows the drop in the mean number of rules tried per log line for log files
-containing a mail loop, and \graphref{Mean number of rules tried per log
-line for each Postfix component} shows the mean number of rules tried per
-log line for each Postfix component.
+log line reduces, as does the mean parsing time per log line.
+\Tableref{Number of rules per Postfix component} shows the number of rules
+for each Postfix component; \graphref{Mean number of rules tried per log
+line} shows the drop in the mean number of rules tried per log line for log
+files containing a mail loop, and \graphref{Mean number of rules tried per
+log line for each Postfix component} shows the mean number of rules tried
+per log line for each Postfix component.
 
@@
 \showgraph{build/graph-input-size-vs-parsing-time-ratio}{Ratio of log file
-size and number of log lines to parsing time (higher is better)}{parsing
-time vs log file size vs number of log lines factor}
+size and number of log lines to parsing time (higher is more
+efficient)}{parsing time vs log file size vs number of log lines factor}
 
@@
 
-\parsername{} has \numberOFrules{} rules: the top 10\% recognise
+\parsernames{} ruleset has \numberOFrules{} rules: the top 10\% recognise
 \input{build/include-top-ten-hits}\% of the log lines in the
@@
 Assuming that the distribution of log lines is reasonably consistent over
-time, \parsernames{} efficiency should benefit from trying rules that
-recognise log lines more frequently before those rules that recognise log
+time, \parsernames{} efficiency should benefit from using rules that
+recognise log lines more frequently before it uses rules that recognise log
 lines less frequently.  To test this hypothesis, three full test runs were
@@
 in those log files.  A careful examination of \graphref{Parsing time
-relative to shuffled ordering graph} shows that, for the first log file,
-optimal and reverse orderings perform identically: this is because the hits
-field of each rule is zero for the first log file, so optimal and reverse
-orderings produce identical rule orderings.  For the first log file,
-shuffled ordering is the most efficient of the three, but that is
+relative to shuffled ordering graph} shows that, for the first log file
+only, optimal and reverse orderings perform identically: this is because
+the hits field of each rule is zero for the first log file, so optimal and
+reverse orderings produce identical rule orderings.  For the first log
+file, shuffled ordering is the most efficient of the three, but that is
 accidental and cannot be relied upon.
@@
 
-\subsection{Comparing Optimal Ordering Against A Perfect Oracle}
+\subsection{Comparing Optimal Ordering Against An Oracle}
 
@@
 \Graphref{Parsing time of oracle and optimal ordering relative to shuffled
-ordering graph} shows the parsing times of the oracle and the optimal
-ordering, relative to shuffled ordering; \tableref{Parsing time of oracle
-and optimal ordering relative to shuffled ordering table} shows mean and
-standard deviation.  As expected, the oracle is more efficient than optimal
-ordering, but not by much.  \Graphref{Percentage increase in parsing time
-when using optimal ordering instead of oracle graph} shows the percentage
-increase in parsing time when using optimal ordering instead of the oracle,
-with mean and standard deviation in \tableref{Percentage increase in
-parsing time when using optimal ordering instead of oracle table}.
+ordering graph} shows how the oracle and optimal ordering perform relative
+to shuffled ordering; \tableref{Parsing time of oracle and optimal ordering
+relative to shuffled ordering table} shows mean and standard deviation.  As
+expected, the oracle is more efficient than optimal ordering, but not by
+much.  \Graphref{Percentage increase in parsing time when using optimal
+ordering instead of oracle graph} shows the percentage increase in parsing
+time when using optimal ordering instead of the oracle, with mean and
+standard deviation in \tableref{Percentage increase in parsing time when
+using optimal ordering instead of oracle table}.
 
@@
 \& 62--68), because the mean number of rules tried per log line is lower
-(see \graphref{Mean number of rules tried per log line}).  When parsing the
-first log file, the performance of optimal ordering relative to the oracle
-is much worse than for the remainder of the log files, because for the
+(see \graphref{Mean number of rules tried per log line}).  The performance
+of optimal ordering relative to the oracle is much worse when parsing the
+first log file than for the remainder of the log files, because for the
 first log file the hits field of every rule is zero, so optimal ordering
@@
 \input{build/include-perfect-best-vs-optimal-mean} slower than parsing
-using a magical oracle that divines the correct rule to use for each log
+using an oracle that magically divines the correct rule to use for each log
 line.
@@
 How any architecture scales as the number of rules increases is important,
-but it is particularly important in this architecture because it is
-expected that the typical parser will have a large ruleset.  The full
+but it is particularly important for this architecture because it is
+anticipated that the typical parser will have a large ruleset.  The full
 \parsername{} ruleset has \numberOFrules{} rules, whereas the minimum
-number of rules required to parse the \numberOFlogFILES{} log files is
-\numberOFrulesMINIMUM{}, \numberOFrulesMINIMUMpercentage{} of the full
-ruleset.  The full ruleset is larger because \parsername{} is tested with
-\numberOFlogFILESall{} log files; testing with more log files increases the
-chance of finding bugs in the parser or new complications to be overcome.
-The \numberOFlogFILES{} log files were each parsed 10 times using the
-minimum ruleset, and the parsing times compared to those generated using
-the full ruleset: the percentage parsing time increase when using the full
-ruleset instead of the minimal ruleset for optimal, shuffled, and reverse
-orderings is shown in \graphref{Percentage parsing time increase when using
-the maximum ruleset instead of the minimum ruleset}, with mean and standard
-deviation in \tableref{Percentage parsing time increase when using the
-maximum ruleset instead of the minimum ruleset table}.
+ruleset required to parse the \numberOFlogFILES{} log files has
+\numberOFrulesMINIMUM{} rules, \numberOFrulesMINIMUMpercentage{} of the
+full ruleset.  The full ruleset is larger because \parsername{} is tested
+with \numberOFlogFILESall{} log files; testing with more log files
+increases the chance of finding bugs in the parser or new complications to
+be overcome.  The \numberOFlogFILES{} log files were each parsed 10 times
+using the minimum ruleset, and the mean parsing times compared to those
+generated using the full ruleset: the percentage parsing time increase when
+using the full ruleset instead of the minimal ruleset for optimal,
+shuffled, and reverse orderings is shown in \graphref{Percentage parsing
+time increase when using the maximum ruleset instead of the minimum
+ruleset}, with mean and standard deviation in \tableref{Percentage parsing
+time increase when using the maximum ruleset instead of the minimum ruleset
+table}.
 
@@
 file, optimal and reverse orderings have identical performance, because the
-hits field of every rule is zero.
+hits field of every rule starts at zero.
 
@@
 These results show that both the architecture and \parsername{} scale
-extremely well as the number of rules increases, and that optimally sorting
-the rules is an important optimisation contributing to this scalability.
+extremely well as the ruleset increases in size, and that optimally
+ordering the rules is an important optimisation contributing to this
+scalability.
 
@@
 to improve the speed of matching.  This compilation and optimisation takes
-CPU time: in many cases it takes far more CPU time than the actual
+CPU time: for most regexes it takes far more CPU time than the actual
 matching.  If the interpreter is certain that a regex will not change while
@@
 a dynamically generated regex can be cached and used in preference to the
-original regex, but it is the responsibility of the programmer to do this;
+original regex, but it is the responsibility of the programmer to do so;
 \parsername{} does this with every rule's regex when the rules are loaded
@@
 compiled regexes is probably the single most effective optimisation
-possible in \parsername{}, and was quite simple to implement.  As with
-previous optimisations, log files 22 \& 62--68 do not suffer such a large
-increase in parsing time when not caching compiled regexes; this is
-because, on average, fewer regexes are compiled per log line for those log
-files.  The increase in parsing time when parsing the first log file is
-much greater than for the other log files; again, this is because every
-rule's hits field is zero, so optimal ordering is less efficient than
-usual, and the mean number of rules tried for each log line will be higher
-than usual.
+possible in \parsername{}, and was quite simple to implement.  As seen
+previously, log files 22 \& 62--68 do not suffer such a large increase in
+parsing time when the optimisation is disabled; this is because, on
+average, fewer regexes are compiled per log line for those log files.  The
+increase in parsing time when parsing the first log file is much greater
+than for the other log files; again, this is because every rule's hits
+field starts at zero, so optimal ordering is less efficient than usual, the
+mean number of rules tried for each log line will be higher than usual, and
+more regexes will need to be compiled when recognising each log line.
 
@@
 
-    \item With the optimisations described in this chapter enabled,
+    \item Even with the optimisations described in this chapter enabled,
         recognising log lines still dominates the execution time of the
@@
         \input{build/include-percentage-time-spent-processing-log-lines-mean}
-        of parsing time, less if using shuffled ordering, so optimising
-        actions could not possibly provide as big a performance increase as
-        optimally ordering rules.  Similarly, caching compiled regexes
-        provides an
+        of parsing time, less if any optimisations are disabled, so
+        optimising actions could not possibly provide as big a performance
+        increase as optimally ordering rules.  Similarly, caching compiled
+        regexes provides an
         \input{build/include-cached-regexes-vs-discarded-regexes-mean-reduction}
@@
         of recognised log lines was optimised to 1\% of its original
-        parsing time, it would be slightly better than optimal ordering,
-        but it would be vastly harder to implement; optimising actions
-        would not provide enough reduction in parsing time to justify the
-        amount of effort required.
+        parsing time, it would be only slightly more effective than optimal
+        ordering, but it would be vastly harder to implement; optimising
+        actions would not reduce parsing time enough to justify the amount
+        of effort required.
 
@@
 achieved.  Correctly understanding and reconstructing every mail delivery
-attempt, whether it was successful or not, is important so that the
-information in the database is correct and complete.  Improving the
-proportion of log lines correctly recognised is the less difficult of the
-two, because usually it just requires new rules to be written or existing
-rules to be changed.  Improving the proportion of correctly understood and
-reconstructed mail delivery attempts is more difficult and intrusive,
-because it requires adding or changing actions, and it can be much harder
-to realise that a deficiency exists and needs to be addressed.
+attempt is important so that the information in the database is correct and
+complete.  Improving the proportion of log lines correctly recognised is
+the less difficult of the two, because it just requires new rules to be
+written or existing rules to be extended.  Improving the proportion of
+correctly understood and reconstructed mail delivery attempts is more
+difficult and intrusive, because it requires adding or changing actions,
+and it can be much harder to realise that a deficiency exists and needs to
+be addressed.
 
@@
 
-Parsing a log line is a three stage process:
+Parsing a log line is a three step process:
 
@@
 
-Correctly recognising all log lines requires that each Postfix component
-whose log lines are of interest must have at least one rule, or its log
-lines will be silently skipped; in the extreme case of an empty ruleset the
-parser would skip every log line.  \parsername{} skips those log lines
-because there may be any number of log lines from other programs
-intermingled in the log file, and some Postfix components that do not
-produce any log lines of interest.  There must be a rule to recognise each
-log line variant produced by each Postfix component; if a log line is not
-recognised the parser will issue a warning, to inform the user that they
-need to extend their ruleset.  \parsername{} does not parse log lines from
-non-Postfix programs, e.g.\ Amavisd-new or SpamAssassin; it could easily be
-extended to do so, if a method could be developed to correctly associate
-such log lines with existing state table entries.  Each rule's regex should
-be as specific and precise as possible, to ensure accurate parsing: a rule
-with a regex that matches zero or more of any character will recognise
-every log line, but not in a meaningful way.  Most log lines contain fixed
-strings, so this is not a problem in practice.
+Each Postfix component whose log lines are of interest must have at least
+one rule, or its log lines will be silently skipped; in the extreme case of
+an empty ruleset the parser would skip every log line.  \parsername{} skips
+log lines logged by programs that do not have any rules because there may
+be any number of log lines from other programs intermingled in the log
+file, and some Postfix components that do not produce any log lines of
+interest.  \parsername{} does not parse log lines from non-Postfix
+programs, e.g.\ Amavisd-new or SpamAssassin; it could easily be extended to
+do so, if a method could be developed to correctly associate such log lines
+with existing state table entries.  To correctly recognise all log lines,
+there must be a rule to recognise each log line variant produced by each
+Postfix component; if a log line is not recognised the parser will issue a
+warning, to inform the user that they need to extend their ruleset.  Each
+rule's regex should be as specific and precise as possible, to ensure
+accurate parsing: a rule with a regex that matches zero or more of any
+character will recognise every log line, but not in a meaningful way; most
+log lines contain fixed strings, so this is not a problem in practice.
 
-Full coverage of log lines can be achieved without undue effort, yet it is
-hard to maintain.  Maintaining full coverage is hard because log lines
-change over time, e.g.\ administrators add restrictions with custom
-messages, \acronym{DNSBL} messages change, or major releases of Postfix
-change log lines (usually adding more information).  Warnings are issued
-for any log lines that are not recognised; no warnings are issued for
-unrecognised log lines while parsing the \numberOFlogFILES{} log files, so
-it can be safely concluded that zero false negatives arise.  False
-positives are harder to quantify, short of examining each of the 60,721,709
-log lines and ensuring that the correct rule recognised it.  However, a
-random sample of 6039 log lines was parsed, and the results manually
-verified by inspection to ensure that the correct rule recognised each log
-line.  The sample was generated by running the command: \texttt{perl -n -e
-\singlequote{}print if (rand 1 < 0.0001)\singlequote{} LOG\_FILES} to
-randomly extract roughly one log line in every 10,000 (it actually
-extracted 0.00994\% instead of 0.01\%).  Each log line was examined and the
-correct rule identified from the \numberOFrules{} rules in the database;
-the correct rule was then compared to the rule that recognised the log line
-when parsing.  The sample results contained zero false positives, and this
-check has been automated to ensure continued accuracy.  Based on this, the
-author is confident that zero false positives occur when parsing the
-\numberOFlogFILES{} log files.  On initial appearances, exercising only 36
-rules from a total of \numberOFrules{} when parsing 6039 log lines seems
-low, but after examining \graphref{rule hits graph} it becomes apparent
-that parsing using such a low number of rules is to be expected.  The
+Full coverage of log lines can be achieved without undue effort, yet once
+achieved it is hard to maintain.  Maintaining full coverage is hard because
+the set of log line variants changes over time, e.g.\ administrators add
+restrictions with custom messages, \acronym{DNSBL} messages change, or
+major releases of Postfix change log lines (usually by adding more
+information).  Warnings are issued for any log lines that are not
+recognised; no warnings are issued for unrecognised log lines while parsing
+the \numberOFlogFILES{} log files, so it can be safely concluded that zero
+false negatives arise.  False positives are harder to quantify, short of
+examining each of the 60,721,709 log lines and checking that the correct
+rule recognised it.  However, a random sample of 6039 log lines was parsed,
+and the results manually verified by inspection to ensure that the correct
+rule recognised each log line.  The sample was generated by running the
+command: \texttt{perl -n -e \singlequote{}print if (rand 1 <
+0.0001)\singlequote{} LOG\_FILES} to randomly extract roughly one log line
+in every 10,000 (it actually extracted 0.00994\% instead of 0.01\%).  Each
+log line was examined and the correct rule identified from the
+\numberOFrules{} rules in the database; the correct rule was then compared
+to the rule that recognised the log line when parsing.  The sample results
+contained zero false positives, and this check has been automated to ensure
+continued accuracy.  Based on these results, and how precise each rule's
+regex is, the author is confident that zero false positives occur when
+parsing the \numberOFlogFILES{} log files.  On initial appearances,
+exercising only 36 rules from a total of \numberOFrules{} when parsing 6039
+log lines seems low, but after examining \graphref{rule hits graph} it
+becomes apparent that such a low number of rules is to be expected.  The
 reader should also bear in mind that even when parsing all
-\numberOFlogFILES{} log files, only \numberOFrulesMINIMUM{} are used,
-because some of the rules are for recognising log lines that only appear in
-other log files.
+\numberOFlogFILES{} log files, only \numberOFrulesMINIMUM{} rules are used,
+because many of the rules recognise log lines that only appear in other log
+files.
 
@@
 \sectionref{complications} were detected in this way).  \parsername{}
-issues warnings when it detects any errors or discrepancies, alerting the
-user to the problem, e.g.\ when a queueid is reused but the previous mail
-remains in the state tables, when a queueid or \acronym{pid} is not found
-in the state tables, or when a mail does not include sufficient data to
-satisfy the database schema.  The parser should produce few or no warnings
-during parsing, and when finished parsing the state tables should only
-contain entries for mails that have log lines in subsequent log files.
-There will often be warnings about a missing queueid or \acronym{pid} in
-the first few thousand log lines, because the earlier log lines for those
+issues warnings when it detects any errors or discrepancies, e.g.\ when a
+queueid is reused but the previous mail remains in the state tables, when a
+queueid or \acronym{pid} is not found in the state tables, or when an entry
+in the state tables does not include sufficient data to satisfy the
+database schema.  The parser should produce few or no warnings during
+parsing, and when finished parsing the state tables should only contain
+entries for mails that have log lines in subsequent log files.  There will
+often be warnings about a missing queueid or \acronym{pid} when parsing the
+first few thousand log lines, because the earlier log lines for those
 connections or mails are in a previous log file; loading saved state tables
@@
 The data used in this chapter is generated by parsing \numberOFlogFILES{}
-log files.  5 are warnings produced, but because \parsername{} errs on the
+log files.  5 warnings are produced, but because \parsername{} errs on the
 side of producing more warnings rather than fewer, those 5 warnings
@@
 parsing the \numberOFlogFILES{} log files, the state tables contain 18
-entries, breaking down into:
+entries:
 
@@
         child mails waiting to be tracked with a parent; see
-        \sectionref{Re-injected mails} for details.
+        \sectionref{Re-injected mails} for details.  XXX CHECK IF THERE IS
+        ANY MENTION OF THIS IN THE COMPLICATION, AND ADD IF MISSING\@.
 
@@
 
-None of the mails in the state tables should not be present, thus it can be
-concluded that zero false negatives occur when parsing the
-\numberOFlogFILES{} log files.  Once again, determining the false positive
-rate is much harder, as manually checking the results of parsing 13,850,793
-connections and mails accepted, rejected, bounced, or delivered is
-infeasible.  Considerable evidence exists that the false positive rate is
-extremely low:
+All of the mails remaining in the state tables have valid reasons for being
+present, so it can be concluded that zero false negatives occur when
+parsing the \numberOFlogFILES{} log files.  Once again, determining the
+false positive rate is much harder, because manually checking the results
+of parsing \numberOFconnectionsINlogFILES{} connections and mails accepted,
+rejected, bounced, or delivered is infeasible.  Considerable evidence
+exists that the false positive rate is extremely low, if not zero:
 
@@
 
-    \item The parser is quite verbose when complaining about known
-        problems, e.g.\ if a mail is missing required data, and no such
-        warnings are produced during the test runs.
+    \item \parsername{} performs many checks to detect known problems
+        occurring, e.g.\ if a mail is missing required data, and no
+        warnings are produced during the test runs other than the five
+        described above.
 
@@
 
-Verifying by inspection that the parser correctly deals with all 60,721,709
-log lines in the \numberOFlogFILES{} log files is infeasible, but verifying
-the parsing of a sample from those log lines is a tractable, if extremely
-time consuming, task.  A sample of log lines was obtained by randomly
-selecting a log file, and then randomly selecting a block of 6000
-contiguous log lines from it (0.00988\% of the total number of log lines).
-It is important that the log lines are contiguous, so that all log lines
-are present for as many of the mail delivery attempts contained in the
-block as possible.  This log segment was parsed with all debugging options
-enabled, resulting in 167,448 lines of output.\footnote{A mean of 27.908
-lines of output per log line; each connection has 30 debugging lines, plus
-21 debugging lines per result.  Connections which have been cloned will
-have the cloned connection in their debugging output, plus another 33
-debugging lines.  Those numbers are approximate, and may vary $\pm{}$ 2.
-An approximate linear relationship between the number of log lines and
-debugging lines is: $33(connections) + 30(accepted~~mails) + 21(results)$.}
-All 167,448 lines were examined in conjunction with the log file segment
-and a dump of the resulting database, verifying that for each of the log
-lines \parsername{} used the correct rule and invoked the correct action,
-which in turn produced the correct result, and the correct data was
-inserted in the database.  The log file segment produced 4 warnings, 10
-mails remaining in the state tables, 1625 mail delivery attempts correctly
+Verifying by inspection that the parser correctly processes all
+\numberOFconnectionsINlogFILES{} mail delivery attempts in the
+\numberOFlogFILES{} log files is infeasible, but verifying the parsing of a
+sample from those log lines is a tractable albeit extremely time consuming
+task.  A sample of log lines was obtained by randomly selecting a log file,
+and then randomly selecting a block of 6000 contiguous log lines from it
+(0.00988\% of the total number of log lines).  It is important that the log
+lines are contiguous, so that all log lines are present for as many of the
+mail delivery attempts contained in the block as possible.  This log
+segment was parsed with all debugging options enabled, resulting in 167,448
+lines of output.\footnote{A mean of 27.908 lines of output per log line;
+each connection has 30 debugging lines, plus 21 debugging lines per result.
+Connections which have been cloned will have the cloned connection in their
+debugging output, plus another 33 debugging lines.  Those numbers are
+approximate, and may vary $\pm{}$ 2.  An approximate linear relationship
+between the number of log lines and debugging lines is: $33(connections) +
+30(accepted~~mails) + 21(results)$.} All 167,448 lines were examined in
+conjunction with the log file segment and a dump of the resulting database,
+verifying that for each of the log lines \parsername{} recognised it with
+the correct rule and invoked the correct action, which in turn correctly
+processed the log line and saved the correct data to be inserted into the
+database.  The log file segment produced 4 warnings, 10 mails correctly
+remaining in the state tables, 1625 mail delivery attempts correctly
 entered in the database, zero false positives, and zero false negatives.
@@
 server the log files were taken from, the computer used to generate the
-statistics in this chapter, and how the parser scales as the size of log
-files increases, including why performance is better on the larger log
-files.  The framework optimises the order in which rules are used when
-trying to recognise each log line, and the effect that optimisation has on
-parsing time is explored; this is followed by the effect that adding more
-rules to the ruleset has on parser performance.  The simplest and most
-effective of the optimisations, caching compiled regexes, is described
-next, and the efficiency evaluation concludes with an examination of where
-parsing time is spent: recognising log lines or processing them?
+statistics in this chapter, and the characteristics of the
+\numberOFlogFILES{} log files used to generate statistics and test
+coverage, including why performance is better on the larger log files.  The
+framework optimises the order in which rules are used when trying to
+recognise each log line, and the effect that optimisation has on parsing
+time is explored; this is followed by comparing optimal ordering with an
+oracle.  How \parsername{} scales as the size of the ruleset increases is
+addressed, followed by a description and analysis of the simplest and most
+effective of the optimisations, caching compiled regexes, and the
+efficiency evaluation concludes with an examination of where parsing time
+is spent: recognising log lines or processing them?
 
@@
 modifying existing rules is very easy because of the separation of rules,
-actions, and framework.  Improving the understanding and reconstruction of
-mail delivery attempts is harder, because Postfix's behaviour must be
-analysed and figured out, and the new behaviour integrated into the actions
+actions, and framework in both the architecture and \parsername{}.
+Improving the understanding and reconstruction of mail delivery attempts is
+harder, because Postfix's behaviour must be analysed and figured out, and
+support for the newly understood behaviour integrated into the actions
 without breaking the existing parsing.  Detecting a deficiency in the
-parser is also significantly harder, because the parser will warn about
-unrecognised log lines, whereas discovering a flaw in the parser requires
-careful study of any warnings produced and the entries remaining in the
-state table.  Rectifying a flaw in the parser requires a deep understanding
-of both the parser and Postfix's log files, investigative work to determine
-the cause of the deficiency, further examination of the log files to aid in
-developing a solution, and finally implementation, integration, and testing
-of the solution.
+parser's understanding of mail delivery attempts requires careful study of
+any warnings produced and the entries remaining in the state tables.
+Rectifying a flaw in the parser requires a deep understanding of Postfix's
+log files, and a working knowledge of the framework, actions, and rules;
+investigative work will be needed to determine the cause of the deficiency,
+followed by further examination of the log files to aid in developing a
+solution, and finally implementation, integration, and testing of the
+solution.
 

### 2009-04-15 doc/logparser-conclusion.tex
Corrections from proofreading.

--- 2009-04-11 doc/logparser-conclusion.tex
@@
 \label{conclusion}
-
-XXX THIS NEEDS TO BE EXTENDED\@.
-
-XXX THERE IS NO MENTION OF RESULTS/EFFICIENCY\@.
-
-XXX PULL CONTENT FROM THE PAPER\@.
-
-XXX INCLUDE A TIGHTLY FOCUSED FUTURE WORK SECTION\@.  PARSE LOG LINES FROM
-OTHER PROGRAMS IF A WAY CAN BE FOUND TO ASSOCIATE THEM WITH EXISTING DATA
-STRUCTURES\@.
 
@@
 because a quick survey of sample log files gives the impression that the
-number of distinct log lines is quite small; this impression is due to the
-uneven distribution exhibited by log lines produced in normal operation
-(see \graphref{rule hits graph} for a vivid illustration of this).
+number of distinct log line variants is quite small; this mistaken
+impression is because of the uneven distribution exhibited by log lines
+produced in normal operation, vividly illustrated by \graphref{rule hits
+graph}.
 
-Given the diverse nature of the log lines and the ease with which
+Given the diverse nature of Postfix log lines, and the ease with which
 administrators can cause new log lines to be logged (\sectionref{postfix
 background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Despite the
-resulting initial increase in complexity the task is quite tractable,
-though it does raise efficiency and optimisation questions (answered in
-\sectionref{parser efficiency}).  Providing a tool to ease the generation
-of regexes from unparsed log lines should greatly help users add rules
-to parse previously unparsed log lines (\sectionref{creating new rules in
-implementation}).
+log lines is a design imperative (\sectionref{parser design}).  Providing a
+tool to ease the generation of regexes from unrecognised log lines should
+greatly help users to extend their ruleset to recognise previously
+unrecognised log lines (\sectionref{creating new rules in implementation}).
+The architecture developed for this project not only makes it trivial to
+recognise new inputs, adding new actions requires the minimum effort
+possible, allowing the implementer to focus on correctly implementing their
+new action.  The separation of the architecture into rules, actions, and
+framework (\sectionref{parser design}) is unusual, partly because the three
+are separated so completely.  Although parsers are often divided into
+separate source code files (the combination of lex \&
+yacc~\cite{lex-and-yacc-book} being a common example), the parts are
+usually quite internally interdependent, and will be combined into a
+complete parser by the compilation process; in contrast \parsername{} keeps
+the rules and actions separate until the parser runs.
 
-The architecture not only substantially eases the parsing of new log lines,
-it makes adding new actions (\sectionref{adding new actions in
-implementation}) a relatively simple task.  The simplicity of adding a new
-action frees the implementer from worrying about how their action might
-interfere with existing parsing, so they can focus on correctly
-implementing their new action.
+The flow of control in this architecture is quite different from most other
+architectures.  Most traditional parsers have a current state: each state
+has a fixed set of acceptable next states, and processing is determined by
+the state transition that takes place; unacceptable state transitions cause
+parsing to fail.  For example, when a \textbf{C} parser sees the keyword
+\textbf{for} it expects to immediately see a left parenthesis, with any
+other input causing parsing to fail (comments will already have been
+removed by the preprocessor).  This architecture is different: the rule
+that recognises the input dictates the action that will be invoked.  Rule
+conditions (\sectionref{rule conditions in architecture}) allow a stateful
+implementation of this architecture, where the list of rules used to
+recognise an input is constrained by the parser's current state, but the
+recognising rule still dictates the action that is invoked, and whether
+directly or indirectly, the next state.
 
-The division of the parser into rules, actions, and framework is unusual
-because rules are separated so completely from the actions and framework;
-although parsers are often divided (the combination of \texttt{lex} and
-\texttt{yacc}~\cite{lex-and-yacc-book} being a common example), the parts
-are usually quite internally interdependent, and will be combined into a
-complete parser by the compilation process; in contrast \parsername{} keeps
-the rules and actions separate until the parser runs.  The actions and
-framework are not as completely separated, because the actions depend on
-services provided by the framework; however the actions and framework are
-not tightly integrated: it would be possible, with some work, to completely
-separate the two.
+XXX SHOULD THIS NEXT PARAGRAPH BE IN THE PAST TENSE\@?
 
-The flow of control in this parser is quite different from that of most
-other parsers.  In most traditional parsers the parser has a current state,
-and each state has a fixed set of acceptable next states, with unacceptable
-states causing parsing to fail --- e.g.\ when a \textbf{C} parser sees the
-keyword \textbf{for} it expects to immediately see a left parenthesis, with
-any other input causing parsing to fail (comments will already have been
-removed by the preprocessor).  This parser is different: the rule which
-recognises the next log line for a mail dictates the action that will be
-invoked, which is equivalent to the next state in other parsers; thus the
-next log line for a mail dictates the next state for that mail.
+The real difficulties arose once the parser was successfully dealing with
+90\% of the log lines, because most of the irregularities and complications
+explained in \sectionref{complications} only became apparent when the vast
+majority of log lines were being parsed successfully.  Adding new rules to
+deal with the numerous infrequently occurring log line variants is a simple
+if tiresome task, whereas dealing with mails that are missing information
+or where Postfix's actions are not being correctly reconstructed is much
+more grueling.  Trawling through log files, searching for something out of
+the ordinary, possibly hundreds or even thousands of log lines away from
+the last occurrence of the queueid for the mail in question, is extremely
+time consuming and quite error prone.  Sometimes the task is not to spot
+the unusual log line, but to spot that a log line that is normally present
+is missing, i.e.\ to realise that one log line amongst thousands is absent.
+In all cases the evidence must be used to construct a hypothesis to explain
+the irregularities, and that hypothesis has to be tested in \parsername{};
+if successful, the parser will be modified to deal with the irregularities,
+without adversely affecting the existing parsing.  The complications
+described in \sectionref{complications} are described in the order they
+were solved in, and that order closely resembles the frequency with which
+they occur; the most frequently occurring complications dominate the
+warning messages produced, and so naturally they were the first
+complications to be dealt with.
 
-The real difficulties arise once the parser is successfully dealing with
-90\% of the log lines, as the irregularities and complications explained in
-\sectionref{complications} only become apparent once the vast majority of
-log lines have been parsed successfully.  Adding new rules to deal with
-numerous, infrequently occurring log lines is a simple, albeit tiresome,
-task, (though alleviated by the tool described in \sectionref{adding new
-actions in implementation}), whereas dealing with mails where information
-appears to be missing is much more grueling.  Trawling through the log
-files, looking for something out of the ordinary, possibly hundreds or even
-thousands of log lines away from the last mention of the queueid in
-question, is extremely time consuming and error prone.  Sometimes the task
-is not to spot the line that is unusual, but to spot that a line that is
-normally present is missing, i.e.\ to realise that one line amongst
-thousands is absent.  In all cases the evidence must be used to construct a
-hypothesis to explain the irregularities, and that hypothesis must then be
-tested in the parser; if successful, the parser must be modified to deal
-with the irregularities, without adversely affecting the existing
-functionality.  The complications described in \sectionref{complications}
-were solved in the order they are described in, and that order closely
-resembles the frequency with which they occur; the most frequently
-occurring complications dominate the warning messages produced, and so
-naturally they are the first complications to be dealt with.
+XXX WRITE ABOUT EFFICIENCY\@.
 
 A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of its test log files is discussed in \sectionref{parsing
-coverage}.  Both its success at parsing individual log lines
-(\sectionref{log-lines-covered}) and its correctness in reconstructing each
-mail's journey through Postfix (\sectionref{mails-covered}) are described
-in detail, including the results of manually verifying the correct parsing
-of a subset of the test log files.
-
+parser's coverage of \numberOFlogFILES{} log files is discussed in
+\sectionref{parsing coverage}.  Both its success at recognising individual
+log lines (\sectionref{log-lines-covered}) and its correctness in
+reconstructing each mail's journey through Postfix
+(\sectionref{mails-covered}) are described in detail, including the results
+of manually verifying the correct parsing of a subset of a log file.
 Experience implementing \parsername{} shows that full input coverage is
@@
 software engineering disciplines.
+
+This architecture's greatest strength is the ease with which
+implementations of it can be adapted to deal with new requirements and
+inputs.  Parsing a variation of an existing input is a trivial task: simply
+modify an existing rule or add a new rule and the task is complete.
+Parsing a new category of input is achieved by writing a new action and
+appropriate rules; quite often the new action will not need to interact
+with existing actions, but when interaction is required the framework
+provides the necessary facilities.  The decoupling of rules from actions
+allows different sets of rules to be used with one set of actions, e.g.\ a
+parser might have actions to process versions one and two of a file format;
+by choosing the appropriate ruleset the parser will parse version one, or
+version two, or both versions.  Decoupling also allows other approaches to
+rule management, as discussed in \sectionref{adding new rules in
+architecture}.  The architecture makes it possible to apply commonly used
+programming techniques (such as object orientation, inheritance,
+composition, delegation, roles, modularisation, or closures) when designing
+and implementing a parser, simplifying the process of working within a team
+or when developing and testing additional functionality.  This architecture
+is ideally suited to parsing inputs that are not fully understood or do not
+follow a fixed grammar: the architecture warns about unparsed inputs and
+other errors, but continues parsing as best it can, allowing the developer
+of a new parser to decide which deficiencies are most important and require
+attention first, rather than being forced to fix the first error that
+arises.
+
+The data gathered by \parsername{} provides the foundation for the future
+of this project: applying machine-learning algorithms to the data to
+analyse and optimise the set of anti-spam defences in use, followed by
+identifying patterns in the data that could be used to write new anti-spam
+techniques to recognise and reject spam rather than accepting it.  The
+parser provides the data in a normalised form that is far easier to use as
+input to new or existing implementations of machine-learning algorithms
+than trying to adapt each algorithm to extract data directly from the log
+files.  New policy servers (\sectionref{policy servers}), written to
+implement new anti-spam measures, could be tested or trained by using the
+data to simulate mail delivery attempts; this would allow simple, fast,
+reproducible testing, without the risk of adversely affecting a production
+mail server.  Development of \parsername{} is complete, i.e.\ it correctly
+parses Postfix log files, and in future it will require only maintenance;
+however, one avenue of future development under consideration is to extend
+it to parse non-Postfix log lines, e.g.\ SpamAssassin or Amavisd-new log
+lines.  \parsername{} can easily be extended to do this, but it lacks a
+method of associating the non-Postfix log lines with the existing data
+structures and state tables, so that all of the data for a mail delivery
+attempt can be stored together.
+
+The Postfix log file parser based on this architecture provides a basis for
+systems administrators to monitor the effectiveness of their anti-spam
+measures and adapt their defences to combat new techniques used by those
+sending spam.  \parsername{} is a fully usable application, built to
+address a genuine need, rather than a proof of concept whose sole purpose
+is to illustrate a new idea; it deals with the oddities and difficulties
+that occur in the real world, rather than a clean, idealised scenario
+developed to showcase the best features of a new approach.

### 2009-04-15 doc/logparser.tex
Corrections from proofreading.

@@
 \newcommand{\numberOFlogLINEShuman}[0]{%
-    60.72 million%
+    60.722 million%
 }
@@
     16,836%
+}
+
+\newcommand{\numberOFconnectionsINlogFILES}[0]{%
+    13,850,793%
 }
@@
 
-    XXX REWRITE WHEN THESIS IS FINISHED\@.
-
     Parsing Postfix log files is much more difficult than it first appears,
     but it \textit{is\/} possible to achieve a high degree of accuracy in
-    understanding Postfix log files and reconstructing the actions taken by
-    Postfix that generated the log files.  This thesis describes the
-    creation of a parser for Postfix log files, documenting the
-    architecture developed and the parser implementing it, explaining the
-    difficulties encountered and the solutions developed, with reference to
-    an implementation that stores data gleaned from the log files in an SQL
-    database.  The gathered data can be used to optimise current anti-spam
-    measures, to produce statistics showing how effective those measures
-    are, or to provide a baseline to test new anti-spam measures against.
-    The Postfix log file parser needs to be very precise and strict when
-    parsing, yet must allow users to easily adapt or extend it to parse new
-    inputs, without requiring that the user have an in-depth knowledge and
+    understanding Postfix log files, and subsequently in reconstructing the
+    actions taken by Postfix when processing mail delivery attempts.  This
+    thesis describes the creation of a parser for Postfix log files,
+    documenting the architecture developed for this project and the parser
+    that implements it, explaining the difficulties encountered and the
+    solutions developed, with reference to an implementation that stores
+    data gleaned from the log files in an SQL database.  The gathered data
+    can be used to optimise current anti-spam measures, to produce
+    statistics showing how effective those measures are, or to provide a
+    baseline to test new anti-spam measures against.  The Postfix log file
+    parser needs to be very precise and strict when parsing log files, yet
+    must allow users to easily adapt or extend it to parse new log lines,
+    without requiring that the user have an in-depth knowledge and
     understanding of the parser's internal workings.  The newly developed
-    architecture aims to make the process of parsing new inputs as simple
-    as possible, enabling users to trivially add new rules (to parse
-    variants of existing inputs) and relatively easily add new actions (to
+    architecture is designed to make the process of parsing new inputs as
+    simple as possible, enabling users to trivially add new rules (to parse
+    variants of known inputs) and relatively easily add new actions (to
     process a previously unknown category of inputs).  The architecture

### 2009-04-15 doc/logparser.tex
Changed title page, added declaration and acknowledgements pages.

@@
 \documentclass[a4paper,12pt,draft]{report}
-\pagestyle{headings}
 
@@
 
+\newcommand{\specialpage}[1]{%
+    \phantomsection{}
+    \addcontentsline{toc}{chapter}{#1}%
+}
+\newcommand{\specialpageheading}[1]{%
+    \begin{center}
+        \textbf{\Large #1}
+    \end{center}
+}
+\newcommand{\specialpageandheading}[1]{%
+    \specialpage{#1}
+    \specialpageheading{#1}
+}
+
 \begin{document}
 
-\title{Parsing Postfix Log Files}
-\author{John Tobin \\ School of Computer Science and Statistics \\
-Trinity College \\ Dublin 2 \\ Ireland \\ tobinjt@cs.tcd.ie}
-\maketitle
+\pagestyle{empty}
+
+\specialpage{Title}
+
+\begin{center}
+
+    {\Huge Parsing Postfix Log Files}
+
+    \vfill{}
+
+    {\LARGE John Tobin}
+
+    \vfill{}
+
+    A thesis submitted to the University of Dublin, in fulfilment of the
+    requirements for the degree of Master of Science in Computer Science.
+
+    \vfill{}
+
+    April 2009
+
+    \vfill{}
+
+    \vfill{}
+
+    \vfill{}
+
+\end{center}
+
+\newpage{}
+
+% Declaration
+\specialpageandheading{Declaration}
+
+\vfill{}
+
+\noindent{}I declare that this thesis, and the work described herein, is
+entirely my own work, and has not been submitted as an exercise for a
+degree at this or any other university.  This thesis may be borrowed or
+copied upon request with the permission of the Librarian, Trinity College,
+University of Dublin.
+
+\vfill{}
+
+\begin{flushright}
+    \underline{\hspace*{10em}} \\~\\
+    John Tobin \\
+    \today{}
+\end{flushright}
+
+\vfill{}
+
+\vfill{}
+
+\vfill{}
+
+\vfill{}
+
+\newpage{}
+
+% Acknowledgements
+\specialpageandheading{Acknowledgements}
+
+\noindent{}I would like to express my appreciation to my supervisor, Dr.\
+Carl Vogel, for his advice, assistance, and guidance.
+
+\bigskip{}
+
+\noindent{}I would like to thank my wife, Ariane Tobin, for her patience,
+good will, and encouragement.
 
@@
 
-    \setcounter{page}{2}
+    \setcounter{page}{4}
     \phantomsection{}
@@
     variants of known inputs) and relatively easily add new actions (to
-    process a previously unknown category of inputs).  The architecture
-    scales linearly or better as the number of rules and size of input
-    increases, making it suitable for parsing large corpora or months of
-    accumulated data.
+    process a previously unknown category of inputs).  The parser
+    implemented for this project is evaluated on the criteria of efficiency
+    and coverage of Postfix log files, demonstrating that the conflicting
+    goals of efficiency and accuracy can be balanced, and one need not be
+    sacrificed to achieve the other.
 
@@
 
-    XXX MAIL MADS ABOUT EXAMINERS\@.
+\end{abstract}
 
-\end{abstract}
+% Include the header across the top of each page, after the declaration.
+\pagestyle{headings}
 
@@
 
-\setcounter{page}{3}
+\setcounter{page}{5}
 \tableofcontents

### 2009-04-17 doc/logparser-background.tex
Some minor changes.
Corrections based on proofreading to the preface and chapters 1-3.

@@
 
-A short example of the optimisation possible using data from the database
-is determining which Postfix restrictions reject the highest number of
-mails:
+Determining which Postfix restrictions reject the highest number of
+delivery attempts is a short example of the optimisation possible using
+data from the database:
 
@@
 used as a parameter in another), percentages can be obtained for the top
-ten restrictions (sample output shown in \tableref{Sample output from SQL
-query}):
+ten restrictions using the query in \figureref{Sample SQL query}, producing
+output similar to \tableref{Sample output from SQL query}.
 
+\begin{figure}[thbp]
+    \caption{Sample SQL query showing the percentage of delivery attempts
+    rejected by the top ten restrictions}
+    \empty{}\label{Sample SQL query}
     \input{build/include-sample-sql-query}
+\end{figure}
 
 \begin{table}[thbp]
-    \caption{Sample output from SQL query, showing the percentage of
-    delivery attempts rejected by top ten restrictions}
+    \caption{Sample output from an SQL query showing the percentage of
+    delivery attempts rejected by the top ten restrictions}
     \empty{}\label{Sample output from SQL query}
@@
 
-    All percentages above are exclusive, i.e.\ the first restriction's
-    number of rejections does not include the second restriction's number
-    of rejections.
+    All percentages in \tableref{Sample output from SQL query} are
+    exclusive, i.e.\ the first restriction's number of rejections does not
+    include the second restriction's number of rejections.
 
@@
 
-Another example is determining which restrictions are not effective: this
-sample query shows which restrictions had fewer than 100 rejections in the
-last log file parsed, and the percentage of total rejections each of those
-restrictions represents.
+Another example is determining which restrictions are not effective:
+\figureref{Sample SQL query showing ineffective restrictions} shows which
+restrictions rejected fewer than 100 delivery attempts in the last log file
+parsed, and the percentage of total rejections in that log file that each
+of those restrictions represents.
 
+\begin{figure}[thbp]
+    \caption{Sample SQL query showing ineffective restrictions}
+    \empty{}\label{Sample SQL query showing ineffective restrictions}
     \begin{verbatim}
@@
     \end{verbatim}
+\end{figure}
 
-The database queries above yield summary statistics about the efficiency of
+The sample database queries yield summary statistics about the efficiency of
 anti-spam techniques.  These statistics are far less feasible to assess
@@
 Message headers and body sent here.
-.
 502 5.5.2 Error: command not recognized
-QUIT
-221 2.0.0 Bye
 \end{verbatim}
@@
 returned by the server and carries on regardless.  Many spam and virus
-sending programs have serious deficiencies; unfortunately, other programs,
-particularly newer programs, were written by competent programmers or send
-mail using well written programs (e.g.\ Postfix or Sendmail on Unix hosts,
-Microsoft Outlook on Windows hosts).  Traditionally a mail server would
-have done its best to deal with deficient clients, with the intention of
-accepting as much mail sent to its users as possible, e.g.\ by accepting
-sender or recipient addresses that were not enclosed in \texttt{<>}, or by
-ignoring the absence of a HELO command.  The HELO command is the first
-command sent by the client in the \acronym{SMTP} conversation; it is
-required by the protocol, but its original purpose was to detect a server
-trying to deliver mail to itself, so some implementations omitted it.
+sending programs have serious deficiencies; unfortunately, newer programs
+tend to be written by competent programmers or send mail using well written
+programs (e.g.\ Postfix or Sendmail on Unix hosts, Microsoft Outlook on
+Windows hosts).  Traditionally a mail server would have done its best to
+deal with deficient clients, with the intention of accepting as much mail
+sent to its users as possible, e.g.\ by accepting sender or recipient
+addresses that were not enclosed in \texttt{<>}, or by ignoring the absence
+of a HELO command.\footnote{The HELO command is the first command sent by
+the client in the \acronym{SMTP} conversation; it is required by the
+protocol, but its original purpose was to detect a server trying to deliver
+mail to itself, so some implementations omitted it.}  Given the volume of
+spam sent daily, this liberal approach is no longer viable.
 
@@
 different levels of stricture and permissiveness.  Administrators can
-supply their own rejection messages to make it clear to senders exactly why
-their mail was rejected.  Policy servers (\sectionref{policy servers})
-provide a simple way to write new restrictions without having to modify
-Postfix's source code.  Unfortunately this flexibility has a cost:
-complexity in the log files generated.  Although it is relatively simple in
-most cases to use standard Unix text processing utilities to determine the
-fate of an individual mail, sometimes it can be quite difficult.  For
-most mails the journey is simple and brief, but the remainder can have
-quite complex journeys (see \sectionref{complications} for details).
+supply their own rejection messages to make it clear to senders why their
+mail was rejected.  Policy servers (\sectionref{policy servers}) provide a
+simple way to write new restrictions without having to modify Postfix's
+source code.  Unfortunately this flexibility has a cost: the complexity
+Postfix's log files.  Although it is usually relatively simple to use
+standard Unix text processing utilities to determine the fate of an
+individual mail, sometimes it can be quite difficult.  For most mails the
+journey is simple and brief, but the remainder can be quite complex (see
+\sectionref{complications} for details).
 
@@
 
-Postfix restrictions are documented fully in~\cite{smtpd_access_readme,
-smtpd_per_user_control, policy-servers}; the following is only a brief
+Postfix restrictions are documented fully in~\cite{smtpd_per_user_control,
+policy-servers, smtpd_access_readme}; the following is only a brief
 introduction.
@@
 
-\end{eqlist}
+\end{boldeqlist}
 
-The restriction \texttt{check\_client\_access} checks if the \acronym{IP}
-address of the connected client is found in the specified table and returns
-the associated action if found; the method of searching the file is
-dependant on the type of the file (\texttt{cidr} in the example) --- see
-\cite{postfix-lookup-tables} for more details.  Other restrictions
-determine their result by consulting external sources, e.g.\newline{}
-\tab{}\texttt{reject\_rbl\_client dnsbl.example.com}\newline{} checks the
-\acronym{DNSBL} \texttt{dnsbl.example.com} and rejects the command if the
-client's \acronym{IP} address is listed.
+The restriction \texttt{check\_client\_access} checks if the IP address of
+the connected client is found in the specified table and returns the
+associated action if found; the method of searching the file is dependant
+on the type of the lookup table~\cite{postfix-lookup-tables}.  Other
+restrictions determine their result by consulting external sources,
+e.g.\newline{} \tab{}\texttt{reject\_rbl\_client
+dnsbl.example.com}\newline{} checks the \acronym{DNSBL}
+\texttt{dnsbl.example.com} and rejects the command if the client's IP
+address is listed.
 
@@
 evaluated; other less commonly used results are possible as described
-in~\cite{smtpd_access_readme,smtpd_per_user_control,policy-servers}.  The
-meaning of \textit{permit\/} and \textit{reject\/} is fairly obvious;
-\textit{dunno\/} means to stop evaluating the current restriction and
-continue processing the remainder of the restriction list, allowing
-exceptions to more general rules.  E.g.\ the general rule is that machines
-on the local network must authenticate before sending mail, except for the
-web server, because the legacy applications running on it lack
-authentication support; this configuration example shows how:
+in~\cite{access-man-page}.  The meaning of \textit{permit\/} and
+\textit{reject\/} is obvious; \textit{dunno\/} means to stop evaluating the
+current restriction and continue processing the remainder of the
+restriction list, allowing exceptions to more general rules.  The
+configuration example below shows how to require that all machines on the
+local network except for the web server authenticate before sending mail;
+the web server is exempt because the legacy applications running on it lack
+authentication support.  The restriction list is evaluated from top to
+bottom: \texttt{permit\_sasl\_authenticated} permits authenticated clients;
+an exception is made for the web server in the
+\texttt{check\_client\_access} check, and other local machines are
+rejected, because they must be unauthenticated to reach that point in the
+restriction list.
 
@@
 evaluate that restriction, allowing restrictions to be chosen based on the
-client \acronym{IP} address, client hostname, HELO hostname, sender
-address, or recipient address.  E.g.\ the administrator may require that
-all clients on the local network have valid DNS entries, to prevent people
-sending mail from unknown machines; one example of how to achieve this is:
+client IP address, client hostname, HELO hostname, sender address, or
+recipient address.  E.g.\ the administrator may require that all clients on
+the local network have valid DNS entries, to prevent people sending mail
+from unknown machines; one example of how to achieve this is:
 
@@
 shown in \tableref{Example attributes sent to policy servers}) and returns
-a result (reject, permit, dunno, a rejection name) as described in
+a result (reject, permit, dunno, or a rejection name) as described in
 \sectionref{Mixing and matching Postfix restrictions}.  A policy server can
@@
         rejects a delivery attempt when the triple of \newline{}
-        \tab{}\texttt{(sender address, recipient address, remote
-        \acronym{IP} address)}\newline{} has not been seen before; on
-        second and subsequent delivery attempts from that triple the mail
-        will be accepted.  This blocks spam from some senders because
-        maintaining a list of failed addresses and retrying after a
-        temporary failure is uneconomical for a spam sender, but a
-        legitimate mail server must retry deliveries that failed
-        temporarily.  Sadly spam senders are using increasingly complex and
-        well written programs to distribute spam, frequently using an ISP
-        provided \acronym{SMTP} server from a compromised machine on the
-        ISP's network.  Greylisting will slowly become less effective as
-        spam senders adapt, but it does block a large percentage of spam
-        mail at the moment; the most effective restrictions from the
-        \numberOFlogFILES{} log files used when generating the results in
-        \sectionref{Evaluation} are shown in \tableref{Most effective
-        Postfix restrictions}.  The table shows that greylisting is worth
-        using at the moment, particularly when you take into account its
-        position as the final restriction that a mail must overcome in the
-        configuration used on the mail server that generated the log files:
-        on that server greylisting only takes effect for mails that have
-        passed all other restrictions.  Some problems may be encountered
-        when using greylisting: some servers fail to retry after a
-        temporary failure, or legitimate mail may be delayed, particularly
-        when coming from a pool of servers.
+        \tab{}\texttt{(sender address, recipient address, remote IP
+        address)}\newline{} has not been seen before; on second and
+        subsequent delivery attempts from that triple the mail will be
+        accepted.  This blocks spam from some senders because maintaining a
+        list of failed addresses and retrying after a temporary failure is
+        uneconomical for a spam sender, but a legitimate mail server must
+        retry deliveries that fail temporarily.  Sadly spam senders are
+        using increasingly complex and well written programs to distribute
+        spam, frequently using an ISP provided \acronym{SMTP} server from a
+        compromised machine on the ISP's network.  Greylisting will slowly
+        become less effective as spam senders adapt, but it does block a
+        large percentage of spam mail at the moment; the most effective
+        restrictions from the \numberOFlogFILES{} log files used when
+        generating the results in \sectionref{Evaluation} are shown in
+        \tableref{Most effective Postfix restrictions}.  The table shows
+        that greylisting is worth using at the moment, particularly when
+        you take into account its position as the final restriction that a
+        mail must overcome in the configuration used on the mail server
+        that generated the log files: on that server greylisting only takes
+        effect for mails that have passed all other restrictions.  Some
+        problems may be encountered when using greylisting: some servers
+        fail to retry after a temporary failure, or legitimate mail may be
+        delayed, particularly when coming from a pool of servers.
 
@@
     \item Scoring systems such as Policyd-weight~\cite{policyd-weight}
-        perform tests on features of the delivery attempt (e.g.\
-        \acronym{IP} address, sender address), incrementing or decrementing
-        a score based on the results; if the eventual score is higher than
-        a threshold the mail is rejected.  The administrator must manually
+        perform tests on features of the delivery attempt (e.g.\ IP
+        address, sender address), incrementing or decrementing a score
+        based on the results; if the final score is higher than a threshold
+        the delivery attempt is rejected.  The administrator must manually
         whitelist clients if they are to bypass a Postfix restriction,
@@
         several scored restrictions will allow mail from clients that would
-        fall foul of one binary restriction only.
+        be rejected by a binary restriction.
 

### 2009-04-17 doc/logparser-complications.tex
Corrections based on proofreading to the preface and chapters 1-3.

@@
 that re-injected mails appear in the log files without explicit logging
-indicating their source.
+indicating their source.  Unfortunately, it is not possible to distinguish
+between child mails waiting to be tracked and other mails whose origin is
+unknown (except for bounce notifications, as described in
+\sectionref{identifying bounce notifications}); postponing the entry into
+the database of mails whose origin is unknown can result in those mails
+remaining indefinitely in the state tables, causing a queueid clash if the
+queueid is reused, resource usage for the period they remain in the state
+tables, and, most importantly, \parsernames{} understanding of such mails
+is incorrect.  This problem has only been observed when a state table entry
+is missing its initial log line, usually because it is in an earlier log
+file; this specific instance is not a serious problem, because
+\parsername{} can not be expected to fully understand and reconstruct a
+mail when it is lacking some of that mail's log lines.
 
@@
 reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and \acronym{IP} address, but the database
-schema requires many more fields be populated (see \sectionref{connections
-table} and \sectionref{results table}).  These heuristics are quite
-restrictive, and it appears to have little scope for false positives; any
-false positives will cause a warning when the next log line for that mail
-is parsed.  False negatives are less likely to be detected: there may be
+available is the client hostname and IP address, but the database schema
+requires many more fields be populated (see \sectionref{connections table}
+and \sectionref{results table}).  These heuristics are quite restrictive,
+and it appears to have little scope for false positives; any false
+positives will cause a warning when the next log line for that mail is
+parsed.  False negatives are less likely to be detected: there may be
 queueid clashes (and warnings) if mails remain in the state tables after
@@
         that passes normally through Postfix will have a \daemon{cleanup}
-        line; lack of one is a sure sign the mail did not make it too far.
+        log line; lack of one is a sure sign the mail did not make it too
+        far.
 

### 2009-04-17 doc/logparser-implementation.tex
Some minor changes.
Corrections based on proofreading to the preface and chapters 1-3.

@@
 \sectionref{results table}.  A diagram of the database schema is provided
-in \figureref{Diagram of the Database Schema picture}, to complement the
+in \figureref{Diagram of the database schema picture}, to complement the
 in-depth descriptions of each table.
@@
 \setlength{\belowcaptionskip}{10pt}
-\showgraph{build/database-schema}{Diagram of the Database Schema}{Diagram
-of the Database Schema picture}
+\showgraph{build/database-schema}{Diagram of the database schema}{Diagram
+of the database schema picture}
 \setlength{\belowcaptionskip}{\belowcaptionskipORIG}
@@
     \item [description] Something must have occurred to cause Postfix to
-        log each line, this field should describe the event causing the log
-        lines this rule recognises.
+        output each log line; this field should describe the event causing
+        the log lines this rule recognises.
 
@@
 
-    \item [server\_ip] The \acronym{IP} address of the server: the local
-        address when receiving mail, the remote address when sending mail.
+    \item [server\_ip] The IP address of the server: the local address when
+        receiving mail, the remote address when sending mail.
 
     \item [server\_hostname] The hostname of the server, it will be
-        \texttt{unknown} if the \acronym{IP} address could not be resolved
-        to a hostname via DNS\@.
-
-    \item [client\_ip] The client \acronym{IP} address: the remote address
-        when receiving mail, the local address when sending mail.
+        \texttt{unknown} if the IP address could not be resolved to a
+        hostname via DNS\@.
+
+    \item [client\_ip] The client IP address: the remote address when
+        receiving mail, the local address when sending mail.
 
     \item [client\_hostname] The hostname of the client, it will be
-        \texttt{unknown} if the \acronym{IP} address could not be resolved
-        to a hostname via DNS\@.
+        \texttt{unknown} if the IP address could not be resolved to a
+        hostname via DNS\@.
 
@@
     \item How long it takes to parse each log file (\graphref{parsing time
-        vs file size vs number of log lines graph}).
-
-    \item The number of rules tried when recognising log lines
-        (\graphref{Mean number of rules tried per log line}); the framework
+        vs log file size vs number of log lines graph}).
+
+    \item The number of rules used when recognising log lines
+        (\graphref{Mean number of rules used per log line}); the framework
         may need to try multiple rules for a given log line before it finds
@@
 
-    \item The number of log lines and rules tried for each Postfix
-        component (\graphref{Mean number of rules tried per log line for
-        each Postfix component}).
+    \item The number of log lines and rules used for each Postfix component
+        (\graphref{Mean number of rules used per log line for each Postfix
+        component}).
 
@@
         it performs many substitutions on the input log lines, replacing
-        commonly occurring variable terms (e.g.\ email addresses,
-        \acronym{IP} addresses, the standard start of rejection messages)
-        with keywords described in \sectionref{regex components}.  The
-        purpose of this step is to utilise existing knowledge to create
-        more accurate regexes; it replaces many variable words with fixed
-        words, improving the subsequent classification of words as fixed or
+        commonly occurring variable terms (e.g.\ email addresses, IP
+        addresses, the standard start of rejection messages) with keywords
+        described in \sectionref{regex components}.  The purpose of this
+        step is to utilise existing knowledge to create more accurate
+        regexes; it replaces many variable words with fixed words,
+        improving the subsequent classification of words as fixed or
         variable.  The altered log lines are written to a temporary file,

### 2009-04-17 doc/logparser-introduction.tex
Corrections based on proofreading to the preface and chapters 1-3.

@@
 The approach chosen to measure performance is to analyse the log files
-produced by the \acronym{SMTP} server in use, Postfix, rather than
-modifying its source code to generate statistics: this approach improves
-the chances of other Postfix users testing and using the software developed
-for this project.  The need arose for a parser capable of dealing with the
-great number and variety of log lines produced by Postfix: the parser must
-be designed and implemented so that adding support for parsing new inputs
-is a simple task, because the log lines to be parsed will change over time.
+produced by the \acronym{MTA} in use, Postfix, rather than modifying its
+source code to generate statistics: this approach improves the chances of
+other Postfix users testing and using the software developed for this
+project.  The need arose for a parser capable of dealing with the great
+number and variety of log lines produced by Postfix: the parser must be
+designed and implemented so that adding support for parsing new inputs is a
+simple task, because the log lines to be parsed will change over time.
 This variety in log lines occurs for several reasons:
@@
 
-    \item Policy servers~\cite{policy-servers} may log different messages
-        depending on the characteristics of the connection.
+    \item Policy servers (see \sectionref{policy servers}) may log
+        different messages depending on the characteristics of the
+        connection.
 
@@
 processing is the unstructured nature of Postfix log files, where each log
-line was added on an ad hoc basis as a requirement was discovered or new
+line was added on an ad hoc basis when a requirement was discovered or new
 functionality was added.\footnote{A history of all changes made to Postfix
@@
 
-It was hoped to reuse an existing parser rather than writing one from
-scratch, but the effort required to adapt and improve an existing parser
-was judged to be greater than the effort to write a new one, because the
-techniques used by the existing parsers severely limited their potential:
-
-\begin{itemize}
-
-    \item Some parsers ignored the majority of log lines, parsing specific
-        log lines accurately, but without any provision for parsing new or
-        similar log lines.
-
-    \item Other parsers sloppily parsed the majority of log lines, but were
-        incapable of distinguishing between log lines in one same category,
-        e.g.\ not distinguishing between different anti-spam techniques
-        causing the rejection of delivery attempts.
-
-    \item None of the reviewed parsers extracted enough data for future
-        in-depth analysis.
-
-\end{itemize}
-
 The only prior published work on the subject of parsing Postfix log files
@@
 Practical Utilizations\/}~\cite{log-mail-analyser}, which aims to extract
-data from log files, correlate it, and present it in a form suitable for a
-systems administrator to search using the myriad of standard Unix text
-processing utilities already available; it is reviewed alongside the other
-parsers in the State of the Art Review in chapter~\ref{state of the art
-review}.
+data from Postfix log files, correlate it, and present it in a form
+suitable for a systems administrator to search using the myriad of standard
+Unix text processing utilities already available; it is reviewed alongside
+the other parsers in the State of the Art Review in chapter~\ref{state of
+the art review}.  It was hoped to reuse an existing parser rather than
+writing one from scratch, but the effort required to adapt and improve an
+existing parser was judged to be greater than the effort to write a new
+one, as described in the State of the Art Review.
 
@@
 \acronym{DNSBL}), and specifies which action will process the inputs it
-recognises.  This architecture is ideally suited to parsing inputs
-where the input is not fully understood or does not conform to a fixed
-grammar: the architecture warns about unparsed inputs and other errors, but
-continues parsing as best it can, allowing the developer of a new parser to
-decide which deficiencies are most important and require immediate
-attention, rather than being forced to fix the first error that arises.
-
-This thesis documents the difficult process of parsing Postfix log files,
-presenting an architecture designed to enable the users of a parser to
+recognises.  This architecture is ideally suited to parsing inputs that are
+not fully understood or do not conform to a fixed grammar: the architecture
+warns about unrecognised inputs and other errors, but continues parsing as
+best it can, allowing the developer of a new parser to decide which
+deficiencies are most important and require immediate attention, rather
+than being forced to fix the first error that arises.
+The architecture is designed to enable the users of a parser to
 easily extend it to parse their particular inputs, without requiring much
 work or a high level of understanding of the parsing process and the
-parser's internal workings.  This architecture is the basis of
-\parsername{}, a program that parses Postfix log files and places the
-resulting data into a database for later analysis.  The gathered data can
-be used to optimise current anti-spam defences, to produce statistics
-showing how effective each technique in use is, or to provide a baseline to
-test new anti-spam measures against.  Numerous other uses are possible for
-such data: improving server performance by identifying troublesome
-destinations and reconfiguring appropriately; identifying regular high
-volume uses (e.g.\ customer newsletters) and restricting those uses to
-off-peak times, or providing a dedicated service for them; detecting virus
-outbreaks that propagate via mail; as a base for billing customers on a
-shared server.  Preserving the raw data enables users to develop a
-multitude of uses far beyond those conceived of by the author.
+parser's internal workings.
+
+This architecture is the basis of \parsername{}, a program that parses
+Postfix log files and places the resulting data into a database for later
+analysis.  The gathered data can be used to optimise current anti-spam
+defences, to produce statistics showing how effective each technique in use
+is, or to provide a baseline to test new anti-spam measures against.
+Numerous other uses are possible for such data: improving server
+performance by identifying troublesome destinations and reconfiguring
+appropriately; identifying regular high volume uses (e.g.\ customer
+newsletters) and restricting those uses to off-peak times, or providing a
+dedicated service for them; detecting virus outbreaks that propagate via
+mail; as a base for billing customers on a shared server.  Preserving the
+raw data enables a multitude of uses far beyond those conceived of by the
+author.
 
@@
 Chapter~\ref{Evaluation} evaluates \parsernames{} efficiency, discussing
-the various optimisations implemented in the framework, and the effect they
+the various optimisations implemented in the parser, and the effect they
 have.  It also discusses the coverage achieved by \parsername{} over
@@
 
-Appendix~\ref{Postfix Daemons} concludes this thesis with a brief
-description of Postfix daemons.
+Appendix~\ref{Postfix Daemons} provides a brief description of Postfix
+daemons.
 

### 2009-04-17 doc/logparser-other-parsers-review.tex
Corrections from proofreading.
Corrections based on proofreading to the preface and chapters 1-3.

--- 2009-04-12 doc/logparser-other-parsers-review.tex
@@
 At the start of this project ten Postfix log file parsers were tested, with
-the hope of finding a suitable parser to build upon, rather than starting
-from scratch.  It was quite difficult to find ten parsers to review for
+the hope of finding a suitable parser to build upon, rather than writing
+one from scratch.  It was quite difficult to find ten parsers to review for
 this project, and the functionality offered by those parsers ranges from
@@
 others sloppily parsed the majority of log lines, but were incapable of
-distinguishing between log lines of the same category, e.g.\ rejecting a
-mail delivery attempt.  The first parser reviewed~\cite{log-mail-analyser}
-is the only previously published research in this area that the author is
-aware of; their research aims to show that providing the data from log
-files in a more accessible form is helpful to systems administrators.
+distinguishing between log lines within one category, e.g.\ not
+distinguishing between different anti-spam techniques causing delivery
+attempts to be rejected.  The first parser
+reviewed~\cite{log-mail-analyser} is the only previously published research
+in this area that the author is aware of; that research aims to show that
+providing the data from log files in a more accessible form is helpful to
+systems administrators.
 
 The ten parsers have been reviewed and compared with \parsername{}, this
-project's finished parser, to show how much effort would have been required
-to use those to fulfil the aims and requirements of this project.  It is
-important to compare and contrast newly developed algorithms and parsers
-against those already available, to accurately judge what improvements, if
-any, are delivered by the newcomers.
+project's parser, to show how much effort would have been required to use
+one of those parsers to fulfil the aims and requirements of this project.
+It is important to compare and contrast newly developed algorithms and
+parsers against those already available, to accurately judge what
+improvements, if any, are delivered by the newcomers.
 
 Some important differences exist between \parsername{} and most or all of
-the parsers reviewed here:
+the parsers reviewed in this chapter:
 
@@
 
-    \item None of the parsers reviewed perform the kind of advanced parsing
-        required for this project or deal with the complications described
+    \item None of the reviewed parsers perform the kind of advanced parsing
+        required for this project, or deal with the complications described
         in \sectionref{complications}.
@@
     \item Only \parsername{} enables parsing of new log lines without
-        extensive and intrusive modifications to the parser; \parsernames{}
-        architecture is described in \sectionref{parser architecture}.
+        extensive and intrusive modifications to the parser; the
+        architecture that enables this is documented in \sectionref{parser
+        architecture}.
 
-    \item The parsers reviewed all produce a report of varying complexity
+    \item The reviewed parsers all produce a report of varying complexity
         and detail, whereas \parsername{} does not; it extracts data and
-        leaves generation of reports from the data to other programs.
-        Using an \acronym{SQL} database simplifies the process of
-        generating such reports (discussed in \sectionref{database as
-        API}); some sample queries are given in \sectionref{motivation}.
-        The parser developed for this project is designed to enable much
-        more detailed log file analysis by providing a stable platform for
-        subsequent programs to develop upon.
+        leaves generation of reports to other programs.  Using a
+        \acronym{SQL} database simplifies the process of generating such
+        reports (discussed in \sectionref{database as API}); some sample
+        queries are given in \sectionref{motivation}.  The parser developed
+        for this project is designed to enable much more detailed log file
+        analysis by providing a stable platform for subsequent programs to
+        build upon.
 
     \item Most of the reviewed parsers silently ignore log lines they
-        cannot handle, whereas \parsername{} complains loudly about every
-        single log line it fails to parse.  The exception is AWStats, which
+        cannot handle, whereas \parsername{} warns about every single log
+        line it fails to recognise.  The exception is AWStats, which
         outputs the percentage of log lines it was unable to parse, but
@@
         increase in parsing time when doing so.  Support for reading
-        compressed log files is quite helpful, as it dramatically reduces
-        the disk space required to store historical log files.
+        compressed log files is quite helpful, because it dramatically
+        reduces the disk space required to store historical log files.
 
@@
 \sectionref{connections table} and \sectionref{results table}; for
-convenience that list is repeated here: client and server \acronym{IP}
-address and hostname, HELO hostname, queueid, start time, end time,
-\acronym{SMTP} code, enhanced status code, sender, recipient, size, message
-ID, delay and delays.
+convenience that list is repeated here: client and server IP address and
+hostname, HELO hostname, queueid, start time, end time, \acronym{SMTP}
+code, enhanced status code, sender, recipient, size, message-id, delay, and
+delays.
 
@@
 
-There only appears to be one prior published paper about parsing Postfix
+There appears to be only one prior published paper about parsing Postfix
 log files: \textit{Log Mail Analyzer: Architecture and Practical
@@
 available.  It produces a \acronym{CSV} file and either a MySQL or Berkeley
-DB database.  The decision to support both \acronym{CSV} and Berkeley DB
-appears to have been a serious limitation: both formats have limitations
-that will be explained later.  Hardly any documentation is provided with
-\acronym{LMA}, though some documentation is available
-in~\cite{log-mail-analyser}.  Studying the source code is informative,
-though this author had difficulty as the authors of \acronym{LMA} wrote in
-Italian.
+DB database.  Hardly any documentation is provided with \acronym{LMA}, but
+some documentation is available in~\cite{log-mail-analyser}.  Studying the
+source code is informative, though this author had difficulty as the
+authors of \acronym{LMA} wrote in Italian.
 
@@
 standard mechanism to document the fields or the separator, unlike
-\acronym{SQL} databases where every database includes a schema naming the
-fields and the type of data they store (e.g.\ integer, text, timestamp).
-The \acronym{CSV} record format is not documented, but the output file
-contains a comment giving the format:\newline{} \texttt{\# Timestamp|Nome
-Client|IP Client|IP Server|From|To|Status|Size} \newline{}\acronym{LMA}
-treats \acronym{CSV} lines starting with \texttt{\#} as comments, but not
-all \acronym{CSV} parsers will.
+\acronym{SQL} where every database includes a schema naming the fields and
+the type of data they store (e.g.\ integer, text, timestamp).  The
+\acronym{CSV} record format used by \acronym{LMA} is not documented, but
+the output file contains a comment giving the format:\newline{} \texttt{\#
+Timestamp|Nome Client|IP Client|IP Server|From|To|Status|Size}
+\newline{}\acronym{LMA} treats \acronym{CSV} lines starting with
+\texttt{\#} as comments, but not all \acronym{CSV} parsers will.
 
@@
 tables, and the value is a \acronym{CSV} line containing all of the data
-for that row.  The secondary by-sender, by-recipient, by-date, and
-by-\acronym{IP} tables use the sender/recipient/date/\acronym{IP} address
-as the key, and the value is a \acronym{CSV} list of integers referring to
-the main table.  This effectively re-implements \acronym{SQL} foreign keys,
-but without the functionality offered by even the most basic of
-\acronym{SQL} databases (e.g.\ joins, ordering, searches).  It also
-requires custom code to search on some combination of the above, though the
-authors of \acronym{LMA} did provide some queries: IP-STORY, FROM-STORY,
-DAILY-EMAIL, and DAILY-REJECT\@.  Berkeley DB appears to be the least
-useful of the three output formats: it does not provide the functionality
-of a basic \acronym{SQL} database, and unlike \acronym{CSV} files it cannot
-be used with standard Unix text processing tools.
+for that row.  The secondary by-sender, by-recipient, by-date, and by-IP
+tables use the sender/recipient/date/IP address as the key, and the value
+is a \acronym{CSV} list of integers referring to the keys in the main
+table.  This effectively re-implements \acronym{SQL} foreign keys, but
+without the functionality offered by even the most basic of \acronym{SQL}
+databases, e.g.\ joins, ordering, searches.  It also requires custom code
+to search on a combination of attributes, though the authors of
+\acronym{LMA} did provide some simple reports: IP-STORY, FROM-STORY,
+DAILY-EMAIL, and DAILY-REJECT (all described later).  Berkeley DB appears
+to be the least useful of the three output formats: it does not provide the
+functionality of a basic \acronym{SQL} database, and unlike \acronym{CSV}
+files it cannot be used with standard Unix text processing tools.
 
@@
 value)} pairs, so any additional structuring of the data is imposed by the
-application, thus the application must document this structure.  MySQL
-support was not tested because the schema required is not documented.
+application, and it is the responsibility of the author of the application
+to document this structure.  MySQL support was not tested because the
+schema is not documented, so the required tables could not be created.
 
@@
 \acronym{CSV} output, \acronym{LMA} stores the following data: time and
-date of the log line, client hostname and \acronym{IP} address, server
-\acronym{IP} address, sender and recipient addresses, \acronym{SMTP} code,
-and size (for accepted mails only).  Unlike \parsername{} it does not store
-the server hostname, HELO hostname, queueid, start and end times,
-timestamps for each log line, enhanced status code, delivery delays, or
-message id (for accepted mails only).  Handling of multiple recipients,
-\acronym{SMTP} codes, or remote servers\footnote{A single mail may be sent
-to multiple remote servers if it was addressed to recipients in different
-domains, or Postfix needs to try multiple servers for one or more
-recipients.} is not explained; experimental observation shows that multiple
-records are added when a mail had multiple recipients (sadly the records
-are not associated or linked in any way), and presumably the same approach
-is taken when there were multiple destination servers.
+date of the log line, client hostname and IP address, server IP address,
+sender and recipient addresses, \acronym{SMTP} code, and size (for accepted
+mails only).  Unlike \parsername{} it does not store the server hostname,
+HELO hostname, queueid, start and end times, timestamps for each log line,
+enhanced status code, delivery delays, or message id (for accepted mails
+only).  Handling of multiple recipients, \acronym{SMTP} codes, or remote
+servers\footnote{A single mail may be sent to multiple remote servers if it
+was addressed to recipients in different domains, or Postfix needs to try
+multiple servers for one or more recipients.} is not explained;
+experimental observation shows that multiple records are added when a mail
+had multiple recipients, but the records are not associated or linked in
+any way, and presumably the same approach was taken when there were
+multiple destination servers.
 
@@
 new log lines requires modifying an existing regex or carefully inserting a
-new block in the correct place; extracting extra data will require
+new block in the correct place; extracting extra data would require
 modifying multiple blocks, regexes, or both.
@@
 \sectionref{complications}, except for correlating log lines by queueid;
-not correlating log lines by pid means it cannot correlate most rejections.
-It does not differentiate between different types of rejections, so it is
-not suitable for the purposes of this project; the data about which
-restriction caused the rejection is discarded, whereas the main goal of
-this project is to retain that data to aid optimisation and evaluation of
-anti-spam techniques.  \acronym{LMA} fails to parse Postfix log files
-generated on Solaris hosts because the fields automatically prepended to
-each log line differ from those added on Linux hosts; log files from
-Solaris hosts (and possibly other operating systems) thus require
-pre-processing before parsing by \acronym{LMA}.  The \numberOFlogFILES{}
-pre-processed test log files were parsed without complaint by
-\acronym{LMA}, although it produced 32 entries in the output \acronym{CSV}
-file for every rejection in the input log file; it also missed some 40\% of
-delivered mail.  Once these deficiencies were discovered the author did not
-spend any more time checking the results.
+not correlating log lines by \acronym{pid} means it cannot correlate most
+rejected delivery attempts.  It does not differentiate between different
+types of rejections, so it is not suitable for the purposes of this
+project; the data about which restriction caused the rejection is
+discarded, whereas the main goal of this project is to retain that
+information to aid optimisation and evaluation of anti-spam techniques.
+\acronym{LMA} fails to parse Postfix log files generated on Solaris hosts
+because the fields automatically prepended to each log line differ from
+those prepended on Linux hosts; log files from Solaris hosts (and possibly
+other operating systems) thus require pre-processing before parsing by
+\acronym{LMA}.  The \numberOFlogFILES{} pre-processed test log files were
+parsed without complaint by \acronym{LMA}, although it produced 32 entries
+in the output \acronym{CSV} file for every rejection in the input log
+files; it also missed some 40\% of delivered mail.  Once these deficiencies
+were discovered the author did not spend any more time checking the
+results.
 
@@
 DAILY-EMAIL and DAILY-REJECT\@.  These reports search the Berkeley DB files
-for matching records: the first three extract \acronym{CSV} lines for the
-specified client \acronym{IP} address, sender address, or date
-respectively.  DAILY-REJECT initially failed with an error message from the
-Perl interpreter;\footnote{The error messages were:
-\newline{}\texttt{Undefined subroutine \&main::LIST called at queryDB.pl
-line 372.}\newline{}\texttt{Undefined subroutine \&main::EXTRACT\_FROM\_DB
+for matching records: the first three produce \acronym{CSV} lines for the
+specified client IP address, sender address, or date respectively.
+DAILY-REJECT initially failed with an error message from the Perl
+interpreter;\footnote{The error messages were: \newline{}\texttt{Undefined
+subroutine \&main::LIST called at queryDB.pl line
+372.}\newline{}\texttt{Undefined subroutine \&main::EXTRACT\_FROM\_DB
 called at queryDB.pl line 379.}} after correcting the errors in the code it
-worked, extracting the \acronym{CSV} lines for the specified day where the
+worked, producing \acronym{CSV} lines for the specified day where the
 \acronym{SMTP} code signifies a rejection.  All of these reports are
@@
 The output from DAILY-REJECT and the \texttt{awk} command is not exactly
-the same; the author did not spend substantial time attempting to explain
+the same; the author did not spend substantial time attempting to rectify
 these differences.
@@
     \item The output from DAILY-REJECT is missing some records that are
-        present in the \acronym{CSV} file; this may be because it uses the
-        Berkeley DB files instead, and there may be differences between the
-        contents.
+        present in the \acronym{CSV} file; this may indicate differences
+        between the data stored in the \acronym{CSV} and Berkeley DB files.
 
     \item Some records output by DAILY-REJECT are truncated: they are
-        missing the last $|$ that separates the fields and the newline
+        missing the last $|$ that separates fields and also the newline
         following it, so the line containing only a single space is
@@
 In summary, \acronym{LMA} appears to be a proof of concept, written to
-demonstrate the point of their paper (that having this information in an
-accessible fashion is useful to systems administrators), rather than a
-program designed to be useful in a production environment.
+demonstrate the point of their paper, that making the information contained
+in log files available in an accessible fashion is useful to systems
+administrators, rather than a program intended to be useful in a production
+environment.
 
@@
 % Hard to extend/expand/understand.  Appears to only save: date and hour,
-% DNS name and \acronym{IP} address host, mail server \acronym{IP} address,
-% sender, receiver and e-mail status (sent, rejected).  Undocumented
-% schema.  Design decision to use \acronym{CSV} as an intermediate format
-% between the log file and the database seems to have been restrictive.
-% Appears to require a queueid but majority of log lines (e.g.\ rejections)
-% lack a queueid.  Supports whitelisting \acronym{IP} addresses when
-% parsing logs, but whitelisting when generating reports/data mining would
-% be preferable.  Supporting Berkeley DB is probably limiting the software;
-% an example is the difficulty in searching a pipe-delimited string, so
-% they have re-implemented foreign keys with tables keyed by ip address and
-% so on pointing at the main table - this also will not scale well.  There
-% does not appear to be any attempt to deal with the complications I have
+% DNS name and IP address host, mail server IP address, sender, receiver
+% and e-mail status (sent, rejected).  Undocumented schema.  Design
+% decision to use \acronym{CSV} as an intermediate format between the log
+% file and the database seems to have been restrictive.  Appears to require
+% a queueid but majority of log lines (e.g.\ rejections) lack a queueid.
+% Supports whitelisting IP addresses when parsing log filess, but
+% whitelisting when generating reports/data mining would be preferable.
+% Supporting Berkeley DB is probably limiting the software; an example is
+% the difficulty in searching a pipe-delimited string, so they have
+% re-implemented foreign keys with tables keyed by ip address and so on
+% pointing at the main table - this also will not scale well.  There does
+% not appear to be any attempt to deal with the complications I have
 % encountered: their parsing is not detailed enough to encounter them.  It
@@
 Pflogsumm produces a report designed for troubleshooting rather than
-in-depth analysis.  It does not support saving any data, and it does not
-extract any data that it does not require to produce its report; e.g.\ it
-does not extract the HELO hostname, queueid, start and end times,
+in-depth analysis.  It does not support saving any of the data it extracts
+from log files, and it does not extract any data that it does not require
+to produce its report: the HELO hostname, queueid, start and end times,
 timestamps for each log line, or message id.  Both the parsing and
@@
 Pflogsumm produces a useful report, and successfully parsed the
-\numberOFlogFILES{} log files it was tested with.\footnote{The results it
-reported were not verified in detail, but it did not report any errors, and
-has an excellent reputation amongst Postfix users.}  Pflogsumm has many
-options to include or exclude certain sections of the report, all clearly
-documented; by default it includes the following:
+\numberOFlogFILES{} log files it was tested with.  The results it reported
+were not verified in detail, but it did not report any errors, and has an
+excellent reputation amongst Postfix users.  Pflogsumm has many options to
+include or exclude certain sections of the report, all clearly documented;
+by default its report includes the following:
 
@@
 
-    \item For received mail: per-domain totals for mails sent, deferred,
-        average delay, maximum delay, and bytes delivered.  For received
-        mail: per-domain totals for size and number of mails received.
+    \item For received mail: per-domain totals for the number of mails
+        sent, number of mails deferred, average delay, maximum delay, and
+        bytes delivered.  For received mail: per-domain totals for size and
+        number of mails received.
 
@@
     \item Summary of why mail delivery was deferred or failed, why mails
-        were bounced, why mails were rejected, and warning messages.
+        were bounced, why mails were rejected, and warning messages from
+        Postfix.
 
@@
 Pflogsumm is the only reviewed parser that distinguishes between different
-rejections.
+rejected delivery attempts, an important requirement for this project.
 
@@
 enough data for the purposes of this project: although it can extract three
-different sets of data from Postfix log files, they are not interlinked in
-any way.  The documentation does not suggest that Sawmill correlates log
-lines by either queueid or pid or deals with the other difficulties
+different sets of data from Postfix log files, they are not linked in any
+way.  The documentation does not suggest that Sawmill correlates log lines
+by either queueid or \acronym{pid}, or deals with the other difficulties
 documented in \sectionref{complications}.
@@
 
-    \item \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/postfix.html}{2008/11/23}.
+    \item
+        \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/postfix.html}{2008/11/23}.
         Fields extracted: from, to, server, UID, relay, status, number of
-        recipients, origin hostname, origin \acronym{IP} address, and
-        virus.  It also counts the number of and total size of all mails
-        delivered.  The fields \texttt{server}, \texttt{uid},
-        \texttt{relay}, and \texttt{virus} are not explained in the
-        documentation: \texttt{server} is probably the hostname or
-        \acronym{IP} address of the server the mail is delivered to;
-        \texttt{relay} might be the delivery method: \acronym{SMTP}, local
-        delivery, or \acronym{LMTP}; \texttt{uid} might be the uid of the
-        user submitting mail locally.  Postfix does not do any form of
-        virus checking (though it has many options for cooperating with an
-        external virus scanner), so the \texttt{virus} field is a mystery.
+        recipients, origin hostname, origin IP address, and virus.  It also
+        counts the number of and total size of all mails delivered.  The
+        fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
+        \texttt{virus} are not explained in the documentation:
+        \texttt{server} is probably the hostname or IP address of the
+        server the mail is delivered to; \texttt{relay} might be the
+        delivery method: \acronym{SMTP}, local delivery, or \acronym{LMTP};
+        \texttt{uid} might be the uid of the user submitting mail locally.
+        Postfix does not do any form of virus checking (though it has many
+        options for cooperating with an external virus scanner), so the
+        \texttt{virus} field is a mystery.
 
-    \item \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/postfix_ii.html}{2008/11/23}.
+    \item
+        \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/postfix_ii.html}{2008/11/23}.
         Fields extracted: from, to, RBL list, client hostname, and client
-        \acronym{IP} address.  It also counts the number and total size of
-        all mails delivered.
+        IP address.  It also counts the number and total size of all mails
+        delivered.
 
-    \item \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/beta_postfix.html}{2008/11/23}.
-        Fields extracted: from, to, client hostname, client \acronym{IP}
-        address, relay hostname, relay \acronym{IP} address, status,
-        response code, RBL list, and message id.  It also counts the number
-        and size of all mails delivered, processed, blocked, expired, and
-        bounced.
+    \item
+        \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/beta_postfix.html}{2008/11/23}.
+        Fields extracted: from, to, client hostname, client IP address,
+        relay hostname, relay IP address, status, response code, RBL list,
+        and message id.  It also counts the number and size of all mails
+        delivered, processed, blocked, expired, and bounced.
 
@@
 searching is neither as simple to use nor as informative as the interface
-provided by Splunk (see \sectionref{Splunk review}).  The administrative
-interface is much easier to use than Splunk's: it took only five minutes to
-start parsing a whole directory of log files.
+provided by Splunk (see \sectionref{Splunk review}).  However, the
+administrative interface is much easier to use than Splunk's: it took only
+five minutes to start parsing all of the log files in a directory.
 
@@
 parse the first half of the log files, and twelve hours to parse the next
-third; after twenty four hours parsing the remaining sixth it crashed due
-to lack of disk space.  On the second parsing attempt the log files were
-uncompressed beforehand and parsing took eight hours.
+third; after twenty four hours spent parsing the remaining sixth it crashed
+due to lack of disk space.  On the second parsing attempt the log files
+were uncompressed beforehand and parsing took eight hours.
 
 In summary, Sawmill suffers from supporting so many types of log files: it
-parses many types of log files, but none of them very well; it is probably
-much more useful when parsing log files where each log line is
-self-contained (e.g.\ web server log files), rather than log files
-containing interlinked log lines.  It is not suitable as a base for this
-parser, as the source code made available is encrypted and not intended
-for modification; in addition the architecture would probably need to be
-overhauled or replaced to deal with correlating log lines.
+is probably much more useful when parsing log files where each log line is
+self-contained (e.g.\ web server log files), rather than log files where
+data is spread across multiple log lines.  It is not suitable as a base for
+this parser, because the source code made available is encrypted and not
+intended for modification; in addition the architecture would probably need
+to be overhauled or replaced to deal with correlating log lines.
 
@@
 graphs in real time, a facility not provided by \parsername{}.  Splunk
-allows quite complicated searches, based on the fields extracted by Splunk
+allows quite complicated searches, based on the fields it extracts
 (described later) or the full text of the log line, though it is not
@@
 searches, though that may have been because of limitations in the free
-version.  The data store is not available for use by external programs,
-whereas \parsername{} provides the database and leaves it to the user to
-utilise it without limit or restriction.  The web interface is optimised
-for interactive use rather than automated queries, and it does not appear
-to be possible to write independent tools to utilise the Splunk database.
-Some additional Postfix reports are supposedly available on
-\urlLastChecked{http://www.splunkbase.com/}{2008/11/23}, but the author was
-unable to find any Postfix reports, or indeed reports for any other log
+version.  The web interface is optimised for interactive use rather than
+automated queries, and it does not appear to be possible to write
+independent tools to utilise the Splunk database, whereas \parsername{}
+provides the database and leaves it to the user to utilise it without limit
+or restriction.  Some additional Postfix reports are supposedly available
+on \urlLastChecked{http://www.splunkbase.com/}{2008/11/23}, but the author
+was unable to find any Postfix reports, or indeed reports for any other log
 file types: every category was empty, even those that the web site claimed
 had numerous reports available.  Many types of graphs can be generated,
-though most are variations of a bar or pie chart, except the bubble and
-heatmap graphs.  It is trivial to drill down through the graphs to select a
-portion of the data (e.g.\ select the hour with the largest number of
+though all except the bubble and heatmap graphs are variations of a bar or
+pie chart.  It is trivial to drill down through the graphs to select a
+portion of the data, e.g.\ select the hour with the largest number of
 events, then select a particular host, and finally a specific sender
-address).  All searches performed using the indexed data returned
-reasonable results.
+address.  All searches performed using the indexed data returned reasonable
+results.  The full power of \acronym{SQL} is available when searching the
+data extracted by \parsername{}, allowing the user to search on arbitrarily
+complicated conditions.
 
@@
 tasks.  When testing Splunk it took roughly 30 minutes to figure out how to
-add a single log file to be indexed so that it could be searched, with the
+add a single log file to be indexed for later searching, with the added
 downside that the log file was copied into a spool directory before
 indexing, doubling the disk space usage.  The next test was to index all
-the log files in a particular directory, but after three hours, numerous
-futile attempts, and reading all the available documentation, the author
-was still unable to index all the log files in a directory using the web
-interface.  Using the \acronym{CLI} rather than the web interface was
-partially successful: the command ``\texttt{splunk find logs
+the log files in a particular directory, but after three hours, reading all
+the available documentation, and numerous futile attempts, the author was
+still unable to index all the log files in a directory using the web
+interface.  Using Splunk's \acronym{CLI} rather than the web interface was
+more successful: the command ``\texttt{splunk find logs
 }\textit{log-directory\/}'' added 40 of the \numberOFlogFILES{} log files
@@
 files, without explaining why the others were excluded.\footnote{The log
-files appear to have been indexed once only; presumably Splunk keeps track
-of the log files it has indexed and discards requests to index log files
-for a second time.  This may or may not be a useful feature for
-\parsername{}.} There did not appear to be an option to ensure the log
-files would be processed in the order they were created, though this may be
-neither necessary nor beneficial with Splunk.  Subsequently the author was
-successful in adding a single file at a time using the \acronym{CLI} and
-the command ``\texttt{splunk add tail }\textit{filename\/}''.  A simple
-loop using that command was enough to add all the desired log files.
+files that were added multiple times appear to have been indexed once only;
+presumably Splunk keeps track of the log files it has indexed and discards
+requests to index log files for a second time.  This may or may not be a
+useful feature for \parsername{}.}  The command did not have an option to
+ensure the log files would be processed in the order they were created,
+though such an option may be neither necessary nor beneficial with Splunk.
+Subsequently the author was successful in adding a single log file at a
+time using the command ``\texttt{splunk add tail }\textit{filename\/}''.  A
+simple loop using that command was enough to add all the desired log files.
 Splunk will periodically check all indexed log files for updates unless
@@
 behaviour.  Splunk did not appear to have any difficulty in indexing the
-log files, once they had been successfully added to its queue.
-\parsername{} parses the logs it is instructed to parse, in the order
-given; periodic parsing of logs is a task an administrator can easily
-achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.
+log files once they had been successfully added to its queue.
+\parsername{} parses the log files it is instructed to parse, in the order
+they are given; periodic parsing of log files is a task an administrator
+can easily achieve with \texttt{cron(8)} and \texttt{logrotate(8)}.
 
@@
 material and lack of organisation makes it hard to find the topic being
-sought.  Searches confusingly tended to return results from old
+sought, and searches confusingly tended to return results from old
 documentation rather than new.  In general the documentation appears to
@@
 
-Splunk supports reading compressed log files without any configuration by
-the user.  The free version of Splunk limits the volume of data indexed per
-day to 500MB, though a trial Enterprise licence is available that allows
-indexing of up to 5GB of data per day.  In 2007, the cheapest licenced
-version cost \$5000 plus \$1000 support, and limited the volume of data
-indexed per day to 500MB\@.  Prices were removed from the Splunk website
-during 2008; now Splunk's sales team must be asked for a quote.  Typical
-log file sizes for a small scale mail server are given in
-\sectionref{parser efficiency}.
+Splunk supports reading compressed log files without any extra
+configuration by the user, like \parsername{}.  The free version of Splunk
+limits the volume of data indexed per day to 500MB, though a trial
+Enterprise licence is available that allows indexing of up to 5GB of data
+per day.  In 2007, the cheapest licenced version cost \$5000 plus \$1000
+support per annum, and limited the volume of data indexed per day to
+500MB\@.  Prices were removed from the Splunk website during 2008; now
+Splunk's sales team must be asked for a quote.  Median log file size for
+the log files used when evaluating \parsername{} is
+\input{build/include-median-file-size}.
 
@@
 syslog\glsadd{syslog} fields at the beginning of the log line, and extracts
-any \texttt{key=value} pairs occurring after the standard syslog prologue:
-to and from addresses, HELO hostname, and protocol (\acronym{SMTP} or
-\acronym{ESMTP}).  \parsername{} extracts noticeably more data (client and
-server \acronym{IP} address and hostname, queueid, start and end times,
-timestamps for each log line, \acronym{SMTP} and enhanced status codes,
-delivery delays, and message ID), though it does not make the full text of
-the line available (this could be trivially added if desired, but would
-greatly increase the size of the resulting database).  The full power of
-\acronym{SQL} is available when searching the data extracted by
-\parsername{}, allowing the user to search on arbitrarily complicated
-conditions.
+any \texttt{key=value} pairs occurring after the syslog prologue: to and
+from addresses, HELO hostname, and protocol (\acronym{SMTP},
+\acronym{LMTP}, or \acronym{ESMTP}).  \parsername{} extracts noticeably
+more data (client and server IP address and hostname, queueid, start and
+end times, timestamps for each log line, \acronym{SMTP} and enhanced status
+codes, delivery delays, and message-id), though it does not make the full
+text of the log line available; this could be trivially added if desired,
+but would greatly increase the size of the resulting database.
 
-Splunk is a generic tool, so it lacks any Postfix specific support over and
-above extracting the \texttt{key=value} fields from a log line; it makes no
-attempt to correlate log lines by queueid or \acronym{pid}, or to handle
+Splunk is a generic tool, so it lacks any Postfix-specific support over and
+above extracting the \texttt{key=value} fields from each log line; it makes
+no attempt to correlate log lines by queueid or \acronym{pid}, or to handle
 any of the other myriad complications discussed in
 \sectionref{complications}.  Its source code is unavailable, so it could
-not be used as a base for this project, even if it fulfilled all other
+not be used as a base for this project, even if it fulfilled all the other
 requirements.
@@
 
-Isoqlog's report misses most of the information gathered by \parsername{}:
-the data extracted is limited to the number of mails sent by each sender,
+Isoqlog's report lacks most of the information gathered by \parsername{}:
+the data it extracts is limited to the number of mails sent by each sender,
 and it only reports on senders from the domains listed in its configuration
@@
 files was pointless.  It does maintain a record of data previously
-extracted and the newly extracted data is merged into it; the format of the
-data store is undocumented.  Almost no documentation is provided with
-Isoqlog, little more than installation instructions.  It does not utilise
-rejection log lines in any way, so is unsuitable for the purposes of this
-project.  Its parsing is completely inextensible, indeed is almost
-incomprehensible, relying on \texttt{scanf(3)}, unexplained fixed offsets,
-and low level string manipulation; it is the opposite end of the spectrum
-to \parsernames{} parsing.  It does not handle any of the complications
+extracted and newly extracted data is added to it; the format of the data
+store is undocumented.  Almost no documentation is provided with Isoqlog,
+little more than installation instructions.  It does not utilise rejection
+log lines in any way, so is unsuitable for the purposes of this project.
+Its parsing is completely inextensible, indeed is almost incomprehensible,
+relying on \texttt{scanf(3)}, unexplained fixed offsets, and low level
+string manipulation; it is the opposite end of the spectrum to
+\parsernames{} parsing.  It does not handle any of the complications
 discussed in \sectionref{complications}, does not gather the breadth of
@@
 AWStats can produce simple graphs from many different services' log files,
-but supporting numerous log files formats without special purpose code
-limits its functionality.  The data it can extract from an \acronym{MTA}
-log file is limited in comparison to \parsername{}: time2, email, email\_r,
-host, host\_r, method, url, code, and bytesd.  No explanation for any of
-those fields is provided in the documentation at
+but supporting numerous log file formats without special purpose code
+limits its functionality.  The data it can extract from Postfix log files
+is limited in comparison to \parsername{}: time2, email, email\_r, host,
+host\_r, method, url, code, and bytesd.  No explanation for any of those
+fields is provided in the documentation at
 \urlLastChecked{http://awstats.sourceforge.net/docs/awstats_faq.html\#MAIL}{2008/11/23},
@@
 determine what data is missing in comparison to \parsername{}, which fully
-documents all data that it extracts.  AWStats coerces Postfix log files
-into the log file format used by the Apache web server, for analysis by
-AWStats' HTTP log file parser.  The converting parser only deals with a
-small portion of the log lines generated by Postfix, silently skipping
-those it cannot deal with, and does not distinguish between different types
-of rejection; extending it to parse all log lines would be at least as much
+documents all the data it extracts.  AWStats coerces Postfix log files into
+the log file format used by the Apache web server, for analysis by AWStats'
+HTTP log file parser.  The converting parser only deals with a small
+portion of the log lines generated by Postfix, silently skipping those it
+cannot parse, and does not distinguish between different delivery attempt
+rejections; extending it to parse all log lines would be at least as much
 work as writing a new parser.  It does correlate log lines by queueid (not
 by pid), but it does not deal with any of the other complications described
-in \sectionref{complications}.  AWStats supports saving data, but the
-format of the data store is not documented.  It also supports reading
-compressed log files, but that functionality was not tested.
+in \sectionref{complications}.  AWStats supports saving data extracted from
+log files, but the format of the data store is not documented.  It also
+supports reading compressed log files, but that was not tested.
 
-When tested with the \numberOFlogFILES{} test log files AWStats' reported
-that it parsed 9,240,075 (88.70\%) of 10,416,129 log lines, skipping
-1,176,050 (11.29\%) corrupt log lines; the \numberOFlogFILES{} log files
-contain \numberOFlogLINES{} log lines, so AWStats parsed only 15.21\% of
-the log lines, declared 1.93\% were corrupt, and ignored the remaining
-82.85\%.  The author did not verify the correctness of the parsing of those
-log lines AWStats did parse.
+When tested with the \numberOFlogFILES{} test log files, AWStats reported
+that it parsed 9,240,075 (88.709\%) of 10,416,129 log lines, skipping
+1,176,050 (11.290\%) corrupt log lines.  However, the \numberOFlogFILES{}
+test log files contain \numberOFlogLINES{} log lines, so AWStats parsed
+only 15.217\% of the log lines, declared that 1.936\% were corrupt, and
+ignored the remaining 82.846\%.  The parsing results were not examined in
+detail or verified.
 
@@
 \begin{itemize}
+
+    \squeezeitems{}
 
@@
 
-AWStats was not a suitable base for this project, because it assumes that
+AWStats was not a suitable base for this project because it assumes that
 all log files can be rewritten to be compatible with web server log files,
@@
 Postfix log files without using AWStats, i.e.\ would require writing a
-separate parser.  It may be possible to use AWStats' graphing capabilities
-to generate reports, by generating input for AWStats from the data
-extracted by \parsername{}.
+separate parser anyway.  It may be possible to use AWStats' graphing
+capabilities to generate reports, by generating input for AWStats from the
+database populated by \parsername{}.
 
@@
 
-YAALA uses a plugin-based system to analyse log files and produce reports
-in HTML format, with all the parsing and report generation handled by
-plugins.  Using YAALA as a base for this project would have been as much
-work as starting from scratch, as both the input and output modules would
-need to be written specially; it might be more work to implement a parser
-within the constraints of YAALA\@.  YAALA supports storing previously
-gathered data using Perl's Storable module, so other Perl programs can use
-Storable to load, examine, and optionally modify the data; \parsername{}
-uses a well documented database that is accessible from most common
-programming languages.  Information about how YAALA stores data was gleaned
-from the source code, as the documentation is sadly lacking.
+YAALA uses plugins to analyse log files and produce reports in HTML format.
+Using YAALA as a base for this project would have been as much work as
+starting from scratch, because both the input and output modules would need
+to be written specially; it might be more work to implement a parser within
+the constraints of YAALA rather than independently.  YAALA supports storing
+previously gathered data using Perl's Storable module, so other Perl
+programs could use Storable to load, examine, and optionally modify the
+data; \parsername{} uses a well documented database that is accessible from
+most common programming languages.  Information about how YAALA stores data
+was gleaned from the source code, because the format is undocumented and
+differs amongst plugins.
 
-YAALA provides a Postfix parser that extracts the following of fields from
+YAALA provides a Postfix parser that extracts the following fields from
 specific log lines:
 
-\begin{eqlist}
+\begin{boldeqlist}
 
@@
 
-\end{eqlist}
+\end{boldeqlist}
 
-\noindent{}YAALA's Postfix parser extracts some of the fields \parsername{}
-does: for client and server it stores either the \acronym{IP} address or
-the hostname, not both; it omits the HELO hostname, queueid, \acronym{SMTP}
-and enhanced status codes, size of each accepted mail, start and end times,
-timestamps for each log line, and message ID\@.  It extracts some data that
-\parsername{} does not: how many times delivery was deferred for each mail;
-this information can be determined from the database populated by
-\parsername{} if desired.  Unlike \parsername{}, YAALA does not maintain
-separate counters for each restriction; this rules out the possibility of
-using the collected data for optimisation, testing, or understanding of
-restrictions.  YAALA's Postfix parser does not deal with the complications
-explained in \sectionref{complications}, except it does correlate log lines
-by queueid.
+YAALA's Postfix log file parser extracts some of the fields \parsername{}
+does: for client and server it stores either the IP address or the
+hostname, not both; it omits the HELO hostname, queueid, \acronym{SMTP} and
+enhanced status codes, size of each accepted mail, start and end times,
+timestamps for each log line, and message-id\@.  It extracts one piece of
+data that \parsername{} does not: how many times delivery was deferred for
+each mail; this information can be determined from the database populated
+by \parsername{} if desired.  Unlike \parsername{}, YAALA does not maintain
+separate counters for different delivery attempt rejections, precluding the
+possibility of using the collected data for optimisation, testing, or
+understanding of restrictions.  YAALA's Postfix log file parser does not
+deal with the complications explained in \sectionref{complications}, except
+that it does correlate log lines by queueid.
 
@@
 \newline{}  The underlying reason for this is that YAALA only parsed 408
-(0.11\%) of 360632 log lines in the first log file; it was not tested with
+(0.11\%) of 360,632 log lines in the first log file; it was not tested with
 the remainder of the \numberOFlogFILES{} log files.
@@
 
-In summary, YAALA provides a Postfix parser that tries to parse the most
-common Postfix log lines only, provides reasonably flexible report
-generation from the limited data extracted, but has no facilities to
-extract data for use in other tools.
+In summary, YAALA provides a Postfix log file parser that unsuccessfully
+attempts to parse only the most common Postfix log lines, provides
+reasonably flexible report generation from the limited data extracted, but
+has no facilities to extract data for use in other tools.
 
@@
 of log file.  It takes a similar approach to YAALA, using plugins to parse
-different log file types.  The data extracted by its Postfix parser is not
-clearly documented:
+different log file types.  The data extracted by its Postfix log file
+parser is not clearly documented: \textit{The email servers' reports will
+show you the number of deliveries and the volume of email delivered by day,
+the domains from which you receive or send the most emails, the relays most
+used, etc\empty{}.\/}
 
-\begin{quotation}
-
-    \noindent{} The email servers' reports will show you the number of
-    deliveries and the volume of email delivered by day, the domains from
-    which you receive or send the most emails, the relays most used,
-    etc\empty{}.
-
-\end{quotation}
-
-\noindent{}Examining the source code reveals that the parser looks for
+Examining the source code reveals that Lire looks for
 \texttt{<key>=<value>} pairs in each log line, extracts them, and
@@
 data: HELO hostname, queueid, \acronym{SMTP} code, sender and recipient
-addresses, and size of accepted mails.  It is unclear if the parser will
-extract any further data.  Lire misses the following fields extracted by
-\parsername{}: client and server \acronym{IP} address and hostname, start
-and end times, enhanced status code, delivery delays, timestamps of each
-log line, and message ID\@.
+addresses, and size of accepted mails.  Lire misses the following fields
+extracted by \parsername{}: client and server IP address and hostname,
+start and end times, enhanced status code, delivery delays, timestamps of
+each log line, and message-id\@.
 
-Lire supports multiple output formats for generated reports (text, HTML,
-PDF, and Excel 95) but the reports do not appear to be customisable;
-\parsername{} does not produce any reports.  Lire's report is not as
-detailed as Pflogsumm's, and it is considerable harder to configure.  Lire
-supports saving extracted data for later report generation, but the format
-of this data store is undocumented; given the source code it should be
-possible, with enough time and effort, to understand the format.
+Lire supports multiple output formats for the reports it generates (text,
+HTML, PDF, and Excel 95) but the reports do not appear to be customisable,
+and are not as detailed as Pflogsumm's; \parsername{} does not produce any
+reports.  Lire supports saving extracted data for later report generation,
+but the format of this data store is undocumented; given the source code it
+should be possible, with enough time and effort, to understand the format.
 \parsername{} uses an \acronym{SQL} database to make accessing the
@@
 mail should be available; its parser extracts only part of the available
-data and makes no attempt to deal with the other complications described in
+data and does not attempt to deal with the other complications described in
 \sectionref{complications}.  When testing Lire on the \numberOFlogFILES{}
 test log files it performed reasonably well: the numbers it reports appear
-reasonable, and the subset verified by the author were correct.  Its report
-provided summaries of:
+accurate, and the subset verified by the author were correct.  Its report
+provides summaries of:
 
 \begin{itemize}
+
+    \squeezeitems{}
 
@@
 Lire would not be a suitable base for this project: it does not extract
-enough data; does not deal with rejections in any way; does not make the
-extracted data easily available to other programs.  Its parser is
-in-extensible but could easily be replaced, but that would require writing
-a parser from scratch, so would not be any less work.  \parsername{} could
-possibly be used to parse Postfix log files for Lire, but the difficulty
-may outweigh the benefits.  Lire's data store may not be suitable for
-storing the data extracted by \parsername{}, but the lack of documentation
-hinders any evaluation.  As with YAALA this approach has not been attempted
-by the author.
+enough data; does not deal with rejected delivery attempts in any way; does
+not make the extracted data easily available to other programs.  Its parser
+is in-extensible; it could easily be replaced, but that would require
+writing a parser from scratch, so would not be any less work.
+\parsername{} could possibly be used to parse Postfix log files for Lire,
+but the difficulty may outweigh the benefits, e.g.\ Lire's data store may
+not be capable of storing the data extracted by \parsername{}, but the lack
+of documentation hinders any evaluation.  As with YAALA this approach has
+not been attempted by the author.
 
@@
 It also counts the number of log lines parsed and skipped.  It omits client
-\acronym{IP} address and hostname, server \acronym{IP} address, HELO
-hostname, queueid, timestamps of each log line, enhanced status code, and
-message ID\@.  Log lines are correlated based on the queueid (called
-sessionname [sic] within Logrep), but not by \acronym{pid}.  The parsing is
-error prone: empty fields are saved when the log line does not match the
-regex, though it appears that they will not overwrite existing data.  Most
-notably rejections are completely ignored, making it unsuitable for the
+IP address and hostname, server IP address, HELO hostname, queueid,
+timestamps of each log line, enhanced status code, and message-id\@.  Log
+lines are correlated based on the queueid (called sessionname [sic] within
+Logrep), but not by \acronym{pid}.  The parsing is error prone: empty
+fields are saved when the log line does not match the regex, though it
+appears that they will not overwrite existing data.  Most notably, rejected
+delivery attempts are completely ignored, making it unsuitable for the
 purposes of this project.  It does not try to address any of the
-complications in \sectionref{complications} except for correlating by
-queueid.
+complications in \sectionref{complications} except for correlating log
+lines by queueid.
 
@@
 documentation).  It requires a web browser to interact with it, so
-automated log file processing will be difficult, whereas enabling automated
-processing is a key part of \parsernames{} design.  Sadly all the author's
-attempts to use Logrep failed, as it was unable to access the log files
-selected; this appears to be a bug rather than operator error.  If it was
-caused by operator error, the interface needs improvement as the (minimal)
-instructions were followed as closely as possible, and multiple attempts
-were made.  Because parsing failed it was not possible to review the
-reports Logrep can generate (available in HTML only), or to examine the
-(undocumented) format it uses to save extracted data for subsequent reuse.
+automated log file processing would be difficult, whereas enabling
+automated processing is a key part of \parsernames{} design.  Sadly, all
+the author's attempts to use Logrep failed, as it was unable to access the
+log files selected for parsing; this appears to be a bug rather than
+operator error.  If the problem was caused by operator error, the interface
+needs improvement because the (minimal) instructions were followed as
+closely as possible, and multiple attempts were made.  Because parsing
+failed it was not possible to review the reports Logrep can generate
+(available in HTML only), or to examine the (undocumented) format it uses
+to save extracted data for subsequent reuse.
 
 Logrep extracts far less data from Postfix log files than \parsername{},
-completely ignores rejections, is effectively undocumented, does not deal
-with the more complicated aspects of Postfix log files, and at the time of
-writing does not work properly.
+completely ignores rejected delivery attempts, is effectively undocumented,
+does not deal with the more complicated aspects of Postfix log files, and
+does not work properly.
 
@@
 This chapter has reviewed ten programs that perform basic Postfix log file
-parsing (some to a greater level of detail than others).   None of the
+parsing, some to a greater level of detail than others.  None of the
 reviewed parsers collect the breadth of information gathered by
-\parsername{}, and are not designed to be extensible to handle new log
-lines.  Some correlate log lines by queueid (none correlate by pid); none
-deal with any of the other complications described in
-\sectionref{complications}.  All of the reviewed parsers generate a report,
-and some provide a greater or lesser degree of customisation.  Most have a
-data store, but only \acronym{LMA} provides any documentation on its
+\parsername{}, and none are designed to be extensible to handle new log
+lines.  Some correlate log lines by queueid, but none correlate by pid;
+none deal with any of the other complications described in
+\sectionref{complications}.  All of the reviewed parsers generate a
+report, and some provide a greater or lesser degree of customisation.  Most
+have a data store, but only \acronym{LMA} provides any documentation on its
 format; some deliberately make the data store inaccessible to other tools.
@@
         \item [\nibeta{}] \parsername{} is the only parser with
-            documentation for its data store.
+            documentation for the format of its data store.
 

### 2009-04-17 doc/logparser-results.tex
Corrections based on proofreading to the preface and chapters 1-3.

@@
 \Tableref{Number of rules per Postfix component} shows the number of rules
-for each Postfix component; \graphref{Mean number of rules tried per log
+for each Postfix component; \graphref{Mean number of rules used per log
 line} shows the drop in the mean number of rules tried per log line for log
-files containing a mail loop, and \graphref{Mean number of rules tried per
+files containing a mail loop, and \graphref{Mean number of rules used per
 log line for each Postfix component} shows the mean number of rules tried
@@
 \showgraph{build/graph-average-number-of-rules-tried-per-log-line}{Mean
-number of rules tried per log line}{Mean number of rules tried per log
-line}
+number of rules used per log line}{Mean number of rules used per log line}
 
 \showgraph{build/graph-average-number-of-rules-tried-per-program}{Mean
-number of rules tried per log line for each Postfix component}{Mean number
-of rules tried per log line for each Postfix component}
+number of rules used per log line for each Postfix component}{Mean number
+of rules used per log line for each Postfix component}
 
@@
 
-\Graphref{Parsing time relative to shuffled ordering graph} shows the
-parsing times of optimal and reverse orderings relative to shuffled
-ordering; the mean relative parsing times for different groupings of log
-files are given in \tableref{Parsing time relative to shuffled ordering
-table}.  This optimisation provides a mean reduction in parsing time of
+\Graphref{Parsing time of optimal and reverse orderings relative to
+shuffled ordering graph} shows the parsing times of optimal and reverse
+orderings relative to shuffled ordering; the mean relative parsing times
+for different groupings of log files are given in \tableref{Parsing time of
+optimal and reverse orderings relative to shuffled ordering table}.  This
+optimisation provides a mean reduction in parsing time of
 \input{build/include-optimal-ordering-parsing-time-reduction-other-logs}\%
@@
 when a mail loop occurs and the distribution of log lines is unusual.
-\Tableref{Parsing time relative to shuffled ordering table} shows that
-differences in rule ordering have less effect on parsing time when parsing
-log files 22 \& 62--68, because of the different distribution of log lines
-in those log files.  A careful examination of \graphref{Parsing time
+\Tableref{Parsing time of optimal and reverse orderings relative to
+shuffled ordering table} shows that differences in rule ordering have less
+effect on parsing time when parsing log files 22 \& 62--68, because of the
+different distribution of log lines in those log files.  A careful
+examination of \graphref{Parsing time of optimal and reverse orderings
 relative to shuffled ordering graph} shows that, for the first log file
@@
 
-\showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time
-relative to shuffled ordering}{Parsing time relative to shuffled ordering
-graph}
+\showgraph{build/graph-optimal-and-reverse-vs-shuffle}{Parsing time of
+optimal and reverse orderings relative to shuffled ordering}{Parsing time
+of optimal and reverse orderings relative to shuffled ordering graph}
 
-\showtable{build/include-optimal-and-reverse-vs-shuffle}{Parsing time
-relative to shuffled ordering}{Parsing time relative to shuffled ordering
-table}
+\showtable{build/include-optimal-and-reverse-vs-shuffle}{Parsing time of
+optimal and reverse orderings relative to shuffled ordering}{Parsing time
+of optimal and reverse orderings relative to shuffled ordering table}
 
@@
 \& 62--68), because the mean number of rules tried per log line is lower
-(see \graphref{Mean number of rules tried per log line}).  The performance
+(see \graphref{Mean number of rules used per log line}).  The performance
 of optimal ordering relative to the oracle is much worse when parsing the
@@
 
-\subsection{Scalability As The Number Of Rules Rises}
+\subsection{Scalability As The Ruleset Grows}
 
@@
         child mails waiting to be tracked with a parent; see
-        \sectionref{Re-injected mails} for details.  XXX CHECK IF THERE IS
-        ANY MENTION OF THIS IN THE COMPLICATION, AND ADD IF MISSING\@.
+        \sectionref{Re-injected mails} for details.
 

### 2009-04-17 doc/logparser.tex
Corrections based on proofreading to the preface and chapters 1-3.

@@
     that implements it, explaining the difficulties encountered and the
-    solutions developed, with reference to an implementation that stores
-    data gleaned from the log files in an SQL database.  The gathered data
-    can be used to optimise current anti-spam measures, to produce
-    statistics showing how effective those measures are, or to provide a
-    baseline to test new anti-spam measures against.  The Postfix log file
-    parser needs to be very precise and strict when parsing log files, yet
-    must allow users to easily adapt or extend it to parse new log lines,
-    without requiring that the user have an in-depth knowledge and
-    understanding of the parser's internal workings.  The newly developed
-    architecture is designed to make the process of parsing new inputs as
-    simple as possible, enabling users to trivially add new rules (to parse
-    variants of known inputs) and relatively easily add new actions (to
-    process a previously unknown category of inputs).  The parser
-    implemented for this project is evaluated on the criteria of efficiency
-    and coverage of Postfix log files, demonstrating that the conflicting
-    goals of efficiency and accuracy can be balanced, and one need not be
-    sacrificed to achieve the other.
+    solutions developed, and evaluating the parser on the criteria of
+    efficiency and Postfix log file coverage.  The parser stores data
+    gleaned from the log files in an SQL database; the gathered data can be
+    used to optimise current anti-spam measures, to produce statistics
+    showing how effective those measures are, or to provide a baseline to
+    test new anti-spam measures against.  The Postfix log file parser needs
+    to be very precise and strict when parsing, yet must allow users to
+    easily adapt or extend it to parse new log lines, without requiring
+    that the user have an in-depth knowledge and understanding of the
+    parser's internal workings.  The newly developed architecture is
+    designed to make the process of parsing new inputs as simple as
+    possible, enabling users to trivially add new rules (to parse variants
+    of known inputs) and relatively easily add new actions (to process a
+    previously unknown category of inputs).  The parser implemented for
+    this project is evaluated on the criteria of efficiency and coverage of
+    Postfix log files, demonstrating that the conflicting goals of
+    efficiency and accuracy can be balanced, and one need not be sacrificed
+    to achieve the other.
 
     \SVN$Id$
-    \begin{center}
-        \SVNId{}
-    \end{center}
 

### 2009-04-20 doc/logparser-architecture.tex
Corrections from proofreading.
Corrections from proofreading for chapters 4 and 5.

@@
 
-To avoid cluttering the explanation of the parser architecture with the
-details involved in implementing a parser for Postfix log files, the two
-topics have been separated.  This chapter presents the architecture
+To avoid cluttering the description of the parser architecture with the
+details involved in implementing a parser for Postfix log files, each topic
+has been given its own chapter.  This chapter presents the architecture
 developed for this project, beginning with the overall architecture and
-design, followed by the three components of the architecture: Framework,
-Actions, and Rules.  This chapter centers on the theoretical,
-implementation-independent aspects of the architecture; the practical
-difficulties of writing a parser for Postfix log files are covered in
-detail in \sectionref{Postfix Parser Implementation}.
+design, followed by detailed documentation of the three components of the
+architecture: Framework, Actions, and Rules.  This chapter centers on the
+theoretical, implementation-independent aspects of the architecture; the
+practical difficulties of writing a parser for Postfix log files are
+covered in detail in \sectionref{Postfix Parser Implementation}.
 
@@
 background}) that log files produced by Postfix may vary widely from host
-to host, depending on the set of restrictions chosen by the administrator.
-With this in mind, one of the architecture's design aims was to make adding
-new rules to parse new inputs as effortless as possible, to enable
-administrators to properly parse their own log files.  The solution
-developed is to divide the architecture into three parts: Framework,
-Actions, and Rules.  Each will be discussed separately, but first an
-overview:
+to host.  With this in mind, one of the architecture's design aims was to
+make parsing new inputs as effortless as possible, to enable administrators
+to properly parse their own log files.  The solution developed is to divide
+the architecture into three parts: Framework, Actions, and Rules.  Each
+will be discussed separately, but first an overview:
 
-\begin{eqlist}[\def\makelabel#1{\bfseries#1}]
+\begin{boldeqlist}
 
@@
 
-    \item [Actions] Each action performs the work required to deal with a
+    \item [Actions] Each action performs the processing required for a
         single \textit{category\/} of inputs, e.g.\ rejecting a delivery
@@
 If the input is not recognised by any of the rules, the framework issues a
-warning; the framework can continue parsing after this, although for some
-inputs stopping immediately may be preferable.
+warning; the framework will usually continue parsing after this, although
+for some inputs stopping immediately may be preferable.
 
@@
 provided by the framework, freeing them to concentrate on the task of
-accurately and correctly processing the inputs and the information provided
-by rules; rules are responsible for classifying inputs, and extracting data
+accurately and correctly processing inputs and the information provided by
+rules; rules are responsible for recognising inputs, and extracting data
 from those inputs for processing by actions.
@@
 parse new inputs without modifying the core parsing algorithm.  Adding a
-new rule with the action to invoke and a regex to recognise the inputs is
+new rule with the action to invoke and a regex to recognise inputs is
 trivial in comparison to understanding an entire parser, identifying the
-correct location to change, and making the appropriate changes.  Additions
-to a parser must be made without adversely affecting existing parsing,
+correct location to change, and making the appropriate changes.  Changes to
+a parser must be made without adversely affecting existing parsing,
 including any edge cases that are not immediately obvious; an edge case
 that occurs only four times in \numberOFlogFILES{} log files is described
-in Yet More Aborted Delivery Attempts
-(\sectionref{yet more aborted delivery attempts}).  More intrusive changes
-are more likely to introduce a bug, so reducing the extent of the changes
-is important.  Requiring changes to a parser's source code also complicates
+in Yet More Aborted Delivery Attempts (\sectionref{yet more aborted
+delivery attempts}).  The more intrusive the changes are, the more likely
+they are to introduce a bug, so reducing the extent of the changes is
+important.  Requiring changes to a parser's source code also complicates
 upgrades of the parser, because the changes must be preserved during the
 upgrade, and they may clash with changes made by the developer.  This
-architecture allows the user to add new rules without having to edit a
-parser, unless the new inputs cannot be processed by the existing actions.
+architecture allows the user to add new rules to a parser without having to
+edit it, unless the new inputs cannot be processed by the existing actions.
 If the new inputs do require new functionality, new actions can be added to
-the parser without having to modify existing actions; only in the rare
-event that the new actions need to cooperate with existing actions will
-more extensive changes be required.
+the parser without having to modify existing actions; only when the new
+actions need to cooperate with existing actions will more extensive changes
+be required.
 
-Some similarity exists between this architecture and William Wood's
-\acronym{ATN}~\cite{atns,nlpip}, used in Computational Linguistics for
-creating grammars to parse or generate sentences.  The resemblance between
+Some similarity exists between this architecture and William A.\ Wood's
+\acronym{ATN}~\cite{nlpip,atns}, used in Computational Linguistics to
+create grammars that parse or generate sentences.  The resemblance between
 the two (shown in \tableref{Similarities between ATN and this
@@
 
-The framework manages the parsing process and provides support functions,
-freeing the programmers writing actions to concentrate on writing
-productive code.  It links actions and rules, allowing either to be
+The framework manages the parsing process and provides support functions
+for the actions, freeing the programmers writing actions to concentrate on
+writing productive code.  It links actions and rules, allowing either to be
 improved independently of the other, and allows new rules to be written
@@
 intricacies and complications encountered during parsing.  The function
-that finds the rule recognising the input and invokes the requested action
-can be expressed in pseudo-code as:
+that finds the rule that recognises the input and invokes the requested
+action can be expressed in pseudo-code as:
 
@@
         if this rule recognises the input {
-            perform the action specified by the rule
+            invoke the action specified by the rule
             next INPUT
@@
     }
-    warn the user that the input was not parsed
+    warn the user that the input was not recognised
 }
@@
 
-            \item Some optimisation steps may also be performed during
-                loading of rules, as discussed in \sectionref{parser
-                efficiency}.
+        \end{itemize}
 
-        \end{itemize}
+        Some optimisation steps may also be performed during loading of
+        rules, as discussed in \sectionref{parser efficiency}.
 
@@
         \sectionref{actions in architecture}) to effectively push input
-        back onto the input stream.  A physical input (i.e.\ a single line
+        back onto the input stream.  A physical input (e.g.\ a single line
         read from the input stream) may contain multiple or partial logical
@@
         equivalent to parsing the physical inputs; in those cases the
-        framework should behave similarly to other architectures: present
-        the entire input stream at once, and discard the portion recognised
-        by the successful rule.
+        framework present the entire input stream at once, and discard the
+        recognised portion.
 
@@
 
-    \item [Save and load state]  The architecture can save its current
-        state, i.e.\ the shared storage it provides for actions, and reload
-        it later, so that information is not lost between parsing runs.
-        \parsername{} uses this functionality because mails may take some
-        time to deliver and thus have their log lines split between log
-        files; a compiler might store data structures it builds as it
-        parses different files.
+    \item [Save and load state]  The architecture can save the contents of
+        the shared storage it provides for actions, and reload it later, so
+        that information is not lost between parsing runs.  \parsername{}
+        uses this functionality because mails may take some time to deliver
+        and thus have their log lines split between log files; a compiler
+        might store data structures it builds as it parses different files.
 
@@
 Each action is a separate procedure written to process a particular
-category of input, e.g.\ rejecting a delivery attempt.  There may be many
-input variants within one input category; in general each action will
+category of input, e.g.\ rejection of a delivery attempt.  There may be
+many input variants within one input category; in general each action will
 handle one input category, with each rule recognising one input variant.
@@
 process different input variants in slightly different ways, but large
-changes in processing indicate the need for a new action and a new category
-of input; if an action becomes overly complicated it starts to turn into a
-monolithic parser, with too much logic contained in a single procedure.
+changes in processing within an action indicate the need for a new action
+and a new category of input; if an action becomes overly complicated it
+starts to turn into a monolithic parser, with too much logic contained in a
+single procedure.
 
@@
 and new requirements that are discovered during parser development is one
-of the strengths of this architecture.  Instead of writing a single
-monolithic function that processes every input and must be modified to
-support any new behaviour, with all the attendant risks of adversely
-affecting the existing parsing, when a new requirement arises an
-independent action can be written to satisfy it.  Sometimes the new action
-will require the cooperation of other actions, e.g.\ to set or check a
-flag, so actions are not always self-contained, but there will still be a
-far lower degree of coupling and interdependency than in a monolithic
-parser.
+of the strengths of this architecture.  When a new requirement arises an
+independent action can be written to satisfy it, instead of writing a
+single monolithic function that processes every input and must be modified
+to support any change in behaviour, with all the attendant risks of
+adversely affecting the existing parsing.  Sometimes the new action will
+require the cooperation of other actions, e.g.\ to set or check a flag, so
+actions are not always self-contained, but there will still be a far lower
+degree of coupling and interdependency than in a monolithic parser.
 
@@
 
-Actions may return a modified input line to the framework that will be
-parsed as if read from the input stream, allowing for a simplified version
-of cascaded parsing~\cite{cascaded-parsing}.  This powerful facility allows
-several rules and actions to parse a single input, potentially simplifying
-both rules and actions.  A simple example is to have one rule and action
-removing comments from inputs, so that other rules and actions do not
-have to handle comments at all; obviously if comment characters can be
-escaped or embedded in quoted strings the implementation must be careful
-not to remove those.  For some inputs this kind of pre-processing can
-greatly simplify parsing, echoing the simplification provided by the
-framework presenting rules and actions with logical inputs rather than
-physical inputs.  A more complex example use of cascaded parsing is
+Actions may modify the input they process and return it to the framework,
+where it will be parsed as if read from the input stream, allowing for a
+simplified version of cascaded parsing~\cite{cascaded-parsing}.  This
+powerful facility allows several rules and actions to parse a single input,
+potentially simplifying both rules and actions.  A simple example is to
+have one rule and action removing comments from inputs, so that other rules
+and actions do not have to handle comments at all; obviously if comment
+characters can be escaped or embedded in quoted strings the implementation
+must be careful not to remove those.  For some inputs this kind of
+pre-processing can greatly simplify parsing, echoing the simplification
+provided by the framework presenting rules and actions with logical inputs
+rather than physical inputs.  A more complex example of cascaded parsing is
 evaluating simple arithmetic expressions, where sub-expressions enclosed in
@@
 
-Classifying inputs using the rules is simple: the first
-rule to recognise the input determines the action that will be invoked;
-there is no backtracking to try alternate rules; no attempt is made to pick
-a \textit{best\/} rule.  \sectionref{attaching conditions to rules}
-contains an example which requires that the rules are used in a specific
-order to correctly parse the input, so a mechanism is needed to allow the
-author of the rules to specify that ordering.  The framework should sort
-the rules according to a priority attribute specified by the rule author,
-allowing fine-grained control over the order that rules are used in.  The
-priority attribute may be implemented as a number, or as a range of values,
-e.g.\ low, medium, and high, or in a different fashion entirely if it suits
-the implementation.  Rule ordering for efficiency is a separate topic that
-is covered in \sectionref{rule ordering for efficiency}; overlapping rules
-are discussed in \sectionref{overlapping rules in architecture}.
+Using the rules is simple: the first rule to recognise the input determines
+the action that will be invoked; there is no backtracking to try alternate
+rules; no attempt is made to pick a \textit{best\/} rule.
+\sectionref{attaching conditions to rules} contains an example which
+requires that the rules are used in a specific order to correctly parse the
+input, so a mechanism is needed to allow the author of the rules to specify
+that ordering.  The framework should sort the rules according to a priority
+attribute specified by the rule author, allowing fine-grained control over
+the order that rules are used in.  The priority attribute may be
+implemented as a number, or as a range of values, e.g.\ low, medium, and
+high, or in a different fashion entirely if it suits the implementation.
+Rule ordering for efficiency is a separate topic that is covered in
+\sectionref{rule ordering for efficiency}; overlapping rules are discussed
+in \sectionref{overlapping rules in architecture}.
 
@@
 
-$\text{\textless{}input\textgreater{}}~\mapsto{}~\text{rule-1}~|~\text{rule-2}~|~\text{rule-3}~|~\dots~|~\text{rule-n}$
+\tab{}$\text{\textless{}input\textgreater{}}~\mapsto{}~\text{rule-1}~|~\text{rule-2}~|~\text{rule-3}~|~\dots~|~\text{rule-n}$
 
@@
 \subsection{Adding New Rules}
+
+\label{adding new rules in architecture}
 
@@
 
-    \item Create a new category of inputs and a new action to process
+    \item Create a new category of inputs, write a new action to process
         inputs from the new category, and write a new rule pairing the new
@@
 e.g.\ taking advantage of dynamic languages' support for evaluating code at
-run-time, or embedding a language like Lua.  If an implementation is going
-to use conditions so complex that they will require a programming language
-capable of implementing a Turing machine, the design may need to be
-re-thought, including the decision to use this architecture --- there may
-be other architectures more suitable.
+run-time, or embedding a language like Lua into the parser.  If an
+implementation is going to use conditions so complex that they will require
+a programming language capable of implementing a Turing machine, the design
+may need to be revisited, including the decision to use this architecture
+--- there may be other architectures more suitable.
 
@@
 specifying the component whose log lines it recognises, reducing the number
-of rules that will be used when classifying an input (see \sectionref{rules
-in implementation} for details) and increasing the chance that the input
-will be correctly recognised.
+of rules that will be used when classifying a log line (see
+\sectionref{rules in implementation} for details) and increasing the chance
+that the log line will be correctly recognised.
 
@@
 either actions or rules; it is easier to understand how a variable's value
-will be used and changed if it is set by rules only rather than by actions,
-because the chain of checking and setting can be followed from rule to
-rule.  The downside to actions setting variables used in rule conditions is
-action at a distance: understanding when a rule's condition will be true or
-false requires understanding not just every other rule but also every
-action.  The framework probably does not need to support a high level of
-complexity and flexibility when rules are setting variables; however, if
-the framework supports complex conditions, that code can probably be easily
-extended to support complex variable assignments too.  The level of
-complexity the framework supports when evaluating conditions and setting
-variables has two costs that must be taken into account when designing a
-parser: the difficulty of implementation, and the difficulty of
-understanding writing correct rules; the flexibility provided by complex
-conditions may be outweighed by the difficulty in understanding the
-interactions between rules that use complex conditions.
+will be used and changed if it is set by rules only, rather than by
+actions, because the chain of checking and setting variables can be
+followed from rule to rule.  The downside to actions setting variables that
+are used in rule conditions is action at a distance: understanding when a
+rule's condition will be true or false requires understanding not just
+every other rule but also every action.  The framework probably does not
+need to support a high level of complexity and flexibility when rules are
+setting variables; however, if the framework supports complex conditions,
+that code can probably be easily extended to support complex variable
+assignments too.  The level of complexity the framework supports when
+evaluating conditions and setting variables has two costs that must be
+taken into account when designing a parser: the difficulty of
+implementation, and the difficulty of understanding or writing correct
+rules; the flexibility provided by complex conditions may be outweighed by
+the difficulty in understanding the interactions between rules that use
+them.
 
@@
 incorrect; the rules could be extended with new rules to parse quoted
-strings, excluding comments from within strings.
-
-Rules one and two will be used when the parser is parsing code, not
-comments: rule one recognises a comment that is contained within one line
-and leaves the parser's state unchanged; rule two recognises the start of a
-comment and changes the parser's state to parsing comments.  Rules three
+strings, excluding comments from within strings.  Rules one and two will be
+used when the parser is parsing code, not comments: rule one recognises a
+comment that is contained within one line and leaves the parser's state
+unchanged; rule two recognises the start of a comment and changes the
+parser's state to parsing comments instead of parsing code.  Rules three
 and four will be used when the parser is parsing a comment: rule three
@@
         \tablemiddleline{}%
-        1     & \verb!/\*.*?\*/!  & expecting == code     & expecting = code      \\
-        2     & \verb!/\*.*!      & expecting == code     & expecting = comment   \\
-        3     & \verb!.*\*/!      & expecting == comment  & expecting = code      \\
-        4     & \verb!.*!         & expecting == comment  & expecting = comment   \\
+        1     & \verb!/\*.*?\*/!  & state == parsing code     & state = parsing code      \\
+        2     & \verb!/\*.*!      & state == parsing code     & state = parsing comment   \\
+        3     & \verb!.*\*/!      & state == parsing comment  & state = parsing code      \\
+        4     & \verb!.*!         & state == parsing comment  & state = parsing comment   \\
         \tablebottomline{}%
     \end{tabular}
+
+    Note that ``state'' is merely a descriptive name for the variable; the
+    variable can called anything at all.
+
 \end{table}
@@
 inconsistent parsing and data extraction because the first rule to
-recognise the input wins, and the order in which rules are tried against
-each input might change between parser invocations.  Overlapping rules are
-frequently a requirement, allowing a more specific rule to recognise some
-inputs and a more general rule to recognise the remainder, e.g.\ separating
+recognise the input wins, and the order in which rules are used might
+change between parser invocations.  Overlapping rules are frequently a
+requirement, allowing a more specific rule to recognise some inputs and a
+more general rule to recognise the remainder, e.g.\ separating
 \acronym{SMTP} delivery to specific sites from \acronym{SMTP} delivery to
-the rest of the world.  Allowing overlapping rules simplifies both the
-general rule and the more specific rule; additionally rules from different
-sources can be combined with a minimum of prior cooperation or modification
+the rest of the world.  Using overlapping rules simplifies both the general
+rule and the more specific rule; additionally rules from different sources
+can be combined with a minimum of prior cooperation or modification
 required.  Overlapping rules should have a priority attribute to specify
@@
 Overlapping rules can be detected by visual inspection, or a program may be
-written to analyse the rule's regexes.  Traditional regexes are equivalent
-in computational power to \acronym{FA} and can be converted to
+written to analyse the regexes in a ruleset.  Traditional regexes are
+equivalent in computational power to \acronym{FA} and can be converted to
 \acronym{FA}, so regex overlap can be detected by finding a non-empty
@@
 cannot always be implemented using a \acronym{PDA}, so intersection may be
-intractable for some combinations, e.g.:
+intractable sometimes, e.g.:
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
@@
 
-When detecting overlap any conditions attached to rules must be taken into
+When detecting overlap, any conditions attached to rules must be taken into
 account, because two rules whose regexes overlap may have conditions
@@
 rules, and then check for overlap between the regexes of each pair of rules
-with overlapping conditions.  Depending on the conditions employed a rule
-may overlap with multiple rules, and rule overlap is not transitive, e.g.\
-given these three artificial conditions:
+with overlapping conditions.  Rule overlap is not transitive, e.g.\ given
+these three conditions:
 
@@
 
-    $x$ & = & cost of checking for overlap between two regexes    \\
     $y$ & = & cost of checking for overlap between two conditions \\
     $o$ & = & set of pairs of rules with overlapping conditions   \\
+    $x$ & = & cost of checking for overlap between two regexes    \\
 
@@
 
-In the worst case no overlap will be found between conditions, so all pairs
-or regexes must be checked for overlap and $|o|$ above will be equal to
-$n^{2}$.  For this approach to be more efficient than the naive approach
-$y$ must be significantly lower than $x$.  If $y$ is higher than $x$ then
-the checks for overlap should be performed in the opposite order: check for
-regex overlap first, then check for condition overlap only between pairs of
-rules with overlapping regexes.
+In the worst case $|o|$ above will be equal to $n^{2}$.  For this approach
+to be more efficient than the naive approach, $y$ must be significantly
+lower than $x$.  If $y$ is higher than $x$ then the checks for overlap
+should be performed in the opposite order: check for regex overlap first,
+then check for condition overlap only between pairs of rules with
+overlapping regexes.
 
-Once conditions pass a certain level of complexity determining if two
+Once conditions pass a certain level of complexity, determining if two
 conditions overlap becomes intractable, because it requires so much
-knowledge of other rules and possibly even actions.  E.g.\ given two rules
-with conditions \verb!verbose == true! and \verb!silent == true!, logically
-these rules should not overlap, yet there is nothing to stop both variables
-being set to true by one or more rules or actions.  If variables used in
-conditions can be set by actions, determining if two conditions overlap is
-impossible: the Halting Problem show that it is impossible for one program
-to answer questions about another program's behaviour if the inquiring
-program is implemented on a computational machine of the equal or lesser
-power to the program being inquired into.
+knowledge of other rules and possibly even actions.  For example, given two
+rules with conditions \verb!verbose == true! and \verb!silent == true!,
+logically these rules should not overlap, yet there is nothing to stop both
+variables being set to true by one or more rules or actions.  If variables
+used in conditions can be set by actions, determining if two conditions
+overlap is impossible: the Halting Problem shows that it is impossible for
+one program to answer questions about another program's behaviour if the
+inquiring program is implemented on a computational machine whose power is
+equal or less than the power of the computational machine the other program
+is implemented upon.
 
@@
 
-It is possible to define pathological regexes, which fall into two main
+It is possible to write pathological regexes, which fall into two main
 categories: regexes that match inputs they should not, and regexes that
@@
 overlapping rules, and would easily be noticed by visual inspection, but
-more complex regexes would be harder to spot.  Regexes that match more than
-they should are a problem not because of excessive resource usage, but
-because they may prevent the correct rule from recognising the input.  If
-an adaptive ordering system is used that prioritises rules that frequently
-recognise inputs (see \sectionref{rule ordering for efficiency}), then a
-rule with a regex that matches inputs it should not may be promoted up
-through the list, displacing an increasing number of correct rules as it
-moves.
+more complex regexes would be harder to find.  Regexes that match inputs
+more than they should are a problem not because of excessive resource
+usage, but because they may prevent the correct rule from recognising the
+input.  If an adaptive ordering system is used that prioritises rules that
+frequently recognise inputs (see \sectionref{rule ordering for
+efficiency}), then a rule with a regex that matches inputs it should not
+may be promoted up through the list, displacing an increasing number of
+correct rules as it rises.
 
-Usually excessive CPU time is consumed when a regex fails to match, and the
-regex engine backtracks many times because of alteration or nested
-quantifiers; successful matching is generally quite fast with such regexes,
-so problematic regexes may go unnoticed for some time.  E.g.\ with most
-regex engines matching double quoted strings with \verb!"([^"\\]+|\\.)*"!
-is very fast when there is a match, but when the match fails its
-computational complexity for a string of length $n$ is $O(2^{n})$;
-see~\cite{mastering-regular-expressions} for in-depth discussion of nested
-quantifiers, backtracking, alteration, and capturing groups.  Pathological
-regexes that consume excessive CPU time can be difficult to detect, both by
-visual inspection and by machine inspection, but if a regex is converted to
-a \acronym{FA} or the internal representation used by the regex engine, it
-may be possible to determine if nested quantifiers or other troublesome
-constructs are present.  Modern regex engines have addressed many of these
-problems, e.g.\ the regex to match double quoted strings given above fails
-immediately with Perl 5.10, regardless of the input length, because the
-regex engine looks for both of the required double quotes first.  Similarly
-Perl 5.10's regex engine optimises alterations starting with literal text
-into a trie, which has matching time proportional to the length of the
-alternatives, rather than the number of alternatives.  Perl regexes can use
-\verb!(?>pattern)!, which matches \verb!pattern! the first time the regex
-engine passes over it, and does not change what it originally matched if
-the regex engine backtracks over it, so troublesome regexes can use this to
-reduce the impact of backtracking; Prolog users will notice a similarity to
-the \verb'!' (cut) operator.  A presentation showing some of Perl 5.10's
-new regex features is available at
+Excessive CPU time is usually consumed when a regex fails to match an
+input, and the regex engine backtracks many times because of alteration or
+nested quantifiers; successful matching is generally quite fast with such
+regexes, so problematic regexes may go unnoticed for a long time.  For
+example, with most regex engines matching double quoted strings with
+\verb!"([^"\\]+|\\.)*"!  is very fast when there is a match, but when the
+match fails its computational complexity for a string of length $n$ is
+$O(2^{n})$; see~\cite{mastering-regular-expressions} for in-depth
+discussion of nested quantifiers, backtracking, alteration, and capturing
+groups.  Pathological regexes that consume excessive CPU time can be
+difficult to detect, both by visual inspection and by machine inspection,
+but if a regex is converted to a \acronym{FA} or the internal
+representation used by the regex engine, it may be possible to determine if
+nested quantifiers or other troublesome constructs are present.  Modern
+regex engines have addressed many of these problems, e.g.\ the regex to
+match double quoted strings given above fails immediately with Perl 5.10,
+regardless of the input length, because the regex engine looks for both of
+the required double quotes first.  Similarly, Perl 5.10's regex engine
+optimises alterations starting with literal text into a trie, which has
+matching time proportional to the length of the alternatives, rather than
+the number of alternatives.  Perl regexes can use \verb!(?>pattern)!, which
+matches \verb!pattern! the first time the regex engine passes over it, and
+does not change what it originally matched if the regex engine backtracks
+over it, alleviating problems caused by excessive backtracking; Prolog
+users will notice a similarity to the \verb'!' (cut) operator.  A
+presentation showing some of Perl 5.10's new regex features is available at
 \urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.
@@
 Turing-equivalent language, so enumerating pathological conditions is
-difficult if not pointless.  Conditions that examine the input using
-regexes can suffer from the same problems as rule regexes; conditions that
-check variables in uncomplicated ways may exhibit unexpected or incorrect
-behaviour, but are unlikely to exhibit pathological behaviour.  Deciding if
-a more complex condition's behaviour is pathological or simply a bug is
-difficult and to some extent is a matter of opinion.  When this
-architecture has received more widespread usage consensus should be reached
-on the topic of pathological conditions.
+difficult if not pointless.  Conditions that check variables or the input
+in uncomplicated ways may exhibit unexpected or incorrect behaviour, but
+are unlikely to exhibit pathological behaviour.  Deciding if a more complex
+condition's behaviour is pathological or simply a bug is difficult and to
+some extent is a matter of opinion.  When this architecture has received
+more widespread usage, consensus should be reached on the topic of
+pathological conditions.
 
@@
 how it achieves its design aim of being easily extensible for users, and
-the advantages that being easily extensible brings to parser authors too.
-The three main components of the architecture were documented in detail,
+the advantages that being easily extensible brings to parser authors.  The
+three main components of the architecture were documented in detail,
 explaining each component's responsibilities and the functionality it
@@
 how it can be useful for general parsing.  The rules section was the
-longest, because although the rules appear conceptually simple ---
+longest in this chapter, because although the rules appear to be simple ---
 recognise an input and specify the action that handles that input --- they
 have subtle behaviour that needs to be clearly explained.  When extending a
-ruleset a decision needs to be taken about whether the input should be
+ruleset, a decision needs to be taken about whether the input should be
 recognised by extending an existing rule, by adding a new rule to an
-existing rule category, or by adding a new rule category, rule, and action.
-Rules can have conditions attached to them, restricting the set of rules
-used to recognise an input; the complexity of the conditions used greatly
-influences the difficulty of writing a correct ruleset or understanding and
-extending an existing ruleset.  Overlapping rules are frequently a
-requirement in a parser, and their use can greatly simplify some rules, but
-they can be a source of bugs because they can recognise inputs
-unexpectedly.  The framework does not try to detect overlapping rules,
-because overlap amongst rules may be valid and is quite often intentional;
-that responsibility falls to the author of the rules.  The difficulty
-involved in detecting overlap is proportional to the complexity of a
-ruleset's regexes and conditions, and may be possible for a human yet
-intractable or impossible for a program.  The rules section finishes with a
-discussion of pathological rules, concentrating on pathological regexes.
+existing rule category, or by adding a new input category, rule, and
+action.  Rules can have conditions attached to them, restricting the set of
+rules used to recognise an input; the complexity of the conditions used
+greatly influences the difficulty of writing a correct ruleset or
+understanding and extending an existing ruleset.  Overlapping rules are
+frequently a requirement in a parser, and their use can greatly simplify
+some rules, but they can be a source of bugs because they can recognise
+inputs unexpectedly.  The framework does not try to detect overlapping
+rules, because overlap amongst rules may be valid and is quite often
+intentional; that responsibility falls to the author of the rules.  The
+difficulty involved in detecting overlap is proportional to the complexity
+of a ruleset's regexes and conditions, and may be possible for a human yet
+intractable or impossible for a program.  The rules section concludes with
+a discussion of pathological rules, concentrating on pathological regexes.

### 2009-04-20 doc/logparser-background.tex
Corrections from proofreading for chapters 4 and 5.

@@
 client IP address, client hostname, HELO hostname, sender address, or
-recipient address.  E.g.\ the administrator may require that all clients on
-the local network have valid DNS entries, to prevent people sending mail
-from unknown machines; one example of how to achieve this is:
+recipient address.  For example, the administrator may require that all
+clients on the local network have valid DNS entries, to prevent people
+sending mail from unknown machines; one example of how to achieve this is:
 
@@
 mail on the third Tuesday after pay day only, to help prevent problems from
-phishing mails using faked sender addresses.  E.g.\ a phishing mail might
-claim that the payroll system had a disastrous disk failure, and until the
-server is replaced all salary payments will have to be processed manually,
-so please reply to this mail with your name, address, and bank account
-details; the criminal can then use any details sent to him to help with
-identity theft.
+phishing mails using faked sender addresses.  For example, a phishing mail
+might claim that the payroll system had a disastrous disk failure, and
+until the server is replaced all salary payments will have to be processed
+manually, so please reply to this mail with your name, address, and bank
+account details; the criminal can then use any details sent to him to help
+with identity theft.
 

### 2009-04-20 doc/logparser-complications.tex
Corrections from proofreading for chapters 4 and 5.

@@
 had found Postfix log files useful when investigating problems reported by
-users, and an examination of the log files gave the impression that they
-would be straightforward to parse, process, and understand.  The large
+users, and an examination of several log files gave the impression that
+they would be straightforward to parse, process, and understand.  The large
 variation in log lines was not apparent, because most log lines are
@@
 
-    \item [Recreating Postfix behaviour] Three complications are caused by
-        \parsername{} accurately recreating Postfix's behaviour.
+    \item [Recreating Postfix behaviour] Three complications needed to be
+        solved for \parsername{} to accurately recreate Postfix's
+        behaviour.
 
     \item [Lack of information in the log files] Seven complications are
-        caused by lack of information in the log files.  In general some
+        caused by lack of information in the log files.  In general, some
         extra logging by Postfix could remove these complications, and for
@@
 
-    \item [Order log lines are found in log files] Five complications are
-        caused by out of order log lines.  Postfix is not to blame for
-        these, they are caused by process scheduling, inter-process
-        communication delays, and are more likely to occur when the mail
-        server is heavily loaded.
+    \item [The order that log lines are found in log files] Five
+        complications are caused by out of order log lines.  Postfix is not
+        to blame for these: they are caused by process scheduling,
+        inter-process communication delays, and are more likely to occur
+        when the mail server is heavily loaded.
 
@@
 
+    \squeezeitems{}
+
     \item Two versions of several functions: \texttt{by\_pid} and
@@
         functions fall into two groups: those that deal with state tables,
-        and so need to know which one to operate on; the other group are
-        passed a specific mail or connection to work with, and do not need
-        to know about state tables at all.
+        and so need to know which one to operate on; the functions in the
+        other group are passed a specific mail or connection to work with,
+        and do not need to know about state tables at all.
 
@@
 accepted, so that subsequent mails will not overwrite each other's data.
-This must be done every time a mail is accepted, as it is impossible to
-tell in advance which connections will accept multiple mails.  Once a mail
-has been accepted its log lines will not be correlated by \acronym{pid} any
-more, its queueid will be used instead, as described in \sectionref{queueid
-vs pid}.  If the original connection has any useful data (e.g.\ rejections)
-it will be saved to the database when the client disconnects.  One unsolved
-difficulty is distinguishing between different groups of rejections, e.g.\
-when dealing with the following sequence:
+This must be done every time a mail is accepted, because it is impossible
+to tell in advance which connections will accept multiple mails.  Once a
+mail has been accepted its log lines will not be correlated by
+\acronym{pid} any more: its queueid will be used instead, as described in
+\sectionref{queueid vs pid}.  If the original connection has any useful
+data (e.g.\ rejections) it will be saved to the database when the client
+disconnects.  One unsolved difficulty is distinguishing between different
+groups of rejections, e.g.\ when dealing with the following sequence:
 
 \begin{enumerate}
+
+    \squeezeitems{}
 
@@
 
-There should ideally be two separate entries in the database resulting from
-the above sequence, but there will only be one.
+Ideally, the sequence above would result in two separate entries in the
+connections table, but only one will be created.
 
@@
 mailbox: sometimes they are sent to and accepted for a local address, but
-need to be delivered to one or more remote addresses due to aliasing.  When
-this occurs, a child mail will be injected into the Postfix queue, but
+need to be delivered to one or more remote addresses because of aliasing.
+When this occurs, a child mail will be injected into the Postfix queue, but
 without the explicit logging that mails injected by \daemon{smtpd} or
@@
 from a strictly chronological reading of the log lines it usually appears
-as if the child mail has been created without a source.  Subsequently the
+as if the child mail has been created without a source.  Subsequently, the
 parent mail will log the creation of the child mail, e.g.\ parent mail
@@
 
+The sample log line above would be processed by the \action{TRACK} action,
+which creates the child mail if it does not exist in the state tables,
+links it to the parent mail, and checks that the child is not being tracked
+for a second time.
+
 Unfortunately, although all log lines from an individual process appear in
 chronological order, the order in which log lines from different processes
-are interleaved is subject to the vagaries of process scheduling.  In
-addition, the first log line belonging to the child mail (the example log
-line above belongs to the parent mail) is logged by either \daemon{qmgr} or
-\daemon{cleanup}, so the order also depends on how soon they process the
-new mail.
+are interleaved in a log file is subject to the vagaries of process
+scheduling.  In addition, the first log line belonging to the child mail
+(the example log line above belongs to the parent mail) is logged by either
+\daemon{qmgr} or \daemon{cleanup}, so the order also depends on how soon
+they process the new mail.
 
@@
 cannot complain when it encounters a log line from either \daemon{qmgr} or
-\daemon{cleanup} for a previously unseen mail; instead it must flag the
-mail as coming from an unknown origin, and subsequently clear that flag if
-the origin of the mail becomes clear.  Obviously the parser could omit
+\daemon{cleanup} for a mail that does not exist in the state tables;
+instead it must create the state table entry and flag the mail as coming
+from an unknown origin.  Subsequently invoked actions will clear that flag
+if the origin of the mail becomes clear.  Obviously the parser could omit
 checking where mails originate from, but requiring an explicit source helps
@@
 
-Process scheduling can have a still more confusing effect: quite often the
-child mail will be created, delivered, and entirely finished with,
+Process scheduling can have a still more confusing effect: the child mail
+will often be created, delivered, and entirely finished with,
 \textit{before\/} the parent mail logs its creation!  Thus, mails flagged
@@
 instead they must be marked as ready for entry and subsequently entered
-once their origin has been identified.  The crux of this complication is
-that re-injected mails appear in the log files without explicit logging
-indicating their source.  Unfortunately, it is not possible to distinguish
-between child mails waiting to be tracked and other mails whose origin is
-unknown (except for bounce notifications, as described in
+once their origin has been identified.  Unfortunately, it is not possible
+to distinguish between child mails waiting to be tracked and other mails
+whose origin is unknown (except for bounce notifications, as described in
 \sectionref{identifying bounce notifications}); postponing the entry into
@@
         mail is tagged as ready to be entered in the database, but entry is
-        postponed until the parent is identified.  The child mail should
-        not have any subsequent log lines: only its parent will refer to
-        it.
+        postponed until the parent is identified.  The child mail will not
+        have any subsequent log lines: only its parent will refer to it.
 
@@
 Postfix 2.2.x (and presumably previous versions) does not generate a log
-line when it generates a bounce notification; suddenly there will be log
-entries for a mail that lacks an obvious source.  Similarities exist to the
+line when it generates a bounce notification; the lof file will have log
+lines for a mail that lacks an obvious source.  Similarities exist to the
 problem of identifying re-injected mails discussed in \sectionref{tracking
@@
     \item Neither \daemon{smtpd} nor \daemon{pickup} have logged any
-        messages associated with the mail, indicating it was generated
-        internally by Postfix, not accepted via \acronym{SMTP} or submitted
-        locally by \daemon{postdrop}.
+        messages associated with the mail, indicating that it was generated
+        internally by Postfix, rather than accepted via \acronym{SMTP} or
+        submitted locally by \daemon{postdrop}.
 
@@
 
-    \item The queueid in the message-id must be the same as the queueid of
-        the mail: this is what distinguishes a new bounce notification from
-        a bounce notification that is being re-injected as a result of
-        aliasing.  For the latter, the message-id will be unchanged from
-        the original bounce notification, and so even if it happens to be
-        in the correct format, e.g.\ if it was generated by Postfix on this
-        or another server, it will not correspond with the queueid of the
-        mail.
+    \item The queueid embedded in the message-id must be the same as the
+        queueid of the mail: this is what distinguishes a new bounce
+        notification from a bounce notification that is being re-injected
+        as a result of aliasing.  For the latter, the message-id will be
+        unchanged from the original bounce notification, and so even if it
+        happens to be in the correct format, i.e.\ if it was generated by
+        Postfix on this or another server, it will not equal the queueid of
+        the mail.
 
@@
 \begin{enumerate}
+
+    \squeezeitems{}
 
@@
 have been generated internally by Postfix; arguably misclassification of
-this kind is a benefit rather than a drawback, as other mails generated
-internally by Postfix will be handled correctly.  Postfix 2.3 and
+this kind is a benefit rather than a drawback, because other mails
+generated internally by Postfix will be handled correctly.  Postfix 2.3 and
 subsequent versions log the creation of a bounce message.
@@
 saved in the connection's state table entry, and the accepted mails are
-examined when the disconnection action is invoked.  Each mail is checked
-for the following:
+examined when the disconnection action is invoked.  Each accepted mail is
+checked for the following:
 
@@
 If all the checks above are successful then the mail is assumed to be an
-aborted delivery attempt and is discarded.  There will be no further
-entries logged for such mails, so without identifying and discarding them
-they accumulate in the state table and will cause clashes if the queueid is
-reused.  The mail cannot be entered in the database as the only data
-available is the client hostname and IP address, but the database schema
-requires many more fields be populated (see \sectionref{connections table}
-and \sectionref{results table}).  These heuristics are quite restrictive,
-and it appears to have little scope for false positives; any false
-positives will cause a warning when the next log line for that mail is
-parsed.  False negatives are less likely to be detected: there may be
-queueid clashes (and warnings) if mails remain in the state tables after
-they should have been removed, otherwise the only way to detect false
-negatives is to examine the state tables after each parsing run.
+aborted delivery attempt and is removed form the state tables.  There will
+be no further entries logged for such mails, so without identifying and
+discarding them they accumulate in the state table and will cause clashes
+if the queueid is reused.  The mail cannot be entered in the database
+because the only data available is the client hostname and IP address, but
+the database schema requires many more fields be populated (see
+\sectionref{connections table} and \sectionref{results table}).  These
+heuristics are quite restrictive, and appear to have little scope for false
+positives; any false positives would cause a warning when the next log line
+for that mail is parsed.  False negatives are less likely to be detected:
+there may be queueid clashes (and warnings) if mails remain in the state
+tables after they should have been removed, otherwise the only way to
+detect false negatives is to examine the state tables after each parsing
+run.
 
@@
 If both checks are successful then the mail is assumed to have been
-discarded when the client disconnected.  There will be no further log lines
-for such mails, so without identifying and entering them in the database
-immediately they accumulate in the state table and will cause clashes if
-the queueid is reused.
+discarded by Postfix when the client disconnected; \parsername{} will
+remove it from the state tables.  There will be no further log lines for
+such mails, so if \parsername{} does not deal with them immediately they
+accumulate in the state table and will cause clashes if the queueid is
+reused.
 
@@
         allocated a queueid for the incoming mail, and there may be a mail
-        in the state tables that needs to be discarded.  The timeout may
+        in the state tables that needs to be discarded by \parsername{}.  The timeout may
         have interrupted transfer of an accepted delivery attempt, or it
@@
 
-        If all the checks are passed the mail will be discarded from the
-        state table, and will not be entered in the database.  If one or
-        more checks are not passed it is assumed that the timeout happened
-        after a rejected delivery attempt.  This assumption is not
-        necessarily correct, because Postfix may have accepted an earlier
-        recipient and rejected a later one, so the timeout applies to the
-        partially accepted mail, which should be discarded; this has not
-        been a problem in practice.  Processing timeouts is further
+        If all the checks pass the mail will be discarded from the state
+        tables, and will not be entered in the database.  If one or more
+        checks do not pass it is assumed that the timeout happened after a
+        rejected delivery attempt.  This assumption is not necessarily
+        correct, because Postfix could have accepted an earlier recipient,
+        rejected a later one, and continued to accept delivery of the mail
+        for the first recipient.  Tn that case the timeout applies to the
+        partially accepted mail, which should be discarded; however, this
+        has not occurred in practice.  Processing timeouts is further
         complicated by the presence of out of order \daemon{cleanup} log
@@
 This complication is dealt with by the \action{TIMEOUT} action, with help
-from the \action{CLONE} action.
+from the \action{CLONE} action.  When a client tries to delivery a mail
+larger than the sever will accept, the \action{MAIL\_TOO\_LARGE} action
+will perform the same processing as the \action{TIMEOUT} action.
 
@@
 still pending, the action updates the \texttt{timeout\_queueids} state
-table, adding the queueid and the timestamp from the log line.  When the
-next \daemon{cleanup} log line is parsed for that queueid, the list will be
-checked and if the log line meets the following requirements it will be
-deemed part of the mail where the timeout occurred and discarded.
+table, adding the queueid and the timestamp from the log line.  The
+following checks will be performed for each \daemon{cleanup} log line
+that is processed:
 
@@
 
-    \item The queueid must not have been reused yet, i.e.\ it does not have
-        a mail in the state tables.
+    \item Does the \texttt{timeout\_queueids} state table have an entry for
+        the queueid in the log line?  If there is an entry, it will be
+        removed, regardless of whether the remaining criteria are
+        satisfied.
+
+    \item Has the queueid been reused yet, i.e.\ does it have an entry in
+        the state tables?
 
@@
 \daemon{cleanup} log line after a timeout it would sometimes be mistaken.
-When the next \daemon{pickup} log line containing that queueid is processed
-the queueid will be removed from the cache of timeout queueids, regardless
-of whether it meets the criteria above.
 
@@
 
-When mail is submitted locally, \daemon{pickup} accepts the new mail and
+When mail is submitted locally, \daemon{pickup} processes the new mail and
 generates a log line.  Occasionally this log line will occur later in the
@@
 
-Occasionally a \daemon{smtpd} will just stop logging, without an
-immediately obvious reason.  After poring over log files for some time
-several reasons have been found for this rare event:
+Occasionally a \daemon{smtpd} will stop logging without an immediately
+obvious reason.  After poring over log files for some time, several reasons
+have been found for this rare event:
 
@@
 
-    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by Postfix
-        for some reason, by the administrator, or by the OS), so its active
+    \item Sometimes a \daemon{smtpd} is killed by a signal (sent by
+        Postfix, by the administrator, or by the OS), so its active
         connection must be cleaned up, entered in the database if it has
@@
 logging after cleanup}.  In the \numberOFlogFILES{} log files used for
-testing this problem occurs only five times in 60,721,709 log lines, but
-for parser correctness it must be dealt with.  All five occurrences have
-the same characteristics: the \daemon{local} log line showing delivery to a
-local mailbox comes after the \daemon{qmgr} log line showing removal of the
-mail from the queue because delivery is complete.  This causes problems:
-the data in the state tables for the mail is not complete, so it cannot be
-entered into the database; a new mail is created when the \daemon{local}
-log line is processed and remains in the state tables; four warnings are
-issued per pair of out of order log lines.
+testing this problem occurs only five times in 60,721,709 log lines.  All
+five occurrences have the same characteristics: the \daemon{local} log line
+showing delivery to a local mailbox comes after the \daemon{qmgr} log line
+showing removal of the mail from the queue because delivery is complete.
+This causes problems: the data in the state tables for the mail is not
+complete, so it cannot be entered into the database; a new mail is created
+when the \daemon{local} log line is processed and remains in the state
+tables; four warnings are issued per pair of out of order log lines.
 
@@
 Postfix component combinations.  If the mail's combination is found in the
-table the mail can be entered in the database; if the combination is not
-found entry must be postponed and the mail flagged for later entry.  The
+table it can be entered in the database; if the combination is not found
+entry must be postponed and the mail flagged for later entry.  The
 \action{MAIL\_DELIVERED} action checks for that flag; if the additional log
-lines have caused the mail to reach a valid combination then entry in the
-database will proceed, otherwise it must be postponed once more.
+lines have caused the mail to reach a valid combination then its entry in
+the database will proceed, otherwise it must be postponed once more.
 
-The list of valid combinations is explained below.  Every mail will
-additionally have log lines from \daemon{cleanup} and \daemon{qmgr}; any
-mail may also have a log line from \daemon{bounce}, \daemon{postsuper}, or
-both.
+The valid combinations are explained below.  In addition to the components
+shown in each combination, every mail will have log lines from both
+\daemon{cleanup} and \daemon{qmgr}, and any mail may also have a log line
+from \daemon{bounce}, \daemon{postsuper}, or both.
 
@@
     \item [\daemon{smtp}, \daemon{smtpd}:] Mail accepted from a remote
-        client, then remotely delivered (typically relaying mail for
-        clients on the local network to addresses outside the local
-        network).
+        client, then remotely delivered.  Typically this is a mail server
+        relaying mail for clients on the local network to addresses outside
+        the local network.
 
@@
         client, then deleted by the administrator before any delivery
-        attempt was made.  Notice that \daemon{postsuper} is required, not
+        attempt was made.  Note that \daemon{postsuper} is required, not
         optional, for this combination.
@@
 symptoms are the same as in \sectionref{aborted delivery attempts}, except
-that the mail has a \daemon{cleanup} log line; nothing can be found in the
-log file to explain why this mail does not have any further log lines.  The
-only way to detect these mails is to periodically scan all mails in the
-state tables, deleting any mails with the following characteristics:
+that the mail \textit{has\/} a \daemon{cleanup} log line; nothing can be
+found in the log files to explain why this mail does not have any further
+log lines.  The only way to detect these mails is to periodically scan all
+mails in the state tables, deleting any mails with the following
+characteristics:
 
@@
 faster than Postfix can attempt delivery, and the administrator deletes
-some of the mail before Postfix has had a chance to try to deliver it.  The
-recipient address will not have been logged, so a dummy recipient address
-needs to be added, because every mail is required by the database schema
-(\sectionref{results table}) to have at least one recipient.  Typically
-when this complication occurs the log file will show many instances of it,
-closely grouped.
+some of the mail before Postfix has had a chance to attempt delivery.  The
+recipient address will not have been logged yet, and the deleted mails will
+not have any more log lines for it to be logged in.  A dummy recipient
+address needs to be added, because every mail is required by the database
+schema (\sectionref{results table}) to have at least one recipient.
+Typically when this complication occurs the log file will show many
+instances of it, closely grouped.
 
@@
 is logged, so the warning will be associated with the connection rather
-than the accepted mail.  A queue file and queueid will be allocated after
-Postfix accepts the MAIL FROM command if
-\texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is set to ``no'', but that
-setting will cause disk IO for every delivery attempt, instead of just for
-delivery attempts where recipients are accepted, and consequently a drastic
-reduction in the performance of the mail server.
+than the accepted mail.  A queueid will be allocated after Postfix accepts
+the MAIL FROM command if \texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is
+set to ``no'', but that setting will cause disk IO for every delivery
+attempt, instead of just for delivery attempts where recipients are
+accepted, and consequently a drastic reduction in the performance of the
+mail server.
 
@@
 results in the log lines from the delivery agent (\daemon{local},
-\daemon{virtual} or \daemon{smtp}) appearing in the log file
+\daemon{smtp}, or \daemon{virtual}) appearing in the log file
 \textit{after\/} the mail has been removed from the state tables and saved
@@
 heavy system load: these complications typically caused warnings and a new
-state table entry.  Three of the fifteen complications are required to
-correctly recreate Postfix's behaviour.  The remaining seven complications
-are caused by deficiencies in Postfix's logging, and some could easily be
-resolved by adding additional logging to Postfix; these seven are dealt
-with by applying heuristics to specific log lines or mails, sometimes in
-conjunction with information cached when mails are removed from the state
-tables.
+state table entry.  Three of the fifteen complications must be dealt with
+to correctly recreate Postfix's behaviour.  The remaining seven
+complications are caused by deficiencies in Postfix's logging, and some
+could easily be resolved by adding additional logging to Postfix; these
+seven are dealt with by applying heuristics to specific log lines or mails,
+sometimes in conjunction with information cached when mails are removed
+from the state tables.

### 2009-04-20 doc/logparser-implementation.tex
Corrections from proofreading for chapters 4 and 5.

@@
 
-This chapter documents the implementation of \parsername{}, based on the
-architecture described in \sectionref{parser architecture}.  Any design may
-look good when examined in the abstract, but the real test of the design
-comes with the first concrete implementation; only then do any practical
-difficulties come to light.  Implementing this architecture was relatively
-straightforward: the difficulties came not from the architecture and
-design, but from anomalies in the log files and Postfix's behaviour, as
+This chapter documents the implementation of \parsername{}, a parsed based
+on the architecture described in \sectionref{parser architecture}.  Any
+design may look good when examined in the abstract, but the real test of
+the design comes with the first concrete implementation; only then do any
+practical difficulties come to light.  Implementing this architecture was
+relatively straightforward: the difficulties came not from the architecture
+and design, but from anomalies in the log files and Postfix's behaviour, as
 described in \sectionref{complications}.  \parsername{} successfully deals
@@
 all potential difficulties that may arise is an intractable task, but the
-descriptions of the solutions developed thus far (see
-\sectionref{complications}) should help if someone needs to develop a
-solution of their own.
-
-\parsername{} deals with all the complications of parsing, the
-eccentricities and oddities of Postfix log files, presenting the resulting
-data in a normalised, simple to use representation.  Unfortunately dealing
-with the complications that arise sometimes requires the parser to discard
-log lines (e.g.\ \sectionref{aborted delivery attempts}), and, less
-frequently, to discard a data structure (e.g.\ \sectionref{timeouts during
-data phase}).  \parsername{} can parse log files from Postfix version 2.2
-through to version 2.5, and should be capable of parsing log files from
-later versions with only minor modifications or an updated ruleset.  The
-parser's task is to follow the journey each mail takes through Postfix,
-combining the data captured by rules into a coherent whole, saving it in a
-useful and consistent form, and performing housekeeping duties.  The
-intermingling of log lines from different mails immediately rules out the
-possibility of handling each mail in isolation; the parser must handle
-multiple mails in parallel, each potentially at a different stage in its
-journey, without any interference between mails --- except in the minority
-of cases where intra-mail interference is required, e.g.\ mail re-injected
-for forwarding (\sectionref{Re-injected mails}).  The best way to deal with
-intermingling of log lines is to maintain state information for every
-unfinished mail, and to manipulate the appropriate mail correctly for each
-log line encountered.
+descriptions of the solutions developed thus far should help if someone
+needs to develop a solution of their own.
+
+\parsername{} deals with all the eccentricities and oddities of parsing
+Postfix log files, presenting the resulting data in a normalised, simple to
+use representation.  Unfortunately, dealing with the complications that
+arise sometimes requires the parser to discard log lines (e.g.\
+\sectionref{aborted delivery attempts}), and, less frequently, to discard a
+data structure (e.g.\ \sectionref{timeouts during data phase}).
+\parsername{} can parse log files from Postfix version 2.2 through to
+version 2.5, and should be capable of parsing log files from later versions
+with only minor modifications or an updated ruleset.  The parser's task is
+to follow the journey each mail takes through Postfix, combining the data
+captured by rules into a coherent whole, saving it in a useful and
+consistent form, and performing housekeeping duties.  The intermingling of
+log lines from different mails immediately rules out the possibility of
+handling each mail in isolation; the parser must handle multiple mails in
+parallel, each potentially at a different stage in its journey, without any
+interference between mails --- except in the minority of cases where
+intra-mail interference is required, e.g.\ mail re-injected for forwarding
+(\sectionref{Re-injected mails}).  The best way to deal with intermingling
+of log lines is to maintain state information for every unfinished mail,
+and to manipulate the appropriate mail correctly for each log line
+encountered.
 
@@
 taken through Postfix and \parsername{}, with a description of the stages
-and transitions.  The topic of the next section is the database that
+and stage transitions.  The second topic is the \acronym{SQL} database that
 provides storage for the parser: any future work that analyses gathered
@@
 of the architecture.  First is the framework, including its initialisation
-phase, the parsing process, and the conveniences it offers to users; that
-section finishes with the features that are important for this thesis but
-not strictly necessary for parsing: the performance data collected by the
-framework, the optimisations that can be disabled to show their effect, and
-the debugging options it provides.  The second component of the
-architecture is the actions: their implementation is documented, with a
-graph of how often each action is specified by rules, and a description of
-why some actions are more popular than others.  Every action that is part
-of \parsername{} is documented, and the actions section concludes with the
-process of adding a new action, and how to determine the values for each of
-the required fields.  Rules are the final component of the architecture,
-and also the most visible to advanced users, e.g.\ systems administrators,
-because it is likely that they will need to add their own rules to
-recognise their log lines.  An example rule used by \parsername{} is shown,
-with every field clearly documented; this is followed by a description of
-adding new rules, and how to determine the value of each of the fields.
-\parsername{} provides a utility to create regexes from unrecognised log
-lines; the algorithm it uses is documented, including the differences
-between it and the original algorithm.  The rules section finishes with a
-discussion of how \parsername{} uses rule conditions and overlapping rules,
-plus a description of the regex snippets provided to ease the process of
-writing regexes.
+phase, the parsing process, and the conveniences it offers to users of
+\parsername{}; that section finishes with the features that are important
+for this thesis but not strictly necessary for parsing: the performance
+data collected by the framework, the optimisations that can be disabled to
+show their effect, and the debugging options it provides.  The second
+component of the architecture is the actions, starting with a graph of how
+often each action is specified by rules, a description of why some actions
+are more popular than others, and how this popularity has no influence on
+how often an action is invoked.  Every action that is part of \parsername{}
+is documented, and the actions section concludes with a description of the
+process of adding a new action.  Rules are the final component of the
+architecture, and also the most visible to advanced users, e.g.\ systems
+administrators, because it is likely that they will need to add their own
+rules to recognise their log lines.  An example rule used by \parsername{}
+is examined, with every field clearly documented; that is followed by a
+description of adding new rules, and how to determine the value of each of
+the fields.  \parsername{} provides a utility to create regexes from
+unrecognised log lines; the algorithm it uses is documented, including the
+differences between it and the original algorithm it is based on.  The
+rules section finishes with a discussion of how \parsername{} uses rule
+conditions and overlapping rules, plus a description of the regex snippets
+provided to ease the process of writing regexes.
 
@@
         awry, file system filling up, logging system unable to cope with
-        the volume of log messages).  On a well run system it is extremely
-        unlikely that any of these problems will arise, though it is of
-        course possible, particularly when suffering from a deluge of spam
-        or a mail loop.  When parsing individual log files in isolation,
-        it is highly likely that some mails will have log lines in previous
-        log files, and others will have log lines in subsequent log files;
-        to alleviate this problem \parsername{} supports saving and loading
+        the volume of log messages).  On a well run mail server it is
+        extremely unlikely that any of these problems will arise, though
+        the likelihood increases when suffering from a deluge of spam or a
+        mail loop.  When parsing individual log files in isolation, it is
+        highly likely that some mails will have log lines in previous log
+        files, and others will have log lines in subsequent log files; to
+        alleviate this problem \parsername{} supports saving and loading
         its state tables, so they can be preserved between log files.
@@
 analysis of the log files has been enough to reconstruct Postfix's
-behaviour, but for more difficult programs the techniques described in
+behaviour, but for other programs the techniques described in
 \cite{black-box-error-reporting} may be useful.  Some advantages come from
@@
     \item The parser acts as a second source of information about Postfix's
-        operation, based on empirical evidence.  A separate project could
-        compare the empirical knowledge inherent in \parsername{} with
-        Postfix's documentation and source code to see how closely the two
-        agree.
+        operation, based on empirical evidence from log files.  A separate
+        project could compare the empirical knowledge inherent in
+        \parsername{} with Postfix's documentation and source code to see
+        how closely the two agree.
 
@@
 implementation}) and complications (\sectionref{complications}).  The
-states and transitions shown in \figureref{flow chart image} will be
-explained later in this section.
+states and state transitions shown in the flow chart will be explained
+later in this section.
 
@@
 
-    \item This is followed by rejection of \acronym{SMTP} commands by the
-        restrictions configured by the administrator
+    \item This is followed by rejection of one or more mail delivery attempts
         \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
@@
         normally or with an error.  If Postfix has rejected any
-        \acronym{SMTP} commands the resulting data will be saved to the
-        database; if there were no rejections there will not be any data to
-        save.  Any mails accepted will already have been cloned so their
-        data is in another data structure, and will be delivered in the
-        same way as mails that entered the system by any other route.
+        \acronym{SMTP} commands the data gathered from those rejections
+        will be saved to the database; if there were no rejections there
+        will not be any data to save.  Any mails accepted will already have
+        been cloned so their data is in another data structure, and will be
+        delivered in the same way as mails that entered the system by any
+        other route.
 
@@
 The obvious counterpart to mail entering the system is mail leaving the
-system, whether by deletion, bouncing, local delivery, or remote delivery.
-All four are handled in exactly the same way:
+system, whether by deletion, bouncing, expiry, local delivery, or remote
+delivery.  All five are handled in the same way:
 
@@
 
-    \item Sometimes mail is re-injected and the child mail needs to be
-        tracked with the parent mail \flowchart{TRACK}{9}; the handling of
-        re-injected mails is described in \sectionref{tracking re-injected
-        mail}.
+    \item Sometimes mail is re-injected for forwarding and the child mail
+        needs to be tracked with the parent mail \flowchart{TRACK}{9}; the
+        handling of re-injected mails is described in \sectionref{tracking
+        re-injected mail}.
 
@@
 
-    \item This is the last log line for this particular mail, though it may
-        be indirectly referred to if it was re-injected.  The mail is
-        cleaned up and entered in the database, then deleted from the state
-        tables \flowchart{COMMIT}{11}.
+    \item The mail is removed from the Postfix queue.  This is the last log
+        line for this particular mail, though it may be indirectly referred
+        to if it was re-injected.  The mail is cleaned up and entered in
+        the database, then deleted from the state tables
+        \flowchart{COMMIT}{11}.
 
@@
 
-It should be emphasised that the mail delivery process above happens
-whether the mail is delivered to a mailbox, piped to a command, delivered
-to a remote server, bounced (because of a mail loop, delivery failure, or
-five day timeout), or deleted by the administrator, \textit{unless\/} the
-mail is either parent or child of re-injection, as explained in
-\sectionref{tracking re-injected mail}.
+It should be emphasised that the sequence above happens whether the mail is
+delivered to a mailbox, piped to a command, delivered to a remote server,
+bounced (because of a mail loop, delivery failure, or five day timeout), or
+deleted by the administrator, \textit{unless\/} the mail is either parent
+or child of re-injection, as explained in \sectionref{tracking re-injected
+mail}.
 
@@
 understanding the actions of the parser, and essential to developing
-further applications that utilise the data gathered; \sectionref{database
-as API} describes how the database schema functions as an \acronym{API}.
+further applications that utilise the data; \sectionref{database as API}
+describes how the database schema functions as an \acronym{API}.
 
@@
 
-The data saved from parsing the log files is also divided into two tables:
+The data saved from parsing the log files is divided into two tables:
 connections and results.  The connections table contains an entry for every
-mail accepted and every connection that rejected an \acronym{SMTP} command;
-the individual fields will be described in \sectionref{connections table}.
-The results table will have at least one entry for each entry in the
+mail accepted and every connection that rejected a delivery attempt; the
+individual fields will be described in \sectionref{connections table}.  The
+results table will have at least one entry for each entry in the
 connections table; its fields will be covered in detail in
@@
 available for inspection because each result references the rule that
-created it.  No ambiguity or confusion is possible about which rule
-resulted in a particular result, eliminating one potential source of
-confusion.
+created it.  No ambiguity is possible about which rule resulted in a
+particular result, eliminating one potential source of confusion.
 
@@
 from log lines; it is more important when using the data than when storing
-it, because storing the data is a write-once operation, whereas utilising
+it, because storing the data is a once-off operation, whereas utilising
 the data requires frequent searching, sorting, and manipulation of the data
@@
 Postfix log files.  Although the interface is a database schema rather than
-a set of functions in a library, it gives the same benefits as any other
-\acronym{API}: it provides a stable interface between the user and the
-creator of the data, allowing code on either side of the interface to be
-changed without adverse effects, as long as the interface is adhered to.
-Programs that use the database can range from the simple examples in
-\sectionref{motivation} to far more complex data mining tools.
+a set of functions in a library, it provides the same benefits as any other
+\acronym{API}: a stable interface between the user and the creator of the
+data, allowing code on either side of the interface to be changed without
+adverse effects, as long as the interface is adhered to.  Programs that use
+the database can range from the simple examples in \sectionref{motivation}
+to far more complex data mining tools and machine learning algorithms.
 
@@
         the user without requiring large amounts of programming.  All
-        databases provide a program, of varying complexity and
+        databases have one or more programs, of varying complexity and
         sophistication, that can be used for ad hoc queries with minimal
@@
     WHERE sender = "boss@example.com"
-        AND action = "DELIVERY_REJECTED"
+        AND rule_id = rules.id
+        AND rules.action = "DELIVERY_REJECTED"
     SEND PANIC EMAIL TO "postmaster@example.com";
@@
             \item Other tables can be added to the database, e.g.\ to cache
-                historical or computed data, or to link in data from other
-                sources.
+                historical or computed data, or to incorporate data from
+                other sources.
 
@@
 Storing the results in a database will also increase the efficiency of
-using those results, because the log files need only be parsed once rather
-than each time the data is used; indeed the results may be used by someone
-with no access to the original log files.
+using those results, because the log files only need to be parsed once
+rather than each time the data is used; indeed the database may be used by
+someone with no access to the original log files.
 
@@
 Rules are discussed in detail in \sectionref{rules in implementation}, but
-the structure of the rules table is documented here along with the other
-tables in the database.  Rules are created by the user, not the parser, and
-will not be modified by \parsername{} (except for the hits and hits\_total
-fields).  Rules classify the individual log lines, capturing data to be
+the structure of the rules table is documented here alongside the other
+tables in the database.  Rules are created by the user, and will not be
+modified by \parsername{}, except when it updates the hits and hits\_total
+fields.  Rules recognise the individual log lines, capturing data to be
 saved in the connections and results tables, and specifying the action to
-take for each recognised log line.
-
-Each rule some or all of these fields; all fields are required unless
-otherwise stated.
+invoke for each recognised log line.
+
+Each rule must provide values for most of the fields below; all fields are
+required unless otherwise stated in the description.
 
@@
 
-    \item [id] A unique identifier for each rule that other tables can use
-        when referring to a specific rule.
+    \item [id] A unique identifier that other tables can use when referring
+        to a specific rule.
 
@@
 
-    \item [description] Something must have occurred to cause Postfix to
-        output each log line; this field should describe the event causing
-        the log lines this rule recognises.
+    \item [description] This field should describe the event causing the
+        log lines this rule recognises, e.g.\ ``Mail has been delivered to
+        the LDA (typically procmail)''.
 
     \item [restriction\_name] The name of the Postfix restriction that
-        caused the mail to be rejected.  This field is valid only for rules
-        that recognise rejection log lines, i.e.\ rules that have an action
-        of \action{DELIVERY\_REJECTED}.
+        caused the rejection of the mail delivery attempt.  This field is
+        valid only for rules that recognise rejection log lines, i.e.\
+        rules that have an action of \action{DELIVERY\_REJECTED}.
 
     \item [program] The Postfix component (e.g.\ \daemon{smtpd}) whose log
-        lines the rule applies to; see \sectionref{rule conditions in
+        lines the rule recognises; see \sectionref{rule conditions in
         implementation} for full details of how this attribute is used.
@@
     \item [connection\_data] Sometimes rules need to provide data that is
-        not present in the log line: e.g.\ setting \texttt{client\_ip} when
-        a mail is being delivered to another server.  Any field in the
+        not present in the log line, e.g.\ setting \texttt{client\_ip} when
+        a mail is being delivered to another server; any field in the
         connections table can be set in this way.  The format is:
@@
         and semi-colons cannot be escaped and thus cannot be included in
-        data, this feature is intended for use with small amounts of data,
-        so dealing with escape sequences was deemed unnecessary.  This
-        field is optional.
+        data, because this feature is intended for use with small amounts
+        of data and dealing with escape sequences was deemed unnecessary.
+        This field is optional.
 
@@
 
-    \item [action] The action that will be invoked when this rule
-        recognises a log line; a full list of actions and the parameters
-        they are invoked with can be found in \sectionref{actions in detail
-        in implementation}.
+    \item [action] The action to be invoked when this rule recognises a log
+        line; a full list of actions and the parameters they are invoked
+        with can be found in \sectionref{actions in detail in
+        implementation}.
 
@@
 
-    \item [hits\_total] The total number of hits for this rule over all
-        runs of the parser; hits is reset to zero each time the parser is
-        run, but hits\_total is not.  This field will be set by the parser
-        rather than the rule author.
+    \item [hits\_total] The total number of log lines recognised by this
+        rule over all runs of the parser; hits is reset to zero each time
+        the parser is run, but hits\_total is not.  This field will be set
+        by the parser rather than the rule author.
 
     \item [priority] This is the user-configurable companion to hits: when
-        the list of rules is sorted, priority overrides hits.  This allows
-        more specific rules to take precedence over more general rules, as
-        described in \sectionref{rules in architecture}.  This field is
-        optional.
+        the list of rules is sorted by the parser, priority overrides hits.
+        This allows more specific rules to take precedence over more
+        general rules, as described in \sectionref{rules in architecture}.
+        This field is optional.
 
@@
 
-Every accepted mail and every connection that rejected an \acronym{SMTP}
-command will have a single entry in the connections table containing all of
-the following fields:
+Every accepted mail and every connection that rejected a mail delivery
+attempt will have a single entry in the connections table containing all of
+the following below.  For an incoming connection, the client is the remote
+machine, and the server is the local machine; for outbound mail delivery
+attempts, the roles are reversed.
 
@@
 
-    \item [server\_ip] The IP address of the server: the local address when
-        receiving mail, the remote address when sending mail.
+    \item [server\_ip] The IP address of the server.
 
@@
 
-    \item [client\_ip] The client IP address: the remote address when
-        receiving mail, the local address when sending mail.
+    \item [client\_ip] The client IP address.
 
@@
         or virus senders think it is a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects an \acronym{SMTP} command,
-        but it is quite simple to rectify this as described in
+        logs the HELO hostname when it rejects a mail delivery attempt, but
+        it is quite simple to rectify this as described in
         \sectionref{logging helo}.
@@
 with a single connection; a single connection will have at least one result
-associated with it, but will usually have several more, and may have
-hundreds.
+associated with it, but will usually have several, and may have hundreds.
 
@@
     \item [rule\_id] The id of the rule in the rules table that recognised
-        the log line and created this result.
+        the log line.
 
@@
     \item [warning] Administrators can configure Postfix to log a warning
-        instead of enforcing a restriction that would reject an
-        \acronym{SMTP} command --- a mechanism that is quite useful for
-        testing new restrictions.  This field will be false for a real
-        rejection, or true if the log line was a warning.  This field
-        should be ignored if the result is not a rejection, i.e.\ the
-        action field of the associated rule is not
-        \action{DELIVERY\_REJECTED}.
+        instead of enforcing a restriction that would reject a mail
+        delivery attempt --- a mechanism that is quite useful for testing
+        new restrictions.  This field will be false for a real rejection,
+        or true if the log line was a warning.  This field should be
+        ignored if the result is not a rejection, i.e.\ the action field of
+        the associated rule is not \action{DELIVERY\_REJECTED}.
 
@@
         final delivery log lines do not contain an \acronym{SMTP} code,
-        e.g.\ when Postfix delivers to a command; in those cases the
+        e.g.\ when Postfix delivers to a user's mailbox; in those cases the
         \acronym{SMTP} code is specified by the rule's
@@
         connection if the client uses different sender addresses for
-        multiple rejected delivery attempts; however there should not be
+        multiple rejected delivery attempts; however there will not be
         different sender addresses in the results for one accepted mail.
-        Mails sent over the same connection can be distinguished by their
-        queueids.
 
@@
 
-    \item [size] The size of the mail, present for delivered mails only.
+    \item [size] The size of the mail, only available for delivered mails.
 
@@
     \item [data] A field available to store a piece of captured data that
-        does not have a specific field, e.g.\ the rejection message from a
-        \acronym{DNSBL}\@.
-
-    \item [timestamp] The time at which the log line was logged.
+        does not have its own specific field, e.g.\ the rejection message
+        from a \acronym{DNSBL}\@.
+
+    \item [timestamp] The log line's timestamp.
 
@@
 Each time the parser is run the framework performs some initialisation
-tasks, setting up data structures that will later be used by actions,
-including several state tables.  Data about the connections and mails being
-processed is held in \texttt{connections} and \texttt{queueids}
-respectively.  The other state tables are used when working around
-complications in parsing: \texttt{timeout\_queueids} is used when dealing
-with connections that time out during the DATA phase (\sectionref{timeouts
-during data phase}); \texttt{bounce\_queueids} is part of the solution to
-bounce notification mails being delivered before their creation is logged
-(\sectionref{Bounce notification mails delivered before their creation is
-logged}); and \texttt{postsuper\_deleted\_queueids} caches information
-about mails recently deleted by the administrator so that subsequent log
-lines processed by the \action{SAVE\_DATA} action can be discarded
+tasks, setting up several state tables that will later be used by actions.
+Data about the connections and mails being processed is held in
+\texttt{connections} and \texttt{queueids} respectively.  The other state
+tables are used when working around complications in parsing:
+\texttt{timeout\_queueids} is used when dealing with connections that time
+out during the DATA phase (\sectionref{timeouts during data phase});
+\texttt{bounce\_queueids} is part of the solution to bounce notification
+mails being delivered before their creation is logged (\sectionref{Bounce
+notification mails delivered before their creation is logged}); and
+\texttt{postsuper\_deleted\_queueids} caches information about mails that
+were recently deleted by the administrator so that subsequent log lines
+processed by the \action{SAVE\_DATA} action can be discarded
 (\sectionref{Mail deleted before delivery is attempted}).
 
-When the framework loads the ruleset, it verifies each of the rules,
+The framework verifies each of the rules when it loads the ruleset,
 checking:
@@
 
-    \item That the regex is valid.  The regex will also be compiled for
-        efficiency (\sectionref{Caching compiled regexes}).
+    \item That the regex is valid.  The regex will have regex components
+        expanded (see \sectionref{regex components}), and will also be
+        compiled for efficiency (\sectionref{Caching compiled regexes}).
 
@@
 (\sectionref{Re-injected mails}) complicates the process of saving state,
-as the relationships between mails must be maintained; loading state is
-also complicated by dealing with aborted delivery attempts
+because the relationships between mails must be maintained; loading state
+is also complicated by dealing with aborted delivery attempts
 (\sectionref{aborted delivery attempts}), because a separate set of
 relationships between connections and mails must be re-created when the
-state tables are restored.
+state tables are restored.  The last step in loading the ruleset is to sort
+the rules as described in \sectionref{rule ordering for efficiency}.
 
@@
 (previously described in \sectionref{rule conditions in implementation}),
-falling back to generic rules if necessary, and warning if the log line is
-unrecognised.  The rules are sorted for increased efficiency; see
-\sectionref{rule ordering for efficiency} for details.  The repetitive
-nature of log files gives them high compression ratios; the framework uses
-a Perl module named \texttt{IO::Uncompress::AnyUncompress} to read
-compressed log files; this saves users the trouble of uncompressing them to
-a temporary file before parsing begins.  When used interactively, the
-framework also displays a progress bar to show how far parsing has
-progressed through the log file and how long the remainder of the parsing
-process is likely to take.  The progress bar is not as reliable when
-parsing compressed log files, because a compressed block may uncompress to
-a variable number of log lines; variation between the recognition and
-processing time of individual log lines also affects its accuracy, but
-overall the progress bar is a useful addition.  \parsername{} is primarily
-intended for parsing complete log files, but with minor modifications it
-could be used to parse a live log file, periodically checking it for new
-log lines; this could be very useful for programs that work best with up to
-date data, e.g.\ a program for monitoring the health of the mail system.
+falling back to generic rules if necessary, and finally warning if the log
+line is unrecognised.  The repetitive nature of log files gives them high
+compression ratios; the framework uses a Perl module named
+\texttt{IO::Uncompress::AnyUncompress} to read compressed log files, saving
+users the trouble of uncompressing them to a temporary file before parsing
+begins.  When used interactively, the framework displays a progress bar to
+show how far parsing has progressed through the log file and how long the
+remainder of the parsing process is likely to take.  The progress bar is
+not as accurate when parsing compressed log files, because a compressed
+block uncompresses to a variable number of log lines; variation between the
+recognition and processing time of individual log lines also affects its
+accuracy, but overall the progress bar is a useful addition.  \parsername{}
+is primarily intended for parsing complete log files, but with minor
+modifications it could be used to parse a live log file, periodically
+checking it for new log lines; this could be very useful for programs that
+work best with up to date data, e.g.\ a program for monitoring the health
+of the mail system, or graphs showing activity over the last five minutes.
 
@@
         (\graphref{Mean number of rules used per log line}); the framework
-        may need to try multiple rules for a given log line before it finds
+        may need to try multiple rules for each log line before it finds
         one that recognises it.
@@
         the correct rule for each log line.  This gives the best possible
-        running time, because only one rule is used for each log line.  A
-        variation of this is to use the correct rule last, to get the worst
-        possible running time.  \sectionref{perfect rule ordering}
-        discusses this in detail.
-
-    \item Each regex is compiled and the result cached when the rules are
-        loaded; this can be disabled and the regex compiled every time it
-        is used when recognising log lines.  The effect of this
-        optimisation is shown in \sectionref{Caching compiled regexes}.
-
-    \item Normally when a log line has been recognised the framework
+        running time, because only one rule is used to recognise each log
+        line, as discussed in \sectionref{perfect rule ordering}.
+
+    \item Each regex is compiled and the result cached when the ruleset is
+        loaded; this optimisation can be disabled and the regex compiled
+        every time it is used when recognising log lines.  The effect of
+        this optimisation is shown in \sectionref{Caching compiled
+        regexes}.
+
+    \item Normally, when a log line has been recognised the framework
         invokes the action specified by the rule.  Invoking the action can
@@
         storing results in the database is described at the beginning of
-        \sectionref{Evaluation}.
+        \sectionref{parser efficiency}.
 
@@
         which is useful when a warning dumps a mail's data for inspection.
-        The information that is added is: the log line's timestamp in human
+        The extra information added is: the log line's timestamp in human
         readable form; the entire log line; the name of the log file and
@@
         and the log line can be printed, so that the user can verify that
-        the correct rule recognised each log line.
+        the correct rule recognised each log line.  This is equivalent to
+        setting every rule's debug field to true.
 
@@
 A user, typically a mail administrator, would use these options when having
-difficulty adding or modifying rules to recognise log lines not recognised
-by \parsernames{} \numberOFrules{} rules.  Option 1 is useful for debugging
-a single rule, to check if it is recognising log lines it should not;
+difficulty extending the ruleset to recognise log lines not recognised by
+\parsernames{} \numberOFrules{} rules.  Option 1 is useful for debugging a
+single rule, to check if it is recognising log lines it should not;
 figuring out that a rule is not recognising log lines that it should
 recognise is a more difficult task.  Option 2 is useful when \parsername{}
-is warning about parsing problems, because all the information about a mail
-will be present in the warning message: the full text of every log line
-that is part of the mail; the rule that recognised each log line; the
-filename of the log file and the line number within it for each log line.
-Option 3 is less useful because of the volume of data it produces, though
-for small log file snippets, e.g.\ 1000 log lines, it is possible to
-manually verify that the correct rule recognised each log line.  Option 4
-is not useful for most users: it is for debugging problems with the data
-that is saved to the database, and sufficient safeguards are in place that
-there should be plenty of warning messages explaining what is wrong without
-using this option.
+is warning about parsing problems, because extra information about a mail
+will be present in the warning message.  Option 3 is less useful because of
+the volume of data it produces, though for small log file snippets, e.g.\
+1000 log lines, it is possible to manually verify that the correct rule
+recognised each log line.  Option 4 is not useful for most users: it is for
+debugging problems with the data that is saved to the database, and
+sufficient safeguards are in place that there should be plenty of warning
+messages explaining what is wrong without using this option.
 
@@
 this action does nothing when invoked, allowing uninteresting log lines to
-be parsed without any effects (it does not imply that the input is
-ungrammatical or unparsed).  Generally rules specifying the
+be parsed without any effects; it does not imply that the input is
+ungrammatical or unparsed.  Generally rules specifying the
 \action{UNINTERESTING} action recognise log lines that are not associated
@@
 these log lines are recognised and processed so that the framework can warn
-about unrecognised log lines, informing the user that they need to add to
-or extend their ruleset.  Most of the remaining actions have only one or
-two associated rules, because they will only ever have one or two log line
-variants, e.g.\ all log lines showing that a remote client has connected
-are recognised by a single rule and handled by the \action{CONNECT} action.
+about unrecognised log lines, informing the user that they need to extend
+the ruleset.  Most of the remaining actions have only one or two associated
+rules, because they will only ever have one or two log line variants, e.g.\
+all log lines showing that a remote client has connected are recognised by
+a single rule and processed by the \action{CONNECT} action.
 
@@
 \numberOFlogFILES{} log files used to generate the statistics in
-\sectionref{parser efficiency}, \textit{excluding\/} log files 22 and
+\sectionref{parser efficiency}, \textit{excluding\/} log files 22 \&
 62--68, because their content is extremely skewed by two mail loops.  The
@@
 (\action{CLEANUP\_PROCESSING}), and again the two have only a 0.01648\%
-difference between their invocations.  Confusingly,
-\action{CLEANUP\_PROCESSING} is invoked more often than
-\action{MAIL\_QUEUED}: occasionally, accepted mail in the process of being
-transferred is interrupted by a timeout, and there is a log line from
+difference between their number of invocations.
+\action{CLEANUP\_PROCESSING} is invoked slightly more often than
+\action{MAIL\_QUEUED}: occasionally, an accepted mail in the process of
+being transferred is interrupted by a timeout, and there is a log line from
 \daemon{cleanup} but not from \daemon{qmgr}, as described in
@@
 of times each action was invoked when parsing \numberOFlogFILES{} log
-files, excluding log files 22 and 62--68}{Number of times each action was
+files, excluding log files 22 \& 62--68}{Number of times each action was
 invoked excluding 22 and 62--68}
@@
 
-This section explains the actions found in \parsername{}; it may help to
+This section documents the actions found in \parsername{}; it may help to
 revisit the flow chart in \sectionref{flow chart} to see how a mail passes
@@
 or mail, each action will save all the data captured by the recognising
-rule's regex.  Each action is passed the same arguments:
+rule's regex; usually, log lines that lack identifying information will be
+processed by the \action{UNINTERESTING} action.  Each action is passed the
+same arguments:
 
@@
     \item [BOUNCE\_CREATED] Postfix 2.3 and subsequent versions log the
-        creation of bounce messages.  This action creates a new mail if
-        necessary; if the mail already exists the unknown origin flag will
-        be removed from it.  The action also marks the mail as a bounce
-        notification.  To deal with complication \sectionref{Bounce
-        notification mails delivered before their creation is logged} this
-        action checks a cache of recent bounce mails, to avoid creating
-        bogus bounce mails when log lines are out of order.
+        creation of bounce messages, and this action processes those log
+        lines.  This action creates a new mail if necessary; if the mail
+        already exists the unknown origin flag will be removed from it.
+        The mail will be marked as a bounce notification.  To deal with
+        complication \sectionref{Bounce notification mails delivered before
+        their creation is logged}, this action checks a cache of recent
+        bounce mails, to avoid incorrectly creating bounce mails when log
+        lines are out of order.
 
@@
         the mail will be usually be deleted from the state tables, but
-        deletion will be postponed if the mail is the parent of re-injected
-        mail (\sectionref{Re-injected mails}).
-
-    \item [CONNECT] Handle a remote client connecting: create a new
+        deletion will be postponed if the mail is the parent of mail
+        re-injected for forwarding (\sectionref{Re-injected mails}).
+
+    \item [CONNECT] Process a remote client connecting: create a new
         connection, indexed by \daemon{smtpd} \acronym{pid}.  If a
@@
 
-    \item [DELIVERY\_REJECTED] Postfix rejected an \acronym{SMTP} command
+    \item [DELIVERY\_REJECTED] Postfix rejected a mail delivery attempt
         from the remote client.  This is the action most frequently
@@
     \item [MAIL\_TOO\_LARGE] When a client tries to send a mail larger than
-        the local server accepts, the mail will be discarded and the client
-        informed of the problem.  The mail may have been accepted and
-        partially transferred; if so a data structure must be disposed of.
-        See \sectionref{timeouts during data phase} for full details;
-        although that describes timeouts the processing is the same for
-        mails that are too large.
+        the local server accepts, the mail will be discarded by Postfix and
+        the client informed of the problem.  The mail may have been
+        accepted and partially transferred; if so the parser will have a
+        data structure that must be disposed of.  See \sectionref{timeouts
+        during data phase} for full details; although that describes
+        timeouts, the processing is the same for mails that are too large.
 
     \item [PICKUP] The \action{PICKUP} action corresponds to the
-        \daemon{pickup} service dealing with a locally submitted mail.  A
-        new mail will be created, although out of order log lines may have
-        caused it to already exist, as documented in \sectionref{pickup
-        logging after cleanup}.
-
-    \item [POSTFIX\_RELOAD] When Postfix stops or reloads its
+        \daemon{pickup} service processing a locally submitted mail.  A new
+        mail will usually be created, although out of order log lines may
+        have caused it to already exist, as documented in
+        \sectionref{pickup logging after cleanup}.
+
+    \item [POSTFIX\_RELOAD] When Postfix stops running or reloads its
         configuration, it kills all \daemon{smtpd} processes, requiring all
-        connections to be cleaned up, entered in the database, and deleted
-        from the state tables.  Postfix probably kills all the other
-        components too, but \parsername{} is only affected by
-        \daemon{smtpd} processes exiting.
+        of the connections in \parsernames{} state tables to be cleaned up,
+        entered in the database, and deleted from the state tables.
+        Postfix probably kills all the other components too, but
+        \parsername{} is only affected by \daemon{smtpd} processes exiting.
 
@@
         entered into the database for future analysis; unfortunately this
-        means that the small amount of data that has been gathered will be
-        lost.
+        means that the small amount of data that has been gathered by
+        \parsername{} will be lost.
 
@@
         expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, as there are many other timeouts that should occur in
-        the intervening hours, e.g.\ timeouts for DNS requests or timeouts
-        reading data from the client.  The active connection for that
-        \daemon{smtpd} must be cleaned up, entered in the database, and
-        deleted from the state tables.
+        infrequently, because there are many other timeouts that should
+        occur in the intervening hours, e.g.\ timeouts for DNS requests or
+        timeouts reading data from the client.  The active connection for
+        that \daemon{smtpd} must be cleaned up, entered in the database,
+        and deleted from the state tables.
 
     \item [TIMEOUT] The connection with the remote client timed out, so the
-        mail being transferred must be discarded.  The mail may have been
-        accepted: if so there is a data structure to dispose of.  See
-        \sectionref{timeouts during data phase} for full details.
+        mail being transferred must be discarded by Postfix.  The mail may
+        have been accepted: if so the parser will have a data structure to
+        dispose of.  See \sectionref{timeouts during data phase} for full
+        details.
 
@@
     \item [UNINTERESTING] This action just returns successfully: it is used
-        when a log line needs to be parsed for completeness, but does not
-        either provide any useful data to be saved or require anything to
-        be done.
+        when a log line needs to be recognised to avoid warning about
+        unrecognised log lines, but does not either provide any useful data
+        to be saved or require any processing.
 
@@
 
-The rules are responsible for classifying each log line and specifying the
-correct action to be invoked.  In any parser implemented using this
-architecture, the rules will be the most visible component and most likely
-to be modified by users.  Rules need to be as simple as possible so that
-users can easily modify or add to them, but the implementation must balance
-that simplicity with the need to provide enough flexibility and power to
-successfully parse inputs.
+The rules are responsible for recognising each log line and specifying the
+correct action to be invoked.  The rules will be the most visible component
+in any parser implemented using this architecture, and also the component
+most likely to be modified by users.  Rules need to be as simple as
+possible so that users can easily modify them or add new rules, but each
+implementation must balance that simplicity with the need to provide enough
+flexibility and power to successfully parse inputs.
 
@@
 implementation}.  The process of creating new rules from unparsed log lines
-is dealt with in \sectionref{creating new rules in implementation}; the
-implementation of the utility supplied with \parsername{} to create regexes
-from unparsed log lines is also described in that section.  The rule
-conditions used in \parsername{} are explained next, and the final portion
-of this section provides some suggestions for how to detect overlapping
-rules.
+is dealt with in \sectionref{creating new rules in implementation},
+followed by the implementation of the utility supplied with \parsername{}
+to create regexes from unparsed log lines; the regex components provided by
+\parsername{} to ease writing of complex regexes are covered in
+\sectionref{regex components}.  The rule conditions used in \parsername{}
+are the penultimate topic (\sectionref{rule conditions in implementation}),
+and the section concludes with some suggestions for how to detect
+overlapping rules.
 
@@
 The example rule in \tableref{Example rule in implementation table}
-recognises the log line resulting from Postfix rejecting a delivery attempt
-because the domain in the sender address does not have an A, AAAA, or MX
-DNS entry, i.e.\ mail could not be delivered to any address in the sender's
-domain (for full details see~\cite{reject-unknown-sender-domain}).
-
-Example log line recognised by this rule:
+recognises the log line that results from Postfix rejecting a delivery
+attempt because the domain part of the sender address does not have an A,
+AAAA, or MX DNS entry, i.e.\ mail could not be delivered to any address in
+the sender's domain (for full details
+see~\cite{reject-unknown-sender-domain}).  A sample log line that would be
+recognised by this rule:
 
@@
 NOQUEUE: reject: RCPT from smtp.example.com[192.0.2.1]:
-  550 5.1.8 <foo@example.com>:
+  550 5.1.8 <alice@example.com>:
   Sender address rejected: Domain not found;
-  from=<foo@example.com> to=<info@example.net>
+  from=<alice@example.com> to=<bob@example.net>
   proto=SMTP helo=<smtp.example.com>
@@
 
-The fields in \tableref{Example rule in implementation table} are used as
-follows:
+\noindent{}The fields in \tableref{Example rule in implementation table}
+are used as follows:
 
@@
         (e.g.\ sender, recipient) will be automatically saved to the
-        results and connections tables.
+        results and connections tables.  The regex components used in the
+        regex are described in \sectionref{regex components}.
 
@@
         \sectionref{actions in architecture} for the role of actions in the
-        parser architecture.
+        architecture.
 
@@
 
-    \item [debug] controls the debugging information that can be generated
+    \item [debug] enables or disables printing of debugging information
         when this rule recognises a log line.
@@
 ease the process of parsing new log lines, the architecture separates rules
-from actions: adding new actions can be difficult, but adding new rules to
-recognise new log lines is trivial, and occurs much more frequently.
+from actions: adding new actions requires some effort, but adding new rules
+to recognise new log lines is trivial, and occurs much more frequently.
 
@@
 database, containing all the required attributes: action, name,
-description, program, and regex.  The name and description fields should be
-set based on the meaning of the log line, to help others understand which
-log lines this rule will recognise; the program will be obvious from the
-unrecognised log lines.  The action depends on the what the log line
-represents, e.g.\ a delivery rejection, a mail being delivered, some useful
-information, or something else; examine the list of actions in
-\sectionref{actions in detail in implementation} to determine the correct
-one.  The regex needs to be constructed based on the log line, but see
-below for a tool to ease the process.
+description, program, and regex; all the other attributes are either
+optional, set by the parser, or have sensible defaults.  The name and
+description fields should be set based on the meaning of the log line, to
+help others understand which log lines this rule will recognise; the
+program will be obvious from the unrecognised log lines.  The action
+depends on the what the log line represents, e.g.\ a delivery rejection, a
+mail being delivered, some useful information, or something else; examine
+the list of actions in \sectionref{actions in detail in implementation} to
+determine the correct one.  The regex needs to be constructed based on the
+log line, but see below for a tool to ease the process.
 
@@
 required when \parsername{} warns about an entry in the connections or
-results tables that is missing some required fields, because values for the
-missing fields are not present in any of the log lines.  For example, the
-rule recognising the \daemon{pickup} component processing a mail sets the
-client\_hostname to \texttt{localhost} and the client\_ip to
-\texttt{127.0.0.1}, because the mail originates on the local server.  If
-the new rule deliberately uses the architecture's overlapping rules feature
-it will be obvious that the priority field needs to be set, on this rule
-and possibly others; the priority field may be needed on unintentionally
-overlapping rules too, but that is more difficult to determine.  Finally
-the restriction\_name field should be set when the rule's action is
+results tables that is missing some required fields, because values for
+those fields are not present in any of the log lines for that connection or
+mail.  For example, the rule that recognises the \daemon{pickup} component
+processing a mail sets client\_hostname to \texttt{localhost} and
+client\_ip to \texttt{127.0.0.1}, because the mail originates on the local
+server.  If the new rule deliberately uses the architecture's overlapping
+rules feature the priority field needs to be set, on this rule and possibly
+others; the priority field may be needed on unintentionally overlapping
+rules too, but that is more difficult to determine.  Finally, the
+restriction\_name field should be set if the rule's action is
 \action{DELIVERY\_REJECTED}; the name of the restriction should be clear
@@
 
-    \item [Pre-process the file.]  The new algorithm leverages the
-        knowledge gained when writing rules and developing \parsername{};
-        it performs many substitutions on the input log lines, replacing
-        commonly occurring variable terms (e.g.\ email addresses, IP
-        addresses, the standard start of rejection messages) with keywords
-        described in \sectionref{regex components}.  The purpose of this
-        step is to utilise existing knowledge to create more accurate
-        regexes; it replaces many variable words with fixed words,
-        improving the subsequent classification of words as fixed or
-        variable.  The altered log lines are written to a temporary file,
-        which subsequent stages will use instead of the original input
-        file.
+    \item [Pre-process the file.]  The modified algorithm begins by
+        leveraging the knowledge gained when writing rules and developing
+        \parsername{}; it performs many substitutions on the input log
+        lines, replacing commonly occurring variable terms (e.g.\ email
+        addresses, IP addresses, the standard start of rejection messages)
+        with keywords described in \sectionref{regex components}.  The
+        purpose of this step is to utilise existing knowledge to create
+        more accurate regexes; it replaces many variable words with fixed
+        words, improving the subsequent classification of words as fixed or
+        variable.  Regex metacharacters in the log line will be escaped, to
+        avoid generating invalid or incorrect regexes.  The altered log
+        lines are written to a temporary file, which the next stage will
+        use instead of the original input file.
 
@@
         pass to filter out words: if the word's hash does not have a high
-        frequency, the word itself cannot have a high frequency, so there
-        is no need to maintain a counter for it, reducing the number of
-        counters required and thus the program's memory consumption.
+        frequency, the word itself cannot have a high frequency, so it must
+        be variable and does not need a counter maintained for it, reducing
+        the number of counters required and thus the program's memory
+        consumption.
 
@@
         indicate similarity, but the algorithm does not attempt to deal
-        with this possibility.  The algorithm maintains a counter for each
-        \texttt{(word, word's position within the log line)} tuple,
-        incrementing it each time that word occurs in that position.
+        with this possibility.  The modified algorithm maintains a counter
+        for each \texttt{(word, word's position within the log line)}
+        tuple, incrementing it each time that word occurs in that position.
 
         The original algorithm hashes each word and checks that result's
-        counter from the previous pass; if that counter has a high
+        counter from the previous pass: if that counter has a high
         frequency a separate counter will be maintained for this
-        \texttt{(word, word's position within the log line)} tuple.  This
-        reduces the number of counters maintained by the algorithm in this
-        step, reducing the memory requirements of the algorithm at the cost
-        of increased CPU usage.
-
-        As time goes by the amount of memory typically available to a
+        \texttt{(word, word's position within the log line)} tuple; if not,
+        the tuple will not have a counter.  This reduces the number of
+        counters maintained in this step, reducing memory requirements at
+        the cost of increased CPU usage.
+
+        As time goes by, the amount of memory typically available to a
         program or algorithm increases, and the need to reduce memory
         requirements correspondingly decreases, so the modified algorithm
-        omits the hashing step and maintains counters for all tokens.  Most
+        omits the hashing step and maintains counters for all tuples.  Most
         of the infrequently occurring words will have been substituted with
@@
         tuples to maintain counters for; the original algorithm does not
-        have the detailed knowledge leveraged by the new algorithm, because
-        it is a generic tool.
+        have the detailed knowledge leveraged by the modified algorithm,
+        because it is a generic tool.
 
@@
         are collapsed to a single \texttt{.+}; any resulting duplicate
-        regexes are combined, and their counters added together.  If a
-        regex's counter is lower than the threshold supplied by the user
-        the regex is discarded; this second threshold is independent of the
-        threshold used to differentiate between fixed and variable words,
-        but once again 1\% of log lines is a good starting point.
+        regexes and their counters are combined.  If a regex's counter is
+        lower than the threshold supplied by the user the regex is
+        discarded; this second threshold is independent of the threshold
+        used to differentiate between fixed and variable words, but once
+        again 1\% of log lines is a good starting point.
 
@@
 
-The new regexes are displayed for the user to add to the database, either
-as new rules or merged into the regexes of existing rules; also displayed
-are the number of log lines each regex was expected to match, and the
-number it actually matched, to help the user notice problems in the new
-regexes.  Discarding regexes will result in some of the log lines not being
-matched; when the rules have been augmented with the new regexes, the
-original log files should be parsed again, and any remaining unparsed log
-lines used as input to this utility.
-
-This utility is not intended to create perfect regexes, but it greatly
+The new regexes are displayed for the user to add to the ruleset, either as
+new rules or merged into the regexes of existing rules; also displayed are
+the number of unrecognised log lines each regex was expected to match, and
+the number it actually matched, to help the user notice problems in the new
+regexes.  Discarding regexes will result in some of the unrecognised log
+lines not being matched; when the ruleset has been augmented with the new
+regexes, the original log files should be parsed again, and any remaining
+unparsed log lines used as input to this utility.
+
+This utility is not expected to create perfect regexes, but it greatly
 reduces the effort required to deal with unrecognised log lines.  The
@@
 of the algorithm.
+
+\subsection{Regex Components}
+
+\label{regex components}
+
+Each rule's regex will have keywords expanded when the ruleset is loaded,
+for several reasons:
+
+\begin{itemize}
+
+    \item It simplifies both reading and writing of regexes and helps to
+        make each regex largely self-documenting.  For example, the meaning
+        of \_\_CLIENT\_HOSTNAME\_\_ is immediately clear, whereas its
+        expansion \verb!(?:unknown|(?:[-.\w]+))! needs to be deciphered
+        each time it is encountered.
+
+    \item It avoids needless repetition of complex regex components, and
+        allows the components to be corrected or improved in one location.
+        For example, \_\_SENDER\_\_ is used in 68 rules; if a mistake is
+        discovered in it the mistake only needs to be corrected in one
+        place.
+
+    \item It enables automatic extraction and saving of captured data.  The
+        regex snippets use Perl 5.10's named capture buffers~\cite{perlre}
+        to capture data, so the mapping between captures and fields does
+        not need to be explicitly specified by the rule.
+
+\end{itemize}
+
+To improve efficiency, the keywords are expanded and every rule's regex is
+compiled before attempting to parse the log file, otherwise every regex
+would be recompiled each time it was used, resulting in a large, data
+dependent slowdown, as described in \sectionref{Caching compiled regexes}.
+Most of the keywords are named after the fields in the connections or
+results tables they populate: \_\_CLIENT\_HOSTNAME\_\_, \_\_CLIENT\_IP\_\_,
+\_\_DELAY\_\_, \_\_DELAYS\_\_, \_\_ENHANCED\_STATUS\_CODE\_\_,
+\_\_HELO\_\_, \_\_MESSAGE\_ID\_\_, \newline{} \_\_QUEUEID\_\_,
+\_\_RECIPIENT\_\_, \_\_SENDER\_\_, \_\_SERVER\_HOSTNAME\_\_,
+\_\_SERVER\_IP\_\_, \_\_SIZE\_\_, and \_\_SMTP\_CODE\_\_.
+
+The other keywords need more explanation:
+
+\begin{eqlist}
+
+    \squeezeitems{}
+
+    \item [\_\_CHILD\_\_]  The queueid of a child mail; see
+        \sectionref{Re-injected mails}.
+
+    \item [\_\_COMMAND\_\_]  All \acronym{SMTP} commands.
+
+    \item [\_\_CONN\_USE\_\_]  How many times the connection was reused;
+        Postfix tries to reuse connections whenever possible to reduce the
+        load on both the sending and receiving servers.
+
+    \item [\_\_DATA\_\_]  This snippet is special: it does not match
+        anything by itself, so it must be followed by a pattern written by
+        the rule author, but it captures whatever is matched by that
+        pattern.  For example, \verb!/(__DATA__connection! \newline{}
+        \tab{}\verb!(?:refused|reset by peer))/! matches either
+        ``connection refused'' or ``connection reset by peer'' and causes
+        it to be saved to the data field of that log line's result.
+
+    \item [\_\_PID\_\_]  The \acronym{pid} of a \daemon{smtpd} process that
+        dies or is killed; see the \action{SMTPD\_DIED} action in
+        \sectionref{actions in detail in implementation}.
+
+    \item [\_\_RESTRICTION\_START\_\_]  Matches the standard information
+        Postfix includes at the start of almost all log lines resulting
+        from rejecting a delivery attempt.
+
+    \item [\_\_SHORT\_CMD\_\_]  Postfix sometimes logs \acronym{SMTP}
+        commands in a short, single word form; this snippet matches all of
+        those, except \texttt{DATA}, which typically has a more specific
+        rule.  Priorities could have been used instead of excluding
+        \texttt{DATA}.
+
+\end{eqlist}
+
+Some similarity exists between regex components and cascaded parsing: each
+regex component resembles a rule that recognises part of a log line and
+consumes it, leaving the remainder to be recognised by other components or
+cascaded rules.  The major difference between the two is that regex
+components are explicitly used by the author of the ruleset, whereas
+cascaded parsing would be dynamically applied by the framework.
 
@@
 
-\subsection{Regex Components}
-
-\label{regex components}
-
-Each rule's regex will have keywords expanded when the rules are loaded,
-for several reasons:
-
-\begin{itemize}
-
-    \item It simplifies both reading and writing of regexes and makes each
-        regex largely self-documenting.  E.g.\ the meaning of
-        \_\_CLIENT\_HOSTNAME\_\_ is immediately clear, whereas
-        \verb!(?:unknown|(?:[-.\w]+))! needs to be deciphered each time it
-        is encountered.
-
-    \item It avoids needless repetition of complex regex components, and
-        allows the components to be corrected or improved in one location.
-        E.g.\ \_\_SENDER\_\_ is used in 68 rules; if a mistake is
-        discovered it only needs to be corrected in one place.
-
-    \item It enables automatic extraction and saving of captured data.  The
-        regex snippets use Perl 5.10's named capture buffers~\cite{perlre}
-        to capture data, so the mapping between captures and fields does
-        not need to be explicitly specified by the rule.
-
-\end{itemize}
-
-For efficiency the keywords are expanded and every rule's regex is compiled
-before attempting to parse the log file, otherwise every regex would be
-recompiled each time it was used, resulting in a large, data dependent
-slowdown, as described in \sectionref{Caching compiled regexes}.  Most of
-the keywords are named after the fields in the connections or results
-tables they populate: \_\_CLIENT\_HOSTNAME\_\_, \_\_CLIENT\_IP\_\_,
-\_\_DELAY\_\_, \_\_DELAYS\_\_, \_\_ENHANCED\_STATUS\_CODE\_\_,
-\_\_HELO\_\_, \_\_MESSAGE\_ID\_\_, \_\_QUEUEID\_\_, \_\_RECIPIENT\_\_,
-\_\_SENDER\_\_, \_\_SERVER\_HOSTNAME\_\_, \_\_SERVER\_IP\_\_, \_\_SIZE\_\_,
-and \_\_SMTP\_CODE\_\_.
-
-Some of the keywords need more explanation:
-
-\begin{eqlist}
-
-    \squeezeitems{}
-
-    \item [\_\_CHILD\_\_]  The queueid of a child mail; see
-        \sectionref{Re-injected mails}.
-
-    \item [\_\_COMMAND\_\_]  All \acronym{SMTP} commands.
-
-    \item [\_\_CONN\_USE\_\_]  How many times the connection was reused;
-        Postfix tries to reuse connections whenever possible to reduce the
-        load on both the sending and receiving servers.
-
-    \item [\_\_DATA\_\_]  This snippet is special: it matches a zero-length
-        string, so it needs to be followed by a pattern that matches
-        appropriately, but it captures whatever is matched so that it can
-        be saved.  E.g.\ \verb!/(__DATA__connection (?:refused|reset!
-        \newline{} \tab{}\verb!by peer))/! matches either ``connection
-        refused'' or ``connection reset by peer'' and causes it to be saved
-        to the data field of that log line's result.
-
-    \item [\_\_PID\_\_]  The \acronym{pid} of a \daemon{smtpd} process that
-        dies or is killed; see the \action{SMTPD\_DIED} action in
-        \sectionref{actions in detail in implementation}.
-
-    \item [\_\_RESTRICTION\_START\_\_]  Matches the standard information
-        Postfix includes at the start of almost all log lines resulting
-        from rejecting a delivery attempt.
-
-    \item [\_\_SHORT\_CMD\_\_]  Postfix sometimes logs \acronym{SMTP}
-        commands in a short, single word form; this snippet matches all of
-        those, except \texttt{DATA}, which typically has a more specific
-        rule.  Priorities could have been used instead of excluding
-        \texttt{DATA}.
-
-\end{eqlist}
-
-Some similarity exists between regex components and cascaded parsing: each
-regex component resembles a rule that recognises part of a log line and
-consumes it, leaving the remainder to be recognised by other components or
-cascaded rules.  The major difference between the two is that regex
-components are explicitly used by the author of the ruleset, whereas
-cascaded parsing would be dynamically applied by the framework.
-
 \subsection{Overlapping Rules}
@@
     \item Sort rules by program and regex, then visually inspect the list,
-        e.g.\ with \acronym{SQL} similar to:    \newline{}
+        e.g.\ with an \acronym{SQL} query similar to:   \newline{}
         \verb!SELECT program, regex!                    \newline{}
@@
         The rules are sorted first by program, then by regex, because rules
-        with overlapping regexes will not overlap if their programs are
-        different.  Note that this does not properly deal with generic
-        rules whose program is \texttt{*}; those rules will be used on all
-        log lines that have not been recognised by program-specific rules.
+        cannot overlap if their programs are different.  Note that this
+        query does not properly deal with generic rules whose program is
+        \texttt{*}; those rules will be used on all log lines that have not
+        been recognised by program-specific rules.
 
@@
         existing rules to cope with the new or changed log lines.
-        Similarly using a new \acronym{DNSBL}, a new policy server, or new
+        Similarly, using a new \acronym{DNSBL}, a new policy server, or new
         administrator-defined rejection messages also requires new rules.
@@
         delivery is attempted} from occurring, but only when the mail has a
-        single recipient; when a mail has a single recipient address it
+        single recipient.  When a mail has a single recipient address it
         will be logged, but when a mail has multiple recipients no
@@
 
-    \item \parsername{} does not create separate mails where one or more
-        delivery attempts are rejected and subsequently a mail is accepted;
-        it will appear in the database as one mail with lots of rejections
-        followed by acceptance (this has already been mentioned in
-        \sectionref{connection reuse}).  It does not appear to be possible
-        to make this distinction given the data Postfix logs, though it
-        might be possible to write a policy server to provide additional
-        logging.
-
     \item \parsername{} will not detect that it is parsing the same log file
@@
     \item \parsername{} does not distinguish between log files produced by
-        different sources when parsing; all results will be saved to the
+        different servers when parsing; all results will be saved to the
         same database.  This may be viewed as an advantage, because log
-        files from different sources can be combined in the same database,
+        files from different servers can be combined in the same database,
         or it may be viewed as a limitation because it is impossible to
-        distinguish between log files from different sources in the same
+        distinguish between log files from different servers in the same
         database.  If the results of parsing log files from different
-        sources must remain separate, \parsername{} can easily be
+        servers must remain separate, \parsername{} can easily be
         instructed to use a different database.
@@
 \begin{enumerate}
-
-    \item Investigate and write the policy server referred to in limitation
-        3 above.
 
@@
 database schema, so the resemblance between a database schema and an
-\acronym{API} is described, followed by a detailed description and a
-diagram of the database schema.  The framework is documented next,
+\acronym{API} is described, followed by a diagram and a detailed
+description of the database schema.  The framework is documented next,
 including the steps it takes during initialisation, the parsing process,
@@
 component, and its implementation is examined in detail with: an example
-rule and an explanation of every field; a description of how to add new
-rules and determine the values that should be used for each field; an
-explanation of the algorithm used by the utility that creates new regexes
-based on unparsed log lines, and how it differs from the original algorithm
-it is based on.  How \parsername{} uses rule conditions and overlapping
-rules is discussed, accompanied by a description of the regex snippets the
-framework provides to simplify the regexes used in rules.  The many
-complications and difficulties encountered while writing \parsername{}, and
-the solutions developed to overcome them, are documented in depth; also
-documented are how some solutions interact, and which action or actions the
-solution is implemented in.  This chapter concludes with a list of the
-limitations identified in \parsername{}, followed by some possible
-improvements suggested by the author.  The next chapter will evaluate this
-implementation, examining both \parsernames{} efficiency and the coverage
-it achieves when parsing log files.
+rule and an explanation of how every field in that rule is used; a
+description of how to add new rules and determine the values that should be
+used for each field; an explanation of the algorithm used by the utility
+that creates new regexes based on unparsed log lines, and how it differs
+from the original algorithm it is based on.  How \parsername{} uses rule
+conditions and overlapping rules is discussed, accompanied by a description
+of the regex snippets the framework provides to simplify the regexes used
+in rules.  The many complications and difficulties encountered while
+writing \parsername{}, and the solutions developed to overcome them, are
+documented in depth; also documented are how some solutions interact, and
+which action or actions the solution is implemented in.  This chapter
+concludes with a list of the limitations identified in \parsername{},
+followed by some possible improvements suggested by the author.  The next
+chapter will evaluate this implementation, examining both \parsernames{}
+efficiency and the coverage it achieves when parsing log files.

### 2009-04-22 doc/logparser-acronyms.tex
Corrections based on proofreading to the preface and chapters 1-3.
Corrections after proofreading chapters 6 & 7 and the appendices.

@@
     description={
-        A DNS Blacklist is a simple collaborative anti-spam technique used
-        to reject or penalise mail sent from mail servers believed to be
-        the source of large volumes of spam.  The criteria used when
-        deciding if an IP address should be included on a DNSBL vary widely
-        between DNSBLs, so before using one it is very important to check
-        their listing policies.  To use a DNSBL the mail server makes a DNS
-        request incorporating the IP address of the client; if the
-        requested hostname exists the client is on the DNSBL, and the mail
-        server can decide what course of action to take.
+        A \acronym{DNSBL} is a simple collaborative anti-spam technique
+        used to reject or penalise mail sent from IP addresses believed to
+        be the source of large volumes of spam.  The criteria used when
+        deciding if an IP address should be included vary widely between
+        \acronym[s]{DNSBL}, so before using one it is very important to
+        check their listing policies.  To use a \acronym{DNSBL}, Postfix
+        makes a DNS request incorporating the IP address of the client; if
+        the requested hostname is found the client is on the
+        \acronym{DNSBL}.
     }
@@
         of existing code, so that each new program does not reinvent
-        existing wheels.  An API defines the interface provided to the user
-        of the existing code, and acts as a contract between the user and
-        the provider: if the user adheres to the API the provider
-        guarantees it will work, but is free to change the implementation
-        if the API is preserved.
+        existing wheels.  An \acronym{API} defines the interface provided
+        to the user of the existing code, and acts as a contract between
+        the user and the provider: if the user adheres to the \acronym{API}
+        the provider guarantees it will work, but is free to change the
+        underlying implementation if the \acronym{API} is preserved.
     }
@@
 \newcommand{\SMTPglossaryDescription}[0]{%
-    \acronym{SMTP} is the protocol used for transferring mail between the
-    sending and receiving \acronym{MTA}\@.  It is a simple, human readable,
-    plain text protocol, making it quite simple to test and debug problems
-    with it.  A detailed description of SMTP is beyond the scope of this
-    thesis: the original protocol definition is in
-    \acronym{RFC}~821~\cite{RFC821}, which was updated in
-    \acronym{RFC}~2821~\cite{RFC2821}.%
+    \glsreset{SMTP}\acronym{SMTP} is the protocol used for transferring
+    mail between the sending and receiving \acronym{MTA}\@.  It is a
+    simple, human readable, plain text protocol, making it quite simple to
+    test and debug problems with it.  A detailed description of
+    \acronym{SMTP} is beyond the scope of this thesis: the original
+    protocol definition is in \acronym{RFC}~821~\cite{RFC821}, which was
+    superceded by \acronym{RFC}~2821~\cite{RFC2821}.%
 }
@@
     description={
-        LMTP is a protocol derived from SMTP that removes the need for the
-        server to maintain a mail delivery queue, instead relying on the
-        client to maintain it.  Typically the client is an MTA, and the
-        server a delivery agent or a mail store.  Full details are
-        available in~\cite{lmtp-rfc}.
+        \acronym{LMTP} is a protocol derived from \acronym{SMTP} that
+        removes the need for the server to maintain a mail delivery queue,
+        instead relying on the client to maintain it.  Typically the client
+        is an \acronym{MTA}, and the server a delivery agent or a mail
+        store.  Full details are available in~\cite{lmtp-rfc}.
     }
@@
     description={
-        A Mail Transfer Agent sends and receives mail via SMTP\@.  Users
-        submit mail to an MTA via their mail client (e.g.\ Microsoft
-        Outlook, Thunderbird, webmail services); the sending MTA transfers
-        the mail to the receiving MTA (which may in turn forward it to
-        another MTA).
+        \glsreset{MTA}A \acronym{MTA} sends and receives mail via
+        \acronym{SMTP}\@.  Users submit mail to an \acronym{MTA} via their
+        mail client (e.g.\ Microsoft Outlook, Thunderbird, webmail
+        services); the sending \acronym{MTA} transfers the mail to the
+        receiving \acronym{MTA}.
     }
@@
     description={
-        The Request For Comments series is a series of proposals defining
-        various protocols and file formats, e.g.\ SMTP\@.  The name is
-        somewhat misleading nowadays: initially the authors were asking for
-        peer review, but these documents are now the de facto standards the
-        Internet runs on.
+        \glsreset{RFC}The \acronym{RFC} series is a series of proposals
+        defining various protocols and file formats, e.g.\
+        \acronym{SMTP}\@.  The name is somewhat misleading nowadays:
+        initially the authors were asking for peer review, but these
+        documents are now the de facto standards the Internet runs on.
     }
@@
     description={
-        An FQDN is a hostname plus domain name, e.g.\ \newline{}
+        An \acronym{FQDN} is a hostname plus domain name, e.g.\ \newline{}
             example.com     is a domain name                    \newline{}
             www             is a hostname                       \newline{}
-            www.example.com is a FQDN
+            www.example.com is a \acronym{FQDN}
     }
@@
         time, so the program's name is not suitable as a distinguishing
-        identifier; instead each process executing is given a pid that is
+        identifier; instead, each process is given a \acronym{pid} that is
         guaranteed to be unique for the lifetime of the process.  Once the
-        process has completed, the pid may be reused, as they are drawn
-        from a finite pool.
+        process has completed, the pid may be reused, because they are
+        drawn from a finite pool.
     }
@@
     description={
-        UCE is a more restrictive definition of spam than most people use:
-        it only covers mail that is explicitly commercial, thus excluding
-        viruses, Bayesian poisoning mails, backscatter\glsadd{Backscatter},
-        and those annoying chain letters you get from friends.
+        \acronym{UCE} is a more restrictive definition of spam than most
+        people use: it only covers mail that is explicitly commercial, thus
+        excluding viruses, Bayesian poisoning mails,
+        backscatter\glsadd{Backscatter}, and those annoying chain letters
+        you get from friends.
     }
@@
     description={
-        \acronym{ESMTP}~\cite{RFC1869}, provides a flexible mechanism for
+        \acronym{ESMTP}~\cite{RFC1869} provides a flexible mechanism for
         \acronym{SMTP} to be extended with new functionality, allowing new
@@
     }
-}
-
-\newglossaryentry{IP}{
-    type=\acronymtype,
-    name={IP},
-    description={Internet Protocol},
-    first={Internet Protocol (IP)}
 }
@@
     description={
-        Finite automata are computing devices that accept or recognize
+        \acronym{FA} are computing devices that accept or recognize
         regular languages.  They lack any form of storage, so although they
@@
     description={
-        A CFG is a grammar in which every transition or production takes
-        the form $L\rightarrow{}R$, where $L$ is a single non-terminal
-        symbol, and $R$ is a (possibly empty) string of terminal or
-        non-terminal symbols.  Every CFG can be recognised by a
-        \acronym{PDA}, and \acronyms{PDA} cannot recognise more complicated
-        grammars, so \acronyms{CFG} and \acronyms{PDA} are equivalent.
+        A \acronym{CFG} is a grammar in which every transition or
+        production takes the form $L\rightarrow{}R$, where $L$ is a single
+        non-terminal symbol, and $R$ is a (possibly empty) string of
+        terminal or non-terminal symbols.  Every \acronym{CFG} can be
+        recognised by a \acronym{PDA}, and \acronyms{PDA} cannot recognise
+        more complicated grammars, so \acronyms{CFG} and \acronyms{PDA} are
+        equivalent.
     }
@@
     description={
-        A Push-Down Automation is a computational device similar to a
-        \acronym{FA}, but it can additionally use a stack to store data.  A
-        \acronym{PDA} can manipulate the stack during state transitions
-        (adding or removing a single piece of data), and when determining
-        which state transition to take it can use the piece of data on the
-        top of the stack in addition to the input.  \acronyms{PDA} can
-        recognise languages of the form $a^{n}b^{n}$, i.e.\ any number of
-        one symbol followed by the same number of another symbol.
+        \glsreset{PDA}A \acronym{PDA} is a computational device similar to
+        a \acronym{FA}, but it can additionally use a stack to store data.
+        A \acronym{PDA} can manipulate the stack during state transitions
+        by adding or removing a single piece of data, and when determining
+        which state transition to take, the piece of data on the top of the
+        stack can be used in addition to the input.  In addition to the
+        languages recognised by \acronyms{FA}, \acronyms{PDA} can recognise
+        languages of the form $a^{n}b^{n}$, i.e.\ any number of one symbol
+        followed by the same number of another symbol.
     }

### 2009-04-22 doc/logparser-background.tex
Corrections after proofreading chapters 6 & 7 and the appendices.

@@
 
-    \item \acronym{SPF}~\cite{openspf}.  \acronym{SPF} records specify
-        which mail servers are allowed to send mail using sender addresses
-        from a particular domain.  The intention is to reduce spam from
-        faked sender addresses, backscatter~\cite{postfix-backscatter}, and
+    \item \acronym{SPF} checks:
+        \urlLastChecked{http://www.openspf.org/}{2009/04/21}.
+        \acronym{SPF} records specify which mail servers are allowed to
+        send mail using sender addresses from a particular domain.  The
+        intention is to reduce spam from faked sender addresses,
+        backscatter~\cite{postfix-backscatter}, and
         Joe~jobs~\glsadd{Joe-job}.  There has been considerable resistance
@@
 
-    \item Greylisting~\cite{greylisting} is a technique that temporarily
-        rejects a delivery attempt when the triple of \newline{}
-        \tab{}\texttt{(sender address, recipient address, remote IP
-        address)}\newline{} has not been seen before; on second and
-        subsequent delivery attempts from that triple the mail will be
-        accepted.  This blocks spam from some senders because maintaining a
-        list of failed addresses and retrying after a temporary failure is
-        uneconomical for a spam sender, but a legitimate mail server must
-        retry deliveries that fail temporarily.  Sadly spam senders are
-        using increasingly complex and well written programs to distribute
-        spam, frequently using an ISP provided \acronym{SMTP} server from a
-        compromised machine on the ISP's network.  Greylisting will slowly
-        become less effective as spam senders adapt, but it does block a
-        large percentage of spam mail at the moment; the most effective
-        restrictions from the \numberOFlogFILES{} log files used when
-        generating the results in \sectionref{Evaluation} are shown in
-        \tableref{Most effective Postfix restrictions}.  The table shows
-        that greylisting is worth using at the moment, particularly when
-        you take into account its position as the final restriction that a
-        mail must overcome in the configuration used on the mail server
-        that generated the log files: on that server greylisting only takes
-        effect for mails that have passed all other restrictions.  Some
-        problems may be encountered when using greylisting: some servers
-        fail to retry after a temporary failure, or legitimate mail may be
-        delayed, particularly when coming from a pool of servers.
+    \item Greylisting,
+        \urlLastChecked{http://www.greylisting.org/}{2009/04/21}, is a
+        technique that temporarily rejects a delivery attempt when the
+        triple of \newline{} \tab{}\texttt{(sender address, recipient
+        address, remote IP address)}\newline{} has not been seen before; on
+        second and subsequent delivery attempts from that triple the mail
+        will be accepted.  This blocks spam from some senders because
+        maintaining a list of failed addresses and retrying after a
+        temporary failure is uneconomical for a spam sender, but a
+        legitimate mail server must retry deliveries that fail temporarily.
+        Sadly spam senders are using increasingly complex and well written
+        programs to distribute spam, frequently using an ISP provided
+        \acronym{SMTP} server from a compromised machine on the ISP's
+        network.  Greylisting will slowly become less effective as spam
+        senders adapt, but it does block a large percentage of spam mail at
+        the moment; the most effective restrictions from the
+        \numberOFlogFILES{} log files used when generating the results in
+        \sectionref{Evaluation} are shown in \tableref{Most effective
+        Postfix restrictions}.  The table shows that greylisting is worth
+        using at the moment, particularly when you take into account its
+        position as the final restriction that a mail must overcome in the
+        configuration used on the mail server that generated the log files:
+        on that server greylisting only takes effect for mails that have
+        passed all other restrictions.  Some problems may be encountered
+        when using greylisting: some servers fail to retry after a
+        temporary failure, or legitimate mail may be delayed, particularly
+        when coming from a pool of servers.
 
@@
 
-    \item Scoring systems such as Policyd-weight~\cite{policyd-weight}
-        perform tests on features of the delivery attempt (e.g.\ IP
-        address, sender address), incrementing or decrementing a score
-        based on the results; if the final score is higher than a threshold
-        the delivery attempt is rejected.  The administrator must manually
-        whitelist clients if they are to bypass a Postfix restriction,
-        whereas using a threshold that requires a delivery attempt to hit
-        several scored restrictions will allow mail from clients that would
-        be rejected by a binary restriction.
+    \item Scoring systems such as postfwd,
+        \urlLastChecked{http://postfwd.org/}{2009/04/21}, perform tests on
+        features of the delivery attempt (e.g.\ IP address, sender
+        address), incrementing or decrementing a score based on the
+        results; if the final score is higher than a threshold the delivery
+        attempt is rejected.  The administrator must manually whitelist
+        clients if they are to bypass a Postfix restriction, whereas using
+        a threshold that requires a delivery attempt to hit several scored
+        restrictions will allow mail from clients that would be rejected by
+        a binary restriction.
 

### 2009-04-22 doc/logparser-conclusion.tex
Corrections after proofreading chapters 6 & 7 and the appendices.

@@
 
-Parsing Postfix log files appears at first sight to be an almost trivial
+Parsing Postfix log files appears at first sight to be an uncomplicated
 task, especially if one has previous experience in parsing log files, but
@@
 number of distinct log line variants is quite small; this mistaken
-impression is because of the uneven distribution exhibited by log lines
-produced in normal operation, vividly illustrated by \graphref{rule hits
-graph}.
+impression comes from the uneven distribution exhibited by log lines
+produced in normal operation, vividly illustrated in \graphref{rule hits
+graph}.  Given the diverse nature of Postfix log lines, and the ease with
+which administrators can cause new log lines to be logged
+(\sectionref{postfix background}), enabling users to easily extend the
+parser to deal with new log lines is a design imperative
+(\sectionref{parser design}).  Providing a tool to ease the generation of
+regexes from unrecognised log lines (\sectionref{creating new rules in
+implementation}) should greatly help users who need to extend their ruleset
+to recognise previously unrecognised log lines.
 
-Given the diverse nature of Postfix log lines, and the ease with which
-administrators can cause new log lines to be logged (\sectionref{postfix
-background}), enabling users to easily extend the parser to deal with new
-log lines is a design imperative (\sectionref{parser design}).  Providing a
-tool to ease the generation of regexes from unrecognised log lines should
-greatly help users to extend their ruleset to recognise previously
-unrecognised log lines (\sectionref{creating new rules in implementation}).
-The architecture developed for this project not only makes it trivial to
-recognise new inputs, adding new actions requires the minimum effort
-possible, allowing the implementer to focus on correctly implementing their
-new action.  The separation of the architecture into rules, actions, and
-framework (\sectionref{parser design}) is unusual, partly because the three
-are separated so completely.  Although parsers are often divided into
+This architecture's greatest strength is the ease with which parsers based
+on it can be adapted to deal with new requirements and inputs.  Parsing a
+variation of an existing input is a trivial task: simply modify an existing
+rule or add a new rule and the task is complete.  Parsing a new category of
+input is achieved by writing a new action and a rule for each input
+variant; quite often the new action will not need to interact with existing
+actions, but when interaction is required the framework provides the
+necessary facilities.  The architecture imposes very little red tape when
+writing new actions, allowing the implementer to focus their time and
+energy on correctly implementing their new action (\sectionref{actions in
+architecture}).  The separation of the architecture into rules, actions,
+and framework (\sectionref{parser design}) is unusual, partly because the
+three are separated so completely.  Although parsers are often divided into
 separate source code files (the combination of lex \&
@@
 complete parser by the compilation process; in contrast \parsername{} keeps
-the rules and actions separate until the parser runs.
+the rules and actions separate until the parser runs.  This separation
+enables the optimisations discussed in \sectionref{parser efficiency}, and
+it also allows different approaches to ruleset management, e.g.\ using
+machine learning techniques to seamlessly create or alter rules to
+recognise new inputs (\sectionref{adding new rules in architecture}).  The
+decoupling of rules from actions allows different sets of rules to be used
+with one set of actions, e.g.\ a parser might have actions to process
+versions one and two of a file format; by choosing the appropriate ruleset
+the parser will parse version one, or version two, or both versions.  A
+general purpose framework can be written, so that writing a parser just
+requires writing actions and rules.  The architecture makes it possible to
+apply commonly used programming techniques (such as object orientation,
+inheritance, composition, delegation, roles, modularisation, or closures)
+when designing and implementing a parser, simplifying the process of
+working within a team or developing and testing additional functionality.
+This architecture is ideally suited to parsing inputs that are not fully
+understood or do not follow a fixed grammar: the architecture warns about
+unrecognised inputs and errors encountered by actions, but continues
+parsing as best it can, allowing the developer of a new parser to decide
+which deficiencies are most important and require attention first, rather
+than being forced to fix the first error that arises.
 
@@
 architectures.  Most traditional parsers have a current state: each state
-has a fixed set of acceptable next states, and processing is determined by
-the state transition that takes place; unacceptable state transitions cause
-parsing to fail.  For example, when a \textbf{C} parser sees the keyword
-\textbf{for} it expects to immediately see a left parenthesis, with any
-other input causing parsing to fail (comments will already have been
-removed by the preprocessor).  This architecture is different: the rule
-that recognises the input dictates the action that will be invoked.  Rule
-conditions (\sectionref{rule conditions in architecture}) allow a stateful
-implementation of this architecture, where the list of rules used to
-recognise an input is constrained by the parser's current state, but the
-recognising rule still dictates the action that is invoked, and whether
-directly or indirectly, the next state.
+has a fixed set of acceptable next states, processing is determined by the
+state transition that takes place, and unacceptable state transitions cause
+parsing to fail.  This architecture is different: the rule that recognises
+the input dictates the action that will be invoked.  Rule conditions
+(\sectionref{rule conditions in architecture}) enable stateful parsing,
+where the list of rules used to recognise an input is constrained by the
+parser's current state, but the recognising rule still dictates the action
+that is invoked and, whether directly or indirectly, the next state.
 
-XXX SHOULD THIS NEXT PARAGRAPH BE IN THE PAST TENSE\@?
+When writing \parsername{}, the real difficulties arose once the parser was
+successfully recognising almost all of the log lines, because most of the
+irregularities and complications explained in \sectionref{complications}
+started to become apparent then.  Adding new rules to deal with numerous
+infrequently occurring log line variants was a simple if tiresome task,
+whereas dealing with mails that were missing information or where Postfix's
+actions were not being correctly reconstructed was much more grueling.
+Trawling through log files was extremely time consuming and quite error
+prone, searching for something out of the ordinary that might help diagnose
+the problem, and eventually finding it --- sometimes hundreds or even
+thousands of log lines away from the last occurrence of the queueid for the
+mail in question.  Sometimes the task was not to identify the unusual log
+line, but to spot that a log line normally present was missing, i.e.\ to
+realise that one log line amongst thousands was absent.  In all cases the
+evidence was used to construct a hypothesis to explain the irregularities,
+and that hypothesis was tested in \parsername{}; if successful, the parser
+was modified to deal with the irregularities, without adversely affecting
+existing parsing.  The complications documented in
+\sectionref{complications} are presented in the order they were solved in,
+and that order closely resembles the frequency in which they occur; the
+most frequently occurring complications dominate the warning messages
+produced, and so naturally they were the first complications to be dealt
+with.
 
-The real difficulties arose once the parser was successfully dealing with
-90\% of the log lines, because most of the irregularities and complications
-explained in \sectionref{complications} only became apparent when the vast
-majority of log lines were being parsed successfully.  Adding new rules to
-deal with the numerous infrequently occurring log line variants is a simple
-if tiresome task, whereas dealing with mails that are missing information
-or where Postfix's actions are not being correctly reconstructed is much
-more grueling.  Trawling through log files, searching for something out of
-the ordinary, possibly hundreds or even thousands of log lines away from
-the last occurrence of the queueid for the mail in question, is extremely
-time consuming and quite error prone.  Sometimes the task is not to spot
-the unusual log line, but to spot that a log line that is normally present
-is missing, i.e.\ to realise that one log line amongst thousands is absent.
-In all cases the evidence must be used to construct a hypothesis to explain
-the irregularities, and that hypothesis has to be tested in \parsername{};
-if successful, the parser will be modified to deal with the irregularities,
-without adversely affecting the existing parsing.  The complications
-described in \sectionref{complications} are described in the order they
-were solved in, and that order closely resembles the frequency with which
-they occur; the most frequently occurring complications dominate the
-warning messages produced, and so naturally they were the first
-complications to be dealt with.
+\parsername{} is not merely a proof of concept: it is intended to be used
+for parsing real-world log files from production mail servers, and the
+resulting data used to improve anti-spam defences.  This means that
+efficiency is important: parsing must complete in an reasonable period of
+time, so that the results can be used in a timely manner.  \parsernames{}
+efficiency is evaluated in \sectionref{parser efficiency}, where
+optimisations and the effect they have are explored.
 
-XXX WRITE ABOUT EFFICIENCY\@.
-
-A parser's ability to correctly parse its input is extremely important; the
-parser's coverage of \numberOFlogFILES{} log files is discussed in
+A parser's ability to correctly parse its inputs is extremely important;
+\parsernames{} coverage of \numberOFlogFILES{} log files is discussed in
 \sectionref{parsing coverage}.  Both its success at recognising individual
-log lines (\sectionref{log-lines-covered}) and its correctness in
-reconstructing each mail's journey through Postfix
-(\sectionref{mails-covered}) are described in detail, including the results
-of manually verifying the correct parsing of a subset of a log file.
-Experience implementing \parsername{} shows that full input coverage is
-relatively easy to achieve with this architecture, and that with enough
-time and effort a full understanding of the input is possible.  Postfix log
-files would require substantial time and effort to correctly parse
-regardless of the architecture used; this architecture enables an iterative
-approach to be used (similar to Stepwise
-Refinement~\cite{stepwise-refinement}), as is practiced in many other
-software engineering disciplines.
-
-This architecture's greatest strength is the ease with which
-implementations of it can be adapted to deal with new requirements and
-inputs.  Parsing a variation of an existing input is a trivial task: simply
-modify an existing rule or add a new rule and the task is complete.
-Parsing a new category of input is achieved by writing a new action and
-appropriate rules; quite often the new action will not need to interact
-with existing actions, but when interaction is required the framework
-provides the necessary facilities.  The decoupling of rules from actions
-allows different sets of rules to be used with one set of actions, e.g.\ a
-parser might have actions to process versions one and two of a file format;
-by choosing the appropriate ruleset the parser will parse version one, or
-version two, or both versions.  Decoupling also allows other approaches to
-rule management, as discussed in \sectionref{adding new rules in
-architecture}.  The architecture makes it possible to apply commonly used
-programming techniques (such as object orientation, inheritance,
-composition, delegation, roles, modularisation, or closures) when designing
-and implementing a parser, simplifying the process of working within a team
-or when developing and testing additional functionality.  This architecture
-is ideally suited to parsing inputs that are not fully understood or do not
-follow a fixed grammar: the architecture warns about unparsed inputs and
-other errors, but continues parsing as best it can, allowing the developer
-of a new parser to decide which deficiencies are most important and require
-attention first, rather than being forced to fix the first error that
-arises.
+log lines  and its correctness in reconstructing each mail's journey
+through Postfix are described in detail, including the results of manually
+verifying that a randomly selected portion of a log file was correctly
+parsed.  Experience implementing \parsername{} shows that full input
+coverage is relatively easy to achieve with this architecture, and that
+with enough time and effort a full understanding of the input is possible.
+Postfix log files would require substantial time and effort to correctly
+parse regardless of the architecture used; this architecture enables an
+iterative approach to be used~\cite{stepwise-refinement}, as is practiced
+in many other software engineering disciplines.
 
 The data gathered by \parsername{} provides the foundation for the future
-of this project: applying machine-learning algorithms to the data to
-analyse and optimise the set of anti-spam defences in use, followed by
-identifying patterns in the data that could be used to write new anti-spam
-techniques to recognise and reject spam rather than accepting it.  The
-parser provides the data in a normalised form that is far easier to use as
-input to new or existing implementations of machine-learning algorithms
-than trying to adapt each algorithm to extract data directly from the log
-files.  New policy servers (\sectionref{policy servers}), written to
-implement new anti-spam measures, could be tested or trained by using the
-data to simulate mail delivery attempts; this would allow simple, fast,
-reproducible testing, without the risk of adversely affecting a production
-mail server.  Development of \parsername{} is complete, i.e.\ it correctly
-parses Postfix log files, and in future it will require only maintenance;
-however, one avenue of future development under consideration is to extend
-it to parse non-Postfix log lines, e.g.\ SpamAssassin or Amavisd-new log
-lines.  \parsername{} can easily be extended to do this, but it lacks a
-method of associating the non-Postfix log lines with the existing data
-structures and state tables, so that all of the data for a mail delivery
-attempt can be stored together.
+of this project: using machine-learning algorithms to analyse the data and
+optimise the set of anti-spam defences in use, followed by identifying
+patterns in the data that could be used to write new anti-spam techniques
+to recognise and reject spam rather than accepting it.  The database
+provides the data in a normalised form that is far easier to use as input
+to new or existing implementations of machine-learning algorithms than
+trying to adapt each algorithm to extract data directly from log files.
+New policy servers, written to implement new anti-spam measures, can be
+tested or trained by using the collected data to simulate mail delivery
+attempts; this would allow simple, fast, reproducible testing, without the
+risk of adversely affecting a production mail server.  Development of
+\parsername{} is finished, i.e.\ it correctly parses Postfix log files, and
+in future it will only require maintenance; however, one avenue of future
+development under consideration is to extend it to parse non-Postfix log
+lines, e.g.\ SpamAssassin or Amavisd-new log lines.  \parsername{} can
+easily be extended to do this, but it requires a method of associating the
+non-Postfix log lines with the existing data structures and state tables,
+so that all of the data for a mail delivery attempt can be stored together.
 
-The Postfix log file parser based on this architecture provides a basis for
-systems administrators to monitor the effectiveness of their anti-spam
-measures and adapt their defences to combat new techniques used by those
-sending spam.  \parsername{} is a fully usable application, built to
-address a genuine need, rather than a proof of concept whose sole purpose
-is to illustrate a new idea; it deals with the oddities and difficulties
-that occur in the real world, rather than a clean, idealised scenario
-developed to showcase the best features of a new approach.
+\parsername{} provides a basis for systems administrators to monitor the
+effectiveness of their anti-spam measures and adapt their defences to
+combat new techniques used by those sending spam.  \parsername{} is a fully
+usable application, built to address a genuine need, rather than a proof of
+concept whose sole purpose is to illustrate a new idea; it deals with the
+oddities and difficulties that occur in the real world, rather than a
+clean, idealised scenario developed to showcase the best features of a new
+approach.

### 2009-04-22 doc/logparser-glossary.tex
Corrections after proofreading chapters 6 & 7 and the appendices.

@@
 \newglossaryentry{Backscatter}{name={Backscatter},description={
-    When a spam sender or worm sends mail with forged sender addresses,
-    innocent sites are flooded with undeliverable mail notifications; this
-    is called backscatter mail.
+    When a spam sender or virus sends mail with forged sender addresses,
+    innocent servers are flooded with undeliverable mail notifications;
+    this is called backscatter.
 }}
@@
 \newglossaryentry{Joe-job}{name={Joe~Job},description={
-    A joe~job describes spam mail sent using a faked sender address with
-    the intention of sullying the good name of the user of that address.
-    Joe~jobs are one cause of backscatter\glsadd{Backscatter}.
+    A joe~job describes a large amount of spam mail sent using a faked
+    sender address with the intention of sullying the good name of the user
+    of that address.  Joe~jobs are one cause of
+    backscatter\glsadd{Backscatter}.
 }}
@@
 \newglossaryentry{<>}{name={$<>$},sort={<>},description={
-    $<>$ is the sender address used for mail that should not be replied to,
-    e.g.\ bounce notifications.  In the \acronym{SMTP} dialog all addresses
-    are enclosed in $<>$, so \texttt{username@domain} is sent as
-    \texttt{$<$username@domain$>$}; thus $<>$ is actually an empty address,
-    but is always written as $<>$ for clarity.  All mail servers must
-    accept mail sent from $<>$, or they are in violation of
-    \acronym{RFC}~2821~\cite{RFC2821}.
+    $<>$ is the sender address used for sending bounce notifications.  In
+    the \acronym{SMTP} conversation all addresses are enclosed in $<>$, so
+    \texttt{username@domain} is sent as \texttt{$<$username@domain$>$};
+    thus $<>$ is actually an empty address, but it is always written as
+    $<>$ for clarity.  All mail servers must accept mail sent from $<>$, or
+    they are in violation of \acronym{RFC}~2821~\cite{RFC2821}.
 }}
@@
     it.  Queueids are assigned from a limited pool, so although they are
-    guaranteed to be unique for the lifetime of the mail, they may be
-    reused later.
+    guaranteed to be unique for the lifetime of the mail they are assigned
+    to, they may be reused later.
 }}
@@
     efficient manner.  A full description is beyond the scope of this
-    paper, further information can be found at~\cite{hash-functions}.
+    thesis, further information can be found at~\cite{hash-functions}.
 }}
@@
     AWK is a general purpose programming language that is designed for
-    processing text, and is available as a standard utility on all Unix
+    processing text and is available as a standard utility on all Unix
     systems.

### 2009-04-22 doc/logparser-other-parsers-review.tex
Corrections after proofreading chapters 6 & 7 and the appendices.

@@
 or restriction.  Some additional Postfix reports are supposedly available
-on \urlLastChecked{http://www.splunkbase.com/}{2008/11/23}, but the author
+on \urlLastChecked{http://www.splunkbase.com/}{2009/04/21}, but the author
 was unable to find any Postfix reports, or indeed reports for any other log
@@
 Copious documentation is made available on
-\urlLastChecked{http://www.splunk.com/}{2008/11/23}, but the abundance of
+\urlLastChecked{http://www.splunk.com/}{2009/04/21}, but the abundance of
 material and lack of organisation makes it hard to find the topic being
@@
 fields is provided in the documentation at
-\urlLastChecked{http://awstats.sourceforge.net/docs/awstats_faq.html\#MAIL}{2008/11/23},
+\urlLastChecked{http://awstats.sourceforge.net/docs/awstats_faq.html\#MAIL}{2009/04/21},
 so the author could neither understand what data is extracted, nor
@@
 YAALA provides a mini-language based on \acronym{SQL} that is used when
-generating reports; sample reports can be seen
-at~\urlLastChecked{http://www.yaala.org/samples.html}{2008/11/23}.  Example
+generating reports; sample reports can be seen at
+\urlLastChecked{http://www.yaala.org/samples.html}{2009/04/21}.  Example
 query for HTTP proxy servers: \newline{} \tab{} \texttt{requests BY file

### 2009-04-22 doc/logparser-postfix-daemons.tex
Corrections from proofreading for chapters 4 and 5.
Corrections after proofreading chapters 6 & 7 and the appendices.

@@
 \newglossaryentry{bounce}{name={bounce},type={postfix},description={
-    The bounce daemon is responsible for sending bounce notifications in
+    The bounce daemon is responsible for generating bounce notifications in
     Postfix version 2.3 and later.
@@
 \newglossaryentry{local}{name={local},type={postfix},description={
-    Local is the Postfix component responsible for local delivery of mail
-    (i.e.\ mail delivered on the server Postfix is running on); this
-    includes alias expansion, processing of a user's \texttt{.forward}
-    file, and delivery of the mail, whether to a user's mailbox or a
-    program such as a mailing list manager.
+    The Postfix component responsible for local delivery of mail (i.e.\
+    mail delivered on the server Postfix is running on); this includes
+    alias expansion, processing of a user's \texttt{.forward} file, and
+    delivery of the mail, whether to a user's mailbox or a program such as
+    a mailing list manager.
     \daemonDocURL{http://www.postfix.org/local.8.html}{2009/02/23}.
@@
     Pickup is the daemon that deals with mail submitted locally via
-    \daemon{postdrop} and \daemon{sendmail,} passing the mail on to
-    \daemon{cleanup} for further processing.
+    \daemon{postdrop}, passing the mail on to \daemon{cleanup} for further
+    processing.
     \daemonDocURL{http://www.postfix.org/pickup.8.html}{2009/02/23}.
@@
     its input into a newly created mail in the queue, for processing by
-    \daemon{pickup} and subsequent delivery by one or more of
-    \daemon{lmtp}, \daemon{local}, \daemon{smtp}, or \daemon{virtual}.
+    \daemon{pickup} and subsequent delivery.
     \daemonDocURL[\newline{}]{http://www.postfix.org/postdrop.1.html}{2009/02/23}.
@@
     The Postfix component responsible for delivery of mails to virtual
-    domains.  When \daemon{local} delivers mail the destination is
+    domains.  When \daemon{local} delivers mail, the destination is
     determined only by the portion of the email address on the left side of
-    the \texttt{@}, whereas with \daemon{virtual} delivery the destination
-    is determined by the entire email address.  E.g.\ if the server is
-    responsible for both the \texttt{example.org} and \texttt{example.net}
-    domains: \daemon{local} would deliver mail for
+    the \texttt{@}, whereas when \daemon{virtual} delivers mail, the
+    destination is determined by the entire email address.  For example, if
+    the server is responsible for both the \texttt{example.org} and
+    \texttt{example.net} domains: \daemon{local} would deliver mail for
     \texttt{john@example.org} and \texttt{john@example.net} to the same

### 2009-04-22 doc/logparser-results.tex
Corrections after proofreading chapters 6 & 7 and the appendices.

@@
 This chapter evaluates \parsername{} on two criteria: efficiency, and
-coverage of  Postfix log files.  \parsername{} is intended to be used in a
-production environment, and must be capable of parsing log files generated
-by a high volume mail server in a reasonable time period.  Accurate but
-slow parsing is preferable to sloppy but quick parsing, so the goal of
-better parser efficiency must be balanced against the requirement for
-precise and correct processing of log files.
+coverage of  Postfix log files.  Efficiency is important because
+\parsername{} is intended to be used in a production environment, and must
+be capable of parsing log files generated by a high volume mail server in a
+reasonable period of time.  Full coverage of Postfix log files is important
+because the data gathered by the parser must be accurate and complete for
+it to be useful and reliable.  The two goals are mutually incompatible:
+every extra measure implemented to improve accuracy slows down parsing.
+Some of the complications described in \sectionref{complications} occur
+fewer than ten times when parsing the \numberOFlogFILES{} log files used to
+evaluate \parsername{} in this chapter, yet their solutions slow down
+parsing of every log file.  However, accurate but slow parsing is
+preferable to sloppy but quick parsing.
 
 The performance evaluation begins by describing the characteristics of the
-mail server that produced the log files used evaluate the parser's
+mail server that produced the log files used to evaluate the parser's
 performance and coverage; the computer that the tests were run on is also
@@
 described next, with an explanation of why \parsername{} has better
-performance with the larger log files in the group.  The effect of using
-different rule orderings is explored, and the use of optimal rule ordering
-is compared to use of an oracle that allows the parser to use only one rule
-when recognising each log line.  When \parsername{} is used by other mail
-administrators, they will need to extend the ruleset to parse their own log
-lines, so the next section addresses the question of how parser performance
-is affected by an increase in the number of rules in the ruleset.  The
-penultimate topic to be covered is the simple optimisation of caching the
-results of compiling each regex, and the huge effect it has on parser
-efficiency.  The performance evaluation concludes by examining where
-parsing time is spent: recognition of log lines or their subsequent
-processing?
+performance when parsing the larger log files in the group.  The effect of
+using different rule orderings is explored, and the use of optimal rule
+ordering is compared to use of an oracle that allows the parser to use only
+one rule when recognising each log line.  When \parsername{} is used by
+other mail administrators, they will need to extend the ruleset to parse
+their own log lines, so the next section addresses the question of how
+parser performance is affected by an increase in the number of rules in the
+ruleset.  The penultimate topic covered is the simple optimisation of
+caching the results of compiling each rule's regex, and the huge effect it
+has on parser efficiency.  The performance evaluation concludes by
+examining where parsing time is spent: recognition of log lines or their
+subsequent processing?
 
 The second criterion the parser is evaluated on is its coverage of Postfix
-log files.  This topic consists of two sections: what proportion of log
+log files.  This subject contains two sections: what proportion of log
 lines are correctly recognised by the ruleset, and what proportion of mail
@@
 Efficiency is an obvious concern when the parser routinely needs to parse
-large log files.  The server that generated the log files used in testing
-this parser accepts approximately 10,000 mails for 700 users each weekday;
-\graphref{Mails received per day} and \tableref{Number of mails received
-per day: statistics} show that, as expected, far more mails are received on
+large log files.  The mail server that generated the \numberOFlogFILES{}
+log files is a production mail server handling mail for a university
+department; the benefit of using this mail server is that its log files
+exhibit the idiosyncrasies and peculiarities that a mail server in the wild
+must deal with, but the downside is that significantly altering the
+configuration to accommodate this project is not an option.
+\Graphref{Mails received per day} and \tableref{Number of mails received
+per day: statistics} show that, as expected, it accepts far more mails on
 weekdays than at weekends.  Note that these figures only count mails
 received by \acronym{SMTP}, and the mail loops noticeable in later graphs
-do not affect the figures.  Median log file size is
-\input{build/include-median-file-size}, containing
-\input{build/include-median-number-of-log-lines} log lines; large scale
-mail servers would have much larger log files.  The mail server in question
-is a production mail server handling mail for a university department; the
-benefit of using this mail server is that its log files exhibit the
-idiosyncrasies and peculiarities a mail server in the wild must deal with,
-but the downside is that significantly altering the configuration to
-accommodate this project is not an option.
+do not affect the figures.
 
@@
         CPU         & One dual core 2.40GHz Intel\textregistered{}
-                        Core\texttrademark{}~2 CPU,                     \\
+                        Core$^{\textsc{TM}}$2 CPU,                      \\
                     & with 32KB L1 cache and 4MB L2 cache.              \\
@@
 \numberOFlogLINEShuman{} log lines) were each parsed 10 times, and the mean
-parsing time.  The computer used for test runs was a Dell Optiplex 745,
-shown in \tableref{Details of the computer used to generate statistics}; it
-was dedicated to the task of gathering statistics from test runs, and did
-not run any other programs while test runs were in progress.  Saving
-results to the database was disabled for the test runs, because that
+parsing time calculated.  The computer used for test runs was a Dell
+Optiplex 745, shown in \tableref{Details of the computer used to generate
+statistics}; it was dedicated to the task of gathering statistics from test
+runs, and did not run any other programs while test runs were in progress.
+Saving results to the database was disabled for the test runs, because that
 dominates the run time of the parser, and the tests are aimed at measuring
@@
 \input{build/include-insert-results-duration}, with mean throughput of
-\input{build/include-insert-results-throughput} --- parsing makes up only
+\input{build/include-insert-results-throughput} --- parsing takes up only
 \input{build/include-skip-inserting-as-percentage-of-inserting} of the
@@
 fairly consistent sizes and contents, except for two groups of log files:
-22 \& 62--68.  \Graphref{parsing time vs log file size vs number of log
-lines graph} shows the parsing time in seconds, log file size in MB, and
-number of log lines in tens of thousands, for each of the
-\numberOFlogFILES{} log files.  \Graphref{parsing time vs log file size vs
-number of log lines factor} plots both the ratio of log file size vs
-parsing time, and the ratio of number of log lines vs parsing time (higher
-is more efficient in both cases); \tableref{parsing time vs log file size
-vs number of log lines factor table} shows the same ratios for different
-groups of log files.  The ratios are quite tightly banded, except they
-increase (i.e.\ improve) for log files 22 \& 62--68, despite their larger
-than usual size.  Both groups of log files are much larger than usual
-because of mail loops caused by users who set up mail forwarding
-incorrectly, resulting in a very different distribution of log lines:
-normally most log lines are generated by mail delivery attempts from other
-hosts, but when the mail loops occurred most of the log lines resulted from
-failed delivery of mail generated on the server itself.  The Postfix
-components that generated most of the log lines during the mail loops have
-fewer associated rules than the Postfix components whose log lines normally
-make up the bulk of each log file, so the mean number of rules tried per
-log line reduces, as does the mean parsing time per log line.
-\Tableref{Number of rules per Postfix component} shows the number of rules
-for each Postfix component; \graphref{Mean number of rules used per log
-line} shows the drop in the mean number of rules tried per log line for log
-files containing a mail loop, and \graphref{Mean number of rules used per
-log line for each Postfix component} shows the mean number of rules tried
-per log line for each Postfix component.
-
-\showtable{build/include-file-size-and-number-of-log-lines-vs-parsing-time}{Ratio
-of log file size and number of log lines to parsing time}{parsing time vs
-log file size vs number of log lines factor table}
+22 \& 62--68.  Median log file size is
+\input{build/include-median-file-size}, containing
+\input{build/include-median-number-of-log-lines} log lines, with a median
+of \input{build/include-median-mails-accepted-over-smtp} mails accepted
+each weekday; large scale mail servers would generate much larger log
+files.  \Graphref{parsing time vs log file size vs number of log lines
+graph} shows the parsing time in seconds, log file size in MB, and number
+of log lines in tens of thousands, for each of the \numberOFlogFILES{} log
+files.  \Graphref{parsing time vs log file size vs number of log lines
+factor} plots both the ratio of log file size vs parsing time, and the
+ratio of number of log lines vs parsing time (higher is more efficient in
+both cases); \tableref{parsing time vs log file size vs number of log lines
+factor table} shows the same ratios for different groups of log files.  The
+ratios are quite tightly banded, except they increase (i.e.\ improve) for
+log files 22 \& 62--68, despite their larger than usual size.  Both groups
+of log files are much larger than usual because of mail loops caused by
+users who set up mail forwarding incorrectly, resulting in a very different
+distribution of log lines: normally most log lines are generated by mail
+delivery attempts from other hosts, but when the mail loops occurred most
+of the log lines resulted from failed delivery of mail generated on the
+server itself.  The Postfix components that generated most of the log lines
+during the mail loops have fewer associated rules than the Postfix
+components whose log lines normally make up the bulk of each log file, so
+the mean number of rules tried per log line is lower for these log files,
+as is the mean parsing time per log line.  \Tableref{Number of rules per
+Postfix component} shows the number of rules for each Postfix component;
+\graphref{Mean number of rules used per log line} shows the drop in the
+mean number of rules tried per log line for log files containing a mail
+loop, and \graphref{Mean number of rules used per log line for each Postfix
+component} shows the mean number of rules tried per log line for each
+Postfix component for all log files.
 
@@
 
+\showtable{build/include-file-size-and-number-of-log-lines-vs-parsing-time}{Ratio
+of log file size and number of log lines to parsing time}{parsing time vs
+log file size vs number of log lines factor table}
+
 \showgraph{build/graph-average-number-of-rules-tried-per-log-line}{Mean
@@
 \showgraph{build/graph-average-number-of-rules-tried-per-program}{Mean
-number of rules used per log line for each Postfix component}{Mean number
-of rules used per log line for each Postfix component}
+number of rules used per log line for each Postfix component over all log
+files}{Mean number of rules used per log line for each Postfix component}
 
@@
 the other 90\% of the rules, as shown in \graphref{rule hits graph}.
-Assuming that the distribution of log lines is reasonably consistent over
-time, \parsernames{} efficiency should benefit from using rules that
+Assuming that the distribution of log lines is reasonably consistent across
+log files, \parsernames{} efficiency should benefit from using rules that
 recognise log lines more frequently before it uses rules that recognise log
@@
     \item [Optimal]  The most optimal order, according to the hypothesis:
-        rules that recognise log lines most often will be tried first.
+        rules that recognise log lines most often will be used first.
 
@@
     \item [Reverse] Hypothetically the worst order: rules that recognise
-        log lines most frequently will be tried last.
+        log lines most frequently will be used last.
 
@@
 ordering graph} shows how the oracle and optimal ordering perform relative
-to shuffled ordering; \tableref{Parsing time of oracle and optimal ordering
-relative to shuffled ordering table} shows mean and standard deviation.  As
-expected, the oracle is more efficient than optimal ordering, but not by
+to shuffled ordering, with mean and standard deviation in \tableref{Parsing
+time of oracle and optimal ordering relative to shuffled ordering table}.
+As expected, the oracle is more efficient than optimal ordering, but not by
 much.  \Graphref{Percentage increase in parsing time when using optimal
@@
 its lowest when parsing log files resulting from a mail loop (log files 22
-\& 62--68), because the mean number of rules tried per log line is lower
+\& 62--68), because the mean number of rules used per log line is lower
 (see \graphref{Mean number of rules used per log line}).  The performance
 of optimal ordering relative to the oracle is much worse when parsing the
-first log file than for the remainder of the log files, because for the
-first log file the hits field of every rule is zero, so optimal ordering
-does not provide any benefit for that log file; the oracle, in contrast, is
-flawless for every log file.
+first log file than for the remainder of the log files, because the hits
+field of every rule starts at zero, so optimal ordering does not provide
+any benefit for the first log file; the oracle, in contrast, is flawless
+for every log file.
 
@@
 component in the maximum and minimum rulesets} shows the number of rules
-per Postfix component for each ruleset.  Once again, for the first log
-file, optimal and reverse orderings have identical performance, because the
-hits field of every rule starts at zero.
+per Postfix component for each ruleset.  Once again, \graphref{Percentage
+parsing time increase when using the maximum ruleset instead of the minimum
+ruleset} shows that optimal and reverse orderings have identical
+performance for the first log file, because the hits field of every rule
+starts at zero.
 
-The optimal ordering has a mean increase of just
+The optimal ordering shows a mean increase of just
 \input{build/include-full-ruleset-vs-minimum-ruleset-mean} in parsing time
@@
     \centering{}
-    \begin{tabular}{lrr}
+    \begin{tabular}{lrrr}
         \tabletopline{}%
-        Postfix component & Maximum ruleset & Minimum ruleset \\
+        Postfix component & Maximum ruleset & Minimum ruleset & Difference \\
         \tablemiddleline{}%
@@
 CPU time: for most regexes it takes far more CPU time than the actual
-matching.  If the interpreter is certain that a regex will not change while
-the program is running, it will automatically cache the results of
+matching does.  If the interpreter is certain that a regex will not change
+while the program is running, it will automatically cache the results of
 compiling and optimising the regex for later use.  The results of compiling
@@
 parsing time when the optimisation is disabled; this is because, on
-average, fewer regexes are compiled per log line for those log files.  The
-increase in parsing time when parsing the first log file is much greater
-than for the other log files; again, this is because every rule's hits
-field starts at zero, so optimal ordering is less efficient than usual, the
-mean number of rules tried for each log line will be higher than usual, and
-more regexes will need to be compiled when recognising each log line.
+average, fewer rules are used so fewer regexes are compiled per log line
+for those log files.  The increase in parsing time when parsing the first
+log file is much greater than for the other log files; again, this is
+because every rule's hits field starts at zero, so optimal ordering is less
+efficient than usual, the mean number of rules used for each log line is
+higher than usual, and more regexes will need to be compiled when
+recognising each log line.
 
@@
     \item Even with the optimisations described in this chapter enabled,
-        recognising log lines still dominates the execution time of the
-        parser.  \Graphref{Percentage of parsing time spent recognising log
-        lines graph} shows the percentage of parsing time spent recognising
-        log lines for each of the \numberOFlogFILES{} log files, with mean
-        and standard deviation shown in \tableref{Percentage of parsing
-        time spent recognising log lines table}; for normal log files,
+        recognising log lines still accounts for two thirds of the parser's
+        execution time.  \Graphref{Percentage of parsing time spent
+        recognising log lines graph} shows the percentage of parsing time
+        spent recognising log lines for each of the \numberOFlogFILES{} log
+        files, with mean and standard deviation shown in
+        \tableref{Percentage of parsing time spent recognising log lines
+        table}; for normal log files,
         \input{build/include-percentage-time-spent-recognising-log-lines-mean}
@@
         \input{build/include-optimal-ordering-parsing-time-reduction-other-logs}\%
-        relative to shuffled ordering; processing recognised log lines
-        occupies
+        relative to shuffled ordering; actions occupy
         \input{build/include-percentage-time-spent-processing-log-lines-mean}
@@
 plenty of existing literature is available on the topic of optimising
-programs, so the subject will not be dealt with here.
+programs, so the subject is not dealt with here.
 
@@
 achieved.  Correctly understanding and reconstructing every mail delivery
-attempt is important so that the information in the database is correct and
+attempt is important so that the information in the database is accurate and
 complete.  Improving the proportion of log lines correctly recognised is
@@
     \item Skip the log line if the ruleset does not contain any rules for
-        the Postfix component that produced the log line.
+        the Postfix component that produced it.
 
-    \item Try each rule until a recognising rule is found; if the log line
-        is not recognised, issue a warning and move on to the next log
-        line.
+    \item Try each rule for that Postfix component, then generic rules,
+        until a recognising rule is found; if the log line is not
+        recognised, issue a warning and move on to the next log line.
 
@@
 Each Postfix component whose log lines are of interest must have at least
-one rule, or its log lines will be silently skipped; in the extreme case of
-an empty ruleset the parser would skip every log line.  \parsername{} skips
-log lines logged by programs that do not have any rules because there may
-be any number of log lines from other programs intermingled in the log
-file, and some Postfix components that do not produce any log lines of
-interest.  \parsername{} does not parse log lines from non-Postfix
-programs, e.g.\ Amavisd-new or SpamAssassin; it could easily be extended to
-do so, if a method could be developed to correctly associate such log lines
-with existing state table entries.  To correctly recognise all log lines,
-there must be a rule to recognise each log line variant produced by each
-Postfix component; if a log line is not recognised the parser will issue a
-warning, to inform the user that they need to extend their ruleset.  Each
-rule's regex should be as specific and precise as possible, to ensure
-accurate parsing: a rule with a regex that matches zero or more of any
-character will recognise every log line, but not in a meaningful way; most
-log lines contain fixed strings, so this is not a problem in practice.
+one rule, or all of its log lines will be silently skipped; in the extreme
+case of an empty ruleset the parser would skip every log line.
+\parsername{} skips log lines logged by programs that do not have any
+associated rules, because there may be any number of log lines from other
+programs in the log file, and some Postfix components do not produce any
+log lines of interest.  \parsername{} does not parse log lines from
+non-Postfix programs, e.g.\ Amavisd-new or SpamAssassin; it could easily be
+extended to do so, if a method could be developed to correctly associate
+such log lines with existing state table entries.  To correctly recognise
+all log lines, there must be a rule to recognise each log line variant
+produced by each Postfix component; if a log line is not recognised the
+parser will issue a warning, to inform the user that they need to extend
+their ruleset.  Each rule's regex should be as specific and precise as
+possible, to ensure accurate parsing: a rule with a regex that matches zero
+or more of any character will recognise every log line, but not in a
+meaningful way; most log lines contain fixed strings, so this is not a
+problem in practice.
 
 Full coverage of log lines can be achieved without undue effort, yet once
-achieved it is hard to maintain.  Maintaining full coverage is hard because
+achieved it requires maintenance.  Maintaining full coverage takes effort because
 the set of log line variants changes over time, e.g.\ administrators add
@@
 regex is, the author is confident that zero false positives occur when
-parsing the \numberOFlogFILES{} log files.  On initial appearances,
-exercising only 36 rules from a total of \numberOFrules{} when parsing 6039
-log lines seems low, but after examining \graphref{rule hits graph} it
-becomes apparent that such a low number of rules is to be expected.  The
-reader should also bear in mind that even when parsing all
-\numberOFlogFILES{} log files, only \numberOFrulesMINIMUM{} rules are used,
-because many of the rules recognise log lines that only appear in other log
-files.
+parsing the \numberOFlogFILES{} log files.
 
@@
 reference to the log files is the best way to detect mails that were not
-handled properly (many of the complications discussed in
-\sectionref{complications} were detected in this way).  \parsername{}
-issues warnings when it detects any errors or discrepancies, e.g.\ when a
-queueid is reused but the previous mail remains in the state tables, when a
-queueid or \acronym{pid} is not found in the state tables, or when an entry
-in the state tables does not include sufficient data to satisfy the
-database schema.  The parser should produce few or no warnings during
-parsing, and when finished parsing the state tables should only contain
-entries for mails that have log lines in subsequent log files.  There will
-often be warnings about a missing queueid or \acronym{pid} when parsing the
-first few thousand log lines, because the earlier log lines for those
-connections or mails are in a previous log file; loading saved state tables
-from the previous log file will solve this problem.
+reconstructed properly; many of the complications discussed in
+\sectionref{complications} were detected in this way.  \parsername{} issues
+warnings when it detects any errors or discrepancies, e.g.\ when a queueid
+is reused but the previous mail remains in the state tables, when a queueid
+or \acronym{pid} is not found in the state tables, or when an entry in the
+state tables does not include sufficient data to satisfy the database
+schema.  The parser should produce few or no warnings during parsing, and
+when finished parsing the state tables should only contain entries for
+mails that have log lines in subsequent log files.  There will often be
+warnings about a missing queueid or \acronym{pid} when parsing the first
+few thousand log lines, because the earlier log lines for those connections
+or mails are in a previous log file; loading saved state tables from the
+previous log file will solve this problem.
 
@@
 The state tables will contain entries for mails not yet delivered when the
-parser finishes execution.  Ideally, that is all they will contain, though
-they may also contain mails whose initial log lines are not contained in
-the log files.  Any other entries in the state tables are evidence of
-either a failure in parsing, or an aberration in the log files.  After
-parsing the \numberOFlogFILES{} log files, the state tables contain 18
-entries:
+parser finishes parsing the log file.  Ideally, those are the only entries
+they will contain, though they may also contain mails whose initial log
+lines are not contained in the log files.  Any other entries in the state
+tables are evidence of either a failure in parsing, or an aberration in the
+log files.  After parsing the \numberOFlogFILES{} log files, the state
+tables contain 18 entries:
 
@@
 
-    \item 1 connection that started only seconds before the log files
-        ended and had not yet been fully transferred from client to server.
+    \item 1 connection that started only seconds before the log files ended
+        and the mail had not yet been fully transferred from client to
+        server.
 
@@
     \item \parsername{} performs many checks to detect known problems
-        occurring, e.g.\ if a mail is missing required data, and no
+        occurring, e.g.\ if a mail is missing required data.  No such
         warnings are produced during the test runs other than the five
@@
         described in \sectionref{out of order log lines} imposes conditions
-        that each reconstructed mail must comply with to be acceptable.
+        that every reconstructed mail must comply with to be acceptable,
+        not just the five mails that exhibit that complication.
 
@@
 \numberOFlogFILES{} log files is infeasible, but verifying the parsing of a
-sample from those log lines is a tractable albeit extremely time consuming
+sample from those log files is a tractable albeit extremely time consuming
 task.  A sample of log lines was obtained by randomly selecting a log file,
@@
 conjunction with the log file segment and a dump of the resulting database,
-verifying that for each of the log lines \parsername{} recognised it with
-the correct rule and invoked the correct action, which in turn correctly
+verifying that \parsername{} recognised each of the log lines with the
+correct rule and invoked the correct action, which in turn correctly
 processed the log line and saved the correct data to be inserted into the
@@
 remaining in the state tables, 1625 mail delivery attempts correctly
-entered in the database, zero false positives, and zero false negatives.
+entered in the database, 0 false positives, and 0 false negatives.
 
@@
 reconstructing the journey a mail delivery attempt takes through Postfix is
-the aim of the parser.  Increasing the proportion of log lines correctly
-recognised is relatively simple and non-intrusive: adding new rules or
-modifying existing rules is very easy because of the separation of rules,
-actions, and framework in both the architecture and \parsername{}.
+the purpose of the parser.  Increasing the proportion of log lines
+correctly recognised is relatively simple and non-intrusive: adding new
+rules or modifying existing rules is very easy because of the separation of
+rules, actions, and framework in both the architecture and \parsername{}.
 Improving the understanding and reconstruction of mail delivery attempts is

### 2009-04-23 doc/logparser-conclusion.tex
Changes made based on Carl's comments.

@@
 
-The flow of control in this architecture is quite different from most other
-architectures.  Most traditional parsers have a current state: each state
-has a fixed set of acceptable next states, processing is determined by the
-state transition that takes place, and unacceptable state transitions cause
-parsing to fail.  This architecture is different: the rule that recognises
-the input dictates the action that will be invoked.  Rule conditions
-(\sectionref{rule conditions in architecture}) enable stateful parsing,
-where the list of rules used to recognise an input is constrained by the
-parser's current state, but the recognising rule still dictates the action
-that is invoked and, whether directly or indirectly, the next state.
+The flow of control in this architecture is quite different from other
+architectures, e.g.\ those used for compiling a programming language.
+Typically, those parsers have a current state: each state has a fixed set
+of acceptable next states, processing is determined by the state transition
+that takes place, and unacceptable state transitions cause parsing to fail.
+This architecture is different: the rule that recognises the input dictates
+the action that will be invoked.  Rule conditions (\sectionref{rule
+conditions in architecture}) enable stateful parsing, where the list of
+rules used to recognise an input is constrained by the parser's current
+state, but the recognising rule still dictates the action that is invoked
+and, whether directly or indirectly, the next state.
 
@@
 A parser's ability to correctly parse its inputs is extremely important;
-\parsernames{} coverage of \numberOFlogFILES{} log files is discussed in
-\sectionref{parsing coverage}.  Both its success at recognising individual
-log lines  and its correctness in reconstructing each mail's journey
-through Postfix are described in detail, including the results of manually
-verifying that a randomly selected portion of a log file was correctly
-parsed.  Experience implementing \parsername{} shows that full input
-coverage is relatively easy to achieve with this architecture, and that
-with enough time and effort a full understanding of the input is possible.
-Postfix log files would require substantial time and effort to correctly
-parse regardless of the architecture used; this architecture enables an
-iterative approach to be used~\cite{stepwise-refinement}, as is practiced
-in many other software engineering disciplines.
+\parsernames{} coverage of \numberOFlogFILES{} log files, each containing
+one day's log lines, is discussed in \sectionref{parsing coverage}.  Both
+its success at recognising individual log lines  and its correctness in
+reconstructing each mail's journey through Postfix are described in detail,
+including the results of manually verifying that a randomly selected
+portion of a log file was correctly parsed.  Experience implementing
+\parsername{} shows that full input coverage is relatively easy to achieve
+with this architecture, and that with enough time and effort a full
+understanding of the input is possible.  Postfix log files would require
+substantial time and effort to correctly parse regardless of the
+architecture used; this architecture enables an iterative approach to be
+used~\cite{stepwise-refinement}, as is practiced in many other software
+engineering disciplines.
 

### 2009-04-23 doc/logparser-results.tex
Changes made based on Carl's comments.

@@
 per day: statistics} show that, as expected, it accepts far more mails on
-weekdays than at weekends.  Note that these figures only count mails
-received by \acronym{SMTP}, and the mail loops noticeable in later graphs
-do not affect the figures.
+weekdays than at weekends.  Note that mail generated on the server does not
+contribute to the figures: in particular, the vast amount of mail resulting
+from the mail loops noticeable in later graphs is not reflected in these
+figures.
 
@@
 When generating the timing data used in this section, \numberOFlogFILES{}
-log files (totaling \input{build/include-total-size-of-93-log-files},
-\numberOFlogLINEShuman{} log lines) were each parsed 10 times, and the mean
-parsing time calculated.  The computer used for test runs was a Dell
-Optiplex 745, shown in \tableref{Details of the computer used to generate
-statistics}; it was dedicated to the task of gathering statistics from test
-runs, and did not run any other programs while test runs were in progress.
-Saving results to the database was disabled for the test runs, because that
-dominates the run time of the parser, and the tests are aimed at measuring
-the speed of \parsername{} rather than the speed of the database and the
-disks it is stored on.  Parsing all \numberOFlogFILES{} log files in one
-run without saving results to the database took
+log files were used, each containing one day's log lines, totalling
+\input{build/include-total-size-of-93-log-files}, \numberOFlogLINEShuman{}
+log lines.  Each log file was parsed 10 times, and the mean parsing time
+calculated.  The computer used for test runs was a Dell Optiplex 745, shown
+in \tableref{Details of the computer used to generate statistics}; it was
+dedicated to the task of gathering statistics from test runs, and did not
+run any other programs while test runs were in progress.  Saving results to
+the database was disabled for the test runs, because that dominates the run
+time of the parser, and the tests are aimed at measuring the speed of
+\parsername{} rather than the speed of the database and the disks it is
+stored on.  Parsing all \numberOFlogFILES{} log files in one run without
+saving results to the database took
 \input{build/include-full-run-duration}, with mean throughput of
@@
 of log lines in tens of thousands, for each of the \numberOFlogFILES{} log
-files.  \Graphref{parsing time vs log file size vs number of log lines
+files.  Parsing time is plotted against number of log lines in
+\graphref{Parsing time plotted against number of log lines}, and against
+log file size in \graphref{Parsing time plotted against log file size};
+the points on both graphs form almost perfectly straight lines.
+
+\Graphref{parsing time vs log file size vs number of log lines
 factor} plots both the ratio of log file size vs parsing time, and the
@@
 
+The three plots on \graphref{parsing time vs log file size vs number of log
+lines graph} are quite close together, and the graph can be difficult to
+read.  Log files 62--68 show a large gap between log file size and number
+of log lines, whereas those plots appear to have a small gap or to overlap
+for all the other log files, suggesting that the ratio of log file size to
+number of log lines is lower for log files 62--68.  This appearance is
+misleading: \graphref{Mean log line size in bytes} shows the mean log line
+size for all \numberOFlogFILES{} log files, and it reduces for log files 22
+\& 62--68, i.e.\ there are more log lines per MB in those log files.
+
 \showgraph{build/graph-input-size-vs-parsing-time}{Parsing time, log file
@@
 time vs log file size vs number of log lines graph}
+
+\showgraph{build/graph-parsing-time-vs-file-size}{Parsing time plotted
+against log file size}{Parsing time plotted against log file size}
+
+\showgraph{build/graph-parsing-time-vs-number-of-log-lines}{Parsing time
+plotted against number of log lines}{Parsing time plotted against number of
+log lines}
 
@@
 log file size vs number of log lines factor table}
-
-\showgraph{build/graph-average-number-of-rules-tried-per-log-line}{Mean
-number of rules used per log line}{Mean number of rules used per log line}
-
-\showgraph{build/graph-average-number-of-rules-tried-per-program}{Mean
-number of rules used per log line for each Postfix component over all log
-files}{Mean number of rules used per log line for each Postfix component}
 
@@
 \end{table}
+
+\showgraph{build/graph-average-number-of-rules-tried-per-log-line}{Mean
+number of rules used per log line}{Mean number of rules used per log line}
+
+\showgraph{build/graph-average-number-of-rules-tried-per-program}{Mean
+number of rules used per log line for each Postfix component over all log
+files}{Mean number of rules used per log line for each Postfix component}
+
+\showgraph{build/graph-mean-bytes-per-log-line}{Mean log line size in
+bytes}{Mean log line size in bytes}
 
@@
 
-\Graphref{Parsing time of optimal and reverse orderings relative to
-shuffled ordering graph} shows the parsing times of optimal and reverse
-orderings relative to shuffled ordering; the mean relative parsing times
-for different groupings of log files are given in \tableref{Parsing time of
-optimal and reverse orderings relative to shuffled ordering table}.  This
-optimisation provides a mean reduction in parsing time of
+The parsing time of the three orderings are plotted against log file size
+in \graphref{Parsing time plotted against log file size for optimal,
+shuffled, and reverse orderings}; \graphref{Parsing time of optimal and
+reverse orderings relative to shuffled ordering graph} shows the parsing
+times of optimal and reverse orderings relative to shuffled ordering, and
+mean relative parsing times for different groupings of log files are given
+in \tableref{Parsing time of optimal and reverse orderings relative to
+shuffled ordering table}.  This optimisation provides a mean reduction in
+parsing time of
 \input{build/include-optimal-ordering-parsing-time-reduction-other-logs}\%
@@
 \showgraph{build/graph-hits}{Log lines recognised per rule}{rule hits graph}
+
+\showgraph{build/graph-file-size-optimal-shuffle-reverse}{Parsing time
+plotted against log file size for optimal, shuffled, and reverse
+orderings}{Parsing time plotted against log file size for optimal,
+shuffled, and reverse orderings}
 
@@
 time increase when using the maximum ruleset instead of the minimum ruleset
-table}.
+table}.  Note that in both \graphref{Percentage parsing time increase when
+using the maximum ruleset instead of the minimum ruleset} and
+\tableref{Percentage parsing time increase when using the maximum ruleset
+instead of the minimum ruleset table}, for each ordering the parsing time
+using the maximum ruleset is compared to the ordering's parsing time using
+the minimum ruleset: the three orderings are not compared to a common
+baseline.
 
@@
 extremely well as the ruleset increases in size, and that optimally
-ordering the rules is an important optimisation contributing to this
-scalability.
+ordering the rules makes an important contribution to this scalability.
 
@@
 compiled regexes is probably the single most effective optimisation
-possible in \parsername{}, and was quite simple to implement.  As seen
-previously, log files 22 \& 62--68 do not suffer such a large increase in
-parsing time when the optimisation is disabled; this is because, on
+possible in \parsername{}, and was quite simple to implement: the framework
+compiles each rule's regex when the ruleset is loaded, and uses the
+compiled regex instead of the source regex when recognising log lines.  As
+seen previously, log files 22 \& 62--68 do not suffer such a large increase
+in parsing time when the optimisation is disabled; this is because, on
 average, fewer rules are used so fewer regexes are compiled per log line
@@
 
-The optimisations described in this chapter have optimised the process of
+The optimisations described in this chapter have addresses the process of
 recognising log lines, but have not optimised actions at all.  Optimisation
@@
         \input{build/include-percentage-time-spent-recognising-log-lines-mean}
-        of parsing time is spent recognising log lines.  Optimal ordering
-        reduces parsing time by
+        of parsing time is spent recognising log lines.  To measure how
+        long recognition of log lines takes, a full test run was performed
+        as previously described in \sectionref{parser efficiency}, but with
+        one difference: actions were not invoked when a log line was
+        recognised, so the parsing time did not include the time normally
+        taken by invocation of actions; the mean parsing times for each log
+        file were subtracted from the mean parsing time from a normal test
+        run to obtain the parsing time taken by actions.
+
+        Optimal ordering reduces parsing time by
         \input{build/include-optimal-ordering-parsing-time-reduction-other-logs}\%

### 2009-04-23 doc/logparser-architecture.tex
Changes based on comments by Ariane.

@@
     \empty{}\label{Similarities between ATN and this architecture}
+    \centering{}
     \begin{tabular}[]{lll}
@@
         \tablemiddleline{}%
-        Networks        & Parser        & Determines the sequence
+        Networks        & Framework     & Determines the sequence
                                           of transitions              \\
@@
 data in one form and transforms it to another form; \parsername{}
-transforms log files to a database.
+transforms log files to a \acronym{SQL} database.
 
@@
 that finds the rule that recognises the input and invokes the requested
-action can be expressed in pseudo-code as:
+action can be expressed in pseudo-code as shown in \figureref{Pseudo-code
+describing how inputs are recognised}.
 
+\begin{figure}[thbp]
+    \caption{Pseudo-code describing how inputs are recognised}
+    \empty{}\label{Pseudo-code describing how inputs are recognised}
 \begin{verbatim}
@@
 \end{verbatim}
+\end{figure}
 
@@
 
-    \item [Classify the input]  The pseudo-code above shows how rules are
+    \item [Classify the input]  The pseudo-code in \figureref{Pseudo-code
+        describing how inputs are recognised} shows how rules are
         successively tried until one is found that recognises the current

### 2009-04-23 doc/logparser-other-parsers-review.tex
Changes based on comments by Ariane.

@@
 parsers against those already available, to accurately judge what
-improvements, if any, are delivered by the newcomers.
+improvements, if any, are delivered by the newcomers.  \Tableref{Summary of
+parsers' features} summarises the results of this review.
 
@@
         required for this project, or deal with the complications described
-        in \sectionref{complications}.
+        in \sectionref{complications}.  Some correlate log lines by
+        queueid, but none correlate log lines by \acronym{pid}
+        (\sectionref{queueid vs pid}).
 
@@
 
-Each of the reviewed parsers was tested with the \numberOFlogFILES{} test
-log files described in \sectionref{parser efficiency}; \tableref{Summary of
-parsers' features} summarises the results of this review.
+Each of the reviewed parsers was tested with the three months
+(\numberOFlogFILES{} days) of contiguous log files described in
+\sectionref{parser efficiency}.
 
@@
 code, enhanced status code, sender, recipient, size, message-id, delay, and
-delays.
+delays.  Note: \parsername{} correlates log lines by both queueid and
+\acronym{pid} (\sectionref{queueid vs pid}), and it stores each mail's
+queueid, but it does not store each connection's \acronyms{pid}, because a
+use has not been identified for that data.
 
@@
 \acronym{CSV} record format used by \acronym{LMA} is not documented, but
-the output file contains a comment giving the format:\newline{} \texttt{\#
-Timestamp|Nome Client|IP Client|IP Server|From|To|Status|Size}
+the output file contains a comment in Italian giving the format:\newline{}
+\texttt{\# Timestamp|Nome Client|IP Client|IP Server|From|To|Status|Size}
 \newline{}\acronym{LMA} treats \acronym{CSV} lines starting with
@@
 to document this structure.  MySQL support was not tested because the
-schema is not documented, so the required tables could not be created.
+schema is not documented, and no existing database was available to extract
+the schema from, so the required tables could not be created.
 
@@
 \acronym{SMTP} code signifies a rejection.  All of these reports are
-trivially simple to produce from the \acronym{CSV} file using the standard
+extremely simple to produce from the \acronym{CSV} file using the standard
 Unix tool \texttt{awk}\glsadd{awk}; the most complicated, DAILY-REJECT, is
@@
 supplied with thorough and well written documentation; everything the
-author searched for was documented, except the MySQL database schema and
-some details of the data extracted by the parser.  A commercial version of
-MySQL is required because of MySQL licensing restrictions, but Sawmill's
-documentation explains why and includes instructions on how to compile
-Sawmill so that it can use a non-commercial version of MySQL (this was not
-attempted during the review process).
+author looked for was documented, except for the MySQL database schema and
+some details of the data extracted by the parser, e.g.\ what the virus
+field stores.  A commercial version of MySQL is required because of MySQL
+licensing restrictions, but Sawmill's documentation explains why and
+includes instructions on how to compile Sawmill so that it can use a
+non-commercial version of MySQL (this was not attempted during the review
+process).
 
@@
 (described later) or the full text of the log line, though it is not
-possible to search on partial words.  Search results are not available for
-use with other tools.  Searches can be saved for reuse; saved searches can
-be run periodically and the results mailed to a recipient or sent to an
-external program for further processing.  The author was unable to save
-searches, though that may have been because of limitations in the free
-version.  The web interface is optimised for interactive use rather than
-automated queries, and it does not appear to be possible to write
-independent tools to utilise the Splunk database, whereas \parsername{}
-provides the database and leaves it to the user to utilise it without limit
-or restriction.  Some additional Postfix reports are supposedly available
-on \urlLastChecked{http://www.splunkbase.com/}{2009/04/21}, but the author
-was unable to find any Postfix reports, or indeed reports for any other log
+possible to search on partial words.  Searches can be saved for reuse;
+saved searches can be run periodically and the results mailed to a
+recipient or sent to an external program for further processing.  The
+author was unable to save searches, possibly because of limitations in the
+free version, and so was unable to examine the format of the data.  The web
+interface is optimised for interactive use rather than automated queries,
+and it does not appear to be possible to write independent tools to utilise
+the Splunk database, whereas \parsername{} provides the database and leaves
+it to the user to utilise it without limit or restriction.  Some additional
+Postfix reports are supposedly available on
+\urlLastChecked{http://www.splunkbase.com/}{2009/04/21}, but the author was
+unable to find any Postfix reports, or indeed reports for any other log
 file types: every category was empty, even those that the web site claimed
@@
 though all except the bubble and heatmap graphs are variations of a bar or
-pie chart.  It is trivial to drill down through the graphs to select a
-portion of the data, e.g.\ select the hour with the largest number of
-events, then select a particular host, and finally a specific sender
+pie chart.  Drilling down through the graphs to select a portion of the
+data is simple and intuitive, e.g.\ select the hour with the largest number
+of events, then select a particular host, and finally a specific sender
 address.  All searches performed using the indexed data returned reasonable
@@
 codes, delivery delays, and message-id), though it does not make the full
-text of the log line available; this could be trivially added if desired,
-but would greatly increase the size of the resulting database.
+text of the log line available; a few minutes work could add this to
+\parsername{} if desired, but it would greatly increase the size of the
+database.
 
@@
         Pflogsumm       & Yes           & No                & Partial \dag{}    & Good           & Yes              \\
-        Sawmill         & Yes           & Yes               & Searches          & Excellent      & \nialpha{}   \\
+        Sawmill         & Yes           & Yes               & Searches          & Excellent      & Yes \nialpha{}   \\
         Splunk          & Yes           & Yes               & Searches          & Abundant       & No               \\
@@
                         &               &                   & \& reports        &                &                  \\
-        Lire            & Yes           & Yes               & Yes               & Reasonable     & Yes          \\
+        Lire            & Yes           & Yes               & Yes               & Poor           & Yes              \\
         Logrep          & No            & Yes               & No                & None           & Yes              \\
@@
 
-        \item [\nialpha{}] Sawmill's source code is available in an
-            encrypted form, so that customers can compile it on platforms
-            that pre-compiled binaries are not available for.
+        \item [\nialpha{}] Sawmill's source code is available, but in an
+            encrypted or obfuscated form.
 

### 2009-04-23 doc/logparser-results.tex
Changes based on comments by Ariane.

@@
 When generating the timing data used in this section, \numberOFlogFILES{}
-log files were used, each containing one day's log lines, totalling
+log files were used, each containing one day's log lines, totalling three
+months of contiguous log files,
 \input{build/include-total-size-of-93-log-files}, \numberOFlogLINEShuman{}

### 2009-04-23 doc/logparser-acronyms.tex
Changes after running diction. Changed acknowledgements.

@@
         deciding if an IP address should be included vary widely between
-        \acronym[s]{DNSBL}, so before using one it is very important to
-        check their listing policies.  To use a \acronym{DNSBL}, Postfix
-        makes a DNS request incorporating the IP address of the client; if
-        the requested hostname is found the client is on the
-        \acronym{DNSBL}.
+        \acronym[s]{DNSBL}, so before using one it is essential to check
+        their listing policies.  To use a \acronym{DNSBL}, Postfix makes a
+        DNS request incorporating the IP address of the client; if the
+        requested hostname is found the client is on the \acronym{DNSBL}.
     }
@@
     \acronym{SMTP} is beyond the scope of this thesis: the original
-    protocol definition is in \acronym{RFC}~821~\cite{RFC821}, which was
+    protocol definition is in \acronym{RFC}~821~\cite{RFC821}, later
     superceded by \acronym{RFC}~2821~\cite{RFC2821}.%

### 2009-04-23 doc/logparser-background.tex
Changes after running diction. Changed acknowledgements.

@@
 only when legitimate mail is impeded.\/}  It is neither as friendly nor as
-catchy, but it more accurately reflects the current situation.} A typical
-\acronym{SMTP} conversation resembles the following (the lines starting
-with a three digit number are sent by the server, all other lines are sent
-by the client):
+catchy, but it more accurately reflects the current circumstances.} A
+typical \acronym{SMTP} conversation resembles the following (the lines
+starting with a three digit number are sent by the server, all other lines
+are sent by the client):
 
@@
 phishing mails using faked sender addresses.  For example, a phishing mail
-might claim that the payroll system had a disastrous disk failure, and
+might pretend that the payroll system had a disastrous disk failure, and
 until the server is replaced all salary payments will have to be processed

### 2009-04-23 doc/logparser-complications.tex
Changes after running diction. Changed acknowledgements.

@@
 file; this specific instance is not a serious problem, because
-\parsername{} can not be expected to fully understand and reconstruct a
-mail when it is lacking some of that mail's log lines.
+\parsername{} cannot be expected to fully understand and reconstruct a mail
+when it is lacking some of that mail's log lines.
 
@@
         rejected a later one, and continued to accept delivery of the mail
-        for the first recipient.  Tn that case the timeout applies to the
+        for the first recipient.  In that case the timeout applies to the
         partially accepted mail, which should be discarded; however, this
@@
     \item Does the \texttt{timeout\_queueids} state table have an entry for
-        the queueid in the log line?  If there is an entry, it will be
+        the queueid in the log line?  If an entry is found it will be
         removed, regardless of whether the remaining criteria are
@@
         five minutes, but some data may have been transferred slowly, and
-        empirical evidence shows that ten minutes is not unreasonable;
-        hopefully it is a good compromise between false positives (log
-        lines incorrectly discarded) and false negatives (new state table
-        entries incorrectly created).
+        empirical evidence shows that ten minutes is not unreasonable; it
+        appears to be a good compromise between false positives (log lines
+        incorrectly discarded) and false negatives (new state table entries
+        incorrectly created).
 

### 2009-04-23 doc/logparser-conclusion.tex
Changes after running diction. Changed acknowledgements.

@@
 portion of a log file was correctly parsed.  Experience implementing
-\parsername{} shows that full input coverage is relatively easy to achieve
+\parsername{} shows that full input coverage is not difficult to achieve
 with this architecture, and that with enough time and effort a full

### 2009-04-23 doc/logparser-implementation.tex
Changes made based on Carl's comments.
Changes after running diction. Changed acknowledgements.

@@
 \parsername{} can parse log files from Postfix version 2.2 through to
-version 2.5, and should be capable of parsing log files from later versions
-with only minor modifications or an updated ruleset.  The parser's task is
-to follow the journey each mail takes through Postfix, combining the data
+version 2.5, and should be able to parse log files from later versions with
+only minor modifications or an updated ruleset.  The parser's task is to
+follow the journey each mail takes through Postfix, combining the data
 captured by rules into a coherent whole, saving it in a useful and
@@
 
-\parsername{} makes a small number of (hopefully safe and reasonable)
-assumptions:
+\parsername{} makes a few (hopefully safe and reasonable) assumptions:
 
@@
 \section{Framework}
+
+\label{framework in implementation}
 
@@
 
-Adding new actions is not as easy as adding new rules, though care has been
-taken in the architecture and implementation to make adding new actions as
-painless as possible; one of the few limitations is that \parsername{} is
-written in Perl, so new actions must also be written in Perl.  The
-implementer writes a subroutine that accepts the standard arguments given
-to actions, and registers it with the framework by calling the framework's
-\texttt{add\_actions()} subroutine.  The new action must be registered
-before the rules are loaded, because it is an error for a rule to specify
-an unregistered action; this helps catch mistakes made when adding new
-rules.  No other work is required from the implementer to integrate the
-action into the parser; all of their attention and effort can be focused on
-correctly implementing their new action.  The action may need to extend the
-list of valid combinations described in \sectionref{out of order log lines}
-if the new action creates a different set of acceptable programs, but this
-would only be necessary if the new action processes log lines from Postfix
-components that \parsername{} does not have rules for, e.g.\
-\daemon{virtual} or \daemon{lmtp}.
+Adding new actions is not as facile as adding new rules, though care has
+been taken in the architecture and implementation to make adding new
+actions as painless as possible; one of the few limitations is that
+\parsername{} is written in Perl, so new actions must also be written in
+Perl.  The implementer writes a subroutine that accepts the standard
+arguments given to actions, and registers it with the framework by calling
+the framework's \texttt{add\_actions()} subroutine.  The new action must be
+registered before the rules are loaded, because it is an error for a rule
+to specify an unregistered action; this helps catch mistakes made when
+adding new rules.  No other work is required from the implementer to
+integrate the action into the parser; all of their attention and effort can
+be focused on correctly implementing their new action.  The action may need
+to extend the list of valid combinations described in \sectionref{out of
+order log lines} if the new action creates a different set of acceptable
+programs, but this would only be necessary if the new action processes log
+lines from Postfix components that \parsername{} does not have rules for,
+e.g.\ \daemon{virtual} or \daemon{lmtp}.
 

### 2009-04-23 doc/logparser.tex
Corrections after proofreading chapters 6 & 7 and the appendices.
Changes after running diction. Changed acknowledgements.

@@
 
-\noindent{}I would like to express my appreciation to my supervisor, Dr.\
-Carl Vogel, for his advice, assistance, and guidance.
+\bigskip{}
+
+\noindent{}I am indebted to my supervisor, Dr.\ Carl Vogel, for his advice,
+assistance, and guidance.
 
@@
 
-\noindent{}I would like to thank my wife, Ariane Tobin, for her patience,
-good will, and encouragement.
+\noindent{}I am grateful to my wife, Ariane Tobin, for her patience, good
+will, and encouragement.
 
@@
 \glsadd{queueid}
+\glsresetall{}
 \renewcommand{\glossarypostamble}{\label{Glossary}}
@@
 \printglossary[type=\acronymtype,style=eqlist]{}
+\glsresetall{}
 \renewcommand{\glossarypostamble}{\label{Postfix Daemons}}

### 2009-04-24 doc/logparser-background.tex
Changes made after proofreading the preface and chapters 1-3.

@@
 Postfix-based anti-spam restrictions, generate statistics and graphs, and
-provide a platform on which new restrictions can be evaluated to determine
+provide a platform on which new restrictions can be developed and evaluated to determine
 if they are beneficial in the fight against spam.  The program written for
-this project, \parsername{}, parses Postfix log files and populates a
+this project, \parsername{}, parses Postfix log files and populates an
+\acronym{SQL}
 database with the data gleaned from those log files, providing a consistent
@@
 
-Determining which Postfix restrictions reject the highest number of
-delivery attempts is a short example of the optimisation possible using
-data from the database:
+Determining the ten Postfix restrictions that reject the highest number of
+delivery attempts is a short example of the analysis possible using data
+from the database:
 
@@
     WHERE action = 'DELIVERY_REJECTED'
-    ORDER BY hits_total DESC;
+    ORDER BY hits_total DESC
+    LIMIT 10;
 \end{verbatim}
@@
 
-Another example is determining which restrictions are not effective:
-\figureref{Sample SQL query showing ineffective restrictions} shows which
-restrictions rejected fewer than 100 delivery attempts in the last log file
-parsed, and the percentage of total rejections in that log file that each
-of those restrictions represents.
+Another example is determining which restrictions are not effective: the
+query in \figureref{Sample SQL query showing ineffective restrictions}
+shows which restrictions rejected fewer than 100 delivery attempts in the
+last log file parsed, and the percentage of total rejections in that log
+file that each of those restrictions represents.
 
@@
 The sample database queries yield summary statistics about the efficiency of
-anti-spam techniques.  These statistics are far less feasible to assess
-directly from log files without prior pre-processing into a database in the
-fashion proposed, implemented, and tested herein.
+anti-spam techniques.  Analysis of this kind is much easier to perform when
+the data source is a database that when trying to directly analyse log
+files.
 
@@
 \SMTPglossaryDescription{} Despite the simplicity of the protocol, many
-virus and spam sending programs fail to implement it properly, so requiring
-strict adherence to the protocol specification is beneficial in protecting
-against spam and viruses.\footnote{Originally all mail servers adhered to
-the principle of \textit{Be liberal in what you accept, and conservative in
-what you send\/} from \acronym{RFC}~760~\cite{rfc760}, but unfortunately
-that principle was written in a friendlier time.  Given the deluge of spam
-that mail servers are subjected to daily, a more appropriate maxim could
-be: \textit{Require strict adherence to relevant \acronyms{RFC}; implement
-the strongest restrictions you can; relax the restrictions and adherence
-only when legitimate mail is impeded.\/}  It is neither as friendly nor as
-catchy, but it more accurately reflects the current circumstances.} A
-typical \acronym{SMTP} conversation resembles the following (the lines
-starting with a three digit number are sent by the server, all other lines
-are sent by the client):
+viruses and spam sending programs fail to implement it properly, so
+requiring strict adherence to the protocol specification is beneficial in
+protecting against spam and viruses.\footnote{Originally all mail servers
+adhered to the principle of \textit{Be liberal in what you accept, and
+conservative in what you send\/} from \acronym{RFC}~760~\cite{rfc760}, but
+unfortunately that principle was written in a friendlier time.  Given the
+deluge of spam that mail servers are subjected to daily, a more appropriate
+maxim could be: \textit{Require strict adherence to relevant
+\acronyms{RFC}; implement the strongest restrictions you can; relax the
+restrictions and adherence only when legitimate mail is impeded.\/}  It is
+neither as friendly nor as catchy, but it more accurately reflects the
+current circumstances.} A typical \acronym{SMTP} conversation resembles the
+following (the lines starting with a three digit number are sent by the
+server, all other lines are sent by the client):
 
@@
 the client in the \acronym{SMTP} conversation; it is required by the
-protocol, but its original purpose was to detect a server trying to deliver
-mail to itself, so some implementations omitted it.}  Given the volume of
-spam sent daily, this liberal approach is no longer viable.
+protocol, but its original purpose was to prevent mail loops by detecting a
+server trying to deliver mail to itself, so some implementations omitted
+it.}  Given the volume of spam sent daily, this liberal approach is no
+longer viable.
 
@@
 
-Postfix is a \glsfirst{MTA} with the following design aims (in order of
+Postfix is a \acronym{MTA} with the following design aims (in order of
 importance): security, flexibility of configuration, scalability, and high
@@
 supply their own rejection messages to make it clear to senders why their
-mail was rejected.  Policy servers (\sectionref{policy servers}) provide a
-simple way to write new restrictions without having to modify Postfix's
-source code.  Unfortunately this flexibility has a cost: the complexity
-Postfix's log files.  Although it is usually relatively simple to use
-standard Unix text processing utilities to determine the fate of an
-individual mail, sometimes it can be quite difficult.  For most mails the
-journey is simple and brief, but the remainder can be quite complex (see
-\sectionref{complications} for details).
+mail was rejected.  Policy servers (soon to be described in
+\sectionref{policy servers}) provide a simple way to write new restrictions
+without having to modify Postfix's source code.  Unfortunately this
+flexibility has a cost: the complexity of Postfix's log files.  Although it
+is usually relatively simple to use standard Unix text processing utilities
+to determine the fate of an individual mail, sometimes it can be quite
+difficult.  For most mails the journey through Postfix is simple and brief,
+but the remainder can be quite complex (see \sectionref{complications} for
+details).
 
@@
 separated into multiple components that each perform one of the tasks
-required of an \acronym{MTA}, e.g.\ receive mail, send mail, local delivery
-of mail;\ full details can be found in~\cite{postfix-overview}.  Postfix's
-design is strongly influenced by security concerns: those components that
-interact with other hosts are not privileged,\footnote{Privilege means the
-power to perform actions that are limited to the administrator, and not
-available to ordinary users.} so bugs in those components will not give an
-attacker extra privileges; those components that are privileged do not
-interact with other hosts, making it much more difficult for an attacker to
-exploit any bugs that may exist in those components.
+required of an \acronym{MTA}, e.g.\ receive mail, send mail, deliver mail
+to a user's mailbox;\ full details can be found in~\cite{postfix-overview}.
+Postfix's design is strongly influenced by security concerns: those
+components that interact with other hosts are not
+privileged,\footnote{Privilege means the power to perform actions that are
+limited to the administrator, and are not available to ordinary users.} so
+bugs in those components will not give an attacker extra privileges; those
+components that are privileged do not interact with other hosts, making it
+much more difficult for an attacker to exploit any bugs that may exist in
+those components.
 
@@
 the connected client is found in the specified table and returns the
-associated action if found; the method of searching the file is dependant
+associated result if found; the method of searching the file is dependant
 on the type of the lookup table~\cite{postfix-lookup-tables}.  Other
@@
 an exception is made for the web server in the
-\texttt{check\_client\_access} check, and other local machines are
-rejected, because they must be unauthenticated to reach that point in the
-restriction list.
+\texttt{check\_client\_access} check, and then other local machines are
+rejected, because to continue past \texttt{permit\_sasl\_authenticated} the
+client must not have authenticated.
 
@@
 
-That example also shows that it is trivial to supply a custom rejection
-message.  When the result is the name of another restriction Postfix will
-evaluate that restriction, allowing restrictions to be chosen based on the
-client IP address, client hostname, HELO hostname, sender address, or
-recipient address.  For example, the administrator may require that all
-clients on the local network have valid DNS entries, to prevent people
-sending mail from unknown machines; one example of how to achieve this is:
+That example also shows how to supply a custom rejection message.  When the
+result of a lookup is the name of another restriction Postfix will evaluate
+that restriction; this allows restrictions to be chosen based on the client
+IP address, client hostname, HELO hostname, sender address, or recipient
+address.  For example, the administrator may require that all clients on
+the local network have valid DNS entries, to prevent people sending mail
+from unknown machines; one example of how to achieve this is:
 
@@
 of lookups, restrictions, and exceptions.  Postfix tries to protect the
-administrator from misconfiguration in as far as is reasonable, e.g.\ the
-restriction \texttt{check\_helo\_mx\_access} cannot cause a mail to be
+administrator from misconfiguration in as far as it reasonably can, e.g.\
+the restriction \texttt{check\_helo\_mx\_access} cannot cause a mail to be
 accepted, because the parameter it checks (the hostname given in the HELO
@@
 \sectionref{Mixing and matching Postfix restrictions}.  A policy server can
-perform more complex checks than those provided by Postfix: a trivial
-example is allowing addresses associated with the payroll system to send
-mail on the third Tuesday after pay day only, to help prevent problems from
-phishing mails using faked sender addresses.  For example, a phishing mail
-might pretend that the payroll system had a disastrous disk failure, and
-until the server is replaced all salary payments will have to be processed
+perform more complex checks than those provided by Postfix, such as
+allowing addresses associated with the payroll system to send mail on the
+third Tuesday after pay day only, to help prevent problems from phishing
+mails using faked sender addresses.  For example, a phishing mail might
+pretend that the payroll system had a disastrous disk failure, and until
+the server is replaced all salary payments will have to be processed
 manually, so please reply to this mail with your name, address, and bank
@@
 
-    \item \acronym{SPF} checks:
-        \urlLastChecked{http://www.openspf.org/}{2009/04/21}.
+    \item Checking if the client satisfies a domain's \acronym{SPF}
+        records, \urlLastChecked{http://www.openspf.org/}{2009/04/21}.
         \acronym{SPF} records specify which mail servers are allowed to
@@
         intention is to reduce spam from faked sender addresses,
-        backscatter~\cite{postfix-backscatter}, and
-        Joe~jobs~\glsadd{Joe-job}.  There has been considerable resistance
-        to \acronym{SPF} because it breaks or vastly complicates some
-        commonly-used features of \acronym{SMTP}, e.g.\ forwarding mail
-        from one domain to another when a user moves.
+        backscatter, and joe~jobs~\glsadd{Joe-job}.  There has been
+        considerable resistance to \acronym{SPF} because it breaks or
+        vastly complicates some commonly-used features of \acronym{SMTP},
+        e.g.\ forwarding mail from one domain to another when a user moves.
 
@@
         technique that temporarily rejects a delivery attempt when the
-        triple of \newline{} \tab{}\texttt{(sender address, recipient
+        tuple of \newline{} \tab{}\texttt{(sender address, recipient
         address, remote IP address)}\newline{} has not been seen before; on
-        second and subsequent delivery attempts from that triple the mail
+        second and subsequent delivery attempts from that tuple the mail
         will be accepted.  This blocks spam from some senders because
@@
         \numberOFlogFILES{} log files used when generating the results in
-        \sectionref{Evaluation} are shown in \tableref{Most effective
-        Postfix restrictions}.  The table shows that greylisting is worth
-        using at the moment, particularly when you take into account its
-        position as the final restriction that a mail must overcome in the
+        \sectionref{Evaluation} are shown in \tableref{Sample output from
+        SQL query}.  That table shows that greylisting is worth using at
+        the moment, particularly when you take into account its position as
+        the final restriction that a mail must overcome in the
         configuration used on the mail server that generated the log files:
@@
         when coming from a pool of servers.
-
-        \showtable{build/include-restriction-summary-table}{Most effective
-        Postfix restrictions}{Most effective Postfix restrictions}
 
@@
         a threshold that requires a delivery attempt to hit several scored
-        restrictions will allow mail from clients that would be rejected by
-        a binary restriction.
+        restrictions will allow delivery attempt that would be rejected by
+        a boolean restriction.
 
@@
 
-This section has provided background information useful in understanding
+This chapter has provided background information useful in understanding
 this thesis, starting with the motivation behind the project, continuing

### 2009-04-24 doc/logparser-implementation.tex
Changes made after proofreading the preface and chapters 1-3.

@@
-\chapter{Postfix Parser Implementation}
+\chapter{Postfix Log Parser}
 
@@
 
-\subsection{Actions In The \parsernamelong{}}
+\subsection{Description Of Each Action}
 
@@
 
-\subsection{Example Rule}
+\subsection{Sample Rule}
 
@@
 
-The example rule in \tableref{Example rule in implementation table}
+The sample rule in \tableref{Example rule in implementation table}
 recognises the log line that results from Postfix rejecting a delivery
@@
 \begin{table}[thbp]
-    \caption{Example of a rule used by \parsernameshort{}}
+    \caption{Sample rule used by \parsernameshort{}}
     \empty{}\label{Example rule in implementation table}

### 2009-04-24 doc/logparser-introduction.tex
Changes made after proofreading the preface and chapters 1-3.

@@
 source code to generate statistics: this approach improves the chances of
-other Postfix users testing and using the software developed for this
-project.  The need arose for a parser capable of dealing with the great
-number and variety of log lines produced by Postfix: the parser must be
-designed and implemented so that adding support for parsing new inputs is a
-simple task, because the log lines to be parsed will change over time.
-This variety in log lines occurs for several reasons:
+other Postfix users testing, using, and benefiting from the software
+developed for this project.  The need arose for a parser capable of dealing
+with the great number and variety of log lines produced by Postfix: the
+parser must be designed and implemented so that adding support for parsing
+new inputs is a simple task, because the log lines to be parsed will change
+over time.  This variety in log lines occurs for several reasons:
 
@@
 in-depth parsing and analysis of their log files, where the parsing must
-correlate the log lines per-connection or per-queueid rather than
-processing log lines independently.  One of the barriers to this kind of
+correlate each connection or mail's log lines rather than
+processing each log line in isolation.  One of the barriers to this kind of
 processing is the unstructured nature of Postfix log files, where each log
@@
 This architecture is the basis of \parsername{}, a program that parses
-Postfix log files and places the resulting data into a database for later
-analysis.  The gathered data can be used to optimise current anti-spam
-defences, to produce statistics showing how effective each technique in use
-is, or to provide a baseline to test new anti-spam measures against.
-Numerous other uses are possible for such data: improving server
-performance by identifying troublesome destinations and reconfiguring
-appropriately; identifying regular high volume uses (e.g.\ customer
-newsletters) and restricting those uses to off-peak times, or providing a
-dedicated service for them; detecting virus outbreaks that propagate via
-mail; as a base for billing customers on a shared server.  Preserving the
+Postfix log files and places the resulting data into a \acronym{SQL}
+database for later analysis.  The gathered data can be used to optimise
+current anti-spam defences, to produce statistics showing how effective
+each technique in use is, or to provide a baseline to test new anti-spam
+measures against.  Numerous other uses are possible for such data:
+improving server performance by identifying troublesome destinations and
+reconfiguring appropriately; identifying regular high volume uses (e.g.\
+customer newsletters) and restricting those uses to off-peak times or
+providing a dedicated service for them; detecting virus outbreaks that
+propagate via mail; billing customers on a shared server.  Preserving the
 raw data enables a multitude of uses far beyond those conceived of by the
@@
 
-Chapter~\ref{Evaluation} evaluates \parsernames{} efficiency, discussing
-the various optimisations implemented in the parser, and the effect they
+Chapter~\ref{Evaluation} evaluates \parsernames{} efficiency, exploring
+the various optimisations implemented in the parser and the effect they
 have.  It also discusses the coverage achieved by \parsername{} over
@@
 
-Appendix~\ref{Postfix Daemons} provides a brief description of Postfix
-daemons.
+Appendix~\ref{Postfix Daemons} provides a brief description of each Postfix
+daemon.
 

### 2009-04-24 doc/logparser-other-parsers-review.tex
Changes made after proofreading the preface and chapters 1-3.

@@
 
-At the start of this project ten Postfix log file parsers were tested, with
+Ten Postfix log file parsers were tested at the start of this project, with
 the hope of finding a suitable parser to build upon, rather than writing
-one from scratch.  It was quite difficult to find ten parsers to review for
-this project, and the functionality offered by those parsers ranges from
-quite basic to much more mature, depending on the needs of the author of
-the parser.
+one from scratch.  It was quite difficult to find ten parsers to review,
+and the functionality offered by those parsers ranges from quite basic to
+more developed, depending on the needs of the parser's author.
 
@@
 reviewed~\cite{log-mail-analyser} is the only previously published research
-in this area that the author is aware of; that research aims to show that
+in this area that the author could find; that research aims to show that
 providing the data from log files in a more accessible form is helpful to
@@
 
-The ten parsers have been reviewed and compared with \parsername{}, this
+The ten parsers have been compared and contrasted with \parsername{}, this
 project's parser, to show how much effort would have been required to use
@@
         extensive and intrusive modifications to the parser; the
-        architecture that enables this is documented in \sectionref{parser
+        architecture enabling this is documented in \sectionref{parser
         architecture}.
@@
     \item The reviewed parsers all produce a report of varying complexity
-        and detail, whereas \parsername{} does not; it extracts data and
+        and detail, whereas \parsername{} does not: it extracts data and
         leaves generation of reports to other programs.  Using a
@@
         for this project is designed to enable much more detailed log file
-        analysis by providing a stable platform for subsequent programs to
-        build upon.
+        analysis than other parsers by providing a stable platform for
+        subsequent programs to build upon.
 
@@
 The data extracted by \parsername{} is documented in
-\sectionref{connections table} and \sectionref{results table}; for
-convenience that list is repeated here: client and server IP address and
+\sectionref{connections table} and \sectionref{results table}, but for
+convenience the list is repeated here: client and server IP address and
 hostname, HELO hostname, queueid, start time, end time, \acronym{SMTP}
@@
 \acronym{pid} (\sectionref{queueid vs pid}), and it stores each mail's
-queueid, but it does not store each connection's \acronyms{pid}, because a
+queueid, but it does not store each connection's \acronym{pid}, because a
 use has not been identified for that data.
@@
 some documentation is available in~\cite{log-mail-analyser}.  Studying the
-source code is informative, though this author had difficulty as the
-authors of \acronym{LMA} wrote in Italian.
+source code is informative, though this author had some difficulty because
+the authors of \acronym{LMA} wrote in Italian.
 
-\acronym{CSV} is a very simple format where each record is stored in a
-single line, with fields separated by a comma or other punctuation symbol.
-Problems with \acronym{CSV} files include the need to escape separators in
-the data stored, providing multiple values for a field (e.g.\ multiple
-recipients), and adding new fields.  \acronym{CSV} files do not have a
-standard mechanism to document the fields or the separator, unlike
-\acronym{SQL} where every database includes a schema naming the fields and
-the type of data they store (e.g.\ integer, text, timestamp).  The
-\acronym{CSV} record format used by \acronym{LMA} is not documented, but
-the output file contains a comment in Italian giving the format:\newline{}
-\texttt{\# Timestamp|Nome Client|IP Client|IP Server|From|To|Status|Size}
+\begin{description}
+
+    \item [CSV] \acronym{CSV} is a very simple format where each record is
+        stored in a single line, with fields separated by a comma or other
+        punctuation symbol.  Problems with \acronym{CSV} include the need
+        to escape separators in the data stored, providing multiple values
+        for a field (e.g.\ multiple recipients), and adding new fields.
+        \acronym{CSV} does not have a standard mechanism to document the
+        fields or the separator, unlike \acronym{SQL} where every database
+        includes a schema naming the fields and the type of data they store
+        (e.g.\ integer, text, timestamp).  The \acronym{CSV} record format
+        used by \acronym{LMA} is not documented
+        in~\cite{log-mail-analyser}, but the output file contains a comment
+        in Italian giving the format:\newline{} \tab{} \texttt{\#
+        Timestamp|Nome Client|IP Client|IP
+        Server|From|To\newline{}\tab{}\tab{}|Status|Size}
         \newline{}\acronym{LMA} treats \acronym{CSV} lines starting with
@@
 
-Berkeley DB only supports storing simple \textbf{(key, value)} pairs,
-unlike \acronym{SQL} databases that store arbitrary tuples.  In
-\acronym{LMA}'s main table the key is an integer referred to by secondary
-tables, and the value is a \acronym{CSV} line containing all of the data
-for that row.  The secondary by-sender, by-recipient, by-date, and by-IP
-tables use the sender/recipient/date/IP address as the key, and the value
-is a \acronym{CSV} list of integers referring to the keys in the main
-table.  This effectively re-implements \acronym{SQL} foreign keys, but
-without the functionality offered by even the most basic of \acronym{SQL}
-databases, e.g.\ joins, ordering, searches.  It also requires custom code
-to search on a combination of attributes, though the authors of
-\acronym{LMA} did provide some simple reports: IP-STORY, FROM-STORY,
-DAILY-EMAIL, and DAILY-REJECT (all described later).  Berkeley DB appears
-to be the least useful of the three output formats: it does not provide the
-functionality of a basic \acronym{SQL} database, and unlike \acronym{CSV}
-files it cannot be used with standard Unix text processing tools.
+    \item [Berkeley DB] Berkeley DB only supports storing simple
+        \textbf{(key, value)} pairs, unlike \acronym{SQL} databases that
+        store arbitrary tuples.  In \acronym{LMA}'s main table the key is
+        an integer referred to by secondary tables, and the value is a
+        \acronym{CSV} line containing all of the data for that row.  The
+        secondary by-sender, by-recipient, by-date, and by-IP tables use
+        the sender/recipient/date/IP address as the key, and the value is a
+        \acronym{CSV} list of integers referring to the keys in the main
+        table.  This effectively re-implements \acronym{SQL} foreign keys,
+        but without the functionality offered by even the most basic of
+        \acronym{SQL} databases, e.g.\ joins, ordering, searches.  It also
+        requires custom code to search on a combination of attributes, and
+        the authors of \acronym{LMA} did provide some simple reports:
+        IP-STORY, FROM-STORY, DAILY-EMAIL, and DAILY-REJECT (all described
+        later).  Berkeley DB appears to be the least useful of the three
+        output formats: it does not provide the functionality of a basic
+        \acronym{SQL} database, and unlike \acronym{CSV} files it cannot be
+        used with standard Unix text processing tools.
 
-The schema used with the MySQL database is undocumented, but at least it is
-possible to discover the schema with an existing \acronym{SQL} database,
-unlike with Berkeley DB\@; all \acronym{SQL} databases embed the schema
-into the database and provide commands for displaying it.  Berkeley DB does
-not embed a schema in its files, because it does not need one, and
-embedding one would not provide any benefit; it only provides \textbf{(key,
-value)} pairs, so any additional structuring of the data is imposed by the
-application, and it is the responsibility of the author of the application
-to document this structure.  MySQL support was not tested because the
-schema is not documented, and no existing database was available to extract
-the schema from, so the required tables could not be created.
+    \item [MySQL] The schema used with the MySQL database is undocumented,
+        but at least it is possible to discover the schema if one has an
+        existing \acronym{SQL} database, unlike with Berkeley DB\@; all
+        \acronym{SQL} databases embed the schema into the database and
+        provide commands for displaying it.  Berkeley DB does not embed a
+        schema in its database files, because it does not need one, and
+        embedding one would not provide any benefit; it only provides
+        \textbf{(key, value)} pairs, so any additional structuring of the
+        data is imposed by the application, and it is the responsibility of
+        the author of the application to document this structure.
+        \acronym['s]{LMA} MySQL support was not tested because the schema
+        is not documented, and no existing database was available to
+        extract the schema from, so the required database could not be
+        created.
 
-Whether a MySQL database or Berkeley DB table is chosen in addition to the
+\end{description}
+
+Whether a MySQL or Berkeley DB database is chosen in addition to the
 \acronym{CSV} output, \acronym{LMA} stores the following data: time and
@@
 HELO hostname, queueid, start and end times, timestamps for each log line,
-enhanced status code, delivery delays, or message id (for accepted mails
+enhanced status code, delivery delays, or message-id (for accepted mails
 only).  Handling of multiple recipients, \acronym{SMTP} codes, or remote
@@
 multiple servers for one or more recipients.} is not explained;
-experimental observation shows that multiple records are added when a mail
+experimental observation shows that multiple records were added when a mail
 had multiple recipients, but the records are not associated or linked in
@@
 \acronym{LMA} does not deal with any of the complications discussed in
-\sectionref{complications}, except for correlating log lines by queueid;
-not correlating log lines by \acronym{pid} means it cannot correlate most
-rejected delivery attempts.  It does not differentiate between different
-types of rejections, so it is not suitable for the purposes of this
-project; the data about which restriction caused the rejection is
+\sectionref{complications}, except for correlating log lines by queueid; it
+cannot correlate most rejected delivery attempts because it does not
+correlate log lines by \acronym{pid}.  It does not differentiate between
+different types of rejections, so it is not suitable for the purposes of
+this project; the data about which restriction caused the rejection is
 discarded, whereas the main goal of this project is to retain that
@@
         suppresses duplicate sequential lines.  DAILY-REJECT produces the
-        correct number of output lines, probably because it uses the
-        Berkeley DB files as input rather than the \acronym{CSV} file.
+        correct number of output lines, probably because it searches the
+        Berkeley DB database rather than the \acronym{CSV} file.
 
@@
     \item Some records output by DAILY-REJECT are truncated: they are
-        missing the last $|$ that separates fields and also the newline
+        missing the last ``$|$'' that separates fields and also the newline
         following it, so the line containing only a single space is
@@
 to produce its report: the HELO hostname, queueid, start and end times,
-timestamps for each log line, or message id.  Both the parsing and
+timestamps for each log line, or message-id.  Both the parsing and
 reporting are difficult to extend because it is a specialised tool, unlike
@@
     reports, all through a web interface.  Sawmill can perform Postfix
-    analysis on any platform, including Window~[sic], Linux, FreeBSD,
-    OpenBSD, Mac OS, Solaris, other UNIX, and more.
+    analysis on any platform, including Window, Linux, FreeBSD, OpenBSD,
+    Mac OS, Solaris, other UNIX, and more.
 }
@@
         \urlLastChecked[\hfill{}\newline{}]{http://www.thesawmill.co.uk/formats/postfix.html}{2008/11/23}.
-        Fields extracted: from, to, server, UID, relay, status, number of
-        recipients, origin hostname, origin IP address, and virus.  It also
-        counts the number of and total size of all mails delivered.  The
-        fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
+        Fields extracted: from, to, server, uid, relay, status, number of
+        recipients, origin hostname, origin IP address, and virus.  It
+        also counts the number of and total size of all mails delivered.
+        The fields \texttt{server}, \texttt{uid}, \texttt{relay}, and
         \texttt{virus} are not explained in the documentation:
@@
         \texttt{uid} might be the uid of the user submitting mail locally.
-        Postfix does not do any form of virus checking (though it has many
-        options for cooperating with an external virus scanner), so the
+        Postfix does not do any form of virus checking itself, so the
         \texttt{virus} field is a mystery.
@@
         relay hostname, relay IP address, status, response code, RBL list,
-        and message id.  It also counts the number and size of all mails
+        and message-id.  It also counts the number and size of all mails
         delivered, processed, blocked, expired, and bounced.
@@
 When tested with the \numberOFlogFILES{} test log files it performed
-adequately, though the rate it processed log files at did slow down
+adequately, though the rate it processed log files at did slowdown
 noticeably as it progressed.  Sawmill supports reading compressed log files
@@
 still unable to index all the log files in a directory using the web
-interface.  Using Splunk's \acronym{CLI} rather than the web interface was
-more successful: the command ``\texttt{splunk find logs
+interface.  Using Splunk's command line interface rather than the web
+interface was more successful: the command ``\texttt{splunk find logs
 }\textit{log-directory\/}'' added 40 of the \numberOFlogFILES{} log files
@@
 Subsequently the author was successful in adding a single log file at a
-time using the command ``\texttt{splunk add tail }\textit{filename\/}''.  A
-simple loop using that command was enough to add all the desired log files.
-Splunk will periodically check all indexed log files for updates unless
-they are manually removed from its list; this may or may not be useful
-behaviour.  Splunk did not appear to have any difficulty in indexing the
-log files once they had been successfully added to its queue.
+time using the command ``\texttt{splunk add tail }\textit{filename\/}'',
+and then a simple loop using that command was enough to add all the desired
+log files.  Splunk will periodically check all indexed log files for
+updates unless they are manually removed from its list; this may or may not
+be useful behaviour.  Splunk did not appear to have any difficulty in
+indexing the log files once they had been successfully added to its queue.
 \parsername{} parses the log files it is instructed to parse, in the order
@@
 Splunk's sales team must be asked for a quote.  Median log file size for
-the log files used when evaluating \parsername{} is
+the \numberOFlogFILES{} log files used when evaluating \parsername{} is
 \input{build/include-median-file-size}.
@@
 \acronym{LMTP}, or \acronym{ESMTP}).  \parsername{} extracts noticeably
-more data (client and server IP address and hostname, queueid, start and
+more data: client and server IP address and hostname, queueid, start and
 end times, timestamps for each log line, \acronym{SMTP} and enhanced status
-codes, delivery delays, and message-id), though it does not make the full
+codes, delivery delays, and message-id.  \parsername{} does not make the full
 text of the log line available; a few minutes work could add this to
@@
 capabilities to generate reports, by generating input for AWStats from the
-database populated by \parsername{}.
+database populated by \parsername{}, but the author has not attempted that.
 
@@
 data that \parsername{} does not: how many times delivery was deferred for
-each mail; this information can be determined from the database populated
+each mail; this information can be calculated from the database populated
 by \parsername{} if desired.  Unlike \parsername{}, YAALA does not maintain
@@
 deal with the complications explained in \sectionref{complications}, except
-that it does correlate log lines by queueid.
+that it does correlate log lines by queueid (but not by \acronym{pid}).
 
@@
 
-Testing YAALA was unsuccessful because all the select clauses tried
-produced a similar error message:
+Testing YAALA was unsuccessful because all the queries produced a similar
+error message:
 \newline{}\tab{}\texttt{lib/Yaala/Data/Core.pm: Unavailable aggregation
@@
 \texttt{<key>=<value>} pairs in each log line, extracts them, and
-correlates the data by queueid.  This approach will extract the following
-data: HELO hostname, queueid, \acronym{SMTP} code, sender and recipient
-addresses, and size of accepted mails.  Lire misses the following fields
-extracted by \parsername{}: client and server IP address and hostname,
-start and end times, enhanced status code, delivery delays, timestamps of
-each log line, and message-id\@.
+correlates the data by queueid (but not by \acronym{pid}).  This approach
+will extract the following data: HELO hostname, queueid, \acronym{SMTP}
+code, sender and recipient addresses, and size of accepted mails.  Lire
+misses the following fields extracted by \parsername{}: client and server
+IP address and hostname, start and end times, enhanced status code,
+delivery delays, timestamps of each log line, and message-id\@.
 
@@
 reports.  Lire supports saving extracted data for later report generation,
-but the format of this data store is undocumented; given the source code it
-should be possible, with enough time and effort, to understand the format.
-\parsername{} uses an \acronym{SQL} database to make accessing the
-extracted data as effortless as possible.  In general, Lire has poor
-documentation.
+but the format of this data store is undocumented.  \parsername{} uses an
+\acronym{SQL} database to make accessing the extracted data as effortless
+as possible.  In general, Lire has poor documentation.
 
 Similar to AWStats and Logrep, Lire attempts to correlate log lines by
-queueid, but not by \acronym{pid}, so the complete list of recipients for a
-mail should be available; its parser extracts only part of the available
-data and does not attempt to deal with the other complications described in
-\sectionref{complications}.  When testing Lire on the \numberOFlogFILES{}
-test log files it performed reasonably well: the numbers it reports appear
-accurate, and the subset verified by the author were correct.  Its report
-provides summaries of:
+queueid, but not by \acronym{pid}, so the complete list of recipients for
+each delivered mail should be available; it does not attempt to deal with
+the other complications described in \sectionref{complications}.  When
+testing Lire on the \numberOFlogFILES{} test log files it performed
+reasonably well: the numbers it reports appear accurate, and the subset
+verified by the author were correct.  Its report provides summaries of:
 
@@
 not make the extracted data easily available to other programs.  Its parser
-is in-extensible; it could easily be replaced, but that would require
-writing a parser from scratch, so would not be any less work.
-\parsername{} could possibly be used to parse Postfix log files for Lire,
-but the difficulty may outweigh the benefits, e.g.\ Lire's data store may
-not be capable of storing the data extracted by \parsername{}, but the lack
-of documentation hinders any evaluation.  As with YAALA this approach has
-not been attempted by the author.
+is not extensible; it could easily be replaced, but that would require
+writing a parser from scratch, so would not be any less work than writing
+\parsername{}.  \parsername{} could possibly be used to parse Postfix log
+files for Lire, but the difficulty may outweigh the benefits, e.g.\ Lire's
+data store may not be capable of storing the data extracted by
+\parsername{}, but the lack of documentation hinders any evaluation.  As
+with YAALA this approach has not been attempted by the author.
 
@@
 automated processing is a key part of \parsernames{} design.  Sadly, all
-the author's attempts to use Logrep failed, as it was unable to access the
-log files selected for parsing; this appears to be a bug rather than
+the author's attempts to use Logrep failed, because it was unable to access
+the log files selected for parsing; this appears to be a bug rather than
 operator error.  If the problem was caused by operator error, the interface
@@
         \item [\nichi{}] \parsername{} defers report generation to
-            subsequent programs, but all the necessary data and
-            documentation to produce reports is provided.
+            subsequent programs, but all the necessary documentation to
+            produce reports is provided.
 

### 2009-04-24 doc/logparser-results.tex
Changes made after proofreading the preface and chapters 1-3.

@@
 
-\subsection{Characteristics Of The Input Log Files}
+\subsection{Characteristics Of The \numberOFlogFILES{} Log Files}
 
@@
 \showgraph{build/graph-average-number-of-rules-tried-per-program}{Mean
-number of rules used per log line for each Postfix component over all log
-files}{Mean number of rules used per log line for each Postfix component}
+number of rules used per log line for each Postfix component over
+\numberOFlogFILES{} log files}{Mean number of rules used per log line for
+each Postfix component}
 
@@
 
-\Graphref{Increase in parsing time when not caching compiled regexes graph}
-shows the effect that not caching compiled regexes has on parser
-performance, with mean and standard deviation in \tableref{Increase in
-parsing time when not caching compiled regexes table}.  For typical log
-files, the mean increase in parsing time when not caching compiled regexes
-is \input{build/include-cached-regexes-vs-discarded-regexes-mean}.  Caching
+\Graphref{Percentage increase in parsing time when not caching compiled
+regexes graph} shows the effect that not caching compiled regexes has on
+parser performance, with mean and standard deviation in
+\tableref{Percentage increase in parsing time when not caching compiled
+regexes table}.  For typical log files, the mean increase in parsing time
+when not caching compiled regexes is
+\input{build/include-cached-regexes-vs-discarded-regexes-mean}.  Caching
 compiled regexes is probably the single most effective optimisation
@@
 
-\showgraph{build/graph-cached-regexes-vs-discarded-regexes}{Increase in
-parsing time when not caching compiled regexes}{Increase in parsing time
-when not caching compiled regexes graph}
+\showgraph{build/graph-cached-regexes-vs-discarded-regexes}{Percentage
+increase in parsing time when not caching compiled regexes}{Percentage
+increase in parsing time when not caching compiled regexes graph}
 
-\showtable{build/include-cached-regexes-vs-discarded-regexes}{Increase in
-parsing time when not caching compiled regexes}{Increase in parsing time
-when not caching compiled regexes table}
+\showtable{build/include-cached-regexes-vs-discarded-regexes}{Percentage
+increase in parsing time when not caching compiled regexes}{Percentage
+increase in parsing time when not caching compiled regexes table}
 

### 2009-04-24 doc/logparser.tex
Changes made after proofreading the preface and chapters 1-3.

@@
     documenting the architecture developed for this project and the parser
-    that implements it, explaining the difficulties encountered and the
-    solutions developed, and evaluating the parser on the criteria of
-    efficiency and Postfix log file coverage.  The parser stores data
-    gleaned from the log files in an SQL database; the gathered data can be
-    used to optimise current anti-spam measures, to produce statistics
-    showing how effective those measures are, or to provide a baseline to
-    test new anti-spam measures against.  The Postfix log file parser needs
-    to be very precise and strict when parsing, yet must allow users to
-    easily adapt or extend it to parse new log lines, without requiring
-    that the user have an in-depth knowledge and understanding of the
-    parser's internal workings.  The newly developed architecture is
-    designed to make the process of parsing new inputs as simple as
-    possible, enabling users to trivially add new rules (to parse variants
-    of known inputs) and relatively easily add new actions (to process a
-    previously unknown category of inputs).  The parser implemented for
-    this project is evaluated on the criteria of efficiency and coverage of
-    Postfix log files, demonstrating that the conflicting goals of
-    efficiency and accuracy can be balanced, and one need not be sacrificed
-    to achieve the other.
+    that implements it, the difficulties encountered and the solutions
+    developed.  The parser stores data gleaned from the log files in an SQL
+    database; the gathered data can be used to optimise current anti-spam
+    measures, to produce statistics showing how effective those measures
+    are, or to provide a baseline to test new anti-spam measures against.
+    The Postfix log file parser needs to be very precise and strict when
+    parsing, yet must allow users to easily adapt or extend it to parse new
+    log lines, without requiring that the user have an in-depth knowledge
+    and understanding of the parser's internal workings.  The newly
+    developed architecture is designed to make the process of parsing new
+    inputs as simple as possible, enabling users to trivially add new rules
+    (to parse variants of known inputs) and relatively easily add new
+    actions (to process a previously unknown category of inputs).  The
+    parser implemented for this project is evaluated on the criteria of
+    efficiency and coverage of Postfix log files, demonstrating that the
+    conflicting goals of efficiency and accuracy can be balanced, and that
+    one need not be sacrificed to achieve the other.
 

### 2009-04-26 doc/logparser-architecture.tex
Changes based on proofreading chapters 4 and 5.

@@
 To avoid cluttering the description of the parser architecture with the
-details involved in implementing a parser for Postfix log files, each topic
-has been given its own chapter.  This chapter presents the architecture
-developed for this project, beginning with the overall architecture and
-design, followed by detailed documentation of the three components of the
-architecture: Framework, Actions, and Rules.  This chapter centers on the
-theoretical, implementation-independent aspects of the architecture; the
-practical difficulties of writing a parser for Postfix log files are
+details of implementing a parser for Postfix log files, each topic has been
+given its own chapter.  This chapter presents the architecture developed
+for this project, beginning with the overall architecture and design,
+followed by detailed documentation of the three components of the
+architecture: Framework, Actions, and Rules.  This chapter is focused on
+the theoretical, implementation-independent aspects of the architecture;
+the practical difficulties of writing a parser for Postfix log files are
 covered in detail in \sectionref{Postfix Parser Implementation}.
@@
 It should be clear from the earlier Postfix background (\sectionref{postfix
-background}) that log files produced by Postfix may vary widely from host
-to host.  With this in mind, one of the architecture's design aims was to
-make parsing new inputs as effortless as possible, to enable administrators
-to properly parse their own log files.  The solution developed is to divide
+background}) that Postfix log files may vary widely from host to host.
+With this in mind, one of the architecture's design aims was to make
+parsing new inputs as effortless as possible, to enable administrators to
+properly parse their own log files.  The solution developed is to divide
 the architecture into three parts: Framework, Actions, and Rules.  Each
-will be discussed separately, but first an overview:
+will be documented separately, but first an overview:
 
@@
     \item [Actions] Each action performs the processing required for a
-        single \textit{category\/} of inputs, e.g.\ rejecting a delivery
+        single \textit{category\/} of inputs, e.g.\ rejection of a delivery
         attempt.  Actions are invoked to process an input once it has been
@@
     \item [Rules]  Rules are responsible for classifying inputs: each rule
-        recognises one input \textit{variant\/} (a single input category
-        may have many input variants).  Each rule also specifies the action
+        recognises one input \textit{variant\/} --- a single input category
+        may have many input variants.  Each rule also specifies the action
         to be invoked when an input has been recognised; rules thus provide
@@
 warning; the framework will usually continue parsing after this, although
-for some inputs stopping immediately may be preferable.
+some parsers might prefer to stop immediately.
 
@@
         Abbreviations   & Rules         & Responsible for
-                                          classifying input.          \\
+                                          recognising inputs.         \\
         \tablebottomline{}%
@@
 deal with the variation in inputs, and the actions deal with the
-intricacies and complications encountered during parsing.  The function
-that finds the rule that recognises the input and invokes the requested
-action can be expressed in pseudo-code as shown in \figureref{Pseudo-code
-describing how inputs are recognised}.
+intricacies and complications encountered during parsing.  Finding the rule
+that recognises the input is a very simple process, as shown by the
+pseudo-code in \figureref{Pseudo-code describing how inputs are
+recognised}.  The framework tries each rule until it finds one that
+recognises the input, then it invokes the action specified by the rule.
+The framework issues a warning if the input is not recognised by any of the
+rules.
 
 \begin{figure}[thbp]
-    \caption{Pseudo-code describing how inputs are recognised}
+    \caption{Pseudo-code showing how the framework finds the rule that
+    recognises the input}
     \empty{}\label{Pseudo-code describing how inputs are recognised}
@@
     \item [Register actions]  Each action needs to be registered with the
-        framework so that the framework knows about it; the list of
-        registered actions will also be used when validating rules.
+        framework so that the framework knows about it: the list of
+        registered actions will be used when validating rules.
 
@@
         checks will be implementation-specific to some extent, but will
-        generally include some of the following:
+        generally include the following:
 
@@
         Some optimisation steps may also be performed during loading of
-        rules, as discussed in \sectionref{parser efficiency}.
+        rules, as described in \sectionref{parser efficiency}.
 
@@
         although actions may use cascaded parsing (explained in
-        \sectionref{actions in architecture}) to effectively push input
-        back onto the input stream.  A physical input (e.g.\ a single line
-        read from the input stream) may contain multiple or partial logical
-        inputs, and the framework must pre-process these to provide a
-        logical input for the rules to recognise.  This simplifies the
-        rules and actions considerably, at the cost of added complexity in
-        the framework; during the design phase it was decided that it was
-        easier to deal with the problem of parsing multiple or partial
-        inputs once, rather than dealing with it in every rule and action.
-        This is trivial for Postfix log files because they have a
-        one-to-one mapping between physical and logical inputs; mapping
-        between physical and logical inputs may be more difficult for other
-        types of input.  Some input types may require pre-processing
-        equivalent to parsing the physical inputs; in those cases the
-        framework present the entire input stream at once, and discard the
-        recognised portion.
+        \sectionref{actions in architecture}) to push input back onto the
+        input stream.  A physical input (e.g.\ a single line read from the
+        input stream) may contain multiple or partial logical inputs, and
+        the framework must pre-process these to provide a logical input for
+        the rules to recognise.  This simplifies the rules and actions
+        considerably, at the cost of added complexity in the framework;
+        during the design phase it was decided that it was easier to deal
+        with the problem of parsing multiple or partial inputs once, rather
+        than dealing with it in every rule and action.  This is trivial for
+        Postfix log files because they have a one-to-one mapping between
+        physical and logical inputs; mapping between physical and logical
+        inputs may be more difficult for other types of input.  Some input
+        types may require pre-processing equivalent to parsing the physical
+        inputs; in such cases the framework should take the approach
+        adopted by many other parsers: combine the physical inputs into one
+        complete input, use the rules to recognise the start of the input,
+        discard the recognised portion if successful, and repeat until the
+        input has been exhausted.
 
@@
         describing how inputs are recognised} shows how rules are
-        successively tried until one is found that recognises the current
-        input.  That pseudo-code is very simple: there may be efficiency
-        concerns (\sectionref{parser efficiency}), rule conditions
+        successively tried until one is found that recognises the input.
+        That pseudo-code is very simple: there may be efficiency concerns
+        (\sectionref{parser efficiency}), rule conditions
         (\sectionref{attaching conditions to rules}), or rule priorities
@@
     \item [Invoke actions]  Once a rule has been found that recognises the
-        current input, the specified action must be invoked.  The framework
+        input, the specified action will be invoked.  The framework
         marshals the data extracted by the rule, invokes the action, and
-        repeats the parsing process if cascaded parsing (see
-        \sectionref{actions in architecture}) is used.
+        pushes the modified input onto the input stream if the action uses
+        cascaded parsing (see \sectionref{actions in architecture}).
 
@@
         block to another.  The framework provides shared storage to deal
-        with this and any other storage needs the actions have.  Actions
-        may need to exchange data to correctly parse the input, e.g.\
-        setting or clearing flags, maintaining a list of previously used
-        identifiers, or ensuring at a higher level that the input being
-        parsed meets requirements.
+        with this and any other storage needs the actions may have.
+        Actions may need to exchange data to correctly parse the input,
+        e.g.\ setting or clearing flags, maintaining a list of previously
+        used identifiers, or ensuring at a higher level that the input
+        being parsed meets certain requirements.
 
     \item [Save and load state]  The architecture can save the contents of
-        the shared storage it provides for actions, and reload it later, so
+        the shared storage it provides for actions, and reload it later so
         that information is not lost between parsing runs.  \parsername{}
-        uses this functionality because mails may take some time to deliver
+        does this because mails may take some time to deliver
         and thus have their log lines split between log files; a compiler
@@
         functions, but if another way exists to make those functions
-        available to all actions it may be preferable to use it instead,
-        maintaining a clear separation of concerns.
+        available to all actions it may be preferable to use that way
+        instead, maintaining a clear separation of concerns.
 
@@
 Each action is a separate procedure written to process a particular
-category of input, e.g.\ rejection of a delivery attempt.  There may be
-many input variants within one input category; in general each action will
-handle one input category, with each rule recognising one input variant.
-It is anticipated that parsers based on this architecture will have many
-actions, and each action may be specified by many rules, with the aim of
-having simple rules and tightly focused actions.  An action may need to
-process different input variants in slightly different ways, but large
-changes in processing within an action indicate the need for a new action
-and a new category of input; if an action becomes overly complicated it
-starts to turn into a monolithic parser, with too much logic contained in a
-single procedure.
+category of input, e.g.\ rejection of a delivery attempt.  One input
+category may have many input variants; in general each action will handle
+one input category, with each rule recognising one input variant.  It is
+anticipated that parsers based on this architecture will have many actions,
+and each action may be invoked by many rules, with the aim of having simple
+rules and tightly focused actions.  An action may need to process different
+input variants in slightly different ways, but large variation in the
+processing performed by an action indicates the need for a new action and a
+new category of input; if an action becomes overly complicated it starts to
+turn into a monolithic parser, with too much logic contained in a single
+procedure.
 
@@
 of the strengths of this architecture.  When a new requirement arises an
-independent action can be written to satisfy it, instead of writing a
-single monolithic function that processes every input and must be modified
-to support any change in behaviour, with all the attendant risks of
-adversely affecting the existing parsing.  Sometimes the new action will
-require the cooperation of other actions, e.g.\ to set or check a flag, so
-actions are not always self-contained, but there will still be a far lower
-degree of coupling and interdependency than in a monolithic parser.
+independent action can be written to satisfy it, instead of having to
+modify a single monolithic function that processes every input, with all
+the attendant risks of adversely affecting the existing parsing.  Sometimes
+the new action will require the cooperation of other actions, e.g.\ to set
+or check a flag, so actions are not always self-contained, but there will
+still be a far lower degree of coupling and interdependency than in a
+monolithic parser.
 
 During development of \parsername{} it became apparent that in addition to
-the obvious variety in log lines there were many complications to be
-overcome.  Some were the result of deficiencies in Postfix's logging, and
-some of those deficiencies were rectified by later versions of Postfix,
-e.g.\ identifying bounce notifications (\sectionref{identifying bounce
+the obvious variety in log lines there were many complications to overcome.
+Some were the result of deficiencies in Postfix's logging, and some of
+those deficiencies were rectified by later versions of Postfix, e.g.\
+identifying bounce notifications (\sectionref{identifying bounce
 notifications}); others were due to the vagaries of process scheduling,
@@
 
+This section is quite brief, because the architecture deliberately imposes
+as few restrictions, conditions, and conventions as possible on actions, to
+allow maximum flexibility for parsers based on this architecture.
+
 \section{Rules}
@@
 
-Rules are responsible for categorising inputs: each rule should recognise
+Rules are responsible for recognising inputs: each rule should recognise
 one and only one input variant; an input category with multiple input
@@
 input, or checking the length of the input; for the remainder of this
-thesis it will be assumed that a regex is used.  Each rule must specify at
-least the regex to recognise inputs and the action to invoke when
-recognition is successful, but implementations are free to add any other
-attributes they require; \sectionref{rules table} describes the attributes
-used in \parsername{}, and some generally useful attributes will be
-discussed later in this section.
+thesis it will be assumed that a regex is used.  Each rule must specify the
+regex to recognise inputs and the action to invoke when recognition is
+successful, but implementations are free to add any other attributes they
+require; \sectionref{rules table} describes the attributes used in
+\parsername{}, and some generally useful attributes will be discussed later
+in this section.
 
@@
 the action that will be invoked; there is no backtracking to try alternate
-rules; no attempt is made to pick a \textit{best\/} rule.
+rules, and no attempt is made to pick a \textit{best\/} rule.
 \sectionref{attaching conditions to rules} contains an example which
@@
 input, so a mechanism is needed to allow the author of the rules to specify
-that ordering.  The framework should sort the rules according to a priority
-attribute specified by the rule author, allowing fine-grained control over
+that ordering.  Each rule can have a priority attribute: when recognising
+inputs, the framework should try the rules in the order specified by their
+priority attributes, giving the ruleset author fine-grained control over
 the order that rules are used in.  The priority attribute may be
@@
 existing rules or adding new rules, machine learning techniques could be
-used to automate the process.  If this approach was taken the choice of
+used to automate the process.  If this approach was taken, the choice of
 machine learning technique would be constrained by the size of typical data
@@
 can be as simple or complex as the parser requires, though naturally as the
-complexity rises so too does the difficulty in understanding how everything
-interacts.  The framework has to evaluate each condition, so as the
-complexity of conditions increases so will the complexity of the code
-required to evaluate them.  Beyond a certain level of complexity,
+complexity rises so too does the difficulty in understanding how different
+rules and actions interact.  The framework has to evaluate each condition,
+so as the complexity of conditions increases so will the complexity of the
+code required to evaluate them.  Beyond a certain level of complexity,
 conditions should probably be written in a proper programming language,
@@
 implementation is going to use conditions so complex that they will require
-a programming language capable of implementing a Turing machine, the design
-may need to be revisited, including the decision to use this architecture
---- there may be other architectures more suitable.
+a Turing-complete programming language, the design may need to be
+revisited, including the decision to use this architecture --- there may be
+other architectures more suitable.
 
-Conditions that examine the input will be the easiest to understand,
-because they can be understood in isolation; they do not depend on
-variables set by actions or other rules.  Conditions that examine the input
-can be complex if required, but simple conditions can be quite useful too,
-e.g.\ every Postfix log line contains the name of the Postfix component
-that produced it, so every rule used in \parsername{} has a condition
-specifying the component whose log lines it recognises, reducing the number
-of rules that will be used when classifying a log line (see
-\sectionref{rules in implementation} for details) and increasing the chance
-that the log line will be correctly recognised.
+Conditions that only examine the input will be the easiest conditions to
+understand, because they can be understood in isolation; they do not depend
+on variables set by actions or other rules.  Conditions that examine the
+input can be complex if required, but simple conditions can be quite useful
+too, e.g.\ every Postfix log line contains the name of the Postfix
+component that produced it, so every rule used in \parsername{} has a
+condition specifying the component whose log lines it recognises, reducing
+the number of rules that will be used when recognising a log line (see
+\sectionref{rule conditions in implementation} for details) and increasing
+the chance that the log line will be correctly recognised.
 
@@
 
-An example of how rule conditions can be useful is parsing C-style
-comments, which start with \texttt{/*} and end with \texttt{*/}; the start
-and end tokens can be on one line, or may have many lines between them.
+An example of how rule conditions can be used is parsing C-style comments,
+which start with \texttt{/*} and end with \texttt{*/}; the start and end
+tokens can be on one line, or may have many lines between them.
 \Tableref{Rules to parse C-style comments} shows the regexes, conditions,
 and state changes of the four rules required to parse C-style comments.
-These rules are somewhat simplified, e.g.\ rules one and two will recognise
-the comment start token even if it is within a quoted string, which is
-incorrect; the rules could be extended with new rules to parse quoted
-strings, excluding comments from within strings.  Rules one and two will be
-used when the parser is parsing code, not comments: rule one recognises a
-comment that is contained within one line and leaves the parser's state
-unchanged; rule two recognises the start of a comment and changes the
-parser's state to parsing comments instead of parsing code.  Rules three
-and four will be used when the parser is parsing a comment: rule three
-recognises the end of a comment and switches the parser's state back to
-parsing code; rule four recognises a comment line without an end token and
-keeps the parser's state unchanged.  It is important that the rules are
-applied in the order listed in \tableref{Rules to parse C-style comments}
-because rule 2 overlaps with rule 1, so rule 1 must be used first to
-correctly parse comments that are contained within one line;
-\sectionref{rules in architecture} has explained how this is achieved.
-\sectionref{overlapping rules in architecture} will discuss the benefits
-and difficulties of using overlapping rules.
+These are simplified rules, e.g.\ rules one and two will incorrectly
+recognise the comment start token if it is within a quoted string.  Rules
+one and two will be used when the parser is parsing code, not comments:
+rule one recognises a comment that is contained within one line and leaves
+the parser's state unchanged; rule two recognises the start of a comment
+and changes the parser's state to parsing comments instead of parsing code.
+Rules three and four will be used when the parser is parsing a comment:
+rule three recognises the end of a comment and switches the parser's state
+back to parsing code; rule four recognises a comment line without an end
+token and keeps the parser's state unchanged.  It is important that the
+rules are applied in the order listed in \tableref{Rules to parse C-style
+comments} because rule two overlaps with rule one, and rule four overlaps
+with rule three; \sectionref{rules in architecture} has explained how this
+is achieved.  \sectionref{overlapping rules in architecture} will discuss
+the benefits and difficulties of using overlapping rules.
 
@@
         \tablemiddleline{}%
-        1     & \verb!/\*.*?\*/!  & state == parsing code     & state = parsing code      \\
+        1     & \verb!/\*.*?\*/!  & state == parsing code     &                           \\
         2     & \verb!/\*.*!      & state == parsing code     & state = parsing comment   \\
         3     & \verb!.*\*/!      & state == parsing comment  & state = parsing code      \\
-        4     & \verb!.*!         & state == parsing comment  & state = parsing comment   \\
+        4     & \verb!.*!         & state == parsing comment  &                           \\
         \tablebottomline{}%
@@
     Note that ``state'' is merely a descriptive name for the variable; the
-    variable can called anything at all.
+    variable can be called anything at all.
 
@@
 the rest of the world.  Using overlapping rules simplifies both the general
-rule and the more specific rule; additionally rules from different sources
-can be combined with a minimum of prior cooperation or modification
-required.  Overlapping rules should have a priority attribute to specify
-their relative ordering; negative priorities may be useful for catchall
-rules.  The architecture does not try to detect overlapping rules: that
-responsibility is left to the author of the rules.
+rule and the more specific rule.  Overlapping rules should have a priority
+attribute to specify their relative ordering; negative priorities may be
+useful for catchall rules.  The architecture does not try to detect
+overlapping rules: that responsibility is left to the author of the rules.
 
-Overlapping rules can be detected by visual inspection, or a program may be
-written to analyse the regexes in a ruleset.  Traditional regexes are
+Overlapping rules can be detected by visual inspection, or a program could
+be written to analyse the regexes in a ruleset.  Traditional regexes are
 equivalent in computational power to \acronym{FA} and can be converted to
@@
 intersection of two \acronym{FA}\@.  Perl 5.10 regexes~\cite{perlre} are
-more powerful than traditional regexes: it is possible to match correctly
-balanced brackets nested to an arbitrary depth, e.g.\
+more powerful than traditional regexes: they can match correctly balanced
+brackets nested to an arbitrary depth, e.g.\
 \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
@@
 
-In the worst case $|o|$ above will be equal to $n^{2}$.  For this approach
+In the worst case, $|o|$ above will be equal to $n^{2}$.  For this approach
 to be more efficient than the naive approach, $y$ must be significantly
@@
 usage, but because they may prevent the correct rule from recognising the
-input.  If an adaptive ordering system is used that prioritises rules that
+input.  If an adaptive ordering system is used to prioritise rules that
 frequently recognise inputs (see \sectionref{rule ordering for
@@
 example, with most regex engines matching double quoted strings with
-\verb!"([^"\\]+|\\.)*"!  is very fast when there is a match, but when the
+\verb!\"([^"\\]+|\\.)*"\!  is very fast when there is a match, but when the
 match fails its computational complexity for a string of length $n$ is
@@
 groups.  Pathological regexes that consume excessive CPU time can be
-difficult to detect, both by visual inspection and by machine inspection,
+difficult to detect, whether by visual inspection or by machine inspection,
 but if a regex is converted to a \acronym{FA} or the internal
@@
 Conditions can vary in complexity from simple equality through to a
-Turing-equivalent language, so enumerating pathological conditions is
+Turing-complete language, so enumerating pathological conditions is
 difficult if not pointless.  Conditions that check variables or the input
@@
 how it can be useful for general parsing.  The rules section was the
-longest in this chapter, because although the rules appear to be simple ---
-recognise an input and specify the action that handles that input --- they
-have subtle behaviour that needs to be clearly explained.  When extending a
-ruleset, a decision needs to be taken about whether the input should be
-recognised by extending an existing rule, by adding a new rule to an
-existing rule category, or by adding a new input category, rule, and
-action.  Rules can have conditions attached to them, restricting the set of
+longest section in this chapter, because although the rules appear to be
+quite simple --- recognise an input and specify the action to invoke ---
+they have subtle behaviour that needs to be clearly explained.  When
+extending a ruleset, a decision needs to be taken about whether the input
+should be recognised by extending an existing rule, by adding a new rule to
+an existing input category, or by adding a new input category, action, and
+rule.  Rules can have conditions attached to them, restricting the set of
 rules used to recognise an input; the complexity of the conditions used
@@
 rules, because overlap amongst rules may be valid and is quite often
-intentional; that responsibility falls to the author of the rules.  The
+intentional; that responsibility falls to the author of the ruleset.  The
 difficulty involved in detecting overlap is proportional to the complexity

### 2009-04-26 doc/logparser-complications.tex
Changes made after proofreading the preface and chapters 1-3.
Changes based on proofreading chapters 4 and 5.

@@
         extra logging by Postfix could remove these complications, and for
-        one complication, extra logging in later versions of Postfix does
-        remove it (\sectionref{identifying bounce notifications}); three of
+        one complication (\sectionref{identifying bounce notifications}),
+        extra logging in later versions of Postfix does remove it; three of
         the complications would be removed if Postfix generated a log line
@@
 
-A delivery attempt lacks a queueid until one recipient has been accepted,
-so log lines must first be correlated by \daemon{smtpd} \acronym{pid}, then
-transition to being correlated by their queueid.  This is relatively minor,
-but does require:
+A delivery attempt lacks a queueid until the first recipient has been
+accepted, so log lines must first be correlated by \daemon{smtpd}
+\acronym{pid}, then transition to being correlated by their queueid.  This
+is relatively minor, but does require:
 
@@
     \item Each action needs to know whether it deals with accepted mails,
-        and should find the mail in the by\_queueid state tables, or deals
-        with delivery attempts, and thus will find the connection in the
-        by\_pid state tables.  Some actions, e.g.\
+        and should find the mail in the \texttt{queueid} state table, or
+        deals with delivery attempts, and thus will find the connection in
+        the \texttt{connections} state table.  Some actions, e.g.\
         \action{DELIVERY\_REJECTED}, need to query both tables.  Support
-        functions fall into two groups: those that deal with state tables,
+        functions fall into two groups: one group deals with state tables,
         and so need to know which one to operate on; the functions in the
@@
 \end{itemize}
+
+The \action{CLONE} action is responsible for copying an entry from the
+\texttt{connections} state table to the \texttt{queueids} state table.
 
@@
 from a strictly chronological reading of the log lines it usually appears
-as if the child mail has been created without a source.  Subsequently, the
-parent mail will log the creation of the child mail, e.g.\ parent mail
+as if the child mail does not have an origin.  Subsequently, the parent
+mail will log the creation of the child mail, e.g.\ parent mail
 \texttt{3FF7C4317} creates child mail \texttt{56F5B43FD}\@:
@@
 from an unknown origin.  Subsequently invoked actions will clear that flag
-if the origin of the mail becomes clear.  Obviously the parser could omit
-checking where mails originate from, but requiring an explicit source helps
-to expose bugs in the parser; such checks helped to identify the
-complications described in \sectionref{discarding cleanup log lines} and
+if the origin of the mail becomes clear.  The parser could omit checking
+where mails originate from, but requiring an explicit source helps to
+expose bugs in the parser; such checks helped to identify the complications
+described in \sectionref{discarding cleanup log lines} and
 \sectionref{pickup logging after cleanup}.
@@
 to distinguish between child mails waiting to be tracked and other mails
-whose origin is unknown (except for bounce notifications, as described in
-\sectionref{identifying bounce notifications}); postponing the entry into
-the database of mails whose origin is unknown can result in those mails
-remaining indefinitely in the state tables, causing a queueid clash if the
-queueid is reused, resource usage for the period they remain in the state
-tables, and, most importantly, \parsernames{} understanding of such mails
+whose origin is unknown, except for bounce notifications, as described in
+\sectionref{identifying bounce notifications}.  Mails whose origin is
+unknown can remain in the state tables indefinitely if their origin is not
+determined at some point; they will cause a queueid clash if the queueid is
+reused, and, most importantly, \parsernames{} understanding of such mails
 is incorrect.  This problem has only been observed when a state table entry
@@
 \parsername{} cannot be expected to fully understand and reconstruct a mail
-when it is lacking some of that mail's log lines.
+when some of that mail's log lines are missing.
 
@@
 Postfix 2.2.x (and presumably previous versions) does not generate a log
-line when it generates a bounce notification; the lof file will have log
-lines for a mail that lacks an obvious source.  Similarities exist to the
+line when it generates a bounce notification; the log file will have log
+lines for a mail whose origin is unknown.  Similarities exist to the
 problem of identifying re-injected mails discussed in \sectionref{tracking
@@
 notifications do not eventually have a log line that identifies their
-source.  Heuristics must be used to identify bounce notifications:
+origin.  Heuristics must be used to identify bounce notifications:
 
@@
 
-    \item The message-id must have the correct format and correspond with
-        the queueid of the mail.  Although a mail sent from elsewhere could
+    \item The message-id must have the correct format and contain the
+        queueid of the mail.  Although a mail sent from elsewhere could
         easily have the correct message-id format, the chance that the
@@
 If a mail is misclassified as a bounce message it will almost certainly
-have been generated internally by Postfix; arguably misclassification of
+have been generated internally by Postfix; arguably, misclassification of
 this kind is a benefit rather than a drawback, because other mails
 generated internally by Postfix will be handled correctly.  Postfix 2.3 and
-subsequent versions log the creation of a bounce message.
+subsequent versions log the creation of a bounce message, so this
+complication does not arise in their log files.  The solution to this
+complication will help with solving the complication in \sectionref{Bounce
+notification mails delivered before their creation is logged}.
 
@@
 Postfix does not log a message making the client's behaviour clear, so
-again heuristics are required to identify when a delivery attempt is
-aborted in this way.  A list of all mails accepted during a connection is
-saved in the connection's state table entry, and the accepted mails are
-examined when the disconnection action is invoked.  Each accepted mail is
-checked for the following:
+heuristics are required to identify when a delivery attempt is aborted in
+this way.  A list of all mails accepted during a connection is saved in the
+connection's state table entry, and the accepted mails are examined when
+the disconnection action is invoked.  Each accepted mail is checked for the
+following:
 
@@
         processed by the \action{SAVE\_DATA} action?  Any log lines after
-        the first two they should be informational only.
+        the first two should be informational only.
 
@@
 
-If all the checks above are successful then the mail is assumed to be an
-aborted delivery attempt and is removed form the state tables.  There will
-be no further entries logged for such mails, so without identifying and
-discarding them they accumulate in the state table and will cause clashes
-if the queueid is reused.  The mail cannot be entered in the database
-because the only data available is the client hostname and IP address, but
-the database schema requires many more fields be populated (see
-\sectionref{connections table} and \sectionref{results table}).  These
-heuristics are quite restrictive, and appear to have little scope for false
-positives; any false positives would cause a warning when the next log line
-for that mail is parsed.  False negatives are less likely to be detected:
-there may be queueid clashes (and warnings) if mails remain in the state
-tables after they should have been removed, otherwise the only way to
-detect false negatives is to examine the state tables after each parsing
+If all the checks above are successful the mail is assumed to be an aborted
+delivery attempt and is removed from the state tables.  There will be no
+further log lines for such mails, so without identifying and discarding
+them they accumulate in the state table and will cause clashes if the
+queueid is reused.  Such mails cannot be entered in the database because
+the only data they contain is the client hostname and IP address, but the
+database schema requires many more fields --- see \sectionref{connections
+table} and \sectionref{results table}.  These heuristics are quite
+restrictive, and appear to have little scope for false positives; any false
+positives would cause a warning when the next log line for such a mail is
+parsed.  False negatives are less likely to be detected: there may be
+queueid clashes (and warnings) if mails remain in the state tables after
+they should have been removed; if the queueid is not reused, the only way
+to detect false negatives is to examine the state tables after each parsing
 run.
@@
 is rejected; they may also disconnect after other errors, but such
-disconnections are either unimportant or are handled elsewhere in the
-parser, e.g.\ \sectionref{timeouts during data phase}.  Postfix does not
-log a message saying the mail has been discarded, as should be expected by
-now.  The checks to identify this happening are:
+disconnections are handled elsewhere in the parser, e.g.\
+\sectionref{timeouts during data phase}.  Postfix does not log a message
+saying the mail has been discarded, as should be expected by now.  The
+checks to identify this happening are:
 
@@
         There should be a connection log line and a mail acceptance log
-        line, followed by one or more delivery rejected log lines.
+        line, followed by one or more delivery attempt rejected log lines.
 
@@
 
-If both checks are successful then the mail is assumed to have been
-discarded by Postfix when the client disconnected; \parsername{} will
-remove it from the state tables.  There will be no further log lines for
-such mails, so if \parsername{} does not deal with them immediately they
-accumulate in the state table and will cause clashes if the queueid is
-reused.
+If both checks are successful the mail is assumed to have been discarded by
+Postfix when the client disconnected; \parsername{} will remove it from the
+state tables.  There will be no further log lines for such mails, so if
+\parsername{} does not deal with them immediately they accumulate in the
+state table and will cause clashes if the queueid is reused.
 
@@
 the mail and \parsername{} needs to discard the data associated with that
-mail.  It seems more intuitive to save the mail's data to the database, but
-if a timeout occurs no data will be available to save; the timeout is
+mail.  It seems more appropriate to save the mail's data to the database,
+but if a timeout occurs no data will be available to save; the timeout is
 recorded and saved with the connection instead.  To deal properly with
@@
 
-    \item Record the timeout and captured data in the connection's results.
+    \item Record the timeout and data extracted from the log line in the
+        connection's results.
 
-    \item If no mails have been accepted yet nothing needs to be done; the
-        \action{TIMEOUT} action ends.
+    \item If no mails have been accepted over the connection, nothing needs
+        to be done; the \action{TIMEOUT} action ends.
 
@@
         allocated a queueid for the incoming mail, and there may be a mail
-        in the state tables that needs to be discarded by \parsername{}.  The timeout may
-        have interrupted transfer of an accepted delivery attempt, or it
-        may have occurred after a mail delivery attempt was rejected.  If a
-        mail needs to be discarded, the following checks will all pass:
+        in the state tables that needs to be discarded by \parsername{}.
+        The timeout may have interrupted transfer of an accepted delivery
+        attempt, or it may have occurred after a mail delivery attempt was
+        rejected.  If a mail needs to be discarded, the following checks
+        will all pass:
 
@@
             \item The timestamp of the log line preceding the timeout log
-                line is earlier than the timestamp of the last accepted
-                delivery attempt, i.e.\ there have not been any rejections
-                since then the delivery attempt was accepted.
+                line must be earlier than the timestamp of the last
+                accepted delivery attempt, i.e.\ there have not been any
+                rejections since then the delivery attempt was accepted.
 
@@
 
-            \item The mail does not have a \daemon{qmgr} log line.
+            \item The mail must not have a \daemon{qmgr} log line.
 
@@
 
-        If all the checks pass the mail will be discarded from the state
-        tables, and will not be entered in the database.  If one or more
-        checks do not pass it is assumed that the timeout happened after a
+        If all checks pass the mail will be discarded from the state tables
+        and will not be entered in the database.  If one or more checks do
+        not pass, \parsername{} assumes that the timeout happened after a
         rejected delivery attempt.  This assumption is not necessarily
@@
         for the first recipient.  In that case the timeout applies to the
-        partially accepted mail, which should be discarded; however, this
-        has not occurred in practice.  Processing timeouts is further
-        complicated by the presence of out of order \daemon{cleanup} log
-        lines: see \sectionref{discarding cleanup log lines} for details.
+        partially accepted mail, which will be discarded by Postfix and
+        should be discarded by \parsername{}; however, this has not
+        occurred in practice.  Processing timeouts is further complicated
+        by the presence of out of order \daemon{cleanup} log lines: see
+        \sectionref{discarding cleanup log lines} for details.
 
@@
 timeout log line, by which time \parsername{} has discarded the mail
-because it will not have enough data to satisfy the database schema (see
-\sectionref{timeouts during data phase}); parsing the \daemon{cleanup} log
-line causes the \action{CLEANUP\_PROCESSING} action to create a new state
-table entry, to help deal with re-injected mails (\sectionref{Re-injected
-mails}).  This is incorrect because the log line actually belongs to the
-mail that has just been discarded; if the queueid is reused there will be a
-queueid clash, otherwise the new mail will just remain in the state tables.
+because it did not have enough data to satisfy the database schema (see
+\sectionref{timeouts during data phase}) and was not expected to have any
+more log lines; parsing the \daemon{cleanup} log line causes the
+\action{CLEANUP\_PROCESSING} action to create a new state table entry, to
+help deal with re-injected mails (\sectionref{Re-injected mails}).  This is
+incorrect because the log line actually belongs to the mail that has just
+been discarded; if the queueid is reused there will be a queueid clash,
+otherwise the new mail will just remain in the state tables.
 
-During the \action{TIMEOUT} action, if the \daemon{cleanup} log line is
-still pending, the action updates the \texttt{timeout\_queueids} state
-table, adding the queueid and the timestamp from the log line.  The
-following checks will be performed for each \daemon{cleanup} log line
-that is processed:
+During the \action{TIMEOUT} action, if the mail's \daemon{cleanup} log line
+is still pending, the \action{TIMEOUT} action updates the
+\texttt{timeout\_queueids} state table, adding the queueid and the
+timestamp from the log line.  To deal with this complication, the following
+checks will be performed for each \daemon{cleanup} log line that is
+processed:
 
@@
         removed, regardless of whether the remaining criteria are
-        satisfied.
+        satisfied.  If the queueid does not exist in the
+        \texttt{timeout\_queueids} state table, the log line being
+        processed cannot belong to a discarded mail.
 
     \item Has the queueid been reused yet, i.e.\ does it have an entry in
-        the state tables?
+        the \texttt{queueids} state table?  If it has an entry in the
+        \texttt{queueids} state table, the log line being processed belongs
+        to that mail, not to a previously discarded mail.
 
@@
 
-The \daemon{cleanup} log line must meet the criteria above for it to be
+The \daemon{cleanup} log line must pass the checks above for it to be
 discarded because some, but not all, connections where a timeout occurs
@@
 When mail is submitted locally, \daemon{pickup} processes the new mail and
-generates a log line.  Occasionally this log line will occur later in the
+generates a log line.  Occasionally, this log line will occur later in the
 log file than the \daemon{cleanup} log line, so the \action{PICKUP} action
@@
 
-    \item The only program that has logged anything thus far for the mail
-        is \daemon{cleanup}.
+    \item The only program that has logged anything so far for the mail is
+        \daemon{cleanup}.
 
@@
 
-As always with heuristics there may be circumstances in which these
+As always with heuristics, there may be circumstances in which these
 heuristics match incorrectly, but none have been identified so far.  This
-complication seems to occur during periods of particularly heavy load, so
-is most likely caused by process scheduling vagaries.
+complication only seems to occur during periods of particularly heavy load,
+so is most likely caused by process scheduling vagaries.
 
@@
 
-    \item Occasionally a \daemon{smtpd} will exit uncleanly, so the active
-        connection must be cleaned up, entered in the database if it has
-        enough data, and deleted from the connections state table.
+    \item Occasionally a \daemon{smtpd} will exit with an error, so the
+        active connection must be cleaned up, entered in the database if it
+        has enough data, and deleted from the connections state table.
 
-    \item Every Postfix process uses a watchdog which kills the process if
-        it is not reset for a considerable period of time (five hours by
-        default).  This safeguard prevents Postfix processes from running
-        indefinitely and consuming resources if a failure causes them to
-        enter a stuck state.
+    \item Every Postfix process uses a watchdog timer that kills the
+        process if it is not reset for a considerable period of time (five
+        hours by default).  This safeguard prevents Postfix processes from
+        running indefinitely and consuming resources if a failure causes
+        them to enter a stuck state.
 
@@
 The circumstances above account for all occasions where a \daemon{smtpd}
-suddenly stops logging.  In addition to removing an active connection the
-last accepted mail may need to be discarded, as detailed in
-\sectionref{timeouts during data phase}; otherwise the queueid state table
-is untouched.
+suddenly stops logging.  In addition to removing an active connection from
+the state tables, the last accepted mail may need to be discarded, as
+described in \sectionref{timeouts during data phase}; otherwise the queueid
+state table is untouched.
 
@@
 
-Occasionally a log file will have out of order log lines that cannot be
+Occasionally, a log file will have out of order log lines that cannot be
 dealt with by the techniques described in \sectionref{tracking re-injected
 mail}, \sectionref{discarding cleanup log lines}, or \sectionref{pickup
-logging after cleanup}.  In the \numberOFlogFILES{} log files used for
-testing this problem occurs only five times in 60,721,709 log lines.  All
-five occurrences have the same characteristics: the \daemon{local} log line
-showing delivery to a local mailbox comes after the \daemon{qmgr} log line
-showing removal of the mail from the queue because delivery is complete.
-This causes problems: the data in the state tables for the mail is not
-complete, so it cannot be entered into the database; a new mail is created
-when the \daemon{local} log line is processed and remains in the state
-tables; four warnings are issued per pair of out of order log lines.
+logging after cleanup}.  In the \numberOFlogFILES{} log files used in
+\sectionref{Evaluation} this problem occurs only five times in 60,721,709
+log lines.  All five occurrences have the same characteristics: the
+\daemon{local} log line showing delivery to a user's mailbox comes after
+the \daemon{qmgr} log line showing removal of the mail from the queue
+because delivery is complete.  This causes several problems: the data in
+the state tables for the mail is not complete, so it cannot be entered into
+the database; a new mail is created when the \daemon{local} log line is
+processed and remains in the state tables; four warnings are issued per
+pair of out of order log lines.
 
@@
 entry must be postponed and the mail flagged for later entry.  The
-\action{MAIL\_DELIVERED} action checks for that flag; if the additional log
-lines have caused the mail to reach a valid combination then its entry in
-the database will proceed, otherwise it must be postponed once more.
+\action{MAIL\_DELIVERED} action checks for that flag; if the log line is
+has just processed has caused the mail to reach a valid combination then
+entry of the mail into the database will proceed, otherwise it must be
+postponed once more.
 
@@
 attempts} occur frequently, but those described in this section only occur
-four times in the \numberOFlogFILES{} log files used for testing.  The
-symptoms are the same as in \sectionref{aborted delivery attempts}, except
-that the mail \textit{has\/} a \daemon{cleanup} log line; nothing can be
-found in the log files to explain why this mail does not have any further
-log lines.  The only way to detect these mails is to periodically scan all
-mails in the state tables, deleting any mails with the following
-characteristics:
+four times in the \numberOFlogFILES{} log files used in
+\sectionref{Evaluation}.  The symptoms are the same as in
+\sectionref{aborted delivery attempts}, except that the mail \textit{has\/}
+a \daemon{cleanup} log line; nothing can be found in the log files to
+explain why this mail does not have any further log lines.  The only way to
+detect these mails is to periodically scan all mails in the state tables,
+deleting any mails with the following characteristics:
 
@@
 if delivery has yet to be attempted \parsername{} cannot determine the
-recipient address or addresses.  This is a problem when mail is arriving
+recipient address or addresses.  This can occur when mail is accepted
 faster than Postfix can attempt delivery, and the administrator deletes
 some of the mail before Postfix has had a chance to attempt delivery.  The
-recipient address will not have been logged yet, and the deleted mails will
-not have any more log lines for it to be logged in.  A dummy recipient
-address needs to be added, because every mail is required by the database
-schema (\sectionref{results table}) to have at least one recipient.
-Typically when this complication occurs the log file will show many
-instances of it, closely grouped.
+deleted mail's recipient address or addresses will not have been logged
+yet, and the deleted mails will not have any more log lines.  A dummy
+recipient address needs to be added by \parsername{}, because every mail is
+required by the database schema (\sectionref{results table}) to have at
+least one recipient.  When this complication occurs the log file will
+typically show many instances of it, closely grouped.  Generally, this
+problem arises because the administrator has deleted some mails from
+Postfix's mail queue to stop a mail loop.
 
-This lack of information cannot easily be overcome: it is trivial to log a
-warning for every accepted recipient, but Postfix will not yet have
-allocated a queueid for the mail when the warning for the first recipient
-is logged, so the warning will be associated with the connection rather
-than the accepted mail.  A queueid will be allocated after Postfix accepts
-the MAIL FROM command if \texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is
-set to ``no'', but that setting will cause disk IO for every delivery
-attempt, instead of just for delivery attempts where recipients are
-accepted, and consequently a drastic reduction in the performance of the
-mail server.
+This lack of information cannot easily be overcome: it is simple to
+configure Postfix to log a warning for every accepted recipient, but
+Postfix will not yet have allocated a queueid for the mail when the warning
+for the first recipient is logged, so the warning will be associated with
+the connection rather than the accepted mail.  A queueid will be allocated
+after Postfix accepts the MAIL FROM command if
+\texttt{smtpd\_delay\_open\_until\_valid\_rcpt} is set to ``no'', but that
+setting will cause disk IO for every delivery attempt, instead of just for
+delivery attempts where recipients are accepted, and consequently a drastic
+reduction in the performance of the mail server.
 
@@
 
-\subsection{Bounce Notification Mails Are Delivered Before Their Creation
-Is Logged}
+\subsection[Bounce Notification Mails Delivered Before Their Creation Is
+Logged]{Bounce Notification Mails Delivered Before \newline{} Their
+Creation Is Logged}
 
@@
 mails are created, delivered, and deleted from the queue, \textit{before\/}
-the log line from \daemon{bounce} that explains their source.  To deal with
-this the \action{COMMIT} action maintains a cache of recently committed
-bounce notification mails, which the \action{BOUNCE\_CREATED} action
-subsequently checks if the bounce mail is not found in the state tables.
-If the queueid exists in the cache, and its start time is less than ten
-seconds before the timestamp of the bounce log line, it is assumed that the
-bounce notification mail has already been processed and the
-\action{BOUNCE\_CREATED} action does not create one.  If the queueid exists
-in the cache it is removed, because it has either just been used or the
-problem did not occur for that mail.  Whether the \action{BOUNCE\_CREATED}
-action creates a mail or finds an existing mail in the state tables, it
-flags the mail as having been seen by the \action{BOUNCE\_CREATED} action;
-if this flag is present the \action{COMMIT} action will not add the mail to
-the cache of recent bounce notification mails.  This is not required to
-correctly deal with the complication, but is an optimisation to reduce the
-parser's memory usage.  On the occasions the author has observed this
-complication occurring there have been a huge number of bounce notification
-mails generated: if every bounce notification mail was cached it would
-dramatically increase the memory requirements of the parser.  The cache of
-bounce notification mails will be pruned whenever the parser's state is
-saved, though if the size of the cache ever becomes a problem it could be
-pruned periodically to keep it in check.
+the log line from \daemon{bounce} that explains their origin.  The origin
+of the mail is unknown when it is created, but when the time comes to enter
+it into the database its origin is correctly identified by the heuristics
+used to identify Postfix 2.2 bounce notifications, described in
+\sectionref{identifying bounce notifications}.  To avoid incorrectly
+creating a new mail when the out of order bounce notification log line is
+processed, the \action{COMMIT} action maintains a cache of recently
+committed bounce notification mails named \texttt{bounce\_queueids}, which
+the \action{BOUNCE\_CREATED} action subsequently checks when processing the
+bounce creation log line.  If the queueid exists in the cache, and its
+start time is less than ten seconds before the timestamp of the bounce log
+line, it is assumed that the bounce notification mail has already been
+processed and the \action{BOUNCE\_CREATED} action does not create one.  If
+the queueid exists in the cache it is removed, because it has either just
+been used or the problem did not occur for the new mail.  Whether the
+\action{BOUNCE\_CREATED} action creates a new mail or finds an existing
+mail in the \texttt{queueids} state table (not the
+\texttt{bounce\_queueids} cache), it flags the mail as having been seen by
+the \action{BOUNCE\_CREATED} action; if this flag is present the
+\action{COMMIT} action will not add the mail to the
+\texttt{bounce\_queueids} cache.  This prevents a bounce notification log
+line being incorrectly discarded if a queueid is reused within 10 seconds.
 
@@
 
-The administrator can delete mails using \daemon{postsuper}; occasionally
-mails that are in the process of being delivered will be deleted.  This
-results in the log lines from the delivery agent (\daemon{local},
-\daemon{smtp}, or \daemon{virtual}) appearing in the log file
-\textit{after\/} the mail has been removed from the state tables and saved
-in the database.  The \action{DELETE} action adds deleted mails to a cache,
-which is checked by the \action{SAVE\_DATA} action, and the current log
-line discarded if the following conditions are met:
+The administrator can delete mails using \daemon{postsuper}; occasionally,
+mails that are in the process of being delivered will be deleted by the
+administrator.  This results in the log lines from the delivery agent
+(\daemon{local}, \daemon{smtp}, or \daemon{virtual}) appearing in the log
+file \textit{after\/} the mail has been removed from the state tables and
+saved in the database.  The \action{DELETE} action adds deleted mails to a
+cache named \textit{postsuper\_deleted\_queueids}, which is checked by the
+\action{MAIL\_DELIVERED} action, and the current log line discarded if the
+following conditions are met:
 
@@
 
-Sadly this solution involves discarding some data, but the complication
+Sadly, this solution involves discarding some data, but the complication
 only arises eight times in quick succession in one log file, which is not

### 2009-04-26 doc/logparser-implementation.tex
Changes based on proofreading chapters 4 and 5.

@@
 
-This chapter documents the implementation of \parsername{}, a parsed based
-on the architecture described in \sectionref{parser architecture}.  Any
-design may look good when examined in the abstract, but the real test of
-the design comes with the first concrete implementation; only then do any
-practical difficulties come to light.  Implementing this architecture was
-relatively straightforward: the difficulties came not from the architecture
-and design, but from anomalies in the log files and Postfix's behaviour, as
-described in \sectionref{complications}.  \parsername{} successfully deals
-with all the difficulties that were discovered during its development, but
-further difficulties may arise during future usage; unfortunately solving
-all potential difficulties that may arise is an intractable task, but the
-descriptions of the solutions developed thus far should help if someone
-needs to develop a solution of their own.
+This chapter documents the implementation of \parsername{}, a parser for
+Postfix log files based on the architecture described in \sectionref{parser
+architecture}.  Any design may look good when examined in the abstract, but
+the real test of the design comes with the first concrete implementation;
+only then do any practical difficulties come to light.  Implementing this
+architecture was straightforward --- the difficulties came not from the
+architecture, but from anomalies in the log files and Postfix's behaviour,
+as described in \sectionref{complications}.  \parsername{} successfully
+deals with all the difficulties that were discovered during its
+development, but further difficulties may arise during future usage;
+unfortunately solving all potential difficulties that may arise is an
+impossible task, but the descriptions of the solutions developed thus far
+should help if someone needs to solve a problem of their own.
 
@@
 follow the journey each mail takes through Postfix, combining the data
-captured by rules into a coherent whole, saving it in a useful and
-consistent form, and performing housekeeping duties.  The intermingling of
+captured by rules into a coherent whole, and saving it in a useful and
+consistent form.  The intermingling of
 log lines from different mails immediately rules out the possibility of
@@
 phase, the parsing process, and the conveniences it offers to users of
-\parsername{}; that section finishes with the features that are important
-for this thesis but not strictly necessary for parsing: the performance
-data collected by the framework, the optimisations that can be disabled to
-show their effect, and the debugging options it provides.  The second
-component of the architecture is the actions, starting with a graph of how
-often each action is specified by rules, a description of why some actions
-are more popular than others, and how this popularity has no influence on
-how often an action is invoked.  Every action that is part of \parsername{}
-is documented, and the actions section concludes with a description of the
-process of adding a new action.  Rules are the final component of the
-architecture, and also the most visible to advanced users, e.g.\ systems
-administrators, because it is likely that they will need to add their own
-rules to recognise their log lines.  An example rule used by \parsername{}
-is examined, with every field clearly documented; that is followed by a
-description of adding new rules, and how to determine the value of each of
-the fields.  \parsername{} provides a utility to create regexes from
-unrecognised log lines; the algorithm it uses is documented, including the
-differences between it and the original algorithm it is based on.  The
-rules section finishes with a discussion of how \parsername{} uses rule
-conditions and overlapping rules, plus a description of the regex snippets
-provided to ease the process of writing regexes.
+\parsername{}; that section finishes by describing features that are
+important for this thesis but not strictly necessary for parsing: the
+performance data collected by the framework, the optimisations that can be
+disabled to show their effect, and the debugging options the framework
+provides.  The second component of the architecture is the actions,
+starting with a graph showing how often each action is specified by rules,
+a description of why some actions are more popular than others, and how
+this popularity has no influence on how often an action is invoked.  Every
+action that is part of \parsername{} is documented, and the actions section
+concludes with a description of the process of adding a new action.  Rules
+are the third component of the architecture, and also the most visible to
+advanced users, e.g.\ systems administrators, because it is likely that
+they will need to add their own rules to recognise their own log lines.  A
+sample rule used by \parsername{} is examined, with every field clearly
+documented; that is followed by a description of adding new rules, and how
+to determine the value of each of the new rule's fields.  \parsername{}
+provides a utility to create regexes from unrecognised log lines; the
+algorithm it uses is documented, including the differences between it and
+the original algorithm it is based on.  The rules section finishes with a
+discussion of how \parsername{} uses rule conditions and overlapping rules,
+plus a description of the regex snippets provided to ease the process of
+writing regexes.
 
@@
     \item The parser acts as a second source of information about Postfix's
-        operation, based on empirical evidence from log files.  A separate
-        project could compare the empirical knowledge inherent in
+        operation, based on empirical evidence gathered from log files.  A
+        separate project could compare the empirical knowledge inherent in
         \parsername{} with Postfix's documentation and source code to see
@@
 states and state transitions shown in the flow chart will be explained
-later in this section.
+in this section.
 
@@
 
-    \item The client disconnects \flowchart{DISCONNECT, TIMEOUT}{6}, either
-        normally or with an error.  If Postfix has rejected any
-        \acronym{SMTP} commands the data gathered from those rejections
-        will be saved to the database; if there were no rejections there
-        will not be any data to save.  Any mails accepted will already have
-        been cloned so their data is in another data structure, and will be
-        delivered in the same way as mails that entered the system by any
-        other route.
+    \item The client disconnects \flowchart{DISCONNECT \textnormal{or}
+        TIMEOUT}{6}, either normally or with an error.  If Postfix has
+        rejected any mail delivery attempts the data gathered from those
+        rejections will be saved to the database \flowchart{DISCONNECT}{7};
+        if there were no rejections there will not be any data to save.
+        Any mails accepted will already have been cloned so their data is
+        in another data structure, and will be delivered in the same way as
+        mails that entered the system by any other route.
 
@@
     \item The mail will have one or more delivery attempts
-        \flowchart{MAIL\_SENT, SAVE\_DATA}{8}.
+        \flowchart{MAIL\_SENT \textnormal{or} SAVE\_DATA}{8}.
 
@@
         the database, then deleted from the state tables
-        \flowchart{COMMIT}{11}.
+        \flowchart{COMMIT}{10}.
 
@@
 
-A database is used to store both the rules and the data gleaned by parsing
-Postfix log files.  Understanding the database schema is important in
-understanding the actions of the parser, and essential to developing
-further applications that utilise the data; \sectionref{database as API}
-describes how the database schema functions as an \acronym{API}.
+An \acronym{SQL} database is used to store both the rules and the data
+gleaned by parsing Postfix log files.  Understanding the database schema is
+helpful in understanding the actions of the parser, and essential to
+developing further applications that utilise the data; \sectionref{database
+as API} describes how the database schema functions as an \acronym{API}.
 
@@
 and specifies the action to be invoked when a log line is recognised; they
-also have several fields that aid the user in understanding the meaning of
-the log lines recognised by each rule.  The rules are described in detail
-in \sectionref{rules in implementation}, but the rules table is documented
-in \sectionref{rules table} with the rest of the database schema.
-
-The data saved from parsing the log files is divided into two tables:
+also have several other fields used by the parser, and several fields that
+aid the user in understanding the meaning of the log lines recognised by
+each rule.  The rules are described in detail in \sectionref{rules in
+implementation}, but the rules table is documented in \sectionref{rules
+table} with the rest of the database schema.
+
+The data saved from parsing log files is divided into two tables:
 connections and results.  The connections table contains an entry for every
@@
 available for inspection because each result references the rule that
-created it.  No ambiguity is possible about which rule resulted in a
+created it.  No ambiguity is possible about which rule or action created a
 particular result, eliminating one potential source of confusion.
@@
 it, because storing the data is a once-off operation, whereas utilising
-the data requires frequent searching, sorting, and manipulation of the data
-to produce customised reports or statistics.
+the data requires frequent searching, sorting, and manipulation of the
+data.
 
@@
 
-            \item New columns can be added to the tables used by the
-                program, using DEFAULT clauses or TRIGGERS to populate
-                them.
+            \item New columns can be added to tables, using DEFAULT clauses
+                or TRIGGERS to populate them.
 
@@
 modified by \parsername{}, except when it updates the hits and hits\_total
-fields.  Rules recognise the individual log lines, capturing data to be
-saved in the connections and results tables, and specifying the action to
-invoke for each recognised log line.
+fields.  Rules recognise individual log lines, capturing data to be saved
+in the connections and results tables, and specifying the action to invoke
+for each recognised log line.
 
@@
     \item [id] A unique identifier that other tables can use when referring
-        to a specific rule.
+        to a specific rule.  This will be assigned by the database.
 
@@
     \item [hits\_total] The total number of log lines recognised by this
-        rule over all runs of the parser; hits is reset to zero each time
+        rule over all runs of the parser; hits starts from zero each time
         the parser is run, but hits\_total is not.  This field will be set
@@
 attempt will have a single entry in the connections table containing all of
-the following below.  For an incoming connection, the client is the remote
+the fields below.  For an incoming connection, the client is the remote
 machine, and the server is the local machine; for outbound mail delivery
@@
 
-    \item [id] This field uniquely identifies the row.
+    \item [id] This field uniquely identifies the row.  This will be
+        assigned by the database.
 
@@
         occasionally changes during a connection, presumably because spam
-        or virus senders think it is a good idea.  By default Postfix only
-        logs the HELO hostname when it rejects a mail delivery attempt, but
-        it is quite simple to rectify this as described in
-        \sectionref{logging helo}.
+        or virus senders think it is a good idea.  Postfix only logs the
+        HELO hostname when it rejects a mail delivery attempt, but it is
+        quite simple to rectify this as described in \sectionref{logging
+        helo}.
 
@@
 
-Every recognised log line will have a row in the results table, associated
-with a single connection; a single connection will have at least one result
-associated with it, but will usually have several, and may have hundreds.
+Every recognised log line will have a row in the results table, and each
+row is associated with a single connection; a single connection will have
+at least one result associated with it, but will usually have several, and
+may have hundreds.
 
@@
 
-    \item [id] A unique identifier for this result.
+    \item [id] A unique identifier for this result.  This will be assigned
+        by the database.
 
@@
         line.  In general an \acronym{SMTP} code is only present in a log
-        line representing a rejection or final delivery; results whose log
-        line did not contain an \acronym{SMTP} code will duplicate the
-        \acronym{SMTP} code of other results in that connection.  Some
-        final delivery log lines do not contain an \acronym{SMTP} code,
-        e.g.\ when Postfix delivers to a user's mailbox; in those cases the
-        \acronym{SMTP} code is specified by the rule's
-        \texttt{result\_data} field, based on the success or failure
-        represented by the log line.
+        line representing a mail being delivered or a mail delivery attempt
+        being rejected; results whose log line did not contain an
+        \acronym{SMTP} code will duplicate the \acronym{SMTP} code of other
+        results in that connection.  Some mail delivery log lines do not
+        contain an \acronym{SMTP} code, e.g.\ when Postfix delivers to a
+        user's mailbox; in those cases the \acronym{SMTP} code is specified
+        by the rule's \texttt{result\_data} field, based on the success or
+        failure represented by the log line.
 
@@
         connection if the client uses different sender addresses for
-        multiple rejected delivery attempts; however there will not be
-        different sender addresses in the results for one accepted mail.
+        multiple rejected delivery attempts; however, the results for one
+        accepted mail will only have one sender address.
 
@@
 
-    \item [message\_id] The message-id of the accepted mail, again present
-        for delivered mails only.
+    \item [message\_id] The message-id of the accepted mail, again it is
+        present for delivered mails only.
 
@@
 manages the parsing process, taking care of the drudgery and boring tasks,
-provides services to the actions, and implements some efficiency measures.
-
-Each time the parser is run the framework performs some initialisation
-tasks, setting up several state tables that will later be used by actions.
+provides services to the actions, and implements some optimisations.
+
+The framework performs some initialisation tasks each time the parser is
+run, setting up several state tables that will later be used by actions.
 Data about the connections and mails being processed is held in
 \texttt{connections} and \texttt{queueids} respectively.  The other state
-tables are used when working around complications in parsing:
+tables are used when solving complications that arise during parsing:
 \texttt{timeout\_queueids} is used when dealing with connections that time
@@
 \texttt{postsuper\_deleted\_queueids} caches information about mails that
-were recently deleted by the administrator so that subsequent log lines
+were recently deleted by the administrator, so that subsequent log lines
 processed by the \action{SAVE\_DATA} action can be discarded
@@
 those rules whose \texttt{program} field equals the program in the log line
-(previously described in \sectionref{rule conditions in implementation}),
-falling back to generic rules if necessary, and finally warning if the log
-line is unrecognised.  The repetitive nature of log files gives them high
+(described in \sectionref{rule conditions in implementation}), falling back
+to generic rules if necessary, and finally warning if the log line is
+unrecognised.  The repetitive nature of log files gives them high
 compression ratios; the framework uses a Perl module named
@@
 remainder of the parsing process is likely to take.  The progress bar is
-not as accurate when parsing compressed log files, because a compressed
+not as accurate when parsing compressed log files, because each compressed
 block uncompresses to a variable number of log lines; variation between the
@@
 
-    \item The number of log lines and rules used for each Postfix component
-        (\graphref{Mean number of rules used per log line for each Postfix
-        component}).
+    \item The number of log lines and rules used per log line for each
+        Postfix component (\graphref{Mean number of rules used per log line
+        for each Postfix component}).
 
@@
         ruleset) or reverse (reverse of optimal, least efficient).  The
-        results of this are shown in \sectionref{rule ordering for
-        efficiency}.
+        results of parsing with the three different orderings are shown in
+        \sectionref{rule ordering for efficiency}.
 
@@
         loaded; this optimisation can be disabled and the regex compiled
-        every time it is used when recognising log lines.  The effect of
-        this optimisation is shown in \sectionref{Caching compiled
+        every time it is used when recognising log lines.  The effect this
+        optimisation has is addressed shown in \sectionref{Caching compiled
         regexes}.
@@
 The framework also provides several debugging options, to aid in writing or
-correcting rules, or figuring out why the parser is not behaving as
+correcting rules, or figuring out why \parsername{} is not behaving as
 expected.  In increasing order of severity they are:
@@
     \item Each result can be extended with extra debugging information,
-        which is useful when a warning dumps a mail's data for inspection.
-        The extra information added is: the log line's timestamp in human
-        readable form; the entire log line; the name of the log file and
-        the number of the log line within it.  This information is not
-        stored in the database.
+        which is useful when a warning dumps a data structure for
+        inspection.  The extra information added is: the log line's
+        timestamp in human readable form; the entire log line; the name of
+        the log file and the line number of the log line within it.  This
+        extra information is not stored in the database.
 
@@
 recognised each log line.  Option 4 is not useful for most users: it is for
-debugging problems with the data that is saved to the database, and
-sufficient safeguards are in place that there should be plenty of warning
-messages explaining what is wrong without using this option.
+debugging problems within \parsername{}, and sufficient safeguards are in
+place that there should be plenty of warning messages explaining what is
+wrong without using this option.
 
@@
 be parsed without any effects; it does not imply that the input is
-ungrammatical or unparsed.  Generally rules specifying the
+ungrammatical or unparsed.  Generally, rules specifying the
 \action{UNINTERESTING} action recognise log lines that are not associated
@@
 these log lines are recognised and processed so that the framework can warn
-about unrecognised log lines, informing the user that they need to extend
+about unrecognised log lines, informing the user that they need to augment
 the ruleset.  Most of the remaining actions have only one or two associated
-rules, because they will only ever have one or two log line variants, e.g.\
-all log lines showing that a remote client has connected are recognised by
-a single rule and processed by the \action{CONNECT} action.
+rules, because that input category will only ever have one or two log line
+variants, e.g.\ all log lines showing that a remote client has connected
+are recognised by a single rule and processed by the \action{CONNECT}
+action.
 
@@
 \sectionref{parser efficiency}, \textit{excluding\/} log files 22 \&
-62--68, because their content is extremely skewed by two mail loops.  The
-action most commonly specified by rules, \action{DELIVERY\_REJECTED}, is
-the third most commonly invoked action; the most commonly invoked actions,
+62--68, because their contents are extremely skewed by two mail loops,
+described in \sectionref{Characteristics of the 93 log files}.  The action
+most commonly specified by rules, \action{DELIVERY\_REJECTED}, is the third
+most commonly invoked action; the most commonly invoked actions,
 \action{CONNECT} and \action{DISCONNECT}, are each specified by only one
-rule.  The \action{UNINTERESTING} action, the third most commonly specified
-action, is halfway down the graph.
-
-As expected, some actions have been invoked almost exactly the same number
-of times: almost every \action{CONNECT} will have a \action{DISCONNECT},
-with only a 0.00042\% difference between the number of times the two were
-invoked; every mail that enters Postfix's queue will be managed by
-\daemon{qmgr} (\action{MAIL\_QUEUED}) and processed by \daemon{cleanup}
-(\action{CLEANUP\_PROCESSING}), and again the two have only a 0.01648\%
-difference between their number of invocations.
+rule.  The \action{SAVE\_DATA} and \action{UNINTERESTING} actions, the
+second and third most commonly specified actions respectively, are halfway
+down the graph.
+
+As should be expected, some actions have been invoked almost exactly the
+same number of times: almost every \action{CONNECT} will have a
+\action{DISCONNECT}, with only a 0.00042\% difference between the number of
+times the two were invoked; every mail that enters Postfix's queue will be
+managed by \daemon{qmgr} (\action{MAIL\_QUEUED}) and processed by
+\daemon{cleanup} (\action{CLEANUP\_PROCESSING}), and again the two have
+only a 0.01648\% difference between their number of invocations.
 \action{CLEANUP\_PROCESSING} is invoked slightly more often than
@@
 from one action to another as its log lines are recognised.  The words
-\textit{mail\/} and \textit{connection\/} are used in the actions
+\textit{mail\/} and \textit{connection\/} are used in the action
 descriptions below because they are less unwieldy and more specific than
@@
 \action{CLONE} action, which processes Postfix accepting a delivery
-attempt, and the data structure moves from the connections state table to
-the queueids state table.
+attempt, and the data structure is copied from the connections state table
+to the queueids state table.
 
@@
 or mail, each action will save all the data captured by the recognising
-rule's regex; usually, log lines that lack identifying information will be
-processed by the \action{UNINTERESTING} action.  Each action is passed the
-same arguments:
+rule's regex to the connection or mail; usually, log lines that lack
+identifying information will be processed by the \action{UNINTERESTING}
+action.  Each action is passed the same arguments:
 
@@
 \end{boldeqlist}
+
+\noindent{}\textbf{\parsernames{} actions:}
 
@@
         lines.  This action creates a new mail if necessary; if the mail
-        already exists the unknown origin flag will be removed from it.
-        The mail will be marked as a bounce notification.  To deal with
-        complication \sectionref{Bounce notification mails delivered before
-        their creation is logged}, this action checks a cache of recent
-        bounce mails, to avoid incorrectly creating bounce mails when log
-        lines are out of order.
+        already exists the unknown origin flag will be removed from it.  To
+        deal with complication \sectionref{Bounce notification mails
+        delivered before their creation is logged}, this action checks a
+        cache of recent bounce mails, to avoid incorrectly creating bounce
+        mails when log lines are out of order, and also marks the mail as a
+        bounce notification.
 
@@
         will be postponed if the mail is a child waiting to be tracked
-        (\sectionref{Re-injected mails}).  Once entered in the database,
+        (\sectionref{Re-injected mails}).  Once entered into the database,
         the mail will be usually be deleted from the state tables, but
@@
 
-    \item [SAVE\_DATA] Every action that can locate the correct entry in
-        the state tables saves any data captured by the recognising rule's
-        regex to it.  The \action{SAVE\_DATA} saves data in this way but
-        does not do anything else; it is invoked for log lines that contain
-        useful data but do not require any further processing.
+    \item [SAVE\_DATA] Every action that can locate the correct data
+        structure in the state tables saves any data captured by the
+        recognising rule's regex to it.  The \action{SAVE\_DATA} saves data
+        in this way but does not do anything else; it is invoked for log
+        lines that contain useful data but do not require any further
+        processing.
 
@@
 
-Adding new actions is not as facile as adding new rules, though care has
+Adding new actions is not as simple as adding new rules, though care has
 been taken in the architecture and implementation to make adding new
@@
 arguments given to actions, and registers it with the framework by calling
-the framework's \texttt{add\_actions()} subroutine.  The new action must be
-registered before the rules are loaded, because it is an error for a rule
-to specify an unregistered action; this helps catch mistakes made when
-adding new rules.  No other work is required from the implementer to
-integrate the action into the parser; all of their attention and effort can
-be focused on correctly implementing their new action.  The action may need
-to extend the list of valid combinations described in \sectionref{out of
-order log lines} if the new action creates a different set of acceptable
-programs, but this would only be necessary if the new action processes log
-lines from Postfix components that \parsername{} does not have rules for,
-e.g.\ \daemon{virtual} or \daemon{lmtp}.
+the framework's \texttt{add\_actions()} subroutine.  No other work is
+required from the implementer to integrate the action into the parser; all
+of their attention and effort can be focused on correctly implementing
+their new action.  The action may need to extend the list of valid
+combinations described in \sectionref{out of order log lines} if the new
+action creates a different set of acceptable programs, but this would only
+be necessary if the new action processes log lines from Postfix components
+that \parsername{} does not have rules for, e.g.\ \daemon{virtual} or
+\daemon{lmtp}.  The new action must be registered before the rules are
+loaded, because it is an error for a rule to specify an unregistered
+action; this helps catch mistakes made when adding new rules.
 
@@
 correct action to be invoked.  The rules will be the most visible component
-in any parser implemented using this architecture, and also the component
-most likely to be modified by users.  Rules need to be as simple as
-possible so that users can easily modify them or add new rules, but each
-implementation must balance that simplicity with the need to provide enough
-flexibility and power to successfully parse inputs.
+in any parser based on this architecture, and also the component most
+likely to be modified by users.  Rules need to be as simple as possible so
+that users can easily modify them or add new rules, but each parser must
+balance that simplicity with the need to provide enough flexibility and
+power to successfully parse inputs.
 
@@
 \sectionref{rules table}; that description will not be repeated here, but
-should be fleshed out by the example rule in \sectionref{example rule in
+should be fleshed out by the sample rule in \sectionref{example rule in
 implementation}.  The process of creating new rules from unparsed log lines
 is dealt with in \sectionref{creating new rules in implementation},
-followed by the implementation of the utility supplied with \parsername{}
-to create regexes from unparsed log lines; the regex components provided by
-\parsername{} to ease writing of complex regexes are covered in
+followed by the algorithm used by the utility supplied with \parsername{}
+that creates regexes from unparsed log lines; the regex components provided
+by \parsername{} to ease writing of complex regexes are covered in
 \sectionref{regex components}.  The rule conditions used in \parsername{}
 are the penultimate topic (\sectionref{rule conditions in implementation}),
-and the section concludes with some suggestions for how to detect
+and this section concludes with some suggestions for how to detect
 overlapping rules.
@@
 the sender's domain (for full details
-see~\cite{reject-unknown-sender-domain}).  A sample log line that would be
-recognised by this rule:
+see~\cite{reject-unknown-sender-domain}).  An example log line that would
+be recognised by this rule:
 
@@
     \empty{}\label{Example rule in implementation table}
+    \centering{}
     \begin{tabular}{ll}
         \tabletopline{}%
-        \textbf{Field}      & \textbf{Value}                                    \\
+        Attribute           & Value                                             \\
         \tablemiddleline{}%
@@
 
-\noindent{}The fields in \tableref{Example rule in implementation table}
-are used as follows:
+\noindent{}The attributes in \tableref{Example rule in implementation
+table} are used as follows:
 
@@
 
-    \item [program and regex:] The program is used to restrict the log
-        lines this rule will be used to recognise; see \sectionref{rule
+    \item [program and regex:] program is used to restrict the log lines
+        this rule will be used to recognise; see \sectionref{rule
         conditions in implementation} for details.  The regex does the
-        actual recognition of the log lines, and data captured by the regex
+        actual recognition of log lines, and data captured by the regex
         (e.g.\ sender, recipient) will be automatically saved to the
@@
     \item [result\_data and connection\_data:] are used to provide data not
-        present in the log line, and are unused in this rule.
+        present in the log line, but are unused in this rule.
 
@@
 description, program, and regex; all the other attributes are either
-optional, set by the parser, or have sensible defaults.  The name and
+optional, are set by the parser, or have sensible defaults.  The name and
 description fields should be set based on the meaning of the log line, to
-help others understand which log lines this rule will recognise; the
-program will be obvious from the unrecognised log lines.  The action
-depends on the what the log line represents, e.g.\ a delivery rejection, a
-mail being delivered, some useful information, or something else; examine
-the list of actions in \sectionref{actions in detail in implementation} to
-determine the correct one.  The regex needs to be constructed based on the
-log line, but see below for a tool to ease the process.
-
-Other attributes may be required to make a rule work correctly:
-connection\_data, result\_data, priority, or restriction\_name.  In general
-it will only become clear that connection\_data or result\_data are
-required when \parsername{} warns about an entry in the connections or
-results tables that is missing some required fields, because values for
-those fields are not present in any of the log lines for that connection or
-mail.  For example, the rule that recognises the \daemon{pickup} component
-processing a mail sets client\_hostname to \texttt{localhost} and
-client\_ip to \texttt{127.0.0.1}, because the mail originates on the local
-server.  If the new rule deliberately uses the architecture's overlapping
-rules feature the priority field needs to be set, on this rule and possibly
-others; the priority field may be needed on unintentionally overlapping
-rules too, but that is more difficult to determine.  Finally, the
-restriction\_name field should be set if the rule's action is
-\action{DELIVERY\_REJECTED}; the name of the restriction should be clear
-from the content of the log line.
+help others understand which log lines this rule will recognise; the value
+for the program field will be obvious from the unrecognised log lines.  The
+action depends on the what the log line represents, e.g.\ a delivery
+rejection, a mail being delivered, some useful information, or something
+else; examine the list of actions in \sectionref{actions in detail in
+implementation} to determine the correct one.  The regex needs to be based
+on the log line, but see below for a tool to ease the process of creating
+regexes from unrecognised log lines.
+
+Other attributes may be required: connection\_data, result\_data, priority,
+or restriction\_name.  In general it will only become clear that
+connection\_data or result\_data are required when \parsername{} warns
+about an entry in the connections or results tables that is missing some
+required fields, because values for those fields are not present in any of
+the log lines for that connection or mail.  For example, the rule that
+recognises the \daemon{pickup} component processing a mail sets
+client\_hostname to \texttt{localhost} and client\_ip to
+\texttt{127.0.0.1}, because the mail originates on the local server.  If
+the new rule deliberately uses the architecture's overlapping rules feature
+the priority field needs to be set, on this rule and possibly others; the
+priority field may be needed on unintentionally overlapping rules too, but
+that is more difficult to determine.  Finally, the restriction\_name field
+should be set if the rule's action is \action{DELIVERY\_REJECTED}; the name
+of the restriction should be clear from the content of the log line.
 
@@
 Risto Vaarandi for his \acronym{SLCT}~\cite{slct-paper}.  The differences
-between the two algorithms will be outlined as part of the general
-explanation below.  The core of the \acronym{SLCT} algorithm is quite
-simple: programs generally create log lines by substituting variable words
-into a fixed pattern, and analysis of the frequency with which each word
-occurs can be used to determine whether the word is variable or part of the
-fixed pattern.  This classification can be used to group similar log lines
-and generate a regex to match each group of log lines.  The algorithm has
-five steps:
+between the two algorithms will be outlined as part of the explanation
+below.  The core of the \acronym{SLCT} algorithm is quite simple: programs
+generally create log lines by substituting variable words into a fixed
+pattern, and analysis of the frequency with which each word occurs can be
+used to determine whether the word is variable or part of the fixed
+pattern.  This classification can be used to group similar log lines and
+generate a regex to match each group of log lines.  The algorithm has five
+steps:
 
@@
     \item [Pre-process the file.]  The modified algorithm begins by
-        leveraging the knowledge gained when writing rules and developing
+        leveraging the knowledge gained from writing rules and developing
         \parsername{}; it performs many substitutions on the input log
@@
         a counter for each hash.  The counters will be used in the next
-        pass to filter out words: if the word's hash does not have a high
-        frequency, the word itself cannot have a high frequency, so it must
-        be variable and does not need a counter maintained for it, reducing
-        the number of counters required and thus the program's memory
-        consumption.
+        pass to filter out words.
 
@@
         The original algorithm hashes each word and checks that result's
-        counter from the previous pass: if that counter has a high
-        frequency a separate counter will be maintained for this
-        \texttt{(word, word's position within the log line)} tuple; if not,
-        the tuple will not have a counter.  This reduces the number of
-        counters maintained in this step, reducing memory requirements at
-        the cost of increased CPU usage.
+        counter from the previous pass: if the word's hash does not have a
+        high frequency, the word itself cannot have a high frequency, so it
+        must be variable and does not need a counter maintained for the
+        \texttt{(word, word's position within the log line)} tuple,
+        reducing the number of counters required and thus the program's
+        memory consumption.  This reduces the number of counters maintained
+        in this step, reducing memory requirements at the cost of increased
+        CPU usage.
 
@@
         each regex to check they are valid.  A match will be attempted
-        between all of the new regexes and each of the unparsed log lines
-        the regexes were built from; the user will be warned if a log line
-        is not matched by any regex, or if a log line is matched by more
-        than one regex.  The number of log lines each regex matches is
-        counted, as is the number of log lines matched by all regexes,
-        though a log line is counted once only, even if matched by more
-        than one regex.
+        between all of the new regexes and each of the original unparsed
+        log lines (not the pre-processed log lines); the user will be
+        warned if a log line is not matched by any regex, or if a log line
+        is matched by more than one regex.  The number of log lines each
+        regex matches is counted, as is the number of log lines matched by
+        all regexes, though a log line is counted once only, even if
+        matched by more than one regex.
 
@@
     \item [\_\_RESTRICTION\_START\_\_]  Matches the standard information
-        Postfix includes at the start of almost all log lines resulting
-        from rejecting a delivery attempt.
+        Postfix includes at the start of most log lines resulting from
+        rejecting a delivery attempt.
 
@@
 informing them that they need to augment their ruleset, and alerting them
-that because \parsername{} failed to recognise some of their log lines, the
-results stored in the database may be incomplete.
+that the results stored in the database may be incomplete because
+\parsername{} failed to recognise some log lines.
 
@@
 that responsibility is left to the author of the rules.  A mechanism is
-provided for ordering overlapping rules: the priority field in each rule;
-negative priorities may be useful for catchall rules.
+provided for ruleset authors to specify the order that overlapping rules
+will be tried in: the priority field in each rule.  Negative priorities may
+be useful for catchall rules.
 
@@
 
-    \item Sort rules by program and regex, then visually inspect the list,
-        e.g.\ with an \acronym{SQL} query similar to:   \newline{}
+    \item Sort rules by program and regex e.g.\ with an \acronym{SQL} query
+        similar to:                                     \newline{}
         \verb!SELECT program, regex!                    \newline{}
@@
         been recognised by program-specific rules.
+
+        Inspect the list to see if any pair of regexes look suspiciously
+        similar; overlapping regexes will often lie beside one another when
+        sorted.
 
@@
         examining the differences in the tables: each result contains a
-        reference to the rule which created it, which will change if that
+        reference to the rule that created it, which will change if that
         rule overlaps with another.  Unfortunately this method cannot prove
@@
 
-\subsection{Limitations}
-
 \label{logging helo}
@@
     \item The hostname used in the HELO command is not logged if the
-        incoming delivery attempt is successful.  Configuring Postfix to do
-        this is relatively simple; create a restriction that is guaranteed
-        to warn for every accepted mail, as follows:
+        incoming delivery attempt is successful.  Configuring Postfix to
+        log the HELO hostname for accepted mails is relatively simple;
+        create a restriction that is guaranteed to warn for every accepted
+        mail, as follows:
 
@@
         recipient there will not yet be a queueid allocated for the
-        delivery attempt, so the log line cannot be associated with the
-        correct mail.  A queueid is guaranteed to have been allocated when
-        the DATA command has been reached, and thus the queueid will be
-        logged by any restrictions taking effect in
+        delivery attempt, so the log line will be associated with the
+        connections rather than the mail.  A queueid is guaranteed to have
+        been allocated when the DATA command has been reached, and thus the
+        queueid will be logged by any restrictions taking effect in
         \texttt{smtpd\_data\_restrictions}, and the log line can be
@@
 
-    \item The solution to complication \sectionref{Mails deleted during
-        delivery} involves discarding data.
-
     \item Further complications may arise when parsing log files, and
@@
         that the user must manage the size of the database.
-
-\end{enumerate}
-
-\subsection{Possible Improvements}
-
-\begin{enumerate}
-
-    \item Improve the solution to complication \sectionref{Mails deleted
-        during delivery} so that data is not discarded.
 
@@
 any further use of that data is dependent on a clear understanding of the
-database schema, so the resemblance between a database schema and an
-\acronym{API} is described, followed by a diagram and a detailed
-description of the database schema.  The framework is documented next,
-including the steps it takes during initialisation, the parsing process,
-and the conveniences it offers to users.  The performance data collected by
-the framework is described, as are the ways in which various optimisations
-can be disabled to demonstrate their effect, and the debugging options the
-framework provides.  The implementation of actions in \parsername{} is
-documented, including how frequently each action is specified by rules, and
-brief descriptions of why some actions are more popular than others.  All
-the actions that are part of \parsername{} are described in detail,
-followed by the process of adding a new action to the parser.  The final
-component of the architecture, the rules, is also the most visible
-component, and its implementation is examined in detail with: an example
-rule and an explanation of how every field in that rule is used; a
-description of how to add new rules and determine the values that should be
-used for each field; an explanation of the algorithm used by the utility
-that creates new regexes based on unparsed log lines, and how it differs
-from the original algorithm it is based on.  How \parsername{} uses rule
-conditions and overlapping rules is discussed, accompanied by a description
-of the regex snippets the framework provides to simplify the regexes used
-in rules.  The many complications and difficulties encountered while
-writing \parsername{}, and the solutions developed to overcome them, are
-documented in depth; also documented are how some solutions interact, and
-which action or actions the solution is implemented in.  This chapter
-concludes with a list of the limitations identified in \parsername{},
-followed by some possible improvements suggested by the author.  The next
-chapter will evaluate this implementation, examining both \parsernames{}
-efficiency and the coverage it achieves when parsing log files.
+database schema, so the role of the database schema as an \acronym{API} is
+described, followed by a diagram and a detailed description of the database
+schema.  The framework is documented next, including the steps it takes
+during initialisation, the parsing process, and the conveniences it offers
+to users.  The performance data collected by the framework is described, as
+are the ways in which various optimisations can be disabled to demonstrate
+their effect, and the debugging options the framework provides.  The
+implementation of actions in \parsername{} is documented, including how
+frequently each action is specified by rules, and brief descriptions of why
+some actions are more popular than others.  All the actions that are part
+of \parsername{} are described in detail, followed by the process of adding
+a new action to the parser.  The final component of the architecture, the
+rules, is also the most visible component, and its implementation is
+examined in detail with: a sample rule and an explanation of how every
+field in that rule is used; a description of how to add new rules and
+determine the values that should be used for each field; an explanation of
+the algorithm used by the utility that creates new regexes from unparsed
+log lines, and how it differs from the original algorithm it is based on.
+How \parsername{} uses rule conditions and overlapping rules is discussed,
+accompanied by a description of the regex snippets the framework provides
+to simplify the regexes used in rules.  The many complications and
+difficulties encountered while writing \parsername{}, and the solutions
+developed to overcome them, are documented in depth; also documented are
+how solutions interact, and which action or actions each solution is
+implemented in.  This chapter concludes with a list of the limitations
+identified in \parsername{}.  The next chapter will evaluate this
+implementation, examining both \parsernames{} efficiency and the coverage
+it achieves when parsing log files.

### 2009-04-26 doc/logparser-implementation.tex
Changes based on comments from Ariane.

@@
 of times each action was invoked when parsing \numberOFlogFILES{} log
-files, excluding log files 22 \& 62--68}{Number of times each action was
-invoked excluding 22 and 62--68}
+files, excluding log files 22 \& 62--68 because their contents are
+extremely skewed by mail loops}{Number of times each action was invoked
+excluding 22 and 62--68}
+
+\FloatBarrier{}
 
@@
     \item [MAIL\_DISCARDED] Sometimes mail is discarded by Postfix, e.g.\
-        mail submitted locally that is larger than the limit configured by
-        the administrator.  This action is used for those mails; it invokes
-        the \action{COMMIT} action, but is a separate action to simplify
-        further analysis.
+        mail submitted locally on the server using \daemon{pickup} that is
+        larger than the limit configured by the administrator.  This action
+        is used for those mails; it invokes the \action{COMMIT} action, but
+        is a separate action to simplify further analysis.
 
@@
 
-    \item [MAIL\_TOO\_LARGE] When a client tries to send a mail larger than
-        the local server accepts, the mail will be discarded by Postfix and
-        the client informed of the problem.  The mail may have been
-        accepted and partially transferred; if so the parser will have a
-        data structure that must be disposed of.  See \sectionref{timeouts
-        during data phase} for full details; although that describes
-        timeouts, the processing is the same for mails that are too large.
+    \item [MAIL\_TOO\_LARGE] When a remote client tries to send a mail over
+        \acronym{SMTP} that is larger than the server accepts, the mail
+        will be discarded by Postfix and the client informed of the
+        problem.  The mail may have been accepted and partially
+        transferred; if so the parser will have a data structure that must
+        be disposed of.  See \sectionref{timeouts during data phase} for
+        full details; although that describes timeouts, the processing is
+        the same for mails that are too large.
 
@@
 \sectionref{rule conditions in architecture}, and they can be very complex
-and difficult to evaluate.  In contrast, \parsername{} uses quite simple
-rule conditions: each rule has a \texttt{program} attribute that specifies
+and difficult to evaluate.  Complex conditions are not required in parsers
+based on this architecture: \parsername{} uses quite simple
+rule conditions.  Each rule has a \texttt{program} attribute that specifies
 the Postfix component whose log lines it recognises, and each log line
@@
         grow without bounds unless the user deletes connections and results
-        from it.  This is both a limitation and a benefit: the benefit is
+        from it.  This is both a benefit and a limitation: the benefit is
         that data will never be unexpectedly deleted, but the limitation is

### 2009-04-26 doc/logparser-results.tex
Changes based on proofreading chapters 4 and 5.
Changes based on comments from Ariane.

@@
 
+The standard deviation for mails received on weekend days is
+quite high, because approximately four times the usual number of mails was
+received on the weekend contained in the $16^{th}$ and $17^{th}$ log files.
+The standard deviation for all days is high because the high number of
+mails received during log files 16, 17, 35, 36, 39, and 40.  Interestingly,
+the extra mail received during log files 35, 36, 39, and 40 stopped during
+log files 37 and 38, which may indicate that it was a virus infected office
+machine that was turned off over the weekend.  Unfortunately, the source
+cannot be investigated properly because all the mails in question were
+relayed via the department's secondary MX (a backup mail server, hosted
+elsewhere), so the log files showing the original source were unavailable.
+
 \showgraph{build/graph-mails-received}{Number of mails received via SMTP
@@
 \subsection{Characteristics Of The \numberOFlogFILES{} Log Files}
+
+\label{Characteristics of the 93 log files}
 
@@
 
+XXX TAKE UP SOME SPACE\@.
+
+XXX TAKE UP SOME SPACE\@.
+
+XXX TAKE UP SOME SPACE\@.
+
 \showgraph{build/graph-perfect-best-and-optimal-vs-shuffled}{Parsing time
@@
         parsing time, it would be only slightly more effective than optimal
-        ordering, but it would be vastly harder to implement; optimising
-        actions would not reduce parsing time enough to justify the amount
-        of effort required.
+        ordering, but it would be vastly more difficult to implement;
+        optimising actions would not reduce parsing time enough to justify
+        the amount of effort required.
 

### 2009-04-27 doc/logparser-results.tex
Corrections to chapter 6 from proofreading.

@@
 because the data gathered by the parser must be accurate and complete for
-it to be useful and reliable.  The two goals are mutually incompatible:
-every extra measure implemented to improve accuracy slows down parsing.
-Some of the complications described in \sectionref{complications} occur
-fewer than ten times when parsing the \numberOFlogFILES{} log files used to
-evaluate \parsername{} in this chapter, yet their solutions slow down
-parsing of every log file.  However, accurate but slow parsing is
-preferable to sloppy but quick parsing.
+it to be useful and reliable.  Progress in achieving one goal usually comes
+at the expense of the other: every extra measure implemented to improve
+accuracy slows down parsing.  Some of the complications described in
+\sectionref{complications} occur fewer than ten times when parsing the
+\numberOFlogFILES{} log files used to evaluate \parsername{} in this
+chapter, yet their solutions slow down parsing of every log file.  However,
+those solutions are retained in \parsername{}, because accurate but slow
+parsing is preferable to sloppy but quick parsing.
 
 The performance evaluation begins by describing the characteristics of the
-mail server that produced the log files used to evaluate the parser's
-performance and coverage; the computer that the tests were run on is also
-described.  The characteristics of the \numberOFlogFILES{} log files are
-described next, with an explanation of why \parsername{} has better
-performance when parsing the larger log files in the group.  The effect of
-using different rule orderings is explored, and the use of optimal rule
-ordering is compared to use of an oracle that allows the parser to use only
-one rule when recognising each log line.  When \parsername{} is used by
-other mail administrators, they will need to extend the ruleset to parse
-their own log lines, so the next section addresses the question of how
-parser performance is affected by an increase in the number of rules in the
-ruleset.  The penultimate topic covered is the simple optimisation of
-caching the results of compiling each rule's regex, and the huge effect it
-has on parser efficiency.  The performance evaluation concludes by
-examining where parsing time is spent: recognition of log lines or their
-subsequent processing?
+mail server that produced the \numberOFlogFILES{} log files used to
+evaluate the parser's performance and coverage; the computer that the tests
+were run on is also described.  The characteristics of the
+\numberOFlogFILES{} log files are described next, with an explanation of
+why \parsername{} has better performance when parsing the larger log files
+in the group.  The effect of using different rule orderings is explored,
+and optimal rule ordering is compared to an oracle that allows the parser
+to use only one rule when recognising each log line.  When \parsername{} is
+used by other mail administrators, they will need to extend the ruleset to
+parse their own log lines, so the next section addresses the question of
+how parser performance is affected by an increase in the number of rules in
+the ruleset.  The penultimate topic is the simple optimisation of caching
+the results of compiling each rule's regex, and the huge effect it has on
+parser efficiency.  The performance evaluation concludes by examining where
+parsing time is spent: recognition of log lines or their subsequent
+processing by actions?
 
 The second criterion the parser is evaluated on is its coverage of Postfix
-log files.  This subject contains two sections: what proportion of log
-lines are correctly recognised by the ruleset, and what proportion of mail
-delivery attempts are correctly understood and reconstructed by the
-actions?  The former is a requirement for the latter to be achieved, and
-the latter is important because the data provided by \parsername{} must be
-both complete and correct for it to be of use to others.
+log files.  There are two types of coverage to be evaluated: what
+proportion of log lines are correctly recognised by the ruleset, and what
+proportion of mail delivery attempts are correctly understood and
+reconstructed by the actions?  The former is a requirement for the latter
+to be achieved, and the latter is important because the data provided by
+\parsername{} must be both complete and correct for it to be of use to
+others.
 
@@
 large log files.  The mail server that generated the \numberOFlogFILES{}
-log files is a production mail server handling mail for a university
-department; the benefit of using this mail server is that its log files
-exhibit the idiosyncrasies and peculiarities that a mail server in the wild
-must deal with, but the downside is that significantly altering the
-configuration to accommodate this project is not an option.
+log files used in this chapter is a production mail server handling mail
+for a university department; the benefit of using this mail server is that
+its log files exhibit the idiosyncrasies and peculiarities that a mail
+server in the wild must deal with, but the downside is that significantly
+altering the configuration to accommodate this project is not an option.
 \Graphref{Mails received per day} and \tableref{Number of mails received
 per day: statistics} show that, as expected, it accepts far more mails on
-weekdays than at weekends.  Note that mail generated on the server does not
-contribute to the figures: in particular, the vast amount of mail resulting
-from the mail loops noticeable in later graphs is not reflected in these
-figures.
+weekdays than at weekends.  Note that mail generated on the server (bounce
+notifications, mail re-injected for forwarding, mail sent from mailing
+lists) does not contribute to these figures: in particular, the vast amount
+of mail resulting from the mail loops noticeable in later graphs is not
+reflected in these figures.
 
@@
 When generating the timing data used in this section, \numberOFlogFILES{}
-log files were used, each containing one day's log lines, totalling three
-months of contiguous log files,
-\input{build/include-total-size-of-93-log-files}, \numberOFlogLINEShuman{}
-log lines.  Each log file was parsed 10 times, and the mean parsing time
-calculated.  The computer used for test runs was a Dell Optiplex 745, shown
-in \tableref{Details of the computer used to generate statistics}; it was
+log files were used, each containing one day's log lines; the three months
+of contiguous log files contain \numberOFlogLINEShuman{} log lines and
+total \input{build/include-total-size-of-93-log-files}.  Each log file was
+parsed 10 times, and the mean parsing time calculated for each log file.
+The computer used for test runs was a Dell Optiplex 745, shown in
+\tableref{Details of the computer used to generate statistics}; it was
 dedicated to the task of gathering statistics from test runs, and did not
@@
 \input{build/include-insert-results-throughput} --- parsing takes up only
-\input{build/include-skip-inserting-as-percentage-of-inserting} of the
+\input{build/include-skip-inserting-as-percentage-of-inserting} of
 execution time when saving results to the database.
@@
 of \input{build/include-median-mails-accepted-over-smtp} mails accepted
-each weekday; large scale mail servers would generate much larger log
-files.  \Graphref{parsing time vs log file size vs number of log lines
-graph} shows the parsing time in seconds, log file size in MB, and number
-of log lines in tens of thousands, for each of the \numberOFlogFILES{} log
-files.  Parsing time is plotted against number of log lines in
-\graphref{Parsing time plotted against number of log lines}, and against
-log file size in \graphref{Parsing time plotted against log file size};
-the points on both graphs form almost perfectly straight lines.
+each weekday; large scale mail servers would accept many more mails and
+consequently generate much larger log files.  \Graphref{parsing time vs log
+file size vs number of log lines graph} shows the parsing time in seconds,
+log file size in MB, and number of log lines in tens of thousands, for each
+of the \numberOFlogFILES{} log files.  Parsing time is plotted against
+number of log lines in \graphref{Parsing time plotted against number of log
+lines}, and against log file size in \graphref{Parsing time plotted against
+log file size}; the points on both graphs appear to form straight lines,
+but they actually curve downwards slightly as they move to the right.
 
-\Graphref{parsing time vs log file size vs number of log lines
-factor} plots both the ratio of log file size vs parsing time, and the
-ratio of number of log lines vs parsing time (higher is more efficient in
-both cases); \tableref{parsing time vs log file size vs number of log lines
+\Graphref{parsing time vs log file size vs number of log lines factor}
+plots both the ratio of log file size to parsing time, and the ratio of
+number of log lines to parsing time (higher is more efficient in both
+cases); \tableref{parsing time vs log file size vs number of log lines
 factor table} shows the same ratios for different groups of log files.  The
@@
 delivery attempts from other hosts, but when the mail loops occurred most
-of the log lines resulted from failed delivery of mail generated on the
-server itself.  The Postfix components that generated most of the log lines
+of the log lines resulted from failed delivery of mail re-injected for
+forwarding.  The Postfix components that generated most of the log lines
 during the mail loops have fewer associated rules than the Postfix
 components whose log lines normally make up the bulk of each log file, so
-the mean number of rules tried per log line is lower for these log files,
-as is the mean parsing time per log line.  \Tableref{Number of rules per
+the mean number of rules used per log line is lower for these log files, as
+is the mean parsing time per log line.  \Tableref{Number of rules per
 Postfix component} shows the number of rules for each Postfix component;
 \graphref{Mean number of rules used per log line} shows the drop in the
-mean number of rules tried per log line for log files containing a mail
+mean number of rules used per log line for log files containing a mail
 loop, and \graphref{Mean number of rules used per log line for each Postfix
-component} shows the mean number of rules tried per log line for each
+component} shows the mean number of rules used per log line for each
 Postfix component for all log files.
@@
     \item [Shuffle] This ordering is intended to represent a randomly
-        ordered rule set.  The rules will be shuffled once before use and
+        ordered ruleset.  The rules will be shuffled once before use and
         will retain that ordering until the parser exits.  Note that the
@@
 
-The parsing time of the three orderings are plotted against log file size
+The parsing times of the three orderings are plotted against log file size
 in \graphref{Parsing time plotted against log file size for optimal,
@@
 \input{build/include-optimal-ordering-parsing-time-reduction-other-logs}\%
-with normal log files,
-\input{build/include-optimal-ordering-parsing-time-reduction-logs-22-62-68}\%
-when a mail loop occurs and the distribution of log lines is unusual.
-\Tableref{Parsing time of optimal and reverse orderings relative to
-shuffled ordering table} shows that differences in rule ordering have less
-effect on parsing time when parsing log files 22 \& 62--68, because of the
-different distribution of log lines in those log files.  A careful
-examination of \graphref{Parsing time of optimal and reverse orderings
-relative to shuffled ordering graph} shows that, for the first log file
-only, optimal and reverse orderings perform identically: this is because
-the hits field of each rule is zero for the first log file, so optimal and
-reverse orderings produce identical rule orderings.  For the first log
-file, shuffled ordering is the most efficient of the three, but that is
-accidental and cannot be relied upon.
+with normal log files.  \Tableref{Parsing time of optimal and reverse
+orderings relative to shuffled ordering table} shows that differences in
+rule ordering have less effect on parsing time when parsing log files 22 \&
+62--68, because of the different distribution of log lines in those log
+files.  A careful examination of \graphref{Parsing time of optimal and
+reverse orderings relative to shuffled ordering graph} shows that, for the
+first log file only, optimal and reverse orderings perform identically:
+this is because the hits field of each rule is zero for the first log file,
+so optimal and reverse orderings produce identical rule orderings.  For the
+first log file, shuffled ordering is the most efficient of the three, but
+that is accidental and cannot be relied upon.
 
@@
 
-Optimal rule ordering, as described in \sectionref{rule ordering for
+Optimal rule ordering, described in \sectionref{rule ordering for
 efficiency}, is the best rule ordering it is possible to achieve without
@@
 
-\Graphref{Parsing time of oracle and optimal ordering relative to shuffled
-ordering graph} shows how the oracle and optimal ordering perform relative
-to shuffled ordering, with mean and standard deviation in \tableref{Parsing
-time of oracle and optimal ordering relative to shuffled ordering table}.
-As expected, the oracle is more efficient than optimal ordering, but not by
-much.  \Graphref{Percentage increase in parsing time when using optimal
-ordering instead of oracle graph} shows the percentage increase in parsing
-time when using optimal ordering instead of the oracle, with mean and
-standard deviation in \tableref{Percentage increase in parsing time when
-using optimal ordering instead of oracle table}.
+\Graphref{Parsing time of the oracle and optimal ordering relative to
+shuffled ordering graph} shows how the oracle and optimal ordering perform
+relative to shuffled ordering, with mean and standard deviation in
+\tableref{Parsing time of the oracle and optimal ordering relative to
+shuffled ordering table}; \graphref{Parsing time plotted against log file
+size for the oracle, optimal ordering, and shuffled ordering} shows parsing
+time plotted against log file size for the oracle, optimal ordering, and
+shuffled ordering.  As expected, the oracle is more efficient than optimal
+ordering, but not by much, particularly when parsing the larger log files.
+\Graphref{Percentage increase in parsing time when using optimal ordering
+instead of the oracle graph} shows the percentage increase in parsing time
+when using optimal ordering instead of the oracle, with mean and standard
+deviation in \tableref{Percentage increase in parsing time when using
+optimal ordering instead of the oracle table}.
 
@@
 Optimal ordering proves to be quite efficient: \tableref{Percentage
-increase in parsing time when using optimal ordering instead of oracle
+increase in parsing time when using optimal ordering instead of the oracle
 table} shows that optimal ordering is less than
@@
 
-XXX TAKE UP SOME SPACE\@.
-
-XXX TAKE UP SOME SPACE\@.
-
-XXX TAKE UP SOME SPACE\@.
-
 \showgraph{build/graph-perfect-best-and-optimal-vs-shuffled}{Parsing time
-of oracle and optimal ordering relative to shuffled ordering}{Parsing time
-of oracle and optimal ordering relative to shuffled ordering graph}
+of the oracle and optimal ordering relative to shuffled ordering}{Parsing
+time of the oracle and optimal ordering relative to shuffled ordering
+graph}
 
 \showtable{build/include-perfect-best-and-optimal-vs-shuffle}{Parsing time
-of oracle and optimal ordering relative to shuffled ordering}{Parsing time
-of oracle and optimal ordering relative to shuffled ordering table}
+of the oracle and optimal ordering relative to shuffled ordering}{Parsing
+time of the oracle and optimal ordering relative to shuffled ordering
+table}
+
+\showgraph{build/graph-file-size-oracle-optimal-shuffle}{Parsing time
+plotted against log file size for the oracle, optimal ordering, and
+shuffled ordering}{Parsing time plotted against log file size for the
+oracle, optimal ordering, and shuffled ordering}
 
 \showgraph{build/graph-perfect-best-vs-optimal}{Percentage increase in
-parsing time when using optimal ordering instead of oracle}{Percentage
-increase in parsing time when using optimal ordering instead of oracle
+parsing time when using optimal ordering instead of the oracle}{Percentage
+increase in parsing time when using optimal ordering instead of the oracle
 graph}
@@
 \showtable{build/include-perfect-best-vs-optimal-stddev}{Percentage
-increase in parsing time when using optimal ordering instead of
+increase in parsing time when using optimal ordering instead of the
 oracle}{Percentage increase in parsing time when using optimal ordering
-instead of oracle table}
+instead of the oracle table}
 
@@
 full ruleset.  The full ruleset is larger because \parsername{} is tested
-with \numberOFlogFILESall{} log files; testing with more log files
-increases the chance of finding bugs in the parser or new complications to
-be overcome.  The \numberOFlogFILES{} log files were each parsed 10 times
-using the minimum ruleset, and the mean parsing times compared to those
-generated using the full ruleset: the percentage parsing time increase when
-using the full ruleset instead of the minimal ruleset for optimal,
-shuffled, and reverse orderings is shown in \graphref{Percentage parsing
-time increase when using the maximum ruleset instead of the minimum
-ruleset}, with mean and standard deviation in \tableref{Percentage parsing
-time increase when using the maximum ruleset instead of the minimum ruleset
-table}.  Note that in both \graphref{Percentage parsing time increase when
-using the maximum ruleset instead of the minimum ruleset} and
+with \numberOFlogFILESall{} log files (\numberOFlogFILESallYEARS{} of log
+files); testing with more log files increases the chance of finding bugs in
+the parser or new complications to be overcome.  The \numberOFlogFILES{}
+log files were each parsed 10 times using the minimum ruleset, and the mean
+parsing times compared to those generated using the full ruleset: the
+percentage parsing time increase when using the full ruleset instead of the
+minimal ruleset for optimal, shuffled, and reverse orderings is shown in
+\graphref{Percentage parsing time increase when using the maximum ruleset
+instead of the minimum ruleset}, with mean and standard deviation in
 \tableref{Percentage parsing time increase when using the maximum ruleset
-instead of the minimum ruleset table}, for each ordering the parsing time
-using the maximum ruleset is compared to the ordering's parsing time using
-the minimum ruleset: the three orderings are not compared to a common
-baseline.
+instead of the minimum ruleset table}.  For each ordering, the parsing time
+using the maximum ruleset is compared to the parsing time using the minimum
+ruleset: the three orderings are not compared to a common baseline.
 
-Clearly the increased number of rules causes a noticeable performance
+Clearly, the increased number of rules causes a noticeable performance
 decrease with reverse ordering, and a lesser decrease with shuffled
@@
 performance for the first log file, because the hits field of every rule
-starts at zero.
+starts at zero, and the two rule orderings are identical.
 
@@
 
-\Graphref{Percentage increase in parsing time when not caching compiled
-regexes graph} shows the effect that not caching compiled regexes has on
-parser performance, with mean and standard deviation in
-\tableref{Percentage increase in parsing time when not caching compiled
-regexes table}.  For typical log files, the mean increase in parsing time
-when not caching compiled regexes is
-\input{build/include-cached-regexes-vs-discarded-regexes-mean}.  Caching
-compiled regexes is probably the single most effective optimisation
+A test run using optimal ordering was performed as described in
+\sectionref{parser efficiency}, with one difference: regexes were not
+compiled and cached when the ruleset was loaded from the database, so
+Perl's regex engine had to compile each regex each time it was used when
+trying to recognise a log line.  \Graphref{Percentage increase in parsing
+time when not caching compiled regexes graph} shows the effect that not
+caching compiled regexes has on parser performance, with mean and standard
+deviation in \tableref{Percentage increase in parsing time when not caching
+compiled regexes table}.  For typical log files, the mean increase in
+parsing time when not caching compiled regexes is
+\input{build/include-cached-regexes-vs-discarded-regexes-mean}; looking at
+it from the opposite direction, caching compiled regexes reduces parsing
+time by
+\input{build/include-cached-regexes-vs-discarded-regexes-mean-reduction}.
+Caching compiled regexes is probably the single most effective optimisation
 possible in \parsername{}, and was quite simple to implement: the framework
@@
 in parsing time when the optimisation is disabled; this is because, on
-average, fewer rules are used so fewer regexes are compiled per log line
-for those log files.  The increase in parsing time when parsing the first
-log file is much greater than for the other log files; again, this is
+average, fewer rules are used per log line, so fewer regexes are compiled
+per log line for those log files.
+
+The increase in parsing time when parsing the first log file is much
+greater than for the other log files (see \graphref{Percentage increase in
+parsing time when not caching compiled regexes graph}); again, this is
 because every rule's hits field starts at zero, so optimal ordering is less
@@
 higher than usual, and more regexes will need to be compiled when
-recognising each log line.
+recognising each log line.  Parsing time is plotted against log file size
+for cached and discarded regexes in \graphref{Parsing time plotted against
+log file size for cached regexes and discarded regexes}; the plot for
+discarding compiled regexes is quite uneven compared to caching compiled
+regexes --- some log files are particularly slow, whereas others are not as
+badly affected, but these differences have not been investigated.
 
@@
 
+\showgraph{build/graph-file-size-optimal-discarded}{Parsing time plotted
+against log file size for cached regexes and discarded regexes}{Parsing
+time plotted against log file size for cached regexes and discarded
+regexes}
+
 \FloatBarrier{}
 
-\subsection{Where Is The Time Spent: Recognising Or Processing Log Lines?}
+\subsection{Where Is Parsing Time Spent: Recognising Or Processing Log
+Lines?}
 
@@
 
-The optimisations described in this chapter have addresses the process of
+The optimisations described in this chapter have addressed the process of
 recognising log lines, but have not optimised actions at all.  Optimisation
@@
         taken by invocation of actions; the mean parsing times for each log
-        file were subtracted from the mean parsing time from a normal test
-        run to obtain the parsing time taken by actions.
+        file were subtracted from the mean parsing times from a normal test
+        run to obtain the parsing time taken by actions.  The parsing time
+        exclude the time taken for framework initialisation, loading and
+        saving state tables, and loading the ruleset: in ad far as
+        possible, the times are just for recognising log lines or
+        recognising log lines plus invoking actions.
 
@@
 
-Individual actions or the framework could and have been optimised, but
+Individual actions and the framework could and have been optimised, but
 plenty of existing literature is available on the topic of optimising
@@
 
-    \item Try each rule for that Postfix component, then generic rules,
-        until a recognising rule is found; if the log line is not
-        recognised, issue a warning and move on to the next log line.
+    \item Try each rule that recognises log lines from that Postfix
+        component, then any generic rules, until a recognising rule is
+        found; if the log line is not recognised, issue a warning and move
+        on to the next log line.
 
@@
 Each Postfix component whose log lines are of interest must have at least
-one rule, or all of its log lines will be silently skipped; in the extreme
-case of an empty ruleset the parser would skip every log line.
-\parsername{} skips log lines logged by programs that do not have any
-associated rules, because there may be any number of log lines from other
-programs in the log file, and some Postfix components do not produce any
-log lines of interest.  \parsername{} does not parse log lines from
-non-Postfix programs, e.g.\ Amavisd-new or SpamAssassin; it could easily be
-extended to do so, if a method could be developed to correctly associate
-such log lines with existing state table entries.  To correctly recognise
-all log lines, there must be a rule to recognise each log line variant
-produced by each Postfix component; if a log line is not recognised the
-parser will issue a warning, to inform the user that they need to extend
-their ruleset.  Each rule's regex should be as specific and precise as
-possible, to ensure accurate parsing: a rule with a regex that matches zero
-or more of any character will recognise every log line, but not in a
+one rule that recognises its log lines, or all of its log lines will be
+silently skipped; in the extreme case of an empty ruleset the parser would
+skip every log line.  \parsername{} skips log lines from programs that do
+not have any associated rules, because there may be any number of log lines
+from other programs in the log file, and some Postfix components do not
+produce any log lines of interest.  \parsername{} does not parse log lines
+from non-Postfix programs, e.g.\ Amavisd-new or SpamAssassin; it could
+easily be extended to do so, if a method could be developed to correctly
+associate such log lines with existing state table entries.  To correctly
+recognise all log lines, there must be a rule to recognise each log line
+variant produced by each Postfix component; if a log line is not recognised
+the parser will issue a warning, to inform the user that they need to
+extend their ruleset.  Each rule's regex should be as specific and precise
+as possible, to ensure accurate parsing: a rule with a regex that matches
+zero or more of any character will recognise every log line, but not in a
 meaningful way; most log lines contain fixed strings, so this is not a
@@
 Full coverage of log lines can be achieved without undue effort, yet once
-achieved it requires maintenance.  Maintaining full coverage takes effort because
-the set of log line variants changes over time, e.g.\ administrators add
-restrictions with custom messages, \acronym{DNSBL} messages change, or
-major releases of Postfix change log lines (usually by adding more
-information).  Warnings are issued for any log lines that are not
-recognised; no warnings are issued for unrecognised log lines while parsing
-the \numberOFlogFILES{} log files, so it can be safely concluded that zero
-false negatives arise.  False positives are harder to quantify, short of
-examining each of the 60,721,709 log lines and checking that the correct
-rule recognised it.  However, a random sample of 6039 log lines was parsed,
-and the results manually verified by inspection to ensure that the correct
-rule recognised each log line.  The sample was generated by running the
-command: \texttt{perl -n -e \singlequote{}print if (rand 1 <
-0.0001)\singlequote{} LOG\_FILES} to randomly extract roughly one log line
-in every 10,000 (it actually extracted 0.00994\% instead of 0.01\%).  Each
-log line was examined and the correct rule identified from the
-\numberOFrules{} rules in the database; the correct rule was then compared
-to the rule that recognised the log line when parsing.  The sample results
-contained zero false positives, and this check has been automated to ensure
-continued accuracy.  Based on these results, and how precise each rule's
-regex is, the author is confident that zero false positives occur when
-parsing the \numberOFlogFILES{} log files.
+achieved it requires maintenance.  Maintaining full coverage is an ongoing
+task because the set of log line variants changes over time, e.g.\
+administrators add restrictions with custom messages, \acronym{DNSBL}
+messages change, or major releases of Postfix change log lines (usually by
+adding more information).  Warnings are issued for any log lines that are
+not recognised; no warnings are issued for unrecognised log lines while
+parsing the \numberOFlogFILES{} log files, so it can be safely concluded
+that zero false negatives arise.  False positives are harder to quantify,
+short of examining each of the 60,721,709 log lines and checking that the
+correct rule recognised it.  However, a random sample of 6039 log lines was
+parsed, and the results manually verified by inspection to ensure that the
+correct rule recognised each log line.  The sample was generated by running
+the command: \newline{} \tab{} \texttt{perl -n -e \singlequote{}print if
+(rand 1 < 0.0001)\singlequote{} LOG\_FILES} \newline{} to randomly extract
+roughly one log line in every 10,000 (it actually extracted 0.00994\%
+instead of 0.010\%).  Each log line was examined and the correct rule
+identified from the \numberOFrules{} rules in the database; the correct
+rule was then compared to the rule that recognised the log line when
+parsing.  The sample results contained zero false positives, and this check
+has been automated to ensure continued accuracy.  Based on these results,
+and how precise each rule's regex is, the author is confident that zero
+false positives occur when parsing the \numberOFlogFILES{} log files.
 
@@
 warnings about a missing queueid or \acronym{pid} when parsing the first
-few thousand log lines, because the earlier log lines for those connections
+few thousand log lines, because some earlier log lines for those connections
 or mails are in a previous log file; loading saved state tables from the
@@
 
-The data used in this chapter is generated by parsing \numberOFlogFILES{}
-log files.  5 warnings are produced, but because \parsername{} errs on the
-side of producing more warnings rather than fewer, those 5 warnings
+5 warnings are produced when parsing the \numberOFlogFILES{} log files to
+generate the data used in this chapter, but because \parsername{} errs on
+the side of producing more warnings rather than fewer, those 5 warnings
 represent 3 instances of 1 problem: 3 connections that started before the
@@
 parser finishes parsing the log file.  Ideally, those are the only entries
-they will contain, though they may also contain mails whose initial log
-lines are not contained in the log files.  Any other entries in the state
-tables are evidence of either a failure in parsing, or an aberration in the
-log files.  After parsing the \numberOFlogFILES{} log files, the state
-tables contain 18 entries:
+the state tables will contain, though they may also contain mails whose
+initial log lines are not contained in the log files.  Any other entries in
+the state tables are evidence of either a failure in parsing, or an
+aberration in the log files.  After parsing the \numberOFlogFILES{} log
+files, the state tables contain 18 entries:
 
@@
     \item 9 mails whose initial log lines were not present in the log
-        files.  These mails did not produce warnings because they resemble
-        child mails waiting to be tracked with a parent; see
+        files.  Some of those mails did not produce warnings because they
+        resemble child mails waiting to be tracked with a parent; see
         \sectionref{Re-injected mails} for details.
@@
     \item 7 mails that had yet to be delivered because of repeated
-        failures.
+        delivery failures.
 
@@
 
-    \item \parsername{} performs many checks to detect known problems
-        occurring, e.g.\ if a mail is missing required data.  No such
-        warnings are produced during the test runs other than the five
-        described above.
+    \item \parsername{} performs many checks to detect known problems,
+        e.g.\ if a mail is missing required data.  No such warnings are
+        produced during the test runs other than the five described above.
 
@@
         to one mail or connection respectively; it is extremely unlikely
-        that a log line would be associated with the wrong connection.
+        that a log line would not be associated with the right connection.
 
@@
 
-%\verb!perl -Mstrict -Mwarnings -MList::Util=shuffle -e! \newline{}
-%\verb!     '@ARGV = [shuffle(@ARGV)]->[0];!             \newline{}
-%\verb!      while (<>) {!                               \newline{}
-%\verb!          if (rand 1 < 0.0001) {!                 \newline{}
-%\verb!              foreach my $i (1 .!.  6000) {!        \newline{}
-%\verb!                  print scalar <>;!               \newline{}
-%\verb!              }!                                  \newline{}
-%\verb!              exit;!                              \newline{}
-%\verb!          }!                                      \newline{}
-%\verb!      }' LOG_FILES!
+\renewcommand{\figurename}{Figure}
+
+\begin{figure}[thbp]
+    \caption{The command used to extract the log segment used to verify
+        \parsernameshort{'s} parsing}
+    \empty{}\label{Command to extract log segment used to verify parsing}
+
+\begin{verbatim}
+perl -Mstrict -Mwarnings -e '
+    while (<>) {
+        if (rand 1 < 0.0001) {
+            my $count = 0;
+            while ($count < 6000) {
+                print scalar <>;
+                $count++;
+            }
+            exit;
+        }
+    }' LOG_FILES
+\end{verbatim}
+
+\end{figure}
 
@@
 sample from those log files is a tractable albeit extremely time consuming
-task.  A sample of log lines was obtained by randomly selecting a log file,
-and then randomly selecting a block of 6000 contiguous log lines from it
-(0.00988\% of the total number of log lines).  It is important that the log
-lines are contiguous, so that all log lines are present for as many of the
-mail delivery attempts contained in the block as possible.  This log
-segment was parsed with all debugging options enabled, resulting in 167,448
-lines of output.\footnote{A mean of 27.908 lines of output per log line;
-each connection has 30 debugging lines, plus 21 debugging lines per result.
-Connections which have been cloned will have the cloned connection in their
-debugging output, plus another 33 debugging lines.  Those numbers are
-approximate, and may vary $\pm{}$ 2.  An approximate linear relationship
-between the number of log lines and debugging lines is: $33(connections) +
-30(accepted~~mails) + 21(results)$.} All 167,448 lines were examined in
-conjunction with the log file segment and a dump of the resulting database,
-verifying that \parsername{} recognised each of the log lines with the
-correct rule and invoked the correct action, which in turn correctly
-processed the log line and saved the correct data to be inserted into the
-database.  The log file segment produced 4 warnings, 10 mails correctly
-remaining in the state tables, 1625 mail delivery attempts correctly
-entered in the database, 0 false positives, and 0 false negatives.
+task.  A sample of log lines was obtained by randomly selecting a block of
+6000 contiguous log lines from the \numberOFlogFILES{} log files (0.00988\%
+of the total number of log lines), using the command shown in
+\figureref{Command to extract log segment used to verify parsing}.  It is
+important that the log lines are contiguous, so that all log lines are
+present for as many of the mail delivery attempts contained in the block as
+possible.  This log segment was parsed with all debugging options enabled,
+resulting in 167,448 lines of output.\footnote{A mean of 27.908 lines of
+output per log line; each connection has 30 debugging lines, plus 21
+debugging lines per result.  Connections which have been cloned will have
+the cloned connection in their debugging output, plus another 33 debugging
+lines.  Those numbers are approximate, and may vary $\pm{}$ 2.  An
+approximate linear relationship between the number of log lines and
+debugging lines is: $33(connections) + 30(accepted~~mails) + 21(results)$.}
+All 167,448 lines were examined in conjunction with the log file segment
+and a dump of the resulting database, verifying that \parsername{}
+recognised each of the log lines with the correct rule and invoked the
+correct action, which in turn correctly processed the log line and saved
+the correct data to be inserted into the database.  The log file segment
+produced 4 warnings, 10 mails correctly remaining in the state tables, 1625
+mail delivery attempts correctly entered in the database, 0 false
+positives, and 0 false negatives.
 

### 2009-04-27 doc/logparser-results.tex
Changes made to chapter 6 after proofreading.

@@
 server in the wild must deal with, but the downside is that significantly
-altering the configuration to accommodate this project is not an option.
-\Graphref{Mails received per day} and \tableref{Number of mails received
-per day: statistics} show that, as expected, it accepts far more mails on
-weekdays than at weekends.  Note that mail generated on the server (bounce
-notifications, mail re-injected for forwarding, mail sent from mailing
-lists) does not contribute to these figures: in particular, the vast amount
-of mail resulting from the mail loops noticeable in later graphs is not
-reflected in these figures.
+altering its Postfix configuration to accommodate this project is not an
+option.  \Graphref{Mails received per day} and \tableref{Number of mails
+received per day: statistics} show that, as expected, it accepts far more
+mails on weekdays than at weekends.  Note that mail generated on the server
+(e.g.\ bounce notifications, mail re-injected for forwarding, mail sent
+from mailing lists) does not contribute to these figures: in particular,
+the vast amounts of mail resulting from the mail loops noticeable in later
+graphs are not reflected in these figures.
 
-The standard deviation for mails received on weekend days is
-quite high, because approximately four times the usual number of mails was
-received on the weekend contained in the $16^{th}$ and $17^{th}$ log files.
-The standard deviation for all days is high because the high number of
-mails received during log files 16, 17, 35, 36, 39, and 40.  Interestingly,
-the extra mail received during log files 35, 36, 39, and 40 stopped during
-log files 37 and 38, which may indicate that it was a virus infected office
-machine that was turned off over the weekend.  Unfortunately, the source
-cannot be investigated properly because all the mails in question were
-relayed via the department's secondary MX (a backup mail server, hosted
+The standard deviation for mails received on weekend days is quite high,
+because approximately four times the usual number of mails were received on
+the weekend contained in log files 16 \& 17.  The standard deviation for
+all days is high because of the unusually high number of mails received
+during log files 16, 17, 35, 36, 39 \& 40.  Interestingly, the extra mails
+received during log files 35, 36, 39 \& 40 stopped during log files 37 \&
+38, which may indicate that the source was a virus infected office machine
+that was turned off over the weekend.  Unfortunately, the source cannot be
+investigated properly because all of the mails in question were relayed via
+the department's secondary \acronym{MTA} (a backup mail server, hosted
 elsewhere), so the log files showing the original source were unavailable.
+Relaying spam mail through a secondary \acronym{MTA} is a common practice
+amongst spam senders, because historically, primary \acronyms{MTA} had
+better anti-spam defences than secondary \acronyms{MTA}.
 
@@
 cases); \tableref{parsing time vs log file size vs number of log lines
-factor table} shows the same ratios for different groups of log files.  The
-ratios are quite tightly banded, except they increase (i.e.\ improve) for
-log files 22 \& 62--68, despite their larger than usual size.  Both groups
-of log files are much larger than usual because of mail loops caused by
-users who set up mail forwarding incorrectly, resulting in a very different
-distribution of log lines: normally most log lines are generated by mail
-delivery attempts from other hosts, but when the mail loops occurred most
-of the log lines resulted from failed delivery of mail re-injected for
-forwarding.  The Postfix components that generated most of the log lines
-during the mail loops have fewer associated rules than the Postfix
-components whose log lines normally make up the bulk of each log file, so
-the mean number of rules used per log line is lower for these log files, as
-is the mean parsing time per log line.  \Tableref{Number of rules per
-Postfix component} shows the number of rules for each Postfix component;
+factor table} shows mean and standard deviation.  The ratios are quite
+tightly banded, except they increase (i.e.\ improve) for log files 22 \&
+62--68, despite their larger than usual size.  The log files in both groups
+are much larger than usual because of mail loops caused by users who set up
+mail forwarding incorrectly, resulting in a very different distribution of
+log lines: normally most log lines are generated by mail delivery attempts
+from other hosts, but during the mail loops most of the log lines resulted
+from failed delivery of bounce notifications re-injected for forwarding.
+The Postfix components that generated most of the log lines during the mail
+loops have fewer associated rules than the Postfix components whose log
+lines normally make up the bulk of each log file, so the mean number of
+rules used per log line is lower for these log files, as is the mean
+parsing time per log line.  \Tableref{Number of rules per Postfix
+component} shows the number of rules for each Postfix component;
 \graphref{Mean number of rules used per log line} shows the drop in the
@@
 component} shows the mean number of rules used per log line for each
-Postfix component for all log files.
+Postfix component for \numberOFlogFILES{} log files.
 
@@
 read.  Log files 62--68 show a large gap between log file size and number
-of log lines, whereas those plots appear to have a small gap or to overlap
-for all the other log files, suggesting that the ratio of log file size to
-number of log lines is lower for log files 62--68.  This appearance is
-misleading: \graphref{Mean log line size in bytes} shows the mean log line
-size for all \numberOFlogFILES{} log files, and it reduces for log files 22
-\& 62--68, i.e.\ there are more log lines per MB in those log files.
+of log lines, whereas those plots appear to have a small gap for all the
+other log files, suggesting that the ratio of number of log lines to log
+file size is lower for log files 62--68 (i.e.\ fewer log lines per MB).
+This appearance is misleading: \graphref{Mean log line size in bytes} shows
+the mean log line size for all \numberOFlogFILES{} log files, and it
+reduces for log files 22 \& 62--68, i.e.\ there are more log lines per MB
+in those log files.
 
@@
 
-\parsernames{} ruleset has \numberOFrules{} rules: the top 10\% recognise
-\input{build/include-top-ten-hits}\% of the log lines in the
+\parsernames{} ruleset has \numberOFrules{} rules: the top 10 rules
+recognise \input{build/include-top-ten-hits}\% of the log lines in the
 \numberOFlogFILES{} log files, with the remaining log lines split across
-the other 90\% of the rules, as shown in \graphref{rule hits graph}.
-Assuming that the distribution of log lines is reasonably consistent across
-log files, \parsernames{} efficiency should benefit from using rules that
-recognise log lines more frequently before it uses rules that recognise log
-lines less frequently.  To test this hypothesis, three full test runs were
-performed with different rule orderings:
+the other \numberOFrulesMINUSten{} of the rules, as shown in \graphref{rule
+hits graph}.  Assuming that the distribution of log lines is reasonably
+consistent across log files, \parsernames{} efficiency should benefit from
+using rules that recognise log lines more frequently before it uses rules
+that recognise log lines less frequently.  To test this hypothesis, three
+full test runs were performed with different rule orderings:
 
@@
 reverse orderings relative to shuffled ordering graph} shows the parsing
-times of optimal and reverse orderings relative to shuffled ordering, and
-mean relative parsing times for different groupings of log files are given
-in \tableref{Parsing time of optimal and reverse orderings relative to
-shuffled ordering table}.  This optimisation provides a mean reduction in
-parsing time of
+times of optimal and reverse orderings relative to shuffled ordering, with
+mean and standard deviation in \tableref{Parsing time of optimal and
+reverse orderings relative to shuffled ordering table}.  \Graphref{Number
+of rules used by optimal, shuffled, and reverse orderings} shows the number
+of rules used by each ordering while parsing the \numberOFlogFILES{} log
+files.  This optimisation provides a mean reduction in parsing time of
 \input{build/include-optimal-ordering-parsing-time-reduction-other-logs}\%
-with normal log files.  \Tableref{Parsing time of optimal and reverse
-orderings relative to shuffled ordering table} shows that differences in
-rule ordering have less effect on parsing time when parsing log files 22 \&
-62--68, because of the different distribution of log lines in those log
-files.  A careful examination of \graphref{Parsing time of optimal and
-reverse orderings relative to shuffled ordering graph} shows that, for the
-first log file only, optimal and reverse orderings perform identically:
-this is because the hits field of each rule is zero for the first log file,
-so optimal and reverse orderings produce identical rule orderings.  For the
-first log file, shuffled ordering is the most efficient of the three, but
-that is accidental and cannot be relied upon.
+with normal log files, making it a very worthwhile and effective
+optimisation.
+
+\Tableref{Parsing time of optimal and reverse orderings relative to
+shuffled ordering table} shows that differences in rule ordering have less
+effect on parsing time when parsing log files 22 \& 62--68, because of the
+different distribution of log lines in those log files.  A careful
+examination of \graphref{Parsing time of optimal and reverse orderings
+relative to shuffled ordering graph} shows that, for the first log file
+only, optimal and reverse orderings perform identically: this is because
+the hits field of each rule is zero for the first log file, so optimal and
+reverse orderings produce identical rule orderings.  For the first log
+file, shuffled ordering is the most efficient of the three, but that is
+accidental and cannot be relied upon.
 
@@
 of optimal and reverse orderings relative to shuffled ordering table}
+
+\showgraph{build/graph-number-or-rules-used-by-different-orderings}{Number
+of rules used by optimal, shuffled, and reverse orderings}{Number of rules
+used by optimal, shuffled, and reverse orderings}
 
@@
 \& 62--68), because the mean number of rules used per log line is lower
-(see \graphref{Mean number of rules used per log line}).  The performance
-of optimal ordering relative to the oracle is much worse when parsing the
-first log file than for the remainder of the log files, because the hits
-field of every rule starts at zero, so optimal ordering does not provide
-any benefit for the first log file; the oracle, in contrast, is flawless
-for every log file.
+when parsing these log files (see \graphref{Mean number of rules used per
+log line}).  The performance of optimal ordering relative to the oracle is
+much worse when parsing the first log file than for the remainder of the
+log files, because the hits field of every rule starts at zero, so optimal
+ordering does not provide any benefit for the first log file; the oracle,
+in contrast, is flawless for every log file.
 
@@
 performance for the first log file, because the hits field of every rule
-starts at zero, and the two rule orderings are identical.
+starts at zero, and so the two rule orderings are identical for the first
+log file.
 
@@
 \sectionref{parser efficiency}, with one difference: regexes were not
-compiled and cached when the ruleset was loaded from the database, so
-Perl's regex engine had to compile each regex each time it was used when
+compiled and cached when the ruleset was loaded from the database, so the
+Perl interpreter had to compile each regex each time it was used when
 trying to recognise a log line.  \Graphref{Percentage increase in parsing
@@
 recognising each log line.  Parsing time is plotted against log file size
-for cached and discarded regexes in \graphref{Parsing time plotted against
-log file size for cached regexes and discarded regexes}; the plot for
-discarding compiled regexes is quite uneven compared to caching compiled
-regexes --- some log files are particularly slow, whereas others are not as
-badly affected, but these differences have not been investigated.
+for caching and not caching compiled regexes in \graphref{Parsing time
+plotted against log file size when caching and not caching compiled
+regexes}; the plot for not caching compiled regexes is quite uneven
+compared to caching compiled regexes --- some log files are particularly
+slow, whereas others are not as badly affected, but these differences have
+not been investigated.
 
@@
 \showgraph{build/graph-file-size-optimal-discarded}{Parsing time plotted
-against log file size for cached regexes and discarded regexes}{Parsing
-time plotted against log file size for cached regexes and discarded
-regexes}
+against log file size when caching and not caching compiled
+regexes}{Parsing time plotted against log file size when caching and not
+caching compiled regexes}
 
@@
         of parsing time is spent recognising log lines.  To measure how
-        long recognition of log lines takes, a full test run was performed
-        as previously described in \sectionref{parser efficiency}, but with
-        one difference: actions were not invoked when a log line was
+        long recognition of log lines takes, a full test run using optimal
+        ordering and caching compiled regexes was performed as previously
+        described in \sectionref{parser efficiency}, but with one
+        difference: actions were not invoked when a log line was
         recognised, so the parsing time did not include the time normally
@@
         file were subtracted from the mean parsing times from a normal test
-        run to obtain the parsing time taken by actions.  The parsing time
+        run to obtain the parsing time taken by actions.  The parsing times
         exclude the time taken for framework initialisation, loading and
-        saving state tables, and loading the ruleset: in ad far as
-        possible, the times are just for recognising log lines or
-        recognising log lines plus invoking actions.
+        saving state tables, loading the ruleset, and other housekeeping
+        tasks: in as far as possible, the times are just for recognising
+        log lines or recognising log lines plus invoking actions.
 
@@
         reduction in parsing time; not invoking actions at all reduces the
-        most optimised parsing time by less than half that.  If processing
-        of recognised log lines was optimised to 1\% of its original
-        parsing time, it would be only slightly more effective than optimal
-        ordering, but it would be vastly more difficult to implement;
-        optimising actions would not reduce parsing time enough to justify
-        the amount of effort required.
+        most optimised parsing time by less than half that.  Optimising
+        actions to 1\% of their original parsing time would be only
+        slightly more effective than optimal ordering is, but would be
+        vastly more difficult to implement; optimising actions would not
+        reduce parsing time enough to justify the amount of effort
+        required.
 
@@
 \FloatBarrier{}
+
+\newpage{}
 
@@
 associate such log lines with existing state table entries.  To correctly
-recognise all log lines, there must be a rule to recognise each log line
-variant produced by each Postfix component; if a log line is not recognised
-the parser will issue a warning, to inform the user that they need to
-extend their ruleset.  Each rule's regex should be as specific and precise
-as possible, to ensure accurate parsing: a rule with a regex that matches
-zero or more of any character will recognise every log line, but not in a
-meaningful way; most log lines contain fixed strings, so this is not a
-problem in practice.
+recognise all log lines that are not skipped, there must be a rule to
+recognise each log line variant produced by each Postfix component; if a
+log line is not recognised the parser will issue a warning, to inform the
+user that they need to extend their ruleset.  Each rule's regex should be
+as specific and precise as possible, to ensure accurate parsing: a rule
+with a regex that matches zero or more of any character will recognise
+every log line, but not in a meaningful way; most log lines contain fixed
+strings, so this is not a problem in practice.
 
@@
 The proportion of mail delivery attempts that are correctly understood and
-reconstructed is much more difficult to determine accurately than the
-proportion of log lines that are correctly recognised.  The parser can dump
-its state tables in a human readable form; examining those tables with
-reference to the log files is the best way to detect mails that were not
-reconstructed properly; many of the complications discussed in
+reconstructed is more difficult to determine accurately than the proportion
+of log lines that are correctly recognised.  The parser can dump its state
+tables in a human readable form; examining those tables with reference to
+the log files and the database is the best way to detect mails that were
+not reconstructed properly; many of the complications documented in
 \sectionref{complications} were detected in this way.  \parsername{} issues
@@
 warnings about a missing queueid or \acronym{pid} when parsing the first
-few thousand log lines, because some earlier log lines for those connections
-or mails are in a previous log file; loading saved state tables from the
-previous log file will solve this problem.
+few thousand log lines, because some earlier log lines for those
+connections or mails are in a previous log file; loading state tables saved
+when parsing the log file containing those log lines will solve this
+problem.
 
@@
     \item 9 mails whose initial log lines were not present in the log
-        files.  Some of those mails did not produce warnings because they
+        files.  Six of those mails did not produce warnings because they
         resemble child mails waiting to be tracked with a parent; see
-        \sectionref{Re-injected mails} for details.
+        \sectionref{Re-injected mails} for details.  Three of these mails
+        were missing more log lines, and so did five produce warnings, as
+        documented previously.
 
@@
     \item \parsername{} performs many checks to detect known problems,
-        e.g.\ if a mail is missing required data.  No such warnings are
-        produced during the test runs other than the five described above.
+        e.g.\ a queueid missing from the state tables.  No such warnings
+        are produced during the test runs other than the five described
+        previously.
 
@@
         that every reconstructed mail must comply with to be acceptable,
-        not just the five mails that exhibit that complication.
+        not just the five mails exhibiting that complication.
 
@@
 \numberOFlogFILES{} log files used to generate statistics and test
-coverage, including why performance is better on the larger log files.  The
-framework optimises the order in which rules are used when trying to
-recognise each log line, and the effect that optimisation has on parsing
-time is explored; this is followed by comparing optimal ordering with an
-oracle.  How \parsername{} scales as the size of the ruleset increases is
-addressed, followed by a description and analysis of the simplest and most
-effective of the optimisations, caching compiled regexes, and the
-efficiency evaluation concludes with an examination of where parsing time
-is spent: recognising log lines or processing them?
+coverage, including why performance is better when parsing the larger log
+files.  The framework optimises the order in which rules are used when
+trying to recognise each log line, and the effect that optimisation has on
+parsing time is explored; this is followed by comparing optimal ordering
+with an oracle.  How \parsername{} scales as the size of the ruleset
+increases is addressed, followed by a description and analysis of the
+simplest and most effective of the optimisations, caching compiled regexes,
+and the efficiency evaluation concludes with an examination of where
+parsing time is spent: recognising log lines or processing them?
 

### 2009-04-28 doc/logparser-acronyms.tex
Changes made after proofreading the preface and chapters 1-3.
Changes made after proofreading chapter 7 and the appendices.

@@
         instead relying on the client to maintain it.  Typically the client
-        is an \acronym{MTA}, and the server a delivery agent or a mail
+        is an \acronym{MTA}, and the server is a delivery agent or a mail
         store.  Full details are available in~\cite{lmtp-rfc}.
@@
         services); the sending \acronym{MTA} transfers the mail to the
-        receiving \acronym{MTA}.
+        receiving \acronym{MTA}, which forwards the mail to another
+        recipient, or delivers the mail to a user's mailbox or a program
+        such as a mailing list manager.
     }
@@
         guaranteed to be unique for the lifetime of the process.  Once the
-        process has completed, the pid may be reused, because they are
-        drawn from a finite pool.
+        process has completed, the \acronym{pid} may be reused, because
+        they are drawn from a finite pool.
     }
@@
         more complicated grammars, so \acronyms{CFG} and \acronyms{PDA} are
-        equivalent.
+        equivalent in computational power.
     }
@@
         which state transition to take, the piece of data on the top of the
-        stack can be used in addition to the input.  In addition to the
-        languages recognised by \acronyms{FA}, \acronyms{PDA} can recognise
-        languages of the form $a^{n}b^{n}$, i.e.\ any number of one symbol
-        followed by the same number of another symbol.
+        stack can be used in addition to the input.  \acronym{PDA} can
+        recognise all the languages recognised by \acronym{FA}, and can
+        also recognised languages of the form $a^{n}b^{n}$.
     }

### 2009-04-28 doc/logparser-conclusion.tex
Changes made after proofreading chapter 7 and the appendices.

@@
 yacc~\cite{lex-and-yacc-book} being a common example), the parts are
-usually quite internally interdependent, and will be combined into a
-complete parser by the compilation process; in contrast \parsername{} keeps
-the rules and actions separate until the parser runs.  This separation
-enables the optimisations discussed in \sectionref{parser efficiency}, and
-it also allows different approaches to ruleset management, e.g.\ using
-machine learning techniques to seamlessly create or alter rules to
-recognise new inputs (\sectionref{adding new rules in architecture}).  The
-decoupling of rules from actions allows different sets of rules to be used
-with one set of actions, e.g.\ a parser might have actions to process
+usually quite internally interdependent, and will be combined by the
+compilation process; in contrast, \parsername{} keeps the rules and actions
+separate until the parser runs (\sectionref{framework in implementation}).
+This separation enables the optimisations discussed in \sectionref{parser
+efficiency}, and it also allows different approaches to ruleset management,
+e.g.\ using machine learning techniques to seamlessly create or alter rules
+to recognise new inputs (\sectionref{adding new rules in architecture}).
+The decoupling of rules from actions allows different sets of rules to be
+used with one set of actions, e.g.\ a parser might have actions to process
 versions one and two of a file format; by choosing the appropriate ruleset
@@
 successfully recognising almost all of the log lines, because most of the
-irregularities and complications explained in \sectionref{complications}
+irregularities and complications documented in \sectionref{complications}
 started to become apparent then.  Adding new rules to deal with numerous
@@
 and that order closely resembles the frequency in which they occur; the
-most frequently occurring complications dominate the warning messages
+most frequently occurring complications dominated the warning messages
 produced, and so naturally they were the first complications to be dealt
@@
 to recognise and reject spam rather than accepting it.  The database
-provides the data in a normalised form that is far easier to use as input
-to new or existing implementations of machine-learning algorithms than
-trying to adapt each algorithm to extract data directly from log files.
-New policy servers, written to implement new anti-spam measures, can be
-tested or trained by using the collected data to simulate mail delivery
-attempts; this would allow simple, fast, reproducible testing, without the
-risk of adversely affecting a production mail server.  Development of
-\parsername{} is finished, i.e.\ it correctly parses Postfix log files, and
-in future it will only require maintenance; however, one avenue of future
-development under consideration is to extend it to parse non-Postfix log
-lines, e.g.\ SpamAssassin or Amavisd-new log lines.  \parsername{} can
-easily be extended to do this, but it requires a method of associating the
-non-Postfix log lines with the existing data structures and state tables,
-so that all of the data for a mail delivery attempt can be stored together.
+(\sectionref{database}) provides the data in a normalised form that is far
+easier to use as input to new or existing implementations of
+machine-learning algorithms than trying to adapt each algorithm to extract
+data directly from log files.  New policy servers, written to implement new
+anti-spam measures, can be tested or trained by using the collected data to
+simulate mail delivery attempts; this would allow simple, fast,
+reproducible testing, without the risk of adversely affecting a production
+mail server.  Development of \parsername{} is finished, i.e.\ it correctly
+parses Postfix log files, and in future it will only require maintenance;
+however, one avenue of future development under consideration is to extend
+it to parse non-Postfix log lines, e.g.\ SpamAssassin or Amavisd-new log
+lines.  \parsername{} can easily be extended to do this, but it requires a
+method of associating the non-Postfix log lines with the existing data
+structures and state tables, so that all of the data for a mail delivery
+attempt can be stored together.
 

### 2009-04-28 doc/logparser-glossary.tex
Changes made after proofreading chapter 7 and the appendices.

@@
     When a spam sender or virus sends mail with forged sender addresses,
-    innocent servers are flooded with undeliverable mail notifications;
-    this is called backscatter.
+    innocent mail servers are flooded with undeliverable mail notifications
+    from badly configured mail servers that either do not verify recipient
+    addresses before accepting deliver attempts, or automatically reply to
+    mail identified as spam; this is called backscatter.
 }}
@@
     $<>$ is the sender address used for sending bounce notifications.  In
-    the \acronym{SMTP} conversation all addresses are enclosed in $<>$, so
+    the \acronym{SMTP} conversation, all addresses are enclosed in $<>$, so
     \texttt{username@domain} is sent as \texttt{$<$username@domain$>$};
     thus $<>$ is actually an empty address, but it is always written as
-    $<>$ for clarity.  All mail servers must accept mail sent from $<>$, or
-    they are in violation of \acronym{RFC}~2821~\cite{RFC2821}.
+    $<>$ for clarity.  Mail servers must not unconditionally reject mail
+    sent from $<>$, or they are in violation of
+    \acronym{RFC}~2821~\cite{RFC2821}.
 }}
@@
     efficient manner.  A full description is beyond the scope of this
-    thesis, further information can be found at~\cite{hash-functions}.
+    thesis, further information can be found in~\cite{hash-functions}.
 }}
@@
 \newglossaryentry{awk}{name={AWK},description={
-    AWK is a general purpose programming language that is designed for
-    processing text and is available as a standard utility on all Unix
-    systems.
+    AWK is a general purpose programming language designed for processing
+    text that is available as a standard utility on all Unix systems.
 }}
@@
 \newglossaryentry{syslog}{name={Syslog},description={
-    Syslog is the standard logging mechanism on Unix systems: a program
-    sends log messages to syslog, then syslog filters and stores the
-    messages as configured by the administrator.
+    Syslog is the standard logging mechanism used on Unix systems: a
+    program sends log messages to syslog, then syslog filters and stores
+    the messages as configured by the administrator.
 }}

### 2009-04-28 doc/logparser-postfix-daemons.tex
Changes made after proofreading chapter 7 and the appendices.

@@
     \daemon{pickup} and subsequent delivery.
-    \daemonDocURL[\newline{}]{http://www.postfix.org/postdrop.1.html}{2009/02/23}.
+    \daemonDocURL{http://www.postfix.org/postdrop.1.html}{2009/02/23}.
 }}
@@
     consistency checking of the mail queue.
-    \daemonDocURL[\newline{}]{http://www.postfix.org/postsuper.1.html}{2009/02/23}.
+    \daemonDocURL{http://www.postfix.org/postsuper.1.html}{2009/02/23}.
 }}
@@
     it executes \daemon{postdrop} to place a new mail in the queue.
-    \daemonDocURL[\newline{}]{http://www.postfix.org/sendmail.1.html}{2009/02/23}.
+    \daemonDocURL{http://www.postfix.org/sendmail.1.html}{2009/02/23}.
 }}

### 2009-04-28 doc/logparser-architecture.tex
Changes made after final checks.

@@
 
-    % Reduce inter-item spacing because each item is a single line.
     \squeezeitems{}
@@
 example, with most regex engines matching double quoted strings with
-\verb!\"([^"\\]+|\\.)*"\!  is very fast when there is a match, but when the
-match fails its computational complexity for a string of length $n$ is
+\verb!\"([^"\\]+|\\.)*"\!  is very fast when a match can be found, but when
+the match fails its computational complexity for a string of length $n$ is
 $O(2^{n})$; see~\cite{mastering-regular-expressions} for in-depth

### 2009-04-28 doc/logparser-background.tex
Changes made after final checks.

@@
         intention is to reduce spam from faked sender addresses,
-        backscatter, and joe~jobs~\glsadd{Joe-job}.  There has been
-        considerable resistance to \acronym{SPF} because it breaks or
-        vastly complicates some commonly-used features of \acronym{SMTP},
-        e.g.\ forwarding mail from one domain to another when a user moves.
+        backscatter\glsadd{Backscatter}, and joe~jobs\glsadd{Joe-job}.
+        There has been considerable resistance to \acronym{SPF} because it
+        breaks or vastly complicates some commonly-used features of
+        \acronym{SMTP}, e.g.\ forwarding mail from one domain to another
+        when a user moves.
 

### 2009-04-28 doc/logparser-implementation.tex
Changes made after final checks.

@@
 \parsername{} can parse log files from Postfix version 2.2 through to
-version 2.5, and should be able to parse log files from later versions with
-only minor modifications or an updated ruleset.  The parser's task is to
-follow the journey each mail takes through Postfix, combining the data
-captured by rules into a coherent whole, and saving it in a useful and
-consistent form.  The intermingling of
-log lines from different mails immediately rules out the possibility of
-handling each mail in isolation; the parser must handle multiple mails in
-parallel, each potentially at a different stage in its journey, without any
-interference between mails --- except in the minority of cases where
-intra-mail interference is required, e.g.\ mail re-injected for forwarding
-(\sectionref{Re-injected mails}).  The best way to deal with intermingling
-of log lines is to maintain state information for every unfinished mail,
-and to manipulate the appropriate mail correctly for each log line
-encountered.
+version 2.5, and should parse log files from later versions with only minor
+modifications or an updated ruleset.  The parser's task is to follow the
+journey each mail takes through Postfix, combining the data captured by
+rules into a coherent whole, and saving it in a useful and consistent form.
+The intermingling of log lines from different mails immediately rules out
+the possibility of handling each mail in isolation; the parser must handle
+multiple mails in parallel, each potentially at a different stage in its
+journey, without any interference between mails --- except in the minority
+of cases where intra-mail interference is required, e.g.\ mail re-injected
+for forwarding (\sectionref{Re-injected mails}).  The best way to deal with
+intermingling of log lines is to maintain state information for every
+unfinished mail, and to manipulate the appropriate mail correctly for each
+log line encountered.
 
@@
         either deliberately or accidentally (e.g.\ log file rotation gone
-        awry, file system filling up, logging system unable to cope with
-        the volume of log messages).  On a well run mail server it is
-        extremely unlikely that any of these problems will arise, though
-        the likelihood increases when suffering from a deluge of spam or a
-        mail loop.  When parsing individual log files in isolation, it is
-        highly likely that some mails will have log lines in previous log
-        files, and others will have log lines in subsequent log files; to
+        awry, filesystem filling up, logging system unable to cope with the
+        volume of log messages).  On a well run mail server it is extremely
+        unlikely that any of these problems will arise, though the
+        likelihood increases when suffering from a deluge of spam or a mail
+        loop.  When parsing individual log files in isolation, it is highly
+        likely that some mails will have log lines in previous log files,
+        and others will have log lines in subsequent log files; to
         alleviate this problem \parsername{} supports saving and loading
@@
 
-            \item [timestamp] The time the line was logged at.
+            \item [timestamp] The time the log line was written to the log
+                file.
 
@@
 
-            \item [pid] The \acronym{pid} of the process that logged the
-                line.
-
-            \item [host] The hostname of the server the line was logged on.
+            \item [pid] The \acronym{pid} of the process that generated the
+                log line.
+
+            \item [host] The hostname of the server the log line was
+                generated on.
 
@@
         expire and the \daemon{smtpd} will exit.  This occurs very
-        infrequently, because there are many other timeouts that should
-        occur in the intervening hours, e.g.\ timeouts for DNS requests or
-        timeouts reading data from the client.  The active connection for
-        that \daemon{smtpd} must be cleaned up, entered in the database,
-        and deleted from the state tables.
+        infrequently, because many other timeouts should occur in the
+        intervening hours, e.g.\ timeouts for DNS requests or timeouts
+        reading data from the client.  The active connection for that
+        \daemon{smtpd} must be cleaned up, entered in the database, and
+        deleted from the state tables.
 
@@
 
-    \item [Pre-process the file.]  The modified algorithm begins by
+    \item [Pre-process the input.]  The modified algorithm begins by
         leveraging the knowledge gained from writing rules and developing

### 2009-04-28 doc/logparser-other-parsers-review.tex
Changes made after final checks.

@@
 
-Similar to AWStats and Logrep, Lire attempts to correlate log lines by
-queueid, but not by \acronym{pid}, so the complete list of recipients for
-each delivered mail should be available; it does not attempt to deal with
-the other complications described in \sectionref{complications}.  When
-testing Lire on the \numberOFlogFILES{} test log files it performed
-reasonably well: the numbers it reports appear accurate, and the subset
-verified by the author were correct.  Its report provides summaries of:
+Similar to AWStats and Logrep (\sectionref{Logrep review}), Lire attempts
+to correlate log lines by queueid, but not by \acronym{pid}, so the
+complete list of recipients for each delivered mail should be available; it
+does not attempt to deal with the other complications described in
+\sectionref{complications}.  When testing Lire on the \numberOFlogFILES{}
+test log files it performed reasonably well: the numbers it reports appear
+accurate, and the subset verified by the author were correct.  Its report
+provides summaries of:
 
@@
 \section{Logrep}
+
+\label{Logrep review}
 

### 2009-04-28 doc/logparser-results.tex
Changes made after final checks.

@@
 The second criterion the parser is evaluated on is its coverage of Postfix
-log files.  There are two types of coverage to be evaluated: what
-proportion of log lines are correctly recognised by the ruleset, and what
-proportion of mail delivery attempts are correctly understood and
-reconstructed by the actions?  The former is a requirement for the latter
-to be achieved, and the latter is important because the data provided by
-\parsername{} must be both complete and correct for it to be of use to
-others.
+log files.  Two kinds of coverage need to be evaluated: what proportion of
+log lines are correctly recognised by the ruleset, and what proportion of
+mail delivery attempts are correctly understood and reconstructed by the
+actions?  The former is a requirement for the latter to be achieved, and
+the latter is important because the data provided by \parsername{} must be
+both complete and correct for it to be of use to others.
 

### 2009-04-28 doc/logparser-complications.tex
A change to the abstract suggested by Carl. Switch to one-and-a-half spacing for the initial submission.

@@
 Occasionally, a log file will have out of order log lines that cannot be
-dealt with by the techniques described in \sectionref{tracking re-injected
-mail}, \sectionref{discarding cleanup log lines}, or \sectionref{pickup
-logging after cleanup}.  In the \numberOFlogFILES{} log files used in
-\sectionref{Evaluation} this problem occurs only five times in 60,721,709
-log lines.  All five occurrences have the same characteristics: the
-\daemon{local} log line showing delivery to a user's mailbox comes after
-the \daemon{qmgr} log line showing removal of the mail from the queue
-because delivery is complete.  This causes several problems: the data in
-the state tables for the mail is not complete, so it cannot be entered into
-the database; a new mail is created when the \daemon{local} log line is
-processed and remains in the state tables; four warnings are issued per
-pair of out of order log lines.
+dealt with by the various techniques described in \sectionref{tracking
+re-injected mail}, \sectionref{discarding cleanup log lines}, or
+\sectionref{pickup logging after cleanup}.  In the \numberOFlogFILES{} log
+files used in \sectionref{Evaluation} this problem occurs only five times
+in 60,721,709 log lines.  All five occurrences have the same
+characteristics: the \daemon{local} log line showing delivery to a user's
+mailbox comes after the \daemon{qmgr} log line showing removal of the mail
+from the queue because delivery is complete.  This causes several problems:
+the data in the state tables for the mail is not complete, so it cannot be
+entered into the database; a new mail is created when the \daemon{local}
+log line is processed and remains in the state tables; four warnings are
+issued per pair of out of order log lines.
 

### 2009-04-28 doc/logparser.tex
Changes based on proofreading chapters 4 and 5.
Corrections to chapter 6 from proofreading.
Changes made to chapter 6 after proofreading.
Changes made after proofreading chapter 7 and the appendices.
Changes made after final checks.
Laid out my own abstract so that I could reuse it for my loose abstract page.
A change to the abstract suggested by Carl. Switch to one-and-a-half spacing for the initial submission.

@@
 \usepackage{lastpage}
+
+% College expect one and a half or double spacing for the initial
+% submission.
+\usepackage{setspace}
+\onehalfspacing{}
 
@@
     \hfill{} \newline{} \url{#2}%
-    \hfill{} #1 Last checked #3%
+    \hfill{} \newline{} #1 Last checked #3%
 }
@@
 
+\newcommand{\numberOFlogFILESallYEARS}[0]{%
+    2 years, 1\textonehalf{} months%
+}
+
 \newcommand{\numberOFrules}[0]{%
     184%
+}
+
+\newcommand{\numberOFrulesMINUSten}[0]{%
+    174%
 }
@@
 % The name of the program, so I only have to change it in one place.
-\newcommand{\parsername}{\acronym{PLP}}
-\newcommand{\parsernames}{\acronym['s]{PLP}}
-\newcommand{\parsernamelong}{Postfix Log Parser}
-\newcommand{\parsernameshort}{PLP}
+\newcommand{\parsername}[1][]{\acronym[#1]{PLP}}
+\newcommand{\parsernames}[0]{\acronym['s]{PLP}}
+\newcommand{\parsernamelong}[0]{Postfix Log Parser}
+\newcommand{\parsernameshort}[1][]{PLP#1}
 
@@
 
-\begin{document}
-
-\pagestyle{empty}
-
-\specialpage{Title}
-
+\newcommand{\titleandauthor}[0]{%
 \begin{center}
@@
     \vfill{}
+
+\end{center}
+}
+
+\begin{document}
+
+\pagestyle{empty}
+
+\specialpage{Title}
+
+\titleandauthor{}
+
+\begin{center}
 
@@
 \begin{flushright}
-    \underline{\hspace*{10em}} \\~\\
+    \underline{\hspace*{15em}} \\~\\
     John Tobin \\
@@
 \noindent{}I am grateful to my wife, Ariane Tobin, for her patience, good
-will, and encouragement.
+will, and encouragement; without her, I would not have accomplished this.
 
-\begin{abstract}
+\newpage{}
 
-    \setcounter{page}{4}
-    \phantomsection{}
-    \addcontentsline{toc}{chapter}{Abstract}
+% Abstract
+\specialpage{Abstract}
 
-    Parsing Postfix log files is much more difficult than it first appears,
-    but it \textit{is\/} possible to achieve a high degree of accuracy in
+~\empty{}
+
+\vfill{}
+\vfill{}
+\vfill{}
+
+%\titleandauthor{}
+
+\begin{center}
+    \textbf{Abstract}
+\end{center}
+
+Parsing Postfix log files is much more difficult than it first appears, but
+it \textit{is\/} possible to achieve a high degree of accuracy in
 understanding Postfix log files, and subsequently in reconstructing the
@@
 thesis describes the creation of a parser for Postfix log files,
-    documenting the architecture developed for this project and the parser
-    that implements it, the difficulties encountered and the solutions
-    developed.  The parser stores data gleaned from the log files in an SQL
-    database; the gathered data can be used to optimise current anti-spam
-    measures, to produce statistics showing how effective those measures
-    are, or to provide a baseline to test new anti-spam measures against.
-    The Postfix log file parser needs to be very precise and strict when
-    parsing, yet must allow users to easily adapt or extend it to parse new
-    log lines, without requiring that the user have an in-depth knowledge
-    and understanding of the parser's internal workings.  The newly
-    developed architecture is designed to make the process of parsing new
-    inputs as simple as possible, enabling users to trivially add new rules
-    (to parse variants of known inputs) and relatively easily add new
-    actions (to process a previously unknown category of inputs).  The
-    parser implemented for this project is evaluated on the criteria of
-    efficiency and coverage of Postfix log files, demonstrating that the
-    conflicting goals of efficiency and accuracy can be balanced, and that
-    one need not be sacrificed to achieve the other.
+documenting the architecture developed for this project and the parser that
+implements it, the difficulties encountered and the solutions developed.
+The parser stores data gleaned from the log files in an SQL database;
+future projects or programs could use the gathered data to optimise current
+anti-spam measures, to produce statistics showing how effective those
+measures are, or to provide a baseline to test new anti-spam measures
+against.  The Postfix log file parser needs to be very precise and strict
+when parsing, yet must allow users to easily adapt or extend it to parse
+new log lines, without requiring that the user have an in-depth knowledge
+and understanding of the parser's internal workings.  The newly developed
+architecture is designed to make the process of parsing new inputs as
+simple as possible, enabling users to trivially add new rules (to parse
+variants of known inputs) and relatively easily add new actions (to process
+a previously unknown category of inputs).  The parser implemented for this
+project is evaluated on the criteria of efficiency and coverage of Postfix
+log files, demonstrating that the conflicting goals of efficiency and
+accuracy can be balanced, and that one need not be sacrificed to achieve
+the other.
 
@@
 
-\end{abstract}
+\vfill{}
+\vfill{}
+\vfill{}
+
+~\empty{}
+
+\newpage{}
 

### 2009-04-29 doc/logparser-results.tex
Changes made to improve appearance after switching to one-and-a-half spacing.

@@
 altering its Postfix configuration to accommodate this project is not an
-option.  \Graphref{Mails received per day} and \tableref{Number of mails
-received per day: statistics} show that, as expected, it accepts far more
-mails on weekdays than at weekends.  Note that mail generated on the server
-(e.g.\ bounce notifications, mail re-injected for forwarding, mail sent
-from mailing lists) does not contribute to these figures: in particular,
-the vast amounts of mail resulting from the mail loops noticeable in later
-graphs are not reflected in these figures.
+option.  \Graphref{Mails received per day} shows the number of mails
+accepted over \acronym{SMTP}, with mean and standard deviation in
+\tableref{Number of mails received per day: statistics}; as expected, far
+more mails are accepted on weekdays than at weekends.  Note that mail
+generated on the server (e.g.\ bounce notifications, mail re-injected for
+forwarding, mail sent from mailing lists) does not contribute to these
+figures: in particular, the vast amounts of mail resulting from the mail
+loops noticeable in later graphs are not reflected in these figures.
 
@@
 parsed 10 times, and the mean parsing time calculated for each log file.
-The computer used for test runs was a Dell Optiplex 745, shown in
-\tableref{Details of the computer used to generate statistics}; it was
-dedicated to the task of gathering statistics from test runs, and did not
-run any other programs while test runs were in progress.  Saving results to
-the database was disabled for the test runs, because that dominates the run
-time of the parser, and the tests are aimed at measuring the speed of
-\parsername{} rather than the speed of the database and the disks it is
-stored on.  Parsing all \numberOFlogFILES{} log files in one run without
-saving results to the database took
+The computer used for test runs was a Dell Optiplex 745, with components
+shown in \tableref{Details of the computer used to generate statistics}; it
+was dedicated to the task of gathering statistics from test runs, and did
+not run any other programs while test runs were in progress.  Saving
+results to the database was disabled for the test runs, because that
+dominates the run time of the parser, and the tests are aimed at measuring
+the speed of \parsername{} rather than the speed of the database and the
+disks it is stored on.  Parsing all \numberOFlogFILES{} log files in one
+run without saving results to the database took
 \input{build/include-full-run-duration}, with mean throughput of

### 2009-08-24 doc/logparser-complications.tex
Changes made to improve appearance after switching to one-and-a-half spacing.
Changes to capitalisation of titles, as specified by Howard Hamilton. Fixed that/than type identified by Joe Carthy.

@@
-\section{Complications Encountered While Writing The \parsernamelong{}}
+\section{Complications Encountered During Development of the \parsernamelong{}}
 
@@
 
-\texttt{3FF7C4317: to=<username@example.com>, relay=local, \hfill{}
-\newline{} \tab{} \tab{} delay=0, status=sent (forwarded as 56F5B43FD)}
+\texttt{\tab{} 3FF7C4317: to=<username@example.com>, relay=local, \hfill{}
+\newline{} \tab{} \tab{} \tab{} delay=0, status=sent (forwarded as
+56F5B43FD)}
 
@@
 
-\subsection{Out Of Order Log Lines}
+\subsection{Out of Order Log Lines}
 
@@
 
-\subsection{Mail Deleted Before Delivery Is Attempted}
+\subsection{Mail Deleted Before Delivery is Attempted}
 
@@
 
-\subsection[Bounce Notification Mails Delivered Before Their Creation Is
-Logged]{Bounce Notification Mails Delivered Before \newline{} Their
+\subsection[Bounce Notification Mails Delivered Before their Creation is
+Logged]{Bounce Notification Mails Delivered Before \newline{} their
 Creation Is Logged}
@@
 
-\subsection{Mails Deleted During Delivery}
+\subsection{Mails Deleted from the Mail Queue During Delivery}
 

### 2009-08-24 doc/logparser-results.tex
Changes to capitalisation of titles, as specified by Howard Hamilton. Fixed that/than type identified by Joe Carthy.

@@
 
-\subsection{Characteristics Of The \numberOFlogFILES{} Log Files}
+\subsection{Characteristics of the \numberOFlogFILES{} Log Files}
 
@@
 
-\subsection{Rule Ordering For Efficiency}
+\subsection{Rule Ordering for Efficiency}
 
@@
 
-\subsection{Comparing Optimal Ordering Against An Oracle}
+\subsection{Comparing Optimal Ordering Against an Oracle}
 
@@
 
-\subsection{Scalability As The Ruleset Grows}
+\subsection{Scalability as the Ruleset Grows}
 
@@
 
-\subsection{Where Is Parsing Time Spent: Recognising Or Processing Log
+\subsection{Where is Parsing Time Spent: Recognising or Processing Log
 Lines?}
@@
 
-\subsection{Mail Delivery Attempts Correctly Understood And Reconstructed}
+\subsection{Mail Delivery Attempts Correctly Understood and Reconstructed}
 

### 2009-08-24 doc/logparser-introduction.tex
Provide a brief introduction to Postfix in the abstract and introduction, as instructed by Howard Hamilton.

@@
 The approach chosen to measure performance is to analyse the log files
-produced by the \acronym{MTA} in use, Postfix, rather than modifying its
-source code to generate statistics: this approach improves the chances of
-other Postfix users testing, using, and benefiting from the software
-developed for this project.  The need arose for a parser capable of dealing
-with the great number and variety of log lines produced by Postfix: the
-parser must be designed and implemented so that adding support for parsing
-new inputs is a simple task, because the log lines to be parsed will change
-over time.  This variety in log lines occurs for several reasons:
+produced by
+Postfix,\footnote{\urlLastChecked{http://www.postfix.org/}{2009/04/21}.  An
+introduction to Postfix is provided in \sectionref{postfix background}.} the
+\acronym{MTA} used by the School of Computer Science and Statistics, rather
+than modifying its source code to generate statistics: this approach
+improves the chances of other Postfix users testing, using, and benefiting
+from the software developed for this project.  The need arose for a parser
+capable of dealing with the great number and variety of log lines produced
+by Postfix: the parser must be designed and implemented so that adding
+support for parsing new inputs is a simple task, because the log lines to
+be parsed will change over time.  This variety in log lines occurs for
+several reasons:
 

### 2009-08-24 doc/logparser.tex
Changes made to improve appearance after switching to one-and-a-half spacing.
Provide a brief introduction to Postfix in the abstract and introduction, as instructed by Howard Hamilton.

@@
 \fi{}
+\newcommand{\mytitle}[0]{%
+    Parsing Log Files Produced by the Postfix Mail Transfer Agent%
+}
 \hypersetup{
-    pdftitle    = {Parsing Postfix Log Files},
+    pdftitle    = {\mytitle},
     pdfauthor   = {John Tobin},
@@
 
-    {\Huge Parsing Postfix Log Files}
+    {\Huge \mytitle{}}
 
@@
 
-Parsing Postfix log files is much more difficult than it first appears, but
+Postfix is a widely used Mail Transfer Agent, transferring hundreds of
+millions of mails between senders and recipients every day.  Parsing log
+files produced by Postfix is much more difficult than it first appears, but
 it \textit{is\/} possible to achieve a high degree of accuracy in
@@
 
-\newpage
+\newpage{}
 

### 2009-08-24 doc/logparser-background.tex
Changes to capitalisation of titles, as specified by Howard Hamilton. Fixed that/than type identified by Joe Carthy.
Improve the description of Postfix restrictions, as specified by Howard Hamilton.

@@
 anti-spam techniques.  Analysis of this kind is much easier to perform when
-the data source is a database that when trying to directly analyse log
+the data source is a database than when trying to directly analyse log
 files.
@@
 
-\section{Postfix}
+\section{The Postfix Mail Transfer Agent}
 
@@
 
-\subsection{Mixing And Matching Postfix Restrictions}
+\subsection{Mixing and Matching Postfix Restrictions}
 
@@
 stage; a benefit of this delay is that Postfix has more information
-available when logging rejections.
+available when logging rejections.  Each restriction is evaluated to
+produce a result of \textit{reject}, \textit{permit}, \textit{dunno}, or
+the name of another restriction to be evaluated; other less commonly used
+results are possible as described in~\cite{access-man-page}.  The meaning
+of \textit{permit\/} and \textit{reject\/} is obvious; \textit{dunno\/}
+means to stop evaluating the current restriction and continue processing
+the remainder of the restriction list, allowing exceptions to more general
+rules.  The administrator can define new restrictions as a list of existing
+restrictions, allowing arbitrarily long and complex user-defined sequences
+of lookups, restrictions, and exceptions.
 
@@
 
-Each restriction is evaluated to produce a result of \textit{reject},
-\textit{permit}, \textit{dunno}, or the name of another restriction to be
-evaluated; other less commonly used results are possible as described
-in~\cite{access-man-page}.  The meaning of \textit{permit\/} and
-\textit{reject\/} is obvious; \textit{dunno\/} means to stop evaluating the
-current restriction and continue processing the remainder of the
-restriction list, allowing exceptions to more general rules.  The
-configuration example below shows how to require that all machines on the
-local network except for the web server authenticate before sending mail;
-the web server is exempt because the legacy applications running on it lack
-authentication support.  The restriction list is evaluated from top to
-bottom: \texttt{permit\_sasl\_authenticated} permits authenticated clients;
-an exception is made for the web server in the
+The configuration example below shows how to require that all machines on
+the local network except for the web server authenticate before sending
+mail; the web server is exempt because the legacy applications running on
+it lack authentication support.  The restriction list is evaluated from top
+to bottom: \texttt{permit\_sasl\_authenticated} permits authenticated
+clients; an exception is made for the web server in the
 \texttt{check\_client\_access} check, and then other local machines are
@@
 
-The administrator can define new restrictions as a list of existing
-restrictions, allowing arbitrarily long and complex user-defined sequences
-of lookups, restrictions, and exceptions.  Postfix tries to protect the
-administrator from misconfiguration in as far as it reasonably can, e.g.\
-the restriction \texttt{check\_helo\_mx\_access} cannot cause a mail to be
-accepted, because the parameter it checks (the hostname given in the HELO
-command) is under the control of the remote client.  Despite this, it is
-possible for the administrator to make catastrophic mistakes, e.g.\
-rejecting all mail --- the administrator must be cognisant of the effects
-their configuration changes will have.  This is similar to one of UNIX's
-design philosophies: \textit{``UNIX was not designed to stop its users from
-doing stupid things, as that would also stop them from doing clever
+Postfix tries to protect the administrator from misconfiguration in as far
+as it reasonably can, e.g.\ the restriction
+\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
+because the parameter it checks (the hostname given in the HELO command) is
+under the control of the remote client.  Despite this, it is possible for
+the administrator to make catastrophic mistakes, e.g.\ rejecting all mail
+--- the administrator must be cognisant of the effects their configuration
+changes will have.  This is similar to one of UNIX's design philosophies:
+\textit{``UNIX was not designed to stop its users from doing stupid things,
+as that would also stop them from doing clever
 things''\/}~\cite{unix-philosophy}.
@@
 shown in \tableref{Example attributes sent to policy servers}) and returns
-a result (reject, permit, dunno, or a rejection name) as described in
+a result (reject, permit, dunno, or a restriction name) as described in
 \sectionref{Mixing and matching Postfix restrictions}.  A policy server can

### 2009-08-26 doc/logparser-other-parsers-review.tex
Changes made to improve appearance after switching to one-and-a-half spacing.
Changes to capitalisation of titles, as specified by Howard Hamilton. Fixed that/than type identified by Joe Carthy.
Corrections from Howard Hamilton: * Added a brief introduction to MySQL. * Removed the MySQL-BerkeleyDB comparison, because I didn't think it was   appropriate.  Howard's instruction was to move it into a separate paragraph,   but there wasn't enough content to justify keeping it.

@@
-\chapter{State Of The Art Review}
+\chapter{State of the Art Review}
 
@@
 
-    \item [MySQL] The schema used with the MySQL database is undocumented,
-        but at least it is possible to discover the schema if one has an
-        existing \acronym{SQL} database, unlike with Berkeley DB\@; all
-        \acronym{SQL} databases embed the schema into the database and
-        provide commands for displaying it.  Berkeley DB does not embed a
-        schema in its database files, because it does not need one, and
-        embedding one would not provide any benefit; it only provides
-        \textbf{(key, value)} pairs, so any additional structuring of the
-        data is imposed by the application, and it is the responsibility of
-        the author of the application to document this structure.
-        \acronym['s]{LMA} MySQL support was not tested because the schema
-        is not documented, and no existing database was available to
-        extract the schema from, so the required database could not be
-        created.
+    \item [MySQL] MySQL is a widely used, open source, relational database.
+        \acronym['s]{LMA} MySQL support was not tested because the database
+        schema used by \acronym{LMA} is not documented, so the required
+        database could not be created.
 
@@
 
-\section{Sawmill Universal Log File Analysis And Reporting}
+\section{Sawmill Universal Log File Analysis and Reporting}
 
@@
 
+% Added because extra spacing causes the new section to be badly
+% positioned.
+\newpage{}
+
 \section{Summary}

### 2009-08-26 doc/logparser-acronyms.tex
Changes made after final checks.
Added a brief explanation of regex after its first occurrence, and added it to the glossary, as required by Howard Hamilton.

@@
     first={Postfix Log Parser (PLP)}
+}
+
+\newglossaryentry{regex}{
+    type=\acronymtype,
+    name={regex},
+    description={Regular Expression},
+    first={Regular Expression (regex)}
+}
+\newcommand{\regexglossarydescription}[0]{
+    Regular Expressions are a compact, powerful method of specifying
+    patterns that describe a set of strings.  The regex \texttt{aa*b*b}
+    describes a set of strings, all of which start with \texttt{a},
+    followed by any number of \texttt{a}, then any number of \texttt{b},
+    and finish with \texttt{b}; the strings \texttt{ab},
+    \texttt{aaaaaaaab}, and \texttt{aaabbbbb} are members of that set,
+    whereas the strings \texttt{abba}, \texttt{abcd}, and \texttt{qwerty}
+    are not.  A string can be checked against a regex to determine if the
+    string is a member of the set of strings described by that regex.
+}
+\newglossaryentry{regex glossary}{
+    name={Regular Expression},
+    description={
+        \regexglossarydescription{}%
+    }
 }
@@
         \glsreset{RFC}The \acronym{RFC} series is a series of proposals
-        defining various protocols and file formats, e.g.\
-        \acronym{SMTP}\@.  The name is somewhat misleading nowadays:
-        initially the authors were asking for peer review, but these
-        documents are now the de facto standards the Internet runs on.
+        defining various protocols and formats, e.g.\ \acronym{SMTP}\@.
+        The name is somewhat misleading nowadays: initially the authors
+        were asking for peer review, but these documents are now the de
+        facto standards the Internet runs on.
     }

### 2009-08-26 doc/logparser-other-parsers-review.tex
Added a brief explanation of regex after its first occurrence, and added it to the glossary, as required by Howard Hamilton.

@@
 
-\acronym{LMA} requires major changes to the parser code to parse new log lines
-or to extract additional data.  The code is structured as a long series of
-blocks that each handle all log lines matching a single regex, so parsing
-new log lines requires modifying an existing regex or carefully inserting a
-new block in the correct place; extracting extra data would require
-modifying multiple blocks, regexes, or both.
+\acronym{LMA} requires major changes to the parser code to parse new log
+lines or to extract additional data.  The code is structured as a long
+series of blocks that each handle all log lines matching a single
+\acronym{regex}, so parsing new log lines requires modifying an existing
+regex or carefully inserting a new block in the correct place; extracting
+extra data would require modifying multiple blocks, regexes, or both.
+\regexglossarydescription{}
 

### 2009-08-27 doc/logparser-acronyms.tex
Turned regex into an acronym everywhere.

@@
     Regular Expressions are a compact, powerful method of specifying
-    patterns that describe a set of strings.  The regex \texttt{aa*b*b}
-    describes a set of strings, all of which start with \texttt{a},
-    followed by any number of \texttt{a}, then any number of \texttt{b},
-    and finish with \texttt{b}; the strings \texttt{ab},
+    patterns that describe a set of strings.  The \acronym{regex}
+    \texttt{aa*b*b} describes a set of strings, all of which start with
+    \texttt{a}, followed by any number of \texttt{a}, then any number of
+    \texttt{b}, and finish with \texttt{b}; the strings \texttt{ab},
     \texttt{aaaaaaaab}, and \texttt{aaabbbbb} are members of that set,
     whereas the strings \texttt{abba}, \texttt{abcd}, and \texttt{qwerty}
-    are not.  A string can be checked against a regex to determine if the
-    string is a member of the set of strings described by that regex.
+    are not.  A string can be checked against a \acronym{regex} to
+    determine if the string is a member of the set of strings described by
+    that \acronym{regex}.
 }
@@
     }
+}
+\newcommand{\regexes}[0]{
+    \acronym[es]{regex}%
 }

### 2009-08-27 doc/logparser-architecture.tex
Changes to capitalisation of titles, as specified by Howard Hamilton. Fixed that/than type identified by Joe Carthy.
Turned regex into an acronym everywhere.

@@
 parse new inputs without modifying the core parsing algorithm.  Adding a
-new rule with the action to invoke and a regex to recognise inputs is
-trivial in comparison to understanding an entire parser, identifying the
-correct location to change, and making the appropriate changes.  Changes to
-a parser must be made without adversely affecting existing parsing,
-including any edge cases that are not immediately obvious; an edge case
-that occurs only four times in \numberOFlogFILES{} log files is described
-in Yet More Aborted Delivery Attempts (\sectionref{yet more aborted
-delivery attempts}).  The more intrusive the changes are, the more likely
-they are to introduce a bug, so reducing the extent of the changes is
-important.  Requiring changes to a parser's source code also complicates
-upgrades of the parser, because the changes must be preserved during the
-upgrade, and they may clash with changes made by the developer.  This
-architecture allows the user to add new rules to a parser without having to
-edit it, unless the new inputs cannot be processed by the existing actions.
-If the new inputs do require new functionality, new actions can be added to
-the parser without having to modify existing actions; only when the new
-actions need to cooperate with existing actions will more extensive changes
-be required.
+new rule with the action to invoke and a \acronym{regex} to recognise
+inputs is trivial in comparison to understanding an entire parser,
+identifying the correct location to change, and making the appropriate
+changes.  Changes to a parser must be made without adversely affecting
+existing parsing, including any edge cases that are not immediately
+obvious; an edge case that occurs only four times in \numberOFlogFILES{}
+log files is described in Yet More Aborted Delivery Attempts
+(\sectionref{yet more aborted delivery attempts}).  The more intrusive the
+changes are, the more likely they are to introduce a bug, so reducing the
+extent of the changes is important.  Requiring changes to a parser's source
+code also complicates upgrades of the parser, because the changes must be
+preserved during the upgrade, and they may clash with changes made by the
+developer.  This architecture allows the user to add new rules to a parser
+without having to edit it, unless the new inputs cannot be processed by the
+existing actions.  If the new inputs do require new functionality, new
+actions can be added to the parser without having to modify existing
+actions; only when the new actions need to cooperate with existing actions
+will more extensive changes be required.
 
@@
 
-            \item Checking that the regex in the rule is valid.
+            \item Checking that the \acronym{regex} in the rule is valid.
 
@@
 variants should have multiple rules, one for each variant.  Rules will
-typically use a regex when recognising inputs, but other approaches may
-prove useful for some applications, e.g.\ comparing fixed strings to the
-input, or checking the length of the input; for the remainder of this
-thesis it will be assumed that a regex is used.  Each rule must specify the
-regex to recognise inputs and the action to invoke when recognition is
-successful, but implementations are free to add any other attributes they
-require; \sectionref{rules table} describes the attributes used in
-\parsername{}, and some generally useful attributes will be discussed later
-in this section.
+typically use a \acronym{regex} when recognising inputs, but other
+approaches may prove useful for some applications, e.g.\ comparing fixed
+strings to the input, or checking the length of the input; for the
+remainder of this thesis it will be assumed that a \acronym{regex} is used.
+Each rule must specify the \acronym{regex} to recognise inputs and the
+action to invoke when recognition is successful, but implementations are
+free to add any other attributes they require; \sectionref{rules table}
+describes the attributes used in \parsername{}, and some generally useful
+attributes will be discussed later in this section.
 
@@
 
-    \item Modify an existing rule's regex, because the new input is part of
-        an existing variant.
+    \item Modify an existing rule's \acronym{regex}, because the new input
+        is part of an existing variant.
 
-    \item Write a new rule that pairs an existing action with a new regex,
-        adding a new variant to an existing category.
+    \item Write a new rule that pairs an existing action with a new
+        \acronym{regex}, adding a new variant to an existing category.
 
@@
         inputs from the new category, and write a new rule pairing the new
-        action with a new regex.
+        action with a new \acronym{regex}.
 
@@
 
-\subsection{Attaching Conditions To Rules}
+\subsection{Attaching Conditions to Rules}
 
@@
 tokens can be on one line, or may have many lines between them.
-\Tableref{Rules to parse C-style comments} shows the regexes, conditions,
-and state changes of the four rules required to parse C-style comments.
-These are simplified rules, e.g.\ rules one and two will incorrectly
-recognise the comment start token if it is within a quoted string.  Rules
-one and two will be used when the parser is parsing code, not comments:
-rule one recognises a comment that is contained within one line and leaves
-the parser's state unchanged; rule two recognises the start of a comment
-and changes the parser's state to parsing comments instead of parsing code.
-Rules three and four will be used when the parser is parsing a comment:
-rule three recognises the end of a comment and switches the parser's state
-back to parsing code; rule four recognises a comment line without an end
-token and keeps the parser's state unchanged.  It is important that the
-rules are applied in the order listed in \tableref{Rules to parse C-style
-comments} because rule two overlaps with rule one, and rule four overlaps
-with rule three; \sectionref{rules in architecture} has explained how this
-is achieved.  \sectionref{overlapping rules in architecture} will discuss
-the benefits and difficulties of using overlapping rules.
+\Tableref{Rules to parse C-style comments} shows the \regexes{},
+conditions, and state changes of the four rules required to parse C-style
+comments.  These are simplified rules, e.g.\ rules one and two will
+incorrectly recognise the comment start token if it is within a quoted
+string.  Rules one and two will be used when the parser is parsing code,
+not comments: rule one recognises a comment that is contained within one
+line and leaves the parser's state unchanged; rule two recognises the start
+of a comment and changes the parser's state to parsing comments instead of
+parsing code.  Rules three and four will be used when the parser is parsing
+a comment: rule three recognises the end of a comment and switches the
+parser's state back to parsing code; rule four recognises a comment line
+without an end token and keeps the parser's state unchanged.  It is
+important that the rules are applied in the order listed in \tableref{Rules
+to parse C-style comments} because rule two overlaps with rule one, and
+rule four overlaps with rule three; \sectionref{rules in architecture} has
+explained how this is achieved.  \sectionref{overlapping rules in
+architecture} will discuss the benefits and difficulties of using
+overlapping rules.
 
@@
 Overlapping rules can be detected by visual inspection, or a program could
-be written to analyse the regexes in a ruleset.  Traditional regexes are
-equivalent in computational power to \acronym{FA} and can be converted to
-\acronym{FA}, so regex overlap can be detected by finding a non-empty
-intersection of two \acronym{FA}\@.  Perl 5.10 regexes~\cite{perlre} are
-more powerful than traditional regexes: they can match correctly balanced
-brackets nested to an arbitrary depth, e.g.\
-\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
+be written to analyse the \regexes{} in a ruleset.  Traditional \regexes{}
+are equivalent in computational power to \acronym{FA} and can be converted
+to \acronym{FA}, so \acronym{regex} overlap can be detected by finding a
+non-empty intersection of two \acronym{FA}\@.  Perl 5.10
+\regexes{}~\cite{perlre} are more powerful than traditional \regexes{}:
+they can match correctly balanced brackets nested to an arbitrary depth,
+e.g.\ \verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
 \verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
-regex engine to maintain state on a stack, so Perl 5.10 regexes are
-equivalent in computational power to \acronym{PDA}; detecting overlap may
-require calculating the intersection of two \acronym{PDA} instead of two
-\acronym{FA}.  \acronym{PDA} intersection is not closed, i.e.\ the result
-cannot always be implemented using a \acronym{PDA}, so intersection may be
-intractable sometimes, e.g.:
+\acronym{regex} engine to maintain state on a stack, so Perl 5.10
+\regexes{} are equivalent in computational power to \acronym{PDA};
+detecting overlap may require calculating the intersection of two
+\acronym{PDA} instead of two \acronym{FA}.  \acronym{PDA} intersection is
+not closed, i.e.\ the result cannot always be implemented using a
+\acronym{PDA}, so intersection may be intractable sometimes, e.g.:
 $a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
-Detecting overlap amongst $n$ regexes requires calculating
+Detecting overlap amongst $n$ \regexes{} requires calculating
 $\frac{n\left(n-1\right)}{2}$ intersections, resulting in
@@
 When detecting overlap, any conditions attached to rules must be taken into
-account, because two rules whose regexes overlap may have conditions
+account, because two rules whose \regexes{} overlap may have conditions
 attached that prevent the rules overlapping.  A less naive approach to
 overlap detection would first check for overlapping conditions amongst
-rules, and then check for overlap between the regexes of each pair of rules
-with overlapping conditions.  Rule overlap is not transitive, e.g.\ given
-these three conditions:
+rules, and then check for overlap between the \regexes{} of each pair of
+rules with overlapping conditions.  Rule overlap is not transitive, e.g.\
+given these three conditions:
 
@@
     $o$ & = & set of pairs of rules with overlapping conditions   \\
-    $x$ & = & cost of checking for overlap between two regexes    \\
+    $x$ & = & cost of checking for overlap between two \regexes{}    \\
 
@@
 lower than $x$.  If $y$ is higher than $x$ then the checks for overlap
-should be performed in the opposite order: check for regex overlap first,
-then check for condition overlap only between pairs of rules with
-overlapping regexes.
+should be performed in the opposite order: check for \acronym{regex}
+overlap first, then check for condition overlap only between pairs of rules
+with overlapping \regexes{}.
 
@@
 
-It is possible to write pathological regexes, which fall into two main
-categories: regexes that match inputs they should not, and regexes that
-consume excessive amounts of CPU time during matching.  Defining a regex
-that matches inputs it should not is trivial: \verb!/^/! matches the start
-of every input.  This regex would be found by a tool that detects
-overlapping rules, and would easily be noticed by visual inspection, but
-more complex regexes would be harder to find.  Regexes that match inputs
-more than they should are a problem not because of excessive resource
-usage, but because they may prevent the correct rule from recognising the
-input.  If an adaptive ordering system is used to prioritise rules that
-frequently recognise inputs (see \sectionref{rule ordering for
-efficiency}), then a rule with a regex that matches inputs it should not
-may be promoted up through the list, displacing an increasing number of
-correct rules as it rises.
+It is possible to write pathological \regexes{}, which fall into two main
+categories: \regexes{} that match inputs they should not, and \regexes{}
+that consume excessive amounts of CPU time during matching.  Defining a
+\acronym{regex} that matches inputs it should not is trivial: \verb!/^/!
+matches the start of every input.  This \acronym{regex} would be found by a
+tool that detects overlapping rules, and would easily be noticed by visual
+inspection, but more complex \regexes{} would be harder to find.  Regexes
+that match inputs more than they should are a problem not because of
+excessive resource usage, but because they may prevent the correct rule
+from recognising the input.  If an adaptive ordering system is used to
+prioritise rules that frequently recognise inputs (see \sectionref{rule
+ordering for efficiency}), then a rule with a \acronym{regex} that matches
+inputs it should not may be promoted up through the list, displacing an
+increasing number of correct rules as it rises.
 
-Excessive CPU time is usually consumed when a regex fails to match an
-input, and the regex engine backtracks many times because of alteration or
-nested quantifiers; successful matching is generally quite fast with such
-regexes, so problematic regexes may go unnoticed for a long time.  For
-example, with most regex engines matching double quoted strings with
-\verb!\"([^"\\]+|\\.)*"\!  is very fast when a match can be found, but when
-the match fails its computational complexity for a string of length $n$ is
-$O(2^{n})$; see~\cite{mastering-regular-expressions} for in-depth
-discussion of nested quantifiers, backtracking, alteration, and capturing
-groups.  Pathological regexes that consume excessive CPU time can be
-difficult to detect, whether by visual inspection or by machine inspection,
-but if a regex is converted to a \acronym{FA} or the internal
-representation used by the regex engine, it may be possible to determine if
-nested quantifiers or other troublesome constructs are present.  Modern
-regex engines have addressed many of these problems, e.g.\ the regex to
-match double quoted strings given above fails immediately with Perl 5.10,
-regardless of the input length, because the regex engine looks for both of
-the required double quotes first.  Similarly, Perl 5.10's regex engine
-optimises alterations starting with literal text into a trie, which has
-matching time proportional to the length of the alternatives, rather than
-the number of alternatives.  Perl regexes can use \verb!(?>pattern)!, which
-matches \verb!pattern! the first time the regex engine passes over it, and
-does not change what it originally matched if the regex engine backtracks
-over it, alleviating problems caused by excessive backtracking; Prolog
-users will notice a similarity to the \verb'!' (cut) operator.  A
-presentation showing some of Perl 5.10's new regex features is available at
+Excessive CPU time is usually consumed when a \acronym{regex} fails to
+match an input, and the \acronym{regex} engine backtracks many times
+because of alteration or nested quantifiers; successful matching is
+generally quite fast with such \regexes{}, so problematic \regexes{} may go
+unnoticed for a long time.  For example, with most \acronym{regex} engines
+matching double quoted strings with \verb!\"([^"\\]+|\\.)*"\!  is very fast
+when a match can be found, but when the match fails its computational
+complexity for a string of length $n$ is $O(2^{n})$;
+see~\cite{mastering-regular-expressions} for in-depth discussion of nested
+quantifiers, backtracking, alteration, and capturing groups.  Pathological
+\regexes{} that consume excessive CPU time can be difficult to detect,
+whether by visual inspection or by machine inspection, but if a
+\acronym{regex} is converted to a \acronym{FA} or the internal
+representation used by the \acronym{regex} engine, it may be possible to
+determine if nested quantifiers or other troublesome constructs are
+present.  Modern \acronym{regex} engines have addressed many of these
+problems, e.g.\ the \acronym{regex} to match double quoted strings given
+above fails immediately with Perl 5.10, regardless of the input length,
+because the \acronym{regex} engine looks for both of the required double
+quotes first.  Similarly, Perl 5.10's \acronym{regex} engine optimises
+alterations starting with literal text into a trie, which has matching time
+proportional to the length of the alternatives, rather than the number of
+alternatives.  Perl \regexes{} can use \verb!(?>pattern)!, which matches
+\verb!pattern! the first time the \acronym{regex} engine passes over it,
+and does not change what it originally matched if the \acronym{regex}
+engine backtracks over it, alleviating problems caused by excessive
+backtracking; Prolog users will notice a similarity to the \verb'!' (cut)
+operator.  A presentation showing some of Perl 5.10's new \acronym{regex}
+features is available at
 \urlLastChecked{http://www.regex-engineer.org/slides/perl510_regex.html}{2009/03/03}.
@@
 difficulty involved in detecting overlap is proportional to the complexity
-of a ruleset's regexes and conditions, and may be possible for a human yet
-intractable or impossible for a program.  The rules section concludes with
-a discussion of pathological rules, concentrating on pathological regexes.
+of a ruleset's \regexes{} and conditions, and may be possible for a human
+yet intractable or impossible for a program.  The rules section concludes
+with a discussion of pathological rules, concentrating on pathological
+\regexes{}.

### 2009-08-27 doc/logparser-conclusion.tex
Turned regex into an acronym everywhere.

@@
 (\sectionref{parser design}).  Providing a tool to ease the generation of
-regexes from unrecognised log lines (\sectionref{creating new rules in
+\regexes{} from unrecognised log lines (\sectionref{creating new rules in
 implementation}) should greatly help users who need to extend their ruleset

### 2009-08-27 doc/logparser-implementation.tex
Changes made to improve appearance after switching to one-and-a-half spacing.
Changes to capitalisation of titles, as specified by Howard Hamilton. Fixed that/than type identified by Joe Carthy.
Turned regex into an acronym everywhere.

@@
 to determine the value of each of the new rule's fields.  \parsername{}
-provides a utility to create regexes from unrecognised log lines; the
+provides a utility to create \regexes{} from unrecognised log lines; the
 algorithm it uses is documented, including the differences between it and
@@
 discussion of how \parsername{} uses rule conditions and overlapping rules,
-plus a description of the regex snippets provided to ease the process of
-writing regexes.
+plus a description of the \acronym{regex} snippets provided to ease the
+process of writing \regexes{}.
 
@@
 
-    \item This is followed by rejection of one or more mail delivery attempts
-        \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
-        \flowchart{CLONE}{5}; failure of the remote client's connection
-        \flowchart{DELIVERY\_ERROR}{6}; or some random interleaving of two
-        or more of the above.
+    \item This is followed by rejection of one or more mail delivery
+        attempts \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or
+        more mails \flowchart{CLONE}{5}; failure of the remote client's
+        connection \flowchart{DELIVERY\_ERROR}{6}; or some random
+        interleaving of two or more of the above.
 
@@
 recognise log lines, and the data saved from the parsing of log files.
-Each rule has a regex to recognise log lines and capture data from them,
-and specifies the action to be invoked when a log line is recognised; they
-also have several other fields used by the parser, and several fields that
-aid the user in understanding the meaning of the log lines recognised by
-each rule.  The rules are described in detail in \sectionref{rules in
-implementation}, but the rules table is documented in \sectionref{rules
-table} with the rest of the database schema.
+Each rule has a \acronym{regex} to recognise log lines and capture data
+from them, and specifies the action to be invoked when a log line is
+recognised; they also have several other fields used by the parser, and
+several fields that aid the user in understanding the meaning of the log
+lines recognised by each rule.  The rules are described in detail in
+\sectionref{rules in implementation}, but the rules table is documented in
+\sectionref{rules table} with the rest of the database schema.
 
@@
 
-\subsection{Using A Database To Provide An Application Programming Interface}
+\subsection{Using a Database to Provide an Application Programming
+Interface}
 
@@
     \item \acronym{SQL} is reasonably standard and many people will already
-        be familiar with it; for those unfamiliar with \acronym{SQL}, lots of
-        resources are available from which to learn, e.g.\
+        be familiar with it; for those unfamiliar with \acronym{SQL}, lots
+        of resources are available from which to learn, e.g.\
         \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
@@
 
-    \item [regex] The regex to recognise log lines with, as documented in
-        \sectionref{regex components}.
+    \item [regex] The \acronym{regex} to recognise log lines with, as
+        documented in \sectionref{regex components}.
 
@@
         each time the rule successfully recognises a log line.  At the
-        start of each run \parsername{} sorts the rules by hits, and at the end of the run it updates
-        every rule's hits field in the database.  Assuming that the
-        distribution of log lines is reasonably consistent across log
-        files, ordering rules by their recognition frequency will reduce
-        the parser's execution time.  Rule ordering for efficiency is
-        discussed in \sectionref{rule ordering for efficiency}.  This field
-        will be set by the parser rather than the rule author.
+        start of each run \parsername{} sorts the rules by hits, and at the
+        end of the run it updates every rule's hits field in the database.
+        Assuming that the distribution of log lines is reasonably
+        consistent across log files, ordering rules by their recognition
+        frequency will reduce the parser's execution time.  Rule ordering
+        for efficiency is discussed in \sectionref{rule ordering for
+        efficiency}.  This field will be set by the parser rather than the
+        rule author.
 
@@
 
-    \item That the regex is valid.  The regex will have regex components
-        expanded (see \sectionref{regex components}), and will also be
-        compiled for efficiency (\sectionref{Caching compiled regexes}).
-
-    \item For overlap between the data captured by the regex and additional
-        data specified in either \texttt{connection\_data} or
+    \item That the \acronym{regex} is valid.  The \acronym{regex} will have
+        \acronym{regex} components expanded (see \sectionref{regex
+        components}), and will also be compiled for efficiency
+        (\sectionref{Caching compiled regexes}).
+
+    \item For overlap between the data captured by the \acronym{regex} and
+        additional data specified in either \texttt{connection\_data} or
         \texttt{result\_data}.
@@
     \item That \texttt{connection\_data}, \texttt{result\_data}, and the
-        regex captures specify valid fields to save data to.
+        \acronym{regex} captures specify valid fields to save data to.
 
@@
 
-    \item Each regex is compiled and the result cached when the ruleset is
-        loaded; this optimisation can be disabled and the regex compiled
-        every time it is used when recognising log lines.  The effect this
-        optimisation has is addressed shown in \sectionref{Caching compiled
-        regexes}.
+    \item Each \acronym{regex} is compiled and the result cached when the
+        ruleset is loaded; this optimisation can be disabled and the
+        \acronym{regex} compiled every time it is used when recognising log
+        lines.  The effect this optimisation has is addressed shown in
+        \sectionref{Caching compiled regexes}.
 
@@
 
-    \item Every time a log line is recognised, the recognising rule's regex
-        and the log line can be printed, so that the user can verify that
-        the correct rule recognised each log line.  This is equivalent to
-        setting every rule's debug field to true.
+    \item Every time a log line is recognised, the recognising rule's
+        \acronym{regex} and the log line can be printed, so that the user
+        can verify that the correct rule recognised each log line.  This is
+        equivalent to setting every rule's debug field to true.
 
@@
 
-\subsection{Description Of Each Action}
+\subsection{Description of Each Action}
 
@@
 or mail, each action will save all the data captured by the recognising
-rule's regex to the connection or mail; usually, log lines that lack
-identifying information will be processed by the \action{UNINTERESTING}
-action.  Each action is passed the same arguments:
+rule's \acronym{regex} to the connection or mail; usually, log lines that
+lack identifying information will be processed by the
+\action{UNINTERESTING} action.  Each action is passed the same arguments:
 
@@
 
-    \item [data] The data captured from the log line by the rule's regex.
+    \item [data] The data captured from the log line by the rule's
+        \acronym{regex}.
 
@@
 \end{boldeqlist}
+
+\newpage{}
 
@@
         \sectionref{Postfix Daemons}.  This action saves all data captured
-        by the rule's regex if the log line has not come after a timeout
-        (see \sectionref{discarding cleanup log lines}); it also creates
-        the mail if necessary, setting its unknown origin flag (see
+        by the rule's \acronym{regex} if the log line has not come after a
+        timeout (see \sectionref{discarding cleanup log lines}); it also
+        creates the mail if necessary, setting its unknown origin flag (see
         \sectionref{pickup logging after cleanup}).
@@
         log line contains a queueid, save the data captured by the rule's
-        regex to the mail identified by that queueid; otherwise save it to
-        the connection identified by the \acronym{pid} in the log line.
+        \acronym{regex} to the mail identified by that queueid; otherwise
+        save it to the connection identified by the \acronym{pid} in the
+        log line.
 
@@
         the \action{SAVE\_DATA} action; it saves all data captured by the
-        recognising rule's regex, and does nothing more.  It is a separate
-        action to distinguish delivery attempts that bounce from other
-        delivery attempts.
+        recognising rule's \acronym{regex}, and does nothing more.  It is a
+        separate action to distinguish delivery attempts that bounce from
+        other delivery attempts.
 
@@
         \action{SAVE\_DATA} action; it saves all data captured by the
-        recognising rule's regex, and does nothing more.  It is a separate
-        action to distinguish successful delivery attempts from other
-        delivery attempts.
+        recognising rule's \acronym{regex}, and does nothing more.  It is a
+        separate action to distinguish successful delivery attempts from
+        other delivery attempts.
 
@@
         structure in the state tables saves any data captured by the
-        recognising rule's regex to it.  The \action{SAVE\_DATA} saves data
-        in this way but does not do anything else; it is invoked for log
-        lines that contain useful data but do not require any further
-        processing.
+        recognising rule's \acronym{regex} to it.  The \action{SAVE\_DATA}
+        saves data in this way but does not do anything else; it is invoked
+        for log lines that contain useful data but do not require any
+        further processing.
 
@@
 followed by the algorithm used by the utility supplied with \parsername{}
-that creates regexes from unparsed log lines; the regex components provided
-by \parsername{} to ease writing of complex regexes are covered in
-\sectionref{regex components}.  The rule conditions used in \parsername{}
-are the penultimate topic (\sectionref{rule conditions in implementation}),
-and this section concludes with some suggestions for how to detect
-overlapping rules.
+that creates \regexes{} from unparsed log lines; the \acronym{regex}
+components provided by \parsername{} to ease writing of complex \regexes{}
+are covered in \sectionref{regex components}.  The rule conditions used in
+\parsername{} are the penultimate topic (\sectionref{rule conditions in
+implementation}), and this section concludes with some suggestions for how
+to detect overlapping rules.
 
@@
         this rule will be used to recognise; see \sectionref{rule
-        conditions in implementation} for details.  The regex does the
-        actual recognition of log lines, and data captured by the regex
-        (e.g.\ sender, recipient) will be automatically saved to the
-        results and connections tables.  The regex components used in the
-        regex are described in \sectionref{regex components}.
+        conditions in implementation} for details.  The \acronym{regex}
+        does the actual recognition of log lines, and data captured by the
+        \acronym{regex} (e.g.\ sender, recipient) will be automatically
+        saved to the results and connections tables.  The \acronym{regex}
+        components used in the \acronym{regex} are described in
+        \sectionref{regex components}.
 
@@
 database, containing all the required attributes: action, name,
-description, program, and regex; all the other attributes are either
-optional, are set by the parser, or have sensible defaults.  The name and
-description fields should be set based on the meaning of the log line, to
-help others understand which log lines this rule will recognise; the value
-for the program field will be obvious from the unrecognised log lines.  The
-action depends on the what the log line represents, e.g.\ a delivery
-rejection, a mail being delivered, some useful information, or something
-else; examine the list of actions in \sectionref{actions in detail in
-implementation} to determine the correct one.  The regex needs to be based
-on the log line, but see below for a tool to ease the process of creating
-regexes from unrecognised log lines.
+description, program, and \acronym{regex}; all the other attributes are
+either optional, are set by the parser, or have sensible defaults.  The
+name and description fields should be set based on the meaning of the log
+line, to help others understand which log lines this rule will recognise;
+the value for the program field will be obvious from the unrecognised log
+lines.  The action depends on the what the log line represents, e.g.\ a
+delivery rejection, a mail being delivered, some useful information, or
+something else; examine the list of actions in \sectionref{actions in
+detail in implementation} to determine the correct one.  The
+\acronym{regex} needs to be based on the log line, but see below for a tool
+to ease the process of creating \regexes{} from unrecognised log lines.
 
@@
 pattern.  This classification can be used to group similar log lines and
-generate a regex to match each group of log lines.  The algorithm has five
-steps:
+generate a \acronym{regex} to match each group of log lines.  The algorithm
+has five steps:
 
@@
         purpose of this step is to utilise existing knowledge to create
-        more accurate regexes; it replaces many variable words with fixed
-        words, improving the subsequent classification of words as fixed or
-        variable.  Regex metacharacters in the log line will be escaped, to
-        avoid generating invalid or incorrect regexes.  The altered log
-        lines are written to a temporary file, which the next stage will
-        use instead of the original input file.
+        more accurate \regexes{}; it replaces many variable words with
+        fixed words, improving the subsequent classification of words as
+        fixed or variable.  Regex metacharacters in the log line will be
+        escaped, to avoid generating invalid or incorrect \regexes{}.  The
+        altered log lines are written to a temporary file, which the next
+        stage will use instead of the original input file.
 
@@
         will be misclassified as a fixed term, but that should be noticed
-        by the user when reviewing the new regexes, and will be obvious
+        by the user when reviewing the new \regexes{}, and will be obvious
         when the new rules do not recognise some log lines they are
@@
 
-    \item [Build regexes.]  The words are reassembled to produce a regex
-        matching the log line, and a counter is maintained for each regex.
-        Contiguous sequences of \texttt{.+} in the newly assembled regexes
-        are collapsed to a single \texttt{.+}; any resulting duplicate
-        regexes and their counters are combined.  If a regex's counter is
-        lower than the threshold supplied by the user the regex is
-        discarded; this second threshold is independent of the threshold
-        used to differentiate between fixed and variable words, but once
-        again 1\% of log lines is a good starting point.
+    \item [Build regexes.]  The words are reassembled to produce a
+        \acronym{regex} matching the log line, and a counter is maintained
+        for each \acronym{regex}.  Contiguous sequences of \texttt{.+} in
+        the newly assembled \regexes{} are collapsed to a single
+        \texttt{.+}; any resulting duplicate \regexes{} and their counters
+        are combined.  If a \acronym{regex}'s counter is lower than the
+        threshold supplied by the user the \acronym{regex} is discarded;
+        this second threshold is independent of the threshold used to
+        differentiate between fixed and variable words, but once again 1\%
+        of log lines is a good starting point.
 
     \item [Test the new regexes.]  The final step replaces keywords in the
-        new regexes in the same way as \parsername{} does, and compiles
-        each regex to check they are valid.  A match will be attempted
-        between all of the new regexes and each of the original unparsed
-        log lines (not the pre-processed log lines); the user will be
-        warned if a log line is not matched by any regex, or if a log line
-        is matched by more than one regex.  The number of log lines each
-        regex matches is counted, as is the number of log lines matched by
-        all regexes, though a log line is counted once only, even if
-        matched by more than one regex.
+        new \regexes{} in the same way as \parsername{} does, and compiles
+        each \acronym{regex} to check they are valid.  A match will be
+        attempted between all of the new \regexes{} and each of the
+        original unparsed log lines (not the pre-processed log lines); the
+        user will be warned if a log line is not matched by any
+        \acronym{regex}, or if a log line is matched by more than one
+        \acronym{regex}.  The number of log lines each \acronym{regex}
+        matches is counted, as is the number of log lines matched by all
+        \regexes{}, though a log line is counted once only, even if matched
+        by more than one \acronym{regex}.
 
@@
 
-The new regexes are displayed for the user to add to the ruleset, either as
-new rules or merged into the regexes of existing rules; also displayed are
-the number of unrecognised log lines each regex was expected to match, and
-the number it actually matched, to help the user notice problems in the new
-regexes.  Discarding regexes will result in some of the unrecognised log
-lines not being matched; when the ruleset has been augmented with the new
-regexes, the original log files should be parsed again, and any remaining
-unparsed log lines used as input to this utility.
-
-This utility is not expected to create perfect regexes, but it greatly
+The new \regexes{} are displayed for the user to add to the ruleset, either
+as new rules or merged into the \regexes{} of existing rules; also
+displayed are the number of unrecognised log lines each \acronym{regex} was
+expected to match, and the number it actually matched, to help the user
+notice problems in the new \regexes{}.  Discarding \regexes{} will result
+in some of the unrecognised log lines not being matched; when the ruleset
+has been augmented with the new \regexes{}, the original log files should
+be parsed again, and any remaining unparsed log lines used as input to this
+utility.
+
+This utility is not expected to create perfect \regexes{}, but it greatly
 reduces the effort required to deal with unrecognised log lines.  The
-regexes it generates will be self-contained: a parser that relies on using
-cascaded parsing would require a modified algorithm, perhaps by replacing
-the pre-processing stage with one that applies existing cascading rules to
-each log line, and uses the resulting modified log lines for the remainder
-of the algorithm.
+\regexes{} it generates will be self-contained: a parser that relies on
+using cascaded parsing would require a modified algorithm, perhaps by
+replacing the pre-processing stage with one that applies existing cascading
+rules to each log line, and uses the resulting modified log lines for the
+remainder of the algorithm.
 
@@
 
-Each rule's regex will have keywords expanded when the ruleset is loaded,
-for several reasons:
+Each rule's \acronym{regex} will have keywords expanded when the ruleset is
+loaded, for several reasons:
 
@@
 
-    \item It simplifies both reading and writing of regexes and helps to
-        make each regex largely self-documenting.  For example, the meaning
-        of \_\_CLIENT\_HOSTNAME\_\_ is immediately clear, whereas its
-        expansion \verb!(?:unknown|(?:[-.\w]+))! needs to be deciphered
-        each time it is encountered.
-
-    \item It avoids needless repetition of complex regex components, and
-        allows the components to be corrected or improved in one location.
-        For example, \_\_SENDER\_\_ is used in 68 rules; if a mistake is
-        discovered in it the mistake only needs to be corrected in one
-        place.
+    \item It simplifies both reading and writing of \regexes{} and helps to
+        make each \acronym{regex} largely self-documenting.  For example,
+        the meaning of \_\_CLIENT\_HOSTNAME\_\_ is immediately clear,
+        whereas its expansion \verb!(?:unknown|(?:[-.\w]+))! needs to be
+        deciphered each time it is encountered.
+
+    \item It avoids needless repetition of complex \acronym{regex}
+        components, and allows the components to be corrected or improved
+        in one location.  For example, \_\_SENDER\_\_ is used in 68 rules;
+        if a mistake is discovered in it the mistake only needs to be
+        corrected in one place.
 
     \item It enables automatic extraction and saving of captured data.  The
-        regex snippets use Perl 5.10's named capture buffers~\cite{perlre}
-        to capture data, so the mapping between captures and fields does
-        not need to be explicitly specified by the rule.
+        \acronym{regex} snippets use Perl 5.10's named capture
+        buffers~\cite{perlre} to capture data, so the mapping between
+        captures and fields does not need to be explicitly specified by the
+        rule.
 
@@
 
-To improve efficiency, the keywords are expanded and every rule's regex is
-compiled before attempting to parse the log file, otherwise every regex
-would be recompiled each time it was used, resulting in a large, data
-dependent slowdown, as described in \sectionref{Caching compiled regexes}.
-Most of the keywords are named after the fields in the connections or
-results tables they populate: \_\_CLIENT\_HOSTNAME\_\_, \_\_CLIENT\_IP\_\_,
-\_\_DELAY\_\_, \_\_DELAYS\_\_, \_\_ENHANCED\_STATUS\_CODE\_\_,
-\_\_HELO\_\_, \_\_MESSAGE\_ID\_\_, \newline{} \_\_QUEUEID\_\_,
-\_\_RECIPIENT\_\_, \_\_SENDER\_\_, \_\_SERVER\_HOSTNAME\_\_,
-\_\_SERVER\_IP\_\_, \_\_SIZE\_\_, and \_\_SMTP\_CODE\_\_.
+To improve efficiency, the keywords are expanded and every rule's
+\acronym{regex} is compiled before attempting to parse the log file,
+otherwise every \acronym{regex} would be recompiled each time it was used,
+resulting in a large, data dependent slowdown, as described in
+\sectionref{Caching compiled regexes}.  Most of the keywords are named
+after the fields in the connections or results tables they populate:
+\_\_CLIENT\_HOSTNAME\_\_, \_\_CLIENT\_IP\_\_, \_\_DELAY\_\_,
+\_\_DELAYS\_\_, \_\_ENHANCED\_STATUS\_CODE\_\_, \_\_HELO\_\_,
+\_\_MESSAGE\_ID\_\_, \newline{} \_\_QUEUEID\_\_, \_\_RECIPIENT\_\_,
+\_\_SENDER\_\_, \_\_SERVER\_HOSTNAME\_\_, \_\_SERVER\_IP\_\_, \_\_SIZE\_\_,
+and \_\_SMTP\_CODE\_\_.
 
@@
 
-Some similarity exists between regex components and cascaded parsing: each
-regex component resembles a rule that recognises part of a log line and
-consumes it, leaving the remainder to be recognised by other components or
-cascaded rules.  The major difference between the two is that regex
-components are explicitly used by the author of the ruleset, whereas
-cascaded parsing would be dynamically applied by the framework.
+Some similarity exists between \acronym{regex} components and cascaded
+parsing: each \acronym{regex} component resembles a rule that recognises
+part of a log line and consumes it, leaving the remainder to be recognised
+by other components or cascaded rules.  The major difference between the
+two is that \acronym{regex} components are explicitly used by the author of
+the ruleset, whereas cascaded parsing would be dynamically applied by the
+framework.
 
@@
 
-    \item Sort rules by program and regex e.g.\ with an \acronym{SQL} query
-        similar to:                                     \newline{}
+    \item Sort rules by program and \acronym{regex} e.g.\ with an
+        \acronym{SQL} query similar to:                 \newline{}
         \verb!SELECT program, regex!                    \newline{}
@@
         \verb!    ORDER BY program, regex;!             \newline{}
-        The rules are sorted first by program, then by regex, because rules
-        cannot overlap if their programs are different.  Note that this
-        query does not properly deal with generic rules whose program is
-        \texttt{*}; those rules will be used on all log lines that have not
-        been recognised by program-specific rules.
-
-        Inspect the list to see if any pair of regexes look suspiciously
-        similar; overlapping regexes will often lie beside one another when
-        sorted.
+        The rules are sorted first by program, then by \acronym{regex},
+        because rules cannot overlap if their programs are different.  Note
+        that this query does not properly deal with generic rules whose
+        program is \texttt{*}; those rules will be used on all log lines
+        that have not been recognised by program-specific rules.
+
+        Inspect the list to see if any pair of \regexes{} look suspiciously
+        similar; overlapping \regexes{} will often lie beside one another
+        when sorted.
 
@@
 
-\section{Limitations And Possible Improvements}
+% Move section start onto next page because extra spacing causes it to be
+% badly positioned.
+\newpage{}
+
+\section{Limitations and Possible Improvements}
 
@@
 
-    \item \parsername{} will not detect that it is parsing the same log file
-        twice, resulting in the database containing duplicate entries.
+    \item \parsername{} will not detect that it is parsing the same log
+        file twice, resulting in the database containing duplicate entries.
 
@@
 determine the values that should be used for each field; an explanation of
-the algorithm used by the utility that creates new regexes from unparsed
+the algorithm used by the utility that creates new \regexes{} from unparsed
 log lines, and how it differs from the original algorithm it is based on.
 How \parsername{} uses rule conditions and overlapping rules is discussed,
-accompanied by a description of the regex snippets the framework provides
-to simplify the regexes used in rules.  The many complications and
-difficulties encountered while writing \parsername{}, and the solutions
+accompanied by a description of the \acronym{regex} snippets the framework
+provides to simplify the \regexes{} used in rules.  The many complications
+and difficulties encountered while writing \parsername{}, and the solutions
 developed to overcome them, are documented in depth; also documented are

### 2009-08-27 doc/logparser-other-parsers-review.tex
Turned regex into an acronym everywhere.

@@
 \acronym{regex}, so parsing new log lines requires modifying an existing
-regex or carefully inserting a new block in the correct place; extracting
-extra data would require modifying multiple blocks, regexes, or both.
-\regexglossarydescription{}
+\acronym{regex} or carefully inserting a new block in the correct place;
+extracting extra data would require modifying multiple blocks, \regexes{},
+or both.  \regexglossarydescription{}
 
@@
 end times, timestamps for each log line, \acronym{SMTP} and enhanced status
-codes, delivery delays, and message-id.  \parsername{} does not make the full
-text of the log line available; a few minutes work could add this to
+codes, delivery delays, and message-id.  \parsername{} does not make the
+full text of the log line available; a few minutes work could add this to
 \parsername{} if desired, but it would greatly increase the size of the
@@
     track of in an active network than just webservers.  Lire is hands down
-    the most versatile log analysis software available today.  Lire not only
-    keeps you informed about your HTTP, FTP, and mail traffic, it also
+    the most versatile log analysis software available today.  Lire not
+    only keeps you informed about your HTTP, FTP, and mail traffic, it also
     reports on your firewalls, your print servers, and your DNS activity.
@@
     other software, in large part thanks to the numerous volunteers who
-    have pioneered many new services and features.  Lire is a total solution
-    for your log analysis needs.
+    have pioneered many new services and features.  Lire is a total
+    solution for your log analysis needs.
 }
@@
 Logrep), but not by \acronym{pid}.  The parsing is error prone: empty
-fields are saved when the log line does not match the regex, though it
-appears that they will not overwrite existing data.  Most notably, rejected
-delivery attempts are completely ignored, making it unsuitable for the
-purposes of this project.  It does not try to address any of the
-complications in \sectionref{complications} except for correlating log
-lines by queueid.
+fields are saved when the log line does not match the \acronym{regex},
+though it appears that they will not overwrite existing data.  Most
+notably, rejected delivery attempts are completely ignored, making it
+unsuitable for the purposes of this project.  It does not try to address
+any of the complications in \sectionref{complications} except for
+correlating log lines by queueid.
 

### 2009-08-27 doc/logparser-results.tex
Turned regex into an acronym everywhere.

@@
 the ruleset.  The penultimate topic is the simple optimisation of caching
-the results of compiling each rule's regex, and the huge effect it has on
-parser efficiency.  The performance evaluation concludes by examining where
-parsing time is spent: recognition of log lines or their subsequent
-processing by actions?
+the results of compiling each rule's \acronym{regex}, and the huge effect
+it has on parser efficiency.  The performance evaluation concludes by
+examining where parsing time is spent: recognition of log lines or their
+subsequent processing by actions?
 
@@
 
-Before the Perl interpreter attempts to match a regex against a piece of
-text, the regex is compiled into an internal representation and optimised
-to improve the speed of matching.  This compilation and optimisation takes
-CPU time: for most regexes it takes far more CPU time than the actual
-matching does.  If the interpreter is certain that a regex will not change
-while the program is running, it will automatically cache the results of
-compiling and optimising the regex for later use.  The results of compiling
-a dynamically generated regex can be cached and used in preference to the
-original regex, but it is the responsibility of the programmer to do so;
-\parsername{} does this with every rule's regex when the rules are loaded
-from the database.
+Before the Perl interpreter attempts to match a \acronym{regex} against a
+piece of text, the \acronym{regex} is compiled into an internal
+representation and optimised to improve the speed of matching.  This
+compilation and optimisation takes CPU time: for most \regexes{} it takes
+far more CPU time than the actual matching does.  If the interpreter is
+certain that a \acronym{regex} will not change while the program is
+running, it will automatically cache the results of compiling and
+optimising the \acronym{regex} for later use.  The results of compiling a
+dynamically generated \acronym{regex} can be cached and used in preference
+to the original \acronym{regex}, but it is the responsibility of the
+programmer to do so; \parsername{} does this with every rule's
+\acronym{regex} when the rules are loaded from the database.
 
 A test run using optimal ordering was performed as described in
-\sectionref{parser efficiency}, with one difference: regexes were not
+\sectionref{parser efficiency}, with one difference: \regexes{} were not
 compiled and cached when the ruleset was loaded from the database, so the
-Perl interpreter had to compile each regex each time it was used when
-trying to recognise a log line.  \Graphref{Percentage increase in parsing
-time when not caching compiled regexes graph} shows the effect that not
-caching compiled regexes has on parser performance, with mean and standard
-deviation in \tableref{Percentage increase in parsing time when not caching
-compiled regexes table}.  For typical log files, the mean increase in
-parsing time when not caching compiled regexes is
+Perl interpreter had to compile each \acronym{regex} each time it was used
+when trying to recognise a log line.  \Graphref{Percentage increase in
+parsing time when not caching compiled regexes graph} shows the effect that
+not caching compiled \regexes{} has on parser performance, with mean and
+standard deviation in \tableref{Percentage increase in parsing time when
+not caching compiled regexes table}.  For typical log files, the mean
+increase in parsing time when not caching compiled \regexes{} is
 \input{build/include-cached-regexes-vs-discarded-regexes-mean}; looking at
-it from the opposite direction, caching compiled regexes reduces parsing
+it from the opposite direction, caching compiled \regexes{} reduces parsing
 time by
 \input{build/include-cached-regexes-vs-discarded-regexes-mean-reduction}.
-Caching compiled regexes is probably the single most effective optimisation
-possible in \parsername{}, and was quite simple to implement: the framework
-compiles each rule's regex when the ruleset is loaded, and uses the
-compiled regex instead of the source regex when recognising log lines.  As
-seen previously, log files 22 \& 62--68 do not suffer such a large increase
-in parsing time when the optimisation is disabled; this is because, on
-average, fewer rules are used per log line, so fewer regexes are compiled
-per log line for those log files.
+Caching compiled \regexes{} is probably the single most effective
+optimisation possible in \parsername{}, and was quite simple to implement:
+the framework compiles each rule's \acronym{regex} when the ruleset is
+loaded, and uses the compiled \acronym{regex} instead of the source
+\acronym{regex} when recognising log lines.  As seen previously, log files
+22 \& 62--68 do not suffer such a large increase in parsing time when the
+optimisation is disabled; this is because, on average, fewer rules are used
+per log line, so fewer \regexes{} are compiled per log line for those log
+files.
 
@@
 efficient than usual, the mean number of rules used for each log line is
-higher than usual, and more regexes will need to be compiled when
+higher than usual, and more \regexes{} will need to be compiled when
 recognising each log line.  Parsing time is plotted against log file size
-for caching and not caching compiled regexes in \graphref{Parsing time
+for caching and not caching compiled \regexes{} in \graphref{Parsing time
 plotted against log file size when caching and not caching compiled
-regexes}; the plot for not caching compiled regexes is quite uneven
-compared to caching compiled regexes --- some log files are particularly
+regexes}; the plot for not caching compiled \regexes{} is quite uneven
+compared to caching compiled \regexes{} --- some log files are particularly
 slow, whereas others are not as badly affected, but these differences have
@@
         long recognition of log lines takes, a full test run using optimal
-        ordering and caching compiled regexes was performed as previously
-        described in \sectionref{parser efficiency}, but with one
-        difference: actions were not invoked when a log line was
+        ordering and caching compiled \regexes{} was performed as
+        previously described in \sectionref{parser efficiency}, but with
+        one difference: actions were not invoked when a log line was
         recognised, so the parsing time did not include the time normally
@@
         increase as optimally ordering rules.  Similarly, caching compiled
-        regexes provides an
+        \regexes{} provides an
         \input{build/include-cached-regexes-vs-discarded-regexes-mean-reduction}
@@
 achieved.  Correctly understanding and reconstructing every mail delivery
-attempt is important so that the information in the database is accurate and
-complete.  Improving the proportion of log lines correctly recognised is
-the less difficult of the two, because it just requires new rules to be
+attempt is important so that the information in the database is accurate
+and complete.  Improving the proportion of log lines correctly recognised
+is the less difficult of the two, because it just requires new rules to be
 written or existing rules to be extended.  Improving the proportion of
@@
 log line is not recognised the parser will issue a warning, to inform the
-user that they need to extend their ruleset.  Each rule's regex should be
-as specific and precise as possible, to ensure accurate parsing: a rule
-with a regex that matches zero or more of any character will recognise
-every log line, but not in a meaningful way; most log lines contain fixed
-strings, so this is not a problem in practice.
+user that they need to extend their ruleset.  Each rule's \acronym{regex}
+should be as specific and precise as possible, to ensure accurate parsing:
+a rule with a \acronym{regex} that matches zero or more of any character
+will recognise every log line, but not in a meaningful way; most log lines
+contain fixed strings, so this is not a problem in practice.
 
@@
 has been automated to ensure continued accuracy.  Based on these results,
-and how precise each rule's regex is, the author is confident that zero
-false positives occur when parsing the \numberOFlogFILES{} log files.
+and how precise each rule's \acronym{regex} is, the author is confident
+that zero false positives occur when parsing the \numberOFlogFILES{} log
+files.
 
@@
 increases is addressed, followed by a description and analysis of the
-simplest and most effective of the optimisations, caching compiled regexes,
-and the efficiency evaluation concludes with an examination of where
-parsing time is spent: recognising log lines or processing them?
+simplest and most effective of the optimisations, caching compiled
+\regexes{}, and the efficiency evaluation concludes with an examination of
+where parsing time is spent: recognising log lines or processing them?
 

### 2009-08-27 doc/logparser-architecture.tex
Correction from Howard Hamilton: emphasise the distinction between architecture and implementation at the start of chapter 4.

@@
 
-To avoid cluttering the description of the parser architecture with the
-details of implementing a parser for Postfix log files, each topic has been
-given its own chapter.  This chapter presents the architecture developed
-for this project, beginning with the overall architecture and design,
-followed by detailed documentation of the three components of the
-architecture: Framework, Actions, and Rules.  This chapter is focused on
-the theoretical, implementation-independent aspects of the architecture;
+The parser architecture described in this chapter is flexible enough to be
+used as the basis of other parsers.  Obviously it is particularly suitable
+for writing parsers for log files; with judicious use of cascaded parsing
+(see \sectionref{actions in architecture}), a calculator could easily be
+implemented; a more ambitious project might attempt to parse a programming
+language.\footnote{Lisp, where every statement is enclosed in parentheses
+and can easily be isolated from the surrounding statements, might be
+particularly amenable.}  To avoid cluttering the description of the
+architecture with the details of implementing a parser for Postfix log
+files, each topic has been given its own chapter.  This chapter is focused
+on the theoretical, implementation-independent aspects of the architecture;
 the practical difficulties of writing a parser for Postfix log files are
-covered in detail in \sectionref{Postfix Parser Implementation}.
+covered in detail in \sectionref{Postfix Parser Implementation}.  This
+chapter presents the architecture developed for this project, beginning
+with the overall architecture and design, followed by detailed
+documentation of the three components of the architecture: Framework,
+Actions, and Rules.
 

### 2009-08-27 doc/logparser-implementation.tex
Improvements to the flow chart, required by Howard Hamilton.

@@
 Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection
-because of forwarding, or generated internally by Postfix.  Local
-submission is the simplest of the four: a queueid is assigned immediately
-\flowchart{PICKUP}{2}, and the mail moves on to the delivery stage.
-Re-injection because of forwarding lacks explicit log lines of its own; it
-is explained fully in \sectionref{Re-injected mails}.  Internally generated
-mails lack any explicit origin in Postfix 2.2.x and must be detected using
-heuristics as described in \sectionref{identifying bounce notifications};
-later versions of Postfix do provide log lines for internally generated
-mails \flowchart{BOUNCE\_CREATED}{3}.  Bounce notifications are the primary
-example of internally generated mails, though other types exist, e.g.\
-Postfix may generate mails to the administrator when it encounters
-configuration errors.
+submission via \daemon{postdrop} or \daemon{sendmail}, by \acronym{SMTP},
+by re-injection because of forwarding, or generated internally by Postfix.
+Local submission is the simplest of the four: a queueid is assigned
+immediately \flowchart{PICKUP}{2}, and the mail moves on to the delivery
+stage.  Re-injection because of forwarding lacks explicit log lines of its
+own; it is explained fully in \sectionref{Re-injected mails}.  Internally
+generated mails lack any explicit origin in Postfix 2.2.x and must be
+detected using heuristics as described in \sectionref{identifying bounce
+notifications}; later versions of Postfix do provide log lines for
+internally generated mails \flowchart{BOUNCE\_CREATED}{3}.  Bounce
+notifications are the primary example of internally generated mails, though
+other types exist, e.g.\ Postfix may generate mails to the administrator
+when it encounters configuration errors.
 
@@
         if there were no rejections there will not be any data to save.
-        Any mails accepted will already have been cloned so their data is
-        in another data structure, and will be delivered in the same way as
-        mails that entered the system by any other route.
+        Any accepted mails will already have a separate data structure, and
+        will be delivered in the same way as mails that entered the system
+        by any other route.
 

### 2009-08-27 doc/logparser-results.tex
Explain where the 93 log files come from; required by Howard Hamilton.

@@
 
-The \numberOFlogFILES{} log files used when generating statistics have
-fairly consistent sizes and contents, except for two groups of log files:
-22 \& 62--68.  Median log file size is
-\input{build/include-median-file-size}, containing
+The \numberOFlogFILES{} log files used in this chapter were generated on
+the School of Computer Science and Statistics' mail server from 2007/01/26
+to 2007/04/28.  The \numberOFlogFILES{} log files have fairly consistent
+sizes and contents, except for two groups of log files: 22 \& 62--68.
+Median log file size is \input{build/include-median-file-size}, containing
 \input{build/include-median-number-of-log-lines} log lines, with a median

### 2009-08-28 doc/logparser-results.tex
Required by Howard Hamilton: * Changed "most optimal" to "optimal". * Reviewed use of the word "optimal".

@@
 
-    \item [Optimal]  The most optimal order, according to the hypothesis:
-        rules that recognise log lines most often will be used first.
+    \item [Optimal]  The optimal order, according to the hypothesis: rules
+        that recognise log lines most frequently will be used first.
 
-    \item [Shuffle] This ordering is intended to represent a randomly
+    \item [Shuffled] This ordering is intended to represent a randomly
         ordered ruleset.  The rules will be shuffled once before use and

### 2009-09-15 doc/logparser-implementation.tex
An improvement to the flow chart and accompanying text, showing which data is copied to the new data structure when a mail is accepted.

@@
 Everything starts off with a mail entering the system, whether by local
-submission via \daemon{postdrop} or \daemon{sendmail}, by \acronym{SMTP},
+submission via \daemon{sendmail} or \daemon{postdrop}, by \acronym{SMTP},
 by re-injection because of forwarding, or generated internally by Postfix.
